### 2022-02-27 chapters/conclusion.tex
Initial Overleaf Import

%!TEX root = ../Thesis.tex
\chapter{Conclusion}
A conclusion with some citations \cite{kingma_auto-encoding_2014,rezende_stochastic_2014}.

### 2022-02-27 chapters/dummy_introduction.tex
Initial Overleaf Import

%!TEX root = ../Thesis.tex
\chapter{Introduction}

\begin{itemize}
    \item \textup{Upright shape}
    \item \textit{Italic shape}
    \item \textsl{Slanted shape}
    \item \textsc{Small Caps shape}
    \item \textmd{Medium series}
    \item \textbf{Bold sereies}
    \item \textrm{Roman family}
    \item \textsf{Sans serif family}
    \item \texttt{Typewriter family}
\end{itemize}

I love to write special characters like øæå indside my \TeX{} document. Also á, à, ü, û, ë, ê, î, ï could be nice. So waht about the ``π'' chracter. What about ° é ® † ¥ ü | œ π ‘ @ ö ä ¬ ∆ ‹ « © ƒ ∂ ß ª Ω … ç √ ∫ ñ µ ‚ · ¡ “ £ ∞ ™ [ ] ≠ ± '.

Some dashes - – —, and the latex form - -- ---
\begin{equation*}
    x = \mathtt{x}, \mathbf{x}, \mathit{x}, x_{1_{2_{3_{4}}}}^{1^{2^{3^{4}}}} \cdot hello * \text{hello world} ⋅ \text{my world} · \text{third world} ⊗ t
\end{equation*}

Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum \cite{adams1980hitchhiker}.

Mauris id quam non magna fermentum malesuada id mattis lorem. In a dapibus neque. Etiam lacus dui, malesuada ac eleifend imperdiet, imperdiet ut ipsum. Vestibulum id ultricies est. Phasellus augue mauris, semper a luctus vel, faucibus in risus. Fusce commodo augue quis elit sagittis non viverra turpis bibendum. Nunc placerat sem non sapien malesuada malesuada ullamcorper orci luctus \cite{adams1980hitchhiker}. Morbi pharetra ligula integer mollis mi nec neque ultrices vitae volutpat leo ullamcorper. In at tellus magna. Curabitur quis posuere purus. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Suspendisse tristique placerat feugiat. Aliquam vitae est at enim auctor ultrices eleifend a urna. Donec non tincidunt felis. Maecenas at suscipit orci. See \cref{myFigure}.
\begin{figure}
    \centering
        \missingfigure[figwidth=6cm]{}
    \caption[Short caption to special figure]{This is my special figure. Aliquam ultricies, arcu quis tempor rhoncus, tellus nisl tempus justo, condimentum tempor erat odio ac purus. Integer quis ipsum felis. Aliquam volutpat, leo ac consequat egestas, lectus lacus adipiscing quam, id iaculis dolor quam in erat. Phasellus tempor interdum arcu quis vestibulum}
    \label{myFigure}
\end{figure}

Fusce id suscipit sem. Aliquam venenatis nibh nec nisl luctus vel consectetur neque dapibus. Nulla feugiat egestas turpis, ac viverra eros cursus sit amet. Cras tincidunt felis vel tellus ultrices condimentum. Quisque vehicula, arcu vitae interdum dignissim, purus tortor cursus libero, sit amet accumsan quam magna in neque. Phasellus luctus leo odio. Aliquam ultricies, arcu quis tempor rhoncus, tellus nisl tempus justo, condimentum tempor erat odio ac purus. Integer quis ipsum felis. Aliquam volutpat, leo ac consequat egestas, lectus lacus adipiscing quam, id iaculis dolor quam in erat. Phasellus tempor interdum arcu quis vestibulum. Pellentesque sit amet augue purus. See \cref{myTable}.

\begin{table}
    \centering
        \begin{tabular}{c | r l}
            h & h & h \\
            e & e & e \\
        \end{tabular}
    \caption{This is a caption to the table}
    \label{myTable}
\end{table}

\section{Torquent Arcu}
Curabitur condimentum suscipit arcu, sit amet convallis urna pellentesque ac. Quisque fringilla tincidunt risus nec accumsan. Curabitur vel sagittis ante. Integer eget placerat leo. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos. Vestibulum quis risus in nulla fermentum pellentesque dictum et erat. Nulla vel pretium nunc. Integer tortor lorem, suscipit sit amet ultricies non, porta at metus. Sed pharetra, ante facilisis interdum porta, mi dolor fringilla quam, ac porttitor urna dolor quis massa. Proin viverra semper tincidunt. Vivamus pulvinar pharetra condimentum. Pellentesque rutrum mollis tellus ac scelerisque.

\begin{figure}
    \centering
        \subbottom[1 pass]{
            \missingfigure[figwidth=6cm]{}
            \label{twoFigures1}
        }
        ~
        \subbottom[5 passes]{
            \missingfigure[figwidth=6cm]{}
            \label{twoFigures2}
        }
        \caption{loop performance comparison}
    \label{twoFigures}
\end{figure}

\subsection{Vestibulum}
Mauris luctus sollicitudin vestibulum. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos \cref{twoFigures2}. Duis eu nisl nec turpis porttitor bibendum eget sed orci. Aliquam consequat lorem a dui viverra porta facilisis augue rutrum. Cras luctus tellus in lectus egestas eu consequat magna cursus. Aenean aliquam neque a nibh elementum ornare. Integer eleifend imperdiet commodo. Morbi auctor, dui vel laoreet congue, purus est accumsan augue, sit amet feugiat neque nisl vel lorem. Curabitur ante sem, lacinia id adipiscing quis, viverra tristique nulla. Pellentesque ullamcorper pellentesque metus varius facilisis. Cras ac dui id odio tempor scelerisque. Curabitur a egestas risus. Pellentesque quis velit in sapien accumsan auctor. Phasellus aliquam, sapien eget lobortis volutpat, libero metus porttitor nisl, sed hendrerit urna dolor nec mi. See \cref{fibonacci}.

\begin{adjustwidth*}{0cm}{-0.4cm}
\begin{lstlisting}[language=Python,caption=Fibonacci,label=fibonacci]
# This is a comment
import easy
str = "I am a string"
str2 = "Now i have an awsome string with ´ '' `` which are not TeX'ed"
str3 = "What about awsome unicode characters? Like “, π, ”, Ω, ç. \" This"
def fib(n):
    if n == 0:
        return 0
    elif n == 1:
        return 1
    else:
        return fib(n-1) + fib(n-2)
str4 = "Yes it is possible with 80 charactes. Which this string proves. Wiiii."
str5 = "It adjusts according to the spine"
\end{lstlisting}
\end{adjustwidth*}

\section{Luctus}
Praesent et pellentesque arcu. Phasellus venenatis mi eu lorem convallis et iaculis ante aliquet. Aenean rhoncus placerat metus, vel convallis leo suscipit eu. Integer dapibus venenatis commodo. Cras laoreet faucibus sem nec luctus. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos. Cras consectetur lacinia dolor at gravida. Phasellus ipsum arcu, vulputate fermentum ultricies eget, tempor eu odio. Aenean accumsan vestibulum risus a mattis. See it on \cref{modifiedminibatch}.

\begin{algorithm}
\caption{Modified mini-batch $K$-means} \label{modifiedminibatch}
\begin{algorithmic}[1]
\State Given: $K$, mini-batch size $B$, iterations $T$, dataset $X$, correlation~matrix~$\mathrm{P}$.
\State Initialize $C = \{\mathbf{c}^{(1)}, \mathbf{c}^{(2)}, \ldots, \mathbf{c}^{(K)}\}$ with random $\mathbf{x}$'es picked from $X$.
\State $A \gets B \cdot T$ sorted random indexes to $X$, denoted $a_1, a_2, \ldots, a_{B\cdot T}$.
\State $X' \gets \{\mathbf{x}^{(a_1)}, \mathbf{x}^{(a_2)}, \ldots, \mathbf{x}^{(a_{B\cdot T})}\}$ \Comment{Cache all points}
\State $\textbf{size} \gets 0$
\For {$i = 1$ to $T$}
    \State $M \gets B$ examples picked randomly from $X'$

    \For{$\mathbf{x} \in M$} \Comment{\textit{Assignment step}}
        \State $\textbf{d}[\textbf{x}] \gets f(C,\mathbf{x}, \mathrm{P})$ \Comment{Cache closest center}
    \EndFor

    \For {$\mathbf{x} \in M$} \Comment{\textit{Update step}}
        \State $\textbf{c} \gets \textbf{d[x]}$ \Comment{Get cached center for current \textbf{x}}
        \State $\textbf{size}[\textbf{c}] \gets \textbf{size}[\textbf{c}] + 1$ \Comment{Update cluster size}
        \State $\eta \gets \frac{1}{\textbf{size}[\textbf{c}]}$       \Comment{Get learning rate}
        \State $\textbf{c} \gets (1 - \eta)\textbf{c}+\eta\textbf{x}$ \Comment{Take gradient step}
    \EndFor
\EndFor
\State \Return {$C$, \textbf{size}}
\end{algorithmic}
\end{algorithm}

Fusce id suscipit sem. Aliquam venenatis nibh nec nisl luctus vel consectetur neque dapibus. Nulla feugiat egestas turpis, ac viverra eros cursus sit amet. Cras tincidunt felis vel tellus ultrices condimentum. Quisque vehicula, arcu vitae interdum dignissim, purus tortor cursus libero, sit amet accumsan quam magna in neque. Phasellus luctus leo odio. Aliquam ultricies, arcu quis tempor rhoncus, tellus nisl tempus justo, condimentum tempor erat odio ac purus. Integer quis ipsum felis. Aliquam volutpat, leo ac consequat egestas, lectus lacus adipiscing quam, id iaculis dolor quam in erat. Phasellus tempor interdum arcu quis vestibulum. Pellentesque sit amet augue purus. 
\begin{figure}[htbp]
    \centering
        \missingfigure[figwidth=6cm]{}
    \caption{This is the caption I wrote}
    \label{fig:label}
\end{figure}
Curabitur condimentum suscipit arcu, sit amet convallis urna pellentesque ac. Quisque fringilla tincidunt risus nec accumsan. Curabitur vel sagittis ante. Integer eget placerat leo. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos. Vestibulum quis risus in nulla fermentum pellentesque dictum et erat. Nulla vel pretium nunc. Integer tortor lorem, suscipit sit amet ultricies non, porta at metus. Sed pharetra, ante facilisis interdum porta, mi dolor fringilla quam, ac porttitor urna dolor quis massa. Proin viverra semper tincidunt. Vivamus pulvinar pharetra condimentum. Pellentesque rutrum mollis tellus ac scelerisque.

\section{Sollicitudin vestibulum}
Mauris luctus sollicitudin vestibulum. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos. Duis eu nisl nec turpis porttitor bibendum eget sed orci. Aliquam consequat lorem a dui viverra porta facilisis augue rutrum. Cras luctus tellus in lectus egestas eu consequat magna cursus. Aenean aliquam neque a nibh elementum ornare. Integer eleifend imperdiet commodo. Morbi auctor, dui vel laoreet congue, purus est accumsan augue, sit amet feugiat neque nisl vel lorem. Curabitur ante sem, lacinia id adipiscing quis, viverra tristique nulla. Pellentesque ullamcorper pellentesque metus varius facilisis. Cras ac dui id odio tempor scelerisque. Curabitur a egestas risus. Pellentesque quis velit in sapien accumsan auctor. Phasellus aliquam, sapien eget lobortis volutpat, libero metus porttitor nisl, sed hendrerit urna dolor nec mi.

Praesent et pellentesque arcu. Phasellus venenatis mi eu lorem convallis et iaculis ante aliquet. Aenean rhoncus placerat metus, vel convallis leo suscipit eu. Integer dapibus venenatis commodo. Cras laoreet faucibus sem nec luctus. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos. Cras consectetur lacinia dolor at gravida. Phasellus ipsum arcu, vulputate fermentum ultricies eget, tempor eu odio. Aenean accumsan vestibulum risus a mattis.

\begin{adjustwidth*}{0cm}{-0.4cm}
\begin{lstlisting}[language=Python,caption=Fibonacci2,label=Fibonacci2]
# This is a comment
import easy
str = "I am a string"
str2 = "Now i have an awsome string with ´ '' `` which are not TeX'ed"
str3 = "What about awsome unicode characters? Like “, π, ”, Ω, ç. \" This"
def fib(n):
    if n == 0:
        return 0
    elif n == 1:
        return 1
    else:
        return fib(n-1) + fib(n-2)
str4 = "Yes it is possible with 80 charactes. Which this string proves. Wiiii."
str5 = "It adjusts according to the spine"
\end{lstlisting}
\end{adjustwidth*}

### 2022-02-27 chapters/layouts.tex
Initial Overleaf Import

\chapter{Layouts}

\begin{figure}
\listdiagram
\caption{The \LaTeX{} parameters which define typesetting and layout of lists.} 
\label{fig:layout-list}
\end{figure}

% \begin{figure}
% \oddpagelayoutfalse
% \pagediagram
% \caption{Left-hand page layout parameters}
% \label{fig:layout-left-page}
% \end{figure}

% \begin{figure}
% \oddpagelayouttrue
% \pagediagram
% \caption{Right-hand page layout parameters}
% \label{fig:layout-right-page}
% \end{figure}

\begin{figure}
\oddpagelayoutfalse
\twocolumnlayoutfalse
\stockdiagram
\caption{Left-hand page major layout parameters for \file{memoir} class}
\label{fig:layout-left-memoir}
\end{figure}

\begin{figure}
\oddpagelayouttrue
\twocolumnlayoutfalse
\stockdiagram
\caption{Right-hand page major layout parameters for \file{memoir} class}
\label{fig:layout-right-memoir}
\end{figure}

### 2022-02-27 frontmatter/abstract-da.tex
Initial Overleaf Import

--- 2022-02-27 chapters/dummy_introduction.tex
@@
 %!TEX root = ../Thesis.tex
-\chapter{Introduction}
-
-\begin{itemize}
-    \item \textup{Upright shape}
-    \item \textit{Italic shape}
-    \item \textsl{Slanted shape}
-    \item \textsc{Small Caps shape}
-    \item \textmd{Medium series}
-    \item \textbf{Bold sereies}
-    \item \textrm{Roman family}
-    \item \textsf{Sans serif family}
-    \item \texttt{Typewriter family}
-\end{itemize}
-
-I love to write special characters like øæå indside my \TeX{} document.
-Also á, à, ü, û, ë, ê, î, ï could be nice.
-So waht about the ``π'' chracter.
-What about ° é ® † ¥ ü | œ π ‘ @ ö ä ¬ ∆ ‹ « © ƒ ∂ ß ª Ω … ç √ ∫ ñ µ ‚ · ¡ “ £ ∞ ™ [ ] ≠ ± '.
-
-Some dashes - – —, and the latex form - -- ---
-\begin{equation*}
-    x = \mathtt{x}, \mathbf{x}, \mathit{x}, x_{1_{2_{3_{4}}}}^{1^{2^{3^{4}}}} \cdot hello * \text{hello world} ⋅ \text{my world} · \text{third world} ⊗ t
-\end{equation*}
-
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum \cite{adams1980hitchhiker}.
-
-Mauris id quam non magna fermentum malesuada id mattis lorem.
-In a dapibus neque.
-Etiam lacus dui, malesuada ac eleifend imperdiet, imperdiet ut ipsum.
-Vestibulum id ultricies est.
-Phasellus augue mauris, semper a luctus vel, faucibus in risus.
-Fusce commodo augue quis elit sagittis non viverra turpis bibendum.
-Nunc placerat sem non sapien malesuada malesuada ullamcorper orci luctus \cite{adams1980hitchhiker}.
-Morbi pharetra ligula integer mollis mi nec neque ultrices vitae volutpat leo ullamcorper.
-In at tellus magna.
-Curabitur quis posuere purus.
-Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
-Suspendisse tristique placerat feugiat.
-Aliquam vitae est at enim auctor ultrices eleifend a urna.
-Donec non tincidunt felis.
-Maecenas at suscipit orci.
-See \cref{myFigure}.
-\begin{figure}
-    \centering
-        \missingfigure[figwidth=6cm]{}
-    \caption[Short caption to special figure]{This is my special figure.
-Aliquam ultricies, arcu quis tempor rhoncus, tellus nisl tempus justo, condimentum tempor erat odio ac purus.
-Integer quis ipsum felis.
-Aliquam volutpat, leo ac consequat egestas, lectus lacus adipiscing quam, id iaculis dolor quam in erat.
-Phasellus tempor interdum arcu quis vestibulum}
-    \label{myFigure}
-\end{figure}
-
-Fusce id suscipit sem.
-Aliquam venenatis nibh nec nisl luctus vel consectetur neque dapibus.
-Nulla feugiat egestas turpis, ac viverra eros cursus sit amet.
-Cras tincidunt felis vel tellus ultrices condimentum.
-Quisque vehicula, arcu vitae interdum dignissim, purus tortor cursus libero, sit amet accumsan quam magna in neque.
-Phasellus luctus leo odio.
-Aliquam ultricies, arcu quis tempor rhoncus, tellus nisl tempus justo, condimentum tempor erat odio ac purus.
-Integer quis ipsum felis.
-Aliquam volutpat, leo ac consequat egestas, lectus lacus adipiscing quam, id iaculis dolor quam in erat.
-Phasellus tempor interdum arcu quis vestibulum.
-Pellentesque sit amet augue purus.
-See \cref{myTable}.
-
-\begin{table}
-    \centering
-        \begin{tabular}{c | r l}
-            h & h & h \\
-            e & e & e \\
-        \end{tabular}
-    \caption{This is a caption to the table}
-    \label{myTable}
-\end{table}
-
-\section{Torquent Arcu}
-Curabitur condimentum suscipit arcu, sit amet convallis urna pellentesque ac.
-Quisque fringilla tincidunt risus nec accumsan.
-Curabitur vel sagittis ante.
-Integer eget placerat leo.
-Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos.
-Vestibulum quis risus in nulla fermentum pellentesque dictum et erat.
-Nulla vel pretium nunc.
-Integer tortor lorem, suscipit sit amet ultricies non, porta at metus.
-Sed pharetra, ante facilisis interdum porta, mi dolor fringilla quam, ac porttitor urna dolor quis massa.
-Proin viverra semper tincidunt.
-Vivamus pulvinar pharetra condimentum.
-Pellentesque rutrum mollis tellus ac scelerisque.
-
-\begin{figure}
-    \centering
-        \subbottom[1 pass]{
-            \missingfigure[figwidth=6cm]{}
-            \label{twoFigures1}
-        }
-        ~
-        \subbottom[5 passes]{
-            \missingfigure[figwidth=6cm]{}
-            \label{twoFigures2}
-        }
-        \caption{loop performance comparison}
-    \label{twoFigures}
-\end{figure}
-
-\subsection{Vestibulum}
-Mauris luctus sollicitudin vestibulum.
-Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos \cref{twoFigures2}.
-Duis eu nisl nec turpis porttitor bibendum eget sed orci.
-Aliquam consequat lorem a dui viverra porta facilisis augue rutrum.
-Cras luctus tellus in lectus egestas eu consequat magna cursus.
-Aenean aliquam neque a nibh elementum ornare.
-Integer eleifend imperdiet commodo.
-Morbi auctor, dui vel laoreet congue, purus est accumsan augue, sit amet feugiat neque nisl vel lorem.
-Curabitur ante sem, lacinia id adipiscing quis, viverra tristique nulla.
-Pellentesque ullamcorper pellentesque metus varius facilisis.
-Cras ac dui id odio tempor scelerisque.
-Curabitur a egestas risus.
-Pellentesque quis velit in sapien accumsan auctor.
-Phasellus aliquam, sapien eget lobortis volutpat, libero metus porttitor nisl, sed hendrerit urna dolor nec mi.
-See \cref{fibonacci}.
-
-\begin{adjustwidth*}{0cm}{-0.4cm}
-\begin{lstlisting}[language=Python,caption=Fibonacci,label=fibonacci]
-# This is a comment
-import easy
-str = "I am a string"
-str2 = "Now i have an awsome string with ´ '' `` which are not TeX'ed"
-str3 = "What about awsome unicode characters? Like “, π, ”, Ω, ç. \" This"
-def fib(n):
-    if n == 0:
-        return 0
-    elif n == 1:
-        return 1
-    else:
-        return fib(n-1) + fib(n-2)
-str4 = "Yes it is possible with 80 charactes. Which this string proves. Wiiii."
-str5 = "It adjusts according to the spine"
-\end{lstlisting}
-\end{adjustwidth*}
-
-\section{Luctus}
-Praesent et pellentesque arcu.
-Phasellus venenatis mi eu lorem convallis et iaculis ante aliquet.
-Aenean rhoncus placerat metus, vel convallis leo suscipit eu.
-Integer dapibus venenatis commodo.
-Cras laoreet faucibus sem nec luctus.
-Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos.
-Cras consectetur lacinia dolor at gravida.
-Phasellus ipsum arcu, vulputate fermentum ultricies eget, tempor eu odio.
-Aenean accumsan vestibulum risus a mattis.
-See it on \cref{modifiedminibatch}.
-
-\begin{algorithm}
-\caption{Modified mini-batch $K$-means} \label{modifiedminibatch}
-\begin{algorithmic}[1]
-\State Given: $K$, mini-batch size $B$, iterations $T$, dataset $X$, correlation~matrix~$\mathrm{P}$.
-\State Initialize $C = \{\mathbf{c}^{(1)}, \mathbf{c}^{(2)}, \ldots, \mathbf{c}^{(K)}\}$ with random $\mathbf{x}$'es picked from $X$.
-\State $A \gets B \cdot T$ sorted random indexes to $X$, denoted $a_1, a_2, \ldots, a_{B\cdot T}$.
-\State $X' \gets \{\mathbf{x}^{(a_1)}, \mathbf{x}^{(a_2)}, \ldots, \mathbf{x}^{(a_{B\cdot T})}\}$ \Comment{Cache all points}
-\State $\textbf{size} \gets 0$
-\For {$i = 1$ to $T$}
-    \State $M \gets B$ examples picked randomly from $X'$
-
-    \For{$\mathbf{x} \in M$} \Comment{\textit{Assignment step}}
-        \State $\textbf{d}[\textbf{x}] \gets f(C,\mathbf{x}, \mathrm{P})$ \Comment{Cache closest center}
-    \EndFor
-
-    \For {$\mathbf{x} \in M$} \Comment{\textit{Update step}}
-        \State $\textbf{c} \gets \textbf{d[x]}$ \Comment{Get cached center for current \textbf{x}}
-        \State $\textbf{size}[\textbf{c}] \gets \textbf{size}[\textbf{c}] + 1$ \Comment{Update cluster size}
-        \State $\eta \gets \frac{1}{\textbf{size}[\textbf{c}]}$       \Comment{Get learning rate}
-        \State $\textbf{c} \gets (1 - \eta)\textbf{c}+\eta\textbf{x}$ \Comment{Take gradient step}
-    \EndFor
-\EndFor
-\State \Return {$C$, \textbf{size}}
-\end{algorithmic}
-\end{algorithm}
-
-Fusce id suscipit sem.
-Aliquam venenatis nibh nec nisl luctus vel consectetur neque dapibus.
-Nulla feugiat egestas turpis, ac viverra eros cursus sit amet.
-Cras tincidunt felis vel tellus ultrices condimentum.
-Quisque vehicula, arcu vitae interdum dignissim, purus tortor cursus libero, sit amet accumsan quam magna in neque.
-Phasellus luctus leo odio.
-Aliquam ultricies, arcu quis tempor rhoncus, tellus nisl tempus justo, condimentum tempor erat odio ac purus.
-Integer quis ipsum felis.
-Aliquam volutpat, leo ac consequat egestas, lectus lacus adipiscing quam, id iaculis dolor quam in erat.
-Phasellus tempor interdum arcu quis vestibulum.
-Pellentesque sit amet augue purus.
-\begin{figure}[htbp]
-    \centering
-        \missingfigure[figwidth=6cm]{}
-    \caption{This is the caption I wrote}
-    \label{fig:label}
-\end{figure}
-Curabitur condimentum suscipit arcu, sit amet convallis urna pellentesque ac.
-Quisque fringilla tincidunt risus nec accumsan.
-Curabitur vel sagittis ante.
-Integer eget placerat leo.
-Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos.
-Vestibulum quis risus in nulla fermentum pellentesque dictum et erat.
-Nulla vel pretium nunc.
-Integer tortor lorem, suscipit sit amet ultricies non, porta at metus.
-Sed pharetra, ante facilisis interdum porta, mi dolor fringilla quam, ac porttitor urna dolor quis massa.
-Proin viverra semper tincidunt.
-Vivamus pulvinar pharetra condimentum.
-Pellentesque rutrum mollis tellus ac scelerisque.
-
-\section{Sollicitudin vestibulum}
-Mauris luctus sollicitudin vestibulum.
-Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos.
-Duis eu nisl nec turpis porttitor bibendum eget sed orci.
-Aliquam consequat lorem a dui viverra porta facilisis augue rutrum.
-Cras luctus tellus in lectus egestas eu consequat magna cursus.
-Aenean aliquam neque a nibh elementum ornare.
-Integer eleifend imperdiet commodo.
-Morbi auctor, dui vel laoreet congue, purus est accumsan augue, sit amet feugiat neque nisl vel lorem.
-Curabitur ante sem, lacinia id adipiscing quis, viverra tristique nulla.
-Pellentesque ullamcorper pellentesque metus varius facilisis.
-Cras ac dui id odio tempor scelerisque.
-Curabitur a egestas risus.
-Pellentesque quis velit in sapien accumsan auctor.
-Phasellus aliquam, sapien eget lobortis volutpat, libero metus porttitor nisl, sed hendrerit urna dolor nec mi.
-
-Praesent et pellentesque arcu.
-Phasellus venenatis mi eu lorem convallis et iaculis ante aliquet.
-Aenean rhoncus placerat metus, vel convallis leo suscipit eu.
-Integer dapibus venenatis commodo.
-Cras laoreet faucibus sem nec luctus.
-Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos.
-Cras consectetur lacinia dolor at gravida.
-Phasellus ipsum arcu, vulputate fermentum ultricies eget, tempor eu odio.
-Aenean accumsan vestibulum risus a mattis.
-
-\begin{adjustwidth*}{0cm}{-0.4cm}
-\begin{lstlisting}[language=Python,caption=Fibonacci2,label=Fibonacci2]
-# This is a comment
-import easy
-str = "I am a string"
-str2 = "Now i have an awsome string with ´ '' `` which are not TeX'ed"
-str3 = "What about awsome unicode characters? Like “, π, ”, Ω, ç. \" This"
-def fib(n):
-    if n == 0:
-        return 0
-    elif n == 1:
-        return 1
-    else:
-        return fib(n-1) + fib(n-2)
-str4 = "Yes it is possible with 80 charactes. Which this string proves. Wiiii."
-str5 = "It adjusts according to the spine"
-\end{lstlisting}
-\end{adjustwidth*}
+\chapter{Resumé (Danish)}
+\blindtext[5]

### 2022-02-27 frontmatter/acknowledgements.tex
Initial Overleaf Import

%!TEX root = ../Thesis.tex
\chapter{Acknowledgements}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\vspace*{\fill}
\noindent \textit{The project was funded by Corti and Innovation Fund Denmark through an industrial PhD grant (no.\@ 0153-00167B).}

### 2022-02-27 preamble/etc.tex
Initial Overleaf Import

%!TEX root = ../Thesis.tex

% Content specific packages.

\usepackage{blindtext}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multicol}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{layouts}
\usepackage[showframe=\showframe,xetex]{geometry}

%\usepackage{pgfplots}                 % Plot tools
%\pgfplotsset{compat=1.7}
\usetikzlibrary{
    arrows,
    matrix,
    positioning,
    shapes,
    topaths,
    bayesnet,
}

% Listings
\lstset{
    basicstyle=\footnotesize\ttfamily,% the size of the fonts that are used for the code
    breakatwhitespace=false,          % sets if automatic breaks should only happen at whitespace
    breaklines=true,                  % sets automatic line breaking
    captionpos=b,                     % sets the caption-position to bottom
    commentstyle=\color{s14a},        % comment style
    deletekeywords={},                % if you want to delete keywords from the given language
    escapeinside={\%*}{*)},           % if you want to add LaTeX within your code
    frame=single,                     % adds a frame around the code
    keywordstyle=\bfseries\ttfamily\color{s09}, % keyword style
    language=Python,                  % the language of the code
    morekeywords={*,...},             % if you want to add more keywords to the set
    numbers=left,                     % where to put the line-numbers; possible values are (none, left, right)
    numbersep=5pt,                    % how far the line-numbers are from the code
    numberstyle=\sffamily\tiny\color{dtugray}, % the style that is used for the line-numbers
    rulecolor=\color{dtugray},        % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
    showspaces=false,                 % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
    showstringspaces=false,           % underline spaces within strings only
    showtabs=false,                   % show tabs within strings adding particular underscores
    stepnumber=1,                     % the step between two line-numbers. If it's 1, each line will be numbered
    stringstyle=\color{s07},          % string literal style
    tabsize=2,                        % sets default tabsize to 2 spaces
    title=\lstname,                   % show the filename of files included with \lstinputlisting; also try caption instead of title
}

### 2022-02-27 preamble/static.tex
Initial Overleaf Import

%!TEX root = ../Thesis.tex

% This information is used in titlepage, colophon, preface and hyperref setup (pdf metainfo), and other options.

%\def\thesistypeabbr{B.Eng.}
%\def\thesistype    {Bachelor of Engineering}
%\def\thesistypeabbr{B.Sc.Eng.}
%\def\thesistype    {Bachelor of Science in Engineering}
%\def\thesistypeabbr{M.Sc.}
%\def\thesistype    {Master of Science in Engineering}
\def\thesistypeabbr{Ph.D.}
\def\thesistype    {Doctor of Philosophy}

\def\thesisdepshort{DTU Compute}
\def\thesisdep     {Department of Applied Mathematics and Computer Science}

\def\thesisauthor  {Jakob Drachmann Havtorn}
\def\thesistitle   {Uncertainty Quantification in Machine Learning}
\def\thesissubtitle{}
\def\thesislocation{Copenhagen}
\def\thesisyear    {2023}
\def\thesismonth   {August}
\def\thesisday     {31}

\def\papersize     {b5paper} % Final papersize (b5paper/a4paper), recommended papersize for DTU Compute is b5paper
\def\showtrims     {false}   % Print on larger paper than \papersize and show trim marks (true/false)?
\def\showframe     {false}   % Show frame of writeable area, marginparsep and marginpar (true/false)?

\def\showtodos     {true}    % Show todos (true/false)?
\def\confidential  {false}   % Confidential thesis (true/false)?

### 2022-02-27 prefrontmatter/colophon.tex
Initial Overleaf Import

%!TEX root = ../Thesis.tex

\thispagestyle{empty} % No page numbers
\frieze{0}{0.5\paperheight}

\vspace*{\fill}
% \vspace*{0.15\textheight}
\small\sffamily
\noindent
\textsc{\textbf{DTU Compute}}\\
\textsc{\textbf{Department of Applied Mathematics and Computer Science}}\\
\textsc{\textbf{Technical University of Denmark}}
\smallskip\\
Matematiktorvet, Building 303B\\
2800 Kongens Lyngby, Denmark\\
% Phone +45 4525 3031\\
compute@compute.dtu.dk\\
www.compute.dtu.dk
\bigskip\\
\noindent
\textsc{\textbf{Corti}}\\
\textsc{\textbf{Research Department}}
\smallskip\\
Store Strandstræde 21. 4.tv\\
1255 København K, Denmark\\
info@corti.ai\\
www.corti.ai\\

\normalsize
\normalfont

### 2022-02-27 appendices/appendix.tex
Make project run in VS Code

--- 2022-02-27 frontmatter/acknowledgements.tex
@@
-%!TEX root = ../Thesis.tex
-\chapter{Acknowledgements}
+%!TEX root = ../thesis.tex
+\chapter{An Appendix}
 Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
@@
 Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-
-\vspace*{\fill}
-\noindent \textit{The project was funded by Corti and Innovation Fund Denmark through an industrial PhD grant (no.\@ 0153-00167B).}

### 2022-02-27 frontmatter/notation.tex
Make project run in VS Code

%!TEX root = ../thesis.tex

\makenomenclature
%\markboth{\MakeUppercase\nomname}{\MakeUppercase\nomname}
% Correct header on multiple page nomenclature
\patchcmd{\thenomenclature}
  {\chapter*{\nomname}}% usually only \chapter*{\nomname} is issued
  {\chapter*{\nomname}\markboth{\MakeUppercase\nomname}{\MakeUppercase\nomname}}
  {}{}
% Rename Nomenclature to Notation
\renewcommand{\nomname}{Notation}
% Increase width of box for mathematical symbol on left
\setlength{\nomlabelwidth}{2.5cm}
% \setlength{\nomitemsep}{-\parsep}

% Command for the group subtitles
\newcommand{\nommenclaturesubtitle}[1]{%
    \bigskip
    \item[\bfseries \hspace{2.5cm}\: #1
        % \vspace{1cm}\\
        % \hspace{2.4cm} #1 \hfill\\
        % \par\noindent\rule{\widthof{#1}}{1pt}
    ]
    %\par\noindent\rule{\widthof{#1}}{1pt}%
    %   \hspace*{-\leftmargin}%
    %   \rule[2pt]{0.45\linewidth}{1pt}%
    %   \hfill #1\hfill
    %   \rule[2pt]{0.45\linewidth}{1pt}%
}

% Introductory text to notation section
\renewcommand{\nompreamble}{%
    This section collects definitions of the notation used in the thesis. Much of this is defined in the thesis in the order that it is used but is collected here for reference and convenience. There are no major departures from convention.
}

% Definition of the groups
\renewcommand\nomgroup[1]{%
      \ifstrequal{#1}{A}{\nommenclaturesubtitle{Numbers and arrays}}{%
      \ifstrequal{#1}{B}{\nommenclaturesubtitle{Sets}}{%
      \ifstrequal{#1}{C}{\nommenclaturesubtitle{Indexing}}{%
      \ifstrequal{#1}{D}{\nommenclaturesubtitle{Linear algebra operations}}{%
      \ifstrequal{#1}{E}{\nommenclaturesubtitle{Calculus}}{%
      \ifstrequal{#1}{F}{\nommenclaturesubtitle{Probability theory}}{%
      \ifstrequal{#1}{G}{\nommenclaturesubtitle{Functions}}{%
      }}}}}}}
}

% Notation entries
\nomenclature[A, 01]{$a$}{A scalar; integer or real}
\nomenclature[A, 02]{$\a$}{A column vector}
\nomenclature[A, 03]{$\A$}{A matrix}
\nomenclature[A, 04]{$\text{diag}(\a)$}{A square diagonal matrix of the elements of $\a$}
\nomenclature[A, 05]{$\e$}{A vector of ones, $\e=\bmat{1 & \cdots & 1}\transpose$}
\nomenclature[A, 06]{$\I$}{The identity matrix, $\I=\text{diag}\pa{\e}$}

\nomenclature[B, 01]{$\mathcal{A}, \mathbb{A}$}{A set}
\nomenclature[B, 02]{$\mathbb{R}$}{Set of real numbers}
\nomenclature[B, 03]{$\cbra{s_1, s_2, \dots, s_n}$}{Set of scalars $s_1, s_2, \dots, s_n$; also written as $\cbra{s_i}_{i=1}^n$}
%\nomenclature[B, 04]{$\bra{a, b}$}{Inclusive real interval between $a$ and $b$, $\bra{a, b}=\cbra{x\in\mathbb{R}\mid a\leq x \leq b}$}
%\nomenclature[B, 05]{$\pa{a, b}$}{Exclusive real interval between $a$ and $b$, $\pa{a, b}=\cbra{x\in\mathbb{R}\mid a<x<b}$}
\nomenclature[B, 06]{$\left(a, b\right]$}{Semi-inclusive real interval between $a$ and $b$, $\left(a, b\right]=\cbra{x\in\mathbb{R}\mid a<x\leq b}$}

\nomenclature[C, 01]{$a_i$}{Element $i$ of vector $\a$ with first index $1$}
\nomenclature[C, 02]{$A_{i,j}$}{Element $i,j$ of matrix $\A$. Comma omitted when unambiguous.}
\nomenclature[C, 03]{$A_{i,:}$}{Row $i$ of matrix $\A$ (a row vector)}
\nomenclature[C, 04]{$A_{:,j}$}{Column $j$ of matrix $\A$ (a column vector)}
\nomenclature[C, 05]{$\a^\bra{l}$}{Neural network layer indexing; a vector in layer $l$}
\nomenclature[C, 06]{$\a^\vbra{t}$}{Recurrent neural network sequence indexing; vector at timestep $t$ in sequence $\cbra{\a^\vbra{t}}_{t=1}^T$}

\nomenclature[D, 01]{$\a\transpose$}{Transpose of column vector $\a$ to give row vector}
\nomenclature[D, 02]{$\A\transpose$}{Transpose of matrix $\A$}
\nomenclature[D, 03]{$\a\transpose\b$}{Inner product of $\a$ and $\b$, $\a\transpose\b=\a\cdot\b$}
\nomenclature[D, 04]{$\a\b\transpose$}{Outer product of $\a$ and $\b$}
\nomenclature[D, 05]{$\A\B$}{Matrix product of $\A$ and $\B$}
\nomenclature[D, 06]{$\A*\B$}{Convolution (actually cross-correlation) of $\A$ and $\B$}
\nomenclature[D, 07]{$\A\odot\B$}{Elementwise, or Hadamard, product of $\A$ and $\B$}
\nomenclature[D, 08]{$\det{\A}$}{Determinant of matrix $\A$}
\nomenclature[D, 09]{$\norm{\a}$}{$L^2$ norm of $\a$}
\nomenclature[D, 10]{$\norm{\a}_p$}{$L^p$ norm of $\a$}
\nomenclature[D, 11]{$\Tr{\A}$}{Trace of matrix $\A$, $\Tr{\A}=\sum_i A_{i,i}$}

\nomenclature[E, 01]{$\frac{\text{d}y}{\text{d}x}$}{Derivative of $y$ with respect to $x$}
\nomenclature[E, 02]{$\pderiv{y}{x}$}{Partial derivative of $y$ with respect to $x$}
\nomenclature[E, 03]{$\nabla_\x$}{Del, nabla or Laplacian operator, $\nabla_\x = \bmat{\pderiv{}{x_1} & \cdots & \pderiv{}{x_n}}^\text{T}, \; \x\in\mathbb{R}^{n}$}
\nomenclature[E, 04]{$\nabla_\x y$}{Gradient of $y$ with respect to $\x$, $\nabla_\x y = \bmat{\pderiv{y}{x_1} & \cdots & \pderiv{y}{x_n}}^\text{T}, \; \x\in\mathbb{R}^{n}$}
\nomenclature[E, 05]{$\H_\x$}{Hessian operator, $\H_\x = \nabla_\x\nabla_\x\transpose = \bmat{
	    \pderiv[2]{}{x_1} & \pderiv[1,1]{}{x_1,x_2} & \cdots & \pderiv[1,1]{}{x_1,x_n}\\
	    \pderiv[1,1]{}{x_2,x_1} & \pderiv[2]{}{x_2} & \cdots & \pderiv[1,1]{}{x_2,x_n}\\
	    \vdots & \vdots & \ddots & \vdots\\
	    \pderiv[1,1]{}{x_n,x_1} & \pderiv[1,1]{}{x_d,x_2} & \cdots & \pderiv[2]{}{x_n}
	    }$}
\nomenclature[E, 06]{$\H_\x y$}{Hessian matrix of $y$ with respect to $\x$; matrix of all second order partial derivatives}
\nomenclature[E, 07]{$\nabla^2_\x$}{Laplacian operator, $\nabla^2_\x=\nabla_\x\transpose\nabla_\x = \nabla_\x\cdot\nabla_\x=\sum_i \pderiv[2]{}{x_i}=\Tr{\H_\x}$}
\nomenclature[E, 08]{$\nabla^2_\x y$}{Laplacian of $y$ with respect to $\x$; sum of the unmixed second order partial derivatives}
\nomenclature[E, 09]{$\pderiv{\y}{\x}$}{Jacobian matrix of $\y$ w.r.t. $\x$, $\pderiv{\y}{\x} = \bmat{
	    \nabla_\x^\text{T} y_1\\
	    \vdots \\
	    \nabla_\x^\text{T} y_n}, \; \y\in\mathbb{R}^{n}$; matrix of all possible first order partial derivatives}
\nomenclature[E, 10]{$\int f(\x)\,\text{d}\x$}{Definite integral over the domain of $\x$}
\nomenclature[E, 11]{$\int_\mathcal{D} f(\x)\,\text{d}\x$}{Definite integral over the domain $\mathcal{D}$}

\nomenclature[F, 01]{$a\sim p$}{$a$ is a random variable with distribution $p$}
\nomenclature[F, 02]{$\text{E}\bra{f(x)}$}{Expectation of $f(x)$ with respect to $p(x)$; sometimes also $\text{E}\bra{f(x)}_{x\sim p(x)}$}
\nomenclature[F, 03]{$\text{Var}\bra{f(x)}$}{Variance of $f(x)$ under $p(x)$}
\nomenclature[F, 04]{$\text{Cov}\bra{f(x), g(x)}$}{Covariance of $f(x)$ and $g(x)$ under $p(x)$}
\nomenclature[F, 05]{$D_\text{KL}\pa{p\mid\mid q}$}{Kullback-Leibler divergence of $p$ and $q$}
\nomenclature[F, 06]{$D_\text{KL}\pa{p,q}$}{Symmetrized Kullback-Leibler divergence of $p$ and $q$}
\nomenclature[F, 07]{$\mathcal{N}\pa{\x\mid\mub,\Sigmab}$}{Gaussian distribution over $\x$ with mean $\mub$ and covariance $\Sigmab$}

\nomenclature[G, 01]{$f:\mathcal{D}\rightarrow\mathcal{C}$}{Function $f$ with domain $\mathcal{D}$ and co-domain $\mathcal{C}$}
\nomenclature[G, 02]{$\log x$}{Natural logarithm of $x$}
\nomenclature[G, 03]{$\text{ReLU}\pa{x}$}{Linear rectifier function, $\text{ReLU}\pa{x}=\max\cbra{0,x}$ (Rectified Linear Unit)}

### 2023-01-06 frontmatter/acknowledgements.tex
Make project run in VS Code
Fix name highlighting and update font

@@
-%!TEX root = ../Thesis.tex
-\chapter{Acknowledgements}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+%!TEX root = ../thesis.tex
+\chapter[acknowledgements]{Acknowledgements}
+First of all, I want to thank all my supervisors...
 
 \vspace*{\fill}
-\noindent \textit{The project was funded by Corti and Innovation Fund Denmark through an industrial PhD grant (no.\@ 0153-00167B).}
+\noindent \textit{The project was funded by Corti and Innovation Fund Denmark through industrial PhD grant no.\@ 0153-00167B).}

### 2023-01-06 prefrontmatter/colophon.tex
Make project run in VS Code
Fix name highlighting and update font

@@
-%!TEX root = ../Thesis.tex
+%!TEX root = ../thesis.tex
 
@@
 \noindent
-\textsc{\textbf{DTU Compute}}\\
-\textsc{\textbf{Department of Applied Mathematics and Computer Science}}\\
-\textsc{\textbf{Technical University of Denmark}}
+\textsc{\textbf{Technical University of Denmark}}\\
+\textsc{\textbf{Department of Applied Mathematics and Computer Science}}
+% \textsc{\textbf{DTU Compute}}
 \smallskip\\
-Matematiktorvet, Building 303B\\
-2800 Kongens Lyngby, Denmark\\
+\href{https://goo.gl/maps/Dx6rNjGT9Ad5xKgZ6}{%
+Richard Petersens Plads, Building 324\\
+2800 Kongens Lyngby, Denmark}\\
 % Phone +45 4525 3031\\
-compute@compute.dtu.dk\\
-www.compute.dtu.dk
+\href{mailto:compute@compute.dtu.dk}{compute@compute.dtu.dk}\\
+\href{www.compute.dtu.dk}{www.compute.dtu.dk}
 \bigskip\\
@@
 \smallskip\\
+\href{https://goo.gl/maps/X7en4p8vaHkfVuyJ7}{%
 Store Strandstræde 21. 4.tv\\
-1255 København K, Denmark\\
-info@corti.ai\\
-www.corti.ai\\
+1255 København K, Denmark}\\
+\href{mailto:info@corti.ai}{info@corti.ai}\\
+\href{www.corti.ai}{www.corti.ai}\\
 

### 2023-02-24 appendices/appendix.tex
Sync-up

@@
 %!TEX root = ../thesis.tex
-\chapter{An Appendix}
+\chapter[an appendix]{An Appendix}
 Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.

### 2023-02-24 chapters/background.tex
Sync-up

\chapter[background]{Background}\label{chp:background}

\begin{enumerate}
    \item Variational inference and variational autoencoders
    \item Automatic speech recognition
    \item Self-supervised learning
    \item 
\end{enumerate}

\cite{kingma_autoencoding_2014}
\cite{rezende_stochastic_2014}

\cite{bahdanau_neural_2014}

\begin{enumerate}
    \item Introduce variational inference in relation to variational autoencoders
    \item Introduce ASR and CTC
\end{enumerate}

### 2023-02-24 chapters/discussion.tex
Sync-up

%!TEX root = ../thesis.tex
\chapter[discussion]{Discussion}\label{chp:discussion}

\section{}

### 2023-02-24 chapters/introduction.tex
Sync-up

\chapter[introduction]{Introduction}\label{chp:introduction}

### 2023-02-24 chapters/main-contributions.tex
Sync-up

\chapter[main contribution]{Main contributions}\label{chp:main-contributions}

### 2023-02-24 chapters/paper_hierarchical.tex
Sync-up

\chapter[hierarchical vaes know what they don't know]{Hierarchical VAEs Know What They Don't Know}
\label{chp:hierarchical}

\section*{Abstract}
Deep generative models have been demonstrated as state-of-the-art density estimators.
Yet, recent work has found that they often assign a higher likelihood to data from outside the training distribution.
This seemingly paradoxical behavior has caused concerns over the quality of the attained density estimates.
In the context of hierarchical variational autoencoders, we provide evidence to explain this behavior by out-of-distribution data having in-distribution low-level features.
We argue that this is both expected and desirable behavior.
With this insight in hand, we develop a fast, scalable and fully unsupervised likelihood-ratio score for OOD detection that requires data to be in-distribution across all feature-levels.
We benchmark the method on a vast set of data and model combinations and achieve state-of-the-art results on out-of-distribution detection.

\section{Introduction}
\begin{figure}[t]
    \centering
    \includegraphics[width=0.7\columnwidth]{reconstructions-front-page-4-samples-2-recons2.pdf}
    \caption[Reconstructions of a hierarchical VAE trained on FashionMNIST.]{%
        Reconstructions using a hierarchical VAE trained on FashionMNIST.
        Reconstruction quality of OOD data is comparable to in-distribution data, resulting in high likelihoods and poor OOD discrimination.
        By sampling the $k$ bottom-most latent variables from the conditional prior distribution $p(\z_{\geq l}|\z_{>l})$ (latent reconstructions) instead of the approximate posterior $q(\z_{>l}|\z_{<l})$, the model reconstructs from the training distribution resulting in lower $p(\x|\z)$ for OOD data.
    }
    \label{fig:reconstructions-fashionmnist}
\end{figure}
% First paragraph: Explains why OOD detection is important in a broad perspective
The reliability and safety of machine learning systems applied in the real-world is contingent on the ability to detect when an input is different from the training distribution. %, an anomaly.
Supervised classifiers built as deep neural networks are well-known to misclassify such \textit{out-of-distribution} (OOD) inputs to known classes with high confidence \cite{goodfellow_explaining_2015, nguyen_deep_2015}.
Several approaches have been suggested to equip deep classifiers with OOD detection capabilities \cite{hendrycks_baseline_2017, lakshminarayanan_simple_2017, hendrycks_deep_2019, devries_learning_2018}.
% However, such methods are inherently supervised and require in-distribution labels or examples of OOD data from an a priori known distribution which limit their applicability.
But, such methods are inherently supervised and require in-distribution labels or examples of OOD data limiting their applicability and generality.

% Second paragraph: Outline deep generative modeling approach and their failure
Unsupervised generative models that estimate an explicit likelihood should understand what it means to be in- and out-of-distribution without requiring labels or examples of OOD data.
By directly modeling the training distribution, such models are expected to assign low likelihoods to OOD data as it originates from regions of little or no support under the learned density \cite{bishop_novelty_1994}.
Recent advances in deep generative models \cite{kingma_autoencoding_2014, rezende_stochastic_2014, oord_pixel_2016, salimans_pixelcnn_2017, kingma_glow_2018} have enabled learning high quality generative models on complex data such as natural images, sequences including audio \cite{oord_wavenet_2016} and graphs \cite{kipf_variational_2016}.
However, recent observations have brought into question the quality of the learned density estimates by showing that they often assign higher likelihoods to OOD data than to in-distribution data \cite{nalisnick_deep_2019, choi_waic_2019}.
Many complex data distributions can be explained to a large degree by low-level features, e.g. edges in images.
However, such features do not explain high-level semantics of the data and may inhibit OOD detection \cite{ren_likelihood_2019, nalisnick_deep_2019}

\textbf{In this paper}, we examine the failure cases of deep generative models on OOD detection tasks within the context of hierarchical VAEs, and make the following contributions:
\begin{itemize}
    \item[(i)] We provide evidence that the root cause of OOD failures is that learned low-level features generalize well across datasets and dominate the estimated likelihoods.
    \item[(ii)] We then propose a fast, scalable, and fully unsupervised likelihood-ratio score for OOD detection that is explicitly developed to ensure that data should be in-distribution across all feature levels, which prevents the low-level features from dominating.
    \item[(iii)] With the likelihood-ratio score, we demonstrate state-of-the-art performance across a wide range of known OOD failure cases.
\end{itemize}

\section{Why does OOD detection fail?}\label{sec:why-does-ood-fail}
\begin{figure}[t]
    \centering
    \includegraphics[width=1\columnwidth]{feature-correlation-heatmap2.pdf}
    \vspace{-5mm}
    \caption[Absolute correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST.]{
    Absolute correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST.
    We compute the correlation between the representations of the two different models given the same data, FashionMNIST (top) and MNIST (bottom).
    %We reduce the representations to a single number by summing over all dimensions and compute the correlation between these numbers for 256 examples.
    %To reduce de-correlation due to the stochastic latent variables, we encode using the mode of the approximate posterior distributions.
    }
    \vspace{-3mm}
    \label{fig:correlations-heatmap}
\end{figure}

The inability to detect out-of-distribution data with deep generative models is surprising.
Before the advent of deep generative models, this was not considered a major issue for probabilistic models \cite{bishop_novelty_1994}.
Is the failure due to model pathologies or something different?

Deep learning models are generally believed to form hierarchies of representations that range from low-level features to more conceptual ones related to semantics \parencite{bengio_representation_2013}.
This has also been observed within deep generative models \parencite{maaloe_biva_2019, child_very_2021}.
%Recent research \cite{maaloe_biva_2019, child_very_2021, vahdat_nvae_2020}, has shown that deep latent variable models are able to learn rich hierarchical representations with increasing levels of abstraction and global structure towards the top of the hierarchy.
For image data there is a trend that the low-level features are quite similar across models (edge detectors, etc.). This raises the question to what extend such features are relevant when detecting OOD data, also suggested  by \cite{nalisnick_deep_2019} and examined for Glow and PixelCNN in \cite{schirrmeister_understanding_2020}.
To investigate, we train two hierarchical VAEs (\cref{sec:background-hie-VAE}) on FashionMNIST and MNIST, respectively, and compute the between-models correlation of the extracted features of in-distribution data and OOD data.
The result appears in \cref{fig:correlations-heatmap}.
We observe that features extracted in the early layers (low-level features) correlate strongly between the two models, and that this correlation drops as we get into later layers.
This suggests that low-level features do not carry much information for OOD detection.

To shed further light on the impact of semantic versus low-level features, we look at model reconstructions of images with a hierarchical VAE (\cref{fig:biva-reconstructions-celeba}).
To study the feature hierarchy, we replace the inference distribution with the corresponding conditional prior in the first layers of the model to see what information is lost.
We observe that as more layers rely on the prior, more details are lost.
Sunglasses, which are uncommon, are first replaced by more common glasses, and then finally disappear.
%Likewise, a long-haired man eventually becomes woman.
This suggests that as we fall back to the conditional priors of each layer, we are pushed closer to local modes of the modeled distribution.

\begin{figure}[t]
    \centering
    %\vspace{0.17cm}
    \includegraphics[width=1\columnwidth]{biva_reconstructions.pdf}
    % https://docs.google.com/drawings/d/1UXdA3c18Oek9_env1vEB9mqOAHZ-tu0ijWmvqmapWCk/edit?usp=sharing
    \vspace{-5.2mm}
    \caption[Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables]{Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables  \cite{maaloe_biva_2019}.
    The higher the latent variable, the more the reconstructions fall into the mode of the learned distribution.
    It is more common to wear regular glasses than sunglasses but most common not to wear glasses at all.
    A man with long hair collapses into the mode of the more common long-haired woman.}
    \vspace{-3mm}
    \label{fig:biva-reconstructions-celeba}
\end{figure}
Finally, we look at reconstructions of out-of-distribution data.
\cref{fig:reconstructions-fashionmnist} illustrates that MNIST data is surprisingly well reconstructed by a hierarchical VAE trained on FashionMNIST.
Similar results have been found elsewhere \cite{xiao_likelihood_2020}.
We repeat the previous experiment and replace inference distributions by their corresponding conditional prior, and now observe that reconstructions from higher latent layers become increasingly similar to the data on which the model was trained.
The reliance on conditional priors seems to prevent accurate reconstruction of out-of-distribution data.
Some details are lost on in-distribution data too, but the distinction between that and out-of-distribution data becomes more clear.

\textbf{These observations lead to our main hypothesis.}
The lowest latent variables in a hierarchical VAE learn generic features that can describe a wide range of data.
This enables the model to achieve high rates of compression and high likelihoods, even on out-of-distribution data as long as the learned low-level features are appropriate.
We further suggest that OOD data are in-distribution with respect to these low-level features, but not with respect to semantic ones.

\vspace{-0.35cm}
\section{Background and related work}

\subsection{Variational autoencoders}
The variational autoencoder (VAE) \cite{kingma_autoencoding_2014, rezende_stochastic_2014} is a framework for constructing deep generative models defined by an observed variable $\mathbf{x}$ and a stochastic latent variable $\mathbf{z}$.
Typically, a neural network with parameters $\theta$ is chosen to parameterize the generative distribution $p_\theta(\x,\z)=p_\theta(\x|\z)p(\z)$, where the prior $p(\z)$ is commonly a standard Gaussian $\mathcal{N}(\0, \I)$.
The true posterior $p(\z|\x)$ is generally not analytically tractable and is approximated by a variational distribution $q_\phi(\z|\x)$ parameterized via another neural network with parameters $\phi$. The approximate posterior $q_\phi(\z|\x)$ is most often  a diagonal covariance Gaussian.
The model parameters $\theta$ and variational parameters $\phi$ are jointly optimized by maximizing the \textit{evidence lower bound} (ELBO),
\begin{equation}\label{eq:elbo}
    \log p_\theta(\x) \geq \mathbb{E}_{q_\phi(\z|\x)} \left[ \log \frac{p_\theta(\x,\z)}{q_\phi(\z|\x)} \right] \equiv \mathcal{L}(\x; \theta, \phi)\ .
\end{equation}
For brevity, we will denote $\mathcal{L}(\x; \theta, \phi)$ as $\mathcal{L}(\x)$ or $\mathcal{L}$. The reparameterization trick is used to backpropagate gradients through the stochastic latent variables with low variance.

The VAE is defined with a single latent variable which limits the ability to learn a high likelihood representation of complex input distributions, e.g.\ natural images.
There exists a few complementary approaches to make the VAE more flexible: (i) model a more expressive variational distribution $q_\phi(\z|\x)$ or prior distribution $p_\theta(\z)$ \cite{rezende_variational_2015, kingma_improved_2016}, (ii) model a more expressive posterior distribution $p_\theta(\x|\z)$ e.g. with a autoregressive decoder \cite{oord_conditional_2016} and (iii) learn a deeper hierarchy of latent variables \cite{burda_importance_2016, sonderby_ladder_2016}.
Here, we focus on the latter.

\subsection{Hierarchical variational autoencoders}\label{sec:background-hie-VAE}
Hierarchical VAEs are a family of probabilistic latent variable models which extends the basic VAE by introducing a hierarchy of $L$ latent variables $\z=\z_1, \dots, \z_L$.
The most common generative model is defined from the top down as $p_\theta(\x|\z)=p(\x|\z_1)p_\theta(\z_1|\z_2)\cdots p_\theta(\z_{L-1}|\z_L)$.
The inference model can then be defined in two ways respectively referred to as \textit{bottom-up} \cite{burda_importance_2016}
\begin{equation}
    %q_\phi(\z|\x)=q_\phi(\z_1|\x)q_\phi(\z_2|\z_1)\cdots q_\phi(\z_L|\z_{L-1})
    q_\phi(\z|\x) = q_\phi(\z_1|\x)\textstyle\prod_{i=2}^{L} q_\phi(\z_i|\z_{i-1})
\end{equation}
and \textit{top-down} \cite{sonderby_ladder_2016}
\begin{equation}
    % q_\phi(\z|\x)=q_\phi(\z_{L}|\x)q_\phi(\z_{L-1}|\z_L)\cdots q_\phi(\z_2|\z_1),
    q_\phi(\z|\x) = q_\phi(\z_L|\x)\textstyle\prod_{i=L-1}^{1} q_\phi(\z_{i}|\z_{i+1}) \ .
\end{equation}
Regardless of the choice of inference model, a hierarchical VAE is still trained using the ELBO \eqref{eq:elbo}.

Until recently, hierarchical VAEs gave inferior likelihoods compared to state-of-the-art autoregressive \cite{ho_flow_2019} and flow-based models \cite{salimans_pixelcnn_2017}.
This was changed by \textcite{maaloe_biva_2019}, \textcite{vahdat_nvae_2020}, and \textcite{child_very_2021}, which introduced complementary methods to extend the number of latent variables to a very deep hierarchy resulting in state-of-the-art likelihood performance.

In this paper we employ a simple hierarchical VAE with bottom-up inference paths and the more powerful BIVA variant with a bidirectional (top-down and bottom-up) inference model \parencite{maaloe_biva_2019}. We employ skip connections between latent variables but omit them for brevity.

\subsection{Out-of-distribution detection}\label{sec:background-ood-detection}
% Intro to alternative scores
So far, no reliable direct likelihood-based method has been found for fully unsupervised deep generative model OOD detection.
A major line of work considers developing new scores that are more reliable than the likelihood.
This includes the \textit{typicality} test presented by \textcite{nalisnick_detecting_2019} which is an OOD detection test based on the typicality of a batch of potentially OOD examples.
This approach however requires a batch of examples from the same class (OOD or not) which limits its practical applicability.
In \textcite{ren_likelihood_2019}, the \textit{likelihood ratio} between a primary model and a background model was shown to be an effective score for OOD detection.
However, to train the background model, the in-distribution data is perturbed via a data augmentation technique that is designed with knowledge about the confounding factors between the in-distribution data and the OOD data. Furthermore, it is tuned towards high performance on a known OOD dataset.
\textcite{serra_input_2020} take a similar approach and attribute the failure to detect OOD data to the high influence of the input complexity on the likelihood and choose a generic lossless compression algorithm as the background model.
Although this method gives good results, no single best choice of compression algorithm exists for all types of OOD data, and any particular choice encodes prior knowledge about the data into the detection method.
Both these methods can be seen as correcting for low-level features of the OOD data being assigned high model likelihood by using a second model focused exclusively on these features.

Similar to these methods, the majority of the approaches to OOD detection make assumptions about the nature of the OOD data.
The assumptions encompass using labels on the in-distribution data \cite{hendrycks_baseline_2017, liang_enhancing_2018, alemi_uncertainty_2018, lee_simple_2018, lakshminarayanan_simple_2017}, examples of OOD data \cite{hendrycks_deep_2019}, augmenting in-distribution data to mimic it \cite{ren_likelihood_2019}, or assuming a certain data type \cite{serra_input_2020}.
Any of these assumptions encode implicit biases into the model about the attributes of OOD data which, in turn, might impair performance on truly unknown data examples (unknown unknowns).

% Introduce the contrast to results with VAEs
While some of these methods achieve very good results on OOD detection with autoregressive models \cite{oord_pixel_2016, salimans_pixelcnn_2017} and invertible flow-based models \cite{kingma_glow_2018}, it was recently shown that they can be much less effective for VAEs \cite{xiao_likelihood_2020} highlighting the need for a more reliable OOD score for VAEs.
Although VAEs have the same failure cases as autoregressive and flow-based models, the caveat is that the difference in the likelihood is generally not as big and reconstructions of OOD can be surprisingly good \cite{xiao_likelihood_2020}.
\textcite{xiao_likelihood_2020} alleviate this by refitting the inference network, as previously proposed by \textcite{cremer_inference_2018, mattei_refit_2018}, to a potentially OOD example and measuring the so-called \textit{likelihood regret}.
However, refitting the inference network can be computationally expensive, especially for the large hierarchical VAEs that are used to model complex data \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}. Furthermore, this scales poorly to large amounts of potentially OOD examples as the optimization is done per example.

A few methods have approached OOD detection in a completely unsupervised fashion \cite{maaloe_biva_2019, choi_waic_2019, xiao_likelihood_2020}.
The work of \textcite{maaloe_biva_2019} is the most related to ours. They introduce BIVA, a deep hierarchy of stochastic latent variables with a top-down and bottom-up inference model and achieve state-of-the-art likelihood scores. 
They also provide early results indicative that a looser likelihood bound may have value in OOD detection.
In this paper, we provide an explanation of those results, and significantly improve upon them.

\section{OOD detection with hierarchical VAEs}
\subsection{A bound for semantic OOD detection}
If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \cref{fig:reconstructions-fashionmnist}-\ref{fig:biva-reconstructions-celeba}, then $p_\theta(\x|\z_1)$ will be high for both in- and out-of-distribution data.
Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phi(\z|\x)}[\log p_\theta(\x|\z_1)] - D_{\mathrm{KL}}( q_\phi(\z|\x) || p(\z))$ from \eqref{eq:elbo} relies on the KL-term for OOD detection. For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phi(\z|\x) || p(\z))$ can be expressed by a hierarchical sum% over the hierarchy
% \begin{multline}
%     \mathcal{L}(\x) =  \mathbb{E}_{q_\phi(\z|\x)} \Big[ \log p_\theta(\x|\z_1) \\
%                      + \textstyle\sum_{i=1}^{L-1} \log \frac{p_\theta(\z_i|\z_{i+1})}{q_\phi(\z_i|\z_{i-1})} 
%                      + \log \frac{p_\theta(\z_L)}{q_\phi(\z_L|\z_{L-1})} \Big] .
% \end{multline}
% \begin{multline}
%     D_{\mathrm{KL}}( q(\z|\x) || p(\z)) = \\
%     \mathbb{E}_{q_\phi(\z|\x)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\theta(\z_i|\z_{i+1})}{q_\phi(\z_i|\z_{i-1})} + \log \frac{p_\theta(\z_L)}{q_\phi(\z_L|\z_{L-1})} \Big] .
% \end{multline}
\begin{equation}
    \mathbb{E}_{q_\phi(\z|\x)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\theta(\z_i|\z_{i+1})}{q_\phi(\z_i|\z_{i-1})} + \log \frac{p_\theta(\z_L)}{q_\phi(\z_L|\z_{L-1})} \Big] \ .
\end{equation}
In general, the absolute log-ratios grow with $\mathrm{dim}(\z_i)$ as the individual log probability terms are computed by summing over the dimensionality of $\z_i$.
This means that the value of the KL-term is dominated by terms where $\z_i$ is high-dimensional. We refer to Appendix C for a more detailed argument.
Since hierarchical VAEs are generally constructed with a bottleneck type structure, the terms corresponding to latent variables towards the top of the hierarchy will have a vanishing influence on the value of the KL-term.
However, as the semantic information most relevant for OOD detection has a tendency to be represented in the top-most latent variables, this makes OOD detection using the regular ELBO difficult, even for state-of-the-art models.
This behavior has also been reported by \textcite{xiao_likelihood_2020}.

To shift the ELBO from primarily being based on the approximate posterior of the lowest latent variables to instead focus on the conditional prior, \textcite{maaloe_biva_2019} introduced slightly different likelihood lower bound defined as
\begin{equation}\label{eq:biva->k}
    \mathcal{L}^{>k} = \mathbb{E}_{p_\theta(\z_{\leq k}|\z_{>k}) q_\phi(\z_{>k}|\x)} \left[ \log \frac{p_\theta(\x|\z)p_\theta(\z_{>k})}{q_\phi(\z_{>k}|\x)} \right]
\end{equation}
where $k\in\{0,1,\dots,L\}$ (see Appendix for the derivation).
We note that $\mathcal{L}^{>0}$ is the regular ELBO (\ref{eq:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\z_1,\dots,\z_l \sim p_\theta(\z_{\leq k}|\z_{>k})$ and only the $L-k$ highest from the approximate posterior $\z_{k+1},\dots,\z_L \sim q_\phi(\z_{>k}|\x)$.
Importantly, this has the effect that the data likelihood $p(\x|\z)$ is dependent on the approximate posterior through a latent variable $\z_{k+1}$ different from $\z_1$ for all $k \geq 1$.
Thereby, the likelihood can be evaluated with a reconstruction from each of the latent variables $\z_k$ of the hierarchical VAE.
Hence, we can now test how well the input $\x$ is reconstructed from each latent variable.
The notation $\mathcal{L}^{>k}$ highlights that for latent variables $\z_{>k}$, the bound is the regular ELBO while for the latent variables $\z_{\leq k}$, the bound is evaluated using the (conditional) prior rather than the approximate posterior as the proposal distribution.

\subsection{A likelihood-ratio score for all feature levels}
While the $\mathcal{L}^{>k}$ bound provides a score for performing semantic OOD detection, it still relies on the data space likelihood function (see equation \eqref{eq:likelihoods-as-exact} below), which is known to be problematic for OOD detection (section \ref{sec:background-ood-detection}). To alleviate this, we phrase OOD detection as a likelihood ratio test of being \emph{semantically} in-distribution.
A standard likelihood ratio test \parencite{buse_likelihood_1982} suggests to consider the ratio between the associated likelihoods, which we can approximate on a log-scale by the corresponding lower bounds $\mathcal{L}$ and $\mathcal{L}^{>k}$,
\begin{equation}\label{eq:llr-as-difference-in-likelihoods}
    % LLR = - 2 log(L0 / L1)
    % where L0 < L1 so log(L0 / L1) < 0 and LLR > 0
    % LLR = 2 log(L1 / L0)
    %     = 2 (log(L1) - log(L0))
    %     = 2 (ELBO - L^{>k})  % LLR^{>a,>b} where a=0
    % L1 = ELBO
    % L0 = L^{>k}
    LLR^{>k}(\x) = \mathcal{L}(\x) - \mathcal{L}^{>k}(\x) \ .
\end{equation}
%This likelihood ratio contrasts the ELBO with $\mathcal{L}^{>k}$ for a variable choice of $k$.
Since, empirically, $\mathcal{L}\geq\mathcal{L}^{>k}$, the ratio is always positive as is standard for likelihood ratio tests.
A low value of $LLR^{>k}(\x)$ means that the ELBO and $\mathcal{L}^{>k}$ are almost equally tight for the data.
On the contrary, a high value indicates that $\mathcal{L}^{>k}$ is looser on the data than the ELBO; hence, the data may be OOD.

We can gather further insights about this score if we write the regular ELBO and the $\mathcal{L}^{>k}$ bounds in the exact form that includes the intractable KL-divergence between the approximate and true posteriors,
\begin{align}
    \mathcal{L}      &= \log p_\theta(\x) - D_{\mathrm{KL}}\left( q_\phi(\z|\x) || p_\theta(\z|\x)\right), \label{eq:likelihoods-as-exact} \\ 
    \mathcal{L}^{>k} &= \log p_\theta(\x) - D_{\mathrm{KL}}\left( p_\theta(\z_{\leq k }|\z_{>k}) q_\phi(\z_{>k}|\x) || p_\theta(\z|\x)\right) \nonumber \ .
\end{align}
Subtracting these cancel out the two data likelihood terms $\log p_\theta(\x)$ and only the KL-divergences from the approximate to the true posterior remain,
\begin{align}\label{eq:llr-as-kls}
    LLR^{>k}(\x) &= - D_{\mathrm{KL}}\left( q_\phi(\z|\x) || p_\theta(\z|\x)\right) \\
                 &\quad + D_{\mathrm{KL}}\left( p_\theta(\z_{\leq k}|\z_{>k}) q_\phi(\z_{>k}|\x) || p_\theta(\z|\x)\right) \ . \notag
\end{align}

Hence, it is clear that compared to the likelihood bound $\mathcal{L}^{>k}$, this likelihood-ratio measures divergence exclusively in the latent space whereas $\mathcal{L}^{>k}$ includes the $\log p_\theta(\x)$ term similar to the ELBO.
Therefore, the $LLR^{>k}$ score should be an improved method for semantic OOD detection compared to $\mathcal{L}^{>k}$.
Now, it can be noted that if we replace the regular ELBO, $\mathcal{L}$, in \eqref{eq:likelihoods-as-exact} with the strictly tighter importance weighted bound \cite{burda_importance_2016},
\begin{equation}\label{eq:iw-bound}
    \mathcal{L}_{S} = \mathbb{E}_{q(\z|\x)}\left[ \log \frac{1}{N} \sum_{s=1}^{S} \frac{p(\x, \z^{(s)})}{q(\z^{(s)}|\x)} \right] \ ,
\end{equation}
then, in the limit $S\rightarrow\infty$, we have $\mathcal{L}_{S} \rightarrow \log p_\theta(\x)$ and the likelihood ratio reduces to
\begin{equation}\label{eq:llr-as-kls-iwae-reduced}
    LLR^{>k}_{S}(\x) \rightarrow D_{\mathrm{KL}}( p(\z_{\leq k}|\z_{>k}) q(\z_{>k}|\x) || p(\z|\x))
\end{equation}
which, in practice, is well-approximated for a finite $S$. We expect this importance weighted likelihood ratio to monotonically improve upon the one in (\ref{eq:llr-as-kls}) as $S$ increases and the KL-divergence in the regular ELBO that contains terms for which $\z_i$ is high-dimensional goes to zero.

Since the scores in (\ref{eq:llr-as-kls}) and (\ref{eq:llr-as-kls-iwae-reduced}) are estimated by sampling their estimators are stochastic objects with nonzero variance.
We note that $\text{Var}(\widehat{LLR}^{>k}) = \text{Var}(\hat{\mathcal{L}}) + \text{Var}(\hat{\mathcal{L}}^{>k}) - 2\, \text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$.
Since $\log p_\theta(\x)$ and part of the KL divergence are identical in the expressions of $\mathcal{L}$ and $\mathcal{L}^{>k}$ we expect $\text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$ to be positive which reduces the total variance. 
Empirical results indeed show that $\text{Var}(\widehat{LLR}^{>k})$ is larger than $\text{Var}(\hat{\mathcal{L}})$ but smaller than $\text{Var}(\hat{\mathcal{L}}^{>k})$.
%Whether this decreases the variance to below the variance of any of the two bounds, we do not know, and we believe it to be difficult to verify mathematically.
Nevertheless, the variance of the estimators is guaranteed to go to zero as the number of samples is increased.

The OOD scores considered in this research all assume that what discriminates an out-of-distribution from an in-distribution data point are semantic, high-level features. Clearly, if this is not the case and the difference instead lies in low-level statistics, the scores would likely fail. We hypothesize that a complementary bound to (\ref{eq:biva->k}), $\mathcal{L}^{<l}$ described in Appendix E, might be useful in these cases, but leave further examination to future work.

\section{Experimental setup}

\paragraph{Tasks} We follow existing literature \cite{nalisnick_deep_2019, hendrycks_deep_2019} and evaluate our method by setting up OOD detection tasks from FashionMNIST \cite{xiao_fashionmnist_2017} to MNIST \cite{lecun_gradientbased_1998} and from CIFAR10 \cite{krizhevsky_learning_2009} to SVHN \cite{netzer_reading_2011}.
For each experiment we train our model on the train split of the former dataset and test its ability to recognize the test split of the latter dataset as OOD from the test split of the former dataset.
We use the standard train/test splits for the datasets.
More details on the datasets can be found in the Appendix.

% Our model
%\begin{wrapfigure}{R}{0.5\columnwidth}
% \begin{SCfigure}[50][t!]
\begin{figure}[t]
    % \vspace{-1.3cm}
    %\begin{figure}
    \centering
    \tikz{
        % inference
        % nodes
        \node[obs] (x_inf) {$\x$};%
        \node[latent,above=.75cm of x_inf](z1_inf){$\z_1$}; %
        \node[latent,above=.75cm of z1_inf](z2_inf){$\z_2$}; %
        % \node[latent,above=.75cm of z2_inf](z3_inf){$\z_3$}; %
        \node[above=of z2_inf, yshift=-1.cm] (phi) {$q_\phi(\z|\x)$};

        % edges
        \edge[]{x_inf}{z1_inf};
        \edge[]{z1_inf}{z2_inf};
        % \edge[]{z2_inf}{z3_inf};
        \edge[dashed, bend left]{x_inf}{z2_inf};
        % \edge[dashed, bend left]{x_inf}{z3_inf};

        % generative
        % nodes$
        \node[obs,right=0.75cm of x_inf] (x_gen) {$\x$};%
        \node[latent,above=.75cm of x_gen](z1_gen){$\z_1$}; %
        \node[latent,above=.75cm of z1_gen](z2_gen){$\z_2$}; %
        % \node[latent,above=.75cm of z2_gen](z3_gen){$\z_3$}; %
        \node[above=of z2_gen, yshift=-1.cm] (theta) {$p_\theta(\x,\z)$};

        % edges
        % \edge[]{z3_gen}{z2_gen};
        \edge[]{z2_gen}{z1_gen};
        \edge[]{z1_gen}{x_gen};
        \edge[dashed, bend left]{z2_gen}{x_gen};
        % \edge[dashed, bend left]{z3_gen}{x_gen};
    }
    \caption[Inference and generative models for a bottom-up hierarchical VAS]{The inference and generative models, $q_\phi$ and $p_\theta$, for an $L=2$ layered bottom-up hierarchical VAE as the one used in our experiments.
    Dashed lines indicate deterministic skip connections which are employed in both networks. Skip connections are found to be useful for optimizing latent variable models \parencite{dieng_avoiding_2019, maaloe_biva_2019}.}
    \label{fig:hvae-graphical-model}
    % \end{figure}
    %\vspace{-0.8cm}
%\end{wrapfigure}
% \end{SCfigure}
\end{figure}

\paragraph{Models} For each OOD task, we train a simple bottom-up hierarchical VAE with $L$ stochastic layers which we will refer to as ``HVAE''.
To alleviate posterior collapse we include skip-connections that connect $\z_i$ to $\z_{i+2}$ for $i\in\{0, L-2\}$ and $\z_0\equiv\x$ in both the inference and generative models \cite{dieng_avoiding_2019} and employ the \textit{free bits} scheme with $\lambda=2$ \cite{kingma_improved_2016}.
We use weight-normalization \cite{salimans_weight_2016} on all weights and residual networks in the deterministic paths. 
A graphical representation of this model can be seen in \cref{fig:hvae-graphical-model}.
We use a Bernoulli output distribution for FashionMNIST/MNIST and a discretized mixture of logistics output distribution \cite{salimans_pixelcnn_2017} for CIFAR10/SVHN.
We use $L=3$ for grey-scale images and $L=4$ for natural images.
For CIFAR/SVHN, we also train a BIVA model \cite{maaloe_biva_2019} with $L=10$ and similar configuration as used by the original paper\footnote{Source code available at \url{github.com/larsmaaloee/BIVA} and \url{github.com/vlievin/biva-pytorch}}.
All models are trained by optimizing the ELBO in (\ref{eq:elbo}). We implement our models in PyTorch \cite{paszke_automatic_2017}\footnote{Source code available at \url{github.com/jakobhavtorn/hvae-oodd}}.
Full model details are in the Appendix.

% Baselines
\paragraph{Baselines} We group baselines into those that use prior knowledge about OOD data, ones that use labels associated with the in-distribution data and purely unsupervised approaches that do not make such assumptions.
Our method falls into the latter category.
For more information on each baseline, we refer to the original literature.

% Metrics and Evaluation
\paragraph{Evaluation} Following previous work \cite{hendrycks_baseline_2017, hendrycks_deep_2019, alemi_uncertainty_2018, ren_likelihood_2019, choi_waic_2019} we use the threshold-independent evaluation metrics of Area Under the Receiver Operator Characteristic (AUROC$\uparrow$), Area Under the Precision Recall Curve (AUPRC$\uparrow$) and False Positive Rate at 80\% true positive rate (FPR80$\downarrow$) where the arrow indicates the direction of improvement.
Note that these metrics are only computable given examples of OOD data but faced with truly OOD data (unknown unknowns), there are many ways to select thresholds to use in practice e.g.\ as the one that yields a specific tolerable false positive rate on the in-distribution test data.
To compute the metrics, we use an equal number of samples from the in-distribution and OOD datasets by including all examples in the smallest of the two sets and randomly sampling equally many from the larger. We compute the $LLR^{>k}$ score with one and $S$ importance samples denoted by $LLR^{>k}_S$.

% The value of k
\paragraph{Selection of $k$} To determine whether an example is OOD in practice, the value of $LLR^{>k}$ is computed on the in-distribution test set for all $k$ and the resulting empirical distribution is used as reference.
If for any value of $k$, the $LLR^{>k}$ score of a new input differs significantly from the empirical distribution, it is regarded OOD.
If it differs for multiple values of $k$, the value for which it differs the most is selected.
In our experiments, we consider an entire dataset at a time and report the results of $LLR^{>k}$ with the value of $k$ that yielded the highest AUROC$\uparrow$ for that dataset in a threshold-free manner.
In practice, slightly better performance may be achieved by choosing $k$ per example.
This would not exclude the use of batching in our method, since $LLR^{>k}$ is computed after the forward pass.

\section{Results}

The likelihoods for our trained models are in \cref{tab:bits-per-dim-ood} alongside baseline results for in-distribution and OOD data.
The main results of the paper on the OOD tasks can be seen along with comparisons to the baseline methods in \cref{tab:rocauc-ood}.
We note that for all our results, the value of the score ($\mathcal{L}^{>k}$ and $LLR^{>k}$) for the training and test splits of the in-distribution data was observed to have the same empirical distribution to within sampling error hence yielding an AUROC score of $\approx0.5$ as expected.
Results on additional commonly used datasets are found in Appendix G.

\begin{table}[t!]
    \centering
    \resizebox{0.8\columnwidth}{!}{%
    \begin{tabular}{lrrrrr}
        \toprule
         Method & Dataset & \multicolumn{4}{c}{Avg. bits/dim}\\
        %   & & $\log p(x)$ & $\mathcal{L}^{>k_1}$ & $\mathcal{L}^{>k_2}$ & $\mathcal{L}^{>k_3}$\\
          & & $\log p(x)$ & $\mathcal{L}^{>1}$ & $\mathcal{L}^{>2}$ & $\mathcal{L}^{>3}$\\
         \midrule
         \multicolumn{6}{c}{\textbf{Trained on FashionMNIST}} \\
         \midrule
         \multirow{2}{*}{Glow}
            & FashionMNIST & 2.96 & - & - & \\
            & MNIST & 1.83 & - & - & \\
         \multirow{2}{*}{HVAE (Ours)}
            & FashionMNIST & 0.420 & 0.476 & 0.579 & - \\
            & MNIST & 0.317 & 0.601 & 0.881 & - \\
         \midrule
         \multicolumn{6}{c}{\textbf{Trained on CIFAR10}} \\
         \midrule
         \multirow{2}{*}{Glow}
          & CIFAR10 & 3.46 & - & - & \\
          & SVHN & 2.39 & - & - & \\
         \multirow{2}{*}{HVAE (Ours)}
            & CIFAR10 & 3.74 & 17.8 & 54.3 & 75.7 \\  % log p(x) = 8010.03
            & SVHN & 2.62 & 10.2 & 64.0 & 93.9 \\
         \multirow{2}{*}{BIVA (Ours)}
          & CIFAR10 & 3.46 & 8.74 & 19.7 & 37.3 \\
          & SVHN & 2.35 & 6.62 & 25.1 & 59.0 \\
         \bottomrule
    \end{tabular}
    }
    \caption{
    Average bits per dimension of different datasets for models trained on FashionMNIST and CIFAR10.
    For the hierarchical models we include the $\mathcal{L}^{>k}$ bounds.
    The likelihoods of training and test splits of the in-distribution data are all cases close.
    Since we train on dynamically binarized FashionMNIST, our bits/dim are smaller than for Glow.
    As $k$ is increased for the $L^{>k}$ bound, the bound gets looser but the model eventually assigns higher likelihood to the in distribution data than to the OOD data.
    Glow refers to \textcite{kingma_glow_2018, nalisnick_deep_2019}.
    BIVA refers to our implementation of \textcite{maaloe_biva_2019}.}
    \label{tab:bits-per-dim-ood}
    \vspace{-3mm}
\end{table}

\footnotetext{
    \textcite{serra_input_2020} performs the best when high likelihoods are assigned to OOD data such that the overlap with in-distribution data is low.
    Performance is worse when the overlap is high, cf. \textcite[Table 1]{serra_input_2020}, as seen with complex images. 
}
\begin{table}[t!]
    \centering
    \resizebox*{!}{0.9\textheight}{%
    \begin{tabular}{lrrr}
        \toprule
         Method & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
         \midrule
         \multicolumn{4}{c}{\textbf{FashionMNIST (in) / MNIST (out)}} \\
         \midrule
         \multicolumn{4}{l}{\textbf{Use prior knowledge of OOD}} \\
Backgr. contrast. LR (PixelCNN) {\cite{ren_likelihood_2019}}               & $0.994$ & $0.993$ & $0.001$ \\
Backgr. contrast. LR (VAE) {\cite{choi_waic_2019}}                    & $0.924$ & - & - \\
Binary classiﬁer {\cite{ren_likelihood_2019}}                              & $0.455$ & $0.505$ & $0.886$ \\ % 6
$p(\hat{y} | \x)$ with OOD as noise class {\cite{ren_likelihood_2019}}     & $0.877$ & $0.871$ & $0.195$ \\ % 7
$p(\hat{y} | \x)$ with calibration on OOD {\cite{ren_likelihood_2019}}     & $0.904$ & $0.895$ & $0.139$ \\ % 8
Input complexity ($S$, Glow) \cite{hendrycks_deep_2019}                    & $0.998$ & - & - \\
Input complexity ($S$, PixelCNN++) \cite{hendrycks_deep_2019}              & $0.967$ & - & - \\
         \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
$p(\hat{y} | \x)$ {\cite{ren_likelihood_2019, hendrycks_baseline_2017}}                        & $0.734$ & $0.702$ & $0.506$ \\
Entropy of $p(y | \x)$ {\cite{ren_likelihood_2019}}                        & $0.746$ & $0.726$ & $0.448$ \\
ODIN {\cite{ren_likelihood_2019, liang_enhancing_2018}}                                       & $0.752$ & $0.763$ & $0.432$ \\
VIB \cite{alemi_uncertainty_2018, choi_waic_2019}                                          & $0.941$ & - & - \\
Mahalanobis distance, CNN {\cite{ren_likelihood_2019}}                     & $0.942$ & $0.928$ & $0.088$ \\
Mahalanobis distance, DenseNet {\cite{lee_simple_2018}}                & $0.986$ & - & - \\
Ensemble, 20 classiﬁers {\cite{ren_likelihood_2019, lakshminarayanan_simple_2017}}                  & $0.857$ & $0.849$ & $0.240$ \\
         \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
         \multicolumn{4}{l}{\textit{- Ensembles}} \\
WAIC, 5 models, VAE {\cite{choi_waic_2019}}                          & $0.766$ & - & - \\
WAIC, 5 models, PixelCNN {\cite{ren_likelihood_2019}}                      & $0.221$ & $0.401$ & $0.911$ \\
        \multicolumn{4}{l}{\textit{- Not ensembles}} \\
Likelihood regret \cite{xiao_likelihood_2020}                               & $\mathbf{0.988}$ & - & - \\
$\mathcal{L}^{>0}$ + HVAE (ours)                    & $0.268$ & $0.363$ & $0.882$ \\
$\mathcal{L}^{>1}$ + HVAE (ours)                    & $0.593$ & $0.591$ & $0.658$ \\
$\mathcal{L}^{>2}$ + HVAE (ours)                    & $0.712$ & $0.750$ & $0.548$ \\
$LLR^{>1}$ + HVAE (ours)                            & $0.964$ & $0.961$ & $0.036$ \\
$LLR^{>1}_{250}$ + HVAE (ours)                      & $0.984$ & $\mathbf{0.984}$ & $\mathbf{0.013}$ \\
         \midrule
         \multicolumn{4}{c}{\textbf{CIFAR10 (in) / SVHN (out)}} \\
         \midrule
         \multicolumn{4}{l}{\textbf{Use prior knowledge of OOD}} \\
Backgr. contrast. LR (PixelCNN) {\cite{ren_likelihood_2019}}               & $0.930$ & $0.881$ & $0.066$ \\
Backgr. contrast. LR (VAE) {\cite{xiao_likelihood_2020}}                    & $0.265$ & - & - \\
Outlier exposure {\cite{hendrycks_deep_2019}}                              & $0.984$ & - & - \\
Input complexity ($S$, Glow) \cite{serra_input_2020}                   & $0.950$ & - & - \\
Input complexity ($S$, PixelCNN++) \cite{serra_input_2020}             & $0.929$ & - & - \\
Input complexity ($S$, HVAE) (Ours) \cite{serra_input_2020}\footnotemark & $0.833$ & $0.855$ & $0.344$ \\
         \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
Mahalanobis distance {\cite{lee_simple_2018}}                          & $0.991$ & - & -  \\
         \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
         \multicolumn{4}{l}{\textit{- Ensembles}} \\
WAIC, 5 models, Glow {\cite{choi_waic_2019}}                          & $1.000$ & - & - \\
WAIC, 5 models, PixelCNN {\cite{ren_likelihood_2019}}                      & $0.628$ & $0.616$ & $0.657$ \\
         \multicolumn{4}{l}{\textit{- Not ensembles}} \\
Likelihood regret \cite{xiao_likelihood_2020}                               & $0.875$ & - & - \\
$LLR^{>2}$ + HVAE (ours)                            & $0.811$ & $0.837$ & $0.394$ \\
$LLR^{>2}$ + BIVA (ours)                            & $\mathbf{0.891}$ & $\mathbf{0.875}$ & $\mathbf{0.172}$ \\
         \bottomrule
    \end{tabular}%
    }
    \caption{%
        AUROC$\uparrow$, AUPRC$\uparrow$ and FPR80$\downarrow$ for OOD detection for a FashionMNIST model using scores on the FashionMNIST test set as reference. We bold the best results within the "No OOD-specific assumptions" group since we only compare directly to those.
        HVAE (ours) refers to our hierarchical bottom-up VAE.
        BIVA (ours) refers to our implementation of the hierarchical BIVA model \cite{maaloe_biva_2019}.
    }
    \label{tab:rocauc-ood}
\end{table}

\subsection{Likelihood-based OOD detection}
% \begin{sidewaysfigure}
\begin{figure}
    %\captionstyle{centerlast}
    \centering
    \begin{subfigure}[l]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{densities-FashionMNIST-test-MNIST-test-bpp-k-0_sohau.pdf}
        \caption{}
        \label{fig:FMNIST-elbo-k0}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{densities-FashionMNIST-test-MNIST-test-bpp-k-2_sohau.pdf}
        \caption{}
        \label{fig:FMNIST-elbo-k2}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[r]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{densities-FashionMNIST-test-MNIST-test-LLR-0-1_sohau.pdf}
        \caption{}
        \label{fig:FMNIST-llr}
    \end{subfigure}
    \caption{
    Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood \subref{fig:FMNIST-elbo-k0}, the $\mathcal{L}^{>2}$ bound \subref{fig:FMNIST-elbo-k2} and the $LLR^{>1}$ score \subref{fig:FMNIST-llr}. All densities are computed using the HVAE model.).
    For the regular likelihood MNIST is very clearly more likely on average than the FashionMNIST test data while with the $\mathcal{L}^{>2}$ bound separation is better but significant overlap remains.
    The $LLR^{>1}$ provides a high degree of separation. Likelihoods are reported in units of the natural log of the number of bits per dimension.
    }
    \label{fig:FMNIST-ood-densities}
\end{figure}
% \end{sidewaysfigure}

We first report the results of the different variations of the $\mathcal{L}^{>k}$ bound for OOD detection. 
We reconfirm the results of \textcite{nalisnick_deep_2019} by observing that our hierarchical latent variable models also assign higher $\mathcal{L}^{>0}$ to the OOD dataset in the FashionMNIST/MNIST and CIFAR10/SVHN cases resulting in an AUROC$\uparrow$ inferior to random (\cref{tab:rocauc-ood}).
Switching the in-distribution data for the OOD data in both cases result in correctly detecting the OOD data; an asymmetry also reported by \textcite{nalisnick_deep_2019}.
\cref{fig:FMNIST-elbo-k0} shows the density of $\mathcal{L}^{>0}$ in bits per dimension \cite{theis_note_2016} by the model trained on FashionMNIST when evaluated on the FashionMNIST and MNIST test sets.
We observe a high degree of overlap, with less separation of the OOD data compared to similar results of autoregressive and flow-based models, like \textcite{xiao_likelihood_2020}.

We then evaluate the looser $\mathcal{L}^{>k}$ \eqref{eq:biva->k} for $k\in\{1,L\}$.
\cref{fig:FMNIST-elbo-k2} shows the result for $\mathcal{L}^{>2}$, which yielded the highest AUCROC$\uparrow$, only slightly better than random.
Like \textcite{maaloe_biva_2019}, we see that increasing the value of $k$ generally leads to improved OOD detection.
However, we also observe that the two empirical distributions never cease to overlap.
Importantly, depending on the OOD dataset, the amount of remaining overlap can be high which limits the discriminatory power of the likelihood-based $\mathcal{L}^{>k}$ bound.
This is in-line with the pathological behavior of the raw likelihood of latent variable models when used for OOD detection \cite{xiao_likelihood_2020}.
Since a high degree of overlap also seems present in \textcite{maaloe_biva_2019}, and we see the same problem for our BIVA model trained on CIFAR10, we do not expect this to be due to the less expressive HVAE.

\subsection{Likelihood-ratio-based OOD detection}
\begin{figure}
    \centering
    \begin{subfigure}[l]{0.48\columnwidth}
        \includegraphics[width=1\columnwidth]{roc-FashionMNIST-test-MNIST-test-ll-and-llr-IW250_sohau.pdf}
        \caption{}
        % \caption{ROC curves with AUROC score for detecting MNIST as OOD with the HVAE model trained on FashionMNIST.
        % A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>1}$.}
        \label{fig:FMNIST-roc-llr}
    \end{subfigure}
    \hfill
    \begin{subfigure}[r]{0.48\columnwidth}
        \includegraphics[width=1\columnwidth]{roc-biva-CIFAR10-SVHN-ll-and-llr_sohau.pdf}
        \caption{}
        % \caption{ROC curves with AUROC score for detecting SVHN as OOD with the BIVA model trained on CIFAR10.
        % A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>2}$.}
        \label{fig:CIFAR10-roc-llr}
    \end{subfigure}
    \caption{%
        ROC curves with AUROC score for detecting \subref{fig:FMNIST-roc-llr} MNIST as OOD with the HVAE model trained on FashionMNIST and \subref{fig:CIFAR10-roc-llr} SVHN as OOD with the BIVA model trained on CIFAR10. 
        A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>1}$.
    }
\end{figure}
We now move to the likelihood ratio-based score.
We find that $LLR^{>k}$ separates the OOD MNIST data from in-distribution FashionMNIST to a higher degree than the likelihood estimates as can be seen by the empirical densities of the score in \cref{fig:FMNIST-llr}.
We note that the likelihood ratio between the ELBO and the $\mathcal{L}^{>k}$ bound provides the highest degree of separation of MNIST and FashionMNIST as measured by the AUROC$\uparrow$ for $k=1$ smaller than $L$.
This is not surprising since the value of $k$ that provides the maximal separation to the reference in-distribution dataset need not be the one for which $\mathcal{LLR}^{>k}$ is overall maximal for the OOD dataset.
We also visualize the ROC curves resulting from using the $LLR^{>k}$ score for OOD detection on both FashionMNIST/MNIST and CIFAR10/SVHN and compare it to the ROC curves resulting from the different $\mathcal{L}^{>k}$ bounds in Figures \ref{fig:FMNIST-roc-llr} and \ref{fig:CIFAR10-roc-llr}, respectively.
On both datasets we see significantly better discriminatory performance when using the $LLR^{>k}$ score.

\cref{tab:rocauc-ood} shows that BIVA improves upon the HVAE model for OOD detection on CIFAR while \cref{tab:bits-per-dim-ood} shows that the BIVA model also improves upon the HVAE in terms of likelihood.
We hypothesize that models larger than our implementation of BIVA, with better likelihood scores may perform even better \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.

\subsection{Comparison to baselines}
\paragraph{Performance} \cref{tab:rocauc-ood} summarize our results compared to baselines based on the commonly used AUROC$\uparrow$, AUPRC$\uparrow$ and FPR80$\downarrow$ metrics.
Our method outperforms other generative model-based methods such as WAIC \cite{choi_waic_2019} with Glow model and performs similarly to the likelihood regret method of \cite{xiao_likelihood_2020}.
Furthermore, our method performs similarly to the background constrative likelihood ratio method of \textcite{ren_likelihood_2019} on FashionMNIST/MNIST but contrary to the failure of that method on CIFAR10/SVHN reported by \cite{xiao_likelihood_2020}, our method performs very well on this task too.
Our approach outperforms all supervised approaches that use in-distribution labels or synthetic examples of OOD data derived from the in-distribution data including ODIN \cite{liang_enhancing_2018} and the predictive distribution of a classifier $p(\hat{y}|\x)$ trained and evaluated in various ways (see \textcite{ren_likelihood_2019}).

\paragraph{Runtime} For a full evaluation of a single example across all feature levels of a model with $L$ stochastic layers, our method requires $L-1$ forward passes through the inference and generative networks as well as computing the likelihood ratio, of which the forward passes are dominant.
For a typical forward pass that is linear in the input dimensionality, $D$, and the number of stochastic layers, $L$, this amounts to computation of $\mathcal{O}(DL)$.
Compared to some related work that either requires an $M>1$ sized batch of inputs of which either all or none are OOD \cite{nalisnick_detecting_2019} or cannot be applied to batches due to the required per-example optimization \cite{xiao_likelihood_2020}, our method additionally is applicable to batches of any size that may consist of both OOD and in-distribution examples which provides drastic speed-ups via vectorization and parallelization.
Furthermore, the method of \textcite{xiao_fashionmnist_2017} requires refitting the inference network of a VAE which can be computationally demanding.
Compared to the likelihood ratio proposed in \textcite{ren_likelihood_2019}, our method requires training only a single model on a single dataset.

\section{Discussion}
Deep generative models are state-of-the-art density estimators, but the OOD failures reported in recent years have raised concerns about the limitations of such density estimates. Recent work on improving OOD detection has largely sidestepped this concern by relying on additional assumptions that strictly should not be needed for models with explicit likelihoods.
While the engineering challenge of building reliable OOD detection schemes is important, it is of more fundamental importance to understand \emph{why} the naive likelihood test fails.
We have provided evidence that low-level features of the neural nets dominate the likelihood, which gives a \emph{cause} to the \emph{why}.
The fact that a simple score for measuring the importance of semantic features yield state-of-the-art results on OOD detection without access to additional information gives validity to our hypothesis.

The findings from, amongst others, \textcite{nalisnick_deep_2019, serra_input_2020} have a clear relation to information theory and compression. 
Semantically complex in-distribution data yields models with diverse low-level feature sets that enable generalization across datasets.
Simpler datasets can only yield models with less diverse low-level feature sets compared to complex training data.
Hence, there can be an asymmetry where the likelihoods of simple OOD data can be high for a model trained on complex data, but not the other way around.
Loosely put, the minimal number of bits required to losslessly compress data sampled from some distribution is the entropy of the generating process \cite{shannon_mathematical_1948, mackay_information_2003}.
\textcite{townsend_practical_2019} recently showed that VAEs can be used for lossless compression at rates superior to more generic algorithms.

We also note that since the hierarchical VAE is a probabilistic graphical latent variable model, it lends itself very naturally to manipulation at the feature level \parencite{kingma_semi-supervised_2014, maaloe_auxiliary_2016, maaloe_semi-supervised_2017}.
This property sets it apart from other generative models that do not explicitly define such a hierarchy of features.
This in turn enables reliable OOD detection with our methodology while making no explicit assumptions about the nature of OOD data and only using a single model. This has not been achieved with autoregressive or flow-based models.

\section{Conclusion}
In this paper we study unsupervised out-of-distribution detection using hierarchical variational autoencoders.
We provide evidence that highly generalizeable low-level features contribute greatly to estimated likelihoods resulting in poor OOD detection performance.
We proceed to develop a likelihood-ratio based score for OOD detection and define it to explicitly ensure that data must be in-distribution across all feature levels to be regarded in-distribution.
This ratio is mathematically shown to perform OOD detection in the latent space of the model, removing the reliance on the troublesome input-space likelihood.
We point out that contrary to much recent literature on OOD detection, our approach is fully unsupervised and does not make assumptions about the nature of OOD data.
Finally, we demonstrate state-of-the-art performance on a wide range of OOD failure cases.

\section*{Acknowledgements}
This research was partially funded by the Innovation Fund Denmark via the Industrial PhD Programme (grant no.\@ 0153-00167B). JF and SH were funded in part by the Novo Nordisk Foundation (grant no.\@ NNF20OC0062606) via the Center for Basic Machine Learning Research in Life Science (MLLS, \hyperlink{https://www.mlls.dk}{https://www.mlls.dk}). JF was further funded by the Novo Nordisk Foundation (grant no.\@ NNF20OC0065611) and the Independent Research Fund Denmark (grant no.\@ 9131-00082B). SH was further funded by VILLUM FONDEN (15334) and the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement no. 757360).

### 2023-02-24 frontmatter/notation.tex
Sync-up

@@
 %\markboth{\MakeUppercase\nomname}{\MakeUppercase\nomname}
+
 % Correct header on multiple page nomenclature
@@
   {}{}
-% Rename Nomenclature to Notation
-\renewcommand{\nomname}{Notation}
-% Increase width of box for mathematical symbol on left
-\setlength{\nomlabelwidth}{2.5cm}
+
+\renewcommand{\nomname}{Notation}  % Rename Nomenclature to Notation
+
+\setlength{\nomlabelwidth}{3cm}  % Increase width of box for mathematical symbol on left
 % \setlength{\nomitemsep}{-\parsep}
@@
     \bigskip
-    \item[\bfseries \hspace{2.5cm}\: #1
+    \item[\bfseries \hspace{3cm}\: #1
         % \vspace{1cm}\\
@@
 \nomenclature[E, 01]{$\frac{\text{d}y}{\text{d}x}$}{Derivative of $y$ with respect to $x$}
-\nomenclature[E, 02]{$\pderiv{y}{x}$}{Partial derivative of $y$ with respect to $x$}
-\nomenclature[E, 03]{$\nabla_\x$}{Del, nabla or Laplacian operator, $\nabla_\x = \bmat{\pderiv{}{x_1} & \cdots & \pderiv{}{x_n}}^\text{T}, \; \x\in\mathbb{R}^{n}$}
-\nomenclature[E, 04]{$\nabla_\x y$}{Gradient of $y$ with respect to $\x$, $\nabla_\x y = \bmat{\pderiv{y}{x_1} & \cdots & \pderiv{y}{x_n}}^\text{T}, \; \x\in\mathbb{R}^{n}$}
+\nomenclature[E, 02]{$\pd{y}{x}$}{Partial derivative of $y$ with respect to $x$}
+\nomenclature[E, 03]{$\nabla_\x$}{Del, nabla or Laplacian operator, $\nabla_\x = \bmat{\pd{}{x_1} & \cdots & \pd{}{x_n}}^\text{T}, \; \x\in\mathbb{R}^{n}$}
+\nomenclature[E, 04]{$\nabla_\x y$}{Gradient of $y$ with respect to $\x$, $\nabla_\x y = \bmat{\pd{y}{x_1} & \cdots & \pd{y}{x_n}}^\text{T}, \; \x\in\mathbb{R}^{n}$}
 \nomenclature[E, 05]{$\H_\x$}{Hessian operator, $\H_\x = \nabla_\x\nabla_\x\transpose = \bmat{
-	    \pderiv[2]{}{x_1} & \pderiv[1,1]{}{x_1,x_2} & \cdots & \pderiv[1,1]{}{x_1,x_n}\\
-	    \pderiv[1,1]{}{x_2,x_1} & \pderiv[2]{}{x_2} & \cdots & \pderiv[1,1]{}{x_2,x_n}\\
+	    \pd[2]{}{x_1} & \frac{\partial^2}{\partial x_1 \partial x_2} & \cdots & \frac{\partial^2}{\partial x_1 \partial x_n}\\
+	    \frac{\partial^2}{\partial x_2 \partial x_1} & \pd[2]{}{x_2} & \cdots & \frac{\partial^2}{\partial x_2 \partial x_n}\\
 	    \vdots & \vdots & \ddots & \vdots\\
-	    \pderiv[1,1]{}{x_n,x_1} & \pderiv[1,1]{}{x_d,x_2} & \cdots & \pderiv[2]{}{x_n}
+	    \frac{\partial^2}{\partial x_n \partial x_1} & \frac{\partial^2}{\partial x_n \partial x_2} & \cdots & \pd[2]{}{x_n}
 	    }$}
 \nomenclature[E, 06]{$\H_\x y$}{Hessian matrix of $y$ with respect to $\x$; matrix of all second order partial derivatives}
-\nomenclature[E, 07]{$\nabla^2_\x$}{Laplacian operator, $\nabla^2_\x=\nabla_\x\transpose\nabla_\x = \nabla_\x\cdot\nabla_\x=\sum_i \pderiv[2]{}{x_i}=\Tr{\H_\x}$}
+\nomenclature[E, 07]{$\nabla^2_\x$}{Laplacian operator, $\nabla^2_\x=\nabla_\x\transpose\nabla_\x = \nabla_\x\cdot\nabla_\x=\sum_i \pd[2]{}{x_i}=\Tr{\H_\x}$}
 \nomenclature[E, 08]{$\nabla^2_\x y$}{Laplacian of $y$ with respect to $\x$; sum of the unmixed second order partial derivatives}
-\nomenclature[E, 09]{$\pderiv{\y}{\x}$}{Jacobian matrix of $\y$ w.r.t. $\x$, $\pderiv{\y}{\x} = \bmat{
-	    \nabla_\x^\text{T} y_1\\
-	    \vdots \\
-	    \nabla_\x^\text{T} y_n}, \; \y\in\mathbb{R}^{n}$; matrix of all possible first order partial derivatives}
+\nomenclature[E, 09]{$\pd{\y}{\x}$}{Jacobian matrix of $\y$ w.r.t. $\x$, $\pd{\y}{\x} = \bmat{
+	    \nabla_\x y_1 &
+	    \cdots &
+	    \nabla_x y_n}^\text{T}, \; \y\in\mathbb{R}^{n}$; matrix of all possible first order partial derivatives}
 \nomenclature[E, 10]{$\int f(\x)\,\text{d}\x$}{Definite integral over the domain of $\x$}

### 2023-02-25 chapters/paper_benchmarking.tex
Add benchmarking paper

%!TEX root = ../thesis.tex

\chapter[benchmarking generative latent variable models for speech]{Benchmarking Generative Latent Variable Models for Speech}
\label{chp:benchmarking}

\section{Abstract}
Stochastic latent variable models (LVMs) achieve state-of-the-art performance on natural image generation but are still inferior to deterministic models on speech. In this paper, we develop a speech benchmark of popular temporal LVMs and compare them against state-of-the-art deterministic models. We report the likelihood, which is a much used metric in the image domain, but rarely, or incomparably, reported for speech models. To assess the quality of the learned representations, we also compare their usefulness for phoneme recognition. Finally, we adapt the Clockwork VAE, a state-of-the-art temporal LVM for video generation, to the speech domain. Despite being autoregressive only in latent space, we find that the Clockwork VAE can outperform previous LVMs and reduce the gap to deterministic models by using a hierarchy of latent variables.

\section{Introduction}
After the introduction of the variational autoencoder (VAE,  \textcite{kingma_autoencoding_2014, rezende_stochastic_2014}) quickly came two temporal extensions for modeling speech data \cite{chung_recurrent_2015, fraccaro_sequential_2016}. Since then, temporal LVMs have undergone little development compared to their counterparts in the image domain, where LVMs recently showed superior performance to autoregressive models such as PixelCNN \cite{oord_pixel_2016, oord_conditional_2016, salimans_pixelcnn_2017}. The improvements in the image domain have been driven mainly by top-down inference models and deeper latent hierarchies \cite{sonderby_ladder_2016, maaloe_biva_2019, vahdat_nvae_2020, child_very_2021,sinha_consistency_2021,kingma_variational_2021}. In speech modeling however, autoregressive models such as the WaveNet remain state-of-the-art \cite{oord_wavenet_2016}.

To compare and develop LVMs for speech, we need good benchmarks similar to those in the image domain. Image benchmarks commonly compare likelihood scores, but research in the speech domain often omits reporting a likelihood \cite{oord_wavenet_2016, hsu_unsupervised_2017, oord_neural_2018} or report likelihoods that are incomparable due to subtle differences in the assumed data distribution \cite{chung_recurrent_2015, fraccaro_sequential_2016, hsu_unsupervised_2017, aksan_stcn_2019}. Without a proper standard, it is difficult to compare explicit likelihood models for speech and develop them in an informed manner.

To advance the state of LVMs for speech, this paper (i) develops a benchmark for LVMs based on model likelihood, (ii) introduces a hierarchical LVM architecture without autoregressive decoder, (iii) compares LVMs to deterministic counterparts including WaveNet, and (iv) qualitatively and quantitatively evaluates the latent variables learned by different LVMs based on their usefulness for phoneme recognition. We find that:
\begin{itemize}[leftmargin=8mm]
    \vspace{-3mm}
    \setlength\itemsep{-0em}
    \item [(I)] State-of-the-art LVMs achieve likelihoods that are inferior to WaveNet at high temporal resolution but are superior at lower resolutions. Interestingly, we find that a standard LSTM \cite{hochreiter_long_1997} almost matches the likelihood of WaveNet.
    \item [(II)] LVMs with powerful autoregressive decoders achieve better likelihoods than the non-autoregressive LVM.
    \item[(III)] The expressiveness of LVMs for speech increases with a deeper hierarchy of stochastic latent variables, similar to conclusions within image modeling. % \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
    \item[(IV)] LVMs learn rich representations that are as good or better than Mel spectrograms for phoneme recognition also when using only 10 minutes of labeled data.
\end{itemize}\vspace{-3mm}
At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models
\footnote{\href{https://github.com/JakobHavtorn/benchmarking-lvms}{\url{github.com/JakobHavtorn/benchmarking-lvms}}}.
% \footnote{\fontsize{8.5}{0}\selectfont\texttt{ \url{github.com/anonymous/available-later}}}.
Before presenting the results, we provide a brief survey of existing LVMs for speech in a coherent notation.

\section{Latent variable models for speech}
\begin{figure*}[t]
    \centering
    \begin{multicols}{5}
        \resizebox{!}{0.18\paperheight}{
            \begin{tikzpicture}
                \node[obs] (x_t) {$\xb_t$};

                % \node[above=0.75cm of x_t, xshift=0cm] (cap1) {$p(\xb_t|\xb_{<t})$};
                \node[above=0.4cm of x_t] (caption) {LSTM};

                \node[det,below=2.55cm of x_t] (d_t) {$\db_{t}$};
                \node[det,left=0.75cm of d_t] (d_tm1) {$\db_{t-1}$};
                \node[obs,below=0.75cm of d_t] (x_tm1) {$\xb_{t-1}$};

                \edge[deterministic-skip-color]{x_tm1}{d_t};
                \edge[deterministic-skip-color]{d_t}{x_t};
                \edge[deterministic-skip-color]{d_tm1}{d_t};
                \edge[draw=white,bend right]{d_t}{x_t};  % for identical width to others
            \end{tikzpicture}
        }
        \\
        \resizebox{!}{0.18\paperheight}{
            \begin{tikzpicture}
                \node[obs] (x_t) {$\xb_t$};

                % \node[above=0.75cm of x_t, xshift=0cm] (cap1) {$p(\xb_t|\zb_{\leq t},\xb_{<t}) p(\zb_t|\xb_{<t},\zb_{<t})$};
                \node[above=0.4cm of x_t] (caption) {VRNN};

                \node[latent,below=0.75cm of x_t] (z_t) {$\zb_t$};
                \node[latent,left=0.75cm of z_t] (z_tm1) {$\zb_{t-1}$};

                \node[det,below=0.75cm of z_t] (d_t) {$ b_{t}$};
                \node[det,below=0.75cm of z_tm1] (d_tm1) {$\db_{t-1}$};
                \node[obs,below=0.75cm of d_t] (x_tm1) {$\xb_{t-1}$};

                \edge[deterministic-skip-color]{x_tm1}{d_t};
                \edge[]{d_t}{z_t};
                \edge[]{z_tm1}{d_t};
                \edge[deterministic-skip-color]{d_tm1}{d_t};
                \edge[]{z_t}{x_t};
                \edge[deterministic-skip-color,bend right]{d_t}{x_t};
            \end{tikzpicture}
        }
        \\
        \resizebox{!}{0.18\paperheight}{
            \begin{tikzpicture}
                \node[obs] (x_t) {$\xb_t$};

                % \node[above=0.75cm of x_t, xshift=0cm] (cap1) {$p(\xb_t|\zb_{\leq t},\xb_{<t}) p(\zb_t|\xb_{<t},\zb_{<t})$};
                \node[above=0.4cm of x_t] (caption) {SRNN};

                \node[latent,below=0.75cm of x_t] (z_t) {$\zb_t$};
                \node[latent,left=0.75cm of z_t] (z_tm1) {$\zb_{t-1}$};

                \node[det,below=0.75cm of z_t] (d_t) {$\db_{t}$};
                \node[det,below=0.75cm of z_tm1] (d_tm1) {$\db_{t-1}$};
                \node[obs,below=0.75cm of d_t] (x_tm1) {$\xb_{t-1}$};

                \edge[deterministic-skip-color]{x_tm1}{d_t};
                \edge[]{d_t}{z_t};
                \edge[]{z_tm1}{z_t};
                \edge[deterministic-skip-color]{d_tm1}{d_t};
                \edge[]{z_t}{x_t};
                \edge[deterministic-skip-color,bend right]{d_t}{x_t};
            \end{tikzpicture}
        }
        \\
        \resizebox{!}{0.18\paperheight}{
            \begin{tikzpicture}
                \node[obs] (x_t) {$\xb_t$};
                % \node[above=0.75cm of x_t, xshift=0cm] (cap1) {$p(\xb_t|\zb_t,\xb_{t-r:t-1})$};
                \node[above=0.4cm of x_t] (caption) {STCN};

                \node[latent,below=0.75cm of x_t] (z_t) {$\zb_t$};
                \node[left=0.025cm of z_t] (dots) {$\cdots$};
                \node[latent,left=0.75cm of z_t] (z_tmr) {$\zb_{t-r'}$};

                \node[det,below=0.75cm of z_t] (d_t) {$\db_{t}$};
                \node[obs,below=0.75cm of d_t] (x_tm1) {$\xb_{t-1}$};
                \node[left=0.025cm of x_tm1] (dots) {$\cdots$};
                \node[obs,left=0.75cm of x_tm1] (x_tmr) {$\xb_{t-r}$};

                \edge[]{x_tm1}{d_t};
                \edge[]{x_tmr}{d_t};
                \edge[]{d_t}{z_t};
                \edge[]{z_t}{x_t};
                \edge[]{z_tmr}{x_t};
                \edge[draw=white,bend right]{d_t}{x_t};  % for identical width to others 
            \end{tikzpicture}
        }
        \\
        \resizebox{!}{0.18\paperheight}{
            \begin{tikzpicture}
                \node[obs] (x_t) {$\xb_t$};%

                % \node[above=0.75cm of x_t, xshift=0cm] (cap1) {$p(\xb_t|\zb_{\leq t}) p(\zb_t|\zb_{<t})$};
                \node[above=0.4cm of x_t] (caption) {CW-VAE};

                \node[latent,below=0.75cm of x_t] (z_t) {$\zb_t$};
                \node[latent,left=0.75cm of z_t] (z_tm1) {$\zb_{t-1}$};

                \node[det,below=0.75cm of z_t] (d_t) {$\db_{t}$};
                \node[det,below=0.75cm of z_tm1] (d_tm1) {$\db_{t-1}$};
                \node[obs,below=0.75cm of d_t,draw=white,fill=white] (x_tm1) {};

                \edge[]{d_t}{z_t};
                \edge[bend right]{d_t}{x_t};
                \edge[]{z_tm1}{d_t};
                \edge[]{d_tm1}{d_t};
                \edge[]{z_t}{x_t};
            \end{tikzpicture}
        }
        \end{multicols}
    \caption{
        Generative models for a single time step of a deterministic autoregressive LSTM, the VRNN and SRNN as well as the STCN and CW-VAE both with a single layer of latent variables. 
        Red arrows indicate purely deterministic paths from the output $\xb_t$ to previous input $\xb_{<t}$ without passing a stochastic node. 
        The models differ in their use of latent variables and dependencies especially autoregressive dependencies and skip connections. 
        % In an LSTM, information flows deterministically from previous observed variables to the next. 
        % The VRNN and SRNN use stochastic latent variables but also include deterministic skip connections from previous observed variables. 
        % The STCN is also autoregressive in $\xb$ but does not use deterministic skip connections from $\xb_{t-1}$ to $\xb_t$. 
        % The CW-VAE is not autoregressive on the observed variable which forces information to flow through the latent variables. 
        We provide more elaborate graphical illustrations including inference models in \cref{fig: clockwork vae inference generative models 2 layered} and appendix~\cref{app: additional graphical models}.
        }
    \label{fig: autoregressive vrnn srnn cwvae 1L graphs}
\end{figure*}

% \linepenalty=1000
% \subsection{Related work}
LVMs formulated as VAEs continue to be of interest since they are able to learn an approximation to the posterior distribution of assumed latent variables. The posterior is usually of a reduced dimensionality compared to the input and lies close to a known prior distribution. Approximate posteriors have various applications e.g. semi-supervised learning \cite{kingma_semi-supervised_2014} and anomaly detection \cite{havtorn_hierarchical_2021}.

In recent years, several complementary methods have been proposed to improve the expressiveness of VAEs. These include building more expressive priors via methods such as normalizing flows \cite{rezende_variational_2015} and building a deeper hierarchy of stochastic latent variables such as the Ladder VAE \cite{sonderby_ladder_2016}. In this research, we focus on the latter due to the recent breakthroughs in image modeling using VAEs without costly autoregressive dependencies on the observed variable \cite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.

Several works have applied LVMs to speech. Among the first contributions were the VRNN \cite{chung_recurrent_2015} and SRNN \cite{fraccaro_sequential_2016} which can be seen as conditional VAEs per timestep. Other recent LVMs include the FH-VAE \cite{hsu_unsupervised_2017}, which leverages an additional latent variable to capture global features, and Z-forcing \cite{goyal_z-forcing_2017}, which resembles the SRNN but includes an auxiliary task in the latent space to increase its utilization. The VQ-VAE \cite{oord_neural_2018} is a hybrid between an LVM and an autoregressive model which uses a quantized latent space to improve the quality of generated samples. The Stochastic WaveNet \cite{lai_stochastic_2018} and STCN \cite{aksan_stcn_2019} use WaveNet encoder and decoders and temporally independent latent variables.

In this paper, we focus on the VRNN, SRNN and STCN. We exclude the Stochastic WaveNet as it is similar to STCN and achieves inferior likelihoods \cite{aksan_stcn_2019}. The FH-VAE, with disjoint latent variables and discriminative objective, Z-forcing, with an auxiliary task, and the VQ-VAE, with a quantized latent space and autoregressive prior fitted after training, all introduce significant changes to the original VAE framework and are also not included here.

All selected models have autoregressive generative models which let future observed variables be generated by conditioning on previously generated values. Inspired by recent progress in the image domain, we therefore formulate and benchmark a novel temporal LVM which does not rely on an autoregressive decoder. We do so by adapting the hierarchical Clockwork Variational Autoencoder \cite{saxena_clockwork_2021}, originally proposed for video generation, to speech.

\subsection{Sequential deep latent variable models}
The selected models are all sequential deep latent variable models trained with variational inference and the reparameterization trick \cite{kingma_auto-encoding_2014}.
% The VRNN, SRNN, STCN and CW-VAE are all autoencoders trained with variational inference and the reparameterization trick \cite{kingma_auto-encoding_2014}. 
They take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$. 
% The VRNN, SRNN, STCN and CW-VAE are all VAEs \cite{kingma_auto-encoding_2014} that take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$. 
We let $\xb_{1:T}$ refer to the observed variable or a downsampled version of it. We will sometimes use $\xb$ to refer to the sequence $\xb_{1:T}$ when it is not ambiguous.
% In general, $\xb_{1:T}$ may refer to the original observed variable or a deterministically downsampled representation of the observed variable. We will sometimes use $\xb$ to refer to the full sequence $\xb_{1:T}$ when it is not ambiguous.

First, $\xb_{1:T}$ is encoded to a temporal stochastic latent representation $\zb_{1:T} = \left( \zb_1, \zb_2, \dots, \zb_{T} \right)$ with $\zb_t\in \mathbb{R}^{D_z}$. This representation is then used to reconstruct the original input $\xb_{1:T}$.
The latent variable is assumed to follow some prior distribution $p(\zb_t|\cdot)$ where the dot indicates that it may depend on latent and observed variables at previous time steps, $\zb_{<t}$ and $\xb_{<t}$ where $\zb_{<t} := \left( \zb_1, \zb_2, \dots, \zb_{t-1} \right)$.

The models are trained to maximize a likelihood objective. The exact marginal likelihood $\log p_\thetab(\xb)$, where $\thetab$ are parameters of the generative model, is intractable to optimize due to the integration over the latent space. Instead, we introduce the variational approximation $q_\phib(\zb|\xb)$ to the true posterior. Via Jensen's inequality this yields the well-known evidence lower bound (ELBO) on the exact likelihood $\mathcal{L}(\thetab,\phib;\xb)$
which can be jointly optimized with respect to  $\{\thetab, \phib\}$ using stochastic gradient descent methods. We omit the $\thetab$ and $\phib$ subscripts for the remainder of the paper.
\begin{equation}
    \log p_\thetab(\xb) = \log \int p_\thetab(\xb,\zb) \, \mathrm{d}\zb  \geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] =: \mathcal{L}(\thetab,\phib;\xb) \label{eq: elbo} \enspace,
\end{equation}
Graphical illustrations of the models can be seen in \cref{fig: autoregressive vrnn srnn cwvae 1L graphs} with more illustrations in appendix~\cref{app: additional graphical models}.

\subsection{Variational recurrent neural network ({\small VRNN})}
The VRNN \cite{chung_recurrent_2015} is essentially a VAE per timestep. At timestep $t$, the VAE is conditioned on the hidden state of a recurrent neural network (RNN), $\db_{t}\in\mathbb{R}^{D_d}$, with state transition $\db_{t} = f([\xb_{t-1}, \zb_{t-1}], \db_{t-1})$ where $[\cdot, \cdot]$ denotes concatenation. 
The VRNN uses a Gated Recurrent Unit (GRU, \textcite{cho_properties_2014}) for $f$.
The joint distribution factorizes over time and the latent variables are autoregressive in both the observed and latent space,
\begin{equation}
    p(\xb_{1:T},\zb_{1:T}) = \prod_{t=1}^{T} p(\xb_t|\xb_{<t},\zb_{\leq t}) p(\zb_t|\xb_{<t},\zb_{<t}) \enspace . \label{eq: vrnn joint factorization}
\end{equation}
The approximate posterior similarly factorizes over time,
\begin{equation}
    q(\zb_{1:T}|\xb_{1:T}) = \prod_{t=1}^{T} q(\zb_t|\xb_{\leq t},\zb_{<t}) \enspace . \label{eq: vrnn inference factorization}
\end{equation}
From this, the ELBO for the VRNN is
\begin{equation}
    \mathcal{L}(\xb) = ~ \mathbb{E}_{q(\zb_{1:T}|\xb_{1:T})} \bigg[ \sum_{t} \log p(\xb_t|\xb_{<t},\zb_{\leq t}) - D_\text{KL}\left( q(\zb_t|\xb_{\leq t},\zb_{<t}) \, || \, p(\zb_t|\xb_{<t},\zb_{<t}) \right)  \bigg] \label{eq: vrnn elbo} \enspace .
\end{equation}
% \begin{equation}
% \begin{split}
%     \mathcal{L}(\xb) =&~ \mathbb{E}_{q(\zb_{1:T}|\xb_{1:T})} \bigg[ \sum_{t} \log p(\xb_t|\xb_{<t},\zb_{\leq t}) \\ & - \text{KL}\left( q(\zb_t|\xb_{\leq t},\zb_{<t}) \, || \, p(\zb_t|\xb_{<t},\zb_{<t}) \right)  \bigg] \enspace .
%     \label{eq: vrnn elbo}
% \end{split}
% \end{equation}
The VRNN uses diagonal covariance Gaussian distributions $\mathcal{N}$ for the prior and posterior distributions. We denote the output distribution of choice by $\mathcal{D}$. 
% \begin{align}
%     q(\zb_t|\xb_{\leq t},\zb_{<t}) &= \mathcal{N}\left(\alphab_{q}(\xb_{t},\db_{t})\right) \label{eq: vrnn parameterization inference} \\
%     p(\zb_t|\xb_{<t},\zb_{<t}) &= \mathcal{N}\left(\alphab_{p}(\db_t)\right) \label{eq: vrnn parameterization prior} \\
%     p(\xb_t|\xb_{<t},\zb_{\leq t}) &= \mathcal{D}\left(\betab(\zb_{t},\db_{t})\right) \enspace . \label{eq: vrnn parameterization observation}
% \end{align}
\begin{gather}
    p(\zb_t|\xb_{<t},\zb_{<t}) = \mathcal{N}\left(\alphab_{p}(\db_t)\right) \enspace , \nonumber \\
    p(\xb_t|\xb_{<t},\zb_{\leq t}) = \mathcal{D}\left(\betab(\zb_{t},\db_{t})\right) \enspace , \label{eq: vrnn parameterization generative} \\
    q(\zb_t|\xb_{\leq t},\zb_{<t}) = \mathcal{N}\left(\alphab_{q}(\xb_{t},\db_{t})\right) \enspace . \nonumber 
    \label{eq: vrnn parameterization inference}
\end{gather}
All sets of distributional parameters, $\alphab_q,\alphab_p$ and $\betab$, are the outputs of densely connected neural networks which we notationally overload as functions in equations~\cref{eq: vrnn parameterization generative} and \cref{eq: vrnn parameterization inference}. It is common to refer to $\alphab_q$ as the inference model or encoder and $\betab$ as the decoder. Together with $\betab$, the structural model $\alphab_p$ forms the generative model.

Since the decoder is dependent on $\db_t$, the transition function $f$ allows the VRNN to learn to ignore parts of or the entire latent variable and establish a purely deterministic transition from $\xb_{t-1}$ to $\db_t$ (\cref{fig: autoregressive vrnn srnn cwvae 1L graphs}). This failure mode is commonly referred to as posterior collapse and is a well-known weakness of VAEs with powerful decoders \cite{bowman_generating_2016, sonderby_ladder_2016}.

\subsection{Stochastic recurrent neural network ({\small SRNN})}
The SRNN \cite{fraccaro_sequential_2016} is similar to the VRNN but differs by separating the stochastic latent variables from the deterministic representations (\cref{fig: autoregressive vrnn srnn cwvae 1L graphs}). That is, the GRU state transition is independent of $\zb_{1:T}$ such that $\db_{t} = f(\xb_{t-1}, \db_{t-1})$. With this, the joint $p(\xb_{1:T},\zb_{1:T})$ can be written as for the VRNN in \cref{eq: vrnn joint factorization}. The approximate posterior of $\zb_t$ is conditioned on the full observed sequence,
\begin{align}
    \label{eq: srnn inference factorization}
    q(\zb_{1:T}|\xb_{1:T}) = \prod_{t=1}^{T} q(\zb_t|\xb_{1:T},\zb_{t-1}) \enspace .
\end{align}
This is achieved by introducing a second GRU that runs backwards in time with transition $\ab_t = g([\xb_t,\db_t], \ab_{t+1})$. While $p(\xb_t|\xb_{<t},\zb_{\leq t})$ remains as in \cref{eq: vrnn parameterization generative}, we have
\begin{gather}
        p(\zb_t|\xb_{<t},\zb_{<t}) = \mathcal{N}\left(\alphab_{p}(\zb_{t-1},\db_t)\right) \enspace , \nonumber \\
        q(\zb_t|\xb_{1:T},\zb_{<t}) = \mathcal{N}\left(\alphab_{q}(\zb_{t-1},\ab_{t})\right) \enspace . \label{eq: srnn parameterization}
    % p(\xb_t|\xb_{<t},\zb_{\leq t}) &= \mathcal{D}\left(\vrho(\zb_{t},\db_{t})\right) \enspace . \label{eq: srnn parameterization observation}
\end{gather}
By inferring $\zb_t$ conditioned on the full sequence $\xb_{1:T}$, the SRNN performs smoothing. This has been noted to better approximate the true posterior of $\zb_t$ which can be shown to depend on the full observed sequence \cite{bayer_mind_2021}. Comparatively, the VRNN performs filtering.

\begin{figure*}[t]
    \centering
    % \h3ill
    \resizebox{!}{0.155\paperheight}{
        \begin{tikzpicture}
            % nodes
            \node[obs] (x_1) {$\xb_{t-3}$};
            \node[obs,right=.75cm of x_1] (x_2) {$\xb_{t-2}$};
            \node[obs,right=.75cm of x_2] (x_3) {$\xb_{t-1}$};
            \node[obs,right=.75cm of x_3] (x_4) {$\xb_{t}$};
            \node[obs,right=.75cm of x_4,fill=white,draw=white] (x_5) {$\dots$};

            \node[latent,above=.75cm of x_1](s_1_1){$\zb^{(1)}_{t-3}$};
            \node[latent,above=.75cm of x_2](s_1_2){$\zb^{(1)}_{t-2}$};
            \node[latent,above=.75cm of x_3](s_1_3){$\zb^{(1)}_{t-1}$};
            \node[latent,above=.75cm of x_4](s_1_4){$\zb^{(1)}_{t}$};
            \node[latent,above=.75cm of x_5,fill=white,draw=white](s_1_5){$\dots$};

            \node[latent,above=.75cm of s_1_1](s_2_1){$\zb^{(2)}_{t-3}$};
            \node[latent,above=.75cm of s_1_3](s_2_3){$\zb^{(2)}_{t-1}$};
            \node[latent,above=.75cm of s_1_5,fill=white,draw=white](s_2_5){$\dots$};

            % x to s first layer
            \edge[]{x_1}{s_1_1};
            \edge[]{x_2}{s_1_2};
            \edge[]{x_3}{s_1_3};
            \edge[]{x_4}{s_1_4};

            % x to s above layer
            \edge[bend left=40]{x_1}{s_2_1};
            \edge[]{x_2}{s_2_1};
            \edge[bend left=40]{x_3}{s_2_3};
            \edge[]{x_4}{s_2_3};

            % s per layer forward in time
            \edge[shared-function-color]{s_1_1}{s_1_2};
            \edge[shared-function-color]{s_1_2}{s_1_3};
            \edge[shared-function-color]{s_1_3}{s_1_4};
            \edge[shared-function-color]{s_1_4}{s_1_5};

            \edge[shared-function-color]{s_2_1}{s_2_3};
            \edge[shared-function-color]{s_2_3}{s_2_5};

            % s across layers on time
            \edge[shared-function-color]{s_2_1}{s_1_1};
            \edge[shared-function-color]{s_2_3}{s_1_3};

            % s across layers forward in time
            \edge[shared-function-color]{s_2_1}{s_1_2};
            \edge[shared-function-color]{s_2_3}{s_1_4};

            % \node[above=of s_2_1, yshift=-1.cm, xshift=2cm] (phi) {$q\pa{\zb_{1:T}^{(1:L)}|\xb_{1:T}}$};
            \node[above=of s_2_1, yshift=-0.5cm, xshift=2.5cm] (phi) {CW-VAE inference};
        \end{tikzpicture}
    }
    \hfill
    \resizebox{!}{0.155\paperheight}{
        \begin{tikzpicture}
            % nodes
            \node[obs] (x_1) {$\xb_{t-3}$};
            \node[obs,right=.75cm of x_1] (x_2) {$\xb_{t-2}$};
            \node[obs,right=.75cm of x_2] (x_3) {$\xb_{t-1}$};
            \node[obs,right=.75cm of x_3] (x_4) {$\xb_{t}$};
            \node[obs,right=.75cm of x_4,fill=white,draw=white] (x_5) {$\dots$};

            \node[latent,above=.75cm of x_1](s_1_1){$\zb^{(1)}_{t-3}$};
            \node[latent,above=.75cm of x_2](s_1_2){$\zb^{(1)}_{t-2}$};
            \node[latent,above=.75cm of x_3](s_1_3){$\zb^{(1)}_{t-1}$};
            \node[latent,above=.75cm of x_4](s_1_4){$\zb^{(1)}_{t}$};
            \node[latent,above=.75cm of x_5,fill=white,draw=white](s_1_5){$\dots$};

            \node[latent,above=.75cm of s_1_1](s_2_1){$\zb^{(2)}_{t-3}$};
            \node[latent,above=.75cm of s_1_3](s_2_3){$\zb^{(2)}_{t-1}$};
            \node[latent,above=.75cm of s_1_5,fill=white,draw=white](s_2_5){$\dots$};

            % x to s first layer
            \edge[]{s_1_1}{x_1};
            \edge[]{s_1_2}{x_2};
            \edge[]{s_1_3}{x_3};
            \edge[]{s_1_4}{x_4};

            % s per layer forward in time
            \edge[shared-function-color]{s_1_1}{s_1_2};
            \edge[shared-function-color]{s_1_2}{s_1_3};
            \edge[shared-function-color]{s_1_3}{s_1_4};
            \edge[shared-function-color]{s_1_4}{s_1_5};

            \edge[shared-function-color]{s_2_1}{s_2_3};
            \edge[shared-function-color]{s_2_3}{s_2_5};

            % s across layers on time
            \edge[shared-function-color]{s_2_1}{s_1_1};
            \edge[shared-function-color]{s_2_3}{s_1_3};

            % s across layers forward in time
            \edge[shared-function-color]{s_2_1}{s_1_2};
            \edge[shared-function-color]{s_2_3}{s_1_4};

            % \node[above=of s_2_1, yshift=-1.cm, xshift=2cm] (phi) {$p\pa{\xb_{1:T},\zb_{1:T}^{(1:L)}}$};
            \node[above=of s_2_1, yshift=-0.5cm, xshift=2.5cm] (phi) {CW-VAE generative};
        \end{tikzpicture}
    % \hfill
    }
\caption{
Inference (left) and generative (right) models for the CW-VAE with a hierarchy of $L=2$ latent variables, $s_1=1$ and $s_2=2$. 
The models are unrolled over four consecutive time steps but note that the graph continues towards $t=0$ and $t=T$. 
Blue arrows indicate parameter sharing between inference and generative models. 
We omit the deterministic variable of \cref{fig: autoregressive vrnn srnn cwvae 1L graphs} for clarity.
}
\label{fig: clockwork vae inference generative models 2 layered}
\vspace{-0mm}
\end{figure*}

\subsection{Stochastic temporal convolutional network ({\small STCN})}
The STCN \cite{aksan_stcn_2019} is a hierarchical latent variable model with an autoregressive generative model based on WaveNet \cite{oord_wavenet_2016}. Contrary to VRNN and SRNN, the latent variables have no transition functions connecting them over time.
Instead, a latent $\zb_t^{(l)}$ at layer $l$ is conditioned on a window of observed variables $\xb_{\scriptscriptstyle\mathcal{R}_t^{(l)}}$ via a WaveNet encoder where $\mathcal{R}_t^{(l)} = \cbra{t-r_l+1,\dots,t}$ and $r_l$ is the receptive field of the encoder at layer $l$. The window size grows exponentially with $l$.
\begin{equation}
    p(\xb_{1:T},\zb_{1:T}^{(1:L)}) = \prod_{t=1}^{T} p(\xb_t|\zb^{(1:L)}_{{\scriptscriptstyle \mathcal{R}}_{t}^{(1)}}) \prod_{l=1}^{L} p(\zb^{(l)}_{t}|\xb_{\scriptscriptstyle\mathcal{R}_{t-1}^{(l)}},\zb^{(l+1)}_{t}) \enspace , \nonumber
    \label{eq: stcn joint factorization}
\end{equation}
where $\zb_t^{(L+1)}:=\emptyset$ for notational convenience. 
The deterministic encoding is $\db_t^{(l)} = h(\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}})$ where $h$ is the encoder and $\db_t^{(l)}$ is extracted from the $l$'th layer similar to a Ladder VAE \cite{sonderby_ladder_2016}. 
The approximate posterior is
\begin{equation}
    q\pa{\zb_{1:T}^{(1:L)}|\xb_{1:T}} = \prod_{t=1}^{T}\prod_{l=1}^{L} q\pa{\zb_{t}^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}},\zb_t^{(l+1)}} \enspace .
\end{equation}
The factorized distributions are given as
% \begin{align}
%     q\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}},\zb_t^{(l+1)}\right) &= \mathcal{N}\pa{\alphab_{q}^{(l)}\pa{\zb_t^{(l+1)},\db_t^{(l)}}} \enspace , \\
%     p\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t-1}^{(l)}},\zb_{t}^{(l+1)}\right) &= \mathcal{N}\pa{\alphab_{p}^{(l)}\pa{\zb_t^{(l+1)},\db_{t-1}^{(l)}}} \enspace \nonumber , \\
%     p\left(\xb_t|\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}\right) &= \mathcal{D}\pa{\betab\pa{\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}}} \enspace . \nonumber \label{eq: stcn parameterization} \\
% \end{align}
\begin{gather}
    p\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t-1}^{(l)}},\zb_{t}^{(l+1)}\right) = \mathcal{N}\pa{\alphab_{p}^{(l)}\pa{\zb_t^{(l+1)},\db_{t-1}^{(l)}}} \, \enspace , \nonumber \\
    p\left(\xb_t|\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}\right) = \mathcal{D}\pa{\betab\pa{\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}}} \enspace , \label{eq: stcn parameterization} \\
    q\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}},\zb_t^{(l+1)}\right) = \mathcal{N}\pa{\alphab_{q}^{(l)}\pa{\zb_t^{(l+1)},\db_t^{(l)}}} \enspace . \nonumber
\end{gather}
The decoder $\betab\left(\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}\right)$ is also a WaveNet.

\subsection{Clockwork variational autoencoder ({\small CW-VAE})}
The CW-VAE \cite{saxena_clockwork_2021} is a hierarchical latent variable model recently introduced for video generation. As illustrated in figures~\cref{fig: autoregressive vrnn srnn cwvae 1L graphs} and \cref{fig: clockwork vae inference generative models 2 layered}, it is autoregressive in the latent space but not in the observed space, contrary to the VRNN, SRNN and STCN. Additionally, each latent variable is updated only every $s_l$ timesteps, where $s_l$ is a layer-dependent integer, or stride, and $s_1<s_2<\dots<s_L$.
This imposes the inductive bias that latent variables exist at different temporal resolutions with $\zb^{(l)}$ changing at lower frequency than $\zb^{(l-1)}$. 
In speech, phonetic variation between $\SI{10}{}-\SI{400}{ms}$, morphological and semantic features at the word level and speaker-related variation at the global scale make this a reasonable assumption.

To simplify temporal notation, we define the timesteps at which a layer updates its latent state as $\mathcal{T}_l := \{t \in \bra{1,T} \,|\, (t-1)\,\text{mod}\, s_l = 0\}$. We then define the set of layers that update at a given timestep as $\mathcal{J}_t := \{ l \,|\, t\in\mathcal{T}_l \}$. The joint distribution can then be written as,
% The timesteps at which a layer updates its latent state are given by $\mathcal{T}_l := \{t \in \cbra{1,\dots,T} ~|~ t\mod s_l = 1\}$.
% To simplify temporal notation, we equivalently define this by copying references to the unique states over time, $\zb^{(l)}_t := \zb^{(l)}_{u}$ with $u=\max_\tau \cbra{\tau \in \mathcal{T}_l ~|~ \tau \leq t}$.
% The joint distribution factorizes over time and the hierarchy.
\begin{equation}
    % p\pa{\xb_{1:T},\zb_{1:T}} = \prod_{t} p\pa{\xb_t | \zb_t^{(1)}} \prod_{l=1}^L\prod_{t\in\mathcal{T}_l} p\pa{\zb_t^{(l)} | \zb_{t-1}^{(l)}, \zb_t^{(l+1)}} \enspace .
    p(\xb_{1:T},\zb_{1:T}^{(1:L)}) = 
    \prod_{t=1}^{T} p(\xb_t | \zb_t^{(1)})
    \prod\limits_{l\in\mathcal{J}_t} p(\zb_t^{(l)} | \zb_{t-1}^{(l)}, \zb_t^{(l+1)}) \enspace . \nonumber
    \label{eq: cwvae joint}
\end{equation}
The inference model is conditioned on a window of the observed variable $\xb_{t:t+s_l}$ depending on the layer's stride $s_l$.
\begin{equation}
    q\pa{\zb_{1:T} | \xb_{1:T}} = \prod_{t=1}^T \prod\limits_{l\in\mathcal{J}_t} q\pa{\zb_t^{(l)} | \zb_{t-1}^{(l)}, \zb_t^{(l+1)}, \xb_{t:t+s_l}} \enspace . \nonumber
    \label{eq: cwvae inference}
\end{equation}
The dependency on $\xb_{t:t+s_l}$ is encoded via a convolutional ladder network similar to the STCN with $\db_t^{(l)} = e(\xb_{t:t+s_l})$. We define $\xb_{\scriptscriptstyle\mathcal{S}_t^l}:=\xb_{t:t+s_l}$ for compactness. The latent state transitions are densely connected and the decoder is also a convolutional network. 
% \begin{align}
%     q\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{S}_t^l},\zb_{t-1}^{(l)},\zb_t^{(l+1)}\right) &= \mathcal{N}\pa{\alphab_{q}^{l}\pa{\zb_{t-1}^{(l)},\zb_{t}^{(l+1)},\db_t^{(l)}}} \nonumber \\ % \label{eq: cwvae parameterization inference} \\
%     p\left(\zb_t^{(l)}|\zb_{t-1}^{(l)},\zb_{t}^{(l+1)}\right) &= \mathcal{N}\pa{\alphab_{p}^{l}\pa{\zb_{t-1}^{(l)},\zb_t^{(l+1)}}} \nonumber \\% \label{eq: cwvae parameterization prior} \\
%     p\left(\xb_t|\zb^{(1)}_{t-s_l/2:t+s_l/2}\right) &= \mathcal{D}\pa{\betab\pa{\zb^{(1)}_{t-s_l/2:t+s_l/2}}} \nonumber % \enspace . \label{eq: cwvae parameterization observation}
% \end{align}
\begin{gather}
    p\left(\zb_t^{(l)}|\zb_{t-1}^{(l)},\zb_{t}^{(l+1)}\right) = \mathcal{N}\pa{\alphab_{p}^{l}\pa{\zb_{t-1}^{(l)},\zb_t^{(l+1)}}} \, , \\ \nonumber
    p\left(\xb_t|\zb^{(1)}_{t-s_l/2:t+s_l/2}\right) = \mathcal{D}\pa{\betab\pa{\zb^{(1)}_{t-s_l/2:t+s_l/2}}} \, , \\
    q\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{S}_t^l},\zb_{t-1}^{(l)},\zb_t^{(l+1)}\right) = \mathcal{N}\pa{\alphab_{q}^{l}\pa{\zb_{t-1}^{(l)},\zb_{t}^{(l+1)},\db_t^{(l)}}} \nonumber
\end{gather}

\subsection{Speech modeling with Clockwork VAEs} \label{sec: cwvae on speech}
The video and speech modalities differ in the sampling rates normally used to digitize the natural signals. Sampling rates of common video codecs typically range from $\SI{24}{Hz}$ up to $\SI{60}{}$ or $\SI{120}{Hz}$. In the speech domain, sampling rates range from $\SI{8000}{Hz}$ up to e.g.\@ $\SI{44100}{Hz}$ commonly used for music recordings. In the original work, $s_l$ is defined as $s_l:= k^{l-1}$ for some constant $k$ which makes it exponentially dependent on the layer index $l$ and forces $s_1 = 1$. While this is reasonable for the sample rates of video, training a model at this resolution is infeasible for audio waveform modeling. For this reason, we chose $s_1 \gg 1$ to achieve an initial temporal downsampling and let $s_l:= c^{l-1}s_1$ for $l>1$ and some constant $c$.

The encoder and decoder of the original CW-VAE are not applicable to speech. Hence, we parameterize them using 1D convolutions operating on the raw waveform. We use a ladder-network, similar to \textcite{sonderby_ladder_2016, aksan_stcn_2019}, for the encoder. A ladder-network leverages parameter sharing across the latent hierarchy and importantly processes the full observed sequence only once, sharing the resulting representations between latent variables. This yields a computationally efficient encoder and more activity in latent variables towards the top of the hierarchy.

\subsection{Output distribution}
Audio, as well as image data, are naturally continuous signals that are represented in discrete form in computers. The signals are sampled with some bit depth $b$ which defines the range of attainable values, $\xb\in\{0,1,\dots,2^b-1\}$. The bit depth typically used in audio and image samplers is between $\SI{8}{}$ and $\SI{32}{bit}$ with $\SI{8}{bit}$ and $\SI{16}{bit}$ being the most common in the literature (MNIST \cite{lecun_gradientbased_1998}, CelebA \cite{liu_deep_2015}, CIFAR10 \cite{krizhevsky_learning_2009}, TIMIT \cite{garofolo_timit_1993}, LibriSpeech \cite{panayotov_librispeech_2015}).

In the image domain, the discrete nature of the data is usually modeled in one of two ways; either by using discrete distributions \cite{salimans_pixelcnn_2017, maaloe_biva_2019, child_very_2021} or by dequantizing the data and using a continuous distribution \cite{dinh_nice_2015, sonderby_ladder_2016, ho_flow_2019}, which yields a lower bound on the discrete distribution likelihood \cite{theis_note_2016}. 
In the speech domain, however, the output distribution is often taken to be a continuous Gaussian \cite{hsu_unsupervised_2017,lai_stochastic_2018,zhu_s3vae_2020} which was also originally done in VRNN, SRNN and STCN. 
This choice generally results in an ill-posed problem with a likelihood that is unbounded from above unless the variance is lower bounded \cite{mattei_leveraging_2018}. As a result, \emph{reported likelihoods can be sensitive to hyperparameter settings and be hard to compare}. We discuss this phenomenon further in appendix~\cref{app: gaussian likelihood unboundedness discussion}.

Most work normalizes the audio or standardizes it to values in a bounded interval $\xb\in[-1,1]$. Since $\xb$ becomes approximately continuous as the bit depth $b$ becomes large and the range of possible values becomes small, this alleviates the issue. However, commonly used datasets with bit-depths of $\SI{16}{}$ still result in a discretization gap between values that remains much larger than the gap between the almost continuous 32 bit floating point numbers which reinforces the problem \cite{bishop_pattern_2006}.

In this work, we therefore benchmark models using a discretized mixture of logistics (DMoL) as output distribution. The DMoL was introduced for image modeling with autoregressive models \cite{salimans_pixelcnn_2017} but has become standard in other generative models \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}. 
It was recently applied to autoregressive speech modeling of raw waveforms \cite{oord_parallel_2018}. 
As opposed to e.g.\@ a categorical distribution, the DMoL induces ordinality on the observed space such that values that are numerically close are also close in a probabilistic sense. This is a sensible inductive bias for images as well as audio where individual samples represent the amplitude of light or pressure, respectively. We discuss the DMoL for audio further in appendix~\cref{app: additional discussion on output distribution}.

\section{Speech modeling benchmark}\label{sec:benchmark}
\paragraph{Data}
We train models on TIMIT \cite{garofolo_timit_1993}, LibriSpeech \cite{panayotov_librispeech_2015} and LibriLight \cite{kahn_libri-light_2020}. For TIMIT, we randomly sample 5\% of the training split for validation. 
We represent the audio as $\mu$-law encoded PCM standardized to values in $\bra{-1, 1}$ with discretization gap of ${2^{-b+1}}$. We use the original bit depth of $\SI{16}{bits}$ and sample rate of $\SI{16000}{Hz}$. We use this representation both as the input and the reconstruction target. We provide more details on the datasets in appendix~\cref{app: dataset details} and additional results on linear PCM in appendix~\cref{app: additional likelihood results appendix}.
%
\paragraph{Likelihood}
We report likelihoods in units of bits per frame (bpf) as this yields a more interpretable and comparable likelihood than total likelihood in nats. It also has direct connections with information theory and compression \cite{shannon_mathematical_1948, townsend_practical_2019}. In units of bits per frame, lower is better. For LVMs, we report the one-sample ELBO. The likelihoods can be seen in \cref{tab: likelihoods timit,tab: likelihoods librispeech}. We describe how to convert likelihood to bpf in appendix~\cref{app: likelihood in bits per frame}.
%
\paragraph{Models}
Architecture and training details are sketched below, while the full details are in appendices~\cref{app: model architectures} and \cref{app: training details} along with additional results for some alternative model configurations in appendix~\cref{app: additional likelihood results appendix}.
We select model configurations that can be trained on GPUs with a maximum of 12GB of RAM and train all models until convergence on the validation set.
We use a DMoL with 10 components for the output distribution of all models and model all datasets at their full bit depth of $\SI{16}{bits}$. We train and evaluate models on stacked waveforms similar to previous work \cite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019} with stack sizes of $s=1$, $s=64$ and $s=256$. Hence, every model input $\xb_t$ is composed of $\tilde{x}_{t:t+s}$ where $\tilde{x}$ are waveform frames. We provide additional results with a Gaussian output distribution in appendix~\cref{app: additional likelihood results appendix}.

We configure WaveNet as in the original paper using ten layers per block and five blocks. We use $D_c=96$ channels. 
We also train an LSTM model \cite{hochreiter_long_1997} which has fully connected encoders and decoders, as the VRNN and SRNN, but a deterministic recurrent cell and much fewer parameters than WaveNet. We report on LSTM models with $D_d=256$ hidden units. 
%Results with differently configured WaveNet and LSTM models that did not improve results can be found in appendix~\cref{app: additional likelihood results appendix}.

The configuration of the VRNN and SRNN models is similar to the LSTM. For both models we set the latent variable equal in size to the hidden units, $D_z=256$. At stack size $s=1$, the models are computationally demanding and hence we train them on randomly sampled segments from each training example and only on TIMIT.

The STCN is used in the ``dense" configuration of the original work \cite{aksan_stcn_2019}. It uses 256 convolutional channels and $L=5$ latent variables of dimensions 16, 32, 64, 128, 256 from the top down. We also run a one-layered ablation with the same architecture but only one latent variable of dimension $256$ at the top. 
The CW-VAE is configured similar to the VRNN and SRNN models and with $c=8$. 
We run the CW-VAE with $L=1$ and $2$ layers of latent variables. 
The number of convolutional channels and is set equal to $D_z$ which is set to 96.
%
\paragraph{Baselines} We supply three elementary baselines that form approximate upper and lower bounds on the likelihood for arbitrary $s$. Specifically, we evaluate an uninformed per-frame discrete uniform distribution and a two-component DMoL fitted to the training set to benchmark worst case performance. 
We also report the likelihood achieved by the lossless compression algorithm, FLAC \cite{coalson_free_2019} which establishes a notion of good performance, although not a strict best case. We report FLAC on linear PCM since it was designed for this encoding.

\subsection{Likelihood results}
%
\begin{table}[t]
    \hspace{4mm}
    \begin{minipage}[t]{.45\textwidth}
    \centering
    \begin{adjustbox}{center, totalheight=0.35\paperheight}
    \begin{tabular}[t]{lll|r}
        \toprule
        $s$ & \bfseries Model         & \bfseries Configuration           & \bfseries $\mathcal{L}$ [bpf] \\
        \midrule
        1 & Uniform             & Uninformed                  & 16.00 \\
        1 & DMoL                & Optimal                     & 15.60 \\   % $\vmu=[-0.426, 0.426], \zb=[0.11, 0.11]$
        %- & FLAC                & Linear PCM                  & \textbf{8.582} \\ % Maybe replace this with FLAC compression of mu-law encoded audio 13.33
        \midrule
        1 & WaveNet             & $D_c=96$                    & \textbf{10.88} \\  % s2ibtri3
        1 & LSTM                & $D_d=256$                   & 10.97 \\  % k8i28qxf
        1 & VRNN                & $D_z=256$                   & $\leq$11.09 \\  % 2j6lyktk
        1 & SRNN                & $D_z=256$                   & $\leq$11.19 \\  % 34twkofx
        1 & STCN                & $D_z=256,L=5$               & $\leq$11.77 \\  % 39u25hyi
        \midrule
        64 & WaveNet            & $D_c=96$                    & 13.30 \\  % pbethudv
        64 & LSTM               & $D_d=256$                   & 13.34 \\  % qteylgf1
        64 & VRNN               & $D_z=256$                   & $\leq$12.54 \\  % 3b6u5nf1
        64 & SRNN               & $D_z=256$                   & $\leq$12.42 \\  % 5uws1w5e
        64 & CW-VAE             & $D_z=96,L=1$                & $\leq$12.44 \\
        64 & CW-VAE             & $D_z=96,L=2$                & $\leq$12.17 \\
        % 64 & CW-VAE             & $D_z=96,L=3$                & $\leq$12.18 \\
        64 & STCN               & $D_z=256,L=1$               & $\leq$12.32 \\  % 2ss9fvxd
        64 & STCN               & $D_z=256,L=5$               & $\leq$\textbf{11.78} \\
        \midrule
        256 & WaveNet           & $D_c=96$                    & 14.11 \\  % 1gqazojy
        256 & LSTM              & $D_d=256$                   & 14.20 \\  % wce8d70i
        256 & VRNN              & $D_z=256$                   & $\leq$13.27 \\  % 3t9fbn9d
        256 & SRNN              & $D_z=256$                   & $\leq$13.14 \\  % 2axu3hso
        256 & CW-VAE            & $D_z=96,L=1$                & $\leq$13.11 \\
        256 & CW-VAE            & $D_z=96,L=2$                & $\leq$12.97 \\
        % 256 & CW-VAE            & $D_z=96,L=3$                & $\leq$12.87 \\
        256 & STCN              & $D_z=256,L=1$               & $\leq$13.07 \\  % 2h5ns3zw
        256 & STCN              & $D_z=256,L=5$               & $\leq$\textbf{12.52} \\
        \bottomrule
    \end{tabular}
    \end{adjustbox}
    \vspace{1mm}
    \caption{}
    \vspace{-1mm}
    \label{tab: likelihoods timit}
    \end{minipage}%
    \hfill
    \begin{minipage}[t]{.45\textwidth}
    \centering
    \begin{adjustbox}{center, totalheight=0.35\paperheight}
    \begin{tabular}[t]{cll|c}
        \toprule
        \multicolumn{3}{c}{\bfseries ASR configuration} & \multicolumn{1}{c}{\textbf{Result}} \\
        \midrule
        Data   &  Model           & Input         &  PER [\%]  \\
                \midrule
        % 3.7h   &  Waveform        & stacked       &   84,1   \\  %           0,8405363165254586
        3.7h   &  Spectrogram     & Mel           &   24.1   \\  % 2wjqhwf7  0.24149049930081432
        3.7h   &  WaveNet         & $\hb^{(15)}$  &   27.7   \\  % 36ccyxf4  0.27743686764826847
        3.7h   &  LSTM            & $\hb$         &   23.0   \\  % 3bh4vgrx  0.2301719174138357
        3.7h   &  VRNN            & $\zb$         &   23.2   \\  % 3stwdgas  0.23229415151764415
        3.7h   &  SRNN            & $\zb$         &   26.0   \\  % 2bkvv5i0  0.2596528748868964
        3.7h   &  CW-VAE          & $\zb^{(1)}$   &   36.4   \\  % 3m4f3f7z
        3.7h   &  STCN            & $\zb^{(2)}$   &   \textbf{21.9}   \\  % rkukk5bn  0.2189849469441474
        \midrule
        % 1h     &  Waveform        & stacked       &   59.3   \\  %           0.5926626634860575
        1.0h   &  Spectrogram     & Mel           &   30.8   \\  % 3kmfpljj  0.3083655507115242
        1.0h   &  WaveNet         & $\hb^{(15)}$  &   34.7   \\  % 36ccyxf4  0.34743769021962656
        1.0h   &  LSTM            & $\hb$         &   30.1   \\  % 3bh4vgrx  0.30137369416796905
        1.0h   &  VRNN            & $\zb$         &   30.4   \\  % 3stwdgas  0.31284033889939955
        1.0h   &  SRNN            & $\zb$         &   31.7   \\  % 2bkvv5i0  0.3167064242823065
        1.0h   &  CW-VAE          & $\zb^{(1)}$   &   40.0   \\  % 3m4f3f7z
        1.0h   &  STCN            & $\zb^{(2)}$   &   \textbf{26.7}   \\  % rkukk5bn  0.2671547256724521
        \midrule
        10m    &  Waveform        & stacked       &   85.6   \\  %           0.8564448465904417
        10m    &  Spectrogram     & Mel           &   47.1   \\  % 1sep2uq8  0.4706753310849716
        10m    &  WaveNet         & $\hb^{(15)}$  &   52.8   \\  % 36ccyxf4  0.5276137204902526
        10m    &  LSTM            & $\hb$         &   46.1   \\  % 3bh4vgrx  0.46073866907954264
        10m    &  VRNN            & $\zb$         &   44.6   \\  % 3stwdgas  0.4601299662745743
        10m    &  SRNN            & $\zb$         &   47.3   \\  % 2bkvv5i0  0.47317594801349017
        10m    &  CW-VAE          & $\zb^{(1)}$   &   54.9   \\  % 3m4f3f7z
        10m    &  STCN            & $\zb^{(2)}$   &   \textbf{42.7}   \\  % rkukk5bn  0,4270625976803488
\bottomrule
    \end{tabular}
    \end{adjustbox}
    \vspace{1mm}
    \caption{}
    \vspace{-1mm}
    \label{tab: phoneme recognition (PER)}
    \end{minipage}%
    \caption*{%
    In \cref{tab: likelihoods timit} we report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines. 
    In \cref{tab: phoneme recognition (PER)} we report phoneme error rate (PER) of different representations for phoneme recognition on TIMIT. 
    }
\end{table}
%
\begin{table*}[t]
    \centering
    \begin{adjustbox}{center, totalheight=0.182\paperheight}
    \begin{tabular}{lll|rrrr}  % https://wandb.ai/vseq/icml_asr
        \toprule
        $s$ & \bfseries Model                   & \bfseries Configuration       & \multicolumn{4}{c}{\bfseries Likelihood $\mathcal{L}$ [bpf]} \\
            & & & dev-clean & dev-other & test-clean & test-other\\
            & & & 10h/100h & 10h/100h & 10h/100h & 10h/100h\\
        \midrule
        1 & Uniform             & Uninformed     & \multicolumn{1}{c}{16.00} & \multicolumn{1}{c}{16.00} & \multicolumn{1}{c}{16.00} & \multicolumn{1}{c}{16.00} \\
        1 & DMoL                & Optimal        & \multicolumn{1}{c}{15.66} & \multicolumn{1}{c}{15.70} & \multicolumn{1}{c}{15.62} & \multicolumn{1}{c}{15.71}\\   % $\vmu=[-0.551, 0.551], \zb=[0.11, 0.11]$
        % - & FLAC                & Linear PCM     & \multicolumn{1}{c}{\bfseries 9.390} & \multicolumn{1}{c}{\bfseries 9.292} & \multicolumn{1}{c}{\bfseries 9.700} & \multicolumn{1}{c}{\bfseries 9.272} \\
        \midrule
        1 & Wavenet             & $D_c=96$       & \bfseries 10.96/10.89 & \bfseries 10.85/10.76 & \bfseries 11.12/11.01 & \bfseries 11.05/10.85 \\
        1 & LSTM                & $D_d=256$      & 11.21/11.17 & 11.10/11.06 & 11.35/11.29 & 11.28/11.23 \\  % 15ua3hau/(lmssrk5k,248n2ylu)
        \midrule
        64 & Wavenet            & $D_c=96$       & 13.61/13.24 & 13.58/13.21 & 13.61/13.22 & 13.60/13.21 \\  % 2ghkybtd/3iieasdt
        64 & LSTM               & $D_d=256$      & 13.56/13.25 & 13.52/13.24 & 13.55/13.23 & 13.56/13.25 \\  % 10fpovu3/2bu1izin
        64 & CW-VAE             & $D_z=96,L=1$   & $\leq$12.32/12.24 & 12.32/12.23 & 12.43/12.33 & 12.43/12.33 \\
        64 & CW-VAE             & $D_z=96,L=2$   & $\leq$12.30/12.22 & 12.30/12.21 & 12.40/12.31 & 12.39/12.32 \\
        64 & STCN-dense         & $D_z=256,L=5$  & $\leq$\bfseries 11.98/11.47 & \bfseries 11.98/11.46 & \bfseries 12.08/11.58 & \bfseries 12.09/11.60 \\  % 3vivjzrn/3t0xg7pb
        \bottomrule
    \end{tabular}
    \end{adjustbox}
    \caption{
    Model likelihoods $\mathcal{L}$ on LibriSpeech test sets represented as $\SI{16}{bit}$ $\mu$-law encoded PCM.
    For the CW-VAE, $s$ refers to $s_1$ and the two-layered models have $s_2=8s_1$.
    The models are trained on either the $\SI{10}{h}$ LibriLight subset or the $\SI{100}{h}$ LibriSpeech \textrm{train-clean-100} subset as indicated.
    Likelihoods are given in units of bits per frame (bpf).
    }
    \label{tab: likelihoods librispeech}
\end{table*}
%
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.555\textwidth]{paper_benchmarking/1xrnjn5y_speaker_phonemes_drw0_latent_0_samples_100_lda_linear_subspace.pdf}
    \hfill
    \includegraphics[width=0.425\textwidth]{paper_benchmarking/1xrnjn5y_speaker_phonemes_drw0_samples_100_knn_correct_fraction_model_vs_mel_30_neighbors_lda_subspace_5d.pdf}
    \caption{
    (left) Clustering of phonemes in a 2D Linear Discriminant Analysis (LDA) subspace of a CW-VAE latent space ($\zb^{(1)}$).
    (right) Leave-one-out phoneme classification accuracy for a KNN classifier at different $K$ in a 5D LDA subspace of a CW-VAE latent space.
    }
    \label{fig: latent space phoneme and knn}
\end{figure*}
%

\paragraph{TIMIT}
For temporal resolutions of $s=1$, the deterministic autoregressive models yield the best likelihoods with WaveNet achieving $\SI{10.88}{bpf}$ on TIMIT as seen in \cref{tab: likelihoods timit}. 
Somewhat surprisingly, the LSTM baseline almost matches WaveNet with a likelihood of $\SI{11.11}{bpf}$ at $s=1$. However, due to being autoregressive in training, the LSTM trains considerably slower than the parallel convolutional WaveNet; something not conveyed by \cref{tab: likelihoods timit}. 
Notably, the VRNN and SRNN models achieve likelihoods close to that of WaveNet and the LSTM at around $\SI{11.09}{bpf}$. The STCN exhibited instability when trained at $s=1$ and tended to diverge.

At $s=64$, Wavenet and the LSTM yield significantly worse likelihoods than all LVMs separated by $\sim\SI{1}{bit}$.
The CW-VAE outperforms the VRNN and SRNN when configured with a hierarchy of latent variables. 
With a single layer of latent variables, the CW-VAE is inferior to both SRNN and VRNN but notably still better than the LSTM. 
These observations carry to $s=256$, where a multilayered CW-VAE outperforms the LSTM, VRNN and SRNN. 
The STCN yields the best results at both $s=64$ and $s=256$. 
For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables, such as WaveNet and the LSTM, to the ability of LVMs to model intra-step correlations \cite{lai_re-examination_2019}.

Decreasing the resolution $s$ improves the likelihood for all LVMs. However, the best performing models, STCN and CW-VAE are not yet scalable to this regime for reasons related to numerical instability and computational infeasibility. This seems to indicate that LVMs may be able to outperform autoregressive models at $s=1$ in the future.
\paragraph{LibriSpeech}
On LibriSpeech (\cref{tab: likelihoods librispeech}), results are similar to TIMIT. The STCN achieves the best likelihood at $s=64$ and the CW-VAE surpasses WaveNet and the LSTM. 
\paragraph{Compression}
A connection can be made between the model likelihoods and the compression rates of audio compression algorithms. 
Lossy compression algorithms, such as MP3, exploit the dynamic range of human hearing to achieve 70-95\% reduction in bit rate \cite{brandenburg_mpeg-2_1998} while lossless compression algorithms, such as FLAC, achieve 50-70\% reduction \cite{coalson_free_2019} independent of audio content. 
% On LibriSpeech, FLAC achieves an overall compression to about 58\% of the original bit rate.
Although both the autoregressive models and the LVMs are lossy, their objective minimizes the amount of incurred loss. 
The best likelihoods reported in tables~\cref{tab: likelihoods timit} and \cref{tab: likelihoods librispeech} correspond to about a 30\% reduction in bit rate which indicates that there are significant gains in likelihood to be made in speech modeling.

\section{Phoneme recognition}\label{sec: phoneme recognition}
Although the likelihood is a practical metric for model comparison and selection, a high likelihood does not guarantee that a model has learned useful representations \cite{huszar_is_2017}. 
For speech data, we would expect models to learn features related to phonetics which would make them useful for tasks such as automatic speech recognition (ASR). 
The Mel spectrogram is a well-established representation of audio designed for speech recognition and is predefined rather than learned. % \cite{??}. Basic Correlates of the Auditory Stimulus, https://psycnet.apa.org/record/1951-07758-017, modern mel scale https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=1170013&casa_token=UIyQaj3tBRAAAAAA:GtttZedngBkaxVumjnXctc5zS50fptexPMiWhr-7gDrmIVq3JqJPnGCWYedoymb0X5JTGElmHf8&tag=1,
To assess the usefulness of the representations learned by the benchmarked models, we compare them to the highly useful Mel spectrogram on the task of phoneme recognition. 
Phonemes are fundamental units of speech that relate to how parts of words are pronounced (see also appendix~\cref{app: timit phoneme distributions}).
% rather than characters or words themselves (see also appendix~\cref{apsp: timit phoneme distributions}).

\paragraph{Quantitative}
We train an ASR model to recognize phonemes and compare its performance when using input representations obtained from different unsupervised models. 
For WaveNet and the LSTM, we use the hidden state as the representation. For all LVMs, we use the latent variable. For the hierarchical LVMs and WaveNet, we run the experiment using each possible representation and report only the best one here. 
We compare the learned representations to a log Mel spectrogram with 80 filterbanks, hop length 64 and window size 128. 
We also compared to using raw PCM in vectors of 64 elements standardized to $[-1,1]$ but found that the ASR did not reliably converge at all which highlights the importance of input representation. 
The ASR model is a three-layered bidirectional LSTM with 256 hidden units. It is trained with the connectionist temporal classification (CTC) loss \cite{graves_connectionist_2006} which lets it jointly learn to align and classify without using label alignments. 
We pre-train all unsupervised models at $s=64$ on the full TIMIT training dataset excluding the validation data (3.7h) as in \cref{tab: likelihoods timit}. 
We then train the ASR model on all 3.7h as well as 1h and 10m subsets. We report results on the test set in terms of phoneme error rates (PER) in \cref{tab: phoneme recognition (PER)}.

As expected, Mel spectrogram performs well achieving 24.1\% PER using 3.7 hours of labeled data. However, the ASR trained on STCN representations outperforms the Mel spectrogram with a PER of 21.9\%. 
This indicates that unsupervised STCN representations are phonetically rich and potentially better suited for speech modeling than the engineered Mel spectrogram. 
When the amount of labeled data is reduced, LVM representations suffer slightly less than deterministic ones. WaveNet representations are interestingly outperformed by both the LSTM and all LVMs.

% FH-VAE uses 3 layer bidirectional LSTM on 80 mel features. LSTM hidden state dim is 512 or 1024(?)
% FH-VAE https://arxiv.org/pdf/1709.07902.pdf
% [37] https://static.googleusercontent.com/media/research.google.com/da//pubs/archive/43905.pdf
% [13] https://www.cs.toronto.edu/~graves/asru_2013.pdf

\paragraph{Qualitative}
We qualitatively asses the learned latent representations for the CW-VAE. 
We infer the latent variables of all utterances by a single speaker from the TIMIT test set. We sample the latent sequence $100$ times to estimate the mean representation per time step. We then compute the average latent representation over the duration of each phoneme using aligned phoneme labels. This approximately marginalizes variation during the phoneme. We use linear discriminant analysis (LDA) \cite{fisher_use_1936} to obtain a low-dimensional linear subspace of the latent space. 
We visualize the resulting representations colored according to test set phoneme classes in \cref{fig: latent space phoneme and knn}. We note that many phonemes are separable in the linear subspace and that related phonemes such as ``s" and ``sh" are close.

We also show the average accuracy of a leave-one-out $k$-nearest-neighbor (KNN) classifier on the single left-out latent representation reduced with a 5-dimensional LDA as a function of the number of neighbors. 
We compare accuracy to a Mel-spectrogram averaged over each phoneme duration and LDA reduced. The spectrogram is computed with hop length set to 64, equal to $s_1$ for the CW-VAE, window size 256 and 80 Mel bins.
We see that both latent spaces yield significantly better KNN accuracies than the Mel features.

\section{Conclusion}
In this paper, we developed a benchmark for speech modeling with stochastic latent variable models (LVMs). 
We compared LVMs and deterministic autoregressive models on equal footing and found that LVMs achieve inferior likelihood compared to deterministic WaveNet and LSTM baselines. Surprisingly, the LSTM almost matched the popular WaveNet model. 
We saw that hierarchical LVMs, such as STCN and CW-VAE, outperformed non-hierarchical versions of themselves in ablation experiments as well as non-hierarchical LVMs such as VRNN and SRNN. This matches recent observations in the image domain. 
We noted that the STCN with an autoregressive decoder outperforms the non-autoregressive CW-VAE, which we adapted to speech. 
Finally, we found that LVMs can learn latent representations that are useful for phoneme recognition and better than Mel spectrograms, which are tailored for the task, when identical models are trained on top of the representations.
While the best performing models are not yet scalable to the highest temporal resolution, these results indicate that they might improve upon deterministic models in the future.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Acknowledgements}
This research was partially funded by the Innovation Fund Denmark via the Industrial PhD Programme (grant no.\@ 0153-00167B). JF and SH were funded in part by the Novo Nordisk Foundation (grant no.\@ NNF20OC0062606) via the Center for Basic Machine Learning Research in Life Science (MLLS, \hyperlink{https://www.mlls.dk}{https://www.mlls.dk}). JF was further funded by the Novo Nordisk Foundation (grant no.\@ NNF20OC0065611) and the Independent Research Fund Denmark (grant no.\@ 9131-00082B). SH was further funded by VILLUM FONDEN (15334) and the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement no. 757360).

\iffalse

\fi

### 2023-02-25 chapters/paper_benchmarking_workshop.tex
Add benchmarking paper

%!TEX root = ../thesis.tex

\chapter[benchmarking generative latent variable models for speech]{Benchmarking Generative Latent Variable Models for Speech}
\label{chp:benchmarking}

\section{Abstract}
Stochastic latent variable models (LVMs) achieve state-of-the-art performance on natural image generation but are still inferior to deterministic models on speech. In this paper, we develop a speech benchmark of popular temporal LVMs and compare them against state-of-the-art deterministic models. We report the likelihood, which is a much used metric in the image domain, but rarely, or incomparably, reported for speech models. To assess the quality of the learned representations, we also compare their usefulness for phoneme recognition. Finally, we adapt the Clockwork VAE, a state-of-the-art temporal LVM for video generation, to the speech domain. Despite being autoregressive only in latent space, we find that the Clockwork VAE can outperform previous LVMs and reduce the gap to deterministic models by using a hierarchy of latent variables.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
Since their introduction, temporal latent variable models (LVMs) for speech \cite{chung_recurrent_2015, fraccaro_sequential_2016} based on the variational autoencoder (VAE,  \textcite{kingma_auto-encoding_2014, rezende_stochastic_2014}) have seen little development compared to their image domain counterparts. While LVMs now achieve superior likelihoods on images compared to deterministic models \cite{child_very_2021,sinha_consistency_2021,kingma_variational_2021}, they are still inferior to deterministic models such as WaveNet on speech \cite{oord_wavenet_2016}.
Development in the image domain has been driven by established likelihood benchmarks, but in the speech domain, likelihoods are often not reported \cite{oord_wavenet_2016, hsu_unsupervised_2017, oord_neural_2018} or are incomparable due to subtle differences in the chosen data distributions \cite{chung_recurrent_2015, fraccaro_sequential_2016, hsu_unsupervised_2017, aksan_stcn_2019}. 
This makes it hard to develop explicit likelihood models for speech.

In this paper, we develop a likelihood benchmark for recent temporal LVMs and compare to deterministic counterparts including WaveNet. We introduce a hierarchical LVM without autoregressive decoder and finally evaluate the learned representations for phoneme recognition. We find that 
(i) LVMs achieve likelihoods superior to WaveNet at low temporal resolution, 
(ii) LVMs with autoregressive decoders achieve better likelihoods than a non-autoregressive LVM, 
(iii) LVM likelihoods improve when using hierarchies of latent variables as also seen for images and, 
(iv) LVM representations are as good or better than Mel spectrograms for phoneme recognition.

\iffalse

\begin{figure*}[t!]
    \centering
    \begin{multicols}{5}
    \setlength\columnsep{0mm}
    \resizebox{!}{4.5cm}{
    \begin{tikzpicture}
        \node[obs] (x_t) {$\vx_t$};

        % \node[above=0.75cm of x_t, xshift=0cm] (cap1) {$p(\vx_t|\vx_{<t})$};
        \node[above=0.3cm of x_t] (caption) {LSTM};

        \node[det,below=2.55cm of x_t] (d_t) {$\vd_{t}$};
        \node[det,left=0.75cm of d_t] (d_tm1) {$\vd_{t-1}$};
        \node[obs,below=0.75cm of d_t] (x_tm1) {$\vx_{t-1}$};

        \edge[deterministic-skip-color]{x_tm1}{d_t};
        \edge[deterministic-skip-color]{d_t}{x_t};
        \edge[deterministic-skip-color]{d_tm1}{d_t};
        \edge[draw=white,bend right]{d_t}{x_t};  % for identical width to others
    \end{tikzpicture}
    }
    \\
    \resizebox{!}{4.5cm}{
    \begin{tikzpicture}
        \node[obs] (x_t) {$\vx_t$};

        % \node[above=0.75cm of x_t, xshift=0cm] (cap1) {$p(\vx_t|\vz_{\leq t},\vx_{<t}) p(\vz_t|\vx_{<t},\vz_{<t})$};
        \node[above=0.3cm of x_t] (caption) {VRNN};

        \node[latent,below=0.75cm of x_t] (z_t) {$\vz_t$};
        \node[latent,left=0.75cm of z_t] (z_tm1) {$\vz_{t-1}$};

        \node[det,below=0.75cm of z_t] (d_t) {$\vd_{t}$};
        \node[det,below=0.75cm of z_tm1] (d_tm1) {$\vd_{t-1}$};
        \node[obs,below=0.75cm of d_t] (x_tm1) {$\vx_{t-1}$};

        \edge[deterministic-skip-color]{x_tm1}{d_t};
        \edge[]{d_t}{z_t};
        \edge[]{z_tm1}{d_t};
        \edge[deterministic-skip-color]{d_tm1}{d_t};
        \edge[]{z_t}{x_t};
        \edge[deterministic-skip-color,bend right]{d_t}{x_t};
    \end{tikzpicture}
    }
    \\
    \resizebox{!}{4.5cm}{
    \begin{tikzpicture}
        \node[obs] (x_t) {$\vx_t$};

        % \node[above=0.75cm of x_t, xshift=0cm] (cap1) {$p(\vx_t|\vz_{\leq t},\vx_{<t}) p(\vz_t|\vx_{<t},\vz_{<t})$};
        \node[above=0.3cm of x_t] (caption) {SRNN};

        \node[latent,below=0.75cm of x_t] (z_t) {$\vz_t$};
        \node[latent,left=0.75cm of z_t] (z_tm1) {$\vz_{t-1}$};

        \node[det,below=0.75cm of z_t] (d_t) {$\vd_{t}$};
        \node[det,below=0.75cm of z_tm1] (d_tm1) {$\vd_{t-1}$};
        \node[obs,below=0.75cm of d_t] (x_tm1) {$\vx_{t-1}$};

        \edge[deterministic-skip-color]{x_tm1}{d_t};
        \edge[]{d_t}{z_t};
        \edge[]{z_tm1}{z_t};
        \edge[deterministic-skip-color]{d_tm1}{d_t};
        \edge[]{z_t}{x_t};
        \edge[deterministic-skip-color,bend right]{d_t}{x_t};
    \end{tikzpicture}
    }
    \\
    \resizebox{!}{4.5cm}{
    \begin{tikzpicture}
        \node[obs] (x_t) {$\vx_t$};
        % \node[above=0.75cm of x_t, xshift=0cm] (cap1) {$p(\vx_t|\vz_t,\vx_{t-r:t-1})$};
        \node[above=0.3cm of x_t] (caption) {STCN};

        \node[latent,below=0.75cm of x_t] (z_t) {$\vz_t$};
        \node[left=0.025cm of z_t] (dots) {$\cdots$};
        \node[latent,left=0.75cm of z_t] (z_tmr) {$\vz_{t-r'}$};

        \node[det,below=0.75cm of z_t] (d_t) {$\vd_{t}$};
        \node[obs,below=0.75cm of d_t] (x_tm1) {$\vx_{t-1}$};
        \node[left=0.025cm of x_tm1] (dots) {$\cdots$};
        \node[obs,left=0.75cm of x_tm1] (x_tmr) {$\vx_{t-r}$};

        \edge[]{x_tm1}{d_t};
        \edge[]{x_tmr}{d_t};
        \edge[]{d_t}{z_t};
        \edge[]{z_t}{x_t};
        \edge[]{z_tmr}{x_t};
        \edge[draw=white,bend right]{d_t}{x_t};  % for identical width to others 
    \end{tikzpicture}
    }
    \\
    \resizebox{!}{4.5cm}{
    \begin{tikzpicture}
        \node[obs] (x_t) {$\vx_t$};%

        % \node[above=0.75cm of x_t, xshift=0cm] (cap1) {$p(\vx_t|\vz_{\leq t}) p(\vz_t|\vz_{<t})$};
        \node[above=0.3cm of x_t] (caption) {CW-VAE};

        \node[latent,below=0.75cm of x_t] (z_t) {$\vz_t$};
        \node[latent,left=0.75cm of z_t] (z_tm1) {$\vz_{t-1}$};

        \node[det,below=0.75cm of z_t] (d_t) {$\vd_{t}$};
        \node[det,below=0.75cm of z_tm1] (d_tm1) {$\vd_{t-1}$};
        \node[obs,below=0.75cm of d_t,draw=white,fill=white] (x_tm1) {};

        \edge[]{d_t}{z_t};
        \edge[bend right]{d_t}{x_t};
        \edge[]{z_tm1}{d_t};
        \edge[]{d_tm1}{d_t};
        \edge[]{z_t}{x_t};
    \end{tikzpicture}
    }
    \end{multicols}
    \vspace{-4mm}
\caption{
Generative models of an LSTM, VRNN, SRNN, STCN and CW-VAE for a single time step. The STCN and CW-VAE are illustrated with a single latent variable. 
Red arrows indicate purely deterministic paths from the output $\vx_t$ to previous input $\vx_{<t}$ without passing a stochastic node. 
We provide additional graphical illustrations including inference models in appendix~\ref{app: additional graphical models}.
}
\label{fig: autoregressive vrnn srnn cwvae 1L graphs}
\vspace{-3mm}
\end{figure*}

\paragraph{Scope and related work} 
At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models\footnote{\texttt{ \href{https://github.com/JakobHavtorn/benchmarking-lvms}{github.com/JakobHavtorn/benchmarking-lvms}}}. 
An extended version of this paper is available at \href{https://arxiv.org/abs/2202.12707}{arXiv:2202.12707}.
LVMs based on the VAE are of interest due to their ability to learn an approximate posterior distribution over latent variables. This makes them useful beyond generation for e.g. semi-supervised learning \cite{kingma_semi-supervised_2014} and anomaly detection \cite{havtorn_hierarchical_2021}. 
In this paper, we focus on the VRNN \cite{chung_recurrent_2015}, SRNN \cite{fraccaro_sequential_2016} and STCN \cite{aksan_stcn_2019} since they are similar to the original VAE framework. 
We exclude some related work from the benchmark. Specifically, the FH-VAE \cite{hsu_unsupervised_2017} adds a global latent variable but we exclude it due to it's segmentation of the training data. We also exclude Z-forcing \cite{goyal_z-forcing_2017} which resembles the SRNN, but deviates from maximum likelihood training by using an auxiliary loss. We exclude the VQ-VAE \cite{oord_neural_2018} which is a hybrid LVM and autoregressive model, since it uses a quantized latent space. Finally, we exclude the Stochastic WaveNet \cite{lai_stochastic_2018} which is similar, but inferior, to the STCN \cite{aksan_stcn_2019}.

All selected models have autoregressive generative models. We therefore formulate and benchmark a novel temporal LVM which does not use an autoregressive decoder. We do so by adapting the hierarchical Clockwork Variational Autoencoder \cite{saxena_clockwork_2021}, originally proposed for video generation, to speech.
Before presenting the results, we provide a brief survey of existing LVMs for speech in a coherent notation.

\section{Models}
\paragraph{Sequential deep latent variable models}
The models we consider are all sequential deep latent variable models trained with variational inference and the reparameterization trick \cite{kingma_auto-encoding_2014}.
The input is a variable-length sequence $\vx = \vx_{1:T} = \left(\vx_1, \vx_2, \dots, \vx_{T}\right)$ with $\vx_t\in\mathbb{R}^{D_x}$. 
The models first encode $\vx_{1:T}$ to a temporal posterior distribution $q(\vz|\vx)$ and sample a latent representation $\vz_{1:T}$ where $\vz_t\in\mathbb{R}^{D_z}$. This is then used to reconstruct the input while the posterior distribution is regularized to be close to a prior distribution $p(\vz_t|\cdot)$ where the dot indicates that it may depend on latent and observed variables at previous time steps, $\vz_{<t}$ and $\vx_{<t}$ where $\vz_{<t} := \left( \vz_1, \vz_2, \dots, \vz_{t-1} \right)$. 
The models are trained to maximize a lower bound $\mathcal{L}(\thetab,\phib;\vx)$ on the marginal likelihood $\log p_\thetab(\vx)$,
\begin{equation}
    \log p_\thetab(\vx) = \log \int p_\thetab(\vx,\vz) \, \mathrm{d}\vz  \geq \int q_\phib(\vz|\vx) \log \frac{p_\thetab(\vx|\vz)p(\vz)}{q_\phib(\vz|\vx)}  \, \mathrm{d}\vz \triangleq \mathcal{L}(\thetab,\phib;\vx)  \enspace , \label{eq: elbo}
\end{equation}
where $\thetab$ are parameters of the generative model and $q_\phib(\vz|\vx)$ is the variational approximation to the true posterior.
Graphical illustrations of the models can be seen in \cref{fig: autoregressive vrnn srnn cwvae 1L graphs} and appendix~\cref{app: additional graphical models}.

\paragraph{Variational recurrent neural network (VRNN)}
The VRNN \cite{chung_recurrent_2015} is essentially a VAE per timestep. As seen in \cref{fig: autoregressive vrnn srnn cwvae 1L graphs}, it is conditioned on the hidden state of a Gated Recurrent Unit (GRU, \textcite{cho_properties_2014}) with state transition $\vd_{t} = f([\vx_{t-1}, \vz_{t-1}], \vd_{t-1})\in\mathbb{R}^{D_d}$ where $[\cdot, \cdot]$ denotes concatenation. 
The joint and variational posterior distributions are given by
\begin{equation}
    p(\vx,\vz) = \prod_{t=1}^{T} p(\vx_t|\vx_{<t},\vz_{\leq t}) p(\vz_t|\vx_{<t},\vz_{<t}) \, ,\quad 
    q(\vz|\vx) = \prod_{t=1}^{T} q(\vz_t|\vx_{\leq t},\vz_{<t}) \enspace . \label{eq: vrnn joint and inference}
\end{equation}
%
\paragraph{Stochastic recurrent neural network (SRNN)}
The SRNN \cite{fraccaro_sequential_2016} is similar to the VRNN and its joint can be written as in \eqref{eq: vrnn joint and inference}. Contrary to the VRNN, the SRNN has GRU state transitions that are independent of $\vz_{1:T}$ such that $\vd_{t} = f(\vx_{t-1}, \vd_{t-1})$. Furthermore, the SRNN conditions on the full observed sequence for inference, $q(\vz_t|\vx_{1:T},\vz_{t-1})$ via a second GRU with transition $\va_t = g([\vx_t,\vd_t], \va_{t+1})$. Then $\vz_t$ is inferred from $\va_t$ and $\vz_{t-1}$. This better approximates the true posterior which can be shown to depend on the full observed sequence \cite{bayer_mind_2021}.

\paragraph{Stochastic temporal convolutional network (STCN)}
Contrary to VRNN and SRNN, the latent variables of the STCN are conditionally independent given $\vx_{1:T}$ since there are no transition functions connecting them over time.
Instead, a latent $\vz_t^{(l)}$ at layer $l$ is conditioned on the latent variable above it $\vz^{(l+1)}_t$ and a window of $\vx_{1:T}$ defined by $\mathcal{R}_t^{(l)} = \cbra{t-r_l+1,\dots,t}$ via a WaveNet encoder with receptive field $r_l$ at layer $l$ \cite{oord_wavenet_2016}. The joint and variational posterior are, 
\begin{equation}
\resizebox{.99\textwidth}{!}{%
$
p(\vx,\vz) = \prod\limits_{t=1}^{T} p(\vx_t|\vz_{{\scriptscriptstyle \mathcal{R}}_{t}}) \prod\limits_{l=1}^{L} p(\vz^{(l)}_{t}|\vx_{\scriptscriptstyle\mathcal{R}_{t-1}^{(l)}},\vz^{(l+1)}_{t}) \, , \,\, q(\vz|\vx) = \prod\limits_{t=1}^{T}\prod\limits_{l=1}^{L} q(\vz_{t}^{(l)}|\vx_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}},\vz_t^{(l+1)}) \, , \nonumber
$
}
\label{eq: stcn joint and inference}
\end{equation}
where $\vz_t^{(L+1)}:=\emptyset$ and $\vz=\vz_{1:T}^{(1:L)}$ for notational convenience and inference is done top-down \cite{sonderby_ladder_2016}. The observation model $p(\vx_t|\vz^{(1:L)}_{{\scriptscriptstyle \mathcal{R}}_{t}^{(1)}})$ is also parameterized by a WaveNet.

\paragraph{Clockwork variational autoencoder (CW-VAE)}
The CW-VAE \cite{saxena_clockwork_2021} is a hierarchical LVM originally introduced for video generation. As seen in figure~\cref{fig: autoregressive vrnn srnn cwvae 1L graphs}, it is autoregressive in the latent space but not in the observed space, contrary to the VRNN, SRNN and STCN. Additionally, each latent variable is updated only every $s_l$ timesteps, where $s_l$ is a layer-dependent stride and $s_1<s_2<\dots<s_L$. 
We define $\mathcal{J}_t := \{ l \,|\, t\in\mathcal{T}_l \}$ and $\mathcal{T}_l := \{t \in \bra{1,T} \,|\, (t-1)\,\text{mod}\, s_l = 0\}$. Then,
\begin{equation}
\resizebox{0.99\textwidth}{!}{%
$
    p(\vx,\vz) = 
    \prod\limits_{t=1}^{T} p(\vx_t | \vz_t^{(1)})
    \prod\limits_{l\in\mathcal{J}_t} p(\vz_t^{(l)} | \vz_{t-s_l}^{(l)}, \vz_t^{(l+1)}) \, , \,\,
    q(\vz | \vx) = \prod\limits_{t=1}^T \prod\limits_{l\in\mathcal{J}_t} q(\vz_t^{(l)} | \vz_{t-s_l}^{(l)}, \vz_t^{(l+1)}, \vx_{t:t+s_l}) \, .
    \nonumber
$
}
\label{eq: cwvae joint and inference}
\end{equation}
The original encoder and decoder are not directly applicable to speech, since the sampling rates of speech are much higher than those of video (e.g. $\SI{16000}{Hz}$ compared to  $\SI{30}{Hz}$). Hence, we use a convolutional ladder network similar to the STCN for inference and downsample the waveform.

\paragraph{Output distribution} Recent work on LVMs for speech modeling, including those considered here, often uses a Gaussian output distribution \cite{chung_recurrent_2015, fraccaro_sequential_2016,hsu_unsupervised_2017,lai_stochastic_2018,aksan_stcn_2019,zhu_s3vae_2020}. 
Since audio is a naturally continuous signal, this may seem like an appropriate modeling choice. However, common audio datasets are sampled at bit-depths of $\SI{16}{bit}$ (TIMIT, \citealt{garofolo_timit_1993}; LibriSpeech, \citealt{panayotov_librispeech_2015}). This results in a quantization gap between unique values of $1\times10^{-5}$ which increases the risk of an ill-posed problem by a likelihood that is unbounded from above unless the variance is lower bounded \cite{mattei_leveraging_2018}. As a result, \emph{reported likelihoods can be sensitive to hyperparameter settings and be hard to compare}. We discuss this phenomenon further in appendix~\cref{app: gaussian likelihood unboundedness discussion}.

In this work, we therefore benchmark models using a discretized mixture of logistics (DMoL) as output distribution. The DMoL was introduced for image modeling with autoregressive models \cite{salimans_pixelcnn_2017} but has become standard in other generative models \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}. Although continuous distributions can be used if the data is dequantized \cite{dinh_nice_2015,theis_note_2016,ho_flow_2019}, we do not consider this option here.

\section{Speech modeling benchmark}

\input{tables/results}

\paragraph{Data}
We train models on TIMIT \cite{garofolo_timit_1993} and randomly sample 5\% of the training split for validation. 
Both input and target are $\mu$-law PCM standardized to $\bra{-1, 1}$. 
We report on LibriSpeech \cite{panayotov_librispeech_2015} and linear PCM in appendix~\cref{app: additional likelihood results appendix}. We further describe datasets in appendix~\cref{app: dataset details}.

\paragraph{Likelihood}
We report likelihoods in units of bits per frame (bpf; lower is better) as this yields a more interpretable likelihood that has connections to information theory and compression \cite{shannon_mathematical_1948, townsend_practical_2019} compared to total likelihood in nats. For LVMs, we report the one-sample ELBO. The likelihoods can be seen in \cref{tab: likelihoods timit}. We describe how to convert to bpf in appendix~\cref{app: likelihood in bits per frame}.

\paragraph{Models}
We configure WaveNet, VRNN, SRNN, STCN and CW-VAE as in the original papers  with the choices described here. Specifically, WaveNet uses ten layers, five blocks and $D_c=96$ channels. VRNN and SRNN use latent dimension $D_z=256$ and an equal number of hidden units. The STCN is in the dense configuration and uses 256 convolution channels, $L=5$ layers and latent variables of dimensions $16, 32, 64, 128, 256$ from the top down. We also run a one-layered ablation with the same architecture but only one latent variable of dimension $256$ at the top. The CW-VAE has $L=1$ or $2$ latent variables of dimension 96 equal to the number of convolution channels and we let $s$ refer to $s_1$ and set $s_2=8s_1$. We also train an LSTM baseline \cite{hochreiter_long_1997} configured similar to the VRNN.
%
All models use a 10 component DMoL output distribution. All LVMs use diagonal covariance Gaussian priors and posteriors. We train and evaluate models with waveforms stacked similar to previous work \cite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019} with stack sizes of $s=1$, $s=64$ and $s=256$. 
Finally, we evaluate a per-frame discrete uniform distribution and a two-component DMoL fitted to the training set to estimate worst case performance. 
We further describe models and training in appendix~\cref{app: model architectures} and \cref{app: training details} and provide results with a Gaussian output distribution in appendix~\cref{app: additional likelihood results appendix}.

\paragraph{Results}
We present the model likelihoods for $s=1$ and $64$ in \cref{tab: likelihoods timit} and for $s=256$ in appendix~\cref{app: additional likelihood results appendix}. It is clear that deterministic autoregressive models are superior to LVMs at $s=1$, but inferior to them at $s=64$ and $256$. For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables to the ability of LVMs to model intra-step correlations \cite{lai_re-examination_2019}. It should be noted that at $s=1$, the STCN was numerically unstable and that the CW-VAE was computationally infeasible to train.
%
Surprisingly, the simple LSTM performs almost on par with WaveNet. The autoregressive STCN is the best-performing LVM. For both CW-VAE and STCN, increasing the number of latent variables in the hierarchy improves performance, similar to results in the image domain. 
%The best likelihoods correspond to about a 30\% reduction in bit rate which is inferior to FLAC indicating that significant gains in likelihood might be possible. 
The best performing LVMs, STCN and CW-VAE, are not yet scalable to $s=1$ resolution where the highest likelihoods are achieved. Hence, LVMs may be able to outperform autoregressive models at $s=1$ in the future.

\paragraph{Phoneme recognition}
Although the likelihood is a practical metric for model comparison, a high likelihood does not guarantee that a model has learned useful representations \cite{huszar_is_2017}. To assess the usefulness of the representations, we train a model to recognize phonemes on the TIMIT data. We compare the phoneme error rate (PER) of the model when using input representations obtained from different unsupervised models, trained on the full TIMIT dataset (3.7h) at $s=64$, as in \cref{tab: likelihoods timit}, and when using a Mel spectrogram (80 filterbanks, hop length 64, window size 128). The recognition model is a three-layered bidirectional LSTM with 256 hidden units. It is trained with the connectionist temporal classification (CTC) loss \cite{graves_connectionist_2006}. We report the PER in \cref{tab: phoneme recognition (PER)}.

As expected, Mel spectrograms perform well achieving 24.1\% PER using 3.7 hours of labeled data. However, the ASR trained on STCN representations outperforms the Mel spectrogram with a PER of 21.9\%. 
This indicates that unsupervised STCN representations are phonetically rich and potentially better suited for speech modeling than the engineered Mel spectrogram. 
When the amount of labeled data is reduced, LVM representations suffer slightly less than deterministic ones. Interestingly, representations from WaveNet are outperformed by all LVMs and the LSTM.

\subsubsection*{Acknowledgments}
This research was partially funded by the Innovation Fund Denmark via the Industrial PhD Programme (grant no.\@ 0153-00167B). JF and SH were funded in part by the Novo Nordisk Foundation (grant no.\@ NNF20OC0062606) via the Center for Basic Machine Learning Research in Life Science (MLLS, \hyperlink{https://www.mlls.dk}{https://www.mlls.dk}). JF was further funded by the Novo Nordisk Foundation (grant no.\@ NNF20OC0065611) and the Independent Research Fund Denmark (grant no.\@ 9131-00082B). SH was further funded by VILLUM FONDEN (15334) and the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement no. 757360).

\fi

### 2023-02-25 chapters/paper_hierarchical.tex
Add benchmarking paper

@@
     \centering
-    \includegraphics[width=0.7\columnwidth]{reconstructions-front-page-4-samples-2-recons2.pdf}
+    \includegraphics[width=0.7\columnwidth]{paper_hierarchical/reconstructions-front-page-4-samples-2-recons2.pdf}
     \caption[Reconstructions of a hierarchical VAE trained on FashionMNIST.]{%
@@
         Reconstruction quality of OOD data is comparable to in-distribution data, resulting in high likelihoods and poor OOD discrimination.
-        By sampling the $k$ bottom-most latent variables from the conditional prior distribution $p(\z_{\geq l}|\z_{>l})$ (latent reconstructions) instead of the approximate posterior $q(\z_{>l}|\z_{<l})$, the model reconstructs from the training distribution resulting in lower $p(\x|\z)$ for OOD data.
+        By sampling the $k$ bottom-most latent variables from the conditional prior distribution $p(\zb_{\geq l}|\zb_{>l})$ (latent reconstructions) instead of the approximate posterior $q(\zb_{>l}|\zb_{<l})$, the model reconstructs from the training distribution resulting in lower $p(\xb|\zb)$ for OOD data.
     }
@@
     \centering
-    \includegraphics[width=1\columnwidth]{feature-correlation-heatmap2.pdf}
+    \includegraphics[width=1\columnwidth]{paper_hierarchical/feature-correlation-heatmap2.pdf}
     \vspace{-5mm}
@@
     %\vspace{0.17cm}
-    \includegraphics[width=1\columnwidth]{biva_reconstructions.pdf}
+    \includegraphics[width=1\columnwidth]{paper_hierarchical/biva_reconstructions.pdf}
     % https://docs.google.com/drawings/d/1UXdA3c18Oek9_env1vEB9mqOAHZ-tu0ijWmvqmapWCk/edit?usp=sharing
@@
 The variational autoencoder (VAE) \cite{kingma_autoencoding_2014, rezende_stochastic_2014} is a framework for constructing deep generative models defined by an observed variable $\mathbf{x}$ and a stochastic latent variable $\mathbf{z}$.
-Typically, a neural network with parameters $\theta$ is chosen to parameterize the generative distribution $p_\theta(\x,\z)=p_\theta(\x|\z)p(\z)$, where the prior $p(\z)$ is commonly a standard Gaussian $\mathcal{N}(\0, \I)$.
-The true posterior $p(\z|\x)$ is generally not analytically tractable and is approximated by a variational distribution $q_\phi(\z|\x)$ parameterized via another neural network with parameters $\phi$.
-The approximate posterior $q_\phi(\z|\x)$ is most often  a diagonal covariance Gaussian.
-The model parameters $\theta$ and variational parameters $\phi$ are jointly optimized by maximizing the \textit{evidence lower bound} (ELBO),
+Typically, a neural network with parameters $\thetab$ is chosen to parameterize the generative distribution $p_\thetab(\xb,\zb)=p_\thetab(\xb|\zb)p(\zb)$, where the prior $p(\zb)$ is commonly a standard Gaussian $\mathcal{N}(\0, \Ib)$.
+The true posterior $p(\zb|\xb)$ is generally not analytically tractable and is approximated by a variational distribution $q_\phib(\zb|\xb)$ parameterized via another neural network with parameters $\phib$.
+The approximate posterior $q_\phib(\zb|\xb)$ is most often  a diagonal covariance Gaussian.
+The model parameters $\thetab$ and variational parameters $\phib$ are jointly optimized by maximizing the \textit{evidence lower bound} (ELBO),
 \begin{equation}\label{eq:elbo}
-    \log p_\theta(\x) \geq \mathbb{E}_{q_\phi(\z|\x)} \left[ \log \frac{p_\theta(\x,\z)}{q_\phi(\z|\x)} \right] \equiv \mathcal{L}(\x; \theta, \phi)\ .
+    \log p_\thetab(\xb) \geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}(\xb; \thetab, \phib)\ .
 \end{equation}
-For brevity, we will denote $\mathcal{L}(\x; \theta, \phi)$ as $\mathcal{L}(\x)$ or $\mathcal{L}$.
+For brevity, we will denote $\mathcal{L}(\xb; \thetab, \phib)$ as $\mathcal{L}(\xb)$ or $\mathcal{L}$.
 The reparameterization trick is used to backpropagate gradients through the stochastic latent variables with low variance.
@@
 The VAE is defined with a single latent variable which limits the ability to learn a high likelihood representation of complex input distributions, e.g.\ natural images.
-There exists a few complementary approaches to make the VAE more flexible: (i) model a more expressive variational distribution $q_\phi(\z|\x)$ or prior distribution $p_\theta(\z)$ \cite{rezende_variational_2015, kingma_improved_2016}, (ii) model a more expressive posterior distribution $p_\theta(\x|\z)$ e.g.
+There exists a few complementary approaches to make the VAE more flexible: (i) model a more expressive variational distribution $q_\phib(\zb|\xb)$ or prior distribution $p_\thetab(\zb)$ \cite{rezende_variational_2015, kingma_improved_2016}, (ii) model a more expressive posterior distribution $p_\thetab(\xb|\zb)$ e.g.
 with a autoregressive decoder \cite{oord_conditional_2016} and (iii) learn a deeper hierarchy of latent variables \cite{burda_importance_2016, sonderby_ladder_2016}.
@@
 \subsection{Hierarchical variational autoencoders}\label{sec:background-hie-VAE}
-Hierarchical VAEs are a family of probabilistic latent variable models which extends the basic VAE by introducing a hierarchy of $L$ latent variables $\z=\z_1, \dots, \z_L$.
-The most common generative model is defined from the top down as $p_\theta(\x|\z)=p(\x|\z_1)p_\theta(\z_1|\z_2)\cdots p_\theta(\z_{L-1}|\z_L)$.
+Hierarchical VAEs are a family of probabilistic latent variable models which extends the basic VAE by introducing a hierarchy of $L$ latent variables $\zb=\zb_1, \dots, \zb_L$.
+The most common generative model is defined from the top down as $p_\thetab(\xb|\zb)=p(\xb|\zb_1)p_\thetab(\zb_1|\zb_2)\cdots p_\thetab(\zb_{L-1}|\zb_L)$.
 The inference model can then be defined in two ways respectively referred to as \textit{bottom-up} \cite{burda_importance_2016}
 \begin{equation}
-    %q_\phi(\z|\x)=q_\phi(\z_1|\x)q_\phi(\z_2|\z_1)\cdots q_\phi(\z_L|\z_{L-1})
-    q_\phi(\z|\x) = q_\phi(\z_1|\x)\textstyle\prod_{i=2}^{L} q_\phi(\z_i|\z_{i-1})
+    %q_\phib(\zb|\xb)=q_\phib(\zb_1|\xb)q_\phib(\zb_2|\zb_1)\cdots q_\phib(\zb_L|\zb_{L-1})
+    q_\phib(\zb|\xb) = q_\phib(\zb_1|\xb)\textstyle\prod_{i=2}^{L} q_\phib(\zb_i|\zb_{i-1})
 \end{equation}
@@
 \begin{equation}
-    % q_\phi(\z|\x)=q_\phi(\z_{L}|\x)q_\phi(\z_{L-1}|\z_L)\cdots q_\phi(\z_2|\z_1),
-    q_\phi(\z|\x) = q_\phi(\z_L|\x)\textstyle\prod_{i=L-1}^{1} q_\phi(\z_{i}|\z_{i+1}) \ .
+    % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb_{L-1}|\zb_L)\cdots q_\phib(\zb_2|\zb_1),
+    q_\phib(\zb|\xb) = q_\phib(\zb_L|\xb)\textstyle\prod_{i=L-1}^{1} q_\phib(\zb_{i}|\zb_{i+1}) \ .
 \end{equation}
@@
 \subsection{A bound for semantic OOD detection}
-If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \cref{fig:reconstructions-fashionmnist}-\ref{fig:biva-reconstructions-celeba}, then $p_\theta(\x|\z_1)$ will be high for both in- and out-of-distribution data.
-Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phi(\z|\x)}[\log p_\theta(\x|\z_1)] - D_{\mathrm{KL}}( q_\phi(\z|\x) || p(\z))$ from \eqref{eq:elbo} relies on the KL-term for OOD detection.
-For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phi(\z|\x) || p(\z))$ can be expressed by a hierarchical sum% over the hierarchy
+If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \cref{fig:reconstructions-fashionmnist}-\ref{fig:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb_1)$ will be high for both in- and out-of-distribution data.
+Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb_1)] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) || p(\zb))$ from \eqref{eq:elbo} relies on the KL-term for OOD detection.
+For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) || p(\zb))$ can be expressed by a hierarchical sum% over the hierarchy
 % \begin{multline}
-%     \mathcal{L}(\x) =  \mathbb{E}_{q_\phi(\z|\x)} \Big[ \log p_\theta(\x|\z_1) \\
-%                      + \textstyle\sum_{i=1}^{L-1} \log \frac{p_\theta(\z_i|\z_{i+1})}{q_\phi(\z_i|\z_{i-1})} 
-%                      + \log \frac{p_\theta(\z_L)}{q_\phi(\z_L|\z_{L-1})} \Big] .
+%     \mathcal{L}(\xb) =  \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \log p_\thetab(\xb|\zb_1) \\
+%                      + \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} 
+%                      + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
 % \end{multline}
 % \begin{multline}
-%     D_{\mathrm{KL}}( q(\z|\x) || p(\z)) = \\
-%     \mathbb{E}_{q_\phi(\z|\x)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\theta(\z_i|\z_{i+1})}{q_\phi(\z_i|\z_{i-1})} + \log \frac{p_\theta(\z_L)}{q_\phi(\z_L|\z_{L-1})} \Big] .
+%     D_{\mathrm{KL}}( q(\zb|\xb) || p(\zb)) = \\
+%     \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
 % \end{multline}
 \begin{equation}
-    \mathbb{E}_{q_\phi(\z|\x)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\theta(\z_i|\z_{i+1})}{q_\phi(\z_i|\z_{i-1})} + \log \frac{p_\theta(\z_L)}{q_\phi(\z_L|\z_{L-1})} \Big] \ .
+    \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] \ .
 \end{equation}
-In general, the absolute log-ratios grow with $\mathrm{dim}(\z_i)$ as the individual log probability terms are computed by summing over the dimensionality of $\z_i$.
-This means that the value of the KL-term is dominated by terms where $\z_i$ is high-dimensional. We refer to Appendix C for a more detailed argument.
+In general, the absolute log-ratios grow with $\mathrm{dim}(\zb_i)$ as the individual log probability terms are computed by summing over the dimensionality of $\zb_i$.
+This means that the value of the KL-term is dominated by terms where $\zb_i$ is high-dimensional. We refer to Appendix C for a more detailed argument.
 Since hierarchical VAEs are generally constructed with a bottleneck type structure, the terms corresponding to latent variables towards the top of the hierarchy will have a vanishing influence on the value of the KL-term.
@@
 \begin{equation}\label{eq:biva->k}
-    \mathcal{L}^{>k} = \mathbb{E}_{p_\theta(\z_{\leq k}|\z_{>k}) q_\phi(\z_{>k}|\x)} \left[ \log \frac{p_\theta(\x|\z)p_\theta(\z_{>k})}{q_\phi(\z_{>k}|\x)} \right]
+    \mathcal{L}^{>k} = \mathbb{E}_{p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p_\thetab(\zb_{>k})}{q_\phib(\zb_{>k}|\xb)} \right]
 \end{equation}
@@
 We note that $\mathcal{L}^{>0}$ is the regular ELBO (\ref{eq:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
-The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\z_1,\dots,\z_l \sim p_\theta(\z_{\leq k}|\z_{>k})$ and only the $L-k$ highest from the approximate posterior $\z_{k+1},\dots,\z_L \sim q_\phi(\z_{>k}|\x)$.
-Importantly, this has the effect that the data likelihood $p(\x|\z)$ is dependent on the approximate posterior through a latent variable $\z_{k+1}$ different from $\z_1$ for all $k \geq 1$.
-Thereby, the likelihood can be evaluated with a reconstruction from each of the latent variables $\z_k$ of the hierarchical VAE.
-Hence, we can now test how well the input $\x$ is reconstructed from each latent variable.
-The notation $\mathcal{L}^{>k}$ highlights that for latent variables $\z_{>k}$, the bound is the regular ELBO while for the latent variables $\z_{\leq k}$, the bound is evaluated using the (conditional) prior rather than the approximate posterior as the proposal distribution.
+The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\zb_1,\dots,\zb_l \sim p_\thetab(\zb_{\leq k}|\zb_{>k})$ and only the $L-k$ highest from the approximate posterior $\zb_{k+1},\dots,\zb_L \sim q_\phib(\zb_{>k}|\xb)$.
+Importantly, this has the effect that the data likelihood $p(\xb|\zb)$ is dependent on the approximate posterior through a latent variable $\zb_{k+1}$ different from $\zb_1$ for all $k \geq 1$.
+Thereby, the likelihood can be evaluated with a reconstruction from each of the latent variables $\zb_k$ of the hierarchical VAE.
+Hence, we can now test how well the input $\xb$ is reconstructed from each latent variable.
+The notation $\mathcal{L}^{>k}$ highlights that for latent variables $\zb_{>k}$, the bound is the regular ELBO while for the latent variables $\zb_{\leq k}$, the bound is evaluated using the (conditional) prior rather than the approximate posterior as the proposal distribution.
 
@@
     % L0 = L^{>k}
-    LLR^{>k}(\x) = \mathcal{L}(\x) - \mathcal{L}^{>k}(\x) \ .
+    LLR^{>k}(\xb) = \mathcal{L}(\xb) - \mathcal{L}^{>k}(\xb) \ .
 \end{equation}
@@
 Since, empirically, $\mathcal{L}\geq\mathcal{L}^{>k}$, the ratio is always positive as is standard for likelihood ratio tests.
-A low value of $LLR^{>k}(\x)$ means that the ELBO and $\mathcal{L}^{>k}$ are almost equally tight for the data.
+A low value of $LLR^{>k}(\xb)$ means that the ELBO and $\mathcal{L}^{>k}$ are almost equally tight for the data.
 On the contrary, a high value indicates that $\mathcal{L}^{>k}$ is looser on the data than the ELBO; hence, the data may be OOD.
@@
 \begin{align}
-    \mathcal{L}      &= \log p_\theta(\x) - D_{\mathrm{KL}}\left( q_\phi(\z|\x) || p_\theta(\z|\x)\right), \label{eq:likelihoods-as-exact} \\ 
-    \mathcal{L}^{>k} &= \log p_\theta(\x) - D_{\mathrm{KL}}\left( p_\theta(\z_{\leq k }|\z_{>k}) q_\phi(\z_{>k}|\x) || p_\theta(\z|\x)\right) \nonumber \ .
+    \mathcal{L}      &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) || p_\thetab(\zb|\xb)\right), \label{eq:likelihoods-as-exact} \\ 
+    \mathcal{L}^{>k} &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k }|\zb_{>k}) q_\phib(\zb_{>k}|\xb) || p_\thetab(\zb|\xb)\right) \nonumber \ .
 \end{align}
-Subtracting these cancel out the two data likelihood terms $\log p_\theta(\x)$ and only the KL-divergences from the approximate to the true posterior remain,
+Subtracting these cancel out the two data likelihood terms $\log p_\thetab(\xb)$ and only the KL-divergences from the approximate to the true posterior remain,
 \begin{align}\label{eq:llr-as-kls}
-    LLR^{>k}(\x) &= - D_{\mathrm{KL}}\left( q_\phi(\z|\x) || p_\theta(\z|\x)\right) \\
-                 &\quad + D_{\mathrm{KL}}\left( p_\theta(\z_{\leq k}|\z_{>k}) q_\phi(\z_{>k}|\x) || p_\theta(\z|\x)\right) \ . \notag
+    LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) || p_\thetab(\zb|\xb)\right) \\
+                 &\quad + D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb) || p_\thetab(\zb|\xb)\right) \ . \notag
 \end{align}
 
-Hence, it is clear that compared to the likelihood bound $\mathcal{L}^{>k}$, this likelihood-ratio measures divergence exclusively in the latent space whereas $\mathcal{L}^{>k}$ includes the $\log p_\theta(\x)$ term similar to the ELBO.
+Hence, it is clear that compared to the likelihood bound $\mathcal{L}^{>k}$, this likelihood-ratio measures divergence exclusively in the latent space whereas $\mathcal{L}^{>k}$ includes the $\log p_\thetab(\xb)$ term similar to the ELBO.
 Therefore, the $LLR^{>k}$ score should be an improved method for semantic OOD detection compared to $\mathcal{L}^{>k}$.
@@
 \begin{equation}\label{eq:iw-bound}
-    \mathcal{L}_{S} = \mathbb{E}_{q(\z|\x)}\left[ \log \frac{1}{N} \sum_{s=1}^{S} \frac{p(\x, \z^{(s)})}{q(\z^{(s)}|\x)} \right] \ ,
+    \mathcal{L}_{S} = \mathbb{E}_{q(\zb|\xb)}\left[ \log \frac{1}{N} \sum_{s=1}^{S} \frac{p(\xb, \zb^{(s)})}{q(\zb^{(s)}|\xb)} \right] \ ,
 \end{equation}
-then, in the limit $S\rightarrow\infty$, we have $\mathcal{L}_{S} \rightarrow \log p_\theta(\x)$ and the likelihood ratio reduces to
+then, in the limit $S\rightarrow\infty$, we have $\mathcal{L}_{S} \rightarrow \log p_\thetab(\xb)$ and the likelihood ratio reduces to
 \begin{equation}\label{eq:llr-as-kls-iwae-reduced}
-    LLR^{>k}_{S}(\x) \rightarrow D_{\mathrm{KL}}( p(\z_{\leq k}|\z_{>k}) q(\z_{>k}|\x) || p(\z|\x))
+    LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq k}|\zb_{>k}) q(\zb_{>k}|\xb) || p(\zb|\xb))
 \end{equation}
 which, in practice, is well-approximated for a finite $S$.
-We expect this importance weighted likelihood ratio to monotonically improve upon the one in (\ref{eq:llr-as-kls}) as $S$ increases and the KL-divergence in the regular ELBO that contains terms for which $\z_i$ is high-dimensional goes to zero.
+We expect this importance weighted likelihood ratio to monotonically improve upon the one in (\ref{eq:llr-as-kls}) as $S$ increases and the KL-divergence in the regular ELBO that contains terms for which $\zb_i$ is high-dimensional goes to zero.
 
@@
 We note that $\text{Var}(\widehat{LLR}^{>k}) = \text{Var}(\hat{\mathcal{L}}) + \text{Var}(\hat{\mathcal{L}}^{>k}) - 2\, \text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$.
-Since $\log p_\theta(\x)$ and part of the KL divergence are identical in the expressions of $\mathcal{L}$ and $\mathcal{L}^{>k}$ we expect $\text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$ to be positive which reduces the total variance.
+Since $\log p_\thetab(\xb)$ and part of the KL divergence are identical in the expressions of $\mathcal{L}$ and $\mathcal{L}^{>k}$ we expect $\text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$ to be positive which reduces the total variance.
 Empirical results indeed show that $\text{Var}(\widehat{LLR}^{>k})$ is larger than $\text{Var}(\hat{\mathcal{L}})$ but smaller than $\text{Var}(\hat{\mathcal{L}}^{>k})$.
@@
         % nodes
-        \node[obs] (x_inf) {$\x$};%
-        \node[latent,above=.75cm of x_inf](z1_inf){$\z_1$}; %
-        \node[latent,above=.75cm of z1_inf](z2_inf){$\z_2$}; %
-        % \node[latent,above=.75cm of z2_inf](z3_inf){$\z_3$}; %
-        \node[above=of z2_inf, yshift=-1.cm] (phi) {$q_\phi(\z|\x)$};
+        \node[obs] (x_inf) {$\xb$};%
+        \node[latent,above=.75cm of x_inf](z1_inf){$\zb_1$}; %
+        \node[latent,above=.75cm of z1_inf](z2_inf){$\zb_2$}; %
+        % \node[latent,above=.75cm of z2_inf](z3_inf){$\zb_3$}; %
+        \node[above=of z2_inf, yshift=-1.cm] (phi) {$q_\phib(\zb|\xb)$};
 
@@
         % nodes$
-        \node[obs,right=0.75cm of x_inf] (x_gen) {$\x$};%
-        \node[latent,above=.75cm of x_gen](z1_gen){$\z_1$}; %
-        \node[latent,above=.75cm of z1_gen](z2_gen){$\z_2$}; %
-        % \node[latent,above=.75cm of z2_gen](z3_gen){$\z_3$}; %
-        \node[above=of z2_gen, yshift=-1.cm] (theta) {$p_\theta(\x,\z)$};
+        \node[obs,right=0.75cm of x_inf] (x_gen) {$\xb$};%
+        \node[latent,above=.75cm of x_gen](z1_gen){$\zb_1$}; %
+        \node[latent,above=.75cm of z1_gen](z2_gen){$\zb_2$}; %
+        % \node[latent,above=.75cm of z2_gen](z3_gen){$\zb_3$}; %
+        \node[above=of z2_gen, yshift=-1.cm] (theta) {$p_\thetab(\xb,\zb)$};
 
@@
     }
-    \caption[Inference and generative models for a bottom-up hierarchical VAS]{The inference and generative models, $q_\phi$ and $p_\theta$, for an $L=2$ layered bottom-up hierarchical VAE as the one used in our experiments.
+    \caption[Inference and generative models for a bottom-up hierarchical VAS]{The inference and generative models, $q_\phib$ and $p_\thetab$, for an $L=2$ layered bottom-up hierarchical VAE as the one used in our experiments.
     Dashed lines indicate deterministic skip connections which are employed in both networks.
@@
 \paragraph{Models} For each OOD task, we train a simple bottom-up hierarchical VAE with $L$ stochastic layers which we will refer to as ``HVAE''.
-To alleviate posterior collapse we include skip-connections that connect $\z_i$ to $\z_{i+2}$ for $i\in\{0, L-2\}$ and $\z_0\equiv\x$ in both the inference and generative models \cite{dieng_avoiding_2019} and employ the \textit{free bits} scheme with $\lambda=2$ \cite{kingma_improved_2016}.
+To alleviate posterior collapse we include skip-connections that connect $\zb_i$ to $\zb_{i+2}$ for $i\in\{0, L-2\}$ and $\zb_0\equiv\xb$ in both the inference and generative models \cite{dieng_avoiding_2019} and employ the \textit{free bits} scheme with $\lambda=2$ \cite{kingma_improved_2016}.
 We use weight-normalization \cite{salimans_weight_2016} on all weights and residual networks in the deterministic paths. 
@@
 We use $L=3$ for grey-scale images and $L=4$ for natural images.
-For CIFAR/SVHN, we also train a BIVA model \cite{maaloe_biva_2019} with $L=10$ and similar configuration as used by the original paper\footnote{Source code available at \url{github.com/larsmaaloee/BIVA} and \url{github.com/vlievin/biva-pytorch}}.
-All models are trained by optimizing the ELBO in (\ref{eq:elbo}).
-We implement our models in PyTorch \cite{paszke_automatic_2017}\footnote{Source code available at \url{github.com/jakobhavtorn/hvae-oodd}}.
+% For CIFAR/SVHN, we also train a BIVA model \cite{maaloe_biva_2019} with $L=10$ and similar configuration as used by the original paper\footnote{Code available at \url{github.com/larsmaaloee/BIVA} and \url{github.com/vlievin/biva-pytorch}}.
+% All models are trained by optimizing the ELBO in (\ref{eq:elbo}). We implement our models in PyTorch \cite{paszke_automatic_2017}\footnote{Code available at \url{github.com/jakobhavtorn/hvae-oodd}}.
 Full model details are in the Appendix.
@@
     }
-    \caption{
+    \caption[Average bits per dimension of different datasets for generative models trained on FashionMNIST and CIFAR10]{%
     Average bits per dimension of different datasets for models trained on FashionMNIST and CIFAR10.
@@
 Binary classiﬁer {\cite{ren_likelihood_2019}}                              & $0.455$ & $0.505$ & $0.886$ \\ % 6
-$p(\hat{y} | \x)$ with OOD as noise class {\cite{ren_likelihood_2019}}     & $0.877$ & $0.871$ & $0.195$ \\ % 7
-$p(\hat{y} | \x)$ with calibration on OOD {\cite{ren_likelihood_2019}}     & $0.904$ & $0.895$ & $0.139$ \\ % 8
+$p(\hat{y} | \xb)$ with OOD as noise class {\cite{ren_likelihood_2019}}     & $0.877$ & $0.871$ & $0.195$ \\ % 7
+$p(\hat{y} | \xb)$ with calibration on OOD {\cite{ren_likelihood_2019}}     & $0.904$ & $0.895$ & $0.139$ \\ % 8
 Input complexity ($S$, Glow) \cite{hendrycks_deep_2019}                    & $0.998$ & - & - \\
@@
          \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
-$p(\hat{y} | \x)$ {\cite{ren_likelihood_2019, hendrycks_baseline_2017}}                        & $0.734$ & $0.702$ & $0.506$ \\
-Entropy of $p(y | \x)$ {\cite{ren_likelihood_2019}}                        & $0.746$ & $0.726$ & $0.448$ \\
+$p(\hat{y} | \xb)$ {\cite{ren_likelihood_2019, hendrycks_baseline_2017}}                        & $0.734$ & $0.702$ & $0.506$ \\
+Entropy of $p(y | \xb)$ {\cite{ren_likelihood_2019}}                        & $0.746$ & $0.726$ & $0.448$ \\
 ODIN {\cite{ren_likelihood_2019, liang_enhancing_2018}}                                       & $0.752$ & $0.763$ & $0.432$ \\
@@
         \centering
-        \includegraphics[width=1\columnwidth]{densities-FashionMNIST-test-MNIST-test-bpp-k-0_sohau.pdf}
+        \includegraphics[width=1\columnwidth]{paper_hierarchical/densities-FashionMNIST-test-MNIST-test-bpp-k-0_sohau.pdf}
         \caption{}
@@
         \centering
-        \includegraphics[width=1\columnwidth]{densities-FashionMNIST-test-MNIST-test-bpp-k-2_sohau.pdf}
+        \includegraphics[width=1\columnwidth]{paper_hierarchical/densities-FashionMNIST-test-MNIST-test-bpp-k-2_sohau.pdf}
         \caption{}
@@
         \centering
-        \includegraphics[width=1\columnwidth]{densities-FashionMNIST-test-MNIST-test-LLR-0-1_sohau.pdf}
+        \includegraphics[width=1\columnwidth]{paper_hierarchical/densities-FashionMNIST-test-MNIST-test-LLR-0-1_sohau.pdf}
         \caption{}
@@
     \end{subfigure}
-    \caption{
+    \caption[Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood and $\mathcal{L}^{>2}$ bound]{%
     Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood \subref{fig:FMNIST-elbo-k0}, the $\mathcal{L}^{>2}$ bound \subref{fig:FMNIST-elbo-k2} and the $LLR^{>1}$ score \subref{fig:FMNIST-llr}.
@@
     \centering
-    \begin{subfigure}[l]{0.48\columnwidth}
-        \includegraphics[width=1\columnwidth]{roc-FashionMNIST-test-MNIST-test-ll-and-llr-IW250_sohau.pdf}
+    \begin{subfigure}[l]{0.495\columnwidth}
+        \includegraphics[width=1\columnwidth]{paper_hierarchical/roc-FashionMNIST-test-MNIST-test-ll-and-llr-IW250_sohau.pdf}
         \caption{}
@@
     \hfill
-    \begin{subfigure}[r]{0.48\columnwidth}
-        \includegraphics[width=1\columnwidth]{roc-biva-CIFAR10-SVHN-ll-and-llr_sohau.pdf}
+    \begin{subfigure}[r]{0.495\columnwidth}
+        \includegraphics[width=1\columnwidth]{paper_hierarchical/roc-biva-CIFAR10-SVHN-ll-and-llr_sohau.pdf}
         \caption{}
@@
     \end{subfigure}
-    \caption{%
+    \caption[ROC curves for out of distribution detection (MNIST/FashionMNIST and SVHN/CIFAR10)]{%
         ROC curves with AUROC score for detecting \subref{fig:FMNIST-roc-llr} MNIST as OOD with the HVAE model trained on FashionMNIST and \subref{fig:CIFAR10-roc-llr} SVHN as OOD with the BIVA model trained on CIFAR10.
@@
 Furthermore, our method performs similarly to the background constrative likelihood ratio method of \textcite{ren_likelihood_2019} on FashionMNIST/MNIST but contrary to the failure of that method on CIFAR10/SVHN reported by \cite{xiao_likelihood_2020}, our method performs very well on this task too.
-Our approach outperforms all supervised approaches that use in-distribution labels or synthetic examples of OOD data derived from the in-distribution data including ODIN \cite{liang_enhancing_2018} and the predictive distribution of a classifier $p(\hat{y}|\x)$ trained and evaluated in various ways (see \textcite{ren_likelihood_2019}).
+Our approach outperforms all supervised approaches that use in-distribution labels or synthetic examples of OOD data derived from the in-distribution data including ODIN \cite{liang_enhancing_2018} and the predictive distribution of a classifier $p(\hat{y}|\xb)$ trained and evaluated in various ways (see \textcite{ren_likelihood_2019}).
 

### 2023-02-25 frontmatter/notation.tex
Add benchmarking paper

@@
 \nomenclature[A, 01]{$a$}{A scalar; integer or real}
-\nomenclature[A, 02]{$\a$}{A column vector}
-\nomenclature[A, 03]{$\A$}{A matrix}
-\nomenclature[A, 04]{$\text{diag}(\a)$}{A square diagonal matrix of the elements of $\a$}
+\nomenclature[A, 02]{$\ab$}{A column vector}
+\nomenclature[A, 03]{$\Ab$}{A matrix}
+\nomenclature[A, 04]{$\text{diag}(\ab)$}{A square diagonal matrix of the elements of $\ab$}
 \nomenclature[A, 05]{$\e$}{A vector of ones, $\e=\bmat{1 & \cdots & 1}\transpose$}
-\nomenclature[A, 06]{$\I$}{The identity matrix, $\I=\text{diag}\pa{\e}$}
+\nomenclature[A, 06]{$\Ib$}{The identity matrix, $\Ib=\text{diag}\pa{\e}$}
 
@@
 
-\nomenclature[C, 01]{$a_i$}{Element $i$ of vector $\a$ with first index $1$}
-\nomenclature[C, 02]{$A_{i,j}$}{Element $i,j$ of matrix $\A$. Comma omitted when unambiguous.}
-\nomenclature[C, 03]{$A_{i,:}$}{Row $i$ of matrix $\A$ (a row vector)}
-\nomenclature[C, 04]{$A_{:,j}$}{Column $j$ of matrix $\A$ (a column vector)}
-\nomenclature[C, 05]{$\a^\bra{l}$}{Neural network layer indexing; a vector in layer $l$}
-\nomenclature[C, 06]{$\a^\vbra{t}$}{Recurrent neural network sequence indexing; vector at timestep $t$ in sequence $\cbra{\a^\vbra{t}}_{t=1}^T$}
+\nomenclature[C, 01]{$a_i$}{Element $i$ of vector $\ab$ with first index $1$}
+\nomenclature[C, 02]{$A_{i,j}$}{Element $i,j$ of matrix $\Ab$. Comma omitted when unambiguous.}
+\nomenclature[C, 03]{$A_{i,:}$}{Row $i$ of matrix $\Ab$ (a row vector)}
+\nomenclature[C, 04]{$A_{:,j}$}{Column $j$ of matrix $\Ab$ (a column vector)}
+\nomenclature[C, 05]{$\ab^\bra{l}$}{Neural network layer indexing; a vector in layer $l$}
+\nomenclature[C, 06]{$\ab^\vbra{t}$}{Recurrent neural network sequence indexing; vector at timestep $t$ in sequence $\cbra{\ab^\vbra{t}}_{t=1}^T$}
 
-\nomenclature[D, 01]{$\a\transpose$}{Transpose of column vector $\a$ to give row vector}
-\nomenclature[D, 02]{$\A\transpose$}{Transpose of matrix $\A$}
-\nomenclature[D, 03]{$\a\transpose\b$}{Inner product of $\a$ and $\b$, $\a\transpose\b=\a\cdot\b$}
-\nomenclature[D, 04]{$\a\b\transpose$}{Outer product of $\a$ and $\b$}
-\nomenclature[D, 05]{$\A\B$}{Matrix product of $\A$ and $\B$}
-\nomenclature[D, 06]{$\A*\B$}{Convolution (actually cross-correlation) of $\A$ and $\B$}
-\nomenclature[D, 07]{$\A\odot\B$}{Elementwise, or Hadamard, product of $\A$ and $\B$}
-\nomenclature[D, 08]{$\det{\A}$}{Determinant of matrix $\A$}
-\nomenclature[D, 09]{$\norm{\a}$}{$L^2$ norm of $\a$}
-\nomenclature[D, 10]{$\norm{\a}_p$}{$L^p$ norm of $\a$}
-\nomenclature[D, 11]{$\Tr{\A}$}{Trace of matrix $\A$, $\Tr{\A}=\sum_i A_{i,i}$}
+\nomenclature[D, 01]{$\ab\transpose$}{Transpose of column vector $\ab$ to give row vector}
+\nomenclature[D, 02]{$\Ab\transpose$}{Transpose of matrix $\Ab$}
+\nomenclature[D, 03]{$\ab\transpose\bb$}{Inner product of $\ab$ and $\bb$, $\ab\transpose\bb=\ab\cdot\bb$}
+\nomenclature[D, 04]{$\ab\bb\transpose$}{Outer product of $\ab$ and $\bb$}
+\nomenclature[D, 05]{$\Ab\Bb$}{Matrix product of $\Ab$ and $\Bb$}
+\nomenclature[D, 06]{$\Ab*\Bb$}{Convolution (actually cross-correlation) of $\Ab$ and $\Bb$}
+\nomenclature[D, 07]{$\Ab\odot\Bb$}{Elementwise, or Hadamard, product of $\Ab$ and $\Bb$}
+\nomenclature[D, 08]{$\det{\Ab}$}{Determinant of matrix $\Ab$}
+\nomenclature[D, 09]{$\norm{\ab}$}{$L^2$ norm of $\ab$}
+\nomenclature[D, 10]{$\norm{\ab}_p$}{$L^p$ norm of $\ab$}
+\nomenclature[D, 11]{$\Tr{\Ab}$}{Trace of matrix $\Ab$, $\Tr{\Ab}=\sum_i A_{i,i}$}
 
@@
 \nomenclature[E, 02]{$\pd{y}{x}$}{Partial derivative of $y$ with respect to $x$}
-\nomenclature[E, 03]{$\nabla_\x$}{Del, nabla or Laplacian operator, $\nabla_\x = \bmat{\pd{}{x_1} & \cdots & \pd{}{x_n}}^\text{T}, \; \x\in\mathbb{R}^{n}$}
-\nomenclature[E, 04]{$\nabla_\x y$}{Gradient of $y$ with respect to $\x$, $\nabla_\x y = \bmat{\pd{y}{x_1} & \cdots & \pd{y}{x_n}}^\text{T}, \; \x\in\mathbb{R}^{n}$}
-\nomenclature[E, 05]{$\H_\x$}{Hessian operator, $\H_\x = \nabla_\x\nabla_\x\transpose = \bmat{
+\nomenclature[E, 03]{$\nabla_\xb$}{Del, nabla or Laplacian operator, $\nabla_\xb = \bmat{\pd{}{x_1} & \cdots & \pd{}{x_n}}^\text{T}, \; \xb\in\mathbb{R}^{n}$}
+\nomenclature[E, 04]{$\nabla_\xb y$}{Gradient of $y$ with respect to $\xb$, $\nabla_\xb y = \bmat{\pd{y}{x_1} & \cdots & \pd{y}{x_n}}^\text{T}, \; \xb\in\mathbb{R}^{n}$}
+\nomenclature[E, 05]{$\H_\xb$}{Hessian operator, $\H_\xb = \nabla_\xb\nabla_\xb\transpose = \bmat{
 	    \pd[2]{}{x_1} & \frac{\partial^2}{\partial x_1 \partial x_2} & \cdots & \frac{\partial^2}{\partial x_1 \partial x_n}\\
@@
 	    }$}
-\nomenclature[E, 06]{$\H_\x y$}{Hessian matrix of $y$ with respect to $\x$; matrix of all second order partial derivatives}
-\nomenclature[E, 07]{$\nabla^2_\x$}{Laplacian operator, $\nabla^2_\x=\nabla_\x\transpose\nabla_\x = \nabla_\x\cdot\nabla_\x=\sum_i \pd[2]{}{x_i}=\Tr{\H_\x}$}
-\nomenclature[E, 08]{$\nabla^2_\x y$}{Laplacian of $y$ with respect to $\x$; sum of the unmixed second order partial derivatives}
-\nomenclature[E, 09]{$\pd{\y}{\x}$}{Jacobian matrix of $\y$ w.r.t. $\x$, $\pd{\y}{\x} = \bmat{
-	    \nabla_\x y_1 &
+\nomenclature[E, 06]{$\H_\xb y$}{Hessian matrix of $y$ with respect to $\xb$; matrix of all second order partial derivatives}
+\nomenclature[E, 07]{$\nabla^2_\xb$}{Laplacian operator, $\nabla^2_\xb=\nabla_\xb\transpose\nabla_\xb = \nabla_\xb\cdot\nabla_\xb=\sum_i \pd[2]{}{x_i}=\Tr{\H_\xb}$}
+\nomenclature[E, 08]{$\nabla^2_\xb y$}{Laplacian of $y$ with respect to $\xb$; sum of the unmixed second order partial derivatives}
+\nomenclature[E, 09]{$\pd{\y}{\xb}$}{Jacobian matrix of $\y$ w.r.t. $\xb$, $\pd{\y}{\xb} = \bmat{
+	    \nabla_\xb y_1 &
 	    \cdots &
 	    \nabla_x y_n}^\text{T}, \; \y\in\mathbb{R}^{n}$; matrix of all possible first order partial derivatives}
-\nomenclature[E, 10]{$\int f(\x)\,\text{d}\x$}{Definite integral over the domain of $\x$}
-\nomenclature[E, 11]{$\int_\mathcal{D} f(\x)\,\text{d}\x$}{Definite integral over the domain $\mathcal{D}$}
+\nomenclature[E, 10]{$\int f(\xb)\,\text{d}\xb$}{Definite integral over the domain of $\xb$}
+\nomenclature[E, 11]{$\int_\mathcal{D} f(\xb)\,\text{d}\xb$}{Definite integral over the domain $\mathcal{D}$}
 
@@
 \nomenclature[F, 06]{$D_\text{KL}\pa{p,q}$}{Symmetrized Kullback-Leibler divergence of $p$ and $q$}
-\nomenclature[F, 07]{$\mathcal{N}\pa{\x\mid\mub,\Sigmab}$}{Gaussian distribution over $\x$ with mean $\mub$ and covariance $\Sigmab$}
+\nomenclature[F, 07]{$\mathcal{N}\pa{\xb\mid\mub,\Sigmab}$}{Gaussian distribution over $\xb$ with mean $\mub$ and covariance $\Sigmab$}
 

### 2023-02-25 appendices/appendix_benchmarking.tex
Add appendices for benchmarking and hierarchical

%!TEX root = ../thesis.tex

\chapter[benchmarking generative latent variable models for speech]{Benchmarking Generative Latent Variable Models for Speech}

\section{Reproducibility statement}\label{app: reproducibility statement}
The source code used for the work presented in this paper will be made available before the conference. 
This code provides all details, practical and otherwise, needed to reproduce the results in this paper including data preprocessing, model training, model likelihood and latent space evaluation.
The source code also includes scripts for downloading and preparing the LibriSpeech, LibriLight and TIMIT datasets. The LibriSpeech and LibriLight datasets are open source and can be downloaded with the preparation scripts. They are also available at \url{https://www.openslr.org/12} and \url{https://github.com/facebookresearch/libri-light}, respectively. The TIMIT dataset is commercial and must be purchased and downloaded from \url{https://catalog.ldc.upenn.edu/LDC93S1} before running the preparation script.

The stochastic latent variable models considered in this work do not provide an exact likelihood estimate nor an exact latent space representation. For the likelihood, they provide a stochastic lower bound and some variation in the reproduced likelihoods as well as latent representations must be expected between otherwise completely identical forward passes. This variance is fairly small in practice when averaging over large datasets such as those considered in this work. We seed our experiments to reduce the randomness to a minimum, but parts of the algorithms underlying the CUDA framework are stochastic for efficiency. To retain computational feasibility, we  do not run experiments with a deterministic CUDA backend.

\section{Ethics statement}\label{app: ethics statement}
The work presented here fundamentally deals with automated perception of speech and generation of speech. These applications of machine learning potentially raise a number of ethical concerns. For instance, the these models might see possibly adverse use in automated surveillance and generation of deep fakes. To counter some of these effects, this work has focused on openness by using publicly available datasets for model development and benchmarking. Additionally, the work will open source the source code used to create these results. 
Ensuring the net positive effect of the development of these technologies is and must continue to be an ongoing effort.

We do not associate any significant ethical concerns with the datasets used in this work. However, one might note that the TIMIT dataset has somewhat skewed distributions in terms of gender and race diversity. Specifically, the male to female ratio is about two to one while the vast majority of speakers are Caucasian. Such statistics might have an effect of some ethical concern on downstream applications derived from such a dataset as also highlighted in recent research \cite{koenecke_racial_2020}. In LibriSpeech, there is an approximately equal number of female and male speakers while the diversity in race is unknown to the authors.

\section{Datasets}\label{app: dataset details}
\paragraph{TIMIT} 
TIMIT \cite{garofolo_timit_1993} is a speech dataset which contains $\SI{16}{kHz}$ recordings of 630 speakers of eight major dialects of American English, each reading ten phonetically rich sentences. It amounts to 6300 total recordings splits approximately in 3.94 hours of audio for training and 1.43 hours of audio for testing. No speakers or sentences in the test set are in the training set. 
The full train and test subsets of TIMIT are as in previous work \cite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019}. 
We randomly sample 5\% of the training set to use as a validation set. 
TIMIT includes temporally aligned annotations of phonemes and words as well as speaker metadata such as gender, height, age, race, education level and dialect region \cite{garofolo_timit_1993}.

\paragraph{LibriSpeech and LibriLight}
The LibriSpeech dataset \cite{panayotov_librispeech_2015} consists of readings of public domain audio books amounting to approximately $\SI{1000}{h}$ of audio. The data is derived from the LibriVox project.
LibriLight \cite{kahn_libri-light_2020} is a subset of LibriSpeech created as an automatic speech transcription (ASR) benchmark with limited or no supervision.
We specifically train on the $\SI{100}{h}$ \textrm{train-clean-100} subset of LibriSpeech and the $\SI{10}{h}$ subset of LibriLight. 
In all cases we evaluate on all the test splits \textrm{dev-clean}, \textrm{dev-other}, \textrm{test-clean}, \textrm{test-other}.

Both datasets represent the audio as $\SI{16}{bit}$ pulse code modulation (PCM) sampled at $\SI{16000}{Hz}$.

\section{Model architectures}\label{app: model architectures}
This section details model architectures. See appendix~\cref{app: additional graphical models} for graphical models and appendix~\cref{app: training details} for training details.

\paragraph{WaveNet}
We implement WaveNet as described in the original work \cite{oord_wavenet_2016} but use a discretized mixture of logistics as the output distribution as also done in other work \cite{oord_parallel_2018}. Our WaveNet is not conditioned on any signal other than the raw waveform. The model applies the causal convolution directly to the raw waveform frames (i.e. one input channel). An alternative option that we did not examine is to replace the initial convolution with an embedding lookup with a learnable vector for each waveform frame value.

\paragraph{LSTM}
The LSTM baseline uses an MLP encoder to embed the waveform subsegment $\xb_{t:t+s-1}$ to a feature vector before feeding it to the LSTM cell. The encoder is similar to the parameterization of $\phi_\text{vrnn}^\text{enc}$ for the VRNN described above. The LSTM cell produces the hidden state $\db_t$ from $\xb_{t:t+s-1}$ and passes it to a decoder. Like the encoder, the decoder is parameterized like $\phi_\text{vrnn}^\text{dec}$ of the VRNN. It outputs the waveform predictions $\xb_{t+s:t+2s-1}$ from the hidden state $\db_t$. The LSTM model uses a single vanilla unidirectional LSTM cell.

\paragraph{VRNN}
We implement the VRNN as described in the original work \cite{chung_recurrent_2015} and verify that we can reproduce the original Gaussian likelihood TIMIT results. We replace the Gaussian output distribution with the DMoL.

\paragraph{SRNN}
We implement the VRNN as described in the original work \cite{fraccaro_sequential_2016} and verify that we can reproduce the original Gaussian likelihood TIMIT results. We replace the Gaussian output distribution with the DMoL.

\paragraph{CW-VAE} We implement the CW-VAE based on the original work \cite{saxena_clockwork_2021} but with some modifications also briefly described in section~\cref{sec: cwvae on speech}. We replace the encoder/decoder model architectures of the original work with architectures designed for waveform modeling. Specifically, the encoder and decoder are based on the Conv-TasNet \cite{luo_conv-tasnet_2019} and uses similar residual block structure. However, contrary to the Conv-TasNet, we require downsampling factors larger than two. In order to achieve this we use strides of two in the separable convolution of each block. With e.g. six blocks we hence get an overall stride of $2^6=64$. We can then add additional blocks with unit stride.
We also need to modify the residual connections that skip strided convolutions. Specifically, we replace the residual with a single convolution with stride equal to the stride used in the separable convolution. This convolution uses no nonlinearity and hence simply learns a local linear downsampling.

\paragraph{STCN} We implement the STCN as described in the original work \cite{aksan_stcn_2019} and verify that we can reproduce the original Gaussian likelihood TIMIT results. We replace the Gaussian output distribution with the DMoL. We use the best-performing version of the STCN reported in the original paper, namedly the ``STCN-dense" variant which conditions the observed variable on all five latent variables in the hierarchy. For the ablation experiment, we remove the bottom four latent variables. That is, we completely remove the corresponding four small densely connected networks that parameterize the prior and posterior distributions based on deterministic representations of the WaveNet encoder. We keep the top most prior and posterior networks and use them to parameterize a latent variable of $256$. This maintains the widest bottleneck of the model as well as almost all of the model's capacity.

\paragraph{ASR model}
The ASR model used for the phoneme recognition experiments is a three-layered bidirectional LSTM. We apply temporal dropout between the LSTM layers and also after the final layer. Temporal dropout works similar to regular dropout but samples the entries of the hidden state to mask only once and apply it to all timesteps, i.e. masking $\hb_{t}$ at vector index $i$ for all $t$ (and $i$). 
We mask by zeroing vector elements. We never mask the first timestep. 
We apply temporal dropout with masking probability of 0.3 for the 3.7h subset, 0.35 for the 1h subset and 0.4 for the 10m subset. 
The only difference in model architecture between the evaluation of different representations is the first affine transformation; from the dimensionality of the representation to the hidden state size of the LSTM. This gives rise to a very small difference in model capacity and parameter count which we find is negligible. We set the hidden unit size to 256.

\section{Training details}\label{app: training details}
\paragraph{Likelihood benchmark}
We implement all models and training scripts in PyTorch 1.9 \cite{paszke_automatic_2017}.
For both datasets we use the Adam optimizer \cite{kingma_adam_2015} with default parameters as given in PyTorch. We use learning rate $3\text{e}-4$ and no learning rate schedule.
We use PyTorch automatic mixed precision (AMP) to significantly reduce memory consumption. We did not observe any significant difference in final model performance compared to full ($\SI{32}{bit}$) precision.

We train stateful models (LSTM, VRNN, SRNN and CW-VAE) on the full sequence lengths padding batches with zeros when examples are not of equal length. We sample batches such that they consist of examples that are approximately the same length to minimize the amount of computation wasted on padding.

For $s=1$, we train stateless models (WaveNet, STCN) on random subsegments of the training examples and resample every epoch. This reduces memory requirements but does not bias the gradient. The subsequences are chosen to be of length 16000 which is larger than the receptive fields of the models and corresponds to one second of audio in TIMIT and LibriSpeech. 
For $s=64$ and $s=256$ we train the stateless models on the full example lengths similar to the stateful models since the receptive field is effectively $s$ times larger and the shorter sequence length reduces memory requirements.

In testing, we evaluate on the full sequences. Due to memory constrains, for LibriSpeech, we need to split the test examples into subsegments since the average sequence length in Librispeech is about 4 times longer than that of TIMIT. Hence, we do multiple forward passes per test example, one for each of several subsegments. We carry along the internal state for models that are autoregressive in training (LSTM, VRNN, SRNN, CW-VAE) and define segments to overlap according to model architecture.

\paragraph{Phoneme recognition}
The ASR experiment consists of two stages: 1) pre-training of the unsupervised model and 2) training of the ASR model. The pre-training is done as for the likelihood benchmark above. The ASR model is trained using the Adam optimizer \cite{kingma_adam_2015} with default parameters as given in PyTorch. We use learning rate $3\text{e}-4$ and no learning rate schedule.

For the spectrogram, WaveNet and the LSTM, we extract the representation only once and train the ASR model on these. Since the models are deterministic and do not parameterize distributions, this is the only option. For the LVMs, we resample the latent representation of a training example at every epoch. This is the most principled approach as these models parameterize probability distributions. Furthermore, using a single sample would be subject to artificially high variance in the representations while it is not straightforward to establish a sound mean representation for sequential models.

\section{Converting the likelihood to units of bits per frame}\label{app: likelihood in bits per frame}
Here we briefly describe how to compute a likelihood in units of bits per frame (bpf). In the main text, we use $\log$ to mean $\log_e$, but here we will be explicit. In general, conversion from nats to bits (i.e., from $\log_e$ to $\log_2$) is achieved by $\log_2(x) = \log_e(x)  / \log_2(e) $. Remember that $\log_2 p(\xb_{1:T})$ generally factorizes as $\sum_t \log_2 p(x_t|\cdot)$. In sequence modeling, it is important to remember that each example $\xb^{i}$ must be weighted differently according the sequence length of that specific example. This is in contrast to computing bits per dimension in the image domain where images in a dataset are usually of the same dimensions. Thus, we compute the log-likelihood in bits per frame over the entire dataset as
\begin{equation}
    \mathcal{L}(\xb^{i}) = \frac{1}{\sum_i T_i} \sum_i \sum_t \log_2 p(\xb_t^{i}) \enspace ,
\end{equation}
where $i$ denotes the example index, $T_i$ is the length of example $\xb^i$ in waveform frames and $t$ is the time index. If a single timestep $\xb_t^i$ represents multiple waveform frames stacked with some stack size $s$, it is important to note that the sum over $t$ only has $T_i/s$ elements. 
For the LVMs, the term $\log_2 p(\xb_t^{i})$ is lower bounded by the ELBO in \eqref{eq: elbo}.

\section{Additional likelihood results} \label{app: additional likelihood results appendix}

\paragraph{TIMIT, $\mu$-law, DMoL} We provide additional results on TIMIT with audio represented as $\mu$-law encoded PCM in \cref{tab: timit likelihoods dmol mu-law appendix}. Details are as presented in the main paper.

\paragraph{TIMIT, linear, DMoL}: We provide results on TIMIT with audio represented as linear PCM (raw PCM) in \cref{tab: timit likelihoods dmol linear appendix}. Except for the encoding, details are as for $\mu$-law encoded TIMIT

\paragraph{TIMIT, linear, Gaussian} We also provide some results on TIMIT with the audio instead represented as linear PCM (linearly encoded) and using Gaussian output distributions as has been done previously in the literature \cite{chung_recurrent_2015, fraccaro_sequential_2016, lai_stochastic_2018, aksan_stcn_2019}. We use $s=200$ for comparability to the previous work. We provide the results in \cref{tab: timit likelihoods gaussian appendix} and include likelihoods reported in the literature for reference. For our models, we use the same architectures as before but replace the discretized mixture of logistics with either a Gaussian distribution or a mixture of Gaussian distributions.

We constrain the variance of the Gaussians used with our models to be at least $\sigma^2_\text{min} = 0.01^2$ in order to avoid the variance going to zero, the likelihood going to infinity and optimization becoming unstable. The Gaussian standard deviation is clamped at minimum $0.001$ by \cite{aksan_stcn_2019}.

From \cref{tab: timit likelihoods gaussian appendix} we note that the performance of the CW-VAE with Gaussian output distribution when modeling linear PCM (i.e. not $\mu$-law encoded) does not compare as favorably to the other baselines as it did with the discretized mixture of logistics distribution. We hypothesize that this has to do with using a Gaussian output distribution in latent variable models which, as has been reported elsewhere \cite{mattei_leveraging_2018}, leads to a likelihood function that is unbounded above and can grow arbitrarily high. We discuss this phenomenon in further detail in section~\cref{app: gaussian likelihood unboundedness discussion}.

We specifically hypothesize that models that are autoregressive in the observed variable (VRNN, SRNN, Stochastic WaveNet, STCN) are well-equipped to utilize local smoothness to put very high density on the correct next value and that this in turn leads to a high degree of exploitation of the unboundedness of the likelihood. Not being autoregressive in the observed variable, the CW-VAE cannot exploit this local smoothness in the same way. Instead, the reconstruction is conditioned on a stochastic latent variable, $p(\xb_t|\zb^1_t)$, which introduces uncertainty and likely larger reconstruction variances.

\begin{table}[t!]
    \centering
    \begin{tabular}{ll|lr}
        \toprule
        $s$    & \bfseries Model           & \bfseries Configuration           & \bfseries $\mathcal{L}$ [bpf] \\
        \midrule
        1         & Uniform             & Uninformed            & 16.00 \\
        1         & DMoL                & Optimal               & 10.70 \\   % $mean[0.00], scale=[0.005]$
        -         & FLAC                & Linear PCM            & 8.582 \\
        \midrule
        $1$       & Wavenet             & $D_C=96$              & \textbf{7.246} \\
        $1$       & LSTM                & $D_d=256, L=1$        & 7.295 \\
        $1$       & VRNN                & $D_z=256$             & $\leq$7.316 \\
        $1$       & SRNN                & $D_z=256$             & $\leq$7.501 \\
        $1$       & STCN                & $D_z=256,L=5$         & $\leq$9.970 \\
        \midrule
        $64$      & WaveNet             & $D_c=96$              & 8.402 \\
        $64$      & LSTM                & $D_d=256, L=1$        & 8.357 \\
        $64$      & VRNN                & $D_z=256$             & $\leq$8.103 \\
        $64$      & SRNN                & $D_z=256$             & $\leq$8.036 \\
        $64$      & CW-VAE              & $D_z=96, L=1$         & $\leq$7.989 \\
        % $64$      & CW-VAE              & $D_z=96, L=2$         & $\leq$XX \\
        % $64$      & CW-VAE              & $D_z=96, L=3$         & $\leq$XX \\
        $64$      & STCN                & $D_z=256,L=5$  & $\leq$\textbf{7.768} \\
        \midrule
        $256$      & WaveNet             & $D_c=96$              & 9.018 \\
        $256$      & LSTM                & $D_d=256, L=1$        & 8.959 \\  % overfits slightly
        $256$      & VRNN                & $D_z=256$             & $\leq$8.739 \\
        $256$      & SRNN                & $D_z=256$             & $\leq$8.674 \\
        $256$      & CW-VAE              & $D_z=96, L=1$         & $\leq$8.406 \\
        % $256$      & CW-VAE              & $D_z=96, L=2$         & $\leq$XX \\
        % $256$      & CW-VAE              & $D_z=96, L=3$         & $\leq$XX \\
        $256$      & STCN                & $D_z=256,L=5$  & $\leq$\textbf{8.196} \\
        \bottomrule
    \end{tabular}
    \vspace{2mm}
    \caption{
    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit linear PCM}. The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
    }
    \label{tab: timit likelihoods dmol linear appendix}
\end{table}

\begin{table}[p]
    \centering
    \begin{tabular}{ll|lr}
        \toprule
        $s$    & \bfseries Model           & \bfseries Configuration           & \bfseries $\mathcal{L}$ [bpf] \\
        \midrule
        $1$       & Wavenet             & $D_C=16$              & 11.27 \\
        $1$       & Wavenet             & $D_C=24$              & 11.14 \\
        $1$       & Wavenet             & $D_C=32$              & 11.03 \\
        $1$       & Wavenet             & $D_C=96$              & 10.88 \\
        $1$       & Wavenet             & $D_C=128$             & 10.98 \\
        $1$       & Wavenet             & $D_C=160$             & 10.91 \\
        $1$       & LSTM                & $D_d=128, L=1$        & 11.40 \\
        $1$       & LSTM                & $D_d=256, L=1$        & 11.11 \\
        $1$       & VRNN                & $D_z=256$             & $\leq$11.09 \\
        $1$       & SRNN                & $D_z=256$             & $\leq$11.19 \\
        1 & STCN                & $D_z=256,L=5$               & $\leq$11.77 \\  % 39u25hyi
        \midrule
        $4$       & LSTM                & $D_d=256, L=1$        & 11.65 \\
        \midrule
        $16$      & LSTM                & $D_d=256, L=1$        & 12.54 \\
        $16$      & LSTM                & $D_d=256, L=2$        & 12.54 \\
        $16$      & LSTM                & $D_d=256, L=3$        & 12.44 \\
        \midrule
        $64$      & WaveNet             & $D_c=96$              & 13.30 \\
        $64$      & LSTM                & $D_d=96, L=1$         & 13.49 \\
        $64$      & LSTM                & $D_d=96, L=2$         & 13.46 \\
        $64$      & LSTM                & $D_d=96, L=3$         & 13.40 \\
        $64$      & LSTM                & $D_d=256, L=1$        & 13.27 \\
        $64$      & LSTM                & $D_d=256, L=2$        & 13.29 \\
        $64$      & LSTM                & $D_d=256, L=3$        & 13.31 \\
        $64$      & LSTM                & $D_d=512, L=1$        & 13.37 \\
        $64$      & LSTM                & $D_d=512, L=2$        & 13.37 \\
        $64$      & LSTM                & $D_d=512, L=3$        & 13.41 \\
        $64$      & VRNN                & $D_z=96$              & $\leq$12.93 \\
        $64$      & VRNN                & $D_z=256$             & $\leq$12.54 \\
        $64$      & SRNN                & $D_z=96$              & $\leq$12.87 \\
        $64$      & SRNN                & $D_z=256$             & $\leq$12.42 \\
        $64$      & CW-VAE              & $D_z=96, L=1$         & $\leq$12.44 \\
        $64$      & CW-VAE              & $D_z=96, L=2$         & $\leq$12.17 \\
        $64$      & CW-VAE              & $D_z=96, L=3$         & $\leq$12.15 \\
        $64$      & CW-VAE              & $D_z=256, L=2$        & $\leq$12.10 \\
        $64$ & STCN               & $D_z=256,L=1$               & $\leq$12.32 \\  % 2ss9fvxd
        $64$ & STCN               & $D_z=256,L=5$               & $\leq$11.78 \\
        \midrule
        $256$     & WaveNet             & $D_c=96$              & 14.11 \\
        $256$     & LSTM                & $D_d=256, L=1$        & 14.20 \\
        $256$     & LSTM                & $D_d=256, L=2$        & 14.17 \\
        $256$     & LSTM                & $D_d=256, L=3$        & 14.26 \\
        $256$     & VRNN                & $D_z=96$              & $\leq$13.51 \\
        $256$     & VRNN                & $D_z=256$             & $\leq$13.27 \\
        $256$     & SRNN                & $D_z=96$              & $\leq$13.28 \\
        $256$     & SRNN                & $D_z=256$             & $\leq$13.14 \\
        $256$     & CW-VAE              & $D_z=96, L=1$         & $\leq$13.11 \\
        $256$     & CW-VAE              & $D_z=96, L=2$         & $\leq$12.97 \\
        $256$     & CW-VAE              & $D_z=96, L=3$         & $\leq$12.87 \\
        $256$     & STCN                & $D_z=256,L=1$         & $\leq$13.07 \\  % 2h5ns3zw
        $256$     & STCN                & $D_z=256,L=5$         & $\leq$12.52 \\
        \bottomrule
    \end{tabular}
    \vspace{2mm}
    \caption{
    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}.
    }
    \label{tab: timit likelihoods dmol mu-law appendix}
\end{table}

\begin{table}[t!]
    \centering
    \begin{tabular}{ll|lr}
        \toprule
        $s$ & \bfseries Model           & \bfseries Configuration           & \bfseries $\mathcal{L}$ [nats] \\
        \midrule
        1 & WaveNet                                                               & Normal & 119656 \\
        1 & WaveNet                                                               & GMM-2  & 120699 \\
        1 & WaveNet                                                               & GMM-20 & 121681 \\
        \midrule
        % 200 & WaveNet                                                              & GMM    & XX.XX
        200 & WaveNet {\scriptsize \cite{aksan_stcn_2019}}                        & GMM-20    & 30188 \\
        200 & WaveNet {\scriptsize \cite{aksan_stcn_2019}}                        & Normal & -7443 \\
        200 & Stochastic WaveNet$^*$ {\scriptsize \cite{lai_stochastic_2018}}     & Normal & $\geq$72463\\
        % 200 & VRNN {\scriptsize \cite{chung_recurrent_2015}}                          & & \\
        200 & VRNN {\scriptsize \cite{chung_recurrent_2015}}                      & Normal & $\approx$28982\\
        % 200 & SRNN {\scriptsize \cite{fraccaro_sequential_2016}}                      & & \\
        200 & SRNN {\scriptsize \cite{fraccaro_sequential_2016}}                  & Normal & $\geq$60550 \\
        200 & STCN {\scriptsize \cite{aksan_stcn_2019}}                           & GMM-20 & $\geq$69195\\
        200 & STCN {\scriptsize \cite{aksan_stcn_2019}}                           & Normal & $\geq$64913\\
        200 & STCN-dense {\scriptsize \cite{aksan_stcn_2019}}                     & GMM-20 & $\geq$71386\\
        200 & STCN-dense {\scriptsize \cite{aksan_stcn_2019}}                     & Normal & $\geq$70294\\
        200 & STCN-dense-large {\scriptsize \cite{aksan_stcn_2019}}               & GMM-20 & $\geq$77438\\
        200 & CW-VAE$^*$                                         & $L=1, D_z=96$, Normal & $\geq$41629 \\  % https://wandb.ai/vseq/cwvae/runs/n8nu778t?workspace=user-jaksen
        % 200 & CW-VAE$^*$                                         & $L=2, D_z=96$, Normal & $\geq$2090 \\  % https://wandb.ai/vseq/cwvae/runs/1so36fii?workspace=user-jaksen
        \bottomrule
    \end{tabular}
    \vspace{2mm}
    \caption{
    Model likelihoods on \textbf{TIMIT} represented as \textbf{globally normalized 16 bit linear PCM}. Contrary to the other likelihoods reported in this paper, here they are given in units of nats and obtained by summing the likelihood over time and over all examples in the dataset and dividing by the total number of examples. In the table, Normal refers to using a Gaussian likelihood and GMM refers to using a  Gaussian Mixture Model likelihood with 20 components. Models with asterisks $^*$ are our implementations while remaining results are as reported in the referenced work.
    }
    \label{tab: timit likelihoods gaussian appendix}
\end{table}

\section{Additional discussion on Gaussian likelihoods in LVMs} \label{app: gaussian likelihood unboundedness discussion}
As noted in section~\cref{app: additional likelihood results appendix}, we constrain the variance of the output distribution of our models to be $\sigma^2_\text{min} = 0.01^2$ for the additional results on TIMIT with Gaussian outputs. This limits the maximum value attainable by the prediction/reconstruction density of a single waveform frame $x_t$.

Specifically, we can see that since
\begin{equation}
    \log p(x_t|\cdot) = \log \mathcal{N}\pa{x_t; \mu_t, \max\cbra{\sigma_\text{min}^2, \sigma_t^2}}\enspace \label{eq: gaussian ll with minimum variance} ,
\end{equation}
the best prediction/reconstruction density is achieved when $\sigma^2 \leq \sigma_\text{min}^2$ and $\mu=x_t$. 
Here $\cdot$ indicates any variables we might condition on such as the previous input frame $x_{t-1}$ or some latent variables.
We can evaluate this best case scenario for $\sigma_\text{min}^2 = 0.01^2$,
\begin{align}
    \log \mathcal{N}\pa{x_t; x_t, \sigma_\text{min}^2} &= -\frac{1}{2}\log 2\pi - \frac{1}{2}\log \sigma_\text{min}^2 - \frac{1}{2\sigma_\text{min}^2} (x_t - x_t) \nonumber \\
    &= -\frac{1}{2}\log 2\pi - \frac{1}{2}\log 0.01^2 \nonumber \\
    &= 3.686 \enspace . 
\end{align}
Hence, with perfect prediction/reconstruction and the minimal variance ($0.01^2$), a waveform frame contributes to the likelihood with $\SI{3.686}{nats}$. With an average test set example length of $\SI{49367.3}{frames}$ frames this leads to a best-case likelihood of 181967. We provide a list of maximally attainable Gaussian likelihoods on TIMIT for different minimal variances in \cref{tab:timit best gaussian ll}. One can note that the maximal likelihood at $\sigma_\text{min}^2=0.1^2$ is lower than the likelihoods achieved by some models in \cref{tab: timit likelihoods gaussian appendix}. This indicates that the models learn to use very small variances in order to increase the likelihood. Empirically, standard deviations smaller than approximately $0.001$ can result in numerical instability.

\begin{table}[t!]
    \centering
    \begin{tabular}{rrr}
    \toprule
    $\sigma_\text{min}$ & $\sigma_\text{min}^2$ & $\max\mathcal{L}$ \\
    \midrule
    $1$      & $1$           &    $-45367$ \\
    $0.5$    & $0.25$        &    $-11146$ \\
    $0.1$    & $0.01$        &     $68307$ \\
    $0.05$   & $0.0025$      &    $102525$ \\
    $0.01$   & $0.0001$      &    $181979$ \\
    $0.005$  & $0.000025$    &    $216198$ \\
    $0.001$  & $0.000001$    &    $295651$ \\
    \bottomrule
    \end{tabular}
    \vspace{2mm}
    \caption{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
    \label{tab:timit best gaussian ll}
\end{table}

\section{Additional discussion on the choice of output distribution}\label{app: additional discussion on output distribution}
The DMoL uses a discretization of the continuous logistic distribution to define a mixture model over a discrete random variable. This allows it to parameterize multimodal distributions which can express ambiguity about the value of $\xb_t$.
The model can learn to maximize likelihood by assigning a bit of probability mass to multiple potential values of $\xb_t$.

While this is well-suited for autoregressive modeling, for which the distribution was developed, the potential multimodality poses a challenge for non-autoregressive latent variable models which independently sample multiple neighboring observations at the output.
In fact, if multiple neighboring outputs defined by the subsequence $\xb_{t_1:t_2}$ have multimodal $p(\xb_t|\cdot)$, we risk sampling a subsequence where each neighboring value expresses different potential realities, independently.

Interestingly, most work on latent variable models with non-autoregressive output distributions seem to ignore this fact and simply employ the mixture distribution with 10 mixture components \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
However, given the empirically good results of latent variable models for image generation, this seems to have posed only a minor problem in practice. We speculate that this is due to the high degree of similarity between neighbouring pixels in images. I.e. if the neighboring pixels are nuances of red, then, in all likelihood, so is the central pixel.

In the audio domain, however, neighbouring waveform frames can take wildly different values, especially at low sample rates. Furthermore, waveforms exhibit a natural symmetry between positive and negative amplitudes.
Hence, it seems plausible that multimodality may pose a larger problem in non-autoregressive speech generation by causing locally incoherent samples than it seems to do in image modelling.

\section{Additional graphical models}\label{app: additional graphical models}
In \cref{fig: cwvae cell state graph} we show the graphical model of the recurrent cell of the CW-VAE for a single time step. As noted in \cite{saxena_clockwork_2021}, this cell is very similar to the one of the Recurrent State Space Model (RSSM) \cite{hafner_learning_2019}.
In \cref{fig: cwvae three-layered graphical models (inference and generative)} we show the unrolled graphical models of a three-layered CW-VAE with $k_1=1$ and $c=2$ yielding $k_2=2$ and $k_3=4$. 
We show both the generative and inference models and highlight in blue the parameter sharing between the two models due to top-down inference.
In \cref{fig: stcn graphical models (inference and generative)} we show the graphical models of the STCN \cite{aksan_stcn_2019} at a single timestep. The model has three layers and shares the parameters of the WaveNet encoder between the inference and generative models.
In \cref{fig: vrnn graphical models (inference and generative)} we illustrate the unrolled graphical models of the inference and generative models of the VRNN \cite{chung_recurrent_2015}. We include the deterministic variable $\db_t$ in order to illustrate the difference to other latent variable models.
Likewise, in \cref{fig: srnn graphical models (inference and generative)} we illustrate the unrolled graphical models the SRNN \cite{fraccaro_sequential_2016}.

\begin{figure}[t!]
    \centering
    \def\col{blue}
    \begin{tikzpicture}
        % LATENT STATE SEGMENTATION
        % internal latent state
        \node[det] (h_t_l) {$\db_t^l$};
        \node[latent, below right=1cm of h_t_l] (z_t_l) {$\zb_t^l$};
        \node[latent, right=1.5cm of h_t_l] (s_t_l) {$\vs_t^l$};

        % \plate [options] {name} {fitlist} {caption}
        \plate [inner sep=0.4cm] {cell} {(h_t_l)(z_t_l)(s_t_l)} {};

        % inputs
        \node[latent,left=0.75cm of h_t_l] (s_tm1_l) {$\vs_{t-1}^l$};
        \node[latent,above=0.75cm of h_t_l] (s_t_lp1) {$\vs_t^{l+1}$};%
        \node[latent,below=1cm of z_t_l] (e_t_l) {$\eb_t^l$};%
        % \node[left=0.5cm of e_t_l] (e_t_l_caption) {($\eb_t^l$ conditioned only in inference)};%

        % outputs
        \node[right=1.0cm of s_t_l] (s_t_l_copy1) {};%
        \node[below=2.5cm of s_t_l] (s_t_l_copy2) {};%

        % edges
        \edge[blue] {s_tm1_l} {h_t_l}
        \edge[blue] {s_t_lp1} {h_t_l}

        \edge[] {h_t_l} {z_t_l}
        \edge[dashed] {e_t_l} {z_t_l}

        \edge[blue] {h_t_l} {s_t_l}
        \edge[blue] {z_t_l} {s_t_l}

        \edge[blue] {s_t_l} {s_t_l_copy1}
        \edge[blue] {s_t_l} {s_t_l_copy2}
    \end{tikzpicture}
\caption{CW-VAE cell state $\vs_t^l$ update. The cell state is given as $\vs_t^l=(\zb_t^l, \db_t^l)$ where $\db_t^l$ is the deterministic hidden state of a Gated Recurrent Unit \cite{cho_properties_2014}. The vector $\eb_t^l$ is computed from $\xb_t$ by the encoder network which outputs $L$ encodings, one for each latent variable, similar to that of a Ladder VAE \cite{sonderby_ladder_2016}. All blue arrows are shared between generation and inference. The dashed arrow is used only during inference. The solid arrow has unique transformations during inference and generation.}
\label{fig: cwvae cell state graph}
\end{figure}

\begin{figure}[t!]
    \begin{subfigure}[b]{\textwidth}
    \flushright
    \def\col{blue}
    \resizebox{!}{5.3cm}{
    \begin{tikzpicture}
        % GENERATIVE
        % nodes
        \node[obs] (x_1) {$\xb_t$};%
        \node[obs,right=1.00cm of x_1] (x_2) {$\xb_{t+1}$};%
        \node[obs,right=1.00cm of x_2] (x_3) {$\xb_{t+2}$};%
        \node[obs,right=1.00cm of x_3] (x_4) {$\xb_{t+3}$};%
        \node[obs,right=1.00cm of x_4] (x_5) {$\xb_{t+4}$};%
        \node[obs,right=1.00cm of x_5] (x_6) {$\xb_{t+5}$};%
        \node[obs,right=1.00cm of x_6] (x_7) {$\xb_{t+6}$};%
        \node[obs,right=1.00cm of x_7] (x_8) {$\xb_{t+7}$};%

        \node[latent,above=1.00cm of x_1](s_1_1){$\zb^{(1)}_{t}$}; %
        \node[latent,above=1.00cm of x_2](s_1_2){$\zb^{(1)}_{t+1}$}; %
        \node[latent,above=1.00cm of x_3](s_1_3){$\zb^{(1)}_{t+2}$}; %
        \node[latent,above=1.00cm of x_4](s_1_4){$\zb^{(1)}_{t+3}$}; %
        \node[latent,above=1.00cm of x_5](s_1_5){$\zb^{(1)}_{t+4}$}; %
        \node[latent,above=1.00cm of x_6](s_1_6){$\zb^{(1)}_{t+5}$}; %
        \node[latent,above=1.00cm of x_7](s_1_7){$\zb^{(1)}_{t+6}$}; %
        \node[latent,above=1.00cm of x_8](s_1_8){$\zb^{(1)}_{t+7}$}; %

        \node[latent,above=1.00cm of s_1_1](s_2_1){$\zb^{(2)}_{t}$}; %
        \node[latent,above=1.00cm of s_1_3](s_2_3){$\zb^{(2)}_{t+2}$}; %
        \node[latent,above=1.00cm of s_1_5](s_2_5){$\zb^{(2)}_{t+4}$}; %
        \node[latent,above=1.00cm of s_1_7](s_2_7){$\zb^{(2)}_{t+6}$}; %

        \node[latent,above=1.00cm of s_2_1](s_3_1){$\zb^{(3)}_{t}$}; %
        \node[latent,above=1.00cm of s_2_5](s_3_5){$\zb^{(3)}_{t+4}$}; %

        % % x to s first layer
        \edge[]{s_1_1}{x_1};
        \edge[]{s_1_2}{x_2};
        \edge[]{s_1_3}{x_3};
        \edge[]{s_1_4}{x_4};
        \edge[]{s_1_5}{x_5};
        \edge[]{s_1_6}{x_6};
        \edge[]{s_1_7}{x_7};
        \edge[]{s_1_8}{x_8};

        % % s per layer forward in time
        \edge[blue]{s_1_1}{s_1_2};
        \edge[blue]{s_1_2}{s_1_3};
        \edge[blue]{s_1_3}{s_1_4};
        \edge[blue]{s_1_4}{s_1_5};
        \edge[blue]{s_1_5}{s_1_6};
        \edge[blue]{s_1_6}{s_1_7};
        \edge[blue]{s_1_7}{s_1_8};

        \edge[blue]{s_2_1}{s_2_3};
        \edge[blue]{s_2_3}{s_2_5};
        \edge[blue]{s_2_5}{s_2_7};

        \edge[blue]{s_3_1}{s_3_5};

        % % s across layers on time
        \edge[blue]{s_2_1}{s_1_1};
        \edge[blue]{s_2_3}{s_1_3};
        \edge[blue]{s_2_5}{s_1_5};
        \edge[blue]{s_2_7}{s_1_7};

        \edge[blue]{s_3_1}{s_2_1};
        \edge[blue]{s_3_5}{s_2_5};

        % % s across layers forward in time
        \edge[blue]{s_2_1}{s_1_2};
        \edge[blue]{s_2_3}{s_1_4};
        \edge[blue]{s_2_5}{s_1_6};
        \edge[blue]{s_2_7}{s_1_8};

        \edge[blue]{s_3_1}{s_2_3};
        \edge[blue]{s_3_5}{s_2_7};

        % \node[above=of s_3_5, yshift=-1.cm, xshift=-2cm] (phi) {$p_\theta(\xb,\zb)$};
    \end{tikzpicture}
    \hspace{2cm}
    }
    \caption{}
    \label{fig: cwvae three-layered graphical model generative}
    \end{subfigure}
    \\
    \vspace{0.25cm}
    \\
    \begin{subfigure}[b]{\textwidth}
    \flushright
    \def\col{blue}
    \resizebox{!}{5.3cm}{
    \begin{tikzpicture}
        % INFERENCE
        % nodes
        \node[obs] (x_1) {$\xb_t$};%
        \node[obs,right=1.00cm of x_1] (x_2) {$\xb_{t+1}$};%
        \node[obs,right=1.00cm of x_2] (x_3) {$\xb_{t+2}$};%
        \node[obs,right=1.00cm of x_3] (x_4) {$\xb_{t+3}$};%
        \node[obs,right=1.00cm of x_4] (x_5) {$\xb_{t+4}$};%
        \node[obs,right=1.00cm of x_5] (x_6) {$\xb_{t+5}$};%
        \node[obs,right=1.00cm of x_6] (x_7) {$\xb_{t+6}$};%
        \node[obs,right=1.00cm of x_7] (x_8) {$\xb_{t+7}$};%

        \node[latent,above=1.00cm of x_1](s_1_1){$\zb^{(1)}_{t}$}; %
        \node[latent,above=1.00cm of x_2](s_1_2){$\zb^{(1)}_{t+1}$}; %
        \node[latent,above=1.00cm of x_3](s_1_3){$\zb^{(1)}_{t+2}$}; %
        \node[latent,above=1.00cm of x_4](s_1_4){$\zb^{(1)}_{t+3}$}; %
        \node[latent,above=1.00cm of x_5](s_1_5){$\zb^{(1)}_{t+4}$}; %
        \node[latent,above=1.00cm of x_6](s_1_6){$\zb^{(1)}_{t+5}$}; %
        \node[latent,above=1.00cm of x_7](s_1_7){$\zb^{(1)}_{t+6}$}; %
        \node[latent,above=1.00cm of x_8](s_1_8){$\zb^{(1)}_{t+7}$}; %

        \node[latent,above=1.00cm of s_1_1](s_2_1){$\zb^{(2)}_{t}$}; %
        \node[latent,above=1.00cm of s_1_3](s_2_3){$\zb^{(2)}_{t+2}$}; %
        \node[latent,above=1.00cm of s_1_5](s_2_5){$\zb^{(2)}_{t+4}$}; %
        \node[latent,above=1.00cm of s_1_7](s_2_7){$\zb^{(2)}_{t+6}$}; %

        \node[latent,above=1.00cm of s_2_1](s_3_1){$\zb^{(3)}_{t}$}; %
        \node[latent,above=1.00cm of s_2_5](s_3_5){$\zb^{(3)}_{t+4}$}; %

        % x to s first layer
        \edge[]{x_1}{s_1_1};
        \edge[]{x_2}{s_1_2};
        \edge[]{x_3}{s_1_3};
        \edge[]{x_4}{s_1_4};
        \edge[]{x_5}{s_1_5};
        \edge[]{x_6}{s_1_6};
        \edge[]{x_7}{s_1_7};
        \edge[]{x_8}{s_1_8};

        % x to s above layer
        \edge[bend left=40]{x_1}{s_2_1};
        \edge[bend left=40]{x_1}{s_3_1};
        \edge[]{x_2}{s_2_1};
        \edge[]{x_2}{s_3_1};
        \edge[bend left=40]{x_3}{s_2_3};
        \edge[]{x_3}{s_3_1};
        \edge[]{x_4}{s_2_3};
        \edge[]{x_4}{s_3_1};
        \edge[bend left=40]{x_5}{s_2_5};
        \edge[bend left=40]{x_5}{s_3_5};
        \edge[]{x_6}{s_2_5};
        \edge[]{x_6}{s_3_5};
        \edge[bend left=40]{x_7}{s_2_7};
        \edge[]{x_7}{s_3_5};
        \edge[]{x_8}{s_2_7};
        \edge[]{x_8}{s_3_5};

        % s per layer forward in time
        \edge[\col]{s_1_1}{s_1_2};
        \edge[\col]{s_1_2}{s_1_3};
        \edge[\col]{s_1_3}{s_1_4};
        \edge[\col]{s_1_4}{s_1_5};
        \edge[\col]{s_1_5}{s_1_6};
        \edge[\col]{s_1_6}{s_1_7};
        \edge[\col]{s_1_7}{s_1_8};

        \edge[\col]{s_2_1}{s_2_3};
        \edge[\col]{s_2_3}{s_2_5};
        \edge[\col]{s_2_5}{s_2_7};

        \edge[\col]{s_3_1}{s_3_5};

        % s across layers on time
        \edge[\col]{s_2_1}{s_1_1};
        \edge[\col]{s_2_3}{s_1_3};
        \edge[\col]{s_2_5}{s_1_5};
        \edge[\col]{s_2_7}{s_1_7};

        \edge[\col]{s_3_1}{s_2_1};
        \edge[\col]{s_3_5}{s_2_5};

        % s across layers forward in time
        \edge[\col]{s_2_1}{s_1_2};
        \edge[\col]{s_2_3}{s_1_4};
        \edge[\col]{s_2_5}{s_1_6};
        \edge[\col]{s_2_7}{s_1_8};

        \edge[\col]{s_3_1}{s_2_3};
        \edge[\col]{s_3_5}{s_2_7};

        % \node[above=of s_3_5, yshift=-1.cm, xshift=-2cm] (phi) {$q_\phi(\zb|\xb)$};
    \end{tikzpicture}
    \hspace{2cm}
    }
    \caption{}
    \label{fig: cwvae three-layered graphical model inference}
    \end{subfigure}
\caption{
    CW-VAE \cite{saxena_clockwork_2021} generative model $p(\xb,\zb)$ in (\subref{fig: cwvae three-layered graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: cwvae three-layered graphical model inference}) for a three-layered model with $k_1=1$ and $c=2$ giving $k_2=2$ and $k_3=4$ unrolled over eight steps in the observed variable. Blue arrows are (mostly) shared between the inference and generative models. See \cref{fig: cwvae cell state graph} for a detailed graphical model expanding on the latent nodes $\zb^l_t$ and parameter sharing.
}
\label{fig: cwvae three-layered graphical models (inference and generative)}
\end{figure}

\begin{figure}[t!]
    \begin{subfigure}[b]{\textwidth}
    \centering
    \def\col{blue}
    \resizebox{!}{5.0cm}{
    \begin{tikzpicture}
        % GENERATIVE
        % nodes
        \node[obs] (x_1) {$\xb_t$};%
        \node[obs,left=.75cm of x_1] (x_2) {$\xb_{t-1}$};
        \node[obs,left=.75cm of x_2] (x_3) {$\xb_{t-2}$};
        \node[obs,left=.75cm of x_3] (x_4) {$\xb_{t-3}$};
        \node[obs,left=.75cm of x_4] (x_5) {$\xb_{t-4}$};
        \node[obs,left=.75cm of x_5] (x_6) {$\xb_{t-5}$};
        \node[obs,left=.75cm of x_6] (x_7) {$\xb_{t-6}$};
        \node[obs,left=.75cm of x_7] (x_8) {$\xb_{t-7}$};

        \node[det,above=.75cm of x_1](d_1_1){$\db_{t}^{(1)}$};
        \node[det,above=.75cm of x_3](d_1_3){$\db_{t-2}^{(1)}$};
        \node[det,above=.75cm of x_5](d_1_5){$\db_{t-4}^{(1)}$};
        \node[det,above=.75cm of x_7](d_1_7){$\db_{t-6}^{(1)}$};

        \node[det,above=.75cm of d_1_1](d_2_1){$\db_{t}^{(2)}$};
        \node[det,above=.75cm of d_1_5](d_2_5){$\db_{t-4}^{(2)}$};

        \node[det,above=.75cm of d_2_1](d_3_1){$\db_{t}^{(3)}$};

        \node[latent,right=.75cm of d_1_1](z_1){$\zb_{t+1}^{(1)}$};
        \node[latent,right=.75cm of d_2_1](z_2){$\zb_{t+1}^{(2)}$};
        \node[latent,right=.75cm of d_3_1](z_3){$\zb_{t+1}^{(3)}$};

        \node[latent,right=.75cm of z_2](z){$\zb_{t+1}$};

        \node[obs,below=2.8cm of z] (x_0) {$\xb_{t+1}$};

        \edge[\col]{x_1}{d_1_1};
        \edge[\col]{x_2}{d_1_1};
        \edge[\col]{x_3}{d_1_3};
        \edge[\col]{x_4}{d_1_3};
        \edge[\col]{x_5}{d_1_5};
        \edge[\col]{x_6}{d_1_5};
        \edge[\col]{x_7}{d_1_7};
        \edge[\col]{x_8}{d_1_7};

        \edge[\col]{d_1_1}{d_2_1};
        \edge[\col]{d_1_3}{d_2_1};
        \edge[\col]{d_1_5}{d_2_5};
        \edge[\col]{d_1_7}{d_2_5};

        \edge[\col]{d_2_1}{d_3_1};
        \edge[\col]{d_2_5}{d_3_1};

        \edge[]{d_1_1}{z_1};
        \edge[]{d_2_1}{z_2};
        \edge[]{d_3_1}{z_3};

        \edge[]{z_3}{z_2}
        \edge[]{z_2}{z_1}

        \edge[]{z_1}{z}
        \edge[]{z_2}{z}
        \edge[]{z_3}{z}

        \edge[]{z}{x_0}

        % \node[above=of d_1_3, yshift=3.cm] (theta) {$p_\theta(\zb,\xb)$};
    \end{tikzpicture}
    }
    \caption{}
    \label{fig: stcn graphical model generative}
    \end{subfigure}
    \\
    \vspace{0.25cm}
    \\
    \begin{subfigure}[b]{\textwidth}
    \centering
    \def\col{blue}
    \resizebox{!}{5.0cm}{
    \begin{tikzpicture}
        % INFERENCE
        % nodes
        \node[obs] (x_1) {$\xb_t$};%
        \node[obs,left=.75cm of x_1] (x_2) {$\xb_{t-1}$};
        \node[obs,left=.75cm of x_2] (x_3) {$\xb_{t-2}$};
        \node[obs,left=.75cm of x_3] (x_4) {$\xb_{t-3}$};
        \node[obs,left=.75cm of x_4] (x_5) {$\xb_{t-4}$};
        \node[obs,left=.75cm of x_5] (x_6) {$\xb_{t-5}$};
        \node[obs,left=.75cm of x_6] (x_7) {$\xb_{t-6}$};
        \node[obs,left=.75cm of x_7] (x_8) {$\xb_{t-7}$};

        \node[det,above=.75cm of x_1](d_1_1){$\db_{t}^{(1)}$};
        \node[det,above=.75cm of x_3](d_1_3){$\db_{t-2}^{(1)}$};
        \node[det,above=.75cm of x_5](d_1_5){$\db_{t-4}^{(1)}$};
        \node[det,above=.75cm of x_7](d_1_7){$\db_{t-6}^{(1)}$};

        \node[det,above=.75cm of d_1_1](d_2_1){$\db_{t}^{(2)}$};
        \node[det,above=.75cm of d_1_5](d_2_5){$\db_{t-4}^{(2)}$};

        \node[det,above=.75cm of d_2_1](d_3_1){$\db_{t}^{(3)}$};

        \node[latent,right=.75cm of d_1_1](z_1){$\zb_{t}^{(1)}$};
        \node[latent,right=.75cm of d_2_1](z_2){$\zb_{t}^{(2)}$};
        \node[latent,right=.75cm of d_3_1](z_3){$\zb_{t}^{(3)}$};

        \node[latent,right=.75cm of z_2](z){$\zb_{t}$};

        \edge[\col]{x_1}{d_1_1};
        \edge[\col]{x_2}{d_1_1};
        \edge[\col]{x_3}{d_1_3};
        \edge[\col]{x_4}{d_1_3};
        \edge[\col]{x_5}{d_1_5};
        \edge[\col]{x_6}{d_1_5};
        \edge[\col]{x_7}{d_1_7};
        \edge[\col]{x_8}{d_1_7};

        \edge[\col]{d_1_1}{d_2_1};
        \edge[\col]{d_1_3}{d_2_1};
        \edge[\col]{d_1_5}{d_2_5};
        \edge[\col]{d_1_7}{d_2_5};

        \edge[\col]{d_2_1}{d_3_1};
        \edge[\col]{d_2_5}{d_3_1};

        \edge[]{d_1_1}{z_1};
        \edge[]{d_2_1}{z_2};
        \edge[]{d_3_1}{z_3};

        \edge[]{z_3}{z_2}
        \edge[]{z_2}{z_1}

        \edge[]{z_1}{z}
        \edge[]{z_2}{z}
        \edge[]{z_3}{z}

        % \node[above=of d_1_3, yshift=3.cm] (phi){$q_\phi(\zb|\xb)$};
    \end{tikzpicture}
    }
    \caption{}
    \label{fig: stcn graphical model inference}
    \end{subfigure}
\caption{
    STCN \cite{aksan_stcn_2019} generative model $p(\xb,\zb)$ in (\subref{fig: stcn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: stcn graphical model inference}) for a single time-step.
    The WaveNet autoregressive encoder is shared between generative and inference models. It is depicted here with only one stack of three layers in order to illustrate the dilated convolution with limited space. In practice, the model uses ten layers in each of five stacks/cycles resulting in a much larger receptive field. Importantly, the model parameterizes the five latent variables using the last deterministic representation $\db^{(l)}$ from each stack, i.e. only every fifth $l$ starting from $l=5$ and ending at $l=25$.
    Note that the generative model uses the prior to transform the WaveNet hidden states $\db_{t}^{(l)}$ into the latent variable $\zb_{t+1}^{(l)}$ one step ahead in time compared to the approximate posterior which infers $\zb_{t}^{(l)}$.
    Also note that $\zb_t$ is constructed by concatenating all $\zb_t^{(l)}$. The original paper explores setting $\zb_t$ equal to $\zb_t^{(1)}$. The best-performing STCN for speech, which also the one we implement, uses a WaveNet decoder to predict $\xb_{t+1}$ from a sequence of $\zb_t$ rather than a per-timestep transform.
    Blue arrows are shared between the inference and generative models.
}
\label{fig: stcn graphical models (inference and generative)}
\end{figure}

\begin{figure}[t!]
    \centering
    \hfill
    \def\col{blue}
    \begin{subfigure}[b]{0.48\textwidth}
    \centering
    \begin{tikzpicture}
        % GENERATIVE
        % nodes
        \node[obs] (x_1) {$\xb_t$};%
        \node[obs,right=1.00cm of x_1] (x_2) {$\xb_{t+1}$};
        \node[obs,right=1.00cm of x_2] (x_3) {$\xb_{t+2}$};

        \node[latent,above=1.00cm of x_1](z_1_1){$\zb_{t}$};
        \node[latent,above=1.00cm of x_2](z_1_2){$\zb_{t+1}$};
        \node[latent,above=1.00cm of x_3](z_1_3){$\zb_{t+2}$};

        \node[det,above=1.00cm of z_1_1](d_1_1){$\db_{t}$};
        \node[det,above=1.00cm of z_1_2](d_1_2){$\db_{t+1}$};
        \node[det,above=1.00cm of z_1_3](d_1_3){$\db_{t+2}$};

        % x to latent
        \edge[]{z_1_1}{x_1};
        \edge[]{z_1_2}{x_2};
        \edge[]{z_1_3}{x_3};

        % deterministic to latent
        \edge[\col]{d_1_1}{z_1_1};
        \edge[\col]{d_1_2}{z_1_2};
        \edge[\col]{d_1_3}{z_1_3};

        % deterministic to x (skip)
        \edge[bend left]{d_1_1}{x_1};
        \edge[bend left]{d_1_2}{x_2};
        \edge[bend left]{d_1_3}{x_3};

        % deterministic forward in time
        \edge[\col]{d_1_1}{d_1_2};
        \edge[\col]{d_1_2}{d_1_3};

        % x to deterministic forward in time
        \edge[\col]{x_1.35}{d_1_2};
        \edge[\col]{x_2.35}{d_1_3};

        % latent to deterministic forward in time
        \edge[\col]{z_1_1}{d_1_2};
        \edge[\col]{z_1_2}{d_1_3};

        \node[above=of d_1_2, yshift=-1.cm] (phi) {$p_\theta(\zb,\xb)$};
    \end{tikzpicture}
    \caption{}
    \label{fig: vrnn graphical model generative}
    \end{subfigure}
    \begin{subfigure}[b]{0.48\textwidth}
    \centering
    \begin{tikzpicture}
        % INFERENCE
        % nodes
        \node[obs] (x_1) {$\xb_t$};%
        \node[obs,right=1.00cm of x_1] (x_2) {$\xb_{t+1}$};
        \node[obs,right=1.00cm of x_2] (x_3) {$\xb_{t+2}$};

        \node[latent,above=1.00cm of x_1](z_1_1){$\zb_{t}$};
        \node[latent,above=1.00cm of x_2](z_1_2){$\zb_{t+1}$};
        \node[latent,above=1.00cm of x_3](z_1_3){$\zb_{t+2}$};

        \node[det,above=1.00cm of z_1_1](d_1_1){$\db_{t}$};
        \node[det,above=1.00cm of z_1_2](d_1_2){$\db_{t+1}$};
        \node[det,above=1.00cm of z_1_3](d_1_3){$\db_{t+2}$};

        % x to latent
        \edge[]{x_1}{z_1_1};
        \edge[]{x_2}{z_1_2};
        \edge[]{x_3}{z_1_3};

        % deterministic to latent
        \edge[\col]{d_1_1}{z_1_1};
        \edge[\col]{d_1_2}{z_1_2};
        \edge[\col]{d_1_3}{z_1_3};

        % deterministic forward in time
        \edge[\col]{d_1_1}{d_1_2};
        \edge[\col]{d_1_2}{d_1_3};

        % x to deterministic forward in time
        \edge[\col]{x_1}{d_1_2};
        \edge[\col]{x_2}{d_1_3};

        % latent to deterministic forward in time
        \edge[\col]{z_1_1}{d_1_2};
        \edge[\col]{z_1_2}{d_1_3};

        \node[above=of d_1_2, yshift=-1.cm] (phi) {$q_\phi(\zb|\xb)$};
    \end{tikzpicture}
    \caption{}
    \label{fig: vrnn graphical model inference}
    \end{subfigure}
\caption{
    VRNN \cite{chung_recurrent_2015} generative model $p(\xb,\zb)$ in (\subref{fig: vrnn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: vrnn graphical model inference}) unrolled over three steps in the observed variable. 
    Blue arrows are shared between the inference and generative models.
}
\label{fig: vrnn graphical models (inference and generative)}
\end{figure}

\begin{figure}[t!]
    \centering
    \hfill
    \def\col{blue}
    \begin{subfigure}[b]{0.48\textwidth}
    \centering
    \begin{tikzpicture}
        % GENERATIVE
        % nodes
        \node[obs] (x_1) {$\xb_t$};%
        \node[obs,right=1.00cm of x_1] (x_2) {$\xb_{t+1}$};
        \node[obs,right=1.00cm of x_2] (x_3) {$\xb_{t+2}$};

        \node[latent,above=1.00cm of x_1](z_1_1){$\zb_{t}$};
        \node[latent,above=1.00cm of x_2](z_1_2){$\zb_{t+1}$};
        \node[latent,above=1.00cm of x_3](z_1_3){$\zb_{t+2}$};

        \node[det,above=1.00cm of z_1_1](d_1_1){$\db_{t}$};
        \node[det,above=1.00cm of z_1_2](d_1_2){$\db_{t+1}$};
        \node[det,above=1.00cm of z_1_3](d_1_3){$\db_{t+2}$};

        % x to latent
        \edge[]{z_1_1}{x_1};
        \edge[]{z_1_2}{x_2};
        \edge[]{z_1_3}{x_3};

        % deterministic to latent
        \edge[]{d_1_1}{z_1_1};
        \edge[]{d_1_2}{z_1_2};
        \edge[]{d_1_3}{z_1_3};

        % deterministic to x (skip)
        \edge[bend left=35]{d_1_1}{x_1};
        \edge[bend left=35]{d_1_2}{x_2};
        \edge[bend left=35]{d_1_3}{x_3};

        % deterministic forward in time
        \edge[\col]{d_1_1}{d_1_2};
        \edge[\col]{d_1_2}{d_1_3};

        % x to deterministic forward in time
        \edge[\col]{x_1.35}{d_1_2};
        \edge[\col]{x_2.35}{d_1_3};

        % latent forward in time
        \edge[\col]{z_1_1}{z_1_2};
        \edge[\col]{z_1_2}{z_1_3};

        \node[above=of d_1_2, yshift=-1.cm] (phi) {$p_\theta(\zb,\xb)$};
    \end{tikzpicture}
    \caption{}
    \label{fig: srnn graphical model generative}
    \end{subfigure}
    \begin{subfigure}[b]{0.48\textwidth}
    \centering
    \begin{tikzpicture}
        % INFERENCE
        % nodes
        \node[obs] (x_1) {$\xb_t$};%
        \node[obs,right=1.00cm of x_1] (x_2) {$\xb_{t+1}$};
        \node[obs,right=1.00cm of x_2] (x_3) {$\xb_{t+2}$};

        \node[latent,above=1.00cm of x_1](z_1_1){$\zb_{t}$};
        \node[latent,above=1.00cm of x_2](z_1_2){$\zb_{t+1}$};
        \node[latent,above=1.00cm of x_3](z_1_3){$\zb_{t+2}$};

        \node[det,above=1.00cm of z_1_1](a_1_1){$\ab_{t}$};
        \node[det,above=1.00cm of z_1_2](a_1_2){$\ab_{t+1}$};
        \node[det,above=1.00cm of z_1_3](a_1_3){$\ab_{t+2}$};

        \node[det,above=1.00cm of a_1_1](d_1_1){$\db_{t}$};
        \node[det,above=1.00cm of a_1_2](d_1_2){$\db_{t+1}$};
        \node[det,above=1.00cm of a_1_3](d_1_3){$\db_{t+2}$};

        % x to deterministic a
        \edge[bend left=40]{x_1}{a_1_1};
        \edge[bend left=40]{x_2}{a_1_2};
        \edge[bend left=40]{x_3}{a_1_3};

        % x to deterministic d
        \edge[\col]{x_1}{d_1_2};
        \edge[\col]{x_2}{d_1_3};

        % deterministic d to deterministic a
        \edge[]{d_1_1}{a_1_1};
        \edge[]{d_1_2}{a_1_2};
        \edge[]{d_1_3}{a_1_3};

        % deterministic to latent
        \edge[]{a_1_1}{z_1_1};
        \edge[]{a_1_2}{z_1_2};
        \edge[]{a_1_3}{z_1_3};

        % deterministic d forward in time
        \edge[\col]{d_1_1}{d_1_2};
        \edge[\col]{d_1_2}{d_1_3};

        % deterministic a backward in time
        \edge[]{a_1_2}{a_1_1};
        \edge[]{a_1_3}{a_1_2};

        % latent forward in time
        \edge[\col]{z_1_1}{z_1_2};
        \edge[\col]{z_1_2}{z_1_3};

        \node[above=of d_1_2, yshift=-1.cm] (phi) {$q_\phi(\zb|\xb)$};
    \end{tikzpicture}
    \caption{}
    \label{fig: srnn graphical model inference}
    \end{subfigure}
\caption{
    SRNN \cite{fraccaro_sequential_2016} generative model $p(\xb,\zb)$ in (\subref{fig: srnn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: srnn graphical model inference}) unrolled over three steps in the observed variable. 
    Blue arrows are shared between the inference and generative models.
}
\label{fig: srnn graphical models (inference and generative)}
\end{figure}

\section{Additional latent evaluation}\label{app: additional latent space clustering}
We visualize the performance of a $k$-nearest-neighbour classifier for classification of speaker gender and height in \cref{fig: knn fraction latent space height and gender}.
The classifier is fitted to time-averaged latent representations and Mel-features. 
We divide the height into three classes: below $\SI{175}{cm}$, above $\SI{185}{cm}$ and in-between. Compared to phonemes, the gender and height of a speaker are global attributes that affect the entire signal. 
In both cases, we see improved performance from using the learned latent space over Mel-features. Notably, $\zb^2$ is outperformed by the Mel-features for gender identification which may indicate that $\zb^2$ learns to ignore this attribute compared to $\zb^1$.

We provide some additional latent space clustering of speaker gender in \cref{fig: latent space visualization gender} and of speaker height in \cref{fig: cwvae latent height}.

All results presented here are obtained with a 2-layered CW-VAE trained on $\mu$-law encoded PCM similar to the one in \cref{tab: likelihoods timit}.

\begin{figure*}[t!]
    \centering
    \includegraphics[width=0.477\textwidth]{paper_benchmarking/1xrnjn5y_speaker_gender_10_samples_knn_correct_fraction_model_vs_mel_100_neighbors_lda_subspace_1d.pdf}
    \hfill
    \includegraphics[width=0.483\textwidth]{paper_benchmarking/1xrnjn5y_speaker_height_m_10_samples_knn_correct_fraction_model_vs_mel_50_neighbors_lda_subspace_2d.pdf}
    \caption{
    Leave-one-out $k$-nearest-neighbor accuracy with different $k$ for
    (a) the speaker's gender and
    (b) the height of male speakers (female speakers yield a similar result).
    }
    \label{fig: knn fraction latent space height and gender}
\end{figure*}
\begin{figure}[t!]
     \centering
     \hfill
     \begin{subfigure}[b]{0.48\textwidth}
         \centering
         \includegraphics[width=\textwidth]{paper_benchmarking/1xrnjn5y_speaker_gender_latent_0_1_samples_lda_linear_subspace.pdf}
         \caption{}
         \label{fig: latent space gender}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.48\textwidth}
         \centering
         \includegraphics[width=\textwidth]{paper_benchmarking/1xrnjn5y_speaker_gender_mel_features_lda_linear_subspace.pdf}
         \caption{}
         \label{fig: mel features gender}
     \end{subfigure}
    \caption{Clustering of speaker gender in an one-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space and of a time-averaged mel spectrogram. The total overlap is slightly smaller in the subspace of the CW-VAE latent space and the separation between the distribution peaks is larger.}
    \label{fig: latent space visualization gender}
\end{figure}

\begin{figure}[t!]
     \centering
     \hfill
     \begin{subfigure}[b]{0.48\textwidth}
         \centering
         \includegraphics[width=\textwidth]{paper_benchmarking/1xrnjn5y_speaker_height_m_1_samples_latent_0_lda_linear_subspace.pdf}
         \caption{}
         \label{fig: cwvae latent z1 height}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.48\textwidth}
         \centering
         \includegraphics[width=\textwidth]{paper_benchmarking/1xrnjn5y_speaker_age_f_1_samples_latent_1_lda_linear_subspace.pdf}
         \caption{}
         \label{fig: cwvae latent z0 age}
     \end{subfigure}
    \caption{(a) Clustering of speaker height for male speakers and (b) speaker age for female speakers in an two-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space.}
    \label{fig: cwvae latent height}
\end{figure}

\section{Distribution of phoneme duration in TIMIT}\label{app: timit phoneme distributions}
In \cref{fig: timit validation set phoneme duration} we plot a boxplots of the duration of each phoneme in the TIMIT dataset. We do this globally as well as for a single speaker to show that phoneme duration can vary between individual speakers.

A description of the phonemes used for the TIMIT dataset can be found at \url{https://catalog.ldc.upenn.edu/docs/LDC93S1/PHONCODE.TXT}.

\begin{figure}[t!]
    \centering
    \hfill
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{paper_benchmarking/timit_validation_set_phoneme_duration_boxplot_speaker_id_drw0.pdf}
        \caption{}
        \label{fig: timit validation set phoneme duration DRW0}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{paper_benchmarking/timit_validation_set_phoneme_duration_boxplot_speaker_id_none.pdf}
        \caption{}
        \label{fig: timit validation set phoneme duration global}
    \end{subfigure}
    \caption{Boxplots of the duration of the pronunciation of phonemes in TIMIT for a specific speaker DRW0 in (\subref{fig: timit validation set phoneme duration DRW0}) and globally in (\subref{fig: timit validation set phoneme duration global}). Not all phonemes are pronounced by speaker DRW0 over the course of their 10 test set sentences and hence they are missing from the x-axis compared to the global durations.}
    \label{fig: timit validation set phoneme duration}
\end{figure}

\section{Model samples and reconstructions}\label{app: model samples and reconstructions}
We provide samples and reconstructions for some of the models considered here at the following URL: \url{https://doi.org/10.5281/zenodo.5911899}.
The samples are generated from the prior of Clockwork VAE, SRNN and VRNN and from a WaveNet by conditioning on pure zeros. All models are configured as those reported in \cref{tab: likelihoods timit}.
Importantly, the samples are unconditional. Hence they are \emph{not} reconstructions inferred from a given input nor are they conditioned on any auxiliary data like text.

Although sample quality is a somewhat subjective matter, we find the quality of the unconditional Clockwork VAE to be better than those of our VRNN and SRNN. WaveNet is known to produce samples with intelligible speech when conditioned on e.g. text, but unconditional samples from WaveNet lack semantic content such as words as do VRNN, SRNN and Clockwork VAE.

\iffalse

\fi

### 2023-02-25 appendices/appendix_benchmarking_workshop.tex
Add appendices for benchmarking and hierarchical

--- 2023-02-25 appendices/appendix_benchmarking.tex
@@
 
-\chapter[benchmarking generative latent variable models for speech]{Benchmarking Generative Latent Variable Models for Speech}
+\chapter[appendix for benchmarking generative latent variable models for speech]{Benchmarking Generative Latent Variable Models for Speech}
+
+\iffalse
 
@@
 \paragraph{LSTM}
-The LSTM baseline uses an MLP encoder to embed the waveform subsegment $\xb_{t:t+s-1}$ to a feature vector before feeding it to the LSTM cell.
+The LSTM baseline uses an MLP encoder to embed the waveform subsegment $\vx_{t:t+s-1}$ to a feature vector before feeding it to the LSTM cell.
 The encoder is similar to the parameterization of $\phi_\text{vrnn}^\text{enc}$ for the VRNN described above.
-The LSTM cell produces the hidden state $\db_t$ from $\xb_{t:t+s-1}$ and passes it to a decoder.
+The LSTM cell produces the hidden state $\vd_t$ from $\vx_{t:t+s-1}$ and passes it to a decoder.
 Like the encoder, the decoder is parameterized like $\phi_\text{vrnn}^\text{dec}$ of the VRNN.
-It outputs the waveform predictions $\xb_{t+s:t+2s-1}$ from the hidden state $\db_t$.
+It outputs the waveform predictions $\vx_{t+s:t+2s-1}$ from the hidden state $\vd_t$.
 The LSTM model uses a single vanilla unidirectional LSTM cell.
@@
 
-\paragraph{CW-VAE} We implement the CW-VAE based on the original work \cite{saxena_clockwork_2021} but with some modifications also briefly described in section~\cref{sec: cwvae on speech}.
+\paragraph{CW-VAE} We implement the CW-VAE based on the original work \cite{saxena_clockwork_2021} but with some modifications also briefly described in the paper.
 We replace the encoder/decoder model architectures of the original work with architectures designed for waveform modeling.
@@
 Temporal dropout works similar to regular dropout but samples the entries of the hidden state to mask only once and apply it to all timesteps, i.e.
-masking $\hb_{t}$ at vector index $i$ for all $t$ (and $i$).
+masking $\vh_{t}$ at vector index $i$ for all $t$ (and $i$).
 We mask by zeroing vector elements. We never mask the first timestep. 
@@
 In general, conversion from nats to bits (i.e., from $\log_e$ to $\log_2$) is achieved by $\log_2(x) = \log_e(x)  / \log_2(e) $.
-Remember that $\log_2 p(\xb_{1:T})$ generally factorizes as $\sum_t \log_2 p(x_t|\cdot)$.
-In sequence modeling, it is important to remember that each example $\xb^{i}$ must be weighted differently according the sequence length of that specific example.
+Remember that $\log_2 p(\vx_{1:T})$ generally factorizes as $\sum_t \log_2 p(x_t|\cdot)$.
+In sequence modeling, it is important to remember that each example $\vx^{i}$ must be weighted differently according the sequence length of that specific example.
 This is in contrast to computing bits per dimension in the image domain where images in a dataset are usually of the same dimensions.
@@
 \begin{equation}
-    \mathcal{L}(\xb^{i}) = \frac{1}{\sum_i T_i} \sum_i \sum_t \log_2 p(\xb_t^{i}) \enspace ,
+    \mathcal{L}(\vx^{i}) = \frac{1}{\sum_i T_i} \sum_i \sum_t \log_2 p(\vx_t^{i}) \enspace ,
 \end{equation}
-where $i$ denotes the example index, $T_i$ is the length of example $\xb^i$ in waveform frames and $t$ is the time index.
-If a single timestep $\xb_t^i$ represents multiple waveform frames stacked with some stack size $s$, it is important to note that the sum over $t$ only has $T_i/s$ elements.
-For the LVMs, the term $\log_2 p(\xb_t^{i})$ is lower bounded by the ELBO in \eqref{eq: elbo}.
+where $i$ denotes the example index, $T_i$ is the length of example $\vx^i$ in waveform frames and $t$ is the time index.
+If a single timestep $\vx_t^i$ represents multiple waveform frames stacked with some stack size $s$, it is important to note that the sum over $t$ only has $T_i/s$ elements.
+For the LVMs, the term $\log_2 p(\vx_t^{i})$ is lower bounded by the ELBO in \eqref{eq: elbo}.
 
 \section{Additional likelihood results} \label{app: additional likelihood results appendix}
+
+\paragraph{LibriSpeech, $\mu$-law, DMoL} We provide additional results on LibriSpeech with audio represented as $\mu$-law encoded PCM in \cref{tab: likelihoods librispeech}.
+See appendix~\cref{app: dataset details}, \cref{app: model architectures} and \cref{app: training details} for additional details.
 
@@
 
-We constrain the variance of the Gaussians used with our models to be at least $\sigma^2_\text{min} = 0.01^2$ in order to avoid the variance going to zero, the likelihood going to infinity and optimization becoming unstable.
-The Gaussian standard deviation is clamped at minimum $0.001$ by \cite{aksan_stcn_2019}.
+We constrain the standard deviation of the Gaussians used with our models to be at least $\sigma_\text{min} = 0.01$ in order to avoid it going to zero, the likelihood going to infinity and optimization becoming unstable.
+The minimum Gaussian standard deviation of \citet{aksan_stcn_2019} is $\sigma_\text{min} = 0.001$.
 
@@
 Not being autoregressive in the observed variable, the CW-VAE cannot exploit this local smoothness in the same way.
-Instead, the reconstruction is conditioned on a stochastic latent variable, $p(\xb_t|\zb^1_t)$, which introduces uncertainty and likely larger reconstruction variances.
+Instead, the reconstruction is conditioned on a stochastic latent variable, $p(\vx_t|\vz^1_t)$, which introduces uncertainty and likely larger reconstruction variances.
+
+\begin{table*}[t]
+    \centering
+    \resizebox{0.98\textwidth}{!}{
+    \begin{tabular}{lll|rrrr}  % https://wandb.ai/vseq/icml_asr
+        \toprule
+        $s$ & \bf Model                   & \bf Configuration       & \multicolumn{4}{c}{\bf Likelihood $\mathcal{L}$ [bpf]} \\
+         & & & dev-clean & dev-other & test-clean & test-other\\
+         & & & 10h/100h & 10h/100h & 10h/100h & 10h/100h\\
+        \midrule
+        1 & Uniform             & Uninformed     & \multicolumn{1}{c}{16.00} & \multicolumn{1}{c}{16.00} & \multicolumn{1}{c}{16.00} & \multicolumn{1}{c}{16.00} \\
+        1 & DMoL                & Optimal        & \multicolumn{1}{c}{15.66} & \multicolumn{1}{c}{15.70} & \multicolumn{1}{c}{15.62} & \multicolumn{1}{c}{15.71}\\   % $\vmu=[-0.551, 0.551], \vz=[0.11, 0.11]$
+        - & FLAC                & Linear PCM     & \multicolumn{1}{c}{\bf 9.390} & \multicolumn{1}{c}{\bf 9.292} & \multicolumn{1}{c}{\bf 9.700} & \multicolumn{1}{c}{\bf 9.272} \\
+        \midrule
+        1 & Wavenet             & $D_c=96$       & \bf 10.96/10.89 & \bf 10.85/10.76 & \bf 11.12/11.01 & \bf 11.05/10.85 \\
+        1 & LSTM                & $D_d=256$      & 11.21/11.17 & 11.10/11.06 & 11.35/11.29 & 11.28/11.23 \\  % 15ua3hau/(lmssrk5k,248n2ylu)
+        \midrule
+        64 & Wavenet            & $D_c=96$       & 13.61/13.24 & 13.58/13.21 & 13.61/13.22 & 13.60/13.21 \\  % 2ghkybtd/3iieasdt
+        64 & LSTM               & $D_d=256$      & 13.56/13.25 & 13.52/13.24 & 13.55/13.23 & 13.56/13.25 \\  % 10fpovu3/2bu1izin
+        64 & CW-VAE             & $D_z=96,L=1$   & $\leq$12.32/12.24 & 12.32/12.23 & 12.43/12.33 & 12.43/12.33 \\
+        64 & CW-VAE             & $D_z=96,L=2$   & $\leq$12.30/12.22 & 12.30/12.21 & 12.40/12.31 & 12.39/12.32 \\
+        64 & STCN               & $D_z=256,L=5$  & $\leq$\bf 11.83/11.47 & \bf 11.82/11.46 & \bf 11.94/11.58 & \bf 11.94/11.60 \\  % kutxm2hx/3t0xg7pb
+        \bottomrule
+    \end{tabular}
+    }
+    \caption{
+    Model likelihoods $\mathcal{L}$ on \textbf{LibriSpeech} test sets represented as \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}.
+    For the CW-VAE, $s$ refers to $s_1$ and the two-layered models have $s_2=8s_1$.
+    The models are trained on either the $\SI{10}{h}$ LibriLight subset or the $\SI{100}{h}$ LibriSpeech \textrm{train-clean-100} subset as indicated.
+    Likelihoods are given in units of bits per frame (bpf).
+    }
+    \label{tab: likelihoods librispeech}
+\end{table*}
 
@@
     \centering
-    \begin{tabular}{ll|lr}
+    \begin{tabular}{lll|r}
         \toprule
-        $s$    & \bfseries Model           & \bfseries Configuration           & \bfseries $\mathcal{L}$ [bpf] \\
+        $s$    & \bf Model              & \bf Configuration     & \bf $\mathcal{L}$ [bpf] \\
         \midrule
@@
         1         & DMoL                & Optimal               & 10.70 \\   % $mean[0.00], scale=[0.005]$
-        -         & FLAC                & Linear PCM            & 8.582 \\
+        %-         & FLAC                & Linear PCM            & 8.582 \\
         \midrule
@@
     \end{tabular}
-    \vspace{2mm}
     \caption{
-    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit linear PCM}. The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
+    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit linear PCM}, obtained by different latent variable models and compared to autoregressive baselines all using a discretized mixture of logistics with 10 components as output distribution.
+Likelihoods are given in units of bits per frame (bpf) and obtained by normalizing the total likelihood of each sequence with the individual sequence length and then averaging over the dataset.
+The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
     }
@@
     \centering
-    \begin{tabular}{ll|lr}
+    \begin{tabular}{lll|r}
         \toprule
-        $s$    & \bfseries Model           & \bfseries Configuration           & \bfseries $\mathcal{L}$ [bpf] \\
+        $s$    & \bf Model           & \bf Configuration           & \bf $\mathcal{L}$ [bpf] \\
         \midrule
@@
     \end{tabular}
-    \vspace{2mm}
     \caption{
-    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}.
+    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}, obtained by different latent variable models and compared to autoregressive baselines all using a discretized mixture of logistics with 10 components as output distribution.
+Likelihoods are given in units of bits per frame (bpf) and obtained by normalizing the total likelihood of each sequence with the individual sequence length and then averaging over the dataset.
     }
@@
     \centering
-    \begin{tabular}{ll|lr}
+    \begin{tabular}{lll|r}
         \toprule
-        $s$ & \bfseries Model           & \bfseries Configuration           & \bfseries $\mathcal{L}$ [nats] \\
+        $s$ & \bf Model           & \bf Configuration           & \bf $\mathcal{L}$ [nats] \\
         \midrule
@@
     \end{tabular}
-    \vspace{2mm}
     \caption{
     Model likelihoods on \textbf{TIMIT} represented as \textbf{globally normalized 16 bit linear PCM}.
-Contrary to the other likelihoods reported in this paper, here they are given in units of nats and obtained by summing the likelihood over time and over all examples in the dataset and dividing by the total number of examples.
+Likelihoods are given in units of nats and obtained by summing the likelihood over time and over all examples in the dataset and dividing by the total number of examples.
 In the table, Normal refers to using a Gaussian likelihood and GMM refers to using a  Gaussian Mixture Model likelihood with 20 components.
@@
     \end{tabular}
-    \vspace{2mm}
     \caption{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
@@
 The DMoL uses a discretization of the continuous logistic distribution to define a mixture model over a discrete random variable.
-This allows it to parameterize multimodal distributions which can express ambiguity about the value of $\xb_t$.
-The model can learn to maximize likelihood by assigning a bit of probability mass to multiple potential values of $\xb_t$.
+This allows it to parameterize multimodal distributions which can express ambiguity about the value of $\vx_t$.
+The model can learn to maximize likelihood by assigning a bit of probability mass to multiple potential values of $\vx_t$.
 
 While this is well-suited for autoregressive modeling, for which the distribution was developed, the potential multimodality poses a challenge for non-autoregressive latent variable models which independently sample multiple neighboring observations at the output.
-In fact, if multiple neighboring outputs defined by the subsequence $\xb_{t_1:t_2}$ have multimodal $p(\xb_t|\cdot)$, we risk sampling a subsequence where each neighboring value expresses different potential realities, independently.
+In fact, if multiple neighboring outputs defined by the subsequence $\vx_{t_1:t_2}$ have multimodal $p(\vx_t|\cdot)$, we risk sampling a subsequence where each neighboring value expresses different potential realities, independently.
 
@@
 As noted in \cite{saxena_clockwork_2021}, this cell is very similar to the one of the Recurrent State Space Model (RSSM) \cite{hafner_learning_2019}.
+
 In \cref{fig: cwvae three-layered graphical models (inference and generative)} we show the unrolled graphical models of a three-layered CW-VAE with $k_1=1$ and $c=2$ yielding $k_2=2$ and $k_3=4$. 
 We show both the generative and inference models and highlight in blue the parameter sharing between the two models due to top-down inference.
+
 In \cref{fig: stcn graphical models (inference and generative)} we show the graphical models of the STCN \cite{aksan_stcn_2019} at a single timestep.
 The model has three layers and shares the parameters of the WaveNet encoder between the inference and generative models.
+
 In \cref{fig: vrnn graphical models (inference and generative)} we illustrate the unrolled graphical models of the inference and generative models of the VRNN \cite{chung_recurrent_2015}.
-We include the deterministic variable $\db_t$ in order to illustrate the difference to other latent variable models.
+We include the deterministic variable $\vd_t$ in order to illustrate the difference to other latent variable models.
+
 Likewise, in \cref{fig: srnn graphical models (inference and generative)} we illustrate the unrolled graphical models the SRNN \cite{fraccaro_sequential_2016}.
@@
         % internal latent state
-        \node[det] (h_t_l) {$\db_t^l$};
-        \node[latent, below right=1cm of h_t_l] (z_t_l) {$\zb_t^l$};
+        \node[det] (h_t_l) {$\vd_t^l$};
+        \node[latent, below right=1cm of h_t_l] (z_t_l) {$\vz_t^l$};
         \node[latent, right=1.5cm of h_t_l] (s_t_l) {$\vs_t^l$};
@@
         \node[latent,above=0.75cm of h_t_l] (s_t_lp1) {$\vs_t^{l+1}$};%
-        \node[latent,below=1cm of z_t_l] (e_t_l) {$\eb_t^l$};%
-        % \node[left=0.5cm of e_t_l] (e_t_l_caption) {($\eb_t^l$ conditioned only in inference)};%
+        \node[latent,below=1cm of z_t_l] (e_t_l) {$\ve_t^l$};%
+        % \node[left=0.5cm of e_t_l] (e_t_l_caption) {($\ve_t^l$ conditioned only in inference)};%
 
@@
 \caption{CW-VAE cell state $\vs_t^l$ update.
-The cell state is given as $\vs_t^l=(\zb_t^l, \db_t^l)$ where $\db_t^l$ is the deterministic hidden state of a Gated Recurrent Unit \cite{cho_properties_2014}.
-The vector $\eb_t^l$ is computed from $\xb_t$ by the encoder network which outputs $L$ encodings, one for each latent variable, similar to that of a Ladder VAE \cite{sonderby_ladder_2016}.
+The cell state is given as $\vs_t^l=(\vz_t^l, \vd_t^l)$ where $\vd_t^l$ is the deterministic hidden state of a Gated Recurrent Unit \cite{cho_properties_2014}.
+The vector $\ve_t^l$ is computed from $\vx_t$ by the encoder network which outputs $L$ encodings, one for each latent variable, similar to that of a Ladder VAE \cite{sonderby_ladder_2016}.
 All blue arrows are shared between generation and inference.
@@
     \def\col{blue}
-    \resizebox{!}{5.3cm}{
+    \resizebox{!}{5.25cm}{
     \begin{tikzpicture}
@@
         % nodes
-        \node[obs] (x_1) {$\xb_t$};%
-        \node[obs,right=1.00cm of x_1] (x_2) {$\xb_{t+1}$};%
-        \node[obs,right=1.00cm of x_2] (x_3) {$\xb_{t+2}$};%
-        \node[obs,right=1.00cm of x_3] (x_4) {$\xb_{t+3}$};%
-        \node[obs,right=1.00cm of x_4] (x_5) {$\xb_{t+4}$};%
-        \node[obs,right=1.00cm of x_5] (x_6) {$\xb_{t+5}$};%
-        \node[obs,right=1.00cm of x_6] (x_7) {$\xb_{t+6}$};%
-        \node[obs,right=1.00cm of x_7] (x_8) {$\xb_{t+7}$};%
-
-        \node[latent,above=1.00cm of x_1](s_1_1){$\zb^{(1)}_{t}$}; %
-        \node[latent,above=1.00cm of x_2](s_1_2){$\zb^{(1)}_{t+1}$}; %
-        \node[latent,above=1.00cm of x_3](s_1_3){$\zb^{(1)}_{t+2}$}; %
-        \node[latent,above=1.00cm of x_4](s_1_4){$\zb^{(1)}_{t+3}$}; %
-        \node[latent,above=1.00cm of x_5](s_1_5){$\zb^{(1)}_{t+4}$}; %
-        \node[latent,above=1.00cm of x_6](s_1_6){$\zb^{(1)}_{t+5}$}; %
-        \node[latent,above=1.00cm of x_7](s_1_7){$\zb^{(1)}_{t+6}$}; %
-        \node[latent,above=1.00cm of x_8](s_1_8){$\zb^{(1)}_{t+7}$}; %
-
-        \node[latent,above=1.00cm of s_1_1](s_2_1){$\zb^{(2)}_{t}$}; %
-        \node[latent,above=1.00cm of s_1_3](s_2_3){$\zb^{(2)}_{t+2}$}; %
-        \node[latent,above=1.00cm of s_1_5](s_2_5){$\zb^{(2)}_{t+4}$}; %
-        \node[latent,above=1.00cm of s_1_7](s_2_7){$\zb^{(2)}_{t+6}$}; %
-
-        \node[latent,above=1.00cm of s_2_1](s_3_1){$\zb^{(3)}_{t}$}; %
-        \node[latent,above=1.00cm of s_2_5](s_3_5){$\zb^{(3)}_{t+4}$}; %
+        \node[obs] (x_1) {$\vx_t$};%
+        \node[obs,right=1.00cm of x_1] (x_2) {$\vx_{t+1}$};%
+        \node[obs,right=1.00cm of x_2] (x_3) {$\vx_{t+2}$};%
+        \node[obs,right=1.00cm of x_3] (x_4) {$\vx_{t+3}$};%
+        \node[obs,right=1.00cm of x_4] (x_5) {$\vx_{t+4}$};%
+        \node[obs,right=1.00cm of x_5] (x_6) {$\vx_{t+5}$};%
+        \node[obs,right=1.00cm of x_6] (x_7) {$\vx_{t+6}$};%
+        \node[obs,right=1.00cm of x_7] (x_8) {$\vx_{t+7}$};%
+
+        \node[latent,above=1.00cm of x_1](s_1_1){$\vz^{(1)}_{t}$}; %
+        \node[latent,above=1.00cm of x_2](s_1_2){$\vz^{(1)}_{t+1}$}; %
+        \node[latent,above=1.00cm of x_3](s_1_3){$\vz^{(1)}_{t+2}$}; %
+        \node[latent,above=1.00cm of x_4](s_1_4){$\vz^{(1)}_{t+3}$}; %
+        \node[latent,above=1.00cm of x_5](s_1_5){$\vz^{(1)}_{t+4}$}; %
+        \node[latent,above=1.00cm of x_6](s_1_6){$\vz^{(1)}_{t+5}$}; %
+        \node[latent,above=1.00cm of x_7](s_1_7){$\vz^{(1)}_{t+6}$}; %
+        \node[latent,above=1.00cm of x_8](s_1_8){$\vz^{(1)}_{t+7}$}; %
+
+        \node[latent,above=1.00cm of s_1_1](s_2_1){$\vz^{(2)}_{t}$}; %
+        \node[latent,above=1.00cm of s_1_3](s_2_3){$\vz^{(2)}_{t+2}$}; %
+        \node[latent,above=1.00cm of s_1_5](s_2_5){$\vz^{(2)}_{t+4}$}; %
+        \node[latent,above=1.00cm of s_1_7](s_2_7){$\vz^{(2)}_{t+6}$}; %
+
+        \node[latent,above=1.00cm of s_2_1](s_3_1){$\vz^{(3)}_{t}$}; %
+        \node[latent,above=1.00cm of s_2_5](s_3_5){$\vz^{(3)}_{t+4}$}; %
 
@@
 
-        % \node[above=of s_3_5, yshift=-1.cm, xshift=-2cm] (phi) {$p_\theta(\xb,\zb)$};
+        % \node[above=of s_3_5, yshift=-1.cm, xshift=-2cm] (phi) {$p_\theta(\vx,\vz)$};
     \end{tikzpicture}
@@
 
-        % \node[above=of s_3_5, yshift=-1.cm, xshift=-2cm] (phi) {$q_\phi(\zb|\xb)$};
+        % \node[above=of s_3_5, yshift=-1.cm, xshift=-2cm] (phi) {$q_\phi(\vz|\vx)$};
     \end{tikzpicture}
@@
 \caption{
-    CW-VAE \cite{saxena_clockwork_2021} generative model $p(\xb,\zb)$ in (\subref{fig: cwvae three-layered graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: cwvae three-layered graphical model inference}) for a three-layered model with $k_1=1$ and $c=2$ giving $k_2=2$ and $k_3=4$ unrolled over eight steps in the observed variable.
+    CW-VAE \cite{saxena_clockwork_2021} generative model $p(\vx,\vz)$ in (\subref{fig: cwvae three-layered graphical model generative}) and inference model $q(\vz|\vx)$ in (\subref{fig: cwvae three-layered graphical model inference}) for a three-layered model with $k_1=1$ and $c=2$ giving $k_2=2$ and $k_3=4$ unrolled over eight steps in the observed variable.
 Blue arrows are (mostly) shared between the inference and generative models.
-See \cref{fig: cwvae cell state graph} for a detailed graphical model expanding on the latent nodes $\zb^l_t$ and parameter sharing.
+See \cref{fig: cwvae cell state graph} for a detailed graphical model expanding on the latent nodes $\vz^l_t$ and parameter sharing.
 }
@@
     \def\col{blue}
-    \resizebox{!}{5.0cm}{
+    \resizebox{!}{5.5cm}{
     \begin{tikzpicture}
@@
         % nodes
-        \node[obs] (x_1) {$\xb_t$};%
-        \node[obs,left=.75cm of x_1] (x_2) {$\xb_{t-1}$};
-        \node[obs,left=.75cm of x_2] (x_3) {$\xb_{t-2}$};
-        \node[obs,left=.75cm of x_3] (x_4) {$\xb_{t-3}$};
-        \node[obs,left=.75cm of x_4] (x_5) {$\xb_{t-4}$};
-        \node[obs,left=.75cm of x_5] (x_6) {$\xb_{t-5}$};
-        \node[obs,left=.75cm of x_6] (x_7) {$\xb_{t-6}$};
-        \node[obs,left=.75cm of x_7] (x_8) {$\xb_{t-7}$};
-
-        \node[det,above=.75cm of x_1](d_1_1){$\db_{t}^{(1)}$};
-        \node[det,above=.75cm of x_3](d_1_3){$\db_{t-2}^{(1)}$};
-        \node[det,above=.75cm of x_5](d_1_5){$\db_{t-4}^{(1)}$};
-        \node[det,above=.75cm of x_7](d_1_7){$\db_{t-6}^{(1)}$};
-
-        \node[det,above=.75cm of d_1_1](d_2_1){$\db_{t}^{(2)}$};
-        \node[det,above=.75cm of d_1_5](d_2_5){$\db_{t-4}^{(2)}$};
-
-        \node[det,above=.75cm of d_2_1](d_3_1){$\db_{t}^{(3)}$};
-
-        \node[latent,right=.75cm of d_1_1](z_1){$\zb_{t+1}^{(1)}$};
-        \node[latent,right=.75cm of d_2_1](z_2){$\zb_{t+1}^{(2)}$};
-        \node[latent,right=.75cm of d_3_1](z_3){$\zb_{t+1}^{(3)}$};
-
-        \node[latent,right=.75cm of z_2](z){$\zb_{t+1}$};
-
-        \node[obs,below=2.8cm of z] (x_0) {$\xb_{t+1}$};
+        \node[obs] (x_1) {$\vx_t$};%
+        \node[obs,left=.75cm of x_1] (x_2) {$\vx_{t-1}$};
+        \node[obs,left=.75cm of x_2] (x_3) {$\vx_{t-2}$};
+        \node[obs,left=.75cm of x_3] (x_4) {$\vx_{t-3}$};
+        \node[obs,left=.75cm of x_4] (x_5) {$\vx_{t-4}$};
+        \node[obs,left=.75cm of x_5] (x_6) {$\vx_{t-5}$};
+        \node[obs,left=.75cm of x_6] (x_7) {$\vx_{t-6}$};
+        \node[obs,left=.75cm of x_7] (x_8) {$\vx_{t-7}$};
+
+        \node[det,above=.75cm of x_1](d_1_1){$\vd_{t}^{(1)}$};
+        \node[det,above=.75cm of x_3](d_1_3){$\vd_{t-2}^{(1)}$};
+        \node[det,above=.75cm of x_5](d_1_5){$\vd_{t-4}^{(1)}$};
+        \node[det,above=.75cm of x_7](d_1_7){$\vd_{t-6}^{(1)}$};
+
+        \node[det,above=.75cm of d_1_1](d_2_1){$\vd_{t}^{(2)}$};
+        \node[det,above=.75cm of d_1_5](d_2_5){$\vd_{t-4}^{(2)}$};
+
+        \node[det,above=.75cm of d_2_1](d_3_1){$\vd_{t}^{(3)}$};
+
+        \node[latent,right=.75cm of d_1_1](z_1){$\vz_{t+1}^{(1)}$};
+        \node[latent,right=.75cm of d_2_1](z_2){$\vz_{t+1}^{(2)}$};
+        \node[latent,right=.75cm of d_3_1](z_3){$\vz_{t+1}^{(3)}$};
+
+        \node[latent,right=.75cm of z_2](z){$\vz_{t+1}$};
+
+        \node[obs,below=2.8cm of z] (x_0) {$\vx_{t+1}$};
 
@@
 
-        % \node[above=of d_1_3, yshift=3.cm] (theta) {$p_\theta(\zb,\xb)$};
+        % \node[above=of d_1_3, yshift=3.cm] (theta) {$p_\theta(\vz,\vx)$};
     \end{tikzpicture}
@@
         % nodes
-        \node[obs] (x_1) {$\xb_t$};%
-        \node[obs,left=.75cm of x_1] (x_2) {$\xb_{t-1}$};
-        \node[obs,left=.75cm of x_2] (x_3) {$\xb_{t-2}$};
-        \node[obs,left=.75cm of x_3] (x_4) {$\xb_{t-3}$};
-        \node[obs,left=.75cm of x_4] (x_5) {$\xb_{t-4}$};
-        \node[obs,left=.75cm of x_5] (x_6) {$\xb_{t-5}$};
-        \node[obs,left=.75cm of x_6] (x_7) {$\xb_{t-6}$};
-        \node[obs,left=.75cm of x_7] (x_8) {$\xb_{t-7}$};
-
-        \node[det,above=.75cm of x_1](d_1_1){$\db_{t}^{(1)}$};
-        \node[det,above=.75cm of x_3](d_1_3){$\db_{t-2}^{(1)}$};
-        \node[det,above=.75cm of x_5](d_1_5){$\db_{t-4}^{(1)}$};
-        \node[det,above=.75cm of x_7](d_1_7){$\db_{t-6}^{(1)}$};
-
-        \node[det,above=.75cm of d_1_1](d_2_1){$\db_{t}^{(2)}$};
-        \node[det,above=.75cm of d_1_5](d_2_5){$\db_{t-4}^{(2)}$};
-
-        \node[det,above=.75cm of d_2_1](d_3_1){$\db_{t}^{(3)}$};
-
-        \node[latent,right=.75cm of d_1_1](z_1){$\zb_{t}^{(1)}$};
-        \node[latent,right=.75cm of d_2_1](z_2){$\zb_{t}^{(2)}$};
-        \node[latent,right=.75cm of d_3_1](z_3){$\zb_{t}^{(3)}$};
-
-        \node[latent,right=.75cm of z_2](z){$\zb_{t}$};
+        \node[obs] (x_1) {$\vx_t$};%
+        \node[obs,left=.75cm of x_1] (x_2) {$\vx_{t-1}$};
+        \node[obs,left=.75cm of x_2] (x_3) {$\vx_{t-2}$};
+        \node[obs,left=.75cm of x_3] (x_4) {$\vx_{t-3}$};
+        \node[obs,left=.75cm of x_4] (x_5) {$\vx_{t-4}$};
+        \node[obs,left=.75cm of x_5] (x_6) {$\vx_{t-5}$};
+        \node[obs,left=.75cm of x_6] (x_7) {$\vx_{t-6}$};
+        \node[obs,left=.75cm of x_7] (x_8) {$\vx_{t-7}$};
+
+        \node[det,above=.75cm of x_1](d_1_1){$\vd_{t}^{(1)}$};
+        \node[det,above=.75cm of x_3](d_1_3){$\vd_{t-2}^{(1)}$};
+        \node[det,above=.75cm of x_5](d_1_5){$\vd_{t-4}^{(1)}$};
+        \node[det,above=.75cm of x_7](d_1_7){$\vd_{t-6}^{(1)}$};
+
+        \node[det,above=.75cm of d_1_1](d_2_1){$\vd_{t}^{(2)}$};
+        \node[det,above=.75cm of d_1_5](d_2_5){$\vd_{t-4}^{(2)}$};
+
+        \node[det,above=.75cm of d_2_1](d_3_1){$\vd_{t}^{(3)}$};
+
+        \node[latent,right=.75cm of d_1_1](z_1){$\vz_{t}^{(1)}$};
+        \node[latent,right=.75cm of d_2_1](z_2){$\vz_{t}^{(2)}$};
+        \node[latent,right=.75cm of d_3_1](z_3){$\vz_{t}^{(3)}$};
+
+        \node[latent,right=.75cm of z_2](z){$\vz_{t}$};
 
@@
 
-        % \node[above=of d_1_3, yshift=3.cm] (phi){$q_\phi(\zb|\xb)$};
+        % \node[above=of d_1_3, yshift=3.cm] (phi){$q_\phi(\vz|\vx)$};
     \end{tikzpicture}
@@
 \caption{
-    STCN \cite{aksan_stcn_2019} generative model $p(\xb,\zb)$ in (\subref{fig: stcn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: stcn graphical model inference}) for a single time-step.
+    STCN \cite{aksan_stcn_2019} generative model $p(\vx,\vz)$ in (\subref{fig: stcn graphical model generative}) and inference model $q(\vz|\vx)$ in (\subref{fig: stcn graphical model inference}) for a single time-step.
     The WaveNet autoregressive encoder is shared between generative and inference models.
@@
 In practice, the model uses ten layers in each of five stacks/cycles resulting in a much larger receptive field.
-Importantly, the model parameterizes the five latent variables using the last deterministic representation $\db^{(l)}$ from each stack, i.e.
+Importantly, the model parameterizes the five latent variables using the last deterministic representation $\vd^{(l)}$ from each stack, i.e.
 only every fifth $l$ starting from $l=5$ and ending at $l=25$.
-    Note that the generative model uses the prior to transform the WaveNet hidden states $\db_{t}^{(l)}$ into the latent variable $\zb_{t+1}^{(l)}$ one step ahead in time compared to the approximate posterior which infers $\zb_{t}^{(l)}$.
-    Also note that $\zb_t$ is constructed by concatenating all $\zb_t^{(l)}$.
-The original paper explores setting $\zb_t$ equal to $\zb_t^{(1)}$.
-The best-performing STCN for speech, which also the one we implement, uses a WaveNet decoder to predict $\xb_{t+1}$ from a sequence of $\zb_t$ rather than a per-timestep transform.
+    Note that the generative model uses the prior to transform the WaveNet hidden states $\vd_{t}^{(l)}$ into the latent variable $\vz_{t+1}^{(l)}$ one step ahead in time compared to the approximate posterior which infers $\vz_{t}^{(l)}$.
+    Also note that $\vz_t$ is constructed by concatenating all $\vz_t^{(l)}$.
+The original paper explores setting $\vz_t$ equal to $\vz_t^{(1)}$.
+The best-performing STCN for speech, which also the one we implement, uses a WaveNet decoder to predict $\vx_{t+1}$ from a sequence of $\vz_t$ rather than a per-timestep transform.
     Blue arrows are shared between the inference and generative models.
@@
         % nodes
-        \node[obs] (x_1) {$\xb_t$};%
-        \node[obs,right=1.00cm of x_1] (x_2) {$\xb_{t+1}$};
-        \node[obs,right=1.00cm of x_2] (x_3) {$\xb_{t+2}$};
-
-        \node[latent,above=1.00cm of x_1](z_1_1){$\zb_{t}$};
-        \node[latent,above=1.00cm of x_2](z_1_2){$\zb_{t+1}$};
-        \node[latent,above=1.00cm of x_3](z_1_3){$\zb_{t+2}$};
-
-        \node[det,above=1.00cm of z_1_1](d_1_1){$\db_{t}$};
-        \node[det,above=1.00cm of z_1_2](d_1_2){$\db_{t+1}$};
-        \node[det,above=1.00cm of z_1_3](d_1_3){$\db_{t+2}$};
+        \node[obs] (x_1) {$\vx_t$};%
+        \node[obs,right=1.00cm of x_1] (x_2) {$\vx_{t+1}$};
+        \node[obs,right=1.00cm of x_2] (x_3) {$\vx_{t+2}$};
+
+        \node[latent,above=1.00cm of x_1](z_1_1){$\vz_{t}$};
+        \node[latent,above=1.00cm of x_2](z_1_2){$\vz_{t+1}$};
+        \node[latent,above=1.00cm of x_3](z_1_3){$\vz_{t+2}$};
+
+        \node[det,above=1.00cm of z_1_1](d_1_1){$\vd_{t}$};
+        \node[det,above=1.00cm of z_1_2](d_1_2){$\vd_{t+1}$};
+        \node[det,above=1.00cm of z_1_3](d_1_3){$\vd_{t+2}$};
 
@@
 
-        \node[above=of d_1_2, yshift=-1.cm] (phi) {$p_\theta(\zb,\xb)$};
+        \node[above=of d_1_2, yshift=-1.cm] (phi) {$p_\theta(\vz,\vx)$};
     \end{tikzpicture}
@@
 
-        \node[above=of d_1_2, yshift=-1.cm] (phi) {$q_\phi(\zb|\xb)$};
+        \node[above=of d_1_2, yshift=-1.cm] (phi) {$q_\phi(\vz|\vx)$};
     \end{tikzpicture}
@@
 \caption{
-    VRNN \cite{chung_recurrent_2015} generative model $p(\xb,\zb)$ in (\subref{fig: vrnn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: vrnn graphical model inference}) unrolled over three steps in the observed variable.
+    VRNN \cite{chung_recurrent_2015} generative model $p(\vx,\vz)$ in (\subref{fig: vrnn graphical model generative}) and inference model $q(\vz|\vx)$ in (\subref{fig: vrnn graphical model inference}) unrolled over three steps in the observed variable.
     Blue arrows are shared between the inference and generative models.
@@
         % nodes
-        \node[obs] (x_1) {$\xb_t$};%
-        \node[obs,right=1.00cm of x_1] (x_2) {$\xb_{t+1}$};
-        \node[obs,right=1.00cm of x_2] (x_3) {$\xb_{t+2}$};
-
-        \node[latent,above=1.00cm of x_1](z_1_1){$\zb_{t}$};
-        \node[latent,above=1.00cm of x_2](z_1_2){$\zb_{t+1}$};
-        \node[latent,above=1.00cm of x_3](z_1_3){$\zb_{t+2}$};
-
-        \node[det,above=1.00cm of z_1_1](a_1_1){$\ab_{t}$};
-        \node[det,above=1.00cm of z_1_2](a_1_2){$\ab_{t+1}$};
-        \node[det,above=1.00cm of z_1_3](a_1_3){$\ab_{t+2}$};
-
-        \node[det,above=1.00cm of a_1_1](d_1_1){$\db_{t}$};
-        \node[det,above=1.00cm of a_1_2](d_1_2){$\db_{t+1}$};
-        \node[det,above=1.00cm of a_1_3](d_1_3){$\db_{t+2}$};
+        \node[obs] (x_1) {$\vx_t$};%
+        \node[obs,right=1.00cm of x_1] (x_2) {$\vx_{t+1}$};
+        \node[obs,right=1.00cm of x_2] (x_3) {$\vx_{t+2}$};
+
+        \node[latent,above=1.00cm of x_1](z_1_1){$\vz_{t}$};
+        \node[latent,above=1.00cm of x_2](z_1_2){$\vz_{t+1}$};
+        \node[latent,above=1.00cm of x_3](z_1_3){$\vz_{t+2}$};
+
+        \node[det,above=1.00cm of z_1_1](a_1_1){$\va_{t}$};
+        \node[det,above=1.00cm of z_1_2](a_1_2){$\va_{t+1}$};
+        \node[det,above=1.00cm of z_1_3](a_1_3){$\va_{t+2}$};
+
+        \node[det,above=1.00cm of a_1_1](d_1_1){$\vd_{t}$};
+        \node[det,above=1.00cm of a_1_2](d_1_2){$\vd_{t+1}$};
+        \node[det,above=1.00cm of a_1_3](d_1_3){$\vd_{t+2}$};
 
@@
 \caption{
-    SRNN \cite{fraccaro_sequential_2016} generative model $p(\xb,\zb)$ in (\subref{fig: srnn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: srnn graphical model inference}) unrolled over three steps in the observed variable.
+    SRNN \cite{fraccaro_sequential_2016} generative model $p(\vx,\vz)$ in (\subref{fig: srnn graphical model generative}) and inference model $q(\vz|\vx)$ in (\subref{fig: srnn graphical model inference}) unrolled over three steps in the observed variable.
     Blue arrows are shared between the inference and generative models.
@@
 \section{Additional latent evaluation}\label{app: additional latent space clustering}
+Here we present additional qualitative assesessment of the learned latent representations selectively for the CW-VAE.
+
+In \cref{fig: latent space phoneme and knn} we present evaluation in terms of phoneme clustering. 
+Specifically, we infer the latent variables of all utterances by a single speaker from the TIMIT test set.
+We sample the latent sequence $100$ times to estimate the mean representation per time step.
+We then compute the average latent representation over the duration of each phoneme using aligned phoneme labels.
+This approximately marginalizes out variation during the phoneme.
+We use linear discriminant analysis (LDA) \cite{fisher_use_1936} to obtain a low-dimensional linear subspace of the latent space.
+We visualize the resulting representations colored according to test set phoneme classes in \cref{fig: latent space phoneme and knn}. 
+In the left plot, many phonemes are separable in the linear subspace and that related phonemes such as ``s" and ``sh" are close. 
+In the right plot, we show the average accuracy of a leave-one-out $k$-nearest-neighbor (KNN) classifier on the single left-out latent representation reduced with a 5-dimensional LDA as a function of the number of neighbors.
+We compare accuracy to a Mel-spectrogram averaged over each phoneme duration and LDA reduced.
+The spectrogram is computed with hop length set to 64, equal to $s_1$ for the CW-VAE, window size 256 and 80 Mel bins.
+We see that both latent spaces yield significantly better KNN accuracies than the Mel features.
+
 We visualize the performance of a $k$-nearest-neighbour classifier for classification of speaker gender and height in \cref{fig: knn fraction latent space height and gender}.
@@
 In both cases, we see improved performance from using the learned latent space over Mel-features.
-Notably, $\zb^2$ is outperformed by the Mel-features for gender identification which may indicate that $\zb^2$ learns to ignore this attribute compared to $\zb^1$.
+Notably, $\vz^2$ is outperformed by the Mel-features for gender identification which may indicate that $\vz^2$ learns to ignore this attribute compared to $\vz^1$.
 
@@
     \centering
-    \includegraphics[width=0.477\textwidth]{paper_benchmarking/1xrnjn5y_speaker_gender_10_samples_knn_correct_fraction_model_vs_mel_100_neighbors_lda_subspace_1d.pdf}
+    \includegraphics[width=0.555\textwidth]{
+        graphics/1xrnjn5y_speaker_phonemes_drw0_latent_0_samples_100_lda_linear_subspace.pdf
+    }
     \hfill
-    \includegraphics[width=0.483\textwidth]{paper_benchmarking/1xrnjn5y_speaker_height_m_10_samples_knn_correct_fraction_model_vs_mel_50_neighbors_lda_subspace_2d.pdf}
+    \includegraphics[width=0.425\textwidth]{
+        graphics/1xrnjn5y_speaker_phonemes_drw0_samples_100_knn_correct_fraction_model_vs_mel_30_neighbors_lda_subspace_5d.pdf
+    }
+    \caption{
+    (left) Clustering of phonemes in a 2D Linear Discriminant Analysis (LDA) subspace of a CW-VAE latent space ($\vz^{(1)}$).
+    (right) Leave-one-out phoneme classification accuracy for a KNN classifier at different $K$ in a 5D LDA subspace of a CW-VAE latent space.
+    }
+    \label{fig: latent space phoneme and knn}
+\end{figure*}
+
+\begin{figure*}[t!]
+     \centering
+     \includegraphics[width=0.477\textwidth]{graphics/1xrnjn5y_speaker_gender_10_samples_knn_correct_fraction_model_vs_mel_100_neighbors_lda_subspace_1d.pdf}
+     \hfill
+     \includegraphics[width=0.483\textwidth]{graphics/1xrnjn5y_speaker_height_m_10_samples_knn_correct_fraction_model_vs_mel_50_neighbors_lda_subspace_2d.pdf}
     \caption{
@@
 \end{figure*}
+
 \begin{figure}[t!]
@@
          \centering
-         \includegraphics[width=\textwidth]{paper_benchmarking/1xrnjn5y_speaker_gender_latent_0_1_samples_lda_linear_subspace.pdf}
+         \includegraphics[width=\textwidth]{graphics/1xrnjn5y_speaker_gender_latent_0_1_samples_lda_linear_subspace.pdf}
          \caption{}
@@
          \centering
-         \includegraphics[width=\textwidth]{paper_benchmarking/1xrnjn5y_speaker_gender_mel_features_lda_linear_subspace.pdf}
+         \includegraphics[width=\textwidth]{graphics/1xrnjn5y_speaker_gender_mel_features_lda_linear_subspace.pdf}
          \caption{}
@@
          \centering
-         \includegraphics[width=\textwidth]{paper_benchmarking/1xrnjn5y_speaker_height_m_1_samples_latent_0_lda_linear_subspace.pdf}
+         \includegraphics[width=\textwidth]{graphics/1xrnjn5y_speaker_height_m_1_samples_latent_0_lda_linear_subspace.pdf}
          \caption{}
@@
          \centering
-         \includegraphics[width=\textwidth]{paper_benchmarking/1xrnjn5y_speaker_age_f_1_samples_latent_1_lda_linear_subspace.pdf}
+         \includegraphics[width=\textwidth]{graphics/1xrnjn5y_speaker_age_f_1_samples_latent_1_lda_linear_subspace.pdf}
          \caption{}
@@
         \centering
-        \includegraphics[width=\textwidth]{paper_benchmarking/timit_validation_set_phoneme_duration_boxplot_speaker_id_drw0.pdf}
+        \includegraphics[width=\textwidth]{graphics/timit_validation_set_phoneme_duration_boxplot_speaker_id_drw0.pdf}
         \caption{}
@@
         \centering
-        \includegraphics[width=\textwidth]{paper_benchmarking/timit_validation_set_phoneme_duration_boxplot_speaker_id_none.pdf}
+        \includegraphics[width=\textwidth]{graphics/timit_validation_set_phoneme_duration_boxplot_speaker_id_none.pdf}
         \caption{}
@@
 \section{Model samples and reconstructions}\label{app: model samples and reconstructions}
-We provide samples and reconstructions for some of the models considered here at the following URL: \url{https://doi.org/10.5281/zenodo.5911899}.
+We provide samples and reconstructions for some of the models considered here at the following URL: \url{https://doi.org/10.5281/zenodo.5704512}.
 The samples are generated from the prior of Clockwork VAE, SRNN and VRNN and from a WaveNet by conditioning on pure zeros. All models are configured as those reported in \cref{tab: likelihoods timit}.
@@
 
-\iffalse
-
 \fi

### 2023-02-25 appendices/appendix_hierarchical.tex
Add appendices for benchmarking and hierarchical

%!TEX root = ../thesis.tex

\chapter[hierarchical vaes know what they don't know]{Hierarchical VAEs Know What They Don't Know}

\section{Datasets}
\cref{tab:datasets-overview} lists the datasets used in the paper. We use the predefined train/test splits for the datasets.

For SmallNORB and Omniglot we resize the original grey-scale images to $28\times28$ with ordinary bi-linear interpolation.
For each of these datasets, we also create a version where the grey-scale is inverted. We do this because, the overall white nature of the images tends to make detecting them as OOD from FashionMNIST artificially easy.
The inversion is done via the simple transformation $\xb_\text{inverted} = 255 - \xb_\text{original}$ since images are encoded as 8 bit unsigned integers.

% Similar to \citet{serra_input_2020} and \citet{xiao_likelihood_2020}, we also create a uniform noise dataset ``UniformNoise'' and benchmark our model against this. We dynamically sample each pixel value uniformly as an integer between 0 and 255.

\begin{table}[h!]
    \centering
    % \resizebox{0.7\columnwidth}{!}{%
    \begin{tabular}{llr}
        \toprule
        Dataset & Dimensionality & Examples \\
        \midrule
        FashionMNIST \cite{xiao_fashionmnist_2017} & $28\times28\times1$ & 70,000 \\
        MNIST \cite{lecun_gradientbased_1998} & $28\times28\times1$ & 70,000 \\
        notMNIST \cite{bulatov_notmnist_2011} & $28\times28\times1$ & 547,838 \\
        KMNIST \cite{clanuwat_deep_2018} & $28\times28\times1$ & 70,000 \\
        Omniglot \cite{lake_humanlevel_2015} & $28\times28\times1$ & 32,460 \\
        SmallNORB \cite{lecun_learning_2004} & $28\times28\times1$ & 97,200 \\
        % UniformNoise & $28\times28\times1$ & dynamic \\
        \midrule
        CIFAR10 \cite{krizhevsky_learning_2009} & $32\times32\times3$ & 60,000 \\
        SVHN \cite{netzer_reading_2011} & $32\times32\times3$ & 99,289 \\
        \bottomrule
    \end{tabular}
    % }
    \caption{Overview of the used datasets.}
    \label{tab:datasets-overview}
\end{table}

\section{Model details}\label{sec:model-details}
In \cref{tab:hyperparameters} we specify the hyperparameters used when training our models.
We make our source code available at \url{https://github.com/JakobHavtorn/hvae-oodd}.

\subsection{Hierarchical VAE}
Our Hierarchical VAE (HVAE) model uses bottom-up inference and top-down generative paths as specified in the paper.
For grey-scale images, the output is parameterized by a Bernoulli distribution while for natural images we use a Discretized Logistic Mixture \cite{salimans_pixelcnn_2017}.
The latent variables are parameterized by stochastic layers that output the mean and log-variance of a diagonal covariance Gaussian. The prior distribution on the top-most latent is a standard Gaussian.
For grey-scale images, the lowest latent space is parameterized by a convolutional neural network and has dimensions $14\times14\times8$ interpreted as (height $\times$ width $\times$ latent dimension). The highest two latent variables are parameterized by dense transformations with $16$ and $8$ units, respectively.
For natural images, the bottom-two latent variables are parameterized by convolutional neural networks and have dimensions $(16\times16)\times128$, $(8\times8)\times64$, respectively for $\zb_1, \zb_2$. The top-most latent, $\zb_3$, is densely connected with dimension $32$.

Each stochastic layer is preceeded by a determininistic transformation.
For both grey-scale and natural images, each deterministic transformation consists of three residual blocks of the same type used by \textcite{maaloe_biva_2019}. The structure of a residual block is:
\begin{equation}
    \yb = \textsc{conv}\left( \textsc{act} \left( \textsc{conv}_s \left( \textsc{act}(\xb) \right) \right) \right) + \xb \ ,
\end{equation}
where \textsc{conv} refers to a same-padded convolution and \textsc{act} to the activation function. Within a residual block, the first convolution always has stride 1 while the second convolution has stride $s$. In a deterministic transformation, any non-unit stride is performed in the third residual block. For grey-scale images, we stride by 2 in the first and second deterministic transformations but not the third. For natural images, we similarly stride by 2 in the first and second deterministic transformations. For grey-scale we use 64 channels while we use 256 for natural images.
In both cases, the first deterministic block uses a kernel size of 5 and the latter two a kernel of size 3. We use the ReLU activation function \cite{fukushima_neocognitron_1980, nair_rectified_2010}.

Since the benefits and drawbacks of using batch normalization \cite{ioffe_batch_2015} in hierarchical VAEs is still the matter of some debate \cite{sonderby_ladder_2016, vahdat_nvae_2020, child_very_2021} we choose to use weight normalization \cite{salimans_weight_2016} as in other work \cite{maaloe_biva_2019} and initialize the model using the originally proposed data-dependent initialization.
To have the stochastic layers initialize to standard Gaussian distributions (zero mean, unit variance), with this initialization, we select the activation function for the variance as a Softplus,
\begin{equation*}
    \textsc{softplus}(\xb) = \frac{1}{\beta}\log\left(1 + \exp(\beta\xb)\right) \ ,
\end{equation*}
with $\beta = \log(2) \approx 0.693$ to output 1 for $\xb = 0$.

Training of a HVAE model took approximately two days on a single NVIDIA GTX 1080 Ti graphics card.

\subsection{BIVA}
For the BIVA model \cite{maaloe_biva_2019}, we use a specification that is very similar to that of the HVAE above, and to that of the original paper.
The model has 10 latent variables the lowest 3 of which are spatial and the rest are densely connected in order to have an architecture similar to the HVAE.
The model uses an overall stride of 8, achieved by striding by 2 in the first, fourth and sixth deterministic transformations.
From $\zb_1$ to $\zb_{10}$, the latents have the following dimensions: The lowest three latents are spatial $(16\times16)\times8$, $(16\times16)\times16$ and $(16\times16)\times32$, given as $(\text{height}\times\text{width})\times\text{dim})$, while the rest are dense vectors with dimensions of $42, 40, 38, 36, 34, 32, 30$.

Training of a BIVA model took approximately a week on a single NVIDIA GTX 1080 Ti graphics card.

\begin{table}[h]
    \centering
    % \resizebox{\columnwidth}{!}{%
    \begin{tabular}{ll}
        \toprule
        Hyperparameter & Setting/Range \\
        \midrule
        \multicolumn{2}{l}{\textbf{All}} \\
        \midrule
        Optimization & Adam \cite{kingma_adam_2015} \\
        Learning rate & $3e-4$ \\
        Batch size & 128 \\
        Epochs & 2000 \\
        Free bits & $\SI{2}{nats}$ per $\zb_i$ shared across latent dim. \\
        Free bits constant & 200 epochs \\
        Free bits annealed & 200 epochs \\
        Activation & ReLU \\
        \multirow{2}{*}{Initialization} & Data-dependent \\
        & \cite{salimans_weight_2016} \\
        \midrule
        \multicolumn{2}{l}{\textbf{HVAE}} \\
        Latent dimensionality & 128-64-32 (natural) / 8-16-8 (grey) \\
        Convolution kernel & 5-3-3 \\
        Stride & 2-2-1 \\
        Warmup anneal period & 200 epochs \\
        \midrule
        \multicolumn{2}{l}{\textbf{BIVA}} \\
        \multirow{2}{*}{Latent dimensionality} & 10-8-6 (spatial) \\
        & 42-40-38-36-34-32-30 (dense) \\
        Convolution kernel & 5-3-3-3-3-3-3-3-3-3 \\
        Stride & {2-1-1-2-1-2-1-1-1-1} \\
        \bottomrule
    \end{tabular}
    % }
    \caption{Selection of most important hyperparameters and their setting. Convolutional kernels are square and latent dimensions are given without spatial dimensions which are given in the text. See \cref{sec:model-details} for more details.}
    \label{tab:hyperparameters}
\end{table}

\section{Analysis of the influence of latent variables on the marginal likelihood}
In the paper, we argue that the lowest level latent variables, which have the highest dimensionality, contribute the most to the approximate likelihood.
Here, we provide a stringent mathematical argument that generalizes this to the exact marginal likelihood in a model with a deterministic decoder.

\subsection{Model specification}
For an arbitrary hierarchical latent variable model, we have a prior $p(\zb_L)$ and a generative mapping $f: \mathbb{R}^d \rightarrow \mathbb{R}^D$, such that $\xb = f(\zb_L)$ and $D>d$.
Note that we will assume that $f$ is deterministic, such that we are effectively working with $p(\xb|\zb) = \delta_{f(\zb)}(\xb)$.
This is a limiting assumption, but it allows working through the following. For shorthand we will simply write $\zb=\zb_L$.

Let $f$ have a bottleneck architecture, i.e.\
\begin{align}
f(\zb) &= f_1(\ldots f_{L-1}(f_L(\zb))) \ ,
\end{align}
where
\begin{align}
    f_i: \mathbb{R}^{d_{i}} \rightarrow \mathbb{R}^{d_{i-1}}, \qquad i = L, \ldots, 1 \ .
\end{align}
Here we use the notation $d_0=D=\vert\xb\vert$ and $d_L=d=\vert\zb\vert$ and further assume $d_0\geq d_1\geq \ldots \geq d_{L-1} \geq d_L$ which gives the bottleneck.
%$d_1 = d$ and $d_{L+1} = D$ and further assume $d = d_1 \leq d_2 \leq \ldots \leq d_L \leq d_{L+1} = D$ which gives the bottleneck.

Assuming $\xb$ is such that a corresponding latent variable $\zb$ exists, i.e.\ that there exists $\zb$ such that $\xb = f(\zb)$, then we can write the likelihood of $\xb$ through a standard change of variables (similar to flow-based models),
\begin{align}
    p(\xb) &= p(\zb) \prod_{i = 1}^L \left(\sqrt{\det \Jb_i^T \Jb_i}\right)^{-1} \ ,
\end{align}
where $\Jb_i$ is the Jacobian of $f_i$, i.e.
\begin{align}
    \Jb_i &= \frac{\partial f_i}{\partial \zb_i} \in \mathbb{R}^{d_i\times d_{i-1}} \ .
\end{align}
Here we use the notation that $\zb_i$ is the representation at layer $i$.
Note that $\Jb_i^T \Jb_i$ is a $d_{i-1} \times d_{i-1}$ symmetric positive semidefinite matrix (determinant $\geq 0$).

The log-likelihood can be written as
\begin{equation}
    \log p(\xb) = \log p(\zb) - \frac{1}{2} \sum_{i=1}^L \log \det \Jb_i^T\Jb_i \ .
\end{equation}

By construction of determinants, we can generally expect these determinants to grow with the dimensionality of the matrix.
We should expect the determinant of a $d \times d$ matrix to be of the order $\mathcal{O}(\lambda^d)$ for some number $\lambda>0$.
With that in mind, we should generally expect that
\begin{align}
    \det \Jb_{i+1}^T \Jb_{i+1} < \det \Jb_i^T \Jb_i \ ,
\end{align}
due to the bottleneck assumption.
If so, we see that the marginal likelihood $p(\xb)$ will be dominated by $\left(\sqrt{\det \Jb_1^T \Jb_1}\right)^{-1}$, i.e.\ low-level features have a higher influence on the likelihood than more important semantic ones.

\subsection{The Gaussian case}
%
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{paper_hierarchical/logdet2.pdf}
    \caption{The expected inverse volume change for Gaussian Jacobians \eqref{eq:Edet} on a log-scale.}
    \label{fig:Edet}
\end{figure}
The previous remarks can be made more precise if we make distributional assumptions on the Jacobians.
Here we will assume that the Jacobians of each layer follow a Gaussian distribution.
Specifically, we will assume that each entry in $\Jb_i$ is distributed as $\mathcal{N}(0, \sigma^2)$.
The analysis below extends to nonzero means and more general covariance structure, but this comes with a cost of less transparent notation.
In this setting, $\Jb_i^T \Jb_i$ follows a Wishart distribution (in the general setting it would follow a non-central Wishart distribution).
\textcite{muirhead_aspects_2009} tells us that the expected multiplicative contribution to the likelihood of each layer is
\begin{align}
    \mathbb{E}\left[ \left(\sqrt{\det \Jb_i^T \Jb_i}\right)^{-1} \right]
    &= \sigma^{-d_{i-1}}  2^{-\frac{d_{i-1}}{2}} \frac{\Gamma_{d_{i-1}}\left( \frac{1}{2} d_{i} - \frac{1}{2} \right)}{\Gamma_{d_{i-1}}\left( \frac{1}{2} d_{i} \right)} \notag \\
    &= \sigma^{-d_{i-1}}  2^{-\frac{d_{i-1}}{2}} \frac{\Gamma\left( \frac{1}{2} (d_{i} - d_{i-1}) \right)}{\Gamma\left( \frac{1}{2} d_{i} \right)}
    \label{eq:Edet}
\end{align}
where $\Gamma_d$ is the multivariate Gamma function.
Assuming that the increase in layer dimension $d_{i} - d_{i-1}$ is constant, then we see that \eqref{eq:Edet} goes to zero as $d_{i}$ goes to infinity as the $\Gamma$ function grows super-exponentially to infinity.
This super-exponential growth further implies that the first layers dominate the marginal likelihood $p(\xb)$.
This is also visually evident in \cref{fig:Edet}.

\section{Derivation of the $\mathcal{L}^{>k}$ bound}
In this section we present the derivation of $\mathcal{L}^{>k}$ and show that it is a lower bound on the marginal likelihood.

First, we consider a two-layered VAE with bottom-up inference.
We proceed very similarly to the derivation of the regular ELBO and also use Jensen's inequality.
\begin{align}
\log p(\xb) &= \log \int\int p(\xb|\zb_1)p(\zb_1|\zb_2)p(\zb_2) \text{d}\zb_1\text{d}\zb_2 \\
&= \log \int\int \frac{q(\zb_2|\xb)}{q(\zb_2|\xb)} p(\xb|\zb_1)p(\zb_1|\zb_2)p(\zb_2) \text{d}\zb_1\text{d}\zb_2 \notag \\
&= \log \int\int q(\zb_2|\xb) p(\zb_1|\zb_2) \frac{p(\xb|\zb_1)p(\zb_2)}{q(\zb_2|\xb)} \text{d}\zb_1\text{d}\zb_2 \notag \\
%&\ge \mathbb{E}_{p(\zb_1|\zb_2)} \left [ \log q(\zb_2|\xb) \frac{p(\xb|\zb_1)p(\zb_2)}{q(\zb_2|\xb)} \right ] \notag \\
&\ge \mathbb{E}_{p(\zb_1|\zb_2)q(\zb_2|\xb)} \left [ \log \frac{p(\xb|\zb_1)p(\zb_2)}{q(\zb_2|\xb)} \right ] \equiv \mathcal{L}^{> 1} \notag \ .
\end{align}
Here, we have introduced the variational distribution $q(\zb_2|\xb)$ which, naively, is different from any of the available variational distributions $q(\zb_1|\xb)$ and $q(\zb_2|\zb_1)$.
However, it's easy to see that we can simply define $q(\zb_2|\xb)=q(\zb_2|d_1(\xb))$ where $d_1(\xb) = \mathbb{E}\left[q(\zb_1|\xb)\right]$. I.e.\@ we compute the distribution over $\zb_2$ via the mode of $q(\zb_1|\xb)$.
This is possible since we exclusively manipulate the variational proposal distribution without altering the generative model $p(\xb,\zb)$.

In general, the derivation of $\mathcal{L}^{>k}$ for an $L$-layered hierarchical VAE with $\zb=\zb_1,\dots,\zb_L$ is as follows:
\begin{align}
    \log p(\xb) &= \log \int p(\xb|\zb)p(\zb) \text{d}\zb \\
    &= \log \int \frac{q(\zb_{>k}|\xb)}{q(\zb_{>k}|\xb)} p(\xb|\zb)p(\zb) \text{d}\zb \notag \\
    &= \log \int q(\zb_{>k}|\xb) p(\zb) \frac{p(\xb|\zb)}{q(\zb_{>k}|\xb)} \text{d}\zb \notag \\
    &= \log \int q(\zb_{>k}|\xb) p(\zb_{\leq k}|\zb_{>k}) p(\zb_{>k}) \frac{p(\xb|\zb)}{q(\zb_{>k}|\xb)} \text{d}\zb \notag \\
    &= \log \int q(\zb_{>k}|\xb) p(\zb_{\leq k}|\zb_{>k}) \frac{p(\xb|\zb)p(\zb_{>k})}{q(\zb_{>k}|\xb)} \text{d}\zb \notag \\
    &\ge \mathbb{E}_{p(\zb_{\leq k}|\zb_{>k})} \left[ \log q(\zb_{>k}|\xb) \frac{p(\xb|\zb)p(\zb_{>k})}{q(\zb_{>k}|\xb)} \right] \notag \\
    &\ge \mathbb{E}_{p(\zb_{\leq k}|\zb_{>k})q(\zb_{>k}|\xb)} \left[ \log \frac{p(\xb|\zb)p(\zb_{>k})}{q(\zb_{>k}|\xb)} \right] \equiv \mathcal{L}^{>k} \notag \ .
\end{align}
Similar to the $L=2$ case above, we have defined
$$q(\zb_{>k}|\xb) = q(\zb_{>k}|d_k(\xb))$$
with $d_k$ defined recursively as
$$d_k(\xb) = \mathbb{E}\left[q(\zb_k|d_{k-1}(\xb))\right], \qquad d_0(\xb) = \xb\ .$$
That is, we simply consider the inference network below $\zb_{k+1}$ to be a deterministic encoder and forward pass the mode of each preceding variational distribution.

Additionally, we obtain $p(\zb_{\leq k}|\zb_{>k}) p(\zb_{>k})$ by splitting
$$p(\zb)=p(\zb_L)p(\zb_{L-1}|\zb_L)\cdots p(\zb_{1}|\zb_2)$$
at index $k$. Importantly, we then evaluate
$$p(\zb_{>k})=p(\zb_L)p(\zb_{L-1}|\zb_L)\cdots p(\zb_{k+1}|\zb_{k+2})$$
with samples from $q(\zb_{>k}|\xb)$ while
$$p(\zb_{\leq k}|\zb_{>k})=p(\zb_{k}|\zb_{k+1})p(\zb_{k-1}|\zb_{k})\cdots p(\zb_{1}|\zb_{2})$$
is evaluated for $\zb_k$ with $\zb_{k+1} \sim q(\zb_{>k}|\xb)$ and for $\zb_{<k}$ with $\zb_{>k}$ obtained conditionally from itself.

\section{The complementary $\mathcal{L}^{<l}$ bound}
We can generalize the $\mathcal{L}^{>k}$ bound by introducing the flipped version, $\mathcal{L}^{<l}$, which compared to $\mathcal{L}^{>k}$, instead samples the $L-l$ \textit{highest} latent variables in the hierarchy from the prior $\zb_{l},\dots,\zb_L \sim p_\theta(\zb_{\geq l})=p_\theta(\zb_l|\zb_{l+1})\cdots p_\theta(\zb_L)$ and the remaining lower latents from the approximate posterior $\hat{\zb}_{1},\dots,\hat{\zb}_{l-1} \sim q_\phi(\zb_{<l}|\xb)=q_\phi(\zb_{1}|\xb)q_\phi(\zb_{2}|\zb_1)\cdots q_\phi(\zb_{l-1}|\zb_{l-2})$,
\begin{equation}\label{eq:biva-<l}
    \mathcal{L}^{<l} = \mathbb{E}_{p_\theta(\zb_{\geq l}) q_\phi(\zb_{<l}|\xb)} \left[ \log \frac{p_\theta(\xb,\zb)p_\theta(\zb_{<l})}{q_\phi(\zb_{<l}|\xb)} \right].
\end{equation}
Similar to $\mathcal{L}^{>k}$, we recover the regular ELBO for $l=L$. Contrary to $\mathcal{L}^{>k}$, this bound puts as much emphasis on the lowest latent variables as the regular ELBO but keeps track of large deviation from the unconditional prior in the top $L-l$ KL-terms since it is not guided by the approximate posterior for $\zb_{>l}$. We hypothesize that this bound might be useful for OOD detection in cases where the discriminating factor is to be found in low-level statistics rather than high-level features.

Additionally, we can incorporate it in a generalized log likelihood-ratio between $\mathcal{L}^{<l}$ and $\mathcal{L}^{>k}$
\begin{equation}
    LLR^{>k}_{<l} = \mathcal{L}^{<l} - \mathcal{L}^{>k}.
\end{equation}
We hypothesize that this score, or the other possible permutations of it, might be useful for OOD detection but leave further examination to future work.

% While this bound is not expected to improve OOD detection by itself, mainly since the data likelihood $p(\xb|\zb)$ is the same as for the ELBO, we introduce a log likelihood-ratio between $\mathcal{L}^{<l}$ and the regular ELBO, as we did for $\mathcal{L}^{>k}$.
% \begin{equation}
    %     LLR^{<l} = \mathcal{L}^{<l} - \mathcal{L}
    % \end{equation}
    % We hypothesize that this log-likelihood ratio might also be useful for OOD detection based on a more low-level set of semantics learned by the bottom-most latents in the bottom-up hierarchy.

    % \section{Generalized log-likelihood ratio $LLR^{>k}_{<l}$}
    % We can further generalize the log-likelhood ratio $LLR^{>k}$ by replacing the ELBO $\mathcal{L}$ with the bound in \eqref{eq:biva-<l} to obtain
    % \begin{equation}
        %     LLR^{>k}_{<l} = \mathcal{L}^{<l} - \mathcal{L}^{>k}.
% \end{equation}

\section{Note on the KL-term of hierarchical VAEs}
In this research we choose model parameterizations relying on bottom-up inference \cite{burda_importance_2016},
\begin{equation}
    q_\phi(\zb|\xb) = q_\phi(\zb_1|\xb)\textstyle\prod_{i=2}^{L} q_\phi(\zb_i|\zb_{i-1}) \ .
\end{equation}
We do this because bottom-up inference enables the model to learn covariance between the latent variables in the hierarchy.
In the inference model, any latent variable is dependent on the latent variables below it in the hierarchy and, importantly, the top most latent variable is dependent on all other latent variables.

In contrast, a top-down inference model \cite{sonderby_ladder_2016} has a topmost latent variable $\zb_L$ that is independent of the other latent variables and is directly given by $\xb$.
\begin{equation}
    q_\phi(\zb|\xb) = q_\phi(\zb_L|\xb)\textstyle\prod_{i=L-1}^{1} q_\phi(\zb_{i}|\zb_{i+1}) \ .
\end{equation}
This, in essence, makes $\zb_L$ a mean-field approximation without any covariance structure tying it to the other latent variables, $\text{Cov}(z_{L,i}, z_{k,j})=0$ for $k<L$.
Furthermore, since the approximate posterior (and the prior) typically have diagonal covariance, $\zb_L$ is also mean-field within its own elements, $\text{Cov}(z_{L,i}, z_{L,j})=0$ for $i\ne j$.

We hypothesize that the covariance of latent variables towards the top of the hierarchy with other latent variables is important for learning semantic representations.
However, top-down inference models are easier to optimize as has recently been demonstrated \cite{sonderby_ladder_2016, vahdat_nvae_2020, child_very_2021}.

In the following, we inspect the differences between the ELBO used for bottom-up inference and the ELBO used for top-down inference and show that it is not generally possible to decompose the total KL-divergence into separate KL-divergences per latent variable.
Specifically, for top-down inference it is possible to obtain KL-divergence at the top-most latent variable and an expectation of a KL-divergence for the other latent variables.
For bottom-up inference, the resulting terms are no longer KL-divergences except at the top-most latent variable.

We ask the question whether models relying on top-down inference are impeded in their use for semantic OOD detection, or whether they still learn to assign a more semantic representation in the top-most variables simply due to the flexibility of the deterministic neural network layers.
This remains an open research question.

\subsection{Bottom-up inference}
% The ELBO for the bottom-up VAE is given by:
% \begin{align}
    %     \log p(\xb) &\ge \mathbb{E}_{q(\zb_1, \zb_2|\xb)} \left[ \log \left(p(\xb|\zb_1) p(\zb_1|\zb_2) p(\zb_2) \right) \notag \\
    %     &- \log \left( q(\zb_1|\xb) q(\zb_2|\zb_1)\right) \right]\ .
    % \end{align}
    By splitting up the expectation, we can write the ELBO of a two-layer bottom-up hierarchical VAE as
    \begin{align}
        \log p(\xb) &\ge \mathbb{E}_{q(\zb_1, \zb_2|\xb)} \left[ \log p(\xb|\zb_1) \right] \\
    &+ \mathbb{E}_{q(\zb_1,\zb_2|\xb)} \left[ \log p(\zb_1|\zb_2) - \log q(\zb_1|\xb) \right] \notag \\
    &+ \mathbb{E}_{q(\zb_1,\zb_2|\xb)} \left[ \log p(\zb_2) - \log q(\zb_2|\zb_1) \right] \notag \ .
\end{align}
We can write out the expectations in order to derive the KL-divergence terms of the bottom-up ELBO:
\begin{align}
    \log p(\xb) &\ge \int \int \log p(\xb|\zb_1) \text{d}\zb_2\zb_1 \label{eq:2 layer VAE bottom up ELBO integrals} \\
    &+ \int q(\zb_1|\xb) \int q(\zb_2|\zb_1) \log \frac{p(\zb_1|\zb_2)}{q(\zb_1|\xb)} \text{d}\zb_2\zb_1 \notag \\ 
    &+ \int q(\zb_1|\xb) \int q(\zb_2|\zb_1) \log \frac{p(\zb_2)}{q(\zb_2|\zb_1)} \text{d}\zb_2\zb_1 \notag  \ .
\end{align}
From the above, we can see that since the decomposition is in a reverse order, we cannot derive the KL-divergence for the second term. This will hold in general for $L$-layered models for any latent variables  $\zb_1,...,\zb_{L-1}$:
\begin{align}
    \log p(\xb) &\ge \mathbb{E}_{q(\zb_1, \zb_2|\xb)} \left[ \log p(\xb|\zb_1) \right] \\
    &+ \mathbb{E}_{q(\zb_1|\xb)}\left[ \mathbb{E}_{q(\zb_2|\zb_1)}\left[\log \frac{p(\zb_1|\zb_2)}{q(\zb_1|\xb)}\right]\right] \notag \\
    &+ \mathbb{E}_{q(\zb_1|\xb)}\left[-D_\text{KL}[q(\zb_2|\zb_1)||p(\zb_2)]\right] \notag \ .
\end{align}

\subsection{Top-down inference}
% The ELBO for the top-down VAE is given by:
% \begin{align}
    %     \log p(\xb) \ge \mathbb{E}_{q(\zb_1, \zb_2|\xb)} \left[ \log \left(p(\xb|\zb_1) p(\zb_1|\zb_2) p(\zb_2) \right)- \log \left( q(\zb_2|\xb) q(\zb_1|\zb_2)\right) \right]\ .
    % \end{align}
    % By splitting up the expectation we can easily rewrite this ELBO:
    By splitting up the expectation, we can write the ELBO of a two-layer top-down hierarchical VAE as
    \begin{align}
        \log p(\xb) &\ge \mathbb{E}_{q(\zb_1, \zb_2|\xb)} \left[ \log p(\xb|\zb_1) \right] \\ 
        &+ \mathbb{E}_{q(\zb_1,\zb_2|\xb)} \left[ \log p(\zb_2|\xb) - \log q(\zb_2|\xb) \right] \notag \\
        &+ \mathbb{E}_{q(\zb_1,\zb_2|\xb)} \left[ \log p(\zb_1|\zb_2) - \log q(\zb_1|\zb_2) \right] \notag \ .
    \end{align}
    We can write out the expectations in order to derive the KL-divergence terms:
    \begin{align}
        \log p(\xb) &\ge \int \int \log p(\xb|\zb_1) d\zb_1\zb_2 \\
        &+ \int q(\zb_2|\xb) \log \frac{p(\zb_2|\xb)}{q(\zb_2|\xb)} d\zb_2 \notag \\ 
        &+ \int q(\zb_2|\xb) \int q(\zb_1|\zb_2) \log \frac{p(\zb_1|\zb_2)}{q(\zb_1|\zb_2)} d\zb_1\zb_2 \notag \ .
    \end{align}
    The KL-divergence terms can now easily be computed by:
    \begin{align}
        \log p(\xb) &\ge \mathbb{E}_{q(\zb_1, \zb_2|\xb)} \left[ \log p(\xb|\zb_1) \right] \\
    &- D_\text{KL}[q(\zb_2|\xb)||p(\zb_2)] \notag \\ 
    &- \mathbb{E}_{q(\zb_2|\xb)}\left[D_\text{KL}[q(\zb_1|\zb_2)||p(\zb_1|\zb_2)\right] \notag \ .
\end{align}
Note that the KL-divergence in the second layer is not exact since it is dependent on the sample-noise from the layer below. An exact solution can only be derived if the latent variables $\mathbf{z}$ are all conditionally independent. However, this comes at the cost of not learning a covariance structure.

\section{Additional results}
We provide additional results for a model trained on FashionMNIST in \cref{tab:additional-results-fashionmnist}, a model trained on MNIST in \cref{tab:additional-results-mnist}, a model trained on CIFAR10 in \cref{tab:additional-results-cifar} and a model trained on SVHN in \cref{tab:additional-results-svhn}.

We note that while the likelihood is highly unreliable across the datasets, the proposed log likelihood-ratio score is consistent and always allows correct OOD detection with high AUROC$\uparrow$.

\begin{table}[t]
    \centering
    % \resizebox{\columnwidth}{!}{%
    \begin{tabular}{llrrr}
        \toprule
        OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
        \midrule
        \multicolumn{5}{c}{\textbf{Trained on SVHN}} \\
        \midrule
        CIFAR10            &  $L^{>0}$         &  $0.992$  &  $0.993$  &  $0.004$  \\
        CIFAR10            &  $L^{>1}$         &  $0.988$  &  $0.990$  &  $0.002$  \\
        CIFAR10            &  $L^{>2}$         &  $0.746$  &  $0.756$  &  $0.468$  \\
        CIFAR10            &  $LLR^{>1}$       &  $0.939$  &  $0.950$  &  $0.052$  \\
        % \midrule
        % UniformNoise       &  $L^{>0}$         &  $1.000$  &  $1.000$  &  $0.000$  \\
        % UniformNoise       &  $L^{>1}$         &  $1.000$  &  $1.000$  &  $0.000$  \\
        % UniformNoise       &  $L^{>2}$         &  $0.965$  &  $0.957$  &  $0.055$  \\
        % UniformNoise       &  $LLR^{>1}$       &  $0.984$  &  $0.992$  &  $0.000$  \\
        \midrule
        SVHN               &  $L^{>0}$         &  $0.599$  &  $0.587$  &  $0.702$  \\
        SVHN               &  $L^{>1}$         &  $0.555$  &  $0.543$  &  $0.755$  \\
        SVHN               &  $L^{>2}$         &  $0.403$  &  $0.431$  &  $0.869$  \\
        SVHN               &  $LLR^{>1}$       &  $0.489$  &  $0.484$  &  $0.799$  \\
        \bottomrule
    \end{tabular}
    % }
    \caption{Additional results for the HVAE model trained on SVHN. All results computed with 1000 importance samples.}
    \label{tab:additional-results-svhn}
\end{table}
% Results of SVHN model. (d64aa83013304864b974a3fca31e44f1)

\begin{table}[t]
    \centering
    % \resizebox{\columnwidth}{!}{%
    \begin{tabular}{llrrr}
        \toprule
        OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
        \midrule
        \multicolumn{5}{c}{\textbf{Trained on CIFAR10}} \\
        \midrule
        SVHN          &  $L^{>0}$    &  0.083  &  0.318  &  0.974  \\
        SVHN          &  $L^{>1}$    &  0.097  &  0.320  &  0.972  \\
        SVHN          &  $L^{>2}$    &  0.693  &  0.725  &  0.599  \\
        SVHN          &  $LLR^{>2}$  &  0.811  &  0.837  &  0.394  \\
        % \midrule
        % UniformNoise  &  $L^{>0}$    &  1.000  &  1.000  &  0.000  \\
% UniformNoise  &  $L^{>1}$    &  0.987  &  0.984  &  0.025  \\
% UniformNoise  &  $L^{>2}$    &  0.031  &  0.313  &  0.971  \\
% UniformNoise  &  $LLR^{>1}$  &  0.206  &  0.436  &  0.999  \\
\midrule
CIFAR10       &  $L^{>0}$    &  0.485  &  0.488  &  0.817  \\
CIFAR10       &  $L^{>1}$    &  0.467  &  0.476  &  0.822  \\
CIFAR10       &  $L^{>2}$    &  0.411  &  0.433  &  0.869  \\
CIFAR10       &  $LLR^{>1}$  &  0.469  &  0.479  &  0.835  \\
         \bottomrule
    \end{tabular}
    % }
    \caption{Additional results for the HVAE model trained on CIFAR10. All results computed with 1000 importance samples.}
    \label{tab:additional-results-cifar}
\end{table}
% CIFAR10 model: 94f41e80fde444478403496071289a75

\begin{table}[t]
    \centering
    % \resizebox{\columnwidth}{!}{%
    \begin{tabular}{llrrr}
        \toprule
         OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
         \midrule
         \multicolumn{5}{c}{\textbf{Trained on FashionMNIST}} \\
         \midrule
MNIST                    & $\mathcal{L}^{>0}$     &  0.268  &  0.363  &  0.882 \\
MNIST                    & $\mathcal{L}^{>1}$     &  0.593  &  0.591  &  0.658 \\
MNIST                    & $\mathcal{L}^{>2}$     &  0.712  &  0.750  &  0.548 \\
MNIST                    & $LLR^{>1}$             &  0.986  &  0.987  &  0.011 \\
\midrule
notMNIST                 &  $\mathcal{L}^{>0}$    &  0.916  &  0.932  &  0.116 \\
notMNIST                 &  $\mathcal{L}^{>1}$    &  0.983  &  0.986  &  0.000 \\
notMNIST                 &  $\mathcal{L}^{>2}$    &  0.997  &  0.997  &  0.000 \\
notMNIST                 &  $LLR^{>1}$            &  0.998  &  0.998  &  0.000 \\
\midrule
KMNIST                   &  $\mathcal{L}^{>0}$    &  0.690  &  0.694  &  0.554 \\
KMNIST                   &  $\mathcal{L}^{>1}$    &  0.835  &  0.863  &  0.359 \\
KMNIST                   &  $\mathcal{L}^{>2}$    &  0.844  &  0.875  &  0.339 \\
KMNIST                   &  $LLR^{>1}$            &  0.974  &  0.977  &  0.017 \\
\midrule
Omniglot28x28            &  $\mathcal{L}^{>0}$    &  0.898  &  0.837  &  0.166 \\
Omniglot28x28            &  $\mathcal{L}^{>1}$    &  0.991  &  0.989  &  0.011 \\
Omniglot28x28            &  $\mathcal{L}^{>2}$    &  1.000  &  1.000  &  0.000 \\
Omniglot28x28            &  $LLR^{>2}$            &  1.000  &  1.000  &  0.000 \\
\midrule
Omniglot28x28Inverted    &  $\mathcal{L}^{>0}$    &  0.261  &  0.361  &  0.879 \\
Omniglot28x28Inverted    &  $\mathcal{L}^{>1}$    &  0.450  &  0.431  &  0.709 \\
Omniglot28x28Inverted    &  $\mathcal{L}^{>2}$    &  0.557  &  0.574  &  0.678 \\
Omniglot28x28Inverted    &  $LLR^{>1}$            &  0.954  &  0.954  &  0.050 \\
\midrule
SmallNORB28x28           &  $\mathcal{L}^{>0}$    &  0.982  &  0.984  &  0.000 \\
SmallNORB28x28           &  $\mathcal{L}^{>1}$    &  0.998  &  0.998  &  0.000 \\
SmallNORB28x28           &  $\mathcal{L}^{>2}$    &  1.000  &  1.000  &  0.000 \\
SmallNORB28x28           &  $LLR^{>2}$            &  0.999  &  0.999  &  0.002 \\
\midrule
SmallNORB28x28Inverted   &  $\mathcal{L}^{>0}$    &  0.965  &  0.971  &  0.000 \\
SmallNORB28x28Inverted   &  $\mathcal{L}^{>1}$    &  0.997  &  0.992  &  0.000 \\
SmallNORB28x28Inverted   &  $\mathcal{L}^{>2}$    &  0.981  &  0.985  &  0.000 \\
SmallNORB28x28Inverted   &  $LLR^{>2}$            &  0.941  &  0.946  &  0.069 \\
% \midrule
% UniformNoise             &  $\mathcal{L}^{>0}$    &  1.000  &  1.000  &  0.000 \\
% UniformNoise             &  $\mathcal{L}^{>1}$    &  1.000  &  1.000  &  0.000 \\
% UniformNoise             &  $\mathcal{L}^{>2}$    &  1.000  &  1.000  &  0.000 \\
% UniformNoise             &  $LLR^{>2}$            &  0.997  &  0.996  &  0.004 \\
\midrule
FashionMNIST             &  $\mathcal{L}^{>0}$    &  0.476  &  0.484  &  0.816 \\
FashionMNIST             &  $\mathcal{L}^{>1}$    &  0.475  &  0.482  &  0.817 \\
FashionMNIST             &  $\mathcal{L}^{>2}$    &  0.475  &  0.484  &  0.823 \\
FashionMNIST             &  $LLR^{>1}$            &  0.488  &  0.496  &  0.811 \\
         \bottomrule
    \end{tabular}
    % }
    \caption{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
    \label{tab:additional-results-fashionmnist}
\end{table}
% FMNIST model: d65f60927f174ab786a817024a0a4d75.  (alt 0e34596bfe6a4e60a5c0d2bf9440c70d)

\begin{table}[t]
    \centering
    % \resizebox{\columnwidth}{!}{%
    \begin{tabular}{llrrr}
        \toprule
         OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
         \midrule
         \multicolumn{5}{c}{\textbf{Trained on MNIST}} \\
         \midrule
FashionMNIST                     &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
FashionMNIST                     &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
FashionMNIST                     &  $L^{>2}$  &  $0.981$  &  $0.983$  &  $0.003$ \\
FashionMNIST                   &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
\midrule
notMNIST                         &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
notMNIST                         &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
notMNIST                         &  $L^{>2}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
notMNIST                       &  $LLR^{>1}$  &  $1.000$  &  $0.999$  &  $0.000$ \\
\midrule
KMNIST                           &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
KMNIST                           &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
KMNIST                           &  $L^{>2}$  &  $0.987$  &  $0.987$  &  $0.011$ \\
KMNIST                         &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
\midrule
Omniglot28x28                    &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
Omniglot28x28                    &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
Omniglot28x28                    &  $L^{>2}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
Omniglot28x28                  &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
\midrule
Omniglot28x28Inverted            &  $L^{>0}$  &  $0.862$  &  $0.902$  &  $0.205$ \\
Omniglot28x28Inverted            &  $L^{>1}$  &  $0.923$  &  $0.943$  &  $0.056$ \\
Omniglot28x28Inverted            &  $L^{>2}$  &  $0.749$  &  $0.691$  &  $0.411$ \\
Omniglot28x28Inverted          &  $LLR^{>1}$  &  $0.944$  &  $0.953$  &  $0.057$ \\
\midrule
SmallNORB28x28                   &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
SmallNORB28x28                   &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
SmallNORB28x28                   &  $L^{>2}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
SmallNORB28x28                 &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
\midrule
SmallNORB28x28Inverted           &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
SmallNORB28x28Inverted           &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
SmallNORB28x28Inverted           &  $L^{>2}$  &  $0.977$  &  $0.980$  &  $0.001$ \\
SmallNORB28x28Inverted         &  $LLR^{>1}$  &  $0.985$  &  $0.987$  &  $0.000$ \\
% \midrule
% UniformNoise                     &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
% UniformNoise                     &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
% UniformNoise                     &  $L^{>2}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
% UniformNoise                   &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
\midrule
MNIST                            &  $L^{>0}$  &  $0.488$  &  $0.486$  &  $0.807$ \\
MNIST                            &  $L^{>1}$  &  $0.469$  &  $0.469$  &  $0.816$ \\
MNIST                            &  $L^{>2}$  &  $0.514$  &  $0.505$  &  $0.791$ \\
MNIST                          &  $LLR^{>2}$  &  $0.515$  &  $0.507$  &  $0.792$ \\
         \bottomrule
    \end{tabular}
    % }
    \caption{Additional results for the HVAE model trained on MNIST. All results computed with 1000 importance samples.}
    \label{tab:additional-results-mnist}
\end{table}
% Results of MNIST model. (57bed090cf4a4e44b896b4c2cce55500)

\iffalse

\fi

### 2023-03-04 chapters/paper_hierarchical.tex
Add review paper (WIP)

@@
 where $k\in\{0,1,\dots,L\}$ (see Appendix for the derivation).
-We note that $\mathcal{L}^{>0}$ is the regular ELBO (\ref{eq:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
+We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
 The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\zb_1,\dots,\zb_l \sim p_\thetab(\zb_{\leq k}|\zb_{>k})$ and only the $L-k$ highest from the approximate posterior $\zb_{k+1},\dots,\zb_L \sim q_\phib(\zb_{>k}|\xb)$.
@@
 \subsection{A likelihood-ratio score for all feature levels}
-While the $\mathcal{L}^{>k}$ bound provides a score for performing semantic OOD detection, it still relies on the data space likelihood function (see equation \eqref{eq:likelihoods-as-exact} below), which is known to be problematic for OOD detection (section \ref{sec:background-ood-detection}).
+While the $\mathcal{L}^{>k}$ bound provides a score for performing semantic OOD detection, it still relies on the data space likelihood function (see equation \eqref{eq:likelihoods-as-exact} below), which is known to be problematic for OOD detection (\cref{sec:background-ood-detection}).
 To alleviate this, we phrase OOD detection as a likelihood ratio test of being \emph{semantically} in-distribution.
@@
 Subtracting these cancel out the two data likelihood terms $\log p_\thetab(\xb)$ and only the KL-divergences from the approximate to the true posterior remain,
-\begin{align}\label{eq:llr-as-kls}
+\begin{align}
     LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) || p_\thetab(\zb|\xb)\right) \\
                  &\quad + D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb) || p_\thetab(\zb|\xb)\right) \ . \notag
-\end{align}
+\end{align}\label{eq:llr-as-kls}
 
@@
 Now, it can be noted that if we replace the regular ELBO, $\mathcal{L}$, in \eqref{eq:likelihoods-as-exact} with the strictly tighter importance weighted bound \cite{burda_importance_2016},
-\begin{equation}\label{eq:iw-bound}
-    \mathcal{L}_{S} = \mathbb{E}_{q(\zb|\xb)}\left[ \log \frac{1}{N} \sum_{s=1}^{S} \frac{p(\xb, \zb^{(s)})}{q(\zb^{(s)}|\xb)} \right] \ ,
+\begin{equation}
+    \mathcal{L}_{S} = \mathbb{E}_{q(\zb|\xb)}\left[ \log \frac{1}{N} \sum_{s=1}^{S} \frac{p(\xb, \zb^{(s)})}{q(\zb^{(s)}|\xb)} \right] \ , \label{eq:iw-bound}
 \end{equation}
 then, in the limit $S\rightarrow\infty$, we have $\mathcal{L}_{S} \rightarrow \log p_\thetab(\xb)$ and the likelihood ratio reduces to
-\begin{equation}\label{eq:llr-as-kls-iwae-reduced}
-    LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq k}|\zb_{>k}) q(\zb_{>k}|\xb) || p(\zb|\xb))
+\begin{equation}
+    LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq k}|\zb_{>k}) q(\zb_{>k}|\xb) || p(\zb|\xb)) \label{eq:llr-as-kls-iwae-reduced}
 \end{equation}
 which, in practice, is well-approximated for a finite $S$.
-We expect this importance weighted likelihood ratio to monotonically improve upon the one in (\ref{eq:llr-as-kls}) as $S$ increases and the KL-divergence in the regular ELBO that contains terms for which $\zb_i$ is high-dimensional goes to zero.
+We expect this importance weighted likelihood ratio to monotonically improve upon the one in \cref{eq:llr-as-kls} as $S$ increases and the KL-divergence in the regular ELBO that contains terms for which $\zb_i$ is high-dimensional goes to zero.
 
-Since the scores in (\ref{eq:llr-as-kls}) and (\ref{eq:llr-as-kls-iwae-reduced}) are estimated by sampling their estimators are stochastic objects with nonzero variance.
+Since the scores in \cref{eq:llr-as-kls, eq:llr-as-kls-iwae-reduced} are estimated by sampling their estimators are stochastic objects with nonzero variance.
 We note that $\text{Var}(\widehat{LLR}^{>k}) = \text{Var}(\hat{\mathcal{L}}) + \text{Var}(\hat{\mathcal{L}}^{>k}) - 2\, \text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$.
@@
 Clearly, if this is not the case and the difference instead lies in low-level statistics, the scores would likely fail.
-We hypothesize that a complementary bound to (\ref{eq:biva->k}), $\mathcal{L}^{<l}$ described in Appendix E, might be useful in these cases, but leave further examination to future work.
+We hypothesize that a complementary bound to \cref{eq:biva->k}, $\mathcal{L}^{<l}$ described in Appendix E, might be useful in these cases, but leave further examination to future work.
 
@@
 % For CIFAR/SVHN, we also train a BIVA model \cite{maaloe_biva_2019} with $L=10$ and similar configuration as used by the original paper\footnote{Code available at \url{github.com/larsmaaloee/BIVA} and \url{github.com/vlievin/biva-pytorch}}.
-% All models are trained by optimizing the ELBO in (\ref{eq:elbo}). We implement our models in PyTorch \cite{paszke_automatic_2017}\footnote{Code available at \url{github.com/jakobhavtorn/hvae-oodd}}.
 Full model details are in the Appendix.

### 2023-03-04 chapters/paper_review.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\chapter[self-supervised speech representation learning: a review]{Self-Supervised Speech Representation Learning: A Review}
\label{chp:review}

% no-op commands
\newcommand{\edit}[1]{#1}
\newcommand{\ci}[1]{#1}
\newcommand{\red}[1]{#1}
\newcommand{\TNS}[1]{#1}
\newcommand{\kl}[1]{#1}
\newcommand{\abdo}[1]{#1}

\section*{Abstract}
Although supervised deep learning has revolutionized speech and audio processing, it has necessitated the building of specialist models for individual tasks and application scenarios. It is likewise difficult to apply this to dialects and languages for which only limited labeled data is available. Self-supervised representation learning methods promise a single universal model that would benefit a wide variety of tasks and domains. Such methods have shown success in natural language processing and computer vision domains, achieving new levels of performance while reducing the number of labels required for many downstream scenarios. Speech representation learning is experiencing similar progress in three main categories: generative, contrastive, and predictive methods. Other approaches rely on multi-modal data for pre-training, mixing text or visual data streams with speech. Although self-supervised speech representation is still a nascent research area, it is closely related to acoustic word embedding and learning with zero lexical resources, both of which have seen active research for many years. This review presents approaches for self-supervised speech representation learning and their connection to other research areas. Since many current methods focus solely on automatic speech recognition as a downstream task, we review recent efforts on benchmarking learned representations to extend the application beyond speech recognition.

% TODO (JDH): Revise all references in this chapter
\input{chapters/paper_review/01_introduction}
\input{chapters/paper_review/02_three_waves}
\input{chapters/paper_review/03_approaches}  % TODO (JDH): Compilation errors completed until here
% \input{chapters/paper_review/04_benchmarking}
% \input{chapters/paper_review/04b_analysis}
% \input{chapters/paper_review/05_zero}
% \input{chapters/paper_review/06_conclusions}

### 2023-03-04 chapters/paper_review/01_introduction.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\section{Introduction}

%Hung-yi Lee: I think maybe we need a basic frame image. I happen to have an image that can be used here.
\begin{figure}
    \centering
    \includegraphics[width=0.80\textwidth]{paper_review/SSL_framework.png}
	 \caption{Framework for using self-supervised representation learning in
	 downstream applications}
    \label{fig:SSL_framework}
\end{figure}

% {\color{blue} Reviewer: Daniel, KL, SW}\\
% {\color{blue} Abdo -- this section is a work-in-progress}\\

% {\color{red} Hung-yi: Last time we have decided to use "pre-train" (with "-")} -- Done\\

% {\color{red} Hung-yi: Fine-tune? or finetune?} --Fine-tune Done\\

% {\color{red} Hung-yi: Fig. 1 or Figure 1} -- Fig. 1  Done\\

% \abdo{comment from Shinji: bring references to intro.}

% \kl{Comments from Nov. 10 meeting: 
% \begin{itemize}
%     \item \abdo{added.} would be good to define representation learning a bit more thoroughly in the intro, and mention that we are considering both discrete and continuous representations.
%     \item \abdo{added.} it would also be good to mention and broadly define "self-supervised" in the intro (as of now it's first mentioned in Section~\ref{sec:thirdwave}), and mention that the term is now used (and we use it) for some techniques that in the past would have just been called "unsupervised".
% \end{itemize}}

% \kl{New comments: 
% \begin{itemize}
%     \item It would be good to include somewhere in the paper a summary of tasks for which SSL pre-trained models have the SOTA performance.  Maybe it is in the paper already but I don't see it.
% \end{itemize}}

Over the past decade, deep learning approaches have revolutionized speech processing
through a giant leap in performance, enabling various real-world applications.
Supervised learning of deep neural networks has been the cornerstone of this
transformation, offering impressive gains for scenarios rich in labeled
data~\cite{lecun2015deep, hinton2012deep, bourlard2012connectionist}. 
Paradoxically, this heavy reliance on supervised learning has restricted progress in
languages and domains that do not attract the same level of labeling
investment. 
%Paradoxically, the heavy reliance on supervised learning bottlenecked the development of novel industrial speech applications and restricted progress in languages and domains that do not entertain the same level of labeling investment.

To overcome the need for labeled data, researchers have explored approaches that use
unpaired audio-only data to open up new industrial speech use-cases and
low-resource languages~\cite{Kemp1999, csl01_limsi, ma_bbn_06}. Inspired by how
children learn their first language through listening and interacting with
family and surroundings, scientists seek to use raw waveforms and
spectral signals to learn speech representations that capture low-level
acoustic events, lexical knowledge, all the way to syntactic and semantic
information. These learned representations are then used for target downstream
applications requiring a minimal number of labeled data~\cite{Hinton_2007,
LeCun06atutorial, bengio_representation_2013}. 
Formally, representation learning refers to algorithms for extracting latent
features that capture the underlying explanatory factors for the observed
input~\cite{bengio_representation_2013}. 
% Previous research efforts explored latent representations that are binary \cite{}, discrete \cite{} and continuous \cite{}. 
% These latent representations are generally used as inputs to a supervised predictor or a generation module. The whole network generating these representations is frequently used to initialize a supervised classifier~\cite{hinton_2006}.

%Learning representations approaches generally fall within \textit{unsupervised learning}, which refers to the family of machine learning methods that discover naturally occurring patterns in the training samples without any pre-assigned labels or scores associated with these samples~\cite{jordan_2015}. 
Representation learning approaches are generally considered examples of \textit{unsupervised
learning}, which refers to the family of machine learning methods that discover
naturally occurring patterns in training samples for which there are no pre-assigned
labels or scores~\cite{jordan_2015}. 
The term ``unsupervised'' is used to distinguish this family of methods from
``supervised'' approaches, which assign a label to each training sample, and
``semi-supervised'' approaches, which utilize a small number of training samples
with labels to guide learning using a larger volume of unlabeled samples.
Examples of unsupervised learning techniques include k-means clustering~\cite{vq}, mixture models~\cite{MoE}, autoencoders~\cite{hinton_94},
and non-negative matrix factorization~\cite{nmf}. 
\textit{Self-supervised learning} (SSL) is a fast-growing subcategory of
unsupervised learning approaches, which are techniques that utilize
information extracted from the input data itself as the label to learn
representations useful for 
% one or many 
downstream tasks. {\color{black} For example, unsupervised k-means clustering doesn't adhere to this definition of self-supervision since it iteratively minimizes the within-cluster variance during learning.}
In this review, we
focus on self-supervised learning approaches.

\Cref{fig:SSL_framework} outlines self-supervised representation learning in
relation to downstream applications. 
There are two stages in this framework.
In the first stage, we use SSL to pre-train a \textit{representation model},
also called an \textit{upstream model} or a \textit{foundation model}.
In the second stage, downstream tasks use either the learned
representation from the frozen model, or fine-tune the entire pre-trained model
in a supervised phase~\cite{hinton_2006}. 
%e.g., Automatic Speech Recognition (ASR) and Speaker Identification (SID) are two example downstream applications in \cref{fig:SSL_framework}. 
Automatic speech recognition (ASR) and speaker identification (SID) are 
examples of downstream applications in \cref{fig:SSL_framework}.
% There are some task-specific labeled data for each downstream task, and there is a downstream model on top of the upstream model. The labeled data is used to train the corresponding downstream models and optionally fine-tune upstream models.
% If the pre-training stage only leverages unlabeled data, the framework in \cref{fig:SSL_framework} is named self-supervised learning (SSL), which is the focus of this review paper.
% Pre-training with labeled data is out of the scope. Before the term self-supervised learning was widely used, pre-training without any labeled data was usually known as an \texit{unsupervised learning} approach.

%Lee: in the introduction, we may need a specific example. Below is just an example.
%With unsupervised ASR, imagine that in the future, when an indigenous language family buys an intelligent assistant, even if the assistant does not support the language at the beginning, it would automatically learn to transcribe the new language after learning.

% The speech representation learning approaches evolved over the past few decades through at least three major stages, as discussed in \cref{sec:thirdwave}; static and temporal clustering algorithms \cite{}, energy-based neural models \cite{}, and more recently, self-supervised learning techniques \cite{}. 
% Across all these three waves of representation learning, three main desirable characteristics emerged; features need to be disentangled, invariant, and hierarchical. 
It is considered desirable for learned speech representations to be
disentangled, invariant, and hierarchical.
Since spoken utterances contain much richer information than the corresponding text
transcriptions---e.g., speaker identity, style, emotion, surrounding noise, and
communication channel noise---it is important to learn representations that
disentangle these factors of variation. Furthermore, invariance of the learned
features to changes in background noise and in the communication channel ensures
stability with respect to downstream application scenarios. Learning feature
hierarchies at the acoustic, lexical, and semantic levels supports applications
with different requirements. For instance, whereas a speaker identification task
benefits from a low-level acoustic representation, a speech
translation task requires a more semantic representation of the input
utterance.

Due to the popularity of SSL, reviews have been published about the
technology in general~\cite{bommasani2021opportunities,ericsson2021selfsupervised,LiuSSLsurvey} as well as its application to natural language processing (NLP)~\cite{rogers-etal-2020-primer,liu2021pretrain,xia-etal-2020-bert,QiuSSLNLPsurvey} and computer vision (CV)~\cite{JingSSLCVsurvey}. Recently, a brief overview with a general focus on speech representation learning was published \cite{borgholt_22}. However, none of these overviews focus exclusively on SSL for speech processing. Since the speech signal differs greatly from image and text inputs, many  theories  and technologies have been developed to address the unique challenges of speech. One review addresses speech representation learning based on deep learning models~\cite{latif2021deep}, but does not address recent developments in self-supervised learning. This motivates this overview of speech SSL.

%For survey papers on TL to speech processing, please check the reference~\cite{survey_speech_TL}.
%The representation learning methods covered in this review are primarily task-agnostic methods that only benefit from supervised data at the fine-tuning stage, as shown in \cref{}. %Lee: Similar sentence may have been mentioned in the paragraph describing the figure.

The structure of this paper is arranged as follows. \Cref{sec:thirdwave}
briefly reviews the history of speech representation learning, and
\cref{sec:approach} reviews current speech SSL models.
\Cref{section:benchmark} surveys SSL datasets and benchmarks, 
and discusses and compares results from different works. \Cref{analysis}
analyzes successful SSL approaches and offers insights into the
importance of technological innovations. \Cref{sec:zero} reviews 
zero-resource downstream tasks that utilize SSL. 
Finally, \cref{sec:conclusion} summarizes the paper and suggests 
future research directions.

### 2023-03-04 chapters/paper_review/02_three_waves.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\section{Historical Context of Representation Learning}

\label{sec:thirdwave}

% {\color{blue} Reviewer: KL, KK, Daniel }\\
In this section we present the historical background of the current surge in
self-supervised representation learning methods in the context of two previous
waves of research work in the 1990s and 2000s. The discussed approaches go
beyond speech to describe the overall landscape of machine learning development during the
past few decades.

%%KK: if you adopt the paragraph below I suggest renaming this as "Features based on generative models" or something
\subsection{Clustering and mixture models}
% {\color{blue} Tara}\\

%\kl{to people who are familiar with modern rep learning work, this first wave may not seem like rep learning at all.  readers might also wonder why we are talking about what look like standard supervised ASR in the 3rd paragraph below.  one option is to describe it is as unsupervised learning of latent variables (Gaussian components, HMM states) within a supervised model. (something like that -- I haven't thought about the wording carefully)}
%KK: suggesting the paragraph below for inclusion here
%The initial wave of speech representation {\em learning} focused on extracting speech features from generative models. The main objective was
%to render  the knowledgpe learned by generative models accessible to
%discriminative downstream classifiers, or to map variable-length sequences into  fixed-length representations.
%The generative model, typically a global Gaussian mixture model (GMM) or a Hidden Markov Model (HMM) was trained by
%by maximizing the likelihood of the data given the
%model, which could be done in either an unsupervised or a supervised way. Feature vectors were then derived from the parameters of the trained model. In
%the case of {\em Fisher vectors}, the features were the normalized gradients of the log-likelihood
%with respect to the model parameters (mixture weights, means, variances) of the Gaussian mixtures.
%An extension of this approach (likelihood ratio score space) used the derivative of the log likelihood ratio of	two models,
%e.g., a background model and a foreground model.
%Examples of their use in speech processing include speech recognition \cite{smith01,venkata03} and speaker recognition \cite{wan03}.
%Subsequent techniques in speaker and language verification \cite{dehak11a,dehak11b} similarly extracted parameters (concatenated means) from trained background GMMs as representations that were then %combined with low-rank projections of speaker/session or language specific vectors.

Initial research in learning latent speech and audio representations involved
simple models in which the training data likelihood was optimized directly
or via the expectation--maximization (EM) algorithm.

Early work used simple clustering methods. For example, in work such
as \cite{Rabiner1979,Wilpon1985}, word patterns were clustered
semi-automatically using techniques such as k-means, after which isolated words
were recognized by finding the training cluster closest to 
  the test data.   % AMH: check

Through time, modeling techniques improved such that subword units were
represented by Gaussian mixture models (GMMs) \cite{Gauvain1994}, which facilitated
the modeling of more variability in the input data. GMMs were first built for
context-independent phonemes; state-clustering 
algorithms~\cite{Young1994} then resulted in GMMs for context-dependent phonemes. 
Each latent component of these mixture models acted as a template of a
prototypical speech frame, 
% leading to their deficiency to deal with 
  making it difficult to handle             % AMH: check
large volumes of data with diverse characteristics. 
Furthermore, dynamical models like hidden Markov models (HMMs)~\cite{Bahl1986}
allowed for the processing of continuous speech rather than just isolated word
recognition. These generative GMM and HMM models were trained by maximizing the
likelihood of data given the model, which could be accomplished in either an
unsupervised or a supervised manner.

Another line of research focused on extracting speech features from generative
models. The main objective here was to render the knowledge learned by generative
models accessible to discriminative downstream classifiers, or to map
variable-length sequences to fixed-length representations. Feature vectors
were derived from the parameters of trained GMM models. In the case of {\em
Fisher vectors}, the features were the normalized gradients of the
log-likelihood with respect to the model parameters (mixture weights, means, and
variances) of the Gaussian mixtures. An extension of this approach (likelihood
ratio score space) used the derivative of the log-likelihood ratio of two
models, e.g., a background model and a foreground model. Examples of their use
in speech processing include speech recognition~\cite{smith01,venkata03} and
speaker recognition~\cite{wan03}. Subsequent techniques in speaker and language
verification~\cite{dehak11a,dehak11b} similarly extracted parameters
(concatenated means) from trained background GMMs as representations that were
then combined with low-rank projections of speaker/session- or language-specific
vectors.

\subsection{Stacked neural models}
\label{subsec:stack}

%{\color{blue} Tara}\\
More recently, representation learning has seen a shift of focus towards neural models,
which, compared to GMMs and HMMs, offer distributed representations with more
capacity to model diverse input signals into efficient latent binary codes. 
Examples of early techniques include restricted Boltzmann machines
(RBM)~\cite{hinton_2006}, denoising autoencoders~\cite{DAE}, noise contrastive
estimation (NCE)~\cite{gutmann2012noise}, sparse coding~\cite{Olshausen1996,
sparse_lee, sivaram2010sparse}, and energy-based methods~\cite{Ranzato2007}.
Many of these techniques have also been applied to CV and NLP problems, which
provided inspiration for their application to speech.
%Examples of techniques include Restricted Boltzmann Machine (RBM) \cite{hinton2012practical,Srivastava2013}, Neural Auto Encoders \cite{Lecun2012}, Noise Contrastive Estimation (NCE) \cite{chen2015recurrent},  Sparse Coding \cite{sivaram2010sparse} and Energy-based method \cite{Ranzato2007}. Note that many of these techniques were also tried for CV and NLP problems, which provided inspiration to try them for ASR as well.

Higher-capacity neural models were achieved by stacking several neural network
layers to build progressively higher-level concept representations.
However, these deeper networks also increased the training complexities. For
example, approximate training methods such as contrastive divergence
\cite{hinton_cd_2002} were a practical technique to 
% make       RBM training more efficient. 
  streamline RBM training.                   % AMH: check
Furthermore, deep networks had non-convex objective functions, which
often resulted in long training times compared to GMMs, which are trained
using full batches instead of mini-batch learning.

% http://proceedings.mlr.press/v2/ranzato07a/ranzato07a.pdf, 
%https://auai.org/uai2013/prints/papers/166.pdf, % https://www.cs.cmu.edu/~rsalakhu/papers/neco_DBM.pdf
% problems with this trend: approximate gradient, failure modes, long training time, no clear wins in downstream applications.

\subsection{Learning through pretext task optimization}

% {\color{blue} Hung-yi + Abdo}\\

%Note on 12/09: We will mainly describe the speech part but link to CV and NLP. 
%discuss instance classification, contrastive losses,\\

%Currently, the network learns a function that maps input to desired representations by solving a \textit{pretext} task.
%After pre-training, learned representation models could be applied to downstream tasks through feature-based representation extraction or fine-tuning as part of the downstream model.  
%By optimizing carefully designed pretext tasks, neural models can capture latent representations which generalize well across a wide range of downstream applications, usually under few-shot or even zero-shot conditions. 
A more recent trend is learning networks that map the input to desired
representations by solving a \textit{pretext} task. Such studies have several
characteristics:
(1)~All layers are trained end-to-end to optimize a single pretext task instead
of relying on layer-wise pre-training
(2)~Past stacked networks typically had only a few layers, but 
very deep networks with more than ten layers are now common.
(3)~It is common to evaluate a representation model on a wide range of tasks.
For example, in NLP, a representation model is usually assessed on GLUE,
which comprises nine tasks~\cite{GLUE}, whereas in speech, a representation model can be
evaluated on SUPERB, which comprises ten tasks~\cite{yang21c_interspeech}, 
as described in detail in \cref{sec:benchmark}.

The cornerstone of this third wave is the design of a pretext task, which
allows the model to efficiently leverage knowledge from unlabeled data.
The pretext task should be challenging enough for the model to learn high-level
abstract representations and 
% not be too amiable to exploit low-level shortcuts.
  not be so easy as to encourage the exploitation of low-level shortcuts.  % AMH: check
Early breakthroughs included end-to-end learning of deep neural architectures
via pretext tasks for restoring the true color of black-and-white
images~\cite{colorizing}, joint learning of latent representations and their
cluster assignments~\cite{deepcluster}, and the prediction of the relative positions of
image patches~\cite{context_pred}. Other popular approaches include variational
autoencoders (VAEs)~\cite{vae, rezende2014stochastic}. While typical autoencoders learn data
representations using unsupervised objectives by reconstructing the input
after passing it through an information bottleneck, VAEs estimate a neural model of a probability density function (pdf) that approximates the unknown “true” distribution of the observed data, for which we only have access to independently identically distributed (iid) samples. It is also important to mention dynamical VAEs \cite{Girin2021}, which is an extension of VAE for sequential data such as speech.

In the SSL context, a pretext task related to autoencoding is to generate
an object from its partial information. Such tasks are widely used in NLP, for
example, using 
the previous tokens 
in a sentence to predict the next token such as in
ELMo~\cite{peters2018deep}, the GPT
series~\cite{alex2019GPT2}, and %brown2020gpt3 alex2018GPT,
Megatron~\cite{shoeybi2020megatronlm}, or predicting the masked tokens in a
sentence such as with the bidirectional encoder representations from Transformers
(BERT) series~\cite{jacob2019BERT,Liu2019RoBERTa}. %,albert
%ELMo~\cite{peters2018deep}, GPT series~\cite{alex2018GPT,alex2019GPT2,brown2020gpt3}, Megatron~\cite{shoeybi2020megatronlm} are gigantic LM which predict the next token. 
%BERT~\cite{jacob2019BERT}, RoBERTa~\cite{Liu2019RoBERTa}, ALBERT~\cite{albert} masked the input token, and the models are learned to predict the masked tokens. 
Another common pretext task in the third wave is contrastive
learning~\cite{oord2018representation}, in which a model learns to identify a
target instance from a set of negative samples. 
This approach has become especially popular in the
CV context~\cite{pmlr-v119-chen20j,he2020momentum,chen2020improved,SwAV}.
In this survey, we will mainly focus on techniques for pretext task
optimization for speech processing, and discuss these techniques in detail
in \cref{sec:approach}.

\subsection{Other related work}
A closely related area of research that is not covered in this review is
semi-supervised pre-training methods such as pseudo-labeling (that is,
self-training). Pseudo-labeling (PL) relies on a supervised teacher model to
label a large volume of speech-only data, which is then used to augment the
initial labeled data to train a student model~\cite{Kemp1999, csl01_limsi,
ma_bbn_06, hari_1mhour}. PL has been successful and widely adopted in the
speech community since the 1990s. Other proposed variations of PL include
augmenting speech-only data with noise to improve robustness, iterating
over the PL process to improve teacher labeling quality, and training student
models with more parameters than their original teachers to capture the
complexities in vastly larger speech-only data~\cite{park2020improved,
xu2020iterative, xiao_scaling_2021}. 
Both SSL and PL leverage unlabeled speech-only data.
One distinguishing factor in PL is the utilization of supervised data for a
specific task during model pre-training, which limits the model's focus to
a single (or at best a few) downstream tasks. 
SSL, in turn, is an attempt to learn task-agnostic representations to benefit
a wide range of tasks.

Transfer learning (TL) is another closely related area of research for
pre-training speech models. TL transfers knowledge captured by models
trained on one task to different but related tasks~\cite{caruana1997multitask}. 
The past few decades have seen active research on TL and its
extension to multitask learning for more general representations. 
Multilingual and cross-lingual supervised models have proven superior in
low-resource speech recognition tasks~\cite{Cui2015MultilingualRF}.
SSL can be regarded as a type of TL because knowledge learned from pre-training
is used for different downstream tasks.
This survey paper focuses on SSL, and not all TL technologies for speech. 
One survey indeed addresses TL for speech processing~\cite{survey_speech_TL}
but does not include current SSL technologies for speech.

%Then contrastive loss is applied~\cite{oord2018representation}, in which the model learns to classify the positive instances from a set of unrelated negative samples. SimCLR~\cite{pmlr-v119-chen20j}, MoCo~\cite{he2020momentum,chen2020improved} and SwAV~\cite{SwAV} are representative ones. 
%These approaches learn representations from visual inputs by maximizing agreement between differently augmented views of the same sample while minimizing the agreement between different samples. 
%Then there is a series of work trying to remove the requirement of negative examples. 
%BYOL~\cite{BYOL} trains a student network to predict the representations of the teacher network.  The parameter of the teacher network is the exponential moving average of the previous student network. In such a scenario, negative examples are not required. Then SimSiam~\cite{Chen_2021_CVPR} improved BYOL by removing the moving average and found that good representation can still be learned in this way.
%In Barlow twins~\cite{BarlowTwins}, the cross-correlation matrix between the representations of two identical encoders fed with distorted versions of a sample is measured. The encoder learns to make the cross-correlation matrix as close to the identity matrix as possible.
%VICReg~\cite{bardes2021vicreg}

%The 3rd wave begins from NLP and CV communities. %Is it true?
%Below the progress of self-supervised learning in CV  and NLP are briefly summarized. %(However, we have to consider how to connect this part with other parts.)

%In CV, some earlier approaches make some pretexts to train instance classifiers, and the last hidden layers of the classifiers are used as the representations in the downstream tasks. Below are some representative approaches:
%\begin{itemize}
%\item Position~\cite{Doersch_2015_ICCV}: predicting the relative position between two random patches from one image. A model needs to understand the spatial context of objects to tell the relative position between parts.
%\item Jigsaw~\cite{noroozi2016unsupervised}: The model is trained to place shuffled patches back to the original locations.
%\item Rotation~\cite{gidaris2018unsupervised}: The model is trained to predict which rotation has been applied.
%\end{itemize}

%How to deal with negative examples is critical in this series. SimCLR~\cite{pmlr-v119-chen20j} uses a large batch size to obtain a lot of negative examples, while MoCo uses a momentum encoder to make the negative samples’ representation consistent.

%In ELECTRA~\cite{clark2020electra}, given a sentence with some tokens replaced, the ELECTRA model predicts whether a token is generated or not. 
%XLNet~\cite{XLNet}
%MASS~\cite{MASS}, BART~\cite{lewis-etal-2020-bart}, and T5~\cite{T5} are seq2seq models. The input is disturbed by different approaches, and the seq2seq model learns to reconstruct the disturbed input. %(There are speech T5, should we mention this work?)
%The contrastive approaches have been applied at the sentence level in NLP~\cite{gao-etal-2021-simcse,kim-etal-2021-self,yan-etal-2021-consert,giorgi-etal-2021-declutr}.

%Turing NLG only has webpage: https://www.microsoft.com/en-us/research/blog/turing-nlg-a-17-billion-parameter-language-model-by-microsoft/
%I don't know how to cite it.

%ref: http://speech.ee.ntu.edu.tw/~tlkagk/courses/DLHLP20/BERT%20train%20(v8).pdf

### 2023-03-04 chapters/paper_review/03_approaches.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\section{Speech Representation Learning Paradigms} \label{sec:approach}
% {\color{blue} Reviewer: Hung-yi, Daniel , Abdo}\\
% {\color{blue} Hung-yi + Abdo}\\
%Start by discussing central challenges for self-supervised representation learning in speech: 1)Can't do instance classification in variable-length sequences. 2)No pre-defined vocabulary. 3)No pre-defined segmentation. \\
%New approaches offer deep contextualization of learned representations. \\
% {\color{red} first version for a list of symbols to use -- let's keep updating it as needed -- Abdo}\\
Due to the characteristics of speech, SSL pretext tasks developed for CV
and NLP may not directly apply to speech.
Below we summarize the characteristics of speech as compared to CV and NLP.
\begin{itemize}
\item \textit{Speech is a sequence.} 
Unlike CV, in which an image usually has a fixed size representation, it is
natural to represent a speech utterance as a variable-length sequence. 
Therefore, pretext tasks developed for CV cannot generally be directly
applied to speech.

\item \textit{Speech is a long sequence without segment boundaries.} 
Both text and speech can be represented as sequences. From this viewpoint, it
is natural to apply learning approaches developed for text directly to
speech. 
% However, the sequence representations of speech are usually much longer than those of text and they contain no natural boundaries at which they can be split into segments.
In NLP, morpheme-like tokens are widely used as sequence units in
pre-training. The standard BERT takes 512 morpheme-like tokens as
input, usually covering a paragraph including several sentences. 
% In speech processing, the audio signal is recorded as a waveform represents the amplitude of pressure. 
However, speech signals consist of sound pressure measurements with thousands
of samples per second, resulting in sequences much longer than those for text. Even
spectral representations which reduce the sequence length can have hundreds of
frames per second.
%This raw signal is often converted to a spectral representation where each frame represents 20-25 milliseconds of audio computed with overlaps at 10 millisecond strides.
% a frame is the typical sequence unit in speech, and each frame usually represents a length of 10 milliseconds. 
% Therefore, an utterance is usually represented as a long sequence with more than a thousand frames. 
Processing such sequences with typical neural network architectures like
Transformers can result in problems with running time and memory requirements. 
One could gather consecutive frames to form shorter segments,
but unlike text, there is no obvious segmentation for unlabeled
speech.
\item \textit{Speech is continuous.} 
In NLP, it is common to use a pretext task that models a categorical
distribution of masked or future inputs. Since text is easily broken down into
individual tokens such as words, subwords, or characters, it is
straightforward to define a finite vocabulary for such tasks.
%The generative-based pretexts are widely applied in NLP.
%Because a sentence is composed of tokens, sentence generation can be formulated as a series of classification tasks.  
However, this idea does not apply to speech modeling because speech
signals are continuous; 
% thus, such a vocabulary is unavailable.
  in this sense there is no such thing as a speech vocabulary.   % AMH: check
\item \textit{Speech processing tasks are diverse.}
Building generalizable self-supervised representation models for diverse speech
processing tasks is challenging.
Speech contains rich, hierarchical information, and different speech tasks
may require mutually orthogonal information.
For example, speech recognition requires a model that extracts content information
but ignores speaker information; in contrast, speaker recognition 
requires a model that extracts speaker information but removes content information.
Therefore, it is challenging to define a self-supervised model whose
representations are suitable for both speech recognition and speaker
recognition. Analogous considerations apply within CV and NLP.
\end{itemize}
%(Although we list the problems here, we do not highlight the solutions in the following paragraph.)

In the sections below, we group modern SSL pretext tasks designed for speech
into three main categories: \textit{generative} approaches,
\textit{contrastive} approaches and \textit{predictive} approaches. 
%For each category, we provide a brief motivation and then describe prominent models that belong to it. Lastly, we discuss a set of challenges related to the category.
\Cref{fig:timeline} shows a timeline of the models covered in these sections
with each model colored according to our categorization. 
\Cref{table:pretext} summarizes model pretext tasks along within the categories.

%\input{tables/model}

% \begin{center}
% \begin{tabular}{ |c|c| } 
%  \hline
%  $s$ & a single waveform sample \\
%  $S$ & a sequence of waveform samples \\
%  $x$ & a single feature vector (MFCC or learned)\\
%  $X$ & a sequence of feature vectors \\
%  $\widetilde{X}$ & a data augmented sequence of feature vectors \\
%  $z$ & a single latent vector \\
%  $Z$ & a sequence of latent vectors \\
%  $h$ & a single hidden vector representation \\
%  $H$ & a sequence of hidden vector representation \\
%  $m$ & a binary mask for a single time step \\
%  $M$ & a binary mask for the whole sequence \\
%  $y$ & a single supervised label \\
%  $Y$ & a sequence of supervised labels \\
%  $\tau$ & temperature variable \\ 
%  $f()$ & an encoding network - use $f_1()$ $f_2()$ if multiple\\
%  $g()$ & a decoding network \\
%  $P()$ & Probability \\
%  $E()$ & Expectation \\
%  $L()$ & Loss \\
%  $\mathcal{L}$ & Likelihood \\
%  \hline
% \end{tabular}
% \end{center}

\subsection{Notation}
\label{sec:notation}

To efficiently describe the different approaches, we use a simple
notation. Models are assumed to consist of functions $f(\cdot)$ and $g(\cdot)$, where $f(\cdot)$  denotes the representation model to be used after pre-training and $g(\cdot)$ is an auxiliary module needed only to support the pretext task. For instance, in a classic autoencoder, $f(\cdot)$ would denote the encoder and $g(\cdot)$ the decoder. For more complex models, these functions might consist of several components indicated by sub-indices $f_1(\cdot) \dots f_N(\cdot)$. As we will see, many self-supervised models use masking, which replaces some parts of the input or a hidden representation  by zeros or a learned vector. We use $m(\cdot)$ to denote a function that applies such masking to its input. Similar to $g(\cdot)$, this function is only used during pre-training.

Given an acoustic input $X =\{x_1,x_2, ..., x_T\}$, $f(\cdot)$ outputs a representation $H =\{h_1,h_2,...,h_T\}$. The input~$X$ may be either the raw waveform samples or a sequence of spectral feature vectors. Both are viable options in practice. For simplicity, we do not distinguish between the two in our notation.

While $f(\cdot)$ always takes an acoustic input, the input to $g(\cdot)$ can be either the acoustic signal or another learned representation. Most importantly, $g(\cdot)$ produces an output that is used for the pretext task but is not  used by $f(\cdot)$ to produce the representation $H$. Hence, $g(\cdot)$ can be discarded after pre-training. Finally, $f(\cdot)$ commonly downsamples the temporal dimension, but again, this is not crucial to understand the models, so consider only a single temporal scale $t\in\{1,\dots, T\}$ for notational convenience.

%A few other conventions should be noted here. 
We use $Q = \{q_1,q_2, ..., q_T\}$ to denote representations that are quantized via codebook learning. Alternatively, discrete representations may take the form of one-hot vectors, or the equivalent integer IDs, which we denote by $C = \{c_1,c_2, ..., c_T\}$. We use a circumflex to denote that, for instance, $\hat{x}_t$ is an approximation of $x_t$. Finally, we often use a subscript when defining a loss, $\mathcal{L}_i$, to imply that the total loss is computed as a sum over $i$, unless otherwise stated.

For some models, we will refer to $H$ as a \emph{contextualized} representation which means that each $h_t$ is a function of some, linguistically speaking, long sub-sequence of $X$ spanning at least several phonemes. Usually, $h_t$ depends on the entire input $X$ or all previous timesteps $X_{[1,t]}$. In contrast, a \emph{localized} representation is one that only depends on a short part of the input $X_{[t - u,t + u]}$, where $u \geq 0$. The distinction between contextualized and localized may become fuzzy if $u$ is large, however, this is rarely the case.

\edit{After pre-training,} the representation model $f(\cdot)$ can be fine-tuned for a downstream task directly or used to extract features which are fed to another model, as visualized in \cref{fig:SSL_framework}. It is not uncommon to use the output representation~$H$, but often representations from hidden layers of $f(\cdot)$ are better suited \cite{pasad2021}.

\begin{figure*}[t!]
    \centering
    \includegraphics[width=0.98\textwidth]{paper_review/model_timeline.pdf}
   % \includegraphics[width=0.98\textwidth]{paper_review/model_timeline_wo_dccae.pdf}
	 \caption{A selection of models listed according to first publication date
	 on arXiv or conference submission date when this clearly precedes the
	 former. The models are categorized as generative, contrastive, or predictive.
	 In addition, some models are characterized as embedding models or
	 multi-modal models, although most learn frame-level
	 representations from speech only. Some models use a mixture of generative
	 and contrastive tasks. For instance, PASE and PASE+ use a multi-task setup,
	 but find that generative tasks are the most important for downstream
	 task performance~\cite{pascual2019learning}.}
    \label{fig:timeline}
\end{figure*}

\subsection{Generative approaches}
\label{sec:generative}

% {\color{blue} Hung-yi}\\

\subsubsection{Motivation}

%In this category, the pretext task is to generate speech signals given quantized, past, or unmasked time steps. To solve the generative-based pretext, the model leverages an encoder $f(\cdot)$ and a decoder $g(\cdot)$. Given waveform samples $S$ or its corresponding acoustic feature vectors $X$, the encoder $f(\cdot)$ outputs hidden representations $H$. The waveform samples $S$ or acoustic feature vectors $X$ are usually corrupted before entering the encoder $f(\cdot)$. In this subsection, we only consider the acoustic features $X$ and ignore waveform samples $S$ for simplicity, but taking waveform samples as input and generating waveform samples is possible. Some regularization in the encoder $f(\cdot)$ or decoder $g(\cdot)$ is required to encourage the model to utilize global speech information and prevent naively copying input when reconstructing, especially when the input of $f(\cdot)$ is not corrupted. After pre-training, $f(\cdot)$ is used as the representation model, which takes a sequence of acoustic feature vectors $X=\{x_1,x_2, ..., x_T\}$ as input, and outputs hidden contextual representations $H =\{h_1,h_2,...,h_T\}$ for downstream tasks.

%In this category, the pretext task is to generate, or reconstruct, the input data based on some limited view. This includes predicting future inputs from past, masked from unmasked, or the original from some other corrupted view. Models that solve a generative pretext task typically consist of an encoder $f(\cdot)$ and a decoder $g(\cdot)$. Given an acoustic input $X =\{x_1,x_2, ..., x_T\}$, the encoder $f(\cdot)$ outputs a representation $H =\{h_1,h_2,...,h_T\}$. The decoder $g(\cdot)$ takes $H$ as input to generate parts of the original acoustic input. The input $X$ may be either the raw waveform samples or a sequence of spectral feature vectors. In the following, we will not distinguish between the two, but in practice, both are viable options. After pre-training, $f(\cdot)$ is used as the representation model. The model can be fine-tuned for a downstream task directly or used to extract features which are fed to another model, as visualized in \cref{fig:SSL_framework}. It is common to use the output representations $H$, but often representations from hidden layers of $f(\cdot)$ are more suited.

%All the generative approaches are summarized in \cref{table:generative}. 
%There is little work generating waveform samples $S$ because generating them is computationally expensive.
In this category, the pretext task is to generate, or reconstruct, the input data based on some limited view. This includes predicting future inputs from past inputs, masked from unmasked, or the original from some other corrupted view.  ``Generative'' as used in this paper hence refers to models that target the original input in their pretext task. Note that this differs from generative models, which learn \ci{distributions that allow} to sample new data.
% It is important to note that many of the approaches discussed in this subsection are ''reconstructive``, i.e., they focus on reconstructing the input signal rather than learning a probabilistic model for sampling novel sequences\footnote{Generative self-supervised models are not the same as generative probabilistic models that permit sampling new data.}.

%\footnote{The lengths of $X$ and $H$ are not necessary to be equal. Usually, $H$ is shorter than $X$. For simplicity of discussion, we assume the lengths of $X$ and $H$ are the same in this subsection.} % Any one care about this?

\subsubsection{Approaches}

%\subsubsection{Autoregressive reconstruction}
%\subsubsection{Autoregressive Prediction}
%\label{sssec:autoregressive}

\paragraph{Autoencoding} 
Since their introduction in the mid-1990s~\cite{hinton_94}, autoencoders (AEs)
have played an essential role in learning distributed latent representations of
sensory data. 
As described above, AEs consist of an encoder and decoder; the pretext task
is to reconstruct the given input. The most common type of AE places an
information bottleneck on the latent representation by simply having fewer
hidden units available than input features. This forces the model to discard
low-level details and disccourages the learning of trivial solutions. Other models add
regularization to the latent space to further improve the quality of the
learned representations.
% An AE converts an input utterance $X$ to a sequence of representations $H$ and then uses a decoder network to convert $H$ into an approximate reconstruction of the input $X$. Regularization is required to learn general and efficient latent representations and to avoid learning a trivial identity mapping. 
For instance, denoising autoencoders (DAEs) learn latent representations by
reconstructing from input corrupted by noise~\cite{DAE}. 
\edit{The Variational Autoencoder (VAE) is a probabilistic version of the AE which defines the latent representation via a posterior distribution over stochastic latent variables \cite{vae,rezende2014stochastic}. VAEs have been applied to speech in numerous works \cite{chung_recurrent_2015, fraccaro_sequential_2016, hsu2017learning, hsu2017unsupervised, aksan_stcn_2019}}.
The vector-quantized variational autoencoder (VQ-VAE) is another model in this
category~\cite{vqvae};
% The first method to sucessfully learn meaningful discrete latent representations was the Vector Quantised Variational Autoencoder (VQ-VAE)~\cite{vqvae}. 
it extends the original VAE~\cite{vae} with a novel
parameterization of the posterior distribution for discrete latent
representations. 
\edit{The VQ-VAE has been instrumental in generative speech modelling and recent work on generative spoken language modeling has successfully combined the idea of a discrete latent space with self-supervised learning \cite{SpeechResynthesis_IS21, pGSLM, GSLM}.}
% Forcing the use of discrete latent representations encourages the model to focus on salient input features, e.g., phonemes in speech, rather than using the capacity to model channel noise and imperceptible local input details.

Specifically, in the VQ-VAE, the continuous representation vector $h_t$ at the output of the encoder is quantized by mapping it to a codebook vector, which is then used as the input to the decoder. This operation is non-differentiable and the \edit{gradients of the loss with respect to the encoder parameters} must be obtained by approximation. In the VQ-VAE this is done using the straight-through estimator~\cite{bengio2013estimating}, i.e., the gradients \edit{with respect to the} encoder output are taken to be equal to those \edit{with respect to the} decoder input \ci{(i.e., the quatization step is ignored)}. Given a learned codebook $A\in\mathbb{R}^{K \times D}$, where $K$ is the codebook size and $D$ is the dimensionality of each codebook vector $a_k$, the quantized representation $q_t$ of $h_t$ is obtained as
\begin{align}
    q_t = a_k, \text{ where } k=\arg\min_j \norm{h_t - a_j}_2 .
\end{align}
\edit{The decoder $g(\cdot)$ is an autoregressive model that takes $q_{[1,t]}$ as input to generate $x_t$ \cite{oord2016wavenet}}. 
Codebook learning is facilitated by a two-term auxiliary loss similar to
classical vector quantization dictionary 
learning~\cite{burton_generalization_1983, soong_vector_1985}. 
Gradients for the codebook vectors are given solely by a term that moves
codebook vectors $a_k$ closer to the non-quantized vectors $h_t$. A so-called
\emph{commitment term} is added to ensure that non-quantized vectors do not grow
unboundedly by enforcing the encoder to keep them close to a codebook vector.
This commitment term is optimized only by the encoder. The
total VQ-VAE loss \edit{for a single timestep} is
% \begin{align}
%     %\mathcal{L} = \log p(x | q) +  \underset{\text{vq}}{\underbrace{\norm{\text{sg}\left[v\right] - A}_2^2}} + \underset{\text{commitment}}{\underbrace{\beta\norm{v - \text{sg}\left[A\right]}_2^2}} \enspace ,
%     \resizebox{.44\textwidth}{!}{%
%     $\mathcal{L}_t = \underset{\text{encoder+decoder}}{\underbrace{\log p(x_t | q_{[1,t]})}} +  \underset{\text{codebook}}{\underbrace{\norm{\mathrm{sg}\left[h_t\right] - A}_2^2}} + \beta\underset{\text{encoder}}{\underbrace{\norm{h_t - \mathrm{sg}\left[A\right]}_2^2}} \;,$
%     }
%     \label{eq: vector quantization losses}
% \end{align}
% \begin{align}
%     \mathcal{L}_t =&~
%     \underset{\text{encoder+decoder}}{\underbrace{\log p(x_t | q_{[1,t]})}} + \underset{\text{codebook}}{\underbrace{\text{MSE}\left(\mathrm{sg}\left[H\right], A\right)^2}} + \nonumber\\
%     & \underset{\text{encoder}}{\underbrace{\beta~\text{MSE}\left(H, \mathrm{sg}\left[A\right]\right)^2}} \;,
%     \label{eq: vector quantization losses}
% \end{align}
\edit{
\begin{align}
    \resizebox{.44\textwidth}{!}{%
    $\mathcal{L}_t = \underset{\text{encoder+decoder}}{\underbrace{\log p(x_t | q_{[1,t]})}} + \underset{\text{codebook}}{\underbrace{\text{MSE}\left(\mathrm{sg}\left[h_t\right], A\right)}} + \underset{\text{encoder}}{\underbrace{\alpha~\text{MSE}\left(h_t, \mathrm{sg}\left[A\right]\right)}} \;,$
    }
    \label{eq: vector quantization losses}
\end{align}}%
% \begin{align}
%     \mathcal{L}_t =&~
%     \underset{\text{encoder+decoder}}{\underbrace{\log p(x_t | q_{[1,t]})}} + \nonumber\\
%     & \underset{\text{codebook}}{\underbrace{\frac{1}{KD} \sum_{t=1}^{T}\sum_{k=1}^{K}\sum_{i=1}^{D} \left(\mathrm{sg}\left[h_{t,i}\right] - a_{k,i}\right)^2}} + \nonumber\\
%     & \underset{\text{encoder}}{\underbrace{\frac{\beta}{KD}\sum_{t=1}^{T}\sum_{k=1}^{K}{\sum_{i=1}^{D} \left(h_{t,i} - \mathrm{sg}\left[a_{k,i}\right]\right)^2}}} \;,
%     \label{eq: vector quantization losses}
% \end{align}
\noindent where \edit{$\log p(x_t|q_{[1,t]})$ is a reconstruction likelihood term usually using a categorical distribution,} $\mathrm{sg}[x] = x$ is the so-called stop-gradient operator \edit{which acts as the identity function during the forward pass but \ci{is assumed to have}
partial derivatives all equal to zero during the backward pass}, $\alpha$ is a scalar hyperparameter, 
% and we define $\text{MSE}(H,A) = \frac{1}{KD} \sum_{t=1}^{T}\sum_{k=1}^{K}\sum_{i=1}^{D}\left(h_{t,i} - a_{k,i}\right)^2$. 
\edit{and we define $\text{MSE}(h_t,A) = \frac{1}{KD} \sum_{k=1}^{K}\sum_{i=1}^{D}\left(h_{t,i} - a_{k,i}\right)^2$. 
The loss for a full sequence is the sum or mean over all $\mathcal{L}_t$.} 
% We note that the subtraction of a matrix from a vector in \cref{eq: vector
% quantization losses} implies the use of broadcasting.

These learned discrete representations have been shown to capture high-level
speech information closely related to phonemes, and are useful for
applications such as speaker conversion~\cite{chorowski2019unsupervised}.
Vector quantization is \edit{not} exclusive to VQ-VAE but has seen
widespread application within SSL for regularization purposes and to define
targets for the pretext task. We will cover these applications below.
%We highlight the models using vector-quantization in the column \textbc{{qtz}} in \cref{tab:model-taxonomy}.

\edit{The Gumbel softmax \cite{Gumbel-Softmax} is another frequently used approach for obtaining a discrete representation space, and has also been used for AEs \cite{eloff2019unsupervised}. In addition to the approaches discussed above, several other works on speech representation learning take inspiration from the AE framework \cite{zeiler2013rectified, badino2014auto, badino2015discovering, kamper2015unsupervised, renshaw2015comparison, settle2019_a2w}.}

\begin{comment}
\paragraph{Regularization}
In the generative-based approaches, some regularization in the encoder $f(\cdot)$ or decoder $g(\cdot)$ is required to encourage the model to utilize global speech information and prevent naive copying input when reconstruction. 
The models below have been enhanced by regularization methods. 
\begin{itemize}
%\item VAE, VQ-VAE
    \item DeCoAR 2.0~\cite{ling2020decoar} presents a deep contextualized acoustic representation learning approach with the addition of a vector quantization (VQ) layer.
    \item In VQ-APC~\cite{chung20e_interspeech}, a VQ layer is used with the APC objective, which imposes a bottleneck and forces the model to learn better representations.
    \item Two dropout regularization methods, attention dropout and layer dropout, are introduced to TERA~\cite{luo2021drop}. 
\end{itemize}
\end{comment}

\paragraph{Autoregressive prediction}
\label{par:apc}

Autoregressive predictive coding (APC)~\cite{chung2019unsupervised,
chung2020generative} \edit{takes inspiration from the classic Linear Predictive Coding (LPC) approach for speech feature extraction~\cite{LPC} and autoregressive language models (LM) for text, where the model learns to predict future information from past}.
% This approach can be viewed as training a speech version of an LM.
%Just like an LM for text, this approach uses an encoder model to encode temporal information of past acoustic sequence, and the decoder then predicts future frames.
%In this approach, given $X=\{x_1,x_2, ..., x_T\}$, $\widetilde{X}$ is the acoustic features up to time step $t$, that is, $X_{[0,t]} = \{x_1,x_2, ..., x_{t}\}$.
\edit{A function} $f(\cdot)$ reads the input sequence $X_{[1,t]}$ and \edit{outputs a representation sequence} $H_{[1,t]}$.
The \edit{auxiliary module} $g(\cdot)$ is a linear projection layer which takes the last vector of $H_{[1,t]}$ as input to \edit{approximate} $x_{t+c}$, where $c \geq 1$. Thus, $c$ indicates how many timesteps the model predicts ahead. The \edit{modules} $f(\cdot)$
and $g(\cdot)$ are jointly learned to minimize {the \ensuremath{L_1} loss between $x_{t+c}$ and its approximation $\hat{x}_{t+c}$}. APC is formulated as 
%\begin{align}
    %Z_{[0,t-1]} &= f(X_{[0,t-1]}) \\
    %\hat{\mathbf{x}}_{t+c} &= g(\mathbf{z}_{t-1}) \label{eq:c} \\
    %\mathcal{L}_t &= \lVert \hat{\mathbf{x}}_{t+c} - \mathbf{x}_{t+c} \rVert_1\enspace.
%\end{align}
\begin{align}
    H_{[1,t]} &= f(X_{[1,t]}) , \\
    \hat{x}_{t+c} &= g(h_{t}) \label{eq:c} , \\
    \mathcal{L}_t &= \lVert \hat{x}_{t+c} - x_{t+c} \rVert_1 .
\end{align}
In text-based autoregressive LMs, $c$ is set to $1$ \edit{to enable autoregressive generation}. However, due to the smoothness of the speech signal, neighboring acoustic features are usually similar. Depending on the downstream task, we are often interested in learning so-called \emph{slow features} that typically span multiple input frames~\cite{wiskott2002slow}. Even the smallest linguistic units of speech---phonemes---span $0.07$~seconds on average in the English TIMIT dataset~\cite{garofolo_timit_1993},   whereas spectrogram frames $\mathbf{x}_t$ are typically computed at $0.01$ second intervals. Thus, simply predicting the next frame constitutes a trivial pretext task for APC; the original work finds that $c=3$ performs well. 
%\begin{equation}
  %  L = \sum_{X} \sum_{t} d(x_{t+c}, g(h_t)),
%   L = d(x_{t+c}, g( f(X_{[0,t-1]}) )),
%\end{equation}
%where $d(x_{t+c}, g(h_t))$ is the distance between $x_{t+c}$ and  $g(h_t)$. 
In \cite{chung2020improved}, the APC objective is extended to multi-target
training. The new objective generates both past and future frames conditioned
on previous context. 
In VQ-APC~\cite{chung20e_interspeech}, quantization is used with the APC
objective, which imposes an information bottleneck serving as a regularizer.

A drawback of APC is that it encodes information only from previous timesteps
and not the entire input.
DeCoAR~\cite{ling2020deep} combines the bidirectionality of the popular NLP model
ELMo~\cite{peters2018deep} and the reconstruction objective of APC to alleviate
this issue and allow encoding information from the entire input. 
%, so it is able to learn deep contextualized acoustic representations. 
It uses a forward LSTM $f_1(\cdot)$ to encode $X_{[1,t]}$ and a backward LSTM
$f_2(\cdot)$ to encode $X_{[t+k,T]}$, where $k>1$: 
\begin{gather}
    H_{[1,t]} = f_1(X_{[1,t]}), \\
    H^\prime_{[t+k,T]} = f_2(X_{[t+k,T]}), \\
    \hat{X}_{[t+1,t+k-1]} = g(h_t, h^\prime_{t+k}).
\end{gather}
% The last vector in $H_{[1,t]}$, $h_{t}$, and the first vector in $H^\prime_{[t+k,T]}$, $h_{t+k}$, are the input to the decoder $g$ used to predict $X_{[t+1,t+k-1]}$. 
The input feature vector used in the downstream tasks is the concatenation of
$h_{t}$ and $h^\prime_{t}$.

\paragraph{Masked Reconstruction}
%\label{sssec:time-only-masked}
%In wav2vec 2.0~\cite{wav2vec2}, time masking is applied in the latent space. But we will not mention this in this section.

%Basic idea
Masked reconstruction is largely inspired by the masked language model (MLM) task from BERT~\cite{jacob2019BERT}. During BERT pre-training, some tokens in the input sentences are masked by randomly replacing them by a learned masking token or another input token. The model learns to reconstruct the masked tokens from the non-masked tokens. Recent work has explored similar pretext tasks \edit{for speech representation learning}. Similar to the DeCoAR model described above, this allows a model to learn contextualized representations that encode information from the entire input. While we here focus on the models that reconstruct the masked input, it is important to note that masking has also been used extensively for contrastive (\cref{contrastive_approaches}) and predictive (\cref{predictive_approaches}) models. 
%In \cref{tab:model-taxonomy}, the models that use masking are highlighted by the \textbc{{msk}} column.

From a high-level perspective, the training phase of models using masked
reconstruction can be formulated as
\begin{align}
    H &= f(m(X)), \\
    \hat{x}_t &= g(h_{t}), \\
    \mathcal{L}_t &= \lVert \hat{x}_{t} - x_{t} \rVert_1 .
\end{align}
\edit{The exact masking policy defined by $m(\cdot)$ differs from model to model and will be discussed further below.}
The function $f(\cdot)$ is typically a Transformer encoder~\cite{liu2020mockingjay,jiang2019improving,liu2020masked}, but recurrent neural networks have also been used~\cite{wang2020unsupervised}. \edit{In general, the Transformer encoder architecture has been adopted widely by self-supervised models for speech within all three surveyed categories.} \edit{The function $g(\cdot)$ is usually a linear projection or a multilayer perceptron (MLP)}. Finally, the loss $\mathcal{L}_t$ is commonly computed only for masked timesteps in order to discourage the model from learning an identity mapping.

%Time masking 
%In the time masking approaches~\cite{liu2020mockingjay,jiang2019improving,liu2020masked}, some of the input frames in $X$ are masked to zero or randomly replaced by other frames. $X$ with $t$-th to $t+k$-th frames masked is denoted as $X_{-[t,t+k]}$. The encoder $f(\cdot)$ is usually a Transformer that encodes the whole utterance. It takes $X_{-[t,t+k]}$ as input and generates $H$. The decoder $g(\cdot)$ is a prediction head, which use $H_{[t,t+k]}$ to predict $X_{[t,t+k]}$. The encoder $f(\cdot)$ and decoder $g(\cdot)$ are jointly learned to minimize the distance between the decoder output and $X_{[t,t+k]}$.
%\begin{equation}
%    L = d(X_{[t,t+k]}, g(f( X_{-[t,t+k]} )) ).
%\end{equation}

The masking policies used in NLP can be adapted to speech by considering a speech \edit{segment} equivalent to a token in a sentence; indeed, the masking strategy of BERT has also been used for speech pre-training~\cite{liu2020mockingjay}.
%However, taking inspiration from NLP is not sufficient, as speech data warrants domain-specific considerations.
In the standard BERT masking policy, each token is masked independently at random. However, for speech, masking a single \edit{sample or spectrogram frame results in a largely} trivial reconstruction task \edit{since, as discussed in paragraph \ref{par:apc}, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore it is common to mask chunks of consecutive frames~\cite{liu2020mockingjay,jiang2021further}.

We can bring the pretext task closer to the NLP equivalent by using a masking policy where the masked regions of the input correspond to linguistic units. Instead of just masking a fixed number of consecutive frames, pMPC~\cite{yue2021pMPC} selects masked speech frames according to the phonetic segmentation in an utterance. \edit{However, in order to obtain this segmentation, some labeled data is of course needed.}

Whereas most studies use masking along the temporal dimension of the input, speech can also be masked along the frequency dimension when spectral input features are used~\cite{wang2020unsupervised,liu2021tera}. Frequency masking has been shown to improve representations used for speaker classification~\cite{liu2021tera}. 
%The phonetically motivated self-supervised representation learns the speech representation that benefits downstream speech processing tasks.

%TERA
%Previous work mostly explored masking on the temporal axis, and the model learns to reconstruct from corrupted blocks of time steps. Frequency is another alteration for masking. The model can learn to reconstruct from missing blocks of frequency bins~\cite{wang2020unsupervised,liu2021tera}. In Transformer Encoder Representations from Alteration (TERA), the time and frequency alterations are applied together in the pre-training process. This is simply SpecAugment~\cite{spec_augment}. 
%It has been found that each of the alteration methods guides the model to learn a distinct aspect of speech~\cite{liu2021tera}. The time alteration effectively enforces a more accurate phoneme prediction, keyword detection, and speech recognition, as it leads the model to learn richer phonetic content. The frequency alteration effectively improves speaker prediction accuracy, as it leads the model to encode speaker identity.
%The magnitude alteration effectively improves performance for all tasks, as it potentially increases data diversity for pre-training.
Some studies explore alternatives to masking the input directly. In non-autoregressive predictive coding (NPC)~\cite{liu21l_interspeech}, time masking is introduced through masked convolution blocks. Taking inspiration from XLNet~\cite{XLNet}, it has also been suggested that the input be reconstructed from a shuffled version~\cite{song20d_interspeech} to address the discrepancy between pre-training and fine-tuning of masking-based approaches.
%In NLP, the masking tokens only exist in the pre-training phase and never appear in the downstream tasks, making the pre-training and downstream phases have mismatched inputs. To avoid using masks during pre-training, in NLP, XLNet is proposed, which learns by reconstructing from shuffled input. For speech, the speech version of XLNet, Speech-XLNet~\cite{song20d_interspeech}, has also been proposed.

Regularization methods can further improve on masked reconstruction approaches. \edit{DeCoAR 2.0~\cite{ling2020decoar} uses vector quantization}, which is shown to improve the learned representations. Furthermore, two dropout regularization methods---attention dropout and layer dropout---are introduced with the TERA model~\cite{liu2021tera, luo2021drop}. \edit{Both methods are variations on the original dropout method \cite{srivastava_dropout_2014}.}

% PASE, WaveNet autoencoders, Phase reconstruction, Audio2Vec
%\label{sssec:autoencoder}
%\subsubsection{Autoencoder} 
%TBD: 
% -- Do we have to mention ConvDMM? %The ConvDMM~\cite{khurana20_interspeech} approach learns speech representations with convolutional neural networks and Markov Models.
% -- More about auto-encoder

\paragraph{More Generative Approaches}
Other than the autoregressive and masked reconstruction tasks discussed above, various studies have explored the reconstruction of other targets derived from the input. PASE and PASE+~\cite{pascual2019learning,ravanelli2020multi} use multiple targets, including the waveform, log power spectrum, \edit{mel cepstral coefficients (MFCCs)}, and prosody features. Models that learn acoustic embeddings of small speech segments have targeted future and past spectrogram segments~\cite{chung2018speech2vec, tagliasacchi2019self, tagliasacchi2020pre}, phase information~\cite{quitry2019learning}, and the temporal gap between two segments~\cite{tagliasacchi2019self, tagliasacchi2020pre}.

%\begin{itemize}
% \item Autoencoder~\cite{chorowski2019unsupervised}: The input and output are the same. Usually the regularization introduced above is required. %In these works, the autoencoder framework is designed to encode only phonetic content in latent representation and remove other confounding detail such as speaker identity.
%\item The model learns through reconstructing a spectrogram slice from past and future slices~\cite{tagliasacchi2019self, tagliasacchi2020pre}. %Similar to DecoAR?
%\item The TemporalGap~\cite{tagliasacchi2019self, tagliasacchi2020pre} approach learns through estimating the temporal gap between two short audio segments extracted at random from the same audio clip. %Can this approach be considered as generative?
%\item Representations are learned through reconstructing the phase of the short-time Fourier transform from its magnitude~\cite{quitry2019learning}. %This is audio, not speech.
%\item In PASE~\cite{pascual2019learning,ravanelli2020multi}, a single neural encoder learns to solve multiple self-supervised tasks at once, including reconstruction of waveform, Log power spectrum, MFCC, prosody, and other binary discrimination tasks. %Is this PASE+? I cite PASE+ here, are all the tasks of PASE+ included here. Is this approach generative? 
%\end{itemize}

\subsubsection{Challenges}

Although successful NLP models like BERT and GPT are based on generative pretext tasks, the progress have not been translated directly to the speech domain. A speech signal encodes more information than text, such as speaker identity and prosodic features, which makes it harder to generate. However, in order to generate all details of the input, the model must encode all information in the speech signal. Hence, a model that learns to perfectly reconstruct its input may not necessarily have learned to isolate the features of interest \edit{\ci{and will encode} redundant information for a given downstream task.}

There are many choices involved in designing a generative pretext task. For instance, \edit{masking strategy and the choice of input and target representation (e.g., waveform samples or spectral
features)}. These choices influence what the model learns through the pretext task. However, there is little research on the relationship between task design and the information encoded in the learned representations.

%In \cite{slu_bert}, phoneme posterior vectors are used to train a standard BERT~\cite{bert, xlnet} model.
%The phoneme posterior vectors are output from a supervised acoustic model, which requires CTC loss training over the ground-truth phonemes.

% This is about network compression. Perhaps we will have a session about network compression.
%In Audio ALBERT~\cite{audioalbert}, Mockingjay is modified to have shared parameters across Transformer layers.

%Perhpas we will talk about attack defense
%In \cite{mockingjay_defense}, Mockingjay is shown to be effective in defending adversarial black-box attacks.

\input{chapters/paper_review/pretext}

\subsection{Contrastive approaches}
\label{contrastive_approaches}
\subsubsection{Motivation}
As discussed above, speech contains many entangled features. Thus, \edit{learning to reconstruct the raw speech signal} might not be the best way to discover contextualized latent factors of variations. 
Contrastive models learn representations by distinguishing a target sample (positive) from distractor samples (negatives) given an \emph{anchor representation}. \edit{The pretext task is to maximize latent space similarity between the anchor and positive samples while minimizing the similarity between the anchor and negative samples}. This approach has been used extensively in the general \edit{ML} community \cite{Schultz_Joachims_NeurIPS2004}.

\subsubsection{Approaches}
\paragraph{CPC}
\label{par:cpc}
Contrastive Predictive Coding (CPC)~\cite{oord2018representation} is a prominent example of a contrastive model. 
\edit{CPC uses a convolutional module $f_1(\cdot)$ to produce localized representations $z_t$ with a recurrent module $f_2(\cdot)$ on top that outputs a contextualized representation $h_t$. An anchor representation $\hat{z}_{t,k}$ is obtained via a linear projection $g_k(\cdot)$ of $h_t$. The positives and negatives are sampled from the localized representation $Z$.
%\edit{CPC forms localized representations $z_t$ that depend only on a local window of $X$ and then creates a contextualized representation $h_t$ by running an autoregressive module $f_2(\cdot)$ over $Z_{[1,t]}$. The positives and negatives are then sampled from the localized representation $Z$.
%while the contextualized representation $h_t$ is transformed to an anchor representation $\hat{z}_{t,k}$.
Hence, at a single timestep $t$, CPC forms multiple anchor representations $\hat{z}_{t,k}$ for $k\in\{1,\dots,K\}$ and associates with each one a single positive sample at the corresponding timestep, $z_{t+k}$, $k$ steps in the future:
\begin{align}
    z_t &= f_{1}(X_{[t-u,t+u]}) \label{eq: cpc local representation} \enspace ,\\
    H_{[1,t]} &= f_{2}(Z_{[1,t]}) \enspace , \\
    \hat{z}_{t,k} &= g_k(h_{t}) \enspace.
\end{align}
%
\noindent Each $z_{t}$ only encodes information from a limited receptive field, while $f_2(\cdot)$ is limited to condition each $h_t$ on previous timesteps $Z_{[1,t]}$. \ci{Without these restrictions}, the model could collapse to a trivial solution. $g_k$ is a unique transformation per \ci{offset} $k$ (e.g., a linear projection).
%where $f_1(\cdot)$ is a convolutional neural network, such that each $z_{t}$ only encodes information from a limited receptive field in $X$, $f_2(\cdot)$ is limited to condition each $h_t$ only on previous timesteps $Z_{[1,t]}$ and $g_k$ is a unique transformation per step $k$ (e.g. a linear projection). 
The loss function measures the similarity between the anchor representation $\hat{z}_{t,k}$ and the positive $z_{t+k}$ normalized by the total similarity to the positive and negatives.}
The approach is similar to previous work on Noise-Contrastive Estimation (NCE)~\cite{gutmann2010noise}. \edit{Minimizing} the loss corresponds to maximizing a lower bound on the mutual information between $h_t$ and $z_{t+k}$ (and in turn $x_{t+k-u:t+k+u}$) and is hence called InfoNCE:
\begin{align}
    \mathcal{L}_{t,k} &= - \log \left(\frac{\exp(\hat{z}_{t,k}^{\text{\tiny T}}z_{t+k})}{\sum_{\ci{i \in \mathcal{I}}} \exp(\hat{z}_{t,k}^{\text{\tiny T}}z_{i})} \right)\enspace .
    \label{eq: cpc loss}
\end{align}
%
\noindent \edit{Here, $\mathcal{I}$ \ci{is a random subset of $N$} indices  which includes the target index $t+k$ and $N-1$ negative samples drawn from a proposal distribution, e.g., a uniform distribution over $\{1,\dots,T\}$. Including the target index in $\mathcal{I}$ ensures that the loss is a proper categorical cross-entropy and that minimizing it has the previously stated relation to mutual information maximization}. This corresponds to sampling negatives from the same sequence and has been \edit{shown} to give good performance for phoneme classification~\cite{oord2018representation}. The loss is indexed by $k$ to show that CPC targets multiple offsets using different projection layers $g_k(\cdot)$. The authors find \edit{$K=12$} to work well for phoneme classification.

The wav2vec model \cite{schneider2019wav2vec} extends the CPC approach \edit{and uses fully convolutional parameterizations for the modules $f_1(\cdot)$ and $f_2(\cdot)$ with receptive fields of \SI{30}{ms} and \SI{210}{ms}, respectively. While the CPC loss solves a 1-of-$N$ classification task per $(t,k)$, either assigning the anchor to the positive class or (wrongly) to one of the $N-1$ negative classes, the wav2vec loss considers a sequence of $N$ independent binary classifications. That is, the anchor is compared independently to the positive and each negative, and the loss is computed as a sum of the associated log-probabilities,
\begin{align}
    %g_k(\mathbf{c}_{t}, \mathbf{v}_{t+k}) &= \log(\sigma(\mathbf{v}_{t+k}^{\intercal} (\mathbf{W}_k \mathbf{c}_{t} + \mathbf{b}_k)))\\
    \mathcal{L}_{t,k} &= - \log(\sigma(\hat{z}_{t,k}^{\text{\tiny T}}z_{t+k})) + \sum_{i \in \mathcal{I}} \log(1 - \sigma(\hat{z}_{t,k}^{\text{\tiny T}}z_{i})) \enspace .
    \label{eq: wav2vec loss}
\end{align}
\noindent Here, $\sigma(x)=1/(1+\exp(-x))$ is the sigmoid function, $\sigma(\hat{z}_{t,k}^{\text{\tiny T}}z_{t+k})$ is the probability of the anchor being the positive sample and $\sigma(\hat{z}_{t,k}^{\text{\tiny T}}z_{i})$ is the probability of the anchor being the negative sample. Evidently and contrary to CPC, $\mathcal{I}$ must not include the target index $t+k$ as this would cancel out the positive term.}

\paragraph{wav2vec 2.0} 
\edit{The wav2vec 2.0 model combines contrastive learning with masking}. As the CPC model, it uses the InfoNCE loss \cite{oord2018representation} to maximize the similarity between a contextualized representation and a localized representation. \edit{However, instead of using the $z_t$ directly as positive and negatives, it uses a quantization module $g(\cdot)$ to obtain a discrete representation. This has the practical implication that one can avoid sampling negatives from the same category as the positive}. \edit{The model} takes as input a waveform and uses a convolutional module \edit{$f_1(\cdot)$} followed by a \edit{Transformer encoder} \edit{$f_2(\cdot)$}. Masking is applied to the output of the convolutional module: 
\begin{align}
    z_t &= f_1(X_{[t-u,t+u]}) \label{w2v2 f_v} \enspace , \\
    H &= f_2(m(Z)) \label{w2v2 f_c} \enspace , \\
    q_t &= g(z_t) \enspace . \label{w2v2 qtz}
\end{align}
\noindent
%Here, $m(\cdot)$ again defines a masking policy 
% Here the $\circ$ operator denotes the Hadamard product with the mask $M$,
% , $f_1(\cdot)$ is a convolutional neural network with limited receptive field, $f_2(\cdot)$ is a Transformer encoder, 
The quantization module $g(\cdot)$ uses a Gumbel softmax~\cite{Gumbel-Softmax} with a straight-through estimator. Since the quality of the learned representations is contingent on the quality of the quantization, wav2vec 2.0 combines two techniques to learn high-quality codebooks. First, wav2vec 2.0 concatenates quantized representations from multiple codebooks at each timestep, so-called Product Quantization (PQ)~\cite{ProductQuantization}. Also, the \edit{primary} training loss \edit{described below} is augmented with an auxiliary term designed to encourage equal use of all codebook entries.

In wav2vec 2.0, anchors are taken to be $h_t$ at masked timesteps only, the positive sample is chosen as the quantized vector, $q_t$, at the same timestep, and negatives are sampled from other masked timesteps. The loss is
% For the InfoNCE loss, anchors features are taken from the top layer contextualized Transformer representations for masked regions,  positive samples are the corresponding quantized features, and negative samples are selected from other masked locations in the same utterance. 
%
\begin{align}
    \mathcal{L}_t &= - \log \left(\frac{\exp(S_{\text{c}}(h_{t}, q_{t}))}{\sum_{i \in \mathcal{I}} \exp(S_{\text{c}}(h_{t}, q_{i}))} \right) \enspace , \label{w2v2 loss}
\end{align}
%
\noindent where $S_{\text{c}}(\cdot)$ is the cosine similarity and $\mathcal{I}$ contains the target index $t$ and negative indices sampled from other masked timesteps.
%\edit{Notably, since the positives and negatives are quantized representations, we can guarentee that representations of the same class as the positive are excluded when sampling the negatives.}

The wav2vec 2.0 approach was the first to reach single-digit \edit{word error rate (WER) on LibriSpeech using only the low-resource Libri-light subsets for fine-tuning a pre-trained model (see \cref{sec:datasets}). It has subsequently} inspired many follow-up studies. The wav2vec-C~\cite{sadhu21_interspeech} approach extends wav2vec 2.0 with a consistency term in the loss that aims to reconstruct the input features from the learned quantized representations, similar to VQ-VAE~\cite{VQVAE2}.
%The wav2vec-C representations achieve, on average, twice the error reduction over baseline and a higher codebook utilization in comparison to wav2vec 2.0 for far-field English voice commands and voice queries.

\subsubsection{Challenges}
Although representations learned using contrastive approaches have proved effective across a wide range of downstream applications, they face many challenges when applied to speech data. 
One challenging aspect is that the strategy used to define positive and negative samples can also indirectly impose invariances on the learned representations. For example, sampling negatives exclusively from the same utterance as the positive biases the features towards speaker invariance, which may or may not be desired for downstream applications. 
Another standing challenge is that since speech input does not have explicit segmentation of acoustic units, the negative and positive samples do not represent a whole unit of language but rather partial or multiple units, depending on the span covered by each sample. 
Finally, since speech input is smooth and lacks natural segmentation, it can be difficult to define a contrastive sampling strategy that is guaranteed to provide samples that always relate to the anchor as truly positives and negatives in a sound way.

\subsection{Predictive approaches}
\label{predictive_approaches}
\subsubsection{Motivation}
\edit{Similar to the contrastive approaches discussed above, predictive approaches are defined by using a learned target for the pretext task. However, unlike the contrastive approaches, they do not employ a contrastive loss and instead use a loss function such as squared error and cross-entropy. 
Whereas a contrastive loss discourages the model from learning a trivial solution by the use of negative samples, this must be circumvented differently for predictive methods.
%However, such losses do not discourage trivial solutions, in the same way that contrastive losses do.
For this reason, predictive methods compute the targets outside the model's computational graph; usually with a completely separate model. Thus, the predictive setup is somewhat akin to teacher-student training. The first predictive approaches were motivated by the success of BERT-like methods in NLP \cite{jacob2019BERT} as well as the DeepCluster method in CV \cite{Caron_2018_ECCV}.}
%where the model learns a contextual representation that aggregates information from distant timesteps. Predictive approaches for speech representation learning also draw inspiration from the DeepCluster method \cite{Caron_2018_ECCV} for self-supervised visual learning.

\subsubsection{Approaches}

\paragraph{Discrete BERT} \edit{Applying BERT-type training directly to speech input is not possible due to its continuous nature}. The Discrete BERT approach~\cite{baevski2019effectiveness} uses a pre-trained vq-wav2vec model to derive a discrete vocabulary \cite{Baevski2020vq-wav2vec}. The vq-wav2vec model is similar to wav2vec mentioned in Paragraph~\ref{par:cpc} but uses quantization to learn discrete representations. 
%For the LibriSpeech dataset, the discrete vocabulary contains about 13.5k unique codes. 
\edit{Specifically, discrete units $c_t$ are first extracted with the vq-wav2vec model $f_1(\cdot)$ and then used as inputs \emph{and} targets in a standard BERT model $f_2(\cdot)$ with a softmax normalized output layer $g(\cdot)$, 
\begin{align}
    c_t &= f_1(X_{[t-u,t+u]}) \enspace , \\
    H &= f_2(m(C)) \enspace , \\
    \hat{c}_t &= g(h_t) \enspace .
\end{align}
Similar to BERT, the model can then be trained with a categorical cross-entropy loss,
\begin{align}
    \mathcal{L} &= \sum_{t\in \mathcal{M}} - \log p(c_t \mid X) \enspace ,
\end{align}
\noindent where $\mathcal{M}$ is the set of all masked timesteps. During training, only the BERT model's parameters are updated, while the vq-wav2vec model parameters are frozen.} 
% After pre-training, the BERT Transformer model is fine-tuned using a Connectionist Temporal Classification (CTC) \cite{graves2006connectionist} loss for ASR. 
%Following the Libri-light near-zero evaluation policy \cite{kahn2020libri}, fine-tuning is done using 10 minute, 1 hour and 10 hour subsets of LibriSpeech balanced for gender and recording conditions, as well as the standard 100 hour-clean part of LibriSpeech.
Discrete BERT was the first model to demonstrate the effectiveness of self-supervised speech representation learning by achieving a WER of 25\% on the standard test-other subset using a 10-minute fine-tuning set, \edit{setting the direction for many approaches to follow}.

\paragraph{HuBERT} Rather than relying on an advanced representation learning model for discretizing continuous inputs, as Discrete BERT, the Hidden Unit BERT (HuBERT) approach \cite{hsu2021hubert} \edit{uses quantized MFCC features as targets learned with classic k-means. Thus, to compute the targets, the k-means model $g_1(\cdot)$ assigns a cluster center to each timestep.  
% \begin{align}
%     c_t &= g_1(X_{[t-w,t+w]}) \enspace .
% \end{align}
% \noindent where $w$ defines the window size used to compute the MFCCs. 
Different from Discrete BERT, HuBERT takes the raw waveform as input, rather than discrete units. This helps to prevent loss of any relevant information due to input quantization.
%Specifically, HuBERT predicts the \edit{pre-computed} k-means cluster assignment given the masked continuous speech features.
HuBERT uses an architecture similar to that of wav2vec 2.0, with a convolutional module $f_1(\cdot)$ and a Transformer encoder $f_2(\cdot)$, as well as a softmax normalized output layer $g_2(\cdot)$:
%Similar to wav2vec 2.0, the HuBERT model uses a convolutional encoder for the input. Masking is then applied before the Transformer network as in wav2vec 2.0 \namecrefs{w2v2 f_v} \ref{w2v2 f_v} and \ref{w2v2 f_c}.
\begin{align}
    c_t &= g_1(X_{[t-w,t+w]}) \enspace , \\
    z_t &= f_1(X_{[t-u,t+u]}) \enspace , \\
    H &= f_2(m(Z)) \enspace , \\
    \hat{c}_t &= g_2(h_t) \enspace ,
\end{align}
\noindent where $w$ defines the window size used to compute the MFCCs. 
%Since the targets are pre-computed cluster identities, the HuBERT model can directly evaluate the regular categorical cross-entropy loss between the correct k-means cluster and the predicted one. This is opposed to constrastive methods which require negative samples to avoid degenerating to trivial solutions.
The categorical cross-entropy loss is} computed on both masked, $\mathcal{L}_m$, and unmasked, $\mathcal{L}_u$, timesteps:  
\begin{align}
 %   \mathcal{L}_m &= \sum_{t\in M} - \log p(c_t | \tilde{X}, t)
%    \mathcal{L}_m &= \sum_{t\in M} - \log p(c_t \mid H, t) \enspace ,
    % \mathcal{L}_m &= \alpha\sum_{t\in M} - \log p(c_t \mid H) + (1-\alpha)\sum_{t\notin M} - \log p(c_t \mid H) \enspace , \\
    \mathcal{L}_m &= \sum_{t\in \mathcal{M}} - \log p(c_t \mid X) \enspace , \\
    \mathcal{L} &= \beta
    \mathcal{L}_m + (1 - \beta) \mathcal{L}_u \enspace .
\end{align}
%
\noindent Again, $\mathcal{M}$ is the set of all masked timesteps, $\beta$ is a scalar hyperparameter and $\mathcal{L}_u$ is computed as $\mathcal{L}_m$ but summing over $t\notin \mathcal{M}$.

Intuitively, the HuBERT model is forced to learn both \edit{an acoustic and a language model. 
First, the model needs to learn a meaningful continuous latent representation for unmasked timesteps which are mapped to discrete units, similar to a classical frame-based acoustic modeling problem. Second, similar to other masked pre-training approaches, the model needs to capture long-range temporal dependencies to make correct predictions for masked timesteps}.

One crucial insight motivating this work is the importance of consistency of the targets which enables the model to focus on modeling the sequential structure of the input. \edit{Importantly though, for HuBERT, pre-training is a two-step procedure. The first iteration is described above.
%As described, in the first iteration of pre-training, k-means clustered MFCC features are used as targets, $c_t$.
Once completed, a second iteration of pre-training follows. Here, representations from a hidden layer of the model from the first iteration are clustered with k-means to obtain new targets $c_t$.}
%Here, the final model of the first iteration is used to encode latent representations of the full dataset. These are then quantized using k-means to form a new set of target vectors, replacing $c_t$ from the first iteration.
%Alternating between learning offline discrete targets using k-means and training the model using masked prediction ensures training stability while preserving representation quality.

\edit{For HuBERT, only two iterations are needed to match or outperform the previous state-of-the-art results for low-resource speech recognition. And combining the HuBERT approach with the wav2vec 2.0 approach, the w2v-BERT model has managed to improve results even further~\cite{w2vbert}.}

%\edit{The w2v-BERT approach~\cite{w2vbert} combines contrastive and predictive approach and optimizes the wav2vec 2.0 contrastive loss and the HuBERT masked prediction loss jointly}. It showed 5\% to 10\% relative WER reduction on the test-clean and test-other subsets of Librispeech, and more than 30\% relative improvement for real-world in-house voice search compared to a wav2vec 2.0 baseline.

\paragraph{WavLM}

\edit{WavLM emphasizes spoken content modeling and speaker identity preservation~\cite{chen2021wavlm}. It is largely identical to HuBERT, but introduces two useful extensions.}

\edit{First, it extends the Transformer self-attention mechanism with a so-called \emph{gated relative position bias}. The bias is added prior to the softmax normalization of the attention weights. For the attention weight at $i,j$, the bias is computed based on the input to the Transformer layer at the current time step $i$ and also incorporates a relative positional embedding for $i-j$. The authors find that this extension improves performance on phoneme and speech recognition tasks.}

%First, it extends the Transformer self-attention with a content-based gated relative position bias which improves the model's capability on recognition tasks compared to the convolutional context embedding \edit{used} in HuBERT and wav2vec 2.0. The gates adjust the relative position bias by conditioning on the current speech content \edit{which differs from the usual approach of learning a fixed positional encoding that is independent of the input}. Intuitively, the same distance offset between two frames tends to play different roles if one frame is silent and the other belongs to a speech segment.
%WavLM uses 320 bucketed positions for the range of possible offsets.
%The relative positional embedding parameters are shared across all layers.

\edit{Second, it uses an utterance mixing strategy where signals from different speakers are combined to augment the training data.} \edit{Specifically, random subsequences from other examples in the same batch are scaled and added to each input example.
%each example in a batch is combined with random subsequences from other examples in the same batch.
%Mixed utterances are selected from the same training minibatch, then randomly cropped and scaled by a sampled source energy ratio.
Only the targets corresponding to the original example are predicted during pre-training. Thus, the model learns to filter out the added overlapping speech.}

Most SSL methods are trained on data \edit{where each example only contains speech from a single person}; therefore, they can perform subpar on multispeaker tasks like speaker separation and diarization. 
%The content information corresponding to the main speaker is used as the target during the masked prediction pre-training.

%The pre-training data used for WavLM extends the sources used for HuBERT and wav2vec 2.0 to reach a total of 94k hours of public audio.
The WavLM model achieved substantial improvements on the speech separation, speaker verification and speaker diarization tasks in the SUPERB benchmark, while also performing well on many other tasks compared to HuBERT and wav2vec 2.0.
%It also improved upon \edit{the} HuBERT and wav2vec 2.0 models across all ten tasks of the SUPERB benchmark while maintaining a comparable performance on ASR.

\paragraph{data2vec} \edit{Motivated by the success of using an exponential moving average (EMA) teacher for self-supervised visual representations~\cite{grill2020byol, dino}, the data2vec model~\cite{data2vec} computes targets $Y$ using an EMA of its own parameters. The targets are constructed by averaging hidden representations of the top $k$ layers of the EMA teacher network applied to unmasked inputs. Here, we denote this jointly as $\bar{f}_2(\cdot)$.

The data2vec model was proposed for different data modalities, but for audio it uses an architecture similar to wav2vec 2.0 and HuBERT with a convolutional module $f_1(\cdot)$, a Transformer $f_2(\cdot)$ and masking applied to the Transformer input.
\begin{align}
    z_t &= f_1(X_{[t-u,t+u]}) \enspace , \\
    H &= f_2(m(Z)) \enspace , \\
    Y &= \bar{f}_2(Z) \enspace .
\end{align}
The teacher network $\bar{f}_2(\cdot)$ is a copy of the Transformer of the student network but with the parameters at training step $i$, $\theta_{\text{teacher},i}$, given by an EMA of the student parameters over all previous training steps.
\begin{equation}
    \theta_{\text{teacher},i} = \begin{cases}
        \theta_{\text{student},0} & i = 0 \\
        \gamma\theta_{\text{student},i} + (1-\gamma)\theta_{\text{teacher},i-1} & i > 0 
    \end{cases} \enspace ,
\end{equation}
\noindent where $\theta_{\text{student},i}$ are the parameters of the student network at training step $i$, updated via gradient descent, and $\gamma$ is the EMA decay rate.

The data2vec model uses a regression loss between target and prediction. Specifically, to reduce sensitivity to outliers and prevent exploding gradients, it uses the smoothed \ensuremath{L_1} loss \cite{fastrcnn},
\begin{equation}
\mathcal{L}_t =
    \begin{cases}
      \frac{1}{2}(y_t - h_t)^2 / \eta\text{,} & |y_t - h_t| \leq \eta\\
      |y_t - h_t| - \frac{1}{2}\eta\text{,} & \text{otherwise}
    \end{cases} \enspace ,
\end{equation}
\noindent where the hyperparameter $\eta$ controls the transition from a squared loss to an \ensuremath{L_1} loss.}

\edit{The data2vec approach was shown to work well for representation learning with either speech, images or text data.} It is the first approach to achieve competitive results when trained on \edit{any} one of the three modalities.

\subsubsection{Challenges}
The iterative nature of pre-training for the HuBERT and wavLM could present a practical inconvenience when working with large volumes of data. Another challenge for these models centers around the quality of the initial vocabulary generated from MFCC features. 
% Although \edit{a} masked prediction loss enables the models not to be \edit{limited} by the quality of their initial target units, experiments show that initial teachers with higher quality, e.g., Gaussian Mixture Models (GMMs) instead of k-means, provide non-negligible performance gains.
The data2vec approach improves over other predictive models \edit{by allowing the targets to improve continuously via the EMA teacher network}; however, \edit{student-teacher approaches inflate the existing computational challenges of very large models and may necessitate the use of methods that decrease instantaneous memory utilization such as mixed precision training, model parallelism and model sharding \cite{paszke2019pytorch}.}

\subsection{Learning from multi-modal data}
% {\color{blue} Karen}\\
\subsubsection{Motivation}
%Benefit from the naturally parallel data without annotation
Multiple modalities are useful in many settings, where each modality provides information that is complementary to other modalities.  Multi-modal work includes supervised settings, such as audio-visual ASR~\cite{petajan1984automatic,potamianos2003recent} and person identification~\cite{aleksic2006audio} which have been studied for decades.  In this section, we focus only on unsupervised representation learning from multi-modal data.

One of the motivations for learning from multiple modalities is that it can reduce the effect of noise, since noise in different modalities is likely to be largely independent or uncorrelated.  In addition, learning from speech data with accompanying signals such as images or video can help learn representations that encode more semantic information. Such ``grounding" signals can contain supplementary information that can be used by models to infer the content of the speech. Human language learning provides a proof of concept for this, as it is believed that infants benefit from the visual modality when learning language~\cite{legerstee1990infants}. Early computational models of multi-modal language learning were motivated by (and tried to emulate) human learning of language in the context of the visual surroundings~\cite{roy1999learning}.

\subsubsection{Approaches}
We define two broad classes of approaches in this area. Specifically, depending on what type of multi-modal data is involved we refer to ``intrinsic" or ``extrinsic" modalities.

%\paragraph{Intrinsic modalities} 
\edit{\textit{Intrinsic modalities} are} 
%refer to 
modalities produced directly by the speech source.  Examples of intrinsic modalities (besides the speech audio) include images or video of the speaker's face~\cite{lee2004avicar,chung2016lip}, lip-movement~\cite{shi2022learning}, articulatory flesh point measurements~\cite{westbury1990x,wrench2001new}, or simultaneous \edit{magnetic resonance imaging (MRI)} scans~\cite{narayanan2011multimodal}.  Typically, learning from multiple intrinsic modalities is done so as to improve robustness to noise, since acoustic noise is likely to be uncorrelated with the other \edit{modality(ies)}.  This type of representation learning is often referred to as ``multi-view learning" because the multiple intrinsic modalities can be seen as multiple views of the same content. Some typical approaches in this category include
\begin{itemize}
    \item Multi-view autoencoders and variations~\cite{ngiam2011multimodal,badino2016integrating},
    \item Multi-modal deep Boltzmann machines~\cite{srivastava2012multimodal},
    \item Canonical correlation analysis (CCA)~\cite{hotelling1936relations} and its nonlinear extensions~\cite{andrew2013deep,wang2015deep,wang2016deep,michaeli2016nonparametric,Melzer_01a,LaiFyfe00a,LaiFyfe99a,BachJordan05a,wang_15a},
    \item Multi-view contrastive losses~\cite{HermanBlunsom14a,Huang_13b},
    \item More recently, audio-visual extensions of masked prediction methods~\cite{shi2022learning,shi2022robust}, specifically Audio-Visual HuBERT (AV-HuBERT)~\cite{shi2022learning}.
\end{itemize}

%\paragraph{Extrinsic modalities} refer to
\textit{Extrinsic modalities} are modalities that are not produced by the same source but still provide context for each other. A typical example is an image and its spoken caption: The image tells us what the speech is likely describing, so a representation model that takes both modalities into account will hopefully encode more of the meaning of the speech than a single-modality model. 
There has recently been \edit{a} surge of datasets collected for this purpose, usually consisting of images and spoken captions, the audio and image frames in a video, or video clips with their spoken descriptions. A recent review of datasets, as well as methods, in this category is provided by Chrupa\l{}a~\cite{chrupala2021visually}.

Typical approaches involve learning a neural representation model for each modality, with a multi-modal contrastive loss that encourages paired examples in the two modalities to have similar representations while unpaired examples remain different~\cite{synnaeve+etal_nipsworkshop14,harwath+etal_nips16,harwath+glass_asru15,merkx2019language,rouditchenko2020avlnet,peng2022fast}. 
Other choices include training with a masked margin softmax loss~\cite{ilharco2019large,sanabria2021talk} or a masked prediction loss~\cite{MultimodelAmazonICASSP22}.  Such models are typically evaluated on cross-modal retrieval, although some work has also used the models for other downstream tasks such as the ZeroSpeech and SUPERB benchmark tasks~\cite{peng2022self}. 
Analyses of such models have found that, despite the very high-level learning objective of matching speech with a corresponding image (or other contextual modality), such models often learn multiple levels of linguistic representations from the shallowest to the deepest model layers~\cite{harwath2019learning,chrupala2017representations,scharenborg+etal_icassp18}.  They are also able to learn word-like units~\cite{harwath2018jointly,peng2022word,wang2020dnn} and can be used for cross-lingual retrieval, by considering the visual signal as an ``interlingua"~\cite{harwath2018vision,havard2019models,kamper+roth_sltu18}. 
In some settings, even in the presence of some amount of textual supervision (i.e., the speech is transcribed), visual grounding still helps learn a better representation for retrieval~\cite{pasad2019contributions}.

There has also been growing interest in learning joint speech and text representations using paired and unpaired data. The SLAM approach~\cite{bapna2021slam} is an example where speech and text are first represented using two separate pre-trained encoders followed by a multi-modal encoder to build the joint representations. The entire model is trained using a multi-task loss including two supervised and two self-supervised tasks.

%Benefiting from natural pairing of modalities in videos. Audio and text meta-data in FB videos. \\
%oWE can add it here or in the TTS subsection:
% link in comment to avoid compilation errors: https://arxiv.org/pdf/2011.11564.pdf \\

\subsubsection{Challenges}
One of the challenges of using multi-modal approaches is that the multi-modal data they rely on is often in shorter supply than single-modality data. In addition, multi-modal data is typically drawn from specific domains, for example domains involving descriptions of visual scenes. It is not clear how well the learned speech representations apply to other speech domains that are not necessarily describing or situated in a visual scene, and this question requires further study.

\subsection{Acoustic Word Embeddings}

% {\color{blue} Karen.} \\
% \kl{I moved this section here from the zero-resource section.}\\

Most of the representation learning techniques discussed in the preceding sections are aimed at learning frame-level representations. For some purposes, however, it may be useful to explicitly represent longer spans of speech audio of arbitrary duration, such as phone, word, or phrase-level segments.  For example, searching within a corpus of recorded speech for segments that match a given (written or spoken) query can be seen as finding segments whose representations are most similar to that of the query~\cite{levin+etal_icassp15,guoguo+etal_icassp15,Chung2016AudioWord2Vec,settle2017query}; word embeddings can be defined by pooling representations of instances of a given word~\cite{chung2018speech2vec}; unsupervised segmentation and spoken term discovery can be seen as a problem of detecting and clustering segments~\cite{kamper2017embedded,Kamper17BESGMM}; and even ASR can be viewed as the problem of matching written word representations to representations of audio spans~\cite{maas+etal_icmlwrl12,bengio+heigold_interspeech14,settle2019_a2w}.

Several lines of work have begun to address the problem of learning representations of spans of speech, especially word segments, typically referred to as \textit{acoustic word embeddings}.  Early work on unsupervised acoustic word embeddings defined them as vectors of distances from the target segment to a number of pre-defined ``template" segments~\cite{levin+etal_asru13}. Later work used variants of neural autoencoders~\cite{Chung2016AudioWord2Vec,holzenberger2018learning,kamper2019truly,pengcorrespondence}. These are often evaluated on word discrimination, that is the task of determining whether two word segments correspond to the same word or not~\cite{carlin2011rapid}. This task can be thought of as a proxy for query-by-example search, since the basic operation in search is to determine whether a segment in the search database matches a query segment, and has been used for evaluation of both frame-level (e.g.,~\cite{kamper2015unsupervised}) and word-level~\cite{levin+etal_asru13,kamper2016deep} representations.

% \kl{I left out much of the work on acoustic word embeddings, since it is in a supervised or at least weakly supervised setting.  I could add a quick note that a lot of other work exists, and add citations, but not sure that's needed for this paper.}

Since most work on acoustic word embeddings preceded the very recent wave of new self-supervised frame-level representations, one question is whether word (or more generally segment) embeddings could be derived more simply by pooling self-supervised frame-level representations, as has been done for text span embeddings by pooling over word embeddings~\cite{toshniwal2020cross,wang2021phrase}. Some initial results suggest that at least very simple pooling approaches like downsampling and mean or max pooling are not successful~\cite{van2021comparison,pengcorrespondence}, but more work is needed to reach conclusive results.

% The model learns through reconstructing a spectrogram slice from past and future slices~\cite{tagliasacchi2019self, tagliasacchi2020pre}. %Similar to DecoAR?
% The TemporalGap~\cite{tagliasacchi2019self, tagliasacchi2020pre} approach learns through estimating the temporal gap between two short audio segments extracted at random from the same audio clip. %Can this approach be considered as generative?
% Representations are learned through reconstructing the phase of the short-time Fourier transform from its magnitude~\cite{quitry2019learning}. %This is audio, not speech.

%We put the work of audio word embedding here.
%This can be a link to zero speech and further speech2semantic applications.
%Originally, audio word embedding was targeted at query-by-example, so the researchers want the embedding to contain the phoneme information (for example,  https://arxiv.org/abs/1510.01032, https://arxiv.org/abs/1603.00982). 
%Then there is some work targeting at including semantic information in the embedding (for example, https://arxiv.org/abs/1803.08976).

% should we add a section on Learning representation through multi-tasking}

### 2023-03-04 chapters/paper_review/04_benchmarking.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\section{Benchmarks for Self-supervised Learning}
\label{section:benchmark}
% {\color{blue} Daniel}\\

% {\color{blue} Reviewer: SW, Hung-yi Lee, Abdo }\\

%\subsection{Benchmarks for self-supervised learning approaches}

%This paper is about the size of self-supervised model. I think it can also be considered as benchmark.
%https://www.amazon.science/publications/scaling-effect-of-self-supervised-speech-models

%We have shown in the previous sections various methodologies to learn speech representations from unlabeled corpora. This section surveys the available datasets to learn the representations and evaluate the learned ones. We also summarize several works and their results to demonstrate how effective the learned representations are via providing robust model initialization for various downstream tasks.
The previous sections presented various methodologies by which to learn speech
representations from unlabeled corpora. This section surveys the datasets
available to learn and evaluate these representations. We also summarize
several studies and their results to demonstrate the usefulness of the learned
representations for various downstream tasks.

\subsection{Datasets only for pre-training} 
\Cref{table:datasets} summarizes datasets used for pre-training SSL techniques
in the literature. These datasets are usually large but with limited or no
labels. Libri-light (LL)~\cite{kahn2020libri}, one of these datasets, is
derived from audiobooks that are part of the
LibriVox\footnote{https://librivox.org/ \label{librivox}} project. LL contains
60k hours of spoken English audio tagged with SNR, speaker ID, and genre
descriptions. The speech examples in Audioset~\cite{gemmeke2017audio}, which
consists of over 2M 10-second YouTube video clips human-annotated with 632
audio events, have also been used for pre-training. Audioset has 2.5k hours
of audio of varying quality, different languages, and sometimes multiple sound
sources. AVSpeech~\cite{ephrat2018looking} is another large-scale audio-visual
dataset used in SSL research, comprising 4.7k hours of clips from a wide
variety of languages. 
%from 150k distinct speakers
Each clip contains a visible face and audible sound originating from a single
speaker without interfering background signals. The 3100-hour audio part of
AVSpeech has been used to learn audio-only 
representations~\cite{kawakami2020learning}. The Fisher corpus~\cite{cieri2004fisher} collects
over 2k hours of conversational telephone speech, 1k hours of which is utilized
for pre-training~\cite{jiang2021further}. Industrial researchers have also
begun to build large-scale datasets for learning speech representations.
For instance, 10k hours of real-world far-field English voice commands for
self-supervised pre-training have been collected at 
Amazon~\cite{sadhu21_interspeech}.

In addition to these English and multilingual efforts, researchers have also
collected corpora for pre-training Chinese speech representations. Didi Dictation
and Didi Callcenter~\cite{jiang2019improving, jiang2021further} are two
internal datasets containing respectively 10k hours of read speech collected
from mobile dictation application and 10k hours of spontaneous phone calls
between users and customer service staff.

\subsection{Datasets for both pre-training and evaluation}\label{sec:datasets}
Several datasets that provide both speech and associated transcripts and
speaker labels have also been used to develop SSL techniques by enabling
in-domain pre-training and evaluation. Such datasets are also listed in
\cref{table:datasets}. One of the most commonly used datasets in this category
is LibriSpeech (LS)~\cite{panayotov2015librispeech}, a labeled corpus
containing 960 hours of
%16 kHz sampled 
read English speech, which is also derived from an open-source audiobooks
project.\textsuperscript{\ref{librivox}} The corpus consists of subsets
\textit{train-clean-100}, \textit{train-clean-360}, \textit{train-other-500},
\textit{dev-clean}, \textit{dev-other}, \textit{test-clean}, and
\textit{test-other} used for training, development, and testing, respectively.
Subsets tagged with \textit{other} are more challenging utterances from
speakers that yield higher WER as measured with previously built models. LS is
used for unsupervised representation pre-training by ignoring its labels, and
can also be utilized to evaluate the performance of representation on ASR, phoneme recognition (PR),
phoneme classification (PC), and speaker identification (SID) tasks. Wall Street Journal (WSJ)~\cite{paul1992design} is another
widely adopted, labeled corpus for pre-training. Its labels can evaluate
performance for ASR, PR, PC, and SID. The original WSJ corpus contains 400 hours
of English read speech data, and today its \textit{si284} (81 hours),
\textit{dev93}, \textit{eval92} subsets are the most-used partitions for
unsupervised training, development, and test, respectively.  The \textit{si84}
(15 hours) partition is also used for training.

The speech community also utilizes multilingual corpora. These are often
large-scale, which facilitates pre-training, but are also partially labeled for ASR
evaluation (PC and PR can be enabled via phone-level forced alignment). These
corpora include Common Voice (CV)~\cite{ardila2020common}, Multilingual
LibriSpeech (MLS)~\cite{pratap20_interspeech}, VoxPopuli 
(VP)~\cite{wang-etal-2021-voxpopuli}, and BABEL (BBL)~\cite{gales2014speech}. CV is
an open-source, multi-language, growing dataset of voices containing 11k hours
of audio from 76 languages as of the date this review was written (Common Voice
corpus~7.0). Researchers usually use part of this for pre-training (e.g., 7k
hours/60 languages in \cite{babu2021xlsr} and 430 hours/29 languages in
\cite{kawakami2020learning}) or evaluation. 
%Demographic metadata like age, sex, and accent is also provided.
MLS derives content from read audiobooks of LibriVox and contains data in eight
European languages for a total of 50k hours of audio. VP comprises a total of 400k
hours of parliamentary speech from the European parliament in 23 European
languages.
The entire dataset~\cite{babu2021xlsr} or a 24k-hour 
portion~\cite{chen2021unispeechsat, chen2021wavlm} thereof has been used for pre-training. BBL
consists of 1k hours of conversational telephone speech in 17 African and Asian
languages.

Several datasets, including GigaSpeech~\cite{chen21o_interspeech}, TED-LIUM~3
(TED3)~\cite{hernandez2018ted}, TED-LIUM~2 (TED2)~\cite{rousseau2012ted},
Switchboard (SWB)~\cite{godfrey1992switchboard}, 
TIMIT~\cite{garofolo_timit_1993}, and VoxLingua107~\cite{valk2021voxlingua107}, are
labeled and conventionally used for evaluation, while their audio streams are
also aggregated to build diversified and large-scale corpora for unsupervised
pre-training~\cite{kawakami2020learning, song20d_interspeech, babu2021xlsr}.
GigaSpeech is a multi-domain English ASR corpus with 33k hours of audio
collected from audiobooks, podcasts, and YouTube. A subset of 10k audio is
transcribed. TED2 comes with 118 hours of English speech extracted from
TED conference talks and its transcription for evaluating ASR. Its recordings
are clear but with some reverberation. TED3 is an extension of TED2 and
comprises 450 hours of talks. SWB is a 260-hour conversational speech
recognition dataset containing two-sided telephone conversations.
%The dataset was recorded in 1990 with a lower (8k) sampling rate than the other corpora, and thus presents a challenging recognition problem.
The TIMIT corpus was designed to provide read speech data and its word and
phone-level transcriptions for acoustic-phonetic studies. It contains recordings
%from 630 speakers with 8 dialects 
in American English. Compared to the previous corpora labeled for ASR
evaluation, VoxLingua107 consists of 6.6k hours of audio in 107 languages and
is annotated for language identification. Beyond the original purpose of
evaluation, these corpora are also used in pre-training to improve the
generalizability of learned representations.

%, and each person read 10 phonetically-rich sentences.

For the purpose of pre-training and evaluating Mandarin speech representations,
the authors of \cite{jiang2019improving, jiang2021further} also compiled Open Mandarin, 
an open-source Mandarin dataset of 1.5k hours of speech from
the Linguistic Data Consortium (LDC) and
OpenSLR.\footnote{\url{https://openslr.org}} Open Mandarin consists of the HKUST
Mandarin Telephone Speech Corpus (HKUST, 200 hours of spontaneous 
speech, of which   % AMH: check
%from over 2100 speakers in mainland China. 
168 hours of audio is used for pre-training; the development and test sets are
excluded.)~\cite{liu2006hkust}, AISHELL-1~\cite{bu2017aishell} (178 hours of
read speech),
%recorded in a quiet indoor environment from 400 speakers with different accents
aidatatang 200zh (200 hours, read speech)~\cite{aidatatang}, MAGICDATA
Mandarin Chinese Read Speech Corpus (755 hours, read speech)~\cite{magic_data},
Free ST Chinese Mandarin Corpus (ST-CMDS, 100 hours, read speech)~\cite{st_cmds},
and Primewords Chinese Corpus Set 1 (100 hours, read speech)~\cite{primewords_201801}. 
Both HKUST and AISHELL-1 are labeled and are suitable for ASR evaluation.

\subsection{Datasets for evaluation}

Besides the aforementioned datasets, conventional speech processing benchmarks
are also used to evaluate self-supervised representations. Studies leverage
% Hub5'00, 
  Hub5,     % AMH: check
DIRHA, and CHiME-5 to measure the efficacy of representations in ASR.
The Hub5 evaluation (LDC2002T43 and LDC2002S09, also referred to as the NIST 2000
Hub5 English evaluation set) contains 40 transcribed English telephone
conversations only for testing, where 20 are from conversations collected in
SWB studies but not released with the SWB dataset, and the rest are from CallHome
American English Speech (LDC97S42). DIRHA~\cite{ravanelli2015dirha}, short for
Distant-speech Interaction for Robust Home Applications, is a database composed
of utterances sampled from WSJ, speech of keywords and commands, and
phonetically-rich sentences. These utterances are read by UK and US English
speakers and recorded with microphone arrays. CHiME-5~\cite{barker2018fifth} is
a challenge that aims to advance robust ASR and presents a dataset of natural
conversational speech collected under a dinner party scenario with microphone
arrays. A team at Amazon Alexa also recorded and transcribed a corpus of 1k
hours of audio for model training and evaluation~\cite{sadhu21_interspeech}.

Researchers also evaluate representations for sentiment analysis with the
INTERFACE~\cite{hozjan2002interface} and MOSEI (CMU Multimodal Opinion
Sentiment and Emotion Intensity)~\cite{zadeh2018multimodal} datasets. INTERFACE
is an emotional speech database for Slovenian, English, Spanish, and French,
and contains six emotions: anger, sadness, joy, fear, disgust, and
surprise, plus neutral. MOSEI is comprised of sentence-level sentiment
annotations of 65 hours of YouTube videos
%from 1000 speakers having 
using emotion categories similar to INTERFACE, but replacing joy with
happiness.

In addition, datasets employed to demonstrate the benefit of SSL
representations on various tasks include VCTK \cite{veaux2016superseded} and VoxCeleb1 \cite{nagrani17_interspeech} for SID/ASV (automatic speaker verification) tasks, FSC (Fluent Speech Commands) \cite{lugosch19_interspeech} for IC (intent classification), QUESST
(QUESST 2014) \cite{anguera2015quesst2014} for QbE (query by example), LS En-Fr \cite{kocabiyikoglu2018augmenting} and CoVoST-2 \cite{wang2020covost} for ST (speech translation), and ALFFA and OpenSLR-multi for multilingual ASR. The VCTK corpus includes speech
data with 109 English speakers of various accents, each reading out about 400
sentences sampled from newspapers. VoxCeleb1 is an audio-visual dataset
comprised of short YouTube clips containing human speech. It consists of 1251
unique speakers and 352 hours of audio. FSC contains utterances of spoken
English commands that one might use for a smart home or virtual assistant, and
is used to evaluate the performance of a spoken language understanding system.
The QUESST search dataset comprises spoken documents and queries in 6~languages
to measure the capability of models in spotting spoken keywords from documents.
LS En-Fr is a dataset augmenting existing LS monolingual utterances with
corresponding French translations to train and evaluate English-French machine
translators. CoVoST-2 is a multilingual speech translation benchmark based on
CV. It provides data for translating from English into 15 languages and from 21
languages into English, and has a total of 2.9k hours of speech. The ALFFA
project\footnote{\url{http://alffa.imag.fr}} collects speech of African
languages to 
% drive 
  promote    % AMH: check
the development of speech technologies in Africa, and
\cite{kawakami2020learning} leverages four African languages collected in the
project for evaluation: Amharic~\cite{tachbelie2014using}, Fongbe~\cite{laleye2016first},
Swahili~\cite{gelas2012developments}, and Wolof~\cite{gauthier2016collecting}.
In the same work~\cite{kawakami2020learning}, the authors
further select 21 phonetically diverse languages from OpenSLR to evaluate the
generalizability of SSL representations across languages. We denote the
collection as OpenSLR-multi below.

Last, \cite{tagliasacchi2020pre} puts together five datasets 
(MUSAN~\cite{snyder2015musan}, Bird Audio Detection~\cite{stowell2019automatic},
Speech Commands~\cite{warden2018speech}, Spoken Language 
Identification~\cite{oponowiczspoken}, and TUT Urban Acoustic Scenes 2018~\cite{dcase2018}) plus
an SID task built with the LS \textit{train-clean-100} subset to evaluate the
capability of representations on audio event detection.
\cite{quitry2019learning} employs the NSynth dataset~\cite{engel2017neural} on top
of the six for benchmarking. As many of the datasets are built for research in
audio processing, we here provide only a list of these datasets for reference.

\begin{table*}[ht]
  \centering
  \scriptsize %footnotesize %small, scriptsize, tiny
  \caption{Summary of datasets used in pre-training (denoted as PT) or
  evaluating (denoted as EV) SSL techniques in the literature. The languages and
  sizes of the datasets are provided in columns 3 and 4. Column 5 lists the tasks each
  dataset is used to evaluate. We use the following abbreviations:
  \textbf{EN}: English; 
% \textbf{Multi}: multilinguality; 
  \textbf{Multi}: multilingual;     % AMH: check
  \textbf{ZH}: Chinese;
  \textbf{Fr}: French; \textbf{ASR}: automatic speech recognition; \textbf{PR}:
  phoneme recognition; \textbf{PC}: phoneme classification; \textbf{SID}:
  speaker identification; \textbf{ASV}: automatic speaker verification;
  \textbf{Sentiment}: sentiment analysis; \textbf{ST}: speech translation;
  \textbf{QbE}: query by example or spoken term detection; \textbf{IC}: intent
  classification; \textbf{AED}: audio event detection; and \textbf{LID}: language
  identification. We distinguish \textbf{PR} from \textbf{PC} based on whether
  the inference is made at the phone level sequentially or the frame level separately.
  \textbf{SID} and \textbf{ASV} both evaluate model capability in encoding
  speaker information; \textbf{SID} classifies one utterance into a pre-defined
  set of speaker labels, whereas \textbf{ASV} infers whether a given pair of
  utterances was uttered by the same speaker.}
  \label{table:datasets}
  {\renewcommand*\arraystretch{1.2}
  \begin{tabular}{llllll}
    \toprule
    Dataset & Purpose & Lang. & Size [hours] & Task & License \\
    \midrule
    % \multirow{2}{*}[0em]{wav2vec 2.0}
    LibriLight (LL) & PT & EN & 60k & - & MIT License \\ \hline
    AudioSet & PT & Multi & 2.5k & - & CC BY 4.0 \\ \hline %Bidir-CPC, Audio2Vec
    AVSpeech & PT & Multi & 3.1k & - & CC BY 4.0 \\ \hline %Bidir-CPC
    Fisher & PT & EN & 2k/1k \cite{jiang2021further}  & - & Linguistic Data Consortium (LDC) \\ \hline % MPC

    Alexa-10k & PT & EN & 10k & - & Not released \\ \hline % wav2vec-c
    Didi Callcenter & PT & ZH & 10k & - & Not released \\ \hline % MPC
    Didi Dictation & PT & ZH & 10k & - & Not released \\ \hline % MPC

    LibriSpeech (LS) & PT/EV & EN & 960 & ASR/PR/PC/SID & CC BY 4.0 \\ \hline
    Wall Street Journal (WSJ) & PT/EV & EN & 81 & ASR/PR/PC/SID & Linguistic Data Consortium (LDC)  \\ \hline

    Common Voice (CV-dataset) & PT/EV & Multi & 11k/7k \cite{babu2021xlsr}/430 \cite{kawakami2020learning} & ASR/PR/PC & CC0 \\ \hline % mention the split, CPC-modified, Bidir-CPC
    Multilingual LS (MLS) & PT/EV & Multi & 50k & ASR & CC BY 4.0 \\ \hline
    VoxPopuli (VP) & PT/EV & Multi & 400k \cite{babu2021xlsr}/24k \cite{chen2021unispeechsat, chen2021wavlm} & ASR & CC0 \\ \hline
    BABEL (BBL) & PT/EV & Multi & 1k & ASR & IARPA Babel Agreement  \\ \hline

    GigaSpeech & PT/EV & EN & 40k/10k \cite{chen2021unispeechsat, chen2021wavlm} & ASR & Apache-2.0 License \\ \hline
    TED-LIUM 3 (TED3) & PT/EV & EN & 450 & ASR & CC BY-NC-ND 3.0 \\ \hline %Bidir-CPC
    TED-LIUM 2 (TED2) & PT/EV & EN & 118 & ASR & CC BY-NC-ND 3.0  \\ \hline % speech-XLNet
    Switchboard (SWB) & PT/EV & EN & 260 & ASR & Linguistic Data Consortium (LDC) \\ \hline % MPC, Bidir-CPC
    TIMIT & PT/EV & EN & 4 & ASR/PR/PC & Linguistic Data Consortium (LDC) \\ \hline
    VoxLingua107 & PT/EV & Multi & 6.6k & LID & CC BY 4.0 \\ \hline

    Open Mandarin & PT/EV & ZH & 1.5k & ASR & \makecell[l]{CC BY-NC-ND 4.0, Apache License v.2.0,\\ Linguistic Data Consortium (LDC)} \\ \hline % MPC
    HKUST & PT/EV & ZH & 168/200 & ASR & Linguistic Data Consortium (LDC) \\ \hline % MPC
    AISHELL-1 & PT/EV & ZH & 178 & ASR & Apache License v.2.0 \\ \hline % MPC
    Hub5'00 & EV & EN & 13 & ASR & Linguistic Data Consortium (LDC) \\ \hline % MPC - https://arxiv.org/pdf/1801.00059.pdf
    DIRHA & EV & EN & 11 & ASR & See link for details\textsuperscript{\ref{footnote:DIRHA_license}} \\ \hline % PASE
    % from the paper Learning Problem-agnostic Speech Representations from Multiple Self-supervised Tasks - "in Section 4.3 we use the DIRHA dataset [36]. Training and validation sets are based on the original WSJ-5k corpus (consisting of 7138 sentences uttered by 83 speakers) that is contaminated with a set of impulse re- sponses measured in a real apartment. The test set is composed of 409 WSJ sentences uttered by six American speakers and is based on real recordings in a domestic environment with a reverberation time of 0.7 s and an average signal-to-noise ratio of about 10 dB"
    CHiME-5 & EV & EN & 50 & ASR & See link for details\textsuperscript{\ref{footnote:CHiME_license}}\\ \hline % PASE+
    Alexa-eval & EV & EN & 1k & ASR & Not released \\ \hline % wav2vec-c
    INTERFACE & EV & Multi & 16 & Sentiment & No information \\ \hline % PASE
    MOSEI & EV & EN & 65 & Sentiment & See link for details\textsuperscript{\ref{footnote:MOSEI_license}}\\ \hline % MockingJay
    VCTK & EV & EN & 44 & SID/ASV & CC BY 4.0 \\ \hline % PASE
    VoxCeleb1 & EV & Multi & 352 & SID/ASV & CC BY 4.0 \\ \hline
    Fluent Speech Commands (FSC) & EV & EN & 14.7 & IC & CC BY-NC-ND 4.0 \\ \hline
    QUESST 2014 (QUESST) & EV & Multi & 23 & QbE & No information \\ \hline
    LS En-Fr & EV & En-Fr & 236 & ST & CC BY 4.0 \\ \hline % APC (tera-9), https://github.com/alicank/Translation-Augmented-LibriSpeech-Corpus
    CoVoST-2 & EV & Multi & 2.9k & ST & CC0 \\ \hline

    ALFFA & EV & Multi & 5.2--18.3 & ASR-multi & MIT License \\ \hline %Bidir-CPC
    OpenSLR-multi & EV & Multi & 4.4--265.9 & ASR-multi & \makecell[l]{CC BY-SA 3.0 US, CC BY-SA 4.0, CC BY 4.0, \\ CC BY-NC-ND 4.0, Apache License v.2.0}\\ \hline %Bidir-CPC
    AED datasets & EV & - & - & AED & \makecell[l]{CC BY 4.0 (MUSAN, Speech Commands, NSynth, Bird Audio\\ Detection),  CC0 (Spoken Language Identification),\\ Non-Commercial (TUT)} \\
    \bottomrule
  \end{tabular}}
\end{table*}

\addtocounter{footnote}{1}
\footnotetext{\label{footnote:DIRHA_license}\url{https://dirha.fbk.eu/node/107}}
\addtocounter{footnote}{1}
\footnotetext{\label{footnote:CHiME_license}\url{https://chimechallenge.github.io/chime6/download.html}}
\addtocounter{footnote}{1}
\footnotetext{\label{footnote:MOSEI_license}\url{https://github.com/A2Zadeh/CMU-MultimodalSDK/blob/master/LICENSE.txt}}

\begin{table*}[ht]
  \centering
  \scriptsize
  \caption{A summary of common experiment settings for various SSL
  evaluations (Part 1). Networks are usually pre-trained with SSL
  techniques, augmented with prediction heads, and fine-tuned (or trained) with
  labeled data in downstream tasks for benchmarking. The 
  \textbf{Pre-training corpus}, \textbf{Training (fine-tuning)}, and
  \textbf{Test} columns list the datasets used in each work, and the \textbf{Task}
  column lists the tasks performed in the corresponding papers. We follow
  the abbreviation introduced in \cref{table:datasets}. The \textbf{Transfer}
  column indicates whether the SSL technique is evaluated by its capability for
  transfer learning, i.e., different datasets are utilized for pre-training and
  fine-tuning. The \textbf{Fine-tuning labels used} column summarizes the
  amount of labeled examples used in downstream fine-tuning.}
  \label{table:method_data_setting}
  {%\tabulinesep=1mm
  \renewcommand*\arraystretch{1.4}
  \begin{tabular}{p{0.12\textwidth}p{0.12\textwidth}p{0.06\textwidth}p{0.13\textwidth}p{0.13\textwidth}p{0.06\textwidth}p{0.20\textwidth}}
    \toprule
    \multirow{2}{*}{Work} & \multirow{2}{*}{\makecell[l]{Pre-training\\ corpus}} & \multirow{2}{*}{Task} & \multicolumn{2}{c}{Dataset} & \multirow{2}{*}{Transfer} & \multirow{2}{*}{Fine-tuning labels used} \\ \cline{4-5}
     &  &  & Training (fine-tuning) & Test & & \\
    \midrule

    % CPC
    \multirow{2}{*}[0mm]{CPC \cite{oord2018representation}} & \multirow{2}{*}[0mm]{LS 100 hrs} & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
    & & SID & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline

    % PASE
    \multirow{2}{*}[-5mm]{PASE\cite{pascual2019learning}} & \multirow{2}{*}[-5mm]{LS 50 hrs \cite{ravanelli2018learning}} & SID & VCTK\textsuperscript{\ref{footnote:official-split}} & VCTK\textsuperscript{\ref{footnote:official-split}} & \checkmark & 44 hrs \\ \cline{3-7}
    & & Sentiment & INTERFACE\textsuperscript{\ref{footnote:interface-split}} & INTERFACE\textsuperscript{\ref{footnote:interface-split}} & \checkmark & 3 hrs \\ \cline{3-7}
    & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
    & & ASR & DIRHA\textsuperscript{\ref{footnote:official-split}} & DIRHA\textsuperscript{\ref{footnote:official-split}} & \checkmark & 11 hrs \\ \hline
    % Audio2Vec
    Audio2Vec \cite{tagliasacchi2020pre} & AudioSet & AED & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & \checkmark & See \cite{tagliasacchi2020pre} for details \\ \hline
    % APC
    \multirow{3}{*}[0mm]{APC \cite{chung2019unsupervised, chung2020generative}} & \multirow{3}{*}[0mm]{LS 360 hrs} & ASR & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & WSJ dev93 & \checkmark & 72 hrs \\ \cline{3-7}
    & & ST & LS En-Fr\textsuperscript{\ref{footnote:official-split}} & LS En-Fr\textsuperscript{\ref{footnote:official-split}} & - & 236 hrs \\ \cline{3-7}
    & & SID & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & \checkmark & 65\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline
    % wav2vec
    \multirow{2}{*}[0mm]{wav2vec \cite{schneider2019wav2vec}} & \multirow{2}{*}[-0.5mm]{\makecell[l]{LS 80/960 hrs,\\ LS 960 hrs \\ \:+ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & \checkmark & 81 hrs\vspace{1mm} \\ \cline{3-7}
    & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hr\vspace{1mm} \\ \hline
    % Phase predict
    PhasePredict \cite{quitry2019learning} & AudioSet & AED & 7 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & 7 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & \checkmark & See \cite{quitry2019learning} for details \\ \hline
    % Bidir-CPC
    \multirow{3}{*}{Bidir-CPC\cite{kawakami2020learning}} & \multirow{3}{*}{\makecell[l]{LS 960 hrs,\\ CPC-8k\textsuperscript{\ref{footnote:CPC-8k}}}} & ASR & \makecell[l]{WSJ si284,\\LS 960 hrs,\\TED3\textsuperscript{\ref{footnote:official-split}}} & \makecell[l]{WSJ eval92,\\LS test-clean,\\LS test-other,\\TED3\textsuperscript{\ref{footnote:official-split}}, SWB\textsuperscript{\ref{footnote:official-split}}}  & \makecell[l]{Different\\ datasets for\\ training\\ and test} & 81/960/450 hrs \\ \cline{3-7} % cross domain transferring, SWB eval2000, TED3 training set size = 450 hrs
    & & \makecell[l]{ASR-multi} & ALFFA\textsuperscript{\ref{footnote:official-split}} & ALFFA\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 languages, 5.2--18.3 hrs\\ \cline{3-7}
    & & \makecell[l]{ASR-multi} & OpenSLR-multi\textsuperscript{\ref{footnote:official-split}} & OpenSLR-multi\textsuperscript{\ref{footnote:official-split}} & \checkmark & 21 languages, 4.4--265.9 hrs \\ \hline
    % vq-VAE

    % MockingJay
    \multirow{3}{*}{MockingJay \cite{liu2020mockingjay}} & \multirow{3}{*}{\makecell[l]{LS 360 hrs}} & PC & LS 360 hrs & LS test-clean & - & 0.36\textsuperscript{\ref{footnote:low-resource}}/1.8\textsuperscript{\ref{footnote:low-resource}}/3.6\textsuperscript{\ref{footnote:low-resource}}/18\textsuperscript{\ref{footnote:low-resource}}/45\textsuperscript{\ref{footnote:low-resource}}/360  hrs \\ \cline{3-7}
    & & SID & LS 100 hrs\textsuperscript{\ref{footnote:91-split-test}} & LS 100 hrs\textsuperscript{\ref{footnote:91-split-test}} & - & 90\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
    & & Sentiment & MOSEI\textsuperscript{\ref{footnote:official-split}} & MOSEI\textsuperscript{\ref{footnote:official-split}} & \checkmark & 65 hrs \\ \hline

    % CPC modified
    \multirow{2}{*}{CPC modified \cite{riviere2020unsupervised}} & LS 100 hrs & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & \makecell[l]{LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}}} & - & \makecell[l]{80\textsuperscript{\ref{footnote:low-resource}} hrs} \\ \cline{2-7}
    & \makecell[l]{LS 100 hrs,\\LS 960 hrs,\\LL 60k hrs} & PC & CV-dataset\textsuperscript{\ref{footnote:official-split}} & CV-dataset\textsuperscript{\ref{footnote:official-split}} & \checkmark & 1 hrs \\ \hline
    % vq-wav2vec
    \multirow{2}{*}[0mm]{vq-wav2vec \cite{Baevski2020vq-wav2vec}} & \multirow{2}{*}[0mm]{LS 960 hrs} & ASR & WSJ si284 & \makecell[l]{WSJ eval92} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
    & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \hline
    % DeCoAR
    \multirow{2}{*}[0mm]{DeCoAR \cite{ling2020deep}} & \multirow{2}{*}[0mm]{\makecell[l]{LS 100/360/460/\\\:\:\:960 hrs,\\ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{25\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}}/81 hrs} \\ \cline{3-7}
    & & ASR & \makecell[l]{LS 100/360/\\\:\:\:460/960 hrs} & \makecell[l]{LS test-clean,\\ LS test-other} & - & 100/360/460/960 hrs \\ \hline
    % MT-APC
    \multirow{2}{*}[0mm]{MT-APC\cite{chung2020improved}} & \multirow{2}{*}[0mm]{LS 360 hrs} & ASR & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & WSJ dev93 & \checkmark & 72 hrs \\ \cline{3-7}
    & & ST & LS En-Fr\textsuperscript{\ref{footnote:official-split}} & LS En-Fr\textsuperscript{\ref{footnote:official-split}} & - & 236 hrs \\ \hline
    % PASE+
    \multirow{3}{*}[0mm]{PASE+\cite{ravanelli2020multi}} & \multirow{3}{*}[0mm]{LS 50 hrs \cite{ravanelli2018learning}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
    & & ASR & DIRHA\textsuperscript{\ref{footnote:official-split}} & DIRHA\textsuperscript{\ref{footnote:official-split}} & \checkmark & 11 hrs \\ \cline{3-7}
    & & ASR & CHiME-5\textsuperscript{\ref{footnote:official-split}} & CHiME-5\textsuperscript{\ref{footnote:official-split}} & \checkmark & 50 hrs \\ \hline

    % AALBERT
    \multirow{2}{*}[0mm]{AALBERT \cite{chi2020audio}} & \multirow{2}{*}[0mm]{\makecell[l]{LS 360 hrs}} & PC & LS 100 hrs\textsuperscript{\ref{footnote:APC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:APC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
    & & SID & LS 360 hrs\textsuperscript{\ref{footnote:APC-split}} & LS 360 hrs\textsuperscript{\ref{footnote:APC-split}} & - & 288\textsuperscript{\ref{footnote:low-resource}} hrs \\
    \bottomrule
  \end{tabular}}
\end{table*}

\begin{table*}[ht]
  \centering
  \scriptsize
  \caption{A summary of common experiment settings for various SSL
  evaluations (Part 2). See the caption of \cref{table:method_data_setting} for a
  detailed description of all the abbreviations used in this table.}
  \label{table:method_data_setting_2}
  {%\tabulinesep=1mm
  \renewcommand*\arraystretch{1.4}
  \begin{tabular}{p{0.12\textwidth}p{0.12\textwidth}p{0.05\textwidth}p{0.18\textwidth}p{0.18\textwidth}p{0.05\textwidth}p{0.15\textwidth}}
    \toprule
    \multirow{2}{*}{Work} & \multirow{2}{*}{\makecell[l]{Pre-training\\ corpus}} & \multirow{2}{*}{Task} & \multicolumn{2}{c}{Dataset} & \multirow{2}{*}{Transfer} & \multirow{2}{*}{Fine-tuning labels used} \\ \cline{4-5}
     &  &  & Training (fine-tuning) & Test & & \\
    \midrule

    % BMR
    \multirow{2}{*}[0mm]{BMR \cite{wang2020unsupervised}} & \multirow{2}{*}[0mm]{\makecell[l]{WSJ si284,\\LS 960 hrs}} & ASR & WSJ si284 & WSJ eval92 & - & 81 hrs \\ \cline{3-7}
    & & PR & WSJ si84/si284 & WSJ dev93 & - & 15/81 hrs \\ \hline

    % vq-APC
    \multirow{2}{*}[0mm]{vq-APC \cite{chung20e_interspeech}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
    & & SID & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & \checkmark & 65\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline

    % vq-wav2vec + Discrete BERT
    \makecell[l]{vq-wav2vec + \\\:DiscreteBERT \cite{baevski2019effectiveness}} & LS 960 hrs & ASR & LS 100 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline

    % speech-XLNet
    \multirow{2}{*}[-1mm]{speech-XLNet \cite{song20d_interspeech}} & \multirow{2}{*}[-1mm]{\makecell[l]{LS 960 hrs\\\;\;\;+ WSJ si284\\\;\;\;+ TED2}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs\vspace{1mm} \\ \cline{3-7}
    & & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{7\textsuperscript{\ref{footnote:low-resource}}/14\textsuperscript{\ref{footnote:low-resource}}/30\textsuperscript{\ref{footnote:low-resource}}/81 hrs}\vspace{1mm} \\ \hline

    % MPC
    \multirow{3}{*}[0mm]{MPC \cite{jiang2019improving, jiang2021further}} & \multirow{2}{*}[0mm]{\makecell[l]{Didi Callcenter,\\Didi Dictation,\\Open Mandarin}} & ASR-zh & HKUST\textsuperscript{\ref{footnote:official-split}} & HKUST\textsuperscript{\ref{footnote:official-split}} & \checkmark & 168 hrs \\ \cline{3-7}
    & & ASR-zh & AISHELL-1\textsuperscript{\ref{footnote:official-split}} & AISHELL-1\textsuperscript{\ref{footnote:official-split}} & \checkmark & 178 hrs\vspace{1mm} \\ \cline{2-7}
    & \multirow{1}{*}[-0.5mm]{\makecell[l]{SWB, Fisher 1k,\\LS 960 hrs}} & ASR & SWB & Hub5'00 & - & 260 hrs \vspace{2.5mm} \\ \hline

    % MPE
    \multirow{2}{*}[0mm]{MPE \cite{liu2020masked}} & \multirow{2}{*}[0mm]{\makecell[l]{WSJ si284,\\LS 960 hrs}} & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{25\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}}/81 hrs} \\ \cline{3-7}
    & & ASR & LS 100/360/960 hrs & LS test-clean & - & 100/360/960 hrs \\ \hline

    % ConvDMM
    ConvDMM \cite{khurana20_interspeech} & \makecell[l]{LS 50\cite{ravanelli2018learning}/360/\\\:\:\:960 hrs} & PC/PR & WSJ si284 & WSJ eval92 & \checkmark & \makecell[l]{5\textsuperscript{\ref{footnote:low-resource}}/50\textsuperscript{\ref{footnote:low-resource}}/100\textsuperscript{\ref{footnote:low-resource}} mins,\\ 4\textsuperscript{\ref{footnote:low-resource}}/8\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}} hrs} \\ \hline
    % wav2vec 2.0
    \multirow{2}{*}{wav2vec 2.0 \cite{baevski2020wav2vec}} & \multirow{2}{*}{\makecell[l]{LS 960 hrs,\\ LL 60k hrs}} & ASR & LS 960 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \cline{3-7}
    & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \hline
    % NPC
    \multirow{2}{*}[0mm]{NPC \cite{liu21l_interspeech}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
    & & SID & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & \checkmark & 65\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline
    % DeCoAR 2.0
    DeCoAR 2.0 \cite{ling2020decoar} & LS 960 hrs & ASR & \makecell[l]{LS 100 hrs} & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline

    % TERA
    \multirow{4}{*}[0mm]{TERA \cite{liu2021tera}} & \multirow{4}{*}[0mm]{\makecell[l]{LS 100/360/\\\:\:\:960 hrs}} & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
    & & SID & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
    & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
    & & ASR & LS 100 hrs & LS test-clean & - & 100 hrs \\ \hline
    % HuBERT
    HuBERT \cite{hsu2021hubert} & \makecell[l]{LS 960 hrs,\\ LL 60k hrs} & ASR & LS 960 hrs & \makecell[l]{LS test-clean\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \hline
    %wav2vec-c
    wav2vec-c \cite{sadhu21_interspeech} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
    UniSpeech-SAT \cite{chen2021unispeechsat} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See SUPERB \cite{yang21c_interspeech} paper for details \\ \hline
    WavLM \cite{chen2021wavlm} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See \cite{yang21c_interspeech} for details \\ \hline
    \multirow{4}{*}[0mm]{XLS-R \cite{babu2021xlsr}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;\;+ CV-dataset-7k + VL\\\;\;\;+ BBL}} & ASR & VP, MLS, CV-dataset, BBL, LS & VP, MLS, CV-dataset, BBL, LS & - & \multirow{4}{*}[0mm]{See \cite{babu2021xlsr} for details} \\ \cline{3-6}
    & & SID & VoxCeleb1 & VoxCeleb1 & \checkmark & \\ \cline{3-6}
    & & ST & CoVoST-2 & CoVoST-2 & \checkmark & \\ \cline{3-6}
    & & LID & VL & VL & - & \\

    % more & & & & & & \\ \hline
    % https://tex.stackexchange.com/questions/66596/vertical-alignment-in-multirow-using-cells-with-1-lines
    % https://texblog.org/2019/06/03/control-the-width-of-table-columns-tabular-in-latex/
    %\multirow{2}{*}{wav2vec 2.0} & \multirow{2}{*}{wav2vec 2.0} & Population Population Population Population & 2 & 3 & 4 \\\cline{3-6}
    %http://www.emerson.emory.edu/services/latex/latex_119.html
    %& &2016&2017& 5 & 6 \\ \hline
    \bottomrule
  \end{tabular}}
\end{table*}

\addtocounter{footnote}{1}
\footnotetext{\label{footnote:CPC-split}
Train/test split made available
by \cite{oord2018representation} on Google drive
https://drive.google.com/drive/folders/1BhJ2umKH3whguxMwifaKtSra0TgAbtfb.}
\addtocounter{footnote}{1}
\footnotetext{\label{footnote:official-split}
Utilizes official training or test split.}

\addtocounter{footnote}{1}
\footnotetext{\label{footnote:interface-split}
English utterances used in experiments. The utterances correspond to
approximately 3~hours for training, 40~minutes for development, and 30~minutes
for testing.}

\addtocounter{footnote}{1}
\footnotetext{\label{footnote:AED-datasets}
The 6 AED datasets used in
\cite{tagliasacchi2020pre} are MUSAN~\cite{snyder2015musan}, Bird Audio
Detection~\cite{stowell2019automatic}, Speech Commands~\cite{warden2018speech},
Spoken Language Identification~\cite{oponowiczspoken}, TUT Urban Acoustic
Scenes 2018~\cite{dcase2018} plus an SID task built with LS
\textit{train-clean-100}. In addition to the 6 datasets,
\cite{quitry2019learning} use the NSynth dataset~\cite{engel2017neural} for
evaluation.}
\addtocounter{footnote}{1}
\footnotetext{\label{footnote:CPC-8k}
A collection of AudioSet, AVSpeech,
CV-dataset, LS, WSJ, TIMIT, Speech Accent Archive 
(SSA)~\cite{weinberger2009towards}, TED3, and SWB. SSA is a growing annotated corpus of
English speech with various accents. Among the papers studied in this review, SSA
is used in \cite{kawakami2020learning} only for pre-training, and only 1 hour
of audio is utilized. Thus, we exclude it from our discussion in
\cref{section:benchmark}.}
\addtocounter{footnote}{1}
\footnotetext{\label{footnote:low-resource}
A subset of the official training
split is sampled, usually to mimic low-resource learning conditions or to
quickly evaluate for training and testing on the same split but disjoint
subsets.}
\addtocounter{footnote}{1}
\footnotetext{\label{footnote:APC-split}
Dataset split into training,
validation, and test subsets at a ratio of 8:1:1.}
\addtocounter{footnote}{1}
\footnotetext{\label{footnote:91-split-dev}
Dataset split into training
and validation subsets at a ratio of 9:1.}
\addtocounter{footnote}{1}
\footnotetext{\label{footnote:91-split-test}
The dataset split into training
and test subsets at a ratio of 9:1.}

\subsection{Experiment settings for evaluating SSL techniques}
A common way to benchmark SSL techniques and show their efficacy is to
fine-tune a pre-trained SSL model for a supervised downstream task.
% taking a network pre-trained with SSL as initialization, and then fine-tuning and evaluating the network with labeled data in downstream tasks.
Depending on the corpora used in pre-training and fine-tuning, techniques can
be benchmarked in terms of their capability to transfer knowledge across datasets
(i.e., using pre-training corpora that differ from the fine-tuning ones), their
benefit when training with limited labeled examples (i.e., sampling a subset of
labeled examples for fine-tuning), or their improvement over a fully supervised
baseline (i.e., using the entire training split of downstream datasets for
fine-tuning). \cref{table:method_data_setting,table:method_data_setting_2} summarize experiment settings used in the
SSL literature, including the pre-training corpora, downstream tasks and datasets,
and the amount of fine-tuning labels used, which indicates the targeted
benchmarking scenario as discussed above. Note that there are a variety of ways
to fine-tune pre-trained networks (e.g., fine-tune the entire network,
freeze certain layers during fine-tuning, and add various architectures of
prediction layers to pre-trained networks). We here omit
descriptions of these choices; readers can consult the
original publications for details.

As observed in \cref{table:method_data_setting,table:method_data_setting_2}, LS and WSJ are the most commonly used
pre-training corpora. At the same time, we observe a
growing industry investment in pre-training with larger datasets, e.g.,
CPC-8k (8k hours) for Bidir-CPC~\cite{kawakami2020learning}, LL (60k hours) for
CPC modified~\cite{riviere2020unsupervised}, 
wav2vec 2.0~\cite{baevski2020wav2vec}, and HuBERT~\cite{hsu2021hubert}, Alexa internal
datasets (10k hours) for wav2vec-c~\cite{sadhu21_interspeech}, Didi internal
datasets (10k hours) for MPC~\cite{jiang2019improving, jiang2021further}, the
combination of Gigaspeech, VP-24k, and LL (94k hours in total) for
UniSpeech-SAT~\cite{chen2021unispeechsat} and WavLM~\cite{chen2021wavlm}, and
the combination of VP-400K, MLS, CV-dataset, VL and BBL (436k hours in total)
for XLS-R~\cite{babu2021xlsr}. We expect this trend to continue with the growth
in available computing power. Most studies focuses on learning representations
for English, whereas Chinese~\cite{jiang2019improving, jiang2021further} and
multilinguality~\cite{kawakami2020learning, babu2021xlsr} are also gaining
attention. Compared to pre-training, datasets used for fine-tuning are more
diverse and cover downstream tasks as varied as ASR, PR, PC, SID, AED,
Sentiment, ST, and LID. For benchmarking training scenarios covering full
supervision as well as limited resources, the amount of labeled examples used for
fine-tuning also varies from several minutes up to 1k hours. Recent
benchmarks such as SUPERB~\cite{yang21c_interspeech} that consolidate multiple 
downstream tasks have gained attention for evaluating SSL
methodologies~\cite{chen2021unispeechsat, chen2021wavlm}. The goal of such benchmarks is
to provide a holistic evaluation of the performance of learned representations; 
we discuss these in detail in \cref{sec:benchmark}. With the increasing
popularity of SSL research, we expect future experiment settings to proliferate
and cover more languages, downstream tasks, and pre-training/fine-tuning
datasets.

\begin{figure*}
\centering
%\includegraphics[width=1.0\textwidth]{paper_review/ls-test-clean-other.pdf}
\includegraphics[width=0.9\textwidth]{paper_review/ls-test-clean-v2.pdf}
\caption{SSL performance on ASR WER (vertical axis) evaluated with
LS \textit{test-clean} split. Techniques are sorted based on the number of
model parameters along the horizontal axis. Markers in blue correspond to models initialized with various SSL
techniques and then fine-tuned using 960, 100, 10, 1 hour(s), and 10 minutes
respectively. The 960-hour training set is the aggregation of
\textit{train-clean-100}, \textit{train-clean-360}, and
\textit{train-other-500} splits. The 100-, 10-, 1-hour, and 10-minute sets
leverage \textit{train-clean-100} or its sampling, except for Bidir-CPC, which
samples 10\% of the training examples from the entire 960-hour corpus. For simplicity,
several SSL techniques are appended with suffixes \textit{B}, \textit{L},
\textit{XL}, or \textit{XXL} indicating the \textit{Base}, \textit{Large},
\textit{X-Large}, or \textit{XX-Large} variants specified in the original
publication. We also compare with baselines including the log mel filterbank (fBANK) and semi-supervised, self-training
approaches (iterative pseudo labeling (IPL)~\cite{xu2020iterative}, 
slimIPL~\cite{likhomanenko2020slimipl}, noisy student~\cite{park2020improved}). These approaches are visualized in black. Also, note that the current state of the art---conformer XXL + noisy student~\cite{zhang2020pushing}---is a combination of self-training and SSL techniques. Given the
diversity of the listed methods in experiment settings (e.g., pre-training corpora
and objectives, whether a language model is used in decoding, whether model
parameters are frozen in fine-tuning), readers should be careful that the superiority
of methods cannot be decided only based on lower WER numbers.}
\label{fig:asr_result_no_limit}
\end{figure*}

\subsection{Benchmark results and discussion} \label{sec:benchmark} 
Given the diversity of datasets and downstream tasks used to evaluate SSL
techniques in the literature, it is infeasible to discuss all 
experiment settings in this survey. Hence, due to their wide adoption for
experiments conducted by studies in both SSL and the speech community in
general, we focus first on ASR on the LS dataset to understand the
efficacy of SSL. We examine SSL techniques which report ASR results on the LS
\textit{test-clean} split, and summarize the published WER in
\cref{fig:asr_result_no_limit}. The ASR models were obtained first by using
unlabeled speech to pre-train a model with each SSL technique. The model was
then fine-tuned on labeled data by utilizing a supervised training objective.
Respectively, 960, 100, 10, 1 hour(s), and 10 minutes of labeled LS training data
were used for fine-tuning, as indicated in different panels of
\cref{fig:asr_result_no_limit} (see
the caption of \cref{fig:asr_result_no_limit} for more details).
Semi-supervised methods such as self-training, where a model is first trained
on labeled data to annotate unlabeled speech, and then subsequently trained on
combined golden and self-annotated label-speech pairs, are gaining popularity
in the speech community and have yielded competitive results. For comparison, we also
show performance from such methods (iterative pseudo labeling 
(IPL)~\cite{xu2020iterative}, slimIPL~\cite{likhomanenko2020slimipl}, noisy 
student~\cite{park2020improved}), as well as the current state of the art---conformer XXL + noisy
student~\cite{zhang2020pushing}---which augments SSL with various advanced
techniques including self-training. Furthermore, we illustrate in the figure
the performance of a baseline system \cite{yang21c_interspeech} based on log mel filterbank (fBANK), which is one of the most commonly used features designed by domain experts.
As observed in the figure, most SSL techniques outperform fBANK
features, and with the growing investment in model size, better performance is
achieved. The largest ones, such as wav2vec 2.0-L and HuBERT-L/XL, yield
% results competitive to the state of the art 
  competitive results                         % AMH: check
when the entire 960-hour of labeled data is used in
training/fine-tuning. The benefit of SSL, especially models with more parameters
like wav2vec 2.0 and HuBERT, becomes more evident when the labeling resources
become scarce. Compared to popular semi-supervised methods such as IPL,
slimIPL, and noisy student using 100 hours of labels, wav2vec~2.0 and HuBERT
achieve lower or competitive WERs with 1 hour or even 10 minutes of labeled
examples. The results are highly favorable for low-resource use cases, for instance when
expanding systems to new domains or languages for which large amounts of unlabeled
audio are available, since collecting labels for new conditions is often prohibitively
slow or costly.
% discuss the trend
%[Pick representative tasks and datasets and show the benefit of SSL, e.g., fewer fine-tuning/labeled data is required to achieve better/comparable performance] [talk about trend as well]  The plot suggests that ASR results have been increased significantly with the progress in SSL. Near SOTA performance can be achieved with much fewer training examples.

In addition to the ASR task, where the current state of the art is achieved by a method
combining SSL pre-training and self-training 
techniques~\cite{zhang2020pushing}, SSL models 
% approach the state of the art 
  are competitive                % AMH: check
in other tasks, including IC,
SID, ASV, and QbE. We summarize the performance of these models and previous
non-SSL methods in \cref{table:sota_performance}. The results suggest that the
benefit of SSL is generalizable among tasks that require encoding 
information such as content, speaker, and semantics. As SSL research
gains more attention, we expect that SSL pre-trained models will 
achieve state-of-the-art results on an increasing number of tasks.

\begin{table}[ht]
  \centering
  \footnotesize
  \caption{Tasks where the state of the art is models with SSL pre-training.}
  \label{table:sota_performance}
  \renewcommand*\arraystretch{1.2}
  \begin{tabular}{llllll}  
    \toprule
    Tasks & Dataset & non-SSL & SSL \\
    \midrule
    ASR (WER $\downarrow$) & LS test-clean/other & 2.1/4.0 \cite{xu2020iterative} & 1.4/2.6 \cite{zhang2020pushing} \\ \hline
    IC (Acc $\uparrow$) & FSC & 98.8 \cite{lugosch19_interspeech} & 99.3\cite{chen2021unispeechsat} \\ \hline
    SID (Acc $\uparrow$) & VoxCeleb1 & 94.8 \cite{hajibabaei2018unified} & 95.5 \cite{chen2021wavlm} \\ \hline
    ASV (EER $\downarrow$) & VoxCeleb1 & 3.1 \cite{hajavi2021siamese} & 2.4 \cite{wang2021fine} \\ \hline
    QbE (MTWV $\uparrow$) & QUESST (EN) & 10.6 \cite{rodriguez2014gtts} & 11.2\cite{chen2021unispeechsat} \\

    %wav2vec-c
    %wav2vec-c \cite{sadhu21_interspeech} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
    %UniSpeech-SAT \cite{chen2021unispeechsat} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & Check SUPERB \cite{yang21c_interspeech} paper for details \\ \hline
    %WavLM \cite{chen2021wavlm} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & Check \cite{yang21c_interspeech} for details \\ \hline
    %\multirow{4}{*}[0mm]{XLS-R \cite{babu2021xlsr}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;\;+ CV-dataset-7k + VL\\\;\;\;+ BBL}} & ASR & VP, MLS, CV-dataset, BBL, LS & VP, MLS, CV-dataset, BBL, LS & - & \multirow{4}{*}[0mm]{Check \cite{babu2021xlsr} for details} \\ \cline{3-6}
    %& & SID & VoxCeleb1 & VoxCeleb1 & \checkmark & \\ \cline{3-6}
    %& & ST & CoVoST-2 & CoVoST-2 & \checkmark & \\ \cline{3-6}
    %& & LID & VL & VL & - & \\ \hline

    \bottomrule
  \end{tabular}
\end{table}

Despite the obvious trend of increasing performance as more parameters and SSL
pre-training data are being used, numbers in \cref{fig:asr_result_no_limit} 
and \cref{table:sota_performance} are less comparable than might be expected.
The task performance is obtained from the original papers and is often
achieved with different downstream fine-tuning recipes, including various
language models (used in the ASR system), prediction heads (networks added to
SSL for downstream inference), or choices between fine-tuning the whole
networks or freezing the SSL encoders. For example, in the ASR task, HuBERT-L
and wav2vec~2.0-L leverage Transformer as their language model, while a 4-gram
language model trained on LS is used in DeCoAR~2.0. The lack of common and
established mechanisms to evaluate SSL techniques in downstream applications
makes it difficult to compare techniques fairly and understand their
capabilities. To address this challenge, there are increasing efforts to establish
benchmarks with shared downstream tasks, datasets, and downstream recipes. Such
efforts include SUPERB~\cite{yang21c_interspeech}, 
LeBenchmark~\cite{evain21_interspeech}, ZeroSpeech~\cite{dunbar2020zero},
HEAR \cite{pmlr-v176-turian22a},
NOSS~\cite{shor20_interspeech}, and HARES~\cite{wang2021towards}.

SUPERB~\cite{yang21c_interspeech} is a benchmarking platform that allows the
SSL community to train, evaluate, and compare speech representations on
diverse downstream speech processing tasks, from acoustic and speaker identity
to paralinguistic and semantics. SUPERB consolidates downstream recipes to
focus on common and straightforward settings (e.g., prediction head
architectures, language models, hyperparameter spaces) to facilitate generalizable
and reproducible benchmarking of SSL techniques. SUPERB also encourages
researchers to innovate for efficient use of model parameters and computation
resources 
to democratize SSL beyond race among Big Tech.   % AMH: what does this mean?
LeBenchmark~\cite{evain21_interspeech} shares a vision similar to SUPERB and provides a
reproducible framework for assessing SSL in French with ASR, spoken language
understanding, speech translation, and emotion recognition. 
ZeroSpeech~\cite{dunbar2020zero} (described in more detail in \cref{zero_speech})
challenges the scientific community to build speech and language understanding
systems using zero expert resources 
%with the ultimate goal of bringing the systems to the services 
for millions of users of ``low-resource" languages.
%or in abnormal condition (dyslexia, autism, etc). ZeroSpeech offers evaluation for independent units in the system as well aggregation of multiple components. 
SSL techniques are also benchmarked with the ZeroSpeech 
challenge~\cite{tjandra20_interspeech, niekerk20b_interspeech}. Apart from the speech
community, researchers have also established HEAR (holistic evaluation of audio
representations) \cite{pmlr-v176-turian22a}, NOSS (non-semantic
speech benchmark)~\cite{shor20_interspeech}, and HARES (holistic audio
representation evaluation suite)~\cite{wang2021towards} to benchmark audio
representations. These efforts promote the creation of an audio embedding
that is as holistic as the human ear in interpreting speech, environmental
sound, and music. Given the significant need to understand and compare SSL techniques
fairly and comprehensively, we expect SSL benchmarking to remain an
active research area.

%SUPERB motivation: consolidating evaluation recipes to understand the capability of various SSL in a fair way. SUPERB designates the architectures of prediction heads for each downstream tasks [common architectures] [frozen] [generalizable, comparable, reproducible results]. [Results on tasks with such constraint conditions (mainly SUPERB but also from literatures imposing similar constraint).] [discussion on the results - competitive performance with simple downstream recipes] [More challenges, ZeroSpeech, HEAR ...]

%Librispeech 100h-960h \\
%Libri-light 10m, 1h, 10h - 60k \\
%SUPERB \\
%Add more
%Hear 2021 Challenge? (focus on audio, not only speech)

### 2023-03-04 chapters/paper_review/04b_analysis.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\section{Analysis of Self-Supervised Representations}
\label{analysis}

% {\color{blue} Katrin}\\
The previous sections have shown how self-supervised learning can result in
powerful representations that provide a robust starting point for several
downstream tasks. It is natural to ask if we can gain an even deeper
understanding of the nature of these representations, in order to further
optimize them or apply them to different problems.
What is the information encoded in these representations? How robust are they
to distributional shifts, and how dependent are they on the size of the
training data? Do they generalize across languages? What are the key
ingredients for training powerful representations: input data, network
architecture, training criterion, or all three? Can we predict their
performance on downstream tasks from their training behavior? This section
tries to answer these questions by summarizing several studies that analyze
self-supervised representations.

\subsection{Information content}

In \cite{pasad2021} wav2vec~2.0 representations were analyzed with
respect to their acoustic-linguistic information content at different
network layers. Three different mechanisms were used for this
purpose. The first of these is canonical correlation analysis (CCA),
which computes similarity scores between two continuous vectors
based on the maximum correlation of their linear projections.  These
can be used to judge the similarity of embeddings at different layers
with each other, with standard acoustic representations such as mel
filterbank features, or word embeddings derived from text. The second
method clusters continuous representation vectors and computes the
discrete mutual information between cluster IDs and phone or word
labels. The third method involves probing tasks: representation
vectors extracted from the network are used to perform simple
downstream tasks, in particular determining whether two acoustic segments
correspond to the same word, and a standard benchmark of 11 word
similarity tasks~\cite{faruqi14}. These are mostly used to gauge the
amount of lexical information present in the embeddings.  Using this
battery of tests the authors compared pre-trained models of varying
sizes as well as models fine-tuned for ASR. They found that pre-trained
models show an autoencoder-style behavior, with early layers showing
strong similarity with input features, intermediate layers diverging
more, and final layers reverting back to higher similarity with input
features and early layers. Generally, the earlier layers in wav2vec~2.0 
models encode acoustic information. The next set of layers encodes
phonetic class information, followed by word meaning information,
before reverting back to encoding phonetic/acoustic information. Thus,
extracting representations from the last layers for tasks that require
phonetic or word-related information may not be the best
strategy. Indeed, the authors of \cite{baevski2021unsupervised} show that a
phone classifier trained on each of the 24 frozen layers of a wav2vec~2.0 model
showed the lowest phone error rates
for layers 10--21 and higher error rates for the other layers. 
\cite{pasad2021} further show that fine-tuning the pre-trained model with a
character-level CTC training criterion changes the behavior of the
last layers (especially the final two layers), breaking the
autoencoder-style behavior and focusing the information encoded in the
last layers on orthographic-phonetic and word information.

The peaking of class-relevant information in intermediate layers seems to be
common across different self-supervised learners and different modalities. In
an analysis of text-based Transformers trained with a masked language model
criterion~\cite{voita2019} observed a similar compression plus reconstruction
pattern. Interestingly, similar network behavior was also recently described
for self-supervised learners in computer vision: using a contrastive
self-supervised learner (SimCLR) that optimizes for augmentation invariance,
\cite{grigg2021self} show that it is the intermediate representations that most
closely approximate information learned in a supervised way, i.e., they
provide more class information than the representations from final layers. This
is similar to the findings described above for wav2vec~2.0 without fine-tuning,
where intermediate layers provide more information about phone and word
classes.

Self-supervised representations may encode other information besides phonetic
classes or words, for example, channel, language, speaker, and sentiment
information. 
\edit{It is shown that the per-utterance mean of CPC features captures speaker information to a large extent\cite{van2021analyzing}.}
Location of information pertaining to speakers vs.\ language classes was
analyzed in \cite{ling20odyssey} for a 12-layer BERTphone model. This model
combines a self-supervised masked reconstruction loss with a phone-based CTC
loss to produce representations	for speaker recognition and language
identification. By	analyzing the weights	of a linear combination	of layer
representations for these two downstream tasks, it was	shown that language
recognition draws on representation	from higher layers (peaking at layer~10)
whereas speaker recognition benefited from layers at positions 6, 9, and 12.
This may indicate that language recognition relies more on higher-level
phonetic information whereas speaker recognition uses a combination of
acoustic and phonetic information. In a recent study~\cite{chen2021wavlm} the
same technique was used to identify layer contributions for the downstream
SUPERB benchmark tasks in the WavLM model. For a smaller model (95M parameters)
it was again confirmed that lower layers encode speaker-related information
necessary for speaker diarization and verification whereas higher layers encode
phonetic and semantic information. Another study~\cite{wang2021layersup} used
explicit self-supervised loss at the intermediate layers rather than just the
output layer of a HuBERT model in order to enforce better learning of phonetic
information. The resulting model was indeed better at downstream tasks
requiring information about phonetic content, such as phone recognition, ASR,
and keyword spotting, but worse at speaker-related tasks like speaker
diarization and verification.

%Hung-yi Lee: My group also has a paper analyzing the attention of self-supervised speech model (https://arxiv.org/abs/2006.03265). I will include some findings in this subsection later.
Most self-supervised learning approaches rely on a Transformer architecture for
the representation model. In \cite{AnalyzeAttention} the attention patterns in
generatively trained Transformer representation models were analyzed.
Self-attention heads were grouped into three categories: diagonal, vertical,
and global. It was found that the diagonal head focuses on neighbors and is
highly correlated with phoneme boundaries, whereas the vertical head focuses on
specific phonemes in the utterance. Global heads were found to be redundant as
removing them resulted in faster inference time and higher performance.

\subsection{Training criterion}
In \cite{chung2021}, representations based on different training criteria
(masked predictive coding, contrastive predictive coding, and autoregressive
predictive coding) were compared  and analyzed with respect to the correlation
between their training loss and performance on both phone discrimination and
speaker classification probing tasks. It was observed that the autoregressive
predictive coding loss showed the strongest correlation with downstream
performance on both tasks; however, models were not further analyzed
internally. An evaluation of the similarity of representations trained
according to the three criteria above (but with different architectures and
directionality of contextual information) also showed that it is the training
criterion that most influences the information encoded in the representations,
not the architecture of the learner or the directionality of the input.

A similar insight was obtained in \cite{zhou2020}, which compared vq-vae and
vq-wav2vec with respect to their ability to discover phonetic units.
The vq-vae model extracts continuous features from the audio signal; a
quantizer then  maps them into a discrete space, and a decoder is trained to
reconstruct the original audio conditioned on the latent discrete
representation and the past acoustic observations. By contrast, vq-wav2vec
predicts future latent discrete representations based on contextualized
embeddings of past discrete representations, in a CPC-style way. The models
were evaluated according to their ability to discover phonetic units (as
measured by phone recognition error rate on TIMIT, and the ZeroSpeech ABX task
(see \cref{sec:zero} for more details)), and it was found that the predictive vq-wav2vec
model fared better than the autoencoder-like vq-vae model, most likely due to
its superior ability to model temporal dynamics.

\subsection{Effects of data and model size}
\label{subsec:modelsize}

How does the performance of self-supervised models change in relation to the
amount of training data, and in relation to the size (number of parameters) of
the model?
Several studies have demonstrated better downstream performance when using
larger datasets~\cite{rivi20, kawakami2020learning, chen2021wavlm}. For
example,
\cite{kawakami2020learning} compared representations learned by a bidirectional
CPC model from the standard 960 hour LS corpus and a corpus of 8,000 hours of
diverse speech from multiple sources.\textsuperscript{\ref{footnote:CPC-8k}}
Not surprisingly, an ASR model trained on top of these representations
performed better when representations were learned from the larger dataset. 
Although the precise relationship between data size and performance has not
been quantified, we can assume that it follows a law of diminishing returns (or
power law), 
similar to observations for most data-intensive machine learning tasks.
In addition to the size of the dataset, the diversity of the data also seems
to play a role, although this was not quantified in this study. However, recent
experiments with larger and more diverse data collections~\cite{chen2021wavlm}
confirm this assumption, as do 
explicit investigations of domain shift robustness (see
\cref{subsec:robustness} below).

\edit{The relation between model sizes and downstream performances have also been investigated \cite{pue2021scaling,versteegh2015zero}.}
Using the Mockingjay 
model~\cite{liu2020mockingjay}, the authors in \cite{pue2021scaling} attempt to establish a relationship between
model size and self-supervised \ensuremath{L_1} loss and demonstrate that it approximately
follows a power law. Model size and accuracy on downstream phone
classification and speaker recognition tasks are positively correlated but do not
exactly follow a power law; rather, the accuracy saturates as models increase in
size, possibly due to the lack of a corresponding expansion in training data
size.

\subsection{Robustness and transferability}
\label{subsec:robustness}

It is well known that traditional speech features like MFCCs lack
robustness against environmental effects such as additive noise,
reverberation, accents, etc., that cause differences in the distributions of
speech features.
Do pre-trained representations offer greater robustness against distributional
shifts? 
One study~\cite{kawakami2020learning} compared pre-trained
representations from a CPC model against MFCCs and 
found pre-trained representations to be more robust to mismatches between
training and test data. The 
training data consisted of clean, read speech (LS)
whereas test data consisted of the Switchboard corpus and TED talks. The
distributional shifts here may stem from both the acoustics (microphone, room
reverberation) as well as
lexical effects related to topic and style, as well as differences in speaker
characteristics such as accent. 
Similar problems were also investigated using HuBERT and wav2vec~2.0 models in
\cite{chang2021exploration}.
In \cite{robustw2v2} domain effects were studied in greater detail using 
datasets from six different domains. In particular, the authors focused on the
usefulness of adding out-of-domain data to pre-training. The general
conclusions are that pre-training on more and diverse domains is preferable:
models pre-trained on more domains performed better than those pre-trained on
fewer when tested on held-out domains, regardless of which additional
labeled data was used for fine-tuning. Adding in-domain unlabeled data---if
available---to pre-training improves performance robustly; however, even
out-of-domain unlabeled data is helpful and closes  66--73\% of the performance
gap between the ideal setting of in-domain labeled data and a competitive
supervised out-of-domain model.

In \cite{rivi20} the effectiveness of CPC-trained representations for
phone discrimination tasks was compared across several languages. It
was found that representations pre-trained only on English successfully
enabled phone discrimination in 10 other languages, rivaling
supervised methods in accuracy in low-data regimes (1h of labeled data
per language). Thus, self-supervised pre-training enables the model to
learn contextualized speech features that generalize across different
languages. In \cite{conneau20xling}, a wav2vec~2.0 model was trained on data
from multiple different languages and different corpora (Babel, Common Voice,
and multilingual LS)
jointly, followed by fine-tuning for each individual language. The largest
model covers 53 languages in total
and consists of 56,000 hours of speech. Compared to monolingual pre-training,
even smaller models trained on only ten languages improve performance
substantially on a downstream character-based ASR task. Low-resource languages
with little labeled data improve the most under this training regime.
Multilingual representations also resulted in competitive performance (lower
character error rate than monolingual representations) for languages not
present in the training dataset, again showing that unsupervised pre-trained
representations can learn generic features of the speech signal that generalize
across different languages. The study also found that sharing data from closely
related languages is more beneficial than combining distant languages. An
analysis of language clusters in the shared discrete latent representation
space revealed that similar languages do indeed show a higher degree of sharing
of discrete tokens. 
Finally, one might ask whether the interpretation of representations extracted
from different layers of a self-supervised models also generalizes to the
multilingual setting. Experiments in \cite{baevski2021unsupervised} on phone
recognition in eight languages based on the different layers of the
multilingual wav2vec~2.0 XLSR-53 model indicate that this is indeed the case:
phone error rates showed the same pattern as in the monolingual (English)
scenario, with lower phone error rates for middle layers as opposed to
earlier/later layers.

### 2023-03-04 chapters/paper_review/05_zero.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\section{From Representation Learning to Zero Resources}
\label{sec:zero}

% {\color{black} Reviewer: TNS, Abdo, Hung-yi Lee }\\
% \kl{We discussed in our last meeting that I would add a pointer in this section to the acoustic word embedding section.  However, I'm not quite sure what the right spot for it is, so waiting for this section to stabilize a bit more first.}

In the SSL framework, speech representations can be
learned and used in various downstream tasks to achieve competitive, robust,
and transferable performance, as shown in
\crefrange{section:benchmark}{analysis}. 
However, labeled data is still required. 
For example, in ASR, utterances and their manual transcriptions are needed to learn downstream models or fine-tune representation models. 
Can a model learn without any labeled data? 
In \cref{secsec:unpaired}, we show how to learn ASR models without any paired audio and text and how SSL improves the framework.
In addition, many languages have no writing system. 
In \cref{subsec:zero}, the SSL representation is further used in scenarios where text data is unavailable.

\subsection{Unpaired text and audio} \label{secsec:unpaired}
% What are we going to do % how to match
%In the self-supervised learning framework, speech representations can be learned and used in the downstream tasks, but some labeled data for finetuning is needed.

\begin{table*}[ht]
\centering
    {\renewcommand*\arraystretch{1.4}}
\caption{
Unsupervised ASR.
TIMIT numbers are phoneme error rates (PER), while the numbers for
LibriSpeech are word error rates (WER).
SWC $=$ spoken word classifier, ST $=$ speech translation.
All speech and text are in English if not specified. 
\textcolor{black}{
The references in the table are sorted according to the date of publication.}
}
\label{fig:unsupervisedASR}
\resizebox{0.99\textwidth}{!}{
\begin{tabular}{cllllll}
    \toprule
Reference & Speech representation & Speech segmentation  & Text token/representation & Mapping approach & Refinement  & Results  \\ 
    \midrule
\cite{liu2018completely}& Audio word2vec~\cite{Wang2018SegAudioWord2Vec} & Oracle & Phoneme & \edit{Adversarial Training~\cite{gulrajani2017improved}} & - & TIMIT (PER): 63.6\%  \\ %Submitted on 1 Apr 2018
  \midrule
  \cite{chung2018unsupervised} & Speech2vec~\cite{chung2018speech2vec} & BES-GMM~\cite{Kamper17BESGMM} &  Word2Vec & \edit{Adversarial Training~\cite{conneau2017word}} & Self-training & SWC (Acc): 10.9\%  \\ %Submitted on 18 May 2018
%https://arxiv.org/pdf/1805.07467.pdf %The text embeddings were obtained by training Word2Vec on the transcriptions using the fastText implementation without subword information [3].
  \midrule

    \cite{chung2019towards} & \tabincell{l}{Speech2vec\\(English)} & Oracle & \tabincell{c}{Word2Vec\\(French)} & VecMap~\cite{artetxe-etal-2018-robust} & \tabincell{l}{LM rescore,\\sequence DAE} & ST (BLEU): 10.8\%   \\ %[Submitted on 4 Nov 2018]
\midrule

\cite{yeh2018unsupervised} & MFCC & GAS~\cite{wang2017gate} & Phoneme & Empirical-ODM~\cite{Liu2017ODM} & Self-training &  TIMIT (PER):  41.6\%  \\ %Submitted on 23 Dec 2018
  \midrule

\cite{chen2019completely} & MFCC & GAS  & Phoneme & \edit{Adversarial Training~\cite{gulrajani2017improved}} & Self-training &  TIMIT (PER):  33.1\%   \\
  \midrule

\cite{baevski2021unsupervised} & Wav2vec 2.0~\cite{baevski2020wav2vec} & \kmeans{} & Phoneme &\edit{Adversarial Training~\cite{gulrajani2017improved}} & Self-training & \tabincell{l}{TIMIT (PER):  18.6\%, \\ LibriSpeech (WER): 5.9\%}   \\
  \midrule

\textcolor{black}{\cite{Deciphering2022}} & \textcolor{black}{\tabincell{l}{Universal Phone \\ Recogniser}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Decipherment~\cite{ravi-knight-2011-deciphering}} & \textcolor{black}{Self-training} & \textcolor{black}{\tabincell{l}{GlobalPhone: 32.5\% to just 1.9\% \\ worse than supervised models} }  \\ %ubmitted on 12 Nov 2021 %m 32.5% to just 1.9% absolute worse than the equivalent fully supervised models trained on the same data
  \midrule
\textcolor{black}{\cite{w2v-u2}} & \textcolor{black}{Wav2vec 2.0~\cite{baevski2020wav2vec} }& - & \textcolor{black}{Phoneme} & \textcolor{black}{Adversarial Training~\cite{gulrajani2017improved}} & \textcolor{black}{Self-training} & \textcolor{black}{LibriSpeech (WER): 6.3\%}   \\
  \midrule

\textcolor{black}{\cite{w2v-u2}} & \textcolor{black}{Wav2vec 2.0~\cite{baevski2020wav2vec}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Adversarial Training~\cite{gulrajani2017improved}} & \textcolor{black}{Self-training} & \textcolor{black}{LJSpeech (WER): 64.0\%}  \\

%e MUSE [7] and VecMap [8], n
%The numbers in the section of unsupervised methods denoted as BLEU score (%) of VecMap /
%BLEU score (%) of MUSE 10.8 / 6.2
%11.3 / 7.3
%https://arxiv.org/pdf/1811.01307.pdf
    \bottomrule
\end{tabular}
}
\end{table*}

\subsubsection{Unsupervised ASR}
% {\color{black} Hung-yi}\\

If only unpaired speech and text are available, that is, the text is not a
manual transcription of speech, can the machine learn how to transcribe speech
into text?
This scenario is called \textit{unsupervised ASR}, and the framework is as
below. 
Given a set of unlabeled utterances $\mathcal{S}=\{S_1, S_2, ..., S_N\}$  and a
set of sentences $\mathcal{Y}=\{Y_1, Y_2, ..., Y_M\}$,\footnote{Note that
the speech and text are not paired, that is, $Y_i$ is not the transcription of
$S_i$.} a mapping function~$F$, which can take an utterance~$S$ as input and
generate its transcription, is learned from data. 
\Cref{fig:unsupervisedASR} summarizes recent work on unsupervised ASR,
including the speech representation used, the algorithm used to learn the
mapping without supervision, and the results. Below, we will discuss these
methods in more detail.

Adversarial training~\cite{goodfellow2014generative,arjovsky2017wasserstein, gulrajani2017improved} is one common way to learn such a 
mapping function. 
The framework includes a discriminator and a generator.
The mapping function~$F$ plays the role of the generator, which takes speech utterances as input and outputs text.
The discriminator learns to distinguish real text from the generated
output; the generator learns to ``fool'' the discriminator.
The generator and the discriminator are trained in an iterative, 
interleaved way. 
After the training, the generator serves as the speech recognition model.
\edit{
There is a large amount of work using gradient penalty in the objective of training discriminators~\cite{liu2018completely,chen2019completely,baevski2021unsupervised,w2v-u2}, which is inspired by Improved Wasserstein Generative Adversarial Network (WGAN)~\cite{gulrajani2017improved}.
}
\textcolor{black}{
Other ways to map speech and text include via segmental empirical output distribution matching (segmental empirical-ODM)~\cite{yeh2018unsupervised} and decipherment algorithm~\cite{Deciphering2022}.}

% Difficulty of unsupervised ASR (v.s. unsupervised MT)
Success in unsupervised neural machine translation
(MT)~\cite{artetxe2017unsupervised, conneau2017word, lample2017unsupervised}
has inspired innovative exploration of various unsupervised ASR algorithms.
If learning a translation model from unaligned sentences in two languages is
possible, considering speech and text as two different languages, learning the
mapping relationship from speech space to text space without an alignment 
% is not impossible.
  should likewise be possible.   % AMH: check
However, there are differences between unsupervised MT and unsupervised
ASR.
In unsupervised MT, most discrete source tokens can be mapped to specific
target tokens representing the same meaning. 
However, because speech has segmental structures, in unsupervised ASR, each text
token maps to a segment of consecutive acoustic features of variable length in
an utterance.
The generator is supposed to learn the segmental structure of an utterance
because information like token boundaries is not directly available.
This makes unsupervised ASR more challenging than unsupervised MT.

% how to represent audio
For unsupervised ASR to be feasible, the common idea is to make the speech and text units close to each other.
For the text side, word sequences can be transformed into phoneme sequences if a lexicon is available. 
On the other hand, we must first convert the speech signal into something close to phonemes. 
To achieve that, most studies on unsupervised ASR use a phoneme
segmentation module before the generator to segment utterances into
phoneme-level
segments~\cite{liu2018completely,chen2019completely,yeh2018unsupervised}. 
A representation vector or a token then represents each phoneme-level segment. 
It is easier for the generator to map each segment-level representation 
or token to the correct phoneme when the representation or token is highly correlated to the phonemes.
Wav2vec-U~\cite{baevski2021unsupervised} selects the input feature from different layers of wave2vec~2.0~\cite{baevski2020wav2vec}.
The selection criterion is based on analysis of the phonetic information in each layer.
\textcolor{black}{
If a universal phone recognizer trained from a diverse set of languages is available, it is another way to transcribe speech into phone-level tokens~\cite{Deciphering2022}.}
Another series of work is to transform a word into a word embedding. 
\cite{chung2018unsupervised,chung2019towards} use adversarial training to map the word-level
speech embedding space~\cite{chung2018speech2vec} to the word embedding space
and achieve promising performance on spoken word classification, speech
translation, and spoken word retrieval.
\Cref{fig:unsupervisedASR} summarizes the various ways to segment speech and represent speech and text in each reference.

%Hung-yi Lee: I comment this paragraph first.
\begin{comment}
%Hung-yi Lee: This paragraph is not finished yet.
\TNS{if we are limited on space,we can consider removing some of the detail of the various algorithms}
\TNS{split into 2 sentences, also have a connection to the previous para}
\cite{liu2018completely} is the first work to successfully achieve unsupervised
phoneme recognition by first clustering phoneme-level
embeddings~\cite{Chung2016AudioWord2Vec,Wang2018SegAudioWord2Vec} into a set of
tokens and then using GAN to learn the mapping relationship between tokens and
phonemes. 
However, in \cite{liu2018completely}, oracle phoneme segmentation is still
required to get reasonable performance. 
\cite{chen2019completely,yeh2018unsupervised} uses gate activation signals
(GAS)~\cite{wang2017gate} to get the phoneme-level segmentation, which is
unsupervised. 
%However, the loss term in~\cite{yeh2018unsupervised} includes an empirical average over all segments inside a logarithmic function, which can be biased if we sample this empirical average by a mini-batch average. \cite{liu2017unsupervised} therefore use an extremely large batch size during training to reduce the biasing problem.
%These problems are handled by another work~\cite{chen2019completely}.
%\cite{chen2019completely} removes the quantization step.
%Compared to~\cite{yeh2018unsupervised}, the discriminator in~\cite{chen2019completely} considers all possible information from the text dataset, not limited to the n-gram local statistics.
%Besides, \cite{chen2019completely} works well with a reasonable batch size, which makes it feasible when the computational resource is limited.
Wav2vec-U~\cite{baevski2021unsupervised} selects the input feature from
different layers of wave2vec 2.0~\cite{baevski2020wav2vec}, a self-supervised
model.
The selecting criterion is the PER by training a linear model in a supervised manner. 
%Should we mention this?
Wav2vec-U uses \kmeans{} to cluster the selected features, and the boundaries
are drawn whenever the clustered index changes.
%Wave2vec-U tends to generate more segments compared to the oracle phoneme-level segments. For example, their segmentation gives 0.93 precision, the ratio of the oracle phoneme-level boundaries being hit, and 0.38 recall, the ratio of the generated boundaries hit the oracle phoneme-level boundaries.
While Wav2vec-U merges the neighboring segments containing the same predicted
labels in each step of GAN training, this design can be viewed as refining the
segmentation implicitly.  
%Should we mention this?
%At generator output, the wav2vec-U will combine the neighboring predicted posterior containing the same predicted label into a predicted posterior.
\end{comment}

%In \cite{chung2018unsupervised, chung2019towards}, the word boundaries are also generated automatically with speech only. 
%However, \cite{chung2018unsupervised, chung2019towards} need oracle boundaries, therefore not completely unsupervised as the framework proposed in this paper.

% more improvement: iterative training, development 
As shown in \cref{fig:unsupervisedASR}, most studies use
\textit{self-training} to refine the models.
% to achieve better performance.  % AMH: redundant
In self-training, the generator serves as the first-version phoneme recognition model.
Inputting unpaired speech to the generator generates the corresponding
``pseudo transcription''. 
We then view the speech utterances and their pseudo transcriptions as paired
data which we use to train a model in a supervised manner.
Although the pseudo transcriptions have more errors than oracle
transcriptions, experiments show that training models on pseudo
transcriptions still significantly boosts performance compared to the
first-version model.
%With the new model, we can further obtain new transcriptions.
%The iteration will continue until the performance converges.

%Using performance to end this section.
Wav2vec-U~\cite{baevski2021unsupervised} achieved state-of-the-art results at the time, which suggests that representation learning is
essential for the success of unsupervised ASR.
It achieved an 11.3\% phoneme error rate on the TIMIT benchmark. 
On the LS benchmark, wav2vec-U achieved a 5.9\% \edit{WER} on
\emph{test-other}, rivaling some of the best published systems trained on 960 hours of labeled data from only two years earlier. 
\textcolor{black}{
And wav2vec-U 2.0~\cite{w2v-u2} further removes the requirement of the segmentation stage, so the unsupervised ASR model can be learned in an end-to-end style.}
The robustness of wav2vec-U was further analyzed with respect to 
domain-mismatch scenarios in which the domains of unpaired
speech and text were different~\cite{lin2021analyzing}.
Experimental results showed that domain mismatch leads to inferior performance, but a representation model pre-trained on the targeted speech domain extracts better representations and reduces this drop in performance.

%Hung-yi: The technique below is very important. But I think it is fine to ignore this part in a reviewe paper due to space limitation. 
\begin{comment}
In unsupervised learning, how to determine the hyperparameters is an issue. 
In a typical supervised framework, hyperparameter tuning is done by a
development set, which includes labeled data.
In an unsupervised setting, the existence of this kind of development is tricky. %Is tricky a correct term?
The existence of such kind of development set means the presence of the labeled
data.
Therefore, wav2vec-U~\cite{baevski2021unsupervised} develops a new approach to
determine the hyperparameters without a labeled development set. 
%should I provide more details? 
\end{comment}

%Hung-yi: A series of work (most from MS) focuses on meager resources but not wholly unsupervised. Perhaps we have to mention the papers.

\subsubsection{ASR-TTS}
%{\color{black} Shinji}\\
%{\color{black} Hung-yi: In the last meeting, we discussed the connection between this section and self-supervised learning. In addition to looking for papers using self-supervision to improve the ASR-TTS framework here, I think there may be another direction. Can we regard this ASR-TTS framework as a kind of self-supervised learning? ASR-TTS is similar to an autoencoder, and the output of ASR is "latent representation". In other words, this section describes a special self-supervised learning method, whose representation is expressed as "text."}\\
%{\color{black} Hung-yi: However, the above idea sounds tricky. In addition, the ``ASR-TTS'' autoencoder is learned by semi-supervised, not self-supervised.}
% AMH: to make the paper more consistent, I recommend changing all $f^{\mathrm{asr}}$ to $f_{\mathrm{asr}}$
Here we describe an alternative approach by which to train an ASR and \edit{text-to-speech (TTS)} system
based on unpaired text and audio. The ASR-TTS framework, which \edit{combines the ASR} and
TTS systems in a cascaded manner,
can be regarded as an autoencoder, where the encoder~$f$
corresponds to the ASR module and the decoder~$g$ corresponds to the TTS module.
In this framework, we consider the intermediate ASR output as a latent
representation; the framework as a whole can be regarded as a variant of
self-supervised learning.\footnote{However, to make this complicated
system work, we often require that data is paired. Therefore, in practice,
ASR-TTS and other methods described in this section are categorized as
semi-supervised learning.}

The ASR-TTS framework can jointly optimize both ASR and TTS \edit{without using paired data}~\cite{tjandra2017listening,hori2019cycle,wang2020improving}.
A speech chain~\cite{tjandra2017listening,tjandra2018machine} is one 
successful way to utilize audio-only and text-only data to train both
end-to-end ASR/TTS models.
This approach first prepares pre-trained ASR model \edit{
$f_{\mathrm{asr}}(X)$ with acoustic input $X$ and pre-trained TTS model $g_{\mathrm{tts}}(Y)$ with text input $Y$.
By following the TTS system with an ASR system, we generate new 
acoustic feature sequence~$\hat{X}$, which must be close to the original input~$X$.
Thus, we design a loss function $\mathcal{L}_{\mathrm{asr} \rightarrow \mathrm{tts}}(X, \hat{X})$, where $\hat{X}$ is generated by
%and~$\hat{X}$:
%\begin{equation}
%    \mathcal{L}_{\mathrm{asr} \rightarrow \mathrm{tts}} (X, \hat{X}),
%\end{equation}    
%where
\begin{equation}
    \hat{X} = g_{\mathrm{tts}}(f_{\mathrm{asr}}(X)).
    \label{eq:asr-tts}
\end{equation}
Thus, we train the ASR model (or both ASR and TTS models) using only
the acoustic input by minimizing $\mathcal{L}_{\mathrm{asr} \rightarrow \mathrm{tts}}$.  
%\begin{equation}
%\hat{\theta}_{\mathrm{asr}} = \arg \min _{\theta_{\mathrm{asr}}} \mathcal{L}(Y, \hat{Y}; \theta_{\mathrm{asr}}, \theta_{\mathrm{tts}}).
%\end{equation}
Note that this approach does not require the supervised text data~$Y$}.
As an analogy to the generative approach in \cref{sec:generative}, the
intermediate ASR output~$\hat{Y}$ can be regarded as the latent representation~$Z$.

The other cycle with the text-only data~$Y$ is also accomplished by the
concatenated TTS-ASR systems:\edit{
\begin{equation}
    \hat{Y} = f_{\mathrm{asr}}(g_{\mathrm{tts}}(Y)).
    \label{eq:tts-asr}
\end{equation}
Similarly, this approach does not require the supervised audio data~$X$, and the
intermediate TTS output~$\hat{X}$ can be regarded as the latent representation~$Z$}.
Although this approach initially freezes either the ASR or TTS model, 
extensions of this study~\cite{hori2019cycle,tjandra2019end,baskar2019semi}
implement the joint training of both ASR and TTS parameters using 
REINFORCE~\cite{williams1992simple} and straight-through estimators.

An emerging technique uses a well-trained TTS system to generate
speech and text data from text-only data.
This technique is a sub-problem of the TTS-ASR approach formulated in
\eqref{eq:tts-asr} in which we fix the TTS system part and estimate only the ASR
parameters. 
For example, a huge amount of text resources can be obtained from the web
and document archives without corresponding audio data.
%Self-supervised learning in this paper mostly focuses on representation learning perspectives of such unpaired data.
The typical use case scenario of such a text resource for ASR is through \edit{the
language model}. 
We combine the ASR and language model via a noisy channel 
model~\cite{jelinek1997statistical}, a weighted finite state 
transducer~\cite{mohri2002weighted}, or shallow 
fusion~\cite{gulcehre2015using,chorowski2016towards}.
However, the progress of TTS systems boosted by deep 
learning~\cite{oord2016wavenet,shen2018natural} has inspired another interesting and
straightforward research direction: \edit{artificially creating paired text and audio
data $\{\hat{X}, Y\}$ with only text data~$Y$ by generating the corresponding
audio data~$\hat{X}$ with TTS.} %\begin{equation}
%    \{\hat{X}, Y\} \text{ where } \hat{X} = g_{\mathrm{tts}} (Y).
%\end{equation}
The most straightforward approach is to simply use multi-speaker TTS to
generate the waveform with various acoustic 
variations~\cite{li2018training,ueno2019multi,rosenberg2019speech,laptev2020you,huang2020using}.
The other approaches are based on the generation of high-level (more
linguistic) features instead of generating the waveform, e.g., encoder
features~\cite{hayashi2018back} and phoneme 
features~\cite{renduchintala18_interspeech,masumura20_interspeech}.
This approach is similar to the back-translation technique developed in \edit{neural machine translation}~\cite{sennrich2016improving}.
One benefit of the above data generation approaches is that it can be used to feed
unseen word or context phrases to end-to-end ASR.

% {\color{black} Shinji: puts some notions to encourage the community to contributions to work on this direction and has more connections to the current SSL.}

\subsection{No text or lexicon} \label{subsec:zero}

% \kl{moved the acoustic word embedding subsection from here to the rep learning methods section.}\\

\subsubsection{Zero-resource speech technologies and challenges}
% {\color{black} Shinji}\\
\label{zero_speech}
Zero-resource speech technologies, which seek to discover linguistic concepts
from audio only (no text nor lexicon), are one of the most active applications
of unsupervised/self-supervised speech processing.
Zero-resource speech technologies were initially studied for acoustic and
linguistic unit discovery from speech data without linguistic resources,
e.g., transcriptions and other annotations~\cite{jansen2013summary}.
This study was motivated by unsupervised query-by-example, applications of
non-parametric Bayesian machine learning to speech processing, and low-resource
speech recognition, and was also inspired by the learning process of infants.
The goal of this type of work is to build spoken dialog systems in a zero-resource
setup for any language.
% To facilitate the zero-resource problem in the community, 
  To encourage      zero-resource research,                  % AMH: check
zero-resource speech challenges have been organized since 2015.

In this section, we describe the research directions of zero-resource speech
technologies by following the series of zero-resource speech challenges.
\begin{itemize}
	 \item \edit{Zero Resource Speech Challenge 2015~\cite{versteegh2015zero} mainly focused on building an acoustic model without using any
	 linguistic annotations based on subword unit modeling and spoken term
	 discovery tracks}.
	 \edit{For the subword unit modeling track, the ABX score for the within- and across-speaker
	 tasks was used as an evaluation metric.}
	 The spoken term discovery track used the normalized edit distance and coverage
	 scores in addition to the precision, recall, and F1 scores for types,
	 tokens, and boundaries.
    Both tracks were based on the English and Xitsonga languages.
	 \item The Zero Resource Speech Challenge 2017~\cite{dunbar2017zero} focused on 
	 unseen language and speaker aspects from the previous challenge. For
	 example, to demonstrate the robustness against unseen languages, the systems
	 were developed with English, French, and Mandarin and tested on
	 two ``surprise'' languages: German and Wolof.
	 Similarly, robustness against unseen speakers was demonstrated by varying
	 the amount of speech available for each speaker.
	 \item The Zero Resource Speech Challenge 2019~\cite{dunbar2019zero} extended a goal of
	 previous challenges by synthesizing speech without
	 text or phonetic labels but with acoustic units obtained using
	 zero-resource techniques.
	 The evaluation metrics were also extended to subjectively evaluate the
	 quality of synthesized speech, including its intelligibility, naturalness,
	 and speaker similarity.
	 \item The Zero Resource Speech Challenge 2020~\cite{dunbar2020zero} was based on two
	 tracks, revisiting previous challenges with different evaluation metrics. 
	 The first task revisited the 2019 challenge with low bit-rate subword
	 representations that optimize the quality of speech synthesis. The second
	 task revisited the 2017 challenge by focusing on the discovery of word-like
	 units from unsegmented raw speech.
	 \item The Zero Resource Speech Challenge 2021~\cite{nguyen2020zero}, the latest
	 \edit{challenge}, expanded the scope to include language modeling tasks.
	 In addition to phoneme-level ABX, the challenge includes lexical,
	 semantic, and syntactic evaluation metrics computed via a language model of
	 pseudo-acoustic labels.
\end{itemize}
These challenges have facilitated the tracking of technical trends in
zero-resource speech technologies.
For example, research directions thereof 
have expanded to various speech processing components to cover the entire
spoken dialogue systems.
To keep up with this expansion, the challenge has continued to develop
appropriate evaluation metrics for zero-resource scenarios.
Following the success of representation learning, baseline and challenge
techniques have shifted from purely generative 
models~\cite{ondel2016variational,heck2017feature}, deep 
autoencoders~\cite{tjandra19_interspeech,chorowski2019unsupervised}, and incorporation of
neural-network-based TTS/VC techniques~\cite{niekerk20b_interspeech} to
self-supervised learning~\cite{maekaku21_interspeech}.
The latest challenge included the visual modality, continuing the
expansion to include more aspects of human interaction.

\subsubsection{Textless NLP}
% {\color{black} Abdo}\\

Textless NLP is a new research direction that leverages the progress mentioned
above in self-supervised speech representation learning to model language
directly from audio, bypassing the need for text or 
labels~\cite{GSLM,SpeechResynthesis_IS21,pGSLM,emotion_conversion}. 
Not only does this open the gate for language and dialect modeling without
orthographic rules, but it also offers the opportunity to model other
non-lexical information about how speech is delivered, e.g., speaker identity,
emotion, hesitation, interruptions. 
The generative spoken language model (GSLM)~\cite{GSLM} utilizes discrete
representations from wav2vec~2.0, HuBERT, and CPC algorithms as inputs to an
autoregressive language model trained by \edit{using the cross-entropy function} to maximize the
probability of predicting the next discrete speech token. A \edit{synthesis} module
follows the language model to produce speech waveforms given the generated
discrete speech units. The generated spoken continuations compete with
supervised generations and synthesis using a character language model in
subjective human evaluations. The model completes incomplete words
(pow[..] $\rightarrow$ POWER) and continues using words in the same general mood (dark $\rightarrow$
BLACKNESS)\footnote{https://speechbot.github.io/gslm/} and has been extended to
model and generate
dialogues~\cite{dialogue_GSLM}.\footnote{https://speechbot.github.io/dgslm/}
%\cite{}. 
Given its flexibility in modeling spoken content, the GSLM has been further extended
to jointly model content and prosody~\cite{pGSLM}. This prosodic-GSLM model
introduced a multistream causal Transformer, where the input and output layers
use multiple heads to model three channels:  discrete speech units, duration,
and quantized pitch. The prosodic-GSLM model jointly generates novel content
and prosody congruently in the expressive 
style of the prompt.\footnote{https://speechbot.github.io/pgslm/}
%\cite{}. 
Going one step further, \cite{emotion_conversion} used a speech emotion
conversion framework to modify the perceived emotion of a speech utterance
while preserving its lexical content and speaker identity. Other studies have
extended the idea of textless language processing or audio discrete representation to applications such as 
spoken question answering~\cite{spoken_qa_dual}, speech separation~\cite{shi2021discretization}, TTS~\cite{hayashi2020discretalk}, and speech-to-speech
translation~\cite{textless_translation}.

### 2023-03-04 chapters/paper_review/06_conclusions.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\section{Discussion and Conclusion}
\label{sec:conclusion}
% {\color{blue} Reviewer: Daniel, SW }\\
% {\color{blue} Abdo}\\

In this overview, we have presented the historical context of self-supervised learning and provided a thorough methodological review of important  self-supervised speech representation models. Specifically, we have categorized the approaches into three categories, generative, contrastive and predictive, differing in terms of how the pretext task is defined.
We have presented an overview of existing benchmarks and reviewed the efforts towards efficient zero-resource learning.
\edit{Although the field is progressing rapidly, with new approaches reaching higher levels of performance, a couple of patterns have emerged: (1) The solid performance of Wav2vec 2.0 for speech recognition and many downstream tasks, as well as the public availability of its pre-trained multilingual variants, enabled wide adoption in the community making it a ``standard'' go-to model. (2) The simplicity and stability of the HuBERT approach, as well as the resemblance of its training procedure to classic frame-level ASR systems, made it an easy choice for research extensions on improving representation quality, speech translation, and textless NLP.}

Below we \edit{highlight} various \edit{shortcomings of existing work and } \edit{future} research directions:
\begin{itemize}

\item \textbf{Using the representation model.} So far, there are two main ways to use representation models: Freeze the representation models and use them as feature extractors, or fine-tune the representation models \edit{on} downstream tasks. 
\edit{Some efficient methods for leveraging SSL models exist in the NLP community. 
Adapters~\cite{pmlr-v97-houlsby19a,zaken2021bitfit,guo-etal-2021-parameter} are lightweight modules inserted into SSL models, and in downstream tasks, the parameters of SSL models are frozen, and only the adapters are trained. 
The prompt/instruction learning methods~\cite{liu2021pretrain} also freeze the SSL parameters and control the output of SSL by adding additional information, which is called \textit{prompt}, in the input.
Both adapter-based methods and prompt/instruction learning yield competitive performance compared with fine-tuning in NLP applications, but there is only little related work for speech~\cite{SpeechAdapter,SpeechPrompt}.
In addition, prompt for speech SSL does not achieve comparable performance on sequence generation tasks like phoneme recognition and slot filling, so how to use prompt is still an open question.}

\item \textbf{Increasing the efficiency of the representation model.} As discussed in \cref{subsec:modelsize}, larger representation models lead to better downstream performance. Despite the success of these \edit{large} models, they \edit{incur high costs in terms of memory and time for pre-training, fine-tuning, and even when used only to extract representations without gradient calculation. This} makes them unsuitable for edge devices \edit{but also limits the ability to scale these models to very large datasets} \ci{ -- and leads to a large energy consumption}. 
Preliminary studies have been conducted on compressing speech representation models through network pruning~\cite{PARP} or knowledge distillation~\cite{chang2021distilhubert}. 
\edit{There has been quite some effort towards more efficient general neural network models via conditional computing \cite{bengio_conditional_2016} and neural network quantization \cite{gholami_survey_2021} as well as extensive work on improving the specific efficiency of Transformer models, especially with the focus on self-attention \cite{tay_efficient_2022}, but these technology has not been widely used in speech SSL.}
\edit{Because speech is intrinsically represented as sequence, one way to reduce computation is to reduce the length of speech representation sequence but still keep the vital information in speech. But we have not been aware of any publication in this direction when writing this paper.}
On the other hand, non-streaming architectures in models such as the bidirectional Transformer have hindered the representation model used in streaming scenarios, leading to studies that address these problems~\cite{StreamingW2v}. 
We anticipate research in these directions to continue in the future.

\edit{\item \textbf{Data-efficient approaches.} SOTA representation learning methods require large volumes of unlabeled speech during pre-training, going way beyond what babies need to understand language. 
Different learning approaches have different data needs, e.g., generative approaches could be more data efficient than contrastive or predictive approaches since they are constrained by more bits of information to reconstruct their inputs. Comprehensive research is needed to study the data efficiency of different approaches. }

\item \textbf{Feature Disentanglement.} 
Speech SSL models show strengths on a surprisingly wide range of tasks~\cite{yang21c_interspeech}, suggesting that representations contain different information.
One way to further improve downstream tasks is to disentangle different information from the representation.
For example, we can decompose the representation into content embedding and speaker embedding and use content embedding for ASR and speaker embedding for SID.
Some work has been in this direction~\cite{pmlr-v162-qian22b,NANCY,MultimodelAmazonIS22}.

\item \textbf{Creating robust models.} \edit{As discussed in \cref{subsec:robustness}, studies have been conducted on the robustness of representation models \cite{wu2021characterizing}. However, the failure modes of SSL models are still poorly understood, and it remains unclear whether they provide more or less robustness to adversarial attacks than fully supervised models. Due to the importance of this research direction, while writing this paper, there is already some related research about enhancing the robustness of SSL models~\cite{robustw2v2,huang2022improving,wang2022improving,zhu2022noise} and identifying their vulnerability to adversarial attack~\cite{wu2021characterizing}. }

\edit{\item \textbf{Capturing higher-level semantic information.} Although many representation learning approaches can go beyond low-level phonetic modeling to capture some lexical information~\cite{TuAnh_2022_discrete}, they still struggle in higher-level semantic tasks easily captured by word-level counterparts like BERT. One workaround is two-stage training~\cite{pGSLM, dialogue_GSLM}; however, this prevents propagating rich lexical and semantic knowledge modeled in the second stage to benefit the phonetically focused first stage. }

\item \textbf{Using text representation models to improve speech representation.} The amount of content information in speech corpora used to train speech representation models is far less than that of text representation models. Noting that the BERT training corpus exceeds 3 billion words~\cite{jacob2019BERT}, and assuming a typical speaking rate of 120 words per minute, a speech corpus containing the same content as the BERT training data would include 400,000 hours of audio, which exceeds the \edit{accumulated} training data of all current speech representation models. Therefore, to enable speech representation models to better learn human language, for instance by extracting semantic information from acoustic signals, the use of text models such as BERT and GPT  \edit{seems} key: nevertheless, how to use these to improve speech representation model pre-training remains an open question. 
%\edit{ There is already some study using both speech and text data to pre-training, but some paired data is still required to achieve good performance~\cite{bapna2021slam}.  }

\end{itemize}

We believe SSL representation models have considerable room to grow. The relationship between representation models and downstream tasks can be compared to the relationship between operating systems and applications. Today, even individuals can build applications with desired functions on a smartphone because the smartphone's operating system handles the complex communication with the hardware and provides a convenient developer interface. Likewise, as SSL representation models learn general knowledge from human speech, it is easy to develop new speech processing applications on this basis. From this viewpoint, \edit{speech representation} models will play the role of operating systems in speech processing and further facilitate the continued development of speech technology. 
%Suppose, for instance, a speaker of an indigenous language seeks to purchase an intelligent assistant, but discovers that it does not yet support the indigenous language. Because the smart assistant has a built-in representation model, though, it can quickly learn a new language with little supervision. We believe SSL representation models will bring the benefits of speech technology to more people.

### 2023-03-04 chapters/paper_review/main_hi.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\begin{table*}[t]
    \centering
    \begin{tabular}{lcccccc}
        \toprule
        Model & Unlabeled Data & LM & dev-clean & dev-other & test-clean & test-other \\
        \midrule\midrule
        \multicolumn{7}{c}{\textit{\textbf{Superivsed}}} \\
        Conformer L~\cite{gulati2020conformer} & - & LSTM & - & - & 1.9 & 3.9 \\

        \midrule\midrule
        \multicolumn{7}{c}{\textit{\textbf{Self-Training}}} \\
        IPL~\cite{xu2020iterative} & LL-60k & 4-gram + Transformer & 1.85 & 3.26 & 2.10 & 4.01 \\
        Noisy Student~\cite{park2020improved} & LV-60k & LSTM & 1.6 & 3.4 & 1.7 & 3.4 \\

        \midrule\midrule
        \multicolumn{7}{c}{\textit{\textbf{Pre-Training}}} \\
        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec} & LL-60k & Transformer & 1.6 & 3.0 & 1.8 & 3.3 \\
        pre-trained Conformer XXL~\cite{zhang2020pushing} & LL-60k & LSTM & 1.5 & 3.0 & 1.5 & 3.1 \\

        \midrule\midrule
        \multicolumn{7}{c}{\textit{\textbf{Pre-Training + Self-Training}}} \\
        wav2vec 2.0 + self-training~\cite{xu2020self} & LL-60k & Transformer & 1.1 & 2.7 & 1.5 & 3.1\\
        pre-trained Conformer XXL + Noisy Student~\cite{zhang2020pushing} & LL-60k & LSTM & 1.3 & 2.6 & 1.4 & 2.6 \\

        \midrule\midrule
        \multicolumn{7}{c}{\textit{\textbf{This work (Pre-Training)}}} \\
        HUBERT \textsc{Large}   & LL-60k & Transformer & 1.5 & 3.0 & 1.9 & 3.3 \\
        % HUBERT \textsc{X-Large} & LL-60k & Transformer & 1.6 & 2.6 & 1.8 & 3.0 \\ % old number
        HUBERT \textsc{X-Large} & LL-60k & Transformer & 1.5 & 2.5 & 1.8 & 2.9 \\

        \bottomrule
    \end{tabular}
    \caption{Comparison with the literature on high resource setups using all 960 hours of labeled LibriSpeech data.}
    \label{tab:main_hi}
\end{table*}

### 2023-03-04 chapters/paper_review/main_lo.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\begin{table*}[t]
    \centering
    \begin{tabular}{lcccccc}
        \toprule
        Model & Unlabeled Data & LM & dev-clean & dev-other & test-clean & test-other \\
        \midrule\midrule
        \multicolumn{7}{c}{\textit{\textbf{10-min labeled}}} \\
        % DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & - & - \\
        DiscreteBERT~\cite{baevski2019effectiveness} & LS-960 & 4-gram & 15.7 & 24.1 & 16.3 & 25.2 \\
        wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 8.9 & 15.7 & 9.1 & 15.6 \\
        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & 4-gram & 6.3 & 9.8 & 6.6 & 10.3  \\
        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & Transformer & 4.6 & 7.9 & 4.8 & 8.2 \\
        \midrule
        HUBERT \textsc{Base}    & LS-960 & 4-gram & 9.1 & 15.0 & 9.7 & 15.3  \\
        HUBERT \textsc{Large}   & LL-60k & 4-gram & 6.1 & 9.4 & 6.6 & 10.1 \\
        HUBERT \textsc{Large}   & LL-60k & Transformer & 4.3 & 7.0 & 4.7 & 7.6 \\
        HUBERT \textsc{X-Large} & LL-60k & Transformer & 4.4 & 6.1 & 4.6 & 6.8 \\

        \midrule\midrule
        \multicolumn{7}{c}{\textit{\textbf{1-hour labeled}}} \\
        DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & 13.8 & 29.1 \\
        DiscreteBERT~\cite{baevski2019effectiveness} & LS-960 & 4-gram & 8.5 & 16.4 & 9.0 & 17.6 \\
        wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 5.0 & 10.8 & 5.5 & 11.3 \\
        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & Transformer & 2.9 & 5.4 & 2.9 & 5.8 \\
        \midrule
        HUBERT \textsc{Base}    & LS-960 & 4-gram & 5.6 & 10.9 & 6.1 & 11.3  \\
        HUBERT \textsc{Large}   & LL-60k & Transformer & 2.6 & 4.9 & 2.9 & 5.4 \\
        HUBERT \textsc{X-Large} & LL-60k & Transformer & 2.6 & 4.2 & 2.8 & 4.8 \\

        \midrule\midrule
        \multicolumn{7}{c}{\textit{\textbf{10-hour labeled}}} \\
        SlimIPL~\cite{likhomanenko2020slimipl} & LS-960 & 4-gram + Transformer & 5.3 & 7.9 & 5.5 & 9.0 \\
        DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & 5.4 & 13.3 \\
        DiscreteBERT~\cite{baevski2019effectiveness} & LS-960 & 4-gram & 5.3 & 13.2 & 5.9 & 14.1 \\
        wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 3.8 & 9.1 & 4.3 & 9.5 \\
        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & Transformer & 2.4 & 4.8 & 2.6 & 4.9 \\
        \midrule
        HUBERT \textsc{Base}    & LS-960 & 4-gram & 3.9 & 9.0 & 4.3 & 9.4 \\
        HUBERT \textsc{Large}   & LL-60k & Transformer & 2.2 & 4.3 & 2.4 & 4.6 \\
        HUBERT \textsc{X-Large} & LL-60k & Transformer & 2.1 & 3.6 & 2.3 & 4.0 \\

        \midrule\midrule
        \multicolumn{7}{c}{\textit{\textbf{100-hour labeled}}} \\
        % LPM~\cite{hsu2020semi} & LL-60K & ConvLM & 4.87 & 13.84 & 4.88 & 15.28 \\
        IPL~\cite{xu2020iterative} & LL-60k & 4-gram + Transformer & 3.19 & 6.14 & 3.72 & 7.11 \\
        SlimIPL~\cite{likhomanenko2020slimipl} & LS-860 & 4-gram + Transformer & 2.2 & 4.6 & 2.7 & 5.2 \\
        Noisy Student\cite{park2020improved} & LS-860 & LSTM & 3.9 & 8.8 & 4.2 & 8.6 \\
        DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & 5.0 & 12.1 \\
        DiscreteBERT~\cite{baevski2019effectiveness} & LS-960 & 4-gram & 4.0 & 10.9 & 4.5 & 12.1 \\
        wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 2.7 & 7.9 & 3.4 & 8.0 \\
        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & Transformer & 1.9 & 4.0 & 2.0 & 4.0 \\

        \midrule
        HUBERT \textsc{Base}    & LS-960 & 4-gram & 2.7 & 7.8 & 3.4 & 8.1  \\
        HUBERT \textsc{Large}   & LL-60k & Transformer & 1.8 & 3.7 & 2.1 & 3.9 \\
        HUBERT \textsc{X-Large} & LL-60k & Transformer & 1.7 & 3.0 & 1.9 & 3.5 \\

        \bottomrule
    \end{tabular}
    \caption{Results and comparison with the literature on low resource setups (10-min, 1-hour, 10-hour, and 100-hour of labeled data).}
    \label{tab:main_lo}
\end{table*}

### 2023-03-04 chapters/paper_review/model.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\begin{table*}[h!]
\caption{
The models are sorted according to publication date of the first version on arXiv which might differ from the year indicated by the citation. 
\textbc{{reg}} means that the model uses a regression loss, 
\textbc{{clf}} that it uses a classification loss,
\textbc{{con}} that it uses a contrastive loss, 
\textbc{{msk}} that it uses masking to define the target region, 
\textbc{{ltg}} that it uses a learned target, 
\textbc{{qtz}} that it uses a quantized target, and
\textbc{{emb}} that it embeds a variable length segment.
%Lee: I think contrastive always have learned target. Am I correct???
}
\label{tab:model-taxonomy}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{ l l | c c c | c c c | c } 
\toprule
\multicolumn{2}{c}{} & 
\multicolumn{3}{c}{\small\textsc{Loss}} & 
\multicolumn{3}{c}{\small\textsc{Target}} &
 \\
\textbc{{model}} &
\textbc{{pub. date}} &
\textbc{{reg}} & 
\textbc{{clf}} & 
\textbc{{con}} &
\textbc{{msk}} &
\textbc{{ltg}} &  
\textbc{{qtz}} & 
\textbc{{emb}} \\
\midrule
%                                                         REG      CLF      CON      MSK      LTG      QTZ      EMB

Audio Word2vec \cite{Chung2016AudioWord2Vec}      & 2016 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\

VQ-VAE \cite{vqvae} & 2017 Nov. & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark \\ %2 Nov 2017

Speech2Vec \cite{chung2018speech2vec}     & 2018 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
Unspeech \cite{milde2018unspeech}         & 2018 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\

CPC \cite{oord2018representation}         & 2018 Jul. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\

PASE \cite{pascual2019learning}                   & 2019 Apr. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 6/4

APC \cite{chung2019unsupervised}          & 2019 Oct. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 5/4

wav2vec \cite{schneider2019wav2vec}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4

vq-wav2vec \cite{Baevski2020vq-wav2vec} & 2019 Oct. & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ %12 Oct 2019 %Lee: vq-wav2vec is CON or CLF??? %Lee: Discrete BERT use masking, but vq-wav2vec does not It is CON, not CLF.

Discrete BERT \cite{Baevski2020vq-wav2vec,baevski2019effectiveness} & 2019 Oct. & \xmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ %12 Oct 2019 %10 Nov 2019 %In the paper of vq-wav2vec, it also includes the approach of Discrete BERT. I know this is tricky ....

Mockingjay  \cite{liu2020mockingjay}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 25/10-19

DeCoAR \cite{ling2020deep}                & 2019 Dec. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 3/12-19

PASE+ \cite{ravanelli2020multi}                & 2020 Jan. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 25/1 % I think contrastive always have learned target.

Bidir CPC \cite{kawakami2020learning} & 2020 Jan. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\% 29/1

Modified CPC \cite{riviere2020unsupervised} & 2020 Feb. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 7/2

VQ-APC \cite{chung20e_interspeech}  & 2020 May  & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark \\ %17/05

AALBERT \cite{chi2020audio}  & 2019 May & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ %18 May 2020

speech-XLNet \cite{song20d_interspeech} & 2019 Oct. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\%23 Oct 2019 %Lee: It is speech version of XLNet. I am not sure we should consider this approach has mask or not.

wav2vec 2.0 \cite{baevski2020wav2vec}     & 2020 Jun. & \xmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ % 20/6
%Lee: In my point of view, wav2vec 2.0 has learned target. The target is from the CNN encoder.

MPE \cite{liu2020masked} & 2020 May & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 25 May 2020 %Lee: I think it has learned target. But please check.

TERA \cite{liu2021tera}                   & 2020 Jul. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 12/7

Speech SimCLR \cite{SpeechSimCLR} &  2020 Oct. & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \xmark \\ %Lee: It is TERA+ SIMCLR. I considered it has learned target because contrastive always have learned target.
%27 Oct 2020

NPC \cite{liu21l_interspeech}                     & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ % 1/11

DeCoAR 2.0 \cite{ling2020decoar}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ % 11/12 %Lee:  DeCoAR 2.0 has VQ layer although it is not used as learning target.

wav2vec-C \cite{sadhu21_interspeech} & 2021 Mar. & \cmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ %9 Mar 2021 %Lee:  consistency loss is reconstruction

HuBERT \cite{hsu2021hubert}               & 2021 Jun. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ % 14/6

W2v-bert \cite{w2vbert} & 2021 Aug & \xmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark \\
%7 Aug 2021

pMPC \cite{yue2021pMPC}  & 2021 Aug. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ %30 August – 3 September, 2021 (I do not find arXiv version.)

WavLM \cite{chen2021wavlm} & 2021 Oct. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ %26 Oct 2021

data2vec \cite{data2vec}     & 2022 Jan. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 20/1

BEST-RQ \cite{BEST-RQ} & 2022 Feb.  & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ %Lee: no learned target. The target is random ...
% 3 Feb 2022

\bottomrule
\end{tabular}
\end{center}
\end{table*}

\begin{comment}
\begin{table*}[h!]
\caption{
The models are sorted according to the publication date of the first version on arXiv, which might differ from the year indicated by the citation. 
\textbc{{reg}} means that the model uses a regression loss, 
\textbc{{clf}} that it uses a classification loss,
\textbc{{con}} that it uses a contrastive loss, 
\textbc{{msk}} that it uses masking to define the target region, 
\textbc{{ltg}} that it uses a learned target, %lasseborgholt: We could consider using the inverse of this attribute. I.e., that the model uses the input as taget and belongs in the generative category.
\textbc{{qtz}} that it uses a quantized target, and
\textbc{{emb}} that it embeds a variable-length segment
}
\label{tab:model-taxonomy}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{ l l | c c c | c c c | c } 
\toprule
\multicolumn{2}{c}{} & 
\multicolumn{3}{c}{\small\textsc{Loss}} & 
\multicolumn{3}{c}{\small\textsc{Target}} &
 \\
\textbc{{model}} &
\textbc{{pub. date}} &
\textbc{{reg}} & 
\textbc{{clf}} & 
\textbc{{con}} &
\textbc{{msk}} &
\textbc{{ltg}} &  
\textbc{{qtz}} & 
\textbc{{emb}} \\
\midrule
%                                                         REG      CLF      CON      MSK      LTG      QTZ      EMB  
Audio Word2vec \cite{Chung2016AudioWord2Vec}      & 2016 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
Speech2Vec \cite{chung2018speech2vec}     & 2018 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
% Unspeech \cite{milde2018unspeech}         & 2018 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
CPC \cite{oord2018representation}         & 2018 Jul. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\

PASE \cite{pascual2019learning}                   & 2019 Apr. & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark \\ % 6/4

APC \cite{chung2019unsupervised}          & 2019 Oct. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 5/4
wav2vec \cite{schneider2019wav2vec}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
%VQ-APC \cite{chung2020vector}             & 2020 May  & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ 
Mockingjay  \cite{liu2020mockingjay}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 25/10-19
%DeCoAR \cite{ling2020deep}                & 2019 Dec. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 3/12-19
%PASE+ \cite{Ravanelli2020}                & 2020 Jan. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 25/1
wav2vec 2.0 \cite{baevski2020wav2vec}     & 2020 Jun. & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark \\ % 20/6
%TERA \cite{liu2021tera}                   & 2021 Jul. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 12/7

NPC \cite{liu21l_interspeech}             & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 1/11

DeCoAR 2.0 \cite{ling2020decoar}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 11/12
HuBERT \cite{hsu2021hubert}               & 2021 Jun. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ % 14/6
data2vec \cite{data2vec}     & 2022 Jan. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 20/1
\bottomrule
\end{tabular}
\end{center}
\end{table*}

\end{comment}

### 2023-03-04 chapters/paper_review/pretext.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\begin{table*}[!htb]
    \centering
    \caption{
    A summary of the approaches in the three categories of
    self-supervised learning. 
    Column~(a) lists the names of the models and related references, 
    column~(b) defines the model input, 
    column~(c) defines any corruption of the input or hidden representation, and
    column~(d) defines the target of the pretext task; the pretext task itself
    is described by the overall model category and the main text.
    $X=\{x_1,x_2,...,x_T\}$ is the input sequence in which $x_t$ can be an
    acoustic feature vector (e.g., MFCC, filterbank, or spectrogram features)
    or a waveform sample. 
    $X_{[t_1:t_2]}$ represents $\{x_{t_1},x_{t_1+1},...,x_{t_2}\}$.
    $X_{-[t_1:t_2]}$ represents $X$ in which the segment
    $X_{[t_1:t_2]}=\{x_{t_1},x_{t_1+1},...,x_{t_2}\}$ is masked.
    $x_{t}^{i}$ represents the $i$-th dimension of $x_t$.
    If $x_t$ is a frame in a spectrogram, then the $i$-th dimension corresponds
    to a specific frequency bin.
    $X^{-[f,f+j]}$ refers to a spectrogram $X$ which is masked along the frequency
    axis from the $f$-th to $(f+j)$-th bin. 
    % $X_{-[t:t+k]}^{-[f,f+j]}$ refers to a spectrogram $X$ but masked along both time and frequency dimensions. 
    % $X^*$ is a temporally permuted version of $X$, that is, the $x_t$ are randomly shuffled to form $X^*$. 
    We indicate random temporal permutation of a sequence by indexing it with
    the set $\mathcal{P}_t\triangleq\textsc{permute}([0,t])$, where
    $\textsc{permute}(\cdot)$ returns a permutation of the given list. 
    We indicate data augmentation (e.g., reverberation) by the function
    $\textsc{augment}(\cdot)$. Subscripts indicate different augmentations. 
    % $X^\prime$ is an augmented version of $X$ (e.g., $X$ with reverberation), while $X^{\prime\prime}$ is $X$ adding distortion different from  $X^\prime$.
    $Z$ represents a localized latent representation sequence of $X$. %, and $Z^\prime$ and $Z^{\prime\prime}$ are the latent representation sequences of  $X^\prime$ and $X^{\prime\prime}$, respectively. 
    % $Z$ represents a localized latent representation sequence and $Z^\prime$ is the latent representation sequence of $X^\prime$. 
    $Z^{(l)}$ is $Z$ at the $l$-th layer of the model used to compute it.
    $\bar{H}$ is the contextualized sequence $H$ obtained from an exponential
    moving average (EMA) of the model undergoing training with no masking
    applied.
    $Q$ represents a sequence of quantized learned representations, and $C$ is
    a sequence of discrete cluster IDs.
    For contrastive models, we specify only positive targets.
    }
    % \begin{minipage}{\textwidth}  
\centering
\renewcommand*\arraystretch{1.2}{
\begin{tabular}{l|c|c|c}
    \toprule
    \textbf{Model} (a) & \textbf{Input} (b) & \textbf{Corruption} (c) & \textbf{Target} (d) \\
    \midrule
    \midrule
    \multicolumn{4}{c}{\textsc{Generative models}} \\
    \midrule
    \midrule
    Audio Word2vec~\cite{Chung2016AudioWord2Vec}, VQ-VAE~\cite{vqvae}     & $X$ &   \textsc{-}    &  $X$  \\ %It has VQ as extra constraint.
    \midrule  
    Speech2Vec~\cite{chung2018speech2vec}, Audio2Vec~\cite{tagliasacchi2020pre} - skip-gram    & $X_{[t_1,t_2]}$  &     \textsc{-}   &    $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$     \\
    \midrule 
    Speech2Vec~\cite{chung2018speech2vec}, Audio2Vec~\cite{tagliasacchi2020pre} - cbow    & $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$   &     \textsc{-}   &    $X_{[t_1,t_2]}$   \\
    \midrule 
    PASE~\cite{pascual2019learning}, PASE+~\cite{ravanelli2020multi}\footnote{PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
    \midrule 
    APC~\cite{chung2019unsupervised,chung20e_interspeech}         & $X_{[1,t]}$   & \textsc{-}              & $x_{t+c},\, c\geq1$    \\ %VQ-APC is the same as APC
    \midrule
    Speech-XLNet \cite{song20d_interspeech}     & \multicolumn{2}{c|}{$X_{\mathcal{P}_{t}}$}   &     $x_{i\sim\mathcal{P}^c_{t}}$  \\ %Lee: Use * to represent the permutation, hope it is not strange.
    \midrule  
    DeCoAR~\cite{ling2020deep}     & $X_{[1,t-1]}, X_{[t+k+1,T]}$ & \textsc{-} & $X_{[t,t+k]}$   \\
    \midrule
    Mockingjay~\cite{liu2020mockingjay}, Audio ALBERT~\cite{chi2020audio}, DeCoAR 2.0~\cite{ling2020decoar}   & \multicolumn{2}{c|}{$X_{-[t,t+k]}$}   & $X_{[t,t+k]}$    \\
    \midrule 
    TERA~\cite{liu2021tera}, BMR~\cite{wang2020unsupervised}  & \multicolumn{2}{c|}{$X_{-[t,t+k]}^{-[f,f+j]}$}       & $X$       \\
    \midrule
    pMPC~\cite{yue2021pMPC}      &  \multicolumn{2}{c|}{$X_{-[t,t+k^\prime]}$ ($X_{[t,t+k^\prime]}$ is a phoneme)}       & $X_{[t,t+k^\prime]}$    \\
    \midrule 
    MPE~\cite{liu2020masked} & $X$ &  $Z_{-[t,t+k]}$  & $Z$    \\ %Lee: What is the difference between MPE and NPC??? And I believe it reconstruct the convolutional blocks output (learned target?)
    \midrule
    NPC~\cite{liu21l_interspeech}      & $X$  &   $Z_{-[t,t+k]}$  &    $X$   \\ %Lee: I am not 100% sure it is correct. Please check.
    \midrule
    \midrule
    \multicolumn{4}{c}{\textsc{Contrastive models}} \\
    \midrule
    \midrule
    Unspeech \cite{milde2018unspeech}       &   $X_{[t_1,t_2]}$ &   \textsc{-}   &  $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$ \\
    \midrule 
    CPC~\cite{oord2018representation}, wav2vec \cite{schneider2019wav2vec}, Modified CPC \cite{riviere2020unsupervised}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
    \midrule 
    Bidirectional CPC \cite{kawakami2020learning}      & $X_{[1,t]}$ or $X_{[t,T]}$ &  \textsc{-}    &    $z_{t+c}$ or $z_{t-c},\, c\geq1$   \\
    \midrule 
    vq-wav2vec \cite{Baevski2020vq-wav2vec}     &   $X_{[1,t]}$ &   \textsc{-}    &   $q_{t+c},\, c\geq1$   \\ 
    \midrule 
    wav2vec 2.0 \cite{baevski2020wav2vec}, wav2vec-C \cite{sadhu21_interspeech}\footnote{wav2vec-C adds reconstruction loss to wav2vec 2.0.}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
    \midrule 
    w2v-BERT \cite{w2vbert}     &$X$ &    $Z_{-[t,t+k]}$   &     $Q_{[t,t+k]}$ and $C_{[t,t+k]}$     \\
    \midrule
    Speech SimCLR \cite{SpeechSimCLR}\footnote{Speech SimCLR targets the latent representation of an augmented version of $X$ using a differently augmented $X$, and vice-versa.}    & \multicolumn{2}{c|}{$\textsc{augment}_1(X)$ and $\textsc{augment}_2(X)$}     &    $\textsc{augment}_2(Z)$ and $\textsc{augment}_1(Z)$   \\ 
    \midrule
    \midrule 
    \multicolumn{4}{c}{\textsc{Predictive models}} \\
    \midrule
    \midrule
    Discrete BERT~\cite{Baevski2020vq-wav2vec,baevski2019effectiveness} \footnote{Discrete BERT obtains codes $C$ from vq-wav2vec.}      &   \multicolumn{2}{c|}{$C_{-[t,t+k]}$}   & $C_{[t,t+k]}$  \\
    \midrule 
    HuBERT \cite{hsu2021hubert}\footnote{HuBERT is trained first using cluster IDs of the MFCCs as target and subsequently clusters IDs of the model representations from the last iteration.}, WavLM \cite{chen2021wavlm}\footnote{WavLM simulates noisy/overlapped speech as inputs.}  & $X$             & $Z_{-[t,t+k]}$          & $C_{[t,t+k]}$  \\ 
    \midrule
    data2vec \cite{data2vec}    & $X$             & $Z_{-[t,t+k]}$          & $\sum_{l}\bar{H}^{(l)}_{[t,t+k]}$  \\ 
    \midrule 
    BEST-RQ \cite{BEST-RQ}\footnote{BEST-RQ obtains codes $C$ by quantizing acoustic features using a random projection quantizer.}     &  \multicolumn{2}{c|}{$X_{-[t,t+k]}$}      &  $C_{[t,t+k]}$   \\ 
    \midrule 
    \bottomrule
\end{tabular}
}

% \end{minipage}
\label{table:pretext}
\end{table*}

\begin{comment}
\begin{table*}[ht!]
    \centering
    \caption{
    This table summarizes the approaches of the three categories of self-supervised learning. 
    Column (a) lists the names of the models and related references. 
    Column (b) defines the input to the models. 
    Column (c) defines any corruption of the input or some hidden representation. 
    Column (d) defines the target of the pretext task; the pretext task itself is described by the overall model category and the main text.
    $X=\{x_1,x_2,...,x_T\}$ is the input sequence in which $x_t$ can be an acoustic feature vector (e.g., MFCC, filterbank, or spectrogram features) or a waveform sample. 
    $X_{[t_1:t_2]}$ represents $\{x_{t_1},x_{t_1+1},...,x_{t_2}\}$.
    $X_{-[t_1:t_2]}$ represents $X$ with the segment $X_{[t_1:t_2]}=\{x_{t_1},x_{t_1+1},...,x_{t_2}\}$ masked.
    $x_{t}^{i}$ represents the $i$-th dimension of $x_t$.
    If $x_t$ is a frame in a spectrogram, then the $i$-th dimension corresponds to a specific frequency bin.
    $X^{-[f,f+j]}$ refers to a spectrogram $X$ but masked along the frequency axis from the $f$-th to $f+j$-th bin. 
    % $X_{-[t:t+k]}^{-[f,f+j]}$ refers to a spectrogram $X$ but masked along both time and frequency dimensions. 
    % $X^*$ is a temporally permuted version of $X$, that is, the $x_t$ are randomly shuffled to form $X^*$. 
    We indicate random temporal permutation of a sequence by indexing it with the set $\mathcal{P}_t\triangleq\textsc{permute}([0,t])$ where $\textsc{permute}(\cdot)$ returns a permutation of the given list. 
    We indicate data augmentation (e.g. reverberation) by the function $\textsc{augment}(\cdot)$. Subscripts indicate different augmentations. 
    % $X^\prime$ is an augmented version of $X$ (e.g., $X$ with reverberation), while $X^{\prime\prime}$ is $X$ adding distortion different from  $X^\prime$.
    $Z$ represents a localized latent representation sequence of $X$. %, and $Z^\prime$ and $Z^{\prime\prime}$ are the latent representation sequences of  $X^\prime$ and $X^{\prime\prime}$, respectively. 
    % $Z$ represents a localized latent representation sequence and $Z^\prime$ is the latent representation sequence of $X^\prime$. 
    $Z^{(l)}$ is $Z$ at the $l$-th layer of the model used to compute it.
    $\bar{H}$ is the contextualized sequence $H$ obtained from an exponential moving average (EMA) of the model undergoing training with no masking applied.
    $Q$ represents a sequence of quantized learned representations, and $C$ is a sequence of discrete cluster IDs.
    For contrastive models, we specify only positive targets.
    }
    \begin{minipage}{\textwidth}  
    \centering
    \renewcommand*\arraystretch{1.1}{
    \begin{tabular}{l|c|c|c}
    \toprule
    \textbf{Model} (a) & \textbf{Input} (b) & \textbf{Corruption} (c) & \textbf{Target} (d) \\
    \midrule
    \midrule
    \multicolumn{4}{c}{\textsc{Generative models}} \\
    \midrule
    \midrule
    Audio Word2vec~\cite{Chung2016AudioWord2Vec}, VQ-VAE \cite{vqvae}     & $X$ &   \textsc{-}    &  $X$  \\ %It has VQ as extra constraint.
    \midrule  
    Speech2Vec~\cite{chung2018speech2vec}, Audio2Vec~\cite{tagliasacchi2020pre} - skip-gram    & $X_{[t_1,t_2]}$  &     \textsc{-}   &    $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$     \\
    \midrule 
   Speech2Vec~\cite{chung2018speech2vec}, Audio2Vec~\cite{tagliasacchi2020pre} - cbow    & $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$   &     \textsc{-}   &    $X_{[t_1,t_2]}$   \\
    \midrule 
     PASE~\cite{pascual2019learning}, PASE+~\cite{ravanelli2020multi}\footnote{PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
    \midrule 
    APC~\cite{chung2019unsupervised,chung20e_interspeech}         & $X_{[1,t]}$   & \textsc{-}              & $x_{t+c},\, c\geq1$    \\ %VQ-APC is the same as APC
    \midrule
    Speech-XLNet \cite{song20d_interspeech}     & \multicolumn{2}{c|}{$X_{\mathcal{P}_{t}}$}   &     $x_{i\sim\mathcal{P}^c_{t}}$  \\ %Lee: Use * to represent the permutation, hope it is not strange.
    \midrule  
    DeCoAR~\cite{ling2020deep}     & $X_{[1,t-1]}, X_{[t+k+1,T]}$ & \textsc{-} & $X_{[t,t+k]}$   \\
    \midrule
    Mockingjay~\cite{liu2020mockingjay}, Audio ALBERT~\cite{chi2020audio}, DeCoAR 2.0~\cite{ling2020decoar}   & \multicolumn{2}{c|}{$X_{-[t,t+k]}$}   & $X_{[t,t+k]}$    \\
    \midrule 
    TERA~\cite{liu2021tera}, BMR~\cite{wang2020unsupervised}  & \multicolumn{2}{c|}{$X_{-[t,t+k]}^{-[f,f+j]}$}       & $X$       \\
    \midrule
    pMPC~\cite{yue2021pMPC}      &  \multicolumn{2}{c|}{$X_{-[t,t+k^\prime]}$ ($X_{[t,t+k^\prime]}$ is a phoneme)}       & $X_{[t,t+k^\prime]}$    \\
    \midrule 
    MPE~\cite{liu2020masked} & $X$ &  $Z_{-[t,t+k]}$  & $Z$    \\ %Lee: What is the difference between MPE and NPC??? And I believe it reconstruct the convolutional blocks output (learned target?)
    \midrule
    NPC~\cite{liu21l_interspeech}      & $X$  &   $Z_{-[t,t+k]}$  &    $X$   \\ %Lee: I am not 100% sure it is correct. Please check.
    \midrule
    \midrule
    \multicolumn{4}{c}{\textsc{Contrastive models}} \\
    \midrule
    \midrule
    Unspeech \cite{milde2018unspeech}       &   $X_{[t_1,t_2]}$ &   \textsc{-}   &  $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$ \\
    \midrule 
    CPC~\cite{oord2018representation}, wav2vec \cite{schneider2019wav2vec}, Modified CPC \cite{riviere2020unsupervised}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
        \midrule 
    Bidirectional CPC \cite{kawakami2020learning}      & $X_{[1,t]}$ or $X_{[t,T]}$ &  \textsc{-}    &    $z_{t+c}$ or $z_{t-c},\, c\geq1$   \\
    \midrule 
    vq-wav2vec \cite{Baevski2020vq-wav2vec}     &   $X_{[1,t]}$ &   \textsc{-}    &   $q_{t+c},\, c\geq1$   \\ 
    \midrule 
    wav2vec 2.0 \cite{baevski2020wav2vec}, wav2vec-C \cite{sadhu21_interspeech}\footnote{wav2vec-C adds the reconstruction loss to wav2vec 2.0.}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
    \midrule 
    w2v-BERT \cite{w2vbert}     &$X$ &    $Z_{-[t,t+k]}$   &     $Q_{[t,t+k]}$ and $C_{[t,t+k]}$     \\
    \midrule
    Speech SimCLR \cite{SpeechSimCLR}\footnote{Speech SimCLR targets the latent representation of an augmented version of $X$ using a differently augmented $X$, and vice-versa.}    & \multicolumn{2}{c|}{$\textsc{augment}_1(X)$ and $\textsc{augment}_2(X)$}     &    $\textsc{augment}_2(Z)$ and $\textsc{augment}_1(Z)$   \\ 
   \midrule
   \midrule 
    \multicolumn{4}{c}{\textsc{Predictive models}} \\
    \midrule
    \midrule
    Discrete BERT~\cite{Baevski2020vq-wav2vec,baevski2019effectiveness} \footnote{Discrete BERT obtains codes $C$ from vq-wav2vec.}      &   \multicolumn{2}{c|}{$C_{-[t,t+k]}$}   & $C_{[t,t+k]}$  \\
    \midrule 
    HuBERT \cite{hsu2021hubert}\footnote{HuBERT is trained first using cluster IDs of MFCCs as target and subsequently cluster IDs of model representations from the last iteration.}, WavLM \cite{chen2021wavlm}\footnote{WavLM simulates noisy/overlapped speech as inputs.}  & $X$             & $Z_{-[t,t+k]}$          & $C_{[t,t+k]}$  \\ 
    \midrule
    data2vec \cite{data2vec}    & $X$             & $Z_{-[t,t+k]}$          & $\sum_{l}\bar{H}^{(l)}_{[t,t+k]}$  \\ 
        \midrule 
BEST-RQ \cite{BEST-RQ}\footnote{BEST-RQ obtains codes $C$ by quantizing acoustic features using a random projection quantizer.}     &  \multicolumn{2}{c|}{$X_{-[t,t+k]}$}      &  $C_{[t,t+k]}$   \\ 
        \midrule 
    \bottomrule
    \end{tabular}
    }

    \end{minipage}
    \label{table:pretext}
\end{table*}
\end{comment}

\begin{comment}
\begin{table*}[h!]
\centering

\setlength{\tabcolsep}{3pt}
\caption{
This table summarizes the generative approaches. 
Column (a) is the name of the models (if any) and their reference.
Column (b) is the corrupted speech signals for encoder input.
Column (c) is the target output of decoder.
Column (d) is regularization methods used if any.
Column (e) are the models using the same pretext in NLP or CV.
$S$ is a sequence of waveform samples, and $s_t$ is a single waveform sample.
$S^*$ is the degradation of $S$ with reverberation or additive noises. 
$S$ can be represented by a sequence of acoustic feature, $X=\{x_1,x_2,...,x_T\}$, in which $x_t$ is an acoustic feature vector like MFCC, fbank, spectrogram, etc. 
$X^*$ is the permuted version of $X$. That is, the frames in $X$ are randomly shuffled to form $X^*$. 
$X_{[t_1:t_2]}$ represents $\{x_{t_1},x_{t_1+1},...,x_{t_2}\}$.
$X_{-[t_1:t_2]}$ represents the segment $\{x_{t_1},x_{t_1+1},...,x_{t_2}\}$ in $X$ is masked, that is, the frames in the region is replaced by zero vectors or random sampled vectors.
$x_{t}^{i}$ represents the $i$-th dimension of frame $x_t$.
If $x_t$ is a frame in a spectrogram, then the $i$-th dimension corresponds to a specific frequency bin.
$X^{-[f_1,f_2]}$ means that for all the frames $x$ in $X$, their $f_1$-th to $f_2$-th dimensions are all masked. 
$X_{-[t_1:t_2]}^{-[f_1,f_2]}$ means $X$ are masked in two directions. 
}
\label{table:generative}
  {\renewcommand*\arraystretch{1.2}
\begin{tabular}{|l|l|l|l|l|}
\hline
(a) Model & (b) Corrupted Input & (c) Target Output & (d) Regularization &   (e) Counterparts \\ 
\hline \hline

APC~\cite{chung2019unsupervised, chung2020generative,chung2020improved} & $X_{[1,t-1]}$  & $x_{t+c},c\geq0$ & VQ~\cite{chung20e_interspeech} &  GPT~\cite{alex2018GPT,alex2019GPT2,brown2020gpt3} \\ \hline

%$X_{[1,t-1]}$  & $x_{t+c},c\geq0$ & VQ & VQ-APC~\cite{chung20e_interspeech} & -  \\

DeCoAR~\cite{ling2020deep} & $X_{[1,t-1]}$, $X_{[t+k+1,T]}$ &  $X_{[t,t+k]}$ & VQ~\cite{ling2020deep} &  ELMo~\cite{Matthew2018ELMO} \\
\hline

\tabincell{l}{ Mockingjay~\cite{liu2020mockingjay} \\ MPC~\cite{jiang2019improving,jiang2021further} \\ AALBERT~\cite{chi2021aalbert} \\
DeCoAR 2.0~\cite{ling2020decoar} } &
$X_{-[t,t+k]}$ & $X_{[t,t+k]}$ & VQ~\cite{ling2020decoar} &  \tabincell{l}{BERT ~\cite{jacob2019BERT}\\RoBERTa~\cite{Liu2019RoBERTa}} \\ \hline

pMPC~\cite{yue2021pMPC}   &
\tabincell{l}{$X_{-[t,t+k^\prime]}$\\($X_{[t,t+k^\prime]}$: phoneme-level segment)} & $X_{[t,t+k^\prime]}$ & - &  SpanBert~\cite{joshi-etal-2020-spanbert} \\ \hline

speech-XLNet~\cite{song20d_interspeech} &
$X^*_{[1:t-1]}$ & $x^*_t$ & - &  XLNet~\cite{Yang2019LNet}   \\ \hline

\tabincell{l}{NPC~\cite{liu21l_interspeech}\\MPE~\cite{liu2020masked}} &
\tabincell{l}{$X_{-[t,t+k]}$\\(mask on hidden layer)} & $X_{[t,t+k]}$ & - &   - \\ \hline

\tabincell{l}{BMR~\cite{wang2020unsupervised}\\TERA~\cite{liu2021tera}} &
$X_{-[t_1:t_2]}^{-[f_1,f_2]}$ & $X$ & dropout~\cite{luo2021drop} &  -  \\ \hline

\cite{chorowski2019unsupervised} &
$S$ & $S$ & \tabincell{l}{VQ-VAE~\cite{chorowski2019unsupervised}\\time-jitter regularizer~\cite{chorowski2019unsupervised}} & 
\tabincell{l}{\cite{van2017neural,VQVAE2} \\ (just name a few)} \\ \hline

\cite{tagliasacchi2019self, tagliasacchi2020pre} & 
$X_{[t-k,t-1]},X_{[t+k+1,t+2k]}$ & $X_{[t,t+k]}$  & - & CBoW~\cite{mikolov2013efficient}  \\ \hline

\cite{tagliasacchi2019self, tagliasacchi2020pre} &
$X_{[t,t+k]}$ & $X_{[t-k,t-1]},X_{[t+k+1,t+2k]}$ & - &  Skipgram~\cite{mikolov2013efficient} \\ \hline

\cite{quitry2019learning}  &
Magnitude of $X$ & Phase of $X$ & - &  - \\ \hline

TemporalGap~\cite{tagliasacchi2019self, tagliasacchi2020pre} &
$X_{[t_1,t_1+k]}$, $X_{[t_2,t_2+k]}$ & $|t_1 - t_2|$ & - &  -  \\ \hline

\cite{Carr2021Ranking} &
$X^*$ & correct order & - &  Jigsaw~\cite{noroozi2016unsupervised} \\ 
\hline

PASE~\cite{pascual2019learning} &
$S$ & $X$: LPS, MFCC, Prosody & - &  -  \\ \hline

PASE+~\cite{ravanelli2020multi} & 
$S^*$ & \tabincell{l}{$X$: LPS, MFCC, Prosody, \\ fbank, gammatone} & -  &  -  \\ \hline
\end{tabular}
}
\end{table*}
\end{comment}

### 2023-03-04 chapters/paper_review/stability.tex
Add review paper (WIP)

%!TEX root = ../thesis.tex

\begin{table}[ht]
    \centering
    \resizebox{\linewidth}{!}{
    \begin{tabular}{cc|ccc}
        \toprule
        \multirow{2}{*}{feature} & \multirow{2}{*}{C} 
        & \multicolumn{3}{c}{PNMI (mean $\pm$ std) with K-means Training Size = } \\
        & & 1h & 10h & 100h \\
        \midrule\midrule
        \multirow{2}{*}{MFCC}
        & 100 & 0.251 $\pm$ 0.001 & 0.253 $\pm$ 0.001 & 0.253 $\pm$ 0.001 \\
        & 500 & 0.283 $\pm$ 0.001 & 0.285 $\pm$ 0.000 & 0.287 $\pm$ 0.001 \\
        \midrule
        \multirow{2}{*}{\textsc{Base}-it1-L6}
        & 100 & 0.563 $\pm$ 0.012 & 0.561 $\pm$ 0.012 & 0.575 $\pm$ 0.008 \\
        & 500 & 0.680 $\pm$ 0.005 & 0.684 $\pm$ 0.003 & 0.686 $\pm$ 0.004 \\
        \bottomrule
    \end{tabular}
    }
    \caption{Stability of K-means as an unsupervised unit discovery algorithm with respect to different features, numbers of clusters, and training data sizes. PNMI stands for phone-normalized mutual information.}
    \label{tab:stability}
\end{table}

### 2023-03-16 appendices/appendix_benchmarking.tex
Minor updates

@@
     \caption{
-    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit linear PCM}. The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
+    Model likelihoods on TIMIT represented as a 16 bit linear PCM. The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
     }
@@
     \caption{
-    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}.
+    Model likelihoods on TIMIT represented as a \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}.
     }
@@
     \caption{
-    Model likelihoods on \textbf{TIMIT} represented as \textbf{globally normalized 16 bit linear PCM}.
+    Model likelihoods on TIMIT represented as globally normalized 16 bit linear PCM.
 Contrary to the other likelihoods reported in this paper, here they are given in units of nats and obtained by summing the likelihood over time and over all examples in the dataset and dividing by the total number of examples.

### 2023-03-16 appendices/appendix_benchmarking_workshop.tex
Minor updates

@@
     \caption{
-    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit linear PCM}, obtained by different latent variable models and compared to autoregressive baselines all using a discretized mixture of logistics with 10 components as output distribution.
+    Model likelihoods on TIMIT represented as a 16 bit linear PCM, obtained by different latent variable models and compared to autoregressive baselines all using a discretized mixture of logistics with 10 components as output distribution.
 Likelihoods are given in units of bits per frame (bpf) and obtained by normalizing the total likelihood of each sequence with the individual sequence length and then averaging over the dataset.
@@
     \caption{
-    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}, obtained by different latent variable models and compared to autoregressive baselines all using a discretized mixture of logistics with 10 components as output distribution.
+    Model likelihoods on TIMIT represented as a \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}, obtained by different latent variable models and compared to autoregressive baselines all using a discretized mixture of logistics with 10 components as output distribution.
 Likelihoods are given in units of bits per frame (bpf) and obtained by normalizing the total likelihood of each sequence with the individual sequence length and then averaging over the dataset.
@@
     \caption{
-    Model likelihoods on \textbf{TIMIT} represented as \textbf{globally normalized 16 bit linear PCM}.
+    Model likelihoods on TIMIT represented as globally normalized 16 bit linear PCM.
 Likelihoods are given in units of nats and obtained by summing the likelihood over time and over all examples in the dataset and dividing by the total number of examples.

### 2023-03-16 chapters/paper_review/01_introduction.tex
Change reference keys for review paper (WIP)

@@
 
-%Hung-yi Lee: I think maybe we need a basic frame image. I happen to have an image that can be used here.
 \begin{figure}
@@
 
-% {\color{blue} Reviewer: Daniel, KL, SW}\\
-% {\color{blue} Abdo -- this section is a work-in-progress}\\
-
-% {\color{red} Hung-yi: Last time we have decided to use "pre-train" (with "-")} -- Done\\
-
-% {\color{red} Hung-yi: Fine-tune? or finetune?} --Fine-tune Done\\
-
-% {\color{red} Hung-yi: Fig. 1 or Figure 1} -- Fig. 1  Done\\
-
-% \abdo{comment from Shinji: bring references to intro.}
-
-% \kl{Comments from Nov. 10 meeting: 
-% \begin{itemize}
-%     \item \abdo{added.} would be good to define representation learning a bit more thoroughly in the intro, and mention that we are considering both discrete and continuous representations.
-%     \item \abdo{added.} it would also be good to mention and broadly define "self-supervised" in the intro (as of now it's first mentioned in Section~\ref{sec:thirdwave}), and mention that the term is now used (and we use it) for some techniques that in the past would have just been called "unsupervised".
-% \end{itemize}}
-
-% \kl{New comments: 
-% \begin{itemize}
-%     \item It would be good to include somewhere in the paper a summary of tasks for which SSL pre-trained models have the SOTA performance.  Maybe it is in the paper already but I don't see it.
-% \end{itemize}}
-
 Over the past decade, deep learning approaches have revolutionized speech processing
@@
 transformation, offering impressive gains for scenarios rich in labeled
-data~\cite{lecun2015deep, hinton2012deep, bourlard2012connectionist}. 
+% data~\cite{lecun_deep_2015}. 
+% data~\cite{lecun_deep_2015, hinton_deep_2012}. 
+% data~\cite{bourlard_connectionist_1994}. 
+% data~\cite{lecun_deep_2015,bourlard_connectionist_1994}. 
+data~\cite{lecun_deep_2015,hinton_deep_2012,bourlard_connectionist_1994}. 
 Paradoxically, this heavy reliance on supervised learning has restricted progress in
@@
 investment.
-%Paradoxically, the heavy reliance on supervised learning bottlenecked the development of novel industrial speech applications and restricted progress in languages and domains that do not entertain the same level of labeling investment.
 
@@
 unpaired audio-only data to open up new industrial speech use-cases and
-low-resource languages~\cite{Kemp1999, csl01_limsi, ma_bbn_06}. Inspired by how
+low-resource languages~\cite{kemp_unsupervised_1999, lamel_lightly_2002, ma_unsupervised_2006}. Inspired by how
 children learn their first language through listening and interacting with
@@
 information. These learned representations are then used for target downstream
-applications requiring a minimal number of labeled data~\cite{Hinton_2007,
-LeCun06atutorial, bengio_representation_2013}. 
+applications requiring a minimal number of labeled data~\cite{hinton_learning_2007,
+lecun_tutorial_2006, bengio_representation_2013}. 
 Formally, representation learning refers to algorithms for extracting latent
@@
 input~\cite{bengio_representation_2013}.
-% Previous research efforts explored latent representations that are binary \cite{}, discrete \cite{} and continuous \cite{}. 
-% These latent representations are generally used as inputs to a supervised predictor or a generation module.
-The whole network generating these representations is frequently used to initialize a supervised classifier~\cite{hinton_2006}.
 
-%Learning representations approaches generally fall within \textit{unsupervised learning}, which refers to the family of machine learning methods that discover naturally occurring patterns in the training samples without any pre-assigned labels or scores associated with these samples~\cite{jordan_2015}.
 Representation learning approaches are generally considered examples of \textit{unsupervised
@@
 naturally occurring patterns in training samples for which there are no pre-assigned
-labels or scores~\cite{jordan_2015}. 
+labels or scores~\cite{jordan_machine_2015}. 
 The term ``unsupervised'' is used to distinguish this family of methods from
@@
 with labels to guide learning using a larger volume of unlabeled samples.
-Examples of unsupervised learning techniques include k-means clustering~\cite{vq}, mixture models~\cite{MoE}, autoencoders~\cite{hinton_94},
-and non-negative matrix factorization~\cite{nmf}. 
+Examples of unsupervised learning techniques include k-means clustering~\cite{gray_vector_1984}, mixture models~\cite{jordan_hierarchical_1994}, autoencoders~\cite{hinton_autoencoders_1994},
+and non-negative matrix factorization~\cite{lee_learning_1999}. 
 \textit{Self-supervised learning} (SSL) is a fast-growing subcategory of
@@
 representations useful for 
-% one or many 
 downstream tasks.
@@
 representation from the frozen model, or fine-tune the entire pre-trained model
-in a supervised phase~\cite{hinton_2006}. 
-%e.g., Automatic Speech Recognition (ASR) and Speaker Identification (SID) are two example downstream applications in \cref{fig:SSL_framework}. 
+in a supervised phase~\cite{hinton_reducing_2006}. 
 Automatic speech recognition (ASR) and speaker identification (SID) are 
 examples of downstream applications in \cref{fig:SSL_framework}.
-% There are some task-specific labeled data for each downstream task, and there is a downstream model on top of the upstream model.
-The labeled data is used to train the corresponding downstream models and optionally fine-tune upstream models.
-% If the pre-training stage only leverages unlabeled data, the framework in \cref{fig:SSL_framework} is named self-supervised learning (SSL), which is the focus of this review paper.
-% Pre-training with labeled data is out of the scope.
-Before the term self-supervised learning was widely used, pre-training without any labeled data was usually known as an \texit{unsupervised learning} approach.
 
-%Lee: in the introduction, we may need a specific example. Below is just an example.
-%With unsupervised ASR, imagine that in the future, when an indigenous language family buys an intelligent assistant, even if the assistant does not support the language at the beginning, it would automatically learn to transcribe the new language after learning.
-
-% The speech representation learning approaches evolved over the past few decades through at least three major stages, as discussed in \cref{sec:thirdwave}; static and temporal clustering algorithms \cite{}, energy-based neural models \cite{}, and more recently, self-supervised learning techniques \cite{}.
-% Across all these three waves of representation learning, three main desirable characteristics emerged; features need to be disentangled, invariant, and hierarchical. 
 It is considered desirable for learned speech representations to be
@@
 Due to the popularity of SSL, reviews have been published about the
-technology in general~\cite{bommasani2021opportunities,ericsson2021selfsupervised,LiuSSLsurvey} as well as its application to natural language processing (NLP)~\cite{rogers-etal-2020-primer,liu2021pretrain,xia-etal-2020-bert,QiuSSLNLPsurvey} and computer vision (CV)~\cite{JingSSLCVsurvey}.
-Recently, a brief overview with a general focus on speech representation learning was published \cite{borgholt_22}.
+technology in general~\cite{bommasani_opportunities_2021,ericsson_selfsupervised_2022,liu_selfsupervised_2021} as well as its application to natural language processing (NLP)~\cite{rogers_primer_2020,liu_pretrain_2021,xia_which_2020,qiu_pretrained_2020} and computer vision (CV)~\cite{jing_selfsupervised_2021}.
+Recently, a brief overview with a general focus on speech representation learning was published \cite{borgholt_brief_2022}.
 However, none of these overviews focus exclusively on SSL for speech processing.
 Since the speech signal differs greatly from image and text inputs, many  theories  and technologies have been developed to address the unique challenges of speech.
-One review addresses speech representation learning based on deep learning models~\cite{latif2021deep}, but does not address recent developments in self-supervised learning.
+One review addresses speech representation learning based on deep learning models~\cite{latif_deep_2021}, but does not address recent developments in self-supervised learning.
 This motivates this overview of speech SSL.
-
-%For survey papers on TL to speech processing, please check the reference~\cite{survey_speech_TL}.
-%The representation learning methods covered in this review are primarily task-agnostic methods that only benefit from supervised data at the fine-tuning stage, as shown in \cref{}.
-%Lee: Similar sentence may have been mentioned in the paragraph describing the figure.
 

### 2023-03-16 chapters/paper_review/02_three_waves.tex
Change reference keys for review paper (WIP)

@@
 
-% {\color{blue} Reviewer: KL, KK, Daniel }\\
 In this section we present the historical background of the current surge in
@@
 
-%%KK: if you adopt the paragraph below I suggest renaming this as "Features based on generative models" or something
 \subsection{Clustering and mixture models}
-% {\color{blue} Tara}\\
-
-%\kl{to people who are familiar with modern rep learning work, this first wave may not seem like rep learning at all.
-readers might also wonder why we are talking about what look like standard supervised ASR in the 3rd paragraph below.
-one option is to describe it is as unsupervised learning of latent variables (Gaussian components, HMM states) within a supervised model.
-(something like that -- I haven't thought about the wording carefully)}
-%KK: suggesting the paragraph below for inclusion here
-%The initial wave of speech representation {\em learning} focused on extracting speech features from generative models. The main objective was
-%to render  the knowledgpe learned by generative models accessible to
-%discriminative downstream classifiers, or to map variable-length sequences into  fixed-length representations.
-%The generative model, typically a global Gaussian mixture model (GMM) or a Hidden Markov Model (HMM) was trained by
-%by maximizing the likelihood of the data given the
-%model, which could be done in either an unsupervised or a supervised way. Feature vectors were then derived from the parameters of the trained model. In
-%the case of {\em Fisher vectors}, the features were the normalized gradients of the log-likelihood
-%with respect to the model parameters (mixture weights, means, variances) of the Gaussian mixtures.
-%An extension of this approach (likelihood ratio score space) used the derivative of the log likelihood ratio of	two models,
-%e.g., a background model and a foreground model.
-%Examples of their use in speech processing include speech recognition \cite{smith01,venkata03} and speaker recognition \cite{wan03}.
-%Subsequent techniques in speaker and language verification \cite{dehak11a,dehak11b} similarly extracted parameters (concatenated means) from trained background GMMs as representations that were then %combined with low-rank projections of speaker/session or language specific vectors.
 
@@
 Early work used simple clustering methods. For example, in work such
-as \cite{Rabiner1979,Wilpon1985}, word patterns were clustered
+as \cite{rabiner_considerations_1979,wilpon_modified_1985}, word patterns were clustered
 semi-automatically using techniques such as k-means, after which isolated words
 were recognized by finding the training cluster closest to 
-  the test data.   % AMH: check
+  the test data.
 
 Through time, modeling techniques improved such that subword units were
-represented by Gaussian mixture models (GMMs) \cite{Gauvain1994}, which facilitated
+represented by Gaussian mixture models (GMMs) \cite{gauvain_maximum_1994}, which facilitated
 the modeling of more variability in the input data. GMMs were first built for
 context-independent phonemes; state-clustering 
-algorithms~\cite{Young1994} then resulted in GMMs for context-dependent phonemes. 
+algorithms~\cite{young_state_1994} then resulted in GMMs for context-dependent phonemes. 
 Each latent component of these mixture models acted as a template of a
 prototypical speech frame, 
-% leading to their deficiency to deal with 
-  making it difficult to handle             % AMH: check
+making it difficult to handle 
 large volumes of data with diverse characteristics. 
-Furthermore, dynamical models like hidden Markov models (HMMs)~\cite{Bahl1986}
+Furthermore, dynamical models like hidden Markov models (HMMs)~\cite{bell_adaptation_2021}
 allowed for the processing of continuous speech rather than just isolated word
@@
 models, e.g., a background model and a foreground model. Examples of their use
-in speech processing include speech recognition~\cite{smith01,venkata03} and
-speaker recognition~\cite{wan03}. Subsequent techniques in speaker and language
-verification~\cite{dehak11a,dehak11b} similarly extracted parameters
+in speech processing include speech recognition~\cite{smith_speech_2001,venkataramani_support_2003} and
+speaker recognition~\cite{wan_svmsvm_2003}. Subsequent techniques in speaker and language
+verification~\cite{dehak_frontend_2011,dehak_language_2011} similarly extracted parameters
 (concatenated means) from trained background GMMs as representations that were
@@
 
-%{\color{blue} Tara}\\
 More recently, representation learning has seen a shift of focus towards neural models,
@@
 Examples of early techniques include restricted Boltzmann machines
-(RBM)~\cite{hinton_2006}, denoising autoencoders~\cite{DAE}, noise contrastive
-estimation (NCE)~\cite{gutmann2012noise}, sparse coding~\cite{Olshausen1996,
-sparse_lee, sivaram2010sparse}, and energy-based methods~\cite{Ranzato2007}.
+(RBM)~\cite{hinton_reducing_2006}, denoising autoencoders~\cite{vincent_stacked_2010}, noise contrastive
+estimation (NCE)~\cite{gutmann_noisecontrastive_2012}, sparse coding~\cite{olshausen_emergence_1996,
+lee_efficient_2006, sivaram_sparse_2010}, and energy-based methods~\cite{ranzato_unified_2007}.
 Many of these techniques have also been applied to CV and NLP problems, which
 provided inspiration for their application to speech.
-%Examples of techniques include Restricted Boltzmann Machine (RBM) \cite{hinton2012practical,Srivastava2013}, Neural Auto Encoders \cite{Lecun2012}, Noise Contrastive Estimation (NCE) \cite{chen2015recurrent},  Sparse Coding \cite{sivaram2010sparse} and Energy-based method \cite{Ranzato2007}.
-Note that many of these techniques were also tried for CV and NLP problems, which provided inspiration to try them for ASR as well.
 
@@
 example, approximate training methods such as contrastive divergence
-\cite{hinton_cd_2002} were a practical technique to 
-% make       RBM training more efficient. 
-  streamline RBM training.                   % AMH: check
+\cite{hinton_training_2002} were a practical technique to streamline RBM training. 
 Furthermore, deep networks had non-convex objective functions, which
@@
 
-% http://proceedings.mlr.press/v2/ranzato07a/ranzato07a.pdf, 
-%https://auai.org/uai2013/prints/papers/166.pdf, % https://www.cs.cmu.edu/~rsalakhu/papers/neco_DBM.pdf
-% problems with this trend: approximate gradient, failure modes, long training time, no clear wins in downstream applications.
-
 \subsection{Learning through pretext task optimization}
 
-% {\color{blue} Hung-yi + Abdo}\\
-
-%Note on 12/09: We will mainly describe the speech part but link to CV and NLP. 
-%discuss instance classification, contrastive losses,\\
-
-%Currently, the network learns a function that maps input to desired representations by solving a \textit{pretext} task.
-%After pre-training, learned representation models could be applied to downstream tasks through feature-based representation extraction or fine-tuning as part of the downstream model.  
-%By optimizing carefully designed pretext tasks, neural models can capture latent representations which generalize well across a wide range of downstream applications, usually under few-shot or even zero-shot conditions.
 A more recent trend is learning networks that map the input to desired
@@
 For example, in NLP, a representation model is usually assessed on GLUE,
-which comprises nine tasks~\cite{GLUE}, whereas in speech, a representation model can be
-evaluated on SUPERB, which comprises ten tasks~\cite{yang21c_interspeech}, 
+which comprises nine tasks~\cite{wang_glue_2018}, whereas in speech, a representation model can be
+evaluated on SUPERB, which comprises ten tasks~\cite{yang_superb_2021}, 
 as described in detail in \cref{sec:benchmark}.
@@
 via pretext tasks for restoring the true color of black-and-white
-images~\cite{colorizing}, joint learning of latent representations and their
-cluster assignments~\cite{deepcluster}, and the prediction of the relative positions of
-image patches~\cite{context_pred}. Other popular approaches include variational
-autoencoders (VAEs)~\cite{vae, rezende2014stochastic}. While typical autoencoders learn data
+images~\cite{zhang_colorful_2016}, joint learning of latent representations and their
+cluster assignments~\cite{caron_deep_2018}, and the prediction of the relative positions of
+image patches~\cite{doersch_unsupervised_2016}. Other popular approaches include variational
+autoencoders (VAEs)~\cite{kingma_autoencoding_2014, rezende_stochastic_2014}. While typical autoencoders learn data
 representations using unsupervised objectives by reconstructing the input
 after passing it through an information bottleneck, VAEs estimate a neural model of a probability density function (pdf) that approximates the unknown “true” distribution of the observed data, for which we only have access to independently identically distributed (iid) samples.
-It is also important to mention dynamical VAEs \cite{Girin2021}, which is an extension of VAE for sequential data such as speech.
+It is also important to mention dynamical VAEs \cite{girin_dynamical_2021}, which is an extension of VAE for sequential data such as speech.
 
@@
 in a sentence to predict the next token such as in
-ELMo~\cite{peters2018deep}, the GPT
-series~\cite{alex2019GPT2}, and %brown2020gpt3 alex2018GPT,
-Megatron~\cite{shoeybi2020megatronlm}, or predicting the masked tokens in a
+ELMo~\cite{peters_deep_2018}, the GPT
+series~\cite{radford_language_2019}, and 
+Megatron~\cite{shoeybi_megatronlm_2020}, or predicting the masked tokens in a
 sentence such as with the bidirectional encoder representations from Transformers
-(BERT) series~\cite{jacob2019BERT,Liu2019RoBERTa}. %,albert
-%ELMo~\cite{peters2018deep}, GPT series~\cite{alex2018GPT,alex2019GPT2,brown2020gpt3}, Megatron~\cite{shoeybi2020megatronlm} are gigantic LM which predict the next token. 
-%BERT~\cite{jacob2019BERT}, RoBERTa~\cite{Liu2019RoBERTa}, ALBERT~\cite{albert} masked the input token, and the models are learned to predict the masked tokens. 
+(BERT) series~\cite{devlin_bert:_2018,liu_roberta_2019}. 
 Another common pretext task in the third wave is contrastive
-learning~\cite{oord2018representation}, in which a model learns to identify a
+learning~\cite{oord_representation_2018}, in which a model learns to identify a
 target instance from a set of negative samples. 
 This approach has become especially popular in the
-CV context~\cite{pmlr-v119-chen20j,he2020momentum,chen2020improved,SwAV}.
+CV context~\cite{chen_simple_2020,he_momentum_2020,chen_improved_2020,caron_unsupervised_2020}.
 In this survey, we will mainly focus on techniques for pretext task
@@
 label a large volume of speech-only data, which is then used to augment the
-initial labeled data to train a student model~\cite{Kemp1999, csl01_limsi,
-ma_bbn_06, hari_1mhour}. PL has been successful and widely adopted in the
+initial labeled data to train a student model~\cite{kemp_unsupervised_1999, lamel_lightly_2002,
+ma_unsupervised_2006, parthasarathi_lessons_2019}. PL has been successful and widely adopted in the
 speech community since the 1990s. Other proposed variations of PL include
@@
 models with more parameters than their original teachers to capture the
-complexities in vastly larger speech-only data~\cite{park2020improved,
-xu2020iterative, xiao_scaling_2021}. 
+complexities in vastly larger speech-only data~\cite{park_improved_2020,
+xu_iterative_2020, xiao_scaling_2021}. 
 Both SSL and PL leverage unlabeled speech-only data.
@@
 pre-training speech models. TL transfers knowledge captured by models
-trained on one task to different but related tasks~\cite{caruana1997multitask}. 
+trained on one task to different but related tasks~\cite{caruana_multitask_1997}. 
 The past few decades have seen active research on TL and its
@@
 Multilingual and cross-lingual supervised models have proven superior in
-low-resource speech recognition tasks~\cite{Cui2015MultilingualRF}.
+low-resource speech recognition tasks~\cite{cui_multilingual_2015}.
 SSL can be regarded as a type of TL because knowledge learned from pre-training
@@
 This survey paper focuses on SSL, and not all TL technologies for speech. 
-One survey indeed addresses TL for speech processing~\cite{survey_speech_TL}
+One survey indeed addresses TL for speech processing~\cite{bell_adaptation_2021}
 but does not include current SSL technologies for speech.
-
-%Then contrastive loss is applied~\cite{oord2018representation}, in which the model learns to classify the positive instances from a set of unrelated negative samples.
-SimCLR~\cite{pmlr-v119-chen20j}, MoCo~\cite{he2020momentum,chen2020improved} and SwAV~\cite{SwAV} are representative ones.
-%These approaches learn representations from visual inputs by maximizing agreement between differently augmented views of the same sample while minimizing the agreement between different samples. 
-%Then there is a series of work trying to remove the requirement of negative examples. 
-%BYOL~\cite{BYOL} trains a student network to predict the representations of the teacher network.
-The parameter of the teacher network is the exponential moving average of the previous student network.
-In such a scenario, negative examples are not required.
-Then SimSiam~\cite{Chen_2021_CVPR} improved BYOL by removing the moving average and found that good representation can still be learned in this way.
-%In Barlow twins~\cite{BarlowTwins}, the cross-correlation matrix between the representations of two identical encoders fed with distorted versions of a sample is measured.
-The encoder learns to make the cross-correlation matrix as close to the identity matrix as possible.
-%VICReg~\cite{bardes2021vicreg}
-
-%The 3rd wave begins from NLP and CV communities. %Is it true?
-%Below the progress of self-supervised learning in CV  and NLP are briefly summarized. %(However, we have to consider how to connect this part with other parts.)
-
-%In CV, some earlier approaches make some pretexts to train instance classifiers, and the last hidden layers of the classifiers are used as the representations in the downstream tasks.
-Below are some representative approaches:
-%\begin{itemize}
-%\item Position~\cite{Doersch_2015_ICCV}: predicting the relative position between two random patches from one image.
-A model needs to understand the spatial context of objects to tell the relative position between parts.
-%\item Jigsaw~\cite{noroozi2016unsupervised}: The model is trained to place shuffled patches back to the original locations.
-%\item Rotation~\cite{gidaris2018unsupervised}: The model is trained to predict which rotation has been applied.
-%\end{itemize}
-
-%How to deal with negative examples is critical in this series.
-SimCLR~\cite{pmlr-v119-chen20j} uses a large batch size to obtain a lot of negative examples, while MoCo uses a momentum encoder to make the negative samples’ representation consistent.
-
-%In ELECTRA~\cite{clark2020electra}, given a sentence with some tokens replaced, the ELECTRA model predicts whether a token is generated or not. 
-%XLNet~\cite{XLNet}
-%MASS~\cite{MASS}, BART~\cite{lewis-etal-2020-bart}, and T5~\cite{T5} are seq2seq models.
-The input is disturbed by different approaches, and the seq2seq model learns to reconstruct the disturbed input.
-%(There are speech T5, should we mention this work?)
-%The contrastive approaches have been applied at the sentence level in NLP~\cite{gao-etal-2021-simcse,kim-etal-2021-self,yan-etal-2021-consert,giorgi-etal-2021-declutr}.
-
-%Turing NLG only has webpage: https://www.microsoft.com/en-us/research/blog/turing-nlg-a-17-billion-parameter-language-model-by-microsoft/
-%I don't know how to cite it.
-
-%ref: http://speech.ee.ntu.edu.tw/~tlkagk/courses/DLHLP20/BERT%20train%20(v8).pdf

### 2023-03-16 chapters/paper_review/03_approaches.tex
Change reference keys for review paper (WIP)

@@
 \section{Speech Representation Learning Paradigms} \label{sec:approach}
-% {\color{blue} Reviewer: Hung-yi, Daniel , Abdo}\\
-% {\color{blue} Hung-yi + Abdo}\\
-%Start by discussing central challenges for self-supervised representation learning in speech: 1)Can't do instance classification in variable-length sequences.
-2)No pre-defined vocabulary.
-3)No pre-defined segmentation.
-\\
-%New approaches offer deep contextualization of learned representations. \\
-% {\color{red} first version for a list of symbols to use -- let's keep updating it as needed -- Abdo}\\
 Due to the characteristics of speech, SSL pretext tasks developed for CV
@@
 speech. 
-% However, the sequence representations of speech are usually much longer than those of text and they contain no natural boundaries at which they can be split into segments.
 In NLP, morpheme-like tokens are widely used as sequence units in
@@
 input, usually covering a paragraph including several sentences. 
-% In speech processing, the audio signal is recorded as a waveform represents the amplitude of pressure. 
 However, speech signals consist of sound pressure measurements with thousands
@@
 frames per second.
-%This raw signal is often converted to a spectral representation where each frame represents 20-25 milliseconds of audio computed with overlaps at 10 millisecond strides.
-% a frame is the typical sequence unit in speech, and each frame usually represents a length of 10 milliseconds. 
-% Therefore, an utterance is usually represented as a long sequence with more than a thousand frames. 
 Processing such sequences with typical neural network architectures like
@@
 straightforward to define a finite vocabulary for such tasks.
-%The generative-based pretexts are widely applied in NLP.
-%Because a sentence is composed of tokens, sentence generation can be formulated as a series of classification tasks.  
 However, this idea does not apply to speech modeling because speech
 signals are continuous; 
-% thus, such a vocabulary is unavailable.
-  in this sense there is no such thing as a speech vocabulary.   % AMH: check
+in this sense there is no such thing as a speech vocabulary. 
 \item \textit{Speech processing tasks are diverse.}
@@
 \end{itemize}
-%(Although we list the problems here, we do not highlight the solutions in the following paragraph.)
 
@@
 \textit{contrastive} approaches and \textit{predictive} approaches. 
-%For each category, we provide a brief motivation and then describe prominent models that belong to it. Lastly, we discuss a set of challenges related to the category.
 \Cref{fig:timeline} shows a timeline of the models covered in these sections
@@
 \Cref{table:pretext} summarizes model pretext tasks along within the categories.
-
-%\input{tables/model}
-
-% \begin{center}
-% \begin{tabular}{ |c|c| } 
-%  \hline
-%  $s$ & a single waveform sample \\
-%  $S$ & a sequence of waveform samples \\
-%  $x$ & a single feature vector (MFCC or learned)\\
-%  $X$ & a sequence of feature vectors \\
-%  $\widetilde{X}$ & a data augmented sequence of feature vectors \\
-%  $z$ & a single latent vector \\
-%  $Z$ & a sequence of latent vectors \\
-%  $h$ & a single hidden vector representation \\
-%  $H$ & a sequence of hidden vector representation \\
-%  $m$ & a binary mask for a single time step \\
-%  $M$ & a binary mask for the whole sequence \\
-%  $y$ & a single supervised label \\
-%  $Y$ & a sequence of supervised labels \\
-%  $\tau$ & temperature variable \\ 
-%  $f()$ & an encoding network - use $f_1()$ $f_2()$ if multiple\\
-%  $g()$ & a decoding network \\
-%  $P()$ & Probability \\
-%  $E()$ & Expectation \\
-%  $L()$ & Loss \\
-%  $\mathcal{L}$ & Likelihood \\
-%  \hline
-% \end{tabular}
-% \end{center}
 
@@
 
-%A few other conventions should be noted here. 
 We use $Q = \{q_1,q_2, ..., q_T\}$ to denote representations that are quantized via codebook learning.
@@
 \edit{After pre-training,} the representation model $f(\cdot)$ can be fine-tuned for a downstream task directly or used to extract features which are fed to another model, as visualized in \cref{fig:SSL_framework}.
-It is not uncommon to use the output representation~$H$, but often representations from hidden layers of $f(\cdot)$ are better suited \cite{pasad2021}.
+It is not uncommon to use the output representation~$H$, but often representations from hidden layers of $f(\cdot)$ are better suited \cite{pasad_layerwise_2021}.
 
@@
 	 but find that generative tasks are the most important for downstream
-	 task performance~\cite{pascual2019learning}.}
+	 task performance~\cite{pascual_learning_2019}.}
     \label{fig:timeline}
@@
 
-% {\color{blue} Hung-yi}\\
-
 \subsubsection{Motivation}
 
-%In this category, the pretext task is to generate speech signals given quantized, past, or unmasked time steps.
-To solve the generative-based pretext, the model leverages an encoder $f(\cdot)$ and a decoder $g(\cdot)$.
-Given waveform samples $S$ or its corresponding acoustic feature vectors $X$, the encoder $f(\cdot)$ outputs hidden representations $H$.
-The waveform samples $S$ or acoustic feature vectors $X$ are usually corrupted before entering the encoder $f(\cdot)$.
-In this subsection, we only consider the acoustic features $X$ and ignore waveform samples $S$ for simplicity, but taking waveform samples as input and generating waveform samples is possible.
-Some regularization in the encoder $f(\cdot)$ or decoder $g(\cdot)$ is required to encourage the model to utilize global speech information and prevent naively copying input when reconstructing, especially when the input of $f(\cdot)$ is not corrupted.
-After pre-training, $f(\cdot)$ is used as the representation model, which takes a sequence of acoustic feature vectors $X=\{x_1,x_2, ..., x_T\}$ as input, and outputs hidden contextual representations $H =\{h_1,h_2,...,h_T\}$ for downstream tasks.
-
-%In this category, the pretext task is to generate, or reconstruct, the input data based on some limited view.
-This includes predicting future inputs from past, masked from unmasked, or the original from some other corrupted view.
-Models that solve a generative pretext task typically consist of an encoder $f(\cdot)$ and a decoder $g(\cdot)$.
-Given an acoustic input $X =\{x_1,x_2, ..., x_T\}$, the encoder $f(\cdot)$ outputs a representation $H =\{h_1,h_2,...,h_T\}$.
-The decoder $g(\cdot)$ takes $H$ as input to generate parts of the original acoustic input.
-The input $X$ may be either the raw waveform samples or a sequence of spectral feature vectors.
-In the following, we will not distinguish between the two, but in practice, both are viable options.
-After pre-training, $f(\cdot)$ is used as the representation model.
-The model can be fine-tuned for a downstream task directly or used to extract features which are fed to another model, as visualized in \cref{fig:SSL_framework}.
-It is common to use the output representations $H$, but often representations from hidden layers of $f(\cdot)$ are more suited.
-
-%All the generative approaches are summarized in \cref{table:generative}. 
-%There is little work generating waveform samples $S$ because generating them is computationally expensive.
 In this category, the pretext task is to generate, or reconstruct, the input data based on some limited view.
@@
 Note that this differs from generative models, which learn \ci{distributions that allow} to sample new data.
-% It is important to note that many of the approaches discussed in this subsection are ''reconstructive``, i.e., they focus on reconstructing the input signal rather than learning a probabilistic model for sampling novel sequences\footnote{Generative self-supervised models are not the same as generative probabilistic models that permit sampling new data.}.
-
-%\footnote{The lengths of $X$ and $H$ are not necessary to be equal.
-Usually, $H$ is shorter than $X$.
-For simplicity of discussion, we assume the lengths of $X$ and $H$ are the same in this subsection.} % Any one care about this?
 
@@
 
-%\subsubsection{Autoregressive reconstruction}
-%\subsubsection{Autoregressive Prediction}
-%\label{sssec:autoregressive}
-
 \paragraph{Autoencoding} 
-Since their introduction in the mid-1990s~\cite{hinton_94}, autoencoders (AEs)
+Since their introduction in the mid-1990s~\cite{hinton_autoencoders_1994}, autoencoders (AEs)
 have played an essential role in learning distributed latent representations of
@@
 learned representations.
-% An AE converts an input utterance $X$ to a sequence of representations $H$ and then uses a decoder network to convert $H$ into an approximate reconstruction of the input $X$.
-Regularization is required to learn general and efficient latent representations and to avoid learning a trivial identity mapping.
 For instance, denoising autoencoders (DAEs) learn latent representations by
-reconstructing from input corrupted by noise~\cite{DAE}. 
-\edit{The Variational Autoencoder (VAE) is a probabilistic version of the AE which defines the latent representation via a posterior distribution over stochastic latent variables \cite{vae,rezende2014stochastic}.
-VAEs have been applied to speech in numerous works \cite{chung_recurrent_2015, fraccaro_sequential_2016, hsu2017learning, hsu2017unsupervised, aksan_stcn_2019}}.
+reconstructing from input corrupted by noise~\cite{vincent_stacked_2010}. 
+\edit{The Variational Autoencoder (VAE) is a probabilistic version of the AE which defines the latent representation via a posterior distribution over stochastic latent variables \cite{kingma_autoencoding_2014,rezende_stochastic_2014}.
+VAEs have been applied to speech in numerous works \cite{chung_recurrent_2015, fraccaro_sequential_2016, hsu_learning_2017, hsu_unsupervised_2017, aksan_stcn_2019}}.
 The vector-quantized variational autoencoder (VQ-VAE) is another model in this
-category~\cite{vqvae};
-% The first method to sucessfully learn meaningful discrete latent representations was the Vector Quantised Variational Autoencoder (VQ-VAE)~\cite{vqvae}. 
-it extends the original VAE~\cite{vae} with a novel
+category~\cite{oord_neural_2018};
+it extends the original VAE~\cite{kingma_autoencoding_2014} with a novel
 parameterization of the posterior distribution for discrete latent
 representations. 
-\edit{The VQ-VAE has been instrumental in generative speech modelling and recent work on generative spoken language modeling has successfully combined the idea of a discrete latent space with self-supervised learning \cite{SpeechResynthesis_IS21, pGSLM, GSLM}.}
-% Forcing the use of discrete latent representations encourages the model to focus on salient input features, e.g., phonemes in speech, rather than using the capacity to model channel noise and imperceptible local input details.
+\edit{The VQ-VAE has been instrumental in generative speech modelling and recent work on generative spoken language modeling has successfully combined the idea of a discrete latent space with self-supervised learning \cite{polyak_speech_2021, kharitonov_textfree_2021, nguyen_generative_2022}.}
 
@@
 This operation is non-differentiable and the \edit{gradients of the loss with respect to the encoder parameters} must be obtained by approximation.
-In the VQ-VAE this is done using the straight-through estimator~\cite{bengio2013estimating}, i.e., the gradients \edit{with respect to the} encoder output are taken to be equal to those \edit{with respect to the} decoder input \ci{(i.e., the quatization step is ignored)}.
+In the VQ-VAE this is done using the straight-through estimator~\cite{bengio_estimating_2013}, i.e., the gradients \edit{with respect to the} encoder output are taken to be equal to those \edit{with respect to the} decoder input \ci{(i.e., the quatization step is ignored)}.
 Given a learned codebook $A\in\mathbb{R}^{K \times D}$, where $K$ is the codebook size and $D$ is the dimensionality of each codebook vector $a_k$, the quantized representation $q_t$ of $h_t$ is obtained as
@@
 \end{align}
-\edit{The decoder $g(\cdot)$ is an autoregressive model that takes $q_{[1,t]}$ as input to generate $x_t$ \cite{oord2016wavenet}}. 
+\edit{The decoder $g(\cdot)$ is an autoregressive model that takes $q_{[1,t]}$ as input to generate $x_t$ \cite{oord_wavenet_2016}}. 
 Codebook learning is facilitated by a two-term auxiliary loss similar to
@@
 This commitment term is optimized only by the encoder. The
-total VQ-VAE loss \edit{for a single timestep} is
-% \begin{align}
-%     %\mathcal{L} = \log p(x | q) +  \underset{\text{vq}}{\underbrace{\norm{\text{sg}\left[v\right] - A}_2^2}} + \underset{\text{commitment}}{\underbrace{\beta\norm{v - \text{sg}\left[A\right]}_2^2}} \enspace ,
-%     \resizebox{.44\textwidth}{!}{%
-%     $\mathcal{L}_t = \underset{\text{encoder+decoder}}{\underbrace{\log p(x_t | q_{[1,t]})}} +  \underset{\text{codebook}}{\underbrace{\norm{\mathrm{sg}\left[h_t\right] - A}_2^2}} + \beta\underset{\text{encoder}}{\underbrace{\norm{h_t - \mathrm{sg}\left[A\right]}_2^2}} \;,$
-%     }
-%     \label{eq: vector quantization losses}
-% \end{align}
-% \begin{align}
-%     \mathcal{L}_t =&~
-%     \underset{\text{encoder+decoder}}{\underbrace{\log p(x_t | q_{[1,t]})}} + \underset{\text{codebook}}{\underbrace{\text{MSE}\left(\mathrm{sg}\left[H\right], A\right)^2}} + \nonumber\\
-%     & \underset{\text{encoder}}{\underbrace{\beta~\text{MSE}\left(H, \mathrm{sg}\left[A\right]\right)^2}} \;,
-%     \label{eq: vector quantization losses}
-% \end{align}
-\edit{
+total VQ-VAE loss for a single timestep is
+
 \begin{align}
-    \resizebox{.44\textwidth}{!}{%
-    $\mathcal{L}_t = \underset{\text{encoder+decoder}}{\underbrace{\log p(x_t | q_{[1,t]})}} + \underset{\text{codebook}}{\underbrace{\text{MSE}\left(\mathrm{sg}\left[h_t\right], A\right)}} + \underset{\text{encoder}}{\underbrace{\alpha~\text{MSE}\left(h_t, \mathrm{sg}\left[A\right]\right)}} \;,$
-    }
+    \mathcal{L}_t = \underset{\text{encoder+decoder}}{\underbrace{\log p(x_t | q_{[1,t]})}} + \underset{\text{codebook}}{\underbrace{\text{MSE}\left(\mathrm{sg}\left[h_t\right], A\right)}} + \underset{\text{encoder}}{\underbrace{\alpha~\text{MSE}\left(h_t, \mathrm{sg}\left[A\right]\right)}}
     \label{eq: vector quantization losses}
-\end{align}}%
-% \begin{align}
-%     \mathcal{L}_t =&~
-%     \underset{\text{encoder+decoder}}{\underbrace{\log p(x_t | q_{[1,t]})}} + \nonumber\\
-%     & \underset{\text{codebook}}{\underbrace{\frac{1}{KD} \sum_{t=1}^{T}\sum_{k=1}^{K}\sum_{i=1}^{D} \left(\mathrm{sg}\left[h_{t,i}\right] - a_{k,i}\right)^2}} + \nonumber\\
-%     & \underset{\text{encoder}}{\underbrace{\frac{\beta}{KD}\sum_{t=1}^{T}\sum_{k=1}^{K}{\sum_{i=1}^{D} \left(h_{t,i} - \mathrm{sg}\left[a_{k,i}\right]\right)^2}}} \;,
-%     \label{eq: vector quantization losses}
-% \end{align}
+\end{align}%
 \noindent where \edit{$\log p(x_t|q_{[1,t]})$ is a reconstruction likelihood term usually using a categorical distribution,} $\mathrm{sg}[x] = x$ is the so-called stop-gradient operator \edit{which acts as the identity function during the forward pass but \ci{is assumed to have}
 partial derivatives all equal to zero during the backward pass}, $\alpha$ is a scalar hyperparameter, 
-% and we define $\text{MSE}(H,A) = \frac{1}{KD} \sum_{t=1}^{T}\sum_{k=1}^{K}\sum_{i=1}^{D}\left(h_{t,i} - a_{k,i}\right)^2$. 
 \edit{and we define $\text{MSE}(h_t,A) = \frac{1}{KD} \sum_{k=1}^{K}\sum_{i=1}^{D}\left(h_{t,i} - a_{k,i}\right)^2$. 
 The loss for a full sequence is the sum or mean over all $\mathcal{L}_t$.}
-% We note that the subtraction of a matrix from a vector in \cref{eq: vector
-% quantization losses} implies the use of broadcasting.
 
@@
 speech information closely related to phonemes, and are useful for
-applications such as speaker conversion~\cite{chorowski2019unsupervised}.
+applications such as speaker conversion~\cite{chorowski_unsupervised_2019}.
 Vector quantization is \edit{not} exclusive to VQ-VAE but has seen
@@
 targets for the pretext task. We will cover these applications below.
-%We highlight the models using vector-quantization in the column \textbc{{qtz}} in \cref{tab:model-taxonomy}.
 
-\edit{The Gumbel softmax \cite{Gumbel-Softmax} is another frequently used approach for obtaining a discrete representation space, and has also been used for AEs \cite{eloff2019unsupervised}.
-In addition to the approaches discussed above, several other works on speech representation learning take inspiration from the AE framework \cite{zeiler2013rectified, badino2014auto, badino2015discovering, kamper2015unsupervised, renshaw2015comparison, settle2019_a2w}.}
+\edit{The Gumbel softmax \cite{jang_categorical_2017} is another frequently used approach for obtaining a discrete representation space, and has also been used for AEs \cite{eloff_unsupervised_2019}.
+In addition to the approaches discussed above, several other works on speech representation learning take inspiration from the AE framework \cite{zeiler_rectified_2013, badino_autoencoder_2014, badino_discovering_2015, kamper_unsupervised_2015, renshaw_comparison_2015, settle_acoustically_2019}.}
 
@@
     \item DeCoAR 2.0~\cite{ling2020decoar} presents a deep contextualized acoustic representation learning approach with the addition of a vector quantization (VQ) layer.
-    \item In VQ-APC~\cite{chung20e_interspeech}, a VQ layer is used with the APC objective, which imposes a bottleneck and forces the model to learn better representations.
-    \item Two dropout regularization methods, attention dropout and layer dropout, are introduced to TERA~\cite{luo2021drop}. 
+    \item In VQ-APC~\cite{chung_vector-quantized_2020}, a VQ layer is used with the APC objective, which imposes a bottleneck and forces the model to learn better representations.
+    \item Two dropout regularization methods, attention dropout and layer dropout, are introduced to TERA~\cite{luo_drop_2021}. 
 \end{itemize}
@@
 
-Autoregressive predictive coding (APC)~\cite{chung2019unsupervised,
-chung2020generative} \edit{takes inspiration from the classic Linear Predictive Coding (LPC) approach for speech feature extraction~\cite{LPC} and autoregressive language models (LM) for text, where the model learns to predict future information from past}.
-% This approach can be viewed as training a speech version of an LM.
-%Just like an LM for text, this approach uses an encoder model to encode temporal information of past acoustic sequence, and the decoder then predicts future frames.
-%In this approach, given $X=\{x_1,x_2, ..., x_T\}$, $\widetilde{X}$ is the acoustic features up to time step $t$, that is, $X_{[0,t]} = \{x_1,x_2, ..., x_{t}\}$.
+Autoregressive predictive coding (APC)~\cite{chung_unsupervised_2019,
+chung_generative_2020} \edit{takes inspiration from the classic Linear Predictive Coding (LPC) approach for speech feature extraction~\cite{oshaughnessy_linear_1988} and autoregressive language models (LM) for text, where the model learns to predict future information from past}.
 \edit{A function} $f(\cdot)$ reads the input sequence $X_{[1,t]}$ and \edit{outputs a representation sequence} $H_{[1,t]}$.
@@
 and $g(\cdot)$ are jointly learned to minimize {the \ensuremath{L_1} loss between $x_{t+c}$ and its approximation $\hat{x}_{t+c}$}. APC is formulated as 
-%\begin{align}
-    %Z_{[0,t-1]} &= f(X_{[0,t-1]}) \\
-    %\hat{\mathbf{x}}_{t+c} &= g(\mathbf{z}_{t-1}) \label{eq:c} \\
-    %\mathcal{L}_t &= \lVert \hat{\mathbf{x}}_{t+c} - \mathbf{x}_{t+c} \rVert_1\enspace.
-%\end{align}
 \begin{align}
@@
 However, due to the smoothness of the speech signal, neighboring acoustic features are usually similar.
-Depending on the downstream task, we are often interested in learning so-called \emph{slow features} that typically span multiple input frames~\cite{wiskott2002slow}.
+Depending on the downstream task, we are often interested in learning so-called \emph{slow features} that typically span multiple input frames~\cite{wiskott_slow_2002}.
 Even the smallest linguistic units of speech---phonemes---span $0.07$~seconds on average in the English TIMIT dataset~\cite{garofolo_timit_1993},   whereas spectrogram frames $\mathbf{x}_t$ are typically computed at $0.01$ second intervals.
 Thus, simply predicting the next frame constitutes a trivial pretext task for APC; the original work finds that $c=3$ performs well.
-%\begin{equation}
-  %  L = \sum_{X} \sum_{t} d(x_{t+c}, g(h_t)),
-%   L = d(x_{t+c}, g( f(X_{[0,t-1]}) )),
-%\end{equation}
-%where $d(x_{t+c}, g(h_t))$ is the distance between $x_{t+c}$ and  $g(h_t)$. 
-In \cite{chung2020improved}, the APC objective is extended to multi-target
+In \cite{chung_improved_2020}, the APC objective is extended to multi-target
 training. The new objective generates both past and future frames conditioned
 on previous context. 
-In VQ-APC~\cite{chung20e_interspeech}, quantization is used with the APC
+In VQ-APC~\cite{chung_vector-quantized_2020}, quantization is used with the APC
 objective, which imposes an information bottleneck serving as a regularizer.
@@
 and not the entire input.
-DeCoAR~\cite{ling2020deep} combines the bidirectionality of the popular NLP model
-ELMo~\cite{peters2018deep} and the reconstruction objective of APC to alleviate
+DeCoAR~\cite{ling_deep_2020} combines the bidirectionality of the popular NLP model
+ELMo~\cite{peters_deep_2018} and the reconstruction objective of APC to alleviate
 this issue and allow encoding information from the entire input. 
-%, so it is able to learn deep contextualized acoustic representations. 
 It uses a forward LSTM $f_1(\cdot)$ to encode $X_{[1,t]}$ and a backward LSTM
@@
 \end{gather}
-% The last vector in $H_{[1,t]}$, $h_{t}$, and the first vector in $H^\prime_{[t+k,T]}$, $h_{t+k}$, are the input to the decoder $g$ used to predict $X_{[t+1,t+k-1]}$. 
 The input feature vector used in the downstream tasks is the concatenation of
@@
 \paragraph{Masked Reconstruction}
-%\label{sssec:time-only-masked}
-%In wav2vec 2.0~\cite{wav2vec2}, time masking is applied in the latent space. But we will not mention this in this section.
 
-%Basic idea
-Masked reconstruction is largely inspired by the masked language model (MLM) task from BERT~\cite{jacob2019BERT}.
+Masked reconstruction is largely inspired by the masked language model (MLM) task from BERT~\cite{devlin_bert:_2018}.
 During BERT pre-training, some tokens in the input sentences are masked by randomly replacing them by a learned masking token or another input token.
@@
 While we here focus on the models that reconstruct the masked input, it is important to note that masking has also been used extensively for contrastive (\cref{contrastive_approaches}) and predictive (\cref{predictive_approaches}) models.
-%In \cref{tab:model-taxonomy}, the models that use masking are highlighted by the \textbc{{msk}} column.
 
@@
 \edit{The exact masking policy defined by $m(\cdot)$ differs from model to model and will be discussed further below.}
-The function $f(\cdot)$ is typically a Transformer encoder~\cite{liu2020mockingjay,jiang2019improving,liu2020masked}, but recurrent neural networks have also been used~\cite{wang2020unsupervised}.
+The function $f(\cdot)$ is typically a Transformer encoder~\cite{liu_mockingjay_2020,jiang_improving_2019,liu_masked_2020}, but recurrent neural networks have also been used~\cite{wang_unsupervised_2020}.
 \edit{In general, the Transformer encoder architecture has been adopted widely by self-supervised models for speech within all three surveyed categories.} \edit{The function $g(\cdot)$ is usually a linear projection or a multilayer perceptron (MLP)}.
@@
 
-%Time masking 
-%In the time masking approaches~\cite{liu2020mockingjay,jiang2019improving,liu2020masked}, some of the input frames in $X$ are masked to zero or randomly replaced by other frames.
-$X$ with $t$-th to $t+k$-th frames masked is denoted as $X_{-[t,t+k]}$.
-The encoder $f(\cdot)$ is usually a Transformer that encodes the whole utterance.
-It takes $X_{-[t,t+k]}$ as input and generates $H$.
-The decoder $g(\cdot)$ is a prediction head, which use $H_{[t,t+k]}$ to predict $X_{[t,t+k]}$.
-The encoder $f(\cdot)$ and decoder $g(\cdot)$ are jointly learned to minimize the distance between the decoder output and $X_{[t,t+k]}$.
-%\begin{equation}
-%    L = d(X_{[t,t+k]}, g(f( X_{-[t,t+k]} )) ).
-%\end{equation}
-
-The masking policies used in NLP can be adapted to speech by considering a speech \edit{segment} equivalent to a token in a sentence; indeed, the masking strategy of BERT has also been used for speech pre-training~\cite{liu2020mockingjay}.
-%However, taking inspiration from NLP is not sufficient, as speech data warrants domain-specific considerations.
+The masking policies used in NLP can be adapted to speech by considering a speech \edit{segment} equivalent to a token in a sentence; indeed, the masking strategy of BERT has also been used for speech pre-training~\cite{liu_mockingjay_2020}.
 In the standard BERT masking policy, each token is masked independently at random.
-However, for speech, masking a single \edit{sample or spectrogram frame results in a largely} trivial reconstruction task \edit{since, as discussed in paragraph \ref{par:apc}, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore it is common to mask chunks of consecutive frames~\cite{liu2020mockingjay,jiang2021further}.
+However, for speech, masking a single \edit{sample or spectrogram frame results in a largely} trivial reconstruction task \edit{since, as discussed in paragraph \ref{par:apc}, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore it is common to mask chunks of consecutive frames~\cite{liu_mockingjay_2020,jiang_further_2021}.
 
 We can bring the pretext task closer to the NLP equivalent by using a masking policy where the masked regions of the input correspond to linguistic units.
-Instead of just masking a fixed number of consecutive frames, pMPC~\cite{yue2021pMPC} selects masked speech frames according to the phonetic segmentation in an utterance.
+Instead of just masking a fixed number of consecutive frames, pMPC~\cite{yue_pmpc_2021} selects masked speech frames according to the phonetic segmentation in an utterance.
 \edit{However, in order to obtain this segmentation, some labeled data is of course needed.}
 
-Whereas most studies use masking along the temporal dimension of the input, speech can also be masked along the frequency dimension when spectral input features are used~\cite{wang2020unsupervised,liu2021tera}.
-Frequency masking has been shown to improve representations used for speaker classification~\cite{liu2021tera}.
-%The phonetically motivated self-supervised representation learns the speech representation that benefits downstream speech processing tasks.
+Whereas most studies use masking along the temporal dimension of the input, speech can also be masked along the frequency dimension when spectral input features are used~\cite{wang_unsupervised_2020,liu_tera_2021}.
+Frequency masking has been shown to improve representations used for speaker classification~\cite{liu_tera_2021}.
 
-%TERA
-%Previous work mostly explored masking on the temporal axis, and the model learns to reconstruct from corrupted blocks of time steps.
-Frequency is another alteration for masking.
-The model can learn to reconstruct from missing blocks of frequency bins~\cite{wang2020unsupervised,liu2021tera}.
-In Transformer Encoder Representations from Alteration (TERA), the time and frequency alterations are applied together in the pre-training process.
-This is simply SpecAugment~\cite{spec_augment}.
-%It has been found that each of the alteration methods guides the model to learn a distinct aspect of speech~\cite{liu2021tera}.
-The time alteration effectively enforces a more accurate phoneme prediction, keyword detection, and speech recognition, as it leads the model to learn richer phonetic content.
-The frequency alteration effectively improves speaker prediction accuracy, as it leads the model to encode speaker identity.
-%The magnitude alteration effectively improves performance for all tasks, as it potentially increases data diversity for pre-training.
 Some studies explore alternatives to masking the input directly.
@@
 Taking inspiration from XLNet~\cite{XLNet}, it has also been suggested that the input be reconstructed from a shuffled version~\cite{song20d_interspeech} to address the discrepancy between pre-training and fine-tuning of masking-based approaches.
-%In NLP, the masking tokens only exist in the pre-training phase and never appear in the downstream tasks, making the pre-training and downstream phases have mismatched inputs.
-To avoid using masks during pre-training, in NLP, XLNet is proposed, which learns by reconstructing from shuffled input.
-For speech, the speech version of XLNet, Speech-XLNet~\cite{song20d_interspeech}, has also been proposed.
 
@@
 \edit{DeCoAR 2.0~\cite{ling2020decoar} uses vector quantization}, which is shown to improve the learned representations.
-Furthermore, two dropout regularization methods---attention dropout and layer dropout---are introduced with the TERA model~\cite{liu2021tera, luo2021drop}.
+Furthermore, two dropout regularization methods---attention dropout and layer dropout---are introduced with the TERA model~\cite{liu_tera_2021, luo_drop_2021}.
 \edit{Both methods are variations on the original dropout method \cite{srivastava_dropout_2014}.}
-
-% PASE, WaveNet autoencoders, Phase reconstruction, Audio2Vec
-%\label{sssec:autoencoder}
-%\subsubsection{Autoencoder} 
-%TBD: 
-% -- Do we have to mention ConvDMM? %The ConvDMM~\cite{khurana20_interspeech} approach learns speech representations with convolutional neural networks and Markov Models.
-% -- More about auto-encoder
 
@@
 Other than the autoregressive and masked reconstruction tasks discussed above, various studies have explored the reconstruction of other targets derived from the input.
-PASE and PASE+~\cite{pascual2019learning,ravanelli2020multi} use multiple targets, including the waveform, log power spectrum, \edit{mel cepstral coefficients (MFCCs)}, and prosody features.
-Models that learn acoustic embeddings of small speech segments have targeted future and past spectrogram segments~\cite{chung2018speech2vec, tagliasacchi2019self, tagliasacchi2020pre}, phase information~\cite{quitry2019learning}, and the temporal gap between two segments~\cite{tagliasacchi2019self, tagliasacchi2020pre}.
-
-%\begin{itemize}
-% \item Autoencoder~\cite{chorowski2019unsupervised}: The input and output are the same.
-Usually the regularization introduced above is required.
-%In these works, the autoencoder framework is designed to encode only phonetic content in latent representation and remove other confounding detail such as speaker identity.
-%\item The model learns through reconstructing a spectrogram slice from past and future slices~\cite{tagliasacchi2019self, tagliasacchi2020pre}. %Similar to DecoAR?
-%\item The TemporalGap~\cite{tagliasacchi2019self, tagliasacchi2020pre} approach learns through estimating the temporal gap between two short audio segments extracted at random from the same audio clip.
-%Can this approach be considered as generative?
-%\item Representations are learned through reconstructing the phase of the short-time Fourier transform from its magnitude~\cite{quitry2019learning}. %This is audio, not speech.
-%\item In PASE~\cite{pascual2019learning,ravanelli2020multi}, a single neural encoder learns to solve multiple self-supervised tasks at once, including reconstruction of waveform, Log power spectrum, MFCC, prosody, and other binary discrimination tasks.
-%Is this PASE+?
-I cite PASE+ here, are all the tasks of PASE+ included here.
-Is this approach generative?
-%\end{itemize}
+PASE and PASE+~\cite{pascual_learning_2019,ravanelli_multi_2020} use multiple targets, including the waveform, log power spectrum, \edit{mel cepstral coefficients (MFCCs)}, and prosody features.
+Models that learn acoustic embeddings of small speech segments have targeted future and past spectrogram segments~\cite{chung_speech2vec_2018, tagliasacchi_self_2019, tagliasacchi_pre_2020}, phase information~\cite{quitry2019learning}, and the temporal gap between two segments~\cite{tagliasacchi_self_2019, tagliasacchi_pre_2020}.
 
@@
 
-%In \cite{slu_bert}, phoneme posterior vectors are used to train a standard BERT~\cite{bert, xlnet} model.
-%The phoneme posterior vectors are output from a supervised acoustic model, which requires CTC loss training over the ground-truth phonemes.
-
-% This is about network compression. Perhaps we will have a session about network compression.
-%In Audio ALBERT~\cite{audioalbert}, Mockingjay is modified to have shared parameters across Transformer layers.
-
-%Perhpas we will talk about attack defense
-%In \cite{mockingjay_defense}, Mockingjay is shown to be effective in defending adversarial black-box attacks.
-
 \input{chapters/paper_review/pretext}
@@
 \edit{The pretext task is to maximize latent space similarity between the anchor and positive samples while minimizing the similarity between the anchor and negative samples}.
-This approach has been used extensively in the general \edit{ML} community \cite{Schultz_Joachims_NeurIPS2004}.
+This approach has been used extensively in the general \edit{ML} community \cite{schultz_learning_2003}.
 
@@
 \label{par:cpc}
-Contrastive Predictive Coding (CPC)~\cite{oord2018representation} is a prominent example of a contrastive model. 
+Contrastive Predictive Coding (CPC)~\cite{oord_representation_2018} is a prominent example of a contrastive model. 
 \edit{CPC uses a convolutional module $f_1(\cdot)$ to produce localized representations $z_t$ with a recurrent module $f_2(\cdot)$ on top that outputs a contextualized representation $h_t$.
@@
 The positives and negatives are sampled from the localized representation $Z$.
-%\edit{CPC forms localized representations $z_t$ that depend only on a local window of $X$ and then creates a contextualized representation $h_t$ by running an autoregressive module $f_2(\cdot)$ over $Z_{[1,t]}$.
-The positives and negatives are then sampled from the localized representation $Z$.
-%while the contextualized representation $h_t$ is transformed to an anchor representation $\hat{z}_{t,k}$.
 Hence, at a single timestep $t$, CPC forms multiple anchor representations $\hat{z}_{t,k}$ for $k\in\{1,\dots,K\}$ and associates with each one a single positive sample at the corresponding timestep, $z_{t+k}$, $k$ steps in the future:
@@
 $g_k$ is a unique transformation per \ci{offset} $k$ (e.g., a linear projection).
-%where $f_1(\cdot)$ is a convolutional neural network, such that each $z_{t}$ only encodes information from a limited receptive field in $X$, $f_2(\cdot)$ is limited to condition each $h_t$ only on previous timesteps $Z_{[1,t]}$ and $g_k$ is a unique transformation per step $k$ (e.g.
-a linear projection).
 The loss function measures the similarity between the anchor representation $\hat{z}_{t,k}$ and the positive $z_{t+k}$ normalized by the total similarity to the positive and negatives.}
-The approach is similar to previous work on Noise-Contrastive Estimation (NCE)~\cite{gutmann2010noise}.
+The approach is similar to previous work on Noise-Contrastive Estimation (NCE)~\cite{gutmann_noisecontrastive_2012}.
 \edit{Minimizing} the loss corresponds to maximizing a lower bound on the mutual information between $h_t$ and $z_{t+k}$ (and in turn $x_{t+k-u:t+k+u}$) and is hence called InfoNCE:
@@
 \end{align}
-%
 \noindent \edit{Here, $\mathcal{I}$ \ci{is a random subset of $N$} indices  which includes the target index $t+k$ and $N-1$ negative samples drawn from a proposal distribution, e.g., a uniform distribution over $\{1,\dots,T\}$.
 Including the target index in $\mathcal{I}$ ensures that the loss is a proper categorical cross-entropy and that minimizing it has the previously stated relation to mutual information maximization}.
-This corresponds to sampling negatives from the same sequence and has been \edit{shown} to give good performance for phoneme classification~\cite{oord2018representation}.
+This corresponds to sampling negatives from the same sequence and has been \edit{shown} to give good performance for phoneme classification~\cite{oord_representation_2018}.
 The loss is indexed by $k$ to show that CPC targets multiple offsets using different projection layers $g_k(\cdot)$.
@@
 
-The wav2vec model \cite{schneider2019wav2vec} extends the CPC approach \edit{and uses fully convolutional parameterizations for the modules $f_1(\cdot)$ and $f_2(\cdot)$ with receptive fields of \SI{30}{ms} and \SI{210}{ms}, respectively.
+The wav2vec model \cite{schneider_wav2vec:_2019} extends the CPC approach \edit{and uses fully convolutional parameterizations for the modules $f_1(\cdot)$ and $f_2(\cdot)$ with receptive fields of \SI{30}{ms} and \SI{210}{ms}, respectively.
 While the CPC loss solves a 1-of-$N$ classification task per $(t,k)$, either assigning the anchor to the positive class or (wrongly) to one of the $N-1$ negative classes, the wav2vec loss considers a sequence of $N$ independent binary classifications.
@@
 \begin{align}
-    %g_k(\mathbf{c}_{t}, \mathbf{v}_{t+k}) &= \log(\sigma(\mathbf{v}_{t+k}^{\intercal} (\mathbf{W}_k \mathbf{c}_{t} + \mathbf{b}_k)))\\
     \mathcal{L}_{t,k} &= - \log(\sigma(\hat{z}_{t,k}^{\text{\tiny T}}z_{t+k})) + \sum_{i \in \mathcal{I}} \log(1 - \sigma(\hat{z}_{t,k}^{\text{\tiny T}}z_{i})) \enspace .
@@
 \edit{The wav2vec 2.0 model combines contrastive learning with masking}.
-As the CPC model, it uses the InfoNCE loss \cite{oord2018representation} to maximize the similarity between a contextualized representation and a localized representation.
+As the CPC model, it uses the InfoNCE loss \cite{oord_representation_2018} to maximize the similarity between a contextualized representation and a localized representation.
 \edit{However, instead of using the $z_t$ directly as positive and negatives, it uses a quantization module $g(\cdot)$ to obtain a discrete representation.
@@
 \noindent
-%Here, $m(\cdot)$ again defines a masking policy 
-% Here the $\circ$ operator denotes the Hadamard product with the mask $M$,
-% , $f_1(\cdot)$ is a convolutional neural network with limited receptive field, $f_2(\cdot)$ is a Transformer encoder, 
-The quantization module $g(\cdot)$ uses a Gumbel softmax~\cite{Gumbel-Softmax} with a straight-through estimator.
+The quantization module $g(\cdot)$ uses a Gumbel softmax~\cite{jang_categorical_2017} with a straight-through estimator.
 Since the quality of the learned representations is contingent on the quality of the quantization, wav2vec 2.0 combines two techniques to learn high-quality codebooks.
-First, wav2vec 2.0 concatenates quantized representations from multiple codebooks at each timestep, so-called Product Quantization (PQ)~\cite{ProductQuantization}.
+First, wav2vec 2.0 concatenates quantized representations from multiple codebooks at each timestep, so-called Product Quantization (PQ)~\cite{jegou_product_2011}.
 Also, the \edit{primary} training loss \edit{described below} is augmented with an auxiliary term designed to encourage equal use of all codebook entries.
@@
 The loss is
-% For the InfoNCE loss, anchors features are taken from the top layer contextualized Transformer representations for masked regions,  positive samples are the corresponding quantized features, and negative samples are selected from other masked locations in the same utterance.
 %
@@
 \noindent where $S_{\text{c}}(\cdot)$ is the cosine similarity and $\mathcal{I}$ contains the target index $t$ and negative indices sampled from other masked timesteps.
-%\edit{Notably, since the positives and negatives are quantized representations, we can guarentee that representations of the same class as the positive are excluded when sampling the negatives.}
 
@@
 It has subsequently} inspired many follow-up studies.
-The wav2vec-C~\cite{sadhu21_interspeech} approach extends wav2vec 2.0 with a consistency term in the loss that aims to reconstruct the input features from the learned quantized representations, similar to VQ-VAE~\cite{VQVAE2}.
-%The wav2vec-C representations achieve, on average, twice the error reduction over baseline and a higher codebook utilization in comparison to wav2vec 2.0 for far-field English voice commands and voice queries.
+The wav2vec-C~\cite{sadhu_wav2vec-c_2021} approach extends wav2vec 2.0 with a consistency term in the loss that aims to reconstruct the input features from the learned quantized representations, similar to VQ-VAE~\cite{razavi_generating_2019}.
 
@@
 Whereas a contrastive loss discourages the model from learning a trivial solution by the use of negative samples, this must be circumvented differently for predictive methods.
-%However, such losses do not discourage trivial solutions, in the same way that contrastive losses do.
 For this reason, predictive methods compute the targets outside the model's computational graph; usually with a completely separate model.
 Thus, the predictive setup is somewhat akin to teacher-student training.
-The first predictive approaches were motivated by the success of BERT-like methods in NLP \cite{jacob2019BERT} as well as the DeepCluster method in CV \cite{Caron_2018_ECCV}.}
-%where the model learns a contextual representation that aggregates information from distant timesteps.
-Predictive approaches for speech representation learning also draw inspiration from the DeepCluster method \cite{Caron_2018_ECCV} for self-supervised visual learning.
+The first predictive approaches were motivated by the success of BERT-like methods in NLP \cite{devlin_bert:_2018} as well as the DeepCluster method in CV \cite{caron_deep_2018}.}
 
@@
 \paragraph{Discrete BERT} \edit{Applying BERT-type training directly to speech input is not possible due to its continuous nature}.
-The Discrete BERT approach~\cite{baevski2019effectiveness} uses a pre-trained vq-wav2vec model to derive a discrete vocabulary \cite{Baevski2020vq-wav2vec}.
-The vq-wav2vec model is similar to wav2vec mentioned in Paragraph~\ref{par:cpc} but uses quantization to learn discrete representations.
-%For the LibriSpeech dataset, the discrete vocabulary contains about 13.5k unique codes. 
+The Discrete BERT approach~\cite{baevski_effectiveness_2020} uses a pre-trained vq-wav2vec model to derive a discrete vocabulary \cite{baevski_vq-wav2vec_2020}.
+The vq-wav2vec model is similar to wav2vec mentioned in \cref{par:cpc} but uses quantization to learn discrete representations.
 \edit{Specifically, discrete units $c_t$ are first extracted with the vq-wav2vec model $f_1(\cdot)$ and then used as inputs \emph{and} targets in a standard BERT model $f_2(\cdot)$ with a softmax normalized output layer $g(\cdot)$,
@@
 \noindent where $\mathcal{M}$ is the set of all masked timesteps. During training, only the BERT model's parameters are updated, while the vq-wav2vec model parameters are frozen.} 
-% After pre-training, the BERT Transformer model is fine-tuned using a Connectionist Temporal Classification (CTC) \cite{graves2006connectionist} loss for ASR. 
-%Following the Libri-light near-zero evaluation policy \cite{kahn2020libri}, fine-tuning is done using 10 minute, 1 hour and 10 hour subsets of LibriSpeech balanced for gender and recording conditions, as well as the standard 100 hour-clean part of LibriSpeech.
 Discrete BERT was the first model to demonstrate the effectiveness of self-supervised speech representation learning by achieving a WER of 25\% on the standard test-other subset using a 10-minute fine-tuning set, \edit{setting the direction for many approaches to follow}.
 
-\paragraph{HuBERT} Rather than relying on an advanced representation learning model for discretizing continuous inputs, as Discrete BERT, the Hidden Unit BERT (HuBERT) approach \cite{hsu2021hubert} \edit{uses quantized MFCC features as targets learned with classic k-means.
+\paragraph{HuBERT} Rather than relying on an advanced representation learning model for discretizing continuous inputs, as Discrete BERT, the Hidden Unit BERT (HuBERT) approach \cite{hsu_hubert_2021} \edit{uses quantized MFCC features as targets learned with classic k-means.
 Thus, to compute the targets, the k-means model $g_1(\cdot)$ assigns a cluster center to each timestep.
-% \begin{align}
-%     c_t &= g_1(X_{[t-w,t+w]}) \enspace .
-% \end{align}
-% \noindent where $w$ defines the window size used to compute the MFCCs. 
 Different from Discrete BERT, HuBERT takes the raw waveform as input, rather than discrete units. This helps to prevent loss of any relevant information due to input quantization.
-%Specifically, HuBERT predicts the \edit{pre-computed} k-means cluster assignment given the masked continuous speech features.
 HuBERT uses an architecture similar to that of wav2vec 2.0, with a convolutional module $f_1(\cdot)$ and a Transformer encoder $f_2(\cdot)$, as well as a softmax normalized output layer $g_2(\cdot)$:
-%Similar to wav2vec 2.0, the HuBERT model uses a convolutional encoder for the input.
-Masking is then applied before the Transformer network as in wav2vec 2.0 \namecrefs{w2v2 f_v} \ref{w2v2 f_v} and \ref{w2v2 f_c}.
 \begin{align}
@@
 \noindent where $w$ defines the window size used to compute the MFCCs. 
-%Since the targets are pre-computed cluster identities, the HuBERT model can directly evaluate the regular categorical cross-entropy loss between the correct k-means cluster and the predicted one.
-This is opposed to constrastive methods which require negative samples to avoid degenerating to trivial solutions.
 The categorical cross-entropy loss is} computed on both masked, $\mathcal{L}_m$, and unmasked, $\mathcal{L}_u$, timesteps:  
 \begin{align}
- %   \mathcal{L}_m &= \sum_{t\in M} - \log p(c_t | \tilde{X}, t)
-%    \mathcal{L}_m &= \sum_{t\in M} - \log p(c_t \mid H, t) \enspace ,
-    % \mathcal{L}_m &= \alpha\sum_{t\in M} - \log p(c_t \mid H) + (1-\alpha)\sum_{t\notin M} - \log p(c_t \mid H) \enspace , \\
     \mathcal{L}_m &= \sum_{t\in \mathcal{M}} - \log p(c_t \mid X) \enspace , \\
@@
 The first iteration is described above.
-%As described, in the first iteration of pre-training, k-means clustered MFCC features are used as targets, $c_t$.
 Once completed, a second iteration of pre-training follows. Here, representations from a hidden layer of the model from the first iteration are clustered with k-means to obtain new targets $c_t$.}
-%Here, the final model of the first iteration is used to encode latent representations of the full dataset.
-These are then quantized using k-means to form a new set of target vectors, replacing $c_t$ from the first iteration.
-%Alternating between learning offline discrete targets using k-means and training the model using masked prediction ensures training stability while preserving representation quality.
 
 \edit{For HuBERT, only two iterations are needed to match or outperform the previous state-of-the-art results for low-resource speech recognition.
-And combining the HuBERT approach with the wav2vec 2.0 approach, the w2v-BERT model has managed to improve results even further~\cite{w2vbert}.}
-
-%\edit{The w2v-BERT approach~\cite{w2vbert} combines contrastive and predictive approach and optimizes the wav2vec 2.0 contrastive loss and the HuBERT masked prediction loss jointly}.
-It showed 5\% to 10\% relative WER reduction on the test-clean and test-other subsets of Librispeech, and more than 30\% relative improvement for real-world in-house voice search compared to a wav2vec 2.0 baseline.
+And combining the HuBERT approach with the wav2vec 2.0 approach, the w2v-BERT model has managed to improve results even further~\cite{chung_w2vbert_2021}.}
 
@@
 
-\edit{WavLM emphasizes spoken content modeling and speaker identity preservation~\cite{chen2021wavlm}. It is largely identical to HuBERT, but introduces two useful extensions.}
+\edit{WavLM emphasizes spoken content modeling and speaker identity preservation~\cite{chen_wavlm_2021}. It is largely identical to HuBERT, but introduces two useful extensions.}
 
@@
 
-%First, it extends the Transformer self-attention with a content-based gated relative position bias which improves the model's capability on recognition tasks compared to the convolutional context embedding \edit{used} in HuBERT and wav2vec 2.0.
-The gates adjust the relative position bias by conditioning on the current speech content \edit{which differs from the usual approach of learning a fixed positional encoding that is independent of the input}.
-Intuitively, the same distance offset between two frames tends to play different roles if one frame is silent and the other belongs to a speech segment.
-%WavLM uses 320 bucketed positions for the range of possible offsets.
-%The relative positional embedding parameters are shared across all layers.
-
 \edit{Second, it uses an utterance mixing strategy where signals from different speakers are combined to augment the training data.} \edit{Specifically, random subsequences from other examples in the same batch are scaled and added to each input example.
-%each example in a batch is combined with random subsequences from other examples in the same batch.
-%Mixed utterances are selected from the same training minibatch, then randomly cropped and scaled by a sampled source energy ratio.
 Only the targets corresponding to the original example are predicted during pre-training. Thus, the model learns to filter out the added overlapping speech.}
@@
 Most SSL methods are trained on data \edit{where each example only contains speech from a single person}; therefore, they can perform subpar on multispeaker tasks like speaker separation and diarization.
-%The content information corresponding to the main speaker is used as the target during the masked prediction pre-training.
 
-%The pre-training data used for WavLM extends the sources used for HuBERT and wav2vec 2.0 to reach a total of 94k hours of public audio.
 The WavLM model achieved substantial improvements on the speech separation, speaker verification and speaker diarization tasks in the SUPERB benchmark, while also performing well on many other tasks compared to HuBERT and wav2vec 2.0.
-%It also improved upon \edit{the} HuBERT and wav2vec 2.0 models across all ten tasks of the SUPERB benchmark while maintaining a comparable performance on ASR.
 
-\paragraph{data2vec} \edit{Motivated by the success of using an exponential moving average (EMA) teacher for self-supervised visual representations~\cite{grill2020byol, dino}, the data2vec model~\cite{data2vec} computes targets $Y$ using an EMA of its own parameters.
+\paragraph{data2vec} \edit{Motivated by the success of using an exponential moving average (EMA) teacher for self-supervised visual representations~\cite{grill_bootstrap_2020, caron_emerging_2021}, the data2vec model~\cite{baevski_data2vec_2022} computes targets $Y$ using an EMA of its own parameters.
 The targets are constructed by averaging hidden representations of the top $k$ layers of the EMA teacher network applied to unmasked inputs.
@@
 The data2vec model uses a regression loss between target and prediction.
-Specifically, to reduce sensitivity to outliers and prevent exploding gradients, it uses the smoothed \ensuremath{L_1} loss \cite{fastrcnn},
+Specifically, to reduce sensitivity to outliers and prevent exploding gradients, it uses the smoothed \ensuremath{L_1} loss \cite{girshick_fast_2015},
 \begin{equation}
@@
 Another challenge for these models centers around the quality of the initial vocabulary generated from MFCC features.
-% Although \edit{a} masked prediction loss enables the models not to be \edit{limited} by the quality of their initial target units, experiments show that initial teachers with higher quality, e.g., Gaussian Mixture Models (GMMs) instead of k-means, provide non-negligible performance gains.
-The data2vec approach improves over other predictive models \edit{by allowing the targets to improve continuously via the EMA teacher network}; however, \edit{student-teacher approaches inflate the existing computational challenges of very large models and may necessitate the use of methods that decrease instantaneous memory utilization such as mixed precision training, model parallelism and model sharding \cite{paszke2019pytorch}.}
+The data2vec approach improves over other predictive models \edit{by allowing the targets to improve continuously via the EMA teacher network}; however, \edit{student-teacher approaches inflate the existing computational challenges of very large models and may necessitate the use of methods that decrease instantaneous memory utilization such as mixed precision training, model parallelism and model sharding \cite{paszke_automatic_2017}.}
 
 \subsection{Learning from multi-modal data}
-% {\color{blue} Karen}\\
 \subsubsection{Motivation}
-%Benefit from the naturally parallel data without annotation
 Multiple modalities are useful in many settings, where each modality provides information that is complementary to other modalities.
-Multi-modal work includes supervised settings, such as audio-visual ASR~\cite{petajan1984automatic,potamianos2003recent} and person identification~\cite{aleksic2006audio} which have been studied for decades.
+Multi-modal work includes supervised settings, such as audio-visual ASR~\cite{petajan_automatic_1984,potamianos_recent_2003} and person identification~\cite{aleksic_audiovisual_2006} which have been studied for decades.
 In this section, we focus only on unsupervised representation learning from multi-modal data.
@@
 Such ``grounding" signals can contain supplementary information that can be used by models to infer the content of the speech.
-Human language learning provides a proof of concept for this, as it is believed that infants benefit from the visual modality when learning language~\cite{legerstee1990infants}.
-Early computational models of multi-modal language learning were motivated by (and tried to emulate) human learning of language in the context of the visual surroundings~\cite{roy1999learning}.
+Human language learning provides a proof of concept for this, as it is believed that infants benefit from the visual modality when learning language~\cite{legerstee_infants_1990}.
+Early computational models of multi-modal language learning were motivated by (and tried to emulate) human learning of language in the context of the visual surroundings~\cite{roy_learning_1999}.
 
@@
 
-%\paragraph{Intrinsic modalities} 
 \edit{\textit{Intrinsic modalities} are} 
-%refer to 
 modalities produced directly by the speech source.
-Examples of intrinsic modalities (besides the speech audio) include images or video of the speaker's face~\cite{lee2004avicar,chung2016lip}, lip-movement~\cite{shi2022learning}, articulatory flesh point measurements~\cite{westbury1990x,wrench2001new}, or simultaneous \edit{magnetic resonance imaging (MRI)} scans~\cite{narayanan2011multimodal}.
+Examples of intrinsic modalities (besides the speech audio) include images or video of the speaker's face~\cite{lee_avicar_2004,chung_lip_2016}, lip-movement~\cite{shi_learning_2022}, articulatory flesh point measurements~\cite{westbury_xray_1990,wrench_new_2001}, or simultaneous \edit{magnetic resonance imaging (MRI)} scans~\cite{narayanan_multimodal_2011}.
 Typically, learning from multiple intrinsic modalities is done so as to improve robustness to noise, since acoustic noise is likely to be uncorrelated with the other \edit{modality(ies)}.
@@
 \begin{itemize}
-    \item Multi-view autoencoders and variations~\cite{ngiam2011multimodal,badino2016integrating},
-    \item Multi-modal deep Boltzmann machines~\cite{srivastava2012multimodal},
-    \item Canonical correlation analysis (CCA)~\cite{hotelling1936relations} and its nonlinear extensions~\cite{andrew2013deep,wang2015deep,wang2016deep,michaeli2016nonparametric,Melzer_01a,LaiFyfe00a,LaiFyfe99a,BachJordan05a,wang_15a},
-    \item Multi-view contrastive losses~\cite{HermanBlunsom14a,Huang_13b},
-    \item More recently, audio-visual extensions of masked prediction methods~\cite{shi2022learning,shi2022robust}, specifically Audio-Visual HuBERT (AV-HuBERT)~\cite{shi2022learning}.
+    \item Multi-view autoencoders and variations~\cite{ngiam_multimodal_2011,badino_integrating_2016},
+    \item Multi-modal deep Boltzmann machines~\cite{srivastava_multimodal_2012},
+    \item Canonical correlation analysis (CCA)~\cite{hotelling_relations_1936} and its nonlinear extensions~\cite{andrew_deep_2013,wang_deep_2015,wang_deep_2016,michaeli_nonparametric_2016,melzer_nonlinear_2001,lai_kernel_2000,lai_neural_1999,bach_probabilistic_2005,wang_unsupervised_2015},
+    \item Multi-view contrastive losses~\cite{hermann_multilingual_2013,huang_learning_2013},
+    \item More recently, audio-visual extensions of masked prediction methods~\cite{shi_learning_2022,shi_robust_2022}, specifically Audio-Visual HuBERT (AV-HuBERT)~\cite{shi_learning_2022}.
 \end{itemize}
 
-%\paragraph{Extrinsic modalities} refer to
 \textit{Extrinsic modalities} are modalities that are not produced by the same source but still provide context for each other.
@@
 There has recently been \edit{a} surge of datasets collected for this purpose, usually consisting of images and spoken captions, the audio and image frames in a video, or video clips with their spoken descriptions.
-A recent review of datasets, as well as methods, in this category is provided by Chrupa\l{}a~\cite{chrupala2021visually}.
+A recent review of datasets, as well as methods, in this category is provided by Chrupa\l{}a~\cite{chrupala_visually_2021}.
 
-Typical approaches involve learning a neural representation model for each modality, with a multi-modal contrastive loss that encourages paired examples in the two modalities to have similar representations while unpaired examples remain different~\cite{synnaeve+etal_nipsworkshop14,harwath+etal_nips16,harwath+glass_asru15,merkx2019language,rouditchenko2020avlnet,peng2022fast}.
-Other choices include training with a masked margin softmax loss~\cite{ilharco2019large,sanabria2021talk} or a masked prediction loss~\cite{MultimodelAmazonICASSP22}.
-Such models are typically evaluated on cross-modal retrieval, although some work has also used the models for other downstream tasks such as the ZeroSpeech and SUPERB benchmark tasks~\cite{peng2022self}.
-Analyses of such models have found that, despite the very high-level learning objective of matching speech with a corresponding image (or other contextual modality), such models often learn multiple levels of linguistic representations from the shallowest to the deepest model layers~\cite{harwath2019learning,chrupala2017representations,scharenborg+etal_icassp18}.
-They are also able to learn word-like units~\cite{harwath2018jointly,peng2022word,wang2020dnn} and can be used for cross-lingual retrieval, by considering the visual signal as an ``interlingua"~\cite{harwath2018vision,havard2019models,kamper+roth_sltu18}.
-In some settings, even in the presence of some amount of textual supervision (i.e., the speech is transcribed), visual grounding still helps learn a better representation for retrieval~\cite{pasad2019contributions}.
+Typical approaches involve learning a neural representation model for each modality, with a multi-modal contrastive loss that encourages paired examples in the two modalities to have similar representations while unpaired examples remain different~\cite{synnaeve_learning_2014,harwath_unsupervised_2016,harwath_deep_2015,merkx_language_2019,rouditchenko_avlnet_2021,peng_fastslow_2022}.
+Other choices include training with a masked margin softmax loss~\cite{ilharco_largescale_2019,sanabria_talk_2021} or a masked prediction loss~\cite{chan_multimodal_2022}.
+Such models are typically evaluated on cross-modal retrieval, although some work has also used the models for other downstream tasks such as the ZeroSpeech and SUPERB benchmark tasks~\cite{peng_selfsupervised_2022}.
+Analyses of such models have found that, despite the very high-level learning objective of matching speech with a corresponding image (or other contextual modality), such models often learn multiple levels of linguistic representations from the shallowest to the deepest model layers~\cite{harwath_learning_2019,chrupala_representations_2017,scharenborg_linguistic_2018}.
+They are also able to learn word-like units~\cite{harwath_jointly_2018,peng_word_2022,wang_dnnhmmdnn_2020} and can be used for cross-lingual retrieval, by considering the visual signal as an ``interlingua"~\cite{harwath_vision_2018,havard_models_2019,kamper_visually_2018}.
+In some settings, even in the presence of some amount of textual supervision (i.e., the speech is transcribed), visual grounding still helps learn a better representation for retrieval~\cite{pasad_contributions_2019}.
 
 There has also been growing interest in learning joint speech and text representations using paired and unpaired data.
-The SLAM approach~\cite{bapna2021slam} is an example where speech and text are first represented using two separate pre-trained encoders followed by a multi-modal encoder to build the joint representations.
+The SLAM approach~\cite{bapna_slam_2021} is an example where speech and text are first represented using two separate pre-trained encoders followed by a multi-modal encoder to build the joint representations.
 The entire model is trained using a multi-task loss including two supervised and two self-supervised tasks.
-
-%Benefiting from natural pairing of modalities in videos. Audio and text meta-data in FB videos. \\
-%oWE can add it here or in the TTS subsection:
-% link in comment to avoid compilation errors: https://arxiv.org/pdf/2011.11564.pdf \\
 
@@
 
-% {\color{blue} Karen.} \\
-% \kl{I moved this section here from the zero-resource section.}\\
-
 Most of the representation learning techniques discussed in the preceding sections are aimed at learning frame-level representations.
 For some purposes, however, it may be useful to explicitly represent longer spans of speech audio of arbitrary duration, such as phone, word, or phrase-level segments.
-For example, searching within a corpus of recorded speech for segments that match a given (written or spoken) query can be seen as finding segments whose representations are most similar to that of the query~\cite{levin+etal_icassp15,guoguo+etal_icassp15,Chung2016AudioWord2Vec,settle2017query}; word embeddings can be defined by pooling representations of instances of a given word~\cite{chung2018speech2vec}; unsupervised segmentation and spoken term discovery can be seen as a problem of detecting and clustering segments~\cite{kamper2017embedded,Kamper17BESGMM}; and even ASR can be viewed as the problem of matching written word representations to representations of audio spans~\cite{maas+etal_icmlwrl12,bengio+heigold_interspeech14,settle2019_a2w}.
+For example, searching within a corpus of recorded speech for segments that match a given (written or spoken) query can be seen as finding segments whose representations are most similar to that of the query~\cite{levin_segmental_2015,chen_querybyexample_2015,chung_audio_2016,settle_querybyexample_2017}; word embeddings can be defined by pooling representations of instances of a given word~\cite{chung_speech2vec_2018}; unsupervised segmentation and spoken term discovery can be seen as a problem of detecting and clustering segments~\cite{kamper_embedded_2017,kamper_segmental_2017}; and even ASR can be viewed as the problem of matching written word representations to representations of audio spans~\cite{maas_wordlevel_2012,bengio_word_2014,settle_acoustically_2019}.
 
 Several lines of work have begun to address the problem of learning representations of spans of speech, especially word segments, typically referred to as \textit{acoustic word embeddings}.
-Early work on unsupervised acoustic word embeddings defined them as vectors of distances from the target segment to a number of pre-defined ``template" segments~\cite{levin+etal_asru13}.
-Later work used variants of neural autoencoders~\cite{Chung2016AudioWord2Vec,holzenberger2018learning,kamper2019truly,pengcorrespondence}.
-These are often evaluated on word discrimination, that is the task of determining whether two word segments correspond to the same word or not~\cite{carlin2011rapid}.
-This task can be thought of as a proxy for query-by-example search, since the basic operation in search is to determine whether a segment in the search database matches a query segment, and has been used for evaluation of both frame-level (e.g.,~\cite{kamper2015unsupervised}) and word-level~\cite{levin+etal_asru13,kamper2016deep} representations.
+Early work on unsupervised acoustic word embeddings defined them as vectors of distances from the target segment to a number of pre-defined ``template" segments~\cite{levin_fixed-dimensional_2013}.
+Later work used variants of neural autoencoders~\cite{chung_audio_2016,holzenberger_learning_2018,kamper_truly_2019,peng_correspondence_2020}.
+These are often evaluated on word discrimination, that is the task of determining whether two word segments correspond to the same word or not~\cite{carlin_rapid_2011}.
+This task can be thought of as a proxy for query-by-example search, since the basic operation in search is to determine whether a segment in the search database matches a query segment, and has been used for evaluation of both frame-level (e.g.,~\cite{kamper_unsupervised_2015}) and word-level~\cite{levin_fixed-dimensional_2013,kamper2016deep} representations.
 
-% \kl{I left out much of the work on acoustic word embeddings, since it is in a supervised or at least weakly supervised setting.
-I could add a quick note that a lot of other work exists, and add citations, but not sure that's needed for this paper.}
-
-Since most work on acoustic word embeddings preceded the very recent wave of new self-supervised frame-level representations, one question is whether word (or more generally segment) embeddings could be derived more simply by pooling self-supervised frame-level representations, as has been done for text span embeddings by pooling over word embeddings~\cite{toshniwal2020cross,wang2021phrase}.
-Some initial results suggest that at least very simple pooling approaches like downsampling and mean or max pooling are not successful~\cite{van2021comparison,pengcorrespondence}, but more work is needed to reach conclusive results.
-
-% The model learns through reconstructing a spectrogram slice from past and future slices~\cite{tagliasacchi2019self, tagliasacchi2020pre}. %Similar to DecoAR?
-% The TemporalGap~\cite{tagliasacchi2019self, tagliasacchi2020pre} approach learns through estimating the temporal gap between two short audio segments extracted at random from the same audio clip.
-%Can this approach be considered as generative?
-% Representations are learned through reconstructing the phase of the short-time Fourier transform from its magnitude~\cite{quitry2019learning}. %This is audio, not speech.
-
-%We put the work of audio word embedding here.
-%This can be a link to zero speech and further speech2semantic applications.
-%Originally, audio word embedding was targeted at query-by-example, so the researchers want the embedding to contain the phoneme information (for example,  https://arxiv.org/abs/1510.01032, https://arxiv.org/abs/1603.00982).
-%Then there is some work targeting at including semantic information in the embedding (for example, https://arxiv.org/abs/1803.08976).
-
-% should we add a section on Learning representation through multi-tasking}
+Since most work on acoustic word embeddings preceded the very recent wave of new self-supervised frame-level representations, one question is whether word (or more generally segment) embeddings could be derived more simply by pooling self-supervised frame-level representations, as has been done for text span embeddings by pooling over word embeddings~\cite{toshniwal_crosstask_2020,wang_phrasebert_2021}.
+Some initial results suggest that at least very simple pooling approaches like downsampling and mean or max pooling are not successful~\cite{vanstaden_comparison_2021,peng_correspondence_2020}, but more work is needed to reach conclusive results.

### 2023-03-16 chapters/paper_review/04_benchmarking.tex
Change reference keys for review paper (WIP)

@@
 over 2k hours of conversational telephone speech, 1k hours of which is utilized
-for pre-training~\cite{jiang2021further}. Industrial researchers have also
+for pre-training~\cite{jiang_further_2021}. Industrial researchers have also
 begun to build large-scale datasets for learning speech representations.
@@
 self-supervised pre-training have been collected at 
-Amazon~\cite{sadhu21_interspeech}.
+Amazon~\cite{sadhu_wav2vec-c_2021}.
 
@@
 collected corpora for pre-training Chinese speech representations. Didi Dictation
-and Didi Callcenter~\cite{jiang2019improving, jiang2021further} are two
+and Didi Callcenter~\cite{jiang2019improving, jiang_further_2021} are two
 internal datasets containing respectively 10k hours of read speech collected
@@
 The entire dataset~\cite{babu2021xlsr} or a 24k-hour 
-portion~\cite{chen2021unispeechsat, chen2021wavlm} thereof has been used for pre-training. BBL
+portion~\cite{chen2021unispeechsat, chen_wavlm_2021} thereof has been used for pre-training. BBL
 consists of 1k hours of conversational telephone speech in 17 African and Asian
@@
 For the purpose of pre-training and evaluating Mandarin speech representations,
-the authors of \cite{jiang2019improving, jiang2021further} also compiled Open Mandarin, 
+the authors of \cite{jiang2019improving, jiang_further_2021} also compiled Open Mandarin, 
 an open-source Mandarin dataset of 1.5k hours of speech from
@@
 arrays. A team at Amazon Alexa also recorded and transcribed a corpus of 1k
-hours of audio for model training and evaluation~\cite{sadhu21_interspeech}.
+hours of audio for model training and evaluation~\cite{sadhu_wav2vec-c_2021}.
 
@@
 
-Last, \cite{tagliasacchi2020pre} puts together five datasets 
+Last, \cite{tagliasacchi_pre_2020} puts together five datasets 
 (MUSAN~\cite{snyder2015musan}, Bird Audio Detection~\cite{stowell2019automatic},
@@
     AVSpeech & PT & Multi & 3.1k & - & CC BY 4.0 \\ \hline %Bidir-CPC
-    Fisher & PT & EN & 2k/1k \cite{jiang2021further}  & - & Linguistic Data Consortium (LDC) \\ \hline % MPC
+    Fisher & PT & EN & 2k/1k \cite{jiang_further_2021}  & - & Linguistic Data Consortium (LDC) \\ \hline % MPC
 
@@
     Multilingual LS (MLS) & PT/EV & Multi & 50k & ASR & CC BY 4.0 \\ \hline
-    VoxPopuli (VP) & PT/EV & Multi & 400k \cite{babu2021xlsr}/24k \cite{chen2021unispeechsat, chen2021wavlm} & ASR & CC0 \\ \hline
+    VoxPopuli (VP) & PT/EV & Multi & 400k \cite{babu2021xlsr}/24k \cite{chen2021unispeechsat, chen_wavlm_2021} & ASR & CC0 \\ \hline
     BABEL (BBL) & PT/EV & Multi & 1k & ASR & IARPA Babel Agreement  \\ \hline
 
-    GigaSpeech & PT/EV & EN & 40k/10k \cite{chen2021unispeechsat, chen2021wavlm} & ASR & Apache-2.0 License \\ \hline
+    GigaSpeech & PT/EV & EN & 40k/10k \cite{chen2021unispeechsat, chen_wavlm_2021} & ASR & Apache-2.0 License \\ \hline
     TED-LIUM 3 (TED3) & PT/EV & EN & 450 & ASR & CC BY-NC-ND 3.0 \\ \hline %Bidir-CPC
@@
     % CPC
-    \multirow{2}{*}[0mm]{CPC \cite{oord2018representation}} & \multirow{2}{*}[0mm]{LS 100 hrs} & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{CPC \cite{oord_representation_2018}} & \multirow{2}{*}[0mm]{LS 100 hrs} & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
     & & SID & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline
@@
     % PASE
-    \multirow{2}{*}[-5mm]{PASE\cite{pascual2019learning}} & \multirow{2}{*}[-5mm]{LS 50 hrs \cite{ravanelli2018learning}} & SID & VCTK\textsuperscript{\ref{footnote:official-split}} & VCTK\textsuperscript{\ref{footnote:official-split}} & \checkmark & 44 hrs \\ \cline{3-7}
+    \multirow{2}{*}[-5mm]{PASE\cite{pascual_learning_2019}} & \multirow{2}{*}[-5mm]{LS 50 hrs \cite{ravanelli2018learning}} & SID & VCTK\textsuperscript{\ref{footnote:official-split}} & VCTK\textsuperscript{\ref{footnote:official-split}} & \checkmark & 44 hrs \\ \cline{3-7}
     & & Sentiment & INTERFACE\textsuperscript{\ref{footnote:interface-split}} & INTERFACE\textsuperscript{\ref{footnote:interface-split}} & \checkmark & 3 hrs \\ \cline{3-7}
@@
     % Audio2Vec
-    Audio2Vec \cite{tagliasacchi2020pre} & AudioSet & AED & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & \checkmark & See \cite{tagliasacchi2020pre} for details \\ \hline
+    Audio2Vec \cite{tagliasacchi_pre_2020} & AudioSet & AED & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & \checkmark & See \cite{tagliasacchi_pre_2020} for details \\ \hline
     % APC
-    \multirow{3}{*}[0mm]{APC \cite{chung2019unsupervised, chung2020generative}} & \multirow{3}{*}[0mm]{LS 360 hrs} & ASR & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & WSJ dev93 & \checkmark & 72 hrs \\ \cline{3-7}
+    \multirow{3}{*}[0mm]{APC \cite{chung_unsupervised_2019, chung_generative_2020}} & \multirow{3}{*}[0mm]{LS 360 hrs} & ASR & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & WSJ dev93 & \checkmark & 72 hrs \\ \cline{3-7}
     & & ST & LS En-Fr\textsuperscript{\ref{footnote:official-split}} & LS En-Fr\textsuperscript{\ref{footnote:official-split}} & - & 236 hrs \\ \cline{3-7}
@@
     % wav2vec
-    \multirow{2}{*}[0mm]{wav2vec \cite{schneider2019wav2vec}} & \multirow{2}{*}[-0.5mm]{\makecell[l]{LS 80/960 hrs,\\ LS 960 hrs \\ \:+ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & \checkmark & 81 hrs\vspace{1mm} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{wav2vec \cite{schneider_wav2vec:_2019}} & \multirow{2}{*}[-0.5mm]{\makecell[l]{LS 80/960 hrs,\\ LS 960 hrs \\ \:+ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & \checkmark & 81 hrs\vspace{1mm} \\ \cline{3-7}
     & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hr\vspace{1mm} \\ \hline
@@
     % MockingJay
-    \multirow{3}{*}{MockingJay \cite{liu2020mockingjay}} & \multirow{3}{*}{\makecell[l]{LS 360 hrs}} & PC & LS 360 hrs & LS test-clean & - & 0.36\textsuperscript{\ref{footnote:low-resource}}/1.8\textsuperscript{\ref{footnote:low-resource}}/3.6\textsuperscript{\ref{footnote:low-resource}}/18\textsuperscript{\ref{footnote:low-resource}}/45\textsuperscript{\ref{footnote:low-resource}}/360  hrs \\ \cline{3-7}
+    \multirow{3}{*}{MockingJay \cite{liu_mockingjay_2020}} & \multirow{3}{*}{\makecell[l]{LS 360 hrs}} & PC & LS 360 hrs & LS test-clean & - & 0.36\textsuperscript{\ref{footnote:low-resource}}/1.8\textsuperscript{\ref{footnote:low-resource}}/3.6\textsuperscript{\ref{footnote:low-resource}}/18\textsuperscript{\ref{footnote:low-resource}}/45\textsuperscript{\ref{footnote:low-resource}}/360  hrs \\ \cline{3-7}
     & & SID & LS 100 hrs\textsuperscript{\ref{footnote:91-split-test}} & LS 100 hrs\textsuperscript{\ref{footnote:91-split-test}} & - & 90\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
@@
     % vq-wav2vec
-    \multirow{2}{*}[0mm]{vq-wav2vec \cite{Baevski2020vq-wav2vec}} & \multirow{2}{*}[0mm]{LS 960 hrs} & ASR & WSJ si284 & \makecell[l]{WSJ eval92} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{vq-wav2vec \cite{baevski_vq-wav2vec_2020}} & \multirow{2}{*}[0mm]{LS 960 hrs} & ASR & WSJ si284 & \makecell[l]{WSJ eval92} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
     & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \hline
@@
     % PASE+
-    \multirow{3}{*}[0mm]{PASE+\cite{ravanelli2020multi}} & \multirow{3}{*}[0mm]{LS 50 hrs \cite{ravanelli2018learning}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
+    \multirow{3}{*}[0mm]{PASE+\cite{ravanelli_multi_2020}} & \multirow{3}{*}[0mm]{LS 50 hrs \cite{ravanelli2018learning}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
     & & ASR & DIRHA\textsuperscript{\ref{footnote:official-split}} & DIRHA\textsuperscript{\ref{footnote:official-split}} & \checkmark & 11 hrs \\ \cline{3-7}
@@
     % BMR
-    \multirow{2}{*}[0mm]{BMR \cite{wang2020unsupervised}} & \multirow{2}{*}[0mm]{\makecell[l]{WSJ si284,\\LS 960 hrs}} & ASR & WSJ si284 & WSJ eval92 & - & 81 hrs \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{BMR \cite{wang_unsupervised_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{WSJ si284,\\LS 960 hrs}} & ASR & WSJ si284 & WSJ eval92 & - & 81 hrs \\ \cline{3-7}
     & & PR & WSJ si84/si284 & WSJ dev93 & - & 15/81 hrs \\ \hline
@@
     % vq-APC
-    \multirow{2}{*}[0mm]{vq-APC \cite{chung20e_interspeech}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{vq-APC \cite{chung_vector-quantized_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
     & & SID & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & \checkmark & 65\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline
@@
     % vq-wav2vec + Discrete BERT
-    \makecell[l]{vq-wav2vec + \\\:DiscreteBERT \cite{baevski2019effectiveness}} & LS 960 hrs & ASR & LS 100 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline
+    \makecell[l]{vq-wav2vec + \\\:DiscreteBERT \cite{baevski_effectiveness_2020}} & LS 960 hrs & ASR & LS 100 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline
 
@@
     % MPC
-    \multirow{3}{*}[0mm]{MPC \cite{jiang2019improving, jiang2021further}} & \multirow{2}{*}[0mm]{\makecell[l]{Didi Callcenter,\\Didi Dictation,\\Open Mandarin}} & ASR-zh & HKUST\textsuperscript{\ref{footnote:official-split}} & HKUST\textsuperscript{\ref{footnote:official-split}} & \checkmark & 168 hrs \\ \cline{3-7}
+    \multirow{3}{*}[0mm]{MPC \cite{jiang2019improving, jiang_further_2021}} & \multirow{2}{*}[0mm]{\makecell[l]{Didi Callcenter,\\Didi Dictation,\\Open Mandarin}} & ASR-zh & HKUST\textsuperscript{\ref{footnote:official-split}} & HKUST\textsuperscript{\ref{footnote:official-split}} & \checkmark & 168 hrs \\ \cline{3-7}
     & & ASR-zh & AISHELL-1\textsuperscript{\ref{footnote:official-split}} & AISHELL-1\textsuperscript{\ref{footnote:official-split}} & \checkmark & 178 hrs\vspace{1mm} \\ \cline{2-7}
@@
     % TERA
-    \multirow{4}{*}[0mm]{TERA \cite{liu2021tera}} & \multirow{4}{*}[0mm]{\makecell[l]{LS 100/360/\\\:\:\:960 hrs}} & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
+    \multirow{4}{*}[0mm]{TERA \cite{liu_tera_2021}} & \multirow{4}{*}[0mm]{\makecell[l]{LS 100/360/\\\:\:\:960 hrs}} & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
     & & SID & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
@@
     % HuBERT
-    HuBERT \cite{hsu2021hubert} & \makecell[l]{LS 960 hrs,\\ LL 60k hrs} & ASR & LS 960 hrs & \makecell[l]{LS test-clean\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \hline
+    HuBERT \cite{hsu_hubert_2021} & \makecell[l]{LS 960 hrs,\\ LL 60k hrs} & ASR & LS 960 hrs & \makecell[l]{LS test-clean\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \hline
     %wav2vec-c
-    wav2vec-c \cite{sadhu21_interspeech} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
-    UniSpeech-SAT \cite{chen2021unispeechsat} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See SUPERB \cite{yang21c_interspeech} paper for details \\ \hline
-    WavLM \cite{chen2021wavlm} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See \cite{yang21c_interspeech} for details \\ \hline
+    wav2vec-c \cite{sadhu_wav2vec-c_2021} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
+    UniSpeech-SAT \cite{chen2021unispeechsat} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See SUPERB \cite{yang_superb_2021} paper for details \\ \hline
+    WavLM \cite{chen_wavlm_2021} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See \cite{yang_superb_2021} for details \\ \hline
     \multirow{4}{*}[0mm]{XLS-R \cite{babu2021xlsr}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;\;+ CV-dataset-7k + VL\\\;\;\;+ BBL}} & ASR & VP, MLS, CV-dataset, BBL, LS & VP, MLS, CV-dataset, BBL, LS & - & \multirow{4}{*}[0mm]{See \cite{babu2021xlsr} for details} \\ \cline{3-6}
@@
 Train/test split made available
-by \cite{oord2018representation} on Google drive
+by \cite{oord_representation_2018} on Google drive
 https://drive.google.com/drive/folders/1BhJ2umKH3whguxMwifaKtSra0TgAbtfb.}
@@
 The 6 AED datasets used in
-\cite{tagliasacchi2020pre} are MUSAN~\cite{snyder2015musan}, Bird Audio
+\cite{tagliasacchi_pre_2020} are MUSAN~\cite{snyder2015musan}, Bird Audio
 Detection~\cite{stowell2019automatic}, Speech Commands~\cite{warden2018speech},
@@
 CPC modified~\cite{riviere2020unsupervised}, 
-wav2vec 2.0~\cite{baevski2020wav2vec}, and HuBERT~\cite{hsu2021hubert}, Alexa internal
-datasets (10k hours) for wav2vec-c~\cite{sadhu21_interspeech}, Didi internal
-datasets (10k hours) for MPC~\cite{jiang2019improving, jiang2021further}, the
+wav2vec 2.0~\cite{baevski2020wav2vec}, and HuBERT~\cite{hsu_hubert_2021}, Alexa internal
+datasets (10k hours) for wav2vec-c~\cite{sadhu_wav2vec-c_2021}, Didi internal
+datasets (10k hours) for MPC~\cite{jiang2019improving, jiang_further_2021}, the
 combination of Gigaspeech, VP-24k, and LL (94k hours in total) for
-UniSpeech-SAT~\cite{chen2021unispeechsat} and WavLM~\cite{chen2021wavlm}, and
+UniSpeech-SAT~\cite{chen2021unispeechsat} and WavLM~\cite{chen_wavlm_2021}, and
 the combination of VP-400K, MLS, CV-dataset, VL and BBL (436k hours in total)
@@
 in available computing power. Most studies focuses on learning representations
-for English, whereas Chinese~\cite{jiang2019improving, jiang2021further} and
+for English, whereas Chinese~\cite{jiang2019improving, jiang_further_2021} and
 multilinguality~\cite{kawakami2020learning, babu2021xlsr} are also gaining
@@
 fine-tuning also varies from several minutes up to 1k hours. Recent
-benchmarks such as SUPERB~\cite{yang21c_interspeech} that consolidate multiple 
+benchmarks such as SUPERB~\cite{yang_superb_2021} that consolidate multiple 
 downstream tasks have gained attention for evaluating SSL
-methodologies~\cite{chen2021unispeechsat, chen2021wavlm}. The goal of such benchmarks is
+methodologies~\cite{chen2021unispeechsat, chen_wavlm_2021}. The goal of such benchmarks is
 to provide a holistic evaluation of the performance of learned representations; 
@@
 publication. We also compare with baselines including the log mel filterbank (fBANK) and semi-supervised, self-training
-approaches (iterative pseudo labeling (IPL)~\cite{xu2020iterative}, 
-slimIPL~\cite{likhomanenko2020slimipl}, noisy student~\cite{park2020improved}).
+approaches (iterative pseudo labeling (IPL)~\cite{xu_iterative_2020}, 
+slimIPL~\cite{likhomanenko2020slimipl}, noisy student~\cite{park_improved_2020}).
 These approaches are visualized in black.
@@
 show performance from such methods (iterative pseudo labeling 
-(IPL)~\cite{xu2020iterative}, slimIPL~\cite{likhomanenko2020slimipl}, noisy 
-student~\cite{park2020improved}), as well as the current state of the art---conformer XXL + noisy
+(IPL)~\cite{xu_iterative_2020}, slimIPL~\cite{likhomanenko2020slimipl}, noisy 
+student~\cite{park_improved_2020}), as well as the current state of the art---conformer XXL + noisy
 student~\cite{zhang2020pushing}---which augments SSL with various advanced
 techniques including self-training. Furthermore, we illustrate in the figure
-the performance of a baseline system \cite{yang21c_interspeech} based on log mel filterbank (fBANK), which is one of the most commonly used features designed by domain experts.
+the performance of a baseline system \cite{yang_superb_2021} based on log mel filterbank (fBANK), which is one of the most commonly used features designed by domain experts.
 As observed in the figure, most SSL techniques outperform fBANK
@@
     \midrule
-    ASR (WER $\downarrow$) & LS test-clean/other & 2.1/4.0 \cite{xu2020iterative} & 1.4/2.6 \cite{zhang2020pushing} \\ \hline
+    ASR (WER $\downarrow$) & LS test-clean/other & 2.1/4.0 \cite{xu_iterative_2020} & 1.4/2.6 \cite{zhang2020pushing} \\ \hline
     IC (Acc $\uparrow$) & FSC & 98.8 \cite{lugosch19_interspeech} & 99.3\cite{chen2021unispeechsat} \\ \hline
-    SID (Acc $\uparrow$) & VoxCeleb1 & 94.8 \cite{hajibabaei2018unified} & 95.5 \cite{chen2021wavlm} \\ \hline
+    SID (Acc $\uparrow$) & VoxCeleb1 & 94.8 \cite{hajibabaei2018unified} & 95.5 \cite{chen_wavlm_2021} \\ \hline
     ASV (EER $\downarrow$) & VoxCeleb1 & 3.1 \cite{hajavi2021siamese} & 2.4 \cite{wang2021fine} \\ \hline
@@
     %wav2vec-c
-    %wav2vec-c \cite{sadhu21_interspeech} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
-    %UniSpeech-SAT \cite{chen2021unispeechsat} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & Check SUPERB \cite{yang21c_interspeech} paper for details \\ \hline
-    %WavLM \cite{chen2021wavlm} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & Check \cite{yang21c_interspeech} for details \\ \hline
+    %wav2vec-c \cite{sadhu_wav2vec-c_2021} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
+    %UniSpeech-SAT \cite{chen2021unispeechsat} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & Check SUPERB \cite{yang_superb_2021} paper for details \\ \hline
+    %WavLM \cite{chen_wavlm_2021} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & Check \cite{yang_superb_2021} for details \\ \hline
     %\multirow{4}{*}[0mm]{XLS-R \cite{babu2021xlsr}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;\;+ CV-dataset-7k + VL\\\;\;\;+ BBL}} & ASR & VP, MLS, CV-dataset, BBL, LS & VP, MLS, CV-dataset, BBL, LS & - & \multirow{4}{*}[0mm]{Check \cite{babu2021xlsr} for details} \\ \cline{3-6}
@@
 benchmarks with shared downstream tasks, datasets, and downstream recipes. Such
-efforts include SUPERB~\cite{yang21c_interspeech}, 
+efforts include SUPERB~\cite{yang_superb_2021}, 
 LeBenchmark~\cite{evain21_interspeech}, ZeroSpeech~\cite{dunbar2020zero},
@@
 
-SUPERB~\cite{yang21c_interspeech} is a benchmarking platform that allows the
+SUPERB~\cite{yang_superb_2021} is a benchmarking platform that allows the
 SSL community to train, evaluate, and compare speech representations on

### 2023-03-16 chapters/paper_review/04b_analysis.tex
Change reference keys for review paper (WIP)

@@
 
-In \cite{pasad2021} wav2vec~2.0 representations were analyzed with
+In \cite{pasad_layerwise_2021} wav2vec~2.0 representations were analyzed with
 respect to their acoustic-linguistic information content at different
@@
 for layers 10--21 and higher error rates for the other layers. 
-\cite{pasad2021} further show that fine-tuning the pre-trained model with a
+\cite{pasad_layerwise_2021} further show that fine-tuning the pre-trained model with a
 character-level CTC training criterion changes the behavior of the
@@
 phonetic information whereas speaker recognition uses a combination of
-acoustic and phonetic information. In a recent study~\cite{chen2021wavlm} the
+acoustic and phonetic information. In a recent study~\cite{chen_wavlm_2021} the
 same technique was used to identify layer contributions for the downstream
@@
 Several studies have demonstrated better downstream performance when using
-larger datasets~\cite{rivi20, kawakami2020learning, chen2021wavlm}. For
+larger datasets~\cite{rivi20, kawakami2020learning, chen_wavlm_2021}. For
 example,
@@
 to play a role, although this was not quantified in this study. However, recent
-experiments with larger and more diverse data collections~\cite{chen2021wavlm}
+experiments with larger and more diverse data collections~\cite{chen_wavlm_2021}
 confirm this assumption, as do 
@@
 Using the Mockingjay 
-model~\cite{liu2020mockingjay}, the authors in \cite{pue2021scaling} attempt to establish a relationship between
+model~\cite{liu_mockingjay_2020}, the authors in \cite{pue2021scaling} attempt to establish a relationship between
 model size and self-supervised \ensuremath{L_1} loss and demonstrate that it approximately

### 2023-03-16 chapters/paper_review/05_zero.tex
Change reference keys for review paper (WIP)

@@
   \midrule
-  \cite{chung2018unsupervised} & Speech2vec~\cite{chung2018speech2vec} & BES-GMM~\cite{Kamper17BESGMM} &  Word2Vec & \edit{Adversarial Training~\cite{conneau2017word}} & Self-training & SWC (Acc): 10.9\%  \\ %Submitted on 18 May 2018
+  \cite{chung2018unsupervised} & Speech2vec~\cite{chung_speech2vec_2018} & BES-GMM~\cite{kamper_segmental_2017} &  Word2Vec & \edit{Adversarial Training~\cite{conneau2017word}} & Self-training & SWC (Acc): 10.9\%  \\ %Submitted on 18 May 2018
 %https://arxiv.org/pdf/1805.07467.pdf %The text embeddings were obtained by training Word2Vec on the transcriptions using the fastText implementation without subword information [3].
@@
 \cite{chung2018unsupervised,chung2019towards} use adversarial training to map the word-level
-speech embedding space~\cite{chung2018speech2vec} to the word embedding space
+speech embedding space~\cite{chung_speech2vec_2018} to the word embedding space
 and achieve promising performance on spoken word classification, speech
@@
 phoneme recognition by first clustering phoneme-level
-embeddings~\cite{Chung2016AudioWord2Vec,Wang2018SegAudioWord2Vec} into a set of
+embeddings~\cite{chung_audio_2016,Wang2018SegAudioWord2Vec} into a set of
 tokens and then using GAN to learn the mapping relationship between tokens and
@@
 However, the progress of TTS systems boosted by deep 
-learning~\cite{oord2016wavenet,shen2018natural} has inspired another interesting and
+learning~\cite{oord_wavenet_2016,shen2018natural} has inspired another interesting and
 straightforward research direction: \edit{artificially creating paired text and audio
@@
 models~\cite{ondel2016variational,heck2017feature}, deep 
-autoencoders~\cite{tjandra19_interspeech,chorowski2019unsupervised}, and incorporation of
+autoencoders~\cite{tjandra19_interspeech,chorowski_unsupervised_2019}, and incorporation of
 neural-network-based TTS/VC techniques~\cite{niekerk20b_interspeech} to
@@
 directly from audio, bypassing the need for text or 
-labels~\cite{GSLM,SpeechResynthesis_IS21,pGSLM,emotion_conversion}. 
+labels~\cite{nguyen_generative_2022,polyak_speech_2021,kharitonov_textfree_2021,emotion_conversion}. 
 Not only does this open the gate for language and dialect modeling without
@@
 emotion, hesitation, interruptions. 
-The generative spoken language model (GSLM)~\cite{GSLM} utilizes discrete
+The generative spoken language model (nguyen_generative_2022)~\cite{nguyen_generative_2022} utilizes discrete
 representations from wav2vec~2.0, HuBERT, and CPC algorithms as inputs to an
@@
 model and generate
-dialogues~\cite{dialogue_GSLM}.\footnote{https://speechbot.github.io/dgslm/}
+dialogues~\cite{dialogue_nguyen_generative_2022}.\footnote{https://speechbot.github.io/dgslm/}
 %\cite{}. 
-Given its flexibility in modeling spoken content, the GSLM has been further extended
-to jointly model content and prosody~\cite{pGSLM}. This prosodic-GSLM model
+Given its flexibility in modeling spoken content, the nguyen_generative_2022 has been further extended
+to jointly model content and prosody~\cite{kharitonov_textfree_2021}. This prosodic-nguyen_generative_2022 model
 introduced a multistream causal Transformer, where the input and output layers
 use multiple heads to model three channels:  discrete speech units, duration,
-and quantized pitch. The prosodic-GSLM model jointly generates novel content
+and quantized pitch. The prosodic-nguyen_generative_2022 model jointly generates novel content
 and prosody congruently in the expressive 

### 2023-03-16 chapters/paper_review/06_conclusions.tex
Change reference keys for review paper (WIP)

@@
 Adapters~\cite{pmlr-v97-houlsby19a,zaken2021bitfit,guo-etal-2021-parameter} are lightweight modules inserted into SSL models, and in downstream tasks, the parameters of SSL models are frozen, and only the adapters are trained.
-The prompt/instruction learning methods~\cite{liu2021pretrain} also freeze the SSL parameters and control the output of SSL by adding additional information, which is called \textit{prompt}, in the input.
+The prompt/instruction learning methods~\cite{liu_pretrain_2021} also freeze the SSL parameters and control the output of SSL by adding additional information, which is called \textit{prompt}, in the input.
 Both adapter-based methods and prompt/instruction learning yield competitive performance compared with fine-tuning in NLP applications, but there is only little related work for speech~\cite{SpeechAdapter,SpeechPrompt}.
@@
 \item \textbf{Feature Disentanglement.} 
-Speech SSL models show strengths on a surprisingly wide range of tasks~\cite{yang21c_interspeech}, suggesting that representations contain different information.
+Speech SSL models show strengths on a surprisingly wide range of tasks~\cite{yang_superb_2021}, suggesting that representations contain different information.
 One way to further improve downstream tasks is to disentangle different information from the representation.
@@
 \edit{\item \textbf{Capturing higher-level semantic information.} Although many representation learning approaches can go beyond low-level phonetic modeling to capture some lexical information~\cite{TuAnh_2022_discrete}, they still struggle in higher-level semantic tasks easily captured by word-level counterparts like BERT.
-One workaround is two-stage training~\cite{pGSLM, dialogue_GSLM}; however, this prevents propagating rich lexical and semantic knowledge modeled in the second stage to benefit the phonetically focused first stage.
+One workaround is two-stage training~\cite{kharitonov_textfree_2021, dialogue_nguyen_generative_2022}; however, this prevents propagating rich lexical and semantic knowledge modeled in the second stage to benefit the phonetically focused first stage.
 }
@@
 \item \textbf{Using text representation models to improve speech representation.} The amount of content information in speech corpora used to train speech representation models is far less than that of text representation models.
-Noting that the BERT training corpus exceeds 3 billion words~\cite{jacob2019BERT}, and assuming a typical speaking rate of 120 words per minute, a speech corpus containing the same content as the BERT training data would include 400,000 hours of audio, which exceeds the \edit{accumulated} training data of all current speech representation models.
+Noting that the BERT training corpus exceeds 3 billion words~\cite{devlin_bert:_2018}, and assuming a typical speaking rate of 120 words per minute, a speech corpus containing the same content as the BERT training data would include 400,000 hours of audio, which exceeds the \edit{accumulated} training data of all current speech representation models.
 Therefore, to enable speech representation models to better learn human language, for instance by extracting semantic information from acoustic signals, the use of text models such as BERT and GPT  \edit{seems} key: nevertheless, how to use these to improve speech representation model pre-training remains an open question.
-%\edit{ There is already some study using both speech and text data to pre-training, but some paired data is still required to achieve good performance~\cite{bapna2021slam}.  }
+%\edit{ There is already some study using both speech and text data to pre-training, but some paired data is still required to achieve good performance~\cite{bapna_slam_2021}.  }
 

### 2023-03-16 chapters/paper_review/main_hi.tex
Change reference keys for review paper (WIP)

@@
         \multicolumn{7}{c}{\textit{\textbf{Self-Training}}} \\
-        IPL~\cite{xu2020iterative} & LL-60k & 4-gram + Transformer & 1.85 & 3.26 & 2.10 & 4.01 \\
-        Noisy Student~\cite{park2020improved} & LV-60k & LSTM & 1.6 & 3.4 & 1.7 & 3.4 \\
+        IPL~\cite{xu_iterative_2020} & LL-60k & 4-gram + Transformer & 1.85 & 3.26 & 2.10 & 4.01 \\
+        Noisy Student~\cite{park_improved_2020} & LV-60k & LSTM & 1.6 & 3.4 & 1.7 & 3.4 \\
 

### 2023-03-16 chapters/paper_review/main_lo.tex
Change reference keys for review paper (WIP)

@@
         % DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & - & - \\
-        DiscreteBERT~\cite{baevski2019effectiveness} & LS-960 & 4-gram & 15.7 & 24.1 & 16.3 & 25.2 \\
+        DiscreteBERT~\cite{baevski_effectiveness_2020} & LS-960 & 4-gram & 15.7 & 24.1 & 16.3 & 25.2 \\
         wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 8.9 & 15.7 & 9.1 & 15.6 \\
@@
         DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & 13.8 & 29.1 \\
-        DiscreteBERT~\cite{baevski2019effectiveness} & LS-960 & 4-gram & 8.5 & 16.4 & 9.0 & 17.6 \\
+        DiscreteBERT~\cite{baevski_effectiveness_2020} & LS-960 & 4-gram & 8.5 & 16.4 & 9.0 & 17.6 \\
         wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 5.0 & 10.8 & 5.5 & 11.3 \\
@@
         DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & 5.4 & 13.3 \\
-        DiscreteBERT~\cite{baevski2019effectiveness} & LS-960 & 4-gram & 5.3 & 13.2 & 5.9 & 14.1 \\
+        DiscreteBERT~\cite{baevski_effectiveness_2020} & LS-960 & 4-gram & 5.3 & 13.2 & 5.9 & 14.1 \\
         wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 3.8 & 9.1 & 4.3 & 9.5 \\
@@
         % LPM~\cite{hsu2020semi} & LL-60K & ConvLM & 4.87 & 13.84 & 4.88 & 15.28 \\
-        IPL~\cite{xu2020iterative} & LL-60k & 4-gram + Transformer & 3.19 & 6.14 & 3.72 & 7.11 \\
+        IPL~\cite{xu_iterative_2020} & LL-60k & 4-gram + Transformer & 3.19 & 6.14 & 3.72 & 7.11 \\
         SlimIPL~\cite{likhomanenko2020slimipl} & LS-860 & 4-gram + Transformer & 2.2 & 4.6 & 2.7 & 5.2 \\
-        Noisy Student\cite{park2020improved} & LS-860 & LSTM & 3.9 & 8.8 & 4.2 & 8.6 \\
+        Noisy Student\cite{park_improved_2020} & LS-860 & LSTM & 3.9 & 8.8 & 4.2 & 8.6 \\
         DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & 5.0 & 12.1 \\
-        DiscreteBERT~\cite{baevski2019effectiveness} & LS-960 & 4-gram & 4.0 & 10.9 & 4.5 & 12.1 \\
+        DiscreteBERT~\cite{baevski_effectiveness_2020} & LS-960 & 4-gram & 4.0 & 10.9 & 4.5 & 12.1 \\
         wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 2.7 & 7.9 & 3.4 & 8.0 \\

### 2023-03-16 chapters/paper_review/model.tex
Change reference keys for review paper (WIP)

@@
 
-Audio Word2vec \cite{Chung2016AudioWord2Vec}      & 2016 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
+Audio Word2vec \cite{chung_audio_2016}      & 2016 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
 
-VQ-VAE \cite{vqvae} & 2017 Nov. & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark \\ %2 Nov 2017
+VQ-VAE \cite{oord_neural_2018} & 2017 Nov. & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark \\ %2 Nov 2017
 
-Speech2Vec \cite{chung2018speech2vec}     & 2018 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
+Speech2Vec \cite{chung_speech2vec_2018}     & 2018 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
 Unspeech \cite{milde2018unspeech}         & 2018 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
 
-CPC \cite{oord2018representation}         & 2018 Jul. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\
+CPC \cite{oord_representation_2018}         & 2018 Jul. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\
 
-PASE \cite{pascual2019learning}                   & 2019 Apr. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 6/4
+PASE \cite{pascual_learning_2019}                   & 2019 Apr. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 6/4
 
-APC \cite{chung2019unsupervised}          & 2019 Oct. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 5/4
+APC \cite{chung_unsupervised_2019}          & 2019 Oct. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 5/4
 
-wav2vec \cite{schneider2019wav2vec}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
+wav2vec \cite{schneider_wav2vec:_2019}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
 
-vq-wav2vec \cite{Baevski2020vq-wav2vec} & 2019 Oct.
+vq-wav2vec \cite{baevski_vq-wav2vec_2020} & 2019 Oct.
 & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ %12 Oct 2019 %Lee: vq-wav2vec is CON or CLF???
@@
 
-Discrete BERT \cite{Baevski2020vq-wav2vec,baevski2019effectiveness} & 2019 Oct.
+Discrete BERT \cite{baevski_vq-wav2vec_2020,baevski_effectiveness_2020} & 2019 Oct.
 & \xmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ %12 Oct 2019 %10 Nov 2019 %In the paper of vq-wav2vec, it also includes the approach of Discrete BERT.
@@
 
-Mockingjay  \cite{liu2020mockingjay}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 25/10-19
+Mockingjay  \cite{liu_mockingjay_2020}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 25/10-19
 
@@
 
-PASE+ \cite{ravanelli2020multi}                & 2020 Jan. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 25/1 % I think contrastive always have learned target.
+PASE+ \cite{ravanelli_multi_2020}                & 2020 Jan. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 25/1 % I think contrastive always have learned target.
 
@@
 
-VQ-APC \cite{chung20e_interspeech}  & 2020 May  & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark \\ %17/05
+VQ-APC \cite{chung_vector-quantized_2020}  & 2020 May  & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark \\ %17/05
 
@@
 
-TERA \cite{liu2021tera}                   & 2020 Jul. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 12/7
+TERA \cite{liu_tera_2021}                   & 2020 Jul. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 12/7
 
@@
 
-wav2vec-C \cite{sadhu21_interspeech} & 2021 Mar. & \cmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ %9 Mar 2021 %Lee:  consistency loss is reconstruction
+wav2vec-C \cite{sadhu_wav2vec-c_2021} & 2021 Mar. & \cmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ %9 Mar 2021 %Lee:  consistency loss is reconstruction
 
-HuBERT \cite{hsu2021hubert}               & 2021 Jun. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ % 14/6
+HuBERT \cite{hsu_hubert_2021}               & 2021 Jun. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ % 14/6
 
-W2v-bert \cite{w2vbert} & 2021 Aug & \xmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark \\
+W2v-bert \cite{chung_w2vbert_2021} & 2021 Aug & \xmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark \\
 %7 Aug 2021
 
-pMPC \cite{yue2021pMPC}  & 2021 Aug. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ %30 August – 3 September, 2021 (I do not find arXiv version.)
+pMPC \cite{yue_pmpc_2021}  & 2021 Aug. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ %30 August – 3 September, 2021 (I do not find arXiv version.)
 
-WavLM \cite{chen2021wavlm} & 2021 Oct. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ %26 Oct 2021
+WavLM \cite{chen_wavlm_2021} & 2021 Oct. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ %26 Oct 2021
 
-data2vec \cite{data2vec}     & 2022 Jan. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 20/1
+data2vec \cite{baevski_data2vec_2022}     & 2022 Jan. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 20/1
 
@@
 %                                                         REG      CLF      CON      MSK      LTG      QTZ      EMB  
-Audio Word2vec \cite{Chung2016AudioWord2Vec}      & 2016 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
-Speech2Vec \cite{chung2018speech2vec}     & 2018 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
+Audio Word2vec \cite{chung_audio_2016}      & 2016 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
+Speech2Vec \cite{chung_speech2vec_2018}     & 2018 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
 % Unspeech \cite{milde2018unspeech}         & 2018 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
-CPC \cite{oord2018representation}         & 2018 Jul. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\
+CPC \cite{oord_representation_2018}         & 2018 Jul. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\
 
-PASE \cite{pascual2019learning}                   & 2019 Apr. & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark \\ % 6/4
+PASE \cite{pascual_learning_2019}                   & 2019 Apr. & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark \\ % 6/4
 
-APC \cite{chung2019unsupervised}          & 2019 Oct. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 5/4
-wav2vec \cite{schneider2019wav2vec}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
+APC \cite{chung_unsupervised_2019}          & 2019 Oct. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 5/4
+wav2vec \cite{schneider_wav2vec:_2019}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
 %VQ-APC \cite{chung2020vector}             & 2020 May  & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ 
-Mockingjay  \cite{liu2020mockingjay}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 25/10-19
+Mockingjay  \cite{liu_mockingjay_2020}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 25/10-19
 %DeCoAR \cite{ling2020deep}                & 2019 Dec. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 3/12-19
@@
 wav2vec 2.0 \cite{baevski2020wav2vec}     & 2020 Jun. & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark \\ % 20/6
-%TERA \cite{liu2021tera}                   & 2021 Jul. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 12/7
+%TERA \cite{liu_tera_2021}                   & 2021 Jul. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 12/7
 
@@
 DeCoAR 2.0 \cite{ling2020decoar}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 11/12
-HuBERT \cite{hsu2021hubert}               & 2021 Jun. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ % 14/6
-data2vec \cite{data2vec}     & 2022 Jan. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 20/1
+HuBERT \cite{hsu_hubert_2021}               & 2021 Jun. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ % 14/6
+data2vec \cite{baevski_data2vec_2022}     & 2022 Jan. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 20/1
 \bottomrule

### 2023-03-16 chapters/paper_review/pretext.tex
Change reference keys for review paper (WIP)

@@
     \midrule
-    Audio Word2vec~\cite{Chung2016AudioWord2Vec}, VQ-VAE~\cite{vqvae}     & $X$ &   \textsc{-}    &  $X$  \\ %It has VQ as extra constraint.
+    Audio Word2vec~\cite{chung_audio_2016}, VQ-VAE~\cite{oord_neural_2018}     & $X$ &   \textsc{-}    &  $X$  \\ %It has VQ as extra constraint.
     \midrule  
-    Speech2Vec~\cite{chung2018speech2vec}, Audio2Vec~\cite{tagliasacchi2020pre} - skip-gram    & $X_{[t_1,t_2]}$  &     \textsc{-}   &    $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$     \\
+    Speech2Vec~\cite{chung_speech2vec_2018}, Audio2Vec~\cite{tagliasacchi_pre_2020} - skip-gram    & $X_{[t_1,t_2]}$  &     \textsc{-}   &    $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$     \\
     \midrule 
-    Speech2Vec~\cite{chung2018speech2vec}, Audio2Vec~\cite{tagliasacchi2020pre} - cbow    & $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$   &     \textsc{-}   &    $X_{[t_1,t_2]}$   \\
+    Speech2Vec~\cite{chung_speech2vec_2018}, Audio2Vec~\cite{tagliasacchi_pre_2020} - cbow    & $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$   &     \textsc{-}   &    $X_{[t_1,t_2]}$   \\
     \midrule 
-    PASE~\cite{pascual2019learning}, PASE+~\cite{ravanelli2020multi}\footnote{PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
+    PASE~\cite{pascual_learning_2019}, PASE+~\cite{ravanelli_multi_2020}\footnote{PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
     \midrule 
-    APC~\cite{chung2019unsupervised,chung20e_interspeech}         & $X_{[1,t]}$   & \textsc{-}              & $x_{t+c},\, c\geq1$    \\ %VQ-APC is the same as APC
+    APC~\cite{chung_unsupervised_2019,chung_vector-quantized_2020}         & $X_{[1,t]}$   & \textsc{-}              & $x_{t+c},\, c\geq1$    \\ %VQ-APC is the same as APC
     \midrule
@@
     \midrule
-    Mockingjay~\cite{liu2020mockingjay}, Audio ALBERT~\cite{chi2020audio}, DeCoAR 2.0~\cite{ling2020decoar}   & \multicolumn{2}{c|}{$X_{-[t,t+k]}$}   & $X_{[t,t+k]}$    \\
+    Mockingjay~\cite{liu_mockingjay_2020}, Audio ALBERT~\cite{chi2020audio}, DeCoAR 2.0~\cite{ling2020decoar}   & \multicolumn{2}{c|}{$X_{-[t,t+k]}$}   & $X_{[t,t+k]}$    \\
     \midrule 
-    TERA~\cite{liu2021tera}, BMR~\cite{wang2020unsupervised}  & \multicolumn{2}{c|}{$X_{-[t,t+k]}^{-[f,f+j]}$}       & $X$       \\
+    TERA~\cite{liu_tera_2021}, BMR~\cite{wang_unsupervised_2020}  & \multicolumn{2}{c|}{$X_{-[t,t+k]}^{-[f,f+j]}$}       & $X$       \\
     \midrule
-    pMPC~\cite{yue2021pMPC}      &  \multicolumn{2}{c|}{$X_{-[t,t+k^\prime]}$ ($X_{[t,t+k^\prime]}$ is a phoneme)}       & $X_{[t,t+k^\prime]}$    \\
+    pMPC~\cite{yue_pmpc_2021}      &  \multicolumn{2}{c|}{$X_{-[t,t+k^\prime]}$ ($X_{[t,t+k^\prime]}$ is a phoneme)}       & $X_{[t,t+k^\prime]}$    \\
     \midrule 
@@
     \midrule 
-    CPC~\cite{oord2018representation}, wav2vec \cite{schneider2019wav2vec}, Modified CPC \cite{riviere2020unsupervised}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
+    CPC~\cite{oord_representation_2018}, wav2vec \cite{schneider_wav2vec:_2019}, Modified CPC \cite{riviere2020unsupervised}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
     \midrule 
@@
     \midrule 
-    vq-wav2vec \cite{Baevski2020vq-wav2vec}     &   $X_{[1,t]}$ &   \textsc{-}    &   $q_{t+c},\, c\geq1$   \\ 
+    vq-wav2vec \cite{baevski_vq-wav2vec_2020}     &   $X_{[1,t]}$ &   \textsc{-}    &   $q_{t+c},\, c\geq1$   \\ 
     \midrule 
-    wav2vec 2.0 \cite{baevski2020wav2vec}, wav2vec-C \cite{sadhu21_interspeech}\footnote{wav2vec-C adds reconstruction loss to wav2vec 2.0.}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
+    wav2vec 2.0 \cite{baevski2020wav2vec}, wav2vec-C \cite{sadhu_wav2vec-c_2021}\footnote{wav2vec-C adds reconstruction loss to wav2vec 2.0.}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
     \midrule 
-    w2v-BERT \cite{w2vbert}     &$X$ &    $Z_{-[t,t+k]}$   &     $Q_{[t,t+k]}$ and $C_{[t,t+k]}$     \\
+    w2v-BERT \cite{chung_w2vbert_2021}     &$X$ &    $Z_{-[t,t+k]}$   &     $Q_{[t,t+k]}$ and $C_{[t,t+k]}$     \\
     \midrule
@@
     \midrule
-    Discrete BERT~\cite{Baevski2020vq-wav2vec,baevski2019effectiveness} \footnote{Discrete BERT obtains codes $C$ from vq-wav2vec.}      &   \multicolumn{2}{c|}{$C_{-[t,t+k]}$}   & $C_{[t,t+k]}$  \\
+    Discrete BERT~\cite{baevski_vq-wav2vec_2020,baevski_effectiveness_2020} \footnote{Discrete BERT obtains codes $C$ from vq-wav2vec.}      &   \multicolumn{2}{c|}{$C_{-[t,t+k]}$}   & $C_{[t,t+k]}$  \\
     \midrule 
-    HuBERT \cite{hsu2021hubert}\footnote{HuBERT is trained first using cluster IDs of the MFCCs as target and subsequently clusters IDs of the model representations from the last iteration.}, WavLM \cite{chen2021wavlm}\footnote{WavLM simulates noisy/overlapped speech as inputs.}  & $X$             & $Z_{-[t,t+k]}$          & $C_{[t,t+k]}$  \\
+    HuBERT \cite{hsu_hubert_2021}\footnote{HuBERT is trained first using cluster IDs of the MFCCs as target and subsequently clusters IDs of the model representations from the last iteration.}, WavLM \cite{chen_wavlm_2021}\footnote{WavLM simulates noisy/overlapped speech as inputs.}  & $X$             & $Z_{-[t,t+k]}$          & $C_{[t,t+k]}$  \\
     \midrule
-    data2vec \cite{data2vec}    & $X$             & $Z_{-[t,t+k]}$          & $\sum_{l}\bar{H}^{(l)}_{[t,t+k]}$  \\ 
+    data2vec \cite{baevski_data2vec_2022}    & $X$             & $Z_{-[t,t+k]}$          & $\sum_{l}\bar{H}^{(l)}_{[t,t+k]}$  \\ 
     \midrule 
@@
     \midrule
-    Audio Word2vec~\cite{Chung2016AudioWord2Vec}, VQ-VAE \cite{vqvae}     & $X$ &   \textsc{-}    &  $X$  \\ %It has VQ as extra constraint.
+    Audio Word2vec~\cite{chung_audio_2016}, VQ-VAE \cite{oord_neural_2018}     & $X$ &   \textsc{-}    &  $X$  \\ %It has VQ as extra constraint.
     \midrule  
-    Speech2Vec~\cite{chung2018speech2vec}, Audio2Vec~\cite{tagliasacchi2020pre} - skip-gram    & $X_{[t_1,t_2]}$  &     \textsc{-}   &    $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$     \\
+    Speech2Vec~\cite{chung_speech2vec_2018}, Audio2Vec~\cite{tagliasacchi_pre_2020} - skip-gram    & $X_{[t_1,t_2]}$  &     \textsc{-}   &    $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$     \\
     \midrule 
-   Speech2Vec~\cite{chung2018speech2vec}, Audio2Vec~\cite{tagliasacchi2020pre} - cbow    & $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$   &     \textsc{-}   &    $X_{[t_1,t_2]}$   \\
+   Speech2Vec~\cite{chung_speech2vec_2018}, Audio2Vec~\cite{tagliasacchi_pre_2020} - cbow    & $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$   &     \textsc{-}   &    $X_{[t_1,t_2]}$   \\
     \midrule 
-     PASE~\cite{pascual2019learning}, PASE+~\cite{ravanelli2020multi}\footnote{PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
+     PASE~\cite{pascual_learning_2019}, PASE+~\cite{ravanelli_multi_2020}\footnote{PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
     \midrule 
-    APC~\cite{chung2019unsupervised,chung20e_interspeech}         & $X_{[1,t]}$   & \textsc{-}              & $x_{t+c},\, c\geq1$    \\ %VQ-APC is the same as APC
+    APC~\cite{chung_unsupervised_2019,chung_vector-quantized_2020}         & $X_{[1,t]}$   & \textsc{-}              & $x_{t+c},\, c\geq1$    \\ %VQ-APC is the same as APC
     \midrule
@@
     \midrule 
-    CPC~\cite{oord2018representation}, wav2vec \cite{schneider2019wav2vec}, Modified CPC \cite{riviere2020unsupervised}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
+    CPC~\cite{oord_representation_2018}, wav2vec \cite{schneider_wav2vec:_2019}, Modified CPC \cite{riviere2020unsupervised}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
         \midrule 
@@
     \midrule 
-    vq-wav2vec \cite{Baevski2020vq-wav2vec}     &   $X_{[1,t]}$ &   \textsc{-}    &   $q_{t+c},\, c\geq1$   \\ 
+    vq-wav2vec \cite{baevski_vq-wav2vec_2020}     &   $X_{[1,t]}$ &   \textsc{-}    &   $q_{t+c},\, c\geq1$   \\ 
     \midrule 
-    wav2vec 2.0 \cite{baevski2020wav2vec}, wav2vec-C \cite{sadhu21_interspeech}\footnote{wav2vec-C adds the reconstruction loss to wav2vec 2.0.}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
+    wav2vec 2.0 \cite{baevski2020wav2vec}, wav2vec-C \cite{sadhu_wav2vec-c_2021}\footnote{wav2vec-C adds the reconstruction loss to wav2vec 2.0.}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
     \midrule 
-    w2v-BERT \cite{w2vbert}     &$X$ &    $Z_{-[t,t+k]}$   &     $Q_{[t,t+k]}$ and $C_{[t,t+k]}$     \\
+    w2v-BERT \cite{chung_w2vbert_2021}     &$X$ &    $Z_{-[t,t+k]}$   &     $Q_{[t,t+k]}$ and $C_{[t,t+k]}$     \\
     \midrule
@@
     \midrule
-    Discrete BERT~\cite{Baevski2020vq-wav2vec,baevski2019effectiveness} \footnote{Discrete BERT obtains codes $C$ from vq-wav2vec.}      &   \multicolumn{2}{c|}{$C_{-[t,t+k]}$}   & $C_{[t,t+k]}$  \\
+    Discrete BERT~\cite{baevski_vq-wav2vec_2020,baevski_effectiveness_2020} \footnote{Discrete BERT obtains codes $C$ from vq-wav2vec.}      &   \multicolumn{2}{c|}{$C_{-[t,t+k]}$}   & $C_{[t,t+k]}$  \\
     \midrule 
-    HuBERT \cite{hsu2021hubert}\footnote{HuBERT is trained first using cluster IDs of MFCCs as target and subsequently cluster IDs of model representations from the last iteration.}, WavLM \cite{chen2021wavlm}\footnote{WavLM simulates noisy/overlapped speech as inputs.}  & $X$             & $Z_{-[t,t+k]}$          & $C_{[t,t+k]}$  \\
+    HuBERT \cite{hsu_hubert_2021}\footnote{HuBERT is trained first using cluster IDs of MFCCs as target and subsequently cluster IDs of model representations from the last iteration.}, WavLM \cite{chen_wavlm_2021}\footnote{WavLM simulates noisy/overlapped speech as inputs.}  & $X$             & $Z_{-[t,t+k]}$          & $C_{[t,t+k]}$  \\
     \midrule
-    data2vec \cite{data2vec}    & $X$             & $Z_{-[t,t+k]}$          & $\sum_{l}\bar{H}^{(l)}_{[t,t+k]}$  \\ 
+    data2vec \cite{baevski_data2vec_2022}    & $X$             & $Z_{-[t,t+k]}$          & $\sum_{l}\bar{H}^{(l)}_{[t,t+k]}$  \\ 
         \midrule 
@@
 
-APC~\cite{chung2019unsupervised, chung2020generative,chung2020improved} & $X_{[1,t-1]}$  & $x_{t+c},c\geq0$ & VQ~\cite{chung20e_interspeech} &  GPT~\cite{alex2018GPT,alex2019GPT2,brown2020gpt3} \\ \hline
+APC~\cite{chung_unsupervised_2019, chung_generative_2020,chung2020improved} & $X_{[1,t-1]}$  & $x_{t+c},c\geq0$ & VQ~\cite{chung_vector-quantized_2020} &  GPT~\cite{alex2018GPT,radford_language_2019,brown2020gpt3} \\ \hline
 
-%$X_{[1,t-1]}$  & $x_{t+c},c\geq0$ & VQ & VQ-APC~\cite{chung20e_interspeech} & -  \\
+%$X_{[1,t-1]}$  & $x_{t+c},c\geq0$ & VQ & VQ-APC~\cite{chung_vector-quantized_2020} & -  \\
 
@@
 
-\tabincell{l}{ Mockingjay~\cite{liu2020mockingjay} \\ MPC~\cite{jiang2019improving,jiang2021further} \\ AALBERT~\cite{chi2021aalbert} \\
+\tabincell{l}{ Mockingjay~\cite{liu_mockingjay_2020} \\ MPC~\cite{jiang2019improving,jiang_further_2021} \\ AALBERT~\cite{chi2021aalbert} \\
 DeCoAR 2.0~\cite{ling2020decoar} } &
-$X_{-[t,t+k]}$ & $X_{[t,t+k]}$ & VQ~\cite{ling2020decoar} &  \tabincell{l}{BERT ~\cite{jacob2019BERT}\\RoBERTa~\cite{Liu2019RoBERTa}} \\ \hline
+$X_{-[t,t+k]}$ & $X_{[t,t+k]}$ & VQ~\cite{ling2020decoar} &  \tabincell{l}{BERT ~\cite{devlin_bert:_2018}\\RoBERTa~\cite{liu_roberta_2019}} \\ \hline
 
-pMPC~\cite{yue2021pMPC}   &
+pMPC~\cite{yue_pmpc_2021}   &
 \tabincell{l}{$X_{-[t,t+k^\prime]}$\\($X_{[t,t+k^\prime]}$: phoneme-level segment)} & $X_{[t,t+k^\prime]}$ & - &  SpanBert~\cite{joshi-etal-2020-spanbert} \\ \hline
@@
 
-\tabincell{l}{BMR~\cite{wang2020unsupervised}\\TERA~\cite{liu2021tera}} &
-$X_{-[t_1:t_2]}^{-[f_1,f_2]}$ & $X$ & dropout~\cite{luo2021drop} &  -  \\ \hline
+\tabincell{l}{BMR~\cite{wang_unsupervised_2020}\\TERA~\cite{liu_tera_2021}} &
+$X_{-[t_1:t_2]}^{-[f_1,f_2]}$ & $X$ & dropout~\cite{luo_drop_2021} &  -  \\ \hline
 
-\cite{chorowski2019unsupervised} &
-$S$ & $S$ & \tabincell{l}{VQ-VAE~\cite{chorowski2019unsupervised}\\time-jitter regularizer~\cite{chorowski2019unsupervised}} & 
-\tabincell{l}{\cite{van2017neural,VQVAE2} \\ (just name a few)} \\ \hline
+\cite{chorowski_unsupervised_2019} &
+$S$ & $S$ & \tabincell{l}{VQ-VAE~\cite{chorowski_unsupervised_2019}\\time-jitter regularizer~\cite{chorowski_unsupervised_2019}} & 
+\tabincell{l}{\cite{van2017neural,razavi_generating_2019} \\ (just name a few)} \\ \hline
 
-\cite{tagliasacchi2019self, tagliasacchi2020pre} & 
+\cite{tagliasacchi_self_2019, tagliasacchi_pre_2020} & 
 $X_{[t-k,t-1]},X_{[t+k+1,t+2k]}$ & $X_{[t,t+k]}$  & - & CBoW~\cite{mikolov2013efficient}  \\ \hline
 
-\cite{tagliasacchi2019self, tagliasacchi2020pre} &
+\cite{tagliasacchi_self_2019, tagliasacchi_pre_2020} &
 $X_{[t,t+k]}$ & $X_{[t-k,t-1]},X_{[t+k+1,t+2k]}$ & - &  Skipgram~\cite{mikolov2013efficient} \\ \hline
@@
 
-TemporalGap~\cite{tagliasacchi2019self, tagliasacchi2020pre} &
+TemporalGap~\cite{tagliasacchi_self_2019, tagliasacchi_pre_2020} &
 $X_{[t_1,t_1+k]}$, $X_{[t_2,t_2+k]}$ & $|t_1 - t_2|$ & - &  -  \\ \hline
@@
 
-PASE~\cite{pascual2019learning} &
+PASE~\cite{pascual_learning_2019} &
 $S$ & $X$: LPS, MFCC, Prosody & - &  -  \\ \hline
 
-PASE+~\cite{ravanelli2020multi} & 
+PASE+~\cite{ravanelli_multi_2020} & 
 $S^*$ & \tabincell{l}{$X$: LPS, MFCC, Prosody, \\ fbank, gammatone} & -  &  -  \\ \hline

### 2023-03-17 chapters/paper_review/03_approaches.tex
Change reference keys (WIP)

@@
 %\item VAE, VQ-VAE
-    \item DeCoAR 2.0~\cite{ling2020decoar} presents a deep contextualized acoustic representation learning approach with the addition of a vector quantization (VQ) layer.
+    \item DeCoAR 2.0~\cite{ling_decoar_2020} presents a deep contextualized acoustic representation learning approach with the addition of a vector quantization (VQ) layer.
     \item In VQ-APC~\cite{chung_vector-quantized_2020}, a VQ layer is used with the APC objective, which imposes a bottleneck and forces the model to learn better representations.
-    \item Two dropout regularization methods, attention dropout and layer dropout, are introduced to TERA~\cite{luo_drop_2021}. 
+    \item Two dropout regularization methods, attention dropout and layer dropout, are introduced to TERA~\cite{luo_dropout_2021}. 
 \end{itemize}
@@
 We can bring the pretext task closer to the NLP equivalent by using a masking policy where the masked regions of the input correspond to linguistic units.
-Instead of just masking a fixed number of consecutive frames, pMPC~\cite{yue_pmpc_2021} selects masked speech frames according to the phonetic segmentation in an utterance.
+Instead of just masking a fixed number of consecutive frames, pMPC~\cite{yue_phonetically_2021} selects masked speech frames according to the phonetic segmentation in an utterance.
 \edit{However, in order to obtain this segmentation, some labeled data is of course needed.}
@@
 Some studies explore alternatives to masking the input directly.
-In non-autoregressive predictive coding (NPC)~\cite{liu21l_interspeech}, time masking is introduced through masked convolution blocks.
-Taking inspiration from XLNet~\cite{XLNet}, it has also been suggested that the input be reconstructed from a shuffled version~\cite{song20d_interspeech} to address the discrepancy between pre-training and fine-tuning of masking-based approaches.
+In non-autoregressive predictive coding (NPC)~\cite{liu_non-autoregressive_2020}, time masking is introduced through masked convolution blocks.
+Taking inspiration from XLNet~\cite{yang_xlnet_2019}, it has also been suggested that the input be reconstructed from a shuffled version~\cite{song_speech-xlnet_2020} to address the discrepancy between pre-training and fine-tuning of masking-based approaches.
 
 Regularization methods can further improve on masked reconstruction approaches.
-\edit{DeCoAR 2.0~\cite{ling2020decoar} uses vector quantization}, which is shown to improve the learned representations.
-Furthermore, two dropout regularization methods---attention dropout and layer dropout---are introduced with the TERA model~\cite{liu_tera_2021, luo_drop_2021}.
+\edit{DeCoAR 2.0~\cite{ling_decoar_2020} uses vector quantization}, which is shown to improve the learned representations.
+Furthermore, two dropout regularization methods---attention dropout and layer dropout---are introduced with the TERA model~\cite{liu_tera_2021, luo_dropout_2021}.
 \edit{Both methods are variations on the original dropout method \cite{srivastava_dropout_2014}.}
@@
 Other than the autoregressive and masked reconstruction tasks discussed above, various studies have explored the reconstruction of other targets derived from the input.
-PASE and PASE+~\cite{pascual_learning_2019,ravanelli_multi_2020} use multiple targets, including the waveform, log power spectrum, \edit{mel cepstral coefficients (MFCCs)}, and prosody features.
-Models that learn acoustic embeddings of small speech segments have targeted future and past spectrogram segments~\cite{chung_speech2vec_2018, tagliasacchi_self_2019, tagliasacchi_pre_2020}, phase information~\cite{quitry2019learning}, and the temporal gap between two segments~\cite{tagliasacchi_self_2019, tagliasacchi_pre_2020}.
+PASE and PASE+~\cite{pascual_learning_2019,ravanelli_multi-task_2020} use multiple targets, including the waveform, log power spectrum, \edit{mel cepstral coefficients (MFCCs)}, and prosody features.
+Models that learn acoustic embeddings of small speech segments have targeted future and past spectrogram segments~\cite{chung_speech2vec_2018, tagliasacchi_selfsupervised_2019, tagliasacchi_pretraining_2020}, phase information~\cite{quitry_learning_2019}, and the temporal gap between two segments~\cite{tagliasacchi_selfsupervised_2019, tagliasacchi_pretraining_2020}.
 
@@
 
-Typical approaches involve learning a neural representation model for each modality, with a multi-modal contrastive loss that encourages paired examples in the two modalities to have similar representations while unpaired examples remain different~\cite{synnaeve_learning_2014,harwath_unsupervised_2016,harwath_deep_2015,merkx_language_2019,rouditchenko_avlnet_2021,peng_fastslow_2022}.
+Typical approaches involve learning a neural representation model for each modality, with a multi-modal contrastive loss that encourages paired examples in the two modalities to have similar representations while unpaired examples remain different~\cite{synnaeve_learning_2014,harwath_unsupervised_2016,harwath_deep_2015,merkx_language_2019,rouditchenko_avlnet_2021,peng_fastslow_2022}
+
 Other choices include training with a masked margin softmax loss~\cite{ilharco_largescale_2019,sanabria_talk_2021} or a masked prediction loss~\cite{chan_multimodal_2022}.
@@
 These are often evaluated on word discrimination, that is the task of determining whether two word segments correspond to the same word or not~\cite{carlin_rapid_2011}.
-This task can be thought of as a proxy for query-by-example search, since the basic operation in search is to determine whether a segment in the search database matches a query segment, and has been used for evaluation of both frame-level (e.g.,~\cite{kamper_unsupervised_2015}) and word-level~\cite{levin_fixed-dimensional_2013,kamper2016deep} representations.
+This task can be thought of as a proxy for query-by-example search, since the basic operation in search is to determine whether a segment in the search database matches a query segment, and has been used for evaluation of both frame-level (e.g.,~\cite{kamper_unsupervised_2015}) and word-level~\cite{levin_fixed-dimensional_2013,kamper_deep_2016} representations.
 

### 2023-03-17 chapters/paper_review/04_benchmarking.tex
Change reference keys (WIP)

@@
 AVSpeech has been used to learn audio-only 
-representations~\cite{kawakami2020learning}. The Fisher corpus~\cite{cieri2004fisher} collects
+representations~\cite{kawakami_learning_2020}. The Fisher corpus~\cite{cieri2004fisher} collects
 over 2k hours of conversational telephone speech, 1k hours of which is utilized
@@
 hours/60 languages in \cite{babu2021xlsr} and 430 hours/29 languages in
-\cite{kawakami2020learning}) or evaluation. 
+\cite{kawakami_learning_2020}) or evaluation. 
 %Demographic metadata like age, sex, and accent is also provided.
@@
 also aggregated to build diversified and large-scale corpora for unsupervised
-pre-training~\cite{kawakami2020learning, song20d_interspeech, babu2021xlsr}.
+pre-training~\cite{kawakami_learning_2020, song_speech-xlnet_2020, babu2021xlsr}.
 GigaSpeech is a multi-domain English ASR corpus with 33k hours of audio
@@
 the development of speech technologies in Africa, and
-\cite{kawakami2020learning} leverages four African languages collected in the
+\cite{kawakami_learning_2020} leverages four African languages collected in the
 project for evaluation: Amharic~\cite{tachbelie2014using}, Fongbe~\cite{laleye2016first},
 Swahili~\cite{gelas2012developments}, and Wolof~\cite{gauthier2016collecting}.
-In the same work~\cite{kawakami2020learning}, the authors
+In the same work~\cite{kawakami_learning_2020}, the authors
 further select 21 phonetically diverse languages from OpenSLR to evaluate the
@@
 
-Last, \cite{tagliasacchi_pre_2020} puts together five datasets 
+Last, \cite{tagliasacchi_pretraining_2020} puts together five datasets 
 (MUSAN~\cite{snyder2015musan}, Bird Audio Detection~\cite{stowell2019automatic},
@@
 capability of representations on audio event detection.
-\cite{quitry2019learning} employs the NSynth dataset~\cite{engel2017neural} on top
+\cite{quitry_learning_2019} employs the NSynth dataset~\cite{engel2017neural} on top
 of the six for benchmarking. As many of the datasets are built for research in
@@
 
-    Common Voice (CV-dataset) & PT/EV & Multi & 11k/7k \cite{babu2021xlsr}/430 \cite{kawakami2020learning} & ASR/PR/PC & CC0 \\ \hline % mention the split, CPC-modified, Bidir-CPC
+    Common Voice (CV-dataset) & PT/EV & Multi & 11k/7k \cite{babu2021xlsr}/430 \cite{kawakami_learning_2020} & ASR/PR/PC & CC0 \\ \hline % mention the split, CPC-modified, Bidir-CPC
     Multilingual LS (MLS) & PT/EV & Multi & 50k & ASR & CC BY 4.0 \\ \hline
@@
     % Audio2Vec
-    Audio2Vec \cite{tagliasacchi_pre_2020} & AudioSet & AED & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & \checkmark & See \cite{tagliasacchi_pre_2020} for details \\ \hline
+    Audio2Vec \cite{tagliasacchi_pretraining_2020} & AudioSet & AED & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & \checkmark & See \cite{tagliasacchi_pretraining_2020} for details \\ \hline
     % APC
@@
     % Phase predict
-    PhasePredict \cite{quitry2019learning} & AudioSet & AED & 7 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & 7 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & \checkmark & See \cite{quitry2019learning} for details \\ \hline
+    PhasePredict \cite{quitry_learning_2019} & AudioSet & AED & 7 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & 7 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & \checkmark & See \cite{quitry_learning_2019} for details \\ \hline
     % Bidir-CPC
-    \multirow{3}{*}{Bidir-CPC\cite{kawakami2020learning}} & \multirow{3}{*}{\makecell[l]{LS 960 hrs,\\ CPC-8k\textsuperscript{\ref{footnote:CPC-8k}}}} & ASR & \makecell[l]{WSJ si284,\\LS 960 hrs,\\TED3\textsuperscript{\ref{footnote:official-split}}} & \makecell[l]{WSJ eval92,\\LS test-clean,\\LS test-other,\\TED3\textsuperscript{\ref{footnote:official-split}}, SWB\textsuperscript{\ref{footnote:official-split}}}  & \makecell[l]{Different\\ datasets for\\ training\\ and test} & 81/960/450 hrs \\ \cline{3-7} % cross domain transferring, SWB eval2000, TED3 training set size = 450 hrs
+    \multirow{3}{*}{Bidir-CPC\cite{kawakami_learning_2020}} & \multirow{3}{*}{\makecell[l]{LS 960 hrs,\\ CPC-8k\textsuperscript{\ref{footnote:CPC-8k}}}} & ASR & \makecell[l]{WSJ si284,\\LS 960 hrs,\\TED3\textsuperscript{\ref{footnote:official-split}}} & \makecell[l]{WSJ eval92,\\LS test-clean,\\LS test-other,\\TED3\textsuperscript{\ref{footnote:official-split}}, SWB\textsuperscript{\ref{footnote:official-split}}}  & \makecell[l]{Different\\ datasets for\\ training\\ and test} & 81/960/450 hrs \\ \cline{3-7} % cross domain transferring, SWB eval2000, TED3 training set size = 450 hrs
     & & \makecell[l]{ASR-multi} & ALFFA\textsuperscript{\ref{footnote:official-split}} & ALFFA\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 languages, 5.2--18.3 hrs\\ \cline{3-7}
@@
     % CPC modified
-    \multirow{2}{*}{CPC modified \cite{riviere2020unsupervised}} & LS 100 hrs & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & \makecell[l]{LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}}} & - & \makecell[l]{80\textsuperscript{\ref{footnote:low-resource}} hrs} \\ \cline{2-7}
+    \multirow{2}{*}{CPC modified \cite{riviere_unsupervised_2020}} & LS 100 hrs & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & \makecell[l]{LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}}} & - & \makecell[l]{80\textsuperscript{\ref{footnote:low-resource}} hrs} \\ \cline{2-7}
     & \makecell[l]{LS 100 hrs,\\LS 960 hrs,\\LL 60k hrs} & PC & CV-dataset\textsuperscript{\ref{footnote:official-split}} & CV-dataset\textsuperscript{\ref{footnote:official-split}} & \checkmark & 1 hrs \\ \hline
@@
     % PASE+
-    \multirow{3}{*}[0mm]{PASE+\cite{ravanelli_multi_2020}} & \multirow{3}{*}[0mm]{LS 50 hrs \cite{ravanelli2018learning}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
+    \multirow{3}{*}[0mm]{PASE+\cite{ravanelli_multi-task_2020}} & \multirow{3}{*}[0mm]{LS 50 hrs \cite{ravanelli2018learning}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
     & & ASR & DIRHA\textsuperscript{\ref{footnote:official-split}} & DIRHA\textsuperscript{\ref{footnote:official-split}} & \checkmark & 11 hrs \\ \cline{3-7}
@@
     % AALBERT
-    \multirow{2}{*}[0mm]{AALBERT \cite{chi2020audio}} & \multirow{2}{*}[0mm]{\makecell[l]{LS 360 hrs}} & PC & LS 100 hrs\textsuperscript{\ref{footnote:APC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:APC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{AALBERT \cite{chi_audio_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{LS 360 hrs}} & PC & LS 100 hrs\textsuperscript{\ref{footnote:APC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:APC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
     & & SID & LS 360 hrs\textsuperscript{\ref{footnote:APC-split}} & LS 360 hrs\textsuperscript{\ref{footnote:APC-split}} & - & 288\textsuperscript{\ref{footnote:low-resource}} hrs \\
@@
     % speech-XLNet
-    \multirow{2}{*}[-1mm]{speech-XLNet \cite{song20d_interspeech}} & \multirow{2}{*}[-1mm]{\makecell[l]{LS 960 hrs\\\;\;\;+ WSJ si284\\\;\;\;+ TED2}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs\vspace{1mm} \\ \cline{3-7}
+    \multirow{2}{*}[-1mm]{speech-XLNet \cite{song_speech-xlnet_2020}} & \multirow{2}{*}[-1mm]{\makecell[l]{LS 960 hrs\\\;\;\;+ WSJ si284\\\;\;\;+ TED2}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs\vspace{1mm} \\ \cline{3-7}
     & & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{7\textsuperscript{\ref{footnote:low-resource}}/14\textsuperscript{\ref{footnote:low-resource}}/30\textsuperscript{\ref{footnote:low-resource}}/81 hrs}\vspace{1mm} \\ \hline
@@
     % MPE
-    \multirow{2}{*}[0mm]{MPE \cite{liu2020masked}} & \multirow{2}{*}[0mm]{\makecell[l]{WSJ si284,\\LS 960 hrs}} & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{25\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}}/81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{MPE \cite{liu_masked_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{WSJ si284,\\LS 960 hrs}} & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{25\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}}/81 hrs} \\ \cline{3-7}
     & & ASR & LS 100/360/960 hrs & LS test-clean & - & 100/360/960 hrs \\ \hline
@@
     % wav2vec 2.0
-    \multirow{2}{*}{wav2vec 2.0 \cite{baevski2020wav2vec}} & \multirow{2}{*}{\makecell[l]{LS 960 hrs,\\ LL 60k hrs}} & ASR & LS 960 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \cline{3-7}
+    \multirow{2}{*}{wav2vec 2.0 \cite{baevski_wav2vec_2020}} & \multirow{2}{*}{\makecell[l]{LS 960 hrs,\\ LL 60k hrs}} & ASR & LS 960 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \cline{3-7}
     & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \hline
     % NPC
-    \multirow{2}{*}[0mm]{NPC \cite{liu21l_interspeech}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{NPC \cite{liu_non-autoregressive_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
     & & SID & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & \checkmark & 65\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline
     % DeCoAR 2.0
-    DeCoAR 2.0 \cite{ling2020decoar} & LS 960 hrs & ASR & \makecell[l]{LS 100 hrs} & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline
+    DeCoAR 2.0 \cite{ling_decoar_2020} & LS 960 hrs & ASR & \makecell[l]{LS 100 hrs} & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline
 
@@
 The 6 AED datasets used in
-\cite{tagliasacchi_pre_2020} are MUSAN~\cite{snyder2015musan}, Bird Audio
+\cite{tagliasacchi_pretraining_2020} are MUSAN~\cite{snyder2015musan}, Bird Audio
 Detection~\cite{stowell2019automatic}, Speech Commands~\cite{warden2018speech},
@@
 \textit{train-clean-100}. In addition to the 6 datasets,
-\cite{quitry2019learning} use the NSynth dataset~\cite{engel2017neural} for
+\cite{quitry_learning_2019} use the NSynth dataset~\cite{engel2017neural} for
 evaluation.}
@@
 English speech with various accents. Among the papers studied in this review, SSA
-is used in \cite{kawakami2020learning} only for pre-training, and only 1 hour
+is used in \cite{kawakami_learning_2020} only for pre-training, and only 1 hour
 of audio is utilized. Thus, we exclude it from our discussion in
@@
 growing industry investment in pre-training with larger datasets, e.g.,
-CPC-8k (8k hours) for Bidir-CPC~\cite{kawakami2020learning}, LL (60k hours) for
-CPC modified~\cite{riviere2020unsupervised}, 
-wav2vec 2.0~\cite{baevski2020wav2vec}, and HuBERT~\cite{hsu_hubert_2021}, Alexa internal
+CPC-8k (8k hours) for Bidir-CPC~\cite{kawakami_learning_2020}, LL (60k hours) for
+CPC modified~\cite{riviere_unsupervised_2020}, 
+wav2vec 2.0~\cite{baevski_wav2vec_2020}, and HuBERT~\cite{hsu_hubert_2021}, Alexa internal
 datasets (10k hours) for wav2vec-c~\cite{sadhu_wav2vec-c_2021}, Didi internal
@@
 for English, whereas Chinese~\cite{jiang2019improving, jiang_further_2021} and
-multilinguality~\cite{kawakami2020learning, babu2021xlsr} are also gaining
+multilinguality~\cite{kawakami_learning_2020, babu2021xlsr} are also gaining
 attention. Compared to pre-training, datasets used for fine-tuning are more

### 2023-03-17 chapters/paper_review/04b_analysis.tex
Change reference keys (WIP)

@@
 Several studies have demonstrated better downstream performance when using
-larger datasets~\cite{rivi20, kawakami2020learning, chen_wavlm_2021}. For
+larger datasets~\cite{rivi20, kawakami_learning_2020, chen_wavlm_2021}. For
 example,
-\cite{kawakami2020learning} compared representations learned by a bidirectional
+\cite{kawakami_learning_2020} compared representations learned by a bidirectional
 CPC model from the standard 960 hour LS corpus and a corpus of 8,000 hours of
@@
 shifts? 
-One study~\cite{kawakami2020learning} compared pre-trained
+One study~\cite{kawakami_learning_2020} compared pre-trained
 representations from a CPC model against MFCCs and 

### 2023-03-17 chapters/paper_review/05_zero.tex
Change reference keys (WIP)

@@
 
-\cite{baevski2021unsupervised} & Wav2vec 2.0~\cite{baevski2020wav2vec} & \kmeans{} & Phoneme &\edit{Adversarial Training~\cite{gulrajani2017improved}} & Self-training & \tabincell{l}{TIMIT (PER):  18.6\%, \\ LibriSpeech (WER): 5.9\%}   \\
+\cite{baevski2021unsupervised} & Wav2vec 2.0~\cite{baevski_wav2vec_2020} & \kmeans{} & Phoneme &\edit{Adversarial Training~\cite{gulrajani2017improved}} & Self-training & \tabincell{l}{TIMIT (PER):  18.6\%, \\ LibriSpeech (WER): 5.9\%}   \\
   \midrule
@@
   \midrule
-\textcolor{black}{\cite{w2v-u2}} & \textcolor{black}{Wav2vec 2.0~\cite{baevski2020wav2vec} }& - & \textcolor{black}{Phoneme} & \textcolor{black}{Adversarial Training~\cite{gulrajani2017improved}} & \textcolor{black}{Self-training} & \textcolor{black}{LibriSpeech (WER): 6.3\%}   \\
+\textcolor{black}{\cite{w2v-u2}} & \textcolor{black}{Wav2vec 2.0~\cite{baevski_wav2vec_2020} }& - & \textcolor{black}{Phoneme} & \textcolor{black}{Adversarial Training~\cite{gulrajani2017improved}} & \textcolor{black}{Self-training} & \textcolor{black}{LibriSpeech (WER): 6.3\%}   \\
   \midrule
 
-\textcolor{black}{\cite{w2v-u2}} & \textcolor{black}{Wav2vec 2.0~\cite{baevski2020wav2vec}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Adversarial Training~\cite{gulrajani2017improved}} & \textcolor{black}{Self-training} & \textcolor{black}{LJSpeech (WER): 64.0\%}  \\
+\textcolor{black}{\cite{w2v-u2}} & \textcolor{black}{Wav2vec 2.0~\cite{baevski_wav2vec_2020}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Adversarial Training~\cite{gulrajani2017improved}} & \textcolor{black}{Self-training} & \textcolor{black}{LJSpeech (WER): 64.0\%}  \\
 
@@
 or token to the correct phoneme when the representation or token is highly correlated to the phonemes.
-Wav2vec-U~\cite{baevski2021unsupervised} selects the input feature from different layers of wave2vec~2.0~\cite{baevski2020wav2vec}.
+Wav2vec-U~\cite{baevski2021unsupervised} selects the input feature from different layers of wave2vec~2.0~\cite{baevski_wav2vec_2020}.
 The selection criterion is based on analysis of the phonetic information in each layer.
@@
 Wav2vec-U~\cite{baevski2021unsupervised} selects the input feature from
-different layers of wave2vec 2.0~\cite{baevski2020wav2vec}, a self-supervised
+different layers of wave2vec 2.0~\cite{baevski_wav2vec_2020}, a self-supervised
 model.

### 2023-03-17 chapters/paper_review/main_hi.tex
Change reference keys (WIP)

@@
         \multicolumn{7}{c}{\textit{\textbf{Pre-Training}}} \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec} & LL-60k & Transformer & 1.6 & 3.0 & 1.8 & 3.3 \\
+        wav2vec 2.0 \textsc{Large}~\cite{baevski_wav2vec_2020} & LL-60k & Transformer & 1.6 & 3.0 & 1.8 & 3.3 \\
         pre-trained Conformer XXL~\cite{zhang2020pushing} & LL-60k & LSTM & 1.5 & 3.0 & 1.5 & 3.1 \\

### 2023-03-17 chapters/paper_review/main_lo.tex
Change reference keys (WIP)

--- 2023-03-04 chapters/paper_review/main_lo.tex
@@
         \multicolumn{7}{c}{\textit{\textbf{10-min labeled}}} \\
-        % DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & - & - \\
-        DiscreteBERT~\cite{baevski2019effectiveness} & LS-960 & 4-gram & 15.7 & 24.1 & 16.3 & 25.2 \\
-        wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 8.9 & 15.7 & 9.1 & 15.6 \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & 4-gram & 6.3 & 9.8 & 6.6 & 10.3  \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & Transformer & 4.6 & 7.9 & 4.8 & 8.2 \\
+        % DeCoAR 2.0~\cite{ling_decoar_2020} & LS-960 & 4-gram & - & - & - & - \\
+        DiscreteBERT~\cite{baevski_effectiveness_2020} & LS-960 & 4-gram & 15.7 & 24.1 & 16.3 & 25.2 \\
+        wav2vec 2.0 \textsc{Base}~\cite{baevski_wav2vec_2020}   & LS-960 & 4-gram & 8.9 & 15.7 & 9.1 & 15.6 \\
+        wav2vec 2.0 \textsc{Large}~\cite{baevski_wav2vec_2020}  & LL-60k & 4-gram & 6.3 & 9.8 & 6.6 & 10.3  \\
+        wav2vec 2.0 \textsc{Large}~\cite{baevski_wav2vec_2020}  & LL-60k & Transformer & 4.6 & 7.9 & 4.8 & 8.2 \\
         \midrule
@@
         \multicolumn{7}{c}{\textit{\textbf{1-hour labeled}}} \\
-        DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & 13.8 & 29.1 \\
-        DiscreteBERT~\cite{baevski2019effectiveness} & LS-960 & 4-gram & 8.5 & 16.4 & 9.0 & 17.6 \\
-        wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 5.0 & 10.8 & 5.5 & 11.3 \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & Transformer & 2.9 & 5.4 & 2.9 & 5.8 \\
+        DeCoAR 2.0~\cite{ling_decoar_2020} & LS-960 & 4-gram & - & - & 13.8 & 29.1 \\
+        DiscreteBERT~\cite{baevski_effectiveness_2020} & LS-960 & 4-gram & 8.5 & 16.4 & 9.0 & 17.6 \\
+        wav2vec 2.0 \textsc{Base}~\cite{baevski_wav2vec_2020}   & LS-960 & 4-gram & 5.0 & 10.8 & 5.5 & 11.3 \\
+        wav2vec 2.0 \textsc{Large}~\cite{baevski_wav2vec_2020}  & LL-60k & Transformer & 2.9 & 5.4 & 2.9 & 5.8 \\
         \midrule
@@
         SlimIPL~\cite{likhomanenko2020slimipl} & LS-960 & 4-gram + Transformer & 5.3 & 7.9 & 5.5 & 9.0 \\
-        DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & 5.4 & 13.3 \\
-        DiscreteBERT~\cite{baevski2019effectiveness} & LS-960 & 4-gram & 5.3 & 13.2 & 5.9 & 14.1 \\
-        wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 3.8 & 9.1 & 4.3 & 9.5 \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & Transformer & 2.4 & 4.8 & 2.6 & 4.9 \\
+        DeCoAR 2.0~\cite{ling_decoar_2020} & LS-960 & 4-gram & - & - & 5.4 & 13.3 \\
+        DiscreteBERT~\cite{baevski_effectiveness_2020} & LS-960 & 4-gram & 5.3 & 13.2 & 5.9 & 14.1 \\
+        wav2vec 2.0 \textsc{Base}~\cite{baevski_wav2vec_2020}   & LS-960 & 4-gram & 3.8 & 9.1 & 4.3 & 9.5 \\
+        wav2vec 2.0 \textsc{Large}~\cite{baevski_wav2vec_2020}  & LL-60k & Transformer & 2.4 & 4.8 & 2.6 & 4.9 \\
         \midrule
@@
         % LPM~\cite{hsu2020semi} & LL-60K & ConvLM & 4.87 & 13.84 & 4.88 & 15.28 \\
-        IPL~\cite{xu2020iterative} & LL-60k & 4-gram + Transformer & 3.19 & 6.14 & 3.72 & 7.11 \\
+        IPL~\cite{xu_iterative_2020} & LL-60k & 4-gram + Transformer & 3.19 & 6.14 & 3.72 & 7.11 \\
         SlimIPL~\cite{likhomanenko2020slimipl} & LS-860 & 4-gram + Transformer & 2.2 & 4.6 & 2.7 & 5.2 \\
-        Noisy Student\cite{park2020improved} & LS-860 & LSTM & 3.9 & 8.8 & 4.2 & 8.6 \\
-        DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & 5.0 & 12.1 \\
-        DiscreteBERT~\cite{baevski2019effectiveness} & LS-960 & 4-gram & 4.0 & 10.9 & 4.5 & 12.1 \\
-        wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 2.7 & 7.9 & 3.4 & 8.0 \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & Transformer & 1.9 & 4.0 & 2.0 & 4.0 \\
+        Noisy Student\cite{park_improved_2020} & LS-860 & LSTM & 3.9 & 8.8 & 4.2 & 8.6 \\
+        DeCoAR 2.0~\cite{ling_decoar_2020} & LS-960 & 4-gram & - & - & 5.0 & 12.1 \\
+        DiscreteBERT~\cite{baevski_effectiveness_2020} & LS-960 & 4-gram & 4.0 & 10.9 & 4.5 & 12.1 \\
+        wav2vec 2.0 \textsc{Base}~\cite{baevski_wav2vec_2020}   & LS-960 & 4-gram & 2.7 & 7.9 & 3.4 & 8.0 \\
+        wav2vec 2.0 \textsc{Large}~\cite{baevski_wav2vec_2020}  & LL-60k & Transformer & 1.9 & 4.0 & 2.0 & 4.0 \\
 

### 2023-03-17 chapters/paper_review/model.tex
Change reference keys (WIP)

@@
 Speech2Vec \cite{chung_speech2vec_2018}     & 2018 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
-Unspeech \cite{milde2018unspeech}         & 2018 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
+Unspeech \cite{milde_unspeech_2018}         & 2018 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
 
@@
 
-PASE+ \cite{ravanelli_multi_2020}                & 2020 Jan. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 25/1 % I think contrastive always have learned target.
+PASE+ \cite{ravanelli_multi-task_2020}                & 2020 Jan. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 25/1 % I think contrastive always have learned target.
 
-Bidir CPC \cite{kawakami2020learning} & 2020 Jan. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\% 29/1
+Bidir CPC \cite{kawakami_learning_2020} & 2020 Jan. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\% 29/1
 
-Modified CPC \cite{riviere2020unsupervised} & 2020 Feb. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 7/2
+Modified CPC \cite{riviere_unsupervised_2020} & 2020 Feb. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 7/2
 
@@
 
-AALBERT \cite{chi2020audio}  & 2019 May & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ %18 May 2020
+AALBERT \cite{chi_audio_2020}  & 2019 May & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ %18 May 2020
 
-speech-XLNet \cite{song20d_interspeech} & 2019 Oct.
+speech-XLNet \cite{song_speech-xlnet_2020} & 2019 Oct.
 & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\%23 Oct 2019 %Lee: It is speech version of XLNet.
@@
 
-wav2vec 2.0 \cite{baevski2020wav2vec}     & 2020 Jun. & \xmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ % 20/6
+wav2vec 2.0 \cite{baevski_wav2vec_2020}     & 2020 Jun. & \xmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ % 20/6
 %Lee: In my point of view, wav2vec 2.0 has learned target. The target is from the CNN encoder.
 
-MPE \cite{liu2020masked} & 2020 May & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 25 May 2020 %Lee: I think it has learned target. But please check.
+MPE \cite{liu_masked_2020} & 2020 May & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 25 May 2020 %Lee: I think it has learned target. But please check.
 
@@
 
-Speech SimCLR \cite{SpeechSimCLR} &  2020 Oct.
+Speech SimCLR \cite{jiang_speech_2021} &  2020 Oct.
 & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \xmark \\ %Lee: It is TERA+ SIMCLR.
@@
 
-NPC \cite{liu21l_interspeech}                     & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ % 1/11
+NPC \cite{liu_non-autoregressive_2020}                     & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ % 1/11
 
-DeCoAR 2.0 \cite{ling2020decoar}          & 2020 Dec.
+DeCoAR 2.0 \cite{ling_decoar_2020}          & 2020 Dec.
 & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ % 11/12 %Lee:  DeCoAR 2.0 has VQ layer although it is not used as learning target.
@@
 
-pMPC \cite{yue_pmpc_2021}  & 2021 Aug. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ %30 August – 3 September, 2021 (I do not find arXiv version.)
+pMPC \cite{yue_phonetically_2021}  & 2021 Aug. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ %30 August – 3 September, 2021 (I do not find arXiv version.)
 
@@
 Speech2Vec \cite{chung_speech2vec_2018}     & 2018 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
-% Unspeech \cite{milde2018unspeech}         & 2018 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
+% Unspeech \cite{milde_unspeech_2018}         & 2018 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
 CPC \cite{oord_representation_2018}         & 2018 Jul. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\
@@
 %PASE+ \cite{Ravanelli2020}                & 2020 Jan. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 25/1
-wav2vec 2.0 \cite{baevski2020wav2vec}     & 2020 Jun. & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark \\ % 20/6
+wav2vec 2.0 \cite{baevski_wav2vec_2020}     & 2020 Jun. & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark \\ % 20/6
 %TERA \cite{liu_tera_2021}                   & 2021 Jul. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 12/7
 
-NPC \cite{liu21l_interspeech}             & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 1/11
+NPC \cite{liu_non-autoregressive_2020}             & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 1/11
 
-DeCoAR 2.0 \cite{ling2020decoar}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 11/12
+DeCoAR 2.0 \cite{ling_decoar_2020}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 11/12
 HuBERT \cite{hsu_hubert_2021}               & 2021 Jun. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ % 14/6

### 2023-03-17 chapters/paper_review/pretext.tex
Change reference keys (WIP)

@@
 \centering
-\renewcommand*\arraystretch{1.2}{
+\resizebox{0.98\textwidth}{!}{
 \begin{tabular}{l|c|c|c}
@@
     \midrule  
-    Speech2Vec~\cite{chung_speech2vec_2018}, Audio2Vec~\cite{tagliasacchi_pre_2020} - skip-gram    & $X_{[t_1,t_2]}$  &     \textsc{-}   &    $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$     \\
+    Speech2Vec~\cite{chung_speech2vec_2018}, Audio2Vec~\cite{tagliasacchi_pretraining_2020} - skip-gram    & $X_{[t_1,t_2]}$  &     \textsc{-}   &    $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$     \\
     \midrule 
-    Speech2Vec~\cite{chung_speech2vec_2018}, Audio2Vec~\cite{tagliasacchi_pre_2020} - cbow    & $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$   &     \textsc{-}   &    $X_{[t_1,t_2]}$   \\
+    Speech2Vec~\cite{chung_speech2vec_2018}, Audio2Vec~\cite{tagliasacchi_pretraining_2020} - cbow    & $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$   &     \textsc{-}   &    $X_{[t_1,t_2]}$   \\
     \midrule 
-    PASE~\cite{pascual_learning_2019}, PASE+~\cite{ravanelli_multi_2020}\footnote{PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
+    PASE~\cite{pascual_learning_2019}, PASE+~\cite{ravanelli_multi-task_2020}\footnote{PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
     \midrule 
@@
     \midrule
-    Speech-XLNet \cite{song20d_interspeech}     & \multicolumn{2}{c|}{$X_{\mathcal{P}_{t}}$}   &     $x_{i\sim\mathcal{P}^c_{t}}$  \\ %Lee: Use * to represent the permutation, hope it is not strange.
+    Speech-XLNet \cite{song_speech-xlnet_2020}     & \multicolumn{2}{c|}{$X_{\mathcal{P}_{t}}$}   &     $x_{i\sim\mathcal{P}^c_{t}}$  \\ %Lee: Use * to represent the permutation, hope it is not strange.
     \midrule  
@@
     \midrule
-    Mockingjay~\cite{liu_mockingjay_2020}, Audio ALBERT~\cite{chi2020audio}, DeCoAR 2.0~\cite{ling2020decoar}   & \multicolumn{2}{c|}{$X_{-[t,t+k]}$}   & $X_{[t,t+k]}$    \\
+    Mockingjay~\cite{liu_mockingjay_2020}, Audio ALBERT~\cite{chi_audio_2020}, DeCoAR 2.0~\cite{ling_decoar_2020}   & \multicolumn{2}{c|}{$X_{-[t,t+k]}$}   & $X_{[t,t+k]}$    \\
     \midrule 
@@
     \midrule
-    pMPC~\cite{yue_pmpc_2021}      &  \multicolumn{2}{c|}{$X_{-[t,t+k^\prime]}$ ($X_{[t,t+k^\prime]}$ is a phoneme)}       & $X_{[t,t+k^\prime]}$    \\
+    pMPC~\cite{yue_phonetically_2021}      &  \multicolumn{2}{c|}{$X_{-[t,t+k^\prime]}$ ($X_{[t,t+k^\prime]}$ is a phoneme)}       & $X_{[t,t+k^\prime]}$    \\
     \midrule 
-    MPE~\cite{liu2020masked} & $X$ &  $Z_{-[t,t+k]}$  & $Z$    \\ %Lee: What is the difference between MPE and NPC??? And I believe it reconstruct the convolutional blocks output (learned target?)
+    MPE~\cite{liu_masked_2020} & $X$ &  $Z_{-[t,t+k]}$  & $Z$    \\ %Lee: What is the difference between MPE and NPC??? And I believe it reconstruct the convolutional blocks output (learned target?)
     \midrule
-    NPC~\cite{liu21l_interspeech}      & $X$  &   $Z_{-[t,t+k]}$  &    $X$   \\ %Lee: I am not 100% sure it is correct. Please check.
+    NPC~\cite{liu_non-autoregressive_2020}      & $X$  &   $Z_{-[t,t+k]}$  &    $X$   \\ %Lee: I am not 100% sure it is correct. Please check.
     \midrule
@@
     \midrule
-    Unspeech \cite{milde2018unspeech}       &   $X_{[t_1,t_2]}$ &   \textsc{-}   &  $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$ \\
+    Unspeech \cite{milde_unspeech_2018}       &   $X_{[t_1,t_2]}$ &   \textsc{-}   &  $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$ \\
     \midrule 
-    CPC~\cite{oord_representation_2018}, wav2vec \cite{schneider_wav2vec:_2019}, Modified CPC \cite{riviere2020unsupervised}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
+    CPC~\cite{oord_representation_2018}, wav2vec \cite{schneider_wav2vec:_2019}, Modified CPC \cite{riviere_unsupervised_2020}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
     \midrule 
-    Bidirectional CPC \cite{kawakami2020learning}      & $X_{[1,t]}$ or $X_{[t,T]}$ &  \textsc{-}    &    $z_{t+c}$ or $z_{t-c},\, c\geq1$   \\
+    Bidirectional CPC \cite{kawakami_learning_2020}      & $X_{[1,t]}$ or $X_{[t,T]}$ &  \textsc{-}    &    $z_{t+c}$ or $z_{t-c},\, c\geq1$   \\
     \midrule 
@@
     \midrule 
-    wav2vec 2.0 \cite{baevski2020wav2vec}, wav2vec-C \cite{sadhu_wav2vec-c_2021}\footnote{wav2vec-C adds reconstruction loss to wav2vec 2.0.}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
+    wav2vec 2.0 \cite{baevski_wav2vec_2020}, wav2vec-C \cite{sadhu_wav2vec-c_2021}\footnote{wav2vec-C adds reconstruction loss to wav2vec 2.0.}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
     \midrule 
@@
     \midrule
-    Speech SimCLR \cite{SpeechSimCLR}\footnote{Speech SimCLR targets the latent representation of an augmented version of $X$ using a differently augmented $X$, and vice-versa.}    & \multicolumn{2}{c|}{$\textsc{augment}_1(X)$ and $\textsc{augment}_2(X)$}     &    $\textsc{augment}_2(Z)$ and $\textsc{augment}_1(Z)$   \\
+    Speech SimCLR \cite{jiang_speech_2021}\footnote{Speech SimCLR targets the latent representation of an augmented version of $X$ using a differently augmented $X$, and vice-versa.}    & \multicolumn{2}{c|}{$\textsc{augment}_1(X)$ and $\textsc{augment}_2(X)$}     &    $\textsc{augment}_2(Z)$ and $\textsc{augment}_1(Z)$   \\
     \midrule
@@
 \end{table*}
-
-\begin{comment}
-\begin{table*}[ht!]
-    \centering
-    \caption{
-    This table summarizes the approaches of the three categories of self-supervised learning. 
-    Column (a) lists the names of the models and related references. 
-    Column (b) defines the input to the models. 
-    Column (c) defines any corruption of the input or some hidden representation. 
-    Column (d) defines the target of the pretext task; the pretext task itself is described by the overall model category and the main text.
-    $X=\{x_1,x_2,...,x_T\}$ is the input sequence in which $x_t$ can be an acoustic feature vector (e.g., MFCC, filterbank, or spectrogram features) or a waveform sample. 
-    $X_{[t_1:t_2]}$ represents $\{x_{t_1},x_{t_1+1},...,x_{t_2}\}$.
-    $X_{-[t_1:t_2]}$ represents $X$ with the segment $X_{[t_1:t_2]}=\{x_{t_1},x_{t_1+1},...,x_{t_2}\}$ masked.
-    $x_{t}^{i}$ represents the $i$-th dimension of $x_t$.
-    If $x_t$ is a frame in a spectrogram, then the $i$-th dimension corresponds to a specific frequency bin.
-    $X^{-[f,f+j]}$ refers to a spectrogram $X$ but masked along the frequency axis from the $f$-th to $f+j$-th bin. 
-    % $X_{-[t:t+k]}^{-[f,f+j]}$ refers to a spectrogram $X$ but masked along both time and frequency dimensions. 
-    % $X^*$ is a temporally permuted version of $X$, that is, the $x_t$ are randomly shuffled to form $X^*$. 
-    We indicate random temporal permutation of a sequence by indexing it with the set $\mathcal{P}_t\triangleq\textsc{permute}([0,t])$ where $\textsc{permute}(\cdot)$ returns a permutation of the given list.
-    We indicate data augmentation (e.g. reverberation) by the function $\textsc{augment}(\cdot)$. Subscripts indicate different augmentations. 
-    % $X^\prime$ is an augmented version of $X$ (e.g., $X$ with reverberation), while $X^{\prime\prime}$ is $X$ adding distortion different from  $X^\prime$.
-    $Z$ represents a localized latent representation sequence of $X$.
-%, and $Z^\prime$ and $Z^{\prime\prime}$ are the latent representation sequences of  $X^\prime$ and $X^{\prime\prime}$, respectively.
-    % $Z$ represents a localized latent representation sequence and $Z^\prime$ is the latent representation sequence of $X^\prime$. 
-    $Z^{(l)}$ is $Z$ at the $l$-th layer of the model used to compute it.
-    $\bar{H}$ is the contextualized sequence $H$ obtained from an exponential moving average (EMA) of the model undergoing training with no masking applied.
-    $Q$ represents a sequence of quantized learned representations, and $C$ is a sequence of discrete cluster IDs.
-    For contrastive models, we specify only positive targets.
-    }
-    \begin{minipage}{\textwidth}  
-    \centering
-    \renewcommand*\arraystretch{1.1}{
-    \begin{tabular}{l|c|c|c}
-    \toprule
-    \textbf{Model} (a) & \textbf{Input} (b) & \textbf{Corruption} (c) & \textbf{Target} (d) \\
-    \midrule
-    \midrule
-    \multicolumn{4}{c}{\textsc{Generative models}} \\
-    \midrule
-    \midrule
-    Audio Word2vec~\cite{chung_audio_2016}, VQ-VAE \cite{oord_neural_2018}     & $X$ &   \textsc{-}    &  $X$  \\ %It has VQ as extra constraint.
-    \midrule  
-    Speech2Vec~\cite{chung_speech2vec_2018}, Audio2Vec~\cite{tagliasacchi_pre_2020} - skip-gram    & $X_{[t_1,t_2]}$  &     \textsc{-}   &    $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$     \\
-    \midrule 
-   Speech2Vec~\cite{chung_speech2vec_2018}, Audio2Vec~\cite{tagliasacchi_pre_2020} - cbow    & $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$   &     \textsc{-}   &    $X_{[t_1,t_2]}$   \\
-    \midrule 
-     PASE~\cite{pascual_learning_2019}, PASE+~\cite{ravanelli_multi_2020}\footnote{PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
-    \midrule 
-    APC~\cite{chung_unsupervised_2019,chung_vector-quantized_2020}         & $X_{[1,t]}$   & \textsc{-}              & $x_{t+c},\, c\geq1$    \\ %VQ-APC is the same as APC
-    \midrule
-    Speech-XLNet \cite{song20d_interspeech}     & \multicolumn{2}{c|}{$X_{\mathcal{P}_{t}}$}   &     $x_{i\sim\mathcal{P}^c_{t}}$  \\ %Lee: Use * to represent the permutation, hope it is not strange.
-    \midrule  
-    DeCoAR~\cite{ling2020deep}     & $X_{[1,t-1]}, X_{[t+k+1,T]}$ & \textsc{-} & $X_{[t,t+k]}$   \\
-    \midrule
-    Mockingjay~\cite{liu_mockingjay_2020}, Audio ALBERT~\cite{chi2020audio}, DeCoAR 2.0~\cite{ling2020decoar}   & \multicolumn{2}{c|}{$X_{-[t,t+k]}$}   & $X_{[t,t+k]}$    \\
-    \midrule 
-    TERA~\cite{liu_tera_2021}, BMR~\cite{wang_unsupervised_2020}  & \multicolumn{2}{c|}{$X_{-[t,t+k]}^{-[f,f+j]}$}       & $X$       \\
-    \midrule
-    pMPC~\cite{yue_pmpc_2021}      &  \multicolumn{2}{c|}{$X_{-[t,t+k^\prime]}$ ($X_{[t,t+k^\prime]}$ is a phoneme)}       & $X_{[t,t+k^\prime]}$    \\
-    \midrule 
-    MPE~\cite{liu2020masked} & $X$ &  $Z_{-[t,t+k]}$  & $Z$    \\ %Lee: What is the difference between MPE and NPC??? And I believe it reconstruct the convolutional blocks output (learned target?)
-    \midrule
-    NPC~\cite{liu21l_interspeech}      & $X$  &   $Z_{-[t,t+k]}$  &    $X$   \\ %Lee: I am not 100% sure it is correct. Please check.
-    \midrule
-    \midrule
-    \multicolumn{4}{c}{\textsc{Contrastive models}} \\
-    \midrule
-    \midrule
-    Unspeech \cite{milde2018unspeech}       &   $X_{[t_1,t_2]}$ &   \textsc{-}   &  $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$ \\
-    \midrule 
-    CPC~\cite{oord_representation_2018}, wav2vec \cite{schneider_wav2vec:_2019}, Modified CPC \cite{riviere2020unsupervised}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
-        \midrule 
-    Bidirectional CPC \cite{kawakami2020learning}      & $X_{[1,t]}$ or $X_{[t,T]}$ &  \textsc{-}    &    $z_{t+c}$ or $z_{t-c},\, c\geq1$   \\
-    \midrule 
-    vq-wav2vec \cite{baevski_vq-wav2vec_2020}     &   $X_{[1,t]}$ &   \textsc{-}    &   $q_{t+c},\, c\geq1$   \\ 
-    \midrule 
-    wav2vec 2.0 \cite{baevski2020wav2vec}, wav2vec-C \cite{sadhu_wav2vec-c_2021}\footnote{wav2vec-C adds the reconstruction loss to wav2vec 2.0.}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
-    \midrule 
-    w2v-BERT \cite{chung_w2vbert_2021}     &$X$ &    $Z_{-[t,t+k]}$   &     $Q_{[t,t+k]}$ and $C_{[t,t+k]}$     \\
-    \midrule
-    Speech SimCLR \cite{SpeechSimCLR}\footnote{Speech SimCLR targets the latent representation of an augmented version of $X$ using a differently augmented $X$, and vice-versa.}    & \multicolumn{2}{c|}{$\textsc{augment}_1(X)$ and $\textsc{augment}_2(X)$}     &    $\textsc{augment}_2(Z)$ and $\textsc{augment}_1(Z)$   \\
-   \midrule
-   \midrule 
-    \multicolumn{4}{c}{\textsc{Predictive models}} \\
-    \midrule
-    \midrule
-    Discrete BERT~\cite{baevski_vq-wav2vec_2020,baevski_effectiveness_2020} \footnote{Discrete BERT obtains codes $C$ from vq-wav2vec.}      &   \multicolumn{2}{c|}{$C_{-[t,t+k]}$}   & $C_{[t,t+k]}$  \\
-    \midrule 
-    HuBERT \cite{hsu_hubert_2021}\footnote{HuBERT is trained first using cluster IDs of MFCCs as target and subsequently cluster IDs of model representations from the last iteration.}, WavLM \cite{chen_wavlm_2021}\footnote{WavLM simulates noisy/overlapped speech as inputs.}  & $X$             & $Z_{-[t,t+k]}$          & $C_{[t,t+k]}$  \\
-    \midrule
-    data2vec \cite{baevski_data2vec_2022}    & $X$             & $Z_{-[t,t+k]}$          & $\sum_{l}\bar{H}^{(l)}_{[t,t+k]}$  \\ 
-        \midrule 
-BEST-RQ \cite{BEST-RQ}\footnote{BEST-RQ obtains codes $C$ by quantizing acoustic features using a random projection quantizer.}     &  \multicolumn{2}{c|}{$X_{-[t,t+k]}$}      &  $C_{[t,t+k]}$   \\ 
-        \midrule 
-    \bottomrule
-    \end{tabular}
-    }
-
-    \end{minipage}
-    \label{table:pretext}
-\end{table*}
-\end{comment}
-
-\begin{comment}
-\begin{table*}[h!]
-\centering
-
-\setlength{\tabcolsep}{3pt}
-\caption{
-This table summarizes the generative approaches. 
-Column (a) is the name of the models (if any) and their reference.
-Column (b) is the corrupted speech signals for encoder input.
-Column (c) is the target output of decoder.
-Column (d) is regularization methods used if any.
-Column (e) are the models using the same pretext in NLP or CV.
-$S$ is a sequence of waveform samples, and $s_t$ is a single waveform sample.
-$S^*$ is the degradation of $S$ with reverberation or additive noises. 
-$S$ can be represented by a sequence of acoustic feature, $X=\{x_1,x_2,...,x_T\}$, in which $x_t$ is an acoustic feature vector like MFCC, fbank, spectrogram, etc. 
-$X^*$ is the permuted version of $X$. That is, the frames in $X$ are randomly shuffled to form $X^*$. 
-$X_{[t_1:t_2]}$ represents $\{x_{t_1},x_{t_1+1},...,x_{t_2}\}$.
-$X_{-[t_1:t_2]}$ represents the segment $\{x_{t_1},x_{t_1+1},...,x_{t_2}\}$ in $X$ is masked, that is, the frames in the region is replaced by zero vectors or random sampled vectors.
-$x_{t}^{i}$ represents the $i$-th dimension of frame $x_t$.
-If $x_t$ is a frame in a spectrogram, then the $i$-th dimension corresponds to a specific frequency bin.
-$X^{-[f_1,f_2]}$ means that for all the frames $x$ in $X$, their $f_1$-th to $f_2$-th dimensions are all masked. 
-$X_{-[t_1:t_2]}^{-[f_1,f_2]}$ means $X$ are masked in two directions. 
-}
-\label{table:generative}
-  {\renewcommand*\arraystretch{1.2}
-\begin{tabular}{|l|l|l|l|l|}
-\hline
-(a) Model & (b) Corrupted Input & (c) Target Output & (d) Regularization &   (e) Counterparts \\ 
-\hline \hline
-
-APC~\cite{chung_unsupervised_2019, chung_generative_2020,chung2020improved} & $X_{[1,t-1]}$  & $x_{t+c},c\geq0$ & VQ~\cite{chung_vector-quantized_2020} &  GPT~\cite{alex2018GPT,radford_language_2019,brown2020gpt3} \\ \hline
-
-%$X_{[1,t-1]}$  & $x_{t+c},c\geq0$ & VQ & VQ-APC~\cite{chung_vector-quantized_2020} & -  \\
-
-DeCoAR~\cite{ling2020deep} & $X_{[1,t-1]}$, $X_{[t+k+1,T]}$ &  $X_{[t,t+k]}$ & VQ~\cite{ling2020deep} &  ELMo~\cite{Matthew2018ELMO} \\
-\hline
-
-\tabincell{l}{ Mockingjay~\cite{liu_mockingjay_2020} \\ MPC~\cite{jiang2019improving,jiang_further_2021} \\ AALBERT~\cite{chi2021aalbert} \\
-DeCoAR 2.0~\cite{ling2020decoar} } &
-$X_{-[t,t+k]}$ & $X_{[t,t+k]}$ & VQ~\cite{ling2020decoar} &  \tabincell{l}{BERT ~\cite{devlin_bert:_2018}\\RoBERTa~\cite{liu_roberta_2019}} \\ \hline
-
-pMPC~\cite{yue_pmpc_2021}   &
-\tabincell{l}{$X_{-[t,t+k^\prime]}$\\($X_{[t,t+k^\prime]}$: phoneme-level segment)} & $X_{[t,t+k^\prime]}$ & - &  SpanBert~\cite{joshi-etal-2020-spanbert} \\ \hline
-
-speech-XLNet~\cite{song20d_interspeech} &
-$X^*_{[1:t-1]}$ & $x^*_t$ & - &  XLNet~\cite{Yang2019LNet}   \\ \hline
-
-\tabincell{l}{NPC~\cite{liu21l_interspeech}\\MPE~\cite{liu2020masked}} &
-\tabincell{l}{$X_{-[t,t+k]}$\\(mask on hidden layer)} & $X_{[t,t+k]}$ & - &   - \\ \hline
-
-\tabincell{l}{BMR~\cite{wang_unsupervised_2020}\\TERA~\cite{liu_tera_2021}} &
-$X_{-[t_1:t_2]}^{-[f_1,f_2]}$ & $X$ & dropout~\cite{luo_drop_2021} &  -  \\ \hline
-
-\cite{chorowski_unsupervised_2019} &
-$S$ & $S$ & \tabincell{l}{VQ-VAE~\cite{chorowski_unsupervised_2019}\\time-jitter regularizer~\cite{chorowski_unsupervised_2019}} & 
-\tabincell{l}{\cite{van2017neural,razavi_generating_2019} \\ (just name a few)} \\ \hline
-
-\cite{tagliasacchi_self_2019, tagliasacchi_pre_2020} & 
-$X_{[t-k,t-1]},X_{[t+k+1,t+2k]}$ & $X_{[t,t+k]}$  & - & CBoW~\cite{mikolov2013efficient}  \\ \hline
-
-\cite{tagliasacchi_self_2019, tagliasacchi_pre_2020} &
-$X_{[t,t+k]}$ & $X_{[t-k,t-1]},X_{[t+k+1,t+2k]}$ & - &  Skipgram~\cite{mikolov2013efficient} \\ \hline
-
-\cite{quitry2019learning}  &
-Magnitude of $X$ & Phase of $X$ & - &  - \\ \hline
-
-TemporalGap~\cite{tagliasacchi_self_2019, tagliasacchi_pre_2020} &
-$X_{[t_1,t_1+k]}$, $X_{[t_2,t_2+k]}$ & $|t_1 - t_2|$ & - &  -  \\ \hline
-
-\cite{Carr2021Ranking} &
-$X^*$ & correct order & - &  Jigsaw~\cite{noroozi2016unsupervised} \\ 
-\hline
-
-PASE~\cite{pascual_learning_2019} &
-$S$ & $X$: LPS, MFCC, Prosody & - &  -  \\ \hline
-
-PASE+~\cite{ravanelli_multi_2020} & 
-$S^*$ & \tabincell{l}{$X$: LPS, MFCC, Prosody, \\ fbank, gammatone} & -  &  -  \\ \hline
-\end{tabular}
-}
-\end{table*}
-\end{comment}

### 2023-03-17 chapters/paper_benchmarking.tex
Add review paper (WIP)
Change citation keys (WIP)

@@
 \begin{itemize}[leftmargin=8mm]
-    \vspace{-3mm}
     \setlength\itemsep{-0em}
@@
     \item[(IV)] LVMs learn rich representations that are as good or better than Mel spectrograms for phoneme recognition also when using only 10 minutes of labeled data.
-\end{itemize}\vspace{-3mm}
+\end{itemize}
 At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models
@@
 \subsection{Sequential deep latent variable models}
-The selected models are all sequential deep latent variable models trained with variational inference and the reparameterization trick \cite{kingma_auto-encoding_2014}.
+The selected models are all sequential deep latent variable models trained with variational inference and the reparameterization trick \cite{kingma_autoencoding_2014}.
 % The VRNN, SRNN, STCN and CW-VAE are all autoencoders trained with variational inference and the reparameterization trick \cite{kingma_auto-encoding_2014}. 
@@
 \label{fig: clockwork vae inference generative models 2 layered}
-\vspace{-0mm}
 \end{figure*}
@@
 
-\section{Speech modeling benchmark}\label{sec:benchmark}
+\section{Speech modeling benchmark}\label{sec:glvm_benchmark}
 \paragraph{Data}

### 2023-03-17 chapters/paper_review/04_benchmarking.tex
Change citation keys (WIP)

@@
 \label{section:benchmark}
-% {\color{blue} Daniel}\\
 
-% {\color{blue} Reviewer: SW, Hung-yi Lee, Abdo }\\
-
-%\subsection{Benchmarks for self-supervised learning approaches}
-
-%This paper is about the size of self-supervised model. I think it can also be considered as benchmark.
-%https://www.amazon.science/publications/scaling-effect-of-self-supervised-speech-models
-
-%We have shown in the previous sections various methodologies to learn speech representations from unlabeled corpora.
-This section surveys the available datasets to learn the representations and evaluate the learned ones.
-We also summarize several works and their results to demonstrate how effective the learned representations are via providing robust model initialization for various downstream tasks.
 The previous sections presented various methodologies by which to learn speech
@@
 in the literature. These datasets are usually large but with limited or no
-labels. Libri-light (LL)~\cite{kahn2020libri}, one of these datasets, is
+labels. Libri-light (LL)~\cite{kahn_libri-light_2020}, one of these datasets, is
 derived from audiobooks that are part of the
@@
 60k hours of spoken English audio tagged with SNR, speaker ID, and genre
-descriptions. The speech examples in Audioset~\cite{gemmeke2017audio}, which
+descriptions. The speech examples in Audioset~\cite{gemmeke_audio_2017}, which
 consists of over 2M 10-second YouTube video clips human-annotated with 632
@@
 of audio of varying quality, different languages, and sometimes multiple sound
-sources. AVSpeech~\cite{ephrat2018looking} is another large-scale audio-visual
+sources. AVSpeech~\cite{ephrat_looking_2018} is another large-scale audio-visual
 dataset used in SSL research, comprising 4.7k hours of clips from a wide
 variety of languages. 
-%from 150k distinct speakers
 Each clip contains a visible face and audible sound originating from a single
@@
 AVSpeech has been used to learn audio-only 
-representations~\cite{kawakami_learning_2020}. The Fisher corpus~\cite{cieri2004fisher} collects
+representations~\cite{kawakami_learning_2020}. The Fisher corpus~\cite{cieri_fisher_2004} collects
 over 2k hours of conversational telephone speech, 1k hours of which is utilized
@@
 collected corpora for pre-training Chinese speech representations. Didi Dictation
-and Didi Callcenter~\cite{jiang2019improving, jiang_further_2021} are two
+and Didi Callcenter~\cite{jiang_improving_2019, jiang_further_2021} are two
 internal datasets containing respectively 10k hours of read speech collected
@@
 \cref{table:datasets}. One of the most commonly used datasets in this category
-is LibriSpeech (LS)~\cite{panayotov2015librispeech}, a labeled corpus
+is LibriSpeech (LS)~\cite{panayotov_librispeech_2015}, a labeled corpus
 containing 960 hours of
-%16 kHz sampled 
 read English speech, which is also derived from an open-source audiobooks
@@
 can also be utilized to evaluate the performance of representation on ASR, phoneme recognition (PR),
-phoneme classification (PC), and speaker identification (SID) tasks. Wall Street Journal (WSJ)~\cite{paul1992design} is another
+phoneme classification (PC), and speaker identification (SID) tasks. Wall Street Journal (WSJ)~\cite{paul_design_1992} is another
 widely adopted, labeled corpus for pre-training. Its labels can evaluate
@@
 evaluation (PC and PR can be enabled via phone-level forced alignment). These
-corpora include Common Voice (CV)~\cite{ardila2020common}, Multilingual
-LibriSpeech (MLS)~\cite{pratap20_interspeech}, VoxPopuli 
-(VP)~\cite{wang-etal-2021-voxpopuli}, and BABEL (BBL)~\cite{gales2014speech}. CV is
+corpora include Common Voice (CV)~\cite{ardila_common_2020}, Multilingual
+LibriSpeech (MLS)~\cite{pratap_mls_2020}, VoxPopuli 
+(VP)~\cite{wang_voxpopuli_2021}, and BABEL (BBL)~\cite{gales_speech_2014}. CV is
 an open-source, multi-language, growing dataset of voices containing 11k hours
@@
 corpus~7.0). Researchers usually use part of this for pre-training (e.g., 7k
-hours/60 languages in \cite{babu2021xlsr} and 430 hours/29 languages in
+hours/60 languages in \cite{babu_xlsr_2021} and 430 hours/29 languages in
 \cite{kawakami_learning_2020}) or evaluation. 
-%Demographic metadata like age, sex, and accent is also provided.
 MLS derives content from read audiobooks of LibriVox and contains data in eight
@@
 languages.
-The entire dataset~\cite{babu2021xlsr} or a 24k-hour 
-portion~\cite{chen2021unispeechsat, chen_wavlm_2021} thereof has been used for pre-training. BBL
+The entire dataset~\cite{babu_xlsr_2021} or a 24k-hour 
+portion~\cite{chen_unispeechsat_2021, chen_wavlm_2021} thereof has been used for pre-training. BBL
 consists of 1k hours of conversational telephone speech in 17 African and Asian
@@
 
-Several datasets, including GigaSpeech~\cite{chen21o_interspeech}, TED-LIUM~3
-(TED3)~\cite{hernandez2018ted}, TED-LIUM~2 (TED2)~\cite{rousseau2012ted},
-Switchboard (SWB)~\cite{godfrey1992switchboard}, 
-TIMIT~\cite{garofolo_timit_1993}, and VoxLingua107~\cite{valk2021voxlingua107}, are
+Several datasets, including GigaSpeech~\cite{chen_gigaspeech_2021}, TED-LIUM~3
+(TED3)~\cite{hernandez_tedlium_2018}, TED-LIUM~2 (TED2)~\cite{rousseau_tedlium_2012},
+Switchboard (SWB)~\cite{godfrey_switchboard_1992}, 
+TIMIT~\cite{garofolo_timit_1993}, and VoxLingua107~\cite{valk_voxlingua107_2021}, are
 labeled and conventionally used for evaluation, while their audio streams are
 also aggregated to build diversified and large-scale corpora for unsupervised
-pre-training~\cite{kawakami_learning_2020, song_speech-xlnet_2020, babu2021xlsr}.
+pre-training~\cite{kawakami_learning_2020, song_speech-xlnet_2020, babu_xlsr_2021}.
 GigaSpeech is a multi-domain English ASR corpus with 33k hours of audio
@@
 recognition dataset containing two-sided telephone conversations.
-%The dataset was recorded in 1990 with a lower (8k) sampling rate than the other corpora, and thus presents a challenging recognition problem.
 The TIMIT corpus was designed to provide read speech data and its word and
 phone-level transcriptions for acoustic-phonetic studies. It contains recordings
-%from 630 speakers with 8 dialects 
 in American English. Compared to the previous corpora labeled for ASR
@@
 
-%, and each person read 10 phonetically-rich sentences.
-
 For the purpose of pre-training and evaluating Mandarin speech representations,
-the authors of \cite{jiang2019improving, jiang_further_2021} also compiled Open Mandarin, 
+the authors of \cite{jiang_improving_2019, jiang_further_2021} also compiled Open Mandarin, 
 an open-source Mandarin dataset of 1.5k hours of speech from
@@
 Mandarin Telephone Speech Corpus (HKUST, 200 hours of spontaneous 
-speech, of which   % AMH: check
-%from over 2100 speakers in mainland China. 
+speech, of which
 168 hours of audio is used for pre-training; the development and test sets are
-excluded.)~\cite{liu2006hkust}, AISHELL-1~\cite{bu2017aishell} (178 hours of
+excluded.)~\cite{liu_hkust_2006}, AISHELL-1~\cite{bu_aishell1_2017} (178 hours of
 read speech),
-%recorded in a quiet indoor environment from 400 speakers with different accents
-aidatatang 200zh (200 hours, read speech)~\cite{aidatatang}, MAGICDATA
-Mandarin Chinese Read Speech Corpus (755 hours, read speech)~\cite{magic_data},
-Free ST Chinese Mandarin Corpus (ST-CMDS, 100 hours, read speech)~\cite{st_cmds},
-and Primewords Chinese Corpus Set 1 (100 hours, read speech)~\cite{primewords_201801}. 
+aidatatang 200zh (200 hours, read speech)~\cite{beijingdatatangtechnologycoltd_aidatatang_}, MAGICDATA
+Mandarin Chinese Read Speech Corpus (755 hours, read speech)~\cite{magicdatatechnologycoltd_magicdata_2019},
+Free ST Chinese Mandarin Corpus (ST-CMDS, 100 hours, read speech)~\cite{surfingtechcoltd_stcmds_},
+and Primewords Chinese Corpus Set 1 (100 hours, read speech)~\cite{primewordsinformationtechnologycoltd_primewords_2018}. 
 Both HKUST and AISHELL-1 are labeled and are suitable for ASR evaluation.
@@
 Besides the aforementioned datasets, conventional speech processing benchmarks
-are also used to evaluate self-supervised representations. Studies leverage
-% Hub5'00, 
-  Hub5,     % AMH: check
+are also used to evaluate self-supervised representations. Studies leverage Hub5,
 DIRHA, and CHiME-5 to measure the efficacy of representations in ASR.
@@
 SWB studies but not released with the SWB dataset, and the rest are from CallHome
-American English Speech (LDC97S42). DIRHA~\cite{ravanelli2015dirha}, short for
+American English Speech (LDC97S42). DIRHA~\cite{ravanelli_dirhaenglish_2015}, short for
 Distant-speech Interaction for Robust Home Applications, is a database composed
@@
 phonetically-rich sentences. These utterances are read by UK and US English
-speakers and recorded with microphone arrays. CHiME-5~\cite{barker2018fifth} is
+speakers and recorded with microphone arrays. CHiME-5~\cite{barker_fifth_2018} is
 a challenge that aims to advance robust ASR and presents a dataset of natural
@@
 Researchers also evaluate representations for sentiment analysis with the
-INTERFACE~\cite{hozjan2002interface} and MOSEI (CMU Multimodal Opinion
-Sentiment and Emotion Intensity)~\cite{zadeh2018multimodal} datasets. INTERFACE
+INTERFACE~\cite{hozjan_interface_2002} and MOSEI (CMU Multimodal Opinion
+Sentiment and Emotion Intensity)~\cite{zadeh_multimodal_2018} datasets. INTERFACE
 is an emotional speech database for Slovenian, English, Spanish, and French,
@@
 annotations of 65 hours of YouTube videos
-%from 1000 speakers having 
 using emotion categories similar to INTERFACE, but replacing joy with
@@
 In addition, datasets employed to demonstrate the benefit of SSL
-representations on various tasks include VCTK \cite{veaux2016superseded} and VoxCeleb1 \cite{nagrani17_interspeech} for SID/ASV (automatic speaker verification) tasks, FSC (Fluent Speech Commands) \cite{lugosch19_interspeech} for IC (intent classification), QUESST
-(QUESST 2014) \cite{anguera2015quesst2014} for QbE (query by example), LS En-Fr \cite{kocabiyikoglu2018augmenting} and CoVoST-2 \cite{wang2020covost} for ST (speech translation), and ALFFA and OpenSLR-multi for multilingual ASR.
+representations on various tasks include VCTK \cite{veaux_cstr_2016} and VoxCeleb1 \cite{nagrani_voxceleb_2017} for SID/ASV (automatic speaker verification) tasks, FSC (Fluent Speech Commands) \cite{lugosch_speech_2019} for IC (intent classification), QUESST
+(QUESST 2014) \cite{anguera_quesst2014_2015} for QbE (query by example), LS En-Fr \cite{kocabiyikoglu_augmenting_2018} and CoVoST-2 \cite{wang_covost_2020} for ST (speech translation), and ALFFA and OpenSLR-multi for multilingual ASR.
 The VCTK corpus includes speech
@@
 project\footnote{\url{http://alffa.imag.fr}} collects speech of African
-languages to 
-% drive 
-  promote    % AMH: check
-the development of speech technologies in Africa, and
+languages to promote the development of speech technologies in Africa, and
 \cite{kawakami_learning_2020} leverages four African languages collected in the
-project for evaluation: Amharic~\cite{tachbelie2014using}, Fongbe~\cite{laleye2016first},
-Swahili~\cite{gelas2012developments}, and Wolof~\cite{gauthier2016collecting}.
+project for evaluation: Amharic~\cite{tachbelie_using_2014}, Fongbe~\cite{laleye_first_2016},
+Swahili~\cite{gelas_developments_2012}, and Wolof~\cite{gauthier_collecting_2016}.
 In the same work~\cite{kawakami_learning_2020}, the authors
@@
 Last, \cite{tagliasacchi_pretraining_2020} puts together five datasets 
-(MUSAN~\cite{snyder2015musan}, Bird Audio Detection~\cite{stowell2019automatic},
-Speech Commands~\cite{warden2018speech}, Spoken Language 
-Identification~\cite{oponowiczspoken}, and TUT Urban Acoustic Scenes 2018~\cite{dcase2018}) plus
+(MUSAN~\cite{snyder_musan_2015}, Bird Audio Detection~\cite{stowell_automatic_2019},
+Speech Commands~\cite{warden_speech_2018}, Spoken Language 
+Identification~\cite{oponowicz_spoken_2018}, and TUT Urban Acoustic Scenes 2018~\cite{mesaros_detection_2018}) plus
 an SID task built with the LS \textit{train-clean-100} subset to evaluate the
 capability of representations on audio event detection.
-\cite{quitry_learning_2019} employs the NSynth dataset~\cite{engel2017neural} on top
+\cite{quitry_learning_2019} employs the NSynth dataset~\cite{engel_neural_2017} on top
 of the six for benchmarking. As many of the datasets are built for research in
@@
   \textbf{EN}: English; 
-% \textbf{Multi}: multilinguality; 
-  \textbf{Multi}: multilingual;     % AMH: check
+  \textbf{Multi}: multilingual;
   \textbf{ZH}: Chinese;
@@
   \label{table:datasets}
+  \resizebox*{0.99\textwidth}{!}{%
   {\renewcommand*\arraystretch{1.2}
@@
     \midrule
-    % \multirow{2}{*}[0em]{wav2vec 2.0}
     LibriLight (LL) & PT & EN & 60k & - & MIT License \\ \hline
@@
 
-    Common Voice (CV-dataset) & PT/EV & Multi & 11k/7k \cite{babu2021xlsr}/430 \cite{kawakami_learning_2020} & ASR/PR/PC & CC0 \\ \hline % mention the split, CPC-modified, Bidir-CPC
+    Common Voice (CV-dataset) & PT/EV & Multi & 11k/7k \cite{babu_xlsr_2021}/430 \cite{kawakami_learning_2020} & ASR/PR/PC & CC0 \\ \hline % mention the split, CPC-modified, Bidir-CPC
     Multilingual LS (MLS) & PT/EV & Multi & 50k & ASR & CC BY 4.0 \\ \hline
-    VoxPopuli (VP) & PT/EV & Multi & 400k \cite{babu2021xlsr}/24k \cite{chen2021unispeechsat, chen_wavlm_2021} & ASR & CC0 \\ \hline
+    VoxPopuli (VP) & PT/EV & Multi & 400k \cite{babu_xlsr_2021}/24k \cite{chen_unispeechsat_2021, chen_wavlm_2021} & ASR & CC0 \\ \hline
     BABEL (BBL) & PT/EV & Multi & 1k & ASR & IARPA Babel Agreement  \\ \hline
 
-    GigaSpeech & PT/EV & EN & 40k/10k \cite{chen2021unispeechsat, chen_wavlm_2021} & ASR & Apache-2.0 License \\ \hline
+    GigaSpeech & PT/EV & EN & 40k/10k \cite{chen_unispeechsat_2021, chen_wavlm_2021} & ASR & Apache-2.0 License \\ \hline
     TED-LIUM 3 (TED3) & PT/EV & EN & 450 & ASR & CC BY-NC-ND 3.0 \\ \hline %Bidir-CPC
@@
     DIRHA & EV & EN & 11 & ASR & See link for details\textsuperscript{\ref{footnote:DIRHA_license}} \\ \hline % PASE
-    % from the paper Learning Problem-agnostic Speech Representations from Multiple Self-supervised Tasks - "in Section 4.3 we use the DIRHA dataset [36].
-Training and validation sets are based on the original WSJ-5k corpus (consisting of 7138 sentences uttered by 83 speakers) that is contaminated with a set of impulse re- sponses measured in a real apartment.
-The test set is composed of 409 WSJ sentences uttered by six American speakers and is based on real recordings in a domestic environment with a reverberation time of 0.7 s and an average signal-to-noise ratio of about 10 dB"
     CHiME-5 & EV & EN & 50 & ASR & See link for details\textsuperscript{\ref{footnote:CHiME_license}}\\ \hline % PASE+
@@
     \bottomrule
-  \end{tabular}}
+  \end{tabular}}}
 \end{table*}
@@
   \label{table:method_data_setting}
+  \resizebox*{0.99\textwidth}{!}{%
   {%\tabulinesep=1mm
@@
     \midrule
-
     % CPC
@@
     & & SID & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline
-
     % PASE
-    \multirow{2}{*}[-5mm]{PASE\cite{pascual_learning_2019}} & \multirow{2}{*}[-5mm]{LS 50 hrs \cite{ravanelli2018learning}} & SID & VCTK\textsuperscript{\ref{footnote:official-split}} & VCTK\textsuperscript{\ref{footnote:official-split}} & \checkmark & 44 hrs \\ \cline{3-7}
+    \multirow{2}{*}[-5mm]{PASE\cite{pascual_learning_2019}} & \multirow{2}{*}[-5mm]{LS 50 hrs \cite{ravanelli_learning_2018}} & SID & VCTK\textsuperscript{\ref{footnote:official-split}} & VCTK\textsuperscript{\ref{footnote:official-split}} & \checkmark & 44 hrs \\ \cline{3-7}
     & & Sentiment & INTERFACE\textsuperscript{\ref{footnote:interface-split}} & INTERFACE\textsuperscript{\ref{footnote:interface-split}} & \checkmark & 3 hrs \\ \cline{3-7}
@@
     % vq-VAE
-
     % MockingJay
@@
     & & Sentiment & MOSEI\textsuperscript{\ref{footnote:official-split}} & MOSEI\textsuperscript{\ref{footnote:official-split}} & \checkmark & 65 hrs \\ \hline
-
     % CPC modified
@@
     % PASE+
-    \multirow{3}{*}[0mm]{PASE+\cite{ravanelli_multi-task_2020}} & \multirow{3}{*}[0mm]{LS 50 hrs \cite{ravanelli2018learning}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
+    \multirow{3}{*}[0mm]{PASE+\cite{ravanelli_multi-task_2020}} & \multirow{3}{*}[0mm]{LS 50 hrs \cite{ravanelli_learning_2018}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
     & & ASR & DIRHA\textsuperscript{\ref{footnote:official-split}} & DIRHA\textsuperscript{\ref{footnote:official-split}} & \checkmark & 11 hrs \\ \cline{3-7}
     & & ASR & CHiME-5\textsuperscript{\ref{footnote:official-split}} & CHiME-5\textsuperscript{\ref{footnote:official-split}} & \checkmark & 50 hrs \\ \hline
-
     % AALBERT
@@
   \label{table:method_data_setting_2}
+  \resizebox*{0.99\textwidth}{!}{%
   {%\tabulinesep=1mm
@@
     \midrule
-
     % BMR
@@
     & & PR & WSJ si84/si284 & WSJ dev93 & - & 15/81 hrs \\ \hline
-
     % vq-APC
@@
     & & SID & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & \checkmark & 65\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline
-
     % vq-wav2vec + Discrete BERT
     \makecell[l]{vq-wav2vec + \\\:DiscreteBERT \cite{baevski_effectiveness_2020}} & LS 960 hrs & ASR & LS 100 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline
-
     % speech-XLNet
@@
     & & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{7\textsuperscript{\ref{footnote:low-resource}}/14\textsuperscript{\ref{footnote:low-resource}}/30\textsuperscript{\ref{footnote:low-resource}}/81 hrs}\vspace{1mm} \\ \hline
-
     % MPC
-    \multirow{3}{*}[0mm]{MPC \cite{jiang2019improving, jiang_further_2021}} & \multirow{2}{*}[0mm]{\makecell[l]{Didi Callcenter,\\Didi Dictation,\\Open Mandarin}} & ASR-zh & HKUST\textsuperscript{\ref{footnote:official-split}} & HKUST\textsuperscript{\ref{footnote:official-split}} & \checkmark & 168 hrs \\ \cline{3-7}
+    \multirow{3}{*}[0mm]{MPC \cite{jiang_improving_2019, jiang_further_2021}} & \multirow{2}{*}[0mm]{\makecell[l]{Didi Callcenter,\\Didi Dictation,\\Open Mandarin}} & ASR-zh & HKUST\textsuperscript{\ref{footnote:official-split}} & HKUST\textsuperscript{\ref{footnote:official-split}} & \checkmark & 168 hrs \\ \cline{3-7}
     & & ASR-zh & AISHELL-1\textsuperscript{\ref{footnote:official-split}} & AISHELL-1\textsuperscript{\ref{footnote:official-split}} & \checkmark & 178 hrs\vspace{1mm} \\ \cline{2-7}
     & \multirow{1}{*}[-0.5mm]{\makecell[l]{SWB, Fisher 1k,\\LS 960 hrs}} & ASR & SWB & Hub5'00 & - & 260 hrs \vspace{2.5mm} \\ \hline
-
     % MPE
@@
     & & ASR & LS 100/360/960 hrs & LS test-clean & - & 100/360/960 hrs \\ \hline
-
     % ConvDMM
-    ConvDMM \cite{khurana20_interspeech} & \makecell[l]{LS 50\cite{ravanelli2018learning}/360/\\\:\:\:960 hrs} & PC/PR & WSJ si284 & WSJ eval92 & \checkmark & \makecell[l]{5\textsuperscript{\ref{footnote:low-resource}}/50\textsuperscript{\ref{footnote:low-resource}}/100\textsuperscript{\ref{footnote:low-resource}} mins,\\ 4\textsuperscript{\ref{footnote:low-resource}}/8\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}} hrs} \\ \hline
+    ConvDMM \cite{khurana20_interspeech} & \makecell[l]{LS 50\cite{ravanelli_learning_2018}/360/\\\:\:\:960 hrs} & PC/PR & WSJ si284 & WSJ eval92 & \checkmark & \makecell[l]{5\textsuperscript{\ref{footnote:low-resource}}/50\textsuperscript{\ref{footnote:low-resource}}/100\textsuperscript{\ref{footnote:low-resource}} mins,\\ 4\textsuperscript{\ref{footnote:low-resource}}/8\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}} hrs} \\ \hline
     % wav2vec 2.0
@@
     DeCoAR 2.0 \cite{ling_decoar_2020} & LS 960 hrs & ASR & \makecell[l]{LS 100 hrs} & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline
-
     % TERA
@@
     HuBERT \cite{hsu_hubert_2021} & \makecell[l]{LS 960 hrs,\\ LL 60k hrs} & ASR & LS 960 hrs & \makecell[l]{LS test-clean\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \hline
-    %wav2vec-c
+    % wav2vec-c
     wav2vec-c \cite{sadhu_wav2vec-c_2021} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
-    UniSpeech-SAT \cite{chen2021unispeechsat} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See SUPERB \cite{yang_superb_2021} paper for details \\ \hline
+    UniSpeech-SAT \cite{chen_unispeechsat_2021} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See SUPERB \cite{yang_superb_2021} paper for details \\ \hline
     WavLM \cite{chen_wavlm_2021} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See \cite{yang_superb_2021} for details \\ \hline
-    \multirow{4}{*}[0mm]{XLS-R \cite{babu2021xlsr}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;\;+ CV-dataset-7k + VL\\\;\;\;+ BBL}} & ASR & VP, MLS, CV-dataset, BBL, LS & VP, MLS, CV-dataset, BBL, LS & - & \multirow{4}{*}[0mm]{See \cite{babu2021xlsr} for details} \\ \cline{3-6}
+    \multirow{4}{*}[0mm]{XLS-R \cite{babu_xlsr_2021}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;\;+ CV-dataset-7k + VL\\\;\;\;+ BBL}} & ASR & VP, MLS, CV-dataset, BBL, LS & VP, MLS, CV-dataset, BBL, LS & - & \multirow{4}{*}[0mm]{See \cite{babu_xlsr_2021} for details} \\ \cline{3-6}
     & & SID & VoxCeleb1 & VoxCeleb1 & \checkmark & \\ \cline{3-6}
@@
     & & LID & VL & VL & - & \\
-
-    % more & & & & & & \\ \hline
-    % https://tex.stackexchange.com/questions/66596/vertical-alignment-in-multirow-using-cells-with-1-lines
-    % https://texblog.org/2019/06/03/control-the-width-of-table-columns-tabular-in-latex/
-    %\multirow{2}{*}{wav2vec 2.0} & \multirow{2}{*}{wav2vec 2.0} & Population Population Population Population & 2 & 3 & 4 \\\cline{3-6}
-    %http://www.emerson.emory.edu/services/latex/latex_119.html
-    %& &2016&2017& 5 & 6 \\ \hline
     \bottomrule
-  \end{tabular}}
+  \end{tabular}}}
 \end{table*}
@@
 The 6 AED datasets used in
-\cite{tagliasacchi_pretraining_2020} are MUSAN~\cite{snyder2015musan}, Bird Audio
-Detection~\cite{stowell2019automatic}, Speech Commands~\cite{warden2018speech},
-Spoken Language Identification~\cite{oponowiczspoken}, TUT Urban Acoustic
-Scenes 2018~\cite{dcase2018} plus an SID task built with LS
+\cite{tagliasacchi_pretraining_2020} are MUSAN~\cite{snyder_musan_2015}, Bird Audio
+Detection~\cite{stowell_automatic_2019}, Speech Commands~\cite{warden_speech_2018},
+Spoken Language Identification~\cite{oponowicz_spoken_2018}, TUT Urban Acoustic
+Scenes 2018~\cite{mesaros_detection_2018} plus an SID task built with LS
 \textit{train-clean-100}. In addition to the 6 datasets,
-\cite{quitry_learning_2019} use the NSynth dataset~\cite{engel2017neural} for
+\cite{quitry_learning_2019} use the NSynth dataset~\cite{engel_neural_2017} for
 evaluation.}
@@
 CV-dataset, LS, WSJ, TIMIT, Speech Accent Archive 
-(SSA)~\cite{weinberger2009towards}, TED3, and SWB. SSA is a growing annotated corpus of
+(SSA)~\cite{weinberger_speech_2011}, TED3, and SWB. SSA is a growing annotated corpus of
 English speech with various accents. Among the papers studied in this review, SSA
@@
 fine-tune a pre-trained SSL model for a supervised downstream task.
-% taking a network pre-trained with SSL as initialization, and then fine-tuning and evaluating the network with labeled data in downstream tasks.
 Depending on the corpora used in pre-training and fine-tuning, techniques can
@@
 datasets (10k hours) for wav2vec-c~\cite{sadhu_wav2vec-c_2021}, Didi internal
-datasets (10k hours) for MPC~\cite{jiang2019improving, jiang_further_2021}, the
+datasets (10k hours) for MPC~\cite{jiang_improving_2019, jiang_further_2021}, the
 combination of Gigaspeech, VP-24k, and LL (94k hours in total) for
-UniSpeech-SAT~\cite{chen2021unispeechsat} and WavLM~\cite{chen_wavlm_2021}, and
+UniSpeech-SAT~\cite{chen_unispeechsat_2021} and WavLM~\cite{chen_wavlm_2021}, and
 the combination of VP-400K, MLS, CV-dataset, VL and BBL (436k hours in total)
-for XLS-R~\cite{babu2021xlsr}. We expect this trend to continue with the growth
+for XLS-R~\cite{babu_xlsr_2021}. We expect this trend to continue with the growth
 in available computing power. Most studies focuses on learning representations
-for English, whereas Chinese~\cite{jiang2019improving, jiang_further_2021} and
-multilinguality~\cite{kawakami_learning_2020, babu2021xlsr} are also gaining
+for English, whereas Chinese~\cite{jiang_improving_2019, jiang_further_2021} and
+multilinguality~\cite{kawakami_learning_2020, babu_xlsr_2021} are also gaining
 attention. Compared to pre-training, datasets used for fine-tuning are more
@@
 downstream tasks have gained attention for evaluating SSL
-methodologies~\cite{chen2021unispeechsat, chen_wavlm_2021}. The goal of such benchmarks is
+methodologies~\cite{chen_unispeechsat_2021, chen_wavlm_2021}. The goal of such benchmarks is
 to provide a holistic evaluation of the performance of learned representations; 
@@
 \centering
-%\includegraphics[width=1.0\textwidth]{paper_review/ls-test-clean-other.pdf}
 \includegraphics[width=0.9\textwidth]{paper_review/ls-test-clean-v2.pdf}
@@
 approaches (iterative pseudo labeling (IPL)~\cite{xu_iterative_2020}, 
-slimIPL~\cite{likhomanenko2020slimipl}, noisy student~\cite{park_improved_2020}).
+slimIPL~\cite{likhomanenko_slimipl_2021}, noisy student~\cite{park_improved_2020}).
 These approaches are visualized in black.
-Also, note that the current state of the art---conformer XXL + noisy student~\cite{zhang2020pushing}---is a combination of self-training and SSL techniques.
+Also, note that the current state of the art---conformer XXL + noisy student~\cite{zhang_pushing_2020}---is a combination of self-training and SSL techniques.
 Given the
@@
 show performance from such methods (iterative pseudo labeling 
-(IPL)~\cite{xu_iterative_2020}, slimIPL~\cite{likhomanenko2020slimipl}, noisy 
+(IPL)~\cite{xu_iterative_2020}, slimIPL~\cite{likhomanenko_slimipl_2021}, noisy 
 student~\cite{park_improved_2020}), as well as the current state of the art---conformer XXL + noisy
-student~\cite{zhang2020pushing}---which augments SSL with various advanced
+student~\cite{zhang_pushing_2020}---which augments SSL with various advanced
 techniques including self-training. Furthermore, we illustrate in the figure
@@
 features, and with the growing investment in model size, better performance is
-achieved. The largest ones, such as wav2vec 2.0-L and HuBERT-L/XL, yield
-% results competitive to the state of the art 
-  competitive results                         % AMH: check
+achieved. The largest ones, such as wav2vec 2.0-L and HuBERT-L/XL, yield competitive results 
 when the entire 960-hour of labeled data is used in
@@
 slow or costly.
-% discuss the trend
-%[Pick representative tasks and datasets and show the benefit of SSL, e.g., fewer fine-tuning/labeled data is required to achieve better/comparable performance] [talk about trend as well]  The plot suggests that ASR results have been increased significantly with the progress in SSL.
-Near SOTA performance can be achieved with much fewer training examples.
 
@@
 combining SSL pre-training and self-training 
-techniques~\cite{zhang2020pushing}, SSL models 
-% approach the state of the art 
-  are competitive                % AMH: check
-in other tasks, including IC,
+techniques~\cite{zhang_pushing_2020}, SSL models are competitive in other tasks, including IC,
 SID, ASV, and QbE. We summarize the performance of these models and previous
@@
     \midrule
-    ASR (WER $\downarrow$) & LS test-clean/other & 2.1/4.0 \cite{xu_iterative_2020} & 1.4/2.6 \cite{zhang2020pushing} \\ \hline
-    IC (Acc $\uparrow$) & FSC & 98.8 \cite{lugosch19_interspeech} & 99.3\cite{chen2021unispeechsat} \\ \hline
-    SID (Acc $\uparrow$) & VoxCeleb1 & 94.8 \cite{hajibabaei2018unified} & 95.5 \cite{chen_wavlm_2021} \\ \hline
-    ASV (EER $\downarrow$) & VoxCeleb1 & 3.1 \cite{hajavi2021siamese} & 2.4 \cite{wang2021fine} \\ \hline
-    QbE (MTWV $\uparrow$) & QUESST (EN) & 10.6 \cite{rodriguez2014gtts} & 11.2\cite{chen2021unispeechsat} \\
-
-    %wav2vec-c
-    %wav2vec-c \cite{sadhu_wav2vec-c_2021} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
-    %UniSpeech-SAT \cite{chen2021unispeechsat} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & Check SUPERB \cite{yang_superb_2021} paper for details \\ \hline
-    %WavLM \cite{chen_wavlm_2021} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & Check \cite{yang_superb_2021} for details \\ \hline
-    %\multirow{4}{*}[0mm]{XLS-R \cite{babu2021xlsr}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;\;+ CV-dataset-7k + VL\\\;\;\;+ BBL}} & ASR & VP, MLS, CV-dataset, BBL, LS & VP, MLS, CV-dataset, BBL, LS & - & \multirow{4}{*}[0mm]{Check \cite{babu2021xlsr} for details} \\ \cline{3-6}
-    %& & SID & VoxCeleb1 & VoxCeleb1 & \checkmark & \\ \cline{3-6}
-    %& & ST & CoVoST-2 & CoVoST-2 & \checkmark & \\ \cline{3-6}
-    %& & LID & VL & VL & - & \\ \hline
-
+    ASR (WER $\downarrow$) & LS test-clean/other & 2.1/4.0 \cite{xu_iterative_2020} & 1.4/2.6 \cite{zhang_pushing_2020} \\ \hline
+    IC (Acc $\uparrow$) & FSC & 98.8 \cite{lugosch_speech_2019} & 99.3\cite{chen_unispeechsat_2021} \\ \hline
+    SID (Acc $\uparrow$) & VoxCeleb1 & 94.8 \cite{hajibabaei_unified_2018} & 95.5 \cite{chen_wavlm_2021} \\ \hline
+    ASV (EER $\downarrow$) & VoxCeleb1 & 3.1 \cite{hajavi_siamese_2021} & 2.4 \cite{wang_finetuned_2021} \\ \hline
+    QbE (MTWV $\uparrow$) & QUESST (EN) & 10.6 \cite{rodriguez-fuentes_gttsehu_2014} & 11.2\cite{chen_unispeechsat_2021} \\
     \bottomrule
@@
 efforts include SUPERB~\cite{yang_superb_2021}, 
-LeBenchmark~\cite{evain21_interspeech}, ZeroSpeech~\cite{dunbar2020zero},
-HEAR \cite{pmlr-v176-turian22a},
-NOSS~\cite{shor20_interspeech}, and HARES~\cite{wang2021towards}.
+LeBenchmark~\cite{evain_lebenchmark_2021}, ZeroSpeech~\cite{dunbar_zero_2020},
+HEAR \cite{turian_hear_2022},
+NOSS~\cite{shor_learning_2020}, and HARES~\cite{wang_learning_2021}.
 
@@
 to democratize SSL beyond race among Big Tech.   % AMH: what does this mean?
-LeBenchmark~\cite{evain21_interspeech} shares a vision similar to SUPERB and provides a
+LeBenchmark~\cite{evain_lebenchmark_2021} shares a vision similar to SUPERB and provides a
 reproducible framework for assessing SSL in French with ASR, spoken language
 understanding, speech translation, and emotion recognition. 
-ZeroSpeech~\cite{dunbar2020zero} (described in more detail in \cref{zero_speech})
+ZeroSpeech~\cite{dunbar_zero_2020} (described in more detail in \cref{zero_speech})
 challenges the scientific community to build speech and language understanding
 systems using zero expert resources 
-%with the ultimate goal of bringing the systems to the services 
 for millions of users of ``low-resource" languages.
-%or in abnormal condition (dyslexia, autism, etc). ZeroSpeech offers evaluation for independent units in the system as well aggregation of multiple components. 
 SSL techniques are also benchmarked with the ZeroSpeech 
-challenge~\cite{tjandra20_interspeech, niekerk20b_interspeech}. Apart from the speech
+challenge~\cite{tjandra_transformer_2020, vanniekerk_vectorquantized_2020}. Apart from the speech
 community, researchers have also established HEAR (holistic evaluation of audio
-representations) \cite{pmlr-v176-turian22a}, NOSS (non-semantic
-speech benchmark)~\cite{shor20_interspeech}, and HARES (holistic audio
-representation evaluation suite)~\cite{wang2021towards} to benchmark audio
+representations) \cite{turian_hear_2022}, NOSS (non-semantic
+speech benchmark)~\cite{shor_learning_2020}, and HARES (holistic audio
+representation evaluation suite)~\cite{wang_learning_2021} to benchmark audio
 representations. These efforts promote the creation of an audio embedding
@@
 active research area.
-
-%SUPERB motivation: consolidating evaluation recipes to understand the capability of various SSL in a fair way.
-SUPERB designates the architectures of prediction heads for each downstream tasks [common architectures] [frozen] [generalizable, comparable, reproducible results].
-[Results on tasks with such constraint conditions (mainly SUPERB but also from literatures imposing similar constraint).] [discussion on the results - competitive performance with simple downstream recipes] [More challenges, ZeroSpeech, HEAR ...]
-
-%Librispeech 100h-960h \\
-%Libri-light 10m, 1h, 10h - 60k \\
-%SUPERB \\
-%Add more
-%Hear 2021 Challenge? (focus on audio, not only speech)

### 2023-03-17 chapters/paper_review/05_zero.tex
Change citation keys (WIP)

@@
     Both tracks were based on the English and Xitsonga languages.
-	 \item The Zero Resource Speech Challenge 2017~\cite{dunbar2017zero} focused on 
+	 \item The Zero Resource Speech Challenge 2017~\cite{dunbar_zero_2017} focused on 
 	 unseen language and speaker aspects from the previous challenge. For
@@
 	 the amount of speech available for each speaker.
-	 \item The Zero Resource Speech Challenge 2019~\cite{dunbar2019zero} extended a goal of
+	 \item The Zero Resource Speech Challenge 2019~\cite{dunbar_zero_2019} extended a goal of
 	 previous challenges by synthesizing speech without
@@
 	 and speaker similarity.
-	 \item The Zero Resource Speech Challenge 2020~\cite{dunbar2020zero} was based on two
+	 \item The Zero Resource Speech Challenge 2020~\cite{dunbar_zero_2020} was based on two
 	 tracks, revisiting previous challenges with different evaluation metrics. 
@@
 autoencoders~\cite{tjandra19_interspeech,chorowski_unsupervised_2019}, and incorporation of
-neural-network-based TTS/VC techniques~\cite{niekerk20b_interspeech} to
+neural-network-based TTS/VC techniques~\cite{vanniekerk_vectorquantized_2020} to
 self-supervised learning~\cite{maekaku21_interspeech}.

### 2023-03-17 chapters/paper_review/main_hi.tex
Change citation keys (WIP)

--- 2023-03-16 chapters/paper_review/main_hi.tex
@@
         \multicolumn{7}{c}{\textit{\textbf{Pre-Training}}} \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec} & LL-60k & Transformer & 1.6 & 3.0 & 1.8 & 3.3 \\
-        pre-trained Conformer XXL~\cite{zhang2020pushing} & LL-60k & LSTM & 1.5 & 3.0 & 1.5 & 3.1 \\
+        wav2vec 2.0 \textsc{Large}~\cite{baevski_wav2vec_2020} & LL-60k & Transformer & 1.6 & 3.0 & 1.8 & 3.3 \\
+        pre-trained Conformer XXL~\cite{zhang_pushing_2020} & LL-60k & LSTM & 1.5 & 3.0 & 1.5 & 3.1 \\
 
@@
         wav2vec 2.0 + self-training~\cite{xu2020self} & LL-60k & Transformer & 1.1 & 2.7 & 1.5 & 3.1\\
-        pre-trained Conformer XXL + Noisy Student~\cite{zhang2020pushing} & LL-60k & LSTM & 1.3 & 2.6 & 1.4 & 2.6 \\
+        pre-trained Conformer XXL + Noisy Student~\cite{zhang_pushing_2020} & LL-60k & LSTM & 1.3 & 2.6 & 1.4 & 2.6 \\
 

### 2023-03-17 chapters/paper_review/main_lo.tex
Change citation keys (WIP)

@@
         \multicolumn{7}{c}{\textit{\textbf{10-hour labeled}}} \\
-        SlimIPL~\cite{likhomanenko2020slimipl} & LS-960 & 4-gram + Transformer & 5.3 & 7.9 & 5.5 & 9.0 \\
+        SlimIPL~\cite{likhomanenko_slimipl_2021} & LS-960 & 4-gram + Transformer & 5.3 & 7.9 & 5.5 & 9.0 \\
         DeCoAR 2.0~\cite{ling_decoar_2020} & LS-960 & 4-gram & - & - & 5.4 & 13.3 \\
@@
         IPL~\cite{xu_iterative_2020} & LL-60k & 4-gram + Transformer & 3.19 & 6.14 & 3.72 & 7.11 \\
-        SlimIPL~\cite{likhomanenko2020slimipl} & LS-860 & 4-gram + Transformer & 2.2 & 4.6 & 2.7 & 5.2 \\
+        SlimIPL~\cite{likhomanenko_slimipl_2021} & LS-860 & 4-gram + Transformer & 2.2 & 4.6 & 2.7 & 5.2 \\
         Noisy Student\cite{park_improved_2020} & LS-860 & LSTM & 3.9 & 8.8 & 4.2 & 8.6 \\

### 2023-04-12 chapters/paper_benchmarking_workshop.tex
Add review paper (WIP)
Change citation keys (WIP)
Update reference keys for review paper

@@
 \section{Introduction}
-Since their introduction, temporal latent variable models (LVMs) for speech \cite{chung_recurrent_2015, fraccaro_sequential_2016} based on the variational autoencoder (VAE,  \textcite{kingma_auto-encoding_2014, rezende_stochastic_2014}) have seen little development compared to their image domain counterparts.
+Since their introduction, temporal latent variable models (LVMs) for speech \cite{chung_recurrent_2015, fraccaro_sequential_2016} based on the variational autoencoder (VAE,  \textcite{kingma_autoencoding_2014, rezende_stochastic_2014}) have seen little development compared to their image domain counterparts.
 While LVMs now achieve superior likelihoods on images compared to deterministic models \cite{child_very_2021,sinha_consistency_2021,kingma_variational_2021}, they are still inferior to deterministic models such as WaveNet on speech \cite{oord_wavenet_2016}.
@@
     \end{multicols}
-    \vspace{-4mm}
 \caption{
@@
 Red arrows indicate purely deterministic paths from the output $\vx_t$ to previous input $\vx_{<t}$ without passing a stochastic node. 
-We provide additional graphical illustrations including inference models in appendix~\ref{app: additional graphical models}.
+We provide additional graphical illustrations including inference models in \cref{app: additional graphical models}.
 }
 \label{fig: autoregressive vrnn srnn cwvae 1L graphs}
-\vspace{-3mm}
 \end{figure*}
@@
 \paragraph{Sequential deep latent variable models}
-The models we consider are all sequential deep latent variable models trained with variational inference and the reparameterization trick \cite{kingma_auto-encoding_2014}.
+The models we consider are all sequential deep latent variable models trained with variational inference and the reparameterization trick \cite{kingma_autoencoding_2014}.
 The input is a variable-length sequence $\vx = \vx_{1:T} = \left(\vx_1, \vx_2, \dots, \vx_{T}\right)$ with $\vx_t\in\mathbb{R}^{D_x}$. 

### 2023-04-12 chapters/paper_hierarchical.tex
Change citation keys (WIP)
Update reference keys for review paper

@@
     \includegraphics[width=1\columnwidth]{paper_hierarchical/feature-correlation-heatmap2.pdf}
-    \vspace{-5mm}
+    \vspace{0mm}
     \caption[Absolute correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST.]{
@@
     }
-    \vspace{-3mm}
+    \vspace{0mm}
     \label{fig:correlations-heatmap}
@@
     % https://docs.google.com/drawings/d/1UXdA3c18Oek9_env1vEB9mqOAHZ-tu0ijWmvqmapWCk/edit?usp=sharing
-    \vspace{-5.2mm}
+    \vspace{0mm}
     \caption[Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables]{Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables  \cite{maaloe_biva_2019}.
@@
     A man with long hair collapses into the mode of the more common long-haired woman.}
-    \vspace{-3mm}
+    \vspace{0mm}
     \label{fig:biva-reconstructions-celeba}
@@
 
-\vspace{-0.35cm}
+\vspace{0cm}
 \section{Background and related work}
@@
 \begin{equation}
-    LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq k}|\zb_{>k}) q(\zb_{>k}|\xb) || p(\zb|\xb)) \label{eq:llr-as-kls-iwae-reduced}
-\end{equation}
+    LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq k}|\zb_{>k}) q(\zb_{>k}|\xb) || p(\zb|\xb))
+\end{equation}\label{eq:llr-as-kls-iwae-reduced}
 which, in practice, is well-approximated for a finite $S$.
@@
 
-Since the scores in \cref{eq:llr-as-kls, eq:llr-as-kls-iwae-reduced} are estimated by sampling their estimators are stochastic objects with nonzero variance.
+Since the scores in \cref{eq:llr-as-kls,eq:llr-as-kls-iwae-reduced} are estimated by sampling their estimators are stochastic objects with nonzero variance.
 We note that $\text{Var}(\widehat{LLR}^{>k}) = \text{Var}(\hat{\mathcal{L}}) + \text{Var}(\hat{\mathcal{L}}^{>k}) - 2\, \text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$.
@@
 \begin{figure}[t]
-    % \vspace{-1.3cm}
     %\begin{figure}
@@
     % \end{figure}
-    %\vspace{-0.8cm}
 %\end{wrapfigure}
@@
     \label{tab:bits-per-dim-ood}
-    \vspace{-3mm}
+    \vspace{0mm}
 \end{table}
 
-\footnotetext{
+\footnotetext{%
     \textcite{serra_input_2020} performs the best when high likelihoods are assigned to OOD data such that the overlap with in-distribution data is low.
-    Performance is worse when the overlap is high, cf. \textcite[Table 1]{serra_input_2020}, as seen with complex images. 
-}
+    Performance is worse when the overlap is high, cf. \textcite[Table 1]{serra_input_2020}, as seen with complex images.}
 \begin{table}[t!]

### 2023-04-12 chapters/paper_review/01_introduction.tex
Update reference keys for review paper

@@
 with labels to guide learning using a larger volume of unlabeled samples.
-Examples of unsupervised learning techniques include k-means clustering~\cite{gray_vector_1984}, mixture models~\cite{jordan_hierarchical_1994}, autoencoders~\cite{hinton_autoencoders_1994},
+Examples of unsupervised learning techniques include $k$-means clustering~\cite{gray_vector_1984}, mixture models~\cite{jordan_hierarchical_1994}, autoencoders~\cite{hinton_autoencoders_1994},
 and non-negative matrix factorization~\cite{lee_learning_1999}. 
@@
 downstream tasks.
-{\color{black} For example, unsupervised k-means clustering doesn't adhere to this definition of self-supervision since it iteratively minimizes the within-cluster variance during learning.}
+{\color{black} For example, unsupervised $k$-means clustering doesn't adhere to this definition of self-supervision since it iteratively minimizes the within-cluster variance during learning.}
 In this review, we

### 2023-04-12 chapters/paper_review/02_three_waves.tex
Update reference keys for review paper

@@
 as \cite{rabiner_considerations_1979,wilpon_modified_1985}, word patterns were clustered
-semi-automatically using techniques such as k-means, after which isolated words
+semi-automatically using techniques such as $k$-means, after which isolated words
 were recognized by finding the training cluster closest to 
@@
 sentence such as with the bidirectional encoder representations from Transformers
-(BERT) series~\cite{devlin_bert:_2018,liu_roberta_2019}. 
+(BERT) series~\cite{devlin_bert_2018,liu_roberta_2019}. 
 Another common pretext task in the third wave is contrastive

### 2023-04-12 chapters/paper_review/03_approaches.tex
Update reference keys for review paper

@@
 
-Masked reconstruction is largely inspired by the masked language model (MLM) task from BERT~\cite{devlin_bert:_2018}.
+Masked reconstruction is largely inspired by the masked language model (MLM) task from BERT~\cite{devlin_bert_2018}.
 During BERT pre-training, some tokens in the input sentences are masked by randomly replacing them by a learned masking token or another input token.
@@
 In non-autoregressive predictive coding (NPC)~\cite{liu_non-autoregressive_2020}, time masking is introduced through masked convolution blocks.
-Taking inspiration from XLNet~\cite{yang_xlnet_2019}, it has also been suggested that the input be reconstructed from a shuffled version~\cite{song_speech-xlnet_2020} to address the discrepancy between pre-training and fine-tuning of masking-based approaches.
+Taking inspiration from XLNet~\cite{yang_xlnet_2019}, it has also been suggested that the input be reconstructed from a shuffled version~\cite{song_speechxlnet_2020} to address the discrepancy between pre-training and fine-tuning of masking-based approaches.
 
@@
 Other than the autoregressive and masked reconstruction tasks discussed above, various studies have explored the reconstruction of other targets derived from the input.
-PASE and PASE+~\cite{pascual_learning_2019,ravanelli_multi-task_2020} use multiple targets, including the waveform, log power spectrum, \edit{mel cepstral coefficients (MFCCs)}, and prosody features.
+PASE and PASE+~\cite{pascual_learning_2019,ravanelli_multitask_2020} use multiple targets, including the waveform, log power spectrum, \edit{mel cepstral coefficients (MFCCs)}, and prosody features.
 Models that learn acoustic embeddings of small speech segments have targeted future and past spectrogram segments~\cite{chung_speech2vec_2018, tagliasacchi_selfsupervised_2019, tagliasacchi_pretraining_2020}, phase information~\cite{quitry_learning_2019}, and the temporal gap between two segments~\cite{tagliasacchi_selfsupervised_2019, tagliasacchi_pretraining_2020}.
@@
 
-The wav2vec model \cite{schneider_wav2vec:_2019} extends the CPC approach \edit{and uses fully convolutional parameterizations for the modules $f_1(\cdot)$ and $f_2(\cdot)$ with receptive fields of \SI{30}{ms} and \SI{210}{ms}, respectively.
+The wav2vec model \cite{schneider_wav2vec_2019} extends the CPC approach \edit{and uses fully convolutional parameterizations for the modules $f_1(\cdot)$ and $f_2(\cdot)$ with receptive fields of \SI{30}{ms} and \SI{210}{ms}, respectively.
 While the CPC loss solves a 1-of-$N$ classification task per $(t,k)$, either assigning the anchor to the positive class or (wrongly) to one of the $N-1$ negative classes, the wav2vec loss considers a sequence of $N$ independent binary classifications.
@@
 Thus, the predictive setup is somewhat akin to teacher-student training.
-The first predictive approaches were motivated by the success of BERT-like methods in NLP \cite{devlin_bert:_2018} as well as the DeepCluster method in CV \cite{caron_deep_2018}.}
+The first predictive approaches were motivated by the success of BERT-like methods in NLP \cite{devlin_bert_2018} as well as the DeepCluster method in CV \cite{caron_deep_2018}.}
 
@@
 
-\paragraph{HuBERT} Rather than relying on an advanced representation learning model for discretizing continuous inputs, as Discrete BERT, the Hidden Unit BERT (HuBERT) approach \cite{hsu_hubert_2021} \edit{uses quantized MFCC features as targets learned with classic k-means.
-Thus, to compute the targets, the k-means model $g_1(\cdot)$ assigns a cluster center to each timestep.
+\paragraph{HuBERT} Rather than relying on an advanced representation learning model for discretizing continuous inputs, as Discrete BERT, the Hidden Unit BERT (HuBERT) approach \cite{hsu_hubert_2021} \edit{uses quantized MFCC features as targets learned with classic $k$-means.
+Thus, to compute the targets, the $k$-means model $g_1(\cdot)$ assigns a cluster center to each timestep.
 Different from Discrete BERT, HuBERT takes the raw waveform as input, rather than discrete units. This helps to prevent loss of any relevant information due to input quantization.
@@
 The first iteration is described above.
-Once completed, a second iteration of pre-training follows. Here, representations from a hidden layer of the model from the first iteration are clustered with k-means to obtain new targets $c_t$.}
+Once completed, a second iteration of pre-training follows. Here, representations from a hidden layer of the model from the first iteration are clustered with $k$-means to obtain new targets $c_t$.}
 

### 2023-04-12 chapters/paper_review/04_benchmarking.tex
Update reference keys for review paper

@@
 also aggregated to build diversified and large-scale corpora for unsupervised
-pre-training~\cite{kawakami_learning_2020, song_speech-xlnet_2020, babu_xlsr_2021}.
+pre-training~\cite{kawakami_learning_2020, song_speechxlnet_2020, babu_xlsr_2021}.
 GigaSpeech is a multi-domain English ASR corpus with 33k hours of audio
@@
     % wav2vec
-    \multirow{2}{*}[0mm]{wav2vec \cite{schneider_wav2vec:_2019}} & \multirow{2}{*}[-0.5mm]{\makecell[l]{LS 80/960 hrs,\\ LS 960 hrs \\ \:+ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & \checkmark & 81 hrs\vspace{1mm} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{wav2vec \cite{schneider_wav2vec_2019}} & \multirow{2}{*}[-0.5mm]{\makecell[l]{LS 80/960 hrs,\\ LS 960 hrs \\ \:+ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & \checkmark & 81 hrs\vspace{1mm} \\ \cline{3-7}
     & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hr\vspace{1mm} \\ \hline
@@
     % DeCoAR
-    \multirow{2}{*}[0mm]{DeCoAR \cite{ling2020deep}} & \multirow{2}{*}[0mm]{\makecell[l]{LS 100/360/460/\\\:\:\:960 hrs,\\ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{25\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}}/81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{DeCoAR \cite{ling_deep_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{LS 100/360/460/\\\:\:\:960 hrs,\\ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{25\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}}/81 hrs} \\ \cline{3-7}
     & & ASR & \makecell[l]{LS 100/360/\\\:\:\:460/960 hrs} & \makecell[l]{LS test-clean,\\ LS test-other} & - & 100/360/460/960 hrs \\ \hline
     % MT-APC
-    \multirow{2}{*}[0mm]{MT-APC\cite{chung2020improved}} & \multirow{2}{*}[0mm]{LS 360 hrs} & ASR & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & WSJ dev93 & \checkmark & 72 hrs \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{MT-APC\cite{chung_improved_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & ASR & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & WSJ dev93 & \checkmark & 72 hrs \\ \cline{3-7}
     & & ST & LS En-Fr\textsuperscript{\ref{footnote:official-split}} & LS En-Fr\textsuperscript{\ref{footnote:official-split}} & - & 236 hrs \\ \hline
     % PASE+
-    \multirow{3}{*}[0mm]{PASE+\cite{ravanelli_multi-task_2020}} & \multirow{3}{*}[0mm]{LS 50 hrs \cite{ravanelli_learning_2018}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
+    \multirow{3}{*}[0mm]{PASE+\cite{ravanelli_multitask_2020}} & \multirow{3}{*}[0mm]{LS 50 hrs \cite{ravanelli_learning_2018}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
     & & ASR & DIRHA\textsuperscript{\ref{footnote:official-split}} & DIRHA\textsuperscript{\ref{footnote:official-split}} & \checkmark & 11 hrs \\ \cline{3-7}
@@
     % speech-XLNet
-    \multirow{2}{*}[-1mm]{speech-XLNet \cite{song_speech-xlnet_2020}} & \multirow{2}{*}[-1mm]{\makecell[l]{LS 960 hrs\\\;\;\;+ WSJ si284\\\;\;\;+ TED2}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs\vspace{1mm} \\ \cline{3-7}
+    \multirow{2}{*}[-1mm]{speech-XLNet \cite{song_speechxlnet_2020}} & \multirow{2}{*}[-1mm]{\makecell[l]{LS 960 hrs\\\;\;\;+ WSJ si284\\\;\;\;+ TED2}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs\vspace{1mm} \\ \cline{3-7}
     & & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{7\textsuperscript{\ref{footnote:low-resource}}/14\textsuperscript{\ref{footnote:low-resource}}/30\textsuperscript{\ref{footnote:low-resource}}/81 hrs}\vspace{1mm} \\ \hline
@@
     % ConvDMM
-    ConvDMM \cite{khurana20_interspeech} & \makecell[l]{LS 50\cite{ravanelli_learning_2018}/360/\\\:\:\:960 hrs} & PC/PR & WSJ si284 & WSJ eval92 & \checkmark & \makecell[l]{5\textsuperscript{\ref{footnote:low-resource}}/50\textsuperscript{\ref{footnote:low-resource}}/100\textsuperscript{\ref{footnote:low-resource}} mins,\\ 4\textsuperscript{\ref{footnote:low-resource}}/8\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}} hrs} \\ \hline
+    ConvDMM \cite{khurana_convolutional_2020} & \makecell[l]{LS 50\cite{ravanelli_learning_2018}/360/\\\:\:\:960 hrs} & PC/PR & WSJ si284 & WSJ eval92 & \checkmark & \makecell[l]{5\textsuperscript{\ref{footnote:low-resource}}/50\textsuperscript{\ref{footnote:low-resource}}/100\textsuperscript{\ref{footnote:low-resource}} mins,\\ 4\textsuperscript{\ref{footnote:low-resource}}/8\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}} hrs} \\ \hline
     % wav2vec 2.0

### 2023-04-12 chapters/paper_review/04b_analysis.tex
Update reference keys for review paper

@@
 correspond to the same word, and a standard benchmark of 11 word
-similarity tasks~\cite{faruqi14}. These are mostly used to gauge the
+similarity tasks~\cite{faruqi_community_2014}. These are mostly used to gauge the
 amount of lexical information present in the embeddings.  Using this
@@
 phonetic or word-related information may not be the best
-strategy. Indeed, the authors of \cite{baevski2021unsupervised} show that a
+strategy. Indeed, the authors of \cite{baevski_unsupervised_2021} show that a
 phone classifier trained on each of the 24 frozen layers of a wav2vec~2.0 model
@@
 an analysis of text-based Transformers trained with a masked language model
-criterion~\cite{voita2019} observed a similar compression plus reconstruction
+criterion~\cite{voita_bottomup_2019} observed a similar compression plus reconstruction
 pattern. Interestingly, similar network behavior was also recently described
@@
 self-supervised learner (SimCLR) that optimizes for augmentation invariance,
-\cite{grigg2021self} show that it is the intermediate representations that most
+\cite{grigg_selfsupervised_2021} show that it is the intermediate representations that most
 closely approximate information learned in a supervised way, i.e., they
@@
 information. 
-\edit{It is shown that the per-utterance mean of CPC features captures speaker information to a large extent\cite{van2021analyzing}.}
+\edit{It is shown that the per-utterance mean of CPC features captures speaker information to a large extent\cite{vanniekerk_analyzing_2021}.}
 Location of information pertaining to speakers vs.\ language classes was
-analyzed in \cite{ling20odyssey} for a 12-layer BERTphone model. This model
+analyzed in \cite{ling_bertphone_2020} for a 12-layer BERTphone model. This model
 combines a self-supervised masked reconstruction loss with a phone-based CTC
@@
 necessary for speaker diarization and verification whereas higher layers encode
-phonetic and semantic information. Another study~\cite{wang2021layersup} used
+phonetic and semantic information. Another study~\cite{wang_selfsupervised_2021} used
 explicit self-supervised loss at the intermediate layers rather than just the
@@
 Most self-supervised learning approaches rely on a Transformer architecture for
-the representation model. In \cite{AnalyzeAttention} the attention patterns in
+the representation model. In \cite{yang_understanding_2020} the attention patterns in
 generatively trained Transformer representation models were analyzed.
@@
 \subsection{Training criterion}
-In \cite{chung2021}, representations based on different training criteria
+In \cite{chung_similarity_2021}, representations based on different training criteria
 (masked predictive coding, contrastive predictive coding, and autoregressive
@@
 
-A similar insight was obtained in \cite{zhou2020}, which compared vq-vae and
+A similar insight was obtained in \cite{zhou_comparison_2020}, which compared vq-vae and
 vq-wav2vec with respect to their ability to discover phonetic units.
@@
 Several studies have demonstrated better downstream performance when using
-larger datasets~\cite{rivi20, kawakami_learning_2020, chen_wavlm_2021}. For
+larger datasets~\cite{riviere_unsupervised_2020, kawakami_learning_2020, chen_wavlm_2021}. For
 example,
@@
 
-\edit{The relation between model sizes and downstream performances have also been investigated \cite{pue2021scaling,versteegh2015zero}.}
+\edit{The relation between model sizes and downstream performances have also been investigated \cite{pu_scaling_2021,versteegh_zero_2015}.}
 Using the Mockingjay 
-model~\cite{liu_mockingjay_2020}, the authors in \cite{pue2021scaling} attempt to establish a relationship between
+model~\cite{liu_mockingjay_2020}, the authors in \cite{pu_scaling_2021} attempt to establish a relationship between
 model size and self-supervised \ensuremath{L_1} loss and demonstrate that it approximately
@@
 Similar problems were also investigated using HuBERT and wav2vec~2.0 models in
-\cite{chang2021exploration}.
-In \cite{robustw2v2} domain effects were studied in greater detail using 
+\cite{chang_exploration_2021}.
+In \cite{hsu_robust_2021} domain effects were studied in greater detail using 
 datasets from six different domains. In particular, the authors focused on the
@@
 
-In \cite{rivi20} the effectiveness of CPC-trained representations for
+In \cite{riviere_unsupervised_2020} the effectiveness of CPC-trained representations for
 phone discrimination tasks was compared across several languages. It
@@
 learn contextualized speech features that generalize across different
-languages. In \cite{conneau20xling}, a wav2vec~2.0 model was trained on data
+languages. In \cite{conneau_unsupervised_2020}, a wav2vec~2.0 model was trained on data
 from multiple different languages and different corpora (Babel, Common Voice,
@@
 from different layers of a self-supervised models also generalizes to the
-multilingual setting. Experiments in \cite{baevski2021unsupervised} on phone
+multilingual setting. Experiments in \cite{baevski_unsupervised_2021} on phone
 recognition in eight languages based on the different layers of the

### 2023-04-12 chapters/paper_review/05_zero.tex
Update reference keys for review paper

@@
 \centering
-    {\renewcommand*\arraystretch{1.4}}
 \caption{
@@
 	All speech and text are in English if not specified. 
-\textcolor{black}{
-The references in the table are sorted according to the date of publication.}
-}
+	\textcolor{black}{The references in the table are sorted according to the date of publication.}}
 	\label{fig:unsupervisedASR}
-\resizebox{0.99\textwidth}{!}{
+	\resizebox*{0.99\textwidth}{!}{%
+	{\renewcommand*\arraystretch{1.4}}
 	\begin{tabular}{cllllll}
 	\toprule
-Reference & Speech representation & Speech segmentation  & Text token/representation & Mapping approach & Refinement  & Results  \\ 
+	Reference & Speech representation & Speech segmentation  & Token & Mapping approach & Refinement  & Results  \\ 
 	\midrule
-\cite{liu2018completely}& Audio word2vec~\cite{Wang2018SegAudioWord2Vec} & Oracle & Phoneme & \edit{Adversarial Training~\cite{gulrajani2017improved}} & - & TIMIT (PER): 63.6\%  \\ %Submitted on 1 Apr 2018
+	\cite{liu_completely_2018}& Audio word2vec~\cite{wang_segmental_2018} & Oracle & Phoneme & \edit{Adversarial Training~\cite{gulrajani_improved_2017}} & - & TIMIT (PER): 63.6\%  \\ %Submitted on 1 Apr 2018
 	\midrule
-  \cite{chung2018unsupervised} & Speech2vec~\cite{chung_speech2vec_2018} & BES-GMM~\cite{kamper_segmental_2017} &  Word2Vec & \edit{Adversarial Training~\cite{conneau2017word}} & Self-training & SWC (Acc): 10.9\%  \\ %Submitted on 18 May 2018
+	\cite{chung_unsupervised_2018} & Speech2vec~\cite{chung_speech2vec_2018} & BES-GMM~\cite{kamper_segmental_2017} &  Word2Vec & \edit{Adversarial Training~\cite{conneau_word_2018}} & Self-training & SWC (Acc): 10.9\%  \\ %Submitted on 18 May 2018
 	%https://arxiv.org/pdf/1805.07467.pdf %The text embeddings were obtained by training Word2Vec on the transcriptions using the fastText implementation without subword information [3].
@@
 
-    \cite{chung2019towards} & \tabincell{l}{Speech2vec\\(English)} & Oracle & \tabincell{c}{Word2Vec\\(French)} & VecMap~\cite{artetxe-etal-2018-robust} & \tabincell{l}{LM rescore,\\sequence DAE} & ST (BLEU): 10.8\%   \\ %[Submitted on 4 Nov 2018]
+	\cite{chung_unsupervised_2019a} & \tabincell{l}{Speech2vec\\(English)} & Oracle & \tabincell{c}{Word2Vec\\(French)} & VecMap~\cite{artetxe_robust_2018} & \tabincell{l}{LM rescore,\\sequence DAE} & ST (BLEU): 10.8\%   \\ %[Submitted on 4 Nov 2018]
 	\midrule
 
-\cite{yeh2018unsupervised} & MFCC & GAS~\cite{wang2017gate} & Phoneme & Empirical-ODM~\cite{Liu2017ODM} & Self-training &  TIMIT (PER):  41.6\%  \\ %Submitted on 23 Dec 2018
+	\cite{yeh_unsupervised_2019} & MFCC & GAS~\cite{wang_gate_2017} & Phoneme & Empirical-ODM~\cite{liu_unsupervised_2017} & Self-training &  TIMIT (PER):  41.6\%  \\ %Submitted on 23 Dec 2018
 	\midrule
 
-\cite{chen2019completely} & MFCC & GAS  & Phoneme & \edit{Adversarial Training~\cite{gulrajani2017improved}} & Self-training &  TIMIT (PER):  33.1\%   \\
+	\cite{chen_completely_2019} & MFCC & GAS  & Phoneme & \edit{Adversarial Training~\cite{gulrajani_improved_2017}} & Self-training &  TIMIT (PER):  33.1\%   \\
 	\midrule
 
-\cite{baevski2021unsupervised} & Wav2vec 2.0~\cite{baevski_wav2vec_2020} & \kmeans{} & Phoneme &\edit{Adversarial Training~\cite{gulrajani2017improved}} & Self-training & \tabincell{l}{TIMIT (PER):  18.6\%, \\ LibriSpeech (WER): 5.9\%}   \\
+	\cite{baevski_unsupervised_2021} & Wav2vec 2.0~\cite{baevski_wav2vec_2020} & $k$-means & Phoneme &\edit{Adversarial Training~\cite{gulrajani_improved_2017}} & Self-training & \tabincell{l}{TIMIT (PER):  18.6\%, \\ LibriSpeech (WER): 5.9\%}   \\
 	\midrule
 
-\textcolor{black}{\cite{Deciphering2022}} & \textcolor{black}{\tabincell{l}{Universal Phone \\ Recogniser}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Decipherment~\cite{ravi-knight-2011-deciphering}} & \textcolor{black}{Self-training} & \textcolor{black}{\tabincell{l}{GlobalPhone: 32.5\% to just 1.9\% \\ worse than supervised models} }  \\ %ubmitted on 12 Nov 2021 %m 32.5% to just 1.9% absolute worse than the equivalent fully supervised models trained on the same data
+	\textcolor{black}{\cite{klejch_deciphering_2022}} & \textcolor{black}{\tabincell{l}{Universal Phone \\ Recogniser}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Decipherment~\cite{ravi_deciphering_2011}} & \textcolor{black}{Self-training} & \textcolor{black}{\tabincell{l}{GlobalPhone: 32.5\% to just 1.9\% \\ worse than supervised models} }  \\ %ubmitted on 12 Nov 2021 %m 32.5% to just 1.9% absolute worse than the equivalent fully supervised models trained on the same data
 	\midrule
-\textcolor{black}{\cite{w2v-u2}} & \textcolor{black}{Wav2vec 2.0~\cite{baevski_wav2vec_2020} }& - & \textcolor{black}{Phoneme} & \textcolor{black}{Adversarial Training~\cite{gulrajani2017improved}} & \textcolor{black}{Self-training} & \textcolor{black}{LibriSpeech (WER): 6.3\%}   \\
+	\textcolor{black}{\cite{liu_endtoend_2023}} & \textcolor{black}{Wav2vec 2.0~\cite{baevski_wav2vec_2020} }& - & \textcolor{black}{Phoneme} & \textcolor{black}{Adversarial Training~\cite{gulrajani_improved_2017}} & \textcolor{black}{Self-training} & \textcolor{black}{LibriSpeech (WER): 6.3\%}   \\
 	\midrule
 
-\textcolor{black}{\cite{w2v-u2}} & \textcolor{black}{Wav2vec 2.0~\cite{baevski_wav2vec_2020}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Adversarial Training~\cite{gulrajani2017improved}} & \textcolor{black}{Self-training} & \textcolor{black}{LJSpeech (WER): 64.0\%}  \\
+	\textcolor{black}{\cite{liu_endtoend_2023}} & \textcolor{black}{Wav2vec 2.0~\cite{baevski_wav2vec_2020}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Adversarial Training~\cite{gulrajani_improved_2017}} & \textcolor{black}{Self-training} & \textcolor{black}{LJSpeech (WER): 64.0\%}  \\
 
@@
 	\bottomrule
-\end{tabular}
-}
+	\end{tabular}}
 \end{table*}
@@
 
-Adversarial training~\cite{goodfellow2014generative,arjovsky2017wasserstein, gulrajani2017improved} is one common way to learn such a 
+Adversarial training~\cite{goodfellow_generative_2014,arjovsky_wasserstein_2017, gulrajani_improved_2017} is one common way to learn such a 
 mapping function. 
@@
 \edit{
-There is a large amount of work using gradient penalty in the objective of training discriminators~\cite{liu2018completely,chen2019completely,baevski2021unsupervised,w2v-u2}, which is inspired by Improved Wasserstein Generative Adversarial Network (WGAN)~\cite{gulrajani2017improved}.
+There is a large amount of work using gradient penalty in the objective of training discriminators~\cite{liu_completely_2018,chen_completely_2019,baevski_unsupervised_2021,liu_endtoend_2023}, which is inspired by Improved Wasserstein Generative Adversarial Network (WGAN)~\cite{gulrajani_improved_2017}.
 }
 \textcolor{black}{
-Other ways to map speech and text include via segmental empirical output distribution matching (segmental empirical-ODM)~\cite{yeh2018unsupervised} and decipherment algorithm~\cite{Deciphering2022}.}
+Other ways to map speech and text include via segmental empirical output distribution matching (segmental empirical-ODM)~\cite{yeh_unsupervised_2019} and decipherment algorithm~\cite{klejch_deciphering_2022}.}
 
@@
 Success in unsupervised neural machine translation
-(MT)~\cite{artetxe2017unsupervised, conneau2017word, lample2017unsupervised}
+(MT)~\cite{artetxe_unsupervised_2018, conneau_word_2018, lample_unsupervised_2018}
 has inspired innovative exploration of various unsupervised ASR algorithms.
@@
 phoneme-level
-segments~\cite{liu2018completely,chen2019completely,yeh2018unsupervised}. 
+segments~\cite{liu_completely_2018,chen_completely_2019,yeh_unsupervised_2019}. 
 A representation vector or a token then represents each phoneme-level segment. 
@@
 or token to the correct phoneme when the representation or token is highly correlated to the phonemes.
-Wav2vec-U~\cite{baevski2021unsupervised} selects the input feature from different layers of wave2vec~2.0~\cite{baevski_wav2vec_2020}.
+Wav2vec-U~\cite{baevski_unsupervised_2021} selects the input feature from different layers of wave2vec~2.0~\cite{baevski_wav2vec_2020}.
 The selection criterion is based on analysis of the phonetic information in each layer.
 \textcolor{black}{
-If a universal phone recognizer trained from a diverse set of languages is available, it is another way to transcribe speech into phone-level tokens~\cite{Deciphering2022}.}
+If a universal phone recognizer trained from a diverse set of languages is available, it is another way to transcribe speech into phone-level tokens~\cite{klejch_deciphering_2022}.}
 Another series of work is to transform a word into a word embedding. 
-\cite{chung2018unsupervised,chung2019towards} use adversarial training to map the word-level
+\cite{chung_unsupervised_2018,chung_unsupervised_2019a} use adversarial training to map the word-level
 speech embedding space~\cite{chung_speech2vec_2018} to the word embedding space
@@
 \TNS{split into 2 sentences, also have a connection to the previous para}
-\cite{liu2018completely} is the first work to successfully achieve unsupervised
+\cite{liu_completely_2018} is the first work to successfully achieve unsupervised
 phoneme recognition by first clustering phoneme-level
-embeddings~\cite{chung_audio_2016,Wang2018SegAudioWord2Vec} into a set of
+embeddings~\cite{chung_audio_2016,wang_segmental_2018} into a set of
 tokens and then using GAN to learn the mapping relationship between tokens and
 phonemes. 
-However, in \cite{liu2018completely}, oracle phoneme segmentation is still
+However, in \cite{liu_completely_2018}, oracle phoneme segmentation is still
 required to get reasonable performance. 
-\cite{chen2019completely,yeh2018unsupervised} uses gate activation signals
-(GAS)~\cite{wang2017gate} to get the phoneme-level segmentation, which is
+\cite{chen_completely_2019,yeh_unsupervised_2019} uses gate activation signals
+(GAS)~\cite{wang_gate_2017} to get the phoneme-level segmentation, which is
 unsupervised. 
-%However, the loss term in~\cite{yeh2018unsupervised} includes an empirical average over all segments inside a logarithmic function, which can be biased if we sample this empirical average by a mini-batch average.
+%However, the loss term in~\cite{yeh_unsupervised_2019} includes an empirical average over all segments inside a logarithmic function, which can be biased if we sample this empirical average by a mini-batch average.
 \cite{liu2017unsupervised} therefore use an extremely large batch size during training to reduce the biasing problem.
-%These problems are handled by another work~\cite{chen2019completely}.
-%\cite{chen2019completely} removes the quantization step.
-%Compared to~\cite{yeh2018unsupervised}, the discriminator in~\cite{chen2019completely} considers all possible information from the text dataset, not limited to the n-gram local statistics.
-%Besides, \cite{chen2019completely} works well with a reasonable batch size, which makes it feasible when the computational resource is limited.
-Wav2vec-U~\cite{baevski2021unsupervised} selects the input feature from
+%These problems are handled by another work~\cite{chen_completely_2019}.
+%\cite{chen_completely_2019} removes the quantization step.
+%Compared to~\cite{yeh_unsupervised_2019}, the discriminator in~\cite{chen_completely_2019} considers all possible information from the text dataset, not limited to the n-gram local statistics.
+%Besides, \cite{chen_completely_2019} works well with a reasonable batch size, which makes it feasible when the computational resource is limited.
+Wav2vec-U~\cite{baevski_unsupervised_2021} selects the input feature from
 different layers of wave2vec 2.0~\cite{baevski_wav2vec_2020}, a self-supervised
@@
 %Should we mention this?
-Wav2vec-U uses \kmeans{} to cluster the selected features, and the boundaries
+Wav2vec-U uses $k$-means to cluster the selected features, and the boundaries
 are drawn whenever the clustered index changes.
@@
 
-%In \cite{chung2018unsupervised, chung2019towards}, the word boundaries are also generated automatically with speech only. 
-%However, \cite{chung2018unsupervised, chung2019towards} need oracle boundaries, therefore not completely unsupervised as the framework proposed in this paper.
+%In \cite{chung_unsupervised_2018, chung_unsupervised_2019a}, the word boundaries are also generated automatically with speech only. 
+%However, \cite{chung_unsupervised_2018, chung_unsupervised_2019a} need oracle boundaries, therefore not completely unsupervised as the framework proposed in this paper.
 
@@
 %Using performance to end this section.
-Wav2vec-U~\cite{baevski2021unsupervised} achieved state-of-the-art results at the time, which suggests that representation learning is
+Wav2vec-U~\cite{baevski_unsupervised_2021} achieved state-of-the-art results at the time, which suggests that representation learning is
 essential for the success of unsupervised ASR.
@@
 \textcolor{black}{
-And wav2vec-U 2.0~\cite{w2v-u2} further removes the requirement of the segmentation stage, so the unsupervised ASR model can be learned in an end-to-end style.}
+And wav2vec-U 2.0~\cite{liu_endtoend_2023} further removes the requirement of the segmentation stage, so the unsupervised ASR model can be learned in an end-to-end style.}
 The robustness of wav2vec-U was further analyzed with respect to 
 domain-mismatch scenarios in which the domains of unpaired
-speech and text were different~\cite{lin2021analyzing}.
+speech and text were different~\cite{lin_analyzing_2022}.
 Experimental results showed that domain mismatch leads to inferior performance, but a representation model pre-trained on the targeted speech domain extracts better representations and reduces this drop in performance.
@@
 data.
-Therefore, wav2vec-U~\cite{baevski2021unsupervised} develops a new approach to
+Therefore, wav2vec-U~\cite{baevski_unsupervised_2021} develops a new approach to
 determine the hyperparameters without a labeled development set. 
@@
 
-The ASR-TTS framework can jointly optimize both ASR and TTS \edit{without using paired data}~\cite{tjandra2017listening,hori2019cycle,wang2020improving}.
-A speech chain~\cite{tjandra2017listening,tjandra2018machine} is one 
+The ASR-TTS framework can jointly optimize both ASR and TTS \edit{without using paired data}~\cite{tjandra_listening_2017,hori_cycleconsistency_2019,wang_improving_2020}.
+A speech chain~\cite{tjandra_listening_2017,tjandra_machine_2018} is one 
 successful way to utilize audio-only and text-only data to train both
@@
 Although this approach initially freezes either the ASR or TTS model, 
-extensions of this study~\cite{hori2019cycle,tjandra2019end,baskar2019semi}
+extensions of this study~\cite{hori_cycleconsistency_2019,tjandra_endtoend_2019,baskar_semisupervised_2019}
 implement the joint training of both ASR and TTS parameters using 
-REINFORCE~\cite{williams1992simple} and straight-through estimators.
+REINFORCE~\cite{williams_simple_1992} and straight-through estimators.
 
@@
 We combine the ASR and language model via a noisy channel 
-model~\cite{jelinek1997statistical}, a weighted finite state 
-transducer~\cite{mohri2002weighted}, or shallow 
-fusion~\cite{gulcehre2015using,chorowski2016towards}.
+model~\cite{jelinek_statistical_1997}, a weighted finite state 
+transducer~\cite{mohri_weighted_2002}, or shallow 
+fusion~\cite{gulcehre_using_2015,chorowski_better_2017}.
 However, the progress of TTS systems boosted by deep 
-learning~\cite{oord_wavenet_2016,shen2018natural} has inspired another interesting and
+learning~\cite{oord_wavenet_2016,shen_natural_2018} has inspired another interesting and
 straightforward research direction: \edit{artificially creating paired text and audio
@@
 generate the waveform with various acoustic 
-variations~\cite{li2018training,ueno2019multi,rosenberg2019speech,laptev2020you,huang2020using}.
+variations~\cite{li_training_2018,ueno_multispeaker_2019,rosenberg_speech_2019,laptev_you_2020,huang_using_2020}.
 The other approaches are based on the generation of high-level (more
 linguistic) features instead of generating the waveform, e.g., encoder
-features~\cite{hayashi2018back} and phoneme 
-features~\cite{renduchintala18_interspeech,masumura20_interspeech}.
-This approach is similar to the back-translation technique developed in \edit{neural machine translation}~\cite{sennrich2016improving}.
+features~\cite{hayashi_backtranslationstyle_2018} and phoneme 
+features~\cite{renduchintala_multimodal_2018,masumura_phonemetographeme_2020}.
+This approach is similar to the back-translation technique developed in \edit{neural machine translation}~\cite{sennrich_improving_2016}.
 One benefit of the above data generation approaches is that it can be used to feed
@@
 linguistic unit discovery from speech data without linguistic resources,
-e.g., transcriptions and other annotations~\cite{jansen2013summary}.
+e.g., transcriptions and other annotations~\cite{jansen_summary_2013}.
 This study was motivated by unsupervised query-by-example, applications of
@@
 \begin{itemize}
-	 \item \edit{Zero Resource Speech Challenge 2015~\cite{versteegh2015zero} mainly focused on building an acoustic model without using any
+	 \item \edit{Zero Resource Speech Challenge 2015~\cite{versteegh_zero_2015} mainly focused on building an acoustic model without using any
 	 linguistic annotations based on subword unit modeling and spoken term
@@
 	 units from unsegmented raw speech.
-	 \item The Zero Resource Speech Challenge 2021~\cite{nguyen2020zero}, the latest
+	 \item The Zero Resource Speech Challenge 2021~\cite{nguyen_zero_2020}, the latest
 	 \edit{challenge}, expanded the scope to include language modeling tasks.
@@
 techniques have shifted from purely generative 
-models~\cite{ondel2016variational,heck2017feature}, deep 
-autoencoders~\cite{tjandra19_interspeech,chorowski_unsupervised_2019}, and incorporation of
+models~\cite{ondel_variational_2016,heck_feature_2017}, deep 
+autoencoders~\cite{tjandra_vqvae_2019,chorowski_unsupervised_2019}, and incorporation of
 neural-network-based TTS/VC techniques~\cite{vanniekerk_vectorquantized_2020} to
-self-supervised learning~\cite{maekaku21_interspeech}.
+self-supervised learning~\cite{maekaku_speech_2021}.
 The latest challenge included the visual modality, continuing the
@@
 directly from audio, bypassing the need for text or 
-labels~\cite{nguyen_generative_2022,polyak_speech_2021,kharitonov_textfree_2021,emotion_conversion}. 
+labels~\cite{nguyen_generative_2022,polyak_speech_2021,kharitonov_textfree_2021,kreuk_textless_2022}. 
 Not only does this open the gate for language and dialect modeling without
@@
 emotion, hesitation, interruptions. 
-The generative spoken language model (nguyen_generative_2022)~\cite{nguyen_generative_2022} utilizes discrete
+The generative spoken language model (GSLM)~\cite{nguyen_generative_2022} utilizes discrete
 representations from wav2vec~2.0, HuBERT, and CPC algorithms as inputs to an
@@
 model and generate
-dialogues~\cite{dialogue_nguyen_generative_2022}.\footnote{https://speechbot.github.io/dgslm/}
+dialogues~\cite{nguyen_generative_2022}.\footnote{https://speechbot.github.io/dgslm/}
 %\cite{}. 
-Given its flexibility in modeling spoken content, the nguyen_generative_2022 has been further extended
-to jointly model content and prosody~\cite{kharitonov_textfree_2021}. This prosodic-nguyen_generative_2022 model
+Given its flexibility in modeling spoken content, the GSLM has been further extended
+to jointly model content and prosody~\cite{kharitonov_textfree_2021}. This prosodic-GSLM model
 introduced a multistream causal Transformer, where the input and output layers
 use multiple heads to model three channels:  discrete speech units, duration,
-and quantized pitch. The prosodic-nguyen_generative_2022 model jointly generates novel content
+and quantized pitch. The prosodic-GSLM model jointly generates novel content
 and prosody congruently in the expressive 
@@
 %\cite{}. 
-Going one step further, \cite{emotion_conversion} used a speech emotion
+Going one step further, \cite{kreuk_textless_2022} used a speech emotion
 conversion framework to modify the perceived emotion of a speech utterance
@@
 extended the idea of textless language processing or audio discrete representation to applications such as 
-spoken question answering~\cite{spoken_qa_dual}, speech separation~\cite{shi2021discretization}, TTS~\cite{hayashi2020discretalk}, and speech-to-speech
-translation~\cite{textless_translation}.
+spoken question answering~\cite{lin_dual_2022}, speech separation~\cite{shi_discretization_2022}, TTS~\cite{hayashi_discretalk_2020}, and speech-to-speech
+translation~\cite{lee_textless_2022}.

### 2023-04-12 chapters/paper_review/06_conclusions.tex
Update reference keys for review paper

@@
 \edit{Some efficient methods for leveraging SSL models exist in the NLP community. 
-Adapters~\cite{pmlr-v97-houlsby19a,zaken2021bitfit,guo-etal-2021-parameter} are lightweight modules inserted into SSL models, and in downstream tasks, the parameters of SSL models are frozen, and only the adapters are trained.
+Adapters~\cite{houlsby_parameterefficient_2019,zaken_bitfit_2022,guo_parameterefficient_2021} are lightweight modules inserted into SSL models, and in downstream tasks, the parameters of SSL models are frozen, and only the adapters are trained.
 The prompt/instruction learning methods~\cite{liu_pretrain_2021} also freeze the SSL parameters and control the output of SSL by adding additional information, which is called \textit{prompt}, in the input.
-Both adapter-based methods and prompt/instruction learning yield competitive performance compared with fine-tuning in NLP applications, but there is only little related work for speech~\cite{SpeechAdapter,SpeechPrompt}.
+Both adapter-based methods and prompt/instruction learning yield competitive performance compared with fine-tuning in NLP applications, but there is only little related work for speech~\cite{thomas_efficient_2022,chang_speechprompt_2022}.
 In addition, prompt for speech SSL does not achieve comparable performance on sequence generation tasks like phoneme recognition and slot filling, so how to use prompt is still an open question.}
@@
 This} makes them unsuitable for edge devices \edit{but also limits the ability to scale these models to very large datasets} \ci{ -- and leads to a large energy consumption}.
-Preliminary studies have been conducted on compressing speech representation models through network pruning~\cite{PARP} or knowledge distillation~\cite{chang2021distilhubert}. 
+Preliminary studies have been conducted on compressing speech representation models through network pruning~\cite{lai_parp_2021} or knowledge distillation~\cite{chang_distilhubert_2021}. 
 \edit{There has been quite some effort towards more efficient general neural network models via conditional computing \cite{bengio_conditional_2016} and neural network quantization \cite{gholami_survey_2021} as well as extensive work on improving the specific efficiency of Transformer models, especially with the focus on self-attention \cite{tay_efficient_2022}, but these technology has not been widely used in speech SSL.}
@@
 But we have not been aware of any publication in this direction when writing this paper.}
-On the other hand, non-streaming architectures in models such as the bidirectional Transformer have hindered the representation model used in streaming scenarios, leading to studies that address these problems~\cite{StreamingW2v}.
+On the other hand, non-streaming architectures in models such as the bidirectional Transformer have hindered the representation model used in streaming scenarios, leading to studies that address these problems~\cite{cao_improving_2021}.
 We anticipate research in these directions to continue in the future.
@@
 For example, we can decompose the representation into content embedding and speaker embedding and use content embedding for ASR and speaker embedding for SID.
-Some work has been in this direction~\cite{pmlr-v162-qian22b,NANCY,MultimodelAmazonIS22}.
+Some work has been in this direction~\cite{qian_contentvec_2022,choi_neural_2021,chan_contentcontext_2022}.
 
-\item \textbf{Creating robust models.} \edit{As discussed in \cref{subsec:robustness}, studies have been conducted on the robustness of representation models \cite{wu2021characterizing}.
+\item \textbf{Creating robust models.} \edit{As discussed in \cref{subsec:robustness}, studies have been conducted on the robustness of representation models \cite{wu_characterizing_2022}.
 However, the failure modes of SSL models are still poorly understood, and it remains unclear whether they provide more or less robustness to adversarial attacks than fully supervised models.
-Due to the importance of this research direction, while writing this paper, there is already some related research about enhancing the robustness of SSL models~\cite{robustw2v2,huang2022improving,wang2022improving,zhu2022noise} and identifying their vulnerability to adversarial attack~\cite{wu2021characterizing}.
+Due to the importance of this research direction, while writing this paper, there is already some related research about enhancing the robustness of SSL models~\cite{hsu_robust_2021,huang_improving_2022,wang_improving_2022,zhu_noiserobust_2022} and identifying their vulnerability to adversarial attack~\cite{wu_characterizing_2022}.
 }
 
-\edit{\item \textbf{Capturing higher-level semantic information.} Although many representation learning approaches can go beyond low-level phonetic modeling to capture some lexical information~\cite{TuAnh_2022_discrete}, they still struggle in higher-level semantic tasks easily captured by word-level counterparts like BERT.
-One workaround is two-stage training~\cite{kharitonov_textfree_2021, dialogue_nguyen_generative_2022}; however, this prevents propagating rich lexical and semantic knowledge modeled in the second stage to benefit the phonetically focused first stage.
-}
+\item \textbf{Capturing higher-level semantic information.} Although many representation learning approaches can go beyond low-level phonetic modeling to capture some lexical information~\cite{nguyen_are_2022}, they still struggle in higher-level semantic tasks easily captured by word-level counterparts like BERT.
+One workaround is two-stage training~\cite{kharitonov_textfree_2021, nguyen_generative_2022}; however, this prevents propagating rich lexical and semantic knowledge modeled in the second stage to benefit the phonetically focused first stage.
 
 \item \textbf{Using text representation models to improve speech representation.} The amount of content information in speech corpora used to train speech representation models is far less than that of text representation models.
-Noting that the BERT training corpus exceeds 3 billion words~\cite{devlin_bert:_2018}, and assuming a typical speaking rate of 120 words per minute, a speech corpus containing the same content as the BERT training data would include 400,000 hours of audio, which exceeds the \edit{accumulated} training data of all current speech representation models.
+Noting that the BERT training corpus exceeds 3 billion words~\cite{devlin_bert_2018}, and assuming a typical speaking rate of 120 words per minute, a speech corpus containing the same content as the BERT training data would include 400,000 hours of audio, which exceeds the \edit{accumulated} training data of all current speech representation models.
 Therefore, to enable speech representation models to better learn human language, for instance by extracting semantic information from acoustic signals, the use of text models such as BERT and GPT  \edit{seems} key: nevertheless, how to use these to improve speech representation model pre-training remains an open question.

### 2023-04-12 chapters/paper_review/model.tex
Update reference keys for review paper

@@
 
-wav2vec \cite{schneider_wav2vec:_2019}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
+wav2vec \cite{schneider_wav2vec_2019}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
 
@@
 
-DeCoAR \cite{ling2020deep}                & 2019 Dec. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 3/12-19
+DeCoAR \cite{ling_deep_2020}                & 2019 Dec. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 3/12-19
 
-PASE+ \cite{ravanelli_multi-task_2020}                & 2020 Jan. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 25/1 % I think contrastive always have learned target.
+PASE+ \cite{ravanelli_multitask_2020}                & 2020 Jan. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 25/1 % I think contrastive always have learned target.
 
@@
 
-speech-XLNet \cite{song_speech-xlnet_2020} & 2019 Oct.
+speech-XLNet \cite{song_speechxlnet_2020} & 2019 Oct.
 & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\%23 Oct 2019 %Lee: It is speech version of XLNet.
@@
 
-BEST-RQ \cite{BEST-RQ} & 2022 Feb.  & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ %Lee: no learned target. The target is random ...
+BEST-RQ \cite{chiu_selfsupervised_2022} & 2022 Feb.  & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ %Lee: no learned target. The target is random ...
 % 3 Feb 2022
@@
 APC \cite{chung_unsupervised_2019}          & 2019 Oct. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 5/4
-wav2vec \cite{schneider_wav2vec:_2019}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
+wav2vec \cite{schneider_wav2vec_2019}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
 %VQ-APC \cite{chung2020vector}             & 2020 May  & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ 
 Mockingjay  \cite{liu_mockingjay_2020}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 25/10-19
-%DeCoAR \cite{ling2020deep}                & 2019 Dec. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 3/12-19
+%DeCoAR \cite{ling_deep_2020}                & 2019 Dec. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 3/12-19
 %PASE+ \cite{Ravanelli2020}                & 2020 Jan. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 25/1

### 2023-04-12 chapters/paper_review/pretext.tex
Update reference keys for review paper

@@
     \midrule 
-    PASE~\cite{pascual_learning_2019}, PASE+~\cite{ravanelli_multi-task_2020}\footnote{PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
+    PASE~\cite{pascual_learning_2019}, PASE+~\cite{ravanelli_multitask_2020}\footnote{PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
     \midrule 
@@
     \midrule
-    Speech-XLNet \cite{song_speech-xlnet_2020}     & \multicolumn{2}{c|}{$X_{\mathcal{P}_{t}}$}   &     $x_{i\sim\mathcal{P}^c_{t}}$  \\ %Lee: Use * to represent the permutation, hope it is not strange.
+    Speech-XLNet \cite{song_speechxlnet_2020}     & \multicolumn{2}{c|}{$X_{\mathcal{P}_{t}}$}   &     $x_{i\sim\mathcal{P}^c_{t}}$  \\ %Lee: Use * to represent the permutation, hope it is not strange.
     \midrule  
-    DeCoAR~\cite{ling2020deep}     & $X_{[1,t-1]}, X_{[t+k+1,T]}$ & \textsc{-} & $X_{[t,t+k]}$   \\
+    DeCoAR~\cite{ling_deep_2020}     & $X_{[1,t-1]}, X_{[t+k+1,T]}$ & \textsc{-} & $X_{[t,t+k]}$   \\
     \midrule
@@
     \midrule 
-    CPC~\cite{oord_representation_2018}, wav2vec \cite{schneider_wav2vec:_2019}, Modified CPC \cite{riviere_unsupervised_2020}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
+    CPC~\cite{oord_representation_2018}, wav2vec \cite{schneider_wav2vec_2019}, Modified CPC \cite{riviere_unsupervised_2020}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
     \midrule 
@@
     \midrule 
-    BEST-RQ \cite{BEST-RQ}\footnote{BEST-RQ obtains codes $C$ by quantizing acoustic features using a random projection quantizer.}     &  \multicolumn{2}{c|}{$X_{-[t,t+k]}$}      &  $C_{[t,t+k]}$   \\
+    BEST-RQ \cite{chiu_selfsupervised_2022}\footnote{BEST-RQ obtains codes $C$ by quantizing acoustic features using a random projection quantizer.}     &  \multicolumn{2}{c|}{$X_{-[t,t+k]}$}      &  $C_{[t,t+k]}$   \\
     \midrule 

### 2023-04-12 appendices/appendix_hierarchical.tex
Proposal for indent for multiline equations

@@
 \log p(\xb) &= \log \int\int p(\xb|\zb_1)p(\zb_1|\zb_2)p(\zb_2) \text{d}\zb_1\text{d}\zb_2 \\
-&= \log \int\int \frac{q(\zb_2|\xb)}{q(\zb_2|\xb)} p(\xb|\zb_1)p(\zb_1|\zb_2)p(\zb_2) \text{d}\zb_1\text{d}\zb_2 \notag \\
-&= \log \int\int q(\zb_2|\xb) p(\zb_1|\zb_2) \frac{p(\xb|\zb_1)p(\zb_2)}{q(\zb_2|\xb)} \text{d}\zb_1\text{d}\zb_2 \notag \\
+&\qquad = \log \int\int \frac{q(\zb_2|\xb)}{q(\zb_2|\xb)} p(\xb|\zb_1)p(\zb_1|\zb_2)p(\zb_2) \text{d}\zb_1\text{d}\zb_2 \notag \\
+&\qquad = \log \int\int q(\zb_2|\xb) p(\zb_1|\zb_2) \frac{p(\xb|\zb_1)p(\zb_2)}{q(\zb_2|\xb)} \text{d}\zb_1\text{d}\zb_2 \notag \\
 %&\ge \mathbb{E}_{p(\zb_1|\zb_2)} \left [ \log q(\zb_2|\xb) \frac{p(\xb|\zb_1)p(\zb_2)}{q(\zb_2|\xb)} \right ] \notag \\
@@
     \log p(\xb) &= \log \int p(\xb|\zb)p(\zb) \text{d}\zb \\
-    &= \log \int \frac{q(\zb_{>k}|\xb)}{q(\zb_{>k}|\xb)} p(\xb|\zb)p(\zb) \text{d}\zb \notag \\
-    &= \log \int q(\zb_{>k}|\xb) p(\zb) \frac{p(\xb|\zb)}{q(\zb_{>k}|\xb)} \text{d}\zb \notag \\
-    &= \log \int q(\zb_{>k}|\xb) p(\zb_{\leq k}|\zb_{>k}) p(\zb_{>k}) \frac{p(\xb|\zb)}{q(\zb_{>k}|\xb)} \text{d}\zb \notag \\
-    &= \log \int q(\zb_{>k}|\xb) p(\zb_{\leq k}|\zb_{>k}) \frac{p(\xb|\zb)p(\zb_{>k})}{q(\zb_{>k}|\xb)} \text{d}\zb \notag \\
-    &\ge \mathbb{E}_{p(\zb_{\leq k}|\zb_{>k})} \left[ \log q(\zb_{>k}|\xb) \frac{p(\xb|\zb)p(\zb_{>k})}{q(\zb_{>k}|\xb)} \right] \notag \\
-    &\ge \mathbb{E}_{p(\zb_{\leq k}|\zb_{>k})q(\zb_{>k}|\xb)} \left[ \log \frac{p(\xb|\zb)p(\zb_{>k})}{q(\zb_{>k}|\xb)} \right] \equiv \mathcal{L}^{>k} \notag \ .
+    &\qquad = \log \int \frac{q(\zb_{>k}|\xb)}{q(\zb_{>k}|\xb)} p(\xb|\zb)p(\zb) \text{d}\zb \notag \\
+    &\qquad = \log \int q(\zb_{>k}|\xb) p(\zb) \frac{p(\xb|\zb)}{q(\zb_{>k}|\xb)} \text{d}\zb \notag \\
+    &\qquad = \log \int q(\zb_{>k}|\xb) p(\zb_{\leq k}|\zb_{>k}) p(\zb_{>k}) \frac{p(\xb|\zb)}{q(\zb_{>k}|\xb)} \text{d}\zb \notag \\
+    &\qquad = \log \int q(\zb_{>k}|\xb) p(\zb_{\leq k}|\zb_{>k}) \frac{p(\xb|\zb)p(\zb_{>k})}{q(\zb_{>k}|\xb)} \text{d}\zb \notag \\
+    &\qquad \ge \mathbb{E}_{p(\zb_{\leq k}|\zb_{>k})} \left[ \log q(\zb_{>k}|\xb) \frac{p(\xb|\zb)p(\zb_{>k})}{q(\zb_{>k}|\xb)} \right] \notag \\
+    &\qquad \ge \mathbb{E}_{p(\zb_{\leq k}|\zb_{>k})q(\zb_{>k}|\xb)} \left[ \log \frac{p(\xb|\zb)p(\zb_{>k})}{q(\zb_{>k}|\xb)} \right] \equiv \mathcal{L}^{>k} \notag \ .
 \end{align}
@@
     %     \log p(\xb) &\ge \mathbb{E}_{q(\zb_1, \zb_2|\xb)} \left[ \log \left(p(\xb|\zb_1) p(\zb_1|\zb_2) p(\zb_2) \right) \notag \\
-    %     &- \log \left( q(\zb_1|\xb) q(\zb_2|\zb_1)\right) \right]\ .
+    %     &\qquad - \log \left( q(\zb_1|\xb) q(\zb_2|\zb_1)\right) \right]\ .
     % \end{align}
@@
         \log p(\xb) &\ge \mathbb{E}_{q(\zb_1, \zb_2|\xb)} \left[ \log p(\xb|\zb_1) \right] \\
-    &+ \mathbb{E}_{q(\zb_1,\zb_2|\xb)} \left[ \log p(\zb_1|\zb_2) - \log q(\zb_1|\xb) \right] \notag \\
-    &+ \mathbb{E}_{q(\zb_1,\zb_2|\xb)} \left[ \log p(\zb_2) - \log q(\zb_2|\zb_1) \right] \notag \ .
+    &\qquad + \mathbb{E}_{q(\zb_1,\zb_2|\xb)} \left[ \log p(\zb_1|\zb_2) - \log q(\zb_1|\xb) \right] \notag \\
+    &\qquad + \mathbb{E}_{q(\zb_1,\zb_2|\xb)} \left[ \log p(\zb_2) - \log q(\zb_2|\zb_1) \right] \notag \ .
 \end{align}
@@
     \log p(\xb) &\ge \int \int \log p(\xb|\zb_1) \text{d}\zb_2\zb_1 \label{eq:2 layer VAE bottom up ELBO integrals} \\
-    &+ \int q(\zb_1|\xb) \int q(\zb_2|\zb_1) \log \frac{p(\zb_1|\zb_2)}{q(\zb_1|\xb)} \text{d}\zb_2\zb_1 \notag \\ 
-    &+ \int q(\zb_1|\xb) \int q(\zb_2|\zb_1) \log \frac{p(\zb_2)}{q(\zb_2|\zb_1)} \text{d}\zb_2\zb_1 \notag  \ .
+    &\qquad + \int q(\zb_1|\xb) \int q(\zb_2|\zb_1) \log \frac{p(\zb_1|\zb_2)}{q(\zb_1|\xb)} \text{d}\zb_2\zb_1 \notag \\ 
+    &\qquad + \int q(\zb_1|\xb) \int q(\zb_2|\zb_1) \log \frac{p(\zb_2)}{q(\zb_2|\zb_1)} \text{d}\zb_2\zb_1 \notag  \ .
 \end{align}
@@
     \log p(\xb) &\ge \mathbb{E}_{q(\zb_1, \zb_2|\xb)} \left[ \log p(\xb|\zb_1) \right] \\
-    &+ \mathbb{E}_{q(\zb_1|\xb)}\left[ \mathbb{E}_{q(\zb_2|\zb_1)}\left[\log \frac{p(\zb_1|\zb_2)}{q(\zb_1|\xb)}\right]\right] \notag \\
-    &+ \mathbb{E}_{q(\zb_1|\xb)}\left[-D_\text{KL}[q(\zb_2|\zb_1)||p(\zb_2)]\right] \notag \ .
+    &\qquad + \mathbb{E}_{q(\zb_1|\xb)}\left[ \mathbb{E}_{q(\zb_2|\zb_1)}\left[\log \frac{p(\zb_1|\zb_2)}{q(\zb_1|\xb)}\right]\right] \notag \\
+    &\qquad + \mathbb{E}_{q(\zb_1|\xb)}\left[-D_\text{KL}[q(\zb_2|\zb_1)||p(\zb_2)]\right] \notag \ .
 \end{align}
@@
         \log p(\xb) &\ge \mathbb{E}_{q(\zb_1, \zb_2|\xb)} \left[ \log p(\xb|\zb_1) \right] \\ 
-        &+ \mathbb{E}_{q(\zb_1,\zb_2|\xb)} \left[ \log p(\zb_2|\xb) - \log q(\zb_2|\xb) \right] \notag \\
-        &+ \mathbb{E}_{q(\zb_1,\zb_2|\xb)} \left[ \log p(\zb_1|\zb_2) - \log q(\zb_1|\zb_2) \right] \notag \ .
+        &\qquad + \mathbb{E}_{q(\zb_1,\zb_2|\xb)} \left[ \log p(\zb_2|\xb) - \log q(\zb_2|\xb) \right] \notag \\
+        &\qquad + \mathbb{E}_{q(\zb_1,\zb_2|\xb)} \left[ \log p(\zb_1|\zb_2) - \log q(\zb_1|\zb_2) \right] \notag \ .
     \end{align}
@@
         \log p(\xb) &\ge \int \int \log p(\xb|\zb_1) d\zb_1\zb_2 \\
-        &+ \int q(\zb_2|\xb) \log \frac{p(\zb_2|\xb)}{q(\zb_2|\xb)} d\zb_2 \notag \\ 
-        &+ \int q(\zb_2|\xb) \int q(\zb_1|\zb_2) \log \frac{p(\zb_1|\zb_2)}{q(\zb_1|\zb_2)} d\zb_1\zb_2 \notag \ .
+        &\qquad + \int q(\zb_2|\xb) \log \frac{p(\zb_2|\xb)}{q(\zb_2|\xb)} d\zb_2 \notag \\ 
+        &\qquad + \int q(\zb_2|\xb) \int q(\zb_1|\zb_2) \log \frac{p(\zb_1|\zb_2)}{q(\zb_1|\zb_2)} d\zb_1\zb_2 \notag \ .
     \end{align}
@@
         \log p(\xb) &\ge \mathbb{E}_{q(\zb_1, \zb_2|\xb)} \left[ \log p(\xb|\zb_1) \right] \\
-    &- D_\text{KL}[q(\zb_2|\xb)||p(\zb_2)] \notag \\ 
-    &- \mathbb{E}_{q(\zb_2|\xb)}\left[D_\text{KL}[q(\zb_1|\zb_2)||p(\zb_1|\zb_2)\right] \notag \ .
+    &\qquad - D_\text{KL}[q(\zb_2|\xb)||p(\zb_2)] \notag \\ 
+    &\qquad - \mathbb{E}_{q(\zb_2|\xb)}\left[D_\text{KL}[q(\zb_1|\zb_2)||p(\zb_1|\zb_2)\right] \notag \ .
 \end{align}

### 2023-04-25 appendices/appendix_benchmarking.tex
Add and revise short table and figure captions

@@
 
-\chapter[benchmarking generative latent variable models for speech]{Benchmarking Generative Latent Variable Models for Speech}
+\chapter[supplementary material for benchmarking generative latent variable models for speech]{Supplementary material for Benchmarking generative latent variable models for speech}
 
@@
     \vspace{2mm}
-    \caption{
+    \caption[Model likelihoods on TIMIT represented as a 16 bit linear PCM]{
     Model likelihoods on TIMIT represented as a 16 bit linear PCM. The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
@@
     \vspace{2mm}
-    \caption{
-    Model likelihoods on TIMIT represented as a \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}.
+    \caption[Model likelihoods on TIMIT represented as a 16 bit $\boldsymbol{\mu}$-law PCM]{
+    Model likelihoods on TIMIT represented as a 16 bit $\boldsymbol{\mu}$-law encoded PCM.
     }
@@
     \vspace{2mm}
-    \caption{
+    \caption[Model likelihoods on TIMIT represented as globally normalized 16 bit linear PCM]{
     Model likelihoods on TIMIT represented as globally normalized 16 bit linear PCM.
@@
     \vspace{2mm}
-    \caption{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
+    \caption[The highest possible Gaussian log-likelihoods attainable on the TIMIT test]{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
     \label{tab:timit best gaussian ll}
@@
     \end{tikzpicture}
-\caption{CW-VAE cell state $\vs_t^l$ update.
+\caption[CW-VAE cell state $\vs_t^l$ update]{CW-VAE cell state $\vs_t^l$ update.
 The cell state is given as $\vs_t^l=(\zb_t^l, \db_t^l)$ where $\db_t^l$ is the deterministic hidden state of a Gated Recurrent Unit \cite{cho_properties_2014}.
@@
     \end{subfigure}
-\caption{
+\caption[Generative and inference models for three-layered CW-VAE]{
     CW-VAE \cite{saxena_clockwork_2021} generative model $p(\xb,\zb)$ in (\subref{fig: cwvae three-layered graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: cwvae three-layered graphical model inference}) for a three-layered model with $k_1=1$ and $c=2$ giving $k_2=2$ and $k_3=4$ unrolled over eight steps in the observed variable.
@@
     \end{subfigure}
-\caption{
+\caption[Generative and inference models for three-layered STCN]{
     STCN \cite{aksan_stcn_2019} generative model $p(\xb,\zb)$ in (\subref{fig: stcn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: stcn graphical model inference}) for a single time-step.
@@
     \end{subfigure}
-\caption{
+\caption[Generative and inference models for VRNN]{
     VRNN \cite{chung_recurrent_2015} generative model $p(\xb,\zb)$ in (\subref{fig: vrnn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: vrnn graphical model inference}) unrolled over three steps in the observed variable.
@@
     \end{subfigure}
-\caption{
+\caption[Generative and inference models for SRNN]{
     SRNN \cite{fraccaro_sequential_2016} generative model $p(\xb,\zb)$ in (\subref{fig: srnn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: srnn graphical model inference}) unrolled over three steps in the observed variable.
@@
     \includegraphics[width=0.483\textwidth]{paper_benchmarking/1xrnjn5y_speaker_height_m_10_samples_knn_correct_fraction_model_vs_mel_50_neighbors_lda_subspace_2d.pdf}
-    \caption{
+    \caption[Leave-one-out $k$-nearest-neighbour accuracy for speaker gender and height]{
     Leave-one-out $k$-nearest-neighbor accuracy with different $k$ for
@@
      \end{subfigure}
-    \caption{Clustering of speaker gender in an one-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space and of a time-averaged mel spectrogram.
+    \caption[Clustering of speaker gender in 1D linear subspace of a CW-VAE latent space and a time-averaged mel spectrogram.]{Clustering of speaker gender in an one-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space and of a time-averaged mel spectrogram.
 The total overlap is slightly smaller in the subspace of the CW-VAE latent space and the separation between the distribution peaks is larger.}
@@
      \end{subfigure}
-    \caption{(a) Clustering of speaker height for male speakers and (b) speaker age for female speakers in an two-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space.}
+    \caption[Clustering of speaker gender in 2D linear subspace of a CW-VAE latent space]{(a) Clustering of speaker height for male speakers and (b) speaker age for female speakers in an two-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space.}
     \label{fig: cwvae latent height}
@@
     \end{subfigure}
-    \caption{Boxplots of the duration of the pronunciation of phonemes in TIMIT for a specific speaker DRW0 in (\subref{fig: timit validation set phoneme duration DRW0}) and globally in (\subref{fig: timit validation set phoneme duration global}).
+    \caption[Boxplots of duration of phoneme pronunciation in TIMIT]{Boxplots of the duration of the pronunciation of phonemes in TIMIT for a specific speaker DRW0 in (\subref{fig: timit validation set phoneme duration DRW0}) and globally in (\subref{fig: timit validation set phoneme duration global}).
 Not all phonemes are pronounced by speaker DRW0 over the course of their 10 test set sentences and hence they are missing from the x-axis compared to the global durations.}

### 2023-04-25 appendices/appendix_hierarchical.tex
Add and revise short table and figure captions

@@
 
-\chapter[hierarchical vaes know what they don't know]{Hierarchical VAEs Know What They Don't Know}
+\chapter[supplementary material for hierarchical vaes know what they don't know]{Supplementary material for Hierarchical VAEs know what they don't know}
 
@@
     % }
-    \caption{Selection of most important hyperparameters and their setting.
+    \caption[Selection of most important hyperparameters]{Selection of most important hyperparameters and their setting.
 Convolutional kernels are square and latent dimensions are given without spatial dimensions which are given in the text.
@@
     % }
-    \caption{Additional results for the HVAE model trained on SVHN. All results computed with 1000 importance samples.}
+    \caption[Additional results for the HVAE model trained on SVHN]{Additional results for the HVAE model trained on SVHN. All results computed with 1000 importance samples.}
     \label{tab:additional-results-svhn}
@@
     % }
-    \caption{Additional results for the HVAE model trained on CIFAR10. All results computed with 1000 importance samples.}
+    \caption[Additional results for the HVAE model trained on CIFAR10]{Additional results for the HVAE model trained on CIFAR10. All results computed with 1000 importance samples.}
     \label{tab:additional-results-cifar}
@@
     % }
-    \caption{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
+    \caption[Additional results for the HVAE model trained on FashionMNIST]{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
     \label{tab:additional-results-fashionmnist}
@@
     % }
-    \caption{Additional results for the HVAE model trained on MNIST. All results computed with 1000 importance samples.}
+    \caption[Additional results for the HVAE model trained on MNIST]{Additional results for the HVAE model trained on MNIST. All results computed with 1000 importance samples.}
     \label{tab:additional-results-mnist}

### 2023-04-25 chapters/paper_benchmarking.tex
Update reference keys for review paper
Add and revise short table and figure captions

@@
         \end{multicols}
-    \caption{
+    \caption[Generative models for an LSTM, the VRNN, SRNN, STCN, and CW-VAE models.]{
         Generative models for a single time step of a deterministic autoregressive LSTM, the VRNN and SRNN as well as the STCN and CW-VAE both with a single layer of latent variables. 
@@
 The selected models are all sequential deep latent variable models trained with variational inference and the reparameterization trick \cite{kingma_autoencoding_2014}.
-% The VRNN, SRNN, STCN and CW-VAE are all autoencoders trained with variational inference and the reparameterization trick \cite{kingma_auto-encoding_2014}. 
+% The VRNN, SRNN, STCN and CW-VAE are all autoencoders trained with variational inference and the reparameterization trick \cite{kingma_autoencoding_2014}. 
 They take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$. 
-% The VRNN, SRNN, STCN and CW-VAE are all VAEs \cite{kingma_auto-encoding_2014} that take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$.
+% The VRNN, SRNN, STCN and CW-VAE are all VAEs \cite{kingma_autoencoding_2014} that take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$.
 We let $\xb_{1:T}$ refer to the observed variable or a downsampled version of it. We will sometimes use $\xb$ to refer to the sequence $\xb_{1:T}$ when it is not ambiguous.
@@
     }
-\caption{
+\caption[Generative and inference models for a two-layered CW-VAE.]{
 Inference (left) and generative (right) models for the CW-VAE with a hierarchy of $L=2$ latent variables, $s_1=1$ and $s_2=2$. 
@@
     \end{minipage}%
-    \caption*{%
+    \caption*[Model likelihoods and phoneme error rate for TIMIT.]{%
     In \cref{tab: likelihoods timit} we report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines.
@@
     \end{adjustbox}
-    \caption{
+    \caption[Model likelihoods on LibriSpeech test sets.]{
     Model likelihoods $\mathcal{L}$ on LibriSpeech test sets represented as $\SI{16}{bit}$ $\mu$-law encoded PCM.
@@
     \includegraphics[width=0.425\textwidth]{paper_benchmarking/1xrnjn5y_speaker_phonemes_drw0_samples_100_knn_correct_fraction_model_vs_mel_30_neighbors_lda_subspace_5d.pdf}
-    \caption{
+    \caption[Clustering of phonemes in 2D subspace of CW-VAE latent space and KNN classification accuracy.]{
     (left) Clustering of phonemes in a 2D Linear Discriminant Analysis (LDA) subspace of a CW-VAE latent space ($\zb^{(1)}$).

### 2023-04-25 chapters/paper_hierarchical.tex
Add and revise short table and figure captions

@@
 
-Deep learning models are generally believed to form hierarchies of representations that range from low-level features to more conceptual ones related to semantics \parencite{bengio_representation_2013}.
-This has also been observed within deep generative models \parencite{maaloe_biva_2019, child_very_2021}.
+Deep learning models are generally believed to form hierarchies of representations that range from low-level features to more conceptual ones related to semantics \cite{bengio_representation_2013}.
+This has also been observed within deep generative models \cite{maaloe_biva_2019, child_very_2021}.
 %Recent research \cite{maaloe_biva_2019, child_very_2021, vahdat_nvae_2020}, has shown that deep latent variable models are able to learn rich hierarchical representations with increasing levels of abstraction and global structure towards the top of the hierarchy.
@@
     \vspace{0mm}
-    \caption[Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables]{Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables  \cite{maaloe_biva_2019}.
+    \caption[Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables.]{Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables  \cite{maaloe_biva_2019}.
     The higher the latent variable, the more the reconstructions fall into the mode of the learned distribution.
@@
 
-In this paper we employ a simple hierarchical VAE with bottom-up inference paths and the more powerful BIVA variant with a bidirectional (top-down and bottom-up) inference model \parencite{maaloe_biva_2019}.
+In this paper we employ a simple hierarchical VAE with bottom-up inference paths and the more powerful BIVA variant with a bidirectional (top-down and bottom-up) inference model \cite{maaloe_biva_2019}.
 We employ skip connections between latent variables but omit them for brevity.
@@
 To alleviate this, we phrase OOD detection as a likelihood ratio test of being \emph{semantically} in-distribution.
-A standard likelihood ratio test \parencite{buse_likelihood_1982} suggests to consider the ratio between the associated likelihoods, which we can approximate on a log-scale by the corresponding lower bounds $\mathcal{L}$ and $\mathcal{L}^{>k}$,
+A standard likelihood ratio test \cite{buse_likelihood_1982} suggests to consider the ratio between the associated likelihoods, which we can approximate on a log-scale by the corresponding lower bounds $\mathcal{L}$ and $\mathcal{L}^{>k}$,
 \begin{equation}\label{eq:llr-as-difference-in-likelihoods}
@@
     }
-    \caption[Inference and generative models for a bottom-up hierarchical VAS]{The inference and generative models, $q_\phib$ and $p_\thetab$, for an $L=2$ layered bottom-up hierarchical VAE as the one used in our experiments.
+    \caption[Inference and generative models for a bottom-up hierarchical VAEs.]{The inference and generative models, $q_\phib$ and $p_\thetab$, for an $L=2$ layered bottom-up hierarchical VAE as the one used in our experiments.
     Dashed lines indicate deterministic skip connections which are employed in both networks.
-Skip connections are found to be useful for optimizing latent variable models \parencite{dieng_avoiding_2019, maaloe_biva_2019}.}
+Skip connections are found to be useful for optimizing latent variable models \cite{dieng_avoiding_2019, maaloe_biva_2019}.}
     \label{fig:hvae-graphical-model}
@@
     }
-    \caption[Average bits per dimension of different datasets for generative models trained on FashionMNIST and CIFAR10]{%
+    \caption[Average bits per dimension of different datasets for generative models trained on FashionMNIST and CIFAR10.]{%
     Average bits per dimension of different datasets for models trained on FashionMNIST and CIFAR10.
@@
     }
-    \caption{%
+    \caption[AUROC, AUPRC, and FPR80 of generative models for OOD detection (MNIST/FashionMNIST and SVHN/CIFAR10).]{%
         AUROC$\uparrow$, AUPRC$\uparrow$ and FPR80$\downarrow$ for OOD detection for a FashionMNIST model using scores on the FashionMNIST test set as reference.
@@
     \end{subfigure}
-    \caption[Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood and $\mathcal{L}^{>2}$ bound]{%
+    \caption[Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood and $\mathcal{L}^{>2}$ bound.]{%
     Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood \subref{fig:FMNIST-elbo-k0}, the $\mathcal{L}^{>2}$ bound \subref{fig:FMNIST-elbo-k2} and the $LLR^{>1}$ score \subref{fig:FMNIST-llr}.
-All densities are computed using the HVAE model.).
+All densities are computed using the HVAE model.
     For the regular likelihood MNIST is very clearly more likely on average than the FashionMNIST test data while with the $\mathcal{L}^{>2}$ bound separation is better but significant overlap remains.
@@
     \end{subfigure}
-    \caption[ROC curves for out of distribution detection (MNIST/FashionMNIST and SVHN/CIFAR10)]{%
+    \caption[ROC curves for out of distribution detection (MNIST/FashionMNIST and SVHN/CIFAR10).]{%
         ROC curves with AUROC score for detecting \subref{fig:FMNIST-roc-llr} MNIST as OOD with the HVAE model trained on FashionMNIST and \subref{fig:CIFAR10-roc-llr} SVHN as OOD with the BIVA model trained on CIFAR10.
@@
 
-We also note that since the hierarchical VAE is a probabilistic graphical latent variable model, it lends itself very naturally to manipulation at the feature level \parencite{kingma_semi-supervised_2014, maaloe_auxiliary_2016, maaloe_semi-supervised_2017}.
+We also note that since the hierarchical VAE is a probabilistic graphical latent variable model, it lends itself very naturally to manipulation at the feature level \cite{kingma_semi-supervised_2014, maaloe_auxiliary_2016, maaloe_semi-supervised_2017}.
 This property sets it apart from other generative models that do not explicitly define such a hierarchy of features.

### 2023-04-25 chapters/paper_review/03_approaches.tex
Add and revise short table and figure captions

@@
 
-\section{Speech Representation Learning Paradigms} \label{sec:approach}
+\section{Speech representation learning paradigms} \label{sec:approach}
 Due to the characteristics of speech, SSL pretext tasks developed for CV
@@
    % \includegraphics[width=0.98\textwidth]{paper_review/model_timeline_wo_dccae.pdf}
-	 \caption{A selection of models listed according to first publication date
+	 \caption[Selection of self-supervised models for speech.]{A selection of models listed according to first publication date
 	 on arXiv or conference submission date when this clearly precedes the

### 2023-04-25 chapters/paper_review/04_benchmarking.tex
Add and revise short table and figure captions

@@
 
-\section{Benchmarks for Self-supervised Learning}
+\section{Benchmarks for self-supervised learning}
 \label{section:benchmark}
@@
   \scriptsize %footnotesize %small, scriptsize, tiny
-  \caption{Summary of datasets used in pre-training (denoted as PT) or
-  evaluating (denoted as EV) SSL techniques in the literature. The languages and
+  \caption[Summary of datasets used for pre-training or evaluation of SSL techniques in the literature.]%
+  {Summary of datasets used for pre-training (denoted as PT) or
+  evaluation (denoted as EV) of SSL techniques in the literature. The languages and
   sizes of the datasets are provided in columns 3 and 4. Column 5 lists the tasks each
@@
   \scriptsize
-  \caption{A summary of common experiment settings for various SSL
+  \caption[Summary of common experiment settings for various SSL
+  evaluations (Part 1).]{A summary of common experiment settings for various SSL
   evaluations (Part 1). Networks are usually pre-trained with SSL
@@
   \scriptsize
-  \caption{A summary of common experiment settings for various SSL
+  \caption[Summary of common experiment settings for various SSL
+  evaluations (Part 2).]{A summary of common experiment settings for various SSL
   evaluations (Part 2). See the caption of \cref{table:method_data_setting} for a
@@
 \includegraphics[width=0.9\textwidth]{paper_review/ls-test-clean-v2.pdf}
-\caption{SSL performance on ASR WER (vertical axis) evaluated with
+\caption[SSL performance for ASR evaluated on
+LS \textit{test-clean}.]{SSL performance on ASR WER (vertical axis) evaluated with
 LS \textit{test-clean} split. Techniques are sorted based on the number of

### 2023-07-17 chapters/main-contributions.tex
Add benchmarking paper
Updates

@@
+%!TEX root = ../thesis.tex
+
+% ~2 pages
+
 \chapter[main contribution]{Main contributions}\label{chp:main-contributions}
+
+As previously mentioned, the chapters of the thesis are self-contained studies and therefore detail their own contributions. 
+They are however written without consideration to the other chapters. 
+For that reason, this chapter will provide a high-level overview of the organization of the thesis and describe its main
+contributions in relation to the overall research topic.
+
+% \cref{chp} proposes a new hierarchical latent variable model for speech inspired by the Clockwork VAE [238]. 
+% The model is benchmarked against other latent variable models and autoregressive models for speech. 
+% It is shown that a hierarchy of latent variables improves the likelihood of the model. 
+% Finally, the latent space of the models are analyzed in terms of phonetic content.

### 2023-07-17 chapters/paper_review/03_approaches.tex
Updates

@@
 \begin{align}
-    q_t = a_k, \text{ where } k=\arg\min_j \norm{h_t - a_j}_2 .
+    q_t = a_k, \text{ where } k=\arg\min_j \norm{h_t - a_j}_2 \enspace .
 \end{align}
@@
 \begin{align}
-    \mathcal{L}_t = \underset{\text{encoder+decoder}}{\underbrace{\log p(x_t | q_{[1,t]})}} + \underset{\text{codebook}}{\underbrace{\text{MSE}\left(\mathrm{sg}\left[h_t\right], A\right)}} + \underset{\text{encoder}}{\underbrace{\alpha~\text{MSE}\left(h_t, \mathrm{sg}\left[A\right]\right)}}
+    \mathcal{L}_t = \underset{\text{encoder+decoder}}{\underbrace{\log p(x_t | q_{[1,t]})}}
+     + \underset{\text{codebook}}{\underbrace{\text{MSE}\left(\mathrm{sg}\left[h_t\right], A\right)}}
+     + \underset{\text{encoder}}{\underbrace{\alpha~\text{MSE}\left(h_t, \mathrm{sg}\left[A\right]\right)}} \enspace ,
     \label{eq: vector quantization losses}
@@
     \hat{x}_{t+c} &= g(h_{t}) \label{eq:c} , \\
-    \mathcal{L}_t &= \lVert \hat{x}_{t+c} - x_{t+c} \rVert_1 .
+    \mathcal{L}_t &= \lVert \hat{x}_{t+c} - x_{t+c} \rVert_1 \enspace .
 \end{align}
@@
     H^\prime_{[t+k,T]} = f_2(X_{[t+k,T]}), \\
-    \hat{X}_{[t+1,t+k-1]} = g(h_t, h^\prime_{t+k}).
+    \hat{X}_{[t+1,t+k-1]} = g(h_t, h^\prime_{t+k}) \enspace .
 \end{gather}

### 2023-07-20 chapters/paper_hierarchical.tex
Some formatting

@@
 \chapter[hierarchical vaes know what they don't know]{Hierarchical VAEs Know What They Don't Know}
-\label{chp:hierarchical}
+\label{chp:paper-hierarchical}
+% \ifthenelse{\equal{\skippaperbody}{true}}{}{
 
@@
     }
-    \label{fig:reconstructions-fashionmnist}
+    \label{fig_hierarchical:reconstructions-fashionmnist}
 \end{figure}
@@
 
-\section{Why does OOD detection fail?}\label{sec:why-does-ood-fail}
+\section{Why does OOD detection fail?}\label{sec_paper_hierarchical:why-does-ood-fail}
 \begin{figure}[t]
@@
     \vspace{0mm}
-    \label{fig:correlations-heatmap}
+    \label{fig_hierarchical:correlations-heatmap}
 \end{figure}
@@
 This raises the question to what extend such features are relevant when detecting OOD data, also suggested  by \cite{nalisnick_deep_2019} and examined for Glow and PixelCNN in \cite{schirrmeister_understanding_2020}.
-To investigate, we train two hierarchical VAEs (\cref{sec:background-hie-VAE}) on FashionMNIST and MNIST, respectively, and compute the between-models correlation of the extracted features of in-distribution data and OOD data.
-The result appears in \cref{fig:correlations-heatmap}.
+To investigate, we train two hierarchical VAEs (\cref{sec_paper_hierarchical:background-hie-VAE}) on FashionMNIST and MNIST, respectively, and compute the between-models correlation of the extracted features of in-distribution data and OOD data.
+The result appears in \cref{fig_hierarchical:correlations-heatmap}.
 We observe that features extracted in the early layers (low-level features) correlate strongly between the two models, and that this correlation drops as we get into later layers.
@@
 
-To shed further light on the impact of semantic versus low-level features, we look at model reconstructions of images with a hierarchical VAE (\cref{fig:biva-reconstructions-celeba}).
+To shed further light on the impact of semantic versus low-level features, we look at model reconstructions of images with a hierarchical VAE (\cref{fig_hierarchical:biva-reconstructions-celeba}).
 To study the feature hierarchy, we replace the inference distribution with the corresponding conditional prior in the first layers of the model to see what information is lost.
@@
     \vspace{0mm}
-    \label{fig:biva-reconstructions-celeba}
+    \label{fig_hierarchical:biva-reconstructions-celeba}
 \end{figure}
 Finally, we look at reconstructions of out-of-distribution data.
-\cref{fig:reconstructions-fashionmnist} illustrates that MNIST data is surprisingly well reconstructed by a hierarchical VAE trained on FashionMNIST.
+\cref{fig_hierarchical:reconstructions-fashionmnist} illustrates that MNIST data is surprisingly well reconstructed by a hierarchical VAE trained on FashionMNIST.
 Similar results have been found elsewhere \cite{xiao_likelihood_2020}.
@@
 The model parameters $\thetab$ and variational parameters $\phib$ are jointly optimized by maximizing the \textit{evidence lower bound} (ELBO),
-\begin{equation}\label{eq:elbo}
+\begin{equation}\label{eq_hierarchical:elbo}
     \log p_\thetab(\xb) \geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}(\xb; \thetab, \phib)\ .
@@
 
-\subsection{Hierarchical variational autoencoders}\label{sec:background-hie-VAE}
+\subsection{Hierarchical variational autoencoders}\label{sec_paper_hierarchical:background-hie-VAE}
 Hierarchical VAEs are a family of probabilistic latent variable models which extends the basic VAE by introducing a hierarchy of $L$ latent variables $\zb=\zb_1, \dots, \zb_L$.
@@
 \end{equation}
-Regardless of the choice of inference model, a hierarchical VAE is still trained using the ELBO \eqref{eq:elbo}.
+Regardless of the choice of inference model, a hierarchical VAE is still trained using the ELBO \cref{eq_hierarchical:elbo}.
 
@@
 
-\subsection{Out-of-distribution detection}\label{sec:background-ood-detection}
+\subsection{Out-of-distribution detection}\label{sec_paper_hierarchical:background-ood-detection}
 % Intro to alternative scores
@@
 \subsection{A bound for semantic OOD detection}
-If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \cref{fig:reconstructions-fashionmnist}-\ref{fig:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb_1)$ will be high for both in- and out-of-distribution data.
-Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb_1)] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) || p(\zb))$ from \eqref{eq:elbo} relies on the KL-term for OOD detection.
+If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \cref{fig_hierarchical:reconstructions-fashionmnist}-\ref{fig_hierarchical:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb_1)$ will be high for both in- and out-of-distribution data.
+Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb_1)] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) || p(\zb))$ from \cref{eq_hierarchical:elbo} relies on the KL-term for OOD detection.
 For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) || p(\zb))$ can be expressed by a hierarchical sum% over the hierarchy
@@
 To shift the ELBO from primarily being based on the approximate posterior of the lowest latent variables to instead focus on the conditional prior, \textcite{maaloe_biva_2019} introduced slightly different likelihood lower bound defined as
-\begin{equation}\label{eq:biva->k}
+\begin{equation}\label{eq_hierarchical:biva->k}
     \mathcal{L}^{>k} = \mathbb{E}_{p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p_\thetab(\zb_{>k})}{q_\phib(\zb_{>k}|\xb)} \right]
@@
 where $k\in\{0,1,\dots,L\}$ (see Appendix for the derivation).
-We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
+We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq_hierarchical:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
 The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\zb_1,\dots,\zb_l \sim p_\thetab(\zb_{\leq k}|\zb_{>k})$ and only the $L-k$ highest from the approximate posterior $\zb_{k+1},\dots,\zb_L \sim q_\phib(\zb_{>k}|\xb)$.
@@
 \subsection{A likelihood-ratio score for all feature levels}
-While the $\mathcal{L}^{>k}$ bound provides a score for performing semantic OOD detection, it still relies on the data space likelihood function (see equation \eqref{eq:likelihoods-as-exact} below), which is known to be problematic for OOD detection (\cref{sec:background-ood-detection}).
+While the $\mathcal{L}^{>k}$ bound provides a score for performing semantic OOD detection, it still relies on the data space likelihood function (see equation \cref{eq_hierarchical:likelihoods-as-exact} below), which is known to be problematic for OOD detection (\cref{sec_paper_hierarchical:background-ood-detection}).
 To alleviate this, we phrase OOD detection as a likelihood ratio test of being \emph{semantically} in-distribution.
 A standard likelihood ratio test \cite{buse_likelihood_1982} suggests to consider the ratio between the associated likelihoods, which we can approximate on a log-scale by the corresponding lower bounds $\mathcal{L}$ and $\mathcal{L}^{>k}$,
-\begin{equation}\label{eq:llr-as-difference-in-likelihoods}
+\begin{equation}\label{eq_hierarchical:llr-as-difference-in-likelihoods}
     % LLR = - 2 log(L0 / L1)
@@
 \begin{align}
-    \mathcal{L}      &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) || p_\thetab(\zb|\xb)\right), \label{eq:likelihoods-as-exact} \\ 
+    \mathcal{L}      &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) || p_\thetab(\zb|\xb)\right), \label{eq_hierarchical:likelihoods-as-exact} \\ 
     \mathcal{L}^{>k} &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k }|\zb_{>k}) q_\phib(\zb_{>k}|\xb) || p_\thetab(\zb|\xb)\right) \nonumber \ .
@@
                  &\quad + D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb) || p_\thetab(\zb|\xb)\right) \ . \notag
-\end{align}\label{eq:llr-as-kls}
+\end{align}\label{eq_hierarchical:llr-as-kls}
 
@@
 Therefore, the $LLR^{>k}$ score should be an improved method for semantic OOD detection compared to $\mathcal{L}^{>k}$.
-Now, it can be noted that if we replace the regular ELBO, $\mathcal{L}$, in \eqref{eq:likelihoods-as-exact} with the strictly tighter importance weighted bound \cite{burda_importance_2016},
+Now, it can be noted that if we replace the regular ELBO, $\mathcal{L}$, in \cref{eq_hierarchical:likelihoods-as-exact} with the strictly tighter importance weighted bound \cite{burda_importance_2016},
 \begin{equation}
-    \mathcal{L}_{S} = \mathbb{E}_{q(\zb|\xb)}\left[ \log \frac{1}{N} \sum_{s=1}^{S} \frac{p(\xb, \zb^{(s)})}{q(\zb^{(s)}|\xb)} \right] \ , \label{eq:iw-bound}
+    \mathcal{L}_{S} = \mathbb{E}_{q(\zb|\xb)}\left[ \log \frac{1}{N} \sum_{s=1}^{S} \frac{p(\xb, \zb^{(s)})}{q(\zb^{(s)}|\xb)} \right] \ , \label{eq_hierarchical:iw-bound}
 \end{equation}
@@
     LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq k}|\zb_{>k}) q(\zb_{>k}|\xb) || p(\zb|\xb))
-\end{equation}\label{eq:llr-as-kls-iwae-reduced}
+\end{equation}\label{eq_hierarchical:llr-as-kls-iwae-reduced}
 which, in practice, is well-approximated for a finite $S$.
-We expect this importance weighted likelihood ratio to monotonically improve upon the one in \cref{eq:llr-as-kls} as $S$ increases and the KL-divergence in the regular ELBO that contains terms for which $\zb_i$ is high-dimensional goes to zero.
+We expect this importance weighted likelihood ratio to monotonically improve upon the one in \cref{eq_hierarchical:llr-as-kls} as $S$ increases and the KL-divergence in the regular ELBO that contains terms for which $\zb_i$ is high-dimensional goes to zero.
 
-Since the scores in \cref{eq:llr-as-kls,eq:llr-as-kls-iwae-reduced} are estimated by sampling their estimators are stochastic objects with nonzero variance.
+Since the scores in \cref{eq_hierarchical:llr-as-kls,eq_hierarchical:llr-as-kls-iwae-reduced} are estimated by sampling their estimators are stochastic objects with nonzero variance.
 We note that $\text{Var}(\widehat{LLR}^{>k}) = \text{Var}(\hat{\mathcal{L}}) + \text{Var}(\hat{\mathcal{L}}^{>k}) - 2\, \text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$.
@@
 Clearly, if this is not the case and the difference instead lies in low-level statistics, the scores would likely fail.
-We hypothesize that a complementary bound to \cref{eq:biva->k}, $\mathcal{L}^{<l}$ described in Appendix E, might be useful in these cases, but leave further examination to future work.
+We hypothesize that a complementary bound to \cref{eq_hierarchical:biva->k}, $\mathcal{L}^{<l}$ described in Appendix E, might be useful in these cases, but leave further examination to future work.
 
@@
     \centering
+    \resizebox{0.25\textwidth}{!}{
     \tikz{
@@
     }
+    }
     \caption[Inference and generative models for a bottom-up hierarchical VAEs.]{The inference and generative models, $q_\phib$ and $p_\thetab$, for an $L=2$ layered bottom-up hierarchical VAE as the one used in our experiments.
@@
 Skip connections are found to be useful for optimizing latent variable models \cite{dieng_avoiding_2019, maaloe_biva_2019}.}
-    \label{fig:hvae-graphical-model}
+    \label{fig_hierarchical:hvae-graphical-model}
     % \end{figure}
@@
 We use weight-normalization \cite{salimans_weight_2016} on all weights and residual networks in the deterministic paths. 
-A graphical representation of this model can be seen in \cref{fig:hvae-graphical-model}.
+A graphical representation of this model can be seen in \cref{fig_hierarchical:hvae-graphical-model}.
 We use a Bernoulli output distribution for FashionMNIST/MNIST and a discretized mixture of logistics output distribution \cite{salimans_pixelcnn_2017} for CIFAR10/SVHN.
@@
 
-The likelihoods for our trained models are in \cref{tab:bits-per-dim-ood} alongside baseline results for in-distribution and OOD data.
-The main results of the paper on the OOD tasks can be seen along with comparisons to the baseline methods in \cref{tab:rocauc-ood}.
+The likelihoods for our trained models are in \cref{tab_hierarchical:bits-per-dim-ood} alongside baseline results for in-distribution and OOD data.
+The main results of the paper on the OOD tasks can be seen along with comparisons to the baseline methods in \cref{tab_hierarchical:rocauc-ood}.
 We note that for all our results, the value of the score ($\mathcal{L}^{>k}$ and $LLR^{>k}$) for the training and test splits of the in-distribution data was observed to have the same empirical distribution to within sampling error hence yielding an AUROC score of $\approx0.5$ as expected.
@@
     BIVA refers to our implementation of \textcite{maaloe_biva_2019}.}
-    \label{tab:bits-per-dim-ood}
+    \label{tab_hierarchical:bits-per-dim-ood}
     \vspace{0mm}
@@
     }
-    \label{tab:rocauc-ood}
+    \label{tab_hierarchical:rocauc-ood}
 \end{table}
@@
         \caption{}
-        \label{fig:FMNIST-elbo-k0}
+        \label{fig_hierarchical:FMNIST-elbo-k0}
     \end{subfigure}
@@
         \caption{}
-        \label{fig:FMNIST-elbo-k2}
+        \label{fig_hierarchical:FMNIST-elbo-k2}
     \end{subfigure}
@@
         \caption{}
-        \label{fig:FMNIST-llr}
+        \label{fig_hierarchical:FMNIST-llr}
     \end{subfigure}
     \caption[Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood and $\mathcal{L}^{>2}$ bound.]{%
-    Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood \subref{fig:FMNIST-elbo-k0}, the $\mathcal{L}^{>2}$ bound \subref{fig:FMNIST-elbo-k2} and the $LLR^{>1}$ score \subref{fig:FMNIST-llr}.
+    Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood \subref{fig_hierarchical:FMNIST-elbo-k0}, the $\mathcal{L}^{>2}$ bound \subref{fig_hierarchical:FMNIST-elbo-k2} and the $LLR^{>1}$ score \subref{fig_hierarchical:FMNIST-llr}.
 All densities are computed using the HVAE model.
@@
     }
-    \label{fig:FMNIST-ood-densities}
+    \label{fig_hierarchical:FMNIST-ood-densities}
 \end{figure}
@@
 We first report the results of the different variations of the $\mathcal{L}^{>k}$ bound for OOD detection. 
-We reconfirm the results of \textcite{nalisnick_deep_2019} by observing that our hierarchical latent variable models also assign higher $\mathcal{L}^{>0}$ to the OOD dataset in the FashionMNIST/MNIST and CIFAR10/SVHN cases resulting in an AUROC$\uparrow$ inferior to random (\cref{tab:rocauc-ood}).
+We reconfirm the results of \textcite{nalisnick_deep_2019} by observing that our hierarchical latent variable models also assign higher $\mathcal{L}^{>0}$ to the OOD dataset in the FashionMNIST/MNIST and CIFAR10/SVHN cases resulting in an AUROC$\uparrow$ inferior to random (\cref{tab_hierarchical:rocauc-ood}).
 Switching the in-distribution data for the OOD data in both cases result in correctly detecting the OOD data; an asymmetry also reported by \textcite{nalisnick_deep_2019}.
-\cref{fig:FMNIST-elbo-k0} shows the density of $\mathcal{L}^{>0}$ in bits per dimension \cite{theis_note_2016} by the model trained on FashionMNIST when evaluated on the FashionMNIST and MNIST test sets.
+\cref{fig_hierarchical:FMNIST-elbo-k0} shows the density of $\mathcal{L}^{>0}$ in bits per dimension \cite{theis_note_2016} by the model trained on FashionMNIST when evaluated on the FashionMNIST and MNIST test sets.
 We observe a high degree of overlap, with less separation of the OOD data compared to similar results of autoregressive and flow-based models, like \textcite{xiao_likelihood_2020}.
 
-We then evaluate the looser $\mathcal{L}^{>k}$ \eqref{eq:biva->k} for $k\in\{1,L\}$.
-\cref{fig:FMNIST-elbo-k2} shows the result for $\mathcal{L}^{>2}$, which yielded the highest AUCROC$\uparrow$, only slightly better than random.
+We then evaluate the looser $\mathcal{L}^{>k}$ \cref{eq_hierarchical:biva->k} for $k\in\{1,L\}$.
+\cref{fig_hierarchical:FMNIST-elbo-k2} shows the result for $\mathcal{L}^{>2}$, which yielded the highest AUCROC$\uparrow$, only slightly better than random.
 Like \textcite{maaloe_biva_2019}, we see that increasing the value of $k$ generally leads to improved OOD detection.
@@
         % A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>1}$.}
-        \label{fig:FMNIST-roc-llr}
+        \label{fig_hierarchical:FMNIST-roc-llr}
     \end{subfigure}
@@
         % A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>2}$.}
-        \label{fig:CIFAR10-roc-llr}
+        \label{fig_hierarchical:CIFAR10-roc-llr}
     \end{subfigure}
     \caption[ROC curves for out of distribution detection (MNIST/FashionMNIST and SVHN/CIFAR10).]{%
-        ROC curves with AUROC score for detecting \subref{fig:FMNIST-roc-llr} MNIST as OOD with the HVAE model trained on FashionMNIST and \subref{fig:CIFAR10-roc-llr} SVHN as OOD with the BIVA model trained on CIFAR10.
+        ROC curves with AUROC score for detecting \subref{fig_hierarchical:FMNIST-roc-llr} MNIST as OOD with the HVAE model trained on FashionMNIST and \subref{fig_hierarchical:CIFAR10-roc-llr} SVHN as OOD with the BIVA model trained on CIFAR10.
         A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>1}$.
@@
 We now move to the likelihood ratio-based score.
-We find that $LLR^{>k}$ separates the OOD MNIST data from in-distribution FashionMNIST to a higher degree than the likelihood estimates as can be seen by the empirical densities of the score in \cref{fig:FMNIST-llr}.
+We find that $LLR^{>k}$ separates the OOD MNIST data from in-distribution FashionMNIST to a higher degree than the likelihood estimates as can be seen by the empirical densities of the score in \cref{fig_hierarchical:FMNIST-llr}.
 We note that the likelihood ratio between the ELBO and the $\mathcal{L}^{>k}$ bound provides the highest degree of separation of MNIST and FashionMNIST as measured by the AUROC$\uparrow$ for $k=1$ smaller than $L$.
 This is not surprising since the value of $k$ that provides the maximal separation to the reference in-distribution dataset need not be the one for which $\mathcal{LLR}^{>k}$ is overall maximal for the OOD dataset.
-We also visualize the ROC curves resulting from using the $LLR^{>k}$ score for OOD detection on both FashionMNIST/MNIST and CIFAR10/SVHN and compare it to the ROC curves resulting from the different $\mathcal{L}^{>k}$ bounds in Figures \ref{fig:FMNIST-roc-llr} and \ref{fig:CIFAR10-roc-llr}, respectively.
+We also visualize the ROC curves resulting from using the $LLR^{>k}$ score for OOD detection on both FashionMNIST/MNIST and CIFAR10/SVHN and compare it to the ROC curves resulting from the different $\mathcal{L}^{>k}$ bounds in Figures \ref{fig_hierarchical:FMNIST-roc-llr} and \ref{fig_hierarchical:CIFAR10-roc-llr}, respectively.
 On both datasets we see significantly better discriminatory performance when using the $LLR^{>k}$ score.
 
-\cref{tab:rocauc-ood} shows that BIVA improves upon the HVAE model for OOD detection on CIFAR while \cref{tab:bits-per-dim-ood} shows that the BIVA model also improves upon the HVAE in terms of likelihood.
+\cref{tab_hierarchical:rocauc-ood} shows that BIVA improves upon the HVAE model for OOD detection on CIFAR while \cref{tab_hierarchical:bits-per-dim-ood} shows that the BIVA model also improves upon the HVAE in terms of likelihood.
 We hypothesize that models larger than our implementation of BIVA, with better likelihood scores may perform even better \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
@@
 \subsection{Comparison to baselines}
-\paragraph{Performance} \cref{tab:rocauc-ood} summarize our results compared to baselines based on the commonly used AUROC$\uparrow$, AUPRC$\uparrow$ and FPR80$\downarrow$ metrics.
+\paragraph{Performance} \cref{tab_hierarchical:rocauc-ood} summarize our results compared to baselines based on the commonly used AUROC$\uparrow$, AUPRC$\uparrow$ and FPR80$\downarrow$ metrics.
 Our method outperforms other generative model-based methods such as WAIC \cite{choi_waic_2019} with Glow model and performs similarly to the likelihood regret method of \cite{xiao_likelihood_2020}.
@@
 757360).
+
+% }

### 2023-07-20 chapters/main-contributions.tex
First go at main contributions

@@
 
-\chapter[main contribution]{Main contributions}\label{chp:main-contributions}
+\chapter[main contributions]{Main contributions}\label{chp:main-contributions}
 
@@
 
-% \cref{chp} proposes a new hierarchical latent variable model for speech inspired by the Clockwork VAE [238]. 
-% The model is benchmarked against other latent variable models and autoregressive models for speech. 
-% It is shown that a hierarchy of latent variables improves the likelihood of the model. 
-% Finally, the latent space of the models are analyzed in terms of phonetic content.
+\newcommand{\contribitem}[1]{{\vspace{1em}\noindent\scshape\bfseries{\color{dtured}\cref{#1}}\\\ucnameref{#1}\\}}
+
+\contribitem{part:background}
+This part of the thesis provides the necessary background for the rest of the thesis.
+
+\contribitem{part:unsupervised-learning}
+
+\contribitem{chp:paper-hierarchical}
+This paper proposes a new method for unsupervised out-of-distribution detection using hierarchical latent variable models.
+The paper provides evidence that the failure of deep generative models to assign lower likelihoods to data from outside the training distribution is due to overemphasis on low-level features that generalize between distributions.
+To solve this, the proposed method computes a likelihood-ratio score for OOD detection that requires data to be in-distribution across all feature-levels.
+The method is computationally efficient, scalable, fully unsupervised and achieves state-of-the-art performance on several image-based OOD detection benchmarks.
+
+\contribitem{chp:paper-benchmarking} This paper proposes a new hierarchical latent variable model for speech inspired by the Clockwork VAE \cite{saxena_clockwork_2021}.
+The model is benchmarked against other latent variable models and autoregressive models for speech.
+It is shown that a hierarchy of latent variables improves the likelihood of the model.
+Finally, the latent space of the models are analyzed in terms of phonetic content.
+
+\contribitem{chp:paper-review} This paper provides a comprehensive review of self-supervised speech representation learning.
+Previous work is grouped into three main categories of methods: contrastive, autoregressive, and generative, and the methods are compared in terms of their training objectives, model architectures, and performance on downstream tasks.
+The review also provides a discussion of the current state of the field and promising future directions of research.

### 2023-07-20 chapters/background.tex
Add benchmarking paper
Updates
Add WIP for introduction and background

@@
-\chapter[background]{Background}\label{chp:background}
+%!TEX root = ../thesis.tex
 
+% ~8 pages
+
+\chapter[technical background]{Technical Background}\label{chp:technical-background}
+
+\section{Uncertainty}
+
+% https://en.wikipedia.org/wiki/Uncertainty#cite_note-3
+% https://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=8371683
+
+In the general public, the concept of uncertainty goes by many names and its meaning can be dependent on a specific context.
+However, in most quantitative scientific fields, there seems to be a consensus definition along the lines of the following \cite{hubbard_how_2014}:
+
+\begin{center}
+    \textit{Uncertainty is the lack of certainty; a state of limited knowledge where it is impossible to exactly describe the existing state, a future outcome, or more than one possible outcome.}
+\end{center}
+
+While a definition of uncertainty made using only words easily begs philosophical inquisition, information theory provides a rigourous method for quantifying uncertainty \cite{kabir_neural_2018} by the language of probabilities.
+In this context, uncertainty is understood as the information entropy of a random variable \cite{mackay_information_2003}.
+
+\subsection{Information entropy}
+First characterised by Claude Shannon in 1948 \cite{shannon_mathematical_1948}, information entropy $H(X)$ is a measure of the average amount of information contained in a random variable $X$.
+Shannon's definition follows from three fundamental axioms of information theory.
+Let $I_X(x)$ be the information carried by a specific outcome $x$ when sampling the random variable $X$ with probability mass function $p_X(x)$.
+Then, the axioms are as follows:
+\begin{enumerate}
+    \item  $I_X(x)$ is a monotonically decreasing function in $p_X(x)$. The more likely an outcome is, the less information it carries.
+    \item Outcomes that are certain to happen carry no information. $I_X(1) = 0$.
+    \item If an outcome $x_i$ is independent of another outcome $x_j$, then $I(x_i x_j) = I(x_i) + I(x_j)$.
+The information carried by independent events is the sum of the information carried by each event.
+\end{enumerate}
+
+From these axioms, Shannon derived the following expression for the information content of an event $x$ with probability $p_X(x)$:
+\begin{equation}
+    I_X(x) = -\log_b p_X(x)
+\end{equation}
+
+information entropy of a random variable $X$ with $n$ possible outcomes:
+\begin{equation}
+    H(X) = -\sum_{i=1}^n p_i \log p_i
+\end{equation}
+
+% Two kinds of uncertainty:\\
+% - Aleatoric: Uncertainty inherent to the data. Irreducible. "Known unknowns".\\
+% - Epistemic: Uncertainty due to lack of knowledge. Can be reduced with more data. "Unknown unknowns".\\
+
+% https://arxiv.org/pdf/2306.05674.pdf
+
+% We compare our framework with several major related lines of work.
+First, our work focuses on the quantification of epistemic uncertainty, which refers to the errors coming from the inadequacy of the model or data noises.
+This is different from aleatoric uncertainty, which refers to the intrinsic stochasticity of the problem [87, 91, 112, 59, 34], or predictive uncertainty which captures the sum of epistemic and aleatoric uncertainties (but not their dissection) [89, 92, 12, 3, 22].
+Regarding epistemic uncertainty, a related line of study is deep ensemble that aggregates predictions from multiple independent training replications [76, 70, 38, 8, 89].
+This approach, as we will make clear later, can reduce and potentially quantify procedural variability, but a naive use would require demanding retraining effort and does not address data variability.
+Another line is the Bayesian UQ approach on neural networks [41, 2].
+This regards network weights as parameters subject to common priors such as Gaussian.
+Because of the computation difficulties in exact inference, an array of studies investigate efficient approximate inference approaches to estimate the posteriors [40, 46, 16, 31, 30, 90, 74, 53].
+While powerful, these approaches nonetheless possess inference error that could be hard to quantify, and ultimately finding rigorous guarantees on the performance of these approximate posteriors remains open to our best knowledge.
+
+% 87: 
+% 91: 
+% 112:
+% 59: 
+% 34:
+
+% 89: 
+% 92: 
+% 12:
+% 3: 
+% 22:
+
+% 76: 
+% 70: 
+% 38: 
+% 8: 
+% 89:
+
+\section{Probabilistic latent variable models}
+
+\section{Speech representation learning}
+
+Modelling paradigms
 \begin{enumerate}
@@
     \item Self-supervised learning
-    \item 
 \end{enumerate}
@@
 
-\cite{bahdanau_neural_2014}
+Two approaches to statistics:\\
+- Bayesian\\
+- Frequentist\\
 
-\begin{enumerate}
-    \item Introduce variational inference in relation to variational autoencoders
-    \item Introduce ASR and CTC
-\end{enumerate}
+Two kinds of uncertainty:\\
+- Aleatoric: Uncertainty inherent to the data. Irreducible. "Known unknowns".\\
+- Epistemic: Uncertainty due to lack of knowledge. Can be reduced with more data. "Unknown unknowns".\\
+
+Approaches to uncertainty in deep learning:\\
+- Tuning of supervised classifiers to output well-calibrated probabilities that correspond to the actual likelihood of the prediction being correct (on some validation set).\\
+- Learning rich unsupervised representations of the data that can be used to estimate the uncertainty of the prediction (e.g.
+by measuring the distance to the nearest training example in the latent space).\\
+- Learning of a probability distribution over the parameters of the model, which can then be used to compute the uncertainty of the prediction (Bayesian neural networks).\\
+
+In the field of machine learning, uncertainty is often represented by a probability distribution.
+The most common approach is to use Bayesian methods, where uncertainty is captured by posterior distributions over model parameters and predictions.
+Bayesian neural networks, for instance, offer a powerful framework to model uncertainty in deep learning architectures.
+By incorporating prior beliefs and updating them based on observed data, these networks can produce probabilistic predictions that provide a measure of uncertainty.
+This is particularly useful in applications where high-confidence predictions are required, and the consequences of errors can be significant.
+
+Recently, another promising approach to uncertainty estimation in machine learning is the use of Monte Carlo Dropout \cite{gal_dropout_2016}.
+Monte Carlo Dropout leverages the idea of dropout regularization, originally employed during training to prevent overfitting.
+To form a prediction with associated uncertainty, Monte Carlo Dropout proposes to make multiple forward passes through the network while sampling different dropout masks for each one.
+This leads to obtaining a distribution of outputs for each input sample and the variance among these sampled predictions serves as a measure of uncertainty.
+Monte Carlo Dropout has shown remarkable success in various tasks such as image classification, object detection, and natural language processing.
+It is computationally efficient and can be easily incorporated into existing deep learning architectures, making it an attractive choice for uncertainty estimation.
+
+Furthermore, there has been a surge of interest in ensemble methods for uncertainty estimation.
+Ensembles combine the predictions of multiple models to obtain a more robust and calibrated uncertainty measure.
+Bagging and boosting techniques, which have been widely used in the field of statistics, have found their way into the realm of machine learning for uncertainty estimation as well.
+By training multiple models with different initializations or using diverse learning algorithms, ensembles can capture different sources of uncertainty, thereby providing a comprehensive assessment of the overall uncertainty in the predictions.
+
+Despite the progress made in uncertainty estimation for machine learning models, challenges still remain.
+One key concern is the interpretability of uncertainty measures.
+While probabilistic outputs are more informative than point estimates, effectively communicating uncertainty to end-users and decision-makers is not trivial.
+Developing visualization techniques and intuitive explanations for uncertainty is an ongoing area of research.
+Additionally, quantifying uncertainty in high-dimensional data or complex models with massive amounts of parameters requires careful consideration and efficient computational strategies.

### 2023-07-20 chapters/introduction.tex
Add benchmarking paper
Updates
Add WIP for introduction and background

@@
+%!TEX root = ../thesis.tex
+
 \chapter[introduction]{Introduction}\label{chp:introduction}
+
+% ~9 pages
+
+Uncertainty is a fundamental part of human experience.
+Understanding and quantifying uncertainty is crucial in making informed decisions, drawing meaningful conclusions, and assessing the reliability of predictions made from observations.
+Yet, capturing it in a mathematical model is a difficult task.
+
+Historically, uncertainty estimation has been a central topic in the field of statistics.
+Uncertainty has traditionally been addressed through methods like confidence intervals, hypothesis testing, and Bayesian inference.
+Confidence intervals provide a range of plausible values for a parameter, while hypothesis testing evaluates the significance of differences between groups or variables \cite{blitzstein_introduction_2019}.
+Bayesian inference, on the other hand, offers a powerful framework to incorporate prior knowledge and update beliefs based on observed data using probability theory.
+These approaches have been extensively studied and applied in various domains, from social sciences to natural sciences, to improve the robustness and generalizability of statistical analyses \cite{gelman_bayesian_2013}.
+However, the widespread adoption in diverse applications of machine learning, and in particular deep neural networks, has posed new challenges in dealing with uncertainty.
+
+Machine learning algorithms are designed to learn patterns from data and make predictions or decisions based on those patterns.
+Since many applications of machine learning are concerned with real-world phenomena, it is important to understand the uncertainty associated with the predictions made by the model.
+For instance, in medical diagnosis, knowing the uncertainty in a model's prediction is crucial when the consequences of a false positive or false negative can be significant.
+In speech recognition, knowing that a certain word was hard to understand for the model can help avoid misinterpretation of the transcribed text.
+
+Speech processing is
+
+% The performance of machine learning models is highly dependent on the quality and quantity of data used for training.
+In many cases, the data used to train machine learning models are collected from the real world, and therefore, may contain biases and errors which models may learn to mirror.
+In addition, the complexity of modern machine learning models makes it difficult to understand how they arrive at their predictions.
+This is especially true for deep neural networks, which are often referred to as black-box models.
+The lack of interpretability makes it difficult to assess the reliability of the model's predictions.
+
+% Modern speech processing relies on high-performance parallel computing [40, 157], large volumes of data [136, 213], and years of innovation in model design [109, 266].
+
+% 40: High performance convolutional neural networks for document processing \cite{chellapilla_high_2006}
+% 157: Imagenet classification with deep convolutional neural networks \cite{krizhevsky_imagenet_2012}
+% 136: Libri-light: A benchmark for asr with limited or no supervision \cite{kahn_libri-light_2020}
+% 213: Librispeech: an ASR corpus based on public domain audio books \cite{panayotov_librispeech_2015}
+% 109:  Long short-term memory \cite{hochreiter_long_1997}
+% 266: Attention Is All You Need \cite{vaswani_attention_2017}
+
+The research in this thesis covers a number of different modelling approaches and tasks, including
+
+The research in this thesis covers a varied set of topics.
+One group of the produced studies are concerned with generative latent variable models and their applications to uncertainty estimation and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
+Another group of studies are concerned with self-supervised speech representation learning and automatic speech recognition \cite{borgholt_scaling_2021,borgholt_we_2021,mohamed_selfsupervised_2022,borgholt_brief_2022}.
+A final group
+
+This first part of the thesis ties together these different studies by providing a high-level overview of the research topic and the main contributions of the thesis.
+
+covering both representation learning and uncertainty estimation and their applications to speech recognition and automated medical coding.
+
+\subsection{Motivation: Speech recognition}
+
+\subsection{Motivation: Automated medical coding}
+
+\subsection{Challenges in uncertainty estimation}
+
+General challenges in uncertainty estimation:
+\begin{enumerate}
+    \item \textbf{Accuracy}: The uncertainty estimation method should be accurate.
+    \item \textbf{Interpretability}: The uncertainty estimation method should be interpretable.
+    \item \textbf{Robustness}: The uncertainty estimation method should be robust to adversarial attacks.
+    \item \textbf{Applicability}: The uncertainty estimation method should be applicable to a wide range of models without requiring extensive modifications.
+    \item \textbf{Efficiency}: The uncertainty estimation method should be computationally efficient at training and prediction time.
+    \item \textbf{Scalability}: The uncertainty estimation method should scale to large datasets.
+\end{enumerate}
+
+\begin{enumerate}
+    \item How can we learn representations that can form the basis of both prediction and uncertainty estimation?
+    \item 
+\end{enumerate}
+
+\subsection{Scope of the thesis}
+
+\iffalse
+
+Uncertainty is a fundamental part of human experience.
+Yet, capturing it in a mathematical model is a difficult task.
+Historically, uncertainty estimation has been a central topic in the field of statistics.
+While classical statistical methods have provided valuable insights, they often assume well-defined probability distributions and may not fully capture the complexities of uncertainty in real-world scenarios.
+Moreover, they tend to rely heavily on parametric assumptions, which might not hold true for complex data with high-dimensional features and non-linear relationships.
+Therefore, in recent years, researchers have turned their attention towards more flexible and data-driven approaches to model uncertainty in various domains.
+
+In the field of machine learning, uncertainty is often represented by a probability distribution.
+The most common approach is to use Bayesian methods, where uncertainty is captured by posterior distributions over model parameters and predictions.
+Bayesian neural networks, for instance, offer a powerful framework to model uncertainty in deep learning architectures.
+By incorporating prior beliefs and updating them based on observed data, these networks can produce probabilistic predictions that provide a measure of uncertainty.
+This is particularly useful in applications where high-confidence predictions are required, and the consequences of errors can be significant.
+
+Recently, another promising approach to uncertainty estimation in machine learning is the use of Monte Carlo Dropout (MC Dropout).
+MC Dropout leverages the idea of dropout regularization, originally employed during training to prevent overfitting.
+During prediction, MC Dropout takes multiple forward passes through the network with dropout enabled, which leads to obtaining a distribution of outputs for each input sample.
+The variance among these sampled predictions serves as a measure of uncertainty.
+MC Dropout has shown remarkable success in various tasks such as image classification, object detection, and natural language processing.
+It is computationally efficient and can be easily incorporated into existing deep learning architectures, making it an attractive choice for uncertainty estimation.
+
+Furthermore, there has been a surge of interest in ensemble methods for uncertainty estimation.
+Ensembles combine the predictions of multiple models to obtain a more robust and calibrated uncertainty measure.
+Bagging and boosting techniques, which have been widely used in the field of statistics, have found their way into the realm of machine learning for uncertainty estimation as well.
+By training multiple models with different initializations or using diverse learning algorithms, ensembles can capture different sources of uncertainty, thereby providing a comprehensive assessment of the overall uncertainty in the predictions.
+
+Despite the progress made in uncertainty estimation for machine learning models, challenges still remain.
+One key concern is the interpretability of uncertainty measures.
+While probabilistic outputs are more informative than point estimates, effectively communicating uncertainty to end-users and decision-makers is not trivial.
+Developing visualization techniques and intuitive explanations for uncertainty is an ongoing area of research.
+Additionally, quantifying uncertainty in high-dimensional data or complex models with massive amounts of parameters requires careful consideration and efficient computational strategies.
+
+In conclusion, uncertainty estimation is a crucial aspect of machine learning models, enabling them to make more informed and reliable predictions.
+Researchers continue to explore innovative techniques to better represent and quantify uncertainty, allowing for safer and more trustworthy deployment of machine learning systems in real-world scenarios.
+
+Uncertainty is a fundamental part of human experience.
+Yet, capturing it in a mathematical model is a difficult task.
+Historically, uncertainty estimation has been a central topic in the field of statistics.
+Understanding and quantifying uncertainty is crucial in making informed decisions, drawing meaningful conclusions, and assessing the reliability of predictions and inferences derived from data.
+
+In the field of statistics, uncertainty is traditionally addressed through methods like confidence intervals, hypothesis testing, and Bayesian inference.
+Confidence intervals provide a range of plausible values for a parameter, while hypothesis testing evaluates the significance of differences between groups or variables.
+Bayesian inference, on the other hand, offers a powerful framework to incorporate prior knowledge and update beliefs based on observed data using probability theory.
+These approaches have been extensively studied and applied in various domains, from social sciences to natural sciences, to improve the robustness and generalizability of statistical analyses.
+However, the advent of machine learning and its widespread adoption in diverse applications has posed new challenges and opportunities in dealing with uncertainty.
+
+Machine learning algorithms are designed to learn patterns from data and make predictions or decisions based on that knowledge.
+In many cases, it is not enough to provide a single deterministic prediction; it is equally important to convey the level of uncertainty associated with the prediction.
+For instance, in medical diagnosis, knowing the uncertainty in a model's prediction is crucial when the consequences of a false positive or false negative can be significant.
+In autonomous driving systems, understanding uncertainty becomes essential to ensure the safety of passengers and pedestrians.
+
+To address the uncertainty challenge in machine learning, researchers have developed various techniques to represent and quantify uncertainty.
+One of the common approaches is the use of probabilistic models, where uncertainty is captured in the form of probability distributions.
+Instead of providing a single point prediction, probabilistic models generate a range of possible outcomes along with their associated probabilities.
+Bayesian neural networks, Gaussian processes, and variational autoencoders are some examples of such models that have gained prominence in the quest for uncertainty-aware machine learning.
+
+Another avenue for dealing with uncertainty in machine learning is through ensemble methods.
+Ensemble learning combines multiple models to improve predictive accuracy and can also provide uncertainty estimates.
+By aggregating the outputs of diverse models, ensemble methods can identify situations where predictions are consistent among the models, leading to higher confidence, and cases where the models disagree, indicating higher uncertainty.
+
+As the intersection of statistics and machine learning grows increasingly important, the integration of these two domains has become an active area of research.
+The incorporation of probabilistic modeling and Bayesian approaches into machine learning methods opens up new possibilities for developing more robust, interpretable, and calibrated models.
+Moreover, the synergy of statistical techniques with deep learning, reinforcement learning, and other advanced machine learning paradigms has the potential to push the boundaries of uncertainty quantification and application even further.
+
+In this thesis, we aim to explore and contribute to the field of uncertainty in statistics and machine learning.
+We will investigate state-of-the-art methods for uncertainty estimation, delve into their theoretical underpinnings, and evaluate their performance in various real-world scenarios.
+By gaining a deeper understanding of uncertainty and its implications, we aspire to pave the way for more reliable, trustworthy, and accountable AI systems in the future.
+
+In the following chapters, we will begin by reviewing the relevant literature on uncertainty in both statistics and machine learning, providing a comprehensive overview of the current landscape.
+We will then present our proposed methodologies and empirical studies, illustrating how uncertainty quantification can enhance decision-making processes and mitigate risks in practical applications.
+Finally, we will conclude with a discussion of our findings, potential limitations, and future research directions in this exciting and rapidly evolving field.
+
+\fi

### 2023-07-20 appendices/appendix_hierarchical.tex
Change reference keys

@@
 \section{Datasets}
-\cref{tab:datasets-overview} lists the datasets used in the paper. We use the predefined train/test splits for the datasets.
+\cref{tab_hierarchical:datasets-overview} lists the datasets used in the paper. We use the predefined train/test splits for the datasets.
 
@@
     \caption{Overview of the used datasets.}
-    \label{tab:datasets-overview}
+    \label{tab_hierarchical:datasets-overview}
 \end{table}
@@
 \section{Model details}\label{sec:model-details}
-In \cref{tab:hyperparameters} we specify the hyperparameters used when training our models.
+In \cref{tab_hierarchical:hyperparameters} we specify the hyperparameters used when training our models.
 We make our source code available at \url{https://github.com/JakobHavtorn/hvae-oodd}.
@@
 See \cref{sec:model-details} for more details.}
-    \label{tab:hyperparameters}
+    \label{tab_hierarchical:hyperparameters}
 \end{table}
@@
     \caption{The expected inverse volume change for Gaussian Jacobians \eqref{eq:Edet} on a log-scale.}
-    \label{fig:Edet}
+    \label{fig_hierarchical:Edet}
 \end{figure}
@@
 This super-exponential growth further implies that the first layers dominate the marginal likelihood $p(\xb)$.
-This is also visually evident in \cref{fig:Edet}.
+This is also visually evident in \cref{fig_hierarchical:Edet}.
 
@@
 \section{Additional results}
-We provide additional results for a model trained on FashionMNIST in \cref{tab:additional-results-fashionmnist}, a model trained on MNIST in \cref{tab:additional-results-mnist}, a model trained on CIFAR10 in \cref{tab:additional-results-cifar} and a model trained on SVHN in \cref{tab:additional-results-svhn}.
+We provide additional results for a model trained on FashionMNIST in \cref{tab_hierarchical:additional-results-fashionmnist}, a model trained on MNIST in \cref{tab_hierarchical:additional-results-mnist}, a model trained on CIFAR10 in \cref{tab_hierarchical:additional-results-cifar} and a model trained on SVHN in \cref{tab_hierarchical:additional-results-svhn}.
 
@@
     \caption[Additional results for the HVAE model trained on SVHN]{Additional results for the HVAE model trained on SVHN. All results computed with 1000 importance samples.}
-    \label{tab:additional-results-svhn}
+    \label{tab_hierarchical:additional-results-svhn}
 \end{table}
@@
     \caption[Additional results for the HVAE model trained on CIFAR10]{Additional results for the HVAE model trained on CIFAR10. All results computed with 1000 importance samples.}
-    \label{tab:additional-results-cifar}
+    \label{tab_hierarchical:additional-results-cifar}
 \end{table}
@@
     \caption[Additional results for the HVAE model trained on FashionMNIST]{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
-    \label{tab:additional-results-fashionmnist}
+    \label{tab_hierarchical:additional-results-fashionmnist}
 \end{table}
@@
     \caption[Additional results for the HVAE model trained on MNIST]{Additional results for the HVAE model trained on MNIST. All results computed with 1000 importance samples.}
-    \label{tab:additional-results-mnist}
+    \label{tab_hierarchical:additional-results-mnist}
 \end{table}
 % Results of MNIST model. (57bed090cf4a4e44b896b4c2cce55500)
-
-\iffalse
-
-\fi

### 2023-07-20 chapters/paper_modelagnostic.tex
Add WIP model-agnostic paper

%!TEX root = ../thesis.tex

\chapter[model-agnostic out-of-distribution detection using combined statistical tests]{Model-Agnostic Out-of-Distribution Detection Using Combined Statistical Tests}
\label{chp:paper-modelagnostic}
\ifthenelse{\equal{\skippaperbody}{true}}{}{

\section*{Abstract}
We present simple methods for out-of-distribution detection using a trained generative model. These techniques, based on classical statistical tests, are model-agnostic in the sense that they can be applied to any differentiable generative model. The idea is to combine a classical parametric test (Rao's score test) with the recently introduced typicality test. These two test statistics are both theoretically well-founded and exploit different sources of information based on the likelihood for the typicality test and its gradient for the score test. We show that combining them using Fisher's method overall leads to a more accurate out-of-distribution test. We also discuss the benefits of casting out-of-distribution detection as a statistical testing problem, noting in particular that false positive rate control can be valuable for practical out-of-distribution detection. Despite their simplicity and generality, these methods can be competitive with model-specific out-of-distribution detection algorithms without any assumptions on the out-distribution. %out-of-distribution data.\looseness=-1

\section{Introduction}

The ability to recognise when data are anomalous, i.e.\@ if they originate from a distribution different from that of the training data, is a necessary property for machine learning models for safe and reliable applications in the real world. Historically, \citet{bishop1994novelty} proposed to use a one-sided threshold on the log-likelihoods of a learned model as a decision rule to identify outliers in a dataset. However, recently, \citet{nalisnick2018deep, hendrycks2018deep} showed that state-of-the-art deep generative models (DGMs) failed in this task, assigning higher a likelihood to out-of-distribution (OOD) data than in-distribution data. Most of the recent works focused on proposing new test statistics to alleviate the problem of using the plain likelihood, see \cref{sec:related_works} for details.

We believe that OOD detection should be formulated as statistical hypothesis testing \citep{nalisnick2019detecting, ijcai2021-292, haroush2021statistical}. Since the power of a single test depends on the out-distribution \citep{zhang2021understanding}, we propose to approach this problem by using a combination of multiple statistical tests. %While this does not solve completely the problem, because the power of the combined tests will also depend on the out-distribution, we think that we can get better performance than using a single-statistics especially in situations where one statistic fails. 
While the power of the combined test also depends on the out-distribution, we hypothesise that the combined test empirically will perform better, especially in situations where one of the statistics fails.
%
%Furthermore, the use of the statistical testing framework, allows us to correct for the multiple comparisons problem when identifying outliers in a dataset by controlling the number of type I errors through the false discovery rate (FDR).
Furthermore, the use of the statistical testing framework has several advantages. Since we obtain a $p$-value, it is more natural deciding on a threshold as this corresponds to the significance level. In addition to that, it also allow us to correct for the multiple comparisons problem when identifying outliers in a dataset by controlling the number of Type I errors through the false discovery rate (FDR).

In summary, our contributions are the following:
\begin{itemize}
    \item We illustrate the benefits of combining multiple statistical tests to perform OOD detection with DGMs using well-established methods. This allows for a proper decision procedure to control the FDR in a real outlier detection setting.
    \item We revisit some proposed detection scores and highlight their alternative formulation as classical significance tests.
    \item Empirically we show the complementarity of the typicality and the score statistics and that their combination leads to a robust score for anomaly detection. %We evaluate our proposed method with two PixelCNN++ \cite{salimans2017pixelcnn++}, two Glow models \cite{kingma2018glow} and one Hierarchical-VAE (HVAE)\cite{kingma2013auto}.
\end{itemize}
%

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  SECTION 2   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Using statistical tests for out-of-distribution detection}

We consider some data of interest that live in a space $\mathcal{X}$. Assume that we have a curated dataset $x_1,\ldots,x_m$, i.e.\@ there are no outliers, and we are interested in understanding if some new data $\tilde{x}_1,\ldots,\tilde{x}_n$ are collectively anomalies. In other words, we wonder whether or not $\tilde{x}_1,\ldots,\tilde{x}_n$ are likely to come from the same distribution that generated our curated dataset. We present in this section two different approaches for doing out-of-distribution detection using statistical tests: one based on classical parametric tests and one based on maximum mean discrepancy. A convenient property of the tests we consider is that they are all one-sided, which means we can expect them to be larger when the data are more likely to be OOD. This allows us to compute $p$-values by simply using the empirical CDF, which is hyperparameter-free.

Note that in this problem formulation, the case $n=1$ corresponds to the situation where we need to decide if a \emph{single} data point is out-of-distribution. This hardest setting will be of particular interest, and this is also the main focus of recent work, see \cref{sec:related_works}.

\subsection{Parametric tests for out-of-distribution detection}\label{sec:parametric}
The typical approach is to consider a parametric family $(p_\theta)_{\theta \in  \Theta}$ of probability densities over $\mathcal{X}$ and learn a suitable $\theta_0 \in  \Theta$ using any inference technique, for example maximum likelihood, and the clean data $x_1,\ldots,x_m$. Depending on the input domain, $(p_\theta)_{\theta \in  \Theta}$ could be composed of DGMs (in that case, $\theta$ would be neural network weights) or Gaussian mixture models (in that case, $\theta$ would be composed of means, covariances, and proportions). The question we wish to answer may then be phrased: \emph{is $p_{\theta_0}$ an appropriate model for $\tilde{x}_1,\ldots,\tilde{x}_n$?}
%We consider some data of interest that live in a space $\mathcal{X}$. We are given a parametric family $(p_\theta)_{\theta \in  \Theta}$ of probability densities over $\mathcal{X}$. For instance $(p_\theta)_{\theta \in  \Theta}$ could be composed of DGMs (in that case, $\theta$ would be neural network weights) or Gaussian mixture models (in that case, $\theta$ would be composed of means, covariances, and proportions).
%
%Upon analysis of an initial, in-distribution training dataset, we learn a suitable $\theta_0 \in  \Theta $ using any inference technique, for example maximum likelihood. We are now given additional data points $\tilde{x}_1,\ldots,\tilde{x}_n$ and wonder whether or not they are likely to come from the same distribution that generated our training dataset. The case $n=1$ corresponds to the case where we need to decide if a \emph{single} data point is out-of-distribution; this hardest setting will be of particular interest.
%The question we wish to answer may then be phrased: \emph{is $p_{\theta_0}$ an appropriate model for $\tilde{x}_1,\ldots,\tilde{x}_n$?}

We choose to formalize this problem as a \emph{parametric test} whose alternative hypothesis is that $\tilde{x}$ is \emph{out-of-distribution}. More specifically, if we assume that $\tilde{x}_1,\ldots,\tilde{x}_n \sim_{\textup{i.i.d.}} p_{\tilde{\theta}}$ for some unknown $\tilde{\theta} \in \Theta$, we wish to test $\mathcal{H}_0 : \tilde{\theta} = \theta_0$ against $\mathcal{H} : \tilde{\theta} \neq \theta_0$, where the alternative hypothesis $\mathcal{H}$ is that the test points are OOD.
%we wish to test $\mathcal{H}_\textup{id} : \tilde{\theta} = \theta_0$ against $\mathcal{H}_\textup{ood} : \tilde{\theta} \neq \theta_0$.

Many tests have been proposed for this purpose. The three most famous are the \emph{likelihood ratio test} of \citet{neyman1928use}, Rao's \citeyearpar{rao1948large} \emph{score test},
%(also called \emph{Lagrange multiplier test}, mostly by econometricians),
and \emph{the Wald test} \citep{wald1943tests}. These three classics are nicely reviewed by \citet{buse1982likelihood} or by \citet{rao2005score}, who called them the ``Holy Trinity''. A recent and interesting one is the \emph{gradient test} of \citet{terrell2002gradient}, which is reviewed in great detail in Lemonte's \citeyearpar{lemonte2016gradient} monograph.

Let us review the statistics of these four tests:
\begin{itemize}
    \item likelihood ratio statistic is $S_{LR} = 2 ( \ell ( \hat{\theta}) -\ell ( \theta_0))$,
    \item Wald statistic is $S_{W} = (\hat{\theta} - \theta_0)^T I(\hat{\theta}) (\hat{\theta} - \theta_0)$,
    \item score statistic is $S_{S} = \nabla \ell ( \theta_0)^T I(\theta_0)^{-1}  \nabla \ell ( \theta_0)$,
    \item gradient statistic is $S_{G} = \nabla \ell ( \theta_0)^T (\hat{\theta} - \theta_0)$,
\end{itemize}
where $ \ell ( \theta) = \log p_\theta (\tilde{x}_1,\ldots,\tilde{x}_n)$ is the likelihood function, 
$I({\theta}) = \mathbb{E}_{p_\theta}[\nabla \ell ( \theta)\nabla \ell ( \theta)^T]$ is the Fisher information matrix (FIM), and $ \hat{\theta} \in \argmax_{\theta \in \Theta} \ell ( \theta)$.

The likelihood ratio statistic, the Wald statistic and the gradient statistic all require to fit a model on the additional datapoints $\tilde{x}_1,\ldots,\tilde{x}_n$ in order to compute either $\ell(\hat{\theta})$ or $\hat{\theta}$. 
%
%In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint. \cite{xiao2020likelihood} try to do that in the context of VAEs by re-fitting only the inference network, or encoder, of a VAE to every single additional example. This is the typical thing to do when dealing with out-of-sample data in VAEs, as argued by \cite{mattei2018refit, cremer2018inference}. 
In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint. \cite{xiao2020likelihood} did this for a variational autoencoder (VAE, \citealp{kingma2013auto,rezende2014stochastic}) by only re-fitting inference network (or encoder) to the additional example, which is a typical approach to dealing with out-of-sample data in VAEs, as argued by \cite{cremer2018inference} and \citet{mattei2018refit}.
However, much of the recent works in the literature \citep{ren2019likelihood, schirrmeister2020understanding, serra2019input} mainly focus on deriving different versions of what they call a likelihood ratio statistic.

%In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint, which barely sounds serious. Most of the recent works in the literature mainly focus on deriving different versions of what they call a likelihood ratio statistic. Among those that trained a likelihood-based model on $\tilde{x}_1,\ldots,\tilde{x}_n$, we note that the work by \cite{xiao2020likelihood} is the one that is closer to the real definition of the likelihood ratio statistic given by \cite{neyman1928use}. Indeed, they propose to compute $\ell(\hat{\theta})$ by refitting the inference network of a VAE to every single additional example as proposed in \cite{mattei2018refit, cremer2018inference}. 
%While a similar approach could potentially be used to compute also the gradient statistic and the Wald statistic, this will be possible only for a VAE model. 
%
We tried to derive a general way to compute both the Wald statistic and the gradient statistic, by computing $ \hat{\theta}$ with a few steps of a gradient-based optimization algorithm initialized at $ \theta_0$, but this resulted in a very unstable update leading to computational issues (results not shown). Therefore, in this work we focus on studying the relevance of the score statistic for performing out-of-distribution detection since it is the only statistic that does not require fitting an additional model to the OOD data.

\subsection{Maximum mean discrepancy for out-of-distribution detection}\label{sec:MMD}
Another way of approaching out-of-distribution detection from a testing perspective is through a \emph{two-sample test}. Denoting $p_\textup{data}$ the true training data distribution, the goal is to test $\mathcal{H}_0: \tilde{x}_1,\ldots,\tilde{x}_n \sim p_\textup{data}$ against $\mathcal{H}: \tilde{x}_1,\ldots,\tilde{x}_n \not\sim p_\textup{data}$, where the alternative hypothesis $\mathcal{H}$ again is that the test points are OOD.

A popular way of building statistics for two-sample tests is to use a measure of distance between $p_\textup{data}$ and the distribution of $\tilde{x}_1,\ldots,\tilde{x}_n$. The key idea here will be to use the trained generative model to build this measure of distance. To this end, we will use the \emph{maximum mean discrepancy (MMD)} of \citet{gretton2012kernel}, which is a kernel-based measure of distance. Then, $p_\theta$ will be used to specify an appropriate kernel.

More specifically, given a kernel whose feature map is $\Phi : \mathcal{X} \to \mathcal{H}$, the MMD between two distributions $P$ and $Q$ over $\mathcal{X}$ is defined as
\begin{equation}
    \textup{MMD}_\Phi(P,Q) = \| E_{X\sim P}[\Phi(X)] - E_{Y\sim Q}[\Phi(Y)] \|_\mathcal{H}. \hspace{-.1em}
\end{equation}
In our context, the test statistics will be of the form
\begin{multline}
    \label{eq:mmd}
    \textup{MMD}_\Phi\left(\frac{1}{m}\sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \\\left\|\frac{1}{m}\sum_{i=1}^{m} \Phi(x_i) -  \frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)\right\|_\mathcal{H},
\end{multline}
where $\Phi$ is a kernel feature map built using the generative model and $x_1,\ldots,x_m$ is the training data, i.e.\@ samples from  $p_\textup{data}$. When $\mathcal{H}$ is a simple finite-dimensional Hilbert space and $\Phi$ can be computed easily, then \eqref{eq:mmd} can be computed by going through the data and computing the means in an online fashion.% While the definition seems to imply that we can only perform group-sample OOD detection, it is straightforward to notice that any test statistic of the form of \cref{eq:mmd} can be used also for single-test OOD detection when $n=1$.%by just substituting  $\frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)$ with simply $\Phi(\tilde{x})$.

%\subsection{How to choose $\Phi$?}
As always with kernel methods, a key question is how to choose the kernel, or its feature map $\Phi$.
Here, we want to use the trained generative model $p_\theta$ to build our kernel feature map $\Phi$.

\paragraph{The Fisher kernel} An important example of kernel based on a generative model is the \emph{Fisher kernel} of \citet{jaakkola1999exploiting}. The embedding of this kernel is the Fisher score 
\begin{equation}
    \label{eq:Fisher_score}
    \Phi_\textup{Fisher} (x) = I(\theta)^{-\frac{1}{2}} \nabla \log p_\theta (x),
\end{equation}

and the corresponding reproducing kernel Hilbert space norm is just the $\ell_2$ norm: $|| \cdot || _\mathcal{H} = || \cdot ||_2$. In the case of the Fisher kernel, this means that \cref{eq:mmd} becomes:
\begin{multline}
    \label{eq:mmd_fisher}
    \textup{MMD}_{\Phi_\textup{Fisher}}\left(\frac{1}{m} \sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \\ \left\|\frac{I(\theta)^{-\frac{1}{2}}}{m} \sum_{i=1}^{m}  \nabla \log p_\theta (x_i) - \frac{I(\theta)^{-\frac{1}{2}}}{n}  \sum_{i=1}^{n}  \nabla \log p_\theta (\tilde{x}_i)\right\|_2.
\end{multline}
We will see later that MMD with a Fisher kernel is closely related to the score statistic. In \cref{appendix:mahalanobis}, we additionally show that another popular OOD metric known as the \emph{Mahalanobis score} \citep{lee2018simple} can be interpreted as a MMD statistic with a certain Fisher kernel.

\paragraph{The typicality kernel} A very simple approach of embedding the data using $p_\theta$ is to choose $\Phi_\textup{Typical}(x) = \log p_\theta (x)$. Then, MMD is exactly equivalent to the \emph{typicality test statistic} of \citet{nalisnick2019detecting}, although this connection was not explicitly stated  by \citet{nalisnick2019detecting}. Because of this, we call the kernel $k(x,y) = \log p_\theta (x) \cdot \log p_\theta (y)$ the \emph{typicality kernel}.  %We could also call this "new" kernel the "likelihood kernel".
%While $\Phi_\textup{Typical}$ is not as well motivated as the Fisher kernel, we found that it generally gives good results. 
While $\Phi_\textup{Typical}$ is not as well motivated as a kernel as $\Phi_\textup{Fisher}$, the concepts of typicality and typical set can be used to explain unintuitive behaviours of probability distributions in high-dimensional space as highlighted by \cite{nalisnick2018deep}. We also found that using this kernel generally gives good results for OOD tasks. An interesting analysis that we d not consider in this paper would be to study the properties of this kernel.

In general, neither of these two kernels are characteristic, meaning that our MMD can be zero even if the distributions are not identical. This could be solved by combining them with a characteristic kernel, as in \cite{liu2020learning}, at the price of including a new hyperparameter.

%By looking at the definition of the two considered test statistics we can see that these are both one-sided. We can expect them to be larger when the data are more likely to be OOD. This allows us to compute $p$-values by simply using the empirical CDF, which is hyperparameter-free.

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  SECTION 3   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Combining different test statistics}
\label{sec:combination_p_values}
For single-sample OOD detection, \citet{zhang2021understanding} proved that there is not a single statistic that is constantly better compared to all the possible alternatives of interest. For this reason, we believe that using a combination of different test statistics should lead to an overall better OOD detection in settings where a single statistic might fail. Assume we compute $k$ different test statistics $T_1,\dots, T_k$, each testing $\mathcal{H}_0$ against $\mathcal{H}$ as defined in \cref{sec:parametric,sec:MMD}. The goal is to combine these different tests into a single statistical test that ideally will perform better than the initial single tests. However, different tests can have different magnitudes and they can differ also in the direction of out-of-distribution detection, i.e.\@ for some statistics having a higher values is associated with being OOD, while for other smaller values are OOD. This makes a combination non-trivial.

\citet{morningstar2021density} proposed the density of states estimator (DoSE) to overcome this problem. They only focused on the single-sample detection task, i.e.\@ $n=1$ following our problem formulation. Their idea is to fit different nonparametric density estimators, such as a kernel-density estimator (KDE) or a one-class support vector machines (SVM), for each different statistic $T_1,\dots, T_k$ by using the values computed on the training set examples. For a single test example, $\tilde{x}_1$, they first compute $T_1,\dots, T_k$ and then combine those statistics by summing the different KDEs log-density. %This approach can be used when combining both one-sided and two-sided test statistic. 
While this approach can be used for any type of statistic, and thus is more general, it uses less prior information. Indeed, if we use only statistics that are truly one-sided, then we assume that a method that leverages the true nature of the statistics should work better. In addition to that, fitting a KDE introduces an additional hyperparameter.

In our work, instead, we propose a different approach and leverage the fact that we use only one-sided test statistics. This setting is a well-studied problem in the literature both for independent \citep{fisher1925statistical, folks1971asymptotic} and dependent one-sided test statistics \citep{brown1975400, wilson2019harmonic}. All these approaches rely on the computation of $p$-values of each statistic for the test set $\tilde{x}_1,\ldots,\tilde{x}_n$. This corresponds to computing
$p_j = \textup{Pr}(T_j > t_j \mid \mathcal{H}_0)$, i.e.\@ the probability that the $j$'th test is bigger than the observed value under the null hypothesis $\mathcal{H}_0$, where we assume that each $T_j$ has a continuous distribution. Using $p$-values also solves the problem of the statistics having different scales. Indeed, $p$-values transform the different test statistics into the unit interval.

\paragraph{Computation of $p$-values} 
We want to approximate the distribution of the $p$-values $p_1, \ldots, p_k$ of $\tilde{x}_1,\ldots,\tilde{x}_n$ under the null hypothesis $\mathcal{H}_0$. When $\mathcal{H}_0$ is true, then $p_j$ is uniformly distributed on the interval $[0,1]$. To succeed in this, we should be able to compute $p_j = \textup{Pr}(T_j > t_j \mid \mathcal{H}_0)$, therefore we need to estimate the distribution of each statistic $T_j$ under $\mathcal{H}_0$. %As done by \citet{nalisnick2019detecting}, we assume the existence of a validation set $\mathbf{X}'$ that was not used to train our generative model. Instead of bootstrapping $K$ new datasets $\{X^{'}_k\}_{k=1}^{K}$ of size $M$, as they did in their work, we directly evaluate each test statistic $T_j$ on every single validation example. Asymptotically, this is equivalent to creating $K$ new datasets of size $M=1$ when $K \rightarrow \infty$. 
As done by \citet{nalisnick2019detecting}, we assume the existence of a validation set $\mathbf{X}'$ that was not used to train our generative model. From $\mathbf{X}'$ we bootstrap $S$ new datasets $\{\mathbf{X}^{'}_s\}_{s=1}^{S}$ of size $M'$ by using bootstrap resampling. When $n$ is small, for example $n=1$ or $n=2$, where $n=1$ corresponds to single-sample OOD detection, and the validation set is big, a convenient alternative to bootstrapping is to directly evaluate each test statistic $T_j$ on every single validation example. Asymptotically, this is equivalent to creating $S$ new datasets of size $M'=1$ when $S \rightarrow \infty$. In case of $n=2$, i.e.\@ two-samples OOD detection, and a big validation set we can simply bootstrap without resampling.
We then use these values to estimate the empirical distribution function (eCDF) of the considered statistic $T_j$ under $\mathcal{H}_0$. To obtain the $p$-values of test examples $\tilde{x}_1,\ldots, \tilde{x}_n$ for the test statistic $T_j = t_j$, we simply compute $p_j = 1 - \textup{Pr}(T_j < t_j \mid \mathcal{H}_0)$ using the eCDF.

\paragraph{Combining test statistics by combining $p$-values}
Fisher's \citeyearpar{fisher1925statistical} method is a procedure to combine different $p$-values $p_1,\dots, p_k$. This method assumes that all the considered test statistics are independent, and \citet{folks1971asymptotic} proved that it is asymptotically optimal among all methods of combining independent tests.
%
Given $T_1,\dots, T_k$ and corresponding $p$-values $p_1,\dots, p_k$, Fisher's method combines the $p$-values into a test statistic $X^2$ defined as
\begin{equation}
\label{eq:Fisher_method}
    X^2 \sim -2 \sum_{j=1}^{k} \ln (p_j).
\end{equation}
In case all null-hypotheses are accepted, the resulting test statistic $X^2$ follows a chi-squared distribution with $2k$ degrees of freedom. 
%
In the \cref{appendix:harmonic} , we also consider the Harmonic mean $p$-value \citep{wilson2019harmonic} as a way to combine $p$-values from different statistics. This method usually works best when the statistics are not independent.

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  SECTION 4   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{From test statistics to practical out-of-distribution scores}
\label{sec:section4}

Several of the test statistics that we consider make use of the inverse of the Fisher information matrix $I(\theta)$. The true Fisher information matrix requires an identifiable model to be invertible \citep{watanabe2009algebraic} and computing its inverse is $\mathcal{O}(m^3)$, where $m$ is the number of model parameters. For DGMs, the Fisher information matrix might not be invertible due to the fact that DGMs typically do not satisfy the identifiability condition. Also, the inversion may be computationally impractical, since state-of-the-art DGMs involve very high-dimensional parameter spaces $\Theta$. For the same reason, storing $I(\theta)$ can also be challenging.

We replace it by using a proxy matrix that has to be easy to compute and invert. A first idea is to simply replace $I(\theta)$ by the identity matrix. A more refined way is to look for a diagonal approximation. In \cref{appendix:cheap}, we describe cheap ways of computing such approximations. In particular, we will study two cases: the case where $I(\theta)$ is replaced by the identity matrix and the case where $I(\theta)$ is replaced by a diagonal matrix estimated using the training data.

A possible third option would be to estimate the diagonal of $I(\theta)$ using samples from the model. However, for autoregressive models as the PixelCNN, sampling is a sequential procedure and therefore it is computationally expensive to generate many samples when the input-space is high-dimensional. For this reason, we do not consider it in this work. More complex and precise approximations of the FIM exists, such as the Kronecker-factored Approximate Curvature (K-FAC, \citealp{martens2015optimizing}), but these are not defined for all types of layers used by state-of-the-art models.

\paragraph{On the difficulty of computing per-example gradients} Both the diagonal approximation of the FIM and the computation of the MMD with Fisher kernel of \cref{eq:mmd_fisher} require the gradient computation for all training and test examples. This is known as a costly procedure. For example, if we have to compute the gradient for $N$ examples using a simple fully connected network with $l$ layers of size $p$, the naive procedure of using a batch-size of dimension 1 is $\mathcal{O}(Nlp^2)$ \citep{goodfellow2015efficient}. While more efficient per-example gradient computations were proposed \citep{goodfellow2015efficient, rochette2019efficient}, these techniques can only be applied on simple fully connected or convolutional networks. While for this paper we relied on the naive solution of looping through every example one at the time, a more efficient solution is provided by the BackPACK library \citep{dangel2020backpack} which allows to compute the gradient with respect each sample in a minibatch.

\subsection{Relationship between MMD with Fisher kernel and the score statistic and gradient norm} 
Depending on the choice of the Fisher information approximation, we can notice that there is a strong connection between the MMD using a Fisher kernel, the score statistic and the gradient norm in terms of expected OOD performance. Let us start by looking at the case where we approximate $I(\theta)$ with a diagonal matrix estimated using the training data. At the maximum likelihood estimate, we have that $\mathbb{E}[ \nabla \log p_\theta (x)] = 0$, i.e.\@ the first term inside the norm is $0$. Therefore, we expect that the differences between the OOD scores computed by using \cref{eq:mmd_fisher} will be preserved if we only consider $\left\|I(\theta)^{-1/2} \nabla \log p_\theta (\tilde{x}_1,\dots,\tilde{x}_n)\right\|_2$,
%\begin{equation}
%\left\|I(\theta)^{-1/2} \nabla \log p_\theta (\tilde{x}_1,\dots,\tilde{x}_n)\right\|_2 =  \sqrt{\nabla \log p_\theta (\tilde{x}_1,\dots,\tilde{x}_n)^T I(\theta)^{-1}  \nabla \log p_\theta (\tilde{x}_1,\dots,\tilde{x}_n)},
%\end{equation}
which corresponds to the square root of the score statistic. Since taking the square root still preserves the difference between values, we can expect that the MMD using a Fisher kernel will perform closely to the score statistic. The same reasoning also holds in case we replace the FIM with an identity matrix. In this specific case, instead, we will get that $\left\|{I}(\theta)^{-1/2} \nabla \log p_\theta (\tilde{x}_1,\dots,\tilde{x}_n)\right\|_2 = \left\|\nabla \log p_\theta (\tilde{x}_1,\dots,\tilde{x}_n)\right\|_2 $, which corresponds to considering the gradient norm.

Computationally speaking, considering the score statistic instead of the MMD Fisher lets us avoid going through the entire training set to compute the average gradient (first term in \cref{eq:mmd_fisher}) while carrying the same information. Therefore, in this paper, we will mainly focus on the combination of the typicality test and the score statistic.

\subsection{Why does it make sense to combine the score statistic and the typicality test?}

Let us discuss our choice of combining the score statistic and the typicality test. We will try to look in which situations one of the test fails and the other works and vice versa. Both examples assume that the in-distribution data follows a $\mathcal{N}(0,I_D)$ distribution, and that the correct model has been learned by fitting $(\mathcal{N}(\theta,I_D))_{\theta \in \mathbb{R}^D}$ via maximum-likelihood. Even in this simple setting with no model misspecification, we will see that the two statistics that we consider may have very different strengths.

In this simple Gaussian case, the score statistic can be computed exactly and will be $||\tilde{x}_1 + \ldots + \tilde{x}_n ||_2^2 $. On the other hand, the typicality statistic will be $| (||\tilde{x}_1||_2^2 + \ldots + ||\tilde{x}_n||_2^2)/(2\cdot n) -  D/ 2| $. One interesting regime is the very high-dimensional one ($D \to \infty$). Indeed, by the law of large numbers, these random statistics become deterministic quantities.

\paragraph{Typicality fails, the score succeeds}

Assume that we have two independent OOD data samples that follow a product of truncated normal distributions, with density proportional to $$\mathcal{N}(x | 0,I_D) \cdot  \mathbf{1}\{x_1>0,\ldots,x_D>0\}.$$

We denote by $T_\text{score}^\text{ood},T_\text{score}^\text{id}$ and $T_\text{typicality}^\text{ood},T_\text{typicality}^\text{id}$ the statistics obtained when confronted with either OOD data from the truncated normal, or the in-distribution data. While these statistics are random in general, they will become deterministic when $D \to \infty$, by virtue of the law of large numbers.

For the typicality statistic, these two OOD samples will be indistinguishable from Gaussian ones. Indeed, when $D \to \infty$, both $T_\text{typicality}^\text{ood}$ and $T_\text{typicality}^\text{id}$ will be $\mathcal{O}(D)$.
%
On the other hand, for the score, one can show that
\begin{equation}
T_\text{score}^\text{ood} - T_\text{score}^\text{id} \sim 2 D \mu_\text{TN}^2,
\end{equation}
where $\mu_\text{TN}>0$ is the mean of the truncated normal distribution.

\paragraph{Typicality succeeds, the score fails}

Let us now consider as the OOD distribution a Dirac distribution with mean $0$. Suppose that we see a single sample from this distribution. In this case, the score statistic will be $0$, and will therefore not detect that the point is actually OOD. However, when $D$ is large, the typicality test will be able to declare that this point is anomalous, as shown by \citet{nalisnick2019detecting}.

Therefore, we have that the typicality test and the score statistic are complementary and measure a different type of information. In \cref{appendix:correlation}, we empirically show that they are not correlated, by plotting the two measures against each other and by computing the correlation matrix.

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  SECTION 5: RELATED WORKS   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related works}
\label{sec:related_works}

Since \citet{nalisnick2018deep} and \cite{hendrycks2018deep}, different test statistics or methodologies for OOD detection using DGMs were proposed. Most of the recent solutions were highly influenced by three major lines of work: \emph{typicality set}, \emph{likelihood ratio} test statistics, and \emph{model misestimation}.

The typicality set hypothesis was introduced by \citet{nalisnick2019detecting} as a possible explanation for the DGMs assigning higher likelihood to OOD data. The typicality set is the subset of the model full support where the model samples from and this does not intersect with the region of higher likelihood. While the typicality test was introduced for batch-OOD detection, \cite{morningstar2021density} shows that it also works well in the single-sample case. This is also confirmed by our own experiments.

The likelihood ratio test statistic method by \citet{ren2019likelihood} assumes that every input is composed by a background component and a semantic component. For OOD detection, only the semantic component matters. In addition to a model trained on the in-distribution data, they proposed to train a background model on perturbed inputs data and then for each test example consider as OOD score the likelihood ratio between the two models. \cite{schirrmeister2020understanding}, instead, trained the background model on a more general distribution of images by considering 80 million general tiny images. Similarly to these approaches, \cite{serra2019input} argued that the failure of DGMs is due to the high-influence that the input complexity has on the likelihood. Therefore, they proposed to use a general lossless image compression algorithm as a background model.
All these methods, however, require additional knowledge of the OOD data for either choosing an image augmentation procedure to perturb the input data or for choosing a specific compressor.

Another line of works blame the models themselves and not the test statistics. \cite{zhang2021understanding} argued that model misestimation is the main cause of higher likelihood assigned to OOD data. This can be due to both the model architecture and the maximum likelihood objective. \cite{kirichenko2020normalizing} and \cite{schirrmeister2020understanding} showed that normalizing flows can achieve better OOD performance despite achieving a worse likelihood if one changes some model design choices.
%
Other works in the literature focused on deriving specific test statistics that works only for a specific model, for example for VAEs \citep{xiao2020likelihood, maaloe2019biva, havtorn2021hierarchical}, or for normalizing flows \citep{kirichenko2020normalizing, ijcai2021-292}.

%All the solutions we presented can be described in terms of the following two categories: \emph{model-agnostic} vs \emph{model-specific} test statistics and statistics \emph{with OOD assumptions} vs statistics \emph{with no OOD assumptions}. In this work, we are mostly interested in model-agnostic scores with no OOD assumptions, i.e.\@ scores that can be directly computed with any models and that do not need any explicit assumption on the OOD data.

As mentioned in the introduction, we frame the OOD detection problem in terms of statistical tests problem. Recently, \cite{haroush2021statistical} showed that adopting hypothesis testing at the layer and channel level of a neural network can be used for OOD detection in the discriminative setting. They used both Fisher's method and Simes' method to combine class-conditional $p$-values computed for each convolutional and dense layer of a deep neural network. We focus on the unsupervised setting using DGMs and use hypothesis testing on statistics that can be computed on all differentiable DGM. As already explained in section \cref{sec:combination_p_values}, \cite{morningstar2021density} considered the combination of different statistics for OOD detection. The main difference with their approach is that we propose statistics that can be applied to any differentiable generative model and combine them by using Fisher's method, which takes advantage of using only one-sided independent statistics. Concurrently, \cite{choi2021robust} derived the score statistic by starting from the likelihood ratio statistic and applying a Laplace approximation. They computed the score statistic only for certain layers of the model and for a specific example, the OOD score is given by the infinity norm of these different layer scores after a ReLU operation. Our procedure differs both in the derivation of the score statistic and its usage since we compute the score statistic for the entire model.

\begin{figure*}[tbp]
    \includegraphics[width=\textwidth]{imgs/pvalues_comb_no_drop_correct_font.pdf}
    \caption{\emph{First plot}: $p$-values of the typicality test on the two test sets. We can see that under $\mathcal{H}_0$, they should be uniformly distributed. \emph{Second plot}: $p$-values of the score statistic. \emph{Third plot}: values obtained by the Fisher's method. In red, we plot the density function of a $\chi^2$-distribution with 4 degrees. This shows that the statistics are independent. \emph{Fourth plot}: $p$-values obtained of the combination. These plots refer to a PixelCNN++ trained on FashionMNIST without dropout.}
    \label{fig:comb_p_values}
\end{figure*}

\begin{table*}[tb]
    %\renewcommand{\thetable}{4}
    \caption{AUROC$\uparrow$ for single-sample OOD detection. For Fisher's method we mean the combination of the typicality test and the test statistic. These are also combined using DoSE.}
    \resizebox{\textwidth}{!}{
        \scriptsize
        \begin{tabular}{ccccccc}
            \toprule
            &\multicolumn{6}{c}{\textsc{FashionMNIST (in) / MNIST (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{PixelCNN++} (dropout)  &  0.0762 & 0.8709 & 0.8314 & \textbf{0.8822} & \textbf{0.9369}  & 0.8822 \\
            \textsc{PixelCNN++} (no dropout)  & 0.1048 &  \textbf{0.9532} &  0.7575 &  0.9381 & \textbf{0.9536} &  0.9382 \\
            \textsc{Glow} (RMSProp) &  0.1970 & 0.8904  & 0.4807 &  \textbf{0.9114} & 0.8598  &  \textbf{0.8901}\\
            \textsc{Glow} (Adam)  & 0.1223 & 0.7705 & 0.6987 &  \textbf{0.8745} & \textbf{0.8839}  & 0.8752 \\
            \textsc{HVAE}  & 0.2620 & 0.8714 &0.4884 & \textbf{0.9578} & 0.9383 & \textbf{0.9498} \\
            \bottomrule
            & & & & & & \\
            \toprule
            &\multicolumn{6}{c}{\textsc{CIFAR10 (in) / SVHN (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{PixelCNN++} (model1)  & 0.1553 & \textbf{0.8006} &  0.6457 & 0.6407 & \textbf{0.6826} & 0.6571 \\
            \textsc{PixelCNN++} (model2) & 0.1567 & \textbf{0.7923} & 0.6498 &  0.7067 & \textbf{0.7300} & 0.7243 \\
            \textsc{Glow} (RMSProp)  &  0.0630 & 0.8585 & \textbf{0.8651} &  0.7940 &  \textbf{0.8683} & 0.8510 \\
            \textsc{Glow} (Adam)   &  0.0627 & 0.7844 &  \textbf{0.8624} &  0.7655 &  \textbf{0.8613} &  0.8588 \\
            \textsc{HVAE}  & 0.0636 & 0.8067  & \textbf{0.8679} & 0.7335 & \textbf{0.8603} & 0.8179 \\
            \bottomrule
            & & & & & & \\
            \toprule
            &\multicolumn{6}{c}{\textsc{CIFAR10 (in) / CIFAR100 (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{PixelCNN++} (model1)  & 0.5153 & 0.5306 & \textbf{0.5458} & 0.5362 & \textbf{0.5563}  & 0.5477\\
            \textsc{PixelCNN++} (model2) & 0.5150 &  0.5230 & \textbf{0.5455} & 0.5325  & \textbf{0.5543} & 0.5453 \\
            \textsc{Glow} (RMSProp)  &  0.5206 & 0.5547 & 0.5507 & 0.\textbf{5801} & \textbf{0.5844} &  \textbf{0.5842} \\
            \textsc{Glow} (Adam)   & 0.5206 & 0.5593 & 0.5508  & \textbf{0.5692} & \textbf{0.5775} &  \textbf{0.5767} \\
            \textsc{HVAE}  & 0.5340 & 0.5280 &  0.5493 &  \textbf{0.5798} &  0.5879 &  \textbf{0.5941}\\
            \bottomrule 
        \end{tabular}
        \label{tab:single_sample_results}
    }
    \vspace*{-\baselineskip}
\end{table*}

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  SECTION 5: EXP SETUP   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental Setup}

To evaluate the performance of the combination of the typicality test and the score statistic in detecting OOD data, we follow the experiments of \citet{nalisnick2018deep, hendrycks2018deep} and considered the OOD detection task on three image dataset pairs that have been proven challenging for DGMs, i.e.\@ FashionMNIST \citep{xiao2017fashion} vs MNIST \citep{lecun1998mnist}, CIFAR10 \citep{krizhevsky2009learning} vs SVHN \citep{netzer2011reading}, and CIFAR10 vs CIFAR100. \cite{winkens2020contrastive} divide these tasks into \textit{far}-OOD tasks, where the in-distribution and out-distribution are different such as in the case of CIFAR10 against SVHN, and \textit{near}-OOD where the two distributions are pretty similar, such as CIFAR10 and CIFAR100. \textit{Near}-OOD tasks are usually most challenging.

For each task, we trained three different state-of-the-art DGMs, a PixelCNN++ \citep{salimans2017pixelcnn++}, a Glow model \citep{kingma2018glow}, and a hierarchical variational autoencoder \citep{kingma2013auto,rezende2014stochastic} with bottom-up inference (HVAE, \citealp{burda2015importance}). These are DGMs parametrized by neural networks that make different assumptions in the modelling choice of the target distribution. In addition to that, for PixelCNN++ and Glow we have a tractable likelihood while for HVAE we can only estimate a lower bound. A more in-depth description of these methods and additional results testing MNIST against FashionMNIST and SVHN against CIFAR10 can be found in \cref{appendix:mnist_svhn}. We also extensively analyzed, focusing mostly in the influence of the preprocessing, the results on CIFAR10 vs CelebA \citep{liu2015faceattributes} in \cref{appendix:celeba}. In \cref{appendix:gmm_ppca}, we also considered a Gaussian Mixture Model and a Probabilistic PCA as simple generative models.

\paragraph{Models} To analyze the effect of model architecture choices and optimization choice, we also consider different versions of the same model that reaches a similar log-likelihood. We consider 5 different models for each dataset pair. On FashionMNIST, we consider two Glow models, one trained using Adam and one using RMSProp and two PixelCNN++, trained with and without dropout. For CIFAR10, we consider two different PixelCNN++, one trained by us (model1) and one using a checkpoint given by the repository we used\footnote{ \href{https://github.com/pclucas14/pixel-cnn-pp}{\texttt{https://github.com/pclucas14/pixel-cnn-pp}}} (model2), and two Glow models (Adam and RMSProp). For both datasets, instead, we consider only one HVAE.

\paragraph{Baselines} We are mostly interested in testing our methods with other model-agnostic test statistics in the literature. Apart from using the plain likelihood as an OOD score, the only test statistic we are aware of that can be applied to any generative model without requiring any background model or OOD assumptions is the typicality test statistic of \cite{nalisnick2019detecting}. We also considered the gradient norm, which in general seem to work well but fails in the case of SVHN vs CIFAR10 (see \cref{appendix:mnist_svhn}). In addition to that, we compare our methods to a model-agnostic version of DoSE by \cite{morningstar2021density}, where we used KDEs to combine the score statistic and the typicality test statistic.

\paragraph{Evaluation} We compare our methods with the baselines by computing the area under the receiver operating characteristic curve (AUROC) as done in previous works \citep{hendrycks2018deep, ren2019likelihood, morningstar2021density}. We also evaluate our methods in terms of False Discovery Rate (FDR) control \cite{benjamini1995controlling}, i.e.\@ the proportion of false positive among the rejected hypothesis. Note that both quantities need to know the true label (OOD or in-distribution) to be computed.

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  SECTION 5: RESULTS   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

\paragraph{One-sample OOD}
We first evaluate our proposed method in the single-sample OOD detection task. Results are summarized in \cref{tab:single_sample_results}. We start by considering the OOD task on FashionMNIST against MNIST. Looking at the single statistics, we notice that the score statistic is the one that works the best and the combination of the typicality test and the score statistic usually improve the AUROC than the two standalone statistics. In addition to that, it is better than the combination of the two statistics by using a KDE. DoSE seems to perform better on Glow trained with RMSProp, where the typicality is failing.

On natural images, instead, we have a different trend. The typicality test is better than the score statistic overall. The gradient norm surprisingly performs well in the two dataset pairs, but it fails badly when the model is trained on SVHN (see \cref{appendix:mnist_svhn}). Regarding the combination of the two statistics, the Fisher's method is always better than DoSE, but in this setting, it improves over the best of the single statistics three out of five times.
%
In the \textit{near}-OOD task, we have that both our method and DoSE using our suggested statistics perform closely. We want to highlight that for this challenging task we get results that are comparable with those reported in \cite{morningstar2021density}, but by using two model-agnostic statistics instead of three model-specific ones.
%
It can be noticed that the way we train our models has a strong influence on both the typicality test and the score statistic, although the models get the same test log-likelihood. In \cref{appendix:checkpoints}, we also show that this can happen between different checkpoints of the same model.

In \cref{fig:comb_p_values}, we show that the $p$-values distributions for both the typicality and the score statistic are uniformly distributed under the null-hypothesis and that the combination under the null follows a $\chi^2$ distribution with 4 degrees of freedom. This also supports the fact that the typicality test and the score statistic are independent.

\paragraph{Two-sample OOD}
As \citet{nalisnick2019detecting}, we consider how these test statistics change when performing two-sample OOD detection. Results are summarized in \cref{tab:two_sample_results}. As shown by \citet{nalisnick2019detecting}, the typicality improves but also the score statistic gets better if we consider more samples. Combining those leads to an improvement of performance in terms of AUROC with almost all the models. When training on FashionMNIST, the model can almost perfectly distinguish between the in-distribution test set and the OOD test set. While the performance improves for the two \textit{far}-OOD task, we have that the improvement is slightly less evident in the \textit{near}-OOD task of CIFAR10 vs CIFAR100.

\begin{table}[t]
\scriptsize
\caption{AUROC$\uparrow$ for two-sample OOD detection using the usual considered model.}\normalsize
    %\renewcommand{\thetable}{4}
    \resizebox{0.5\textwidth}{!}{
        \scriptsize
        \begin{tabular}{ccccc}
            \toprule
            &\multicolumn{4}{c}{\textsc{FashionMNIST (in) / MNIST (out)}}\\
            \cmidrule{2-5}
            \textsc{models}  & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{PCNN++} (drop.) &  0.9514 & 0.9828 & \textbf{0.9934} & \textbf{0.9912} \\
            \textsc{PCNN++} (no drop)  & 0.9081 &   0.9853 & \textbf{0.9916} &  \textbf{0.9921} \\
            \textsc{Glow} (RMSProp) & 0.6190 &  \textbf{0.9588} &  0.9187 & 0.7201  \\
            \textsc{Glow} (Adam)  & 0.8525 &  \textbf{0.9716} & \textbf{0.9708} & \textbf{0.9736} \\
            \textsc{HVAE}  & 0.6634 & \textbf{0.9881} & \textbf{0.9837} &  \textbf{0.9889}\\
            \bottomrule
            & & & &\\
            \toprule
            &\multicolumn{4}{c}{\textsc{CIFAR10 (in) / SVHN (out)}}\\
            \cmidrule{2-5}
            \textsc{models}  & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{PCNN++} (m1) & 0.7675 & 0.6555 & \textbf{0.7800} &  0.7046\\
            \textsc{PCNN++} (m2) & 0.7720 &   0.7235 & \textbf{0.8227} & 0.7850\\
            \textsc{Glow} (RMSProp)  & 0.9497 &  0.8624 & \textbf{0.9536} & 0.9379\\
            \textsc{Glow} (Adam) & 0.9480 &  0.8370 & \textbf{0.9519} & 0.9329\\
            \textsc{HVAE} &  \textbf{0.9623} &  0.7754 &  0.9560 & 0.9133\\
            \bottomrule
            & & & &\\
            \toprule
            &\multicolumn{4}{c}{\textsc{CIFAR10 (in) / CIFAR100 (out)}}\\
            \cmidrule{2-5}
            \textsc{models}  & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{PCNN++} (m1) &  0.5433 & 0.5450 &  \textbf{0.5540} &  \textbf{0.5508}\\
            \textsc{PCNN++} (m2) &  0.5435 &  0.5370 &  \textbf{0.5533} & 0.5470 \\
            \textsc{Glow} (RMSProp)  & 0.5550 &  \textbf{0.6211}  & 0.6165 &  \textbf{0.6233}\\
            \textsc{Glow} (Adam) & 0.5558  &   0.6073 & 0.6083 &   \textbf{0.6117}\\
            \textsc{HVAE} & 0.5594  & 0.6188  &  \textbf{0.6218} & \textbf{0.6273} \\
            \bottomrule
        \end{tabular}
        \label{tab:two_sample_results}
    }
    \vspace*{-\baselineskip}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%% FIGURE
\begin{figure}[tb]
    \centering
    \includegraphics[scale=0.47]{imgs/fashion_hvae_type1_correct_font.pdf}
    \caption{Type I (probability of an inlier to be classified as outlier) and Type II (probability of an outlier to be considered as inlier) errors versus the significance level $\alpha$ on the combination values. By using Benjamini-Hochberg correction, we get that the Type I error stays below identity line.}
    \label{fig:type1}
\end{figure}

\subsection{Practical OOD detection with FDR control}
One of the advantages of framing the problem as multiple testing is that we have a well-defined procedure to decide on which hypotheses to reject while controlling the False Discovery Rate (FDR, \citealp{benjamini1995controlling}). Imagine we are interested in finding the outliers from the dataset given by the combination of the two test-sets but we do not want to discard too many inliers, then we can use the Benjamini-Hochberg (BH) procedure \citep{benjamini1995controlling} to decide a threshold and reject all hypothesis below that threshold. For a specific significance level $\alpha$, the procedure guarantees that the FDR stays below that level. Therefore, we can guarantee that the rate of inliers that are classified as outliers is less than the chosen $\alpha$.

%As we have mentioned before, when combining $p$-values of $k$ different statistics using the Fisher's method if the null hypothesis is true, then these new scores are $\chi^2_{2k}$. This allows us to easily compute the $p$-values associate to those new values and then apply the Benjamini-Hochberg procedure. Alternatively, one can directly apply the procedure to the $p$-values of a single test-statistic.

We leverage the fact that when the null hypothesis is true and the $p$-values are independent, then the scores obtained by combining $k$ different statistics are $\chi^2_{2k}$ distributed to compute the $p$-values. Alternatively, the procedure can be also applied to the $p$-values of a single test-statistic.
%
Usually, it is better to use a FDR control when it is actually possible to make few false discoveries, i.e.\@ when we have a strong statistic. Therefore, we expect the procedure to work well when the AUROC is good, for examples on models trained on FashionMNIST.

As can be seen in \cref{fig:type1}, we have that the Type I ratio line stays below the identity line, meaning that the BH correction is working. When deciding for a specific threshold $\alpha$, we usually have to trade-off between Type I and Type II error and in most cases the threshold to choose depends on the application domain. Ideally, we would like to have a low Type I and a low Type II error rate, meaning that we are not considering a lot of in-distribution examples as OOD and at the same time considering a lot of outliers as in-distribution. \cref{fig:type1} shows that we can achieve this for low values of $\alpha$. When training on CIFAR, instead, we are able to control the FDR only from a certain significance level (see \cref{appendix:BH_cifar}). This is expected given that the AUROC is not as good as when testing on MNIST.

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  SECTION 5: CONCLUSION   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion and Conclusions}
In this paper we studied the task of out-of-distribution detection using deep generative models and a combination of multiple statistical tests. We tested our method using different state-of-the art DGMs on classic image benchmark for OOD detection. We found that combining the two statistic leads to a more robust score that in some cases is close to state-of-the-art model-specific scores that require more assumptions. We also noticed that both the model design choice and the optimization choices have an influence on the score we are computing.

When considering only one-sided independent statistics, we showed that the Fisher's method tends to works better than combine them by summing the log-density of a KDE. We also noticed that the score statistic tends to perform a bit worse when the number of parameters of the models increases, i.e.\@ in the context of natural images. One possible reason can be that in this setting the diagonal approximation is not good, and therefore one could consider different approximations, such as K-FAC.

DGMs have recently been used for handling missing data (see e.g.\@ \citealp{mattei2019miwae, ma2018eddi, nazabal2020handling, ipsen2020not}). An interesting future direction would be to extend these OOD detection methods to handle missing values.

The methods  presented in this paper can also easily be applied when using model-specific one-sided statistics. In addition to obtain a more accurate score if one want to combine the test statistics, this also allows one to use well-defined procedure to control the FDR when choosing a which example to mark as outliers. Having this control, is necessary when we want to apply these methods in real settings.

\subsubsection*{Acknowledgements}
Federico Bergamin and Pierre-Alexandre Mattei contributed eqaully to this paper, which is indicated by the asterisk (*) in the author list. The work was supported by the Innovation Fund Denmark (0175-00014B and 0153-00167B), the Independent Research Fund Denmark (9131-00082B) and the Novo Nordisk Foundation (NNF20OC0062606 and NNF20OC0065611). Furthermore, it was supported by the French government, through the 3IA Côte d’Azur Investments in the Future project managed by the National Research Agency (ANR) with the reference number ANR-19-P3IA-0002.

}

### 2023-07-20 appendices/appendix_modelagnostic.tex
Add appendix for Bergamin

\chapter[supplementary material for model-agnostic out-of-distribution detection using combined statistical tests]{Supplementary material for Model-Agnostic Out-of-Distribution Detection Using Combined Statistical Tests}
\label{app:modelagnostic}
\ifthenelse{\equal{\skipappendices}{true}}{}{

\section{Crude approximation of the Fisher information}
\label{appendix_modelagnostic:cheap}

The Fisher information is defined as:
\begin{equation}
    I({\theta}) =  \mathbb{E}_{x \sim p_\theta}[\nabla \log {p_\theta}(x) \nabla  \log {p_\theta}(x) ^T].
\end{equation}

A crude diagonal approximation can be computed by simply estimating the diagonal of $I({\theta})$ and setting all off-diagonal elements to zero. Such diagonal approximations have been used in machine learning for decades: for instance, \citet[Section 3.12.2]{lecun1987} used a similar approximation of the Hessian matrix, and called it ``outrageously simplifying". Much more complex approximations have been derived, although diagonal approximations have been consistently used (e.g. by \citealp{kirkpatrick2017overcoming}, who used essentially the same approximation in a supervised context), and are linked to several adaptive optimisation techniques like Adam \citep{kingma2014adam} or RMSProp \citep{tieleman2012lecture}. A good discussion on these issues is provided in Martens's \citeyearpar{martens2020new} recent review.

The approximation we used in the paper works as follows:
\begin{itemize}
    %\item We first sample $x_1,...,x_T\sim p_\theta $ according to the learned generative model (which will be relatively cheap in most cases).
    \item By using the training examples $x_1,...,x_T$, we form the estimate
    $$ {D}_T({\theta}) = \frac{1}{T} \sum_{t=1}^T \textup{Diag}(\nabla \log {p_\theta}(x_t)^2),  $$
    where the square in $\nabla \log p_\theta(x_t)^2$ is computed elementwise.
    \item While we could directly use ${D}_T({\theta})$ as an estimate. A slightly more refined approach is to slightly regularise $D_T({\theta})$. Following \citet{martens2020new}, our final estimate of the Fisher information matrix is
    \begin{equation}
    	\hat{I}_T({\theta}) = (D_T({\theta}) + \varepsilon)^\xi,
    \end{equation}
    with all operations performed elementwise. The diagonal matrix $\hat{I}_T({\theta})$ is then easy to invert and can be used to compute our statistics.
\end{itemize}

\paragraph{How to choose $\varepsilon$ and $\xi$?} The Adam optimizer uses a similar estimate, with default hyperparameters $\varepsilon = 10^{-8}$ and $\xi = 1$. As argued by \citet{martens2020new}, it can be interesting to use $\xi <1$  in order to diminish the influence of extreme values of $D_T({\theta})$. In particular, \citet{martens2020new} suggests taking $\xi = 0.75$. When $\xi \longrightarrow 0$, then $\hat{I}_T({\theta})$ will approach the identity matrix. We tested the two settings by using a PixelCNN++ trained on CIFAR. Results are shown in table~\ref{tab:martens}. In terms of OOD detection, it seems that using $\varepsilon = 10^{-8}$ and $\xi = 1$ is slightly better. All results presented in the paper and in the supplementary material are computed by using $\varepsilon = 10^{-8}$ and $\xi = 1$.

\paragraph{A few notes on the computation of $D_T({\theta})$} While it seems more sensible to use samples $x_1,...,x_m\sim p_\theta $ from the model, we decided to simply reuse the training data $x_1,...,x_T $ instead. There are two computational advantages to this. The first one is that sampling many data points can be expensive (in particular for deep autoregressive models à la PixelCNN). The second advantage is that, if we wish to compute a MMD statistic, such as the MMD with the Fisher kernel or the MMD typicality (that require the average of gradient or the average log-likelihood over the training), computing the average of the square of the gradient costs very little. One can just do a single loop over the data, and use the usual formulas for online estimation of a mean, see \cref{algo_description}.

\paragraph{Do we really need to approximate the diagonal of $I({\theta})$?} 
Another possibility is to just use the identity matrix as FIM instead of of approximating the diagonal through the procedure explained above. In our experiments (see table~\ref{tab:mmd} and table~\ref{tab:mmd_pair2}), we can see that sometimes using the identity matrix seems to work equally well or a bit better for some models trained on FashionMNIST and CIFAR10. However, when we train on SVHN or MNIST, there are a cases where the statistic that is using the identity matrix as approximation fails, sometimes being worse than random chance. In those setting, using the diagonal approximation leads to way better results. Therefore, considering a test statistic that uses the diagonal approximation of the FIM is more robust for OOD detection.

\begin{table*}[tb]
    \centering
    %\renewcommand{\thetable}{4}
    \caption{AUROC$\uparrow$ for single-sample OOD detection. Comparison between two different estimates of the Fisher information matrix. For ($\ddagger$) we used the Adam parameter choice, i.e.\@ $\varepsilon = 10^{-8}$ and $\xi = 1$. For ($\mathsection$), instead, we used  $\varepsilon = 10^{-8}$ and $\xi = 0.75$, as suggested by \citet{martens2020new}. As a results we have that using Adam parameters choice is slightly better for our task.}
    \small
    \begin{tabular}{ccccc}
        \toprule
        &\multicolumn{4}{c}{\textsc{CIFAR10 (in) / SVHN (out)}}\\
        \cmidrule{2-5}
        \textsc{models}  & \textcolor{blue}{\textsc{MMD Diagonal}} &  \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} \\
        \midrule
        \textsc{PixelCNN++} (model2) ($\ddagger$) & 0.7070 & 0.6498  &  0.7067 & 0.7300\\
        \textsc{PixelCNN++} (model2) ($\mathsection$) & 0.6881 &  0.6498  & 0.6878 & 0.7176\\
        \bottomrule
        ($\ddagger$) With $\varepsilon = 10^{-8}$ and $\xi = 1$ \\
        \hspace{0.4cm}($\mathsection$) With $\varepsilon = 10^{-8}$ and $\xi = 0.75$
    \end{tabular}
    \label{tab:martens}
\end{table*}

% \section{Typicality test and score statistic}

% Let's us look at the case when the typicality test fail and the score statistic works example introduced in Section 4.2 of the paper.

% \paragraph{Typicality test and score statistic are uncorrelated} To test if the typicality test and the score statistic are uncorrelated we plot the two scores computed on the validation set. As can be seen from figure \ref{fig:correlatio}, we have that the two measures are not correlated as it is also highlight by the correlation matrix.

% \begin{figure}[htbp]
% \centering
% \begin{subfigure}{.5\textwidth}
    %   \centering
    %   \includegraphics[scale=0.5]{imgs/correlation.png}
    %   %\caption{Kalman Filter $p(x_t\mid y_{1:t})$ and true state observations.}
    % \end{subfigure}%
    % \begin{subfigure}{.5\textwidth}
        %       \centering
        %       \begin{tabular}{|c|c|}
            %       \hline
            %           1. &  -0.014\\ 
            %           \hline
            %           -0.014 & 1. \\ 
            %         \hline
            %       \end{tabular}
            % \end{subfigure}
            % \end{figure}

            % TO DO:
            % \begin{itemize}
                %     \item WRITE "PROOF" OF WHEN THE TYPICALITY FAILS (SKETCHED PROOF ON PIERRE'S MOLESKINE)
                % \end{itemize}

                \section{The Mahalanobis score as MMD}
                \label{appendix_modelagnostic:mahalanobis}

                \citet{lee_simple_2018} introduced a simple metric to perform OOD detection with a trained deep classifier. The key idea is to train a simple generative model (linear discriminant analysis) in the feature space of the classifier. Let $y$ denote the labels, and $z = f(x)$ the data in feature space. In the simplest case, $f$ is just the trained deep net devoid of the last softmax layer. The linear discriminant analysis model is
                \begin{equation}
                    y \sim \textup{Cat}(\pi), \; \; z | y \sim \mathcal{N}(\mu_y, \Sigma),
                \end{equation}
                where $\mu_1,...,\mu_K$ are class-dependent means, $\Sigma$ a common covariance matrix, and $\pi_1,...,\pi_K$ are the class proportions, estimated by maximum-likelihood. The \emph{Mahalanobis score} is then
                \begin{equation}
                    M(x) = \max_{k \in \{1,...,K \}} - (z - \mu_k)^T \Sigma^{-1} (z - \mu_k),
                \end{equation}
                which may be rewritten
                \begin{equation}
                    M(x) = \max_{k \in \{1,...,K \}} p(z|k),
                \end{equation}
                under the assumption of equal class proportions (i.e.\@ $\pi_1 = ... = \pi_K = 1/K$).

We show here that it is possible to re-interpret this score as a MMD score with a certain Fisher kernel. 
The generative model induced on $z$ by linear disciminant analysis is a Gaussian mixture:
\begin{equation}
    p_{\pi, \mu , \Sigma }(z) = \sum_{k=1}^K \pi_k \mathcal{N}(z | \mu_k, \Sigma).
\end{equation}
If we want a powerful deep kernel, it seems somewhat natural to consider the Fisher kernel associated with this generative model. The most important part of this mixture model are arguably the class-specific means (indeed, the model has been trained to discriminate the classes as well as possible). Therefore, we will only include these means in the Fisher kernel, and look at
\begin{equation}
	\Phi_\textup{Fisher} (x) = I(\mu)^{-1/2} \nabla_\mu  \log p_{\pi, \mu , \Sigma }(z),
\end{equation}
assuming that $\pi$ and $\Sigma$ are fixed at their maximum likelihood estimates. Similar mixture-based Fisher kernels have been very popular in the past, and were actually a key element of state-of-the art classification models on Imagenet before deep nets won the competition \citep{perronnin2010improving}. Our idea is to re-use ideas introduced by this computer vision litterature. Under the assumption that the Gaussian clusters are well-separated, \citet{tanaka2013fisher}, extending an earlier analysis of \citet[Appendix A]{sanchez2013image}, showed that
\begin{equation}
	[\Phi_\textup{Fisher} (x)]_{\mu_k} \approx \sqrt{\frac{p(z|k)}{\pi_k }}  \Sigma^{-1/2} (z - \mu_k).
\end{equation}
Now, using the fact that the expected value of the score is approximatively zero, we can write that
\begin{equation}
    \textup{MMD}_{\Phi_\textup{Fisher}}^2 \approx \sum_{k=1}^K  || [\Phi_\textup{Fisher} (x)]_{\mu_k} ||_2^2 \approx \sum_{k=1}^K \frac{p(z|k)}{\pi_k} (z - \mu_k)^T \Sigma^{-1} (z - \mu_k).
\end{equation}

Using again the fact that the clusters are well-separated, we may say that $z|k$ is approximatively a point mass at the most probable label,  i.e.\@ that $p(z|k) \approx \delta_k^{\text{argmax}_c p(z|c)}$. This leads to the approximation 
\begin{equation}
    \textup{MMD}_{\Phi_\textup{Fisher}}^2 \approx \max_{k \in \{1,...,K\}} \frac{1}{\pi_k} (z - \mu_k)^T \Sigma^{-1} (z - \mu_k).
\end{equation}
Finally, assuming that the class proportions are equal leads to the equivalence of $\textup{MMD}_{\Phi_\textup{Fisher}}$ and the Mahalanobis score.

\section{More Information on the experimental setup}

\subsection{A bit more background}
The three considered DGMs are both parametrized by neural networks but they differ in the way they model the data distribution of interest. Assume we are interested in approximating a target distribution $p^*(\mathbf{x})$, for example a distribution of natural images, as it is done when using CIFAR10. PixelCNN++ is an autoregressive model and it models $p^*(\mathbf{x})$ as a product of conditional distribution over the variables, i.e.\@ $p(\mathbf{x}) = p(x_1) \prod_{d=2}^D p(x_d \mid \mathbf{x}_{<d})$, where $\mathbf{x}_{<d} = [x_1, \dots, x_{d-1}]^T$. Glow is a normalizing flow model and it approximate $p^*(\mathbf{x})$ by using a sequence of bijiective transformations starting from a simple distribution, also called base distribution. If we use only a single invertible transformation $f$, the normalizing  flow is defined as $\mathbf{x} = f(\mathbf{z})$, where $\mathbf{z} \sim p_{Z}(\mathbf{z})$, and $p_{X}(\mathbf{x}) = p_{Z}(\mathbf{z})|\det J_f(\mathbf{z}) |^{-1}$, where we used the change of variable formula. For these two types of model we have a tractable likelihood that can be used to optimize the model parameters. The Variational Autoencoder (VAE), instead, is a framework to model the data with a latent variable model, i.e.\@ $p(\mathbf{x}, \mathbf{z}) = p(\mathbf{x} \mid \mathbf{z}) p(\mathbf{z})$, where $\mathbf{x}$ is the observed input data and $\mathbf{z}$ is a stochastic latent variable and the prior distribution $p(\mathbf{z})$ is usually a standard Normal. Since the posterior $p(\mathbf{z}\mid\mathbf{x})$ is not tractable, a variational distribution $q_{\phi}(\mathbf{z}\mid\mathbf{x})$ is used as an approximation. Due to the intractability of the posterior, we cannot directly optimize the likelihood of the model, but instead the model parameters are optimized by maximizing the evidence lower bound (ELBO): $\log p_{\theta}(\mathbf{x}) \geq \mathbb{E}_{q_{\phi}(\mathbf{z}\mid \mathbf{x})} \left[ \log \frac{p(\mathbf{x}, \mathbf{z})}{q_{\phi}(\mathbf{z}\mid \mathbf{x})} \right] \equiv \mathcal{L} $. In this work we are considering an Hierarchical VAE (HVAE) with bottom-up inference as done in \cite{havtorn_hierarchical_2021}. This is an extension of the VAE framework that consider an hierarchy of $L$ latent variables $\mathbf{z} = \mathbf{z}_1,\dots, \mathbf{z}_L$. The bottom-up inference is defined as $q_{\phi}(\mathbf{z}\mid \mathbf{x}) = q_{\phi}(\mathbf{z}_1\mid \mathbf{x})\prod_{i=2}^L q_{\theta}(\mathbf{z}_i \mid \mathbf{z}_{i-1})$, while the generative path is top-down, meaning $p_{\theta}(\mathbf{x}\mid \mathbf{z}) = p(\mathbf{x}\mid \mathbf{z}_1)p_{\theta}(\mathbf{z}_1\mid \mathbf{z}_2)\cdots p_{\theta}(\mathbf{z}_{L-1}\mid \mathbf{z}_L)$. This is still trained by maximizing the ELBO. For a more in-depth explanation of these models we refer to their papers.

\subsection{Generative model details}
We will briefly describe the different model architectures and training procedures used in this paper. Since most of the models are taken from public code repositories and related papers, we will mostly invite the reader to have a look at the cited paper for a more in-depth description of the training details.
For MNIST, CIFAR10, and FashionMNIST we used 3000 examples from the test set as validation set. For SVHN, instead, we used 6032 datapoints from the test set as validation, leaving the remaining 20000 examples as test set. In Table~\ref{tab:log_like}, we reported test log-likelihood of the models used in this paper.

\paragraph{PixelCNN++} For PixelCNN++ we used the code available in this repository\footnote{\href{https://github.com/pclucas14/pixel-cnn-pp}{\texttt{https://github.com/pclucas14/pixel-cnn-pp}}}. For the greyscale images, we used one residual block per stage with 32 filters and 5 logistic components in the discretized mixture of logistics. For natural images, instead, we used 5 residual blocks per stage with 160 filters and 10 components in the mixture. We trained all the models using Adam optimizer.

\paragraph{Glow} For training Glow models we follow \citet{kirichenko_why_2020} and their repository\footnote{\href{https://github.com/PolinaKirichenko/flows_ood}{\texttt{https://github.com/PolinaKirichenko/flows\_ood}}}. They closely follow \citet{nalisnick_deep_2019} and \cite{kingma_glow_2018} implementation for multi-scale Glow, where a scale is defined as the sequence of actorm, invertible $1\times1$ convolution and coupling layers. While \citet{kirichenko_why_2020} only considers the RMSProp optimizer, we trained two different models, one using RMSProp and one using Adam with batch-size 32. For the greyscale dataset our Glow is made up of 2 scales with 16 coupling layers, and a 3-layers highway network with 200 hidden units is used to predict the scale and shift parameters. For CIFAR10 and SVHN, instead, we used 3 scales with 8 coupling layers, and 400 hidden units for the 3-layers highway network. %Since training flow-based models is a bit unstable, in some cases we had to restart the training when the training failed. 
For a more in-depth description, we refer to the codebase and the Appendix C of \citet{kirichenko_why_2020}.

\paragraph{Hierarchical VAE} We follow \cite{havtorn_hierarchical_2021} for both model architecture design and training choices for our hierarchical VAEs. We used their open-sourced repository\footnote{\href{https://github.com/JakobHavtorn/hvae-oodd}{\texttt{https://github.com/JakobHavtorn/hvae-oodd}}}. As mentioned in the paper, the HVAE model we used has a bottom-up inference path and a top-down generative path. We trained each model for 1000 epochs using Adam optimizer with learning rate $3e-4$ and a batch-size of $128$. All models were initialized using the data-dependent initialization and they used weight-normalization \citep{salimans2016weight}. In addition to that, we always consider a hierarchy of three latent variables. For greyscale images (MNIST and FashionMNIST) we used a latent dimension of $8-16-8$ for $\mathbf{z}_1, \mathbf{z}_2, \mathbf{z}_3$ respectively, while for natural images (CIFAR10 and SVHN) we used $8-16-32$. For a more in depth description of the model, we refer to Appendix B of \citet{havtorn_hierarchical_2021}.

% table with log-likelihood (bits/dim) of the models
\begin{table*}[tb]
    %\renewcommand{\thetable}{4}
    \caption{Test log-likelihood (bits/dim) achieved by the models used in the paper.}
    \begin{subtable}{.5\textwidth}
        \centering
        \scriptsize
        \begin{tabular}{cc}
            \toprule
            \multicolumn{2}{c}{\textsc{Models trained on FashionMNIST)}}\\
            \midrule
            \textsc{models}  & \textcolor{black}{\textsc{Log-Likelihood (bits/dim)}}\\
            \midrule
            \textsc{PixelCNN++} (dropout) & 2.75 \\
            \textsc{PixelCNN++} (no dropout) & 2.72 \\
            \textsc{Glow} (RMSProp) &  3.04\\
            \textsc{Glow} (Adam)  & 3.02 \\
            \textsc{HVAE} ($**$) & 0.43 \\
            \bottomrule
            &   \\
            \toprule
            \multicolumn{2}{c}{\textsc{Models trained on CIFAR10}}\\
            \midrule
            \textsc{models}  & \textcolor{black}{\textsc{Log-Likelihood (bits/dim)}}\\
            \midrule

            \textsc{PixelCNN++} (model1)  & 2.94  \\
            \textsc{PixelCNN++} (model2)  & 2.94 \\
            \textsc{Glow} (RMSProp)   & 3.62 \\
            \textsc{Glow} (Adam)  & 3.62  \\
            \textsc{HVAE}  & 3.87 \\
            \bottomrule
            ($**$) Binarized FashionMNIST
        \end{tabular}
    \end{subtable}
    \begin{subtable}{.5\textwidth}
        \centering
        \scriptsize
        \begin{tabular}{cc}
            \toprule
            \multicolumn{2}{c}{\textsc{Models trained on MNIST}}\\
            \midrule
            \textsc{models}  & \textcolor{black}{\textsc{Log-Likelihood (bits/dim)}}\\
            \midrule
            \textsc{PixelCNN++} (dropout) & 0.90 \\
            \textsc{Glow} (RMSProp)  &  1.32 \\
            \textsc{Glow} (Adam)   & 1.30 \\
            \textsc{HVAE} ($**$)   &  0.16\\
            \bottomrule
            &   \\
            \toprule
            \multicolumn{2}{c}{\textsc{Models trained on SVHN}}\\
            \midrule
            \textsc{models}  & \textcolor{black}{\textsc{Log-Likelihood (bits/dim)}}\\
            \midrule
            \textsc{PixelCNN++} (dropout)  & 1.58 \\
            \textsc{Glow} (RMSProp)   & 2.23 \\
            \textsc{Glow} (Adam)  & 2.21  \\
            \textsc{HVAE}  & 2.38 \\
            \bottomrule
            ($**$) Binarized MNIST
        \end{tabular}

    \end{subtable} 
    \label{tab:log_like}
    %\vspace*{-\baselineskip}
\end{table*}

\section{Additional results}

\subsection{Typicality test and score statistic are uncorrelated}
\label{appendix_modelagnostic:correlation}
To test if the typicality test and the score statistic are uncorrelated, we plot the two scores computed on the validation set. As can be seen from figure \ref{fig:corralation}, we have that the two measures are not correlated as it is also highlight by the correlation coefficient.

\begin{figure}[tb]
    \centering
    \includegraphics[scale=0.5]{imgs/correlation_correct_font.pdf}
    \caption{Correlation of Typicality Test and Score Statistic computed on the validation set using a PixelCNN++ trained on FashionMNIST. The correlation coefficient is $-0.014$. This can also be seen by looking at the regression line, which is almost straight. }
    \label{fig:corralation}
\end{figure}

\subsection{Harmonic Mean}
\label{appendix_modelagnostic:harmonic}
In the paper we mentioned that another way to combine $p$-values from different test statistics is the Harmonic mean \citep{wilson_harmonic_2019}. This is defined as:
\begin{equation}
    \mathring{p} = \frac{\sum_{i=1}^{k} w_i}{\sum_{i=1}^{k} w_i / p_i},
\end{equation}
where $w_1,\dots,w_k$ are weights that sum up to 1. In our setting, we considered equal weights, i.e.\@ $w_i = 1 / k$. Therefore, if we simply consider two test statistics $T_1$ and $T_2$ and corresponding $p$-values $p_1$ and $p_2$, the harmonic mean $p$-values becomes:
\begin{equation}
    \mathring{p} = \frac{2 p_1 p_2}{p_1 + p_2}.
\end{equation}

As expected, this combination should work better when the statistics that we are combining are somewhat correlated. Indeed, since in our setting we have that the typicality and the score statistic are independent, we would expect this to work worse than the Fisher's combination. This is confirmed by table~\ref{tab:all_combinations}, where we are reporting the results when combining the two statistics using the three different ways we analyzed.

\subsection{Results considering maximum-mean-discrepancy}
In Section~\ref{sec:section4}, we discussed the relationship between the maximum-mean-discrepancy with a Fisher kernel and the score statistic and the gradient norm, which depends on the choice of approximation of the Fisher information matrix we use. In table~\ref{tab:mmd} we reported also the AUROC scores for the MMD with Fisher kernel considering both the diagonal approximation of the FIM (called \emph{MMD Diagonal} in the table) and the FIM being the identity matrix (called \emph{MMD Identity}). As expected, we have that the AUROC of the MMD with the diagonal approximated FIM is pretty close to the AUROC we obtained by using the score statistic. Likewise, we have that the AUROC of MMD with the identity matrix as FIM is close to the gradient norm when we trained on FashionMNIST and CIFAR10.

So, why did we decide to use the score statistic instead of the MMD with Fisher kernel and diagonal approximation of the FIM? The main reason is Occam's razor. If we have two things that work equally well, we should keep the simplest one. In our case, we have that for computing the MMD with the Fisher kernel, we need to compute both the average gradient and the FIM using the training set. For the score statistic, instead, we just need the FIM. In addition to that, from all our experiments (see table~\ref{tab:mmd} and table~\ref{tab:mmd_pair2}) we do not have any evidence for one statistic working better than the other, because they are always pretty close to each other.

% \begin{table*}[]
%     \centering
%     \begin{tabular}{ |p{3.5cm}||p{1cm}|p{1cm}|p{1cm}|p{1cm}|p{1cm}|p{1cm}||p{1cm}|p{1cm}|p{1cm}|}
%     \hline
%  \multicolumn{10}{|c|}{\textbf{FashionMNIST (in) / MNIST (out)}} \\
%  \hline
%  \multicolumn{1}{|c||}{} &
%  \multicolumn{6}{|c||}{\hspace{1cm} \textbf{Single statistics}} &
% \multicolumn{3}{|c|}{\textbf{Combinations}} \\
% \hline
%         Model & $\log p(x)$ & $\left\|\nabla\right\|$ & MMD F & MMD F id &  TT & Score & FC & HM & $\text{DoSE}_{\text{KDE}}$ \\
%         \hline 
%         PixelCNN++ (dropout) &  0.0762 & 0.8709 & 0.8903 & 0.8690 & 0.8314 & 0.8822 & \textbf{0.9369} & 0.9148 & 0.8822 \\
%         \hline
%         PixelCNN++ (no drop) & 0.1048 &  0.9532 & 0.9393 & 0.9539 & 0.7575 &  0.9381 & \textbf{0.9536} & 0.9392 & 0.9382 \\
%         \hline
%         Glow (RMSProp) &  0.1970 & 0.8904 & \textbf{0.9115} &  0.8986 & 0.4807 &  0.9114 & 0.8598 & 0.8853 &  0.8901\\
%         \hline
%         Glow (Adam) & 0.1223 & 0.7705 & 0.8540 & 0.7217 & 0.6987 &  0.8745 & \textbf{0.8839} & 0.8632 & 0.8752 \\
%         \hline
%         HVAE & 0.0653 & 0.8714 & 0.9574 & 0.8726 & 0.8336 & 0.9578 & \textbf{0.9708} &  0.9569 & 0.9630  \\
%         \hline
%     \end{tabular}
%     \caption{AUROC for single-sample OOD detection}
%     \label{tab:my_label}
% \end{table*}

% \begin{table*}[]
%     \centering
%     \begin{tabular}{ |p{3.5cm}||p{1cm}|p{1cm}|p{1cm}|p{1cm}|p{1cm}|p{1cm}||p{1cm}|p{1cm}|p{1cm}|}
%     \hline
%  \multicolumn{10}{|c|}{\textbf{CIFAR10 (in) / SVHN (out)}} \\
%   \hline
%  \multicolumn{1}{|c||}{} &
%  \multicolumn{6}{|c||}{\hspace{1cm} \textbf{Single statistics}} &
% \multicolumn{3}{|c|}{\textbf{Combinations}} \\
% \hline
%         Model & $\log p(x)$ & $\left\|\nabla\right\|$ & MMD F & MMD F id &  TT & Score & FC & HM & $\text{DoSE}_{\text{KDE}}$ \\
%         \hline 
%         PixelCNN++ (model1) & 0.1553 & 0.8006 & 0.6406 & 0.8126 & 0.6457 & 0.6407 & \textbf{0.6826} & 0.6667 & 0.6571 \\
%         \hline
%         PixelCNN++ (model2) & 0.1567 & 0.7923 & 0.7070 & 0.7955 & 0.6498 &  0.7067 & \textbf{0.7300} & 0.7105 & 0.7243 \\
%         \hline
%         Glow (RMSProp) &  0.0630 & 0.8585 & 0.7929 & 0.8621 & 0.8651 &  0.7940 &  \textbf{0.8683} & 0.8551 & 0.8510 \\
%         \hline
%         Glow (Adam) &  0.0627 & 0.7844 & 0.7620 &  0.7838 &  \textbf{0.8624} &  0.7655 &  \textbf{0.8613} & 0.8493 & 0.8588 \\
%         \hline
%         HVAE & 0.0455 & 0.8041 &  0.7268 & 0.7634 & \textbf{0.8845} & 0.7334 & 0.8699 &  0.8525 & 0.8245  \\
%         \hline
%     \end{tabular}
%     \caption{AUROC for single-sample OOD detection}
%     \label{tab:my_label}
% \end{table*}

\begin{table*}[tb]
    %\renewcommand{\thetable}{4}
    \caption{AUROC$\uparrow$ for single-sample OOD detection. In this table we consider all the different single statistics we mentioned in the paper. One can notice that MMD Diagonal is pretty close to the score statistic and the MMD Identity is close to the gradient norm, as expected (see Section 4.1 in the paper).}
    \resizebox{\textwidth}{!}{
        \scriptsize
        \begin{tabular}{ccccccc}
            \toprule
            &\multicolumn{6}{c}{\textsc{FashionMNIST (in) / MNIST (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{6}{c}{\textcolor{blue}{\textsc{single statistics}}}\\
            \cmidrule{2-7}
            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{MMD Diagonal}}  & \textcolor{blue}{\textsc{MMD Identity}} &\textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}}  \\
            \midrule
            \textsc{PixelCNN++} (dropout) & 0.0762 & 0.8709 & 0.8903 & 0.8690 & 0.8314 & 0.8822 \\
            \textsc{PixelCNN++} (no dropout)  & 0.1048 &  0.9532 & 0.9393 & 0.9539 & 0.7575 &  0.9381\\
            \textsc{Glow} (RMSProp) &  0.1970 & 0.8904 & 0.9115 &  0.8986 & 0.4807 &  0.9114\\
            \textsc{Glow} (Adam)  &0.1223 & 0.7705 & 0.8540 & 0.7217 & 0.6987 &  0.8745\\
            \textsc{HVAE}  &0.0653 & 0.8714 & 0.9574 & 0.8726 & 0.8336 & 0.9578 \\
            \bottomrule
            & & & & & & \\
            \toprule
            &\multicolumn{6}{c}{\textsc{CIFAR10 (in) / SVHN (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{6}{c}{\textcolor{blue}{\textsc{single statistics}}}\\
            \cmidrule{2-7}
            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{MMD Diagonal}}  & \textcolor{blue}{\textsc{MMD Identity}} &\textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}}  \\
            \midrule

            \textsc{PixelCNN++} (model1)  & 0.1553 & 0.8006 & 0.6406 & 0.8126 & 0.6457 & 0.6407 \\
            \textsc{PixelCNN++} (model2) & 0.1567 & 0.7923 & 0.7070 & 0.7955 & 0.6498 &  0.7067\\
            \textsc{Glow} (RMSProp)  & 0.0630 & 0.8585 & 0.7929 & 0.8621 & 0.8651 &  0.7940\\
            \textsc{Glow} (Adam)   & 0.0627 & 0.7844 & 0.7620 &  0.7838 &  0.8624 &  0.7655 \\
            \textsc{HVAE}  & 0.0455 & 0.8041 &  0.7268 & 0.7634 & 0.8845 & 0.7334\\
            \bottomrule
        \end{tabular}
        \label{tab:mmd}
    }
    \vspace*{-0.85\baselineskip}
\end{table*}

\begin{table*}[tb]
    \centering
    %\renewcommand{\thetable}{4}
    \caption{AUROC$\uparrow$ for single-sample OOD detection. Comparison between the three method we mentioned to combine different statistics. Since the typicality and the score statistic are not correlated, we have that the Fisher's method is mostly working better than the other two methods.}
        \scriptsize
        \begin{tabular}{cccc}
            \toprule
            &\multicolumn{3}{c}{\textsc{FashionMNIST (in) / MNIST (out)}}\\
            \cmidrule{2-4}
            & \multicolumn{3}{c}{\textcolor{red}{\textsc{combinations}}}\\
            \cmidrule{2-4}
            \textsc{models}  & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{Harmonic mean}} &  \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}}  \\
            \midrule
            \textsc{PixelCNN++} (dropout) & 0.9369 & 0.9148 & 0.8822 \\
            \textsc{PixelCNN++} (no dropout)  & 0.9536 & 0.9392 & 0.9382\\
            \textsc{Glow} (RMSProp) &  0.8598 & 0.8853 &  0.8901\\
            \textsc{Glow} (Adam)  & 0.8839 & 0.8632 & 0.8752\\
            \textsc{HVAE}  & 0.9708 &  0.9569 & 0.9630\\
            \bottomrule
            & & &  \\
            \toprule
            & \multicolumn{3}{c}{\textsc{CIFAR10 (in) / SVHN (out)}}\\
            \cmidrule{2-4}
            & \multicolumn{3}{c}{\textcolor{red}{\textsc{combinations}}}\\
            \cmidrule{2-4}
            \textsc{models}  & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{Harmonic mean}} &  \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}}  \\
            \midrule

            \textsc{PixelCNN++} (model1)  & 0.6826 & 0.6667 & 0.6571  \\
            \textsc{PixelCNN++} (model2) & 0.7300 & 0.7105 & 0.7243 \\
            \textsc{Glow} (RMSProp)  &  0.8683 & 0.8551 & 0.8510\\
            \textsc{Glow} (Adam) & 0.8613 & 0.8493 & 0.8588  \\
            \textsc{HVAE}  & 0.8699 &  0.8525 & 0.8245 \\
            \bottomrule
        \end{tabular}
        \label{tab:all_combinations}
    \vspace*{-\baselineskip}
\end{table*}

\subsection{Variability within the same model in different checkpoints}
\label{appendix_modelagnostic:checkpoints}
As mentioned in the paper, we noticed that all statistics depend on choices we made about our model and the training procedure, such as deciding between Adam or RMSProp, or between using dropout or not. In addition to that, we find out that they can differ also within the same model at different checkpoints that obtain almost the same log-likelihood. Here we consider two Glow models, one trained with Adam and one using RMSProp on CIFAR10. For both, we consider two checkpoints that achieve the same test log-likelihood. Those trained with Adam get a log-likelihood of $3.63$ bits/dim, while the ones trained with RMSProp get $3.62$ bits/dim. Results are shown in Table~\ref{tab:variability}. It can be noticed, that although the models are similar in terms of test bits/dim the statistics vary a lot, mostly when training with RMSProp.

\begin{table*}[tb]
    \centering
    %\renewcommand{\thetable}{4}
    \caption{AUROC$\uparrow$ for single-sample OOD detection. In this table we are comparing two different Glow models trained on CIFAR10 by considering two different checkpoints with almost the same test log-likelihood. We can see that both statistics vary a bit. }
        \scriptsize
        \begin{tabular}{ccccc}
            \toprule
            &\multicolumn{4}{c}{\textsc{CIFAR10 (in) / SVHN (out)}}\\
            \cmidrule{2-5}
            & \multicolumn{2}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-3}\cmidrule(l){4-5}
            \textsc{models}  &  \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{Glow} (RMSProp) \{\emph{check1}\} & 0.8651 &  0.7940  &  0.8683 &  0.8510\\
            \textsc{Glow} (RMSProp) \{\emph{check2}\} & 0.8532 &  0.6894  & 0.8275 & 0.7815\\
            \midrule
            \textsc{Glow} (Adam)  \{\emph{check1}\} & 0.8624 &  0.7655 & 0.8613 & 0.8588\\
            \textsc{Glow} (Adam) \{\emph{check2}\} & 0.8558 & 0.7327 & 0.8402 & 0.8303\\
            \bottomrule
        \end{tabular}
        \label{tab:variability}
    \vspace*{-\baselineskip}
\end{table*}

\subsection{Benjamini-Hochberg procedure when training on CIFAR10}
\label{appendix_modelagnostic:BH_cifar}
In the main paper we focused on the Benjamini-Hochberg procedure applied to a model trained on FashionMNIST. Although one should use a False Discovery Rate control procedure when the statistics we are using are strong, for completeness, we will present what happens when we apply the BH procedure on a model trained on CIFAR10. In Fig.~\ref{fig:type1_cifar}, we report the Type I error ratio and the Type II error ratio for different significance levels $\alpha$. We can see that we can actually control the FDR for $\alpha > 0.2$, and for these significance levels we are actually controlling the FDR. What is happening for $\alpha < 0.2$? We have that the procedure is only rejecting 5 hypotheses and all these hypotheses corresponds to in-distribution examples. Therefore, we have that the ratio of Type I error is still low, but we are making a lot of Type II errors because we are accepting all the examples whose hypotheses should be rejected.

%%%%%%%%%%%%%%%%%%%%%%% FIGURE
\begin{figure}[tb]
    \centering
    \includegraphics[scale=0.5]{imgs/cifar_hvae_type1_correct_font.pdf}
    \caption{Type I and Type II errors versus the significance level $\alpha$ on the combination values. We can control the FDR only for $\alpha > 0.2$ in this case. For $\alpha >0.2$, since we are using Benjamini-Hochberg procedure, we get that the Type I error stays below identity line.}
    \label{fig:type1_cifar}
     \vspace*{-\baselineskip}
\end{figure}

% \subsection{False positive and False negative examples}

\subsection{Results when training on MNIST and SVHN}
\label{appendix_modelagnostic:mnist_svhn}
We also evaluated our methods in the two dataset pairs, MNIST against FashionMNIST and SVHN against CIFAR10, that are usually considered easier than the tasks presented in the main paper. For both tasks, we trained two Glow models, one trained with Adam and one trained with RMSProp, one PixelCNN++ trained with dropout and a hierarchical-VAE. Results are reported in table~\ref{tab:results_second}. We can see that almost all the statistics we considered are able to almost perfectly distinguish between the in-distribution test-set and the OOD test-set. However, we can notice that the gradient norm is failing sometimes both when we trained on CIFAR10 and when we trained on FashionMNIST. From table~\ref{tab:mmd_pair2}, instead, it is clear that we need to approximate the diagonal of the Fisher Information Matrix because if we simply consider the identity matrix, this will also fail as the gradient norm is doing.

\begin{table*}[tb]
    %\renewcommand{\thetable}{4}
    \caption{AUROC$\uparrow$ for single-sample OOD detection when training on MNIST and testing again FashionMNIST and when training on SVHN and testing against CIFAR10. As before, Fisher's method is the combination of the typicality test and the test statistic. These are also combined using DoSE.}
    \resizebox{\textwidth}{!}{
        \scriptsize
        \begin{tabular}{ccccccc}
            \toprule
            &\multicolumn{6}{c}{\textsc{MNIST (in) / FashionMNIST (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{PixelCNN++} (dropout) ($\dagger$)  & 0.9999  & 0.8534 & 0.9996 & 0.9993 & 0.9999  & 0.9999 \\
            \textsc{Glow} (RMSProp) & 0.9997  & 0.9936  & 0.9991 & 0.9936 &  0.9992 & 0.9994\\
            \textsc{Glow} (Adam)  &0.9999 & 0.6506 & 0.9995 &  0.9992 &  0.9998 &  0.9999\\
            \textsc{HVAE}  & 0.9999 & 0.9998 &  0.9997 &  0.9999 &  0.9999  & 0.9999\\
            \bottomrule
            & & & & & & \\
            \toprule
            &\multicolumn{6}{c}{\textsc{SVHN (in) / CIFAR10 (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{PixelCNN++} (dropout)  & 0.9820 & 0.2670 &  0.9590 &  0.9543 & 0.9914 & 0.9824 \\
            \textsc{Glow} (RMSProp)  &  0.9917  & 0.9180 & 0.9830 & 0.9823 & 0.9913  & 0.9913 \\
            \textsc{Glow} (Adam)   &0.9913 &  0.5658 & 0.9779 & 0.9641 & 0.9883 & 0.9863 \\
            \textsc{HVAE}  & 0.9943 &  0.1011  &  0.9857 &  0.9862 & 0.9934 &  0.9862 \\
            \bottomrule
            ($\dagger$) Trained using 50000 datapoints
        \end{tabular}
        \label{tab:results_second}
    }
\end{table*}

\begin{table*}[tb]
    %\renewcommand{\thetable}{4}
    \caption{AUROC$\uparrow$ for single-sample OOD detection. In this table we consider all the different single statistics we mentioned in the paper but for the models trained on MNIST and SVHN this time. In this case, it is important to notice that the gradient norm and the MMD identity sometimes fail to a different extent.}
    \resizebox{\textwidth}{!}{
        \scriptsize
        \begin{tabular}{ccccccc}
            \toprule
            &\multicolumn{6}{c}{\textsc{MNIST (in) / FashionMNIST (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{6}{c}{\textcolor{blue}{\textsc{single statistics}}}\\
            \cmidrule{2-7}
            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{MMD Diagonal}}  & \textcolor{blue}{\textsc{MMD Identity}} &\textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}}  \\
            \midrule
            \textsc{PixelCNN++} (dropout) ($\dagger$)  & 0.9999  & 0.8534 & 0.9993 & 0.8608 & 0.9996  & 0.9993 \\
            \textsc{Glow} (RMSProp) & 0.9997  & 0.9936  & 0.9942 & 0.6609 &  0.9991 & 0.9936\\
            \textsc{Glow} (Adam)  &0.9999 & 0.6506 & 0.9993 &  0.9124 &  0.9997 &  0.9992\\
            \textsc{HVAE}  & 0.9999 & 0.9998 &  0.9999 &  0.9999 &  0.9999  & 0.9999\\
            \bottomrule
            & & & & & & \\
            \toprule
            &\multicolumn{6}{c}{\textsc{SVHN (in) / CIFAR10 (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{6}{c}{\textcolor{blue}{\textsc{single statistics}}}\\
            \cmidrule{2-7}
            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{MMD Diagonal}}  & \textcolor{blue}{\textsc{MMD Identity}} &\textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}}  \\
            \midrule
            \textsc{PixelCNN++} (dropout)  & 0.9820 & 0.2670 &   0.9543 & 0.3185 & 0.9590 & 0.9543 \\
            \textsc{Glow} (RMSProp)  &  0.9917  & 0.9180 & 0.9824 & 0.9317 & 0.9830  &  0.9823 \\
            \textsc{Glow} (Adam)   &0.9913 &  0.5658 & 0.9653 & 0.7096 & 0.9779 &  0.9641 \\
            \textsc{HVAE}  & 0.9943 &  0.1011  &  0.9865 &  0.4508 & 0.9857 &  0.9862\\
            \bottomrule
            ($\dagger$) Trained using 50000 datapoints
        \end{tabular}
        \label{tab:mmd_pair2}
    }
\end{table*}

\subsection{Application of our method to Gaussian Mixture Model and Probabilistic PCA}
\label{appendix_modelagnostic:gmm_ppca}
Since the method we propose is model-agnostic, we show that it can be used for out-of-distribution detection also using two simple generative models, Gaussian Mixture Model (GMM) and Probabilistic PCA (PPCA). We consider the two pairs of datasets as before, i.e.\@ FashionMNIST vs MNIST and CIFAR10 vs SVHN. Results can be seen in Table~\ref{tab:gmm_results} and Table~\ref{tab:ppca_results}. For both GMM and PPCA trained on FashionMNIST the likelihood can be used to perform OOD detection. Indeed, in this setting, they are not assigning higher likelihood to OOD data as it is the case for DGMs. This happens instead when we fit these models on CIFAR10. However, this behaviour can be due to the fact that they are really poor generative models for this dataset. It is also surprising that when training on CIFAR10 the score statistic is failing in both models. We think that this is also due to the fact that both the GMM and the PPCA are far from being good generative models for this dataset.

\begin{table*}[tb]
    %\renewcommand{\thetable}{4}
    \caption{AUROC$\uparrow$ for single-sample OOD detection using a Gaussian mixture model (GMM). For Fisher's method we mean the combination of the typicality test and the test statistic. These are also combined using DoSE.}
    \resizebox{\textwidth}{!}{
        \scriptsize
        \begin{tabular}{ccccccc}
            \toprule
            &\multicolumn{6}{c}{\textsc{FashionMNIST (in) / MNIST (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{components}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{50}  & 0.6627 & 0.5514  & 0.5196 & 0.8777  & 0.7689 & 0.8152 \\
            \textsc{100}   &  0.6872 & 0.5509 & 0.5575 & 0.8742 & 0.7965 & 0.7989 \\
            \bottomrule
            &  &  &   &   &  & \\
            \toprule
            &\multicolumn{6}{c}{\textsc{CIFAR10 (in) / SVHN (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{components}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{50}   & 0.2335  &   0.6087 & 0.6759 & 0.3512 &  0.6098 & 0.6569 \\
            \textsc{100} & 0.2372 &  0.6136 & 0.6714 & 0.3294  &  0.5898  & 0.6573 \\
            \bottomrule
        \end{tabular}
        \label{tab:gmm_results}
    }
    \vspace*{-\baselineskip}
\end{table*}

\begin{table*}[tb]
    %\renewcommand{\thetable}{4}
    \caption{AUROC$\uparrow$ for single-sample OOD detection using a Probabilistic PCA. For Fisher's method we mean the combination of the typicality test and the test statistic. These are also combined using DoSE.}
    \resizebox{\textwidth}{!}{
        \scriptsize
        \begin{tabular}{ccccccc}
            \toprule
            &\multicolumn{6}{c}{\textsc{FashionMNIST (in) / MNIST (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{components}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{50}  & 0.9727 & 0.9637 & 0.9587 & 0.9505  & 0.9635 & 0.9610 \\
            \textsc{100}   & 0.9557 &  0.9715 & 0.9309 & 0.9626 & 0.9566 & 0.9585\\
            \bottomrule
            &  &  &   &   &  & \\
            \toprule
            &\multicolumn{6}{c}{\textsc{CIFAR10 (in) / SVHN (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{components}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{50}  & 0.0770 & 0.1494  &  0.8468 & 0.1308 & 0.7568 & 0.8210 \\
            \textsc{100}   &  0.0357 &  0.0778 & 0.8944 & 0.0755  &  0.7966 &  0.8830 \\
            \bottomrule
        \end{tabular}
        \label{tab:ppca_results}
    }
\end{table*}

\subsection{More in depth analysis of the variability of the results for different HVAE}
As we have pointed out before, test statistics and consequentially out-of-distribution performances can vary between the same model trained several times on the same dataset. To test the variability of the results shown in the main paper, we trained five different hierarchical VAEs and compute mean and standard deviations of the final AUROC scores. All models have the same architecture and were trained with the same procedure. Results can be found in Table~\ref{table:consistency_results}. For the models trained on CIFAR10, most of the variability in terms of performance is due to the score statistic, which has the highest standard deviation. When training on FashionMNIST, instead, it seems that the typicality performance is the one varying the most between the five models.

\begin{table*}[tb]
    % \vspace{-1.3em}
    \caption{Mean and standard deviation of the performance in terms of AUROC of our method. Quantities are computed by taking the performance of 5 different trained HVAEs both trained on CIFAR10 and FashionMNIST.}
        \centering
        \scriptsize
        \begin{tabular}{cccccc}
            \toprule
            \textsc{$D_{\text{out}}$}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}}  \\
            \midrule
            &\multicolumn{5}{c}{\textsc{HVAE trained on CIFAR10}} \\
            \cmidrule{2-6}
            \textsc{SVHN} & 0.0631 (0.0008) & 0.8711 (0.0028)  & 0.7808 (0.0255) & 0.8844 (0.0140)  & 0.8519 (0.0194)\\
            \textsc{CIFAR100} & 0.5349 (0.0007) & 0.5496 (0.0003)  & 0.5857 (0.0042) & 0.5924 (0.0029)  & 0.5985 (0.0028)\\
            \textsc{CelebA} & 0.9004 (0.0035) & 0.8203 (0.0046) & 0.7565 (0.0369) & 0.8505 (0.0138) & 0.8247 (0.0228)\\
            %\bottomrule
            %\toprule
            \cmidrule{2-6}
            &\multicolumn{5}{c}{\textsc{HVAE trained on FashionMNIST}} \\
            % \cmidrule{2-3}
            % \textsc{$D_{\text{out}}$}  & \textcolor{red}{\textsc{Our method}} & \textcolor{red}{\textsc{DoSE$_{\textup{orig}}$}} \\
            \cmidrule{2-6}%\midrule
            \textsc{MNIST} & 0.2487 (0.0152) & 0.5064 (0.0245) & 0.9532 (0.0084) & 0.9220 (0.01491) & 0.9377 (0.0126)\\
            %\bottomrule
            \bottomrule%\toprule
    \end{tabular}
    \label{table:consistency_results}
    % \vspace*{-\baselineskip}
\end{table*}

\section{Yes, we should talk about CelebA}
\label{appendix_modelagnostic:celeba}
Out-of-distribution detection performance is not only influenced by the model architecture or the training process. Indeed, transformations applied to the input data play an important role by transforming a difficult task into an easier problem where the likelihood can detect OOD data. By looking at the different results for Glow trained on CIFAR10 and tested on CelebA shown in \cite{hendrycks_deep_2019}, \cite{kirichenko_why_2020}, \cite{morningstar_density_2021}, and \cite{ahmadian_likelihoodfree_2021} we can see that the AUROC scores obtain by the plain log-likelihood are pretty different. In \cite{hendrycks_deep_2019} and  \cite{kirichenko_why_2020} the log-likelihood gets a poor performance, confirming that CIFAR10-CelebA is a challenging pair for DGMs, while in \cite{morningstar_density_2021} the likelihood is able to distinguish OOD data. While the main reason for these different results can be due to model implementation and training procedure, we decided to investigate how different transformations can influence OOD detection. Indeed, CelebA examples originally have a shape of $(218, 178, 3)$ and to transform them into $(32, 32, 3)$-shaped images, as CIFAR10, we have to resize them and then crop their center. The resize function is performing an interpolation, therefore we analyze how different interpolation strategies influence the OOD task.

We considered three different interpolations: bilinear (default in PyTorch), Lanczos, and nearest. As can be seen from Fig.~\ref{fig:celeba}, these transformations mostly affect the sharpness of the images. In Table~\ref{table:celeba} we show how the OOD performance changes for our considered models when testing on CelebA where we applied different interpolations. We can notice that when using the bilinear interpolation we get results that are pretty similar to \cite{hendrycks_deep_2019}, \cite{kirichenko_why_2020}, and \cite{ahmadian_likelihoodfree_2021} in terms of likelihood OOD performance. When using the nearest interpolation, instead, we get results that are closer to \cite{morningstar_density_2021}.

In conclusion, with these experiments, we wanted to highlight the importance of reporting the preprocessing steps used in loading CelebA in order to be able to make a fair comparison with the other proposed methods in the literature.

%%%%%%%%%%%%%%%%%%%%%%% FIGURE
\begin{figure}[t]
    \centering
    \includegraphics[scale=0.5]{imgs/celeba_difference_correct_font.pdf}
    \caption{Comparison of different interpolation methods for CelebA dataset.}
    \label{fig:celeba}
     \vspace*{-\baselineskip}
\end{figure}

\begin{table*}[tb]
    %\renewcommand{\thetable}{4}
    \caption{AUROC$\uparrow$ for single-sample OOD detection for CIFAR10 vs CelebA considering all the three interpolations when using CelebA.}
    \resizebox{\textwidth}{!}{
        \scriptsize
        \begin{tabular}{ccccccc}
            \toprule
            &\multicolumn{6}{c}{\textsc{CIFAR10 (in) / CelebA (out) ($\dagger$)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{PixelCNN++} (model1)  &  0.7027 & 0.5856 & 0.5581  & 0.7001 & 0.6450 & 0.6931 \\
            \textsc{PixelCNN++} (model2) & 0.7034  & 0.4298 & 0.5554 & 0.7505 & 0.6879 & 0.7430\\
            \textsc{Glow} (RMSProp)  & 0.5337  & 0.5616 &  0.3926 & 0.6561  & 0.5400 & 0.5866 \\
            \textsc{Glow} (Adam)   & 0.5308  & 0.5820 & 0.3914 & 0.5850  & 0.4818 & 0.5212 \\
            \textsc{HVAE}  &  0.5643 & 0.5214  & 0.4011 & 0.6712 & 0.5483 & 0.5987 \\
            \bottomrule
            & & & & & & \\
            \toprule
            &\multicolumn{6}{c}{\textsc{CIFAR10 (in) / CelebA (out) ($\nparallel$)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{PixelCNN++} (model1)  &  0.8284  & 0.5035 & 0.7399 & 0.6714  & 0.7477 & 0.7123 \\
            \textsc{PixelCNN++} (model2) & 0.8284 & 0.3530 &   0.7370 & 0.70088 & 0.7631 & 0.7446\\
            \textsc{Glow} (RMSProp)  & 0.7556 & 0.4427 &  0.6222 & 0.7865  & 0.7423 & 0.7632 \\
            \textsc{Glow} (Adam)   & 0.7499 & 0.4800 &  0.6177 &  0.6442  &  0.6460 &  0.6467 \\
            \textsc{HVAE}  & 0.7561 &   0.4097 & 0.6051 &  0.6779 &  0.6775 &  0.6772  \\
            \bottomrule
             & & & & & & \\
            \toprule
            &\multicolumn{6}{c}{\textsc{CIFAR10 (in) / CelebA (out) ($\ddagger$)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{PixelCNN++} (model1)  & 0.9270  & 0.4196 & 0.8902 &  0.8320 & 0.9287 &  0.8908 \\
            \textsc{PixelCNN++} (model2) & 0.9270 & 0.3065 & 0.8886 & 0.8448 & 0.9339 & 0.9236 \\
            \textsc{Glow} (RMSProp)  & 0.9364 & 0.5345 &  0.8880 & 0.9286  & 0.9390 & 0.9423 \\
            \textsc{Glow} (Adam)   &  0.9322 & 0.5957 & 0.8829  &  0.8350  & 0.9017 &  0.8933 \\
            \textsc{HVAE}  & 0.8964 & 0.3515  & 0.8158 & 0.7952 &  0.8620 &   0.8455 \\
            \bottomrule
            ($\dagger$) Bilinear interpolation \\
            ($\nparallel$) Lanczos interpolation \\
            ($\ddagger$) Nearest interpolation\\
        \end{tabular}
        \label{table:celeba}
    }
    \vspace*{-\baselineskip}
\end{table*}

\section{Comparison with the original DoSE statistics}
As the last experiment, we study how our proposed method with our model agnostic statistic performs against DoSE using the original statistics proposed in \cite{morningstar_density_2021}. For the VAEs model, they suggested to use the following 5 statistics: the posterior/prior cross-entropy $\text{H}[q_{\phi}( \textbf{z}\mid \textbf{x}), p(\textbf{z})]$, the posterior entropy $\text{H}[q_{\phi}(\textbf{z}\mid \textbf{x})]$, the posterior/prior KL divergence $\text{D}_{\text{KL}}[q_{\phi}(\textbf{z}\mid \textbf{x}) \mid \mid p(\textbf{z})]$, the posterior expected log-likelihood $\mathbb{E}_{q_{\phi}(\textbf{z}\mid \textbf{x})}[\log q_{\phi}(\textbf{z}\mid \mathbf{x})]$, and the log-likelihood $\log \mathbb{E}_{q_{\phi}(\textbf{z}\mid \textbf{x})} \left[ \frac{p_{\theta}(\textbf{x}, \textbf{z})}{q_{\phi}(\textbf{z}\mid \textbf{x})}\right]$. For DoSE on Glow, instead, they considered three metrics: the log-likelihood $p_{X}(\textbf{x}\mid \theta_n)$ and its two components, i.e.\@ the log-probability of the latent variable $p_{Z}(\textbf{z}\mid \textbf{x}, \theta_n)$ and the log-determinant of the Jacobian $\log | \text{J}_{f}(\textbf{x})|$.

In this setting, since DoSE is using statistics that are HVAE and Glow specific, it is not model agnostic anymore. Indeed, we cannot use those statistics also for a PixelCNN++ for example or any other DGM. We want also to highlight that the models used in \cite{morningstar_density_2021} are a bit different from the ones used in this work. For example, they are considering a beta-VAE with only one stochastic layer, while in our case we used a HVAE with 3-stochastic layers.

\begin{table}%[15]
    \vspace{-1.3em}
    \caption{Comparison between our method and DoSE using the original statistics. In these experiments we considered only Glow trained with Adam.}
        \centering
        \scriptsize
        \begin{tabular}{ccc}
            \toprule
            \textsc{$D_{\text{out}}$}  & \textcolor{red}{\textsc{Our method}} & \textcolor{red}{\textsc{DoSE$_{\textup{orig}}$}} \\
            \midrule
            &\multicolumn{2}{c}{\textsc{GLOW trained on CIFAR10}} \\
            \cmidrule{2-3}
            \textsc{SVHN} & \textbf{0.8613}  &  0.7819  \\
            \textsc{CIFAR100} & \textbf{0.5775} & 0.5700 \\
            \textsc{CelebA} & 0.9017 & \textbf{0.9663}  \\
            %\bottomrule
            %\toprule
            \cmidrule{2-3}
            &\multicolumn{2}{c}{\textsc{GLOW trained on FashionMNIST}} \\
            % \cmidrule{2-3}
            % \textsc{$D_{\text{out}}$}  & \textcolor{red}{\textsc{Our method}} & \textcolor{red}{\textsc{DoSE$_{\textup{orig}}$}} \\
            \cmidrule{2-3}%\midrule
            \textsc{MNIST} & 0.8839 & \textbf{0.9568}   \\
            %\bottomrule
            \midrule%\toprule
            &\multicolumn{2}{c}{\textsc{HVAE trained on FashionMNIST}} \\
            % \cmidrule{2-3}
            % \textsc{$D_{\text{out}}$}  & \textcolor{red}{\textsc{Our method}} & \textcolor{red}{\textsc{DoSE$_{\textup{orig}}$}}  \\
            \cmidrule{2-3}%\midrule
            \textsc{MNIST} & 0.9383 & \textbf{0.9762} \\
            %\bottomrule
            %\toprule
            &\multicolumn{2}{c}{\textsc{HVAE trained on CIFAR10}} \\
            % \cmidrule{2-3}
            % \textsc{$D_{\text{out}}$}  & \textcolor{red}{\textsc{Our method}} & \textcolor{red}{\textsc{DoSE$_{\textup{orig}}$}} \\
            \cmidrule{2-3}%\midrule
            \textsc{SVHN} & 0.8605 & \textbf{0.8823}  \\
            \textsc{CIFAR100} & \textbf{0.5888} & 0.5608  \\
            \textsc{CelebA} & \textbf{0.8620} & 0.8203  \\
            \bottomrule
    \end{tabular}
    \label{table:original_dose}
    \vspace*{-\baselineskip}
\end{table}

\section{Algorithmic implementation}
A pseudocode describing step-by-step how to implement our method is given in Algorithm~\ref{algo_description}.

\begin{algorithm}[ht]
%\begin{small}
\begin{algorithmic}
\STATE \textbf{Input}: Training data $\textbf{X} = (x_1,\ldots,x_m)^T$, validation data $\textbf{X}'$, trained model $p_{\theta}(x)$.\\
\textbf{} \\
\textit{Approximation of the diagonal of the Fisher Information Matrix $I(\theta)$ and average log-likelihood $(1/m)\log p_{\theta}(x_1,\ldots,x_m)$, indicated by $L(\theta)$. We do it in an online fashion.} \\
\textbf{Initialize}  ${I(\theta)} = {0}$ and $L(\theta)=0$ \\
\textbf{For all} $i \in \{1,\ldots ,m\}$: \\
\hspace{1.5em} \textbf{Compute} $\log p_{\theta} ({x}_i)$ \\
\hspace{1.5em} \textbf{Compute} $\nabla_{{\theta}}\log p({x}_i \mid \theta)$ \\
\hspace{1.5em} \textbf{Set}  ${I(\theta)} = \frac{1}{i+1} \cdot (i \cdot {I(\theta)} + (\nabla_{{\theta}}\log p_{\theta}({x}_i))^2$) \\
\hspace{1.5em} \textbf{Set}  $L(\theta) = \frac{1}{i+1} \cdot (i \cdot L(\theta) + \log p_{\theta}({x}_i)$) \\
\textbf{} \\
\textit{Estimation of distributions over the test statistics} \\
\textbf{Sample} $S$ $M'$-sized datasets from $\textbf{X}'$ using bootstrap resampling. \\
\textit{(For single-sample OOD we just cycle through each example, see Sec.~\ref{sec:combination_p_values})}\\
\textbf{Initialize}   $T^{\text{typicality}} = [\hspace{0.5mm}]$ and $T^{\text{score}} = [\hspace{0.5mm}]$ \\
\textbf{For every} bootstrapped dataset $\textbf{X}_s' = (x_1, \dots, x_{M'})^T$: \\
\hspace{1.5em} \textbf{Compute} $\frac{1}{m'} \sum_{m'=1}^{M'}\log p_{\theta}({x}_{m'})$ \\
\hspace{1.5em} \textbf{Compute} $\frac{1}{m'} \sum_{m'=1}^{M'}\nabla_{{\theta}}\log p_{\theta}({x}_{m'})$ \\
\hspace{1.5em} \textbf{Compute} MMD Typicality for ${x}_{m'}$ by $\left\| \frac{1}{m'} \sum_{m'=1}^{M'}\log p_\theta ({x}_{m'}) - L(\theta) \right\|_2$ and add it to $T^{\text{typicality}}$\\
\hspace{1.5em} \textbf{Compute} Score Statistic for ${x}_{m'}$ by $\left\|{I(\theta)}^{-1/2} \frac{1}{m'} \sum_{m'=1}^{M'}\nabla \log p_\theta ({x}_{m'})\right\|_2$ and add it to $T^{\text{score}}$\\
\textbf{Return} Two vectors of size $S$ containing the two statistics for $T^{\text{typicality}}$ and $T^{\text{score}}$  \\
\textbf{} \\
\textbf{Compute} $\hat{F}^{\text{typicality}}$ and $\hat{F}^{\text{score}}$, the two empirical CDFs, from $T^{\text{typicality}}$ and $T^{\text{score}}$. For example, we used \texttt{statsmodels} library \citep{seabold2010statsmodels}. \\
\textbf{} \\  
%\textit{For any test sample} \\
\textbf{Given} a test set $\tilde{x}_1, \dots, \tilde{x}_n$:\\
\textit{($n=1$ corresponds to perform single-sample OOD detection)} \\ 
\hspace{1.5em} \textbf{Compute} $\frac{1}{n} \sum_{i=1}^{n}\log p_{\theta}(\tilde{{x}}_i)$ and $\frac{1}{n} \sum_{i=1}^{n}\nabla_{{\theta}}\log p_{\theta}(\tilde{{x}}_i)$ \\
\hspace{1.5em} \textbf{Compute} MMD Typicality $\tilde{t}$ and Score Statistic $\tilde{s}$ \\
\hspace{1.5em} \textbf{Compute} $p$-values $p_{T} = 1 - \hat{F}^{\text{typicality}}(\hat{t})$ and $p_{S} = 1 - \hat{F}^{\text{score}}(\tilde{s})$ \\
\hspace{1.5em} \textbf{Combine} the two $p$-values using Fisher's method Eq.~\ref{eq:Fisher_method}
\end{algorithmic}
\caption{Computing $p$-values for OOD detection using a trained generative model.}
\label{algo_description}
%\end{small}
\end{algorithm}

}

### 2023-07-20 chapters/paper_modelagnostic.tex
Revise skipping

@@
 \label{chp:paper-modelagnostic}
-\ifthenelse{\equal{\skippaperbody}{true}}{}{
+\ifthenelse{\equal{\skippapers}{true}}{}{
 
@@
 
-The ability to recognise when data are anomalous, i.e.\@ if they originate from a distribution different from that of the training data, is a necessary property for machine learning models for safe and reliable applications in the real world.
-Historically, \citet{bishop1994novelty} proposed to use a one-sided threshold on the log-likelihoods of a learned model as a decision rule to identify outliers in a dataset.
-However, recently, \citet{nalisnick2018deep, hendrycks2018deep} showed that state-of-the-art deep generative models (DGMs) failed in this task, assigning higher a likelihood to out-of-distribution (OOD) data than in-distribution data.
-Most of the recent works focused on proposing new test statistics to alleviate the problem of using the plain likelihood, see \cref{sec:related_works} for details.
+The ability to recognise when data are anomalous, i.e.
+if they originate from a distribution different from that of the training data, is a necessary property for machine learning models for safe and reliable applications in the real world.
+Historically, \textcite{bishop_novelty_1994} proposed to use a one-sided threshold on the log-likelihoods of a learned model as a decision rule to identify outliers in a dataset.
+However, recently, \textcite{nalisnick_deep_2019,hendrycks_deep_2019} showed that state-of-the-art deep generative models (DGMs) failed in this task, assigning higher a likelihood to out-of-distribution (OOD) data than in-distribution data.
+Most of the recent works focused on proposing new test statistics to alleviate the problem of using the plain likelihood, see \cref{sec_modelagnostic:related_works} for details.
 
-We believe that OOD detection should be formulated as statistical hypothesis testing \citep{nalisnick2019detecting, ijcai2021-292, haroush2021statistical}.
-Since the power of a single test depends on the out-distribution \citep{zhang2021understanding}, we propose to approach this problem by using a combination of multiple statistical tests.
+We believe that OOD detection should be formulated as statistical hypothesis testing \parencite{nalisnick_detecting_2019, ahmadian_likelihoodfree_2021, haroush_statistical_2021}.
+Since the power of a single test depends on the out-distribution \parencite{zhang_understanding_2021}, we propose to approach this problem by using a combination of multiple statistical tests.
 %While this does not solve completely the problem, because the power of the combined tests will also depend on the out-distribution, we think that we can get better performance than using a single-statistics especially in situations where one statistic fails.
@@
     \item Empirically we show the complementarity of the typicality and the score statistics and that their combination leads to a robust score for anomaly detection.
-%We evaluate our proposed method with two PixelCNN++ \cite{salimans2017pixelcnn++}, two Glow models \cite{kingma2018glow} and one Hierarchical-VAE (HVAE)\cite{kingma2013auto}.
+%We evaluate our proposed method with two PixelCNN++ \textcite{salimans_pixelcnn_2017}, two Glow models \textcite{kingma_glow_2018} and one Hierarchical-VAE (HVAE)\textcite{kingma_autoencoding_2014}.
 \end{itemize}
@@
 Note that in this problem formulation, the case $n=1$ corresponds to the situation where we need to decide if a \emph{single} data point is out-of-distribution.
-This hardest setting will be of particular interest, and this is also the main focus of recent work, see \cref{sec:related_works}.
+This hardest setting will be of particular interest, and this is also the main focus of recent work, see \cref{sec_modelagnostic:related_works}.
 
-\subsection{Parametric tests for out-of-distribution detection}\label{sec:parametric}
+\subsection{Parametric tests for out-of-distribution detection}\label{sec_modelagnostic:parametric}
 The typical approach is to consider a parametric family $(p_\theta)_{\theta \in  \Theta}$ of probability densities over $\mathcal{X}$ and learn a suitable $\theta_0 \in  \Theta$ using any inference technique, for example maximum likelihood, and the clean data $x_1,\ldots,x_m$.
@@
 
-Many tests have been proposed for this purpose. The three most famous are the \emph{likelihood ratio test} of \citet{neyman1928use}, Rao's \citeyearpar{rao1948large} \emph{score test},
+Many tests have been proposed for this purpose. The three most famous are the \emph{likelihood ratio test} of \textcite{neyman_use_1928}, Rao's \citeyearpar{rao_large_1948} \emph{score test},
 %(also called \emph{Lagrange multiplier test}, mostly by econometricians),
-and \emph{the Wald test} \citep{wald1943tests}.
-These three classics are nicely reviewed by \citet{buse1982likelihood} or by \citet{rao2005score}, who called them the ``Holy Trinity''.
-A recent and interesting one is the \emph{gradient test} of \citet{terrell2002gradient}, which is reviewed in great detail in Lemonte's \citeyearpar{lemonte2016gradient} monograph.
+and \emph{the Wald test} \parencite{wald_tests_1943}.
+These three classics are nicely reviewed by \textcite{buse_likelihood_1982} or by \textcite{rao_score_2005}, who called them the ``Holy Trinity''.
+A recent and interesting one is the \emph{gradient test} of \textcite{terrell_gradient_2002}, which is reviewed in great detail in Lemonte's \citeyearpar{lemonte_gradient_2016} monograph.
 
@@
 where $ \ell ( \theta) = \log p_\theta (\tilde{x}_1,\ldots,\tilde{x}_n)$ is the likelihood function, 
-$I({\theta}) = \mathbb{E}_{p_\theta}[\nabla \ell ( \theta)\nabla \ell ( \theta)^T]$ is the Fisher information matrix (FIM), and $ \hat{\theta} \in \argmax_{\theta \in \Theta} \ell ( \theta)$.
+$I({\theta}) = \mathbb{E}_{p_\theta}[\nabla \ell ( \theta)\nabla \ell ( \theta)^T]$ is the Fisher information matrix (FIM), and $ \hat{\theta} \in \arg\max_{\theta \in \Theta} \ell ( \theta)$.
 
@@
 %In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint.
-\cite{xiao2020likelihood} try to do that in the context of VAEs by re-fitting only the inference network, or encoder, of a VAE to every single additional example.
-This is the typical thing to do when dealing with out-of-sample data in VAEs, as argued by \cite{mattei2018refit, cremer2018inference}.
+\textcite{xiao_likelihood_2020} try to do that in the context of VAEs by re-fitting only the inference network, or encoder, of a VAE to every single additional example.
+This is the typical thing to do when dealing with out-of-sample data in VAEs, as argued by \textcite{mattei2018refit, cremer2018inference}.
 In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint.
-\cite{xiao2020likelihood} did this for a variational autoencoder (VAE, \citealp{kingma2013auto,rezende2014stochastic}) by only re-fitting inference network (or encoder) to the additional example, which is a typical approach to dealing with out-of-sample data in VAEs, as argued by \cite{cremer2018inference} and \citet{mattei2018refit}.
-However, much of the recent works in the literature \citep{ren2019likelihood, schirrmeister2020understanding, serra2019input} mainly focus on deriving different versions of what they call a likelihood ratio statistic.
+\textcite{xiao_likelihood_2020} did this for a variational autoencoder (VAE, \citealp{kingma_autoencoding_2014,rezende_stochastic_2014}) by only re-fitting inference network (or encoder) to the additional example, which is a typical approach to dealing with out-of-sample data in VAEs, as argued by \textcite{cremer_inference_2018} and \textcite{mattei_refit_2018}.
+However, much of the recent works in the literature \parencite{ren_likelihood_2019, schirrmeister_understanding_2020, serra_input_2020} mainly focus on deriving different versions of what they call a likelihood ratio statistic.
 
@@
 Most of the recent works in the literature mainly focus on deriving different versions of what they call a likelihood ratio statistic.
-Among those that trained a likelihood-based model on $\tilde{x}_1,\ldots,\tilde{x}_n$, we note that the work by \cite{xiao2020likelihood} is the one that is closer to the real definition of the likelihood ratio statistic given by \cite{neyman1928use}.
-Indeed, they propose to compute $\ell(\hat{\theta})$ by refitting the inference network of a VAE to every single additional example as proposed in \cite{mattei2018refit, cremer2018inference}.
+Among those that trained a likelihood-based model on $\tilde{x}_1,\ldots,\tilde{x}_n$, we note that the work by \textcite{xiao_likelihood_2020} is the one that is closer to the real definition of the likelihood ratio statistic given by \textcite{neyman_use_1928}.
+Indeed, they propose to compute $\ell(\hat{\theta})$ by refitting the inference network of a VAE to every single additional example as proposed in \textcite{mattei2018refit, cremer2018inference}.
 %While a similar approach could potentially be used to compute also the gradient statistic and the Wald statistic, this will be possible only for a VAE model. 
@@
 
-\subsection{Maximum mean discrepancy for out-of-distribution detection}\label{sec:MMD}
+\subsection{Maximum mean discrepancy for out-of-distribution detection}\label{sec_modelagnostic:MMD}
 Another way of approaching out-of-distribution detection from a testing perspective is through a \emph{two-sample test}.
@@
 The key idea here will be to use the trained generative model to build this measure of distance.
-To this end, we will use the \emph{maximum mean discrepancy (MMD)} of \citet{gretton2012kernel}, which is a kernel-based measure of distance.
+To this end, we will use the \emph{maximum mean discrepancy (MMD)} of \textcite{gretton_kernel_2012}, which is a kernel-based measure of distance.
 Then, $p_\theta$ will be used to specify an appropriate kernel.
@@
 In our context, the test statistics will be of the form
-\begin{multline}
-    \label{eq:mmd}
-    \textup{MMD}_\Phi\left(\frac{1}{m}\sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \\\left\|\frac{1}{m}\sum_{i=1}^{m} \Phi(x_i) -  \frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)\right\|_\mathcal{H},
-\end{multline}
+\begin{equation}
+    \label{eq_modelagnostic:mmd}
+    \textup{MMD}_\Phi\left(\frac{1}{m}\sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \left\|\frac{1}{m}\sum_{i=1}^{m} \Phi(x_i) -  \frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)\right\|_\mathcal{H},
+\end{equation}
 where $\Phi$ is a kernel feature map built using the generative model and $x_1,\ldots,x_m$ is the training data, i.e.\@ samples from  $p_\textup{data}$.
-When $\mathcal{H}$ is a simple finite-dimensional Hilbert space and $\Phi$ can be computed easily, then \eqref{eq:mmd} can be computed by going through the data and computing the means in an online fashion.% While the definition seems to imply that we can only perform group-sample OOD detection, it is straightforward to notice that any test statistic of the form of \cref{eq:mmd} can be used also for single-test OOD detection when $n=1$.%by just substituting  $\frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)$ with simply $\Phi(\tilde{x})$.
+When $\mathcal{H}$ is a simple finite-dimensional Hilbert space and $\Phi$ can be computed easily, then \cref{eq_modelagnostic:mmd} can be computed by going through the data and computing the means in an online fashion.% While the definition seems to imply that we can only perform group-sample OOD detection, it is straightforward to notice that any test statistic of the form of \cref{eq_modelagnostic:mmd} can be used also for single-test OOD detection when $n=1$.%by just substituting  $\frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)$ with simply $\Phi(\tilde{x})$.
 
@@
 
-\paragraph{The Fisher kernel} An important example of kernel based on a generative model is the \emph{Fisher kernel} of \citet{jaakkola1999exploiting}.
+\paragraph{The Fisher kernel} An important example of kernel based on a generative model is the \emph{Fisher kernel} of \textcite{jaakkola_exploiting_1999}.
 The embedding of this kernel is the Fisher score
 \begin{equation}
-    \label{eq:Fisher_score}
+    \label{eq_modelagnostic:Fisher_score}
     \Phi_\textup{Fisher} (x) = I(\theta)^{-\frac{1}{2}} \nabla \log p_\theta (x),
@@
 and the corresponding reproducing kernel Hilbert space norm is just the $\ell_2$ norm: $|| \cdot || _\mathcal{H} = || \cdot ||_2$.
-In the case of the Fisher kernel, this means that \cref{eq:mmd} becomes:
+In the case of the Fisher kernel, this means that \cref{eq_modelagnostic:mmd} becomes:
 \begin{multline}
-    \label{eq:mmd_fisher}
+    \label{eq_modelagnostic:mmd_fisher}
     \textup{MMD}_{\Phi_\textup{Fisher}}\left(\frac{1}{m} \sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \\ \left\|\frac{I(\theta)^{-\frac{1}{2}}}{m} \sum_{i=1}^{m}  \nabla \log p_\theta (x_i) - \frac{I(\theta)^{-\frac{1}{2}}}{n}  \sum_{i=1}^{n}  \nabla \log p_\theta (\tilde{x}_i)\right\|_2.
@@
 We will see later that MMD with a Fisher kernel is closely related to the score statistic.
-In \cref{appendix:mahalanobis}, we additionally show that another popular OOD metric known as the \emph{Mahalanobis score} \citep{lee2018simple} can be interpreted as a MMD statistic with a certain Fisher kernel.
+In \cref{appendix_modelagnostic:mahalanobis}, we additionally show that another popular OOD metric known as the \emph{Mahalanobis score} \parencite{lee_simple_2018} can be interpreted as a MMD statistic with a certain Fisher kernel.
 
 \paragraph{The typicality kernel} A very simple approach of embedding the data using $p_\theta$ is to choose $\Phi_\textup{Typical}(x) = \log p_\theta (x)$.
-Then, MMD is exactly equivalent to the \emph{typicality test statistic} of \citet{nalisnick2019detecting}, although this connection was not explicitly stated  by \citet{nalisnick2019detecting}.
+Then, MMD is exactly equivalent to the \emph{typicality test statistic} of \textcite{nalisnick_detecting_2019}, although this connection was not explicitly stated  by \textcite{nalisnick_detecting_2019}.
 Because of this, we call the kernel $k(x,y) = \log p_\theta (x) \cdot \log p_\theta (y)$ the \emph{typicality kernel}.
@@
 %While $\Phi_\textup{Typical}$ is not as well motivated as the Fisher kernel, we found that it generally gives good results. 
-While $\Phi_\textup{Typical}$ is not as well motivated as a kernel as $\Phi_\textup{Fisher}$, the concepts of typicality and typical set can be used to explain unintuitive behaviours of probability distributions in high-dimensional space as highlighted by \cite{nalisnick2018deep}.
+While $\Phi_\textup{Typical}$ is not as well motivated as a kernel as $\Phi_\textup{Fisher}$, the concepts of typicality and typical set can be used to explain unintuitive behaviours of probability distributions in high-dimensional space as highlighted by \textcite{nalisnick_deep_2019}.
 We also found that using this kernel generally gives good results for OOD tasks.
@@
 In general, neither of these two kernels are characteristic, meaning that our MMD can be zero even if the distributions are not identical.
-This could be solved by combining them with a characteristic kernel, as in \cite{liu2020learning}, at the price of including a new hyperparameter.
+This could be solved by combining them with a characteristic kernel, as in \textcite{liu_learning_2020}, at the price of including a new hyperparameter.
 
@@
 \section{Combining different test statistics}
-\label{sec:combination_p_values}
-For single-sample OOD detection, \citet{zhang2021understanding} proved that there is not a single statistic that is constantly better compared to all the possible alternatives of interest.
+\label{sec_modelagnostic:combination_p_values}
+For single-sample OOD detection, \textcite{zhang_understanding_2021} proved that there is not a single statistic that is constantly better compared to all the possible alternatives of interest.
 For this reason, we believe that using a combination of different test statistics should lead to an overall better OOD detection in settings where a single statistic might fail.
-Assume we compute $k$ different test statistics $T_1,\dots, T_k$, each testing $\mathcal{H}_0$ against $\mathcal{H}$ as defined in \cref{sec:parametric,sec:MMD}.
+Assume we compute $k$ different test statistics $T_1,\dots, T_k$, each testing $\mathcal{H}_0$ against $\mathcal{H}$ as defined in \cref{sec_modelagnostic:parametric,sec_modelagnostic:MMD}.
 The goal is to combine these different tests into a single statistical test that ideally will perform better than the initial single tests.
@@
 
-\citet{morningstar2021density} proposed the density of states estimator (DoSE) to overcome this problem.
+\textcite{morningstar_density_2021} proposed the density of states estimator (DoSE) to overcome this problem.
 They only focused on the single-sample detection task, i.e.\@ $n=1$ following our problem formulation.
@@
 In our work, instead, we propose a different approach and leverage the fact that we use only one-sided test statistics.
-This setting is a well-studied problem in the literature both for independent \citep{fisher1925statistical, folks1971asymptotic} and dependent one-sided test statistics \citep{brown1975400, wilson2019harmonic}.
+This setting is a well-studied problem in the literature both for independent \parencite{fisher_statistical_1925, folks_asymptotic_1971} and dependent one-sided test statistics \parencite{brown_400_1975, wilson_harmonic_2019}.
 All these approaches rely on the computation of $p$-values of each statistic for the test set $\tilde{x}_1,\ldots,\tilde{x}_n$.
@@
 To succeed in this, we should be able to compute $p_j = \textup{Pr}(T_j > t_j \mid \mathcal{H}_0)$, therefore we need to estimate the distribution of each statistic $T_j$ under $\mathcal{H}_0$.
-%As done by \citet{nalisnick2019detecting}, we assume the existence of a validation set $\mathbf{X}'$ that was not used to train our generative model.
+%As done by \textcite{nalisnick_detecting_2019}, we assume the existence of a validation set $\mathbf{X}'$ that was not used to train our generative model.
 Instead of bootstrapping $K$ new datasets $\{X^{'}_k\}_{k=1}^{K}$ of size $M$, as they did in their work, we directly evaluate each test statistic $T_j$ on every single validation example.
 Asymptotically, this is equivalent to creating $K$ new datasets of size $M=1$ when $K \rightarrow \infty$.
-As done by \citet{nalisnick2019detecting}, we assume the existence of a validation set $\mathbf{X}'$ that was not used to train our generative model.
+As done by \textcite{nalisnick_detecting_2019}, we assume the existence of a validation set $\mathbf{X}'$ that was not used to train our generative model.
 From $\mathbf{X}'$ we bootstrap $S$ new datasets $\{\mathbf{X}^{'}_s\}_{s=1}^{S}$ of size $M'$ by using bootstrap resampling.
@@
 \paragraph{Combining test statistics by combining $p$-values}
-Fisher's \citeyearpar{fisher1925statistical} method is a procedure to combine different $p$-values $p_1,\dots, p_k$.
-This method assumes that all the considered test statistics are independent, and \citet{folks1971asymptotic} proved that it is asymptotically optimal among all methods of combining independent tests.
+Fisher's \citeyearpar{fisher_statistical_1925} method is a procedure to combine different $p$-values $p_1,\dots, p_k$.
+This method assumes that all the considered test statistics are independent, and \textcite{folks_asymptotic_1971} proved that it is asymptotically optimal among all methods of combining independent tests.
 %
@@
 \begin{equation}
-\label{eq:Fisher_method}
+    \label{eq_modelagnostic:Fisher_method}
     X^2 \sim -2 \sum_{j=1}^{k} \ln (p_j).
@@
 %
-In the \cref{appendix:harmonic} , we also consider the Harmonic mean $p$-value \citep{wilson2019harmonic} as a way to combine $p$-values from different statistics.
+In the \cref{appendix_modelagnostic:harmonic} , we also consider the Harmonic mean $p$-value \parencite{wilson_harmonic_2019} as a way to combine $p$-values from different statistics.
 This method usually works best when the statistics are not independent.
@@
 \section{From test statistics to practical out-of-distribution scores}
-\label{sec:section4}
+\label{sec_modelagnostic:section4}
 
 Several of the test statistics that we consider make use of the inverse of the Fisher information matrix $I(\theta)$.
-The true Fisher information matrix requires an identifiable model to be invertible \citep{watanabe2009algebraic} and computing its inverse is $\mathcal{O}(m^3)$, where $m$ is the number of model parameters.
+The true Fisher information matrix requires an identifiable model to be invertible \parencite{watanabe_algebraic_2009} and computing its inverse is $\mathcal{O}(m^3)$, where $m$ is the number of model parameters.
 For DGMs, the Fisher information matrix might not be invertible due to the fact that DGMs typically do not satisfy the identifiability condition.
@@
 A more refined way is to look for a diagonal approximation.
-In \cref{appendix:cheap}, we describe cheap ways of computing such approximations.
+In \cref{appendix_modelagnostic:cheap}, we describe cheap ways of computing such approximations.
 In particular, we will study two cases: the case where $I(\theta)$ is replaced by the identity matrix and the case where $I(\theta)$ is replaced by a diagonal matrix estimated using the training data.
@@
 For this reason, we do not consider it in this work.
-More complex and precise approximations of the FIM exists, such as the Kronecker-factored Approximate Curvature (K-FAC, \citealp{martens2015optimizing}), but these are not defined for all types of layers used by state-of-the-art models.
+More complex and precise approximations of the FIM exists, such as the Kronecker-factored Approximate Curvature (K-FAC, \citealp{martens_optimizing_2015}), but these are not defined for all types of layers used by state-of-the-art models.
 
-\paragraph{On the difficulty of computing per-example gradients} Both the diagonal approximation of the FIM and the computation of the MMD with Fisher kernel of \cref{eq:mmd_fisher} require the gradient computation for all training and test examples.
+\paragraph{On the difficulty of computing per-example gradients} Both the diagonal approximation of the FIM and the computation of the MMD with Fisher kernel of \cref{eq_modelagnostic:mmd_fisher} require the gradient computation for all training and test examples.
 This is known as a costly procedure.
-For example, if we have to compute the gradient for $N$ examples using a simple fully connected network with $l$ layers of size $p$, the naive procedure of using a batch-size of dimension 1 is $\mathcal{O}(Nlp^2)$ \citep{goodfellow2015efficient}.
-While more efficient per-example gradient computations were proposed \citep{goodfellow2015efficient, rochette2019efficient}, these techniques can only be applied on simple fully connected or convolutional networks.
-While for this paper we relied on the naive solution of looping through every example one at the time, a more efficient solution is provided by the BackPACK library \citep{dangel2020backpack} which allows to compute the gradient with respect each sample in a minibatch.
+For example, if we have to compute the gradient for $N$ examples using a simple fully connected network with $l$ layers of size $p$, the naive procedure of using a batch-size of dimension 1 is $\mathcal{O}(Nlp^2)$ \parencite{goodfellow_efficient_2015}.
+While more efficient per-example gradient computations were proposed \parencite{goodfellow_efficient_2015, rochette_efficient_2019}, these techniques can only be applied on simple fully connected or convolutional networks.
+While for this paper we relied on the naive solution of looping through every example one at the time, a more efficient solution is provided by the BackPACK library \parencite{dangel_backpack_2020} which allows to compute the gradient with respect each sample in a minibatch.
 
@@
 At the maximum likelihood estimate, we have that $\mathbb{E}[ \nabla \log p_\theta (x)] = 0$, i.e.\@ the first term inside the norm is $0$.
-Therefore, we expect that the differences between the OOD scores computed by using \cref{eq:mmd_fisher} will be preserved if we only consider $\left\|I(\theta)^{-1/2} \nabla \log p_\theta (\tilde{x}_1,\dots,\tilde{x}_n)\right\|_2$,
+Therefore, we expect that the differences between the OOD scores computed by using \cref{eq_modelagnostic:mmd_fisher} will be preserved if we only consider $\left\|I(\theta)^{-1/2} \nabla \log p_\theta (\tilde{x}_1,\dots,\tilde{x}_n)\right\|_2$,
 %\begin{equation}
@@
 
-Computationally speaking, considering the score statistic instead of the MMD Fisher lets us avoid going through the entire training set to compute the average gradient (first term in \cref{eq:mmd_fisher}) while carrying the same information.
+Computationally speaking, considering the score statistic instead of the MMD Fisher lets us avoid going through the entire training set to compute the average gradient (first term in \cref{eq_modelagnostic:mmd_fisher}) while carrying the same information.
 Therefore, in this paper, we will mainly focus on the combination of the typicality test and the score statistic.
@@
 In this case, the score statistic will be $0$, and will therefore not detect that the point is actually OOD.
-However, when $D$ is large, the typicality test will be able to declare that this point is anomalous, as shown by \citet{nalisnick2019detecting}.
+However, when $D$ is large, the typicality test will be able to declare that this point is anomalous, as shown by \textcite{nalisnick_detecting_2019}.
 
 Therefore, we have that the typicality test and the score statistic are complementary and measure a different type of information.
-In \cref{appendix:correlation}, we empirically show that they are not correlated, by plotting the two measures against each other and by computing the correlation matrix.
+In \cref{appendix_modelagnostic:correlation}, we empirically show that they are not correlated, by plotting the two measures against each other and by computing the correlation matrix.
 
@@
 \section{Related works}
-\label{sec:related_works}
+\label{sec_modelagnostic:related_works}
 
-Since \citet{nalisnick2018deep} and \cite{hendrycks2018deep}, different test statistics or methodologies for OOD detection using DGMs were proposed.
+Since \textcite{nalisnick_deep_2019} and \textcite{hendrycks_deep_2019}, different test statistics or methodologies for OOD detection using DGMs were proposed.
 Most of the recent solutions were highly influenced by three major lines of work: \emph{typicality set}, \emph{likelihood ratio} test statistics, and \emph{model misestimation}.
 
-The typicality set hypothesis was introduced by \citet{nalisnick2019detecting} as a possible explanation for the DGMs assigning higher likelihood to OOD data.
+The typicality set hypothesis was introduced by \textcite{nalisnick_detecting_2019} as a possible explanation for the DGMs assigning higher likelihood to OOD data.
 The typicality set is the subset of the model full support where the model samples from and this does not intersect with the region of higher likelihood.
-While the typicality test was introduced for batch-OOD detection, \cite{morningstar2021density} shows that it also works well in the single-sample case.
+While the typicality test was introduced for batch-OOD detection, \textcite{morningstar_density_2021} shows that it also works well in the single-sample case.
 This is also confirmed by our own experiments.
 
-The likelihood ratio test statistic method by \citet{ren2019likelihood} assumes that every input is composed by a background component and a semantic component.
+The likelihood ratio test statistic method by \textcite{ren_likelihood_2019} assumes that every input is composed by a background component and a semantic component.
 For OOD detection, only the semantic component matters.
 In addition to a model trained on the in-distribution data, they proposed to train a background model on perturbed inputs data and then for each test example consider as OOD score the likelihood ratio between the two models.
-\cite{schirrmeister2020understanding}, instead, trained the background model on a more general distribution of images by considering 80 million general tiny images.
-Similarly to these approaches, \cite{serra2019input} argued that the failure of DGMs is due to the high-influence that the input complexity has on the likelihood.
+\textcite{schirrmeister_understanding_2020}, instead, trained the background model on a more general distribution of images by considering 80 million general tiny images.
+Similarly to these approaches, \textcite{serra_input_2020} argued that the failure of DGMs is due to the high-influence that the input complexity has on the likelihood.
 Therefore, they proposed to use a general lossless image compression algorithm as a background model.
@@
 Another line of works blame the models themselves and not the test statistics.
-\cite{zhang2021understanding} argued that model misestimation is the main cause of higher likelihood assigned to OOD data.
+\textcite{zhang_understanding_2021} argued that model misestimation is the main cause of higher likelihood assigned to OOD data.
 This can be due to both the model architecture and the maximum likelihood objective.
-\cite{kirichenko2020normalizing} and \cite{schirrmeister2020understanding} showed that normalizing flows can achieve better OOD performance despite achieving a worse likelihood if one changes some model design choices.
+\textcite{kirichenko_why_2020} and \textcite{schirrmeister_understanding_2020} showed that normalizing flows can achieve better OOD performance despite achieving a worse likelihood if one changes some model design choices.
 %
-Other works in the literature focused on deriving specific test statistics that works only for a specific model, for example for VAEs \citep{xiao2020likelihood, maaloe2019biva, havtorn2021hierarchical}, or for normalizing flows \citep{kirichenko2020normalizing, ijcai2021-292}.
+Other works in the literature focused on deriving specific test statistics that works only for a specific model, for example for VAEs \parencite{xiao_likelihood_2020, maaloe_biva_2019, havtorn_hierarchical_2021}, or for normalizing flows \parencite{kirichenko_why_2020, ahmadian_likelihoodfree_2021}.
 
@@
 As mentioned in the introduction, we frame the OOD detection problem in terms of statistical tests problem.
-Recently, \cite{haroush2021statistical} showed that adopting hypothesis testing at the layer and channel level of a neural network can be used for OOD detection in the discriminative setting.
+Recently, \textcite{haroush_statistical_2021} showed that adopting hypothesis testing at the layer and channel level of a neural network can be used for OOD detection in the discriminative setting.
 They used both Fisher's method and Simes' method to combine class-conditional $p$-values computed for each convolutional and dense layer of a deep neural network.
 We focus on the unsupervised setting using DGMs and use hypothesis testing on statistics that can be computed on all differentiable DGM.
-As already explained in section \cref{sec:combination_p_values}, \cite{morningstar2021density} considered the combination of different statistics for OOD detection.
+As already explained in section \cref{sec_modelagnostic:combination_p_values}, \textcite{morningstar_density_2021} considered the combination of different statistics for OOD detection.
 The main difference with their approach is that we propose statistics that can be applied to any differentiable generative model and combine them by using Fisher's method, which takes advantage of using only one-sided independent statistics.
-Concurrently, \cite{choi2021robust} derived the score statistic by starting from the likelihood ratio statistic and applying a Laplace approximation.
+Concurrently, \textcite{choi_robust_2021} derived the score statistic by starting from the likelihood ratio statistic and applying a Laplace approximation.
 They computed the score statistic only for certain layers of the model and for a specific example, the OOD score is given by the infinity norm of these different layer scores after a ReLU operation.
@@
 \begin{figure*}[tbp]
-    \includegraphics[width=\textwidth]{imgs/pvalues_comb_no_drop_correct_font.pdf}
+    \includegraphics[width=\textwidth]{graphics/paper_modelagnostic/pvalues_comb_no_drop_correct_font.pdf}
     \caption{\emph{First plot}: $p$-values of the typicality test on the two test sets.
@@
 These plots refer to a PixelCNN++ trained on FashionMNIST without dropout.}
-    \label{fig:comb_p_values}
+    \label{fig_modelagnostic:comb_p_values}
 \end{figure*}
@@
         \end{tabular}
-        \label{tab:single_sample_results}
+        \label{tab_modelagnostic:single_sample_results}
     }
@@
 
-To evaluate the performance of the combination of the typicality test and the score statistic in detecting OOD data, we follow the experiments of \citet{nalisnick2018deep, hendrycks2018deep} and considered the OOD detection task on three image dataset pairs that have been proven challenging for DGMs, i.e.\@ FashionMNIST \citep{xiao2017fashion} vs MNIST \citep{lecun1998mnist}, CIFAR10 \citep{krizhevsky2009learning} vs SVHN \citep{netzer2011reading}, and CIFAR10 vs CIFAR100.
-\cite{winkens2020contrastive} divide these tasks into \textit{far}-OOD tasks, where the in-distribution and out-distribution are different such as in the case of CIFAR10 against SVHN, and \textit{near}-OOD where the two distributions are pretty similar, such as CIFAR10 and CIFAR100.
+To evaluate the performance of the combination of the typicality test and the score statistic in detecting OOD data, we follow the experiments of \textcite{nalisnick_deep_2019, hendrycks_deep_2019} and considered the OOD detection task on three image dataset pairs that have been proven challenging for DGMs, i.e.\@ FashionMNIST \parencite{xiao_fashionmnist_2017} vs MNIST \parencite{lecun_mnist_1998}, CIFAR10 \parencite{krizhevsky_learning_2009} vs SVHN \parencite{netzer_reading_2011}, and CIFAR10 vs CIFAR100.
+\textcite{winkens_contrastive_2020} divide these tasks into \textit{far}-OOD tasks, where the in-distribution and out-distribution are different such as in the case of CIFAR10 against SVHN, and \textit{near}-OOD where the two distributions are pretty similar, such as CIFAR10 and CIFAR100.
 \textit{Near}-OOD tasks are usually most challenging.
 
-For each task, we trained three different state-of-the-art DGMs, a PixelCNN++ \citep{salimans2017pixelcnn++}, a Glow model \citep{kingma2018glow}, and a hierarchical variational autoencoder \citep{kingma2013auto,rezende2014stochastic} with bottom-up inference (HVAE, \citealp{burda2015importance}).
+For each task, we trained three different state-of-the-art DGMs, a PixelCNN++ \parencite{salimans_pixelcnn_2017}, a Glow model \parencite{kingma_glow_2018}, and a hierarchical variational autoencoder \parencite{kingma_autoencoding_2014,rezende_stochastic_2014} with bottom-up inference (HVAE, \citealp{burda_importance_2016}).
 These are DGMs parametrized by neural networks that make different assumptions in the modelling choice of the target distribution.
 In addition to that, for PixelCNN++ and Glow we have a tractable likelihood while for HVAE we can only estimate a lower bound.
-A more in-depth description of these methods and additional results testing MNIST against FashionMNIST and SVHN against CIFAR10 can be found in \cref{appendix:mnist_svhn}.
-We also extensively analyzed, focusing mostly in the influence of the preprocessing, the results on CIFAR10 vs CelebA \citep{liu2015faceattributes} in \cref{appendix:celeba}.
-In \cref{appendix:gmm_ppca}, we also considered a Gaussian Mixture Model and a Probabilistic PCA as simple generative models.
+A more in-depth description of these methods and additional results testing MNIST against FashionMNIST and SVHN against CIFAR10 can be found in \cref{appendix_modelagnostic:mnist_svhn}.
+We also extensively analyzed, focusing mostly in the influence of the preprocessing, the results on CIFAR10 vs CelebA \parencite{liu_deep_2015} in \cref{appendix_modelagnostic:celeba}.
+In \cref{appendix_modelagnostic:gmm_ppca}, we also considered a Gaussian Mixture Model and a Probabilistic PCA as simple generative models.
 
@@
 \paragraph{Baselines} We are mostly interested in testing our methods with other model-agnostic test statistics in the literature.
-Apart from using the plain likelihood as an OOD score, the only test statistic we are aware of that can be applied to any generative model without requiring any background model or OOD assumptions is the typicality test statistic of \cite{nalisnick2019detecting}.
-We also considered the gradient norm, which in general seem to work well but fails in the case of SVHN vs CIFAR10 (see \cref{appendix:mnist_svhn}).
-In addition to that, we compare our methods to a model-agnostic version of DoSE by \cite{morningstar2021density}, where we used KDEs to combine the score statistic and the typicality test statistic.
+Apart from using the plain likelihood as an OOD score, the only test statistic we are aware of that can be applied to any generative model without requiring any background model or OOD assumptions is the typicality test statistic of \textcite{nalisnick_detecting_2019}.
+We also considered the gradient norm, which in general seem to work well but fails in the case of SVHN vs CIFAR10 (see \cref{appendix_modelagnostic:mnist_svhn}).
+In addition to that, we compare our methods to a model-agnostic version of DoSE by \textcite{morningstar_density_2021}, where we used KDEs to combine the score statistic and the typicality test statistic.
 
-\paragraph{Evaluation} We compare our methods with the baselines by computing the area under the receiver operating characteristic curve (AUROC) as done in previous works \citep{hendrycks2018deep, ren2019likelihood, morningstar2021density}.
-We also evaluate our methods in terms of False Discovery Rate (FDR) control \cite{benjamini1995controlling}, i.e.\@ the proportion of false positive among the rejected hypothesis.
+\paragraph{Evaluation} We compare our methods with the baselines by computing the area under the receiver operating characteristic curve (AUROC) as done in previous works \parencite{hendrycks_deep_2019, ren_likelihood_2019, morningstar_density_2021}.
+We also evaluate our methods in terms of False Discovery Rate (FDR) control \textcite{benjamini_controlling_1995}, i.e.\@ the proportion of false positive among the rejected hypothesis.
 Note that both quantities need to know the true label (OOD or in-distribution) to be computed.
@@
 We first evaluate our proposed method in the single-sample OOD detection task.
-Results are summarized in \cref{tab:single_sample_results}.
+Results are summarized in \cref{tab_modelagnostic:single_sample_results}.
 We start by considering the OOD task on FashionMNIST against MNIST.
@@
 The typicality test is better than the score statistic overall.
-The gradient norm surprisingly performs well in the two dataset pairs, but it fails badly when the model is trained on SVHN (see \cref{appendix:mnist_svhn}).
+The gradient norm surprisingly performs well in the two dataset pairs, but it fails badly when the model is trained on SVHN (see \cref{appendix_modelagnostic:mnist_svhn}).
 Regarding the combination of the two statistics, the Fisher's method is always better than DoSE, but in this setting, it improves over the best of the single statistics three out of five times.
@@
 In the \textit{near}-OOD task, we have that both our method and DoSE using our suggested statistics perform closely.
-We want to highlight that for this challenging task we get results that are comparable with those reported in \cite{morningstar2021density}, but by using two model-agnostic statistics instead of three model-specific ones.
+We want to highlight that for this challenging task we get results that are comparable with those reported in \textcite{morningstar_density_2021}, but by using two model-agnostic statistics instead of three model-specific ones.
 %
 It can be noticed that the way we train our models has a strong influence on both the typicality test and the score statistic, although the models get the same test log-likelihood.
-In \cref{appendix:checkpoints}, we also show that this can happen between different checkpoints of the same model.
+In \cref{appendix_modelagnostic:checkpoints}, we also show that this can happen between different checkpoints of the same model.
 
-In \cref{fig:comb_p_values}, we show that the $p$-values distributions for both the typicality and the score statistic are uniformly distributed under the null-hypothesis and that the combination under the null follows a $\chi^2$ distribution with 4 degrees of freedom.
+In \cref{fig_modelagnostic:comb_p_values}, we show that the $p$-values distributions for both the typicality and the score statistic are uniformly distributed under the null-hypothesis and that the combination under the null follows a $\chi^2$ distribution with 4 degrees of freedom.
 This also supports the fact that the typicality test and the score statistic are independent.
@@
 \paragraph{Two-sample OOD}
-As \citet{nalisnick2019detecting}, we consider how these test statistics change when performing two-sample OOD detection.
-Results are summarized in \cref{tab:two_sample_results}.
-As shown by \citet{nalisnick2019detecting}, the typicality improves but also the score statistic gets better if we consider more samples.
+As \textcite{nalisnick_detecting_2019}, we consider how these test statistics change when performing two-sample OOD detection.
+Results are summarized in \cref{tab_modelagnostic:two_sample_results}.
+As shown by \textcite{nalisnick_detecting_2019}, the typicality improves but also the score statistic gets better if we consider more samples.
 Combining those leads to an improvement of performance in terms of AUROC with almost all the models.
@@
 \scriptsize
+\centering
 \caption{AUROC$\uparrow$ for two-sample OOD detection using the usual considered model.}\normalsize
     %\renewcommand{\thetable}{4}
-    \resizebox{0.5\textwidth}{!}{
+    \resizebox{0.80\textwidth}{!}{
         \scriptsize
@@
         \end{tabular}
-        \label{tab:two_sample_results}
+        \label{tab_modelagnostic:two_sample_results}
     }
@@
 
-%%%%%%%%%%%%%%%%%%%%%%% FIGURE
-\begin{figure}[tb]
-    \centering
-    \includegraphics[scale=0.47]{imgs/fashion_hvae_type1_correct_font.pdf}
-    \caption{Type I (probability of an inlier to be classified as outlier) and Type II (probability of an outlier to be considered as inlier) errors versus the significance level $\alpha$ on the combination values.
-By using Benjamini-Hochberg correction, we get that the Type I error stays below identity line.}
-    \label{fig:type1}
-\end{figure}
-
 \subsection{Practical OOD detection with FDR control}
-One of the advantages of framing the problem as multiple testing is that we have a well-defined procedure to decide on which hypotheses to reject while controlling the False Discovery Rate (FDR, \citealp{benjamini1995controlling}).
-Imagine we are interested in finding the outliers from the dataset given by the combination of the two test-sets but we do not want to discard too many inliers, then we can use the Benjamini-Hochberg (BH) procedure \citep{benjamini1995controlling} to decide a threshold and reject all hypothesis below that threshold.
+One of the advantages of framing the problem as multiple testing is that we have a well-defined procedure to decide on which hypotheses to reject while controlling the False Discovery Rate (FDR, \citealp{benjamini_controlling_1995}).
+Imagine we are interested in finding the outliers from the dataset given by the combination of the two test-sets but we do not want to discard too many inliers, then we can use the Benjamini-Hochberg (BH) procedure \parencite{benjamini_controlling_1995} to decide a threshold and reject all hypothesis below that threshold.
 For a specific significance level $\alpha$, the procedure guarantees that the FDR stays below that level.
@@
 We leverage the fact that when the null hypothesis is true and the $p$-values are independent, then the scores obtained by combining $k$ different statistics are $\chi^2_{2k}$ distributed to compute the $p$-values.
-Alternatively, the procedure can be also applied to the $p$-values of a single test-statistic.
+Alternatively, th e procedure can be also applied to the $p$-values of a single test-statistic.
 %
@@
 
-As can be seen in \cref{fig:type1}, we have that the Type I ratio line stays below the identity line, meaning that the BH correction is working.
+As can be seen in \cref{fig_modelagnostic:type1}, we have that the Type I ratio line stays below the identity line, meaning that the BH correction is working.
 When deciding for a specific threshold $\alpha$, we usually have to trade-off between Type I and Type II error and in most cases the threshold to choose depends on the application domain.
 Ideally, we would like to have a low Type I and a low Type II error rate, meaning that we are not considering a lot of in-distribution examples as OOD and at the same time considering a lot of outliers as in-distribution.
-\cref{fig:type1} shows that we can achieve this for low values of $\alpha$.
-When training on CIFAR, instead, we are able to control the FDR only from a certain significance level (see \cref{appendix:BH_cifar}).
+\cref{fig_modelagnostic:type1} shows that we can achieve this for low values of $\alpha$.
+When training on CIFAR, instead, we are able to control the FDR only from a certain significance level (see \cref{appendix_modelagnostic:BH_cifar}).
 This is expected given that the AUROC is not as good as when testing on MNIST.
+
+%%%%%%%%%%%%%%%%%%%%%%% FIGURE
+\begin{figure}[tb]
+    \centering
+    \includegraphics[scale=0.48]{graphics/paper_modelagnostic/fashion_hvae_type1_correct_font.pdf}
+    \caption{Type I (probability of an inlier to be classified as outlier) and Type II (probability of an outlier to be considered as inlier) errors versus the significance level $\alpha$ on the combination values.
+By using Benjamini-Hochberg correction, we get that the Type I error stays below identity line.}
+    \label{fig_modelagnostic:type1}
+\end{figure}
 
@@
 
-DGMs have recently been used for handling missing data (see e.g.\@ \citealp{mattei2019miwae, ma2018eddi, nazabal2020handling, ipsen2020not}).
+DGMs have recently been used for handling missing data (see e.g.\@ \citealp{mattei_miwae_2019, ma_eddi_2019, nazabal_handling_2020, ipsen_not-miwae_2021}).
 An interesting future direction would be to extend these OOD detection methods to handle missing values.

### 2023-07-20 appendices/appendix_modelagnostic.tex
Format Bergamin appendix

@@
 A crude diagonal approximation can be computed by simply estimating the diagonal of $I({\theta})$ and setting all off-diagonal elements to zero.
-Such diagonal approximations have been used in machine learning for decades: for instance, \citet[Section 3.12.2]{lecun1987} used a similar approximation of the Hessian matrix, and called it ``outrageously simplifying".
+Such diagonal approximations have been used in machine learning for decades: for instance, \textcite[Section 3.12.2]{lecun1987} used a similar approximation of the Hessian matrix, and called it ``outrageously simplifying".
 Much more complex approximations have been derived, although diagonal approximations have been consistently used (e.g.
-by \citealp{kirkpatrick2017overcoming}, who used essentially the same approximation in a supervised context), and are linked to several adaptive optimisation techniques like Adam \citep{kingma2014adam} or RMSProp \citep{tieleman2012lecture}.
-A good discussion on these issues is provided in Martens's \citeyearpar{martens2020new} recent review.
+by \citealp{kirkpatrick_overcoming_2017}, who used essentially the same approximation in a supervised context), and are linked to several adaptive optimisation techniques like Adam \citep{kingma_adam_2015} or RMSProp \citep{tieleman_lecture_2012}.
+A good discussion on these issues is provided in Martens's \citeyearpar{martens_new_2020} recent review.
 
@@
     \item By using the training examples $x_1,...,x_T$, we form the estimate
-    $$ {D}_T({\theta}) = \frac{1}{T} \sum_{t=1}^T \textup{Diag}(\nabla \log {p_\theta}(x_t)^2),  $$
+    $$ {D}_T({\theta}) = \frac{1}{T} \sum_{t=1}^T \textup{diag}(\nabla \log {p_\theta}(x_t)^2),  $$
     where the square in $\nabla \log p_\theta(x_t)^2$ is computed elementwise.
@@
 A slightly more refined approach is to slightly regularise $D_T({\theta})$.
-Following \citet{martens2020new}, our final estimate of the Fisher information matrix is
+Following \textcite{martens_new_2020}, our final estimate of the Fisher information matrix is
     \begin{equation}
@@
 \paragraph{How to choose $\varepsilon$ and $\xi$?} The Adam optimizer uses a similar estimate, with default hyperparameters $\varepsilon = 10^{-8}$ and $\xi = 1$.
-As argued by \citet{martens2020new}, it can be interesting to use $\xi <1$  in order to diminish the influence of extreme values of $D_T({\theta})$.
-In particular, \citet{martens2020new} suggests taking $\xi = 0.75$.
+As argued by \textcite{martens_new_2020}, it can be interesting to use $\xi <1$  in order to diminish the influence of extreme values of $D_T({\theta})$.
+In particular, \textcite{martens_new_2020} suggests taking $\xi = 0.75$.
 When $\xi \longrightarrow 0$, then $\hat{I}_T({\theta})$ will approach the identity matrix.
 We tested the two settings by using a PixelCNN++ trained on CIFAR.
-Results are shown in table~\ref{tab:martens}.
+Results are shown in \cref{tab_modelagnostic:martens}.
 In terms of OOD detection, it seems that using $\varepsilon = 10^{-8}$ and $\xi = 1$ is slightly better.
@@
 The second advantage is that, if we wish to compute a MMD statistic, such as the MMD with the Fisher kernel or the MMD typicality (that require the average of gradient or the average log-likelihood over the training), computing the average of the square of the gradient costs very little.
-One can just do a single loop over the data, and use the usual formulas for online estimation of a mean, see \cref{algo_description}.
+One can just do a single loop over the data, and use the usual formulas for online estimation of a mean, see \cref{alg_modelagnostic:algo_description}.
 
@@
 Another possibility is to just use the identity matrix as FIM instead of of approximating the diagonal through the procedure explained above.
-In our experiments (see table~\ref{tab:mmd} and table~\ref{tab:mmd_pair2}), we can see that sometimes using the identity matrix seems to work equally well or a bit better for some models trained on FashionMNIST and CIFAR10.
+In our experiments (see \cref{tab_modelagnostic:mmd} and \cref{tab_modelagnostic:mmd_pair2}), we can see that sometimes using the identity matrix seems to work equally well or a bit better for some models trained on FashionMNIST and CIFAR10.
 However, when we train on SVHN or MNIST, there are a cases where the statistic that is using the identity matrix as approximation fails, sometimes being worse than random chance.
@@
     \centering
-    %\renewcommand{\thetable}{4}
     \caption{AUROC$\uparrow$ for single-sample OOD detection.
@@
 For ($\ddagger$) we used the Adam parameter choice, i.e.\@ $\varepsilon = 10^{-8}$ and $\xi = 1$.
-For ($\mathsection$), instead, we used  $\varepsilon = 10^{-8}$ and $\xi = 0.75$, as suggested by \citet{martens2020new}.
+For ($\mathsection$), instead, we used  $\varepsilon = 10^{-8}$ and $\xi = 0.75$, as suggested by \textcite{martens_new_2020}.
 As a results we have that using Adam parameters choice is slightly better for our task.}
+    \resizebox*{\textwidth}{!}{
     \small
@@
         \bottomrule
+        \vspace{-0.2cm}\\
         ($\ddagger$) With $\varepsilon = 10^{-8}$ and $\xi = 1$ \\
@@
     \end{tabular}
-    \label{tab:martens}
+    \label{tab_modelagnostic:martens}
+    }
 \end{table*}
-
-% \section{Typicality test and score statistic}
-
-% Let's us look at the case when the typicality test fail and the score statistic works example introduced in Section 4.2 of the paper.
-
-% \paragraph{Typicality test and score statistic are uncorrelated} To test if the typicality test and the score statistic are uncorrelated we plot the two scores computed on the validation set.
-As can be seen from figure \ref{fig:correlatio}, we have that the two measures are not correlated as it is also highlight by the correlation matrix.
-
-% \begin{figure}[htbp]
-% \centering
-% \begin{subfigure}{.5\textwidth}
-    %   \centering
-    %   \includegraphics[scale=0.5]{imgs/correlation.png}
-    %   %\caption{Kalman Filter $p(x_t\mid y_{1:t})$ and true state observations.}
-    % \end{subfigure}%
-    % \begin{subfigure}{.5\textwidth}
-        %       \centering
-        %       \begin{tabular}{|c|c|}
-            %       \hline
-            %           1. &  -0.014\\ 
-            %           \hline
-            %           -0.014 & 1. \\ 
-            %         \hline
-            %       \end{tabular}
-            % \end{subfigure}
-            % \end{figure}
-
-            % TO DO:
-            % \begin{itemize}
-                %     \item WRITE "PROOF" OF WHEN THE TYPICALITY FAILS (SKETCHED PROOF ON PIERRE'S MOLESKINE)
-                % \end{itemize}
 
@@
 
-                \citet{lee_simple_2018} introduced a simple metric to perform OOD detection with a trained deep classifier.
+\textcite{lee_simple_2018} introduced a simple metric to perform OOD detection with a trained deep classifier.
 The key idea is to train a simple generative model (linear discriminant analysis) in the feature space of the classifier.
@@
 Our idea is to re-use ideas introduced by this computer vision litterature.
-Under the assumption that the Gaussian clusters are well-separated, \citet{tanaka2013fisher}, extending an earlier analysis of \citet[Appendix A]{sanchez2013image}, showed that
+Under the assumption that the Gaussian clusters are well-separated, \textcite{tanaka2013fisher}, extending an earlier analysis of \textcite[Appendix A]{sanchez2013image}, showed that
 \begin{equation}
@@
 For SVHN, instead, we used 6032 datapoints from the test set as validation, leaving the remaining 20000 examples as test set.
-In Table~\ref{tab:log_like}, we reported test log-likelihood of the models used in this paper.
+In \cref{tab_modelagnostic:log_like}, we reported test log-likelihood of the models used in this paper.
 
@@
 
-\paragraph{Glow} For training Glow models we follow \citet{kirichenko_why_2020} and their repository\footnote{\href{https://github.com/PolinaKirichenko/flows_ood}{\texttt{https://github.com/PolinaKirichenko/flows\_ood}}}.
-They closely follow \citet{nalisnick_deep_2019} and \cite{kingma_glow_2018} implementation for multi-scale Glow, where a scale is defined as the sequence of actorm, invertible $1\times1$ convolution and coupling layers.
-While \citet{kirichenko_why_2020} only considers the RMSProp optimizer, we trained two different models, one using RMSProp and one using Adam with batch-size 32.
+\paragraph{Glow} For training Glow models we follow \textcite{kirichenko_why_2020} and their repository\footnote{\href{https://github.com/PolinaKirichenko/flows_ood}{\texttt{https://github.com/PolinaKirichenko/flows\_ood}}}.
+They closely follow \textcite{nalisnick_deep_2019} and \cite{kingma_glow_2018} implementation for multi-scale Glow, where a scale is defined as the sequence of actorm, invertible $1\times1$ convolution and coupling layers.
+While \textcite{kirichenko_why_2020} only considers the RMSProp optimizer, we trained two different models, one using RMSProp and one using Adam with batch-size 32.
 For the greyscale dataset our Glow is made up of 2 scales with 16 coupling layers, and a 3-layers highway network with 200 hidden units is used to predict the scale and shift parameters.
@@
 %Since training flow-based models is a bit unstable, in some cases we had to restart the training when the training failed.
-For a more in-depth description, we refer to the codebase and the Appendix C of \citet{kirichenko_why_2020}.
+For a more in-depth description, we refer to the codebase and the Appendix C of \textcite{kirichenko_why_2020}.
 
@@
 We trained each model for 1000 epochs using Adam optimizer with learning rate $3e-4$ and a batch-size of $128$.
-All models were initialized using the data-dependent initialization and they used weight-normalization \citep{salimans2016weight}.
+All models were initialized using the data-dependent initialization and they used weight-normalization \citep{salimans_weight_2016}.
 In addition to that, we always consider a hierarchy of three latent variables.
 For greyscale images (MNIST and FashionMNIST) we used a latent dimension of $8-16-8$ for $\mathbf{z}_1, \mathbf{z}_2, \mathbf{z}_3$ respectively, while for natural images (CIFAR10 and SVHN) we used $8-16-32$.
-For a more in depth description of the model, we refer to Appendix B of \citet{havtorn_hierarchical_2021}.
+For a more in depth description of the model, we refer to Appendix B of \textcite{havtorn_hierarchical_2021}.
 
@@
     \end{subtable} 
-    \label{tab:log_like}
+    \label{tab_modelagnostic:log_like}
     %\vspace*{-\baselineskip}
@@
 To test if the typicality test and the score statistic are uncorrelated, we plot the two scores computed on the validation set.
-As can be seen from figure \ref{fig:corralation}, we have that the two measures are not correlated as it is also highlight by the correlation coefficient.
+As can be seen from \cref{fig_modelagnostic:corralation}, we have that the two measures are not correlated as it is also highlight by the correlation coefficient.
 
@@
     \centering
-    \includegraphics[scale=0.5]{imgs/correlation_correct_font.pdf}
+    \includegraphics[scale=0.5]{graphics/paper_modelagnostic/correlation_correct_font.pdf}
     \caption{Correlation of Typicality Test and Score Statistic computed on the validation set using a PixelCNN++ trained on FashionMNIST.
@@
 }
-    \label{fig:corralation}
+    \label{fig_modelagnostic:corralation}
 \end{figure}
@@
 Indeed, since in our setting we have that the typicality and the score statistic are independent, we would expect this to work worse than the Fisher's combination.
-This is confirmed by table~\ref{tab:all_combinations}, where we are reporting the results when combining the two statistics using the three different ways we analyzed.
+This is confirmed by \cref{tab_modelagnostic:all_combinations}, where we are reporting the results when combining the two statistics using the three different ways we analyzed.
 
 \subsection{Results considering maximum-mean-discrepancy}
-In Section~\ref{sec:section4}, we discussed the relationship between the maximum-mean-discrepancy with a Fisher kernel and the score statistic and the gradient norm, which depends on the choice of approximation of the Fisher information matrix we use.
-In table~\ref{tab:mmd} we reported also the AUROC scores for the MMD with Fisher kernel considering both the diagonal approximation of the FIM (called \emph{MMD Diagonal} in the table) and the FIM being the identity matrix (called \emph{MMD Identity}).
+In \cref{sec_modelagnostic:section4}, we discussed the relationship between the maximum-mean-discrepancy with a Fisher kernel and the score statistic and the gradient norm, which depends on the choice of approximation of the Fisher information matrix we use.
+In \cref{tab_modelagnostic:mmd} we reported also the AUROC scores for the MMD with Fisher kernel considering both the diagonal approximation of the FIM (called \emph{MMD Diagonal} in the table) and the FIM being the identity matrix (called \emph{MMD Identity}).
 As expected, we have that the AUROC of the MMD with the diagonal approximated FIM is pretty close to the AUROC we obtained by using the score statistic.
@@
 For the score statistic, instead, we just need the FIM.
-In addition to that, from all our experiments (see table~\ref{tab:mmd} and table~\ref{tab:mmd_pair2}) we do not have any evidence for one statistic working better than the other, because they are always pretty close to each other.
+In addition to that, from all our experiments (see \cref{tab_modelagnostic:mmd} and \cref{tab_modelagnostic:mmd_pair2}) we do not have any evidence for one statistic working better than the other, because they are always pretty close to each other.
 
@@
 %     \caption{AUROC for single-sample OOD detection}
-%     \label{tab:my_label}
+%     \label{tab_modelagnostic:my_label}
 % \end{table*}
@@
         \end{tabular}
-        \label{tab:mmd}
+        \label{tab_modelagnostic:mmd}
     }
@@
         \end{tabular}
-        \label{tab:all_combinations}
+        \label{tab_modelagnostic:all_combinations}
     \vspace*{-\baselineskip}
@@
 Those trained with Adam get a log-likelihood of $3.63$ bits/dim, while the ones trained with RMSProp get $3.62$ bits/dim.
-Results are shown in Table~\ref{tab:variability}.
+Results are shown in \cref{tab_modelagnostic:variability}.
 It can be noticed, that although the models are similar in terms of test bits/dim the statistics vary a lot, mostly when training with RMSProp.
@@
         \end{tabular}
-        \label{tab:variability}
+        \label{tab_modelagnostic:variability}
     \vspace*{-\baselineskip}
@@
 Although one should use a False Discovery Rate control procedure when the statistics we are using are strong, for completeness, we will present what happens when we apply the BH procedure on a model trained on CIFAR10.
-In Fig.~\ref{fig:type1_cifar}, we report the Type I error ratio and the Type II error ratio for different significance levels $\alpha$.
+In \cref{fig_modelagnostic:type1_cifar}, we report the Type I error ratio and the Type II error ratio for different significance levels $\alpha$.
 We can see that we can actually control the FDR for $\alpha > 0.2$, and for these significance levels we are actually controlling the FDR.
@@
     \centering
-    \includegraphics[scale=0.5]{imgs/cifar_hvae_type1_correct_font.pdf}
+    \includegraphics[scale=0.5]{graphics/paper_modelagnostic/cifar_hvae_type1_correct_font.pdf}
     \caption{Type I and Type II errors versus the significance level $\alpha$ on the combination values.
@@
 For $\alpha >0.2$, since we are using Benjamini-Hochberg procedure, we get that the Type I error stays below identity line.}
-    \label{fig:type1_cifar}
+    \label{fig_modelagnostic:type1_cifar}
      \vspace*{-\baselineskip}
@@
 For both tasks, we trained two Glow models, one trained with Adam and one trained with RMSProp, one PixelCNN++ trained with dropout and a hierarchical-VAE.
-Results are reported in table~\ref{tab:results_second}.
+Results are reported in \cref{tab_modelagnostic:results_second}.
 We can see that almost all the statistics we considered are able to almost perfectly distinguish between the in-distribution test-set and the OOD test-set.
 However, we can notice that the gradient norm is failing sometimes both when we trained on CIFAR10 and when we trained on FashionMNIST.
-From table~\ref{tab:mmd_pair2}, instead, it is clear that we need to approximate the diagonal of the Fisher Information Matrix because if we simply consider the identity matrix, this will also fail as the gradient norm is doing.
+From \cref{tab_modelagnostic:mmd_pair2}, instead, it is clear that we need to approximate the diagonal of the Fisher Information Matrix because if we simply consider the identity matrix, this will also fail as the gradient norm is doing.
 
@@
         \end{tabular}
-        \label{tab:results_second}
+        \label{tab_modelagnostic:results_second}
     }
@@
         \end{tabular}
-        \label{tab:mmd_pair2}
+        \label{tab_modelagnostic:mmd_pair2}
     }
@@
 We consider the two pairs of datasets as before, i.e.\@ FashionMNIST vs MNIST and CIFAR10 vs SVHN.
-Results can be seen in Table~\ref{tab:gmm_results} and Table~\ref{tab:ppca_results}.
+Results can be seen in \cref{tab_modelagnostic:gmm_results} and \cref{tab_modelagnostic:ppca_results}.
 For both GMM and PPCA trained on FashionMNIST the likelihood can be used to perform OOD detection.
@@
         \end{tabular}
-        \label{tab:gmm_results}
+        \label{tab_modelagnostic:gmm_results}
     }
@@
         \end{tabular}
-        \label{tab:ppca_results}
+        \label{tab_modelagnostic:ppca_results}
     }
@@
 All models have the same architecture and were trained with the same procedure.
-Results can be found in Table~\ref{table:consistency_results}.
+Results can be found in \cref{table:consistency_results}.
 For the models trained on CIFAR10, most of the variability in terms of performance is due to the score statistic, which has the highest standard deviation.
@@
 We considered three different interpolations: bilinear (default in PyTorch), Lanczos, and nearest.
-As can be seen from Fig.~\ref{fig:celeba}, these transformations mostly affect the sharpness of the images.
-In Table~\ref{table:celeba} we show how the OOD performance changes for our considered models when testing on CelebA where we applied different interpolations.
+As can be seen from \cref{fig_modelagnostic:celeba}, these transformations mostly affect the sharpness of the images.
+In \cref{table:celeba} we show how the OOD performance changes for our considered models when testing on CelebA where we applied different interpolations.
 We can notice that when using the bilinear interpolation we get results that are pretty similar to \cite{hendrycks_deep_2019}, \cite{kirichenko_why_2020}, and \cite{ahmadian_likelihoodfree_2021} in terms of likelihood OOD performance.
@@
     \centering
-    \includegraphics[scale=0.5]{imgs/celeba_difference_correct_font.pdf}
+    \includegraphics[scale=0.5]{graphics/paper_modelagnostic/celeba_difference_correct_font.pdf}
     \caption{Comparison of different interpolation methods for CelebA dataset.}
-    \label{fig:celeba}
+    \label{fig_modelagnostic:celeba}
      \vspace*{-\baselineskip}
@@
 \section{Algorithmic implementation}
-A pseudocode describing step-by-step how to implement our method is given in Algorithm~\ref{algo_description}.
+A pseudocode describing step-by-step how to implement our method is given in \cref{alg_modelagnostic:algo_description}.
 
@@
 \begin{algorithmic}
-\STATE \textbf{Input}: Training data $\textbf{X} = (x_1,\ldots,x_m)^T$, validation data $\textbf{X}'$, trained model $p_{\theta}(x)$.\\
+\State \textbf{Input}: Training data $\textbf{X} = (x_1,\ldots,x_m)^T$, validation data $\textbf{X}'$, trained model $p_{\theta}(x)$.\\
 \textbf{} \\
@@
 \textbf{Sample} $S$ $M'$-sized datasets from $\textbf{X}'$ using bootstrap resampling. \\
-\textit{(For single-sample OOD we just cycle through each example, see Sec.~\ref{sec:combination_p_values})}\\
+\textit{(For single-sample OOD we just cycle through each example, see \cref{sec_modelagnostic:combination_p_values})}\\
 \textbf{Initialize}   $T^{\text{typicality}} = [\hspace{0.5mm}]$ and $T^{\text{score}} = [\hspace{0.5mm}]$ \\
@@
 \textbf{Compute} $\hat{F}^{\text{typicality}}$ and $\hat{F}^{\text{score}}$, the two empirical CDFs, from $T^{\text{typicality}}$ and $T^{\text{score}}$.
-For example, we used \texttt{statsmodels} library \citep{seabold2010statsmodels}.
+For example, we used \texttt{statsmodels} library \citep{seabold_statsmodels_2010}.
 \\
@@
 \hspace{1.5em} \textbf{Compute} $p$-values $p_{T} = 1 - \hat{F}^{\text{typicality}}(\hat{t})$ and $p_{S} = 1 - \hat{F}^{\text{score}}(\tilde{s})$ \\
-\hspace{1.5em} \textbf{Combine} the two $p$-values using Fisher's method Eq.~\ref{eq:Fisher_method}
+\hspace{1.5em} \textbf{Combine} the two $p$-values using Fisher's method \cref{eq_modelagnostic:Fisher_method}
 \end{algorithmic}
 \caption{Computing $p$-values for OOD detection using a trained generative model.}
-\label{algo_description}
+\label{alg_modelagnostic:algo_description}
 %\end{small}

### 2023-07-20 chapters/paper_review/03_approaches.tex
Formatting
Fix some citation references

@@
     \item DeCoAR 2.0~\cite{ling_decoar_2020} presents a deep contextualized acoustic representation learning approach with the addition of a vector quantization (VQ) layer.
-    \item In VQ-APC~\cite{chung_vector-quantized_2020}, a VQ layer is used with the APC objective, which imposes a bottleneck and forces the model to learn better representations.
+    \item In VQ-APC~\cite{chung_vectorquantized_2020}, a VQ layer is used with the APC objective, which imposes a bottleneck and forces the model to learn better representations.
     \item Two dropout regularization methods, attention dropout and layer dropout, are introduced to TERA~\cite{luo_dropout_2021}. 
@@
 on previous context. 
-In VQ-APC~\cite{chung_vector-quantized_2020}, quantization is used with the APC
+In VQ-APC~\cite{chung_vectorquantized_2020}, quantization is used with the APC
 objective, which imposes an information bottleneck serving as a regularizer.
@@
 Some studies explore alternatives to masking the input directly.
-In non-autoregressive predictive coding (NPC)~\cite{liu_non-autoregressive_2020}, time masking is introduced through masked convolution blocks.
+In non-autoregressive predictive coding (NPC)~\cite{liu_nonautoregressive_2020}, time masking is introduced through masked convolution blocks.
 Taking inspiration from XLNet~\cite{yang_xlnet_2019}, it has also been suggested that the input be reconstructed from a shuffled version~\cite{song_speechxlnet_2020} to address the discrepancy between pre-training and fine-tuning of masking-based approaches.
@@
 It has subsequently} inspired many follow-up studies.
-The wav2vec-C~\cite{sadhu_wav2vec-c_2021} approach extends wav2vec 2.0 with a consistency term in the loss that aims to reconstruct the input features from the learned quantized representations, similar to VQ-VAE~\cite{razavi_generating_2019}.
+The wav2vec-C~\cite{sadhu_wav2vecc_2021} approach extends wav2vec 2.0 with a consistency term in the loss that aims to reconstruct the input features from the learned quantized representations, similar to VQ-VAE~\cite{razavi_generating_2019}.
 
@@
 \paragraph{Discrete BERT} \edit{Applying BERT-type training directly to speech input is not possible due to its continuous nature}.
-The Discrete BERT approach~\cite{baevski_effectiveness_2020} uses a pre-trained vq-wav2vec model to derive a discrete vocabulary \cite{baevski_vq-wav2vec_2020}.
+The Discrete BERT approach~\cite{baevski_effectiveness_2020} uses a pre-trained vq-wav2vec model to derive a discrete vocabulary \cite{baevski_vqwav2vec_2020}.
 The vq-wav2vec model is similar to wav2vec mentioned in \cref{par:cpc} but uses quantization to learn discrete representations.

### 2023-07-20 chapters/paper_review/04_benchmarking.tex
Fix some citation references

@@
 self-supervised pre-training have been collected at 
-Amazon~\cite{sadhu_wav2vec-c_2021}.
+Amazon~\cite{sadhu_wav2vecc_2021}.
 
@@
 arrays. A team at Amazon Alexa also recorded and transcribed a corpus of 1k
-hours of audio for model training and evaluation~\cite{sadhu_wav2vec-c_2021}.
+hours of audio for model training and evaluation~\cite{sadhu_wav2vecc_2021}.
 
@@
     % vq-wav2vec
-    \multirow{2}{*}[0mm]{vq-wav2vec \cite{baevski_vq-wav2vec_2020}} & \multirow{2}{*}[0mm]{LS 960 hrs} & ASR & WSJ si284 & \makecell[l]{WSJ eval92} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{vq-wav2vec \cite{baevski_vqwav2vec_2020}} & \multirow{2}{*}[0mm]{LS 960 hrs} & ASR & WSJ si284 & \makecell[l]{WSJ eval92} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
     & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \hline
@@
     % vq-APC
-    \multirow{2}{*}[0mm]{vq-APC \cite{chung_vector-quantized_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{vq-APC \cite{chung_vectorquantized_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
     & & SID & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & \checkmark & 65\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline
@@
     % NPC
-    \multirow{2}{*}[0mm]{NPC \cite{liu_non-autoregressive_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{NPC \cite{liu_nonautoregressive_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
     & & SID & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & \checkmark & 65\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline
@@
     % wav2vec-c
-    wav2vec-c \cite{sadhu_wav2vec-c_2021} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
+    wav2vec-c \cite{sadhu_wav2vecc_2021} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
     UniSpeech-SAT \cite{chen_unispeechsat_2021} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See SUPERB \cite{yang_superb_2021} paper for details \\ \hline
@@
 wav2vec 2.0~\cite{baevski_wav2vec_2020}, and HuBERT~\cite{hsu_hubert_2021}, Alexa internal
-datasets (10k hours) for wav2vec-c~\cite{sadhu_wav2vec-c_2021}, Didi internal
+datasets (10k hours) for wav2vec-c~\cite{sadhu_wav2vecc_2021}, Didi internal
 datasets (10k hours) for MPC~\cite{jiang_improving_2019, jiang_further_2021}, the

### 2023-07-20 chapters/paper_review/model.tex
Fix some citation references

@@
 
-vq-wav2vec \cite{baevski_vq-wav2vec_2020} & 2019 Oct.
+vq-wav2vec \cite{baevski_vqwav2vec_2020} & 2019 Oct.
 & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ %12 Oct 2019 %Lee: vq-wav2vec is CON or CLF???
@@
 
-Discrete BERT \cite{baevski_vq-wav2vec_2020,baevski_effectiveness_2020} & 2019 Oct.
+Discrete BERT \cite{baevski_vqwav2vec_2020,baevski_effectiveness_2020} & 2019 Oct.
 & \xmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ %12 Oct 2019 %10 Nov 2019 %In the paper of vq-wav2vec, it also includes the approach of Discrete BERT.
@@
 
-VQ-APC \cite{chung_vector-quantized_2020}  & 2020 May  & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark \\ %17/05
+VQ-APC \cite{chung_vectorquantized_2020}  & 2020 May  & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark \\ %17/05
 
@@
 
-NPC \cite{liu_non-autoregressive_2020}                     & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ % 1/11
+NPC \cite{liu_nonautoregressive_2020}                     & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ % 1/11
 
@@
 
-wav2vec-C \cite{sadhu_wav2vec-c_2021} & 2021 Mar. & \cmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ %9 Mar 2021 %Lee:  consistency loss is reconstruction
+wav2vec-C \cite{sadhu_wav2vecc_2021} & 2021 Mar. & \cmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ %9 Mar 2021 %Lee:  consistency loss is reconstruction
 
@@
 
-NPC \cite{liu_non-autoregressive_2020}             & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 1/11
+NPC \cite{liu_nonautoregressive_2020}             & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 1/11
 

### 2023-07-20 chapters/paper_review/pretext.tex
Add and revise short table and figure captions
Fix some citation references

@@
     \centering
-    \caption{
-    A summary of the approaches in the three categories of
+    \caption[An overview of approaches within the three main categories of
+    self-supervised learning.]{
+    An overview of approaches within the three categories of
     self-supervised learning. 
@@
     \midrule 
-    APC~\cite{chung_unsupervised_2019,chung_vector-quantized_2020}         & $X_{[1,t]}$   & \textsc{-}              & $x_{t+c},\, c\geq1$    \\ %VQ-APC is the same as APC
+    APC~\cite{chung_unsupervised_2019,chung_vectorquantized_2020}         & $X_{[1,t]}$   & \textsc{-}              & $x_{t+c},\, c\geq1$    \\ %VQ-APC is the same as APC
     \midrule
@@
     \midrule
-    NPC~\cite{liu_non-autoregressive_2020}      & $X$  &   $Z_{-[t,t+k]}$  &    $X$   \\ %Lee: I am not 100% sure it is correct. Please check.
+    NPC~\cite{liu_nonautoregressive_2020}      & $X$  &   $Z_{-[t,t+k]}$  &    $X$   \\ %Lee: I am not 100% sure it is correct. Please check.
     \midrule
@@
     \midrule 
-    vq-wav2vec \cite{baevski_vq-wav2vec_2020}     &   $X_{[1,t]}$ &   \textsc{-}    &   $q_{t+c},\, c\geq1$   \\ 
+    vq-wav2vec \cite{baevski_vqwav2vec_2020}     &   $X_{[1,t]}$ &   \textsc{-}    &   $q_{t+c},\, c\geq1$   \\ 
     \midrule 
-    wav2vec 2.0 \cite{baevski_wav2vec_2020}, wav2vec-C \cite{sadhu_wav2vec-c_2021}\footnote{wav2vec-C adds reconstruction loss to wav2vec 2.0.}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
+    wav2vec 2.0 \cite{baevski_wav2vec_2020}, wav2vec-C \cite{sadhu_wav2vecc_2021}\footnote{wav2vec-C adds reconstruction loss to wav2vec 2.0.}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
     \midrule 
@@
     \midrule
-    Discrete BERT~\cite{baevski_vq-wav2vec_2020,baevski_effectiveness_2020} \footnote{Discrete BERT obtains codes $C$ from vq-wav2vec.}      &   \multicolumn{2}{c|}{$C_{-[t,t+k]}$}   & $C_{[t,t+k]}$  \\
+    Discrete BERT~\cite{baevski_vqwav2vec_2020,baevski_effectiveness_2020} \footnote{Discrete BERT obtains codes $C$ from vq-wav2vec.}      &   \multicolumn{2}{c|}{$C_{-[t,t+k]}$}   & $C_{[t,t+k]}$  \\
     \midrule 

### 2023-07-21 appendices/appendix_hierarchical.tex
Add option to not compile appendices
Add comment about freezing compiler

@@
 \chapter[supplementary material for hierarchical vaes know what they don't know]{Supplementary material for Hierarchical VAEs know what they don't know}
+\label{chp:appendix-hierarchical}
+\ifthenelse{\equal{\skipappendices}{true}}{}{
 
@@
     \includegraphics[width=\columnwidth]{paper_hierarchical/logdet2.pdf}
-    \caption{The expected inverse volume change for Gaussian Jacobians \eqref{eq:Edet} on a log-scale.}
+    \caption{The expected inverse volume change for Gaussian Jacobians \eqref{eq_hierarchical:Edet} on a log-scale.}
     \label{fig_hierarchical:Edet}
@@
     &= \sigma^{-d_{i-1}}  2^{-\frac{d_{i-1}}{2}} \frac{\Gamma\left( \frac{1}{2} (d_{i} - d_{i-1}) \right)}{\Gamma\left( \frac{1}{2} d_{i} \right)}
-    \label{eq:Edet}
+    \label{eq_hierarchical:Edet}
 \end{align}
 where $\Gamma_d$ is the multivariate Gamma function.
-Assuming that the increase in layer dimension $d_{i} - d_{i-1}$ is constant, then we see that \eqref{eq:Edet} goes to zero as $d_{i}$ goes to infinity as the $\Gamma$ function grows super-exponentially to infinity.
+Assuming that the increase in layer dimension $d_{i} - d_{i-1}$ is constant, then we see that \eqref{eq_hierarchical:Edet} goes to zero as $d_{i}$ goes to infinity as the $\Gamma$ function grows super-exponentially to infinity.
 This super-exponential growth further implies that the first layers dominate the marginal likelihood $p(\xb)$.
@@
 
+\moretodo[inline]{LaTeX won't compile beyond this point and it is unclear why.}
+\iffalse
+% TODO (JDH): LaTeX won't compile beyond this point and it is unclear why. 
 \section{Derivation of the $\mathcal{L}^{>k}$ bound}
@@
 We can generalize the $\mathcal{L}^{>k}$ bound by introducing the flipped version, $\mathcal{L}^{<l}$, which compared to $\mathcal{L}^{>k}$, instead samples the $L-l$ \textit{highest} latent variables in the hierarchy from the prior $\zb_{l},\dots,\zb_L \sim p_\theta(\zb_{\geq l})=p_\theta(\zb_l|\zb_{l+1})\cdots p_\theta(\zb_L)$ and the remaining lower latents from the approximate posterior $\hat{\zb}_{1},\dots,\hat{\zb}_{l-1} \sim q_\phi(\zb_{<l}|\xb)=q_\phi(\zb_{1}|\xb)q_\phi(\zb_{2}|\zb_1)\cdots q_\phi(\zb_{l-1}|\zb_{l-2})$,
-\begin{equation}\label{eq:biva-<l}
+\begin{equation}\label{eq_hierarchical:biva-<l}
     \mathcal{L}^{<l} = \mathbb{E}_{p_\theta(\zb_{\geq l}) q_\phi(\zb_{<l}|\xb)} \left[ \log \frac{p_\theta(\xb,\zb)p_\theta(\zb_{<l})}{q_\phi(\zb_{<l}|\xb)} \right].
@@
     % \section{Generalized log-likelihood ratio $LLR^{>k}_{<l}$}
-    % We can further generalize the log-likelhood ratio $LLR^{>k}$ by replacing the ELBO $\mathcal{L}$ with the bound in \eqref{eq:biva-<l} to obtain
+    % We can further generalize the log-likelhood ratio $LLR^{>k}$ by replacing the ELBO $\mathcal{L}$ with the bound in \eqref{eq_hierarchical:biva-<l} to obtain
     % \begin{equation}
@@
 \begin{align}
-    \log p(\xb) &\ge \int \int \log p(\xb|\zb_1) \text{d}\zb_2\zb_1 \label{eq:2 layer VAE bottom up ELBO integrals} \\
+    \log p(\xb) &\ge \int \int \log p(\xb|\zb_1) \text{d}\zb_2\zb_1 \label{eq_hierarchical:2 layer VAE bottom up ELBO integrals} \\
     &\qquad + \int q(\zb_1|\xb) \int q(\zb_2|\zb_1) \log \frac{p(\zb_1|\zb_2)}{q(\zb_1|\xb)} \text{d}\zb_2\zb_1 \notag \\ 
@@
 % Results of MNIST model. (57bed090cf4a4e44b896b4c2cce55500)
+\fi
+}

### 2023-08-08 appendices/appendix_benchmarking.tex
Add option to not compile appendices
Fix compilation issue and revise captions

@@
 \chapter[supplementary material for benchmarking generative latent variable models for speech]{Supplementary material for Benchmarking generative latent variable models for speech}
+\label{app:benchmarking}
+\ifthenelse{\equal{\skipappendices}{true}}{}{
 
@@
     \vspace{2mm}
-    \caption[The highest possible Gaussian log-likelihoods attainable on the TIMIT test]{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
+    \caption[The highest possible Gaussian log-likelihoods attainable on the TIMIT test set]{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
     \label{tab:timit best gaussian ll}
@@
 
-\iffalse
-
-\fi
+}

### 2023-08-08 appendices/appendix_hierarchical.tex
Fix compilation issue and revise captions

@@
 
-\moretodo[inline]{LaTeX won't compile beyond this point and it is unclear why.}
-\iffalse
-% TODO (JDH): LaTeX won't compile beyond this point and it is unclear why. 
-\section{Derivation of the $\mathcal{L}^{>k}$ bound}
+\section{Derivation of the \texorpdfstring{$\mathcal{L}^{>k}$}{Lk} bound}
 In this section we present the derivation of $\mathcal{L}^{>k}$ and show that it is a lower bound on the marginal likelihood.
@@
 
-\section{The complementary $\mathcal{L}^{<l}$ bound}
+\section{The complementary \texorpdfstring{$\mathcal{L}^{<l}$}{Ll} bound}
 We can generalize the $\mathcal{L}^{>k}$ bound by introducing the flipped version, $\mathcal{L}^{<l}$, which compared to $\mathcal{L}^{>k}$, instead samples the $L-l$ \textit{highest} latent variables in the hierarchy from the prior $\zb_{l},\dots,\zb_L \sim p_\theta(\zb_{\geq l})=p_\theta(\zb_l|\zb_{l+1})\cdots p_\theta(\zb_L)$ and the remaining lower latents from the approximate posterior $\hat{\zb}_{1},\dots,\hat{\zb}_{l-1} \sim q_\phi(\zb_{<l}|\xb)=q_\phi(\zb_{1}|\xb)q_\phi(\zb_{2}|\zb_1)\cdots q_\phi(\zb_{l-1}|\zb_{l-2})$,
@@
         \midrule
-        CIFAR10            &  $L^{>0}$         &  $0.992$  &  $0.993$  &  $0.004$  \\
-        CIFAR10            &  $L^{>1}$         &  $0.988$  &  $0.990$  &  $0.002$  \\
-        CIFAR10            &  $L^{>2}$         &  $0.746$  &  $0.756$  &  $0.468$  \\
+        CIFAR10            &  $\mathcal{L}^{>0}$         &  $0.992$  &  $0.993$  &  $0.004$  \\
+        CIFAR10            &  $\mathcal{L}^{>1}$         &  $0.988$  &  $0.990$  &  $0.002$  \\
+        CIFAR10            &  $\mathcal{L}^{>2}$         &  $0.746$  &  $0.756$  &  $0.468$  \\
         CIFAR10            &  $LLR^{>1}$       &  $0.939$  &  $0.950$  &  $0.052$  \\
         % \midrule
-        % UniformNoise       &  $L^{>0}$         &  $1.000$  &  $1.000$  &  $0.000$  \\
-        % UniformNoise       &  $L^{>1}$         &  $1.000$  &  $1.000$  &  $0.000$  \\
-        % UniformNoise       &  $L^{>2}$         &  $0.965$  &  $0.957$  &  $0.055$  \\
+        % UniformNoise       &  $\mathcal{L}^{>0}$         &  $1.000$  &  $1.000$  &  $0.000$  \\
+        % UniformNoise       &  $\mathcal{L}^{>1}$         &  $1.000$  &  $1.000$  &  $0.000$  \\
+        % UniformNoise       &  $\mathcal{L}^{>2}$         &  $0.965$  &  $0.957$  &  $0.055$  \\
         % UniformNoise       &  $LLR^{>1}$       &  $0.984$  &  $0.992$  &  $0.000$  \\
         \midrule
-        SVHN               &  $L^{>0}$         &  $0.599$  &  $0.587$  &  $0.702$  \\
-        SVHN               &  $L^{>1}$         &  $0.555$  &  $0.543$  &  $0.755$  \\
-        SVHN               &  $L^{>2}$         &  $0.403$  &  $0.431$  &  $0.869$  \\
+        SVHN               &  $\mathcal{L}^{>0}$         &  $0.599$  &  $0.587$  &  $0.702$  \\
+        SVHN               &  $\mathcal{L}^{>1}$         &  $0.555$  &  $0.543$  &  $0.755$  \\
+        SVHN               &  $\mathcal{L}^{>2}$         &  $0.403$  &  $0.431$  &  $0.869$  \\
         SVHN               &  $LLR^{>1}$       &  $0.489$  &  $0.484$  &  $0.799$  \\
@@
         \midrule
-        SVHN          &  $L^{>0}$    &  0.083  &  0.318  &  0.974  \\
-        SVHN          &  $L^{>1}$    &  0.097  &  0.320  &  0.972  \\
-        SVHN          &  $L^{>2}$    &  0.693  &  0.725  &  0.599  \\
+        SVHN          &  $\mathcal{L}^{>0}$    &  0.083  &  0.318  &  0.974  \\
+        SVHN          &  $\mathcal{L}^{>1}$    &  0.097  &  0.320  &  0.972  \\
+        SVHN          &  $\mathcal{L}^{>2}$    &  0.693  &  0.725  &  0.599  \\
         SVHN          &  $LLR^{>2}$  &  0.811  &  0.837  &  0.394  \\
         % \midrule
-        % UniformNoise  &  $L^{>0}$    &  1.000  &  1.000  &  0.000  \\
-% UniformNoise  &  $L^{>1}$    &  0.987  &  0.984  &  0.025  \\
-% UniformNoise  &  $L^{>2}$    &  0.031  &  0.313  &  0.971  \\
+        % UniformNoise  &  $\mathcal{L}^{>0}$    &  1.000  &  1.000  &  0.000  \\
+% UniformNoise  &  $\mathcal{L}^{>1}$    &  0.987  &  0.984  &  0.025  \\
+% UniformNoise  &  $\mathcal{L}^{>2}$    &  0.031  &  0.313  &  0.971  \\
 % UniformNoise  &  $LLR^{>1}$  &  0.206  &  0.436  &  0.999  \\
 \midrule
-CIFAR10       &  $L^{>0}$    &  0.485  &  0.488  &  0.817  \\
-CIFAR10       &  $L^{>1}$    &  0.467  &  0.476  &  0.822  \\
-CIFAR10       &  $L^{>2}$    &  0.411  &  0.433  &  0.869  \\
+CIFAR10       &  $\mathcal{L}^{>0}$    &  0.485  &  0.488  &  0.817  \\
+CIFAR10       &  $\mathcal{L}^{>1}$    &  0.467  &  0.476  &  0.822  \\
+CIFAR10       &  $\mathcal{L}^{>2}$    &  0.411  &  0.433  &  0.869  \\
 CIFAR10       &  $LLR^{>1}$  &  0.469  &  0.479  &  0.835  \\
@@
          \midrule
-FashionMNIST                     &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-FashionMNIST                     &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-FashionMNIST                     &  $L^{>2}$  &  $0.981$  &  $0.983$  &  $0.003$ \\
+FashionMNIST                     &  $\mathcal{L}^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+FashionMNIST                     &  $\mathcal{L}^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+FashionMNIST                     &  $\mathcal{L}^{>2}$  &  $0.981$  &  $0.983$  &  $0.003$ \\
 FashionMNIST                   &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
 \midrule
-notMNIST                         &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-notMNIST                         &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-notMNIST                         &  $L^{>2}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+notMNIST                         &  $\mathcal{L}^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+notMNIST                         &  $\mathcal{L}^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+notMNIST                         &  $\mathcal{L}^{>2}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
 notMNIST                       &  $LLR^{>1}$  &  $1.000$  &  $0.999$  &  $0.000$ \\
 \midrule
-KMNIST                           &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-KMNIST                           &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-KMNIST                           &  $L^{>2}$  &  $0.987$  &  $0.987$  &  $0.011$ \\
+KMNIST                           &  $\mathcal{L}^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+KMNIST                           &  $\mathcal{L}^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+KMNIST                           &  $\mathcal{L}^{>2}$  &  $0.987$  &  $0.987$  &  $0.011$ \\
 KMNIST                         &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
 \midrule
-Omniglot28x28                    &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-Omniglot28x28                    &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-Omniglot28x28                    &  $L^{>2}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+Omniglot28x28                    &  $\mathcal{L}^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+Omniglot28x28                    &  $\mathcal{L}^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+Omniglot28x28                    &  $\mathcal{L}^{>2}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
 Omniglot28x28                  &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
 \midrule
-Omniglot28x28Inverted            &  $L^{>0}$  &  $0.862$  &  $0.902$  &  $0.205$ \\
-Omniglot28x28Inverted            &  $L^{>1}$  &  $0.923$  &  $0.943$  &  $0.056$ \\
-Omniglot28x28Inverted            &  $L^{>2}$  &  $0.749$  &  $0.691$  &  $0.411$ \\
+Omniglot28x28Inverted            &  $\mathcal{L}^{>0}$  &  $0.862$  &  $0.902$  &  $0.205$ \\
+Omniglot28x28Inverted            &  $\mathcal{L}^{>1}$  &  $0.923$  &  $0.943$  &  $0.056$ \\
+Omniglot28x28Inverted            &  $\mathcal{L}^{>2}$  &  $0.749$  &  $0.691$  &  $0.411$ \\
 Omniglot28x28Inverted          &  $LLR^{>1}$  &  $0.944$  &  $0.953$  &  $0.057$ \\
 \midrule
-SmallNORB28x28                   &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-SmallNORB28x28                   &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-SmallNORB28x28                   &  $L^{>2}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+SmallNORB28x28                   &  $\mathcal{L}^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+SmallNORB28x28                   &  $\mathcal{L}^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+SmallNORB28x28                   &  $\mathcal{L}^{>2}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
 SmallNORB28x28                 &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
 \midrule
-SmallNORB28x28Inverted           &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-SmallNORB28x28Inverted           &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-SmallNORB28x28Inverted           &  $L^{>2}$  &  $0.977$  &  $0.980$  &  $0.001$ \\
+SmallNORB28x28Inverted           &  $\mathcal{L}^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+SmallNORB28x28Inverted           &  $\mathcal{L}^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+SmallNORB28x28Inverted           &  $\mathcal{L}^{>2}$  &  $0.977$  &  $0.980$  &  $0.001$ \\
 SmallNORB28x28Inverted         &  $LLR^{>1}$  &  $0.985$  &  $0.987$  &  $0.000$ \\
 % \midrule
-% UniformNoise                     &  $L^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-% UniformNoise                     &  $L^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-% UniformNoise                     &  $L^{>2}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+% UniformNoise                     &  $\mathcal{L}^{>0}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+% UniformNoise                     &  $\mathcal{L}^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
+% UniformNoise                     &  $\mathcal{L}^{>2}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
 % UniformNoise                   &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
 \midrule
-MNIST                            &  $L^{>0}$  &  $0.488$  &  $0.486$  &  $0.807$ \\
-MNIST                            &  $L^{>1}$  &  $0.469$  &  $0.469$  &  $0.816$ \\
-MNIST                            &  $L^{>2}$  &  $0.514$  &  $0.505$  &  $0.791$ \\
+MNIST                            &  $\mathcal{L}^{>0}$  &  $0.488$  &  $0.486$  &  $0.807$ \\
+MNIST                            &  $\mathcal{L}^{>1}$  &  $0.469$  &  $0.469$  &  $0.816$ \\
+MNIST                            &  $\mathcal{L}^{>2}$  &  $0.514$  &  $0.505$  &  $0.791$ \\
 MNIST                          &  $LLR^{>2}$  &  $0.515$  &  $0.507$  &  $0.792$ \\
@@
 % Results of MNIST model. (57bed090cf4a4e44b896b4c2cce55500)
-\fi
 }

### 2023-08-08 appendices/appendix_modelagnostic.tex
Fix compilation issue and revise captions

@@
     \centering
-    \caption{AUROC$\uparrow$ for single-sample OOD detection.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing two different estimates of the Fisher information matrix]{AUROC$\uparrow$ for single-sample OOD detection.
 Comparison between two different estimates of the Fisher information matrix.
@@
 For ($\mathsection$), instead, we used  $\varepsilon = 10^{-8}$ and $\xi = 0.75$, as suggested by \textcite{martens_new_2020}.
-As a results we have that using Adam parameters choice is slightly better for our task.}
+As a result, we have that using Adam parameters choice is slightly better for our task.}
     \resizebox*{\textwidth}{!}{
@@
         \cmidrule{2-5}
-        \textsc{models}  & \textcolor{blue}{\textsc{MMD Diagonal}} &  \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} \\
+        \textsc{models}  & \textcolor{blue}{\textsc{MMD diagonal}} &  \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}} & \textcolor{red}{\textsc{Fisher's method}} \\
         \midrule
@@
     \caption{Test log-likelihood (bits/dim) achieved by the models used in the paper.}
-    \begin{subtable}{.5\textwidth}
+    \begin{subtable}{.45\textwidth}
         \centering
@@
             \toprule
-            \multicolumn{2}{c}{\textsc{Models trained on FashionMNIST)}}\\
+            \multicolumn{2}{c}{\textsc{Models trained on FashionMNIST}}\\
             \midrule
@@
     \end{subtable}
-    \begin{subtable}{.5\textwidth}
+    \begin{subtable}{.45\textwidth}
         \centering
@@
     \includegraphics[scale=0.5]{graphics/paper_modelagnostic/correlation_correct_font.pdf}
-    \caption{Correlation of Typicality Test and Score Statistic computed on the validation set using a PixelCNN++ trained on FashionMNIST.
+    \caption[Correlation of typicality test and score statistic computed on the validation set using a PixelCNN++ trained on FashionMNIST]{Correlation of typicality test and score statistic computed on the validation set using a PixelCNN++ trained on FashionMNIST.
 The correlation coefficient is $-0.014$.
@@
 In \cref{sec_modelagnostic:section4}, we discussed the relationship between the maximum-mean-discrepancy with a Fisher kernel and the score statistic and the gradient norm, which depends on the choice of approximation of the Fisher information matrix we use.
-In \cref{tab_modelagnostic:mmd} we reported also the AUROC scores for the MMD with Fisher kernel considering both the diagonal approximation of the FIM (called \emph{MMD Diagonal} in the table) and the FIM being the identity matrix (called \emph{MMD Identity}).
+In \cref{tab_modelagnostic:mmd} we reported also the AUROC scores for the MMD with Fisher kernel considering both the diagonal approximation of the FIM (called \emph{MMD diagonal} in the table) and the FIM being the identity matrix (called \emph{MMD identity}).
 As expected, we have that the AUROC of the MMD with the diagonal approximated FIM is pretty close to the AUROC we obtained by using the score statistic.
@@
     %\renewcommand{\thetable}{4}
-    \caption{AUROC$\uparrow$ for single-sample OOD detection.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing individual statistics.]{AUROC$\uparrow$ for single-sample OOD detection.
 In this table we consider all the different single statistics we mentioned in the paper.
-One can notice that MMD Diagonal is pretty close to the score statistic and the MMD Identity is close to the gradient norm, as expected (see Section 4.1 in the paper).}
+One can notice that MMD diagonal is pretty close to the score statistic and the MMD identity is close to the gradient norm, as expected (see Section 4.1 in the paper).}
     \resizebox{\textwidth}{!}{
@@
             \cmidrule{2-7}
-            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{MMD Diagonal}}  & \textcolor{blue}{\textsc{MMD Identity}} &\textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}}  \\
+            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{MMD diagonal}}  & \textcolor{blue}{\textsc{MMD identity}} &\textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}}  \\
             \midrule
@@
     %\renewcommand{\thetable}{4}
-    \caption{AUROC$\uparrow$ for single-sample OOD detection.
-Comparison between the three method we mentioned to combine different statistics.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing three methods of combining different statistics.]{AUROC$\uparrow$ for single-sample OOD detection.
+Comparison between the three methods we mentioned to combine different statistics.
 Since the typicality and the score statistic are not correlated, we have that the Fisher's method is mostly working better than the other two methods.}
@@
     %\renewcommand{\thetable}{4}
-    \caption{AUROC$\uparrow$ for single-sample OOD detection.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing two different Glow models.]{AUROC$\uparrow$ for single-sample OOD detection.
 In this table we are comparing two different Glow models trained on CIFAR10 by considering two different checkpoints with almost the same test log-likelihood.
-We can see that both statistics vary a bit.
-}
+We can see that both statistics vary a bit.}
         \scriptsize
@@
             \cmidrule(r){2-3}\cmidrule(l){4-5}
-            \textsc{models}  &  \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
+            \textsc{models}  &  \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
             \midrule
@@
     \includegraphics[scale=0.5]{graphics/paper_modelagnostic/cifar_hvae_type1_correct_font.pdf}
-    \caption{Type I and Type II errors versus the significance level $\alpha$ on the combination values.
+    \caption[Type I and Type II errors versus the significance level $\alpha$ on the combination values.]{Type I and Type II errors versus the significance level $\alpha$ on the combination values.
 We can control the FDR only for $\alpha > 0.2$ in this case.
@@
 We also evaluated our methods in the two dataset pairs, MNIST against FashionMNIST and SVHN against CIFAR10, that are usually considered easier than the tasks presented in the main paper.
-For both tasks, we trained two Glow models, one trained with Adam and one trained with RMSProp, one PixelCNN++ trained with dropout and a hierarchical-VAE.
+For both tasks, we trained two Glow models, one trained with Adam and one trained with RMSProp, one PixelCNN++ trained with dropout and a hierarchical VAE.
 Results are reported in \cref{tab_modelagnostic:results_second}.
@@
     %\renewcommand{\thetable}{4}
-    \caption{AUROC$\uparrow$ for single-sample OOD detection when training on MNIST and testing again FashionMNIST and when training on SVHN and testing against CIFAR10.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection when training on MNIST and testing against FashionMNIST and when training on SVHN and testing against CIFAR10.]{AUROC$\uparrow$ for single-sample OOD detection when training on MNIST and testing against FashionMNIST and when training on SVHN and testing against CIFAR10.
 As before, Fisher's method is the combination of the typicality test and the test statistic.
@@
             \cmidrule(r){2-5}\cmidrule(l){6-7}
-            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
+            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
             \midrule
@@
     %\renewcommand{\thetable}{4}
-    \caption{AUROC$\uparrow$ for single-sample OOD detection.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection.
+In this table we consider all the different single statistics we mentioned in the paper but for the models trained on MNIST and SVHN this time]{AUROC$\uparrow$ for single-sample OOD detection.
 In this table we consider all the different single statistics we mentioned in the paper but for the models trained on MNIST and SVHN this time.
@@
     %\renewcommand{\thetable}{4}
-    \caption{AUROC$\uparrow$ for single-sample OOD detection using a Gaussian mixture model (GMM).
+    \caption[AUROC$\uparrow$ for single-sample OOD detection using a Gaussian mixture model (GMM).]{AUROC$\uparrow$ for single-sample OOD detection using a Gaussian mixture model (GMM).
 For Fisher's method we mean the combination of the typicality test and the test statistic.
@@
             \cmidrule(r){2-5}\cmidrule(l){6-7}
-            \textsc{components}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
+            \textsc{components}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
             \midrule
@@
     %\renewcommand{\thetable}{4}
-    \caption{AUROC$\uparrow$ for single-sample OOD detection using a Probabilistic PCA.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection using a Probabilistic PCA.]{AUROC$\uparrow$ for single-sample OOD detection using a Probabilistic PCA.
 For Fisher's method we mean the combination of the typicality test and the test statistic.
@@
     % \vspace{-1.3em}
-    \caption{Mean and standard deviation of the performance in terms of AUROC of our method.
+    \caption[Mean and standard deviation of the performance in terms of AUROC of combined statistics for a HVAE.]{Mean and standard deviation of the performance in terms of AUROC of our method.
 Quantities are computed by taking the performance of 5 different trained HVAEs both trained on CIFAR10 and FashionMNIST.}
@@
             \toprule
-            \textsc{$D_{\text{out}}$}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score Stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}}  \\
+            \textsc{$D_{\text{out}}$}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}}  \\
             \midrule
@@
     \vspace{-1.3em}
-    \caption{Comparison between our method and DoSE using the original statistics. In these experiments we considered only Glow trained with Adam.}
+    \caption[Comparison between our method and DoSE using the original statistics]{Comparison between our method and DoSE using the original statistics.
+In these experiments we considered only Glow trained with Adam.}
         \centering
@@
 \hspace{1.5em} \textbf{Compute} MMD Typicality for ${x}_{m'}$ by $\left\| \frac{1}{m'} \sum_{m'=1}^{M'}\log p_\theta ({x}_{m'}) - L(\theta) \right\|_2$ and add it to $T^{\text{typicality}}$\\
-\hspace{1.5em} \textbf{Compute} Score Statistic for ${x}_{m'}$ by $\left\|{I(\theta)}^{-1/2} \frac{1}{m'} \sum_{m'=1}^{M'}\nabla \log p_\theta ({x}_{m'})\right\|_2$ and add it to $T^{\text{score}}$\\
+\hspace{1.5em} \textbf{Compute} Score statistic for ${x}_{m'}$ by $\left\|{I(\theta)}^{-1/2} \frac{1}{m'} \sum_{m'=1}^{M'}\nabla \log p_\theta ({x}_{m'})\right\|_2$ and add it to $T^{\text{score}}$\\
 \textbf{Return} Two vectors of size $S$ containing the two statistics for $T^{\text{typicality}}$ and $T^{\text{score}}$  \\
@@
 \hspace{1.5em} \textbf{Compute} $\frac{1}{n} \sum_{i=1}^{n}\log p_{\theta}(\tilde{{x}}_i)$ and $\frac{1}{n} \sum_{i=1}^{n}\nabla_{{\theta}}\log p_{\theta}(\tilde{{x}}_i)$ \\
-\hspace{1.5em} \textbf{Compute} MMD Typicality $\tilde{t}$ and Score Statistic $\tilde{s}$ \\
+\hspace{1.5em} \textbf{Compute} MMD Typicality $\tilde{t}$ and Score statistic $\tilde{s}$ \\
 \hspace{1.5em} \textbf{Compute} $p$-values $p_{T} = 1 - \hat{F}^{\text{typicality}}(\hat{t})$ and $p_{S} = 1 - \hat{F}^{\text{score}}(\tilde{s})$ \\

### 2023-08-08 chapters/paper_benchmarking.tex
Conditional skipping of compilation
Revise skipping
Fix compilation issue and revise captions

@@
 \chapter[benchmarking generative latent variable models for speech]{Benchmarking Generative Latent Variable Models for Speech}
-\label{chp:benchmarking}
+\label{chp:paper-benchmarking}
+\ifthenelse{\equal{\skippapers}{true}}{}{
 
@@
 \begin{table}[t]
-    \hspace{4mm}
-    \begin{minipage}[t]{.45\textwidth}
+    \begin{subtable}[t]{.45\textwidth}
     \centering
@@
     \label{tab: likelihoods timit}
-    \end{minipage}%
+    \end{subtable}%
     \hfill
-    \begin{minipage}[t]{.45\textwidth}
+    \begin{subtable}[t]{.45\textwidth}
     \centering
@@
     \label{tab: phoneme recognition (PER)}
-    \end{minipage}%
-    \caption*[Model likelihoods and phoneme error rate for TIMIT.]{%
-    In \cref{tab: likelihoods timit} we report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines.
+    \end{subtable}%
+    \caption[Model likelihoods and phoneme error rate for TIMIT.]{In \cref{tab: likelihoods timit} we report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines.
     In \cref{tab: phoneme recognition (PER)} we report phoneme error rate (PER) of different representations for phoneme recognition on TIMIT. 

### 2023-08-08 chapters/paper_modelagnostic.tex
Fix compilation issue and revise captions

@@
     \includegraphics[width=\textwidth]{graphics/paper_modelagnostic/pvalues_comb_no_drop_correct_font.pdf}
-    \caption{\emph{First plot}: $p$-values of the typicality test on the two test sets.
+    \caption[Distribution of the p-values of the typicality test, score statistic, Fisher's method and a combination of all.]{\emph{First plot}: $p$-values of the typicality test on the two test sets.
 We can see that under $\mathcal{H}_0$, they should be uniformly distributed.
@@
     \includegraphics[scale=0.48]{graphics/paper_modelagnostic/fashion_hvae_type1_correct_font.pdf}
-    \caption{Type I (probability of an inlier to be classified as outlier) and Type II (probability of an outlier to be considered as inlier) errors versus the significance level $\alpha$ on the combination values.
+    \caption[Type I and Type II errors versus the significance level $\alpha$ on the combination values]{Type I (probability of an inlier to be classified as outlier) and Type II (probability of an outlier to be considered as inlier) errors versus the significance level $\alpha$ on the combination values.
 By using Benjamini-Hochberg correction, we get that the Type I error stays below identity line.}

### 2023-08-09 appendices/appendix_modelagnostic.tex
Finalize model-agnostic paper and add stroke paper

@@
 A crude diagonal approximation can be computed by simply estimating the diagonal of $I({\theta})$ and setting all off-diagonal elements to zero.
-Such diagonal approximations have been used in machine learning for decades: for instance, \textcite[Section 3.12.2]{lecun1987} used a similar approximation of the Hessian matrix, and called it ``outrageously simplifying".
+Such diagonal approximations have been used in machine learning for decades: for instance, \textcite[Section 3.12.2]{lecun_modeles_1987} used a similar approximation of the Hessian matrix, and called it ``outrageously simplifying".
 Much more complex approximations have been derived, although diagonal approximations have been consistently used (e.g.
@@
 assuming that $\pi$ and $\Sigma$ are fixed at their maximum likelihood estimates.
-Similar mixture-based Fisher kernels have been very popular in the past, and were actually a key element of state-of-the art classification models on Imagenet before deep nets won the competition \citep{perronnin2010improving}.
+Similar mixture-based Fisher kernels have been very popular in the past, and were actually a key element of state-of-the art classification models on Imagenet before deep nets won the competition \citep{perronnin_improving_2010}.
 Our idea is to re-use ideas introduced by this computer vision litterature.
-Under the assumption that the Gaussian clusters are well-separated, \textcite{tanaka2013fisher}, extending an earlier analysis of \textcite[Appendix A]{sanchez2013image}, showed that
+Under the assumption that the Gaussian clusters are well-separated, \textcite{tanaka_fisher_2013}, extending an earlier analysis of \textcite[Appendix A]{sanchez_image_2013}, showed that
 \begin{equation}
@@
 \begin{table*}[tb]
-    %\renewcommand{\thetable}{4}
-    \caption{Test log-likelihood (bits/dim) achieved by the models used in the paper.}
-    \begin{subtable}{.45\textwidth}
+    \caption{Test log-likelihood (bits/dim) on MNIST, FashionMNIST, SVHN, and CIFAR10 achieved by the models used in the paper.}
+    \label{tab_modelagnostic:log_like}
+    \begin{subtable}{0.48\textwidth}
         \centering
         \scriptsize
+        \resizebox*{1.05\textwidth}{!}{
         \begin{tabular}{cc}
@@
             \bottomrule
+            \vspace{-0.2cm}\\
             ($**$) Binarized FashionMNIST
-        \end{tabular}
-    \end{subtable}
-    \begin{subtable}{.45\textwidth}
+        \end{tabular}%
+        }
+    \end{subtable}%
+    \hspace{0.02\textwidth}
+    \begin{subtable}{0.48\textwidth}
         \centering
         \scriptsize
+        \resizebox*{0.95\textwidth}{!}{
         \begin{tabular}{cc}
@@
             \bottomrule
+            \vspace{-0.2cm}\\
             ($**$) Binarized MNIST
-        \end{tabular}
-
+        \end{tabular}%
+        }
     \end{subtable}
-    \label{tab_modelagnostic:log_like}
     %\vspace*{-\baselineskip}
@@
     %\renewcommand{\thetable}{4}
-    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing individual statistics.]{AUROC$\uparrow$ for single-sample OOD detection.
-In this table we consider all the different single statistics we mentioned in the paper.
-One can notice that MMD diagonal is pretty close to the score statistic and the MMD identity is close to the gradient norm, as expected (see Section 4.1 in the paper).}
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing all single statistics with models trained on FashionMNIST (tested on MNIST) and CIFAR10 (tested on SVHN)]{AUROC$\uparrow$ for single-sample OOD detection.
+In this table we consider all the different single statistics we mentioned in the paper with models trained on FashionMNIST and CIFAR10.
+One can notice that MMD diagonal is pretty close to the score statistic and the MMD identity is close to the gradient norm, as expected (see Section 4.1 in the paper).
+Complementary results for models trained on MNIST and SVHN are in \cref{tab_modelagnostic:mmd_pair2}.}
     \resizebox{\textwidth}{!}{
@@
     %\renewcommand{\thetable}{4}
-    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing three methods of combining different statistics.]{AUROC$\uparrow$ for single-sample OOD detection.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing three methods of combining different statistics with models trained on FashionMNIST (tested on MNIST) and CIFAR10 (tested on SVHN).]{AUROC$\uparrow$ for single-sample OOD detection.
 Comparison between the three methods we mentioned to combine different statistics.
@@
     %\renewcommand{\thetable}{4}
-    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing two different Glow models.]{AUROC$\uparrow$ for single-sample OOD detection.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing two different Glow models on single and combined statistics.]{AUROC$\uparrow$ for single-sample OOD detection.
 In this table we are comparing two different Glow models trained on CIFAR10 by considering two different checkpoints with almost the same test log-likelihood.
@@
 \begin{table*}[tb]
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing all single statistics with models trained on MNIST (tested on FashionMNIST) SVHN (tested on CIFAR10)]{AUROC$\uparrow$ for single-sample OOD detection.
+In this table we consider all the different single statistics we mentioned in the paper with models trained on MNIST and SVHN.
+In this case, it is important to notice that the gradient norm and the MMD identity sometimes fail to a different extent.
+Complementary results for models trained on MNIST and SVHN are in \cref{tab_modelagnostic:mmd}}
+    \resizebox{\textwidth}{!}{
+        \scriptsize
+        \begin{tabular}{ccccccc}
+            \toprule
+            &\multicolumn{6}{c}{\textsc{MNIST (in) / FashionMNIST (out)}}\\
+            \cmidrule{2-7}
+            & \multicolumn{6}{c}{\textcolor{blue}{\textsc{single statistics}}}\\
+            \cmidrule{2-7}
+            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{MMD diagonal}}  & \textcolor{blue}{\textsc{MMD identity}} &\textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}}  \\
+            \midrule
+            \textsc{PixelCNN++} (dropout) ($\dagger$)  & 0.9999  & 0.8534 & 0.9993 & 0.8608 & 0.9996  & 0.9993 \\
+            \textsc{Glow} (RMSProp) & 0.9997  & 0.9936  & 0.9942 & 0.6609 &  0.9991 & 0.9936\\
+            \textsc{Glow} (Adam)  &0.9999 & 0.6506 & 0.9993 &  0.9124 &  0.9997 &  0.9992\\
+            \textsc{HVAE}  & 0.9999 & 0.9998 &  0.9999 &  0.9999 &  0.9999  & 0.9999\\
+            \bottomrule
+            & & & & & & \\
+            \toprule
+            &\multicolumn{6}{c}{\textsc{SVHN (in) / CIFAR10 (out)}}\\
+            \cmidrule{2-7}
+            & \multicolumn{6}{c}{\textcolor{blue}{\textsc{single statistics}}}\\
+            \cmidrule{2-7}
+            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{MMD diagonal}}  & \textcolor{blue}{\textsc{MMD identity}} &\textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}}  \\
+            \midrule
+            \textsc{PixelCNN++} (dropout)  & 0.9820 & 0.2670 &   0.9543 & 0.3185 & 0.9590 & 0.9543 \\
+            \textsc{Glow} (RMSProp)  &  0.9917  & 0.9180 & 0.9824 & 0.9317 & 0.9830  &  0.9823 \\
+            \textsc{Glow} (Adam)   &0.9913 &  0.5658 & 0.9653 & 0.7096 & 0.9779 &  0.9641 \\
+            \textsc{HVAE}  & 0.9943 &  0.1011  &  0.9865 &  0.4508 & 0.9857 &  0.9862\\
+            \bottomrule
+            \vspace{-0.2cm}\\
+            ($\dagger$) Trained using 50000 datapoints
+        \end{tabular}
+        \label{tab_modelagnostic:mmd_pair2}
+    }
+\end{table*}
+
+\begin{table*}[tb]
     %\renewcommand{\thetable}{4}
-    \caption[AUROC$\uparrow$ for single-sample OOD detection when training on MNIST and testing against FashionMNIST and when training on SVHN and testing against CIFAR10.]{AUROC$\uparrow$ for single-sample OOD detection when training on MNIST and testing against FashionMNIST and when training on SVHN and testing against CIFAR10.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection considering single and combined statistics with models trained on MNIST (tested on FashionMNIST) SVHN (tested on CIFAR10).]{AUROC$\uparrow$ for single-sample OOD detection when training on MNIST and testing against FashionMNIST and when training on SVHN and testing against CIFAR10.
 As before, Fisher's method is the combination of the typicality test and the test statistic.
-These are also combined using DoSE.}
+These are also combined using DoSE.
+Complementary results are in \cref{tab_modelagnostic:all_combinations}.}
     \resizebox{\textwidth}{!}{
@@
             \bottomrule
+            \vspace{-0.2cm}\\
             ($\dagger$) Trained using 50000 datapoints
@@
         \label{tab_modelagnostic:results_second}
-    }
-\end{table*}
-
-\begin{table*}[tb]
-    %\renewcommand{\thetable}{4}
-    \caption[AUROC$\uparrow$ for single-sample OOD detection.
-In this table we consider all the different single statistics we mentioned in the paper but for the models trained on MNIST and SVHN this time]{AUROC$\uparrow$ for single-sample OOD detection.
-In this table we consider all the different single statistics we mentioned in the paper but for the models trained on MNIST and SVHN this time.
-In this case, it is important to notice that the gradient norm and the MMD identity sometimes fail to a different extent.}
-    \resizebox{\textwidth}{!}{
-        \scriptsize
-        \begin{tabular}{ccccccc}
-            \toprule
-            &\multicolumn{6}{c}{\textsc{MNIST (in) / FashionMNIST (out)}}\\
-            \cmidrule{2-7}
-            & \multicolumn{6}{c}{\textcolor{blue}{\textsc{single statistics}}}\\
-            \cmidrule{2-7}
-            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{MMD diagonal}}  & \textcolor{blue}{\textsc{MMD identity}} &\textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}}  \\
-            \midrule
-            \textsc{PixelCNN++} (dropout) ($\dagger$)  & 0.9999  & 0.8534 & 0.9993 & 0.8608 & 0.9996  & 0.9993 \\
-            \textsc{Glow} (RMSProp) & 0.9997  & 0.9936  & 0.9942 & 0.6609 &  0.9991 & 0.9936\\
-            \textsc{Glow} (Adam)  &0.9999 & 0.6506 & 0.9993 &  0.9124 &  0.9997 &  0.9992\\
-            \textsc{HVAE}  & 0.9999 & 0.9998 &  0.9999 &  0.9999 &  0.9999  & 0.9999\\
-            \bottomrule
-            & & & & & & \\
-            \toprule
-            &\multicolumn{6}{c}{\textsc{SVHN (in) / CIFAR10 (out)}}\\
-            \cmidrule{2-7}
-            & \multicolumn{6}{c}{\textcolor{blue}{\textsc{single statistics}}}\\
-            \cmidrule{2-7}
-            \textsc{models}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{MMD diagonal}}  & \textcolor{blue}{\textsc{MMD identity}} &\textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}}  \\
-            \midrule
-            \textsc{PixelCNN++} (dropout)  & 0.9820 & 0.2670 &   0.9543 & 0.3185 & 0.9590 & 0.9543 \\
-            \textsc{Glow} (RMSProp)  &  0.9917  & 0.9180 & 0.9824 & 0.9317 & 0.9830  &  0.9823 \\
-            \textsc{Glow} (Adam)   &0.9913 &  0.5658 & 0.9653 & 0.7096 & 0.9779 &  0.9641 \\
-            \textsc{HVAE}  & 0.9943 &  0.1011  &  0.9865 &  0.4508 & 0.9857 &  0.9862\\
-            \bottomrule
-            ($\dagger$) Trained using 50000 datapoints
-        \end{tabular}
-        \label{tab_modelagnostic:mmd_pair2}
     }    
@@
 \begin{table*}[tb]
-    %\renewcommand{\thetable}{4}
-    \caption[AUROC$\uparrow$ for single-sample OOD detection using a Gaussian mixture model (GMM).]{AUROC$\uparrow$ for single-sample OOD detection using a Gaussian mixture model (GMM).
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing single and combined statistics using a Gaussian mixture model (GMM) trained on FashionMNIST (tested on MNIST) and CIFAR10 (tested on SVHN).]{AUROC$\uparrow$ for single-sample OOD detection using a Gaussian mixture model (GMM).
 For Fisher's method we mean the combination of the typicality test and the test statistic.
@@
     %\renewcommand{\thetable}{4}
-    \caption[AUROC$\uparrow$ for single-sample OOD detection using a Probabilistic PCA.]{AUROC$\uparrow$ for single-sample OOD detection using a Probabilistic PCA.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing single and combined statistics using Probabilistic PCA trained on FashionMNIST (tested on MNIST) and CIFAR10 (tested on SVHN).]{AUROC$\uparrow$ for single-sample OOD detection using a Probabilistic PCA.
 For Fisher's method we mean the combination of the typicality test and the test statistic.
@@
     % \vspace{-1.3em}
-    \caption[Mean and standard deviation of the performance in terms of AUROC of combined statistics for a HVAE.]{Mean and standard deviation of the performance in terms of AUROC of our method.
-Quantities are computed by taking the performance of 5 different trained HVAEs both trained on CIFAR10 and FashionMNIST.}
+    \caption[Mean and standard deviation of the performance in terms of AUROC of combined statistics for different HVAE models.]{Mean and standard deviation of the performance in terms of AUROC of our method.
+Quantities are computed by taking the performance of five different trained HVAEs both trained on CIFAR10 and FashionMNIST.}
         \centering
@@
     %\renewcommand{\thetable}{4}
-    \caption{AUROC$\uparrow$ for single-sample OOD detection for CIFAR10 vs CelebA considering all the three interpolations when using CelebA.}
+    \caption{AUROC$\uparrow$ for single-sample OOD detection training on CIFAR10 and testing on CelebA considering all the three interpolations for CelebA.}
     \resizebox{\textwidth}{!}{
@@
             \bottomrule
+            \vspace{-0.2cm}\\
             ($\dagger$) Bilinear interpolation \\
-            ($\nparallel$) Lanczos interpolation \\
+            \hspace{0.06cm}($\nparallel$) Lanczos interpolation \\
             ($\ddagger$) Nearest interpolation\\

### 2023-08-09 chapters/introduction.tex
Finalize model-agnostic paper and add stroke paper

@@
 
-Historically, uncertainty estimation has been a central topic in the field of statistics.
-Uncertainty has traditionally been addressed through methods like confidence intervals, hypothesis testing, and Bayesian inference.
+Uncertainty estimation has been a central topic in the field of statistics.
+It has traditionally been addressed through methods like confidence intervals, hypothesis testing, and Bayesian inference.
 Confidence intervals provide a range of plausible values for a parameter, while hypothesis testing evaluates the significance of differences between groups or variables \cite{blitzstein_introduction_2019}.
@@
 Machine learning algorithms are designed to learn patterns from data and make predictions or decisions based on those patterns.
-Since many applications of machine learning are concerned with real-world phenomena, it is important to understand the uncertainty associated with the predictions made by the model.
+Since many applications of machine learning are concerned with real-world phenomena, it is important to understand the uncertainty associated with the predictions made by such a model.
 For instance, in medical diagnosis, knowing the uncertainty in a model's prediction is crucial when the consequences of a false positive or false negative can be significant.
@@
 
-The research in this thesis covers a varied set of topics.
-One group of the produced studies are concerned with generative latent variable models and their applications to uncertainty estimation and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
+The research done over the course of the project at hand covers a varied set of topics.
+One group of the produced studies is concerned with generative latent variable models and their applications to uncertainty estimation and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
 Another group of studies are concerned with self-supervised speech representation learning and automatic speech recognition \cite{borgholt_scaling_2021,borgholt_we_2021,mohamed_selfsupervised_2022,borgholt_brief_2022}.
-A final group
+A final group deals with applications within the medical domain including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.
 
 This first part of the thesis ties together these different studies by providing a high-level overview of the research topic and the main contributions of the thesis.
-
-covering both representation learning and uncertainty estimation and their applications to speech recognition and automated medical coding.
 

### 2023-08-09 chapters/paper_automated.tex
Finalize model-agnostic paper and add stroke paper

%!TEX root = ../thesis.tex

\chapter[automated medical coding on mimic-iii and mimic-iv: a critical review and replicability study]{Automated Medical Coding on MIMIC-III and MIMIC-IV: A Critical Review and Replicability Study}
\label{chp:paper-retrospective}
\ifthenelse{\equal{\skippapers}{true}}{}{

\section*{Abstract}

}

### 2023-08-09 chapters/paper_benchmarking.tex
Finalize model-agnostic paper and add stroke paper

@@
 Among the first contributions were the VRNN \cite{chung_recurrent_2015} and SRNN \cite{fraccaro_sequential_2016} which can be seen as conditional VAEs per timestep.
-Other recent LVMs include the FH-VAE \cite{hsu_unsupervised_2017}, which leverages an additional latent variable to capture global features, and Z-forcing \cite{goyal_z-forcing_2017}, which resembles the SRNN but includes an auxiliary task in the latent space to increase its utilization.
+Other recent LVMs include the FH-VAE \cite{hsu_unsupervised_2017}, which leverages an additional latent variable to capture global features, and Z-forcing \cite{goyal_zforcing_2017}, which resembles the SRNN but includes an auxiliary task in the latent space to increase its utilization.
 The VQ-VAE \cite{oord_neural_2018} is a hybrid between an LVM and an autoregressive model which uses a quantized latent space to improve the quality of generated samples.

### 2023-08-09 chapters/paper_benchmarking_workshop.tex
Conditional skipping of compilation
Revise skipping
Finalize model-agnostic paper and add stroke paper

@@
 \chapter[benchmarking generative latent variable models for speech]{Benchmarking Generative Latent Variable Models for Speech}
-\label{chp:benchmarking}
+\label{chp:paper-benchmarking-workshop}
+\ifthenelse{\equal{\skippapers}{true}}{}{
 
@@
 (iv) LVM representations are as good or better than Mel spectrograms for phoneme recognition.
-
-\iffalse
 
@@
 Specifically, the FH-VAE \cite{hsu_unsupervised_2017} adds a global latent variable but we exclude it due to it's segmentation of the training data.
-We also exclude Z-forcing \cite{goyal_z-forcing_2017} which resembles the SRNN, but deviates from maximum likelihood training by using an auxiliary loss.
+We also exclude Z-forcing \cite{goyal_zforcing_2017} which resembles the SRNN, but deviates from maximum likelihood training by using an auxiliary loss.
 We exclude the VQ-VAE \cite{oord_neural_2018} which is a hybrid LVM and autoregressive model, since it uses a quantized latent space.
@@
 
-\fi
+}

### 2023-08-09 chapters/paper_retrospective.tex
Finalize model-agnostic paper and add stroke paper

%!TEX root = ../thesis.tex

\chapter[a retrospective study on machine learning-assisted stroke recognition for medical helpline calls]{A Retrospective Study on Machine Learning-Assisted Stroke Recognition for Medical Helpline Calls}
\label{chp:paper-retrospective}
% \ifthenelse{\equal{\skippapers}{true}}{}{

% cite21 = borgholt_endtoend_2020
% cite23 = hochreiter_long_1997
% cite24 = graves_connectionist_2006
% cite25 = hansen_neural_1990
% cite26 = rosenblatt_perceptron_1958
% cite27 = dwass_modified_1957
% cite28 = eden_validity_1933

\section*{Abstract}
\paragraph{Background:} Effective stroke treatment depends on recognition by call-takers at prehospital telehealth services. This study aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.
\paragraph{Methods:} We used calls from 1 January 2015 to 31 December 2020 in Copenhagen to develop a machine learning-based classification pipeline. Calls from 2021 were used for testing. Calls were transcribed using an automatic speech recognition model and then categorized as stroke or non-stroke using a text classification model.
\paragraph{Findings:} Call-takers achieved a sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%). The machine learning framework performed significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
\paragraph{Interpretation:} A machine learning framework for recognizing stroke in prehospital medical helpline calls may become a supportive tool for call-takers, aiding in early and accurate stroke recognition.

\section{Introduction}
Stroke is a leading cause of disability and death worldwide \cite{cite1,cite2,cite3}. Effective treatment is time- sensitive, and an optimal outcome is more likely when treatment is administered within the first four and a half hours from stroke onset \cite{cite4,cite5}. The gateway to ambulance transport and hospital admittance is through prehospital telehealth services, including emergency medical call centres, nurse advice call lines, and out-of-hours health services. In the pre-hospital setting, the use of Mobile Stroke Units has made it possible to deliver advanced treatment faster \cite{cite6,cite7}. As the Mobile Stroke Unit is only dispatched to patients with a suspected stroke, the impact of Mobile Stroke Units is directly influenced by accurate call-taker recognition of stroke.6,7 Call-takers who can rapidly and accurately recognize stroke are therefore crucial in facilitating prompt care in both pre-hospital and in-hospital settings.

Despite initiatives to improve stroke recognition \cite{cite8,cite9}, approximately half of all patients with stroke do not receive the correct triage for their condition from call-takers \cite{cite10,cite11,cite12}. Most initiatives aim to improve stroke recognition by call-takers via introducing more specific assessment tools \cite{cite8,cite9} or providing specialized training \cite{cite13}. Recent advances in machine learning technology might be applied to improve stroke recognition without requiring changes to the triaging approach, and machine learning aided identification of stroke has been suggested as a means of improving Mobile Stroke Unit effectiveness \cite{cite7}. Real-time feedback from a machine learning model can improve the recognition of out-of-hospital cardiac arrest \cite{cite14,cite15}. Therefore, this study aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.

\section{Results}

\subsection{Main results}

The classification model outperformed the call-takers (\cref{tab_retrospective:table1}), with significant differences in all metrics (p < 0.0001). Excluding the 1-1-2 call line training data significantly degraded the model's performance (p < 0.0001), despite the domain mismatch with the MH-1813 call line test data. The performance on the 2021 calls without a diagnostic category was significantly worse than that of the test set regarding F1 score, sensitivity, FPR, and FOR (p < 0.0001).

The ROC curve (\cref{fig_retrospective:figure1}, left) illustrates the potential to increase the sensitivity while maintaining a FPR lower than or equal to that of the call-takers. Similarly, the PPV sensitivity curve (\cref{fig_retrospective:figure1}, right) demonstrates that sensitivity can be improved while retaining a PPV higher than that of the call-takers. The framework can thus be tuned to a sensitivity of around 73\%, while still having a higher positive predictive value than the human call-taker (\cref{fig_retrospective:figure1}, right) The ensemble model outperformed the individual models regardless of the threshold, except for one that exhibited a slightly better sensitivity at a high FPR exceeding 1.5\%. The confusion matrices (\cref{fig_retrospective:figure2}) illustrate the performance differences in absolute numbers, with the model exhibiting more true positives and fewer false positives than the call-takers.

% \todo[inline]{Insert Table 1}
\begin{table}
    \centering
    \caption[Performance metrics for model stroke recognition.]{Performance metrics. Performance metrics for the model and call-takers on the test set. The model was trained on the training set, which included calls from 2015 to 2020. The call-takers were trained on the same data. The 2021 data were used for testing. The 2021 data without diagnostic categories were excluded}
    \label{tab_retrospective:table1}
    \resizebox*{0.98\textwidth}{!}{%
    \begin{tabular}{l|ccccc}
        \toprule
                              & Training (112) & Training (MH-1813) & Validation & Test & 2021 w/o category \\

        \midrule
        \multicolumn{6}{c}{All calls} \\
        \midrule
        Num. calls            & 155,696 & 1,391,301 & 155,825 & 344,030 & 231,009 \\
        Female                & 74,640 (47.94\%) & 792,783 (56.98\%) & 86,959 (55.81\%) & 190,974 (55.51\%) & 134,324 (58.14\%) \\
        Male                  & 79,564 (51.10\%) & 596,760 (42.89\%) & 68,866 (44.19\%) & 153,050 (44.49\%) & 96,258 (41.67\%) \\
        65+ years             & 72,930 (46.84\%) & 335,146 (24.09\%) & 30,313 (19.45\%) & 65,652 (19.08\%) & 81,488 (35.27\%) \\
        Age (mean $\pm$ std.) & 59.47 ± 21.24 & 47.12 ± 21.38 & 44.63 ± 20.08 & 44.31 ± 20.10 & 50.36 ± 22.77 \\

        \midrule
        \multicolumn{6}{c}{Stroke calls} \\
        \midrule
        Num. calls            & 3,899 & 3,471 & 360 & 757 & 679 \\
        Female                & 1,784 (45.76\%) & 1,654 (47.65\%) & 161 (44.72\%) & 349 (46.10\%) & 366 (53.90\%) \\
        Male                  & 2,115 (54.24\%) & 1,815 (52.29\%) & 199 (55.28\%) & 408 (53.90\%) & 313 (46.10\%) \\
        65+ years             & 2,968 (76.12\%) & 2,421 (69.75\%) & 250 (69.44\%) & 555 (73.32\%) & 567 (83.51\%) \\
        Age (mean $\pm$ std.) & 72.91 ± 12.77 & 70.68 ± 13.85 & 70.93 ± 13.83 & 71.51 ± 13.41 & 73.41 ± 14.11 \\

        \midrule
        \multicolumn{6}{c}{Non-stroke calls} \\
        \midrule
        Num. calls            & 151,797 & 1,387,830 & 155,465 & 343,273 & 230,330 \\
        Female                & 72,856 (48.00\%) & 791,129 (57.00\%) & 86,798 (55.83\%) & 190,625 (55.53\%) & 133,958 (58.16\%) \\
        Male                  & 77,449 (51.02\%) & 594,945 (42.87\%) & 68,667 (44.17\%) & 152,642 (44.47\%) & 95,945 (41.66\%) \\
        65+ years             & 69,962 (46.09\%) & 332,725 (23.97\%) & 30,063 (19.34\%) & 65,097 (18.96\%) & 80,921 (35.13\%) \\
        Age (mean $\pm$ std.) & 59.12 ± 21.30 & 47.06 ± 21.36 & 44.57 ± 20.05 & 44.25 ± 20.08 & 50.29 ± 22.76 \\

        \bottomrule
    \end{tabular}%
    }
\end{table}

\subsection{Sex and age}

The model and call-takers exhibited significantly higher PPV and F1-score in men than in women (p < 0.0001) (\cref{tab_retrospective:table1}). The model significantly outperformed the call-takers on all metrics for each sex (p < 0.0001).
The model performed significantly better in the 65+ group than in the 18-64 year group regarding sensitivity, PPV, and F1-score (p < 0.0001). Similarly, the call-takers performed significantly better in the 65+ group than in the 18-64 group regarding PPV and F1-score (p < 0.0001). Finally, the model significantly outperformed the call-takers on all metrics in both age groups (p < 0.0001).

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{paper_retrospective/figure1.pdf}
    \caption[Receiver operator characteristic (ROC) curve and PPV-sensitivity curve for model stroke recognition.]{Receiver operator characteristic (ROC) curve and PPV-sensitivity curve. Left, the ROC curve and, right, PPV-sensitivity curve (precision-recall curve). Models 1-5 are the individual models that make up the ensemble model.}
    \label{fig_retrospective:figure1}
\end{figure}
% \todo[inline]{Insert Figure 1}

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{paper_retrospective/figure2.pdf}
    \caption[Prediction confusion matrices for model stroke recognition.]{Prediction confusion matrices. Confusion matrices of predictions for call takers and the model on the test set. Numbers for the model are given as the rounded mean over eleven runs.}
    \label{fig_retrospective:figure2}
\end{figure}
% \todo[inline]{Insert Figure 2}

\subsection{Model explainability}

Among the words with a positive rank score, several words are synonymous with stroke, such as ``blood clot", ``haemorrhagic stroke", and ``stroke". Ambulances are rarely dispatched because the MH-1813 is not intended for emergencies. Therefore, a word like ``ambulance" may also be a strong indicator of call-taker recognition, which the model has learned to mimic. Additionally, most of the remaining words can be linked to stroke-related symptoms such as ``double vision", ``difficulties speaking", and ``hangs". Particularly, words describing the side of the body where symptoms occur ranked high (such as ``left", ``right", and ``side"). Finally, some words were also related to the sudden onset of symptoms (including ``suddenly" and ``minutes").
% \todo[inline]{Insert Table 2}
\begin{table}
    \centering
    \caption[Overall stroke recognition performance of model compared to call takers.]{Overall performance on MH-1813 test data, performance without 1-1-2 training data, and performance on data from 2021 without diagnostic categories as well as performance on MH-1813 based on demographic subgroups (age/sex) [mean (95\% CI)].}
    \label{tab_retrospective:table2}
    \resizebox*{0.98\textwidth}{!}{%
    \begin{tabular}{l|ccccc}
        \toprule
                    & F1-score [\%] $\uparrow$ & Sensitivity [\%] $\uparrow$ & PPV [\%] $\uparrow$ & \makecell{FOR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell{FPR [\%] $\downarrow$ \\ (1 - NPV)} \\

        \midrule
        \multicolumn{6}{c}{Overall} \\
        \midrule
        Call takers & 25.8 (23.7-27.9) & 52.7 (49.2-56.4) & 17.1 (15.5-18.6) & 0.105 (0.094-0.116) & 0.565 (0.539-0.590) \\
        Model       & 35.7 (35.0-36.4) & 63.0 (62.0-64.1) & 24.9 (24.3-25.5) & 0.082 (0.079-0.085) & 0.419 (0.413-0.426) \\

        \midrule
        \multicolumn{6}{c}{Without 112 training data} \\
        \midrule
        Model       & 32.4 (31.8-33.1) & 60.4 (59.3-61.4) & 22.2 (21.6-22.7) & 0.088 (0.085-0.091) & 0.467 (0.460-0.474) \\

        \midrule
        \multicolumn{6}{c}{On MH-1813 data without diagnostic category} \\
        \midrule
        Model       & 32.6 (31.9-33.4) & 48.3 (47.2-49.4) & 24.7 (23.9-25.3) & 0.153 (0.148-0.158) & 0.435 (0.427-0.443) \\

        \midrule
        \multicolumn{6}{c}{18-64 years} \\
        \midrule
        Call takers & 15.9 (13.1-18.5) & 50.5 (43.6-57.2) & 9.40 (7.61-11.18) & 0.036 (0.028-0.043) & 0.353 (0.331-0.375) \\
        Model       & 22.9 (21.8-24.0) & 54.1 (52.1-56.3) & 14.5 (13.8-15.3) & 0.033 (0.031-0.035) & 0.231 (0.226-0.236) \\

        \midrule
        \multicolumn{6}{c}{65+ years} \\
        \midrule
        Call takers & 32.9 (30.1-35.7) & 53.5 (49.4-57.6) & 23.7 (21.4-26.0) & 0.401 (0.352-0.449) & 1.467 (1.373-1.560) \\
        Model       & 42.8 (41.9-43.7) & 66.3 (65.1-67.5) & 31.6 (30.8-32.4) & 0.290 (0.278-0.303) & 1.224 (1.198-1.249) \\

        \midrule
        \multicolumn{6}{c}{Male} \\
        \midrule
        Call takers & 30.2 (27.2-33.3) & 53.9 (49.1-58.9) & 21.0 (18.5-23.5) & 0.124 (0.105-0.141) & 0.542 (0.506-0.580) \\
        Model       & 39.0 (38.0-40.1) & 63.7 (62.3-65.2) & 28.1 (27.3-29.0) & 0.097 (0.093-0.102) & 0.435 (0.425-0.445) \\

        \midrule
        \multicolumn{6}{c}{Female} \\
        \midrule
        Call takers & 21.9 (19.1-24.6) & 51.3 (46.0-56.6) & 13.9 (12.0-15.8) & 0.090 (0.076-0.103) & 0.582 (0.547-0.616) \\
        Model       & 32.4 (31.4-33.4) & 62.3 (60.7-63.8) & 21.9 (21.1-22.7) & 0.069 (0.066-0.073) & 0.407 (0.399-0.416) \\

        \bottomrule
    \end{tabular}%
    }
\end{table}

Among the words with a negative rank score, most were strong indicators for specific conditions, symptoms, or body parts that are unrelated to stroke (such as ``tetanus", ``pregnant", ``swollen", ``fever", and ``the knee"). Another group of words used to describe aspects of treatment that are unlikely to be addressed in a stroke call included ``prescription", ``bandage", and ``OTC". Finally, a small group of words described institutions that are not commonly involved in stroke treatment (such as ``psychiatric", ``the emergency room", and ``the police").

\section{Discussion}

Our results showed that a machine learning framework can substantially improve stroke recognition in medical helpline calls compared to solely relying on human call-takers. This improvement was observed across all performance metrics and for basic patient demographics (age and sex). Our occlusion analysis revealed that the model relied on the relevant predictive features associated with call-taker triaging, patient symptoms, and treatment.

This study does not imply that a machine learning model can replace medical call-takers. The effectiveness of the model is fully reliant on the conversation between the call-taker and caller and the call-taker's ability to skillfully triage the patient. Instead, the model should be used as a supportive tool for call-takers in the decision-making process, contributing to a higher recognition of patients with stroke and potentially boosting the confidence of call- takers in their decisions. A similar machine learning model designed to predict cardiac arrest was tested in a randomised controlled trial (RCT) at CEMS.15 The results highlighted the necessity of incorporating input from call-takers. The machine learning model for cardiac arrest has subsequently been implemented in daily practice at CEMS, in a setup similar to the one presented in our study. However, implementation of our framework requires further investigation. The relative performance gap between call-takers and the model was larger in our study than in the cardiac arrest study,15 which may affect the results of a potential RCT. To support future work and discussions beyond the scope of this study, the supplementary material includes the results of a simulation of a live implementation where call-takers are assumed to follow a set of fixed rules based on the output of the machine learning framework.

The performance gap between the model and call-takers could be explained by the rarity of stroke calls to MH-1813 (0.250\% of all calls in 2021), which might affect call-taker awareness of stroke as a possible cause of certain symptoms. Additionally, certain stroke symptoms are so rare that some call-takers may never encounter them, increasing the risk of false negatives. The model was trained on more calls than any single call-taker would handle in a lifetime, enabling it to recognise even rare descriptors of stroke. The model is specifically trained to recognise strokes and exclusively learns from actual stroke descriptions, unlike call-takers, who are trained with generalised teaching materials to triage many different conditions. Therefore, call-takers may not have received specific training for patients with stroke and may never have encountered them.

The model performed significantly better on men than on women. This could be attributed to several factors. First, the model may have learned to mimic call-takers with the same bias. Second, women may experience different and challenging-to-identify symptoms than men \cite{cite29, cite30}. Third, a higher prevalence of male patients with stroke was observed in the training data. Despite these potential sources of bias, the model exhibited less bias than call-takers did. That is, the relative performance improvements were higher for women than for men. This bias could be further reduced using advanced data augmentation and balanced data when training a machine learning model. However, such measures may degrade overall performance.

This study has some limitations. First, the mapping of call recordings to electronic records was incomplete due to technical limitations in the CAD registry, which limited the number of calls available to us. Of note, there was no obvious pattern of bias related to the unmapped calls, and we included all calls with matching audio files, regardless of dispatcher performance. The results could potentially be improved if more calls were available for analysis. Second, calls without a call-taker indicated diagnostic category were not included in the validation and test data because the call-taker's performance could not be evaluated. Moreover, in exploratory analyses, the model performed worse on these calls, which might be attributed to differences in population characteristics (\cref{tab_retrospective:table3}). Finally, the ground truth stroke labelling relied on the patient-reported time of onset being exact; however, estimating the accuracy of the timestamps in DanStroke was impossible..

The proposed machine learning framework can be improved. In preliminary experiments, we tested convolutional, recurrent, and self-attention architectures for text classification; however, this did not improve the model. One promising path is to leverage the abundance of non-stroke calls. Self-supervised learning techniques for speech and text data can be used to improve speech recognition and text classification models. This approach can enable direct classification from audio input without speech recognition. Finally, incorporating additional data from CAD records through multitask learning could be used to inform the classifier and predict the diagnostic category to improve its ability to identify more fine-grained characteristics of non-stroke calls.

In conclusion, using the largest collection of audio calls from patients with stroke to date, we developed a machine learning framework that significantly outperformed human call-takers in stroke recognition in medical helpline calls. The frameworkcan assist human call-takers during medical helpline calls. Ideally, this would enable a higher recognition of patients with stroke in the prehospital setting, benefiting both patient outcomes and health service resource allocation.

\subsection{Contributors}

JW, JDH and LB share lead authorship and contributed equally to the original drafting of the paper, review and editing, data curation, formal analysis, methodology, validation, and conceptualisation. JDH and LB contributed to software development and funding acquisition. SNFB contributed to the conceptualisation, data curation, funding acquisition, supervision, and writing - review and editing. LM contributed to conceptualisation, funding acquisition, project administration, supervision, and writing - review and editing. MS contributed to supervision and writing - review and editing. HC contributed to the methodology, supervision, and writing - review and editing. CK contributed to the conceptualisation, funding acquisition, methodology, project administration, supervision, and writing - review and editing. HCC contributed to the conceptualisation, data curation, funding acquisition, methodology, project administration, supervision, and writing - review and editing.

HCC and CK share last authorship.

The three lead authors directly accessed and verified the underlying data reported in this manuscript. JW's affiliations are exclusively academic.

\section{Methods}

\subsection{Study design}

We used call recordings and registry data from the Copenhagen Emergency Medical Services (CEMS) and the Danish Stroke Registry \cite{cite16} from 2015 to 2020. These data were used to train a machine learning model to classify medical helpline calls as stroke or non-stroke. We compared the performances of the model with that of the call-taker using data from 2021.

\subsection{Data sources}

\subsubsection{CEMS}

The CEMS is responsible for providing prehospital telehealth services in the Capital Region of Denmark, with a catchment area of 1.9 million \cite{cite17}. CEMS operates two call lines: the 1-1-2 emergency line, similar to 9-1-1 in the United States, intended for acute conditions. The other is the medical helpline 1813 (MH-1813, pronounced ``'eighteen-thirteen") intended for non- life-threatening conditions that cannot wait until a general practitioner is available.18

Call-takers for both lines, who are nurses, paramedics, or physicians, can dispatch ambulances. The condition suspected by the call-taker is categorised based on a predefined diagnostic index and stored in an electronic record using a computer-aided dispatch (CAD) system. The CAD records are associated with the Danish civil registration number (CPR number) \cite{cite19} of the patient. The CPR number is a unique identification assigned to all Danish residents. It is used for interactions with health services and registries, enabling cross- referencing of the data sources used in this study. The call audio is recorded and stored separately from the CAD recordings using a telephone system.

\subsubsection{Danish Stroke Registry (DanStroke)}

All patients with stroke admitted to a Danish hospital within 5 days of symptom onset are recorded in the Danish Stroke Registry \cite{cite16}, also known as DanStroke. This record includes the patient-reported time of onset, stroke type (haemorrhagic, ischaemic, or transitory ischaemic attack), and CPR number of the patient.

\subsubsection{Data ethics}

The Danish Data Protection Agency (P-2021-475) approved this study. Danish law did not require approval from the Scientific Ethics Committee because the data were registry-based. CEMS approved the transcription of all calls made to 1-1-2 and MH-1813. All electronic records were anonymised before analysis, and the researchers did not inspect the calls manually.

\subsection{Study scope}

Stroke prevalence in calls made to the MH-1813 is lower than that in calls made to 1-1-2. Patients with stroke may exhibit different symptoms and symptom severity because MH- 1813 is meant for low-acuity incidents, leading to reduced recognition. In addition, MH-1813 call-takers dispatch high-priority transport less frequently, which may affect optimal treatment timing. Therefore, we focused on MH-1813 in this study.

\subsection{Stroke dataset}

\subsubsection{Cross-referencing data sources}

From the CAD medical records, we included all calls that could be matched to a corresponding audio file. for 1-1-2 and MH-1813 from 2015 to 2021 for patients older than 18. The CAD records were matched with the telephone call recordings based on the call start, call duration, and call-taker identity. Due to data incompleteness, and the way the audio data is stored, at CEMS, 2,730,199 contacts could not be matched to their corresponding audiofile, however, 2,361,178 contacts were successfully matched. We found no obvious pattern in the matched and unmatched calls and we included all calls with at matching audio file. Next, a call was regarded as a case of ground truth stroke positive when the CPR number in the CAD record matched that of a DanStroke record, and the patient-reported time of onset was close to the call start time. We allowed a window of 72 hours before and 24 hours after the call starts to account for uncertainty in recording stroke onset time. We excluded calls involving subarachnoid haemorrhage cases. Finally, we considered a call to be a call-taker stroke positive when the call-taker selected the stroke diagnostic category during the call and dispatched an ambulance with the appropriate level of response \cite{cite20}. To ensure that the effect of the machine learning framework was not overestimated, we excluded calls where diagnostic category had not been registered from the final testing. We still reported the population characteristics and model performance of this group of calls to assess potential bias introduced by excluding them. A data-flow diagram is included in the supplementary material. The resulting dataset is the largest dataset of audio files from stroke calls collected to date.

\subsubsection{Dataset splitting}

We reserved all the MH-1813 calls from 2021 for testing. We used stratified sampling to split the MH-1813 calls from 2015 to 2020 into six folds: one for validation and five for training. The calls were stratified based on the ground truth stroke label and the presence of a diagnostic category. Only calls without diagnostic categories were included in the training set. The 1-1-2 calls were used only for training; however, calls from 2021 were discarded to avoid temporal overlap with the test period.

As expected, the 1-1-2 training data differed from the MH-1813 data regarding age, male/female ratio, and stroke prevalence (\cref{tab_retrospective:table3}). We performed an ablation study where 1- 1-2 data were not used for training to assess whether this difference negatively impacted model performance. The training, validation, and test subsets of the MH-1813 data had similar characteristics, whereas the 2021 data without diagnostic categories differed in age and sex.

\subsection{Machine learning pipeline}

We employed a two-step machine learning pipeline. First, a call was transcribed using the speech recognition model. Second, the transcript was used as input for the text classification model. The final output score was used to classify whether the call concerned a stroke. The pipeline is illustrated in the Supplementary Material.

\todo[inline]{Insert Table 3}

\subsubsection{Speech recognition}

The call recordings from the CEMS were stored as 8-bit linear pulse-code modulated audio, sampled at 8 kHz. A call was converted into a log-Mel spectrogram before being input into the speech recognition model. This conversion is a commonly used input representation for speech-processing tasks, which facilitates the identification of linguistic content in audio signals. We used a speech recognition model with a neural network architecture \cite{borgholt_endtoend_2020}, consisting of two-dimensional convolutional layers \cite{cite22} and blocks of bidirectional long short-term memory layers \cite{hochreiter_long_1997} The output is a sequence of probability distributions over characters of the Danish alphabet, which were then converted into a human-readable transcript using a greedy decoder \cite{graves_connectionist_2006}.

\subsubsection{Text classification}

As input for the classification model, each transcript was transformed into a fixed-size bag- of-words vector, which encoded the occurrence of word and character ($n$-grams) in a fixed vocabulary. The feature selection procedure is detailed in the Supplementary Material. The model was constructed as an ensemble \cite{hansen_neural_1990} of five identical, independently trained models. Each consists of a stack of neural network layers commonly referred to as a multi-layer perceptron \cite{rosenblatt_perceptron_1958}. The final layer has a single scalar output and applies a sigmoid nonlinearity to produce an output score between zero and one.

\subsubsection{Threshold calibration and ensembling}

We selected the prediction threshold as the harmonic mean of the thresholds for each model in the ensemble to ensure that the sensitivity and positive predictive value (PPV) were equal to those of the call-takers. This simplifies the comparison by ensuring a trade-off between sensitivity and PPV, similar to that of call-takers.

As the threshold differed for each model in the ensemble, computing the ensemble output score as the average output score of the individual models would not be meaningful. Instead, we first subtracted the threshold from the output score in the logit space (before sigmoid nonlinearity) for each model to obtain the same threshold (0.5). Subsequently, we defined the ensemble output score as the average of the centred output scores. The exact equations are provided in the Supplementary Material.

\subsubsection{Model training}

The speech recognition model was trained on 3,811 manually transcribed random calls (173 h) from the CEMS as part of a previous project.14 These calls exclusively originated from 1- 1-2 between 2015 and 2018, ensuring no overlap with the test data used for the text classification model. The model was trained using a connectionist temporal classification objective \cite{graves_connectionist_2006}.

We trained five models for the text classification ensemble using binary cross-entropy after transcribing all calls in the dataset using the speech recognition model. One training fold was used for early stopping, whereas the remaining four folds and 1-1-2 data were used for training. Thus, each model in the ensemble was trained and validated using different datasets. We ran a grid search with 96 different hyperparameter configurations and selected the ensemble model with the best F1 score for the validation set.

\subsection{Model explainability}

We performed an \emph{occlusion analysis} to better understand the predictions of the text classification model. This involved removing all instances of a given word from the input transcript to evaluate its impact on the model output. The word was removed before vectorisation, such that all word and character $n$-grams associated with the word were discarded. Specifically, let $z^{(n, d, w)}$ be the logit output of model $n$ in the ensemble for transcript $d$ when the word $w$ is occluded. For transcript $d$, we computed the word impact score $i^{(d, w)}$ as the mean difference between the logit before and after occlusion.
%
\begin{equation}
    i^{(d,w)} = \frac{1}{N_d} \sum_{n=1}^{N_d} \left( z^{(n, d)} - z^{(n, d, w)} \right) \enspace .
\end{equation}
%
We used the logit output to compute the impact score because the difference in sigmoid-normalised output is biased towards zero for values close to 0 or 1. To select words for inspection, we computed a ranking score, $r^{(w)}$, as the sum of the signed squares of the impact:
%
\begin{equation}
    r^{(w)} = \sum_{d=1}^{N} \text{sign}\left( i^{(d, w)} \right) \left( i^{(d,w)}\right) ^2 \enspace .
\end{equation}
%
where $\text{sign}$ represents the sign function. Squaring $i^{(d,w)}$ favours rare features with a high impact over common features with a low impact.

\subsection{Statistical analysis}

We report the F1-score, sensitivity, PPV, false omission rate (FOR, equal to 1- negative predictive value), and false positive rate (FPR, equal to 1 - specificity). Due to the imbalanced nature of the dataset, the negative predictive value and specificity were greater than 99\% for all cases. We reported FOR and FPR instead because such large numerical values exhibit low relative variance, thereby obfuscating comparisons. Finally, we report the prediction confusion matrices, receiver operator characteristic (ROC) curve, and PPV- sensitivity curve, commonly known as the precision-recall curve. All results are reported with up to three significant digits.

We present the results with and without 1-1-2 training data, subgroup analyses based on age (18-64/65+) and sex (male/female), and call-takers performance. We also report the model performance on calls without a diagnostic category from the test year 2021 to assess potential data bias. We tested our results for statistical significance using approximate permutation tests, paired or independent, as appropriate and computed 95\% confidence intervals (CIs) using bootstrapping \cite{dwass_modified_1957,eden_validity_1933}. In our assessment, we accounted for random variation associated with model training by basing the means, tests, and CIs on the predictions of 11 randomly initialised training runs. Statistical significance was defined as a two-sided p-value of < 0.05.

We used the model with the median F1-score out of the 11 runs for the occlusion analysis. We listed the 30 words with the highest positive ranking scores for calls classified as stroke and the 30 words with the highest negative ranking scores for calls classified as non-stroke.

\subsection{Role of the funding source}

The grant providers had no role in the study design, data collection, analysis, interpretation, manuscript writing, or publication decision. Corti provided additional funds and technical expertise to develop the models. Corti was not financially compensated for this, and the project was part of its research initiatives, which were conducted in cooperation with several universities and the Innovation Fund Denmark. Corti owns the rights to the models and source code.

\section{Data availability}

The datasets used to evaluate call-taker performance and to train and evaluate the machine learning framework are legally restricted by Danish patient privacy and secrecy laws and are therefore, not publicly available. The data can be made available from the publication date but requires a Data Access Agreement, which is examined and approved by the ethics committees that approved this research. For the same reason, the machine learning framework trained in this study is not publicly available; however, instructions on how to train it are included in the main manuscript and supplementary material. The source code can be shared using a Creative Commons NC-ND 4.0, an international licence, upon reasonable written request to the corresponding author, and requires a data use agreement.

\section{Declaration of interests}

JDH and LB received funding from the Innovation Fund Denmark. JDH and LB used Corti and hold stock warrants. LM is a co-founder, stockholder, and the Chief Technology Officer of Corti. JW received funding from Trygfonden. SNFB has no conflicts of interest to declare. HC has received funding from the Velux Foundation, Tværsfonden, Helsefonden, Hartmann Fonden, Lundbeck Foundation, and Novo Nordisk Foundation; royalties from Gyldendal; honoraria from Bayer and Bristol Meyers Squibb, and is chair of Action Plan for stroke in Europe Implementation, Co-chair of the Scientific Stroke Panel EAN and Senior Guest Editor of AHA Stroke. MS has no conflicts of interest to declare. HCC has no conflicts of interest to declare. CK received funding from the Novo Nordisk Foundation and is the chair of the Danish Resuscitation Council and vice chair of the Danish Stroke Society. Both positions are unpaid.

\section*{Acknowledgements}

We thank the staff of the CEMS for their role in generating the data used in this study. We thank Emilie Grunddal Pedersen, Mette Bjerg Lindhøj, and Jens Morten Haugård for their help and cooperation in accessing the data sources. We also thank the Centre for IT and Medical Technology (CIMT) and Corti employees Akihiro Inui and Nathaniel Joselson for their assistance in setting up and using the cloud-computing environment for training and evaluating the machine learning framework of this study.

% }

### 2023-08-09 chapters/paper_review/04_benchmarking.tex
Finalize model-agnostic paper and add stroke paper

@@
   \centering
-  \scriptsize
+  \tiny
   \caption[Summary of common experiment settings for various SSL
@@
   \label{table:method_data_setting}
-  \resizebox*{0.99\textwidth}{!}{%
+  \resizebox*{0.97\textwidth}{!}{%
   {%\tabulinesep=1mm
@@
     % MockingJay
-    \multirow{3}{*}{MockingJay \cite{liu_mockingjay_2020}} & \multirow{3}{*}{\makecell[l]{LS 360 hrs}} & PC & LS 360 hrs & LS test-clean & - & 0.36\textsuperscript{\ref{footnote:low-resource}}/1.8\textsuperscript{\ref{footnote:low-resource}}/3.6\textsuperscript{\ref{footnote:low-resource}}/18\textsuperscript{\ref{footnote:low-resource}}/45\textsuperscript{\ref{footnote:low-resource}}/360  hrs \\ \cline{3-7}
+    \multirow{3}{*}{MockingJay \cite{liu_mockingjay_2020}} & \multirow{3}{*}{\makecell[l]{LS 360 hrs}} & PC & LS 360 hrs & LS test-clean & - & 0.36\textsuperscript{\ref{footnote:low-resource}} /1.8\textsuperscript{\ref{footnote:low-resource}} /3.6\textsuperscript{\ref{footnote:low-resource}} /18\textsuperscript{\ref{footnote:low-resource}} /45\textsuperscript{\ref{footnote:low-resource}} /360  hrs \\ \cline{3-7}
     & & SID & LS 100 hrs\textsuperscript{\ref{footnote:91-split-test}} & LS 100 hrs\textsuperscript{\ref{footnote:91-split-test}} & - & 90\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
@@
     % DeCoAR
-    \multirow{2}{*}[0mm]{DeCoAR \cite{ling_deep_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{LS 100/360/460/\\\:\:\:960 hrs,\\ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{25\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}}/81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{DeCoAR \cite{ling_deep_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{LS 100/360/\\\:\:\:460/960 hrs,\\ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{25\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}}/81 hrs} \\ \cline{3-7}
     & & ASR & \makecell[l]{LS 100/360/\\\:\:\:460/960 hrs} & \makecell[l]{LS test-clean,\\ LS test-other} & - & 100/360/460/960 hrs \\ \hline
@@
   \label{table:method_data_setting_2}
-  \resizebox*{0.99\textwidth}{!}{%
+  \resizebox*{0.97\textwidth}{!}{%
   {%\tabulinesep=1mm
@@
     % vq-wav2vec + Discrete BERT
-    \makecell[l]{vq-wav2vec + \\\:DiscreteBERT \cite{baevski_effectiveness_2020}} & LS 960 hrs & ASR & LS 100 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline
+    \makecell[l]{vq-wav2vec + \\DiscreteBERT \cite{baevski_effectiveness_2020}} & LS 960 hrs & ASR & LS 100 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline
     % speech-XLNet
@@
     % HuBERT
-    HuBERT \cite{hsu_hubert_2021} & \makecell[l]{LS 960 hrs,\\ LL 60k hrs} & ASR & LS 960 hrs & \makecell[l]{LS test-clean\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \hline
+    HuBERT \cite{hsu_hubert_2021} & \makecell[l]{LS 960 hrs,\\ LL 60k hrs} & ASR & LS 960 hrs & \makecell[l]{LS test-clean\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}}, 1\textsuperscript{\ref{footnote:low-resource}}\\/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \hline
     % wav2vec-c
     wav2vec-c \cite{sadhu_wav2vecc_2021} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
-    UniSpeech-SAT \cite{chen_unispeechsat_2021} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See SUPERB \cite{yang_superb_2021} paper for details \\ \hline
-    WavLM \cite{chen_wavlm_2021} & \makecell[l]{LL 60k hrs\\\;\;\;+ GigaSpeech-10k\\\;\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See \cite{yang_superb_2021} for details \\ \hline
-    \multirow{4}{*}[0mm]{XLS-R \cite{babu_xlsr_2021}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;\;+ CV-dataset-7k + VL\\\;\;\;+ BBL}} & ASR & VP, MLS, CV-dataset, BBL, LS & VP, MLS, CV-dataset, BBL, LS & - & \multirow{4}{*}[0mm]{See \cite{babu_xlsr_2021} for details} \\ \cline{3-6}
+    UniSpeech-SAT \cite{chen_unispeechsat_2021} & \makecell[l]{LL 60k hrs\\\;\;+ GigaSpeech-10k\\\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See SUPERB \cite{yang_superb_2021} paper for details \\ \hline
+    WavLM \cite{chen_wavlm_2021} & \makecell[l]{LL 60k hrs\\\;\;+ GigaSpeech-10k\\\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See \cite{yang_superb_2021} for details \\ \hline
+    \multirow{4}{*}[0mm]{XLS-R \cite{babu_xlsr_2021}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;+ CV-dataset-7k\\\;\;+ VL\\\;\;+ BBL}} & ASR & VP, MLS, CV-dataset, BBL, LS & VP, MLS, CV-dataset, BBL, LS & - & \multirow{4}{*}[0mm]{See \cite{babu_xlsr_2021} for details} \\ \cline{3-6}
     & & SID & VoxCeleb1 & VoxCeleb1 & \checkmark & \\ \cline{3-6}
@@
 
-\begin{table}[ht]
+\begin{table}[t!]
   \centering

### 2023-08-10 appendices/appendix_retrospective.tex
Add more of stroke paper

\chapter[supplementary material for a retrospective study on machine learning-assisted stroke recognition for medical helpline calls]{Supplementary Material for A Retrospective Study on Machine Learning-Assisted Stroke Recognition for Medical Helpline Calls}
\label{app:retrospective}
\ifthenelse{\equal{\skipappendices}{true}}{}{

\section{Data flow digram}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{paper_retrospective/data_flowchart.pdf}
    \caption[Overview of data flow from the initial data sources to the final stroke dataset.]{Overview of data flow from the initial data sources to the final stroke dataset.}
    \label{fig_retrospective:data_flowchart}
\end{figure}

\section{Machine learning pipeline}

\subsection{Model training}

We used stochastic gradient descent on mini-batches of data to train the stroke classification model. We used the Adam (adaptive moment estimation) optimisation algorithm and ensured an equal number of stroke positives and negatives in each batch by stratifying the class labels during sampling. We saved the model parameters after each epoch if the maximum F1-score (across all possible thresholds) improved in the validation dataset. We used the latest saved parameters as the final result of the run.

\subsection{Hyperparameters}\label{sec_retrospective:hyperparameters}

The selection of hyperparameters followed a simple two-stage process using validation data (table A1). First, a manual search was conducted by running different model configurations with varying numbers of epochs, updates per epoch, batch sizes, vectoriser types, and hyperparameters. Subsequently, a structured grid search was performed to further tune a subset of these hyperparameters.

\begin{table}[h]
    \centering
    \caption{Overview of hyperparameters used for training the text classification models.}
    \label{tab_retrospective:tableA1-hyperparameters}
    \resizebox*{0.98\textwidth}{!}{%
    \begin{tabular}{l|l|l}
        \toprule

        Name & Chosen value & Grid search range \\

        \midrule

        \multicolumn{3}{c}{\emph{Determined from initial heuristic, manual hyperparameter search on validation fold}}\\

        \midrule

        Epochs & 30 & - \\
        Parameter updates per epoch & 500 & - \\
        Batch size & 128 & - \\
        Batch sampling & Label stratified (balanced) & - \\
        Type of text vectoriser & Count vectorisation & - \\
        Size of bag-of-words vector & 10,000 & - \\
        Optimisation algorithm & Adam $(\beta_1=0.9, \beta_2=0.999)$ & - \\
        Learning rate schedule & Cosine annealing from start & - \\

        \midrule

        \multicolumn{3}{c}{\emph{Determined from grid search on validation fold}} \\

        \midrule

        Learning rate start & 0.0003 & \{0.003, 0.0003\} \\
        Learning rate end & 0.0003 & \{0.0003, 0.00003, 0.000003\} \\
        Model input dropout & 0.50 & \{0.25, 0.50\} \\
        Model dropout & 0.25 & \{0.25, 0.50\} \\
        Model configuration & [256, 128, 64, 32, 16] & \{[64, 32, 16], [256, 128, 64, 32, 16]\} \\

        \bottomrule
    \end{tabular}%
    }
\end{table}

\subsubsection{Bag-of-words selection}

Each transcript was transformed into a fixed-size bag-of-words vector to serve as input for the classification model. These vectors encode the occurrence of words and character n-grams within a fixed vocabulary. We selected vocabulary by first computing the χ2-statistics for all word uni-and bi-grams and character three-, four-, and fivegrams that occurred in more than ten training calls. We then retained the M highest-scoring word n-grams and $M$ highest-scoring character n-grams, yielding $2M$ input features, where $M$ represents a tuned hyperparameter. By complementing word n-grams with character n-grams, the model can use out-of-vocabulary words not included in the word n-grams and robustly represent words misspelt by the speech recogniser. The feature vector was input into the classification model (\cref{fig_retrospective:model_sketch}).

As part of our manual hyperparameter search, we trained the models using vectorisers of different sizes. We discovered that using 5,000-word n-grams and 5,000-character n-grams (M=5,000) struck a good trade-off between size, feature quality, and model performance, yielding 10,000 bag-of-words features.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{paper_retrospective/model_sketch.pdf}
    \caption[Overview of machine learning pipeline for stroke recognition.]{Overview of machine learning pipeline. Panel A presents a schematic overview of the machine learning pipeline. The individual models are broken down in panels B and C. The 2D convolutional layers have parameters [input channels, output channels]-[kernel width, kernel height]-[stride width, stride height]. The long short-term memory (LSTM) blocks have parameters hidden units-number of layers. The rectified linear unit (ReLU), sigmoid, and softmax layers have parameters [input features, output features]. Joining arrows indicate the concatenation of two vector sequences along the feature dimension. The full set of model hyperparameters is listed in \cref{sec_retrospective:hyperparameters}.}
    \label{fig_retrospective:model_sketch}
\end{figure}

\subsection{Ensembling details}

A common way to combine individual classification models into an ensemble is to use a voting scheme, such as majority voting, where a combined prediction is made based on the consensus among the individual models. However, this approach makes the ensemble not have a continuous output score. This is problematic for two reasons.
%
\begin{itemize}
    \item The lack of a continuous output score prevents the evaluation of the model's performance across a continuous range of thresholds required for plotting the receiver operating characteristic and precisionrecall curves.
    \item The lack of a continuous output score deteriorates the quality of assessing the effect of different words on ensemble performances (see \cref{sec_retrospective:model_explainability}).
\end{itemize}
%
Therefore, we used a different ensemble method, which is briefly described in the main text. Herein, we provide a mathematically rigorous definition of the proposed method.

Let $z^{(n,d)}$ be the logit output of model $n$ for transcript $d$, $t^{(n)}$ be the tuned logit threshold of model $n$, and $N$ be the number of models in the ensemble. The output score $p^{(n)}$ is then given by 
%
\begin{equation}
    p^{(d)} = \frac{1}{N} \sum_{n=1}^N \sigma(z^{(n,d)} - t^{(n)}) \enspace ,
\end{equation}
%
where $\sigma(\cdot)$ is the sigmoid function (or standard logistic function). The final ensembel predction $s^{(d)}$ is then simply 
%
\begin{equation}
    s^{(d)} = I_{p^{(d)} > 0.5}(p^{(d)}) \enspace ,
\end{equation}
%
where $I_{\cdot}(\cdot)$ is the indicator function that returns 1 if the subscript condition is satisfied and 0 otherwise.

\section{Significance testing and confidence intervals}

We used standard methods for significance testing and computing the confidence intervals. We used approximate methods owing to data size and to maintain computational feasibility \cite{cite25,cite26}

We performed \emph{paired approximate permutation} tests by pairing each observation from the first sample to a random observation from the other sample (without replacement), while keeping each observation within its original sample. This allowed us to test the significance of the observed pairings on the chosen statistics, i.e. whether a significant difference was observed in the test statistics depending on whether the call taker or model made the prediction. We used this approach to test whether
%
\begin{itemize}
    \item the model performed better than the call takers on the 2021 test set,
    \item including 112 training data improved the model's performance on the 2021 test set.
\end{itemize}
%
We performed \emph{independent approximate permutation} tests by randomly assigning observations to either of the two samples (without replacement) while maintaining any differences in sample size. This approach allowed us to test the significance of the observed sample assignments, i.e. whether there was a significant difference in test statistics depending on whether the predictions were assigned to the model or call taker. We used this approach to test whether.
%
\begin{itemize}
    \item the model performed better on the 2021 test set with diagnostic categories than on the test set without diagnostic categories,
    \item the model performed better on men than women on the 2021 test set,
    \item the model performed better on the 65+ group than on the 18-64 group on the 2021 test set,
    \item the call taker performed better on men than women on the 2021 test set,
    \item the call taker performed better on the 65+ group than on the 18-64 group on the 2021 test set.
\end{itemize}
%
The p-values were not exact because we used approximate permutation tests. However, owing to the large dataset size and substantial number of observations, the estimated p-values had tight confidence intervals. We reported the upper bound of the 99\% confidence interval on the p-value computed as the usual binomial distribution confidence interval.
%
\begin{equation}
    \text{CI}(p) = \hat{p} \pm z \sqrt{\frac{\hat{p}(1-\hat{p})}{N}} \enspace ,
\end{equation}
%
where $N$ is the number of resamplings.

We computed \emph{bootstrapped confidence intervals} for the statistics by resampling (with replacement) the predictions made by model or call taker on a relevant subgroup and recomputing the relevant statistics for each bootstrap sample. This process established a bootstrap distribution of the statistic that was then used to estimate the standard error and compute the confidence intervals. We computed confidence intervals using the bootstrap distribution percentiles. This method yielded reliable results because our dataset was large, and the bootstrap distribution was symmetrical and centred on the observed statistic. No observed bootstrap distributions differed significantly from normal distributions (according to Anderson-Darling and Shapiro-Wilk tests). Therefore, confidence intervals computed alternatively as studentised bootstrap intervals (t-intervals) did not differ substantially from percentile confidence intervals. This further validated our tests.

We used $N=15,000$ resamplings for permutation tests and $N=8,192$ for confidence intervals.

\section{Software}

We used Python version 3.8.10. PyTorch version 1.12.1 + cu113 was used to train the neural network models. We used SciKit-Learn (version 1.2.2) to perform bag-of-words vectorisation. We used NumPy version 1.23.5, Pandas version 1.5.3, Matplotlib version 3.7.1, and SciPy version 1.10.1 to perform data analysis, plotting, and testing.

\section{Additional results: Model performance across demographics}

When 1-1-2 data were not used for training (\cref{tab_retrospective:tableA2}), the model performed significantly better in men than in women on the test set in terms of all metrics (p < 0.0001) and significantly better on the 65+ group than on the 1864 group in terms of all metrics (p < 0.0001). As noted in the main text, both these statements were also true when 1-1-2 data were used for the training (\cref{tab_retrospective:table2}) and that including 1-1-2 data significantly improved overall performance (p < 0.0001).

Compared to the test set (\cref{tab_retrospective:table2}), model performance on the 2021 calls without diagnostic category was significantly worse in all demographic subgroups (\cref{tab_retrospective:tableA3}) in terms of all metrics (p < 0.0001), except for positive predictive value (PPV) for male (where p = 0.0056 [significant]) and for the false positive rate (FPR) on females and individuals aged 65+ years (where p = 0.213 and p = 0.362, respectively [insignificant]).

\todo[inline]{Insert table A2 here}
\begin{table}[h]
    \centering
    \caption{}
    \label{tab_retrospective:tableA2}
    % \resizebox*{0.98\textwidth}{!}{%
    % \begin{tabular}{l|l|l}
    %     \toprule

    %     \midrule

    %     \bottomrule
    % \end{tabular}%
    % }
\end{table}

\todo[inline]{Insert table A3 here}
\begin{table}[h]
    \centering
    \caption{}
    \label{tab_retrospective:tableA3}
    % \resizebox*{0.98\textwidth}{!}{%
    % \begin{tabular}{l|l|l}
    %     \toprule

    %     \midrule

    %     \bottomrule
    % \end{tabular}%
    % }
\end{table}

\section{Additional results: Detailed model explainability tables}

\todo[inline]{Insert table A4 here}
\begin{table}[h]
    \centering
    \caption{}
    \label{tab_retrospective:tableA4}
    % \resizebox*{0.98\textwidth}{!}{%
    % \begin{tabular}{l|l|l}
    %     \toprule

    %     \midrule

    %     \bottomrule
    % \end{tabular}%
    % }
\end{table}

\todo[inline]{Insert table A5 here}
\begin{table}[h]
    \centering
    \caption{}
    \label{tab_retrospective:tableA5}
    % \resizebox*{0.98\textwidth}{!}{%
    % \begin{tabular}{l|l|l}
    %     \toprule

    %     \midrule

    %     \bottomrule
    % \end{tabular}%
    % }
\end{table}

\section{Simulation of a prospective study on 2021 data}

\subsection{Method}

The machine learning frameworks can be deployed in different forms in clinical practice. To assess the potential outcomes of deploying this framework in a future prospective study, we performed an experiment using the 2021 test data to simulate different scenarios. Each scenario included two main variables.
%
\begin{enumerate}[label=\Roman*.]
    \item Presentation of the model prediction to the call taker.
    \item How does prediction influence the diagnostic code the call taker assigns to the call?
\end{enumerate}
%
There are two primary options as to when the model prediction is presented (I):
%
\begin{enumerate}
    \item Notify the call taker of potential false positive or negative stroke cases after the call ends.
    \item Notify the call taker of potential false positive or negative stroke cases during the call.
\end{enumerate}
%
Option 1 is identical to the method used in the main study. In option 2, predictions are made during the call based only on partial transcriptions. We implemented option 2 in such a manner that the model predicted every time 50 new words were transcribed and added to the transcript. A stroke positive was triggered only when three consecutive positive predictions were made (i.e. without intermediate negative stroke predictions). In other words, the sigmoid activation of the model had to remain above 0.5 for three consecutive predictions, for example, after 150, 200, and 250 words were transcribed.

As we can only assume how call takers are influenced by model predictions (II), precisely evaluating the hypothetical performance of call takers when supported by a machine learning framework is impossible. Furthermore, option 2 may influence the conversation, further complicating matters. Therefore, we report the results combining the call taker and the model under the following two assumptions:
%
\begin{enumerate}[label=\Alph*.]
    \item Call takers change any stroke prediction from negative to positive if the model predicts a positive (call takers mirror model positives).
    \item Call takers change any stroke prediction from positive to negative if the model predicts a negative (call takers mirror model negatives).
\end{enumerate}
%
By definition, method A tended to increase sensitivity and decrease PPV, whereas method B tended to decrease sensitivity and increase PPV.

We also report the results of the model itself (C). This method corresponds to call takers mirroring the model predictions exactly. This is not feasible in practice, although technically possible, because the conversation and instructions given to patients may conflict with the actions taken by the call taker after hanging up. Method 1.C is identical to the method employed in the main text, and we have copied the same results here for easier comparison.

\subsection{Results}

As expected (\cref{tab_retrospective:tableA6}), method 2.C (raw model predictions during calls) yielded slightly worse results than 1.C (raw model predictions after calls). Compared with method C, method A (call takers mirror model positives) led to increased sensitivity and decreased PPV, whereas method B (call takers mirror model negatives) led to decreased sensitivity and increased PPV, as expected. The numerical changes compared with method C are quite large because, in our simulation, the call taker is assumed to strictly follow methods A or B without divergence.

\todo[inline]{Insert table A6 here}
\begin{table}[h]
    \centering
    \caption{}
    \label{tab_retrospective:tableA6}
    % \resizebox*{0.98\textwidth}{!}{%
    % \begin{tabular}{l|l|l}
    %     \toprule

    %     \midrule

    %     \bottomrule
    % \end{tabular}%
    % }
\end{table}

Method 1.A (call takers mirror model positives after a call) yields a better F1-score, sensitivity, PPV, and FOR than call takers alone, although at the cost of a slightly higher FPR. This stands in contrast to methods 1.B, 2.A, and 2.B where either the sensitivity or the PPV is worse for the combined system than for call takers alone. Regardless, the F1-score (harmonic mean of sensitivity and PPV) is higher for all methods of combining call takers and model (1.A through 2.C).

These findings highlight that the implementation strategy selected for practice can substantially affect performance. Therefore, it may be possible to implement the system in a way that improves stroke recognition in practice.

}

### 2023-08-10 chapters/paper_retrospective.tex
Add more of stroke paper

@@
 % \todo[inline]{Insert Table 1}
-\begin{table}
+\begin{table}[t]
     \centering
@@
         \midrule
-        \multicolumn{6}{c}{All calls} \\
+        \multicolumn{6}{c}{\emph{All calls}} \\
         \midrule
@@
         \midrule
-        \multicolumn{6}{c}{Stroke calls} \\
+        \multicolumn{6}{c}{\emph{Stroke calls}} \\
         \midrule
@@
         \midrule
-        \multicolumn{6}{c}{Non-stroke calls} \\
+        \multicolumn{6}{c}{\emph{Non-stroke calls}} \\
         \midrule
@@
 
-\begin{figure}
+\begin{figure}[t]
     \centering
@@
 % \todo[inline]{Insert Table 2}
-\begin{table}
+\begin{table}[t]
     \centering
@@
         \midrule
-        \multicolumn{6}{c}{Overall} \\
+        \multicolumn{6}{c}{\emph{Overall}} \\
         \midrule
@@
         \midrule
-        \multicolumn{6}{c}{Without 112 training data} \\
+        \multicolumn{6}{c}{\emph{Without 112 training data}} \\
         \midrule
@@
         \midrule
-        \multicolumn{6}{c}{On MH-1813 data without diagnostic category} \\
+        \multicolumn{6}{c}{\emph{On MH-1813 data without diagnostic category}} \\
         \midrule
@@
         \midrule
-        \multicolumn{6}{c}{18-64 years} \\
+        \multicolumn{6}{c}{\emph{18-64 years}} \\
         \midrule
@@
         \midrule
-        \multicolumn{6}{c}{65+ years} \\
+        \multicolumn{6}{c}{\emph{65+ years}} \\
         \midrule
@@
         \midrule
-        \multicolumn{6}{c}{Male} \\
+        \multicolumn{6}{c}{\emph{Male}} \\
         \midrule
@@
         \midrule
-        \multicolumn{6}{c}{Female} \\
+        \multicolumn{6}{c}{\emph{Female}} \\
         \midrule
@@
 The relative performance gap between call-takers and the model was larger in our study than in the cardiac arrest study,15 which may affect the results of a potential RCT.
-To support future work and discussions beyond the scope of this study, the supplementary material includes the results of a simulation of a live implementation where call-takers are assumed to follow a set of fixed rules based on the output of the machine learning framework.
+To support future work and discussions beyond the scope of this study, \cref{app:retrospective} includes the results of a simulation of a live implementation where call-takers are assumed to follow a set of fixed rules based on the output of the machine learning framework.
 
@@
 We still reported the population characteristics and model performance of this group of calls to assess potential bias introduced by excluding them.
-A data-flow diagram is included in the supplementary material.
+A data-flow diagram is included in \cref{app:retrospective}.
 The resulting dataset is the largest dataset of audio files from stroke calls collected to date.
@@
 The final output score was used to classify whether the call concerned a stroke.
-The pipeline is illustrated in the Supplementary Material.
+The pipeline is illustrated in \cref{app:retrospective}.
 
-\todo[inline]{Insert Table 3}
+% \todo[inline]{Insert Table 3}
+\begin{table}[t]
+    \centering
+    \caption[Words with the largest positive and negative ranking score in calls predicted as stroke and non-stroke, respectively.]{English translation of words with the largest positive and negative ranking score in calls predicted as stroke and non-stroke, respectively.}
+    \label{tab_retrospective:table3}
+    \resizebox*{0.98\textwidth}{!}{%
+    \begin{tabular}{l|lr|lr}
+        \toprule
+                    & \multicolumn{2}{c}{Positive ranking score} & \multicolumn{2}{c}{Negative ranking score} \\
+        \midrule
+                    & \multicolumn{2}{c}{Stroke predictions, $D=1,897$} & \multicolumn{2}{c}{Non-stroke predictions, $D=342,133$} \\
+        \midrule
+                    & Word, $w$ \textit{(translated)} & Occurences, $D^{(w)}$ & Word, $w$ \textit{(translated)} & Occurences, $D^{(w)}$ \\
+        \midrule
+        1.  & Ambulance & 1,680 & Tetanus & 4,378 \\        
+        2.  & Blood clot & 895 & Pregnant & 8,749 \\        
+        3.  & Left & 1,108 & Cut & 7,592 \\        
+        4.  & Right & 1,050 & Bandage & 4,561 \\        
+        5.  & Double vision & 84 & Amager (a location) & 23,776 \\        
+        6.  & The words & 344 & O'clock & 94,436 \\        
+        7.  & Suddenly & 783 & The emergency room & 42,809 \\        
+        8.  & Arm & 709 & The police & 2,903 \\        
+        9.  & Side & 1,139 & Swollen & 60,559 \\        
+        10. & Stroke & 117 & Over the counter (OTC) & 4,641 \\        
+        11. & Double & 113 & The neck & 30,151 \\        
+        12. & Control & 134 & Fever & 112,586 \\        
+        13. & Call & 39 & Prescription & 5,450 \\        
+        14. & Numb & 94 & Centimetre & 12,026 \\        
+        15. & Minutes & 763 & The knee & 8,875 \\        
+        16. & Difficulties speaking & 44 & The pharmacy & 10,085 \\        
+        17. & Haemorrhagic stroke & 133 & The stomach & 42,105 \\        
+        18. & Hand & 297 & Psychiatric & 3,688 \\        
+        19. & The ambulance & 521 & Pneumonia & 7,597 \\        
+        20. & Slurred speech & 58 & Stomach pain & 10,551 \\        
+        21. & Blood clots & 224 & Stool & 19,155 \\        
+        22. & Fast & 663 & The ribs & 3,928 \\        
+        23. & Express & 44 & Bleed & 10,501 \\        
+        24. & Blood thinner & 259 & Bleeding & 24,313 \\        
+        25. & Incoherent & 15 & Ribs & 2,941 \\        
+        26. & Lopsided & 211 & Broken & 19,415 \\        
+        27. & Reduced & 528 & Inflammation & 10,050 \\        
+        28. & Hangs & 628 & Common cold & 8,127 \\        
+        29. & Transient & 48 & Morning or morrow & 78,558 \\        
+        30. & Not making sense & 14 & Swelling & 17,762 \\        
+        \bottomrule
+    \end{tabular}%
+    }
+\end{table}
 
@@
 As input for the classification model, each transcript was transformed into a fixed-size bag- of-words vector, which encoded the occurrence of word and character ($n$-grams) in a fixed vocabulary.
-The feature selection procedure is detailed in the Supplementary Material.
+The feature selection procedure is detailed in \cref{app:retrospective}.
 The model was constructed as an ensemble \cite{hansen_neural_1990} of five identical, independently trained models.
@@
 Subsequently, we defined the ensemble output score as the average of the centred output scores.
-The exact equations are provided in the Supplementary Material.
+The exact equations are provided in \cref{app:retrospective}.
 
@@
 
-\subsection{Model explainability}
+\subsection{Model explainability}\label{sec_retrospective:model_explainability}
 

### 2023-08-10 appendices/appendix_retrospective.tex
Minor fixes

@@
     \centering
-    \includegraphics[width=0.9\textwidth]{paper_retrospective/model_sketch.pdf}
+    \includegraphics[width=0.95\textwidth]{paper_retrospective/model_sketch.pdf}
     \caption[Overview of machine learning pipeline for stroke recognition.]{Overview of machine learning pipeline.
@@
 
-We used standard methods for significance testing and computing the confidence intervals. We used approximate methods owing to data size and to maintain computational feasibility \cite{cite25,cite26}
+We used standard methods for significance testing and computing the confidence intervals.
+We used approximate methods owing to data size and to maintain computational feasibility \cite{dwass_modified_1957,eden_validity_1933}
 

### 2023-08-11 appendices/appendix_benchmarking.tex
Finalize model-agnostic paper and add stroke paper
Add automated medical coding paper etc

@@
 
-\chapter[supplementary material for benchmarking generative latent variable models for speech]{Supplementary material for Benchmarking generative latent variable models for speech}
+\chapter[supplementary material: benchmarking generative latent variable models for speech]{Supplementary material: Benchmarking generative latent variable models for speech}
 \label{app:benchmarking}
@@
     \centering
+    \resizebox*{!}{\textheight}{
     \begin{tabular}{ll|lr}
@@
         \bottomrule
-    \end{tabular}
-    \vspace{2mm}
+    \end{tabular}%
+    }
     \caption[Model likelihoods on TIMIT represented as a 16 bit $\boldsymbol{\mu}$-law PCM]{
@@
 
-        % % x to s first layer
+        % x to s first layer
         \edge[]{s_1_1}{x_1};
@@
 
-        % % s per layer forward in time
+        % s per layer forward in time
         \edge[blue]{s_1_1}{s_1_2};
@@
 
-        % % s across layers on time
+        % s across layers on time
         \edge[blue]{s_2_1}{s_1_1};
@@
 
-        % % s across layers forward in time
+        % s across layers forward in time
         \edge[blue]{s_2_1}{s_1_2};
@@
     \end{tikzpicture}
-    \hspace{2cm}
+    % \hspace{2cm}
     }

### 2023-08-11 appendices/appendix_modelagnostic.tex
Add automated medical coding paper etc

@@
-\chapter[supplementary material for model-agnostic out-of-distribution detection using combined statistical tests]{Supplementary material for Model-Agnostic Out-of-Distribution Detection Using Combined Statistical Tests}
+\chapter[supplementary material: model-agnostic out-of-distribution detection using combined statistical tests]{Supplementary material: Model-Agnostic Out-of-Distribution Detection Using Combined Statistical Tests}
 \label{app:modelagnostic}

### 2023-08-11 appendices/appendix_retrospective.tex
Add automated medical coding paper etc

@@
-\chapter[supplementary material for a retrospective study on machine learning-assisted stroke recognition for medical helpline calls]{Supplementary Material for A Retrospective Study on Machine Learning-Assisted Stroke Recognition for Medical Helpline Calls}
+\chapter[supplementary material: a retrospective study on machine learning-assisted stroke recognition for medical helpline calls]{Supplementary Material for: A Retrospective Study on Machine Learning-Assisted Stroke Recognition for Medical Helpline Calls}
 \label{app:retrospective}
@@
 
-\todo[inline]{Insert table A2 here}
+% \todo[inline]{Insert table A2 here}
 \begin{table}[h]
     \centering
-    \caption{}
+    \caption{Model performance based on sex and age [mean (95\% CI)] when 1-1-2 training data is not used for training.}
     \label{tab_retrospective:tableA2}
     % \resizebox*{0.98\textwidth}{!}{%
-    % \begin{tabular}{l|l|l}
-    %     \toprule
+    \begin{tabular}{l|cc|cc}
+        \toprule
 
-    %     \midrule
+         & \multicolumn{2}{c|}{Model} & \multicolumn{2}{c}{Model} \\
+        \midrule
+         & Female & Male & 18-64 years & 65+ years \\
 
-    %     \bottomrule
-    % \end{tabular}%
+        \midrule
+
+        F1-score [\%] $\uparrow$ & \makecell{28.7 \\ (27.8-29.6)} & \makecell{36.2 \\ (35.3-37.2)} & \makecell{20.5 \\ (19.5-21.5)} & \makecell{39.4 \\ (38.6-40.3)} \\
+        \midrule
+        Sensitivity [\%] $\uparrow$ & \makecell{58.2 \\ (56.7-59.9)} & \makecell{62.1 \\ (60.7-63.6)} & \makecell{53.0 \\ (50.9-55.0)} & \makecell{63.0 \\ (61.9-64.3)} \\
+        \midrule
+        PPV [\%] $\uparrow$ & \makecell{19.0 \\ (18.3-19.7)} & \makecell{25.6 \\ (24.8-26.4)} & \makecell{12.7 \\ (12.0-13.4)} & \makecell{28.7 \\ (27.9-29.5)} \\
+        \midrule
+        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - specificity)}  & \makecell{0.077 \\ (0.073-0.080)} & \makecell{0.102 \\ (0.097-0.106)} & \makecell{0.034 \\ (0.032-0.036)} & \makecell{0.318 \\ (0.305-0.331)} \\
+        \midrule
+        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - NPV)} & \makecell{0.45 \\ (0.446-0.463)} & \makecell{0.483 \\ (0.473-0.494)} & \makecell{0.264 \\ (0.258-0.270)} & \makecell{1.335 \\ (1.308-1.362)} \\
+
+        \bottomrule
+    \end{tabular}%
     % }
@@
 
-\todo[inline]{Insert table A3 here}
+% \todo[inline]{Insert table A3 here}
 \begin{table}[h]
     \centering
-    \caption{}
+    \caption{Model performance based on sex and age [mean (95\% CI)] on the 2021 data without diagnostic category.}
     \label{tab_retrospective:tableA3}
     % \resizebox*{0.98\textwidth}{!}{%
-    % \begin{tabular}{l|l|l}
-    %     \toprule
+    \begin{tabular}{l|cc|cc}
+        \toprule
 
-    %     \midrule
+         & \multicolumn{2}{c|}{Model} & \multicolumn{2}{c}{Model} \\
+        \midrule
+         & Female & Male & 18-64 years & 65+ years \\
 
-    %     \bottomrule
-    % \end{tabular}%
+        \midrule
+
+        F1-score [\%] $\uparrow$ & \makecell{30.4 \\ (29.3-31.5)} & \makecell{35.2 \\ (34.0-36.4)} & \makecell{24.3 \\ (23.0-25.7)} & \makecell{36.6 \\ (35.6-37.6)} \\
+        \midrule
+        Sensitivity [\%] $\uparrow$ & \makecell{44.5 \\ (42.9-46.0)} & \makecell{52.8 \\ (51.1-54.4)} & \makecell{45.6 \\ (43.4-47.9)} & \makecell{49.2 \\ (47.9-50.5)} \\
+        \midrule
+        PPV [\%] $\uparrow$ & \makecell{23.1 \\ (22.2-24.1)} & \makecell{26.4 \\ (25.3-27.4)} & \makecell{16.6 \\ (15.5-17.5)} & \makecell{29.1 \\ (28.2-30.0)} \\
+        \midrule
+        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell{0.152 \\ (0.146-0.159)} & \makecell{0.155 \\ (0.147-0.162)} & \makecell{0.058 \\ (0.055-0.062)} & \makecell{0.372 \\ (0.359-0.386)} \\
+        \midrule
+        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - NPV)} & \makecell{0.404 \\ (0.394-0.414)} & \makecell{0.480 \\ (0.467-0.494)} & \makecell{0.246 \\ (0.239-0.254)} & \makecell{0.873 \\ (0.852-0.894)} \\
+
+        \bottomrule
+    \end{tabular}%
     % }
@@
 
-\todo[inline]{Insert table A4 here}
+% % \todo[inline]{Insert table A4 here}
 \begin{table}[h]
     \centering
-    \caption{}
+    \caption{Mean impact for words with the largest positive rank score in calls predicted as stroke.}
     \label{tab_retrospective:tableA4}
-    % \resizebox*{0.98\textwidth}{!}{%
-    % \begin{tabular}{l|l|l}
-    %     \toprule
-
-    %     \midrule
-
-    %     \bottomrule
-    % \end{tabular}%
-    % }
+    \resizebox*{\textwidth}{!}{%
+    \begin{tabular}{l|llccc}
+        \toprule
+        \multicolumn{6}{c}{Stroke predictions, $D=1,897$} \\
+        \midrule
+         & Word, $w$ (\textit{Danish}) & Translation (\textit{English}) & Rank, $r^{(w)}$ & Occurences, $D^{(w)}$ & \makecell{Impact, $i^{(d,w)}$ \\ mean $\pm$ std.} \\
+        \midrule    
+        1. & Ambulance & Ambulance & $1.000$ & $1,$680 & $0.52 \pm 0.51$ \\
+        2. & Blodprop & Blood clot & $0.599$ & $89$5 & $0.51 \pm 0.58$ \\
+        3. & Venstre & Left & $0.381$ & $1,$108 & $0.38 \pm 0.4$ \\
+        4. & Højre & Right & $0.326$ & $1,$050 & $0.31 \pm 0.42$ \\
+        5. & Dobbeltsyn & Double vision & $0.247$ & $84$ & $1.01 \pm 1.26$ \\
+        6. & Ordene & The words & $0.217$ & $34$4 & $0.6 \pm 0.45$ \\
+        7. & Pludselig & Suddenly & $0.142$ & $78$3 & $0.29 \pm 0.28$ \\
+        8. & Arm & Arm & $0.140$ & $70$9 & $0.3 \pm 0.3$ \\
+        9. & Side & Side & $0.125$ & $1,$139 & $0.23 \pm 0.21$ \\
+        10. & Apopleksi & Stroke & $0.102$ & $11$7 & $0.33 \pm 0.82$ \\
+        11. & Dobbelt & Double & $0.102$ & $11$3 & $0.54 \pm 0.72$ \\
+        12. & Styre & Control & $0.092$ & $13$4 & $0.63 \pm 0.46$ \\
+        13. & Opkald & Call & $0.067$ & $39$ & $0.18 \pm 1.22$ \\
+        14. & Følelsesløs & Numb & $0.065$ & $94$ & $0.53 \pm 0.58$ \\
+        15. & Minutter & Minutes & $0.064$ & $76$3 & $0.22 \pm 0.16$ \\
+        16. & Talebesvær & Difficulties speaking & $0.063$ & $44$ & $0.87 \pm 0.72$ \\
+        17. & Hjerneblødning & Haemorrhagic stroke & $0.060$ & $13$3 & $0.4 \pm 0.49$ \\
+        18. & Hånd & Hand & $0.057$ & $29$7 & $0.28 \pm 0.31$ \\
+        19. & Ambulancen & The ambulance & $0.055$ & $52$1 & $0.21 \pm 0.23$ \\
+        20. & Snøvler & Slurred speech & $0.052$ & $58$ & $0.71 \pm 0.54$ \\
+        21. & Blodpropper & Blood clots & $0.051$ & $22$4 & $0.27 \pm 0.36$ \\
+        22. & Hurtigt & Fast & $0.048$ & $66$3 & $0.18 \pm 0.18$ \\
+        23. & Udtrykke & Express & $0.044$ & $44$ & $0.59 \pm 0.74$ \\
+        24. & Blodfortyndende & Blood thinner & $0.044$ & $25$9 & $0.32 \pm 0.22$ \\
+        25. & Usammenhængende & Incoherent & $0.043$ & $15$ & $1.14 \pm 1.13$ \\
+        26. & Skæv & Lopsided & $0.039$ & $21$1 & $0.29 \pm 0.28$ \\
+        27. & Nedsat & Reduced & $0.038$ & $52$8 & $0.14 \pm 0.21$ \\
+        28. & Hænger & Hangs & $0.036$ & $62$8 & $0.15 \pm 0.17$ \\
+        29. & Forbigående & Transient & $0.035$ & $48$ & $0.52 \pm 0.62$ \\
+        30. & Vrøvler & Not making sense & $0.033$ & $14$ & $1.13 \pm 0.89$ \\
+        \bottomrule
+    \end{tabular}%
+    }
 \end{table}
 
-\todo[inline]{Insert table A5 here}
+% \todo[inline]{Insert table A5 here}
 \begin{table}[h]
     \centering
-    \caption{}
+    \caption{Mean impact for words with the largest negative rank score in calls predicted as non-stroke.}
     \label{tab_retrospective:tableA5}
-    % \resizebox*{0.98\textwidth}{!}{%
-    % \begin{tabular}{l|l|l}
-    %     \toprule
+    \resizebox*{\textwidth}{!}{%
+    \begin{tabular}{l|llcrr}
+        \toprule
+        \multicolumn{6}{c}{Non-stroke predictions, $D=342,133$} \\
+        \midrule
+         & Word, $w$ (\textit{Danish}) & Translation (\textit{English}) & Rank, $r^{(w)}$ & Occurences, $D^{(w)}$ & \makecell{Impact, $i^{(d,w)}$ \\ mean $\pm$ std.} \\
+        \midrule    
+        1. & Stivkrampe & Tetanus & $-1.000$ & $4378$ & $-19.40 \pm 10.61$ \\
+        2. & Gravid & Pregnant & $-0.901$ & $8749$ & $-12.08 \pm 8.64$ \\
+        3. & Skåret & Cut & $-0.772$ & $7592$ & $-11.98 \pm 8.61$ \\
+        4. & Forbinding & Bandage & $-0.569$ & $4561$ & $-12.87 \pm 10.08$ \\
+        5. & Amager & Amager (a location) & $-0.566$ & $23776$ & $-5.60 \pm 4.43$ \\
+        6. & Klokken & O'clock & $-0.535$ & $94436$ & $-2.22 \pm 2.69$ \\
+        7. & Skadestuen & The emergency room & $-0.486$ & $42809$ & $-3.72 \pm 3.23$ \\
+        8. & Politiet & The police & $-0.413$ & $2903$ & $-10.73 \pm 13.77$ \\
+        9. & Hævet & Swollen & $-0.388$ & $60559$ & $-2.84 \pm 2.38$ \\
+        10. & Håndkøb & over the counter (otc) & $-0.372$ & $4641$ & $-11.64 \pm 6.00$ \\
+        11. & Halsen & The neck & $-0.366$ & $30151$ & $-3.33 \pm 3.86$ \\
+        12. & Feber & Fever & $-0.361$ & $112586$ & $-1.94 \pm 1.76$ \\
+        13. & Recept & Prescription & $-0.334$ & $5450$ & $-9.87 \pm 5.82$ \\
+        14. & Centimetre & Centimetre & $-0.311$ & $12026$ & $-6.01 \pm 4.39$ \\
+        15. & Knæet & The knee & $-0.300$ & $8875$ & $-6.12 \pm 5.91$ \\
+        16. & Apoteket & The pharmacy & $-0.267$ & $10085$ & $-6.05 \pm 4.49$ \\
+        17. & Maven & The stomach & $-0.267$ & $42105$ & $-2.36 \pm 2.82$ \\
+        18. & Psykiatrisk & Psychiatric & $-0.263$ & $3688$ & $-8.99 \pm 8.49$ \\
+        19. & Lungebetændelse & Pneumonia & $-0.231$ & $7597$ & $-5.79 \pm 5.62$ \\
+        20. & Mavesmerter & Stomach pain & $-0.209$ & $10551$ & $-5.12 \pm 4.02$ \\
+        21. & Afføring & Stool & $-0.199$ & $19155$ & $-3.40 \pm 3.27$ \\
+        22. & Ribbenene & The ribs & $-0.195$ & $3928$ & $-8.26 \pm 6.18$ \\
+        23. & Bløde & Bleed & $-0.194$ & $10501$ & $-4.88 \pm 3.97$ \\
+        24. & Bløder & Bleeding & $-0.193$ & $24313$ & $-2.90 \pm 2.93$ \\
+        25. & Ribben & Ribs & $-0.189$ & $2941$ & $-8.96 \pm 7.56$ \\
+        26. & Brækket & Broken & $-0.183$ & $19415$ & $-3.49 \pm 2.83$ \\
+        27. & Betændelse & Inflammation & $-0.181$ & $10050$ & $-5.27 \pm 3.30$ \\
+        28. & Forkølet & Common cold & $-0.161$ & $8127$ & $-5.31 \pm 3.75$ \\
+        29. & Morgen & Morning or morrow & $-0.160$ & $78558$ & $-1.23 \pm 1.70$ \\
+        30. & Hævelse & Swelling & $-0.159$ & $17762$ & $-3.71 \pm 2.32$ \\
 
-    %     \midrule
-
-    %     \bottomrule
-    % \end{tabular}%
-    % }
+        \bottomrule
+    \end{tabular}%
+    }
 \end{table}
@@
 
-\todo[inline]{Insert table A6 here}
+% \todo[inline]{Insert table A6 here}
 \begin{table}[h]
     \centering
-    \caption{}
+    \caption{Overall performance of model, call takers and simulated combinations of model and call takers on MH-1813 test data.}
     \label{tab_retrospective:tableA6}
-    % \resizebox*{0.98\textwidth}{!}{%
-    % \begin{tabular}{l|l|l}
-    %     \toprule
+    \resizebox*{0.98\textwidth}{!}{%
+    \begin{tabular}{c|c|cc|cc|cc}
+        \toprule
 
-    %     \midrule
+        Predictor & Call taker & \multicolumn{2}{c|}{Model} & \multicolumn{4}{c}{Call taker supported by the model (simulated)} \\
+        \midrule
+        When & - & After call & During call & After call & During call & After call & During call \\
+        \midrule
+        Method & - & 1.C & 2.C & 1.A & 1.B & 2.A & 2.B \\
+        \midrule
 
-    %     \bottomrule
-    % \end{tabular}%
-    % }
+        F1-score [\%] $\uparrow$ & \makecell{25.8 \\ (23.7-27.9)} & \makecell{35.7 \\ (35.0-36.4)} & \makecell{33.1 \\ (32.4-33.7)} & \makecell{28.9 \\ (28.3-29.5)} & \makecell{33.3 \\ (32.5-34.1)} & \makecell{27.6 \\ (27.0-28.1)} & \makecell{32.7 \\ (31.8-33.5)} \\
+        \midrule
+        Sensitivity [\%] $\uparrow$ & \makecell{52.7 \\ (49.2-56.4)} & \makecell{63.0 \\ (62.0-64.1)} & \makecell{58.7 \\ (57.7-59.8)} & \makecell{72.4 \\ (71.5-73.3)} & \makecell{43.4 \\ (42.3-44.5)} & \makecell{72.3 \\ (71.4-73.3)} & \makecell{39.1 \\ (38.1-40.1)} \\
+        \midrule
+        PPV [\%] $\uparrow$ & \makecell{17.1 \\ (15.5-18.6)} & \makecell{24.9 \\ (24.3-25.5)} & \makecell{23.0 \\ (22.5-23.6)} & \makecell{18.0 \\ (17.6-18.4)} & \makecell{27.0 \\ (26.3-27.8)} & \makecell{17.0 \\ (16.7-17.4)} & \makecell{28.1 \\ (27.3-28.9)} \\
+        \midrule
+        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell{0.105 \\ (0.094-0.116)} & \makecell{0.082 \\ (0.079-0.085)} & \makecell{0.091 \\ (0.088-0.094)} & \makecell{0.061 \\ (0.059-0.064)} & \makecell{0.125 \\ (0.121-0.129)} & \makecell{0.061 \\ (0.059-0.064)} & \makecell{0.134 \\ (0.131-0.138)} \\
+        \midrule
+        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - NPV)} & \makecell{0.565 \\ (0.539-0.590)} & \makecell{0.419 \\ (0.413-0.426)} & \makecell{0.432 \\ (0.426-0.439)} & \makecell{0.726 \\ (0.717-0.735)} & \makecell{0.258 \\ (0.253-0.263)} & \makecell{0.776 \\ (0.767-0.786)} & \makecell{0.221 \\ (0.216-0.226)} \\
+
+        \bottomrule
+    \end{tabular}%
+    }
 \end{table}
@@
 
+\section{Research in context}
+
+\subsection{Evidence before this study}
+
+We searched the PubMed database for articles published in any language up to 9 May 2023 using the following terms: stroke AND (artificial intelligence OR machine learning OR deep learning) AND (EMS OR emergency medical services OR dispatch OR telephone).
+We identified 88 articles, none of which reported the results of machine learning-based stroke recognition during telephone calls.
+
+One study assessed the potential impact of using speech classification software for stroke recognition by extrapolating results from a similar solution for recognising cardiac arrest.
+Several authors of this study co-authored the article; however, it did not report the results of an actual novel machine learning framework.
+The remaining articles primarily reported the use of machine learning in imaging diagnostics, stroke recognition using movement-tracking mobile devices, and the development of stroke recognition tools using other non-audio data.
+
+\subsection{Added value of this study}
+
+This study is the first to investigate the use of a machine learning framework for stroke recognition in medical helpline calls.
+The study's results can be replicated in other call lines and for other acute illnesses.
+A machine learning framework has been previously described for outof-hospital cardiac arrest; however, our results illustrate the feasibility of employing machine learning for detecting stroke - a more complex acute condition.
+
+\subsection{Implications of all the available evidence}
+
+Implementing the framework described in this study could lead to improved recognition of patients with stroke during initial contact with health services.
+Improving recognition would result in more patients being eligible for advanced stroke treatment and better overall outcomes owing to faster referral to a stroke unit.
+
 }

### 2023-08-11 chapters/paper_automated.tex
Fix label
Add automated medical coding paper etc

@@
 \chapter[automated medical coding on mimic-iii and mimic-iv: a critical review and replicability study]{Automated Medical Coding on MIMIC-III and MIMIC-IV: A Critical Review and Replicability Study}
-\label{chp:paper-retrospective}
+\label{chp:paper-automated}
 \ifthenelse{\equal{\skippapers}{true}}{}{
 
+\newcommand{\lacaml}{LA$_{\textsc{caml}}$}
+\newcommand{\lalaat}{LA$_{\textsc{laat}}$}
+
 \section*{Abstract}
+Medical coding is the task of assigning medical codes to clinical free-text documentation.
+Healthcare professionals manually assign such codes to track patient diagnoses and treatments.
+Automated medical coding can considerably alleviate this administrative burden.
+In this paper, we reproduce, compare, and analyze state-of-the-art automated medical coding machine learning models.
+We show that several models underperform due to weak configurations, poorly sampled train-test splits, and insufficient evaluation.
+In previous work, the macro F1 score has been calculated sub-optimally, and our correction doubles it.
+We contribute a revised model comparison using stratified sampling and identical experimental setups, including hyperparameters and decision boundary tuning.
+We analyze prediction errors to validate and falsify assumptions of previous works.
+The analysis confirms that all models struggle with rare codes, while long documents only have a negligible impact.
+Finally, we present the first comprehensive results on the newly released MIMIC-IV dataset using the reproduced models.
+We release our code, model parameters, and new MIMIC-III and MIMIC-IV training and evaluation pipelines to accommodate fair future comparisons.\footnote{\label{footnote:source_code}\url{https://github.com/JoakimEdin/medical-coding-reproducibility}}
+
+\section{Introduction}
+
+Medical coding is the task of assigning diagnosis and procedure codes to free-text medical documentation \cite{dongAutomatedClinicalCoding2022,tengReviewDeepNeural2022}.
+
+These codes ensure that patients receive the correct level of care and that healthcare providers are accurately compensated for their services.
+However, this is a costly manual process prone to error \cite{tsengAdministrativeCostsAssociated2018, omalleyMeasuringDiagnosesICD2005, burnsSystematicReviewDischarge2012}.
+
+The goal of automated medical coding (AMC) is to predict a set of codes or provide a list of codes ranked by relevance for a medical document.
+Numerous machine learning models have been developed for AMC \cite{jiUnifiedReviewDeep2022, tengReviewDeepNeural2022, stanfillSystematicLiteratureReview2010}.
+These models are trained on datasets of medical documents, typically discharge summaries, each labeled with a set of medical codes.
+While some models treat AMC as an ad-hoc information retrieval problem \cite{rizzoICDCodeRetrieval2015,parkInformationRetrievalApproach2019}, it is more commonly posed as a multi-label classification problem \cite{tengReviewDeepNeural2022, jiUnifiedReviewDeep2022}.
+
+While most research in AMC has been conducted on the third version of the Medical Information Mart for Intensive Care dataset (MIMIC-III) \cite{tengReviewDeepNeural2022, venkateshAutomatingOverburdenedClinical2023}, it remains challenging to compare the results of different models.
+Performance improvements are commonly attributed to model design, but differences in experimental setups make these claims hard to validate.
+In addition, long documents, rare codes, and lack of training data are often cited as core research challenges \cite{baoMedicalCodePrediction2021,dongAutomatedClinicalCoding2022,dongExplainableAutomatedCoding2021,feuchtDescriptionbasedLabelAttention2021,gaoLimitationsTransformersClinical2021,huangPLMICDAutomaticICD2022,jiDoesMagicBERT2021,jiUnifiedReviewDeep2022,kavuluruEmpiricalEvaluationSupervised2015,kimReadAttendCode2021,liICDCodingClinical2020,liuEffectiveConvolutionalAttention2021,michalopoulosICDBigBirdContextualEmbedding2022,moonsComparisonDeepLearning2020,pascualBERTbasedAutomaticICD2021,tengReviewDeepNeural2022,tengExplainablePredictionMedical2020,xieEHRCodingMultiscale2019,yangKnowledgeInjectedPrompt2022,zhangBERTXMLLargeScale2020,zhouAutomaticICDCoding2021, vuLabelAttentionModel2020, venkateshAutomatingOverburdenedClinical2023}.
+However, except for a few studies demonstrating performance drops on rare codes, the number of studies containing in-depth error analyses is limited \cite{baoMedicalCodePrediction2021,dongExplainableAutomatedCoding2021,jiDoesMagicBERT2021}.
+
+We address the above challenges. Our major contributions are:
+\begin{enumerate}
+    \item We reproduce the performance of state-of-the-art models on MIMIC-III. We find that evaluation methods are flawed and propose corrections that double the macro F1 scores.
+    \item We find the original split of MIMIC-III to introduce strong biases in results due to missing classes in the test set.
+We create a new split with full class representation using stratified sampling.
+    \item We perform a revised model comparison on MIMIC-III \textit{clean} using the same training, evaluation, and experimental setup for all models.
+We find that models previously reported as low-performing improve considerably, demonstrating the importance of hyperparameters and decision boundary tuning.
+    \item We report the first results of current state-of-the-art models on the newly released MIMIC-IV dataset \cite{johnsonMIMICIVFreelyAccessible2023, goldbergerPhysioBankPhysioToolkitPhysioNet2000}.
+We find that previous conclusions generalize to the new dataset.
+    \item Through error analysis, we provide empirical evidence for multiple model weaknesses.
+Most importantly, we find that rare codes harm performance, while, in contrast to previous claims, long documents only have a negligible performance impact.
+\end{enumerate}
+We release our source code and new splits for MIMIC-III and IV$^\text{1}$
+and hope these contributions will aid future research in AMC.
+
+\section{Previous work}
+
+In the following, we review datasets, model architectures, training, and evaluation of the models we compare in this study.
+Our criteria for selecting these models are presented in \cref{sec: inclusion and exclusion criteria}.
+
+\subsection{Datasets}
+\label{sec:dataset}
+The International Classification of Diseases (ICD) is the most popular medical coding system worldwide \cite{tengReviewDeepNeural2022}.
+It follows a tree-like hierarchical structure, also known as a medical ontology, to ensure the functional and structural integrity of the classification.
+Chapters are the highest level in the hierarchy, followed by categories, sub-categories, and codes.
+The World Health Organization (WHO) revises ICD periodically.
+Each revision introduces new codes.
+For instance, ICD-9 contains 18,000 codes, while ICD-10 contains 142,000.\footnote{\url{https://www.cdc.gov/nchs/icd/icd10cm_pcs_background.htm}} MIMIC-II and MIMIC-III are the most widely used open-access datasets for research on ICD coding and are provided by the Beth Israel Deaconess Medical Center \cite{tengReviewDeepNeural2022, johnsonMIMICIIIFreelyAccessible2016, leeOpenaccessMIMICIIDatabase2011}.
+
+MIMIC-III contains medical documents annotated with ICD-9 codes collected between 2001 and 2012 \cite{johnsonMIMICIIIFreelyAccessible2016}.
+Usually, discharge summaries---free-text notes on patient and hospitalization history---are the only documents used for AMC \cite{tengReviewDeepNeural2022}.
+MIMIC-III \textit{full} and \textit{50} are commonly used splits.
+MIMIC-III \textit{full} contains all ICD-9 codes, while \textit{50} only contains the top 50 most frequent codes \cite{mullenbachExplainablePredictionMedical2018, shiAutomatedICDCoding2018}.
+
+MIMIC-IV was released on January 6th, 2023, and has not previously been used for AMC.
+It contains data for patients admitted to the Beth Israel Deaconess Medical Center emergency department or ICU between 2008-2019  annotated with either ICD-9 or ICD-10 codes \cite{johnsonMIMICIVFreelyAccessible2023}.
+The empirical frequencies of codes of each ICD version in MIMIC-IV are shown in \cref{fig:mimiciv_code_dist}.
+Statistics for the MIMIC-III \textit{50}, \textit{full}, and MIMIC-IV datasets are listed in \cref{tab:subsets}.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.9\linewidth]{paper_automated/mimiciv_code_freq.png}
+    \caption[The frequency of ICD-9 and ICD-10 codes in MIMIC-IV before pre-processing.]{The frequency of ICD-9 and ICD-10 codes in MIMIC-IV before pre-processing.
+As discussed in \cref{sec: splits}, we removed codes with fewer than ten instances (dashed line).}
+    \label{fig:mimiciv_code_dist}
+\end{figure}
+
+\begin{sidewaystable}[t]
+    \centering
+    \caption{Comparison of the previously defined MIMIC-III splits \textit{full} and \textit{50} \cite{mullenbachExplainablePredictionMedical2018} and our proposed MIMIC-III \textit{clean} split along with similarly defined splits for MIMIC-IV \textit{ICD-9} and \textit{ICD-10} after pre-processing.}
+    \label{tab:subsets}
+    \resizebox*{\textwidth}{!}{%
+    \begin{tabular}{lccccc}
+        \toprule
+        & \multicolumn{2}{c}{\textit{Previous work}} & \multicolumn{3}{c}{\textit{Our work}} \\
+        \cmidrule(lr){2-3}\cmidrule(lr){4-6}
+        & {MIMIC-III \textit{full}} & {MIMIC-III 50}  & {MIMIC-III \textit{clean}}  & {MIMIC-IV \textit{ICD-9}}  & {MIMIC-IV \textit{ICD-10}}\\
+        \midrule
+        Number of documents  & 52,723 & 11,368 &  52,712 & 209,326 & 122,279 \\
+        Number of patients & 41,126 & 10,356 & 41,118 & 97,709 & 65,659 \\
+        Number of unique codes  & 8,929 & 50  & 3,681 & 6,150 & 7,942\\
+        Codes pr. instance: Median (IQR)    & 14 (10-20) & 5 (3-8) & 14 (10-20) & 12 (8-17) & 14 (9-20)\\
+        Words pr. document: Median (IQR) & 1,375 (965-1,900) & 1,478 (1,065-1,992) & 1,311 (917-1,822) & 1,320 (997-1,715) & 1,492 (1,147-1,931)\\
+        Documents: Train/val/test [\%] & 90.5/3.1/6.4 & 71.0/13.8/15.2 & 72.9/10.6/16.6  & 73.8/10.5/15.7& 72.9/10.9/16.2  \\
+        Missing codes: Train/val/test [\%] & 2.7/66.4/54.3 & 0.0/0.0/0.0 & 0.0/0.1/0.0 & 0.0/0.5/0.2 & 0.0/0.5/0.1\\
+        \bottomrule
+    \end{tabular}%
+    }
+\end{sidewaystable}
+
+\subsection{Model architectures}
+
+Most recent state-of-the-art AMC models use an encoder-decoder architecture.
+The encoder takes a sequence of tokens $T \in \mathbb{Z}^{n}$ as input and outputs a sequence of hidden representations $H \in \mathbb{R}^{d_h \times n}$, where $n$ is the number of tokens in a sequence, and $d_h$ is the hidden dimension.
+The decoder takes $H$ as input and outputs the code probability distributions.
+For the task of ranking, codes are sorted by decreasing probability.
+For classification, code probabilities larger than a set decision boundary are predicted.
+
+\subsubsection{Encoders:}
+The encoder usually consists of pre-trained non-contextualized word embeddings (e.g., Word2Vec) and a neural network for encoding context.
+More recently, pre-trained masked language models (e.g., BERT) have gained popularity~\cite{tengReviewDeepNeural2022}.
+The MIMIC-III training set or PubMed articles are commonly used for pre-training.
+
+\subsubsection{Decoders:}
+The most common decoder architectures can be grouped into three primary types. 
+The simplest decoder is a pooling layer (e.g., max pooling) followed by a feed-forward neural network.
+More recently, label-wise attention (LA) \cite{mullenbachExplainablePredictionMedical2018} has replaced pooling \cite{vuLabelAttentionModel2020, liuEffectiveConvolutionalAttention2021, huangPLMICDAutomaticICD2022, liICDCodingClinical2020}.
+LA transforms a sequence of hidden representations $H$ into label-specific representations $V \in \mathbb{R}^{d_h \times L}$, where $L$ is the number of unique medical codes in the dataset.
+It is computed as
+\begin{equation}
+    A = \text{softmax}(WH) \, , \quad
+    V = HA^{\top} \enspace ,
+\end{equation}
+where the softmax normalizes each column of $WH$, $W \in \mathbb{R}^{L \times d_h}$ is an embedding matrix that learns label-specific queries, and $A \in \mathbb{R}^{L \times n}$ is the attention matrix.
+Then, $V$ is used to compute class-wise probabilities via a feedforward neural network. 
+As LA was first used in the \textit{convolutional attention for multi-label classification} (CAML) model~\cite{mullenbachExplainablePredictionMedical2018}, we refer to this method as \lacaml.
+
+An updated label-wise attention module was introduced in the \textit{label attention model} (LAAT) \cite{vuLabelAttentionModel2020}.
+We refer to this attention module as \lalaat.
+In \lalaat, the label-specific attention is computed similarly to  \lacaml~ as $A = \text{softmax}(UZ)$, where $U \in \mathbb{R}^{L \times d_p}$ is a learnable embedding matrix, but with
+$Z = \text{tanh}(PH)$ 
+where $P \in \mathbb{R}^{d_p \times d_h}$ is a learnable matrix, $Z \in \mathbb{R}^{d_p \times n}$ and $d_p$ is a hyperparameter.
+
+\subsection{Training and evaluation methods}\label{sec: training and evaluation methods}
+
+\citet{mullenbachExplainablePredictionMedical2018} released code for pre-processing the discharge summaries, generating the train-test split, and evaluating model performance on MIMIC-III which many subsequent papers have used \cite{vuLabelAttentionModel2020, huangPLMICDAutomaticICD2022, liICDCodingClinical2020, baoMedicalCodePrediction2021, yuanCodeSynonymsMatter2022, kimReadAttendCode2021}.
+Pre-processing consisted of lowercasing all text and removing words that only contain out-of-alphabet characters. Predicting procedures and diagnosis codes were treated as a single task.
+The dataset was split into training, validation, and test sets using random sampling, ensuring that no patient occurred in both the training and test set.
+The (non-stratified) random sampling lead to 54\% of the ICD codes in MIMIC-III \textit{full} not being sampled in the test set.
+This complicates the interpretation of results since these codes only contribute true negatives or false positives.
+Models are evaluated using the micro and macro average of the area under the curve of the receiver operating characteristics (AUC-ROC), F1 score, and Precision@k.
+
+While most papers use the pre-processing, train-test split, and evaluation metrics described above, they differ in several aspects of training.
+This may lead to performance differences unrelated to modeling choices which are undesirable when seeking to compare models.
+For instance, due to varying memory constraints of different models, documents are usually truncated to some maximum length.
+In the literature, this maximum varies from 2,500 to 4,000 words \cite{mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, huangPLMICDAutomaticICD2022}.
+Furthermore, not all papers tune the prediction decision boundary but simply set it to 0.5,
+ hyperparameter search ranges and sampling methods  vary between works, and learning rate schedulers are only used in LAAT and PLM-ICD\cite{mullenbachExplainablePredictionMedical2018, liICDCodingClinical2020}.
+In LAAT, the learning rate was decreased by 90\% when the F1 score had not increased for five epochs.
+PLM-ICD used a schedule with linear warmup followed by linear decay.
+
+All models except for PLM-ICD use Word2Vec embeddings pre-trained on the MIMIC-III training set.
+PLM-ICD uses a BERT encoder pre-trained on PubMed to encode the text in chunks of 128 tokens, and these contextualized embeddings are fed to a \lalaat~ layer.
+
+Finally, all models compute independent code probabilities using sigmoid activation functions and optimize the binary cross entropy loss function during training.
+\cref{tab:model_facts} presents the selected models.
+
+\section{Methods}
+
+\begin{table}[t]
+    \centering
+    \caption{An overview of the compared models.}
+    \label{tab:model_facts}
+    \begin{tabular}{lcccc}
+        \toprule
+        Model   & Encoder & Decoder  & Param \\
+        \midrule
+        Bi-GRU \cite{mullenbachExplainablePredictionMedical2018}   &  Word2Vec, Bi-GRU & Max-pool  &  9.9M\\
+        CNN \cite{mullenbachExplainablePredictionMedical2018}   & Word2Vec, CNN & Max-pool &  10.3M\\
+        CAML \cite{mullenbachExplainablePredictionMedical2018}   & Word2Vec, CNN  & \lacaml &  6.1M\\
+        MultiResCNN \cite{liICDCodingClinical2020} &  Word2Vec, ResNet & \lacaml &  11.9M\\
+        LAAT \cite{vuLabelAttentionModel2020} & Word2Vec, Bi-LSTM  & \lalaat  &  21.9M\\
+        PLM-ICD \cite{huangPLMICDAutomaticICD2022} &  BERT & \lalaat &  138.8M\\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+\subsection{Selection criteria}\label{sec: inclusion and exclusion criteria}
+
+In this study, we included both models trained from scratch and models with components pre-trained on external corpora. 
+We excluded models that use multi-modal inputs, such as medical code descriptions \cite{kimReadAttendCode2021, mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, caoHyperCoreHyperbolicCograph2020, baoMedicalCodePrediction2021}, code synonyms \cite{yuanCodeSynonymsMatter2022}, code hierarchies \cite{caoHyperCoreHyperbolicCograph2020, xieEHRCodingMultiscale2019}, or associated Wikipedia articles \cite{baiImprovingMedicalCode2019}, because they introduced additional complexity without providing evidence for significant performance improvements \cite{mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, tengReviewDeepNeural2022}.
+We excluded works without publically available source code as the experiment descriptions often lacked important implementation details.
+
+\subsection{Evaluation metrics}
+\label{sec:metrics}
+Similar to previous work, we evaluated models using AUC-ROC, F1 score, and precision@$k$.
+Additionally, we introduced exact match ratio (EMR), R-precision, and mean average precision (MAP).
+The EMR is the percentage of instances where all codes were predicted correctly.
+This allowed us to measure how many documents were predicted perfectly, which is important for \textit{fully automated} medical coding.
+Where precision@$k$ is computed based on the top-$k$ codes (i.e., $k$ is fixed), R-precision considers a number of codes equal to the true number of relevant codes.
+Thus, R-precision is useful when the number of relevant codes varies considerably between documents, which is the case for the MIMIC datasets.
+Finally, in contrast to all other metrics, MAP considers the exact rank of all relevant codes in a document.
+
+\begin{sidewaystable}[t]
+    \centering
+    \caption[Hyperparameters, maximum document lengths, and decision boundary tuning strategies used in the original works compared to improved settings]{Hyperparameters, maximum document lengths, and decision boundary tuning strategies used in the original works compared to the optimal settings found in this paper (marked with *).
+LR is the learning rate scheduler.
+``Length" is the maximum number of words a document can contain before being truncated.
+$\dagger$ applies to models using word-piece tokenization.
+These models were filtered on the number of sub-words instead of full words.
+``DB tune" is whether the optimal decision boundary was found using the validation set.
+If a paper did not tune the decision boundary, it was set to 0.5.}
+    \label{tab:hyperparameters}
+    \resizebox*{\textwidth}{!}{%
+    \begin{tabular}{lccccccccc}
+        \toprule
+        & \multicolumn{7}{c}{Hyperparameters} & & \\
+        \cmidrule(lr){2-8}
+        Model & Batch Size & Weight Decay & Learning Rate & Dropout & LR Scheduler & Optimizer & Epochs & Length & DB tune \\
+        \midrule
+        Bi-GRU  & 16 &  0.0 & 0.003 &  0.2   & no  & Adam &  100 & 2500 & no\\
+        Bi-GRU$^*$  & 8 &  0.0001 & 0.001 &  0   & yes  & AdamW &  20 & 4000 & yes\\
+        \hline
+        CNN  & 16 &  0.0 & 0.003  &  0.2 & no  & Adam &  100 & 2500 & no\\
+        CNN$^*$    & 8 &  0.00001 & 0.001  &  0 & yes  & AdamW &  20 & 4000 & yes \\
+        \hline
+        CAML  & 16 &  0.0 & 0.0001  &  0.2  & no  & Adam  &  200 & 2500 & no\\
+        CAML$^*$    & 8 &  0.001 & 0.005  &  0.2  & yes  & AdamW  &  20 & 4000 & yes\\
+        \hline
+        MultiResCNN & 16 &  0.0 & 0.0001  &  0.2  & no  & Adam &  200 & 2500 & no\\
+        MultiResCNN$^*$  & 16 &  0.0001 & 0.0005  &  0.2  & yes  & AdamW &  20 & 4000 & yes\\
+        \hline
+        LAAT   & 8 &  0.0 & 0.0001  &  0.3  & yes  & AdamW &  50 & 4000 & no\\
+        LAAT$^*$    & 8 &  0.001 & 0.001  &  0.2  & yes  & AdamW &  20 & 4000 & yes\\
+        \hline
+        PLM-ICD  & 8 &  0.0 & 0.00005  &  0.2  & yes  & AdamW &  20 & 3072$^\dagger$ & yes \\
+        PLM-ICD$^*$   & 16 &  0.0 & 0.00005  &  0.2  & yes  & AdamW &  20 & 4000 & yes \\
+        \bottomrule
+    \end{tabular}%
+    }
+\end{sidewaystable}
+
+Previous works calculated the macro F1 score as the harmonic mean of the macro precision and macro recall \cite{mullenbachExplainablePredictionMedical2018, liICDCodingClinical2020, vuLabelAttentionModel2020, huangPLMICDAutomaticICD2022}.
+\citet{opitzMacroF1Macro2021} analyze macro F1 formulas common in multi-class and multi-label classification.
+They demonstrate that the above formulation is sub-optimal, as it rewards heavily biased classifiers in unbalanced datasets.
+Therefore, as recommended by the authors, we calculated the macro F1 score as the arithmetic mean of the F1 score for each class.
+As seen in \cref{tab:subsets}, 54\% of codes in MIMIC-III \textit{full} are missing in the test set.
+Previous works set the F1 score of all the missing codes in the test set to 0, resulting in a misleadingly low macro F1 score.
+Because 54\% of the codes are missing, the maximum possible macro F1 score is 46\%.
+We ignored all codes not in the test set for our reproduction, essentially trading bias for variance. 
+For our revised comparison, we resolved the issue by instead sampling new splits that reduce missing codes to a negligible fraction (see \cref{sec: splits}) and ignoring the few that were still missing.
+
+\subsection{Definition of splits}
+\label{sec: splits}
+
+% \cref{tab:subsets}
+
+We define three new splits: MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9}, and \textit{ICD-10}.
+As described in \cref{sec:metrics}, 54\% of the codes in MIMIC-III \textit{full} are absent from the test set, which introduces significant bias in the model evaluation metrics.
+Therefore, we created a new MIMIC-III split to ensure that most codes are present in both the training and test set.
+Specifically, we removed codes with fewer than ten occurrences, doubled the test set size, and sampled the documents using multi-label stratified sampling \cite{sechidisStratificationMultilabelData2011}.
+We ensured that no patient occurred in both the training and test set, preprocessed the text, and considered procedures and diagnosis codes as a single task as done by \citet{mullenbachExplainablePredictionMedical2018}.
+We based our new split on the v1.4 version of the dataset and refer to it as MIMIC-III \textit{clean}.
+Using the same method, we created two splits for MIMIC-IV v2.2: one containing all documents labeled with ICD-9 codes and one with ICD-10 codes.
+
+\subsection{Reproducibility experiments}
+
+We ran reproducibility experiments with all models to evaluate whether the results in the original works could be reproduced and to validate our reimplementations.
+We ran these experiments on MIMIC-III \textit{full}, and \textit{50} as in the original works \cite{mullenbachExplainablePredictionMedical2018,liICDCodingClinical2020,vuLabelAttentionModel2020,huangPLMICDAutomaticICD2022}.
+We used the hyperparameters reported in each paper (see \cref{tab:hyperparameters}) and report both the original and the revised macro F1 scores discussed in \cref{sec:metrics}.
+
+\subsection{Revised comparison}
+To address the issues associated with comparing results reported by previous works described in \cref{sec: training and evaluation methods,sec:metrics}, we perform a revised model comparison.
+We run experiments on the new MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9}, and \textit{ICD-10}.
+All models were trained for 20 epochs using a learning rate schedule with linear warmup for the first 2K updates followed by linear decay \cite{huangPLMICDAutomaticICD2022}.
+We found this schedule to speed up the training convergence of all the models.
+Whereas original works use Adam or AdamW, we used AdamW for all experiments as it corrects the weight decay implementation of Adam \cite{kingmaAdamMethodStochastic2017, loshchilovDecoupledWeightDecay2022}.
+For each model, we tuned the decision boundary to maximize the micro F1 score on the validation set.
+We used randomized sampling to find optimal settings for dropout, weight decay, learning rate, and batch size.
+The hyperparameter search was performed on MIMIC-III \textit{clean}, and the MIMIC-IV splits.
+We found that the best setting for each model generalized across datasets.
+Using this setting, we ran each model ten times with different seeds on each dataset.
+All documents were truncated to a maximum of 4000 words.
+The hyperparameters, maximum document lengths, and decision boundary tuning strategy are summarised in \cref{tab:hyperparameters}.
+
+We performed an ablation study to analyze which changes had the largest impact on performance. 
+Specifically, we evaluated the effect of truncation, hyperparameter search, and decision boundary tuning.
+We modified one of these at a time: We ran one experiment where documents were truncated to a maximum length of 2,500 words, a second experiment where the models were trained with the hyperparameters, number of epochs, and learning rate schedule used in the original works, and a third experiment where the decision boundary was set to 0.5 instead of tuned.
+
+\subsection{Error analysis}
+To validate and falsify the commonly cited challenges of AMC, which include a lack of training data, long documents, and rare codes, we performed an error analysis.
+In addition to analyzing rare codes, we contribute an in-depth code analysis aiming to identify the attributes that make certain codes challenging to predict.
+
+\subsubsection{Amount of training data:}
+Multiple studies attribute poor performance to data sparsity of MIMIC-III, which contains only fifty thousand examples \cite{kavuluruEmpiricalEvaluationSupervised2015,tengExplainablePredictionMedical2020,yanMedicalCodingClassification2010,yangKnowledgeInjectedPrompt2022}.
+MIMIC-IV \textit{ICD-9} contains four times as many examples, which allows analyzing the effect of training set size.
+We train each model on 25k, 50k, 75k, 100k, and 125k examples and report micro and macro F1 on the fixed test set.
+The training subsets were sampled from the training set using multi-label stratified sampling to ensure the same code distributions \cite{sechidisStratificationMultilabelData2011}.
+
+\subsubsection{Document length:}
+We analyzed whether model performance correlates with document length on MIMIC-IV \textit{ICD-9}.
+Specifically, we calculated the Pearson and Spearman correlation between the number of words in the documents and the micro F1 score for all models.
+For each model, we used the best seed from the revised comparison.
+
+\subsubsection{Code analysis:}
+To analyze the performance impact of rare codes, we first calculated the Pearson and Spearman correlation between model performance on each code and the corresponding code frequency in the training data.
+We calculated these correlations for all splits.
+To identify attributes of challenging codes, we analyzed model performance on the chapter level of the ICD-10 classification system.
+Using high-level chapters instead of codes allows us to group examples into categories, which we use as a starting point for further analysis.
+We limit the scope of the analysis to diagnosis codes.
+We focused on ICD-10 because it is the classification system currently in use at most hospitals.
+
+\begin{sidewaystable}[t]
+    \centering
+    \caption[Reproduced test set results compared with those from the original works]{Reproduced test set results compared with those from the original works.
+Our reproduced results are indicated with *.
+The results were reproduced on MIMIC-III v1.4 with the preprocessing pipeline and splits of \citet{mullenbachExplainablePredictionMedical2018}.
+Each model was reproduced using the hyperparameters presented in the respective paper.
+We use both macro F1 formulas: Macro$^\dagger$ refers to the method used in the original work, while Macro refers to the corrected version used in this paper.}
+    \label{tab:reproduced_results}
+    \resizebox*{\textwidth}{!}{%
+    \begin{tabular}{l  cc  ccc  cc  cc  ccc  c}
+        \toprule
+        & \multicolumn{7}{c}{MIMIC-III \textit{full}} & \multicolumn{6}{c}{MIMIC-III \textit{50}} \\
+        \cmidrule(lr){2-8}\cmidrule(lr){9-14}
+        & \multicolumn{2}{c}{AUC-ROC} & \multicolumn{3}{c}{F1} & \multicolumn{2}{c}{Precision@$k$}  & \multicolumn{2}{c}{AUC-ROC} & \multicolumn{3}{c}{F1} & Precision@$k$  \\
+        & Micro & Macro & Micro & Macro$^\dagger$ & Macro & 8 & 15 & Micro & Macro & Micro & Macro$^\dagger$ & Macro & 5 \\
+        \midrule
+        CNN &  96.9 & 80.6 &  41.9& 4.2 & -& 58.1 &  48.8  & 90.7 & 87.6 & 62.5 & 57.6 & - & 62.0  \\
+        CNN$^*$ & 97.3 & 83.1 & 41.5 & 3.4 & 6.7 & 61.9 & 47.2 & 91.9 & 89.2 & 64.9 & 58.8 & 58.0 &  62.6 \\
+        \hline
+        Bi-GRU & 97.1 & 82.2 & 41.7 & 3.8 & - & 58.5 & 44.5  &  89.2 & 82.8 & 54.9 &  48.4 & - &  59.1  \\
+        Bi-GRU$^*$  & 98.0 & 87.1 & 42.6 & 3.6 & 7.0 & 65.0 & 49.8 & 89.3 & 85.2 & 56.1 & 46.2 & 43.1 & 57.9  \\
+        \hline
+        CAML  &  98.6 &  89.5 &  53.9 & 8.8 & - &  70.9 &  56.1  & 90.9 & 87.5 &  61.4 &  53.2 & - &   60.9  \\
+        CAML$^*$ & 98.4 & 88.4 & 49.5 & 5.6 & 11.3 & 69.9 & 54.9 & 91.1 & 87.5 & 60.6 & 52.4 & 51.0 & 61.1  \\
+        \hline
+        MultiResCNN  &  98.6 &  91.0 & 55.2 & 8.6 & - & 73.4 & 58.4  &  93.8 &  89.9 &  67.0 &  60.6 & - &  64.1 \\
+        MultiResCNN$^*$  & 98.6 & 90.8 & 56.5 & 9.2 & 18.5 & 73.4 & 58.4 & 92.4 & 89.7 & 67.3 & 62.2 & 61.1 & 63.4  \\
+        \hline
+        LAAT  &  98.8 &  91.9 &  57.5 & 9.9 & - &  74.5 &  59.1  &  94.6 &  92.5 &  71.5 &  66.6 & - &  67.5 \\
+        LAAT$^*$ & 98.6 & 89.5 & 56.1 & 8.2 & 16.2 & 73.9 & 58.7 & 92.8 & 90.5 & 66.8 & 60.8 & 59.2 & 64.0  \\
+        \hline
+        PLM-ICD  &  98.9 & 92.6 &  59.8 & 10.4 & - &  77.1 &  61.3  & - & - & - & - & - & -  \\
+        PLM-ICD$^*$ & 98.8 & 92.3 & 58.9 & 11.1 & 22.8 & 75.7 & 60.5 & 93.8 & 91.7 & 70.5 & 66.3 & 65.4 & 65.7 \\
+        \bottomrule
+    \end{tabular}%
+    }
+\end{sidewaystable}
+
+\section{Results}
+\subsection{Reproduced results}
+
+In \cref{tab:reproduced_results}, we report the reproduced results on MIMIC-III \textit{full} and \textit{50} using hyperparameters as reported in the original papers.
+We list the original and corrected macro F1 score described in \cref{sec:metrics}.
+In most cases, our corrections doubled the macro F1 scores on MIMIC-III \textit{full}.
+The differences were smaller on MIMIC-III \textit{50} because all included codes are in the test set.
+
+\subsection{Revised comparison}
+
+\begin{sidewaystable}[t]
+    \centering
+    \caption[Results on the MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9} and MIMIC-IV  \textit{ICD-10} test sets presented as percentages.]{Results on the MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9} and MIMIC-IV  \textit{ICD-10} test sets presented as percentages.
+Micro F1 scores rank the table in ascending order.
+Each model was trained ten times with different seeds.
+We performed a McNemar's test with Bonferroni correction and found that all the models are significantly different ($p < 0.001$).}
+    \label{tab:fair_results}
+    \resizebox*{\textwidth}{!}{%
+    \begin{tabular}{llccccccccc}
+        \toprule
+        & & \multicolumn{5}{c}{Classification} & \multicolumn{4}{c}{Ranking} \\ 
+        \cmidrule(lr){3-7}\cmidrule(lr){8-11}
+         & & \multicolumn{2}{c}{AUC-ROC} & \multicolumn{2}{c}{F1} & EMR & \multicolumn{2}{c}{Precision@k} & R-precision & MAP \\
+         & & Micro & Macro & Micro & Macro &  & 8 & 15 &  &  \\
+        \midrule
+        \multirow{6}{*}{\shortstack[c]{MIMIC-III \\ clean}} & CNN & 97.1$\pm$0.0 & 88.1$\pm$0.2 & 48.0$\pm$0.3 & 9.9$\pm$0.4 & 0.1$\pm$0.0 & 61.6$\pm$0.2 & 46.6$\pm$0.1 & 49.1$\pm$0.2 & 50.6$\pm$0.2 \\
+        & Bi-GRU  & 97.8$\pm$0.1 & 91.1$\pm$0.2 & 49.7$\pm$0.4 & 12.2$\pm$0.2 & 0.1$\pm$0.0 & 62.8$\pm$0.4 & 47.6$\pm$0.4 & 50.1$\pm$0.4 & 52.1$\pm$0.4 \\
+        & CAML & 98.2$\pm$0.0 & 91.4$\pm$0.2 & 55.4$\pm$0.1 & 20.4$\pm$0.3 & 0.1$\pm$0.0 & 67.7$\pm$0.2 & 52.8$\pm$0.1 & 55.8$\pm$0.1 & 58.9$\pm$0.2 \\
+        & MultiResCNN & 98.5$\pm$0.0 & 93.1$\pm$0.3 & 56.4$\pm$0.2 & 22.9$\pm$0.6 & 0.1$\pm$0.0 & 68.5$\pm$0.2 & 53.5$\pm$0.1 & 56.7$\pm$0.2 & 59.9$\pm$0.3 \\
+        & LAAT & 98.6$\pm$0.1 & 94.0$\pm$0.3 & 57.8$\pm$0.2 & 22.6$\pm$0.6 & 0.2$\pm$0.1 & 70.1$\pm$0.2 & 54.8$\pm$0.2 & 58.0$\pm$0.2 & 61.7$\pm$0.3 \\
+        & PLM-ICD & \bfseries 98.9$\pm$0.0 & \bfseries 95.9$\pm$0.1 & \bfseries 59.6$\pm$0.2 & \bfseries 26.6$\pm$0.8 & \bfseries 0.4$\pm$0.0 & \bfseries 72.1$\pm$0.2 & \bfseries 56.5$\pm$0.1 & \bfseries 60.1$\pm$0.1 & \bfseries 64.6$\pm$0.2 \\
+        \hline
+        \multirow{6}{*}{\shortstack[c]{MIMIC-IV \\ ICD-9}} & CNN  & 98.1$\pm$0.1 & 89.4$\pm$0.5 & 52.4$\pm$0.1 & 12.6$\pm$0.4 & 0.6$\pm$0.0 & 61.3$\pm$0.1 & 45.6$\pm$0.0 & 52.9$\pm$0.1 & 55.2$\pm$0.1 \\
+        & Bi-GRU  & 98.8$\pm$0.0 & 93.8$\pm$0.1 & 55.5$\pm$0.1 & 16.6$\pm$0.2 & 0.7$\pm$0.0 & 64.1$\pm$0.1 & 47.8$\pm$0.1 & 55.8$\pm$0.1 & 58.9$\pm$0.1 \\
+        & CAML & 98.8$\pm$0.0 & 90.7$\pm$0.3 & 58.6$\pm$0.1 & 19.3$\pm$0.1 & 0.6$\pm$0.0 & 66.3$\pm$0.1 & 50.3$\pm$0.0 & 58.5$\pm$0.1 & 62.4$\pm$0.1 \\
+        & MultiResCNN  & 99.2$\pm$0.0 & 95.1$\pm$0.1 & 60.4$\pm$0.0 & 27.7$\pm$0.3 & 0.8$\pm$0.0 & 67.6$\pm$0.0 & 51.8$\pm$0.0 & 60.4$\pm$0.0 & 64.7$\pm$0.1 \\
+        & LAAT  & 99.3$\pm$0.0 & 96.0$\pm$0.3 & 61.7$\pm$0.1 & 26.4$\pm$0.9 & 0.9$\pm$0.0 & 68.9$\pm$0.1 & 52.7$\pm$0.1 & 61.7$\pm$0.2 & 66.3$\pm$0.2 \\
+        & PLM-ICD  & \bfseries 99.4$\pm$0.0 & \bfseries 97.2$\pm$0.2 & \bfseries 62.6$\pm$0.3 & \bfseries 29.8$\pm$1.0 & \bfseries 1.0$\pm$0.1 & \bfseries 70.0$\pm$0.2 & \bfseries 53.5$\pm$0.2 & \bfseries 62.7$\pm$0.3 & \bfseries 68.0$\pm$0.3 \\
+        \hline
+        \multirow{6}{*}{\shortstack[c]{MIMIC-IV \\ ICD-10}} & CNN & 97.5$\pm$0.1 & 87.9$\pm$0.4 & 47.2$\pm$0.6 & 8.0$\pm$0.4 & 0.3$\pm$0.0 & 60.3$\pm$0.1 & 45.7$\pm$0.1 & 47.3$\pm$0.2 & 48.2$\pm$0.2 \\
+        & Bi-GRU  & 98.3$\pm$0.0 & 92.4$\pm$0.2 & 50.1$\pm$0.2 & 10.6$\pm$0.4 & 0.3$\pm$0.0 & 62.6$\pm$0.2 & 47.7$\pm$0.2 & 49.6$\pm$0.1 & 51.1$\pm$0.2 \\
+        & CAML  & 98.5$\pm$0.0 & 91.1$\pm$0.1 & 55.4$\pm$0.2 & 16.0$\pm$0.3 & 0.3$\pm$0.0 & 66.8$\pm$0.2 & 52.2$\pm$0.1 & 54.5$\pm$0.2 & 57.4$\pm$0.2 \\
+        & MultiResCNN  & 99.0$\pm$0.0 & 94.5$\pm$0.2 & 56.9$\pm$0.1 & \bfseries 21.1$\pm$0.2 & \bfseries 0.4$\pm$0.0 & 67.8$\pm$0.1 & 53.5$\pm$0.1 & 56.1$\pm$0.1 & 59.3$\pm$0.2 \\
+        & LAAT  & 99.0$\pm$0.1 & 95.4$\pm$0.3 & 57.9$\pm$0.1 & 20.3$\pm$0.4 & \bfseries 0.4$\pm$0.0 & 68.9$\pm$0.1 & 54.3$\pm$0.1 & 57.2$\pm$0.1 & 60.6$\pm$0.2 \\
+        & PLM-ICD   & \bfseries 99.2$\pm$0.0 & \bfseries 96.6$\pm$0.2 & \bfseries 58.5$\pm$0.7 & \bfseries 21.1$\pm$2.3 & \bfseries 0.4$\pm$0.0 & \bfseries 69.9$\pm$0.6 & \bfseries 55.0$\pm$0.6 & \bfseries 57.9$\pm$0.8 & \bfseries 61.9$\pm$0.9 \\
+
+        \bottomrule
+    \end{tabular}%
+    }
+\end{sidewaystable}
+
+The results of our revised comparison on MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9}, and \textit{ICD-10} are shown in \cref{tab:fair_results}.
+Contrary to the originally reported results, Bi-GRU performs better than CNN in all metrics.
+Otherwise, the model performance ranking is unchanged from the original works.
+PLM-ICD outperformed the other models on all metrics and all datasets.
+The models previously reported as least performant improved the most.
+
+The ablation study results are shown in \cref{tab:ablation_study} for MIMIC-III \textit{clean}.
+Truncating the documents to 2,500 words instead of 4,000 had little impact on the performance.
+Using the hyperparameters from the original works degraded the performance substantially for CAML, Bi-GRU, and CNN but had a smaller effect on the other models.
+Not tuning the decision boundary had the largest negative effect on all models except MultiResCNN.
+In \cref{fig:threshold_tuning}, we plot the relationship between the decision boundary and F1 scores.
+LAAT and MultiResCNN perform similarly when using a decision boundary of 0.5.
+However, when tuning the decision boundary, LAAT outperforms MultiResCNN considerably.
+Similar results were obtained on the other datasets.
+
+\begin{sidewaystable}[t]
+    \centering
+    \caption[Ablation study on MIMIC-III \textit{clean}.]{Ablation study on MIMIC-III \textit{clean}. The numbers are the micro/macro F1 scores on the test set.}
+    \label{tab:ablation_study}
+    \resizebox*{\textwidth}{!}{%
+    \begin{tabular}{lcccccc}
+        \toprule
+         & PLM-ICD & LAAT & MultiResCNN & CAML & Bi-GRU & CNN \\
+        \midrule
+        Our result & 59.6/26.6 & 57.8/22.6 & 56.4/22.9 & 55.4/20.4 & 49.7/12.2 & 48.0/9.9 \\
+        \hline
+        Input length truncated at 2500 words & 59.4/26.2 & 57.6/22.3 & 56.0/23.2 & 54.8/19.7 & 49.4/12.0 & 47.9/9.8 \\
+        No decision boundary tuning & 58.7/23.0 & 56.2/19.0 & 56.2/22.6 & 53.3/17.1 & 45.3/8.1 & 43.8/7.0 \\
+        Original hyperparameters & 59.6/27.0 & 57.5/21.6 & 56.4/20.0 & 52.8/17.3 & 48.1/11.2 & 46.9/10.2 \\
+        \bottomrule
+        \end{tabular}%
+    }
+\end{sidewaystable}
+
+\begin{figure}[t]
+    \centering
+    \begin{subfigure}[b]{0.8\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{paper_automated/threshold_tuning_micro_full.png}
+        % \caption{}
+        % \label{fig:threshold_tuning_micro}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.8\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{paper_automated/threshold_tuning_macro_full.png}
+        % \caption{}
+        % \label{fig:threshold_tuning_macro}
+    \end{subfigure}%
+    \caption[The relationship between chosen threshold and F1 score of every reproduced model in \cref{tab:reproduced_results}]{The relationship between chosen threshold and F1 score of every reproduced model in \cref{tab:reproduced_results}.
+The left figure shows the micro F1 score, and the right shows the macro F1 score.
+The models were evaluated on MIMIC-III \textit{clean}.}
+    \label{fig:threshold_tuning}
+\end{figure}
+
+\subsection{Error analysis}
+
+\subsubsection{Amount of training data:}
+
+\cref{fig:train_size} shows the relationship between the number of training examples and the micro and macro F1 scores for all models.
+In most cases, increasing the training data had a larger effect on the macro F1 score than the micro F1 score, indicating more extensive improvements in rare codes than common codes.
+The curve for macro F1 is less smooth because the decision boundary was tuned on the micro F1 scores.
+
+\begin{figure}[t]
+    \centering
+    \begin{subfigure}[b]{0.8\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{paper_automated/train_size_f1_micro.png}
+        % \caption{}
+        % \label{fig:train_size_micro}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.8\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{paper_automated/train_size_f1_macro.png}
+        % \caption{}
+        % \label{fig:train_size_macro}
+    \end{subfigure}
+    \caption[The relationship between the number of training examples and F1 score on MIMIC-IV \textit{ICD-9}]{The relationship between the number of training examples and F1 score on MIMIC-IV \textit{ICD-9}.
+The left figure shows the F1 Micro score on the y-axis, while the right figure shows the F1 Macro score.}
+    \label{fig:train_size}
+\end{figure}
+
+\subsubsection{Document length:}\label{subsubsec:document-length}
+
+We plot the micro F1 score for all models as a function of the number of words per document in \cref{fig:text_length}. 
+We note that all models underperformed on documents with fewer than 1000 words. 
+By manual inspection, we found that most of these documents missed the information necessary to predict their labeled codes, leading to underperformance. 
+In \cref{tab:correlations}, we list the Pearson and Spearman correlations.
+We excluded documents shorter than 1000 words to avoid confounding with missing information and longer than 4000 words due to the truncation limit.
+We observe a very small negative correlation between document length and micro F1 which matches the downward trend in micro F1, starting from approximately 1000 words in \cref{fig:text_length}.
+Although document length may itself be the cause of the slightly lower performance for long documents, there may be other factors correlated with document length impacting performance, such as the number of codes per document and code frequency.
+As there are few long documents, the effect on average micro F1 for each dataset is negligible; hence, previous claims that long documents lead to poor performance in AMC could not be validated.
+Results on MIMIC-IV \textit{ICD-10} and MIMIC-III \textit{clean} were similar.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.8\linewidth]{paper_automated/f1_micro_vs_num_words.png}
+    \caption[Relationship between the lengths of the clinical notes and the micro F1 score for each model on MIMIC-IV \textit{ICD-9}]{Relationship between the lengths of the clinical notes and the micro F1 score for each model on MIMIC-IV \textit{ICD-9}.
+The vertical line indicates the maximum length of the notes after truncation.
+The histogram at the top visualizes the document length distribution.}
+    \label{fig:text_length}
+\end{figure}
+
+\begin{table}[t]
+    \centering
+    \caption[Correlation between the F1 score and the logarithm of code frequency and document length on MIMIC-IV \textit{ICD-9}.]{Correlation between the F1 score and the logarithm of code frequency and document length on MIMIC-IV \textit{ICD-9}.
+As discussed in \cref{subsubsec:document-length}, we only considered document lengths between 1000 and 4000 words.
+All correlations are statistically significant ($p<0.001$).}
+    \label{tab:correlations}
+    \begin{tabular}{lcccc}
+        \toprule
+        & \multicolumn{2}{c}{Code frequency} & \multicolumn{2}{c}{Document lengths}\\
+        \cmidrule(lr){2-3}\cmidrule(lr){4-5}
+        & Pearson & Spearman & Pearson & Spearman \\
+        \midrule
+        CNN  & 0.61 & 0.68 & -0.09 & -0.08 \\
+        Bi-GRU & 0.57 & 0.65 & -0.08 & -0.07 \\
+        CAML & 0.56 & 0.60 & -0.03 & -0.03   \\
+        MultiResCNN & 0.47 & 0.53 & -0.02 & -0.03  \\
+        LAAT & 0.52 & 0.57 & -0.02 & -0.02 \\
+        PLM-ICD & 0.48 & 0.52 & -0.02 & -0.02   \\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+\subsubsection{Code analysis:}
+
+\Cref{fig:icd9_vs_icd10} compares the best performing model, PLM-ICD, trained and evaluated on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}. Similar results were obtained on MIMIC-III \textit{clean}.
+The comparison shows the relationship between code frequencies in the training set and macro F1 scores.
+As shown in \cref{tab:fair_results}, all models perform worse on \textit{ICD-10} compared to \textit{ICD-9}.
+However, \cref{fig:icd9_vs_icd10} demonstrates that performance on codes with similar frequencies is comparable between the two splits.
+This suggests that the performance differences in \cref{tab:fair_results} are due to \textit{ICD-10} containing a higher fraction of rare codes as shown in \cref{fig:mimiciv_code_dist,fig:icd9_vs_icd10}.
+
+The Pearson and Spearman correlations between the logarithm of code frequency and F1 score are shown in \cref{tab:correlations} for MIMIC-IV \textit{ICD-9}.
+Similar correlations were observed for the other datasets.
+All the models show moderately high correlation confirming that performance on rare codes is generally lower than on common codes.
+To further our understanding of the problem, we computed the percentage of unique codes in each dataset that the models never predicted.
+As seen in \cref{tab:missed_classes}, no model correctly predicted more than 50\% of the ICD-10 codes.
+
+\Cref{fig:chapter_performance} shows the performance of PLM-ICD on each ICD-10 chapter---the top-most level in the tree-like hierarchy. 
+For our analysis, we limited the scope to only focus on the diagnosis codes. We also excluded codes with fewer than one hundred training examples to control for some chapters having many rare codes.
+
+Overall, PLM-ICD never correctly predicted 2,928 of the 5,794 ICD-10 diagnosis codes in our split.
+Of these codes, only 110 had over a hundred training examples, and 58 belong to only two of the 20 chapters in MIMIC-IV \textit{ICD-10}.
+Specifically, 45 belong to the chapter relating to ``external causes of morbidity" (Z00-Z99), while 13 relate to ``factors influencing health status and contact with health services" (V00-Y99).
+To further investigate why most non-predicted codes with more than 100 training examples belong to only two chapters, we manually inspected a selection of codes in these chapters, as described in the following.
+
+The Z68 category, part of the Z00-Z99 chapter, contains codes related to the patient's body mass index (BMI).
+Codes within this category occur more than 17,000 times in the MIMIC-IV training data, but PLM-ICD never predicts 20 out of the 26 codes of Z68.
+One possible hypothesis is that PLM-ICD struggles with extracting the BMI from the discharge summaries, as all digits have been removed in the pre-processing.
+We found several other codes containing digits in the code descriptions that the model failed to detect, e.g., ``Blood alcohol level of less than 20 mg/100 ml" (Y90.0), ``34 weeks gestation of pregnancy" (Z3A.34), and ``NIHSS score 15" (R29.715).
+These observations support our hypothesis that removing digits in the pre-processing makes certain codes challenging to predict.
+
+The Y92 category, part of the V00-Y99 chapter, contains codes related to the physical location of occurrence of the external cause.
+It is a large category of 246 unique codes occurring 27,870 times in the training set.
+The category is challenging due to locations being very specific.
+For instance, there are unique codes for whether an incident occurred on a tennis court, squash court, art gallery, or museum.
+We hypothesize that the level of detail in the discharge summaries does not always match the fine-grained code differences.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.8\linewidth]{paper_automated/code_count_vs_f1_all_datasets.png}
+    \caption[Relationship between the code frequencies in the training set and the macro F1 score for PLM-ICD on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}]{Relationship between the code frequencies in the training set and the macro F1 score for PLM-ICD on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}.
+The shaded area indicates the standard deviation of the score computed for codes within the bin.}
+    \label{fig:icd9_vs_icd10}
+\end{figure}
+
+There are ten different codes in MIMIC-IV \textit{ICD-10} relating to nicotine dependence and tobacco use.
+The three most common are Z87.891 (``Personal history of nicotine dependence"), F17.210 (``Nicotine dependence, cigarettes, uncomplicated"), and Z72.0 (``Tobacco use"), with 26,427, 8,486, and 1,914 training examples, respectively.
+Among these, Z72.0 was the third most common single code in the training set that PLM-ICD never predicted correctly.
+PLM-ICD achieved an F1 score of 53\% for Z87.891, 51\% for F17.210, and 0\% for Z72.0 and all other nicotine-related codes.
+These findings suggest that when there is a class imbalance among highly similar codes, PLM-ICD is strongly biased toward the most frequent ones.
+
+ \begin{table}[t]
+    \centering
+    \caption{Percentage of ICD diagnosis codes in the test set that the models never predicted correctly.}
+    \label{tab:missed_classes}
+    \begin{tabular}{lccc}
+        \toprule
+        & MIMIC-III & \multicolumn{2}{c}{MIMIC-IV}\\
+        \cmidrule(lr){2-2}\cmidrule(lr){3-4}
+        & \textit{clean} &  \textit{ICD-9} & \textit{ICD-10} \\
+        \midrule
+        CNN  & 68.2 & 61.5 & 72.0\\
+        Bi-GRU  & 65.0 & 54.3 & 67.1\\
+        CAML & 52.8 & 57.0 & 62.0  \\
+        MultiResCNN  & 48.8 & 40.3 & 53.5 \\
+        LAAT  & 50.4 & 43.6 & 55.0\\
+        PLM-ICD   & 44.3 & 39.3 & 51.8\\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.8\linewidth]{paper_automated/chapters.png}
+    \caption[Performance of PLM-ICD on ICD-10 chapters]{Performance of PLM-ICD on ICD-10 chapters.
+Only codes with more than a hundred occurrences in the MIIMC-IV ICD-10 training set were considered, leaving 20 chapters.
+We found Z00-Z99 and V00-Y99 to be the most challenging.}
+    \label{fig:chapter_performance}
+\end{figure}
+
+\section{Discussion}
+\subsection{Lessons learned}
+We found reproducing the results of CNN, Bi-GRU, CAML, and LAAT challenging.
+While we expected discrepancies due to random weight initializations and data shuffling, the differences from the original works exceeded our presuppositions.
+Our reproduced results were better than originally reported for Bi-GRU and CNN and worse for CAML and LAAT on most metrics.
+There have been multiple reports of issues in reproducing the results of \citet{mullenbachExplainablePredictionMedical2018}.\footnote{\url{https://github.com/jamesmullenbach/caml-mimic}} Additionally, most previous works did not report which version of MIMIC-III they used, and the code and hyperparameter configurations were not documented in detail.
+Therefore, we hypothesize that our results differ because previous works report incorrect hyperparameters or use an earlier version of MIMIC-III.
+
+We showed that models previously reported as low-performing underperformed partly due to a poor selection of hyperparameters and not tuning the decision boundary.
+In our revised comparison, we demonstrated that training the models using our setup decreased the difference between the best and worst micro F1 scores by $5.8$ percentage points.
+\citet{mullenbachExplainablePredictionMedical2018} concluded that CNN outperformed Bi-GRU.
+However, in our revised comparison, Bi-GRU outperformed CNN on all metrics on MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9}, and MIMIC-IV \textit{ICD-10}.
+
+Even though MultiResCNN contains more parameters than CAML, \citet{liICDCodingClinical2020} concluded that MultiResCNN was faster to train because it converged in fewer epochs.
+However, this was only true when using the original setup where CAML converged after 84 epochs.
+We found that when using a learning rate schedule and appropriate hyperparameters, it was possible to train all the models in 20 epochs without sacrificing performance.
+Therefore, with our setup, CAML was faster to train than MultiResCNN.
+
+We demonstrated that the macro F1 score had been underestimated in prior works due to the poorly sampled MIMIC-III \textit{full} split and the practice of setting the F1 score of all codes absent in the test set to 0.
+Since 54\% of the codes in MIMIC-III \textit{full} are missing in the test set, the maximum possible macro F1 score is 46\%.
+The previously highest reported macro F1 score on MIMIC-III \textit{full} is 12.7\% for PLM-ICD~\cite{kimReadAttendCode2021}.
+Using our corrected macro F1 score on the same split, PLM-ICD achieved a macro F1 score of 22.8\%.
+This large difference from previous state-of-the-art seems to indicate that all previous work on AMC used the sub-optimally calculated macro F1 score, including works not reproduced in this paper.
+Many studies use the macro F1 score to evaluate the ability of their models to predict rare codes \cite{kimReadAttendCode2021, yuanCodeSynonymsMatter2022}.
+If it has indeed been incorrectly calculated in these studies, some conclusions drawn in previous work regarding rare code prediction may have been misguided.
+
+Multiple studies mention lack of training data, rare codes, and long documents as the main challenges of AMC \cite{dongExplainableAutomatedCoding2021,feuchtDescriptionbasedLabelAttention2021,huangPLMICDAutomaticICD2022,jiDoesMagicBERT2021,liICDCodingClinical2020,liuEffectiveConvolutionalAttention2021,moonsComparisonDeepLearning2020,pascualBERTbasedAutomaticICD2021,tengReviewDeepNeural2022,tengExplainablePredictionMedical2020, vuLabelAttentionModel2020, venkateshAutomatingOverburdenedClinical2023}.
+In the error analysis, we aimed to validate or falsify these assumptions.
+We found that rare codes were challenging for all models and observed that more than half of all ICD-10 codes were never predicted correctly.
+Furthermore, in \cref{fig:train_size}, we showed that when adding more training data, most models see a greater performance improvement on rare codes than on common codes.
+These findings suggest that medical coding is fundamentally challenged by a lack of training data that, in turn, gives rise to many rare codes.
+We found that document length and model performance only exhibited a weak correlation.
+Specifically, the low number of very long documents was insufficient to affect the average performance on the dataset.
+
+\subsection{Future work}
+We recommend future work within AMC use our revised comparison method, including stratified sampled splits of MIMIC datasets, corrected evaluation metrics, hyperparameter search, and decision boundary tuning to avoid reporting suboptimal or biased results.
+Furthermore, for AMC to become a viable solution for ICD-10, future research should focus on improving performance on rare codes while, in the shorter term, developing methods to detect codes that are too challenging for automated coding and, therefore, should be coded manually.
+Finally, while PLM-ICD outperforms the other models in this paper, the improvements are limited compared to the effect of pre-training in other domains \cite{mohamed_selfsupervised_2022, linPretrainedTransformersText2021,baevski_wav2vec_2020,devlin_bert_2018,dosovitskiy_image_2021}.
+Notably, there have been several unsuccessful attempts at using pre-trained transformers for medical coding \cite{jiDoesMagicBERT2021,gaoLimitationsTransformersClinical2021,michalopoulosICDBigBirdContextualEmbedding2022,pascualBERTbasedAutomaticICD2021,zhangBERTXMLLargeScale2020}.
+In future work, we want to investigate why pre-trained transformers underperform in medical coding.
+
+\subsection{Limitations}
+We presented findings and analyses on MIMIC-III and MIMIC-IV. It is unclear how our findings generalize to medical coding in real-world settings.
+For instance, since MIMIC-III and IV contain data from the emergency department and ICU of a single hospital, the findings in this paper may not generalize to other departments or hospitals. 
+For instance, discharge summaries from outpatient care are often easier to code than summaries from inpatient care as they are shorter with fewer codes per document \cite{zhangBERTXMLLargeScale2020, liuEffectiveConvolutionalAttention2021, tsengAdministrativeCostsAssociated2018}.
+
+The medical code labeling of MIMIC is used as a gold standard in this paper.
+However, medical coding is error-prone, and, in many cases, deciding between certain codes can be a subjective matter \cite{nouraeiAuditNatureImpact2013, lloydPhysicianCodingErrors1985}.
+\citet{burnsSystematicReviewDischarge2012} systematically reviewed studies assessing the accuracy of human medical coders and found an overall median accuracy of 83.2\% (IQR: 67.3–92.1\%).
+\citet{searleExperimentalEvaluationDevelopment2020} investigated the quality of the human annotations in MIMIC-III and concluded that 35\% of the common codes were under-coded.
+Such errors and subjectivity in manual medical coding make model training and evaluation challenging and suggests that additional evaluation methods using, e.g., a human-in-the-loop, could be useful to increase the reliability of results.
+
+\section{Conclusion}
+In this paper, we first reproduced the results of selected state-of-the-art models focusing on unimodal models with publically available source code. 
+We found that model evaluation in original works was biased by an inappropriate formulation of the macro F1 score and treatment of missing classes in the test set.
+By fixing the macro F1 computation, we approximately doubled the macro F1 of the reproduced models on MIMIC-III \textit{full}.
+We introduced a new \textit{clean} split for MIMIC-III that contains all classes in the test set and performed a revised comparison of all models under the same training, evaluation, and experimental setup, including hyperparameter and decision boundary tuning.
+We observed a significant performance improvement for all models, with those previously reported as low-performing improving the most.
+We reported the first results of current state-of-the-art models on the newly released MIMIC-IV dataset \cite{johnsonMIMICIVFreelyAccessible2023, goldbergerPhysioBankPhysioToolkitPhysioNet2000} and provided splits for the \textit{ICD-9} and \textit{ICD-10} coded subsets using the same method as for MIMIC-III \textit{clean}.
+Through error analysis, we provided empirical evidence for multiple model weaknesses. Specifically, models underperform severely on rare codes 
+and, in contrast to previous claims, long documents only have a negligible negative performance impact. 
+We release our source code, model parameters, and the new MIMIC-III \textit{clean} and MIMIC-IV \textit{ICD-9} and \textit{ICD-10} splits.$^\text{1}$
+
+\section*{Acknowledgments}
+This research was partially funded by the Innovation Fund Denmark via the Industrial Ph.D. Program (grant no. 2050-00040B, 0153-00167B, 2051-00015B) and Academy of Finland (grant no. 322653).
+We thank Sotiris Lamprinidis for implementing our stratification algorithm and data preprocessing helper functions.
 

### 2023-08-11 chapters/paper_retrospective.tex
Add automated medical coding paper etc

@@
     \centering
-    \caption[Performance metrics for model stroke recognition.]{Performance metrics.
+    \caption[Performance metrics for stroke recognition.]{Performance metrics.
 Performance metrics for the model and call-takers on the test set.
@@
     \includegraphics[width=0.9\textwidth]{paper_retrospective/figure1.pdf}
-    \caption[Receiver operator characteristic (ROC) curve and PPV-sensitivity curve for model stroke recognition.]{Receiver operator characteristic (ROC) curve and PPV-sensitivity curve.
+    \caption[Receiver operator characteristic (ROC) curve and PPV-sensitivity curve for stroke recognition.]{Receiver operator characteristic (ROC) curve and PPV-sensitivity curve.
 Left, the ROC curve and, right, PPV-sensitivity curve (precision-recall curve).
@@
     \includegraphics[width=0.9\textwidth]{paper_retrospective/figure2.pdf}
-    \caption[Prediction confusion matrices for model stroke recognition.]{Prediction confusion matrices.
+    \caption[Prediction confusion matrices for stroke recognition.]{Prediction confusion matrices.
 Confusion matrices of predictions for call takers and the model on the test set.

### 2023-08-11 chapters/background.tex
Some work on the background

@@
 
-In the general public, the concept of uncertainty goes by many names and its meaning can be dependent on a specific context.
-However, in most quantitative scientific fields, there seems to be a consensus definition along the lines of the following \cite{hubbard_how_2014}:
+To the general public, the concept of uncertainty goes by many names, and its meaning can vary depending on the specific context.
+However, across most quantitative scientific fields, a consensus definition appears to align with the following \cite{hubbard_how_2014}:
 
@@
 
-While a definition of uncertainty made using only words easily begs philosophical inquisition, information theory provides a rigourous method for quantifying uncertainty \cite{kabir_neural_2018} by the language of probabilities.
+While a purely lexical definition of uncertainty might trigger a philosophical inquisition, information theory provides a mathematically rigorous method for quantifying uncertainty through the language of probabilities.
 In this context, uncertainty is understood as the information entropy of a random variable \cite{mackay_information_2003}.
@@
 \subsection{Information entropy}
-First characterised by Claude Shannon in 1948 \cite{shannon_mathematical_1948}, information entropy $H(X)$ is a measure of the average amount of information contained in a random variable $X$.
+First characterised by Claude Shannon in 1948 \cite{shannon_mathematical_1948}, information entropy, $H(X)$, is a measure of the average amount of information contained in a discrete random variable $X$.
 Shannon's definition follows from three fundamental axioms of information theory.
-Let $I_X(x)$ be the information carried by a specific outcome $x$ when sampling the random variable $X$ with probability mass function $p_X(x)$.
-Then, the axioms are as follows:
-\begin{enumerate}
-    \item  $I_X(x)$ is a monotonically decreasing function in $p_X(x)$. The more likely an outcome is, the less information it carries.
-    \item Outcomes that are certain to happen carry no information. $I_X(1) = 0$.
-    \item If an outcome $x_i$ is independent of another outcome $x_j$, then $I(x_i x_j) = I(x_i) + I(x_j)$.
-The information carried by independent events is the sum of the information carried by each event.
+Let $I_X(x)$ be the information carried by a specific outcome $x$ of sampling the random variable $X$ with probability mass function $p_X(x)$.
+Then, these axioms are as follows:
+\begin{enumerate}[label=(\Roman*)]
+    \item The more likely an outcome is, the less information it carries; $I_X(x)$ is a monotonically decreasing function in $p_X(x)$.
+    \item Outcomes that are certain to happen carry no information; $I_X(1) = 0$.
+    \item The total information carried by independent outcomes is the sum of the information carried by each outcome; $I(x_i, x_j) = I(x_i) + I(x_j)$ if $x_i$ and $x_j$ are independent. 
 \end{enumerate}
 
-From these axioms, Shannon derived the following expression for the information content of an event $x$ with probability $p_X(x)$:
+From these axioms, Shannon found that a suitable function for the \emph{information content} of an outcome $x$ with probability $p_X(x)$ is the logarithm of the inverse probability,
+%
 \begin{equation}
-    I_X(x) = -\log_b p_X(x)
+    I_X(x) = -\log_b p_X(x) \enspace .
 \end{equation}
+%
+Information content is said to be measured in units of bits when using a logarithm with base $b=2$, or nats if $b=e$.
 
-information entropy of a random variable $X$ with $n$ possible outcomes:
+To illustrate the \dots \lesstodo[inline]{Come up with a suitable example explaining the intuition behind information entropy.}
+
+The information entropy of a random variable $X$ is defined as the \emph{expected information content} of an outcome of $X$,
+%
 \begin{equation}
-    H(X) = -\sum_{i=1}^n p_i \log p_i
+    H(X) = \mathbb{E}_X\left[I_X(x)\right] = - \sum_{x\in X} p_X(x) \log p_X(x) \enspace .
 \end{equation}
+%
+Intuitively, the entropy $H(X)$ of a discrete random variable $X$ is a measure of the amount of uncertainty associated with the value of $X$ when only its distribution is known. \todo{Rewrite}
+
+\subsection{Continuous entropy}
+\lesstodo[inline]{Discuss differential entropy as a generalization to continuous case.}
+% Other important information theoretic quantities include Rényi entropy (a generalization of entropy), differential entropy (a generalization of quantities of information to continuous distributions), and the conditional mutual information.
+
+% \cite{kabir_neural_2018} Survey of uncertainty quantification in deep learning.
 

### 2023-08-11 chapters/conclusion.tex
Make project run in VS Code
Sync-up
Add review paper (WIP)
Add page estimate to conclusion
Some work on the background

@@
-%!TEX root = ../Thesis.tex
-\chapter{Conclusion}
-A conclusion with some citations \cite{kingma_auto-encoding_2014,rezende_stochastic_2014}.
+%!TEX root = ../thesis.tex
+
+\chapter[conclusions and outlook]{Conclusions and outlook}\label{chp:conclusion}
+
+% ~3 pages
+
+A conclusion with some abbreviations and citations, \gls{VAE} \cite{kingma_autoencoding_2014,rezende_stochastic_2014}.
+
+\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}

### 2023-08-11 chapters/introduction.tex
Some work on the background

@@
 Uncertainty is a fundamental part of human experience.
-Understanding and quantifying uncertainty is crucial in making informed decisions, drawing meaningful conclusions, and assessing the reliability of predictions made from observations.
-Yet, capturing it in a mathematical model is a difficult task.
+The ability to understand and quantify uncertainty is crucial in making informed decisions, drawing meaningful conclusions, and assessing the reliability of predictions made from observations.
+Yet, capturing this ability in a mathematical model is a difficult task.
 
-Uncertainty estimation has been a central topic in the field of statistics.
+Uncertainty estimation has long been a central topic in the field of statistics.
 It has traditionally been addressed through methods like confidence intervals, hypothesis testing, and Bayesian inference.
@@
 In speech recognition, knowing that a certain word was hard to understand for the model can help avoid misinterpretation of the transcribed text.
-
-Speech processing is
 
@@
 
-The research in this thesis covers a number of different modelling approaches and tasks, including
-
-The research done over the course of the project at hand covers a varied set of topics.
+The research done during this project covers a varied set of topics.
 One group of the produced studies is concerned with generative latent variable models and their applications to uncertainty estimation and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.

### 2023-08-11 chapters/background.tex
Minor revisions

@@
 
-To the general public, the concept of uncertainty goes by many names, and its meaning can vary depending on the specific context.
+In society, the concept of uncertainty goes by many names, and its meaning can vary depending on the specific context.
 However, across most quantitative scientific fields, a consensus definition appears to align with the following \cite{hubbard_how_2014}:
@@
 
-While a purely lexical definition of uncertainty might trigger a philosophical inquisition, information theory provides a mathematically rigorous method for quantifying uncertainty through the language of probabilities.
+While such a purely lexical definition of uncertainty might trigger a philosophical inquisition, information theory provides a mathematically rigorous method for quantifying uncertainty through the language of probabilities.
 In this context, uncertainty is understood as the information entropy of a random variable \cite{mackay_information_2003}.

### 2023-08-14 chapters/background.tex
Updates

@@
 \subsection{Information entropy}
-First characterised by Claude Shannon in 1948 \cite{shannon_mathematical_1948}, information entropy, $H(X)$, is a measure of the average amount of information contained in a discrete random variable $X$.
+First characterized by Claude Shannon in 1948 \cite{shannon_mathematical_1948}, information entropy, $H(X)$, is a measure of the average amount of information contained in a discrete random variable $X$.
 Shannon's definition follows from three fundamental axioms of information theory.
@@
     \item Outcomes that are certain to happen carry no information; $I_X(1) = 0$.
-    \item The total information carried by independent outcomes is the sum of the information carried by each outcome; $I(x_i, x_j) = I(x_i) + I(x_j)$ if $x_i$ and $x_j$ are independent. 
+    \item The joint information carried by independent outcomes is the sum of the information carried by each outcome; $I(x_i, x_j) = I(x_i) + I(x_j)$ if $x_i$ and $x_j$ are independent. 
 \end{enumerate}
@@
 
+\iffalse
+
 Modelling paradigms
@@
 Additionally, quantifying uncertainty in high-dimensional data or complex models with massive amounts of parameters requires careful consideration and efficient computational strategies.
+
+\fi

### 2023-08-14 chapters/conclusion.tex
Updates

@@
 
-A conclusion with some abbreviations and citations, \gls{VAE} \cite{kingma_autoencoding_2014,rezende_stochastic_2014}.
-
 \lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}
+Since the main focus within self-supervised learning has been on improving downstream task performance, very limited work, if any, has investigated self-supervised representations in terms of uncertainty estimation.
+However, in the context of medical applications where data can be abundant but labels are sparse, unsupervised uncertainty estimation is a very interesting direction for future work.

### 2023-08-14 chapters/discussion.tex
Add benchmarking paper
Add todo to discussion
Updates

@@
 %!TEX root = ../thesis.tex
+
 \chapter[discussion]{Discussion}\label{chp:discussion}
 
-\section{}
+% ~5 pages
+
+\section{\cref{chp:paper-hierarchical} revisited: \dots}
+
+\lesstodo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other).
+Include reference to \cite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
+
+\section{\cref{chp:paper-review} revisited: \dots}
+
+\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}

### 2023-08-14 chapters/introduction.tex
Minor revisions
Updates

@@
 
-The research done during this project covers a varied set of topics.
+\section{Machine learning and societal risk}
+\todo[inline]{Discuss the role of uncertainty aware models in the context of risk.}
+\cite{europeancommission_briefing_2021}
+
+\section{Motivation: Speech recognition}
+
+\section{Motivation: Automated medical coding}
+
+\section{Scope of the thesis}
+
+The complete set of research published as part of this project covers a varied set of topics.
 One group of the produced studies is concerned with generative latent variable models and their applications to uncertainty estimation and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
@@
 
-This first part of the thesis ties together these different studies by providing a high-level overview of the research topic and the main contributions of the thesis.
+This first part of the thesis ties together these different studies by providing a high-level overview of the research topic and summarizing the main contributions of the thesis.
 
-\subsection{Motivation: Speech recognition}
-
-\subsection{Motivation: Automated medical coding}
-
-\subsection{Challenges in uncertainty estimation}
+\iffalse
 
@@
 \subsection{Scope of the thesis}
-
-\iffalse
 

### 2023-08-14 chapters/main-contributions.tex
Updates

@@
 \chapter[main contributions]{Main contributions}\label{chp:main-contributions}
+\newcommand{\contribitem}[1]{{\vspace{1em}\noindent\scshape\bfseries{\color{dtured}\cref{#1}}\\\ucnameref{#1}\\}}
 
-As previously mentioned, the chapters of the thesis are self-contained studies and therefore detail their own contributions. 
+The chapters of the thesis are self-contained studies and therefore detail their own contributions. 
 They are however written without consideration to the other chapters. 
@@
 
-\newcommand{\contribitem}[1]{{\vspace{1em}\noindent\scshape\bfseries{\color{dtured}\cref{#1}}\\\ucnameref{#1}\\}}
+\contribitem{part:background}
+This is the part of the thesis you are currently reading.
+\Cref{chp:introduction} provided a general introduction to the thesis and gave motivating examples for speech recognition and assistance in medical diagnostics.
+\Cref{chp:technical-background} provided the necessary technical background for the rest of the thesis by introducing uncertainty as a concept in the context of information theory and probability theory and providing a brief overview of generative latent variable models.
+\lesstodo[inline]{Update the contributions of the technical background section accordingly.}
 
-\contribitem{part:background}
-This part of the thesis provides the necessary background for the rest of the thesis.
-
-\contribitem{part:unsupervised-learning}
+\contribitem{part:unsupervised-uncertainty-estimation}
+This part of the thesis is concerned with unsupervised uncertainty estimation and consists of three papers.
+The first two papers are concerned with out-of-distribution detection using generative models.
+The first paper proposes a new method for out-of-distribution detection using hierarchical latent variable models.
+The second paper builds on the first and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable, explicit likelihood generative model.
+The third paper proposes a new hierarchical latent variable model for speech
 
@@
 The paper provides evidence that the failure of deep generative models to assign lower likelihoods to data from outside the training distribution is due to overemphasis on low-level features that generalize between distributions.
-To solve this, the proposed method computes a likelihood-ratio score for OOD detection that requires data to be in-distribution across all feature-levels.
-The method is computationally efficient, scalable, fully unsupervised and achieves state-of-the-art performance on several image-based OOD detection benchmarks.
+To solve this, the proposed method computes a likelihood-ratio score for out-of-distribution detection that requires data to be in-distribution across all feature-levels.
+The proposed method is computationally efficient, fully unsupervised, and performs well on several image-based out-of-distribution detection benchmarks.
+Methods like this are important for the development of safe and reliable machine learning systems, especially in medical applications where the consequences of errors can be significant.
 
-\contribitem{chp:paper-benchmarking} This paper proposes a new hierarchical latent variable model for speech inspired by the Clockwork VAE \cite{saxena_clockwork_2021}.
-The model is benchmarked against other latent variable models and autoregressive models for speech.
-It is shown that a hierarchy of latent variables improves the likelihood of the model.
-Finally, the latent space of the models are analyzed in terms of phonetic content.
+\contribitem{chp:paper-modelagnostic}
+This paper builds on the previous paper and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable, explicit likelihood generative model.
+The method is based on combining a classical parametric test with the recently introduced typicality test using Fisher's method.
+We show that this leads to a more accurate out-of-distribution test.
+We also discuss the benefits of casting out-of-distribution detection as a statistical testing problem, for instance enabling false positive rate control which is valuable for practical out-of-distribution detection, especially in high-risk applications such as in medical devices or services.
 
-\contribitem{chp:paper-review} This paper provides a comprehensive review of self-supervised speech representation learning.
+\contribitem{chp:paper-benchmarking}
+This paper proposes a new hierarchical latent variable model for speech inspired by the Clockwork VAE \cite{saxena_clockwork_2021}.
+The model is benchmarked against other latent variable models and autoregressive models for speech abnd we show that using a hierarchy of latent variables improves the likelihood of the model.
+Finally, we show that the learned latent space captures information relevant for speech recognition such as phonetic content.
+
+\contribitem{part:self-supervised-speech-representation-learning}
+This part of the thesis is concerned with self-supervised speech representation learning and consists of a single paper that provides a substantial review of the field \cite{mohamed_selfsupervised_2022}.
+Self-supervised learning is a promising approach to speech representation learning that has shown a fast, wide and successful adoption across speech modelling over the course of the last few years. 
+As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self-supervised representations generally outperform representations learned via probabilistic methods such as those investigated in \cref{part:unsupervised-uncertainty-estimation}.
+In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self-supervised learning as an alternative paradigm to generative models for learning representations for unsupervised uncertainty estimation.
+We shall return to this in the discussion of \cref{chp:paper-review}.
+
+\todo[options]{Update this paragraph according to the discussion of self-supervised learning for uncertainty estimation}
+% \cite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.
+
+\contribitem{chp:paper-review}
+This paper provides a comprehensive review of self-supervised speech representation learning.
 Previous work is grouped into three main categories of methods: contrastive, autoregressive, and generative, and the methods are compared in terms of their training objectives, model architectures, and performance on downstream tasks.
 The review also provides a discussion of the current state of the field and promising future directions of research.
+
+An early version of this review paper \cite{borgholt_brief_2022} included a review of generative latent variable models for speech representation learning and a comparison between the two modelling paradigms.
+Due to the superior performance of self-supervised methods and a desire to focus on self-supervised methods, the paper in \cref{chp:paper-review} does not focus on generative latent variable models.
+The early version is included in \cref{app:paper-brief} for reference.
+
+\todo[inline]{Should we include the brief review paper as a chapter instead of the long review paper?}
+
+\contribitem{part:medical-applications}
+This part contains studies on machine learning methods applied for tasks in a medical setting.
+While uncertainty estimation is not a central theme to any of the two papers, the first paper performs a substantial evaluation of the explainability of the proposed model via an occlusion analysis on the text input.
+Later, in the discussion, we shall further consider uncertainty estimation in relation to these two papers.
+\lesstodo[inline]{Revise and add more details to this section.}
+
+Prior to this thesis, but relevant to mention in this section, we published work on question tracking for calls to emergency services using a multimodal approach that combines automatic transcripts with the raw audio \cite{havtorn_multiqt_2020}.
+
+\contribitem{chp:paper-retrospective}
+
+\contribitem{chp:paper-automated}

### 2023-08-14 chapters/main-contributions.tex
Update

@@
 
-\todo[options]{Update this paragraph according to the discussion of self-supervised learning for uncertainty estimation}
+\todo[inline]{Update this paragraph according to the discussion of self-supervised learning for uncertainty estimation}
 % \cite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.
@@
 Due to the superior performance of self-supervised methods and a desire to focus on self-supervised methods, the paper in \cref{chp:paper-review} does not focus on generative latent variable models.
-The early version is included in \cref{app:paper-brief} for reference.
+The early version is included in \cref{chp:appendix-paper-brief} for reference.
 
@@
 
-Prior to this thesis, but relevant to mention in this section, we published work on question tracking for calls to emergency services using a multimodal approach that combines automatic transcripts with the raw audio \cite{havtorn_multiqt_2020}.
+Prior to this thesis, we published related work on question tracking for calls to emergency services using a multimodal approach that combines automatic transcripts with the raw audio \cite{havtorn_multiqt_2020}.
 
 \contribitem{chp:paper-retrospective}
+Effective stroke treatment depends on recognition by call-takers at prehospital telehealth services.
+This paper aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.
+We used calls from Copenhagen during 2015 to 2020, to develop a machine learning-based classification pipeline.
+First, calls are transcribed by a speech recognition model and then categorised as stroke or non-stroke using a text classification model.
+On test data from 2021, call-takers achieved an overall sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%) while the machine learning framework performed significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
+This study provides preliminary evidence that a machine learning framework for recognizing stroke in prehospital medical helpline calls could become a supportive tool for call-takers, aiding in early and accurate stroke recognition.
 

### 2023-08-15 chapters/background.tex
Update
Add some more backgroundf

@@
 
-\section{Uncertainty}
+\section{Uncertainty and information}
 
@@
 However, across most quantitative scientific fields, a consensus definition appears to align with the following \cite{hubbard_how_2014}:
-
+%
 \begin{center}
@@
 \end{center}
+%
+While such a purely lexical definition of uncertainty might trigger a philosophical inquisition, it highlights an important connection between uncertainty, information and probability.
+When information is not complete, we can describe the state of the world only with uncertainty; we are simply not certain.
+In that sense, uncertainty is a word we use to refer to information that is missing.
+A natural way to describe a state of the world with uncertainty is to use probabilities.
+To our luck, information theory provides exactly that; a mathematically rigorous method for quantifying information through the language of probabilities.
+In this context, uncertainty can be understood as the information entropy of a random variable \cite{mackay_information_2003}.
 
-While such a purely lexical definition of uncertainty might trigger a philosophical inquisition, information theory provides a mathematically rigorous method for quantifying uncertainty through the language of probabilities.
-In this context, uncertainty is understood as the information entropy of a random variable \cite{mackay_information_2003}.
-
-\subsection{Information entropy}
+\subsection{Entropy for discrete random variables}
 First characterized by Claude Shannon in 1948 \cite{shannon_mathematical_1948}, information entropy, $H(X)$, is a measure of the average amount of information contained in a discrete random variable $X$.
 Shannon's definition follows from three fundamental axioms of information theory.
-Let $I_X(x)$ be the information carried by a specific outcome $x$ of sampling the random variable $X$ with probability mass function $p_X(x)$.
+Let $I_X(x)$ be the information carried by a specific outcome $x$ of sampling the discrete random variable $X$ with probability mass function $p_X(x)$.
 Then, these axioms are as follows:
@@
     \item The more likely an outcome is, the less information it carries; $I_X(x)$ is a monotonically decreasing function in $p_X(x)$.
-    \item Outcomes that are certain to happen carry no information; $I_X(1) = 0$.
-    \item The joint information carried by independent outcomes is the sum of the information carried by each outcome; $I(x_i, x_j) = I(x_i) + I(x_j)$ if $x_i$ and $x_j$ are independent. 
+    \item Outcomes that are certain to happen carry no information; if $p_X(x) = 1$ then $I_X(x) = 0$.
+    \item The joint information carried by independent outcomes is the sum of the information carried by each outcome; if $x_i$ and $x_j$ are independent then $I(x_i, x_j) = I(x_i) + I(x_j)$. 
 \end{enumerate}
 
-From these axioms, Shannon found that a suitable function for the \emph{information content} of an outcome $x$ with probability $p_X(x)$ is the logarithm of the inverse probability,
+From these axioms, Shannon found that the \emph{information content} $I_X(x)$ of an outcome $x$ with probability $p_X(x)$ can suitably be defined as the negative logarithm of the probability of the outcome,
 %
-\begin{equation}
+\begin{equation} \label{eq:information-content}
     I_X(x) = -\log_b p_X(x) \enspace .
@@
 %
-Information content is said to be measured in units of bits when using a logarithm with base $b=2$, or nats if $b=e$.
+% Information content is measured in units of bits when the logarithm base $b$ is $2$, or nats when $b=e$.
 
-To illustrate the \dots \lesstodo[inline]{Come up with a suitable example explaining the intuition behind information entropy.}
-
-The information entropy of a random variable $X$ is defined as the \emph{expected information content} of an outcome of $X$,
+The \emph{information entropy} of a random variable $X$ is defined as the \emph{expected information content} of an outcome of $X$,
 %
-\begin{equation}
+\begin{equation} \label{eq:information-entropy}
     H(X) = \mathbb{E}_X\left[I_X(x)\right] = - \sum_{x\in X} p_X(x) \log p_X(x) \enspace .
@@
 %
-Intuitively, the entropy $H(X)$ of a discrete random variable $X$ is a measure of the amount of uncertainty associated with the value of $X$ when only its distribution is known. \todo{Rewrite}
+This means that $H(X)$ measures the amount of information we can expect to gain from observing an outcome of $X$, when we know only its distribution $p_X(x)$.
 
-\subsection{Continuous entropy}
-\lesstodo[inline]{Discuss differential entropy as a generalization to continuous case.}
-% Other important information theoretic quantities include Rényi entropy (a generalization of entropy), differential entropy (a generalization of quantities of information to continuous distributions), and the conditional mutual information.
+\lesstodo[inline]{Come up with a suitable example explaining the intuition behind information entropy.}
+
+\subsection{Entropy for continuous random variables}
+To generalize the concept of information entropy to continuous random variables, Shannon originally replaced the sum over the probability mass function in \eqref{eq:information-entropy} with an integral over the probability density function, as suggested by the definition $H(X) = \mathbb{E}_X\left[I_X(x)\right]$.
+This leads to the definition of the information entropy of a continuous random variable $X$ as the \emph{differential entropy} \cite{shannon_mathematical_1948},
+%
+\begin{equation} \label{eq:differential-entropy}
+    H(X) = \mathbb{E}_X\left[I_X(x)\right] = - \int_{\mathcal{X}} p_X(x) \log p_X(x) \, dx \enspace ,
+\end{equation}
+%
+where $\mathcal{X}$ is the support of $X$.
+
+However, the differential entropy does not have several of the desired attributes of the discrete version; it can be negative, it is not invariant under a change of variables, and it is not dimensionally correct%
+\footnote{
+    Information entropy has the same dimensionality as information content which is typically measured in units of bits, depending on the base of the logarithm. 
+    However, the bit is not a base unit of the International System of Units (SI), nor is it an official derived unit.
+    A bit is quite simply a number, specifically 0 or 1. If something is said to have a size of 4 bits, it means that it can be described with 4 binary digits, i.e. 4 numbers that are either 0 or 1.
+
+    As such, it might be useful to think of information as a dimensionless quantity rather than a quantity with units.
+    Following this interpretation, the dimensionality of the discrete information entropy in \cref{eq:information-entropy} is also dimensionless, since a probability mass function is dimensionless. 
+    For probability density functions, which have dimensionality of the inverse of some quantity, e.g.
+inverse length, the dimensionality of differential entropy in \cref{eq:differential-entropy} becomes,
+    \begin{equation}
+        \text{dim} \left( H(X) \right) = \text{dim} \left( \int_{\mathcal{X}} p_X(x) \log p_X(x) \, dx \right) = \frac{1}{\text{length}} \log\left(\text{length}\right) \enspace .
+    \end{equation}
+    This is clearly not dimensionless and therefore does not correspond to the discrete information entropy.
+
+    % As a final twist on this story, note that the SI system does actually have a unit for an amount of substance: the mole \cite{internationalbureauofweightsandmeasures_systeme_2019}. 
+    % The mole is defined as the amount of substance that contains as many elementary entities as there are atoms in $\SI{12}{g}$ of carbon-12, that is $\SI{1}{mol} = 6.02214076\times10^{23}$.
+    % Taking the position that a mole is quite simply a number (not a number of particles) \cite{debievre_atomic_1992,baranski_atomic_2012}, and a bit is a number of elementary (information) entities, we could define a bit as a derived unit from the mole as $\SI{1}{bit} = \SI{1}{mol} / 6.02214076\times10^{23}$.
+}. 
+For these reasons, the differential entropy may not be a suitable measure of information, or uncertainty, for continuous random variables.
+
+Instead, it can be argued that the information entropy of a continuous distribution should be defined via the limiting density for discrete points \cite{jaynes_information_1957,jaynes_prior_1968}.
+Following the derivation of \textcite{jaynes_information_1957}, in the discrete entropy expression in \cref{eq:information-entropy}, we assume that the number of discrete points $x_i, i=1,2,\dots,N$, in the support of $X$ grows larger such that in the limit $N\rightarrow\infty$, the density of points becomes a continuous function $m(x)$ and we can replace the sum with an integral.
+Denoting the number of points satisfying $a<x_i<b$ by $N_{a,b}$, we have
+%
+\begin{equation}
+    \lim_{N\rightarrow\infty} \frac{N_{a,b}}{N} = \int_a^b m(x) \, dx \enspace .
+\end{equation}
+%
+If this limit behaves sufficiently well, adjacent differences $(x_{i+1} - x_{i})$ in the neighbourhood of any particular value of $x$ will also tend to zero, so that
+%
+\begin{equation} \label{eq:limiting-density-discrete-to-continuous}
+    \lim_{N\rightarrow\infty} \left[ N(x_{i+1} - x_{i}) \right] = \left[ m(x_{i}) \right]^{-1} \enspace .
+\end{equation}
+%
+In the limit, the value of the discrete probability mass function will become a continuous probability density function $w(x)$, according to the limiting form of
+%
+\begin{equation}
+    p_i = w(x_i)\left(x_{i+1} - x_{i}\right) \enspace ,
+\end{equation}
+%
+where $p_i$ is the value of the probability mass function at $x_i$ which via \cref{eq:limiting-density-discrete-to-continuous} becomes,
+%
+\begin{equation}
+    p_i = w(x_i)\left[Nm(x_i)\right]^{-1} \enspace .
+\end{equation}
+%
+Consequently, the discrete entropy expression in \cref{eq:information-entropy} becomes,
+%
+\begin{align}
+    H_{N,\text{discrete}}(X) &= - \sum_{i} p_i \log p_i \nonumber \\
+    &= - \sum_{i\in\mathcal{X}} w(x_i)\left[Nm(x_i)\right]^{-1} \log \left( w(x_i)\left[Nm(x_i)\right]^{-1} \right) \nonumber \\
+    &= - \sum_{i\in\mathcal{X}} w(x_i)\left[Nm(x_i)\right]^{-1} (\log w(x_i) - \log N - \log m(x_i)) \nonumber  \\
+    &= 1
+\end{align}
+
+%
+\begin{align}
+    % H_{N,\text{discrete}}(X) &= - \int_\mathcal{X} w(x)\left[Nm(x)\right]^{-1} \, dx \, \log \left[\frac{w(x)}{Nm(x)}\right] \nonumber \\
+    % &= - \int_\mathcal{X} w(x) \, dx \, \log \left[\frac{w(x)}{Nm(x)}\right]
+    H_{N,\text{discrete}}(X) &= - \int_\mathcal{X} w(x) \, dx \, \log \left[\frac{w(x)}{Nm(x)}\right] \nonumber \\
+\end{align}
+%
+and taking the limit $N\rightarrow\infty$,
+%
+\begin{equation}
+    \lim_{N\rightarrow\infty} H_{N,\text{discrete}}(X) = H_{N} = \log N - \int_\mathcal{X} w(x) \log \left[\frac{w(x)}{m(x)}\right] \enspace .
+\end{equation}
+%
+Note that since $\lim_{N\rightarrow\infty} \log N = \infty$, this procedure suggests that the entropy in the discrete sense of a continuous random variable should be infinite.
+
+% I apologize for any confusion. Let's clarify step 6, focusing on the transition from the discrete entropy expression to the continuous entropy expression. I'll provide a more detailed explanation.
+
+% Step 6: Expression for Discrete Entropy
+% - The discrete entropy $H_{N,\text{discrete}}(X)$ using the probability mass function $p_i$ is given by:
+%   \[ H_{N,\text{discrete}}(X) = - \sum_{i} p_i \log p_i. \]
+% - We transition from discrete to continuous terms, keeping in mind that $p_i = w(x_i)[Nm(x_i)]^{-1}$.
+% - So, the discrete entropy expression becomes:
+%   \[ H_{N,\text{discrete}}(X) = - \sum_{i} w(x_i)\left[Nm(x_i)\right]^{-1} \log \left( w(x_i)\left[Nm(x_i)\right]^{-1} \right). \]
+
+% Now, let's simplify the above expression by breaking it down step by step:
+
+% 1. **Simplify the term inside the logarithm:**
+%    \[ w(x_i)\left[Nm(x_i)\right]^{-1} = \frac{w(x_i)}{Nm(x_i)}. \]
+
+% 2. **Apply the logarithm property $\log(ab) = \log a + \log b$:**
+%    \[ \log \left( \frac{w(x_i)}{Nm(x_i)} \right) = \log w(x_i) - \log (Nm(x_i)). \]
+
+% 3. **Distribute the negative sign from the entropy definition:**
+%    \[ -w(x_i)\left[Nm(x_i)\right]^{-1} \log \left( \frac{w(x_i)}{Nm(x_i)} \right) = -w(x_i) \log w(x_i) + w(x_i) \log (Nm(x_i)). \]
+
+% 4. **Combine the logarithms using properties of logarithms:**
+%    \[ w(x_i) \log (Nm(x_i)) = \log \left( [Nm(x_i)]^{w(x_i)} \right) = \log \left( N^{w(x_i)} \cdot [m(x_i)]^{w(x_i)} \right). \]
+
+% 5. **Apply the property $\log(a^b) = b \log a$:**
+%    \[ \log \left( N^{w(x_i)} \cdot [m(x_i)]^{w(x_i)} \right) = w(x_i) \log N + w(x_i) \log m(x_i). \]
+
+% 6. **Substitute this result back into the expression:**
+%    \[ -w(x_i) \log w(x_i) + w(x_i) \log (Nm(x_i)) = -w(x_i) \log w(x_i) + w(x_i) \log N + w(x_i) \log m(x_i). \]
+
+% 7. **Use the sum notation to represent the terms:**
+%    \[ H_{N,\text{discrete}}(X) = - \sum_{i} \left( -w(x_i) \log w(x_i) + w(x_i) \log N + w(x_i) \log m(x_i) \right). \]
+
+% 8. **Combine the terms inside the sum:**
+%    \[ H_{N,\text{discrete}}(X) = \sum_{i} w(x_i) \log w(x_i) - \sum_{i} w(x_i) \log N - \sum_{i} w(x_i) \log m(x_i). \]
+
+% 9. **Recognize that the first term is the entropy of a discrete distribution:**
+%    \[ \sum_{i} w(x_i) \log w(x_i) = -\sum_{i} w(x_i) \log \frac{1}{w(x_i)} = -H(X). \]
+
+% 10. **Recognize that the second term is $\log N$ multiplied by the sum of probabilities, which is 1:**
+%    \[ \sum_{i} w(x_i) \log N = \log N \sum_{i} w(x_i) = \log N \cdot 1 = \log N. \]
+
+% 11. **Substitute back into the expression:**
+%    \[ H_{N,\text{discrete}}(X) = -H(X) - \log N - \sum_{i} w(x_i) \log m(x_i). \]
+
+% This should give you a more detailed understanding of how the transition from the discrete entropy expression to the continuous entropy expression is made while taking into account the continuous limit.
 

### 2023-08-25 chapters/background.tex
Write introduction, revise title and some formatting

@@
 
+While the papers in this thesis were written to be self-contained, page limits and the need to maintain a specific focus, forced us to limit the scope of technical background each paper provides.
+In this chapter, we present a more in-depth overview of the technical background relevant to the papers in this thesis.
+We start by introducing the concept of uncertainty, and how it relates to information and probability. We then introduce the task of out-of-distribution detection and a few ways to tackle it.
+Then, we present variational autoencoders, and how they can be used to learn representations of speech. Finally, we introduce the concept of self-supervised learning.
+
 \section{Uncertainty and information}
-
-% https://en.wikipedia.org/wiki/Uncertainty#cite_note-3
-% https://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=8371683
 
@@
 \subsection{Entropy for discrete random variables}
+
 First characterized by Claude Shannon in 1948 \cite{shannon_mathematical_1948}, information entropy, $H(X)$, is a measure of the average amount of information contained in a discrete random variable $X$.
@@
 %
-% Information content is measured in units of bits when the logarithm base $b$ is $2$, or nats when $b=e$.
 
@@
 This means that $H(X)$ measures the amount of information we can expect to gain from observing an outcome of $X$, when we know only its distribution $p_X(x)$.
+If all possible outcomes are equally likely, then $H(X)$ is maximized, and if only one outcome is possible, then $H(X)$ is minimized.
 
@@
 \subsection{Entropy for continuous random variables}
+
 To generalize the concept of information entropy to continuous random variables, Shannon originally replaced the sum over the probability mass function in \eqref{eq:information-entropy} with an integral over the probability density function, as suggested by the definition $H(X) = \mathbb{E}_X\left[I_X(x)\right]$.
-This leads to the definition of the information entropy of a continuous random variable $X$ as the \emph{differential entropy} \cite{shannon_mathematical_1948},
+This approach leads to the definition of the information entropy of a continuous random variable $X$ as,
 %
 \begin{equation} \label{eq:differential-entropy}
-    H(X) = \mathbb{E}_X\left[I_X(x)\right] = - \int_{\mathcal{X}} p_X(x) \log p_X(x) \, dx \enspace ,
+    H(X) = - \int_{\mathcal{X}} p_X(x) \log p_X(x) \, dx \enspace ,
 \end{equation}
 %
-where $\mathcal{X}$ is the support of $X$.
+which is called the \emph{differential entropy} \cite{shannon_mathematical_1948}. Here, $\mathcal{X}$ is the support of $X$.
 
-However, the differential entropy does not have several of the desired attributes of the discrete version; it can be negative, it is not invariant under a change of variables, and it is not dimensionally correct%
-\footnote{
+However, the differential entropy does not have several of the desirable attributes of the discrete version; it can be negative, it is not invariant under a change of variables, and it is not dimensionally correct%
+\footnote{\label{fn:dimensional-analysis-of-differential-entropy}
     Information entropy has the same dimensionality as information content which is typically measured in units of bits, depending on the base of the logarithm. 
     However, the bit is not a base unit of the International System of Units (SI), nor is it an official derived unit.
-    A bit is quite simply a number, specifically 0 or 1. If something is said to have a size of 4 bits, it means that it can be described with 4 binary digits, i.e. 4 numbers that are either 0 or 1.
+    A bit is quite simply a number, specifically 0 or 1. If something is said to have a size of 4 bits, it means that it can be described with 4 binary digits, i.e. 4 numbers, each either 0 or 1.
 
     As such, it might be useful to think of information as a dimensionless quantity rather than a quantity with units.
-    Following this interpretation, the dimensionality of the discrete information entropy in \cref{eq:information-entropy} is also dimensionless, since a probability mass function is dimensionless. 
-    For probability density functions, which have dimensionality of the inverse of some quantity, e.g.
-inverse length, the dimensionality of differential entropy in \cref{eq:differential-entropy} becomes,
-    \begin{equation}
-        \text{dim} \left( H(X) \right) = \text{dim} \left( \int_{\mathcal{X}} p_X(x) \log p_X(x) \, dx \right) = \frac{1}{\text{length}} \log\left(\text{length}\right) \enspace .
-    \end{equation}
-    This is clearly not dimensionless and therefore does not correspond to the discrete information entropy.
-
+    Following this interpretation, the discrete information entropy in \cref{eq:information-entropy} is also dimensionless, since a probability mass function is dimensionless. 
+    Since probability density functions have dimensionality of the inverse of some quantity, e.g.
+inverse length, the dimensionality of the continuous differential entropy in \cref{eq:differential-entropy} becomes,
+    \begin{equation*}
+        \text{dim} \left( H(X) \right) = \text{dim} \left( \int_{\mathcal{X}} p_X(x) \log p_X(x) \, dx \right) = \frac{1}{\text{length}} \log\left(\text{length}\right) \enspace ,
+    \end{equation*}
+    which is clearly not dimensionless and therefore does not correspond to the discrete information entropy.
     % As a final twist on this story, note that the SI system does actually have a unit for an amount of substance: the mole \cite{internationalbureauofweightsandmeasures_systeme_2019}. 
@@
 
-Instead, it can be argued that the information entropy of a continuous distribution should be defined via the limiting density for discrete points \cite{jaynes_information_1957,jaynes_prior_1968}.
-Following the derivation of \textcite{jaynes_information_1957}, in the discrete entropy expression in \cref{eq:information-entropy}, we assume that the number of discrete points $x_i, i=1,2,\dots,N$, in the support of $X$ grows larger such that in the limit $N\rightarrow\infty$, the density of points becomes a continuous function $m(x)$ and we can replace the sum with an integral.
+Instead, \textcite{jaynes_information_1957} argued that the information entropy of a continuous distribution should be defined as the limiting density of increasingly dense discrete distributions \cite{jaynes_prior_1968}.
+This argument leads to an expression for the information entropy of a continuous random variable $X$ as,
+%
+\begin{equation} \label{eq:corrected-differential-entropy-jaynes}
+    H_{N\rightarrow\infty}(X) = \lim_{N\rightarrow\infty} [H_{N}(X) - \log N] = - \int_\mathcal{X} p(x) \log \left(\frac{p(x)}{q(x)}\right) \enspace ,
+\end{equation}
+%
+where $H_N(X)$ is the limiting density for discrete points, $q(x)$ is a uniform density over the quantization of the continuous space, and we have subtracted a $\log N$ term that would go to infinity in the limit of infinite points.
+By doing this, the information entropy becomes dimensionally correct, and invariant under a change of variables%
+\footnote{\label{fn:change-of-variables-invariance-of-corrected-differential-entropy}
+    Under a change of independent variable from $x$ to $y(x)$, we have that
+    \begin{align*}
+        \widetilde{w}(y)\,dy = w(x)\,dx \enspace , \quad
+        \widetilde{q}(y)\,dy = q(x)\,dx \enspace .
+    \end{align*}
+    Plugging this into the differential entropy \cref{eq:corrected-differential-entropy-jaynes} we arrive at the original expression but now with $y$ as the independent variable,
+    \begin{equation*}
+        H_{N\rightarrow\infty}(y(x)) =  = - \int_\mathcal{Y} \widetilde{w}(y)\,\frac{dy}{dx} \log \frac{\widetilde{w}(y)\frac{dy}{dx}}{\widetilde{q}(y)\frac{dy}{dx}} \, dx = - \int_\mathcal{Y} w(y) \log \frac{w(y)}{q(y)} \, dy \enspace .
+    \end{equation*}
+}.
+
+The form on the right-hand side of \cref{eq:corrected-differential-entropy-jaynes} can be recognized as the negative Kullback-Leibler divergence of $p(x)$ to $q(x)$ \cite{kullback_information_1959},
+%
+\begin{equation} \label{eq:kullback-leibler-divergence}
+    D_{KL}(P\parallel Q) = \int_\mathcal{X} p(x) \log \left(\frac{p(x)}{q(x)}\right) \enspace .
+\end{equation}
+%
+The Kullback-Leibler divergence $D_{KL}(p\parallel q)$ is often interpreted as the amount of additional information required to represent events from distribution the $P$ with density $p(x)$ using a code optimized for distribution $Q$ with density $q(x)$.
+In other words, it measures how surprised one would be if they used distribution $Q$ to represent events from distribution $P$.
+The KL-divergence is also known as the \emph{relative entropy} of $P$ with respect to $Q$, or simply as the \emph{information gain}.
+
+Returning to Jaynes and \cref{eq:corrected-differential-entropy-jaynes}, the argument can be interpreted to say that the information entropy of a continuous random variable $X$ should be defined in relative terms as the expected difference in information entropy between its actual density and a uniform density.
+The Kullback-Leibler divergence as a measure of (relative) information entropy also provides a natural point of convergence of this section since it is equivalently defined for both discrete and continuous random variables.
+
+% In fact, Kullback initially motivated the divergence as an expected likelihood ratio \cref{kullback_information_1959}.
+
+\subsection{Two kinds of uncertainty}
+
+Uncertainty can be divided into two kinds: aleatoric and epistemic.
+
+Aleatoric uncertainty is defined as being inherent to the data, and is irreducible.
+
+Epistemic uncertainty is uncertainty due to lack of knowledge, and can be reduced with more data.
+
+\section{Unsupervised out-of-distribution detection}
+
+Unsupervised out-of-distribution detection (OOD) is the task of identifying data that is unlikely to originate from the distribution of the training data.
+
+% Out-of-distribution detection and anomaly detection are both related concepts in machine learning and deep learning, but they address slightly different problems and have distinct focuses.
+
+% Out-of-Distribution Detection (OOD Detection):
+% Out-of-distribution detection involves determining whether a given input or data point belongs to the same distribution as the training data.
+In other words, it helps identify whether an input is coming from a category or distribution that the model has never encountered during training.
+This is especially important for safety-critical applications where the model should not make predictions on inputs that are significantly different from what it has seen before.
+OOD detection aims to prevent the model from making confident predictions on unfamiliar inputs that might lead to incorrect results.
+% Anomaly Detection:
+% Anomaly detection, on the other hand, focuses on identifying rare and unusual instances within the data, regardless of whether they belong to the same distribution as the training data or not.
+Anomalies can be defined as data points that deviate significantly from the norm or expected behavior.
+Anomaly detection is used to find instances that are different from the majority and might indicate potential errors, fraud, faults, or other unusual occurrences.
+% In summary, the key differences are:
+
+% Focus:
+% OOD Detection: Focuses on detecting inputs that are coming from distributions that differ significantly from the training data distribution.
+% Anomaly Detection: Focuses on identifying rare and unusual instances, regardless of whether they come from a different distribution or not.
+% Purpose:
+% OOD Detection: Primarily used for safety and robustness, ensuring that the model doesn't make confident predictions on unfamiliar data.
+% Anomaly Detection: Used for various applications such as fraud detection, fault detection, and identifying outliers in the data.
+% Detection Approach:
+% OOD Detection: Often involves measuring the uncertainty of the model's predictions. High uncertainty indicates that the input might be out-of-distribution.
+% Anomaly Detection: Focuses on detecting data points that are statistically rare or far from the norm.
+% It's worth noting that while the concepts are distinct, they can sometimes overlap.
+An out-of-distribution data point could also be considered an anomaly, but not all anomalies are necessarily out-of-distribution.
+Anomaly detection methods might use OOD detection techniques as part of their process to flag anomalous instances that are not in line with the expected distribution.
+
+% Nice review:
+% https://openaccess.thecvf.com/content/CVPR2023W/VAND/papers/Graham_Denoising_Diffusion_Models_for_Out-of-Distribution_Detection_CVPRW_2023_paper.pdf
+
+% \cite{kabir_neural_2018} Survey of uncertainty quantification in deep learning.
+
+% Two kinds of uncertainty:\\
+% - Aleatoric: Uncertainty inherent to the data. Irreducible. "Known unknowns". \\
+% - Epistemic: Uncertainty due to lack of knowledge. Can be reduced with more data. "Unknown unknowns".\\
+
+% https://arxiv.org/pdf/2306.05674.pdf
+
+\section{Variational autoencoders}
+
+% Trouble in paradise: factoriing the ELBO to show that we minimize the mutual information between z and x. 
+% Use this as transition to self-supervised methods as alternative for representation learning: 
+% https://jmtomczak.github.io/blog/13/13_trouble_in_paradise.html
+
+\section{Speech representation learning}
+
+\iffalse
+
+% TODO Delete this probably
+Following the derivation of \textcite{jaynes_information_1957}, in the discrete entropy expression in \cref{eq:information-entropy}, we assume that the number of discrete points, $x_i, i=1,2,\dots,N$, in the support of $X$ grows larger such that in the limit $N\rightarrow\infty$, the density of points becomes a continuous function $m(x)$, and we can replace the sum with an integral.
 Denoting the number of points satisfying $a<x_i<b$ by $N_{a,b}$, we have
@@
 %
-If this limit behaves sufficiently well, adjacent differences $(x_{i+1} - x_{i})$ in the neighbourhood of any particular value of $x$ will also tend to zero, so that
+If this limit behaves sufficiently well, adjacent differences $(x_{i+1} - x_{i})$ in the neighborhood of any particular value of $x$ will also tend to zero, so that
 %
@@
 Consequently, the discrete entropy expression in \cref{eq:information-entropy} becomes,
-%
-\begin{align}
-    H_{N,\text{discrete}}(X) &= - \sum_{i} p_i \log p_i \nonumber \\
-    &= - \sum_{i\in\mathcal{X}} w(x_i)\left[Nm(x_i)\right]^{-1} \log \left( w(x_i)\left[Nm(x_i)\right]^{-1} \right) \nonumber \\
-    &= - \sum_{i\in\mathcal{X}} w(x_i)\left[Nm(x_i)\right]^{-1} (\log w(x_i) - \log N - \log m(x_i)) \nonumber  \\
-    &= 1
-\end{align}
-
-%
 \begin{align}
@@
 Note that since $\lim_{N\rightarrow\infty} \log N = \infty$, this procedure suggests that the entropy in the discrete sense of a continuous random variable should be infinite.
-
-% I apologize for any confusion. Let's clarify step 6, focusing on the transition from the discrete entropy expression to the continuous entropy expression. I'll provide a more detailed explanation.
-
-% Step 6: Expression for Discrete Entropy
-% - The discrete entropy $H_{N,\text{discrete}}(X)$ using the probability mass function $p_i$ is given by:
-%   \[ H_{N,\text{discrete}}(X) = - \sum_{i} p_i \log p_i. \]
-% - We transition from discrete to continuous terms, keeping in mind that $p_i = w(x_i)[Nm(x_i)]^{-1}$.
-% - So, the discrete entropy expression becomes:
-%   \[ H_{N,\text{discrete}}(X) = - \sum_{i} w(x_i)\left[Nm(x_i)\right]^{-1} \log \left( w(x_i)\left[Nm(x_i)\right]^{-1} \right). \]
-
-% Now, let's simplify the above expression by breaking it down step by step:
-
-% 1. **Simplify the term inside the logarithm:**
-%    \[ w(x_i)\left[Nm(x_i)\right]^{-1} = \frac{w(x_i)}{Nm(x_i)}. \]
-
-% 2. **Apply the logarithm property $\log(ab) = \log a + \log b$:**
-%    \[ \log \left( \frac{w(x_i)}{Nm(x_i)} \right) = \log w(x_i) - \log (Nm(x_i)). \]
-
-% 3. **Distribute the negative sign from the entropy definition:**
-%    \[ -w(x_i)\left[Nm(x_i)\right]^{-1} \log \left( \frac{w(x_i)}{Nm(x_i)} \right) = -w(x_i) \log w(x_i) + w(x_i) \log (Nm(x_i)). \]
-
-% 4. **Combine the logarithms using properties of logarithms:**
-%    \[ w(x_i) \log (Nm(x_i)) = \log \left( [Nm(x_i)]^{w(x_i)} \right) = \log \left( N^{w(x_i)} \cdot [m(x_i)]^{w(x_i)} \right). \]
-
-% 5. **Apply the property $\log(a^b) = b \log a$:**
-%    \[ \log \left( N^{w(x_i)} \cdot [m(x_i)]^{w(x_i)} \right) = w(x_i) \log N + w(x_i) \log m(x_i). \]
-
-% 6. **Substitute this result back into the expression:**
-%    \[ -w(x_i) \log w(x_i) + w(x_i) \log (Nm(x_i)) = -w(x_i) \log w(x_i) + w(x_i) \log N + w(x_i) \log m(x_i). \]
-
-% 7. **Use the sum notation to represent the terms:**
-%    \[ H_{N,\text{discrete}}(X) = - \sum_{i} \left( -w(x_i) \log w(x_i) + w(x_i) \log N + w(x_i) \log m(x_i) \right). \]
-
-% 8. **Combine the terms inside the sum:**
-%    \[ H_{N,\text{discrete}}(X) = \sum_{i} w(x_i) \log w(x_i) - \sum_{i} w(x_i) \log N - \sum_{i} w(x_i) \log m(x_i). \]
-
-% 9. **Recognize that the first term is the entropy of a discrete distribution:**
-%    \[ \sum_{i} w(x_i) \log w(x_i) = -\sum_{i} w(x_i) \log \frac{1}{w(x_i)} = -H(X). \]
-
-% 10. **Recognize that the second term is $\log N$ multiplied by the sum of probabilities, which is 1:**
-%    \[ \sum_{i} w(x_i) \log N = \log N \sum_{i} w(x_i) = \log N \cdot 1 = \log N. \]
-
-% 11. **Substitute back into the expression:**
-%    \[ H_{N,\text{discrete}}(X) = -H(X) - \log N - \sum_{i} w(x_i) \log m(x_i). \]
-
-% This should give you a more detailed understanding of how the transition from the discrete entropy expression to the continuous entropy expression is made while taking into account the continuous limit.
-
-% \cite{kabir_neural_2018} Survey of uncertainty quantification in deep learning.
-
-% Two kinds of uncertainty:\\
-% - Aleatoric: Uncertainty inherent to the data. Irreducible. "Known unknowns".\\
-% - Epistemic: Uncertainty due to lack of knowledge. Can be reduced with more data. "Unknown unknowns".\\
-
-% https://arxiv.org/pdf/2306.05674.pdf
-
-% We compare our framework with several major related lines of work.
-First, our work focuses on the quantification of epistemic uncertainty, which refers to the errors coming from the inadequacy of the model or data noises.
-This is different from aleatoric uncertainty, which refers to the intrinsic stochasticity of the problem [87, 91, 112, 59, 34], or predictive uncertainty which captures the sum of epistemic and aleatoric uncertainties (but not their dissection) [89, 92, 12, 3, 22].
-Regarding epistemic uncertainty, a related line of study is deep ensemble that aggregates predictions from multiple independent training replications [76, 70, 38, 8, 89].
-This approach, as we will make clear later, can reduce and potentially quantify procedural variability, but a naive use would require demanding retraining effort and does not address data variability.
-Another line is the Bayesian UQ approach on neural networks [41, 2].
-This regards network weights as parameters subject to common priors such as Gaussian.
-Because of the computation difficulties in exact inference, an array of studies investigate efficient approximate inference approaches to estimate the posteriors [40, 46, 16, 31, 30, 90, 74, 53].
-While powerful, these approaches nonetheless possess inference error that could be hard to quantify, and ultimately finding rigorous guarantees on the performance of these approximate posteriors remains open to our best knowledge.
-
-% 87: 
-% 91: 
-% 112:
-% 59: 
-% 34:
-
-% 89: 
-% 92: 
-% 12:
-% 3: 
-% 22:
-
-% 76: 
-% 70: 
-% 38: 
-% 8: 
-% 89:
-
-\section{Probabilistic latent variable models}
-
-\section{Speech representation learning}
-
-\iffalse
 
@@
 
-In the field of machine learning, uncertainty is often represented by a probability distribution.
-The most common approach is to use Bayesian methods, where uncertainty is captured by posterior distributions over model parameters and predictions.
-Bayesian neural networks, for instance, offer a powerful framework to model uncertainty in deep learning architectures.
-By incorporating prior beliefs and updating them based on observed data, these networks can produce probabilistic predictions that provide a measure of uncertainty.
-This is particularly useful in applications where high-confidence predictions are required, and the consequences of errors can be significant.
-
-Recently, another promising approach to uncertainty estimation in machine learning is the use of Monte Carlo Dropout \cite{gal_dropout_2016}.
-Monte Carlo Dropout leverages the idea of dropout regularization, originally employed during training to prevent overfitting.
-To form a prediction with associated uncertainty, Monte Carlo Dropout proposes to make multiple forward passes through the network while sampling different dropout masks for each one.
-This leads to obtaining a distribution of outputs for each input sample and the variance among these sampled predictions serves as a measure of uncertainty.
-Monte Carlo Dropout has shown remarkable success in various tasks such as image classification, object detection, and natural language processing.
-It is computationally efficient and can be easily incorporated into existing deep learning architectures, making it an attractive choice for uncertainty estimation.
-
-Furthermore, there has been a surge of interest in ensemble methods for uncertainty estimation.
-Ensembles combine the predictions of multiple models to obtain a more robust and calibrated uncertainty measure.
-Bagging and boosting techniques, which have been widely used in the field of statistics, have found their way into the realm of machine learning for uncertainty estimation as well.
-By training multiple models with different initializations or using diverse learning algorithms, ensembles can capture different sources of uncertainty, thereby providing a comprehensive assessment of the overall uncertainty in the predictions.
-
-Despite the progress made in uncertainty estimation for machine learning models, challenges still remain.
-One key concern is the interpretability of uncertainty measures.
-While probabilistic outputs are more informative than point estimates, effectively communicating uncertainty to end-users and decision-makers is not trivial.
-Developing visualization techniques and intuitive explanations for uncertainty is an ongoing area of research.
-Additionally, quantifying uncertainty in high-dimensional data or complex models with massive amounts of parameters requires careful consideration and efficient computational strategies.
-
 \fi

### 2023-08-25 chapters/introduction.tex
Write introduction, revise title and some formatting

@@
 % ~9 pages
+
+% Information overlead + administrative burden + lack of resources + rapid scientific progress in medicine and healthcare
+In an era where technology increasingly intertwines with healthcare, artificial intelligence is on the verge of becoming part of standard medical practice. 
+Growing administrative burdens, rapid scientific progress in medicine, and ageing populations have created a need to rethink how medical professionals are best enabled to successfully do their job. 
+The application of artificial intelligence in medical decision-making, particularly in diagnosing and recommending treatments, is likely to hold significant potential for improving patient outcomes by reducing the effect of these issues.
+However, this prospect also raises concerns related to the reliability of such models as their decisions impact critical aspects of human health and well-being.
+
+\todo[inline]{Can we find a better example than the Rhind Papyrus? It could be more on-point with decision-making and more well-known. Ideally, weave in an example of a failure of such technology.}
+The evolution of technology to support and automate human decision-making can be traced back to the development of algorithms and mathematical techniques by many ancient civilizations. 
+One example is the Egyptian Rhind Mathematical Papyrus, dated around 1550 BCE.
+It contains various mathematical techniques for calculations involving as multiplication, division, and fractions which were used for practical purposes like measuring land, constructing buildings, and sharing resources \cite{georges_universal_2001}.
+Concurrently, the Babylonians developed a sophisticated system of mathematics, including algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \cite{fowler_square_1998}.
+
+However, these new tools also brought new challenges.
+The Babylonian system of mathematics did not formalize the difference between rational and irrational numbers often leading to approximation errors and the common practice of approximating $\pi$ with 3.
+Besides system limitations, human error was also a factor in the application of the new tools.
+The Plimpton 322 clay tablet, for instance, contains several errors presumably made by a novice in Babylonian mathematics \cite{britton_plimpton_2011} (\cref{fig:plimpton_332}).
+These examples highlight the importance of understanding the limitations of technology, even in ancient times.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.80\textwidth]{plimpton_322.png}
+    \caption{Technology is difficult.
+The author of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grasp Babylonian mathematics.}
+    \label{fig:plimpton_332}
+\end{figure}
+
+While the quest to ease human decision-making has been ongoing for thousands of years, it was not until 1642 that the first mechanical calculator was successfully built by Blaise Pascal.
+And while the idea of a programmable computer with applications beyond pure arithmetic was first proposed by Ada Lovelace in 1843,
+it was not until the 1940s and 50s that % / 
+% it took one-hundred years before / 
+the first programmable, electronic, general-purpose digital computers were developed, starting with the ENIAC in 1945 \cite{georges_universal_2001} (\cref{fig:eniac_programmers}). 
+Initially used for computing artillery trajectories, as they became more accessible general-purpose computers were adopted widely across society for various applications in the sciences, in industry and in healthcare.
+The societal changes brought about by the widespread adoption of computer technology through the latter half of the 20th century have been so profound that some scholars refer to the current era as the information age \cite{georges_universal_2001, harari_sapiens_2011}.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.95\textwidth]{eniac_programmers.jpeg}
+    \caption{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the female programmers of the ENIAC. Photo: \copyright Corbis/ Getty Images.}
+    \label{fig:eniac_programmers}
+\end{figure}
+
+With the advent of the internet and the widespread adoption of digital technologies, the amount of data available to decision-makers has grown exponentially. 
+This has led to a surge of interest in the scientific field of machine learning, specifically deep learning, subfields of artificial intelligence that focus on developing algorithms that can learn from data and make predictions to guide decisions based on that knowledge.
+Such methods have set new standards for the performance of computer systems in various domains, including speech recognition, computer vision, and natural language processing \cite{lecun_deep_2015}.
+In recent years, machine learning has also been applied within healthcare, where it has shown promising results in tasks such as medical imaging \cite{lundervold_overview_2019}, drug discovery \cite{chen_rise_2018}, and clinical decision support \cite{cite15, cite14}.
+
+However, these technologies come with the same risks as did previous generations, and in some cases, they may exacerbate them.
+In 2018, the first incident of a self-driving car killing a pedestrian took place in Tempe, Arizona. 
+The car, an SUV operated by Uber, was driving in autonomous mode when it struck a pedestrian crossing the street, pushing her bicycle. 
+According to the investigation by the National Transportation Safety Board, the car's sensors detected the pedestrian $5.6$ seconds before the crash, but it did not correctly predict her path or reduce the SUV's speed (\cref{fig:uber_nhsa_accident}).
+Specifically, during those seconds, the system incorrectly classified the pedestrian more than ten times, first as a vehicle, then as an unknown object and ultimately as a bicyclist, each time changing, or resetting, the predicted path.
+About one second before the crash, the system determined that a collision was imminent, but the situation exceeded the constraints within which the autonomous driving system was allowed to operate, and the car's safety driver failed to intervene \cite{nationaltransportationsafetyboardnhsa_collision_2019}.
+This incident highlights the need for autonomous systems to be able to accurately assess their own uncertainty and act accordingly; requirements that European policymakers seek to impose on applications of machine learning via the European Artificial Inteliigence Act \cite{europeancommission_briefing_2021}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.95\textwidth]{uber_nhsa_accident.png}
+    \caption{A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
+The car's sensors detected the pedestrian $5.6$ seconds before the crash, but it did not correctly predict her path or reduce the SUV's speed \cite{nationaltransportationsafetyboardnhsa_collision_2019}.}
+    % \caption{Aerial view of crash location showing path of pedestrian as she attempted to cross N.
+Mill Avenue and movement and speed of SUV at three points before impact.
+Pedestrian's path shows her position from initial detection ($5.6$ seconds before impact) until impact; SUV's position is shown at corresponding times beginning $4.2$ seconds before impact \cite{nationaltransportationsafetyboardnhsa_collision_2019}.}
+    \label{fig:uber_nhsa_accident}
+\end{figure}
+
+\todo[inline]{Add something here about probability theory, Bayesian modelling etc. as a warm-up for potential solutions.}
+
+\section{Motivation: Medical conversations}
+
+This research project was formulated and carried out in collaboration with the Danish company Corti, which develops computer software for medical decision support.
+
+% The ancient Egyptians used mathematics for practical purposes like measuring land, constructing buildings, and managing resources.
+The Rhind Mathematical Papyrus, dated around 1550 BCE, contains various mathematical problems and techniques for calculations, such as multiplication, division, and calculations involving fractions.
+
+% The Babylonians, around 300 BCE, developed a sophisticated system of mathematics, including algorithms for solving quadratic equations and calculating areas of shapes.
+The Babylonians also used numerical methods for approximating square roots.
+
+% Greek mathematicians like Euclid (circa 300 BCE) formulated mathematical proofs and geometrical concepts that laid the foundation for algorithmic thinking.
+The "Euclidean algorithm," used to find the greatest common divisor of two numbers, is still taught in schools today.
+
+% Scholars in the Islamic world made significant contributions to mathematics.
+Al-Khwarizmi (circa 800 CE) wrote a book called "Al-Kitab al-Mukhtasar fi Hisab al-Jabr wal-Muqabala," from which the term "algebra" is derived.
+The book introduced systematic methods for solving linear and quadratic equations.
+
+\iffalse
 
@@
 For instance, in medical diagnosis, knowing the uncertainty in a model's prediction is crucial when the consequences of a false positive or false negative can be significant.
-In speech recognition, knowing that a certain word was hard to understand for the model can help avoid misinterpretation of the transcribed text.
+In speech recognition, knowing that a certain word was hard to transcribe for the model can help avoid misinterpretation of the transcribed text.
 
@@
 
+\subsection{Uncertainty in machine learning}
+
+\todo[inline]{Revise this section with less focus on individual methods and more focus on overall trends.}
+
+In the context of machine learning, uncertainty is often represented by a probability distribution.
+The most common approach is to use Bayesian methods, where uncertainty is captured by posterior distributions over model parameters and predictions \cite{gelman_bayesian_2013}.
+Bayesian neural networks, for instance, offer a powerful framework to model uncertainty in deep learning architectures \cite{neal_bayesian_1995}.
+By incorporating prior beliefs and updating them based on observed data, these networks can produce probabilistic predictions that provide a measure of uncertainty.
+This is particularly useful in applications where high-confidence predictions are required, and the consequences of errors can be significant.
+
+Recently, another interesting approach to uncertainty estimation in machine learning is the use of Monte Carlo Dropout \cite{gal_dropout_2016}.
+Monte Carlo Dropout leverages the idea of dropout regularization, originally employed during training to prevent overfitting.
+To form a prediction with associated uncertainty, Monte Carlo Dropout proposes to make multiple forward passes through the network while sampling different dropout masks for each one.
+This leads to obtaining a distribution of outputs for each input sample and the variance among these sampled predictions serves as a measure of uncertainty.
+It is computationally efficient and can be easily incorporated into existing deep learning architectures, making it an attractive choice for uncertainty estimation.
+
+Furthermore, there has been an interest in ensemble methods for uncertainty estimation \cite{lakshminarayanan_simple_2017}.
+Ensembles combine the predictions of multiple models to obtain a more robust and calibrated uncertainty measure.
+Bagging and boosting techniques, which have been widely used in the field of statistics, have found their way into the realm of machine learning for uncertainty estimation as well.
+By training multiple models with different initialization or using diverse learning algorithms, ensembles can capture different sources of uncertainty, thereby providing a comprehensive assessment of the overall uncertainty in the predictions.
+
+% We compare our framework with several major related lines of work.
+First, our work focuses on the quantification of epistemic uncertainty, which refers to the errors coming from the inadequacy of the model or data noises.
+This is different from aleatoric uncertainty, which refers to the intrinsic stochasticity of the problem [87, 91, 112, 59, 34], or predictive uncertainty which captures the sum of epistemic and aleatoric uncertainties (but not their dissection) [89, 92, 12, 3, 22].
+Regarding epistemic uncertainty, a related line of study is deep ensemble that aggregates predictions from multiple independent training replications [76, 70, 38, 8, 89].
+This approach, as we will make clear later, can reduce and potentially quantify procedural variability, but a naive use would require demanding retraining effort and does not address data variability.
+Another line is the Bayesian UQ approach on neural networks [41, 2].
+This regards network weights as parameters subject to common priors such as Gaussian.
+Because of the computation difficulties in exact inference, an array of studies investigate efficient approximate inference approaches to estimate the posteriors [40, 46, 16, 31, 30, 90, 74, 53].
+While powerful, these approaches nonetheless possess inference error that could be hard to quantify, and ultimately finding rigorous guarantees on the performance of these approximate posteriors remains open to our best knowledge.
+
+% 87: 
+% 91: 
+% 112:
+% 59: 
+% 34:
+
+% 89: 
+% 92: 
+% 12:
+% 3: 
+% 22:
+
+% 76: 
+% 70: 
+% 38: 
+% 8: 
+% 89: X
+
+% 41: 
+% 2:
+
+% 40:
+% 46:
+% 16:
+% 31:
+% 30:
+% 90:
+% 74:
+% 53:
+
 \section{Scope of the thesis}
@@
 One group of the produced studies is concerned with generative latent variable models and their applications to uncertainty estimation and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
-Another group of studies are concerned with self-supervised speech representation learning and automatic speech recognition \cite{borgholt_scaling_2021,borgholt_we_2021,mohamed_selfsupervised_2022,borgholt_brief_2022}.
+Another group of studies is concerned with self-supervised speech representation learning and automatic speech recognition \cite{borgholt_scaling_2021,borgholt_we_2021,mohamed_selfsupervised_2022,borgholt_brief_2022}.
 A final group deals with applications within the medical domain including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.
@@
 \end{enumerate}
-
-\begin{enumerate}
-    \item How can we learn representations that can form the basis of both prediction and uncertainty estimation?
-    \item 
-\end{enumerate}
-
-\subsection{Scope of the thesis}
 
@@
 \fi
+\fi

### 2023-08-25 chapters/paper_hierarchical.tex
Revise skipping
Add automated medical coding paper etc
Write introduction, revise title and some formatting

@@
+%!TEX root = ../thesis.tex
+
 \chapter[hierarchical vaes know what they don't know]{Hierarchical VAEs Know What They Don't Know}
 \label{chp:paper-hierarchical}
-% \ifthenelse{\equal{\skippaperbody}{true}}{}{
+\ifthenelse{\equal{\skippapers}{true}}{}{
 
@@
 If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \cref{fig_hierarchical:reconstructions-fashionmnist}-\ref{fig_hierarchical:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb_1)$ will be high for both in- and out-of-distribution data.
-Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb_1)] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) || p(\zb))$ from \cref{eq_hierarchical:elbo} relies on the KL-term for OOD detection.
-For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) || p(\zb))$ can be expressed by a hierarchical sum% over the hierarchy
+Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb_1)] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ from \cref{eq_hierarchical:elbo} relies on the KL-term for OOD detection.
+For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ can be expressed by a hierarchical sum% over the hierarchy
 % \begin{multline}
@@
 % \begin{multline}
-%     D_{\mathrm{KL}}( q(\zb|\xb) || p(\zb)) = \\
+%     D_{\mathrm{KL}}( q(\zb|\xb) \parallel p(\zb)) = \\
 %     \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
@@
 \begin{align}
-    \mathcal{L}      &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) || p_\thetab(\zb|\xb)\right), \label{eq_hierarchical:likelihoods-as-exact} \\ 
-    \mathcal{L}^{>k} &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k }|\zb_{>k}) q_\phib(\zb_{>k}|\xb) || p_\thetab(\zb|\xb)\right) \nonumber \ .
+    \mathcal{L}      &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) \parallel p_\thetab(\zb|\xb)\right), \label{eq_hierarchical:likelihoods-as-exact} \\ 
+    \mathcal{L}^{>k} &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k }|\zb_{>k}) q_\phib(\zb_{>k}|\xb) \parallel p_\thetab(\zb|\xb)\right) \nonumber \ .
 \end{align}
@@
 \begin{align}
-    LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) || p_\thetab(\zb|\xb)\right) \\
-                 &\quad + D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb) || p_\thetab(\zb|\xb)\right) \ . \notag
+    LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) \parallel p_\thetab(\zb|\xb)\right) \\
+                 &\quad + D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb) \parallel p_\thetab(\zb|\xb)\right) \ . \notag
 \end{align}\label{eq_hierarchical:llr-as-kls}
@@
 \begin{equation}
-    LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq k}|\zb_{>k}) q(\zb_{>k}|\xb) || p(\zb|\xb))
+    LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq k}|\zb_{>k}) q(\zb_{>k}|\xb) \parallel p(\zb|\xb))
 \end{equation}\label{eq_hierarchical:llr-as-kls-iwae-reduced}
@@
 % \begin{sidewaysfigure}
-\begin{figure}
+\begin{figure}[t]
     %\captionstyle{centerlast}
@@
 \subsection{Likelihood-ratio-based OOD detection}
-\begin{figure}
+\begin{figure}[t]
     \centering
@@
 
-% }
+}

### 2023-08-26 chapters/background.tex
Write more on the introduction

@@
 
-\chapter[technical background]{Technical Background}\label{chp:technical-background}
+\chapter[technical background]{Technical background}\label{chp:technical-background}
 
@@
 %
-While such a purely lexical definition of uncertainty might trigger a philosophical inquisition, it highlights an important connection between uncertainty, information and probability.
+While such a purely lexical definition of uncertainty might prompt a philosophical inquiry, it highlights an important connection between uncertainty, information and probability.
 When information is not complete, we can describe the state of the world only with uncertainty; we are simply not certain.
@@
     As such, it might be useful to think of information as a dimensionless quantity rather than a quantity with units.
-    Following this interpretation, the discrete information entropy in \cref{eq:information-entropy} is also dimensionless, since a probability mass function is dimensionless. 
+    Following this interpretation, the discrete information entropy in \cref{eq:information-entropy} is also dimensionless, because a probability mass function is dimensionless. 
     Since probability density functions have dimensionality of the inverse of some quantity, e.g.
@@
 \subsection{Two kinds of uncertainty}
+
+\todo[inline]{Consider not using the terms aleatoric and epistemic uncertainty.
+Instead use the terms ``uncertainty due to lack of knowledge (reducible)" and ``uncertainty inherent to the data (irreducible)".}
 

### 2023-08-26 chapters/introduction.tex
Write more on the introduction

@@
 In an era where technology increasingly intertwines with healthcare, artificial intelligence is on the verge of becoming part of standard medical practice. 
-Growing administrative burdens, rapid scientific progress in medicine, and ageing populations have created a need to rethink how medical professionals are best enabled to successfully do their job. 
+Growing administrative burdens, rapid scientific progress in medicine, and aging populations have created a need to rethink how medical professionals are best enabled to successfully do their job. 
 The application of artificial intelligence in medical decision-making, particularly in diagnosing and recommending treatments, is likely to hold significant potential for improving patient outcomes by reducing the effect of these issues.
@@
 
-\todo[inline]{Can we find a better example than the Rhind Papyrus? It could be more on-point with decision-making and more well-known. Ideally, weave in an example of a failure of such technology.}
-The evolution of technology to support and automate human decision-making can be traced back to the development of algorithms and mathematical techniques by many ancient civilizations. 
+\todo[inline]{Can we find a better example than the Rhind Papyrus, or an additional one?
+It could be more on-point with decision-making and more well-known.
+Ideally, weave in an example of a failure of such technology.}
+The evolution of technology to support and automate decision-making can be traced back to the development of algorithms and mathematical techniques by ancient civilizations. 
 One example is the Egyptian Rhind Mathematical Papyrus, dated around 1550 BCE.
 It contains various mathematical techniques for calculations involving as multiplication, division, and fractions which were used for practical purposes like measuring land, constructing buildings, and sharing resources \cite{georges_universal_2001}.
-Concurrently, the Babylonians developed a sophisticated system of mathematics, including algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \cite{fowler_square_1998}.
+Around the same time, the Babylonians had developed a sophisticated system of mathematics, including algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \cite{fowler_square_1998}.
 
-However, these new tools also brought new challenges.
+However, these new computing tools also brought new challenges.
 The Babylonian system of mathematics did not formalize the difference between rational and irrational numbers often leading to approximation errors and the common practice of approximating $\pi$ with 3.
-Besides system limitations, human error was also a factor in the application of the new tools.
-The Plimpton 322 clay tablet, for instance, contains several errors presumably made by a novice in Babylonian mathematics \cite{britton_plimpton_2011} (\cref{fig:plimpton_332}).
+Apart from limitations inherent to the tools, inadequate human proficiency in their use was also a source of error.
+The Plimpton 322 clay tablet, for instance, contains several errors presumably made by a novice in Babylonian mathematics (\cref{fig:plimpton_332}) \cite{britton_plimpton_2011}.
 These examples highlight the importance of understanding the limitations of technology, even in ancient times.
@@
     \caption{Technology is difficult.
-The author of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grasp Babylonian mathematics.}
+The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grasp Babylonian mathematics.}
     \label{fig:plimpton_332}
@@
 % it took one-hundred years before / 
-the first programmable, electronic, general-purpose digital computers were developed, starting with the ENIAC in 1945 \cite{georges_universal_2001} (\cref{fig:eniac_programmers}). 
-Initially used for computing artillery trajectories, as they became more accessible general-purpose computers were adopted widely across society for various applications in the sciences, in industry and in healthcare.
-The societal changes brought about by the widespread adoption of computer technology through the latter half of the 20th century have been so profound that some scholars refer to the current era as the information age \cite{georges_universal_2001, harari_sapiens_2011}.
+the first programmable, electronic, general-purpose digital computers were developed, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}) \cite{georges_universal_2001}. 
+As they became more accessible, general-purpose computers were adopted widely across society for various applications in science and industry, including healthcare.
+The societal changes brought about by the widespread adoption of computer technology through the latter half of the 20th century were so profound that some scholars refer to the current era as the information age \cite{georges_universal_2001, harari_sapiens_2011}.
+\todo[inline]{Expand on the history of modern computing.}
 
@@
 With the advent of the internet and the widespread adoption of digital technologies, the amount of data available to decision-makers has grown exponentially. 
-This has led to a surge of interest in the scientific field of machine learning, specifically deep learning, subfields of artificial intelligence that focus on developing algorithms that can learn from data and make predictions to guide decisions based on that knowledge.
+This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that focus on developing algorithms that can learn from data and make predictions to guide decisions based on that knowledge.
 Such methods have set new standards for the performance of computer systems in various domains, including speech recognition, computer vision, and natural language processing \cite{lecun_deep_2015}.
@@
 
-However, these technologies come with the same risks as did previous generations, and in some cases, they may exacerbate them.
+However, these technologies come with the same risks as previous generations did, and in some cases, they may exacerbate them.
+Inherent limitations of the technology itself, human error, and the complexity of the real world are all factors that can lead to failures.
+A pertinent example is the use of machine learning in autonomous driving systems:
 In 2018, the first incident of a self-driving car killing a pedestrian took place in Tempe, Arizona. 
@@
 About one second before the crash, the system determined that a collision was imminent, but the situation exceeded the constraints within which the autonomous driving system was allowed to operate, and the car's safety driver failed to intervene \cite{nationaltransportationsafetyboardnhsa_collision_2019}.
-This incident highlights the need for autonomous systems to be able to accurately assess their own uncertainty and act accordingly; requirements that European policymakers seek to impose on applications of machine learning via the European Artificial Inteliigence Act \cite{europeancommission_briefing_2021}.
 
-\begin{figure}
+\begin{figure}[t]
     \centering
@@
     \caption{A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
-The car's sensors detected the pedestrian $5.6$ seconds before the crash, but it did not correctly predict her path or reduce the SUV's speed \cite{nationaltransportationsafetyboardnhsa_collision_2019}.}
+The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize the uncertainty of multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \cite{nationaltransportationsafetyboardnhsa_collision_2019}.}
     % \caption{Aerial view of crash location showing path of pedestrian as she attempted to cross N.
@@
 
-\todo[inline]{Add something here about probability theory, Bayesian modelling etc. as a warm-up for potential solutions.}
+Though this incident is an extreme one, it highlights the challenges of applying machine learning in real-world scenarios and underlines the need for machine learning systems to be able to accurately assess their own uncertainty and operate accordingly.
+Recently, European policymakers proposed a set of requirements for applications of machine learning to ensure that they are safe, reliable, and trustworthy \cite{europeancommission_briefing_2021}.
+This AI Act groups applications into three categories based on the potential risk they pose to society: limited risk, high risk, and unacceptable risk.
+Limited risk applications, for instance within gaming, must meet transparency requirements, and unacceptable risk applications, such as public use of facial recognition, are banned.
+Applications within healthcare are considered high risk and must meet a strict set of requirements focusing on anti-discrimination and robustness.
+A recent study by the European Parliamentary Research Services concluded: \textit{"Future AI solutions for healthcare should be implemented by integrating uncertainty estimation, a relatively new field of research that aims to provide clinicians with clinically useful indications on the degree of confidence in AI predictions"} \cite{europeanparliament_artificial_2022}.
 
-\section{Motivation: Medical conversations}
+% \todo[inline]{Add something here about probability theory, Bayesian modelling etc. as a warm-up for potential solutions.}
 
-This research project was formulated and carried out in collaboration with the Danish company Corti, which develops computer software for medical decision support.
+\vspace{1em}
+\begin{center}
+\noindent\rule{0.2\textwidth}{0.5pt}
+\end{center}
+\vspace{1em}
+
+\noindent The research presented in this thesis will focus on the use of machine learning in relation to uncertainty estimation, speech processing, and medical conversations. 
+The research project of this thesis was formulated and carried out in collaboration with the Danish company Corti, which develops computer software for medical decision support. 
+In the rest of this chapter, we will provide a high-level introduction to these topics and discuss the motivation behind the research project.
+
+\section{Motivation and Corti's use-cases}
+
+At the start of this project in 2020, Corti's main product was a software system providing decision-support for emergency medical dispatchers.
+The system 
+% was in use in several countries, including the 1-1-2 line in Copenhagen, Denmark and the 9-1-1 line in Seattle, Washington in the United States and 
+provides an interface for navigating a graph-based protocol, defined by the local emergency medical services, which dispatchers use to triage patients during emergency calls.
+Calls are transcribed in real-time using customer-domain automatic speech recognition and the system makes suggestions for the dispatcher to use in the conversation with the caller, including notifications about forgotten protocol questions \cite{havtorn_multiqt_2020} and potential cases of cardiac arrests \cite{cite15, cite14}.
+The system logs the details about the call and the dispatcher's actions to be used for later review and quality assurance.
+
+\subsection{Stroke recognition in emergency calls}
+In 2021, Corti entered a research collaboration with the Capital Region of Denmark and the Copenhagen Emergency Services to develop a system for recognizing stroke cases in calls to the 1-1-2 emergency line and the 1813 medical helpline.
+Stroke is one of the biggest causes of disability and death worldwide \cite{cite1,cite2,cite3}.
+Effective treatment is highly time-sensitive \cite{cite4,cite5} and the most common gateway to specialized treatment and hospital admittance is through prehospital telehealth services like emergency medical call centers, nurse advice call lines, and out-of-hours health services \cite{cite6,cite7}.
+However, studies have found that approximately half of all patients with stroke do not receive the correct triage for their condition from call-takers \cite{cite10,cite11,cite12}.
+This may be due to the fact that stroke is a complex condition with a wide range of symptoms and that the symptoms may be difficult to recognize over the phone.
+Additionally, out of total call volume, stroke cases are relatively rare, occurring in only $0.25\%$ of all calls to the Copenhagen 1-1-2 emergency line in 2021, which makes it difficult for call-takers to gain practical experience in recognizing stroke cases.
+
+A reasonable machine learning pipeline for recognizing stroke might consist of an automatic speech recognition model to transcribe the conversation and a binary classifier to estimate the probability that what the transcript describes is a case of stroke.
+% The performance of such a system would depend on the quality of the automatic speech recognition model, the quality of the classifier, and the quality of the data used to train them. 
+In such a system, many factors can lead to classification errors, and when that happens, we would like the model to express high degree of uncertainty. 
+The automatic speech recognition model might be make an error in the transcription of particular word or phrase due to a noisy environment, overlapping, slurred or mumbling speech, or words outside the training distribution of the model.
+The classifier might misclassify a conversation due to ambiguous symptoms, a general lack of information, or errors of the speech recognizer propagated via the transcription. 
+How to best make such models accurately represent and estimate the uncertainty of given predictions, and how to use it to improve the value of such a system, especially in cooperation with the call taker is an open question.
+What is clear, however, is that the ability of such a system to accurately estimate the uncertainty of given predictions is crucial for its safe and reliable deployment in the real world.
+Overconfident predictions could lead to misdiagnoses and potentially fatal consequences for patients.
+Underconfident predictions could lead to unnecessary delays in treatment and increased costs for the healthcare system.
+
+% Corti's machine learning systems are not exempt from the challenges that come with real-world applications.
+The development is an ongoing effort to improve performance and to ensure robustness, reliability and compliance with current regulations.
+This thesis
+
+% One of the challenges faced by Corti's systems is the need to be able to recognize and handle rare events.
+% For instance, the system for emergency medical dispatchers is designed to recognize cases of cardiac arrest in emergency calls.
+% Cardiac arrest is a rare event, occurring in only $0.1\%$ of emergency calls \cite{cite14}, and the system must be able to recognize it in order to provide the dispatcher with the correct protocol to follow.
+% The system is trained on a dataset of calls where the dispatcher has indicated that the patient was in cardiac arrest, but the dataset is small and imbalanced, making it difficult to train a model that can reliably recognize cardiac arrest cases.
+% In addition, the system must be able to recognize cardiac arrest cases in calls from different countries, where the protocols and the way dispatchers handle calls may differ.
+% This is a challenging task as the system must be able to generalize to unseen data and adapt to different contexts.
+
+\subsection{Medical coding of clinical notes}
+
+During the course of the project presented in this thesis, Corti's product portfolio has expanded to include several new products. 
+One of these is a software system for medical coding of clinical notes. 
+When a patient is admitted to a hospital, the medical staff will write a clinical note describing the patient's diagnosis and the procedures they underwent, including drug prescriptions. 
+These notes are then used for billing and reimbursement purposes, as well as for research and quality assurance. 
+The clinical notes are written in natural language, usually following a certain structure, and a medical coder will assign a set of medical codes to the note based on its content. 
+This process is time-consuming and error-prone task due to the complexity of medical diagnoses and procedures and the immense number of medical codes in the most common standards such as the Internation Classification of Diseases (ICD) versions 10 and 11 with 55,000 and 85,000 codes, respectively \cite{worldhealthorganisationwho_international_2023}.
+
+\todo[inline]{
+Discuss the role of uncertainty estimation in medical coding. 
+Special challenges include significant class imbalance that cannot be removed by stratification since each example is associated with multiple labels.
+Mutually dependent classes also complicates uncertainty estimation. Some classes are mutually exclusive, while others are not. Some classes are easily confused with each other.
+}
+
+\section{Machine learning trustworthiness}
+
+Many modern machine learning models are deep neural networks with millions or even billions of parameters. 
+These models are often referred to as black-box models as it is difficult to understand how they arrive at their predictions.
+The lack of interpretability makes it difficult to assess the reliability of the model's predictions.
+Furthermore, a number of factors including the ability of deep neural networks to overfit, or even memorize, the training data \cite{arpit_closer_2017, burg_memorization_2021}, as well as the use of training objectives that are proxies for the evaluation metrics of interest, often lead to models that are miscalibrated; that is, the probability of a class predicted by the model does not reflect the true probability of the predicted class \cite{guo_calibration_2017, kull_temperature_2019}.
+Commonly, the predicted probabilities are too extreme, leading to overconfident predictions.
+This means a model will often assign a high probability to the predicted class, even when it is wrong.
+
+\todo[inline]{Expand in this discussion of overconfidence and calibration.
+E.g.: Calibration is fairly well understood in the context of classification, especially binary, but is harder in structured prediction such as automated speech recognition.}
+
+\section{Scope of the thesis}
+
+The complete set of research published as part of this project covers a varied set of topics.
+One group of the produced studies is concerned with generative latent variable models and their applications to uncertainty estimation and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
+Another group of studies is concerned with self-supervised speech representation learning and automatic speech recognition \cite{borgholt_scaling_2021,borgholt_we_2021,mohamed_selfsupervised_2022,borgholt_brief_2022}.
+A final group deals with applications within the medical domain including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.
+
+This first part of the thesis ties together these different studies by providing a high-level overview of the research topic and summarizing the main contributions of the thesis.
 
@@
 
-\section{Scope of the thesis}
-
-The complete set of research published as part of this project covers a varied set of topics.
-One group of the produced studies is concerned with generative latent variable models and their applications to uncertainty estimation and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
-Another group of studies is concerned with self-supervised speech representation learning and automatic speech recognition \cite{borgholt_scaling_2021,borgholt_we_2021,mohamed_selfsupervised_2022,borgholt_brief_2022}.
-A final group deals with applications within the medical domain including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.
-
-This first part of the thesis ties together these different studies by providing a high-level overview of the research topic and summarizing the main contributions of the thesis.
-
 \iffalse

### 2023-08-26 chapters/paper_retrospective.tex
Updates
Write more on the introduction

@@
 \label{chp:paper-retrospective}
-% \ifthenelse{\equal{\skippapers}{true}}{}{
+\ifthenelse{\equal{\skippapers}{true}}{}{
 
@@
 Stroke is a leading cause of disability and death worldwide \cite{cite1,cite2,cite3}.
-Effective treatment is time- sensitive, and an optimal outcome is more likely when treatment is administered within the first four and a half hours from stroke onset \cite{cite4,cite5}.
+Effective treatment is time-sensitive, and an optimal outcome is more likely when treatment is administered within the first four and a half hours from stroke onset \cite{cite4,cite5}.
 The gateway to ambulance transport and hospital admittance is through prehospital telehealth services, including emergency medical call centres, nurse advice call lines, and out-of-hours health services.
-In the pre-hospital setting, the use of Mobile Stroke Units has made it possible to deliver advanced treatment faster \cite{cite6,cite7}.
-As the Mobile Stroke Unit is only dispatched to patients with a suspected stroke, the impact of Mobile Stroke Units is directly influenced by accurate call-taker recognition of stroke.6,7 Call-takers who can rapidly and accurately recognize stroke are therefore crucial in facilitating prompt care in both pre-hospital and in-hospital settings.
+In the pre-hospital setting, the use of mobile stroke units has made it possible to deliver advanced treatment faster.
+As the mobile stroke unit is only dispatched to patients with a suspected stroke, the impact of mobile stroke units is directly influenced by accurate call-taker recognition of stroke \cite{cite6,cite7}.
+Call-takers who can rapidly and accurately recognize stroke are therefore crucial in facilitating prompt care in both pre-hospital and in-hospital settings.
 
@@
 Most initiatives aim to improve stroke recognition by call-takers via introducing more specific assessment tools \cite{cite8,cite9} or providing specialized training \cite{cite13}.
-Recent advances in machine learning technology might be applied to improve stroke recognition without requiring changes to the triaging approach, and machine learning aided identification of stroke has been suggested as a means of improving Mobile Stroke Unit effectiveness \cite{cite7}.
+Recent advances in machine learning technology might be applied to improve stroke recognition without requiring changes to the triaging approach, and machine learning aided identification of stroke has been suggested as a means of improving mobile stroke unit effectiveness \cite{cite7}.
 Real-time feedback from a machine learning model can improve the recognition of out-of-hospital cardiac arrest \cite{cite14,cite15}.
@@
 Moreover, in exploratory analyses, the model performed worse on these calls, which might be attributed to differences in population characteristics (\cref{tab_retrospective:table3}).
-Finally, the ground truth stroke labelling relied on the patient-reported time of onset being exact; however, estimating the accuracy of the timestamps in DanStroke was impossible..
+Finally, the ground truth stroke labelling relied on the patient-reported time of onset being exact; however, estimating the accuracy of the timestamps in DanStroke was impossible.
 
@@
 In conclusion, using the largest collection of audio calls from patients with stroke to date, we developed a machine learning framework that significantly outperformed human call-takers in stroke recognition in medical helpline calls.
-The frameworkcan assist human call-takers during medical helpline calls.
+The framework can assist human call-takers during medical helpline calls.
 Ideally, this would enable a higher recognition of patients with stroke in the prehospital setting, benefiting both patient outcomes and health service resource allocation.
@@
 HCC contributed to the conceptualisation, data curation, funding acquisition, methodology, project administration, supervision, and writing - review and editing.
-
 HCC and CK share last authorship.
-
-The three lead authors directly accessed and verified the underlying data reported in this manuscript. JW's affiliations are exclusively academic.
+The three lead authors directly accessed and verified the underlying data reported in this manuscript.
+JW's affiliations are exclusively academic.
 
@@
         \toprule
-                    & \multicolumn{2}{c}{Positive ranking score} & \multicolumn{2}{c}{Negative ranking score} \\
+                    & \multicolumn{2}{c|}{Positive ranking score} & \multicolumn{2}{c}{Negative ranking score} \\
         \midrule
-                    & \multicolumn{2}{c}{Stroke predictions, $D=1,897$} & \multicolumn{2}{c}{Non-stroke predictions, $D=342,133$} \\
+                    & \multicolumn{2}{c|}{Stroke predictions, $D=1,897$} & \multicolumn{2}{c}{Non-stroke predictions, $D=342,133$} \\
         \midrule

### 2023-08-26 appendices/paper_brief.tex
Some formatting and changes to labels

%!TEX root = ../thesis.tex

\chapter[a brief overview of unsupervised neural speech representation learning]{A Brief Overview of Unsupervised Neural Speech Representation Learning}
\label{app:paper-brief}
\ifthenelse{\equal{\skipappendices}{true}}{}{

\todo[inline]{Include brief overview of unsupervised neural speech representation learning in appendix.}

}

### 2023-08-26 chapters/main-contributions.tex
Some formatting and changes to labels

@@
 \Cref{chp:technical-background} provided the necessary technical background for the rest of the thesis by introducing uncertainty as a concept in the context of information theory and probability theory and providing a brief overview of generative latent variable models.
-\lesstodo[inline]{Update the contributions of the technical background section accordingly.}
+\todo[inline]{Update the contributions of the technical background section accordingly.}
 
@@
 Due to the superior performance of self-supervised methods and a desire to focus on self-supervised methods, the paper in \cref{chp:paper-review} does not focus on generative latent variable models.
-The early version is included in \cref{chp:appendix-paper-brief} for reference.
+The early version is included in \cref{app:paper-brief} for reference.
 
@@
 Later, in the discussion, we shall further consider uncertainty estimation in relation to these two papers.
-\lesstodo[inline]{Revise and add more details to this section.}
+\todo[inline]{Revise and add more details to this section.}
 

### 2023-08-26 chapters/paper_retrospective.tex
Some formatting and changes to labels

@@
 The relative performance gap between call-takers and the model was larger in our study than in the cardiac arrest study,15 which may affect the results of a potential RCT.
-To support future work and discussions beyond the scope of this study, \cref{app:retrospective} includes the results of a simulation of a live implementation where call-takers are assumed to follow a set of fixed rules based on the output of the machine learning framework.
+To support future work and discussions beyond the scope of this study, \cref{app:supplementary-retrospective} includes the results of a simulation of a live implementation where call-takers are assumed to follow a set of fixed rules based on the output of the machine learning framework.
 
@@
 We still reported the population characteristics and model performance of this group of calls to assess potential bias introduced by excluding them.
-A data-flow diagram is included in \cref{app:retrospective}.
+A data-flow diagram is included in \cref{app:supplementary-retrospective}.
 The resulting dataset is the largest dataset of audio files from stroke calls collected to date.
@@
 The final output score was used to classify whether the call concerned a stroke.
-The pipeline is illustrated in \cref{app:retrospective}.
+The pipeline is illustrated in \cref{app:supplementary-retrospective}.
 
@@
 As input for the classification model, each transcript was transformed into a fixed-size bag- of-words vector, which encoded the occurrence of word and character ($n$-grams) in a fixed vocabulary.
-The feature selection procedure is detailed in \cref{app:retrospective}.
+The feature selection procedure is detailed in \cref{app:supplementary-retrospective}.
 The model was constructed as an ensemble \cite{hansen_neural_1990} of five identical, independently trained models.
@@
 Subsequently, we defined the ensemble output score as the average of the centred output scores.
-The exact equations are provided in \cref{app:retrospective}.
+The exact equations are provided in \cref{app:supplementary-retrospective}.
 

### 2023-09-02 chapters/background.tex
Some formatting and changes to labels
Write on introduction

@@
 %
-While such a purely lexical definition of uncertainty might prompt a philosophical inquiry, it highlights an important connection between uncertainty, information and probability.
-When information is not complete, we can describe the state of the world only with uncertainty; we are simply not certain.
-In that sense, uncertainty is a word we use to refer to information that is missing.
-A natural way to describe a state of the world with uncertainty is to use probabilities.
-To our luck, information theory provides exactly that; a mathematically rigorous method for quantifying information through the language of probabilities.
-In this context, uncertainty can be understood as the information entropy of a random variable \cite{mackay_information_2003}.
+Although such a purely lexical definition of uncertainty might prompt a philosophical inquiry, it highlights an important connection between uncertainty, information and probability.
+When information is limited, we can describe the state of the world only with uncertainty; we are simply not certain.
+In that sense, uncertainty is a word we use to refer to the information that is missing, whether we know what is missing or not.
+A natural way to describe an uncertain world state is to use probabilities.
+To our luck, information theory provides exactly that; a mathematically rigorous method for quantifying information through the language of probability theory.
+In the following, we shall see that, in this context, uncertainty can be understood as the information entropy of a random variable \cite{mackay_information_2003}.
 
@@
 
-\lesstodo[inline]{Come up with a suitable example explaining the intuition behind information entropy.}
+\todo[inline]{Come up with a suitable example explaining the intuition behind information entropy.}
 

### 2023-09-02 chapters/introduction.tex
Write on introduction

@@
 \chapter[introduction]{Introduction}\label{chp:introduction}
+%s
+% SIZE: ~9 pages
+%
+% OUTLINE: Broad introduction to the topic.
+% - Motivate the reader.
+% - Why should they care about this topic? 
+% - What is the historical context of decision-support, what are the practical implications of it, and what happens when it fails.
 
-% ~9 pages
-
-% Information overlead + administrative burden + lack of resources + rapid scientific progress in medicine and healthcare
 In an era where technology increasingly intertwines with healthcare, artificial intelligence is on the verge of becoming part of standard medical practice. 
@@
 The Plimpton 322 clay tablet, for instance, contains several errors presumably made by a novice in Babylonian mathematics (\cref{fig:plimpton_332}) \cite{britton_plimpton_2011}.
-These examples highlight the importance of understanding the limitations of technology, even in ancient times.
+It is easy to forget that the challenges we face today in adopting new technologies are not new but have been faced by humans for thousands of years.
+Even in ancient times, it was important to understand the limitations of technology and the need for human expertise in using it.
 
@@
     \centering
-    \includegraphics[width=0.80\textwidth]{plimpton_322.png}
+    \includegraphics[width=0.90\textwidth]{plimpton_322.png}
     \caption{Technology is difficult.
-The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grasp Babylonian mathematics.}
+The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics.}
     \label{fig:plimpton_332}
@@
 As they became more accessible, general-purpose computers were adopted widely across society for various applications in science and industry, including healthcare.
-The societal changes brought about by the widespread adoption of computer technology through the latter half of the 20th century were so profound that some scholars refer to the current era as the information age \cite{georges_universal_2001, harari_sapiens_2011}.
+The societal changes brought about by the widespread adoption of computer technology through the latter half of the 20th century were so profound that some scholars refer to the era as the information age \cite{georges_universal_2001, harari_sapiens_2011}.
 \todo[inline]{Expand on the history of modern computing.}
@@
     \includegraphics[width=0.95\textwidth]{eniac_programmers.jpeg}
-    \caption{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the female programmers of the ENIAC. Photo: \copyright Corbis/ Getty Images.}
+    \caption{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the female programmers of the ENIAC.} %Photo: \copyright Corbis/ Getty Images.}
     \label{fig:eniac_programmers}
@@
 With the advent of the internet and the widespread adoption of digital technologies, the amount of data available to decision-makers has grown exponentially. 
-This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that focus on developing algorithms that can learn from data and make predictions to guide decisions based on that knowledge.
-Such methods have set new standards for the performance of computer systems in various domains, including speech recognition, computer vision, and natural language processing \cite{lecun_deep_2015}.
+This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that focuses on developing algorithms that can learn from data to recognize salient patterns and make predictions to guide decisions.
+Enabled by a massive increase in the power and availability of parallel computing resources, such methods have set new standards for the performance of computer systems in various domains, including speech recognition, computer vision, and natural language processing \cite{lecun_deep_2015}.
 In recent years, machine learning has also been applied within healthcare, where it has shown promising results in tasks such as medical imaging \cite{lundervold_overview_2019}, drug discovery \cite{chen_rise_2018}, and clinical decision support \cite{cite15, cite14}.
@@
 Inherent limitations of the technology itself, human error, and the complexity of the real world are all factors that can lead to failures.
-A pertinent example is the use of machine learning in autonomous driving systems:
+A pertinent example is the nascent use of machine learning in autonomous driving systems:
 In 2018, the first incident of a self-driving car killing a pedestrian took place in Tempe, Arizona. 
-The car, an SUV operated by Uber, was driving in autonomous mode when it struck a pedestrian crossing the street, pushing her bicycle. 
+The car, an SUV operated by Uber, was driving in autonomous mode when it struck a pedestrian crossing the street with her bicycle. 
 According to the investigation by the National Transportation Safety Board, the car's sensors detected the pedestrian $5.6$ seconds before the crash, but it did not correctly predict her path or reduce the SUV's speed (\cref{fig:uber_nhsa_accident}).
@@
 
-Though this incident is an extreme one, it highlights the challenges of applying machine learning in real-world scenarios and underlines the need for machine learning systems to be able to accurately assess their own uncertainty and operate accordingly.
+Though this incident is an extreme one, it highlights the challenges of applying machine learning in real-world scenarios and underlines the need for machine learning systems to be robust to adverse or difficult situations.
 Recently, European policymakers proposed a set of requirements for applications of machine learning to ensure that they are safe, reliable, and trustworthy \cite{europeancommission_briefing_2021}.
@@
 Limited risk applications, for instance within gaming, must meet transparency requirements, and unacceptable risk applications, such as public use of facial recognition, are banned.
-Applications within healthcare are considered high risk and must meet a strict set of requirements focusing on anti-discrimination and robustness.
-A recent study by the European Parliamentary Research Services concluded: \textit{"Future AI solutions for healthcare should be implemented by integrating uncertainty estimation, a relatively new field of research that aims to provide clinicians with clinically useful indications on the degree of confidence in AI predictions"} \cite{europeanparliament_artificial_2022}.
+High risk applications include those in healthcare and must meet a strict set of requirements focusing on anti-discrimination and robustness.
+%
+One of the key ways to improve the safety, reliability and trustworthiness of machine learning systems is to equip them with the ability accurately estimate the uncertainty of their predictions. 
+A model that can accurately estimate the uncertainty of its predictions can communicate to other systems or humans when it is unsure about a prediction, and so avoid making potentially catastrophic mistakes.
+Fittingly, a recent study by the European Parliamentary Research Services concluded:
+
+\begin{center}
+
+\textit{"Future AI solutions for healthcare should be implemented by integrating uncertainty estimation, a relatively new field of research that aims to provide clinicians with clinically useful indications on the degree of confidence in AI predictions"} \cite{europeanparliament_artificial_2022}.
+
+\end{center}
 
@@
 
+% OUTLINE: Primer overview of the thesis. What are the topics and what is the context of the research.
 \noindent The research presented in this thesis will focus on the use of machine learning in relation to uncertainty estimation, speech processing, and medical conversations. 
@@
 \section{Motivation and Corti's use-cases}
-
-At the start of this project in 2020, Corti's main product was a software system providing decision-support for emergency medical dispatchers.
-The system 
-% was in use in several countries, including the 1-1-2 line in Copenhagen, Denmark and the 9-1-1 line in Seattle, Washington in the United States and 
-provides an interface for navigating a graph-based protocol, defined by the local emergency medical services, which dispatchers use to triage patients during emergency calls.
+%
+% OUTLINE: Motivate the research project. 
+% - What is the context of the research and why is it important? 
+% - What are the current applications of machine learning in the domain (decision-support in healthcare)?
+% - What are the practical implications of the research? 
+% - What are the challenges and limitations of the current state-of-the-art?
+%
+At the start of this project in 2020, Corti's main product was a software system providing decision-support for emergency medical dispatchers which operate the 1-1-2 emergency line in Copenhagen, Denmark and the 9-1-1 emergency line in Seattle, Washington in the United States.
+The system, having been further developed, is still in use in 2023, and provides an interface for navigating a graph-based protocol, defined by the local emergency medical services, which dispatchers use to triage patients during emergency calls.
 Calls are transcribed in real-time using customer-domain automatic speech recognition and the system makes suggestions for the dispatcher to use in the conversation with the caller, including notifications about forgotten protocol questions \cite{havtorn_multiqt_2020} and potential cases of cardiac arrests \cite{cite15, cite14}.
@@
 \subsection{Stroke recognition in emergency calls}
+%
 In 2021, Corti entered a research collaboration with the Capital Region of Denmark and the Copenhagen Emergency Services to develop a system for recognizing stroke cases in calls to the 1-1-2 emergency line and the 1813 medical helpline.
-Stroke is one of the biggest causes of disability and death worldwide \cite{cite1,cite2,cite3}.
-Effective treatment is highly time-sensitive \cite{cite4,cite5} and the most common gateway to specialized treatment and hospital admittance is through prehospital telehealth services like emergency medical call centers, nurse advice call lines, and out-of-hours health services \cite{cite6,cite7}.
-However, studies have found that approximately half of all patients with stroke do not receive the correct triage for their condition from call-takers \cite{cite10,cite11,cite12}.
-This may be due to the fact that stroke is a complex condition with a wide range of symptoms and that the symptoms may be difficult to recognize over the phone.
-Additionally, out of total call volume, stroke cases are relatively rare, occurring in only $0.25\%$ of all calls to the Copenhagen 1-1-2 emergency line in 2021, which makes it difficult for call-takers to gain practical experience in recognizing stroke cases.
+Stroke is one of the biggest causes of disability and death worldwide \cite{cite1,cite2,cite3} and effective treatment is highly time-sensitive \cite{cite4,cite5}.
+The most common gateway to specialized treatment and hospital admittance is through prehospital telehealth services like emergency medical call centers, nurse advice call lines, and out-of-hours health services \cite{cite6,cite7}, however, studies have found that approximately half of all patients with stroke do not receive the correct triage for their condition from call-takers \cite{cite10,cite11,cite12}.
+This is likely due to the complexity of stroke cases which exhibit a wide range of symptoms that can be difficult to recognize over the phone.
+Additionally, out of total call volume, stroke cases are relatively rare, occurring in only $0.25\%$ of all calls to the Copenhagen 1-1-2 emergency line in 2021.
+This makes it difficult for call-takers to gain practical experience in recognizing stroke cases which exacerbates the general difficulty of recognizing the cases in the first place.
+Although several efforts have been made to improve the recognition of stroke cases in emergency calls \cite{cite13,cite14,cite15}, there is still a need for better tools to support call-takers in recognizing stroke cases.
 
-A reasonable machine learning pipeline for recognizing stroke might consist of an automatic speech recognition model to transcribe the conversation and a binary classifier to estimate the probability that what the transcript describes is a case of stroke.
+A reasonable machine learning pipeline to assist in recognizing stroke on calls to emergency services might consist of an automatic speech recognition model to transcribe the conversation and a binary classifier to estimate the probability that what the transcript describes is a case of stroke.
+Such a system might be trained on a dataset calls with verified stroke and non-stroke cases and evaluated against a held-out test set of calls where the call-taker has indicated that they suspect a stroke.
+Due to the low prevalence of stroke cases, the dataset would likely be imbalanced, and the number of stroke cases limited.
 % The performance of such a system would depend on the quality of the automatic speech recognition model, the quality of the classifier, and the quality of the data used to train them. 
 In such a system, many factors can lead to classification errors, and when that happens, we would like the model to express high degree of uncertainty. 
-The automatic speech recognition model might be make an error in the transcription of particular word or phrase due to a noisy environment, overlapping, slurred or mumbling speech, or words outside the training distribution of the model.
-The classifier might misclassify a conversation due to ambiguous symptoms, a general lack of information, or errors of the speech recognizer propagated via the transcription. 
-How to best make such models accurately represent and estimate the uncertainty of given predictions, and how to use it to improve the value of such a system, especially in cooperation with the call taker is an open question.
+The automatic speech recognition model might be make an error in the transcription of particular word or phrase due to a noisy environment, overlapping, slurred or mumbling speech, or words not in the model's training data.
+The classifier might misclassify a conversation due to medically ambiguous symptoms, a general lack of information in the conversation, or errors of the speech recognizer propagated via the transcription.
+How to best make such models accurately represent and estimate the uncertainty of given predictions, and how to use it to improve the value of such a system, especially in cooperation with the call taker, is an open question.
 What is clear, however, is that the ability of such a system to accurately estimate the uncertainty of given predictions is crucial for its safe and reliable deployment in the real world.
+Underconfident predictions could lead to unnecessary delays in treatment and increased costs for the healthcare system.
 Overconfident predictions could lead to misdiagnoses and potentially fatal consequences for patients.
-Underconfident predictions could lead to unnecessary delays in treatment and increased costs for the healthcare system.
-
-% Corti's machine learning systems are not exempt from the challenges that come with real-world applications.
-The development is an ongoing effort to improve performance and to ensure robustness, reliability and compliance with current regulations.
-This thesis
-
-% One of the challenges faced by Corti's systems is the need to be able to recognize and handle rare events.
-% For instance, the system for emergency medical dispatchers is designed to recognize cases of cardiac arrest in emergency calls.
-% Cardiac arrest is a rare event, occurring in only $0.1\%$ of emergency calls \cite{cite14}, and the system must be able to recognize it in order to provide the dispatcher with the correct protocol to follow.
-% The system is trained on a dataset of calls where the dispatcher has indicated that the patient was in cardiac arrest, but the dataset is small and imbalanced, making it difficult to train a model that can reliably recognize cardiac arrest cases.
-% In addition, the system must be able to recognize cardiac arrest cases in calls from different countries, where the protocols and the way dispatchers handle calls may differ.
-% This is a challenging task as the system must be able to generalize to unseen data and adapt to different contexts.
 
 \subsection{Medical coding of clinical notes}
-
+%
 During the course of the project presented in this thesis, Corti's product portfolio has expanded to include several new products. 
 One of these is a software system for medical coding of clinical notes. 
-When a patient is admitted to a hospital, the medical staff will write a clinical note describing the patient's diagnosis and the procedures they underwent, including drug prescriptions. 
+When a patient is admitted to a hospital, the medical staff will write a clinical note describing the patient's diagnosis and the procedures they underwent, which includes drug prescriptions. 
 These notes are then used for billing and reimbursement purposes, as well as for research and quality assurance. 
-The clinical notes are written in natural language, usually following a certain structure, and a medical coder will assign a set of medical codes to the note based on its content. 
-This process is time-consuming and error-prone task due to the complexity of medical diagnoses and procedures and the immense number of medical codes in the most common standards such as the Internation Classification of Diseases (ICD) versions 10 and 11 with 55,000 and 85,000 codes, respectively \cite{worldhealthorganisationwho_international_2023}.
+The clinical notes are written in natural language adhering to a certain structure, usually by a doctor. Later, a medical coder will assign a set of medical codes to the note based on its content. 
+The process of medical coding is time-consuming and error-prone due to the number and complexity of medical diagnoses and procedures.
+For instance, the widely used International Classification of Diseases (ICD) standard consist of 55,000 medical codes in version 10 and 85,000 in version 11 \cite{worldhealthorganisationwho_international_2023}.
+Additionally, a single medical note will usually contain several diagnoses and procedures which must all be inferred from the clinical note.
+In public benchmarking datasets such as MIMIC-III and IV the number is around 15.
+Official guidelines describe how to code specific diagnoses and procedures and combinations thereof \cite{centersformedicaremedicaidservicesus_icd10cm_2023}.
+For instance, a code may have several exclusion or inclusion criteria which define when certain other codes must be coded along with it, or when they must not.
 
@@
 Special challenges include significant class imbalance that cannot be removed by stratification since each example is associated with multiple labels.
-Mutually dependent classes also complicates uncertainty estimation. Some classes are mutually exclusive, while others are not. Some classes are easily confused with each other.
+Mutually dependent classes also complicates uncertainty estimation.
+Some classes are mutually exclusive, while others are not.
+Some classes are easily confused with each other.
+Uncertainty can be helpful in identifying code candidates as uncertain codes may need human review.
 }
 
-\section{Machine learning trustworthiness}
+Many codes are rarely used and datasets for medical coding usually have a considerable class-imbalance which can be impossible to exactly correct for by stratified sampling since each example is associated with multiple labels.
 
+\todo[inline]{Write paragraph on using machine learning for medical coding and the role of uncertainty estimation in that.}
+
+\section{Machine learning reliability}
+%
 Many modern machine learning models are deep neural networks with millions or even billions of parameters. 
@@
 The lack of interpretability makes it difficult to assess the reliability of the model's predictions. 
-Furthermore, a number of factors including the ability of deep neural networks to overfit, or even memorize, the training data \cite{arpit_closer_2017, burg_memorization_2021}, as well as the use of training objectives that are proxies for the evaluation metrics of interest, often lead to models that are miscalibrated; that is, the probability of a class predicted by the model does not reflect the true probability of the predicted class \cite{guo_calibration_2017, kull_temperature_2019}.
-Commonly, the predicted probabilities are too extreme, leading to overconfident predictions.
+Furthermore, a number of factors, including the ability of deep neural networks to overfit, or even memorize, the training data \cite{arpit_closer_2017, burg_memorization_2021}, and the use of training objectives that are proxies for the evaluation metrics of interest, lead to models that often are miscalibrated; that is, the probability of a class predicted by the model does not reflect the true probability of the predicted class under the circumstances \cite{guo_calibration_2017, kull_temperature_2019}.
+Usually, the predicted probabilities are too extreme which leads to overconfident predictions.
 This means a model will often assign a high probability to the predicted class, even when it is wrong.
+In practice, such a model can be hard to trust.
+No model is perfect, but the inability to judge the certainty of a model's predictions makes it difficult to know how to weigh the model's predictions in a decision-making process.
 
-\todo[inline]{Expand in this discussion of overconfidence and calibration.
-E.g.: Calibration is fairly well understood in the context of classification, especially binary, but is harder in structured prediction such as automated speech recognition.}
+A connection can be made back to the incident with the Uber self-driving car. 
+The multitude of rapid misclassifications made by the Uber self-driving car in Tempe is likely an indication that the model responsible for object-classification was overconfident in its predictions and unable to represent or communicate uncertainty to other systems in the car, or the safety driver.
+Instead, the model's predictions were taken at face value and subsystems in the car acted on them without appropriate consideration to their reliability.
+
+\subsection{Model calibration}
+
+\todo[inline]{
+    Expand in this discussion of overconfidence and calibration.
+E.g.: Calibration is fairly well understood in the context of classification, especially binary, but is harder in structured prediction such as automated speech recognition.
+Also:
+    Even if a model is well-calibrated, it can still be overconfident in wrong predictions - this is a problem in particular for rare events or adverse examples that were not presented to the model during training.
+Such examples are sometimes called out-of-distribution examples with reference to the training distribution.
+    For instance, a model trained to classify images of cats and dogs may have been well-calibrated, but if it is presented with an image of a horse, it has no option but to assign 100\% total probability to the cat and dog categories, even though that is clearly wrong.
+    Worse yet, we might expect the model to then at least assign 50\% probability to each of the cat and dog categories, but that behavior is not guaranteed.
+Specifically, since no horses were in the training data, the model will assign arbitrary probabilities to either class and so still risks being confidently wrong.
+}
+
+\subsection{Understanding uncertainty}
+% 
+% OUTLINE: Introduce aleatoric and epistemic uncertainty. 
+% - Discuss the distinction between them and how they relate to the problem of uncertainty estimation in machine learning.
+% - 
+% 
+%
+The distinction in uncertainty estimation between known and unknown unknowns is sometimes referred to in terms of \textit{aleatoric uncertainty} and \textit{epistemic uncertainty} \cite{kendall_what_2017}.
+Both categories of uncertainty are usually used to describe uncertainties in a model's predictions or the model itself including its parameters, but they differ in their sources.
+Aleatoric uncertainty (from latin (aleatorius) 'dice player' or 'gambler') refers to uncertainty in the data itself due to true randomness which is irreducible even with infinite data. 
+Epistemic uncertainty (from Greek (epistḗmē) 'knowledge') refers to uncertainty due to things one could in principle know but does not in practice. 
+% It could arise because a measurement is not accurate, because the model neglects certain effects, or because particular data have been deliberately hidden. 
+Epistemic uncertainty is sometimes referred to as \textit{systematic uncertainty} and aleatoric uncertainty as \textit{stochastic uncertainty} \cite{kendall_what_2017}. 
+%, or, \textit{reducible} and \textit{irreducible} uncertainty, respectively.
+This latter terminology is somewhat misleading since epistemic uncertainty might not be reducible by additional data collection if the model is not sufficiently expressive to represent the data distribution, and aleatoric uncertainty is not necessarily irreducible if we can reduce inherent noise by improving the instruments used to collect the data \cite{kendall_what_2017}.
+Although the distinction between aleatoric and epistemic uncertainty is important, it is not always useful.
+For instance, a model may be uncertain about a particular prediction due to any of these two kinds of uncertainty.
+For that reason, the quantity of interest is usually the ``total" uncertainty, often referred to as \textit{predictive uncertainty}, which is simply the combined epistemic and aleatoric uncertainties.
+
+An example of aleatoric uncertainty is the uncertainty in the transcription of a word due to a noisy environment, overlapping, slurred or mumbling speech.
+We can reduce this uncertainty by improving the automatic speech recognition model, but we cannot eliminate it completely.
+
+Words not present in the training data are arguably sources of epistemic uncertainty although a good speech recognition model may generalize well to unseen words if its spelling and pronunciation follow the same patterns as the words in the training data.
+
+A stronger example of epistemic uncertainty is the occurrence of truly out-of-distribution examples, such as the horse in the image classifier example above: inputs unlike anything the model has seen during training that potentially require outputs unavailable to the model.
+In high-dimensional data with a practically unlimited diversity in possible out-of-distribution examples, it is impossible to collect enough data to eliminate epistemic uncertainty.
+If we included some horses in the training data, the model would likely be able to recognize them, but it would still be unable to recognize giraffes, or zebras, or unicorns.
+Ultimately, any practical machine learning model will have at least some sources of epistemic uncertainty i.e. unknown sources of uncertainty.
+
+% Aleatory uncertainty is often pretty straightforward to model since it is available in the data.
+Epistemic is much harder since we don't know much about it, other than we have only examples of where it is low.
+
+% In real life applications, both kinds of uncertainties are present.
+Uncertainty quantification intends to explicitly express both types of uncertainty separately.
+The quantification for the aleatoric uncertainties can be relatively straightforward, where traditional (frequentist) probability is the most basic form.
+Techniques such as the Monte Carlo method are frequently used.
+A probability distribution can be represented by its moments (in the Gaussian case, the mean and covariance suffice, although, in general, even knowledge of all moments to arbitrarily high order still does not specify the distribution function uniquely), or more recently, by techniques such as Karhunen–Loève and polynomial chaos expansions.
+To evaluate epistemic uncertainties, the efforts are made to understand the (lack of) knowledge of the system, process or mechanism.
+Epistemic uncertainty is generally understood through the lens of Bayesian probability, where probabilities are interpreted as indicating how certain a rational person could be regarding a specific claim.
+
+% In mathematics, uncertainty is often characterized in terms of a probability distribution.
+From that perspective, epistemic uncertainty means not being certain what the relevant probability distribution is, and aleatoric uncertainty means not being certain what a random sample drawn from a probability distribution will be.
+
+\todo[inline]{Wrap this up.
+How to model aleatoric and epistemic uncertainty.
+Why do we care most about epistemic uncertainty (because we care about correctly estimating predictive uncertainty, and aleatoric is easy to model).}
 
 \section{Scope of the thesis}
-
-The complete set of research published as part of this project covers a varied set of topics.
+%
+% OUTLINE: 
+% - Tie together the different topics of the thesis and point briefly to the relevant papers.
+%
+The complete set of research published as part of this project covers a number of topics related to decision-support in medical conversations. 
 One group of the produced studies is concerned with generative latent variable models and their applications to uncertainty estimation and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
 Another group of studies is concerned with self-supervised speech representation learning and automatic speech recognition \cite{borgholt_scaling_2021,borgholt_we_2021,mohamed_selfsupervised_2022,borgholt_brief_2022}.
-A final group deals with applications within the medical domain including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.
+A final group deals with applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.
 
 This first part of the thesis ties together these different studies by providing a high-level overview of the research topic and summarizing the main contributions of the thesis.
-
-% The ancient Egyptians used mathematics for practical purposes like measuring land, constructing buildings, and managing resources.
-The Rhind Mathematical Papyrus, dated around 1550 BCE, contains various mathematical problems and techniques for calculations, such as multiplication, division, and calculations involving fractions.
-
-% The Babylonians, around 300 BCE, developed a sophisticated system of mathematics, including algorithms for solving quadratic equations and calculating areas of shapes.
-The Babylonians also used numerical methods for approximating square roots.
-
-% Greek mathematicians like Euclid (circa 300 BCE) formulated mathematical proofs and geometrical concepts that laid the foundation for algorithmic thinking.
-The "Euclidean algorithm," used to find the greatest common divisor of two numbers, is still taught in schools today.
-
-% Scholars in the Islamic world made significant contributions to mathematics.
-Al-Khwarizmi (circa 800 CE) wrote a book called "Al-Kitab al-Mukhtasar fi Hisab al-Jabr wal-Muqabala," from which the term "algebra" is derived.
-The book introduced systematic methods for solving linear and quadratic equations.
 

### 2023-09-02 chapters/background.tex
Minor updates

@@
 In this chapter, we present a more in-depth overview of the technical background relevant to the papers in this thesis.
-We start by introducing the concept of uncertainty, and how it relates to information and probability. We then introduce the task of out-of-distribution detection and a few ways to tackle it.
-Then, we present variational autoencoders, and how they can be used to learn representations of speech. Finally, we introduce the concept of self-supervised learning.
+We start by quantifying the concept of uncertainty, and how it relates to information and probability. We then introduce the task of out-of-distribution detection and a few ways to tackle it. 
+Then, we present variational autoencoders, and how they can be used to learn representations of speech.
+Finally, we introduce the concept of self-supervised learning and highlight important differences to variational autoencoders.
 

### 2023-09-02 chapters/discussion.tex
Update
Write introduction, revise title and some formatting
Write more on the introduction
Minor updates

@@
 
-\section{\cref{chp:paper-hierarchical} revisited: \dots}
+\section{\Cref{chp:paper-hierarchical} revisited: \dots}
 
@@
 Include reference to \cite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
+\lesstodo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimzing a mutual information term in the ELBO (derive this form).}
+\lesstodo[inline]{Discuss some recent work e.g. \cref{morningstar_density_2021}}
 
-\section{\cref{chp:paper-review} revisited: \dots}
+\section{\Cref{chp:paper-review} revisited: \dots}
 
 \lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}
+
+\section{\Cref{chp:paper-retrospective} revisited: \dots}
+
+\lesstodo[inline]{Discuss the calibration of the stroke model and report calibration curve.}
+\lesstodo[inline]{Report the results of calibrating the stroke model using e.g. Platt scaling or Isotonic scaling.} % https://scikit-learn.org/stable/modules/calibration.html

### 2023-09-02 chapters/introduction.tex
Minor updates

@@
 Both categories of uncertainty are usually used to describe uncertainties in a model's predictions or the model itself including its parameters, but they differ in their sources.
-Aleatoric uncertainty (from latin (aleatorius) 'dice player' or 'gambler') refers to uncertainty in the data itself due to true randomness which is irreducible even with infinite data. 
+Aleatoric uncertainty (from Latin (aleatorius) 'dice player' or 'gambler') refers to uncertainty in the data itself due to true randomness which is irreducible even with infinite data. 
 Epistemic uncertainty (from Greek (epistḗmē) 'knowledge') refers to uncertainty due to things one could in principle know but does not in practice. 

### 2023-09-10 chapters/background.tex
First complete introduction

@@
 
-The \emph{information entropy} of a random variable $X$ is defined as the \emph{expected information content} of an outcome of $X$,
+The \emph{information entropy of a discrete random variable} $X$ is defined as the \emph{expected information content} of an outcome of $X$,
 %
@@
 To generalize the concept of information entropy to continuous random variables, Shannon originally replaced the sum over the probability mass function in \eqref{eq:information-entropy} with an integral over the probability density function, as suggested by the definition $H(X) = \mathbb{E}_X\left[I_X(x)\right]$.
-This approach leads to the definition of the information entropy of a continuous random variable $X$ as,
+This approach leads to the definition of the \emph{information entropy of a continuous random variable} $X$ as,
 %
@@
 
-Instead, \textcite{jaynes_information_1957} argued that the information entropy of a continuous distribution should be defined as the limiting density of increasingly dense discrete distributions \cite{jaynes_prior_1968}.
-This argument leads to an expression for the information entropy of a continuous random variable $X$ as,
+Instead, \textcite{jaynes_information_1957,jaynes_prior_1968} argued that the information entropy of a continuous distribution should be defined as the limiting density of increasingly dense discrete distributions, $H_{N\rightarrow\infty}(X)$.
+This argument leads to,
 %
@@
 where $H_N(X)$ is the limiting density for discrete points, $q(x)$ is a uniform density over the quantization of the continuous space, and we have subtracted a $\log N$ term that would go to infinity in the limit of infinite points.
-By doing this, the information entropy becomes dimensionally correct, and invariant under a change of variables%
+By doing this, the information entropy becomes positive, dimensionally correct, and invariant under a change of variables%
 \footnote{\label{fn:change-of-variables-invariance-of-corrected-differential-entropy}
@@
 %
-The Kullback-Leibler divergence $D_{KL}(p\parallel q)$ is often interpreted as the amount of additional information required to represent events from distribution the $P$ with density $p(x)$ using a code optimized for distribution $Q$ with density $q(x)$.
-In other words, it measures how surprised one would be if they used distribution $Q$ to represent events from distribution $P$.
-The KL-divergence is also known as the \emph{relative entropy} of $P$ with respect to $Q$, or simply as the \emph{information gain}.
+The Kullback-Leibler divergence $D_{KL}(P\parallel Q)$ is often interpreted as the amount of additional information required to represent events from distribution $P$ with density $p(x)$ using a code optimized for distribution $Q$ with density $q(x)$.
+In other words, it measures how surprised one would be if they used distribution $Q$ to represent events from distribution $P$, i.e.
+the \emph{relative entropy} of $P$ with respect to $Q$.
+The KL-divergence is also known as the \emph{information gain}.
 
-Returning to Jaynes and \cref{eq:corrected-differential-entropy-jaynes}, the argument can be interpreted to say that the information entropy of a continuous random variable $X$ should be defined in relative terms as the expected difference in information entropy between its actual density and a uniform density.
-The Kullback-Leibler divergence as a measure of (relative) information entropy also provides a natural point of convergence of this section since it is equivalently defined for both discrete and continuous random variables.
+Returning to Jaynes' argument and \cref{eq:corrected-differential-entropy-jaynes}, we can interpret this to say that the information entropy of a continuous random variable $X$ should be defined as the expected difference in information entropy between its density and a uniform density.
+Identifying the Kullback-Leibler divergence as a measure of (relative) information entropy provides a natural point of convergence for this section since it is equivalently defined for both discrete and continuous random variables as opposed to the differential entropy.
+In this view, information entropy (relative or absolute) is a quantity that distills the distribution of a random variable into a single number that describes the diversity of the potential outcomes of the random variable.
 
@@
 
-\subsection{Two kinds of uncertainty}
+% \subsection{Two kinds of uncertainty}
 
-\todo[inline]{Consider not using the terms aleatoric and epistemic uncertainty.
+% \todo[inline]{Consider not using the terms aleatoric and epistemic uncertainty.
 Instead use the terms ``uncertainty due to lack of knowledge (reducible)" and ``uncertainty inherent to the data (irreducible)".}
 
-Uncertainty can be divided into two kinds: aleatoric and epistemic.
+% Uncertainty can be divided into two kinds: aleatoric and epistemic.
 
-Aleatoric uncertainty is defined as being inherent to the data, and is irreducible.
+% Aleatoric uncertainty is defined as being inherent to the data, and is irreducible.
 
-Epistemic uncertainty is uncertainty due to lack of knowledge, and can be reduced with more data.
+% Epistemic uncertainty is uncertainty due to lack of knowledge, and can be reduced with more data.
+
+% Two kinds of uncertainty:\\
+% - Aleatoric: Uncertainty inherent to the data. Irreducible. "Known unknowns". \\
+% - Epistemic: Uncertainty due to lack of knowledge. Can be reduced with more data. "Unknown unknowns".\\
 
@@
 % \cite{kabir_neural_2018} Survey of uncertainty quantification in deep learning.
-
-% Two kinds of uncertainty:\\
-% - Aleatoric: Uncertainty inherent to the data. Irreducible. "Known unknowns". \\
-% - Epistemic: Uncertainty due to lack of knowledge. Can be reduced with more data. "Unknown unknowns".\\
 

### 2023-09-10 chapters/introduction.tex
First complete introduction

@@
 \chapter[introduction]{Introduction}\label{chp:introduction}
-%s
+%
 % SIZE: ~9 pages
@@
 Growing administrative burdens, rapid scientific progress in medicine, and aging populations have created a need to rethink how medical professionals are best enabled to successfully do their job. 
-The application of artificial intelligence in medical decision-making, particularly in diagnosing and recommending treatments, is likely to hold significant potential for improving patient outcomes by reducing the effect of these issues.
-However, this prospect also raises concerns related to the reliability of such models as their decisions impact critical aspects of human health and well-being.
+The application of artificial intelligence in medical decision-making and particularly for diagnosing and recommending treatments is likely to hold significant potential to reduce the effect of these issues and improve patient outcomes.
+However, as such decisions impact critical aspects of human health and well-being, this prospect also raises concerns related to the reliability of artificial intelligence systems.
 
@@
 One example is the Egyptian Rhind Mathematical Papyrus, dated around 1550 BCE.
-It contains various mathematical techniques for calculations involving as multiplication, division, and fractions which were used for practical purposes like measuring land, constructing buildings, and sharing resources \cite{georges_universal_2001}.
-Around the same time, the Babylonians had developed a sophisticated system of mathematics, including algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \cite{fowler_square_1998}.
+It describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \cite{georges_universal_2001}.
+Around the same time, the Babylonians had developed a sophisticated system of mathematics that included algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \cite{fowler_square_1998}.
 
-However, these new computing tools also brought new challenges.
-The Babylonian system of mathematics did not formalize the difference between rational and irrational numbers often leading to approximation errors and the common practice of approximating $\pi$ with 3.
+However, these computing tools brought with them novel challenges.
+The Babylonian mathematics, for instance, did not formalize the difference between rational and irrational numbers often leading to approximation errors such as the common practice of approximating $\pi$ with 3 \cite{georges_universal_2001}.
 Apart from limitations inherent to the tools, inadequate human proficiency in their use was also a source of error.
 The Plimpton 322 clay tablet, for instance, contains several errors presumably made by a novice in Babylonian mathematics (\cref{fig:plimpton_332}) \cite{britton_plimpton_2011}.
-It is easy to forget that the challenges we face today in adopting new technologies are not new but have been faced by humans for thousands of years.
-Even in ancient times, it was important to understand the limitations of technology and the need for human expertise in using it.
+% Easy to overlook, the challenges we face in adopting new technologies today are not new but have been faced by humans for thousands of years.
+Today, as in ancient times, it was important to understand the limitations of technology and the need for human expertise in its use.
 
@@
     \centering
-    \includegraphics[width=0.90\textwidth]{plimpton_322.png}
+    % \includegraphics[width=0.90\textwidth]{plimpton_322.png}
+    \includegraphics[height=0.4\textheight]{plimpton_322.png}
     \caption{Technology is difficult.
@@
 
-While the quest to ease human decision-making has been ongoing for thousands of years, it was not until 1642 that the first mechanical calculator was successfully built by Blaise Pascal. 
-And while the idea of a programmable computer with applications beyond pure arithmetic was first proposed by Ada Lovelace in 1843, 
-it was not until the 1940s and 50s that % / 
+While the quest to ease human decision-making has been ongoing for thousands of years, it was not until 1642 that the first mechanical calculator was successfully designed by Blaise Pascal.
+Refined by Gottfried Leibniz during the latter half of the 17th century, the mechanical calculator marked the beginning of a new era where rudimentary calculations could be automated and performed with greater speed and accuracy than humans.
+In the following centuries, the development of computing machines continued, and in the 1820s, Charles Babbage proposed the first programmable mechanical computer, the Difference Engine.
+Although never completed, it laid the foundation for his later design of the Analytical Engine in 1837 which is considered to be the first general-purpose computer design.
+Working with Babbage, Ada Lovelace is widely regarded as the first to have recognized that general-purpose programmable computers could have applications beyond pure arithmetic.
+However, it was not until the advent of electronics that the first successful general-purpose programmable computers were developed, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}).
+% While the potential of a general-purpose programmable computer with applications beyond pure arithmetic was recognized by Ada Lovelace,  
+% it was not until the 1940s and 50s that % / 
 % it took one-hundred years before / 
-the first programmable, electronic, general-purpose digital computers were developed, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}) \cite{georges_universal_2001}. 
-As they became more accessible, general-purpose computers were adopted widely across society for various applications in science and industry, including healthcare.
+% the first programmable, electronic, general-purpose digital computers were developed, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}) \cite{georges_universal_2001}. 
+As manufacturing processes were developed and scaled, electronic computers were adopted widely across society for various applications in science and industry, including healthcare. 
 The societal changes brought about by the widespread adoption of computer technology through the latter half of the 20th century were so profound that some scholars refer to the era as the information age \cite{georges_universal_2001, harari_sapiens_2011}.
-\todo[inline]{Expand on the history of modern computing.}
 
@@
     \centering
-    \includegraphics[width=0.95\textwidth]{eniac_programmers.jpeg}
-    \caption{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the female programmers of the ENIAC.} %Photo: \copyright Corbis/ Getty Images.}
+    % \includegraphics[width=0.95\textwidth]{eniac_programmers.jpeg}
+    \includegraphics[height=0.4\textheight]{eniac_programmers.jpeg}
+    \caption{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.} %Photo: \copyright Corbis/ Getty Images.}
     \label{fig:eniac_programmers}
@@
 
-With the advent of the internet and the widespread adoption of digital technologies, the amount of data available to decision-makers has grown exponentially. 
+With the advent of electronic computers, the internet, and the widespread adoption of digital technologies, the amount of data available to decision-makers has grown exponentially. 
 This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that focuses on developing algorithms that can learn from data to recognize salient patterns and make predictions to guide decisions.
-Enabled by a massive increase in the power and availability of parallel computing resources, such methods have set new standards for the performance of computer systems in various domains, including speech recognition, computer vision, and natural language processing \cite{lecun_deep_2015}.
+Enabled by a massive increase in the performance and availability of parallel computing resources, such methods have set new standards for the abilities of computer systems in various domains, including speech recognition, computer vision, and natural language processing \cite{lecun_deep_2015}.
 In recent years, machine learning has also been applied within healthcare, where it has shown promising results in tasks such as medical imaging \cite{lundervold_overview_2019}, drug discovery \cite{chen_rise_2018}, and clinical decision support \cite{cite15, cite14}.
@@
 
-Though this incident is an extreme one, it highlights the challenges of applying machine learning in real-world scenarios and underlines the need for machine learning systems to be robust to adverse or difficult situations.
+Though this incident is an extreme one, it highlights the challenges of applying machine learning in real-world scenarios and underlines the need for machine learning systems to be robust in rare, adverse or difficult situations.
 Recently, European policymakers proposed a set of requirements for applications of machine learning to ensure that they are safe, reliable, and trustworthy \cite{europeancommission_briefing_2021}.
-This AI Act groups applications into three categories based on the potential risk they pose to society: limited risk, high risk, and unacceptable risk.
+This AI Act groups applications of machine learning into three categories based on the potential risk they pose to society: limited risk, high risk, and unacceptable risk.
 Limited risk applications, for instance within gaming, must meet transparency requirements, and unacceptable risk applications, such as public use of facial recognition, are banned.
@@
 One of the key ways to improve the safety, reliability and trustworthiness of machine learning systems is to equip them with the ability accurately estimate the uncertainty of their predictions. 
-A model that can accurately estimate the uncertainty of its predictions can communicate to other systems or humans when it is unsure about a prediction, and so avoid making potentially catastrophic mistakes.
+A model that can accurately estimate the uncertainty of its predictions is able to communicate to other systems or humans when it is not to be trusted, and so, is able to reduce the risk of wrong predictions leading to catastrophic mistakes.
 Fittingly, a recent study by the European Parliamentary Research Services concluded:
@@
 \noindent The research presented in this thesis will focus on the use of machine learning in relation to uncertainty estimation, speech processing, and medical conversations. 
-The research project of this thesis was formulated and carried out in collaboration with the Danish company Corti, which develops computer software for medical decision support. 
+The research project was partially funded by a grant from the national Danish Innovation Fund (grant no.\@ 0153-00167B) and defined and carried out in collaboration with the Danish company Corti, which develops computer software for medical decision support.
 In the rest of this chapter, we will provide a high-level introduction to these topics and discuss the motivation behind the research project.
@@
 %
-At the start of this project in 2020, Corti's main product was a software system providing decision-support for emergency medical dispatchers which operate the 1-1-2 emergency line in Copenhagen, Denmark and the 9-1-1 emergency line in Seattle, Washington in the United States.
+At the start of this project in 2020, Corti's main product was a software system providing decision-support for emergency medical dispatchers operating the 1-1-2 emergency line in Copenhagen, Denmark, and the 9-1-1 emergency line in Seattle, Washington, in the United States.
 The system, having been further developed, is still in use in 2023, and provides an interface for navigating a graph-based protocol, defined by the local emergency medical services, which dispatchers use to triage patients during emergency calls.
-Calls are transcribed in real-time using customer-domain automatic speech recognition and the system makes suggestions for the dispatcher to use in the conversation with the caller, including notifications about forgotten protocol questions \cite{havtorn_multiqt_2020} and potential cases of cardiac arrests \cite{cite15, cite14}.
-The system logs the details about the call and the dispatcher's actions to be used for later review and quality assurance.
+Calls are transcribed in real-time using automatic speech recognition built for the customer-domain and the system makes suggestions for the dispatcher to use in the conversation with the caller, including notifications about forgotten protocol questions \cite{havtorn_multiqt_2020} and potential cases of cardiac arrests \cite{cite15, cite14}.
+The system logs the details about the call and the actions taken by the dispatcher for later use in review, training and quality assurance.
 
@@
 This is likely due to the complexity of stroke cases which exhibit a wide range of symptoms that can be difficult to recognize over the phone.
-Additionally, out of total call volume, stroke cases are relatively rare, occurring in only $0.25\%$ of all calls to the Copenhagen 1-1-2 emergency line in 2021.
-This makes it difficult for call-takers to gain practical experience in recognizing stroke cases which exacerbates the general difficulty of recognizing the cases in the first place.
-Although several efforts have been made to improve the recognition of stroke cases in emergency calls \cite{cite13,cite14,cite15}, there is still a need for better tools to support call-takers in recognizing stroke cases.
+Additionally, stroke cases are relatively rare, occurring in only $0.25\%$ of all calls made to the Copenhagen 1-1-2 emergency line in 2021.
+This makes it difficult for call-takers to gain practical experience with stroke cases which exacerbates the difficulty of recognizing them in the first place.
+Although several efforts have been made to improve the recognition of stroke cases in emergency calls, there is still a need for better tools to support call-takers in recognizing stroke cases \cite{cite13,cite14,cite15}.
 
-A reasonable machine learning pipeline to assist in recognizing stroke on calls to emergency services might consist of an automatic speech recognition model to transcribe the conversation and a binary classifier to estimate the probability that what the transcript describes is a case of stroke.
-Such a system might be trained on a dataset calls with verified stroke and non-stroke cases and evaluated against a held-out test set of calls where the call-taker has indicated that they suspect a stroke.
+A reasonable machine learning pipeline to assist in recognizing stroke cases on calls to emergency services might consist of an automatic speech recognition model that transcribes the conversation and a binary classifier that estimates the probability that the transcript describes a case of stroke.
+Such a system might be trained on a dataset calls with verified stroke and non-stroke cases and evaluated against a held-out test set of calls where the call-taker has indicated whether they suspect a stroke.
 Due to the low prevalence of stroke cases, the dataset would likely be imbalanced, and the number of stroke cases limited.
@@
 In such a system, many factors can lead to classification errors, and when that happens, we would like the model to express high degree of uncertainty. 
-The automatic speech recognition model might be make an error in the transcription of particular word or phrase due to a noisy environment, overlapping, slurred or mumbling speech, or words not in the model's training data.
-The classifier might misclassify a conversation due to medically ambiguous symptoms, a general lack of information in the conversation, or errors of the speech recognizer propagated via the transcription.
+For instance, the automatic speech recognition model might make an error in the transcription of particular word or phrase due to a noisy environment, overlapping, slurred or mumbling speech, or words not in the model's training data.
+The classifier might misclassify a conversation due to medically ambiguous symptoms, a general lack of information given in the conversation, or errors of the speech recognizer propagated via the transcription.
 How to best make such models accurately represent and estimate the uncertainty of given predictions, and how to use it to improve the value of such a system, especially in cooperation with the call taker, is an open question.
@@
 %
-During the course of the project presented in this thesis, Corti's product portfolio has expanded to include several new products. 
+During the course of this project, Corti's product portfolio has expanded to include several new products. 
 One of these is a software system for medical coding of clinical notes. 
-When a patient is admitted to a hospital, the medical staff will write a clinical note describing the patient's diagnosis and the procedures they underwent, which includes drug prescriptions. 
+When a patient is admitted to a hospital, the medical staff will write a clinical note describing the patient's diagnosis and the procedures they underwent, including drug prescriptions. 
 These notes are then used for billing and reimbursement purposes, as well as for research and quality assurance. 
-The clinical notes are written in natural language adhering to a certain structure, usually by a doctor. Later, a medical coder will assign a set of medical codes to the note based on its content. 
-The process of medical coding is time-consuming and error-prone due to the number and complexity of medical diagnoses and procedures.
+The clinical notes are written, usually by a doctor, in natural language adhering to a certain structure. Later, a medical coder will assign a set of medical codes to the note based on its content. 
+The process of medical coding is time-consuming and error-prone due to the vast number and high complexity of medical diagnoses and procedures.
 For instance, the widely used International Classification of Diseases (ICD) standard consist of 55,000 medical codes in version 10 and 85,000 in version 11 \cite{worldhealthorganisationwho_international_2023}.
 Additionally, a single medical note will usually contain several diagnoses and procedures which must all be inferred from the clinical note.
-In public benchmarking datasets such as MIMIC-III and IV the number is around 15.
+In public benchmarking datasets such as MIMIC-III and IV the number is around 15 \cite{johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}.
 Official guidelines describe how to code specific diagnoses and procedures and combinations thereof \cite{centersformedicaremedicaidservicesus_icd10cm_2023}.
-For instance, a code may have several exclusion or inclusion criteria which define when certain other codes must be coded along with it, or when they must not.
+For instance, a code may have several exclusion or inclusion criteria which define which other codes must be coded along with it, or when they must not.
 
-\todo[inline]{
-Discuss the role of uncertainty estimation in medical coding. 
-Special challenges include significant class imbalance that cannot be removed by stratification since each example is associated with multiple labels.
-Mutually dependent classes also complicates uncertainty estimation.
-Some classes are mutually exclusive, while others are not.
-Some classes are easily confused with each other.
-Uncertainty can be helpful in identifying code candidates as uncertain codes may need human review.
-}
-
-Many codes are rarely used and datasets for medical coding usually have a considerable class-imbalance which can be impossible to exactly correct for by stratified sampling since each example is associated with multiple labels.
-
-\todo[inline]{Write paragraph on using machine learning for medical coding and the role of uncertainty estimation in that.}
+A reasonable machine learning pipeline to assist in medical coding might consist of a natural language processing model that extracts relevant information from the clinical note and outputs a number of probability distributions that each estimate the probability that the note should be assigned one of the medical codes.
+Training and using such as system comes with several challenges, however.
+For instance, the prevalence of different medical codes in the training data is highly imbalanced; it is common to have several orders of magnitude difference between the frequency of the most frequent code and the least frequent codes.
+Since each clinical note is associated with multiple codes that have complex co-occurence patterns, it is often impossible to exactly correct for this class imbalance by stratified sampling, which is an otherwise common approach to deal with class imbalance in machine learning.
+This also gives rise to
+Furthermore, some codes are easily confused with each other, for instance the ICD-10 codes Z87.891 ``Personal history of nicotine dependence'' and F17.210 ``Nicotine dependence, cigarettes, uncomplicated", but in practice only one of them should be assigned to a given note.
+As with the stroke recognition system, we would like the model to express high degree of uncertainty when it makes a mistake.
+For example, if the model is uncertain about two similar codes, it might be appropriate to ask a human expert to review the note and make the final decision.
+Additionally, the lack of training data for the rarest codes makes it difficult to learn to robustly predict those codes, so we would in general want the model to be uncertain about such codes.
 
@@
 %
-Many modern machine learning models are deep neural networks with millions or even billions of parameters. 
-These models are often referred to as black-box models as it is difficult to understand how they arrive at their predictions. 
-The lack of interpretability makes it difficult to assess the reliability of the model's predictions. 
-Furthermore, a number of factors, including the ability of deep neural networks to overfit, or even memorize, the training data \cite{arpit_closer_2017, burg_memorization_2021}, and the use of training objectives that are proxies for the evaluation metrics of interest, lead to models that often are miscalibrated; that is, the probability of a class predicted by the model does not reflect the true probability of the predicted class under the circumstances \cite{guo_calibration_2017, kull_temperature_2019}.
+Several factors contribute to the reliability of a machine learning model. 
+Besides model performance and accuracy, important factors include the interpretability of how the model functions, the explainability of its predictions, fairness in its treatment of different groups, and robustness to noise, outliers and adversarial attacks.
+% In medical applications, it has been argued that another factor
+Since many modern machine learning models are deep neural networks with millions or billions of parameters, their size and complexity make them inherently difficult to interpret, and their predictions hard to explain.
+Due to high cost and practical infeasibility, the vast amounts of data needed to train such models are often not manually curated for quality, and so, may contain biases and errors which models are well-equipped to learn to mirror \cite{burkart_survey_2021}.
+Finally, a number of factors, including the ability of deep neural networks to overfit, or even memorize, the training data \cite{arpit_closer_2017, burg_memorization_2021}, can lead to models that are sensitive adverse noise conditions and outliers.
+
+The ability of a model to accurately estimate the uncertainty of its predictions plays an important role in its reliability.
+However, the tendency of deep neural networks to overfit or memorize the training data \cite{arpit_closer_2017, burg_memorization_2021} and the use of training objectives that are proxies for the evaluation metrics of interest lead to models that often are miscalibrated; that is, the predicted probability of a class does not reflect the true probability of the class.
+This means, for instance, that if for a certain input the model assigns 70\% probability to a class, it would not generally be right 70\% of the time, if we could repeat the prediction multiple times under similar circumstances \cite{guo_calibration_2017, kull_temperature_2019}.
 Usually, the predicted probabilities are too extreme which leads to overconfident predictions.
 This means a model will often assign a high probability to the predicted class, even when it is wrong.
-In practice, such a model can be hard to trust.
-No model is perfect, but the inability to judge the certainty of a model's predictions makes it difficult to know how to weigh the model's predictions in a decision-making process.
 
+In practice, miscalibrated models are not reliable in high risk applications.
+Although no model is perfect, the inability of a human, or another model, to assess the certainty with which a prediction was made, makes it impossible to know how to weigh it in a decision-making process.
 A connection can be made back to the incident with the Uber self-driving car.
@@
 \subsection{Model calibration}
+% 
+The problem of miscalibration in machine learning models is well-known and has been studied extensively \cite{lewis_sequential_1995, platt_probabilistic_1999, garczarek_classification_2002, zadrozny_transforming_2002, bennett_using_2003}.
+One of the best known methods for calibrating a binary classifier is Platt scaling which fits a logistic regression model to the model outputs, assuming that the miscalibration can be corrected by a logarithmic function \cite{platt_probabilistic_1999}.
+Another method is isotonic regression which instead fits a non-parametric, monotonic function \cite{zadrozny_transforming_2002}.
+More recently, \textcite{guo_calibration_2017} proposed a single-parameter variant of Platt-scaling that fits only a temperature parameter on the logits of a neural network classifier. 
+These methods are generally simple and effective, but not without limitations. 
+To perform the calibration, most of the methods require a held-out validation set on which the model was not trained.
+Platt scaling and isotonic regression are not naively applicable to multi-class classification problems.
+How to calibrate models for structured prediction tasks such as speech recognition and machine translation remains an open problem \cite{astudillo_uncertainty_2010, astudillo_integration_2013, jayashankar_detecting_2020}.
 
-\todo[inline]{
-    Expand in this discussion of overconfidence and calibration.
-E.g.: Calibration is fairly well understood in the context of classification, especially binary, but is harder in structured prediction such as automated speech recognition.
-Also:
-    Even if a model is well-calibrated, it can still be overconfident in wrong predictions - this is a problem in particular for rare events or adverse examples that were not presented to the model during training.
-Such examples are sometimes called out-of-distribution examples with reference to the training distribution.
-    For instance, a model trained to classify images of cats and dogs may have been well-calibrated, but if it is presented with an image of a horse, it has no option but to assign 100\% total probability to the cat and dog categories, even though that is clearly wrong.
-    Worse yet, we might expect the model to then at least assign 50\% probability to each of the cat and dog categories, but that behavior is not guaranteed.
-Specifically, since no horses were in the training data, the model will assign arbitrary probabilities to either class and so still risks being confidently wrong.
-}
+It is also important to realize that correct calibration does not guarantee that the probability of a prediction is accurate. 
+Specifically, even if a model is well-calibrated, it can still be overconfident for data that were not presented to the model during training such as rare events, outliers, and adverse examples.
+Such data are sometimes called out-of-distribution examples with reference to the training distribution.
+For instance, a model trained to classify images of cats and dogs and provide calibrated class probabilities will still fail if presented with an image of a horse.
+In this case, it has no option but to distribute 100\% total probability across the cat and dog categories, even though that is clearly wrong.
+Worse yet, we might expect the model to then at least assign 50\% probability to each of the cat and dog categories, to indicate the uncertainty, but that behavior is not guaranteed.
+Indeed, since no horses were in the training data, the model will not have learned specialized features for horses, nor learned to associate known features with horses.
+So, if some features of a particular horse resemble those learned for a dog more than for a cat, the model might assign arbitrarily high probability to the dog class and therefore risks being confidently wrong \cite{zhou_survey_2022}.
 
@@
 % 
-%
-The distinction in uncertainty estimation between known and unknown unknowns is sometimes referred to in terms of \textit{aleatoric uncertainty} and \textit{epistemic uncertainty} \cite{kendall_what_2017}.
-Both categories of uncertainty are usually used to describe uncertainties in a model's predictions or the model itself including its parameters, but they differ in their sources.
-Aleatoric uncertainty (from Latin (aleatorius) 'dice player' or 'gambler') refers to uncertainty in the data itself due to true randomness which is irreducible even with infinite data. 
-Epistemic uncertainty (from Greek (epistḗmē) 'knowledge') refers to uncertainty due to things one could in principle know but does not in practice. 
-% It could arise because a measurement is not accurate, because the model neglects certain effects, or because particular data have been deliberately hidden. 
-Epistemic uncertainty is sometimes referred to as \textit{systematic uncertainty} and aleatoric uncertainty as \textit{stochastic uncertainty} \cite{kendall_what_2017}. 
-%, or, \textit{reducible} and \textit{irreducible} uncertainty, respectively.
-This latter terminology is somewhat misleading since epistemic uncertainty might not be reducible by additional data collection if the model is not sufficiently expressive to represent the data distribution, and aleatoric uncertainty is not necessarily irreducible if we can reduce inherent noise by improving the instruments used to collect the data \cite{kendall_what_2017}.
+As hinted at in the previous section, there are different sources of uncertainty in machine learning models. 
+At a fundamental level, we can distinguish between uncertainty that is present in the knowledge we have, and uncertainty that originates from the knowledge that we do not have.
+This is sometimes called known unknowns and unknown unknowns, and is referred to in terms of \textit{aleatoric uncertainty} and \textit{epistemic uncertainty} \cite{kendall_what_2017}. 
+Aleatoric uncertainty (from Latin (aleatorius) 'dice player' or 'gambler') refers to uncertainty present in the data itself due to randomness in the process that generated the data.
+This kind of uncertainty is commonly seen as irreducible but, since it is represented in the collected data, it can usually be modelled directly.
+Epistemic uncertainty (from Greek (epistḗmē) 'knowledge') refers to uncertainty due to things one could in principle know, but does not in practice.
+This lack of knowledge could be due to a lack of data, or an improper model specification.
+Aleatoric uncertainty is sometimes referred to as \textit{stochastic uncertainty} and epistemic uncertainty as \textit{systematic uncertainty} \cite{kendall_what_2017}. 
 Although the distinction between aleatoric and epistemic uncertainty is important, it is not always useful.
@@
 An example of aleatoric uncertainty is the uncertainty in the transcription of a word due to a noisy environment, overlapping, slurred or mumbling speech. 
-We can reduce this uncertainty by improving the automatic speech recognition model, but we cannot eliminate it completely.
+Unknown words are arguably sources of epistemic uncertainty although a good speech recognition model may generalize well if the word's spelling and pronunciation follow the same patterns as the words in the training data.
+Another example of epistemic uncertainty is the occurrence of truly out-of-distribution examples, such as the horse in the image classifier example above: inputs unlike anything the model has seen during training, especially if they require outputs unavailable to the model.
+In high-dimensional data with a practically unlimited diversity in possible out-of-distribution examples, it is impossible to collect enough data to eliminate all sources of epistemic uncertainty.
+If we included horses in the training data, the model would likely learn to recognize them, but it would still be unable to recognize giraffes, or zebras, or unicorns.
+Ultimately, any practical machine learning model will have some non-negligible sources of epistemic uncertainty.
 
-Words not present in the training data are arguably sources of epistemic uncertainty although a good speech recognition model may generalize well to unseen words if its spelling and pronunciation follow the same patterns as the words in the training data.
+In practice, we are interested in the predictive uncertainty, so we need to include both aleatoric and epistemic uncertainty in the estimate. 
+Since sources of epistemic uncertainty are by definition those sources not represented in the data, it is not possible to model them directly, as it is for aleatoric uncertainty.
+This generally makes epistemic uncertainty more difficult to model and quantify than aleatoric uncertainty.
+Some methods for estimating epistemic uncertainty include Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995}, Monte Carlo dropout \cite{gal_dropout_2016}, and ensembling \cite{lakshminarayanan_simple_2017}.
+Common to these methods is that they represent epistemic uncertainty by an implicit or explicit distribution over the model parameters.
+This is in contrast to aleatoric uncertainty, which is usually represented by a distribution over the data.
+The methods often require multiple forward passes through the model to estimate the predictive uncertainty.
+This makes them computationally expensive and impractical for real-time applications such as speech recognition.
 
-A stronger example of epistemic uncertainty is the occurrence of truly out-of-distribution examples, such as the horse in the image classifier example above: inputs unlike anything the model has seen during training that potentially require outputs unavailable to the model.
-In high-dimensional data with a practically unlimited diversity in possible out-of-distribution examples, it is impossible to collect enough data to eliminate epistemic uncertainty.
-If we included some horses in the training data, the model would likely be able to recognize them, but it would still be unable to recognize giraffes, or zebras, or unicorns.
-Ultimately, any practical machine learning model will have at least some sources of epistemic uncertainty i.e. unknown sources of uncertainty.
-
-% Aleatory uncertainty is often pretty straightforward to model since it is available in the data.
-Epistemic is much harder since we don't know much about it, other than we have only examples of where it is low.
-
-% In real life applications, both kinds of uncertainties are present.
-Uncertainty quantification intends to explicitly express both types of uncertainty separately.
-The quantification for the aleatoric uncertainties can be relatively straightforward, where traditional (frequentist) probability is the most basic form.
-Techniques such as the Monte Carlo method are frequently used.
-A probability distribution can be represented by its moments (in the Gaussian case, the mean and covariance suffice, although, in general, even knowledge of all moments to arbitrarily high order still does not specify the distribution function uniquely), or more recently, by techniques such as Karhunen–Loève and polynomial chaos expansions.
-To evaluate epistemic uncertainties, the efforts are made to understand the (lack of) knowledge of the system, process or mechanism.
-Epistemic uncertainty is generally understood through the lens of Bayesian probability, where probabilities are interpreted as indicating how certain a rational person could be regarding a specific claim.
-
-% In mathematics, uncertainty is often characterized in terms of a probability distribution.
-From that perspective, epistemic uncertainty means not being certain what the relevant probability distribution is, and aleatoric uncertainty means not being certain what a random sample drawn from a probability distribution will be.
-
-\todo[inline]{Wrap this up.
-How to model aleatoric and epistemic uncertainty.
-Why do we care most about epistemic uncertainty (because we care about correctly estimating predictive uncertainty, and aleatoric is easy to model).}
-
-\section{Scope of the thesis}
+\section{Thesis scope and outline}
 %
@@
 %
-The complete set of research published as part of this project covers a number of topics related to decision-support in medical conversations. 
-One group of the produced studies is concerned with generative latent variable models and their applications to uncertainty estimation and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
-Another group of studies is concerned with self-supervised speech representation learning and automatic speech recognition \cite{borgholt_scaling_2021,borgholt_we_2021,mohamed_selfsupervised_2022,borgholt_brief_2022}.
-A final group deals with applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.
+% The complete set of research published as part of this project covers a number of topics related to decision-support in medical conversations. 
+% One group of the primary papers is concerned with generative latent variable models and their applications to estimation of epistemic uncertainty and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
+% Another group of studies examines self-supervised speech representation learning as an alternative to generative latent variable models \cite{mohamed_selfsupervised_2022,borgholt_brief_2022} and explores them for speech recognition \cite{borgholt_scaling_2021} and spoken language understanding \cite{borgholt_we_2021}.
+% Finally, a group deals with applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.
 
-This first part of the thesis ties together these different studies by providing a high-level overview of the research topic and summarizing the main contributions of the thesis.
+This introduction provided a high-level overview of the motivation behind the research project and highlighted important use-cases of speech processing, natural language processing and the importance of uncertainty quantification.
+The remainder of \cref{part:background} consists of two chapters.
+\Cref{chp:technical-background} provides relevant technical background to the papers of the thesis, including uncertainty, out-of-distribution detection, and variational autoencoders.
+Finally, \cref{chp:main-contributions} outlines the main contributions of the thesis in relation to the published papers and the research project as a whole.
+
+\Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} each contain a number of chapters that each correspond to a primary paper.
+\Cref{part:unsupervised-uncertainty-estimation} contains papers on variational autoencoders and their applications to estimation of epistemic uncertainty and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
+\Cref{part:self-supervised-speech-representation-learning} contains a review paper on self-supervised speech representation learning \cite{mohamed_selfsupervised_2022}.
+\Cref{part:medical-applications} contains papers on applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.
+
+Finally, \cref{part:discussion-and-conclusion} concludes the thesis by summarizing the main contributions and discussing the presented work and future directions for research.
 

### 2023-09-11 chapters/paper_modelagnostic.tex
Fix typo

@@
 %
-Other works in the literature focused on deriving specific test statistics that works only for a specific model, for example for VAEs \parencite{xiao_likelihood_2020, maaloe_biva_2019, havtorn_hierarchical_2021}, or for normalizing flows \parencite{kirichenko_why_2020, ahmadian_likelihoodfree_2021}.
+Other works in the literature focused on deriving specific test statistics that work only for a specific model, for example for VAEs \parencite{xiao_likelihood_2020, maaloe_biva_2019, havtorn_hierarchical_2021}, or for normalizing flows \parencite{kirichenko_why_2020, ahmadian_likelihoodfree_2021}.
 

### 2023-09-11 chapters/background.tex
Start outline of OODD section
Work on background section

@@
 
-\section{Unsupervised out-of-distribution detection}
+\section{Out-of-distribution detection}
+% 
+% OUTLINE: C.f. https://openaccess.thecvf.com/content/CVPR2023W/VAND/papers/Graham_Denoising_Diffusion_Models_for_Out-of-Distribution_Detection_CVPRW_2023_paper.pdf
+% - Formalize the problem. https://arxiv.org/pdf/2302.10326.pdf
+%   - Assume we have a dataset $\mathcal{D}_{\text{train}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\vx$ is sampled from $\mathcal{D}_{\text{train}}$ or not.
+Given a test point $\vx$, the model outputs a score $s(\vx)$ that indicates whether $\vx$ is in-distribution or out-of-distribution.
+%   - In the context of neural networks, this problem dates back to at least 1994 \cite{bishop_novelty_1994}. 
+% - Categorisation into supervised and unsupervised. 
+%   - Supervised: Whether p(y|x) should be trusted for a given x. 
+%   - Unsupervised: Whether x should be trusted and used at all. 
+% - Supervised OODD:
+%   - Requires some examples of (synthetic) OODD data, which in some sense makes it inherently flawed. We can never really be sure that we have good enough coverage of OODD data.
+%   - One limitation of model-dependent OoD techniques is that they may discard information about p(x) in learning the task-specific model p(y|x). That information may be useful for OoD detection.
+%   - Likelihood Ratios for Out-of-Distribution Detection \cite{ren_likelihood_2019} provides a number of baselines for supervised OODD:
+%       - 1 Maximum class probability of the predictive distribution p(y|x) \cite{hendrycks_baseline_2017}
+%       - 2 Entropy of predictive distribution p(y|x) \cite{ren_likelihood_2019}
+%       - 3 ODIN \cite{liang_enhancing_2018} (Uses synthetic OOD data by adding noise to in-distribution)
+%       - 4 Mahalanobis distance to class conditional Gaussian distribution \cite{lee_training_2018}
+%       - 5 DeepEnsemble \cite{lakshminarayanan_simple_2017} The classifier-based ensemble method that uses the average of the predictions from multiple independently trained models with random initialization of network parameters and random shuffling of training inputs
+%       - 6 The log-odds of a binary classifier trained to distinguish between in-distribution inputs from all classes as one class and randomly perturbed in-distribution inputs as the other.
+\cite{ren_likelihood_2019}
+%       - 7 The maximum class probability over K in-distribution classes of a (K + 1)-class classifier where the additional class is perturbed in-distribution. \cite{ren_likelihood_2019}
+%       - 8 The maximum class probability of a K-class classifier for in-distribution classes but the predicted class distribution is explicitly trained to output uniform distribution on perturbed in-distribution inputs.
+\cite{ren_likelihood_2019}.
+This is similar to using simulated OOD inputs from GAN \cite{lee_training_2018} or using auxiliary datasets of outliers \cite{hendrycks_deep_2019} for calibration purpose.
+%   - Variational information bottleneck (VIB) (Alemi et al., 2018b) performs divergence estimation in latent space to detect OoD, but is technically a model-dependent technique because the latent code is trained jointly with the downstream classification task.
+\cite{alemi_deep_2017}
+% - Unsupervised ODDD
+%   - No OODD data required. Isolates uncertainty related to the input data.
+%   - Unsupervised can be categorised into generative and reconstructive.
+%   - Generative
+%       - 9 Compute WAIC (\cite{watanabe_algebraic_2009, watanabe_asymptotic_2010}) using ensemble of generative models \cite{choi_waic_2019}
+%       - Likelihood Ratios for Out-of-Distribution Detection \cite{ren_likelihood_2019}
+%       - Input complexity and out-of-distribution detection with likelihood-based generative models \cite{serra_input_2020}
+%       - Likelihood regret for variational autoencoders \cite{xiao_likelihood_2020}
+%       - Typicality test for generative OODD (Detecting out-of-distribution inputs to deep generative models using typicality) \cite{nalisnick_detecting_2019}
+%       - Density of states estimation for out of distribution detection \cite{morningstar_density_2021}
+%   - Reconstructive
+%       - A review of novelty detection \cite{pimentel_review_2014}
+%       - Outlier detection using autoencoders \cite{lyudchik_outlier_2016} and ensembles of autoencoders \cite{chen_outlier_2017}
+%       - Improving reconstruction autoencoder out-of-distribution detection with mahalanobis distance \cite{denouden_improving_2018}
+%       - Denosing diffusion models for out-of-distribution detection \cite{graham_denoising_2023}
+%       - Diffusion inpainting \cite{liu_unsupervised_2023a}
 
-Unsupervised out-of-distribution detection (OOD) is the task of identifying data that is unlikely to originate from the distribution of the training data.
+% , MC Dropout (Gal & Ghahramani, 2016) and  model a calibrated predictive distribution for a classification task.
+
+In this section we will introduce the task of out-of-distribution detection, relate it to the concept of uncertainty and relate it to the previous section on information entropy.
+We will also provide a brief overview of some of the methods that have been proposed to tackle the task of out-of-distribution detection to supplement the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}.
+
+Out-of-distribution detection (OOD) is the task of identifying data that is unlikely to originate from the distribution of the training data. 
+Assume we have a dataset $\mathcal{D}_{\text{train}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\vx$ is sampled from $\mathcal{D}_{\text{train}}$ or not.
+Given a test point $\vx$, the model outputs a score $s(\vx)$ that indicates whether $\vx$ is in-distribution or out-of-distribution.
+In the context of neural networks, this problem dates back to at least 1994 \cite{bishop_novelty_1994}.
+
+Out-of-distribution detection is closely related to the concept of uncertainty. 
+\todo[inline]{Relate to epistemic uncertainty}
+
+One way to subdivide approaches to out-of-distribution detection is based on whether the model estimates a conditional probability distribution $p(y|\vx)$ over some target given the input, or whether it estimates a probability distribution $p(\vx)$ over the input itself.
+We might refer to the former as \emph{supervised} out-of-distribution detection, and the latter as \emph{unsupervised} out-of-distribution detection.
 

### 2023-09-16 chapters/background.tex
Updates

@@
 \footnote{\label{fn:dimensional-analysis-of-differential-entropy}
-    Information entropy has the same dimensionality as information content which is typically measured in units of bits, depending on the base of the logarithm. 
+    Information entropy has the same dimensionality as information content which is typically measured in units of bits, but depends on the base of the logarithm. 
     However, the bit is not a base unit of the International System of Units (SI), nor is it an official derived unit.
@@
     \end{equation*}
-    which is clearly not dimensionless and therefore does not correspond to the discrete information entropy.
+    which is clearly not dimensionless and therefore cannot correspond directly to the discrete information entropy.
     % As a final twist on this story, note that the SI system does actually have a unit for an amount of substance: the mole \cite{internationalbureauofweightsandmeasures_systeme_2019}. 
@@
 \begin{equation} \label{eq:corrected-differential-entropy-jaynes}
-    H_{N\rightarrow\infty}(X) = \lim_{N\rightarrow\infty} [H_{N}(X) - \log N] = - \int_\mathcal{X} p(x) \log \left(\frac{p(x)}{q(x)}\right) \enspace ,
+    H_{N\rightarrow\infty}(X) \equiv \lim_{N\rightarrow\infty} [H_{N}(X) - \log N] = - \int_\mathcal{X} p(x) \log \left(\frac{p(x)}{q(x)}\right) \enspace ,
 \end{equation}
@@
 %
-The Kullback-Leibler divergence $D_{KL}(P\parallel Q)$ is often interpreted as the amount of additional information required to represent events from distribution $P$ with density $p(x)$ using a code optimized for distribution $Q$ with density $q(x)$.
+The Kullback-Leibler divergence $D_{KL}(P\parallel Q)$, also known as the \emph{information gain}, is often interpreted as the amount of additional information required to represent events from distribution $P$ with density $p(x)$ using a code optimized for distribution $Q$ with density $q(x)$.
 In other words, it measures how surprised one would be if they used distribution $Q$ to represent events from distribution $P$, i.e.
 the \emph{relative entropy} of $P$ with respect to $Q$.
-The KL-divergence is also known as the \emph{information gain}.
 
 Returning to Jaynes' argument and \cref{eq:corrected-differential-entropy-jaynes}, we can interpret this to say that the information entropy of a continuous random variable $X$ should be defined as the expected difference in information entropy between its density and a uniform density.
-Identifying the Kullback-Leibler divergence as a measure of (relative) information entropy provides a natural point of convergence for this section since it is equivalently defined for both discrete and continuous random variables as opposed to the differential entropy.
-In this view, information entropy (relative or absolute) is a quantity that distills the distribution of a random variable into a single number that describes the diversity of the potential outcomes of the random variable.
+Identifying the Kullback-Leibler divergence as a measure of this relative information entropy provides a natural point of convergence for this section since it is equivalently defined for both discrete and continuous random variables, contrary to the differential entropy.
+In this view, information entropy is a quantity that distills the distribution of a random variable into a single number that describes the diversity of the potential outcomes of the random variable. 
+% In the following, we will see that this interpretation of information entropy as a measure of diversity is useful for understanding the concept of uncertainty in machine learning.
 
 % In fact, Kullback initially motivated the divergence as an expected likelihood ratio \cref{kullback_information_1959}.
-
-% \subsection{Two kinds of uncertainty}
-
-% \todo[inline]{Consider not using the terms aleatoric and epistemic uncertainty.
-Instead use the terms ``uncertainty due to lack of knowledge (reducible)" and ``uncertainty inherent to the data (irreducible)".}
-
-% Uncertainty can be divided into two kinds: aleatoric and epistemic.
-
-% Aleatoric uncertainty is defined as being inherent to the data, and is irreducible.
-
-% Epistemic uncertainty is uncertainty due to lack of knowledge, and can be reduced with more data.
-
-% Two kinds of uncertainty:\\
-% - Aleatoric: Uncertainty inherent to the data. Irreducible. "Known unknowns". \\
-% - Epistemic: Uncertainty due to lack of knowledge. Can be reduced with more data. "Unknown unknowns".\\
 
@@
 % - Formalize the problem. https://arxiv.org/pdf/2302.10326.pdf
-%   - Assume we have a dataset $\mathcal{D}_{\text{train}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\vx$ is sampled from $\mathcal{D}_{\text{train}}$ or not.
-Given a test point $\vx$, the model outputs a score $s(\vx)$ that indicates whether $\vx$ is in-distribution or out-of-distribution.
+%   - Assume we have a dataset $\mathcal{D}_{\text{train}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from $\mathcal{D}_{\text{train}}$ or not.
+Given a test point $\xb$, the model outputs a score $s(\xb)$ that indicates whether $\xb$ is in-distribution or out-of-distribution.
 %   - In the context of neural networks, this problem dates back to at least 1994 \cite{bishop_novelty_1994}. 
@@
 This is similar to using simulated OOD inputs from GAN \cite{lee_training_2018} or using auxiliary datasets of outliers \cite{hendrycks_deep_2019} for calibration purpose.
+%   - X Deep semi-supervised anomaly detection \cite{ruff_deep_2020}.
 %   - Variational information bottleneck (VIB) (Alemi et al., 2018b) performs divergence estimation in latent space to detect OoD, but is technically a model-dependent technique because the latent code is trained jointly with the downstream classification task.
@@
 
-In this section we will introduce the task of out-of-distribution detection, relate it to the concept of uncertainty and relate it to the previous section on information entropy.
-We will also provide a brief overview of some of the methods that have been proposed to tackle the task of out-of-distribution detection to supplement the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}.
+In this section we will introduce out-of-distribution detection, relate it to the concept of uncertainty and to the previous section on information entropy.
+To supplement the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}, we will also provide a brief overview of some of the related methods that have been proposed in the literature.
 
-Out-of-distribution detection (OOD) is the task of identifying data that is unlikely to originate from the distribution of the training data. 
-Assume we have a dataset $\mathcal{D}_{\text{train}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\vx$ is sampled from $\mathcal{D}_{\text{train}}$ or not.
-Given a test point $\vx$, the model outputs a score $s(\vx)$ that indicates whether $\vx$ is in-distribution or out-of-distribution.
+Out-of-distribution detection is the task of identifying data that is unlikely to originate from the distribution of the training data. 
+In the general case, we assume that we have a dataset $\mathcal{D}_{\text{training}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from $\mathcal{D}_{\text{training}}$ or not.
+Out-of-distribution detection is closely related to the concept of uncertainty.
+Specifically, since such data is unknown to the model, it is, by definition, a source of epistemic uncertainty.
+Only in cases where strong prior knowledge of the data is incorporated into the model, can we expect the model to be able to make predictions on out-of-distribution data.
+
+One way to categorize the many approaches to out-of-distribution detection is based on whether the method estimates a conditional probability distribution $p(y|\xb)$ over some target variable given the input, or a probability distribution $p(\xb)$ over the input itself.
+We might refer to the former as \emph{supervised} out-of-distribution detection, and the latter as \emph{unsupervised} out-of-distribution detection \cite{graham_denoising_2023,liu_unsupervised_2023a}.
+Supervised out-of-distribution detection tries to assess whether the model's prediction $p(y|\xb)$ should be trusted for a given input $\xb$, whereas unsupervised out-of-distribution detection seeks to assess whether the input $\xb$ should be trusted and used at all.
+
+\subsection{Supervised out-of-distribution detection}
+
+In the supervised setting the supervision can come from different sources that result in different levels of informedness. 
+If one assumes access to representative out-of-distribution data, then one can train a model to distinguish between in-distribution and out-of-distribution data. 
+The out-of-distribution data can either be synthetic or real.
+Such approaches have achieved high performance \cite{lee_training_2018,hendrycks_deep_2019,ruff_deep_2020} but require access to out-of-distribution data, which in some sense makes it inherently flawed; it is often not possible to guarantee good enough coverage of out-of-distribution data to cover all possibilities.
+\todo[inline]{Describe an example method or two...}
+
+Less informed approaches only assume access to in-distribution data when learning the task-specific model $p(y|\xb)$. 
+One particularly simple approach requiring only in-distribution data uses the maximum class probability of the predictive distribution $p(y|\xb)$ to evaluate whether the example is in-distribution or not \cite{hendrycks_baseline_2017}.
+Other simple methods presented by \textcite{ren_likelihood_2019} include using the entropy of $p(y|\xb)$, the maximum class probability with an additional class trained to predict perturbed in-distribution data, the maximum class probability where the predicted class distribution is explicitly trained to output uniform distribution on perturbed in-distribution inputs, and the log-odds of a binary classifier trained to distinguish between original and perturbed in-distribution data.
+
+In ODIN \cite{liang_enhancing_2018}, the authors propose to add input-dependent noise to the inputs of a pre-trained classifier and use the calibrated maximum class probability of the predictive distribution $p(y|\xb)$ to evaluate whether the example is in-distribution or not.
+This method does not require retraining the classifier and generally improves on the simpler methods mentioned above.
+In \textcite{lee_training_2018}, the authors fit a multivariate Gaussian distribution to the activations of the penultimate layer of a pre-trained classifier and use the Mahalanobis distance to this distribution to evaluate whether the example is in-distribution or not.
+\todo[inline]{Continue here ...}
+
+%       - 2 Entropy of predictive distribution p(y|x) \cite{ren_likelihood_2019}
+%       - 6 The log-odds of a binary classifier trained to distinguish between in-distribution inputs from all classes as one class and randomly perturbed in-distribution inputs as the other.
+\cite{ren_likelihood_2019}
+%       - 7 The maximum class probability over K in-distribution classes of a (K + 1)-class classifier where the additional class is perturbed in-distribution. \cite{ren_likelihood_2019}
+%       - 8 The maximum class probability of a K-class classifier for in-distribution classes but the predicted class distribution is explicitly trained to output uniform distribution on perturbed in-distribution inputs.
+\cite{ren_likelihood_2019}.
+This is similar to using simulated OOD inputs from GAN \cite{lee_training_2018} or using auxiliary datasets of outliers \cite{hendrycks_deep_2019} for calibration purpose.
+
+%       - 3 ODIN \cite{liang_enhancing_2018} (Uses synthetic OOD data by adding noise to in-distribution)
+%       - 4 Mahalanobis distance to class conditional Gaussian distribution \cite{lee_training_2018}
+%       - 5 DeepEnsemble \cite{lakshminarayanan_simple_2017} The classifier-based ensemble method that uses the average of the predictions from multiple independently trained models with random initialization of network parameters and random shuffling of training inputs
+
+%   - Variational information bottleneck (VIB) (Alemi et al., 2018b) performs divergence estimation in latent space to detect OoD, but is technically a model-dependent technique because the latent code is trained jointly with the downstream classification task.
+\cite{alemi_deep_2017}
+
+\cite{hendrycks_baseline_2017, lakshminarayanan_simple_2017, lliang_enhancing_2018, lee_simple_2018, huang_importance_2021, wang_vim_2022}
+
+ either synthetic out-of-distribution data or from real out-of-distribution data.
+Synthetic out-of-distribution data is data that is generated by some process that is known to be out-of-distribution.
+For example, if we are training a model to classify images of cats and dogs, we could generate synthetic out-of-distribution data by adding noise to the images.
+Real out-of-distribution data is data that is known to be out-of-distribution, but is not generated synthetically.
+For example, if we are training a model to classify images of cats and dogs, we could use images of cars as real out-of-distribution data.
+
+ requires some examples of out-of-distribution data,
+
+ which in some sense makes it inherently flawed.
+As we discussed earlier, for many modern applications of machine learning, the data is high dimensional and complex which makes sufficiently sampling the space of out-of-distribution data intractable and makes it impossible to guarantee robust out-of-distribution detection.
+Another weakness of supervised out-of-distribution detection is that in learning the task-specific model $p(y|\xb)$ a model may discard information about $p(\xb)$ which could be useful for out-of-distribution detection.
+
+ \subsection{Unsupervised out-of-distribution detection}
+
+% Assume we have a dataset $\mathcal{D}_{\text{training}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from $\mathcal{D}_{\text{training}}$ or not.
+Given a test point $\xb$, the model outputs a score $s(\xb)$ that indicates whether $\xb$ is in-distribution or out-of-distribution.
 In the context of neural networks, this problem dates back to at least 1994 \cite{bishop_novelty_1994}.
 
-Out-of-distribution detection is closely related to the concept of uncertainty. 
-\todo[inline]{Relate to epistemic uncertainty}
-
-One way to subdivide approaches to out-of-distribution detection is based on whether the model estimates a conditional probability distribution $p(y|\vx)$ over some target given the input, or whether it estimates a probability distribution $p(\vx)$ over the input itself.
-We might refer to the former as \emph{supervised} out-of-distribution detection, and the latter as \emph{unsupervised} out-of-distribution detection.
+Continual novelty detection \cite{aljundi_continual_2022}
 

### 2023-09-16 chapters/background.tex
Work on background

@@
 %   - In the context of neural networks, this problem dates back to at least 1994 \cite{bishop_novelty_1994}. 
+%
 % - Categorisation into supervised and unsupervised. 
@@
 %   - Unsupervised: Whether x should be trusted and used at all. 
+%
+In this section we will introduce out-of-distribution (OOD) detection, relate it to the concept of uncertainty and to the previous section on information entropy. 
+To supplement the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}, we will also provide a brief overview of some of the related methods that have been proposed in the literature.
+
+OOD detection is the task of identifying data that is unlikely to originate from the distribution of the training data and, in the context of neural networks, the task dates back to at least 1994 \cite{bishop_novelty_1994}.
+In the general case, we assume that we have a dataset $\mathcal{D}_{\text{training}}$ of in-distribution data, and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from $\mathcal{D}_{\text{training}}$ or not.
+Since such OOD data is unknown to the model, it is, by definition, a source of epistemic uncertainty.
+This makes OOD detection closely related to uncertainty quantification.
+Only in cases where strong prior knowledge of the data is incorporated into a model, can we expect it to be able to make reliable predictions on OOD data.
+
+Many approaches for OOD detection have been proposed in the literature.
+One way to categorize the different approaches is based on whether the method estimates a conditional probability distribution $p(y|\xb)$ over some target variable given the input, or a probability distribution $p(\xb)$ over the input itself.
+We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \cite{graham_denoising_2023,liu_unsupervised_2023a}.
+Supervised OOD detection tries to assess whether the model's prediction $p(y|\xb)$ should be trusted for a given input $\xb$, whereas unsupervised OOD detection judges whether the input $\xb$ should be trusted and used at all.
+
+\subsection{Supervised out-of-distribution detection}
+% OUTLINE:
 % - Supervised OODD:
@@
 %       - 3 ODIN \cite{liang_enhancing_2018} (Uses synthetic OOD data by adding noise to in-distribution)
-%       - 4 Mahalanobis distance to class conditional Gaussian distribution \cite{lee_training_2018}
+%       - 4 Mahalanobis distance to class conditional Gaussian distribution \cite{lee_simple_2018}
 %       - 5 DeepEnsemble \cite{lakshminarayanan_simple_2017} The classifier-based ensemble method that uses the average of the predictions from multiple independently trained models with random initialization of network parameters and random shuffling of training inputs
@@
 \cite{alemi_deep_2017}
+%
+% - ViM: Out-Of-Distribution with Virtual-logit Matchinghttps://arxiv.org/pdf/2203.10807.pdf in cludes a short overview of many modern methods
+%
+In the supervised setting the supervision can come from different sources depending on how much we know, or assume to know, about the OOD data. 
+The most informed approaches assume we have access to representative OOD data such that we can modify the training procedure to explicitly model the OOD data. 
+Such approaches have achieved high performance, but are inherently limited by the availability of such OOD data and how well it covers the large variation in OOD data \cite{hendrycks_deep_2019,ruff_deep_2020}.
+As we discussed earlier, for many modern applications of machine learning, the data is high dimensional and complex which makes sufficiently sampling the space of out-of-distribution data intractable.
+
+A number of less informed methods do not require access to actual OOD data but instead use synthetic OOD data. 
+Several methods synthesize this OOD data by adding noise to in-distribution data \cite{liang_enhancing_2018, lee_simple_2018, ren_likelihood_2019}. 
+For instance, \textcite{ren_likelihood_2019} propose a number of baselines including training a binary classifier to distinguish between original and perturbed in-distribution data, using an additional class trained to predict perturbed in-distribution data, and training the predicted class distribution to output uniform distribution for perturbed in-distribution inputs \cite{dhamija_reducing_2018}.
+In ODIN \cite{liang_enhancing_2018}, the authors propose to add gradient-based, input-dependent noise to the inputs of a pre-trained classifier and use the calibrated maximum class probability of the predictive distribution $p(y|\xb)$ to evaluate whether the example is in-distribution or not.
+This method does not require retraining the classifier and generally improves on the previous, simple methods.
+A different approach generates OOD inputs using a GAN \cite{lee_training_2018}. 
+As the most informed approaches, this group of methods is also limited by the intractability of sampling the space of OOD data.
+
+The least informed supervised OOD detection methods do not require instantiations of real or synthetic OOD data. 
+A generalized version of ODIN \cite{hsu_generalized_2020} removes the need for simulating OOD data by following a probabilistic motivation and computing output logits as a fraction between two carefully designed network branches.
+The authors note that most current methods make a closed world assumption and implicitly condition on the in-domain $d_\text{in}$ in the form of the predictive distribution $p(y|\xb, d_\text{in})$.
+With this observation, the authors decompose the $p(y|\xb, d_\text{in})$ into a joint class-domain probability and a domain probability,
+\begin{equation}
+    p(y|d_\text{in},\xb) = \frac{p(y, d_\text{in}|\xb)}{p(d_\text{in}|\xb)} \enspace .
+\end{equation}
+
+\todo[inline]{This here is WIP}
+
+One particularly simple approach uses the maximum class probability of $p(y|\xb)$ to evaluate whether the example is in-distribution or not by noting that it tends to be larger for correctly classified examples \cite{hendrycks_baseline_2017}.
+Another simple method proposes that a high entropy of $p(y|\xb)$ indicates an OOD $\xb$ \cite{ren_likelihood_2019} while \textcite{hendrycks_scaling_2019} suggested using the maximum logit value and KL-divergence matching.
+
+\textcite{lee_simple_2018} fit a multivariate Gaussian distribution to the activations of the penultimate layer of a pre-trained classifier and use the Mahalanobis distance to this distribution to evaluate whether inputs are in-distribution or not.
+\textcite{lakshminarayanan_simple_2017} propose to use an ensemble of independently trained classifiers to discriminate between in-distribution and OOD data by evaluating the agreement between the classifiers.
+\textcite{huang_importance_2021} compute the gradient of the KL-divergence of the predictive distribution $p(y|\xb)$ to a uniform distribution noting that the magnitude of gradients is higher for in-distribution data than for OOD data.
+
+Other approaches  \cite{devries_learning_2018, huang_mos_2021, zaeemzadeh_outofdistribution_2021}
+
+\cite{liu_energy-based_2020} propose to use an energy-based model to detect OOD data. UNSUPERVISED?
+
+%       - 2 Entropy of predictive distribution p(y|x) \cite{ren_likelihood_2019}
+%       - 6 The log-odds of a binary classifier trained to distinguish between in-distribution inputs from all classes as one class and randomly perturbed in-distribution inputs as the other.
+\cite{ren_likelihood_2019}
+%       - 7 The maximum class probability over K in-distribution classes of a (K + 1)-class classifier where the additional class is perturbed in-distribution. \cite{ren_likelihood_2019}
+%       - 8 The maximum class probability of a K-class classifier for in-distribution classes but the predicted class distribution is explicitly trained to output uniform distribution on perturbed in-distribution inputs.
+\cite{ren_likelihood_2019}.
+This is similar to using simulated OOD inputs from GAN \cite{lee_training_2018} or using auxiliary datasets of outliers \cite{hendrycks_deep_2019} for calibration purpose.
+
+%       - 3 ODIN \cite{liang_enhancing_2018} (Uses synthetic OOD data by adding noise to in-distribution)
+%       - 4 Mahalanobis distance to class conditional Gaussian distribution \cite{lee_training_2018}
+%       - 5 DeepEnsemble \cite{lakshminarayanan_simple_2017} The classifier-based ensemble method that uses the average of the predictions from multiple independently trained models with random initialization of network parameters and random shuffling of training inputs
+
+%   - Variational information bottleneck (VIB) (Alemi et al., 2018b) performs divergence estimation in latent space to detect OoD, but is technically a model-dependent technique because the latent code is trained jointly with the downstream classification task.
+\cite{alemi_deep_2017}
+
+\cite{lakshminarayanan_simple_2017, liang_enhancing_2018 , huang_importance_2021, wang_vim_2022} % lee_simple_2018, hendrycks_baseline_2017
+
+A general weakness of all supervised out-of-distribution detection is that in learning the task-specific model $p(y|\xb)$ a model may discard information about $p(\xb)$ which could be useful for out-of-distribution detection.
+
+either synthetic out-of-distribution data or from real out-of-distribution data.
+Synthetic out-of-distribution data is data that is generated by some process that is known to be out-of-distribution.
+For example, if we are training a model to classify images of cats and dogs, we could generate synthetic out-of-distribution data by adding noise to the images.
+Real out-of-distribution data is data that is known to be out-of-distribution, but is not generated synthetically.
+For example, if we are training a model to classify images of cats and dogs, we could use images of cars as real out-of-distribution data.
+
+\subsection{Unsupervised out-of-distribution detection}
+% OUTLINE:
 % - Unsupervised ODDD
@@
 
-In this section we will introduce out-of-distribution detection, relate it to the concept of uncertainty and to the previous section on information entropy.
-To supplement the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}, we will also provide a brief overview of some of the related methods that have been proposed in the literature.
-
-Out-of-distribution detection is the task of identifying data that is unlikely to originate from the distribution of the training data. 
-In the general case, we assume that we have a dataset $\mathcal{D}_{\text{training}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from $\mathcal{D}_{\text{training}}$ or not.
-Out-of-distribution detection is closely related to the concept of uncertainty.
-Specifically, since such data is unknown to the model, it is, by definition, a source of epistemic uncertainty.
-Only in cases where strong prior knowledge of the data is incorporated into the model, can we expect the model to be able to make predictions on out-of-distribution data.
-
-One way to categorize the many approaches to out-of-distribution detection is based on whether the method estimates a conditional probability distribution $p(y|\xb)$ over some target variable given the input, or a probability distribution $p(\xb)$ over the input itself.
-We might refer to the former as \emph{supervised} out-of-distribution detection, and the latter as \emph{unsupervised} out-of-distribution detection \cite{graham_denoising_2023,liu_unsupervised_2023a}.
-Supervised out-of-distribution detection tries to assess whether the model's prediction $p(y|\xb)$ should be trusted for a given input $\xb$, whereas unsupervised out-of-distribution detection seeks to assess whether the input $\xb$ should be trusted and used at all.
-
-\subsection{Supervised out-of-distribution detection}
-
-In the supervised setting the supervision can come from different sources that result in different levels of informedness. 
-If one assumes access to representative out-of-distribution data, then one can train a model to distinguish between in-distribution and out-of-distribution data. 
-The out-of-distribution data can either be synthetic or real.
-Such approaches have achieved high performance \cite{lee_training_2018,hendrycks_deep_2019,ruff_deep_2020} but require access to out-of-distribution data, which in some sense makes it inherently flawed; it is often not possible to guarantee good enough coverage of out-of-distribution data to cover all possibilities.
-\todo[inline]{Describe an example method or two...}
-
-Less informed approaches only assume access to in-distribution data when learning the task-specific model $p(y|\xb)$. 
-One particularly simple approach requiring only in-distribution data uses the maximum class probability of the predictive distribution $p(y|\xb)$ to evaluate whether the example is in-distribution or not \cite{hendrycks_baseline_2017}.
-Other simple methods presented by \textcite{ren_likelihood_2019} include using the entropy of $p(y|\xb)$, the maximum class probability with an additional class trained to predict perturbed in-distribution data, the maximum class probability where the predicted class distribution is explicitly trained to output uniform distribution on perturbed in-distribution inputs, and the log-odds of a binary classifier trained to distinguish between original and perturbed in-distribution data.
-
-In ODIN \cite{liang_enhancing_2018}, the authors propose to add input-dependent noise to the inputs of a pre-trained classifier and use the calibrated maximum class probability of the predictive distribution $p(y|\xb)$ to evaluate whether the example is in-distribution or not.
-This method does not require retraining the classifier and generally improves on the simpler methods mentioned above.
-In \textcite{lee_training_2018}, the authors fit a multivariate Gaussian distribution to the activations of the penultimate layer of a pre-trained classifier and use the Mahalanobis distance to this distribution to evaluate whether the example is in-distribution or not.
-\todo[inline]{Continue here ...}
-
-%       - 2 Entropy of predictive distribution p(y|x) \cite{ren_likelihood_2019}
-%       - 6 The log-odds of a binary classifier trained to distinguish between in-distribution inputs from all classes as one class and randomly perturbed in-distribution inputs as the other.
-\cite{ren_likelihood_2019}
-%       - 7 The maximum class probability over K in-distribution classes of a (K + 1)-class classifier where the additional class is perturbed in-distribution. \cite{ren_likelihood_2019}
-%       - 8 The maximum class probability of a K-class classifier for in-distribution classes but the predicted class distribution is explicitly trained to output uniform distribution on perturbed in-distribution inputs.
-\cite{ren_likelihood_2019}.
-This is similar to using simulated OOD inputs from GAN \cite{lee_training_2018} or using auxiliary datasets of outliers \cite{hendrycks_deep_2019} for calibration purpose.
-
-%       - 3 ODIN \cite{liang_enhancing_2018} (Uses synthetic OOD data by adding noise to in-distribution)
-%       - 4 Mahalanobis distance to class conditional Gaussian distribution \cite{lee_training_2018}
-%       - 5 DeepEnsemble \cite{lakshminarayanan_simple_2017} The classifier-based ensemble method that uses the average of the predictions from multiple independently trained models with random initialization of network parameters and random shuffling of training inputs
-
-%   - Variational information bottleneck (VIB) (Alemi et al., 2018b) performs divergence estimation in latent space to detect OoD, but is technically a model-dependent technique because the latent code is trained jointly with the downstream classification task.
-\cite{alemi_deep_2017}
-
-\cite{hendrycks_baseline_2017, lakshminarayanan_simple_2017, lliang_enhancing_2018, lee_simple_2018, huang_importance_2021, wang_vim_2022}
-
- either synthetic out-of-distribution data or from real out-of-distribution data.
-Synthetic out-of-distribution data is data that is generated by some process that is known to be out-of-distribution.
-For example, if we are training a model to classify images of cats and dogs, we could generate synthetic out-of-distribution data by adding noise to the images.
-Real out-of-distribution data is data that is known to be out-of-distribution, but is not generated synthetically.
-For example, if we are training a model to classify images of cats and dogs, we could use images of cars as real out-of-distribution data.
-
- requires some examples of out-of-distribution data,
-
- which in some sense makes it inherently flawed.
-As we discussed earlier, for many modern applications of machine learning, the data is high dimensional and complex which makes sufficiently sampling the space of out-of-distribution data intractable and makes it impossible to guarantee robust out-of-distribution detection.
-Another weakness of supervised out-of-distribution detection is that in learning the task-specific model $p(y|\xb)$ a model may discard information about $p(\xb)$ which could be useful for out-of-distribution detection.
-
- \subsection{Unsupervised out-of-distribution detection}
-
 % Assume we have a dataset $\mathcal{D}_{\text{training}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from $\mathcal{D}_{\text{training}}$ or not.
 Given a test point $\xb$, the model outputs a score $s(\xb)$ that indicates whether $\xb$ is in-distribution or out-of-distribution.
-In the context of neural networks, this problem dates back to at least 1994 \cite{bishop_novelty_1994}.
 

### 2023-09-17 chapters/background.tex
Fix order
More work on background and introduction

@@
 
-\section{Out-of-distribution detection}
+\section{Out-of-distribution detection} \label{sec:out-of-distribution-detection}
 % 
@@
 In this section we will introduce out-of-distribution (OOD) detection, relate it to the concept of uncertainty and to the previous section on information entropy. 
-To supplement the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}, we will also provide a brief overview of some of the related methods that have been proposed in the literature.
+To supplement the short related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}, we will then provide a concise review of the literature on out-of-distribution detection.
 
-OOD detection is the task of identifying data that is unlikely to originate from the distribution of the training data and, in the context of neural networks, the task dates back to at least 1994 \cite{bishop_novelty_1994}.
-In the general case, we assume that we have a dataset $\mathcal{D}_{\text{training}}$ of in-distribution data, and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from $\mathcal{D}_{\text{training}}$ or not.
+\subsection{An overview of the task}
+%
+OOD detection is the task of identifying data that is unlikely to originate from the distribution of the training data and, in the context of neural networks, dates back several decades \cite{bishop_novelty_1994, chang_figure_1993}.
+In the literature, the task is also referred to as open-set \cite{bendale_open_2016, cardoso_weightless_2017, panaredabusto_open_2017}, outlier-rejection \cite{xu_deep_2014, mor_confidence_2018}, or selective prediction \cite{geifman_selective_2017}.
+A closely related field is that of network-based uncertainty estimation [14, 10, 36, 20].
+In the general case, we assume that we have a domain of in-distribution data, $\mathcal{D}_{\text{in}}$, and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from that domain or not.
 Since such OOD data is unknown to the model, it is, by definition, a source of epistemic uncertainty.
@@
 
-Many approaches for OOD detection have been proposed in the literature.
+Recently, several approaches for deep neural networks have been developed that formally address the rejection of samples $\xb\notin\mathcal{D}_{\text{in}}$. 
 One way to categorize the different approaches is based on whether the method estimates a conditional probability distribution $p(y|\xb)$ over some target variable given the input, or a probability distribution $p(\xb)$ over the input itself.
-We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \cite{graham_denoising_2023,liu_unsupervised_2023a}.
+We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \cite{graham_denoising_2023,liu_unsupervised_2023a}, although it is important to note that this distinction relates only to the availability of some target value, $y$ - not whether OOD data is available for supervision\footnote{
+    In our overview, we distinguish between supervised and unsupervised out-of-distribution by letting methods be classified as supervised that use any kind of target value $y$, whether it relates to the original task, available OOD data, or both.
+This is the same distinction made by \textcite{graham_denoising_2023} and \textcite{liu_unsupervised_2023a}.
+    However, it is important to note that in other works, the distinction is made based on whether the model is trained on OOD data or not \cite{hendrycks_baseline_2017,liu_energy-based_2020}.
+This difference in nomenclature is currently unresolved in the literature.
+}. 
 Supervised OOD detection tries to assess whether the model's prediction $p(y|\xb)$ should be trusted for a given input $\xb$, whereas unsupervised OOD detection judges whether the input $\xb$ should be trusted and used at all.
+
+Central to any OOD detection method is the ability to assign a score $s(\xb) \in \mathbb{R}$ to a given input $\xb$ that indicates the degree to which the input is likely to be OOD.
+After obtaining such a scoring method, we typically use a validation set to tune a threshold $\tau$ such that $\xb$ is considered OOD if $s(\xb) > \tau$.
+The threshold is typically chosen such that performance on the validation set is above some standard, for instance by imposing constraints on the recall and precision.
+
+There are many ways to obtain such a scoring method. 
+Unsupervised methods often derive the score 
+from the \emph{likelihood} assigned to the input \cite{bishop_novelty_1994,choi_waic_2019,kirichenko_why_2020,ren_likelihood_2019,serra_input_2020,xiao_likelihood_2020,morningstar_density_2021,nalisnick_detecting_2019,bergamin_modelagnostic_2022,maaloe_biva_2019,havtorn_hierarchical_2021},
+from a \emph{reconstruction} of the input $\xb$ \cite{sakurada_anomaly_2014,xia_learning_2015,lyudchik_outlier_2016,zhou_anomaly_2017,chen_outlier_2017,schlegl_unsupervised_2017,zong_deep_2018,li_madgan_2019,graham_denoising_2023,liu_unsupervised_2023a}, or
+from a latent \emph{representation} of the input \cite{denouden_improving_2018,hendrycks_using_2019,ahmadian_likelihoodfree_2021,bergman_classificationbased_2020,tack_csi_2020,sehwag_ssd_2021,xiao_we_2021}.
+% https://arxiv.org/pdf/2302.10326.pdf
+Supervised methods often derive the score 
+from the \emph{probabilities} given by the predictive distribution $p(y|\xb)$ \cite{hendrycks_baseline_2017,hendrycks_scaling_2022} or 
+from the \emph{logits} of $p(y|\xb)$ \cite{hendrycks_2022_scaling,liu_energy-based_2020,}. 
+Other methods use a latent \emph{representation} of the input \cite{lee_simple_2018,li_anomaly_2019,ndiour_outofdistribution_2020,cook_outlier_2020,zaeemzadeh_outofdistribution_2021}, similar to unsupervised methods.
+In the following sections, we provide a more in-depth overview of the different approaches to OOD detection following the above categorization into supervised and unsupervised methods.
+For the supervised methods we further distinguish between whether methods use real OOD data, synthetic OOD data, or no OOD data at all.
 
@@
 In the supervised setting the supervision can come from different sources depending on how much we know, or assume to know, about the OOD data.
-The most informed approaches assume we have access to representative OOD data such that we can modify the training procedure to explicitly model the OOD data. 
-Such approaches have achieved high performance, but are inherently limited by the availability of such OOD data and how well it covers the large variation in OOD data \cite{hendrycks_deep_2019,ruff_deep_2020}.
-As we discussed earlier, for many modern applications of machine learning, the data is high dimensional and complex which makes sufficiently sampling the space of out-of-distribution data intractable.
+In the least informed setting, the supervision is from the target variable $y$ only.
+In the most informed setting we also assume access to representative OOD data.
 
-A number of less informed methods do not require access to actual OOD data but instead use synthetic OOD data. 
+% Actual OOD data
+\paragraph{Methods using real OOD data}
+Methods that use representative OOD data have achieved high performance since they can directly learn to distinguish between in-distribution and OOD data. 
+\textcite{hendrycks_deep_2019} augments the original training objective with a task-dependent outlier exposure loss that aims to make the output logits different for outlier data. 
+In a similar vein, \textcite{dhamija_reducing_2018} propose losses designed to maximize the entropy of $p(y|\xb)$ and decrease feature magnitudes for OOD data sampled from other datasets. 
+\textcite{ruff_deep_2020} use semi-supervised learning and learn representations of in-distribution data that concentrate close to a centroid in latent space while labelled outliers are pushed away from the centroid.
+Other methods including MCD \cite{yu_unsupervised_2019}, NGC \cite{wu_ngc_2021}, and UDG \cite{yang_semantically_2021} utilize external, unlabeled, noisy data to improve OOD detection performance without requiring cleanly labelled OOD examples.
+As we discussed earlier, for many modern applications of machine learning, input data is often high dimensional and complex making sufficiently sampling the space of out-of-distribution data intractable.
+Therefore, methods that rely on samples of OOD data are inherently limited by its availability and how well it covers the large variation in OOD data.
+
+% Synthetic OOD data
+\paragraph{Methods using synthetic OOD data}
+A number of methods do not require access to actual OOD data but synthesize it instead. 
 Several methods synthesize this OOD data by adding noise to in-distribution data \cite{liang_enhancing_2018, lee_simple_2018, ren_likelihood_2019}. 
-For instance, \textcite{ren_likelihood_2019} propose a number of baselines including training a binary classifier to distinguish between original and perturbed in-distribution data, using an additional class trained to predict perturbed in-distribution data, and training the predicted class distribution to output uniform distribution for perturbed in-distribution inputs \cite{dhamija_reducing_2018}.
-In ODIN \cite{liang_enhancing_2018}, the authors propose to add gradient-based, input-dependent noise to the inputs of a pre-trained classifier and use the calibrated maximum class probability of the predictive distribution $p(y|\xb)$ to evaluate whether the example is in-distribution or not.
-This method does not require retraining the classifier and generally improves on the previous, simple methods.
-A different approach generates OOD inputs using a GAN \cite{lee_training_2018}. 
-As the most informed approaches, this group of methods is also limited by the intractability of sampling the space of OOD data.
+For instance, \textcite{ren_likelihood_2019} propose a number of baselines including training a binary classifier to distinguish between original and perturbed in-distribution data.
+They also propose adding an OOD class to softmax classifiers and training it to predict perturbed in-distribution data, or alternatively, training the predicted class distribution to output uniform distribution for perturbed in-distribution inputs.
+With ODIN \cite{liang_enhancing_2018}, the authors propose to calibrate $p(y|\xb)$ with temperature scaling \cite{guo_calibration_2017} and add gradient-based, input-dependent perturbations to the inputs to use the calibrated maximum class probability as the OOD score.
+This method does not require retraining the classifier and generally improves on the baselines of \textcite{ren_likelihood_2019}.
+\textcite{vyas_outofdistribution_2018} train an ensemble of classifiers on different subsets of the training data, with the left out data taken as OOD, and propose novel loss over $p(y|\xb)$ that seeks to maintain a set margin between its average entropy for the OOD and in-distribution examples.
+Another approach generates OOD inputs using a GAN \cite{lee_training_2018}. 
+Similar to actual OOD data, the usefulness of synthetic OOD data is also fundamentally limited by the intractability of sampling the complete space of OOD data.
 
-The least informed supervised OOD detection methods do not require instantiations of real or synthetic OOD data. 
-A generalized version of ODIN \cite{hsu_generalized_2020} removes the need for simulating OOD data by following a probabilistic motivation and computing output logits as a fraction between two carefully designed network branches.
-The authors note that most current methods make a closed world assumption and implicitly condition on the in-domain $d_\text{in}$ in the form of the predictive distribution $p(y|\xb, d_\text{in})$.
-With this observation, the authors decompose the $p(y|\xb, d_\text{in})$ into a joint class-domain probability and a domain probability,
-\begin{equation}
-    p(y|d_\text{in},\xb) = \frac{p(y, d_\text{in}|\xb)}{p(d_\text{in}|\xb)} \enspace .
+% No OOD data, probs
+\paragraph{Methods not using OOD data}
+The least informed supervised OOD detection methods do not require instantiations of any kind of OOD data, real or synthetic. 
+A baseline approach uses the maximum class probability of $p(y|\xb)$ directly by noting that it tends to be larger for correctly classified examples \cite{hendrycks_baseline_2017}.
+Another baseline method proposes that a high entropy of $p(y|\xb)$ indicates an OOD input \cite{ren_likelihood_2019}
+Other methods that derive the score from the classifier probabilities include \textcite{lakshminarayanan_simple_2017} who propose to use an ensemble of independently trained classifiers to discriminate between in-distribution and OOD data by evaluating the agreement between the classifiers, \textcite{devries_learning_2018} who augment the network with a confidence estimation branch that learns to estimate the confidence of the classifier separately from the probability, and \textcite{huang_importance_2021} who compute the gradient of the KL-divergence of the predictive distribution $p(y|\xb)$ to a uniform distribution noting that the magnitude of gradients is higher for in-distribution data than for OOD data.
+The Variational Information Bottleneck \cite{alemi_deep_2017} jointly learns a probabilistic latent representation, $\zb$ and $p(y|\xb)$, using a generalized variational autoencoder \cite{kingma_autoencoding_2014} that tries to maximize the mutual information between $\zb$ and $y$.
+
+\textcite{hsu_generalized_2020} proposes a generalized version of ODIN that removes the need for simulating OOD data.
+The authors note that most current methods make a closed world assumption and implicitly condition on the in-domain $\mathcal{D}_{\text{in}}$ in the form of the predictive distribution $p(y|\xb, \mathcal{D}_{\text{in}})$.
+With this observation, the authors decompose the $p(y|\xb, \mathcal{D}_{\text{in}})$ into a joint class-domain probability and a domain probability,
+%
+\begin{equation} \label{eq: joint class-domain and domain probability}
+    p(y|\xb,\mathcal{D}_{\text{in}}) = \frac{p(y, \mathcal{D}_{\text{in}}|\xb)}{p(\mathcal{D}_{\text{in}}|\xb)} \enspace .
 \end{equation}
+%
+Without data from the out-domain, it is not possible to directly learn either $p(y, \mathcal{D}_{\text{in}}|\xb)$ or $p(\mathcal{D}_{\text{in}}|\xb)$.
+Instead, the authors use this observation to impose the inductive bias of predicting logits as a fraction between two carefully designed network branches, imitating the form of \cref{eq: joint class-domain and domain probability}.
 
-\todo[inline]{This here is WIP}
+% No OOD data, logits
+A number of works have noted that the maximum softmax probability is not generally a reliable score for OOD detection \cite{hendrycks_scaling_2022,liu_energy-based_2020}. 
+\textcite{liu_energy-based_2020} make an interesting argument as to why based on the energy $E(\xb; f)$ of a softmax classifier $f(\xb)$ \cite{lecun_tutorial_2006},
+%
+\begin{equation} \label{eq:softmax-classifier-energy}
+    E(\xb; f) = - \log \sum_{i=1}^K \exp \left( f_i(\xb) \right) \enspace .
+\end{equation}
+%
+Specifically, the authors relate the maximum softmax probability,
+%
+\begin{align} \label{eq:softmax-classifier-maximum-probability}
+    \max_y p(y|\xb;f) 
+    = \max_y \frac{\exp f_y(\xb)}{\sum_{i=1}^K \exp f_i(\xb)} 
+    = \frac{\exp f^\text{max}(\xb)}{\sum_{i=1}^K \exp f_i(\xb)} \enspace ,
+\end{align}
+%
+to the energy by noting that,
+%
+\begin{align}
+    \log \max_y p(y|\xb;f) = E(\xb; f(\xb) - f^\text{max}(\xb)) = E(\xb; f) - f^\text{max}(\xb) \enspace .
+\end{align}
+%
+This shows that log of the softmax confidence score is equivalent to the special case of the energy score, where all the logits are shifted by their maximum logit value. 
+The authors empirically show that $E(\xb;f)$ is a more reliable score for OOD detection than the maximum softmax probability and note that the energy $E(\xb; f)$ tends to be larger for OOD data than for in-distribution data - contrary to the maximum softmax probability.
+They conclude that this shift results in $\max_y p(y|\xb;f)$ being a biased score for OOD detection and propose instead to use the energy directly as the OOD score.
+This energy-score is improved in ReAct by feature clipping \cite{sun_react_2021}.
 
-One particularly simple approach uses the maximum class probability of $p(y|\xb)$ to evaluate whether the example is in-distribution or not by noting that it tends to be larger for correctly classified examples \cite{hendrycks_baseline_2017}.
-Another simple method proposes that a high entropy of $p(y|\xb)$ indicates an OOD $\xb$ \cite{ren_likelihood_2019} while \textcite{hendrycks_scaling_2019} suggested using the maximum logit value and KL-divergence matching.
+% No OOD, feature
+The final category of supervised methods derive the score from a latent representation of the input. 
+Some methods define OOD classes.
+\textcite{huang_mos_2021} groups the classes of the target variable $y$ and defines an OOD class for each group.
+Each training example is then the correct target for one group and an OOD example for all other groups; a kind of hierarchical version of the OOD class of simpler baselines based on noise augmentation \cite{ren_likelihood_2019}.
+Similarly, to represent a virtual OOD class, \textcite{wang_vim_2022} generates an additional logit by first computing the residual of the input's latent space representation against a principal feature space and then converting it to a valid logit by matching its mean over training samples to the average maximum logits.
 
-\textcite{lee_simple_2018} fit a multivariate Gaussian distribution to the activations of the penultimate layer of a pre-trained classifier and use the Mahalanobis distance to this distribution to evaluate whether inputs are in-distribution or not.
-\textcite{lakshminarayanan_simple_2017} propose to use an ensemble of independently trained classifiers to discriminate between in-distribution and OOD data by evaluating the agreement between the classifiers.
-\textcite{huang_importance_2021} compute the gradient of the KL-divergence of the predictive distribution $p(y|\xb)$ to a uniform distribution noting that the magnitude of gradients is higher for in-distribution data than for OOD data.
+Other methods note that the difficulty of detecting OOD data might be attributed to the curse of dimensionality in the learned feature spaces and propose to use dimensionality reduction techniques.
+In \textcite{ndiour_outofdistribution_2020}, apply dimensionality reduction on learned, high-dimensional features to capture the true feature subspace and compute the norm of the difference between the original feature and the pre-image of its low-dimensional manifold embedding.
+\textcite{zaeemzadeh_outofdistribution_2021} forces the ID samples to embed into a union of 1-dimensional subspaces during training and computes the minimum angular distance from the feature to the class-wise subspaces.
+NuSA \cite{cook_outlier_2020} uses projects features onto the column space of the classification weight matrix and computes the ratio of the norm the projected and original features.
+\textcite{lee_simple_2018} fit a multivariate Gaussian distribution to the activations of the penultimate layer of a pre-trained classifier and use the Mahalanobis distance to this distribution to evaluate whether inputs are OOD.
+This method can also be seen as ameliorating the curse of dimensionality by clustering the high-dimensional feature space.
 
-Other approaches  \cite{devries_learning_2018, huang_mos_2021, zaeemzadeh_outofdistribution_2021}
+% Summary and weaknesses of supervised OOD detection approaches
+% Methods that use real OOD data achieve high performance but are limited by the availability of OOD data and how well it covers the large variation in OOD data.
+% A general weakness of all supervised out-of-distribution detection is that in learning the task-specific model $p(y|\xb)$ a model may discard information about $p(\xb)$ which could be useful for out-of-distribution detection.
 
-\cite{liu_energy-based_2020} propose to use an energy-based model to detect OOD data. UNSUPERVISED?
+% This makes them inherently flawed since we can never really be sure that we have good enough coverage of OOD data.
+% Furthermore, the methods that use synthetic OOD data are limited by the intractability of sampling the complete space of OOD data.
 
@@
 
-\cite{lakshminarayanan_simple_2017, liang_enhancing_2018 , huang_importance_2021, wang_vim_2022} % lee_simple_2018, hendrycks_baseline_2017
-
-A general weakness of all supervised out-of-distribution detection is that in learning the task-specific model $p(y|\xb)$ a model may discard information about $p(\xb)$ which could be useful for out-of-distribution detection.
-
-either synthetic out-of-distribution data or from real out-of-distribution data.
+% either synthetic out-of-distribution data or from real out-of-distribution data.
 Synthetic out-of-distribution data is data that is generated by some process that is known to be out-of-distribution.
@@
 
-Continual novelty detection \cite{aljundi_continual_2022}
+Unsupervised methods often derive the score ...
+
+\paragraph{Likelihood-based}
+from the likelihood assigned to the input \cite{bishop_novelty_1994,choi_waic_2019,kirichenko_why_2020,ren_likelihood_2019,serra_input_2020,xiao_likelihood_2020,morningstar_density_2021,nalisnick_detecting_2019,bergamin_modelagnostic_2022,maaloe_biva_2019,havtorn_hierarchical_2021},
+
+\paragraph{Reconstruction-based}
+from a reconstruction of the input $\xb$ \cite{sakurada_anomaly_2014,xia_learning_2015,lyudchik_outlier_2016,zhou_anomaly_2017,chen_outlier_2017,schlegl_unsupervised_2017,zong_deep_2018,li_madgan_2019,graham_denoising_2023,liu_unsupervised_2023a}, or
+
+\paragraph{Representation-based}
+from a latent representation of the input \cite{denouden_improving_2018, hendrycks_using_2019, ahmadian_likelihoodfree_2021,bergman_classificationbased_2020, tack_csi_2020, sehwag_ssd_2021, xiao_we_2021}.
+% https://arxiv.org/pdf/2302.10326.pdf
 
@@
 \section{Variational autoencoders}
+
+VAE \cite{kingma_autoencoding_2014,rezende_stochastic_2014}
+
+Derive ELBO
+
+Importance sampling \cite{burda_importance_2016}
+
+Semi-supervised learning \cite{kingma_semi-supervised_2014,maaloe_semi-supervised_2017,siddharth_learning_2017}
+
+Trouble in paradise \cite{tomczak_trouble_2022}
 

### 2023-09-17 chapters/introduction.tex
More work on background and introduction

@@
     \caption{A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
-The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize the uncertainty of multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \cite{nationaltransportationsafetyboardnhsa_collision_2019}.}
+The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \cite{nationaltransportationsafetyboardnhsa_collision_2019}.}
     % \caption{Aerial view of crash location showing path of pedestrian as she attempted to cross N.
@@
 
-Though this incident is an extreme one, it highlights the challenges of applying machine learning in real-world scenarios and underlines the need for machine learning systems to be robust in rare, adverse or difficult situations.
+Uber is not the only actor facing the challenges of applying machine learning in real-world scenarios.
+A recent report by the Washington Post found that Tesla's Autopilot system has been involved in at least 736 crashes in the last four years with 16 fatalities \cite{siddiqui_17_2023}.
+Although recent statistics published by Waymo, the self-driving car company owned by Google's parent company Alphabet, revealed far fewer crashes, with only 2 of 20 incidents over 1 million miles meeting reporting criteria, it is interesting to note that the cars are all operated in the sunny states of California and Arizona, where driving conditions vary relatively little \cite{hawkins_waymo_2023}.
+While driver-assistance systems undoubtedly help reduce accidents in many case, and has great future potential, the issues of current systems highlight the need for machine learning systems to be robust in rare, adverse or difficult situations.
+
 Recently, European policymakers proposed a set of requirements for applications of machine learning to ensure that they are safe, reliable, and trustworthy \cite{europeancommission_briefing_2021}.
@@
 
-In practice, we are interested in the predictive uncertainty, so we need to include both aleatoric and epistemic uncertainty in the estimate. 
-Since sources of epistemic uncertainty are by definition those sources not represented in the data, it is not possible to model them directly, as it is for aleatoric uncertainty.
-This generally makes epistemic uncertainty more difficult to model and quantify than aleatoric uncertainty.
-Some methods for estimating epistemic uncertainty include Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995}, Monte Carlo dropout \cite{gal_dropout_2016}, and ensembling \cite{lakshminarayanan_simple_2017}.
-Common to these methods is that they represent epistemic uncertainty by an implicit or explicit distribution over the model parameters.
-This is in contrast to aleatoric uncertainty, which is usually represented by a distribution over the data.
-The methods often require multiple forward passes through the model to estimate the predictive uncertainty.
-This makes them computationally expensive and impractical for real-time applications such as speech recognition.
+In practice, we are interested in the predictive uncertainty, so we need to include both aleatoric and epistemic parts in the uncertainty estimate. 
+Since sources of epistemic uncertainty are by definition those not represented in the data, it is not generally possible to model them directly, as it is for aleatoric uncertainty.
+This makes epistemic uncertainty more difficult to model and quantify than aleatoric uncertainty.
+% Methods for estimating epistemic uncertainty include Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995}, ensembling \cite{gal_dropout_2016,lakshminarayanan_simple_2017} and a suite of recent research into out-of-distribution detection (see \cref{sec:out-of-distribution-detection}).
+In contrast to aleatoric uncertainty, which is usually represented by a distribution over the data, epistemic uncertainty is often represented by an implicit or explicit distribution over the model parameters, such as in Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995} and ensemble models \cite{gal_dropout_2016,lakshminarayanan_simple_2017}.
+In most cases, such methods require multiple forward passes through the model to estimate the predictive uncertainty making them computationally expensive and impractical for real-time applications.
+Within the field of out-of-distribution detection, the epistemic uncertainty is not always expressed in model parameters, but sometimes by a distribution over data representations, or output probabilities with special classes for out-of-distribution data.
+In such cases, the uncertainty can be estimated in a single forward pass through the model.
 
@@
 % Finally, a group deals with applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.
-
+%
 This introduction provided a high-level overview of the motivation behind the research project and highlighted important use-cases of speech processing, natural language processing and the importance of uncertainty quantification.

### 2023-09-19 chapters/background.tex
Work on background

@@
 Recently, several approaches for deep neural networks have been developed that formally address the rejection of samples $\xb\notin\mathcal{D}_{\text{in}}$. 
-One way to categorize the different approaches is based on whether the method estimates a conditional probability distribution $p(y|\xb)$ over some target variable given the input, or a probability distribution $p(\xb)$ over the input itself.
+One way to categorize the different approaches is based on whether the underlying model is a classifier that estimates a conditional probability distribution $p(y|\xb)$ over some target variable given the input, or a model that learns a probability density $p(\xb)$ over the input itself.
 We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \cite{graham_denoising_2023,liu_unsupervised_2023a}, although it is important to note that this distinction relates only to the availability of some target value, $y$ - not whether OOD data is available for supervision\footnote{
@@
 }. 
-Supervised OOD detection tries to assess whether the model's prediction $p(y|\xb)$ should be trusted for a given input $\xb$, whereas unsupervised OOD detection judges whether the input $\xb$ should be trusted and used at all.
+Supervised OOD detection tries to assess whether the model's prediction $\hat{y}$ via $p(y|\xb)$ should be trusted for a given input $\xb$, whereas unsupervised OOD detection judges whether the input $\xb$ should be trusted and used at all.
 
 Central to any OOD detection method is the ability to assign a score $s(\xb) \in \mathbb{R}$ to a given input $\xb$ that indicates the degree to which the input is likely to be OOD.
-After obtaining such a scoring method, we typically use a validation set to tune a threshold $\tau$ such that $\xb$ is considered OOD if $s(\xb) > \tau$.
+After defining a score, we typically use a validation set to tune a threshold $\tau$ such that $\xb$ is considered OOD if $s(\xb) > \tau$.
 The threshold is typically chosen such that performance on the validation set is above some standard, for instance by imposing constraints on the recall and precision.
+Many works also evaluate the performance of OOD detection methods using the area under the receiver operating characteristic (AUROC) curve which does not require a threshold to be chosen.
 
-There are many ways to obtain such a scoring method. 
+The many different ways of defining a score $s(\xb)$ can be used to further categorize OOD detection methods.
 Unsupervised methods often derive the score 
@@
 from the \emph{probabilities} given by the predictive distribution $p(y|\xb)$ \cite{hendrycks_baseline_2017,hendrycks_scaling_2022} or 
-from the \emph{logits} of $p(y|\xb)$ \cite{hendrycks_2022_scaling,liu_energy-based_2020,}. 
+from the \emph{logits} of $p(y|\xb)$ \cite{hendrycks_scaling_2022,liu_energy-based_2020}. 
 Other methods use a latent \emph{representation} of the input \cite{lee_simple_2018,li_anomaly_2019,ndiour_outofdistribution_2020,cook_outlier_2020,zaeemzadeh_outofdistribution_2021}, similar to unsupervised methods.
@@
 This method can also be seen as ameliorating the curse of dimensionality by clustering the high-dimensional feature space.
+Finally, Bayesian neural networks have also been proposed for OOD detection although their performance is not yet competitive with other methods \cite{henning_are_2021,dangelo_outofdistribution_2022,nguyen_out_2022}.
 
@@
 
-Unsupervised methods often derive the score ...
+For high-dimensional data, the most successful unsupervised models for OOD detection are deep autoencoders \cite{hinton_reducing_2006} and deep generative models. 
+Deep generative models are a class of models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution. 
+There exist at least six main classes of deep generative models: variational autoencoders \cite{kingma_autoencoding_2014,rezende_stochastic_2014}, flow-based models \cite{dinh_nice_2015,kingma_glow_2018}, autoregressive models \cite{oord_conditional_2016,oord_wavenet_2016}, diffusion models \cite{sohl-dickstein_deep_2015,ho_denoising_2020}, energy-based models \cite{lecun_tutorial_2006,salakhutdinov_efficient_2010,grathwohl_your_2020}, and generative adversarial networks \cite{goodfellow_generative_2014,arjovsky_wasserstein_2017}.
+Besides data generation, common properties of deep generative models include the ability to compute an exact or approximate likelihood of input $p(\xb)$, reconstruct an input $\xb$ from its latent representation $\zb$, and represent an input $\xb$ in a latent space.
+These properties enable various methods for their use in OOD detection.
+
+Other important density estimation methods that have been applied to OOD detection include kernel density estimation \cite{parzen_estimation_1962}, nearest neighbor methods \cite{cover_nearest_1967}, support vector machines \cite{cortes_supportvector_1995,scholkopf_estimating_2001}, and Gaussian mixture models \cite{dempster_maximum_1977}.
+However, compared to deep generative models, these methods are not well suited for high-dimensional data such as images, text or audio, and have had little direct impact on the recent work on OOD detection.
+
+% In the following we will review the literature on unsupervised OOD detection using deep generative models and categorize the different approaches based on whether they derive their OOD score from the likelihood, a reconstruction, or a latent representation of the input.
 
 \paragraph{Likelihood-based}
-from the likelihood assigned to the input \cite{bishop_novelty_1994,choi_waic_2019,kirichenko_why_2020,ren_likelihood_2019,serra_input_2020,xiao_likelihood_2020,morningstar_density_2021,nalisnick_detecting_2019,bergamin_modelagnostic_2022,maaloe_biva_2019,havtorn_hierarchical_2021},
+% \cite{
+%     X bishop_novelty_1994,
+%     X choi_waic_2019,
+%     X kirichenko_why_2020,
+%     X ren_likelihood_2019,
+%     X serra_input_2020,
+%     X xiao_likelihood_2020,
+%     X morningstar_density_2021,
+%     X nalisnick_detecting_2019,
+%     X bergamin_modelagnostic_2022,
+%     X maaloe_biva_2019,
+%     X havtorn_hierarchical_2021
+% },
+%
+Among the deep generative models that compute exact likelihoods are flow-based models, autoregressive models, and diffusion models, whereas variational autoencoders compute a lower bound on the likelihood, energy-based models usually require approximate methods, and generative adversarial networks do not compute an explicit likelihood at all.
+\textcite{bishop_novelty_1994} first proposed to use the likelihood assigned to data by a generative model as a measure for detecting anomalous data.
+Simply put, since the likelihood measures ``how probable the data is'', OOD data is expected to give lower likelihoods than in-distribution data.
+Although, this method originally gave encouraging results, the advent of deep generative models and their application to high-dimensional data has lead many to observe likelihoods for OOD data that are higher than for in-distribution data \cite{choi_waic_2019,nalisnick_detecting_2019,kirichenko_why_2020}.
+Such results have sparked interest in trying to explain this phenomenon and many works have proposed new scores for OOD detection derived from the likelihood.
+
+In \textcite{ren_likelihood_2019}, the authors propose to use the likelihood ratio between a model trained on in-distribution data and a background model trained on perturbed in-distribution data as the OOD score.
+\textcite{serra_input_2020} argue that the failure of DGMs is due to the high-influence that the input complexity has on the likelihood.
+Therefore, they propose to use a general lossless image compression algorithm as a background model.
+\textcite{choi_waic_2019} propose to use the Watanabe information criterion (WAIC) computed from the likelihood \cite{watanabe_algebraic_2009, watanabe_asymptotic_2010}. 
+For variational autoencoders, other work proposes to refit the encoder on a test data example, hypothesizing that the likelihood of OOD data will improve more than for in-distribution data, and use this ``likelihood regret" as the OOD score \cite{xiao_likelihood_2020}.
+\textcite{maaloe_biva_2019} provide initial results that a loosened variational bound on the likelihood, using only encoded representations from the top-most latent variables in a hierarchical variationa autoencoder, can improve OOD detection performance.
+In \textcite{havtorn_hierarchical_2021}, we show that variational autoencoders are surprisingly good at reconstructing OOD data and propose an improved score based on the likelihood ratio of such loosened bounds \textcite{maaloe_biva_2019}.
+
+A different approach is to use the typicality set hypothesis \cite{nalisnick_detecting_2019}.
+The typicality set is the subset of the model full support in data space, where the model samples from, that does not overlap with regions of maximal likelihood.
+\textcite{nalisnick_detecting_2019} propose to use the typicality set as a test statistic for OOD detection while \textcite{morningstar_density_2021} propose to use the related idea of density of states of the model.
+In \textcite{bergamin_modelagnostic_2022} we use Fisher's method \cite{fisher_statistical_1925} to combine Rao's score test statistic \cite{rao_large_1948} with the typicality set test statistic hence including information from both the gradient and the likelihood.
 
 \paragraph{Reconstruction-based}
-from a reconstruction of the input $\xb$ \cite{sakurada_anomaly_2014,xia_learning_2015,lyudchik_outlier_2016,zhou_anomaly_2017,chen_outlier_2017,schlegl_unsupervised_2017,zong_deep_2018,li_madgan_2019,graham_denoising_2023,liu_unsupervised_2023a}, or
+% \cite{
+%     X sakurada_anomaly_2014                 encode and decode data with autoencoders.
+%     X xia_learning_2015                     encode and decode data with autoencoders.
+%     zhou_anomaly_2017                     encode and decode data with autoencoders.
+%     zong_deep_2018                        encode and decode data with autoencoders.
+%     lyudchik_outlier_2016                 encode and decode data with autoencoders.
+%     chen_outlier_2017                     encode and decode data with an ensemble of autoencoders.
+%     schlegl_unsupervised_2017             perform GAN inversion for a data point, and evaluate its reconstruction error and discriminator confidence under the inverted latent variable.
+%     li_madgan_2019                        perform GAN inversion for a data point, and evaluate its reconstruction error and discriminator confidence under the inverted latent variable.
+%     graham_denoising_2023                 leverages diffusion models to reconstruct images at varied diffusion steps, while we mask and inpaint an image repeatedly with fixed steps.
+%     liu_unsupervised_2023a                LMD lifts an image off its original manifold by corrupting it, and maps it towards the in-domain manifold with a diffusion model.
+For an out-of-domain image, the mapped image would have a large distance away from its original manifold, and LMD would identify it as OOD accordingly.
+% }
+%
+A number of methods derive the OOD score from a reconstruction error. 
+Among the first methods are \textcite{sakurada_anomaly_2014} and \textcite{xia_learning_2015} who use deep autoencoders to reconstruct the input and use the reconstruction error as the OOD score. 
+% In \textcite{zhou_anomaly_2017}
 
 \paragraph{Representation-based}
-from a latent representation of the input \cite{denouden_improving_2018, hendrycks_using_2019, ahmadian_likelihoodfree_2021,bergman_classificationbased_2020, tack_csi_2020, sehwag_ssd_2021, xiao_we_2021}.
 % https://arxiv.org/pdf/2302.10326.pdf
+% \cite{
+%     denouden_improving_2018,              leverages the latent variables of an autoencoder, and evaluates the Mahalanobis distance in the latent space along with the data reconstruction error
+%     ahmadian_likelihoodfree_2021,         extracts low-level features from the encoder of an invertible generative model.
+%     hendrycks_using_2019,                 learn a representation over the in-domain data through self-supervised training
+%     bergman_classificationbased_2020,     learn a representation over the in-domain data through self-supervised training
+%     tack_csi_2020,                        learn a representation over the in-domain data through self-supervised training
+%     sehwag_ssd_2021,                      learn a representation over the in-domain data through self-supervised training
+%     xiao_we_2021                          further shows that one can instead use a strong pretrained feature extractor while maintaining comparable performance.
+% }
+%
+
+% \subsubsection{Model agnostic: Related works}
+% Since \textcite{nalisnick_deep_2019} and \textcite{hendrycks_deep_2019}, different test statistics or methodologies for OOD detection using DGMs were proposed.
+Most of the recent solutions were highly influenced by three major lines of work: \emph{typicality set}, \emph{likelihood ratio} test statistics, and \emph{model misestimation}.
+
+% The typicality set hypothesis was introduced by \textcite{nalisnick_detecting_2019} as a possible explanation for the DGMs assigning higher likelihood to OOD data.
+The typicality set is the subset of the model full support where the model samples from and this does not intersect with the region of higher likelihood.
+While the typicality test was introduced for batch-OOD detection, \textcite{morningstar_density_2021} shows that it also works well in the single-sample case.
+This is also confirmed by our own experiments.
+
+% The likelihood ratio test statistic method by \textcite{ren_likelihood_2019} assumes that every input is composed by a background component and a semantic component.
+For OOD detection, only the semantic component matters.
+In addition to a model trained on the in-distribution data, they proposed to train a background model on perturbed inputs data and then for each test example consider as OOD score the likelihood ratio between the two models.
+\textcite{schirrmeister_understanding_2020}, instead, trained the background model on a more general distribution of images by considering 80 million general tiny images.
+Similarly to these approaches, \textcite{serra_input_2020} argued that the failure of DGMs is due to the high-influence that the input complexity has on the likelihood.
+Therefore, they proposed to use a general lossless image compression algorithm as a background model.
+% All these methods, however, require additional knowledge of the OOD data for either choosing an image augmentation procedure to perturb the input data or for choosing a specific compressor.
+
+% Another line of works blame the models themselves and not the test statistics.
+\textcite{zhang_understanding_2021} argued that model misestimation is the main cause of higher likelihood assigned to OOD data.
+This can be due to both the model architecture and the maximum likelihood objective.
+\textcite{kirichenko_why_2020} and \textcite{schirrmeister_understanding_2020} showed that normalizing flows can achieve better OOD performance despite achieving a worse likelihood if one changes some model design choices.
+% %
+% Other works in the literature focused on deriving specific test statistics that work only for a specific model, for example for VAEs \parencite{xiao_likelihood_2020, maaloe_biva_2019, havtorn_hierarchical_2021}, or for normalizing flows \parencite{kirichenko_why_2020, ahmadian_likelihoodfree_2021}.
+
+% %All the solutions we presented can be described in terms of the following two categories: \emph{model-agnostic} vs \emph{model-specific} test statistics and statistics \emph{with OOD assumptions} vs statistics \emph{with no OOD assumptions}.
+In this work, we are mostly interested in model-agnostic scores with no OOD assumptions, i.e.\@ scores that can be directly computed with any models and that do not need any explicit assumption on the OOD data.
+
+% As mentioned in the introduction, we frame the OOD detection problem in terms of statistical tests problem.
+Recently, \textcite{haroush_statistical_2021} showed that adopting hypothesis testing at the layer and channel level of a neural network can be used for OOD detection in the discriminative setting.
+They used both Fisher's method and Simes' method to combine class-conditional $p$-values computed for each convolutional and dense layer of a deep neural network.
+We focus on the unsupervised setting using DGMs and use hypothesis testing on statistics that can be computed on all differentiable DGM.
+As already explained in section \cref{sec_modelagnostic:combination_p_values}, \textcite{morningstar_density_2021} considered the combination of different statistics for OOD detection.
+The main difference with their approach is that we propose statistics that can be applied to any differentiable generative model and combine them by using Fisher's method, which takes advantage of using only one-sided independent statistics.
+Concurrently, \textcite{choi_robust_2021} derived the score statistic by starting from the likelihood ratio statistic and applying a Laplace approximation.
+They computed the score statistic only for certain layers of the model and for a specific example, the OOD score is given by the infinity norm of these different layer scores after a ReLU operation.
+Our procedure differs both in the derivation of the score statistic and its usage since we compute the score statistic for the entire model.
+
+% \subsubsection{Hierarchical: Related works}
+
+% So far, no reliable direct likelihood-based method has been found for fully unsupervised deep generative model OOD detection.
+% A major line of work considers developing new scores that are more reliable than the likelihood.
+% This includes the \textit{typicality} test presented by \textcite{nalisnick_detecting_2019} which is an OOD detection test based on the typicality of a batch of potentially OOD examples.
+% This approach however requires a batch of examples from the same class (OOD or not) which limits its practical applicability.
+% In \textcite{ren_likelihood_2019}, the \textit{likelihood ratio} between a primary model and a background model was shown to be an effective score for OOD detection.
+% However, to train the background model, the in-distribution data is perturbed via a data augmentation technique that is designed with knowledge about the confounding factors between the in-distribution data and the OOD data.
+Furthermore, it is tuned towards high performance on a known OOD dataset.
+% \textcite{serra_input_2020} take a similar approach and attribute the failure to detect OOD data to the high influence of the input complexity on the likelihood and choose a generic lossless compression algorithm as the background model.
+% Although this method gives good results, no single best choice of compression algorithm exists for all types of OOD data, and any particular choice encodes prior knowledge about the data into the detection method.
+% Both these methods can be seen as correcting for low-level features of the OOD data being assigned high model likelihood by using a second model focused exclusively on these features.
+
+% Similar to these methods, the majority of the approaches to OOD detection make assumptions about the nature of the OOD data.
+% The assumptions encompass using labels on the in-distribution data \cite{hendrycks_baseline_2017, liang_enhancing_2018, alemi_uncertainty_2018, lee_simple_2018, lakshminarayanan_simple_2017}, examples of OOD data \cite{hendrycks_deep_2019}, augmenting in-distribution data to mimic it \cite{ren_likelihood_2019}, or assuming a certain data type \cite{serra_input_2020}.
+% Any of these assumptions encode implicit biases into the model about the attributes of OOD data which, in turn, might impair performance on truly unknown data examples (unknown unknowns).
+
+% % Introduce the contrast to results with VAEs
+% While some of these methods achieve very good results on OOD detection with autoregressive models \cite{oord_pixel_2016, salimans_pixelcnn_2017} and invertible flow-based models \cite{kingma_glow_2018}, it was recently shown that they can be much less effective for VAEs \cite{xiao_likelihood_2020} highlighting the need for a more reliable OOD score for VAEs.
+% Although VAEs have the same failure cases as autoregressive and flow-based models, the caveat is that the difference in the likelihood is generally not as big and reconstructions of OOD can be surprisingly good \cite{xiao_likelihood_2020}.
+% \textcite{xiao_likelihood_2020} alleviate this by refitting the inference network, as previously proposed by \textcite{cremer_inference_2018, mattei_refit_2018}, to a potentially OOD example and measuring the so-called \textit{likelihood regret}.
+% However, refitting the inference network can be computationally expensive, especially for the large hierarchical VAEs that are used to model complex data \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
+Furthermore, this scales poorly to large amounts of potentially OOD examples as the optimization is done per example.
+
+% A few methods have approached OOD detection in a completely unsupervised fashion \cite{maaloe_biva_2019, choi_waic_2019, xiao_likelihood_2020}.
+% The work of \textcite{maaloe_biva_2019} is the most related to ours.
+They introduce BIVA, a deep hierarchy of stochastic latent variables with a top-down and bottom-up inference model and achieve state-of-the-art likelihood scores.
+% They also provide early results indicative that a looser likelihood bound may have value in OOD detection.
+% In this paper, we provide an explanation of those results, and significantly improve upon them.
 

### 2023-09-19 chapters/paper_modelagnostic.tex
Work on background

@@
 
-The likelihood ratio statistic, the Wald statistic and the gradient statistic all require to fit a model on the additional datapoints $\tilde{x}_1,\ldots,\tilde{x}_n$ in order to compute either $\ell(\hat{\theta})$ or $\hat{\theta}$.
+The likelihood ratio statistic, the Wald statistic and the gradient statistic all require to fit a model on the additional data points $\tilde{x}_1,\ldots,\tilde{x}_n$ in order to compute either $\ell(\hat{\theta})$ or $\hat{\theta}$.
 %
-%In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint.
+%In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single data point.
 \textcite{xiao_likelihood_2020} try to do that in the context of VAEs by re-fitting only the inference network, or encoder, of a VAE to every single additional example.
 This is the typical thing to do when dealing with out-of-sample data in VAEs, as argued by \textcite{mattei2018refit, cremer2018inference}.
-In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint.
+In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single data point.
 \textcite{xiao_likelihood_2020} did this for a variational autoencoder (VAE, \citealp{kingma_autoencoding_2014,rezende_stochastic_2014}) by only re-fitting inference network (or encoder) to the additional example, which is a typical approach to dealing with out-of-sample data in VAEs, as argued by \textcite{cremer_inference_2018} and \textcite{mattei_refit_2018}.
@@
 
-%In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint, which barely sounds serious.
+%In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single data point, which barely sounds serious.
 Most of the recent works in the literature mainly focus on deriving different versions of what they call a likelihood ratio statistic.
@@
 %
-In the \cref{appendix_modelagnostic:harmonic} , we also consider the Harmonic mean $p$-value \parencite{wilson_harmonic_2019} as a way to combine $p$-values from different statistics.
+In the \cref{appendix_modelagnostic:harmonic}, we also consider the Harmonic mean $p$-value \parencite{wilson_harmonic_2019} as a way to combine $p$-values from different statistics.
 This method usually works best when the statistics are not independent.

### 2023-09-21 appendices/supplementary_retrospective.tex
Some formatting and changes to labels
Update retrospective paper supplementary

--- 2023-08-11 appendices/appendix_retrospective.tex
@@
 \chapter[supplementary material: a retrospective study on machine learning-assisted stroke recognition for medical helpline calls]{Supplementary Material for: A Retrospective Study on Machine Learning-Assisted Stroke Recognition for Medical Helpline Calls}
-\label{app:retrospective}
+\label{app:supplementary-retrospective}
 \ifthenelse{\equal{\skipappendices}{true}}{}{
@@
         \midrule
-        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - specificity)}  & \makecell{0.077 \\ (0.073-0.080)} & \makecell{0.102 \\ (0.097-0.106)} & \makecell{0.034 \\ (0.032-0.036)} & \makecell{0.318 \\ (0.305-0.331)} \\
+        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - NPV)}  & \makecell{0.077 \\ (0.073-0.080)} & \makecell{0.102 \\ (0.097-0.106)} & \makecell{0.034 \\ (0.032-0.036)} & \makecell{0.318 \\ (0.305-0.331)} \\
         \midrule
-        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - NPV)} & \makecell{0.45 \\ (0.446-0.463)} & \makecell{0.483 \\ (0.473-0.494)} & \makecell{0.264 \\ (0.258-0.270)} & \makecell{1.335 \\ (1.308-1.362)} \\
+        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell{0.45 \\ (0.446-0.463)} & \makecell{0.483 \\ (0.473-0.494)} & \makecell{0.264 \\ (0.258-0.270)} & \makecell{1.335 \\ (1.308-1.362)} \\
 
@@
         \midrule
-        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell{0.152 \\ (0.146-0.159)} & \makecell{0.155 \\ (0.147-0.162)} & \makecell{0.058 \\ (0.055-0.062)} & \makecell{0.372 \\ (0.359-0.386)} \\
+        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - NPV)}         & \makecell{0.152 \\ (0.146-0.159)} & \makecell{0.155 \\ (0.147-0.162)} & \makecell{0.058 \\ (0.055-0.062)} & \makecell{0.372 \\ (0.359-0.386)} \\
         \midrule
-        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - NPV)} & \makecell{0.404 \\ (0.394-0.414)} & \makecell{0.480 \\ (0.467-0.494)} & \makecell{0.246 \\ (0.239-0.254)} & \makecell{0.873 \\ (0.852-0.894)} \\
+        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell{0.404 \\ (0.394-0.414)} & \makecell{0.480 \\ (0.467-0.494)} & \makecell{0.246 \\ (0.239-0.254)} & \makecell{0.873 \\ (0.852-0.894)} \\
 
@@
     % }
+\end{table}
+
+\section{Additional results: Model with patient age and sex as explicit inputs}
+
+To assess the importance of patient age and sex for accurate model stroke recognition we have performed an experiment to test whether explicitly adding the age and sex of the patient as inputs to the model improves performance.
+
+Specifically, we encode patient sex as two binary numbers which we concatenate to the bag-of-words input representation.
+The first has the value of 1 if the patient is female, and the second has the value of 1 if the patient is male.
+Neither is 1 if the patient's sex is unknown/undisclosed.
+Similarly, we include patient age as a 15-dimensional one-hot vector where the first value represents patients with an age below 25 and the last value represents patients with an age of 90 or above.
+Values in between represent 5-year intervals.
+
+We trained the model similarly to how we performed our ablation experiments.
+That is, we used the hyperparameters that gave the best performance on the validation set to train 11 differently seeded versions of the model and then report mean metrics and associated CIs on the MH-1813 test data.
+The results are as shown in \cref{tab_retrospective:tableA4} below.
+
+\begin{table}[h]
+    \centering
+    \caption{Overall performance on MH-1813 test data for the model that also takes patient age and sex as direct inputs.
+We also list the original performance of call-takers and the model (w/o sex and age) from the main manuscript for ease of comparison [mean (95\% CI)].}
+    \label{tab_retrospective:tableA4}
+    \resizebox*{0.98\textwidth}{!}{%
+    \begin{tabular}{l|ccccc}
+        \toprule
+        & F1-score [\%] $\uparrow$ & Sensitivity [\%] $\uparrow$ & PPV [\%] $\uparrow$ & \makecell[c]{FOR [\%] $\downarrow$ \\ (1 - NPV)} & \makecell[c]{FPR [\%] $\downarrow$ \\ (1 - specificity)} \\
+        \midrule
+        & \multicolumn{5}{c}{\textit{Overall}} \\
+        \midrule
+
+        Call-takers                             & \makecell[c]{25.8 \\ (23.7-27.9)} & \makecell[c]{52.7 \\ (49.2-56.4)} & \makecell[c]{17.1 \\ (15.5-18.6)} & \makecell[c]{0.105 \\ (0.094-0.116)} & \makecell[c]{0.565 \\ (0.539-0.590)} \\
+        \midrule
+        \makecell[l]{Model \\ w/o sex and age}  & \makecell[c]{35.7 \\ (35.0-36.4)} & \makecell[c]{63.0 \\ (62.0-64.1)} & \makecell[c]{24.9 \\ (24.3-25.5)} & \makecell[c]{0.082 \\ (0.079-0.085)} & \makecell[c]{0.419 \\ (0.413-0.426)} \\
+        \midrule
+        \makecell[l]{Model \\ w/ sex and age}   & \makecell[c]{35.8 \\ (35.1-36.5)} & \makecell[c]{64.1 \\ (63.1-65.1)} & \makecell[c]{24.9 \\ (24.3-25.4)} & \makecell[c]{0.080 \\ (0.077-0.082)} & \makecell[c]{0.427 \\ (0.421-0.434)} \\
+
+        \bottomrule
+    \end{tabular}%
+    }
+\end{table}
+
+We note that the performance difference between the original model and the models with age and sex inputs is statistically insignificant (p > 0.05).
+We hypothesize this might be because information regarding the age and sex of the patient is, in many cases, already present in the transcript.
+Another reason might be that the patient's age and sex are less useful discriminators than other indicative factors.
+This latter hypothesis is supported by our occlusion analysis and \cref{tab_retrospective:table3} where no highly ranked words directly refer to the patient's age or sex.
+
+It is important to note that in practice the information about patient sex and age is extracted from the patient's CPR number which is typically entered by the patient themselves while queuing for MH-1813.
+However, at call lines in countries without similar systems in place, such information may not be readily available and cannot be incorporated as an explicit input feature.
+The same is the case for the 1-1-2 call line in Denmark.
+
+\section{Additional results: Model without MH-1813 training data}
+%
+The ablation study in the main manuscript that examines the importance of the two different source domains (1-1-2 and MH-1813) includes only two of the three possible combinations of data sources, specifically: training on both 1-1-2 and MH-1813 data, and training only on MH-1813 excluding 1-1-2 data.
+For this reason, we have conducted an experiment that includes the remaining combination: training only on 1-1-2 excluding MH-1813 data.
+
+Since the manuscript focuses on the MH-1813 line, the main purpose of experimenting with including and excluding the 1-1-2 data was to examine whether using the out-of-domain 1-1-2 data could improve stroke recognition performance of the machine learning framework on the MH-1813 data.
+This has interest since many prehospital call centres operate both a high acuity emergency line (like 1-1-2) and a low-acuity medical helpline (like MH-1813) which makes high-acuity data available for modelling and a potentially valuable data source.
+The ablation experiment performed here, on the other hand, has a different purpose.
+It aims to show whether a model to assist the MH-1813 helpline could be developed also in the hypothetical case of only having access to out-of-domain, high-acuity training data from 1-1-2.
+This prospect may be interesting for some call centres that, for instance, have only recently started operating a medical helpline, and so, do not have in-domain training data available.
+
+We trained the models and provided the results of this experiment in \cref{tab_retrospective:tableA5} below using the same methods as in the main manuscript.
+We tested the significance of these results using the same statistical significance testing methods as used and described in the main manuscript.
+
+We see that the performance of the ensemble model trained only with 1-1-2 data compared to training only with MH-1813 data was worse in terms of sensitivity and FOR (p < 0.0001), but on-par in terms of F1-score and PPV, and better in terms of FPR (p < 0.0001).
+Compared to training with all data, training only on 1-1-2 was worse on all metrics (p < 0.0001) except FPR, where it was better (p < 0.0001).
+We note that the model still performed better than the call-takers in terms of F1-score, PPV, and FPR (p < 0.0001) and was on-par in terms of sensitivity and FOR.
+
+In summary, training on only 1-1-2 data was only somewhat worse than training on only MH-1813 data, and still outperformed call-takers to some degree.
+This indicates that the domain shift between different call lines, even with different acuity levels, is small enough that naive domain transfer of models works fairly well.
+
+\begin{table}[h]
+    \centering
+    \caption{Overall performance on MH-1813 test data, performance without 1-1-2 training data, and performance without 1813 training data [mean (95\% CI)].}
+    \label{tab_retrospective:tableA5}
+    \resizebox*{0.98\textwidth}{!}{%
+    \begin{tabular}{l|ccccc}
+        \toprule
+        & F1-score [\%] $\uparrow$ & Sensitivity [\%] $\uparrow$ & PPV [\%] $\uparrow$ & \makecell[c]{FOR [\%] $\downarrow$ \\ (1 - NPV)} & \makecell[c]{FPR [\%] $\downarrow$ \\ (1 - specificity)} \\
+        \midrule
+        & \multicolumn{5}{c}{\textit{Overall}} \\
+        \midrule
+
+        Call-takers  & \makecell[c]{25.8 \\ (23.7-27.9)} & \makecell[c]{52,7 \\ (49,2-56.4)} & \makecell[c]{17.1 \\ (15.5-18.6)} & \makecell[c]{0.105 \\ (0.094-0.116)} & \makecell[c]{0.565 \\ (0.539-0.590)} \\
+        \midrule
+        Model        & \makecell[c]{35.7 \\ (35.0-36.4)} & \makecell[c]{63.0 \\ (62.0-64.1)} & \makecell[c]{24.9 \\ (24.3-25.5)} & \makecell[c]{0.082 \\ (0.079-0.085)} & \makecell[c]{0.419 \\ (0.413-0.426)} \\
+        \midrule
+        & \multicolumn{5}{c}{\textit{Without 1-1-2 training data (only MH-1813 data)}} \\
+        \midrule
+
+        Model        & \makecell[c]{32.4 \\ (31.8-33.1)} & \makecell[c]{60.4 \\ (59.3-61.4)} & \makecell[c]{22.2 \\ (21.6-22.7)} & \makecell[c]{0.088 \\ (0.085-0.091)} & \makecell[c]{0.467 \\ (0.460-0.474)} \\
+
+        \midrule
+        & \multicolumn{5}{c}{\textit{Without MH-1813 training data (only 1-1-2 data)}} \\
+        \midrule
+
+        Model        & \makecell[c]{31.4 \\ (30.7-32.1)} & \makecell[c]{50.4 \\ (49.3-51.4)} & \makecell[c]{22.8 \\ (22.2-23.4)} & \makecell[c]{0.110 \\ (0.106-0.113)} & \makecell[c]{0.375 \\ (0.369-0.381)} \\
+
+        \bottomrule
+    \end{tabular}%
+    }
 \end{table}
@@
     \caption{Mean impact for words with the largest positive rank score in calls predicted as stroke.}
-    \label{tab_retrospective:tableA4}
+    \label{tab_retrospective:tableA6}
     \resizebox*{\textwidth}{!}{%
@@
     \caption{Mean impact for words with the largest negative rank score in calls predicted as non-stroke.}
-    \label{tab_retrospective:tableA5}
+    \label{tab_retrospective:tableA7}
     \resizebox*{\textwidth}{!}{%
@@
 
+\section{Additional results: Fine-tuning of Danish BERT model for stroke recognition}
+%
+During the preliminary experimental phase, we fine-tuned a BERT model pre-trained on Danish text from CommonCrawl, Wikipedia, OpenSubtitles, and other Danish online forums (available at \url{https://github.com/certainlyio/nordic_bert}).
+The model was fine-tuned for 10,000 updates using linear learning rate warm-up (1,000 updates) and decay (9,000 updates). 
+The maximum learning rate was set to $5\times 10^{-5}$ and an accumulated batch size of 128. 
+The maximum sequence length of the pre-trained model was 512 input tokens. 
+To accommodate longer input sequences, which was necessary for our dataset, we concatenated several copies of the original positional embedding matrix. 
+The results of the final model are presented in \cref{tab_retrospective:tableA8}.
+
+We see that the fine-tuned BERT model performs slightly worse across F1-score, sensitivity, PPV and FOR (p < 0.0001), but better in terms of FPR (p < 0.0001), compared the MLP model presented in the main manuscript.
+As described in the discussion section, we hypothesize that the number of stroke positives was too small for these advanced models to learn more complex patterns than the MLP ensemble.
+In addition, the BERT model would likely benefit from pre-training on text data from the target domain, or a domain close to it, rather than various online fora.
+
+\begin{table}[h]
+    \centering
+    \caption{Overall performance on MH-1813 test data for the model that also takes patient age and sex as direct inputs.
+We also list the original performance of call-takers and the model (w/o sex and age) from the main manuscript for ease of comparison [mean (95\% CI)].}
+    \label{tab_retrospective:tableA8}
+    \resizebox*{0.98\textwidth}{!}{%
+    \begin{tabular}{l|ccccc}
+        \toprule
+        & F1-score [\%] $\uparrow$ & Sensitivity [\%] $\uparrow$ & PPV [\%] $\uparrow$ & \makecell[c]{FOR [\%] $\downarrow$ \\ (1 - NPV)} & \makecell[c]{FPR [\%] $\downarrow$ \\ (1 - specificity)} \\
+        \midrule
+        & \multicolumn{5}{c}{\textit{Overall}} \\
+        \midrule
+
+        Call-takers                             & \makecell[c]{25.8 \\ (23.7-27.9)} & \makecell[c]{52.7 \\ (49.2-56.4)} & \makecell[c]{17.1 \\ (15.5-18.6)} & \makecell[c]{0.105 \\ (0.094-0.116)} & \makecell[c]{0.565 \\ (0.539-0.590)} \\
+        \midrule
+        \makecell[l]{MLP}                       & \makecell[c]{35.7 \\ (35.0-36.4)} & \makecell[c]{63.0 \\ (62.0-64.1)} & \makecell[c]{24.9 \\ (24.3-25.5)} & \makecell[c]{0.082 \\ (0.079-0.085)} & \makecell[c]{0.419 \\ (0.413-0.426)} \\
+        \midrule
+        \makecell[l]{BERT \\ (fine-tuned)}      & \makecell[c]{33.8 \\ (31.5-36.2)} & \makecell[c]{57.5 \\ (53.9-60.9)} & \makecell[c]{23.9 \\ (21.9-25.9)} & \makecell[c]{0.094 \\ (0.084-0.104)} & \makecell[c]{0.403 \\ (0.381-0.424)} \\
+
+        \bottomrule
+    \end{tabular}%
+    }
+\end{table}
+
 \section{Simulation of a prospective study on 2021 data}
@@
 \begin{enumerate}[label=\Roman*.]
-    \item Presentation of the model prediction to the call taker.
+    \item When is the model prediction presented to the call taker?
     \item How does prediction influence the diagnostic code the call taker assigns to the call?
@@
 
-As expected (\cref{tab_retrospective:tableA6}), method 2.C (raw model predictions during calls) yielded slightly worse results than 1.C (raw model predictions after calls).
+As expected (\cref{tab_retrospective:tableA9}), method 2.C (raw model predictions during calls) yielded slightly worse results than 1.C (raw model predictions after calls).
 Compared with method C, method A (call takers mirror model positives) led to increased sensitivity and decreased PPV, whereas method B (call takers mirror model negatives) led to decreased sensitivity and increased PPV, as expected.
@@
     \caption{Overall performance of model, call takers and simulated combinations of model and call takers on MH-1813 test data.}
-    \label{tab_retrospective:tableA6}
+    \label{tab_retrospective:tableA9}
     \resizebox*{0.98\textwidth}{!}{%
@@
 
-        F1-score [\%] $\uparrow$ & \makecell{25.8 \\ (23.7-27.9)} & \makecell{35.7 \\ (35.0-36.4)} & \makecell{33.1 \\ (32.4-33.7)} & \makecell{28.9 \\ (28.3-29.5)} & \makecell{33.3 \\ (32.5-34.1)} & \makecell{27.6 \\ (27.0-28.1)} & \makecell{32.7 \\ (31.8-33.5)} \\
+        \makecell[l]{F1-score [\%] $\uparrow$}                   & \makecell[c]{25.8 \\ (23.7-27.9)} & \makecell[c]{35.7 \\ (35.0-36.4)} & \makecell[c]{33.1 \\ (32.4-33.7)} & \makecell[c]{28.9 \\ (28.3-29.5)} & \makecell[c]{33.3 \\ (32.5-34.1)} & \makecell[c]{27.6 \\ (27.0-28.1)} & \makecell[c]{32.7 \\ (31.8-33.5)} \\
         \midrule
-        Sensitivity [\%] $\uparrow$ & \makecell{52.7 \\ (49.2-56.4)} & \makecell{63.0 \\ (62.0-64.1)} & \makecell{58.7 \\ (57.7-59.8)} & \makecell{72.4 \\ (71.5-73.3)} & \makecell{43.4 \\ (42.3-44.5)} & \makecell{72.3 \\ (71.4-73.3)} & \makecell{39.1 \\ (38.1-40.1)} \\
+        \makecell[l]{Sensitivity [\%] $\uparrow$}                & \makecell[c]{52.7 \\ (49.2-56.4)} & \makecell[c]{63.0 \\ (62.0-64.1)} & \makecell[c]{58.7 \\ (57.7-59.8)} & \makecell[c]{72.4 \\ (71.5-73.3)} & \makecell[c]{43.4 \\ (42.3-44.5)} & \makecell[c]{72.3 \\ (71.4-73.3)} & \makecell[c]{39.1 \\ (38.1-40.1)} \\
         \midrule
-        PPV [\%] $\uparrow$ & \makecell{17.1 \\ (15.5-18.6)} & \makecell{24.9 \\ (24.3-25.5)} & \makecell{23.0 \\ (22.5-23.6)} & \makecell{18.0 \\ (17.6-18.4)} & \makecell{27.0 \\ (26.3-27.8)} & \makecell{17.0 \\ (16.7-17.4)} & \makecell{28.1 \\ (27.3-28.9)} \\
+        \makecell[l]{PPV [\%] $\uparrow$}                        & \makecell[c]{17.1 \\ (15.5-18.6)} & \makecell[c]{24.9 \\ (24.3-25.5)} & \makecell[c]{23.0 \\ (22.5-23.6)} & \makecell[c]{18.0 \\ (17.6-18.4)} & \makecell[c]{27.0 \\ (26.3-27.8)} & \makecell[c]{17.0 \\ (16.7-17.4)} & \makecell[c]{28.1 \\ (27.3-28.9)} \\
         \midrule
-        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell{0.105 \\ (0.094-0.116)} & \makecell{0.082 \\ (0.079-0.085)} & \makecell{0.091 \\ (0.088-0.094)} & \makecell{0.061 \\ (0.059-0.064)} & \makecell{0.125 \\ (0.121-0.129)} & \makecell{0.061 \\ (0.059-0.064)} & \makecell{0.134 \\ (0.131-0.138)} \\
+        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - NPV)}         & \makecell[c]{0.105 \\ (0.094-0.116)} & \makecell[c]{0.082 \\ (0.079-0.085)} & \makecell[c]{0.091 \\ (0.088-0.094)} & \makecell[c]{0.061 \\ (0.059-0.064)} & \makecell[c]{0.125 \\ (0.121-0.129)} & \makecell[c]{0.061 \\ (0.059-0.064)} & \makecell[c]{0.134 \\ (0.131-0.138)} \\
         \midrule
-        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - NPV)} & \makecell{0.565 \\ (0.539-0.590)} & \makecell{0.419 \\ (0.413-0.426)} & \makecell{0.432 \\ (0.426-0.439)} & \makecell{0.726 \\ (0.717-0.735)} & \makecell{0.258 \\ (0.253-0.263)} & \makecell{0.776 \\ (0.767-0.786)} & \makecell{0.221 \\ (0.216-0.226)} \\
+        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell[c]{0.565 \\ (0.539-0.590)} & \makecell[c]{0.419 \\ (0.413-0.426)} & \makecell[c]{0.432 \\ (0.426-0.439)} & \makecell[c]{0.726 \\ (0.717-0.735)} & \makecell[c]{0.258 \\ (0.253-0.263)} & \makecell[c]{0.776 \\ (0.767-0.786)} & \makecell[c]{0.221 \\ (0.216-0.226)} \\
 

### 2023-09-22 appendices/supplementary_retrospective.tex
Updates to table formatting retrospective

@@
 \label{app:supplementary-retrospective}
-\ifthenelse{\equal{\skipappendices}{true}}{}{
+% \ifthenelse{\equal{\skipappendices}{true}}{}{
 
@@
 \begin{equation}
-    p^{(d)} = \frac{1}{N} \sum_{n=1}^N \sigma(z^{(n,d)} - t^{(n)}) \enspace ,
+    p^{(d)} = \frac{1}{N} \sum_{n=1}^N \sigma\left(z^{(n,d)} - t^{(n)}\right) \enspace ,
 \end{equation}
@@
 \begin{equation}
-    s^{(d)} = I_{p^{(d)} > 0.5}(p^{(d)}) \enspace ,
+    s^{(d)} = I_{p^{(d)} > 0.5}\left(p^{(d)}\right) \enspace ,
 \end{equation}
@@
     \centering
-    \caption{Model performance based on sex and age [mean (95\% CI)] when 1-1-2 training data is not used for training.}
+    \caption{Model performance for sex and age subgroups [mean (95\% CI)] on the MH-1813 test data when 1-1-2 training data is not used for training.}
     \label{tab_retrospective:tableA2}
-    % \resizebox*{0.98\textwidth}{!}{%
+    \resizebox*{0.85\textwidth}{!}{%
     \begin{tabular}{l|cc|cc}
@@
 
-         & \multicolumn{2}{c|}{Model} & \multicolumn{2}{c}{Model} \\
+         & \multicolumn{4}{c|}{Model w/o 1-1-2 training data} \\
         \midrule
@@
 
-        F1-score [\%] $\uparrow$ & \makecell{28.7 \\ (27.8-29.6)} & \makecell{36.2 \\ (35.3-37.2)} & \makecell{20.5 \\ (19.5-21.5)} & \makecell{39.4 \\ (38.6-40.3)} \\
+        F1-score [\%] $\uparrow$ & \makecell[c]{28.7 \\ (27.8-29.6)} & \makecell[c]{36.2 \\ (35.3-37.2)} & \makecell[c]{20.5 \\ (19.5-21.5)} & \makecell[c]{39.4 \\ (38.6-40.3)} \\
         \midrule
-        Sensitivity [\%] $\uparrow$ & \makecell{58.2 \\ (56.7-59.9)} & \makecell{62.1 \\ (60.7-63.6)} & \makecell{53.0 \\ (50.9-55.0)} & \makecell{63.0 \\ (61.9-64.3)} \\
+        Sensitivity [\%] $\uparrow$ & \makecell[c]{58.2 \\ (56.7-59.9)} & \makecell[c]{62.1 \\ (60.7-63.6)} & \makecell[c]{53.0 \\ (50.9-55.0)} & \makecell[c]{63.0 \\ (61.9-64.3)} \\
         \midrule
-        PPV [\%] $\uparrow$ & \makecell{19.0 \\ (18.3-19.7)} & \makecell{25.6 \\ (24.8-26.4)} & \makecell{12.7 \\ (12.0-13.4)} & \makecell{28.7 \\ (27.9-29.5)} \\
+        PPV [\%] $\uparrow$ & \makecell[c]{19.0 \\ (18.3-19.7)} & \makecell[c]{25.6 \\ (24.8-26.4)} & \makecell[c]{12.7 \\ (12.0-13.4)} & \makecell[c]{28.7 \\ (27.9-29.5)} \\
         \midrule
-        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - NPV)}  & \makecell{0.077 \\ (0.073-0.080)} & \makecell{0.102 \\ (0.097-0.106)} & \makecell{0.034 \\ (0.032-0.036)} & \makecell{0.318 \\ (0.305-0.331)} \\
+        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - NPV)}  & \makecell[c]{0.077 \\ (0.073-0.080)} & \makecell[c]{0.102 \\ (0.097-0.106)} & \makecell[c]{0.034 \\ (0.032-0.036)} & \makecell[c]{0.318 \\ (0.305-0.331)} \\
         \midrule
-        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell{0.45 \\ (0.446-0.463)} & \makecell{0.483 \\ (0.473-0.494)} & \makecell{0.264 \\ (0.258-0.270)} & \makecell{1.335 \\ (1.308-1.362)} \\
+        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell[c]{0.45 \\ (0.446-0.463)} & \makecell[c]{0.483 \\ (0.473-0.494)} & \makecell[c]{0.264 \\ (0.258-0.270)} & \makecell[c]{1.335 \\ (1.308-1.362)} \\
 
@@
     \end{tabular}%
-    % }
+    }
 \end{table}
@@
     \centering
-    \caption{Model performance based on sex and age [mean (95\% CI)] on the 2021 data without diagnostic category.}
+    \caption{Model performance for sex and age subgroups [mean (95\% CI)] on the 2021 data without diagnostic category.}
     \label{tab_retrospective:tableA3}
-    % \resizebox*{0.98\textwidth}{!}{%
+    \resizebox*{0.85\textwidth}{!}{%
     \begin{tabular}{l|cc|cc}
@@
 
-         & \multicolumn{2}{c|}{Model} & \multicolumn{2}{c}{Model} \\
+         & \multicolumn{4}{c|}{Model} \\
         \midrule
@@
 
-        F1-score [\%] $\uparrow$                                 & \makecell{30.4 \\ (29.3-31.5)} & \makecell{35.2 \\ (34.0-36.4)} & \makecell{24.3 \\ (23.0-25.7)} & \makecell{36.6 \\ (35.6-37.6)} \\
+        F1-score [\%] $\uparrow$                                 & \makecell[c]{30.4 \\ (29.3-31.5)} & \makecell[c]{35.2 \\ (34.0-36.4)} & \makecell[c]{24.3 \\ (23.0-25.7)} & \makecell[c]{36.6 \\ (35.6-37.6)} \\
         \midrule
-        Sensitivity [\%] $\uparrow$                              & \makecell{44.5 \\ (42.9-46.0)} & \makecell{52.8 \\ (51.1-54.4)} & \makecell{45.6 \\ (43.4-47.9)} & \makecell{49.2 \\ (47.9-50.5)} \\
+        Sensitivity [\%] $\uparrow$                              & \makecell[c]{44.5 \\ (42.9-46.0)} & \makecell[c]{52.8 \\ (51.1-54.4)} & \makecell[c]{45.6 \\ (43.4-47.9)} & \makecell[c]{49.2 \\ (47.9-50.5)} \\
         \midrule
-        PPV [\%] $\uparrow$                                      & \makecell{23.1 \\ (22.2-24.1)} & \makecell{26.4 \\ (25.3-27.4)} & \makecell{16.6 \\ (15.5-17.5)} & \makecell{29.1 \\ (28.2-30.0)} \\
+        PPV [\%] $\uparrow$                                      & \makecell[c]{23.1 \\ (22.2-24.1)} & \makecell[c]{26.4 \\ (25.3-27.4)} & \makecell[c]{16.6 \\ (15.5-17.5)} & \makecell[c]{29.1 \\ (28.2-30.0)} \\
         \midrule
-        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - NPV)}         & \makecell{0.152 \\ (0.146-0.159)} & \makecell{0.155 \\ (0.147-0.162)} & \makecell{0.058 \\ (0.055-0.062)} & \makecell{0.372 \\ (0.359-0.386)} \\
+        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - NPV)}         & \makecell[c]{0.152 \\ (0.146-0.159)} & \makecell[c]{0.155 \\ (0.147-0.162)} & \makecell[c]{0.058 \\ (0.055-0.062)} & \makecell[c]{0.372 \\ (0.359-0.386)} \\
         \midrule
-        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell{0.404 \\ (0.394-0.414)} & \makecell{0.480 \\ (0.467-0.494)} & \makecell{0.246 \\ (0.239-0.254)} & \makecell{0.873 \\ (0.852-0.894)} \\
+        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell[c]{0.404 \\ (0.394-0.414)} & \makecell[c]{0.480 \\ (0.467-0.494)} & \makecell[c]{0.246 \\ (0.239-0.254)} & \makecell[c]{0.873 \\ (0.852-0.894)} \\
 
@@
     \centering
-    \caption{Overall performance on MH-1813 test data, performance without 1-1-2 training data, and performance without 1813 training data [mean (95\% CI)].}
+    \caption{Overall performance on MH-1813 test data for the original model, the model without 1-1-2 training data, and the model without 1813 training data along with call-taker performance [mean (95\% CI)].}
     \label{tab_retrospective:tableA5}
@@
         \midrule
-        & \multicolumn{5}{c}{\textit{Without 1-1-2 training data (only MH-1813 data)}} \\
+        \makecell[l]{Model \\ w/o 1-1-2} & \makecell[c]{32.4 \\ (31.8-33.1)} & \makecell[c]{60.4 \\ (59.3-61.4)} & \makecell[c]{22.2 \\ (21.6-22.7)} & \makecell[c]{0.088 \\ (0.085-0.091)} & \makecell[c]{0.467 \\ (0.460-0.474)} \\
         \midrule
+        \makecell[l]{Model \\ w/o MH-1813} & \makecell[c]{31.4 \\ (30.7-32.1)} & \makecell[c]{50.4 \\ (49.3-51.4)} & \makecell[c]{22.8 \\ (22.2-23.4)} & \makecell[c]{0.110 \\ (0.106-0.113)} & \makecell[c]{0.375 \\ (0.369-0.381)} \\
 
-        Model        & \makecell[c]{32.4 \\ (31.8-33.1)} & \makecell[c]{60.4 \\ (59.3-61.4)} & \makecell[c]{22.2 \\ (21.6-22.7)} & \makecell[c]{0.088 \\ (0.085-0.091)} & \makecell[c]{0.467 \\ (0.460-0.474)} \\
+        % & \multicolumn{5}{c}{\textit{Without 1-1-2 training data (only MH-1813 data)}} \\
+        % \midrule
+        % Model        & \makecell[c]{32.4 \\ (31.8-33.1)} & \makecell[c]{60.4 \\ (59.3-61.4)} & \makecell[c]{22.2 \\ (21.6-22.7)} & \makecell[c]{0.088 \\ (0.085-0.091)} & \makecell[c]{0.467 \\ (0.460-0.474)} \\
 
-        \midrule
-        & \multicolumn{5}{c}{\textit{Without MH-1813 training data (only 1-1-2 data)}} \\
-        \midrule
-
-        Model        & \makecell[c]{31.4 \\ (30.7-32.1)} & \makecell[c]{50.4 \\ (49.3-51.4)} & \makecell[c]{22.8 \\ (22.2-23.4)} & \makecell[c]{0.110 \\ (0.106-0.113)} & \makecell[c]{0.375 \\ (0.369-0.381)} \\
+        % & \multicolumn{5}{c}{\textit{Without MH-1813 training data (only 1-1-2 data)}} \\
+        % \midrule
+        % Model        & \makecell[c]{31.4 \\ (30.7-32.1)} & \makecell[c]{50.4 \\ (49.3-51.4)} & \makecell[c]{22.8 \\ (22.2-23.4)} & \makecell[c]{0.110 \\ (0.106-0.113)} & \makecell[c]{0.375 \\ (0.369-0.381)} \\
 
@@
 \section{Additional results: Detailed model explainability tables}
+
+In \cref{tab_retrospective:tableA6,tab_retrospective:tableA7} we provide detailed versions of the data presented in the model explainability section in the main manuscript.
+Specifically, we include the Danish word, its English translation, the rank and impact scores and the number of occurrences of each word.
 
@@
         \midrule
-         & Word, $w$ (\textit{Danish}) & Translation (\textit{English}) & Rank, $r^{(w)}$ & Occurences, $D^{(w)}$ & \makecell{Impact, $i^{(d,w)}$ \\ mean $\pm$ std.} \\
+         & \makecell{Word, $w$ \\ (\textit{Danish})} & \makecell{Translation, $w$ \\ (\textit{English})} & Rank, $r^{(w)}$ & Count, $D^{(w)}$ & \makecell{Impact, $i^{(d,w)}$ \\ mean $\pm$ std.} \\
         \midrule    
@@
 \begin{enumerate}[label=\Alph*.]
-    \item Call takers change any stroke prediction from negative to positive if the model predicts a positive (call takers mirror model positives).
-    \item Call takers change any stroke prediction from positive to negative if the model predicts a negative (call takers mirror model negatives).
+    \item Call-takers change any stroke prediction from negative to positive if the model predicts a positive (call takers mirror model positives).
+    \item Call-takers change any stroke prediction from positive to negative if the model predicts a negative (call takers mirror model negatives).
 \end{enumerate}
@@
 
+\section{Research in context search term results}
+The search terms stroke AND (artificial intelligence OR machine learning OR deep learning) AND (EMS OR emergency medical services OR dispatch OR telephone) yielded 88 articles in PubMed.
+These articles are listed below:
+
+{\small
+\begin{enumerate}
+    \item Soun JE, Chow DS, Nagamine M et al. Artificial intelligence and acute stroke imaging. AJNR. American journal of neuroradiology 2021; 42: 2-11.
+    \item Murray NM, Unberath M, Hager GD, Hui FK.
+Artificial intelligence to diagnose ischemic stroke and identify large vessel occlusions: a systematic review.
+Journal of neurointerventional surgery 2020; 12: 156-64.
+    \item Shlobin NA, Baig AA, Waqas M et al. Artificial intelligence for large-vessel occlusion stroke: a systematic review. World neurosurgery 2022; 159: 207-20.e1.
+    \item Linder SM, Rosenfeldt AB, Bay RC, Sahu K, Wolf SL, Alberts JL.
+Improving quality of life and depression after stroke through telerehabilitation.
+The American journal of occupational therapy: official publication of the American Occupational Therapy Association 2015; 69: 6902290020p1-10.
+    \item Humphreys K, Shover CL, Andrews CM et al.
+Responding to the opioid crisis in North America and beyond: recommendations of the Stanford-Lancet Commission.
+Lancet (London, England) 2022; 399: 555-604.
+    \item Fassbender K, Lesmeister M, Merzou F. Prehospital stroke management and mobile stroke units. Current opinion in neurology 2023; 36: 140-6.
+    \item McDonough RV, Ospel JM, Majoie CBLM et al.
+Clinical outcome of patients with mild pre-stroke morbidity following endovascular treatment: a HERMES substudy.
+Journal of neurointerventional surgery 2023; 15: 214-20.
+    \item Bat-Erdene BO, Saver JL.
+Automatic acute stroke symptom detection and emergency medical systems alerting by mobile health technologies: a review.
+Journal of stroke and cerebrovascular diseases: the official journal of National Stroke Association 2021; 30: 105826.
+    \item Bonini N, Vitolo M, Imberti JF et al. Mobile health technology in atrial fibrillation. Expert review of medical devices 2022; 19: 327-40.
+    \item Scholz ML, Collatz-Christensen H, Blomberg SNF, Boebel S, Verhoeven J, Krafft T.
+Artificial intelligence in Emergency Medical Services dispatching: assessing the potential impact of an automatic speech recognition software on stroke detection taking the Capital Region of Denmark as case in point.
+Scandinavian journal of trauma, resuscitation and emergency medicine 2022; 30: 36.
+    \item Gupta R, Krishnam SP, Schaefer PW, Lev MH, Gonzalez RG.
+An East Coast perspective on artificial intelligence and machine learning: part 2: ischemic stroke imaging and triage.
+Neuroimaging clinics of North America 2020; 30: 467-78.
+    \item Esbjörnsson M, Ullberg T.
+Safety and usability of wearable accelerometers for stroke detection the STROKE ALARM PRO 1 study.
+Journal of stroke and cerebrovascular diseases: the official journal of National Stroke Association 2022; 31: 106762.
+    \item Fast L, Temuulen U, Villringer K et al. Machine learning-based prediction of clinical outcomes after first-ever ischemic stroke. Frontiers in neurology 2023; 14: 1114360.
+    \item Gupta R, Krishnam SP, Schaefer PW, Lev MH, Gilberto Gonzalez R.
+An East Coast perspective on artificial intelligence and machine learning: part 1: hemorrhagic stroke imaging and triage.
+Neuroimaging clinics of North America 2020; 30: 459-66.
+    \item Cai T, Ni H, Yu M et al. DeepStroke: An efficient stroke screening framework for emergency rooms with multimodal adversarial deep learning. Medical image analysis 2022; 80: 102522.
+    \item Zachrison KS, Dhand A, Schwamm LH, Onnela JP. A network approach to stroke systems of care. Circulation. Cardiovascular quality and outcomes 2019; 12: e005526.
+    \item O'Connell GC, Walsh KB, Smothers CG et al. Use of deep artificial neural networks to identify stroke during triage via subtle changes in circulating cell counts. BMC neurology 2022; 22: 206.
+    \item Yang L, Liu Q, Zhao Q, Zhu X, Wang L. Machine learning is a valid method for predicting prehospital delay after acute ischemic stroke. Brain and behavior 2020; 10: e01794.
+    \item Aldridge CM, McDonald MM, Wruble M et al. Human vs. machine learning based detection of facial weakness using video analysis. Frontiers in neurology 2022; 13: 878282.
+    \item Lachance CC, Ford C.
+Portable stroke detection devices for patients with stroke symptoms: a review of diagnostic accuracy and cost-effectiveness.
+Canadian Agency for Drugs and Technologies in Health 2019
+    \item Fardoun HM, Mashat AS. Methodologies, models and algorithms for patients rehabilitation. Methods of information in medicine 2016; 55: 60-4.
+    \item Wibring K, Magnusson C, Axelsson C, Lundgren P, Herlitz J, Andersson Hagiwara M.
+Towards definitions of time-sensitive conditions in prehospital care.
+Scandinavian journal of trauma, resuscitation and emergency medicine 2020; 28: 7.
+    \item Mayampurath A, Parnianpour Z, Richards CT et al. Improving prehospital stroke diagnosis using natural language processing of paramedic reports. Stroke 2021; 52: 2676-9.
+    \item Morey JR, Zhang X, Yaeger KA et al.
+Real-world experience with artificial intelligence-based triage in transferred large vessel occlusion stroke patients.
+Cerebrovascular diseases (Basel, Switzerland) 2021; 50: 450-5.
+    \item Sung SF, Hung LC, Hu YH.
+Developing a stroke alert trigger for clinical decision support at emergency triage using machine learning.
+International journal of medical informatics 2021; 152: 104505.
+    \item Zhang Z, Zhou D, Zhang J et al. Multilayer perceptron-based prediction of stroke mimics in prehospital triage. Scientific reports 2022; 12: 17994.
+    \item Park E, Lee K, Han T, Nam HS.
+Automatic grading of stroke symptoms for rapid assessment using optimized machine learning and 4-limb kinematics: clinical validation study.
+Journal of medical Internet research 2020; 22: e20641.
+    \item Chen M, Tan X, Padman R.
+A machine learning approach to support urgent stroke triage using administrative data and social determinants of health at hospital presentation: retrospective study.
+Journal of medical Internet research 2023; 25: e36477.
+    \item Uchida K, Kouno J, Yoshimura S et al.
+Development of machine learning models to predict probabilities and types of stroke at prehospital stage: the Japan urgent stroke triage score using machine learning (JUST-ML).
+Translational stroke research 2022; 13: 370-81.
+    \item Finck T, Schinz D, Grundl L et al. Automated detection of ischemic stroke and subsequent patient triage in routinely acquired head CT. Clinical neuroradiology 2022; 32: 419-26.
+    \item Sangal RB, Fodeh S, Taylor A et al.
+Identification of patients with nontraumatic intracranial hemorrhage using administrative claims data.
+Journal of stroke and cerebrovascular diseases: the official journal of National Stroke Association 2020; 29: 105306.
+    \item Adedinsewo D, Carter RE, Attia Z et al.
+Artificial intelligence-enabled ECG algorithm to identify patients with left ventricular systolic dysfunction presenting to the emergency department with dyspnea.
+Circulation.
+Arrhythmia and electrophysiology 2020; 13: e008437.
+    \item Liu L, Yu B, Han M, Yuan S, Wang N. Mild cognitive impairment understanding: an empirical study by data-driven approach. BMC bioinformatics 2019; 20: 481.
+    \item Hayashi Y, Shimada T, Hattori N et al. A prehospital diagnostic algorithm for strokes using machine learning: a prospective observational study. Scientific reports 2021; 11: 20519.
+    \item Grant L, Joo P, Nemnom MJ, Thiruganasambandamoorthy V.
+Machine learning versus traditional methods for the development of risk stratification scores: a case study using original Canadian Syncope Risk Score data.
+Internal and emergency medicine 2022; 17: 1145-53.
+    \item van Meenen LCC, van Stigt MN, Marquering HA et al.
+Detection of large vessel occlusion stroke with electroencephalography in the emergency room: first results of the ELECTRA-STROKE study.
+Journal of neurology 2022; 269: 2030-8.
+    \item van Meenen LCC, den Hartog SJ, Groot AE et al.
+Relationship between primary stroke center volume and time to endovascular thrombectomy in acute ischemic stroke.
+European journal of neurology 2021; 28: 4031-8.
+    \item Chien HC, Yang TL, Juang WC, Chen YA, Li YJ, Chen CY.
+Pilot report for intracranial hemorrhage detection with deep learning implanted head computed tomography images at emergency department.
+Journal of medical systems 2022; 46: 49.
+    \item Huo D, Leppert M, Pollard R et al.
+Large vessel occlusion prediction in the emergency department with national institutes of health stroke scale components: a machine learning approach.
+Journal of stroke and cerebrovascular diseases: the official journal of National Stroke Association 2021; 30: 106030.
+    \item Wu MR, Chen YT, Li ZX et al.
+Dysphagia screening and pneumonia after subarachnoid hemorrhage: findings from the Chinese stroke center alliance.
+CNS neuroscience \& therapeutics 2022; 28: 913-21.
+    \item Bacchi S, Oakden-Rayner L, Zerner T, Kleinig T, Patel S, Jannes J.
+Deep learning natural language processing successfully predicts the cerebrovascular cause of transient ischemic attack-like presentations.
+Stroke 2019; 50: 758-60.
+    \item Heldt FS, Vizcaychipi MP, Peacock S et al. Early risk assessment for COVID-19 patients from emergency department data using machine learning. Scientific reports 2021; 11: 4200.
+    \item Lu Z, Xiong Y, Yang K et al. What predicts large vessel occlusion in mild stroke patients. BMC neurology 2023; 23: 29.
+    \item Marijon E, Garcia R, Narayanan K, Karam N, Jouven X.
+Fighting against sudden cardiac death: need for a paradigm shift-adding near-term prevention and pre-emptive action to long-term prevention.
+European heart journal 2022; 43: 1457-64.
+    \item You J, Yu PLH, Tsang ACO et al.
+3D dissimilar-Siamese-u-net for hyperdense Middle cerebral artery sign segmentation.
+Computerized medical imaging and graphics: the official journal of the Computerized Medical Imaging Society 2021; 90: 101898.
+    \item Moral-Munoz JA, Zhang W, Cobo MJ, Herrera-Viedma E, Kaber DB.
+Smartphone-based systems for physical rehabilitation applications: a systematic review.
+Assistive technology: the official journal of RESNA 2021; 33: 223-36.
+    \item Salman OH, Aal-Nouman MI, Taha ZK, Alsabah MQ, Hussein YS, Abdelkareem ZA.
+Formulating multi diseases dataset for identifying, triaging and prioritizing patients to multi medical emergency levels: Simulated dataset accompanied with codes.
+Data in brief 2021; 34: 106576.
+    \item Desai A, Zumbo A, Giordano M et al.
+Word2vec Word embedding-based artificial intelligence model in the triage of patients with suspected diagnosis of major ischemic stroke: a feasibility study.
+International journal of environmental research and public health 2022; 19: 15295.
+    \item Rinkel LA, Prick JCM, Slot RER et al. Impact of the COVID-19 outbreak on acute stroke care. Journal of neurology 2021; 268: 403-8.
+    \item Chilamkurthy S, Ghosh R, Tanamala S et al. Deep learning algorithms for detection of critical findings in head CT scans: a retrospective study. Lancet (London, England) 2018; 392: 2388-96.
+    \item Kramer NM, Demaerschalk BM.
+A novel application of teleneurology: robotic telepresence in supervision of neurology trainees.
+Telemedicine journal and e-health: the official journal of the American Telemedicine Association 2014; 20: 1087-92.
+    \item Yadav K, Sarioglu E, Choi HA, Cartwright WB, Hinds PS, Chamberlain JM.
+Automated outcome classification of computed tomography imaging reports for pediatric traumatic brain injury.
+Academic emergency medicine: official journal of the Society for Academic Emergency Medicine 2016; 23: 171-8.
+    \item van Meenen LCC, Riedijk F, Stolp J et al.
+Pre- and interhospital workflow times for patients with large vessel occlusion stroke transferred for endovasvular thrombectomy.
+Frontiers in neurology 2021; 12: 730250.
+    \item Saberi-Movahed F, Mohammadifard M, Mehrpooya A et al.
+Decoding clinical biomarker space of COVID-19: Exploring matrix factorization-based feature selection methods.
+Computers in biology and medicine 2022; 146: 105426.
+    \item Shimada T, Matsubara K, Koyama D et al.
+Development of evaluation system for cerebral artery occlusion in emergency medical services: noninvasive measurement and utilization of pulse waves.
+Scientific reports 2023; 13: 3339.
+    \item Gil-Jardiné C, Chenais G, Pradeau C et al.
+Trends in reasons for emergency calls during the COVID-19 crisis in the department of Gironde, France using artificial neural network for natural language classification.
+Scandinavian journal of trauma, resuscitation and emergency medicine 2021; 29: 55.
+    \item Santala OE, Lipponen JA, Jäntti H et al.
+Continuous mHealth Patch Monitoring for the Algorithm-Based Detection of Atrial Fibrillation: Feasibility and Diagnostic Accuracy Study.
+JMIR cardio 2022; 6: e31230.
+    \item O'Brien MK, Shawen N, Mummidisetty CK et al.
+Activity recognition for persons with stroke using mobile phone technology: toward improved performance in a home setting.
+Journal of medical Internet research 2017; 19: e184.
+    \item Bateman RM, Sharpe MD, Jagger JE et al.
+36th international symposium on intensive care and emergency medicine: Brussels, Belgium.
+15-18 March 2016.
+Critical care (London, England) 2016; 20: 94.
+    \item Modica CM, Johnson BR, Zalewski C et al. Hearing loss and irritability reporting without vestibular differences in explosive breaching professionals. Frontiers in neurology 2020; 11: 588377.
+    \item Palmcrantz S, Plantin J, Borg J. Factors affecting the usability of an assistive soft robotic glove after stroke or multiple sclerosis. Journal of rehabilitation medicine 2020; 52: jrm00027.
+    \item Mitsopoulos K, Fiska V, Tagaras K et al.
+NeuroSuitUp: System architecture and validation of a motor rehabilitation wearable robotics and serious game platform.
+Sensors (Basel, Switzerland) 2023; 23: 3281.
+    \item Meng G, Tan Y, Fang M, Yang H, Liu X, Zhao Y.
+Meteorological factors related to emergency admission of elderly stroke patients in shanghai: analysis with a multilayer perceptron neural network.
+Medical science monitor: international medical journal of experimental and clinical research 2015; 21: 3600-7.
+    \item Iglesias JE, Schleicher R, Laguna S et al. Quantitative Brain Morphometry of Portable Low-Field-Strength MRI Using Super-Resolution Machine Learning. Radiology 2023; 306: e220522.
+    \item Lai F. Stroke networks based on robotic telepresence. Journal of telemedicine and telecare 2009; 15: 135-6
+    \item Kim D, Oh J, Im H, Yoon M, Park J, Lee J.
+Automatic classification of the Korean triage acuity scale in simulated emergency rooms using speech recognition and natural language processing: a proof of concept study.
+Journal of Korean medical science 2021; 36: e175.
+    \item Jonsson A, Binongo J, Patel P et al. Mastering the learning curve for robotic-assisted coronary artery bypass surgery. The Annals of thoracic surgery 2023; 115: 1118-25.
+    \item Knowlton SM, Sencan I, Aytar Y et al. Sickle cell detection using a smartphone. Scientific reports 2015; 5: 15022.
+    \item Che F, Liu Y, Gong X et al.
+Extracranial carotid plaque hemorrhage is independently associated with poor 3-month functional outcome after acute ischemic stroke-a prospective cohort study.
+Frontiers in neurology 2021; 12: 780436.
+    \item Boers AMM, Jansen IGH, Brown S et al.
+Mediation of the relationship between endovascular therapy and functional outcome by follow-up infarct volume in patients with acute ischemic stroke.
+JAMA neurology 2019; 76: 194-202.
+    \item Stone JR, Avants BB, Tustison NJ et al.
+Functional and structural neuroimaging correlates of repetitive low-level blast exposure in career breachers.
+Journal of neurotrauma 2020; 37: 2468-81.
+    \item Rickards CA, Vyas N, Ryan KL et al.
+Are you bleeding?
+Validation of a machine-learning algorithm for determination of blood volume status: application to remote triage.
+Journal of applied physiology (Bethesda, Md.
+: 1985) 2014; 116: 486-94.
+    \item Edwards KA, Greer K, Leete J et al. Neuronally-derived tau is increased in experienced breachers and is associated with neurobehavioral symptoms. Scientific reports 2021; 11: 19527.
+    \item Cheng AL, Liu J, Bravo S, Miller JC, Pahlevan NM.
+Screening left ventricular systolic dysfunction in children using intrinsic frequencies of carotid pressure waveforms measured by a novel smartphone-based device.
+Physiological measurement 2023; 44.
+    \item Grunwald IQ, Kulikovski J, Reith W et al.
+Collateral automation for triage in stroke: evaluating automated scoring of collaterals in acute stroke on computed tomography scans.
+Cerebrovascular diseases (Basel, Switzerland) 2019; 47: 217-22.
+    \item Hyakutake K, Morishita T, Saita K et al.
+Effects of home-based robotic therapy involving the single-joint hybrid assistive limb robotic suit in the chronic phase of stroke: a pilot study.
+BioMed research international 2019; 2019: 5462694.
+    \item Murray C, Ortiz E, Kubin C. Application of a robot for critical care rounding in small rural hospitals. Critical care nursing clinics of North America 2014; 26: 477-85.
+    \item Edwards KA, Leete JJ, Tschiffely AE et al. Blast exposure results in tau and neurofilament light chain changes in peripheral blood. Brain injury 2020; 34: 1213-21.
+    \item Oresko JJ, Duschl H, Cheng AC.
+A wearable smartphone-based platform for real-time cardiovascular disease detection via electrocardiogram processing.
+IEEE transactions on information technology in biomedicine : a publication of the IEEE Engineering in Medicine and Biology Society 2010; 14: 734-40.
+    \item Alcoceba-Herrero I, Coco-Martín MB, Leal-Vega L et al.
+Randomized controlled trial evaluating the benefit of a novel clinical decision support system for the management of COVID-19 patients in home quarantine: a study protocol.
+International journal of environmental research and public health 2023; 20: 2300.
+    \item van Meenen LCC, Groot AE, Venema E et al.
+Interhospital transfer vs.
+direct presentation of patients with a large vessel occlusion not eligible for IV thrombolysis.
+Journal of neurology 2020; 267: 2142-50.
+    \item Forghani R, Gupta R. Case of the Season: Artificial intelligence in clinical practice-large vessel occlusion triage in stroke imaging. Seminars in roentgenology 2023; 58: 147-51.
+    \item Jabbour P, Gonzalez LF, Tjoumakaris S, Randazzo C, Rosenwasser R. Stroke in the robotic era. World neurosurgery 2010; 73: 603-4.
+    \item Larkin H. On the spot when doc is not. Hospitals \& health networks 2008; 82: 41-2.
+    \item Jiang Y, Guarino P, Ma S et al. Bayesian accrual prediction for interim review of clinical studies: open source R package and smartphone application. Trials 2016; 17: 336.
+    \item Lee JH, Oh BJ, Ahn JY et al.
+Effectiveness of automatic acute stroke alert system based on UMLS mapped local terminology codes at emergency department.
+AMIA ...
+Annual Symposium proceedings.
+AMIA Symposium 2008; 1018.
+    \item Robot saves ED stroke patients, addresses subspecialist shortage. ED management: the monthly update on emergency department management 2011; 23: 13-5.
+    \item Nagy Z, Simon P, Sipos E, Kozmann G.
+The main elements of the information system of the National Stroke Program (Smart Card - Telecommunication - Knowledge Bases).
+Medinfo.
+MEDINFO 1995; 8 Pt 2: 1496-9.
+\end{enumerate}
 }
+
+% }

### 2023-09-22 chapters/introduction.tex
Update wording

@@
 The application of artificial intelligence in medical decision-making and particularly for diagnosing and recommending treatments is likely to hold significant potential to reduce the effect of these issues and improve patient outcomes.
-However, as such decisions impact critical aspects of human health and well-being, this prospect also raises concerns related to the reliability of artificial intelligence systems.
+However, as such decisions impact critical aspects of human health and well-being, this prospect also raises concerns related to the reliability of artificial intelligence systems \cite{chen_ethical_2021, shailaja_machine_2018, ahmad_interpretable_2018}.
 
-\todo[inline]{Can we find a better example than the Rhind Papyrus, or an additional one?
-It could be more on-point with decision-making and more well-known.
-Ideally, weave in an example of a failure of such technology.}
 The evolution of technology to support and automate decision-making can be traced back to the development of algorithms and mathematical techniques by ancient civilizations. 
@@
 Although never completed, it laid the foundation for his later design of the Analytical Engine in 1837 which is considered to be the first general-purpose computer design.
-Working with Babbage, Ada Lovelace is widely regarded as the first to have recognized that general-purpose programmable computers could have applications beyond pure arithmetic.
-However, it was not until the advent of electronics that the first successful general-purpose programmable computers were developed, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}).
+Working with Babbage, Ada Lovelace is widely regarded as the first to have recognized that programmable computers could have applications beyond pure arithmetic.
+However, it was not until the advent of electronics that the first successful general-purpose computers were developed, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}).
 % While the potential of a general-purpose programmable computer with applications beyond pure arithmetic was recognized by Ada Lovelace,  
@@
 
-With the advent of electronic computers, the internet, and the widespread adoption of digital technologies, the amount of data available to decision-makers has grown exponentially. 
-This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that focuses on developing algorithms that can learn from data to recognize salient patterns and make predictions to guide decisions.
-Enabled by a massive increase in the performance and availability of parallel computing resources, such methods have set new standards for the abilities of computer systems in various domains, including speech recognition, computer vision, and natural language processing \cite{lecun_deep_2015}.
-In recent years, machine learning has also been applied within healthcare, where it has shown promising results in tasks such as medical imaging \cite{lundervold_overview_2019}, drug discovery \cite{chen_rise_2018}, and clinical decision support \cite{cite15, cite14}.
+With the advent of electronic computers, the internet, and the widespread adoption of digital technologies continuing into the 21st century, the amount of data available to decision-makers has grown exponentially.
+This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that focuses on developing algorithms that can learn from data to recognize patterns and make predictions to guide decision-making.
+Enabled by a massive increase in the performance and availability of parallel computing resources, such methods have set new standards for the abilities of computer systems in various domains, such as speech recognition, computer vision, and natural language processing \cite{lecun_deep_2015}.
+In recent years, machine learning has made its way into the healthcare sector, where it has shown promising results in tasks such as medical imaging \cite{lundervold_overview_2019}, drug discovery \cite{chen_rise_2018}, and clinical decision support \cite{cite15, cite14}.
 
 However, these technologies come with the same risks as previous generations did, and in some cases, they may exacerbate them. 
-Inherent limitations of the technology itself, human error, and the complexity of the real world are all factors that can lead to failures.
-A pertinent example is the nascent use of machine learning in autonomous driving systems:
+Limitations of the technology itself, human error, and the complexity of the real world are all factors that can lead to failures.
+A pertinent example is the use of machine learning in autonomous driving systems:
 In 2018, the first incident of a self-driving car killing a pedestrian took place in Tempe, Arizona. 
 The car, an SUV operated by Uber, was driving in autonomous mode when it struck a pedestrian crossing the street with her bicycle. 
-According to the investigation by the National Transportation Safety Board, the car's sensors detected the pedestrian $5.6$ seconds before the crash, but it did not correctly predict her path or reduce the SUV's speed (\cref{fig:uber_nhsa_accident}).
+According to the investigation by the National Transportation Safety Board, the car's sensors detected the pedestrian $5.6$ seconds before the crash, but it did not manage to correctly predict her path or reduce the SUV's speed (\cref{fig:uber_nhsa_accident}).
 Specifically, during those seconds, the system incorrectly classified the pedestrian more than ten times, first as a vehicle, then as an unknown object and ultimately as a bicyclist, each time changing, or resetting, the predicted path.
@@
 
-Uber is not the only actor facing the challenges of applying machine learning in real-world scenarios.
+And Uber is not the only actor facing the challenges of applying machine learning in real-world scenarios.
 A recent report by the Washington Post found that Tesla's Autopilot system has been involved in at least 736 crashes in the last four years with 16 fatalities \cite{siddiqui_17_2023}.
-Although recent statistics published by Waymo, the self-driving car company owned by Google's parent company Alphabet, revealed far fewer crashes, with only 2 of 20 incidents over 1 million miles meeting reporting criteria, it is interesting to note that the cars are all operated in the sunny states of California and Arizona, where driving conditions vary relatively little \cite{hawkins_waymo_2023}.
-While driver-assistance systems undoubtedly help reduce accidents in many case, and has great future potential, the issues of current systems highlight the need for machine learning systems to be robust in rare, adverse or difficult situations.
+% And though recent statistics published by Waymo, 
+% %the self-driving car company owned by Google's parent company Alphabet, 
+% revealed far fewer crashes, with only 2 of 20 incidents over 1 million miles meeting reporting criteria, it is interesting to note that the cars are all operated in the sunny states of California and Arizona, where driving conditions vary relatively little \cite{hawkins_waymo_2023}.
+While driver-assistance systems undoubtedly help reduce accidents in many cases, and has great future potential, the issues of current systems highlight the need for machine learning systems that are robust in rare, adverse or difficult situations.
 
@@
 This AI Act groups applications of machine learning into three categories based on the potential risk they pose to society: limited risk, high risk, and unacceptable risk.
-Limited risk applications, for instance within gaming, must meet transparency requirements, and unacceptable risk applications, such as public use of facial recognition, are banned.
-High risk applications include those in healthcare and must meet a strict set of requirements focusing on anti-discrimination and robustness.
-%
+Limited risk applications, such as in computer games, must meet transparency requirements, and unacceptable risk applications, such as public use of facial recognition, are banned.
+High risk applications, such as in autonomous vehicles and healthcare, must meet a strict set of requirements focusing on anti-discrimination and robustness.
+
 One of the key ways to improve the safety, reliability and trustworthiness of machine learning systems is to equip them with the ability accurately estimate the uncertainty of their predictions. 
@@
 
-\vspace{1em}
+\vspace{0.5em}
 \begin{center}
@@
 % OUTLINE: Primer overview of the thesis. What are the topics and what is the context of the research.
-\noindent The research presented in this thesis will focus on the use of machine learning in relation to uncertainty estimation, speech processing, and medical conversations. 
+% \noindent The research presented in this thesis will focus on the use of machine learning in relation to uncertainty estimation, speech processing, and medical conversations. 
+\noindent The research presented in this thesis deals with the use of machine learning for medical conversations with a focus on uncertainty estimation and speech processing. 
 The research project was partially funded by a grant from the national Danish Innovation Fund (grant no.\@ 0153-00167B) and defined and carried out in collaboration with the Danish company Corti, which develops computer software for medical decision support.
 In the rest of this chapter, we will provide a high-level introduction to these topics and discuss the motivation behind the research project.
+
+\todo[inline]{Can we find a better example than the Rhind Papyrus, or an additional one?
+It could be more on-point with decision-making and more well-known.
+Ideally, weave in an example of a failure of such technology.}
 

### 2023-09-22 appendices/supplementary_retrospective.tex
Update retrospective paper with rebuttal version

@@
 \label{app:supplementary-retrospective}
-% \ifthenelse{\equal{\skipappendices}{true}}{}{
+\ifthenelse{\equal{\skipappendices}{true}}{}{
 
@@
 When 1-1-2 data were not used for training (\cref{tab_retrospective:tableA2}), the model performed significantly better in men than in women on the test set in terms of all metrics (p < 0.0001) and significantly better on the 65+ group than on the 1864 group in terms of all metrics (p < 0.0001).
-As noted in the main text, both these statements were also true when 1-1-2 data were used for the training (\cref{tab_retrospective:table2}) and that including 1-1-2 data significantly improved overall performance (p < 0.0001).
+As noted in the main text, both these statements were also true when 1-1-2 data were used for the training (\cref{tab_retrospective:table2-main-results}) and that including 1-1-2 data significantly improved overall performance (p < 0.0001).
 
-Compared to the test set (\cref{tab_retrospective:table2}), model performance on the 2021 calls without diagnostic category was significantly worse in all demographic subgroups (\cref{tab_retrospective:tableA3}) in terms of all metrics (p < 0.0001), except for positive predictive value (PPV) for male (where p = 0.0056 [significant]) and for the false positive rate (FPR) on females and individuals aged 65+ years (where p = 0.213 and p = 0.362, respectively [insignificant]).
+Compared to the test set (\cref{tab_retrospective:table2-main-results}), model performance on the 2021 calls without diagnostic category was significantly worse in all demographic subgroups (\cref{tab_retrospective:tableA3}) in terms of all metrics (p < 0.0001), except for positive predictive value (PPV) for male (where p = 0.0056 [significant]) and for the false positive rate (FPR) on females and individuals aged 65+ years (where p = 0.213 and p = 0.362, respectively [insignificant]).
 
@@
 % \todo[inline]{Insert table A6 here}
-\begin{table}[h]
+\begin{sidewaystable}[h]
     \centering
@@
     }
-\end{table}
+\end{sidewaystable}
 

### 2023-09-22 chapters/paper_retrospective.tex
Updates to table formatting retrospective
Update retrospective paper with rebuttal version

@@
 \section*{Abstract}
-\paragraph{Background:} Effective stroke treatment depends on recognition by call-takers at prehospital telehealth services.
+Effective stroke treatment depends on recognition by call-takers at prehospital telehealth services.
 This study aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.
-\paragraph{Methods:} We used calls from 1 January 2015 to 31 December 2020 in Copenhagen to develop a machine learning-based classification pipeline.
+We used calls from 1 January 2015 to 31 December 2020 in Copenhagen to develop a machine learning-based classification pipeline.
 Calls from 2021 were used for testing.
 Calls were transcribed using an automatic speech recognition model and then categorized as stroke or non-stroke using a text classification model.
-\paragraph{Findings:} Call-takers achieved a sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%).
+Call-takers achieved a sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%).
 The machine learning framework performed significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
-\paragraph{Interpretation:} A machine learning framework for recognizing stroke in prehospital medical helpline calls may become a supportive tool for call-takers, aiding in early and accurate stroke recognition.
+Thus, a machine learning framework for recognising stroke in prehospital medical helpline calls may become a supportive tool for call-takers, aiding in early and accurate stroke recognition.
 
@@
 
+We used call recordings and registry data from the Copenhagen Emergency Medical Services (CEMS) and the Danish Stroke Registry (DanStroke)\cite{cite16} from 2015 to 2020.
+The call recordings were obtained from two call lines: the 1-1-2 emergency line and the medical helpline 1813 (MH-1813).
+These data were used to build a machine learning framework to classify medical helpline calls as stroke or non-stroke.
+Calls were first transcribed using an automatic speech recognition model and then categorised by a text classification model trained as an ensemble of five individual models.
+We compared the performance of the model with that of call-takers using MH-1813 data from 2021.
+
 \section{Results}
 
-\subsection{Main results}
+\subsection{Population characteristics}
+%
+Calls to the MH-1813 were divided into training, validation, and test subsets and calls to the emergency line 1-1-2 were only used as supplementary training data (Table 1).
+Calls from the test year (2021) that were not associated with a diagnostic category code, which we used to evaluate call-taker performance, were separated from our primary test set, but still included to assess potential bias in this group of calls (2021 w/o category, Table 1).
+The 1-1-2 training data differed from the MH-1813 data regarding age, male/female ratio, and stroke prevalence (Table 1).
+We therefore performed an ablation study where 1-1-2 data were not used for training to assess whether this difference negatively impacted model performance.
+The training, validation, and test subsets of the MH-1813 data had similar characteristics, whereas the 2021 data without diagnostic categories differed in age and sex.
 
-The classification model outperformed the call-takers (\cref{tab_retrospective:table1}), with significant differences in all metrics (p < 0.0001).
-Excluding the 1-1-2 call line training data significantly degraded the model's performance (p < 0.0001), despite the domain mismatch with the MH-1813 call line test data.
-The performance on the 2021 calls without a diagnostic category was significantly worse than that of the test set regarding F1 score, sensitivity, FPR, and FOR (p < 0.0001).
-
-The ROC curve (\cref{fig_retrospective:figure1}, left) illustrates the potential to increase the sensitivity while maintaining a FPR lower than or equal to that of the call-takers.
-Similarly, the PPV sensitivity curve (\cref{fig_retrospective:figure1}, right) demonstrates that sensitivity can be improved while retaining a PPV higher than that of the call-takers.
-The framework can thus be tuned to a sensitivity of around 73\%, while still having a higher positive predictive value than the human call-taker (\cref{fig_retrospective:figure1}, right) The ensemble model outperformed the individual models regardless of the threshold, except for one that exhibited a slightly better sensitivity at a high FPR exceeding 1.5\%.
-The confusion matrices (\cref{fig_retrospective:figure2}) illustrate the performance differences in absolute numbers, with the model exhibiting more true positives and fewer false positives than the call-takers.
-
-% \todo[inline]{Insert Table 1}
 \begin{table}[t]
@@
 The 2021 data without diagnostic categories were excluded}
-    \label{tab_retrospective:table1}
+    \label{tab_retrospective:table1-population-characteristics}
     \resizebox*{0.98\textwidth}{!}{%
@@
 
-\subsection{Sex and age}
+\subsection{Main results}
 
-The model and call-takers exhibited significantly higher PPV and F1-score in men than in women (p < 0.0001) (\cref{tab_retrospective:table1}).
-The model significantly outperformed the call-takers on all metrics for each sex (p < 0.0001).
-The model performed significantly better in the 65+ group than in the 18-64 year group regarding sensitivity, PPV, and F1-score (p < 0.0001).
-Similarly, the call-takers performed significantly better in the 65+ group than in the 18-64 group regarding PPV and F1-score (p < 0.0001).
-Finally, the model significantly outperformed the call-takers on all metrics in both age groups (p < 0.0001).
+The classification model outperformed the call-takers (Table 21), with significant differences in all metrics (p < 0.0001).
+Excluding the 1-1-2 call line training data significantly degraded the model's performance (p < 0.0001), despite the domain mismatch with the MH-1813 call line test data.
+The performance on the 2021 calls without a diagnostic category was significantly worse than that of the test set regarding F1-score, sensitivity, false positive rate (FPR), and false omission rate (FOR) (p < 0.0001).
+The difference in positive predictive value (PPV) was not significant (p = 0.298).
 
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.9\textwidth]{paper_retrospective/figure1.pdf}
-    \caption[Receiver operator characteristic (ROC) curve and PPV-sensitivity curve for stroke recognition.]{Receiver operator characteristic (ROC) curve and PPV-sensitivity curve.
-Left, the ROC curve and, right, PPV-sensitivity curve (precision-recall curve).
-Models 1-5 are the individual models that make up the ensemble model.}
-    \label{fig_retrospective:figure1}
-\end{figure}
-% \todo[inline]{Insert Figure 1}
+The receiver operating characteristic (ROC) curve (\cref{fig_retrospective:figure1}, left) illustrates the potential to increase the sensitivity while maintaining a FPR lower than or equal to that of the call-takers.
+Similarly, the PPV sensitivity curve (\cref{fig_retrospective:figure1}, right) demonstrates that sensitivity can be improved while retaining a PPV higher than that of the call-takers.
+The framework can thus be tuned to a sensitivity of around 73\%, while still having a higher positive predictive value than the human call-taker (\cref{fig_retrospective:figure1}, right) The ensemble model outperformed the individual models regardless of the threshold, except for one that exhibited a slightly better sensitivity at a high FPR exceeding 1.5\%.
+The confusion matrices (\cref{fig_retrospective:figure2}) illustrate the performance differences in absolute numbers, with the model exhibiting more true positives and fewer false positives than the call-takers.
 
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.9\textwidth]{paper_retrospective/figure2.pdf}
-    \caption[Prediction confusion matrices for stroke recognition.]{Prediction confusion matrices.
-Confusion matrices of predictions for call takers and the model on the test set.
-Numbers for the model are given as the rounded mean over eleven runs.}
-    \label{fig_retrospective:figure2}
-\end{figure}
-% \todo[inline]{Insert Figure 2}
-
-\subsection{Model explainability}
-
-Among the words with a positive rank score, several words are synonymous with stroke, such as ``blood clot", ``haemorrhagic stroke", and ``stroke".
-Ambulances are rarely dispatched because the MH-1813 is not intended for emergencies.
-Therefore, a word like ``ambulance" may also be a strong indicator of call-taker recognition, which the model has learned to mimic.
-Additionally, most of the remaining words can be linked to stroke-related symptoms such as ``double vision", ``difficulties speaking", and ``hangs".
-Particularly, words describing the side of the body where symptoms occur ranked high (such as ``left", ``right", and ``side").
-Finally, some words were also related to the sudden onset of symptoms (including ``suddenly" and ``minutes").
-% \todo[inline]{Insert Table 2}
 \begin{table}[t]
@@
     \caption[Overall stroke recognition performance of model compared to call takers.]{Overall performance on MH-1813 test data, performance without 1-1-2 training data, and performance on data from 2021 without diagnostic categories as well as performance on MH-1813 based on demographic subgroups (age/sex) [mean (95\% CI)].}
-    \label{tab_retrospective:table2}
+    \label{tab_retrospective:table2-main-results}
     \resizebox*{0.98\textwidth}{!}{%
@@
         \midrule
-        Call takers & 25.8 (23.7-27.9) & 52.7 (49.2-56.4) & 17.1 (15.5-18.6) & 0.105 (0.094-0.116) & 0.565 (0.539-0.590) \\
+        Call-takers                                 & 25.8 (23.7-27.9) & 52.7 (49.2-56.4) & 17.1 (15.5-18.6) & 0.105 (0.094-0.116) & 0.565 (0.539-0.590) \\
         Model                                       & 35.7 (35.0-36.4) & 63.0 (62.0-64.1) & 24.9 (24.3-25.5) & 0.082 (0.079-0.085) & 0.419 (0.413-0.426) \\
+        % \makecell[l]{Model \\ w/o 1-1-2 training data} & 32.4 (31.8-33.1) & 60.4 (59.3-61.4) & 22.2 (21.6-22.7) & 0.088 (0.085-0.091) & 0.467 (0.460-0.474) \\
+        % \midrule
 
@@
         \midrule
-        Call takers & 15.9 (13.1-18.5) & 50.5 (43.6-57.2) & 9.40 (7.61-11.18) & 0.036 (0.028-0.043) & 0.353 (0.331-0.375) \\
+        Call-takers                                 & 15.9 (13.1-18.5) & 50.5 (43.6-57.2) & 9.40 (7.61-11.18) & 0.036 (0.028-0.043) & 0.353 (0.331-0.375) \\
         Model                                       & 22.9 (21.8-24.0) & 54.1 (52.1-56.3) & 14.5 (13.8-15.3) & 0.033 (0.031-0.035) & 0.231 (0.226-0.236) \\
@@
         \midrule
-        Call takers & 32.9 (30.1-35.7) & 53.5 (49.4-57.6) & 23.7 (21.4-26.0) & 0.401 (0.352-0.449) & 1.467 (1.373-1.560) \\
+        Call-takers                                 & 32.9 (30.1-35.7) & 53.5 (49.4-57.6) & 23.7 (21.4-26.0) & 0.401 (0.352-0.449) & 1.467 (1.373-1.560) \\
         Model                                       & 42.8 (41.9-43.7) & 66.3 (65.1-67.5) & 31.6 (30.8-32.4) & 0.290 (0.278-0.303) & 1.224 (1.198-1.249) \\
@@
         \midrule
-        Call takers & 30.2 (27.2-33.3) & 53.9 (49.1-58.9) & 21.0 (18.5-23.5) & 0.124 (0.105-0.141) & 0.542 (0.506-0.580) \\
+        Call-takers                                 & 30.2 (27.2-33.3) & 53.9 (49.1-58.9) & 21.0 (18.5-23.5) & 0.124 (0.105-0.141) & 0.542 (0.506-0.580) \\
         Model                                       & 39.0 (38.0-40.1) & 63.7 (62.3-65.2) & 28.1 (27.3-29.0) & 0.097 (0.093-0.102) & 0.435 (0.425-0.445) \\
@@
         \midrule
-        Call takers & 21.9 (19.1-24.6) & 51.3 (46.0-56.6) & 13.9 (12.0-15.8) & 0.090 (0.076-0.103) & 0.582 (0.547-0.616) \\
+        Call-takers                                 & 21.9 (19.1-24.6) & 51.3 (46.0-56.6) & 13.9 (12.0-15.8) & 0.090 (0.076-0.103) & 0.582 (0.547-0.616) \\
         Model                                       & 32.4 (31.4-33.4) & 62.3 (60.7-63.8) & 21.9 (21.1-22.7) & 0.069 (0.066-0.073) & 0.407 (0.399-0.416) \\
@@
 
+\subsection{Sex and age}
+
+The model and call-takers exhibited significantly higher PPV and F1-score in men than in women (p < 0.0001) (\cref{tab_retrospective:table2-main-results}).
+The model significantly outperformed the call-takers on all metrics for each sex (p < 0.0001).
+The model performed significantly better in the 65+ group than in the 18-64 year group regarding sensitivity, PPV, and F1-score (p < 0.0001).
+Similarly, the call-takers performed significantly better in the 65+ group than in the 18-64 group regarding PPV and F1-score (p < 0.0001).
+Finally, the model significantly outperformed the call-takers on all metrics in both age groups (p < 0.0001).
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.9\textwidth]{paper_retrospective/figure1.pdf}
+    \caption[Receiver operator characteristic (ROC) curve and PPV-sensitivity curve for stroke recognition.]{Receiver operator characteristic (ROC) curve and PPV-sensitivity curve.
+Left, the ROC curve and, right, PPV-sensitivity curve (precision-recall curve).
+Models 1-5 are the individual models that make up the ensemble model.}
+    \label{fig_retrospective:figure1}
+\end{figure}
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.9\textwidth]{paper_retrospective/figure2.pdf}
+    \caption[Prediction confusion matrices for stroke recognition.]{Prediction confusion matrices.
+Confusion matrices of predictions for call takers and the model on the test set.
+Numbers for the model are given as the rounded mean over eleven runs.}
+    \label{fig_retrospective:figure2}
+\end{figure}
+
+\subsection{Model explainability}
+
+We performed an occlusion analysis to evaluate the importance of individual words for both positive and negative classifier predictions \cref{tab_retrospective:table3}
+Among the words with a positive rank score, several words are synonymous with stroke, such as ``blood clot", ``haemorrhagic stroke", and ``stroke".
+Ambulances are rarely dispatched because the MH-1813 is not intended for emergencies.
+Therefore, a word like ``ambulance" may also be a strong indicator of call-taker recognition, which the model has learned to mimic.
+Additionally, most of the remaining words can be linked to stroke-related symptoms such as ``double vision", ``difficulties speaking", and ``hangs".
+Particularly, words describing the side of the body where symptoms occur ranked high (such as ``left", ``right", and ``side").
+Finally, some words were also related to the sudden onset of symptoms (including ``suddenly" and ``minutes").
+
 Among the words with a negative rank score, most were strong indicators for specific conditions, symptoms, or body parts that are unrelated to stroke (such as ``tetanus", ``pregnant", ``swollen", ``fever", and ``the knee").
@@
 
-\section{Discussion}
-
-Our results showed that a machine learning framework can substantially improve stroke recognition in medical helpline calls compared to solely relying on human call-takers.
-This improvement was observed across all performance metrics and for basic patient demographics (age and sex).
-Our occlusion analysis revealed that the model relied on the relevant predictive features associated with call-taker triaging, patient symptoms, and treatment.
-
-This study does not imply that a machine learning model can replace medical call-takers.
-The effectiveness of the model is fully reliant on the conversation between the call-taker and caller and the call-taker's ability to skillfully triage the patient.
-Instead, the model should be used as a supportive tool for call-takers in the decision-making process, contributing to a higher recognition of patients with stroke and potentially boosting the confidence of call- takers in their decisions.
-A similar machine learning model designed to predict cardiac arrest was tested in a randomised controlled trial (RCT) at CEMS.15 The results highlighted the necessity of incorporating input from call-takers.
-The machine learning model for cardiac arrest has subsequently been implemented in daily practice at CEMS, in a setup similar to the one presented in our study.
-However, implementation of our framework requires further investigation.
-The relative performance gap between call-takers and the model was larger in our study than in the cardiac arrest study,15 which may affect the results of a potential RCT.
-To support future work and discussions beyond the scope of this study, \cref{app:supplementary-retrospective} includes the results of a simulation of a live implementation where call-takers are assumed to follow a set of fixed rules based on the output of the machine learning framework.
-
-The performance gap between the model and call-takers could be explained by the rarity of stroke calls to MH-1813 (0.250\% of all calls in 2021), which might affect call-taker awareness of stroke as a possible cause of certain symptoms.
-Additionally, certain stroke symptoms are so rare that some call-takers may never encounter them, increasing the risk of false negatives.
-The model was trained on more calls than any single call-taker would handle in a lifetime, enabling it to recognise even rare descriptors of stroke.
-The model is specifically trained to recognise strokes and exclusively learns from actual stroke descriptions, unlike call-takers, who are trained with generalised teaching materials to triage many different conditions.
-Therefore, call-takers may not have received specific training for patients with stroke and may never have encountered them.
-
-The model performed significantly better on men than on women.
-This could be attributed to several factors.
-First, the model may have learned to mimic call-takers with the same bias.
-Second, women may experience different and challenging-to-identify symptoms than men \cite{cite29, cite30}.
-Third, a higher prevalence of male patients with stroke was observed in the training data.
-Despite these potential sources of bias, the model exhibited less bias than call-takers did.
-That is, the relative performance improvements were higher for women than for men.
-This bias could be further reduced using advanced data augmentation and balanced data when training a machine learning model.
-However, such measures may degrade overall performance.
-
-This study has some limitations.
-First, the mapping of call recordings to electronic records was incomplete due to technical limitations in the CAD registry, which limited the number of calls available to us.
-Of note, there was no obvious pattern of bias related to the unmapped calls, and we included all calls with matching audio files, regardless of dispatcher performance.
-The results could potentially be improved if more calls were available for analysis.
-Second, calls without a call-taker indicated diagnostic category were not included in the validation and test data because the call-taker's performance could not be evaluated.
-Moreover, in exploratory analyses, the model performed worse on these calls, which might be attributed to differences in population characteristics (\cref{tab_retrospective:table3}).
-Finally, the ground truth stroke labelling relied on the patient-reported time of onset being exact; however, estimating the accuracy of the timestamps in DanStroke was impossible.
-
-The proposed machine learning framework can be improved.
-In preliminary experiments, we tested convolutional, recurrent, and self-attention architectures for text classification; however, this did not improve the model.
-One promising path is to leverage the abundance of non-stroke calls.
-Self-supervised learning techniques for speech and text data can be used to improve speech recognition and text classification models.
-This approach can enable direct classification from audio input without speech recognition.
-Finally, incorporating additional data from CAD records through multitask learning could be used to inform the classifier and predict the diagnostic category to improve its ability to identify more fine-grained characteristics of non-stroke calls.
-
-In conclusion, using the largest collection of audio calls from patients with stroke to date, we developed a machine learning framework that significantly outperformed human call-takers in stroke recognition in medical helpline calls.
-The framework can assist human call-takers during medical helpline calls.
-Ideally, this would enable a higher recognition of patients with stroke in the prehospital setting, benefiting both patient outcomes and health service resource allocation.
-
-\subsection{Contributors}
-
-JW, JDH and LB share lead authorship and contributed equally to the original drafting of the paper, review and editing, data curation, formal analysis, methodology, validation, and conceptualisation.
-JDH and LB contributed to software development and funding acquisition.
-SNFB contributed to the conceptualisation, data curation, funding acquisition, supervision, and writing - review and editing.
-LM contributed to conceptualisation, funding acquisition, project administration, supervision, and writing - review and editing.
-MS contributed to supervision and writing - review and editing.
-HC contributed to the methodology, supervision, and writing - review and editing.
-CK contributed to the conceptualisation, funding acquisition, methodology, project administration, supervision, and writing - review and editing.
-HCC contributed to the conceptualisation, data curation, funding acquisition, methodology, project administration, supervision, and writing - review and editing.
-HCC and CK share last authorship.
-The three lead authors directly accessed and verified the underlying data reported in this manuscript.
-JW's affiliations are exclusively academic.
-
-\section{Methods}
-
-\subsection{Study design}
-
-We used call recordings and registry data from the Copenhagen Emergency Medical Services (CEMS) and the Danish Stroke Registry \cite{cite16} from 2015 to 2020.
-These data were used to train a machine learning model to classify medical helpline calls as stroke or non-stroke.
-We compared the performances of the model with that of the call-taker using data from 2021.
-
-\subsection{Data sources}
-
-\subsubsection{CEMS}
-
-The CEMS is responsible for providing prehospital telehealth services in the Capital Region of Denmark, with a catchment area of 1.9 million \cite{cite17}.
-CEMS operates two call lines: the 1-1-2 emergency line, similar to 9-1-1 in the United States, intended for acute conditions.
-The other is the medical helpline 1813 (MH-1813, pronounced ``'eighteen-thirteen") intended for non- life-threatening conditions that cannot wait until a general practitioner is available.18
-
-Call-takers for both lines, who are nurses, paramedics, or physicians, can dispatch ambulances.
-The condition suspected by the call-taker is categorised based on a predefined diagnostic index and stored in an electronic record using a computer-aided dispatch (CAD) system.
-The CAD records are associated with the Danish civil registration number (CPR number) \cite{cite19} of the patient.
-The CPR number is a unique identification assigned to all Danish residents.
-It is used for interactions with health services and registries, enabling cross- referencing of the data sources used in this study.
-The call audio is recorded and stored separately from the CAD recordings using a telephone system.
-
-\subsubsection{Danish Stroke Registry (DanStroke)}
-
-All patients with stroke admitted to a Danish hospital within 5 days of symptom onset are recorded in the Danish Stroke Registry \cite{cite16}, also known as DanStroke.
-This record includes the patient-reported time of onset, stroke type (haemorrhagic, ischaemic, or transitory ischaemic attack), and CPR number of the patient.
-
-\subsubsection{Data ethics}
-
-The Danish Data Protection Agency (P-2021-475) approved this study.
-Danish law did not require approval from the Scientific Ethics Committee because the data were registry-based.
-CEMS approved the transcription of all calls made to 1-1-2 and MH-1813.
-All electronic records were anonymised before analysis, and the researchers did not inspect the calls manually.
-
-\subsection{Study scope}
-
-Stroke prevalence in calls made to the MH-1813 is lower than that in calls made to 1-1-2.
-Patients with stroke may exhibit different symptoms and symptom severity because MH- 1813 is meant for low-acuity incidents, leading to reduced recognition.
-In addition, MH-1813 call-takers dispatch high-priority transport less frequently, which may affect optimal treatment timing.
-Therefore, we focused on MH-1813 in this study.
-
-\subsection{Stroke dataset}
-
-\subsubsection{Cross-referencing data sources}
-
-From the CAD medical records, we included all calls that could be matched to a corresponding audio file.
-for 1-1-2 and MH-1813 from 2015 to 2021 for patients older than 18.
-The CAD records were matched with the telephone call recordings based on the call start, call duration, and call-taker identity.
-Due to data incompleteness, and the way the audio data is stored, at CEMS, 2,730,199 contacts could not be matched to their corresponding audiofile, however, 2,361,178 contacts were successfully matched.
-We found no obvious pattern in the matched and unmatched calls and we included all calls with at matching audio file.
-Next, a call was regarded as a case of ground truth stroke positive when the CPR number in the CAD record matched that of a DanStroke record, and the patient-reported time of onset was close to the call start time.
-We allowed a window of 72 hours before and 24 hours after the call starts to account for uncertainty in recording stroke onset time.
-We excluded calls involving subarachnoid haemorrhage cases.
-Finally, we considered a call to be a call-taker stroke positive when the call-taker selected the stroke diagnostic category during the call and dispatched an ambulance with the appropriate level of response \cite{cite20}.
-To ensure that the effect of the machine learning framework was not overestimated, we excluded calls where diagnostic category had not been registered from the final testing.
-We still reported the population characteristics and model performance of this group of calls to assess potential bias introduced by excluding them.
-A data-flow diagram is included in \cref{app:supplementary-retrospective}.
-The resulting dataset is the largest dataset of audio files from stroke calls collected to date.
-
-\subsubsection{Dataset splitting}
-
-We reserved all the MH-1813 calls from 2021 for testing.
-We used stratified sampling to split the MH-1813 calls from 2015 to 2020 into six folds: one for validation and five for training.
-The calls were stratified based on the ground truth stroke label and the presence of a diagnostic category.
-Only calls without diagnostic categories were included in the training set.
-The 1-1-2 calls were used only for training; however, calls from 2021 were discarded to avoid temporal overlap with the test period.
-
-As expected, the 1-1-2 training data differed from the MH-1813 data regarding age, male/female ratio, and stroke prevalence (\cref{tab_retrospective:table3}).
-We performed an ablation study where 1- 1-2 data were not used for training to assess whether this difference negatively impacted model performance.
-The training, validation, and test subsets of the MH-1813 data had similar characteristics, whereas the 2021 data without diagnostic categories differed in age and sex.
-
-\subsection{Machine learning pipeline}
-
-We employed a two-step machine learning pipeline.
-First, a call was transcribed using the speech recognition model.
-Second, the transcript was used as input for the text classification model.
-The final output score was used to classify whether the call concerned a stroke.
-The pipeline is illustrated in \cref{app:supplementary-retrospective}.
-
-% \todo[inline]{Insert Table 3}
 \begin{table}[t]
@@
 
+\section{Discussion}
+
+Our results showed that a machine learning framework can substantially improve stroke recognition in medical helpline calls compared to solely relying on human call-takers.
+This improvement was observed across all performance metrics and for basic patient demographics (age and sex).
+Our occlusion analysis revealed that the model relied on the relevant predictive features associated with call-taker triaging, patient symptoms, and treatment.
+
+This study does not imply that a machine learning model can replace medical call-takers.
+The effectiveness of the model is fully reliant on the conversation between the call-taker and caller and the call-taker's ability to skillfully triage the patient.
+Instead, the model should be used as a supportive tool for call-takers in the decision-making process, contributing to a higher recognition of patients with stroke and potentially boosting the confidence of call- takers in their decisions.
+A similar machine learning model designed to predict cardiac arrest was tested in a randomised controlled trial (RCT) at CEMS.15 The results highlighted the necessity of incorporating input from call-takers.
+The machine learning model for cardiac arrest has subsequently been implemented in daily practice at CEMS, in a setup similar to the one presented in our study.
+However, implementation of our framework requires further investigation.
+The relative performance gap between call-takers and the model was larger in our study than in the cardiac arrest study,15 which may affect the results of a potential RCT.
+
+To support future work and discussions beyond the scope of this study, \cref{app:supplementary-retrospective} includes the results of a simulation of a live implementation where call-takers are assumed to follow a set of fixed rules based on the output of the machine learning framework.
+For instance, in one simulation call-takers are assumed to change any stroke negative to a positive, if the model predicts a positive.
+While the results of the simulation are encouraging, it is important to stress that it is not practically feasible to use a fixed rule set to overrule the call-taker.
+These results should only be seen as a preliminary indicator of a potential RCT.
+In practice, a nuanced set of guidelines should be developed over several iterations of implementation and testing.
+
+The performance gap between the model and call-takers could be explained by the rarity of stroke calls to MH-1813 (0.250\% of all calls in 2021), which might affect call-taker awareness of stroke as a possible cause of certain symptoms.
+Additionally, certain stroke symptoms are so rare that some call-takers may never encounter them, increasing the risk of false negatives.
+The model was trained on more calls than any single call-taker would handle in a lifetime, enabling it to recognise even rare descriptors of stroke.
+The model is specifically trained to recognise strokes and exclusively learns from actual stroke descriptions, unlike call-takers, who are trained with generalised teaching materials to triage many different conditions.
+Therefore, call-takers may not have received specific training for patients with stroke and may never have encountered them.
+
+The model performed significantly better on men than on women.
+This could be attributed to several factors.
+First, the model may have learned to mimic call-takers with the same bias.
+Second, women may experience different and more challenging-to-identify symptoms than men \cite{cite29, cite30}.
+Third, a higher prevalence of male patients with stroke was observed in the training data.
+Despite these potential sources of bias, the model exhibited less bias than call-takers did.
+That is, the relative performance improvements were higher for women than for men.
+This bias could be further reduced using advanced data augmentation and balanced data when training a machine learning model.
+However, such measures may degrade overall performance.
+
+The improved sensitivity and PPV on the 65+ years group may be explained a higher prior probability of stroke for older patients and stronger evidence from the patient's medical history.
+The relatively high FOR and FPR for the 65+ group is likely to be a result of the much higher prevalence of stroke cases compared to the 18-64 year olds (0.85\% vs.
+0.07\%).
+We did not have data to estimate potential bias related to race, ethnicity, language, accent, or dialects.
+Previous studies on speech recognition for call centres have indeed found that non-native speakers had a higher rate of transcription errors \cite{han_deep_2017}.
+Since our model was trained on a representative - and therefore unbalanced - sample, we expect it to behave similarly.
+Future research should look to address these shortcomings, for example by utilising self-supervised learning on massive amounts of diverse, unlabelled data covering multiple languages, accents, and dialects.
+
+Due to European data regulations (GDPR), it was not possible to manually transcribe MH-1813 calls to train a new speech recognition model, so we had to rely on an existing solution.
+This also meant that we could not evaluate the word error rate (WER) of the model.
+Instead, we used the downstream performance of the text classification model when trained in combination with different speech recognition models to choose the best option.
+Since the focus of this study is the ability to correctly recognise stroke, and not the performance of the speech recognition model alone, this approach is better suited.
+Indeed, the WER might be misleading when choosing a speech recognition model for a specific task.
+For instance, one model might fail to predict redundant minimal response words (e.g.
+“uh” and “uhm”) and make small inflection errors (e.g.
+“clot” instead of “clots”), which results in a relatively high WER, while another model only fails to predict rare, specialised words that are highly indicative of stroke (e.g.
+“haemorrhage” and “thrombolysis”), which results in a relatively low WER.
+
+Although we believe that the proposed machine learning framework can be further improved, several alternatives have already been explored in the preliminary experimental phase.
+The speech recognition model we used was trained on 1-1-2 calls for a previous project \cite{cite14}, and so, was specialised to a domain very similar to that of MH-1813.
+We also tested an open-source, multilingual model from OpenAI called Whisper \cite{radford_robust_2023}, but found that performance degraded slightly compared to the model trained on 1-1-2.
+We hypothesise that this is due to Whisper's inability to handle the specific noise conditions and recognise words from a specialised medical vocabulary.
+
+For text classification, we used an ensemble of multi-layer perceptrons (MLPs).
+We also tested convolutional, recurrent, and self-attention (i.e.
+Transformer) architectures.
+However, this did not improve performance.
+In addition, we tested a pre-trained self-supervised model.
+Although many of these models are freely available to the public, they are primarily trained on English data.
+Only relatively few options exist for the Danish language, none of which are specialised in the medical domain.
+We used a monolingual Danish BERT model, which has previously been shown to outperform a multilingual alternative from Google for Danish named entity recognition \cite{hvingelby_dane_2020}.
+However, this also did not result in a significant performance improvement.
+We hypothesise that the number of ground truth stroke positives was too small for these advanced models to learn more complex patterns than the MLP ensemble.
+In addition, a self-supervised model would likely benefit from being pre-trained on speech or text data from the target domain.
+Although training such large-scale foundation models have the potential to improve the classification model further, it is beyond the scope of this study.
+Thus, we chose the simpler MLP ensemble.
+We have included references to reviews of self-supervised learning for speech and text in the references \cite{gururangan_don_2020,mohamed_selfsupervised_2022}.
+Notably, it is not uncommon for small, simple models to match or outperform large, pre-trained models for text-classification tasks \cite{galke_bagofwords_2022}.
+
+This study has some limitations.
+First, the mapping of call recordings to electronic records was incomplete due to technical limitations in the computer-aided dispatch (CAD) registry, which limited the number of calls available to us.
+Of note, there was no obvious pattern of bias related to the unmapped calls, and we included all calls with matching audio files, regardless of dispatcher performance.
+The results could potentially be improved if more calls were available for analysis.
+Second, calls without a call-taker indicated diagnostic category were not included in the validation and test data because the call-taker's performance could not be evaluated.
+Moreover, in exploratory analyses, the model performed worse on these calls, which might be attributed to differences in population characteristics (\cref{tab_retrospective:table1-population-characteristics,tab_retrospective:table2-main-results}).
+Finally, the ground truth stroke labelling relied on the patient-reported time of onset being exact; however, estimating the accuracy of the timestamps in DanStroke was impossible.
+
+In conclusion, using the largest collection of audio calls from patients with stroke to date, we developed a machine learning framework that significantly outperformed human call-takers in stroke recognition in medical helpline calls.
+The framework can assist human call-takers during medical helpline calls.
+Ideally, this would enable a higher recognition of patients with stroke in the prehospital setting, benefiting both patient outcomes and health service resource allocation.
+
+\section{Methods}
+
+\subsection{Data sources}
+
+\subsubsection{Copenhagen Emergency Medical Services (CEMS)}
+
+The CEMS is responsible for providing prehospital telehealth services in the Capital Region of Denmark, with a catchment area of 1.9 million \cite{cite17}.
+CEMS operates two call lines: the 1-1-2 emergency line, similar to 9-1-1 in the United States, intended for acute conditions.
+The other is the medical helpline 1813 (MH-1813, pronounced ``'eighteen-thirteen") intended for non- life-threatening conditions that cannot wait until a general practitioner is available.18
+
+Call-takers for both lines, who are nurses, paramedics, or physicians, can dispatch ambulances.
+The condition suspected by the call-taker is categorised based on a predefined diagnostic index and stored in an electronic record using a computer-aided dispatch (CAD) system.
+The CAD records are associated with the Danish civil registration number (CPR number) \cite{cite19} of the patient.
+The CPR number is a unique identification assigned to all Danish residents.
+It is used for interactions with health services and registries, enabling cross- referencing of the data sources used in this study.
+The call audio is recorded and stored separately from the CAD recordings using a telephone system.
+
+\subsubsection{Danish Stroke Registry (DanStroke)}
+
+All patients with a final diagnosis of stroke or transient ischaemic attack admitted to a Danish hospital within 5 days of symptom onset are recorded in the Danish Stroke Registry \cite{cite16}, also known as DanStroke.
+This record includes the patient-reported time of onset, stroke type (haemorrhagic, ischaemic, or transient ischaemic attack), and CPR number of the patient.
+The diagnosis is obtained according to the national guidelines \cite{baluenfeldt_national_2021}, which includes cerebral imaging and full diagnostic workup by neurologists.
+The validity of the Danish stroke registry has been shown to be high \cite{wildenschild_registration_2013}, and the number of stroke mimics is therefore minimised in our dataset.
+
+\subsubsection{Inclusion and ethics}
+
+The Danish Data Protection Agency (P-2021-475) approved this study.
+Danish law did not require approval from the Scientific Ethics Committee because the data were registry-based.
+CEMS approved the transcription of all calls made to 1-1-2 and MH-1813.
+All electronic records were anonymised before analysis, and the researchers did not inspect the calls manually.
+
+\subsection{Study scope}
+
+Stroke prevalence in calls made to the MH-1813 is lower than that in calls made to 1-1-2.
+Patients with stroke may exhibit different symptoms and symptom severity because MH- 1813 is meant for low-acuity incidents, leading to reduced recognition.
+In addition, MH-1813 call-takers dispatch high-priority transport less frequently, which may affect optimal treatment timing.
+Therefore, we focused on MH-1813 in this study.
+
+\subsection{Stroke dataset}
+
+\subsubsection{Cross-referencing data sources}
+
+From the CAD medical records, we included all calls that could be matched to a corresponding audio file.
+for 1-1-2 and MH-1813 from 2015 to 2021 for patients older than 18.
+The CAD records were matched with the telephone call recordings based on the call start, call duration, and call-taker identity.
+Due to data incompleteness, and the way the audio data is stored, at CEMS, 2,730,199 contacts could not be matched to their corresponding audiofile, however, 2,361,178 contacts were successfully matched.
+We found no obvious pattern in the matched and unmatched calls and we included all calls with at matching audio file.
+Next, a call was regarded as a case of ground truth stroke positive when the CPR number in the CAD record matched that of a DanStroke record, and the patient-reported time of onset was close to the call start time.
+We allowed a window of 72 hours before and 24 hours after the call starts to account for uncertainty in recording stroke onset time.
+We excluded calls involving subarachnoid haemorrhage cases.
+Finally, we considered a call to be a call-taker stroke positive when the call-taker selected the stroke diagnostic category during the call and dispatched an ambulance with the appropriate level of response \cite{cite20}.
+To ensure that the effect of the machine learning framework was not overestimated, we excluded calls where diagnostic category had not been registered from the test set.
+We still reported the population characteristics and model performance of this group of calls to assess potential bias introduced by excluding them.
+A data-flow diagram is included in \cref{app:supplementary-retrospective}.
+The resulting dataset is the largest dataset of audio files from stroke calls collected to date.
+
+\subsubsection{Dataset splitting}
+
+We reserved all the MH-1813 calls from 2021 for testing.
+We used stratified sampling to divide the MH-1813 calls from 2015 to 2020 into validation and training subsets.
+The training subset was further split into five folds which were used for ensemble training.
+The calls were stratified based on the ground truth stroke label and the presence of a diagnostic category.
+Calls without diagnostic categories were only included in the training set.
+The 1-1-2 calls were used only for training; however, calls from 2021 were discarded to avoid temporal overlap with the test period.
+
+\subsection{Machine learning pipeline}
+
+We employed a two-step machine learning pipeline.
+First, a call was transcribed using the speech recognition model.
+Second, the transcript was used as input for the text classification model.
+The final output score was used to classify whether the call concerned a stroke.
+The pipeline is illustrated in \cref{app:supplementary-retrospective}.
+
 \subsubsection{Speech recognition}
@@
 
-We selected the prediction threshold as the harmonic mean of the thresholds for each model in the ensemble to ensure that the sensitivity and positive predictive value (PPV) were equal to those of the call-takers.
+For each model in the ensemble, we selected the prediction threshold as the harmonic mean of the two thresholds that respectively ensure sensitivity and PPV equal to that of the call-takers.
 This simplifies the comparison by ensuring a trade-off between sensitivity and PPV, similar to that of call-takers.
@@
 We trained five models for the text classification ensemble using binary cross-entropy after transcribing all calls in the dataset using the speech recognition model.
-One training fold was used for early stopping, whereas the remaining four folds and 1-1-2 data were used for training.
+One training fold was used for early stopping using the F1-score, whereas the remaining four folds and 1-1-2 data were used for training.
 Thus, each model in the ensemble was trained and validated using different datasets.
-We ran a grid search with 96 different hyperparameter configurations and selected the ensemble model with the best F1 score for the validation set.
+We ran a grid search with 96 different hyperparameter configurations and selected the ensemble model with the best F1-score for the validation set.
 
@@
 
-We present the results with and without 1-1-2 training data, subgroup analyses based on age (18-64/65+) and sex (male/female), and call-takers performance.
-We also report the model performance on calls without a diagnostic category from the test year 2021 to assess potential data bias.
-We tested our results for statistical significance using approximate permutation tests, paired or independent, as appropriate and computed 95\% confidence intervals (CIs) using bootstrapping \cite{dwass_modified_1957,eden_validity_1933}.
+We tested our results for statistical significance using approximate permutation tests.
+We used one-sided paired approximate permutation tests for model-to-model and model-to-call-taker comparisons when done on the same subset.
+For comparisons across different subsets (e.g.
+male vs.
+female) we used one-sided independent approximate permutation tests.
+We computed 95\% confidence intervals (CIs) using bootstrapping.
+\cite{dwass_modified_1957,eden_validity_1933}.
 In our assessment, we accounted for random variation associated with model training by basing the means, tests, and CIs on the predictions of 11 randomly initialised training runs.
-Statistical significance was defined as a two-sided p-value of < 0.05.
+Statistical significance was defined as a p-value of < 0.05.
 
@@
 The source code can be shared using a Creative Commons NC-ND 4.0, an international licence, upon reasonable written request to the corresponding author, and requires a data use agreement.
+
+\section{Code availability}
+
+The source code can be shared using a Creative Commons NC-ND 4.0, an international license, upon reasonable written request to the corresponding author, and requires a data use agreement.
 
@@
 
+\subsection{Author contributions}
+
+JW, JDH and LB share lead authorship and contributed equally to the original drafting of the paper, review and editing, data curation, formal analysis, methodology, validation, and conceptualisation.
+JDH and LB contributed to software development and funding acquisition.
+SNFB contributed to the conceptualisation, data curation, funding acquisition, supervision, and writing - review and editing.
+LM contributed to conceptualisation, funding acquisition, project administration, supervision, and writing - review and editing.
+MS contributed to supervision and writing - review and editing.
+HC contributed to the methodology, supervision, and writing - review and editing.
+CK contributed to the conceptualisation, funding acquisition, methodology, project administration, supervision, and writing - review and editing.
+HCC contributed to the conceptualisation, data curation, funding acquisition, methodology, project administration, supervision, and writing - review and editing.
+HCC and CK share last authorship.
+The three lead authors directly accessed and verified the underlying data reported in this manuscript.
+JW's affiliations are exclusively academic.
+
 }

### 2023-09-22 appendices/supplementary_retrospective.tex
Update figure placement

@@
 \section{Data flow digram}
+In \cref{fig_retrospective:data_flowchart} we present an overview of the data flow from the initial data sources to the final stroke dataset.
 
-\begin{figure}[h]
+\begin{sidewaysfigure}[h]
     \centering
-    \includegraphics[width=0.9\textwidth]{paper_retrospective/data_flowchart.pdf}
+    \includegraphics[width=0.8\textwidth]{paper_retrospective/data_flowchart.pdf}
     \caption[Overview of data flow from the initial data sources to the final stroke dataset.]{Overview of data flow from the initial data sources to the final stroke dataset.}
     \label{fig_retrospective:data_flowchart}
-\end{figure}
+\end{sidewaysfigure}
 
@@
 
-\begin{table}[h]
+\begin{table}[t]
     \centering
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
     \centering

### 2023-09-22 chapters/background.tex
Work on background

@@
 \cite{ren_likelihood_2019}.
-This is similar to using simulated OOD inputs from GAN \cite{lee_training_2018} or using auxiliary datasets of outliers \cite{hendrycks_deep_2019} for calibration purpose.
+This is similar to using simulated OOD inputs from generative adversarial networks \cite{lee_training_2018} or using auxiliary datasets of outliers \cite{hendrycks_deep_2019} for calibration purpose.
 %   - X Deep semi-supervised anomaly detection \cite{ruff_deep_2020}.
@@
 \textcite{vyas_outofdistribution_2018} train an ensemble of classifiers on different subsets of the training data, with the left out data taken as OOD, and propose novel loss over $p(y|\xb)$ that seeks to maintain a set margin between its average entropy for the OOD and in-distribution examples.
-Another approach generates OOD inputs using a GAN \cite{lee_training_2018}. 
+Another approach generates OOD inputs using a generative adversarial network \cite{lee_training_2018}. 
 Similar to actual OOD data, the usefulness of synthetic OOD data is also fundamentally limited by the intractability of sampling the complete space of OOD data.
@@
 \cite{ren_likelihood_2019}.
-This is similar to using simulated OOD inputs from GAN \cite{lee_training_2018} or using auxiliary datasets of outliers \cite{hendrycks_deep_2019} for calibration purpose.
+This is similar to using simulated OOD inputs from generative adversarial networks \cite{lee_training_2018} or using auxiliary datasets of outliers \cite{hendrycks_deep_2019} for calibration purpose.
 
@@
 % \cite{
-%     X sakurada_anomaly_2014                 encode and decode data with autoencoders.
+%     X sakurada_anomaly_2014                 encode and decode data with autoencoders. Dimensionality reduction
+%     X lyudchik_outlier_2016                 encode and decode data with autoencoders.
 %     X xia_learning_2015                     encode and decode data with autoencoders.
-%     zhou_anomaly_2017                     encode and decode data with autoencoders.
-%     zong_deep_2018                        encode and decode data with autoencoders.
-%     lyudchik_outlier_2016                 encode and decode data with autoencoders.
-%     chen_outlier_2017                     encode and decode data with an ensemble of autoencoders.
-%     schlegl_unsupervised_2017             perform GAN inversion for a data point, and evaluate its reconstruction error and discriminator confidence under the inverted latent variable.
-%     li_madgan_2019                        perform GAN inversion for a data point, and evaluate its reconstruction error and discriminator confidence under the inverted latent variable.
-%     graham_denoising_2023                 leverages diffusion models to reconstruct images at varied diffusion steps, while we mask and inpaint an image repeatedly with fixed steps.
-%     liu_unsupervised_2023a                LMD lifts an image off its original manifold by corrupting it, and maps it towards the in-domain manifold with a diffusion model.
+We observe that when data are reconstructed from low-dimensional representations, the inliers and the outliers can be well separated according to their reconstruction errors.
+Based on this basic observation, we gradually inject discriminative information in the learning process of an autoencoder to make the inliers and the outliers more separable.
+%     X zhou_anomaly_2017                     encode and decode data with autoencoders.
+inspired by Robust Principal Component Analysis, and we split the input data X into two parts, , where  can be effectively reconstructed by a deep autoencoder and  contains the outliers and noise in the original data X
+%     X zong_deep_2018                        encode and decode data with autoencoders.
+jointly learn a deep autoencoder and a Gaussian Mixture Model on the learned hidden representations and link their method to neural variational inference \cite{mnih_neural_2014}.
+%     X chen_outlier_2017                     encode and decode data with an ensemble of autoencoders.
+%     X schlegl_unsupervised_2017             perform generative adversarial networks inversion for a data point, and evaluate its reconstruction error and discriminator confidence under the inverted latent variable.
+%     X li_madgan_2019                        perform generative adversarial networks inversion for a data point, and evaluate its reconstruction error and discriminator confidence under the inverted latent variable.
+%     X graham_denoising_2023                 leverages diffusion models to reconstruct images at varied diffusion steps, while we mask and inpaint an image repeatedly with fixed steps.
+%     X liu_unsupervised_2023a                LMD lifts an image off its original manifold by corrupting it, and maps it towards the in-domain manifold with a diffusion model.
 For an out-of-domain image, the mapped image would have a large distance away from its original manifold, and LMD would identify it as OOD accordingly.
@@
 A number of methods derive the OOD score from a reconstruction error. 
-Among the first methods are \textcite{sakurada_anomaly_2014} and \textcite{xia_learning_2015} who use deep autoencoders to reconstruct the input and use the reconstruction error as the OOD score. 
+Among the first methods are \textcite{sakurada_anomaly_2014, lyudchik_outlier_2016} who note that dimensionality reduction helps separate inliers and outliers and propose to use deep autoencoders to reconstruct the input and evaluate the reconstruction error.
+\textcite{xia_learning_2015} take a similar approach but also propose to inject discriminative information in the learning process.
+Drawing inspiration from Robust Principal Component Analysis, \textcite{zhou_anomaly_2017} propose to first split the input data into a dense low-rank factor and a sparse factor, assuming that outliers are caught in the sparse factor, and then use a deep autoencoder to reconstruct the dense factor.
+\textcite{zong_deep_2018} jointly learn a deep autoencoder and a Gaussian Mixture Model on the learned hidden representations and draw parallels of their method to neural variational inference \cite{mnih_neural_2014}.
+Similarly to the ensemble-based methods for supervised OOD detection, \textcite{chen_outlier_2017} propose to use an ensemble of autoencoders to reconstruct the input and use the median reconstruction error as the OOD score.
+
+Although generative adversarial networks do not have the ability to encode a given data point, methods have been proposed to invert the generator to find a latent representation that can be used to reconstruct the input and use the reconstruction error as well as a discriminator score for OOD detection \cite{schlegl_unsupervised_2017,li_madgan_2019}.
+
+Diffusion models have also been used for OOD detection via a reconstruction-based score.
+\textcite{graham_denoising_2023} add varying amounts of diffusion noise to an input image and show that reconstructions of OOD inputs from appropriate noise levels fall back onto the in-domain manifold resulting in high reconstruction error.
+\textcite{liu_unsupervised_2023a} lift an image off its original manifold by sampling a number of masks, and then maps it towards the in-domain manifold with a diffusion model, using the median reconstruction error as the OOD score.
+
 % In \textcite{zhou_anomaly_2017}
@@
 %     sehwag_ssd_2021,                      learn a representation over the in-domain data through self-supervised training
-%     xiao_we_2021                          further shows that one can instead use a strong pretrained feature extractor while maintaining comparable performance.
+%     X xiao_we_2021                          further shows that one can instead use a strong pretrained feature extractor (self-supervised)  while maintaining comparable performance.
 % }
 %
+
+\texticite{xiao_we_2021} propose to use an existing, strong foundation model, pre-trained with a self-supervised objective, to extract features from the input, fit a parametric density estimator to the features and use the Mahalanobis distance as the OOD score.
 

### 2023-09-22 chapters/background.tex
Work on background

@@
 To supplement the short related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}, we will then provide a concise review of the literature on out-of-distribution detection.
+A
 
@@
 %
-OOD detection is the task of identifying data that is unlikely to originate from the distribution of the training data and, in the context of neural networks, dates back several decades \cite{bishop_novelty_1994, chang_figure_1993}.
-In the literature, the task is also referred to as open-set \cite{bendale_open_2016, cardoso_weightless_2017, panaredabusto_open_2017}, outlier-rejection \cite{xu_deep_2014, mor_confidence_2018}, or selective prediction \cite{geifman_selective_2017}.
-A closely related field is that of network-based uncertainty estimation [14, 10, 36, 20].
-In the general case, we assume that we have a domain of in-distribution data, $\mathcal{D}_{\text{in}}$, and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from that domain or not.
+OOD detection is the task of identifying test data that is unlikely to originate from the distribution of the training data and, in the context of neural networks, dates back several decades \cite{bishop_novelty_1994, chang_figure_1993}.
+In the general case, we assume that we have a domain of in-distribution data, $\mathcal{D}_{\text{in}}$, and we would like to build a model that can be used to assess whether a test data point $\xb$ originates from that domain or not.
 Since such OOD data is unknown to the model, it is, by definition, a source of epistemic uncertainty.
@@
 
+% In the literature, the task is also referred to as open set recognition \cite{bendale_open_2016, cardoso_weightless_2017, panaredabusto_open_2017}, outlier detection \cite{xu_deep_2014, mor_confidence_2018}, or selective prediction \cite{geifman_selective_2017} .
+OOD detection bears many similarities with anomaly detection, novelty detection, open set recognition, and outlier detection \cite{yang_generalized_2022}. 
+Some of these differences are subtle, and the terminology is not always used consistently in the literature.
+To provide some clarity, we will briefly outline the taxonomy of \textcite{yang_generalized_2022}:
+Outlier detection most clearly differs from the other tasks by directly processesing all observations and aiming to select outliers from a single contaminated dataset. 
+The remaining tasks differ in whether they detect both covariate and semantic shift and require the simultaneous classification of in-distribution classes.%, and whether the in-distribution data contains more than one class.
+Anomaly detection deals with multiclass data and detects both covariate and semantic shift without requiring simultaneous classification of in-distribution classes whereas novelty detection, open set recognition and OOD detection are usually concerned with semantic shifts.
+Although a vague difference, novelty detection usually defined in terms of a class of normal data, while OOD detection centers around the distribution of the training data.
+Different from novelty detection, OOD detection methods sometimes draw on examples of OOD data or require the simultaneous classification of in-distribution classes, although this is not always the case.
+OOD detection benchmarks almost always take OOD data to be from external datasets different from the training dataset which distinguishes it from open set recognition that usually benchmarks on a single dataset split into ID and OOD data.
+% Another closely related field is that of network-based uncertainty estimation \cite{neal_bayesian_1995, hernandez-lobato_probabilistic_2015, gawlikowski_survey_2023}. %[14, 10, 36, 20]. 
+Although we share the sentiment of \textcite{yang_generalized_2022}, who propose to unify the tasks as ``generalized out-of-distribution detection'', this thesis will follow the nomenclature used in the works most related to ours and refer to the task as out-of-distribution detection.
+
 Recently, several approaches for deep neural networks have been developed that formally address the rejection of samples $\xb\notin\mathcal{D}_{\text{in}}$. 
 One way to categorize the different approaches is based on whether the underlying model is a classifier that estimates a conditional probability distribution $p(y|\xb)$ over some target variable given the input, or a model that learns a probability density $p(\xb)$ over the input itself.
-We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \cite{graham_denoising_2023,liu_unsupervised_2023a}, although it is important to note that this distinction relates only to the availability of some target value, $y$ - not whether OOD data is available for supervision\footnote{
+We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \cite{graham_denoising_2023,liu_unsupervised_2023a}.
+It is important to note that this distinction relates only to the availability of some target value, $y$ - not whether OOD data is available for supervision\footnote{
     In our overview, we distinguish between supervised and unsupervised out-of-distribution by letting methods be classified as supervised that use any kind of target value $y$, whether it relates to the original task, available OOD data, or both.
@@
 from a \emph{reconstruction} of the input $\xb$ \cite{sakurada_anomaly_2014,xia_learning_2015,lyudchik_outlier_2016,zhou_anomaly_2017,chen_outlier_2017,schlegl_unsupervised_2017,zong_deep_2018,li_madgan_2019,graham_denoising_2023,liu_unsupervised_2023a}, or
-from a latent \emph{representation} of the input \cite{denouden_improving_2018,hendrycks_using_2019,ahmadian_likelihoodfree_2021,bergman_classificationbased_2020,tack_csi_2020,sehwag_ssd_2021,xiao_we_2021}.
+from a hidden \emph{representation} of the input \cite{denouden_improving_2018,hendrycks_using_2019,ahmadian_likelihoodfree_2021,bergman_classificationbased_2020,tack_csi_2020,sehwag_ssd_2021,xiao_we_2021}.
 % https://arxiv.org/pdf/2302.10326.pdf
@@
 % Summary and weaknesses of supervised OOD detection approaches
-% Methods that use real OOD data achieve high performance but are limited by the availability of OOD data and how well it covers the large variation in OOD data.
-% A general weakness of all supervised out-of-distribution detection is that in learning the task-specific model $p(y|\xb)$ a model may discard information about $p(\xb)$ which could be useful for out-of-distribution detection.
+A general weakness of all supervised out-of-distribution detection is that in learning the task-specific model $p(y|\xb)$ a model may discard information about $p(\xb)$ which could be useful for out-of-distribution detection.
+Methods that use real OOD data achieve high performance but are limited by the availability of OOD data and how well it covers the large variation in OOD data.
 
@@
 
-% Assume we have a dataset $\mathcal{D}_{\text{training}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from $\mathcal{D}_{\text{training}}$ or not.
-Given a test point $\xb$, the model outputs a score $s(\xb)$ that indicates whether $\xb$ is in-distribution or out-of-distribution.
-
-For high-dimensional data, the most successful unsupervised models for OOD detection are deep autoencoders \cite{hinton_reducing_2006} and deep generative models. 
-Deep generative models are a class of models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution. 
-There exist at least six main classes of deep generative models: variational autoencoders \cite{kingma_autoencoding_2014,rezende_stochastic_2014}, flow-based models \cite{dinh_nice_2015,kingma_glow_2018}, autoregressive models \cite{oord_conditional_2016,oord_wavenet_2016}, diffusion models \cite{sohl-dickstein_deep_2015,ho_denoising_2020}, energy-based models \cite{lecun_tutorial_2006,salakhutdinov_efficient_2010,grathwohl_your_2020}, and generative adversarial networks \cite{goodfellow_generative_2014,arjovsky_wasserstein_2017}.
-Besides data generation, common properties of deep generative models include the ability to compute an exact or approximate likelihood of input $p(\xb)$, reconstruct an input $\xb$ from its latent representation $\zb$, and represent an input $\xb$ in a latent space.
-These properties enable various methods for their use in OOD detection.
-
+For high-dimensional data, the most successful unsupervised models for OOD detection are deep autoencoders \cite{hinton_reducing_2006}, self-supervised methods \cite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020}, and deep generative models \cite{kingma_autoencoding_2014,rezende_stochastic_2014,dinh_nice_2015,rezende_variational_2015,ho_denoising_2020,hinton_fast_2006,oord_conditional_2016,goodfellow_generative_2014}.
 Other important density estimation methods that have been applied to OOD detection include kernel density estimation \cite{parzen_estimation_1962}, nearest neighbor methods \cite{cover_nearest_1967}, support vector machines \cite{cortes_supportvector_1995,scholkopf_estimating_2001}, and Gaussian mixture models \cite{dempster_maximum_1977}.
-However, compared to deep generative models, these methods are not well suited for high-dimensional data such as images, text or audio, and have had little direct impact on the recent work on OOD detection.
+However, these methods are not well suited for high-dimensional data such as images, text or audio, and have had little direct impact on the recent work on OOD detection.
 
@@
 %
-Among the deep generative models that compute exact likelihoods are flow-based models, autoregressive models, and diffusion models, whereas variational autoencoders compute a lower bound on the likelihood, energy-based models usually require approximate methods, and generative adversarial networks do not compute an explicit likelihood at all.
+% Among the deep generative models that compute exact likelihoods are flow-based models, autoregressive models, and diffusion models, whereas variational autoencoders compute a lower bound on the likelihood, energy-based models usually require approximate methods, and generative adversarial networks do not compute an explicit likelihood at all.
 \textcite{bishop_novelty_1994} first proposed to use the likelihood assigned to data by a generative model as a measure for detecting anomalous data.
@@
 % \cite{
-%     denouden_improving_2018,              leverages the latent variables of an autoencoder, and evaluates the Mahalanobis distance in the latent space along with the data reconstruction error
-%     ahmadian_likelihoodfree_2021,         extracts low-level features from the encoder of an invertible generative model.
-%     hendrycks_using_2019,                 learn a representation over the in-domain data through self-supervised training
-%     bergman_classificationbased_2020,     learn a representation over the in-domain data through self-supervised training
-%     tack_csi_2020,                        learn a representation over the in-domain data through self-supervised training
-%     sehwag_ssd_2021,                      learn a representation over the in-domain data through self-supervised training
+%     X denouden_improving_2018,              leverages the latent variables of an autoencoder, and evaluates the Mahalanobis distance in the latent space along with the data reconstruction error
+%     X ahmadian_likelihoodfree_2021,         extracts low-level features from the encoder of an invertible generative model.
+%     X hendrycks_using_2019,                 learn a representation over the in-domain data through self-supervised training
+%     X bergman_classificationbased_2020,     learn a representation over the in-domain data through self-supervised training
+%     X tack_csi_2020,                        learn a representation over the in-domain data through self-supervised training
+%     X sehwag_ssd_2021,                      learn a representation over the in-domain data through self-supervised training
 %     X xiao_we_2021                          further shows that one can instead use a strong pretrained feature extractor (self-supervised)  while maintaining comparable performance.
@@
 %
+\textcite{denouden_improving_2018} suggest that reconstruction-based approaches fail to capture particular anomalies that lie far from known inlier samples in latent space but near the latent dimension manifold defined by the parameters of the model.
+They propose to measure the Mahalanobis distance between the global Gaussian distribution of training set in latent space and an encoded test input.
+\textcite{xiao_we_2021} instead propose to use an existing, strong foundation model, pre-trained with a self-supervised objective, to extract features from the input, and then fit a Gaussian Mixture Model to the features using the minimal Mahalanobis distance to the mixture components as the OOD score.
 
-\texticite{xiao_we_2021} propose to use an existing, strong foundation model, pre-trained with a self-supervised objective, to extract features from the input, fit a parametric density estimator to the features and use the Mahalanobis distance as the OOD score.
+Several other works also use self-supervised representations of the in-domain data for OOD detection \cite{hendrycks_using_2019,bergman_classificationbased_2020}. 
+\textcite{tack_csi_2020} propose to use a contrastive objective to learn representations of the in-domain data contrasted with data augmented in-samples and use a softmax classifier trained on the representations to compute the OOD score.
+\textcite{sehwag_ssd_2021} present a similar approach but use the feature space Mahalanobis distance similar to \textcite{denouden_improving_2018}.
+\textcite{ahmadian_likelihoodfree_2021} propose to use the latent representation of an invertible generative model to compute the OOD score.
 
@@
 
+Deep generative models are a class of models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution. 
+There exist at least six main classes of deep generative models: variational autoencoders \cite{kingma_autoencoding_2014,rezende_stochastic_2014,ranganath_hierarchical_2016,vahdat_nvae_2020,child_very_2021}, normalizing flow models \cite{dinh_nice_2015,rezende_variational_2015,dinh_density_2017,kingma_glow_2018,grathwohl_ffjord_2018}, diffusion models \cite{sohl-dickstein_deep_2015,song_generative_2019, ho_denoising_2020, vahdat_scorebased_2021}, energy-based models \cite{lecun_tutorial_2006,hinton_fast_2006,salakhutdinov_efficient_2010,du_implicit_2019}, autoregressive models \cite{oord_conditional_2016,oord_wavenet_2016,radford_improving_2018}, and generative adversarial networks \cite{goodfellow_generative_2014,arjovsky_wasserstein_2017,brock_large_2019,karras_stylebased_2019}.
+Besides data generation, common properties of deep generative models include the ability to compute an exact or approximate likelihood of input $p(\xb)$, reconstruct an input $\xb$ from its latent representation $\zb$, and represent an input $\xb$ in a latent space.
+These properties enable various methods for their use in OOD detection.
+
 VAE \cite{kingma_autoencoding_2014,rezende_stochastic_2014}

### 2023-09-22 chapters/introduction.tex
Update

@@
 In most cases, such methods require multiple forward passes through the model to estimate the predictive uncertainty making them computationally expensive and impractical for real-time applications.
-Within the field of out-of-distribution detection, the epistemic uncertainty is not always expressed in model parameters, but sometimes by a distribution over data representations, or output probabilities with special classes for out-of-distribution data.
-In such cases, the uncertainty can be estimated in a single forward pass through the model.
+Within the field of out-of-distribution detection, the epistemic uncertainty is generally not expressed in model parameters, but e.g.
+by a distribution over data representations, output probabilities with special classes for out-of-distribution data, or by distances in feature spaces.
+In such cases, the uncertainty can often be estimated in a single forward pass through the model.
 

### 2023-09-23 appendices/paper_brief.tex
Add brief overview paper

@@
 \label{app:paper-brief}
-\ifthenelse{\equal{\skipappendices}{true}}{}{
+% \ifthenelse{\equal{\skipappendices}{true}}{}{
 
-\todo[inline]{Include brief overview of unsupervised neural speech representation learning in appendix.}
+% \newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
 
+% \newcommand{\norm}[1]{\left\lVert#1\right\rVert}
+
+\begin{abstract}
+Unsupervised representation learning for speech processing has matured greatly in the last few years.
+Work in computer vision and natural language processing has paved the way, but speech data offers unique challenges.
+As a result, methods from other domains rarely translate directly.
+We review the development of unsupervised representation learning for speech over the last decade.
+We identify two primary model categories: self-supervised methods and probabilistic latent variable models.
+We describe the models and develop a comprehensive taxonomy.
+Finally, we discuss and compare models from the two categories.
+\end{abstract}
+
+\section{Introduction}
+Representation learning has shaped modern computer vision \cite{simonyan2014very} and natural language processing \cite{devlin2019bert}, and more recently speech processing has been subject to the same development \cite{baevski2020wav2vec}.
+Representation learning has been defined as \emph{``learning representations of the data that make it easier to extract useful information when building classifiers or other predictors"} \cite{bengio_representation_2013}.
+Unsupervised representation learning is concerned with learning useful representations without the use of human annotations.
+Usually, a model is first pre-trained on a task where plenty of data is available.
+The model is then fine-tuned, or used to extract input representations for a smaller model, targeting a task with limited training data.
+In computer vision, both supervised \cite{simonyan2014very, szegedy2015going, he2016deep} and unsupervised \cite{pathak2016context, doersch2015unsupervised} representation learning have gained attention with supervised representation learning driven by the availability of large annotated datasets \cite{deng2009imagenet}.
+For text and speech, pre-training is usually unsupervised as labeled data is difficult to obtain.
+Although work on text has paved the way, and the two fields share many characteristics, learning representations from speech is a problem faced with a unique set of challenges.
+
+In this paper, we survey work on unsupervised representation learning for speech processing from within the last decade.
+From a methodological perspective, we identify two primary model categories, namely models based on self-supervised learning and probabilistic latent variable models. 
+We provide a methodological review of the design choices related to each of the model categories and develop a model taxonomy that highlights the different directions of work.
+Finally, we compare and discuss models from the two categories and their respective evaluation procedures.
+%We then provide a comparison of the methods based on their respective evaluation procedures.
+%Finally, based on the model taxonomy, we discuss the differences between models from the two categories and identify future challenges and directions for research in speech representation learning.
+
+\begin{figure}[!t]
+    \begin{flushleft}
+        \textsc{\small \hspace{0.16\textwidth} self-supervised \hspace{0.22\textwidth} latent variable}
+    \end{flushleft}
+    \centering
+    \includegraphics[width=0.90\textwidth]{paper_brief/SSL_LVM.pdf}
+    \caption{A schematic overview of the two groups of models covered in this survey.
+\textit{Left}: A model trained with self-supervised learning.
+We take these models to consist of two functions $f(\cdot)$ and $g(\cdot)$ (sec.~\ref{ssec:notation}).
+After pre-training, $f(\cdot)$ is fine-tuned or used for extracting features $\mathbf{c}$.
+$g(\cdot)$ is an auxiliary function used to accommodate the self-supervised pre-training task.
+\textit{Right}: A probabilistic latent variable model.
+In contrast to the self-supervised model, the functions $f(\cdot)$ and $g(\cdot)$ learn the parameters of distributions $q$ and $p$.
+The latent variable $\mathbf{z}$ is commonly used for representation learning.}
+    \label{fig:ssl_lvm}
+\end{figure}
+
+\section{Unsupervised representation learning}
+
+In the following, we group previous work into \textit{self-supervised models} and \textit{probabilistic latent variable models}, and take a \emph{model} to comprise a neural architecture and a corresponding learning algorithm.
+A schematic overview is found in figure \ref{fig:ssl_lvm}.
+These categories are neither exhaustive nor mutually exclusive, but allow us to focus on the characteristics that have shaped different branches of research.
+
+With emphasis on recent successes in the field, we cover literature from the last 10 years.
+While a complete description of all relevant models is not within the scope of this work, we sketch important technicalities when they are particularly descriptive of certain models.
+We first define our high-level notation and conventions to ease discussion.
+
+%Similarly, $\textbf{a}_{<t}$ denotes a vector-sequence containing all elements prior to step $t$ and $\textbf{a}_{t \in \mathbb{A}}$ is a vector sequence containing elements with index contained in the set $\mathbb{A}$.
+%(or a scalar sequence $a_{i:j}$)
+
+%Input to a model is either a spectrogram $\mathbf{x}_{1:T}$ or the raw speech signal $x_{1:L}$ where $L > T$. Commonly $L = T(\frac{S}{100})$ where $S$ is the sample rate of the raw speech signal.
+
+% \IEEEpubidadjcol % related to the IEEEpubid command in the top
+
+\paragraph{Notation} \label{ssec:notation} We use the subscript $i{:}j$ with $i{\leq}j$ to denote a vector sequence $\textbf{a}_{i:j}$ containing elements $\textbf{a}_{i}$ through $\textbf{a}_{j}$.
+We denote model input as $\mathbf{x}_{1:T}$ which, in practice, might be either a spectrogram or the raw speech signal, but we do not distinguish between the two in notation as it is not essential to understand the models.
+Also, models commonly downsample the temporal dimension, but again, this is not crucial to understand the models, so we maintain a notation based on a single temporal dimension $t\in\{1,\dots, T\}$.
+
+When discussing self-supervised models, we use $\mathbf{c}_{1:T}$ to denote a contextualized representation.
+For stochastic latent variable models, we use $\mathbf{z}_{1:T}$ as is customary to the field.
+% The representation at a single time-step, $\mathbf{c}_t$ or $\mathbf{z}_t$, is typically a function of a larger part of the input space than just $\mathbf{x}_t$, in some cases the entire input sequence $\mathbf{x}_{1:T}$.
+While some models are frozen and produce representations used as input for downstream tasks (\textbf{\textsc{frz}}, table \ref{tab:model-taxonomy}), others are designed to be fine-tuned (\textbf{\textsc{ftn}}, table \ref{tab:model-taxonomy}).
+In either case, we use $f(\cdot)$ to denote the model that is used for the downstream task.
+We use $g(\cdot)$ to denote any auxiliary model components (e.g., for a reconstruction task we might have $g: \mathbf{c}_t \mapsto \hat{\mathbf{x}}_t$).
+When a model can be naturally subdivided into multiple components, we simply use $f_*(\cdot)$ where $*$ may be any convenient descriptor.
+Finally, we often use a subscript when defining a loss, $\mathcal{L}_i$, to imply that the total loss is computed as a sum over $i$.
+
+\subsection{Self-supervised models}
+\label{sec:ssmodels}
+
+Self-supervised learning is a subset of unsupervised learning \cite{tsai2020self}.
+Where other unsupervised methods can be seen as a means to an end in itself (e.g., clustering or data generation), self-supervised learning takes the form of a pretext task that only adds value when associated with a downstream task.
+This makes self-supervised learning tie naturally with semi-supervised learning, but it may also be part of a fully unsupervised setup \cite{baevski2021unsupervised}.
+Self-supervised learning is often characterized by automatically deriving the target from the input or other unlabeled examples \cite{ouali2020overview}.
+
+\paragraph{Predictive models} Similar to classic autoregressive language models \cite{mikolov2010recurrent}, contextualized speech representations can be learned by predicting future values of a simple representation \cite{oord2018representation, chung2019unsupervised, schneider2019wav2vec, chung2020generative, jiang2021further} (\textbf{\textsc{prd}}, table \ref{tab:model-taxonomy}).
+Modeling spectrograms directly, autoregressive predictive coding (APC, \citealp{chung2019unsupervised}) is perhaps the simplest example in this category.
+The forward pass and loss are computed as
+\begin{align}
+    \mathbf{c}_{t} &= f(\mathbf{x}_{1:t})  \label{eq:apc_f} \\ 
+    \hat{\mathbf{x}}_{t+k} &= g(\mathbf{c}_{t}) \\
+    %\mathcal{L}_t(\mathbf{\hat{x}}_{t+k}, \mathbf{x}_{t+k}) &= |\mathbf{\hat{x}}_{t+k} - \mathbf{x}_{t+k}|
+    \mathcal{L}_t &= \lVert \hat{\mathbf{x}}_{t+k} - \mathbf{x}_{t+k} \rVert_1\enspace.
+\end{align}
+%
+\noindent Here, $f(\cdot)$ and $g(\cdot)$ are parameterized by neural networks such that each $\mathbf{c}_t$ is only conditioned on previous inputs $\mathbf{x}_{1:t}$ and $\hat{\mathbf{x}}_{t+k}$ is computed step-wise.
+\citet{chung2019unsupervised} use a stack of unidirectional LSTMs for $f(\cdot)$ and a linear regression layer for $g(\cdot)$.
+Tasks that seek to predict or reconstruct the input are very common.
+In the literature, these are often jointly referred to as reconstruction tasks (\textbf{\textsc{rec}}, table \ref{tab:model-taxonomy}) \cite{liu2021tera, wang2021unispeech}, although this is somewhat misleading in the case of prediction.
+
+Contrary to generative models, such as WaveNet \cite{oord_wavenet:_2016}, the APC model is not restricted to next-step prediction.
+Instead, it predicts $k > 0$ steps ahead in order to ensure that the model does not learn a trivial solution by exploiting the smoothness of the signal.
+Depending on the downstream task, we are often interested in learning so-called slow features that will typically span multiple input frames \cite{wiskott2002slow}.
+Even the smallest linguistic units of speech, phonemes, tend to span $0.1$ seconds on average \cite{garofolo_timit_1993}, whereas spectrogram frames $\mathbf{x}_t$ are typically computed at $0.01$ second intervals.
+However, sometimes local smoothness is explicitly used to define the task \cite{badino2014auto, jati2017speaker2vec, jati2019neural}.
+
+%In early work, \citet{badino2014auto} propose a step-wise bottleneck model that learns to predict $\mathbf{x}_{t+1}$ from $\mathbf{x}_{t}$.
+Since phonetic information changes slowly, $\mathbf{x}_{t+1}$ is argued to be a noisy version of $\mathbf{x}_{t}$ encoding almost the same phonetic content.
+Thus, the model is akin to a denoising autoencoder.
+%A similar idea has been proposed by Jati and Georgiou \cite{jati2019neural} to learn speaker embeddings.
+
+% Next-step prediction has also been frequently used for generative models (\textbf{\textsc{gen}}, table \ref{tab:model-taxonomy}).
+The popular speech synthesis model WaveNet is perhaps the most obvious example \cite{oord_wavenet:_2016}.
+While this work does not consider representation learning, the model might learn useful features similar to those learned with APC.
+
+\paragraph{Contrastive models} Speech contains localized noise (e.g., phase shifts) that does not inform slow feature learning.
+Thus, directly modeling speech might not be the best way to learn contextualized representations.
+Contrastive predictive coding (CPC,  \citealp{oord2018representation}) targets a local variable $\mathbf{v}_{1:T}$, learned from the model input $\mathbf{x}_{1:T}$, instead of the input itself.
+The forward pass is
+%
+\begin{align}
+    \mathbf{v}_{t} &= f_{v}(\mathbf{x}_{t-r:t+r}) \label{eq: cpc local representation} \\
+    \mathbf{c}_{t} &= f_{c}(\mathbf{v}_{1:t}) \\
+    \hat{\mathbf{v}}_{t,k} &= g_k(\mathbf{c}_{t}) \enspace,
+    %g_k(\mathbf{c}_{t}, \mathbf{v}_{t+k}) &= \mathbf{v}_{t+k}^{\intercal} \mathbf{W}_k \mathbf{c}_{t}
+    %\mathcal{L}_{t,k} &= - \log \left(\frac{g_k(\mathbf{c}_{t}, \mathbf{v}_{t+k})}{\sum_{n \sim \mathcal{D}} g_k(\mathbf{c}_{t}, \mathbf{v}_{n})} \right)
+    % \mathcal{L}_{t,k} &= -\log g_k + \log \sum_{n \sim \mathcal{D}} g_k
+\end{align}
+%
+\noindent where $f_v(\cdot)$ is a convolutional neural network, such that each $\mathbf{v}_{t}$ only encodes information from a limited receptive field $2r+1$.
+Again, $f_c(\cdot)$ should be limited to condition each $\mathbf{c}_t$ on previous time-steps $\mathbf{v}_{1:t}$ and $g_k(\cdot)$ is a step-wise transformation.
+The loss is based on noise constrastive estimation \cite{gutmann2010noise} and is given by
+\begin{align}
+    \mathcal{L}_{t,k} &= - \log \left(\frac{\exp(\hat{\mathbf{v}}_{t,k}^{\text{\tiny T}}\mathbf{v}_{t+k})}{\sum_{n \sim \mathcal{D}} \exp(\hat{\mathbf{v}}_{t,k}^{\text{\tiny T}}\mathbf{v}_{n})} \right)\enspace .
+    % \mathcal{L}_{t,k} &= - \hat{\mathbf{v}}_{t,k}^{\text{\tiny T}}\mathbf{v}_{t+k} + \log \sum_{n \sim \mathcal{D}} \exp(\hat{\mathbf{v}}_{t,k}^{\text{\tiny T}}\mathbf{v}_{n})\enspace .
+    % \mathcal{L}_{t,k} &= -\log g_k + \log \sum_{n \sim \mathcal{D}} g_k
+    \label{eq: cpc loss}
+\end{align}
+%
+\noindent Here, $\mathcal{D}$ is a set of indices including the target index $t+k$ and negative samples drawn from a proposal distribution, which is typically taken to be a uniform distribution over the set $\{1,\dots,T\}$.
+Note that the loss is also indexed by $k$ to show that CPC targets multiple offsets.
+The APC model is easily extended in a similar way \cite{chung2020improved}.
+
+Crucially, we cannot simply predict $\mathbf{v}_{t+k}$ from $\mathbf{c}_t$ with an $\ell_1$ loss.
+This would cause $f_v(\cdot)$ to collapse to a trivial solution, such as setting all $\mathbf{v}_t$ equal.
+With a contrastive loss on the other hand, setting all $\mathbf{v}_t$ equal would cause $\mathcal{L}_{k,t}$ to be constant at a value no better than a random baseline.
+
+A model closely related to the original CPC model is wav2vec \cite{schneider2019wav2vec}.
+It uses a different parameterization of the functions $f_v(\cdot)$ and $f_c(\cdot)$, and modifies the loss to consider a binary prediction task, such that we have
+%
+\begin{align}
+    %g_k(\mathbf{c}_{t}, \mathbf{v}_{t+k}) &= \log(\sigma(\mathbf{v}_{t+k}^{\intercal} (\mathbf{W}_k \mathbf{c}_{t} + \mathbf{b}_k)))\\
+    \mathcal{L}_{t,k} &= - \log(\sigma(\hat{\mathbf{v}}_{t,k}^{\text{\tiny T}}\mathbf{v}_{t+k})) - \sum_{n \sim \mathcal{D}} \log(\sigma(-\hat{\mathbf{v}}_{t,k}^{\text{\tiny T}}\mathbf{v}_{n})) \,.
+    \label{eq: wav2vec loss}
+\end{align}
+
+\noindent This model was among the first to show that learned representations can be used to improve end-to-end speech recognition.
+As we will see, the wav2vec framework has evolved to shape state-of-the-art representation learning for speech.
+
+\paragraph{Masking-based models} One downside of predictive tasks is that models are primarily unidirectional.
+Some work has extended APC and CPC inspired models with separate encoders operating in opposite directions \cite{ling2020deep, kawakami2020learning, borgholt2021scaling}, but these models are still restricted to process left and right context separately.
+Inspired by the masked language model task used for text-based representation learning \cite{devlin2019bert}, several papers have used masking to overcome this challenge (\textbf{\textsc{msk}}, table \ref{tab:model-taxonomy}).
+Masking refers to replacing parts of the input with zeros or a learned masking vector.
+For zero-masking \cite{jiang2019improving, liu2020mockingjay, wang2020unsupervised, chi2021audio, ling2020decoar}, we have
+\begin{align}
+    \mathbf{c}_{1:T} &= f(\mathbf{x}_{1:T} \circ \mathbf{m}_{1:T})\\
+    \mathbf{\hat{x}}_{t} &= g(\mathbf{c}_{t}) \\
+    \mathcal{L}_t &= \lVert \mathbf{\hat{x}}_{t} - \mathbf{x}_{t} \rVert_1\enspace,
+\end{align}
+where the $\circ$ operator denotes the Hadamard product, $f(\cdot)$ is typically a transformer encoder or a bidirectional recurrent neural network, $g(\cdot)$ is a step-wise transformation, and $\mathbf{m}_{1:T}$ is a mask such that $m_{t,i} \in \{0,1\}$.
+Alternatively, $\mathbf{m}_{1:T}$ is used to select which $\mathbf{x}_t$ are replaced by a learned masking vector.
+The entries of $\mathbf{m}_{1:T}$ are determined by some stochastic policy.
+One frequent inspiration is SpecAugment \cite{park2019specaugment}, which was originally proposed for supervised speech recognition and applies frequency and time masking to spectrogram representations.
+While temporal masking is most common, frequency masking has also been adopted for representation learning  \cite{wang2020unsupervised}.
+A simple, yet popular, masking strategy is to draw a proportion of input indices $t_i\sim\{1,\dots,T-M\}$ without replacement, and then mask $\{t_i, \dots, t_i+M\}$ \cite{baevski2020wav2vec, hsu2021hubert, ling2020decoar}.
+
+Combining masking with a contrastive loss, wav2vec 2.0 was the first work to show that a competitive speech recognition model can be learned by fine-tuning a pre-trained model with as little as 10 minutes of labeled data.
+For this model
+%
+\begin{align}
+    \mathbf{v}_{t} &= f_v(\mathbf{x}_{t-r:t+r}) \\
+    \mathbf{c}_{1:T} &= f_c(\mathbf{v}_{1:T} \circ \mathbf{m}_{1:T}) \\
+    \mathbf{q}_t &= g_q(\mathbf{v}_t) \enspace . \label{w2v2 qtz}
+\end{align}
+
+\noindent Here, $f_v(\cdot)$ is a convolutional neural network, $f_c(\cdot)$ is a transformer encoder \cite{vaswani2017attention} and $g_q(\cdot)$ is a quantization module used to learn targets from the localized variable $\mathbf{v}_{1:T}$.
+Computing quantized targets this way requires an extra loss term, which we will present when we discuss quantization in general below.
+The contrastive loss for wav2vec 2.0 is similar to that of the CPC model,
+%
+\begin{align}
+    \mathcal{L}_t &= - \log \left(\frac{\exp(S_{\text{c}}(\mathbf{c}_{t}, \mathbf{q}_{t}))}{\sum_{n \sim \mathcal{D}} \exp(S_{\text{c}}(\mathbf{c}_{t}, \mathbf{q}_{n}))} \right) \enspace , \label{w2v2 loss}
+\end{align}
+%
+\noindent where $S_{\text{c}}(\cdot)$ is the cosine similarity and the negative samples in $\mathcal{D}$ are sampled from other masked time-steps.
+
+In general, masking is less data efficient than prediction, as only the masked portion of the input is non-trivial to reconstruct. For this reason, the loss might be computed as
+%
+\begin{align}
+    \mathcal{L}_t &= \lVert (\mathbf{\hat{x}}_{t} - \mathbf{x}_{t}) \circ (\mathbf{1} - \mathbf{m}_t) \rVert_1 \enspace .
+\end{align}
+
+\noindent Non-autoregressive predictive coding (NPC, \citealp{liu2020non}) tries to resolve this by using a convolutional neural network where the kernel is masked instead of the input.
+%That is, the model is prevented from using $\mathbf{x}_{t}$ for its corresponding reconstruction $\mathbf{\hat{x}}_{t}$.
+This allows for complete data utilization, but limits the amount of context encoded in the learned representation.
+Figure \ref{fig:ssl_grid} summarizes the models discussed so far.
+
+\begin{figure}[!t]
+    \centering
+    \setlength\tabcolsep{1.5pt}
+    \begin{tabular}{>{\centering\arraybackslash} m{4mm}  >{\centering\arraybackslash} m{0.45\textwidth}|>{\centering\arraybackslash} m{0.45\textwidth}}
+          & {\small \textsc{predictive}} & {\small \textsc{masking-based}} \\
+        %\midrule
+        \rotatebox{90}{{\small \textsc{reconstruct}}} & \includegraphics[width=0.45\textwidth]{paper_brief/REC_PRD.pdf} & \includegraphics[width=0.45\textwidth]{paper_brief/REC_MSK.pdf}  \\
+        \midrule
+        \rotatebox{90}{{\small \textsc{contrastive}}} & \includegraphics[width=0.45\textwidth]{paper_brief/CON_PRD.pdf} & \includegraphics[width=0.45\textwidth]{paper_brief/CON_MSK.pdf}
+    \end{tabular}
+    \caption{
+    Schematic of self-supervised methods. Each subfigure illustrates the loss computation for a single time-step.
+    The temporal subscript has been left out for simplicity. 
+    %Examples of models in each category include APC (\textbf{\textsc{prd}} + \textbf{\textsc{rec}}), Mockingjay (\textbf{\textsc{msk}} + \textbf{\textsc{rec}}), CPC (\textbf{\textsc{prd}} + \textbf{\textsc{con}}) and wav2vec 2.0 (\textbf{\textsc{msk}} + \textbf{\textsc{con}}).
     }
+    \label{fig:ssl_grid}
+\end{figure}
+
+\begin{sidewaystable*}[t]
+    \caption{Selected models classified according to the binary attributes identified throughout the text.
+The models are sorted according to first publication date on arXiv which might differ from the citation year.
+\textbf{\textsc{msk}}: masking, \textbf{\textsc{prd}}: prediction, \textbf{\textsc{con}}: contrastive, \textbf{\textsc{rec}}: reconstruction, \textbf{\textsc{qtz}}: quantization, \textbf{\textsc{gen}}: generative, \textbf{\textsc{frz}}: frozen, \textbf{\textsc{ftn}}: fine-tuned, \textbf{\textsc{loc}}: local, \textbf{\textsc{glo}}: global.}
+    \label{tab:model-taxonomy}
+    \begin{center}
+        \renewcommand{\arraystretch}{1.1}
+        \resizebox{0.925\textwidth}{!}{%
+        \begin{tabular}{ l l | c c c c c c | c c c | c c } 
+            \toprule
+
+            \multicolumn{2}{c}{} & 
+            \multicolumn{6}{c}{\footnotesize\textsc{Model and task design}} & 
+            \multicolumn{3}{c}{\footnotesize\textsc{Resolution}} &
+            \multicolumn{2}{c}{\footnotesize\makebox[0pt][c]{\textsc{Usage}}} \\
+            \textbf{\textsc{model}} &
+            \textbf{\textsc{pub. date}} &
+            \textbf{\textsc{msk}} & 
+            \textbf{\textsc{prd}} & 
+            \textbf{\textsc{con}} &  
+            \textbf{\textsc{rec}} &  
+            \textbf{\textsc{qtz}} & 
+            \textbf{\textsc{gen}} & 
+            \textbf{\textsc{loc}} & 
+            \textbf{\textsc{glb}} & 
+            \textbf{\textsc{var}} & 
+            \textbf{\textsc{frz}} & 
+            \textbf{\textsc{ftn}} \\
+
+            \midrule
+            \multicolumn{12}{c}{\textsc{Self-supervised models}} \\
+            \midrule
+
+            %                                                           MSK      PRD      CON      REC      QTZ      GEN      LOC      GLB     VAR      FRZ      FTN
+            \textbf{Audio Word2vec} \footnotesize\cite{chung2016audio}      & 2016 Mar. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{Speech2Vec} \footnotesize\cite{chung2018speech2vec}     & 2018 Mar. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{Unspeech} \footnotesize\cite{milde2018unspeech}         & 2018 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+
+            \textbf{CPC} (van den Oord et al. 2018)         & 2018 Jul. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+
+            %PASE \footnotesize\cite{Pascual2019}                   & 2019 Apr. & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 6/4
+            \textbf{APC} \footnotesize\cite{chung2019unsupervised}          & 2019 Oct. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 5/4
+            \textbf{wav2vec} \footnotesize\cite{schneider2019wav2vec}       & 2019 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/4
+
+            %VQ-APC \footnotesize\cite{chung2020vector}             & 2020 May  & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+
+            \textbf{Mockingjay}  \footnotesize\cite{liu2020mockingjay}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 25/10-19
+            %DeCoAR \footnotesize\cite{ling2020deep}                & 2019 Dec. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 3/12-19
+            %PASE+ \footnotesize\cite{Ravanelli2020}                & 2020 Jan. & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 25/1
+            \textbf{wav2vec 2.0} \footnotesize\cite{baevski2020wav2vec}     & 2020 Jun. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 20/6
+            %TERA \footnotesize\cite{liu2021tera}                   & 2021 Jul. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 12/7
+            \textbf{NPC} \footnotesize\cite{liu2020non}                     & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 1/11
+            \textbf{DeCoAR 2.0} \footnotesize\cite{ling2020decoar}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/12
+            \textbf{SCPC} \footnotesize\cite{bhati2021segmental}            & 2021 Jun. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ % 3/6
+            \textbf{HuBERT} \footnotesize\cite{hsu2021hubert}               & 2021 Jun. & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 14/6
+
+            \midrule
+            \multicolumn{12}{c}{\textsc{Probabilistic latent variable models}} \\
+            \midrule
+            %                                                           MSK      PRD      CON      REC      QTZ      GEN      LOC      GLO     VAR      FRZ      FTN
+            % Conv. DBN \footnotesize\cite{lee_unsupervised_2009}    & 2009 Jan. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark \\
+            \textbf{VRNN} \footnotesize\cite{chung_recurrent_2015}          & 2015 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{SRNN} \footnotesize\cite{fraccaro_sequential_2016}      & 2016 May  & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{HMM-VAE} \footnotesize\cite{ebbers_hidden_2017}         & 2017 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{ConvVAE} \footnotesize\cite{hsu_learning_2017}          & 2017 Apr. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{FHVAE} \footnotesize\cite{hsu_unsupervised_2017}        & 2017 Sep. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{VQ-VAE} (van den Oord et al. 2018)                      & 2017 Nov. & \xmark & \xmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{BHMM-VAE} \footnotesize\cite{glarner_full_2018}         & 2018 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{STCN} \footnotesize\cite{aksan_stcn_2019}               & 2019 Feb. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{FDMM} \footnotesize\cite{khurana_factorial_2019}        & 2019 Oct. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{ConvDMM} \footnotesize\cite{khurana_convolutional_2020} & 2020 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \bottomrule
+        \end{tabular}
+        }
+    \end{center}
+\end{sidewaystable*}
+
+\paragraph{Quantization} 
+Several models enforce a discrete latent space by quantizing the vector representation (\textbf{\textsc{qtz}}, table \ref{tab:model-taxonomy}).
+The two most popular approaches are the Gumbel-softmax \cite{jang2016categorical, maddison2016concrete} and the quantization used in the VQ-VAE \cite{oord_neural_2018}.
+
+%Several models enforce a discrete latent space by quantizing the vector representation (\textbf{\textsc{qtz}}, table \ref{tab:model-taxonomy}).
+A popular approach for online vector quantization is to use the Gumbel softmax \cite{jang2016categorical, maddison2016concrete}.
+This approach corresponds to approximate differentiable sampling from a categorical distribution.
+Another common approach employs a codebook of learnable representations.
+This approach was popularized with the VQ-VAE \cite{oord_neural_2018}, and thus, we refer to it as the VQ-VAE approach.
+Both approaches are non-differentiable and obtain gradients using the straight-through estimator \cite{bengio2013estimating}.
+
+\textit{Gumbel-softmax approach:} 
+Say we want to quantize a vector $\mathbf{v}$ such that it takes one of $K$ possible values.
+We first map $\mathbf{v}$ to $\mathbf{l} \in \mathbb{R}^K$ and then map $\mathbf{l}$ to a probability vector $\mathbf{p} \in \mathbb{R}^K$ via the Gumbel softmax given by
+\begin{align}
+    p_i &= \frac{\exp(l_i + n_i) / \tau}{\sum_k^K \exp(l_k + n_k) / \tau} 
+\end{align}
+% for $i=1,\dots,K$. Here $\tau$ is a temperature parameter and $\mathbf{n}\in\mathbb R^K$ is a random vector with $n_i = -\log(-\log(u_i))$ for $u_i \sim \mathcal{U}(0, 1)$. 
+% As  $\tau \rightarrow 0$, $\mathbf{p}$ approaches a one-hot vector. The Gumbel noise $\mathbf{n}$ is a practical way to sample from the untempered categorical distribution (i.e., $\tau = 1$). 
+% A function $\varphi(\mathbf{p})$ is then used to map $\mathbf{p}$ to a discrete sample which can then be used to obtain the quantized vector (e.g., $\mathbf{q}_t$ in eq.~\ref{w2v2 qtz}) via a codebook lookup.
+As this function is non-differentiable, we have to rely on the straight-through estimator \cite{bengio2013estimating} which assumes that the Jacobian ${\partial \, \varphi}/{\partial \, \mathbf{p}}$ equals the identity matrix.
+for $i=1,\dots,K$. Here $\tau$ is a temperature parameter and $\mathbf{n}\in\mathbb R^K$ is a random vector with $n_i = -\log(-\log(u_i))$ for $u_i \sim \mathcal{U}(0, 1)$.
+As  $\tau \rightarrow 0$, $\mathbf{p}$ approaches a one-hot vector. The Gumbel noise $\mathbf{n}$ is a practical way to sample from the untempered categorical distribution (i.e., $\tau = 1$).
+$\mathbf{p}$ is mapped to a one-hot vector using a function $\varphi(\cdot)$, such that $\varphi(\mathbf{p})_i = 1$ if $i = \argmax_j p_j$ and $0$ otherwise. 
+As this function is non-differentiable, we must rely on the straight-through gradient estimator \cite{bengio2013estimating} which assumes that the Jacobian ${\partial \, \varphi}/{\partial \, \mathbf{p}}$ equals the identity matrix.
+The one-hot vector can then be used for a codebook lookup to obtain the final quantized vector (e.g., $\mathbf{q}_t$ in eq.~\ref{w2v2 qtz}).
+
+%We choose $\varphi(\cdot)$ to return a one-hot vector with $\varphi(\mathbf{p})_i = 1$ if $i = \argmax_j p_j$ and $0$ otherwise.
+The one-hot vector can then be used for a codebook lookup.
+For non-differentiable $\varphi(\cdot)$, the gradient is usually taken to be that of the input to $\varphi(\cdot)$.
+That is, we use the straight-through estimator and assume during training that the Jacobian ${\partial \, \varphi}/{\partial \, \mathbf{p}}$ equals the identity matrix.
+%\begin{align}
+%    %\mathbf{v} &= \mathbf{p} + \textsc{sg}(\varphi(\mathbf{p}) - %\mathbf{p}) \, .
+%    \frac{\partial \, \varphi}{\partial \, \mathbf{p}} = \mathbf{I}
+%\end{align}
+%where $\textsc{sg}(\cdot)$ is the \textit{stop gradient} function; an identity function that blocks the gradient from the input. .
+
+The wav2vec 2.0 quantization module (eq.~\ref{w2v2 qtz}) uses the Gumbel softmax. To ensure utilization of codebook vectors, a diversity loss is added to the task specific loss (eq. \ref{w2v2 loss})
+%
+\begin{align}
+    % \mathcal{L} =  - H(\widetilde{\mathbf{p}}) \frac{1}{K}\enspace,
+    \mathcal{L} =  - H(\widetilde{\mathbf{p}}) / K \enspace,
+\end{align}
+\noindent where $H(\cdot)$ is the entropy and $\widetilde{\mathbf{p}}$ is the untempered version of $\mathbf{p}$ without Gumbel noise.
+
+\textit{VQ-VAE approach:} Instead of directly parameterizing a probability distribution, as in the Gumbel softmax, a vector $\mathbf{v}$ can be quantized by replacing it with the closest codebook vector $\mathbf{e}_k$.
+Specifically, given a learned codebook $\mathbf{e}\in\mathbb{R}^{K\times D}$, where $K$ is the codebook size and $D$ is the dimensionality of each codebook vector $\mathbf{e}_k$, the quantized representation $\mathbf{q}$ of $\mathbf{v}$ is obtained as,
+\begin{align}
+    \mathbf{q} = \mathbf{e}_k \ , \enspace \text{where} \enspace k=\arg\min_j \norm{\mathbf{v} - \mathbf{e}_j}_2 \enspace .
+\end{align}
+As $\arg\min$ is non-differentiable, the straight-through estimator is used as for the Gumbel-softmax. 
+% The straight-through estimator defines the gradient of this operation by setting the gradient of $\mathbf{v}$ wrt. to the loss equal to that of $\mathbf{q}$. 
+%
+% Codebook learning is facilitated by a two-term auxiliary loss similar to classical vector quantization dictionary learning \cite{burton_generalization_1983, soong_vector_1985}.
+Gradients for the codebook vectors are given solely by a vector quantization term, which moves codebook vectors $\mathbf{e}_k$ closer to the non-quantized vectors $\mathbf{v}$.
+A so-called commitment term is added to ensure that non-quantized vectors do not grow unboundedly by enforcing the encoder to keep them close to a codebook vector.
+%
+Codebook learning is facilitated by a two-term auxiliary loss similar to classical vector quantization dictionary learning \cite{burton_generalization_1983, soong_vector_1985}.
+Gradients for the codebook vectors are given solely by a vector quantization term.
+A so-called commitment term ensures that non-quantized vectors do not grow unboundedly.
+\begin{align}
+    \mathcal{L} = \underset{\text{vq}}{\underbrace{\norm{\text{sg}\left[\mathbf{v}\right] - \mathbf{e}}_2^2}} + \underset{\text{commitment}}{\underbrace{\beta\norm{\mathbf{v} - \text{sg}\left[\mathbf{e}\right]}_2^2}} \enspace , \label{eq: vector quantization losses}
+\end{align}
+where $\text{sg}[\mathbf{x}] = \mathbf{x}$ is the stop-gradient operator with the property $\frac{d}{dx_i}\text{sg}[\mathbf{x}] \equiv 0$ for all $i$ and $\beta$ is a hyperparameter.
+Although vector quantization was introduced by the VQ-VAE which is, in some ways, a latent variable model, it has been applied to self-supervised methods \cite{niekerk_vector_2020, baevski2019vq}.
+
+%but with variations between individual approaches.
+The VQ-CPC model \cite{niekerk_vector_2020} quantizes $\mathbf{v}$ in the original CPC (eq.~\ref{eq: cpc local representation}) and augments its loss (eq.~\ref{eq: cpc loss}) by addition of the commitment term in eq.~\ref{eq: vector quantization losses}.
+Instead of learning the codebook via the VQ term, it is updated as a moving average of $\mathbf{v}$ as initially suggested by \citet{oord_neural_2018}.
+%The vq-wav2vec \cite{baevski2019vq} is defined similarly but adds the full auxiliary loss to the wav2vec loss (eq.~\ref{eq: wav2vec loss}) in place of moving averages.
+
+\textit{Motivation:} Similar to how quantization approaches differ between works, so do the motivations provided for employing them.
+The vq-wav2vec \cite{baevski2019vq, baevski2019effectiveness} %quantizes $\mathbf{v}_{1:T}$ before feeding them to the context network $f_c(\cdot)$.
+Here, the motivation is
+learn quantized representations in order to apply natural language processing models, like BERT \cite{devlin2019bert}, afterwards.
+Other works use quantization for speech segmentation \cite{kamper2020towards, chorowski2019unsupervised} or as a bottleneck in order to ``\textit{limit model capacity}" \cite{chung2020vector, ling2020decoar}.
+%Quantized representations have also been used for speech segmentation \cite{kamper2020towards, chorowski2019unsupervised}. 
+Finally, \citet{chung2020vector} explore quantization between different layers in the APC model, but find that continuous representations consistently perform better than their quantized counterparts on a downstream phoneme classification task
+
+%They find that applying quantization to the output $\mathbf{c}_{1:T}$ yields the best results on a downstream phoneme classification task.
+However, the continuous representations consistently perform better than their quantized counterparts.
+
+Given our previous discussion of how it might not be beneficial to model localized noise, quantization in wav2vec 2.0 seems well motivated, as it enforces the target representation $\mathbf{q}_{1:T}$ to discard such noise.
+Taking this idea further, the HuBERT model \cite{hsu2021hubert} uses offline quantization to learn categorical targets.
+Initially, spectrogram features are used to learn frame-wise labels with $k$-means clustering.
+A model similar to wav2vec 2.0, but without online quantization, is then trained to infer labels for masked time-steps.
+Since quantization is offline, this model does not need to rely on a contrastive loss, but can infer the target class directly.
+The offline quantization also ensures more stable training, as targets do not change abruptly.
+
+\paragraph{Global representations}
+The models covered so far learn representations that maintain a temporal resolution proportional to the input resolution.
+We say that they learn local representations (\textbf{\textsc{loc}}, table \ref{tab:model-taxonomy}).
+Now, we cover models that learn global representations (\textbf{\textsc{glb}}, table \ref{tab:model-taxonomy}).
+%Apart from this attribute, we find that models that learn global representations rely on many of the same techniques described in the previous sections.
+%We first give a broad overview, before discussing how they relate to the self-supervised models.
+
+Early work on global speech representation learning takes inspiration from the autoencoder framework \cite{kramer1991nonlinear}.
+\citet{chung2016audio} propose a simple sequence-to-sequence autoencoder for learning acoustic word embeddings:
+%
+% i/x-vectors here?
+\begin{align}
+    \mathbf{c} &= f(\mathbf{x}_{1:T}) \\
+    \hat{\mathbf{x}}_{1:T} &= g(\mathbf{c}) \enspace , \label{eq:dec-aw2v}
+\end{align}
+
+\noindent where $f(\cdot)$ and $g(\cdot)$ are a recurrent neural networks, such that $\mathbf{c}$ is taken to be the hidden state at the last time-step $T$ of $f(\cdot)$ and used as initial hidden state of $g(\cdot)$.
+%where $f(\cdot)$ is a recurrent neural network encoder, such that $\mathbf{c}$ is taken to be the hidden state at the last time-step $T$.
+The decoder, $g(\cdot)$, is also a recurrent neural network where $\mathbf{c}$ is used to initialize the hidden state.
+The authors also propose a denoising autoencoder with masked inputs $f(\mathbf{x}_{1:T} \circ \mathbf{m}_{1:T})$.
+Similar RNN-based autoencoders have also been explored \cite{kamper2019truly, holzenberger2018learning}.
+%
+%\begin{align}
+%    \mathbf{c} &= f(\mathbf{x}_{1:T} \circ \mathbf{m}_{1:T})
+%\end{align}
+%
+
+Prior to this work, \citet{kamper2015unsupervised} and \citet{renshaw2015comparison} introduced the \textit{correspondence autoencoder}.
+This method uses dynamic time warping to align input-target segment pairs extracted with unsupervised term discovery.
+In more recent work, the need for alignment has been alleviated by adopting the sequence-to-sequence framework \cite{kamper2019truly, jacobs2021acoustic}.
+
+%This method relies on unsupervised term discovery with randomized algorithms \cite{jansen2011efficient} to extract corresponding speech segment pairs for training.
+Thus, one segment is used as input and the other as reconstruction target.
+Because the paired segments do not necessarily have the same length, they need to be aligned with dynamic time warping.
+
+% In more recent work, this need has been alleviated by adopting the sequence-to-sequence framework \cite{kamper2019truly, jacobs2021acoustic}.
+%Unlike the reconstruction tasks reviewed in the previous sections, models presented here all rely on an $\ell_2$ loss, $\mathcal{L} = \lVert \hat{\mathbf{x}}_{1:T} - \mathbf{x}_{1:T} \rVert_2\enspace$.
+%
+% \begin{align}
+%     \mathcal{L} &= \lVert \hat{\mathbf{x}}_{1:T} - \mathbf{x}_{1:T} \rVert_2\enspace.
+% \end{align}
+
+Inspired by the work on semantic word embeddings for text \cite{mikolov2013distributed}, the sequence-to-sequence framework has also been used to implement speech-based versions of the skipgram and continuous bag-of-words models \cite{chung2017learning, chung2018speech2vec}.
+Given a segment corresponding to a single word $\mathbf{x}_{(n)} = \mathbf{x}_{t_{n}:t_{n+1}}$, the skipgram model is trained to predict neighboring words $\mathbf{x}_{(n + k)}$ where $k\ne0$.
+That is, instead of a single decoder, as in eq.
+\ref{eq:dec-aw2v}, the skipgram model employs multiple decoders
+%
+\begin{align}
+    \hat{\mathbf{x}}_{(n + k)} = g_k(\mathbf{c})\enspace.
+\end{align}
+%
+Conversely, the continuous bag-of-words model is trained to predict the target word from the neighboring words, so here multiple encoders sum over several offsets $\mathcal{K}$ to obtain $\mathbf{c}$:
+%
+\begin{align}
+    \mathbf{c} &= \sum_{k\in\mathcal{K}} f_k(\mathbf{x}_{(n + k)}) %\enspace .
+    % \mathbf{c} &= \sum_{k\in\mathcal K_n} \mathbf{c}_k
+\end{align}
+%for different values of $k$.
+The sequence-to-sequence models described above rely on speech segments corresponding to words.
+The segments are obtained by supervised forced alignment, %Thus, since forced alignment requires labeled data, the models might be seen as weakly supervised
+but similar models have been explored without this requirement \cite{jati2017speaker2vec, tagliasacchi2020pre}.
+
+Contrastive learning has also been explored for global speech representation learning \citep{milde2018unspeech, jati2019neural, jansen2018unsupervised}.
+And prior to the widespread adoption of neural networks, \citet{levin2013fixed} explore principal component analysis and Laplacian eigenmaps for learning fixed-sized acoustic embeddings.
+%Comparable to the skipgram model described above, \citet{milde2018unspeech} and \citet{jati2019neural} define a binary contrastive task where a target segment is used to predict neighboring segments.
+
+%For general audio representation learning, \citet{jansen2018unsupervised} explore multiple pre-training tasks using a contrastive triplet loss.
+And prior to the widespread adoption of neural networks, \citet{levin2013fixed} explore principal component analysis and Laplacian eigenmaps for learning fixed-sized acoustic embeddings.
+
+\paragraph{Other work} Some models learn local representations with a variable temporal resolution that is not proportional to the input resolution (\textbf{\textsc{var}}, table \ref{tab:model-taxonomy}).
+In practice, this is often achieved implicitly, by learning segment boundaries or by taking repeated quantized values to belong to the same segment \cite{kamper2020towards, chorowski2019unsupervised, michel2017blind, kreuk2020self, wang2017gate, dieleman_variable-rate_2021}.
+An exception is the recently proposed \emph{segmental contrastive predictive coding} (SCPC, \citealp{bhati2021segmental, bhati2021unsupervised}).
+With this approach, the model explicitly learns segment boundaries, which are used to downsample the representations during training.
+The same segmentation strategy has subsequently been applied in other models \cite{cuervo2021contrastive}.
+
+Most of the work presented so far fits neatly into the taxonomy presented in table \ref{tab:model-taxonomy}.
+One exception is the problem-agnostic speech encoder (PASE, \citealp{Pascual2019, Ravanelli2020}) that combines multiple pre-training tasks.
+Furthermore, many of the presented models have been successfully applied to other use cases.
+For instance, wav2vec 2.0 and related models have been applied to learn cross-lingual and multi-lingual representations \cite{riviere2020unsupervised, conneau2020unsupervised, khurana2021magic} and proven well-suited for concurrently learning with labeled data \cite{talnikar2021joint, wang2021unispeech}.
+%We will discuss the downstream application of self-supervised representations further in section \ref{sec:eval}.
+
+\subsection{Probabilistic latent variable models}
+\label{sec:plvms}
+Another prominent class of models are probabilistic latent variable models (LVMs). 
+Before surveying their application to speech, we briefly review LVMs and their usual specification when applied for representation learning in general.
+We disregard any specific temporal notation without loss of generality.
+We then introduce the variational autoencoder framework (VAE, \citealp{kingma_auto-encoding_2014}).
+We focus on different dependency structures between data and learned representations, in contrast to the more practical view on self-supervised models taken above.
+
+\paragraph{LVMs and inference}
+Fundamental to LVMs is the assumption that the data is produced by a generative process that involves unobserved stochastic latent variables $\textbf{z}$. 
+An LVM aims to model this generative process to enable generation of new data $\mathbf{x}$ (\textbf{\textsc{gen}}, table \ref{tab:model-taxonomy}) and inference of the latent variable associated with a given observed variable $\textbf{x}$.
+For representation learning, the inference of latent variables is of primary interest.
+An LVM is defined by the observation model $p(\mathbf{x}|\mathbf{z})$, which defines the relationship between the observed and latent variables, and the prior $p(\mathbf{z})$, which defines the relationship among the latent variables \cite{bartholomew_latent_2011}.
+An LVM models the generative process via the joint observation and prior model $p(\mathbf{x}, \mathbf{z})$ often referred to as the generative model. 
+The likelihood of an LVM given an example $\textbf{x}$ can be written as
+\begin{equation}
+    \log p(\mathbf{x}) = \log \int p(\mathbf{x}|\mathbf{z}) p(\mathbf{z}) \,\text{d}\mathbf{z} \enspace.
+    \label{eq: lvm log-likelihood}
+\end{equation}
+The latent variable can be inferred with e.g. Markov Chain Monte Carlo (MCMC) methods \cite{mohamed_monte_2019} or variational inference \cite{jordan_introduction_1999}.
+
+For representation learning, LVMs are commonly defined using the VAE framework \cite{kingma_auto-encoding_2014, rezende_stochastic_2014}) which is also the focus of our exposition. 
+In the VAE framework, the observation model $p(\mathbf{x}|\mathbf{z})$ is parameterized using a deep neural network.
+This choice allows modeling complex and high-dimensional data but also makes the integral in eq.~\ref{eq: lvm log-likelihood} analytically intractable.
+MCMC methods can be used to estimate it and the true model posterior $p(\mathbf{z}|\mathbf{x})$, but these methods are usually computationally expensive in this setting \cite{mohamed_monte_2019}.
+To counter this and make gradient-based maximum likelihood training feasible, the VAE instead employs variational inference \cite{jordan_introduction_1999}.
+It approximates the intractable true model posterior by introducing a variational posterior distribution $q(\mathbf{z}|\mathbf{x})$, also parameterized by a deep neural network.
+From eq.~\ref{eq: lvm log-likelihood}, via Jensen's inequality, this gives rise to a variational lower bound on the likelihood, also known as the evidence lower bound (ELBO).
+\begin{equation}
+    \log p(\mathbf{x}) \geq \int q(\mathbf{z}|\mathbf{x}) \log \frac{p(\mathbf{x}| \mathbf{z})p(\mathbf{z})}{q(\mathbf{z}|\mathbf{x})} \,\text{d}\mathbf{z} \equiv \mathcal{L}_{\text{ELBO}} \enspace .
+\label{eq: lvm likelihood bound (elbo)}
+\end{equation}
+The bound can be efficiently evaluated and optimized with Monte Carlo (MC) estimation by sampling from $q(\mathbf{z}|\mathbf{x})$.
+Low-variance gradient estimates are usually obtained via reparameterization of $q(\mathbf{z}|\mathbf{x})$ \cite{kingma_auto-encoding_2014} although alternatives exist (e.g., inverse CDF sampling) \cite{mohamed_monte_2019}.
+The ELBO can also be written as
+\begin{equation}
+    \mathcal{L}_{\text{ELBO}} = \mathbb{E}_{q(\mathbf{z}|\mathbf{x})}\left[ \log p(\mathbf{x}|\mathbf{z}) \right] - D_\text{KL}\left( q(\mathbf{z}|\mathbf{x}) || p(\mathbf{z}) \right) \enspace  \label{eq: lvm likelihood bound recon/kl form (elbo)},
+\end{equation}
+where $\mathbb{E}\left[ \log p(\mathbf{x}|\mathbf{z}) \right]$ can be seen as a reconstruction loss and $D_\text{KL}\left( q(\mathbf{z}|\mathbf{x})||p(\mathbf{z}) \right)$ is the Kullback-Leibler (KL) divergence between the variational posterior distribution and the prior.
+
+In brief, LVMs of the VAE-type consist of a approximate posterior, $q(\mathbf{z}|\mathbf{x})$, an observation model, $p(\mathbf{x}|\mathbf{z})$, and a prior, $p(\mathbf{z})$. 
+% The joint observation model and prior form the generative model which can be efficiently sampled with ancestral sampling which entails first sampling a latent variable $\tilde{\mathbf{z}}\sim p(\mathbf{z})$ followed by sampling the observed variable conditioned on that latent variable, $\tilde{\mathbf{x}}\sim p(\mathbf{x}|\tilde{\mathbf{z}})$.
+With reference to probabilistic coding theory, the approximate posterior is often referred to as the encoder and the observation model as the decoder \cite{kingma_auto-encoding_2014, rezende_stochastic_2014}.
+From a theoretical perspective, the encoder exists solely as the result of choosing to use variational inference to train the decoder rather than e.g.
+MCMC.
+As such, it is also referred to as the inference model.
+However, from a representation learning perspective, the encoder is essential as it can be used to efficiently obtain the representation $\mathbf{z}$ commonly used for downstream tasks.
+It is still possible to evaluate and sample the true posterior distribution $p(\mathbf{z}|\mathbf{x})$ by applying MCMC methods such as Hamiltonian Monte Carlo on the decoder, but for computational reasons this is rarely done in practice.
+
+\begin{table}[t!]
+\caption{
+A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable. 
+The observation, prior and inference models may all belong to one or more of the categories listed under them as detailed in section \ref{sec:plvms}. 
+The types listed here serve as primitives from which more complex structures can be constructed including models with hierarchies of multiple latent variables. 
+% We indicate autoregressiveness (\textbf{\textsc{arx}} and \textbf{\textsc{arz}}) using a ``catch-all" notation $*$, e.g. $\mathbf{x}_{*:t-1}$. 
+% This serves to indicate that autoregressive dependencies can have different span including at the extremes the full sequence $\mathbf{x}_{1:t-1}$ and the last value $\mathbf{x}_{t-1}$.
+% We have not explicitly listed any cases with multiple latent variables as these generally include a multitude of potential dependency structures.
+% We outline notable examples of models with hierarchies of latent variables in section \ref{sec:plvms}.
+}
+\label{tab:lvm-model-primitives}
+\begin{center}
+    % \renewcommand{\arraystretch}{1.1}
+    % \resizebox{0.8\columnwidth}{!}{%
+    \begin{tabular}{ l l l } 
+        \toprule
+        \multicolumn{2}{l}{\textbf{\textsc{Type}}} & \textbf{\textsc{Form}} \\
+        \midrule
+        \multicolumn{3}{c}{\textsc{Observation model}} \\
+        \midrule
+        \textbf{\textsc{arx}} & Autoregressive on $\mathbf{x}_t$      & $p(\mathbf{x}_t|\mathbf{x}_{1:t-1})$ \\
+        \textbf{\textsc{loc}} & Local latent variable                 & $p(\mathbf{x}_{t}|\mathbf{z}_{1:t})$ \\
+        \textbf{\textsc{glb}} & Global latent variable                & $p(\mathbf{x}_{t}|\mathbf{z})$ \\
+        \midrule
+        \multicolumn{3}{c}{\textsc{Prior}} \\
+        \midrule
+        % \textbf{\textsc{aut}} & Autoregressive on $\mathbf{z}_t$      & $p(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ \\
+        % \textbf{\textsc{obs}} & Observation dependent                                & $p(\mathbf{z}_t|\mathbf{x}_{1:t-1})$ \\
+        \textbf{\textsc{arx}} & Autoregressive on $\mathbf{x}_t$      & $p(\mathbf{z}_t|\mathbf{x}_{1:t-1})$ \\
+        \textbf{\textsc{arz}} & Autoregressive on $\mathbf{z}_t$      & $p(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ \\
+        \textbf{\textsc{ind}} & Locally independent $\mathbf{z}_t$                 & $p(\mathbf{z}_t)$ \\
+        \textbf{\textsc{glb}} & Global latent variable                & $p(\mathbf{z})$ \\
+        \midrule
+        \multicolumn{3}{c}{\textsc{Inference model}} \\
+        \midrule
+        \textbf{\textsc{arz}} & Autoregressive on $\mathbf{z}_t$      & $q(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ \\
+        \textbf{\textsc{flt}} & Filtering                             & $q(\mathbf{z}_t|\mathbf{x}_{1:t})$ \\
+        \textbf{\textsc{lsm}} & Local smoothing                       & $q(\mathbf{z}_t|\mathbf{x}_{t-r:t+r})$ \\
+        \textbf{\textsc{gsm}} & Global smoothing                      & $q(\mathbf{z}_t|\mathbf{x}_{1:T})$ \\
+        \textbf{\textsc{glb}} & Global latent variable                & $q(\mathbf{z}|\mathbf{x}_{1:T})$ \\
+        \bottomrule
+    \end{tabular}
+    % }
+    \end{center}
+\end{table}
+
+\begin{sidewaystable*}[t]
+    \caption{
+    Selected latent variable models classified according the attributes defined throughout section \ref{sec:plvms}.
+    %The models are sorted according to the publication date of the first version on arXiv which might differ from the year indicated by the citation. 
+    %This table supplements the classification provided in table \ref{tab:model-taxonomy} which uses binary attributes largely defined for self-supervised methods. 
+    See table~\ref{tab:lvm-model-primitives} for the probability distributions that correspond to each of the attribute short-hands. \textbf{\textsc{hie}} indicates a hierarchical representation.
+    %The mathematical expressions of the observation, prior and inference models for the models considered here can be found in table~\ref{tab:lvm-mathematical-expressions}. 
+    %Figure~\ref{fig:lvm-graphical-models} depicts corresponding graphical model for selected LVMs of this table.
+    }
+    \label{tab:lvm-taxonomy}
+    \begin{center}
+        \renewcommand{\arraystretch}{1.1}
+        \resizebox{0.99\textwidth}{!}{% <------ Don't forget this %
+        \begin{tabular}{ l l | c c c | c c c c | c c c c c | c } 
+            \toprule
+            \multicolumn{2}{c}{} & 
+            \multicolumn{3}{c}{\footnotesize\textsc{Observation}} & 
+            \multicolumn{4}{c}{\footnotesize\textsc{Prior}} & 
+            \multicolumn{5}{c}{\footnotesize\textsc{Inference}} \\
+            \textbf{\textsc{model}} & 
+            \textbf{\textsc{pub. date}} &
+            \textbf{\textsc{arx}} & 
+            \textbf{\textsc{loc}} & 
+            \textbf{\textsc{glb}} &  
+            \textbf{\textsc{arx}} & 
+            \textbf{\textsc{arz}} & 
+            \textbf{\textsc{ind}} & 
+            \textbf{\textsc{glb}} &  
+            \textbf{\textsc{arz}} & 
+            \textbf{\textsc{flt}} & 
+            \textbf{\textsc{lsm}} & 
+            \textbf{\textsc{gsm}} & 
+            \textbf{\textsc{glb}} &
+            \textbf{\textsc{hie}} \\
+            \midrule
+            %                                                              OBSERVATION       |              PRIOR                |                  INFERENCE       
+            %                                                         ARX      LOC      GLB      ARX      ARZ     IND      GLB      ARZ      FLT      LSM      GSM      GLB
+
+            % Conv. DBN \cite{lee_unsupervised_2009}    & 2009 Jan. & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\
+            \textbf{VRNN} \footnotesize\cite{chung_recurrent_2015}          & 2015 Jun.
+& \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark \\
+            \textbf{SRNN} \footnotesize\cite{fraccaro_sequential_2016}      & 2016 May  & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark \\
+            \textbf{HMM-VAE} \footnotesize\cite{ebbers_hidden_2017}         & 2017 Mar.
+& \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
+            \textbf{ConvVAE} \footnotesize\cite{hsu_learning_2017}          & 2017 Apr.
+& \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \cmark & \xmark \\
+            \textbf{FHVAE} \footnotesize\cite{hsu_unsupervised_2017}        & 2017 Sep.
+& \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \cmark & \cmark \\
+            \textbf{VQ-VAE} (van den Oord et al. 2018)            & 2017 Nov. & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\
+            \textbf{BHMM-VAE} \footnotesize\cite{glarner_full_2018}         & 2018 Mar.
+& \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark \\
+            \textbf{STCN} \footnotesize\cite{aksan_stcn_2019}               & 2019 Feb.
+& \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
+            \textbf{FDMM} \footnotesize\cite{khurana_factorial_2019}        & 2019 Oct.
+& \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark \\
+            \textbf{ConvDMM} \footnotesize\cite{khurana_convolutional_2020} & 2020 Jun.
+& \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark \\
+            % WaveNet AE        &  \\
+            % HMM-VAE           &  \\
+            % Convolutional DBN &  \\
+            \bottomrule
+        \end{tabular}
+        }
+    \end{center}
+\end{sidewaystable*}
+
+We next review VAEs applied to speech. We consider the choices of observation, prior and inference models.
+% and introduce attributes throughout the text. An overview of these is found in table~\ref{tab:lvm-model-primitives}.
+We provide a model taxonomy for selected LVMs in table~\ref{tab:lvm-taxonomy}.
+% Supplementing the binary attributes, table~\ref{tab:lvm-mathematical-expressions} explicitly details the observation, prior and inference models of the same selection of LVMs considered in table~\ref{tab:lvm-taxonomy}.
+
+\paragraph{Observation models} 
+A common choice for the observation model $p(\mathbf{x}|\mathbf{z})$ is to include an autoregressive dependency on the observed variable (\textbf{\textsc{arx}}, table \ref{tab:lvm-model-primitives}) that is, $p(\mathbf{x}_t|\mathbf{x}_{1:t-1}, \cdot)$ where $\cdot$ represents some dependency on the latent variable \cite{chung_recurrent_2015, fraccaro_sequential_2016, oord_neural_2018, oord2018representation}.
+This allows the latent representation to focus on correlations that cannot easily be predicted from the observed variable at previous time-steps \cite{oord2018representation}.
+In practice, the dependency on $\mathbf{x}_{1:t-1}$ is often assumed to be Markovian and hence only on $\mathbf{x}_{t-1}$.
+Another common choice is to depend on a local window $\mathbf{x}_{t-r:t-1}$ where $r>1$ is an integer denoting some receptive field.
+We will take a dependency on $\mathbf{x}_{1:t-1}$ to mean any one of these choices unless otherwise specified.
+
+While the autoregressive dependency might be important for learning a powerful generative model, it might not benefit the learned latent representations.
+Specifically encouraging the latent representation to discard correlations across the temporal dimension might degrade the quality of the latent representation.
+Furthermore, since such a decoder can perform quite well by simply solving an autoregressive prediction problem, similar to WaveNet \cite{oord_wavenet:_2016}, it can make the model prone to suffer from posterior collapse.
+This problem arises when the approximate and true posterior distributions collapse into the prior which renders the representations non-informative \cite{bowman_generating_2016, sonderby_ladder_2016}.
+Notably, posterior collapse is a local minimum of the ELBO since the KL-divergence becomes zero.
+Some works alleviate this problem with tricks like KL-annealing and free bits \cite{bowman_generating_2016, sonderby_ladder_2016, kingma_improved_2016}. 
+%KL-annealing down-weighs the KL-divergence in the ELBO (eq.~\ref{eq: lvm likelihood bound recon/kl form (elbo)}) gradually phasing it in during an initial training phase \cite{bowman_generating_2016, sonderby_ladder_2016}.
+Free bits simply returns a zero gradient for KL-divergences smaller than some number.
+\cite{kingma_improved_2016}.
+The VQ-VAE uses a quantized latent space that is not susceptible to posterior collapse \textit{per se} \cite{oord_neural_2018}.
+How to equip LVMs with powerful decoders while avoiding posterior collapse is an open problem.
+
+Some LVMs do not use autoregressive observation models \cite{ebbers_hidden_2017, glarner_full_2018, hsu_unsupervised_2017, hsu_learning_2017, khurana_factorial_2019, khurana_convolutional_2020}.
+These more closely follow the assumption of \emph{local independence} which states that observed variables are conditionally independent given the local (\textbf{\textsc{loc}}, table \ref{tab:lvm-model-primitives}) and/or global (\textbf{\textsc{glb}}, table \ref{tab:lvm-model-primitives}) latent variables \cite{bartholomew_latent_2011}.
+However, this forces the latent variable to encode details about the observed variable to achieve a good reconstruction.
+This is opposite to contrastive self-supervised learning which allows models to discard details in $\mathbf{x}_{1:T}$ that do not inform the training objective \cite{baevski2020wav2vec}.
+
+\paragraph{Priors}
+Priors can be said to belong to one or more of four broad categories. See table~\ref{tab:lvm-model-primitives}. 
+% Autoregressive dependency on the observed variable (\textbf{\textsc{arx}}, table \ref{tab:lvm-model-primitives}), autoregressive dependency on the latent variable (\textbf{\textsc{arz}}, table \ref{tab:lvm-model-primitives}), locally independent (\textbf{\textsc{ind}}, table \ref{tab:lvm-model-primitives}) and global (\textbf{\textsc{glb}}, table \ref{tab:lvm-model-primitives}).
+Priors that are autoregressive on the observed variable (\textbf{\textsc{arx}}, table \ref{tab:lvm-model-primitives}) take the form $p(\mathbf{z}_t|\mathbf{x}_{1:t-1})$.
+This generally results in a slow-down of the generative process which may be of concern if the use-case is data generation.
+%It is also interesting to note that this encourages the latent variables to model less dynamic behaviour in their stochastic transitions and instead rely more on the observed variable.
+Priors that are autoregressive on the latent variable (\textbf{\textsc{arz}}, table \ref{tab:lvm-model-primitives}) take the form $p(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ and enable stochastic temporal transitions similar to hidden Markov models but with potentially nonlinear transition functions \cite{chung_recurrent_2015, fraccaro_sequential_2016, khurana_factorial_2019, khurana_convolutional_2020}.
+Locally independent priors (\textbf{\textsc{ind}}, table \ref{tab:lvm-model-primitives}) are rarely applied to sequential latent variables since they make the prior latent dynamics independent of the value of previous latent variables.
+Models that do impose such priors on sequential latents are quite limited in their generative power, unless they learn the prior dynamics post-hoc as done in the VQ-VAE \cite{oord_neural_2018}.
+Global latent variables (\textbf{\textsc{glb}}, table \ref{tab:lvm-model-primitives}) are fundamentally limited in the amount of information they can encode.
+Hence, models usually use them in combination with another local latent variable, or to encode fixed length input segments \cite{khurana_factorial_2019, hsu_learning_2017, hsu_unsupervised_2017}.
+
+% Many LVMs have priors that do not have autoregressive dependence on the observed variable (table \ref{tab:lvm-model-primitives}).
+This allows generating the latent sequence in full before generating the observed sequence.
+It also permits conditioning past observed variables on future latent variables.
+Despite this, it is usual to let $\mathbf{x}_t$ depend causally on $\mathbf{z}_{1:t}$ in the generative model.
+To our knowledge, so far no work has examined full (non-causal) dependence of $\mathbf{x}_{1:T}$ on $\mathbf{z}_{1:T}$ during generation.% where, for example, $\mathbf{z}_T$ can influence $\mathbf{x}_1$.
+
+\paragraph{Inference models} 
+LVMs based on the VAE perform so-called amortized variational inference. Here, a single inference network is used to infer the latent variables of any $\mathbf{x}$. 
+For this reason, all inference models covered here are conditioned on the observed sequence in some way. Generally, the inference model can be seen as solving either a filtering or smoothing problem.
+In filtering (\textbf{\textsc{flt}}, table \ref{tab:lvm-model-primitives}), the latent variables are assumed to depend only on past and current values of the observed variable, $q(\mathbf{z}_t|\mathbf{x}_{1:t})$ \cite{chung_recurrent_2015, khurana_convolutional_2020}.
+In global smoothing (\textbf{\textsc{gsm}}, table \ref{tab:lvm-model-primitives}), this causal dependency is replaced with a dependency on all observed values, $q(\mathbf{z}_t|\mathbf{x}_{1:T})$ \cite{fraccaro_sequential_2016, hsu_learning_2017}.
+Smoothing can also be done locally (\textbf{\textsc{lsm}}, table \ref{tab:lvm-model-primitives}), where the latent variables then depend on $\mathbf{x}_{t-r:t+r}$ for some integer $r>0$ \cite{oord_neural_2018}.
+Compared to self-supervised models that often use transformer encoders it can be hypothesized that global smoothing offers a stronger case than local smoothing and filtering for representation learning.
+
+The inference model may also be used to infer a global latent variable (\textbf{\textsc{glb}}, table \ref{tab:lvm-model-primitives}) that might encode global information about $\mathbf{x}$. 
+While it must be included in the prior model it might not be in the observation model, if the model also has a local latent variable. 
+Finally, latent variables are often made to depend autoregressively on past inferred values, e.g.
+$q(\mathbf{z}_t|\mathbf{z}_{1:t-1},\mathbf{x}_{1:t})$ (\textbf{\textsc{arz}}, table \ref{tab:lvm-model-primitives}) \cite{chung_recurrent_2015, fraccaro_sequential_2016}.
+% Finally, latent variables are often made to depend autoregressively on past values, $q(\mathbf{z}_t|\mathbf{z}_{1:t-1},\mathbf{x}_{1:t})$ for filtering and $q(\mathbf{z}_t|\mathbf{z}_{1:t-1},\mathbf{x}_{1:T})$ for smoothing (\textbf{\textsc{arz}}, table \ref{tab:lvm-model-primitives}) \cite{chung_recurrent_2015, fraccaro_sequential_2016}.
+%Such autoregressive dependencies help the inference model better match the prior dynamics but notably are not employed in models that learn the prior dynamics post-hoc \cite{oord_neural_2018}.
+
+\paragraph{Multiscale and hierarchical models} Some work has explored using a hierarchy of latent variables (\textbf{\textsc{hie}}, table \ref{tab:lvm-taxonomy}).
+This allows encoding the inductive bias that speech contains information at different temporal scales by letting the latent variables operate at different temporal scales \cite{hsu_unsupervised_2017}.
+\citet{khurana_factorial_2019} propose using a temporal latent variable along with a global latent variable.
+Recent work has focused on learning a deeper latent hierarchy with five latent variables \cite{aksan_stcn_2019}. 
+%Here, the latents have no autoregressive temporal dependencies but instead depend on latent variables further up in the hierarchy with all latent variables operating at the same temporal scale. 
+%Interestingly, this effort within LVMs towards separating learned features into representations at different time scales or learning hierarchical representations stands in contrast to work on self-supervised learning where focus is either on learning local or global representations.
+
+\paragraph{Other work}
+% Most recent works learn latent variable models via amortized variational inference as introduced by the VAE framework \cite{kingma_auto-encoding_2014, rezende_stochastic_2014}. 
+Before the introduction of the VAE, models such as deep belief networks (DBN, \citealp{hinton_fast_2006}) built from stacks of restricted Boltzmann machines \cite{smolensky_parallel_1987,fischer:13} were popular.
+\citet{lee_unsupervised_2009} show the feasibility of using a two-layered DBN for discovering acoustic units of speech, while \citet{deng_binary_2010} show that a DBN can learn a binary coding of spectrograms that has higher signal-to-noise ratio than classical vector-quantization techniques for speech coding.
+DBNs are however notoriously tricky to optimize requiring the use of expensive MCMC sampling techniques for inference or resort to biased gradient estimates \cite{hinton_practical_2012,fischer:10c}.
+Non-neural LVMs for speech representation learning have also been explored
+\cite{lee_nonparametric_2012, ondel_variational_2016, heck_feature_2017, jansen_weak_2013}.
+
+% Non-neural approaches include Bayesian non-parametric models which can be used for unsupervised acoustic unit discovery.
+This is often done using a Dirichlet process prior which has the appealing property of automatically inferring the number of acoustic units \cite{lee_nonparametric_2012, ondel_variational_2016, heck_feature_2017}.
+% Finally, Gaussian mixture models have been used to form so-called universal background models from which acoustic representations can be extracted \cite{jansen_weak_2013}.
+
+\section{Discussion}
+\label{sec:mtax}
+%Throughout the text we defined several binary attributes which are used to characterize a selection of models in table \ref{tab:model-taxonomy} and \ref{tab:lvm-taxonomy}.
+We discuss representation learning models based on this overarching model taxonomy below.
+
+%In table \ref{tab:model-taxonomy}, we use those derived from the self-supervised models to characterize a selection of models within the two surveyed categories.
+To obtain a more detailed understanding of the LVMs, we also derived a set of variable dependencies  presented  in table \ref{tab:lvm-model-primitives}.
+In table \ref{tab:lvm-taxonomy} we provide a specification of how they apply to selected LVMs.
+
+% Throughout the text we defined a set of binary model attributes.
+In table \ref{tab:model-taxonomy}, we use these to characterize a selection of models within the two surveyed categories.
+These attributes have been derived with a focus on the self-supervised models.
+As a result, the characterization of the latent variable models is fairly uniform in comparison.
+Thus, to obtain a more detailed understanding, we also derived a set of dependencies between variables used to specify the graphical models for the LVMs.
+These are presented in table \ref{tab:lvm-model-primitives} and a specification of how they apply to selected models is given in table \ref{tab:lvm-taxonomy}.
+Below, we compare models included in this survey using table \ref{tab:model-taxonomy} and \ref{tab:lvm-taxonomy} as a basis for the discussion.
+
+\paragraph{From global to local}
+
+In table \ref{tab:model-taxonomy}, we see that work on global representations within self-supervised learning precedes work on local representations.
+However, we find that the core ideas underlying the recent successes in learning local representation models have also been used for global representation learning; masking \cite{chung2016audio}, context prediction \cite{chung2018speech2vec}, and contrastive training \cite{milde2018unspeech} have been applied in both settings.
+Furthermore, where work on global representation learning has taken inspiration from Word2vec \cite{mikolov2013distributed}, the techniques used for learning local representations are inspired by contextualized word embeddings \cite{devlin2019bert}.
+Thus, the gap between these two model classes is largely a product of the developments in related fields and the general increase in computational resources.
+
+\paragraph{Representations beyond inference} Predictive tasks are commonly used for self-supervised models, but they are not directly compatible with LVM training. 
+However, an LVM prior with an autoregressive parameterization, $p(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ or $p(\mathbf{z}_t|\mathbf{x}_{1:t-1})$, can be seen as predictive in the sense that it tries to match the approximate posterior.
+%, which conditions on $\mathbf{x}_t$, using only latent dynamics and past observed variables. 
+Hence, the prior might be considered for feature extraction.
+%although it remains unclear whether a representation extracted from the prior is better suited for downstream tasks than one extracted from the posterior.
+\citet{jones2020discrete} examine the importance of the prior in the VQ-VAE and show that the ability of this model to estimate densities $p(\mathbf{x}_{1:T})$ lies solely in its prior.
+Other work has also explored representations beyond the latent variable such as hidden units of the observation model \cite{khurana_convolutional_2020, chorowski_unsupervised_2019}.
+
+% Hidden units from the observation model have also been used as downstream task features in previous work \cite{khurana_convolutional_2020}.
+Similarly, \citet{chorowski_unsupervised_2019} explore representations beyond the latent variable.
+
+\paragraph{Masking and missing data}
+Masking may also improve representations learned with VAEs. 
+Masking in VAEs has already been explored in the literature in the context of missing data imputation.
+Here, $\mathbf{x}$ is only partially observed, and often represented as a segmentation into observed and missing parts and a mask $\mathbf{m}$ indicating where the data is missing.
+The model is then trained to infer the latent variable from the observed data.
+Reconstruction also deals only with the observed data.
+Previous work has largely focused on the ability of these models to yield high-quality imputations within the tabular and image data domains, without probing for the effects on the learned latent representation \cite{mattei_miwae_2019, ipsen_not-miwae_2021}.
+The idea of using VAEs to impute missing data was already examined in the seminal paper by \citet{rezende_stochastic_2014}.
+Here the model was trained with fully observed data and used to impute data in an iterative sampling approach post hoc leaving the learned representations unchanged.
+
+\paragraph{Evaluating representations} 
+Although this review has a primarily methodological focus, we should briefly touch upon evaluation procedures.
+Training metrics for self-supervised tasks and the likelihood of LVMs offer little guidance as to the quality of the learned representations \cite{huszar_is_2017}.
+Thus, a common approach is to evaluate the representations in terms of their usefulness for downstream tasks.
+Such tasks may be chosen to target specific attributes of the representation (e.g.
+semantic or speaker information).
+
+The SUPERB benchmark \cite{yang2021superb} gathers multiple tasks grouped into categories such as \emph{recognition}, \emph{detection}, \emph{semantics}, \emph{speaker}, \emph{paralinguistics} and \emph{generation}.
+The recently proposed SLUE benchmark focuses on spoken language understanding \cite{shon2021slue}.
+The long-standing zero resource speech challenge (ZeroSpeech) offers a new set of tasks for each edition \cite{versteegh2015zero, dunbar2017zero, dunbar2019zero, dunbar2020zero, dunbar2021zero} usually featuring a minimal-pair ABX task \cite{schatz2013evaluating, schatz2014evaluating}.
+
+Tasks that evaluate representations in terms of speaker-related information include speaker verification \cite{hsu_unsupervised_2017, khurana_factorial_2019, milde2018unspeech}, speaker identification \cite{oord2018representation, jati2019neural, chung2019unsupervised, liu2020non}, dialect classification \cite{khurana_factorial_2019}, emotion recognition \cite{Pascual2019, yang2021superb} and gender classification \cite{lee_unsupervised_2009}.
+The semantic content of representations are evaluated using tasks such as intent classification \cite{morais2021end, yang2021superb}, slot filling \cite{lai2021semi, yang2021superb}, sentiment analysis \cite{liu2020mockingjay}, question answering \cite{chung2020splat}, named entity recognition \cite{shon2021slue, borgholt2021we, pasad2021use} and speech translation \cite{bansal2017towards, chung2020generative}.
+Cardiac arrest detection for emergency calls has also been used to evaluate speech representations \cite{borgholt2021we}.
+For local representations, phoneme classification is very common \cite{lee_unsupervised_2009, hsu_learning_2017, chorowski_unsupervised_2019, chung2019unsupervised, liu2021tera}.
+However, automatic speech recognition has become the \textit{de facto} standard benchmark task \cite{ling2020decoar, chung2020generative, hsu2021hubert}.
+
+\paragraph{Moving forward}
+
+Most of the seminal work has focused on improving speech recognition \cite{schneider2019wav2vec, baevski2020wav2vec}.
+This focus has gained traction over the last couple of years, as computational resources have become more accessible and end-to-end models \cite{graves2006connectionist, chan2016listen} have been established as the dominant approach to speech recognition \cite{gulati2020conformer}.
+It is important to stress that self-supervised models, such as wav2vec 2.0 \cite{baevski2020wav2vec}, represent a breakthrough, and recent successful approaches build upon this method.
+That is, deep self-attention models combined with masking \cite{hsu2021hubert, wang2021unispeech, chen2021wavlm}.
+This development mirrors years of rapid progress in masked language modeling within natural language processing \cite{devlin2019bert,clark_2020_electra} and we expect this to continue for unsupervised neural speech representation learning.
+
+\section{Conclusion}
+\label{sec:conc}
+
+We reviewed unsupervised representation learning for speech, focusing on two primary categories: self-supervised methods and probabilistic latent variable models.
+Inspired by the development of self-supervised learning and the dependency structures of latent variable models, we derived a comprehensive model taxonomy.
+Finally, we compare and discuss models from the two categories and their respective evaluation procedures.
+
+% }

### 2023-09-23 appendices/paper_brief.tex
Work on variational autoencoders

@@
 \label{app:paper-brief}
-% \ifthenelse{\equal{\skipappendices}{true}}{}{
-
-% \newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
-
-% \newcommand{\norm}[1]{\left\lVert#1\right\rVert}
+\ifthenelse{\equal{\skipappendices}{true}}{}{
 
@@
 \section{Introduction}
-Representation learning has shaped modern computer vision \cite{simonyan2014very} and natural language processing \cite{devlin2019bert}, and more recently speech processing has been subject to the same development \cite{baevski2020wav2vec}.
+Representation learning has shaped modern computer vision \cite{simonyan_very_2014} and natural language processing \cite{devlin_bert_2018}, and more recently speech processing has been subject to the same development \cite{baevski_wav2vec_2020}.
 Representation learning has been defined as \emph{``learning representations of the data that make it easier to extract useful information when building classifiers or other predictors"} \cite{bengio_representation_2013}.
@@
 The model is then fine-tuned, or used to extract input representations for a smaller model, targeting a task with limited training data.
-In computer vision, both supervised \cite{simonyan2014very, szegedy2015going, he2016deep} and unsupervised \cite{pathak2016context, doersch2015unsupervised} representation learning have gained attention with supervised representation learning driven by the availability of large annotated datasets \cite{deng2009imagenet}.
+In computer vision, both supervised \cite{simonyan_very_2014, szegedy_going_2015, he_deep_2015} and unsupervised \cite{pathak_context_2016, doersch_unsupervised_2015} representation learning have gained attention with supervised representation learning driven by the availability of large annotated datasets \cite{deng_imagenet_2009}.
 For text and speech, pre-training is usually unsupervised as labeled data is difficult to obtain.
@@
 Finally, we compare and discuss models from the two categories and their respective evaluation procedures.
-%We then provide a comparison of the methods based on their respective evaluation procedures.
-%Finally, based on the model taxonomy, we discuss the differences between models from the two categories and identify future challenges and directions for research in speech representation learning.
 
@@
 
-\section{Unsupervised representation learning}
+\section{Unsupervised representation learning}\label{ssec:notation}
 
@@
 
-%Similarly, $\textbf{a}_{<t}$ denotes a vector-sequence containing all elements prior to step $t$ and $\textbf{a}_{t \in \mathbb{A}}$ is a vector sequence containing elements with index contained in the set $\mathbb{A}$.
-%(or a scalar sequence $a_{i:j}$)
-
-%Input to a model is either a spectrogram $\mathbf{x}_{1:T}$ or the raw speech signal $x_{1:L}$ where $L > T$. Commonly $L = T(\frac{S}{100})$ where $S$ is the sample rate of the raw speech signal.
-
-% \IEEEpubidadjcol % related to the IEEEpubid command in the top
-
-\paragraph{Notation} \label{ssec:notation} We use the subscript $i{:}j$ with $i{\leq}j$ to denote a vector sequence $\textbf{a}_{i:j}$ containing elements $\textbf{a}_{i}$ through $\textbf{a}_{j}$.
+\paragraph{Notation}
+We use the subscript $i{:}j$ with $i{\leq}j$ to denote a vector sequence $\textbf{a}_{i:j}$ containing elements $\textbf{a}_{i}$ through $\textbf{a}_{j}$.
 We denote model input as $\mathbf{x}_{1:T}$ which, in practice, might be either a spectrogram or the raw speech signal, but we do not distinguish between the two in notation as it is not essential to understand the models.
@@
 For stochastic latent variable models, we use $\mathbf{z}_{1:T}$ as is customary to the field.
-% The representation at a single time-step, $\mathbf{c}_t$ or $\mathbf{z}_t$, is typically a function of a larger part of the input space than just $\mathbf{x}_t$, in some cases the entire input sequence $\mathbf{x}_{1:T}$.
 While some models are frozen and produce representations used as input for downstream tasks (\textbf{\textsc{frz}}, table \ref{tab:model-taxonomy}), others are designed to be fine-tuned (\textbf{\textsc{ftn}}, table \ref{tab:model-taxonomy}).
@@
 
-Self-supervised learning is a subset of unsupervised learning \cite{tsai2020self}.
+Self-supervised learning is a subset of unsupervised learning \cite{tsai_selfsupervised_2021}.
 Where other unsupervised methods can be seen as a means to an end in itself (e.g., clustering or data generation), self-supervised learning takes the form of a pretext task that only adds value when associated with a downstream task.
-This makes self-supervised learning tie naturally with semi-supervised learning, but it may also be part of a fully unsupervised setup \cite{baevski2021unsupervised}.
-Self-supervised learning is often characterized by automatically deriving the target from the input or other unlabeled examples \cite{ouali2020overview}.
+This makes self-supervised learning tie naturally with semi-supervised learning, but it may also be part of a fully unsupervised setup \cite{baevski_unsupervised_2021}.
+Self-supervised learning is often characterized by automatically deriving the target from the input or other unlabeled examples \cite{ouali_overview_2020}.
 
-\paragraph{Predictive models} Similar to classic autoregressive language models \cite{mikolov2010recurrent}, contextualized speech representations can be learned by predicting future values of a simple representation \cite{oord2018representation, chung2019unsupervised, schneider2019wav2vec, chung2020generative, jiang2021further} (\textbf{\textsc{prd}}, table \ref{tab:model-taxonomy}).
-Modeling spectrograms directly, autoregressive predictive coding (APC, \citealp{chung2019unsupervised}) is perhaps the simplest example in this category.
+\paragraph{Predictive models}
+Similar to classic autoregressive language models \cite{mikolov_recurrent_2010}, contextualized speech representations can be learned by predicting future values of a simple representation \cite{oord_representation_2018, chung_unsupervised_2019, schneider_wav2vec_2019, chung_generative_2020, jiang_further_2021} (\textbf{\textsc{prd}}, table \ref{tab:model-taxonomy}).
+Modeling spectrograms directly, autoregressive predictive coding (APC, \citealp{chung_unsupervised_2019}) is perhaps the simplest example in this category.
 The forward pass and loss are computed as
 \begin{align}
-    \mathbf{c}_{t} &= f(\mathbf{x}_{1:t})  \label{eq:apc_f} \\ 
+    \mathbf{c}_{t} &= f(\mathbf{x}_{1:t})  \label{eq_brief:apc_f} \\ 
     \hat{\mathbf{x}}_{t+k} &= g(\mathbf{c}_{t}) \\
-    %\mathcal{L}_t(\mathbf{\hat{x}}_{t+k}, \mathbf{x}_{t+k}) &= |\mathbf{\hat{x}}_{t+k} - \mathbf{x}_{t+k}|
     \mathcal{L}_t &= \lVert \hat{\mathbf{x}}_{t+k} - \mathbf{x}_{t+k} \rVert_1\enspace.
@@
 \noindent Here, $f(\cdot)$ and $g(\cdot)$ are parameterized by neural networks such that each $\mathbf{c}_t$ is only conditioned on previous inputs $\mathbf{x}_{1:t}$ and $\hat{\mathbf{x}}_{t+k}$ is computed step-wise.
-\citet{chung2019unsupervised} use a stack of unidirectional LSTMs for $f(\cdot)$ and a linear regression layer for $g(\cdot)$.
+\citet{chung_unsupervised_2019} use a stack of unidirectional LSTMs for $f(\cdot)$ and a linear regression layer for $g(\cdot)$.
 Tasks that seek to predict or reconstruct the input are very common.
-In the literature, these are often jointly referred to as reconstruction tasks (\textbf{\textsc{rec}}, table \ref{tab:model-taxonomy}) \cite{liu2021tera, wang2021unispeech}, although this is somewhat misleading in the case of prediction.
+In the literature, these are often jointly referred to as reconstruction tasks (\textbf{\textsc{rec}}, table \ref{tab:model-taxonomy}) \cite{liu_tera_2021, wang_unispeech_2021}, although this is somewhat misleading in the case of prediction.
 
-Contrary to generative models, such as WaveNet \cite{oord_wavenet:_2016}, the APC model is not restricted to next-step prediction.
+Contrary to generative models, such as WaveNet \cite{oord_wavenet_2016}, the APC model is not restricted to next-step prediction.
 Instead, it predicts $k > 0$ steps ahead in order to ensure that the model does not learn a trivial solution by exploiting the smoothness of the signal.
-Depending on the downstream task, we are often interested in learning so-called slow features that will typically span multiple input frames \cite{wiskott2002slow}.
+Depending on the downstream task, we are often interested in learning so-called slow features that will typically span multiple input frames \cite{wiskott_slow_2002}.
 Even the smallest linguistic units of speech, phonemes, tend to span $0.1$ seconds on average \cite{garofolo_timit_1993}, whereas spectrogram frames $\mathbf{x}_t$ are typically computed at $0.01$ second intervals.
-However, sometimes local smoothness is explicitly used to define the task \cite{badino2014auto, jati2017speaker2vec, jati2019neural}.
+However, sometimes local smoothness is explicitly used to define the task \cite{badino_autoencoder_2014, jati_speaker2vec_2017, jati_neural_2019}.
 
-%In early work, \citet{badino2014auto} propose a step-wise bottleneck model that learns to predict $\mathbf{x}_{t+1}$ from $\mathbf{x}_{t}$.
-Since phonetic information changes slowly, $\mathbf{x}_{t+1}$ is argued to be a noisy version of $\mathbf{x}_{t}$ encoding almost the same phonetic content.
-Thus, the model is akin to a denoising autoencoder.
-%A similar idea has been proposed by Jati and Georgiou \cite{jati2019neural} to learn speaker embeddings.
-
-% Next-step prediction has also been frequently used for generative models (\textbf{\textsc{gen}}, table \ref{tab:model-taxonomy}).
-The popular speech synthesis model WaveNet is perhaps the most obvious example \cite{oord_wavenet:_2016}.
-While this work does not consider representation learning, the model might learn useful features similar to those learned with APC.
-
-\paragraph{Contrastive models} Speech contains localized noise (e.g., phase shifts) that does not inform slow feature learning.
+\paragraph{Contrastive models}
+Speech contains localized noise (e.g., phase shifts) that does not inform slow feature learning.
 Thus, directly modeling speech might not be the best way to learn contextualized representations.
-Contrastive predictive coding (CPC,  \citealp{oord2018representation}) targets a local variable $\mathbf{v}_{1:T}$, learned from the model input $\mathbf{x}_{1:T}$, instead of the input itself.
+Contrastive predictive coding (CPC,  \citealp{oord_representation_2018}) targets a local variable $\mathbf{v}_{1:T}$, learned from the model input $\mathbf{x}_{1:T}$, instead of the input itself.
 The forward pass is
@@
 \begin{align}
-    \mathbf{v}_{t} &= f_{v}(\mathbf{x}_{t-r:t+r}) \label{eq: cpc local representation} \\
+    \mathbf{v}_{t} &= f_{v}(\mathbf{x}_{t-r:t+r}) \label{eq_brief: cpc local representation} \\
     \mathbf{c}_{t} &= f_{c}(\mathbf{v}_{1:t}) \\
     \hat{\mathbf{v}}_{t,k} &= g_k(\mathbf{c}_{t}) \enspace,
-    %g_k(\mathbf{c}_{t}, \mathbf{v}_{t+k}) &= \mathbf{v}_{t+k}^{\intercal} \mathbf{W}_k \mathbf{c}_{t}
-    %\mathcal{L}_{t,k} &= - \log \left(\frac{g_k(\mathbf{c}_{t}, \mathbf{v}_{t+k})}{\sum_{n \sim \mathcal{D}} g_k(\mathbf{c}_{t}, \mathbf{v}_{n})} \right)
-    % \mathcal{L}_{t,k} &= -\log g_k + \log \sum_{n \sim \mathcal{D}} g_k
 \end{align}
@@
 Again, $f_c(\cdot)$ should be limited to condition each $\mathbf{c}_t$ on previous time-steps $\mathbf{v}_{1:t}$ and $g_k(\cdot)$ is a step-wise transformation.
-The loss is based on noise constrastive estimation \cite{gutmann2010noise} and is given by
+The loss is based on noise constrastive estimation \cite{gutmann_noisecontrastive_2010} and is given by
 \begin{align}
     \mathcal{L}_{t,k} &= - \log \left(\frac{\exp(\hat{\mathbf{v}}_{t,k}^{\text{\tiny T}}\mathbf{v}_{t+k})}{\sum_{n \sim \mathcal{D}} \exp(\hat{\mathbf{v}}_{t,k}^{\text{\tiny T}}\mathbf{v}_{n})} \right)\enspace .
-    % \mathcal{L}_{t,k} &= - \hat{\mathbf{v}}_{t,k}^{\text{\tiny T}}\mathbf{v}_{t+k} + \log \sum_{n \sim \mathcal{D}} \exp(\hat{\mathbf{v}}_{t,k}^{\text{\tiny T}}\mathbf{v}_{n})\enspace .
-    % \mathcal{L}_{t,k} &= -\log g_k + \log \sum_{n \sim \mathcal{D}} g_k
-    \label{eq: cpc loss}
+    \label{eq_brief: cpc loss}
 \end{align}
@@
 Note that the loss is also indexed by $k$ to show that CPC targets multiple offsets.
-The APC model is easily extended in a similar way \cite{chung2020improved}.
+The APC model is easily extended in a similar way \cite{chung_improved_2020}.
 
@@
 
-A model closely related to the original CPC model is wav2vec \cite{schneider2019wav2vec}.
+A model closely related to the original CPC model is wav2vec \cite{schneider_wav2vec_2019}.
 It uses a different parameterization of the functions $f_v(\cdot)$ and $f_c(\cdot)$, and modifies the loss to consider a binary prediction task, such that we have
@@
 \begin{align}
-    %g_k(\mathbf{c}_{t}, \mathbf{v}_{t+k}) &= \log(\sigma(\mathbf{v}_{t+k}^{\intercal} (\mathbf{W}_k \mathbf{c}_{t} + \mathbf{b}_k)))\\
     \mathcal{L}_{t,k} &= - \log(\sigma(\hat{\mathbf{v}}_{t,k}^{\text{\tiny T}}\mathbf{v}_{t+k})) - \sum_{n \sim \mathcal{D}} \log(\sigma(-\hat{\mathbf{v}}_{t,k}^{\text{\tiny T}}\mathbf{v}_{n})) \,.
-    \label{eq: wav2vec loss}
+    \label{eq_brief: wav2vec loss}
 \end{align}
@@
 \paragraph{Masking-based models} One downside of predictive tasks is that models are primarily unidirectional.
-Some work has extended APC and CPC inspired models with separate encoders operating in opposite directions \cite{ling2020deep, kawakami2020learning, borgholt2021scaling}, but these models are still restricted to process left and right context separately.
-Inspired by the masked language model task used for text-based representation learning \cite{devlin2019bert}, several papers have used masking to overcome this challenge (\textbf{\textsc{msk}}, table \ref{tab:model-taxonomy}).
+Some work has extended APC and CPC inspired models with separate encoders operating in opposite directions \cite{ling_deep_2020, kawakami_learning_2020, borgholt_scaling_2021}, but these models are still restricted to process left and right context separately.
+Inspired by the masked language model task used for text-based representation learning \cite{devlin_bert_2018}, several papers have used masking to overcome this challenge (\textbf{\textsc{msk}}, table \ref{tab:model-taxonomy}).
 Masking refers to replacing parts of the input with zeros or a learned masking vector.
-For zero-masking \cite{jiang2019improving, liu2020mockingjay, wang2020unsupervised, chi2021audio, ling2020decoar}, we have
+For zero-masking \cite{jiang_improving_2019, liu_mockingjay_2020, wang_unsupervised_2020, chi_audio_2020, ling_decoar_2020}, we have
 \begin{align}
@@
 The entries of $\mathbf{m}_{1:T}$ are determined by some stochastic policy.
-One frequent inspiration is SpecAugment \cite{park2019specaugment}, which was originally proposed for supervised speech recognition and applies frequency and time masking to spectrogram representations.
-While temporal masking is most common, frequency masking has also been adopted for representation learning  \cite{wang2020unsupervised}.
-A simple, yet popular, masking strategy is to draw a proportion of input indices $t_i\sim\{1,\dots,T-M\}$ without replacement, and then mask $\{t_i, \dots, t_i+M\}$ \cite{baevski2020wav2vec, hsu2021hubert, ling2020decoar}.
+One frequent inspiration is SpecAugment \cite{park_specaugment_2019}, which was originally proposed for supervised speech recognition and applies frequency and time masking to spectrogram representations.
+While temporal masking is most common, frequency masking has also been adopted for representation learning  \cite{wang_unsupervised_2020}.
+A simple, yet popular, masking strategy is to draw a proportion of input indices $t_i\sim\{1,\dots,T-M\}$ without replacement, and then mask $\{t_i, \dots, t_i+M\}$ \cite{baevski_wav2vec_2020, hsu_hubert_2021, ling_decoar_2020}.
 
@@
     \mathbf{c}_{1:T} &= f_c(\mathbf{v}_{1:T} \circ \mathbf{m}_{1:T}) \\
-    \mathbf{q}_t &= g_q(\mathbf{v}_t) \enspace . \label{w2v2 qtz}
+    \mathbf{q}_t &= g_q(\mathbf{v}_t) \enspace . \label{eq_brief: w2v2 qtz}
 \end{align}
 
-\noindent Here, $f_v(\cdot)$ is a convolutional neural network, $f_c(\cdot)$ is a transformer encoder \cite{vaswani2017attention} and $g_q(\cdot)$ is a quantization module used to learn targets from the localized variable $\mathbf{v}_{1:T}$.
+\noindent Here, $f_v(\cdot)$ is a convolutional neural network, $f_c(\cdot)$ is a transformer encoder \cite{vaswani_attention_2017} and $g_q(\cdot)$ is a quantization module used to learn targets from the localized variable $\mathbf{v}_{1:T}$.
 Computing quantized targets this way requires an extra loss term, which we will present when we discuss quantization in general below.
@@
 \begin{align}
-    \mathcal{L}_t &= - \log \left(\frac{\exp(S_{\text{c}}(\mathbf{c}_{t}, \mathbf{q}_{t}))}{\sum_{n \sim \mathcal{D}} \exp(S_{\text{c}}(\mathbf{c}_{t}, \mathbf{q}_{n}))} \right) \enspace , \label{w2v2 loss}
+    \mathcal{L}_t &= - \log \left(\frac{\exp(S_{\text{c}}(\mathbf{c}_{t}, \mathbf{q}_{t}))}{\sum_{n \sim \mathcal{D}} \exp(S_{\text{c}}(\mathbf{c}_{t}, \mathbf{q}_{n}))} \right) \enspace , \label{eq_brief: w2v2 loss}
 \end{align}
@@
 
-\noindent Non-autoregressive predictive coding (NPC, \citealp{liu2020non}) tries to resolve this by using a convolutional neural network where the kernel is masked instead of the input.
-%That is, the model is prevented from using $\mathbf{x}_{t}$ for its corresponding reconstruction $\mathbf{\hat{x}}_{t}$.
+\noindent Non-autoregressive predictive coding (NPC, \citealp{liu_nonautoregressive_2020}) tries to resolve this by using a convolutional neural network where the kernel is masked instead of the input.
 This allows for complete data utilization, but limits the amount of context encoded in the learned representation.
@@
     \setlength\tabcolsep{1.5pt}
-    \begin{tabular}{>{\centering\arraybackslash} m{4mm}  >{\centering\arraybackslash} m{0.45\textwidth}|>{\centering\arraybackslash} m{0.45\textwidth}}
+    \begin{tabular}{>{\centering\arraybackslash} m{4mm}  >{\centering\arraybackslash} m{0.40\textwidth}|>{\centering\arraybackslash} m{0.40\textwidth}}
           & {\small \textsc{predictive}} & {\small \textsc{masking-based}} \\
-        %\midrule
-        \rotatebox{90}{{\small \textsc{reconstruct}}} & \includegraphics[width=0.45\textwidth]{paper_brief/REC_PRD.pdf} & \includegraphics[width=0.45\textwidth]{paper_brief/REC_MSK.pdf}  \\
+        \rotatebox{90}{{\small \textsc{reconstruct}}} & \includegraphics[width=0.40\textwidth]{paper_brief/REC_PRD.pdf} & \includegraphics[width=0.40\textwidth]{paper_brief/REC_MSK.pdf}  \\
         \midrule
-        \rotatebox{90}{{\small \textsc{contrastive}}} & \includegraphics[width=0.45\textwidth]{paper_brief/CON_PRD.pdf} & \includegraphics[width=0.45\textwidth]{paper_brief/CON_MSK.pdf}
+        \rotatebox{90}{{\small \textsc{contrastive}}} & \includegraphics[width=0.40\textwidth]{paper_brief/CON_PRD.pdf} & \includegraphics[width=0.40\textwidth]{paper_brief/CON_MSK.pdf}
     \end{tabular}
-    \caption{
-    Schematic of self-supervised methods. Each subfigure illustrates the loss computation for a single time-step.
-    The temporal subscript has been left out for simplicity. 
-    %Examples of models in each category include APC (\textbf{\textsc{prd}} + \textbf{\textsc{rec}}), Mockingjay (\textbf{\textsc{msk}} + \textbf{\textsc{rec}}), CPC (\textbf{\textsc{prd}} + \textbf{\textsc{con}}) and wav2vec 2.0 (\textbf{\textsc{msk}} + \textbf{\textsc{con}}).
-    }
+    \caption{Schematic of self-supervised methods. Each subfigure illustrates the loss computation for a single time-step. The temporal subscript has been left out for simplicity.}
     \label{fig:ssl_grid}
@@
             \midrule
-
             %                                                           MSK      PRD      CON      REC      QTZ      GEN      LOC      GLB     VAR      FRZ      FTN
-            \textbf{Audio Word2vec} \footnotesize\cite{chung2016audio}      & 2016 Mar. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
-            \textbf{Speech2Vec} \footnotesize\cite{chung2018speech2vec}     & 2018 Mar. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
-            \textbf{Unspeech} \footnotesize\cite{milde2018unspeech}         & 2018 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{Audio Word2vec} \footnotesize\cite{chung_audio_2016}      & 2016 Mar. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{Speech2Vec} \footnotesize\cite{chung_speech2vec_2018}     & 2018 Mar. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{Unspeech} \footnotesize\cite{milde_unspeech_2018}         & 2018 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
 
@@
 
-            %PASE \footnotesize\cite{Pascual2019}                   & 2019 Apr. & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 6/4
-            \textbf{APC} \footnotesize\cite{chung2019unsupervised}          & 2019 Oct. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 5/4
-            \textbf{wav2vec} \footnotesize\cite{schneider2019wav2vec}       & 2019 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/4
-
-            %VQ-APC \footnotesize\cite{chung2020vector}             & 2020 May  & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-
-            \textbf{Mockingjay}  \footnotesize\cite{liu2020mockingjay}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 25/10-19
-            %DeCoAR \footnotesize\cite{ling2020deep}                & 2019 Dec. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 3/12-19
-            %PASE+ \footnotesize\cite{Ravanelli2020}                & 2020 Jan. & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 25/1
-            \textbf{wav2vec 2.0} \footnotesize\cite{baevski2020wav2vec}     & 2020 Jun. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 20/6
-            %TERA \footnotesize\cite{liu2021tera}                   & 2021 Jul. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 12/7
-            \textbf{NPC} \footnotesize\cite{liu2020non}                     & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 1/11
-            \textbf{DeCoAR 2.0} \footnotesize\cite{ling2020decoar}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/12
-            \textbf{SCPC} \footnotesize\cite{bhati2021segmental}            & 2021 Jun. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ % 3/6
-            \textbf{HuBERT} \footnotesize\cite{hsu2021hubert}               & 2021 Jun. & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 14/6
-
+            \textbf{APC} \footnotesize\cite{chung_unsupervised_2019}          & 2019 Oct. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 5/4
+            \textbf{wav2vec} \footnotesize\cite{schneider_wav2vec_2019}       & 2019 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/4
+            \textbf{Mockingjay}  \footnotesize\cite{liu_mockingjay_2020}      & 2019 Oct.
+& \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 25/10-19
+            \textbf{wav2vec 2.0} \footnotesize\cite{baevski_wav2vec_2020}     & 2020 Jun. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 20/6
+            \textbf{NPC} \footnotesize\cite{liu_nonautoregressive_2020}                     & 2020 Nov.
+& \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 1/11
+            \textbf{DeCoAR 2.0} \footnotesize\cite{ling_decoar_2020}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/12
+            \textbf{SCPC} \footnotesize\cite{bhati_segmental_2021}            & 2021 Jun. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ % 3/6
+            \textbf{HuBERT} \footnotesize\cite{hsu_hubert_2021}               & 2021 Jun. & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 14/6
             \midrule
@@
             %                                                           MSK      PRD      CON      REC      QTZ      GEN      LOC      GLO     VAR      FRZ      FTN
-            % Conv. DBN \footnotesize\cite{lee_unsupervised_2009}    & 2009 Jan. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark \\
             \textbf{VRNN} \footnotesize\cite{chung_recurrent_2015}          & 2015 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
@@
 Several models enforce a discrete latent space by quantizing the vector representation (\textbf{\textsc{qtz}}, table \ref{tab:model-taxonomy}).
-The two most popular approaches are the Gumbel-softmax \cite{jang2016categorical, maddison2016concrete} and the quantization used in the VQ-VAE \cite{oord_neural_2018}.
-
-%Several models enforce a discrete latent space by quantizing the vector representation (\textbf{\textsc{qtz}}, table \ref{tab:model-taxonomy}).
-A popular approach for online vector quantization is to use the Gumbel softmax \cite{jang2016categorical, maddison2016concrete}.
-This approach corresponds to approximate differentiable sampling from a categorical distribution.
-Another common approach employs a codebook of learnable representations.
-This approach was popularized with the VQ-VAE \cite{oord_neural_2018}, and thus, we refer to it as the VQ-VAE approach.
-Both approaches are non-differentiable and obtain gradients using the straight-through estimator \cite{bengio2013estimating}.
+The two most popular approaches are the Gumbel-softmax \cite{jang_categorical_2016, maddison_concrete_2017} and the quantization used in the VQ-VAE \cite{oord_neural_2018}.
 
@@
 \end{align}
-% for $i=1,\dots,K$. Here $\tau$ is a temperature parameter and $\mathbf{n}\in\mathbb R^K$ is a random vector with $n_i = -\log(-\log(u_i))$ for $u_i \sim \mathcal{U}(0, 1)$. 
-% As  $\tau \rightarrow 0$, $\mathbf{p}$ approaches a one-hot vector. The Gumbel noise $\mathbf{n}$ is a practical way to sample from the untempered categorical distribution (i.e., $\tau = 1$). 
-% A function $\varphi(\mathbf{p})$ is then used to map $\mathbf{p}$ to a discrete sample which can then be used to obtain the quantized vector (e.g., $\mathbf{q}_t$ in eq.~\ref{w2v2 qtz}) via a codebook lookup.
-As this function is non-differentiable, we have to rely on the straight-through estimator \cite{bengio2013estimating} which assumes that the Jacobian ${\partial \, \varphi}/{\partial \, \mathbf{p}}$ equals the identity matrix.
 for $i=1,\dots,K$. Here $\tau$ is a temperature parameter and $\mathbf{n}\in\mathbb R^K$ is a random vector with $n_i = -\log(-\log(u_i))$ for $u_i \sim \mathcal{U}(0, 1)$.
@@
 $\mathbf{p}$ is mapped to a one-hot vector using a function $\varphi(\cdot)$, such that $\varphi(\mathbf{p})_i = 1$ if $i = \argmax_j p_j$ and $0$ otherwise. 
-As this function is non-differentiable, we must rely on the straight-through gradient estimator \cite{bengio2013estimating} which assumes that the Jacobian ${\partial \, \varphi}/{\partial \, \mathbf{p}}$ equals the identity matrix.
-The one-hot vector can then be used for a codebook lookup to obtain the final quantized vector (e.g., $\mathbf{q}_t$ in eq.~\ref{w2v2 qtz}).
+As this function is non-differentiable, we must rely on the straight-through gradient estimator \cite{bengio_estimating_2013} which assumes that the Jacobian ${\partial \, \varphi}/{\partial \, \mathbf{p}}$ equals the identity matrix.
+The one-hot vector can then be used for a codebook lookup to obtain the final quantized vector (e.g., $\mathbf{q}_t$ in eq.~\ref{eq_brief: w2v2 qtz}).
 
-%We choose $\varphi(\cdot)$ to return a one-hot vector with $\varphi(\mathbf{p})_i = 1$ if $i = \argmax_j p_j$ and $0$ otherwise.
-The one-hot vector can then be used for a codebook lookup.
-For non-differentiable $\varphi(\cdot)$, the gradient is usually taken to be that of the input to $\varphi(\cdot)$.
-That is, we use the straight-through estimator and assume during training that the Jacobian ${\partial \, \varphi}/{\partial \, \mathbf{p}}$ equals the identity matrix.
-%\begin{align}
-%    %\mathbf{v} &= \mathbf{p} + \textsc{sg}(\varphi(\mathbf{p}) - %\mathbf{p}) \, .
-%    \frac{\partial \, \varphi}{\partial \, \mathbf{p}} = \mathbf{I}
-%\end{align}
-%where $\textsc{sg}(\cdot)$ is the \textit{stop gradient} function; an identity function that blocks the gradient from the input. .
-
-The wav2vec 2.0 quantization module (eq.~\ref{w2v2 qtz}) uses the Gumbel softmax. To ensure utilization of codebook vectors, a diversity loss is added to the task specific loss (eq. \ref{w2v2 loss})
+The wav2vec 2.0 quantization module (eq.~\ref{eq_brief: w2v2 qtz}) uses the Gumbel softmax.
+To ensure utilization of codebook vectors, a diversity loss is added to the task specific loss (eq.
+\ref{eq_brief: w2v2 loss})
 %
 \begin{align}
-    % \mathcal{L} =  - H(\widetilde{\mathbf{p}}) \frac{1}{K}\enspace,
     \mathcal{L} =  - H(\widetilde{\mathbf{p}}) / K \enspace,
@@
 As $\arg\min$ is non-differentiable, the straight-through estimator is used as for the Gumbel-softmax. 
-% The straight-through estimator defines the gradient of this operation by setting the gradient of $\mathbf{v}$ wrt. to the loss equal to that of $\mathbf{q}$. 
-%
-% Codebook learning is facilitated by a two-term auxiliary loss similar to classical vector quantization dictionary learning \cite{burton_generalization_1983, soong_vector_1985}.
-Gradients for the codebook vectors are given solely by a vector quantization term, which moves codebook vectors $\mathbf{e}_k$ closer to the non-quantized vectors $\mathbf{v}$.
-A so-called commitment term is added to ensure that non-quantized vectors do not grow unboundedly by enforcing the encoder to keep them close to a codebook vector.
-%
 Codebook learning is facilitated by a two-term auxiliary loss similar to classical vector quantization dictionary learning \cite{burton_generalization_1983, soong_vector_1985}.
@@
 \begin{align}
-    \mathcal{L} = \underset{\text{vq}}{\underbrace{\norm{\text{sg}\left[\mathbf{v}\right] - \mathbf{e}}_2^2}} + \underset{\text{commitment}}{\underbrace{\beta\norm{\mathbf{v} - \text{sg}\left[\mathbf{e}\right]}_2^2}} \enspace , \label{eq: vector quantization losses}
+    \mathcal{L} = \underset{\text{vq}}{\underbrace{\norm{\text{sg}\left[\mathbf{v}\right] - \mathbf{e}}_2^2}} + \underset{\text{commitment}}{\underbrace{\beta\norm{\mathbf{v} - \text{sg}\left[\mathbf{e}\right]}_2^2}} \enspace , \label{eq_brief: vector quantization losses}
 \end{align}
 where $\text{sg}[\mathbf{x}] = \mathbf{x}$ is the stop-gradient operator with the property $\frac{d}{dx_i}\text{sg}[\mathbf{x}] \equiv 0$ for all $i$ and $\beta$ is a hyperparameter.
-Although vector quantization was introduced by the VQ-VAE which is, in some ways, a latent variable model, it has been applied to self-supervised methods \cite{niekerk_vector_2020, baevski2019vq}.
-
-%but with variations between individual approaches.
-The VQ-CPC model \cite{niekerk_vector_2020} quantizes $\mathbf{v}$ in the original CPC (eq.~\ref{eq: cpc local representation}) and augments its loss (eq.~\ref{eq: cpc loss}) by addition of the commitment term in eq.~\ref{eq: vector quantization losses}.
-Instead of learning the codebook via the VQ term, it is updated as a moving average of $\mathbf{v}$ as initially suggested by \citet{oord_neural_2018}.
-%The vq-wav2vec \cite{baevski2019vq} is defined similarly but adds the full auxiliary loss to the wav2vec loss (eq.~\ref{eq: wav2vec loss}) in place of moving averages.
+Although vector quantization was introduced by the VQ-VAE which is, in some ways, a latent variable model, it has been applied to self-supervised methods \cite{vanniekerk_vectorquantized_2020, baevski_vqwav2vec_2020}.
 
 \textit{Motivation:} Similar to how quantization approaches differ between works, so do the motivations provided for employing them.
-The vq-wav2vec \cite{baevski2019vq, baevski2019effectiveness} %quantizes $\mathbf{v}_{1:T}$ before feeding them to the context network $f_c(\cdot)$.
-Here, the motivation is
-learn quantized representations in order to apply natural language processing models, like BERT \cite{devlin2019bert}, afterwards.
-Other works use quantization for speech segmentation \cite{kamper2020towards, chorowski2019unsupervised} or as a bottleneck in order to ``\textit{limit model capacity}" \cite{chung2020vector, ling2020decoar}.
-%Quantized representations have also been used for speech segmentation \cite{kamper2020towards, chorowski2019unsupervised}. 
-Finally, \citet{chung2020vector} explore quantization between different layers in the APC model, but find that continuous representations consistently perform better than their quantized counterparts on a downstream phoneme classification task
-
-%They find that applying quantization to the output $\mathbf{c}_{1:T}$ yields the best results on a downstream phoneme classification task.
-However, the continuous representations consistently perform better than their quantized counterparts.
+The vq-wav2vec \cite{baevski_vqwav2vec_2020, baevski_effectiveness_2020} learn quantized representations in order to apply natural language processing models, like BERT \cite{devlin_bert_2018}, afterwards.
+Other works use quantization for speech segmentation \cite{kamper_unsupervised_2021, chorowski_unsupervised_2019} or as a bottleneck in order to ``\textit{limit model capacity}" \cite{chung_vectorquantized_2020, ling_decoar_2020}.
+Finally, \citet{chung_vectorquantized_2020} explore quantization between different layers in the APC model, but find that continuous representations consistently perform better than their quantized counterparts on a downstream phoneme classification task
 
 Given our previous discussion of how it might not be beneficial to model localized noise, quantization in wav2vec 2.0 seems well motivated, as it enforces the target representation $\mathbf{q}_{1:T}$ to discard such noise.
-Taking this idea further, the HuBERT model \cite{hsu2021hubert} uses offline quantization to learn categorical targets.
+Taking this idea further, the HuBERT model \cite{hsu_hubert_2021} uses offline quantization to learn categorical targets.
 Initially, spectrogram features are used to learn frame-wise labels with $k$-means clustering.
@@
 Now, we cover models that learn global representations (\textbf{\textsc{glb}}, table \ref{tab:model-taxonomy}).
-%Apart from this attribute, we find that models that learn global representations rely on many of the same techniques described in the previous sections.
-%We first give a broad overview, before discussing how they relate to the self-supervised models.
 
-Early work on global speech representation learning takes inspiration from the autoencoder framework \cite{kramer1991nonlinear}.
-\citet{chung2016audio} propose a simple sequence-to-sequence autoencoder for learning acoustic word embeddings:
-%
-% i/x-vectors here?
+Early work on global speech representation learning takes inspiration from the autoencoder framework \cite{kramer_nonlinear_1991}.
+\citet{chung_audio_2016} propose a simple sequence-to-sequence autoencoder for learning acoustic word embeddings:
 \begin{align}
     \mathbf{c} &= f(\mathbf{x}_{1:T}) \\
-    \hat{\mathbf{x}}_{1:T} &= g(\mathbf{c}) \enspace , \label{eq:dec-aw2v}
+    \hat{\mathbf{x}}_{1:T} &= g(\mathbf{c}) \enspace , \label{eq_brief:dec-aw2v}
 \end{align}
@@
 \noindent where $f(\cdot)$ and $g(\cdot)$ are a recurrent neural networks, such that $\mathbf{c}$ is taken to be the hidden state at the last time-step $T$ of $f(\cdot)$ and used as initial hidden state of $g(\cdot)$.
-%where $f(\cdot)$ is a recurrent neural network encoder, such that $\mathbf{c}$ is taken to be the hidden state at the last time-step $T$.
-The decoder, $g(\cdot)$, is also a recurrent neural network where $\mathbf{c}$ is used to initialize the hidden state.
 The authors also propose a denoising autoencoder with masked inputs $f(\mathbf{x}_{1:T} \circ \mathbf{m}_{1:T})$.
-Similar RNN-based autoencoders have also been explored \cite{kamper2019truly, holzenberger2018learning}.
-%
-%\begin{align}
-%    \mathbf{c} &= f(\mathbf{x}_{1:T} \circ \mathbf{m}_{1:T})
-%\end{align}
-%
+Similar RNN-based autoencoders have also been explored \cite{kamper_truly_2019, holzenberger_learning_2018}.
 
-Prior to this work, \citet{kamper2015unsupervised} and \citet{renshaw2015comparison} introduced the \textit{correspondence autoencoder}.
+Prior to this work, \citet{kamper_unsupervised_2015} and \citet{renshaw_comparison_2015} introduced the \textit{correspondence autoencoder}.
 This method uses dynamic time warping to align input-target segment pairs extracted with unsupervised term discovery.
-In more recent work, the need for alignment has been alleviated by adopting the sequence-to-sequence framework \cite{kamper2019truly, jacobs2021acoustic}.
+In more recent work, the need for alignment has been alleviated by adopting the sequence-to-sequence framework \cite{kamper_truly_2019, jacobs_acoustic_2021}.
 
-%This method relies on unsupervised term discovery with randomized algorithms \cite{jansen2011efficient} to extract corresponding speech segment pairs for training.
-Thus, one segment is used as input and the other as reconstruction target.
-Because the paired segments do not necessarily have the same length, they need to be aligned with dynamic time warping.
-
-% In more recent work, this need has been alleviated by adopting the sequence-to-sequence framework \cite{kamper2019truly, jacobs2021acoustic}.
-%Unlike the reconstruction tasks reviewed in the previous sections, models presented here all rely on an $\ell_2$ loss, $\mathcal{L} = \lVert \hat{\mathbf{x}}_{1:T} - \mathbf{x}_{1:T} \rVert_2\enspace$.
-%
-% \begin{align}
-%     \mathcal{L} &= \lVert \hat{\mathbf{x}}_{1:T} - \mathbf{x}_{1:T} \rVert_2\enspace.
-% \end{align}
-
-Inspired by the work on semantic word embeddings for text \cite{mikolov2013distributed}, the sequence-to-sequence framework has also been used to implement speech-based versions of the skipgram and continuous bag-of-words models \cite{chung2017learning, chung2018speech2vec}.
+Inspired by the work on semantic word embeddings for text \cite{mikolov_distributed_2013}, the sequence-to-sequence framework has also been used to implement speech-based versions of the skipgram and continuous bag-of-words models \cite{chung_learning_2017, chung_speech2vec_2018}.
 Given a segment corresponding to a single word $\mathbf{x}_{(n)} = \mathbf{x}_{t_{n}:t_{n+1}}$, the skipgram model is trained to predict neighboring words $\mathbf{x}_{(n + k)}$ where $k\ne0$.
 That is, instead of a single decoder, as in eq.
-\ref{eq:dec-aw2v}, the skipgram model employs multiple decoders
+\ref{eq_brief:dec-aw2v}, the skipgram model employs multiple decoders
 %
@@
     \mathbf{c} &= \sum_{k\in\mathcal{K}} f_k(\mathbf{x}_{(n + k)}) %\enspace .
-    % \mathbf{c} &= \sum_{k\in\mathcal K_n} \mathbf{c}_k
 \end{align}
-%for different values of $k$.
 The sequence-to-sequence models described above rely on speech segments corresponding to words.
-The segments are obtained by supervised forced alignment, %Thus, since forced alignment requires labeled data, the models might be seen as weakly supervised
-but similar models have been explored without this requirement \cite{jati2017speaker2vec, tagliasacchi2020pre}.
+The segments are obtained by supervised forced alignment, but similar models have been explored without this requirement \cite{jati_speaker2vec_2017, tagliasacchi_pretraining_2020}.
 
-Contrastive learning has also been explored for global speech representation learning \citep{milde2018unspeech, jati2019neural, jansen2018unsupervised}.
-And prior to the widespread adoption of neural networks, \citet{levin2013fixed} explore principal component analysis and Laplacian eigenmaps for learning fixed-sized acoustic embeddings.
-%Comparable to the skipgram model described above, \citet{milde2018unspeech} and \citet{jati2019neural} define a binary contrastive task where a target segment is used to predict neighboring segments.
+Contrastive learning has also been explored for global speech representation learning \citep{milde_unspeech_2018, jati_neural_2019, jansen_unsupervised_2018}.
+And prior to the widespread adoption of neural networks, \citet{levin_fixeddimensional_2013} explore principal component analysis and Laplacian eigenmaps for learning fixed-sized acoustic embeddings.
 
-%For general audio representation learning, \citet{jansen2018unsupervised} explore multiple pre-training tasks using a contrastive triplet loss.
-And prior to the widespread adoption of neural networks, \citet{levin2013fixed} explore principal component analysis and Laplacian eigenmaps for learning fixed-sized acoustic embeddings.
-
-\paragraph{Other work} Some models learn local representations with a variable temporal resolution that is not proportional to the input resolution (\textbf{\textsc{var}}, table \ref{tab:model-taxonomy}).
-In practice, this is often achieved implicitly, by learning segment boundaries or by taking repeated quantized values to belong to the same segment \cite{kamper2020towards, chorowski2019unsupervised, michel2017blind, kreuk2020self, wang2017gate, dieleman_variable-rate_2021}.
-An exception is the recently proposed \emph{segmental contrastive predictive coding} (SCPC, \citealp{bhati2021segmental, bhati2021unsupervised}).
+\paragraph{Other work}
+Some models learn local representations with a variable temporal resolution that is not proportional to the input resolution (\textbf{\textsc{var}}, table \ref{tab:model-taxonomy}).
+In practice, this is often achieved implicitly, by learning segment boundaries or by taking repeated quantized values to belong to the same segment \cite{kamper_unsupervised_2021, chorowski_unsupervised_2019, michel_blind_2017, kreuk_selfsupervised_2020, wang_gate_2017, dieleman_variablerate_2021}.
+An exception is the recently proposed \emph{segmental contrastive predictive coding} (SCPC, \citealp{bhati_segmental_2021, bhati_unsupervised_2022}).
 With this approach, the model explicitly learns segment boundaries, which are used to downsample the representations during training.
-The same segmentation strategy has subsequently been applied in other models \cite{cuervo2021contrastive}.
+The same segmentation strategy has subsequently been applied in other models \cite{cuervo_contrastive_2022}.
 
 Most of the work presented so far fits neatly into the taxonomy presented in table \ref{tab:model-taxonomy}.
-One exception is the problem-agnostic speech encoder (PASE, \citealp{Pascual2019, Ravanelli2020}) that combines multiple pre-training tasks.
+One exception is the problem-agnostic speech encoder (PASE, \citealp{pascual_learning_2019, ravanelli_multitask_2020}) that combines multiple pre-training tasks.
 Furthermore, many of the presented models have been successfully applied to other use cases.
-For instance, wav2vec 2.0 and related models have been applied to learn cross-lingual and multi-lingual representations \cite{riviere2020unsupervised, conneau2020unsupervised, khurana2021magic} and proven well-suited for concurrently learning with labeled data \cite{talnikar2021joint, wang2021unispeech}.
+For instance, wav2vec 2.0 and related models have been applied to learn cross-lingual and multi-lingual representations \cite{riviere_unsupervised_2020, conneau_unsupervised_2020, khurana_magic_2021} and proven well-suited for concurrently learning with labeled data \cite{talnikar_joint_2021, wang_unispeech_2021}.
 %We will discuss the downstream application of self-supervised representations further in section \ref{sec:eval}.
@@
 We disregard any specific temporal notation without loss of generality.
-We then introduce the variational autoencoder framework (VAE, \citealp{kingma_auto-encoding_2014}).
+We then introduce the variational autoencoder framework (VAE, \citealp{kingma_autoencoding_2014}).
 We focus on different dependency structures between data and learned representations, in contrast to the more practical view on self-supervised models taken above.
@@
     \log p(\mathbf{x}) = \log \int p(\mathbf{x}|\mathbf{z}) p(\mathbf{z}) \,\text{d}\mathbf{z} \enspace.
-    \label{eq: lvm log-likelihood}
+    \label{eq_brief: lvm log-likelihood}
 \end{equation}
@@
 
-For representation learning, LVMs are commonly defined using the VAE framework \cite{kingma_auto-encoding_2014, rezende_stochastic_2014}) which is also the focus of our exposition. 
+For representation learning, LVMs are commonly defined using the VAE framework \cite{kingma_autoencoding_2014, rezende_stochastic_2014} which is also the focus of our exposition. 
 In the VAE framework, the observation model $p(\mathbf{x}|\mathbf{z})$ is parameterized using a deep neural network.
-This choice allows modeling complex and high-dimensional data but also makes the integral in eq.~\ref{eq: lvm log-likelihood} analytically intractable.
+This choice allows modeling complex and high-dimensional data but also makes the integral in eq.~\ref{eq_brief: lvm log-likelihood} analytically intractable.
 MCMC methods can be used to estimate it and the true model posterior $p(\mathbf{z}|\mathbf{x})$, but these methods are usually computationally expensive in this setting \cite{mohamed_monte_2019}.
@@
 It approximates the intractable true model posterior by introducing a variational posterior distribution $q(\mathbf{z}|\mathbf{x})$, also parameterized by a deep neural network.
-From eq.~\ref{eq: lvm log-likelihood}, via Jensen's inequality, this gives rise to a variational lower bound on the likelihood, also known as the evidence lower bound (ELBO).
+From eq.~\ref{eq_brief: lvm log-likelihood}, via Jensen's inequality, this gives rise to a variational lower bound on the likelihood, also known as the evidence lower bound (ELBO).
 \begin{equation}
     \log p(\mathbf{x}) \geq \int q(\mathbf{z}|\mathbf{x}) \log \frac{p(\mathbf{x}| \mathbf{z})p(\mathbf{z})}{q(\mathbf{z}|\mathbf{x})} \,\text{d}\mathbf{z} \equiv \mathcal{L}_{\text{ELBO}} \enspace .
-\label{eq: lvm likelihood bound (elbo)}
+\label{eq_brief: lvm likelihood bound (elbo)}
 \end{equation}
 The bound can be efficiently evaluated and optimized with Monte Carlo (MC) estimation by sampling from $q(\mathbf{z}|\mathbf{x})$.
-Low-variance gradient estimates are usually obtained via reparameterization of $q(\mathbf{z}|\mathbf{x})$ \cite{kingma_auto-encoding_2014} although alternatives exist (e.g., inverse CDF sampling) \cite{mohamed_monte_2019}.
+Low-variance gradient estimates are usually obtained via reparameterization of $q(\mathbf{z}|\mathbf{x})$ \cite{kingma_autoencoding_2014} although alternatives exist (e.g., inverse CDF sampling) \cite{mohamed_monte_2019}.
 The ELBO can also be written as
 \begin{equation}
-    \mathcal{L}_{\text{ELBO}} = \mathbb{E}_{q(\mathbf{z}|\mathbf{x})}\left[ \log p(\mathbf{x}|\mathbf{z}) \right] - D_\text{KL}\left( q(\mathbf{z}|\mathbf{x}) || p(\mathbf{z}) \right) \enspace  \label{eq: lvm likelihood bound recon/kl form (elbo)},
+    \mathcal{L}_{\text{ELBO}} = \mathbb{E}_{q(\mathbf{z}|\mathbf{x})}\left[ \log p(\mathbf{x}|\mathbf{z}) \right] - D_\text{KL}\left( q(\mathbf{z}|\mathbf{x}) || p(\mathbf{z}) \right) \enspace  \label{eq_brief: lvm likelihood bound recon/kl form (elbo)},
 \end{equation}
@@
 In brief, LVMs of the VAE-type consist of a approximate posterior, $q(\mathbf{z}|\mathbf{x})$, an observation model, $p(\mathbf{x}|\mathbf{z})$, and a prior, $p(\mathbf{z})$. 
-% The joint observation model and prior form the generative model which can be efficiently sampled with ancestral sampling which entails first sampling a latent variable $\tilde{\mathbf{z}}\sim p(\mathbf{z})$ followed by sampling the observed variable conditioned on that latent variable, $\tilde{\mathbf{x}}\sim p(\mathbf{x}|\tilde{\mathbf{z}})$.
-With reference to probabilistic coding theory, the approximate posterior is often referred to as the encoder and the observation model as the decoder \cite{kingma_auto-encoding_2014, rezende_stochastic_2014}.
+With reference to probabilistic coding theory, the approximate posterior is often referred to as the encoder and the observation model as the decoder \cite{kingma_autoencoding_2014, rezende_stochastic_2014}.
 From a theoretical perspective, the encoder exists solely as the result of choosing to use variational inference to train the decoder rather than e.g.
@@
 \begin{table}[t!]
-\caption{
-A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable. 
+\caption{A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable.
 The observation, prior and inference models may all belong to one or more of the categories listed under them as detailed in section \ref{sec:plvms}.
-The types listed here serve as primitives from which more complex structures can be constructed including models with hierarchies of multiple latent variables. 
-% We indicate autoregressiveness (\textbf{\textsc{arx}} and \textbf{\textsc{arz}}) using a ``catch-all" notation $*$, e.g. $\mathbf{x}_{*:t-1}$. 
-% This serves to indicate that autoregressive dependencies can have different span including at the extremes the full sequence $\mathbf{x}_{1:t-1}$ and the last value $\mathbf{x}_{t-1}$.
-% We have not explicitly listed any cases with multiple latent variables as these generally include a multitude of potential dependency structures.
-% We outline notable examples of models with hierarchies of latent variables in section \ref{sec:plvms}.
-}
+The types listed here serve as primitives from which more complex structures can be constructed including models with hierarchies of multiple latent variables.}
 \label{tab:lvm-model-primitives}
@@
         \midrule
-        % \textbf{\textsc{aut}} & Autoregressive on $\mathbf{z}_t$      & $p(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ \\
-        % \textbf{\textsc{obs}} & Observation dependent                                & $p(\mathbf{z}_t|\mathbf{x}_{1:t-1})$ \\
         \textbf{\textsc{arx}} & Autoregressive on $\mathbf{x}_t$      & $p(\mathbf{z}_t|\mathbf{x}_{1:t-1})$ \\
@@
 \begin{sidewaystable*}[t]
-    \caption{
-    Selected latent variable models classified according the attributes defined throughout section \ref{sec:plvms}.
-    %The models are sorted according to the publication date of the first version on arXiv which might differ from the year indicated by the citation. 
-    %This table supplements the classification provided in table \ref{tab:model-taxonomy} which uses binary attributes largely defined for self-supervised methods. 
-    See table~\ref{tab:lvm-model-primitives} for the probability distributions that correspond to each of the attribute short-hands. \textbf{\textsc{hie}} indicates a hierarchical representation.
-    %The mathematical expressions of the observation, prior and inference models for the models considered here can be found in table~\ref{tab:lvm-mathematical-expressions}. 
-    %Figure~\ref{fig:lvm-graphical-models} depicts corresponding graphical model for selected LVMs of this table.
-    }
+    \caption{Selected latent variable models classified according the attributes defined throughout section \ref{sec:plvms}.
+See table~\ref{tab:lvm-model-primitives} for the probability distributions that correspond to each of the attribute short-hands.
+\textbf{\textsc{hie}} indicates a hierarchical representation.}
     \label{tab:lvm-taxonomy}
@@
             %                                                         ARX      LOC      GLB      ARX      ARZ     IND      GLB      ARZ      FLT      LSM      GSM      GLB
-
-            % Conv. DBN \cite{lee_unsupervised_2009}    & 2009 Jan. & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\
             \textbf{VRNN} \footnotesize\cite{chung_recurrent_2015}          & 2015 Jun.
@@
 & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark \\
-            % WaveNet AE        &  \\
-            % HMM-VAE           &  \\
-            % Convolutional DBN &  \\
             \bottomrule
@@
 We next review VAEs applied to speech. We consider the choices of observation, prior and inference models.
-% and introduce attributes throughout the text. An overview of these is found in table~\ref{tab:lvm-model-primitives}.
 We provide a model taxonomy for selected LVMs in table~\ref{tab:lvm-taxonomy}.
-% Supplementing the binary attributes, table~\ref{tab:lvm-mathematical-expressions} explicitly details the observation, prior and inference models of the same selection of LVMs considered in table~\ref{tab:lvm-taxonomy}.
 
 \paragraph{Observation models} 
-A common choice for the observation model $p(\mathbf{x}|\mathbf{z})$ is to include an autoregressive dependency on the observed variable (\textbf{\textsc{arx}}, table \ref{tab:lvm-model-primitives}) that is, $p(\mathbf{x}_t|\mathbf{x}_{1:t-1}, \cdot)$ where $\cdot$ represents some dependency on the latent variable \cite{chung_recurrent_2015, fraccaro_sequential_2016, oord_neural_2018, oord2018representation}.
-This allows the latent representation to focus on correlations that cannot easily be predicted from the observed variable at previous time-steps \cite{oord2018representation}.
+A common choice for the observation model $p(\mathbf{x}|\mathbf{z})$ is to include an autoregressive dependency on the observed variable (\textbf{\textsc{arx}}, table \ref{tab:lvm-model-primitives}) that is, $p(\mathbf{x}_t|\mathbf{x}_{1:t-1}, \cdot)$ where $\cdot$ represents some dependency on the latent variable \cite{chung_recurrent_2015, fraccaro_sequential_2016, oord_neural_2018, oord_representation_2018}.
+This allows the latent representation to focus on correlations that cannot easily be predicted from the observed variable at previous time-steps \cite{oord_representation_2018}.
 In practice, the dependency on $\mathbf{x}_{1:t-1}$ is often assumed to be Markovian and hence only on $\mathbf{x}_{t-1}$.
@@
 Specifically encouraging the latent representation to discard correlations across the temporal dimension might degrade the quality of the latent representation.
-Furthermore, since such a decoder can perform quite well by simply solving an autoregressive prediction problem, similar to WaveNet \cite{oord_wavenet:_2016}, it can make the model prone to suffer from posterior collapse.
+Furthermore, since such a decoder can perform quite well by simply solving an autoregressive prediction problem, similar to WaveNet \cite{oord_wavenet_2016}, it can make the model prone to suffer from posterior collapse.
 This problem arises when the approximate and true posterior distributions collapse into the prior which renders the representations non-informative \cite{bowman_generating_2016, sonderby_ladder_2016}.
@@
 Some works alleviate this problem with tricks like KL-annealing and free bits \cite{bowman_generating_2016, sonderby_ladder_2016, kingma_improved_2016}. 
-%KL-annealing down-weighs the KL-divergence in the ELBO (eq.~\ref{eq: lvm likelihood bound recon/kl form (elbo)}) gradually phasing it in during an initial training phase \cite{bowman_generating_2016, sonderby_ladder_2016}.
-Free bits simply returns a zero gradient for KL-divergences smaller than some number.
-\cite{kingma_improved_2016}.
 The VQ-VAE uses a quantized latent space that is not susceptible to posterior collapse \textit{per se} \cite{oord_neural_2018}.
@@
 However, this forces the latent variable to encode details about the observed variable to achieve a good reconstruction.
-This is opposite to contrastive self-supervised learning which allows models to discard details in $\mathbf{x}_{1:T}$ that do not inform the training objective \cite{baevski2020wav2vec}.
+This is opposite to contrastive self-supervised learning which allows models to discard details in $\mathbf{x}_{1:T}$ that do not inform the training objective \cite{baevski_wav2vec_2020}.
 
@@
 Priors can be said to belong to one or more of four broad categories. See table~\ref{tab:lvm-model-primitives}. 
-% Autoregressive dependency on the observed variable (\textbf{\textsc{arx}}, table \ref{tab:lvm-model-primitives}), autoregressive dependency on the latent variable (\textbf{\textsc{arz}}, table \ref{tab:lvm-model-primitives}), locally independent (\textbf{\textsc{ind}}, table \ref{tab:lvm-model-primitives}) and global (\textbf{\textsc{glb}}, table \ref{tab:lvm-model-primitives}).
 Priors that are autoregressive on the observed variable (\textbf{\textsc{arx}}, table \ref{tab:lvm-model-primitives}) take the form $p(\mathbf{z}_t|\mathbf{x}_{1:t-1})$.
 This generally results in a slow-down of the generative process which may be of concern if the use-case is data generation.
-%It is also interesting to note that this encourages the latent variables to model less dynamic behaviour in their stochastic transitions and instead rely more on the observed variable.
 Priors that are autoregressive on the latent variable (\textbf{\textsc{arz}}, table \ref{tab:lvm-model-primitives}) take the form $p(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ and enable stochastic temporal transitions similar to hidden Markov models but with potentially nonlinear transition functions \cite{chung_recurrent_2015, fraccaro_sequential_2016, khurana_factorial_2019, khurana_convolutional_2020}.
@@
 Hence, models usually use them in combination with another local latent variable, or to encode fixed length input segments \cite{khurana_factorial_2019, hsu_learning_2017, hsu_unsupervised_2017}.
-
-% Many LVMs have priors that do not have autoregressive dependence on the observed variable (table \ref{tab:lvm-model-primitives}).
-This allows generating the latent sequence in full before generating the observed sequence.
-It also permits conditioning past observed variables on future latent variables.
-Despite this, it is usual to let $\mathbf{x}_t$ depend causally on $\mathbf{z}_{1:t}$ in the generative model.
-To our knowledge, so far no work has examined full (non-causal) dependence of $\mathbf{x}_{1:T}$ on $\mathbf{z}_{1:T}$ during generation.% where, for example, $\mathbf{z}_T$ can influence $\mathbf{x}_1$.
 
@@
 $q(\mathbf{z}_t|\mathbf{z}_{1:t-1},\mathbf{x}_{1:t})$ (\textbf{\textsc{arz}}, table \ref{tab:lvm-model-primitives}) \cite{chung_recurrent_2015, fraccaro_sequential_2016}.
-% Finally, latent variables are often made to depend autoregressively on past values, $q(\mathbf{z}_t|\mathbf{z}_{1:t-1},\mathbf{x}_{1:t})$ for filtering and $q(\mathbf{z}_t|\mathbf{z}_{1:t-1},\mathbf{x}_{1:T})$ for smoothing (\textbf{\textsc{arz}}, table \ref{tab:lvm-model-primitives}) \cite{chung_recurrent_2015, fraccaro_sequential_2016}.
-%Such autoregressive dependencies help the inference model better match the prior dynamics but notably are not employed in models that learn the prior dynamics post-hoc \cite{oord_neural_2018}.
 
-\paragraph{Multiscale and hierarchical models} Some work has explored using a hierarchy of latent variables (\textbf{\textsc{hie}}, table \ref{tab:lvm-taxonomy}).
+\paragraph{Multiscale and hierarchical models} 
+Some work has explored using a hierarchy of latent variables (\textbf{\textsc{hie}}, table \ref{tab:lvm-taxonomy}).
 This allows encoding the inductive bias that speech contains information at different temporal scales by letting the latent variables operate at different temporal scales \cite{hsu_unsupervised_2017}.
@@
 Recent work has focused on learning a deeper latent hierarchy with five latent variables \cite{aksan_stcn_2019}.
-%Here, the latents have no autoregressive temporal dependencies but instead depend on latent variables further up in the hierarchy with all latent variables operating at the same temporal scale. 
-%Interestingly, this effort within LVMs towards separating learned features into representations at different time scales or learning hierarchical representations stands in contrast to work on self-supervised learning where focus is either on learning local or global representations.
 
 \paragraph{Other work}
-% Most recent works learn latent variable models via amortized variational inference as introduced by the VAE framework \cite{kingma_auto-encoding_2014, rezende_stochastic_2014}. 
-Before the introduction of the VAE, models such as deep belief networks (DBN, \citealp{hinton_fast_2006}) built from stacks of restricted Boltzmann machines \cite{smolensky_parallel_1987,fischer:13} were popular.
+Before the introduction of the VAE, models such as deep belief networks (DBN, \citealp{hinton_fast_2006}) built from stacks of restricted Boltzmann machines \cite{smolensky_chapter_1986,fischer_training_2014} were popular.
 \citet{lee_unsupervised_2009} show the feasibility of using a two-layered DBN for discovering acoustic units of speech, while \citet{deng_binary_2010} show that a DBN can learn a binary coding of spectrograms that has higher signal-to-noise ratio than classical vector-quantization techniques for speech coding.
-DBNs are however notoriously tricky to optimize requiring the use of expensive MCMC sampling techniques for inference or resort to biased gradient estimates \cite{hinton_practical_2012,fischer:10c}.
+DBNs are however notoriously tricky to optimize requiring the use of expensive MCMC sampling techniques for inference or resort to biased gradient estimates \cite{montavon_practical_2012,fischer_bounding_2011}.
 Non-neural LVMs for speech representation learning have also been explored
@@
 
-% Non-neural approaches include Bayesian non-parametric models which can be used for unsupervised acoustic unit discovery.
-This is often done using a Dirichlet process prior which has the appealing property of automatically inferring the number of acoustic units \cite{lee_nonparametric_2012, ondel_variational_2016, heck_feature_2017}.
-% Finally, Gaussian mixture models have been used to form so-called universal background models from which acoustic representations can be extracted \cite{jansen_weak_2013}.
-
 \section{Discussion}
 \label{sec:mtax}
-%Throughout the text we defined several binary attributes which are used to characterize a selection of models in table \ref{tab:model-taxonomy} and \ref{tab:lvm-taxonomy}.
-We discuss representation learning models based on this overarching model taxonomy below.
-
-%In table \ref{tab:model-taxonomy}, we use those derived from the self-supervised models to characterize a selection of models within the two surveyed categories.
-To obtain a more detailed understanding of the LVMs, we also derived a set of variable dependencies  presented  in table \ref{tab:lvm-model-primitives}.
-In table \ref{tab:lvm-taxonomy} we provide a specification of how they apply to selected LVMs.
-
-% Throughout the text we defined a set of binary model attributes.
-In table \ref{tab:model-taxonomy}, we use these to characterize a selection of models within the two surveyed categories.
-These attributes have been derived with a focus on the self-supervised models.
-As a result, the characterization of the latent variable models is fairly uniform in comparison.
-Thus, to obtain a more detailed understanding, we also derived a set of dependencies between variables used to specify the graphical models for the LVMs.
-These are presented in table \ref{tab:lvm-model-primitives} and a specification of how they apply to selected models is given in table \ref{tab:lvm-taxonomy}.
-Below, we compare models included in this survey using table \ref{tab:model-taxonomy} and \ref{tab:lvm-taxonomy} as a basis for the discussion.
 
 \paragraph{From global to local} 
-
 In table \ref{tab:model-taxonomy}, we see that work on global representations within self-supervised learning precedes work on local representations.
-However, we find that the core ideas underlying the recent successes in learning local representation models have also been used for global representation learning; masking \cite{chung2016audio}, context prediction \cite{chung2018speech2vec}, and contrastive training \cite{milde2018unspeech} have been applied in both settings.
-Furthermore, where work on global representation learning has taken inspiration from Word2vec \cite{mikolov2013distributed}, the techniques used for learning local representations are inspired by contextualized word embeddings \cite{devlin2019bert}.
+However, we find that the core ideas underlying the recent successes in learning local representation models have also been used for global representation learning; masking \cite{chung_audio_2016}, context prediction \cite{chung_speech2vec_2018}, and contrastive training \cite{milde_unspeech_2018} have been applied in both settings.
+Furthermore, where work on global representation learning has taken inspiration from Word2vec \cite{mikolov_distributed_2013}, the techniques used for learning local representations are inspired by contextualized word embeddings \cite{devlin_bert_2018}.
 Thus, the gap between these two model classes is largely a product of the developments in related fields and the general increase in computational resources.
 
-\paragraph{Representations beyond inference} Predictive tasks are commonly used for self-supervised models, but they are not directly compatible with LVM training. 
+\paragraph{Representations beyond inference} 
+Predictive tasks are commonly used for self-supervised models, but they are not directly compatible with LVM training. 
 However, an LVM prior with an autoregressive parameterization, $p(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ or $p(\mathbf{z}_t|\mathbf{x}_{1:t-1})$, can be seen as predictive in the sense that it tries to match the approximate posterior.
-%, which conditions on $\mathbf{x}_t$, using only latent dynamics and past observed variables. 
 Hence, the prior might be considered for feature extraction.
-%although it remains unclear whether a representation extracted from the prior is better suited for downstream tasks than one extracted from the posterior.
-\citet{jones2020discrete} examine the importance of the prior in the VQ-VAE and show that the ability of this model to estimate densities $p(\mathbf{x}_{1:T})$ lies solely in its prior.
+\citet{jones_is_2020} examine the importance of the prior in the VQ-VAE and show that the ability of this model to estimate densities $p(\mathbf{x}_{1:T})$ lies solely in its prior.
 Other work has also explored representations beyond the latent variable such as hidden units of the observation model \cite{khurana_convolutional_2020, chorowski_unsupervised_2019}.
-
-% Hidden units from the observation model have also been used as downstream task features in previous work \cite{khurana_convolutional_2020}.
-Similarly, \citet{chorowski_unsupervised_2019} explore representations beyond the latent variable.
 
@@
 
-The SUPERB benchmark \cite{yang2021superb} gathers multiple tasks grouped into categories such as \emph{recognition}, \emph{detection}, \emph{semantics}, \emph{speaker}, \emph{paralinguistics} and \emph{generation}.
-The recently proposed SLUE benchmark focuses on spoken language understanding \cite{shon2021slue}.
-The long-standing zero resource speech challenge (ZeroSpeech) offers a new set of tasks for each edition \cite{versteegh2015zero, dunbar2017zero, dunbar2019zero, dunbar2020zero, dunbar2021zero} usually featuring a minimal-pair ABX task \cite{schatz2013evaluating, schatz2014evaluating}.
+The SUPERB benchmark \cite{yang_superb_2021} gathers multiple tasks grouped into categories such as \emph{recognition}, \emph{detection}, \emph{semantics}, \emph{speaker}, \emph{paralinguistics} and \emph{generation}.
+The recently proposed SLUE benchmark focuses on spoken language understanding \cite{shon_slue_2021}.
+The long-standing zero resource speech challenge (ZeroSpeech) offers a new set of tasks for each edition \cite{versteegh_zero_2015, dunbar_zero_2017, dunbar_zero_2019, dunbar_zero_2020, dunbar_zero_2021} usually featuring a minimal-pair ABX task \cite{schatz_evaluating_2013, schatz_evaluating_2014}.
 
-Tasks that evaluate representations in terms of speaker-related information include speaker verification \cite{hsu_unsupervised_2017, khurana_factorial_2019, milde2018unspeech}, speaker identification \cite{oord2018representation, jati2019neural, chung2019unsupervised, liu2020non}, dialect classification \cite{khurana_factorial_2019}, emotion recognition \cite{Pascual2019, yang2021superb} and gender classification \cite{lee_unsupervised_2009}.
-The semantic content of representations are evaluated using tasks such as intent classification \cite{morais2021end, yang2021superb}, slot filling \cite{lai2021semi, yang2021superb}, sentiment analysis \cite{liu2020mockingjay}, question answering \cite{chung2020splat}, named entity recognition \cite{shon2021slue, borgholt2021we, pasad2021use} and speech translation \cite{bansal2017towards, chung2020generative}.
-Cardiac arrest detection for emergency calls has also been used to evaluate speech representations \cite{borgholt2021we}.
-For local representations, phoneme classification is very common \cite{lee_unsupervised_2009, hsu_learning_2017, chorowski_unsupervised_2019, chung2019unsupervised, liu2021tera}.
-However, automatic speech recognition has become the \textit{de facto} standard benchmark task \cite{ling2020decoar, chung2020generative, hsu2021hubert}.
+Tasks that evaluate representations in terms of speaker-related information include speaker verification \cite{hsu_unsupervised_2017, khurana_factorial_2019, milde_unspeech_2018}, speaker identification \cite{oord_representation_2018, jati_neural_2019, chung_unsupervised_2019, liu_nonautoregressive_2020}, dialect classification \cite{khurana_factorial_2019}, emotion recognition \cite{pascual_learning_2019, yang_superb_2021} and gender classification \cite{lee_unsupervised_2009}.
+The semantic content of representations are evaluated using tasks such as intent classification \cite{morais_endtoend_2021, yang_superb_2021}, slot filling \cite{lai_semisupervised_2021, yang_superb_2021}, sentiment analysis \cite{liu_mockingjay_2020}, question answering \cite{chung_splat_2021}, named entity recognition \cite{shon_slue_2021, borgholt_we_2021, pasad_use_2022} and speech translation \cite{bansal_speechtotext_2017, chung_generative_2020}.
+Cardiac arrest detection for emergency calls has also been used to evaluate speech representations \cite{borgholt_we_2021}.
+For local representations, phoneme classification is very common \cite{lee_unsupervised_2009, hsu_learning_2017, chorowski_unsupervised_2019, chung_unsupervised_2019, liu_tera_2021}.
+However, automatic speech recognition has become the \textit{de facto} standard benchmark task \cite{ling_decoar_2020, chung_generative_2020, hsu_hubert_2021}.
 
 \paragraph{Moving forward} 
-
-Most of the seminal work has focused on improving speech recognition \cite{schneider2019wav2vec, baevski2020wav2vec}.
-This focus has gained traction over the last couple of years, as computational resources have become more accessible and end-to-end models \cite{graves2006connectionist, chan2016listen} have been established as the dominant approach to speech recognition \cite{gulati2020conformer}.
-It is important to stress that self-supervised models, such as wav2vec 2.0 \cite{baevski2020wav2vec}, represent a breakthrough, and recent successful approaches build upon this method.
-That is, deep self-attention models combined with masking \cite{hsu2021hubert, wang2021unispeech, chen2021wavlm}.
-This development mirrors years of rapid progress in masked language modeling within natural language processing \cite{devlin2019bert,clark_2020_electra} and we expect this to continue for unsupervised neural speech representation learning.
+Most of the seminal work has focused on improving speech recognition \cite{schneider_wav2vec_2019, baevski_wav2vec_2020}.
+This focus has gained traction over the last couple of years, as computational resources have become more accessible and end-to-end models \cite{graves_connectionist_2006, chan_listen_2015} have been established as the dominant approach to speech recognition \cite{gulati_conformer_2020}.
+It is important to stress that self-supervised models, such as wav2vec 2.0 \cite{baevski_wav2vec_2020}, represent a breakthrough, and recent successful approaches build upon this method.
+That is, deep self-attention models combined with masking \cite{hsu_hubert_2021, wang_unispeech_2021, chen_wavlm_2021}.
+This development mirrors years of rapid progress in masked language modeling within natural language processing \cite{devlin_bert_2018,clark_2020_electra} and we expect this to continue for unsupervised neural speech representation learning.
 
@@
 \label{sec:conc}
-
 We reviewed unsupervised representation learning for speech, focusing on two primary categories: self-supervised methods and probabilistic latent variable models.

### 2023-09-23 chapters/background.tex
Work on variational autoencoders

@@
 
-While the papers in this thesis were written to be self-contained, page limits and the need to maintain a specific focus, forced us to limit the scope of technical background each paper provides.
+While the papers in this thesis were written to be self-contained, page limits and the need to maintain a specific focus forced us to limit the scope of the technical background that each paper provides.
 In this chapter, we present a more in-depth overview of the technical background relevant to the papers in this thesis.
-We start by quantifying the concept of uncertainty, and how it relates to information and probability. We then introduce the task of out-of-distribution detection and a few ways to tackle it. 
-Then, we present variational autoencoders, and how they can be used to learn representations of speech.
+We start by quantifying the concept of uncertainty relating it to information and probability.
+We then introduce the task of out-of-distribution detection and review the existing litterature on the topic.
+Then, we describe variational autoencoders, how they can be used to learn representations of speech, and some of their challenges.
 Finally, we introduce the concept of self-supervised learning and highlight important differences to variational autoencoders.
@@
 When information is limited, we can describe the state of the world only with uncertainty; we are simply not certain.
-In that sense, uncertainty is a word we use to refer to the information that is missing, whether we know what is missing or not.
+In that sense, uncertainty is a word we use to refer to missing information, whether we know what is missing or not.
 A natural way to describe an uncertain world state is to use probabilities.
@@
 
-The \emph{information entropy of a discrete random variable} $X$ is defined as the \emph{expected information content} of an outcome of $X$,
+The information entropy of a discrete random variable $X$ is defined as the \emph{expected information content} of an outcome of $X$,
 %
@@
 To generalize the concept of information entropy to continuous random variables, Shannon originally replaced the sum over the probability mass function in \eqref{eq:information-entropy} with an integral over the probability density function, as suggested by the definition $H(X) = \mathbb{E}_X\left[I_X(x)\right]$.
-This approach leads to the definition of the \emph{information entropy of a continuous random variable} $X$ as,
+This approach leads to the definition of the information entropy of a continuous random variable $X$ as,
 %
@@
 
-However, the differential entropy does not have several of the desirable attributes of the discrete version; it can be negative, it is not invariant under a change of variables, and it is not dimensionally correct%
+However, the differential entropy does not have several of the desirable properties of the discrete version; it can be negative, it is not invariant under a change of variables, and it is not dimensionally correct%
 \footnote{\label{fn:dimensional-analysis-of-differential-entropy}
@@
 %
-In this section we will introduce out-of-distribution (OOD) detection, relate it to the concept of uncertainty and to the previous section on information entropy. 
+In this section we will introduce out-of-distribution (OOD) detection and relate it to the concept of uncertainty. 
 To supplement the short related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}, we will then provide a concise review of the literature on out-of-distribution detection.
-A
 
@@
 In the general case, we assume that we have a domain of in-distribution data, $\mathcal{D}_{\text{in}}$, and we would like to build a model that can be used to assess whether a test data point $\xb$ originates from that domain or not.
-Since such OOD data is unknown to the model, it is, by definition, a source of epistemic uncertainty.
-This makes OOD detection closely related to uncertainty quantification.
-Only in cases where strong prior knowledge of the data is incorporated into a model, can we expect it to be able to make reliable predictions on OOD data.
+Since such OOD data is unknown to the model, it is, by definition, a source of epistemic uncertainty. This makes OOD detection closely related to uncertainty quantification. 
+% Only in cases where strong prior knowledge of the data is incorporated into a model, can we expect it to be able to make reliable predictions on OOD data.
 
@@
 Outlier detection most clearly differs from the other tasks by directly processesing all observations and aiming to select outliers from a single contaminated dataset. 
-The remaining tasks differ in whether they detect both covariate and semantic shift and require the simultaneous classification of in-distribution classes.%, and whether the in-distribution data contains more than one class.
+The remaining tasks differ in whether they detect both covariate and semantic shift and require the simultaneous classification of in-distribution classes.
+%, and whether the in-distribution data contains more than one class.
 Anomaly detection deals with multiclass data and detects both covariate and semantic shift without requiring simultaneous classification of in-distribution classes whereas novelty detection, open set recognition and OOD detection are usually concerned with semantic shifts.
@@
 Different from novelty detection, OOD detection methods sometimes draw on examples of OOD data or require the simultaneous classification of in-distribution classes, although this is not always the case.
-OOD detection benchmarks almost always take OOD data to be from external datasets different from the training dataset which distinguishes it from open set recognition that usually benchmarks on a single dataset split into ID and OOD data.
+OOD detection benchmarks almost always take OOD data to be from external datasets different from the training dataset which distinguishes it from open set recognition that usually uses a single dataset split into ID and OOD data.
 % Another closely related field is that of network-based uncertainty estimation \cite{neal_bayesian_1995, hernandez-lobato_probabilistic_2015, gawlikowski_survey_2023}. %[14, 10, 36, 20]. 
@@
 Recently, several approaches for deep neural networks have been developed that formally address the rejection of samples $\xb\notin\mathcal{D}_{\text{in}}$. 
-One way to categorize the different approaches is based on whether the underlying model is a classifier that estimates a conditional probability distribution $p(y|\xb)$ over some target variable given the input, or a model that learns a probability density $p(\xb)$ over the input itself.
+One way to categorize the different approaches is based on whether the underlying model is a classifier that estimates a conditional probability distribution $p(y|\xb)$ over some target variable $y$, or a model that learns a probability density $p(\xb)$ over the input itself.
 We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \cite{graham_denoising_2023,liu_unsupervised_2023a}.
@@
 After defining a score, we typically use a validation set to tune a threshold $\tau$ such that $\xb$ is considered OOD if $s(\xb) > \tau$.
-The threshold is typically chosen such that performance on the validation set is above some standard, for instance by imposing constraints on the recall and precision.
+The threshold is typically chosen such that performance on the validation set is above some level, for instance by imposing constraints on the recall and precision.
 Many works also evaluate the performance of OOD detection methods using the area under the receiver operating characteristic (AUROC) curve which does not require a threshold to be chosen.
@@
 from the \emph{likelihood} assigned to the input \cite{bishop_novelty_1994,choi_waic_2019,kirichenko_why_2020,ren_likelihood_2019,serra_input_2020,xiao_likelihood_2020,morningstar_density_2021,nalisnick_detecting_2019,bergamin_modelagnostic_2022,maaloe_biva_2019,havtorn_hierarchical_2021},
-from a \emph{reconstruction} of the input $\xb$ \cite{sakurada_anomaly_2014,xia_learning_2015,lyudchik_outlier_2016,zhou_anomaly_2017,chen_outlier_2017,schlegl_unsupervised_2017,zong_deep_2018,li_madgan_2019,graham_denoising_2023,liu_unsupervised_2023a}, or
+from a \emph{reconstruction} of the input \cite{sakurada_anomaly_2014,xia_learning_2015,lyudchik_outlier_2016,zhou_anomaly_2017,chen_outlier_2017,schlegl_unsupervised_2017,zong_deep_2018,li_madgan_2019,graham_denoising_2023,liu_unsupervised_2023a}, or
 from a hidden \emph{representation} of the input \cite{denouden_improving_2018,hendrycks_using_2019,ahmadian_likelihoodfree_2021,bergman_classificationbased_2020,tack_csi_2020,sehwag_ssd_2021,xiao_we_2021}.
@@
 Supervised methods often derive the score 
-from the \emph{probabilities} given by the predictive distribution $p(y|\xb)$ \cite{hendrycks_baseline_2017,hendrycks_scaling_2022} or 
-from the \emph{logits} of $p(y|\xb)$ \cite{hendrycks_scaling_2022,liu_energy-based_2020}. 
-Other methods use a latent \emph{representation} of the input \cite{lee_simple_2018,li_anomaly_2019,ndiour_outofdistribution_2020,cook_outlier_2020,zaeemzadeh_outofdistribution_2021}, similar to unsupervised methods.
+from the \emph{probabilities} given by the predictive distribution $p(y|\xb)$ \cite{hendrycks_baseline_2017,hendrycks_scaling_2022}, 
+from the \emph{logits} of $p(y|\xb)$ \cite{hendrycks_scaling_2022,liu_energy-based_2020}, or 
+use a latent \emph{representation} of the input \cite{lee_simple_2018,li_anomaly_2019,ndiour_outofdistribution_2020,cook_outlier_2020,zaeemzadeh_outofdistribution_2021}, similar to unsupervised methods.
 In the following sections, we provide a more in-depth overview of the different approaches to OOD detection following the above categorization into supervised and unsupervised methods.
@@
 
-% \subsubsection{Model agnostic: Related works}
-% Since \textcite{nalisnick_deep_2019} and \textcite{hendrycks_deep_2019}, different test statistics or methodologies for OOD detection using DGMs were proposed.
-Most of the recent solutions were highly influenced by three major lines of work: \emph{typicality set}, \emph{likelihood ratio} test statistics, and \emph{model misestimation}.
-
-% The typicality set hypothesis was introduced by \textcite{nalisnick_detecting_2019} as a possible explanation for the DGMs assigning higher likelihood to OOD data.
-The typicality set is the subset of the model full support where the model samples from and this does not intersect with the region of higher likelihood.
-While the typicality test was introduced for batch-OOD detection, \textcite{morningstar_density_2021} shows that it also works well in the single-sample case.
-This is also confirmed by our own experiments.
-
-% The likelihood ratio test statistic method by \textcite{ren_likelihood_2019} assumes that every input is composed by a background component and a semantic component.
-For OOD detection, only the semantic component matters.
-In addition to a model trained on the in-distribution data, they proposed to train a background model on perturbed inputs data and then for each test example consider as OOD score the likelihood ratio between the two models.
-\textcite{schirrmeister_understanding_2020}, instead, trained the background model on a more general distribution of images by considering 80 million general tiny images.
-Similarly to these approaches, \textcite{serra_input_2020} argued that the failure of DGMs is due to the high-influence that the input complexity has on the likelihood.
-Therefore, they proposed to use a general lossless image compression algorithm as a background model.
-% All these methods, however, require additional knowledge of the OOD data for either choosing an image augmentation procedure to perturb the input data or for choosing a specific compressor.
-
-% Another line of works blame the models themselves and not the test statistics.
-\textcite{zhang_understanding_2021} argued that model misestimation is the main cause of higher likelihood assigned to OOD data.
-This can be due to both the model architecture and the maximum likelihood objective.
-\textcite{kirichenko_why_2020} and \textcite{schirrmeister_understanding_2020} showed that normalizing flows can achieve better OOD performance despite achieving a worse likelihood if one changes some model design choices.
-% %
-% Other works in the literature focused on deriving specific test statistics that work only for a specific model, for example for VAEs \parencite{xiao_likelihood_2020, maaloe_biva_2019, havtorn_hierarchical_2021}, or for normalizing flows \parencite{kirichenko_why_2020, ahmadian_likelihoodfree_2021}.
-
-% %All the solutions we presented can be described in terms of the following two categories: \emph{model-agnostic} vs \emph{model-specific} test statistics and statistics \emph{with OOD assumptions} vs statistics \emph{with no OOD assumptions}.
-In this work, we are mostly interested in model-agnostic scores with no OOD assumptions, i.e.\@ scores that can be directly computed with any models and that do not need any explicit assumption on the OOD data.
-
-% As mentioned in the introduction, we frame the OOD detection problem in terms of statistical tests problem.
-Recently, \textcite{haroush_statistical_2021} showed that adopting hypothesis testing at the layer and channel level of a neural network can be used for OOD detection in the discriminative setting.
-They used both Fisher's method and Simes' method to combine class-conditional $p$-values computed for each convolutional and dense layer of a deep neural network.
-We focus on the unsupervised setting using DGMs and use hypothesis testing on statistics that can be computed on all differentiable DGM.
-As already explained in section \cref{sec_modelagnostic:combination_p_values}, \textcite{morningstar_density_2021} considered the combination of different statistics for OOD detection.
-The main difference with their approach is that we propose statistics that can be applied to any differentiable generative model and combine them by using Fisher's method, which takes advantage of using only one-sided independent statistics.
-Concurrently, \textcite{choi_robust_2021} derived the score statistic by starting from the likelihood ratio statistic and applying a Laplace approximation.
-They computed the score statistic only for certain layers of the model and for a specific example, the OOD score is given by the infinity norm of these different layer scores after a ReLU operation.
-Our procedure differs both in the derivation of the score statistic and its usage since we compute the score statistic for the entire model.
-
-% \subsubsection{Hierarchical: Related works}
-
-% So far, no reliable direct likelihood-based method has been found for fully unsupervised deep generative model OOD detection.
-% A major line of work considers developing new scores that are more reliable than the likelihood.
-% This includes the \textit{typicality} test presented by \textcite{nalisnick_detecting_2019} which is an OOD detection test based on the typicality of a batch of potentially OOD examples.
-% This approach however requires a batch of examples from the same class (OOD or not) which limits its practical applicability.
-% In \textcite{ren_likelihood_2019}, the \textit{likelihood ratio} between a primary model and a background model was shown to be an effective score for OOD detection.
-% However, to train the background model, the in-distribution data is perturbed via a data augmentation technique that is designed with knowledge about the confounding factors between the in-distribution data and the OOD data.
-Furthermore, it is tuned towards high performance on a known OOD dataset.
-% \textcite{serra_input_2020} take a similar approach and attribute the failure to detect OOD data to the high influence of the input complexity on the likelihood and choose a generic lossless compression algorithm as the background model.
-% Although this method gives good results, no single best choice of compression algorithm exists for all types of OOD data, and any particular choice encodes prior knowledge about the data into the detection method.
-% Both these methods can be seen as correcting for low-level features of the OOD data being assigned high model likelihood by using a second model focused exclusively on these features.
-
-% Similar to these methods, the majority of the approaches to OOD detection make assumptions about the nature of the OOD data.
-% The assumptions encompass using labels on the in-distribution data \cite{hendrycks_baseline_2017, liang_enhancing_2018, alemi_uncertainty_2018, lee_simple_2018, lakshminarayanan_simple_2017}, examples of OOD data \cite{hendrycks_deep_2019}, augmenting in-distribution data to mimic it \cite{ren_likelihood_2019}, or assuming a certain data type \cite{serra_input_2020}.
-% Any of these assumptions encode implicit biases into the model about the attributes of OOD data which, in turn, might impair performance on truly unknown data examples (unknown unknowns).
-
-% % Introduce the contrast to results with VAEs
-% While some of these methods achieve very good results on OOD detection with autoregressive models \cite{oord_pixel_2016, salimans_pixelcnn_2017} and invertible flow-based models \cite{kingma_glow_2018}, it was recently shown that they can be much less effective for VAEs \cite{xiao_likelihood_2020} highlighting the need for a more reliable OOD score for VAEs.
-% Although VAEs have the same failure cases as autoregressive and flow-based models, the caveat is that the difference in the likelihood is generally not as big and reconstructions of OOD can be surprisingly good \cite{xiao_likelihood_2020}.
-% \textcite{xiao_likelihood_2020} alleviate this by refitting the inference network, as previously proposed by \textcite{cremer_inference_2018, mattei_refit_2018}, to a potentially OOD example and measuring the so-called \textit{likelihood regret}.
-% However, refitting the inference network can be computationally expensive, especially for the large hierarchical VAEs that are used to model complex data \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
-Furthermore, this scales poorly to large amounts of potentially OOD examples as the optimization is done per example.
-
-% A few methods have approached OOD detection in a completely unsupervised fashion \cite{maaloe_biva_2019, choi_waic_2019, xiao_likelihood_2020}.
-% The work of \textcite{maaloe_biva_2019} is the most related to ours.
-They introduce BIVA, a deep hierarchy of stochastic latent variables with a top-down and bottom-up inference model and achieve state-of-the-art likelihood scores.
-% They also provide early results indicative that a looser likelihood bound may have value in OOD detection.
-% In this paper, we provide an explanation of those results, and significantly improve upon them.
-
-% Out-of-distribution detection and anomaly detection are both related concepts in machine learning and deep learning, but they address slightly different problems and have distinct focuses.
-
-% Out-of-Distribution Detection (OOD Detection):
-% Out-of-distribution detection involves determining whether a given input or data point belongs to the same distribution as the training data.
-In other words, it helps identify whether an input is coming from a category or distribution that the model has never encountered during training.
-This is especially important for safety-critical applications where the model should not make predictions on inputs that are significantly different from what it has seen before.
-OOD detection aims to prevent the model from making confident predictions on unfamiliar inputs that might lead to incorrect results.
-% Anomaly Detection:
-% Anomaly detection, on the other hand, focuses on identifying rare and unusual instances within the data, regardless of whether they belong to the same distribution as the training data or not.
-Anomalies can be defined as data points that deviate significantly from the norm or expected behavior.
-Anomaly detection is used to find instances that are different from the majority and might indicate potential errors, fraud, faults, or other unusual occurrences.
-% In summary, the key differences are:
-
-% Focus:
-% OOD Detection: Focuses on detecting inputs that are coming from distributions that differ significantly from the training data distribution.
-% Anomaly Detection: Focuses on identifying rare and unusual instances, regardless of whether they come from a different distribution or not.
-% Purpose:
-% OOD Detection: Primarily used for safety and robustness, ensuring that the model doesn't make confident predictions on unfamiliar data.
-% Anomaly Detection: Used for various applications such as fraud detection, fault detection, and identifying outliers in the data.
-% Detection Approach:
-% OOD Detection: Often involves measuring the uncertainty of the model's predictions. High uncertainty indicates that the input might be out-of-distribution.
-% Anomaly Detection: Focuses on detecting data points that are statistically rare or far from the norm.
-% It's worth noting that while the concepts are distinct, they can sometimes overlap.
-An out-of-distribution data point could also be considered an anomaly, but not all anomalies are necessarily out-of-distribution.
-Anomaly detection methods might use OOD detection techniques as part of their process to flag anomalous instances that are not in line with the expected distribution.
-
-% Nice review:
-% https://openaccess.thecvf.com/content/CVPR2023W/VAND/papers/Graham_Denoising_Diffusion_Models_for_Out-of-Distribution_Detection_CVPRW_2023_paper.pdf
-
-% \cite{kabir_neural_2018} Survey of uncertainty quantification in deep learning.
-
-% https://arxiv.org/pdf/2306.05674.pdf
-
 \section{Variational autoencoders}
 
-Deep generative models are a class of models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution. 
-There exist at least six main classes of deep generative models: variational autoencoders \cite{kingma_autoencoding_2014,rezende_stochastic_2014,ranganath_hierarchical_2016,vahdat_nvae_2020,child_very_2021}, normalizing flow models \cite{dinh_nice_2015,rezende_variational_2015,dinh_density_2017,kingma_glow_2018,grathwohl_ffjord_2018}, diffusion models \cite{sohl-dickstein_deep_2015,song_generative_2019, ho_denoising_2020, vahdat_scorebased_2021}, energy-based models \cite{lecun_tutorial_2006,hinton_fast_2006,salakhutdinov_efficient_2010,du_implicit_2019}, autoregressive models \cite{oord_conditional_2016,oord_wavenet_2016,radford_improving_2018}, and generative adversarial networks \cite{goodfellow_generative_2014,arjovsky_wasserstein_2017,brock_large_2019,karras_stylebased_2019}.
-Besides data generation, common properties of deep generative models include the ability to compute an exact or approximate likelihood of input $p(\xb)$, reconstruct an input $\xb$ from its latent representation $\zb$, and represent an input $\xb$ in a latent space.
-These properties enable various methods for their use in OOD detection.
+A variational autoencoder (VAE) \cite{kingma_autoencoding_2014,rezende_stochastic_2014} is a type of deep generative model and an instance of a latent variable model.
+In the following, we will introduce VAEs starting from the general definition of latent variable models.
+We then derive the maximum likelihood objective used to train VAEs, discuss its different components, and highlight some central challenges.
+Finally, we discuss some of the most important extensions to VAEs made model speech data.
+
+The fundamental assumption underlying the definition of latent variable models is that a data point $\xb$ is created via a generative process that involves one or more unobserved, stochastic latent variables  $\zb$.
+Latent variable model design centers around capturing this generative process and learning to generate new data points $\xb$.
+The generative process is usually defined by a joint distribution $p(\xb, \zb)$ over the data and latent variables.
+In many models, the joint distribution is factorized as $p(\xb,\zb) = p(\xb|\zb)p(\zb)$, including Gaussian mixture models \cite{dempster_maximum_1977}, hidden Markov models \cite{rabiner_tutorial_1989}, and probabilstic principal component analysis \cite{tipping_probabilistic_1999}.
+In this case, we can write the marginal distribution of the data as,
+%
+\begin{equation} \label{eq:latent-variable-model-marginal-distribution}
+    p(\xb) = \int p(\xb|\zb)p(\zb) \, d\zb \enspace .
+\end{equation}
+%
+The ability to infer the latent variables $\zb$ from a given data point $\xb$ is of interest in applications that focus on representation learning.
+Inference is usually performed by computing the posterior distribution $p(\zb|\xb)$, which can be computed using Bayes' rule,
+%
+\begin{equation} \label{eq:latent-variable-model-posterior-distribution}
+    p(\zb|\xb) = \frac{p(\xb|\zb)p(\zb)}{p(\xb)} \enspace .
+\end{equation}
+%
+The marginal distribution $p(\xb)$ involves an integral over all possible values of the latent variable $\zb$ that is in many cases analytically intractable.
+This is especially true when the latent variable space is high-dimensional or if the conditional distribution $p(\xb|\zb)$ is complex, as with neural networks or non-conjugate models.
+Approximate methods such as variational inference \cite{jordan_introduction_1999} and Markov Chain Monte Carlo (MCMC) \cite{mohamed_monte_2019} are therefore often used to compute the marginal and posterior distributions but these methods can be computationally expensive.
+
+When the data is high-dimensional, an obvious choice for modelling the generative process is to use a deep neural network.
+This makes the marginal and posterior distributions intractable and requires approximate methods for inference.
+Variational autoencoders \cite{kingma_autoencoding_2014,rezende_stochastic_2014} use variational inference to learn a latent representation of the input data.
+Variational inference introduces an approximate posterior distribution $q(\zb|\xb)$ that is optimized to be close to the true posterior distribution $p(\zb|\xb)$. 
+We can write the marginal distribution of the data as,
+%
+\begin{align}
+    a
+\end{align}
+%
+
+VAEs amortize the cost of inference by parameterizing the approximate posterior with a neural network $q_\phib(\zb|\xb)$ that can produce approximate posterior distributions for any input $\xb$.
+
+trained jointly with the generative model $p_\thetab(\xb|\zb)$ by maximizing the evidence lower bound (ELBO),
+
+Any VAE aims to model this generative process by learning a generative model $p_\thetab(\xb, \zb)$
+
+For a more background on VAEs and a comparison with self-supervised methods, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{app:paper-brief}.
 
@@
 % https://jmtomczak.github.io/blog/13/13_trouble_in_paradise.html
+
+Deep generative models are a class of models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution. 
+There exist at least six main classes of deep generative models: variational autoencoders \cite{kingma_autoencoding_2014,rezende_stochastic_2014,ranganath_hierarchical_2016,vahdat_nvae_2020,child_very_2021}, normalizing flow models \cite{dinh_nice_2015,rezende_variational_2015,dinh_density_2017,kingma_glow_2018,grathwohl_ffjord_2018}, diffusion models \cite{sohl-dickstein_deep_2015,song_generative_2019, ho_denoising_2020, vahdat_scorebased_2021}, energy-based models \cite{lecun_tutorial_2006,hinton_fast_2006,salakhutdinov_efficient_2010,du_implicit_2019}, autoregressive models \cite{oord_conditional_2016,oord_wavenet_2016,radford_improving_2018}, and generative adversarial networks \cite{goodfellow_generative_2014,arjovsky_wasserstein_2017,brock_large_2019,karras_stylebased_2019}.
+Besides data generation, common properties of deep generative models include the ability to compute an exact or approximate likelihood of input $p(\xb)$, reconstruct an input $\xb$ from its latent representation $\zb$, and represent an input $\xb$ in a latent space.
+These properties enable various methods for their use in OOD detection.
 

### 2023-09-23 chapters/introduction.tex
Work on variational autoencoders

--- 2023-09-22 chapters/introduction.tex
@@
 The evolution of technology to support and automate decision-making can be traced back to the development of algorithms and mathematical techniques by ancient civilizations. 
-One example is the Egyptian Rhind Mathematical Papyrus, dated around 1550 BCE.
-It describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \cite{georges_universal_2001}.
+The Egyptian Rhind Mathematical Papyrus, dated around 1550 BCE, describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \cite{georges_universal_2001}.
 Around the same time, the Babylonians had developed a sophisticated system of mathematics that included algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \cite{fowler_square_1998}.
 
-However, these computing tools brought with them novel challenges.
+These computing tools brought with them novel challenges.
 The Babylonian mathematics, for instance, did not formalize the difference between rational and irrational numbers often leading to approximation errors such as the common practice of approximating $\pi$ with 3 \cite{georges_universal_2001}.
-Apart from limitations inherent to the tools, inadequate human proficiency in their use was also a source of error.
-The Plimpton 322 clay tablet, for instance, contains several errors presumably made by a novice in Babylonian mathematics (\cref{fig:plimpton_332}) \cite{britton_plimpton_2011}.
+Apart from such inherent limitations of the tools, inadequate human proficiency in their use was also a source of error.
+The Plimpton 322 clay tablet, for instance, contains several errors, presumably made by a novice in Babylonian mathematics (\cref{fig:plimpton_332}) \cite{neugebauer_mathematical_1945,britton_plimpton_2011}.
 % Easy to overlook, the challenges we face in adopting new technologies today are not new but have been faced by humans for thousands of years.
@@
     \caption{Technology is difficult.
-The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics.}
+The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics.
+\cite[photo credit]{neugebauer_mathematical_1945}}
     \label{fig:plimpton_332}
@@
 Working with Babbage, Ada Lovelace is widely regarded as the first to have recognized that programmable computers could have applications beyond pure arithmetic.
-However, it was not until the advent of electronics that the first successful general-purpose computers were developed, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}).
+However, it was not until the advent of electronics that the first successful general-purpose computers were built, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}).
 % While the potential of a general-purpose programmable computer with applications beyond pure arithmetic was recognized by Ada Lovelace,  
@@
     \includegraphics[height=0.4\textheight]{eniac_programmers.jpeg}
-    \caption{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.} %Photo: \copyright Corbis/ Getty Images.}
+    \caption{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.
+\cite[photo credit]{usarmyresearchlaboratoryarltechnicallibrary_female_1940}} %Photo: \copyright Corbis/ Getty Images.}
     \label{fig:eniac_programmers}
@@
 
-However, these technologies come with the same risks as previous generations did, and in some cases, they may exacerbate them. 
+However, these technologies come with the same risks as previous generations did, and in some cases, may amplify them. 
 Limitations of the technology itself, human error, and the complexity of the real world are all factors that can lead to failures.
@@
     \caption{A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
-The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \cite{nationaltransportationsafetyboardnhsa_collision_2019}.}
+The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \cite[photo credit]{nationaltransportationsafetyboardnhsa_collision_2019}.}
     % \caption{Aerial view of crash location showing path of pedestrian as she attempted to cross N.
@@
 
-Recently, European policymakers proposed a set of requirements for applications of machine learning to ensure that they are safe, reliable, and trustworthy \cite{europeancommission_briefing_2021}.
-This AI Act groups applications of machine learning into three categories based on the potential risk they pose to society: limited risk, high risk, and unacceptable risk.
-Limited risk applications, such as in computer games, must meet transparency requirements, and unacceptable risk applications, such as public use of facial recognition, are banned.
-High risk applications, such as in autonomous vehicles and healthcare, must meet a strict set of requirements focusing on anti-discrimination and robustness.
+Recently, European policymakers proposed a set of requirements to ensure that machine learning applications are safe, reliable, and trustworthy \cite{europeancommission_briefing_2021}. 
+This AI Act groups applications into three categories based on the potential risk they pose to society: limited, high, and unacceptable.
+Limited risk applications, such as in computer games, must meet transparency requirements, and technologies such as public facial recognition are banned due to unacceptable risk.
+High risk applications include autonomous vehicles and systems within healthcare. These must meet a set of strict requirements focusing on anti-discrimination and robustness.
 
 One of the key ways to improve the safety, reliability and trustworthiness of machine learning systems is to equip them with the ability accurately estimate the uncertainty of their predictions. 
-A model that can accurately estimate the uncertainty of its predictions is able to communicate to other systems or humans when it is not to be trusted, and so, is able to reduce the risk of wrong predictions leading to catastrophic mistakes.
+Almost any model will inevitably make mistakes, but if it can accurately estimate the uncertainty of its predictions, it can communicate to other systems or humans when it is not to be trusted, and so, reduce the risk that wrong predictions lead to catastrophic failures.
 Fittingly, a recent study by the European Parliamentary Research Services concluded:
@@
 %
-At the start of this project in 2020, Corti's main product was a software system providing decision-support for emergency medical dispatchers operating the 1-1-2 emergency line in Copenhagen, Denmark, and the 9-1-1 emergency line in Seattle, Washington, in the United States.
-The system, having been further developed, is still in use in 2023, and provides an interface for navigating a graph-based protocol, defined by the local emergency medical services, which dispatchers use to triage patients during emergency calls.
+At the start of this project in 2020, Corti's main product was a software system providing decision-support for call-takers that operate the 1-1-2 emergency line in Copenhagen, Denmark, and the 9-1-1 emergency line in Seattle, Washington, in the United States.
+The system is in continued use in 2023, and provides an interface for navigating a graph-based protocol, defined by the local emergency medical services, which dispatchers use to triage patients during emergency calls.
 Calls are transcribed in real-time using automatic speech recognition built for the customer-domain and the system makes suggestions for the dispatcher to use in the conversation with the caller, including notifications about forgotten protocol questions \cite{havtorn_multiqt_2020} and potential cases of cardiac arrests \cite{cite15, cite14}.
@@
 %
-In 2021, Corti entered a research collaboration with the Capital Region of Denmark and the Copenhagen Emergency Services to develop a system for recognizing stroke cases in calls to the 1-1-2 emergency line and the 1813 medical helpline.
+\paragraph{Motivation} In 2021, Corti entered a research collaboration with the Capital Region of Denmark and the Copenhagen Emergency Services to develop a system for recognizing stroke cases in calls to the 1-1-2 emergency line and the 1813 medical helpline.
 Stroke is one of the biggest causes of disability and death worldwide \cite{cite1,cite2,cite3} and effective treatment is highly time-sensitive \cite{cite4,cite5}.
@@
 Additionally, stroke cases are relatively rare, occurring in only $0.25\%$ of all calls made to the Copenhagen 1-1-2 emergency line in 2021.
-This makes it difficult for call-takers to gain practical experience with stroke cases which exacerbates the difficulty of recognizing them in the first place.
-Although several efforts have been made to improve the recognition of stroke cases in emergency calls, there is still a need for better tools to support call-takers in recognizing stroke cases \cite{cite13,cite14,cite15}.
+This makes it difficult for call-takers to gain practical experience with stroke cases which compounds to the difficulty of recognizing them.
+Although several efforts have been made to improve recognition rates, there is still a need for better tools to support call-takers in recognizing stroke cases \cite{cite13,cite14,cite15}.
 
-A reasonable machine learning pipeline to assist in recognizing stroke cases on calls to emergency services might consist of an automatic speech recognition model that transcribes the conversation and a binary classifier that estimates the probability that the transcript describes a case of stroke.
+\paragraph{Augmentation} A reasonable machine learning pipeline to assist in recognizing stroke cases on calls to emergency services might consist of an automatic speech recognition model that transcribes the conversation and a binary classifier that estimates the probability that the transcript describes a case of stroke.
 Such a system might be trained on a dataset calls with verified stroke and non-stroke cases and evaluated against a held-out test set of calls where the call-taker has indicated whether they suspect a stroke.
@@
 For instance, the automatic speech recognition model might make an error in the transcription of particular word or phrase due to a noisy environment, overlapping, slurred or mumbling speech, or words not in the model's training data.
-The classifier might misclassify a conversation due to medically ambiguous symptoms, a general lack of information given in the conversation, or errors of the speech recognizer propagated via the transcription.
+The classifier might misclassify a conversation due to medically ambiguous symptoms, a general lack of information given in the conversation, or transcription errors made by the speech recognizer. 
 How to best make such models accurately represent and estimate the uncertainty of given predictions, and how to use it to improve the value of such a system, especially in cooperation with the call taker, is an open question.
 What is clear, however, is that the ability of such a system to accurately estimate the uncertainty of given predictions is crucial for its safe and reliable deployment in the real world.
-Underconfident predictions could lead to unnecessary delays in treatment and increased costs for the healthcare system.
-Overconfident predictions could lead to misdiagnoses and potentially fatal consequences for patients.
+Overconfident predictions could lead to unnecessary delays in treatment, misdiagnoses, increased costs for the healthcare system, and potentially fatal consequences for patients.
 
@@
 %
-During the course of this project, Corti's product portfolio has expanded to include several new products. 
-One of these is a software system for medical coding of clinical notes. 
+\paragraph{Motivation} During the course of this project, Corti's portfolio has expanded to include a software system for medical coding of clinical notes. 
 When a patient is admitted to a hospital, the medical staff will write a clinical note describing the patient's diagnosis and the procedures they underwent, including drug prescriptions. 
 These notes are then used for billing and reimbursement purposes, as well as for research and quality assurance. 
-The clinical notes are written, usually by a doctor, in natural language adhering to a certain structure. Later, a medical coder will assign a set of medical codes to the note based on its content. 
+The clinical notes are written, usually by a doctor, in natural language adhering to a certain structure. Later, a medical coder will assign a set of codes to the note based on its content. 
 The process of medical coding is time-consuming and error-prone due to the vast number and high complexity of medical diagnoses and procedures.
 For instance, the widely used International Classification of Diseases (ICD) standard consist of 55,000 medical codes in version 10 and 85,000 in version 11 \cite{worldhealthorganisationwho_international_2023}.
-Additionally, a single medical note will usually contain several diagnoses and procedures which must all be inferred from the clinical note.
-In public benchmarking datasets such as MIMIC-III and IV the number is around 15 \cite{johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}.
-Official guidelines describe how to code specific diagnoses and procedures and combinations thereof \cite{centersformedicaremedicaidservicesus_icd10cm_2023}.
-For instance, a code may have several exclusion or inclusion criteria which define which other codes must be coded along with it, or when they must not.
+Additionally, a single medical note will usually contain several diagnoses and procedures which must all be inferred from the clinical note \cite{johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}.
+Furthermore, any single code can have several exclusion or inclusion criteria defined by official guidelines that determine when other codes must be coded along with it, and when they must not \cite{centersformedicaremedicaidservicesus_icd10cm_2023}.
+% Official guidelines describe how to code specific diagnoses and procedures and combinations thereof . 
+% For instance, a code may have several exclusion or inclusion criteria which define which other codes must be coded along with it, or when they must not.
 
-A reasonable machine learning pipeline to assist in medical coding might consist of a natural language processing model that extracts relevant information from the clinical note and outputs a number of probability distributions that each estimate the probability that the note should be assigned one of the medical codes.
-Training and using such as system comes with several challenges, however.
-For instance, the prevalence of different medical codes in the training data is highly imbalanced; it is common to have several orders of magnitude difference between the frequency of the most frequent code and the least frequent codes.
+\paragraph{Augmentation} A reasonable machine learning pipeline to assist in medical coding might consist of a natural language processing model that extracts relevant information from the clinical note and outputs a number of probability distributions that each estimate the probability that the note should be assigned one of the medical codes.
+However, training and using such as system comes with several challenges. 
+For instance, the prevalence of different medical codes in the training data is highly imbalanced; it is common to have several orders of magnitude difference between the frequency of the most frequent code and the least frequent code.
 Since each clinical note is associated with multiple codes that have complex co-occurence patterns, it is often impossible to exactly correct for this class imbalance by stratified sampling, which is an otherwise common approach to deal with class imbalance in machine learning.
-This also gives rise to
-Furthermore, some codes are easily confused with each other, for instance the ICD-10 codes Z87.891 ``Personal history of nicotine dependence'' and F17.210 ``Nicotine dependence, cigarettes, uncomplicated", but in practice only one of them should be assigned to a given note.
+% This also gives rise to 
+Furthermore, some codes are highly similar, for instance the ICD-10 codes Z87.891 ``Personal history of nicotine dependence'' and F17.210 ``Nicotine dependence, cigarettes, uncomplicated", but in practice, only one of them should be assigned to a given note.
 As with the stroke recognition system, we would like the model to express high degree of uncertainty when it makes a mistake.
 For example, if the model is uncertain about two similar codes, it might be appropriate to ask a human expert to review the note and make the final decision.
-Additionally, the lack of training data for the rarest codes makes it difficult to learn to robustly predict those codes, so we would in general want the model to be uncertain about such codes.
+Additionally, the lack of training data for the rarest codes makes them difficult to learn to robustly predict, and so, we would often want the model to be uncertain for rare codes.
 
@@
 However, the tendency of deep neural networks to overfit or memorize the training data \cite{arpit_closer_2017, burg_memorization_2021} and the use of training objectives that are proxies for the evaluation metrics of interest lead to models that often are miscalibrated; that is, the predicted probability of a class does not reflect the true probability of the class.
-This means, for instance, that if for a certain input the model assigns 70\% probability to a class, it would not generally be right 70\% of the time, if we could repeat the prediction multiple times under similar circumstances \cite{guo_calibration_2017, kull_temperature_2019}.
+This means, for instance, that if for a certain input the model assigns 70\% probability to a class, it would not generally be right 70\% of the time - if we could repeat the prediction multiple times under similar circumstances \cite{guo_calibration_2017, kull_temperature_2019}.
 Usually, the predicted probabilities are too extreme which leads to overconfident predictions.
@@
 Although no model is perfect, the inability of a human, or another model, to assess the certainty with which a prediction was made, makes it impossible to know how to weigh it in a decision-making process.
-A connection can be made back to the incident with the Uber self-driving car.
+% A connection can be made back to the incident with the Uber self-driving car. 
 The multitude of rapid misclassifications made by the Uber self-driving car in Tempe is likely an indication that the model responsible for object-classification was overconfident in its predictions and unable to represent or communicate uncertainty to other systems in the car, or the safety driver.
@@
 % 
-The problem of miscalibration in machine learning models is well-known and has been studied extensively \cite{lewis_sequential_1995, platt_probabilistic_1999, garczarek_classification_2002, zadrozny_transforming_2002, bennett_using_2003}.
+The problem of miscalibration in machine learning models is well-known and has been studied for decades \cite{lewis_sequential_1995, platt_probabilistic_1999, garczarek_classification_2002, zadrozny_transforming_2002, bennett_using_2003}.
 One of the best known methods for calibrating a binary classifier is Platt scaling which fits a logistic regression model to the model outputs, assuming that the miscalibration can be corrected by a logarithmic function \cite{platt_probabilistic_1999}.
@@
 These methods are generally simple and effective, but not without limitations. 
-To perform the calibration, most of the methods require a held-out validation set on which the model was not trained.
-Platt scaling and isotonic regression are not naively applicable to multi-class classification problems.
-How to calibrate models for structured prediction tasks such as speech recognition and machine translation remains an open problem \cite{astudillo_uncertainty_2010, astudillo_integration_2013, jayashankar_detecting_2020}.
+To perform the calibration, most of the methods require a held-out validation set on which the model was not trained, Platt scaling and isotonic regression are not naively applicable to multi-class classification problems, and how to calibrate models for structured prediction tasks, such as speech recognition and machine translation, remains an open problem \cite{astudillo_uncertainty_2010, astudillo_integration_2013, jayashankar_detecting_2020}.
 
-It is also important to realize that correct calibration does not guarantee that the probability of a prediction is accurate. 
-Specifically, even if a model is well-calibrated, it can still be overconfident for data that were not presented to the model during training such as rare events, outliers, and adverse examples.
-Such data are sometimes called out-of-distribution examples with reference to the training distribution.
-For instance, a model trained to classify images of cats and dogs and provide calibrated class probabilities will still fail if presented with an image of a horse.
+It is also important to realize that correct calibration of a machine learning model does not guarantee that the probability of a prediction is accurate. 
+Specifically, even well-calibrated model can be overconfident for data that were not presented to it during training.
+This includes rare events, outliers, and adverse examples which are sometimes collectively referred to as out-of-distribution examples.
+To give an example, a model trained to classify images of cats and dogs and provide calibrated class probabilities will still fail if presented with an image of a horse.
 In this case, it has no option but to distribute 100\% total probability across the cat and dog categories, even though that is clearly wrong.
-Worse yet, we might expect the model to then at least assign 50\% probability to each of the cat and dog categories, to indicate the uncertainty, but that behavior is not guaranteed.
-Indeed, since no horses were in the training data, the model will not have learned specialized features for horses, nor learned to associate known features with horses.
+Worse yet, to indicate the uncertainty, we might expect the model to assign 50\% probability to each of the cat and dog categories, but that behavior is not guaranteed.
+Indeed, since no horses were in the training data, the model will not have learned specialized features for horses, nor learned to associate relevant known features with horses.
 So, if some features of a particular horse resemble those learned for a dog more than for a cat, the model might assign arbitrarily high probability to the dog class and therefore risks being confidently wrong \cite{zhou_survey_2022}.
@@
 At a fundamental level, we can distinguish between uncertainty that is present in the knowledge we have, and uncertainty that originates from the knowledge that we do not have.
-This is sometimes called known unknowns and unknown unknowns, and is referred to in terms of \textit{aleatoric uncertainty} and \textit{epistemic uncertainty} \cite{kendall_what_2017}. 
-Aleatoric uncertainty (from Latin (aleatorius) 'dice player' or 'gambler') refers to uncertainty present in the data itself due to randomness in the process that generated the data.
+These sources are sometimes called \emph{known unknowns} and \emph{unknown unknowns}, or referred to in terms of \textit{aleatoric uncertainty} and \textit{epistemic uncertainty} \cite{kendall_what_2017}.
+Aleatoric comes from the Latin word 'aleatorius' for 'dice player' or 'gambler' and refers to uncertainty present in the data itself due to randomness in the process that generated it.
 This kind of uncertainty is commonly seen as irreducible but, since it is represented in the collected data, it can usually be modelled directly.
-Epistemic uncertainty (from Greek (epistḗmē) 'knowledge') refers to uncertainty due to things one could in principle know, but does not in practice.
-This lack of knowledge could be due to a lack of data, or an improper model specification.
+Epistemic comes from the Greek word 'epistḗmē' which means 'knowledge' and refers to uncertainty due to things one could in principle know, but does not in practice.
+Such lack of knowledge could for instance be due to a lack of data, or an improper model specification.
 Aleatoric uncertainty is sometimes referred to as \textit{stochastic uncertainty} and epistemic uncertainty as \textit{systematic uncertainty} \cite{kendall_what_2017}. 
@@
 For instance, a model may be uncertain about a particular prediction due to any of these two kinds of uncertainty.
-For that reason, the quantity of interest is usually the ``total" uncertainty, often referred to as \textit{predictive uncertainty}, which is simply the combined epistemic and aleatoric uncertainties.
+For that reason, the quantity of interest is usually the combined aleatoric and epistemic uncertainties, often referred to as \textit{predictive uncertainty}.
 
-An example of aleatoric uncertainty is the uncertainty in the transcription of a word due to a noisy environment, overlapping, slurred or mumbling speech. 
+An example of aleatoric uncertainty is the uncertainty in the transcription of a word due to a noisy environment or overlapping, slurred or mumbling speech. 
 Unknown words are arguably sources of epistemic uncertainty although a good speech recognition model may generalize well if the word's spelling and pronunciation follow the same patterns as the words in the training data.
-Another example of epistemic uncertainty is the occurrence of truly out-of-distribution examples, such as the horse in the image classifier example above: inputs unlike anything the model has seen during training, especially if they require outputs unavailable to the model.
-In high-dimensional data with a practically unlimited diversity in possible out-of-distribution examples, it is impossible to collect enough data to eliminate all sources of epistemic uncertainty.
-If we included horses in the training data, the model would likely learn to recognize them, but it would still be unable to recognize giraffes, or zebras, or unicorns.
-Ultimately, any practical machine learning model will have some non-negligible sources of epistemic uncertainty.
+Another example of epistemic uncertainty is the occurrence of truly out-of-distribution examples, such as the horse in the image classifier example from earlier: inputs unlike anything the model has seen during training, especially if they require outputs unavailable to the model.
+In high-dimensional data with a practically unlimited diversity in out-of-distribution examples, it is impossible to collect enough data to eliminate all potential sources of epistemic uncertainty.
+Let us include horses in the training data.
+The model would likely learn to recognize them, but it would still be unable to recognize giraffes, or zebras, or unicorns.
+Ultimately, we must accept that any practical machine learning model will have some sources of epistemic uncertainty.
 
-In practice, we are interested in the predictive uncertainty, so we need to include both aleatoric and epistemic parts in the uncertainty estimate. 
-Since sources of epistemic uncertainty are by definition those not represented in the data, it is not generally possible to model them directly, as it is for aleatoric uncertainty.
+Since aleatoric uncertainty is present in the data, it can usually be modelled directly.
+For instance, in the case of speech recognition, we can model the uncertainty in the transcription of a word by a distribution over the words in the vocabulary.
+But how do we model epistemic uncertainty?
+Since sources of epistemic uncertainty are by definition those not represented in the data, it is not generally possible to model them directly.
+This makes epistemic uncertainty more difficult to quantify than aleatoric uncertainty.
+Methods for estimating epistemic uncertainty include Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995} which represent it by an explicit distribution over learned model parameters.
+Ensembling methods \cite{gal_dropout_2016,lakshminarayanan_simple_2017} take a similar approach but use an implicit distribution.
+Other approaches include anomaly detection and the recent field of out-of-distribution detection which for example represent epistemic uncertainty by special output classes for out-of-distribution data, distributions over data representations, or distances between them (see \cref{sec:out-of-distribution-detection}).
+
+% In practice, we are interested in the predictive uncertainty, so we need to include both aleatoric and epistemic parts in the uncertainty estimate. 
+% Since sources of epistemic uncertainty are by definition those not represented in the data, it is not generally possible to model them directly, as it is for aleatoric uncertainty.
 This makes epistemic uncertainty more difficult to model and quantify than aleatoric uncertainty.
-% Methods for estimating epistemic uncertainty include Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995}, ensembling \cite{gal_dropout_2016,lakshminarayanan_simple_2017} and a suite of recent research into out-of-distribution detection (see \cref{sec:out-of-distribution-detection}).
-In contrast to aleatoric uncertainty, which is usually represented by a distribution over the data, epistemic uncertainty is often represented by an implicit or explicit distribution over the model parameters, such as in Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995} and ensemble models \cite{gal_dropout_2016,lakshminarayanan_simple_2017}.
+% % Methods for estimating epistemic uncertainty include Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995}, ensembling \cite{gal_dropout_2016,lakshminarayanan_simple_2017} and a suite of recent research into out-of-distribution detection (see \cref{sec:out-of-distribution-detection}).
+% In contrast to aleatoric uncertainty, which is usually represented by a distribution over the data, epistemic uncertainty is often represented by an implicit or explicit distribution over the model parameters, such as in Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995} and ensemble models \cite{gal_dropout_2016,lakshminarayanan_simple_2017}.
 In most cases, such methods require multiple forward passes through the model to estimate the predictive uncertainty making them computationally expensive and impractical for real-time applications.
-Within the field of out-of-distribution detection, the epistemic uncertainty is not always expressed in model parameters, but sometimes by a distribution over data representations, or output probabilities with special classes for out-of-distribution data.
-In such cases, the uncertainty can be estimated in a single forward pass through the model.
+Within the field of out-of-distribution detection, the epistemic uncertainty is generally not expressed in model parameters, but e.g.
+by a distribution over data representations, output probabilities with special classes for out-of-distribution data, or by distances in feature spaces.
+In such cases, the uncertainty can often be estimated in a single forward pass through the model.
 
@@
 %
-This introduction provided a high-level overview of the motivation behind the research project and highlighted important use-cases of speech processing, natural language processing and the importance of uncertainty quantification.
+This introduction provided a high-level overview of the motivation behind the research project by focusing on the use-cases of stroke recognition and automated medical coding as well as speech processing and the importance of uncertainty quantification.
 The remainder of \cref{part:background} consists of two chapters.
-\Cref{chp:technical-background} provides relevant technical background to the papers of the thesis, including uncertainty, out-of-distribution detection, and variational autoencoders.
-Finally, \cref{chp:main-contributions} outlines the main contributions of the thesis in relation to the published papers and the research project as a whole.
+\Cref{chp:technical-background} provides relevant technical background to the papers of the thesis, including uncertainty, out-of-distribution detection, and variational autoencoders while \cref{chp:main-contributions} outlines the main contributions of the thesis in relation to the published papers and the research project as a whole.
 
-\Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} each contain a number of chapters that each correspond to a primary paper.
+\Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper.
 \Cref{part:unsupervised-uncertainty-estimation} contains papers on variational autoencoders and their applications to estimation of epistemic uncertainty and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.

### 2023-09-23 chapters/paper_review/03_approaches.tex
Work on variational autoencoders

@@
 
-\edit{The Gumbel softmax \cite{jang_categorical_2017} is another frequently used approach for obtaining a discrete representation space, and has also been used for AEs \cite{eloff_unsupervised_2019}.
+\edit{The Gumbel softmax \cite{jang_categorical_2016} is another frequently used approach for obtaining a discrete representation space, and has also been used for AEs \cite{eloff_unsupervised_2019}.
 In addition to the approaches discussed above, several other works on speech representation learning take inspiration from the AE framework \cite{zeiler_rectified_2013, badino_autoencoder_2014, badino_discovering_2015, kamper_unsupervised_2015, renshaw_comparison_2015, settle_acoustically_2019}.}
@@
 \noindent
-The quantization module $g(\cdot)$ uses a Gumbel softmax~\cite{jang_categorical_2017} with a straight-through estimator.
+The quantization module $g(\cdot)$ uses a Gumbel softmax~\cite{jang_categorical_2016} with a straight-through estimator.
 Since the quality of the learned representations is contingent on the quality of the quantization, wav2vec 2.0 combines two techniques to learn high-quality codebooks.
@@
 Several lines of work have begun to address the problem of learning representations of spans of speech, especially word segments, typically referred to as \textit{acoustic word embeddings}.
-Early work on unsupervised acoustic word embeddings defined them as vectors of distances from the target segment to a number of pre-defined ``template" segments~\cite{levin_fixed-dimensional_2013}.
+Early work on unsupervised acoustic word embeddings defined them as vectors of distances from the target segment to a number of pre-defined ``template" segments~\cite{levin_fixeddimensional_2013}.
 Later work used variants of neural autoencoders~\cite{chung_audio_2016,holzenberger_learning_2018,kamper_truly_2019,peng_correspondence_2020}.
 These are often evaluated on word discrimination, that is the task of determining whether two word segments correspond to the same word or not~\cite{carlin_rapid_2011}.
-This task can be thought of as a proxy for query-by-example search, since the basic operation in search is to determine whether a segment in the search database matches a query segment, and has been used for evaluation of both frame-level (e.g.,~\cite{kamper_unsupervised_2015}) and word-level~\cite{levin_fixed-dimensional_2013,kamper_deep_2016} representations.
+This task can be thought of as a proxy for query-by-example search, since the basic operation in search is to determine whether a segment in the search database matches a query segment, and has been used for evaluation of both frame-level (e.g.,~\cite{kamper_unsupervised_2015}) and word-level~\cite{levin_fixeddimensional_2013,kamper_deep_2016} representations.
 

### 2023-09-24 chapters/background.tex
Work on background and VAEs

@@
 \begin{equation} \label{eq:kullback-leibler-divergence}
-    D_{KL}(P\parallel Q) = \int_\mathcal{X} p(x) \log \left(\frac{p(x)}{q(x)}\right) \enspace .
+    D_{KL}(p(x)\parallel q(x)) = \int_\mathcal{X} p(x) \log \left(\frac{p(x)}{q(x)}\right) \enspace .
 \end{equation}
 %
-The Kullback-Leibler divergence $D_{KL}(P\parallel Q)$, also known as the \emph{information gain}, is often interpreted as the amount of additional information required to represent events from distribution $P$ with density $p(x)$ using a code optimized for distribution $Q$ with density $q(x)$.
-In other words, it measures how surprised one would be if they used distribution $Q$ to represent events from distribution $P$, i.e.
-the \emph{relative entropy} of $P$ with respect to $Q$.
+The Kullback-Leibler divergence $D_{KL}(p(x)\parallel q(x))$, also known as the \emph{information gain}, is often interpreted as the amount of additional information required to represent events from a distribution with density $p(x)$ using a code optimized for a distribution with density $q(x)$.
+In other words, it measures how surprised one would be if they used distribution $q$ to represent events from distribution $p$, i.e.
+the \emph{relative entropy} of $p$ with respect to $q$.
 
@@
 Methods that use representative OOD data have achieved high performance since they can directly learn to distinguish between in-distribution and OOD data. 
-\textcite{hendrycks_deep_2019} augments the original training objective with a task-dependent outlier exposure loss that aims to make the output logits different for outlier data. 
+\textcite{hendrycks_deep_2019} augment the original training objective with a task-dependent outlier exposure loss that aims to make the output logits discriminative of outlier data. 
 In a similar vein, \textcite{dhamija_reducing_2018} propose losses designed to maximize the entropy of $p(y|\xb)$ and decrease feature magnitudes for OOD data sampled from other datasets. 
@@
 Other methods including MCD \cite{yu_unsupervised_2019}, NGC \cite{wu_ngc_2021}, and UDG \cite{yang_semantically_2021} utilize external, unlabeled, noisy data to improve OOD detection performance without requiring cleanly labelled OOD examples.
-As we discussed earlier, for many modern applications of machine learning, input data is often high dimensional and complex making sufficiently sampling the space of out-of-distribution data intractable.
-Therefore, methods that rely on samples of OOD data are inherently limited by its availability and how well it covers the large variation in OOD data.
+As we discussed earlier, for many modern applications of machine learning, input data is often high dimensional and complex which makes it difficult to obtain enough representative OOD data to ensure robust OOD detection capabilities.
 
@@
 A number of methods do not require access to actual OOD data but synthesize it instead. 
-Several methods synthesize this OOD data by adding noise to in-distribution data \cite{liang_enhancing_2018, lee_simple_2018, ren_likelihood_2019}. 
+Several methods do so by adding noise to in-distribution data \cite{liang_enhancing_2018, lee_simple_2018, ren_likelihood_2019}. 
 For instance, \textcite{ren_likelihood_2019} propose a number of baselines including training a binary classifier to distinguish between original and perturbed in-distribution data.
 They also propose adding an OOD class to softmax classifiers and training it to predict perturbed in-distribution data, or alternatively, training the predicted class distribution to output uniform distribution for perturbed in-distribution inputs.
-With ODIN \cite{liang_enhancing_2018}, the authors propose to calibrate $p(y|\xb)$ with temperature scaling \cite{guo_calibration_2017} and add gradient-based, input-dependent perturbations to the inputs to use the calibrated maximum class probability as the OOD score.
-This method does not require retraining the classifier and generally improves on the baselines of \textcite{ren_likelihood_2019}.
-\textcite{vyas_outofdistribution_2018} train an ensemble of classifiers on different subsets of the training data, with the left out data taken as OOD, and propose novel loss over $p(y|\xb)$ that seeks to maintain a set margin between its average entropy for the OOD and in-distribution examples.
+While these methods are appealing, their weaknesses have been pointed out by later work which generally improve on their perforamnce.
+An example is ODIN \cite{liang_enhancing_2018}, in which the authors propose to calibrate $p(y|\xb)$ with temperature scaling \cite{guo_calibration_2017} and add gradient-based, input-dependent perturbations to the inputs and use the calibrated maximum class probability as the OOD score.
+\textcite{vyas_outofdistribution_2018} train an ensemble of classifiers on different subsets of the training data, with the left out data taken as OOD, and propose novel loss over $p(y|\xb)$ that seeks to maintain a predefined margin between its average entropy for the OOD and in-distribution examples.
 Another approach generates OOD inputs using a generative adversarial network \cite{lee_training_2018}. 
@@
 % No OOD data, logits
-A number of works have noted that the maximum softmax probability is not generally a reliable score for OOD detection \cite{hendrycks_scaling_2022,liu_energy-based_2020}. 
+Although some works use the maximum softmax probability as a score for OOD detection \cite{hendrycks_deep_2019,ren_likelihood_2019}, several works have noted that it is not generally reliable \cite{hendrycks_scaling_2022,liu_energy-based_2020}.
 \textcite{liu_energy-based_2020} make an interesting argument as to why based on the energy $E(\xb; f)$ of a softmax classifier $f(\xb)$ \cite{lecun_tutorial_2006},
@@
 %
-Specifically, the authors relate the maximum softmax probability,
+Specifically, the authors write the maximum softmax probability as,
 %
@@
     \max_y p(y|\xb;f) 
-    = \max_y \frac{\exp f_y(\xb)}{\sum_{i=1}^K \exp f_i(\xb)} 
-    = \frac{\exp f^\text{max}(\xb)}{\sum_{i=1}^K \exp f_i(\xb)} \enspace ,
+    &= \max_y \frac{\exp f_y(\xb)}{\sum_{i=1}^K \exp f_i(\xb)} \nonumber \\
+    &= \frac{\exp f^\text{max}(\xb)}{\sum_{i=1}^K \exp f_i(\xb)} \nonumber \\
+    &= \frac{1}{\sum_{i=1}^K \exp \left(f_i(\xb) - f^\text{max}(\xb)\right)} \enspace ,
 \end{align}
-%
-to the energy by noting that,
+where $f^\text{max}(\xb) = \max_i f_i(\xb)$. 
+The authors then relate the maximum softmax probability to the energy by noting that,
 %
 \begin{align}
-    \log \max_y p(y|\xb;f) = E(\xb; f(\xb) - f^\text{max}(\xb)) = E(\xb; f) - f^\text{max}(\xb) \enspace .
+    \log \max_y p(y|\xb;f) = E\left(\xb; f(\xb) - f^\text{max}(\xb)\right) = E(\xb; f) - f^\text{max}(\xb) \enspace .
 \end{align}
@@
 This shows that log of the softmax confidence score is equivalent to the special case of the energy score, where all the logits are shifted by their maximum logit value. 
-The authors empirically show that $E(\xb;f)$ is a more reliable score for OOD detection than the maximum softmax probability and note that the energy $E(\xb; f)$ tends to be larger for OOD data than for in-distribution data - contrary to the maximum softmax probability.
-They conclude that this shift results in $\max_y p(y|\xb;f)$ being a biased score for OOD detection and propose instead to use the energy directly as the OOD score.
-This energy-score is improved in ReAct by feature clipping \cite{sun_react_2021}.
+The authors then note that the energy $E(\xb; f)$ empirically tends to be larger for OOD data than for in-distribution data, while $f^\text{max}(\xb)$ tends to be smaller. 
+They conclude that this shift results in the maximum softmax probability being a biased score for OOD detection and propose to instead use the energy directly.
+This energy-score was improved in ReAct by feature clipping \cite{sun_react_2021}.
 
@@
 
-A variational autoencoder (VAE) \cite{kingma_autoencoding_2014,rezende_stochastic_2014} is a type of deep generative model and an instance of a latent variable model.
-In the following, we will introduce VAEs starting from the general definition of latent variable models.
+% A variational autoencoder (VAE) \cite{kingma_autoencoding_2014,rezende_stochastic_2014} is a type of deep generative model and an instance of a latent variable model. 
+In the following, we will introduce VAEs starting from the definition of the group of latent variable models to which they belong.
 We then derive the maximum likelihood objective used to train VAEs, discuss its different components, and highlight some central challenges.
@@
 
+\paragraph{Latent variable models} 
 The fundamental assumption underlying the definition of latent variable models is that a data point $\xb$ is created via a generative process that involves one or more unobserved, stochastic latent variables  $\zb$.
@@
 The generative process is usually defined by a joint distribution $p(\xb, \zb)$ over the data and latent variables.
-In many models, the joint distribution is factorized as $p(\xb,\zb) = p(\xb|\zb)p(\zb)$, including Gaussian mixture models \cite{dempster_maximum_1977}, hidden Markov models \cite{rabiner_tutorial_1989}, and probabilstic principal component analysis \cite{tipping_probabilistic_1999}.
-In this case, we can write the marginal distribution of the data as,
+We can write the marginal distribution of the data as,
 %
 \begin{equation} \label{eq:latent-variable-model-marginal-distribution}
-    p(\xb) = \int p(\xb|\zb)p(\zb) \, d\zb \enspace .
+    p(\xb) = \int p(\xb,\zb) \, d\zb \enspace .
 \end{equation}
@@
 The ability to infer the latent variables $\zb$ from a given data point $\xb$ is of interest in applications that focus on representation learning.
-Inference is usually performed by computing the posterior distribution $p(\zb|\xb)$, which can be computed using Bayes' rule,
+Inference is performed by computing the posterior distribution $p(\zb|\xb)$ via Bayes' theorem,
 %
 \begin{equation} \label{eq:latent-variable-model-posterior-distribution}
-    p(\zb|\xb) = \frac{p(\xb|\zb)p(\zb)}{p(\xb)} \enspace .
+    p(\zb|\xb) = \frac{p(\xb,\zb)}{p(\xb)} \enspace .
 \end{equation}
 %
-The marginal distribution $p(\xb)$ involves an integral over all possible values of the latent variable $\zb$ that is in many cases analytically intractable.
-This is especially true when the latent variable space is high-dimensional or if the conditional distribution $p(\xb|\zb)$ is complex, as with neural networks or non-conjugate models.
-Approximate methods such as variational inference \cite{jordan_introduction_1999} and Markov Chain Monte Carlo (MCMC) \cite{mohamed_monte_2019} are therefore often used to compute the marginal and posterior distributions but these methods can be computationally expensive.
+In latent variable models such as Gaussian mixture models \cite{dempster_maximum_1977}, hidden Markov models \cite{rabiner_tutorial_1989}, and probabilistic principal component analysis \cite{tipping_probabilistic_1999}, the joint distribution is factorized as $p(\xb,\zb) = p(\xb|\zb)p(\zb)$.
 
-When the data is high-dimensional, an obvious choice for modelling the generative process is to use a deep neural network.
-This makes the marginal and posterior distributions intractable and requires approximate methods for inference.
-Variational autoencoders \cite{kingma_autoencoding_2014,rezende_stochastic_2014} use variational inference to learn a latent representation of the input data.
-Variational inference introduces an approximate posterior distribution $q(\zb|\xb)$ that is optimized to be close to the true posterior distribution $p(\zb|\xb)$. 
-We can write the marginal distribution of the data as,
+The marginal distribution $p(\xb)$ involves an integral over all possible values of the latent variable $\zb$ that is often analytically intractable.
+This is the case when the latent variable space is high-dimensional or the conditional distribution $p(\xb|\zb)$ is complex, as with non-conjugate models or neural networks.
+In these cases, approximate methods such as variational inference \cite{jordan_introduction_1999} and Markov Chain Monte Carlo (MCMC) \cite{mohamed_monte_2019} are used to compute the marginal and posterior distributions.
+
+\paragraph{Variational autoencoders} 
+To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to parameterize the generative model $p_\thetab(\xb|\zb)$ with a neural network with parameters $\thetab$.
+Since this makes the marginal and posterior distributions intractable, both training and inference require approximate methods.
+While MCMC methods are computationally expensive for large neural networks, variational inference provides a more scalable alternative by estimating the posterior distribution $p(\zb|\xb)$ with a variational approximation.
+
+Specifically, VAEs amortize the cost of inference by parameterizing $q_\phib(\zb|\xb)$ as a single neural network which makes it possible to efficiently compute approximate posterior distributions for any input $\xb$.
+With the approximate posterior, we can also compute the so-called evidence lower bound (ELBO) on the marginal distribution $p(\xb)$,
 %
-\begin{align}
-    a
+ \begin{align} \label{eq:variational-autoencoder-elbo}
+    \log p(\xb)
+    &= \log \int p_\thetab(\xb|\zb)p(\zb),\textrm{d}\zb \nonumber \\
+    &= \log \int q_\phib(\zb|\xb) \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)}\,\textrm{d}\zb \nonumber \\
+    &\geq \int q_\phib(\zb|\xb) \log \left( \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right) \,\textrm{d}\zb \nonumber \\
+    &\geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] \triangleq \mathcal{L}_{\thetab,\phib}(\xb) \enspace .
 \end{align}
@@
 
-VAEs amortize the cost of inference by parameterizing the approximate posterior with a neural network $q_\phib(\zb|\xb)$ that can produce approximate posterior distributions for any input $\xb$.
+Ideally, we want to use maximum likelihood estimation to maximize the ELBO with respect to the parameters $\thetab$ and $\phib$ and jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$.
+Computing the gradient of the ELBO with respect to the parameters of the generative model, $\thetab$, is straightforward,
+%
+\begin{equation}
+    % \nabla_\thetab \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
+    \nabla_\thetab \mathcal{L}_{\thetab,\phib}(\xb)
+    = \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \nabla_\thetab \log p_\thetab(\xb|\zb) \right] \enspace .
+\end{equation}
+%
+More troublesome is the gradient with respect to the parameters of the variational approximation, $\phib$.
+Here the parameters occur in the expectation itself and it is not possible to move the gradient inside the expectation.
+However, by reparameterizing the latent variable $\zb$ as a deterministic function of the input $\xb$ and a random variable $\epsilonb\sim p(\epsilonb)$, $\zb = g_\phib(\xb,\epsilonb)$, we can write the gradient as,
+%
+\begin{equation}
+    % \nabla_\phib \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
+    \nabla_\phib \mathcal{L}_{\thetab,\phib}(\xb)
+    = \mathbb{E}_{p(\epsilonb)} \left[ \nabla_\phib \log \frac{p_\thetab\left(\xb|g_\phib(\xb,\epsilonb)\right)p\left(g_\phib(\xb,\epsilonb)\right)}{q_\phib\left(g_\phib(\xb,\epsilonb)|\xb\right)} \right] \enspace .
+\end{equation}
+%
+The most important instantiation of this reparameterization is the diagonal Gaussian case, where $\zb = \mu_\phib(\xb) + \sigma_\phib(\xb) \odot \epsilonb$ with $\epsilonb\sim\mathcal{N}(0,\mathbf{I})$.
+It is usual to assume that the covariance matrix is diagonal which makes the method a mean-field approximation by assuming independence between the latent variables in $\zb$.
 
-trained jointly with the generative model $p_\thetab(\xb|\zb)$ by maximizing the evidence lower bound (ELBO),
+Since the joint factorizes in VAEs, we can also write the ELBO in \cref{eq:variational-autoencoder-elbo} as,
+%
+\begin{align} \label{eq:variational-autoencoder-elbo-factorized}
+    \log p(\xb) \geq \mathcal{L}_{\thetab,\phib}(\xb)
+            % &\triangleq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
+            &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right] + \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
+            &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right] - D_\text{KL}\left( q_\phib(\zb|\xb) \,||\, p(\zb) \right)
+\end{align}
+%
+This reveals an interpretation of the ELBO as the negative KL divergence between the variational approximation $q_\phib(\zb|\xb)$ and the prior distribution $p(\zb)$ plus the expected log-likelihood of the data under the generative model $p_\thetab(\xb|\zb)$.
 
-Any VAE aims to model this generative process by learning a generative model $p_\thetab(\xb, \zb)$
+The training of VAEs is usually performed by stochastic gradient descent on the ELBO.
+This method samples a mini-batch of data points $\xb$ from the empirical training data distribution $\hat{p}(\xb)$ and then evaluates the gradient of the ELBO with respect to the parameters $\thetab$ and $\phib$ on the mini-batch.
+To gain some insight into the objective that we are optimizing, we will see if we can exploit the expectation over the data distribution $\hat{p}(\xb)$ to form a factorization similar to \cref{eq:variational-autoencoder-elbo-factorized} but with the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$ in the KL-divergence.
+It will turn out that we can indeed factor the ELBO into such a KL-divergence, a reconstruction term, and a third term \cite{tomczak_trouble_2022}.
+For ease of notation, we will let $q(\xb) = \hat{p}(\xb)$ so that $q_\phib(\zb,\xb) = q_{\phib}(\zb|\xb)\hat{p}(\xb)$.
+%
+\begin{align}
+    \mathbb{E}_{\hat{p}(\xb)}\left[ \log p(\xb) \right] 
+    &= \mathbb{E}_{\hat{p}(\xb)}\left[ \log \int p_{\thetab}(\xb | \zb)\ p(\zb) \, \mathrm{d}\zb \right] \nonumber \\
+    &\geq \mathbb{E}_{\hat{p}(\xb)}\left[ \mathbb{E}_{q_{\phib}(\zb|\xb)}\left[ \log p_{\thetab}(\xb | \zb) + \log p(\zb) - \log q_{\phib}(\zb | \xb) \right] \right] \nonumber \\
+    &= \mathbb{E}_{q_{\phib}(\zb|\xb)\hat{p}(\xb)}\left[ \log p_{\thetab}(\xb | \zb) + \log p(\zb) - \log q_{\phib}(\zb | \xb) \right] \nonumber \\
+    &= \mathbb{E}_{q_{\phib}(\zb,\xb)}\left[ \log p_{\thetab}(\xb | \zb) \right] + \mathbb{E}_{q_{\phib}(\zb)}\left[ \log p(\zb) \right] \nonumber \\ 
+    &\quad - \mathbb{E}_{q_{\phib}(\zb,\xb)}\left[ \log q_{\phib}(\zb | \xb) \right]
+\end{align}
+%
+% where we let $q(\xb) = \hat{p}(\xb)$ for ease of notation so that $q_\phib(\zb,\xb) = q_{\phib}(\zb|\xb)\hat{p}(\xb)$ and we defined the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$.
+
+\todo[inline]{We can do this without all the special terminology. Just add and subtract E_{q(z)}[log q(z)].}
+
+We can observe that the second term $\mathbb{E}_{q_{\phib}(\zb)}\left[ \log p(\zb) \right]$ is the cross-entropy $H(q_\phib(\zb),p(\zb))$ which can also be written as $D_\text{KL}(q_\phib(\zb)||p(\zb)) + H(q_\phib(\zb))$.
+We can also observe that the third term $-\mathbb{E}_{q_{\phib}(\zb,\xb)}\left[ \log q_{\phib}(\zb | \xb) \right]$ is the conditional entropy $H_{q_\phib(\zb,\xb)}[\zb|\xb]$ which can be factored as $H_{q_\phib(\zb)}[\zb] - I_{q_\phib(\zb,\xb)}[\zb;\xb]$ where $I_{q_\phib(\zb,\xb)}[\zb;\xb]$ is the mutual information between $\zb$ and $\xb$ under the joint distribution $q_\phib(\zb,\xb)$.
+This gives us the following decomposition of the ELBO,
+%
+\begin{align}
+    \mathbb{E}_{\hat{p}(\xb)}\left[ \log p(\xb) \right] 
+    &\geq \mathbb{E}_{q_{\phib}(\zb,\xb)}\left[ \log p_{\thetab}(\xb | \zb) \right] + H(q_\phib(\zb),p(\zb)) + H_{q_\phib(\zb,\xb)}[\zb|\xb]
+\end{align}
+%
+% \begin{align}
+%     a &= \mathbb{E}_{\xb,\zb \sim q_{\phib}(\zb|\xb)\hat{p}(\xb)}\left[ \log p_{\thetab}(\xb | \zb) \right]
+% \end{align}
 
@@
 
-VAE \cite{kingma_autoencoding_2014,rezende_stochastic_2014}
+% VAE \cite{kingma_autoencoding_2014,rezende_stochastic_2014}
 
-Derive ELBO
+% Derive ELBO
 
-Importance sampling \cite{burda_importance_2016}
+% Importance sampling \cite{burda_importance_2016}
 
-Semi-supervised learning \cite{kingma_semi-supervised_2014,maaloe_semi-supervised_2017,siddharth_learning_2017}
+% Semi-supervised learning \cite{kingma_semi-supervised_2014,maaloe_semi-supervised_2017,siddharth_learning_2017}
 
-Trouble in paradise \cite{tomczak_trouble_2022}
+% Trouble in paradise \cite{tomczak_trouble_2022}
 
@@
 
-Deep generative models are a class of models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution. 
+\paragraph{Deep generative models} VAEs also belong to the class of deep generative models which are models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution.
 There exist at least six main classes of deep generative models: variational autoencoders \cite{kingma_autoencoding_2014,rezende_stochastic_2014,ranganath_hierarchical_2016,vahdat_nvae_2020,child_very_2021}, normalizing flow models \cite{dinh_nice_2015,rezende_variational_2015,dinh_density_2017,kingma_glow_2018,grathwohl_ffjord_2018}, diffusion models \cite{sohl-dickstein_deep_2015,song_generative_2019, ho_denoising_2020, vahdat_scorebased_2021}, energy-based models \cite{lecun_tutorial_2006,hinton_fast_2006,salakhutdinov_efficient_2010,du_implicit_2019}, autoregressive models \cite{oord_conditional_2016,oord_wavenet_2016,radford_improving_2018}, and generative adversarial networks \cite{goodfellow_generative_2014,arjovsky_wasserstein_2017,brock_large_2019,karras_stylebased_2019}.

### 2023-09-25 appendices/paper_brief.tex
Work on background and main contributions

@@
 That is, deep self-attention models combined with masking \cite{hsu_hubert_2021, wang_unispeech_2021, chen_wavlm_2021}.
-This development mirrors years of rapid progress in masked language modeling within natural language processing \cite{devlin_bert_2018,clark_2020_electra} and we expect this to continue for unsupervised neural speech representation learning.
+This development mirrors years of rapid progress in masked language modeling within natural language processing \cite{devlin_bert_2018,clark_electra_2020} and we expect this to continue for unsupervised neural speech representation learning.
 

### 2023-09-25 appendices/supplementary_modelagnostic.tex
Some formatting and changes to labels
Work on background and main contributions

--- 2023-08-11 appendices/appendix_modelagnostic.tex
@@
 \chapter[supplementary material: model-agnostic out-of-distribution detection using combined statistical tests]{Supplementary material: Model-Agnostic Out-of-Distribution Detection Using Combined Statistical Tests}
-\label{app:modelagnostic}
+\label{app:supplementary-modelagnostic}
 \ifthenelse{\equal{\skipappendices}{true}}{}{
@@
 \begin{equation}
-    \textup{MMD}_{\Phi_\textup{Fisher}}^2 \approx \sum_{k=1}^K  || [\Phi_\textup{Fisher} (x)]_{\mu_k} ||_2^2 \approx \sum_{k=1}^K \frac{p(z|k)}{\pi_k} (z - \mu_k)^T \Sigma^{-1} (z - \mu_k).
+    \textup{MMD}_{\Phi_\textup{Fisher}}^2 \approx \sum_{k=1}^K  \parallel [\Phi_\textup{Fisher} (x)]_{\mu_k} ||_2^2 \approx \sum_{k=1}^K \frac{p(z|k)}{\pi_k} (z - \mu_k)^T \Sigma^{-1} (z - \mu_k).
 \end{equation}
@@
 As the last experiment, we study how our proposed method with our model agnostic statistic performs against DoSE using the original statistics proposed in \cite{morningstar_density_2021}.
-For the VAEs model, they suggested to use the following 5 statistics: the posterior/prior cross-entropy $\text{H}[q_{\phi}( \textbf{z}\mid \textbf{x}), p(\textbf{z})]$, the posterior entropy $\text{H}[q_{\phi}(\textbf{z}\mid \textbf{x})]$, the posterior/prior KL divergence $\text{D}_{\text{KL}}[q_{\phi}(\textbf{z}\mid \textbf{x}) \mid \mid p(\textbf{z})]$, the posterior expected log-likelihood $\mathbb{E}_{q_{\phi}(\textbf{z}\mid \textbf{x})}[\log q_{\phi}(\textbf{z}\mid \mathbf{x})]$, and the log-likelihood $\log \mathbb{E}_{q_{\phi}(\textbf{z}\mid \textbf{x})} \left[ \frac{p_{\theta}(\textbf{x}, \textbf{z})}{q_{\phi}(\textbf{z}\mid \textbf{x})}\right]$.
+For the VAEs model, they suggested to use the following 5 statistics: the posterior/prior cross-entropy $\text{H}[q_{\phi}( \textbf{z}\mid \textbf{x}), p(\textbf{z})]$, the posterior entropy $\text{H}[q_{\phi}(\textbf{z}\mid \textbf{x})]$, the posterior/prior KL-divergence $\text{D}_{\text{KL}}[q_{\phi}(\textbf{z}\mid \textbf{x}) \mid \mid p(\textbf{z})]$, the posterior expected log-likelihood $\mathbb{E}_{q_{\phi}(\textbf{z}\mid \textbf{x})}[\log q_{\phi}(\textbf{z}\mid \mathbf{x})]$, and the log-likelihood $\log \mathbb{E}_{q_{\phi}(\textbf{z}\mid \textbf{x})} \left[ \frac{p_{\theta}(\textbf{x}, \textbf{z})}{q_{\phi}(\textbf{z}\mid \textbf{x})}\right]$.
 For DoSE on Glow, instead, they considered three metrics: the log-likelihood $p_{X}(\textbf{x}\mid \theta_n)$ and its two components, i.e.\@ the log-probability of the latent variable $p_{Z}(\textbf{z}\mid \textbf{x}, \theta_n)$ and the log-determinant of the Jacobian $\log | \text{J}_{f}(\textbf{x})|$.

### 2023-09-25 chapters/background.tex
Work on background and main contributions

@@
 
-For high-dimensional data, the most successful unsupervised models for OOD detection are deep autoencoders \cite{hinton_reducing_2006}, self-supervised methods \cite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020}, and deep generative models \cite{kingma_autoencoding_2014,rezende_stochastic_2014,dinh_nice_2015,rezende_variational_2015,ho_denoising_2020,hinton_fast_2006,oord_conditional_2016,goodfellow_generative_2014}.
+For high-dimensional data, the most successful unsupervised models for OOD detection are deep autoencoders \cite{hinton_reducing_2006}, self-supervised methods \cite{mikolov_efficient_2013,devlin_bert_2018,schneider_wav2vec_2019,chen_simple_2020}, and deep generative models \cite{kingma_autoencoding_2014,rezende_stochastic_2014,dinh_nice_2015,rezende_variational_2015,ho_denoising_2020,hinton_fast_2006,oord_conditional_2016,goodfellow_generative_2014}.
 Other important density estimation methods that have been applied to OOD detection include kernel density estimation \cite{parzen_estimation_1962}, nearest neighbor methods \cite{cover_nearest_1967}, support vector machines \cite{cortes_supportvector_1995,scholkopf_estimating_2001}, and Gaussian mixture models \cite{dempster_maximum_1977}.
@@
 
+% \paragraph{Deep generative models} 
+VAEs belong to the class of deep generative models which are models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution. 
+There exist at least six main classes of deep generative models: variational autoencoders \cite{kingma_autoencoding_2014,rezende_stochastic_2014,ranganath_hierarchical_2016,vahdat_nvae_2020,child_very_2021}, normalizing flow models \cite{dinh_nice_2015,rezende_variational_2015,dinh_density_2017,kingma_glow_2018,grathwohl_ffjord_2018}, diffusion models \cite{sohl-dickstein_deep_2015,song_generative_2019, ho_denoising_2020, vahdat_scorebased_2021}, energy-based models \cite{lecun_tutorial_2006,hinton_fast_2006,salakhutdinov_efficient_2010,du_implicit_2019}, autoregressive models \cite{oord_conditional_2016,oord_wavenet_2016,radford_improving_2018}, and generative adversarial networks \cite{goodfellow_generative_2014,arjovsky_wasserstein_2017,brock_large_2019,karras_stylebased_2019}.
+% Besides data generation, common properties of deep generative models include the ability to compute an exact or approximate likelihood of input $p(\xb)$, reconstruct an input $\xb$ from its latent representation $\zb$, and represent an input $\xb$ in a latent space.
+These properties enable various methods for their use in OOD detection.
+
 % A variational autoencoder (VAE) \cite{kingma_autoencoding_2014,rezende_stochastic_2014} is a type of deep generative model and an instance of a latent variable model. 
 In the following, we will introduce VAEs starting from the definition of the group of latent variable models to which they belong.
-We then derive the maximum likelihood objective used to train VAEs, discuss its different components, and highlight some central challenges.
-Finally, we discuss some of the most important extensions to VAEs made model speech data.
+We then derive the marginal likelihood objective used to train VAEs, discuss its different components, and highlight some central challenges facing VAEs.
 
-\paragraph{Latent variable models} 
+\subsection{Latent variable models}
 The fundamental assumption underlying the definition of latent variable models is that a data point $\xb$ is created via a generative process that involves one or more unobserved, stochastic latent variables  $\zb$.
@@
 %
-In latent variable models such as Gaussian mixture models \cite{dempster_maximum_1977}, hidden Markov models \cite{rabiner_tutorial_1989}, and probabilistic principal component analysis \cite{tipping_probabilistic_1999}, the joint distribution is factorized as $p(\xb,\zb) = p(\xb|\zb)p(\zb)$.
+In many latent variable models, the joint distribution is factorized as $p(\xb,\zb) = p(\xb|\zb)p(\zb)$ including Gaussian mixture models \cite{dempster_maximum_1977}, hidden Markov models \cite{rabiner_tutorial_1989}, and probabilistic principal component analysis \cite{tipping_probabilistic_1999}.
 
@@
 This is the case when the latent variable space is high-dimensional or the conditional distribution $p(\xb|\zb)$ is complex, as with non-conjugate models or neural networks.
-In these cases, approximate methods such as variational inference \cite{jordan_introduction_1999} and Markov Chain Monte Carlo (MCMC) \cite{mohamed_monte_2019} are used to compute the marginal and posterior distributions.
+In these cases, approximate methods such as variational inference \cite{jordan_introduction_1999} and Markov Chain Monte Carlo (MCMC) \cite{mohamed_monte_2019} are used to approximately compute the marginal and posterior distributions.
 
-\paragraph{Variational autoencoders} 
-To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to parameterize the generative model $p_\thetab(\xb|\zb)$ with a neural network with parameters $\thetab$.
+\subsection{Variational autoencoders} 
+To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to parameterize the generative model $p_\thetab(\xb,\zb)$ with a neural network with parameters $\thetab$.
+Specifically, they use a neural network to parameterize the conditional distribution $p_\thetab(\xb|\zb)$ and a simple distribution for the prior $p(\zb)$.
 Since this makes the marginal and posterior distributions intractable, both training and inference require approximate methods. 
@@
 
-Specifically, VAEs amortize the cost of inference by parameterizing $q_\phib(\zb|\xb)$ as a single neural network which makes it possible to efficiently compute approximate posterior distributions for any input $\xb$.
-With the approximate posterior, we can also compute the so-called evidence lower bound (ELBO) on the marginal distribution $p(\xb)$,
+\paragraph{Evidence lower bound}
+VAEs amortize the cost of variational inference by parameterizing $q_\phib(\zb|\xb)$ as a single neural network which makes it possible to efficiently compute approximate posterior distributions for any input $\xb$.
+The approximate posterior can also be used to compute the so-called evidence lower bound (ELBO) on the marginal distribution $p(\xb)$,
 %
@@
 %
+Although analytically evaluating the expectation is intractable, it can be approximated by sampling from the variational approximation $q_\phib(\zb|\xb)$. 
+Since both the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ are parameterized by neural networks that can be efficiently evaluated, and the prior $p(\zb)$ is simple, this makes it possible to compute the ELBO for any input $\xb$.
 
-Ideally, we want to use maximum likelihood estimation to maximize the ELBO with respect to the parameters $\thetab$ and $\phib$ and jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$.
-Computing the gradient of the ELBO with respect to the parameters of the generative model, $\thetab$, is straightforward,
+Since the joint factorizes in VAEs, we can also write the ELBO in \cref{eq:variational-autoencoder-elbo} as,
+%
+\begin{align} \label{eq:variational-autoencoder-elbo-factorized}
+    \mathcal{L}_{\thetab,\phib}(\xb)
+    % &\triangleq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
+    &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right] + \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
+    &= \underset{\text{reconstruction loss}}{\underbrace{\mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]}} - \underset{\text{KL-divergence to prior}}{\underbrace{D_\text{KL}\left( q_\phib(\zb|\xb) \,||\, p(\zb) \right)}} \enspace .
+\end{align}
+%
+This reveals a commonly used interpretation of the ELBO: 
+The expected log-likelihood of the data under the generative model $p_\thetab(\xb|\zb)$ acts as a reconstruction loss encouraging the model to make $q_\phib(\zb|\xb)$ as peaky as possible to make reconstruction easy.
+The negative KL-divergence between $q_\phib(\zb|\xb)$ and the prior $p(\zb)$ acts as a regularizer that forces the approximate posterior to be within the support of the prior (reverse KL-divergence). 
+A well-fitted VAE will have a large expected log-likelihood of the data under the generative model enabled by an informative latent variable.
+However, an optimal KL-divergence of zero entails that $q_\phib(\zb|\xb) = p(\zb)$ which means that $\zb$ is independent of $\xb$.
+This local optimum is referred to as posterior collapse and is particularly prone to happen for strong generative models, e.g.
+with autoregressive dependencies $p(\xb_{t}|\xb_{1:t},\zb)$.
+Some works have proposed to mitigate posterior collapse by tempering the KL-divergence term \cite{alemi_fixing_2018,higgins_vvae_2017} or by adding additional terms to the ELBO \cite{zhao_infovae_2018}.
+
+\paragraph{Sampling}
+For some choices of prior and approximate posterior, the KL-divergence term can be computed analytically, but we generally must estimate the reconstruction loss by sampling from the approximate posterior.
+This Monte Carlo estimator can be written as,
+%
+\begin{equation} \label{eq:variational-autoencoder-elbo-sampling}
+    \mathcal{L}_{\thetab,\phib}(\xb)
+    = \frac{1}{K} \sum_{k=1}^K \log p_\thetab(\xb|\zb_k) - D_\text{KL}\left( q_\phib(\zb|\xb) \,||\, p(\zb) \right) \enspace ,
+\end{equation}
+%
+where $\zb_k \sim q_\phib(\zb|\xb)$ are samples from the approximate posterior.
+To reduce the variance of the estimate, \textcite{burda_importance_2016} propose to use importance sampling to reweight the samples.
+
+\paragraph{Reparameterization} 
+Via the ELBO, we want to use maximum likelihood estimation to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$.
+Gradient-based optimization works well for neural networks but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
+Computing the gradient of the ELBO with respect to $\thetab$ is straightforward,
 %
@@
 %
-More troublesome is the gradient with respect to the parameters of the variational approximation, $\phib$.
-Here the parameters occur in the expectation itself and it is not possible to move the gradient inside the expectation.
-However, by reparameterizing the latent variable $\zb$ as a deterministic function of the input $\xb$ and a random variable $\epsilonb\sim p(\epsilonb)$, $\zb = g_\phib(\xb,\epsilonb)$, we can write the gradient as,
+The gradient with respect to the $\phib$ is more challenging since $\phib$ occurs in the expectation itself making it impossible to move the gradient inside the expectation. 
+However, by reparameterizing the latent variable $\zb$ as a deterministic function of the input $\xb$ and a random variable $\epsilonb\sim p(\epsilonb)$, $\zb = g_\phib(\xb,\epsilonb)$, we can write the gradient as a path-wise derivative,
 %
@@
     = \mathbb{E}_{p(\epsilonb)} \left[ \nabla_\phib \log \frac{p_\thetab\left(\xb|g_\phib(\xb,\epsilonb)\right)p\left(g_\phib(\xb,\epsilonb)\right)}{q_\phib\left(g_\phib(\xb,\epsilonb)|\xb\right)} \right] \enspace .
+\\
 \end{equation}
@@
 The most important instantiation of this reparameterization is the diagonal Gaussian case, where $\zb = \mu_\phib(\xb) + \sigma_\phib(\xb) \odot \epsilonb$ with $\epsilonb\sim\mathcal{N}(0,\mathbf{I})$.
-It is usual to assume that the covariance matrix is diagonal which makes the method a mean-field approximation by assuming independence between the latent variables in $\zb$.
+It is usual to make a mean-field approximation by assuming the covariance matrix is diagonal imposing independence between the latent variables in $\zb$.
 
-Since the joint factorizes in VAEs, we can also write the ELBO in \cref{eq:variational-autoencoder-elbo} as,
+Although lower variance than alternatives, such as score function estimators, a number of works have attempted to reduce the variance. 
+\textcite{roeder_sticking_2017} note that the path-wise derivative can be factored into a path-wise estimator and a score function estimator and propose to ignore the score function factor to reduce gradient variance.
+\textcite{rainforth_tighter_2019} that this problem is even larger for the importance weighted ELBO \cite{burda_importance_2016} leading to diminishing returns for increasing numbers of samples.
+Instead of removing the score function factor in the gradient, \textcite{tucker_doubly_2019} propose to reparameterize the score function and derive low variance gradient estimators also for importance weighted ELBO.
+
+\paragraph{Mutual information interpretation}
+Training by stochastic gradient descent training involves sampling mini-batches of data points $\xb$ from the empirical training data distribution $\hat{p}(\xb)$ and evaluating the gradient of the ELBO with respect to the parameters $\thetab$ and $\phib$ for each mini-batch.
+To gain some additional insight into the objective that we are optimizing, we will exploit this expectation over the data distribution $\hat{p}(\xb)$ to form a factorization similar to \cref{eq:variational-autoencoder-elbo-factorized} but with the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$ in the KL-divergence \cite{tomczak_trouble_2022}.
+The marginal likelihood under this expectation is lower bounded by the same expectation over the ELBO,
 %
-\begin{align} \label{eq:variational-autoencoder-elbo-factorized}
-    \log p(\xb) \geq \mathcal{L}_{\thetab,\phib}(\xb)
-            % &\triangleq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
-            &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right] + \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
-            &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right] - D_\text{KL}\left( q_\phib(\zb|\xb) \,||\, p(\zb) \right)
+\begin{align} \label{eq:variational-autoencoder-elbo-expected}
+    \mathbb{E}_{\hat{p}(\xb)}\left[ \log p(\xb) \right] 
+    &\geq \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log \frac{p_{}(\xb|\zb)p(\zb)}{q_{}(\zb|\xb)} \right] \enspace .
 \end{align}
 %
-This reveals an interpretation of the ELBO as the negative KL divergence between the variational approximation $q_\phib(\zb|\xb)$ and the prior distribution $p(\zb)$ plus the expected log-likelihood of the data under the generative model $p_\thetab(\xb|\zb)$.
-
-The training of VAEs is usually performed by stochastic gradient descent on the ELBO.
-This method samples a mini-batch of data points $\xb$ from the empirical training data distribution $\hat{p}(\xb)$ and then evaluates the gradient of the ELBO with respect to the parameters $\thetab$ and $\phib$ on the mini-batch.
-To gain some insight into the objective that we are optimizing, we will see if we can exploit the expectation over the data distribution $\hat{p}(\xb)$ to form a factorization similar to \cref{eq:variational-autoencoder-elbo-factorized} but with the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$ in the KL-divergence.
-It will turn out that we can indeed factor the ELBO into such a KL-divergence, a reconstruction term, and a third term \cite{tomczak_trouble_2022}.
-For ease of notation, we will let $q(\xb) = \hat{p}(\xb)$ so that $q_\phib(\zb,\xb) = q_{\phib}(\zb|\xb)\hat{p}(\xb)$.
+where we have defined $q(\xb) = \hat{p}(\xb)$ and $q_{}(\xb,\zb) = q_{}(\zb|\xb)\hat{p}(\xb)$ and dropped the parameter subscripts for ease of notation. 
+This allows us to focus on the KL-divergence in \cref{eq:variational-autoencoder-elbo-factorized} which we can rewrite as follows,
 %
-\begin{align}
-    \mathbb{E}_{\hat{p}(\xb)}\left[ \log p(\xb) \right] 
-    &= \mathbb{E}_{\hat{p}(\xb)}\left[ \log \int p_{\thetab}(\xb | \zb)\ p(\zb) \, \mathrm{d}\zb \right] \nonumber \\
-    &\geq \mathbb{E}_{\hat{p}(\xb)}\left[ \mathbb{E}_{q_{\phib}(\zb|\xb)}\left[ \log p_{\thetab}(\xb | \zb) + \log p(\zb) - \log q_{\phib}(\zb | \xb) \right] \right] \nonumber \\
-    &= \mathbb{E}_{q_{\phib}(\zb|\xb)\hat{p}(\xb)}\left[ \log p_{\thetab}(\xb | \zb) + \log p(\zb) - \log q_{\phib}(\zb | \xb) \right] \nonumber \\
-    &= \mathbb{E}_{q_{\phib}(\zb,\xb)}\left[ \log p_{\thetab}(\xb | \zb) \right] + \mathbb{E}_{q_{\phib}(\zb)}\left[ \log p(\zb) \right] \nonumber \\ 
-    &\quad - \mathbb{E}_{q_{\phib}(\zb,\xb)}\left[ \log q_{\phib}(\zb | \xb) \right]
+\begin{align} \label{eq:variational-autoencoder-kl-factor-expected}
+    \mathbb{E}_{\hat{p}(\xb)} \left[ D_\text{KL}\left( q_{}(\zb|\xb) \,||\, p(\zb) \right) \right]
+    &= \mathbb{E}_{q_{}(\zb,\xb)} \left[ \log \frac{q_{}(\zb|\xb)}{p(\zb)}\right] \nonumber \\
+    % &= - \mathbb{E}_{q_{}(\zb)}\left[ \log p(\zb) \right] + \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log q_{}(\zb | \xb) \right] \nonumber \\
+    &= \mathbb{E}_{q_{}(\zb)}\left[ \log \frac{1}{p(\zb)} \right] + \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log \frac{q_{}(\xb,\zb)}{q(\xb)} \right] \nonumber \\
+    % &= - \mathbb{E}_{q_{}(\zb)}\left[ \log p(\zb) \right] + \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log \frac{q_{}(\xb,\zb)}{q(\xb)} \right] \nonumber \\ 
+    % &\qquad + \mathbb{E}_{q_{}(\zb)}\left[ \log q_{}(\zb) \right] - \mathbb{E}_{q_{}(\zb)}\left[ \log q_{}(\zb) \right] \nonumber \\
+    &= \mathbb{E}_{q_{}(\zb)}\left[ \log \frac{q_{}(\zb)}{p(\zb)} \right] - \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log \frac{q_{}(\xb,\zb)}{q(\xb)q_{}(\zb)} \right] \nonumber \\
+    &= D_\mathrm{KL}\left( q_{}(\zb) \parallel p(\zb) \right) - I_{q_{}(\xb,\zb)}[\xb;\zb] \enspace .
 \end{align}
 %
-% where we let $q(\xb) = \hat{p}(\xb)$ for ease of notation so that $q_\phib(\zb,\xb) = q_{\phib}(\zb|\xb)\hat{p}(\xb)$ and we defined the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$.
-
-\todo[inline]{We can do this without all the special terminology. Just add and subtract E_{q(z)}[log q(z)].}
-
-We can observe that the second term $\mathbb{E}_{q_{\phib}(\zb)}\left[ \log p(\zb) \right]$ is the cross-entropy $H(q_\phib(\zb),p(\zb))$ which can also be written as $D_\text{KL}(q_\phib(\zb)||p(\zb)) + H(q_\phib(\zb))$.
-We can also observe that the third term $-\mathbb{E}_{q_{\phib}(\zb,\xb)}\left[ \log q_{\phib}(\zb | \xb) \right]$ is the conditional entropy $H_{q_\phib(\zb,\xb)}[\zb|\xb]$ which can be factored as $H_{q_\phib(\zb)}[\zb] - I_{q_\phib(\zb,\xb)}[\zb;\xb]$ where $I_{q_\phib(\zb,\xb)}[\zb;\xb]$ is the mutual information between $\zb$ and $\xb$ under the joint distribution $q_\phib(\zb,\xb)$.
-This gives us the following decomposition of the ELBO,
+Inserting this back into the ELBO in \cref{eq:variational-autoencoder-elbo-factorized}, we get,
 %
-\begin{align}
-    \mathbb{E}_{\hat{p}(\xb)}\left[ \log p(\xb) \right] 
-    &\geq \mathbb{E}_{q_{\phib}(\zb,\xb)}\left[ \log p_{\thetab}(\xb | \zb) \right] + H(q_\phib(\zb),p(\zb)) + H_{q_\phib(\zb,\xb)}[\zb|\xb]
-\end{align}
-%
-% \begin{align}
-%     a &= \mathbb{E}_{\xb,\zb \sim q_{\phib}(\zb|\xb)\hat{p}(\xb)}\left[ \log p_{\thetab}(\xb | \zb) \right]
-% \end{align}
-
-For a more background on VAEs and a comparison with self-supervised methods, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{app:paper-brief}.
-
-% VAE \cite{kingma_autoencoding_2014,rezende_stochastic_2014}
-
-% Derive ELBO
-
-% Importance sampling \cite{burda_importance_2016}
-
-% Semi-supervised learning \cite{kingma_semi-supervised_2014,maaloe_semi-supervised_2017,siddharth_learning_2017}
-
-% Trouble in paradise \cite{tomczak_trouble_2022}
-
-% Trouble in paradise: factoriing the ELBO to show that we minimize the mutual information between z and x. 
-% Use this as transition to self-supervised methods as alternative for representation learning: 
-% https://jmtomczak.github.io/blog/13/13_trouble_in_paradise.html
-
-\paragraph{Deep generative models} VAEs also belong to the class of deep generative models which are models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution.
-There exist at least six main classes of deep generative models: variational autoencoders \cite{kingma_autoencoding_2014,rezende_stochastic_2014,ranganath_hierarchical_2016,vahdat_nvae_2020,child_very_2021}, normalizing flow models \cite{dinh_nice_2015,rezende_variational_2015,dinh_density_2017,kingma_glow_2018,grathwohl_ffjord_2018}, diffusion models \cite{sohl-dickstein_deep_2015,song_generative_2019, ho_denoising_2020, vahdat_scorebased_2021}, energy-based models \cite{lecun_tutorial_2006,hinton_fast_2006,salakhutdinov_efficient_2010,du_implicit_2019}, autoregressive models \cite{oord_conditional_2016,oord_wavenet_2016,radford_improving_2018}, and generative adversarial networks \cite{goodfellow_generative_2014,arjovsky_wasserstein_2017,brock_large_2019,karras_stylebased_2019}.
-Besides data generation, common properties of deep generative models include the ability to compute an exact or approximate likelihood of input $p(\xb)$, reconstruct an input $\xb$ from its latent representation $\zb$, and represent an input $\xb$ in a latent space.
-These properties enable various methods for their use in OOD detection.
-
-\section{Speech representation learning}
-
-\iffalse
-
-% TODO Delete this probably
-Following the derivation of \textcite{jaynes_information_1957}, in the discrete entropy expression in \cref{eq:information-entropy}, we assume that the number of discrete points, $x_i, i=1,2,\dots,N$, in the support of $X$ grows larger such that in the limit $N\rightarrow\infty$, the density of points becomes a continuous function $m(x)$, and we can replace the sum with an integral.
-Denoting the number of points satisfying $a<x_i<b$ by $N_{a,b}$, we have
-%
-\begin{equation}
-    \lim_{N\rightarrow\infty} \frac{N_{a,b}}{N} = \int_a^b m(x) \, dx \enspace .
+\begin{equation} \label{eq:variational-autoencoder-elbo-factorized-expected}
+    \mathbb{E}_{q_{}(\xb,\zb)} \left[ \mathcal{L}_{\thetab,\phib}(\xb) \right] = 
+    \underset{\text{average reconstruction}}{\underbrace{
+        \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log p_{}(\xb | \zb) \right]
+    }}
+    - 
+    \underset{\text{marginal KL to prior}}{\underbrace{
+    \vphantom{\mathbb{E}_{q_{}(\xb,\zb)}\left[ \log p_{}(\xb | \zb) \right]}{
+        D_{\mathrm{KL}}\left[ q_{}(\zb) \parallel p(\zb) \right]
+    }
+    }} 
+    -
+    \underset{\text{mutual information}}{\underbrace{
+    I_{q_{}(\xb,\zb)}\left[\xb ; \zb \right]
+    }} \enspace . \nonumber
 \end{equation}
 %
-If this limit behaves sufficiently well, adjacent differences $(x_{i+1} - x_{i})$ in the neighborhood of any particular value of $x$ will also tend to zero, so that
-%
-\begin{equation} \label{eq:limiting-density-discrete-to-continuous}
-    \lim_{N\rightarrow\infty} \left[ N(x_{i+1} - x_{i}) \right] = \left[ m(x_{i}) \right]^{-1} \enspace .
-\end{equation}
-%
-In the limit, the value of the discrete probability mass function will become a continuous probability density function $w(x)$, according to the limiting form of
-%
-\begin{equation}
-    p_i = w(x_i)\left(x_{i+1} - x_{i}\right) \enspace ,
-\end{equation}
-%
-where $p_i$ is the value of the probability mass function at $x_i$ which via \cref{eq:limiting-density-discrete-to-continuous} becomes,
-%
-\begin{equation}
-    p_i = w(x_i)\left[Nm(x_i)\right]^{-1} \enspace .
-\end{equation}
-%
-Consequently, the discrete entropy expression in \cref{eq:information-entropy} becomes,
-\begin{align}
-    % H_{N,\text{discrete}}(X) &= - \int_\mathcal{X} w(x)\left[Nm(x)\right]^{-1} \, dx \, \log \left[\frac{w(x)}{Nm(x)}\right] \nonumber \\
-    % &= - \int_\mathcal{X} w(x) \, dx \, \log \left[\frac{w(x)}{Nm(x)}\right]
-    H_{N,\text{discrete}}(X) &= - \int_\mathcal{X} w(x) \, dx \, \log \left[\frac{w(x)}{Nm(x)}\right] \nonumber \\
-\end{align}
-%
-and taking the limit $N\rightarrow\infty$,
-%
-\begin{equation}
-    \lim_{N\rightarrow\infty} H_{N,\text{discrete}}(X) = H_{N} = \log N - \int_\mathcal{X} w(x) \log \left[\frac{w(x)}{m(x)}\right] \enspace .
-\end{equation}
-%
-Note that since $\lim_{N\rightarrow\infty} \log N = \infty$, this procedure suggests that the entropy in the discrete sense of a continuous random variable should be infinite.
+In this form, the KL-divergence term of \cref{eq:variational-autoencoder-elbo-factorized} has been factored into the marginal KL-divergence of the aggregated posterior to the prior and the mutual information between the data and the latent variables.
+The marginal KL-divergence is minimized by making the aggregated posterior $q_{}(\zb)$ match the prior $p(\zb)$.
+Contrary to the KL-divergence term of \cref{eq:variational-autoencoder-elbo-factorized}, driving the marginal KL-divergence to zero does not enforce independence between $\xb$ and $\zb$.
+However, this alternative form reveals that the ELBO is maximized by minizing the mutual information between $\xb$ and $\zb$, $I_{q_{}(\xb,\zb)}\left[\xb ; \zb \right]$. 
+This result indicates that the usual interpretation of $\zb$ as a representation of $\xb$ is not a fundamental property arising from the ELBO and challenges the usefulness of variational autoencoders as representation learners.
 
-Modelling paradigms
-\begin{enumerate}
-    \item Variational inference and variational autoencoders
-    \item Automatic speech recognition
-    \item Self-supervised learning
-\end{enumerate}
+% \paragraph{Hierarchical models}
+% VAEs can be extended to hierarchical models by introducing additional latent variables $\zb_1,\dots,\zb_L$ that are conditionally dependent on each other \cite{ranganath_hierarchical_2016,sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021,maaloe_biva_2019}.
+The generative model can then be written as,
+% %
+% \begin{equation}
+%     p_\thetab(\xb,\zb_1,\dots,\zb_L) = p_\thetab(\xb|\zb_1)\prod_{l=2}^L p_\thetab(\zb_{l-1}|\zb_{l}) \enspace .
+% \end{equation}
+% %
 
-\cite{kingma_autoencoding_2014}
-\cite{rezende_stochastic_2014}
+\paragraph{Applications of VAEs}
+Despite the challenges facing the training of VAEs, they have been successfully applied to a number of tasks including image generation \cite{kingma_autoencoding_2014,rezende_stochastic_2014}, image inpainting \cite{pathak_context_2016}, image super-resolution \cite{sonderby_amortised_2017,chira_image_2022}, and speech synthesis \cite{hsu_unsupervised_2017,hsu_hierarchical_2019}.
+VAEs have also proven themselves useful for semi-supervised learning \cite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}.
+Recent architectural advances have made it possible to train VAEs with a large number of latent variables \cite{sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021} and have alleviated the posterior collapse problem \cite{maaloe_biva_2019}.
 
-Two approaches to statistics:\\
-- Bayesian\\
-- Frequentist\\
-
-Two kinds of uncertainty:\\
-- Aleatoric: Uncertainty inherent to the data. Irreducible. "Known unknowns".\\
-- Epistemic: Uncertainty due to lack of knowledge. Can be reduced with more data. "Unknown unknowns".\\
-
-Approaches to uncertainty in deep learning:\\
-- Tuning of supervised classifiers to output well-calibrated probabilities that correspond to the actual likelihood of the prediction being correct (on some validation set).\\
-- Learning rich unsupervised representations of the data that can be used to estimate the uncertainty of the prediction (e.g.
-by measuring the distance to the nearest training example in the latent space).\\
-- Learning of a probability distribution over the parameters of the model, which can then be used to compute the uncertainty of the prediction (Bayesian neural networks).\\
-
-\fi
+\paragraph{Self-supervised learning}
+An alternative approach to representation learning is provided by self-supervised learning \cite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020,schneider_wav2vec_2019} which is a form of unsupervised learning where the training objective is derived from the data itself.
+While it is fair to say that not all self-supervised training objectives are as well-motivated as the ELBO is for VAEs, they have shown impressive results within the field of natural language processing \cite{devlin_bert_2018,chen_simple_2020}, speech processing \cite{schneider_wav2vec_2019}, and computer vision \cite{chen_simple_2020}.
+For a deeper comparison of VAEs and self-supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{app:paper-brief}.

### 2023-09-25 chapters/discussion.tex
Work on background and main contributions

@@
 
-\section{\Cref{chp:paper-review} revisited: \dots}
+\section{\Cref{chp:paper-review} revisited: \dots} \label{sec:discussion-paper-review}
 
-\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}
+\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?
+\cite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}
 
@@
 \lesstodo[inline]{Report the results of calibrating the stroke model using e.g. Platt scaling or Isotonic scaling.} % https://scikit-learn.org/stable/modules/calibration.html
+
+\lesstodo[inline]{Maybe mention MultiQT paper \cite{havtorn_multiqt_2020}.}

### 2023-09-25 chapters/main-contributions.tex
First complete introduction
Work on background and main contributions

@@
 
-The chapters of the thesis are self-contained studies and therefore detail their own contributions. 
+The following chapters of the thesis are self-contained studies and therefore detail their own contributions. 
 They are however written without consideration to the other chapters. 
-For that reason, this chapter will provide a high-level overview of the organization of the thesis and describe its main
-contributions in relation to the overall research topic.
+For that reason, this chapter will describe the main contributions of each chapter in relation to the overall research topic of the thesis. 
+The chapters are structured into three parts: \cref{part:unsupervised-uncertainty-estimation} is concerned with unsupervised uncertainty estimation, \cref{part:self-supervised-speech-representation-learning} is concerned with self-supervised speech representation learning, and \cref{part:medical-applications} contains studies on machine learning methods applied for tasks in a medical setting.
+Finally, \cref{part:discussion-and-conclusion} contains the discussion and conclusions of the thesis.
 
 \contribitem{part:background}
-This is the part of the thesis you are currently reading.
 \Cref{chp:introduction} provided a general introduction to the thesis and gave motivating examples for speech recognition and assistance in medical diagnostics.
-\Cref{chp:technical-background} provided the necessary technical background for the rest of the thesis by introducing uncertainty as a concept in the context of information theory and probability theory and providing a brief overview of generative latent variable models.
-\todo[inline]{Update the contributions of the technical background section accordingly.}
+\Cref{chp:technical-background} provided additional technical background not included in the individual studies by introducing uncertainty as a concept in the context of information and probability theory, introducing the task of out-of-distribution detection and reviewing existing work on the problem, and providing an introduction to variational autoencoders.
 
@@
 The first two papers are concerned with out-of-distribution detection using generative models.
-The first paper proposes a new method for out-of-distribution detection using hierarchical latent variable models.
-The second paper builds on the first and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable, explicit likelihood generative model.
-The third paper proposes a new hierarchical latent variable model for speech
+The first paper proposes a new method for out-of-distribution detection using hierarchical variational autoencoders.
+The second paper builds on the first and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable generative model with explicit likelihood. 
+The third paper proposes a new hierarchical variational autoencoder for speech and benchmarks it against other variational autoencoders and autoregressive models.
 
 \contribitem{chp:paper-hierarchical}
-This paper proposes a new method for unsupervised out-of-distribution detection using hierarchical latent variable models.
+This paper proposes a new method for unsupervised out-of-distribution detection using hierarchical variational autoencoders. 
 The paper provides evidence that the failure of deep generative models to assign lower likelihoods to data from outside the training distribution is due to overemphasis on low-level features that generalize between distributions.
 To solve this, the proposed method computes a likelihood-ratio score for out-of-distribution detection that requires data to be in-distribution across all feature-levels. 
-The proposed method is computationally efficient, fully unsupervised, and performs well on several image-based out-of-distribution detection benchmarks.
-Methods like this are important for the development of safe and reliable machine learning systems, especially in medical applications where the consequences of errors can be significant.
+The proposed method is computationally efficient, fully unsupervised, and performs well on several out-of-distribution detection benchmarks. 
+% Methods like this are important for the development of safe and reliable machine learning systems, especially in medical applications where the consequences of errors can be significant.
 
 \contribitem{chp:paper-modelagnostic}
-This paper builds on the previous paper and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable, explicit likelihood generative model.
-The method is based on combining a classical parametric test with the recently introduced typicality test using Fisher's method.
+This paper proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable, explicit likelihood generative model. 
+The method is based on combining a classical parametric test with the recently introduced typicality test via Fisher's method. 
 We show that this leads to a more accurate out-of-distribution test.
-We also discuss the benefits of casting out-of-distribution detection as a statistical testing problem, for instance enabling false positive rate control which is valuable for practical out-of-distribution detection, especially in high-risk applications such as in medical devices or services.
+We also discuss the benefits of casting out-of-distribution detection as a statistical testing problem, for instance enabling false positive rate control which is valuable for practical applications, especially in high-risk settings such as medical decision-making.
 
 \contribitem{chp:paper-benchmarking}
-This paper proposes a new hierarchical latent variable model for speech inspired by the Clockwork VAE \cite{saxena_clockwork_2021}.
-The model is benchmarked against other latent variable models and autoregressive models for speech abnd we show that using a hierarchy of latent variables improves the likelihood of the model.
-Finally, we show that the learned latent space captures information relevant for speech recognition such as phonetic content.
+This paper proposes a new hierarchical variational autoencoder for speech inspired by the Clockwork VAE \cite{saxena_clockwork_2021}. 
+The model is benchmarked against other variational autoencoders and autoregressive models for speech.
+We show that using a hierarchy of latent variables improves the likelihood of the model and that the learned latent space captures information relevant for speech recognition such as phonetic content.
 
@@
 Self-supervised learning is a promising approach to speech representation learning that has shown a fast, wide and successful adoption across speech modelling over the course of the last few years. 
-As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self-supervised representations generally outperform representations learned via probabilistic methods such as those investigated in \cref{part:unsupervised-uncertainty-estimation}.
+As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self-supervised representations generally outperform representations learned via probabilistic generative models such as those investigated in \cref{part:unsupervised-uncertainty-estimation}.
 In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self-supervised learning as an alternative paradigm to generative models for learning representations for unsupervised uncertainty estimation.
-We shall return to this in the discussion of \cref{chp:paper-review}.
-
-\todo[inline]{Update this paragraph according to the discussion of self-supervised learning for uncertainty estimation}
-% \cite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.
+We shall return to this in the discussion of \cref{chp:paper-review} in \cref{sec:discussion-paper-review}.
 
@@
 
-\todo[inline]{Should we include the brief review paper as a chapter instead of the long review paper?}
-
 \contribitem{part:medical-applications}
@@
 
-Prior to this thesis, we published related work on question tracking for calls to emergency services using a multimodal approach that combines automatic transcripts with the raw audio \cite{havtorn_multiqt_2020}.
-
 \contribitem{chp:paper-retrospective}
@@
 We used calls from Copenhagen during 2015 to 2020, to develop a machine learning-based classification pipeline.
-First, calls are transcribed by a speech recognition model and then categorised as stroke or non-stroke using a text classification model.
+First, calls were transcribed by a speech recognition model and then categorised as stroke or non-stroke using a text classification model.
 On test data from 2021, call-takers achieved an overall sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%) while the machine learning framework performed significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
@@
 \contribitem{chp:paper-automated}
+Improving the efficiency of medical coding can alleviate the administrative burden on medical staff and help reduce the rising costs for hospitals. 
+In this paper, we review the current state-of-the-art in automated medical coding of clinical notes. 
+We reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is often attributable to weak configuration, crudely sampled train-test splits, and insufficient evaluation.
+We compare models on data splits created with stratified sampling, use identical experimental setups and tune hyperparameters and decision boundaries. 
+By analyzing prediction errors, we confirm that all models struggle with rare codes, while long documents only have a negligible impact. 
+We also present the first comprehensive results on the newly released MIMIC-IV dataset using the reproduced models. 
+This paper aims to increase reproducibility and comparability of future work in automated medical coding and to provide a baseline for future research.

### 2023-09-25 chapters/paper_hierarchical.tex
Work on background and main contributions

@@
 We note that $\text{Var}(\widehat{LLR}^{>k}) = \text{Var}(\hat{\mathcal{L}}) + \text{Var}(\hat{\mathcal{L}}^{>k}) - 2\, \text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$.
-Since $\log p_\thetab(\xb)$ and part of the KL divergence are identical in the expressions of $\mathcal{L}$ and $\mathcal{L}^{>k}$ we expect $\text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$ to be positive which reduces the total variance.
+Since $\log p_\thetab(\xb)$ and part of the KL-divergence are identical in the expressions of $\mathcal{L}$ and $\mathcal{L}^{>k}$ we expect $\text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$ to be positive which reduces the total variance.
 Empirical results indeed show that $\text{Var}(\widehat{LLR}^{>k})$ is larger than $\text{Var}(\hat{\mathcal{L}})$ but smaller than $\text{Var}(\hat{\mathcal{L}}^{>k})$.

### 2023-09-25 chapters/paper_review/02_three_waves.tex
Add and revise short table and figure captions
Work on background and main contributions

@@
 
-\section{Historical Context of Representation Learning}
+\section{Historical context of representation learning}
 
@@
 (RBM)~\cite{hinton_reducing_2006}, denoising autoencoders~\cite{vincent_stacked_2010}, noise contrastive
-estimation (NCE)~\cite{gutmann_noisecontrastive_2012}, sparse coding~\cite{olshausen_emergence_1996,
+estimation (NCE)~\cite{gutmann_noisecontrastive_2010}, sparse coding~\cite{olshausen_emergence_1996,
 lee_efficient_2006, sivaram_sparse_2010}, and energy-based methods~\cite{ranzato_unified_2007}.

### 2023-09-25 chapters/paper_review/03_approaches.tex
Work on background and main contributions

@@
 The loss function measures the similarity between the anchor representation $\hat{z}_{t,k}$ and the positive $z_{t+k}$ normalized by the total similarity to the positive and negatives.}
-The approach is similar to previous work on Noise-Contrastive Estimation (NCE)~\cite{gutmann_noisecontrastive_2012}.
+The approach is similar to previous work on Noise-Contrastive Estimation (NCE)~\cite{gutmann_noisecontrastive_2010}.
 \edit{Minimizing} the loss corresponds to maximizing a lower bound on the mutual information between $h_t$ and $z_{t+k}$ (and in turn $x_{t+k-u:t+k+u}$) and is hence called InfoNCE:

### 2023-09-26 appendices/paper_brief.tex
Move table captions to top

@@
 
-\begin{abstract}
+\section*{Abstract}
 Unsupervised representation learning for speech processing has matured greatly in the last few years.
@@
 Finally, we discuss and compare models from the two categories.
-\end{abstract}
 
@@
 \textit{Left}: A model trained with self-supervised learning.
-We take these models to consist of two functions $f(\cdot)$ and $g(\cdot)$ (sec.~\ref{ssec:notation}).
+We take these models to consist of two functions $f(\cdot)$ and $g(\cdot)$ (\cref{ssec:notation}).
 After pre-training, $f(\cdot)$ is fine-tuned or used for extracting features $\mathbf{c}$.
@@
 In the following, we group previous work into \textit{self-supervised models} and \textit{probabilistic latent variable models}, and take a \emph{model} to comprise a neural architecture and a corresponding learning algorithm.
-A schematic overview is found in figure \ref{fig:ssl_lvm}.
+A schematic overview is found in \cref{fig:ssl_lvm}.
 These categories are neither exhaustive nor mutually exclusive, but allow us to focus on the characteristics that have shaped different branches of research.
@@
 For stochastic latent variable models, we use $\mathbf{z}_{1:T}$ as is customary to the field.
-While some models are frozen and produce representations used as input for downstream tasks (\textbf{\textsc{frz}}, table \ref{tab:model-taxonomy}), others are designed to be fine-tuned (\textbf{\textsc{ftn}}, table \ref{tab:model-taxonomy}).
+While some models are frozen and produce representations used as input for downstream tasks (\textbf{\textsc{frz}}, \cref{tab:model-taxonomy}), others are designed to be fine-tuned (\textbf{\textsc{ftn}}, \cref{tab:model-taxonomy}).
 In either case, we use $f(\cdot)$ to denote the model that is used for the downstream task.
@@
 \paragraph{Predictive models}
-Similar to classic autoregressive language models \cite{mikolov_recurrent_2010}, contextualized speech representations can be learned by predicting future values of a simple representation \cite{oord_representation_2018, chung_unsupervised_2019, schneider_wav2vec_2019, chung_generative_2020, jiang_further_2021} (\textbf{\textsc{prd}}, table \ref{tab:model-taxonomy}).
+Similar to classic autoregressive language models \cite{mikolov_recurrent_2010}, contextualized speech representations can be learned by predicting future values of a simple representation \cite{oord_representation_2018, chung_unsupervised_2019, schneider_wav2vec_2019, chung_generative_2020, jiang_further_2021} (\textbf{\textsc{prd}}, \cref{tab:model-taxonomy}).
 Modeling spectrograms directly, autoregressive predictive coding (APC, \citealp{chung_unsupervised_2019}) is perhaps the simplest example in this category.
@@
 Tasks that seek to predict or reconstruct the input are very common.
-In the literature, these are often jointly referred to as reconstruction tasks (\textbf{\textsc{rec}}, table \ref{tab:model-taxonomy}) \cite{liu_tera_2021, wang_unispeech_2021}, although this is somewhat misleading in the case of prediction.
+In the literature, these are often jointly referred to as reconstruction tasks (\textbf{\textsc{rec}}, \cref{tab:model-taxonomy}) \cite{liu_tera_2021, wang_unispeech_2021}, although this is somewhat misleading in the case of prediction.
 
@@
 Some work has extended APC and CPC inspired models with separate encoders operating in opposite directions \cite{ling_deep_2020, kawakami_learning_2020, borgholt_scaling_2021}, but these models are still restricted to process left and right context separately.
-Inspired by the masked language model task used for text-based representation learning \cite{devlin_bert_2018}, several papers have used masking to overcome this challenge (\textbf{\textsc{msk}}, table \ref{tab:model-taxonomy}).
+Inspired by the masked language model task used for text-based representation learning \cite{devlin_bert_2018}, several papers have used masking to overcome this challenge (\textbf{\textsc{msk}}, \cref{tab:model-taxonomy}).
 Masking refers to replacing parts of the input with zeros or a learned masking vector.
@@
 This allows for complete data utilization, but limits the amount of context encoded in the learned representation.
-Figure \ref{fig:ssl_grid} summarizes the models discussed so far.
+\cref{fig:ssl_grid} summarizes the models discussed so far.
 
@@
 \paragraph{Quantization} 
-Several models enforce a discrete latent space by quantizing the vector representation (\textbf{\textsc{qtz}}, table \ref{tab:model-taxonomy}).
+Several models enforce a discrete latent space by quantizing the vector representation (\textbf{\textsc{qtz}}, \cref{tab:model-taxonomy}).
 The two most popular approaches are the Gumbel-softmax \cite{jang_categorical_2016, maddison_concrete_2017} and the quantization used in the VQ-VAE \cite{oord_neural_2018}.
@@
 As this function is non-differentiable, we must rely on the straight-through gradient estimator \cite{bengio_estimating_2013} which assumes that the Jacobian ${\partial \, \varphi}/{\partial \, \mathbf{p}}$ equals the identity matrix.
-The one-hot vector can then be used for a codebook lookup to obtain the final quantized vector (e.g., $\mathbf{q}_t$ in eq.~\ref{eq_brief: w2v2 qtz}).
+The one-hot vector can then be used for a codebook lookup to obtain the final quantized vector (e.g., $\mathbf{q}_t$ in \cref{eq_brief: w2v2 qtz}).
 
-The wav2vec 2.0 quantization module (eq.~\ref{eq_brief: w2v2 qtz}) uses the Gumbel softmax.
-To ensure utilization of codebook vectors, a diversity loss is added to the task specific loss (eq.
-\ref{eq_brief: w2v2 loss})
+The wav2vec 2.0 quantization module (\cref{eq_brief: w2v2 qtz}) uses the Gumbel softmax.
+To ensure utilization of codebook vectors, a diversity loss is added to the task specific loss (\cref{eq_brief: w2v2 loss})
 %
@@
 The models covered so far learn representations that maintain a temporal resolution proportional to the input resolution.
-We say that they learn local representations (\textbf{\textsc{loc}}, table \ref{tab:model-taxonomy}).
-Now, we cover models that learn global representations (\textbf{\textsc{glb}}, table \ref{tab:model-taxonomy}).
+We say that they learn local representations (\textbf{\textsc{loc}}, \cref{tab:model-taxonomy}).
+Now, we cover models that learn global representations (\textbf{\textsc{glb}}, \cref{tab:model-taxonomy}).
 
@@
 Given a segment corresponding to a single word $\mathbf{x}_{(n)} = \mathbf{x}_{t_{n}:t_{n+1}}$, the skipgram model is trained to predict neighboring words $\mathbf{x}_{(n + k)}$ where $k\ne0$.
-That is, instead of a single decoder, as in eq.
-\ref{eq_brief:dec-aw2v}, the skipgram model employs multiple decoders
+That is, instead of a single decoder, as in \cref{eq_brief:dec-aw2v}, the skipgram model employs multiple decoders
 %
@@
 \paragraph{Other work}
-Some models learn local representations with a variable temporal resolution that is not proportional to the input resolution (\textbf{\textsc{var}}, table \ref{tab:model-taxonomy}).
+Some models learn local representations with a variable temporal resolution that is not proportional to the input resolution (\textbf{\textsc{var}}, \cref{tab:model-taxonomy}).
 In practice, this is often achieved implicitly, by learning segment boundaries or by taking repeated quantized values to belong to the same segment \cite{kamper_unsupervised_2021, chorowski_unsupervised_2019, michel_blind_2017, kreuk_selfsupervised_2020, wang_gate_2017, dieleman_variablerate_2021}.
@@
 
-Most of the work presented so far fits neatly into the taxonomy presented in table \ref{tab:model-taxonomy}.
+Most of the work presented so far fits neatly into the taxonomy presented in \cref{tab:model-taxonomy}.
 One exception is the problem-agnostic speech encoder (PASE, \citealp{pascual_learning_2019, ravanelli_multitask_2020}) that combines multiple pre-training tasks.
@@
 For instance, wav2vec 2.0 and related models have been applied to learn cross-lingual and multi-lingual representations \cite{riviere_unsupervised_2020, conneau_unsupervised_2020, khurana_magic_2021} and proven well-suited for concurrently learning with labeled data \cite{talnikar_joint_2021, wang_unispeech_2021}.
-%We will discuss the downstream application of self-supervised representations further in section \ref{sec:eval}.
+%We will discuss the downstream application of self-supervised representations further in \cref{sec:eval}.
 
@@
 Fundamental to LVMs is the assumption that the data is produced by a generative process that involves unobserved stochastic latent variables $\textbf{z}$. 
-An LVM aims to model this generative process to enable generation of new data $\mathbf{x}$ (\textbf{\textsc{gen}}, table \ref{tab:model-taxonomy}) and inference of the latent variable associated with a given observed variable $\textbf{x}$.
+An LVM aims to model this generative process to enable generation of new data $\mathbf{x}$ (\textbf{\textsc{gen}}, \cref{tab:model-taxonomy}) and inference of the latent variable associated with a given observed variable $\textbf{x}$.
 For representation learning, the inference of latent variables is of primary interest.
@@
 In the VAE framework, the observation model $p(\mathbf{x}|\mathbf{z})$ is parameterized using a deep neural network.
-This choice allows modeling complex and high-dimensional data but also makes the integral in eq.~\ref{eq_brief: lvm log-likelihood} analytically intractable.
+This choice allows modeling complex and high-dimensional data but also makes the integral in \cref{eq_brief: lvm log-likelihood} analytically intractable.
 MCMC methods can be used to estimate it and the true model posterior $p(\mathbf{z}|\mathbf{x})$, but these methods are usually computationally expensive in this setting \cite{mohamed_monte_2019}.
@@
 It approximates the intractable true model posterior by introducing a variational posterior distribution $q(\mathbf{z}|\mathbf{x})$, also parameterized by a deep neural network.
-From eq.~\ref{eq_brief: lvm log-likelihood}, via Jensen's inequality, this gives rise to a variational lower bound on the likelihood, also known as the evidence lower bound (ELBO).
+From \cref{eq_brief: lvm log-likelihood}, via Jensen's inequality, this gives rise to a variational lower bound on the likelihood, also known as the evidence lower bound (ELBO).
 \begin{equation}
@@
 \caption{A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable.
-The observation, prior and inference models may all belong to one or more of the categories listed under them as detailed in section \ref{sec:plvms}.
+The observation, prior and inference models may all belong to one or more of the categories listed under them as detailed in \cref{sec:plvms}.
 The types listed here serve as primitives from which more complex structures can be constructed including models with hierarchies of multiple latent variables.}
@@
 \begin{sidewaystable*}[t]
-    \caption{Selected latent variable models classified according the attributes defined throughout section \ref{sec:plvms}.
-See table~\ref{tab:lvm-model-primitives} for the probability distributions that correspond to each of the attribute short-hands.
+    \caption{Selected latent variable models classified according the attributes defined throughout \cref{sec:plvms}.
+See \cref{tab:lvm-model-primitives} for the probability distributions that correspond to each of the attribute short-hands.
 \textbf{\textsc{hie}} indicates a hierarchical representation.}
@@
 We next review VAEs applied to speech. We consider the choices of observation, prior and inference models.
-We provide a model taxonomy for selected LVMs in table~\ref{tab:lvm-taxonomy}.
+We provide a model taxonomy for selected LVMs in \cref{tab:lvm-taxonomy}.
 
 \paragraph{Observation models} 
-A common choice for the observation model $p(\mathbf{x}|\mathbf{z})$ is to include an autoregressive dependency on the observed variable (\textbf{\textsc{arx}}, table \ref{tab:lvm-model-primitives}) that is, $p(\mathbf{x}_t|\mathbf{x}_{1:t-1}, \cdot)$ where $\cdot$ represents some dependency on the latent variable \cite{chung_recurrent_2015, fraccaro_sequential_2016, oord_neural_2018, oord_representation_2018}.
+A common choice for the observation model $p(\mathbf{x}|\mathbf{z})$ is to include an autoregressive dependency on the observed variable (\textbf{\textsc{arx}}, \cref{tab:lvm-model-primitives}) that is, $p(\mathbf{x}_t|\mathbf{x}_{1:t-1}, \cdot)$ where $\cdot$ represents some dependency on the latent variable \cite{chung_recurrent_2015, fraccaro_sequential_2016, oord_neural_2018, oord_representation_2018}.
 This allows the latent representation to focus on correlations that cannot easily be predicted from the observed variable at previous time-steps \cite{oord_representation_2018}.
@@
 Some LVMs do not use autoregressive observation models \cite{ebbers_hidden_2017, glarner_full_2018, hsu_unsupervised_2017, hsu_learning_2017, khurana_factorial_2019, khurana_convolutional_2020}.
-These more closely follow the assumption of \emph{local independence} which states that observed variables are conditionally independent given the local (\textbf{\textsc{loc}}, table \ref{tab:lvm-model-primitives}) and/or global (\textbf{\textsc{glb}}, table \ref{tab:lvm-model-primitives}) latent variables \cite{bartholomew_latent_2011}.
+These more closely follow the assumption of \emph{local independence} which states that observed variables are conditionally independent given the local (\textbf{\textsc{loc}}, \cref{tab:lvm-model-primitives}) and/or global (\textbf{\textsc{glb}}, \cref{tab:lvm-model-primitives}) latent variables \cite{bartholomew_latent_2011}.
 However, this forces the latent variable to encode details about the observed variable to achieve a good reconstruction.
@@
 \paragraph{Priors}
-Priors can be said to belong to one or more of four broad categories. See table~\ref{tab:lvm-model-primitives}. 
-Priors that are autoregressive on the observed variable (\textbf{\textsc{arx}}, table \ref{tab:lvm-model-primitives}) take the form $p(\mathbf{z}_t|\mathbf{x}_{1:t-1})$.
+Priors can be said to belong to one or more of four broad categories. See \cref{tab:lvm-model-primitives}. 
+Priors that are autoregressive on the observed variable (\textbf{\textsc{arx}}, \cref{tab:lvm-model-primitives}) take the form $p(\mathbf{z}_t|\mathbf{x}_{1:t-1})$.
 This generally results in a slow-down of the generative process which may be of concern if the use-case is data generation.
-Priors that are autoregressive on the latent variable (\textbf{\textsc{arz}}, table \ref{tab:lvm-model-primitives}) take the form $p(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ and enable stochastic temporal transitions similar to hidden Markov models but with potentially nonlinear transition functions \cite{chung_recurrent_2015, fraccaro_sequential_2016, khurana_factorial_2019, khurana_convolutional_2020}.
-Locally independent priors (\textbf{\textsc{ind}}, table \ref{tab:lvm-model-primitives}) are rarely applied to sequential latent variables since they make the prior latent dynamics independent of the value of previous latent variables.
+Priors that are autoregressive on the latent variable (\textbf{\textsc{arz}}, \cref{tab:lvm-model-primitives}) take the form $p(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ and enable stochastic temporal transitions similar to hidden Markov models but with potentially nonlinear transition functions \cite{chung_recurrent_2015, fraccaro_sequential_2016, khurana_factorial_2019, khurana_convolutional_2020}.
+Locally independent priors (\textbf{\textsc{ind}}, \cref{tab:lvm-model-primitives}) are rarely applied to sequential latent variables since they make the prior latent dynamics independent of the value of previous latent variables.
 Models that do impose such priors on sequential latents are quite limited in their generative power, unless they learn the prior dynamics post-hoc as done in the VQ-VAE \cite{oord_neural_2018}.
-Global latent variables (\textbf{\textsc{glb}}, table \ref{tab:lvm-model-primitives}) are fundamentally limited in the amount of information they can encode.
+Global latent variables (\textbf{\textsc{glb}}, \cref{tab:lvm-model-primitives}) are fundamentally limited in the amount of information they can encode.
 Hence, models usually use them in combination with another local latent variable, or to encode fixed length input segments \cite{khurana_factorial_2019, hsu_learning_2017, hsu_unsupervised_2017}.
@@
 For this reason, all inference models covered here are conditioned on the observed sequence in some way. Generally, the inference model can be seen as solving either a filtering or smoothing problem.
-In filtering (\textbf{\textsc{flt}}, table \ref{tab:lvm-model-primitives}), the latent variables are assumed to depend only on past and current values of the observed variable, $q(\mathbf{z}_t|\mathbf{x}_{1:t})$ \cite{chung_recurrent_2015, khurana_convolutional_2020}.
-In global smoothing (\textbf{\textsc{gsm}}, table \ref{tab:lvm-model-primitives}), this causal dependency is replaced with a dependency on all observed values, $q(\mathbf{z}_t|\mathbf{x}_{1:T})$ \cite{fraccaro_sequential_2016, hsu_learning_2017}.
-Smoothing can also be done locally (\textbf{\textsc{lsm}}, table \ref{tab:lvm-model-primitives}), where the latent variables then depend on $\mathbf{x}_{t-r:t+r}$ for some integer $r>0$ \cite{oord_neural_2018}.
+In filtering (\textbf{\textsc{flt}}, \cref{tab:lvm-model-primitives}), the latent variables are assumed to depend only on past and current values of the observed variable, $q(\mathbf{z}_t|\mathbf{x}_{1:t})$ \cite{chung_recurrent_2015, khurana_convolutional_2020}.
+In global smoothing (\textbf{\textsc{gsm}}, \cref{tab:lvm-model-primitives}), this causal dependency is replaced with a dependency on all observed values, $q(\mathbf{z}_t|\mathbf{x}_{1:T})$ \cite{fraccaro_sequential_2016, hsu_learning_2017}.
+Smoothing can also be done locally (\textbf{\textsc{lsm}}, \cref{tab:lvm-model-primitives}), where the latent variables then depend on $\mathbf{x}_{t-r:t+r}$ for some integer $r>0$ \cite{oord_neural_2018}.
 Compared to self-supervised models that often use transformer encoders it can be hypothesized that global smoothing offers a stronger case than local smoothing and filtering for representation learning.
 
-The inference model may also be used to infer a global latent variable (\textbf{\textsc{glb}}, table \ref{tab:lvm-model-primitives}) that might encode global information about $\mathbf{x}$. 
+The inference model may also be used to infer a global latent variable (\textbf{\textsc{glb}}, \cref{tab:lvm-model-primitives}) that might encode global information about $\mathbf{x}$. 
 While it must be included in the prior model it might not be in the observation model, if the model also has a local latent variable. 
 Finally, latent variables are often made to depend autoregressively on past inferred values, e.g.
-$q(\mathbf{z}_t|\mathbf{z}_{1:t-1},\mathbf{x}_{1:t})$ (\textbf{\textsc{arz}}, table \ref{tab:lvm-model-primitives}) \cite{chung_recurrent_2015, fraccaro_sequential_2016}.
+$q(\mathbf{z}_t|\mathbf{z}_{1:t-1},\mathbf{x}_{1:t})$ (\textbf{\textsc{arz}}, \cref{tab:lvm-model-primitives}) \cite{chung_recurrent_2015, fraccaro_sequential_2016}.
 
 \paragraph{Multiscale and hierarchical models} 
-Some work has explored using a hierarchy of latent variables (\textbf{\textsc{hie}}, table \ref{tab:lvm-taxonomy}).
+Some work has explored using a hierarchy of latent variables (\textbf{\textsc{hie}}, \cref{tab:lvm-taxonomy}).
 This allows encoding the inductive bias that speech contains information at different temporal scales by letting the latent variables operate at different temporal scales \cite{hsu_unsupervised_2017}.
@@
 \paragraph{From global to local} 
-In table \ref{tab:model-taxonomy}, we see that work on global representations within self-supervised learning precedes work on local representations.
+In \cref{tab:model-taxonomy}, we see that work on global representations within self-supervised learning precedes work on local representations.
 However, we find that the core ideas underlying the recent successes in learning local representation models have also been used for global representation learning; masking \cite{chung_audio_2016}, context prediction \cite{chung_speech2vec_2018}, and contrastive training \cite{milde_unspeech_2018} have been applied in both settings.

### 2023-09-26 appendices/supplementary_hierarchical.tex
Some formatting and changes to labels
Move table captions to top

--- 2023-08-08 appendices/appendix_hierarchical.tex
@@
 
-\chapter[supplementary material for hierarchical vaes know what they don't know]{Supplementary material for Hierarchical VAEs know what they don't know}
-\label{chp:appendix-hierarchical}
+\chapter[supplementary material: hierarchical vaes know what they don't know]{Supplementary material: Hierarchical VAEs know what they don't know}
+\label{app:supplementary-hierarchical}
 \ifthenelse{\equal{\skipappendices}{true}}{}{
 
-\section{Datasets}
-\cref{tab_hierarchical:datasets-overview} lists the datasets used in the paper. We use the predefined train/test splits for the datasets.
+\section{Datasets} \label{sec_hierarchical:datasets}
+\Cref{tab_hierarchical:datasets-overview} lists the datasets used in the paper. We use the predefined train/test splits for the datasets.
 
@@
 \begin{table}[h!]
+    \caption{Overview of the used datasets.}
     \centering
@@
     % }
-    \caption{Overview of the used datasets.}
     \label{tab_hierarchical:datasets-overview}
@@
 
-\section{Model details}\label{sec:model-details}
+\section{Model details}\label{sec_hierarchical:model-details}
 In \cref{tab_hierarchical:hyperparameters} we specify the hyperparameters used when training our models.
@@
 \begin{table}[h]
+    \caption[Selection of most important hyperparameters]{Selection of most important hyperparameters and their setting.
+Convolutional kernels are square and latent dimensions are given without spatial dimensions which are given in the text.
+See \cref{sec_hierarchical:model-details} for more details.}
     \centering
@@
     % }
-    \caption[Selection of most important hyperparameters]{Selection of most important hyperparameters and their setting.
-Convolutional kernels are square and latent dimensions are given without spatial dimensions which are given in the text.
-See \cref{sec:model-details} for more details.}
     \label{tab_hierarchical:hyperparameters}
@@
 
-\section{Analysis of the influence of latent variables on the marginal likelihood}
+\section{Analysis of the influence of latent variables on the marginal likelihood} \label{sec_hierarchical:analysis}
 In the paper, we argue that the lowest level latent variables, which have the highest dimensionality, contribute the most to the approximate likelihood.
@@
 
-\section{Derivation of the \texorpdfstring{$\mathcal{L}^{>k}$}{Lk} bound}
+\section{Derivation of the \texorpdfstring{$\mathcal{L}^{>k}$}{Lk} bound} \label{sec_hierarchical:derivation_L_geq_k}
 In this section we present the derivation of $\mathcal{L}^{>k}$ and show that it is a lower bound on the marginal likelihood.
@@
 
-\section{The complementary \texorpdfstring{$\mathcal{L}^{<l}$}{Ll} bound}
+\section{The complementary \texorpdfstring{$\mathcal{L}^{<l}$}{Ll} bound} \label{sec_hierarchical:complementary-bound}
 We can generalize the $\mathcal{L}^{>k}$ bound by introducing the flipped version, $\mathcal{L}^{<l}$, which compared to $\mathcal{L}^{>k}$, instead samples the $L-l$ \textit{highest} latent variables in the hierarchy from the prior $\zb_{l},\dots,\zb_L \sim p_\theta(\zb_{\geq l})=p_\theta(\zb_l|\zb_{l+1})\cdots p_\theta(\zb_L)$ and the remaining lower latents from the approximate posterior $\hat{\zb}_{1},\dots,\hat{\zb}_{l-1} \sim q_\phi(\zb_{<l}|\xb)=q_\phi(\zb_{1}|\xb)q_\phi(\zb_{2}|\zb_1)\cdots q_\phi(\zb_{l-1}|\zb_{l-2})$,
@@
     &\qquad + \mathbb{E}_{q(\zb_1|\xb)}\left[ \mathbb{E}_{q(\zb_2|\zb_1)}\left[\log \frac{p(\zb_1|\zb_2)}{q(\zb_1|\xb)}\right]\right] \notag \\
-    &\qquad + \mathbb{E}_{q(\zb_1|\xb)}\left[-D_\text{KL}[q(\zb_2|\zb_1)||p(\zb_2)]\right] \notag \ .
+    &\qquad + \mathbb{E}_{q(\zb_1|\xb)}\left[-D_\text{KL}[q(\zb_2|\zb_1)\parallel p(\zb_2)]\right] \notag \ .
 \end{align}
@@
         \log p(\xb) &\ge \mathbb{E}_{q(\zb_1, \zb_2|\xb)} \left[ \log p(\xb|\zb_1) \right] \\
-    &\qquad - D_\text{KL}[q(\zb_2|\xb)||p(\zb_2)] \notag \\ 
-    &\qquad - \mathbb{E}_{q(\zb_2|\xb)}\left[D_\text{KL}[q(\zb_1|\zb_2)||p(\zb_1|\zb_2)\right] \notag \ .
+    &\qquad - D_\text{KL}[q(\zb_2|\xb)\parallel p(\zb_2)] \notag \\ 
+    &\qquad - \mathbb{E}_{q(\zb_2|\xb)}\left[D_\text{KL}[q(\zb_1|\zb_2)\parallel p(\zb_1|\zb_2)\right] \notag \ .
 \end{align}
@@
 
-\section{Additional results}
+\section{Additional results} \label{sec_hierarchical:additional-results}
 We provide additional results for a model trained on FashionMNIST in \cref{tab_hierarchical:additional-results-fashionmnist}, a model trained on MNIST in \cref{tab_hierarchical:additional-results-mnist}, a model trained on CIFAR10 in \cref{tab_hierarchical:additional-results-cifar} and a model trained on SVHN in \cref{tab_hierarchical:additional-results-svhn}.
@@
 \begin{table}[t]
+    \caption[Additional results for the HVAE model trained on SVHN]{Additional results for the HVAE model trained on SVHN. All results computed with 1000 importance samples.}
     \centering
@@
     % }
-    \caption[Additional results for the HVAE model trained on SVHN]{Additional results for the HVAE model trained on SVHN. All results computed with 1000 importance samples.}
     \label{tab_hierarchical:additional-results-svhn}
@@
 \begin{table}[t]
+    \caption[Additional results for the HVAE model trained on CIFAR10]{Additional results for the HVAE model trained on CIFAR10. All results computed with 1000 importance samples.}
     \centering
@@
     % }
-    \caption[Additional results for the HVAE model trained on CIFAR10]{Additional results for the HVAE model trained on CIFAR10. All results computed with 1000 importance samples.}
     \label{tab_hierarchical:additional-results-cifar}
@@
 \begin{table}[t]
+    \caption[Additional results for the HVAE model trained on FashionMNIST]{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
     \centering
@@
     % }
-    \caption[Additional results for the HVAE model trained on FashionMNIST]{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
     \label{tab_hierarchical:additional-results-fashionmnist}
@@
 \begin{table}[t]
+    \caption[Additional results for the HVAE model trained on MNIST]{Additional results for the HVAE model trained on MNIST. All results computed with 1000 importance samples.}
     \centering
@@
     % }
-    \caption[Additional results for the HVAE model trained on MNIST]{Additional results for the HVAE model trained on MNIST. All results computed with 1000 importance samples.}
     \label{tab_hierarchical:additional-results-mnist}

### 2023-09-26 chapters/background.tex
Move table captions to top

@@
 
-Recently, several approaches for deep neural networks have been developed that formally address the rejection of samples $\xb\notin\mathcal{D}_{\text{in}}$. 
+Recently, several approaches for deep neural networks have been developed that address the rejection of samples $\xb\notin\mathcal{D}_{\text{in}}$. 
 One way to categorize the different approaches is based on whether the underlying model is a classifier that estimates a conditional probability distribution $p(y|\xb)$ over some target variable $y$, or a model that learns a probability density $p(\xb)$ over the input itself.

### 2023-09-26 chapters/introduction.tex
Move table captions to top

@@
 
-While the quest to ease human decision-making has been ongoing for thousands of years, it was not until 1642 that the first mechanical calculator was successfully designed by Blaise Pascal.
+While efforts to ease human decision-making have been ongoing for thousands of years, it was not until 1642 that the first mechanical calculator was successfully designed by Blaise Pascal.
 Refined by Gottfried Leibniz during the latter half of the 17th century, the mechanical calculator marked the beginning of a new era where rudimentary calculations could be automated and performed with greater speed and accuracy than humans.
@@
 Finally, \cref{part:discussion-and-conclusion} concludes the thesis by summarizing the main contributions and discussing the presented work and future directions for research.
-
-\iffalse
-
-Uncertainty is a fundamental part of human experience.
-The ability to understand and quantify uncertainty is crucial in making informed decisions, drawing meaningful conclusions, and assessing the reliability of predictions made from observations.
-Yet, capturing this ability in a mathematical model is a difficult task.
-
-Uncertainty estimation has long been a central topic in the field of statistics.
-It has traditionally been addressed through methods like confidence intervals, hypothesis testing, and Bayesian inference.
-Confidence intervals provide a range of plausible values for a parameter, while hypothesis testing evaluates the significance of differences between groups or variables \cite{blitzstein_introduction_2019}.
-Bayesian inference, on the other hand, offers a powerful framework to incorporate prior knowledge and update beliefs based on observed data using probability theory.
-These approaches have been extensively studied and applied in various domains, from social sciences to natural sciences, to improve the robustness and generalizability of statistical analyses \cite{gelman_bayesian_2013}.
-However, the widespread adoption in diverse applications of machine learning, and in particular deep neural networks, has posed new challenges in dealing with uncertainty.
-
-Machine learning algorithms are designed to learn patterns from data and make predictions or decisions based on those patterns.
-Since many applications of machine learning are concerned with real-world phenomena, it is important to understand the uncertainty associated with the predictions made by such a model.
-For instance, in medical diagnosis, knowing the uncertainty in a model's prediction is crucial when the consequences of a false positive or false negative can be significant.
-In speech recognition, knowing that a certain word was hard to transcribe for the model can help avoid misinterpretation of the transcribed text.
-
-% The performance of machine learning models is highly dependent on the quality and quantity of data used for training.
-In many cases, the data used to train machine learning models are collected from the real world, and therefore, may contain biases and errors which models may learn to mirror.
-In addition, the complexity of modern machine learning models makes it difficult to understand how they arrive at their predictions.
-This is especially true for deep neural networks, which are often referred to as black-box models.
-The lack of interpretability makes it difficult to assess the reliability of the model's predictions.
-
-% Modern speech processing relies on high-performance parallel computing [40, 157], large volumes of data [136, 213], and years of innovation in model design [109, 266].
-
-% 40: High performance convolutional neural networks for document processing \cite{chellapilla_high_2006}
-% 157: Imagenet classification with deep convolutional neural networks \cite{krizhevsky_imagenet_2012}
-% 136: Libri-light: A benchmark for asr with limited or no supervision \cite{kahn_libri-light_2020}
-% 213: Librispeech: an ASR corpus based on public domain audio books \cite{panayotov_librispeech_2015}
-% 109:  Long short-term memory \cite{hochreiter_long_1997}
-% 266: Attention Is All You Need \cite{vaswani_attention_2017}
-
-\section{Machine learning and societal risk}
-\todo[inline]{Discuss the role of uncertainty aware models in the context of risk.}
-\cite{europeancommission_briefing_2021}
-
-\section{Motivation: Speech recognition}
-
-\section{Motivation: Automated medical coding}
-
-\subsection{Uncertainty in machine learning}
-
-\todo[inline]{Revise this section with less focus on individual methods and more focus on overall trends.}
-
-In the context of machine learning, uncertainty is often represented by a probability distribution.
-The most common approach is to use Bayesian methods, where uncertainty is captured by posterior distributions over model parameters and predictions \cite{gelman_bayesian_2013}.
-Bayesian neural networks, for instance, offer a powerful framework to model uncertainty in deep learning architectures \cite{neal_bayesian_1995}.
-By incorporating prior beliefs and updating them based on observed data, these networks can produce probabilistic predictions that provide a measure of uncertainty.
-This is particularly useful in applications where high-confidence predictions are required, and the consequences of errors can be significant.
-
-Recently, another interesting approach to uncertainty estimation in machine learning is the use of Monte Carlo Dropout \cite{gal_dropout_2016}.
-Monte Carlo Dropout leverages the idea of dropout regularization, originally employed during training to prevent overfitting.
-To form a prediction with associated uncertainty, Monte Carlo Dropout proposes to make multiple forward passes through the network while sampling different dropout masks for each one.
-This leads to obtaining a distribution of outputs for each input sample and the variance among these sampled predictions serves as a measure of uncertainty.
-It is computationally efficient and can be easily incorporated into existing deep learning architectures, making it an attractive choice for uncertainty estimation.
-
-Furthermore, there has been an interest in ensemble methods for uncertainty estimation \cite{lakshminarayanan_simple_2017}.
-Ensembles combine the predictions of multiple models to obtain a more robust and calibrated uncertainty measure.
-Bagging and boosting techniques, which have been widely used in the field of statistics, have found their way into the realm of machine learning for uncertainty estimation as well.
-By training multiple models with different initialization or using diverse learning algorithms, ensembles can capture different sources of uncertainty, thereby providing a comprehensive assessment of the overall uncertainty in the predictions.
-
-% We compare our framework with several major related lines of work.
-First, our work focuses on the quantification of epistemic uncertainty, which refers to the errors coming from the inadequacy of the model or data noises.
-This is different from aleatoric uncertainty, which refers to the intrinsic stochasticity of the problem [87, 91, 112, 59, 34], or predictive uncertainty which captures the sum of epistemic and aleatoric uncertainties (but not their dissection) [89, 92, 12, 3, 22].
-Regarding epistemic uncertainty, a related line of study is deep ensemble that aggregates predictions from multiple independent training replications [76, 70, 38, 8, 89].
-This approach, as we will make clear later, can reduce and potentially quantify procedural variability, but a naive use would require demanding retraining effort and does not address data variability.
-Another line is the Bayesian UQ approach on neural networks [41, 2].
-This regards network weights as parameters subject to common priors such as Gaussian.
-Because of the computation difficulties in exact inference, an array of studies investigate efficient approximate inference approaches to estimate the posteriors [40, 46, 16, 31, 30, 90, 74, 53].
-While powerful, these approaches nonetheless possess inference error that could be hard to quantify, and ultimately finding rigorous guarantees on the performance of these approximate posteriors remains open to our best knowledge.
-
-% 87: 
-% 91: 
-% 112:
-% 59: 
-% 34:
-
-% 89: 
-% 92: 
-% 12:
-% 3: 
-% 22:
-
-% 76: 
-% 70: 
-% 38: 
-% 8: 
-% 89: X
-
-% 41: 
-% 2:
-
-% 40:
-% 46:
-% 16:
-% 31:
-% 30:
-% 90:
-% 74:
-% 53:
-
-\iffalse
-
-General challenges in uncertainty estimation:
-\begin{enumerate}
-    \item \textbf{Accuracy}: The uncertainty estimation method should be accurate.
-    \item \textbf{Interpretability}: The uncertainty estimation method should be interpretable.
-    \item \textbf{Robustness}: The uncertainty estimation method should be robust to adversarial attacks.
-    \item \textbf{Applicability}: The uncertainty estimation method should be applicable to a wide range of models without requiring extensive modifications.
-    \item \textbf{Efficiency}: The uncertainty estimation method should be computationally efficient at training and prediction time.
-    \item \textbf{Scalability}: The uncertainty estimation method should scale to large datasets.
-\end{enumerate}
-
-Uncertainty is a fundamental part of human experience.
-Yet, capturing it in a mathematical model is a difficult task.
-Historically, uncertainty estimation has been a central topic in the field of statistics.
-While classical statistical methods have provided valuable insights, they often assume well-defined probability distributions and may not fully capture the complexities of uncertainty in real-world scenarios.
-Moreover, they tend to rely heavily on parametric assumptions, which might not hold true for complex data with high-dimensional features and non-linear relationships.
-Therefore, in recent years, researchers have turned their attention towards more flexible and data-driven approaches to model uncertainty in various domains.
-
-In the field of machine learning, uncertainty is often represented by a probability distribution.
-The most common approach is to use Bayesian methods, where uncertainty is captured by posterior distributions over model parameters and predictions.
-Bayesian neural networks, for instance, offer a powerful framework to model uncertainty in deep learning architectures.
-By incorporating prior beliefs and updating them based on observed data, these networks can produce probabilistic predictions that provide a measure of uncertainty.
-This is particularly useful in applications where high-confidence predictions are required, and the consequences of errors can be significant.
-
-Recently, another promising approach to uncertainty estimation in machine learning is the use of Monte Carlo Dropout (MC Dropout).
-MC Dropout leverages the idea of dropout regularization, originally employed during training to prevent overfitting.
-During prediction, MC Dropout takes multiple forward passes through the network with dropout enabled, which leads to obtaining a distribution of outputs for each input sample.
-The variance among these sampled predictions serves as a measure of uncertainty.
-MC Dropout has shown remarkable success in various tasks such as image classification, object detection, and natural language processing.
-It is computationally efficient and can be easily incorporated into existing deep learning architectures, making it an attractive choice for uncertainty estimation.
-
-Furthermore, there has been a surge of interest in ensemble methods for uncertainty estimation.
-Ensembles combine the predictions of multiple models to obtain a more robust and calibrated uncertainty measure.
-Bagging and boosting techniques, which have been widely used in the field of statistics, have found their way into the realm of machine learning for uncertainty estimation as well.
-By training multiple models with different initializations or using diverse learning algorithms, ensembles can capture different sources of uncertainty, thereby providing a comprehensive assessment of the overall uncertainty in the predictions.
-
-Despite the progress made in uncertainty estimation for machine learning models, challenges still remain.
-One key concern is the interpretability of uncertainty measures.
-While probabilistic outputs are more informative than point estimates, effectively communicating uncertainty to end-users and decision-makers is not trivial.
-Developing visualization techniques and intuitive explanations for uncertainty is an ongoing area of research.
-Additionally, quantifying uncertainty in high-dimensional data or complex models with massive amounts of parameters requires careful consideration and efficient computational strategies.
-
-In conclusion, uncertainty estimation is a crucial aspect of machine learning models, enabling them to make more informed and reliable predictions.
-Researchers continue to explore innovative techniques to better represent and quantify uncertainty, allowing for safer and more trustworthy deployment of machine learning systems in real-world scenarios.
-
-Uncertainty is a fundamental part of human experience.
-Yet, capturing it in a mathematical model is a difficult task.
-Historically, uncertainty estimation has been a central topic in the field of statistics.
-Understanding and quantifying uncertainty is crucial in making informed decisions, drawing meaningful conclusions, and assessing the reliability of predictions and inferences derived from data.
-
-In the field of statistics, uncertainty is traditionally addressed through methods like confidence intervals, hypothesis testing, and Bayesian inference.
-Confidence intervals provide a range of plausible values for a parameter, while hypothesis testing evaluates the significance of differences between groups or variables.
-Bayesian inference, on the other hand, offers a powerful framework to incorporate prior knowledge and update beliefs based on observed data using probability theory.
-These approaches have been extensively studied and applied in various domains, from social sciences to natural sciences, to improve the robustness and generalizability of statistical analyses.
-However, the advent of machine learning and its widespread adoption in diverse applications has posed new challenges and opportunities in dealing with uncertainty.
-
-Machine learning algorithms are designed to learn patterns from data and make predictions or decisions based on that knowledge.
-In many cases, it is not enough to provide a single deterministic prediction; it is equally important to convey the level of uncertainty associated with the prediction.
-For instance, in medical diagnosis, knowing the uncertainty in a model's prediction is crucial when the consequences of a false positive or false negative can be significant.
-In autonomous driving systems, understanding uncertainty becomes essential to ensure the safety of passengers and pedestrians.
-
-To address the uncertainty challenge in machine learning, researchers have developed various techniques to represent and quantify uncertainty.
-One of the common approaches is the use of probabilistic models, where uncertainty is captured in the form of probability distributions.
-Instead of providing a single point prediction, probabilistic models generate a range of possible outcomes along with their associated probabilities.
-Bayesian neural networks, Gaussian processes, and variational autoencoders are some examples of such models that have gained prominence in the quest for uncertainty-aware machine learning.
-
-Another avenue for dealing with uncertainty in machine learning is through ensemble methods.
-Ensemble learning combines multiple models to improve predictive accuracy and can also provide uncertainty estimates.
-By aggregating the outputs of diverse models, ensemble methods can identify situations where predictions are consistent among the models, leading to higher confidence, and cases where the models disagree, indicating higher uncertainty.
-
-As the intersection of statistics and machine learning grows increasingly important, the integration of these two domains has become an active area of research.
-The incorporation of probabilistic modeling and Bayesian approaches into machine learning methods opens up new possibilities for developing more robust, interpretable, and calibrated models.
-Moreover, the synergy of statistical techniques with deep learning, reinforcement learning, and other advanced machine learning paradigms has the potential to push the boundaries of uncertainty quantification and application even further.
-
-In this thesis, we aim to explore and contribute to the field of uncertainty in statistics and machine learning.
-We will investigate state-of-the-art methods for uncertainty estimation, delve into their theoretical underpinnings, and evaluate their performance in various real-world scenarios.
-By gaining a deeper understanding of uncertainty and its implications, we aspire to pave the way for more reliable, trustworthy, and accountable AI systems in the future.
-
-In the following chapters, we will begin by reviewing the relevant literature on uncertainty in both statistics and machine learning, providing a comprehensive overview of the current landscape.
-We will then present our proposed methodologies and empirical studies, illustrating how uncertainty quantification can enhance decision-making processes and mitigate risks in practical applications.
-Finally, we will conclude with a discussion of our findings, potential limitations, and future research directions in this exciting and rapidly evolving field.
-
-\fi
-\fi

### 2023-09-26 chapters/main-contributions.tex
Move table captions to top

@@
 \contribitem{part:background}
-\Cref{chp:introduction} provided a general introduction to the thesis and gave motivating examples for speech recognition and assistance in medical diagnostics.
-\Cref{chp:technical-background} provided additional technical background not included in the individual studies by introducing uncertainty as a concept in the context of information and probability theory, introducing the task of out-of-distribution detection and reviewing existing work on the problem, and providing an introduction to variational autoencoders.
+\Cref{chp:introduction} provides a general introduction to the thesis and gives motivating examples for speech recognition and assistance in medical diagnostics.
+\Cref{chp:technical-background} provides additional technical background not included in the individual studies by introducing uncertainty as a concept in the context of information and probability theory, introducing the task of out-of-distribution detection and reviewing existing work on the problem, and providing an introduction to variational autoencoders.
 

### 2023-09-26 chapters/paper_hierarchical.tex
Move table captions to top

@@
 \subsection{A bound for semantic OOD detection}
-If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \cref{fig_hierarchical:reconstructions-fashionmnist}-\ref{fig_hierarchical:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb_1)$ will be high for both in- and out-of-distribution data.
+If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \crefrange{fig_hierarchical:reconstructions-fashionmnist}{fig_hierarchical:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb_1)$ will be high for both in- and out-of-distribution data.
 Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb_1)] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ from \cref{eq_hierarchical:elbo} relies on the KL-term for OOD detection.
@@
 In general, the absolute log-ratios grow with $\mathrm{dim}(\zb_i)$ as the individual log probability terms are computed by summing over the dimensionality of $\zb_i$.
-This means that the value of the KL-term is dominated by terms where $\zb_i$ is high-dimensional. We refer to Appendix C for a more detailed argument.
+This means that the value of the KL-term is dominated by terms where $\zb_i$ is high-dimensional. We refer to \cref{sec_hierarchical:analysis} for a more detailed argument.
 Since hierarchical VAEs are generally constructed with a bottleneck type structure, the terms corresponding to latent variables towards the top of the hierarchy will have a vanishing influence on the value of the KL-term.
@@
 \end{equation}
-where $k\in\{0,1,\dots,L\}$ (see Appendix for the derivation).
+where $k\in\{0,1,\dots,L\}$ (see \cref{sec_hierarchical:derivation_L_geq_k} for the derivation).
 We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq_hierarchical:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
@@
 Clearly, if this is not the case and the difference instead lies in low-level statistics, the scores would likely fail.
-We hypothesize that a complementary bound to \cref{eq_hierarchical:biva->k}, $\mathcal{L}^{<l}$ described in Appendix E, might be useful in these cases, but leave further examination to future work.
+We hypothesize that a complementary bound to \cref{eq_hierarchical:biva->k}, $\mathcal{L}^{<l}$ described in \cref{sec_hierarchical:complementary-bound}, might be useful in these cases, but leave further examination to future work.
 
@@
 We use the standard train/test splits for the datasets.
-More details on the datasets can be found in the Appendix.
+More details on the datasets can be found in the \cref{sec_hierarchical:datasets}.
 
@@
 % For CIFAR/SVHN, we also train a BIVA model \cite{maaloe_biva_2019} with $L=10$ and similar configuration as used by the original paper\footnote{Code available at \url{github.com/larsmaaloee/BIVA} and \url{github.com/vlievin/biva-pytorch}}.
-Full model details are in the Appendix.
+Full model details are in the \cref{sec_hierarchical:model-details}.
 
@@
 We note that for all our results, the value of the score ($\mathcal{L}^{>k}$ and $LLR^{>k}$) for the training and test splits of the in-distribution data was observed to have the same empirical distribution to within sampling error hence yielding an AUROC score of $\approx0.5$ as expected.
-Results on additional commonly used datasets are found in Appendix G.
+Results on additional commonly used datasets are found in \cref{sec_hierarchical:additional-results}.
 
 \begin{table}[t!]
+    \caption[Average bits per dimension of different datasets for generative models trained on FashionMNIST and CIFAR10.]{%
+    Average bits per dimension of different datasets for models trained on FashionMNIST and CIFAR10.
+    For the hierarchical models we include the $\mathcal{L}^{>k}$ bounds.
+    The likelihoods of training and test splits of the in-distribution data are all cases close.
+    Since we train on dynamically binarized FashionMNIST, our bits/dim are smaller than for Glow.
+    As $k$ is increased for the $L^{>k}$ bound, the bound gets looser but the model eventually assigns higher likelihood to the in distribution data than to the OOD data.
+    Glow refers to \textcite{kingma_glow_2018, nalisnick_deep_2019}.
+    BIVA refers to our implementation of \textcite{maaloe_biva_2019}.}
     \centering
@@
     }
-    \caption[Average bits per dimension of different datasets for generative models trained on FashionMNIST and CIFAR10.]{%
-    Average bits per dimension of different datasets for models trained on FashionMNIST and CIFAR10.
-    For the hierarchical models we include the $\mathcal{L}^{>k}$ bounds.
-    The likelihoods of training and test splits of the in-distribution data are all cases close.
-    Since we train on dynamically binarized FashionMNIST, our bits/dim are smaller than for Glow.
-    As $k$ is increased for the $L^{>k}$ bound, the bound gets looser but the model eventually assigns higher likelihood to the in distribution data than to the OOD data.
-    Glow refers to \textcite{kingma_glow_2018, nalisnick_deep_2019}.
-    BIVA refers to our implementation of \textcite{maaloe_biva_2019}.}
     \label{tab_hierarchical:bits-per-dim-ood}
@@
 \begin{table}[t!]
+    \caption[AUROC, AUPRC, and FPR80 of generative models for OOD detection (MNIST/FashionMNIST and SVHN/CIFAR10).]{%
+        AUROC$\uparrow$, AUPRC$\uparrow$ and FPR80$\downarrow$ for OOD detection for a FashionMNIST model using scores on the FashionMNIST test set as reference.
+We bold the best results within the "No OOD-specific assumptions" group since we only compare directly to those.
+        HVAE (ours) refers to our hierarchical bottom-up VAE.
+        BIVA (ours) refers to our implementation of the hierarchical BIVA model \cite{maaloe_biva_2019}.
+    }
     \centering
@@
     \end{tabular}%
-    }
-    \caption[AUROC, AUPRC, and FPR80 of generative models for OOD detection (MNIST/FashionMNIST and SVHN/CIFAR10).]{%
-        AUROC$\uparrow$, AUPRC$\uparrow$ and FPR80$\downarrow$ for OOD detection for a FashionMNIST model using scores on the FashionMNIST test set as reference.
-We bold the best results within the "No OOD-specific assumptions" group since we only compare directly to those.
-        HVAE (ours) refers to our hierarchical bottom-up VAE.
-        BIVA (ours) refers to our implementation of the hierarchical BIVA model \cite{maaloe_biva_2019}.
     }
@@
 This is not surprising since the value of $k$ that provides the maximal separation to the reference in-distribution dataset need not be the one for which $\mathcal{LLR}^{>k}$ is overall maximal for the OOD dataset.
-We also visualize the ROC curves resulting from using the $LLR^{>k}$ score for OOD detection on both FashionMNIST/MNIST and CIFAR10/SVHN and compare it to the ROC curves resulting from the different $\mathcal{L}^{>k}$ bounds in Figures \ref{fig_hierarchical:FMNIST-roc-llr} and \ref{fig_hierarchical:CIFAR10-roc-llr}, respectively.
+We also visualize the ROC curves resulting from using the $LLR^{>k}$ score for OOD detection on both FashionMNIST/MNIST and CIFAR10/SVHN and compare it to the ROC curves resulting from the different $\mathcal{L}^{>k}$ bounds in \cref{fig_hierarchical:FMNIST-roc-llr, fig_hierarchical:CIFAR10-roc-llr}, respectively.
 On both datasets we see significantly better discriminatory performance when using the $LLR^{>k}$ score.

### 2023-09-26 chapters/paper_retrospective.tex
Move table captions to top

@@
 %
-Calls to the MH-1813 were divided into training, validation, and test subsets and calls to the emergency line 1-1-2 were only used as supplementary training data (Table 1).
-Calls from the test year (2021) that were not associated with a diagnostic category code, which we used to evaluate call-taker performance, were separated from our primary test set, but still included to assess potential bias in this group of calls (2021 w/o category, Table 1).
-The 1-1-2 training data differed from the MH-1813 data regarding age, male/female ratio, and stroke prevalence (Table 1).
+Calls to the MH-1813 were divided into training, validation, and test subsets and calls to the emergency line 1-1-2 were only used as supplementary training data (\cref{tab_retrospective:table1-population-characteristics}).
+Calls from the test year (2021) that were not associated with a diagnostic category code, which we used to evaluate call-taker performance, were separated from our primary test set, but still included to assess potential bias in this group of calls (2021 w/o category, \cref{tab_retrospective:table1-population-characteristics}).
+The 1-1-2 training data differed from the MH-1813 data regarding age, male/female ratio, and stroke prevalence (\cref{tab_retrospective:table1-population-characteristics}).
 We therefore performed an ablation study where 1-1-2 data were not used for training to assess whether this difference negatively impacted model performance.
@@
 
-The classification model outperformed the call-takers (Table 21), with significant differences in all metrics (p < 0.0001).
+The classification model outperformed the call-takers (\cref{tab_retrospective:table2-main-results}), with significant differences in all metrics (p < 0.0001).
 Excluding the 1-1-2 call line training data significantly degraded the model's performance (p < 0.0001), despite the domain mismatch with the MH-1813 call line test data.

### 2023-09-26 chapters/paper_review/03_approaches.tex
Move table captions to top

@@
 In the standard BERT masking policy, each token is masked independently at random.
-However, for speech, masking a single \edit{sample or spectrogram frame results in a largely} trivial reconstruction task \edit{since, as discussed in paragraph \ref{par:apc}, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore it is common to mask chunks of consecutive frames~\cite{liu_mockingjay_2020,jiang_further_2021}.
+However, for speech, masking a single \edit{sample or spectrogram frame results in a largely} trivial reconstruction task \edit{since, as discussed in \cref{par:apc}, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore it is common to mask chunks of consecutive frames~\cite{liu_mockingjay_2020,jiang_further_2021}.
 

### 2023-09-28 chapters/background.tex
Work on discussion and conclusion, plotting

@@
 
-\subsection{Variational autoencoders} 
+\subsection{Variational autoencoders} \label{sec:variational-autoencoders}
 To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to parameterize the generative model $p_\thetab(\xb,\zb)$ with a neural network with parameters $\thetab$.
@@
 VAEs have also proven themselves useful for semi-supervised learning \cite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}.
-Recent architectural advances have made it possible to train VAEs with a large number of latent variables \cite{sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021} and have alleviated the posterior collapse problem \cite{maaloe_biva_2019}.
+Recent architectural advances have made it possible to train VAEs with many latent variables \cite{sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021} and have alleviated the posterior collapse problem \cite{maaloe_biva_2019}.
 
@@
 For a deeper comparison of VAEs and self-supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{app:paper-brief}. 
+\Cref{chp:paper-review} provides a comprehensive introduction to and review of self-supervised learning in speech recognition.

### 2023-09-28 chapters/conclusion.tex
Work on background
Work on discussion and conclusion, plotting

--- 2023-08-11 chapters/conclusion.tex
@@
 \chapter[conclusions and outlook]{Conclusions and outlook}\label{chp:conclusion}
-
 % ~3 pages
 
-A conclusion with some abbreviations and citations, \gls{VAE} \cite{kingma_autoencoding_2014,rezende_stochastic_2014}.
+In the introductory \cref{chp:introduction} we laid out how, through human history, progress-driving technological development has always introduced new challenges and risks of error whether via misuse, misunderstanding or inherent limitations.
+We emphasized that machine learning comes with these same types of risk may in some cases may amplify their impacts and argued that uncertainty estimation is a key component in developing safe and reliable machine learning systems.
+We provided the motivational cases of recognition of stroke cases in emergency calls \cref{subsec:motivation-stroke-recognition} and automated medical coding \cref{subsec:motivation-medical-coding} and gave examples of potential sources of uncertainty and of how correctly quantifying uncertainty can improve the usefulness of such tools in practice.
+Finally, we ended the introduction by connecting machine learning reliability \cref{sec:machine-learning-reliability} and model calibration \cref{subsec:model-calibration} and defining different types of uncertainty \cref{subsec:types-of-uncertainty}.
 
-\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}
+\Cref{chp:technical-background} provided technical background only covered briefly by the individual studies.
+We first introduced uncertainty as a concept in the context of information and probability theory \cref{sec:information-theory}.
+Then, we defined the task of out-of-distribution detection and reviewed existing work on the problem \cref{\cref{sec:out-of-distribution-detection}}.
+Finally, we provided technical background for variational autoencoders \cref{sec:variational-autoencoders}.
+
+\dots

### 2023-09-28 chapters/discussion.tex
Work on discussion and conclusion, plotting

@@
 \chapter[discussion]{Discussion}\label{chp:discussion}
-
 % ~5 pages
@@
 Include reference to \cite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
-\lesstodo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimzing a mutual information term in the ELBO (derive this form).}
+\lesstodo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimizing a mutual information term in the ELBO (derive this form).}
 \lesstodo[inline]{Discuss some recent work e.g. \cref{morningstar_density_2021}}
@@
 \cite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}
+\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}
+% Since the main focus within self-supervised learning has been on improving downstream task performance, very limited work, if any, has investigated self-supervised representations in terms of uncertainty estimation.
+% However, in the context of medical applications where data can be abundant but labels are sparse, unsupervised uncertainty estimation is a very interesting direction for future work.
+\lesstodo[inline]{OOD data: Generalization or detection (https://arxiv.org/pdf/2110.11334.pdf)?}
 
-\section{\Cref{chp:paper-retrospective} revisited: \dots}
+\section{\Cref{chp:paper-retrospective} revisited: Uncertainty}
+The work in \textcite{wenstrup_retrospective_2023} deals with predictive performance and analyses feature importance but does not explicitly consider uncertainty estimation.
+
+\begin{figure}
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{python_plotting/calibration_curve_ensemble_and_all_models_uncalibrated.pdf}
+        \caption{}
+        \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{python_plotting/histogram_ensemble_and_single_model.pdf}
+        \caption{}
+        \label{fig_hierarchical:histogram_ensemble_and_single_model}
+    \end{subfigure}
+\end{figure}
 
@@
 \lesstodo[inline]{Report the results of calibrating the stroke model using e.g. Platt scaling or Isotonic scaling.} % https://scikit-learn.org/stable/modules/calibration.html
-
 \lesstodo[inline]{Maybe mention MultiQT paper \cite{havtorn_multiqt_2020}.}

### 2023-09-28 chapters/introduction.tex
Work on discussion and conclusion, plotting

@@
 The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics.
-\cite[photo credit]{neugebauer_mathematical_1945}}
+\cite[photo credit][]{neugebauer_mathematical_1945}}
     \label{fig:plimpton_332}
@@
     \caption{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.
-\cite[photo credit]{usarmyresearchlaboratoryarltechnicallibrary_female_1940}} %Photo: \copyright Corbis/ Getty Images.}
+\cite[photo credit][]{usarmyresearchlaboratoryarltechnicallibrary_female_1940}} %Photo: \copyright Corbis/ Getty Images.}
     \label{fig:eniac_programmers}
@@
     \caption{A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
-The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \cite[photo credit]{nationaltransportationsafetyboardnhsa_collision_2019}.}
+The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \cite[photo credit][]{nationaltransportationsafetyboardnhsa_collision_2019}.}
     % \caption{Aerial view of crash location showing path of pedestrian as she attempted to cross N.
@@
 
-\subsection{Stroke recognition in emergency calls}
+\subsection{Stroke recognition in emergency calls} \label{subsec:motivation-stroke-recognition}
 %
@@
 
-\subsection{Medical coding of clinical notes}
+\subsection{Medical coding of clinical notes} \label{subsec:motivation-medical-coding}
 %
@@
 
-\section{Machine learning reliability}
+\section{Machine learning reliability} \label{sec:machine-learning-reliability}
 %
@@
 
-In practice, miscalibrated models are not reliable in high risk applications.
+In practice, miscalibrated models are not reliable, especially in high risk applications.
 Although no model is perfect, the inability of a human, or another model, to assess the certainty with which a prediction was made, makes it impossible to know how to weigh it in a decision-making process.
@@
 
-\subsection{Model calibration}
+\subsection{Model calibration} \label{subsec:model-calibration}
 % 
@@
 
-\subsection{Understanding uncertainty}
+\subsection{Understanding uncertainty} \label{subsec:understanding-uncertainty}
 % 
@@
 Methods for estimating epistemic uncertainty include Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995} which represent it by an explicit distribution over learned model parameters.
-Ensembling methods \cite{gal_dropout_2016,lakshminarayanan_simple_2017} take a similar approach but use an implicit distribution.
-Other approaches include anomaly detection and the recent field of out-of-distribution detection which for example represent epistemic uncertainty by special output classes for out-of-distribution data, distributions over data representations, or distances between them (see \cref{sec:out-of-distribution-detection}).
+Ensemble methods \cite{gal_dropout_2016,lakshminarayanan_simple_2017} take a similar approach but use an implicit distribution.
+Other approaches include anomaly detection and the recent field of out-of-distribution detection which, for example, represent epistemic uncertainty by special output classes for out-of-distribution data, distributions over data representations, or distances between them (see \cref{sec:out-of-distribution-detection}).
 
@@
 \Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper.
-\Cref{part:unsupervised-uncertainty-estimation} contains papers on variational autoencoders and their applications to estimation of epistemic uncertainty and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
+\Cref{part:unsupervised-uncertainty-estimation} contains papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
 \Cref{part:self-supervised-speech-representation-learning} contains a review paper on self-supervised speech representation learning \cite{mohamed_selfsupervised_2022}.

### 2023-09-28 chapters/main-contributions.tex
Work on discussion and conclusion, plotting

@@
 \contribitem{part:background}
-\Cref{chp:introduction} provides a general introduction to the thesis and gives motivating examples for speech recognition and assistance in medical diagnostics.
+\Cref{chp:introduction} provides a general introduction to the thesis and gives motivating examples for speech recognition and assistance in medical encounters including stroke recognition on medical helplines and automation of medical coding.
 \Cref{chp:technical-background} provides additional technical background not included in the individual studies by introducing uncertainty as a concept in the context of information and probability theory, introducing the task of out-of-distribution detection and reviewing existing work on the problem, and providing an introduction to variational autoencoders.
@@
 This part of the thesis is concerned with unsupervised uncertainty estimation and consists of three papers. 
-The first two papers are concerned with out-of-distribution detection using generative models.
+The first two papers are approach unsupervised uncertainty estimation through the task of out-of-distribution detection using generative models.
 The first paper proposes a new method for out-of-distribution detection using hierarchical variational autoencoders.

### 2023-09-29 chapters/discussion.tex
Updates

@@
 \section{\Cref{chp:paper-retrospective} revisited: Uncertainty}
-The work in \textcite{wenstrup_retrospective_2023} deals with predictive performance and analyses feature importance but does not explicitly consider uncertainty estimation.
+The work in \textcite{wenstrup_retrospective_2023} focuses on predictive performance and an analysis of feature importance but does not explicitly consider uncertainty estimation. To
 
@@
         \includegraphics[width=1\columnwidth]{python_plotting/calibration_curve_ensemble_and_all_models_uncalibrated.pdf}
-        \caption{}
-        \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
+        % \caption{}
+        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
     \end{subfigure}
@@
         \includegraphics[width=1\columnwidth]{python_plotting/histogram_ensemble_and_single_model.pdf}
-        \caption{}
-        \label{fig_hierarchical:histogram_ensemble_and_single_model}
+        % \caption{}
+        % \label{fig_discussion:histogram_ensemble_and_single_model}
     \end{subfigure}
+    \caption[Calibration curve for the uncalibrated stroke recognition model and empirical distribution of predicted probabilities.]{\subref{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated} \Cref{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated} shows the calibration curve for the uncalibrated ensemble model with median F1-score also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.
+\subref{fig_discussion:histogram_ensemble_and_single_model} Histogram of predicted probabilities.}
 \end{figure}
+
+% \subsectionP{Ensembling technique}
+% To form an ensemble from several models, we need a method to form a single prediction from the outputs of the individual models.
+% To allow us to evaluate an impact score for each word feature in our ablation study, we wanted our ensemble to provide not only predictions but also associated probability scores on a continuous scale.
+A majority voting scheme would only provide as many different scores as models, in our case five, which is arguably too coarse a resolution.
 

### 2023-09-29 chapters/introduction.tex
Updates

@@
 
+\section{Research hypotheses}
+
+\todo[inline]{Formulate research hypotheses}
+
 \section{Thesis scope and outline}
@@
 The remainder of \cref{part:background} consists of two chapters.
-\Cref{chp:technical-background} provides relevant technical background to the papers of the thesis, including uncertainty, out-of-distribution detection, and variational autoencoders while \cref{chp:main-contributions} outlines the main contributions of the thesis in relation to the published papers and the research project as a whole.
+\Cref{chp:technical-background} provides relevant technical background to the papers of the thesis, including uncertainty, out-of-distribution detection, and variational autoencoders while \cref{chp:main-contributions} outlines the main contributions of the thesis via the published papers in relation to the research hypotheses.
 

### 2023-09-29 chapters/paper_automated.tex
Updates

@@
     \centering
-    \begin{subfigure}[b]{0.8\textwidth}
+    \begin{subfigure}[b]{0.6\textheight}
         \centering
-        \includegraphics[width=\linewidth]{paper_automated/threshold_tuning_micro_full.png}
+        \includegraphics[width=\textwidth]{paper_automated/threshold_tuning_micro_full.png}
         % \caption{}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.8\textwidth}
+    \begin{subfigure}[b]{0.6\textheight}
         \centering
-        \includegraphics[width=\linewidth]{paper_automated/threshold_tuning_macro_full.png}
+        \includegraphics[width=\textwidth]{paper_automated/threshold_tuning_macro_full.png}
         % \caption{}
@@
     \centering
-    \begin{subfigure}[b]{0.8\textwidth}
+    \begin{subfigure}[b]{0.6\textheight}
         \centering
-        \includegraphics[width=\linewidth]{paper_automated/train_size_f1_micro.png}
+        \includegraphics[width=\textwidth]{paper_automated/train_size_f1_micro.png}
         % \caption{}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.8\textwidth}
+    \begin{subfigure}[b]{0.6\textheight}
         \centering
-        \includegraphics[width=\linewidth]{paper_automated/train_size_f1_macro.png}
+        \includegraphics[width=\textwidth]{paper_automated/train_size_f1_macro.png}
         % \caption{}

### 2023-09-29 chapters/paper_benchmarking.tex
Write introduction, revise title and some formatting
Move table captions to top
Updates

@@
 \begin{equation}
-    \mathcal{L}(\xb) = ~ \mathbb{E}_{q(\zb_{1:T}|\xb_{1:T})} \bigg[ \sum_{t} \log p(\xb_t|\xb_{<t},\zb_{\leq t}) - D_\text{KL}\left( q(\zb_t|\xb_{\leq t},\zb_{<t}) \, || \, p(\zb_t|\xb_{<t},\zb_{<t}) \right)  \bigg] \label{eq: vrnn elbo} \enspace .
+    \mathcal{L}(\xb) = ~ \mathbb{E}_{q(\zb_{1:T}|\xb_{1:T})} \bigg[ \sum_{t} \log p(\xb_t|\xb_{<t},\zb_{\leq t}) - D_\text{KL}\left( q(\zb_t|\xb_{\leq t},\zb_{<t}) \parallel p(\zb_t|\xb_{<t},\zb_{<t}) \right)  \bigg] \label{eq: vrnn elbo} \enspace .
 \end{equation}
@@
 % \begin{split}
-%     \mathcal{L}(\xb) =&~ \mathbb{E}_{q(\zb_{1:T}|\xb_{1:T})} \bigg[ \sum_{t} \log p(\xb_t|\xb_{<t},\zb_{\leq t}) \\ & - \text{KL}\left( q(\zb_t|\xb_{\leq t},\zb_{<t}) \, || \, p(\zb_t|\xb_{<t},\zb_{<t}) \right)  \bigg] \enspace .
+%     \mathcal{L}(\xb) =&~ \mathbb{E}_{q(\zb_{1:T}|\xb_{1:T})} \bigg[ \sum_{t} \log p(\xb_t|\xb_{<t},\zb_{\leq t}) \\ & - \text{KL}\left( q(\zb_t|\xb_{\leq t},\zb_{<t}) \parallel p(\zb_t|\xb_{<t},\zb_{<t}) \right)  \bigg] \enspace .
 %     \label{eq: vrnn elbo}
@@
 \begin{table}[t]
+    \caption[Model likelihoods and phoneme error rate for TIMIT.]{We report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines (left) and phoneme error rate (PER) of different representations for phoneme recognition on TIMIT (right).}
     \begin{subtable}[t]{.45\textwidth}
@@
     \end{adjustbox}
-    \vspace{1mm}
-    \caption{}
-    \vspace{-1mm}
+    % \vspace{1mm}
+    % \caption{}
+    % \vspace{-1mm}
     \label{tab: likelihoods timit}
@@
     \end{adjustbox}
-    \vspace{1mm}
-    \caption{}
-    \vspace{-1mm}
+    % \vspace{1mm}
+    % \caption{}
+    % \vspace{-1mm}
     \label{tab: phoneme recognition (PER)}
     \end{subtable}%
-    \caption[Model likelihoods and phoneme error rate for TIMIT.]{In \cref{tab: likelihoods timit} we report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines.
-    In \cref{tab: phoneme recognition (PER)} we report phoneme error rate (PER) of different representations for phoneme recognition on TIMIT. 
-    }
 \end{table}
@@
 \begin{table*}[t]
+    \caption[Model likelihoods on LibriSpeech test sets.]{
+    Model likelihoods $\mathcal{L}$ on LibriSpeech test sets represented as $\SI{16}{bit}$ $\mu$-law encoded PCM.
+    For the CW-VAE, $s$ refers to $s_1$ and the two-layered models have $s_2=8s_1$.
+    The models are trained on either the $\SI{10}{h}$ LibriLight subset or the $\SI{100}{h}$ LibriSpeech \textrm{train-clean-100} subset as indicated.
+    Likelihoods are given in units of bits per frame (bpf).
+    }
     \centering
@@
     \end{adjustbox}
-    \caption[Model likelihoods on LibriSpeech test sets.]{
-    Model likelihoods $\mathcal{L}$ on LibriSpeech test sets represented as $\SI{16}{bit}$ $\mu$-law encoded PCM.
-    For the CW-VAE, $s$ refers to $s_1$ and the two-layered models have $s_2=8s_1$.
-    The models are trained on either the $\SI{10}{h}$ LibriLight subset or the $\SI{100}{h}$ LibriSpeech \textrm{train-clean-100} subset as indicated.
-    Likelihoods are given in units of bits per frame (bpf).
-    }
     \label{tab: likelihoods librispeech}

### 2023-09-29 chapters/paper_retrospective.tex
Work on discussion and conclusion, plotting
Updates

@@
 
-The receiver operating characteristic (ROC) curve (\cref{fig_retrospective:figure1}, left) illustrates the potential to increase the sensitivity while maintaining a FPR lower than or equal to that of the call-takers.
-Similarly, the PPV sensitivity curve (\cref{fig_retrospective:figure1}, right) demonstrates that sensitivity can be improved while retaining a PPV higher than that of the call-takers.
-The framework can thus be tuned to a sensitivity of around 73\%, while still having a higher positive predictive value than the human call-taker (\cref{fig_retrospective:figure1}, right) The ensemble model outperformed the individual models regardless of the threshold, except for one that exhibited a slightly better sensitivity at a high FPR exceeding 1.5\%.
-The confusion matrices (\cref{fig_retrospective:figure2}) illustrate the performance differences in absolute numbers, with the model exhibiting more true positives and fewer false positives than the call-takers.
+The receiver operating characteristic (ROC) curve (\cref{fig_retrospective:figure1-roc-curve}, left) illustrates the potential to increase the sensitivity while maintaining a FPR lower than or equal to that of the call-takers.
+Similarly, the PPV sensitivity curve (\cref{fig_retrospective:figure1-roc-curve}, right) demonstrates that sensitivity can be improved while retaining a PPV higher than that of the call-takers.
+The framework can thus be tuned to a sensitivity of around 73\%, while still having a higher positive predictive value than the human call-taker (\cref{fig_retrospective:figure1-roc-curve}, right) The ensemble model outperformed the individual models regardless of the threshold, except for one that exhibited a slightly better sensitivity at a high FPR exceeding 1.5\%.
+The confusion matrices (\cref{fig_retrospective:figure2-prediction-confusion-matrices}) illustrate the performance differences in absolute numbers, with the model exhibiting more true positives and fewer false positives than the call-takers.
 
@@
     \centering
-    \caption[Overall stroke recognition performance of model compared to call takers.]{Overall performance on MH-1813 test data, performance without 1-1-2 training data, and performance on data from 2021 without diagnostic categories as well as performance on MH-1813 based on demographic subgroups (age/sex) [mean (95\% CI)].}
+    \caption[Overall stroke recognition performance of model compared to call takers.]{Overall performance on MH-1813 test data, performance without 1-1-2 training data, and performance on data from 2021 without diagnostic categories as well as performance on MH-1813 test data based on demographic subgroups (age/sex) [mean (95\% CI)].}
     \label{tab_retrospective:table2-main-results}
@@
 Models 1-5 are the individual models that make up the ensemble model.}
-    \label{fig_retrospective:figure1}
+    \label{fig_retrospective:figure1-roc-curve}
 \end{figure}
@@
 Numbers for the model are given as the rounded mean over eleven runs.}
-    \label{fig_retrospective:figure2}
+    \label{fig_retrospective:figure2-prediction-confusion-matrices}
 \end{figure}
@@
 
-We performed an occlusion analysis to evaluate the importance of individual words for both positive and negative classifier predictions \cref{tab_retrospective:table3}
+We performed an occlusion analysis to evaluate the importance of individual words for both positive and negative classifier predictions \cref{tab_retrospective:table3-occlusion-analysis}
 Among the words with a positive rank score, several words are synonymous with stroke, such as ``blood clot", ``haemorrhagic stroke", and ``stroke".
@@
     \caption[Words with the largest positive and negative ranking score in calls predicted as stroke and non-stroke, respectively.]{English translation of words with the largest positive and negative ranking score in calls predicted as stroke and non-stroke, respectively.}
-    \label{tab_retrospective:table3}
+    \label{tab_retrospective:table3-occlusion-analysis}
     \resizebox*{0.98\textwidth}{!}{%
@@
 
-\subsubsection{Copenhagen Emergency Medical Services (CEMS)}
+\paragraph{Copenhagen Emergency Medical Services (CEMS)}
 
@@
 
-\subsubsection{Danish Stroke Registry (DanStroke)}
+\paragraph{Danish Stroke Registry (DanStroke)}
 
@@
 
-\subsubsection{Inclusion and ethics}
+\paragraph{Inclusion and ethics}
 
@@
 
-\subsubsection{Cross-referencing data sources}
+\paragraph{Cross-referencing data sources}
 
@@
 
-\subsubsection{Dataset splitting}
+\paragraph{Dataset splitting}
 
@@
 
-\subsubsection{Speech recognition}
+\paragraph{Speech recognition}
 
@@
 
-\subsubsection{Text classification}
+\paragraph{Text classification}
 
@@
 
-\subsubsection{Threshold calibration and ensembling}
+\paragraph{Threshold calibration and ensembling}
 
@@
 
-\subsubsection{Model training}
+\paragraph{Model training}
 

### 2023-09-29 appendices/paper_brief.tex
Updates to citation commands and footnotes

@@
 \section{Introduction}
-Representation learning has shaped modern computer vision \cite{simonyan_very_2014} and natural language processing \cite{devlin_bert_2018}, and more recently speech processing has been subject to the same development \cite{baevski_wav2vec_2020}.
-Representation learning has been defined as \emph{``learning representations of the data that make it easier to extract useful information when building classifiers or other predictors"} \cite{bengio_representation_2013}.
+Representation learning has shaped modern computer vision \parencite{simonyan_very_2014} and natural language processing \parencite{devlin_bert_2018}, and more recently speech processing has been subject to the same development \parencite{baevski_wav2vec_2020}.
+Representation learning has been defined as \emph{``learning representations of the data that make it easier to extract useful information when building classifiers or other predictors"} \parencite{bengio_representation_2013}.
 Unsupervised representation learning is concerned with learning useful representations without the use of human annotations.
@@
 The model is then fine-tuned, or used to extract input representations for a smaller model, targeting a task with limited training data.
-In computer vision, both supervised \cite{simonyan_very_2014, szegedy_going_2015, he_deep_2015} and unsupervised \cite{pathak_context_2016, doersch_unsupervised_2015} representation learning have gained attention with supervised representation learning driven by the availability of large annotated datasets \cite{deng_imagenet_2009}.
+In computer vision, both supervised \parencite{simonyan_very_2014, szegedy_going_2015, he_deep_2015} and unsupervised \parencite{pathak_context_2016, doersch_unsupervised_2015} representation learning have gained attention with supervised representation learning driven by the availability of large annotated datasets \parencite{deng_imagenet_2009}.
 For text and speech, pre-training is usually unsupervised as labeled data is difficult to obtain.
@@
 
-Self-supervised learning is a subset of unsupervised learning \cite{tsai_selfsupervised_2021}.
+Self-supervised learning is a subset of unsupervised learning \parencite{tsai_selfsupervised_2021}.
 Where other unsupervised methods can be seen as a means to an end in itself (e.g., clustering or data generation), self-supervised learning takes the form of a pretext task that only adds value when associated with a downstream task.
-This makes self-supervised learning tie naturally with semi-supervised learning, but it may also be part of a fully unsupervised setup \cite{baevski_unsupervised_2021}.
-Self-supervised learning is often characterized by automatically deriving the target from the input or other unlabeled examples \cite{ouali_overview_2020}.
+This makes self-supervised learning tie naturally with semi-supervised learning, but it may also be part of a fully unsupervised setup \parencite{baevski_unsupervised_2021}.
+Self-supervised learning is often characterized by automatically deriving the target from the input or other unlabeled examples \parencite{ouali_overview_2020}.
 
 \paragraph{Predictive models}
-Similar to classic autoregressive language models \cite{mikolov_recurrent_2010}, contextualized speech representations can be learned by predicting future values of a simple representation \cite{oord_representation_2018, chung_unsupervised_2019, schneider_wav2vec_2019, chung_generative_2020, jiang_further_2021} (\textbf{\textsc{prd}}, \cref{tab:model-taxonomy}).
+Similar to classic autoregressive language models \parencite{mikolov_recurrent_2010}, contextualized speech representations can be learned by predicting future values of a simple representation \parencite{oord_representation_2018, chung_unsupervised_2019, schneider_wav2vec_2019, chung_generative_2020, jiang_further_2021} (\textbf{\textsc{prd}}, \cref{tab:model-taxonomy}).
 Modeling spectrograms directly, autoregressive predictive coding (APC, \citealp{chung_unsupervised_2019}) is perhaps the simplest example in this category.
@@
 \noindent Here, $f(\cdot)$ and $g(\cdot)$ are parameterized by neural networks such that each $\mathbf{c}_t$ is only conditioned on previous inputs $\mathbf{x}_{1:t}$ and $\hat{\mathbf{x}}_{t+k}$ is computed step-wise.
-\citet{chung_unsupervised_2019} use a stack of unidirectional LSTMs for $f(\cdot)$ and a linear regression layer for $g(\cdot)$.
+\textcite{chung_unsupervised_2019} use a stack of unidirectional LSTMs for $f(\cdot)$ and a linear regression layer for $g(\cdot)$.
 Tasks that seek to predict or reconstruct the input are very common.
-In the literature, these are often jointly referred to as reconstruction tasks (\textbf{\textsc{rec}}, \cref{tab:model-taxonomy}) \cite{liu_tera_2021, wang_unispeech_2021}, although this is somewhat misleading in the case of prediction.
+In the literature, these are often jointly referred to as reconstruction tasks (\textbf{\textsc{rec}}, \cref{tab:model-taxonomy}) \parencite{liu_tera_2021, wang_unispeech_2021}, although this is somewhat misleading in the case of prediction.
 
-Contrary to generative models, such as WaveNet \cite{oord_wavenet_2016}, the APC model is not restricted to next-step prediction.
+Contrary to generative models, such as WaveNet \parencite{oord_wavenet_2016}, the APC model is not restricted to next-step prediction.
 Instead, it predicts $k > 0$ steps ahead in order to ensure that the model does not learn a trivial solution by exploiting the smoothness of the signal.
-Depending on the downstream task, we are often interested in learning so-called slow features that will typically span multiple input frames \cite{wiskott_slow_2002}.
-Even the smallest linguistic units of speech, phonemes, tend to span $0.1$ seconds on average \cite{garofolo_timit_1993}, whereas spectrogram frames $\mathbf{x}_t$ are typically computed at $0.01$ second intervals.
-However, sometimes local smoothness is explicitly used to define the task \cite{badino_autoencoder_2014, jati_speaker2vec_2017, jati_neural_2019}.
+Depending on the downstream task, we are often interested in learning so-called slow features that will typically span multiple input frames \parencite{wiskott_slow_2002}.
+Even the smallest linguistic units of speech, phonemes, tend to span $0.1$ seconds on average \parencite{garofolo_timit_1993}, whereas spectrogram frames $\mathbf{x}_t$ are typically computed at $0.01$ second intervals.
+However, sometimes local smoothness is explicitly used to define the task \parencite{badino_autoencoder_2014, jati_speaker2vec_2017, jati_neural_2019}.
 
@@
 Again, $f_c(\cdot)$ should be limited to condition each $\mathbf{c}_t$ on previous time-steps $\mathbf{v}_{1:t}$ and $g_k(\cdot)$ is a step-wise transformation.
-The loss is based on noise constrastive estimation \cite{gutmann_noisecontrastive_2010} and is given by
+The loss is based on noise constrastive estimation \parencite{gutmann_noisecontrastive_2010} and is given by
 \begin{align}
@@
 Note that the loss is also indexed by $k$ to show that CPC targets multiple offsets.
-The APC model is easily extended in a similar way \cite{chung_improved_2020}.
+The APC model is easily extended in a similar way \parencite{chung_improved_2020}.
 
@@
 
-A model closely related to the original CPC model is wav2vec \cite{schneider_wav2vec_2019}.
+A model closely related to the original CPC model is wav2vec \parencite{schneider_wav2vec_2019}.
 It uses a different parameterization of the functions $f_v(\cdot)$ and $f_c(\cdot)$, and modifies the loss to consider a binary prediction task, such that we have
@@
 \paragraph{Masking-based models} One downside of predictive tasks is that models are primarily unidirectional.
-Some work has extended APC and CPC inspired models with separate encoders operating in opposite directions \cite{ling_deep_2020, kawakami_learning_2020, borgholt_scaling_2021}, but these models are still restricted to process left and right context separately.
-Inspired by the masked language model task used for text-based representation learning \cite{devlin_bert_2018}, several papers have used masking to overcome this challenge (\textbf{\textsc{msk}}, \cref{tab:model-taxonomy}).
+Some work has extended APC and CPC inspired models with separate encoders operating in opposite directions \parencite{ling_deep_2020, kawakami_learning_2020, borgholt_scaling_2021}, but these models are still restricted to process left and right context separately.
+Inspired by the masked language model task used for text-based representation learning \parencite{devlin_bert_2018}, several papers have used masking to overcome this challenge (\textbf{\textsc{msk}}, \cref{tab:model-taxonomy}).
 Masking refers to replacing parts of the input with zeros or a learned masking vector.
-For zero-masking \cite{jiang_improving_2019, liu_mockingjay_2020, wang_unsupervised_2020, chi_audio_2020, ling_decoar_2020}, we have
+For zero-masking \parencite{jiang_improving_2019, liu_mockingjay_2020, wang_unsupervised_2020, chi_audio_2020, ling_decoar_2020}, we have
 \begin{align}
@@
 The entries of $\mathbf{m}_{1:T}$ are determined by some stochastic policy.
-One frequent inspiration is SpecAugment \cite{park_specaugment_2019}, which was originally proposed for supervised speech recognition and applies frequency and time masking to spectrogram representations.
-While temporal masking is most common, frequency masking has also been adopted for representation learning  \cite{wang_unsupervised_2020}.
-A simple, yet popular, masking strategy is to draw a proportion of input indices $t_i\sim\{1,\dots,T-M\}$ without replacement, and then mask $\{t_i, \dots, t_i+M\}$ \cite{baevski_wav2vec_2020, hsu_hubert_2021, ling_decoar_2020}.
+One frequent inspiration is SpecAugment \parencite{park_specaugment_2019}, which was originally proposed for supervised speech recognition and applies frequency and time masking to spectrogram representations.
+While temporal masking is most common, frequency masking has also been adopted for representation learning  \parencite{wang_unsupervised_2020}.
+A simple, yet popular, masking strategy is to draw a proportion of input indices $t_i\sim\{1,\dots,T-M\}$ without replacement, and then mask $\{t_i, \dots, t_i+M\}$ \parencite{baevski_wav2vec_2020, hsu_hubert_2021, ling_decoar_2020}.
 
@@
 
-\noindent Here, $f_v(\cdot)$ is a convolutional neural network, $f_c(\cdot)$ is a transformer encoder \cite{vaswani_attention_2017} and $g_q(\cdot)$ is a quantization module used to learn targets from the localized variable $\mathbf{v}_{1:T}$.
+\noindent Here, $f_v(\cdot)$ is a convolutional neural network, $f_c(\cdot)$ is a transformer encoder \parencite{vaswani_attention_2017} and $g_q(\cdot)$ is a quantization module used to learn targets from the localized variable $\mathbf{v}_{1:T}$.
 Computing quantized targets this way requires an extra loss term, which we will present when we discuss quantization in general below.
@@
             %                                                           MSK      PRD      CON      REC      QTZ      GEN      LOC      GLB     VAR      FRZ      FTN
-            \textbf{Audio Word2vec} \footnotesize\cite{chung_audio_2016}      & 2016 Mar. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
-            \textbf{Speech2Vec} \footnotesize\cite{chung_speech2vec_2018}     & 2018 Mar. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
-            \textbf{Unspeech} \footnotesize\cite{milde_unspeech_2018}         & 2018 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{Audio Word2vec} \footnotesize\parencite{chung_audio_2016}      & 2016 Mar. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{Speech2Vec} \footnotesize\parencite{chung_speech2vec_2018}     & 2018 Mar. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{Unspeech} \footnotesize\parencite{milde_unspeech_2018}         & 2018 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
 
@@
 
-            \textbf{APC} \footnotesize\cite{chung_unsupervised_2019}          & 2019 Oct. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 5/4
-            \textbf{wav2vec} \footnotesize\cite{schneider_wav2vec_2019}       & 2019 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/4
-            \textbf{Mockingjay}  \footnotesize\cite{liu_mockingjay_2020}      & 2019 Oct.
+            \textbf{APC} \footnotesize\parencite{chung_unsupervised_2019}          & 2019 Oct.
+& \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 5/4
+            \textbf{wav2vec} \footnotesize\parencite{schneider_wav2vec_2019}       & 2019 Apr.
+& \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/4
+            \textbf{Mockingjay}  \footnotesize\parencite{liu_mockingjay_2020}      & 2019 Oct.
 & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 25/10-19
-            \textbf{wav2vec 2.0} \footnotesize\cite{baevski_wav2vec_2020}     & 2020 Jun. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 20/6
-            \textbf{NPC} \footnotesize\cite{liu_nonautoregressive_2020}                     & 2020 Nov.
+            \textbf{wav2vec 2.0} \footnotesize\parencite{baevski_wav2vec_2020}     & 2020 Jun.
+& \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 20/6
+            \textbf{NPC} \footnotesize\parencite{liu_nonautoregressive_2020}                     & 2020 Nov.
 & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 1/11
-            \textbf{DeCoAR 2.0} \footnotesize\cite{ling_decoar_2020}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/12
-            \textbf{SCPC} \footnotesize\cite{bhati_segmental_2021}            & 2021 Jun. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ % 3/6
-            \textbf{HuBERT} \footnotesize\cite{hsu_hubert_2021}               & 2021 Jun. & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 14/6
+            \textbf{DeCoAR 2.0} \footnotesize\parencite{ling_decoar_2020}          & 2020 Dec.
+& \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/12
+            \textbf{SCPC} \footnotesize\parencite{bhati_segmental_2021}            & 2021 Jun.
+& \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ % 3/6
+            \textbf{HuBERT} \footnotesize\parencite{hsu_hubert_2021}               & 2021 Jun.
+& \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 14/6
             \midrule
@@
             %                                                           MSK      PRD      CON      REC      QTZ      GEN      LOC      GLO     VAR      FRZ      FTN
-            \textbf{VRNN} \footnotesize\cite{chung_recurrent_2015}          & 2015 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-            \textbf{SRNN} \footnotesize\cite{fraccaro_sequential_2016}      & 2016 May  & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-            \textbf{HMM-VAE} \footnotesize\cite{ebbers_hidden_2017}         & 2017 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-            \textbf{ConvVAE} \footnotesize\cite{hsu_learning_2017}          & 2017 Apr. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
-            \textbf{FHVAE} \footnotesize\cite{hsu_unsupervised_2017}        & 2017 Sep. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{VRNN} \footnotesize\parencite{chung_recurrent_2015}          & 2015 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{SRNN} \footnotesize\parencite{fraccaro_sequential_2016}      & 2016 May  & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{HMM-VAE} \footnotesize\parencite{ebbers_hidden_2017}         & 2017 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{ConvVAE} \footnotesize\parencite{hsu_learning_2017}          & 2017 Apr. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{FHVAE} \footnotesize\parencite{hsu_unsupervised_2017}        & 2017 Sep. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
             \textbf{VQ-VAE} (van den Oord et al. 2018)                      & 2017 Nov. & \xmark & \xmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-            \textbf{BHMM-VAE} \footnotesize\cite{glarner_full_2018}         & 2018 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-            \textbf{STCN} \footnotesize\cite{aksan_stcn_2019}               & 2019 Feb. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-            \textbf{FDMM} \footnotesize\cite{khurana_factorial_2019}        & 2019 Oct. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
-            \textbf{ConvDMM} \footnotesize\cite{khurana_convolutional_2020} & 2020 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{BHMM-VAE} \footnotesize\parencite{glarner_full_2018}         & 2018 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{STCN} \footnotesize\parencite{aksan_stcn_2019}               & 2019 Feb. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{FDMM} \footnotesize\parencite{khurana_factorial_2019}        & 2019 Oct. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{ConvDMM} \footnotesize\parencite{khurana_convolutional_2020} & 2020 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
             \bottomrule
@@
 Several models enforce a discrete latent space by quantizing the vector representation (\textbf{\textsc{qtz}}, \cref{tab:model-taxonomy}).
-The two most popular approaches are the Gumbel-softmax \cite{jang_categorical_2016, maddison_concrete_2017} and the quantization used in the VQ-VAE \cite{oord_neural_2018}.
+The two most popular approaches are the Gumbel-softmax \parencite{jang_categorical_2016, maddison_concrete_2017} and the quantization used in the VQ-VAE \parencite{oord_neural_2018}.
 
@@
 $\mathbf{p}$ is mapped to a one-hot vector using a function $\varphi(\cdot)$, such that $\varphi(\mathbf{p})_i = 1$ if $i = \argmax_j p_j$ and $0$ otherwise. 
-As this function is non-differentiable, we must rely on the straight-through gradient estimator \cite{bengio_estimating_2013} which assumes that the Jacobian ${\partial \, \varphi}/{\partial \, \mathbf{p}}$ equals the identity matrix.
+As this function is non-differentiable, we must rely on the straight-through gradient estimator \parencite{bengio_estimating_2013} which assumes that the Jacobian ${\partial \, \varphi}/{\partial \, \mathbf{p}}$ equals the identity matrix.
 The one-hot vector can then be used for a codebook lookup to obtain the final quantized vector (e.g., $\mathbf{q}_t$ in \cref{eq_brief: w2v2 qtz}).
@@
 As $\arg\min$ is non-differentiable, the straight-through estimator is used as for the Gumbel-softmax. 
-Codebook learning is facilitated by a two-term auxiliary loss similar to classical vector quantization dictionary learning \cite{burton_generalization_1983, soong_vector_1985}.
+Codebook learning is facilitated by a two-term auxiliary loss similar to classical vector quantization dictionary learning \parencite{burton_generalization_1983, soong_vector_1985}.
 Gradients for the codebook vectors are given solely by a vector quantization term.
@@
 where $\text{sg}[\mathbf{x}] = \mathbf{x}$ is the stop-gradient operator with the property $\frac{d}{dx_i}\text{sg}[\mathbf{x}] \equiv 0$ for all $i$ and $\beta$ is a hyperparameter.
-Although vector quantization was introduced by the VQ-VAE which is, in some ways, a latent variable model, it has been applied to self-supervised methods \cite{vanniekerk_vectorquantized_2020, baevski_vqwav2vec_2020}.
+Although vector quantization was introduced by the VQ-VAE which is, in some ways, a latent variable model, it has been applied to self-supervised methods \parencite{vanniekerk_vectorquantized_2020, baevski_vqwav2vec_2020}.
 
 \textit{Motivation:} Similar to how quantization approaches differ between works, so do the motivations provided for employing them.
-The vq-wav2vec \cite{baevski_vqwav2vec_2020, baevski_effectiveness_2020} learn quantized representations in order to apply natural language processing models, like BERT \cite{devlin_bert_2018}, afterwards.
-Other works use quantization for speech segmentation \cite{kamper_unsupervised_2021, chorowski_unsupervised_2019} or as a bottleneck in order to ``\textit{limit model capacity}" \cite{chung_vectorquantized_2020, ling_decoar_2020}.
-Finally, \citet{chung_vectorquantized_2020} explore quantization between different layers in the APC model, but find that continuous representations consistently perform better than their quantized counterparts on a downstream phoneme classification task
+The vq-wav2vec \parencite{baevski_vqwav2vec_2020, baevski_effectiveness_2020} learn quantized representations in order to apply natural language processing models, like BERT \parencite{devlin_bert_2018}, afterwards.
+Other works use quantization for speech segmentation \parencite{kamper_unsupervised_2021, chorowski_unsupervised_2019} or as a bottleneck in order to ``\textit{limit model capacity}" \parencite{chung_vectorquantized_2020, ling_decoar_2020}.
+Finally, \textcite{chung_vectorquantized_2020} explore quantization between different layers in the APC model, but find that continuous representations consistently perform better than their quantized counterparts on a downstream phoneme classification task
 
 Given our previous discussion of how it might not be beneficial to model localized noise, quantization in wav2vec 2.0 seems well motivated, as it enforces the target representation $\mathbf{q}_{1:T}$ to discard such noise.
-Taking this idea further, the HuBERT model \cite{hsu_hubert_2021} uses offline quantization to learn categorical targets.
+Taking this idea further, the HuBERT model \parencite{hsu_hubert_2021} uses offline quantization to learn categorical targets.
 Initially, spectrogram features are used to learn frame-wise labels with $k$-means clustering.
@@
 
-Early work on global speech representation learning takes inspiration from the autoencoder framework \cite{kramer_nonlinear_1991}.
-\citet{chung_audio_2016} propose a simple sequence-to-sequence autoencoder for learning acoustic word embeddings:
+Early work on global speech representation learning takes inspiration from the autoencoder framework \parencite{kramer_nonlinear_1991}.
+\textcite{chung_audio_2016} propose a simple sequence-to-sequence autoencoder for learning acoustic word embeddings:
 \begin{align}
@@
 The authors also propose a denoising autoencoder with masked inputs $f(\mathbf{x}_{1:T} \circ \mathbf{m}_{1:T})$.
-Similar RNN-based autoencoders have also been explored \cite{kamper_truly_2019, holzenberger_learning_2018}.
+Similar RNN-based autoencoders have also been explored \parencite{kamper_truly_2019, holzenberger_learning_2018}.
 
-Prior to this work, \citet{kamper_unsupervised_2015} and \citet{renshaw_comparison_2015} introduced the \textit{correspondence autoencoder}.
+Prior to this work, \textcite{kamper_unsupervised_2015, renshaw_comparison_2015} introduced the \textit{correspondence autoencoder}.
 This method uses dynamic time warping to align input-target segment pairs extracted with unsupervised term discovery.
-In more recent work, the need for alignment has been alleviated by adopting the sequence-to-sequence framework \cite{kamper_truly_2019, jacobs_acoustic_2021}.
+In more recent work, the need for alignment has been alleviated by adopting the sequence-to-sequence framework \parencite{kamper_truly_2019, jacobs_acoustic_2021}.
 
-Inspired by the work on semantic word embeddings for text \cite{mikolov_distributed_2013}, the sequence-to-sequence framework has also been used to implement speech-based versions of the skipgram and continuous bag-of-words models \cite{chung_learning_2017, chung_speech2vec_2018}.
+Inspired by the work on semantic word embeddings for text \parencite{mikolov_distributed_2013}, the sequence-to-sequence framework has also been used to implement speech-based versions of the skipgram and continuous bag-of-words models \parencite{chung_learning_2017, chung_speech2vec_2018}.
 Given a segment corresponding to a single word $\mathbf{x}_{(n)} = \mathbf{x}_{t_{n}:t_{n+1}}$, the skipgram model is trained to predict neighboring words $\mathbf{x}_{(n + k)}$ where $k\ne0$.
@@
 The sequence-to-sequence models described above rely on speech segments corresponding to words.
-The segments are obtained by supervised forced alignment, but similar models have been explored without this requirement \cite{jati_speaker2vec_2017, tagliasacchi_pretraining_2020}.
+The segments are obtained by supervised forced alignment, but similar models have been explored without this requirement \parencite{jati_speaker2vec_2017, tagliasacchi_pretraining_2020}.
 
-Contrastive learning has also been explored for global speech representation learning \citep{milde_unspeech_2018, jati_neural_2019, jansen_unsupervised_2018}.
-And prior to the widespread adoption of neural networks, \citet{levin_fixeddimensional_2013} explore principal component analysis and Laplacian eigenmaps for learning fixed-sized acoustic embeddings.
+Contrastive learning has also been explored for global speech representation learning \parencite{milde_unspeech_2018, jati_neural_2019, jansen_unsupervised_2018}.
+And prior to the widespread adoption of neural networks, \textcite{levin_fixeddimensional_2013} explore principal component analysis and Laplacian eigenmaps for learning fixed-sized acoustic embeddings.
 
@@
 Some models learn local representations with a variable temporal resolution that is not proportional to the input resolution (\textbf{\textsc{var}}, \cref{tab:model-taxonomy}).
-In practice, this is often achieved implicitly, by learning segment boundaries or by taking repeated quantized values to belong to the same segment \cite{kamper_unsupervised_2021, chorowski_unsupervised_2019, michel_blind_2017, kreuk_selfsupervised_2020, wang_gate_2017, dieleman_variablerate_2021}.
+In practice, this is often achieved implicitly, by learning segment boundaries or by taking repeated quantized values to belong to the same segment \parencite{kamper_unsupervised_2021, chorowski_unsupervised_2019, michel_blind_2017, kreuk_selfsupervised_2020, wang_gate_2017, dieleman_variablerate_2021}.
 An exception is the recently proposed \emph{segmental contrastive predictive coding} (SCPC, \citealp{bhati_segmental_2021, bhati_unsupervised_2022}).
 With this approach, the model explicitly learns segment boundaries, which are used to downsample the representations during training.
-The same segmentation strategy has subsequently been applied in other models \cite{cuervo_contrastive_2022}.
+The same segmentation strategy has subsequently been applied in other models \parencite{cuervo_contrastive_2022}.
 
@@
 Furthermore, many of the presented models have been successfully applied to other use cases.
-For instance, wav2vec 2.0 and related models have been applied to learn cross-lingual and multi-lingual representations \cite{riviere_unsupervised_2020, conneau_unsupervised_2020, khurana_magic_2021} and proven well-suited for concurrently learning with labeled data \cite{talnikar_joint_2021, wang_unispeech_2021}.
+For instance, wav2vec 2.0 and related models have been applied to learn cross-lingual and multi-lingual representations \parencite{riviere_unsupervised_2020, conneau_unsupervised_2020, khurana_magic_2021} and proven well-suited for concurrently learning with labeled data \parencite{talnikar_joint_2021, wang_unispeech_2021}.
 %We will discuss the downstream application of self-supervised representations further in \cref{sec:eval}.
@@
 For representation learning, the inference of latent variables is of primary interest.
-An LVM is defined by the observation model $p(\mathbf{x}|\mathbf{z})$, which defines the relationship between the observed and latent variables, and the prior $p(\mathbf{z})$, which defines the relationship among the latent variables \cite{bartholomew_latent_2011}.
+An LVM is defined by the observation model $p(\mathbf{x}|\mathbf{z})$, which defines the relationship between the observed and latent variables, and the prior $p(\mathbf{z})$, which defines the relationship among the latent variables \parencite{bartholomew_latent_2011}.
 An LVM models the generative process via the joint observation and prior model $p(\mathbf{x}, \mathbf{z})$ often referred to as the generative model. 
@@
 \end{equation}
-The latent variable can be inferred with e.g. Markov Chain Monte Carlo (MCMC) methods \cite{mohamed_monte_2019} or variational inference \cite{jordan_introduction_1999}.
+The latent variable can be inferred with e.g. Markov Chain Monte Carlo (MCMC) methods \parencite{mohamed_monte_2019} or variational inference \parencite{jordan_introduction_1999}.
 
-For representation learning, LVMs are commonly defined using the VAE framework \cite{kingma_autoencoding_2014, rezende_stochastic_2014} which is also the focus of our exposition. 
+For representation learning, LVMs are commonly defined using the VAE framework \parencite{kingma_autoencoding_2014, rezende_stochastic_2014} which is also the focus of our exposition. 
 In the VAE framework, the observation model $p(\mathbf{x}|\mathbf{z})$ is parameterized using a deep neural network.
 This choice allows modeling complex and high-dimensional data but also makes the integral in \cref{eq_brief: lvm log-likelihood} analytically intractable.
-MCMC methods can be used to estimate it and the true model posterior $p(\mathbf{z}|\mathbf{x})$, but these methods are usually computationally expensive in this setting \cite{mohamed_monte_2019}.
-To counter this and make gradient-based maximum likelihood training feasible, the VAE instead employs variational inference \cite{jordan_introduction_1999}.
+MCMC methods can be used to estimate it and the true model posterior $p(\mathbf{z}|\mathbf{x})$, but these methods are usually computationally expensive in this setting \parencite{mohamed_monte_2019}.
+To counter this and make gradient-based maximum likelihood training feasible, the VAE instead employs variational inference \parencite{jordan_introduction_1999}.
 It approximates the intractable true model posterior by introducing a variational posterior distribution $q(\mathbf{z}|\mathbf{x})$, also parameterized by a deep neural network.
@@
 The bound can be efficiently evaluated and optimized with Monte Carlo (MC) estimation by sampling from $q(\mathbf{z}|\mathbf{x})$.
-Low-variance gradient estimates are usually obtained via reparameterization of $q(\mathbf{z}|\mathbf{x})$ \cite{kingma_autoencoding_2014} although alternatives exist (e.g., inverse CDF sampling) \cite{mohamed_monte_2019}.
+Low-variance gradient estimates are usually obtained via reparameterization of $q(\mathbf{z}|\mathbf{x})$ \parencite{kingma_autoencoding_2014} although alternatives exist (e.g., inverse CDF sampling) \parencite{mohamed_monte_2019}.
 The ELBO can also be written as
@@
 In brief, LVMs of the VAE-type consist of a approximate posterior, $q(\mathbf{z}|\mathbf{x})$, an observation model, $p(\mathbf{x}|\mathbf{z})$, and a prior, $p(\mathbf{z})$. 
-With reference to probabilistic coding theory, the approximate posterior is often referred to as the encoder and the observation model as the decoder \cite{kingma_autoencoding_2014, rezende_stochastic_2014}.
+With reference to probabilistic coding theory, the approximate posterior is often referred to as the encoder and the observation model as the decoder \parencite{kingma_autoencoding_2014, rezende_stochastic_2014}.
 From a theoretical perspective, the encoder exists solely as the result of choosing to use variational inference to train the decoder rather than e.g.
@@
             %                                                         ARX      LOC      GLB      ARX      ARZ     IND      GLB      ARZ      FLT      LSM      GSM      GLB
-            \textbf{VRNN} \footnotesize\cite{chung_recurrent_2015}          & 2015 Jun.
+            \textbf{VRNN} \footnotesize\parencite{chung_recurrent_2015}          & 2015 Jun.
 & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark \\
-            \textbf{SRNN} \footnotesize\cite{fraccaro_sequential_2016}      & 2016 May  & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark \\
-            \textbf{HMM-VAE} \footnotesize\cite{ebbers_hidden_2017}         & 2017 Mar.
+            \textbf{SRNN} \footnotesize\parencite{fraccaro_sequential_2016}      & 2016 May  & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark \\
+            \textbf{HMM-VAE} \footnotesize\parencite{ebbers_hidden_2017}         & 2017 Mar.
 & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
-            \textbf{ConvVAE} \footnotesize\cite{hsu_learning_2017}          & 2017 Apr.
+            \textbf{ConvVAE} \footnotesize\parencite{hsu_learning_2017}          & 2017 Apr.
 & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \cmark & \xmark \\
-            \textbf{FHVAE} \footnotesize\cite{hsu_unsupervised_2017}        & 2017 Sep.
+            \textbf{FHVAE} \footnotesize\parencite{hsu_unsupervised_2017}        & 2017 Sep.
 & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \cmark & \cmark \\
             \textbf{VQ-VAE} (van den Oord et al. 2018)            & 2017 Nov. & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\
-            \textbf{BHMM-VAE} \footnotesize\cite{glarner_full_2018}         & 2018 Mar.
+            \textbf{BHMM-VAE} \footnotesize\parencite{glarner_full_2018}         & 2018 Mar.
 & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark \\
-            \textbf{STCN} \footnotesize\cite{aksan_stcn_2019}               & 2019 Feb.
+            \textbf{STCN} \footnotesize\parencite{aksan_stcn_2019}               & 2019 Feb.
 & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
-            \textbf{FDMM} \footnotesize\cite{khurana_factorial_2019}        & 2019 Oct.
+            \textbf{FDMM} \footnotesize\parencite{khurana_factorial_2019}        & 2019 Oct.
 & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark \\
-            \textbf{ConvDMM} \footnotesize\cite{khurana_convolutional_2020} & 2020 Jun.
+            \textbf{ConvDMM} \footnotesize\parencite{khurana_convolutional_2020} & 2020 Jun.
 & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark \\
@@
 \paragraph{Observation models} 
-A common choice for the observation model $p(\mathbf{x}|\mathbf{z})$ is to include an autoregressive dependency on the observed variable (\textbf{\textsc{arx}}, \cref{tab:lvm-model-primitives}) that is, $p(\mathbf{x}_t|\mathbf{x}_{1:t-1}, \cdot)$ where $\cdot$ represents some dependency on the latent variable \cite{chung_recurrent_2015, fraccaro_sequential_2016, oord_neural_2018, oord_representation_2018}.
-This allows the latent representation to focus on correlations that cannot easily be predicted from the observed variable at previous time-steps \cite{oord_representation_2018}.
+A common choice for the observation model $p(\mathbf{x}|\mathbf{z})$ is to include an autoregressive dependency on the observed variable (\textbf{\textsc{arx}}, \cref{tab:lvm-model-primitives}) that is, $p(\mathbf{x}_t|\mathbf{x}_{1:t-1}, \cdot)$ where $\cdot$ represents some dependency on the latent variable \parencite{chung_recurrent_2015, fraccaro_sequential_2016, oord_neural_2018, oord_representation_2018}.
+This allows the latent representation to focus on correlations that cannot easily be predicted from the observed variable at previous time-steps \parencite{oord_representation_2018}.
 In practice, the dependency on $\mathbf{x}_{1:t-1}$ is often assumed to be Markovian and hence only on $\mathbf{x}_{t-1}$.
@@
 Specifically encouraging the latent representation to discard correlations across the temporal dimension might degrade the quality of the latent representation.
-Furthermore, since such a decoder can perform quite well by simply solving an autoregressive prediction problem, similar to WaveNet \cite{oord_wavenet_2016}, it can make the model prone to suffer from posterior collapse.
-This problem arises when the approximate and true posterior distributions collapse into the prior which renders the representations non-informative \cite{bowman_generating_2016, sonderby_ladder_2016}.
+Furthermore, since such a decoder can perform quite well by simply solving an autoregressive prediction problem, similar to WaveNet \parencite{oord_wavenet_2016}, it can make the model prone to suffer from posterior collapse.
+This problem arises when the approximate and true posterior distributions collapse into the prior which renders the representations non-informative \parencite{bowman_generating_2016, sonderby_ladder_2016}.
 Notably, posterior collapse is a local minimum of the ELBO since the KL-divergence becomes zero.
-Some works alleviate this problem with tricks like KL-annealing and free bits \cite{bowman_generating_2016, sonderby_ladder_2016, kingma_improved_2016}. 
-The VQ-VAE uses a quantized latent space that is not susceptible to posterior collapse \textit{per se} \cite{oord_neural_2018}.
+Some works alleviate this problem with tricks like KL-annealing and free bits \parencite{bowman_generating_2016, sonderby_ladder_2016, kingma_improved_2016}. 
+The VQ-VAE uses a quantized latent space that is not susceptible to posterior collapse \textit{per se} \parencite{oord_neural_2018}.
 How to equip LVMs with powerful decoders while avoiding posterior collapse is an open problem.
 
-Some LVMs do not use autoregressive observation models \cite{ebbers_hidden_2017, glarner_full_2018, hsu_unsupervised_2017, hsu_learning_2017, khurana_factorial_2019, khurana_convolutional_2020}.
-These more closely follow the assumption of \emph{local independence} which states that observed variables are conditionally independent given the local (\textbf{\textsc{loc}}, \cref{tab:lvm-model-primitives}) and/or global (\textbf{\textsc{glb}}, \cref{tab:lvm-model-primitives}) latent variables \cite{bartholomew_latent_2011}.
+Some LVMs do not use autoregressive observation models \parencite{ebbers_hidden_2017, glarner_full_2018, hsu_unsupervised_2017, hsu_learning_2017, khurana_factorial_2019, khurana_convolutional_2020}.
+These more closely follow the assumption of \emph{local independence} which states that observed variables are conditionally independent given the local (\textbf{\textsc{loc}}, \cref{tab:lvm-model-primitives}) and/or global (\textbf{\textsc{glb}}, \cref{tab:lvm-model-primitives}) latent variables \parencite{bartholomew_latent_2011}.
 However, this forces the latent variable to encode details about the observed variable to achieve a good reconstruction.
-This is opposite to contrastive self-supervised learning which allows models to discard details in $\mathbf{x}_{1:T}$ that do not inform the training objective \cite{baevski_wav2vec_2020}.
+This is opposite to contrastive self-supervised learning which allows models to discard details in $\mathbf{x}_{1:T}$ that do not inform the training objective \parencite{baevski_wav2vec_2020}.
 
@@
 This generally results in a slow-down of the generative process which may be of concern if the use-case is data generation.
-Priors that are autoregressive on the latent variable (\textbf{\textsc{arz}}, \cref{tab:lvm-model-primitives}) take the form $p(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ and enable stochastic temporal transitions similar to hidden Markov models but with potentially nonlinear transition functions \cite{chung_recurrent_2015, fraccaro_sequential_2016, khurana_factorial_2019, khurana_convolutional_2020}.
+Priors that are autoregressive on the latent variable (\textbf{\textsc{arz}}, \cref{tab:lvm-model-primitives}) take the form $p(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ and enable stochastic temporal transitions similar to hidden Markov models but with potentially nonlinear transition functions \parencite{chung_recurrent_2015, fraccaro_sequential_2016, khurana_factorial_2019, khurana_convolutional_2020}.
 Locally independent priors (\textbf{\textsc{ind}}, \cref{tab:lvm-model-primitives}) are rarely applied to sequential latent variables since they make the prior latent dynamics independent of the value of previous latent variables.
-Models that do impose such priors on sequential latents are quite limited in their generative power, unless they learn the prior dynamics post-hoc as done in the VQ-VAE \cite{oord_neural_2018}.
+Models that do impose such priors on sequential latents are quite limited in their generative power, unless they learn the prior dynamics post-hoc as done in the VQ-VAE \parencite{oord_neural_2018}.
 Global latent variables (\textbf{\textsc{glb}}, \cref{tab:lvm-model-primitives}) are fundamentally limited in the amount of information they can encode.
-Hence, models usually use them in combination with another local latent variable, or to encode fixed length input segments \cite{khurana_factorial_2019, hsu_learning_2017, hsu_unsupervised_2017}.
+Hence, models usually use them in combination with another local latent variable, or to encode fixed length input segments \parencite{khurana_factorial_2019, hsu_learning_2017, hsu_unsupervised_2017}.
 
@@
 For this reason, all inference models covered here are conditioned on the observed sequence in some way. Generally, the inference model can be seen as solving either a filtering or smoothing problem.
-In filtering (\textbf{\textsc{flt}}, \cref{tab:lvm-model-primitives}), the latent variables are assumed to depend only on past and current values of the observed variable, $q(\mathbf{z}_t|\mathbf{x}_{1:t})$ \cite{chung_recurrent_2015, khurana_convolutional_2020}.
-In global smoothing (\textbf{\textsc{gsm}}, \cref{tab:lvm-model-primitives}), this causal dependency is replaced with a dependency on all observed values, $q(\mathbf{z}_t|\mathbf{x}_{1:T})$ \cite{fraccaro_sequential_2016, hsu_learning_2017}.
-Smoothing can also be done locally (\textbf{\textsc{lsm}}, \cref{tab:lvm-model-primitives}), where the latent variables then depend on $\mathbf{x}_{t-r:t+r}$ for some integer $r>0$ \cite{oord_neural_2018}.
+In filtering (\textbf{\textsc{flt}}, \cref{tab:lvm-model-primitives}), the latent variables are assumed to depend only on past and current values of the observed variable, $q(\mathbf{z}_t|\mathbf{x}_{1:t})$ \parencite{chung_recurrent_2015, khurana_convolutional_2020}.
+In global smoothing (\textbf{\textsc{gsm}}, \cref{tab:lvm-model-primitives}), this causal dependency is replaced with a dependency on all observed values, $q(\mathbf{z}_t|\mathbf{x}_{1:T})$ \parencite{fraccaro_sequential_2016, hsu_learning_2017}.
+Smoothing can also be done locally (\textbf{\textsc{lsm}}, \cref{tab:lvm-model-primitives}), where the latent variables then depend on $\mathbf{x}_{t-r:t+r}$ for some integer $r>0$ \parencite{oord_neural_2018}.
 Compared to self-supervised models that often use transformer encoders it can be hypothesized that global smoothing offers a stronger case than local smoothing and filtering for representation learning.
@@
 Finally, latent variables are often made to depend autoregressively on past inferred values, e.g.
-$q(\mathbf{z}_t|\mathbf{z}_{1:t-1},\mathbf{x}_{1:t})$ (\textbf{\textsc{arz}}, \cref{tab:lvm-model-primitives}) \cite{chung_recurrent_2015, fraccaro_sequential_2016}.
+$q(\mathbf{z}_t|\mathbf{z}_{1:t-1},\mathbf{x}_{1:t})$ (\textbf{\textsc{arz}}, \cref{tab:lvm-model-primitives}) \parencite{chung_recurrent_2015, fraccaro_sequential_2016}.
 
@@
 Some work has explored using a hierarchy of latent variables (\textbf{\textsc{hie}}, \cref{tab:lvm-taxonomy}).
-This allows encoding the inductive bias that speech contains information at different temporal scales by letting the latent variables operate at different temporal scales \cite{hsu_unsupervised_2017}.
-\citet{khurana_factorial_2019} propose using a temporal latent variable along with a global latent variable.
-Recent work has focused on learning a deeper latent hierarchy with five latent variables \cite{aksan_stcn_2019}.
+This allows encoding the inductive bias that speech contains information at different temporal scales by letting the latent variables operate at different temporal scales \parencite{hsu_unsupervised_2017}.
+\textcite{khurana_factorial_2019} propose using a temporal latent variable along with a global latent variable.
+Recent work has focused on learning a deeper latent hierarchy with five latent variables \parencite{aksan_stcn_2019}.
 
 \paragraph{Other work}
-Before the introduction of the VAE, models such as deep belief networks (DBN, \citealp{hinton_fast_2006}) built from stacks of restricted Boltzmann machines \cite{smolensky_chapter_1986,fischer_training_2014} were popular.
-\citet{lee_unsupervised_2009} show the feasibility of using a two-layered DBN for discovering acoustic units of speech, while \citet{deng_binary_2010} show that a DBN can learn a binary coding of spectrograms that has higher signal-to-noise ratio than classical vector-quantization techniques for speech coding.
-DBNs are however notoriously tricky to optimize requiring the use of expensive MCMC sampling techniques for inference or resort to biased gradient estimates \cite{montavon_practical_2012,fischer_bounding_2011}.
+Before the introduction of the VAE, models such as deep belief networks (DBN, \citealp{hinton_fast_2006}) built from stacks of restricted Boltzmann machines \parencite{smolensky_chapter_1986,fischer_training_2014} were popular.
+\textcite{lee_unsupervised_2009} show the feasibility of using a two-layered DBN for discovering acoustic units of speech, while \textcite{deng_binary_2010} show that a DBN can learn a binary coding of spectrograms that has higher signal-to-noise ratio than classical vector-quantization techniques for speech coding.
+DBNs are however notoriously tricky to optimize requiring the use of expensive MCMC sampling techniques for inference or resort to biased gradient estimates \parencite{montavon_practical_2012,fischer_bounding_2011}.
 Non-neural LVMs for speech representation learning have also been explored
-\cite{lee_nonparametric_2012, ondel_variational_2016, heck_feature_2017, jansen_weak_2013}.
+\parencite{lee_nonparametric_2012, ondel_variational_2016, heck_feature_2017, jansen_weak_2013}.
 
@@
 In \cref{tab:model-taxonomy}, we see that work on global representations within self-supervised learning precedes work on local representations.
-However, we find that the core ideas underlying the recent successes in learning local representation models have also been used for global representation learning; masking \cite{chung_audio_2016}, context prediction \cite{chung_speech2vec_2018}, and contrastive training \cite{milde_unspeech_2018} have been applied in both settings.
-Furthermore, where work on global representation learning has taken inspiration from Word2vec \cite{mikolov_distributed_2013}, the techniques used for learning local representations are inspired by contextualized word embeddings \cite{devlin_bert_2018}.
+However, we find that the core ideas underlying the recent successes in learning local representation models have also been used for global representation learning; masking \parencite{chung_audio_2016}, context prediction \parencite{chung_speech2vec_2018}, and contrastive training \parencite{milde_unspeech_2018} have been applied in both settings.
+Furthermore, where work on global representation learning has taken inspiration from Word2vec \parencite{mikolov_distributed_2013}, the techniques used for learning local representations are inspired by contextualized word embeddings \parencite{devlin_bert_2018}.
 Thus, the gap between these two model classes is largely a product of the developments in related fields and the general increase in computational resources.
@@
 Hence, the prior might be considered for feature extraction.
-\citet{jones_is_2020} examine the importance of the prior in the VQ-VAE and show that the ability of this model to estimate densities $p(\mathbf{x}_{1:T})$ lies solely in its prior.
-Other work has also explored representations beyond the latent variable such as hidden units of the observation model \cite{khurana_convolutional_2020, chorowski_unsupervised_2019}.
+\textcite{jones_is_2020} examine the importance of the prior in the VQ-VAE and show that the ability of this model to estimate densities $p(\mathbf{x}_{1:T})$ lies solely in its prior.
+Other work has also explored representations beyond the latent variable such as hidden units of the observation model \parencite{khurana_convolutional_2020, chorowski_unsupervised_2019}.
 
@@
 Reconstruction also deals only with the observed data.
-Previous work has largely focused on the ability of these models to yield high-quality imputations within the tabular and image data domains, without probing for the effects on the learned latent representation \cite{mattei_miwae_2019, ipsen_not-miwae_2021}.
-The idea of using VAEs to impute missing data was already examined in the seminal paper by \citet{rezende_stochastic_2014}.
+Previous work has largely focused on the ability of these models to yield high-quality imputations within the tabular and image data domains, without probing for the effects on the learned latent representation \parencite{mattei_miwae_2019, ipsen_not-miwae_2021}.
+The idea of using VAEs to impute missing data was already examined in the seminal paper by \textcite{rezende_stochastic_2014}.
 Here the model was trained with fully observed data and used to impute data in an iterative sampling approach post hoc leaving the learned representations unchanged.
@@
 Although this review has a primarily methodological focus, we should briefly touch upon evaluation procedures.
-Training metrics for self-supervised tasks and the likelihood of LVMs offer little guidance as to the quality of the learned representations \cite{huszar_is_2017}.
+Training metrics for self-supervised tasks and the likelihood of LVMs offer little guidance as to the quality of the learned representations \parencite{huszar_is_2017}.
 Thus, a common approach is to evaluate the representations in terms of their usefulness for downstream tasks.
@@
 
-The SUPERB benchmark \cite{yang_superb_2021} gathers multiple tasks grouped into categories such as \emph{recognition}, \emph{detection}, \emph{semantics}, \emph{speaker}, \emph{paralinguistics} and \emph{generation}.
-The recently proposed SLUE benchmark focuses on spoken language understanding \cite{shon_slue_2021}.
-The long-standing zero resource speech challenge (ZeroSpeech) offers a new set of tasks for each edition \cite{versteegh_zero_2015, dunbar_zero_2017, dunbar_zero_2019, dunbar_zero_2020, dunbar_zero_2021} usually featuring a minimal-pair ABX task \cite{schatz_evaluating_2013, schatz_evaluating_2014}.
+The SUPERB benchmark \parencite{yang_superb_2021} gathers multiple tasks grouped into categories such as \emph{recognition}, \emph{detection}, \emph{semantics}, \emph{speaker}, \emph{paralinguistics} and \emph{generation}.
+The recently proposed SLUE benchmark focuses on spoken language understanding \parencite{shon_slue_2021}.
+The long-standing zero resource speech challenge (ZeroSpeech) offers a new set of tasks for each edition \parencite{versteegh_zero_2015, dunbar_zero_2017, dunbar_zero_2019, dunbar_zero_2020, dunbar_zero_2021} usually featuring a minimal-pair ABX task \parencite{schatz_evaluating_2013, schatz_evaluating_2014}.
 
-Tasks that evaluate representations in terms of speaker-related information include speaker verification \cite{hsu_unsupervised_2017, khurana_factorial_2019, milde_unspeech_2018}, speaker identification \cite{oord_representation_2018, jati_neural_2019, chung_unsupervised_2019, liu_nonautoregressive_2020}, dialect classification \cite{khurana_factorial_2019}, emotion recognition \cite{pascual_learning_2019, yang_superb_2021} and gender classification \cite{lee_unsupervised_2009}.
-The semantic content of representations are evaluated using tasks such as intent classification \cite{morais_endtoend_2021, yang_superb_2021}, slot filling \cite{lai_semisupervised_2021, yang_superb_2021}, sentiment analysis \cite{liu_mockingjay_2020}, question answering \cite{chung_splat_2021}, named entity recognition \cite{shon_slue_2021, borgholt_we_2021, pasad_use_2022} and speech translation \cite{bansal_speechtotext_2017, chung_generative_2020}.
-Cardiac arrest detection for emergency calls has also been used to evaluate speech representations \cite{borgholt_we_2021}.
-For local representations, phoneme classification is very common \cite{lee_unsupervised_2009, hsu_learning_2017, chorowski_unsupervised_2019, chung_unsupervised_2019, liu_tera_2021}.
-However, automatic speech recognition has become the \textit{de facto} standard benchmark task \cite{ling_decoar_2020, chung_generative_2020, hsu_hubert_2021}.
+Tasks that evaluate representations in terms of speaker-related information include speaker verification \parencite{hsu_unsupervised_2017, khurana_factorial_2019, milde_unspeech_2018}, speaker identification \parencite{oord_representation_2018, jati_neural_2019, chung_unsupervised_2019, liu_nonautoregressive_2020}, dialect classification \parencite{khurana_factorial_2019}, emotion recognition \parencite{pascual_learning_2019, yang_superb_2021} and gender classification \parencite{lee_unsupervised_2009}.
+The semantic content of representations are evaluated using tasks such as intent classification \parencite{morais_endtoend_2021, yang_superb_2021}, slot filling \parencite{lai_semisupervised_2021, yang_superb_2021}, sentiment analysis \parencite{liu_mockingjay_2020}, question answering \parencite{chung_splat_2021}, named entity recognition \parencite{shon_slue_2021, borgholt_we_2021, pasad_use_2022} and speech translation \parencite{bansal_speechtotext_2017, chung_generative_2020}.
+Cardiac arrest detection for emergency calls has also been used to evaluate speech representations \parencite{borgholt_we_2021}.
+For local representations, phoneme classification is very common \parencite{lee_unsupervised_2009, hsu_learning_2017, chorowski_unsupervised_2019, chung_unsupervised_2019, liu_tera_2021}.
+However, automatic speech recognition has become the \textit{de facto} standard benchmark task \parencite{ling_decoar_2020, chung_generative_2020, hsu_hubert_2021}.
 
 \paragraph{Moving forward} 
-Most of the seminal work has focused on improving speech recognition \cite{schneider_wav2vec_2019, baevski_wav2vec_2020}.
-This focus has gained traction over the last couple of years, as computational resources have become more accessible and end-to-end models \cite{graves_connectionist_2006, chan_listen_2015} have been established as the dominant approach to speech recognition \cite{gulati_conformer_2020}.
-It is important to stress that self-supervised models, such as wav2vec 2.0 \cite{baevski_wav2vec_2020}, represent a breakthrough, and recent successful approaches build upon this method.
-That is, deep self-attention models combined with masking \cite{hsu_hubert_2021, wang_unispeech_2021, chen_wavlm_2021}.
-This development mirrors years of rapid progress in masked language modeling within natural language processing \cite{devlin_bert_2018,clark_electra_2020} and we expect this to continue for unsupervised neural speech representation learning.
+Most of the seminal work has focused on improving speech recognition \parencite{schneider_wav2vec_2019, baevski_wav2vec_2020}.
+This focus has gained traction over the last couple of years, as computational resources have become more accessible and end-to-end models \parencite{graves_connectionist_2006, chan_listen_2015} have been established as the dominant approach to speech recognition \parencite{gulati_conformer_2020}.
+It is important to stress that self-supervised models, such as wav2vec 2.0 \parencite{baevski_wav2vec_2020}, represent a breakthrough, and recent successful approaches build upon this method.
+That is, deep self-attention models combined with masking \parencite{hsu_hubert_2021, wang_unispeech_2021, chen_wavlm_2021}.
+This development mirrors years of rapid progress in masked language modeling within natural language processing \parencite{devlin_bert_2018,clark_electra_2020} and we expect this to continue for unsupervised neural speech representation learning.
 

### 2023-09-29 appendices/supplementary_benchmarking.tex
Some formatting and changes to labels
Move table captions to top
Updates to citation commands and footnotes

--- 2023-08-11 appendices/appendix_benchmarking.tex
@@
 \chapter[supplementary material: benchmarking generative latent variable models for speech]{Supplementary material: Benchmarking generative latent variable models for speech}
-\label{app:benchmarking}
+\label{app:supplementary-benchmarking}
 \ifthenelse{\equal{\skipappendices}{true}}{}{
@@
 Specifically, the male to female ratio is about two to one while the vast majority of speakers are Caucasian.
-Such statistics might have an effect of some ethical concern on downstream applications derived from such a dataset as also highlighted in recent research \cite{koenecke_racial_2020}.
+Such statistics might have an effect of some ethical concern on downstream applications derived from such a dataset as also highlighted in recent research \parencite{koenecke_racial_2020}.
 In LibriSpeech, there is an approximately equal number of female and male speakers while the diversity in race is unknown to the authors.
@@
 \paragraph{TIMIT} 
-TIMIT \cite{garofolo_timit_1993} is a speech dataset which contains $\SI{16}{kHz}$ recordings of 630 speakers of eight major dialects of American English, each reading ten phonetically rich sentences.
+TIMIT \parencite{garofolo_timit_1993} is a speech dataset which contains $\SI{16}{kHz}$ recordings of 630 speakers of eight major dialects of American English, each reading ten phonetically rich sentences.
 It amounts to 6300 total recordings splits approximately in 3.94 hours of audio for training and 1.43 hours of audio for testing.
 No speakers or sentences in the test set are in the training set.
-The full train and test subsets of TIMIT are as in previous work \cite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019}. 
+The full train and test subsets of TIMIT are as in previous work \parencite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019}. 
 We randomly sample 5\% of the training set to use as a validation set. 
-TIMIT includes temporally aligned annotations of phonemes and words as well as speaker metadata such as gender, height, age, race, education level and dialect region \cite{garofolo_timit_1993}.
+TIMIT includes temporally aligned annotations of phonemes and words as well as speaker metadata such as gender, height, age, race, education level and dialect region \parencite{garofolo_timit_1993}.
 
 \paragraph{LibriSpeech and LibriLight}
-The LibriSpeech dataset \cite{panayotov_librispeech_2015} consists of readings of public domain audio books amounting to approximately $\SI{1000}{h}$ of audio.
+The LibriSpeech dataset \parencite{panayotov_librispeech_2015} consists of readings of public domain audio books amounting to approximately $\SI{1000}{h}$ of audio.
 The data is derived from the LibriVox project.
-LibriLight \cite{kahn_libri-light_2020} is a subset of LibriSpeech created as an automatic speech transcription (ASR) benchmark with limited or no supervision.
+LibriLight \parencite{kahn_libri-light_2020} is a subset of LibriSpeech created as an automatic speech transcription (ASR) benchmark with limited or no supervision.
 We specifically train on the $\SI{100}{h}$ \textrm{train-clean-100} subset of LibriSpeech and the $\SI{10}{h}$ subset of LibriLight. 
@@
 \paragraph{WaveNet}
-We implement WaveNet as described in the original work \cite{oord_wavenet_2016} but use a discretized mixture of logistics as the output distribution as also done in other work \cite{oord_parallel_2018}.
+We implement WaveNet as described in the original work \parencite{oord_wavenet_2016} but use a discretized mixture of logistics as the output distribution as also done in other work \parencite{oord_parallel_2018}.
 Our WaveNet is not conditioned on any signal other than the raw waveform.
@@
 \paragraph{VRNN}
-We implement the VRNN as described in the original work \cite{chung_recurrent_2015} and verify that we can reproduce the original Gaussian likelihood TIMIT results.
+We implement the VRNN as described in the original work \parencite{chung_recurrent_2015} and verify that we can reproduce the original Gaussian likelihood TIMIT results.
 We replace the Gaussian output distribution with the DMoL.
@@
 \paragraph{SRNN}
-We implement the VRNN as described in the original work \cite{fraccaro_sequential_2016} and verify that we can reproduce the original Gaussian likelihood TIMIT results.
+We implement the VRNN as described in the original work \parencite{fraccaro_sequential_2016} and verify that we can reproduce the original Gaussian likelihood TIMIT results.
 We replace the Gaussian output distribution with the DMoL.
 
-\paragraph{CW-VAE} We implement the CW-VAE based on the original work \cite{saxena_clockwork_2021} but with some modifications also briefly described in section~\cref{sec: cwvae on speech}.
+\paragraph{CW-VAE} We implement the CW-VAE based on the original work \parencite{saxena_clockwork_2021} but with some modifications also briefly described in section~\cref{sec: cwvae on speech}.
 We replace the encoder/decoder model architectures of the original work with architectures designed for waveform modeling.
-Specifically, the encoder and decoder are based on the Conv-TasNet \cite{luo_conv-tasnet_2019} and uses similar residual block structure.
+Specifically, the encoder and decoder are based on the Conv-TasNet \parencite{luo_conv-tasnet_2019} and uses similar residual block structure.
 However, contrary to the Conv-TasNet, we require downsampling factors larger than two.
@@
 
-\paragraph{STCN} We implement the STCN as described in the original work \cite{aksan_stcn_2019} and verify that we can reproduce the original Gaussian likelihood TIMIT results.
+\paragraph{STCN} We implement the STCN as described in the original work \parencite{aksan_stcn_2019} and verify that we can reproduce the original Gaussian likelihood TIMIT results.
 We replace the Gaussian output distribution with the DMoL.
@@
 \paragraph{Likelihood benchmark}
-We implement all models and training scripts in PyTorch 1.9 \cite{paszke_automatic_2017}.
-For both datasets we use the Adam optimizer \cite{kingma_adam_2015} with default parameters as given in PyTorch. We use learning rate $3\text{e}-4$ and no learning rate schedule.
+We implement all models and training scripts in PyTorch 1.9 \parencite{paszke_automatic_2017}.
+For both datasets we use the Adam optimizer \parencite{kingma_adam_2015} with default parameters as given in PyTorch. We use learning rate $3\text{e}-4$ and no learning rate schedule.
 We use PyTorch automatic mixed precision (AMP) to significantly reduce memory consumption.
@@
 The pre-training is done as for the likelihood benchmark above.
-The ASR model is trained using the Adam optimizer \cite{kingma_adam_2015} with default parameters as given in PyTorch.
+The ASR model is trained using the Adam optimizer \parencite{kingma_adam_2015} with default parameters as given in PyTorch.
 We use learning rate $3\text{e}-4$ and no learning rate schedule.
@@
 
-\paragraph{TIMIT, linear, Gaussian} We also provide some results on TIMIT with the audio instead represented as linear PCM (linearly encoded) and using Gaussian output distributions as has been done previously in the literature \cite{chung_recurrent_2015, fraccaro_sequential_2016, lai_stochastic_2018, aksan_stcn_2019}.
+\paragraph{TIMIT, linear, Gaussian} We also provide some results on TIMIT with the audio instead represented as linear PCM (linearly encoded) and using Gaussian output distributions as has been done previously in the literature \parencite{chung_recurrent_2015, fraccaro_sequential_2016, lai_stochastic_2018, aksan_stcn_2019}.
 We use $s=200$ for comparability to the previous work.
@@
 We constrain the variance of the Gaussians used with our models to be at least $\sigma^2_\text{min} = 0.01^2$ in order to avoid the variance going to zero, the likelihood going to infinity and optimization becoming unstable.
-The Gaussian standard deviation is clamped at minimum $0.001$ by \cite{aksan_stcn_2019}.
+The Gaussian standard deviation is clamped at minimum $0.001$ by \parencite{aksan_stcn_2019}.
 
@@
 not $\mu$-law encoded) does not compare as favorably to the other baselines as it did with the discretized mixture of logistics distribution.
-We hypothesize that this has to do with using a Gaussian output distribution in latent variable models which, as has been reported elsewhere \cite{mattei_leveraging_2018}, leads to a likelihood function that is unbounded above and can grow arbitrarily high.
+We hypothesize that this has to do with using a Gaussian output distribution in latent variable models which, as has been reported elsewhere \parencite{mattei_leveraging_2018}, leads to a likelihood function that is unbounded above and can grow arbitrarily high.
 We discuss this phenomenon in further detail in section~\cref{app: gaussian likelihood unboundedness discussion}.
@@
 \begin{table}[t!]
+    \caption[Model likelihoods on TIMIT represented as a 16 bit linear PCM]{
+    Model likelihoods on TIMIT represented as a 16 bit linear PCM. The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
+    }
     \centering
@@
     \vspace{2mm}
-    \caption[Model likelihoods on TIMIT represented as a 16 bit linear PCM]{
-    Model likelihoods on TIMIT represented as a 16 bit linear PCM. The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
-    }
     \label{tab: timit likelihoods dmol linear appendix}
@@
 \begin{table}[p]
+    \caption[Model likelihoods on TIMIT represented as a 16 bit $\boldsymbol{\mu}$-law PCM]{
+    Model likelihoods on TIMIT represented as a 16 bit $\boldsymbol{\mu}$-law encoded PCM.
+    }
     \centering
@@
     }
-    \caption[Model likelihoods on TIMIT represented as a 16 bit $\boldsymbol{\mu}$-law PCM]{
-    Model likelihoods on TIMIT represented as a 16 bit $\boldsymbol{\mu}$-law encoded PCM.
-    }
     \label{tab: timit likelihoods dmol mu-law appendix}
@@
 \begin{table}[t!]
+    \caption[Model likelihoods on TIMIT represented as globally normalized 16 bit linear PCM]{
+    Model likelihoods on TIMIT represented as globally normalized 16 bit linear PCM.
+Contrary to the other likelihoods reported in this paper, here they are given in units of nats and obtained by summing the likelihood over time and over all examples in the dataset and dividing by the total number of examples.
+In the table, Normal refers to using a Gaussian likelihood and GMM refers to using a  Gaussian Mixture Model likelihood with 20 components.
+Models with asterisks $^*$ are our implementations while remaining results are as reported in the referenced work.
+    }
     \centering
@@
         % 200 & WaveNet                                                              & GMM    & XX.XX
-        200 & WaveNet {\scriptsize \cite{aksan_stcn_2019}}                        & GMM-20    & 30188 \\
-        200 & WaveNet {\scriptsize \cite{aksan_stcn_2019}}                        & Normal & -7443 \\
-        200 & Stochastic WaveNet$^*$ {\scriptsize \cite{lai_stochastic_2018}}     & Normal & $\geq$72463\\
-        % 200 & VRNN {\scriptsize \cite{chung_recurrent_2015}}                          & & \\
-        200 & VRNN {\scriptsize \cite{chung_recurrent_2015}}                      & Normal & $\approx$28982\\
-        % 200 & SRNN {\scriptsize \cite{fraccaro_sequential_2016}}                      & & \\
-        200 & SRNN {\scriptsize \cite{fraccaro_sequential_2016}}                  & Normal & $\geq$60550 \\
-        200 & STCN {\scriptsize \cite{aksan_stcn_2019}}                           & GMM-20 & $\geq$69195\\
-        200 & STCN {\scriptsize \cite{aksan_stcn_2019}}                           & Normal & $\geq$64913\\
-        200 & STCN-dense {\scriptsize \cite{aksan_stcn_2019}}                     & GMM-20 & $\geq$71386\\
-        200 & STCN-dense {\scriptsize \cite{aksan_stcn_2019}}                     & Normal & $\geq$70294\\
-        200 & STCN-dense-large {\scriptsize \cite{aksan_stcn_2019}}               & GMM-20 & $\geq$77438\\
+        200 & WaveNet {\scriptsize \parencite{aksan_stcn_2019}}                        & GMM-20    & 30188 \\
+        200 & WaveNet {\scriptsize \parencite{aksan_stcn_2019}}                        & Normal & -7443 \\
+        200 & Stochastic WaveNet$^*$ {\scriptsize \parencite{lai_stochastic_2018}}     & Normal & $\geq$72463\\
+        % 200 & VRNN {\scriptsize \parencite{chung_recurrent_2015}}                          & & \\
+        200 & VRNN {\scriptsize \parencite{chung_recurrent_2015}}                      & Normal & $\approx$28982\\
+        % 200 & SRNN {\scriptsize \parencite{fraccaro_sequential_2016}}                      & & \\
+        200 & SRNN {\scriptsize \parencite{fraccaro_sequential_2016}}                  & Normal & $\geq$60550 \\
+        200 & STCN {\scriptsize \parencite{aksan_stcn_2019}}                           & GMM-20 & $\geq$69195\\
+        200 & STCN {\scriptsize \parencite{aksan_stcn_2019}}                           & Normal & $\geq$64913\\
+        200 & STCN-dense {\scriptsize \parencite{aksan_stcn_2019}}                     & GMM-20 & $\geq$71386\\
+        200 & STCN-dense {\scriptsize \parencite{aksan_stcn_2019}}                     & Normal & $\geq$70294\\
+        200 & STCN-dense-large {\scriptsize \parencite{aksan_stcn_2019}}               & GMM-20 & $\geq$77438\\
         200 & CW-VAE$^*$                                         & $L=1, D_z=96$, Normal & $\geq$41629 \\  % https://wandb.ai/vseq/cwvae/runs/n8nu778t?workspace=user-jaksen
@@
     \vspace{2mm}
-    \caption[Model likelihoods on TIMIT represented as globally normalized 16 bit linear PCM]{
-    Model likelihoods on TIMIT represented as globally normalized 16 bit linear PCM.
-Contrary to the other likelihoods reported in this paper, here they are given in units of nats and obtained by summing the likelihood over time and over all examples in the dataset and dividing by the total number of examples.
-In the table, Normal refers to using a Gaussian likelihood and GMM refers to using a  Gaussian Mixture Model likelihood with 20 components.
-Models with asterisks $^*$ are our implementations while remaining results are as reported in the referenced work.
-    }
     \label{tab: timit likelihoods gaussian appendix}
@@
 \begin{table}[t!]
+    \caption[The highest possible Gaussian log-likelihoods attainable on the TIMIT test set]{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
     \centering
@@
     \vspace{2mm}
-    \caption[The highest possible Gaussian log-likelihoods attainable on the TIMIT test set]{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
     \label{tab:timit best gaussian ll}
@@
 
-Interestingly, most work on latent variable models with non-autoregressive output distributions seem to ignore this fact and simply employ the mixture distribution with 10 mixture components \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
+Interestingly, most work on latent variable models with non-autoregressive output distributions seem to ignore this fact and simply employ the mixture distribution with 10 mixture components \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
 However, given the empirically good results of latent variable models for image generation, this seems to have posed only a minor problem in practice.
@@
 In \cref{fig: cwvae cell state graph} we show the graphical model of the recurrent cell of the CW-VAE for a single time step.
-As noted in \cite{saxena_clockwork_2021}, this cell is very similar to the one of the Recurrent State Space Model (RSSM) \cite{hafner_learning_2019}.
+As noted in \parencite{saxena_clockwork_2021}, this cell is very similar to the one of the Recurrent State Space Model (RSSM) \parencite{hafner_learning_2019}.
 In \cref{fig: cwvae three-layered graphical models (inference and generative)} we show the unrolled graphical models of a three-layered CW-VAE with $k_1=1$ and $c=2$ yielding $k_2=2$ and $k_3=4$. 
 We show both the generative and inference models and highlight in blue the parameter sharing between the two models due to top-down inference.
-In \cref{fig: stcn graphical models (inference and generative)} we show the graphical models of the STCN \cite{aksan_stcn_2019} at a single timestep.
+In \cref{fig: stcn graphical models (inference and generative)} we show the graphical models of the STCN \parencite{aksan_stcn_2019} at a single timestep.
 The model has three layers and shares the parameters of the WaveNet encoder between the inference and generative models.
-In \cref{fig: vrnn graphical models (inference and generative)} we illustrate the unrolled graphical models of the inference and generative models of the VRNN \cite{chung_recurrent_2015}.
+In \cref{fig: vrnn graphical models (inference and generative)} we illustrate the unrolled graphical models of the inference and generative models of the VRNN \parencite{chung_recurrent_2015}.
 We include the deterministic variable $\db_t$ in order to illustrate the difference to other latent variable models.
-Likewise, in \cref{fig: srnn graphical models (inference and generative)} we illustrate the unrolled graphical models the SRNN \cite{fraccaro_sequential_2016}.
+Likewise, in \cref{fig: srnn graphical models (inference and generative)} we illustrate the unrolled graphical models the SRNN \parencite{fraccaro_sequential_2016}.
 
@@
 \caption[CW-VAE cell state $\vs_t^l$ update]{CW-VAE cell state $\vs_t^l$ update.
-The cell state is given as $\vs_t^l=(\zb_t^l, \db_t^l)$ where $\db_t^l$ is the deterministic hidden state of a Gated Recurrent Unit \cite{cho_properties_2014}.
-The vector $\eb_t^l$ is computed from $\xb_t$ by the encoder network which outputs $L$ encodings, one for each latent variable, similar to that of a Ladder VAE \cite{sonderby_ladder_2016}.
+The cell state is given as $\vs_t^l=(\zb_t^l, \db_t^l)$ where $\db_t^l$ is the deterministic hidden state of a Gated Recurrent Unit \parencite{cho_properties_2014}.
+The vector $\eb_t^l$ is computed from $\xb_t$ by the encoder network which outputs $L$ encodings, one for each latent variable, similar to that of a Ladder VAE \parencite{sonderby_ladder_2016}.
 All blue arrows are shared between generation and inference.
@@
 \caption[Generative and inference models for three-layered CW-VAE]{
-    CW-VAE \cite{saxena_clockwork_2021} generative model $p(\xb,\zb)$ in (\subref{fig: cwvae three-layered graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: cwvae three-layered graphical model inference}) for a three-layered model with $k_1=1$ and $c=2$ giving $k_2=2$ and $k_3=4$ unrolled over eight steps in the observed variable.
+    CW-VAE \parencite{saxena_clockwork_2021} generative model $p(\xb,\zb)$ in (\subref{fig: cwvae three-layered graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: cwvae three-layered graphical model inference}) for a three-layered model with $k_1=1$ and $c=2$ giving $k_2=2$ and $k_3=4$ unrolled over eight steps in the observed variable.
 Blue arrows are (mostly) shared between the inference and generative models.
@@
 \caption[Generative and inference models for three-layered STCN]{
-    STCN \cite{aksan_stcn_2019} generative model $p(\xb,\zb)$ in (\subref{fig: stcn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: stcn graphical model inference}) for a single time-step.
+    STCN \parencite{aksan_stcn_2019} generative model $p(\xb,\zb)$ in (\subref{fig: stcn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: stcn graphical model inference}) for a single time-step.
     The WaveNet autoregressive encoder is shared between generative and inference models.
@@
 \caption[Generative and inference models for VRNN]{
-    VRNN \cite{chung_recurrent_2015} generative model $p(\xb,\zb)$ in (\subref{fig: vrnn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: vrnn graphical model inference}) unrolled over three steps in the observed variable.
+    VRNN \parencite{chung_recurrent_2015} generative model $p(\xb,\zb)$ in (\subref{fig: vrnn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: vrnn graphical model inference}) unrolled over three steps in the observed variable.
     Blue arrows are shared between the inference and generative models.
@@
 \caption[Generative and inference models for SRNN]{
-    SRNN \cite{fraccaro_sequential_2016} generative model $p(\xb,\zb)$ in (\subref{fig: srnn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: srnn graphical model inference}) unrolled over three steps in the observed variable.
+    SRNN \parencite{fraccaro_sequential_2016} generative model $p(\xb,\zb)$ in (\subref{fig: srnn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: srnn graphical model inference}) unrolled over three steps in the observed variable.
     Blue arrows are shared between the inference and generative models.

### 2023-09-29 appendices/supplementary_benchmarking_workshop.tex
Some formatting and changes to labels
Move table captions to top
Updates to citation commands and footnotes

--- 2023-03-16 appendices/appendix_benchmarking_workshop.tex
@@
 \chapter[appendix for benchmarking generative latent variable models for speech]{Benchmarking Generative Latent Variable Models for Speech}
-
-\iffalse
+\label{app:supplementary-benchmarking_workshop}
+\ifthenelse{\equal{\skipappendices}{true}}{}{
 
@@
 \paragraph{SRNN}
-We implement the VRNN as described in the original work \cite{fraccaro_sequential_2016} and verify that we can reproduce the original Gaussian likelihood TIMIT results.
+We implement the VRNN as described in the original work \parencite{fraccaro_sequential_2016} and verify that we can reproduce the original Gaussian likelihood TIMIT results.
 We replace the Gaussian output distribution with the DMoL.
 
-\paragraph{CW-VAE} We implement the CW-VAE based on the original work \cite{saxena_clockwork_2021} but with some modifications also briefly described in the paper.
+\paragraph{CW-VAE} We implement the CW-VAE based on the original work \parencite{saxena_clockwork_2021} but with some modifications also briefly described in the paper.
 We replace the encoder/decoder model architectures of the original work with architectures designed for waveform modeling.
-Specifically, the encoder and decoder are based on the Conv-TasNet \cite{luo_conv-tasnet_2019} and uses similar residual block structure.
+Specifically, the encoder and decoder are based on the Conv-TasNet \parencite{luo_conv-tasnet_2019} and uses similar residual block structure.
 However, contrary to the Conv-TasNet, we require downsampling factors larger than two.
@@
 We constrain the standard deviation of the Gaussians used with our models to be at least $\sigma_\text{min} = 0.01$ in order to avoid it going to zero, the likelihood going to infinity and optimization becoming unstable.
-The minimum Gaussian standard deviation of \citet{aksan_stcn_2019} is $\sigma_\text{min} = 0.001$.
+The minimum Gaussian standard deviation of \textcite{aksan_stcn_2019} is $\sigma_\text{min} = 0.001$.
 
@@
 \begin{table*}[t]
+    \caption{
+    Model likelihoods $\mathcal{L}$ on \textbf{LibriSpeech} test sets represented as \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}.
+    For the CW-VAE, $s$ refers to $s_1$ and the two-layered models have $s_2=8s_1$.
+    The models are trained on either the $\SI{10}{h}$ LibriLight subset or the $\SI{100}{h}$ LibriSpeech \textrm{train-clean-100} subset as indicated.
+    Likelihoods are given in units of bits per frame (bpf).
+    }
     \centering
@@
     }
-    \caption{
-    Model likelihoods $\mathcal{L}$ on \textbf{LibriSpeech} test sets represented as \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}.
-    For the CW-VAE, $s$ refers to $s_1$ and the two-layered models have $s_2=8s_1$.
-    The models are trained on either the $\SI{10}{h}$ LibriLight subset or the $\SI{100}{h}$ LibriSpeech \textrm{train-clean-100} subset as indicated.
-    Likelihoods are given in units of bits per frame (bpf).
-    }
     \label{tab: likelihoods librispeech}
@@
 \begin{table}[t!]
+    \caption{
+    Model likelihoods on TIMIT represented as a 16 bit linear PCM, obtained by different latent variable models and compared to autoregressive baselines all using a discretized mixture of logistics with 10 components as output distribution.
+Likelihoods are given in units of bits per frame (bpf) and obtained by normalizing the total likelihood of each sequence with the individual sequence length and then averaging over the dataset.
+The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
+    }
     \centering
@@
     \end{tabular}
-    \caption{
-    Model likelihoods on TIMIT represented as a 16 bit linear PCM, obtained by different latent variable models and compared to autoregressive baselines all using a discretized mixture of logistics with 10 components as output distribution.
-Likelihoods are given in units of bits per frame (bpf) and obtained by normalizing the total likelihood of each sequence with the individual sequence length and then averaging over the dataset.
-The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
-    }
     \label{tab: timit likelihoods dmol linear appendix}
@@
 \begin{table}[p]
+    \caption{
+    Model likelihoods on TIMIT represented as a \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}, obtained by different latent variable models and compared to autoregressive baselines all using a discretized mixture of logistics with 10 components as output distribution.
+Likelihoods are given in units of bits per frame (bpf) and obtained by normalizing the total likelihood of each sequence with the individual sequence length and then averaging over the dataset.
+    }
     \centering
@@
     \end{tabular}
-    \caption{
-    Model likelihoods on TIMIT represented as a \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}, obtained by different latent variable models and compared to autoregressive baselines all using a discretized mixture of logistics with 10 components as output distribution.
-Likelihoods are given in units of bits per frame (bpf) and obtained by normalizing the total likelihood of each sequence with the individual sequence length and then averaging over the dataset.
-    }
     \label{tab: timit likelihoods dmol mu-law appendix}
@@
 \begin{table}[t!]
-    \centering
-    \begin{tabular}{lll|r}
-        \toprule
-        $s$ & \bf Model           & \bf Configuration           & \bf $\mathcal{L}$ [nats] \\
-        \midrule
-        1 & WaveNet                                                               & Normal & 119656 \\
-        1 & WaveNet                                                               & GMM-2  & 120699 \\
-        1 & WaveNet                                                               & GMM-20 & 121681 \\
-        \midrule
-        % 200 & WaveNet                                                              & GMM    & XX.XX
-        200 & WaveNet {\scriptsize \cite{aksan_stcn_2019}}                        & GMM-20    & 30188 \\
-        200 & WaveNet {\scriptsize \cite{aksan_stcn_2019}}                        & Normal & -7443 \\
-        200 & Stochastic WaveNet$^*$ {\scriptsize \cite{lai_stochastic_2018}}     & Normal & $\geq$72463\\
-        % 200 & VRNN {\scriptsize \cite{chung_recurrent_2015}}                          & & \\
-        200 & VRNN {\scriptsize \cite{chung_recurrent_2015}}                      & Normal & $\approx$28982\\
-        % 200 & SRNN {\scriptsize \cite{fraccaro_sequential_2016}}                      & & \\
-        200 & SRNN {\scriptsize \cite{fraccaro_sequential_2016}}                  & Normal & $\geq$60550 \\
-        200 & STCN {\scriptsize \cite{aksan_stcn_2019}}                           & GMM-20 & $\geq$69195\\
-        200 & STCN {\scriptsize \cite{aksan_stcn_2019}}                           & Normal & $\geq$64913\\
-        200 & STCN-dense {\scriptsize \cite{aksan_stcn_2019}}                     & GMM-20 & $\geq$71386\\
-        200 & STCN-dense {\scriptsize \cite{aksan_stcn_2019}}                     & Normal & $\geq$70294\\
-        200 & STCN-dense-large {\scriptsize \cite{aksan_stcn_2019}}               & GMM-20 & $\geq$77438\\
-        200 & CW-VAE$^*$                                         & $L=1, D_z=96$, Normal & $\geq$41629 \\  % https://wandb.ai/vseq/cwvae/runs/n8nu778t?workspace=user-jaksen
-        % 200 & CW-VAE$^*$                                         & $L=2, D_z=96$, Normal & $\geq$2090 \\  % https://wandb.ai/vseq/cwvae/runs/1so36fii?workspace=user-jaksen
-        \bottomrule
-    \end{tabular}
     \caption{
@@
     }
+    \centering
+    \begin{tabular}{lll|r}
+        \toprule
+        $s$ & \bf Model           & \bf Configuration           & \bf $\mathcal{L}$ [nats] \\
+        \midrule
+        1 & WaveNet                                                               & Normal & 119656 \\
+        1 & WaveNet                                                               & GMM-2  & 120699 \\
+        1 & WaveNet                                                               & GMM-20 & 121681 \\
+        \midrule
+        % 200 & WaveNet                                                              & GMM    & XX.XX
+        200 & WaveNet {\scriptsize \parencite{aksan_stcn_2019}}                        & GMM-20    & 30188 \\
+        200 & WaveNet {\scriptsize \parencite{aksan_stcn_2019}}                        & Normal & -7443 \\
+        200 & Stochastic WaveNet$^*$ {\scriptsize \parencite{lai_stochastic_2018}}     & Normal & $\geq$72463\\
+        % 200 & VRNN {\scriptsize \parencite{chung_recurrent_2015}}                          & & \\
+        200 & VRNN {\scriptsize \parencite{chung_recurrent_2015}}                      & Normal & $\approx$28982\\
+        % 200 & SRNN {\scriptsize \parencite{fraccaro_sequential_2016}}                      & & \\
+        200 & SRNN {\scriptsize \parencite{fraccaro_sequential_2016}}                  & Normal & $\geq$60550 \\
+        200 & STCN {\scriptsize \parencite{aksan_stcn_2019}}                           & GMM-20 & $\geq$69195\\
+        200 & STCN {\scriptsize \parencite{aksan_stcn_2019}}                           & Normal & $\geq$64913\\
+        200 & STCN-dense {\scriptsize \parencite{aksan_stcn_2019}}                     & GMM-20 & $\geq$71386\\
+        200 & STCN-dense {\scriptsize \parencite{aksan_stcn_2019}}                     & Normal & $\geq$70294\\
+        200 & STCN-dense-large {\scriptsize \parencite{aksan_stcn_2019}}               & GMM-20 & $\geq$77438\\
+        200 & CW-VAE$^*$                                         & $L=1, D_z=96$, Normal & $\geq$41629 \\  % https://wandb.ai/vseq/cwvae/runs/n8nu778t?workspace=user-jaksen
+        % 200 & CW-VAE$^*$                                         & $L=2, D_z=96$, Normal & $\geq$2090 \\  % https://wandb.ai/vseq/cwvae/runs/1so36fii?workspace=user-jaksen
+        \bottomrule
+    \end{tabular}
     \label{tab: timit likelihoods gaussian appendix}
@@
 \begin{table}[t!]
+    \caption{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
     \centering
@@
     \end{tabular}
-    \caption{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
     \label{tab:timit best gaussian ll}
@@
 In \cref{fig: cwvae cell state graph} we show the graphical model of the recurrent cell of the CW-VAE for a single time step.
-As noted in \cite{saxena_clockwork_2021}, this cell is very similar to the one of the Recurrent State Space Model (RSSM) \cite{hafner_learning_2019}.
+As noted in \parencite{saxena_clockwork_2021}, this cell is very similar to the one of the Recurrent State Space Model (RSSM) \parencite{hafner_learning_2019}.
 
@@
 
-In \cref{fig: stcn graphical models (inference and generative)} we show the graphical models of the STCN \cite{aksan_stcn_2019} at a single timestep.
+In \cref{fig: stcn graphical models (inference and generative)} we show the graphical models of the STCN \parencite{aksan_stcn_2019} at a single timestep.
 The model has three layers and shares the parameters of the WaveNet encoder between the inference and generative models.
 
-In \cref{fig: vrnn graphical models (inference and generative)} we illustrate the unrolled graphical models of the inference and generative models of the VRNN \cite{chung_recurrent_2015}.
+In \cref{fig: vrnn graphical models (inference and generative)} we illustrate the unrolled graphical models of the inference and generative models of the VRNN \parencite{chung_recurrent_2015}.
 We include the deterministic variable $\vd_t$ in order to illustrate the difference to other latent variable models.
 
-Likewise, in \cref{fig: srnn graphical models (inference and generative)} we illustrate the unrolled graphical models the SRNN \cite{fraccaro_sequential_2016}.
+Likewise, in \cref{fig: srnn graphical models (inference and generative)} we illustrate the unrolled graphical models the SRNN \parencite{fraccaro_sequential_2016}.
 
@@
 \caption{CW-VAE cell state $\vs_t^l$ update.
-The cell state is given as $\vs_t^l=(\vz_t^l, \vd_t^l)$ where $\vd_t^l$ is the deterministic hidden state of a Gated Recurrent Unit \cite{cho_properties_2014}.
-The vector $\ve_t^l$ is computed from $\vx_t$ by the encoder network which outputs $L$ encodings, one for each latent variable, similar to that of a Ladder VAE \cite{sonderby_ladder_2016}.
+The cell state is given as $\vs_t^l=(\vz_t^l, \vd_t^l)$ where $\vd_t^l$ is the deterministic hidden state of a Gated Recurrent Unit \parencite{cho_properties_2014}.
+The vector $\ve_t^l$ is computed from $\vx_t$ by the encoder network which outputs $L$ encodings, one for each latent variable, similar to that of a Ladder VAE \parencite{sonderby_ladder_2016}.
 All blue arrows are shared between generation and inference.
@@
 \caption{
-    CW-VAE \cite{saxena_clockwork_2021} generative model $p(\vx,\vz)$ in (\subref{fig: cwvae three-layered graphical model generative}) and inference model $q(\vz|\vx)$ in (\subref{fig: cwvae three-layered graphical model inference}) for a three-layered model with $k_1=1$ and $c=2$ giving $k_2=2$ and $k_3=4$ unrolled over eight steps in the observed variable.
+    CW-VAE \parencite{saxena_clockwork_2021} generative model $p(\vx,\vz)$ in (\subref{fig: cwvae three-layered graphical model generative}) and inference model $q(\vz|\vx)$ in (\subref{fig: cwvae three-layered graphical model inference}) for a three-layered model with $k_1=1$ and $c=2$ giving $k_2=2$ and $k_3=4$ unrolled over eight steps in the observed variable.
 Blue arrows are (mostly) shared between the inference and generative models.
@@
 \caption{
-    STCN \cite{aksan_stcn_2019} generative model $p(\vx,\vz)$ in (\subref{fig: stcn graphical model generative}) and inference model $q(\vz|\vx)$ in (\subref{fig: stcn graphical model inference}) for a single time-step.
+    STCN \parencite{aksan_stcn_2019} generative model $p(\vx,\vz)$ in (\subref{fig: stcn graphical model generative}) and inference model $q(\vz|\vx)$ in (\subref{fig: stcn graphical model inference}) for a single time-step.
     The WaveNet autoregressive encoder is shared between generative and inference models.
@@
 \caption{
-    VRNN \cite{chung_recurrent_2015} generative model $p(\vx,\vz)$ in (\subref{fig: vrnn graphical model generative}) and inference model $q(\vz|\vx)$ in (\subref{fig: vrnn graphical model inference}) unrolled over three steps in the observed variable.
+    VRNN \parencite{chung_recurrent_2015} generative model $p(\vx,\vz)$ in (\subref{fig: vrnn graphical model generative}) and inference model $q(\vz|\vx)$ in (\subref{fig: vrnn graphical model inference}) unrolled over three steps in the observed variable.
     Blue arrows are shared between the inference and generative models.
@@
 \caption{
-    SRNN \cite{fraccaro_sequential_2016} generative model $p(\vx,\vz)$ in (\subref{fig: srnn graphical model generative}) and inference model $q(\vz|\vx)$ in (\subref{fig: srnn graphical model inference}) unrolled over three steps in the observed variable.
+    SRNN \parencite{fraccaro_sequential_2016} generative model $p(\vx,\vz)$ in (\subref{fig: srnn graphical model generative}) and inference model $q(\vz|\vx)$ in (\subref{fig: srnn graphical model inference}) unrolled over three steps in the observed variable.
     Blue arrows are shared between the inference and generative models.
@@
 This approximately marginalizes out variation during the phoneme.
-We use linear discriminant analysis (LDA) \cite{fisher_use_1936} to obtain a low-dimensional linear subspace of the latent space.
+We use linear discriminant analysis (LDA) \parencite{fisher_use_1936} to obtain a low-dimensional linear subspace of the latent space.
 We visualize the resulting representations colored according to test set phoneme classes in \cref{fig: latent space phoneme and knn}. 

### 2023-09-29 appendices/supplementary_hierarchical.tex
Updates to citation commands and footnotes

@@
 
-% Similar to \citet{serra_input_2020} and \citet{xiao_likelihood_2020}, we also create a uniform noise dataset ``UniformNoise'' and benchmark our model against this.
+% Similar to \textcite{serra_input_2020} and \textcite{xiao_likelihood_2020}, we also create a uniform noise dataset ``UniformNoise'' and benchmark our model against this.
 We dynamically sample each pixel value uniformly as an integer between 0 and 255.
@@
         \midrule
-        FashionMNIST \cite{xiao_fashionmnist_2017} & $28\times28\times1$ & 70,000 \\
-        MNIST \cite{lecun_gradientbased_1998} & $28\times28\times1$ & 70,000 \\
-        notMNIST \cite{bulatov_notmnist_2011} & $28\times28\times1$ & 547,838 \\
-        KMNIST \cite{clanuwat_deep_2018} & $28\times28\times1$ & 70,000 \\
-        Omniglot \cite{lake_humanlevel_2015} & $28\times28\times1$ & 32,460 \\
-        SmallNORB \cite{lecun_learning_2004} & $28\times28\times1$ & 97,200 \\
+        FashionMNIST \parencite{xiao_fashionmnist_2017} & $28\times28\times1$ & 70,000 \\
+        MNIST \parencite{lecun_gradientbased_1998} & $28\times28\times1$ & 70,000 \\
+        notMNIST \parencite{bulatov_notmnist_2011} & $28\times28\times1$ & 547,838 \\
+        KMNIST \parencite{clanuwat_deep_2018} & $28\times28\times1$ & 70,000 \\
+        Omniglot \parencite{lake_humanlevel_2015} & $28\times28\times1$ & 32,460 \\
+        SmallNORB \parencite{lecun_learning_2004} & $28\times28\times1$ & 97,200 \\
         % UniformNoise & $28\times28\times1$ & dynamic \\
         \midrule
-        CIFAR10 \cite{krizhevsky_learning_2009} & $32\times32\times3$ & 60,000 \\
-        SVHN \cite{netzer_reading_2011} & $32\times32\times3$ & 99,289 \\
+        CIFAR10 \parencite{krizhevsky_learning_2009} & $32\times32\times3$ & 60,000 \\
+        SVHN \parencite{netzer_reading_2011} & $32\times32\times3$ & 99,289 \\
         \bottomrule
@@
 Our Hierarchical VAE (HVAE) model uses bottom-up inference and top-down generative paths as specified in the paper.
-For grey-scale images, the output is parameterized by a Bernoulli distribution while for natural images we use a Discretized Logistic Mixture \cite{salimans_pixelcnn_2017}.
+For grey-scale images, the output is parameterized by a Bernoulli distribution while for natural images we use a Discretized Logistic Mixture \parencite{salimans_pixelcnn_2017}.
 The latent variables are parameterized by stochastic layers that output the mean and log-variance of a diagonal covariance Gaussian.
@@
 For grey-scale we use 64 channels while we use 256 for natural images.
-In both cases, the first deterministic block uses a kernel size of 5 and the latter two a kernel of size 3. We use the ReLU activation function \cite{fukushima_neocognitron_1980, nair_rectified_2010}.
+In both cases, the first deterministic block uses a kernel size of 5 and the latter two a kernel of size 3.
+We use the ReLU activation function \parencite{fukushima_neocognitron_1980, nair_rectified_2010}.
 
-Since the benefits and drawbacks of using batch normalization \cite{ioffe_batch_2015} in hierarchical VAEs is still the matter of some debate \cite{sonderby_ladder_2016, vahdat_nvae_2020, child_very_2021} we choose to use weight normalization \cite{salimans_weight_2016} as in other work \cite{maaloe_biva_2019} and initialize the model using the originally proposed data-dependent initialization.
+Since the benefits and drawbacks of using batch normalization \parencite{ioffe_batch_2015} in hierarchical VAEs is still the matter of some debate \parencite{sonderby_ladder_2016, vahdat_nvae_2020, child_very_2021} we choose to use weight normalization \parencite{salimans_weight_2016} as in other work \parencite{maaloe_biva_2019} and initialize the model using the originally proposed data-dependent initialization.
 To have the stochastic layers initialize to standard Gaussian distributions (zero mean, unit variance), with this initialization, we select the activation function for the variance as a Softplus,
@@
 \subsection{BIVA}
-For the BIVA model \cite{maaloe_biva_2019}, we use a specification that is very similar to that of the HVAE above, and to that of the original paper.
+For the BIVA model \parencite{maaloe_biva_2019}, we use a specification that is very similar to that of the HVAE above, and to that of the original paper.
 The model has 10 latent variables the lowest 3 of which are spatial and the rest are densely connected in order to have an architecture similar to the HVAE.
@@
 
-\begin{table}[h]
+\begin{table}[t]
     \caption[Selection of most important hyperparameters]{Selection of most important hyperparameters and their setting.
@@
         \midrule
-        Optimization & Adam \cite{kingma_adam_2015} \\
+        Optimization & Adam \parencite{kingma_adam_2015} \\
         Learning rate & $3e-4$ \\
@@
         \multirow{2}{*}{Initialization} & Data-dependent \\
-        & \cite{salimans_weight_2016} \\
+        & \parencite{salimans_weight_2016} \\
         \midrule
@@
 \section{Note on the KL-term of hierarchical VAEs}
-In this research we choose model parameterizations relying on bottom-up inference \cite{burda_importance_2016},
+In this research we choose model parameterizations relying on bottom-up inference \parencite{burda_importance_2016},
 \begin{equation}
@@
 
-In contrast, a top-down inference model \cite{sonderby_ladder_2016} has a topmost latent variable $\zb_L$ that is independent of the other latent variables and is directly given by $\xb$.
+In contrast, a top-down inference model \parencite{sonderby_ladder_2016} has a topmost latent variable $\zb_L$ that is independent of the other latent variables and is directly given by $\xb$.
 \begin{equation}
@@
 We hypothesize that the covariance of latent variables towards the top of the hierarchy with other latent variables is important for learning semantic representations.
-However, top-down inference models are easier to optimize as has recently been demonstrated \cite{sonderby_ladder_2016, vahdat_nvae_2020, child_very_2021}.
+However, top-down inference models are easier to optimize as has recently been demonstrated \parencite{sonderby_ladder_2016, vahdat_nvae_2020, child_very_2021}.
 

### 2023-09-29 appendices/supplementary_modelagnostic.tex
Move table captions to top
Updates to citation commands and footnotes

@@
 Much more complex approximations have been derived, although diagonal approximations have been consistently used (e.g.
-by \citealp{kirkpatrick_overcoming_2017}, who used essentially the same approximation in a supervised context), and are linked to several adaptive optimisation techniques like Adam \citep{kingma_adam_2015} or RMSProp \citep{tieleman_lecture_2012}.
+by \citealp{kirkpatrick_overcoming_2017}, who used essentially the same approximation in a supervised context), and are linked to several adaptive optimisation techniques like Adam \parencite{kingma_adam_2015} or RMSProp \parencite{tieleman_lecture_2012}.
 A good discussion on these issues is provided in Martens's \citeyearpar{martens_new_2020} recent review.
@@
 assuming that $\pi$ and $\Sigma$ are fixed at their maximum likelihood estimates.
-Similar mixture-based Fisher kernels have been very popular in the past, and were actually a key element of state-of-the art classification models on Imagenet before deep nets won the competition \citep{perronnin_improving_2010}.
+Similar mixture-based Fisher kernels have been very popular in the past, and were actually a key element of state-of-the art classification models on Imagenet before deep nets won the competition \parencite{perronnin_improving_2010}.
 Our idea is to re-use ideas introduced by this computer vision litterature.
@@
 Due to the intractability of the posterior, we cannot directly optimize the likelihood of the model, but instead the model parameters are optimized by maximizing the evidence lower bound (ELBO): $\log p_{\theta}(\mathbf{x}) \geq \mathbb{E}_{q_{\phi}(\mathbf{z}\mid \mathbf{x})} \left[ \log \frac{p(\mathbf{x}, \mathbf{z})}{q_{\phi}(\mathbf{z}\mid \mathbf{x})} \right] \equiv \mathcal{L} $.
-In this work we are considering an Hierarchical VAE (HVAE) with bottom-up inference as done in \cite{havtorn_hierarchical_2021}.
+In this work we are considering an Hierarchical VAE (HVAE) with bottom-up inference as done in \parencite{havtorn_hierarchical_2021}.
 This is an extension of the VAE framework that consider an hierarchy of $L$ latent variables $\mathbf{z} = \mathbf{z}_1,\dots, \mathbf{z}_L$.
@@
 \paragraph{Glow} For training Glow models we follow \textcite{kirichenko_why_2020} and their repository\footnote{\href{https://github.com/PolinaKirichenko/flows_ood}{\texttt{https://github.com/PolinaKirichenko/flows\_ood}}}.
-They closely follow \textcite{nalisnick_deep_2019} and \cite{kingma_glow_2018} implementation for multi-scale Glow, where a scale is defined as the sequence of actorm, invertible $1\times1$ convolution and coupling layers.
+They closely follow \textcite{nalisnick_deep_2019} and \parencite{kingma_glow_2018} implementation for multi-scale Glow, where a scale is defined as the sequence of actorm, invertible $1\times1$ convolution and coupling layers.
 While \textcite{kirichenko_why_2020} only considers the RMSProp optimizer, we trained two different models, one using RMSProp and one using Adam with batch-size 32.
@@
 
-\paragraph{Hierarchical VAE} We follow \cite{havtorn_hierarchical_2021} for both model architecture design and training choices for our hierarchical VAEs.
+\paragraph{Hierarchical VAE} We follow \parencite{havtorn_hierarchical_2021} for both model architecture design and training choices for our hierarchical VAEs.
 We used their open-sourced repository\footnote{\href{https://github.com/JakobHavtorn/hvae-oodd}{\texttt{https://github.com/JakobHavtorn/hvae-oodd}}}.
@@
 We trained each model for 1000 epochs using Adam optimizer with learning rate $3e-4$ and a batch-size of $128$.
-All models were initialized using the data-dependent initialization and they used weight-normalization \citep{salimans_weight_2016}.
+All models were initialized using the data-dependent initialization and they used weight-normalization \parencite{salimans_weight_2016}.
 In addition to that, we always consider a hierarchy of three latent variables.
@@
 \label{appendix_modelagnostic:harmonic}
-In the paper we mentioned that another way to combine $p$-values from different test statistics is the Harmonic mean \citep{wilson_harmonic_2019}. This is defined as:
+In the paper we mentioned that another way to combine $p$-values from different test statistics is the Harmonic mean \parencite{wilson_harmonic_2019}. This is defined as:
 \begin{equation}
@@
     }
-    \vspace*{-0.85\baselineskip}
+    % \vspace*{-0.85\baselineskip}
 \end{table*}
@@
         \label{tab_modelagnostic:all_combinations}
-    \vspace*{-\baselineskip}
+    % \vspace*{-\baselineskip}
 \end{table*}
@@
         \label{tab_modelagnostic:variability}
-    \vspace*{-\baselineskip}
+    % \vspace*{-\baselineskip}
 \end{table*}
@@
     \label{fig_modelagnostic:type1_cifar}
-    \vspace*{-\baselineskip}
+    % \vspace*{-\baselineskip}
 \end{figure}
@@
     }
-    \vspace*{-\baselineskip}
+    % \vspace*{-\baselineskip}
 \end{table*}
@@
 Indeed, transformations applied to the input data play an important role by transforming a difficult task into an easier problem where the likelihood can detect OOD data.
-By looking at the different results for Glow trained on CIFAR10 and tested on CelebA shown in \cite{hendrycks_deep_2019}, \cite{kirichenko_why_2020}, \cite{morningstar_density_2021}, and \cite{ahmadian_likelihoodfree_2021} we can see that the AUROC scores obtain by the plain log-likelihood are pretty different.
-In \cite{hendrycks_deep_2019} and  \cite{kirichenko_why_2020} the log-likelihood gets a poor performance, confirming that CIFAR10-CelebA is a challenging pair for DGMs, while in \cite{morningstar_density_2021} the likelihood is able to distinguish OOD data.
+By looking at the different results for Glow trained on CIFAR10 and tested on CelebA shown in \parencite{hendrycks_deep_2019}, \parencite{kirichenko_why_2020}, \parencite{morningstar_density_2021}, and \parencite{ahmadian_likelihoodfree_2021} we can see that the AUROC scores obtain by the plain log-likelihood are pretty different.
+In \parencite{hendrycks_deep_2019} and  \parencite{kirichenko_why_2020} the log-likelihood gets a poor performance, confirming that CIFAR10-CelebA is a challenging pair for DGMs, while in \parencite{morningstar_density_2021} the likelihood is able to distinguish OOD data.
 While the main reason for these different results can be due to model implementation and training procedure, we decided to investigate how different transformations can influence OOD detection.
@@
 In \cref{table:celeba} we show how the OOD performance changes for our considered models when testing on CelebA where we applied different interpolations.
-We can notice that when using the bilinear interpolation we get results that are pretty similar to \cite{hendrycks_deep_2019}, \cite{kirichenko_why_2020}, and \cite{ahmadian_likelihoodfree_2021} in terms of likelihood OOD performance.
-When using the nearest interpolation, instead, we get results that are closer to \cite{morningstar_density_2021}.
+We can notice that when using the bilinear interpolation we get results that are pretty similar to \parencite{hendrycks_deep_2019}, \parencite{kirichenko_why_2020}, and \parencite{ahmadian_likelihoodfree_2021} in terms of likelihood OOD performance.
+When using the nearest interpolation, instead, we get results that are closer to \parencite{morningstar_density_2021}.
 
@@
     \label{fig_modelagnostic:celeba}
-     \vspace*{-\baselineskip}
+    %  \vspace*{-\baselineskip}
 \end{figure}
@@
 \section{Comparison with the original DoSE statistics}
-As the last experiment, we study how our proposed method with our model agnostic statistic performs against DoSE using the original statistics proposed in \cite{morningstar_density_2021}.
+As the last experiment, we study how our proposed method with our model agnostic statistic performs against DoSE using the original statistics proposed in \parencite{morningstar_density_2021}.
 For the VAEs model, they suggested to use the following 5 statistics: the posterior/prior cross-entropy $\text{H}[q_{\phi}( \textbf{z}\mid \textbf{x}), p(\textbf{z})]$, the posterior entropy $\text{H}[q_{\phi}(\textbf{z}\mid \textbf{x})]$, the posterior/prior KL-divergence $\text{D}_{\text{KL}}[q_{\phi}(\textbf{z}\mid \textbf{x}) \mid \mid p(\textbf{z})]$, the posterior expected log-likelihood $\mathbb{E}_{q_{\phi}(\textbf{z}\mid \textbf{x})}[\log q_{\phi}(\textbf{z}\mid \mathbf{x})]$, and the log-likelihood $\log \mathbb{E}_{q_{\phi}(\textbf{z}\mid \textbf{x})} \left[ \frac{p_{\theta}(\textbf{x}, \textbf{z})}{q_{\phi}(\textbf{z}\mid \textbf{x})}\right]$.
@@
 Indeed, we cannot use those statistics also for a PixelCNN++ for example or any other DGM.
-We want also to highlight that the models used in \cite{morningstar_density_2021} are a bit different from the ones used in this work.
+We want also to highlight that the models used in \parencite{morningstar_density_2021} are a bit different from the ones used in this work.
 For example, they are considering a beta-VAE with only one stochastic layer, while in our case we used a HVAE with 3-stochastic layers.
 
-\begin{table}%[15]
+\begin{table}[tb]%[15]
     \vspace{-1.3em}
@@
     \label{table:original_dose}
-    \vspace*{-\baselineskip}
+    % \vspace*{-\baselineskip}
 \end{table}
@@
 \textbf{Compute} $\hat{F}^{\text{typicality}}$ and $\hat{F}^{\text{score}}$, the two empirical CDFs, from $T^{\text{typicality}}$ and $T^{\text{score}}$.
-For example, we used \texttt{statsmodels} library \citep{seabold_statsmodels_2010}.
+For example, we used \texttt{statsmodels} library \parencite{seabold_statsmodels_2010}.
 \\

### 2023-09-29 appendices/supplementary_retrospective.tex
Work on background and main contributions
Updates to citation commands and footnotes

@@
     \centering
-    \caption{Overview of hyperparameters used for training the text classification models.}
+    \caption{Overview of hyperparameters for training text classification models.}
     \label{tab_retrospective:tableA1-hyperparameters}
@@
     \end{tabular}%
+    \vspace{-2mm}
     }
@@
 
-\subsubsection{Bag-of-words selection}
+\paragraph{Bag-of-words selection}
 
@@
 We used standard methods for significance testing and computing the confidence intervals.
-We used approximate methods owing to data size and to maintain computational feasibility \cite{dwass_modified_1957,eden_validity_1933}
+We used approximate methods owing to data size and to maintain computational feasibility \parencite{dwass_modified_1957,eden_validity_1933}
 

### 2023-09-29 chapters/background.tex
Updates to citation commands and footnotes

@@
 
-\section{Uncertainty and information}
+\section{Uncertainty and information} \label{sec:uncertainty-information-theory}
 
 In society, the concept of uncertainty goes by many names, and its meaning can vary depending on the specific context.
-However, across most quantitative scientific fields, a consensus definition appears to align with the following \cite{hubbard_how_2014}:
+However, across most quantitative scientific fields, a consensus definition appears to align with the following \parencite{hubbard_how_2014}:
 %
@@
 To our luck, information theory provides exactly that; a mathematically rigorous method for quantifying information through the language of probability theory.
-In the following, we shall see that, in this context, uncertainty can be understood as the information entropy of a random variable \cite{mackay_information_2003}.
+In the following, we shall see that, in this context, uncertainty can be understood as the information entropy of a random variable \parencite{mackay_information_2003}.
 
@@
 
-First characterized by Claude Shannon in 1948 \cite{shannon_mathematical_1948}, information entropy, $H(X)$, is a measure of the average amount of information contained in a discrete random variable $X$.
+First characterized by Claude Shannon in 1948 \parencite{shannon_mathematical_1948}, information entropy, $H(X)$, is a measure of the average amount of information contained in a discrete random variable $X$.
 Shannon's definition follows from three fundamental axioms of information theory.
@@
 %
-which is called the \emph{differential entropy} \cite{shannon_mathematical_1948}. Here, $\mathcal{X}$ is the support of $X$.
+which is called the \emph{differential entropy} \parencite{shannon_mathematical_1948}. Here, $\mathcal{X}$ is the support of $X$.
 
@@
     which is clearly not dimensionless and therefore cannot correspond directly to the discrete information entropy.
-    % As a final twist on this story, note that the SI system does actually have a unit for an amount of substance: the mole \cite{internationalbureauofweightsandmeasures_systeme_2019}. 
+    % As a final twist on this story, note that the SI system does actually have a unit for an amount of substance: the mole \parencite{internationalbureauofweightsandmeasures_systeme_2019}. 
     % The mole is defined as the amount of substance that contains as many elementary entities as there are atoms in $\SI{12}{g}$ of carbon-12, that is $\SI{1}{mol} = 6.02214076\times10^{23}$.
-    % Taking the position that a mole is quite simply a number (not a number of particles) \cite{debievre_atomic_1992,baranski_atomic_2012}, and a bit is a number of elementary (information) entities, we could define a bit as a derived unit from the mole as $\SI{1}{bit} = \SI{1}{mol} / 6.02214076\times10^{23}$.
+    % Taking the position that a mole is quite simply a number (not a number of particles) \parencite{debievre_atomic_1992,baranski_atomic_2012}, and a bit is a number of elementary (information) entities, we could define a bit as a derived unit from the mole as $\SI{1}{bit} = \SI{1}{mol} / 6.02214076\times10^{23}$.
 }. 
@@
 
-The form on the right-hand side of \cref{eq:corrected-differential-entropy-jaynes} can be recognized as the negative Kullback-Leibler divergence of $p(x)$ to $q(x)$ \cite{kullback_information_1959},
+The form on the right-hand side of \cref{eq:corrected-differential-entropy-jaynes} can be recognized as the negative Kullback-Leibler divergence of $p(x)$ to $q(x)$ \parencite{kullback_information_1959},
 %
@@
 Given a test point $\xb$, the model outputs a score $s(\xb)$ that indicates whether $\xb$ is in-distribution or out-of-distribution.
-%   - In the context of neural networks, this problem dates back to at least 1994 \cite{bishop_novelty_1994}. 
+%   - In the context of neural networks, this problem dates back to at least 1994 \parencite{bishop_novelty_1994}. 
 %
@@
 %
-OOD detection is the task of identifying test data that is unlikely to originate from the distribution of the training data and, in the context of neural networks, dates back several decades \cite{bishop_novelty_1994, chang_figure_1993}.
+OOD detection is the task of identifying test data that is unlikely to originate from the distribution of the training data and, in the context of neural networks, dates back several decades \parencite{bishop_novelty_1994, chang_figure_1993}.
 In the general case, we assume that we have a domain of in-distribution data, $\mathcal{D}_{\text{in}}$, and we would like to build a model that can be used to assess whether a test data point $\xb$ originates from that domain or not.
@@
 
-% In the literature, the task is also referred to as open set recognition \cite{bendale_open_2016, cardoso_weightless_2017, panaredabusto_open_2017}, outlier detection \cite{xu_deep_2014, mor_confidence_2018}, or selective prediction \cite{geifman_selective_2017} .
-OOD detection bears many similarities with anomaly detection, novelty detection, open set recognition, and outlier detection \cite{yang_generalized_2022}. 
+% In the literature, the task is also referred to as open set recognition \parencite{bendale_open_2016, cardoso_weightless_2017, panaredabusto_open_2017}, outlier detection \parencite{xu_deep_2014, mor_confidence_2018}, or selective prediction \parencite{geifman_selective_2017} .
+OOD detection bears many similarities with anomaly detection, novelty detection, open set recognition, and outlier detection \parencite{yang_generalized_2022}. 
 Some of these differences are subtle, and the terminology is not always used consistently in the literature.
@@
 OOD detection benchmarks almost always take OOD data to be from external datasets different from the training dataset which distinguishes it from open set recognition that usually uses a single dataset split into ID and OOD data.
-% Another closely related field is that of network-based uncertainty estimation \cite{neal_bayesian_1995, hernandez-lobato_probabilistic_2015, gawlikowski_survey_2023}. %[14, 10, 36, 20]. 
+% Another closely related field is that of network-based uncertainty estimation \parencite{neal_bayesian_1995, hernandez-lobato_probabilistic_2015, gawlikowski_survey_2023}. %[14, 10, 36, 20]. 
 Although we share the sentiment of \textcite{yang_generalized_2022}, who propose to unify the tasks as ``generalized out-of-distribution detection'', this thesis will follow the nomenclature used in the works most related to ours and refer to the task as out-of-distribution detection.
@@
 One way to categorize the different approaches is based on whether the underlying model is a classifier that estimates a conditional probability distribution $p(y|\xb)$ over some target variable $y$, or a model that learns a probability density $p(\xb)$ over the input itself.
-We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \cite{graham_denoising_2023,liu_unsupervised_2023a}.
+We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \parencite{graham_denoising_2023,liu_unsupervised_2023a}.
 It is important to note that this distinction relates only to the availability of some target value, $y$ - not whether OOD data is available for supervision\footnote{
@@
 This is the same distinction made by \textcite{graham_denoising_2023} and \textcite{liu_unsupervised_2023a}.
-    However, it is important to note that in other works, the distinction is made based on whether the model is trained on OOD data or not \cite{hendrycks_baseline_2017,liu_energy-based_2020}.
+    However, it is important to note that in other works, the distinction is made based on whether the model is trained on OOD data or not \parencite{hendrycks_baseline_2017,liu_energy-based_2020}.
 This difference in nomenclature is currently unresolved in the literature.
@@
 Unsupervised methods often derive the score 
-from the \emph{likelihood} assigned to the input \cite{bishop_novelty_1994,choi_waic_2019,kirichenko_why_2020,ren_likelihood_2019,serra_input_2020,xiao_likelihood_2020,morningstar_density_2021,nalisnick_detecting_2019,bergamin_modelagnostic_2022,maaloe_biva_2019,havtorn_hierarchical_2021},
-from a \emph{reconstruction} of the input \cite{sakurada_anomaly_2014,xia_learning_2015,lyudchik_outlier_2016,zhou_anomaly_2017,chen_outlier_2017,schlegl_unsupervised_2017,zong_deep_2018,li_madgan_2019,graham_denoising_2023,liu_unsupervised_2023a}, or
-from a hidden \emph{representation} of the input \cite{denouden_improving_2018,hendrycks_using_2019,ahmadian_likelihoodfree_2021,bergman_classificationbased_2020,tack_csi_2020,sehwag_ssd_2021,xiao_we_2021}.
+from the \emph{likelihood} assigned to the input \parencite{bishop_novelty_1994,choi_waic_2019,kirichenko_why_2020,ren_likelihood_2019,serra_input_2020,xiao_likelihood_2020,morningstar_density_2021,nalisnick_detecting_2019,bergamin_modelagnostic_2022,maaloe_biva_2019,havtorn_hierarchical_2021},
+from a \emph{reconstruction} of the input \parencite{sakurada_anomaly_2014,xia_learning_2015,lyudchik_outlier_2016,zhou_anomaly_2017,chen_outlier_2017,schlegl_unsupervised_2017,zong_deep_2018,li_madgan_2019,graham_denoising_2023,liu_unsupervised_2023a}, or
+from a hidden \emph{representation} of the input \parencite{denouden_improving_2018,hendrycks_using_2019,ahmadian_likelihoodfree_2021,bergman_classificationbased_2020,tack_csi_2020,sehwag_ssd_2021,xiao_we_2021}.
 % https://arxiv.org/pdf/2302.10326.pdf
 Supervised methods often derive the score 
-from the \emph{probabilities} given by the predictive distribution $p(y|\xb)$ \cite{hendrycks_baseline_2017,hendrycks_scaling_2022}, 
-from the \emph{logits} of $p(y|\xb)$ \cite{hendrycks_scaling_2022,liu_energy-based_2020}, or 
-use a latent \emph{representation} of the input \cite{lee_simple_2018,li_anomaly_2019,ndiour_outofdistribution_2020,cook_outlier_2020,zaeemzadeh_outofdistribution_2021}, similar to unsupervised methods.
+from the \emph{probabilities} given by the predictive distribution $p(y|\xb)$ \parencite{hendrycks_baseline_2017,hendrycks_scaling_2022}, 
+from the \emph{logits} of $p(y|\xb)$ \parencite{hendrycks_scaling_2022,liu_energy-based_2020}, or 
+use a latent \emph{representation} of the input \parencite{lee_simple_2018,li_anomaly_2019,ndiour_outofdistribution_2020,cook_outlier_2020,zaeemzadeh_outofdistribution_2021}, similar to unsupervised methods.
 In the following sections, we provide a more in-depth overview of the different approaches to OOD detection following the above categorization into supervised and unsupervised methods.
@@
 %   - One limitation of model-dependent OoD techniques is that they may discard information about p(x) in learning the task-specific model p(y|x). That information may be useful for OoD detection.
-%   - Likelihood Ratios for Out-of-Distribution Detection \cite{ren_likelihood_2019} provides a number of baselines for supervised OODD:
-%       - 1 Maximum class probability of the predictive distribution p(y|x) \cite{hendrycks_baseline_2017}
-%       - 2 Entropy of predictive distribution p(y|x) \cite{ren_likelihood_2019}
-%       - 3 ODIN \cite{liang_enhancing_2018} (Uses synthetic OOD data by adding noise to in-distribution)
-%       - 4 Mahalanobis distance to class conditional Gaussian distribution \cite{lee_simple_2018}
-%       - 5 DeepEnsemble \cite{lakshminarayanan_simple_2017} The classifier-based ensemble method that uses the average of the predictions from multiple independently trained models with random initialization of network parameters and random shuffling of training inputs
+%   - Likelihood Ratios for Out-of-Distribution Detection \parencite{ren_likelihood_2019} provides a number of baselines for supervised OODD:
+%       - 1 Maximum class probability of the predictive distribution p(y|x) \parencite{hendrycks_baseline_2017}
+%       - 2 Entropy of predictive distribution p(y|x) \parencite{ren_likelihood_2019}
+%       - 3 ODIN \parencite{liang_enhancing_2018} (Uses synthetic OOD data by adding noise to in-distribution)
+%       - 4 Mahalanobis distance to class conditional Gaussian distribution \parencite{lee_simple_2018}
+%       - 5 DeepEnsemble \parencite{lakshminarayanan_simple_2017} The classifier-based ensemble method that uses the average of the predictions from multiple independently trained models with random initialization of network parameters and random shuffling of training inputs
 %       - 6 The log-odds of a binary classifier trained to distinguish between in-distribution inputs from all classes as one class and randomly perturbed in-distribution inputs as the other.
-\cite{ren_likelihood_2019}
-%       - 7 The maximum class probability over K in-distribution classes of a (K + 1)-class classifier where the additional class is perturbed in-distribution. \cite{ren_likelihood_2019}
+\parencite{ren_likelihood_2019}
+%       - 7 The maximum class probability over K in-distribution classes of a (K + 1)-class classifier where the additional class is perturbed in-distribution. \parencite{ren_likelihood_2019}
 %       - 8 The maximum class probability of a K-class classifier for in-distribution classes but the predicted class distribution is explicitly trained to output uniform distribution on perturbed in-distribution inputs.
-\cite{ren_likelihood_2019}.
-This is similar to using simulated OOD inputs from generative adversarial networks \cite{lee_training_2018} or using auxiliary datasets of outliers \cite{hendrycks_deep_2019} for calibration purpose.
-%   - X Deep semi-supervised anomaly detection \cite{ruff_deep_2020}.
+\parencite{ren_likelihood_2019}.
+This is similar to using simulated OOD inputs from generative adversarial networks \parencite{lee_training_2018} or using auxiliary datasets of outliers \parencite{hendrycks_deep_2019} for calibration purpose.
+%   - X Deep semi-supervised anomaly detection \parencite{ruff_deep_2020}.
 %   - Variational information bottleneck (VIB) (Alemi et al., 2018b) performs divergence estimation in latent space to detect OoD, but is technically a model-dependent technique because the latent code is trained jointly with the downstream classification task.
-\cite{alemi_deep_2017}
+\parencite{alemi_deep_2017}
 %
@@
 \textcite{ruff_deep_2020} use semi-supervised learning and learn representations of in-distribution data that concentrate close to a centroid in latent space while labelled outliers are pushed away from the centroid.
-Other methods including MCD \cite{yu_unsupervised_2019}, NGC \cite{wu_ngc_2021}, and UDG \cite{yang_semantically_2021} utilize external, unlabeled, noisy data to improve OOD detection performance without requiring cleanly labelled OOD examples.
+Other methods including MCD \parencite{yu_unsupervised_2019}, NGC \parencite{wu_ngc_2021}, and UDG \parencite{yang_semantically_2021} utilize external, unlabeled, noisy data to improve OOD detection performance without requiring cleanly labelled OOD examples.
 As we discussed earlier, for many modern applications of machine learning, input data is often high dimensional and complex which makes it difficult to obtain enough representative OOD data to ensure robust OOD detection capabilities.
@@
 A number of methods do not require access to actual OOD data but synthesize it instead. 
-Several methods do so by adding noise to in-distribution data \cite{liang_enhancing_2018, lee_simple_2018, ren_likelihood_2019}. 
+Several methods do so by adding noise to in-distribution data \parencite{liang_enhancing_2018, lee_simple_2018, ren_likelihood_2019}. 
 For instance, \textcite{ren_likelihood_2019} propose a number of baselines including training a binary classifier to distinguish between original and perturbed in-distribution data.
@@
 While these methods are appealing, their weaknesses have been pointed out by later work which generally improve on their perforamnce.
-An example is ODIN \cite{liang_enhancing_2018}, in which the authors propose to calibrate $p(y|\xb)$ with temperature scaling \cite{guo_calibration_2017} and add gradient-based, input-dependent perturbations to the inputs and use the calibrated maximum class probability as the OOD score.
+An example is ODIN \parencite{liang_enhancing_2018}, in which the authors propose to calibrate $p(y|\xb)$ with temperature scaling \parencite{guo_calibration_2017} and add gradient-based, input-dependent perturbations to the inputs and use the calibrated maximum class probability as the OOD score.
 \textcite{vyas_outofdistribution_2018} train an ensemble of classifiers on different subsets of the training data, with the left out data taken as OOD, and propose novel loss over $p(y|\xb)$ that seeks to maintain a predefined margin between its average entropy for the OOD and in-distribution examples.
-Another approach generates OOD inputs using a generative adversarial network \cite{lee_training_2018}. 
+Another approach generates OOD inputs using a generative adversarial network \parencite{lee_training_2018}. 
 Similar to actual OOD data, the usefulness of synthetic OOD data is also fundamentally limited by the intractability of sampling the complete space of OOD data.
@@
 The least informed supervised OOD detection methods do not require instantiations of any kind of OOD data, real or synthetic. 
-A baseline approach uses the maximum class probability of $p(y|\xb)$ directly by noting that it tends to be larger for correctly classified examples \cite{hendrycks_baseline_2017}.
-Another baseline method proposes that a high entropy of $p(y|\xb)$ indicates an OOD input \cite{ren_likelihood_2019}
+A baseline approach uses the maximum class probability of $p(y|\xb)$ directly by noting that it tends to be larger for correctly classified examples \parencite{hendrycks_baseline_2017}.
+Another baseline method proposes that a high entropy of $p(y|\xb)$ indicates an OOD input \parencite{ren_likelihood_2019}
 Other methods that derive the score from the classifier probabilities include \textcite{lakshminarayanan_simple_2017} who propose to use an ensemble of independently trained classifiers to discriminate between in-distribution and OOD data by evaluating the agreement between the classifiers, \textcite{devries_learning_2018} who augment the network with a confidence estimation branch that learns to estimate the confidence of the classifier separately from the probability, and \textcite{huang_importance_2021} who compute the gradient of the KL-divergence of the predictive distribution $p(y|\xb)$ to a uniform distribution noting that the magnitude of gradients is higher for in-distribution data than for OOD data.
-The Variational Information Bottleneck \cite{alemi_deep_2017} jointly learns a probabilistic latent representation, $\zb$ and $p(y|\xb)$, using a generalized variational autoencoder \cite{kingma_autoencoding_2014} that tries to maximize the mutual information between $\zb$ and $y$.
+The Variational Information Bottleneck \parencite{alemi_deep_2017} jointly learns a probabilistic latent representation, $\zb$ and $p(y|\xb)$, using a generalized variational autoencoder \parencite{kingma_autoencoding_2014} that tries to maximize the mutual information between $\zb$ and $y$.
 
@@
 % No OOD data, logits
-Although some works use the maximum softmax probability as a score for OOD detection \cite{hendrycks_deep_2019,ren_likelihood_2019}, several works have noted that it is not generally reliable \cite{hendrycks_scaling_2022,liu_energy-based_2020}.
-\textcite{liu_energy-based_2020} make an interesting argument as to why based on the energy $E(\xb; f)$ of a softmax classifier $f(\xb)$ \cite{lecun_tutorial_2006},
+Although some works use the maximum softmax probability as a score for OOD detection \parencite{hendrycks_deep_2019,ren_likelihood_2019}, several works have noted that it is not generally reliable \parencite{hendrycks_scaling_2022,liu_energy-based_2020}.
+\textcite{liu_energy-based_2020} make an interesting argument as to why based on the energy $E(\xb; f)$ of a softmax classifier $f(\xb)$ \parencite{lecun_tutorial_2006},
 %
@@
 They conclude that this shift results in the maximum softmax probability being a biased score for OOD detection and propose to instead use the energy directly.
-This energy-score was improved in ReAct by feature clipping \cite{sun_react_2021}.
+This energy-score was improved in ReAct by feature clipping \parencite{sun_react_2021}.
 
@@
 \textcite{huang_mos_2021} groups the classes of the target variable $y$ and defines an OOD class for each group.
-Each training example is then the correct target for one group and an OOD example for all other groups; a kind of hierarchical version of the OOD class of simpler baselines based on noise augmentation \cite{ren_likelihood_2019}.
+Each training example is then the correct target for one group and an OOD example for all other groups; a kind of hierarchical version of the OOD class of simpler baselines based on noise augmentation \parencite{ren_likelihood_2019}.
 Similarly, to represent a virtual OOD class, \textcite{wang_vim_2022} generates an additional logit by first computing the residual of the input's latent space representation against a principal feature space and then converting it to a valid logit by matching its mean over training samples to the average maximum logits.
@@
 \textcite{zaeemzadeh_outofdistribution_2021} forces the ID samples to embed into a union of 1-dimensional subspaces during training and computes the minimum angular distance from the feature to the class-wise subspaces.
-NuSA \cite{cook_outlier_2020} uses projects features onto the column space of the classification weight matrix and computes the ratio of the norm the projected and original features.
+NuSA \parencite{cook_outlier_2020} uses projects features onto the column space of the classification weight matrix and computes the ratio of the norm the projected and original features.
 \textcite{lee_simple_2018} fit a multivariate Gaussian distribution to the activations of the penultimate layer of a pre-trained classifier and use the Mahalanobis distance to this distribution to evaluate whether inputs are OOD.
 This method can also be seen as ameliorating the curse of dimensionality by clustering the high-dimensional feature space.
-Finally, Bayesian neural networks have also been proposed for OOD detection although their performance is not yet competitive with other methods \cite{henning_are_2021,dangelo_outofdistribution_2022,nguyen_out_2022}.
+Finally, Bayesian neural networks have also been proposed for OOD detection although their performance is not yet competitive with other methods \parencite{henning_are_2021,dangelo_outofdistribution_2022,nguyen_out_2022}.
 
@@
 
-%       - 2 Entropy of predictive distribution p(y|x) \cite{ren_likelihood_2019}
+%       - 2 Entropy of predictive distribution p(y|x) \parencite{ren_likelihood_2019}
 %       - 6 The log-odds of a binary classifier trained to distinguish between in-distribution inputs from all classes as one class and randomly perturbed in-distribution inputs as the other.
-\cite{ren_likelihood_2019}
-%       - 7 The maximum class probability over K in-distribution classes of a (K + 1)-class classifier where the additional class is perturbed in-distribution. \cite{ren_likelihood_2019}
+\parencite{ren_likelihood_2019}
+%       - 7 The maximum class probability over K in-distribution classes of a (K + 1)-class classifier where the additional class is perturbed in-distribution. \parencite{ren_likelihood_2019}
 %       - 8 The maximum class probability of a K-class classifier for in-distribution classes but the predicted class distribution is explicitly trained to output uniform distribution on perturbed in-distribution inputs.
-\cite{ren_likelihood_2019}.
-This is similar to using simulated OOD inputs from generative adversarial networks \cite{lee_training_2018} or using auxiliary datasets of outliers \cite{hendrycks_deep_2019} for calibration purpose.
+\parencite{ren_likelihood_2019}.
+This is similar to using simulated OOD inputs from generative adversarial networks \parencite{lee_training_2018} or using auxiliary datasets of outliers \parencite{hendrycks_deep_2019} for calibration purpose.
 
-%       - 3 ODIN \cite{liang_enhancing_2018} (Uses synthetic OOD data by adding noise to in-distribution)
-%       - 4 Mahalanobis distance to class conditional Gaussian distribution \cite{lee_training_2018}
-%       - 5 DeepEnsemble \cite{lakshminarayanan_simple_2017} The classifier-based ensemble method that uses the average of the predictions from multiple independently trained models with random initialization of network parameters and random shuffling of training inputs
+%       - 3 ODIN \parencite{liang_enhancing_2018} (Uses synthetic OOD data by adding noise to in-distribution)
+%       - 4 Mahalanobis distance to class conditional Gaussian distribution \parencite{lee_training_2018}
+%       - 5 DeepEnsemble \parencite{lakshminarayanan_simple_2017} The classifier-based ensemble method that uses the average of the predictions from multiple independently trained models with random initialization of network parameters and random shuffling of training inputs
 
 %   - Variational information bottleneck (VIB) (Alemi et al., 2018b) performs divergence estimation in latent space to detect OoD, but is technically a model-dependent technique because the latent code is trained jointly with the downstream classification task.
-\cite{alemi_deep_2017}
+\parencite{alemi_deep_2017}
 
@@
 %   - Generative
-%       - 9 Compute WAIC (\cite{watanabe_algebraic_2009, watanabe_asymptotic_2010}) using ensemble of generative models \cite{choi_waic_2019}
-%       - Likelihood Ratios for Out-of-Distribution Detection \cite{ren_likelihood_2019}
-%       - Input complexity and out-of-distribution detection with likelihood-based generative models \cite{serra_input_2020}
-%       - Likelihood regret for variational autoencoders \cite{xiao_likelihood_2020}
-%       - Typicality test for generative OODD (Detecting out-of-distribution inputs to deep generative models using typicality) \cite{nalisnick_detecting_2019}
-%       - Density of states estimation for out of distribution detection \cite{morningstar_density_2021}
+%       - 9 Compute WAIC (\parencite{watanabe_algebraic_2009, watanabe_asymptotic_2010}) using ensemble of generative models \parencite{choi_waic_2019}
+%       - Likelihood Ratios for Out-of-Distribution Detection \parencite{ren_likelihood_2019}
+%       - Input complexity and out-of-distribution detection with likelihood-based generative models \parencite{serra_input_2020}
+%       - Likelihood regret for variational autoencoders \parencite{xiao_likelihood_2020}
+%       - Typicality test for generative OODD (Detecting out-of-distribution inputs to deep generative models using typicality) \parencite{nalisnick_detecting_2019}
+%       - Density of states estimation for out of distribution detection \parencite{morningstar_density_2021}
 %   - Reconstructive
-%       - A review of novelty detection \cite{pimentel_review_2014}
-%       - Outlier detection using autoencoders \cite{lyudchik_outlier_2016} and ensembles of autoencoders \cite{chen_outlier_2017}
-%       - Improving reconstruction autoencoder out-of-distribution detection with mahalanobis distance \cite{denouden_improving_2018}
-%       - Denosing diffusion models for out-of-distribution detection \cite{graham_denoising_2023}
-%       - Diffusion inpainting \cite{liu_unsupervised_2023a}
+%       - A review of novelty detection \parencite{pimentel_review_2014}
+%       - Outlier detection using autoencoders \parencite{lyudchik_outlier_2016} and ensembles of autoencoders \parencite{chen_outlier_2017}
+%       - Improving reconstruction autoencoder out-of-distribution detection with mahalanobis distance \parencite{denouden_improving_2018}
+%       - Denosing diffusion models for out-of-distribution detection \parencite{graham_denoising_2023}
+%       - Diffusion inpainting \parencite{liu_unsupervised_2023a}
 
@@
 
-For high-dimensional data, the most successful unsupervised models for OOD detection are deep autoencoders \cite{hinton_reducing_2006}, self-supervised methods \cite{mikolov_efficient_2013,devlin_bert_2018,schneider_wav2vec_2019,chen_simple_2020}, and deep generative models \cite{kingma_autoencoding_2014,rezende_stochastic_2014,dinh_nice_2015,rezende_variational_2015,ho_denoising_2020,hinton_fast_2006,oord_conditional_2016,goodfellow_generative_2014}.
-Other important density estimation methods that have been applied to OOD detection include kernel density estimation \cite{parzen_estimation_1962}, nearest neighbor methods \cite{cover_nearest_1967}, support vector machines \cite{cortes_supportvector_1995,scholkopf_estimating_2001}, and Gaussian mixture models \cite{dempster_maximum_1977}.
+For high-dimensional data, the most successful unsupervised models for OOD detection are deep autoencoders \parencite{hinton_reducing_2006}, self-supervised methods \parencite{mikolov_efficient_2013,devlin_bert_2018,schneider_wav2vec_2019,chen_simple_2020}, and deep generative models \parencite{kingma_autoencoding_2014,rezende_stochastic_2014,dinh_nice_2015,rezende_variational_2015,ho_denoising_2020,hinton_fast_2006,oord_conditional_2016,goodfellow_generative_2014}.
+Other important density estimation methods that have been applied to OOD detection include kernel density estimation \parencite{parzen_estimation_1962}, nearest neighbor methods \parencite{cover_nearest_1967}, support vector machines \parencite{cortes_supportvector_1995,scholkopf_estimating_2001}, and Gaussian mixture models \parencite{dempster_maximum_1977}.
 However, these methods are not well suited for high-dimensional data such as images, text or audio, and have had little direct impact on the recent work on OOD detection.
@@
 \paragraph{Likelihood-based}
-% \cite{
+% \parencite{
 %     X bishop_novelty_1994,
@@
 Simply put, since the likelihood measures ``how probable the data is'', OOD data is expected to give lower likelihoods than in-distribution data.
-Although, this method originally gave encouraging results, the advent of deep generative models and their application to high-dimensional data has lead many to observe likelihoods for OOD data that are higher than for in-distribution data \cite{choi_waic_2019,nalisnick_detecting_2019,kirichenko_why_2020}.
+Although, this method originally gave encouraging results, the advent of deep generative models and their application to high-dimensional data has lead many to observe likelihoods for OOD data that are higher than for in-distribution data \parencite{choi_waic_2019,nalisnick_detecting_2019,kirichenko_why_2020}.
 Such results have sparked interest in trying to explain this phenomenon and many works have proposed new scores for OOD detection derived from the likelihood.
@@
 Therefore, they propose to use a general lossless image compression algorithm as a background model.
-\textcite{choi_waic_2019} propose to use the Watanabe information criterion (WAIC) computed from the likelihood \cite{watanabe_algebraic_2009, watanabe_asymptotic_2010}. 
-For variational autoencoders, other work proposes to refit the encoder on a test data example, hypothesizing that the likelihood of OOD data will improve more than for in-distribution data, and use this ``likelihood regret" as the OOD score \cite{xiao_likelihood_2020}.
+\textcite{choi_waic_2019} propose to use the Watanabe information criterion (WAIC) computed from the likelihood \parencite{watanabe_algebraic_2009, watanabe_asymptotic_2010}. 
+For variational autoencoders, other work proposes to refit the encoder on a test data example, hypothesizing that the likelihood of OOD data will improve more than for in-distribution data, and use this ``likelihood regret" as the OOD score \parencite{xiao_likelihood_2020}.
 \textcite{maaloe_biva_2019} provide initial results that a loosened variational bound on the likelihood, using only encoded representations from the top-most latent variables in a hierarchical variationa autoencoder, can improve OOD detection performance.
@@
 
-A different approach is to use the typicality set hypothesis \cite{nalisnick_detecting_2019}.
+A different approach is to use the typicality set hypothesis \parencite{nalisnick_detecting_2019}.
 The typicality set is the subset of the model full support in data space, where the model samples from, that does not overlap with regions of maximal likelihood.
 \textcite{nalisnick_detecting_2019} propose to use the typicality set as a test statistic for OOD detection while \textcite{morningstar_density_2021} propose to use the related idea of density of states of the model.
-In \textcite{bergamin_modelagnostic_2022} we use Fisher's method \cite{fisher_statistical_1925} to combine Rao's score test statistic \cite{rao_large_1948} with the typicality set test statistic hence including information from both the gradient and the likelihood.
+In \textcite{bergamin_modelagnostic_2022} we use Fisher's method \parencite{fisher_statistical_1925} to combine Rao's score test statistic \parencite{rao_large_1948} with the typicality set test statistic hence including information from both the gradient and the likelihood.
 
 \paragraph{Reconstruction-based}
-% \cite{
+% \parencite{
 %     X sakurada_anomaly_2014                 encode and decode data with autoencoders. Dimensionality reduction
@@
 %     X zong_deep_2018                        encode and decode data with autoencoders.
-jointly learn a deep autoencoder and a Gaussian Mixture Model on the learned hidden representations and link their method to neural variational inference \cite{mnih_neural_2014}.
+jointly learn a deep autoencoder and a Gaussian Mixture Model on the learned hidden representations and link their method to neural variational inference \parencite{mnih_neural_2014}.
 %     X chen_outlier_2017                     encode and decode data with an ensemble of autoencoders.
@@
 Drawing inspiration from Robust Principal Component Analysis, \textcite{zhou_anomaly_2017} propose to first split the input data into a dense low-rank factor and a sparse factor, assuming that outliers are caught in the sparse factor, and then use a deep autoencoder to reconstruct the dense factor.
-\textcite{zong_deep_2018} jointly learn a deep autoencoder and a Gaussian Mixture Model on the learned hidden representations and draw parallels of their method to neural variational inference \cite{mnih_neural_2014}.
+\textcite{zong_deep_2018} jointly learn a deep autoencoder and a Gaussian Mixture Model on the learned hidden representations and draw parallels of their method to neural variational inference \parencite{mnih_neural_2014}.
 Similarly to the ensemble-based methods for supervised OOD detection, \textcite{chen_outlier_2017} propose to use an ensemble of autoencoders to reconstruct the input and use the median reconstruction error as the OOD score.
 
-Although generative adversarial networks do not have the ability to encode a given data point, methods have been proposed to invert the generator to find a latent representation that can be used to reconstruct the input and use the reconstruction error as well as a discriminator score for OOD detection \cite{schlegl_unsupervised_2017,li_madgan_2019}.
+Although generative adversarial networks do not have the ability to encode a given data point, methods have been proposed to invert the generator to find a latent representation that can be used to reconstruct the input and use the reconstruction error as well as a discriminator score for OOD detection \parencite{schlegl_unsupervised_2017,li_madgan_2019}.
 
@@
 % https://arxiv.org/pdf/2302.10326.pdf
-% \cite{
+% \parencite{
 %     X denouden_improving_2018,              leverages the latent variables of an autoencoder, and evaluates the Mahalanobis distance in the latent space along with the data reconstruction error
@@
 
-Several other works also use self-supervised representations of the in-domain data for OOD detection \cite{hendrycks_using_2019,bergman_classificationbased_2020}. 
+Several other works also use self-supervised representations of the in-domain data for OOD detection \parencite{hendrycks_using_2019,bergman_classificationbased_2020}. 
 \textcite{tack_csi_2020} propose to use a contrastive objective to learn representations of the in-domain data contrasted with data augmented in-samples and use a softmax classifier trained on the representations to compute the OOD score.
@@
 VAEs belong to the class of deep generative models which are models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution. 
-There exist at least six main classes of deep generative models: variational autoencoders \cite{kingma_autoencoding_2014,rezende_stochastic_2014,ranganath_hierarchical_2016,vahdat_nvae_2020,child_very_2021}, normalizing flow models \cite{dinh_nice_2015,rezende_variational_2015,dinh_density_2017,kingma_glow_2018,grathwohl_ffjord_2018}, diffusion models \cite{sohl-dickstein_deep_2015,song_generative_2019, ho_denoising_2020, vahdat_scorebased_2021}, energy-based models \cite{lecun_tutorial_2006,hinton_fast_2006,salakhutdinov_efficient_2010,du_implicit_2019}, autoregressive models \cite{oord_conditional_2016,oord_wavenet_2016,radford_improving_2018}, and generative adversarial networks \cite{goodfellow_generative_2014,arjovsky_wasserstein_2017,brock_large_2019,karras_stylebased_2019}.
+There exist at least six main classes of deep generative models: variational autoencoders \parencite{kingma_autoencoding_2014,rezende_stochastic_2014,ranganath_hierarchical_2016,vahdat_nvae_2020,child_very_2021}, normalizing flow models \parencite{dinh_nice_2015,rezende_variational_2015,dinh_density_2017,kingma_glow_2018,grathwohl_ffjord_2018}, diffusion models \parencite{sohl-dickstein_deep_2015,song_generative_2019, ho_denoising_2020, vahdat_scorebased_2021}, energy-based models \parencite{lecun_tutorial_2006,hinton_fast_2006,salakhutdinov_efficient_2010,du_implicit_2019}, autoregressive models \parencite{oord_conditional_2016,oord_wavenet_2016,radford_improving_2018}, and generative adversarial networks \parencite{goodfellow_generative_2014,arjovsky_wasserstein_2017,brock_large_2019,karras_stylebased_2019}.
 % Besides data generation, common properties of deep generative models include the ability to compute an exact or approximate likelihood of input $p(\xb)$, reconstruct an input $\xb$ from its latent representation $\zb$, and represent an input $\xb$ in a latent space.
@@
 
-% A variational autoencoder (VAE) \cite{kingma_autoencoding_2014,rezende_stochastic_2014} is a type of deep generative model and an instance of a latent variable model. 
+% A variational autoencoder (VAE) \parencite{kingma_autoencoding_2014,rezende_stochastic_2014} is a type of deep generative model and an instance of a latent variable model. 
 In the following, we will introduce VAEs starting from the definition of the group of latent variable models to which they belong.
@@
 %
-In many latent variable models, the joint distribution is factorized as $p(\xb,\zb) = p(\xb|\zb)p(\zb)$ including Gaussian mixture models \cite{dempster_maximum_1977}, hidden Markov models \cite{rabiner_tutorial_1989}, and probabilistic principal component analysis \cite{tipping_probabilistic_1999}.
+In many latent variable models, the joint distribution is factorized as $p(\xb,\zb) = p(\xb|\zb)p(\zb)$ including Gaussian mixture models \parencite{dempster_maximum_1977}, hidden Markov models \parencite{rabiner_tutorial_1989}, and probabilistic principal component analysis \parencite{tipping_probabilistic_1999}.
 
@@
 This is the case when the latent variable space is high-dimensional or the conditional distribution $p(\xb|\zb)$ is complex, as with non-conjugate models or neural networks.
-In these cases, approximate methods such as variational inference \cite{jordan_introduction_1999} and Markov Chain Monte Carlo (MCMC) \cite{mohamed_monte_2019} are used to approximately compute the marginal and posterior distributions.
+In these cases, approximate methods such as variational inference \parencite{jordan_introduction_1999} and Markov Chain Monte Carlo (MCMC) \parencite{mohamed_monte_2019} are used to approximately compute the marginal and posterior distributions.
 
@@
 with autoregressive dependencies $p(\xb_{t}|\xb_{1:t},\zb)$.
-Some works have proposed to mitigate posterior collapse by tempering the KL-divergence term \cite{alemi_fixing_2018,higgins_vvae_2017} or by adding additional terms to the ELBO \cite{zhao_infovae_2018}.
+Some works have proposed to mitigate posterior collapse by tempering the KL-divergence term \parencite{alemi_fixing_2018,higgins_vvae_2017} or by adding additional terms to the ELBO \parencite{zhao_infovae_2018}.
 
@@
 \textcite{roeder_sticking_2017} note that the path-wise derivative can be factored into a path-wise estimator and a score function estimator and propose to ignore the score function factor to reduce gradient variance.
-\textcite{rainforth_tighter_2019} that this problem is even larger for the importance weighted ELBO \cite{burda_importance_2016} leading to diminishing returns for increasing numbers of samples.
+\textcite{rainforth_tighter_2019} that this problem is even larger for the importance weighted ELBO \parencite{burda_importance_2016} leading to diminishing returns for increasing numbers of samples.
 Instead of removing the score function factor in the gradient, \textcite{tucker_doubly_2019} propose to reparameterize the score function and derive low variance gradient estimators also for importance weighted ELBO.
@@
 Training by stochastic gradient descent training involves sampling mini-batches of data points $\xb$ from the empirical training data distribution $\hat{p}(\xb)$ and evaluating the gradient of the ELBO with respect to the parameters $\thetab$ and $\phib$ for each mini-batch.
-To gain some additional insight into the objective that we are optimizing, we will exploit this expectation over the data distribution $\hat{p}(\xb)$ to form a factorization similar to \cref{eq:variational-autoencoder-elbo-factorized} but with the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$ in the KL-divergence \cite{tomczak_trouble_2022}.
+To gain some additional insight into the objective that we are optimizing, we will exploit this expectation over the data distribution $\hat{p}(\xb)$ to form a factorization similar to \cref{eq:variational-autoencoder-elbo-factorized} but with the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$ in the KL-divergence \parencite{tomczak_trouble_2022}.
 The marginal likelihood under this expectation is lower bounded by the same expectation over the ELBO,
@@
 % \paragraph{Hierarchical models}
-% VAEs can be extended to hierarchical models by introducing additional latent variables $\zb_1,\dots,\zb_L$ that are conditionally dependent on each other \cite{ranganath_hierarchical_2016,sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021,maaloe_biva_2019}.
+% VAEs can be extended to hierarchical models by introducing additional latent variables $\zb_1,\dots,\zb_L$ that are conditionally dependent on each other \parencite{ranganath_hierarchical_2016,sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021,maaloe_biva_2019}.
 The generative model can then be written as,
@@
 \paragraph{Applications of VAEs}
-Despite the challenges facing the training of VAEs, they have been successfully applied to a number of tasks including image generation \cite{kingma_autoencoding_2014,rezende_stochastic_2014}, image inpainting \cite{pathak_context_2016}, image super-resolution \cite{sonderby_amortised_2017,chira_image_2022}, and speech synthesis \cite{hsu_unsupervised_2017,hsu_hierarchical_2019}.
-VAEs have also proven themselves useful for semi-supervised learning \cite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}.
-Recent architectural advances have made it possible to train VAEs with many latent variables \cite{sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021} and have alleviated the posterior collapse problem \cite{maaloe_biva_2019}.
+Despite the challenges facing the training of VAEs, they have been successfully applied to a number of tasks including image generation \parencite{kingma_autoencoding_2014,rezende_stochastic_2014}, image inpainting \parencite{pathak_context_2016}, image super-resolution \parencite{sonderby_amortised_2017,chira_image_2022}, and speech synthesis \parencite{hsu_unsupervised_2017,hsu_hierarchical_2019}.
+VAEs have also proven themselves useful for semi-supervised learning \parencite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}.
+Recent architectural advances have made it possible to train VAEs with many latent variables \parencite{sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021} and have alleviated the posterior collapse problem \parencite{maaloe_biva_2019}.
 
 \paragraph{Self-supervised learning}
-An alternative approach to representation learning is provided by self-supervised learning \cite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020,schneider_wav2vec_2019} which is a form of unsupervised learning where the training objective is derived from the data itself.
-While it is fair to say that not all self-supervised training objectives are as well-motivated as the ELBO is for VAEs, they have shown impressive results within the field of natural language processing \cite{devlin_bert_2018,chen_simple_2020}, speech processing \cite{schneider_wav2vec_2019}, and computer vision \cite{chen_simple_2020}.
+An alternative approach to representation learning is provided by self-supervised learning \parencite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020,schneider_wav2vec_2019} which is a form of unsupervised learning where the training objective is derived from the data itself.
+While it is fair to say that not all self-supervised training objectives are as well-motivated as the ELBO is for VAEs, they have shown impressive results within the field of natural language processing \parencite{devlin_bert_2018,chen_simple_2020}, speech processing \parencite{schneider_wav2vec_2019}, and computer vision \parencite{chen_simple_2020}.
 For a deeper comparison of VAEs and self-supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{app:paper-brief}. 

### 2023-09-29 chapters/conclusion.tex
Updates to citation commands and footnotes

@@
 We provided the motivational cases of recognition of stroke cases in emergency calls \cref{subsec:motivation-stroke-recognition} and automated medical coding \cref{subsec:motivation-medical-coding} and gave examples of potential sources of uncertainty and of how correctly quantifying uncertainty can improve the usefulness of such tools in practice.
-Finally, we ended the introduction by connecting machine learning reliability \cref{sec:machine-learning-reliability} and model calibration \cref{subsec:model-calibration} and defining different types of uncertainty \cref{subsec:types-of-uncertainty}.
+Finally, we ended the introduction by connecting machine learning reliability \cref{sec:machine-learning-reliability} and model calibration \cref{subsec:model-calibration} and defining different types of uncertainty \cref{subsec:understanding-uncertainty}.
 
 \Cref{chp:technical-background} provided technical background only covered briefly by the individual studies.
-We first introduced uncertainty as a concept in the context of information and probability theory \cref{sec:information-theory}.
-Then, we defined the task of out-of-distribution detection and reviewed existing work on the problem \cref{\cref{sec:out-of-distribution-detection}}.
+We first introduced uncertainty as a concept in the context of information and probability theory \cref{sec:uncertainty-information-theory}.
+Then, we defined the task of out-of-distribution detection and reviewed existing work on the problem \cref{sec:out-of-distribution-detection}.
 Finally, we provided technical background for variational autoencoders \cref{sec:variational-autoencoders}.

### 2023-09-29 chapters/discussion.tex
Updates to citation commands and footnotes

@@
 \lesstodo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other).
-Include reference to \cite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
+Include reference to \parencite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
 \lesstodo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimizing a mutual information term in the ELBO (derive this form).}
@@
 \lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?
-\cite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}
+\parencite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}
 \lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}
@@
     \end{subfigure}
-    \caption[Calibration curve for the uncalibrated stroke recognition model and empirical distribution of predicted probabilities.]{\subref{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated} \Cref{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated} shows the calibration curve for the uncalibrated ensemble model with median F1-score also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.
-\subref{fig_discussion:histogram_ensemble_and_single_model} Histogram of predicted probabilities.}
+    \caption[Calibration curve for the uncalibrated stroke recognition model and empirical distribution of predicted probabilities.]{Calibration curve for the uncalibrated ensemble model with median F1-score also reported in \cref{tab_retrospective:table3-occlusion-analysis, fig_retrospective:figure1-roc-curve} (left) and the histogram of predicted probabilities (right).}
 \end{figure}
@@
 \lesstodo[inline]{Report the results of calibrating the stroke model using e.g. Platt scaling or Isotonic scaling.} % https://scikit-learn.org/stable/modules/calibration.html
-\lesstodo[inline]{Maybe mention MultiQT paper \cite{havtorn_multiqt_2020}.}
+\lesstodo[inline]{Maybe mention MultiQT paper \parencite{havtorn_multiqt_2020}.}

### 2023-09-29 chapters/dummy_chapters/dummy_introduction.tex
Move table captions to top
Updates to citation commands and footnotes

--- 2022-02-27 chapters/dummy_introduction.tex
@@
-%!TEX root = ../Thesis.tex
+%!TEX root = ../thesis.tex
 \chapter{Introduction}
@@
 Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum \cite{adams1980hitchhiker}.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum \parencite{adams1980hitchhiker}.
 
@@
 Fusce commodo augue quis elit sagittis non viverra turpis bibendum.
-Nunc placerat sem non sapien malesuada malesuada ullamcorper orci luctus \cite{adams1980hitchhiker}.
+Nunc placerat sem non sapien malesuada malesuada ullamcorper orci luctus \parencite{adams1980hitchhiker}.
 Morbi pharetra ligula integer mollis mi nec neque ultrices vitae volutpat leo ullamcorper.
@@
 \begin{table}
+    \caption{This is a caption to the table}
     \centering
@@
         \end{tabular}
-    \caption{This is a caption to the table}
     \label{myTable}

### 2023-09-29 chapters/introduction.tex
Updates to citation commands and footnotes

@@
 The application of artificial intelligence in medical decision-making and particularly for diagnosing and recommending treatments is likely to hold significant potential to reduce the effect of these issues and improve patient outcomes.
-However, as such decisions impact critical aspects of human health and well-being, this prospect also raises concerns related to the reliability of artificial intelligence systems \cite{chen_ethical_2021, shailaja_machine_2018, ahmad_interpretable_2018}.
+However, as such decisions impact critical aspects of human health and well-being, this prospect also raises concerns related to the reliability of artificial intelligence systems \parencite{chen_ethical_2021, shailaja_machine_2018, ahmad_interpretable_2018}.
 
 The evolution of technology to support and automate decision-making can be traced back to the development of algorithms and mathematical techniques by ancient civilizations. 
-The Egyptian Rhind Mathematical Papyrus, dated around 1550 BCE, describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \cite{georges_universal_2001}.
-Around the same time, the Babylonians had developed a sophisticated system of mathematics that included algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \cite{fowler_square_1998}.
+The Egyptian Rhind Mathematical Papyrus, dated around 1550 BCE, describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \parencite{georges_universal_2001}.
+Around the same time, the Babylonians had developed a sophisticated system of mathematics that included algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \parencite{fowler_square_1998}.
 
 These computing tools brought with them novel challenges.
-The Babylonian mathematics, for instance, did not formalize the difference between rational and irrational numbers often leading to approximation errors such as the common practice of approximating $\pi$ with 3 \cite{georges_universal_2001}.
+The Babylonian mathematics, for instance, did not formalize the difference between rational and irrational numbers often leading to approximation errors such as the common practice of approximating $\pi$ with 3 \parencite{georges_universal_2001}.
 Apart from such inherent limitations of the tools, inadequate human proficiency in their use was also a source of error.
-The Plimpton 322 clay tablet, for instance, contains several errors, presumably made by a novice in Babylonian mathematics (\cref{fig:plimpton_332}) \cite{neugebauer_mathematical_1945,britton_plimpton_2011}.
+The Plimpton 322 clay tablet, for instance, contains several errors, presumably made by a novice in Babylonian mathematics (\cref{fig:plimpton_332}) \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
 % Easy to overlook, the challenges we face in adopting new technologies today are not new but have been faced by humans for thousands of years.
@@
 The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics.
-\cite[photo credit][]{neugebauer_mathematical_1945}}
+\parencite[photo credit][]{neugebauer_mathematical_1945}}
     \label{fig:plimpton_332}
@@
 % it took one-hundred years before / 
-% the first programmable, electronic, general-purpose digital computers were developed, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}) \cite{georges_universal_2001}. 
+% the first programmable, electronic, general-purpose digital computers were developed, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}) \parencite{georges_universal_2001}. 
 As manufacturing processes were developed and scaled, electronic computers were adopted widely across society for various applications in science and industry, including healthcare. 
-The societal changes brought about by the widespread adoption of computer technology through the latter half of the 20th century were so profound that some scholars refer to the era as the information age \cite{georges_universal_2001, harari_sapiens_2011}.
+The societal changes brought about by the widespread adoption of computer technology through the latter half of the 20th century were so profound that some scholars refer to the era as the information age \parencite{georges_universal_2001, harari_sapiens_2011}.
 
@@
     \caption{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.
-\cite[photo credit][]{usarmyresearchlaboratoryarltechnicallibrary_female_1940}} %Photo: \copyright Corbis/ Getty Images.}
+\parencite[photo credit][]{usarmyresearchlaboratoryarltechnicallibrary_female_1940}} %Photo: \copyright Corbis/ Getty Images.}
     \label{fig:eniac_programmers}
@@
 This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that focuses on developing algorithms that can learn from data to recognize patterns and make predictions to guide decision-making.
-Enabled by a massive increase in the performance and availability of parallel computing resources, such methods have set new standards for the abilities of computer systems in various domains, such as speech recognition, computer vision, and natural language processing \cite{lecun_deep_2015}.
-In recent years, machine learning has made its way into the healthcare sector, where it has shown promising results in tasks such as medical imaging \cite{lundervold_overview_2019}, drug discovery \cite{chen_rise_2018}, and clinical decision support \cite{cite15, cite14}.
+Enabled by a massive increase in the performance and availability of parallel computing resources, such methods have set new standards for the abilities of computer systems in various domains, such as speech recognition, computer vision, and natural language processing \parencite{lecun_deep_2015}.
+In recent years, machine learning has made its way into the healthcare sector, where it has shown promising results in tasks such as medical imaging \parencite{lundervold_overview_2019}, drug discovery \parencite{chen_rise_2018}, and clinical decision support \parencite{cite15, cite14}.
 
@@
 Specifically, during those seconds, the system incorrectly classified the pedestrian more than ten times, first as a vehicle, then as an unknown object and ultimately as a bicyclist, each time changing, or resetting, the predicted path.
-About one second before the crash, the system determined that a collision was imminent, but the situation exceeded the constraints within which the autonomous driving system was allowed to operate, and the car's safety driver failed to intervene \cite{nationaltransportationsafetyboardnhsa_collision_2019}.
+About one second before the crash, the system determined that a collision was imminent, but the situation exceeded the constraints within which the autonomous driving system was allowed to operate, and the car's safety driver failed to intervene \parencite{nationaltransportationsafetyboardnhsa_collision_2019}.
 
@@
     \centering
-    \includegraphics[width=0.95\textwidth]{uber_nhsa_accident.png}
+    \includegraphics[width=0.90\textwidth]{uber_nhsa_accident.png}
     \caption{A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
-The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \cite[photo credit][]{nationaltransportationsafetyboardnhsa_collision_2019}.}
+The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \parencite[photo credit][]{nationaltransportationsafetyboardnhsa_collision_2019}.}
     % \caption{Aerial view of crash location showing path of pedestrian as she attempted to cross N.
 Mill Avenue and movement and speed of SUV at three points before impact.
-Pedestrian's path shows her position from initial detection ($5.6$ seconds before impact) until impact; SUV's position is shown at corresponding times beginning $4.2$ seconds before impact \cite{nationaltransportationsafetyboardnhsa_collision_2019}.}
+Pedestrian's path shows her position from initial detection ($5.6$ seconds before impact) until impact; SUV's position is shown at corresponding times beginning $4.2$ seconds before impact \parencite{nationaltransportationsafetyboardnhsa_collision_2019}.}
     \label{fig:uber_nhsa_accident}
@@
 And Uber is not the only actor facing the challenges of applying machine learning in real-world scenarios.
-A recent report by the Washington Post found that Tesla's Autopilot system has been involved in at least 736 crashes in the last four years with 16 fatalities \cite{siddiqui_17_2023}.
+A recent report by the Washington Post found that Tesla's Autopilot system has been involved in at least 736 crashes in the last four years with 16 fatalities \parencite{siddiqui_17_2023}.
 % And though recent statistics published by Waymo, 
 % %the self-driving car company owned by Google's parent company Alphabet, 
-% revealed far fewer crashes, with only 2 of 20 incidents over 1 million miles meeting reporting criteria, it is interesting to note that the cars are all operated in the sunny states of California and Arizona, where driving conditions vary relatively little \cite{hawkins_waymo_2023}.
+% revealed far fewer crashes, with only 2 of 20 incidents over 1 million miles meeting reporting criteria, it is interesting to note that the cars are all operated in the sunny states of California and Arizona, where driving conditions vary relatively little \parencite{hawkins_waymo_2023}.
 While driver-assistance systems undoubtedly help reduce accidents in many cases, and has great future potential, the issues of current systems highlight the need for machine learning systems that are robust in rare, adverse or difficult situations.
 
-Recently, European policymakers proposed a set of requirements to ensure that machine learning applications are safe, reliable, and trustworthy \cite{europeancommission_briefing_2021}. 
+Recently, European policymakers proposed a set of requirements to ensure that machine learning applications are safe, reliable, and trustworthy \parencite{europeancommission_briefing_2021}. 
 This AI Act groups applications into three categories based on the potential risk they pose to society: limited, high, and unacceptable.
@@
 
-\textit{"Future AI solutions for healthcare should be implemented by integrating uncertainty estimation, a relatively new field of research that aims to provide clinicians with clinically useful indications on the degree of confidence in AI predictions"} \cite{europeanparliament_artificial_2022}.
+\textit{"Future AI solutions for healthcare should be implemented by integrating uncertainty estimation, a relatively new field of research that aims to provide clinicians with clinically useful indications on the degree of confidence in AI predictions"} \parencite{europeanparliament_artificial_2022}.
 
@@
 The system is in continued use in 2023, and provides an interface for navigating a graph-based protocol, defined by the local emergency medical services, which dispatchers use to triage patients during emergency calls.
-Calls are transcribed in real-time using automatic speech recognition built for the customer-domain and the system makes suggestions for the dispatcher to use in the conversation with the caller, including notifications about forgotten protocol questions \cite{havtorn_multiqt_2020} and potential cases of cardiac arrests \cite{cite15, cite14}.
+Calls are transcribed in real-time using automatic speech recognition built for the customer-domain and the system makes suggestions for the dispatcher to use in the conversation with the caller, including notifications about forgotten protocol questions \parencite{havtorn_multiqt_2020} and potential cases of cardiac arrests \parencite{cite15, cite14}.
 The system logs the details about the call and the actions taken by the dispatcher for later use in review, training and quality assurance.
@@
 \paragraph{Motivation} In 2021, Corti entered a research collaboration with the Capital Region of Denmark and the Copenhagen Emergency Services to develop a system for recognizing stroke cases in calls to the 1-1-2 emergency line and the 1813 medical helpline.
-Stroke is one of the biggest causes of disability and death worldwide \cite{cite1,cite2,cite3} and effective treatment is highly time-sensitive \cite{cite4,cite5}.
-The most common gateway to specialized treatment and hospital admittance is through prehospital telehealth services like emergency medical call centers, nurse advice call lines, and out-of-hours health services \cite{cite6,cite7}, however, studies have found that approximately half of all patients with stroke do not receive the correct triage for their condition from call-takers \cite{cite10,cite11,cite12}.
+Stroke is one of the biggest causes of disability and death worldwide \parencite{cite1,cite2,cite3} and effective treatment is highly time-sensitive \parencite{cite4,cite5}.
+The most common gateway to specialized treatment and hospital admittance is through prehospital telehealth services like emergency medical call centers, nurse advice call lines, and out-of-hours health services \parencite{cite6,cite7}, however, studies have found that approximately half of all patients with stroke do not receive the correct triage for their condition from call-takers \parencite{cite10,cite11,cite12}.
 This is likely due to the complexity of stroke cases which exhibit a wide range of symptoms that can be difficult to recognize over the phone.
@@
 This makes it difficult for call-takers to gain practical experience with stroke cases which compounds to the difficulty of recognizing them.
-Although several efforts have been made to improve recognition rates, there is still a need for better tools to support call-takers in recognizing stroke cases \cite{cite13,cite14,cite15}.
+Although several efforts have been made to improve recognition rates, there is still a need for better tools to support call-takers in recognizing stroke cases \parencite{cite13,cite14,cite15}.
 
@@
 The process of medical coding is time-consuming and error-prone due to the vast number and high complexity of medical diagnoses and procedures.
-For instance, the widely used International Classification of Diseases (ICD) standard consist of 55,000 medical codes in version 10 and 85,000 in version 11 \cite{worldhealthorganisationwho_international_2023}.
-Additionally, a single medical note will usually contain several diagnoses and procedures which must all be inferred from the clinical note \cite{johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}.
-Furthermore, any single code can have several exclusion or inclusion criteria defined by official guidelines that determine when other codes must be coded along with it, and when they must not \cite{centersformedicaremedicaidservicesus_icd10cm_2023}.
+For instance, the widely used International Classification of Diseases (ICD) standard consist of 55,000 medical codes in version 10 and 85,000 in version 11 \parencite{worldhealthorganisationwho_international_2023}.
+Additionally, a single medical note will usually contain several diagnoses and procedures which must all be inferred from the clinical note \parencite{johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}.
+Furthermore, any single code can have several exclusion or inclusion criteria defined by official guidelines that determine when other codes must be coded along with it, and when they must not \parencite{centersformedicaremedicaidservicesus_icd10cm_2023}.
 % Official guidelines describe how to code specific diagnoses and procedures and combinations thereof . 
@@
 Since many modern machine learning models are deep neural networks with millions or billions of parameters, their size and complexity make them inherently difficult to interpret, and their predictions hard to explain.
-Due to high cost and practical infeasibility, the vast amounts of data needed to train such models are often not manually curated for quality, and so, may contain biases and errors which models are well-equipped to learn to mirror \cite{burkart_survey_2021}.
-Finally, a number of factors, including the ability of deep neural networks to overfit, or even memorize, the training data \cite{arpit_closer_2017, burg_memorization_2021}, can lead to models that are sensitive adverse noise conditions and outliers.
+Due to high cost and practical infeasibility, the vast amounts of data needed to train such models are often not manually curated for quality, and so, may contain biases and errors which models are well-equipped to learn to mirror \parencite{burkart_survey_2021}.
+Finally, a number of factors, including the ability of deep neural networks to overfit, or even memorize, the training data \parencite{arpit_closer_2017, burg_memorization_2021}, can lead to models that are sensitive adverse noise conditions and outliers.
 
 The ability of a model to accurately estimate the uncertainty of its predictions plays an important role in its reliability.
-However, the tendency of deep neural networks to overfit or memorize the training data \cite{arpit_closer_2017, burg_memorization_2021} and the use of training objectives that are proxies for the evaluation metrics of interest lead to models that often are miscalibrated; that is, the predicted probability of a class does not reflect the true probability of the class.
-This means, for instance, that if for a certain input the model assigns 70\% probability to a class, it would not generally be right 70\% of the time - if we could repeat the prediction multiple times under similar circumstances \cite{guo_calibration_2017, kull_temperature_2019}.
+However, the tendency of deep neural networks to overfit or memorize the training data \parencite{arpit_closer_2017, burg_memorization_2021} and the use of training objectives that are proxies for the evaluation metrics of interest lead to models that often are miscalibrated; that is, the predicted probability of a class does not reflect the true probability of the class.
+This means, for instance, that if for a certain input the model assigns 70\% probability to a class, it would not generally be right 70\% of the time - if we could repeat the prediction multiple times under similar circumstances \parencite{guo_calibration_2017, kull_temperature_2019}.
 Usually, the predicted probabilities are too extreme which leads to overconfident predictions.
@@
 % 
-The problem of miscalibration in machine learning models is well-known and has been studied for decades \cite{lewis_sequential_1995, platt_probabilistic_1999, garczarek_classification_2002, zadrozny_transforming_2002, bennett_using_2003}.
-One of the best known methods for calibrating a binary classifier is Platt scaling which fits a logistic regression model to the model outputs, assuming that the miscalibration can be corrected by a logarithmic function \cite{platt_probabilistic_1999}.
-Another method is isotonic regression which instead fits a non-parametric, monotonic function \cite{zadrozny_transforming_2002}.
+The problem of miscalibration in machine learning models is well-known and has been studied for decades \parencite{lewis_sequential_1995, platt_probabilistic_1999, garczarek_classification_2002, zadrozny_transforming_2002, bennett_using_2003}.
+One of the best known methods for calibrating a binary classifier is Platt scaling which fits a logistic regression model to the model outputs, assuming that the miscalibration can be corrected by a logarithmic function \parencite{platt_probabilistic_1999}.
+Another method is isotonic regression which instead fits a non-parametric, monotonic function \parencite{zadrozny_transforming_2002}.
 More recently, \textcite{guo_calibration_2017} proposed a single-parameter variant of Platt-scaling that fits only a temperature parameter on the logits of a neural network classifier. 
 These methods are generally simple and effective, but not without limitations. 
-To perform the calibration, most of the methods require a held-out validation set on which the model was not trained, Platt scaling and isotonic regression are not naively applicable to multi-class classification problems, and how to calibrate models for structured prediction tasks, such as speech recognition and machine translation, remains an open problem \cite{astudillo_uncertainty_2010, astudillo_integration_2013, jayashankar_detecting_2020}.
+To perform the calibration, most of the methods require a held-out validation set on which the model was not trained, Platt scaling and isotonic regression are not naively applicable to multi-class classification problems, and how to calibrate models for structured prediction tasks, such as speech recognition and machine translation, remains an open problem \parencite{astudillo_uncertainty_2010, astudillo_integration_2013, jayashankar_detecting_2020}.
 
@@
 Indeed, since no horses were in the training data, the model will not have learned specialized features for horses, nor learned to associate relevant known features with horses.
-So, if some features of a particular horse resemble those learned for a dog more than for a cat, the model might assign arbitrarily high probability to the dog class and therefore risks being confidently wrong \cite{zhou_survey_2022}.
+So, if some features of a particular horse resemble those learned for a dog more than for a cat, the model might assign arbitrarily high probability to the dog class and therefore risks being confidently wrong \parencite{zhou_survey_2022}.
 
@@
 At a fundamental level, we can distinguish between uncertainty that is present in the knowledge we have, and uncertainty that originates from the knowledge that we do not have.
-These sources are sometimes called \emph{known unknowns} and \emph{unknown unknowns}, or referred to in terms of \textit{aleatoric uncertainty} and \textit{epistemic uncertainty} \cite{kendall_what_2017}.
+These sources are sometimes called \emph{known unknowns} and \emph{unknown unknowns}, or referred to in terms of \textit{aleatoric uncertainty} and \textit{epistemic uncertainty} \parencite{kendall_what_2017}.
 Aleatoric comes from the Latin word 'aleatorius' for 'dice player' or 'gambler' and refers to uncertainty present in the data itself due to randomness in the process that generated it.
@@
 Such lack of knowledge could for instance be due to a lack of data, or an improper model specification.
-Aleatoric uncertainty is sometimes referred to as \textit{stochastic uncertainty} and epistemic uncertainty as \textit{systematic uncertainty} \cite{kendall_what_2017}. 
+Aleatoric uncertainty is sometimes referred to as \textit{stochastic uncertainty} and epistemic uncertainty as \textit{systematic uncertainty} \parencite{kendall_what_2017}. 
 Although the distinction between aleatoric and epistemic uncertainty is important, it is not always useful.
@@
 This makes epistemic uncertainty more difficult to quantify than aleatoric uncertainty.
-Methods for estimating epistemic uncertainty include Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995} which represent it by an explicit distribution over learned model parameters.
-Ensemble methods \cite{gal_dropout_2016,lakshminarayanan_simple_2017} take a similar approach but use an implicit distribution.
+Methods for estimating epistemic uncertainty include Bayesian neural networks \parencite{mackay_practical_1992, neal_bayesian_1995} which represent it by an explicit distribution over learned model parameters.
+Ensemble methods \parencite{gal_dropout_2016,lakshminarayanan_simple_2017} take a similar approach but use an implicit distribution.
 Other approaches include anomaly detection and the recent field of out-of-distribution detection which, for example, represent epistemic uncertainty by special output classes for out-of-distribution data, distributions over data representations, or distances between them (see \cref{sec:out-of-distribution-detection}).
@@
 This makes epistemic uncertainty more difficult to model and quantify than aleatoric uncertainty.
-% % Methods for estimating epistemic uncertainty include Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995}, ensembling \cite{gal_dropout_2016,lakshminarayanan_simple_2017} and a suite of recent research into out-of-distribution detection (see \cref{sec:out-of-distribution-detection}).
-% In contrast to aleatoric uncertainty, which is usually represented by a distribution over the data, epistemic uncertainty is often represented by an implicit or explicit distribution over the model parameters, such as in Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995} and ensemble models \cite{gal_dropout_2016,lakshminarayanan_simple_2017}.
+% % Methods for estimating epistemic uncertainty include Bayesian neural networks \parencite{mackay_practical_1992, neal_bayesian_1995}, ensembling \parencite{gal_dropout_2016,lakshminarayanan_simple_2017} and a suite of recent research into out-of-distribution detection (see \cref{sec:out-of-distribution-detection}).
+% In contrast to aleatoric uncertainty, which is usually represented by a distribution over the data, epistemic uncertainty is often represented by an implicit or explicit distribution over the model parameters, such as in Bayesian neural networks \parencite{mackay_practical_1992, neal_bayesian_1995} and ensemble models \parencite{gal_dropout_2016,lakshminarayanan_simple_2017}.
 In most cases, such methods require multiple forward passes through the model to estimate the predictive uncertainty making them computationally expensive and impractical for real-time applications.
@@
 
-\section{Research hypotheses}
+\section{Thesis outline and research questions}
 
 \todo[inline]{Formulate research hypotheses}
+
+This introduction provided a high-level overview of the motivation behind the research project by focusing on the use-cases of stroke recognition and automated medical coding as well as speech processing and the importance of uncertainty quantification.
+The remainder of \cref{part:background} consists of two chapters. 
+\Cref{chp:technical-background} provides relevant technical background to the papers of the thesis, including uncertainty, out-of-distribution detection, and variational autoencoders while \cref{chp:main-contributions} outlines the main contributions of the thesis via the published papers in relation to the research hypotheses.
+
+\Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. 
+\Cref{part:unsupervised-uncertainty-estimation} contains papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
+Here we examine whether it is possible to devise a method that utilizes a hierarhy of latent variables in
+
+\Cref{part:self-supervised-speech-representation-learning} contains a review paper on self-supervised speech representation learning \parencite{mohamed_selfsupervised_2022}.
+\Cref{part:medical-applications} contains papers on applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}.
+
+Finally, \cref{part:discussion-and-conclusion} concludes the thesis by summarizing the main contributions and discussing the presented work and future directions for research.
 
@@
 % The complete set of research published as part of this project covers a number of topics related to decision-support in medical conversations. 
-% One group of the primary papers is concerned with generative latent variable models and their applications to estimation of epistemic uncertainty and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
-% Another group of studies examines self-supervised speech representation learning as an alternative to generative latent variable models \cite{mohamed_selfsupervised_2022,borgholt_brief_2022} and explores them for speech recognition \cite{borgholt_scaling_2021} and spoken language understanding \cite{borgholt_we_2021}.
-% Finally, a group deals with applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.
+% One group of the primary papers is concerned with generative latent variable models and their applications to estimation of epistemic uncertainty and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
+% Another group of studies examines self-supervised speech representation learning as an alternative to generative latent variable models \parencite{mohamed_selfsupervised_2022,borgholt_brief_2022} and explores them for speech recognition \parencite{borgholt_scaling_2021} and spoken language understanding \parencite{borgholt_we_2021}.
+% Finally, a group deals with applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}.
 %
-This introduction provided a high-level overview of the motivation behind the research project by focusing on the use-cases of stroke recognition and automated medical coding as well as speech processing and the importance of uncertainty quantification.
-The remainder of \cref{part:background} consists of two chapters.
-\Cref{chp:technical-background} provides relevant technical background to the papers of the thesis, including uncertainty, out-of-distribution detection, and variational autoencoders while \cref{chp:main-contributions} outlines the main contributions of the thesis via the published papers in relation to the research hypotheses.
-
-\Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper.
-\Cref{part:unsupervised-uncertainty-estimation} contains papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
-\Cref{part:self-supervised-speech-representation-learning} contains a review paper on self-supervised speech representation learning \cite{mohamed_selfsupervised_2022}.
-\Cref{part:medical-applications} contains papers on applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.
-
-Finally, \cref{part:discussion-and-conclusion} concludes the thesis by summarizing the main contributions and discussing the presented work and future directions for research.

### 2023-09-29 chapters/main-contributions.tex
Updates to citation commands and footnotes

@@
 \contribitem{chp:paper-benchmarking}
-This paper proposes a new hierarchical variational autoencoder for speech inspired by the Clockwork VAE \cite{saxena_clockwork_2021}. 
+This paper proposes a new hierarchical variational autoencoder for speech inspired by the Clockwork VAE \parencite{saxena_clockwork_2021}. 
 The model is benchmarked against other variational autoencoders and autoregressive models for speech.
@@
 \contribitem{part:self-supervised-speech-representation-learning}
-This part of the thesis is concerned with self-supervised speech representation learning and consists of a single paper that provides a substantial review of the field \cite{mohamed_selfsupervised_2022}.
+This part of the thesis is concerned with self-supervised speech representation learning and consists of a single paper that provides a substantial review of the field \parencite{mohamed_selfsupervised_2022}.
 Self-supervised learning is a promising approach to speech representation learning that has shown a fast, wide and successful adoption across speech modelling over the course of the last few years. 
@@
 
-An early version of this review paper \cite{borgholt_brief_2022} included a review of generative latent variable models for speech representation learning and a comparison between the two modelling paradigms.
+An early version of this review paper \parencite{borgholt_brief_2022} included a review of generative latent variable models for speech representation learning and a comparison between the two modelling paradigms.
 Due to the superior performance of self-supervised methods and a desire to focus on self-supervised methods, the paper in \cref{chp:paper-review} does not focus on generative latent variable models. 

### 2023-09-29 chapters/paper_automated.tex
Updates to citation commands and footnotes

@@
 
-Medical coding is the task of assigning diagnosis and procedure codes to free-text medical documentation \cite{dongAutomatedClinicalCoding2022,tengReviewDeepNeural2022}.
+Medical coding is the task of assigning diagnosis and procedure codes to free-text medical documentation \parencite{dongAutomatedClinicalCoding2022,tengReviewDeepNeural2022}.
 
 These codes ensure that patients receive the correct level of care and that healthcare providers are accurately compensated for their services.
-However, this is a costly manual process prone to error \cite{tsengAdministrativeCostsAssociated2018, omalleyMeasuringDiagnosesICD2005, burnsSystematicReviewDischarge2012}.
+However, this is a costly manual process prone to error \parencite{tsengAdministrativeCostsAssociated2018, omalleyMeasuringDiagnosesICD2005, burnsSystematicReviewDischarge2012}.
 
 The goal of automated medical coding (AMC) is to predict a set of codes or provide a list of codes ranked by relevance for a medical document.
-Numerous machine learning models have been developed for AMC \cite{jiUnifiedReviewDeep2022, tengReviewDeepNeural2022, stanfillSystematicLiteratureReview2010}.
+Numerous machine learning models have been developed for AMC \parencite{jiUnifiedReviewDeep2022, tengReviewDeepNeural2022, stanfillSystematicLiteratureReview2010}.
 These models are trained on datasets of medical documents, typically discharge summaries, each labeled with a set of medical codes.
-While some models treat AMC as an ad-hoc information retrieval problem \cite{rizzoICDCodeRetrieval2015,parkInformationRetrievalApproach2019}, it is more commonly posed as a multi-label classification problem \cite{tengReviewDeepNeural2022, jiUnifiedReviewDeep2022}.
+While some models treat AMC as an ad-hoc information retrieval problem \parencite{rizzoICDCodeRetrieval2015,parkInformationRetrievalApproach2019}, it is more commonly posed as a multi-label classification problem \parencite{tengReviewDeepNeural2022, jiUnifiedReviewDeep2022}.
 
-While most research in AMC has been conducted on the third version of the Medical Information Mart for Intensive Care dataset (MIMIC-III) \cite{tengReviewDeepNeural2022, venkateshAutomatingOverburdenedClinical2023}, it remains challenging to compare the results of different models.
+While most research in AMC has been conducted on the third version of the Medical Information Mart for Intensive Care dataset (MIMIC-III) \parencite{tengReviewDeepNeural2022, venkateshAutomatingOverburdenedClinical2023}, it remains challenging to compare the results of different models.
 Performance improvements are commonly attributed to model design, but differences in experimental setups make these claims hard to validate.
-In addition, long documents, rare codes, and lack of training data are often cited as core research challenges \cite{baoMedicalCodePrediction2021,dongAutomatedClinicalCoding2022,dongExplainableAutomatedCoding2021,feuchtDescriptionbasedLabelAttention2021,gaoLimitationsTransformersClinical2021,huangPLMICDAutomaticICD2022,jiDoesMagicBERT2021,jiUnifiedReviewDeep2022,kavuluruEmpiricalEvaluationSupervised2015,kimReadAttendCode2021,liICDCodingClinical2020,liuEffectiveConvolutionalAttention2021,michalopoulosICDBigBirdContextualEmbedding2022,moonsComparisonDeepLearning2020,pascualBERTbasedAutomaticICD2021,tengReviewDeepNeural2022,tengExplainablePredictionMedical2020,xieEHRCodingMultiscale2019,yangKnowledgeInjectedPrompt2022,zhangBERTXMLLargeScale2020,zhouAutomaticICDCoding2021, vuLabelAttentionModel2020, venkateshAutomatingOverburdenedClinical2023}.
-However, except for a few studies demonstrating performance drops on rare codes, the number of studies containing in-depth error analyses is limited \cite{baoMedicalCodePrediction2021,dongExplainableAutomatedCoding2021,jiDoesMagicBERT2021}.
+In addition, long documents, rare codes, and lack of training data are often cited as core research challenges \parencite{baoMedicalCodePrediction2021,dongAutomatedClinicalCoding2022,dongExplainableAutomatedCoding2021,feuchtDescriptionbasedLabelAttention2021,gaoLimitationsTransformersClinical2021,huangPLMICDAutomaticICD2022,jiDoesMagicBERT2021,jiUnifiedReviewDeep2022,kavuluruEmpiricalEvaluationSupervised2015,kimReadAttendCode2021,liICDCodingClinical2020,liuEffectiveConvolutionalAttention2021,michalopoulosICDBigBirdContextualEmbedding2022,moonsComparisonDeepLearning2020,pascualBERTbasedAutomaticICD2021,tengReviewDeepNeural2022,tengExplainablePredictionMedical2020,xieEHRCodingMultiscale2019,yangKnowledgeInjectedPrompt2022,zhangBERTXMLLargeScale2020,zhouAutomaticICDCoding2021, vuLabelAttentionModel2020, venkateshAutomatingOverburdenedClinical2023}.
+However, except for a few studies demonstrating performance drops on rare codes, the number of studies containing in-depth error analyses is limited \parencite{baoMedicalCodePrediction2021,dongExplainableAutomatedCoding2021,jiDoesMagicBERT2021}.
 
@@
 We find that models previously reported as low-performing improve considerably, demonstrating the importance of hyperparameters and decision boundary tuning.
-    \item We report the first results of current state-of-the-art models on the newly released MIMIC-IV dataset \cite{johnsonMIMICIVFreelyAccessible2023, goldbergerPhysioBankPhysioToolkitPhysioNet2000}.
+    \item We report the first results of current state-of-the-art models on the newly released MIMIC-IV dataset \parencite{johnsonMIMICIVFreelyAccessible2023, goldbergerPhysioBankPhysioToolkitPhysioNet2000}.
 We find that previous conclusions generalize to the new dataset.
@@
 \label{sec:dataset}
-The International Classification of Diseases (ICD) is the most popular medical coding system worldwide \cite{tengReviewDeepNeural2022}.
+The International Classification of Diseases (ICD) is the most popular medical coding system worldwide \parencite{tengReviewDeepNeural2022}.
 It follows a tree-like hierarchical structure, also known as a medical ontology, to ensure the functional and structural integrity of the classification.
@@
 Each revision introduces new codes.
-For instance, ICD-9 contains 18,000 codes, while ICD-10 contains 142,000.\footnote{\url{https://www.cdc.gov/nchs/icd/icd10cm_pcs_background.htm}} MIMIC-II and MIMIC-III are the most widely used open-access datasets for research on ICD coding and are provided by the Beth Israel Deaconess Medical Center \cite{tengReviewDeepNeural2022, johnsonMIMICIIIFreelyAccessible2016, leeOpenaccessMIMICIIDatabase2011}.
+For instance, ICD-9 contains 18,000 codes, while ICD-10 contains 142,000.\footnote{\url{https://www.cdc.gov/nchs/icd/icd10cm_pcs_background.htm}} MIMIC-II and MIMIC-III are the most widely used open-access datasets for research on ICD coding and are provided by the Beth Israel Deaconess Medical Center \parencite{tengReviewDeepNeural2022, johnsonMIMICIIIFreelyAccessible2016, leeOpenaccessMIMICIIDatabase2011}.
 
-MIMIC-III contains medical documents annotated with ICD-9 codes collected between 2001 and 2012 \cite{johnsonMIMICIIIFreelyAccessible2016}.
-Usually, discharge summaries---free-text notes on patient and hospitalization history---are the only documents used for AMC \cite{tengReviewDeepNeural2022}.
+MIMIC-III contains medical documents annotated with ICD-9 codes collected between 2001 and 2012 \parencite{johnsonMIMICIIIFreelyAccessible2016}.
+Usually, discharge summaries---free-text notes on patient and hospitalization history---are the only documents used for AMC \parencite{tengReviewDeepNeural2022}.
 MIMIC-III \textit{full} and \textit{50} are commonly used splits.
-MIMIC-III \textit{full} contains all ICD-9 codes, while \textit{50} only contains the top 50 most frequent codes \cite{mullenbachExplainablePredictionMedical2018, shiAutomatedICDCoding2018}.
+MIMIC-III \textit{full} contains all ICD-9 codes, while \textit{50} only contains the top 50 most frequent codes \parencite{mullenbachExplainablePredictionMedical2018, shiAutomatedICDCoding2018}.
 
 MIMIC-IV was released on January 6th, 2023, and has not previously been used for AMC.
-It contains data for patients admitted to the Beth Israel Deaconess Medical Center emergency department or ICU between 2008-2019  annotated with either ICD-9 or ICD-10 codes \cite{johnsonMIMICIVFreelyAccessible2023}.
+It contains data for patients admitted to the Beth Israel Deaconess Medical Center emergency department or ICU between 2008-2019  annotated with either ICD-9 or ICD-10 codes \parencite{johnsonMIMICIVFreelyAccessible2023}.
 The empirical frequencies of codes of each ICD version in MIMIC-IV are shown in \cref{fig:mimiciv_code_dist}.
@@
     \centering
-    \caption{Comparison of the previously defined MIMIC-III splits \textit{full} and \textit{50} \cite{mullenbachExplainablePredictionMedical2018} and our proposed MIMIC-III \textit{clean} split along with similarly defined splits for MIMIC-IV \textit{ICD-9} and \textit{ICD-10} after pre-processing.}
+    \caption{Comparison of the previously defined MIMIC-III splits \textit{full} and \textit{50} \parencite{mullenbachExplainablePredictionMedical2018} and our proposed MIMIC-III \textit{clean} split along with similarly defined splits for MIMIC-IV \textit{ICD-9} and \textit{ICD-10} after pre-processing.}
     \label{tab:subsets}
@@
 
-\subsubsection{Encoders:}
+\paragraph{Encoders:}
 The encoder usually consists of pre-trained non-contextualized word embeddings (e.g., Word2Vec) and a neural network for encoding context.
-More recently, pre-trained masked language models (e.g., BERT) have gained popularity~\cite{tengReviewDeepNeural2022}.
+More recently, pre-trained masked language models (e.g., BERT) have gained popularity~\parencite{tengReviewDeepNeural2022}.
 The MIMIC-III training set or PubMed articles are commonly used for pre-training.
 
-\subsubsection{Decoders:}
+\paragraph{Decoders:}
 The most common decoder architectures can be grouped into three primary types. 
 The simplest decoder is a pooling layer (e.g., max pooling) followed by a feed-forward neural network.
-More recently, label-wise attention (LA) \cite{mullenbachExplainablePredictionMedical2018} has replaced pooling \cite{vuLabelAttentionModel2020, liuEffectiveConvolutionalAttention2021, huangPLMICDAutomaticICD2022, liICDCodingClinical2020}.
+More recently, label-wise attention (LA) \parencite{mullenbachExplainablePredictionMedical2018} has replaced pooling \parencite{vuLabelAttentionModel2020, liuEffectiveConvolutionalAttention2021, huangPLMICDAutomaticICD2022, liICDCodingClinical2020}.
 LA transforms a sequence of hidden representations $H$ into label-specific representations $V \in \mathbb{R}^{d_h \times L}$, where $L$ is the number of unique medical codes in the dataset.
@@
 Then, $V$ is used to compute class-wise probabilities via a feedforward neural network. 
-As LA was first used in the \textit{convolutional attention for multi-label classification} (CAML) model~\cite{mullenbachExplainablePredictionMedical2018}, we refer to this method as \lacaml.
+As LA was first used in the \textit{convolutional attention for multi-label classification} (CAML) model~\parencite{mullenbachExplainablePredictionMedical2018}, we refer to this method as \lacaml.
 
-An updated label-wise attention module was introduced in the \textit{label attention model} (LAAT) \cite{vuLabelAttentionModel2020}.
+An updated label-wise attention module was introduced in the \textit{label attention model} (LAAT) \parencite{vuLabelAttentionModel2020}.
 We refer to this attention module as \lalaat.
@@
 
-\citet{mullenbachExplainablePredictionMedical2018} released code for pre-processing the discharge summaries, generating the train-test split, and evaluating model performance on MIMIC-III which many subsequent papers have used \cite{vuLabelAttentionModel2020, huangPLMICDAutomaticICD2022, liICDCodingClinical2020, baoMedicalCodePrediction2021, yuanCodeSynonymsMatter2022, kimReadAttendCode2021}.
+\textcite{mullenbachExplainablePredictionMedical2018} released code for pre-processing the discharge summaries, generating the train-test split, and evaluating model performance on MIMIC-III which many subsequent papers have used \parencite{vuLabelAttentionModel2020, huangPLMICDAutomaticICD2022, liICDCodingClinical2020, baoMedicalCodePrediction2021, yuanCodeSynonymsMatter2022, kimReadAttendCode2021}.
 Pre-processing consisted of lowercasing all text and removing words that only contain out-of-alphabet characters. Predicting procedures and diagnosis codes were treated as a single task.
@@
 For instance, due to varying memory constraints of different models, documents are usually truncated to some maximum length.
-In the literature, this maximum varies from 2,500 to 4,000 words \cite{mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, huangPLMICDAutomaticICD2022}.
+In the literature, this maximum varies from 2,500 to 4,000 words \parencite{mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, huangPLMICDAutomaticICD2022}.
 Furthermore, not all papers tune the prediction decision boundary but simply set it to 0.5,
- hyperparameter search ranges and sampling methods  vary between works, and learning rate schedulers are only used in LAAT and PLM-ICD\cite{mullenbachExplainablePredictionMedical2018, liICDCodingClinical2020}.
+ hyperparameter search ranges and sampling methods  vary between works, and learning rate schedulers are only used in LAAT and PLM-ICD\parencite{mullenbachExplainablePredictionMedical2018, liICDCodingClinical2020}.
 In LAAT, the learning rate was decreased by 90\% when the F1 score had not increased for five epochs.
@@
         \midrule
-        Bi-GRU \cite{mullenbachExplainablePredictionMedical2018}   &  Word2Vec, Bi-GRU & Max-pool  &  9.9M\\
-        CNN \cite{mullenbachExplainablePredictionMedical2018}   & Word2Vec, CNN & Max-pool &  10.3M\\
-        CAML \cite{mullenbachExplainablePredictionMedical2018}   & Word2Vec, CNN  & \lacaml &  6.1M\\
-        MultiResCNN \cite{liICDCodingClinical2020} &  Word2Vec, ResNet & \lacaml &  11.9M\\
-        LAAT \cite{vuLabelAttentionModel2020} & Word2Vec, Bi-LSTM  & \lalaat  &  21.9M\\
-        PLM-ICD \cite{huangPLMICDAutomaticICD2022} &  BERT & \lalaat &  138.8M\\
+        Bi-GRU \parencite{mullenbachExplainablePredictionMedical2018}   &  Word2Vec, Bi-GRU & Max-pool  &  9.9M\\
+        CNN \parencite{mullenbachExplainablePredictionMedical2018}   & Word2Vec, CNN & Max-pool &  10.3M\\
+        CAML \parencite{mullenbachExplainablePredictionMedical2018}   & Word2Vec, CNN  & \lacaml &  6.1M\\
+        MultiResCNN \parencite{liICDCodingClinical2020} &  Word2Vec, ResNet & \lacaml &  11.9M\\
+        LAAT \parencite{vuLabelAttentionModel2020} & Word2Vec, Bi-LSTM  & \lalaat  &  21.9M\\
+        PLM-ICD \parencite{huangPLMICDAutomaticICD2022} &  BERT & \lalaat &  138.8M\\
         \bottomrule
@@
 In this study, we included both models trained from scratch and models with components pre-trained on external corpora. 
-We excluded models that use multi-modal inputs, such as medical code descriptions \cite{kimReadAttendCode2021, mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, caoHyperCoreHyperbolicCograph2020, baoMedicalCodePrediction2021}, code synonyms \cite{yuanCodeSynonymsMatter2022}, code hierarchies \cite{caoHyperCoreHyperbolicCograph2020, xieEHRCodingMultiscale2019}, or associated Wikipedia articles \cite{baiImprovingMedicalCode2019}, because they introduced additional complexity without providing evidence for significant performance improvements \cite{mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, tengReviewDeepNeural2022}.
+We excluded models that use multi-modal inputs, such as medical code descriptions \parencite{kimReadAttendCode2021, mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, caoHyperCoreHyperbolicCograph2020, baoMedicalCodePrediction2021}, code synonyms \parencite{yuanCodeSynonymsMatter2022}, code hierarchies \parencite{caoHyperCoreHyperbolicCograph2020, xieEHRCodingMultiscale2019}, or associated Wikipedia articles \parencite{baiImprovingMedicalCode2019}, because they introduced additional complexity without providing evidence for significant performance improvements \parencite{mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, tengReviewDeepNeural2022}.
 We excluded works without publically available source code as the experiment descriptions often lacked important implementation details.
@@
 
-Previous works calculated the macro F1 score as the harmonic mean of the macro precision and macro recall \cite{mullenbachExplainablePredictionMedical2018, liICDCodingClinical2020, vuLabelAttentionModel2020, huangPLMICDAutomaticICD2022}.
-\citet{opitzMacroF1Macro2021} analyze macro F1 formulas common in multi-class and multi-label classification.
+Previous works calculated the macro F1 score as the harmonic mean of the macro precision and macro recall \parencite{mullenbachExplainablePredictionMedical2018, liICDCodingClinical2020, vuLabelAttentionModel2020, huangPLMICDAutomaticICD2022}.
+\textcite{opitzMacroF1Macro2021} analyze macro F1 formulas common in multi-class and multi-label classification.
 They demonstrate that the above formulation is sub-optimal, as it rewards heavily biased classifiers in unbalanced datasets.
@@
 Therefore, we created a new MIMIC-III split to ensure that most codes are present in both the training and test set.
-Specifically, we removed codes with fewer than ten occurrences, doubled the test set size, and sampled the documents using multi-label stratified sampling \cite{sechidisStratificationMultilabelData2011}.
-We ensured that no patient occurred in both the training and test set, preprocessed the text, and considered procedures and diagnosis codes as a single task as done by \citet{mullenbachExplainablePredictionMedical2018}.
+Specifically, we removed codes with fewer than ten occurrences, doubled the test set size, and sampled the documents using multi-label stratified sampling \parencite{sechidisStratificationMultilabelData2011}.
+We ensured that no patient occurred in both the training and test set, preprocessed the text, and considered procedures and diagnosis codes as a single task as done by \textcite{mullenbachExplainablePredictionMedical2018}.
 We based our new split on the v1.4 version of the dataset and refer to it as MIMIC-III \textit{clean}.
@@
 We ran reproducibility experiments with all models to evaluate whether the results in the original works could be reproduced and to validate our reimplementations.
-We ran these experiments on MIMIC-III \textit{full}, and \textit{50} as in the original works \cite{mullenbachExplainablePredictionMedical2018,liICDCodingClinical2020,vuLabelAttentionModel2020,huangPLMICDAutomaticICD2022}.
+We ran these experiments on MIMIC-III \textit{full}, and \textit{50} as in the original works \parencite{mullenbachExplainablePredictionMedical2018,liICDCodingClinical2020,vuLabelAttentionModel2020,huangPLMICDAutomaticICD2022}.
 We used the hyperparameters reported in each paper (see \cref{tab:hyperparameters}) and report both the original and the revised macro F1 scores discussed in \cref{sec:metrics}.
@@
 We run experiments on the new MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9}, and \textit{ICD-10}.
-All models were trained for 20 epochs using a learning rate schedule with linear warmup for the first 2K updates followed by linear decay \cite{huangPLMICDAutomaticICD2022}.
+All models were trained for 20 epochs using a learning rate schedule with linear warmup for the first 2K updates followed by linear decay \parencite{huangPLMICDAutomaticICD2022}.
 We found this schedule to speed up the training convergence of all the models.
-Whereas original works use Adam or AdamW, we used AdamW for all experiments as it corrects the weight decay implementation of Adam \cite{kingmaAdamMethodStochastic2017, loshchilovDecoupledWeightDecay2022}.
+Whereas original works use Adam or AdamW, we used AdamW for all experiments as it corrects the weight decay implementation of Adam \parencite{kingmaAdamMethodStochastic2017, loshchilovDecoupledWeightDecay2022}.
 For each model, we tuned the decision boundary to maximize the micro F1 score on the validation set.
@@
 
-\subsubsection{Amount of training data:}
-Multiple studies attribute poor performance to data sparsity of MIMIC-III, which contains only fifty thousand examples \cite{kavuluruEmpiricalEvaluationSupervised2015,tengExplainablePredictionMedical2020,yanMedicalCodingClassification2010,yangKnowledgeInjectedPrompt2022}.
+\paragraph{Amount of training data:}
+Multiple studies attribute poor performance to data sparsity of MIMIC-III, which contains only fifty thousand examples \parencite{kavuluruEmpiricalEvaluationSupervised2015,tengExplainablePredictionMedical2020,yanMedicalCodingClassification2010,yangKnowledgeInjectedPrompt2022}.
 MIMIC-IV \textit{ICD-9} contains four times as many examples, which allows analyzing the effect of training set size.
 We train each model on 25k, 50k, 75k, 100k, and 125k examples and report micro and macro F1 on the fixed test set.
-The training subsets were sampled from the training set using multi-label stratified sampling to ensure the same code distributions \cite{sechidisStratificationMultilabelData2011}.
+The training subsets were sampled from the training set using multi-label stratified sampling to ensure the same code distributions \parencite{sechidisStratificationMultilabelData2011}.
 
-\subsubsection{Document length:}
+\paragraph{Document length:}
 We analyzed whether model performance correlates with document length on MIMIC-IV \textit{ICD-9}.
@@
 
-\subsubsection{Code analysis:}
+\paragraph{Code analysis:}
 To analyze the performance impact of rare codes, we first calculated the Pearson and Spearman correlation between model performance on each code and the corresponding code frequency in the training data.
@@
 Our reproduced results are indicated with *.
-The results were reproduced on MIMIC-III v1.4 with the preprocessing pipeline and splits of \citet{mullenbachExplainablePredictionMedical2018}.
+The results were reproduced on MIMIC-III v1.4 with the preprocessing pipeline and splits of \textcite{mullenbachExplainablePredictionMedical2018}.
 Each model was reproduced using the hyperparameters presented in the respective paper.
@@
 
-\subsubsection{Amount of training data:}
+\paragraph{Amount of training data:}
 
@@
 
-\subsubsection{Document length:}\label{subsubsec:document-length}
+\paragraph{Document length:}\label{subsubsec:document-length}
 
@@
 
-\subsubsection{Code analysis:}
+\paragraph{Code analysis:}
 
@@
 Our reproduced results were better than originally reported for Bi-GRU and CNN and worse for CAML and LAAT on most metrics.
-There have been multiple reports of issues in reproducing the results of \citet{mullenbachExplainablePredictionMedical2018}.\footnote{\url{https://github.com/jamesmullenbach/caml-mimic}} Additionally, most previous works did not report which version of MIMIC-III they used, and the code and hyperparameter configurations were not documented in detail.
+There have been multiple reports of issues in reproducing the results of \textcite{mullenbachExplainablePredictionMedical2018}.\footnote{\url{https://github.com/jamesmullenbach/caml-mimic}} Additionally, most previous works did not report which version of MIMIC-III they used, and the code and hyperparameter configurations were not documented in detail.
 Therefore, we hypothesize that our results differ because previous works report incorrect hyperparameters or use an earlier version of MIMIC-III.
@@
 In our revised comparison, we demonstrated that training the models using our setup decreased the difference between the best and worst micro F1 scores by $5.8$ percentage points.
-\citet{mullenbachExplainablePredictionMedical2018} concluded that CNN outperformed Bi-GRU.
+\textcite{mullenbachExplainablePredictionMedical2018} concluded that CNN outperformed Bi-GRU.
 However, in our revised comparison, Bi-GRU outperformed CNN on all metrics on MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9}, and MIMIC-IV \textit{ICD-10}.
 
-Even though MultiResCNN contains more parameters than CAML, \citet{liICDCodingClinical2020} concluded that MultiResCNN was faster to train because it converged in fewer epochs.
+Even though MultiResCNN contains more parameters than CAML, \textcite{liICDCodingClinical2020} concluded that MultiResCNN was faster to train because it converged in fewer epochs.
 However, this was only true when using the original setup where CAML converged after 84 epochs.
@@
 Since 54\% of the codes in MIMIC-III \textit{full} are missing in the test set, the maximum possible macro F1 score is 46\%.
-The previously highest reported macro F1 score on MIMIC-III \textit{full} is 12.7\% for PLM-ICD~\cite{kimReadAttendCode2021}.
+The previously highest reported macro F1 score on MIMIC-III \textit{full} is 12.7\% for PLM-ICD~\parencite{kimReadAttendCode2021}.
 Using our corrected macro F1 score on the same split, PLM-ICD achieved a macro F1 score of 22.8\%.
 This large difference from previous state-of-the-art seems to indicate that all previous work on AMC used the sub-optimally calculated macro F1 score, including works not reproduced in this paper.
-Many studies use the macro F1 score to evaluate the ability of their models to predict rare codes \cite{kimReadAttendCode2021, yuanCodeSynonymsMatter2022}.
+Many studies use the macro F1 score to evaluate the ability of their models to predict rare codes \parencite{kimReadAttendCode2021, yuanCodeSynonymsMatter2022}.
 If it has indeed been incorrectly calculated in these studies, some conclusions drawn in previous work regarding rare code prediction may have been misguided.
 
-Multiple studies mention lack of training data, rare codes, and long documents as the main challenges of AMC \cite{dongExplainableAutomatedCoding2021,feuchtDescriptionbasedLabelAttention2021,huangPLMICDAutomaticICD2022,jiDoesMagicBERT2021,liICDCodingClinical2020,liuEffectiveConvolutionalAttention2021,moonsComparisonDeepLearning2020,pascualBERTbasedAutomaticICD2021,tengReviewDeepNeural2022,tengExplainablePredictionMedical2020, vuLabelAttentionModel2020, venkateshAutomatingOverburdenedClinical2023}.
+Multiple studies mention lack of training data, rare codes, and long documents as the main challenges of AMC \parencite{dongExplainableAutomatedCoding2021,feuchtDescriptionbasedLabelAttention2021,huangPLMICDAutomaticICD2022,jiDoesMagicBERT2021,liICDCodingClinical2020,liuEffectiveConvolutionalAttention2021,moonsComparisonDeepLearning2020,pascualBERTbasedAutomaticICD2021,tengReviewDeepNeural2022,tengExplainablePredictionMedical2020, vuLabelAttentionModel2020, venkateshAutomatingOverburdenedClinical2023}.
 In the error analysis, we aimed to validate or falsify these assumptions.
@@
 Furthermore, for AMC to become a viable solution for ICD-10, future research should focus on improving performance on rare codes while, in the shorter term, developing methods to detect codes that are too challenging for automated coding and, therefore, should be coded manually.
-Finally, while PLM-ICD outperforms the other models in this paper, the improvements are limited compared to the effect of pre-training in other domains \cite{mohamed_selfsupervised_2022, linPretrainedTransformersText2021,baevski_wav2vec_2020,devlin_bert_2018,dosovitskiy_image_2021}.
-Notably, there have been several unsuccessful attempts at using pre-trained transformers for medical coding \cite{jiDoesMagicBERT2021,gaoLimitationsTransformersClinical2021,michalopoulosICDBigBirdContextualEmbedding2022,pascualBERTbasedAutomaticICD2021,zhangBERTXMLLargeScale2020}.
+Finally, while PLM-ICD outperforms the other models in this paper, the improvements are limited compared to the effect of pre-training in other domains \parencite{mohamed_selfsupervised_2022, linPretrainedTransformersText2021,baevski_wav2vec_2020,devlin_bert_2018,dosovitskiy_image_2021}.
+Notably, there have been several unsuccessful attempts at using pre-trained transformers for medical coding \parencite{jiDoesMagicBERT2021,gaoLimitationsTransformersClinical2021,michalopoulosICDBigBirdContextualEmbedding2022,pascualBERTbasedAutomaticICD2021,zhangBERTXMLLargeScale2020}.
 In future work, we want to investigate why pre-trained transformers underperform in medical coding.
@@
 For instance, since MIMIC-III and IV contain data from the emergency department and ICU of a single hospital, the findings in this paper may not generalize to other departments or hospitals. 
-For instance, discharge summaries from outpatient care are often easier to code than summaries from inpatient care as they are shorter with fewer codes per document \cite{zhangBERTXMLLargeScale2020, liuEffectiveConvolutionalAttention2021, tsengAdministrativeCostsAssociated2018}.
+For instance, discharge summaries from outpatient care are often easier to code than summaries from inpatient care as they are shorter with fewer codes per document \parencite{zhangBERTXMLLargeScale2020, liuEffectiveConvolutionalAttention2021, tsengAdministrativeCostsAssociated2018}.
 
 The medical code labeling of MIMIC is used as a gold standard in this paper.
-However, medical coding is error-prone, and, in many cases, deciding between certain codes can be a subjective matter \cite{nouraeiAuditNatureImpact2013, lloydPhysicianCodingErrors1985}.
-\citet{burnsSystematicReviewDischarge2012} systematically reviewed studies assessing the accuracy of human medical coders and found an overall median accuracy of 83.2\% (IQR: 67.3–92.1\%).
-\citet{searleExperimentalEvaluationDevelopment2020} investigated the quality of the human annotations in MIMIC-III and concluded that 35\% of the common codes were under-coded.
+However, medical coding is error-prone, and, in many cases, deciding between certain codes can be a subjective matter \parencite{nouraeiAuditNatureImpact2013, lloydPhysicianCodingErrors1985}.
+\textcite{burnsSystematicReviewDischarge2012} systematically reviewed studies assessing the accuracy of human medical coders and found an overall median accuracy of 83.2\% (IQR: 67.3–92.1\%).
+\textcite{searleExperimentalEvaluationDevelopment2020} investigated the quality of the human annotations in MIMIC-III and concluded that 35\% of the common codes were under-coded.
 Such errors and subjectivity in manual medical coding make model training and evaluation challenging and suggests that additional evaluation methods using, e.g., a human-in-the-loop, could be useful to increase the reliability of results.
@@
 We observed a significant performance improvement for all models, with those previously reported as low-performing improving the most.
-We reported the first results of current state-of-the-art models on the newly released MIMIC-IV dataset \cite{johnsonMIMICIVFreelyAccessible2023, goldbergerPhysioBankPhysioToolkitPhysioNet2000} and provided splits for the \textit{ICD-9} and \textit{ICD-10} coded subsets using the same method as for MIMIC-III \textit{clean}.
+We reported the first results of current state-of-the-art models on the newly released MIMIC-IV dataset \parencite{johnsonMIMICIVFreelyAccessible2023, goldbergerPhysioBankPhysioToolkitPhysioNet2000} and provided splits for the \textit{ICD-9} and \textit{ICD-10} coded subsets using the same method as for MIMIC-III \textit{clean}.
 Through error analysis, we provided empirical evidence for multiple model weaknesses. Specifically, models underperform severely on rare codes 

### 2023-09-29 chapters/paper_benchmarking.tex
Updates to citation commands and footnotes

@@
 \section{Introduction}
-After the introduction of the variational autoencoder (VAE,  \textcite{kingma_autoencoding_2014, rezende_stochastic_2014}) quickly came two temporal extensions for modeling speech data \cite{chung_recurrent_2015, fraccaro_sequential_2016}.
-Since then, temporal LVMs have undergone little development compared to their counterparts in the image domain, where LVMs recently showed superior performance to autoregressive models such as PixelCNN \cite{oord_pixel_2016, oord_conditional_2016, salimans_pixelcnn_2017}.
-The improvements in the image domain have been driven mainly by top-down inference models and deeper latent hierarchies \cite{sonderby_ladder_2016, maaloe_biva_2019, vahdat_nvae_2020, child_very_2021,sinha_consistency_2021,kingma_variational_2021}.
-In speech modeling however, autoregressive models such as the WaveNet remain state-of-the-art \cite{oord_wavenet_2016}.
+After the introduction of the variational autoencoder (VAE,  \textcite{kingma_autoencoding_2014, rezende_stochastic_2014}) quickly came two temporal extensions for modeling speech data \parencite{chung_recurrent_2015, fraccaro_sequential_2016}.
+Since then, temporal LVMs have undergone little development compared to their counterparts in the image domain, where LVMs recently showed superior performance to autoregressive models such as PixelCNN \parencite{oord_pixel_2016, oord_conditional_2016, salimans_pixelcnn_2017}.
+The improvements in the image domain have been driven mainly by top-down inference models and deeper latent hierarchies \parencite{sonderby_ladder_2016, maaloe_biva_2019, vahdat_nvae_2020, child_very_2021,sinha_consistency_2021,kingma_variational_2021}.
+In speech modeling however, autoregressive models such as the WaveNet remain state-of-the-art \parencite{oord_wavenet_2016}.
 
 To compare and develop LVMs for speech, we need good benchmarks similar to those in the image domain.
-Image benchmarks commonly compare likelihood scores, but research in the speech domain often omits reporting a likelihood \cite{oord_wavenet_2016, hsu_unsupervised_2017, oord_neural_2018} or report likelihoods that are incomparable due to subtle differences in the assumed data distribution \cite{chung_recurrent_2015, fraccaro_sequential_2016, hsu_unsupervised_2017, aksan_stcn_2019}.
+Image benchmarks commonly compare likelihood scores, but research in the speech domain often omits reporting a likelihood \parencite{oord_wavenet_2016, hsu_unsupervised_2017, oord_neural_2018} or report likelihoods that are incomparable due to subtle differences in the assumed data distribution \parencite{chung_recurrent_2015, fraccaro_sequential_2016, hsu_unsupervised_2017, aksan_stcn_2019}.
 Without a proper standard, it is difficult to compare explicit likelihood models for speech and develop them in an informed manner.
@@
     \item [(I)] State-of-the-art LVMs achieve likelihoods that are inferior to WaveNet at high temporal resolution but are superior at lower resolutions.
-Interestingly, we find that a standard LSTM \cite{hochreiter_long_1997} almost matches the likelihood of WaveNet.
+Interestingly, we find that a standard LSTM \parencite{hochreiter_long_1997} almost matches the likelihood of WaveNet.
     \item [(II)] LVMs with powerful autoregressive decoders achieve better likelihoods than the non-autoregressive LVM.
     \item[(III)] The expressiveness of LVMs for speech increases with a deeper hierarchy of stochastic latent variables, similar to conclusions within image modeling.
-% \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
+% \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
     \item[(IV)] LVMs learn rich representations that are as good or better than Mel spectrograms for phoneme recognition also when using only 10 minutes of labeled data.
@@
 Approximate posteriors have various applications e.g.
-semi-supervised learning \cite{kingma_semi-supervised_2014} and anomaly detection \cite{havtorn_hierarchical_2021}.
+semi-supervised learning \parencite{kingma_semi-supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.
 
 In recent years, several complementary methods have been proposed to improve the expressiveness of VAEs.
-These include building more expressive priors via methods such as normalizing flows \cite{rezende_variational_2015} and building a deeper hierarchy of stochastic latent variables such as the Ladder VAE \cite{sonderby_ladder_2016}.
-In this research, we focus on the latter due to the recent breakthroughs in image modeling using VAEs without costly autoregressive dependencies on the observed variable \cite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
+These include building more expressive priors via methods such as normalizing flows \parencite{rezende_variational_2015} and building a deeper hierarchy of stochastic latent variables such as the Ladder VAE \parencite{sonderby_ladder_2016}.
+In this research, we focus on the latter due to the recent breakthroughs in image modeling using VAEs without costly autoregressive dependencies on the observed variable \parencite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
 
 Several works have applied LVMs to speech.
-Among the first contributions were the VRNN \cite{chung_recurrent_2015} and SRNN \cite{fraccaro_sequential_2016} which can be seen as conditional VAEs per timestep.
-Other recent LVMs include the FH-VAE \cite{hsu_unsupervised_2017}, which leverages an additional latent variable to capture global features, and Z-forcing \cite{goyal_zforcing_2017}, which resembles the SRNN but includes an auxiliary task in the latent space to increase its utilization.
-The VQ-VAE \cite{oord_neural_2018} is a hybrid between an LVM and an autoregressive model which uses a quantized latent space to improve the quality of generated samples.
-The Stochastic WaveNet \cite{lai_stochastic_2018} and STCN \cite{aksan_stcn_2019} use WaveNet encoder and decoders and temporally independent latent variables.
+Among the first contributions were the VRNN \parencite{chung_recurrent_2015} and SRNN \parencite{fraccaro_sequential_2016} which can be seen as conditional VAEs per timestep.
+Other recent LVMs include the FH-VAE \parencite{hsu_unsupervised_2017}, which leverages an additional latent variable to capture global features, and Z-forcing \parencite{goyal_zforcing_2017}, which resembles the SRNN but includes an auxiliary task in the latent space to increase its utilization.
+The VQ-VAE \parencite{oord_neural_2018} is a hybrid between an LVM and an autoregressive model which uses a quantized latent space to improve the quality of generated samples.
+The Stochastic WaveNet \parencite{lai_stochastic_2018} and STCN \parencite{aksan_stcn_2019} use WaveNet encoder and decoders and temporally independent latent variables.
 
 In this paper, we focus on the VRNN, SRNN and STCN.
-We exclude the Stochastic WaveNet as it is similar to STCN and achieves inferior likelihoods \cite{aksan_stcn_2019}.
+We exclude the Stochastic WaveNet as it is similar to STCN and achieves inferior likelihoods \parencite{aksan_stcn_2019}.
 The FH-VAE, with disjoint latent variables and discriminative objective, Z-forcing, with an auxiliary task, and the VQ-VAE, with a quantized latent space and autoregressive prior fitted after training, all introduce significant changes to the original VAE framework and are also not included here.
@@
 Inspired by recent progress in the image domain, we therefore formulate and benchmark a novel temporal LVM which does not rely on an autoregressive decoder.
-We do so by adapting the hierarchical Clockwork Variational Autoencoder \cite{saxena_clockwork_2021}, originally proposed for video generation, to speech.
+We do so by adapting the hierarchical Clockwork Variational Autoencoder \parencite{saxena_clockwork_2021}, originally proposed for video generation, to speech.
 
 \subsection{Sequential deep latent variable models}
-The selected models are all sequential deep latent variable models trained with variational inference and the reparameterization trick \cite{kingma_autoencoding_2014}.
-% The VRNN, SRNN, STCN and CW-VAE are all autoencoders trained with variational inference and the reparameterization trick \cite{kingma_autoencoding_2014}. 
+The selected models are all sequential deep latent variable models trained with variational inference and the reparameterization trick \parencite{kingma_autoencoding_2014}.
+% The VRNN, SRNN, STCN and CW-VAE are all autoencoders trained with variational inference and the reparameterization trick \parencite{kingma_autoencoding_2014}. 
 They take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$. 
-% The VRNN, SRNN, STCN and CW-VAE are all VAEs \cite{kingma_autoencoding_2014} that take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$.
+% The VRNN, SRNN, STCN and CW-VAE are all VAEs \parencite{kingma_autoencoding_2014} that take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$.
 We let $\xb_{1:T}$ refer to the observed variable or a downsampled version of it. We will sometimes use $\xb$ to refer to the sequence $\xb_{1:T}$ when it is not ambiguous.
@@
 \subsection{Variational recurrent neural network ({\small VRNN})}
-The VRNN \cite{chung_recurrent_2015} is essentially a VAE per timestep.
+The VRNN \parencite{chung_recurrent_2015} is essentially a VAE per timestep.
 At timestep $t$, the VAE is conditioned on the hidden state of a recurrent neural network (RNN), $\db_{t}\in\mathbb{R}^{D_d}$, with state transition $\db_{t} = f([\xb_{t-1}, \zb_{t-1}], \db_{t-1})$ where $[\cdot, \cdot]$ denotes concatenation.
@@
 Since the decoder is dependent on $\db_t$, the transition function $f$ allows the VRNN to learn to ignore parts of or the entire latent variable and establish a purely deterministic transition from $\xb_{t-1}$ to $\db_t$ (\cref{fig: autoregressive vrnn srnn cwvae 1L graphs}).
-This failure mode is commonly referred to as posterior collapse and is a well-known weakness of VAEs with powerful decoders \cite{bowman_generating_2016, sonderby_ladder_2016}.
+This failure mode is commonly referred to as posterior collapse and is a well-known weakness of VAEs with powerful decoders \parencite{bowman_generating_2016, sonderby_ladder_2016}.
 
 \subsection{Stochastic recurrent neural network ({\small SRNN})}
-The SRNN \cite{fraccaro_sequential_2016} is similar to the VRNN but differs by separating the stochastic latent variables from the deterministic representations (\cref{fig: autoregressive vrnn srnn cwvae 1L graphs}).
+The SRNN \parencite{fraccaro_sequential_2016} is similar to the VRNN but differs by separating the stochastic latent variables from the deterministic representations (\cref{fig: autoregressive vrnn srnn cwvae 1L graphs}).
 That is, the GRU state transition is independent of $\zb_{1:T}$ such that $\db_{t} = f(\xb_{t-1}, \db_{t-1})$.
@@
 By inferring $\zb_t$ conditioned on the full sequence $\xb_{1:T}$, the SRNN performs smoothing.
-This has been noted to better approximate the true posterior of $\zb_t$ which can be shown to depend on the full observed sequence \cite{bayer_mind_2021}.
+This has been noted to better approximate the true posterior of $\zb_t$ which can be shown to depend on the full observed sequence \parencite{bayer_mind_2021}.
 Comparatively, the VRNN performs filtering.
@@
 \subsection{Stochastic temporal convolutional network ({\small STCN})}
-The STCN \cite{aksan_stcn_2019} is a hierarchical latent variable model with an autoregressive generative model based on WaveNet \cite{oord_wavenet_2016}.
+The STCN \parencite{aksan_stcn_2019} is a hierarchical latent variable model with an autoregressive generative model based on WaveNet \parencite{oord_wavenet_2016}.
 Contrary to VRNN and SRNN, the latent variables have no transition functions connecting them over time.
@@
 where $\zb_t^{(L+1)}:=\emptyset$ for notational convenience. 
-The deterministic encoding is $\db_t^{(l)} = h(\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}})$ where $h$ is the encoder and $\db_t^{(l)}$ is extracted from the $l$'th layer similar to a Ladder VAE \cite{sonderby_ladder_2016}.
+The deterministic encoding is $\db_t^{(l)} = h(\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}})$ where $h$ is the encoder and $\db_t^{(l)}$ is extracted from the $l$'th layer similar to a Ladder VAE \parencite{sonderby_ladder_2016}.
 The approximate posterior is
@@
 \subsection{Clockwork variational autoencoder ({\small CW-VAE})}
-The CW-VAE \cite{saxena_clockwork_2021} is a hierarchical latent variable model recently introduced for video generation.
+The CW-VAE \parencite{saxena_clockwork_2021} is a hierarchical latent variable model recently introduced for video generation.
 As illustrated in figures~\cref{fig: autoregressive vrnn srnn cwvae 1L graphs} and \cref{fig: clockwork vae inference generative models 2 layered}, it is autoregressive in the latent space but not in the observed space, contrary to the VRNN, SRNN and STCN.
@@
 The signals are sampled with some bit depth $b$ which defines the range of attainable values, $\xb\in\{0,1,\dots,2^b-1\}$.
-The bit depth typically used in audio and image samplers is between $\SI{8}{}$ and $\SI{32}{bit}$ with $\SI{8}{bit}$ and $\SI{16}{bit}$ being the most common in the literature (MNIST \cite{lecun_gradientbased_1998}, CelebA \cite{liu_deep_2015}, CIFAR10 \cite{krizhevsky_learning_2009}, TIMIT \cite{garofolo_timit_1993}, LibriSpeech \cite{panayotov_librispeech_2015}).
+The bit depth typically used in audio and image samplers is between $\SI{8}{}$ and $\SI{32}{bit}$ with $\SI{8}{bit}$ and $\SI{16}{bit}$ being the most common in the literature (MNIST \parencite{lecun_gradientbased_1998}, CelebA \parencite{liu_deep_2015}, CIFAR10 \parencite{krizhevsky_learning_2009}, TIMIT \parencite{garofolo_timit_1993}, LibriSpeech \parencite{panayotov_librispeech_2015}).
 
-In the image domain, the discrete nature of the data is usually modeled in one of two ways; either by using discrete distributions \cite{salimans_pixelcnn_2017, maaloe_biva_2019, child_very_2021} or by dequantizing the data and using a continuous distribution \cite{dinh_nice_2015, sonderby_ladder_2016, ho_flow_2019}, which yields a lower bound on the discrete distribution likelihood \cite{theis_note_2016}.
-In the speech domain, however, the output distribution is often taken to be a continuous Gaussian \cite{hsu_unsupervised_2017,lai_stochastic_2018,zhu_s3vae_2020} which was also originally done in VRNN, SRNN and STCN.
-This choice generally results in an ill-posed problem with a likelihood that is unbounded from above unless the variance is lower bounded \cite{mattei_leveraging_2018}.
+In the image domain, the discrete nature of the data is usually modeled in one of two ways; either by using discrete distributions \parencite{salimans_pixelcnn_2017, maaloe_biva_2019, child_very_2021} or by dequantizing the data and using a continuous distribution \parencite{dinh_nice_2015, sonderby_ladder_2016, ho_flow_2019}, which yields a lower bound on the discrete distribution likelihood \parencite{theis_note_2016}.
+In the speech domain, however, the output distribution is often taken to be a continuous Gaussian \parencite{hsu_unsupervised_2017,lai_stochastic_2018,zhu_s3vae_2020} which was also originally done in VRNN, SRNN and STCN.
+This choice generally results in an ill-posed problem with a likelihood that is unbounded from above unless the variance is lower bounded \parencite{mattei_leveraging_2018}.
 As a result, \emph{reported likelihoods can be sensitive to hyperparameter settings and be hard to compare}.
@@
 Since $\xb$ becomes approximately continuous as the bit depth $b$ becomes large and the range of possible values becomes small, this alleviates the issue.
-However, commonly used datasets with bit-depths of $\SI{16}{}$ still result in a discretization gap between values that remains much larger than the gap between the almost continuous 32 bit floating point numbers which reinforces the problem \cite{bishop_pattern_2006}.
+However, commonly used datasets with bit-depths of $\SI{16}{}$ still result in a discretization gap between values that remains much larger than the gap between the almost continuous 32 bit floating point numbers which reinforces the problem \parencite{bishop_pattern_2006}.
 
 In this work, we therefore benchmark models using a discretized mixture of logistics (DMoL) as output distribution.
-The DMoL was introduced for image modeling with autoregressive models \cite{salimans_pixelcnn_2017} but has become standard in other generative models \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
-It was recently applied to autoregressive speech modeling of raw waveforms \cite{oord_parallel_2018}. 
+The DMoL was introduced for image modeling with autoregressive models \parencite{salimans_pixelcnn_2017} but has become standard in other generative models \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
+It was recently applied to autoregressive speech modeling of raw waveforms \parencite{oord_parallel_2018}. 
 As opposed to e.g.\@ a categorical distribution, the DMoL induces ordinality on the observed space such that values that are numerically close are also close in a probabilistic sense.
@@
 \paragraph{Data}
-We train models on TIMIT \cite{garofolo_timit_1993}, LibriSpeech \cite{panayotov_librispeech_2015} and LibriLight \cite{kahn_libri-light_2020}.
+We train models on TIMIT \parencite{garofolo_timit_1993}, LibriSpeech \parencite{panayotov_librispeech_2015} and LibriLight \parencite{kahn_libri-light_2020}.
 For TIMIT, we randomly sample 5\% of the training split for validation.
@@
 We report likelihoods in units of bits per frame (bpf) as this yields a more interpretable and comparable likelihood than total likelihood in nats.
-It also has direct connections with information theory and compression \cite{shannon_mathematical_1948, townsend_practical_2019}.
+It also has direct connections with information theory and compression \parencite{shannon_mathematical_1948, townsend_practical_2019}.
 In units of bits per frame, lower is better.
@@
 We use a DMoL with 10 components for the output distribution of all models and model all datasets at their full bit depth of $\SI{16}{bits}$.
-We train and evaluate models on stacked waveforms similar to previous work \cite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019} with stack sizes of $s=1$, $s=64$ and $s=256$.
+We train and evaluate models on stacked waveforms similar to previous work \parencite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019} with stack sizes of $s=1$, $s=64$ and $s=256$.
 Hence, every model input $\xb_t$ is composed of $\tilde{x}_{t:t+s}$ where $\tilde{x}$ are waveform frames.
@@
 We configure WaveNet as in the original paper using ten layers per block and five blocks. We use $D_c=96$ channels. 
-We also train an LSTM model \cite{hochreiter_long_1997} which has fully connected encoders and decoders, as the VRNN and SRNN, but a deterministic recurrent cell and much fewer parameters than WaveNet.
+We also train an LSTM model \parencite{hochreiter_long_1997} which has fully connected encoders and decoders, as the VRNN and SRNN, but a deterministic recurrent cell and much fewer parameters than WaveNet.
 We report on LSTM models with $D_d=256$ hidden units.
@@
 
-The STCN is used in the ``dense" configuration of the original work \cite{aksan_stcn_2019}.
+The STCN is used in the ``dense" configuration of the original work \parencite{aksan_stcn_2019}.
 It uses 256 convolutional channels and $L=5$ latent variables of dimensions 16, 32, 64, 128, 256 from the top down.
@@
 Specifically, we evaluate an uninformed per-frame discrete uniform distribution and a two-component DMoL fitted to the training set to benchmark worst case performance.
-We also report the likelihood achieved by the lossless compression algorithm, FLAC \cite{coalson_free_2019} which establishes a notion of good performance, although not a strict best case.
+We also report the likelihood achieved by the lossless compression algorithm, FLAC \parencite{coalson_free_2019} which establishes a notion of good performance, although not a strict best case.
 We report FLAC on linear PCM since it was designed for this encoding.
@@
 The STCN yields the best results at both $s=64$ and $s=256$. 
-For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables, such as WaveNet and the LSTM, to the ability of LVMs to model intra-step correlations \cite{lai_re-examination_2019}.
+For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables, such as WaveNet and the LSTM, to the ability of LVMs to model intra-step correlations \parencite{lai_re-examination_2019}.
 
@@
 A connection can be made between the model likelihoods and the compression rates of audio compression algorithms. 
-Lossy compression algorithms, such as MP3, exploit the dynamic range of human hearing to achieve 70-95\% reduction in bit rate \cite{brandenburg_mpeg-2_1998} while lossless compression algorithms, such as FLAC, achieve 50-70\% reduction \cite{coalson_free_2019} independent of audio content.
+Lossy compression algorithms, such as MP3, exploit the dynamic range of human hearing to achieve 70-95\% reduction in bit rate \parencite{brandenburg_mpeg-2_1998} while lossless compression algorithms, such as FLAC, achieve 50-70\% reduction \parencite{coalson_free_2019} independent of audio content.
 % On LibriSpeech, FLAC achieves an overall compression to about 58\% of the original bit rate.
@@
 \section{Phoneme recognition}\label{sec: phoneme recognition}
-Although the likelihood is a practical metric for model comparison and selection, a high likelihood does not guarantee that a model has learned useful representations \cite{huszar_is_2017}. 
+Although the likelihood is a practical metric for model comparison and selection, a high likelihood does not guarantee that a model has learned useful representations \parencite{huszar_is_2017}. 
 For speech data, we would expect models to learn features related to phonetics which would make them useful for tasks such as automatic speech recognition (ASR). 
 The Mel spectrogram is a well-established representation of audio designed for speech recognition and is predefined rather than learned.
-% \cite{??}.
+% \parencite{??}.
 Basic Correlates of the Auditory Stimulus, https://psycnet.apa.org/record/1951-07758-017, modern mel scale https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=1170013&casa_token=UIyQaj3tBRAAAAAA:GtttZedngBkaxVumjnXctc5zS50fptexPMiWhr-7gDrmIVq3JqJPnGCWYedoymb0X5JTGElmHf8&tag=1,
@@
 The ASR model is a three-layered bidirectional LSTM with 256 hidden units.
-It is trained with the connectionist temporal classification (CTC) loss \cite{graves_connectionist_2006} which lets it jointly learn to align and classify without using label alignments.
+It is trained with the connectionist temporal classification (CTC) loss \parencite{graves_connectionist_2006} which lets it jointly learn to align and classify without using label alignments.
 We pre-train all unsupervised models at $s=64$ on the full TIMIT training dataset excluding the validation data (3.7h) as in \cref{tab: likelihoods timit}. 
@@
 This approximately marginalizes variation during the phoneme.
-We use linear discriminant analysis (LDA) \cite{fisher_use_1936} to obtain a low-dimensional linear subspace of the latent space.
+We use linear discriminant analysis (LDA) \parencite{fisher_use_1936} to obtain a low-dimensional linear subspace of the latent space.
 We visualize the resulting representations colored according to test set phoneme classes in \cref{fig: latent space phoneme and knn}.

### 2023-09-29 chapters/paper_benchmarking_workshop.tex
Updates to citation commands and footnotes

@@
 \section{Introduction}
-Since their introduction, temporal latent variable models (LVMs) for speech \cite{chung_recurrent_2015, fraccaro_sequential_2016} based on the variational autoencoder (VAE,  \textcite{kingma_autoencoding_2014, rezende_stochastic_2014}) have seen little development compared to their image domain counterparts.
-While LVMs now achieve superior likelihoods on images compared to deterministic models \cite{child_very_2021,sinha_consistency_2021,kingma_variational_2021}, they are still inferior to deterministic models such as WaveNet on speech \cite{oord_wavenet_2016}.
-Development in the image domain has been driven by established likelihood benchmarks, but in the speech domain, likelihoods are often not reported \cite{oord_wavenet_2016, hsu_unsupervised_2017, oord_neural_2018} or are incomparable due to subtle differences in the chosen data distributions \cite{chung_recurrent_2015, fraccaro_sequential_2016, hsu_unsupervised_2017, aksan_stcn_2019}.
+Since their introduction, temporal latent variable models (LVMs) for speech \parencite{chung_recurrent_2015, fraccaro_sequential_2016} based on the variational autoencoder (VAE,  \textcite{kingma_autoencoding_2014, rezende_stochastic_2014}) have seen little development compared to their image domain counterparts.
+While LVMs now achieve superior likelihoods on images compared to deterministic models \parencite{child_very_2021,sinha_consistency_2021,kingma_variational_2021}, they are still inferior to deterministic models such as WaveNet on speech \parencite{oord_wavenet_2016}.
+Development in the image domain has been driven by established likelihood benchmarks, but in the speech domain, likelihoods are often not reported \parencite{oord_wavenet_2016, hsu_unsupervised_2017, oord_neural_2018} or are incomparable due to subtle differences in the chosen data distributions \parencite{chung_recurrent_2015, fraccaro_sequential_2016, hsu_unsupervised_2017, aksan_stcn_2019}.
 This makes it hard to develop explicit likelihood models for speech.
@@
 This makes them useful beyond generation for e.g.
-semi-supervised learning \cite{kingma_semi-supervised_2014} and anomaly detection \cite{havtorn_hierarchical_2021}.
-In this paper, we focus on the VRNN \cite{chung_recurrent_2015}, SRNN \cite{fraccaro_sequential_2016} and STCN \cite{aksan_stcn_2019} since they are similar to the original VAE framework. 
+semi-supervised learning \parencite{kingma_semi-supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.
+In this paper, we focus on the VRNN \parencite{chung_recurrent_2015}, SRNN \parencite{fraccaro_sequential_2016} and STCN \parencite{aksan_stcn_2019} since they are similar to the original VAE framework.
 We exclude some related work from the benchmark.
-Specifically, the FH-VAE \cite{hsu_unsupervised_2017} adds a global latent variable but we exclude it due to it's segmentation of the training data.
-We also exclude Z-forcing \cite{goyal_zforcing_2017} which resembles the SRNN, but deviates from maximum likelihood training by using an auxiliary loss.
-We exclude the VQ-VAE \cite{oord_neural_2018} which is a hybrid LVM and autoregressive model, since it uses a quantized latent space.
-Finally, we exclude the Stochastic WaveNet \cite{lai_stochastic_2018} which is similar, but inferior, to the STCN \cite{aksan_stcn_2019}.
+Specifically, the FH-VAE \parencite{hsu_unsupervised_2017} adds a global latent variable but we exclude it due to it's segmentation of the training data.
+We also exclude Z-forcing \parencite{goyal_zforcing_2017} which resembles the SRNN, but deviates from maximum likelihood training by using an auxiliary loss.
+We exclude the VQ-VAE \parencite{oord_neural_2018} which is a hybrid LVM and autoregressive model, since it uses a quantized latent space.
+Finally, we exclude the Stochastic WaveNet \parencite{lai_stochastic_2018} which is similar, but inferior, to the STCN \parencite{aksan_stcn_2019}.
 
@@
 We therefore formulate and benchmark a novel temporal LVM which does not use an autoregressive decoder.
-We do so by adapting the hierarchical Clockwork Variational Autoencoder \cite{saxena_clockwork_2021}, originally proposed for video generation, to speech.
+We do so by adapting the hierarchical Clockwork Variational Autoencoder \parencite{saxena_clockwork_2021}, originally proposed for video generation, to speech.
 Before presenting the results, we provide a brief survey of existing LVMs for speech in a coherent notation.
@@
 \paragraph{Sequential deep latent variable models}
-The models we consider are all sequential deep latent variable models trained with variational inference and the reparameterization trick \cite{kingma_autoencoding_2014}.
+The models we consider are all sequential deep latent variable models trained with variational inference and the reparameterization trick \parencite{kingma_autoencoding_2014}.
 The input is a variable-length sequence $\vx = \vx_{1:T} = \left(\vx_1, \vx_2, \dots, \vx_{T}\right)$ with $\vx_t\in\mathbb{R}^{D_x}$. 
@@
 \paragraph{Variational recurrent neural network (VRNN)}
-The VRNN \cite{chung_recurrent_2015} is essentially a VAE per timestep.
+The VRNN \parencite{chung_recurrent_2015} is essentially a VAE per timestep.
 As seen in \cref{fig: autoregressive vrnn srnn cwvae 1L graphs}, it is conditioned on the hidden state of a Gated Recurrent Unit (GRU, \textcite{cho_properties_2014}) with state transition $\vd_{t} = f([\vx_{t-1}, \vz_{t-1}], \vd_{t-1})\in\mathbb{R}^{D_d}$ where $[\cdot, \cdot]$ denotes concatenation.
@@
 \paragraph{Stochastic recurrent neural network (SRNN)}
-The SRNN \cite{fraccaro_sequential_2016} is similar to the VRNN and its joint can be written as in \eqref{eq: vrnn joint and inference}.
+The SRNN \parencite{fraccaro_sequential_2016} is similar to the VRNN and its joint can be written as in \eqref{eq: vrnn joint and inference}.
 Contrary to the VRNN, the SRNN has GRU state transitions that are independent of $\vz_{1:T}$ such that $\vd_{t} = f(\vx_{t-1}, \vd_{t-1})$.
@@
 Then $\vz_t$ is inferred from $\va_t$ and $\vz_{t-1}$.
-This better approximates the true posterior which can be shown to depend on the full observed sequence \cite{bayer_mind_2021}.
+This better approximates the true posterior which can be shown to depend on the full observed sequence \parencite{bayer_mind_2021}.
 
@@
 Contrary to VRNN and SRNN, the latent variables of the STCN are conditionally independent given $\vx_{1:T}$ since there are no transition functions connecting them over time.
-Instead, a latent $\vz_t^{(l)}$ at layer $l$ is conditioned on the latent variable above it $\vz^{(l+1)}_t$ and a window of $\vx_{1:T}$ defined by $\mathcal{R}_t^{(l)} = \cbra{t-r_l+1,\dots,t}$ via a WaveNet encoder with receptive field $r_l$ at layer $l$ \cite{oord_wavenet_2016}.
+Instead, a latent $\vz_t^{(l)}$ at layer $l$ is conditioned on the latent variable above it $\vz^{(l+1)}_t$ and a window of $\vx_{1:T}$ defined by $\mathcal{R}_t^{(l)} = \cbra{t-r_l+1,\dots,t}$ via a WaveNet encoder with receptive field $r_l$ at layer $l$ \parencite{oord_wavenet_2016}.
 The joint and variational posterior are,
@@
 \end{equation}
-where $\vz_t^{(L+1)}:=\emptyset$ and $\vz=\vz_{1:T}^{(1:L)}$ for notational convenience and inference is done top-down \cite{sonderby_ladder_2016}.
+where $\vz_t^{(L+1)}:=\emptyset$ and $\vz=\vz_{1:T}^{(1:L)}$ for notational convenience and inference is done top-down \parencite{sonderby_ladder_2016}.
 The observation model $p(\vx_t|\vz^{(1:L)}_{{\scriptscriptstyle \mathcal{R}}_{t}^{(1)}})$ is also parameterized by a WaveNet.
@@
 \paragraph{Clockwork variational autoencoder (CW-VAE)}
-The CW-VAE \cite{saxena_clockwork_2021} is a hierarchical LVM originally introduced for video generation.
+The CW-VAE \parencite{saxena_clockwork_2021} is a hierarchical LVM originally introduced for video generation.
 As seen in figure~\cref{fig: autoregressive vrnn srnn cwvae 1L graphs}, it is autoregressive in the latent space but not in the observed space, contrary to the VRNN, SRNN and STCN.
@@
 
-\paragraph{Output distribution} Recent work on LVMs for speech modeling, including those considered here, often uses a Gaussian output distribution \cite{chung_recurrent_2015, fraccaro_sequential_2016,hsu_unsupervised_2017,lai_stochastic_2018,aksan_stcn_2019,zhu_s3vae_2020}.
+\paragraph{Output distribution} Recent work on LVMs for speech modeling, including those considered here, often uses a Gaussian output distribution \parencite{chung_recurrent_2015, fraccaro_sequential_2016,hsu_unsupervised_2017,lai_stochastic_2018,aksan_stcn_2019,zhu_s3vae_2020}.
 Since audio is a naturally continuous signal, this may seem like an appropriate modeling choice.
 However, common audio datasets are sampled at bit-depths of $\SI{16}{bit}$ (TIMIT, \citealt{garofolo_timit_1993}; LibriSpeech, \citealt{panayotov_librispeech_2015}).
-This results in a quantization gap between unique values of $1\times10^{-5}$ which increases the risk of an ill-posed problem by a likelihood that is unbounded from above unless the variance is lower bounded \cite{mattei_leveraging_2018}.
+This results in a quantization gap between unique values of $1\times10^{-5}$ which increases the risk of an ill-posed problem by a likelihood that is unbounded from above unless the variance is lower bounded \parencite{mattei_leveraging_2018}.
 As a result, \emph{reported likelihoods can be sensitive to hyperparameter settings and be hard to compare}.
@@
 In this work, we therefore benchmark models using a discretized mixture of logistics (DMoL) as output distribution.
-The DMoL was introduced for image modeling with autoregressive models \cite{salimans_pixelcnn_2017} but has become standard in other generative models \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
-Although continuous distributions can be used if the data is dequantized \cite{dinh_nice_2015,theis_note_2016,ho_flow_2019}, we do not consider this option here.
+The DMoL was introduced for image modeling with autoregressive models \parencite{salimans_pixelcnn_2017} but has become standard in other generative models \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
+Although continuous distributions can be used if the data is dequantized \parencite{dinh_nice_2015,theis_note_2016,ho_flow_2019}, we do not consider this option here.
 
@@
 \paragraph{Data}
-We train models on TIMIT \cite{garofolo_timit_1993} and randomly sample 5\% of the training split for validation. 
+We train models on TIMIT \parencite{garofolo_timit_1993} and randomly sample 5\% of the training split for validation. 
 Both input and target are $\mu$-law PCM standardized to $\bra{-1, 1}$. 
-We report on LibriSpeech \cite{panayotov_librispeech_2015} and linear PCM in appendix~\cref{app: additional likelihood results appendix}.
+We report on LibriSpeech \parencite{panayotov_librispeech_2015} and linear PCM in appendix~\cref{app: additional likelihood results appendix}.
 We further describe datasets in appendix~\cref{app: dataset details}.
@@
 \paragraph{Likelihood}
-We report likelihoods in units of bits per frame (bpf; lower is better) as this yields a more interpretable likelihood that has connections to information theory and compression \cite{shannon_mathematical_1948, townsend_practical_2019} compared to total likelihood in nats.
+We report likelihoods in units of bits per frame (bpf; lower is better) as this yields a more interpretable likelihood that has connections to information theory and compression \parencite{shannon_mathematical_1948, townsend_practical_2019} compared to total likelihood in nats.
 For LVMs, we report the one-sample ELBO.
@@
 The CW-VAE has $L=1$ or $2$ latent variables of dimension 96 equal to the number of convolution channels and we let $s$ refer to $s_1$ and set $s_2=8s_1$.
-We also train an LSTM baseline \cite{hochreiter_long_1997} configured similar to the VRNN.
+We also train an LSTM baseline \parencite{hochreiter_long_1997} configured similar to the VRNN.
 %
@@
 All LVMs use diagonal covariance Gaussian priors and posteriors.
-We train and evaluate models with waveforms stacked similar to previous work \cite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019} with stack sizes of $s=1$, $s=64$ and $s=256$.
+We train and evaluate models with waveforms stacked similar to previous work \parencite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019} with stack sizes of $s=1$, $s=64$ and $s=256$.
 Finally, we evaluate a per-frame discrete uniform distribution and a two-component DMoL fitted to the training set to estimate worst case performance. 
@@
 It is clear that deterministic autoregressive models are superior to LVMs at $s=1$, but inferior to them at $s=64$ and $256$.
-For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables to the ability of LVMs to model intra-step correlations \cite{lai_re-examination_2019}.
+For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables to the ability of LVMs to model intra-step correlations \parencite{lai_re-examination_2019}.
 It should be noted that at $s=1$, the STCN was numerically unstable and that the CW-VAE was computationally infeasible to train.
@@
 \paragraph{Phoneme recognition}
-Although the likelihood is a practical metric for model comparison, a high likelihood does not guarantee that a model has learned useful representations \cite{huszar_is_2017}.
+Although the likelihood is a practical metric for model comparison, a high likelihood does not guarantee that a model has learned useful representations \parencite{huszar_is_2017}.
 To assess the usefulness of the representations, we train a model to recognize phonemes on the TIMIT data.
@@
 The recognition model is a three-layered bidirectional LSTM with 256 hidden units.
-It is trained with the connectionist temporal classification (CTC) loss \cite{graves_connectionist_2006}.
+It is trained with the connectionist temporal classification (CTC) loss \parencite{graves_connectionist_2006}.
 We report the PER in \cref{tab: phoneme recognition (PER)}.
@@
 
-\subsubsection*{Acknowledgments}
+\section*{Acknowledegments}
 This research was partially funded by the Innovation Fund Denmark via the Industrial PhD Programme (grant no.\@ 0153-00167B).

### 2023-09-29 chapters/paper_hierarchical.tex
Updates to citation commands and footnotes

@@
 The reliability and safety of machine learning systems applied in the real-world is contingent on the ability to detect when an input is different from the training distribution. %, an anomaly.
-Supervised classifiers built as deep neural networks are well-known to misclassify such \textit{out-of-distribution} (OOD) inputs to known classes with high confidence \cite{goodfellow_explaining_2015, nguyen_deep_2015}.
-Several approaches have been suggested to equip deep classifiers with OOD detection capabilities \cite{hendrycks_baseline_2017, lakshminarayanan_simple_2017, hendrycks_deep_2019, devries_learning_2018}.
+Supervised classifiers built as deep neural networks are well-known to misclassify such \textit{out-of-distribution} (OOD) inputs to known classes with high confidence \parencite{goodfellow_explaining_2015, nguyen_deep_2015}.
+Several approaches have been suggested to equip deep classifiers with OOD detection capabilities \parencite{hendrycks_baseline_2017, lakshminarayanan_simple_2017, hendrycks_deep_2019, devries_learning_2018}.
 % However, such methods are inherently supervised and require in-distribution labels or examples of OOD data from an a priori known distribution which limit their applicability.
@@
 Unsupervised generative models that estimate an explicit likelihood should understand what it means to be in- and out-of-distribution without requiring labels or examples of OOD data.
-By directly modeling the training distribution, such models are expected to assign low likelihoods to OOD data as it originates from regions of little or no support under the learned density \cite{bishop_novelty_1994}.
-Recent advances in deep generative models \cite{kingma_autoencoding_2014, rezende_stochastic_2014, oord_pixel_2016, salimans_pixelcnn_2017, kingma_glow_2018} have enabled learning high quality generative models on complex data such as natural images, sequences including audio \cite{oord_wavenet_2016} and graphs \cite{kipf_variational_2016}.
-However, recent observations have brought into question the quality of the learned density estimates by showing that they often assign higher likelihoods to OOD data than to in-distribution data \cite{nalisnick_deep_2019, choi_waic_2019}.
+By directly modeling the training distribution, such models are expected to assign low likelihoods to OOD data as it originates from regions of little or no support under the learned density \parencite{bishop_novelty_1994}.
+Recent advances in deep generative models \parencite{kingma_autoencoding_2014, rezende_stochastic_2014, oord_pixel_2016, salimans_pixelcnn_2017, kingma_glow_2018} have enabled learning high quality generative models on complex data such as natural images, sequences including audio \parencite{oord_wavenet_2016} and graphs \parencite{kipf_variational_2016}.
+However, recent observations have brought into question the quality of the learned density estimates by showing that they often assign higher likelihoods to OOD data than to in-distribution data \parencite{nalisnick_deep_2019, choi_waic_2019}.
 Many complex data distributions can be explained to a large degree by low-level features, e.g. edges in images.
-However, such features do not explain high-level semantics of the data and may inhibit OOD detection \cite{ren_likelihood_2019, nalisnick_deep_2019}
+However, such features do not explain high-level semantics of the data and may inhibit OOD detection \parencite{ren_likelihood_2019, nalisnick_deep_2019}
 
@@
 The inability to detect out-of-distribution data with deep generative models is surprising.
-Before the advent of deep generative models, this was not considered a major issue for probabilistic models \cite{bishop_novelty_1994}.
+Before the advent of deep generative models, this was not considered a major issue for probabilistic models \parencite{bishop_novelty_1994}.
 Is the failure due to model pathologies or something different?
 
-Deep learning models are generally believed to form hierarchies of representations that range from low-level features to more conceptual ones related to semantics \cite{bengio_representation_2013}.
-This has also been observed within deep generative models \cite{maaloe_biva_2019, child_very_2021}.
-%Recent research \cite{maaloe_biva_2019, child_very_2021, vahdat_nvae_2020}, has shown that deep latent variable models are able to learn rich hierarchical representations with increasing levels of abstraction and global structure towards the top of the hierarchy.
+Deep learning models are generally believed to form hierarchies of representations that range from low-level features to more conceptual ones related to semantics \parencite{bengio_representation_2013}.
+This has also been observed within deep generative models \parencite{maaloe_biva_2019, child_very_2021}.
+%Recent research \parencite{maaloe_biva_2019, child_very_2021, vahdat_nvae_2020}, has shown that deep latent variable models are able to learn rich hierarchical representations with increasing levels of abstraction and global structure towards the top of the hierarchy.
 For image data there is a trend that the low-level features are quite similar across models (edge detectors, etc.).
-This raises the question to what extend such features are relevant when detecting OOD data, also suggested  by \cite{nalisnick_deep_2019} and examined for Glow and PixelCNN in \cite{schirrmeister_understanding_2020}.
+This raises the question to what extend such features are relevant when detecting OOD data, also suggested  by \parencite{nalisnick_deep_2019} and examined for Glow and PixelCNN in \parencite{schirrmeister_understanding_2020}.
 To investigate, we train two hierarchical VAEs (\cref{sec_paper_hierarchical:background-hie-VAE}) on FashionMNIST and MNIST, respectively, and compute the between-models correlation of the extracted features of in-distribution data and OOD data.
@@
     \vspace{0mm}
-    \caption[Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables.]{Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables  \cite{maaloe_biva_2019}.
+    \caption[Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables.]{Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables  \parencite{maaloe_biva_2019}.
     The higher the latent variable, the more the reconstructions fall into the mode of the learned distribution.
@@
 \cref{fig_hierarchical:reconstructions-fashionmnist} illustrates that MNIST data is surprisingly well reconstructed by a hierarchical VAE trained on FashionMNIST.
-Similar results have been found elsewhere \cite{xiao_likelihood_2020}.
+Similar results have been found elsewhere \parencite{xiao_likelihood_2020}.
 We repeat the previous experiment and replace inference distributions by their corresponding conditional prior, and now observe that reconstructions from higher latent layers become increasingly similar to the data on which the model was trained.
@@
 \subsection{Variational autoencoders}
-The variational autoencoder (VAE) \cite{kingma_autoencoding_2014, rezende_stochastic_2014} is a framework for constructing deep generative models defined by an observed variable $\mathbf{x}$ and a stochastic latent variable $\mathbf{z}$.
+The variational autoencoder (VAE) \parencite{kingma_autoencoding_2014, rezende_stochastic_2014} is a framework for constructing deep generative models defined by an observed variable $\mathbf{x}$ and a stochastic latent variable $\mathbf{z}$.
 Typically, a neural network with parameters $\thetab$ is chosen to parameterize the generative distribution $p_\thetab(\xb,\zb)=p_\thetab(\xb|\zb)p(\zb)$, where the prior $p(\zb)$ is commonly a standard Gaussian $\mathcal{N}(\0, \Ib)$.
@@
 The VAE is defined with a single latent variable which limits the ability to learn a high likelihood representation of complex input distributions, e.g.\ natural images.
-There exists a few complementary approaches to make the VAE more flexible: (i) model a more expressive variational distribution $q_\phib(\zb|\xb)$ or prior distribution $p_\thetab(\zb)$ \cite{rezende_variational_2015, kingma_improved_2016}, (ii) model a more expressive posterior distribution $p_\thetab(\xb|\zb)$ e.g.
-with a autoregressive decoder \cite{oord_conditional_2016} and (iii) learn a deeper hierarchy of latent variables \cite{burda_importance_2016, sonderby_ladder_2016}.
+There exists a few complementary approaches to make the VAE more flexible: (i) model a more expressive variational distribution $q_\phib(\zb|\xb)$ or prior distribution $p_\thetab(\zb)$ \parencite{rezende_variational_2015, kingma_improved_2016}, (ii) model a more expressive posterior distribution $p_\thetab(\xb|\zb)$ e.g.
+with a autoregressive decoder \parencite{oord_conditional_2016} and (iii) learn a deeper hierarchy of latent variables \parencite{burda_importance_2016, sonderby_ladder_2016}.
 Here, we focus on the latter.
@@
 The most common generative model is defined from the top down as $p_\thetab(\xb|\zb)=p(\xb|\zb_1)p_\thetab(\zb_1|\zb_2)\cdots p_\thetab(\zb_{L-1}|\zb_L)$.
-The inference model can then be defined in two ways respectively referred to as \textit{bottom-up} \cite{burda_importance_2016}
+The inference model can then be defined in two ways respectively referred to as \textit{bottom-up} \parencite{burda_importance_2016}
 \begin{equation}
@@
 \end{equation}
-and \textit{top-down} \cite{sonderby_ladder_2016}
+and \textit{top-down} \parencite{sonderby_ladder_2016}
 \begin{equation}
@@
 
-Until recently, hierarchical VAEs gave inferior likelihoods compared to state-of-the-art autoregressive \cite{ho_flow_2019} and flow-based models \cite{salimans_pixelcnn_2017}.
+Until recently, hierarchical VAEs gave inferior likelihoods compared to state-of-the-art autoregressive \parencite{ho_flow_2019} and flow-based models \parencite{salimans_pixelcnn_2017}.
 This was changed by \textcite{maaloe_biva_2019}, \textcite{vahdat_nvae_2020}, and \textcite{child_very_2021}, which introduced complementary methods to extend the number of latent variables to a very deep hierarchy resulting in state-of-the-art likelihood performance.
 
-In this paper we employ a simple hierarchical VAE with bottom-up inference paths and the more powerful BIVA variant with a bidirectional (top-down and bottom-up) inference model \cite{maaloe_biva_2019}.
+In this paper we employ a simple hierarchical VAE with bottom-up inference paths and the more powerful BIVA variant with a bidirectional (top-down and bottom-up) inference model \parencite{maaloe_biva_2019}.
 We employ skip connections between latent variables but omit them for brevity.
@@
 Similar to these methods, the majority of the approaches to OOD detection make assumptions about the nature of the OOD data.
-The assumptions encompass using labels on the in-distribution data \cite{hendrycks_baseline_2017, liang_enhancing_2018, alemi_uncertainty_2018, lee_simple_2018, lakshminarayanan_simple_2017}, examples of OOD data \cite{hendrycks_deep_2019}, augmenting in-distribution data to mimic it \cite{ren_likelihood_2019}, or assuming a certain data type \cite{serra_input_2020}.
+The assumptions encompass using labels on the in-distribution data \parencite{hendrycks_baseline_2017, liang_enhancing_2018, alemi_uncertainty_2018, lee_simple_2018, lakshminarayanan_simple_2017}, examples of OOD data \parencite{hendrycks_deep_2019}, augmenting in-distribution data to mimic it \parencite{ren_likelihood_2019}, or assuming a certain data type \parencite{serra_input_2020}.
 Any of these assumptions encode implicit biases into the model about the attributes of OOD data which, in turn, might impair performance on truly unknown data examples (unknown unknowns).
@@
 % Introduce the contrast to results with VAEs
-While some of these methods achieve very good results on OOD detection with autoregressive models \cite{oord_pixel_2016, salimans_pixelcnn_2017} and invertible flow-based models \cite{kingma_glow_2018}, it was recently shown that they can be much less effective for VAEs \cite{xiao_likelihood_2020} highlighting the need for a more reliable OOD score for VAEs.
-Although VAEs have the same failure cases as autoregressive and flow-based models, the caveat is that the difference in the likelihood is generally not as big and reconstructions of OOD can be surprisingly good \cite{xiao_likelihood_2020}.
+While some of these methods achieve very good results on OOD detection with autoregressive models \parencite{oord_pixel_2016, salimans_pixelcnn_2017} and invertible flow-based models \parencite{kingma_glow_2018}, it was recently shown that they can be much less effective for VAEs \parencite{xiao_likelihood_2020} highlighting the need for a more reliable OOD score for VAEs.
+Although VAEs have the same failure cases as autoregressive and flow-based models, the caveat is that the difference in the likelihood is generally not as big and reconstructions of OOD can be surprisingly good \parencite{xiao_likelihood_2020}.
 \textcite{xiao_likelihood_2020} alleviate this by refitting the inference network, as previously proposed by \textcite{cremer_inference_2018, mattei_refit_2018}, to a potentially OOD example and measuring the so-called \textit{likelihood regret}.
-However, refitting the inference network can be computationally expensive, especially for the large hierarchical VAEs that are used to model complex data \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
+However, refitting the inference network can be computationally expensive, especially for the large hierarchical VAEs that are used to model complex data \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
 Furthermore, this scales poorly to large amounts of potentially OOD examples as the optimization is done per example.
 
-A few methods have approached OOD detection in a completely unsupervised fashion \cite{maaloe_biva_2019, choi_waic_2019, xiao_likelihood_2020}.
+A few methods have approached OOD detection in a completely unsupervised fashion \parencite{maaloe_biva_2019, choi_waic_2019, xiao_likelihood_2020}.
 The work of \textcite{maaloe_biva_2019} is the most related to ours.
@@
 To alleviate this, we phrase OOD detection as a likelihood ratio test of being \emph{semantically} in-distribution.
-A standard likelihood ratio test \cite{buse_likelihood_1982} suggests to consider the ratio between the associated likelihoods, which we can approximate on a log-scale by the corresponding lower bounds $\mathcal{L}$ and $\mathcal{L}^{>k}$,
+A standard likelihood ratio test \parencite{buse_likelihood_1982} suggests to consider the ratio between the associated likelihoods, which we can approximate on a log-scale by the corresponding lower bounds $\mathcal{L}$ and $\mathcal{L}^{>k}$,
 \begin{equation}\label{eq_hierarchical:llr-as-difference-in-likelihoods}
@@
 Therefore, the $LLR^{>k}$ score should be an improved method for semantic OOD detection compared to $\mathcal{L}^{>k}$.
-Now, it can be noted that if we replace the regular ELBO, $\mathcal{L}$, in \cref{eq_hierarchical:likelihoods-as-exact} with the strictly tighter importance weighted bound \cite{burda_importance_2016},
+Now, it can be noted that if we replace the regular ELBO, $\mathcal{L}$, in \cref{eq_hierarchical:likelihoods-as-exact} with the strictly tighter importance weighted bound \parencite{burda_importance_2016},
 \begin{equation}
@@
 
-\paragraph{Tasks} We follow existing literature \cite{nalisnick_deep_2019, hendrycks_deep_2019} and evaluate our method by setting up OOD detection tasks from FashionMNIST \cite{xiao_fashionmnist_2017} to MNIST \cite{lecun_gradientbased_1998} and from CIFAR10 \cite{krizhevsky_learning_2009} to SVHN \cite{netzer_reading_2011}.
+\paragraph{Tasks} We follow existing literature \parencite{nalisnick_deep_2019, hendrycks_deep_2019} and evaluate our method by setting up OOD detection tasks from FashionMNIST \parencite{xiao_fashionmnist_2017} to MNIST \parencite{lecun_gradientbased_1998} and from CIFAR10 \parencite{krizhevsky_learning_2009} to SVHN \parencite{netzer_reading_2011}.
 For each experiment we train our model on the train split of the former dataset and test its ability to recognize the test split of the latter dataset as OOD from the test split of the former dataset.
@@
     Dashed lines indicate deterministic skip connections which are employed in both networks.
-Skip connections are found to be useful for optimizing latent variable models \cite{dieng_avoiding_2019, maaloe_biva_2019}.}
+Skip connections are found to be useful for optimizing latent variable models \parencite{dieng_avoiding_2019, maaloe_biva_2019}.}
     \label{fig_hierarchical:hvae-graphical-model}
@@
 \paragraph{Models} For each OOD task, we train a simple bottom-up hierarchical VAE with $L$ stochastic layers which we will refer to as ``HVAE''.
-To alleviate posterior collapse we include skip-connections that connect $\zb_i$ to $\zb_{i+2}$ for $i\in\{0, L-2\}$ and $\zb_0\equiv\xb$ in both the inference and generative models \cite{dieng_avoiding_2019} and employ the \textit{free bits} scheme with $\lambda=2$ \cite{kingma_improved_2016}.
-We use weight-normalization \cite{salimans_weight_2016} on all weights and residual networks in the deterministic paths. 
+To alleviate posterior collapse we include skip-connections that connect $\zb_i$ to $\zb_{i+2}$ for $i\in\{0, L-2\}$ and $\zb_0\equiv\xb$ in both the inference and generative models \parencite{dieng_avoiding_2019} and employ the \textit{free bits} scheme with $\lambda=2$ \parencite{kingma_improved_2016}.
+We use weight-normalization \parencite{salimans_weight_2016} on all weights and residual networks in the deterministic paths. 
 A graphical representation of this model can be seen in \cref{fig_hierarchical:hvae-graphical-model}.
-We use a Bernoulli output distribution for FashionMNIST/MNIST and a discretized mixture of logistics output distribution \cite{salimans_pixelcnn_2017} for CIFAR10/SVHN.
+We use a Bernoulli output distribution for FashionMNIST/MNIST and a discretized mixture of logistics output distribution \parencite{salimans_pixelcnn_2017} for CIFAR10/SVHN.
 We use $L=3$ for grey-scale images and $L=4$ for natural images.
-% For CIFAR/SVHN, we also train a BIVA model \cite{maaloe_biva_2019} with $L=10$ and similar configuration as used by the original paper\footnote{Code available at \url{github.com/larsmaaloee/BIVA} and \url{github.com/vlievin/biva-pytorch}}.
+% For CIFAR/SVHN, we also train a BIVA model \parencite{maaloe_biva_2019} with $L=10$ and similar configuration as used by the original paper\footnote{Code available at \url{github.com/larsmaaloee/BIVA} and \url{github.com/vlievin/biva-pytorch}}.
 Full model details are in the \cref{sec_hierarchical:model-details}.
@@
 % Metrics and Evaluation
-\paragraph{Evaluation} Following previous work \cite{hendrycks_baseline_2017, hendrycks_deep_2019, alemi_uncertainty_2018, ren_likelihood_2019, choi_waic_2019} we use the threshold-independent evaluation metrics of Area Under the Receiver Operator Characteristic (AUROC$\uparrow$), Area Under the Precision Recall Curve (AUPRC$\uparrow$) and False Positive Rate at 80\% true positive rate (FPR80$\downarrow$) where the arrow indicates the direction of improvement.
+\paragraph{Evaluation} Following previous work \parencite{hendrycks_baseline_2017, hendrycks_deep_2019, alemi_uncertainty_2018, ren_likelihood_2019, choi_waic_2019} we use the threshold-independent evaluation metrics of Area Under the Receiver Operator Characteristic (AUROC$\uparrow$), Area Under the Precision Recall Curve (AUPRC$\uparrow$) and False Positive Rate at 80\% true positive rate (FPR80$\downarrow$) where the arrow indicates the direction of improvement.
 Note that these metrics are only computable given examples of OOD data but faced with truly OOD data (unknown unknowns), there are many ways to select thresholds to use in practice e.g.\ as the one that yields a specific tolerable false positive rate on the in-distribution test data.
@@
 
-\footnotetext{%
-    \textcite{serra_input_2020} performs the best when high likelihoods are assigned to OOD data such that the overlap with in-distribution data is low.
-    Performance is worse when the overlap is high, cf. \textcite[Table 1]{serra_input_2020}, as seen with complex images.}
+\todo[inline]{Check out if we can make underbrace like curly brackets to the left of tab_hierarchical:rocauc-ood to decrease height and fit page better.}
 \begin{table}[t!]
@@
         HVAE (ours) refers to our hierarchical bottom-up VAE.
-        BIVA (ours) refers to our implementation of the hierarchical BIVA model \cite{maaloe_biva_2019}.
+        BIVA (ours) refers to our implementation of the hierarchical BIVA model.
     }
     \centering
-    \resizebox*{!}{0.9\textheight}{%
+    \resizebox*{!}{0.82\textheight}{%
     \begin{tabular}{lrrr}
@@
          \multicolumn{4}{l}{\textbf{Use prior knowledge of OOD}} \\
-Backgr. contrast. LR (PixelCNN) {\cite{ren_likelihood_2019}}               & $0.994$ & $0.993$ & $0.001$ \\
-Backgr. contrast. LR (VAE) {\cite{choi_waic_2019}}                    & $0.924$ & - & - \\
-Binary classiﬁer {\cite{ren_likelihood_2019}}                              & $0.455$ & $0.505$ & $0.886$ \\ % 6
-$p(\hat{y} | \xb)$ with OOD as noise class {\cite{ren_likelihood_2019}}     & $0.877$ & $0.871$ & $0.195$ \\ % 7
-$p(\hat{y} | \xb)$ with calibration on OOD {\cite{ren_likelihood_2019}}     & $0.904$ & $0.895$ & $0.139$ \\ % 8
-Input complexity ($S$, Glow) \cite{hendrycks_deep_2019}                    & $0.998$ & - & - \\
-Input complexity ($S$, PixelCNN++) \cite{hendrycks_deep_2019}              & $0.967$ & - & - \\
+Backgr. contrast. LR (PixelCNN) {\parencite{ren_likelihood_2019}}               & $0.994$ & $0.993$ & $0.001$ \\
+Backgr. contrast. LR (VAE) {\parencite{choi_waic_2019}}                    & $0.924$ & - & - \\
+Binary classiﬁer {\parencite{ren_likelihood_2019}}                              & $0.455$ & $0.505$ & $0.886$ \\ % 6
+$p(\hat{y} | \xb)$ with OOD as noise class {\parencite{ren_likelihood_2019}}     & $0.877$ & $0.871$ & $0.195$ \\ % 7
+$p(\hat{y} | \xb)$ with calibration on OOD {\parencite{ren_likelihood_2019}}     & $0.904$ & $0.895$ & $0.139$ \\ % 8
+Input complexity ($S$, Glow) \parencite{hendrycks_deep_2019}                    & $0.998$ & - & - \\
+Input complexity ($S$, PixelCNN++) \parencite{hendrycks_deep_2019}              & $0.967$ & - & - \\
          \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
-$p(\hat{y} | \xb)$ {\cite{ren_likelihood_2019, hendrycks_baseline_2017}}                        & $0.734$ & $0.702$ & $0.506$ \\
-Entropy of $p(y | \xb)$ {\cite{ren_likelihood_2019}}                        & $0.746$ & $0.726$ & $0.448$ \\
-ODIN {\cite{ren_likelihood_2019, liang_enhancing_2018}}                                       & $0.752$ & $0.763$ & $0.432$ \\
-VIB \cite{alemi_uncertainty_2018, choi_waic_2019}                                          & $0.941$ & - & - \\
-Mahalanobis distance, CNN {\cite{ren_likelihood_2019}}                     & $0.942$ & $0.928$ & $0.088$ \\
-Mahalanobis distance, DenseNet {\cite{lee_simple_2018}}                & $0.986$ & - & - \\
-Ensemble, 20 classiﬁers {\cite{ren_likelihood_2019, lakshminarayanan_simple_2017}}                  & $0.857$ & $0.849$ & $0.240$ \\
+$p(\hat{y} | \xb)$ {\parencite{ren_likelihood_2019, hendrycks_baseline_2017}}                        & $0.734$ & $0.702$ & $0.506$ \\
+Entropy of $p(y | \xb)$ {\parencite{ren_likelihood_2019}}                        & $0.746$ & $0.726$ & $0.448$ \\
+ODIN {\parencite{ren_likelihood_2019, liang_enhancing_2018}}                                       & $0.752$ & $0.763$ & $0.432$ \\
+VIB \parencite{alemi_uncertainty_2018, choi_waic_2019}                                          & $0.941$ & - & - \\
+Mahalanobis distance, CNN {\parencite{ren_likelihood_2019}}                     & $0.942$ & $0.928$ & $0.088$ \\
+Mahalanobis distance, DenseNet {\parencite{lee_simple_2018}}                & $0.986$ & - & - \\
+Ensemble, 20 classiﬁers {\parencite{ren_likelihood_2019, lakshminarayanan_simple_2017}}                  & $0.857$ & $0.849$ & $0.240$ \\
          \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
          \multicolumn{4}{l}{\textit{- Ensembles}} \\
-WAIC, 5 models, VAE {\cite{choi_waic_2019}}                          & $0.766$ & - & - \\
-WAIC, 5 models, PixelCNN {\cite{ren_likelihood_2019}}                      & $0.221$ & $0.401$ & $0.911$ \\
+WAIC, 5 models, VAE {\parencite{choi_waic_2019}}                          & $0.766$ & - & - \\
+WAIC, 5 models, PixelCNN {\parencite{ren_likelihood_2019}}                      & $0.221$ & $0.401$ & $0.911$ \\
         \multicolumn{4}{l}{\textit{- Not ensembles}} \\
-Likelihood regret \cite{xiao_likelihood_2020}                               & $\mathbf{0.988}$ & - & - \\
+Likelihood regret \parencite{xiao_likelihood_2020}                               & $\mathbf{0.988}$ & - & - \\
 $\mathcal{L}^{>0}$ + HVAE (ours)                    & $0.268$ & $0.363$ & $0.882$ \\
@@
          \multicolumn{4}{l}{\textbf{Use prior knowledge of OOD}} \\
-Backgr. contrast. LR (PixelCNN) {\cite{ren_likelihood_2019}}               & $0.930$ & $0.881$ & $0.066$ \\
-Backgr. contrast. LR (VAE) {\cite{xiao_likelihood_2020}}                    & $0.265$ & - & - \\
-Outlier exposure {\cite{hendrycks_deep_2019}}                              & $0.984$ & - & - \\
-Input complexity ($S$, Glow) \cite{serra_input_2020}                   & $0.950$ & - & - \\
-Input complexity ($S$, PixelCNN++) \cite{serra_input_2020}             & $0.929$ & - & - \\
-Input complexity ($S$, HVAE) (Ours) \cite{serra_input_2020}\footnotemark & $0.833$ & $0.855$ & $0.344$ \\
+Backgr. contrast. LR (PixelCNN) {\parencite{ren_likelihood_2019}}               & $0.930$ & $0.881$ & $0.066$ \\
+Backgr. contrast. LR (VAE) {\parencite{xiao_likelihood_2020}}                    & $0.265$ & - & - \\
+Outlier exposure {\parencite{hendrycks_deep_2019}}                              & $0.984$ & - & - \\
+Input complexity ($S$, Glow) \parencite{serra_input_2020}                   & $0.950$ & - & - \\
+Input complexity ($S$, PixelCNN++) \parencite{serra_input_2020}             & $0.929$ & - & - \\
+Input complexity ($S$, HVAE) (Ours) \parencite{serra_input_2020}\textsuperscript{\ref{footnote:serra_hierarchical_likelihood}} & $0.833$ & $0.855$ & $0.344$ \\
+% Input complexity ($S$, HVAE) (Ours) \parencite{serra_input_2020}\% \footnote{\textcite{serra_input_2020} performs the best when high likelihoods are assigned to OOD data such that the overlap with in-distribution data is low.
+Performance is worse when the overlap is high, cf.
+\textcite[Table 1]{serra_input_2020}, as seen with complex images.} & $0.833$ & $0.855$ & $0.344$ \\
 \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
-Mahalanobis distance {\cite{lee_simple_2018}}                          & $0.991$ & - & -  \\
+Mahalanobis distance {\parencite{lee_simple_2018}}                          & $0.991$ & - & -  \\
          \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
          \multicolumn{4}{l}{\textit{- Ensembles}} \\
-WAIC, 5 models, Glow {\cite{choi_waic_2019}}                          & $1.000$ & - & - \\
-WAIC, 5 models, PixelCNN {\cite{ren_likelihood_2019}}                      & $0.628$ & $0.616$ & $0.657$ \\
+WAIC, 5 models, Glow {\parencite{choi_waic_2019}}                          & $1.000$ & - & - \\
+WAIC, 5 models, PixelCNN {\parencite{ren_likelihood_2019}}                      & $0.628$ & $0.616$ & $0.657$ \\
          \multicolumn{4}{l}{\textit{- Not ensembles}} \\
-Likelihood regret \cite{xiao_likelihood_2020}                               & $0.875$ & - & - \\
+Likelihood regret \parencite{xiao_likelihood_2020}                               & $0.875$ & - & - \\
 $LLR^{>2}$ + HVAE (ours)                            & $0.811$ & $0.837$ & $0.394$ \\
@@
 \end{table}
+
+\addtocounter{footnote}{1}
+\footnotetext{
+    \textcite{serra_input_2020} performs the best when high likelihoods are assigned to OOD data such that the overlap with in-distribution data is low.
+    Performance is worse when the overlap is high, cf. \textcite[Table 1]{serra_input_2020}, as seen with complex images.
+    \label{footnote:serra_hierarchical_likelihood}
+}
 
@@
 Switching the in-distribution data for the OOD data in both cases result in correctly detecting the OOD data; an asymmetry also reported by \textcite{nalisnick_deep_2019}.
-\cref{fig_hierarchical:FMNIST-elbo-k0} shows the density of $\mathcal{L}^{>0}$ in bits per dimension \cite{theis_note_2016} by the model trained on FashionMNIST when evaluated on the FashionMNIST and MNIST test sets.
+\cref{fig_hierarchical:FMNIST-elbo-k0} shows the density of $\mathcal{L}^{>0}$ in bits per dimension \parencite{theis_note_2016} by the model trained on FashionMNIST when evaluated on the FashionMNIST and MNIST test sets.
 We observe a high degree of overlap, with less separation of the OOD data compared to similar results of autoregressive and flow-based models, like \textcite{xiao_likelihood_2020}.
@@
 Importantly, depending on the OOD dataset, the amount of remaining overlap can be high which limits the discriminatory power of the likelihood-based $\mathcal{L}^{>k}$ bound.
-This is in-line with the pathological behavior of the raw likelihood of latent variable models when used for OOD detection \cite{xiao_likelihood_2020}.
+This is in-line with the pathological behavior of the raw likelihood of latent variable models when used for OOD detection \parencite{xiao_likelihood_2020}.
 Since a high degree of overlap also seems present in \textcite{maaloe_biva_2019}, and we see the same problem for our BIVA model trained on CIFAR10, we do not expect this to be due to the less expressive HVAE.
@@
         \includegraphics[width=1\columnwidth]{paper_hierarchical/roc-FashionMNIST-test-MNIST-test-ll-and-llr-IW250_sohau.pdf}
-        \caption{}
+        % \caption{}
         % \caption{ROC curves with AUROC score for detecting MNIST as OOD with the HVAE model trained on FashionMNIST.
         % A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>1}$.}
-        \label{fig_hierarchical:FMNIST-roc-llr}
+        % \label{fig_hierarchical:FMNIST-roc-llr}
     \end{subfigure}
@@
         \includegraphics[width=1\columnwidth]{paper_hierarchical/roc-biva-CIFAR10-SVHN-ll-and-llr_sohau.pdf}
-        \caption{}
+        % \caption{}
         % \caption{ROC curves with AUROC score for detecting SVHN as OOD with the BIVA model trained on CIFAR10.
         % A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>2}$.}
-        \label{fig_hierarchical:CIFAR10-roc-llr}
+        % \label{fig_hierarchical:CIFAR10-roc-llr}
     \end{subfigure}
     \caption[ROC curves for out of distribution detection (MNIST/FashionMNIST and SVHN/CIFAR10).]{%
-        ROC curves with AUROC score for detecting \subref{fig_hierarchical:FMNIST-roc-llr} MNIST as OOD with the HVAE model trained on FashionMNIST and \subref{fig_hierarchical:CIFAR10-roc-llr} SVHN as OOD with the BIVA model trained on CIFAR10.
+        ROC curves with AUROC score for detecting MNIST as OOD with the HVAE model trained on FashionMNIST (left) and SVHN as OOD with the BIVA model trained on CIFAR10 (right). 
         A ROC curve is plotted for each of the $\mathcal{L}^{>k}$ bounds including the ELBO along with one for the best-performing log likelihood-ratio $LLR^{>1}$.
     }
+    \label{fig_hierarchical:FMNIST-roc-llr and CIFAR10-roc-llr}
 \end{figure}
@@
 This is not surprising since the value of $k$ that provides the maximal separation to the reference in-distribution dataset need not be the one for which $\mathcal{LLR}^{>k}$ is overall maximal for the OOD dataset.
-We also visualize the ROC curves resulting from using the $LLR^{>k}$ score for OOD detection on both FashionMNIST/MNIST and CIFAR10/SVHN and compare it to the ROC curves resulting from the different $\mathcal{L}^{>k}$ bounds in \cref{fig_hierarchical:FMNIST-roc-llr, fig_hierarchical:CIFAR10-roc-llr}, respectively.
+We also visualize the ROC curves resulting from using the $LLR^{>k}$ score for OOD detection on both FashionMNIST/MNIST and CIFAR10/SVHN and compare it to the ROC curves resulting from the different $\mathcal{L}^{>k}$ bounds in \cref{fig_hierarchical:FMNIST-roc-llr and CIFAR10-roc-llr}, respectively.
 On both datasets we see significantly better discriminatory performance when using the $LLR^{>k}$ score.
@@
 \cref{tab_hierarchical:rocauc-ood} shows that BIVA improves upon the HVAE model for OOD detection on CIFAR while \cref{tab_hierarchical:bits-per-dim-ood} shows that the BIVA model also improves upon the HVAE in terms of likelihood.
-We hypothesize that models larger than our implementation of BIVA, with better likelihood scores may perform even better \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
+We hypothesize that models larger than our implementation of BIVA, with better likelihood scores may perform even better \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
 
@@
 \paragraph{Performance} \cref{tab_hierarchical:rocauc-ood} summarize our results compared to baselines based on the commonly used AUROC$\uparrow$, AUPRC$\uparrow$ and FPR80$\downarrow$ metrics.
-Our method outperforms other generative model-based methods such as WAIC \cite{choi_waic_2019} with Glow model and performs similarly to the likelihood regret method of \cite{xiao_likelihood_2020}.
-Furthermore, our method performs similarly to the background constrative likelihood ratio method of \textcite{ren_likelihood_2019} on FashionMNIST/MNIST but contrary to the failure of that method on CIFAR10/SVHN reported by \cite{xiao_likelihood_2020}, our method performs very well on this task too.
-Our approach outperforms all supervised approaches that use in-distribution labels or synthetic examples of OOD data derived from the in-distribution data including ODIN \cite{liang_enhancing_2018} and the predictive distribution of a classifier $p(\hat{y}|\xb)$ trained and evaluated in various ways (see \textcite{ren_likelihood_2019}).
+Our method outperforms other generative model-based methods such as WAIC \parencite{choi_waic_2019} with Glow model and performs similarly to the likelihood regret method of \parencite{xiao_likelihood_2020}.
+Furthermore, our method performs similarly to the background constrative likelihood ratio method of \textcite{ren_likelihood_2019} on FashionMNIST/MNIST but contrary to the failure of that method on CIFAR10/SVHN reported by \parencite{xiao_likelihood_2020}, our method performs very well on this task too.
+Our approach outperforms all supervised approaches that use in-distribution labels or synthetic examples of OOD data derived from the in-distribution data including ODIN \parencite{liang_enhancing_2018} and the predictive distribution of a classifier $p(\hat{y}|\xb)$ trained and evaluated in various ways (see \textcite{ren_likelihood_2019}).
 
@@
 For a typical forward pass that is linear in the input dimensionality, $D$, and the number of stochastic layers, $L$, this amounts to computation of $\mathcal{O}(DL)$.
-Compared to some related work that either requires an $M>1$ sized batch of inputs of which either all or none are OOD \cite{nalisnick_detecting_2019} or cannot be applied to batches due to the required per-example optimization \cite{xiao_likelihood_2020}, our method additionally is applicable to batches of any size that may consist of both OOD and in-distribution examples which provides drastic speed-ups via vectorization and parallelization.
+Compared to some related work that either requires an $M>1$ sized batch of inputs of which either all or none are OOD \parencite{nalisnick_detecting_2019} or cannot be applied to batches due to the required per-example optimization \parencite{xiao_likelihood_2020}, our method additionally is applicable to batches of any size that may consist of both OOD and in-distribution examples which provides drastic speed-ups via vectorization and parallelization.
 Furthermore, the method of \textcite{xiao_fashionmnist_2017} requires refitting the inference network of a VAE which can be computationally demanding.
@@
 Hence, there can be an asymmetry where the likelihoods of simple OOD data can be high for a model trained on complex data, but not the other way around.
-Loosely put, the minimal number of bits required to losslessly compress data sampled from some distribution is the entropy of the generating process \cite{shannon_mathematical_1948, mackay_information_2003}.
+Loosely put, the minimal number of bits required to losslessly compress data sampled from some distribution is the entropy of the generating process \parencite{shannon_mathematical_1948, mackay_information_2003}.
 \textcite{townsend_practical_2019} recently showed that VAEs can be used for lossless compression at rates superior to more generic algorithms.
 
-We also note that since the hierarchical VAE is a probabilistic graphical latent variable model, it lends itself very naturally to manipulation at the feature level \cite{kingma_semi-supervised_2014, maaloe_auxiliary_2016, maaloe_semi-supervised_2017}.
+We also note that since the hierarchical VAE is a probabilistic graphical latent variable model, it lends itself very naturally to manipulation at the feature level \parencite{kingma_semi-supervised_2014, maaloe_auxiliary_2016, maaloe_semi-supervised_2017}.
 This property sets it apart from other generative models that do not explicitly define such a hierarchy of features.

### 2023-09-29 chapters/paper_modelagnostic.tex
Updates
Updates to citation commands and footnotes

@@
 
-\begin{figure*}[tbp]
+\begin{sidewaysfigure}[tbp]
     \includegraphics[width=\textwidth]{graphics/paper_modelagnostic/pvalues_comb_no_drop_correct_font.pdf}
@@
     \label{fig_modelagnostic:comb_p_values}
-\end{figure*}
+\end{sidewaysfigure}
 
@@
     %\renewcommand{\thetable}{4}
-    \resizebox{0.80\textwidth}{!}{
+    \resizebox{0.90\textwidth}{!}{
         \scriptsize
@@
     \centering
-    \includegraphics[scale=0.48]{graphics/paper_modelagnostic/fashion_hvae_type1_correct_font.pdf}
+    \includegraphics[scale=0.6]{graphics/paper_modelagnostic/fashion_hvae_type1_correct_font.pdf}
     \caption[Type I and Type II errors versus the significance level $\alpha$ on the combination values]{Type I (probability of an inlier to be classified as outlier) and Type II (probability of an outlier to be considered as inlier) errors versus the significance level $\alpha$ on the combination values.
@@
 
-\subsubsection*{Acknowledgements}
+\section*{Acknowledgements}
 Federico Bergamin and Pierre-Alexandre Mattei contributed eqaully to this paper, which is indicated by the asterisk (*) in the author list.

### 2023-09-29 chapters/paper_retrospective.tex
Updates to citation commands and footnotes

@@
 \section{Introduction}
-Stroke is a leading cause of disability and death worldwide \cite{cite1,cite2,cite3}.
-Effective treatment is time-sensitive, and an optimal outcome is more likely when treatment is administered within the first four and a half hours from stroke onset \cite{cite4,cite5}.
+Stroke is a leading cause of disability and death worldwide \parencite{cite1,cite2,cite3}.
+Effective treatment is time-sensitive, and an optimal outcome is more likely when treatment is administered within the first four and a half hours from stroke onset \parencite{cite4,cite5}.
 The gateway to ambulance transport and hospital admittance is through prehospital telehealth services, including emergency medical call centres, nurse advice call lines, and out-of-hours health services.
 In the pre-hospital setting, the use of mobile stroke units has made it possible to deliver advanced treatment faster.
-As the mobile stroke unit is only dispatched to patients with a suspected stroke, the impact of mobile stroke units is directly influenced by accurate call-taker recognition of stroke \cite{cite6,cite7}.
+As the mobile stroke unit is only dispatched to patients with a suspected stroke, the impact of mobile stroke units is directly influenced by accurate call-taker recognition of stroke \parencite{cite6,cite7}.
 Call-takers who can rapidly and accurately recognize stroke are therefore crucial in facilitating prompt care in both pre-hospital and in-hospital settings.
 
-Despite initiatives to improve stroke recognition \cite{cite8,cite9}, approximately half of all patients with stroke do not receive the correct triage for their condition from call-takers \cite{cite10,cite11,cite12}.
-Most initiatives aim to improve stroke recognition by call-takers via introducing more specific assessment tools \cite{cite8,cite9} or providing specialized training \cite{cite13}.
-Recent advances in machine learning technology might be applied to improve stroke recognition without requiring changes to the triaging approach, and machine learning aided identification of stroke has been suggested as a means of improving mobile stroke unit effectiveness \cite{cite7}.
-Real-time feedback from a machine learning model can improve the recognition of out-of-hospital cardiac arrest \cite{cite14,cite15}.
+Despite initiatives to improve stroke recognition \parencite{cite8,cite9}, approximately half of all patients with stroke do not receive the correct triage for their condition from call-takers \parencite{cite10,cite11,cite12}.
+Most initiatives aim to improve stroke recognition by call-takers via introducing more specific assessment tools \parencite{cite8,cite9} or providing specialized training \parencite{cite13}.
+Recent advances in machine learning technology might be applied to improve stroke recognition without requiring changes to the triaging approach, and machine learning aided identification of stroke has been suggested as a means of improving mobile stroke unit effectiveness \parencite{cite7}.
+Real-time feedback from a machine learning model can improve the recognition of out-of-hospital cardiac arrest \parencite{cite14,cite15}.
 Therefore, this study aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.
 
-We used call recordings and registry data from the Copenhagen Emergency Medical Services (CEMS) and the Danish Stroke Registry (DanStroke)\cite{cite16} from 2015 to 2020.
+We used call recordings and registry data from the Copenhagen Emergency Medical Services (CEMS) and the Danish Stroke Registry (DanStroke)\parencite{cite16} from 2015 to 2020.
 The call recordings were obtained from two call lines: the 1-1-2 emergency line and the medical helpline 1813 (MH-1813).
@@
 First, the model may have learned to mimic call-takers with the same bias.
-Second, women may experience different and more challenging-to-identify symptoms than men \cite{cite29, cite30}.
+Second, women may experience different and more challenging-to-identify symptoms than men \parencite{cite29, cite30}.
 Third, a higher prevalence of male patients with stroke was observed in the training data.
@@
 We did not have data to estimate potential bias related to race, ethnicity, language, accent, or dialects.
-Previous studies on speech recognition for call centres have indeed found that non-native speakers had a higher rate of transcription errors \cite{han_deep_2017}.
+Previous studies on speech recognition for call centres have indeed found that non-native speakers had a higher rate of transcription errors \parencite{han_deep_2017}.
 Since our model was trained on a representative - and therefore unbalanced - sample, we expect it to behave similarly.
@@
 Although we believe that the proposed machine learning framework can be further improved, several alternatives have already been explored in the preliminary experimental phase.
-The speech recognition model we used was trained on 1-1-2 calls for a previous project \cite{cite14}, and so, was specialised to a domain very similar to that of MH-1813.
-We also tested an open-source, multilingual model from OpenAI called Whisper \cite{radford_robust_2023}, but found that performance degraded slightly compared to the model trained on 1-1-2.
+The speech recognition model we used was trained on 1-1-2 calls for a previous project \parencite{cite14}, and so, was specialised to a domain very similar to that of MH-1813.
+We also tested an open-source, multilingual model from OpenAI called Whisper \parencite{radford_robust_2023}, but found that performance degraded slightly compared to the model trained on 1-1-2.
 We hypothesise that this is due to Whisper's inability to handle the specific noise conditions and recognise words from a specialised medical vocabulary.
@@
 Only relatively few options exist for the Danish language, none of which are specialised in the medical domain.
-We used a monolingual Danish BERT model, which has previously been shown to outperform a multilingual alternative from Google for Danish named entity recognition \cite{hvingelby_dane_2020}.
+We used a monolingual Danish BERT model, which has previously been shown to outperform a multilingual alternative from Google for Danish named entity recognition \parencite{hvingelby_dane_2020}.
 However, this also did not result in a significant performance improvement.
@@
 Thus, we chose the simpler MLP ensemble.
-We have included references to reviews of self-supervised learning for speech and text in the references \cite{gururangan_don_2020,mohamed_selfsupervised_2022}.
-Notably, it is not uncommon for small, simple models to match or outperform large, pre-trained models for text-classification tasks \cite{galke_bagofwords_2022}.
+We have included references to reviews of self-supervised learning for speech and text in the references \parencite{gururangan_don_2020,mohamed_selfsupervised_2022}.
+Notably, it is not uncommon for small, simple models to match or outperform large, pre-trained models for text-classification tasks \parencite{galke_bagofwords_2022}.
 
@@
 
-The CEMS is responsible for providing prehospital telehealth services in the Capital Region of Denmark, with a catchment area of 1.9 million \cite{cite17}.
+The CEMS is responsible for providing prehospital telehealth services in the Capital Region of Denmark, with a catchment area of 1.9 million \parencite{cite17}.
 CEMS operates two call lines: the 1-1-2 emergency line, similar to 9-1-1 in the United States, intended for acute conditions.
@@
 The condition suspected by the call-taker is categorised based on a predefined diagnostic index and stored in an electronic record using a computer-aided dispatch (CAD) system.
-The CAD records are associated with the Danish civil registration number (CPR number) \cite{cite19} of the patient.
+The CAD records are associated with the Danish civil registration number (CPR number) \parencite{cite19} of the patient.
 The CPR number is a unique identification assigned to all Danish residents.
@@
 
-All patients with a final diagnosis of stroke or transient ischaemic attack admitted to a Danish hospital within 5 days of symptom onset are recorded in the Danish Stroke Registry \cite{cite16}, also known as DanStroke.
+All patients with a final diagnosis of stroke or transient ischaemic attack admitted to a Danish hospital within 5 days of symptom onset are recorded in the Danish Stroke Registry \parencite{cite16}, also known as DanStroke.
 This record includes the patient-reported time of onset, stroke type (haemorrhagic, ischaemic, or transient ischaemic attack), and CPR number of the patient.
-The diagnosis is obtained according to the national guidelines \cite{baluenfeldt_national_2021}, which includes cerebral imaging and full diagnostic workup by neurologists.
-The validity of the Danish stroke registry has been shown to be high \cite{wildenschild_registration_2013}, and the number of stroke mimics is therefore minimised in our dataset.
+The diagnosis is obtained according to the national guidelines \parencite{baluenfeldt_national_2021}, which includes cerebral imaging and full diagnostic workup by neurologists.
+The validity of the Danish stroke registry has been shown to be high \parencite{wildenschild_registration_2013}, and the number of stroke mimics is therefore minimised in our dataset.
 
@@
 We excluded calls involving subarachnoid haemorrhage cases.
-Finally, we considered a call to be a call-taker stroke positive when the call-taker selected the stroke diagnostic category during the call and dispatched an ambulance with the appropriate level of response \cite{cite20}.
+Finally, we considered a call to be a call-taker stroke positive when the call-taker selected the stroke diagnostic category during the call and dispatched an ambulance with the appropriate level of response \parencite{cite20}.
 To ensure that the effect of the machine learning framework was not overestimated, we excluded calls where diagnostic category had not been registered from the test set.
@@
 This conversion is a commonly used input representation for speech-processing tasks, which facilitates the identification of linguistic content in audio signals.
-We used a speech recognition model with a neural network architecture \cite{borgholt_endtoend_2020}, consisting of two-dimensional convolutional layers \cite{cite22} and blocks of bidirectional long short-term memory layers \cite{hochreiter_long_1997} The output is a sequence of probability distributions over characters of the Danish alphabet, which were then converted into a human-readable transcript using a greedy decoder \cite{graves_connectionist_2006}.
+We used a speech recognition model with a neural network architecture \parencite{borgholt_endtoend_2020}, consisting of two-dimensional convolutional layers \parencite{cite22} and blocks of bidirectional long short-term memory layers \parencite{hochreiter_long_1997} The output is a sequence of probability distributions over characters of the Danish alphabet, which were then converted into a human-readable transcript using a greedy decoder \parencite{graves_connectionist_2006}.
 
@@
 The feature selection procedure is detailed in \cref{app:supplementary-retrospective}.
-The model was constructed as an ensemble \cite{hansen_neural_1990} of five identical, independently trained models.
-Each consists of a stack of neural network layers commonly referred to as a multi-layer perceptron \cite{rosenblatt_perceptron_1958}.
+The model was constructed as an ensemble \parencite{hansen_neural_1990} of five identical, independently trained models.
+Each consists of a stack of neural network layers commonly referred to as a multi-layer perceptron \parencite{rosenblatt_perceptron_1958}.
 The final layer has a single scalar output and applies a sigmoid nonlinearity to produce an output score between zero and one.
@@
 The speech recognition model was trained on 3,811 manually transcribed random calls (173 h) from the CEMS as part of a previous project.14 These calls exclusively originated from 1- 1-2 between 2015 and 2018, ensuring no overlap with the test data used for the text classification model.
-The model was trained using a connectionist temporal classification objective \cite{graves_connectionist_2006}.
+The model was trained using a connectionist temporal classification objective \parencite{graves_connectionist_2006}.
 
@@
 We computed 95\% confidence intervals (CIs) using bootstrapping.
-\cite{dwass_modified_1957,eden_validity_1933}.
+\parencite{dwass_modified_1957,eden_validity_1933}.
 In our assessment, we accounted for random variation associated with model training by basing the means, tests, and CIs on the predictions of 11 randomly initialised training runs.

### 2023-09-29 chapters/paper_review/01_introduction.tex
Add and revise short table and figure captions
Updates to citation commands and footnotes

@@
 	 \caption{Framework for using self-supervised representation learning in
-	 downstream applications}
+	 downstream applications.}
     \label{fig:SSL_framework}
@@
 transformation, offering impressive gains for scenarios rich in labeled
-% data~\cite{lecun_deep_2015}. 
-% data~\cite{lecun_deep_2015, hinton_deep_2012}. 
-% data~\cite{bourlard_connectionist_1994}. 
-% data~\cite{lecun_deep_2015,bourlard_connectionist_1994}. 
-data~\cite{lecun_deep_2015,hinton_deep_2012,bourlard_connectionist_1994}. 
+% data~\parencite{lecun_deep_2015}. 
+% data~\parencite{lecun_deep_2015, hinton_deep_2012}. 
+% data~\parencite{bourlard_connectionist_1994}. 
+% data~\parencite{lecun_deep_2015,bourlard_connectionist_1994}. 
+data~\parencite{lecun_deep_2015,hinton_deep_2012,bourlard_connectionist_1994}. 
 Paradoxically, this heavy reliance on supervised learning has restricted progress in
@@
 unpaired audio-only data to open up new industrial speech use-cases and
-low-resource languages~\cite{kemp_unsupervised_1999, lamel_lightly_2002, ma_unsupervised_2006}. Inspired by how
+low-resource languages~\parencite{kemp_unsupervised_1999, lamel_lightly_2002, ma_unsupervised_2006}. Inspired by how
 children learn their first language through listening and interacting with
@@
 information. These learned representations are then used for target downstream
-applications requiring a minimal number of labeled data~\cite{hinton_learning_2007,
+applications requiring a minimal number of labeled data~\parencite{hinton_learning_2007,
 lecun_tutorial_2006, bengio_representation_2013}. 
@@
 features that capture the underlying explanatory factors for the observed
-input~\cite{bengio_representation_2013}.
+input~\parencite{bengio_representation_2013}.
 
@@
 naturally occurring patterns in training samples for which there are no pre-assigned
-labels or scores~\cite{jordan_machine_2015}. 
+labels or scores~\parencite{jordan_machine_2015}. 
 The term ``unsupervised'' is used to distinguish this family of methods from
@@
 with labels to guide learning using a larger volume of unlabeled samples.
-Examples of unsupervised learning techniques include $k$-means clustering~\cite{gray_vector_1984}, mixture models~\cite{jordan_hierarchical_1994}, autoencoders~\cite{hinton_autoencoders_1994},
-and non-negative matrix factorization~\cite{lee_learning_1999}. 
+Examples of unsupervised learning techniques include $k$-means clustering~\parencite{gray_vector_1984}, mixture models~\parencite{jordan_hierarchical_1994}, autoencoders~\parencite{hinton_autoencoders_1994},
+and non-negative matrix factorization~\parencite{lee_learning_1999}. 
 \textit{Self-supervised learning} (SSL) is a fast-growing subcategory of
@@
 representation from the frozen model, or fine-tune the entire pre-trained model
-in a supervised phase~\cite{hinton_reducing_2006}. 
+in a supervised phase~\parencite{hinton_reducing_2006}. 
 Automatic speech recognition (ASR) and speaker identification (SID) are 
@@
 Due to the popularity of SSL, reviews have been published about the
-technology in general~\cite{bommasani_opportunities_2021,ericsson_selfsupervised_2022,liu_selfsupervised_2021} as well as its application to natural language processing (NLP)~\cite{rogers_primer_2020,liu_pretrain_2021,xia_which_2020,qiu_pretrained_2020} and computer vision (CV)~\cite{jing_selfsupervised_2021}.
-Recently, a brief overview with a general focus on speech representation learning was published \cite{borgholt_brief_2022}.
+technology in general~\parencite{bommasani_opportunities_2021,ericsson_selfsupervised_2022,liu_selfsupervised_2021} as well as its application to natural language processing (NLP)~\parencite{rogers_primer_2020,liu_pretrain_2021,xia_which_2020,qiu_pretrained_2020} and computer vision (CV)~\parencite{jing_selfsupervised_2021}.
+Recently, a brief overview with a general focus on speech representation learning was published \parencite{borgholt_brief_2022}.
 However, none of these overviews focus exclusively on SSL for speech processing.
 Since the speech signal differs greatly from image and text inputs, many  theories  and technologies have been developed to address the unique challenges of speech.
-One review addresses speech representation learning based on deep learning models~\cite{latif_deep_2021}, but does not address recent developments in self-supervised learning.
+One review addresses speech representation learning based on deep learning models~\parencite{latif_deep_2021}, but does not address recent developments in self-supervised learning.
 This motivates this overview of speech SSL.

### 2023-09-29 chapters/paper_review/02_three_waves.tex
Updates to citation commands and footnotes

@@
 Early work used simple clustering methods. For example, in work such
-as \cite{rabiner_considerations_1979,wilpon_modified_1985}, word patterns were clustered
+as \parencite{rabiner_considerations_1979,wilpon_modified_1985}, word patterns were clustered
 semi-automatically using techniques such as $k$-means, after which isolated words
@@
 Through time, modeling techniques improved such that subword units were
-represented by Gaussian mixture models (GMMs) \cite{gauvain_maximum_1994}, which facilitated
+represented by Gaussian mixture models (GMMs) \parencite{gauvain_maximum_1994}, which facilitated
 the modeling of more variability in the input data. GMMs were first built for
 context-independent phonemes; state-clustering 
-algorithms~\cite{young_state_1994} then resulted in GMMs for context-dependent phonemes. 
+algorithms~\parencite{young_state_1994} then resulted in GMMs for context-dependent phonemes. 
 Each latent component of these mixture models acted as a template of a
@@
 large volumes of data with diverse characteristics. 
-Furthermore, dynamical models like hidden Markov models (HMMs)~\cite{bell_adaptation_2021}
+Furthermore, dynamical models like hidden Markov models (HMMs)~\parencite{bell_adaptation_2021}
 allowed for the processing of continuous speech rather than just isolated word
@@
 models, e.g., a background model and a foreground model. Examples of their use
-in speech processing include speech recognition~\cite{smith_speech_2001,venkataramani_support_2003} and
-speaker recognition~\cite{wan_svmsvm_2003}. Subsequent techniques in speaker and language
-verification~\cite{dehak_frontend_2011,dehak_language_2011} similarly extracted parameters
+in speech processing include speech recognition~\parencite{smith_speech_2001,venkataramani_support_2003} and
+speaker recognition~\parencite{wan_svmsvm_2003}. Subsequent techniques in speaker and language
+verification~\parencite{dehak_frontend_2011,dehak_language_2011} similarly extracted parameters
 (concatenated means) from trained background GMMs as representations that were
@@
 Examples of early techniques include restricted Boltzmann machines
-(RBM)~\cite{hinton_reducing_2006}, denoising autoencoders~\cite{vincent_stacked_2010}, noise contrastive
-estimation (NCE)~\cite{gutmann_noisecontrastive_2010}, sparse coding~\cite{olshausen_emergence_1996,
-lee_efficient_2006, sivaram_sparse_2010}, and energy-based methods~\cite{ranzato_unified_2007}.
+(RBM)~\parencite{hinton_reducing_2006}, denoising autoencoders~\parencite{vincent_stacked_2010}, noise contrastive
+estimation (NCE)~\parencite{gutmann_noisecontrastive_2010}, sparse coding~\parencite{olshausen_emergence_1996,
+lee_efficient_2006, sivaram_sparse_2010}, and energy-based methods~\parencite{ranzato_unified_2007}.
 Many of these techniques have also been applied to CV and NLP problems, which
@@
 example, approximate training methods such as contrastive divergence
-\cite{hinton_training_2002} were a practical technique to streamline RBM training. 
+\parencite{hinton_training_2002} were a practical technique to streamline RBM training. 
 Furthermore, deep networks had non-convex objective functions, which
@@
 For example, in NLP, a representation model is usually assessed on GLUE,
-which comprises nine tasks~\cite{wang_glue_2018}, whereas in speech, a representation model can be
-evaluated on SUPERB, which comprises ten tasks~\cite{yang_superb_2021}, 
+which comprises nine tasks~\parencite{wang_glue_2018}, whereas in speech, a representation model can be
+evaluated on SUPERB, which comprises ten tasks~\parencite{yang_superb_2021}, 
 as described in detail in \cref{sec:benchmark}.
@@
 via pretext tasks for restoring the true color of black-and-white
-images~\cite{zhang_colorful_2016}, joint learning of latent representations and their
-cluster assignments~\cite{caron_deep_2018}, and the prediction of the relative positions of
-image patches~\cite{doersch_unsupervised_2016}. Other popular approaches include variational
-autoencoders (VAEs)~\cite{kingma_autoencoding_2014, rezende_stochastic_2014}. While typical autoencoders learn data
+images~\parencite{zhang_colorful_2016}, joint learning of latent representations and their
+cluster assignments~\parencite{caron_deep_2018}, and the prediction of the relative positions of
+image patches~\parencite{doersch_unsupervised_2016}. Other popular approaches include variational
+autoencoders (VAEs)~\parencite{kingma_autoencoding_2014, rezende_stochastic_2014}. While typical autoencoders learn data
 representations using unsupervised objectives by reconstructing the input
 after passing it through an information bottleneck, VAEs estimate a neural model of a probability density function (pdf) that approximates the unknown “true” distribution of the observed data, for which we only have access to independently identically distributed (iid) samples.
-It is also important to mention dynamical VAEs \cite{girin_dynamical_2021}, which is an extension of VAE for sequential data such as speech.
+It is also important to mention dynamical VAEs \parencite{girin_dynamical_2021}, which is an extension of VAE for sequential data such as speech.
 
@@
 in a sentence to predict the next token such as in
-ELMo~\cite{peters_deep_2018}, the GPT
-series~\cite{radford_language_2019}, and 
-Megatron~\cite{shoeybi_megatronlm_2020}, or predicting the masked tokens in a
+ELMo~\parencite{peters_deep_2018}, the GPT
+series~\parencite{radford_language_2019}, and 
+Megatron~\parencite{shoeybi_megatronlm_2020}, or predicting the masked tokens in a
 sentence such as with the bidirectional encoder representations from Transformers
-(BERT) series~\cite{devlin_bert_2018,liu_roberta_2019}. 
+(BERT) series~\parencite{devlin_bert_2018,liu_roberta_2019}. 
 Another common pretext task in the third wave is contrastive
-learning~\cite{oord_representation_2018}, in which a model learns to identify a
+learning~\parencite{oord_representation_2018}, in which a model learns to identify a
 target instance from a set of negative samples. 
 This approach has become especially popular in the
-CV context~\cite{chen_simple_2020,he_momentum_2020,chen_improved_2020,caron_unsupervised_2020}.
+CV context~\parencite{chen_simple_2020,he_momentum_2020,chen_improved_2020,caron_unsupervised_2020}.
 In this survey, we will mainly focus on techniques for pretext task
@@
 label a large volume of speech-only data, which is then used to augment the
-initial labeled data to train a student model~\cite{kemp_unsupervised_1999, lamel_lightly_2002,
+initial labeled data to train a student model~\parencite{kemp_unsupervised_1999, lamel_lightly_2002,
 ma_unsupervised_2006, parthasarathi_lessons_2019}. PL has been successful and widely adopted in the
@@
 models with more parameters than their original teachers to capture the
-complexities in vastly larger speech-only data~\cite{park_improved_2020,
+complexities in vastly larger speech-only data~\parencite{park_improved_2020,
 xu_iterative_2020, xiao_scaling_2021}. 
@@
 pre-training speech models. TL transfers knowledge captured by models
-trained on one task to different but related tasks~\cite{caruana_multitask_1997}. 
+trained on one task to different but related tasks~\parencite{caruana_multitask_1997}. 
 The past few decades have seen active research on TL and its
@@
 Multilingual and cross-lingual supervised models have proven superior in
-low-resource speech recognition tasks~\cite{cui_multilingual_2015}.
+low-resource speech recognition tasks~\parencite{cui_multilingual_2015}.
 SSL can be regarded as a type of TL because knowledge learned from pre-training
@@
 This survey paper focuses on SSL, and not all TL technologies for speech. 
-One survey indeed addresses TL for speech processing~\cite{bell_adaptation_2021}
+One survey indeed addresses TL for speech processing~\parencite{bell_adaptation_2021}
 but does not include current SSL technologies for speech.

### 2023-09-29 chapters/paper_review/03_approaches.tex
Updates to citation commands and footnotes

@@
 \edit{After pre-training,} the representation model $f(\cdot)$ can be fine-tuned for a downstream task directly or used to extract features which are fed to another model, as visualized in \cref{fig:SSL_framework}.
-It is not uncommon to use the output representation~$H$, but often representations from hidden layers of $f(\cdot)$ are better suited \cite{pasad_layerwise_2021}.
+It is not uncommon to use the output representation~$H$, but often representations from hidden layers of $f(\cdot)$ are better suited \parencite{pasad_layerwise_2021}.
 
-\begin{figure*}[t!]
+\begin{figure*}[t]
     \centering
@@
 	 but find that generative tasks are the most important for downstream
-	 task performance~\cite{pascual_learning_2019}.}
+	 task performance~\parencite{pascual_learning_2019}.}
     \label{fig:timeline}
@@
 \paragraph{Autoencoding} 
-Since their introduction in the mid-1990s~\cite{hinton_autoencoders_1994}, autoencoders (AEs)
+Since their introduction in the mid-1990s~\parencite{hinton_autoencoders_1994}, autoencoders (AEs)
 have played an essential role in learning distributed latent representations of
@@
 For instance, denoising autoencoders (DAEs) learn latent representations by
-reconstructing from input corrupted by noise~\cite{vincent_stacked_2010}. 
-\edit{The Variational Autoencoder (VAE) is a probabilistic version of the AE which defines the latent representation via a posterior distribution over stochastic latent variables \cite{kingma_autoencoding_2014,rezende_stochastic_2014}.
-VAEs have been applied to speech in numerous works \cite{chung_recurrent_2015, fraccaro_sequential_2016, hsu_learning_2017, hsu_unsupervised_2017, aksan_stcn_2019}}.
+reconstructing from input corrupted by noise~\parencite{vincent_stacked_2010}. 
+\edit{The Variational Autoencoder (VAE) is a probabilistic version of the AE which defines the latent representation via a posterior distribution over stochastic latent variables \parencite{kingma_autoencoding_2014,rezende_stochastic_2014}.
+VAEs have been applied to speech in numerous works \parencite{chung_recurrent_2015, fraccaro_sequential_2016, hsu_learning_2017, hsu_unsupervised_2017, aksan_stcn_2019}}.
 The vector-quantized variational autoencoder (VQ-VAE) is another model in this
-category~\cite{oord_neural_2018};
-it extends the original VAE~\cite{kingma_autoencoding_2014} with a novel
+category~\parencite{oord_neural_2018};
+it extends the original VAE~\parencite{kingma_autoencoding_2014} with a novel
 parameterization of the posterior distribution for discrete latent
 representations. 
-\edit{The VQ-VAE has been instrumental in generative speech modelling and recent work on generative spoken language modeling has successfully combined the idea of a discrete latent space with self-supervised learning \cite{polyak_speech_2021, kharitonov_textfree_2021, nguyen_generative_2022}.}
+\edit{The VQ-VAE has been instrumental in generative speech modelling and recent work on generative spoken language modeling has successfully combined the idea of a discrete latent space with self-supervised learning \parencite{polyak_speech_2021, kharitonov_textfree_2021, nguyen_generative_2022}.}
 
@@
 This operation is non-differentiable and the \edit{gradients of the loss with respect to the encoder parameters} must be obtained by approximation.
-In the VQ-VAE this is done using the straight-through estimator~\cite{bengio_estimating_2013}, i.e., the gradients \edit{with respect to the} encoder output are taken to be equal to those \edit{with respect to the} decoder input \ci{(i.e., the quatization step is ignored)}.
+In the VQ-VAE this is done using the straight-through estimator~\parencite{bengio_estimating_2013}, i.e., the gradients \edit{with respect to the} encoder output are taken to be equal to those \edit{with respect to the} decoder input \ci{(i.e., the quatization step is ignored)}.
 Given a learned codebook $A\in\mathbb{R}^{K \times D}$, where $K$ is the codebook size and $D$ is the dimensionality of each codebook vector $a_k$, the quantized representation $q_t$ of $h_t$ is obtained as
@@
 \end{align}
-\edit{The decoder $g(\cdot)$ is an autoregressive model that takes $q_{[1,t]}$ as input to generate $x_t$ \cite{oord_wavenet_2016}}. 
+\edit{The decoder $g(\cdot)$ is an autoregressive model that takes $q_{[1,t]}$ as input to generate $x_t$ \parencite{oord_wavenet_2016}}. 
 Codebook learning is facilitated by a two-term auxiliary loss similar to
 classical vector quantization dictionary 
-learning~\cite{burton_generalization_1983, soong_vector_1985}. 
+learning~\parencite{burton_generalization_1983, soong_vector_1985}. 
 Gradients for the codebook vectors are given solely by a term that moves
@@
 speech information closely related to phonemes, and are useful for
-applications such as speaker conversion~\cite{chorowski_unsupervised_2019}.
+applications such as speaker conversion~\parencite{chorowski_unsupervised_2019}.
 Vector quantization is \edit{not} exclusive to VQ-VAE but has seen
@@
 
-\edit{The Gumbel softmax \cite{jang_categorical_2016} is another frequently used approach for obtaining a discrete representation space, and has also been used for AEs \cite{eloff_unsupervised_2019}.
-In addition to the approaches discussed above, several other works on speech representation learning take inspiration from the AE framework \cite{zeiler_rectified_2013, badino_autoencoder_2014, badino_discovering_2015, kamper_unsupervised_2015, renshaw_comparison_2015, settle_acoustically_2019}.}
+\edit{The Gumbel softmax \parencite{jang_categorical_2016} is another frequently used approach for obtaining a discrete representation space, and has also been used for AEs \parencite{eloff_unsupervised_2019}.
+In addition to the approaches discussed above, several other works on speech representation learning take inspiration from the AE framework \parencite{zeiler_rectified_2013, badino_autoencoder_2014, badino_discovering_2015, kamper_unsupervised_2015, renshaw_comparison_2015, settle_acoustically_2019}.}
 
@@
 %\item VAE, VQ-VAE
-    \item DeCoAR 2.0~\cite{ling_decoar_2020} presents a deep contextualized acoustic representation learning approach with the addition of a vector quantization (VQ) layer.
-    \item In VQ-APC~\cite{chung_vectorquantized_2020}, a VQ layer is used with the APC objective, which imposes a bottleneck and forces the model to learn better representations.
-    \item Two dropout regularization methods, attention dropout and layer dropout, are introduced to TERA~\cite{luo_dropout_2021}. 
+    \item DeCoAR 2.0~\parencite{ling_decoar_2020} presents a deep contextualized acoustic representation learning approach with the addition of a vector quantization (VQ) layer.
+    \item In VQ-APC~\parencite{chung_vectorquantized_2020}, a VQ layer is used with the APC objective, which imposes a bottleneck and forces the model to learn better representations.
+    \item Two dropout regularization methods, attention dropout and layer dropout, are introduced to TERA~\parencite{luo_dropout_2021}. 
 \end{itemize}
@@
 
-Autoregressive predictive coding (APC)~\cite{chung_unsupervised_2019,
-chung_generative_2020} \edit{takes inspiration from the classic Linear Predictive Coding (LPC) approach for speech feature extraction~\cite{oshaughnessy_linear_1988} and autoregressive language models (LM) for text, where the model learns to predict future information from past}.
+Autoregressive predictive coding (APC)~\parencite{chung_unsupervised_2019,
+chung_generative_2020} \edit{takes inspiration from the classic Linear Predictive Coding (LPC) approach for speech feature extraction~\parencite{oshaughnessy_linear_1988} and autoregressive language models (LM) for text, where the model learns to predict future information from past}.
 \edit{A function} $f(\cdot)$ reads the input sequence $X_{[1,t]}$ and \edit{outputs a representation sequence} $H_{[1,t]}$.
@@
 However, due to the smoothness of the speech signal, neighboring acoustic features are usually similar.
-Depending on the downstream task, we are often interested in learning so-called \emph{slow features} that typically span multiple input frames~\cite{wiskott_slow_2002}.
-Even the smallest linguistic units of speech---phonemes---span $0.07$~seconds on average in the English TIMIT dataset~\cite{garofolo_timit_1993},   whereas spectrogram frames $\mathbf{x}_t$ are typically computed at $0.01$ second intervals.
+Depending on the downstream task, we are often interested in learning so-called \emph{slow features} that typically span multiple input frames~\parencite{wiskott_slow_2002}.
+Even the smallest linguistic units of speech---phonemes---span $0.07$~seconds on average in the English TIMIT dataset~\parencite{garofolo_timit_1993},   whereas spectrogram frames $\mathbf{x}_t$ are typically computed at $0.01$ second intervals.
 Thus, simply predicting the next frame constitutes a trivial pretext task for APC; the original work finds that $c=3$ performs well.
-In \cite{chung_improved_2020}, the APC objective is extended to multi-target
+In \parencite{chung_improved_2020}, the APC objective is extended to multi-target
 training. The new objective generates both past and future frames conditioned
 on previous context. 
-In VQ-APC~\cite{chung_vectorquantized_2020}, quantization is used with the APC
+In VQ-APC~\parencite{chung_vectorquantized_2020}, quantization is used with the APC
 objective, which imposes an information bottleneck serving as a regularizer.
@@
 and not the entire input.
-DeCoAR~\cite{ling_deep_2020} combines the bidirectionality of the popular NLP model
-ELMo~\cite{peters_deep_2018} and the reconstruction objective of APC to alleviate
+DeCoAR~\parencite{ling_deep_2020} combines the bidirectionality of the popular NLP model
+ELMo~\parencite{peters_deep_2018} and the reconstruction objective of APC to alleviate
 this issue and allow encoding information from the entire input. 
@@
 
-\paragraph{Masked Reconstruction}
+\paragraph{Masked reconstruction}
 
-Masked reconstruction is largely inspired by the masked language model (MLM) task from BERT~\cite{devlin_bert_2018}.
+Masked reconstruction is largely inspired by the masked language model (MLM) task from BERT~\parencite{devlin_bert_2018}.
 During BERT pre-training, some tokens in the input sentences are masked by randomly replacing them by a learned masking token or another input token.
@@
 \edit{The exact masking policy defined by $m(\cdot)$ differs from model to model and will be discussed further below.}
-The function $f(\cdot)$ is typically a Transformer encoder~\cite{liu_mockingjay_2020,jiang_improving_2019,liu_masked_2020}, but recurrent neural networks have also been used~\cite{wang_unsupervised_2020}.
+The function $f(\cdot)$ is typically a Transformer encoder~\parencite{liu_mockingjay_2020,jiang_improving_2019,liu_masked_2020}, but recurrent neural networks have also been used~\parencite{wang_unsupervised_2020}.
 \edit{In general, the Transformer encoder architecture has been adopted widely by self-supervised models for speech within all three surveyed categories.} \edit{The function $g(\cdot)$ is usually a linear projection or a multilayer perceptron (MLP)}.
@@
 
-The masking policies used in NLP can be adapted to speech by considering a speech \edit{segment} equivalent to a token in a sentence; indeed, the masking strategy of BERT has also been used for speech pre-training~\cite{liu_mockingjay_2020}.
+The masking policies used in NLP can be adapted to speech by considering a speech \edit{segment} equivalent to a token in a sentence; indeed, the masking strategy of BERT has also been used for speech pre-training~\parencite{liu_mockingjay_2020}.
 In the standard BERT masking policy, each token is masked independently at random.
-However, for speech, masking a single \edit{sample or spectrogram frame results in a largely} trivial reconstruction task \edit{since, as discussed in \cref{par:apc}, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore it is common to mask chunks of consecutive frames~\cite{liu_mockingjay_2020,jiang_further_2021}.
+However, for speech, masking a single \edit{sample or spectrogram frame results in a largely} trivial reconstruction task \edit{since, as discussed in \cref{par:apc}, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore it is common to mask chunks of consecutive frames~\parencite{liu_mockingjay_2020,jiang_further_2021}.
 
 We can bring the pretext task closer to the NLP equivalent by using a masking policy where the masked regions of the input correspond to linguistic units.
-Instead of just masking a fixed number of consecutive frames, pMPC~\cite{yue_phonetically_2021} selects masked speech frames according to the phonetic segmentation in an utterance.
+Instead of just masking a fixed number of consecutive frames, pMPC~\parencite{yue_phonetically_2021} selects masked speech frames according to the phonetic segmentation in an utterance.
 \edit{However, in order to obtain this segmentation, some labeled data is of course needed.}
 
-Whereas most studies use masking along the temporal dimension of the input, speech can also be masked along the frequency dimension when spectral input features are used~\cite{wang_unsupervised_2020,liu_tera_2021}.
-Frequency masking has been shown to improve representations used for speaker classification~\cite{liu_tera_2021}.
+Whereas most studies use masking along the temporal dimension of the input, speech can also be masked along the frequency dimension when spectral input features are used~\parencite{wang_unsupervised_2020,liu_tera_2021}.
+Frequency masking has been shown to improve representations used for speaker classification~\parencite{liu_tera_2021}.
 
 Some studies explore alternatives to masking the input directly.
-In non-autoregressive predictive coding (NPC)~\cite{liu_nonautoregressive_2020}, time masking is introduced through masked convolution blocks.
-Taking inspiration from XLNet~\cite{yang_xlnet_2019}, it has also been suggested that the input be reconstructed from a shuffled version~\cite{song_speechxlnet_2020} to address the discrepancy between pre-training and fine-tuning of masking-based approaches.
+In non-autoregressive predictive coding (NPC)~\parencite{liu_nonautoregressive_2020}, time masking is introduced through masked convolution blocks.
+Taking inspiration from XLNet~\parencite{yang_xlnet_2019}, it has also been suggested that the input be reconstructed from a shuffled version~\parencite{song_speechxlnet_2020} to address the discrepancy between pre-training and fine-tuning of masking-based approaches.
 
 Regularization methods can further improve on masked reconstruction approaches.
-\edit{DeCoAR 2.0~\cite{ling_decoar_2020} uses vector quantization}, which is shown to improve the learned representations.
-Furthermore, two dropout regularization methods---attention dropout and layer dropout---are introduced with the TERA model~\cite{liu_tera_2021, luo_dropout_2021}.
-\edit{Both methods are variations on the original dropout method \cite{srivastava_dropout_2014}.}
+\edit{DeCoAR 2.0~\parencite{ling_decoar_2020} uses vector quantization}, which is shown to improve the learned representations.
+Furthermore, two dropout regularization methods---attention dropout and layer dropout---are introduced with the TERA model~\parencite{liu_tera_2021, luo_dropout_2021}.
+\edit{Both methods are variations on the original dropout method \parencite{srivastava_dropout_2014}.}
 
-\paragraph{More Generative Approaches}
+\paragraph{More generative approaches}
 Other than the autoregressive and masked reconstruction tasks discussed above, various studies have explored the reconstruction of other targets derived from the input.
-PASE and PASE+~\cite{pascual_learning_2019,ravanelli_multitask_2020} use multiple targets, including the waveform, log power spectrum, \edit{mel cepstral coefficients (MFCCs)}, and prosody features.
-Models that learn acoustic embeddings of small speech segments have targeted future and past spectrogram segments~\cite{chung_speech2vec_2018, tagliasacchi_selfsupervised_2019, tagliasacchi_pretraining_2020}, phase information~\cite{quitry_learning_2019}, and the temporal gap between two segments~\cite{tagliasacchi_selfsupervised_2019, tagliasacchi_pretraining_2020}.
+PASE and PASE+~\parencite{pascual_learning_2019,ravanelli_multitask_2020} use multiple targets, including the waveform, log power spectrum, \edit{mel cepstral coefficients (MFCCs)}, and prosody features.
+Models that learn acoustic embeddings of small speech segments have targeted future and past spectrogram segments~\parencite{chung_speech2vec_2018, tagliasacchi_selfsupervised_2019, tagliasacchi_pretraining_2020}, phase information~\parencite{quitry_learning_2019}, and the temporal gap between two segments~\parencite{tagliasacchi_selfsupervised_2019, tagliasacchi_pretraining_2020}.
 
@@
 \edit{The pretext task is to maximize latent space similarity between the anchor and positive samples while minimizing the similarity between the anchor and negative samples}.
-This approach has been used extensively in the general \edit{ML} community \cite{schultz_learning_2003}.
+This approach has been used extensively in the general \edit{ML} community \parencite{schultz_learning_2003}.
 
@@
 \label{par:cpc}
-Contrastive Predictive Coding (CPC)~\cite{oord_representation_2018} is a prominent example of a contrastive model. 
+Contrastive Predictive Coding (CPC)~\parencite{oord_representation_2018} is a prominent example of a contrastive model. 
 \edit{CPC uses a convolutional module $f_1(\cdot)$ to produce localized representations $z_t$ with a recurrent module $f_2(\cdot)$ on top that outputs a contextualized representation $h_t$.
@@
 The loss function measures the similarity between the anchor representation $\hat{z}_{t,k}$ and the positive $z_{t+k}$ normalized by the total similarity to the positive and negatives.}
-The approach is similar to previous work on Noise-Contrastive Estimation (NCE)~\cite{gutmann_noisecontrastive_2010}.
+The approach is similar to previous work on Noise-Contrastive Estimation (NCE)~\parencite{gutmann_noisecontrastive_2010}.
 \edit{Minimizing} the loss corresponds to maximizing a lower bound on the mutual information between $h_t$ and $z_{t+k}$ (and in turn $x_{t+k-u:t+k+u}$) and is hence called InfoNCE:
@@
 Including the target index in $\mathcal{I}$ ensures that the loss is a proper categorical cross-entropy and that minimizing it has the previously stated relation to mutual information maximization}.
-This corresponds to sampling negatives from the same sequence and has been \edit{shown} to give good performance for phoneme classification~\cite{oord_representation_2018}.
+This corresponds to sampling negatives from the same sequence and has been \edit{shown} to give good performance for phoneme classification~\parencite{oord_representation_2018}.
 The loss is indexed by $k$ to show that CPC targets multiple offsets using different projection layers $g_k(\cdot)$.
@@
 
-The wav2vec model \cite{schneider_wav2vec_2019} extends the CPC approach \edit{and uses fully convolutional parameterizations for the modules $f_1(\cdot)$ and $f_2(\cdot)$ with receptive fields of \SI{30}{ms} and \SI{210}{ms}, respectively.
+The wav2vec model \parencite{schneider_wav2vec_2019} extends the CPC approach \edit{and uses fully convolutional parameterizations for the modules $f_1(\cdot)$ and $f_2(\cdot)$ with receptive fields of \SI{30}{ms} and \SI{210}{ms}, respectively.
 While the CPC loss solves a 1-of-$N$ classification task per $(t,k)$, either assigning the anchor to the positive class or (wrongly) to one of the $N-1$ negative classes, the wav2vec loss considers a sequence of $N$ independent binary classifications.
@@
 \edit{The wav2vec 2.0 model combines contrastive learning with masking}.
-As the CPC model, it uses the InfoNCE loss \cite{oord_representation_2018} to maximize the similarity between a contextualized representation and a localized representation.
+As the CPC model, it uses the InfoNCE loss \parencite{oord_representation_2018} to maximize the similarity between a contextualized representation and a localized representation.
 \edit{However, instead of using the $z_t$ directly as positive and negatives, it uses a quantization module $g(\cdot)$ to obtain a discrete representation.
@@
 \noindent
-The quantization module $g(\cdot)$ uses a Gumbel softmax~\cite{jang_categorical_2016} with a straight-through estimator.
+The quantization module $g(\cdot)$ uses a Gumbel softmax~\parencite{jang_categorical_2016} with a straight-through estimator.
 Since the quality of the learned representations is contingent on the quality of the quantization, wav2vec 2.0 combines two techniques to learn high-quality codebooks.
-First, wav2vec 2.0 concatenates quantized representations from multiple codebooks at each timestep, so-called Product Quantization (PQ)~\cite{jegou_product_2011}.
+First, wav2vec 2.0 concatenates quantized representations from multiple codebooks at each timestep, so-called Product Quantization (PQ)~\parencite{jegou_product_2011}.
 Also, the \edit{primary} training loss \edit{described below} is augmented with an auxiliary term designed to encourage equal use of all codebook entries.
@@
 It has subsequently} inspired many follow-up studies.
-The wav2vec-C~\cite{sadhu_wav2vecc_2021} approach extends wav2vec 2.0 with a consistency term in the loss that aims to reconstruct the input features from the learned quantized representations, similar to VQ-VAE~\cite{razavi_generating_2019}.
+The wav2vec-C~\parencite{sadhu_wav2vecc_2021} approach extends wav2vec 2.0 with a consistency term in the loss that aims to reconstruct the input features from the learned quantized representations, similar to VQ-VAE~\parencite{razavi_generating_2019}.
 
@@
 Thus, the predictive setup is somewhat akin to teacher-student training.
-The first predictive approaches were motivated by the success of BERT-like methods in NLP \cite{devlin_bert_2018} as well as the DeepCluster method in CV \cite{caron_deep_2018}.}
+The first predictive approaches were motivated by the success of BERT-like methods in NLP \parencite{devlin_bert_2018} as well as the DeepCluster method in CV \parencite{caron_deep_2018}.}
 
@@
 \paragraph{Discrete BERT} \edit{Applying BERT-type training directly to speech input is not possible due to its continuous nature}.
-The Discrete BERT approach~\cite{baevski_effectiveness_2020} uses a pre-trained vq-wav2vec model to derive a discrete vocabulary \cite{baevski_vqwav2vec_2020}.
+The Discrete BERT approach~\parencite{baevski_effectiveness_2020} uses a pre-trained vq-wav2vec model to derive a discrete vocabulary \parencite{baevski_vqwav2vec_2020}.
 The vq-wav2vec model is similar to wav2vec mentioned in \cref{par:cpc} but uses quantization to learn discrete representations.
@@
 
-\paragraph{HuBERT} Rather than relying on an advanced representation learning model for discretizing continuous inputs, as Discrete BERT, the Hidden Unit BERT (HuBERT) approach \cite{hsu_hubert_2021} \edit{uses quantized MFCC features as targets learned with classic $k$-means.
+\paragraph{HuBERT} Rather than relying on an advanced representation learning model for discretizing continuous inputs, as Discrete BERT, the Hidden Unit BERT (HuBERT) approach \parencite{hsu_hubert_2021} \edit{uses quantized MFCC features as targets learned with classic $k$-means.
 Thus, to compute the targets, the $k$-means model $g_1(\cdot)$ assigns a cluster center to each timestep.
@@
 \edit{For HuBERT, only two iterations are needed to match or outperform the previous state-of-the-art results for low-resource speech recognition.
-And combining the HuBERT approach with the wav2vec 2.0 approach, the w2v-BERT model has managed to improve results even further~\cite{chung_w2vbert_2021}.}
+And combining the HuBERT approach with the wav2vec 2.0 approach, the w2v-BERT model has managed to improve results even further~\parencite{chung_w2vbert_2021}.}
 
@@
 
-\edit{WavLM emphasizes spoken content modeling and speaker identity preservation~\cite{chen_wavlm_2021}. It is largely identical to HuBERT, but introduces two useful extensions.}
+\edit{WavLM emphasizes spoken content modeling and speaker identity preservation~\parencite{chen_wavlm_2021}. It is largely identical to HuBERT, but introduces two useful extensions.}
 
@@
 
-\paragraph{data2vec} \edit{Motivated by the success of using an exponential moving average (EMA) teacher for self-supervised visual representations~\cite{grill_bootstrap_2020, caron_emerging_2021}, the data2vec model~\cite{baevski_data2vec_2022} computes targets $Y$ using an EMA of its own parameters.
+\paragraph{data2vec} \edit{Motivated by the success of using an exponential moving average (EMA) teacher for self-supervised visual representations~\parencite{grill_bootstrap_2020, caron_emerging_2021}, the data2vec model~\parencite{baevski_data2vec_2022} computes targets $Y$ using an EMA of its own parameters.
 The targets are constructed by averaging hidden representations of the top $k$ layers of the EMA teacher network applied to unmasked inputs.
@@
 The data2vec model uses a regression loss between target and prediction.
-Specifically, to reduce sensitivity to outliers and prevent exploding gradients, it uses the smoothed \ensuremath{L_1} loss \cite{girshick_fast_2015},
+Specifically, to reduce sensitivity to outliers and prevent exploding gradients, it uses the smoothed \ensuremath{L_1} loss \parencite{girshick_fast_2015},
 \begin{equation}
@@
 Another challenge for these models centers around the quality of the initial vocabulary generated from MFCC features.
-The data2vec approach improves over other predictive models \edit{by allowing the targets to improve continuously via the EMA teacher network}; however, \edit{student-teacher approaches inflate the existing computational challenges of very large models and may necessitate the use of methods that decrease instantaneous memory utilization such as mixed precision training, model parallelism and model sharding \cite{paszke_automatic_2017}.}
+The data2vec approach improves over other predictive models \edit{by allowing the targets to improve continuously via the EMA teacher network}; however, \edit{student-teacher approaches inflate the existing computational challenges of very large models and may necessitate the use of methods that decrease instantaneous memory utilization such as mixed precision training, model parallelism and model sharding \parencite{paszke_automatic_2017}.}
 
@@
 Multiple modalities are useful in many settings, where each modality provides information that is complementary to other modalities.
-Multi-modal work includes supervised settings, such as audio-visual ASR~\cite{petajan_automatic_1984,potamianos_recent_2003} and person identification~\cite{aleksic_audiovisual_2006} which have been studied for decades.
+Multi-modal work includes supervised settings, such as audio-visual ASR~\parencite{petajan_automatic_1984,potamianos_recent_2003} and person identification~\parencite{aleksic_audiovisual_2006} which have been studied for decades.
 In this section, we focus only on unsupervised representation learning from multi-modal data.
@@
 Such ``grounding" signals can contain supplementary information that can be used by models to infer the content of the speech.
-Human language learning provides a proof of concept for this, as it is believed that infants benefit from the visual modality when learning language~\cite{legerstee_infants_1990}.
-Early computational models of multi-modal language learning were motivated by (and tried to emulate) human learning of language in the context of the visual surroundings~\cite{roy_learning_1999}.
+Human language learning provides a proof of concept for this, as it is believed that infants benefit from the visual modality when learning language~\parencite{legerstee_infants_1990}.
+Early computational models of multi-modal language learning were motivated by (and tried to emulate) human learning of language in the context of the visual surroundings~\parencite{roy_learning_1999}.
 
@@
 modalities produced directly by the speech source.
-Examples of intrinsic modalities (besides the speech audio) include images or video of the speaker's face~\cite{lee_avicar_2004,chung_lip_2016}, lip-movement~\cite{shi_learning_2022}, articulatory flesh point measurements~\cite{westbury_xray_1990,wrench_new_2001}, or simultaneous \edit{magnetic resonance imaging (MRI)} scans~\cite{narayanan_multimodal_2011}.
+Examples of intrinsic modalities (besides the speech audio) include images or video of the speaker's face~\parencite{lee_avicar_2004,chung_lip_2016}, lip-movement~\parencite{shi_learning_2022}, articulatory flesh point measurements~\parencite{westbury_xray_1990,wrench_new_2001}, or simultaneous \edit{magnetic resonance imaging (MRI)} scans~\parencite{narayanan_multimodal_2011}.
 Typically, learning from multiple intrinsic modalities is done so as to improve robustness to noise, since acoustic noise is likely to be uncorrelated with the other \edit{modality(ies)}.
@@
 \begin{itemize}
-    \item Multi-view autoencoders and variations~\cite{ngiam_multimodal_2011,badino_integrating_2016},
-    \item Multi-modal deep Boltzmann machines~\cite{srivastava_multimodal_2012},
-    \item Canonical correlation analysis (CCA)~\cite{hotelling_relations_1936} and its nonlinear extensions~\cite{andrew_deep_2013,wang_deep_2015,wang_deep_2016,michaeli_nonparametric_2016,melzer_nonlinear_2001,lai_kernel_2000,lai_neural_1999,bach_probabilistic_2005,wang_unsupervised_2015},
-    \item Multi-view contrastive losses~\cite{hermann_multilingual_2013,huang_learning_2013},
-    \item More recently, audio-visual extensions of masked prediction methods~\cite{shi_learning_2022,shi_robust_2022}, specifically Audio-Visual HuBERT (AV-HuBERT)~\cite{shi_learning_2022}.
+    \item Multi-view autoencoders and variations~\parencite{ngiam_multimodal_2011,badino_integrating_2016},
+    \item Multi-modal deep Boltzmann machines~\parencite{srivastava_multimodal_2012},
+    \item Canonical correlation analysis (CCA)~\parencite{hotelling_relations_1936} and its nonlinear extensions~\parencite{andrew_deep_2013,wang_deep_2015,wang_deep_2016,michaeli_nonparametric_2016,melzer_nonlinear_2001,lai_kernel_2000,lai_neural_1999,bach_probabilistic_2005,wang_unsupervised_2015},
+    \item Multi-view contrastive losses~\parencite{hermann_multilingual_2013,huang_learning_2013},
+    \item More recently, audio-visual extensions of masked prediction methods~\parencite{shi_learning_2022,shi_robust_2022}, specifically Audio-Visual HuBERT (AV-HuBERT)~\parencite{shi_learning_2022}.
 \end{itemize}
@@
 There has recently been \edit{a} surge of datasets collected for this purpose, usually consisting of images and spoken captions, the audio and image frames in a video, or video clips with their spoken descriptions.
-A recent review of datasets, as well as methods, in this category is provided by Chrupa\l{}a~\cite{chrupala_visually_2021}.
+A recent review of datasets, as well as methods, in this category is provided by Chrupa\l{}a~\parencite{chrupala_visually_2021}.
 
-Typical approaches involve learning a neural representation model for each modality, with a multi-modal contrastive loss that encourages paired examples in the two modalities to have similar representations while unpaired examples remain different~\cite{synnaeve_learning_2014,harwath_unsupervised_2016,harwath_deep_2015,merkx_language_2019,rouditchenko_avlnet_2021,peng_fastslow_2022}
+Typical approaches involve learning a neural representation model for each modality, with a multi-modal contrastive loss that encourages paired examples in the two modalities to have similar representations while unpaired examples remain different~\parencite{synnaeve_learning_2014,harwath_unsupervised_2016,harwath_deep_2015,merkx_language_2019,rouditchenko_avlnet_2021,peng_fastslow_2022}
 
-Other choices include training with a masked margin softmax loss~\cite{ilharco_largescale_2019,sanabria_talk_2021} or a masked prediction loss~\cite{chan_multimodal_2022}.
-Such models are typically evaluated on cross-modal retrieval, although some work has also used the models for other downstream tasks such as the ZeroSpeech and SUPERB benchmark tasks~\cite{peng_selfsupervised_2022}.
-Analyses of such models have found that, despite the very high-level learning objective of matching speech with a corresponding image (or other contextual modality), such models often learn multiple levels of linguistic representations from the shallowest to the deepest model layers~\cite{harwath_learning_2019,chrupala_representations_2017,scharenborg_linguistic_2018}.
-They are also able to learn word-like units~\cite{harwath_jointly_2018,peng_word_2022,wang_dnnhmmdnn_2020} and can be used for cross-lingual retrieval, by considering the visual signal as an ``interlingua"~\cite{harwath_vision_2018,havard_models_2019,kamper_visually_2018}.
-In some settings, even in the presence of some amount of textual supervision (i.e., the speech is transcribed), visual grounding still helps learn a better representation for retrieval~\cite{pasad_contributions_2019}.
+Other choices include training with a masked margin softmax loss~\parencite{ilharco_largescale_2019,sanabria_talk_2021} or a masked prediction loss~\parencite{chan_multimodal_2022}.
+Such models are typically evaluated on cross-modal retrieval, although some work has also used the models for other downstream tasks such as the ZeroSpeech and SUPERB benchmark tasks~\parencite{peng_selfsupervised_2022}.
+Analyses of such models have found that, despite the very high-level learning objective of matching speech with a corresponding image (or other contextual modality), such models often learn multiple levels of linguistic representations from the shallowest to the deepest model layers~\parencite{harwath_learning_2019,chrupala_representations_2017,scharenborg_linguistic_2018}.
+They are also able to learn word-like units~\parencite{harwath_jointly_2018,peng_word_2022,wang_dnnhmmdnn_2020} and can be used for cross-lingual retrieval, by considering the visual signal as an ``interlingua"~\parencite{harwath_vision_2018,havard_models_2019,kamper_visually_2018}.
+In some settings, even in the presence of some amount of textual supervision (i.e., the speech is transcribed), visual grounding still helps learn a better representation for retrieval~\parencite{pasad_contributions_2019}.
 
 There has also been growing interest in learning joint speech and text representations using paired and unpaired data.
-The SLAM approach~\cite{bapna_slam_2021} is an example where speech and text are first represented using two separate pre-trained encoders followed by a multi-modal encoder to build the joint representations.
+The SLAM approach~\parencite{bapna_slam_2021} is an example where speech and text are first represented using two separate pre-trained encoders followed by a multi-modal encoder to build the joint representations.
 The entire model is trained using a multi-task loss including two supervised and two self-supervised tasks.
@@
 
-\subsection{Acoustic Word Embeddings}
+\subsection{Acoustic word embeddings}
 
@@
 For some purposes, however, it may be useful to explicitly represent longer spans of speech audio of arbitrary duration, such as phone, word, or phrase-level segments.
-For example, searching within a corpus of recorded speech for segments that match a given (written or spoken) query can be seen as finding segments whose representations are most similar to that of the query~\cite{levin_segmental_2015,chen_querybyexample_2015,chung_audio_2016,settle_querybyexample_2017}; word embeddings can be defined by pooling representations of instances of a given word~\cite{chung_speech2vec_2018}; unsupervised segmentation and spoken term discovery can be seen as a problem of detecting and clustering segments~\cite{kamper_embedded_2017,kamper_segmental_2017}; and even ASR can be viewed as the problem of matching written word representations to representations of audio spans~\cite{maas_wordlevel_2012,bengio_word_2014,settle_acoustically_2019}.
+For example, searching within a corpus of recorded speech for segments that match a given (written or spoken) query can be seen as finding segments whose representations are most similar to that of the query~\parencite{levin_segmental_2015,chen_querybyexample_2015,chung_audio_2016,settle_querybyexample_2017}; word embeddings can be defined by pooling representations of instances of a given word~\parencite{chung_speech2vec_2018}; unsupervised segmentation and spoken term discovery can be seen as a problem of detecting and clustering segments~\parencite{kamper_embedded_2017,kamper_segmental_2017}; and even ASR can be viewed as the problem of matching written word representations to representations of audio spans~\parencite{maas_wordlevel_2012,bengio_word_2014,settle_acoustically_2019}.
 
 Several lines of work have begun to address the problem of learning representations of spans of speech, especially word segments, typically referred to as \textit{acoustic word embeddings}.
-Early work on unsupervised acoustic word embeddings defined them as vectors of distances from the target segment to a number of pre-defined ``template" segments~\cite{levin_fixeddimensional_2013}.
-Later work used variants of neural autoencoders~\cite{chung_audio_2016,holzenberger_learning_2018,kamper_truly_2019,peng_correspondence_2020}.
-These are often evaluated on word discrimination, that is the task of determining whether two word segments correspond to the same word or not~\cite{carlin_rapid_2011}.
-This task can be thought of as a proxy for query-by-example search, since the basic operation in search is to determine whether a segment in the search database matches a query segment, and has been used for evaluation of both frame-level (e.g.,~\cite{kamper_unsupervised_2015}) and word-level~\cite{levin_fixeddimensional_2013,kamper_deep_2016} representations.
+Early work on unsupervised acoustic word embeddings defined them as vectors of distances from the target segment to a number of pre-defined ``template" segments~\parencite{levin_fixeddimensional_2013}.
+Later work used variants of neural autoencoders~\parencite{chung_audio_2016,holzenberger_learning_2018,kamper_truly_2019,peng_correspondence_2020}.
+These are often evaluated on word discrimination, that is the task of determining whether two word segments correspond to the same word or not~\parencite{carlin_rapid_2011}.
+This task can be thought of as a proxy for query-by-example search, since the basic operation in search is to determine whether a segment in the search database matches a query segment, and has been used for evaluation of both frame-level (e.g.,~\parencite{kamper_unsupervised_2015}) and word-level~\parencite{levin_fixeddimensional_2013,kamper_deep_2016} representations.
 
-Since most work on acoustic word embeddings preceded the very recent wave of new self-supervised frame-level representations, one question is whether word (or more generally segment) embeddings could be derived more simply by pooling self-supervised frame-level representations, as has been done for text span embeddings by pooling over word embeddings~\cite{toshniwal_crosstask_2020,wang_phrasebert_2021}.
-Some initial results suggest that at least very simple pooling approaches like downsampling and mean or max pooling are not successful~\cite{vanstaden_comparison_2021,peng_correspondence_2020}, but more work is needed to reach conclusive results.
+Since most work on acoustic word embeddings preceded the very recent wave of new self-supervised frame-level representations, one question is whether word (or more generally segment) embeddings could be derived more simply by pooling self-supervised frame-level representations, as has been done for text span embeddings by pooling over word embeddings~\parencite{toshniwal_crosstask_2020,wang_phrasebert_2021}.
+Some initial results suggest that at least very simple pooling approaches like downsampling and mean or max pooling are not successful~\parencite{vanstaden_comparison_2021,peng_correspondence_2020}, but more work is needed to reach conclusive results.

### 2023-09-29 chapters/paper_review/04_benchmarking.tex
Move table captions to top
Updates to citation commands and footnotes

@@
 in the literature. These datasets are usually large but with limited or no
-labels. Libri-light (LL)~\cite{kahn_libri-light_2020}, one of these datasets, is
+labels. Libri-light (LL)~\parencite{kahn_libri-light_2020}, one of these datasets, is
 derived from audiobooks that are part of the
@@
 60k hours of spoken English audio tagged with SNR, speaker ID, and genre
-descriptions. The speech examples in Audioset~\cite{gemmeke_audio_2017}, which
+descriptions. The speech examples in Audioset~\parencite{gemmeke_audio_2017}, which
 consists of over 2M 10-second YouTube video clips human-annotated with 632
@@
 of audio of varying quality, different languages, and sometimes multiple sound
-sources. AVSpeech~\cite{ephrat_looking_2018} is another large-scale audio-visual
+sources. AVSpeech~\parencite{ephrat_looking_2018} is another large-scale audio-visual
 dataset used in SSL research, comprising 4.7k hours of clips from a wide
@@
 AVSpeech has been used to learn audio-only 
-representations~\cite{kawakami_learning_2020}. The Fisher corpus~\cite{cieri_fisher_2004} collects
+representations~\parencite{kawakami_learning_2020}. The Fisher corpus~\parencite{cieri_fisher_2004} collects
 over 2k hours of conversational telephone speech, 1k hours of which is utilized
-for pre-training~\cite{jiang_further_2021}. Industrial researchers have also
+for pre-training~\parencite{jiang_further_2021}. Industrial researchers have also
 begun to build large-scale datasets for learning speech representations.
@@
 self-supervised pre-training have been collected at 
-Amazon~\cite{sadhu_wav2vecc_2021}.
+Amazon~\parencite{sadhu_wav2vecc_2021}.
 
@@
 collected corpora for pre-training Chinese speech representations. Didi Dictation
-and Didi Callcenter~\cite{jiang_improving_2019, jiang_further_2021} are two
+and Didi Callcenter~\parencite{jiang_improving_2019, jiang_further_2021} are two
 internal datasets containing respectively 10k hours of read speech collected
@@
 \cref{table:datasets}. One of the most commonly used datasets in this category
-is LibriSpeech (LS)~\cite{panayotov_librispeech_2015}, a labeled corpus
+is LibriSpeech (LS)~\parencite{panayotov_librispeech_2015}, a labeled corpus
 containing 960 hours of
@@
 can also be utilized to evaluate the performance of representation on ASR, phoneme recognition (PR),
-phoneme classification (PC), and speaker identification (SID) tasks. Wall Street Journal (WSJ)~\cite{paul_design_1992} is another
+phoneme classification (PC), and speaker identification (SID) tasks. Wall Street Journal (WSJ)~\parencite{paul_design_1992} is another
 widely adopted, labeled corpus for pre-training. Its labels can evaluate
@@
 evaluation (PC and PR can be enabled via phone-level forced alignment). These
-corpora include Common Voice (CV)~\cite{ardila_common_2020}, Multilingual
-LibriSpeech (MLS)~\cite{pratap_mls_2020}, VoxPopuli 
-(VP)~\cite{wang_voxpopuli_2021}, and BABEL (BBL)~\cite{gales_speech_2014}. CV is
+corpora include Common Voice (CV)~\parencite{ardila_common_2020}, Multilingual
+LibriSpeech (MLS)~\parencite{pratap_mls_2020}, VoxPopuli 
+(VP)~\parencite{wang_voxpopuli_2021}, and BABEL (BBL)~\parencite{gales_speech_2014}. CV is
 an open-source, multi-language, growing dataset of voices containing 11k hours
@@
 corpus~7.0). Researchers usually use part of this for pre-training (e.g., 7k
-hours/60 languages in \cite{babu_xlsr_2021} and 430 hours/29 languages in
-\cite{kawakami_learning_2020}) or evaluation. 
+hours/60 languages in \parencite{babu_xlsr_2021} and 430 hours/29 languages in
+\parencite{kawakami_learning_2020}) or evaluation. 
 MLS derives content from read audiobooks of LibriVox and contains data in eight
@@
 languages.
-The entire dataset~\cite{babu_xlsr_2021} or a 24k-hour 
-portion~\cite{chen_unispeechsat_2021, chen_wavlm_2021} thereof has been used for pre-training. BBL
+The entire dataset~\parencite{babu_xlsr_2021} or a 24k-hour 
+portion~\parencite{chen_unispeechsat_2021, chen_wavlm_2021} thereof has been used for pre-training. BBL
 consists of 1k hours of conversational telephone speech in 17 African and Asian
@@
 
-Several datasets, including GigaSpeech~\cite{chen_gigaspeech_2021}, TED-LIUM~3
-(TED3)~\cite{hernandez_tedlium_2018}, TED-LIUM~2 (TED2)~\cite{rousseau_tedlium_2012},
-Switchboard (SWB)~\cite{godfrey_switchboard_1992}, 
-TIMIT~\cite{garofolo_timit_1993}, and VoxLingua107~\cite{valk_voxlingua107_2021}, are
+Several datasets, including GigaSpeech~\parencite{chen_gigaspeech_2021}, TED-LIUM~3
+(TED3)~\parencite{hernandez_tedlium_2018}, TED-LIUM~2 (TED2)~\parencite{rousseau_tedlium_2012},
+Switchboard (SWB)~\parencite{godfrey_switchboard_1992}, 
+TIMIT~\parencite{garofolo_timit_1993}, and VoxLingua107~\parencite{valk_voxlingua107_2021}, are
 labeled and conventionally used for evaluation, while their audio streams are
 also aggregated to build diversified and large-scale corpora for unsupervised
-pre-training~\cite{kawakami_learning_2020, song_speechxlnet_2020, babu_xlsr_2021}.
+pre-training~\parencite{kawakami_learning_2020, song_speechxlnet_2020, babu_xlsr_2021}.
 GigaSpeech is a multi-domain English ASR corpus with 33k hours of audio
@@
 For the purpose of pre-training and evaluating Mandarin speech representations,
-the authors of \cite{jiang_improving_2019, jiang_further_2021} also compiled Open Mandarin, 
+the authors of \parencite{jiang_improving_2019, jiang_further_2021} also compiled Open Mandarin, 
 an open-source Mandarin dataset of 1.5k hours of speech from
@@
 168 hours of audio is used for pre-training; the development and test sets are
-excluded.)~\cite{liu_hkust_2006}, AISHELL-1~\cite{bu_aishell1_2017} (178 hours of
+excluded.)~\parencite{liu_hkust_2006}, AISHELL-1~\parencite{bu_aishell1_2017} (178 hours of
 read speech),
-aidatatang 200zh (200 hours, read speech)~\cite{beijingdatatangtechnologycoltd_aidatatang_}, MAGICDATA
-Mandarin Chinese Read Speech Corpus (755 hours, read speech)~\cite{magicdatatechnologycoltd_magicdata_2019},
-Free ST Chinese Mandarin Corpus (ST-CMDS, 100 hours, read speech)~\cite{surfingtechcoltd_stcmds_},
-and Primewords Chinese Corpus Set 1 (100 hours, read speech)~\cite{primewordsinformationtechnologycoltd_primewords_2018}. 
+aidatatang 200zh (200 hours, read speech)~\parencite{beijingdatatangtechnologycoltd_aidatatang_}, MAGICDATA
+Mandarin Chinese Read Speech Corpus (755 hours, read speech)~\parencite{magicdatatechnologycoltd_magicdata_2019},
+Free ST Chinese Mandarin Corpus (ST-CMDS, 100 hours, read speech)~\parencite{surfingtechcoltd_stcmds_},
+and Primewords Chinese Corpus Set 1 (100 hours, read speech)~\parencite{primewordsinformationtechnologycoltd_primewords_2018}. 
 Both HKUST and AISHELL-1 are labeled and are suitable for ASR evaluation.
@@
 SWB studies but not released with the SWB dataset, and the rest are from CallHome
-American English Speech (LDC97S42). DIRHA~\cite{ravanelli_dirhaenglish_2015}, short for
+American English Speech (LDC97S42). DIRHA~\parencite{ravanelli_dirhaenglish_2015}, short for
 Distant-speech Interaction for Robust Home Applications, is a database composed
@@
 phonetically-rich sentences. These utterances are read by UK and US English
-speakers and recorded with microphone arrays. CHiME-5~\cite{barker_fifth_2018} is
+speakers and recorded with microphone arrays. CHiME-5~\parencite{barker_fifth_2018} is
 a challenge that aims to advance robust ASR and presents a dataset of natural
@@
 arrays. A team at Amazon Alexa also recorded and transcribed a corpus of 1k
-hours of audio for model training and evaluation~\cite{sadhu_wav2vecc_2021}.
+hours of audio for model training and evaluation~\parencite{sadhu_wav2vecc_2021}.
 
 Researchers also evaluate representations for sentiment analysis with the
-INTERFACE~\cite{hozjan_interface_2002} and MOSEI (CMU Multimodal Opinion
-Sentiment and Emotion Intensity)~\cite{zadeh_multimodal_2018} datasets. INTERFACE
+INTERFACE~\parencite{hozjan_interface_2002} and MOSEI (CMU Multimodal Opinion
+Sentiment and Emotion Intensity)~\parencite{zadeh_multimodal_2018} datasets. INTERFACE
 is an emotional speech database for Slovenian, English, Spanish, and French,
@@
 In addition, datasets employed to demonstrate the benefit of SSL
-representations on various tasks include VCTK \cite{veaux_cstr_2016} and VoxCeleb1 \cite{nagrani_voxceleb_2017} for SID/ASV (automatic speaker verification) tasks, FSC (Fluent Speech Commands) \cite{lugosch_speech_2019} for IC (intent classification), QUESST
-(QUESST 2014) \cite{anguera_quesst2014_2015} for QbE (query by example), LS En-Fr \cite{kocabiyikoglu_augmenting_2018} and CoVoST-2 \cite{wang_covost_2020} for ST (speech translation), and ALFFA and OpenSLR-multi for multilingual ASR.
+representations on various tasks include VCTK \parencite{veaux_cstr_2016} and VoxCeleb1 \parencite{nagrani_voxceleb_2017} for SID/ASV (automatic speaker verification) tasks, FSC (Fluent Speech Commands) \parencite{lugosch_speech_2019} for IC (intent classification), QUESST
+(QUESST 2014) \parencite{anguera_quesst2014_2015} for QbE (query by example), LS En-Fr \parencite{kocabiyikoglu_augmenting_2018} and CoVoST-2 \parencite{wang_covost_2020} for ST (speech translation), and ALFFA and OpenSLR-multi for multilingual ASR.
 The VCTK corpus includes speech
@@
 languages to promote the development of speech technologies in Africa, and
-\cite{kawakami_learning_2020} leverages four African languages collected in the
-project for evaluation: Amharic~\cite{tachbelie_using_2014}, Fongbe~\cite{laleye_first_2016},
-Swahili~\cite{gelas_developments_2012}, and Wolof~\cite{gauthier_collecting_2016}.
-In the same work~\cite{kawakami_learning_2020}, the authors
+\parencite{kawakami_learning_2020} leverages four African languages collected in the
+project for evaluation: Amharic~\parencite{tachbelie_using_2014}, Fongbe~\parencite{laleye_first_2016},
+Swahili~\parencite{gelas_developments_2012}, and Wolof~\parencite{gauthier_collecting_2016}.
+In the same work~\parencite{kawakami_learning_2020}, the authors
 further select 21 phonetically diverse languages from OpenSLR to evaluate the
@@
 
-Last, \cite{tagliasacchi_pretraining_2020} puts together five datasets 
-(MUSAN~\cite{snyder_musan_2015}, Bird Audio Detection~\cite{stowell_automatic_2019},
-Speech Commands~\cite{warden_speech_2018}, Spoken Language 
-Identification~\cite{oponowicz_spoken_2018}, and TUT Urban Acoustic Scenes 2018~\cite{mesaros_detection_2018}) plus
+Last, \parencite{tagliasacchi_pretraining_2020} puts together five datasets 
+(MUSAN~\parencite{snyder_musan_2015}, Bird Audio Detection~\parencite{stowell_automatic_2019},
+Speech Commands~\parencite{warden_speech_2018}, Spoken Language 
+Identification~\parencite{oponowicz_spoken_2018}, and TUT Urban Acoustic Scenes 2018~\parencite{mesaros_detection_2018}) plus
 an SID task built with the LS \textit{train-clean-100} subset to evaluate the
 capability of representations on audio event detection.
-\cite{quitry_learning_2019} employs the NSynth dataset~\cite{engel_neural_2017} on top
+\parencite{quitry_learning_2019} employs the NSynth dataset~\parencite{engel_neural_2017} on top
 of the six for benchmarking. As many of the datasets are built for research in
@@
 
-\begin{table*}[ht]
+\begin{table*}[t]
   \centering
@@
   sizes of the datasets are provided in columns 3 and 4. Column 5 lists the tasks each
-  dataset is used to evaluate. We use the following abbreviations:
-  \textbf{EN}: English; 
-  \textbf{Multi}: multilingual;
-  \textbf{ZH}: Chinese;
-  \textbf{Fr}: French; \textbf{ASR}: automatic speech recognition; \textbf{PR}:
-  phoneme recognition; \textbf{PC}: phoneme classification; \textbf{SID}:
-  speaker identification; \textbf{ASV}: automatic speaker verification;
-  \textbf{Sentiment}: sentiment analysis; \textbf{ST}: speech translation;
-  \textbf{QbE}: query by example or spoken term detection; \textbf{IC}: intent
-  classification; \textbf{AED}: audio event detection; and \textbf{LID}: language
-  identification. We distinguish \textbf{PR} from \textbf{PC} based on whether
-  the inference is made at the phone level sequentially or the frame level separately.
-  \textbf{SID} and \textbf{ASV} both evaluate model capability in encoding
-  speaker information; \textbf{SID} classifies one utterance into a pre-defined
-  set of speaker labels, whereas \textbf{ASV} infers whether a given pair of
-  utterances was uttered by the same speaker.}
+  dataset is used to evaluate. Abbrevations are listed in \cref{footnote:table:datasets-abbrevations}.}
   \label{table:datasets}
@@
     AVSpeech & PT & Multi & 3.1k & - & CC BY 4.0 \\ \hline %Bidir-CPC
-    Fisher & PT & EN & 2k/1k \cite{jiang_further_2021}  & - & Linguistic Data Consortium (LDC) \\ \hline % MPC
+    Fisher & PT & EN & 2k/1k \parencite{jiang_further_2021}  & - & Linguistic Data Consortium (LDC) \\ \hline % MPC
 
@@
 
-    Common Voice (CV-dataset) & PT/EV & Multi & 11k/7k \cite{babu_xlsr_2021}/430 \cite{kawakami_learning_2020} & ASR/PR/PC & CC0 \\ \hline % mention the split, CPC-modified, Bidir-CPC
+    Common Voice (CV-dataset) & PT/EV & Multi & 11k/7k \parencite{babu_xlsr_2021}/430 \parencite{kawakami_learning_2020} & ASR/PR/PC & CC0 \\ \hline % mention the split, CPC-modified, Bidir-CPC
     Multilingual LS (MLS) & PT/EV & Multi & 50k & ASR & CC BY 4.0 \\ \hline
-    VoxPopuli (VP) & PT/EV & Multi & 400k \cite{babu_xlsr_2021}/24k \cite{chen_unispeechsat_2021, chen_wavlm_2021} & ASR & CC0 \\ \hline
+    VoxPopuli (VP) & PT/EV & Multi & 400k \parencite{babu_xlsr_2021}/24k \parencite{chen_unispeechsat_2021, chen_wavlm_2021} & ASR & CC0 \\ \hline
     BABEL (BBL) & PT/EV & Multi & 1k & ASR & IARPA Babel Agreement  \\ \hline
 
-    GigaSpeech & PT/EV & EN & 40k/10k \cite{chen_unispeechsat_2021, chen_wavlm_2021} & ASR & Apache-2.0 License \\ \hline
+    GigaSpeech & PT/EV & EN & 40k/10k \parencite{chen_unispeechsat_2021, chen_wavlm_2021} & ASR & Apache-2.0 License \\ \hline
     TED-LIUM 3 (TED3) & PT/EV & EN & 450 & ASR & CC BY-NC-ND 3.0 \\ \hline %Bidir-CPC
@@
     VoxCeleb1 & EV & Multi & 352 & SID/ASV & CC BY 4.0 \\ \hline
-    Fluent Speech Commands (FSC) & EV & EN & 14.7 & IC & CC BY-NC-ND 4.0 \\ \hline
+    Fluent Speech Commands & EV & EN & 14.7 & IC & CC BY-NC-ND 4.0 \\ \hline
     QUESST 2014 (QUESST) & EV & Multi & 23 & QbE & No information \\ \hline
@@
     ALFFA & EV & Multi & 5.2--18.3 & ASR-multi & MIT License \\ \hline %Bidir-CPC
-    OpenSLR-multi & EV & Multi & 4.4--265.9 & ASR-multi & \makecell[l]{CC BY-SA 3.0 US, CC BY-SA 4.0, CC BY 4.0, \\ CC BY-NC-ND 4.0, Apache License v.2.0}\\ \hline %Bidir-CPC
-    AED datasets & EV & - & - & AED & \makecell[l]{CC BY 4.0 (MUSAN, Speech Commands, NSynth, Bird Audio\\ Detection),  CC0 (Spoken Language Identification),\\ Non-Commercial (TUT)} \\
+    OpenSLR-multi & EV & Multi & 4.4--265.9 & ASR-multi & \makecell[l]{CC BY-SA 3.0 US, CC BY-SA 4.0, \\CC BY 4.0, CC BY-NC-ND 4.0, \\ Apache License v.2.0}\\ \hline %Bidir-CPC
+    AED datasets & EV & - & - & AED & \makecell[l]{CC BY 4.0 (MUSAN, Speech Commands, \\ NSynth, Bird Audio Detection), CC0 (Spoken \\ Language Identification), Non-Commercial \\ (TUT)} \\
     \bottomrule
@@
 \end{table*}
+
+\addtocounter{footnote}{1}
+\footnotetext{\label{footnote:table:datasets-abbrevations}We use the following abbreviations:
+\textbf{EN}: English; 
+\textbf{Multi}: multilingual;
+\textbf{ZH}: Chinese;
+\textbf{Fr}: French; \textbf{ASR}: automatic speech recognition; \textbf{PR}:
+phoneme recognition; \textbf{PC}: phoneme classification; \textbf{SID}:
+speaker identification; \textbf{ASV}: automatic speaker verification;
+\textbf{Sentiment}: sentiment analysis; \textbf{ST}: speech translation;
+\textbf{QbE}: query by example or spoken term detection; \textbf{IC}: intent
+classification; \textbf{AED}: audio event detection; and \textbf{LID}: language
+identification. We distinguish \textbf{PR} from \textbf{PC} based on whether
+the inference is made at the phone level sequentially or the frame level separately.
+\textbf{SID} and \textbf{ASV} both evaluate model capability in encoding
+speaker information; \textbf{SID} classifies one utterance into a pre-defined
+set of speaker labels, whereas \textbf{ASV} infers whether a given pair of
+utterances was uttered by the same speaker.}
 
@@
     % CPC
-    \multirow{2}{*}[0mm]{CPC \cite{oord_representation_2018}} & \multirow{2}{*}[0mm]{LS 100 hrs} & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{CPC \parencite{oord_representation_2018}} & \multirow{2}{*}[0mm]{LS 100 hrs} & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
     & & SID & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline
     % PASE
-    \multirow{2}{*}[-5mm]{PASE\cite{pascual_learning_2019}} & \multirow{2}{*}[-5mm]{LS 50 hrs \cite{ravanelli_learning_2018}} & SID & VCTK\textsuperscript{\ref{footnote:official-split}} & VCTK\textsuperscript{\ref{footnote:official-split}} & \checkmark & 44 hrs \\ \cline{3-7}
+    \multirow{2}{*}[-5mm]{PASE\parencite{pascual_learning_2019}} & \multirow{2}{*}[-5mm]{LS 50 hrs \parencite{ravanelli_learning_2018}} & SID & VCTK\textsuperscript{\ref{footnote:official-split}} & VCTK\textsuperscript{\ref{footnote:official-split}} & \checkmark & 44 hrs \\ \cline{3-7}
     & & Sentiment & INTERFACE\textsuperscript{\ref{footnote:interface-split}} & INTERFACE\textsuperscript{\ref{footnote:interface-split}} & \checkmark & 3 hrs \\ \cline{3-7}
@@
     % Audio2Vec
-    Audio2Vec \cite{tagliasacchi_pretraining_2020} & AudioSet & AED & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & \checkmark & See \cite{tagliasacchi_pretraining_2020} for details \\ \hline
+    Audio2Vec \parencite{tagliasacchi_pretraining_2020} & AudioSet & AED & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & 6 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & \checkmark & See \parencite{tagliasacchi_pretraining_2020} for details \\ \hline
     % APC
-    \multirow{3}{*}[0mm]{APC \cite{chung_unsupervised_2019, chung_generative_2020}} & \multirow{3}{*}[0mm]{LS 360 hrs} & ASR & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & WSJ dev93 & \checkmark & 72 hrs \\ \cline{3-7}
+    \multirow{3}{*}[0mm]{APC \parencite{chung_unsupervised_2019, chung_generative_2020}} & \multirow{3}{*}[0mm]{LS 360 hrs} & ASR & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & WSJ dev93 & \checkmark & 72 hrs \\ \cline{3-7}
     & & ST & LS En-Fr\textsuperscript{\ref{footnote:official-split}} & LS En-Fr\textsuperscript{\ref{footnote:official-split}} & - & 236 hrs \\ \cline{3-7}
@@
     % wav2vec
-    \multirow{2}{*}[0mm]{wav2vec \cite{schneider_wav2vec_2019}} & \multirow{2}{*}[-0.5mm]{\makecell[l]{LS 80/960 hrs,\\ LS 960 hrs \\ \:+ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & \checkmark & 81 hrs\vspace{1mm} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{wav2vec \parencite{schneider_wav2vec_2019}} & \multirow{2}{*}[-0.5mm]{\makecell[l]{LS 80/960 hrs,\\ LS 960 hrs \\ \:+ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & \checkmark & 81 hrs\vspace{1mm} \\ \cline{3-7}
     & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hr\vspace{1mm} \\ \hline
     % Phase predict
-    PhasePredict \cite{quitry_learning_2019} & AudioSet & AED & 7 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & 7 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & \checkmark & See \cite{quitry_learning_2019} for details \\ \hline
+    PhasePredict \parencite{quitry_learning_2019} & AudioSet & AED & 7 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & 7 AED datasets\textsuperscript{\ref{footnote:AED-datasets}} & \checkmark & See \parencite{quitry_learning_2019} for details \\ \hline
     % Bidir-CPC
-    \multirow{3}{*}{Bidir-CPC\cite{kawakami_learning_2020}} & \multirow{3}{*}{\makecell[l]{LS 960 hrs,\\ CPC-8k\textsuperscript{\ref{footnote:CPC-8k}}}} & ASR & \makecell[l]{WSJ si284,\\LS 960 hrs,\\TED3\textsuperscript{\ref{footnote:official-split}}} & \makecell[l]{WSJ eval92,\\LS test-clean,\\LS test-other,\\TED3\textsuperscript{\ref{footnote:official-split}}, SWB\textsuperscript{\ref{footnote:official-split}}}  & \makecell[l]{Different\\ datasets for\\ training\\ and test} & 81/960/450 hrs \\ \cline{3-7} % cross domain transferring, SWB eval2000, TED3 training set size = 450 hrs
+    \multirow{3}{*}{Bidir-CPC\parencite{kawakami_learning_2020}} & \multirow{3}{*}{\makecell[l]{LS 960 hrs,\\ CPC-8k\textsuperscript{\ref{footnote:CPC-8k}}}} & ASR & \makecell[l]{WSJ si284,\\LS 960 hrs,\\TED3\textsuperscript{\ref{footnote:official-split}}} & \makecell[l]{WSJ eval92,\\LS test-clean,\\LS test-other,\\TED3\textsuperscript{\ref{footnote:official-split}}, SWB\textsuperscript{\ref{footnote:official-split}}}  & \makecell[l]{Different\\ datasets for\\ training\\ and test} & 81/960/450 hrs \\ \cline{3-7} % cross domain transferring, SWB eval2000, TED3 training set size = 450 hrs
     & & \makecell[l]{ASR-multi} & ALFFA\textsuperscript{\ref{footnote:official-split}} & ALFFA\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 languages, 5.2--18.3 hrs\\ \cline{3-7}
@@
     % MockingJay
-    \multirow{3}{*}{MockingJay \cite{liu_mockingjay_2020}} & \multirow{3}{*}{\makecell[l]{LS 360 hrs}} & PC & LS 360 hrs & LS test-clean & - & 0.36\textsuperscript{\ref{footnote:low-resource}} /1.8\textsuperscript{\ref{footnote:low-resource}} /3.6\textsuperscript{\ref{footnote:low-resource}} /18\textsuperscript{\ref{footnote:low-resource}} /45\textsuperscript{\ref{footnote:low-resource}} /360  hrs \\ \cline{3-7}
+    \multirow{3}{*}{MockingJay \parencite{liu_mockingjay_2020}} & \multirow{3}{*}{\makecell[l]{LS 360 hrs}} & PC & LS 360 hrs & LS test-clean & - & 0.36\textsuperscript{\ref{footnote:low-resource}} /1.8\textsuperscript{\ref{footnote:low-resource}} /3.6\textsuperscript{\ref{footnote:low-resource}} /18\textsuperscript{\ref{footnote:low-resource}} /45\textsuperscript{\ref{footnote:low-resource}} /360  hrs \\ \cline{3-7}
     & & SID & LS 100 hrs\textsuperscript{\ref{footnote:91-split-test}} & LS 100 hrs\textsuperscript{\ref{footnote:91-split-test}} & - & 90\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
@@
     % CPC modified
-    \multirow{2}{*}{CPC modified \cite{riviere_unsupervised_2020}} & LS 100 hrs & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & \makecell[l]{LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}}} & - & \makecell[l]{80\textsuperscript{\ref{footnote:low-resource}} hrs} \\ \cline{2-7}
+    \multirow{2}{*}{CPC modified \parencite{riviere_unsupervised_2020}} & LS 100 hrs & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & \makecell[l]{LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}}} & - & \makecell[l]{80\textsuperscript{\ref{footnote:low-resource}} hrs} \\ \cline{2-7}
     & \makecell[l]{LS 100 hrs,\\LS 960 hrs,\\LL 60k hrs} & PC & CV-dataset\textsuperscript{\ref{footnote:official-split}} & CV-dataset\textsuperscript{\ref{footnote:official-split}} & \checkmark & 1 hrs \\ \hline
     % vq-wav2vec
-    \multirow{2}{*}[0mm]{vq-wav2vec \cite{baevski_vqwav2vec_2020}} & \multirow{2}{*}[0mm]{LS 960 hrs} & ASR & WSJ si284 & \makecell[l]{WSJ eval92} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{vq-wav2vec \parencite{baevski_vqwav2vec_2020}} & \multirow{2}{*}[0mm]{LS 960 hrs} & ASR & WSJ si284 & \makecell[l]{WSJ eval92} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
     & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \hline
     % DeCoAR
-    \multirow{2}{*}[0mm]{DeCoAR \cite{ling_deep_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{LS 100/360/\\\:\:\:460/960 hrs,\\ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{25\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}}/81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{DeCoAR \parencite{ling_deep_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{LS 100/360/\\\:\:\:460/960 hrs,\\ WSJ si284}} & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{25\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}}/81 hrs} \\ \cline{3-7}
     & & ASR & \makecell[l]{LS 100/360/\\\:\:\:460/960 hrs} & \makecell[l]{LS test-clean,\\ LS test-other} & - & 100/360/460/960 hrs \\ \hline
     % MT-APC
-    \multirow{2}{*}[0mm]{MT-APC\cite{chung_improved_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & ASR & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & WSJ dev93 & \checkmark & 72 hrs \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{MT-APC\parencite{chung_improved_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & ASR & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & WSJ dev93 & \checkmark & 72 hrs \\ \cline{3-7}
     & & ST & LS En-Fr\textsuperscript{\ref{footnote:official-split}} & LS En-Fr\textsuperscript{\ref{footnote:official-split}} & - & 236 hrs \\ \hline
     % PASE+
-    \multirow{3}{*}[0mm]{PASE+\cite{ravanelli_multitask_2020}} & \multirow{3}{*}[0mm]{LS 50 hrs \cite{ravanelli_learning_2018}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
+    \multirow{3}{*}[0mm]{PASE+\parencite{ravanelli_multitask_2020}} & \multirow{3}{*}[0mm]{LS 50 hrs \parencite{ravanelli_learning_2018}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \cline{3-7}
     & & ASR & DIRHA\textsuperscript{\ref{footnote:official-split}} & DIRHA\textsuperscript{\ref{footnote:official-split}} & \checkmark & 11 hrs \\ \cline{3-7}
@@
     % AALBERT
-    \multirow{2}{*}[0mm]{AALBERT \cite{chi_audio_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{LS 360 hrs}} & PC & LS 100 hrs\textsuperscript{\ref{footnote:APC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:APC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{AALBERT \parencite{chi_audio_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{LS 360 hrs}} & PC & LS 100 hrs\textsuperscript{\ref{footnote:APC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:APC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
     & & SID & LS 360 hrs\textsuperscript{\ref{footnote:APC-split}} & LS 360 hrs\textsuperscript{\ref{footnote:APC-split}} & - & 288\textsuperscript{\ref{footnote:low-resource}} hrs \\
@@
     % BMR
-    \multirow{2}{*}[0mm]{BMR \cite{wang_unsupervised_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{WSJ si284,\\LS 960 hrs}} & ASR & WSJ si284 & WSJ eval92 & - & 81 hrs \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{BMR \parencite{wang_unsupervised_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{WSJ si284,\\LS 960 hrs}} & ASR & WSJ si284 & WSJ eval92 & - & 81 hrs \\ \cline{3-7}
     & & PR & WSJ si84/si284 & WSJ dev93 & - & 15/81 hrs \\ \hline
     % vq-APC
-    \multirow{2}{*}[0mm]{vq-APC \cite{chung_vectorquantized_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{vq-APC \parencite{chung_vectorquantized_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
     & & SID & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & \checkmark & 65\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline
     % vq-wav2vec + Discrete BERT
-    \makecell[l]{vq-wav2vec + \\DiscreteBERT \cite{baevski_effectiveness_2020}} & LS 960 hrs & ASR & LS 100 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline
+    \makecell[l]{vq-wav2vec + \\DiscreteBERT \parencite{baevski_effectiveness_2020}} & LS 960 hrs & ASR & LS 100 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline
     % speech-XLNet
-    \multirow{2}{*}[-1mm]{speech-XLNet \cite{song_speechxlnet_2020}} & \multirow{2}{*}[-1mm]{\makecell[l]{LS 960 hrs\\\;\;\;+ WSJ si284\\\;\;\;+ TED2}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs\vspace{1mm} \\ \cline{3-7}
+    \multirow{2}{*}[-1mm]{speech-XLNet \parencite{song_speechxlnet_2020}} & \multirow{2}{*}[-1mm]{\makecell[l]{LS 960 hrs\\\;\;\;+ WSJ si284\\\;\;\;+ TED2}} & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs\vspace{1mm} \\ \cline{3-7}
     & & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{7\textsuperscript{\ref{footnote:low-resource}}/14\textsuperscript{\ref{footnote:low-resource}}/30\textsuperscript{\ref{footnote:low-resource}}/81 hrs}\vspace{1mm} \\ \hline
     % MPC
-    \multirow{3}{*}[0mm]{MPC \cite{jiang_improving_2019, jiang_further_2021}} & \multirow{2}{*}[0mm]{\makecell[l]{Didi Callcenter,\\Didi Dictation,\\Open Mandarin}} & ASR-zh & HKUST\textsuperscript{\ref{footnote:official-split}} & HKUST\textsuperscript{\ref{footnote:official-split}} & \checkmark & 168 hrs \\ \cline{3-7}
+    \multirow{3}{*}[0mm]{MPC \parencite{jiang_improving_2019, jiang_further_2021}} & \multirow{2}{*}[0mm]{\makecell[l]{Didi Callcenter,\\Didi Dictation,\\Open Mandarin}} & ASR-zh & HKUST\textsuperscript{\ref{footnote:official-split}} & HKUST\textsuperscript{\ref{footnote:official-split}} & \checkmark & 168 hrs \\ \cline{3-7}
     & & ASR-zh & AISHELL-1\textsuperscript{\ref{footnote:official-split}} & AISHELL-1\textsuperscript{\ref{footnote:official-split}} & \checkmark & 178 hrs\vspace{1mm} \\ \cline{2-7}
@@
     % MPE
-    \multirow{2}{*}[0mm]{MPE \cite{liu_masked_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{WSJ si284,\\LS 960 hrs}} & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{25\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}}/81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{MPE \parencite{liu_masked_2020}} & \multirow{2}{*}[0mm]{\makecell[l]{WSJ si284,\\LS 960 hrs}} & ASR & WSJ si284 & WSJ eval92 & - & \makecell[l]{25\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}}/81 hrs} \\ \cline{3-7}
     & & ASR & LS 100/360/960 hrs & LS test-clean & - & 100/360/960 hrs \\ \hline
     % ConvDMM
-    ConvDMM \cite{khurana_convolutional_2020} & \makecell[l]{LS 50\cite{ravanelli_learning_2018}/360/\\\:\:\:960 hrs} & PC/PR & WSJ si284 & WSJ eval92 & \checkmark & \makecell[l]{5\textsuperscript{\ref{footnote:low-resource}}/50\textsuperscript{\ref{footnote:low-resource}}/100\textsuperscript{\ref{footnote:low-resource}} mins,\\ 4\textsuperscript{\ref{footnote:low-resource}}/8\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}} hrs} \\ \hline
+    ConvDMM \parencite{khurana_convolutional_2020} & \makecell[l]{LS 50\parencite{ravanelli_learning_2018}/360/\\\:\:\:960 hrs} & PC/PR & WSJ si284 & WSJ eval92 & \checkmark & \makecell[l]{5\textsuperscript{\ref{footnote:low-resource}}/50\textsuperscript{\ref{footnote:low-resource}}/100\textsuperscript{\ref{footnote:low-resource}} mins,\\ 4\textsuperscript{\ref{footnote:low-resource}}/8\textsuperscript{\ref{footnote:low-resource}}/40\textsuperscript{\ref{footnote:low-resource}} hrs} \\ \hline
     % wav2vec 2.0
-    \multirow{2}{*}{wav2vec 2.0 \cite{baevski_wav2vec_2020}} & \multirow{2}{*}{\makecell[l]{LS 960 hrs,\\ LL 60k hrs}} & ASR & LS 960 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \cline{3-7}
+    \multirow{2}{*}{wav2vec 2.0 \parencite{baevski_wav2vec_2020}} & \multirow{2}{*}{\makecell[l]{LS 960 hrs,\\ LL 60k hrs}} & ASR & LS 960 hrs & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}},\\ 1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \cline{3-7}
     & & PR & TIMIT\textsuperscript{\ref{footnote:official-split}} & TIMIT\textsuperscript{\ref{footnote:official-split}} & \checkmark & 4 hrs \\ \hline
     % NPC
-    \multirow{2}{*}[0mm]{NPC \cite{liu_nonautoregressive_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
+    \multirow{2}{*}[0mm]{NPC \parencite{liu_nonautoregressive_2020}} & \multirow{2}{*}[0mm]{LS 360 hrs} & PC & WSJ si284\textsuperscript{\ref{footnote:91-split-dev}} & \makecell[l]{WSJ dev93} & \checkmark & \makecell[l]{81 hrs} \\ \cline{3-7}
     & & SID & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & WSJ si284\textsuperscript{\ref{footnote:APC-split}} & \checkmark & 65\textsuperscript{\ref{footnote:low-resource}} hrs \\ \hline
     % DeCoAR 2.0
-    DeCoAR 2.0 \cite{ling_decoar_2020} & LS 960 hrs & ASR & \makecell[l]{LS 100 hrs} & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline
+    DeCoAR 2.0 \parencite{ling_decoar_2020} & LS 960 hrs & ASR & \makecell[l]{LS 100 hrs} & \makecell[l]{LS test-clean,\\ LS test-other} & - & \makecell[l]{1\textsuperscript{\ref{footnote:low-resource}}/10\textsuperscript{\ref{footnote:low-resource}}/100 hrs} \\ \hline
     % TERA
-    \multirow{4}{*}[0mm]{TERA \cite{liu_tera_2021}} & \multirow{4}{*}[0mm]{\makecell[l]{LS 100/360/\\\:\:\:960 hrs}} & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
+    \multirow{4}{*}[0mm]{TERA \parencite{liu_tera_2021}} & \multirow{4}{*}[0mm]{\makecell[l]{LS 100/360/\\\:\:\:960 hrs}} & PC & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
     & & SID & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & LS 100 hrs\textsuperscript{\ref{footnote:CPC-split}} & - & 80\textsuperscript{\ref{footnote:low-resource}} hrs \\ \cline{3-7}
@@
     % HuBERT
-    HuBERT \cite{hsu_hubert_2021} & \makecell[l]{LS 960 hrs,\\ LL 60k hrs} & ASR & LS 960 hrs & \makecell[l]{LS test-clean\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}}, 1\textsuperscript{\ref{footnote:low-resource}}\\/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \hline
+    HuBERT \parencite{hsu_hubert_2021} & \makecell[l]{LS 960 hrs,\\ LL 60k hrs} & ASR & LS 960 hrs & \makecell[l]{LS test-clean\\ LS test-other} & - & \makecell[l]{10 mins\textsuperscript{\ref{footnote:low-resource}}, 1\textsuperscript{\ref{footnote:low-resource}}\\/10\textsuperscript{\ref{footnote:low-resource}}/100/960 hrs} \\ \hline
     % wav2vec-c
-    wav2vec-c \cite{sadhu_wav2vecc_2021} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
-    UniSpeech-SAT \cite{chen_unispeechsat_2021} & \makecell[l]{LL 60k hrs\\\;\;+ GigaSpeech-10k\\\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See SUPERB \cite{yang_superb_2021} paper for details \\ \hline
-    WavLM \cite{chen_wavlm_2021} & \makecell[l]{LL 60k hrs\\\;\;+ GigaSpeech-10k\\\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See \cite{yang_superb_2021} for details \\ \hline
-    \multirow{4}{*}[0mm]{XLS-R \cite{babu_xlsr_2021}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;+ CV-dataset-7k\\\;\;+ VL\\\;\;+ BBL}} & ASR & VP, MLS, CV-dataset, BBL, LS & VP, MLS, CV-dataset, BBL, LS & - & \multirow{4}{*}[0mm]{See \cite{babu_xlsr_2021} for details} \\ \cline{3-6}
+    wav2vec-c \parencite{sadhu_wav2vecc_2021} & Alexa-10k & ASR & \makecell[l]{Alexa-eval} & \makecell[l]{Alexa-eval} & \checkmark & 1k hrs \\ \hline
+    UniSpeech-SAT \parencite{chen_unispeechsat_2021} & \makecell[l]{LL 60k hrs\\\;\;+ GigaSpeech-10k\\\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See SUPERB \parencite{yang_superb_2021} paper for details \\ \hline
+    WavLM \parencite{chen_wavlm_2021} & \makecell[l]{LL 60k hrs\\\;\;+ GigaSpeech-10k\\\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See \parencite{yang_superb_2021} for details \\ \hline
+    \multirow{4}{*}[0mm]{XLS-R \parencite{babu_xlsr_2021}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;+ CV-dataset-7k\\\;\;+ VL\\\;\;+ BBL}} & ASR & VP, MLS, CV-dataset, BBL, LS & VP, MLS, CV-dataset, BBL, LS & - & \multirow{4}{*}[0mm]{See \parencite{babu_xlsr_2021} for details} \\ \cline{3-6}
     & & SID & VoxCeleb1 & VoxCeleb1 & \checkmark & \\ \cline{3-6}
@@
 Train/test split made available
-by \cite{oord_representation_2018} on Google drive
+by \parencite{oord_representation_2018} on Google drive
 https://drive.google.com/drive/folders/1BhJ2umKH3whguxMwifaKtSra0TgAbtfb.}
@@
 The 6 AED datasets used in
-\cite{tagliasacchi_pretraining_2020} are MUSAN~\cite{snyder_musan_2015}, Bird Audio
-Detection~\cite{stowell_automatic_2019}, Speech Commands~\cite{warden_speech_2018},
-Spoken Language Identification~\cite{oponowicz_spoken_2018}, TUT Urban Acoustic
-Scenes 2018~\cite{mesaros_detection_2018} plus an SID task built with LS
+\parencite{tagliasacchi_pretraining_2020} are MUSAN~\parencite{snyder_musan_2015}, Bird Audio
+Detection~\parencite{stowell_automatic_2019}, Speech Commands~\parencite{warden_speech_2018},
+Spoken Language Identification~\parencite{oponowicz_spoken_2018}, TUT Urban Acoustic
+Scenes 2018~\parencite{mesaros_detection_2018} plus an SID task built with LS
 \textit{train-clean-100}. In addition to the 6 datasets,
-\cite{quitry_learning_2019} use the NSynth dataset~\cite{engel_neural_2017} for
+\parencite{quitry_learning_2019} use the NSynth dataset~\parencite{engel_neural_2017} for
 evaluation.}
@@
 CV-dataset, LS, WSJ, TIMIT, Speech Accent Archive 
-(SSA)~\cite{weinberger_speech_2011}, TED3, and SWB. SSA is a growing annotated corpus of
+(SSA)~\parencite{weinberger_speech_2011}, TED3, and SWB. SSA is a growing annotated corpus of
 English speech with various accents. Among the papers studied in this review, SSA
-is used in \cite{kawakami_learning_2020} only for pre-training, and only 1 hour
+is used in \parencite{kawakami_learning_2020} only for pre-training, and only 1 hour
 of audio is utilized. Thus, we exclude it from our discussion in
@@
 growing industry investment in pre-training with larger datasets, e.g.,
-CPC-8k (8k hours) for Bidir-CPC~\cite{kawakami_learning_2020}, LL (60k hours) for
-CPC modified~\cite{riviere_unsupervised_2020}, 
-wav2vec 2.0~\cite{baevski_wav2vec_2020}, and HuBERT~\cite{hsu_hubert_2021}, Alexa internal
-datasets (10k hours) for wav2vec-c~\cite{sadhu_wav2vecc_2021}, Didi internal
-datasets (10k hours) for MPC~\cite{jiang_improving_2019, jiang_further_2021}, the
+CPC-8k (8k hours) for Bidir-CPC~\parencite{kawakami_learning_2020}, LL (60k hours) for
+CPC modified~\parencite{riviere_unsupervised_2020}, 
+wav2vec 2.0~\parencite{baevski_wav2vec_2020}, and HuBERT~\parencite{hsu_hubert_2021}, Alexa internal
+datasets (10k hours) for wav2vec-c~\parencite{sadhu_wav2vecc_2021}, Didi internal
+datasets (10k hours) for MPC~\parencite{jiang_improving_2019, jiang_further_2021}, the
 combination of Gigaspeech, VP-24k, and LL (94k hours in total) for
-UniSpeech-SAT~\cite{chen_unispeechsat_2021} and WavLM~\cite{chen_wavlm_2021}, and
+UniSpeech-SAT~\parencite{chen_unispeechsat_2021} and WavLM~\parencite{chen_wavlm_2021}, and
 the combination of VP-400K, MLS, CV-dataset, VL and BBL (436k hours in total)
-for XLS-R~\cite{babu_xlsr_2021}. We expect this trend to continue with the growth
+for XLS-R~\parencite{babu_xlsr_2021}. We expect this trend to continue with the growth
 in available computing power. Most studies focuses on learning representations
-for English, whereas Chinese~\cite{jiang_improving_2019, jiang_further_2021} and
-multilinguality~\cite{kawakami_learning_2020, babu_xlsr_2021} are also gaining
+for English, whereas Chinese~\parencite{jiang_improving_2019, jiang_further_2021} and
+multilinguality~\parencite{kawakami_learning_2020, babu_xlsr_2021} are also gaining
 attention. Compared to pre-training, datasets used for fine-tuning are more
@@
 fine-tuning also varies from several minutes up to 1k hours. Recent
-benchmarks such as SUPERB~\cite{yang_superb_2021} that consolidate multiple 
+benchmarks such as SUPERB~\parencite{yang_superb_2021} that consolidate multiple 
 downstream tasks have gained attention for evaluating SSL
-methodologies~\cite{chen_unispeechsat_2021, chen_wavlm_2021}. The goal of such benchmarks is
+methodologies~\parencite{chen_unispeechsat_2021, chen_wavlm_2021}. The goal of such benchmarks is
 to provide a holistic evaluation of the performance of learned representations; 
@@
 \centering
-\includegraphics[width=0.9\textwidth]{paper_review/ls-test-clean-v2.pdf}
+\includegraphics[width=\textwidth]{paper_review/ls-test-clean-v2.pdf}
 \caption[SSL performance for ASR evaluated on
@@
 publication. We also compare with baselines including the log mel filterbank (fBANK) and semi-supervised, self-training
-approaches (iterative pseudo labeling (IPL)~\cite{xu_iterative_2020}, 
-slimIPL~\cite{likhomanenko_slimipl_2021}, noisy student~\cite{park_improved_2020}).
+approaches (iterative pseudo labeling (IPL)~\parencite{xu_iterative_2020}, 
+slimIPL~\parencite{likhomanenko_slimipl_2021}, noisy student~\parencite{park_improved_2020}).
 These approaches are visualized in black.
-Also, note that the current state of the art---conformer XXL + noisy student~\cite{zhang_pushing_2020}---is a combination of self-training and SSL techniques.
+Also, note that the current state of the art---conformer XXL + noisy student~\parencite{zhang_pushing_2020}---is a combination of self-training and SSL techniques.
 Given the
@@
 show performance from such methods (iterative pseudo labeling 
-(IPL)~\cite{xu_iterative_2020}, slimIPL~\cite{likhomanenko_slimipl_2021}, noisy 
-student~\cite{park_improved_2020}), as well as the current state of the art---conformer XXL + noisy
-student~\cite{zhang_pushing_2020}---which augments SSL with various advanced
+(IPL)~\parencite{xu_iterative_2020}, slimIPL~\parencite{likhomanenko_slimipl_2021}, noisy 
+student~\parencite{park_improved_2020}), as well as the current state of the art---conformer XXL + noisy
+student~\parencite{zhang_pushing_2020}---which augments SSL with various advanced
 techniques including self-training. Furthermore, we illustrate in the figure
-the performance of a baseline system \cite{yang_superb_2021} based on log mel filterbank (fBANK), which is one of the most commonly used features designed by domain experts.
+the performance of a baseline system \parencite{yang_superb_2021} based on log mel filterbank (fBANK), which is one of the most commonly used features designed by domain experts.
 As observed in the figure, most SSL techniques outperform fBANK
@@
 combining SSL pre-training and self-training 
-techniques~\cite{zhang_pushing_2020}, SSL models are competitive in other tasks, including IC,
+techniques~\parencite{zhang_pushing_2020}, SSL models are competitive in other tasks, including IC,
 SID, ASV, and QbE. We summarize the performance of these models and previous
@@
 
-\begin{table}[t!]
+\begin{table}[t]
   \centering
@@
     \midrule
-    ASR (WER $\downarrow$) & LS test-clean/other & 2.1/4.0 \cite{xu_iterative_2020} & 1.4/2.6 \cite{zhang_pushing_2020} \\ \hline
-    IC (Acc $\uparrow$) & FSC & 98.8 \cite{lugosch_speech_2019} & 99.3\cite{chen_unispeechsat_2021} \\ \hline
-    SID (Acc $\uparrow$) & VoxCeleb1 & 94.8 \cite{hajibabaei_unified_2018} & 95.5 \cite{chen_wavlm_2021} \\ \hline
-    ASV (EER $\downarrow$) & VoxCeleb1 & 3.1 \cite{hajavi_siamese_2021} & 2.4 \cite{wang_finetuned_2021} \\ \hline
-    QbE (MTWV $\uparrow$) & QUESST (EN) & 10.6 \cite{rodriguez-fuentes_gttsehu_2014} & 11.2\cite{chen_unispeechsat_2021} \\
+    ASR (WER $\downarrow$) & LS test-clean/other & 2.1/4.0 \parencite{xu_iterative_2020} & 1.4/2.6 \parencite{zhang_pushing_2020} \\ \hline
+    IC (Acc $\uparrow$) & FSC & 98.8 \parencite{lugosch_speech_2019} & 99.3\parencite{chen_unispeechsat_2021} \\ \hline
+    SID (Acc $\uparrow$) & VoxCeleb1 & 94.8 \parencite{hajibabaei_unified_2018} & 95.5 \parencite{chen_wavlm_2021} \\ \hline
+    ASV (EER $\downarrow$) & VoxCeleb1 & 3.1 \parencite{hajavi_siamese_2021} & 2.4 \parencite{wang_finetuned_2021} \\ \hline
+    QbE (MTWV $\uparrow$) & QUESST (EN) & 10.6 \parencite{rodriguez-fuentes_gttsehu_2014} & 11.2\parencite{chen_unispeechsat_2021} \\
     \bottomrule
@@
 benchmarks with shared downstream tasks, datasets, and downstream recipes. Such
-efforts include SUPERB~\cite{yang_superb_2021}, 
-LeBenchmark~\cite{evain_lebenchmark_2021}, ZeroSpeech~\cite{dunbar_zero_2020},
-HEAR \cite{turian_hear_2022},
-NOSS~\cite{shor_learning_2020}, and HARES~\cite{wang_learning_2021}.
+efforts include SUPERB~\parencite{yang_superb_2021}, 
+LeBenchmark~\parencite{evain_lebenchmark_2021}, ZeroSpeech~\parencite{dunbar_zero_2020},
+HEAR \parencite{turian_hear_2022},
+NOSS~\parencite{shor_learning_2020}, and HARES~\parencite{wang_learning_2021}.
 
-SUPERB~\cite{yang_superb_2021} is a benchmarking platform that allows the
+SUPERB~\parencite{yang_superb_2021} is a benchmarking platform that allows the
 SSL community to train, evaluate, and compare speech representations on
@@
 to democratize SSL beyond race among Big Tech.   % AMH: what does this mean?
-LeBenchmark~\cite{evain_lebenchmark_2021} shares a vision similar to SUPERB and provides a
+LeBenchmark~\parencite{evain_lebenchmark_2021} shares a vision similar to SUPERB and provides a
 reproducible framework for assessing SSL in French with ASR, spoken language
 understanding, speech translation, and emotion recognition. 
-ZeroSpeech~\cite{dunbar_zero_2020} (described in more detail in \cref{zero_speech})
+ZeroSpeech~\parencite{dunbar_zero_2020} (described in more detail in \cref{zero_speech})
 challenges the scientific community to build speech and language understanding
@@
 SSL techniques are also benchmarked with the ZeroSpeech 
-challenge~\cite{tjandra_transformer_2020, vanniekerk_vectorquantized_2020}. Apart from the speech
+challenge~\parencite{tjandra_transformer_2020, vanniekerk_vectorquantized_2020}. Apart from the speech
 community, researchers have also established HEAR (holistic evaluation of audio
-representations) \cite{turian_hear_2022}, NOSS (non-semantic
-speech benchmark)~\cite{shor_learning_2020}, and HARES (holistic audio
-representation evaluation suite)~\cite{wang_learning_2021} to benchmark audio
+representations) \parencite{turian_hear_2022}, NOSS (non-semantic
+speech benchmark)~\parencite{shor_learning_2020}, and HARES (holistic audio
+representation evaluation suite)~\parencite{wang_learning_2021} to benchmark audio
 representations. These efforts promote the creation of an audio embedding

### 2023-09-29 chapters/paper_review/04b_analysis.tex
Add and revise short table and figure captions
Updates to citation commands and footnotes

--- 2023-03-16 chapters/paper_review/04b_analysis.tex
@@
 
-\section{Analysis of Self-Supervised Representations}
+\section{Analysis of self-supervised representations}
 \label{analysis}
@@
 
-In \cite{pasad_layerwise_2021} wav2vec~2.0 representations were analyzed with
+In \parencite{pasad_layerwise_2021} wav2vec~2.0 representations were analyzed with
 respect to their acoustic-linguistic information content at different
@@
 correspond to the same word, and a standard benchmark of 11 word
-similarity tasks~\cite{faruqi14}. These are mostly used to gauge the
+similarity tasks~\parencite{faruqi_community_2014}. These are mostly used to gauge the
 amount of lexical information present in the embeddings.  Using this
@@
 phonetic or word-related information may not be the best
-strategy. Indeed, the authors of \cite{baevski2021unsupervised} show that a
+strategy. Indeed, the authors of \parencite{baevski_unsupervised_2021} show that a
 phone classifier trained on each of the 24 frozen layers of a wav2vec~2.0 model
@@
 for layers 10--21 and higher error rates for the other layers. 
-\cite{pasad_layerwise_2021} further show that fine-tuning the pre-trained model with a
+\parencite{pasad_layerwise_2021} further show that fine-tuning the pre-trained model with a
 character-level CTC training criterion changes the behavior of the
@@
 an analysis of text-based Transformers trained with a masked language model
-criterion~\cite{voita2019} observed a similar compression plus reconstruction
+criterion~\parencite{voita_bottomup_2019} observed a similar compression plus reconstruction
 pattern. Interestingly, similar network behavior was also recently described
@@
 self-supervised learner (SimCLR) that optimizes for augmentation invariance,
-\cite{grigg2021self} show that it is the intermediate representations that most
+\parencite{grigg_selfsupervised_2021} show that it is the intermediate representations that most
 closely approximate information learned in a supervised way, i.e., they
@@
 information. 
-\edit{It is shown that the per-utterance mean of CPC features captures speaker information to a large extent\cite{van2021analyzing}.}
+\edit{It is shown that the per-utterance mean of CPC features captures speaker information to a large extent\parencite{vanniekerk_analyzing_2021}.}
 Location of information pertaining to speakers vs.\ language classes was
-analyzed in \cite{ling20odyssey} for a 12-layer BERTphone model. This model
+analyzed in \parencite{ling_bertphone_2020} for a 12-layer BERTphone model. This model
 combines a self-supervised masked reconstruction loss with a phone-based CTC
@@
 phonetic information whereas speaker recognition uses a combination of
-acoustic and phonetic information. In a recent study~\cite{chen_wavlm_2021} the
+acoustic and phonetic information. In a recent study~\parencite{chen_wavlm_2021} the
 same technique was used to identify layer contributions for the downstream
@@
 necessary for speaker diarization and verification whereas higher layers encode
-phonetic and semantic information. Another study~\cite{wang2021layersup} used
+phonetic and semantic information. Another study~\parencite{wang_selfsupervised_2021} used
 explicit self-supervised loss at the intermediate layers rather than just the
@@
 Most self-supervised learning approaches rely on a Transformer architecture for
-the representation model. In \cite{AnalyzeAttention} the attention patterns in
+the representation model. In \parencite{yang_understanding_2020} the attention patterns in
 generatively trained Transformer representation models were analyzed.
@@
 \subsection{Training criterion}
-In \cite{chung2021}, representations based on different training criteria
+In \parencite{chung_similarity_2021}, representations based on different training criteria
 (masked predictive coding, contrastive predictive coding, and autoregressive
@@
 
-A similar insight was obtained in \cite{zhou2020}, which compared vq-vae and
+A similar insight was obtained in \parencite{zhou_comparison_2020}, which compared vq-vae and
 vq-wav2vec with respect to their ability to discover phonetic units.
@@
 Several studies have demonstrated better downstream performance when using
-larger datasets~\cite{rivi20, kawakami2020learning, chen_wavlm_2021}. For
+larger datasets~\parencite{riviere_unsupervised_2020, kawakami_learning_2020, chen_wavlm_2021}. For
 example,
-\cite{kawakami2020learning} compared representations learned by a bidirectional
+\parencite{kawakami_learning_2020} compared representations learned by a bidirectional
 CPC model from the standard 960 hour LS corpus and a corpus of 8,000 hours of
@@
 to play a role, although this was not quantified in this study. However, recent
-experiments with larger and more diverse data collections~\cite{chen_wavlm_2021}
+experiments with larger and more diverse data collections~\parencite{chen_wavlm_2021}
 confirm this assumption, as do 
@@
 
-\edit{The relation between model sizes and downstream performances have also been investigated \cite{pue2021scaling,versteegh2015zero}.}
+\edit{The relation between model sizes and downstream performances have also been investigated \parencite{pu_scaling_2021,versteegh_zero_2015}.}
 Using the Mockingjay 
-model~\cite{liu_mockingjay_2020}, the authors in \cite{pue2021scaling} attempt to establish a relationship between
+model~\parencite{liu_mockingjay_2020}, the authors in \parencite{pu_scaling_2021} attempt to establish a relationship between
 model size and self-supervised \ensuremath{L_1} loss and demonstrate that it approximately
@@
 shifts? 
-One study~\cite{kawakami2020learning} compared pre-trained
+One study~\parencite{kawakami_learning_2020} compared pre-trained
 representations from a CPC model against MFCCs and 
@@
 Similar problems were also investigated using HuBERT and wav2vec~2.0 models in
-\cite{chang2021exploration}.
-In \cite{robustw2v2} domain effects were studied in greater detail using 
+\parencite{chang_exploration_2021}.
+In \parencite{hsu_robust_2021} domain effects were studied in greater detail using 
 datasets from six different domains. In particular, the authors focused on the
@@
 
-In \cite{rivi20} the effectiveness of CPC-trained representations for
+In \parencite{riviere_unsupervised_2020} the effectiveness of CPC-trained representations for
 phone discrimination tasks was compared across several languages. It
@@
 learn contextualized speech features that generalize across different
-languages. In \cite{conneau20xling}, a wav2vec~2.0 model was trained on data
+languages. In \parencite{conneau_unsupervised_2020}, a wav2vec~2.0 model was trained on data
 from multiple different languages and different corpora (Babel, Common Voice,
@@
 from different layers of a self-supervised models also generalizes to the
-multilingual setting. Experiments in \cite{baevski2021unsupervised} on phone
+multilingual setting. Experiments in \parencite{baevski_unsupervised_2021} on phone
 recognition in eight languages based on the different layers of the

### 2023-09-29 chapters/paper_review/05_zero.tex
Add and revise short table and figure captions
Updates
Move table captions to top
Updates to citation commands and footnotes

@@
 
-\section{From Representation Learning to Zero Resources}
+\section{From representation learning to zero resources}
 \label{sec:zero}
@@
 
-\begin{table*}[ht]
+\begin{sidewaystable}[t]
 \centering
-\caption{
+\caption[Performance of models within unsupervised ASR.]{
 	Unsupervised ASR.
@@
 	\label{fig:unsupervisedASR}
-	\resizebox*{0.99\textwidth}{!}{%
+	\resizebox*{\textwidth}{!}{%
 	{\renewcommand*\arraystretch{1.4}}
@@
 	\midrule
-	\cite{liu_completely_2018}& Audio word2vec~\cite{wang_segmental_2018} & Oracle & Phoneme & \edit{Adversarial Training~\cite{gulrajani_improved_2017}} & - & TIMIT (PER): 63.6\%  \\ %Submitted on 1 Apr 2018
+	\parencite{liu_completely_2018}& Audio word2vec~\parencite{wang_segmental_2018} & Oracle & Phoneme & \edit{Adversarial Training~\parencite{gulrajani_improved_2017}} & - & TIMIT (PER): 63.6\%  \\ %Submitted on 1 Apr 2018
 	\midrule
-	\cite{chung_unsupervised_2018} & Speech2vec~\cite{chung_speech2vec_2018} & BES-GMM~\cite{kamper_segmental_2017} &  Word2Vec & \edit{Adversarial Training~\cite{conneau_word_2018}} & Self-training & SWC (Acc): 10.9\%  \\ %Submitted on 18 May 2018
+	\parencite{chung_unsupervised_2018} & Speech2vec~\parencite{chung_speech2vec_2018} & BES-GMM~\parencite{kamper_segmental_2017} &  Word2Vec & \edit{Adversarial Training~\parencite{conneau_word_2018}} & Self-training & SWC (Acc): 10.9\%  \\ %Submitted on 18 May 2018
 	%https://arxiv.org/pdf/1805.07467.pdf %The text embeddings were obtained by training Word2Vec on the transcriptions using the fastText implementation without subword information [3].
@@
 
-	\cite{chung_unsupervised_2019a} & \tabincell{l}{Speech2vec\\(English)} & Oracle & \tabincell{c}{Word2Vec\\(French)} & VecMap~\cite{artetxe_robust_2018} & \tabincell{l}{LM rescore,\\sequence DAE} & ST (BLEU): 10.8\%   \\ %[Submitted on 4 Nov 2018]
+	\parencite{chung_unsupervised_2019a} & \tabincell{l}{Speech2vec\\(English)} & Oracle & \tabincell{c}{Word2Vec\\(French)} & VecMap~\parencite{artetxe_robust_2018} & \tabincell{l}{LM rescore,\\sequence DAE} & ST (BLEU): 10.8\%   \\ %[Submitted on 4 Nov 2018]
 	\midrule
 
-	\cite{yeh_unsupervised_2019} & MFCC & GAS~\cite{wang_gate_2017} & Phoneme & Empirical-ODM~\cite{liu_unsupervised_2017} & Self-training &  TIMIT (PER):  41.6\%  \\ %Submitted on 23 Dec 2018
+	\parencite{yeh_unsupervised_2019} & MFCC & GAS~\parencite{wang_gate_2017} & Phoneme & Empirical-ODM~\parencite{liu_unsupervised_2017} & Self-training &  TIMIT (PER):  41.6\%  \\ %Submitted on 23 Dec 2018
 	\midrule
 
-	\cite{chen_completely_2019} & MFCC & GAS  & Phoneme & \edit{Adversarial Training~\cite{gulrajani_improved_2017}} & Self-training &  TIMIT (PER):  33.1\%   \\
+	\parencite{chen_completely_2019} & MFCC & GAS  & Phoneme & \edit{Adversarial Training~\parencite{gulrajani_improved_2017}} & Self-training &  TIMIT (PER):  33.1\%   \\
 	\midrule
 
-	\cite{baevski_unsupervised_2021} & Wav2vec 2.0~\cite{baevski_wav2vec_2020} & $k$-means & Phoneme &\edit{Adversarial Training~\cite{gulrajani_improved_2017}} & Self-training & \tabincell{l}{TIMIT (PER):  18.6\%, \\ LibriSpeech (WER): 5.9\%}   \\
+	\parencite{baevski_unsupervised_2021} & Wav2vec 2.0~\parencite{baevski_wav2vec_2020} & $k$-means & Phoneme &\edit{Adversarial Training~\parencite{gulrajani_improved_2017}} & Self-training & \tabincell{l}{TIMIT (PER):  18.6\%, \\ LibriSpeech (WER): 5.9\%}   \\
 	\midrule
 
-	\textcolor{black}{\cite{klejch_deciphering_2022}} & \textcolor{black}{\tabincell{l}{Universal Phone \\ Recogniser}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Decipherment~\cite{ravi_deciphering_2011}} & \textcolor{black}{Self-training} & \textcolor{black}{\tabincell{l}{GlobalPhone: 32.5\% to just 1.9\% \\ worse than supervised models} }  \\ %ubmitted on 12 Nov 2021 %m 32.5% to just 1.9% absolute worse than the equivalent fully supervised models trained on the same data
+	\textcolor{black}{\parencite{klejch_deciphering_2022}} & \textcolor{black}{\tabincell{l}{Universal Phone \\ Recogniser}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Decipherment~\parencite{ravi_deciphering_2011}} & \textcolor{black}{Self-training} & \textcolor{black}{\tabincell{l}{GlobalPhone: 32.5\% to just 1.9\% \\ worse than supervised models} }  \\ %ubmitted on 12 Nov 2021 %m 32.5% to just 1.9% absolute worse than the equivalent fully supervised models trained on the same data
 	\midrule
-	\textcolor{black}{\cite{liu_endtoend_2023}} & \textcolor{black}{Wav2vec 2.0~\cite{baevski_wav2vec_2020} }& - & \textcolor{black}{Phoneme} & \textcolor{black}{Adversarial Training~\cite{gulrajani_improved_2017}} & \textcolor{black}{Self-training} & \textcolor{black}{LibriSpeech (WER): 6.3\%}   \\
+	\textcolor{black}{\parencite{liu_endtoend_2023}} & \textcolor{black}{Wav2vec 2.0~\parencite{baevski_wav2vec_2020} }& - & \textcolor{black}{Phoneme} & \textcolor{black}{Adversarial Training~\parencite{gulrajani_improved_2017}} & \textcolor{black}{Self-training} & \textcolor{black}{LibriSpeech (WER): 6.3\%}   \\
 	\midrule
 
-	\textcolor{black}{\cite{liu_endtoend_2023}} & \textcolor{black}{Wav2vec 2.0~\cite{baevski_wav2vec_2020}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Adversarial Training~\cite{gulrajani_improved_2017}} & \textcolor{black}{Self-training} & \textcolor{black}{LJSpeech (WER): 64.0\%}  \\
+	\textcolor{black}{\parencite{liu_endtoend_2023}} & \textcolor{black}{Wav2vec 2.0~\parencite{baevski_wav2vec_2020}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Adversarial Training~\parencite{gulrajani_improved_2017}} & \textcolor{black}{Self-training} & \textcolor{black}{LJSpeech (WER): 64.0\%}  \\
 
@@
 	\end{tabular}}
-\end{table*}
+\end{sidewaystable}
 
-\subsubsection{Unsupervised ASR}
+\paragraph{Unsupervised ASR}
 % {\color{black} Hung-yi}\\
@@
 
-Adversarial training~\cite{goodfellow_generative_2014,arjovsky_wasserstein_2017, gulrajani_improved_2017} is one common way to learn such a 
+Adversarial training~\parencite{goodfellow_generative_2014,arjovsky_wasserstein_2017, gulrajani_improved_2017} is one common way to learn such a 
 mapping function. 
@@
 \edit{
-There is a large amount of work using gradient penalty in the objective of training discriminators~\cite{liu_completely_2018,chen_completely_2019,baevski_unsupervised_2021,liu_endtoend_2023}, which is inspired by Improved Wasserstein Generative Adversarial Network (WGAN)~\cite{gulrajani_improved_2017}.
+There is a large amount of work using gradient penalty in the objective of training discriminators~\parencite{liu_completely_2018,chen_completely_2019,baevski_unsupervised_2021,liu_endtoend_2023}, which is inspired by Improved Wasserstein Generative Adversarial Network (WGAN)~\parencite{gulrajani_improved_2017}.
 }
 \textcolor{black}{
-Other ways to map speech and text include via segmental empirical output distribution matching (segmental empirical-ODM)~\cite{yeh_unsupervised_2019} and decipherment algorithm~\cite{klejch_deciphering_2022}.}
+Other ways to map speech and text include via segmental empirical output distribution matching (segmental empirical-ODM)~\parencite{yeh_unsupervised_2019} and decipherment algorithm~\parencite{klejch_deciphering_2022}.}
 
@@
 Success in unsupervised neural machine translation
-(MT)~\cite{artetxe_unsupervised_2018, conneau_word_2018, lample_unsupervised_2018}
+(MT)~\parencite{artetxe_unsupervised_2018, conneau_word_2018, lample_unsupervised_2018}
 has inspired innovative exploration of various unsupervised ASR algorithms.
@@
 phoneme-level
-segments~\cite{liu_completely_2018,chen_completely_2019,yeh_unsupervised_2019}. 
+segments~\parencite{liu_completely_2018,chen_completely_2019,yeh_unsupervised_2019}. 
 A representation vector or a token then represents each phoneme-level segment. 
@@
 or token to the correct phoneme when the representation or token is highly correlated to the phonemes.
-Wav2vec-U~\cite{baevski_unsupervised_2021} selects the input feature from different layers of wave2vec~2.0~\cite{baevski_wav2vec_2020}.
+Wav2vec-U~\parencite{baevski_unsupervised_2021} selects the input feature from different layers of wave2vec~2.0~\parencite{baevski_wav2vec_2020}.
 The selection criterion is based on analysis of the phonetic information in each layer.
 \textcolor{black}{
-If a universal phone recognizer trained from a diverse set of languages is available, it is another way to transcribe speech into phone-level tokens~\cite{klejch_deciphering_2022}.}
+If a universal phone recognizer trained from a diverse set of languages is available, it is another way to transcribe speech into phone-level tokens~\parencite{klejch_deciphering_2022}.}
 Another series of work is to transform a word into a word embedding. 
-\cite{chung_unsupervised_2018,chung_unsupervised_2019a} use adversarial training to map the word-level
-speech embedding space~\cite{chung_speech2vec_2018} to the word embedding space
+\parencite{chung_unsupervised_2018,chung_unsupervised_2019a} use adversarial training to map the word-level
+speech embedding space~\parencite{chung_speech2vec_2018} to the word embedding space
 and achieve promising performance on spoken word classification, speech
@@
 \TNS{split into 2 sentences, also have a connection to the previous para}
-\cite{liu_completely_2018} is the first work to successfully achieve unsupervised
+\parencite{liu_completely_2018} is the first work to successfully achieve unsupervised
 phoneme recognition by first clustering phoneme-level
-embeddings~\cite{chung_audio_2016,wang_segmental_2018} into a set of
+embeddings~\parencite{chung_audio_2016,wang_segmental_2018} into a set of
 tokens and then using GAN to learn the mapping relationship between tokens and
 phonemes. 
-However, in \cite{liu_completely_2018}, oracle phoneme segmentation is still
+However, in \parencite{liu_completely_2018}, oracle phoneme segmentation is still
 required to get reasonable performance. 
-\cite{chen_completely_2019,yeh_unsupervised_2019} uses gate activation signals
-(GAS)~\cite{wang_gate_2017} to get the phoneme-level segmentation, which is
+\parencite{chen_completely_2019,yeh_unsupervised_2019} uses gate activation signals
+(GAS)~\parencite{wang_gate_2017} to get the phoneme-level segmentation, which is
 unsupervised. 
-%However, the loss term in~\cite{yeh_unsupervised_2019} includes an empirical average over all segments inside a logarithmic function, which can be biased if we sample this empirical average by a mini-batch average.
-\cite{liu2017unsupervised} therefore use an extremely large batch size during training to reduce the biasing problem.
-%These problems are handled by another work~\cite{chen_completely_2019}.
-%\cite{chen_completely_2019} removes the quantization step.
-%Compared to~\cite{yeh_unsupervised_2019}, the discriminator in~\cite{chen_completely_2019} considers all possible information from the text dataset, not limited to the n-gram local statistics.
-%Besides, \cite{chen_completely_2019} works well with a reasonable batch size, which makes it feasible when the computational resource is limited.
-Wav2vec-U~\cite{baevski_unsupervised_2021} selects the input feature from
-different layers of wave2vec 2.0~\cite{baevski_wav2vec_2020}, a self-supervised
+%However, the loss term in~\parencite{yeh_unsupervised_2019} includes an empirical average over all segments inside a logarithmic function, which can be biased if we sample this empirical average by a mini-batch average.
+\parencite{liu2017unsupervised} therefore use an extremely large batch size during training to reduce the biasing problem.
+%These problems are handled by another work~\parencite{chen_completely_2019}.
+%\parencite{chen_completely_2019} removes the quantization step.
+%Compared to~\parencite{yeh_unsupervised_2019}, the discriminator in~\parencite{chen_completely_2019} considers all possible information from the text dataset, not limited to the n-gram local statistics.
+%Besides, \parencite{chen_completely_2019} works well with a reasonable batch size, which makes it feasible when the computational resource is limited.
+Wav2vec-U~\parencite{baevski_unsupervised_2021} selects the input feature from
+different layers of wave2vec 2.0~\parencite{baevski_wav2vec_2020}, a self-supervised
 model.
@@
 
-%In \cite{chung_unsupervised_2018, chung_unsupervised_2019a}, the word boundaries are also generated automatically with speech only. 
-%However, \cite{chung_unsupervised_2018, chung_unsupervised_2019a} need oracle boundaries, therefore not completely unsupervised as the framework proposed in this paper.
+%In \parencite{chung_unsupervised_2018, chung_unsupervised_2019a}, the word boundaries are also generated automatically with speech only. 
+%However, \parencite{chung_unsupervised_2018, chung_unsupervised_2019a} need oracle boundaries, therefore not completely unsupervised as the framework proposed in this paper.
 
@@
 %Using performance to end this section.
-Wav2vec-U~\cite{baevski_unsupervised_2021} achieved state-of-the-art results at the time, which suggests that representation learning is
+Wav2vec-U~\parencite{baevski_unsupervised_2021} achieved state-of-the-art results at the time, which suggests that representation learning is
 essential for the success of unsupervised ASR.
@@
 \textcolor{black}{
-And wav2vec-U 2.0~\cite{liu_endtoend_2023} further removes the requirement of the segmentation stage, so the unsupervised ASR model can be learned in an end-to-end style.}
+And wav2vec-U 2.0~\parencite{liu_endtoend_2023} further removes the requirement of the segmentation stage, so the unsupervised ASR model can be learned in an end-to-end style.}
 The robustness of wav2vec-U was further analyzed with respect to 
 domain-mismatch scenarios in which the domains of unpaired
-speech and text were different~\cite{lin_analyzing_2022}.
+speech and text were different~\parencite{lin_analyzing_2022}.
 Experimental results showed that domain mismatch leads to inferior performance, but a representation model pre-trained on the targeted speech domain extracts better representations and reduces this drop in performance.
@@
 data.
-Therefore, wav2vec-U~\cite{baevski_unsupervised_2021} develops a new approach to
+Therefore, wav2vec-U~\parencite{baevski_unsupervised_2021} develops a new approach to
 determine the hyperparameters without a labeled development set. 
@@
 
-\subsubsection{ASR-TTS}
+\paragraph{ASR-TTS}
 %{\color{black} Shinji}\\
@@
 
-The ASR-TTS framework can jointly optimize both ASR and TTS \edit{without using paired data}~\cite{tjandra_listening_2017,hori_cycleconsistency_2019,wang_improving_2020}.
-A speech chain~\cite{tjandra_listening_2017,tjandra_machine_2018} is one 
+The ASR-TTS framework can jointly optimize both ASR and TTS \edit{without using paired data}~\parencite{tjandra_listening_2017,hori_cycleconsistency_2019,wang_improving_2020}.
+A speech chain~\parencite{tjandra_listening_2017,tjandra_machine_2018} is one 
 successful way to utilize audio-only and text-only data to train both
@@
 \begin{equation}
-    \hat{X} = g_{\mathrm{tts}}(f_{\mathrm{asr}}(X)).
+    \hat{X} = g_{\mathrm{tts}}(f_{\mathrm{asr}}(X)) \enspace .
     \label{eq:asr-tts}
@@
 \begin{equation}
-    \hat{Y} = f_{\mathrm{asr}}(g_{\mathrm{tts}}(Y)).
+    \hat{Y} = f_{\mathrm{asr}}(g_{\mathrm{tts}}(Y)) \enspace .
     \label{eq:tts-asr}
@@
 Although this approach initially freezes either the ASR or TTS model, 
-extensions of this study~\cite{hori_cycleconsistency_2019,tjandra_endtoend_2019,baskar_semisupervised_2019}
+extensions of this study~\parencite{hori_cycleconsistency_2019,tjandra_endtoend_2019,baskar_semisupervised_2019}
 implement the joint training of both ASR and TTS parameters using 
-REINFORCE~\cite{williams_simple_1992} and straight-through estimators.
+REINFORCE~\parencite{williams_simple_1992} and straight-through estimators.
 
@@
 We combine the ASR and language model via a noisy channel 
-model~\cite{jelinek_statistical_1997}, a weighted finite state 
-transducer~\cite{mohri_weighted_2002}, or shallow 
-fusion~\cite{gulcehre_using_2015,chorowski_better_2017}.
+model~\parencite{jelinek_statistical_1997}, a weighted finite state 
+transducer~\parencite{mohri_weighted_2002}, or shallow 
+fusion~\parencite{gulcehre_using_2015,chorowski_better_2017}.
 However, the progress of TTS systems boosted by deep 
-learning~\cite{oord_wavenet_2016,shen_natural_2018} has inspired another interesting and
+learning~\parencite{oord_wavenet_2016,shen_natural_2018} has inspired another interesting and
 straightforward research direction: \edit{artificially creating paired text and audio
@@
 generate the waveform with various acoustic 
-variations~\cite{li_training_2018,ueno_multispeaker_2019,rosenberg_speech_2019,laptev_you_2020,huang_using_2020}.
+variations~\parencite{li_training_2018,ueno_multispeaker_2019,rosenberg_speech_2019,laptev_you_2020,huang_using_2020}.
 The other approaches are based on the generation of high-level (more
 linguistic) features instead of generating the waveform, e.g., encoder
-features~\cite{hayashi_backtranslationstyle_2018} and phoneme 
-features~\cite{renduchintala_multimodal_2018,masumura_phonemetographeme_2020}.
-This approach is similar to the back-translation technique developed in \edit{neural machine translation}~\cite{sennrich_improving_2016}.
+features~\parencite{hayashi_backtranslationstyle_2018} and phoneme 
+features~\parencite{renduchintala_multimodal_2018,masumura_phonemetographeme_2020}.
+This approach is similar to the back-translation technique developed in \edit{neural machine translation}~\parencite{sennrich_improving_2016}.
 One benefit of the above data generation approaches is that it can be used to feed
@@
 
-\subsubsection{Zero-resource speech technologies and challenges}
+\paragraph{Zero-resource speech technologies and challenges}
 % {\color{black} Shinji}\\
@@
 linguistic unit discovery from speech data without linguistic resources,
-e.g., transcriptions and other annotations~\cite{jansen_summary_2013}.
+e.g., transcriptions and other annotations~\parencite{jansen_summary_2013}.
 This study was motivated by unsupervised query-by-example, applications of
@@
 \begin{itemize}
-	 \item \edit{Zero Resource Speech Challenge 2015~\cite{versteegh_zero_2015} mainly focused on building an acoustic model without using any
+	 \item \edit{Zero Resource Speech Challenge 2015~\parencite{versteegh_zero_2015} mainly focused on building an acoustic model without using any
 	 linguistic annotations based on subword unit modeling and spoken term
@@
     Both tracks were based on the English and Xitsonga languages.
-	 \item The Zero Resource Speech Challenge 2017~\cite{dunbar_zero_2017} focused on 
+	 \item The Zero Resource Speech Challenge 2017~\parencite{dunbar_zero_2017} focused on 
 	 unseen language and speaker aspects from the previous challenge. For
@@
 	 the amount of speech available for each speaker.
-	 \item The Zero Resource Speech Challenge 2019~\cite{dunbar_zero_2019} extended a goal of
+	 \item The Zero Resource Speech Challenge 2019~\parencite{dunbar_zero_2019} extended a goal of
 	 previous challenges by synthesizing speech without
@@
 	 and speaker similarity.
-	 \item The Zero Resource Speech Challenge 2020~\cite{dunbar_zero_2020} was based on two
+	 \item The Zero Resource Speech Challenge 2020~\parencite{dunbar_zero_2020} was based on two
 	 tracks, revisiting previous challenges with different evaluation metrics. 
@@
 	 units from unsegmented raw speech.
-	 \item The Zero Resource Speech Challenge 2021~\cite{nguyen_zero_2020}, the latest
+	 \item The Zero Resource Speech Challenge 2021~\parencite{nguyen_zero_2020}, the latest
 	 \edit{challenge}, expanded the scope to include language modeling tasks.
@@
 techniques have shifted from purely generative 
-models~\cite{ondel_variational_2016,heck_feature_2017}, deep 
-autoencoders~\cite{tjandra_vqvae_2019,chorowski_unsupervised_2019}, and incorporation of
-neural-network-based TTS/VC techniques~\cite{vanniekerk_vectorquantized_2020} to
-self-supervised learning~\cite{maekaku_speech_2021}.
+models~\parencite{ondel_variational_2016,heck_feature_2017}, deep 
+autoencoders~\parencite{tjandra_vqvae_2019,chorowski_unsupervised_2019}, and incorporation of
+neural-network-based TTS/VC techniques~\parencite{vanniekerk_vectorquantized_2020} to
+self-supervised learning~\parencite{maekaku_speech_2021}.
 The latest challenge included the visual modality, continuing the
@@
 
-\subsubsection{Textless NLP}
+\paragraph{Textless NLP}
 % {\color{black} Abdo}\\
@@
 directly from audio, bypassing the need for text or 
-labels~\cite{nguyen_generative_2022,polyak_speech_2021,kharitonov_textfree_2021,kreuk_textless_2022}. 
+labels~\parencite{nguyen_generative_2022,polyak_speech_2021,kharitonov_textfree_2021,kreuk_textless_2022}. 
 Not only does this open the gate for language and dialect modeling without
@@
 emotion, hesitation, interruptions. 
-The generative spoken language model (GSLM)~\cite{nguyen_generative_2022} utilizes discrete
+The generative spoken language model (GSLM)~\parencite{nguyen_generative_2022} utilizes discrete
 representations from wav2vec~2.0, HuBERT, and CPC algorithms as inputs to an
@@
 model and generate
-dialogues~\cite{nguyen_generative_2022}.\footnote{https://speechbot.github.io/dgslm/}
-%\cite{}. 
+dialogues~\parencite{nguyen_generative_2022}.\footnote{https://speechbot.github.io/dgslm/}
+%\parencite{}. 
 Given its flexibility in modeling spoken content, the GSLM has been further extended
-to jointly model content and prosody~\cite{kharitonov_textfree_2021}. This prosodic-GSLM model
+to jointly model content and prosody~\parencite{kharitonov_textfree_2021}. This prosodic-GSLM model
 introduced a multistream causal Transformer, where the input and output layers
@@
 style of the prompt.\footnote{https://speechbot.github.io/pgslm/}
-%\cite{}. 
-Going one step further, \cite{kreuk_textless_2022} used a speech emotion
+%\parencite{}. 
+Going one step further, \parencite{kreuk_textless_2022} used a speech emotion
 conversion framework to modify the perceived emotion of a speech utterance
@@
 extended the idea of textless language processing or audio discrete representation to applications such as 
-spoken question answering~\cite{lin_dual_2022}, speech separation~\cite{shi_discretization_2022}, TTS~\cite{hayashi_discretalk_2020}, and speech-to-speech
-translation~\cite{lee_textless_2022}.
+spoken question answering~\parencite{lin_dual_2022}, speech separation~\parencite{shi_discretization_2022}, TTS~\parencite{hayashi_discretalk_2020}, and speech-to-speech
+translation~\parencite{lee_textless_2022}.

### 2023-09-29 chapters/paper_review/06_conclusions.tex
Add and revise short table and figure captions
Updates to citation commands and footnotes

--- 2023-03-04 chapters/paper_review/06_conclusions.tex
@@
 
-\section{Discussion and Conclusion}
+\section{Discussion and conclusion}
 \label{sec:conclusion}
@@
 \edit{Some efficient methods for leveraging SSL models exist in the NLP community. 
-Adapters~\cite{pmlr-v97-houlsby19a,zaken2021bitfit,guo-etal-2021-parameter} are lightweight modules inserted into SSL models, and in downstream tasks, the parameters of SSL models are frozen, and only the adapters are trained.
-The prompt/instruction learning methods~\cite{liu2021pretrain} also freeze the SSL parameters and control the output of SSL by adding additional information, which is called \textit{prompt}, in the input.
-Both adapter-based methods and prompt/instruction learning yield competitive performance compared with fine-tuning in NLP applications, but there is only little related work for speech~\cite{SpeechAdapter,SpeechPrompt}.
+Adapters~\parencite{houlsby_parameterefficient_2019,zaken_bitfit_2022,guo_parameterefficient_2021} are lightweight modules inserted into SSL models, and in downstream tasks, the parameters of SSL models are frozen, and only the adapters are trained.
+The prompt/instruction learning methods~\parencite{liu_pretrain_2021} also freeze the SSL parameters and control the output of SSL by adding additional information, which is called \textit{prompt}, in the input.
+Both adapter-based methods and prompt/instruction learning yield competitive performance compared with fine-tuning in NLP applications, but there is only little related work for speech~\parencite{thomas_efficient_2022,chang_speechprompt_2022}.
 In addition, prompt for speech SSL does not achieve comparable performance on sequence generation tasks like phoneme recognition and slot filling, so how to use prompt is still an open question.}
@@
 This} makes them unsuitable for edge devices \edit{but also limits the ability to scale these models to very large datasets} \ci{ -- and leads to a large energy consumption}.
-Preliminary studies have been conducted on compressing speech representation models through network pruning~\cite{PARP} or knowledge distillation~\cite{chang2021distilhubert}. 
-\edit{There has been quite some effort towards more efficient general neural network models via conditional computing \cite{bengio_conditional_2016} and neural network quantization \cite{gholami_survey_2021} as well as extensive work on improving the specific efficiency of Transformer models, especially with the focus on self-attention \cite{tay_efficient_2022}, but these technology has not been widely used in speech SSL.}
+Preliminary studies have been conducted on compressing speech representation models through network pruning~\parencite{lai_parp_2021} or knowledge distillation~\parencite{chang_distilhubert_2021}. 
+\edit{There has been quite some effort towards more efficient general neural network models via conditional computing \parencite{bengio_conditional_2016} and neural network quantization \parencite{gholami_survey_2021} as well as extensive work on improving the specific efficiency of Transformer models, especially with the focus on self-attention \parencite{tay_efficient_2022}, but these technology has not been widely used in speech SSL.}
 \edit{Because speech is intrinsically represented as sequence, one way to reduce computation is to reduce the length of speech representation sequence but still keep the vital information in speech.
 But we have not been aware of any publication in this direction when writing this paper.}
-On the other hand, non-streaming architectures in models such as the bidirectional Transformer have hindered the representation model used in streaming scenarios, leading to studies that address these problems~\cite{StreamingW2v}.
+On the other hand, non-streaming architectures in models such as the bidirectional Transformer have hindered the representation model used in streaming scenarios, leading to studies that address these problems~\parencite{cao_improving_2021}.
 We anticipate research in these directions to continue in the future.
@@
 \item \textbf{Feature Disentanglement.} 
-Speech SSL models show strengths on a surprisingly wide range of tasks~\cite{yang21c_interspeech}, suggesting that representations contain different information.
+Speech SSL models show strengths on a surprisingly wide range of tasks~\parencite{yang_superb_2021}, suggesting that representations contain different information.
 One way to further improve downstream tasks is to disentangle different information from the representation.
 For example, we can decompose the representation into content embedding and speaker embedding and use content embedding for ASR and speaker embedding for SID.
-Some work has been in this direction~\cite{pmlr-v162-qian22b,NANCY,MultimodelAmazonIS22}.
+Some work has been in this direction~\parencite{qian_contentvec_2022,choi_neural_2021,chan_contentcontext_2022}.
 
-\item \textbf{Creating robust models.} \edit{As discussed in \cref{subsec:robustness}, studies have been conducted on the robustness of representation models \cite{wu2021characterizing}.
+\item \textbf{Creating robust models.} \edit{As discussed in \cref{subsec:robustness}, studies have been conducted on the robustness of representation models \parencite{wu_characterizing_2022}.
 However, the failure modes of SSL models are still poorly understood, and it remains unclear whether they provide more or less robustness to adversarial attacks than fully supervised models.
-Due to the importance of this research direction, while writing this paper, there is already some related research about enhancing the robustness of SSL models~\cite{robustw2v2,huang2022improving,wang2022improving,zhu2022noise} and identifying their vulnerability to adversarial attack~\cite{wu2021characterizing}.
+Due to the importance of this research direction, while writing this paper, there is already some related research about enhancing the robustness of SSL models~\parencite{hsu_robust_2021,huang_improving_2022,wang_improving_2022,zhu_noiserobust_2022} and identifying their vulnerability to adversarial attack~\parencite{wu_characterizing_2022}.
 }
 
-\edit{\item \textbf{Capturing higher-level semantic information.} Although many representation learning approaches can go beyond low-level phonetic modeling to capture some lexical information~\cite{TuAnh_2022_discrete}, they still struggle in higher-level semantic tasks easily captured by word-level counterparts like BERT.
-One workaround is two-stage training~\cite{pGSLM, dialogue_GSLM}; however, this prevents propagating rich lexical and semantic knowledge modeled in the second stage to benefit the phonetically focused first stage.
-}
+\item \textbf{Capturing higher-level semantic information.} Although many representation learning approaches can go beyond low-level phonetic modeling to capture some lexical information~\parencite{nguyen_are_2022}, they still struggle in higher-level semantic tasks easily captured by word-level counterparts like BERT.
+One workaround is two-stage training~\parencite{kharitonov_textfree_2021, nguyen_generative_2022}; however, this prevents propagating rich lexical and semantic knowledge modeled in the second stage to benefit the phonetically focused first stage.
 
 \item \textbf{Using text representation models to improve speech representation.} The amount of content information in speech corpora used to train speech representation models is far less than that of text representation models.
-Noting that the BERT training corpus exceeds 3 billion words~\cite{jacob2019BERT}, and assuming a typical speaking rate of 120 words per minute, a speech corpus containing the same content as the BERT training data would include 400,000 hours of audio, which exceeds the \edit{accumulated} training data of all current speech representation models.
+Noting that the BERT training corpus exceeds 3 billion words~\parencite{devlin_bert_2018}, and assuming a typical speaking rate of 120 words per minute, a speech corpus containing the same content as the BERT training data would include 400,000 hours of audio, which exceeds the \edit{accumulated} training data of all current speech representation models.
 Therefore, to enable speech representation models to better learn human language, for instance by extracting semantic information from acoustic signals, the use of text models such as BERT and GPT  \edit{seems} key: nevertheless, how to use these to improve speech representation model pre-training remains an open question.
-%\edit{ There is already some study using both speech and text data to pre-training, but some paired data is still required to achieve good performance~\cite{bapna2021slam}.  }
+%\edit{ There is already some study using both speech and text data to pre-training, but some paired data is still required to achieve good performance~\parencite{bapna_slam_2021}.  }
 

### 2023-09-29 chapters/paper_review/main_hi.tex
Work on variational autoencoders
Move table captions to top
Updates to citation commands and footnotes

--- 2023-03-16 chapters/paper_review/main_hi.tex
@@
 \begin{table*}[t]
+    \caption{Comparison with the literature on high resource setups using all 960 hours of labeled LibriSpeech data.}
     \centering
@@
         \multicolumn{7}{c}{\textit{\textbf{Superivsed}}} \\
-        Conformer L~\cite{gulati2020conformer} & - & LSTM & - & - & 1.9 & 3.9 \\
+        Conformer L~\parencite{gulati_conformer_2020} & - & LSTM & - & - & 1.9 & 3.9 \\
 
@@
         \multicolumn{7}{c}{\textit{\textbf{Self-Training}}} \\
-        IPL~\cite{xu_iterative_2020} & LL-60k & 4-gram + Transformer & 1.85 & 3.26 & 2.10 & 4.01 \\
-        Noisy Student~\cite{park_improved_2020} & LV-60k & LSTM & 1.6 & 3.4 & 1.7 & 3.4 \\
+        IPL~\parencite{xu_iterative_2020} & LL-60k & 4-gram + Transformer & 1.85 & 3.26 & 2.10 & 4.01 \\
+        Noisy Student~\parencite{park_improved_2020} & LV-60k & LSTM & 1.6 & 3.4 & 1.7 & 3.4 \\
 
@@
         \multicolumn{7}{c}{\textit{\textbf{Pre-Training}}} \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec} & LL-60k & Transformer & 1.6 & 3.0 & 1.8 & 3.3 \\
-        pre-trained Conformer XXL~\cite{zhang2020pushing} & LL-60k & LSTM & 1.5 & 3.0 & 1.5 & 3.1 \\
+        wav2vec 2.0 \textsc{Large}~\parencite{baevski_wav2vec_2020} & LL-60k & Transformer & 1.6 & 3.0 & 1.8 & 3.3 \\
+        pre-trained Conformer XXL~\parencite{zhang_pushing_2020} & LL-60k & LSTM & 1.5 & 3.0 & 1.5 & 3.1 \\
 
@@
         \multicolumn{7}{c}{\textit{\textbf{Pre-Training + Self-Training}}} \\
-        wav2vec 2.0 + self-training~\cite{xu2020self} & LL-60k & Transformer & 1.1 & 2.7 & 1.5 & 3.1\\
-        pre-trained Conformer XXL + Noisy Student~\cite{zhang2020pushing} & LL-60k & LSTM & 1.3 & 2.6 & 1.4 & 2.6 \\
+        wav2vec 2.0 + self-training~\parencite{xu2020self} & LL-60k & Transformer & 1.1 & 2.7 & 1.5 & 3.1\\
+        pre-trained Conformer XXL + Noisy Student~\parencite{zhang_pushing_2020} & LL-60k & LSTM & 1.3 & 2.6 & 1.4 & 2.6 \\
 
@@
     \end{tabular}
-    \caption{Comparison with the literature on high resource setups using all 960 hours of labeled LibriSpeech data.}
     \label{tab:main_hi}

### 2023-09-29 chapters/paper_review/main_lo.tex
Move table captions to top
Updates to citation commands and footnotes

--- 2023-03-16 chapters/paper_review/main_lo.tex
@@
 \begin{table*}[t]
+    \caption{Results and comparison with the literature on low resource setups (10-min, 1-hour, 10-hour, and 100-hour of labeled data).}
     \centering
@@
         \multicolumn{7}{c}{\textit{\textbf{10-min labeled}}} \\
-        % DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & - & - \\
-        DiscreteBERT~\cite{baevski_effectiveness_2020} & LS-960 & 4-gram & 15.7 & 24.1 & 16.3 & 25.2 \\
-        wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 8.9 & 15.7 & 9.1 & 15.6 \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & 4-gram & 6.3 & 9.8 & 6.6 & 10.3  \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & Transformer & 4.6 & 7.9 & 4.8 & 8.2 \\
+        % DeCoAR 2.0~\parencite{ling_decoar_2020} & LS-960 & 4-gram & - & - & - & - \\
+        DiscreteBERT~\parencite{baevski_effectiveness_2020} & LS-960 & 4-gram & 15.7 & 24.1 & 16.3 & 25.2 \\
+        wav2vec 2.0 \textsc{Base}~\parencite{baevski_wav2vec_2020}   & LS-960 & 4-gram & 8.9 & 15.7 & 9.1 & 15.6 \\
+        wav2vec 2.0 \textsc{Large}~\parencite{baevski_wav2vec_2020}  & LL-60k & 4-gram & 6.3 & 9.8 & 6.6 & 10.3  \\
+        wav2vec 2.0 \textsc{Large}~\parencite{baevski_wav2vec_2020}  & LL-60k & Transformer & 4.6 & 7.9 & 4.8 & 8.2 \\
         \midrule
@@
         \multicolumn{7}{c}{\textit{\textbf{1-hour labeled}}} \\
-        DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & 13.8 & 29.1 \\
-        DiscreteBERT~\cite{baevski_effectiveness_2020} & LS-960 & 4-gram & 8.5 & 16.4 & 9.0 & 17.6 \\
-        wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 5.0 & 10.8 & 5.5 & 11.3 \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & Transformer & 2.9 & 5.4 & 2.9 & 5.8 \\
+        DeCoAR 2.0~\parencite{ling_decoar_2020} & LS-960 & 4-gram & - & - & 13.8 & 29.1 \\
+        DiscreteBERT~\parencite{baevski_effectiveness_2020} & LS-960 & 4-gram & 8.5 & 16.4 & 9.0 & 17.6 \\
+        wav2vec 2.0 \textsc{Base}~\parencite{baevski_wav2vec_2020}   & LS-960 & 4-gram & 5.0 & 10.8 & 5.5 & 11.3 \\
+        wav2vec 2.0 \textsc{Large}~\parencite{baevski_wav2vec_2020}  & LL-60k & Transformer & 2.9 & 5.4 & 2.9 & 5.8 \\
         \midrule
@@
         \multicolumn{7}{c}{\textit{\textbf{10-hour labeled}}} \\
-        SlimIPL~\cite{likhomanenko2020slimipl} & LS-960 & 4-gram + Transformer & 5.3 & 7.9 & 5.5 & 9.0 \\
-        DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & 5.4 & 13.3 \\
-        DiscreteBERT~\cite{baevski_effectiveness_2020} & LS-960 & 4-gram & 5.3 & 13.2 & 5.9 & 14.1 \\
-        wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 3.8 & 9.1 & 4.3 & 9.5 \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & Transformer & 2.4 & 4.8 & 2.6 & 4.9 \\
+        SlimIPL~\parencite{likhomanenko_slimipl_2021} & LS-960 & 4-gram + Transformer & 5.3 & 7.9 & 5.5 & 9.0 \\
+        DeCoAR 2.0~\parencite{ling_decoar_2020} & LS-960 & 4-gram & - & - & 5.4 & 13.3 \\
+        DiscreteBERT~\parencite{baevski_effectiveness_2020} & LS-960 & 4-gram & 5.3 & 13.2 & 5.9 & 14.1 \\
+        wav2vec 2.0 \textsc{Base}~\parencite{baevski_wav2vec_2020}   & LS-960 & 4-gram & 3.8 & 9.1 & 4.3 & 9.5 \\
+        wav2vec 2.0 \textsc{Large}~\parencite{baevski_wav2vec_2020}  & LL-60k & Transformer & 2.4 & 4.8 & 2.6 & 4.9 \\
         \midrule
@@
         \multicolumn{7}{c}{\textit{\textbf{100-hour labeled}}} \\
-        % LPM~\cite{hsu2020semi} & LL-60K & ConvLM & 4.87 & 13.84 & 4.88 & 15.28 \\
-        IPL~\cite{xu_iterative_2020} & LL-60k & 4-gram + Transformer & 3.19 & 6.14 & 3.72 & 7.11 \\
-        SlimIPL~\cite{likhomanenko2020slimipl} & LS-860 & 4-gram + Transformer & 2.2 & 4.6 & 2.7 & 5.2 \\
-        Noisy Student\cite{park_improved_2020} & LS-860 & LSTM & 3.9 & 8.8 & 4.2 & 8.6 \\
-        DeCoAR 2.0~\cite{ling2020decoar} & LS-960 & 4-gram & - & - & 5.0 & 12.1 \\
-        DiscreteBERT~\cite{baevski_effectiveness_2020} & LS-960 & 4-gram & 4.0 & 10.9 & 4.5 & 12.1 \\
-        wav2vec 2.0 \textsc{Base}~\cite{baevski2020wav2vec}   & LS-960 & 4-gram & 2.7 & 7.9 & 3.4 & 8.0 \\
-        wav2vec 2.0 \textsc{Large}~\cite{baevski2020wav2vec}  & LL-60k & Transformer & 1.9 & 4.0 & 2.0 & 4.0 \\
+        % LPM~\parencite{hsu2020semi} & LL-60K & ConvLM & 4.87 & 13.84 & 4.88 & 15.28 \\
+        IPL~\parencite{xu_iterative_2020} & LL-60k & 4-gram + Transformer & 3.19 & 6.14 & 3.72 & 7.11 \\
+        SlimIPL~\parencite{likhomanenko_slimipl_2021} & LS-860 & 4-gram + Transformer & 2.2 & 4.6 & 2.7 & 5.2 \\
+        Noisy Student\parencite{park_improved_2020} & LS-860 & LSTM & 3.9 & 8.8 & 4.2 & 8.6 \\
+        DeCoAR 2.0~\parencite{ling_decoar_2020} & LS-960 & 4-gram & - & - & 5.0 & 12.1 \\
+        DiscreteBERT~\parencite{baevski_effectiveness_2020} & LS-960 & 4-gram & 4.0 & 10.9 & 4.5 & 12.1 \\
+        wav2vec 2.0 \textsc{Base}~\parencite{baevski_wav2vec_2020}   & LS-960 & 4-gram & 2.7 & 7.9 & 3.4 & 8.0 \\
+        wav2vec 2.0 \textsc{Large}~\parencite{baevski_wav2vec_2020}  & LL-60k & Transformer & 1.9 & 4.0 & 2.0 & 4.0 \\
 
@@
     \end{tabular}
-    \caption{Results and comparison with the literature on low resource setups (10-min, 1-hour, 10-hour, and 100-hour of labeled data).}
     \label{tab:main_lo}

### 2023-09-29 chapters/paper_review/model.tex
Work on variational autoencoders
Move table captions to top
Updates to citation commands and footnotes

--- 2023-03-17 chapters/paper_review/model.tex
@@
 
-\begin{table*}[h!]
+\begin{table*}[t]
 \caption{
@@
 
-Audio Word2vec \cite{chung_audio_2016}      & 2016 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
+Audio Word2vec \parencite{chung_audio_2016}      & 2016 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
 
-VQ-VAE \cite{oord_neural_2018} & 2017 Nov. & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark \\ %2 Nov 2017
+VQ-VAE \parencite{oord_neural_2018} & 2017 Nov. & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark \\ %2 Nov 2017
 
-Speech2Vec \cite{chung_speech2vec_2018}     & 2018 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
-Unspeech \cite{milde_unspeech_2018}         & 2018 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
+Speech2Vec \parencite{chung_speech2vec_2018}     & 2018 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
+Unspeech \parencite{milde_unspeech_2018}         & 2018 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
 
-CPC \cite{oord_representation_2018}         & 2018 Jul. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\
+CPC \parencite{oord_representation_2018}         & 2018 Jul. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\
 
-PASE \cite{pascual_learning_2019}                   & 2019 Apr. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 6/4
+PASE \parencite{pascual_learning_2019}                   & 2019 Apr. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 6/4
 
-APC \cite{chung_unsupervised_2019}          & 2019 Oct. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 5/4
+APC \parencite{chung_unsupervised_2019}          & 2019 Oct. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 5/4
 
-wav2vec \cite{schneider_wav2vec:_2019}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
+wav2vec \parencite{schneider_wav2vec_2019}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
 
-vq-wav2vec \cite{baevski_vq-wav2vec_2020} & 2019 Oct.
+vq-wav2vec \parencite{baevski_vqwav2vec_2020} & 2019 Oct.
 & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ %12 Oct 2019 %Lee: vq-wav2vec is CON or CLF???
@@
 
-Discrete BERT \cite{baevski_vq-wav2vec_2020,baevski_effectiveness_2020} & 2019 Oct.
+Discrete BERT \parencite{baevski_vqwav2vec_2020,baevski_effectiveness_2020} & 2019 Oct.
 & \xmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ %12 Oct 2019 %10 Nov 2019 %In the paper of vq-wav2vec, it also includes the approach of Discrete BERT.
@@
 
-Mockingjay  \cite{liu_mockingjay_2020}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 25/10-19
+Mockingjay  \parencite{liu_mockingjay_2020}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 25/10-19
 
-DeCoAR \cite{ling2020deep}                & 2019 Dec. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 3/12-19
+DeCoAR \parencite{ling_deep_2020}                & 2019 Dec. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 3/12-19
 
-PASE+ \cite{ravanelli_multi-task_2020}                & 2020 Jan. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 25/1 % I think contrastive always have learned target.
+PASE+ \parencite{ravanelli_multitask_2020}                & 2020 Jan. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 25/1 % I think contrastive always have learned target.
 
-Bidir CPC \cite{kawakami_learning_2020} & 2020 Jan. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\% 29/1
+Bidir CPC \parencite{kawakami_learning_2020} & 2020 Jan. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\% 29/1
 
-Modified CPC \cite{riviere_unsupervised_2020} & 2020 Feb. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 7/2
+Modified CPC \parencite{riviere_unsupervised_2020} & 2020 Feb. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 7/2
 
-VQ-APC \cite{chung_vector-quantized_2020}  & 2020 May  & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark \\ %17/05
+VQ-APC \parencite{chung_vectorquantized_2020}  & 2020 May  & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark \\ %17/05
 
-AALBERT \cite{chi_audio_2020}  & 2019 May & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ %18 May 2020
+AALBERT \parencite{chi_audio_2020}  & 2019 May & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ %18 May 2020
 
-speech-XLNet \cite{song_speech-xlnet_2020} & 2019 Oct.
+speech-XLNet \parencite{song_speechxlnet_2020} & 2019 Oct.
 & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\%23 Oct 2019 %Lee: It is speech version of XLNet.
@@
 
-wav2vec 2.0 \cite{baevski_wav2vec_2020}     & 2020 Jun. & \xmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ % 20/6
+wav2vec 2.0 \parencite{baevski_wav2vec_2020}     & 2020 Jun. & \xmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ % 20/6
 %Lee: In my point of view, wav2vec 2.0 has learned target. The target is from the CNN encoder.
 
-MPE \cite{liu_masked_2020} & 2020 May & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 25 May 2020 %Lee: I think it has learned target. But please check.
+MPE \parencite{liu_masked_2020} & 2020 May & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 25 May 2020 %Lee: I think it has learned target. But please check.
 
-TERA \cite{liu_tera_2021}                   & 2020 Jul. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 12/7
+TERA \parencite{liu_tera_2021}                   & 2020 Jul. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 12/7
 
-Speech SimCLR \cite{jiang_speech_2021} &  2020 Oct.
+Speech SimCLR \parencite{jiang_speech_2021} &  2020 Oct.
 & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \xmark \\ %Lee: It is TERA+ SIMCLR.
@@
 
-NPC \cite{liu_non-autoregressive_2020}                     & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ % 1/11
+NPC \parencite{liu_nonautoregressive_2020}                     & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ % 1/11
 
-DeCoAR 2.0 \cite{ling_decoar_2020}          & 2020 Dec.
+DeCoAR 2.0 \parencite{ling_decoar_2020}          & 2020 Dec.
 & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ % 11/12 %Lee:  DeCoAR 2.0 has VQ layer although it is not used as learning target.
 
-wav2vec-C \cite{sadhu_wav2vec-c_2021} & 2021 Mar. & \cmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ %9 Mar 2021 %Lee:  consistency loss is reconstruction
+wav2vec-C \parencite{sadhu_wav2vecc_2021} & 2021 Mar. & \cmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark \\ %9 Mar 2021 %Lee:  consistency loss is reconstruction
 
-HuBERT \cite{hsu_hubert_2021}               & 2021 Jun. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ % 14/6
+HuBERT \parencite{hsu_hubert_2021}               & 2021 Jun. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ % 14/6
 
-W2v-bert \cite{chung_w2vbert_2021} & 2021 Aug & \xmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark \\
+W2v-bert \parencite{chung_w2vbert_2021} & 2021 Aug & \xmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark \\
 %7 Aug 2021
 
-pMPC \cite{yue_phonetically_2021}  & 2021 Aug. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ %30 August – 3 September, 2021 (I do not find arXiv version.)
+pMPC \parencite{yue_phonetically_2021}  & 2021 Aug. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ %30 August – 3 September, 2021 (I do not find arXiv version.)
 
-WavLM \cite{chen_wavlm_2021} & 2021 Oct. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ %26 Oct 2021
+WavLM \parencite{chen_wavlm_2021} & 2021 Oct. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ %26 Oct 2021
 
-data2vec \cite{baevski_data2vec_2022}     & 2022 Jan. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 20/1
+data2vec \parencite{baevski_data2vec_2022}     & 2022 Jan. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 20/1
 
-BEST-RQ \cite{BEST-RQ} & 2022 Feb.  & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ %Lee: no learned target. The target is random ...
+BEST-RQ \parencite{chiu_selfsupervised_2022} & 2022 Feb.  & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark \\ %Lee: no learned target. The target is random ...
 % 3 Feb 2022
@@
 %                                                         REG      CLF      CON      MSK      LTG      QTZ      EMB  
-Audio Word2vec \cite{chung_audio_2016}      & 2016 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
-Speech2Vec \cite{chung_speech2vec_2018}     & 2018 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
-% Unspeech \cite{milde_unspeech_2018}         & 2018 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
-CPC \cite{oord_representation_2018}         & 2018 Jul. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\
+Audio Word2vec \parencite{chung_audio_2016}      & 2016 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
+Speech2Vec \parencite{chung_speech2vec_2018}     & 2018 Mar. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
+% Unspeech \parencite{milde_unspeech_2018}         & 2018 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
+CPC \parencite{oord_representation_2018}         & 2018 Jul. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\
 
-PASE \cite{pascual_learning_2019}                   & 2019 Apr. & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark \\ % 6/4
+PASE \parencite{pascual_learning_2019}                   & 2019 Apr. & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark \\ % 6/4
 
-APC \cite{chung_unsupervised_2019}          & 2019 Oct. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 5/4
-wav2vec \cite{schneider_wav2vec:_2019}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
-%VQ-APC \cite{chung2020vector}             & 2020 May  & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ 
-Mockingjay  \cite{liu_mockingjay_2020}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 25/10-19
-%DeCoAR \cite{ling2020deep}                & 2019 Dec. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 3/12-19
-%PASE+ \cite{Ravanelli2020}                & 2020 Jan. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 25/1
-wav2vec 2.0 \cite{baevski_wav2vec_2020}     & 2020 Jun. & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark \\ % 20/6
-%TERA \cite{liu_tera_2021}                   & 2021 Jul. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 12/7
+APC \parencite{chung_unsupervised_2019}          & 2019 Oct. & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 5/4
+wav2vec \parencite{schneider_wav2vec_2019}       & 2019 Apr. & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark \\ % 11/4
+%VQ-APC \parencite{chung_vectorquantized_2020}             & 2020 May  & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ 
+Mockingjay  \parencite{liu_mockingjay_2020}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 25/10-19
+%DeCoAR \parencite{ling_deep_2020}                & 2019 Dec. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 3/12-19
+%PASE+ \parencite{ravanelli_multitask_2020}                & 2020 Jan. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 25/1
+wav2vec 2.0 \parencite{baevski_wav2vec_2020}     & 2020 Jun. & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark \\ % 20/6
+%TERA \parencite{liu_tera_2021}                   & 2021 Jul. & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ % 12/7
 
-NPC \cite{liu_non-autoregressive_2020}             & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 1/11
+NPC \parencite{liu_nonautoregressive_2020}             & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 1/11
 
-DeCoAR 2.0 \cite{ling_decoar_2020}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 11/12
-HuBERT \cite{hsu_hubert_2021}               & 2021 Jun. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ % 14/6
-data2vec \cite{baevski_data2vec_2022}     & 2022 Jan. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 20/1
+DeCoAR 2.0 \parencite{ling_decoar_2020}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\ % 11/12
+HuBERT \parencite{hsu_hubert_2021}               & 2021 Jun. & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark \\ % 14/6
+data2vec \parencite{baevski_data2vec_2022}     & 2022 Jan. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark \\ % 20/1
 \bottomrule

### 2023-09-29 chapters/paper_review/pretext.tex
Move table captions to top
Updates to citation commands and footnotes

@@
 
-\begin{table*}[!htb]
+\begin{table}[h!]
     \centering
@@
     is described by the overall model category and the main text.
+    Notation details are in \cref{footnote:table:pretext}.
+    }
+    % \begin{minipage}{\textwidth}  
+\centering
+\resizebox{0.98\textwidth}{!}{
+\begin{tabular}{l|c|c|c}
+    \toprule
+    \textbf{Model} (a) & \textbf{Input} (b) & \textbf{Corruption} (c) & \textbf{Target} (d) \\
+    \midrule
+    \midrule
+    \multicolumn{4}{c}{\textsc{Generative models}} \\
+    \midrule
+    \midrule
+    Audio Word2vec~\parencite{chung_audio_2016}, VQ-VAE~\parencite{oord_neural_2018}     & $X$ &   \textsc{-}    &  $X$  \\ %It has VQ as extra constraint.
+    \midrule  
+    Speech2Vec~\parencite{chung_speech2vec_2018}, Audio2Vec~\parencite{tagliasacchi_pretraining_2020} - skip-gram    & $X_{[t_1,t_2]}$  &     \textsc{-}   &    $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$     \\
+    \midrule 
+    Speech2Vec~\parencite{chung_speech2vec_2018}, Audio2Vec~\parencite{tagliasacchi_pretraining_2020} - cbow    & $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$   &     \textsc{-}   &    $X_{[t_1,t_2]}$   \\
+    \midrule 
+    PASE~\parencite{pascual_learning_2019}, PASE+~\parencite{ravanelli_multitask_2020}\textsuperscript{\ref{footnote_review:pase}}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
+    \midrule 
+    APC~\parencite{chung_unsupervised_2019,chung_vectorquantized_2020}         & $X_{[1,t]}$   & \textsc{-}              & $x_{t+c},\, c\geq1$    \\ %VQ-APC is the same as APC
+    \midrule
+    Speech-XLNet \parencite{song_speechxlnet_2020}     & \multicolumn{2}{c|}{$X_{\mathcal{P}_{t}}$}   &     $x_{i\sim\mathcal{P}^c_{t}}$  \\ %Lee: Use * to represent the permutation, hope it is not strange.
+    \midrule  
+    DeCoAR~\parencite{ling_deep_2020}     & $X_{[1,t-1]}, X_{[t+k+1,T]}$ & \textsc{-} & $X_{[t,t+k]}$   \\
+    \midrule
+    Mockingjay~\parencite{liu_mockingjay_2020}, Audio ALBERT~\parencite{chi_audio_2020}, DeCoAR 2.0~\parencite{ling_decoar_2020}   & \multicolumn{2}{c|}{$X_{-[t,t+k]}$}   & $X_{[t,t+k]}$    \\
+    \midrule 
+    TERA~\parencite{liu_tera_2021}, BMR~\parencite{wang_unsupervised_2020}  & \multicolumn{2}{c|}{$X_{-[t,t+k]}^{-[f,f+j]}$}       & $X$       \\
+    \midrule
+    pMPC~\parencite{yue_phonetically_2021}      &  \multicolumn{2}{c|}{$X_{-[t,t+k^\prime]}$ ($X_{[t,t+k^\prime]}$ is a phoneme)}       & $X_{[t,t+k^\prime]}$    \\
+    \midrule 
+    MPE~\parencite{liu_masked_2020} & $X$ &  $Z_{-[t,t+k]}$  & $Z$    \\ %Lee: What is the difference between MPE and NPC???
+And I believe it reconstruct the convolutional blocks output (learned target?)
+    \midrule
+    NPC~\parencite{liu_nonautoregressive_2020}      & $X$  &   $Z_{-[t,t+k]}$  &    $X$   \\ %Lee: I am not 100% sure it is correct. Please check.
+    \midrule
+    \midrule
+    \multicolumn{4}{c}{\textsc{Contrastive models}} \\
+    \midrule
+    \midrule
+    Unspeech \parencite{milde_unspeech_2018}       &   $X_{[t_1,t_2]}$ &   \textsc{-}   &  $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$ \\
+    \midrule 
+    CPC~\parencite{oord_representation_2018}, wav2vec \parencite{schneider_wav2vec_2019}, Modified CPC \parencite{riviere_unsupervised_2020}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
+    \midrule 
+    Bidirectional CPC \parencite{kawakami_learning_2020}      & $X_{[1,t]}$ or $X_{[t,T]}$ &  \textsc{-}    &    $z_{t+c}$ or $z_{t-c},\, c\geq1$   \\
+    \midrule 
+    vq-wav2vec \parencite{baevski_vqwav2vec_2020}     &   $X_{[1,t]}$ &   \textsc{-}    &   $q_{t+c},\, c\geq1$   \\ 
+    \midrule 
+    wav2vec 2.0 \parencite{baevski_wav2vec_2020}, wav2vec-C \parencite{sadhu_wav2vecc_2021}\textsuperscript{\ref{footnote_review:wav2vec-c}}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
+    \midrule 
+    w2v-BERT \parencite{chung_w2vbert_2021}     &$X$ &    $Z_{-[t,t+k]}$   &     $Q_{[t,t+k]}$ and $C_{[t,t+k]}$     \\
+    \midrule
+    Speech SimCLR \parencite{jiang_speech_2021}\footnote{Speech SimCLR targets the latent representation of an augmented version of $X$ using a differently augmented $X$, and vice-versa.}    & \multicolumn{2}{c|}{$\textsc{augment}_1(X)$ and $\textsc{augment}_2(X)$}     &    $\textsc{augment}_2(Z)$ and $\textsc{augment}_1(Z)$   \\
+    \midrule
+    \midrule 
+    \multicolumn{4}{c}{\textsc{Predictive models}} \\
+    \midrule
+    \midrule
+    Discrete BERT~\parencite{baevski_vqwav2vec_2020,baevski_effectiveness_2020} \textsuperscript{\ref{footnote_review:discrete-bert}}      &   \multicolumn{2}{c|}{$C_{-[t,t+k]}$}   & $C_{[t,t+k]}$  \\
+    \midrule 
+    HuBERT \parencite{hsu_hubert_2021}\textsuperscript{\ref{footnote_review:hubert}}, WavLM \parencite{chen_wavlm_2021}\textsuperscript{\ref{footnote_review:wavlm}}  & $X$             & $Z_{-[t,t+k]}$          & $C_{[t,t+k]}$  \\
+    \midrule
+    data2vec \parencite{baevski_data2vec_2022}    & $X$             & $Z_{-[t,t+k]}$          & $\sum_{l}\bar{H}^{(l)}_{[t,t+k]}$  \\ 
+    \midrule 
+    BEST-RQ \parencite{chiu_selfsupervised_2022}\textsuperscript{\ref{footnote_review:best-rq}}    &  \multicolumn{2}{c|}{$X_{-[t,t+k]}$}      &  $C_{[t,t+k]}$   \\ 
+    \midrule 
+    \bottomrule
+\end{tabular}
+}
+
+% \end{minipage}
+\label{table:pretext}
+\end{table}
+
+% \addtocounter{footnote}{1}
+\footnotetext{\label{footnote:table:pretext}
 $X=\{x_1,x_2,...,x_T\}$ is the input sequence in which $x_t$ can be an
@@
 % $X^\prime$ is an augmented version of $X$ (e.g., $X$ with reverberation), while $X^{\prime\prime}$ is $X$ adding distortion different from  $X^\prime$.
-    $Z$ represents a localized latent representation sequence of $X$.
-%, and $Z^\prime$ and $Z^{\prime\prime}$ are the latent representation sequences of  $X^\prime$ and $X^{\prime\prime}$, respectively.
+$Z$ represents a localized latent representation sequence of $X$. %, and $Z^\prime$ and $Z^{\prime\prime}$ are the latent representation sequences of  $X^\prime$ and $X^{\prime\prime}$, respectively. 
 % $Z$ represents a localized latent representation sequence and $Z^\prime$ is the latent representation sequence of $X^\prime$. 
@@
 a sequence of discrete cluster IDs.
-    For contrastive models, we specify only positive targets.
-    }
-    % \begin{minipage}{\textwidth}  
-\centering
-\resizebox{0.98\textwidth}{!}{
-\begin{tabular}{l|c|c|c}
-    \toprule
-    \textbf{Model} (a) & \textbf{Input} (b) & \textbf{Corruption} (c) & \textbf{Target} (d) \\
-    \midrule
-    \midrule
-    \multicolumn{4}{c}{\textsc{Generative models}} \\
-    \midrule
-    \midrule
-    Audio Word2vec~\cite{chung_audio_2016}, VQ-VAE~\cite{oord_neural_2018}     & $X$ &   \textsc{-}    &  $X$  \\ %It has VQ as extra constraint.
-    \midrule  
-    Speech2Vec~\cite{chung_speech2vec_2018}, Audio2Vec~\cite{tagliasacchi_pretraining_2020} - skip-gram    & $X_{[t_1,t_2]}$  &     \textsc{-}   &    $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$     \\
-    \midrule 
-    Speech2Vec~\cite{chung_speech2vec_2018}, Audio2Vec~\cite{tagliasacchi_pretraining_2020} - cbow    & $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$   &     \textsc{-}   &    $X_{[t_1,t_2]}$   \\
-    \midrule 
-    PASE~\cite{pascual_learning_2019}, PASE+~\cite{ravanelli_multitask_2020}\footnote{PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}       & $X$ &  \textsc{-}   &  Different modalities of $X$  \\
-    \midrule 
-    APC~\cite{chung_unsupervised_2019,chung_vectorquantized_2020}         & $X_{[1,t]}$   & \textsc{-}              & $x_{t+c},\, c\geq1$    \\ %VQ-APC is the same as APC
-    \midrule
-    Speech-XLNet \cite{song_speechxlnet_2020}     & \multicolumn{2}{c|}{$X_{\mathcal{P}_{t}}$}   &     $x_{i\sim\mathcal{P}^c_{t}}$  \\ %Lee: Use * to represent the permutation, hope it is not strange.
-    \midrule  
-    DeCoAR~\cite{ling_deep_2020}     & $X_{[1,t-1]}, X_{[t+k+1,T]}$ & \textsc{-} & $X_{[t,t+k]}$   \\
-    \midrule
-    Mockingjay~\cite{liu_mockingjay_2020}, Audio ALBERT~\cite{chi_audio_2020}, DeCoAR 2.0~\cite{ling_decoar_2020}   & \multicolumn{2}{c|}{$X_{-[t,t+k]}$}   & $X_{[t,t+k]}$    \\
-    \midrule 
-    TERA~\cite{liu_tera_2021}, BMR~\cite{wang_unsupervised_2020}  & \multicolumn{2}{c|}{$X_{-[t,t+k]}^{-[f,f+j]}$}       & $X$       \\
-    \midrule
-    pMPC~\cite{yue_phonetically_2021}      &  \multicolumn{2}{c|}{$X_{-[t,t+k^\prime]}$ ($X_{[t,t+k^\prime]}$ is a phoneme)}       & $X_{[t,t+k^\prime]}$    \\
-    \midrule 
-    MPE~\cite{liu_masked_2020} & $X$ &  $Z_{-[t,t+k]}$  & $Z$    \\ %Lee: What is the difference between MPE and NPC??? And I believe it reconstruct the convolutional blocks output (learned target?)
-    \midrule
-    NPC~\cite{liu_nonautoregressive_2020}      & $X$  &   $Z_{-[t,t+k]}$  &    $X$   \\ %Lee: I am not 100% sure it is correct. Please check.
-    \midrule
-    \midrule
-    \multicolumn{4}{c}{\textsc{Contrastive models}} \\
-    \midrule
-    \midrule
-    Unspeech \cite{milde_unspeech_2018}       &   $X_{[t_1,t_2]}$ &   \textsc{-}   &  $X_{[t_0,t_1]}$,$X_{[t_2,t_3]}$ \\
-    \midrule 
-    CPC~\cite{oord_representation_2018}, wav2vec \cite{schneider_wav2vec_2019}, Modified CPC \cite{riviere_unsupervised_2020}         & $X_{[1,t]}$   &    \textsc{-}           & $z_{t+c},\, c\geq1$   \\ %Modified CPC is the same as CPC
-    \midrule 
-    Bidirectional CPC \cite{kawakami_learning_2020}      & $X_{[1,t]}$ or $X_{[t,T]}$ &  \textsc{-}    &    $z_{t+c}$ or $z_{t-c},\, c\geq1$   \\
-    \midrule 
-    vq-wav2vec \cite{baevski_vqwav2vec_2020}     &   $X_{[1,t]}$ &   \textsc{-}    &   $q_{t+c},\, c\geq1$   \\ 
-    \midrule 
-    wav2vec 2.0 \cite{baevski_wav2vec_2020}, wav2vec-C \cite{sadhu_wav2vecc_2021}\footnote{wav2vec-C adds reconstruction loss to wav2vec 2.0.}    & $X$             & $Z_{-[t,t+k]}$          & $Q_{[t,t+k]}$ \\
-    \midrule 
-    w2v-BERT \cite{chung_w2vbert_2021}     &$X$ &    $Z_{-[t,t+k]}$   &     $Q_{[t,t+k]}$ and $C_{[t,t+k]}$     \\
-    \midrule
-    Speech SimCLR \cite{jiang_speech_2021}\footnote{Speech SimCLR targets the latent representation of an augmented version of $X$ using a differently augmented $X$, and vice-versa.}    & \multicolumn{2}{c|}{$\textsc{augment}_1(X)$ and $\textsc{augment}_2(X)$}     &    $\textsc{augment}_2(Z)$ and $\textsc{augment}_1(Z)$   \\
-    \midrule
-    \midrule 
-    \multicolumn{4}{c}{\textsc{Predictive models}} \\
-    \midrule
-    \midrule
-    Discrete BERT~\cite{baevski_vqwav2vec_2020,baevski_effectiveness_2020} \footnote{Discrete BERT obtains codes $C$ from vq-wav2vec.}      &   \multicolumn{2}{c|}{$C_{-[t,t+k]}$}   & $C_{[t,t+k]}$  \\
-    \midrule 
-    HuBERT \cite{hsu_hubert_2021}\footnote{HuBERT is trained first using cluster IDs of the MFCCs as target and subsequently clusters IDs of the model representations from the last iteration.}, WavLM \cite{chen_wavlm_2021}\footnote{WavLM simulates noisy/overlapped speech as inputs.}  & $X$             & $Z_{-[t,t+k]}$          & $C_{[t,t+k]}$  \\
-    \midrule
-    data2vec \cite{baevski_data2vec_2022}    & $X$             & $Z_{-[t,t+k]}$          & $\sum_{l}\bar{H}^{(l)}_{[t,t+k]}$  \\ 
-    \midrule 
-    BEST-RQ \cite{chiu_selfsupervised_2022}\footnote{BEST-RQ obtains codes $C$ by quantizing acoustic features using a random projection quantizer.}     &  \multicolumn{2}{c|}{$X_{-[t,t+k]}$}      &  $C_{[t,t+k]}$   \\
-    \midrule 
-    \bottomrule
-\end{tabular}
+For contrastive models, we specify only positive targets.%
 }
 
-% \end{minipage}
-\label{table:pretext}
-\end{table*}
+\addtocounter{footnote}{1}
+\footnotetext{\label{footnote_review:pase}PASE uses multiple pretext tasks, but the authors find that reconstruction is most important.}
+
+\addtocounter{footnote}{1}
+\footnotetext{\label{footnote_review:wav2vec-c} wav2vec-C adds reconstruction loss to wav2vec 2.0.}
+
+\addtocounter{footnote}{1}
+\footnotetext{\label{footnote_review:discrete-bert}Discrete BERT obtains codes $C$ from vq-wav2vec.}
+
+\addtocounter{footnote}{1}
+\footnotetext{\label{footnote_review:hubert}HuBERT is trained first using cluster IDs of the MFCCs as target and subsequently clusters IDs of the model representations from the last iteration.}
+
+\addtocounter{footnote}{1}
+\footnotetext{\label{footnote_review:wavlm}WavLM simulates noisy/overlapped speech as inputs.}
+
+\addtocounter{footnote}{1}
+\footnotetext{\label{footnote_review:best-rq}BEST-RQ obtains codes $C$ by quantizing acoustic features using a random projection quantizer.}

### 2023-09-29 appendices/supplementary_retrospective.tex
Fix wrong reference key

@@
 Another reason might be that the patient's age and sex are less useful discriminators than other indicative factors.
-This latter hypothesis is supported by our occlusion analysis and \cref{tab_retrospective:table3} where no highly ranked words directly refer to the patient's age or sex.
+This latter hypothesis is supported by our occlusion analysis and \cref{tab_retrospective:table3-occlusion-analysis} where no highly ranked words directly refer to the patient's age or sex.
 

### 2023-10-02 appendices/paper_brief.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

@@
 Most of the seminal work has focused on improving speech recognition \parencite{schneider_wav2vec_2019, baevski_wav2vec_2020}.
-This focus has gained traction over the last couple of years, as computational resources have become more accessible and end-to-end models \parencite{graves_connectionist_2006, chan_listen_2015} have been established as the dominant approach to speech recognition \parencite{gulati_conformer_2020}.
+This focus has gained traction over the last couple of years, as computational resources have become more accessible and end-to-end models \parencite{graves_connectionist_2006, chan_listen_2016} have been established as the dominant approach to speech recognition \parencite{gulati_conformer_2020}.
 It is important to stress that self-supervised models, such as wav2vec 2.0 \parencite{baevski_wav2vec_2020}, represent a breakthrough, and recent successful approaches build upon this method.

### 2023-10-02 chapters/background.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

@@
 This method can also be seen as ameliorating the curse of dimensionality by clustering the high-dimensional feature space.
-Finally, Bayesian neural networks have also been proposed for OOD detection although their performance is not yet competitive with other methods \parencite{henning_are_2021,dangelo_outofdistribution_2022,nguyen_out_2022}.
+Finally, Bayesian neural networks have also been proposed for OOD detection, but their performance is not yet competitive with other methods \parencite{henning_are_2021,dangelo_outofdistribution_2022,nguyen_out_2022}.
 
@@
 The most important instantiation of this reparameterization is the diagonal Gaussian case, where $\zb = \mu_\phib(\xb) + \sigma_\phib(\xb) \odot \epsilonb$ with $\epsilonb\sim\mathcal{N}(0,\mathbf{I})$.
-It is usual to make a mean-field approximation by assuming the covariance matrix is diagonal imposing independence between the latent variables in $\zb$.
+By assuming the covariance matrix is diagonal, this is a mean-field approximation which imposes independence between the latent variables in $\zb$.
 
-Although lower variance than alternatives, such as score function estimators, a number of works have attempted to reduce the variance. 
+Although the path-wise derivative has lower variance than most alternatives, such as score function estimators, a number of works have attempted to reduce its variance.
 \textcite{roeder_sticking_2017} note that the path-wise derivative can be factored into a path-wise estimator and a score function estimator and propose to ignore the score function factor to reduce gradient variance.
-\textcite{rainforth_tighter_2019} that this problem is even larger for the importance weighted ELBO \parencite{burda_importance_2016} leading to diminishing returns for increasing numbers of samples.
-Instead of removing the score function factor in the gradient, \textcite{tucker_doubly_2019} propose to reparameterize the score function and derive low variance gradient estimators also for importance weighted ELBO.
+\textcite{rainforth_tighter_2019} show that this problem is even larger for the importance weighted ELBO \parencite{burda_importance_2016} and leads to diminishing returns when increasing the number of samples.
+Rather than removing the score function factor in the estimate, \textcite{tucker_doubly_2019} propose to reparameterize it and derive low variance gradient estimators, also for importance weighted ELBO.
+
+\paragraph{Hierarchical models}\label{sec:hierarchical-vae-background}
+The usual mean-field assumption imposed by letting $\zb \sim\mathcal{N}(\mub,\mathbf{I}\sigmab^2)$ can often be more restrictive for model expressiveness than we would like. 
+While learning a full covariance is a simple solution, it is not always sufficiently computationally efficient and, in any case, enables learning only linear covariance between elements of $\zb$. 
+Along with the usual motivations of hierarchical models such as efficient representation, feature disentanglement, and data efficiency, this has led to research into using a hierarchy of several conditionally dependent latent variables.
+This idea is commonly formalized by introducing a set of $L$ latent variables  $\zb=\zb_1, \dots, \zb_L$ and letting the generative model be defined as,
+%
+\begin{equation}
+    p_\thetab(\xb|\zb)=p(\xb|\zb_1) p_\thetab(\zb_1|\zb_2) \cdots p_\thetab(\zb_{L-1}|\zb_L) p(\zb_L) \enspace .
+\end{equation}
+%
+This top-down generative model can be efficiently sampled via ancestral sampling; first $\zb_L$ is drawn from the prior $p(\zb_L)$, and then each $\zb_l$ is drawn from the corresponding $p(\zb_l|\zb_{l+1})$ until we can draw $p(\xb|\zb_1)$.
+
+The inference model can then be defined in two ways respectively referred to as \emph{bottom-up} \parencite{burda_importance_2016}
+\begin{equation}
+    %q_\phib(\zb|\xb)=q_\phib(\zb_1|\xb)q_\phib(\zb_2|\zb_1)\cdots q_\phib(\zb_L|\zb_{L-1})
+    q_\phib(\zb|\xb) = q_\phib(\zb_1|\xb)\textstyle\prod_{i=2}^{L} q_\phib(\zb_i|\zb_{i-1})
+\end{equation}
+and \emph{top-down} \parencite{sonderby_ladder_2016}
+\begin{equation}
+    % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb_{L-1}|\zb_L)\cdots q_\phib(\zb_2|\zb_1),
+    q_\phib(\zb|\xb) = q_\phib(\zb_L|\xb)\textstyle\prod_{i=L-1}^{1} q_\phib(\zb_{i}|\zb_{i+1}) \ .
+\end{equation}
+Regardless of the choice of inference model, the resulting hierarchical VAE is trained using the ELBO \cref{eq:variational-autoencoder-elbo} and the reparameterization trick.
+
+% Until recently, hierarchical VAEs gave inferior likelihoods compared to state-of-the-art autoregressive \parencite{ho_flow_2019} and flow-based models \parencite{salimans_pixelcnn_2017}.
+% This was changed by \textcite{maaloe_biva_2019}, \textcite{vahdat_nvae_2020}, and \textcite{child_very_2021}, which introduced complementary methods to extend the number of latent variables to a very deep hierarchy resulting in state-of-the-art likelihood performance.
 
@@
 Contrary to the KL-divergence term of \cref{eq:variational-autoencoder-elbo-factorized}, driving the marginal KL-divergence to zero does not enforce independence between $\xb$ and $\zb$.
-However, this alternative form reveals that the ELBO is maximized by minizing the mutual information between $\xb$ and $\zb$, $I_{q_{}(\xb,\zb)}\left[\xb ; \zb \right]$. 
+However, this alternative form reveals that the ELBO is maximized by minimizing the mutual information between $\xb$ and $\zb$, $I_{q_{}(\xb,\zb)}\left[\xb ; \zb \right]$. 
 This result indicates that the usual interpretation of $\zb$ as a representation of $\xb$ is not a fundamental property arising from the ELBO and challenges the usefulness of variational autoencoders as representation learners.
-
-% \paragraph{Hierarchical models}
-% VAEs can be extended to hierarchical models by introducing additional latent variables $\zb_1,\dots,\zb_L$ that are conditionally dependent on each other \parencite{ranganath_hierarchical_2016,sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021,maaloe_biva_2019}.
-The generative model can then be written as,
-% %
-% \begin{equation}
-%     p_\thetab(\xb,\zb_1,\dots,\zb_L) = p_\thetab(\xb|\zb_1)\prod_{l=2}^L p_\thetab(\zb_{l-1}|\zb_{l}) \enspace .
-% \end{equation}
 % %

### 2023-10-02 chapters/discussion.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

@@
 % ~5 pages
+%
+% OUTLINE:
+% - paper-hierarchical:
+%   - Suitability of VAEs for representation learning (minimization of mutual information and sensitivity to implicit prior such as architecture)
+% - paper-benchmarking
+%   - Inferiority of probabilistic methods compared to self-supervised learning. 
+% -
 
@@
 
+\section{\Cref{chp:paper-benchmarking} revisited: Bested?}
+
+\lesstodo[inline]{Discuss whether variational autoencoders are viable for learning good representations for speech for downstream tasks when alternatives such as SSL method exist.}
+\lesstodo[inline]{Out-of-distribution detection on speech?}
+
 \section{\Cref{chp:paper-review} revisited: \dots} \label{sec:discussion-paper-review}
@@
 \parencite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}
-\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}
 % Since the main focus within self-supervised learning has been on improving downstream task performance, very limited work, if any, has investigated self-supervised representations in terms of uncertainty estimation.
@@
 \section{\Cref{chp:paper-retrospective} revisited: Uncertainty}
-The work in \textcite{wenstrup_retrospective_2023} focuses on predictive performance and an analysis of feature importance but does not explicitly consider uncertainty estimation. To
+The work in \textcite{wenstrup_retrospective_2023} focuses on predictive performance and an analysis of feature importance but does not explicitly consider uncertainty estimation.
 
@@
 \lesstodo[inline]{Report the results of calibrating the stroke model using e.g. Platt scaling or Isotonic scaling.} % https://scikit-learn.org/stable/modules/calibration.html
-\lesstodo[inline]{Maybe mention MultiQT paper \parencite{havtorn_multiqt_2020}.}
+
+% \lesstodo[inline]{Maybe mention MultiQT paper \parencite{havtorn_multiqt_2020}.}

### 2023-10-02 chapters/introduction.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

@@
 
-\section{Thesis outline and research questions}
+\section{Thesis scope and outline}
 
@@
 The remainder of \cref{part:background} consists of two chapters. 
-\Cref{chp:technical-background} provides relevant technical background to the papers of the thesis, including uncertainty, out-of-distribution detection, and variational autoencoders while \cref{chp:main-contributions} outlines the main contributions of the thesis via the published papers in relation to the research hypotheses.
+\Cref{chp:main-contributions} outlines the research questions of the thesis via the published papers and presents their main contributions. 
+\Cref{chp:technical-background} provides relevant technical background to the papers of the thesis, including uncertainty, out-of-distribution detection, and variational autoencoders.
 
@@
 \Cref{part:unsupervised-uncertainty-estimation} contains papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
-Here we examine whether it is possible to devise a method that utilizes a hierarhy of latent variables in
-
 \Cref{part:self-supervised-speech-representation-learning} contains a review paper on self-supervised speech representation learning \parencite{mohamed_selfsupervised_2022}.
@@
 
-\section{Thesis scope and outline}
+% \section{Thesis scope and outline}
 %

### 2023-10-02 chapters/main-contributions.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

@@
 
-\chapter[main contributions]{Main contributions}\label{chp:main-contributions}
+\chapter[research questions and main contributions]{Research questions and main contributions}\label{chp:main-contributions}
 \newcommand{\contribitem}[1]{{\vspace{1em}\noindent\scshape\bfseries{\color{dtured}\cref{#1}}\\\ucnameref{#1}\\}}
 
-The following chapters of the thesis are self-contained studies and therefore detail their own contributions. 
-They are however written without consideration to the other chapters. 
-For that reason, this chapter will describe the main contributions of each chapter in relation to the overall research topic of the thesis. 
-The chapters are structured into three parts: \cref{part:unsupervised-uncertainty-estimation} is concerned with unsupervised uncertainty estimation, \cref{part:self-supervised-speech-representation-learning} is concerned with self-supervised speech representation learning, and \cref{part:medical-applications} contains studies on machine learning methods applied for tasks in a medical setting.
-Finally, \cref{part:discussion-and-conclusion} contains the discussion and conclusions of the thesis.
+The following chapters of the thesis are self-contained studies and therefore detail their own research questions and contributions, but do so without consideration to the other chapters. 
+For that reason, we here lay out the research questions and main contributions of each chapter in relation to the overall research project of the thesis. 
+The chapters are structured into three parts: {\color{dtured}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{dtured}\cref{part:self-supervised-speech-representation-learning}} is concerned with self-supervised speech representation learning, and {\color{dtured}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.
+Finally, {\color{dtured}\cref{part:discussion-and-conclusion}} contains a discussion and the conclusion of the thesis.
 
@@
 \Cref{chp:introduction} provides a general introduction to the thesis and gives motivating examples for speech recognition and assistance in medical encounters including stroke recognition on medical helplines and automation of medical coding.
-\Cref{chp:technical-background} provides additional technical background not included in the individual studies by introducing uncertainty as a concept in the context of information and probability theory, introducing the task of out-of-distribution detection and reviewing existing work on the problem, and providing an introduction to variational autoencoders.
+\Cref{chp:technical-background} provides additional technical background not included in the individual studies.
+It introduces uncertainty as a concept in the context of information and probability theory, introduces the task of out-of-distribution detection and provides a review of existing work on the problem, and lays out the foundations for variational autoencoders.
 
@@
 This part of the thesis is concerned with unsupervised uncertainty estimation and consists of three papers. 
-The first two papers are approach unsupervised uncertainty estimation through the task of out-of-distribution detection using generative models.
-The first paper proposes a new method for out-of-distribution detection using hierarchical variational autoencoders.
-The second paper builds on the first and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable generative model with explicit likelihood. 
-The third paper proposes a new hierarchical variational autoencoder for speech and benchmarks it against other variational autoencoders and autoregressive models.
+The first two papers approach unsupervised uncertainty estimation through the task of out-of-distribution detection using generative models. 
+The first paper proposes a new method for out-of-distribution detection using hierarchical variational autoencoders and the second proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable generative model with explicit likelihood.
+The third paper proposes a new hierarchical variational autoencoder for speech and benchmarks it against other variational and autoregressive models.
 
 \contribitem{chp:paper-hierarchical}
-This paper proposes a new method for unsupervised out-of-distribution detection using hierarchical variational autoencoders. 
-The paper provides evidence that the failure of deep generative models to assign lower likelihoods to data from outside the training distribution is due to overemphasis on low-level features that generalize between distributions.
-To solve this, the proposed method computes a likelihood-ratio score for out-of-distribution detection that requires data to be in-distribution across all feature-levels. 
+In this work we hypothesize that the likelihood estimate of variational autoencoders is a poor score for out-of-distribution due to on overemphasis on low-level features that generalize between distributions.
+We further hypothesize that a well-formed hierarchy of latent variables provides a tool that can be used to select between which features to emphasize for out-of-distribution detection and hence a way to improve the performance of variational autoencoders on this task.
+We proceed to provide empirical and theoretical evidence that low-level features indeed do dominate the likelihood and propose a new method for out-of-distribution detection using hierarchical variational autoencoders.
+The proposed method computes a likelihood-ratio score for out-of-distribution detection that requires data to be in-distribution across all feature-levels. 
 The proposed method is computationally efficient, fully unsupervised, and performs well on several out-of-distribution detection benchmarks. 
@@
 \contribitem{chp:paper-modelagnostic}
-This paper proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable, explicit likelihood generative model. 
-The method is based on combining a classical parametric test with the recently introduced typicality test via Fisher's method. 
+In this follow-up work, we note that many diverse methods have been proposed for out-of-distribution detection using different generative models and seek to develop a model-agnostic approach. 
+We hypothesize that by phrasing the problem as a statistical testing problem and combining different tests, the method's efficacy can be improved and weaknesses inherent to any particular test can be alleviated.
+From this hypothesis, we develop a method that combines a classical parametric test with the recently introduced typicality test via Fisher's method and can be used with any differentiable, explicit likelihood generative model.
 We show that this leads to a more accurate out-of-distribution test. 
-We also discuss the benefits of casting out-of-distribution detection as a statistical testing problem, for instance enabling false positive rate control which is valuable for practical applications, especially in high-risk settings such as medical decision-making.
+Finally, we discuss the benefits of casting out-of-distribution detection as a statistical testing problem, for instance enabling false positive rate control.
+This is valuable for practical applications, especially in high-risk settings such as medical decision-making.
 
 \contribitem{chp:paper-benchmarking}
-This paper proposes a new hierarchical variational autoencoder for speech inspired by the Clockwork VAE \parencite{saxena_clockwork_2021}. 
-The model is benchmarked against other variational autoencoders and autoregressive models for speech.
-We show that using a hierarchy of latent variables improves the likelihood of the model and that the learned latent space captures information relevant for speech recognition such as phonetic content.
+In this paper, we focus on benchmarking hierarchical variational autoencoders for speech modelling. 
+We hypothesize that the improvements in likelihood scores proven by hierarchical models in other domains, most prominently images, can also carry over to speech modelling. 
+To this end, we benchmark a number of existing deterministic and stochastic models and adapt the hierarchical Clockwork VAE developed for video \parencite{saxena_clockwork_2021} to speech data. 
+We show that using a hierarchy of latent variables can improve the likelihood of the model and that the learned latent space captures information relevant for speech recognition, such as phonetic content.
+% This paper proposes a new hierarchical variational autoencoder for speech inspired by the Clockwork VAE \parencite{saxena_clockwork_2021}. 
+% The model is benchmarked against other variational autoencoders and autoregressive models for speech.
 
@@
 As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self-supervised representations generally outperform representations learned via probabilistic generative models such as those investigated in \cref{part:unsupervised-uncertainty-estimation}.
-In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self-supervised learning as an alternative paradigm to generative models for learning representations for unsupervised uncertainty estimation.
+In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self-supervised learning for learning representations for unsupervised uncertainty estimation.
 We shall return to this in the discussion of \cref{chp:paper-review} in \cref{sec:discussion-paper-review}.
+An early version of this review paper included a review of variational autoencoders for speech representation learning and a comparison between the two modelling paradigms \parencite{borgholt_brief_2022}.
+Due to a desire to focus on self-supervised methods, the paper in \cref{chp:paper-review} does not put particular focus on variational methods. 
+The early version is included in \cref{app:paper-brief} for reference.
 
@@
 
-An early version of this review paper \parencite{borgholt_brief_2022} included a review of generative latent variable models for speech representation learning and a comparison between the two modelling paradigms.
-Due to the superior performance of self-supervised methods and a desire to focus on self-supervised methods, the paper in \cref{chp:paper-review} does not focus on generative latent variable models. 
-The early version is included in \cref{app:paper-brief} for reference.
-
 \contribitem{part:medical-applications}
 This part contains studies on machine learning methods applied for tasks in a medical setting.
-While uncertainty estimation is not a central theme to any of the two papers, the first paper performs a substantial evaluation of the explainability of the proposed model via an occlusion analysis on the text input.
+While uncertainty estimation is not a central theme to the two papers, the first paper performs a substantial evaluation of the explainability of the proposed model via an occlusion analysis on the text input.
 Later, in the discussion, we shall further consider uncertainty estimation in relation to these two papers.
@@
 \contribitem{chp:paper-retrospective}
+In this paper we examine the hypothesis that a machine learning framework can learn to recognize cases of stroke in calls made to a prehospital medical helpline. 
+We used calls from Copenhagen during 2015 to 2020, to develop a machine learning-based classification pipeline.
+First, calls were transcribed by a speech recognition model and then categorized as stroke or non-stroke using a text classification model.
+On test data from 2021, call-takers achieved an overall sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%) while the machine learning framework performed significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
 Effective stroke treatment depends on recognition by call-takers at prehospital telehealth services. 
-This paper aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.
-We used calls from Copenhagen during 2015 to 2020, to develop a machine learning-based classification pipeline.
-First, calls were transcribed by a speech recognition model and then categorised as stroke or non-stroke using a text classification model.
-On test data from 2021, call-takers achieved an overall sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%) while the machine learning framework performed significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
 This study provides preliminary evidence that a machine learning framework for recognizing stroke in prehospital medical helpline calls could become a supportive tool for call-takers, aiding in early and accurate stroke recognition.
+% This paper aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.
 
 \contribitem{chp:paper-automated}
-Improving the efficiency of medical coding can alleviate the administrative burden on medical staff and help reduce the rising costs for hospitals. 
 In this paper, we review the current state-of-the-art in automated medical coding of clinical notes. 
-We reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is often attributable to weak configuration, crudely sampled train-test splits, and insufficient evaluation.
-We compare models on data splits created with stratified sampling, use identical experimental setups and tune hyperparameters and decision boundaries. 
-By analyzing prediction errors, we confirm that all models struggle with rare codes, while long documents only have a negligible impact. 
+We hypothesize that several previous works underperform for reasons more related to suboptimal hyperparameter tuning, incorrect evaluation, and crude data handling than to model design, and that performance can be improved by addressing these issues.
+With our work, we aim to increase reproducibility and comparability of future work and to provide a baseline for future comparisons.
+We reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is indeed attributable to weak configuration of the training and crudely sampled train-test splits with many extremely rare classes and several without examples in the training data at all.
+We also find a widespread error in the calculation of the macro F1 score.
+We propose new data splits created with stratified sampling and use identical experimental setups to compare models, tuning hyperparameters and decision boundaries. 
+By analyzing prediction errors, we confirm the observation of previous work that all models struggle with rare codes, although long documents only have a negligible impact on performance, contrary to previous claims.
 We also present the first comprehensive results on the newly released MIMIC-IV dataset using the reproduced models.
-This paper aims to increase reproducibility and comparability of future work in automated medical coding and to provide a baseline for future research.

### 2023-10-02 chapters/paper_automated.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

@@
 
-\begin{figure}[t]
+\begin{figure}
     \centering
-    \includegraphics[width=0.9\linewidth]{paper_automated/mimiciv_code_freq.png}
+    \includegraphics[width=0.9\textwidth]{paper_automated/mimiciv_code_freq.png}
     \caption[The frequency of ICD-9 and ICD-10 codes in MIMIC-IV before pre-processing.]{The frequency of ICD-9 and ICD-10 codes in MIMIC-IV before pre-processing.
@@
 
-\begin{sidewaystable}[t]
+\begin{sidewaystable}
     \centering
@@
 
-\begin{table}[t]
+\begin{table}
     \centering
@@
     \label{tab:model_facts}
-    \begin{tabular}{lcccc}
+    \begin{tabular}{llllr}
         \toprule
@@
 
-\begin{figure}[t]
+\begin{figure}
     \centering
-    \begin{subfigure}[b]{0.6\textheight}
+    \begin{subfigure}[b]{0.9\textwidth}
         \centering
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.6\textheight}
+    \begin{subfigure}[b]{0.9\textwidth}
         \centering
@@
 
-\begin{figure}[t]
+\begin{figure}
     \centering
-    \begin{subfigure}[b]{0.6\textheight}
+    \begin{subfigure}[b]{0.95\textwidth}
         \centering
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.6\textheight}
+    \begin{subfigure}[b]{0.95\textwidth}
         \centering
@@
 
-\begin{figure}[t]
+\begin{figure}
     \centering
-    \includegraphics[width=0.8\linewidth]{paper_automated/f1_micro_vs_num_words.png}
+    \includegraphics[width=0.9\textwidth]{paper_automated/f1_micro_vs_num_words.png}
     \caption[Relationship between the lengths of the clinical notes and the micro F1 score for each model on MIMIC-IV \textit{ICD-9}]{Relationship between the lengths of the clinical notes and the micro F1 score for each model on MIMIC-IV \textit{ICD-9}.
@@
 
-\begin{figure}[t]
+\begin{figure}
     \centering
-    \includegraphics[width=0.8\linewidth]{paper_automated/code_count_vs_f1_all_datasets.png}
+    \includegraphics[width=0.9\textwidth]{paper_automated/code_count_vs_f1_all_datasets.png}
     \caption[Relationship between the code frequencies in the training set and the macro F1 score for PLM-ICD on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}]{Relationship between the code frequencies in the training set and the macro F1 score for PLM-ICD on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}.
@@
 
- \begin{table}[t]
+ \begin{table}
     \centering
@@
 
-\begin{figure}[t]
+\begin{figure}
     \centering
-    \includegraphics[width=0.8\linewidth]{paper_automated/chapters.png}
+    \includegraphics[width=0.9\textwidth]{paper_automated/chapters.png}
     \caption[Performance of PLM-ICD on ICD-10 chapters]{Performance of PLM-ICD on ICD-10 chapters.

### 2023-10-02 chapters/paper_hierarchical.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

@@
 This seemingly paradoxical behavior has caused concerns over the quality of the attained density estimates.
-In the context of hierarchical variational autoencoders, we provide evidence to explain this behavior by out-of-distribution data having in-distribution low-level features.
+In the context of hierarchical variational autoencoders, we provide evidence to explain this behavior by out-of-distribution data having in-distribution, low-level features.
 We argue that this is both expected and desirable behavior.
-With this insight in hand, we develop a fast, scalable and fully unsupervised likelihood-ratio score for OOD detection that requires data to be in-distribution across all feature-levels.
+With this insight in hand, we develop a fast, scalable, and fully unsupervised likelihood-ratio score for OOD detection that requires data to be in-distribution across all feature-levels.
 We benchmark the method on a vast set of data and model combinations and achieve state-of-the-art results on out-of-distribution detection.
@@
 Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb_1)] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ from \cref{eq_hierarchical:elbo} relies on the KL-term for OOD detection.
-For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ can be expressed by a hierarchical sum% over the hierarchy
+For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel p(\zb))$ can be expressed by a hierarchical sum,% over the hierarchy
 % \begin{multline}

### 2023-10-02 chapters/paper_review/04_benchmarking.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

@@
 Switchboard (SWB)~\parencite{godfrey_switchboard_1992}, 
-TIMIT~\parencite{garofolo_timit_1993}, and VoxLingua107~\parencite{valk_voxlingua107_2021}, are
+TIMIT~\parencite{garofolo_timit_1993}, and VoxLingua107 \parencite{valk_voxlingua107_2021}, are
 labeled and conventionally used for evaluation, while their audio streams are
@@
 \addtocounter{footnote}{1}
-\footnotetext{\label{footnote:table:datasets-abbrevations}We use the following abbreviations:
+\footnotetext{\label{footnote:table:datasets-abbrevations}
+In \cref{table:datasets}, we use the following abbreviations:
 \textbf{EN}: English; 
@@
 
-\begin{table*}[t]
+\begin{table*}
   \centering
@@
   evaluations (Part 1).]{A summary of common experiment settings for various SSL
-  evaluations (Part 1). Networks are usually pre-trained with SSL
+  evaluations (Part 1). 
+  Networks are usually pre-trained with SSL
   techniques, augmented with prediction heads, and fine-tuned (or trained) with
-  labeled data in downstream tasks for benchmarking. The 
-  \textbf{Pre-training corpus}, \textbf{Training (fine-tuning)}, and
-  \textbf{Test} columns list the datasets used in each work, and the \textbf{Task}
-  column lists the tasks performed in the corresponding papers. We follow
-  the abbreviation introduced in \cref{table:datasets}. The \textbf{Transfer}
-  column indicates whether the SSL technique is evaluated by its capability for
-  transfer learning, i.e., different datasets are utilized for pre-training and
-  fine-tuning. The \textbf{Fine-tuning labels used} column summarizes the
-  amount of labeled examples used in downstream fine-tuning.}
+  labeled data in downstream tasks for benchmarking. 
+  We follow the abbreviations introduced in \cref{table:datasets}.
+  Column descriptions are in \cref{footnote:table:method_data_setting}.
+  }
   \label{table:method_data_setting}
-  \resizebox*{0.97\textwidth}{!}{%
+  \resizebox*{0.99\textwidth}{!}{%
   {%\tabulinesep=1mm
   \renewcommand*\arraystretch{1.4}
-  \begin{tabular}{p{0.12\textwidth}p{0.12\textwidth}p{0.06\textwidth}p{0.13\textwidth}p{0.13\textwidth}p{0.06\textwidth}p{0.20\textwidth}}
+  \begin{tabular}{p{0.14\textwidth}p{0.12\textwidth}p{0.06\textwidth}p{0.15\textwidth}p{0.10\textwidth}p{0.05\textwidth}p{0.15\textwidth}}
     \toprule
     \multirow{2}{*}{Work} & \multirow{2}{*}{\makecell[l]{Pre-training\\ corpus}} & \multirow{2}{*}{Task} & \multicolumn{2}{c}{Dataset} & \multirow{2}{*}{Transfer} & \multirow{2}{*}{Fine-tuning labels used} \\ \cline{4-5}
-     &  &  & Training (fine-tuning) & Test & & \\
+     &  &  & Fine-tuning & Test & & \\
     \midrule
@@
 
-\begin{table*}[t]
+\addtocounter{footnote}{1}
+\footnotetext{\label{footnote:table:method_data_setting}
+  The \textbf{Pre-training corpus}, \textbf{Fine-tuning}, and
+  \textbf{Test} columns list the datasets used in each work, and the \textbf{Task}
+  column lists the tasks performed in the corresponding papers. 
+  The \textbf{Transfer} column indicates whether the SSL technique is evaluated by its capability for
+  transfer learning, i.e., different datasets are utilized for pre-training and
+  fine-tuning. The \textbf{Fine-tuning labels used} column summarizes the
+  amount of labeled examples used in downstream fine-tuning.
+}
+
+\begin{table*}
   \centering
@@
   \label{table:method_data_setting_2}
-  \resizebox*{0.97\textwidth}{!}{%
+  \resizebox*{0.99\textwidth}{!}{%
   {%\tabulinesep=1mm
   \renewcommand*\arraystretch{1.4}
-  \begin{tabular}{p{0.12\textwidth}p{0.12\textwidth}p{0.05\textwidth}p{0.18\textwidth}p{0.18\textwidth}p{0.05\textwidth}p{0.15\textwidth}}
+  \begin{tabular}{p{0.14\textwidth}p{0.12\textwidth}p{0.06\textwidth}p{0.15\textwidth}p{0.10\textwidth}p{0.05\textwidth}p{0.15\textwidth}}
     \toprule
     \multirow{2}{*}{Work} & \multirow{2}{*}{\makecell[l]{Pre-training\\ corpus}} & \multirow{2}{*}{Task} & \multicolumn{2}{c}{Dataset} & \multirow{2}{*}{Transfer} & \multirow{2}{*}{Fine-tuning labels used} \\ \cline{4-5}
-     &  &  & Training (fine-tuning) & Test & & \\
+     &  &  & Fine-tuning & Test & & \\
     \midrule
@@
     WavLM \parencite{chen_wavlm_2021} & \makecell[l]{LL 60k hrs\\\;\;+ GigaSpeech-10k\\\;\;+ VP-24k} & Multi & SUPERB & SUPERB & \checkmark & See \parencite{yang_superb_2021} for details \\ \hline
-    \multirow{4}{*}[0mm]{XLS-R \parencite{babu_xlsr_2021}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;+ CV-dataset-7k\\\;\;+ VL\\\;\;+ BBL}} & ASR & VP, MLS, CV-dataset, BBL, LS & VP, MLS, CV-dataset, BBL, LS & - & \multirow{4}{*}[0mm]{See \parencite{babu_xlsr_2021} for details} \\ \cline{3-6}
+    \multirow{4}{*}[0mm]{XLS-R \parencite{babu_xlsr_2021}} & \multirow{4}{*}[0mm]{\makecell[l]{VP-400k + MLS\\\;\;+ CV-dataset-7k\\\;\;+ VL\\\;\;+ BBL}} & ASR & \makecell[l]{VP, MLS,\\CV-dataset,\\BBL, LS} & \makecell[l]{VP, MLS,\\CV-dataset,\\BBL, LS} & - & \multirow{4}{*}[0mm]{See \parencite{babu_xlsr_2021} for details} \\ \cline{3-6}
     & & SID & VoxCeleb1 & VoxCeleb1 & \checkmark & \\ \cline{3-6}

### 2023-10-02 chapters/paper_review/05_zero.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

@@
 
-\begin{sidewaystable}[t]
+\begin{sidewaystable}
 \centering
@@
 emotion, hesitation, interruptions. 
-The generative spoken language model (GSLM)~\parencite{nguyen_generative_2022} utilizes discrete
+The generative spoken language model (GSLM) \parencite{nguyen_generative_2022} utilizes discrete
 representations from wav2vec~2.0, HuBERT, and CPC algorithms as inputs to an

### 2023-10-02 chapters/background.tex
Add figures

@@
 \end{equation}
+A variant of the variational autoencoder that employs a bidirectional inference network has also been proposed \parencite{maaloe_biva_2019}.
 Regardless of the choice of inference model, the resulting hierarchical VAE is trained using the ELBO \cref{eq:variational-autoencoder-elbo} and the reparameterization trick.
+Recent architectural advances have alleviated the posterior collapse problem and made it possible to train VAEs with many latent variables.\parencite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
 
@@
 VAEs have also proven themselves useful for semi-supervised learning \parencite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}.
-Recent architectural advances have made it possible to train VAEs with many latent variables \parencite{sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021} and have alleviated the posterior collapse problem \parencite{maaloe_biva_2019}.
 

### 2023-10-02 chapters/background.tex
Work on discussion

@@
 An alternative approach to representation learning is provided by self-supervised learning \parencite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020,schneider_wav2vec_2019} which is a form of unsupervised learning where the training objective is derived from the data itself.
-While it is fair to say that not all self-supervised training objectives are as well-motivated as the ELBO is for VAEs, they have shown impressive results within the field of natural language processing \parencite{devlin_bert_2018,chen_simple_2020}, speech processing \parencite{schneider_wav2vec_2019}, and computer vision \parencite{chen_simple_2020}.
+While it is fair to say that not all self-supervised training objectives are as well-motivated as the ELBO is for VAEs, they have shown impressive results within the fields of natural language processing \parencite{devlin_bert_2018,chen_simple_2020}, speech processing \parencite{schneider_wav2vec_2019}, and computer vision \parencite{chen_simple_2020}.
 For a deeper comparison of VAEs and self-supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{app:paper-brief}. 

### 2023-10-02 chapters/discussion.tex
Work on discussion

@@
 \section{\Cref{chp:paper-retrospective} revisited: Uncertainty}
-The work in \textcite{wenstrup_retrospective_2023} focuses on predictive performance and an analysis of feature importance but does not explicitly consider uncertainty estimation.
+The work in \textcite{wenstrup_retrospective_2023} focuses on the predictive performance of the ensemble model and an analysis of feature importance, but does not explicitly consider uncertainty estimation.
+As we discussed in \cref{subsec:model-calibration}, such a model can be calibrated to predict probabilities that are aligned with the empirical probability of the model being correct on some validation set.
+
+In \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}, we plot the calibration curve for the uncalibrated stroke recognition ensemble and the individual models along with a histogram of its predicted probabilities.
+The calibration curve is computed by sorting the predicted probabilities on test set into a number of bins spanning the range from zero to one.
+For each bin, we compute the fraction of examples for which the model predicted correctly and then plot these pairs of mean predicted probabilities and correctness fractions.
+A perfectly calibrated model would have the same fraction of correct predictions in any given bin, as that bin's mean value.
+\Cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model} clearly shows the miscalibration issue that we previously discussed.
+It is interesting to note that the ensemble model is better calibrated than the individual models that form it.
+Since an ensemble output probability is computed as the harmonic mean of the five individual model probabilities, it is easy to see that the ensembles output probability can never exceed the maximum probability predicted among the individual models.
+This property tends to make ensemble probabilities less extreme and, since the individual models are overconfident, this results in better calibration (see also the histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).
+
+To calibrate the ensemble model, we can use methods such as Platt-scaling \parencite{platt_probabilistic_1999} or isotonic regression \parencite{zadrozny_transforming_2002}. 
+In both cases, we fit a simple regression model (logistic or isotonic) to the predicted probabilities and the target labels on some validation set. 
+In \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic} we have done so for the ensemble model using our validation set for fitting and visualizing the results on the test set.
+On the left, we plot the resulting calibration curves and on the right we show the logistic and isotonic fits.
+We see that both methods result in quite good calibrations\footnote{Brier scores: Uncalibrated = $0.003500$, logistic = $0.001807$, isotonic = $0.001774$.
+Brier skill score compared to uncalibrated (relative improvement): Logistic = $0.4830$, isotonic = $0.4924$} and that the predicted probabilities are shifted towards smaller values.
+
+However, while correct calibration provides some interpretability to the predictions, the resulting probabilities are still functions of the model's performance.
+For example, with the logistic calibration, we can note that the predicted probabilities have been shifted so far towards smaller values that the best the model can do, is a predicted probability of about 50\%.
+This means strokes can be predicted at best with 50\% probability of being correct.
+If this model were to be deployed in a randomized controlled trial, the calibrated probabilities might be useful to incorporate in the design of the user interface such that a user can discern between confident and uncertain model predictions.
+Nonetheless, it is easy to imagine how this generally low confidence might lead the user to ignore the model.
+
+% Brier scores:  {'uncalibrated': 0.0034955788687128925, 'logistic': 0.0018072483306197486, 'isotonic': 0.0017744177396100578, 'average': 0.0021955477869598783}
+% BSS uncalibrated reference:  {'logistic': 0.48299025755205005, 'isotonic': 0.4923822902432705}
+% BSS mean reference:  {'logistic': 0.17685766561145932, 'isotonic': 0.1918109229282361}
 
@@
     \end{subfigure}
-    \caption[Calibration curve for the uncalibrated stroke recognition model and empirical distribution of predicted probabilities.]{Calibration curve for the uncalibrated ensemble model with median F1-score also reported in \cref{tab_retrospective:table3-occlusion-analysis, fig_retrospective:figure1-roc-curve} (left) and the histogram of predicted probabilities (right).}
+    \caption[Calibration curve for the uncalibrated stroke recognition model and empirical distribution of predicted probabilities.]{%
+        Calibration curve for the uncalibrated stroke recognition ensemble model (left) and the histogram of predicted probabilities (right). 
+        We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis, fig_retrospective:figure1-roc-curve}.
+    }
+    \label{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}
+\end{figure}
+
+\begin{figure}
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{python_plotting/calibration_curves_ensemble.pdf}
+        % \caption{}
+        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{python_plotting/calibration_fits_ensemble.pdf}
+        % \caption{}
+        % \label{fig_discussion:histogram_ensemble_and_single_model}
+    \end{subfigure}
+    \caption[Calibration fits and curves for the stroke recognition model using Platt-scaling and isotonic regression for calibration.]{%
+        Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right). 
+        We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis, fig_retrospective:figure1-roc-curve}.
+    }
+    \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
 \end{figure}

### 2023-10-02 chapters/introduction.tex
Work on discussion

@@
 The evolution of technology to support and automate decision-making can be traced back to the development of algorithms and mathematical techniques by ancient civilizations. 
-The Egyptian Rhind Mathematical Papyrus, dated around 1550 BCE, describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \parencite{georges_universal_2001}.
+The Egyptian Rhind Mathematical Papyrus, dated to around 1550 BCE, describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \parencite{georges_universal_2001}.
 Around the same time, the Babylonians had developed a sophisticated system of mathematics that included algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \parencite{fowler_square_1998}.
@@
 % 
-The problem of miscalibration in machine learning models is well-known and has been studied for decades \parencite{lewis_sequential_1995, platt_probabilistic_1999, garczarek_classification_2002, zadrozny_transforming_2002, bennett_using_2003}.
+The problem of miscalibration in machine learning models is well-known and has been studied for decades \parencite{lewis_sequential_1995, platt_probabilistic_1999, garczarek_classification_2002, zadrozny_transforming_2002, bennett_using_2003, niculescu-mizil_predicting_2005}.
 One of the best known methods for calibrating a binary classifier is Platt scaling which fits a logistic regression model to the model outputs, assuming that the miscalibration can be corrected by a logarithmic function \parencite{platt_probabilistic_1999}.
@@
 
-\todo[inline]{Formulate research hypotheses}
-
 This introduction provided a high-level overview of the motivation behind the research project by focusing on the use-cases of stroke recognition and automated medical coding as well as speech processing and the importance of uncertainty quantification.
 The remainder of \cref{part:background} consists of two chapters. 
-\Cref{chp:main-contributions} outlines the research questions of the thesis via the published papers and presents their main contributions. 
-\Cref{chp:technical-background} provides relevant technical background to the papers of the thesis, including uncertainty, out-of-distribution detection, and variational autoencoders.
+\Cref{chp:main-contributions} presents the research questions of the thesis and the main contributions made in the included papers. 
+\Cref{chp:technical-background} provides relevant technical background to the included papers, including uncertainty, out-of-distribution detection, and variational autoencoders.
 
 \Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. 
-\Cref{part:unsupervised-uncertainty-estimation} contains papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
-\Cref{part:self-supervised-speech-representation-learning} contains a review paper on self-supervised speech representation learning \parencite{mohamed_selfsupervised_2022}.
-\Cref{part:medical-applications} contains papers on applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}.
+\Cref{part:unsupervised-uncertainty-estimation} is made up of three papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
+\Cref{part:self-supervised-speech-representation-learning} is a single, comprehensive review paper on self-supervised speech representation learning \parencite{mohamed_selfsupervised_2022}. 
+\Cref{part:medical-applications} consists of two papers on applications of machine learning within the medical domain, specifically recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}.
 

### 2023-10-02 chapters/main-contributions.tex
Work on discussion

@@
 For that reason, we here lay out the research questions and main contributions of each chapter in relation to the overall research project of the thesis. 
-The chapters are structured into three parts: {\color{dtured}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{dtured}\cref{part:self-supervised-speech-representation-learning}} is concerned with self-supervised speech representation learning, and {\color{dtured}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.
-Finally, {\color{dtured}\cref{part:discussion-and-conclusion}} contains a discussion and the conclusion of the thesis.
+The chapters are structured into three parts: {\color{black}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{black}\cref{part:self-supervised-speech-representation-learning}} is concerned with self-supervised speech representation learning, and {\color{black}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.
+Finally, {\color{black}\cref{part:discussion-and-conclusion}} contains a discussion and the conclusion of the thesis.
 

### 2023-10-03 appendices/paper_brief.tex
Work on abstract and discussion

@@
     \includegraphics[width=0.90\textwidth]{paper_brief/SSL_LVM.pdf}
-    \caption{A schematic overview of the two groups of models covered in this survey.
+    \caption[ schematic overview of the self-supervised and probabilistic latent variable models covered in this survey.]{A schematic overview of the two groups of models covered in this survey.
 \textit{Left}: A model trained with self-supervised learning.
@@
 \begin{sidewaystable*}[t]
-    \caption{Selected models classified according to the binary attributes identified throughout the text.
+    \caption[Classification of selected self-supervised and probabilistic latent variable models.]{Selected models classified according to the binary attributes identified throughout the text.
 The models are sorted according to first publication date on arXiv which might differ from the citation year.
@@
 \begin{table}[t!]
-\caption{A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable.
+\caption[A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable.]{A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable.
 The observation, prior and inference models may all belong to one or more of the categories listed under them as detailed in \cref{sec:plvms}.
@@
 \begin{sidewaystable*}[t]
-    \caption{Selected latent variable models classified according the attributes defined throughout \cref{sec:plvms}.
+    \caption[Classification of selected latent variable models.]{Selected latent variable models classified according the attributes defined throughout \cref{sec:plvms}.
 See \cref{tab:lvm-model-primitives} for the probability distributions that correspond to each of the attribute short-hands.

### 2023-10-03 appendices/supplementary_retrospective.tex
Work on abstract and discussion

@@
     \centering
-    \caption{Overall performance on MH-1813 test data for the model that also takes patient age and sex as direct inputs.
+    \caption[Overall performance on MH-1813 test data for the model that also takes patient age and sex as direct inputs.]{Overall performance on MH-1813 test data for the model that also takes patient age and sex as direct inputs.
 We also list the original performance of call-takers and the model (w/o sex and age) from the main manuscript for ease of comparison [mean (95\% CI)].}
@@
     \centering
-    \caption{Overall performance on MH-1813 test data for the model that also takes patient age and sex as direct inputs.
-We also list the original performance of call-takers and the model (w/o sex and age) from the main manuscript for ease of comparison [mean (95\% CI)].}
+    \caption[Overall performance on MH-1813 test data for a fine-tuned BERT model.]{Overall performance on MH-1813 test data for a fine-tuned BERT model.
+We also list the original performance of call-takers and the ensemble model from the main manuscript for ease of comparison [mean (95\% CI)].}
     \label{tab_retrospective:tableA8}

### 2023-10-03 chapters/discussion.tex
Work on abstract and discussion

@@
 
-\section{\Cref{chp:paper-retrospective} revisited: Uncertainty}
+\section{\Cref{chp:paper-retrospective} revisited: Calibration}
 The work in \textcite{wenstrup_retrospective_2023} focuses on the predictive performance of the ensemble model and an analysis of feature importance, but does not explicitly consider uncertainty estimation.
@@
 On the left, we plot the resulting calibration curves and on the right we show the logistic and isotonic fits.
-We see that both methods result in quite good calibrations\footnote{Brier scores: Uncalibrated = $0.003500$, logistic = $0.001807$, isotonic = $0.001774$.
-Brier skill score compared to uncalibrated (relative improvement): Logistic = $0.4830$, isotonic = $0.4924$} and that the predicted probabilities are shifted towards smaller values.
+We see that both methods result in quite good calibrations\footnote{Test set Brier scores: Uncalibrated = $0.003500$, logistic = $0.001807$, isotonic = $0.001774$.
+Brier skill score compared to uncalibrated (relative improvement): Logistic = $0.4830$, isotonic = $0.4924$.} and that the predicted probabilities are shifted towards smaller values.
 
-However, while correct calibration provides some interpretability to the predictions, the resulting probabilities are still functions of the model's performance.
-For example, with the logistic calibration, we can note that the predicted probabilities have been shifted so far towards smaller values that the best the model can do, is a predicted probability of about 50\%.
-This means strokes can be predicted at best with 50\% probability of being correct.
-If this model were to be deployed in a randomized controlled trial, the calibrated probabilities might be useful to incorporate in the design of the user interface such that a user can discern between confident and uncertain model predictions.
-Nonetheless, it is easy to imagine how this generally low confidence might lead the user to ignore the model.
+While correct calibration provides some interpretability to the predictions, the actual values of the probabilities are still tied of the model's performance.
+For example, with the logistic calibration, we can note that the predicted probabilities have been shifted so far towards smaller values that the highest possible predicted probability is about 50\%.
+This means strokes can be predicted at best with a 50\% chance of being correct.
+If this model were to be deployed in a randomized controlled trial, the calibrated probabilities might be incorporated in the design of the user interface such that a user can discern between confident and uncertain model predictions.
+However, it is easy to imagine how this generally low confidence might lead the user to ignore the model.
+This highlights a problem with uncertainties
 

### 2023-10-03 chapters/introduction.tex
Work on abstract and discussion

@@
     \includegraphics[height=0.4\textheight]{plimpton_322.png}
-    \caption{Technology is difficult.
+    \caption[Technology is difficult.
+The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets.]{Technology is difficult.
 The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics.
@@
     \includegraphics[height=0.4\textheight]{eniac_programmers.jpeg}
-    \caption{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.
-\parencite[photo credit][]{usarmyresearchlaboratoryarltechnicallibrary_female_1940}} %Photo: \copyright Corbis/ Getty Images.}
+    \caption[Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.]{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.
+\parencite[photo credit][]{usarmyresearchlaboratoryarltechnicallibrary_female_1940}}
     \label{fig:eniac_programmers}
@@
     \includegraphics[width=0.90\textwidth]{uber_nhsa_accident.png}
-    \caption{A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
+    \caption[A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.]{A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
 The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \parencite[photo credit][]{nationaltransportationsafetyboardnhsa_collision_2019}.}

### 2023-10-04 chapters/background.tex
Work on abstract

@@
 \begin{equation} \label{eq:corrected-differential-entropy-jaynes}
-    H_{N\rightarrow\infty}(X) \equiv \lim_{N\rightarrow\infty} [H_{N}(X) - \log N] = - \int_\mathcal{X} p(x) \log \left(\frac{p(x)}{q(x)}\right) \enspace ,
+    H_{N\rightarrow\infty}(X) \equiv \lim_{N\rightarrow\infty} [H_{N}(X) - \log N] = - \int_{\mathcal{X}} p(x) \log \left(\frac{p(x)}{q(x)}\right) \, dx \enspace ,
 \end{equation}
@@
     \begin{equation*}
-        H_{N\rightarrow\infty}(y(x)) =  = - \int_\mathcal{Y} \widetilde{w}(y)\,\frac{dy}{dx} \log \frac{\widetilde{w}(y)\frac{dy}{dx}}{\widetilde{q}(y)\frac{dy}{dx}} \, dx = - \int_\mathcal{Y} w(y) \log \frac{w(y)}{q(y)} \, dy \enspace .
+        H_{N\rightarrow\infty}(y(x)) =  = - \int_{\mathcal{Y}} \widetilde{w}(y)\,\frac{dy}{dx} \log \frac{\widetilde{w}(y)\frac{dy}{dx}}{\widetilde{q}(y)\frac{dy}{dx}} \, dx = - \int_{\mathcal{Y}} w(y) \log \frac{w(y)}{q(y)} \, dy \enspace .
     \end{equation*}
@@
 \begin{equation} \label{eq:kullback-leibler-divergence}
-    D_{KL}(p(x)\parallel q(x)) = \int_\mathcal{X} p(x) \log \left(\frac{p(x)}{q(x)}\right) \enspace .
+    D_{KL}(p(x)\parallel q(x)) = \int_{\mathcal{X}} p(x) \log \left(\frac{p(x)}{q(x)}\right) \enspace .
 \end{equation}

### 2023-10-04 chapters/introduction.tex
Work on abstract

@@
 
+% \todo[inline]{Lead-in with conversations and the patient interview and how it affects the patient journey.
+How healthcare has seen rapid advances in medicine and medical devices for diagnostics and treatment but the patient interview and triaging has seen little development.}
+
 In an era where technology increasingly intertwines with healthcare, artificial intelligence is on the verge of becoming part of standard medical practice. 
-Growing administrative burdens, rapid scientific progress in medicine, and aging populations have created a need to rethink how medical professionals are best enabled to successfully do their job. 
+Growing administrative burdens, aging populations, and rapid scientific progress in medicine and medical devices have created a need to rethink how medical professionals are best enabled to successfully do their job.
 The application of artificial intelligence in medical decision-making and particularly for diagnosing and recommending treatments is likely to hold significant potential to reduce the effect of these issues and improve patient outcomes.
@@
 Working with Babbage, Ada Lovelace is widely regarded as the first to have recognized that programmable computers could have applications beyond pure arithmetic.
-However, it was not until the advent of electronics that the first successful general-purpose computers were built, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}).
+However, only with the advent of electronics were the first successful general-purpose computers built, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}).
 % While the potential of a general-purpose programmable computer with applications beyond pure arithmetic was recognized by Ada Lovelace,  
@@
 % the first programmable, electronic, general-purpose digital computers were developed, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}) \parencite{georges_universal_2001}. 
-As manufacturing processes were developed and scaled, electronic computers were adopted widely across society for various applications in science and industry, including healthcare. 
-The societal changes brought about by the widespread adoption of computer technology through the latter half of the 20th century were so profound that some scholars refer to the era as the information age \parencite{georges_universal_2001, harari_sapiens_2011}.
+As manufacturing processes were developed and scaled, electronic computers were adopted widely across society for various applications in science and industry, including healthcare \parencite{georges_universal_2001, harari_sapiens_2011}.
+% The societal changes brought about by the widespread adoption of computer technology through the latter half of the 20th century were so profound that some scholars refer to the era as the information age \parencite{georges_universal_2001, harari_sapiens_2011}.
 
@@
 
-With the advent of electronic computers, the internet, and the widespread adoption of digital technologies continuing into the 21st century, the amount of data available to decision-makers has grown exponentially.
+With the digital revolution continuing into the 21st century and the worldwide spread of the internet, the amount of data available to decision-makers has grown exponentially. 
 This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that focuses on developing algorithms that can learn from data to recognize patterns and make predictions to guide decision-making.
@@
 The car, an SUV operated by Uber, was driving in autonomous mode when it struck a pedestrian crossing the street with her bicycle. 
-According to the investigation by the National Transportation Safety Board, the car's sensors detected the pedestrian $5.6$ seconds before the crash, but it did not manage to correctly predict her path or reduce the SUV's speed (\cref{fig:uber_nhsa_accident}).
+According to the investigation by the National Transportation Safety Board, the car's sensors detected the pedestrian more than five seconds before the crash, but the system did not manage to correctly predict her path or reduce the SUV's speed (\cref{fig:uber_nhsa_accident}).
 Specifically, during those seconds, the system incorrectly classified the pedestrian more than ten times, first as a vehicle, then as an unknown object and ultimately as a bicyclist, each time changing, or resetting, the predicted path.
 About one second before the crash, the system determined that a collision was imminent, but the situation exceeded the constraints within which the autonomous driving system was allowed to operate, and the car's safety driver failed to intervene \parencite{nationaltransportationsafetyboardnhsa_collision_2019}.
+%
+And Uber is not the only actor facing the challenges of applying machine learning in real-world scenarios.
+A recent report by the Washington Post found that Tesla's Autopilot system has been activated during at least 736 crashes in the last four years with 16 fatalities \parencite{siddiqui_17_2023}.
+% And though recent statistics published by Waymo, 
+% %the self-driving car company owned by Google's parent company Alphabet, 
+% revealed far fewer crashes, with only 2 of 20 incidents over 1 million miles meeting reporting criteria, it is interesting to note that the cars are all operated in the sunny states of California and Arizona, where driving conditions vary relatively little \parencite{hawkins_waymo_2023}.
+While driver-assistance systems undoubtedly help reduce accidents in many cases, and has great future potential, the issues of current systems highlight the need for machine learning systems that are robust in rare, adverse or difficult situations.
 
@@
 
-And Uber is not the only actor facing the challenges of applying machine learning in real-world scenarios.
-A recent report by the Washington Post found that Tesla's Autopilot system has been involved in at least 736 crashes in the last four years with 16 fatalities \parencite{siddiqui_17_2023}.
-% And though recent statistics published by Waymo, 
-% %the self-driving car company owned by Google's parent company Alphabet, 
-% revealed far fewer crashes, with only 2 of 20 incidents over 1 million miles meeting reporting criteria, it is interesting to note that the cars are all operated in the sunny states of California and Arizona, where driving conditions vary relatively little \parencite{hawkins_waymo_2023}.
-While driver-assistance systems undoubtedly help reduce accidents in many cases, and has great future potential, the issues of current systems highlight the need for machine learning systems that are robust in rare, adverse or difficult situations.
-
 Recently, European policymakers proposed a set of requirements to ensure that machine learning applications are safe, reliable, and trustworthy \parencite{europeancommission_briefing_2021}. 
@@
 Limited risk applications, such as in computer games, must meet transparency requirements, and technologies such as public facial recognition are banned due to unacceptable risk.
-High risk applications include autonomous vehicles and systems within healthcare. These must meet a set of strict requirements focusing on anti-discrimination and robustness.
+High risk applications include autonomous vehicles and systems within healthcare. To avoid critical errors, these must meet a set of strict requirements focusing on anti-discrimination and robustness.
 
 One of the key ways to improve the safety, reliability and trustworthiness of machine learning systems is to equip them with the ability accurately estimate the uncertainty of their predictions. 
-Almost any model will inevitably make mistakes, but if it can accurately estimate the uncertainty of its predictions, it can communicate to other systems or humans when it is not to be trusted, and so, reduce the risk that wrong predictions lead to catastrophic failures.
+While any model will inevitably make a wrong prediction, by accurately estimating the uncertainty it can communicate to other systems or humans when it is not to be trusted, and so, reduce the risk that wrong predictions lead to catastrophic failures.
 Fittingly, a recent study by the European Parliamentary Research Services concluded:
@@
 \end{center}
-
-% \todo[inline]{Add something here about probability theory, Bayesian modelling etc. as a warm-up for potential solutions.}
 
@@
 
-\todo[inline]{Can we find a better example than the Rhind Papyrus, or an additional one?
+% \todo[inline]{Can we find a better example than the Rhind Papyrus, or an additional one?
 It could be more on-point with decision-making and more well-known.
@@
 %
-Several factors contribute to the reliability of a machine learning model. 
+Several factors define the reliability of a machine learning model. 
 Besides model performance and accuracy, important factors include the interpretability of how the model functions, the explainability of its predictions, fairness in its treatment of different groups, and robustness to noise, outliers and adversarial attacks.
-% In medical applications, it has been argued that another factor
 Since many modern machine learning models are deep neural networks with millions or billions of parameters, their size and complexity make them inherently difficult to interpret, and their predictions hard to explain.
-Due to high cost and practical infeasibility, the vast amounts of data needed to train such models are often not manually curated for quality, and so, may contain biases and errors which models are well-equipped to learn to mirror \parencite{burkart_survey_2021}.
-Finally, a number of factors, including the ability of deep neural networks to overfit, or even memorize, the training data \parencite{arpit_closer_2017, burg_memorization_2021}, can lead to models that are sensitive adverse noise conditions and outliers.
+Due to high cost and practical infeasibility, the vast amounts of data needed to train such models are often not manually curated for quality, and so, may contain biases and errors which models are well-equipped to learn to mirror, risking fairness \parencite{burkart_survey_2021}.
+Finally, a number of factors, including the ability of deep neural networks to overfit, or even memorize, their training data \parencite{arpit_closer_2017, burg_memorization_2021}, can lead to models that are sensitive to adverse noise conditions and outliers.
 
 The ability of a model to accurately estimate the uncertainty of its predictions plays an important role in its reliability. 
-However, the tendency of deep neural networks to overfit or memorize the training data \parencite{arpit_closer_2017, burg_memorization_2021} and the use of training objectives that are proxies for the evaluation metrics of interest lead to models that often are miscalibrated; that is, the predicted probability of a class does not reflect the true probability of the class.
-This means, for instance, that if for a certain input the model assigns 70\% probability to a class, it would not generally be right 70\% of the time - if we could repeat the prediction multiple times under similar circumstances \parencite{guo_calibration_2017, kull_temperature_2019}.
-Usually, the predicted probabilities are too extreme which leads to overconfident predictions.
-This means a model will often assign a high probability to the predicted class, even when it is wrong.
+While an accurate uncertainty estimate provides a certain degree of interpretability, it especially improves robustness by indicating that uncertain predictions are not to be trusted. 
+However, overfitting, memorization and the use of training objectives that are proxies for the evaluation metrics of interest often lead to models that are miscalibrated; that is, the predicted probability of a class does not reflect the true probability of the model being correct \parencite{guo_calibration_2017, kull_temperature_2019}.
+% This means, for instance, that if for a certain input the model assigns 70\% probability to a class, it would not generally be right 70\% of the time - if we could repeat the prediction multiple times under similar circumstances \parencite{guo_calibration_2017, kull_temperature_2019}.
+Usually, the predicted probabilities are too extreme which leads to overconfident predictions. This means a model will often assign a high probability to the predicted class, even when it is wrong.
 
-In practice, miscalibrated models are not reliable, especially in high risk applications.
+Miscalibrated models are unreliable, especially in high risk applications.
 Although no model is perfect, the inability of a human, or another model, to assess the certainty with which a prediction was made, makes it impossible to know how to weigh it in a decision-making process.
@@
 
-It is also important to realize that correct calibration of a machine learning model does not guarantee that the probability of a prediction is accurate. 
-Specifically, even well-calibrated model can be overconfident for data that were not presented to it during training.
-This includes rare events, outliers, and adverse examples which are sometimes collectively referred to as out-of-distribution examples.
-To give an example, a model trained to classify images of cats and dogs and provide calibrated class probabilities will still fail if presented with an image of a horse.
-In this case, it has no option but to distribute 100\% total probability across the cat and dog categories, even though that is clearly wrong.
+More importantly, correct calibration of a machine learning model does not guarantee that the predicted probability is accurate. 
+Specifically, even a well-calibrated model can be overconfident for data that were not presented to it during training such as rare events, outliers, and adverse examples, sometimes collectively referred to as out-of-distribution examples.
+For instance, take a perfectly calibrated model trained to classify images of cats and dogs.
+If presented with an image of a horse, it has no option but to distribute 100\% total probability across the cat and dog categories, even though that is clearly wrong.
 Worse yet, to indicate the uncertainty, we might expect the model to assign 50\% probability to each of the cat and dog categories, but that behavior is not guaranteed. 
-Indeed, since no horses were in the training data, the model will not have learned specialized features for horses, nor learned to associate relevant known features with horses.
-So, if some features of a particular horse resemble those learned for a dog more than for a cat, the model might assign arbitrarily high probability to the dog class and therefore risks being confidently wrong \parencite{zhou_survey_2022}.
+Specifically, since no horses were in the training data, the model will not have learned specialized features for horses, nor learned to associate relevant known features with horses.
+So, if a particular horse has features that resemble those learned for a dog more than for a cat, the model might assign arbitrarily high probability to the dog class.
+Therefore, even perfectly calibrated models risks being confidently wrong \parencite{zhou_survey_2022}.
 
@@
 As hinted at in the previous section, there are different sources of uncertainty in machine learning models. 
-At a fundamental level, we can distinguish between uncertainty that is present in the knowledge we have, and uncertainty that originates from the knowledge that we do not have.
+At a fundamental level, we can decompose the \textit{predictive uncertainty} into uncertainty that is present in the knowledge we have, and uncertainty that originates from the knowledge that we do not have.
 These sources are sometimes called \emph{known unknowns} and \emph{unknown unknowns}, or referred to in terms of \textit{aleatoric uncertainty} and \textit{epistemic uncertainty} \parencite{kendall_what_2017}.
@@
 Aleatoric uncertainty is sometimes referred to as \textit{stochastic uncertainty} and epistemic uncertainty as \textit{systematic uncertainty} \parencite{kendall_what_2017}. 
-Although the distinction between aleatoric and epistemic uncertainty is important, it is not always useful.
+% Although the distinction between aleatoric and epistemic uncertainty is important, it is not always useful.
 For instance, a model may be uncertain about a particular prediction due to any of these two kinds of uncertainty.
@@
 Unknown words are arguably sources of epistemic uncertainty although a good speech recognition model may generalize well if the word's spelling and pronunciation follow the same patterns as the words in the training data.
-Another example of epistemic uncertainty is the occurrence of truly out-of-distribution examples, such as the horse in the image classifier example from earlier: inputs unlike anything the model has seen during training, especially if they require outputs unavailable to the model.
-In high-dimensional data with a practically unlimited diversity in out-of-distribution examples, it is impossible to collect enough data to eliminate all potential sources of epistemic uncertainty.
-Let us include horses in the training data.
-The model would likely learn to recognize them, but it would still be unable to recognize giraffes, or zebras, or unicorns.
+Another example of epistemic uncertainty is the occurrence of truly out-of-distribution examples, such as the horse in the image classifier example from earlier: inputs unlike anything the model has seen during training.
+%, especially if they require outputs unavailable to the model.
+In high-dimensional data with a practically unlimited diversity in out-of-distribution examples, it is impossible to collect enough data to eliminate all potential sources of epistemic uncertainty: By including horses in the training data, the model would likely learn to recognize them, but would still be unable to recognize giraffes, or zebras, or unicorns.
 Ultimately, we must accept that any practical machine learning model will have some sources of epistemic uncertainty.
@@
 Other approaches include anomaly detection and the recent field of out-of-distribution detection which, for example, represent epistemic uncertainty by special output classes for out-of-distribution data, distributions over data representations, or distances between them (see \cref{sec:out-of-distribution-detection}).
+In this thesis, we will focus on the latter approach.
 

### 2023-10-04 appendices/supplementary_hierarchical.tex
Work on font and fix typos

@@
 
-For SmallNORB and Omniglot we resize the original grey-scale images to $28\times28$ with ordinary bi-linear interpolation.
-For each of these datasets, we also create a version where the grey-scale is inverted.
+For SmallNORB and Omniglot we resize the original grayscale images to $28\times28$ with ordinary bi-linear interpolation.
+For each of these datasets, we also create a version where the grayscale is inverted.
 We do this because, the overall white nature of the images tends to make detecting them as OOD from FashionMNIST artificially easy.
@@
 Our Hierarchical VAE (HVAE) model uses bottom-up inference and top-down generative paths as specified in the paper.
-For grey-scale images, the output is parameterized by a Bernoulli distribution while for natural images we use a Discretized Logistic Mixture \parencite{salimans_pixelcnn_2017}.
+For grayscale images, the output is parameterized by a Bernoulli distribution while for natural images we use a Discretized Logistic Mixture \parencite{salimans_pixelcnn_2017}.
 The latent variables are parameterized by stochastic layers that output the mean and log-variance of a diagonal covariance Gaussian.
 The prior distribution on the top-most latent is a standard Gaussian.
-For grey-scale images, the lowest latent space is parameterized by a convolutional neural network and has dimensions $14\times14\times8$ interpreted as (height $\times$ width $\times$ latent dimension).
+For grayscale images, the lowest latent space is parameterized by a convolutional neural network and has dimensions $14\times14\times8$ interpreted as (height $\times$ width $\times$ latent dimension).
 The highest two latent variables are parameterized by dense transformations with $16$ and $8$ units, respectively.
@@
 
-Each stochastic layer is preceeded by a determininistic transformation.
-For both grey-scale and natural images, each deterministic transformation consists of three residual blocks of the same type used by \textcite{maaloe_biva_2019}. The structure of a residual block is:
+Each stochastic layer is preceded by a deterministic transformation.
+For both grayscale and natural images, each deterministic transformation consists of three residual blocks of the same type used by \textcite{maaloe_biva_2019}. The structure of a residual block is:
 \begin{equation}
@@
 In a deterministic transformation, any non-unit stride is performed in the third residual block.
-For grey-scale images, we stride by 2 in the first and second deterministic transformations but not the third.
+For grayscale images, we stride by 2 in the first and second deterministic transformations but not the third.
 For natural images, we similarly stride by 2 in the first and second deterministic transformations.
-For grey-scale we use 64 channels while we use 256 for natural images.
+For grayscale we use 64 channels while we use 256 for natural images.
 In both cases, the first deterministic block uses a kernel size of 5 and the latter two a kernel of size 3.

### 2023-10-04 chapters/background.tex
Work on font and fix typos

@@
 We start by quantifying the concept of uncertainty relating it to information and probability.
-We then introduce the task of out-of-distribution detection and review the existing litterature on the topic.
+We then introduce the task of out-of-distribution detection and review the existing literature on the topic.
 Then, we describe variational autoencoders, how they can be used to learn representations of speech, and some of their challenges.
@@
 To provide some clarity, we will briefly outline the taxonomy of \textcite{yang_generalized_2022}:
-Outlier detection most clearly differs from the other tasks by directly processesing all observations and aiming to select outliers from a single contaminated dataset. 
+Outlier detection most clearly differs from the other tasks by directly processing all observations and aiming to select outliers from a single contaminated dataset. 
 The remaining tasks differ in whether they detect both covariate and semantic shift and require the simultaneous classification of in-distribution classes.
@@
 
-The many different ways of defining a score $s(\xb)$ can be used to further categorize OOD detection methods.
+There are several ways to define a score $s(\xb)$, and they be used to further categorize OOD detection methods. 
 Unsupervised methods often derive the score 
@@
 In the supervised setting the supervision can come from different sources depending on how much we know, or assume to know, about the OOD data.
-In the least informed setting, the supervision is from the target variable $y$ only.
-In the most informed setting we also assume access to representative OOD data.
+Taking supervision from the target variable $y$ requires few assumptions while to representative OOD data.
 
@@
 They also propose adding an OOD class to softmax classifiers and training it to predict perturbed in-distribution data, or alternatively, training the predicted class distribution to output uniform distribution for perturbed in-distribution inputs.
-While these methods are appealing, their weaknesses have been pointed out by later work which generally improve on their perforamnce.
+While these methods are appealing, their weaknesses have been pointed out by later work which generally improve on their performance.
 An example is ODIN \parencite{liang_enhancing_2018}, in which the authors propose to calibrate $p(y|\xb)$ with temperature scaling \parencite{guo_calibration_2017} and add gradient-based, input-dependent perturbations to the inputs and use the calibrated maximum class probability as the OOD score.
@@
 In \textcite{ren_likelihood_2019}, the authors propose to use the likelihood ratio between a model trained on in-distribution data and a background model trained on perturbed in-distribution data as the OOD score.
-\textcite{serra_input_2020} argue that the failure of DGMs is due to the high-influence that the input complexity has on the likelihood.
+\textcite{serra_input_2020} argue that the failure of deep generative models is due to the high-influence that the input complexity has on the likelihood.
 Therefore, they propose to use a general lossless image compression algorithm as a background model.
 \textcite{choi_waic_2019} propose to use the Watanabe information criterion (WAIC) computed from the likelihood \parencite{watanabe_algebraic_2009, watanabe_asymptotic_2010}. 
-For variational autoencoders, other work proposes to refit the encoder on a test data example, hypothesizing that the likelihood of OOD data will improve more than for in-distribution data, and use this ``likelihood regret" as the OOD score \parencite{xiao_likelihood_2020}.
-\textcite{maaloe_biva_2019} provide initial results that a loosened variational bound on the likelihood, using only encoded representations from the top-most latent variables in a hierarchical variationa autoencoder, can improve OOD detection performance.
-In \textcite{havtorn_hierarchical_2021}, we show that variational autoencoders are surprisingly good at reconstructing OOD data and propose an improved score based on the likelihood ratio of such loosened bounds \textcite{maaloe_biva_2019}.
+For variational autoencoders, other work proposes to refit the encoder on a test data example, hypothesizing that the likelihood of OOD data will improve more than for in-distribution data, and use this ``likelihood regret'' as the OOD score \parencite{xiao_likelihood_2020}.
+\textcite{maaloe_biva_2019} provide initial results that a loosened variational bound on the likelihood, using only encoded representations from the top-most latent variables in a hierarchical variational autoencoder, can improve OOD detection performance.
+In \textcite{havtorn_hierarchical_2021}, we show that variational autoencoders are surprisingly good at reconstructing OOD data and propose an improved score based on the likelihood ratio of such loosened bounds.
 
@@
 \subsection{Variational autoencoders} \label{sec:variational-autoencoders}
-To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to parameterize the generative model $p_\thetab(\xb,\zb)$ with a neural network with parameters $\thetab$.
-Specifically, they use a neural network to parameterize the conditional distribution $p_\thetab(\xb|\zb)$ and a simple distribution for the prior $p(\zb)$.
+To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to parameterize the generative model $p_\thetab(\xb,\zb)$ as a neural network with parameters $\thetab$.
+Specifically, they use a neural network to parameterize the conditional distribution $p_\thetab(\xb|\zb)$ and choose some simple distribution for the prior $p(\zb)$.
 Since this makes the marginal and posterior distributions intractable, both training and inference require approximate methods. 
@@
 where $\zb_k \sim q_\phib(\zb|\xb)$ are samples from the approximate posterior.
-To reduce the variance of the estimate, \textcite{burda_importance_2016} propose to use importance sampling to reweight the samples.
+To reduce the variance of the estimate, \textcite{burda_importance_2016} propose to use importance sampling to reweigh the samples.
 
@@
 Via the ELBO, we want to use maximum likelihood estimation to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$.
-Gradient-based optimization works well for neural networks but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
+Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
 Computing the gradient of the ELBO with respect to $\thetab$ is straightforward,
@@
 Regardless of the choice of inference model, the resulting hierarchical VAE is trained using the ELBO \cref{eq:variational-autoencoder-elbo} and the reparameterization trick.
-Recent architectural advances have alleviated the posterior collapse problem and made it possible to train VAEs with many latent variables.\parencite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
+Recent architectural advances have alleviated the posterior collapse problem and made it possible to train VAEs with many latent variables \parencite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
 
@@
     \mathbb{E}_{\hat{p}(\xb)}\left[ \log p(\xb) \right] 
-    &\geq \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log \frac{p_{}(\xb|\zb)p(\zb)}{q_{}(\zb|\xb)} \right] \enspace .
+    &\geq \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log \frac{p_{}(\xb|\zb)p(\zb)}{q_{}(\zb|\xb)} \right] \enspace ,
 \end{align}

### 2023-10-04 chapters/introduction.tex
Work on font and fix typos

@@
 Apart from such inherent limitations of the tools, inadequate human proficiency in their use was also a source of error.
-The Plimpton 322 clay tablet, for instance, contains several errors, presumably made by a novice in Babylonian mathematics (\cref{fig:plimpton_332}) \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
+For example, the Plimpton 322 clay tablet (\cref{fig:plimpton_332}) contains several errors, presumably made by a novice in Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
 % Easy to overlook, the challenges we face in adopting new technologies today are not new but have been faced by humans for thousands of years.
@@
 
-\begin{figure}[t]
+\begin{figure}
     \centering
@@
     \caption[Technology is difficult.
-The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets.]{Technology is difficult.
+The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets.]{ Technology is difficult.
 The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics.
@@
     \includegraphics[height=0.4\textheight]{eniac_programmers.jpeg}
-    \caption[Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.]{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.
+    \caption[Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.]{ Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.
 \parencite[photo credit][]{usarmyresearchlaboratoryarltechnicallibrary_female_1940}}
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=0.90\textwidth]{uber_nhsa_accident.png}
+    \caption[A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.]{ A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
+The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \parencite[photo credit][]{nationaltransportationsafetyboardnhsa_collision_2019}.}
+    % \caption{Aerial view of crash location showing path of pedestrian as she attempted to cross N.
+Mill Avenue and movement and speed of SUV at three points before impact.
+Pedestrian's path shows her position from initial detection ($5.6$ seconds before impact) until impact; SUV's position is shown at corresponding times beginning $4.2$ seconds before impact \parencite{nationaltransportationsafetyboardnhsa_collision_2019}.}
+    \label{fig:uber_nhsa_accident}
+\end{figure}
+
 However, these technologies come with the same risks as previous generations did, and in some cases, may amplify them. 
@@
 The car, an SUV operated by Uber, was driving in autonomous mode when it struck a pedestrian crossing the street with her bicycle. 
-According to the investigation by the National Transportation Safety Board, the car's sensors detected the pedestrian more than five seconds before the crash, but the system did not manage to correctly predict her path or reduce the SUV's speed (\cref{fig:uber_nhsa_accident}).
+According to the investigation by the National Transportation Safety Board, the car's sensors detected the pedestrian more than five seconds before the crash, but the system did not manage to correctly predict her path or reduce the SUV's speed.
 Specifically, during those seconds, the system incorrectly classified the pedestrian more than ten times, first as a vehicle, then as an unknown object and ultimately as a bicyclist, each time changing, or resetting, the predicted path.
-About one second before the crash, the system determined that a collision was imminent, but the situation exceeded the constraints within which the autonomous driving system was allowed to operate, and the car's safety driver failed to intervene \parencite{nationaltransportationsafetyboardnhsa_collision_2019}.
+About one second before the crash, the system determined that a collision was imminent, but the situation exceeded the constraints within which the autonomous driving system was allowed to operate, and the car's safety driver failed to intervene (\cref{fig:uber_nhsa_accident}) \parencite{nationaltransportationsafetyboardnhsa_collision_2019}.
 %
@@
 While driver-assistance systems undoubtedly help reduce accidents in many cases, and has great future potential, the issues of current systems highlight the need for machine learning systems that are robust in rare, adverse or difficult situations.
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.90\textwidth]{uber_nhsa_accident.png}
-    \caption[A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.]{A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
-The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \parencite[photo credit][]{nationaltransportationsafetyboardnhsa_collision_2019}.}
-    % \caption{Aerial view of crash location showing path of pedestrian as she attempted to cross N.
-Mill Avenue and movement and speed of SUV at three points before impact.
-Pedestrian's path shows her position from initial detection ($5.6$ seconds before impact) until impact; SUV's position is shown at corresponding times beginning $4.2$ seconds before impact \parencite{nationaltransportationsafetyboardnhsa_collision_2019}.}
-    \label{fig:uber_nhsa_accident}
-\end{figure}
 
@@
 Additionally, a single medical note will usually contain several diagnoses and procedures which must all be inferred from the clinical note \parencite{johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}.
-Furthermore, any single code can have several exclusion or inclusion criteria defined by official guidelines that determine when other codes must be coded along with it, and when they must not \parencite{centersformedicaremedicaidservicesus_icd10cm_2023}.
+Furthermore, any single code can have several criteria defined by official guidelines that determine under which conditions the code is mutually exclusive with other codes, and when other codes must be coded along with it \parencite{centersformedicaremedicaidservicesus_icd10cm_2023}.
 % Official guidelines describe how to code specific diagnoses and procedures and combinations thereof . 
@@
 For instance, the prevalence of different medical codes in the training data is highly imbalanced; it is common to have several orders of magnitude difference between the frequency of the most frequent code and the least frequent code.
-Since each clinical note is associated with multiple codes that have complex co-occurence patterns, it is often impossible to exactly correct for this class imbalance by stratified sampling, which is an otherwise common approach to deal with class imbalance in machine learning.
+Since each clinical note is associated with multiple codes that have complex co-occurrence patterns, it is often impossible to exactly correct for this class imbalance by stratified sampling, which is an otherwise common approach to deal with class imbalance in machine learning.
 % This also gives rise to 
-Furthermore, some codes are highly similar, for instance the ICD-10 codes Z87.891 ``Personal history of nicotine dependence'' and F17.210 ``Nicotine dependence, cigarettes, uncomplicated", but in practice, only one of them should be assigned to a given note.
+Furthermore, some codes are highly similar, for instance the ICD-10 codes Z87.891 ``Personal history of nicotine dependence'' and F17.210 ``Nicotine dependence, cigarettes, uncomplicated'', but in practice, only one of them should be assigned to a given note.
 As with the stroke recognition system, we would like the model to express high degree of uncertainty when it makes a mistake.

### 2023-10-04 chapters/main-contributions.tex
Work on font and fix typos

@@
 \chapter[research questions and main contributions]{Research questions and main contributions}\label{chp:main-contributions}
-\newcommand{\contribitem}[1]{{\vspace{1em}\noindent\scshape\bfseries{\color{dtured}\cref{#1}}\\\ucnameref{#1}\\}}
+
+\newcommand{\contribitem}[1]{{\vspace{1em}\noindent\scshape\bfseries\Large{\color{dtured}\cref{#1}}\\\ucnameref{#1}}\\}
 
@@
 We also find a widespread error in the calculation of the macro F1 score. 
-We propose new data splits created with stratified sampling and use identical experimental setups to compare models, tuning hyperparameters and decision boundaries. 
+To compare models, we propose new data splits created with stratified sampling, use identical experimental setups and tune hyperparameters and decision boundaries. 
 By analyzing prediction errors, we confirm the observation of previous work that all models struggle with rare codes, although long documents only have a negligible impact on performance, contrary to previous claims.

### 2023-10-04 chapters/paper_hierarchical.tex
Work on font and fix typos

@@
 \section{Introduction}
-\begin{figure}[t]
+%
+\begin{figure}
     \centering
     \includegraphics[width=0.7\columnwidth]{paper_hierarchical/reconstructions-front-page-4-samples-2-recons2.pdf}
-    \caption[Reconstructions of a hierarchical VAE trained on FashionMNIST.]{%
-        Reconstructions using a hierarchical VAE trained on FashionMNIST.
+    \caption[Reconstructions of a hierarchical VAE trained on FashionMNIST.]{ Reconstructions using a hierarchical VAE trained on FashionMNIST.
         Reconstruction quality of OOD data is comparable to in-distribution data, resulting in high likelihoods and poor OOD discrimination.
@@
 \end{figure}
+%
 % First paragraph: Explains why OOD detection is important in a broad perspective
@@
 \section{Why does OOD detection fail?}\label{sec_paper_hierarchical:why-does-ood-fail}
-\begin{figure}[t]
+\begin{figure}
     \centering
@@
 For image data there is a trend that the low-level features are quite similar across models (edge detectors, etc.).
-This raises the question to what extend such features are relevant when detecting OOD data, also suggested  by \parencite{nalisnick_deep_2019} and examined for Glow and PixelCNN in \parencite{schirrmeister_understanding_2020}.
+This raises the question to what extent such features are relevant when detecting OOD data, also suggested  by \parencite{nalisnick_deep_2019} and examined for Glow and PixelCNN in \parencite{schirrmeister_understanding_2020}.
 To investigate, we train two hierarchical VAEs (\cref{sec_paper_hierarchical:background-hie-VAE}) on FashionMNIST and MNIST, respectively, and compute the between-models correlation of the extracted features of in-distribution data and OOD data.
@@
 There exists a few complementary approaches to make the VAE more flexible: (i) model a more expressive variational distribution $q_\phib(\zb|\xb)$ or prior distribution $p_\thetab(\zb)$ \parencite{rezende_variational_2015, kingma_improved_2016}, (ii) model a more expressive posterior distribution $p_\thetab(\xb|\zb)$ e.g.
-with a autoregressive decoder \parencite{oord_conditional_2016} and (iii) learn a deeper hierarchy of latent variables \parencite{burda_importance_2016, sonderby_ladder_2016}.
+with an autoregressive decoder \parencite{oord_conditional_2016} and (iii) learn a deeper hierarchy of latent variables \parencite{burda_importance_2016, sonderby_ladder_2016}.
 Here, we focus on the latter.
@@
 To alleviate this, we phrase OOD detection as a likelihood ratio test of being \emph{semantically} in-distribution.
-A standard likelihood ratio test \parencite{buse_likelihood_1982} suggests to consider the ratio between the associated likelihoods, which we can approximate on a log-scale by the corresponding lower bounds $\mathcal{L}$ and $\mathcal{L}^{>k}$,
+A standard likelihood ratio test \parencite{buse_likelihood_1982} suggests considering the ratio between the associated likelihoods, which we can approximate on a log-scale by the corresponding lower bounds $\mathcal{L}$ and $\mathcal{L}^{>k}$,
 \begin{equation}\label{eq_hierarchical:llr-as-difference-in-likelihoods}
@@
 % \begin{SCfigure}[50][t!]
-\begin{figure}[t]
+\begin{figure}
     %\begin{figure}
@@
         % nodes
-        \node[obs] (x_inf) {$\xb$};%
-        \node[latent,above=.75cm of x_inf](z1_inf){$\zb_1$}; %
-        \node[latent,above=.75cm of z1_inf](z2_inf){$\zb_2$}; %
-        % \node[latent,above=.75cm of z2_inf](z3_inf){$\zb_3$}; %
+        \node[obs] (x_inf) {$\xb$};
+        \node[latent,above=.75cm of x_inf](z1_inf){$\zb_1$};
+        \node[latent,above=.75cm of z1_inf](z2_inf){$\zb_2$};
+        % \node[latent,above=.75cm of z2_inf](z3_inf){$\zb_3$};
         \node[above=of z2_inf, yshift=-1.cm] (phi) {$q_\phib(\zb|\xb)$};
@@
         % edges
-        \edge[]{x_inf}{z1_inf};
-        \edge[]{z1_inf}{z2_inf};
-        % \edge[]{z2_inf}{z3_inf};
-        \edge[dashed, bend left]{x_inf}{z2_inf};
-        % \edge[dashed, bend left]{x_inf}{z3_inf};
+        \edge[]{x_inf}{z1_inf}; %
+        \edge[]{z1_inf}{z2_inf}; %
+        % \edge[]{z2_inf}{z3_inf}; %
+        \edge[dashed, bend left]{x_inf}{z2_inf}; %
+        % \edge[dashed, bend left]{x_inf}{z3_inf}; %
 
@@
         % nodes$
-        \node[obs,right=0.75cm of x_inf] (x_gen) {$\xb$};%
-        \node[latent,above=.75cm of x_gen](z1_gen){$\zb_1$}; %
-        \node[latent,above=.75cm of z1_gen](z2_gen){$\zb_2$}; %
-        % \node[latent,above=.75cm of z2_gen](z3_gen){$\zb_3$}; %
+        \node[obs,right=0.75cm of x_inf] (x_gen) {$\xb$};
+        \node[latent,above=.75cm of x_gen](z1_gen){$\zb_1$};
+        \node[latent,above=.75cm of z1_gen](z2_gen){$\zb_2$};
+        % \node[latent,above=.75cm of z2_gen](z3_gen){$\zb_3$};
         \node[above=of z2_gen, yshift=-1.cm] (theta) {$p_\thetab(\xb,\zb)$};
@@
         % edges
-        % \edge[]{z3_gen}{z2_gen};
-        \edge[]{z2_gen}{z1_gen};
-        \edge[]{z1_gen}{x_gen};
-        \edge[dashed, bend left]{z2_gen}{x_gen};
-        % \edge[dashed, bend left]{z3_gen}{x_gen};
+        % \edge[]{z3_gen}{z2_gen}; %
+        \edge[]{z2_gen}{z1_gen}; %
+        \edge[]{z1_gen}{x_gen}; %
+        \edge[dashed, bend left]{z2_gen}{x_gen}; %
+        % \edge[dashed, bend left]{z3_gen}{x_gen}; %
     }
     }
-    \caption[Inference and generative models for a bottom-up hierarchical VAEs.]{The inference and generative models, $q_\phib$ and $p_\thetab$, for an $L=2$ layered bottom-up hierarchical VAE as the one used in our experiments.
+    \caption[Inference and generative models for a bottom-up hierarchical VAEs.]{ The inference and generative models, $q_\phib$ and $p_\thetab$, for an $L=2$ layered bottom-up hierarchical VAE as the one used in our experiments.
     Dashed lines indicate deterministic skip connections which are employed in both networks.
@@
 We use a Bernoulli output distribution for FashionMNIST/MNIST and a discretized mixture of logistics output distribution \parencite{salimans_pixelcnn_2017} for CIFAR10/SVHN.
-We use $L=3$ for grey-scale images and $L=4$ for natural images.
+We use $L=3$ for grayscale images and $L=4$ for natural images.
 % For CIFAR/SVHN, we also train a BIVA model \parencite{maaloe_biva_2019} with $L=10$ and similar configuration as used by the original paper\footnote{Code available at \url{github.com/larsmaaloee/BIVA} and \url{github.com/vlievin/biva-pytorch}}.
@@
 
-\begin{table}[t!]
-    \caption[Average bits per dimension of different datasets for generative models trained on FashionMNIST and CIFAR10.]{%
-    Average bits per dimension of different datasets for models trained on FashionMNIST and CIFAR10.
+\begin{table}
+    \caption[Average bits per dimension of different datasets for generative models trained on FashionMNIST and CIFAR10.]{ Average bits per dimension of different datasets for models trained on FashionMNIST and CIFAR10.
     For the hierarchical models we include the $\mathcal{L}^{>k}$ bounds.
@@
     Since we train on dynamically binarized FashionMNIST, our bits/dim are smaller than for Glow.
-    As $k$ is increased for the $L^{>k}$ bound, the bound gets looser but the model eventually assigns higher likelihood to the in distribution data than to the OOD data.
+    As $k$ is increased for the $L^{>k}$ bound, the bound gets looser, but the model eventually assigns higher likelihood to the in distribution data than to the OOD data.
     Glow refers to \textcite{kingma_glow_2018, nalisnick_deep_2019}.
@@
 \todo[inline]{Check out if we can make underbrace like curly brackets to the left of tab_hierarchical:rocauc-ood to decrease height and fit page better.}
-\begin{table}[t!]
+\begin{table}
     \caption[AUROC, AUPRC, and FPR80 of generative models for OOD detection (MNIST/FashionMNIST and SVHN/CIFAR10).]{%
@@
 Backgr. contrast. LR (VAE) {\parencite{choi_waic_2019}}                    & $0.924$ & - & - \\
-Binary classiﬁer {\parencite{ren_likelihood_2019}}                              & $0.455$ & $0.505$ & $0.886$ \\ % 6
+Binary classifier {\parencite{ren_likelihood_2019}}                              & $0.455$ & $0.505$ & $0.886$ \\ % 6
 $p(\hat{y} | \xb)$ with OOD as noise class {\parencite{ren_likelihood_2019}}     & $0.877$ & $0.871$ & $0.195$ \\ % 7
@@
 Mahalanobis distance, DenseNet {\parencite{lee_simple_2018}}                & $0.986$ & - & - \\
-Ensemble, 20 classiﬁers {\parencite{ren_likelihood_2019, lakshminarayanan_simple_2017}}                  & $0.857$ & $0.849$ & $0.240$ \\
+Ensemble, 20 classifiers {\parencite{ren_likelihood_2019, lakshminarayanan_simple_2017}}                  & $0.857$ & $0.849$ & $0.240$ \\
          \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
@@
 % \begin{sidewaysfigure}
-\begin{figure}[t]
+\begin{figure}
     %\captionstyle{centerlast}
@@
 \subsection{Likelihood-ratio-based OOD detection}
-\begin{figure}[t]
+
+\begin{figure}
     \centering
@@
 \end{figure}
+
 We now move to the likelihood ratio-based score.
@@
 Our method outperforms other generative model-based methods such as WAIC \parencite{choi_waic_2019} with Glow model and performs similarly to the likelihood regret method of \parencite{xiao_likelihood_2020}.
-Furthermore, our method performs similarly to the background constrative likelihood ratio method of \textcite{ren_likelihood_2019} on FashionMNIST/MNIST but contrary to the failure of that method on CIFAR10/SVHN reported by \parencite{xiao_likelihood_2020}, our method performs very well on this task too.
+Furthermore, our method performs similarly to the background contrastive likelihood ratio method of \textcite{ren_likelihood_2019} on FashionMNIST/MNIST but contrary to the failure of that method on CIFAR10/SVHN reported by \parencite{xiao_likelihood_2020}, our method performs very well on this task too.
 Our approach outperforms all supervised approaches that use in-distribution labels or synthetic examples of OOD data derived from the in-distribution data including ODIN \parencite{liang_enhancing_2018} and the predictive distribution of a classifier $p(\hat{y}|\xb)$ trained and evaluated in various ways (see \textcite{ren_likelihood_2019}).
@@
 In this paper we study unsupervised out-of-distribution detection using hierarchical variational autoencoders.
-We provide evidence that highly generalizeable low-level features contribute greatly to estimated likelihoods resulting in poor OOD detection performance.
+We provide evidence that highly generalizable low-level features contribute greatly to estimated likelihoods resulting in poor OOD detection performance.
 We proceed to develop a likelihood-ratio based score for OOD detection and define it to explicitly ensure that data must be in-distribution across all feature levels to be regarded in-distribution.
@@
 JF was further funded by the Novo Nordisk Foundation (grant no.\@ NNF20OC0065611) and the Independent Research Fund Denmark (grant no.\@ 9131-00082B).
-SH was further funded by VILLUM FONDEN (15334) and the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement no.
+SH was further funded by VILLUM FONDEN (15334) and the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement no.
 757360).

### 2023-10-04 chapters/paper_modelagnostic.tex
Work on font and fix typos

@@
 
-The ability to recognise when data are anomalous, i.e.
+The ability to recognize when data are anomalous, i.e.
 if they originate from a distribution different from that of the training data, is a necessary property for machine learning models for safe and reliable applications in the real world.
@@
 %While this does not solve completely the problem, because the power of the combined tests will also depend on the out-distribution, we think that we can get better performance than using a single-statistics especially in situations where one statistic fails.
-While the power of the combined test also depends on the out-distribution, we hypothesise that the combined test empirically will perform better, especially in situations where one of the statistics fails.
+While the power of the combined test also depends on the out-distribution, we hypothesize that the combined test empirically will perform better, especially in situations where one of the statistics fails.
 %
@@
 Since we obtain a $p$-value, it is more natural deciding on a threshold as this corresponds to the significance level.
-In addition to that, it also allow us to correct for the multiple comparisons problem when identifying outliers in a dataset by controlling the number of Type I errors through the false discovery rate (FDR).
+In addition to that, it also allows us to correct for the multiple comparisons problem when identifying outliers in a dataset by controlling the number of Type I errors through the false discovery rate (FDR).
 
@@
 We also found that using this kernel generally gives good results for OOD tasks.
-An interesting analysis that we d not consider in this paper would be to study the properties of this kernel.
+An interesting analysis that we'd not consider in this paper would be to study the properties of this kernel.
 
@@
 The goal is to combine these different tests into a single statistical test that ideally will perform better than the initial single tests.
-However, different tests can have different magnitudes and they can differ also in the direction of out-of-distribution detection, i.e.\@ for some statistics having a higher values is associated with being OOD, while for other smaller values are OOD.
+However, different tests can have different magnitudes, and they can differ also in the direction of out-of-distribution detection, i.e.\@ for some statistics having a higher value is associated with being OOD, while for other smaller values are OOD.
 This makes a combination non-trivial.
@@
 They only focused on the single-sample detection task, i.e.\@ $n=1$ following our problem formulation.
-Their idea is to fit different nonparametric density estimators, such as a kernel-density estimator (KDE) or a one-class support vector machines (SVM), for each different statistic $T_1,\dots, T_k$ by using the values computed on the training set examples.
+Their idea is to fit different non-parametric density estimators, such as a kernel-density estimator (KDE) or a one-class support vector machines (SVM), for each different statistic $T_1,\dots, T_k$ by using the values computed on the training set examples.
 For a single test example, $\tilde{x}_1$, they first compute $T_1,\dots, T_k$ and then combine those statistics by summing the different KDEs log-density.
@@
 A possible third option would be to estimate the diagonal of $I(\theta)$ using samples from the model.
-However, for autoregressive models as the PixelCNN, sampling is a sequential procedure and therefore it is computationally expensive to generate many samples when the input-space is high-dimensional.
+However, for autoregressive models as the PixelCNN, sampling is a sequential procedure, and therefore it is computationally expensive to generate many samples when the input-space is high-dimensional.
 For this reason, we do not consider it in this work.
@@
 While more efficient per-example gradient computations were proposed \parencite{goodfellow_efficient_2015, rochette_efficient_2019}, these techniques can only be applied on simple fully connected or convolutional networks.
-While for this paper we relied on the naive solution of looping through every example one at the time, a more efficient solution is provided by the BackPACK library \parencite{dangel_backpack_2020} which allows to compute the gradient with respect each sample in a minibatch.
+While for this paper we relied on the naive solution of looping through every example one at the time, a more efficient solution is provided by the BackPACK library \parencite{dangel_backpack_2020} which allows to compute the gradient with respect each sample in a mini-batch.
 
@@
 The typicality set is the subset of the model full support where the model samples from and this does not intersect with the region of higher likelihood.
-While the typicality test was introduced for batch-OOD detection, \textcite{morningstar_density_2021} shows that it also works well in the single-sample case.
+While the typicality test was introduced for batch OOD detection, \textcite{morningstar_density_2021} shows that it also works well in the single-sample case.
 This is also confirmed by our own experiments.
@@
 
-As mentioned in the introduction, we frame the OOD detection problem in terms of statistical tests problem.
+As mentioned in the introduction, we frame the OOD detection problem in terms of a statistical test problem.
 Recently, \textcite{haroush_statistical_2021} showed that adopting hypothesis testing at the layer and channel level of a neural network can be used for OOD detection in the discriminative setting.
@@
     \includegraphics[width=\textwidth]{graphics/paper_modelagnostic/pvalues_comb_no_drop_correct_font.pdf}
-    \caption[Distribution of the p-values of the typicality test, score statistic, Fisher's method and a combination of all.]{\emph{First plot}: $p$-values of the typicality test on the two test sets.
+    \caption[Distribution of the p-values of the typicality test, score statistic, Fisher's method and a combination of all.]{ \emph{First plot}: $p$-values of the typicality test on the two test sets.
 We can see that under $\mathcal{H}_0$, they should be uniformly distributed.
@@
 Results are summarized in \cref{tab_modelagnostic:two_sample_results}.
-As shown by \textcite{nalisnick_detecting_2019}, the typicality improves but also the score statistic gets better if we consider more samples.
+As shown by \textcite{nalisnick_detecting_2019}, the typicality improves, but also the score statistic gets better if we consider more samples.
 Combining those leads to an improvement of performance in terms of AUROC with almost all the models.
@@
 One of the advantages of framing the problem as multiple testing is that we have a well-defined procedure to decide on which hypotheses to reject while controlling the False Discovery Rate (FDR, \citealp{benjamini_controlling_1995}).
-Imagine we are interested in finding the outliers from the dataset given by the combination of the two test-sets but we do not want to discard too many inliers, then we can use the Benjamini-Hochberg (BH) procedure \parencite{benjamini_controlling_1995} to decide a threshold and reject all hypothesis below that threshold.
+Imagine we are interested in finding the outliers from the dataset given by the combination of the two test-sets, but we do not want to discard too many inliers, then we can use the Benjamini-Hochberg (BH) procedure \parencite{benjamini_controlling_1995} to decide a threshold and reject all hypothesis below that threshold.
 For a specific significance level $\alpha$, the procedure guarantees that the FDR stays below that level.
@@
 We leverage the fact that when the null hypothesis is true and the $p$-values are independent, then the scores obtained by combining $k$ different statistics are $\chi^2_{2k}$ distributed to compute the $p$-values.
-Alternatively, th e procedure can be also applied to the $p$-values of a single test-statistic.
+Alternatively, the procedure can be also applied to the $p$-values of a single test-statistic.
 %
-Usually, it is better to use a FDR control when it is actually possible to make few false discoveries, i.e.\@ when we have a strong statistic.
+Usually, it is better to use an FDR control when it is actually possible to make few false discoveries, i.e.\@ when we have a strong statistic.
 Therefore, we expect the procedure to work well when the AUROC is good, for examples on models trained on FashionMNIST.
@@
 As can be seen in \cref{fig_modelagnostic:type1}, we have that the Type I ratio line stays below the identity line, meaning that the BH correction is working.
-When deciding for a specific threshold $\alpha$, we usually have to trade-off between Type I and Type II error and in most cases the threshold to choose depends on the application domain.
+When deciding for a specific threshold $\alpha$, we usually have to trade off between Type I and Type II error and in most cases the threshold to choose depends on the application domain.
 Ideally, we would like to have a low Type I and a low Type II error rate, meaning that we are not considering a lot of in-distribution examples as OOD and at the same time considering a lot of outliers as in-distribution.
@@
 
-When considering only one-sided independent statistics, we showed that the Fisher's method tends to works better than combine them by summing the log-density of a KDE.
+When considering only one-sided independent statistics, we showed that the Fisher's method tends to work better than combine them by summing the log-density of a KDE.
 We also noticed that the score statistic tends to perform a bit worse when the number of parameters of the models increases, i.e.\@ in the context of natural images.
@@
 \section*{Acknowledgements}
-Federico Bergamin and Pierre-Alexandre Mattei contributed eqaully to this paper, which is indicated by the asterisk (*) in the author list.
+Federico Bergamin and Pierre-Alexandre Mattei contributed equally to this paper, which is indicated by the asterisk (*) in the author list.
 The work was supported by the Innovation Fund Denmark (0175-00014B and 0153-00167B), the Independent Research Fund Denmark (9131-00082B) and the Novo Nordisk Foundation (NNF20OC0062606 and NNF20OC0065611).
-Furthermore, it was supported by the French government, through the 3IA Côte d’Azur Investments in the Future project managed by the National Research Agency (ANR) with the reference number ANR-19-P3IA-0002.
+Furthermore, it was supported by the French government, through the 3IA Côte d'Azur Investments in the Future project managed by the National Research Agency (ANR) with the reference number ANR-19-P3IA-0002.
 

### 2023-10-04 chapters/discussion.tex
Updates to discussion

@@
 
-In \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}, we plot the calibration curve for the uncalibrated stroke recognition ensemble and the individual models along with a histogram of its predicted probabilities.
-The calibration curve is computed by sorting the predicted probabilities on test set into a number of bins spanning the range from zero to one.
+We compute the calibration curve for the stroke recognition ensemble by sorting the predicted probabilities on test set into a number of bins spanning the range from zero to one.
 For each bin, we compute the fraction of examples for which the model predicted correctly and then plot these pairs of mean predicted probabilities and correctness fractions.
 A perfectly calibrated model would have the same fraction of correct predictions in any given bin, as that bin's mean value.
-\Cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model} clearly shows the miscalibration issue that we previously discussed.
+In \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}, we plot the calibration curve for the uncalibrated stroke recognition ensemble and the individual models along with a histogram of its predicted probabilities.
+The miscalibration issue that we previously discussed is clearly visible as a strong overconfidence.
+
 It is interesting to note that the ensemble model is better calibrated than the individual models that form it.
@@
 We see that both methods result in quite good calibrations\footnote{Test set Brier scores: Uncalibrated = $0.003500$, logistic = $0.001807$, isotonic = $0.001774$.
-Brier skill score compared to uncalibrated (relative improvement): Logistic = $0.4830$, isotonic = $0.4924$.} and that the predicted probabilities are shifted towards smaller values.
+Relative improvement in Brier score compared to uncalibrated (Brier skill score): Logistic = $0.4830$, isotonic = $0.4924$.} and that the predicted probabilities are shifted towards smaller values.
 
@@
         Calibration curve for the uncalibrated stroke recognition ensemble model (left) and the histogram of predicted probabilities (right). 
-        We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis, fig_retrospective:figure1-roc-curve}.
-    }
+        We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
     \label{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}
@@
         Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right). 
-        We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis, fig_retrospective:figure1-roc-curve}.
-    }
+        We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
     \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}

### 2023-10-05 chapters/discussion.tex
Revise abstract

@@
     \end{subfigure}
-    \caption[Calibration fits and curves for the stroke recognition model using Platt-scaling and isotonic regression for calibration.]{%
-        Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right). 
+    \caption[Calibration fits and curves for the stroke recognition model using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right).
 We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
@@
 
-% \subsectionP{Ensembling technique}
-% To form an ensemble from several models, we need a method to form a single prediction from the outputs of the individual models.
-% To allow us to evaluate an impact score for each word feature in our ablation study, we wanted our ensemble to provide not only predictions but also associated probability scores on a continuous scale.
-A majority voting scheme would only provide as many different scores as models, in our case five, which is arguably too coarse a resolution.
-
-\lesstodo[inline]{Discuss the calibration of the stroke model and report calibration curve.}
-\lesstodo[inline]{Report the results of calibrating the stroke model using e.g. Platt scaling or Isotonic scaling.} % https://scikit-learn.org/stable/modules/calibration.html
+% \lesstodo[inline]{Discuss the calibration of the stroke model and report calibration curve.}
+% \lesstodo[inline]{Report the results of calibrating the stroke model using e.g. Platt scaling or Isotonic scaling.} % https://scikit-learn.org/stable/modules/calibration.html
 

### 2023-10-05 chapters/introduction.tex
Revise abstract

@@
 
-\section{Motivation and Corti's use-cases}
+\section{Uncertainty by example: Corti use-cases}
 %
@@
 %
-\paragraph{Motivation} In 2021, Corti entered a research collaboration with the Capital Region of Denmark and the Copenhagen Emergency Services to develop a system for recognizing stroke cases in calls to the 1-1-2 emergency line and the 1813 medical helpline.
+\paragraph{Case background} In 2021, Corti entered a research collaboration with the Capital Region of Denmark and the Copenhagen Emergency Services to develop a system for recognizing stroke cases in calls to the 1-1-2 emergency line and the 1813 medical helpline.
 Stroke is one of the biggest causes of disability and death worldwide \parencite{cite1,cite2,cite3} and effective treatment is highly time-sensitive \parencite{cite4,cite5}.
@@
 
-\paragraph{Augmentation} A reasonable machine learning pipeline to assist in recognizing stroke cases on calls to emergency services might consist of an automatic speech recognition model that transcribes the conversation and a binary classifier that estimates the probability that the transcript describes a case of stroke.
+\paragraph{Uncertainty in stroke recognition} A reasonable machine learning pipeline to assist in recognizing stroke cases on calls to emergency services might consist of an automatic speech recognition model that transcribes the conversation and a binary classifier that estimates the probability that the transcript describes a case of stroke.
 Such a system might be trained on a dataset calls with verified stroke and non-stroke cases and evaluated against a held-out test set of calls where the call-taker has indicated whether they suspect a stroke.
@@
 %
-\paragraph{Motivation} During the course of this project, Corti's portfolio has expanded to include a software system for medical coding of clinical notes. 
+\paragraph{Case background} During the course of this project, Corti's portfolio has expanded to include a software system for medical coding of clinical notes. 
 When a patient is admitted to a hospital, the medical staff will write a clinical note describing the patient's diagnosis and the procedures they underwent, including drug prescriptions. 
@@
 
-\paragraph{Augmentation} A reasonable machine learning pipeline to assist in medical coding might consist of a natural language processing model that extracts relevant information from the clinical note and outputs a number of probability distributions that each estimate the probability that the note should be assigned one of the medical codes.
+\paragraph{Uncertainty in automated medical coding} A reasonable machine learning pipeline to assist in medical coding might consist of a natural language processing model that extracts relevant information from the clinical note and outputs a number of probability distributions that each estimate the probability that the note should be assigned one of the medical codes.
 However, training and using such as system comes with several challenges. 
@@
 The remainder of \cref{part:background} consists of two chapters. 
-\Cref{chp:main-contributions} presents the research questions of the thesis and the main contributions made in the included papers. 
+\Cref{chp:main-contributions} presents the research hypotheses and contributions of the thesis via the included papers. 
 \Cref{chp:technical-background} provides relevant technical background to the included papers, including uncertainty, out-of-distribution detection, and variational autoencoders.

### 2023-10-05 chapters/main-contributions.tex
Revise abstract

@@
 
-\chapter[research questions and main contributions]{Research questions and main contributions}\label{chp:main-contributions}
+\chapter[research hypotheses and contributions]{Research hypotheses and contributions}\label{chp:main-contributions}
 
@@
 
-The following chapters of the thesis are self-contained studies and therefore detail their own research questions and contributions, but do so without consideration to the other chapters. 
-For that reason, we here lay out the research questions and main contributions of each chapter in relation to the overall research project of the thesis. 
-The chapters are structured into three parts: {\color{black}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{black}\cref{part:self-supervised-speech-representation-learning}} is concerned with self-supervised speech representation learning, and {\color{black}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.
+The following chapters of the thesis are self-contained studies and therefore detail their own research hypotheses and contributions, but do so without consideration to the other chapters. 
+For that reason, we here lay out research hypotheses and contributions in relation to the overall research project. 
+% The chapters are structured into three parts: {\color{black}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{black}\cref{part:self-supervised-speech-representation-learning}} is concerned with self-supervised speech representation learning, and {\color{black}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.
 Finally, {\color{black}\cref{part:discussion-and-conclusion}} contains a discussion and the conclusion of the thesis.
@@
 
-\contribitem{chp:paper-retrospective}
-In this paper we examine the hypothesis that a machine learning framework can learn to recognize cases of stroke in calls made to a prehospital medical helpline. 
-We used calls from Copenhagen during 2015 to 2020, to develop a machine learning-based classification pipeline.
-First, calls were transcribed by a speech recognition model and then categorized as stroke or non-stroke using a text classification model.
-On test data from 2021, call-takers achieved an overall sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%) while the machine learning framework performed significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
-Effective stroke treatment depends on recognition by call-takers at prehospital telehealth services. 
-This study provides preliminary evidence that a machine learning framework for recognizing stroke in prehospital medical helpline calls could become a supportive tool for call-takers, aiding in early and accurate stroke recognition.
-% This paper aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.
-
 \contribitem{chp:paper-automated}
@@
 We also present the first comprehensive results on the newly released MIMIC-IV dataset using the reproduced models.
+
+\contribitem{chp:paper-retrospective}
+In this paper we examine the hypothesis that a machine learning framework can learn to recognize cases of stroke in calls made to a prehospital medical helpline. 
+We used calls from Copenhagen during 2015 to 2020, to develop a machine learning-based classification pipeline.
+First, calls were transcribed by a speech recognition model and then categorized as stroke or non-stroke using a text classification model.
+On test data from 2021, call-takers achieved an overall sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%) while the machine learning framework performed significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
+Effective stroke treatment depends on recognition by call-takers at prehospital telehealth services. 
+This study provides preliminary evidence that a machine learning framework for recognizing stroke in prehospital medical helpline calls could become a supportive tool for call-takers, aiding in early and accurate stroke recognition.
+% This paper aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.

### 2023-10-06 frontmatter/abstract-da.tex
Make project run in VS Code
Fix name highlighting and update font
Write more on the introduction
Updates to abstracts

@@
-%!TEX root = ../Thesis.tex
-\chapter{Resumé (Danish)}
-\blindtext[5]
+%!TEX root = ../thesis.tex
+% LTeX: enabled=false
+
+\chapter[resumé (danish)]{Resumé (Danish)}
+
+Naturligt sprog spiller en central rolle for patientinteraktioner i sundhedssystemer verden over. 
+På trods af betydelige fremskridt inden for medicin og medicinsk teknologi er patientinterviews og generel medicinsk kommunikation forblevet stort set uændret.
+Typisk har en enkelt sundhedsperson, hvad enten det er en opkaldstager på en akuttelefon, en paramediciner, en praktiserende læge eller en medicinsk koder, til opgave at forstå patientbehov og agere passende.
+Med aldrende befolkninger, stadigt flere behandlingsmuligheder, administrative byrder og strenge dokumentationskrav står sundhedspersonale over for et stigende pres i et stadigt mere komplekst miljø.
+Disse udfordringer udgør risici for personalefastholdelse, plejekvalitet og potentialet for medicinske fejl.
+
+Maskinlæring udgør en mulighed for at støtte sundhedspersonale i forskellige medicinske rammer, herunder primære klinikker og akuttjenester.
+Tidligere maskinlæringsapplikationer i sundhedsvæsenet har ofte forsømt naturligt sprog og fokuseret på opgaver, der involverer strukturerede data som at forudsige medicinske billeddannelsesresultater og analysere elektroniske sundhedsjournaler.
+Brugen af maskinlæringssystemer til at fortolke naturligt sprog kan dog lette de administrative- og dokumentationsbyrder for sundhedspersonale og hjælpe med at sikre, at kritisk information ikke overses eller glemmes.
+
+Udvikling af maskinlæringssystemer til medicinske samtaler er udfordrende på grund af den høje indvirkning som sundhedsbeslutninger har på patienters liv og kompleksiteten af ustrukturerede sprogdata, herunder tale og tekst.
+Mens maskinlæringssystemer baseret på neurale netværk har vist imponerende resultater på andre domæner, kan deres anvendelse i sundhedsvæsenet blive hindret af deres svært fortolkelige prædiktioner og deres dårlige evne til at kvantificere prædiktiv usikkerhed, som forstærkes i dynamiske miljøer og for sjældne tilfælde, almindeligvist omtalt som ude-af-fordeling data.
+
+Denne afhhandling udforsker maskinlæringsteknikker til medicinske samtaler med primært fokus på usikkerhedsvurdering og naturlig sprogbehandling.
+Den første del undersøger usuperviserede metoder til at detektere ude-af-fordeling data og modellering af tale ved hjælp af variationelle autoindkodere, og giver følgende bidrag:
+%
+\begin{enumerate}[label=(\roman*)] 
+    \item Empirisk og teoretisk evidens der viser, at simple dataegenskaber dominerer 
+    estimatet af likelihoodfunktionen for hierarkiske variationelle autoindkodere, hvilket hindrer detektion af data, der er ude-af-fordeling, og en ny likelihood-ratio score, der kræver at data er indenfor fordeling på tværs af alle dataegenskaber, simple såvel som komplekse.
+    \item En model-agnostisk statistisk test der bruger Fishers metode til at kombinere Rao's score test og den nyligt introducerede typikalitets test og er anvendelig med likelihoods fra enhver differentiabel generativ model.
+    \item Demonstrerede forbedringer i likelihood på taledata ved brug af hierarkiske variationelle autoindkodere sammenlignet med ikke-hierarkiske varianter samt latente rum, der repræsenterer information relevant for tale, såsom fonetisk indhold.
+\end{enumerate}
+%
+Anden del af denne afhandling udforsker selv-superviserede metoder til talerepræsentation og -forståelse og bidrager med:
+%
+\begin{enumerate}[label=(\roman*)] 
+    \item En omfattende litteraturgennemgang af selv-superviseret repræsentationslæring for tale der kategoriserer metoder i kontrastive, autoregressive og generative tilgange, giver en sammenligning af træningsobjektiver, modelarkitekturer og præstationer på downstream opgaver, og diskuterer lovende retninger for fremtidig forksning.
+\end{enumerate}
+%
+Den sidste del fokuserer på medicinske anvendelser af maskinlæring og giver følgende bidrag:
+%
+\begin{enumerate}[label=(\roman*)] 
+    \item En analyse af state-of-the-art modeller for automatiseret medicinsk kodning på MIMIC-III og IV datasættene.
+Denne analyse identificerer suboptimal træningspraksis, evalueringsmetoder og ikke-stratificerede dataopdelinger som faktorer, der bidrager til dårlig præstation.
+Den præsenterer efterfølgende en revideret sammenligning ved hjælp af forbedrede dataopdelinger og identiske eksperimentelle opsætninger der kaster lys over metodernes sande komparative ydeevne.
+    \item Et retrospektivt studie der demonstrerer at et lyd-basereet maskinlærings system betydeligt kan forbedre genkendelsen af stroke i opkald til den præhospitale akuttelefon 1813, der dækker Københavnsområdet.
+\end{enumerate}
+%
+Denne afhandling fremmer forskning i maskinlæring indenfor sundhedsvæsenet ved at adressere udfordringer inden for naturlig sprogbehandling, usikkerhedsvurdering og medicinske anvendelser, hvilket i sidste ende kan bidrage til forbedret patientbehandling og klinisk beslutningstagning.
+
+% LTeX: enabled=true

### 2023-10-06 appendices/paper_brief.tex
Improve hyphenation and add "= for compound words

@@
 We review the development of unsupervised representation learning for speech over the last decade.
-We identify two primary model categories: self-supervised methods and probabilistic latent variable models.
+We identify two primary model categories: self"=supervised methods and probabilistic latent variable models.
 We describe the models and develop a comprehensive taxonomy.
@@
 In computer vision, both supervised \parencite{simonyan_very_2014, szegedy_going_2015, he_deep_2015} and unsupervised \parencite{pathak_context_2016, doersch_unsupervised_2015} representation learning have gained attention with supervised representation learning driven by the availability of large annotated datasets \parencite{deng_imagenet_2009}.
-For text and speech, pre-training is usually unsupervised as labeled data is difficult to obtain.
+For text and speech, pre"=training is usually unsupervised as labeled data is difficult to obtain.
 Although work on text has paved the way, and the two fields share many characteristics, learning representations from speech is a problem faced with a unique set of challenges.
@@
 In this paper, we survey work on unsupervised representation learning for speech processing from within the last decade.
-From a methodological perspective, we identify two primary model categories, namely models based on self-supervised learning and probabilistic latent variable models. 
+From a methodological perspective, we identify two primary model categories, namely models based on self"=supervised learning and probabilistic latent variable models. 
 We provide a methodological review of the design choices related to each of the model categories and develop a model taxonomy that highlights the different directions of work.
@@
 
-\begin{figure}[!t]
+\begin{figure}
     \begin{flushleft}
-        \textsc{\small \hspace{0.16\textwidth} self-supervised \hspace{0.22\textwidth} latent variable}
+        \textsc{\small \hspace{0.16\textwidth} self"=supervised \hspace{0.22\textwidth} latent variable}
     \end{flushleft}
@@
     \includegraphics[width=0.90\textwidth]{paper_brief/SSL_LVM.pdf}
-    \caption[ schematic overview of the self-supervised and probabilistic latent variable models covered in this survey.]{A schematic overview of the two groups of models covered in this survey.
-\textit{Left}: A model trained with self-supervised learning.
+    \caption[ schematic overview of the self"=supervised and probabilistic latent variable models covered in this survey.]{A schematic overview of the two groups of models covered in this survey.
+\textit{Left}: A model trained with self"=supervised learning.
 We take these models to consist of two functions $f(\cdot)$ and $g(\cdot)$ (\cref{ssec:notation}).
-After pre-training, $f(\cdot)$ is fine-tuned or used for extracting features $\mathbf{c}$.
-$g(\cdot)$ is an auxiliary function used to accommodate the self-supervised pre-training task.
+After pre"=training, $f(\cdot)$ is fine-tuned or used for extracting features $\mathbf{c}$.
+$g(\cdot)$ is an auxiliary function used to accommodate the self"=supervised pre"=training task.
 \textit{Right}: A probabilistic latent variable model.
-In contrast to the self-supervised model, the functions $f(\cdot)$ and $g(\cdot)$ learn the parameters of distributions $q$ and $p$.
+In contrast to the self"=supervised model, the functions $f(\cdot)$ and $g(\cdot)$ learn the parameters of distributions $q$ and $p$.
 The latent variable $\mathbf{z}$ is commonly used for representation learning.}
@@
 
-In the following, we group previous work into \textit{self-supervised models} and \textit{probabilistic latent variable models}, and take a \emph{model} to comprise a neural architecture and a corresponding learning algorithm.
+In the following, we group previous work into \textit{self"=supervised models} and \textit{probabilistic latent variable models}, and take a \emph{model} to comprise a neural architecture and a corresponding learning algorithm.
 A schematic overview is found in \cref{fig:ssl_lvm}.
@@
 
-When discussing self-supervised models, we use $\mathbf{c}_{1:T}$ to denote a contextualized representation.
+When discussing self"=supervised models, we use $\mathbf{c}_{1:T}$ to denote a contextualized representation.
 For stochastic latent variable models, we use $\mathbf{z}_{1:T}$ as is customary to the field.
@@
 
-\subsection{Self-supervised models}
+\subsection{Self"=supervised models}
 \label{sec:ssmodels}
 
-Self-supervised learning is a subset of unsupervised learning \parencite{tsai_selfsupervised_2021}.
-Where other unsupervised methods can be seen as a means to an end in itself (e.g., clustering or data generation), self-supervised learning takes the form of a pretext task that only adds value when associated with a downstream task.
-This makes self-supervised learning tie naturally with semi-supervised learning, but it may also be part of a fully unsupervised setup \parencite{baevski_unsupervised_2021}.
-Self-supervised learning is often characterized by automatically deriving the target from the input or other unlabeled examples \parencite{ouali_overview_2020}.
+Self"=supervised learning is a subset of unsupervised learning \parencite{tsai_selfsupervised_2021}.
+Where other unsupervised methods can be seen as a means to an end in itself (e.g., clustering or data generation), self"=supervised learning takes the form of a pretext task that only adds value when associated with a downstream task.
+This makes self"=supervised learning tie naturally with semi"=supervised learning, but it may also be part of a fully unsupervised setup \parencite{baevski_unsupervised_2021}.
+Self"=supervised learning is often characterized by automatically deriving the target from the input or other unlabeled examples \parencite{ouali_overview_2020}.
 
@@
 
-\noindent Non-autoregressive predictive coding (NPC, \citealp{liu_nonautoregressive_2020}) tries to resolve this by using a convolutional neural network where the kernel is masked instead of the input.
+\noindent Non"=autoregressive predictive coding (NPC, \citealp{liu_nonautoregressive_2020}) tries to resolve this by using a convolutional neural network where the kernel is masked instead of the input.
 This allows for complete data utilization, but limits the amount of context encoded in the learned representation.
@@
 
-\begin{figure}[!t]
+\begin{figure}
     \centering
@@
     \end{tabular}
-    \caption{Schematic of self-supervised methods. Each subfigure illustrates the loss computation for a single time-step. The temporal subscript has been left out for simplicity.}
+    \caption{Schematic of self"=supervised methods. Each subfigure illustrates the loss computation for a single time-step. The temporal subscript has been left out for simplicity.}
     \label{fig:ssl_grid}
@@
 
-\begin{sidewaystable*}[t]
-    \caption[Classification of selected self-supervised and probabilistic latent variable models.]{Selected models classified according to the binary attributes identified throughout the text.
+\begin{sidewaystable*}
+    \caption[Classification of selected self"=supervised and probabilistic latent variable models.]{ Selected models classified according to the binary attributes identified throughout the text.
 The models are sorted according to first publication date on arXiv which might differ from the citation year.
@@
         \renewcommand{\arraystretch}{1.1}
-        \resizebox{0.925\textwidth}{!}{%
+        \resizebox{1\textwidth}{!}{%
         \begin{tabular}{ l l | c c c c c c | c c c | c c } 
@@
             \midrule
-            \multicolumn{12}{c}{\textsc{Self-supervised models}} \\
+            \multicolumn{12}{c}{\textsc{Self"=supervised models}} \\
             \midrule
             %                                                           MSK      PRD      CON      REC      QTZ      GEN      LOC      GLB     VAR      FRZ      FTN
-            \textbf{Audio Word2vec} \footnotesize\parencite{chung_audio_2016}      & 2016 Mar. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
-            \textbf{Speech2Vec} \footnotesize\parencite{chung_speech2vec_2018}     & 2018 Mar. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
-            \textbf{Unspeech} \footnotesize\parencite{milde_unspeech_2018}         & 2018 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{Audio Word2vec} \footnotesize{\parencite{chung_audio_2016}}      & 2016 Mar. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{Speech2Vec} \footnotesize{\parencite{chung_speech2vec_2018}}     & 2018 Mar. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{Unspeech} \footnotesize{\parencite{milde_unspeech_2018}}         & 2018 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
 
-            \textbf{CPC} (van den Oord et al. 2018)         & 2018 Jul. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{CPC} \footnotesize{\parencite{oord_representation_2018}}         & 2018 Jul. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
 
-            \textbf{APC} \footnotesize\parencite{chung_unsupervised_2019}          & 2019 Oct.
+            \textbf{APC} \footnotesize{\parencite{chung_unsupervised_2019}}          & 2019 Oct.
 & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 5/4
-            \textbf{wav2vec} \footnotesize\parencite{schneider_wav2vec_2019}       & 2019 Apr.
+            \textbf{wav2vec} \footnotesize{\parencite{schneider_wav2vec_2019}}       & 2019 Apr.
 & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/4
-            \textbf{Mockingjay}  \footnotesize\parencite{liu_mockingjay_2020}      & 2019 Oct.
+            \textbf{Mockingjay}  \footnotesize{\parencite{liu_mockingjay_2020}}      & 2019 Oct.
 & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 25/10-19
-            \textbf{wav2vec 2.0} \footnotesize\parencite{baevski_wav2vec_2020}     & 2020 Jun.
+            \textbf{wav2vec 2.0} \footnotesize{\parencite{baevski_wav2vec_2020}}     & 2020 Jun.
 & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 20/6
-            \textbf{NPC} \footnotesize\parencite{liu_nonautoregressive_2020}                     & 2020 Nov.
+            \textbf{NPC} \footnotesize{\parencite{liu_nonautoregressive_2020}}       & 2020 Nov.
 & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 1/11
-            \textbf{DeCoAR 2.0} \footnotesize\parencite{ling_decoar_2020}          & 2020 Dec.
+            \textbf{DeCoAR 2.0} \footnotesize{\parencite{ling_decoar_2020}}          & 2020 Dec.
 & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/12
-            \textbf{SCPC} \footnotesize\parencite{bhati_segmental_2021}            & 2021 Jun.
+            \textbf{SCPC} \footnotesize{\parencite{bhati_segmental_2021}}            & 2021 Jun.
 & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ % 3/6
-            \textbf{HuBERT} \footnotesize\parencite{hsu_hubert_2021}               & 2021 Jun.
+            \textbf{HuBERT} \footnotesize{\parencite{hsu_hubert_2021}}               & 2021 Jun.
 & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 14/6
@@
             %                                                           MSK      PRD      CON      REC      QTZ      GEN      LOC      GLO     VAR      FRZ      FTN
-            \textbf{VRNN} \footnotesize\parencite{chung_recurrent_2015}          & 2015 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-            \textbf{SRNN} \footnotesize\parencite{fraccaro_sequential_2016}      & 2016 May  & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-            \textbf{HMM-VAE} \footnotesize\parencite{ebbers_hidden_2017}         & 2017 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-            \textbf{ConvVAE} \footnotesize\parencite{hsu_learning_2017}          & 2017 Apr. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
-            \textbf{FHVAE} \footnotesize\parencite{hsu_unsupervised_2017}        & 2017 Sep. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
-            \textbf{VQ-VAE} (van den Oord et al. 2018)                      & 2017 Nov. & \xmark & \xmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-            \textbf{BHMM-VAE} \footnotesize\parencite{glarner_full_2018}         & 2018 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-            \textbf{STCN} \footnotesize\parencite{aksan_stcn_2019}               & 2019 Feb. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
-            \textbf{FDMM} \footnotesize\parencite{khurana_factorial_2019}        & 2019 Oct. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
-            \textbf{ConvDMM} \footnotesize\parencite{khurana_convolutional_2020} & 2020 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{VRNN} \footnotesize{\parencite{chung_recurrent_2015}}          & 2015 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{SRNN} \footnotesize{\parencite{fraccaro_sequential_2016}}      & 2016 May  & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{HMM-VAE} \footnotesize{\parencite{ebbers_hidden_2017}}         & 2017 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{ConvVAE} \footnotesize{\parencite{hsu_learning_2017}}          & 2017 Apr. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{FHVAE} \footnotesize{\parencite{hsu_unsupervised_2017}}        & 2017 Sep. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{VQ-VAE} \footnotesize{\parencite{oord_neural_2018}}            & 2017 Nov. & \xmark & \xmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{BHMM-VAE} \footnotesize{\parencite{glarner_full_2018}}         & 2018 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{STCN} \footnotesize{\parencite{aksan_stcn_2019}}               & 2019 Feb. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
+            \textbf{FDMM} \footnotesize{\parencite{khurana_factorial_2019}}        & 2019 Oct. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
+            \textbf{ConvDMM} \footnotesize{\parencite{khurana_convolutional_2020}} & 2020 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
             \bottomrule
@@
 where $\text{sg}[\mathbf{x}] = \mathbf{x}$ is the stop-gradient operator with the property $\frac{d}{dx_i}\text{sg}[\mathbf{x}] \equiv 0$ for all $i$ and $\beta$ is a hyperparameter.
-Although vector quantization was introduced by the VQ-VAE which is, in some ways, a latent variable model, it has been applied to self-supervised methods \parencite{vanniekerk_vectorquantized_2020, baevski_vqwav2vec_2020}.
+Although vector quantization was introduced by the VQ-VAE which is, in some ways, a latent variable model, it has been applied to self"=supervised methods \parencite{vanniekerk_vectorquantized_2020, baevski_vqwav2vec_2020}.
 
@@
 \textcite{chung_audio_2016} propose a simple sequence-to-sequence autoencoder for learning acoustic word embeddings:
-\begin{align}
+\begin{align}\label{eq_brief:dec-aw2v}
     \mathbf{c} &= f(\mathbf{x}_{1:T}) \\
-    \hat{\mathbf{x}}_{1:T} &= g(\mathbf{c}) \enspace , \label{eq_brief:dec-aw2v}
+    \hat{\mathbf{x}}_{1:T} &= g(\mathbf{c}) \enspace ,
 \end{align}
 
-\noindent where $f(\cdot)$ and $g(\cdot)$ are a recurrent neural networks, such that $\mathbf{c}$ is taken to be the hidden state at the last time-step $T$ of $f(\cdot)$ and used as initial hidden state of $g(\cdot)$.
+\noindent where $f(\cdot)$ and $g(\cdot)$ are recurrent neural networks, such that $\mathbf{c}$ is taken to be the hidden state at the last time-step $T$ of $f(\cdot)$ and used as initial hidden state of $g(\cdot)$.
 The authors also propose a denoising autoencoder with masked inputs $f(\mathbf{x}_{1:T} \circ \mathbf{m}_{1:T})$.
@@
 
-Inspired by the work on semantic word embeddings for text \parencite{mikolov_distributed_2013}, the sequence-to-sequence framework has also been used to implement speech-based versions of the skipgram and continuous bag-of-words models \parencite{chung_learning_2017, chung_speech2vec_2018}.
-Given a segment corresponding to a single word $\mathbf{x}_{(n)} = \mathbf{x}_{t_{n}:t_{n+1}}$, the skipgram model is trained to predict neighboring words $\mathbf{x}_{(n + k)}$ where $k\ne0$.
-That is, instead of a single decoder, as in \cref{eq_brief:dec-aw2v}, the skipgram model employs multiple decoders
+Inspired by the work on semantic word embeddings for text \parencite{mikolov_distributed_2013}, the sequence-to-sequence framework has also been used to implement speech-based versions of the skip-gram and continuous bag-of-words models \parencite{chung_learning_2017, chung_speech2vec_2018}.
+Given a segment corresponding to a single word $\mathbf{x}_{(n)} = \mathbf{x}_{t_{n}:t_{n+1}}$, the skip-gram model is trained to predict neighboring words $\mathbf{x}_{(n + k)}$ where $k\ne0$.
+That is, instead of a single decoder, as in \cref{eq_brief:dec-aw2v}, the skip-gram model employs multiple decoders
 %
@@
 Most of the work presented so far fits neatly into the taxonomy presented in \cref{tab:model-taxonomy}.
-One exception is the problem-agnostic speech encoder (PASE, \citealp{pascual_learning_2019, ravanelli_multitask_2020}) that combines multiple pre-training tasks.
+One exception is the problem-agnostic speech encoder (PASE, \citealp{pascual_learning_2019, ravanelli_multitask_2020}) that combines multiple pre"=training tasks.
 Furthermore, many of the presented models have been successfully applied to other use cases.
-For instance, wav2vec 2.0 and related models have been applied to learn cross-lingual and multi-lingual representations \parencite{riviere_unsupervised_2020, conneau_unsupervised_2020, khurana_magic_2021} and proven well-suited for concurrently learning with labeled data \parencite{talnikar_joint_2021, wang_unispeech_2021}.
-%We will discuss the downstream application of self-supervised representations further in \cref{sec:eval}.
+For instance, wav2vec 2.0 and related models have been applied to learn crosslingual and multilingual representations \parencite{riviere_unsupervised_2020, conneau_unsupervised_2020, khurana_magic_2021} and proven well-suited for concurrently learning with labeled data \parencite{talnikar_joint_2021, wang_unispeech_2021}.
+%We will discuss the downstream application of self"=supervised representations further in \cref{sec:eval}.
 
@@
 We then introduce the variational autoencoder framework (VAE, \citealp{kingma_autoencoding_2014}).
-We focus on different dependency structures between data and learned representations, in contrast to the more practical view on self-supervised models taken above.
+We focus on different dependency structures between data and learned representations, in contrast to the more practical view on self"=supervised models taken above.
 
@@
 The bound can be efficiently evaluated and optimized with Monte Carlo (MC) estimation by sampling from $q(\mathbf{z}|\mathbf{x})$.
-Low-variance gradient estimates are usually obtained via reparameterization of $q(\mathbf{z}|\mathbf{x})$ \parencite{kingma_autoencoding_2014} although alternatives exist (e.g., inverse CDF sampling) \parencite{mohamed_monte_2019}.
+Low-variance gradient estimates are usually obtained via reparameterization of $q(\mathbf{z}|\mathbf{x})$ \parencite{kingma_autoencoding_2014}, although alternatives exist (e.g., inverse CDF sampling) \parencite{mohamed_monte_2019}.
 The ELBO can also be written as
@@
 
-\begin{table}[t!]
+\begin{table}
     \caption[A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable.]{A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable.
@@
 
-\begin{sidewaystable*}[t]
-    \caption[Classification of selected latent variable models.]{Selected latent variable models classified according the attributes defined throughout \cref{sec:plvms}.
+\begin{sidewaystable*}
+    \caption[Classification of selected latent variable models.]{ Selected latent variable models classified according the attributes defined throughout \cref{sec:plvms}.
 See \cref{tab:lvm-model-primitives} for the probability distributions that correspond to each of the attribute short-hands.
@@
             %                                                         ARX      LOC      GLB      ARX      ARZ     IND      GLB      ARZ      FLT      LSM      GSM      GLB
-            \textbf{VRNN} \footnotesize\parencite{chung_recurrent_2015}          & 2015 Jun.
+            \textbf{VRNN} \footnotesize{\parencite{chung_recurrent_2015}}          & 2015 Jun.
 & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark \\
-            \textbf{SRNN} \footnotesize\parencite{fraccaro_sequential_2016}      & 2016 May  & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark \\
-            \textbf{HMM-VAE} \footnotesize\parencite{ebbers_hidden_2017}         & 2017 Mar.
+            \textbf{SRNN} \footnotesize{\parencite{fraccaro_sequential_2016}}      & 2016 May  & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark \\
+            \textbf{HMM-VAE} \footnotesize{\parencite{ebbers_hidden_2017}}         & 2017 Mar.
 & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
-            \textbf{ConvVAE} \footnotesize\parencite{hsu_learning_2017}          & 2017 Apr.
+            \textbf{ConvVAE} \footnotesize{\parencite{hsu_learning_2017}}          & 2017 Apr.
 & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \cmark & \xmark \\
-            \textbf{FHVAE} \footnotesize\parencite{hsu_unsupervised_2017}        & 2017 Sep.
+            \textbf{FHVAE} \footnotesize{\parencite{hsu_unsupervised_2017}}        & 2017 Sep.
 & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \cmark & \cmark \\
-            \textbf{VQ-VAE} (van den Oord et al. 2018)            & 2017 Nov. & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\
-            \textbf{BHMM-VAE} \footnotesize\parencite{glarner_full_2018}         & 2018 Mar.
+            \textbf{VQ-VAE} \footnotesize{\parencite{oord_neural_2018}}            & 2017 Nov.
+& \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\
+            \textbf{BHMM-VAE} \footnotesize{\parencite{glarner_full_2018}}         & 2018 Mar.
 & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark \\
-            \textbf{STCN} \footnotesize\parencite{aksan_stcn_2019}               & 2019 Feb.
+            \textbf{STCN} \footnotesize{\parencite{aksan_stcn_2019}}               & 2019 Feb.
 & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
-            \textbf{FDMM} \footnotesize\parencite{khurana_factorial_2019}        & 2019 Oct.
+            \textbf{FDMM} \footnotesize{\parencite{khurana_factorial_2019}}        & 2019 Oct.
 & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark \\
-            \textbf{ConvDMM} \footnotesize\parencite{khurana_convolutional_2020} & 2020 Jun.
+            \textbf{ConvDMM} \footnotesize{\parencite{khurana_convolutional_2020}} & 2020 Jun.
 & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark \\
@@
 However, this forces the latent variable to encode details about the observed variable to achieve a good reconstruction.
-This is opposite to contrastive self-supervised learning which allows models to discard details in $\mathbf{x}_{1:T}$ that do not inform the training objective \parencite{baevski_wav2vec_2020}.
+This is opposite to contrastive self"=supervised learning which allows models to discard details in $\mathbf{x}_{1:T}$ that do not inform the training objective \parencite{baevski_wav2vec_2020}.
 
@@
 Smoothing can also be done locally (\textbf{\textsc{lsm}}, \cref{tab:lvm-model-primitives}), where the latent variables then depend on $\mathbf{x}_{t-r:t+r}$ for some integer $r>0$ \parencite{oord_neural_2018}.
-Compared to self-supervised models that often use transformer encoders it can be hypothesized that global smoothing offers a stronger case than local smoothing and filtering for representation learning.
+Compared to self"=supervised models that often use transformer encoders it can be hypothesized that global smoothing offers a stronger case than local smoothing and filtering for representation learning.
 
@@
 \paragraph{From global to local} 
-In \cref{tab:model-taxonomy}, we see that work on global representations within self-supervised learning precedes work on local representations.
+In \cref{tab:model-taxonomy}, we see that work on global representations within self"=supervised learning precedes work on local representations.
 However, we find that the core ideas underlying the recent successes in learning local representation models have also been used for global representation learning; masking \parencite{chung_audio_2016}, context prediction \parencite{chung_speech2vec_2018}, and contrastive training \parencite{milde_unspeech_2018} have been applied in both settings.
@@
 \paragraph{Representations beyond inference} 
-Predictive tasks are commonly used for self-supervised models, but they are not directly compatible with LVM training. 
+Predictive tasks are commonly used for self"=supervised models, but they are not directly compatible with LVM training. 
 However, an LVM prior with an autoregressive parameterization, $p(\mathbf{z}_t|\mathbf{z}_{1:t-1})$ or $p(\mathbf{z}_t|\mathbf{x}_{1:t-1})$, can be seen as predictive in the sense that it tries to match the approximate posterior.
@@
 The idea of using VAEs to impute missing data was already examined in the seminal paper by \textcite{rezende_stochastic_2014}.
-Here the model was trained with fully observed data and used to impute data in an iterative sampling approach post hoc leaving the learned representations unchanged.
+Here the model was trained with fully observed data and used to impute data in an iterative sampling approach post hoc, leaving the learned representations unchanged.
 
@@
 Although this review has a primarily methodological focus, we should briefly touch upon evaluation procedures.
-Training metrics for self-supervised tasks and the likelihood of LVMs offer little guidance as to the quality of the learned representations \parencite{huszar_is_2017}.
+Training metrics for self"=supervised tasks and the likelihood of LVMs offer little guidance as to the quality of the learned representations \parencite{huszar_is_2017}.
 Thus, a common approach is to evaluate the representations in terms of their usefulness for downstream tasks.
@@
 This focus has gained traction over the last couple of years, as computational resources have become more accessible and end-to-end models \parencite{graves_connectionist_2006, chan_listen_2016} have been established as the dominant approach to speech recognition \parencite{gulati_conformer_2020}.
-It is important to stress that self-supervised models, such as wav2vec 2.0 \parencite{baevski_wav2vec_2020}, represent a breakthrough, and recent successful approaches build upon this method.
+It is important to stress that self"=supervised models, such as wav2vec 2.0 \parencite{baevski_wav2vec_2020}, represent a breakthrough, and recent successful approaches build upon this method.
 That is, deep self-attention models combined with masking \parencite{hsu_hubert_2021, wang_unispeech_2021, chen_wavlm_2021}.
-This development mirrors years of rapid progress in masked language modeling within natural language processing \parencite{devlin_bert_2018,clark_electra_2020} and we expect this to continue for unsupervised neural speech representation learning.
+This development mirrors years of rapid progress in masked language modeling within natural language processing \parencite{devlin_bert_2018,clark_electra_2020}, and we expect this to continue for unsupervised neural speech representation learning.
 
@@
 \label{sec:conc}
-We reviewed unsupervised representation learning for speech, focusing on two primary categories: self-supervised methods and probabilistic latent variable models.
-Inspired by the development of self-supervised learning and the dependency structures of latent variable models, we derived a comprehensive model taxonomy.
+We reviewed unsupervised representation learning for speech, focusing on two primary categories: self"=supervised methods and probabilistic latent variable models.
+Inspired by the development of self"=supervised learning and the dependency structures of latent variable models, we derived a comprehensive model taxonomy.
 Finally, we compare and discuss models from the two categories and their respective evaluation procedures.

### 2023-10-06 appendices/supplementary_benchmarking.tex
Improve hyphenation and add "= for compound words

@@
 \paragraph{Phoneme recognition}
-The ASR experiment consists of two stages: 1) pre-training of the unsupervised model and 2) training of the ASR model.
-The pre-training is done as for the likelihood benchmark above.
+The ASR experiment consists of two stages: 1) pre"=training of the unsupervised model and 2) training of the ASR model.
+The pre"=training is done as for the likelihood benchmark above.
 The ASR model is trained using the Adam optimizer \parencite{kingma_adam_2015} with default parameters as given in PyTorch.
@@
 
-While this is well-suited for autoregressive modeling, for which the distribution was developed, the potential multimodality poses a challenge for non-autoregressive latent variable models which independently sample multiple neighboring observations at the output.
+While this is well-suited for autoregressive modeling, for which the distribution was developed, the potential multimodality poses a challenge for non"=autoregressive latent variable models which independently sample multiple neighboring observations at the output.
 In fact, if multiple neighboring outputs defined by the subsequence $\xb_{t_1:t_2}$ have multimodal $p(\xb_t|\cdot)$, we risk sampling a subsequence where each neighboring value expresses different potential realities, independently.
 
-Interestingly, most work on latent variable models with non-autoregressive output distributions seem to ignore this fact and simply employ the mixture distribution with 10 mixture components \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
+Interestingly, most work on latent variable models with non"=autoregressive output distributions seem to ignore this fact and simply employ the mixture distribution with 10 mixture components \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
 However, given the empirically good results of latent variable models for image generation, this seems to have posed only a minor problem in practice.
@@
 Furthermore, waveforms exhibit a natural symmetry between positive and negative amplitudes.
-Hence, it seems plausible that multimodality may pose a larger problem in non-autoregressive speech generation by causing locally incoherent samples than it seems to do in image modelling.
+Hence, it seems plausible that multimodality may pose a larger problem in non"=autoregressive speech generation by causing locally incoherent samples than it seems to do in image modelling.
 

### 2023-10-06 appendices/supplementary_benchmarking_workshop.tex
Improve hyphenation and add "= for compound words

@@
 
-While this is well-suited for autoregressive modeling, for which the distribution was developed, the potential multimodality poses a challenge for non-autoregressive latent variable models which independently sample multiple neighboring observations at the output.
+While this is well-suited for autoregressive modeling, for which the distribution was developed, the potential multimodality poses a challenge for non"=autoregressive latent variable models which independently sample multiple neighboring observations at the output.
 In fact, if multiple neighboring outputs defined by the subsequence $\vx_{t_1:t_2}$ have multimodal $p(\vx_t|\cdot)$, we risk sampling a subsequence where each neighboring value expresses different potential realities, independently.
 
-Interestingly, most work on latent variable models with non-autoregressive output distributions seem to ignore this fact and simply employ the mixture distribution with 10 mixture components \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
+Interestingly, most work on latent variable models with non"=autoregressive output distributions seem to ignore this fact and simply employ the mixture distribution with 10 mixture components \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
 However, given the empirically good results of latent variable models for image generation, this seems to have posed only a minor problem in practice.

### 2023-10-06 appendices/supplementary_retrospective.tex
Improve hyphenation and add "= for compound words

@@
 As described in the discussion section, we hypothesize that the number of stroke positives was too small for these advanced models to learn more complex patterns than the MLP ensemble.
-In addition, the BERT model would likely benefit from pre-training on text data from the target domain, or a domain close to it, rather than various online fora.
+In addition, the BERT model would likely benefit from pre"=training on text data from the target domain, or a domain close to it, rather than various online fora.
 

### 2023-10-06 chapters/background.tex
Improve hyphenation and add "= for compound words

@@
 Then, we describe variational autoencoders, how they can be used to learn representations of speech, and some of their challenges.
-Finally, we introduce the concept of self-supervised learning and highlight important differences to variational autoencoders.
+Finally, we introduce the concept of self"=supervised learning and highlight important differences to variational autoencoders.
 
@@
 This is similar to using simulated OOD inputs from generative adversarial networks \parencite{lee_training_2018} or using auxiliary datasets of outliers \parencite{hendrycks_deep_2019} for calibration purpose.
-%   - X Deep semi-supervised anomaly detection \parencite{ruff_deep_2020}.
+%   - X Deep semi"=supervised anomaly detection \parencite{ruff_deep_2020}.
 %   - Variational information bottleneck (VIB) (Alemi et al., 2018b) performs divergence estimation in latent space to detect OoD, but is technically a model-dependent technique because the latent code is trained jointly with the downstream classification task.
@@
 In a similar vein, \textcite{dhamija_reducing_2018} propose losses designed to maximize the entropy of $p(y|\xb)$ and decrease feature magnitudes for OOD data sampled from other datasets. 
-\textcite{ruff_deep_2020} use semi-supervised learning and learn representations of in-distribution data that concentrate close to a centroid in latent space while labelled outliers are pushed away from the centroid.
+\textcite{ruff_deep_2020} use semi"=supervised learning and learn representations of in-distribution data that concentrate close to a centroid in latent space while labelled outliers are pushed away from the centroid.
 Other methods including MCD \parencite{yu_unsupervised_2019}, NGC \parencite{wu_ngc_2021}, and UDG \parencite{yang_semantically_2021} utilize external, unlabeled, noisy data to improve OOD detection performance without requiring cleanly labelled OOD examples.
@@
 
-For high-dimensional data, the most successful unsupervised models for OOD detection are deep autoencoders \parencite{hinton_reducing_2006}, self-supervised methods \parencite{mikolov_efficient_2013,devlin_bert_2018,schneider_wav2vec_2019,chen_simple_2020}, and deep generative models \parencite{kingma_autoencoding_2014,rezende_stochastic_2014,dinh_nice_2015,rezende_variational_2015,ho_denoising_2020,hinton_fast_2006,oord_conditional_2016,goodfellow_generative_2014}.
+For high-dimensional data, the most successful unsupervised models for OOD detection are deep autoencoders \parencite{hinton_reducing_2006}, self"=supervised methods \parencite{mikolov_efficient_2013,devlin_bert_2018,schneider_wav2vec_2019,chen_simple_2020}, and deep generative models \parencite{kingma_autoencoding_2014,rezende_stochastic_2014,dinh_nice_2015,rezende_variational_2015,ho_denoising_2020,hinton_fast_2006,oord_conditional_2016,goodfellow_generative_2014}.
 Other important density estimation methods that have been applied to OOD detection include kernel density estimation \parencite{parzen_estimation_1962}, nearest neighbor methods \parencite{cover_nearest_1967}, support vector machines \parencite{cortes_supportvector_1995,scholkopf_estimating_2001}, and Gaussian mixture models \parencite{dempster_maximum_1977}.
@@
 %     X ahmadian_likelihoodfree_2021,         extracts low-level features from the encoder of an invertible generative model.
-%     X hendrycks_using_2019,                 learn a representation over the in-domain data through self-supervised training
-%     X bergman_classificationbased_2020,     learn a representation over the in-domain data through self-supervised training
-%     X tack_csi_2020,                        learn a representation over the in-domain data through self-supervised training
-%     X sehwag_ssd_2021,                      learn a representation over the in-domain data through self-supervised training
-%     X xiao_we_2021                          further shows that one can instead use a strong pretrained feature extractor (self-supervised)  while maintaining comparable performance.
+%     X hendrycks_using_2019,                 learn a representation over the in-domain data through self"=supervised training
+%     X bergman_classificationbased_2020,     learn a representation over the in-domain data through self"=supervised training
+%     X tack_csi_2020,                        learn a representation over the in-domain data through self"=supervised training
+%     X sehwag_ssd_2021,                      learn a representation over the in-domain data through self"=supervised training
+%     X xiao_we_2021                          further shows that one can instead use a strong pretrained feature extractor (self"=supervised)  while maintaining comparable performance.
 % }
@@
 They propose to measure the Mahalanobis distance between the global Gaussian distribution of training set in latent space and an encoded test input.
-\textcite{xiao_we_2021} instead propose to use an existing, strong foundation model, pre-trained with a self-supervised objective, to extract features from the input, and then fit a Gaussian Mixture Model to the features using the minimal Mahalanobis distance to the mixture components as the OOD score.
+\textcite{xiao_we_2021} instead propose to use an existing, strong foundation model, pre-trained with a self"=supervised objective, to extract features from the input, and then fit a Gaussian Mixture Model to the features using the minimal Mahalanobis distance to the mixture components as the OOD score.
 
-Several other works also use self-supervised representations of the in-domain data for OOD detection \parencite{hendrycks_using_2019,bergman_classificationbased_2020}. 
+Several other works also use self"=supervised representations of the in-domain data for OOD detection \parencite{hendrycks_using_2019,bergman_classificationbased_2020}. 
 \textcite{tack_csi_2020} propose to use a contrastive objective to learn representations of the in-domain data contrasted with data augmented in-samples and use a softmax classifier trained on the representations to compute the OOD score.
@@
 Despite the challenges facing the training of VAEs, they have been successfully applied to a number of tasks including image generation \parencite{kingma_autoencoding_2014,rezende_stochastic_2014}, image inpainting \parencite{pathak_context_2016}, image super-resolution \parencite{sonderby_amortised_2017,chira_image_2022}, and speech synthesis \parencite{hsu_unsupervised_2017,hsu_hierarchical_2019}.
-VAEs have also proven themselves useful for semi-supervised learning \parencite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}.
+VAEs have also proven themselves useful for semi"=supervised learning \parencite{kingma_semi"=supervised_2014,kingma_improved_2016,maaloe_biva_2019}.
 
-\paragraph{Self-supervised learning}
-An alternative approach to representation learning is provided by self-supervised learning \parencite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020,schneider_wav2vec_2019} which is a form of unsupervised learning where the training objective is derived from the data itself.
-While it is fair to say that not all self-supervised training objectives are as well-motivated as the ELBO is for VAEs, they have shown impressive results within the fields of natural language processing \parencite{devlin_bert_2018,chen_simple_2020}, speech processing \parencite{schneider_wav2vec_2019}, and computer vision \parencite{chen_simple_2020}.
-For a deeper comparison of VAEs and self-supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{app:paper-brief}. 
-\Cref{chp:paper-review} provides a comprehensive introduction to and review of self-supervised learning in speech recognition.
+\paragraph{Self"=supervised learning}
+An alternative approach to representation learning is provided by self"=supervised learning \parencite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020,schneider_wav2vec_2019} which is a form of unsupervised learning where the training objective is derived from the data itself.
+While it is fair to say that not all self"=supervised training objectives are as well-motivated as the ELBO is for VAEs, they have shown impressive results within the fields of natural language processing \parencite{devlin_bert_2018,chen_simple_2020}, speech processing \parencite{schneider_wav2vec_2019}, and computer vision \parencite{chen_simple_2020}.
+For a deeper comparison of VAEs and self"=supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{app:paper-brief}. 
+\Cref{chp:paper-review} provides a comprehensive introduction to and review of self"=supervised learning in speech recognition.

### 2023-10-06 chapters/discussion.tex
Improve hyphenation and add "= for compound words

@@
 % - paper-benchmarking
-%   - Inferiority of probabilistic methods compared to self-supervised learning. 
+%   - Inferiority of probabilistic methods compared to self"=supervised learning. 
 % -
+
+Though more than three years have passed since the start of this project, the rapid development of machine learning that started about a decade ago with the ImageNet competation and the work of \textcite{krizhevsky_imagenet_2012} did all but slow down.
+In this section we will briefly revisit selected chapters of the thesis and provide some additional thoughts and discussion
+
+\section{Speech representation learning}
+
+Masking in variational autoencoders
 
@@
 
-\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?
+\lesstodo[inline]{Are self"=supervised speech representations useful for unsupervised uncertainty estimation?
 \parencite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}
-% Since the main focus within self-supervised learning has been on improving downstream task performance, very limited work, if any, has investigated self-supervised representations in terms of uncertainty estimation.
+% Since the main focus within self"=supervised learning has been on improving downstream task performance, very limited work, if any, has investigated self"=supervised representations in terms of uncertainty estimation.
 % However, in the context of medical applications where data can be abundant but labels are sparse, unsupervised uncertainty estimation is a very interesting direction for future work.

### 2023-10-06 chapters/introduction.tex
Remove comments
Improve hyphenation and add "= for compound words

@@
 
-\Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. 
+\Cref{part:unsupervised-uncertainty-estimation,part:self"=supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. 
 \Cref{part:unsupervised-uncertainty-estimation} is made up of three papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
-\Cref{part:self-supervised-speech-representation-learning} is a single, comprehensive review paper on self-supervised speech representation learning \parencite{mohamed_selfsupervised_2022}. 
+\Cref{part:self"=supervised-speech-representation-learning} is a single, comprehensive review paper on self"=supervised speech representation learning \parencite{mohamed_selfsupervised_2022}. 
 \Cref{part:medical-applications} consists of two papers on applications of machine learning within the medical domain, specifically recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}.
@@
 Finally, \cref{part:discussion-and-conclusion} concludes the thesis by summarizing the main contributions and discussing the presented work and future directions for research.
-
-% \section{Thesis scope and outline}
-%
-% OUTLINE: 
-% - Tie together the different topics of the thesis and point briefly to the relevant papers.
-%
-% The complete set of research published as part of this project covers a number of topics related to decision-support in medical conversations. 
-% One group of the primary papers is concerned with generative latent variable models and their applications to estimation of epistemic uncertainty and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
-% Another group of studies examines self-supervised speech representation learning as an alternative to generative latent variable models \parencite{mohamed_selfsupervised_2022,borgholt_brief_2022} and explores them for speech recognition \parencite{borgholt_scaling_2021} and spoken language understanding \parencite{borgholt_we_2021}.
-% Finally, a group deals with applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}.
-%

### 2023-10-06 chapters/main-contributions.tex
Improve hyphenation and add "= for compound words

@@
 For that reason, we here lay out research hypotheses and contributions in relation to the overall research project. 
-% The chapters are structured into three parts: {\color{black}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{black}\cref{part:self-supervised-speech-representation-learning}} is concerned with self-supervised speech representation learning, and {\color{black}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.
+% The chapters are structured into three parts: {\color{black}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{black}\cref{part:self"=supervised-speech-representation-learning}} is concerned with self"=supervised speech representation learning, and {\color{black}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.
 Finally, {\color{black}\cref{part:discussion-and-conclusion}} contains a discussion and the conclusion of the thesis.
@@
 
-\contribitem{part:self-supervised-speech-representation-learning}
-This part of the thesis is concerned with self-supervised speech representation learning and consists of a single paper that provides a substantial review of the field \parencite{mohamed_selfsupervised_2022}.
-Self-supervised learning is a promising approach to speech representation learning that has shown a fast, wide and successful adoption across speech modelling over the course of the last few years. 
-As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self-supervised representations generally outperform representations learned via probabilistic generative models such as those investigated in \cref{part:unsupervised-uncertainty-estimation}.
-In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self-supervised learning for learning representations for unsupervised uncertainty estimation.
+\contribitem{part:self"=supervised-speech-representation-learning}
+This part of the thesis is concerned with self"=supervised speech representation learning and consists of a single paper that provides a substantial review of the field \parencite{mohamed_selfsupervised_2022}.
+Self"=supervised learning is a promising approach to speech representation learning that has shown a fast, wide and successful adoption across speech modelling over the course of the last few years. 
+As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self"=supervised representations generally outperform representations learned via probabilistic generative models such as those investigated in \cref{part:unsupervised-uncertainty-estimation}.
+In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self"=supervised learning for learning representations for unsupervised uncertainty estimation.
 We shall return to this in the discussion of \cref{chp:paper-review} in \cref{sec:discussion-paper-review}.
 An early version of this review paper included a review of variational autoencoders for speech representation learning and a comparison between the two modelling paradigms \parencite{borgholt_brief_2022}.
-Due to a desire to focus on self-supervised methods, the paper in \cref{chp:paper-review} does not put particular focus on variational methods. 
+Due to a desire to focus on self"=supervised methods, the paper in \cref{chp:paper-review} does not put particular focus on variational methods. 
 The early version is included in \cref{app:paper-brief} for reference.
@@
 \contribitem{chp:paper-review}
-This paper provides a comprehensive review of self-supervised speech representation learning.
+This paper provides a comprehensive review of self"=supervised speech representation learning.
 Previous work is grouped into three main categories of methods: contrastive, autoregressive, and generative, and the methods are compared in terms of their training objectives, model architectures, and performance on downstream tasks.

### 2023-10-06 chapters/paper_automated.tex
Improve hyphenation and add "= for compound words

@@
 More recently, pre-trained masked language models (e.g., BERT) have gained popularity~\parencite{tengReviewDeepNeural2022}.
-The MIMIC-III training set or PubMed articles are commonly used for pre-training.
+The MIMIC-III training set or PubMed articles are commonly used for pre"=training.
 
@@
 Furthermore, for AMC to become a viable solution for ICD-10, future research should focus on improving performance on rare codes while, in the shorter term, developing methods to detect codes that are too challenging for automated coding and, therefore, should be coded manually.
-Finally, while PLM-ICD outperforms the other models in this paper, the improvements are limited compared to the effect of pre-training in other domains \parencite{mohamed_selfsupervised_2022, linPretrainedTransformersText2021,baevski_wav2vec_2020,devlin_bert_2018,dosovitskiy_image_2021}.
+Finally, while PLM-ICD outperforms the other models in this paper, the improvements are limited compared to the effect of pre"=training in other domains \parencite{mohamed_selfsupervised_2022, linPretrainedTransformersText2021,baevski_wav2vec_2020,devlin_bert_2018,dosovitskiy_image_2021}.
 Notably, there have been several unsuccessful attempts at using pre-trained transformers for medical coding \parencite{jiDoesMagicBERT2021,gaoLimitationsTransformersClinical2021,michalopoulosICDBigBirdContextualEmbedding2022,pascualBERTbasedAutomaticICD2021,zhangBERTXMLLargeScale2020}.

### 2023-10-06 chapters/paper_benchmarking.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses
Improve hyphenation and add "= for compound words

@@
 Interestingly, we find that a standard LSTM \parencite{hochreiter_long_1997} almost matches the likelihood of WaveNet.
-    \item [(II)] LVMs with powerful autoregressive decoders achieve better likelihoods than the non-autoregressive LVM.
+    \item [(II)] LVMs with powerful autoregressive decoders achieve better likelihoods than the non"=autoregressive LVM.
     \item[(III)] The expressiveness of LVMs for speech increases with a deeper hierarchy of stochastic latent variables, similar to conclusions within image modeling.
@@
 Approximate posteriors have various applications e.g.
-semi-supervised learning \parencite{kingma_semi-supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.
+semi"=supervised learning \parencite{kingma_semi"=supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.
 
@@
 
-\begin{figure*}[t]
+\begin{figure*}
     \centering
@@
 %
-\begin{table}[t]
+\begin{table}
     \caption[Model likelihoods and phoneme error rate for TIMIT.]{We report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines (left) and phoneme error rate (PER) of different representations for phoneme recognition on TIMIT (right).}
@@
 This matches recent observations in the image domain.
-We noted that the STCN with an autoregressive decoder outperforms the non-autoregressive CW-VAE, which we adapted to speech. 
+We noted that the STCN with an autoregressive decoder outperforms the non"=autoregressive CW-VAE, which we adapted to speech. 
 Finally, we found that LVMs can learn latent representations that are useful for phoneme recognition and better than Mel spectrograms, which are tailored for the task, when identical models are trained on top of the representations.

### 2023-10-06 chapters/paper_benchmarking_workshop.tex
Improve hyphenation and add "= for compound words

@@
 (i) LVMs achieve likelihoods superior to WaveNet at low temporal resolution, 
-(ii) LVMs with autoregressive decoders achieve better likelihoods than a non-autoregressive LVM, 
+(ii) LVMs with autoregressive decoders achieve better likelihoods than a non"=autoregressive LVM, 
 (iii) LVM likelihoods improve when using hierarchies of latent variables as also seen for images and, 
@@
 This makes them useful beyond generation for e.g.
-semi-supervised learning \parencite{kingma_semi-supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.
+semi"=supervised learning \parencite{kingma_semi"=supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.
 In this paper, we focus on the VRNN \parencite{chung_recurrent_2015}, SRNN \parencite{fraccaro_sequential_2016} and STCN \parencite{aksan_stcn_2019} since they are similar to the original VAE framework.

### 2023-10-06 chapters/paper_hierarchical.tex
Improve hyphenation and add "= for compound words

@@
     \centering
-    \resizebox*{!}{0.82\textheight}{%
+    \resizebox*{!}{0.83\textheight}{%
     \begin{tabular}{lrrr}
@@
 
+We now move to the likelihood ratio-based score.
+We find that $LLR^{>k}$ separates the OOD MNIST data from in-distribution FashionMNIST to a higher degree than the likelihood estimates as can be seen by the empirical densities of the score in \cref{fig_hierarchical:FMNIST-llr}.
+We note that the likelihood ratio between the ELBO and the $\mathcal{L}^{>k}$ bound provides the highest degree of separation of MNIST and FashionMNIST as measured by the AUROC$\uparrow$ for $k=1$ smaller than $L$.
+This is not surprising since the value of $k$ that provides the maximal separation to the reference in-distribution dataset need not be the one for which $\mathcal{LLR}^{>k}$ is overall maximal for the OOD dataset.
+We also visualize the ROC curves resulting from using the $LLR^{>k}$ score for OOD detection on both FashionMNIST/MNIST and CIFAR10/SVHN and compare it to the ROC curves resulting from the different $\mathcal{L}^{>k}$ bounds in \cref{fig_hierarchical:FMNIST-roc-llr and CIFAR10-roc-llr}, respectively.
+On both datasets we see significantly better discriminatory performance when using the $LLR^{>k}$ score.
+
+\cref{tab_hierarchical:rocauc-ood} shows that BIVA improves upon the HVAE model for OOD detection on CIFAR while \cref{tab_hierarchical:bits-per-dim-ood} shows that the BIVA model also improves upon the HVAE in terms of likelihood.
+We hypothesize that models larger than our implementation of BIVA, with better likelihood scores may perform even better \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
+
 \begin{figure}
@@
 \end{figure}
-
-We now move to the likelihood ratio-based score.
-We find that $LLR^{>k}$ separates the OOD MNIST data from in-distribution FashionMNIST to a higher degree than the likelihood estimates as can be seen by the empirical densities of the score in \cref{fig_hierarchical:FMNIST-llr}.
-We note that the likelihood ratio between the ELBO and the $\mathcal{L}^{>k}$ bound provides the highest degree of separation of MNIST and FashionMNIST as measured by the AUROC$\uparrow$ for $k=1$ smaller than $L$.
-This is not surprising since the value of $k$ that provides the maximal separation to the reference in-distribution dataset need not be the one for which $\mathcal{LLR}^{>k}$ is overall maximal for the OOD dataset.
-We also visualize the ROC curves resulting from using the $LLR^{>k}$ score for OOD detection on both FashionMNIST/MNIST and CIFAR10/SVHN and compare it to the ROC curves resulting from the different $\mathcal{L}^{>k}$ bounds in \cref{fig_hierarchical:FMNIST-roc-llr and CIFAR10-roc-llr}, respectively.
-On both datasets we see significantly better discriminatory performance when using the $LLR^{>k}$ score.
-
-\cref{tab_hierarchical:rocauc-ood} shows that BIVA improves upon the HVAE model for OOD detection on CIFAR while \cref{tab_hierarchical:bits-per-dim-ood} shows that the BIVA model also improves upon the HVAE in terms of likelihood.
-We hypothesize that models larger than our implementation of BIVA, with better likelihood scores may perform even better \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
 
@@
 
-We also note that since the hierarchical VAE is a probabilistic graphical latent variable model, it lends itself very naturally to manipulation at the feature level \parencite{kingma_semi-supervised_2014, maaloe_auxiliary_2016, maaloe_semi-supervised_2017}.
+We also note that since the hierarchical VAE is a probabilistic graphical latent variable model, it lends itself very naturally to manipulation at the feature level \parencite{kingma_semi"=supervised_2014, maaloe_auxiliary_2016, maaloe_semi"=supervised_2017}.
 This property sets it apart from other generative models that do not explicitly define such a hierarchy of features.

### 2023-10-06 chapters/paper_retrospective.tex
Updates to abstracts
Improve hyphenation and add "= for compound words

@@
 Since our model was trained on a representative - and therefore unbalanced - sample, we expect it to behave similarly.
-Future research should look to address these shortcomings, for example by utilising self-supervised learning on massive amounts of diverse, unlabelled data covering multiple languages, accents, and dialects.
+Future research should look to address these shortcomings, for example by utilising self"=supervised learning on massive amounts of diverse, unlabelled data covering multiple languages, accents, and dialects.
 
@@
 However, this did not improve performance.
-In addition, we tested a pre-trained self-supervised model.
+In addition, we tested a pre-trained self"=supervised model.
 Although many of these models are freely available to the public, they are primarily trained on English data.
@@
 We hypothesise that the number of ground truth stroke positives was too small for these advanced models to learn more complex patterns than the MLP ensemble.
-In addition, a self-supervised model would likely benefit from being pre-trained on speech or text data from the target domain.
+In addition, a self"=supervised model would likely benefit from being pre-trained on speech or text data from the target domain.
 Although training such large-scale foundation models have the potential to improve the classification model further, it is beyond the scope of this study.
 Thus, we chose the simpler MLP ensemble.
-We have included references to reviews of self-supervised learning for speech and text in the references \parencite{gururangan_don_2020,mohamed_selfsupervised_2022}.
+We have included references to reviews of self"=supervised learning for speech and text in the references \parencite{gururangan_don_2020,mohamed_selfsupervised_2022}.
 Notably, it is not uncommon for small, simple models to match or outperform large, pre-trained models for text-classification tasks \parencite{galke_bagofwords_2022}.
@@
 
-\subsection{Author contributions}
+\section*{Author contributions}
 

### 2023-10-06 chapters/paper_review.tex
Minor updates
Change reference keys (WIP)
Change citation keys (WIP)
Update reference keys for review paper
Conditional skipping of compilation
Revise skipping
Updates
Improve hyphenation and add "= for compound words

@@
 
-\chapter[self-supervised speech representation learning: a review]{Self-Supervised Speech Representation Learning: A Review}
-\label{chp:review}
+\chapter[self"=supervised speech representation learning: a review]{Self-Supervised Speech Representation Learning: A Review}
+\label{chp:paper-review}
+\ifthenelse{\equal{\skippapers}{true}}{}{
 
@@
 
+\newcommand{\tabincell}[2]{\begin{tabular}{@{}#1@{}}#2\end{tabular}}
+
 \section*{Abstract}
@@
 It is likewise difficult to apply this to dialects and languages for which only limited labeled data is available.
-Self-supervised representation learning methods promise a single universal model that would benefit a wide variety of tasks and domains.
+Self"=supervised representation learning methods promise a single universal model that would benefit a wide variety of tasks and domains. 
 Such methods have shown success in natural language processing and computer vision domains, achieving new levels of performance while reducing the number of labels required for many downstream scenarios.
 Speech representation learning is experiencing similar progress in three main categories: generative, contrastive, and predictive methods.
-Other approaches rely on multi-modal data for pre-training, mixing text or visual data streams with speech.
-Although self-supervised speech representation is still a nascent research area, it is closely related to acoustic word embedding and learning with zero lexical resources, both of which have seen active research for many years.
-This review presents approaches for self-supervised speech representation learning and their connection to other research areas.
+Other approaches rely on multi-modal data for pre"=training, mixing text or visual data streams with speech. 
+Although self"=supervised speech representation is still a nascent research area, it is closely related to acoustic word embedding and learning with zero lexical resources, both of which have seen active research for many years.
+This review presents approaches for self"=supervised speech representation learning and their connection to other research areas. 
 Since many current methods focus solely on automatic speech recognition as a downstream task, we review recent efforts on benchmarking learned representations to extend the application beyond speech recognition.
@@
 \input{chapters/paper_review/02_three_waves}
-\input{chapters/paper_review/03_approaches}  % TODO (JDH): Compilation errors completed until here
-% \input{chapters/paper_review/04_benchmarking}
-% \input{chapters/paper_review/04b_analysis}
-% \input{chapters/paper_review/05_zero}
-% \input{chapters/paper_review/06_conclusions}
+\input{chapters/paper_review/03_approaches}
+\input{chapters/paper_review/04_benchmarking}
+\input{chapters/paper_review/04b_analysis} 
+\input{chapters/paper_review/05_zero}
+\input{chapters/paper_review/06_conclusions}
+
+}

### 2023-10-06 chapters/paper_review/01_introduction.tex
Improve hyphenation and add "= for compound words

@@
     \includegraphics[width=0.80\textwidth]{paper_review/SSL_framework.png}
-	 \caption{Framework for using self-supervised representation learning in
+	 \caption{Framework for using self"=supervised representation learning in
 	 downstream applications.}
@@
 ``supervised'' approaches, which assign a label to each training sample, and
-``semi-supervised'' approaches, which utilize a small number of training samples
+``semi"=supervised'' approaches, which utilize a small number of training samples
 with labels to guide learning using a larger volume of unlabeled samples.
@@
 and non-negative matrix factorization~\parencite{lee_learning_1999}. 
-\textit{Self-supervised learning} (SSL) is a fast-growing subcategory of
+\textit{Self"=supervised learning} (SSL) is a fast-growing subcategory of
 unsupervised learning approaches, which are techniques that utilize
@@
 In this review, we
-focus on self-supervised learning approaches.
+focus on self"=supervised learning approaches.
 
-\Cref{fig:SSL_framework} outlines self-supervised representation learning in
+\Cref{fig:SSL_framework} outlines self"=supervised representation learning in
 relation to downstream applications. 
@@
 Since the speech signal differs greatly from image and text inputs, many  theories  and technologies have been developed to address the unique challenges of speech.
-One review addresses speech representation learning based on deep learning models~\parencite{latif_deep_2021}, but does not address recent developments in self-supervised learning.
+One review addresses speech representation learning based on deep learning models~\parencite{latif_deep_2021}, but does not address recent developments in self"=supervised learning.
 This motivates this overview of speech SSL.

### 2023-10-06 chapters/paper_review/02_three_waves.tex
Improve hyphenation and add "= for compound words

@@
 In this section we present the historical background of the current surge in
-self-supervised representation learning methods in the context of two previous
+self"=supervised representation learning methods in the context of two previous
 waves of research work in the 1990s and 2000s. The discussed approaches go
@@
 (1)~All layers are trained end-to-end to optimize a single pretext task instead
-of relying on layer-wise pre-training
+of relying on layer-wise pre"=training
 (2)~Past stacked networks typically had only a few layers, but 
@@
 A closely related area of research that is not covered in this review is
-semi-supervised pre-training methods such as pseudo-labeling (that is,
-self-training). Pseudo-labeling (PL) relies on a supervised teacher model to
+semi"=supervised pre"=training methods such as pseudo"=labeling (that is,
+self-training). Pseudo"=labeling (PL) relies on a supervised teacher model to
 label a large volume of speech-only data, which is then used to augment the
@@
 One distinguishing factor in PL is the utilization of supervised data for a
-specific task during model pre-training, which limits the model's focus to
+specific task during model pre"=training, which limits the model's focus to
 a single (or at best a few) downstream tasks. 
@@
 Transfer learning (TL) is another closely related area of research for
-pre-training speech models. TL transfers knowledge captured by models
+pre"=training speech models. TL transfers knowledge captured by models
 trained on one task to different but related tasks~\parencite{caruana_multitask_1997}. 
@@
 low-resource speech recognition tasks~\parencite{cui_multilingual_2015}.
-SSL can be regarded as a type of TL because knowledge learned from pre-training
+SSL can be regarded as a type of TL because knowledge learned from pre"=training
 is used for different downstream tasks.

### 2023-10-06 chapters/paper_review/03_approaches.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses
Improve hyphenation and add "= for compound words

@@
 In NLP, morpheme-like tokens are widely used as sequence units in
-pre-training. The standard BERT takes 512 morpheme-like tokens as
+pre"=training. The standard BERT takes 512 morpheme-like tokens as
 input, usually covering a paragraph including several sentences. 
@@
 \item \textit{Speech processing tasks are diverse.}
-Building generalizable self-supervised representation models for diverse speech
+Building generalizable self"=supervised representation models for diverse speech
 processing tasks is challenging. 
@@
 requires a model that extracts speaker information but removes content information.
-Therefore, it is challenging to define a self-supervised model whose
+Therefore, it is challenging to define a self"=supervised model whose
 representations are suitable for both speech recognition and speaker
@@
 notation.
-Models are assumed to consist of functions $f(\cdot)$ and $g(\cdot)$, where $f(\cdot)$  denotes the representation model to be used after pre-training and $g(\cdot)$ is an auxiliary module needed only to support the pretext task.
+Models are assumed to consist of functions $f(\cdot)$ and $g(\cdot)$, where $f(\cdot)$  denotes the representation model to be used after pre"=training and $g(\cdot)$ is an auxiliary module needed only to support the pretext task.
 For instance, in a classic autoencoder, $f(\cdot)$ would denote the encoder and $g(\cdot)$ the decoder.
 For more complex models, these functions might consist of several components indicated by sub-indices $f_1(\cdot) \dots f_N(\cdot)$.
-As we will see, many self-supervised models use masking, which replaces some parts of the input or a hidden representation  by zeros or a learned vector.
+As we will see, many self"=supervised models use masking, which replaces some parts of the input or a hidden representation  by zeros or a learned vector.
 We use $m(\cdot)$ to denote a function that applies such masking to its input.
-Similar to $g(\cdot)$, this function is only used during pre-training.
+Similar to $g(\cdot)$, this function is only used during pre"=training.
 
@@
 Most importantly, $g(\cdot)$ produces an output that is used for the pretext task but is not  used by $f(\cdot)$ to produce the representation $H$.
-Hence, $g(\cdot)$ can be discarded after pre-training.
+Hence, $g(\cdot)$ can be discarded after pre"=training.
 Finally, $f(\cdot)$ commonly downsamples the temporal dimension, but again, this is not crucial to understand the models, so consider only a single temporal scale $t\in\{1,\dots, T\}$ for notational convenience.
@@
 
-\edit{After pre-training,} the representation model $f(\cdot)$ can be fine-tuned for a downstream task directly or used to extract features which are fed to another model, as visualized in \cref{fig:SSL_framework}.
+\edit{After pre"=training,} the representation model $f(\cdot)$ can be fine-tuned for a downstream task directly or used to extract features which are fed to another model, as visualized in \cref{fig:SSL_framework}.
 It is not uncommon to use the output representation~$H$, but often representations from hidden layers of $f(\cdot)$ are better suited \parencite{pasad_layerwise_2021}.
 
-\begin{figure*}[t]
+\begin{figure*}
     \centering
-    \includegraphics[width=0.98\textwidth]{paper_review/model_timeline.pdf}
+    \includegraphics[width=\textwidth]{paper_review/model_timeline.pdf}
    % \includegraphics[width=0.98\textwidth]{paper_review/model_timeline_wo_dccae.pdf}
-	 \caption[Selection of self-supervised models for speech.]{A selection of models listed according to first publication date
+	 \caption[Selection of self"=supervised models for speech.]{A selection of models listed according to first publication date
 	 on arXiv or conference submission date when this clearly precedes the
@@
 representations. 
-\edit{The VQ-VAE has been instrumental in generative speech modelling and recent work on generative spoken language modeling has successfully combined the idea of a discrete latent space with self-supervised learning \parencite{polyak_speech_2021, kharitonov_textfree_2021, nguyen_generative_2022}.}
+\edit{The VQ-VAE has been instrumental in generative speech modelling and recent work on generative spoken language modeling has successfully combined the idea of a discrete latent space with self"=supervised learning \parencite{polyak_speech_2021, kharitonov_textfree_2021, nguyen_generative_2022}.}
 
@@
 Masked reconstruction is largely inspired by the masked language model (MLM) task from BERT~\parencite{devlin_bert_2018}.
-During BERT pre-training, some tokens in the input sentences are masked by randomly replacing them by a learned masking token or another input token.
+During BERT pre"=training, some tokens in the input sentences are masked by randomly replacing them by a learned masking token or another input token.
 The model learns to reconstruct the masked tokens from the non-masked tokens.
@@
 The function $f(\cdot)$ is typically a Transformer encoder~\parencite{liu_mockingjay_2020,jiang_improving_2019,liu_masked_2020}, but recurrent neural networks have also been used~\parencite{wang_unsupervised_2020}.
-\edit{In general, the Transformer encoder architecture has been adopted widely by self-supervised models for speech within all three surveyed categories.} \edit{The function $g(\cdot)$ is usually a linear projection or a multilayer perceptron (MLP)}.
+\edit{In general, the Transformer encoder architecture has been adopted widely by self"=supervised models for speech within all three surveyed categories.} \edit{The function $g(\cdot)$ is usually a linear projection or a multilayer perceptron (MLP)}.
 Finally, the loss $\mathcal{L}_t$ is commonly computed only for masked timesteps in order to discourage the model from learning an identity mapping.
 
-The masking policies used in NLP can be adapted to speech by considering a speech \edit{segment} equivalent to a token in a sentence; indeed, the masking strategy of BERT has also been used for speech pre-training~\parencite{liu_mockingjay_2020}.
+The masking policies used in NLP can be adapted to speech by considering a speech \edit{segment} equivalent to a token in a sentence; indeed, the masking strategy of BERT has also been used for speech pre"=training~\parencite{liu_mockingjay_2020}.
 In the standard BERT masking policy, each token is masked independently at random.
@@
 Some studies explore alternatives to masking the input directly.
-In non-autoregressive predictive coding (NPC)~\parencite{liu_nonautoregressive_2020}, time masking is introduced through masked convolution blocks.
-Taking inspiration from XLNet~\parencite{yang_xlnet_2019}, it has also been suggested that the input be reconstructed from a shuffled version~\parencite{song_speechxlnet_2020} to address the discrepancy between pre-training and fine-tuning of masking-based approaches.
+In non"autoregressive predictive coding (NPC)~\parencite{liu_nonautoregressive_2020}, time masking is introduced through masked convolution blocks.
+Taking inspiration from XLNet~\parencite{yang_xlnet_2019}, it has also been suggested that the input be reconstructed from a shuffled version~\parencite{song_speechxlnet_2020} to address the discrepancy between pre"=training and fine-tuning of masking-based approaches.
 
@@
 Other than the autoregressive and masked reconstruction tasks discussed above, various studies have explored the reconstruction of other targets derived from the input.
-PASE and PASE+~\parencite{pascual_learning_2019,ravanelli_multitask_2020} use multiple targets, including the waveform, log power spectrum, \edit{mel cepstral coefficients (MFCCs)}, and prosody features.
+PASE and PASE+~\parencite{pascual_learning_2019,ravanelli_multitask_2020} use multiple targets, including the waveform, log power spectrum, \edit{Mel cepstral coefficients (MFCCs)}, and prosody features.
 Models that learn acoustic embeddings of small speech segments have targeted future and past spectrogram segments~\parencite{chung_speech2vec_2018, tagliasacchi_selfsupervised_2019, tagliasacchi_pretraining_2020}, phase information~\parencite{quitry_learning_2019}, and the temporal gap between two segments~\parencite{tagliasacchi_selfsupervised_2019, tagliasacchi_pretraining_2020}.
@@
 Since the quality of the learned representations is contingent on the quality of the quantization, wav2vec 2.0 combines two techniques to learn high-quality codebooks.
-First, wav2vec 2.0 concatenates quantized representations from multiple codebooks at each timestep, so-called Product Quantization (PQ)~\parencite{jegou_product_2011}.
+First, wav2vec 2.0 concatenates quantized representations from multiple codebooks at each timestep, so-called Product Quantization (PQ) \parencite{jegou_product_2011}.
 Also, the \edit{primary} training loss \edit{described below} is augmented with an auxiliary term designed to encourage equal use of all codebook entries.
@@
 \noindent where $\mathcal{M}$ is the set of all masked timesteps. During training, only the BERT model's parameters are updated, while the vq-wav2vec model parameters are frozen.} 
-Discrete BERT was the first model to demonstrate the effectiveness of self-supervised speech representation learning by achieving a WER of 25\% on the standard test-other subset using a 10-minute fine-tuning set, \edit{setting the direction for many approaches to follow}.
+Discrete BERT was the first model to demonstrate the effectiveness of self"=supervised speech representation learning by achieving a WER of 25\% on the standard test-other subset using a 10-minute fine-tuning set, \edit{setting the direction for many approaches to follow}.
 
@@
 First, the model needs to learn a meaningful continuous latent representation for unmasked timesteps which are mapped to discrete units, similar to a classical frame-based acoustic modeling problem.
-Second, similar to other masked pre-training approaches, the model needs to capture long-range temporal dependencies to make correct predictions for masked timesteps}.
+Second, similar to other masked pre"=training approaches, the model needs to capture long-range temporal dependencies to make correct predictions for masked timesteps}.
 
 One crucial insight motivating this work is the importance of consistency of the targets which enables the model to focus on modeling the sequential structure of the input.
-\edit{Importantly though, for HuBERT, pre-training is a two-step procedure.
+\edit{Importantly though, for HuBERT, pre"=training is a two-step procedure.
 The first iteration is described above.
-Once completed, a second iteration of pre-training follows. Here, representations from a hidden layer of the model from the first iteration are clustered with $k$-means to obtain new targets $c_t$.}
+Once completed, a second iteration of pre"=training follows. Here, representations from a hidden layer of the model from the first iteration are clustered with $k$-means to obtain new targets $c_t$.}
 
@@
 \edit{Second, it uses an utterance mixing strategy where signals from different speakers are combined to augment the training data.} \edit{Specifically, random subsequences from other examples in the same batch are scaled and added to each input example.
-Only the targets corresponding to the original example are predicted during pre-training. Thus, the model learns to filter out the added overlapping speech.}
+Only the targets corresponding to the original example are predicted during pre"=training. Thus, the model learns to filter out the added overlapping speech.}
 
@@
 
-\paragraph{data2vec} \edit{Motivated by the success of using an exponential moving average (EMA) teacher for self-supervised visual representations~\parencite{grill_bootstrap_2020, caron_emerging_2021}, the data2vec model~\parencite{baevski_data2vec_2022} computes targets $Y$ using an EMA of its own parameters.
+\paragraph{data2vec} \edit{Motivated by the success of using an exponential moving average (EMA) teacher for self"=supervised visual representations~\parencite{grill_bootstrap_2020, caron_emerging_2021}, the data2vec model~\parencite{baevski_data2vec_2022} computes targets $Y$ using an EMA of its own parameters.
 The targets are constructed by averaging hidden representations of the top $k$ layers of the EMA teacher network applied to unmasked inputs.
@@
 \subsubsection{Challenges}
-The iterative nature of pre-training for the HuBERT and wavLM could present a practical inconvenience when working with large volumes of data.
+The iterative nature of pre"=training for the HuBERT and wavLM could present a practical inconvenience when working with large volumes of data.
 Another challenge for these models centers around the quality of the initial vocabulary generated from MFCC features.
@@
 modalities produced directly by the speech source.
-Examples of intrinsic modalities (besides the speech audio) include images or video of the speaker's face~\parencite{lee_avicar_2004,chung_lip_2016}, lip-movement~\parencite{shi_learning_2022}, articulatory flesh point measurements~\parencite{westbury_xray_1990,wrench_new_2001}, or simultaneous \edit{magnetic resonance imaging (MRI)} scans~\parencite{narayanan_multimodal_2011}.
+Examples of intrinsic modalities (besides the speech audio) include images or video of the speaker's face~\parencite{lee_avicar_2004,chung_lip_2016}, lip-movement~\parencite{shi_learning_2022}, articulatory flesh point measurements~\parencite{westbury_xray_1990,wrench_new_2001}, or simultaneous \edit{magnetic resonance imaging (MRI)} scans \parencite{narayanan_multimodal_2011}.
 Typically, learning from multiple intrinsic modalities is done so as to improve robustness to noise, since acoustic noise is likely to be uncorrelated with the other \edit{modality(ies)}.
@@
 The SLAM approach~\parencite{bapna_slam_2021} is an example where speech and text are first represented using two separate pre-trained encoders followed by a multi-modal encoder to build the joint representations.
-The entire model is trained using a multi-task loss including two supervised and two self-supervised tasks.
+The entire model is trained using a multi-task loss including two supervised and two self"=supervised tasks.
 
@@
 
-Since most work on acoustic word embeddings preceded the very recent wave of new self-supervised frame-level representations, one question is whether word (or more generally segment) embeddings could be derived more simply by pooling self-supervised frame-level representations, as has been done for text span embeddings by pooling over word embeddings~\parencite{toshniwal_crosstask_2020,wang_phrasebert_2021}.
+Since most work on acoustic word embeddings preceded the very recent wave of new self"=supervised frame-level representations, one question is whether word (or more generally segment) embeddings could be derived more simply by pooling self"=supervised frame-level representations, as has been done for text span embeddings by pooling over word embeddings~\parencite{toshniwal_crosstask_2020,wang_phrasebert_2021}.
 Some initial results suggest that at least very simple pooling approaches like downsampling and mean or max pooling are not successful~\parencite{vanstaden_comparison_2021,peng_correspondence_2020}, but more work is needed to reach conclusive results.

### 2023-10-06 chapters/paper_review/04_benchmarking.tex
Improve hyphenation and add "= for compound words

@@
 
-\section{Benchmarks for self-supervised learning}
+\section{Benchmarks for self"=supervised learning}
 \label{section:benchmark}
@@
 
-\subsection{Datasets only for pre-training} 
-\Cref{table:datasets} summarizes datasets used for pre-training SSL techniques
+\subsection{Datasets only for pre"=training} 
+\Cref{table:datasets} summarizes datasets used for pre"=training SSL techniques
 in the literature. These datasets are usually large but with limited or no
@@
 consists of over 2M 10-second YouTube video clips human-annotated with 632
-audio events, have also been used for pre-training. Audioset has 2.5k hours
+audio events, have also been used for pre"=training. Audioset has 2.5k hours
 of audio of varying quality, different languages, and sometimes multiple sound
@@
 over 2k hours of conversational telephone speech, 1k hours of which is utilized
-for pre-training~\parencite{jiang_further_2021}. Industrial researchers have also
+for pre"=training~\parencite{jiang_further_2021}. Industrial researchers have also
 begun to build large-scale datasets for learning speech representations.
 For instance, 10k hours of real-world far-field English voice commands for
-self-supervised pre-training have been collected at 
+self"=supervised pre"=training have been collected at 
 Amazon~\parencite{sadhu_wav2vecc_2021}.
@@
 In addition to these English and multilingual efforts, researchers have also
-collected corpora for pre-training Chinese speech representations. Didi Dictation
+collected corpora for pre"=training Chinese speech representations. Didi Dictation
 and Didi Callcenter~\parencite{jiang_improving_2019, jiang_further_2021} are two
@@
 
-\subsection{Datasets for both pre-training and evaluation}\label{sec:datasets}
+\subsection{Datasets for both pre"=training and evaluation}\label{sec:datasets}
 Several datasets that provide both speech and associated transcripts and
 speaker labels have also been used to develop SSL techniques by enabling
-in-domain pre-training and evaluation. Such datasets are also listed in
+in-domain pre"=training and evaluation. Such datasets are also listed in
 \cref{table:datasets}. One of the most commonly used datasets in this category
@@
 speakers that yield higher WER as measured with previously built models. LS is
-used for unsupervised representation pre-training by ignoring its labels, and
+used for unsupervised representation pre"=training by ignoring its labels, and
 can also be utilized to evaluate the performance of representation on ASR, phoneme recognition (PR),
 phoneme classification (PC), and speaker identification (SID) tasks. Wall Street Journal (WSJ)~\parencite{paul_design_1992} is another
-widely adopted, labeled corpus for pre-training. Its labels can evaluate
+widely adopted, labeled corpus for pre"=training. Its labels can evaluate
 performance for ASR, PR, PC, and SID. The original WSJ corpus contains 400 hours
@@
 The speech community also utilizes multilingual corpora. These are often
-large-scale, which facilitates pre-training, but are also partially labeled for ASR
+large-scale, which facilitates pre"=training, but are also partially labeled for ASR
 evaluation (PC and PR can be enabled via phone-level forced alignment). These
@@
 of audio from 76 languages as of the date this review was written (Common Voice
-corpus~7.0). Researchers usually use part of this for pre-training (e.g., 7k
+corpus~7.0). Researchers usually use part of this for pre"=training (e.g., 7k
 hours/60 languages in \parencite{babu_xlsr_2021} and 430 hours/29 languages in
@@
 The entire dataset~\parencite{babu_xlsr_2021} or a 24k-hour 
-portion~\parencite{chen_unispeechsat_2021, chen_wavlm_2021} thereof has been used for pre-training. BBL
+portion~\parencite{chen_unispeechsat_2021, chen_wavlm_2021} thereof has been used for pre"=training. BBL
 consists of 1k hours of conversational telephone speech in 17 African and Asian
@@
 also aggregated to build diversified and large-scale corpora for unsupervised
-pre-training~\parencite{kawakami_learning_2020, song_speechxlnet_2020, babu_xlsr_2021}.
+pre"=training~\parencite{kawakami_learning_2020, song_speechxlnet_2020, babu_xlsr_2021}.
 GigaSpeech is a multi-domain English ASR corpus with 33k hours of audio
@@
 is annotated for language identification. Beyond the original purpose of
-evaluation, these corpora are also used in pre-training to improve the
+evaluation, these corpora are also used in pre"=training to improve the
 generalizability of learned representations.
 
-For the purpose of pre-training and evaluating Mandarin speech representations,
+For the purpose of pre"=training and evaluating Mandarin speech representations,
 the authors of \parencite{jiang_improving_2019, jiang_further_2021} also compiled Open Mandarin, 
@@
 speech, of which
-168 hours of audio is used for pre-training; the development and test sets are
+168 hours of audio is used for pre"=training; the development and test sets are
 excluded.)~\parencite{liu_hkust_2006}, AISHELL-1~\parencite{bu_aishell1_2017} (178 hours of
@@
 Besides the aforementioned datasets, conventional speech processing benchmarks
-are also used to evaluate self-supervised representations. Studies leverage Hub5,
+are also used to evaluate self"=supervised representations. Studies leverage Hub5,
 DIRHA, and CHiME-5 to measure the efficacy of representations in ASR.
@@
 Last, \parencite{tagliasacchi_pretraining_2020} puts together five datasets 
-(MUSAN~\parencite{snyder_musan_2015}, Bird Audio Detection~\parencite{stowell_automatic_2019},
+(MUSAN~\parencite{snyder_musan_2015}, Bird Audio Detection \parencite{stowell_automatic_2019},
 Speech Commands~\parencite{warden_speech_2018}, Spoken Language 
@@
   \scriptsize %footnotesize %small, scriptsize, tiny
-  \caption[Summary of datasets used for pre-training or evaluation of SSL techniques in the literature.]%
-  {Summary of datasets used for pre-training (denoted as PT) or
+  \caption[Summary of datasets used for pre"=training or evaluation of SSL techniques in the literature.]%
+  {Summary of datasets used for pre"=training (denoted as PT) or
   evaluation (denoted as EV) of SSL techniques in the literature. The languages and
@@
     \toprule
-    \multirow{2}{*}{Work} & \multirow{2}{*}{\makecell[l]{Pre-training\\ corpus}} & \multirow{2}{*}{Task} & \multicolumn{2}{c}{Dataset} & \multirow{2}{*}{Transfer} & \multirow{2}{*}{Fine-tuning labels used} \\ \cline{4-5}
+    \multirow{2}{*}{Work} & \multirow{2}{*}{\makecell[l]{pre"=training\\ corpus}} & \multirow{2}{*}{Task} & \multicolumn{2}{c}{Dataset} & \multirow{2}{*}{Transfer} & \multirow{2}{*}{Fine-tuning labels used} \\ \cline{4-5}
      &  &  & Fine-tuning & Test & & \\
@@
 \footnotetext{\label{footnote:table:method_data_setting}
-  The \textbf{Pre-training corpus}, \textbf{Fine-tuning}, and
+  The \textbf{pre"=training corpus}, \textbf{Fine-tuning}, and
   \textbf{Test} columns list the datasets used in each work, and the \textbf{Task}
@@
   The \textbf{Transfer} column indicates whether the SSL technique is evaluated by its capability for
-  transfer learning, i.e., different datasets are utilized for pre-training and
+  transfer learning, i.e., different datasets are utilized for pre"=training and
   fine-tuning. The \textbf{Fine-tuning labels used} column summarizes the
@@
 English speech with various accents. Among the papers studied in this review, SSA
-is used in \parencite{kawakami_learning_2020} only for pre-training, and only 1 hour
+is used in \parencite{kawakami_learning_2020} only for pre"=training, and only 1 hour
 of audio is utilized. Thus, we exclude it from our discussion in
@@
 fine-tune a pre-trained SSL model for a supervised downstream task.
-Depending on the corpora used in pre-training and fine-tuning, techniques can
+Depending on the corpora used in pre"=training and fine-tuning, techniques can
 be benchmarked in terms of their capability to transfer knowledge across datasets
-(i.e., using pre-training corpora that differ from the fine-tuning ones), their
+(i.e., using pre"=training corpora that differ from the fine-tuning ones), their
 benefit when training with limited labeled examples (i.e., sampling a subset of
@@
 fine-tuning). \cref{table:method_data_setting,table:method_data_setting_2} summarize experiment settings used in the
-SSL literature, including the pre-training corpora, downstream tasks and datasets,
+SSL literature, including the pre"=training corpora, downstream tasks and datasets,
 and the amount of fine-tuning labels used, which indicates the targeted
@@
 As observed in \cref{table:method_data_setting,table:method_data_setting_2}, LS and WSJ are the most commonly used
-pre-training corpora. At the same time, we observe a
-growing industry investment in pre-training with larger datasets, e.g.,
-CPC-8k (8k hours) for Bidir-CPC~\parencite{kawakami_learning_2020}, LL (60k hours) for
+pre"=training corpora. At the same time, we observe a
+growing industry investment in pre"=training with larger datasets, e.g.,
+CPC-8k (8k hours) for Bidir-CPC \parencite{kawakami_learning_2020}, LL (60k hours) for
 CPC modified~\parencite{riviere_unsupervised_2020}, 
@@
 multilinguality~\parencite{kawakami_learning_2020, babu_xlsr_2021} are also gaining
-attention. Compared to pre-training, datasets used for fine-tuning are more
+attention. Compared to pre"=training, datasets used for fine-tuning are more
 diverse and cover downstream tasks as varied as ASR, PR, PC, SID, AED,
@@
 popularity of SSL research, we expect future experiment settings to proliferate
-and cover more languages, downstream tasks, and pre-training/fine-tuning
+and cover more languages, downstream tasks, and pre"=training/fine-tuning
 datasets.
@@
 \textit{X-Large}, or \textit{XX-Large} variants specified in the original
-publication. We also compare with baselines including the log mel filterbank (fBANK) and semi-supervised, self-training
+publication. We also compare with baselines including the log mel filterbank (fBANK) and semi"=supervised, self-training
 approaches (iterative pseudo labeling (IPL)~\parencite{xu_iterative_2020}, 
@@
 Given the
-diversity of the listed methods in experiment settings (e.g., pre-training corpora
+diversity of the listed methods in experiment settings (e.g., pre"=training corpora
 and objectives, whether a language model is used in decoding, whether model
@@
 the caption of \cref{fig:asr_result_no_limit} for more details).
-Semi-supervised methods such as self-training, where a model is first trained
+Semi"=supervised methods such as self-training, where a model is first trained
 on labeled data to annotate unlabeled speech, and then subsequently trained on
@@
 like wav2vec 2.0 and HuBERT, becomes more evident when the labeling resources
-become scarce. Compared to popular semi-supervised methods such as IPL,
+become scarce. Compared to popular semi"=supervised methods such as IPL,
 slimIPL, and noisy student using 100 hours of labels, wav2vec~2.0 and HuBERT
@@
 In addition to the ASR task, where the current state of the art is achieved by a method
-combining SSL pre-training and self-training 
+combining SSL pre"=training and self-training 
 techniques~\parencite{zhang_pushing_2020}, SSL models are competitive in other tasks, including IC,
@@
   \footnotesize
-  \caption{Tasks where the state of the art is models with SSL pre-training.}
+  \caption{Tasks where the state of the art is models with SSL pre"=training.}
   \label{table:sota_performance}
@@
 Despite the obvious trend of increasing performance as more parameters and SSL
-pre-training data are being used, numbers in \cref{fig:asr_result_no_limit} 
+pre"=training data are being used, numbers in \cref{fig:asr_result_no_limit} 
 and \cref{table:sota_performance} are less comparable than might be expected.

### 2023-10-06 chapters/paper_review/04b_analysis.tex
Improve hyphenation and add "= for compound words

@@
 
-\section{Analysis of self-supervised representations}
+\section{Analysis of self"=supervised representations}
 \label{analysis}
@@
 % {\color{blue} Katrin}\\
-The previous sections have shown how self-supervised learning can result in
+The previous sections have shown how self"=supervised learning can result in
 powerful representations that provide a robust starting point for several
@@
 tries to answer these questions by summarizing several studies that analyze
-self-supervised representations.
+self"=supervised representations.
 
@@
 The peaking of class-relevant information in intermediate layers seems to be
-common across different self-supervised learners and different modalities. In
+common across different self"=supervised learners and different modalities. In
 an analysis of text-based Transformers trained with a masked language model
@@
 pattern. Interestingly, similar network behavior was also recently described
-for self-supervised learners in computer vision: using a contrastive
-self-supervised learner (SimCLR) that optimizes for augmentation invariance,
+for self"=supervised learners in computer vision: using a contrastive
+self"=supervised learner (SimCLR) that optimizes for augmentation invariance,
 \parencite{grigg_selfsupervised_2021} show that it is the intermediate representations that most
@@
 
-Self-supervised representations may encode other information besides phonetic
+Self"=supervised representations may encode other information besides phonetic
 classes or words, for example, channel, language, speaker, and sentiment
@@
 analyzed in \parencite{ling_bertphone_2020} for a 12-layer BERTphone model. This model
-combines a self-supervised masked reconstruction loss with a phone-based CTC
+combines a self"=supervised masked reconstruction loss with a phone-based CTC
 loss to produce representations	for speaker recognition and language
@@
 phonetic and semantic information. Another study~\parencite{wang_selfsupervised_2021} used
-explicit self-supervised loss at the intermediate layers rather than just the
+explicit self"=supervised loss at the intermediate layers rather than just the
 output layer of a HuBERT model in order to enforce better learning of phonetic
@@
 
-%Hung-yi Lee: My group also has a paper analyzing the attention of self-supervised speech model (https://arxiv.org/abs/2006.03265). I will include some findings in this subsection later.
-Most self-supervised learning approaches rely on a Transformer architecture for
+%Hung-yi Lee: My group also has a paper analyzing the attention of self"=supervised speech model (https://arxiv.org/abs/2006.03265). I will include some findings in this subsection later.
+Most self"=supervised learning approaches rely on a Transformer architecture for
 the representation model. In \parencite{yang_understanding_2020} the attention patterns in
@@
 
-How does the performance of self-supervised models change in relation to the
+How does the performance of self"=supervised models change in relation to the
 amount of training data, and in relation to the size (number of parameters) of
@@
 model~\parencite{liu_mockingjay_2020}, the authors in \parencite{pu_scaling_2021} attempt to establish a relationship between
-model size and self-supervised \ensuremath{L_1} loss and demonstrate that it approximately
+model size and self"=supervised \ensuremath{L_1} loss and demonstrate that it approximately
 follows a power law. Model size and accuracy on downstream phone
@@
 datasets from six different domains. In particular, the authors focused on the
-usefulness of adding out-of-domain data to pre-training. The general
-conclusions are that pre-training on more and diverse domains is preferable:
+usefulness of adding out-of-domain data to pre"=training. The general
+conclusions are that pre"=training on more and diverse domains is preferable:
 models pre-trained on more domains performed better than those pre-trained on
@@
 labeled data was used for fine-tuning. Adding in-domain unlabeled data---if
-available---to pre-training improves performance robustly; however, even
+available---to pre"=training improves performance robustly; however, even
 out-of-domain unlabeled data is helpful and closes  66--73\% of the performance
@@
 supervised methods in accuracy in low-data regimes (1h of labeled data
-per language). Thus, self-supervised pre-training enables the model to
+per language). Thus, self"=supervised pre"=training enables the model to
 learn contextualized speech features that generalize across different
@@
 model covers 53 languages in total
-and consists of 56,000 hours of speech. Compared to monolingual pre-training,
+and consists of 56,000 hours of speech. Compared to monolingual pre"=training,
 even smaller models trained on only ten languages improve performance
@@
 Finally, one might ask whether the interpretation of representations extracted
-from different layers of a self-supervised models also generalizes to the
+from different layers of a self"=supervised models also generalizes to the
 multilingual setting. Experiments in \parencite{baevski_unsupervised_2021} on phone

### 2023-10-06 chapters/paper_review/05_zero.tex
Improve hyphenation and add "= for compound words

@@
 % What are we going to do % how to match
-%In the self-supervised learning framework, speech representations can be learned and used in the downstream tasks, but some labeled data for finetuning is needed.
+%In the self"=supervised learning framework, speech representations can be learned and used in the downstream tasks, but some labeled data for finetuning is needed.
 
@@
 segmentation module before the generator to segment utterances into
-phoneme-level
-segments~\parencite{liu_completely_2018,chen_completely_2019,yeh_unsupervised_2019}. 
+phoneme-level segments \parencite{liu_completely_2018,chen_completely_2019,yeh_unsupervised_2019}. 
 A representation vector or a token then represents each phoneme-level segment. 
@@
 Wav2vec-U~\parencite{baevski_unsupervised_2021} selects the input feature from
-different layers of wave2vec 2.0~\parencite{baevski_wav2vec_2020}, a self-supervised
+different layers of wave2vec 2.0~\parencite{baevski_wav2vec_2020}, a self"=supervised
 model.
@@
 %{\color{black} Shinji}\\
-%{\color{black} Hung-yi: In the last meeting, we discussed the connection between this section and self-supervised learning.
+%{\color{black} Hung-yi: In the last meeting, we discussed the connection between this section and self"=supervised learning.
 In addition to looking for papers using self-supervision to improve the ASR-TTS framework here, I think there may be another direction.
-Can we regard this ASR-TTS framework as a kind of self-supervised learning?
+Can we regard this ASR-TTS framework as a kind of self"=supervised learning?
 ASR-TTS is similar to an autoencoder, and the output of ASR is "latent representation".
-In other words, this section describes a special self-supervised learning method, whose representation is expressed as "text."}\\
-%{\color{black} Hung-yi: However, the above idea sounds tricky. In addition, the ``ASR-TTS'' autoencoder is learned by semi-supervised, not self-supervised.}
+In other words, this section describes a special self"=supervised learning method, whose representation is expressed as "text."}\\
+%{\color{black} Hung-yi: However, the above idea sounds tricky. In addition, the ``ASR-TTS'' autoencoder is learned by semi"=supervised, not self"=supervised.}
 % AMH: to make the paper more consistent, I recommend changing all $f^{\mathrm{asr}}$ to $f_{\mathrm{asr}}$
@@
 representation; the framework as a whole can be regarded as a variant of
-self-supervised learning.\footnote{However, to make this complicated
+self"=supervised learning.\footnote{However, to make this complicated
 system work, we often require that data is paired. Therefore, in practice,
 ASR-TTS and other methods described in this section are categorized as
-semi-supervised learning.}
+semi"=supervised learning.}
 
@@
 and document archives without corresponding audio data.
-%Self-supervised learning in this paper mostly focuses on representation learning perspectives of such unpaired data.
+%Self"=supervised learning in this paper mostly focuses on representation learning perspectives of such unpaired data.
 The typical use case scenario of such a text resource for ASR is through \edit{the
@@
 from audio only (no text nor lexicon), are one of the most active applications
-of unsupervised/self-supervised speech processing.
+of unsupervised/self"=supervised speech processing.
 Zero-resource speech technologies were initially studied for acoustic and
@@
 neural-network-based TTS/VC techniques~\parencite{vanniekerk_vectorquantized_2020} to
-self-supervised learning~\parencite{maekaku_speech_2021}.
+self"=supervised learning~\parencite{maekaku_speech_2021}.
 The latest challenge included the visual modality, continuing the
@@
 Textless NLP is a new research direction that leverages the progress mentioned
-above in self-supervised speech representation learning to model language
+above in self"=supervised speech representation learning to model language
 directly from audio, bypassing the need for text or 

### 2023-10-06 chapters/paper_review/06_conclusions.tex
Improve hyphenation and add "= for compound words

@@
 
-In this overview, we have presented the historical context of self-supervised learning and provided a thorough methodological review of important  self-supervised speech representation models.
+In this overview, we have presented the historical context of self"=supervised learning and provided a thorough methodological review of important  self"=supervised speech representation models.
 Specifically, we have categorized the approaches into three categories, generative, contrastive and predictive, differing in terms of how the pretext task is defined.
@@
 \item \textbf{Increasing the efficiency of the representation model.} As discussed in \cref{subsec:modelsize}, larger representation models lead to better downstream performance.
-Despite the success of these \edit{large} models, they \edit{incur high costs in terms of memory and time for pre-training, fine-tuning, and even when used only to extract representations without gradient calculation.
+Despite the success of these \edit{large} models, they \edit{incur high costs in terms of memory and time for pre"=training, fine-tuning, and even when used only to extract representations without gradient calculation.
 This} makes them unsuitable for edge devices \edit{but also limits the ability to scale these models to very large datasets} \ci{ -- and leads to a large energy consumption}.
@@
 
-\edit{\item \textbf{Data-efficient approaches.} SOTA representation learning methods require large volumes of unlabeled speech during pre-training, going way beyond what babies need to understand language.
+\edit{\item \textbf{Data-efficient approaches.} SOTA representation learning methods require large volumes of unlabeled speech during pre"=training, going way beyond what babies need to understand language.
 Different learning approaches have different data needs, e.g., generative approaches could be more data efficient than contrastive or predictive approaches since they are constrained by more bits of information to reconstruct their inputs.
@@
 Noting that the BERT training corpus exceeds 3 billion words~\parencite{devlin_bert_2018}, and assuming a typical speaking rate of 120 words per minute, a speech corpus containing the same content as the BERT training data would include 400,000 hours of audio, which exceeds the \edit{accumulated} training data of all current speech representation models.
-Therefore, to enable speech representation models to better learn human language, for instance by extracting semantic information from acoustic signals, the use of text models such as BERT and GPT  \edit{seems} key: nevertheless, how to use these to improve speech representation model pre-training remains an open question.
-%\edit{ There is already some study using both speech and text data to pre-training, but some paired data is still required to achieve good performance~\parencite{bapna_slam_2021}.  }
+Therefore, to enable speech representation models to better learn human language, for instance by extracting semantic information from acoustic signals, the use of text models such as BERT and GPT  \edit{seems} key: nevertheless, how to use these to improve speech representation model pre"=training remains an open question.
+%\edit{ There is already some study using both speech and text data to pre"=training, but some paired data is still required to achieve good performance~\parencite{bapna_slam_2021}.  }
 

### 2023-10-06 chapters/paper_review/main_hi.tex
Improve hyphenation and add "= for compound words

@@
         \midrule\midrule
-        \multicolumn{7}{c}{\textit{\textbf{Pre-Training}}} \\
+        \multicolumn{7}{c}{\textit{\textbf{pre"=training}}} \\
         wav2vec 2.0 \textsc{Large}~\parencite{baevski_wav2vec_2020} & LL-60k & Transformer & 1.6 & 3.0 & 1.8 & 3.3 \\
@@
         \midrule\midrule
-        \multicolumn{7}{c}{\textit{\textbf{Pre-Training + Self-Training}}} \\
+        \multicolumn{7}{c}{\textit{\textbf{pre"=training + Self-Training}}} \\
         wav2vec 2.0 + self-training~\parencite{xu2020self} & LL-60k & Transformer & 1.1 & 2.7 & 1.5 & 3.1\\
@@
         \midrule\midrule
-        \multicolumn{7}{c}{\textit{\textbf{This work (Pre-Training)}}} \\
+        \multicolumn{7}{c}{\textit{\textbf{This work (pre"=training)}}} \\
         HUBERT \textsc{Large}   & LL-60k & Transformer & 1.5 & 3.0 & 1.9 & 3.3 \\

### 2023-10-06 frontmatter/abstract-da.tex
Improve hyphenation and add "= for compound words

@@
 
-Maskinlæring udgør en mulighed for at støtte sundhedspersonale i forskellige medicinske rammer, herunder primære klinikker og akuttjenester.
+Maskinlæring udgør en mulighed for at understøtte sundhedspersonale i forskellige medicinske rammer, herunder lægeklinikker og akuttjenester.
 Tidligere maskinlæringsapplikationer i sundhedsvæsenet har ofte forsømt naturligt sprog og fokuseret på opgaver, der involverer strukturerede data som at forudsige medicinske billeddannelsesresultater og analysere elektroniske sundhedsjournaler.
-Brugen af maskinlæringssystemer til at fortolke naturligt sprog kan dog lette de administrative- og dokumentationsbyrder for sundhedspersonale og hjælpe med at sikre, at kritisk information ikke overses eller glemmes.
+Brugen af maskinlæringssystemer til at fortolke naturligt sprog kan dog lette administrations- og dokumentationsbyrder for sundhedspersonale og hjælpe med at sikre, at kritisk information ikke overses eller glemmes.
 
 Udvikling af maskinlæringssystemer til medicinske samtaler er udfordrende på grund af den høje indvirkning som sundhedsbeslutninger har på patienters liv og kompleksiteten af ustrukturerede sprogdata, herunder tale og tekst.
-Mens maskinlæringssystemer baseret på neurale netværk har vist imponerende resultater på andre domæner, kan deres anvendelse i sundhedsvæsenet blive hindret af deres svært fortolkelige prædiktioner og deres dårlige evne til at kvantificere prædiktiv usikkerhed, som forstærkes i dynamiske miljøer og for sjældne tilfælde, almindeligvist omtalt som ude-af-fordeling data.
+Mens maskinlæringssystemer baseret på neurale netværk har vist imponerende resultater på andre domæner, kan deres anvendelse i sundhedsvæsenet blive hindret af deres svært fortolkelige prædiktioner og deres dårlige evne til at kvantificere prædiktiv usikkerhed - en svaghed som forstærkes i dynamiske miljøer og for sjældne tilfælde, også kaldt ude-af-fordeling data.
 
-Denne afhhandling udforsker maskinlæringsteknikker til medicinske samtaler med primært fokus på usikkerhedsvurdering og naturlig sprogbehandling.
+Denne afhandling udforsker maskinlæringsteknikker til medicinske samtaler med primært fokus på usikkerhedsvurdering og naturlig sprogbehandling.
 Den første del undersøger usuperviserede metoder til at detektere ude-af-fordeling data og modellering af tale ved hjælp af variationelle autoindkodere, og giver følgende bidrag:

### 2023-10-06 chapters/discussion.tex
Some work on discussion

@@
 
-Though more than three years have passed since the start of this project, the rapid development of machine learning that started about a decade ago with the ImageNet competation and the work of \textcite{krizhevsky_imagenet_2012} did all but slow down.
-In this section we will briefly revisit selected chapters of the thesis and provide some additional thoughts and discussion
+During the three years since the start of this project, the rapid development of machine learning that started about a decade ago with the ImageNet competation and the work of \textcite{krizhevsky_imagenet_2012} all but slowed down.
+In this section we will briefly revisit selected chapters of the thesis and provide some additional thoughts and discussion.
 
 \section{Speech representation learning}
+%
+In this thesis we have studied two different approaches to learning speech representations.
+Variational autoencoders in \cref{chp:paper-hierarchical,chp:paper-modelagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-review}.
+We found that the probabilistic formulation of variational autoencoders provide benefits for their application to uncertainty quantification but that they are generally inferior to self"=supervised methods when it comes to performance on downstream tasks such as phoneme recognition (see e.g.
+\cref{tab: phoneme recognition (PER), tab:unsupervisedASR}).
+
+VAEs are good at uncertainty quantification (let's at least say that they are).
+
+SSL methods are very good at representation learning as reflected by their dominance on downstream task performance in for text and speech.
+
+It seems like we cannot have a model that has both a principled probabilistic representation of uncertainty and the ability to learn rich representations that are useful for downstream applications.
+
+\subsection{Improved representation learning with variational autoencoders}
+
+One way to improve the representations learned in VAEs is via semi-supervised learning. 
+Here a small number of curated labels are used to inform and enrich the patterns that can be learned from a large unlabelled data set. 
+VAEs are strong for semi-supervised learning but there are other strong alternatives too \cite{jiang_speech_2021}.
+Additionally, semi-supervised learning is more constraining than is self-supervised methods that rely only on pretraining; a single foundation model can be fine-tuned at relatively low cost for a large number of downstream tasks, while semi-supervised models must learn all relevant downstream tasks while simultaneously learning from the unlabelled data, which is expensive.
+
+\subsection{Uncertainty estimation with self-supervised methods}
+%
+
+\textcite{pasad_layerwise_2021} found that layers 6-7 hold the most information about phonetic content and word identity and meaning for \texttt{wav2vec2-base}. 
+For \texttt{wav2vec2-large} the phonetic content is highest in layers 11 and 18-19 with a drop in between, while word identity and meaning are highest between layers 12 and 18.
+
+(layers )
 
 Masking in variational autoencoders
+
+Masked pretrained as maximum likelihood
 
@@
 
+\begin{figure}
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curve_ensemble_and_all_models_uncalibrated.pdf}
+        % \caption{}
+        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/histogram_ensemble_and_single_model.pdf}
+        % \caption{}
+        % \label{fig_discussion:histogram_ensemble_and_single_model}
+    \end{subfigure}
+    \caption[Calibration curve for the uncalibrated stroke recognition model and empirical distribution of predicted probabilities.]{%
+        Calibration curve for the uncalibrated stroke recognition ensemble model (left) and the histogram of predicted probabilities (right). 
+        We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
+    \label{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}
+\end{figure}
+
 It is interesting to note that the ensemble model is better calibrated than the individual models that form it.
@@
 Relative improvement in Brier score compared to uncalibrated (Brier skill score): Logistic = $0.4830$, isotonic = $0.4924$.} and that the predicted probabilities are shifted towards smaller values.
+
+\begin{figure}
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curves_ensemble.pdf}
+        % \caption{}
+        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_fits_ensemble.pdf}
+        % \caption{}
+        % \label{fig_discussion:histogram_ensemble_and_single_model}
+    \end{subfigure}
+    \caption[Calibration fits and curves for the stroke recognition model using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right).
+We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
+    \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
+\end{figure}
 
@@
 
-\begin{figure}
-    \begin{subfigure}[c]{0.48\columnwidth}
-        \centering
-        \includegraphics[width=1\columnwidth]{python_plotting/calibration_curve_ensemble_and_all_models_uncalibrated.pdf}
-        % \caption{}
-        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
-    \end{subfigure}
-    % \hfill
-    \begin{subfigure}[c]{0.48\columnwidth}
-        \centering
-        \includegraphics[width=1\columnwidth]{python_plotting/histogram_ensemble_and_single_model.pdf}
-        % \caption{}
-        % \label{fig_discussion:histogram_ensemble_and_single_model}
-    \end{subfigure}
-    \caption[Calibration curve for the uncalibrated stroke recognition model and empirical distribution of predicted probabilities.]{%
-        Calibration curve for the uncalibrated stroke recognition ensemble model (left) and the histogram of predicted probabilities (right). 
-        We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
-    \label{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}
-\end{figure}
-
-\begin{figure}
-    \begin{subfigure}[c]{0.48\columnwidth}
-        \centering
-        \includegraphics[width=1\columnwidth]{python_plotting/calibration_curves_ensemble.pdf}
-        % \caption{}
-        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
-    \end{subfigure}
-    % \hfill
-    \begin{subfigure}[c]{0.48\columnwidth}
-        \centering
-        \includegraphics[width=1\columnwidth]{python_plotting/calibration_fits_ensemble.pdf}
-        % \caption{}
-        % \label{fig_discussion:histogram_ensemble_and_single_model}
-    \end{subfigure}
-    \caption[Calibration fits and curves for the stroke recognition model using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right).
-We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
-    \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
-\end{figure}
-
 % \lesstodo[inline]{Discuss the calibration of the stroke model and report calibration curve.}

### 2023-10-06 chapters/introduction.tex
Some work on discussion

@@
 While any model will inevitably make a wrong prediction, by accurately estimating the uncertainty it can communicate to other systems or humans when it is not to be trusted, and so, reduce the risk that wrong predictions lead to catastrophic failures.
-Fittingly, a recent study by the European Parliamentary Research Services concluded:
-
-\begin{center}
-
-\textit{"Future AI solutions for healthcare should be implemented by integrating uncertainty estimation, a relatively new field of research that aims to provide clinicians with clinically useful indications on the degree of confidence in AI predictions"} \parencite{europeanparliament_artificial_2022}.
-
-\end{center}
-
+Fittingly, a recent study by the European Parliamentary Research Services \parencite{europeanparliament_artificial_2022} concluded: 
+%
+\begin{quote}
+    \centering\itshape
+    Future AI solutions for healthcare should be implemented by integrating uncertainty estimation, a relatively new field of research that aims to provide clinicians with clinically useful indications on the degree of confidence in AI predictions.
+\end{quote}
+%
 \vspace{0.5em}
@@
 
-\Cref{part:unsupervised-uncertainty-estimation,part:self"=supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. 
+\Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. 
 \Cref{part:unsupervised-uncertainty-estimation} is made up of three papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
-\Cref{part:self"=supervised-speech-representation-learning} is a single, comprehensive review paper on self"=supervised speech representation learning \parencite{mohamed_selfsupervised_2022}. 
+\Cref{part:self-supervised-speech-representation-learning} is a single, comprehensive review paper on self"=supervised speech representation learning \parencite{mohamed_selfsupervised_2022}. 
 \Cref{part:medical-applications} consists of two papers on applications of machine learning within the medical domain, specifically recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}.

### 2023-10-06 chapters/paper_review/05_zero.tex
Some work on discussion

@@
 \centering
-\caption[Performance of models within unsupervised ASR.]{
+\caption[Performance of self"=supervised models for unsupervised ASR.]{
 	Unsupervised ASR.
@@
 	\textcolor{black}{The references in the table are sorted according to the date of publication.}}
-	\label{fig:unsupervisedASR}
+	\label{tab:unsupervisedASR}
 	\resizebox*{\textwidth}{!}{%
@@
 generate its transcription, is learned from data. 
-\Cref{fig:unsupervisedASR} summarizes recent work on unsupervised ASR,
+\Cref{tab:unsupervisedASR} summarizes recent work on unsupervised ASR,
 including the speech representation used, the algorithm used to learn the
@@
 translation, and spoken word retrieval.
-\Cref{fig:unsupervisedASR} summarizes the various ways to segment speech and represent speech and text in each reference.
+\Cref{tab:unsupervisedASR} summarizes the various ways to segment speech and represent speech and text in each reference.
 
@@
 % more improvement: iterative training, development 
-As shown in \cref{fig:unsupervisedASR}, most studies use
+As shown in \cref{tab:unsupervisedASR}, most studies use
 \textit{self-training} to refine the models.

### 2023-10-06 chapters/background.tex

@@
 %
-\begin{center}
-    \textit{Uncertainty is the lack of certainty; a state of limited knowledge where it is impossible to exactly describe the existing state, a future outcome, or more than one possible outcome.}
-\end{center}
+\begin{quote}
+    \centering\itshape
+    Uncertainty is the lack of certainty; a state of limited knowledge where it is impossible to exactly describe the existing state, a future outcome, or more than one possible outcome.
+\end{quote}
 %
@@
 Despite the challenges facing the training of VAEs, they have been successfully applied to a number of tasks including image generation \parencite{kingma_autoencoding_2014,rezende_stochastic_2014}, image inpainting \parencite{pathak_context_2016}, image super-resolution \parencite{sonderby_amortised_2017,chira_image_2022}, and speech synthesis \parencite{hsu_unsupervised_2017,hsu_hierarchical_2019}.
-VAEs have also proven themselves useful for semi"=supervised learning \parencite{kingma_semi"=supervised_2014,kingma_improved_2016,maaloe_biva_2019}.
+VAEs have also proven themselves useful for semi"=supervised learning \parencite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}.
 

### 2023-10-06 chapters/main-contributions.tex

@@
 For that reason, we here lay out research hypotheses and contributions in relation to the overall research project. 
-% The chapters are structured into three parts: {\color{black}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{black}\cref{part:self"=supervised-speech-representation-learning}} is concerned with self"=supervised speech representation learning, and {\color{black}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.
+% The chapters are structured into three parts: {\color{black}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{black}\cref{part:self-supervised-speech-representation-learning}} is concerned with self"=supervised speech representation learning, and {\color{black}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.
 Finally, {\color{black}\cref{part:discussion-and-conclusion}} contains a discussion and the conclusion of the thesis.
@@
 
-\contribitem{part:self"=supervised-speech-representation-learning}
+\contribitem{part:self-supervised-speech-representation-learning}
 This part of the thesis is concerned with self"=supervised speech representation learning and consists of a single paper that provides a substantial review of the field \parencite{mohamed_selfsupervised_2022}.

### 2023-10-06 chapters/paper_benchmarking.tex

@@
 Approximate posteriors have various applications e.g.
-semi"=supervised learning \parencite{kingma_semi"=supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.
+semi"=supervised learning \parencite{kingma_semi-supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.
 
@@
     \caption[Model likelihoods and phoneme error rate for TIMIT.]{We report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines (left) and phoneme error rate (PER) of different representations for phoneme recognition on TIMIT (right).}
-    \begin{subtable}[t]{.45\textwidth}
+    \hfill
+    \begin{subtable}[t]{.49\textwidth}
     \centering
-    \begin{adjustbox}{center, totalheight=0.35\paperheight}
+    \begin{adjustbox}{center, totalheight=0.4\paperheight}
     \begin{tabular}[t]{lll|r}
@@
     \hfill
-    \begin{subtable}[t]{.45\textwidth}
+    \begin{subtable}[t]{.49\textwidth}
     \centering
-    \begin{adjustbox}{center, totalheight=0.35\paperheight}
+    \begin{adjustbox}{center, totalheight=0.4\paperheight}
     \begin{tabular}[t]{cll|c}
@@
 \end{subtable}%
+\hfill
 \end{table}
@@
 \paragraph{TIMIT}
-For temporal resolutions of $s=1$, the deterministic autoregressive models yield the best likelihoods with WaveNet achieving $\SI{10.88}{bpf}$ on TIMIT as seen in \cref{tab: likelihoods timit}. 
+For temporal resolutions of $s=1$, the deterministic autoregressive models yield the best likelihoods with WaveNet achieving $\SI{10.88}{bpf}$ on TIMIT as seen in \cref{tab: likelihoods timit} (left).
 Somewhat surprisingly, the LSTM baseline almost matches WaveNet with a likelihood of $\SI{11.11}{bpf}$ at $s=1$.
-However, due to being autoregressive in training, the LSTM trains considerably slower than the parallel convolutional WaveNet; something not conveyed by \cref{tab: likelihoods timit}.
+However, due to being autoregressive in training, the LSTM trains considerably slower than the parallel convolutional WaveNet; something not conveyed by \cref{tab: likelihoods timit} (left).
 Notably, the VRNN and SRNN models achieve likelihoods close to that of WaveNet and the LSTM at around $\SI{11.09}{bpf}$. The STCN exhibited instability when trained at $s=1$ and tended to diverge.
@@
 Although both the autoregressive models and the LVMs are lossy, their objective minimizes the amount of incurred loss. 
-The best likelihoods reported in tables~\cref{tab: likelihoods timit} and \cref{tab: likelihoods librispeech} correspond to about a 30\% reduction in bit rate which indicates that there are significant gains in likelihood to be made in speech modeling.
+The best likelihoods reported in \cref{tab: likelihoods timit} (left) and \cref{tab: likelihoods librispeech} correspond to about a 30\% reduction in bit rate which indicates that there are significant gains in likelihood to be made in speech modeling.
 
@@
 It is trained with the connectionist temporal classification (CTC) loss \parencite{graves_connectionist_2006} which lets it jointly learn to align and classify without using label alignments.
-We pre-train all unsupervised models at $s=64$ on the full TIMIT training dataset excluding the validation data (3.7h) as in \cref{tab: likelihoods timit}. 
-We then train the ASR model on all 3.7h as well as 1h and 10m subsets. We report results on the test set in terms of phoneme error rates (PER) in \cref{tab: phoneme recognition (PER)}.
+We pre-train all unsupervised models at $s=64$ on the full TIMIT training dataset excluding the validation data (3.7h) as in \cref{tab: likelihoods timit} (left). 
+We then train the ASR model on all 3.7h as well as 1h and 10m subsets. We report results on the test set in terms of phoneme error rates (PER) in \cref{tab: phoneme recognition (PER)} (right).
 

### 2023-10-08 chapters/conclusion.tex
Conclusion for chapter 1

@@
 In the introductory \cref{chp:introduction} we laid out how, through human history, progress-driving technological development has always introduced new challenges and risks of error whether via misuse, misunderstanding or inherent limitations.
-We emphasized that machine learning comes with these same types of risk may in some cases may amplify their impacts and argued that uncertainty estimation is a key component in developing safe and reliable machine learning systems.
-We provided the motivational cases of recognition of stroke cases in emergency calls \cref{subsec:motivation-stroke-recognition} and automated medical coding \cref{subsec:motivation-medical-coding} and gave examples of potential sources of uncertainty and of how correctly quantifying uncertainty can improve the usefulness of such tools in practice.
-Finally, we ended the introduction by connecting machine learning reliability \cref{sec:machine-learning-reliability} and model calibration \cref{subsec:model-calibration} and defining different types of uncertainty \cref{subsec:understanding-uncertainty}.
+We emphasized that machine learning comes with these same types of risk, in some cases amplifying their impacts, and argued that uncertainty estimation is a key component in ensuring that systems are safe and reliable in practice.
+We provided background for the research project by introducing the healthtech company Corti, with which with project has been defined and carried out, and the motivational cases of recognition of stroke cases in emergency calls (\cref{subsec:motivation-stroke-recognition}) and automated medical coding (\cref{subsec:motivation-medical-coding}).
+We described how machine learning systems could be designed to aid healthcare professionals with the tasks and gave examples of potential sources of uncertainty and of how correctly quantifying uncertainty might improve the usefulness of such systems.
+Finally, we connected machine learning reliability (\cref{sec:machine-learning-reliability}) and model calibration (\cref{subsec:model-calibration}) and defined the aleatoric, epistemic, and predictive types of uncertainty (\cref{subsec:understanding-uncertainty}).
 

### 2023-10-10 appendices/paper_brief.tex
Updates to discussion

@@
 Unsupervised representation learning is concerned with learning useful representations without the use of human annotations.
-Usually, a model is first pre-trained on a task where plenty of data is available.
-The model is then fine-tuned, or used to extract input representations for a smaller model, targeting a task with limited training data.
+Usually, a model is first pre"=trained on a task where plenty of data is available.
+The model is then fine"=tuned, or used to extract input representations for a smaller model, targeting a task with limited training data.
 In computer vision, both supervised \parencite{simonyan_very_2014, szegedy_going_2015, he_deep_2015} and unsupervised \parencite{pathak_context_2016, doersch_unsupervised_2015} representation learning have gained attention with supervised representation learning driven by the availability of large annotated datasets \parencite{deng_imagenet_2009}.
@@
     \includegraphics[width=0.90\textwidth]{paper_brief/SSL_LVM.pdf}
-    \caption[ schematic overview of the self"=supervised and probabilistic latent variable models covered in this survey.]{A schematic overview of the two groups of models covered in this survey.
+    \caption[ Schematic overview of the self"=supervised and probabilistic latent variable models covered in this survey.]{A schematic overview of the two groups of models covered in this survey.
 \textit{Left}: A model trained with self"=supervised learning.
 We take these models to consist of two functions $f(\cdot)$ and $g(\cdot)$ (\cref{ssec:notation}).
-After pre"=training, $f(\cdot)$ is fine-tuned or used for extracting features $\mathbf{c}$.
+After pre"=training, $f(\cdot)$ is fine"=tuned or used for extracting features $\mathbf{c}$.
 $g(\cdot)$ is an auxiliary function used to accommodate the self"=supervised pre"=training task.
@@
 For stochastic latent variable models, we use $\mathbf{z}_{1:T}$ as is customary to the field.
-While some models are frozen and produce representations used as input for downstream tasks (\textbf{\textsc{frz}}, \cref{tab:model-taxonomy}), others are designed to be fine-tuned (\textbf{\textsc{ftn}}, \cref{tab:model-taxonomy}).
+While some models are frozen and produce representations used as input for downstream tasks (\textbf{\textsc{frz}}, \cref{tab:model-taxonomy}), others are designed to be fine"=tuned (\textbf{\textsc{ftn}}, \cref{tab:model-taxonomy}).
 In either case, we use $f(\cdot)$ to denote the model that is used for the downstream task.
@@
 
-Combining masking with a contrastive loss, wav2vec 2.0 was the first work to show that a competitive speech recognition model can be learned by fine-tuning a pre-trained model with as little as 10 minutes of labeled data.
+Combining masking with a contrastive loss, wav2vec 2.0 was the first work to show that a competitive speech recognition model can be learned by fine"=tuning a pre"=trained model with as little as 10 minutes of labeled data.
 For this model
@@
 
-\begin{sidewaystable*}
+\begin{table}
     \caption[Classification of selected self"=supervised and probabilistic latent variable models.]{ Selected models classified according to the binary attributes identified throughout the text.
 The models are sorted according to first publication date on arXiv which might differ from the citation year.
-\textbf{\textsc{msk}}: masking, \textbf{\textsc{prd}}: prediction, \textbf{\textsc{con}}: contrastive, \textbf{\textsc{rec}}: reconstruction, \textbf{\textsc{qtz}}: quantization, \textbf{\textsc{gen}}: generative, \textbf{\textsc{frz}}: frozen, \textbf{\textsc{ftn}}: fine-tuned, \textbf{\textsc{loc}}: local, \textbf{\textsc{glo}}: global.}
+\textbf{\textsc{msk}}: masking, \textbf{\textsc{prd}}: prediction, \textbf{\textsc{con}}: contrastive, \textbf{\textsc{rec}}: reconstruction, \textbf{\textsc{qtz}}: quantization, \textbf{\textsc{gen}}: generative, \textbf{\textsc{frz}}: frozen, \textbf{\textsc{ftn}}: fine"=tuned, \textbf{\textsc{loc}}: local, \textbf{\textsc{glo}}: global.}
     \label{tab:model-taxonomy}
     \begin{center}
+        \setlength{\tabcolsep}{3pt}
         \renewcommand{\arraystretch}{1.1}
@@
             \multicolumn{2}{c}{} & 
-            \multicolumn{6}{c}{\footnotesize\textsc{Model and task design}} & 
-            \multicolumn{3}{c}{\footnotesize\textsc{Resolution}} &
-            \multicolumn{2}{c}{\footnotesize\makebox[0pt][c]{\textsc{Usage}}} \\
+            \multicolumn{6}{c}{\textsc{Model and task design}} & 
+            \multicolumn{3}{c}{\textsc{Resolution}} &
+            \multicolumn{2}{c}{\makebox[0pt][c]{\textsc{Usage}}} \\
             \textbf{\textsc{model}} &
@@
     \end{center}
-\end{sidewaystable*}
+\end{table}
 
@@
 $\mathbf{p}$ is mapped to a one-hot vector using a function $\varphi(\cdot)$, such that $\varphi(\mathbf{p})_i = 1$ if $i = \argmax_j p_j$ and $0$ otherwise. 
-As this function is non-differentiable, we must rely on the straight-through gradient estimator \parencite{bengio_estimating_2013} which assumes that the Jacobian ${\partial \, \varphi}/{\partial \, \mathbf{p}}$ equals the identity matrix.
+As this function is non-differentiable, we must rely on the straight-through gradient estimator \parencite{bengio_estimating_2013} which assumes that the Jacobian ${\partial \varphi}/{\partial \mathbf{p}}$ equals the identity matrix.
 The one-hot vector can then be used for a codebook lookup to obtain the final quantized vector (e.g., $\mathbf{q}_t$ in \cref{eq_brief: w2v2 qtz}).
@@
 
-\begin{sidewaystable*}
+\begin{table*}
     \caption[Classification of selected latent variable models.]{ Selected latent variable models classified according the attributes defined throughout \cref{sec:plvms}.
@@
     \begin{center}
+        \setlength{\tabcolsep}{3pt}
         \renewcommand{\arraystretch}{1.1}
-        \resizebox{0.99\textwidth}{!}{% <------ Don't forget this %
+        \resizebox{1\textwidth}{!}{%
         \begin{tabular}{ l l | c c c | c c c c | c c c c c | c } 
@@
             \multicolumn{2}{c}{} & 
-            \multicolumn{3}{c}{\footnotesize\textsc{Observation}} & 
-            \multicolumn{4}{c}{\footnotesize\textsc{Prior}} & 
-            \multicolumn{5}{c}{\footnotesize\textsc{Inference}} \\
+            \multicolumn{3}{c}{\textsc{Observation}} & 
+            \multicolumn{4}{c}{\textsc{Prior}} & 
+            \multicolumn{5}{c}{\textsc{Inference}} \\
             \textbf{\textsc{model}} & 
@@
     \end{center}
-\end{sidewaystable*}
+\end{table*}
 

### 2023-10-10 appendices/supplementary_retrospective.tex
Updates to discussion

@@
 
-\section{Additional results: Fine-tuning of Danish BERT model for stroke recognition}
+\section{Additional results: Fine"=tuning of Danish BERT model for stroke recognition}
 %
-During the preliminary experimental phase, we fine-tuned a BERT model pre-trained on Danish text from CommonCrawl, Wikipedia, OpenSubtitles, and other Danish online forums (available at \url{https://github.com/certainlyio/nordic_bert}).
-The model was fine-tuned for 10,000 updates using linear learning rate warm-up (1,000 updates) and decay (9,000 updates). 
+During the preliminary experimental phase, we fine"=tuned a BERT model pre"=trained on Danish text from CommonCrawl, Wikipedia, OpenSubtitles, and other Danish online forums (available at \url{https://github.com/certainlyio/nordic_bert}).
+The model was fine"=tuned for 10,000 updates using linear learning rate warm-up (1,000 updates) and decay (9,000 updates). 
 The maximum learning rate was set to $5\times 10^{-5}$ and an accumulated batch size of 128. 
-The maximum sequence length of the pre-trained model was 512 input tokens. 
+The maximum sequence length of the pre"=trained model was 512 input tokens. 
 To accommodate longer input sequences, which was necessary for our dataset, we concatenated several copies of the original positional embedding matrix. 
@@
 
-We see that the fine-tuned BERT model performs slightly worse across F1-score, sensitivity, PPV and FOR (p < 0.0001), but better in terms of FPR (p < 0.0001), compared the MLP model presented in the main manuscript.
+We see that the fine"=tuned BERT model performs slightly worse across F1-score, sensitivity, PPV and FOR (p < 0.0001), but better in terms of FPR (p < 0.0001), compared the MLP model presented in the main manuscript.
 As described in the discussion section, we hypothesize that the number of stroke positives was too small for these advanced models to learn more complex patterns than the MLP ensemble.
@@
     \centering
-    \caption[Overall performance on MH-1813 test data for a fine-tuned BERT model.]{Overall performance on MH-1813 test data for a fine-tuned BERT model.
+    \caption[Overall performance on MH-1813 test data for a fine"=tuned BERT model.]{Overall performance on MH-1813 test data for a fine"=tuned BERT model.
 We also list the original performance of call-takers and the ensemble model from the main manuscript for ease of comparison [mean (95\% CI)].}
@@
         \midrule
-        \makecell[l]{BERT \\ (fine-tuned)}      & \makecell[c]{33.8 \\ (31.5-36.2)} & \makecell[c]{57.5 \\ (53.9-60.9)} & \makecell[c]{23.9 \\ (21.9-25.9)} & \makecell[c]{0.094 \\ (0.084-0.104)} & \makecell[c]{0.403 \\ (0.381-0.424)} \\
+        \makecell[l]{BERT \\ (fine"=tuned)}      & \makecell[c]{33.8 \\ (31.5-36.2)} & \makecell[c]{57.5 \\ (53.9-60.9)} & \makecell[c]{23.9 \\ (21.9-25.9)} & \makecell[c]{0.094 \\ (0.084-0.104)} & \makecell[c]{0.403 \\ (0.381-0.424)} \\
 

### 2023-10-10 chapters/background.tex
Updates to discussion

@@
 We then introduce the task of out-of-distribution detection and review the existing literature on the topic.
-Then, we describe variational autoencoders, how they can be used to learn representations of speech, and some of their challenges.
-Finally, we introduce the concept of self"=supervised learning and highlight important differences to variational autoencoders.
+Finally, we describe variational autoencoders, how they can be used to learn representations of speech, and some of their challenges.
+%Finally, we introduce the concept of self"=supervised learning and highlight important differences to variational autoencoders.
 
@@
 In this view, information entropy is a quantity that distills the distribution of a random variable into a single number that describes the diversity of the potential outcomes of the random variable.
+
+The Kullback-Leibler divergence plays a central role in the training of variational autoencoders, which we introduce in \cref{sec:variational-autoencoders}.
+In \cref{chp:paper-hierarchical} we shall see how the Kullback-Leibler divergence emerges in a likelihood-ratio test statistic for out-of-distribution detection using hierarchical variational autoencoders.
 % In the following, we will see that this interpretation of information entropy as a measure of diversity is useful for understanding the concept of uncertainty in machine learning.
@@
 \textcite{ruff_deep_2020} use semi"=supervised learning and learn representations of in-distribution data that concentrate close to a centroid in latent space while labelled outliers are pushed away from the centroid.
-Other methods including MCD \parencite{yu_unsupervised_2019}, NGC \parencite{wu_ngc_2021}, and UDG \parencite{yang_semantically_2021} utilize external, unlabeled, noisy data to improve OOD detection performance without requiring cleanly labelled OOD examples.
+Other methods including MCD \parencite{yu_unsupervised_2019}, NGC \parencite{wu_ngc_2021}, and UDG \parencite{yang_semantically_2021} use external, unlabeled, noisy data to improve OOD detection performance without requiring cleanly labelled OOD examples.
 As we discussed earlier, for many modern applications of machine learning, input data is often high dimensional and complex which makes it difficult to obtain enough representative OOD data to ensure robust OOD detection capabilities.
@@
 NuSA \parencite{cook_outlier_2020} uses projects features onto the column space of the classification weight matrix and computes the ratio of the norm the projected and original features.
-\textcite{lee_simple_2018} fit a multivariate Gaussian distribution to the activations of the penultimate layer of a pre-trained classifier and use the Mahalanobis distance to this distribution to evaluate whether inputs are OOD.
+\textcite{lee_simple_2018} fit a multivariate Gaussian distribution to the activations of the penultimate layer of a pre"=trained classifier and use the Mahalanobis distance to this distribution to evaluate whether inputs are OOD.
 This method can also be seen as ameliorating the curse of dimensionality by clustering the high-dimensional feature space.
@@
 %     X sehwag_ssd_2021,                      learn a representation over the in-domain data through self"=supervised training
-%     X xiao_we_2021                          further shows that one can instead use a strong pretrained feature extractor (self"=supervised)  while maintaining comparable performance.
+%     X xiao_we_2021                          further shows that one can instead use a strong pre"=trained feature extractor (self"=supervised)  while maintaining comparable performance.
 % }
@@
 They propose to measure the Mahalanobis distance between the global Gaussian distribution of training set in latent space and an encoded test input.
-\textcite{xiao_we_2021} instead propose to use an existing, strong foundation model, pre-trained with a self"=supervised objective, to extract features from the input, and then fit a Gaussian Mixture Model to the features using the minimal Mahalanobis distance to the mixture components as the OOD score.
+\textcite{xiao_we_2021} instead propose to use an existing, strong foundation model, pre"=trained with a self"=supervised objective, to extract features from the input, and then fit a Gaussian Mixture Model to the features using the minimal Mahalanobis distance to the mixture components as the OOD score.
 
@@
     &\geq \int q_\phib(\zb|\xb) \log \left( \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right) \,\textrm{d}\zb \nonumber \\
-    &\geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] \triangleq \mathcal{L}_{\thetab,\phib}(\xb) \enspace .
+    &\geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}_{\thetab,\phib}(\xb) \enspace .
 \end{align}
@@
     \mathcal{L}_{\thetab,\phib}(\xb)
-    % &\triangleq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
+    % &\equiv \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
     &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right] + \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\

### 2023-10-10 chapters/discussion.tex
Updates to discussion

@@
 
-During the three years since the start of this project, the rapid development of machine learning that started about a decade ago with the ImageNet competation and the work of \textcite{krizhevsky_imagenet_2012} all but slowed down.
+During the three years since the start of this project, the rapid development of machine learning that started about a decade ago with the 2012 ImageNet competation and the work of \textcite{krizhevsky_imagenet_2012} all but slowed down.
 In this section we will briefly revisit selected chapters of the thesis and provide some additional thoughts and discussion.
 
-\section{Speech representation learning}
+\section{Speech representations and uncertainty}
 %
-In this thesis we have studied two different approaches to learning speech representations.
-Variational autoencoders in \cref{chp:paper-hierarchical,chp:paper-modelagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-review}.
-We found that the probabilistic formulation of variational autoencoders provide benefits for their application to uncertainty quantification but that they are generally inferior to self"=supervised methods when it comes to performance on downstream tasks such as phoneme recognition (see e.g.
+In this thesis we have studied two different approaches to learning speech representations: VAEs in \cref{chp:paper-hierarchical,chp:paper-modelagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-review}.
+We found that the probabilistic formulation of VAEs provide benefits for their application to uncertainty quantification but that they are generally inferior to self"=supervised methods when it comes to performance on downstream tasks, such as phoneme recognition (see e.g.
 \cref{tab: phoneme recognition (PER), tab:unsupervisedASR}).
+Considering VAEs and self"=supervised methods, it seems we cannot build a model that has a principled probabilistic representation of uncertainty without sacrificing some of the ability to learn rich representations that are useful for downstream applications.
 
-VAEs are good at uncertainty quantification (let's at least say that they are).
-
-SSL methods are very good at representation learning as reflected by their dominance on downstream task performance in for text and speech.
-
-It seems like we cannot have a model that has both a principled probabilistic representation of uncertainty and the ability to learn rich representations that are useful for downstream applications.
+% VAEs are good at uncertainty quantification (let's at least say that they are).
+% SSL methods are very good at representation learning as reflected by their dominance on downstream task performance in for text and speech.
 
@@
 One way to improve the representations learned in VAEs is via semi-supervised learning. 
-Here a small number of curated labels are used to inform and enrich the patterns that can be learned from a large unlabelled data set. 
-VAEs are strong for semi-supervised learning but there are other strong alternatives too \cite{jiang_speech_2021}.
-Additionally, semi-supervised learning is more constraining than is self-supervised methods that rely only on pretraining; a single foundation model can be fine-tuned at relatively low cost for a large number of downstream tasks, while semi-supervised models must learn all relevant downstream tasks while simultaneously learning from the unlabelled data, which is expensive.
+Here a small number of labels are used to inform which patterns that are learned from a large, mostly unlabelled, data set. 
+This is usually done by defining a new stochastic variable as the target and deriving a semi-supervised version of the ELBO that accomadates using the labels when they are available, or marginalizing the target variable when they are not.
+The VAE is then trained on the labelled and unlabelled data simultanously.
+
+VAEs are strong models for semi-supervised learning as proven in several works \parencite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}. 
+However, self"=supervised methods methods have established themselves as superior also in this setting \cite{jiang_speech_2021, liu_learning_2023}. 
+Furthermore, by training on labelled and unlabelled data simultaneously, the semi-supervised setting of VAEs is more constraining than it is for self-supervised methods that divide the training into pre"=training and finetuning.
+Here, expensive pre"=training yields a single foundation model that can be fine"=tuned at relatively low cost for a large number of downstream tasks, while VAEs must learn all relevant downstream tasks while also learning from the unlabelled data, which is more expensive, but also requires retraining on the unlabelled data, when a .
+
+Masking in variational autoencoders
+
+Masked pre"=trained as maximum likelihood
 
@@
 For \texttt{wav2vec2-large} the phonetic content is highest in layers 11 and 18-19 with a drop in between, while word identity and meaning are highest between layers 12 and 18.
-
-(layers )
-
-Masking in variational autoencoders
-
-Masked pretrained as maximum likelihood
 
@@
 \section{\Cref{chp:paper-retrospective} revisited: Calibration}
-The work in \textcite{wenstrup_retrospective_2023} focuses on the predictive performance of the ensemble model and an analysis of feature importance, but does not explicitly consider uncertainty estimation.
+Our work on stroke recognition in \cref{chp:paper-retrospective} focuses on the predictive performance of the ensemble model and an analysis of feature importance, but does not explicitly consider uncertainty estimation.
 As we discussed in \cref{subsec:model-calibration}, such a model can be calibrated to predict probabilities that are aligned with the empirical probability of the model being correct on some validation set.
+Here we present and discuss model calibration for the ensemble model.
 
-We compute the calibration curve for the stroke recognition ensemble by sorting the predicted probabilities on test set into a number of bins spanning the range from zero to one.
-For each bin, we compute the fraction of examples for which the model predicted correctly and then plot these pairs of mean predicted probabilities and correctness fractions.
-A perfectly calibrated model would have the same fraction of correct predictions in any given bin, as that bin's mean value.
-In \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}, we plot the calibration curve for the uncalibrated stroke recognition ensemble and the individual models along with a histogram of its predicted probabilities.
+We compute the calibration curve by sorting the probabilities predicted on the test set into a number of bins spanning the range from zero to one.
+For each bin $b$, we compute the mean predicted probability $\bar{p}$ and the fraction of examples for which the model predicted correctly $r_{b}$.
+The calibration curve is the drawn from the $\{(\bar{p}_{b}, r_{b})\}_b$ pairs.
+A perfectly calibrated model would have the same fraction of correct predictions in any given bin, as that bin's mean value, $\bar{p}_{b} = r_{b}\,\forall\,b$.
+
+The calibration curve for the uncalibrated stroke recognition ensemble and its constituent models is plotted in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model} along with a histogram of its predicted probabilities.
 The miscalibration issue that we previously discussed is clearly visible as a strong overconfidence.
+Additionally, the ensemble model is appreciably better calibrated than its constituent models.
+Since the ensemble output probability is computed as the harmonic mean of the five constituent model probabilities, the ensemble's output probability can never exceed the maximum probability predicted between the constituent models.
+This property tends to make ensemble probabilities less extreme and, since the constituent models are overconfident, this results in better calibration (see also the histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).
 
@@
     \caption[Calibration curve for the uncalibrated stroke recognition model and empirical distribution of predicted probabilities.]{%
-        Calibration curve for the uncalibrated stroke recognition ensemble model (left) and the histogram of predicted probabilities (right). 
-        We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
+        Calibration curve for the uncalibrated stroke recognition ensemble model (left) and the histogram of predicted probabilities (right) for the test set. 
+        We use the ensemble that achieved the median F1-score reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
     \label{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}
@@
 
-It is interesting to note that the ensemble model is better calibrated than the individual models that form it.
-Since an ensemble output probability is computed as the harmonic mean of the five individual model probabilities, it is easy to see that the ensembles output probability can never exceed the maximum probability predicted among the individual models.
-This property tends to make ensemble probabilities less extreme and, since the individual models are overconfident, this results in better calibration (see also the histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).
+To calibrate the ensemble model, we can use methods such as Platt-scaling \parencite{platt_probabilistic_1999} or isotonic regression \parencite{zadrozny_transforming_2002}. 
+We fit a simple regression model (logistic or isotonic) to the predicted probabilities and the target labels on the validation set and use it to adjust the probabilities predicted on the test set.
+We show the resulting calibration curves on the left in \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic} and the logistic and isotonic fits on the right.
+% In \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic} we have done so for the ensemble model and visualize the results on the test set. 
+% On the left, we plot the resulting calibration curves and on the right we show the logistic and isotonic fits. 
+We see that both methods result in quite good calibrations\footnote{Brier scores on test set: Uncalibrated = $0.003500$, logistic = $0.001807$, isotonic = $0.001774$.
+Relative improvement in Brier score compared to uncalibrated (Brier skill score): Logistic = $0.4830$, isotonic = $0.4924$.} and that the predicted probabilities are shifted towards smaller values.
+Since there are few stroke cases overall, few examples receive a high predicted probability (see histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).
+This leads to a lack of data for the calibration fits at high predicted probabilities which can be seen to especially affect the nonparametric isotonic regression.
 
-To calibrate the ensemble model, we can use methods such as Platt-scaling \parencite{platt_probabilistic_1999} or isotonic regression \parencite{zadrozny_transforming_2002}. 
-In both cases, we fit a simple regression model (logistic or isotonic) to the predicted probabilities and the target labels on some validation set. 
-In \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic} we have done so for the ensemble model using our validation set for fitting and visualizing the results on the test set.
-On the left, we plot the resulting calibration curves and on the right we show the logistic and isotonic fits.
-We see that both methods result in quite good calibrations\footnote{Test set Brier scores: Uncalibrated = $0.003500$, logistic = $0.001807$, isotonic = $0.001774$.
-Relative improvement in Brier score compared to uncalibrated (Brier skill score): Logistic = $0.4830$, isotonic = $0.4924$.} and that the predicted probabilities are shifted towards smaller values.
+% Brier scores:  {'uncalibrated': 0.0034955788687128925, 'logistic': 0.0018072483306197486, 'isotonic': 0.0017744177396100578, 'average': 0.0021955477869598783}
+% BSS uncalibrated reference:  {'logistic': 0.48299025755205005, 'isotonic': 0.4923822902432705}
+% BSS mean reference:  {'logistic': 0.17685766561145932, 'isotonic': 0.1918109229282361}
 
@@
     \end{subfigure}    
-    \caption[Calibration fits and curves for the stroke recognition model using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right).
-We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
+    \caption[Calibration fits and curves for the stroke recognition model using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right) for the test set.
+We use the ensemble that achieved the median F1-score reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
     \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
@@
 
-While correct calibration provides some interpretability to the predictions, the actual values of the probabilities are still tied of the model's performance.
-For example, with the logistic calibration, we can note that the predicted probabilities have been shifted so far towards smaller values that the highest possible predicted probability is about 50\%.
-This means strokes can be predicted at best with a 50\% chance of being correct.
-If this model were to be deployed in a randomized controlled trial, the calibrated probabilities might be incorporated in the design of the user interface such that a user can discern between confident and uncertain model predictions.
-However, it is easy to imagine how this generally low confidence might lead the user to ignore the model.
-This highlights a problem with uncertainties
+Clinicians in intensive care units and emergency departments have been found to strongly agree that a singular focus on overall accuracy cannot ensure sustained use of and trust in a model \cite{tonekaboni_what_2019}.
+Clinicians expect an alert to present a prediction that aligns with patient status.
+Despite expert-agreed thresholds for when alerts should be triggered, however, many alerts may not be aligned since the unbalanced and ambiguous nature of many predictive problems in healthcare can lead to low predictive precision \parencite{umscheid_development_2015, cite14, cite15, wenstrup_retrospective_2023}.
+This can lead to alarm fatigue \parencite{embi_evaluating_2012} and can undermine the use and endorsement by clinicians of such systems \parencite{guidi_clinician_2015}.
+The expected precision of the stroke recognition model of $24.9\%$ (95\% confidence interval $24.3-25.5\%)$ means it might be wrong three out of four times it predicts a positive.
+This unfortunately makes alarm fatigue likely among its potential users which might diminish it's real-world impact.
 
-% Brier scores:  {'uncalibrated': 0.0034955788687128925, 'logistic': 0.0018072483306197486, 'isotonic': 0.0017744177396100578, 'average': 0.0021955477869598783}
-% BSS uncalibrated reference:  {'logistic': 0.48299025755205005, 'isotonic': 0.4923822902432705}
-% BSS mean reference:  {'logistic': 0.17685766561145932, 'isotonic': 0.1918109229282361}
+If the stroke recognition model were to be deployed in a randomized controlled trial, such issues might be alleviated by incorporating the calibrated probabilities in the alerts presented to users.
+With calibrated probabilities, user would be able to discern between certain and uncertain model predictions or might be presented only with predictions that exceed a specific confidence level.
+Similar approaches have been suggested by clinicians and interviews indicate that predictive uncertainty is perceived as a sort of explanation that complements the prediction \cite{tonekaboni_what_2019}.
+
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=0.5\textwidth]{paper_retrospective_calibration_plots/agreement_fraction.pdf}
+%     \caption[Fractional agreement between stroke recognition model and call-takers as function of predicted probability]{ The fractional agreement between the stroke recognition model and the call-takers as a function of the probability predicted by the model.
+Both the uncalibrated and the isotonically calibrated model agree more with call-takers towards low and high probabilities.
+Logistic calibration yielded almost identical results to isotonic.}
+%     \label{fig_discussion:retrospective-paper-fractional-agreement-vs-probability}
+% \end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{paper_retrospective_calibration_plots/f1_score_ratio.pdf}
+    \caption[]{}
+    \label{fig_discussion:retrospective-paper-f1-score-ratio}
+\end{figure}
+
+Nevertheless, while correct calibration provides some explanation to the predictions, the actual value of the calibrated probabilities is still tied of the model's performance.
+By inspecting \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}, we can see that the logistic calibration shifts the predicted probabilities so far towards zero that the highest possible predicted probability is about 50\%.
+This means strokes can be predicted at best with a 50\% chance of being correct which limits the potential reduction in alarm fatigue.
+
+\todo[inline]{Finish this sentence and analysis.}
+Making matters worse, the most certain model predictions are the ones that agree the most with call-taker judgment (see \cref{fig_discussion:retrospective-paper-fractional-agreement-vs-probability}).
+Hence, keeping the most certain predictions also means reducing the potential for changing call-taker..
+Not really if when the model disagrees at low prob it is always right while when it disagrees at high prob it is always wrong (although unlikely and contrary to calibration - but that was done for the total set of predictions, not subgroups)
+
+This highlights a problem where combining predictions with well-calibrated uncertainties may not be sufficient to ensure sustained use in practice.
 

### 2023-10-10 chapters/main-contributions.tex
Updates to discussion

@@
 \contribitem{chp:paper-hierarchical}
-In this work we hypothesize that the likelihood estimate of variational autoencoders is a poor score for out-of-distribution due to on overemphasis on low-level features that generalize between distributions.
+In this work we hypothesize that the likelihood estimate of variational autoencoders is a poor score for out-of-distribution due to an overemphasis on low-level features that generalize between distributions.
 We further hypothesize that a well-formed hierarchy of latent variables provides a tool that can be used to select between which features to emphasize for out-of-distribution detection and hence a way to improve the performance of variational autoencoders on this task.
@@
 We reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is indeed attributable to weak configuration of the training and crudely sampled train-test splits with many extremely rare classes and several without examples in the training data at all.
-We also find a widespread error in the calculation of the macro F1 score. 
+We also find a widespread error in the calculation of the macro F1-score. 
 To compare models, we propose new data splits created with stratified sampling, use identical experimental setups and tune hyperparameters and decision boundaries. 

### 2023-10-10 chapters/paper_automated.tex
Updates to discussion

@@
 We show that several models underperform due to weak configurations, poorly sampled train-test splits, and insufficient evaluation.
-In previous work, the macro F1 score has been calculated sub-optimally, and our correction doubles it.
+In previous work, the macro F1-score has been calculated sub-optimally, and our correction doubles it.
 We contribute a revised model comparison using stratified sampling and identical experimental setups, including hyperparameters and decision boundary tuning.
@@
 \begin{enumerate}
-    \item We reproduce the performance of state-of-the-art models on MIMIC-III. We find that evaluation methods are flawed and propose corrections that double the macro F1 scores.
+    \item We reproduce the performance of state-of-the-art models on MIMIC-III. We find that evaluation methods are flawed and propose corrections that double the macro F1-scores.
     \item We find the original split of MIMIC-III to introduce strong biases in results due to missing classes in the test set.
@@
 \paragraph{Encoders:}
-The encoder usually consists of pre-trained non-contextualized word embeddings (e.g., Word2Vec) and a neural network for encoding context.
-More recently, pre-trained masked language models (e.g., BERT) have gained popularity~\parencite{tengReviewDeepNeural2022}.
+The encoder usually consists of pre"=trained non-contextualized word embeddings (e.g., Word2Vec) and a neural network for encoding context.
+More recently, pre"=trained masked language models (e.g., BERT) have gained popularity~\parencite{tengReviewDeepNeural2022}.
 The MIMIC-III training set or PubMed articles are commonly used for pre"=training.
@@
 This complicates the interpretation of results since these codes only contribute true negatives or false positives.
-Models are evaluated using the micro and macro average of the area under the curve of the receiver operating characteristics (AUC-ROC), F1 score, and Precision@k.
+Models are evaluated using the micro and macro average of the area under the curve of the receiver operating characteristics (AUC-ROC), F1-score, and precision@k.
 
@@
  hyperparameter search ranges and sampling methods  vary between works, and learning rate schedulers are only used in LAAT and PLM-ICD\parencite{mullenbachExplainablePredictionMedical2018, liICDCodingClinical2020}.
-In LAAT, the learning rate was decreased by 90\% when the F1 score had not increased for five epochs.
+In LAAT, the learning rate was decreased by 90\% when the F1-score had not increased for five epochs.
 PLM-ICD used a schedule with linear warmup followed by linear decay.
 
-All models except for PLM-ICD use Word2Vec embeddings pre-trained on the MIMIC-III training set.
-PLM-ICD uses a BERT encoder pre-trained on PubMed to encode the text in chunks of 128 tokens, and these contextualized embeddings are fed to a \lalaat~ layer.
+All models except for PLM-ICD use Word2Vec embeddings pre"=trained on the MIMIC-III training set.
+PLM-ICD uses a BERT encoder pre"=trained on PubMed to encode the text in chunks of 128 tokens, and these contextualized embeddings are fed to a \lalaat~ layer.
 
@@
 
-In this study, we included both models trained from scratch and models with components pre-trained on external corpora. 
+In this study, we included both models trained from scratch and models with components pre"=trained on external corpora. 
 We excluded models that use multi-modal inputs, such as medical code descriptions \parencite{kimReadAttendCode2021, mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, caoHyperCoreHyperbolicCograph2020, baoMedicalCodePrediction2021}, code synonyms \parencite{yuanCodeSynonymsMatter2022}, code hierarchies \parencite{caoHyperCoreHyperbolicCograph2020, xieEHRCodingMultiscale2019}, or associated Wikipedia articles \parencite{baiImprovingMedicalCode2019}, because they introduced additional complexity without providing evidence for significant performance improvements \parencite{mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, tengReviewDeepNeural2022}.
@@
 \label{sec:metrics}
-Similar to previous work, we evaluated models using AUC-ROC, F1 score, and precision@$k$.
+Similar to previous work, we evaluated models using AUC-ROC, F1-score, and precision@$k$.
 Additionally, we introduced exact match ratio (EMR), R-precision, and mean average precision (MAP).
@@
 
-Previous works calculated the macro F1 score as the harmonic mean of the macro precision and macro recall \parencite{mullenbachExplainablePredictionMedical2018, liICDCodingClinical2020, vuLabelAttentionModel2020, huangPLMICDAutomaticICD2022}.
+Previous works calculated the macro F1-score as the harmonic mean of the macro precision and macro recall \parencite{mullenbachExplainablePredictionMedical2018, liICDCodingClinical2020, vuLabelAttentionModel2020, huangPLMICDAutomaticICD2022}.
 \textcite{opitzMacroF1Macro2021} analyze macro F1 formulas common in multi-class and multi-label classification.
 They demonstrate that the above formulation is sub-optimal, as it rewards heavily biased classifiers in unbalanced datasets.
-Therefore, as recommended by the authors, we calculated the macro F1 score as the arithmetic mean of the F1 score for each class.
+Therefore, as recommended by the authors, we calculated the macro F1-score as the arithmetic mean of the F1-score for each class.
 As seen in \cref{tab:subsets}, 54\% of codes in MIMIC-III \textit{full} are missing in the test set.
-Previous works set the F1 score of all the missing codes in the test set to 0, resulting in a misleadingly low macro F1 score.
-Because 54\% of the codes are missing, the maximum possible macro F1 score is 46\%.
+Previous works set the F1-score of all the missing codes in the test set to 0, resulting in a misleadingly low macro F1-score.
+Because 54\% of the codes are missing, the maximum possible macro F1-score is 46\%.
 We ignored all codes not in the test set for our reproduction, essentially trading bias for variance. 
@@
 We ran these experiments on MIMIC-III \textit{full}, and \textit{50} as in the original works \parencite{mullenbachExplainablePredictionMedical2018,liICDCodingClinical2020,vuLabelAttentionModel2020,huangPLMICDAutomaticICD2022}.
-We used the hyperparameters reported in each paper (see \cref{tab:hyperparameters}) and report both the original and the revised macro F1 scores discussed in \cref{sec:metrics}.
+We used the hyperparameters reported in each paper (see \cref{tab:hyperparameters}) and report both the original and the revised macro F1-scores discussed in \cref{sec:metrics}.
 
@@
 Whereas original works use Adam or AdamW, we used AdamW for all experiments as it corrects the weight decay implementation of Adam \parencite{kingmaAdamMethodStochastic2017, loshchilovDecoupledWeightDecay2022}.
-For each model, we tuned the decision boundary to maximize the micro F1 score on the validation set.
+For each model, we tuned the decision boundary to maximize the micro F1-score on the validation set.
 We used randomized sampling to find optimal settings for dropout, weight decay, learning rate, and batch size.
@@
 We analyzed whether model performance correlates with document length on MIMIC-IV \textit{ICD-9}.
-Specifically, we calculated the Pearson and Spearman correlation between the number of words in the documents and the micro F1 score for all models.
+Specifically, we calculated the Pearson and Spearman correlation between the number of words in the documents and the micro F1-score for all models.
 For each model, we used the best seed from the revised comparison.
@@
 In \cref{tab:reproduced_results}, we report the reproduced results on MIMIC-III \textit{full} and \textit{50} using hyperparameters as reported in the original papers.
-We list the original and corrected macro F1 score described in \cref{sec:metrics}.
-In most cases, our corrections doubled the macro F1 scores on MIMIC-III \textit{full}.
+We list the original and corrected macro F1-score described in \cref{sec:metrics}.
+In most cases, our corrections doubled the macro F1-scores on MIMIC-III \textit{full}.
 The differences were smaller on MIMIC-III \textit{50} because all included codes are in the test set.
@@
     \caption[Results on the MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9} and MIMIC-IV  \textit{ICD-10} test sets presented as percentages.]{Results on the MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9} and MIMIC-IV  \textit{ICD-10} test sets presented as percentages.
-Micro F1 scores rank the table in ascending order.
+Micro F1-scores rank the table in ascending order.
 Each model was trained ten times with different seeds.
@@
 Not tuning the decision boundary had the largest negative effect on all models except MultiResCNN.
-In \cref{fig:threshold_tuning}, we plot the relationship between the decision boundary and F1 scores.
+In \cref{fig:threshold_tuning}, we plot the relationship between the decision boundary and F1-scores.
 LAAT and MultiResCNN perform similarly when using a decision boundary of 0.5.
@@
     \centering
-    \caption[Ablation study on MIMIC-III \textit{clean}.]{Ablation study on MIMIC-III \textit{clean}. The numbers are the micro/macro F1 scores on the test set.}
+    \caption[Ablation study on MIMIC-III \textit{clean}.]{Ablation study on MIMIC-III \textit{clean}. The numbers are the micro/macro F1-scores on the test set.}
     \label{tab:ablation_study}
@@
     \end{subfigure}%
-    \caption[The relationship between chosen threshold and F1 score of every reproduced model in \cref{tab:reproduced_results}]{The relationship between chosen threshold and F1 score of every reproduced model in \cref{tab:reproduced_results}.
-The left figure shows the micro F1 score, and the right shows the macro F1 score.
+    \caption[The relationship between chosen threshold and F1-score of every reproduced model in \cref{tab:reproduced_results}]{The relationship between chosen threshold and F1-score of every reproduced model in \cref{tab:reproduced_results}.
+The left figure shows the micro F1-score, and the right shows the macro F1-score.
 The models were evaluated on MIMIC-III \textit{clean}.}
@@
 
-\cref{fig:train_size} shows the relationship between the number of training examples and the micro and macro F1 scores for all models.
-In most cases, increasing the training data had a larger effect on the macro F1 score than the micro F1 score, indicating more extensive improvements in rare codes than common codes.
-The curve for macro F1 is less smooth because the decision boundary was tuned on the micro F1 scores.
+\cref{fig:train_size} shows the relationship between the number of training examples and the micro and macro F1-scores for all models.
+In most cases, increasing the training data had a larger effect on the macro F1-score than the micro F1-score, indicating more extensive improvements in rare codes than common codes.
+The curve for macro F1 is less smooth because the decision boundary was tuned on the micro F1-scores.
 
@@
     \end{subfigure}
-    \caption[The relationship between the number of training examples and F1 score on MIMIC-IV \textit{ICD-9}]{The relationship between the number of training examples and F1 score on MIMIC-IV \textit{ICD-9}.
+    \caption[The relationship between the number of training examples and F1-score on MIMIC-IV \textit{ICD-9}]{The relationship between the number of training examples and F1-score on MIMIC-IV \textit{ICD-9}.
 The left figure shows the F1 Micro score on the y-axis, while the right figure shows the F1 Macro score.}
@@
 
-We plot the micro F1 score for all models as a function of the number of words per document in \cref{fig:text_length}. 
+We plot the micro F1-score for all models as a function of the number of words per document in \cref{fig:text_length}. 
 We note that all models underperformed on documents with fewer than 1000 words. 
@@
     \includegraphics[width=0.9\textwidth]{paper_automated/f1_micro_vs_num_words.png}
-    \caption[Relationship between the lengths of the clinical notes and the micro F1 score for each model on MIMIC-IV \textit{ICD-9}]{Relationship between the lengths of the clinical notes and the micro F1 score for each model on MIMIC-IV \textit{ICD-9}.
+    \caption[Relationship between the lengths of the clinical notes and the micro F1-score for each model on MIMIC-IV \textit{ICD-9}]{Relationship between the lengths of the clinical notes and the micro F1-score for each model on MIMIC-IV \textit{ICD-9}.
 The vertical line indicates the maximum length of the notes after truncation.
@@
     \centering
-    \caption[Correlation between the F1 score and the logarithm of code frequency and document length on MIMIC-IV \textit{ICD-9}.]{Correlation between the F1 score and the logarithm of code frequency and document length on MIMIC-IV \textit{ICD-9}.
+    \caption[Correlation between the F1-score and the logarithm of code frequency and document length on MIMIC-IV \textit{ICD-9}.]{Correlation between the F1-score and the logarithm of code frequency and document length on MIMIC-IV \textit{ICD-9}.
 As discussed in \cref{subsubsec:document-length}, we only considered document lengths between 1000 and 4000 words.
@@
 \Cref{fig:icd9_vs_icd10} compares the best performing model, PLM-ICD, trained and evaluated on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}. Similar results were obtained on MIMIC-III \textit{clean}.
-The comparison shows the relationship between code frequencies in the training set and macro F1 scores.
+The comparison shows the relationship between code frequencies in the training set and macro F1-scores.
 As shown in \cref{tab:fair_results}, all models perform worse on \textit{ICD-10} compared to \textit{ICD-9}.
@@
 
-The Pearson and Spearman correlations between the logarithm of code frequency and F1 score are shown in \cref{tab:correlations} for MIMIC-IV \textit{ICD-9}.
+The Pearson and Spearman correlations between the logarithm of code frequency and F1-score are shown in \cref{tab:correlations} for MIMIC-IV \textit{ICD-9}.
 Similar correlations were observed for the other datasets.
@@
     \includegraphics[width=0.9\textwidth]{paper_automated/code_count_vs_f1_all_datasets.png}
-    \caption[Relationship between the code frequencies in the training set and the macro F1 score for PLM-ICD on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}]{Relationship between the code frequencies in the training set and the macro F1 score for PLM-ICD on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}.
+    \caption[Relationship between the code frequencies in the training set and the macro F1-score for PLM-ICD on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}]{Relationship between the code frequencies in the training set and the macro F1-score for PLM-ICD on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}.
 The shaded area indicates the standard deviation of the score computed for codes within the bin.}
@@
 Among these, Z72.0 was the third most common single code in the training set that PLM-ICD never predicted correctly.
-PLM-ICD achieved an F1 score of 53\% for Z87.891, 51\% for F17.210, and 0\% for Z72.0 and all other nicotine-related codes.
+PLM-ICD achieved an F1-score of 53\% for Z87.891, 51\% for F17.210, and 0\% for Z72.0 and all other nicotine-related codes.
 These findings suggest that when there is a class imbalance among highly similar codes, PLM-ICD is strongly biased toward the most frequent ones.
@@
 We showed that models previously reported as low-performing underperformed partly due to a poor selection of hyperparameters and not tuning the decision boundary.
-In our revised comparison, we demonstrated that training the models using our setup decreased the difference between the best and worst micro F1 scores by $5.8$ percentage points.
+In our revised comparison, we demonstrated that training the models using our setup decreased the difference between the best and worst micro F1-scores by $5.8$ percentage points.
 \textcite{mullenbachExplainablePredictionMedical2018} concluded that CNN outperformed Bi-GRU.
@@
 
-We demonstrated that the macro F1 score had been underestimated in prior works due to the poorly sampled MIMIC-III \textit{full} split and the practice of setting the F1 score of all codes absent in the test set to 0.
-Since 54\% of the codes in MIMIC-III \textit{full} are missing in the test set, the maximum possible macro F1 score is 46\%.
-The previously highest reported macro F1 score on MIMIC-III \textit{full} is 12.7\% for PLM-ICD~\parencite{kimReadAttendCode2021}.
-Using our corrected macro F1 score on the same split, PLM-ICD achieved a macro F1 score of 22.8\%.
-This large difference from previous state-of-the-art seems to indicate that all previous work on AMC used the sub-optimally calculated macro F1 score, including works not reproduced in this paper.
-Many studies use the macro F1 score to evaluate the ability of their models to predict rare codes \parencite{kimReadAttendCode2021, yuanCodeSynonymsMatter2022}.
+We demonstrated that the macro F1-score had been underestimated in prior works due to the poorly sampled MIMIC-III \textit{full} split and the practice of setting the F1-score of all codes absent in the test set to 0.
+Since 54\% of the codes in MIMIC-III \textit{full} are missing in the test set, the maximum possible macro F1-score is 46\%.
+The previously highest reported macro F1-score on MIMIC-III \textit{full} is 12.7\% for PLM-ICD~\parencite{kimReadAttendCode2021}.
+Using our corrected macro F1-score on the same split, PLM-ICD achieved a macro F1-score of 22.8\%.
+This large difference from previous state-of-the-art seems to indicate that all previous work on AMC used the sub-optimally calculated macro F1-score, including works not reproduced in this paper.
+Many studies use the macro F1-score to evaluate the ability of their models to predict rare codes \parencite{kimReadAttendCode2021, yuanCodeSynonymsMatter2022}.
 If it has indeed been incorrectly calculated in these studies, some conclusions drawn in previous work regarding rare code prediction may have been misguided.
@@
 Finally, while PLM-ICD outperforms the other models in this paper, the improvements are limited compared to the effect of pre"=training in other domains \parencite{mohamed_selfsupervised_2022, linPretrainedTransformersText2021,baevski_wav2vec_2020,devlin_bert_2018,dosovitskiy_image_2021}.
-Notably, there have been several unsuccessful attempts at using pre-trained transformers for medical coding \parencite{jiDoesMagicBERT2021,gaoLimitationsTransformersClinical2021,michalopoulosICDBigBirdContextualEmbedding2022,pascualBERTbasedAutomaticICD2021,zhangBERTXMLLargeScale2020}.
-In future work, we want to investigate why pre-trained transformers underperform in medical coding.
+Notably, there have been several unsuccessful attempts at using pre"=trained transformers for medical coding \parencite{jiDoesMagicBERT2021,gaoLimitationsTransformersClinical2021,michalopoulosICDBigBirdContextualEmbedding2022,pascualBERTbasedAutomaticICD2021,zhangBERTXMLLargeScale2020}.
+In future work, we want to investigate why pre"=trained transformers underperform in medical coding.
 
@@
 In this paper, we first reproduced the results of selected state-of-the-art models focusing on unimodal models with publically available source code. 
-We found that model evaluation in original works was biased by an inappropriate formulation of the macro F1 score and treatment of missing classes in the test set.
+We found that model evaluation in original works was biased by an inappropriate formulation of the macro F1-score and treatment of missing classes in the test set.
 By fixing the macro F1 computation, we approximately doubled the macro F1 of the reproduced models on MIMIC-III \textit{full}.

### 2023-10-10 chapters/paper_benchmarking.tex
Adjust table width
Updates to discussion

@@
     \centering
-    \begin{adjustbox}{center, totalheight=0.4\paperheight}
+    \begin{adjustbox}{center, totalheight=0.37\paperheight}
     \begin{tabular}[t]{lll|r}
@@
     \centering
-    \begin{adjustbox}{center, totalheight=0.4\paperheight}
+    \begin{adjustbox}{center, totalheight=0.37\paperheight}
     \begin{tabular}[t]{cll|c}
@@
 It is trained with the connectionist temporal classification (CTC) loss \parencite{graves_connectionist_2006} which lets it jointly learn to align and classify without using label alignments.
-We pre-train all unsupervised models at $s=64$ on the full TIMIT training dataset excluding the validation data (3.7h) as in \cref{tab: likelihoods timit} (left). 
+We pre"=train all unsupervised models at $s=64$ on the full TIMIT training dataset excluding the validation data (3.7h) as in \cref{tab: likelihoods timit} (left). 
 We then train the ASR model on all 3.7h as well as 1h and 10m subsets. We report results on the test set in terms of phoneme error rates (PER) in \cref{tab: phoneme recognition (PER)} (right).

### 2023-10-10 chapters/paper_modelagnostic.tex
Updates to discussion

@@
 They only focused on the single-sample detection task, i.e.\@ $n=1$ following our problem formulation.
-Their idea is to fit different non-parametric density estimators, such as a kernel-density estimator (KDE) or a one-class support vector machines (SVM), for each different statistic $T_1,\dots, T_k$ by using the values computed on the training set examples.
+Their idea is to fit different nonparametric density estimators, such as a kernel-density estimator (KDE) or a one-class support vector machines (SVM), for each different statistic $T_1,\dots, T_k$ by using the values computed on the training set examples.
 For a single test example, $\tilde{x}_1$, they first compute $T_1,\dots, T_k$ and then combine those statistics by summing the different KDEs log-density.

### 2023-10-10 chapters/paper_retrospective.tex
Updates to discussion

@@
 However, this did not improve performance.
-In addition, we tested a pre-trained self"=supervised model.
+In addition, we tested a pre"=trained self"=supervised model.
 Although many of these models are freely available to the public, they are primarily trained on English data.
@@
 We hypothesise that the number of ground truth stroke positives was too small for these advanced models to learn more complex patterns than the MLP ensemble.
-In addition, a self"=supervised model would likely benefit from being pre-trained on speech or text data from the target domain.
+In addition, a self"=supervised model would likely benefit from being pre"=trained on speech or text data from the target domain.
 Although training such large-scale foundation models have the potential to improve the classification model further, it is beyond the scope of this study.
@@
 We have included references to reviews of self"=supervised learning for speech and text in the references \parencite{gururangan_don_2020,mohamed_selfsupervised_2022}.
-Notably, it is not uncommon for small, simple models to match or outperform large, pre-trained models for text-classification tasks \parencite{galke_bagofwords_2022}.
+Notably, it is not uncommon for small, simple models to match or outperform large, pre"=trained models for text-classification tasks \parencite{galke_bagofwords_2022}.
 

### 2023-10-10 chapters/paper_review/01_introduction.tex
Updates to discussion

@@
 There are two stages in this framework.
-In the first stage, we use SSL to pre-train a \textit{representation model},
+In the first stage, we use SSL to pre"=train a \textit{representation model},
 also called an \textit{upstream model} or a \textit{foundation model}.
 In the second stage, downstream tasks use either the learned
-representation from the frozen model, or fine-tune the entire pre-trained model
+representation from the frozen model, or fine-tune the entire pre"=trained model
 in a supervised phase~\parencite{hinton_reducing_2006}. 

### 2023-10-10 chapters/paper_review/03_approaches.tex
Updates to discussion

@@
 
-\edit{After pre"=training,} the representation model $f(\cdot)$ can be fine-tuned for a downstream task directly or used to extract features which are fed to another model, as visualized in \cref{fig:SSL_framework}.
+\edit{After pre"=training,} the representation model $f(\cdot)$ can be fine"=tuned for a downstream task directly or used to extract features which are fed to another model, as visualized in \cref{fig:SSL_framework}.
 It is not uncommon to use the output representation~$H$, but often representations from hidden layers of $f(\cdot)$ are better suited \parencite{pasad_layerwise_2021}.
@@
 In non"autoregressive predictive coding (NPC)~\parencite{liu_nonautoregressive_2020}, time masking is introduced through masked convolution blocks.
-Taking inspiration from XLNet~\parencite{yang_xlnet_2019}, it has also been suggested that the input be reconstructed from a shuffled version~\parencite{song_speechxlnet_2020} to address the discrepancy between pre"=training and fine-tuning of masking-based approaches.
+Taking inspiration from XLNet~\parencite{yang_xlnet_2019}, it has also been suggested that the input be reconstructed from a shuffled version~\parencite{song_speechxlnet_2020} to address the discrepancy between pre"=training and fine"=tuning of masking-based approaches.
 
@@
 
-The wav2vec 2.0 approach was the first to reach single-digit \edit{word error rate (WER) on LibriSpeech using only the low-resource Libri-light subsets for fine-tuning a pre-trained model (see \cref{sec:datasets}).
+The wav2vec 2.0 approach was the first to reach single-digit \edit{word error rate (WER) on LibriSpeech using only the low-resource Libri-light subsets for fine"=tuning a pre"=trained model (see \cref{sec:datasets}).
 It has subsequently} inspired many follow-up studies.
@@
 \paragraph{Discrete BERT} \edit{Applying BERT-type training directly to speech input is not possible due to its continuous nature}.
-The Discrete BERT approach~\parencite{baevski_effectiveness_2020} uses a pre-trained vq-wav2vec model to derive a discrete vocabulary \parencite{baevski_vqwav2vec_2020}.
+The Discrete BERT approach~\parencite{baevski_effectiveness_2020} uses a pre"=trained vq-wav2vec model to derive a discrete vocabulary \parencite{baevski_vqwav2vec_2020}.
 The vq-wav2vec model is similar to wav2vec mentioned in \cref{par:cpc} but uses quantization to learn discrete representations.
@@
 \noindent where $\mathcal{M}$ is the set of all masked timesteps. During training, only the BERT model's parameters are updated, while the vq-wav2vec model parameters are frozen.} 
-Discrete BERT was the first model to demonstrate the effectiveness of self"=supervised speech representation learning by achieving a WER of 25\% on the standard test-other subset using a 10-minute fine-tuning set, \edit{setting the direction for many approaches to follow}.
+Discrete BERT was the first model to demonstrate the effectiveness of self"=supervised speech representation learning by achieving a WER of 25\% on the standard test-other subset using a 10-minute fine"=tuning set, \edit{setting the direction for many approaches to follow}.
 
@@
 There has also been growing interest in learning joint speech and text representations using paired and unpaired data.
-The SLAM approach~\parencite{bapna_slam_2021} is an example where speech and text are first represented using two separate pre-trained encoders followed by a multi-modal encoder to build the joint representations.
+The SLAM approach~\parencite{bapna_slam_2021} is an example where speech and text are first represented using two separate pre"=trained encoders followed by a multi-modal encoder to build the joint representations.
 The entire model is trained using a multi-task loss including two supervised and two self"=supervised tasks.

### 2023-10-10 chapters/paper_review/04_benchmarking.tex
Updates to discussion

@@
   evaluations (Part 1). 
-  Networks are usually pre-trained with SSL
-  techniques, augmented with prediction heads, and fine-tuned (or trained) with
+  Networks are usually pre"=trained with SSL
+  techniques, augmented with prediction heads, and fine"=tuned (or trained) with
   labeled data in downstream tasks for benchmarking. 
@@
     \toprule
-    \multirow{2}{*}{Work} & \multirow{2}{*}{\makecell[l]{pre"=training\\ corpus}} & \multirow{2}{*}{Task} & \multicolumn{2}{c}{Dataset} & \multirow{2}{*}{Transfer} & \multirow{2}{*}{Fine-tuning labels used} \\ \cline{4-5}
-     &  &  & Fine-tuning & Test & & \\
+    \multirow{2}{*}{Work} & \multirow{2}{*}{\makecell[l]{pre"=training\\ corpus}} & \multirow{2}{*}{Task} & \multicolumn{2}{c}{Dataset} & \multirow{2}{*}{Transfer} & \multirow{2}{*}{Fine"=tuning labels used} \\ \cline{4-5}
+     &  &  & Fine"=tuning & Test & & \\
     \midrule
@@
 \footnotetext{\label{footnote:table:method_data_setting}
-  The \textbf{pre"=training corpus}, \textbf{Fine-tuning}, and
+  The \textbf{pre"=training corpus}, \textbf{Fine"=tuning}, and
   \textbf{Test} columns list the datasets used in each work, and the \textbf{Task}
@@
   transfer learning, i.e., different datasets are utilized for pre"=training and
-  fine-tuning. The \textbf{Fine-tuning labels used} column summarizes the
-  amount of labeled examples used in downstream fine-tuning.
+  fine"=tuning. The \textbf{Fine"=tuning labels used} column summarizes the
+  amount of labeled examples used in downstream fine"=tuning.
 }
@@
 A common way to benchmark SSL techniques and show their efficacy is to
-fine-tune a pre-trained SSL model for a supervised downstream task.
-Depending on the corpora used in pre"=training and fine-tuning, techniques can
+fine-tune a pre"=trained SSL model for a supervised downstream task.
+Depending on the corpora used in pre"=training and fine"=tuning, techniques can
 be benchmarked in terms of their capability to transfer knowledge across datasets
-(i.e., using pre"=training corpora that differ from the fine-tuning ones), their
+(i.e., using pre"=training corpora that differ from the fine"=tuning ones), their
 benefit when training with limited labeled examples (i.e., sampling a subset of
-labeled examples for fine-tuning), or their improvement over a fully supervised
+labeled examples for fine"=tuning), or their improvement over a fully supervised
 baseline (i.e., using the entire training split of downstream datasets for
-fine-tuning). \cref{table:method_data_setting,table:method_data_setting_2} summarize experiment settings used in the
+fine"=tuning). \cref{table:method_data_setting,table:method_data_setting_2} summarize experiment settings used in the
 SSL literature, including the pre"=training corpora, downstream tasks and datasets,
-and the amount of fine-tuning labels used, which indicates the targeted
+and the amount of fine"=tuning labels used, which indicates the targeted
 benchmarking scenario as discussed above. Note that there are a variety of ways
-to fine-tune pre-trained networks (e.g., fine-tune the entire network,
-freeze certain layers during fine-tuning, and add various architectures of
-prediction layers to pre-trained networks). We here omit
+to fine-tune pre"=trained networks (e.g., fine-tune the entire network,
+freeze certain layers during fine"=tuning, and add various architectures of
+prediction layers to pre"=trained networks). We here omit
 descriptions of these choices; readers can consult the
@@
 multilinguality~\parencite{kawakami_learning_2020, babu_xlsr_2021} are also gaining
-attention. Compared to pre"=training, datasets used for fine-tuning are more
+attention. Compared to pre"=training, datasets used for fine"=tuning are more
 diverse and cover downstream tasks as varied as ASR, PR, PC, SID, AED,
@@
 supervision as well as limited resources, the amount of labeled examples used for
-fine-tuning also varies from several minutes up to 1k hours. Recent
+fine"=tuning also varies from several minutes up to 1k hours. Recent
 benchmarks such as SUPERB~\parencite{yang_superb_2021} that consolidate multiple 
@@
 popularity of SSL research, we expect future experiment settings to proliferate
-and cover more languages, downstream tasks, and pre"=training/fine-tuning
+and cover more languages, downstream tasks, and pre"=training/fine"=tuning
 datasets.
@@
 model parameters along the horizontal axis. Markers in blue correspond to models initialized with various SSL
-techniques and then fine-tuned using 960, 100, 10, 1 hour(s), and 10 minutes
+techniques and then fine"=tuned using 960, 100, 10, 1 hour(s), and 10 minutes
 respectively. The 960-hour training set is the aggregation of
@@
 and objectives, whether a language model is used in decoding, whether model
-parameters are frozen in fine-tuning), readers should be careful that the superiority
+parameters are frozen in fine"=tuning), readers should be careful that the superiority
 of methods cannot be decided only based on lower WER numbers.}
@@
 \cref{fig:asr_result_no_limit}. The ASR models were obtained first by using
-unlabeled speech to pre-train a model with each SSL technique. The model was
-then fine-tuned on labeled data by utilizing a supervised training objective.
+unlabeled speech to pre"=train a model with each SSL technique. The model was
+then fine"=tuned on labeled data by utilizing a supervised training objective.
 Respectively, 960, 100, 10, 1 hour(s), and 10 minutes of labeled LS training data
-were used for fine-tuning, as indicated in different panels of
+were used for fine"=tuning, as indicated in different panels of
 \cref{fig:asr_result_no_limit} (see
@@
 when the entire 960-hour of labeled data is used in
-training/fine-tuning. The benefit of SSL, especially models with more parameters
+training/fine"=tuning. The benefit of SSL, especially models with more parameters
 like wav2vec 2.0 and HuBERT, becomes more evident when the labeling resources
@@
 information such as content, speaker, and semantics. As SSL research
-gains more attention, we expect that SSL pre-trained models will 
+gains more attention, we expect that SSL pre"=trained models will 
 achieve state-of-the-art results on an increasing number of tasks.
@@
 The task performance is obtained from the original papers and is often
-achieved with different downstream fine-tuning recipes, including various
+achieved with different downstream fine"=tuning recipes, including various
 language models (used in the ASR system), prediction heads (networks added to
-SSL for downstream inference), or choices between fine-tuning the whole
+SSL for downstream inference), or choices between fine"=tuning the whole
 networks or freezing the SSL encoders. For example, in the ASR task, HuBERT-L

### 2023-10-10 chapters/paper_review/04b_analysis.tex
Updates to discussion

@@
 amount of lexical information present in the embeddings.  Using this
-battery of tests the authors compared pre-trained models of varying
-sizes as well as models fine-tuned for ASR. They found that pre-trained
+battery of tests the authors compared pre"=trained models of varying
+sizes as well as models fine"=tuned for ASR. They found that pre"=trained
 models show an autoencoder-style behavior, with early layers showing
@@
 for layers 10--21 and higher error rates for the other layers. 
-\parencite{pasad_layerwise_2021} further show that fine-tuning the pre-trained model with a
+\parencite{pasad_layerwise_2021} further show that fine"=tuning the pre"=trained model with a
 character-level CTC training criterion changes the behavior of the
@@
 provide more class information than the representations from final layers. This
-is similar to the findings described above for wav2vec~2.0 without fine-tuning,
+is similar to the findings described above for wav2vec~2.0 without fine"=tuning,
 where intermediate layers provide more information about phone and word
@@
 speech features.
-Do pre-trained representations offer greater robustness against distributional
+Do pre"=trained representations offer greater robustness against distributional
 shifts? 
-One study~\parencite{kawakami_learning_2020} compared pre-trained
+One study~\parencite{kawakami_learning_2020} compared pre"=trained
 representations from a CPC model against MFCCs and 
-found pre-trained representations to be more robust to mismatches between
+found pre"=trained representations to be more robust to mismatches between
 training and test data. The 
@@
 conclusions are that pre"=training on more and diverse domains is preferable:
-models pre-trained on more domains performed better than those pre-trained on
+models pre"=trained on more domains performed better than those pre"=trained on
 fewer when tested on held-out domains, regardless of which additional
-labeled data was used for fine-tuning. Adding in-domain unlabeled data---if
+labeled data was used for fine"=tuning. Adding in-domain unlabeled data---if
 available---to pre"=training improves performance robustly; however, even
@@
 phone discrimination tasks was compared across several languages. It
-was found that representations pre-trained only on English successfully
+was found that representations pre"=trained only on English successfully
 enabled phone discrimination in 10 other languages, rivaling
@@
 and multilingual LS)
-jointly, followed by fine-tuning for each individual language. The largest
+jointly, followed by fine"=tuning for each individual language. The largest
 model covers 53 languages in total
@@
 character error rate than monolingual representations) for languages not
-present in the training dataset, again showing that unsupervised pre-trained
+present in the training dataset, again showing that unsupervised pre"=trained
 representations can learn generic features of the speech signal that generalize

### 2023-10-10 chapters/paper_review/05_zero.tex
Updates to discussion

@@
 speech and text were different~\parencite{lin_analyzing_2022}.
-Experimental results showed that domain mismatch leads to inferior performance, but a representation model pre-trained on the targeted speech domain extracts better representations and reduces this drop in performance.
+Experimental results showed that domain mismatch leads to inferior performance, but a representation model pre"=trained on the targeted speech domain extracts better representations and reduces this drop in performance.
 
@@
 end-to-end ASR/TTS models.
-This approach first prepares pre-trained ASR model \edit{
-$f_{\mathrm{asr}}(X)$ with acoustic input $X$ and pre-trained TTS model $g_{\mathrm{tts}}(Y)$ with text input $Y$.
+This approach first prepares pre"=trained ASR model \edit{
+$f_{\mathrm{asr}}(X)$ with acoustic input $X$ and pre"=trained TTS model $g_{\mathrm{tts}}(Y)$ with text input $Y$.
 By following the TTS system with an ASR system, we generate new 
@@
 This study was motivated by unsupervised query-by-example, applications of
-non-parametric Bayesian machine learning to speech processing, and low-resource
+nonparametric Bayesian machine learning to speech processing, and low-resource
 speech recognition, and was also inspired by the learning process of infants.
@@
 	 The spoken term discovery track used the normalized edit distance and coverage
-	 scores in addition to the precision, recall, and F1 scores for types,
+	 scores in addition to the precision, recall, and F1-scores for types,
 	 tokens, and boundaries.

### 2023-10-10 chapters/paper_review/06_conclusions.tex
Updates to discussion

@@
 We have presented an overview of existing benchmarks and reviewed the efforts towards efficient zero-resource learning.
-\edit{Although the field is progressing rapidly, with new approaches reaching higher levels of performance, a couple of patterns have emerged: (1) The solid performance of Wav2vec 2.0 for speech recognition and many downstream tasks, as well as the public availability of its pre-trained multilingual variants, enabled wide adoption in the community making it a ``standard'' go-to model.
+\edit{Although the field is progressing rapidly, with new approaches reaching higher levels of performance, a couple of patterns have emerged: (1) The solid performance of Wav2vec 2.0 for speech recognition and many downstream tasks, as well as the public availability of its pre"=trained multilingual variants, enabled wide adoption in the community making it a ``standard'' go-to model.
 (2) The simplicity and stability of the HuBERT approach, as well as the resemblance of its training procedure to classic frame-level ASR systems, made it an easy choice for research extensions on improving representation quality, speech translation, and textless NLP.}
@@
 The prompt/instruction learning methods~\parencite{liu_pretrain_2021} also freeze the SSL parameters and control the output of SSL by adding additional information, which is called \textit{prompt}, in the input.
-Both adapter-based methods and prompt/instruction learning yield competitive performance compared with fine-tuning in NLP applications, but there is only little related work for speech~\parencite{thomas_efficient_2022,chang_speechprompt_2022}.
+Both adapter-based methods and prompt/instruction learning yield competitive performance compared with fine"=tuning in NLP applications, but there is only little related work for speech~\parencite{thomas_efficient_2022,chang_speechprompt_2022}.
 In addition, prompt for speech SSL does not achieve comparable performance on sequence generation tasks like phoneme recognition and slot filling, so how to use prompt is still an open question.}
@@
 \item \textbf{Increasing the efficiency of the representation model.} As discussed in \cref{subsec:modelsize}, larger representation models lead to better downstream performance.
-Despite the success of these \edit{large} models, they \edit{incur high costs in terms of memory and time for pre"=training, fine-tuning, and even when used only to extract representations without gradient calculation.
+Despite the success of these \edit{large} models, they \edit{incur high costs in terms of memory and time for pre"=training, fine"=tuning, and even when used only to extract representations without gradient calculation.
 This} makes them unsuitable for edge devices \edit{but also limits the ability to scale these models to very large datasets} \ci{ -- and leads to a large energy consumption}.

### 2023-10-10 chapters/paper_review/main_hi.tex
Updates to discussion

@@
         wav2vec 2.0 \textsc{Large}~\parencite{baevski_wav2vec_2020} & LL-60k & Transformer & 1.6 & 3.0 & 1.8 & 3.3 \\
-        pre-trained Conformer XXL~\parencite{zhang_pushing_2020} & LL-60k & LSTM & 1.5 & 3.0 & 1.5 & 3.1 \\
+        pre"=trained Conformer XXL~\parencite{zhang_pushing_2020} & LL-60k & LSTM & 1.5 & 3.0 & 1.5 & 3.1 \\
 
@@
         wav2vec 2.0 + self-training~\parencite{xu2020self} & LL-60k & Transformer & 1.1 & 2.7 & 1.5 & 3.1\\
-        pre-trained Conformer XXL + Noisy Student~\parencite{zhang_pushing_2020} & LL-60k & LSTM & 1.3 & 2.6 & 1.4 & 2.6 \\
+        pre"=trained Conformer XXL + Noisy Student~\parencite{zhang_pushing_2020} & LL-60k & LSTM & 1.3 & 2.6 & 1.4 & 2.6 \\
 

### 2023-10-10 chapters/paper_review/pretext.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses
Improve hyphenation and add "= for compound words
Updates to discussion

@@
 
-\begin{table}[h!]
+\begin{table}
     \centering
     \caption[An overview of approaches within the three main categories of
-    self-supervised learning.]{
+    self"=supervised learning.]{
     An overview of approaches within the three categories of
-    self-supervised learning. 
+    self"=supervised learning. 
     Column~(a) lists the names of the models and related references, 
@@
 \centering
-\resizebox{0.98\textwidth}{!}{
+\resizebox{0.99\textwidth}{!}{
 \begin{tabular}{l|c|c|c}
@@
 We indicate random temporal permutation of a sequence by indexing it with
-the set $\mathcal{P}_t\triangleq\textsc{permute}([0,t])$, where
+the set $\mathcal{P}_t\equiv\textsc{permute}([0,t])$, where
 $\textsc{permute}(\cdot)$ returns a permutation of the given list. 

### 2023-10-11 chapters/discussion.tex
Updates to discussion

@@
     \end{subfigure}
-    \caption[Calibration curve for the uncalibrated stroke recognition model and empirical distribution of predicted probabilities.]{%
-        Calibration curve for the uncalibrated stroke recognition ensemble model (left) and the histogram of predicted probabilities (right) for the test set. 
+    \caption[Calibration curve for the uncalibrated stroke recognition ensemble and empirical distribution of predicted probabilities.]{%
+        Calibration curve for the uncalibrated stroke recognition ensemble (left) and the histogram of predicted probabilities (right) for the test set. 
         We use the ensemble that achieved the median F1-score reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
@@
     \end{subfigure}    
-    \caption[Calibration fits and curves for the stroke recognition model using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right) for the test set.
+    \caption[Calibration fits and curves for the stroke recognition ensemble using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right) for the test set.
 We use the ensemble that achieved the median F1-score reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
@@
 This can lead to alarm fatigue \parencite{embi_evaluating_2012} and can undermine the use and endorsement by clinicians of such systems \parencite{guidi_clinician_2015}.
-The expected precision of the stroke recognition model of $24.9\%$ (95\% confidence interval $24.3-25.5\%)$ means it might be wrong three out of four times it predicts a positive.
+The expected precision of the stroke recognition ensemble of $24.9\%$ (95\% confidence interval $24.3-25.5\%)$ means it might be wrong three out of four times it predicts a positive.
 This unfortunately makes alarm fatigue likely among its potential users which might diminish it's real-world impact.
 
-If the stroke recognition model were to be deployed in a randomized controlled trial, such issues might be alleviated by incorporating the calibrated probabilities in the alerts presented to users.
+If the stroke recognition ensemble were to be deployed in a randomized controlled trial, such issues might be alleviated by incorporating the calibrated probabilities in the alerts presented to users.
 With calibrated probabilities, user would be able to discern between certain and uncertain model predictions or might be presented only with predictions that exceed a specific confidence level.
@@
 
-% \begin{figure}
-%     \centering
-%     \includegraphics[width=0.5\textwidth]{paper_retrospective_calibration_plots/agreement_fraction.pdf}
-%     \caption[Fractional agreement between stroke recognition model and call-takers as function of predicted probability]{ The fractional agreement between the stroke recognition model and the call-takers as a function of the probability predicted by the model.
-Both the uncalibrated and the isotonically calibrated model agree more with call-takers towards low and high probabilities.
-Logistic calibration yielded almost identical results to isotonic.}
-%     \label{fig_discussion:retrospective-paper-fractional-agreement-vs-probability}
-% \end{figure}
+\begin{figure}
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/f1_score_vs_predicted_probability_ensemble_calltaker}
+        % \caption{}
+        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
+    \end{subfigure}    
+    % \hfill
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/predicted_probability_histogram.pdf}
+        % \caption{}
+        % \label{fig_discussion:histogram_ensemble_and_single_model}
+    \end{subfigure}    
+    \caption[Comparison of F1-score of stroke recognition ensemble and call-takers as function of predicted probability.]{ Comparison of the F1-score of the stroke recognition ensemble and call-takers.
+The F1-score is computed on subsets of the test dataset made by binning on the calibrated probabilities of the ensemble.
+We see that the relative performance improvement of the ensemble over call-takers is higher towards more certain predictions.
+We use the ensemble that achieved the median F1-score reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
+    \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
+\end{figure}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{paper_retrospective_calibration_plots/f1_score_ratio.pdf}
-    \caption[]{}
-    \label{fig_discussion:retrospective-paper-f1-score-ratio}
-\end{figure}
+Inspecting the
+
+\todo[inline]{Finish this sentence and analysis.}
 
@@
 \todo[inline]{Finish this sentence and analysis.}
+
 Making matters worse, the most certain model predictions are the ones that agree the most with call-taker judgment (see \cref{fig_discussion:retrospective-paper-fractional-agreement-vs-probability}).

### 2023-10-12 chapters/discussion.tex
Updates to discussion and plotting

@@
         \centering
-        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curves_ensemble.pdf}
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curves_ensemble_with_cis.pdf}
         % \caption{}
@@
         \centering
-        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/f1_score_vs_predicted_probability_ensemble_calltaker}
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/f1_score_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
         % \caption{}
@@
 
-Making matters worse, the most certain model predictions are the ones that agree the most with call-taker judgment (see \cref{fig_discussion:retrospective-paper-fractional-agreement-vs-probability}).
+% Making matters worse, the most certain model predictions are the ones that agree the most with call-taker judgment (see \cref{fig_discussion:retrospective-paper-fractional-agreement-vs-probability}).
 Hence, keeping the most certain predictions also means reducing the potential for changing call-taker..
@@
 
-This highlights a problem where combining predictions with well-calibrated uncertainties may not be sufficient to ensure sustained use in practice.
+% This highlights a problem where combining predictions with well-calibrated uncertainties may not be sufficient to ensure sustained use in practice.
 

### 2023-10-12 chapters/introduction.tex
Updates to discussion
Updates to discussion and plotting

@@
 With the digital revolution continuing into the 21st century and the worldwide spread of the internet, the amount of data available to decision-makers has grown exponentially. 
-This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that focuses on developing algorithms that can learn from data to recognize patterns and make predictions to guide decision-making.
+This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that focuses on developing computer algorithms that can learn from data to recognize patterns and make predictions that guide decision-making.
 Enabled by a massive increase in the performance and availability of parallel computing resources, such methods have set new standards for the abilities of computer systems in various domains, such as speech recognition, computer vision, and natural language processing \parencite{lecun_deep_2015}.
@@
 One of the best known methods for calibrating a binary classifier is Platt scaling which fits a logistic regression model to the model outputs, assuming that the miscalibration can be corrected by a logarithmic function \parencite{platt_probabilistic_1999}.
-Another method is isotonic regression which instead fits a non-parametric, monotonic function \parencite{zadrozny_transforming_2002}.
+Another method is isotonic regression which instead fits a nonparametric, monotonic function \parencite{zadrozny_transforming_2002}.
 More recently, \textcite{guo_calibration_2017} proposed a single-parameter variant of Platt-scaling that fits only a temperature parameter on the logits of a neural network classifier. 

### 2023-10-13 chapters/discussion.tex
Updates to discussion

@@
 The calibration curve is the drawn from the $\{(\bar{p}_{b}, r_{b})\}_b$ pairs.
-A perfectly calibrated model would have the same fraction of correct predictions in any given bin, as that bin's mean value, $\bar{p}_{b} = r_{b}\,\forall\,b$.
+For any given bin, a perfectly calibrated model would have the same fraction of correct predictions as that bin's mean value, $\bar{p}_{b} = r_{b}\,\forall\,b$.
 
 The calibration curve for the uncalibrated stroke recognition ensemble and its constituent models is plotted in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model} along with a histogram of its predicted probabilities.
-The miscalibration issue that we previously discussed is clearly visible as a strong overconfidence.
-Additionally, the ensemble model is appreciably better calibrated than its constituent models.
-Since the ensemble output probability is computed as the harmonic mean of the five constituent model probabilities, the ensemble's output probability can never exceed the maximum probability predicted between the constituent models.
+The miscalibration issue that we previously discussed is clearly visible as a strong overconfidence for both ensemble and constituents, although the ensemble is much better calibrated than its constituents.
+Since the ensemble's output probability is computed as the harmonic mean of the five constituent model probabilities, it can never exceed the maximum probability predicted between the constituent models.
 This property tends to make ensemble probabilities less extreme and, since the constituent models are overconfident, this results in better calibration (see also the histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).
@@
         Calibration curve for the uncalibrated stroke recognition ensemble (left) and the histogram of predicted probabilities (right) for the test set. 
-        We use the ensemble that achieved the median F1-score reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
+        We use the ensemble that achieved the median F1-score reported in \cref{fig_retrospective:figure1-roc-curve} and \cref{tab_retrospective:table3-occlusion-analysis}.}
     \label{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}
@@
 To calibrate the ensemble model, we can use methods such as Platt-scaling \parencite{platt_probabilistic_1999} or isotonic regression \parencite{zadrozny_transforming_2002}. 
-We fit a simple regression model (logistic or isotonic) to the predicted probabilities and the target labels on the validation set and use it to adjust the probabilities predicted on the test set.
-We show the resulting calibration curves on the left in \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic} and the logistic and isotonic fits on the right.
+In either case, we fit a simple regression model (logistic or isotonic) to the predicted probabilities and the target labels on the validation set and use it to adjust the probabilities predicted on the test set.
+We show the resulting calibration curves on the left in \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic} and the logistic and isotonic fits on the right with 95\% bootstrap confidence intervals on the bin centers ($x$ error) and fraction of positives ($y$ error).
 % In \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic} we have done so for the ensemble model and visualize the results on the test set. 
@@
 Relative improvement in Brier score compared to uncalibrated (Brier skill score): Logistic = $0.4830$, isotonic = $0.4924$.} and that the predicted probabilities are shifted towards smaller values.
-Since there are few stroke cases overall, few examples receive a high predicted probability (see histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).
-This leads to a lack of data for the calibration fits at high predicted probabilities which can be seen to especially affect the nonparametric isotonic regression.
+Since stroke cases have low prevalence, high probability is predicted only for a few examples (see histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).
+This leads to a lack of data for the calibration fits at high predicted probabilities which can be seen to result in poor generalization to the test set, especially for the nonparametric isotonic regression.
 
@@
     \caption[Calibration fits and curves for the stroke recognition ensemble using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right) for the test set.
-We use the ensemble that achieved the median F1-score reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
+We use the ensemble that achieved the median F1-score reported in \cref{fig_retrospective:figure1-roc-curve} and \cref{tab_retrospective:table3-occlusion-analysis}.}
     \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
@@
 
-Clinicians in intensive care units and emergency departments have been found to strongly agree that a singular focus on overall accuracy cannot ensure sustained use of and trust in a model \cite{tonekaboni_what_2019}.
+Clinicians in intensive care units and emergency departments have been found to strongly agree that a singular focus on overall accuracy cannot alone ensure sustained trust in a model \cite{tonekaboni_what_2019}.
 Clinicians expect an alert to present a prediction that aligns with patient status.
-Despite expert-agreed thresholds for when alerts should be triggered, however, many alerts may not be aligned since the unbalanced and ambiguous nature of many predictive problems in healthcare can lead to low predictive precision \parencite{umscheid_development_2015, cite14, cite15, wenstrup_retrospective_2023}.
-This can lead to alarm fatigue \parencite{embi_evaluating_2012} and can undermine the use and endorsement by clinicians of such systems \parencite{guidi_clinician_2015}.
-The expected precision of the stroke recognition ensemble of $24.9\%$ (95\% confidence interval $24.3-25.5\%)$ means it might be wrong three out of four times it predicts a positive.
-This unfortunately makes alarm fatigue likely among its potential users which might diminish it's real-world impact.
+Despite expert-agreed thresholds for when alerts should be triggered, however, many alerts may not be aligned since the unbalanced and ambiguous nature of many predictive problems in healthcare can lead to relatively low predictive precision \parencite{umscheid_development_2015, cite14, cite15, wenstrup_retrospective_2023}.
+This is likely to lead to alarm fatigue \parencite{embi_evaluating_2012} and can undermine the sustained use and endorsement by clinicians of such systems \parencite{guidi_clinician_2015}.
+The expected precision of the stroke recognition ensemble presented in \cref{chp:paper-retrospective} is $24.9\%$ (95\% confidence interval $24.3-25.5\%)$.
+This bluntly means it might be wrong three out of four times it predicts a positive.
+This unfortunately risks alarm fatigue among its potential users which might diminish it's real-world impact.
 
-If the stroke recognition ensemble were to be deployed in a randomized controlled trial, such issues might be alleviated by incorporating the calibrated probabilities in the alerts presented to users.
-With calibrated probabilities, user would be able to discern between certain and uncertain model predictions or might be presented only with predictions that exceed a specific confidence level.
+% If the stroke recognition ensemble were to be deployed in a randomized controlled trial, s
+Such issues might be alleviated by incorporating the calibrated probabilities in the alerts presented to users.
+Presented with calibrated probabilities, users would be able to discern between certain and uncertain predictions.
+Furthermore, calibrated probabilities might be used to present users only with predictions that exceed a specific confidence level.
 Similar approaches have been suggested by clinicians and interviews indicate that predictive uncertainty is perceived as a sort of explanation that complements the prediction \cite{tonekaboni_what_2019}.
+In \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability} we show the F1-score of the ensemble model and the call-takers computed on subsets of the data created by binning the predicted probilities of the isotonically calibrated ensemble.
+We note that as might be expected, both call-taker and ensemble model performance increase with increased model certainty.
+This indicates that selecting, based on certainty, which predictions to present to users might indeed help improve trust in the system in practice.
 
 \begin{figure}
+    % \begin{subfigure}[c]{0.48\columnwidth}
+    %     \centering
+    %     \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/precision_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
+    %     % \caption{}
+    %     % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
+    % \end{subfigure}    
+    % % \hfill
+    % \begin{subfigure}[c]{0.48\columnwidth}
+    %     \centering
+    %     \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/recall_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
+    %     % \caption{}
+    %     % \label{fig_discussion:histogram_ensemble_and_single_model}
+    % \end{subfigure}
     \begin{subfigure}[c]{0.48\columnwidth}
@@
     \end{subfigure}    
-    % \hfill
     \begin{subfigure}[c]{0.48\columnwidth}
@@
     \caption[Comparison of F1-score of stroke recognition ensemble and call-takers as function of predicted probability.]{ Comparison of the F1-score of the stroke recognition ensemble and call-takers.
-The F1-score is computed on subsets of the test dataset made by binning on the calibrated probabilities of the ensemble.
+The F1-score is computed on subsets of the test dataset made by binning on the predicted probabilities of the calibrated ensemble.
 We see that the relative performance improvement of the ensemble over call-takers is higher towards more certain predictions.
-We use the ensemble that achieved the median F1-score reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
-    \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
+We use the ensemble that achieved the median F1-score reported in \cref{fig_retrospective:figure1-roc-curve} and \cref{tab_retrospective:table3-occlusion-analysis}.}
+    \label{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability}
 \end{figure}
 
-Inspecting the
+% Inspecting the
 
-\todo[inline]{Finish this sentence and analysis.}
+% \todo[inline]{Finish this sentence and analysis.}
 
-Nevertheless, while correct calibration provides some explanation to the predictions, the actual value of the calibrated probabilities is still tied of the model's performance.
-By inspecting \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}, we can see that the logistic calibration shifts the predicted probabilities so far towards zero that the highest possible predicted probability is about 50\%.
+% Nevertheless, while correct calibration provides some explanation to the predictions, the actual value of the calibrated probabilities is still tied of the model's performance.
+By inspecting \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability}, we can see that the logistic calibration shifts the predicted probabilities so far towards zero that the highest possible predicted probability is about 50\%.
 This means strokes can be predicted at best with a 50\% chance of being correct which limits the potential reduction in alarm fatigue.
 
-\todo[inline]{Finish this sentence and analysis.}
+% \todo[inline]{Finish this sentence and analysis.}
 

### 2023-10-13 chapters/introduction.tex
Updates to discussion

@@
 According to the investigation by the National Transportation Safety Board, the car's sensors detected the pedestrian more than five seconds before the crash, but the system did not manage to correctly predict her path or reduce the SUV's speed.
-Specifically, during those seconds, the system incorrectly classified the pedestrian more than ten times, first as a vehicle, then as an unknown object and ultimately as a bicyclist, each time changing, or resetting, the predicted path.
+Specifically, during those seconds, the system incorrectly classified the pedestrian more than ten times, first as a vehicle, then as an unknown object and ultimately as a bicyclist, each time changing, or resetting, the pedestrian's predicted path.
 About one second before the crash, the system determined that a collision was imminent, but the situation exceeded the constraints within which the autonomous driving system was allowed to operate, and the car's safety driver failed to intervene (\cref{fig:uber_nhsa_accident}) \parencite{nationaltransportationsafetyboardnhsa_collision_2019}.

### 2023-10-13 appendices/supplementary_benchmarking.tex
Add period to short captions

@@
 \begin{table}[t!]
-    \caption[Model likelihoods on TIMIT represented as a 16 bit linear PCM]{
+    \caption[Model likelihoods on TIMIT represented as a 16 bit linear PCM.]{
     Model likelihoods on TIMIT represented as a 16 bit linear PCM. The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
@@
 \begin{table}[p]
-    \caption[Model likelihoods on TIMIT represented as a 16 bit $\boldsymbol{\mu}$-law PCM]{
+    \caption[Model likelihoods on TIMIT represented as a 16 bit $\boldsymbol{\mu}$-law PCM.]{
     Model likelihoods on TIMIT represented as a 16 bit $\boldsymbol{\mu}$-law encoded PCM.
@@
 \begin{table}[t!]
-    \caption[Model likelihoods on TIMIT represented as globally normalized 16 bit linear PCM]{
+    \caption[Model likelihoods on TIMIT represented as globally normalized 16 bit linear PCM.]{
     Model likelihoods on TIMIT represented as globally normalized 16 bit linear PCM.
@@
 \begin{table}[t!]
-    \caption[The highest possible Gaussian log-likelihoods attainable on the TIMIT test set]{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
+    \caption[The highest possible Gaussian log-likelihoods attainable on the TIMIT test set.]{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
     \centering
@@
     \end{tikzpicture}
-\caption[CW-VAE cell state $\vs_t^l$ update]{CW-VAE cell state $\vs_t^l$ update.
+\caption[CW-VAE cell state $\vs_t^l$ update.]{CW-VAE cell state $\vs_t^l$ update.
 The cell state is given as $\vs_t^l=(\zb_t^l, \db_t^l)$ where $\db_t^l$ is the deterministic hidden state of a Gated Recurrent Unit \parencite{cho_properties_2014}.
@@
     \end{subfigure}
-\caption[Generative and inference models for three-layered CW-VAE]{
+\caption[Generative and inference models for three-layered CW-VAE.]{
     CW-VAE \parencite{saxena_clockwork_2021} generative model $p(\xb,\zb)$ in (\subref{fig: cwvae three-layered graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: cwvae three-layered graphical model inference}) for a three-layered model with $k_1=1$ and $c=2$ giving $k_2=2$ and $k_3=4$ unrolled over eight steps in the observed variable.
@@
     \end{subfigure}
-\caption[Generative and inference models for three-layered STCN]{
+\caption[Generative and inference models for three-layered STCN.]{
     STCN \parencite{aksan_stcn_2019} generative model $p(\xb,\zb)$ in (\subref{fig: stcn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: stcn graphical model inference}) for a single time-step.
@@
     \end{subfigure}
-\caption[Generative and inference models for VRNN]{
+\caption[Generative and inference models for VRNN.]{
     VRNN \parencite{chung_recurrent_2015} generative model $p(\xb,\zb)$ in (\subref{fig: vrnn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: vrnn graphical model inference}) unrolled over three steps in the observed variable.
@@
     \end{subfigure}
-\caption[Generative and inference models for SRNN]{
+\caption[Generative and inference models for SRNN.]{
     SRNN \parencite{fraccaro_sequential_2016} generative model $p(\xb,\zb)$ in (\subref{fig: srnn graphical model generative}) and inference model $q(\zb|\xb)$ in (\subref{fig: srnn graphical model inference}) unrolled over three steps in the observed variable.
@@
     \includegraphics[width=0.483\textwidth]{paper_benchmarking/1xrnjn5y_speaker_height_m_10_samples_knn_correct_fraction_model_vs_mel_50_neighbors_lda_subspace_2d.pdf}
-    \caption[Leave-one-out $k$-nearest-neighbour accuracy for speaker gender and height]{
+    \caption[Leave-one-out $k$-nearest-neighbour accuracy for speaker gender and height.]{
     Leave-one-out $k$-nearest-neighbor accuracy with different $k$ for
@@
      \end{subfigure}
-    \caption[Clustering of speaker gender in 2D linear subspace of a CW-VAE latent space]{(a) Clustering of speaker height for male speakers and (b) speaker age for female speakers in an two-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space.}
+    \caption[Clustering of speaker gender in 2D linear subspace of a CW-VAE latent space.]{\subref{fig: cwvae latent z1 height} Clustering of speaker height for male speakers and \subref{fig: cwvae latent z0 age} speaker age for female speakers in an two-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space.}
     \label{fig: cwvae latent height}
@@
     \end{subfigure}
-    \caption[Boxplots of duration of phoneme pronunciation in TIMIT]{Boxplots of the duration of the pronunciation of phonemes in TIMIT for a specific speaker DRW0 in (\subref{fig: timit validation set phoneme duration DRW0}) and globally in (\subref{fig: timit validation set phoneme duration global}).
+    \caption[Boxplots of duration of phoneme pronunciation in TIMIT.]{Boxplots of the duration of the pronunciation of phonemes in TIMIT for a specific speaker DRW0 in (\subref{fig: timit validation set phoneme duration DRW0}) and globally in (\subref{fig: timit validation set phoneme duration global}).
 Not all phonemes are pronounced by speaker DRW0 over the course of their 10 test set sentences and hence they are missing from the x-axis compared to the global durations.}

### 2023-10-13 appendices/supplementary_hierarchical.tex
Add period to short captions

@@
 \begin{table}[t]
-    \caption[Selection of most important hyperparameters]{Selection of most important hyperparameters and their setting.
+    \caption[Selection of most important hyperparameters.]{Selection of most important hyperparameters and their setting.
 Convolutional kernels are square and latent dimensions are given without spatial dimensions which are given in the text.
@@
 \begin{table}[t]
-    \caption[Additional results for the HVAE model trained on SVHN]{Additional results for the HVAE model trained on SVHN. All results computed with 1000 importance samples.}
+    \caption[Additional results for the HVAE model trained on SVHN.]{Additional results for the HVAE model trained on SVHN. All results computed with 1000 importance samples.}
     \centering
@@
 \begin{table}[t]
-    \caption[Additional results for the HVAE model trained on CIFAR10]{Additional results for the HVAE model trained on CIFAR10. All results computed with 1000 importance samples.}
+    \caption[Additional results for the HVAE model trained on CIFAR10.]{Additional results for the HVAE model trained on CIFAR10. All results computed with 1000 importance samples.}
     \centering
@@
 \begin{table}[t]
-    \caption[Additional results for the HVAE model trained on FashionMNIST]{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
+    \caption[Additional results for the HVAE model trained on FashionMNIST.]{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
     \centering
@@
 \begin{table}[t]
-    \caption[Additional results for the HVAE model trained on MNIST]{Additional results for the HVAE model trained on MNIST. All results computed with 1000 importance samples.}
+    \caption[Additional results for the HVAE model trained on MNIST.]{Additional results for the HVAE model trained on MNIST. All results computed with 1000 importance samples.}
     \centering

### 2023-10-13 appendices/supplementary_modelagnostic.tex
Add period to short captions

@@
     \centering
-    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing two different estimates of the Fisher information matrix]{AUROC$\uparrow$ for single-sample OOD detection.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing two different estimates of the Fisher information matrix.]{AUROC$\uparrow$ for single-sample OOD detection.
 Comparison between two different estimates of the Fisher information matrix.
@@
     \includegraphics[scale=0.5]{graphics/paper_modelagnostic/correlation_correct_font.pdf}
-    \caption[Correlation of typicality test and score statistic computed on the validation set using a PixelCNN++ trained on FashionMNIST]{Correlation of typicality test and score statistic computed on the validation set using a PixelCNN++ trained on FashionMNIST.
+    \caption[Correlation of typicality test and score statistic computed on the validation set using a PixelCNN++ trained on FashionMNIST.]{Correlation of typicality test and score statistic computed on the validation set using a PixelCNN++ trained on FashionMNIST.
 The correlation coefficient is $-0.014$.
@@
     %\renewcommand{\thetable}{4}
-    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing all single statistics with models trained on FashionMNIST (tested on MNIST) and CIFAR10 (tested on SVHN)]{AUROC$\uparrow$ for single-sample OOD detection.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing all single statistics with models trained on FashionMNIST (tested on MNIST) and CIFAR10 (tested on SVHN).]{AUROC$\uparrow$ for single-sample OOD detection.
 In this table we consider all the different single statistics we mentioned in the paper with models trained on FashionMNIST and CIFAR10.
@@
 \begin{table*}[tb]
-    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing all single statistics with models trained on MNIST (tested on FashionMNIST) SVHN (tested on CIFAR10)]{AUROC$\uparrow$ for single-sample OOD detection.
+    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing all single statistics with models trained on MNIST (tested on FashionMNIST) SVHN (tested on CIFAR10).]{AUROC$\uparrow$ for single-sample OOD detection.
 In this table we consider all the different single statistics we mentioned in the paper with models trained on MNIST and SVHN.
@@
     \vspace{-1.3em}
-    \caption[Comparison between our method and DoSE using the original statistics]{Comparison between our method and DoSE using the original statistics.
+    \caption[Comparison between our method and DoSE using the original statistics.]{Comparison between our method and DoSE using the original statistics.
 In these experiments we considered only Glow trained with Adam.}

### 2023-10-13 chapters/paper_automated.tex
Add period to short captions

@@
     \centering
-    \caption[Hyperparameters, maximum document lengths, and decision boundary tuning strategies used in the original works compared to improved settings]{Hyperparameters, maximum document lengths, and decision boundary tuning strategies used in the original works compared to the optimal settings found in this paper (marked with *).
+    \caption[Hyperparameters, maximum document lengths, and decision boundary tuning strategies used in the original works compared to improved settings.]{Hyperparameters, maximum document lengths, and decision boundary tuning strategies used in the original works compared to the optimal settings found in this paper (marked with *).
 LR is the learning rate scheduler.
@@
     \centering
-    \caption[Reproduced test set results compared with those from the original works]{Reproduced test set results compared with those from the original works.
+    \caption[Reproduced test set results compared with those from the original works.]{Reproduced test set results compared with those from the original works.
 Our reproduced results are indicated with *.
@@
     \end{subfigure}%
-    \caption[The relationship between chosen threshold and F1-score of every reproduced model in \cref{tab:reproduced_results}]{The relationship between chosen threshold and F1-score of every reproduced model in \cref{tab:reproduced_results}.
+    \caption[Relationship between chosen threshold and F1-score of every reproduced model.]{The relationship between chosen threshold and F1-score of every reproduced model in \cref{tab:reproduced_results}.
 The left figure shows the micro F1-score, and the right shows the macro F1-score.
@@
     \end{subfigure}
-    \caption[The relationship between the number of training examples and F1-score on MIMIC-IV \textit{ICD-9}]{The relationship between the number of training examples and F1-score on MIMIC-IV \textit{ICD-9}.
+    \caption[The relationship between the number of training examples and F1-score on MIMIC-IV \textit{ICD-9}.]{The relationship between the number of training examples and F1-score on MIMIC-IV \textit{ICD-9}.
 The left figure shows the F1 Micro score on the y-axis, while the right figure shows the F1 Macro score.}
@@
     \includegraphics[width=0.9\textwidth]{paper_automated/f1_micro_vs_num_words.png}
-    \caption[Relationship between the lengths of the clinical notes and the micro F1-score for each model on MIMIC-IV \textit{ICD-9}]{Relationship between the lengths of the clinical notes and the micro F1-score for each model on MIMIC-IV \textit{ICD-9}.
+    \caption[Relationship between the lengths of the clinical notes and the micro F1-score for each model on MIMIC-IV \textit{ICD-9}.]{Relationship between the lengths of the clinical notes and the micro F1-score for each model on MIMIC-IV \textit{ICD-9}.
 The vertical line indicates the maximum length of the notes after truncation.
@@
     \includegraphics[width=0.9\textwidth]{paper_automated/code_count_vs_f1_all_datasets.png}
-    \caption[Relationship between the code frequencies in the training set and the macro F1-score for PLM-ICD on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}]{Relationship between the code frequencies in the training set and the macro F1-score for PLM-ICD on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}.
+    \caption[Relationship between code frequencies and macro F1-score for PLM-ICD on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}.]{Relationship between the code frequencies in the training set and the macro F1-score for PLM-ICD on MIMIC-IV \textit{ICD-9} and \textit{ICD-10}.
 The shaded area indicates the standard deviation of the score computed for codes within the bin.}
@@
     \includegraphics[width=0.9\textwidth]{paper_automated/chapters.png}
-    \caption[Performance of PLM-ICD on ICD-10 chapters]{Performance of PLM-ICD on ICD-10 chapters.
+    \caption[Performance of PLM-ICD on ICD-10 chapters.]{Performance of PLM-ICD on ICD-10 chapters.
 Only codes with more than a hundred occurrences in the MIIMC-IV ICD-10 training set were considered, leaving 20 chapters.

### 2023-10-13 chapters/paper_modelagnostic.tex
Add period to short captions

@@
     \includegraphics[scale=0.6]{graphics/paper_modelagnostic/fashion_hvae_type1_correct_font.pdf}
-    \caption[Type I and Type II errors versus the significance level $\alpha$ on the combination values]{Type I (probability of an inlier to be classified as outlier) and Type II (probability of an outlier to be considered as inlier) errors versus the significance level $\alpha$ on the combination values.
+    \caption[Type I and Type II errors versus the significance level $\alpha$ on the combination values.]{Type I (probability of an inlier to be classified as outlier) and Type II (probability of an outlier to be considered as inlier) errors versus the significance level $\alpha$ on the combination values.
 By using Benjamini-Hochberg correction, we get that the Type I error stays below identity line.}

### 2023-10-13 chapters/paper_retrospective.tex
Add period to short captions

@@
     \includegraphics[width=0.9\textwidth]{paper_retrospective/figure1.pdf}
-    \caption[Receiver operator characteristic (ROC) curve and PPV-sensitivity curve for stroke recognition.]{Receiver operator characteristic (ROC) curve and PPV-sensitivity curve.
+    \caption[ROC- and PPV-sensitivity curves for stroke recognition model.]{Receiver operator characteristic (ROC) curve and PPV-sensitivity curve.
 Left, the ROC curve and, right, PPV-sensitivity curve (precision-recall curve).

### 2023-10-13 chapters/background.tex
Work on discussion

@@
 It is important to note that this distinction relates only to the availability of some target value, $y$ - not whether OOD data is available for supervision\footnote{
-    In our overview, we distinguish between supervised and unsupervised out-of-distribution by letting methods be classified as supervised that use any kind of target value $y$, whether it relates to the original task, available OOD data, or both.
+    In our overview, we distinguish between supervised and unsupervised out-of-distribution by letting methods be classified as supervised that use any kind of target value $y$, whether it relates to an original (in-distribution) task, available OOD data, or both.
 This is the same distinction made by \textcite{graham_denoising_2023} and \textcite{liu_unsupervised_2023a}.
-    However, it is important to note that in other works, the distinction is made based on whether the model is trained on OOD data or not \parencite{hendrycks_baseline_2017,liu_energy-based_2020}.
+    It is important to note, however, that in other works, the distinction is made based on whether the model is trained on OOD data or not \parencite{hendrycks_baseline_2017,liu_energy-based_2020}.
 This difference in nomenclature is currently unresolved in the literature.
@@
 Simply put, since the likelihood measures ``how probable the data is'', OOD data is expected to give lower likelihoods than in-distribution data.
-Although, this method originally gave encouraging results, the advent of deep generative models and their application to high-dimensional data has lead many to observe likelihoods for OOD data that are higher than for in-distribution data \parencite{choi_waic_2019,nalisnick_detecting_2019,kirichenko_why_2020}.
+Although, this method originally gave encouraging results, the advent of deep generative models and their application to high-dimensional data has lead many to observe likelihoods for OOD data that are higher than for in-distribution data \parencite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019,kirichenko_why_2020}.
 Such results have sparked interest in trying to explain this phenomenon and many works have proposed new scores for OOD detection derived from the likelihood.

### 2023-10-13 chapters/conclusion.tex
Work on discussion

@@
 
-In the introductory \cref{chp:introduction} we laid out how, through human history, progress-driving technological development has always introduced new challenges and risks of error whether via misuse, misunderstanding or inherent limitations.
+In the introductory \cref{chp:introduction} we laid out how progress-driving technological development has always come with new challenges and risks of error - whether via misuse, misunderstanding or inherent limitations.
 We emphasized that machine learning comes with these same types of risk, in some cases amplifying their impacts, and argued that uncertainty estimation is a key component in ensuring that systems are safe and reliable in practice.
-We provided background for the research project by introducing the healthtech company Corti, with which with project has been defined and carried out, and the motivational cases of recognition of stroke cases in emergency calls (\cref{subsec:motivation-stroke-recognition}) and automated medical coding (\cref{subsec:motivation-medical-coding}).
-We described how machine learning systems could be designed to aid healthcare professionals with the tasks and gave examples of potential sources of uncertainty and of how correctly quantifying uncertainty might improve the usefulness of such systems.
-Finally, we connected machine learning reliability (\cref{sec:machine-learning-reliability}) and model calibration (\cref{subsec:model-calibration}) and defined the aleatoric, epistemic, and predictive types of uncertainty (\cref{subsec:understanding-uncertainty}).
+We also provided the background for the research project by introducing the healthtech company Corti, with which with project has been defined and carried out, and the motivational cases of recognition of stroke cases in emergency calls (\cref{subsec:motivation-stroke-recognition}) and automated medical coding (\cref{subsec:motivation-medical-coding}).
+We described how machine learning systems could be designed to aid healthcare professionals with these tasks and gave examples of potential sources of uncertainty and of how correctly quantifying it might improve the usefulness of such systems.
+Finally, we drew connections between machine learning reliability (\cref{sec:machine-learning-reliability}) and model calibration (\cref{subsec:model-calibration}) and defined the aleatoric, epistemic, and predictive types of uncertainty (\cref{subsec:understanding-uncertainty}).
 
@@
 
-\dots
+In \cref{chp:paper-hierarchical}
+
+In \cref{chp:paper-modelagnostic}
+
+In \cref{chp:paper-benchmarking}
+% In chapter 8, stochastic and deterministic generative models were benchmarked in terms of model likelihood.
+In addition, the chapter presented a hierarchical latent variable model (LVM) for speech inspired by the Clockwork VAE [238].
+In contrast to self-supervised models, LVMs have not yet seen the same success for representation learning.
+As such, they offer little guidance in terms of alleviating the need for labeled data, as discussed in section 1.2.1.
+
+% However, since LVMs learn a distribution over the training data, they are forced to encode all aspects of it.
+While this may be redundant for some tasks, others benefit from a broad variety of features.
+A hierarchical model that operates on multiple temporal scales, such as the Clockwork VAE, offers a natural way to encode different feature categories.
+Pronunciation might be learned at lower layers, speaker identity at the upper layers, and semantic features in between.
+Although some work has successfully separated speaker identity from content [114], models that can learn a deep hierarchy of features for speech remains an open challenge.
+The model presented in chapter 8 represents an effort to reignite this area of research.
+
+\cref{chp:paper-review}
+% Chapter 6 presented an overview of unsupervised neural speech representation learning.
+As emphasized in the overview, the wav2vec 2.0 framework [13], and masked pre-training in general, represent a breakthrough in low-resource speech recognition.
+The challenge associated with obtaining training data for conversational speech recognition, as discussed in section 1.2.1, has become much smaller.
+Even when no labeled in-domain data is available, this framework offers a viable solution [112].
+
+% The general idea for masked pre-training may seem simple; reconstruct masked parts of the input or another representation, given context.
+However, this methodology aligns very well with the general definition of semantics presented in section 1.4: "semantic properties of a lexical item are fully reflected in appropriate aspects of the relations it contracts with actual and potential contexts" ([63], p.
+1).
+Thus, from a philosophical point-of-view, it may be difficult to see how the field should move on from here.
+Of course, context is more than just the neighboring words in a speech segment.
+The next wave of representation learning for speech incorporates multiple modalities, such as video [244, 245] or text [18].
+Furthermore, how to construct targets for pre-training these models and how that affects the learned features is an avenue that warrants more research.
+This is of particular interest in the speech domain, where the input codes for speaker identity and emotional state, as well as the semantic content of the utterance.
+
+\cref{chp:paper-automated}
+
+\cref{chp:paper-retrospective}

### 2023-10-13 chapters/discussion.tex
Work on discussion

@@
 
-During the three years since the start of this project, the rapid development of machine learning that started about a decade ago with the 2012 ImageNet competation and the work of \textcite{krizhevsky_imagenet_2012} all but slowed down.
-In this section we will briefly revisit selected chapters of the thesis and provide some additional thoughts and discussion.
+The last three years have seen the continuation of the rapid progress of machine learning that started about a decade ago with the 2012 ImageNet competation and the work of \textcite{krizhevsky_imagenet_2012}.
+The field of out-of-distribution detection saw a surge of interest following the papers by \textcite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019} and self"=supervised learning for speech has seen rapid advances with works such as CPC \parencite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.
+
+Having focused in the previous chapters on VAEs for out-of-distribution detection and self"=supervised methods for representation learning, this chapter will add perspective on how VAEs might be enabled to learn representations more competetive with self"=supervised approaches, and how self"=supervised approaches might be used for out-of-distribution detection.
+Finally, we will discuss the use of calibration techniques for the stroke recognition model of \cref{chp:paper-retrospective}.
+
+% Following the papers by \textcite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019}, the field of out-of-distribution detection saw a surge of new approaches using both generative models, supervised classifiers and self"=supervised representations.
+% Similarly, though the wave of works in self"=supervised learning might be traced back to \texttt{word2vec} of \textcite{mikolov_efficient_2013}, the impact of later works such as BERT \parencite{devlin_bert_2018} and GPT \parencite{radford_improving_2018} was immense and helped give rise to many approaches for self"=supervised learning for speech, including CPC \parecite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.
+%
 
@@
 \cref{tab: phoneme recognition (PER), tab:unsupervisedASR}).
-Considering VAEs and self"=supervised methods, it seems we cannot build a model that has a principled probabilistic representation of uncertainty without sacrificing some of the ability to learn rich representations that are useful for downstream applications.
+Considering VAEs and self"=supervised methods, it seems we have not yet been able to build a model that has a principled probabilistic representation of uncertainty without sacrificing some of the ability to learn rich representations that are useful for downstream applications.
 
@@
 
-One way to improve the representations learned in VAEs is via semi-supervised learning. 
+\paragraph{Semi-supervised learning} One way to improve the representations learned in VAEs is via semi-supervised learning. 
 Here a small number of labels are used to inform which patterns that are learned from a large, mostly unlabelled, data set. 
@@
 VAEs are strong models for semi-supervised learning as proven in several works \parencite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}. 
-However, self"=supervised methods methods have established themselves as superior also in this setting \cite{jiang_speech_2021, liu_learning_2023}. 
-Furthermore, by training on labelled and unlabelled data simultaneously, the semi-supervised setting of VAEs is more constraining than it is for self-supervised methods that divide the training into pre"=training and finetuning.
-Here, expensive pre"=training yields a single foundation model that can be fine"=tuned at relatively low cost for a large number of downstream tasks, while VAEs must learn all relevant downstream tasks while also learning from the unlabelled data, which is more expensive, but also requires retraining on the unlabelled data, when a .
+However, self"=supervised methods methods have established themselves as superior for most tasks in this setting \cite{jiang_speech_2021, liu_learning_2023}. 
+Furthermore, by training on labelled and unlabelled data simultaneously, the semi-supervised formulation for VAEs is more constraining than it is for self-supervised methods that divide the training into pre"=training and finetuning.
+The expensive pre"=training yields a single foundation model that can be fine"=tuned at relatively low cost for a large number of downstream tasks, whereas VAEs must learn all relevant downstream tasks while also learning from the unlabelled data.
+This is more expensive, but also requires retraining on the unlabelled data, when new supervised data becomes available.
 
-Masking in variational autoencoders
+\paragraph{Masking} As discussed in \cref{chp:paper-review}, one of the driving techniques behind the success of self"=supervised methods is masking \parencite{devlin_bert_2018,baevski_wav2vec_2020}.
+By removing tokens from the input or an early feature extraction layer and tasking a model with inferring their representations, the model is forced to learn how neighbouring tokens relate to those masked.
+Depending on the size of the mask, these dependencies can be more or less semantic in nature.
 
-Masked pre"=trained as maximum likelihood
+Different from self"=supervised methods, since VAEs learn the distribution over the training data, they are forced to encode all aspects important to it. 
+As we also saw in \cref{chp:paper-hierarchical}, this will generally include low-level features that are important for accurate input reconstruction, but usually of lesser interest for downstream tasks.
+Furthermore, VAE reconstruction is performed based on latent variables which are inferred from the full, unmasked input.
+
+Masking has only been sparsely examined for VAEs.
+One of the areas where it has seen the most attention is missing data imputation.
+In this setting, the input is partially observed, and often represented as a segmentation into observed and missing parts via a mask that indicates where the data is missing.
+The model is then trained to infer the latent variable from the observed data and reconstruction also deals only with the observed data.
+By comparison to self"=supervised approaches that use masking for representation learning, the missing data setting of VAEs focuses on reconstructing the observed data rather than the missing, which might lead to less useful representations.
+The idea of using VAEs to impute missing data was already examined in the seminal paper by \textcite{rezende_stochastic_2014}.
+Here the model was trained with fully observed data and used to impute data in an iterative sampling approach post hoc, leaving the learned representations unchanged.
+Previous work that trains on partially observed data has largely focused on the ability of these models to yield high-quality imputations within the tabular and image data domains and have not probed for the effects on the learned latent representation \parencite{mattei_miwae_2019, ipsen_not-miwae_2021}.
 
@@
 
-\section{\Cref{chp:paper-hierarchical} revisited: \dots}
+Uncertainty of representations versus uncertainty of downstream task outputs. 
+% \textcite{wickstrom_relax_2023} propose a method for quantifying uncertainty based on
 
-\lesstodo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other).
+% \section{\Cref{chp:paper-hierarchical} revisited: \dots}
+
+% \lesstodo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other).
 Include reference to \parencite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
-\lesstodo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimizing a mutual information term in the ELBO (derive this form).}
-\lesstodo[inline]{Discuss some recent work e.g. \cref{morningstar_density_2021}}
+% \lesstodo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimizing a mutual information term in the ELBO (derive this form).}
+% \lesstodo[inline]{Discuss some recent work e.g. \cref{morningstar_density_2021}}
 
-\section{\Cref{chp:paper-benchmarking} revisited: Bested?}
+% \section{\Cref{chp:paper-benchmarking} revisited: Bested?}
 
-\lesstodo[inline]{Discuss whether variational autoencoders are viable for learning good representations for speech for downstream tasks when alternatives such as SSL method exist.}
-\lesstodo[inline]{Out-of-distribution detection on speech?}
+% \lesstodo[inline]{Discuss whether variational autoencoders are viable for learning good representations for speech for downstream tasks when alternatives such as SSL method exist.}
+% \lesstodo[inline]{Out-of-distribution detection on speech?}
 
-\section{\Cref{chp:paper-review} revisited: \dots} \label{sec:discussion-paper-review}
+% \section{\Cref{chp:paper-review} revisited: \dots} \label{sec:discussion-paper-review}
 
-\lesstodo[inline]{Are self"=supervised speech representations useful for unsupervised uncertainty estimation?
+% \lesstodo[inline]{Are self"=supervised speech representations useful for unsupervised uncertainty estimation?
 \parencite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}
-% Since the main focus within self"=supervised learning has been on improving downstream task performance, very limited work, if any, has investigated self"=supervised representations in terms of uncertainty estimation.
-% However, in the context of medical applications where data can be abundant but labels are sparse, unsupervised uncertainty estimation is a very interesting direction for future work.
-\lesstodo[inline]{OOD data: Generalization or detection (https://arxiv.org/pdf/2110.11334.pdf)?}
+% % Since the main focus within self"=supervised learning has been on improving downstream task performance, very limited work, if any, has investigated self"=supervised representations in terms of uncertainty estimation.
+% % However, in the context of medical applications where data can be abundant but labels are sparse, unsupervised uncertainty estimation is a very interesting direction for future work.
+% \lesstodo[inline]{OOD data: Generalization or detection (https://arxiv.org/pdf/2110.11334.pdf)?}
 
 \section{\Cref{chp:paper-retrospective} revisited: Calibration}
-Our work on stroke recognition in \cref{chp:paper-retrospective} focuses on the predictive performance of the ensemble model and an analysis of feature importance, but does not explicitly consider uncertainty estimation.
-As we discussed in \cref{subsec:model-calibration}, such a model can be calibrated to predict probabilities that are aligned with the empirical probability of the model being correct on some validation set.
-Here we present and discuss model calibration for the ensemble model.
-
-We compute the calibration curve by sorting the probabilities predicted on the test set into a number of bins spanning the range from zero to one.
-For each bin $b$, we compute the mean predicted probability $\bar{p}$ and the fraction of examples for which the model predicted correctly $r_{b}$.
-The calibration curve is the drawn from the $\{(\bar{p}_{b}, r_{b})\}_b$ pairs.
-For any given bin, a perfectly calibrated model would have the same fraction of correct predictions as that bin's mean value, $\bar{p}_{b} = r_{b}\,\forall\,b$.
-
-The calibration curve for the uncalibrated stroke recognition ensemble and its constituent models is plotted in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model} along with a histogram of its predicted probabilities.
-The miscalibration issue that we previously discussed is clearly visible as a strong overconfidence for both ensemble and constituents, although the ensemble is much better calibrated than its constituents.
-Since the ensemble's output probability is computed as the harmonic mean of the five constituent model probabilities, it can never exceed the maximum probability predicted between the constituent models.
-This property tends to make ensemble probabilities less extreme and, since the constituent models are overconfident, this results in better calibration (see also the histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).
 
@@
 
+Our work on stroke recognition in \cref{chp:paper-retrospective} focuses on the predictive performance of the ensemble model and an analysis of feature importance, but does not explicitly consider uncertainty estimation.
+As we discussed in \cref{subsec:model-calibration}, such a model can be calibrated to predict probabilities that are aligned with the empirical probability of the model being correct on some validation set.
+Here we present and discuss model calibration for the ensemble model.
+
+We compute the calibration curve by sorting the probabilities predicted on the test set into a number of bins spanning the range from zero to one.
+For each bin $b$, we compute the mean predicted probability $\bar{p}$ and the fraction of examples for which the model predicted correctly $r_{b}$.
+The calibration curve is the drawn from the $\{(\bar{p}_{b}, r_{b})\}_b$ pairs.
+For any given bin, a perfectly calibrated model would have the same fraction of correct predictions as that bin's mean value, $\bar{p}_{b} = r_{b}\,\forall\,b$.
+
+The calibration curve for the uncalibrated stroke recognition ensemble and its constituent models is plotted in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model} along with a histogram of its predicted probabilities.
+The miscalibration issue that we previously discussed is clearly visible as a strong overconfidence for both ensemble and constituents, although the ensemble is much better calibrated than its constituents.
+Since the ensemble's output probability is computed as the harmonic mean of the five constituent model probabilities, it can never exceed the maximum probability predicted between the constituent models.
+This property tends to make ensemble probabilities less extreme and, since the constituent models are overconfident, this results in better calibration (see also the histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).
+
+\begin{figure}
+    \centering
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curves_ensemble_with_cis.pdf}
+    \end{subfigure}    
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_fits_ensemble.pdf}
+    \end{subfigure}    
+    \caption[Calibration fits and curves for the stroke recognition ensemble using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right) for the test set.
+We use the ensemble that achieved the median F1-score reported in \cref{fig_retrospective:figure1-roc-curve} and \cref{tab_retrospective:table3-occlusion-analysis}.}
+    \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
+\end{figure}
+\begin{figure}
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curves_ensemble_with_cis.pdf}
+    \end{subfigure}    
+    \begin{subfigure}[c]{0.48\columnwidth}
+        \centering
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_fits_ensemble.pdf}
+    \end{subfigure}    
+    \caption[Calibration fits and curves for the stroke recognition ensemble using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right) for the test set.
+We use the ensemble that achieved the median F1-score reported in \cref{fig_retrospective:figure1-roc-curve} and \cref{tab_retrospective:table3-occlusion-analysis}.}
+    \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
+\end{figure}
+
 To calibrate the ensemble model, we can use methods such as Platt-scaling \parencite{platt_probabilistic_1999} or isotonic regression \parencite{zadrozny_transforming_2002}. 
@@
 \begin{figure}
+    % \begin{subfigure}[c]{0.48\columnwidth}
+    %     \centering
+    %     \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/precision_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
+    %     % \caption{}
+    %     % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
+    % \end{subfigure}    
+    % % \hfill
+    % \begin{subfigure}[c]{0.48\columnwidth}
+    %     \centering
+    %     \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/recall_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
+    %     % \caption{}
+    %     % \label{fig_discussion:histogram_ensemble_and_single_model}
+    % \end{subfigure}
     \begin{subfigure}[c]{0.48\columnwidth}
         \centering
-        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curves_ensemble_with_cis.pdf}
-        % \caption{}
-        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/f1_score_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
     \end{subfigure}    
-    % \hfill
     \begin{subfigure}[c]{0.48\columnwidth}
         \centering
-        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_fits_ensemble.pdf}
-        % \caption{}
-        % \label{fig_discussion:histogram_ensemble_and_single_model}
+        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/predicted_probability_histogram.pdf}
     \end{subfigure}    
-    \caption[Calibration fits and curves for the stroke recognition ensemble using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right) for the test set.
+    \caption[Comparison of F1-score of stroke recognition ensemble and call-takers as function of predicted probability.]{ Comparison of the F1-score of the stroke recognition ensemble and call-takers.
+The F1-score is computed on subsets of the test dataset made by binning on the predicted probabilities of the calibrated ensemble.
+We see that the relative performance improvement of the ensemble over call-takers is higher towards more certain predictions.
 We use the ensemble that achieved the median F1-score reported in \cref{fig_retrospective:figure1-roc-curve} and \cref{tab_retrospective:table3-occlusion-analysis}.}
-    \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
+    \label{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability}
 \end{figure}
@@
 
-\begin{figure}
-    % \begin{subfigure}[c]{0.48\columnwidth}
-    %     \centering
-    %     \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/precision_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
-    %     % \caption{}
-    %     % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
-    % \end{subfigure}    
-    % % \hfill
-    % \begin{subfigure}[c]{0.48\columnwidth}
-    %     \centering
-    %     \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/recall_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
-    %     % \caption{}
-    %     % \label{fig_discussion:histogram_ensemble_and_single_model}
-    % \end{subfigure}
-    \begin{subfigure}[c]{0.48\columnwidth}
-        \centering
-        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/f1_score_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
-        % \caption{}
-        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
-    \end{subfigure}    
-    \begin{subfigure}[c]{0.48\columnwidth}
-        \centering
-        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/predicted_probability_histogram.pdf}
-        % \caption{}
-        % \label{fig_discussion:histogram_ensemble_and_single_model}
-    \end{subfigure}    
-    \caption[Comparison of F1-score of stroke recognition ensemble and call-takers as function of predicted probability.]{ Comparison of the F1-score of the stroke recognition ensemble and call-takers.
-The F1-score is computed on subsets of the test dataset made by binning on the predicted probabilities of the calibrated ensemble.
-We see that the relative performance improvement of the ensemble over call-takers is higher towards more certain predictions.
-We use the ensemble that achieved the median F1-score reported in \cref{fig_retrospective:figure1-roc-curve} and \cref{tab_retrospective:table3-occlusion-analysis}.}
-    \label{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability}
-\end{figure}
-
-% Inspecting the
-
-% \todo[inline]{Finish this sentence and analysis.}
-
-% Nevertheless, while correct calibration provides some explanation to the predictions, the actual value of the calibrated probabilities is still tied of the model's performance.
-By inspecting \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability}, we can see that the logistic calibration shifts the predicted probabilities so far towards zero that the highest possible predicted probability is about 50\%.
-This means strokes can be predicted at best with a 50\% chance of being correct which limits the potential reduction in alarm fatigue.
-
-% \todo[inline]{Finish this sentence and analysis.}
-
-% Making matters worse, the most certain model predictions are the ones that agree the most with call-taker judgment (see \cref{fig_discussion:retrospective-paper-fractional-agreement-vs-probability}).
-Hence, keeping the most certain predictions also means reducing the potential for changing call-taker..
-Not really if when the model disagrees at low prob it is always right while when it disagrees at high prob it is always wrong (although unlikely and contrary to calibration - but that was done for the total set of predictions, not subgroups)
-
-% This highlights a problem where combining predictions with well-calibrated uncertainties may not be sufficient to ensure sustained use in practice.
-
-% \lesstodo[inline]{Discuss the calibration of the stroke model and report calibration curve.}
-% \lesstodo[inline]{Report the results of calibrating the stroke model using e.g. Platt scaling or Isotonic scaling.} % https://scikit-learn.org/stable/modules/calibration.html
-
 % \lesstodo[inline]{Maybe mention MultiQT paper \parencite{havtorn_multiqt_2020}.}

### 2023-10-13 chapters/introduction.tex
Work on discussion

@@
 At the start of this project in 2020, Corti's main product was a software system providing decision-support for call-takers that operate the 1-1-2 emergency line in Copenhagen, Denmark, and the 9-1-1 emergency line in Seattle, Washington, in the United States.
-The system is in continued use in 2023, and provides an interface for navigating a graph-based protocol, defined by the local emergency medical services, which dispatchers use to triage patients during emergency calls.
-Calls are transcribed in real-time using automatic speech recognition built for the customer-domain and the system makes suggestions for the dispatcher to use in the conversation with the caller, including notifications about forgotten protocol questions \parencite{havtorn_multiqt_2020} and potential cases of cardiac arrests \parencite{cite15, cite14}.
-The system logs the details about the call and the actions taken by the dispatcher for later use in review, training and quality assurance.
+The system is in continued use in 2023, and provides an interface for navigating a graph-based protocol, defined by the local emergency medical services, which call-takers use to triage patients during emergency calls.
+Calls are transcribed in real-time using automatic speech recognition built for the customer-domain and the system makes suggestions for the call-taker to use in the conversation with the caller, including notifications about forgotten protocol questions \parencite{havtorn_multiqt_2020} and potential cases of cardiac arrests \parencite{cite15, cite14}.
+The system logs the details about the call and the actions taken by the call-taker for later use in review, training and quality assurance.
 
@@
 Additionally, stroke cases are relatively rare, occurring in only $0.25\%$ of all calls made to the Copenhagen 1-1-2 emergency line in 2021.
-This makes it difficult for call-takers to gain practical experience with stroke cases which compounds to the difficulty of recognizing them.
+This makes it difficult for call-takers to gain practical experience with stroke cases which compounds to the fundamental difficulty of recognizing them.
 Although several efforts have been made to improve recognition rates, there is still a need for better tools to support call-takers in recognizing stroke cases \parencite{cite13,cite14,cite15}.
@@
 For instance, the widely used International Classification of Diseases (ICD) standard consist of 55,000 medical codes in version 10 and 85,000 in version 11 \parencite{worldhealthorganisationwho_international_2023}.
-Additionally, a single medical note will usually contain several diagnoses and procedures which must all be inferred from the clinical note \parencite{johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}.
-Furthermore, any single code can have several criteria defined by official guidelines that determine under which conditions the code is mutually exclusive with other codes, and when other codes must be coded along with it \parencite{centersformedicaremedicaidservicesus_icd10cm_2023}.
+Additionally, a single clinical note will usually contain several diagnoses and procedures which must all be inferred from the natural language text \parencite{johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}.
+Furthermore, any single code can have several criteria defined by official guidelines that determine under which conditions it is mutually exclusive with other codes, and when other codes must be coded along with it.
+These properties and more make medical coding a very complex, multi-label classification problem \parencite{centersformedicaremedicaidservicesus_icd10cm_2023}.
 % Official guidelines describe how to code specific diagnoses and procedures and combinations thereof . 
@@
 
-\paragraph{Uncertainty in automated medical coding} A reasonable machine learning pipeline to assist in medical coding might consist of a natural language processing model that extracts relevant information from the clinical note and outputs a number of probability distributions that each estimate the probability that the note should be assigned one of the medical codes.
+\paragraph{Uncertainty in automated medical coding} A reasonable machine learning pipeline to assist in medical coding might consist of a natural language processing model that extracts relevant information from the clinical note and outputs a probability distribution for each of the potential medical codes.
 However, training and using such as system comes with several challenges. 
@@
 For example, if the model is uncertain about two similar codes, it might be appropriate to ask a human expert to review the note and make the final decision.
-Additionally, the lack of training data for the rarest codes makes them difficult to learn to robustly predict, and so, we would often want the model to be uncertain for rare codes.
+Additionally, the lack of training data for the rarest codes makes them difficult to learn to robustly predict, and so, we might expect the model to often be uncertain for rare codes.
 
@@
 
-Miscalibrated models are unreliable, especially in high risk applications.
-Although no model is perfect, the inability of a human, or another model, to assess the certainty with which a prediction was made, makes it impossible to know how to weigh it in a decision-making process.
+Miscalibrated models are unreliable, especially in high risk applications, and although no model is perfect, the inability of a human, or a model, to assess the certainty with which a prediction was made, makes it impossible to know how to weigh it in a decision-making process.
 % A connection can be made back to the incident with the Uber self-driving car. 
@@
 
-Finally, \cref{part:discussion-and-conclusion} concludes the thesis by summarizing the main contributions and discussing the presented work and future directions for research.
+Finally, \cref{part:discussion-and-conclusion} concludes the thesis by discussing the presented work and presenting future directions for research.

### 2023-10-13 chapters/main-contributions.tex
Work on discussion

@@
 
-The following chapters of the thesis are self-contained studies and therefore detail their own research hypotheses and contributions, but do so without consideration to the other chapters. 
-For that reason, we here lay out research hypotheses and contributions in relation to the overall research project. 
+The following chapters of the thesis are self-contained studies and therefore detail their own research hypotheses and contributions.
+Since each study was written without consideration to the other chapters of the thesis, we here detail them in relation to the overall research project. 
+This constitutes the research hypotheses and contributions of the thesis.
+
 % The chapters are structured into three parts: {\color{black}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{black}\cref{part:self-supervised-speech-representation-learning}} is concerned with self"=supervised speech representation learning, and {\color{black}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.
@@
 The first two papers approach unsupervised uncertainty estimation through the task of out-of-distribution detection using generative models. 
-The first paper proposes a new method for out-of-distribution detection using hierarchical variational autoencoders and the second proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable generative model with explicit likelihood.
+The first paper proposes a new method for out-of-distribution detection using hierarchical variational autoencoders and the second proposes a model-agnostic method for that can be used with any differentiable generative model with explicit likelihood.
 The third paper proposes a new hierarchical variational autoencoder for speech and benchmarks it against other variational and autoregressive models.
@@
 In this work we hypothesize that the likelihood estimate of variational autoencoders is a poor score for out-of-distribution due to an overemphasis on low-level features that generalize between distributions.
-We further hypothesize that a well-formed hierarchy of latent variables provides a tool that can be used to select between which features to emphasize for out-of-distribution detection and hence a way to improve the performance of variational autoencoders on this task.
+We further hypothesize that a well-formed hierarchy of latent variables provides a tool that can be used to select which features to emphasize for out-of-distribution detection and hence a way to improve the performance of variational autoencoders on this task.
 We proceed to provide empirical and theoretical evidence that low-level features indeed do dominate the likelihood and propose a new method for out-of-distribution detection using hierarchical variational autoencoders.
@@
 \contribitem{chp:paper-modelagnostic}
-In this follow-up work, we note that many diverse methods have been proposed for out-of-distribution detection using different generative models and seek to develop a model-agnostic approach. 
-We hypothesize that by phrasing the problem as a statistical testing problem and combining different tests, the method's efficacy can be improved and weaknesses inherent to any particular test can be alleviated.
-From this hypothesis, we develop a method that combines a classical parametric test with the recently introduced typicality test via Fisher's method and can be used with any differentiable, explicit likelihood generative model.
+In this follow-up work, we note that many methods have been proposed for out-of-distribution detection using generative models, with some tailored for specific model types, and seek to develop a model-agnostic approach.
+We hypothesize that by phrasing the problem as a statistical testing problem and combining different tests, the method's efficacy can be improved and weaknesses inherent to any particular test alleviated.
+From this hypothesis, we develop a method that combines a classical parametric test with the recently introduced typicality test and can be used with any differentiable, explicit likelihood generative model.
 We show that this leads to a more accurate out-of-distribution test. 
@@
 Later, in the discussion, we shall further consider uncertainty estimation in relation to these two papers.
-\todo[inline]{Revise and add more details to this section.}
+\todo[inline]{Revise and add more details to this section. Do we discuss uncertainty for medical coding?}
 
@@
 In this paper, we review the current state-of-the-art in automated medical coding of clinical notes. 
-We hypothesize that several previous works underperform for reasons more related to suboptimal hyperparameter tuning, incorrect evaluation, and crude data handling than to model design, and that performance can be improved by addressing these issues.
-With our work, we aim to increase reproducibility and comparability of future work and to provide a baseline for future comparisons. 
-We reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is indeed attributable to weak configuration of the training and crudely sampled train-test splits with many extremely rare classes and several without examples in the training data at all.
-We also find a widespread error in the calculation of the macro F1-score. 
+We hypothesize that several previous works underperform for reasons more related to suboptimal hyperparameter tuning, incorrect evaluation, and crude data handling than to model design, and that performance and comparability can be improved by addressing these issues.
+With our work, we aim to improve reproducibility of future work and to provide a baseline for future comparisons. 
+We first reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is indeed attributable to weak configuration of the training and crudely sampled train-test splits with many extremely rare classes and several without examples in the training data at all.
+We also identify and correct a widespread error in the calculation of the macro F1-score. 
 To compare models, we propose new data splits created with stratified sampling, use identical experimental setups and tune hyperparameters and decision boundaries. 
@@
 On test data from 2021, call-takers achieved an overall sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%) while the machine learning framework performed significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
-Effective stroke treatment depends on recognition by call-takers at prehospital telehealth services. 
-This study provides preliminary evidence that a machine learning framework for recognizing stroke in prehospital medical helpline calls could become a supportive tool for call-takers, aiding in early and accurate stroke recognition.
+Effective treatment of out-of-hospital stroke often hinges on recognition by call-takers at prehospital telehealth services. 
+This study provides preliminary evidence that a machine learning framework could become a supportive tool for call-takers at prehospital medical helplines, aiding in early and accurate stroke recognition.
 % This paper aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.

### 2023-10-14 appendices/paper_brief.tex
Fix many spelling and language errors

@@
     \includegraphics[width=0.90\textwidth]{paper_brief/SSL_LVM.pdf}
-    \caption[ Schematic overview of the self"=supervised and probabilistic latent variable models covered in this survey.]{A schematic overview of the two groups of models covered in this survey.
+    \caption[Schematic overview of the self"=supervised and probabilistic latent variable models covered in this survey.]{ A schematic overview of the two groups of models covered in this survey.
 \textit{Left}: A model trained with self"=supervised learning.
@@
 
-In brief, LVMs of the VAE-type consist of a approximate posterior, $q(\mathbf{z}|\mathbf{x})$, an observation model, $p(\mathbf{x}|\mathbf{z})$, and a prior, $p(\mathbf{z})$. 
+In brief, LVMs of the VAE-type consist of an approximate posterior, $q(\mathbf{z}|\mathbf{x})$, an observation model, $p(\mathbf{x}|\mathbf{z})$, and a prior, $p(\mathbf{z})$. 
 With reference to probabilistic coding theory, the approximate posterior is often referred to as the encoder and the observation model as the decoder \parencite{kingma_autoencoding_2014, rezende_stochastic_2014}.
@@
 \begin{table}
-    \caption[A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable.]{A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable.
+    \caption[A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable.]{ A comprehensive overview of observation, prior and inference models for VAE type latent variable models with a single latent variable.
 The observation, prior and inference models may all belong to one or more of the categories listed under them as detailed in \cref{sec:plvms}.

### 2023-10-14 appendices/supplementary_benchmarking.tex
Fix many spelling and language errors

@@
 \paragraph{LibriSpeech and LibriLight}
-The LibriSpeech dataset \parencite{panayotov_librispeech_2015} consists of readings of public domain audio books amounting to approximately $\SI{1000}{h}$ of audio.
+The LibriSpeech dataset \parencite{panayotov_librispeech_2015} consists of readings of public domain audiobooks amounting to approximately $\SI{1000}{h}$ of audio.
 The data is derived from the LibriVox project.
@@
         \midrule
-        $1$       & Wavenet             & $D_C=96$              & \textbf{7.246} \\
+        $1$       & WaveNet             & $D_C=96$              & \textbf{7.246} \\
         $1$       & LSTM                & $D_d=256, L=1$        & 7.295 \\
@@
         \midrule
-        $1$       & Wavenet             & $D_C=16$              & 11.27 \\
-        $1$       & Wavenet             & $D_C=24$              & 11.14 \\
-        $1$       & Wavenet             & $D_C=32$              & 11.03 \\
-        $1$       & Wavenet             & $D_C=96$              & 10.88 \\
-        $1$       & Wavenet             & $D_C=128$             & 10.98 \\
-        $1$       & Wavenet             & $D_C=160$             & 10.91 \\
+        $1$       & WaveNet             & $D_C=16$              & 11.27 \\
+        $1$       & WaveNet             & $D_C=24$              & 11.14 \\
+        $1$       & WaveNet             & $D_C=32$              & 11.03 \\
+        $1$       & WaveNet             & $D_C=96$              & 10.88 \\
+        $1$       & WaveNet             & $D_C=128$             & 10.98 \\
+        $1$       & WaveNet             & $D_C=160$             & 10.91 \\
         $1$       & LSTM                & $D_d=128, L=1$        & 11.40 \\
@@
 \section{Additional graphical models}\label{app: additional graphical models}
-In \cref{fig: cwvae cell state graph} we show the graphical model of the recurrent cell of the CW-VAE for a single time step.
+In \cref{fig: cwvae cell state graph} we show the graphical model of the recurrent cell of the CW-VAE for a single timestep.
 As noted in \parencite{saxena_clockwork_2021}, this cell is very similar to the one of the Recurrent State Space Model (RSSM) \parencite{hafner_learning_2019}.
@@
      \end{subfigure}
-    \caption[Clustering of speaker gender in 1D linear subspace of a CW-VAE latent space and a time-averaged mel spectrogram.]{Clustering of speaker gender in an one-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space and of a time-averaged mel spectrogram.
+    \caption[Clustering of speaker gender in 1D linear subspace of a CW-VAE latent space and a time-averaged Mel spectrogram.]{ Clustering of speaker gender in a one-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space and of a time-averaged Mel spectrogram.
 The total overlap is slightly smaller in the subspace of the CW-VAE latent space and the separation between the distribution peaks is larger.}
@@
      \end{subfigure}
-    \caption[Clustering of speaker gender in 2D linear subspace of a CW-VAE latent space.]{\subref{fig: cwvae latent z1 height} Clustering of speaker height for male speakers and \subref{fig: cwvae latent z0 age} speaker age for female speakers in an two-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space.}
+    \caption[Clustering of speaker gender in 2D linear subspace of a CW-VAE latent space.]{\subref{fig: cwvae latent z1 height} Clustering of speaker height for male speakers and \subref{fig: cwvae latent z0 age} speaker age for female speakers in a two-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space.}
     \label{fig: cwvae latent height}
@@
 \section{Distribution of phoneme duration in TIMIT}\label{app: timit phoneme distributions}
-In \cref{fig: timit validation set phoneme duration} we plot a boxplots of the duration of each phoneme in the TIMIT dataset.
+In \cref{fig: timit validation set phoneme duration} we plot a box plots of the duration of each phoneme in the TIMIT dataset.
 We do this globally as well as for a single speaker to show that phoneme duration can vary between individual speakers.
@@
     \end{subfigure}
-    \caption[Boxplots of duration of phoneme pronunciation in TIMIT.]{Boxplots of the duration of the pronunciation of phonemes in TIMIT for a specific speaker DRW0 in (\subref{fig: timit validation set phoneme duration DRW0}) and globally in (\subref{fig: timit validation set phoneme duration global}).
+    \caption[Box plots of duration of phoneme pronunciation in TIMIT.]{ Box plots of the duration of the pronunciation of phonemes in TIMIT for a specific speaker DRW0 in (\subref{fig: timit validation set phoneme duration DRW0}) and globally in (\subref{fig: timit validation set phoneme duration global}).
 Not all phonemes are pronounced by speaker DRW0 over the course of their 10 test set sentences and hence they are missing from the x-axis compared to the global durations.}
@@
 The samples are generated from the prior of Clockwork VAE, SRNN and VRNN and from a WaveNet by conditioning on pure zeros. All models are configured as those reported in \cref{tab: likelihoods timit}.
-Importantly, the samples are unconditional. Hence they are \emph{not} reconstructions inferred from a given input nor are they conditioned on any auxiliary data like text.
+Importantly, the samples are unconditional. Hence, they are \emph{not} reconstructions inferred from a given input nor are they conditioned on any auxiliary data like text.
 

### 2023-10-14 appendices/supplementary_benchmarking_workshop.tex
Fix many spelling and language errors

@@
 We replace the Gaussian output distribution with the DMoL.
-We use the best-performing version of the STCN reported in the original paper, namedly the ``STCN-dense" variant which conditions the observed variable on all five latent variables in the hierarchy.
+We use the best-performing version of the STCN reported in the original paper, namely the ``STCN-dense" variant which conditions the observed variable on all five latent variables in the hierarchy.
 For the ablation experiment, we remove the bottom four latent variables.
@@
 
-We constrain the standard deviation of the Gaussians used with our models to be at least $\sigma_\text{min} = 0.01$ in order to avoid it going to zero, the likelihood going to infinity and optimization becoming unstable.
+We constrain the standard deviation of the Gaussian used with our models to be at least $\sigma_\text{min} = 0.01$ in order to avoid it going to zero, the likelihood going to infinity and optimization becoming unstable.
 The minimum Gaussian standard deviation of \textcite{aksan_stcn_2019} is $\sigma_\text{min} = 0.001$.
@@
         \midrule
-        1 & Wavenet             & $D_c=96$       & \bf 10.96/10.89 & \bf 10.85/10.76 & \bf 11.12/11.01 & \bf 11.05/10.85 \\
+        1 & WaveNet             & $D_c=96$       & \bf 10.96/10.89 & \bf 10.85/10.76 & \bf 11.12/11.01 & \bf 11.05/10.85 \\
         1 & LSTM                & $D_d=256$      & 11.21/11.17 & 11.10/11.06 & 11.35/11.29 & 11.28/11.23 \\  % 15ua3hau/(lmssrk5k,248n2ylu)
         \midrule
-        64 & Wavenet            & $D_c=96$       & 13.61/13.24 & 13.58/13.21 & 13.61/13.22 & 13.60/13.21 \\  % 2ghkybtd/3iieasdt
+        64 & WaveNet            & $D_c=96$       & 13.61/13.24 & 13.58/13.21 & 13.61/13.22 & 13.60/13.21 \\  % 2ghkybtd/3iieasdt
         64 & LSTM               & $D_d=256$      & 13.56/13.25 & 13.52/13.24 & 13.55/13.23 & 13.56/13.25 \\  % 10fpovu3/2bu1izin
@@
 However, given the empirically good results of latent variable models for image generation, this seems to have posed only a minor problem in practice.
-We speculate that this is due to the high degree of similarity between neighbouring pixels in images.
+We speculate that this is due to the high degree of similarity between neighboring pixels in images.
 I.e.
@@
 
-In the audio domain, however, neighbouring waveform frames can take wildly different values, especially at low sample rates.
+In the audio domain, however, neighboring waveform frames can take wildly different values, especially at low sample rates.
 Furthermore, waveforms exhibit a natural symmetry between positive and negative amplitudes.
@@
 \section{Additional latent evaluation}\label{app: additional latent space clustering}
-Here we present additional qualitative assesessment of the learned latent representations selectively for the CW-VAE.
+Here we present additional qualitative assessment of the learned latent representations selectively for the CW-VAE.
 
@@
 Specifically, we infer the latent variables of all utterances by a single speaker from the TIMIT test set.
-We sample the latent sequence $100$ times to estimate the mean representation per time step.
+We sample the latent sequence $100$ times to estimate the mean representation per timestep.
 We then compute the average latent representation over the duration of each phoneme using aligned phoneme labels.
@@
      \end{subfigure}
-    \caption{Clustering of speaker gender in an one-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space and of a time-averaged mel spectrogram.
+    \caption{Clustering of speaker gender in a one-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space and of a time-averaged Mel spectrogram.
 The total overlap is slightly smaller in the subspace of the CW-VAE latent space and the separation between the distribution peaks is larger.}
@@
      \end{subfigure}
-    \caption{(a) Clustering of speaker height for male speakers and (b) speaker age for female speakers in an two-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space.}
+    \caption{(a) Clustering of speaker height for male speakers and (b) speaker age for female speakers in a two-dimensional linear subspace defined by a linear discriminant analysis of the CW-VAE latent space.}
     \label{fig: cwvae latent height}
@@
     \end{subfigure}
-    \caption{Boxplots of the duration of the pronunciation of phonemes in TIMIT for a specific speaker DRW0 in (\subref{fig: timit validation set phoneme duration DRW0}) and globally in (\subref{fig: timit validation set phoneme duration global}).
+    \caption{Box plots of the duration of the pronunciation of phonemes in TIMIT for a specific speaker DRW0 in (\subref{fig: timit validation set phoneme duration DRW0}) and globally in (\subref{fig: timit validation set phoneme duration global}).
 Not all phonemes are pronounced by speaker DRW0 over the course of their 10 test set sentences and hence they are missing from the x-axis compared to the global durations.}

### 2023-10-14 appendices/supplementary_hierarchical.tex
Fix many spelling and language errors

@@
 We do this because, the overall white nature of the images tends to make detecting them as OOD from FashionMNIST artificially easy.
-The inversion is done via the simple transformation $\xb_\text{inverted} = 255 - \xb_\text{original}$ since images are encoded as 8 bit unsigned integers.
+The inversion is done via the simple transformation $\xb_\text{inverted} = 255 - \xb_\text{original}$ since images are encoded as 8-bit unsigned integers.
 
@@
 Since the benefits and drawbacks of using batch normalization \parencite{ioffe_batch_2015} in hierarchical VAEs is still the matter of some debate \parencite{sonderby_ladder_2016, vahdat_nvae_2020, child_very_2021} we choose to use weight normalization \parencite{salimans_weight_2016} as in other work \parencite{maaloe_biva_2019} and initialize the model using the originally proposed data-dependent initialization.
-To have the stochastic layers initialize to standard Gaussian distributions (zero mean, unit variance), with this initialization, we select the activation function for the variance as a Softplus,
+To have the stochastic layers initialize to standard Gaussian distributions (zero mean, unit variance), with this initialization, we select the activation function for the variance as a softplus,
 \begin{equation*}
@@
 \begin{table}[t]
-    \caption[Selection of most important hyperparameters.]{Selection of most important hyperparameters and their setting.
+    \caption[Selection of most important hyperparameters.]{ Selection of most important hyperparameters and their setting.
 Convolutional kernels are square and latent dimensions are given without spatial dimensions which are given in the text.
@@
         \multicolumn{2}{l}{\textbf{HVAE}} \\
-        Latent dimensionality & 128-64-32 (natural) / 8-16-8 (grey) \\
+        Latent dimensionality & 128-64-32 (natural) / 8-16-8 (gray) \\
         Convolution kernel & 5-3-3 \\
         Stride & 2-2-1 \\
-        Warmup anneal period & 200 epochs \\
+        Warm up anneal period & 200 epochs \\
         \midrule

### 2023-10-14 chapters/paper_automated.tex
Fix many spelling and language errors

@@
 We show that several models underperform due to weak configurations, poorly sampled train-test splits, and insufficient evaluation.
-In previous work, the macro F1-score has been calculated sub-optimally, and our correction doubles it.
+In previous work, the macro F1-score has been calculated suboptimally, and our correction doubles it.
 We contribute a revised model comparison using stratified sampling and identical experimental setups, including hyperparameters and decision boundary tuning.
@@
     \includegraphics[width=0.9\textwidth]{paper_automated/mimiciv_code_freq.png}
-    \caption[The frequency of ICD-9 and ICD-10 codes in MIMIC-IV before pre-processing.]{The frequency of ICD-9 and ICD-10 codes in MIMIC-IV before pre-processing.
+    \caption[The frequency of ICD-9 and ICD-10 codes in MIMIC-IV before pre-processing.]{ The frequency of ICD-9 and ICD-10 codes in MIMIC-IV before pre-processing.
 As discussed in \cref{sec: splits}, we removed codes with fewer than ten instances (dashed line).}
@@
 \textcite{mullenbachExplainablePredictionMedical2018} released code for pre-processing the discharge summaries, generating the train-test split, and evaluating model performance on MIMIC-III which many subsequent papers have used \parencite{vuLabelAttentionModel2020, huangPLMICDAutomaticICD2022, liICDCodingClinical2020, baoMedicalCodePrediction2021, yuanCodeSynonymsMatter2022, kimReadAttendCode2021}.
-Pre-processing consisted of lowercasing all text and removing words that only contain out-of-alphabet characters. Predicting procedures and diagnosis codes were treated as a single task.
+Pre-processing consisted of lower-casing all text and removing words that only contain out-of-alphabet characters. Predicting procedures and diagnosis codes were treated as a single task.
 The dataset was split into training, validation, and test sets using random sampling, ensuring that no patient occurred in both the training and test set.
@@
 In LAAT, the learning rate was decreased by 90\% when the F1-score had not increased for five epochs.
-PLM-ICD used a schedule with linear warmup followed by linear decay.
+PLM-ICD used a schedule with linear warm up followed by linear decay.
 
@@
     \centering
-    \caption[Hyperparameters, maximum document lengths, and decision boundary tuning strategies used in the original works compared to improved settings.]{Hyperparameters, maximum document lengths, and decision boundary tuning strategies used in the original works compared to the optimal settings found in this paper (marked with *).
+    \caption[Hyperparameters, maximum document lengths, and decision boundary tuning strategies used in the original works compared to improved settings.]{ Hyperparameters, maximum document lengths, and decision boundary tuning strategies used in the original works compared to the optimal settings found in this paper (marked with *).
 LR is the learning rate scheduler.
@@
 \textcite{opitzMacroF1Macro2021} analyze macro F1 formulas common in multi-class and multi-label classification.
-They demonstrate that the above formulation is sub-optimal, as it rewards heavily biased classifiers in unbalanced datasets.
+They demonstrate that the above formulation is suboptimal, as it rewards heavily biased classifiers in unbalanced datasets.
 Therefore, as recommended by the authors, we calculated the macro F1-score as the arithmetic mean of the F1-score for each class.
@@
 We run experiments on the new MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9}, and \textit{ICD-10}.
-All models were trained for 20 epochs using a learning rate schedule with linear warmup for the first 2K updates followed by linear decay \parencite{huangPLMICDAutomaticICD2022}.
+All models were trained for 20 epochs using a learning rate schedule with linear warm up for the first 2K updates followed by linear decay \parencite{huangPLMICDAutomaticICD2022}.
 We found this schedule to speed up the training convergence of all the models.
@@
 All documents were truncated to a maximum of 4000 words.
-The hyperparameters, maximum document lengths, and decision boundary tuning strategy are summarised in \cref{tab:hyperparameters}.
+The hyperparameters, maximum document lengths, and decision boundary tuning strategy are summarized in \cref{tab:hyperparameters}.
 
@@
     \centering
-    \caption[Reproduced test set results compared with those from the original works.]{Reproduced test set results compared with those from the original works.
+    \caption[Reproduced test set results compared with those from the original works.]{ Reproduced test set results compared with those from the original works.
 Our reproduced results are indicated with *.
@@
     \centering
-    \caption[Results on the MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9} and MIMIC-IV  \textit{ICD-10} test sets presented as percentages.]{Results on the MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9} and MIMIC-IV  \textit{ICD-10} test sets presented as percentages.
+    \caption[Results on the MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9} and MIMIC-IV  \textit{ICD-10} test sets presented as percentages.]{ Results on the MIMIC-III \textit{clean}, MIMIC-IV \textit{ICD-9} and MIMIC-IV  \textit{ICD-10} test sets presented as percentages.
 Micro F1-scores rank the table in ascending order.
@@
     \centering
-    \caption[Ablation study on MIMIC-III \textit{clean}.]{Ablation study on MIMIC-III \textit{clean}. The numbers are the micro/macro F1-scores on the test set.}
+    \caption[Ablation study on MIMIC-III \textit{clean}.]{ Ablation study on MIMIC-III \textit{clean}. The numbers are the micro/macro F1-scores on the test set.}
     \label{tab:ablation_study}
@@
     \end{subfigure}%
-    \caption[Relationship between chosen threshold and F1-score of every reproduced model.]{The relationship between chosen threshold and F1-score of every reproduced model in \cref{tab:reproduced_results}.
+    \caption[Relationship between chosen threshold and F1-score of every reproduced model.]{ The relationship between chosen threshold and F1-score of every reproduced model in \cref{tab:reproduced_results}.
 The left figure shows the micro F1-score, and the right shows the macro F1-score.
@@
     \end{subfigure}
-    \caption[The relationship between the number of training examples and F1-score on MIMIC-IV \textit{ICD-9}.]{The relationship between the number of training examples and F1-score on MIMIC-IV \textit{ICD-9}.
+    \caption[The relationship between the number of training examples and F1-score on MIMIC-IV \textit{ICD-9}.]{  The relationship between the number of training examples and F1-score on MIMIC-IV \textit{ICD-9}.
 The left figure shows the F1 Micro score on the y-axis, while the right figure shows the F1 Macro score.}
@@
     \centering
-    \caption[Correlation between the F1-score and the logarithm of code frequency and document length on MIMIC-IV \textit{ICD-9}.]{Correlation between the F1-score and the logarithm of code frequency and document length on MIMIC-IV \textit{ICD-9}.
+    \caption[Correlation between the F1-score and the logarithm of code frequency and document length on MIMIC-IV \textit{ICD-9}.]{ Correlation between the F1-score and the logarithm of code frequency and document length on MIMIC-IV \textit{ICD-9}.
 As discussed in \cref{subsubsec:document-length}, we only considered document lengths between 1000 and 4000 words.
@@
 Using our corrected macro F1-score on the same split, PLM-ICD achieved a macro F1-score of 22.8\%.
-This large difference from previous state-of-the-art seems to indicate that all previous work on AMC used the sub-optimally calculated macro F1-score, including works not reproduced in this paper.
+This large difference from previous state-of-the-art seems to indicate that all previous work on AMC used the suboptimally calculated macro F1-score, including works not reproduced in this paper.
 Many studies use the macro F1-score to evaluate the ability of their models to predict rare codes \parencite{kimReadAttendCode2021, yuanCodeSynonymsMatter2022}.
@@
 and, in contrast to previous claims, long documents only have a negligible negative performance impact. 
-We release our source code, model parameters, and the new MIMIC-III \textit{clean} and MIMIC-IV \textit{ICD-9} and \textit{ICD-10} splits.$^\text{1}$
+We release our source code, model parameters, and the new MIMIC-III \textit{clean} and MIMIC-IV \textit{ICD-9} and \textit{ICD-10} splits.$^(\ref{footnote:source_code})$
 

### 2023-10-14 chapters/paper_benchmarking.tex
Fix many spelling and language errors

@@
 \end{itemize}
-At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models
-\footnote{\href{https://github.com/JakobHavtorn/benchmarking-lvms}{\url{github.com/JakobHavtorn/benchmarking-lvms}}}.
+At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models\footnote{\href{https://github.com/JakobHavtorn/benchmarking-lvms}{\url{github.com/JakobHavtorn/benchmarking-lvms}}}.
 % \footnote{\fontsize{8.5}{0}\selectfont\texttt{ \url{github.com/anonymous/available-later}}}.
@@
     \caption[Generative models for an LSTM, the VRNN, SRNN, STCN, and CW-VAE models.]{
-        Generative models for a single time step of a deterministic autoregressive LSTM, the VRNN and SRNN as well as the STCN and CW-VAE both with a single layer of latent variables. 
+        Generative models for a single timestep of a deterministic autoregressive LSTM, the VRNN and SRNN as well as the STCN and CW-VAE both with a single layer of latent variables. 
         Red arrows indicate purely deterministic paths from the output $\xb_t$ to previous input $\xb_{<t}$ without passing a stochastic node. 
@@
 This representation is then used to reconstruct the original input $\xb_{1:T}$.
-The latent variable is assumed to follow some prior distribution $p(\zb_t|\cdot)$ where the dot indicates that it may depend on latent and observed variables at previous time steps, $\zb_{<t}$ and $\xb_{<t}$ where $\zb_{<t} := \left( \zb_1, \zb_2, \dots, \zb_{t-1} \right)$.
+The latent variable is assumed to follow some prior distribution $p(\zb_t|\cdot)$ where the dot indicates that it may depend on latent and observed variables at previous timesteps, $\zb_{<t}$ and $\xb_{<t}$ where $\zb_{<t} := \left( \zb_1, \zb_2, \dots, \zb_{t-1} \right)$.
 
@@
 Inference (left) and generative (right) models for the CW-VAE with a hierarchy of $L=2$ latent variables, $s_1=1$ and $s_2=2$. 
-The models are unrolled over four consecutive time steps but note that the graph continues towards $t=0$ and $t=T$. 
+The models are unrolled over four consecutive timesteps but note that the graph continues towards $t=0$ and $t=T$. 
 Blue arrows indicate parameter sharing between inference and generative models. 
@@
 We define $\xb_{\scriptscriptstyle\mathcal{S}_t^l}:=\xb_{t:t+s_l}$ for compactness.
-The latent state transitions are densely connected and the decoder is also a convolutional network.
+The latent state transitions are densely connected, and the decoder is also a convolutional network.
 % \begin{align}
@@
 Architecture and training details are sketched below, while the full details are in appendices~\cref{app: model architectures} and \cref{app: training details} along with additional results for some alternative model configurations in appendix~\cref{app: additional likelihood results appendix}.
-We select model configurations that can be trained on GPUs with a maximum of 12GB of RAM and train all models until convergence on the validation set.
+We select model configurations that can be trained on GPUs with a maximum of \SI{12}{GB} of RAM and train all models until convergence on the validation set.
 We use a DMoL with 10 components for the output distribution of all models and model all datasets at their full bit depth of $\SI{16}{bits}$.
@@
 \begin{table}
-    \caption[Model likelihoods and phoneme error rate for TIMIT.]{We report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines (left) and phoneme error rate (PER) of different representations for phoneme recognition on TIMIT (right).}
+    \caption[Model likelihoods and phoneme error rate for TIMIT.]{ We report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines (left) and phoneme error rate (PER) of different representations for phoneme recognition on TIMIT (right).}
     \hfill
@@
         % 3.7h   &  Waveform        & stacked       &   84,1   \\  %           0,8405363165254586
-        3.7h   &  Spectrogram     & Mel           &   24.1   \\  % 2wjqhwf7  0.24149049930081432
-        3.7h   &  WaveNet         & $\hb^{(15)}$  &   27.7   \\  % 36ccyxf4  0.27743686764826847
-        3.7h   &  LSTM            & $\hb$         &   23.0   \\  % 3bh4vgrx  0.2301719174138357
-        3.7h   &  VRNN            & $\zb$         &   23.2   \\  % 3stwdgas  0.23229415151764415
-        3.7h   &  SRNN            & $\zb$         &   26.0   \\  % 2bkvv5i0  0.2596528748868964
-        3.7h   &  CW-VAE          & $\zb^{(1)}$   &   36.4   \\  % 3m4f3f7z
-        3.7h   &  STCN            & $\zb^{(2)}$   &   \textbf{21.9}   \\  % rkukk5bn  0.2189849469441474
+        \SI{3.7}{h}   &  Spectrogram     & Mel           &   24.1   \\  % 2wjqhwf7  0.24149049930081432
+        \SI{3.7}{h}   &  WaveNet         & $\hb^{(15)}$  &   27.7   \\  % 36ccyxf4  0.27743686764826847
+        \SI{3.7}{h}   &  LSTM            & $\hb$         &   23.0   \\  % 3bh4vgrx  0.2301719174138357
+        \SI{3.7}{h}   &  VRNN            & $\zb$         &   23.2   \\  % 3stwdgas  0.23229415151764415
+        \SI{3.7}{h}   &  SRNN            & $\zb$         &   26.0   \\  % 2bkvv5i0  0.2596528748868964
+        \SI{3.7}{h}   &  CW-VAE          & $\zb^{(1)}$   &   36.4   \\  % 3m4f3f7z
+        \SI{3.7}{h}   &  STCN            & $\zb^{(2)}$   &   \textbf{21.9}   \\  % rkukk5bn  0.2189849469441474
         \midrule
         % 1h     &  Waveform        & stacked       &   59.3   \\  %           0.5926626634860575
-        1.0h   &  Spectrogram     & Mel           &   30.8   \\  % 3kmfpljj  0.3083655507115242
-        1.0h   &  WaveNet         & $\hb^{(15)}$  &   34.7   \\  % 36ccyxf4  0.34743769021962656
-        1.0h   &  LSTM            & $\hb$         &   30.1   \\  % 3bh4vgrx  0.30137369416796905
-        1.0h   &  VRNN            & $\zb$         &   30.4   \\  % 3stwdgas  0.31284033889939955
-        1.0h   &  SRNN            & $\zb$         &   31.7   \\  % 2bkvv5i0  0.3167064242823065
-        1.0h   &  CW-VAE          & $\zb^{(1)}$   &   40.0   \\  % 3m4f3f7z
-        1.0h   &  STCN            & $\zb^{(2)}$   &   \textbf{26.7}   \\  % rkukk5bn  0.2671547256724521
+        \SI{1.0}{h}   &  Spectrogram     & Mel           &   30.8   \\  % 3kmfpljj  0.3083655507115242
+        \SI{1.0}{h}   &  WaveNet         & $\hb^{(15)}$  &   34.7   \\  % 36ccyxf4  0.34743769021962656
+        \SI{1.0}{h}   &  LSTM            & $\hb$         &   30.1   \\  % 3bh4vgrx  0.30137369416796905
+        \SI{1.0}{h}   &  VRNN            & $\zb$         &   30.4   \\  % 3stwdgas  0.31284033889939955
+        \SI{1.0}{h}   &  SRNN            & $\zb$         &   31.7   \\  % 2bkvv5i0  0.3167064242823065
+        \SI{1.0}{h}   &  CW-VAE          & $\zb^{(1)}$   &   40.0   \\  % 3m4f3f7z
+        \SI{1.0}{h}   &  STCN            & $\zb^{(2)}$   &   \textbf{26.7}   \\  % rkukk5bn  0.2671547256724521
         \midrule
-        10m    &  Waveform        & stacked       &   85.6   \\  %           0.8564448465904417
-        10m    &  Spectrogram     & Mel           &   47.1   \\  % 1sep2uq8  0.4706753310849716
-        10m    &  WaveNet         & $\hb^{(15)}$  &   52.8   \\  % 36ccyxf4  0.5276137204902526
-        10m    &  LSTM            & $\hb$         &   46.1   \\  % 3bh4vgrx  0.46073866907954264
-        10m    &  VRNN            & $\zb$         &   44.6   \\  % 3stwdgas  0.4601299662745743
-        10m    &  SRNN            & $\zb$         &   47.3   \\  % 2bkvv5i0  0.47317594801349017
-        10m    &  CW-VAE          & $\zb^{(1)}$   &   54.9   \\  % 3m4f3f7z
-        10m    &  STCN            & $\zb^{(2)}$   &   \textbf{42.7}   \\  % rkukk5bn  0,4270625976803488
+        \SI{10}{m}    &  Waveform        & stacked       &   85.6   \\  %           0.8564448465904417
+        \SI{10}{m}    &  Spectrogram     & Mel           &   47.1   \\  % 1sep2uq8  0.4706753310849716
+        \SI{10}{m}    &  WaveNet         & $\hb^{(15)}$  &   52.8   \\  % 36ccyxf4  0.5276137204902526
+        \SI{10}{m}    &  LSTM            & $\hb$         &   46.1   \\  % 3bh4vgrx  0.46073866907954264
+        \SI{10}{m}    &  VRNN            & $\zb$         &   44.6   \\  % 3stwdgas  0.4601299662745743
+        \SI{10}{m}    &  SRNN            & $\zb$         &   47.3   \\  % 2bkvv5i0  0.47317594801349017
+        \SI{10}{m}    &  CW-VAE          & $\zb^{(1)}$   &   54.9   \\  % 3m4f3f7z
+        \SI{10}{m}    &  STCN            & $\zb^{(2)}$   &   \textbf{42.7}   \\  % rkukk5bn  0,4270625976803488
 \bottomrule
@@
         \midrule
-        1 & Wavenet             & $D_c=96$       & \bfseries 10.96/10.89 & \bfseries 10.85/10.76 & \bfseries 11.12/11.01 & \bfseries 11.05/10.85 \\
+        1 & WaveNet             & $D_c=96$       & \bfseries 10.96/10.89 & \bfseries 10.85/10.76 & \bfseries 11.12/11.01 & \bfseries 11.05/10.85 \\
         1 & LSTM                & $D_d=256$      & 11.21/11.17 & 11.10/11.06 & 11.35/11.29 & 11.28/11.23 \\  % 15ua3hau/(lmssrk5k,248n2ylu)
         \midrule
-        64 & Wavenet            & $D_c=96$       & 13.61/13.24 & 13.58/13.21 & 13.61/13.22 & 13.60/13.21 \\  % 2ghkybtd/3iieasdt
+        64 & WaveNet            & $D_c=96$       & 13.61/13.24 & 13.58/13.21 & 13.61/13.22 & 13.60/13.21 \\  % 2ghkybtd/3iieasdt
         64 & LSTM               & $D_d=256$      & 13.56/13.25 & 13.52/13.24 & 13.55/13.23 & 13.56/13.25 \\  % 10fpovu3/2bu1izin
@@
 
-At $s=64$, Wavenet and the LSTM yield significantly worse likelihoods than all LVMs separated by $\sim\SI{1}{bit}$.
+At $s=64$, WaveNet and the LSTM yield significantly worse likelihoods than all LVMs separated by $\sim\SI{1}{bit}$.
 The CW-VAE outperforms the VRNN and SRNN when configured with a hierarchy of latent variables. 
@@
 The STCN yields the best results at both $s=64$ and $s=256$. 
-For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables, such as WaveNet and the LSTM, to the ability of LVMs to model intra-step correlations \parencite{lai_re-examination_2019}.
+For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables, such as WaveNet and the LSTM, to the ability of LVMs to model intrastep correlations \parencite{lai_re-examination_2019}.
 
@@
 % \parencite{??}.
-Basic Correlates of the Auditory Stimulus, https://psycnet.apa.org/record/1951-07758-017, modern mel scale https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=1170013&casa_token=UIyQaj3tBRAAAAAA:GtttZedngBkaxVumjnXctc5zS50fptexPMiWhr-7gDrmIVq3JqJPnGCWYedoymb0X5JTGElmHf8&tag=1,
+Basic Correlates of the Auditory Stimulus, https://psycnet.apa.org/record/1951-07758-017, modern Mel scale https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=1170013&casa_token=UIyQaj3tBRAAAAAA:GtttZedngBkaxVumjnXctc5zS50fptexPMiWhr-7gDrmIVq3JqJPnGCWYedoymb0X5JTGElmHf8&tag=1,
 To assess the usefulness of the representations learned by the benchmarked models, we compare them to the highly useful Mel spectrogram on the task of phoneme recognition. 
@@
 For the hierarchical LVMs and WaveNet, we run the experiment using each possible representation and report only the best one here.
-We compare the learned representations to a log Mel spectrogram with 80 filterbanks, hop length 64 and window size 128. 
+We compare the learned representations to a log Mel spectrogram with 80 filter banks, hop length 64 and window size 128. 
 We also compared to using raw PCM in vectors of 64 elements standardized to $[-1,1]$ but found that the ASR did not reliably converge at all which highlights the importance of input representation. 
@@
 We pre"=train all unsupervised models at $s=64$ on the full TIMIT training dataset excluding the validation data (3.7h) as in \cref{tab: likelihoods timit} (left). 
-We then train the ASR model on all 3.7h as well as 1h and 10m subsets. We report results on the test set in terms of phoneme error rates (PER) in \cref{tab: phoneme recognition (PER)} (right).
+We then train the ASR model on all \SI{3.7}{h} as well as \SI{1}{h} and \SI{10}{m} subsets.
+We report results on the test set in terms of phoneme error rates (PER) in \cref{tab: phoneme recognition (PER)} (right).
 
@@
 \paragraph{Qualitative}
-We qualitatively asses the learned latent representations for the CW-VAE. 
+We qualitatively assess the learned latent representations for the CW-VAE. 
 We infer the latent variables of all utterances by a single speaker from the TIMIT test set.
-We sample the latent sequence $100$ times to estimate the mean representation per time step.
+We sample the latent sequence $100$ times to estimate the mean representation per timestep.
 We then compute the average latent representation over the duration of each phoneme using aligned phoneme labels.

### 2023-10-14 chapters/paper_benchmarking_workshop.tex
Some work on discussion
Updates to discussion
Fix many spelling and language errors

@@
 \caption{
-Generative models of an LSTM, VRNN, SRNN, STCN and CW-VAE for a single time step. The STCN and CW-VAE are illustrated with a single latent variable. 
+Generative models of an LSTM, VRNN, SRNN, STCN and CW-VAE for a single timestep. The STCN and CW-VAE are illustrated with a single latent variable. 
 Red arrows indicate purely deterministic paths from the output $\vx_t$ to previous input $\vx_{<t}$ without passing a stochastic node. 
@@
 This makes them useful beyond generation for e.g.
-semi"=supervised learning \parencite{kingma_semi"=supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.
+semi"=supervised learning \parencite{kingma_semi-supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.
 In this paper, we focus on the VRNN \parencite{chung_recurrent_2015}, SRNN \parencite{fraccaro_sequential_2016} and STCN \parencite{aksan_stcn_2019} since they are similar to the original VAE framework.
@@
 The models first encode $\vx_{1:T}$ to a temporal posterior distribution $q(\vz|\vx)$ and sample a latent representation $\vz_{1:T}$ where $\vz_t\in\mathbb{R}^{D_z}$.
-This is then used to reconstruct the input while the posterior distribution is regularized to be close to a prior distribution $p(\vz_t|\cdot)$ where the dot indicates that it may depend on latent and observed variables at previous time steps, $\vz_{<t}$ and $\vx_{<t}$ where $\vz_{<t} := \left( \vz_1, \vz_2, \dots, \vz_{t-1} \right)$.
+This is then used to reconstruct the input while the posterior distribution is regularized to be close to a prior distribution $p(\vz_t|\cdot)$ where the dot indicates that it may depend on latent and observed variables at previous timesteps, $\vz_{<t}$ and $\vx_{<t}$ where $\vz_{<t} := \left( \vz_1, \vz_2, \dots, \vz_{t-1} \right)$.
 The models are trained to maximize a lower bound $\mathcal{L}(\thetab,\phib;\vx)$ on the marginal likelihood $\log p_\thetab(\vx)$,
 \begin{equation}
-    \log p_\thetab(\vx) = \log \int p_\thetab(\vx,\vz) \, \mathrm{d}\vz  \geq \int q_\phib(\vz|\vx) \log \frac{p_\thetab(\vx|\vz)p(\vz)}{q_\phib(\vz|\vx)}  \, \mathrm{d}\vz \triangleq \mathcal{L}(\thetab,\phib;\vx)  \enspace , \label{eq: elbo}
+    \log p_\thetab(\vx) = \log \int p_\thetab(\vx,\vz) \, \mathrm{d}\vz  \geq \int q_\phib(\vz|\vx) \log \frac{p_\thetab(\vx|\vz)p(\vz)}{q_\phib(\vz|\vx)}  \, \mathrm{d}\vz \equiv \mathcal{L}(\thetab,\phib;\vx)  \enspace , \label{eq: elbo}
 \end{equation}
@@
 It is clear that deterministic autoregressive models are superior to LVMs at $s=1$, but inferior to them at $s=64$ and $256$.
-For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables to the ability of LVMs to model intra-step correlations \parencite{lai_re-examination_2019}.
+For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables to the ability of LVMs to model intrastep correlations \parencite{lai_re-examination_2019}.
 It should be noted that at $s=1$, the STCN was numerically unstable and that the CW-VAE was computationally infeasible to train.
@@
 To assess the usefulness of the representations, we train a model to recognize phonemes on the TIMIT data.
-We compare the phoneme error rate (PER) of the model when using input representations obtained from different unsupervised models, trained on the full TIMIT dataset (3.7h) at $s=64$, as in \cref{tab: likelihoods timit}, and when using a Mel spectrogram (80 filterbanks, hop length 64, window size 128).
+We compare the phoneme error rate (PER) of the model when using input representations obtained from different unsupervised models, trained on the full TIMIT dataset (3.7h) at $s=64$, as in \cref{tab: likelihoods timit}, and when using a Mel spectrogram (80 filter banks, hop length 64, window size 128).
 The recognition model is a three-layered bidirectional LSTM with 256 hidden units.

### 2023-10-14 chapters/paper_hierarchical.tex
Fix many spelling and language errors

--- 2023-10-04 chapters/paper_hierarchical.tex
@@
 where $k\in\{0,1,\dots,L\}$ (see \cref{sec_hierarchical:derivation_L_geq_k} for the derivation).
-We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq_hierarchical:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
+We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq_hierarchical:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$, although this need not hold in general.
 The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\zb_1,\dots,\zb_l \sim p_\thetab(\zb_{\leq k}|\zb_{>k})$ and only the $L-k$ highest from the approximate posterior $\zb_{k+1},\dots,\zb_L \sim q_\phib(\zb_{>k}|\xb)$.
@@
     \end{subfigure}
-    \caption[Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood and $\mathcal{L}^{>2}$ bound.]{%
-    Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood \subref{fig_hierarchical:FMNIST-elbo-k0}, the $\mathcal{L}^{>2}$ bound \subref{fig_hierarchical:FMNIST-elbo-k2} and the $LLR^{>1}$ score \subref{fig_hierarchical:FMNIST-llr}.
+    \caption[Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood and $\mathcal{L}^{>2}$ bound.]{ Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood \subref{fig_hierarchical:FMNIST-elbo-k0}, the $\mathcal{L}^{>2}$ bound \subref{fig_hierarchical:FMNIST-elbo-k2} and the $LLR^{>1}$ score \subref{fig_hierarchical:FMNIST-llr}.
 All densities are computed using the HVAE model.
 For the regular likelihood MNIST is very clearly more likely on average than the FashionMNIST test data while with the $\mathcal{L}^{>2}$ bound separation is better but significant overlap remains.
-    The $LLR^{>1}$ provides a high degree of separation. Likelihoods are reported in units of the natural log of the number of bits per dimension.
+The $LLR^{>1}$ provides a high degree of separation.
+Likelihoods are reported in units of the natural log of the number of bits per dimension.
     }
@@
 \subsection{Likelihood-ratio-based OOD detection}
+
+We now move to the likelihood ratio-based score.
+We find that $LLR^{>k}$ separates the OOD MNIST data from in-distribution FashionMNIST to a higher degree than the likelihood estimates as can be seen by the empirical densities of the score in \cref{fig_hierarchical:FMNIST-llr}.
+We note that the likelihood ratio between the ELBO and the $\mathcal{L}^{>k}$ bound provides the highest degree of separation of MNIST and FashionMNIST as measured by the AUROC$\uparrow$ for $k=1$ smaller than $L$.
+This is not surprising since the value of $k$ that provides the maximal separation to the reference in-distribution dataset need not be the one for which $\mathcal{LLR}^{>k}$ is overall maximal for the OOD dataset.
+We also visualize the ROC curves resulting from using the $LLR^{>k}$ score for OOD detection on both FashionMNIST/MNIST and CIFAR10/SVHN and compare it to the ROC curves resulting from the different $\mathcal{L}^{>k}$ bounds in \cref{fig_hierarchical:FMNIST-roc-llr and CIFAR10-roc-llr}, respectively.
+On both datasets we see significantly better discriminatory performance when using the $LLR^{>k}$ score.
+
+\cref{tab_hierarchical:rocauc-ood} shows that BIVA improves upon the HVAE model for OOD detection on CIFAR while \cref{tab_hierarchical:bits-per-dim-ood} shows that the BIVA model also improves upon the HVAE in terms of likelihood.
+We hypothesize that models larger than our implementation of BIVA, with better likelihood scores may perform even better \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
 

### 2023-10-14 chapters/paper_modelagnostic.tex
Fix many spelling and language errors

--- 2023-10-10 chapters/paper_modelagnostic.tex
@@
 We will try to look in which situations one of the test fails and the other works and vice versa.
-Both examples assume that the in-distribution data follows a $\mathcal{N}(0,I_D)$ distribution, and that the correct model has been learned by fitting $(\mathcal{N}(\theta,I_D))_{\theta \in \mathbb{R}^D}$ via maximum-likelihood.
+Both examples assume that the in-distribution data follows an $\mathcal{N}(0,I_D)$ distribution, and that the correct model has been learned by fitting $(\mathcal{N}(\theta,I_D))_{\theta \in \mathbb{R}^D}$ via maximum-likelihood.
 Even in this simple setting with no model misspecification, we will see that the two statistics that we consider may have very different strengths.
@@
     \includegraphics[scale=0.6]{graphics/paper_modelagnostic/fashion_hvae_type1_correct_font.pdf}
-    \caption[Type I and Type II errors versus the significance level $\alpha$ on the combination values]{Type I (probability of an inlier to be classified as outlier) and Type II (probability of an outlier to be considered as inlier) errors versus the significance level $\alpha$ on the combination values.
+    \caption[Type I and Type II errors versus the significance level $\alpha$ on the combination values.]{ Type I (probability of an inlier to be classified as outlier) and Type II (probability of an outlier to be considered as inlier) errors versus the significance level $\alpha$ on the combination values.
 By using Benjamini-Hochberg correction, we get that the Type I error stays below identity line.}

### 2023-10-14 chapters/paper_review/03_approaches.tex
Fix many spelling and language errors

@@
    % \includegraphics[width=0.98\textwidth]{paper_review/model_timeline_wo_dccae.pdf}
-	 \caption[Selection of self"=supervised models for speech.]{A selection of models listed according to first publication date
+	 \caption[Selection of self"=supervised models for speech.]{ A selection of models listed according to first publication date
 	 on arXiv or conference submission date when this clearly precedes the
@@
 	 representations from speech only. Some models use a mixture of generative
-	 and contrastive tasks. For instance, PASE and PASE+ use a multi-task setup,
+	 and contrastive tasks. For instance, PASE and PASE+ use a multitask setup,
 	 but find that generative tasks are the most important for downstream
@@
 hidden units available than input features. This forces the model to discard
-low-level details and disccourages the learning of trivial solutions. Other models add
+low-level details and discourages the learning of trivial solutions. Other models add
 regularization to the latent space to further improve the quality of the
@@
 This operation is non-differentiable and the \edit{gradients of the loss with respect to the encoder parameters} must be obtained by approximation.
-In the VQ-VAE this is done using the straight-through estimator~\parencite{bengio_estimating_2013}, i.e., the gradients \edit{with respect to the} encoder output are taken to be equal to those \edit{with respect to the} decoder input \ci{(i.e., the quatization step is ignored)}.
+In the VQ-VAE this is done using the straight-through estimator~\parencite{bengio_estimating_2013}, i.e., the gradients \edit{with respect to the} encoder output are taken to be equal to those \edit{with respect to the} decoder input \ci{(i.e., the quantization step is ignored)}.
 Given a learned codebook $A\in\mathbb{R}^{K \times D}$, where $K$ is the codebook size and $D$ is the dimensionality of each codebook vector $a_k$, the quantized representation $q_t$ of $h_t$ is obtained as
@@
 
-The wav2vec 2.0 approach was the first to reach single-digit \edit{word error rate (WER) on LibriSpeech using only the low-resource Libri-light subsets for fine"=tuning a pre"=trained model (see \cref{sec:datasets}).
+The wav2vec 2.0 approach was the first to reach single-digit \edit{word error rate (WER) on LibriSpeech using only the low-resource LibriLight subsets for fine"=tuning a pre"=trained model (see \cref{sec:datasets}).
 It has subsequently} inspired many follow-up studies.
@@
 The bias is added prior to the softmax normalization of the attention weights.
-For the attention weight at $i,j$, the bias is computed based on the input to the Transformer layer at the current time step $i$ and also incorporates a relative positional embedding for $i-j$.
+For the attention weight at $i,j$, the bias is computed based on the input to the Transformer layer at the current timestep $i$ and also incorporates a relative positional embedding for $i-j$.
 The authors find that this extension improves performance on phoneme and speech recognition tasks.}
@@
 Multiple modalities are useful in many settings, where each modality provides information that is complementary to other modalities.
-Multi-modal work includes supervised settings, such as audio-visual ASR~\parencite{petajan_automatic_1984,potamianos_recent_2003} and person identification~\parencite{aleksic_audiovisual_2006} which have been studied for decades.
+Multi-modal work includes supervised settings, such as audiovisual ASR~\parencite{petajan_automatic_1984,potamianos_recent_2003} and person identification~\parencite{aleksic_audiovisual_2006} which have been studied for decades.
 In this section, we focus only on unsupervised representation learning from multi-modal data.
@@
     \item Multi-view contrastive losses~\parencite{hermann_multilingual_2013,huang_learning_2013},
-    \item More recently, audio-visual extensions of masked prediction methods~\parencite{shi_learning_2022,shi_robust_2022}, specifically Audio-Visual HuBERT (AV-HuBERT)~\parencite{shi_learning_2022}.
+    \item More recently, audiovisual extensions of masked prediction methods~\parencite{shi_learning_2022,shi_robust_2022}, specifically Audiovisual HuBERT (AV-HuBERT)~\parencite{shi_learning_2022}.
 \end{itemize}
@@
 The SLAM approach~\parencite{bapna_slam_2021} is an example where speech and text are first represented using two separate pre"=trained encoders followed by a multi-modal encoder to build the joint representations.
-The entire model is trained using a multi-task loss including two supervised and two self"=supervised tasks.
+The entire model is trained using a multitask loss including two supervised and two self"=supervised tasks.
 

### 2023-10-14 chapters/paper_review/04_benchmarking.tex
Fix many spelling and language errors

@@
 in the literature. These datasets are usually large but with limited or no
-labels. Libri-light (LL)~\parencite{kahn_libri-light_2020}, one of these datasets, is
+labels. LibriLight (LL)~\parencite{kahn_libri-light_2020}, one of these datasets, is
 derived from audiobooks that are part of the
@@
 of audio of varying quality, different languages, and sometimes multiple sound
-sources. AVSpeech~\parencite{ephrat_looking_2018} is another large-scale audio-visual
+sources. AVSpeech~\parencite{ephrat_looking_2018} is another large-scale audiovisual
 dataset used in SSL research, comprising 4.7k hours of clips from a wide
@@
 collected corpora for pre"=training Chinese speech representations. Didi Dictation
-and Didi Callcenter~\parencite{jiang_improving_2019, jiang_further_2021} are two
+and Didi Call Center~\parencite{jiang_improving_2019, jiang_further_2021} are two
 internal datasets containing respectively 10k hours of read speech collected
@@
 European languages for a total of 50k hours of audio. VP comprises a total of 400k
-hours of parliamentary speech from the European parliament in 23 European
+hours of parliamentary speech from the European Parliament in 23 European
 languages.
@@
 and contains six emotions: anger, sadness, joy, fear, disgust, and
-surprise, plus neutral. MOSEI is comprised of sentence-level sentiment
+surprise, plus neutral. MOSEI is composed of sentence-level sentiment
 annotations of 65 hours of YouTube videos
@@
 data with 109 English speakers of various accents, each reading out about 400
-sentences sampled from newspapers. VoxCeleb1 is an audio-visual dataset
+sentences sampled from newspapers. VoxCeleb1 is an audiovisual dataset
 comprised of short YouTube clips containing human speech. It consists of 1251
@@
 for XLS-R~\parencite{babu_xlsr_2021}. We expect this trend to continue with the growth
-in available computing power. Most studies focuses on learning representations
+in available computing power. Most studies focus on learning representations
 for English, whereas Chinese~\parencite{jiang_improving_2019, jiang_further_2021} and
-multilinguality~\parencite{kawakami_learning_2020, babu_xlsr_2021} are also gaining
+multilingualism~\parencite{kawakami_learning_2020, babu_xlsr_2021} are also gaining
 attention. Compared to pre"=training, datasets used for fine"=tuning are more
@@
 \textit{X-Large}, or \textit{XX-Large} variants specified in the original
-publication. We also compare with baselines including the log mel filterbank (fBANK) and semi"=supervised, self-training
+publication. We also compare with baselines including the log Mel filter bank (fBANK) and semi"=supervised, self-training
 approaches (iterative pseudo labeling (IPL)~\parencite{xu_iterative_2020}, 
@@
 techniques including self-training. Furthermore, we illustrate in the figure
-the performance of a baseline system \parencite{yang_superb_2021} based on log mel filterbank (fBANK), which is one of the most commonly used features designed by domain experts.
+the performance of a baseline system \parencite{yang_superb_2021} based on log Mel filter bank (fBANK), which is one of the most commonly used features designed by domain experts.
 As observed in the figure, most SSL techniques outperform fBANK
@@
     ASR (WER $\downarrow$) & LS test-clean/other & 2.1/4.0 \parencite{xu_iterative_2020} & 1.4/2.6 \parencite{zhang_pushing_2020} \\ \hline
-    IC (Acc $\uparrow$) & FSC & 98.8 \parencite{lugosch_speech_2019} & 99.3\parencite{chen_unispeechsat_2021} \\ \hline
-    SID (Acc $\uparrow$) & VoxCeleb1 & 94.8 \parencite{hajibabaei_unified_2018} & 95.5 \parencite{chen_wavlm_2021} \\ \hline
+    IC (Acc. $\uparrow$) & FSC & 98.8 \parencite{lugosch_speech_2019} & 99.3\parencite{chen_unispeechsat_2021} \\ \hline
+    SID (Acc. $\uparrow$) & VoxCeleb1 & 94.8 \parencite{hajibabaei_unified_2018} & 95.5 \parencite{chen_wavlm_2021} \\ \hline
     ASV (EER $\downarrow$) & VoxCeleb1 & 3.1 \parencite{hajavi_siamese_2021} & 2.4 \parencite{wang_finetuned_2021} \\ \hline

### 2023-10-14 chapters/paper_review/04b_analysis.tex
Fix many spelling and language errors

@@
 can be used to judge the similarity of embeddings at different layers
-with each other, with standard acoustic representations such as mel
-filterbank features, or word embeddings derived from text. The second
+with each other, with standard acoustic representations such as Mel
+filter bank features, or word embeddings derived from text. The second
 method clusters continuous representation vectors and computes the
@@
 sizes as well as models fine"=tuned for ASR. They found that pre"=trained
-models show an autoencoder-style behavior, with early layers showing
+models show an autoencoder-like behavior, with early layers showing
 strong similarity with input features, intermediate layers diverging
-more, and final layers reverting back to higher similarity with input
+more, and final layers reverting to higher similarity with input
 features and early layers. Generally, the earlier layers in wav2vec~2.0 
@@
 phonetic class information, followed by word meaning information,
-before reverting back to encoding phonetic/acoustic information. Thus,
+before reverting to encoding phonetic/acoustic information. Thus,
 extracting representations from the last layers for tasks that require
@@
 last layers (especially the final two layers), breaking the
-autoencoder-style behavior and focusing the information encoded in the
+autoencoder-like behavior and focusing the information encoded in the
 last layers on orthographic-phonetic and word information.
@@
 
-A similar insight was obtained in \parencite{zhou_comparison_2020}, which compared vq-vae and
+A similar insight was obtained in \parencite{zhou_comparison_2020}, which compared VQ-VAE and
 vq-wav2vec with respect to their ability to discover phonetic units.
-The vq-vae model extracts continuous features from the audio signal; a
+The VQ-VAE model extracts continuous features from the audio signal; a
 quantizer then  maps them into a discrete space, and a decoder is trained to
@@
 (see \cref{sec:zero} for more details)), and it was found that the predictive vq-wav2vec
-model fared better than the autoencoder-like vq-vae model, most likely due to
+model fared better than the autoencoder-like VQ-VAE model, most likely due to
 its superior ability to model temporal dynamics.
@@
 distributional shifts here may stem from both the acoustics (microphone, room
-reverberation) as well as
-lexical effects related to topic and style, as well as differences in speaker
+reverberation),
+lexical effects related to topic and style, and differences in speaker
 characteristics such as accent. 

### 2023-10-14 chapters/paper_review/05_zero.tex
Fix many spelling and language errors

@@
 % What are we going to do % how to match
-%In the self"=supervised learning framework, speech representations can be learned and used in the downstream tasks, but some labeled data for finetuning is needed.
+%In the self"=supervised learning framework, speech representations can be learned and used in the downstream tasks, but some labeled data for fine-tuning is needed.
 
@@
 	\midrule
-	\parencite{chung_unsupervised_2018} & Speech2vec~\parencite{chung_speech2vec_2018} & BES-GMM~\parencite{kamper_segmental_2017} &  Word2Vec & \edit{Adversarial Training~\parencite{conneau_word_2018}} & Self-training & SWC (Acc): 10.9\%  \\ %Submitted on 18 May 2018
+	\parencite{chung_unsupervised_2018} & Speech2vec~\parencite{chung_speech2vec_2018} & BES-GMM~\parencite{kamper_segmental_2017} &  Word2Vec & \edit{Adversarial Training~\parencite{conneau_word_2018}} & Self-training & SWC (Acc.): 10.9\%  \\ %Submitted on 18 May 2018
 	%https://arxiv.org/pdf/1805.07467.pdf %The text embeddings were obtained by training Word2Vec on the transcriptions using the fastText implementation without subword information [3].
@@
 
-	\textcolor{black}{\parencite{klejch_deciphering_2022}} & \textcolor{black}{\tabincell{l}{Universal Phone \\ Recogniser}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Decipherment~\parencite{ravi_deciphering_2011}} & \textcolor{black}{Self-training} & \textcolor{black}{\tabincell{l}{GlobalPhone: 32.5\% to just 1.9\% \\ worse than supervised models} }  \\ %ubmitted on 12 Nov 2021 %m 32.5% to just 1.9% absolute worse than the equivalent fully supervised models trained on the same data
+	\textcolor{black}{\parencite{klejch_deciphering_2022}} & \textcolor{black}{\tabincell{l}{Universal Phone \\ Recognizer}} & - & \textcolor{black}{Grapheme} & \textcolor{black}{Deciphering~\parencite{ravi_deciphering_2011}} & \textcolor{black}{Self-training} & \textcolor{black}{\tabincell{l}{GlobalPhone: 32.5\% to just 1.9\% \\ worse than supervised models} }  \\ %ubmitted on 12 Nov 2021 %m 32.5% to just 1.9% absolute worse than the equivalent fully supervised models trained on the same data
 	\midrule
@@
 \textcolor{black}{
-Other ways to map speech and text include via segmental empirical output distribution matching (segmental empirical-ODM)~\parencite{yeh_unsupervised_2019} and decipherment algorithm~\parencite{klejch_deciphering_2022}.}
+Other ways to map speech and text include segmental empirical output distribution matching (segmental empirical-ODM)~\parencite{yeh_unsupervised_2019} and the deciphering algorithm~\parencite{klejch_deciphering_2022}.}
 
@@
 
-%Hung-yi Lee: I comment this paragraph first.
-\begin{comment}
-%Hung-yi Lee: This paragraph is not finished yet.
-\TNS{if we are limited on space,we can consider removing some of the detail of the various algorithms}
-\TNS{split into 2 sentences, also have a connection to the previous para}
-\parencite{liu_completely_2018} is the first work to successfully achieve unsupervised
-phoneme recognition by first clustering phoneme-level
-embeddings~\parencite{chung_audio_2016,wang_segmental_2018} into a set of
-tokens and then using GAN to learn the mapping relationship between tokens and
-phonemes. 
-However, in \parencite{liu_completely_2018}, oracle phoneme segmentation is still
-required to get reasonable performance. 
-\parencite{chen_completely_2019,yeh_unsupervised_2019} uses gate activation signals
-(GAS)~\parencite{wang_gate_2017} to get the phoneme-level segmentation, which is
-unsupervised. 
-%However, the loss term in~\parencite{yeh_unsupervised_2019} includes an empirical average over all segments inside a logarithmic function, which can be biased if we sample this empirical average by a mini-batch average.
-\parencite{liu2017unsupervised} therefore use an extremely large batch size during training to reduce the biasing problem.
-%These problems are handled by another work~\parencite{chen_completely_2019}.
-%\parencite{chen_completely_2019} removes the quantization step.
-%Compared to~\parencite{yeh_unsupervised_2019}, the discriminator in~\parencite{chen_completely_2019} considers all possible information from the text dataset, not limited to the n-gram local statistics.
-%Besides, \parencite{chen_completely_2019} works well with a reasonable batch size, which makes it feasible when the computational resource is limited.
-Wav2vec-U~\parencite{baevski_unsupervised_2021} selects the input feature from
-different layers of wave2vec 2.0~\parencite{baevski_wav2vec_2020}, a self"=supervised
-model.
-The selecting criterion is the PER by training a linear model in a supervised manner. 
-%Should we mention this?
-Wav2vec-U uses $k$-means to cluster the selected features, and the boundaries
-are drawn whenever the clustered index changes.
-%Wave2vec-U tends to generate more segments compared to the oracle phoneme-level segments.
-For example, their segmentation gives 0.93 precision, the ratio of the oracle phoneme-level boundaries being hit, and 0.38 recall, the ratio of the generated boundaries hit the oracle phoneme-level boundaries.
-While Wav2vec-U merges the neighboring segments containing the same predicted
-labels in each step of GAN training, this design can be viewed as refining the
-segmentation implicitly.  
-%Should we mention this?
-%At generator output, the wav2vec-U will combine the neighboring predicted posterior containing the same predicted label into a predicted posterior.
-\end{comment}
-
-%In \parencite{chung_unsupervised_2018, chung_unsupervised_2019a}, the word boundaries are also generated automatically with speech only. 
-%However, \parencite{chung_unsupervised_2018, chung_unsupervised_2019a} need oracle boundaries, therefore not completely unsupervised as the framework proposed in this paper.
-
-% more improvement: iterative training, development 
 As shown in \cref{tab:unsupervisedASR}, most studies use
 \textit{self-training} to refine the models.
-% to achieve better performance.  % AMH: redundant
 In self-training, the generator serves as the first-version phoneme recognition model.
@@
 first-version model.
-%With the new model, we can further obtain new transcriptions.
-%The iteration will continue until the performance converges.
 
-%Using performance to end this section.
 Wav2vec-U~\parencite{baevski_unsupervised_2021} achieved state-of-the-art results at the time, which suggests that representation learning is
@@
 
-%Hung-yi: The technique below is very important. But I think it is fine to ignore this part in a reviewe paper due to space limitation. 
-\begin{comment}
-In unsupervised learning, how to determine the hyperparameters is an issue. 
-In a typical supervised framework, hyperparameter tuning is done by a
-development set, which includes labeled data.
-In an unsupervised setting, the existence of this kind of development is tricky. %Is tricky a correct term?
-The existence of such kind of development set means the presence of the labeled
-data.
-Therefore, wav2vec-U~\parencite{baevski_unsupervised_2021} develops a new approach to
-determine the hyperparameters without a labeled development set. 
-%should I provide more details? 
-\end{comment}
-
-%Hung-yi: A series of work (most from MS) focuses on meager resources but not wholly unsupervised. Perhaps we have to mention the papers.
-
 \paragraph{ASR-TTS}
-%{\color{black} Shinji}\\
-%{\color{black} Hung-yi: In the last meeting, we discussed the connection between this section and self"=supervised learning.
-In addition to looking for papers using self-supervision to improve the ASR-TTS framework here, I think there may be another direction.
-Can we regard this ASR-TTS framework as a kind of self"=supervised learning?
-ASR-TTS is similar to an autoencoder, and the output of ASR is "latent representation".
-In other words, this section describes a special self"=supervised learning method, whose representation is expressed as "text."}\\
-%{\color{black} Hung-yi: However, the above idea sounds tricky. In addition, the ``ASR-TTS'' autoencoder is learned by semi"=supervised, not self"=supervised.}
-% AMH: to make the paper more consistent, I recommend changing all $f^{\mathrm{asr}}$ to $f_{\mathrm{asr}}$
 Here we describe an alternative approach by which to train an ASR and \edit{text-to-speech (TTS)} system
@@
 Thus, we design a loss function $\mathcal{L}_{\mathrm{asr} \rightarrow \mathrm{tts}}(X, \hat{X})$, where $\hat{X}$ is generated by
-%and~$\hat{X}$:
-%\begin{equation}
-%    \mathcal{L}_{\mathrm{asr} \rightarrow \mathrm{tts}} (X, \hat{X}),
-%\end{equation}    
-%where
 \begin{equation}
@@
 the acoustic input by minimizing $\mathcal{L}_{\mathrm{asr} \rightarrow \mathrm{tts}}$.  
-%\begin{equation}
-%\hat{\theta}_{\mathrm{asr}} = \arg \min _{\theta_{\mathrm{asr}}} \mathcal{L}(Y, \hat{Y}; \theta_{\mathrm{asr}}, \theta_{\mathrm{tts}}).
-%\end{equation}
 Note that this approach does not require the supervised text data~$Y$}.
@@
 and document archives without corresponding audio data.
-%Self"=supervised learning in this paper mostly focuses on representation learning perspectives of such unpaired data.
 The typical use case scenario of such a text resource for ASR is through \edit{the
@@
 audio data~$\hat{X}$ with TTS.} %\begin{equation}
-%    \{\hat{X}, Y\} \text{ where } \hat{X} = g_{\mathrm{tts}} (Y).
-%\end{equation}
 The most straightforward approach is to simply use multi-speaker TTS to
@@
 
-% {\color{black} Shinji: puts some notions to encourage the community to contributions to work on this direction and has more connections to the current SSL.}
-
 \subsection{No text or lexicon} \label{subsec:zero}
-
-% \kl{moved the acoustic word embedding subsection from here to the rep learning methods section.}\\
 
@@
 subjective human evaluations. The model completes incomplete words
-(pow[..] $\rightarrow$ POWER) and continues using words in the same general mood (dark $\rightarrow$
+(pow[...] $\rightarrow$ POWER) and continues using words in the same general mood (dark $\rightarrow$
 BLACKNESS)\footnote{https://speechbot.github.io/gslm/} and has been extended to

### 2023-10-14 chapters/paper_review/pretext.tex
Fix many spelling and language errors

@@
 $X=\{x_1,x_2,...,x_T\}$ is the input sequence in which $x_t$ can be an
-acoustic feature vector (e.g., MFCC, filterbank, or spectrogram features)
+acoustic feature vector (e.g., MFCC, filter bank, or spectrogram features)
 or a waveform sample. 

### 2023-10-14 chapters/discussion.tex
Updates to discussion

@@
 
-The last three years have seen the continuation of the rapid progress of machine learning that started about a decade ago with the 2012 ImageNet competation and the work of \textcite{krizhevsky_imagenet_2012}.
+The last three years have seen the continuation of the rapid progress of machine learning that started about a decade ago with the 2012 ImageNet competition and the work of \textcite{krizhevsky_imagenet_2012}.
 The field of out-of-distribution detection saw a surge of interest following the papers by \textcite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019} and self"=supervised learning for speech has seen rapid advances with works such as CPC \parencite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.
 
-Having focused in the previous chapters on VAEs for out-of-distribution detection and self"=supervised methods for representation learning, this chapter will add perspective on how VAEs might be enabled to learn representations more competetive with self"=supervised approaches, and how self"=supervised approaches might be used for out-of-distribution detection.
+Having focused in the previous chapters on VAEs for out-of-distribution detection and self"=supervised methods for representation learning, this chapter will add perspective on how VAEs might be enabled to learn representations more competitive with self"=supervised approaches, and how self"=supervised approaches might be used for out-of-distribution detection.
 Finally, we will discuss the use of calibration techniques for the stroke recognition model of \cref{chp:paper-retrospective}.
@@
 % Similarly, though the wave of works in self"=supervised learning might be traced back to \texttt{word2vec} of \textcite{mikolov_efficient_2013}, the impact of later works such as BERT \parencite{devlin_bert_2018} and GPT \parencite{radford_improving_2018} was immense and helped give rise to many approaches for self"=supervised learning for speech, including CPC \parecite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.
-%
 
@@
 We found that the probabilistic formulation of VAEs provide benefits for their application to uncertainty quantification but that they are generally inferior to self"=supervised methods when it comes to performance on downstream tasks, such as phoneme recognition (see e.g.
-\cref{tab: phoneme recognition (PER), tab:unsupervisedASR}).
-Considering VAEs and self"=supervised methods, it seems we have not yet been able to build a model that has a principled probabilistic representation of uncertainty without sacrificing some of the ability to learn rich representations that are useful for downstream applications.
+\cref{tab: phoneme recognition (PER),tab:unsupervisedASR}).
+Considering VAEs and self"=supervised methods, it seems we have not yet been able to build a model that has a principled probabilistic representation of uncertainty without compromising its ability to learn rich representations that are useful for downstream applications.
 
-% VAEs are good at uncertainty quantification (let's at least say that they are).
-% SSL methods are very good at representation learning as reflected by their dominance on downstream task performance in for text and speech.
+\subsection{Representation learning with variational autoencoders}
 
-\subsection{Improved representation learning with variational autoencoders}
+Different from self"=supervised methods, VAEs learn the distribution over the training data and are forced to encode all aspects important to it. 
+As we saw in \cref{chp:paper-hierarchical}, this will generally include low-level features that are important for accurate input reconstruction, but usually of lesser interest for downstream tasks.
 
-\paragraph{Semi-supervised learning} One way to improve the representations learned in VAEs is via semi-supervised learning. 
-Here a small number of labels are used to inform which patterns that are learned from a large, mostly unlabelled, data set. 
-This is usually done by defining a new stochastic variable as the target and deriving a semi-supervised version of the ELBO that accomadates using the labels when they are available, or marginalizing the target variable when they are not.
-The VAE is then trained on the labelled and unlabelled data simultanously.
+\paragraph{Masking} 
+Reconstruction in VAEs is done from latent variables inferred from the full, unmasked input, and always targets the input.
+This allows the encoder and decoder models to perform well for reconstruction, even with no, or limited, use of contextual dependencies.
+However, as discussed in \cref{chp:paper-review}, masking is one of the driving techniques behind the success of self"=supervised methods \parencite{devlin_bert_2018,baevski_wav2vec_2020}.
+By removing tokens from the input or an early feature extraction layer and tasking a model with inferring their representations, models are forced to learn how neighboring tokens relate to those masked.
+Depending on the size of the mask, these dependencies can be more or less semantic in nature.
+For instance, the wav2vec 2.0 model is well-known to learn representations that have high similarity to word identity and meaning \parencite{pasad_layerwise_2021}.
+Such
 
-VAEs are strong models for semi-supervised learning as proven in several works \parencite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}. 
-However, self"=supervised methods methods have established themselves as superior for most tasks in this setting \cite{jiang_speech_2021, liu_learning_2023}. 
-Furthermore, by training on labelled and unlabelled data simultaneously, the semi-supervised formulation for VAEs is more constraining than it is for self-supervised methods that divide the training into pre"=training and finetuning.
-The expensive pre"=training yields a single foundation model that can be fine"=tuned at relatively low cost for a large number of downstream tasks, whereas VAEs must learn all relevant downstream tasks while also learning from the unlabelled data.
-This is more expensive, but also requires retraining on the unlabelled data, when new supervised data becomes available.
-
-\paragraph{Masking} As discussed in \cref{chp:paper-review}, one of the driving techniques behind the success of self"=supervised methods is masking \parencite{devlin_bert_2018,baevski_wav2vec_2020}.
-By removing tokens from the input or an early feature extraction layer and tasking a model with inferring their representations, the model is forced to learn how neighbouring tokens relate to those masked.
-Depending on the size of the mask, these dependencies can be more or less semantic in nature.
-
-Different from self"=supervised methods, since VAEs learn the distribution over the training data, they are forced to encode all aspects important to it. 
-As we also saw in \cref{chp:paper-hierarchical}, this will generally include low-level features that are important for accurate input reconstruction, but usually of lesser interest for downstream tasks.
-Furthermore, VAE reconstruction is performed based on latent variables which are inferred from the full, unmasked input.
-
-Masking has only been sparsely examined for VAEs.
-One of the areas where it has seen the most attention is missing data imputation.
+Since the ELBO does not allow for masking as part of the training, it has been only sparsely examined for VAEs.
+Specifically, masking has seen the most attention for VAEs within missing data imputation.
 In this setting, the input is partially observed, and often represented as a segmentation into observed and missing parts via a mask that indicates where the data is missing.
 The model is then trained to infer the latent variable from the observed data and reconstruction also deals only with the observed data.
-By comparison to self"=supervised approaches that use masking for representation learning, the missing data setting of VAEs focuses on reconstructing the observed data rather than the missing, which might lead to less useful representations.
+By comparison to self"=supervised approaches that use masking for representation learning, the missing data setting of VAEs focuses on reconstructing the observed data rather than the missing, which might not lead to the same benefits in representation learning.
 The idea of using VAEs to impute missing data was already examined in the seminal paper by \textcite{rezende_stochastic_2014}.
@@
 Previous work that trains on partially observed data has largely focused on the ability of these models to yield high-quality imputations within the tabular and image data domains and have not probed for the effects on the learned latent representation \parencite{mattei_miwae_2019, ipsen_not-miwae_2021}.
+
+\paragraph{Designing the latent space} 
+Architectural improvements and hierarchies of latent variables may be a way to learn more semantic features in VAEs.
+This approach has seen some success in the image domain with models like BIVA \parencite{maaloe_biva_2019}, NVAE \parencite{vahdat_nvae_2020}, and VD-VAE \parencite{child_very_2021}, with up to 78 stochastic layers, achieving tight likelihoods and high-quality samples, although limited effort has put towards using such models for downstream tasks.
+For speech, a hierarchical model operating on multiple temporal scales, such as the Clockwork VAE \parencite{saxena_clockwork_2021} adapted for speech in \cref{chp:paper-benchmarking}, might help better capture dependencies at longer ranges and encode more semantic features.
+For instance, phonetic content for pronunciation might be learned at lower layers, speaker identity at the upper layers, and semantic features, such as word-meaning, in between. 
+Although some work has successfully separated speaker identity from content by modifying the latent variables and their dependencies \parencite{hsu_unsupervised_2017}, models that can learn a deep hierarchy of features for speech remains an open challenge.
+The work presented in \cref{chp:paper-benchmarking} represents an effort to make progress towards such a model.
+
+\paragraph*{Tighter bounds}
+
+% Although the path-wise derivative has lower variance than most alternatives, such as score function estimators, a number of works have attempted to reduce its variance.
+% \textcite{roeder_sticking_2017} note that the path-wise derivative can be factored into a path-wise estimator and a score function estimator and propose to ignore the score function factor to reduce gradient variance.
+% \textcite{rainforth_tighter_2019} show that this problem is even larger for the importance weighted ELBO \parencite{burda_importance_2016} and leads to diminishing returns when increasing the number of samples.
+% Rather than removing the score function factor in the estimate, \textcite{tucker_doubly_2019} propose to reparameterize it and derive low variance gradient estimators, also for importance weighted ELBO.
+
+\paragraph{Semi-supervised learning} 
+Another way to improve the representations learned in VAEs is via semi-supervised learning. 
+Here, a few labels are used to inform which patterns that are learned from a large, mostly unlabeled, data set. 
+This is usually done by defining a new stochastic variable as the target and deriving a semi-supervised version of the ELBO that accommodates using the labels when they are available, or marginalizing the target variable when they are not.
+The VAE is then trained on the labelled and unlabeled data simultaneously.
+
+VAEs are strong models for semi-supervised learning as proven in several works \parencite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}. 
+However, self"=supervised methods have established themselves as superior for most tasks in this setting \cite{baevski_wav2vec_2020,jiang_speech_2021, liu_learning_2023}. 
+Furthermore, by training on labelled and unlabeled data simultaneously, the semi-supervised formulation for VAEs is less flexible than it is for self"=supervised methods that divide the training into two separate phases; the expensive pre"=training phase yields a single foundation model that can be fine"=tuned at relatively low cost for many downstream tasks, but VAEs must learn labeled downstream tasks while simultaneously training on the unlabeled data.
+This is more expensive, but also requires retraining with the unlabeled data, when new supervised data becomes available or a new task is added.
 
@@
 \begin{figure}
-    \begin{subfigure}[c]{0.48\columnwidth}
+    \centering
+    \begin{subfigure}[c]{0.49\columnwidth}
         \centering
@@
     \end{subfigure}
-    % \hfill
-    \begin{subfigure}[c]{0.48\columnwidth}
+    \hfill
+    \begin{subfigure}[c]{0.49\columnwidth}
         \centering
@@
     \end{subfigure}
-    \caption[Calibration curve for the uncalibrated stroke recognition ensemble and empirical distribution of predicted probabilities.]{%
-        Calibration curve for the uncalibrated stroke recognition ensemble (left) and the histogram of predicted probabilities (right) for the test set. 
+    \caption[Calibration curve for the uncalibrated stroke recognition ensemble and empirical distribution of predicted probabilities.]{ Calibration curve for the uncalibrated stroke recognition ensemble (left) and the histogram of predicted probabilities (right) for the test set.
 We use the ensemble that achieved the median F1-score reported in \cref{fig_retrospective:figure1-roc-curve} and \cref{tab_retrospective:table3-occlusion-analysis}.}
@@
     \centering
-    \begin{subfigure}[c]{0.48\columnwidth}
+    \begin{subfigure}[c]{0.49\columnwidth}
         \centering
@@
     \end{subfigure}
-    \begin{subfigure}[c]{0.48\columnwidth}
-        \centering
-        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_fits_ensemble.pdf}
-    \end{subfigure}    
-    \caption[Calibration fits and curves for the stroke recognition ensemble using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right) for the test set.
-We use the ensemble that achieved the median F1-score reported in \cref{fig_retrospective:figure1-roc-curve} and \cref{tab_retrospective:table3-occlusion-analysis}.}
-    \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
-\end{figure}
-\begin{figure}
-    \begin{subfigure}[c]{0.48\columnwidth}
-        \centering
-        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curves_ensemble_with_cis.pdf}
-    \end{subfigure}    
-    \begin{subfigure}[c]{0.48\columnwidth}
+    \hfill
+    \begin{subfigure}[c]{0.49\columnwidth}
         \centering
@@
 \begin{figure}
-    % \begin{subfigure}[c]{0.48\columnwidth}
-    %     \centering
-    %     \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/precision_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
-    %     % \caption{}
-    %     % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
-    % \end{subfigure}    
-    % % \hfill
-    % \begin{subfigure}[c]{0.48\columnwidth}
-    %     \centering
-    %     \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/recall_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
-    %     % \caption{}
-    %     % \label{fig_discussion:histogram_ensemble_and_single_model}
-    % \end{subfigure}
-    \begin{subfigure}[c]{0.48\columnwidth}
+    \centering
+    \begin{subfigure}[c]{0.49\columnwidth}
         \centering
@@
     \end{subfigure}
-    \begin{subfigure}[c]{0.48\columnwidth}
+    \hfill
+    \begin{subfigure}[c]{0.49\columnwidth}
         \centering
@@
     \label{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability}
+    % \begin{subfigure}[c]{0.49\columnwidth}
+    %     \centering
+    %     \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/precision_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
+    %     % \caption{}
+    %     % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
+    % \end{subfigure}    
+    % % \hfill
+    % \begin{subfigure}[c]{0.49\columnwidth}
+    %     \centering
+    %     \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/recall_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
+    %     % \caption{}
+    %     % \label{fig_discussion:histogram_ensemble_and_single_model}
+    % \end{subfigure}
 \end{figure}
@@
 This is likely to lead to alarm fatigue \parencite{embi_evaluating_2012} and can undermine the sustained use and endorsement by clinicians of such systems \parencite{guidi_clinician_2015}.
-The expected precision of the stroke recognition ensemble presented in \cref{chp:paper-retrospective} is $24.9\%$ (95\% confidence interval $24.3-25.5\%)$.
+The expected precision of the stroke recognition ensemble presented in \cref{chp:paper-retrospective} is $24.9\%$ (95\% confidence interval $24.3-25.5\%$).
 This bluntly means it might be wrong three out of four times it predicts a positive.
@@
 % If the stroke recognition ensemble were to be deployed in a randomized controlled trial, s
-Such issues might be alleviated by incorporating the calibrated probabilities in the alerts presented to users.
+Alarm fatigue might be alleviated by incorporating the calibrated probabilities in the alerts presented to users.
 Presented with calibrated probabilities, users would be able to discern between certain and uncertain predictions.
@@
 Similar approaches have been suggested by clinicians and interviews indicate that predictive uncertainty is perceived as a sort of explanation that complements the prediction \cite{tonekaboni_what_2019}.
-In \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability} we show the F1-score of the ensemble model and the call-takers computed on subsets of the data created by binning the predicted probilities of the isotonically calibrated ensemble.
-We note that as might be expected, both call-taker and ensemble model performance increase with increased model certainty.
-This indicates that selecting, based on certainty, which predictions to present to users might indeed help improve trust in the system in practice.
+In \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability} we show the F1-score of the ensemble model and the call-takers computed on subsets of the data created by binning the calibrated probabilities.
+We note that, as might be expected, both call-taker and ensemble model performance increase with increased model certainty.
+This indicates that selecting, based on certainty, which predictions to present to users might indeed help build trust in the system in practice.
 

### 2023-10-15 chapters/background.tex
Update discussion and background

@@
 %
-In this section we will introduce out-of-distribution (OOD) detection and relate it to the concept of uncertainty. 
-To supplement the short related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}, we will then provide a concise review of the literature on out-of-distribution detection.
+In this section we introduce out-of-distribution (OOD) detection and connect it to the concept of uncertainty. 
+Since the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic} are relatively short, we also provide a concise review of the literature on out-of-distribution detection.
 
@@
 In a similar vein, \textcite{dhamija_reducing_2018} propose losses designed to maximize the entropy of $p(y|\xb)$ and decrease feature magnitudes for OOD data sampled from other datasets. 
-\textcite{ruff_deep_2020} use semi"=supervised learning and learn representations of in-distribution data that concentrate close to a centroid in latent space while labelled outliers are pushed away from the centroid.
-Other methods including MCD \parencite{yu_unsupervised_2019}, NGC \parencite{wu_ngc_2021}, and UDG \parencite{yang_semantically_2021} use external, unlabeled, noisy data to improve OOD detection performance without requiring cleanly labelled OOD examples.
+\textcite{ruff_deep_2020} use semi"=supervised learning and learn representations of in-distribution data that concentrate close to a centroid in latent space while labeled outliers are pushed away from the centroid.
+Other methods including MCD \parencite{yu_unsupervised_2019}, NGC \parencite{wu_ngc_2021}, and UDG \parencite{yang_semantically_2021} use external, unlabeled, noisy data to improve OOD detection performance without requiring cleanly labeled OOD examples.
 As we discussed earlier, for many modern applications of machine learning, input data is often high dimensional and complex which makes it difficult to obtain enough representative OOD data to ensure robust OOD detection capabilities.
@@
 \textcite{tack_csi_2020} propose to use a contrastive objective to learn representations of the in-domain data contrasted with data augmented in-samples and use a softmax classifier trained on the representations to compute the OOD score.
-\textcite{sehwag_ssd_2021} present a similar approach but use the feature space Mahalanobis distance similar to \textcite{denouden_improving_2018}.
+\textcite{sehwag_ssd_2021} present a similar approach but use the feature space Mahalanobis distance, similar to \textcite{denouden_improving_2018}.
 \textcite{ahmadian_likelihoodfree_2021} propose to use the latent representation of an invertible generative model to compute the OOD score.
@@
 \section{Variational autoencoders}
-
+%
 % \paragraph{Deep generative models} 
@@
 \subsection{Variational autoencoders} \label{sec:variational-autoencoders}
-To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to parameterize the generative model $p_\thetab(\xb,\zb)$ as a neural network with parameters $\thetab$.
-Specifically, they use a neural network to parameterize the conditional distribution $p_\thetab(\xb|\zb)$ and choose some simple distribution for the prior $p(\zb)$.
-Since this makes the marginal and posterior distributions intractable, both training and inference require approximate methods. 
+To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to parameterize the generative model $p_\thetab(\xb,\zb)$ using a neural network with parameters $\thetab$.
+Specifically, they let $p(\xb,\zb) = p_\thetab(\xb|\zb)p(\zb)$ and use a neural network for the generative model $p_\thetab(\xb|\zb)$ and any distribution that can be efficiently sampled and evaluated for the prior $p(\zb)$.
+This choice makes both the marginal and posterior distributions intractable and so, training and inference require approximate methods. 
 While MCMC methods are computationally expensive for large neural networks, variational inference provides a more scalable alternative by estimating the posterior distribution $p(\zb|\xb)$ with a variational approximation.
@@
 
-Although the path-wise derivative has lower variance than most alternatives, such as score function estimators, a number of works have attempted to reduce its variance.
-\textcite{roeder_sticking_2017} note that the path-wise derivative can be factored into a path-wise estimator and a score function estimator and propose to ignore the score function factor to reduce gradient variance.
-\textcite{rainforth_tighter_2019} show that this problem is even larger for the importance weighted ELBO \parencite{burda_importance_2016} and leads to diminishing returns when increasing the number of samples.
-Rather than removing the score function factor in the estimate, \textcite{tucker_doubly_2019} propose to reparameterize it and derive low variance gradient estimators, also for importance weighted ELBO.
+Although the variance of the path-wise derivative is lower than most alternatives, such as score function estimators, it is not negligible, and a number of works have attempted to reduce it.
+\textcite{roeder_sticking_2017} note that a score function estimator can be factored out of the path-wise derivative and propose to ignore it to reduce gradient variance. 
+\textcite{rainforth_tighter_2019} show that this problem is exacerbated for the importance weighted ELBO \parencite{burda_importance_2016} and that increasing the number of samples in the importance weighted bound also increases gradient variance for the inference network, hurting its ability to learn useful representations.
+After showing that removing the score function factor introduces bias, \textcite{tucker_doubly_2019} propose to reparameterize it too, giving rise to an unbiased, low variance gradient estimator that improves with more samples.
+Later work has generalized this estimator to hierarchical models \parencite{bauer_generalized_2021}.
 

### 2023-10-15 chapters/discussion.tex
Update discussion and background

@@
 
-The last three years have seen the continuation of the rapid progress of machine learning that started about a decade ago with the 2012 ImageNet competition and the work of \textcite{krizhevsky_imagenet_2012}.
-The field of out-of-distribution detection saw a surge of interest following the papers by \textcite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019} and self"=supervised learning for speech has seen rapid advances with works such as CPC \parencite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.
+The rapid progress of machine learning that started about a decade ago with the 2012 ImageNet competition and the work of \textcite{krizhevsky_imagenet_2012} all but slowed down during the course of this project.
+Following the papers by \textcite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019}, the field of out-of-distribution detection saw a surge of interest that has since grown yearly, while self"=supervised learning for speech has continued to advance with works such as CPC \parencite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.
 
-Having focused in the previous chapters on VAEs for out-of-distribution detection and self"=supervised methods for representation learning, this chapter will add perspective on how VAEs might be enabled to learn representations more competitive with self"=supervised approaches, and how self"=supervised approaches might be used for out-of-distribution detection.
-Finally, we will discuss the use of calibration techniques for the stroke recognition model of \cref{chp:paper-retrospective}.
+Having focused on VAEs for out-of-distribution detection in \cref{part:unsupervised-uncertainty-estimation} and on self"=supervised methods for representation learning in \cref{part:self-supervised-speech-representation-learning}, this discussion will bridge a gap between the two approaches and consider how VAEs might be enabled to learn representations more competitive with self"=supervised methods.
+Furthermore, since the work in \cref{part:medical-applications} only had limited focus on uncertainty, we will also present and discuss the use of calibration techniques for the stroke recognition model drawing connections to how such systems are perceived and used in practice.
 
-% Following the papers by \textcite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019}, the field of out-of-distribution detection saw a surge of new approaches using both generative models, supervised classifiers and self"=supervised representations.
-% Similarly, though the wave of works in self"=supervised learning might be traced back to \texttt{word2vec} of \textcite{mikolov_efficient_2013}, the impact of later works such as BERT \parencite{devlin_bert_2018} and GPT \parencite{radford_improving_2018} was immense and helped give rise to many approaches for self"=supervised learning for speech, including CPC \parecite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.
+\section{Representation learning with variational autoencoders}
+%
+In this thesis we studied two different approaches to speech representation learning: VAEs in \cref{chp:paper-hierarchical,chp:paper-modelagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-review}.
+As we saw, the probabilistic formulation of VAEs provides benefits for their application to uncertainty quantification, although competitive methods that use self"=supervised foundation models have also been proposed \textcite{xiao_we_2021,hendrycks_using_2019,bergman_classificationbased_2020}.
+Self"=supervised methods are generally superior when it comes to performance on downstream tasks.
+While this of course depends on the task and the amount and type of unlabeled and labeled data, self"=supervised methods for speech are commonly evaluated on downstream tasks that are harder and more complex than those used for VAEs.
+For instance, between \cref{tab: phoneme recognition (PER),tab:unsupervisedASR} self"=supervised methods can be seen to outperform VAEs for phoneme recognition while self"=supervised methods are also reported for word-level speech recognition.
+In the following, we will discuss potential directions of future research that might help shed light on why VAEs representations underperform on downstream tasks and how they might be improved.
 
-\section{Speech representations and uncertainty}
-%
-In this thesis we have studied two different approaches to learning speech representations: VAEs in \cref{chp:paper-hierarchical,chp:paper-modelagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-review}.
-We found that the probabilistic formulation of VAEs provide benefits for their application to uncertainty quantification but that they are generally inferior to self"=supervised methods when it comes to performance on downstream tasks, such as phoneme recognition (see e.g.
-\cref{tab: phoneme recognition (PER),tab:unsupervisedASR}).
-Considering VAEs and self"=supervised methods, it seems we have not yet been able to build a model that has a principled probabilistic representation of uncertainty without compromising its ability to learn rich representations that are useful for downstream applications.
+% Considering VAEs and self"=supervised methods, it seems we have not yet been able to build a model that has a principled probabilistic representation of uncertainty without compromising its ability to learn representations that are useful for downstream applications.
 
-\subsection{Representation learning with variational autoencoders}
+% Different from self"=supervised methods, VAEs learn the distribution over the training data and are forced to encode all aspects important to it. 
+% As we saw in \cref{chp:paper-hierarchical}, this will generally include low-level features that are important for accurate input reconstruction, but usually of lesser interest for downstream tasks.
 
-Different from self"=supervised methods, VAEs learn the distribution over the training data and are forced to encode all aspects important to it. 
-As we saw in \cref{chp:paper-hierarchical}, this will generally include low-level features that are important for accurate input reconstruction, but usually of lesser interest for downstream tasks.
+\paragraph{Designing the latent space} 
+Architectural improvements and hierarchies of latent variables are two related, and interesting, approaches to learning more semantic features in VAEs. 
+Some success has been demonstrated in the image domain with models like BIVA \parencite{maaloe_biva_2019}, NVAE \parencite{vahdat_nvae_2020}, and VD-VAE \parencite{child_very_2021}, with up to 78 stochastic layers, achieving tight likelihoods and high-quality samples, although limited effort has put towards using such models for downstream tasks.
+For speech, a hierarchical model operating on multiple temporal scales, such as the Clockwork VAE \parencite{saxena_clockwork_2021} adapted for speech in \cref{chp:paper-benchmarking}, might help better capture dependencies at longer ranges and encode more semantic features.
+For instance, phonetic content for pronunciation might be learned at lower layers, speaker identity at the upper layers, and semantic features, such as word-meaning, in between. 
+Although some work has successfully separated speaker identity from content by modifying the latent variables and their dependencies \parencite{hsu_unsupervised_2017}, models that can learn a deep hierarchy of features for speech remains an open challenge.
+The work presented in \cref{chp:paper-benchmarking} represents an effort to make progress towards such a model.
+
+\paragraph{Adding a few labels} 
+Another way to improve the representations learned in VAEs is via semi-supervised learning. 
+Here, a few labels are used to inform which patterns that are learned from a large, mostly unlabeled, data set. 
+This is usually done by defining a new stochastic variable as the target and deriving a semi-supervised version of the ELBO that accommodates using the labels when they are available, or marginalizing the target variable when they are not.
+The VAE is then trained on the labeled and unlabeled data simultaneously.
+
+Although VAEs are strong models for semi-supervised learning \parencite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}, self"=supervised methods have established themselves as superior for most tasks in this setting \cite{baevski_wav2vec_2020,jiang_speech_2021, liu_learning_2023}.
+Despite being theoretically appealing, joint objectives for semi-supervised learning in VAEs have practical drawbacks.
+By training on labeled and unlabeled data simultaneously, semi-supervised VAEs are often less flexible than are self"=supervised methods that divide the training into two separate phases.
+By first fitting a general foundation model in an expensive pre"=training phase it can then later be fine"=tuned for many downstream tasks, at relatively low cost.
+VAEs, on the other hand, must often learn the labeled downstream tasks while simultaneously training on the unlabeled data.
+This is computationally expensive, but also requires retraining with the unlabeled data, when new supervised data becomes available or a new task is added.
+
+\paragraph*{Approximating less}
+VAE are inherently approximate.
+Training is performed on a lower bound of the likelihood and inference is variational, amortized and for non-hierarchical models, mean field.
+As a consequence, the gradient itself is estimated inexactly, usually by only one posterior sample, leading to non-negligible variance.
+For that reason, it seems plausible that using tighter bounds and reducing gradient variance improve representation learning in VAEs. 
+Importance weighting the ELBO \parencite{burda_importance_2016} provides a tight bound on the likelihood, reduces gradient variance, and induces a complex implicit posterior distribution. 
+Its use has become standard when reporting likelihood benchmarks but as we described in \cref{sec:variational-autoencoders}, \textcite{rainforth_tighter_2019} demonstrated that using it during training introduces high gradient variance for the inference network hurting its ability to learn useful representations.
+Later works largely solved this issue \parencite{roeder_sticking_2017,tucker_doubly_2019,bauer_generalized_2021} and showed that the reduced variance leads to improved likelihoods.
+
+Despite much of this progress being prior to the latest, very large hierarchical VAEs, such as BIVA \parencite{maaloe_biva_2019}, NVAE \parencite{vahdat_nvae_2020}, and VD-VAE \parencite{child_very_2021}, these models are all trained with the single-sample ELBO using regularly reparameterized gradient estimator.
+Instead, these models rely on advanced inference networks \parencite{maaloe_biva_2019} and architectural improvements \parencite{vahdat_nvae_2020, child_very_2021} to overcome the challenges of training hierarchical VAEs.
+This seems to indicate that there might be untapped potential in consolidating the work on low-variance gradient estimators and advanced architectures and inference networks.
+
+% Although the path-wise derivative has lower variance than most alternatives, such as score function estimators, a number of works have attempted to reduce its variance.
+% \textcite{roeder_sticking_2017} note that the path-wise derivative can be factored into a path-wise estimator and a score function estimator and propose to ignore the score function factor to reduce gradient variance.
+% \textcite{rainforth_tighter_2019} show that this problem is even larger for the importance weighted ELBO \parencite{burda_importance_2016} and leads to diminishing returns when increasing the number of samples.
+% Rather than removing the score function factor in the estimate, \textcite{tucker_doubly_2019} propose to reparameterize it and derive low variance gradient estimators, also for importance weighted ELBO.
 
 \paragraph{Masking} 
-Reconstruction in VAEs is done from latent variables inferred from the full, unmasked input, and always targets the input.
-This allows the encoder and decoder models to perform well for reconstruction, even with no, or limited, use of contextual dependencies.
-However, as discussed in \cref{chp:paper-review}, masking is one of the driving techniques behind the success of self"=supervised methods \parencite{devlin_bert_2018,baevski_wav2vec_2020}.
+As discussed in \cref{chp:paper-review}, masking is one of the driving techniques behind the success of self"=supervised methods for speech \parencite{devlin_bert_2018,baevski_wav2vec_2020}.
 By removing tokens from the input or an early feature extraction layer and tasking a model with inferring their representations, models are forced to learn how neighboring tokens relate to those masked.
 Depending on the size of the mask, these dependencies can be more or less semantic in nature.
-For instance, the wav2vec 2.0 model is well-known to learn representations that have high similarity to word identity and meaning \parencite{pasad_layerwise_2021}.
-Such
+For instance, the wav2vec 2.0 model is well-known to learn representations that have high similarity with word identity and meaning \parencite{pasad_layerwise_2021}.
+In comparison, reconstruction in VAEs is done from latent variables that are inferred from the full, unmasked input.
+Since all information is generally available, this might allow the encoder and decoder models to perform well for reconstruction, even with no, or limited, use of contextual dependencies.
+Furthermore, VAE reconstruction almost always targets the direct input in order to learn the distribution over the training data and to enable generating new samples.
+However, compared to many self"=supervised methods that target intermediate, learned representations, this forces VAEs to encode all aspects of the input that are important to its distribution.
+As we saw in \cref{chp:paper-hierarchical}, this will generally include low-level features that are necessary for accurate input reconstruction.
+Such features require large latent space representations and model capacity \parencite{vahdat_nvae_2020,child_very_2021}, but are usually of lesser interest for downstream tasks \parencite{baevski_wav2vec_2020}.
 
-Since the ELBO does not allow for masking as part of the training, it has been only sparsely examined for VAEs.
+Since the ELBO does not immediately allow for masking as part of the training, it has been only sparsely examined for VAEs.
 Specifically, masking has seen the most attention for VAEs within missing data imputation.
@@
 
-\paragraph{Designing the latent space} 
-Architectural improvements and hierarchies of latent variables may be a way to learn more semantic features in VAEs.
-This approach has seen some success in the image domain with models like BIVA \parencite{maaloe_biva_2019}, NVAE \parencite{vahdat_nvae_2020}, and VD-VAE \parencite{child_very_2021}, with up to 78 stochastic layers, achieving tight likelihoods and high-quality samples, although limited effort has put towards using such models for downstream tasks.
-For speech, a hierarchical model operating on multiple temporal scales, such as the Clockwork VAE \parencite{saxena_clockwork_2021} adapted for speech in \cref{chp:paper-benchmarking}, might help better capture dependencies at longer ranges and encode more semantic features.
-For instance, phonetic content for pronunciation might be learned at lower layers, speaker identity at the upper layers, and semantic features, such as word-meaning, in between. 
-Although some work has successfully separated speaker identity from content by modifying the latent variables and their dependencies \parencite{hsu_unsupervised_2017}, models that can learn a deep hierarchy of features for speech remains an open challenge.
-The work presented in \cref{chp:paper-benchmarking} represents an effort to make progress towards such a model.
+Including masked objectives into the principled probabilistic framework of VAEs might require creative thinking, but it seems likely that relaxing the requirement of exact input reconstruction and enforcing the need of encoding (and decoding) contextual information could lead the way for future types of VAEs to learn representations that are competitive with self"=supervised methods.
 
-\paragraph*{Tighter bounds}
+% \subsection{Uncertainty estimation with self-supervised methods}
+% %
 
-% Although the path-wise derivative has lower variance than most alternatives, such as score function estimators, a number of works have attempted to reduce its variance.
-% \textcite{roeder_sticking_2017} note that the path-wise derivative can be factored into a path-wise estimator and a score function estimator and propose to ignore the score function factor to reduce gradient variance.
-% \textcite{rainforth_tighter_2019} show that this problem is even larger for the importance weighted ELBO \parencite{burda_importance_2016} and leads to diminishing returns when increasing the number of samples.
-% Rather than removing the score function factor in the estimate, \textcite{tucker_doubly_2019} propose to reparameterize it and derive low variance gradient estimators, also for importance weighted ELBO.
+% \textcite{pasad_layerwise_2021} found that layers 6-7 hold the most information about phonetic content and word identity and meaning for \texttt{wav2vec2-base}. 
+% For \texttt{wav2vec2-large} the phonetic content is highest in layers 11 and 18-19 with a drop in between, while word identity and meaning are highest between layers 12 and 18.
 
-\paragraph{Semi-supervised learning} 
-Another way to improve the representations learned in VAEs is via semi-supervised learning. 
-Here, a few labels are used to inform which patterns that are learned from a large, mostly unlabeled, data set. 
-This is usually done by defining a new stochastic variable as the target and deriving a semi-supervised version of the ELBO that accommodates using the labels when they are available, or marginalizing the target variable when they are not.
-The VAE is then trained on the labelled and unlabeled data simultaneously.
-
-VAEs are strong models for semi-supervised learning as proven in several works \parencite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}. 
-However, self"=supervised methods have established themselves as superior for most tasks in this setting \cite{baevski_wav2vec_2020,jiang_speech_2021, liu_learning_2023}. 
-Furthermore, by training on labelled and unlabeled data simultaneously, the semi-supervised formulation for VAEs is less flexible than it is for self"=supervised methods that divide the training into two separate phases; the expensive pre"=training phase yields a single foundation model that can be fine"=tuned at relatively low cost for many downstream tasks, but VAEs must learn labeled downstream tasks while simultaneously training on the unlabeled data.
-This is more expensive, but also requires retraining with the unlabeled data, when new supervised data becomes available or a new task is added.
-
-\subsection{Uncertainty estimation with self-supervised methods}
-%
-
-\textcite{pasad_layerwise_2021} found that layers 6-7 hold the most information about phonetic content and word identity and meaning for \texttt{wav2vec2-base}. 
-For \texttt{wav2vec2-large} the phonetic content is highest in layers 11 and 18-19 with a drop in between, while word identity and meaning are highest between layers 12 and 18.
-
-Uncertainty of representations versus uncertainty of downstream task outputs. 
-% \textcite{wickstrom_relax_2023} propose a method for quantifying uncertainty based on
+% Uncertainty of representations versus uncertainty of downstream task outputs. 
+% % \textcite{wickstrom_relax_2023} propose a method for quantifying uncertainty based on
 
@@
 Clinicians expect an alert to present a prediction that aligns with patient status.
-Despite expert-agreed thresholds for when alerts should be triggered, however, many alerts may not be aligned since the unbalanced and ambiguous nature of many predictive problems in healthcare can lead to relatively low predictive precision \parencite{umscheid_development_2015, cite14, cite15, wenstrup_retrospective_2023}.
-This is likely to lead to alarm fatigue \parencite{embi_evaluating_2012} and can undermine the sustained use and endorsement by clinicians of such systems \parencite{guidi_clinician_2015}.
-The expected precision of the stroke recognition ensemble presented in \cref{chp:paper-retrospective} is $24.9\%$ (95\% confidence interval $24.3-25.5\%$).
-This bluntly means it might be wrong three out of four times it predicts a positive.
-This unfortunately risks alarm fatigue among its potential users which might diminish it's real-world impact.
+Despite expert-agreed thresholds for when alerts should be triggered, however, many alerts may not be aligned since class imbalance and ambiguous information in many predictive problems in healthcare can lead to models with relatively low predictive precision \parencite{umscheid_development_2015, cite14, cite15, wenstrup_retrospective_2023}.
+In turn, this is likely to lead to alarm fatigue \parencite{embi_evaluating_2012} and can undermine the sustained use and endorsement by clinicians of such systems \parencite{guidi_clinician_2015}.
+The stroke recognition ensemble presented in \cref{chp:paper-retrospective} is not exempt from this risk.
+With a precision of $24.9\%$ (95\% confidence interval $24.3-25.5\%$) it will on average be wrong three out of four times it predicts a stroke.
+This unfortunately risks alarm fatigue among its potential users diminishing its effect in practice.
 
 % If the stroke recognition ensemble were to be deployed in a randomized controlled trial, s
-Alarm fatigue might be alleviated by incorporating the calibrated probabilities in the alerts presented to users.
-Presented with calibrated probabilities, users would be able to discern between certain and uncertain predictions.
-Furthermore, calibrated probabilities might be used to present users only with predictions that exceed a specific confidence level.
-Similar approaches have been suggested by clinicians and interviews indicate that predictive uncertainty is perceived as a sort of explanation that complements the prediction \cite{tonekaboni_what_2019}.
+Calibrated probabilities in the alerts presented to users might be a way to alleviate alarm fatigue.
+Calibrated probabilities would allow users to discern between certain and uncertain predictions and also enable the system to present users only with predictions that have at least a certain probability of being correct.
+Similar approaches have been suggested by clinicians and interviews indicate that predictive uncertainty is perceived by experts as a sort of explanation that complements the prediction \cite{tonekaboni_what_2019}.
 In \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability} we show the F1-score of the ensemble model and the call-takers computed on subsets of the data created by binning the calibrated probabilities.
 We note that, as might be expected, both call-taker and ensemble model performance increase with increased model certainty.
-This indicates that selecting, based on certainty, which predictions to present to users might indeed help build trust in the system in practice.
+This indicates that selecting, based on certainty, which predictions to present to users might indeed help build trust in the system and ensure a practical impact.
 

### 2023-10-15 chapters/discussion.tex
Add todos

@@
 In the following, we will discuss potential directions of future research that might help shed light on why VAEs representations underperform on downstream tasks and how they might be improved.
-
-% Considering VAEs and self"=supervised methods, it seems we have not yet been able to build a model that has a principled probabilistic representation of uncertainty without compromising its ability to learn representations that are useful for downstream applications.
-
-% Different from self"=supervised methods, VAEs learn the distribution over the training data and are forced to encode all aspects important to it. 
-% As we saw in \cref{chp:paper-hierarchical}, this will generally include low-level features that are important for accurate input reconstruction, but usually of lesser interest for downstream tasks.
 
@@
 
-% Although the path-wise derivative has lower variance than most alternatives, such as score function estimators, a number of works have attempted to reduce its variance.
-% \textcite{roeder_sticking_2017} note that the path-wise derivative can be factored into a path-wise estimator and a score function estimator and propose to ignore the score function factor to reduce gradient variance.
-% \textcite{rainforth_tighter_2019} show that this problem is even larger for the importance weighted ELBO \parencite{burda_importance_2016} and leads to diminishing returns when increasing the number of samples.
-% Rather than removing the score function factor in the estimate, \textcite{tucker_doubly_2019} propose to reparameterize it and derive low variance gradient estimators, also for importance weighted ELBO.
-
 \paragraph{Masking} 
@@
 
+\paragraph{Giving up?}
+\lesstodo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimizing a mutual information term in the ELBO (derive this form).}
+\lesstodo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other).
+Include reference to \parencite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
+
+\lesstodo[inline]{Out-of-distribution detection on speech?}
+
 % \subsection{Uncertainty estimation with self-supervised methods}
@@
 % % \textcite{wickstrom_relax_2023} propose a method for quantifying uncertainty based on
-
-% \section{\Cref{chp:paper-hierarchical} revisited: \dots}
-
-% \lesstodo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other).
-Include reference to \parencite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
-% \lesstodo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimizing a mutual information term in the ELBO (derive this form).}
-% \lesstodo[inline]{Discuss some recent work e.g. \cref{morningstar_density_2021}}
-
-% \section{\Cref{chp:paper-benchmarking} revisited: Bested?}
-
-% \lesstodo[inline]{Discuss whether variational autoencoders are viable for learning good representations for speech for downstream tasks when alternatives such as SSL method exist.}
-% \lesstodo[inline]{Out-of-distribution detection on speech?}
 

### 2023-10-17 appendices/supplementary_retrospective.tex
Updates to discussion
Update bibliography

@@
 
-\begin{sidewaysfigure}[h]
+\begin{sidewaysfigure}
     \centering
@@
 % \todo[inline]{Insert table A2 here}
-\begin{table}[h]
+\begin{table}
     \centering
@@
 % \todo[inline]{Insert table A3 here}
-\begin{table}[h]
+\begin{table}
     \centering
@@
 
-\begin{table}[h]
+\begin{table}
     \centering
@@
 % % \todo[inline]{Insert table A4 here}
-\begin{table}[h]
+\begin{table}
     \centering
@@
 % \todo[inline]{Insert table A5 here}
-\begin{table}[h]
+\begin{table}
     \centering
@@
         30. & Hævelse & Swelling & $-0.159$ & $17762$ & $-3.71 \pm 2.32$ \\
-
         \bottomrule
@@
 % \todo[inline]{Insert table A6 here}
-\begin{sidewaystable}[h]
+\begin{sidewaystable}
     \centering
@@
 \begin{enumerate}
-    \item Soun JE, Chow DS, Nagamine M et al. Artificial intelligence and acute stroke imaging. AJNR. American journal of neuroradiology 2021; 42: 2-11.
-    \item Murray NM, Unberath M, Hager GD, Hui FK.
-Artificial intelligence to diagnose ischemic stroke and identify large vessel occlusions: a systematic review.
-Journal of neurointerventional surgery 2020; 12: 156-64.
-    \item Shlobin NA, Baig AA, Waqas M et al. Artificial intelligence for large-vessel occlusion stroke: a systematic review. World neurosurgery 2022; 159: 207-20.e1.
-    \item Linder SM, Rosenfeldt AB, Bay RC, Sahu K, Wolf SL, Alberts JL.
-Improving quality of life and depression after stroke through telerehabilitation.
-The American journal of occupational therapy: official publication of the American Occupational Therapy Association 2015; 69: 6902290020p1-10.
-    \item Humphreys K, Shover CL, Andrews CM et al.
-Responding to the opioid crisis in North America and beyond: recommendations of the Stanford-Lancet Commission.
-Lancet (London, England) 2022; 399: 555-604.
-    \item Fassbender K, Lesmeister M, Merzou F. Prehospital stroke management and mobile stroke units. Current opinion in neurology 2023; 36: 140-6.
-    \item McDonough RV, Ospel JM, Majoie CBLM et al.
-Clinical outcome of patients with mild pre-stroke morbidity following endovascular treatment: a HERMES substudy.
-Journal of neurointerventional surgery 2023; 15: 214-20.
-    \item Bat-Erdene BO, Saver JL.
-Automatic acute stroke symptom detection and emergency medical systems alerting by mobile health technologies: a review.
-Journal of stroke and cerebrovascular diseases: the official journal of National Stroke Association 2021; 30: 105826.
-    \item Bonini N, Vitolo M, Imberti JF et al. Mobile health technology in atrial fibrillation. Expert review of medical devices 2022; 19: 327-40.
-    \item Scholz ML, Collatz-Christensen H, Blomberg SNF, Boebel S, Verhoeven J, Krafft T.
-Artificial intelligence in Emergency Medical Services dispatching: assessing the potential impact of an automatic speech recognition software on stroke detection taking the Capital Region of Denmark as case in point.
-Scandinavian journal of trauma, resuscitation and emergency medicine 2022; 30: 36.
-    \item Gupta R, Krishnam SP, Schaefer PW, Lev MH, Gonzalez RG.
-An East Coast perspective on artificial intelligence and machine learning: part 2: ischemic stroke imaging and triage.
-Neuroimaging clinics of North America 2020; 30: 467-78.
-    \item Esbjörnsson M, Ullberg T.
-Safety and usability of wearable accelerometers for stroke detection the STROKE ALARM PRO 1 study.
-Journal of stroke and cerebrovascular diseases: the official journal of National Stroke Association 2022; 31: 106762.
-    \item Fast L, Temuulen U, Villringer K et al. Machine learning-based prediction of clinical outcomes after first-ever ischemic stroke. Frontiers in neurology 2023; 14: 1114360.
-    \item Gupta R, Krishnam SP, Schaefer PW, Lev MH, Gilberto Gonzalez R.
-An East Coast perspective on artificial intelligence and machine learning: part 1: hemorrhagic stroke imaging and triage.
-Neuroimaging clinics of North America 2020; 30: 459-66.
-    \item Cai T, Ni H, Yu M et al. DeepStroke: An efficient stroke screening framework for emergency rooms with multimodal adversarial deep learning. Medical image analysis 2022; 80: 102522.
-    \item Zachrison KS, Dhand A, Schwamm LH, Onnela JP. A network approach to stroke systems of care. Circulation. Cardiovascular quality and outcomes 2019; 12: e005526.
-    \item O'Connell GC, Walsh KB, Smothers CG et al. Use of deep artificial neural networks to identify stroke during triage via subtle changes in circulating cell counts. BMC neurology 2022; 22: 206.
-    \item Yang L, Liu Q, Zhao Q, Zhu X, Wang L. Machine learning is a valid method for predicting prehospital delay after acute ischemic stroke. Brain and behavior 2020; 10: e01794.
-    \item Aldridge CM, McDonald MM, Wruble M et al. Human vs. machine learning based detection of facial weakness using video analysis. Frontiers in neurology 2022; 13: 878282.
-    \item Lachance CC, Ford C.
-Portable stroke detection devices for patients with stroke symptoms: a review of diagnostic accuracy and cost-effectiveness.
-Canadian Agency for Drugs and Technologies in Health 2019
-    \item Fardoun HM, Mashat AS. Methodologies, models and algorithms for patients rehabilitation. Methods of information in medicine 2016; 55: 60-4.
-    \item Wibring K, Magnusson C, Axelsson C, Lundgren P, Herlitz J, Andersson Hagiwara M.
-Towards definitions of time-sensitive conditions in prehospital care.
-Scandinavian journal of trauma, resuscitation and emergency medicine 2020; 28: 7.
-    \item Mayampurath A, Parnianpour Z, Richards CT et al. Improving prehospital stroke diagnosis using natural language processing of paramedic reports. Stroke 2021; 52: 2676-9.
-    \item Morey JR, Zhang X, Yaeger KA et al.
-Real-world experience with artificial intelligence-based triage in transferred large vessel occlusion stroke patients.
-Cerebrovascular diseases (Basel, Switzerland) 2021; 50: 450-5.
-    \item Sung SF, Hung LC, Hu YH.
-Developing a stroke alert trigger for clinical decision support at emergency triage using machine learning.
-International journal of medical informatics 2021; 152: 104505.
-    \item Zhang Z, Zhou D, Zhang J et al. Multilayer perceptron-based prediction of stroke mimics in prehospital triage. Scientific reports 2022; 12: 17994.
-    \item Park E, Lee K, Han T, Nam HS.
-Automatic grading of stroke symptoms for rapid assessment using optimized machine learning and 4-limb kinematics: clinical validation study.
-Journal of medical Internet research 2020; 22: e20641.
-    \item Chen M, Tan X, Padman R.
-A machine learning approach to support urgent stroke triage using administrative data and social determinants of health at hospital presentation: retrospective study.
-Journal of medical Internet research 2023; 25: e36477.
-    \item Uchida K, Kouno J, Yoshimura S et al.
-Development of machine learning models to predict probabilities and types of stroke at prehospital stage: the Japan urgent stroke triage score using machine learning (JUST-ML).
-Translational stroke research 2022; 13: 370-81.
-    \item Finck T, Schinz D, Grundl L et al. Automated detection of ischemic stroke and subsequent patient triage in routinely acquired head CT. Clinical neuroradiology 2022; 32: 419-26.
-    \item Sangal RB, Fodeh S, Taylor A et al.
-Identification of patients with nontraumatic intracranial hemorrhage using administrative claims data.
-Journal of stroke and cerebrovascular diseases: the official journal of National Stroke Association 2020; 29: 105306.
-    \item Adedinsewo D, Carter RE, Attia Z et al.
-Artificial intelligence-enabled ECG algorithm to identify patients with left ventricular systolic dysfunction presenting to the emergency department with dyspnea.
-Circulation.
-Arrhythmia and electrophysiology 2020; 13: e008437.
-    \item Liu L, Yu B, Han M, Yuan S, Wang N. Mild cognitive impairment understanding: an empirical study by data-driven approach. BMC bioinformatics 2019; 20: 481.
-    \item Hayashi Y, Shimada T, Hattori N et al. A prehospital diagnostic algorithm for strokes using machine learning: a prospective observational study. Scientific reports 2021; 11: 20519.
-    \item Grant L, Joo P, Nemnom MJ, Thiruganasambandamoorthy V.
-Machine learning versus traditional methods for the development of risk stratification scores: a case study using original Canadian Syncope Risk Score data.
-Internal and emergency medicine 2022; 17: 1145-53.
-    \item van Meenen LCC, van Stigt MN, Marquering HA et al.
-Detection of large vessel occlusion stroke with electroencephalography in the emergency room: first results of the ELECTRA-STROKE study.
-Journal of neurology 2022; 269: 2030-8.
-    \item van Meenen LCC, den Hartog SJ, Groot AE et al.
-Relationship between primary stroke center volume and time to endovascular thrombectomy in acute ischemic stroke.
-European journal of neurology 2021; 28: 4031-8.
-    \item Chien HC, Yang TL, Juang WC, Chen YA, Li YJ, Chen CY.
-Pilot report for intracranial hemorrhage detection with deep learning implanted head computed tomography images at emergency department.
-Journal of medical systems 2022; 46: 49.
-    \item Huo D, Leppert M, Pollard R et al.
-Large vessel occlusion prediction in the emergency department with national institutes of health stroke scale components: a machine learning approach.
-Journal of stroke and cerebrovascular diseases: the official journal of National Stroke Association 2021; 30: 106030.
-    \item Wu MR, Chen YT, Li ZX et al.
-Dysphagia screening and pneumonia after subarachnoid hemorrhage: findings from the Chinese stroke center alliance.
-CNS neuroscience \& therapeutics 2022; 28: 913-21.
-    \item Bacchi S, Oakden-Rayner L, Zerner T, Kleinig T, Patel S, Jannes J.
-Deep learning natural language processing successfully predicts the cerebrovascular cause of transient ischemic attack-like presentations.
-Stroke 2019; 50: 758-60.
-    \item Heldt FS, Vizcaychipi MP, Peacock S et al. Early risk assessment for COVID-19 patients from emergency department data using machine learning. Scientific reports 2021; 11: 4200.
-    \item Lu Z, Xiong Y, Yang K et al. What predicts large vessel occlusion in mild stroke patients. BMC neurology 2023; 23: 29.
-    \item Marijon E, Garcia R, Narayanan K, Karam N, Jouven X.
-Fighting against sudden cardiac death: need for a paradigm shift-adding near-term prevention and pre-emptive action to long-term prevention.
-European heart journal 2022; 43: 1457-64.
-    \item You J, Yu PLH, Tsang ACO et al.
-3D dissimilar-Siamese-u-net for hyperdense Middle cerebral artery sign segmentation.
-Computerized medical imaging and graphics: the official journal of the Computerized Medical Imaging Society 2021; 90: 101898.
-    \item Moral-Munoz JA, Zhang W, Cobo MJ, Herrera-Viedma E, Kaber DB.
-Smartphone-based systems for physical rehabilitation applications: a systematic review.
-Assistive technology: the official journal of RESNA 2021; 33: 223-36.
-    \item Salman OH, Aal-Nouman MI, Taha ZK, Alsabah MQ, Hussein YS, Abdelkareem ZA.
-Formulating multi diseases dataset for identifying, triaging and prioritizing patients to multi medical emergency levels: Simulated dataset accompanied with codes.
-Data in brief 2021; 34: 106576.
-    \item Desai A, Zumbo A, Giordano M et al.
-Word2vec Word embedding-based artificial intelligence model in the triage of patients with suspected diagnosis of major ischemic stroke: a feasibility study.
-International journal of environmental research and public health 2022; 19: 15295.
-    \item Rinkel LA, Prick JCM, Slot RER et al. Impact of the COVID-19 outbreak on acute stroke care. Journal of neurology 2021; 268: 403-8.
-    \item Chilamkurthy S, Ghosh R, Tanamala S et al. Deep learning algorithms for detection of critical findings in head CT scans: a retrospective study. Lancet (London, England) 2018; 392: 2388-96.
-    \item Kramer NM, Demaerschalk BM.
-A novel application of teleneurology: robotic telepresence in supervision of neurology trainees.
-Telemedicine journal and e-health: the official journal of the American Telemedicine Association 2014; 20: 1087-92.
-    \item Yadav K, Sarioglu E, Choi HA, Cartwright WB, Hinds PS, Chamberlain JM.
-Automated outcome classification of computed tomography imaging reports for pediatric traumatic brain injury.
-Academic emergency medicine: official journal of the Society for Academic Emergency Medicine 2016; 23: 171-8.
-    \item van Meenen LCC, Riedijk F, Stolp J et al.
-Pre- and interhospital workflow times for patients with large vessel occlusion stroke transferred for endovasvular thrombectomy.
-Frontiers in neurology 2021; 12: 730250.
-    \item Saberi-Movahed F, Mohammadifard M, Mehrpooya A et al.
-Decoding clinical biomarker space of COVID-19: Exploring matrix factorization-based feature selection methods.
-Computers in biology and medicine 2022; 146: 105426.
-    \item Shimada T, Matsubara K, Koyama D et al.
-Development of evaluation system for cerebral artery occlusion in emergency medical services: noninvasive measurement and utilization of pulse waves.
-Scientific reports 2023; 13: 3339.
-    \item Gil-Jardiné C, Chenais G, Pradeau C et al.
-Trends in reasons for emergency calls during the COVID-19 crisis in the department of Gironde, France using artificial neural network for natural language classification.
-Scandinavian journal of trauma, resuscitation and emergency medicine 2021; 29: 55.
-    \item Santala OE, Lipponen JA, Jäntti H et al.
-Continuous mHealth Patch Monitoring for the Algorithm-Based Detection of Atrial Fibrillation: Feasibility and Diagnostic Accuracy Study.
-JMIR cardio 2022; 6: e31230.
-    \item O'Brien MK, Shawen N, Mummidisetty CK et al.
-Activity recognition for persons with stroke using mobile phone technology: toward improved performance in a home setting.
-Journal of medical Internet research 2017; 19: e184.
-    \item Bateman RM, Sharpe MD, Jagger JE et al.
-36th international symposium on intensive care and emergency medicine: Brussels, Belgium.
-15-18 March 2016.
-Critical care (London, England) 2016; 20: 94.
-    \item Modica CM, Johnson BR, Zalewski C et al. Hearing loss and irritability reporting without vestibular differences in explosive breaching professionals. Frontiers in neurology 2020; 11: 588377.
-    \item Palmcrantz S, Plantin J, Borg J. Factors affecting the usability of an assistive soft robotic glove after stroke or multiple sclerosis. Journal of rehabilitation medicine 2020; 52: jrm00027.
-    \item Mitsopoulos K, Fiska V, Tagaras K et al.
-NeuroSuitUp: System architecture and validation of a motor rehabilitation wearable robotics and serious game platform.
-Sensors (Basel, Switzerland) 2023; 23: 3281.
-    \item Meng G, Tan Y, Fang M, Yang H, Liu X, Zhao Y.
-Meteorological factors related to emergency admission of elderly stroke patients in shanghai: analysis with a multilayer perceptron neural network.
-Medical science monitor: international medical journal of experimental and clinical research 2015; 21: 3600-7.
-    \item Iglesias JE, Schleicher R, Laguna S et al. Quantitative Brain Morphometry of Portable Low-Field-Strength MRI Using Super-Resolution Machine Learning. Radiology 2023; 306: e220522.
-    \item Lai F. Stroke networks based on robotic telepresence. Journal of telemedicine and telecare 2009; 15: 135-6
-    \item Kim D, Oh J, Im H, Yoon M, Park J, Lee J.
-Automatic classification of the Korean triage acuity scale in simulated emergency rooms using speech recognition and natural language processing: a proof of concept study.
-Journal of Korean medical science 2021; 36: e175.
-    \item Jonsson A, Binongo J, Patel P et al. Mastering the learning curve for robotic-assisted coronary artery bypass surgery. The Annals of thoracic surgery 2023; 115: 1118-25.
-    \item Knowlton SM, Sencan I, Aytar Y et al. Sickle cell detection using a smartphone. Scientific reports 2015; 5: 15022.
-    \item Che F, Liu Y, Gong X et al.
-Extracranial carotid plaque hemorrhage is independently associated with poor 3-month functional outcome after acute ischemic stroke-a prospective cohort study.
-Frontiers in neurology 2021; 12: 780436.
-    \item Boers AMM, Jansen IGH, Brown S et al.
-Mediation of the relationship between endovascular therapy and functional outcome by follow-up infarct volume in patients with acute ischemic stroke.
-JAMA neurology 2019; 76: 194-202.
-    \item Stone JR, Avants BB, Tustison NJ et al.
-Functional and structural neuroimaging correlates of repetitive low-level blast exposure in career breachers.
-Journal of neurotrauma 2020; 37: 2468-81.
-    \item Rickards CA, Vyas N, Ryan KL et al.
-Are you bleeding?
-Validation of a machine-learning algorithm for determination of blood volume status: application to remote triage.
-Journal of applied physiology (Bethesda, Md.
-: 1985) 2014; 116: 486-94.
-    \item Edwards KA, Greer K, Leete J et al. Neuronally-derived tau is increased in experienced breachers and is associated with neurobehavioral symptoms. Scientific reports 2021; 11: 19527.
-    \item Cheng AL, Liu J, Bravo S, Miller JC, Pahlevan NM.
-Screening left ventricular systolic dysfunction in children using intrinsic frequencies of carotid pressure waveforms measured by a novel smartphone-based device.
-Physiological measurement 2023; 44.
-    \item Grunwald IQ, Kulikovski J, Reith W et al.
-Collateral automation for triage in stroke: evaluating automated scoring of collaterals in acute stroke on computed tomography scans.
-Cerebrovascular diseases (Basel, Switzerland) 2019; 47: 217-22.
-    \item Hyakutake K, Morishita T, Saita K et al.
-Effects of home-based robotic therapy involving the single-joint hybrid assistive limb robotic suit in the chronic phase of stroke: a pilot study.
-BioMed research international 2019; 2019: 5462694.
-    \item Murray C, Ortiz E, Kubin C. Application of a robot for critical care rounding in small rural hospitals. Critical care nursing clinics of North America 2014; 26: 477-85.
-    \item Edwards KA, Leete JJ, Tschiffely AE et al. Blast exposure results in tau and neurofilament light chain changes in peripheral blood. Brain injury 2020; 34: 1213-21.
-    \item Oresko JJ, Duschl H, Cheng AC.
-A wearable smartphone-based platform for real-time cardiovascular disease detection via electrocardiogram processing.
-IEEE transactions on information technology in biomedicine : a publication of the IEEE Engineering in Medicine and Biology Society 2010; 14: 734-40.
-    \item Alcoceba-Herrero I, Coco-Martín MB, Leal-Vega L et al.
-Randomized controlled trial evaluating the benefit of a novel clinical decision support system for the management of COVID-19 patients in home quarantine: a study protocol.
-International journal of environmental research and public health 2023; 20: 2300.
-    \item van Meenen LCC, Groot AE, Venema E et al.
-Interhospital transfer vs.
-direct presentation of patients with a large vessel occlusion not eligible for IV thrombolysis.
-Journal of neurology 2020; 267: 2142-50.
-    \item Forghani R, Gupta R. Case of the Season: Artificial intelligence in clinical practice-large vessel occlusion triage in stroke imaging. Seminars in roentgenology 2023; 58: 147-51.
-    \item Jabbour P, Gonzalez LF, Tjoumakaris S, Randazzo C, Rosenwasser R. Stroke in the robotic era. World neurosurgery 2010; 73: 603-4.
-    \item Larkin H. On the spot when doc is not. Hospitals \& health networks 2008; 82: 41-2.
-    \item Jiang Y, Guarino P, Ma S et al. Bayesian accrual prediction for interim review of clinical studies: open source R package and smartphone application. Trials 2016; 17: 336.
-    \item Lee JH, Oh BJ, Ahn JY et al.
-Effectiveness of automatic acute stroke alert system based on UMLS mapped local terminology codes at emergency department.
-AMIA ...
-Annual Symposium proceedings.
-AMIA Symposium 2008; 1018.
-    \item Robot saves ED stroke patients, addresses subspecialist shortage. ED management: the monthly update on emergency department management 2011; 23: 13-5.
-    \item Nagy Z, Simon P, Sipos E, Kozmann G.
-The main elements of the information system of the National Stroke Program (Smart Card - Telecommunication - Knowledge Bases).
-Medinfo.
-MEDINFO 1995; 8 Pt 2: 1496-9.
+    \item Soun J.E., Chow D.S., Nagamine M. et al. "Artificial Intelligence and Acute Stroke Imaging". \textit{American Journal of Neuroradiology} (2021); 42: 2-11.
+    \item Murray N.M., Unberath M., Hager G.D., Hui F.K.
+"Artificial Intelligence to Diagnose Ischemic Stroke and Identify Large Vessel Occlusions: A Systematic Review".
+\textit{Journal of Neurointerventional Surgery} (2020); 12: 156-64.
+    \item Shlobin N.A., Baig A.A., Waqas M. et al. "Artificial Intelligence for Large-Vessel Occlusion Stroke: A Systematic Review". \textit{World Neurosurgery} (2022); 159: 207-20.e1.
+    \item Linder S.M., Rosenfeldt A.B., Bay R.C., Sahu K., Wolf S.L., Alberts J.L.
+"Improving Quality of Life and Depression After Stroke Through Telerehabilitation".
+\textit{The American Journal of Occupational Therapy: Official Publication of the American Occupational Therapy Association} (2015); 69: 6902290020p1-10.
+    \item Humphreys K., Shover C.L., Andrews C.M.
+et al.
+"Responding to the Opioid Crisis in North America and Beyond: Recommendations of the Stanford-Lancet Commission".
+\textit{The Lancet (London, England)} (2022); 399: 555-604.
+    \item Fassbender K., Lesmeister M., Merzou F. "Prehospital Stroke Management and Mobile Stroke Units". \textit{Current Opinion in Neurology} (2023); 36: 140-6.
+    \item McDonough R.V., Ospel J.M., Majoie C.B.L.M.
+et al.
+"Clinical Outcome of Patients with Mild Pre-Stroke Morbidity Following Endovascular Treatment: A HERMES Substudy".
+\textit{Journal of Neurointerventional Surgery} (2023); 15: 214-20.
+    \item Bat-Erdene B.O., Saver J.L.
+"Automatic Acute Stroke Symptom Detection and Emergency Medical Systems Alerting by Mobile Health Technologies: A Review".
+\textit{Journal of Stroke and Cerebrovascular Diseases: The Official Journal of National Stroke Association} (2021); 30: 105826.
+    \item Bonini N., Vitolo M., Imberti J.F. et al. "Mobile Health Technology in Atrial Fibrillation". \textit{Expert Review of Medical Devices} (2022); 19: 327-40.
+    \item Scholz M.L., Collatz-Christensen H., Blomberg S.N.F., Boebel S., Verhoeven J., Krafft T.
+"Artificial Intelligence in Emergency Medical Services Dispatching: Assessing the Potential Impact of an Automatic Speech Recognition Software on Stroke Detection Taking the Capital Region of Denmark as Case in Point".
+\textit{Scandinavian Journal of Trauma, Resuscitation, and Emergency Medicine} (2022); 30: 36.
+    \item Gupta R., Krishnam S.P., Schaefer P.W., Lev M.H., Gonzalez R.G.
+"An East Coast Perspective on Artificial Intelligence and Machine Learning: Part 2 - Ischemic Stroke Imaging and Triage".
+\textit{Neuroimaging Clinics of North America} (2020); 30: 467-78.
+    \item Esbjörnsson M., Ullberg T.
+"Safety and Usability of Wearable Accelerometers for Stroke Detection: The STROKE ALARM PRO 1 Study".
+\textit{Journal of Stroke and Cerebrovascular Diseases: The Official Journal of National Stroke Association} (2022); 31: 106762.
+    \item Fast L., Temuulen U., Villringer K. et al. "Machine Learning-Based Prediction of Clinical Outcomes After First-Ever Ischemic Stroke". \textit{Frontiers in Neurology} (2023); 14: 1114360.
+    \item Gupta R., Krishnam S.P., Schaefer P.W., Lev M.H., Gilberto Gonzalez R.
+"An East Coast Perspective on Artificial Intelligence and Machine Learning: Part 1 - Hemorrhagic Stroke Imaging and Triage".
+\textit{Neuroimaging Clinics of North America} (2020); 30: 459-66.
+    \item Cai T., Ni H., Yu M.
+et al.
+"DeepStroke: An Efficient Stroke Screening Framework for Emergency Rooms with Multimodal Adversarial Deep Learning".
+\textit{Medical Image Analysis} (2022); 80: 102522.
+    \item Zachrison K.S., Dhand A., Schwamm L.H., Onnela J.P. "A Network Approach to Stroke Systems of Care". \textit{Circulation. Cardiovascular Quality and Outcomes} (2019); 12: e005526.
+    \item O'Connell G.C., Walsh K.B., Smothers C.G.
+et al.
+"Use of Deep Artificial Neural Networks to Identify Stroke During Triage via Subtle Changes in Circulating Cell Counts".
+\textit{BMC Neurology} (2022); 22: 206.
+    \item Yang L., Liu Q., Zhao Q., Zhu X., Wang L. "Machine Learning Is a Valid Method for Predicting Prehospital Delay After Acute Ischemic Stroke". \textit{Brain and Behavior} (2020); 10: e01794.
+    \item Aldridge C.M., McDonald M.M., Wruble M. et al. "Human vs. Machine Learning-Based Detection of Facial Weakness Using Video Analysis". \textit{Frontiers in Neurology} (2022); 13: 878282.
+    \item Lachance C.C., Ford C.
+"Portable Stroke Detection Devices for Patients with Stroke Symptoms: A Review of Diagnostic Accuracy and Cost-Effectiveness".
+\textit{Canadian Agency for Drugs and Technologies in Health} (2019).
+    \item Fardoun H.M., Mashat A.S. "Methodologies, Models, and Algorithms for Patients Rehabilitation". \textit{Methods of Information in Medicine} (2016); 55: 60-4.
+    \item Wibring K., Magnusson C., Axelsson C., Lundgren P., Herlitz J., Andersson Hagiwara M.
+"Towards Definitions of Time-Sensitive Conditions in Prehospital Care".
+\textit{Scandinavian Journal of Trauma, Resuscitation, and Emergency Medicine} (2020); 28: 7.
+    \item Mayampurath A., Parnianpour Z., Richards C.T. et al. "Improving Prehospital Stroke Diagnosis Using Natural Language Processing of Paramedic Reports". \textit{Stroke} (2021); 52: 2676-9.
+    \item Morey J.R., Zhang X., Yaeger K.A.
+et al.
+"Real-World Experience with Artificial Intelligence-Based Triage in Transferred Large Vessel Occlusion Stroke Patients".
+\textit{Cerebrovascular Diseases (Basel, Switzerland)} (2021); 50: 450-5.
+    \item Sung S.F., Hung L.C., Hu Y.H.
+"Developing a Stroke Alert Trigger for Clinical Decision Support at Emergency Triage Using Machine Learning".
+\textit{International Journal of Medical Informatics} (2021); 152: 104505.
+    \item Zhang Z., Zhou D., Zhang J. et al. "Multilayer Perceptron-Based Prediction of Stroke Mimics in Prehospital Triage". \textit{Scientific Reports} (2022); 12: 17994.
+    \item Park E., Lee K., Han T., Nam H.S.
+"Automatic Grading of Stroke Symptoms for Rapid Assessment Using Optimized Machine Learning and 4-Limb Kinematics: Clinical Validation Study".
+\textit{Journal of Medical Internet Research} (2020); 22: e20641.
+    \item Chen M., Tan X., Padman R.
+"A Machine Learning Approach to Support Urgent Stroke Triage Using Administrative Data and Social Determinants of Health at Hospital Presentation: Retrospective Study".
+\textit{Journal of Medical Internet Research} (2023); 25: e36477.
+    \item Uchida K., Kouno J., Yoshimura S.
+et al.
+"Development of Machine Learning Models to Predict Probabilities and Types of Stroke at Prehospital Stage: The Japan Urgent Stroke Triage Score Using Machine Learning (JUST-ML)".
+\textit{Translational Stroke Research} (2022); 13: 370-81.
+    \item Finck T., Schinz D., Grundl L.
+et al.
+"Automated Detection of Ischemic Stroke and Subsequent Patient Triage in Routinely Acquired Head CT".
+\textit{Clinical Neuroradiology} (2022); 32: 419-26.
+    \item Sangal R.B., Fodeh S., Taylor A.
+et al.
+"Identification of Patients with Nontraumatic Intracranial Hemorrhage Using Administrative Claims Data".
+\textit{Journal of Stroke and Cerebrovascular Diseases: The Official Journal of National Stroke Association} (2020); 29: 105306.
+    \item Adedinsewo D., Carter R.E., Attia Z.
+et al.
+"Artificial Intelligence-Enabled ECG Algorithm to Identify Patients with Left Ventricular Systolic Dysfunction Presenting to the Emergency Department with Dyspnea".
+\textit{Circulation.
+Arrhythmia and Electrophysiology} (2020); 13: e008437.
+    \item Liu L., Yu B., Han M., Yuan S., Wang N. "Mild Cognitive Impairment Understanding: An Empirical Study by Data-Driven Approach". \textit{BMC Bioinformatics} (2019); 20: 481.
+    \item Hayashi Y., Shimada T., Hattori N.
+et al.
+"A Prehospital Diagnostic Algorithm for Strokes Using Machine Learning: A Prospective Observational Study".
+\textit{Scientific Reports} (2021); 11: 20519.
+    \item Grant L., Joo P., Nemnom M.J., Thiruganasambandamoorthy V.
+"Machine Learning Versus Traditional Methods for the Development of Risk Stratification Scores: A Case Study Using Original Canadian Syncope Risk Score Data".
+\textit{Internal and Emergency Medicine} (2022); 17: 1145-53.
+    \item van Meenen L.C.C., van Stigt M.N., Marquering H.A.
+et al.
+"Detection of Large Vessel Occlusion Stroke with Electroencephalography in the Emergency Room: First Results of the ELECTRA-STROKE Study".
+\textit{Journal of Neurology} (2022); 269: 2030-8.
+    \item van Meenen L.C.C., den Hartog S.J., Groot A.E.
+et al.
+"Relationship Between Primary Stroke Center Volume and Time to Endovascular Thrombectomy in Acute Ischemic Stroke".
+\textit{European Journal of Neurology} (2021); 28: 4031-8.
+    \item Chien H.C., Yang T.L., Juang W.C., Chen Y.A., Li Y.J., Chen C.Y.
+"Pilot Report for Intracranial Hemorrhage Detection with Deep Learning Implanted Head Computed Tomography Images at Emergency Department".
+\textit{Journal of Medical Systems} (2022); 46: 49.
+    \item Huo D., Leppert M., Pollard R.
+et al.
+"Large Vessel Occlusion Prediction in the Emergency Department with National Institutes of Health Stroke Scale Components: A Machine Learning Approach".
+\textit{Journal of Stroke and Cerebrovascular Diseases: The Official Journal of National Stroke Association} (2021); 30: 106030.
+    \item Wu M.R., Chen Y.T., Li Z.X.
+et al.
+"Dysphagia Screening and Pneumonia After Subarachnoid Hemorrhage: Findings from the Chinese Stroke Center Alliance".
+\textit{CNS Neuroscience \& Therapeutics} (2022); 28: 913-21.
+    \item Bacchi S., Oakden-Rayner L., Zerner T., Kleinig T., Patel S., Jannes J.
+"Deep Learning Natural Language Processing Successfully Predicts the Cerebrovascular Cause of Transient Ischemic Attack-Like Presentations".
+\textit{Stroke} (2019); 50: 758-60.
+    \item Heldt F.S., Vizcaychipi M.P., Peacock S.
+et al.
+"Early Risk Assessment for COVID-19 Patients from Emergency Department Data Using Machine Learning".
+\textit{Scientific Reports} (2021); 11: 4200.
+    \item Lu Z., Xiong Y., Yang K. et al. "What Predicts Large Vessel Occlusion in Mild Stroke Patients". \textit{BMC Neurology} (2023); 23: 29.
+    \item Marijon E., Garcia R., Narayanan K., Karam N., Jouven X.
+"Fighting Against Sudden Cardiac Death: Need for a Paradigm Shift-Adding Near-Term Prevention and Pre-Emptive Action to Long-Term Prevention".
+\textit{European Heart Journal} (2022); 43: 1457-64.
+    \item You J., Yu P.L.H., Tsang A.C.O.
+et al.
+"3D Dissimilar-Siamese-U-Net for Hyperdense Middle Cerebral Artery Sign Segmentation".
+\textit{Computerized Medical Imaging and Graphics: The Official Journal of the Computerized Medical Imaging Society} (2021); 90: 101898.
+    \item Moral-Munoz J.A., Zhang W., Cobo M.J., Herrera-Viedma E., Kaber D.B.
+"Smartphone-Based Systems for Physical Rehabilitation Applications: A Systematic Review".
+\textit{Assistive Technology: The Official Journal of RESNA} (2021); 33: 223-36.
+    \item Salman O.H., Aal-Nouman M.I., Taha Z.K., Alsabah M.Q., Hussein Y.S., Abdelkareem Z.A.
+"Formulating Multi-Diseases Dataset for Identifying, Triaging, and Prioritizing Patients to Multi-Medical Emergency Levels: Simulated Dataset Accompanied with Codes".
+\textit{Data in Brief} (2021); 34: 106576.
+    \item Desai A., Zumbo A., Giordano M.
+et al.
+"Word2Vec Word Embedding-Based Artificial Intelligence Model in the Triage of Patients with Suspected Diagnosis of Major Ischemic Stroke: A Feasibility Study".
+\textit{International Journal of Environmental Research and Public Health} (2022); 19: 15295.
+    \item Rinkel L.A., Prick J.C.M., Slot R.E.R. et al. "Impact of the COVID-19 Outbreak on Acute Stroke Care". \textit{Journal of Neurology} (2021); 268: 403-8.
+    \item Chilamkurthy S., Ghosh R., Tanamala S.
+et al.
+"Deep Learning Algorithms for Detection of Critical Findings in Head CT Scans: A Retrospective Study".
+\textit{The Lancet (London, England)} (2018); 392: 2388-96.
+    \item Kramer N.M., Demaerschalk B.M.
+"A Novel Application of Teleneurology: Robotic Telepresence in Supervision of Neurology Trainees".
+\textit{Telemedicine Journal and E-Health: The Official Journal of the American Telemedicine Association} (2014); 20: 1087-92.
+    \item Yadav K., Sarioglu E., Choi H.A., Cartwright W.B., Hinds P.S., Chamberlain J.M.
+"Automated Outcome Classification of Computed Tomography Imaging Reports for Pediatric Traumatic Brain Injury".
+\textit{Academic Emergency Medicine: Official Journal of the Society for Academic Emergency Medicine} (2016); 23: 171-8.
+    \item van Meenen L.C.C., Riedijk F., Stolp J.
+et al.
+"Pre- and Interhospital Workflow Times for Patients with Large Vessel Occlusion Stroke Transferred for Endovascular Thrombectomy".
+\textit{Frontiers in Neurology} (2021); 12: 730250.
+    \item Saberi-Movahed F., Mohammadifard M., Mehrpooya A.
+et al.
+"Decoding Clinical Biomarker Space of COVID-19: Exploring Matrix Factorization-Based Feature Selection Methods".
+\textit{Computers in Biology and Medicine} (2022); 146: 105426.
+    \item Shimada T., Matsubara K., Koyama D.
+et al.
+"Development of Evaluation System for Cerebral Artery Occlusion in Emergency Medical Services: Noninvasive Measurement and Utilization of Pulse Waves".
+\textit{Scientific Reports} (2023); 13: 3339.
+    \item Gil-Jardiné C., Chenais G., Pradeau C.
+et al.
+"Trends in Reasons for Emergency Calls During the COVID-19 Crisis in the Department of Gironde, France Using Artificial Neural Network for Natural Language Classification".
+\textit{Scandinavian Journal of Trauma, Resuscitation and Emergency Medicine} (2021); 29: 55.
+    \item Santala O.E., Lipponen J.A., Jäntti H.
+et al.
+"Continuous mHealth Patch Monitoring for the Algorithm-Based Detection of Atrial Fibrillation: Feasibility and Diagnostic Accuracy Study".
+\textit{JMIR Cardio} (2022); 6: e31230.
+    \item O'Brien M.K., Shawen N., Mummidisetty C.K.
+et al.
+"Activity Recognition for Persons with Stroke Using Mobile Phone Technology: Toward Improved Performance in a Home Setting".
+\textit{Journal of Medical Internet Research} (2017); 19: e184.
+    \item Bateman R.M., Sharpe M.D., Jagger J.E.
+et al.
+"36th International Symposium on Intensive Care and Emergency Medicine: Brussels, Belgium.
+15-18 March 2016".
+\textit{Critical Care (London, England)} (2016); 20: 94.
+    \item Modica C.M., Johnson B.R., Zalewski C.
+et al.
+"Hearing Loss and Irritability Reporting Without Vestibular Differences in Explosive Breaching Professionals".
+\textit{Frontiers in Neurology} (2020); 11: 588377.
+    \item Palmcrantz S., Plantin J., Borg J.
+"Factors Affecting the Usability of an Assistive Soft Robotic Glove after Stroke or Multiple Sclerosis".
+\textit{Journal of Rehabilitation Medicine} (2020); 52: jrm00027.
+    \item Mitsopoulos K., Fiska V., Tagaras K.
+et al.
+"NeuroSuitUp: System Architecture and Validation of a Motor Rehabilitation Wearable Robotics and Serious Game Platform".
+\textit{Sensors (Basel, Switzerland)} (2023); 23: 3281.
+    \item Meng G., Tan Y., Fang M., Yang H., Liu X., Zhao Y.
+"Meteorological Factors Related to Emergency Admission of Elderly Stroke Patients in Shanghai: Analysis with a Multilayer Perceptron Neural Network".
+\textit{Medical Science Monitor: International Medical Journal of Experimental and Clinical Research} (2015); 21: 3600-7.
+    \item Iglesias J.E., Schleicher R., Laguna S.
+et al.
+"Quantitative Brain Morphometry of Portable Low-Field-Strength MRI Using Super-Resolution Machine Learning".
+\textit{Radiology} (2023); 306: e220522.
+    \item Lai F. "Stroke Networks Based on Robotic Telepresence". \textit{Journal of Telemedicine and Telecare} (2009); 15: 135-6.
+    \item Kim D., Oh J., Im H., Yoon M., Park J., Lee J.
+"Automatic Classification of the Korean Triage Acuity Scale in Simulated Emergency Rooms Using Speech Recognition and Natural Language Processing: A Proof of Concept Study".
+\textit{Journal of Korean Medical Science} (2021); 36: e175.
+    \item Jonsson A., Binongo J., Patel P. et al. "Mastering the Learning Curve for Robotic-Assisted Coronary Artery Bypass Surgery". \textit{The Annals of Thoracic Surgery} (2023); 115: 1118-25.
+    \item Knowlton S.M., Sencan I., Aytar Y. et al. "Sickle Cell Detection Using a Smartphone". \textit{Scientific Reports} (2015); 5: 15022.
+    \item Che F., Liu Y., Gong X.
+et al.
+"Extracranial Carotid Plaque Hemorrhage Is Independently Associated with Poor 3-Month Functional Outcome after Acute Ischemic Stroke-A Prospective Cohort Study".
+\textit{Frontiers in Neurology} (2021); 12: 780436.
+    \item Boers A.M.M., Jansen I.G.H., Brown S.
+et al.
+"Mediation of the Relationship Between Endovascular Therapy and Functional Outcome by Follow-Up Infarct Volume in Patients with Acute Ischemic Stroke".
+\textit{JAMA Neurology} (2019); 76: 194-202.
+    \item Stone J.R., Avants B.B., Tustison N.J.
+et al.
+"Functional and Structural Neuroimaging Correlates of Repetitive Low-Level Blast Exposure in Career Breachers".
+\textit{Journal of Neurotrauma} (2020); 37: 2468-81.
+    \item Rickards C.A., Vyas N., Ryan K.L.
+et al.
+"Are You Bleeding?
+Validation of a Machine-Learning Algorithm for Determination of Blood Volume Status: Application to Remote Triage".
+\textit{Journal of Applied Physiology (Bethesda, Md.
+: 1985)} (2014); 116: 486-94.
+    \item Edwards K.A., Greer K., Leete J.
+et al.
+"Neuronally-Derived Tau Is Increased in Experienced Breachers and Is Associated with Neurobehavioral Symptoms".
+\textit{Scientific Reports} (2021); 11: 19527.
+    \item Cheng A.L., Liu J., Bravo S., Miller J.C., Pahlevan N.M.
+"Screening Left Ventricular Systolic Dysfunction in Children Using Intrinsic Frequencies of Carotid Pressure Waveforms Measured by a Novel Smartphone-Based Device".
+\textit{Physiological Measurement} (2023); 44.
+    \item Grunwald I.Q., Kulikovski J., Reith W.
+et al.
+"Collateral Automation for Triage in Stroke: Evaluating Automated Scoring of Collaterals in Acute Stroke on Computed Tomography Scans".
+\textit{Cerebrovascular Diseases (Basel, Switzerland)} (2019); 47: 217-22.
+    \item Hyakutake K., Morishita T., Saita K.
+et al.
+"Effects of Home-Based Robotic Therapy Involving the Single-Joint Hybrid Assistive Limb Robotic Suit in the Chronic Phase of Stroke: A Pilot Study".
+\textit{BioMed Research International} (2019); 2019: 5462694.
+    \item Murray C., Ortiz E., Kubin C. "Application of a Robot for Critical Care Rounding in Small Rural Hospitals". \textit{Critical Care Nursing Clinics of North America} (2014); 26: 477-85.
+    \item Edwards K.A., Leete J.J., Tschiffely A.E. et al. "Blast Exposure Results in Tau and Neurofilament Light Chain Changes in Peripheral Blood". \textit{Brain Injury} (2020); 34: 1213-21.
+    \item Oresko J.J., Duschl H., Cheng A.C.
+"A Wearable Smartphone-Based Platform for Real-Time Cardiovascular Disease Detection via Electrocardiogram Processing".
+\textit{IEEE Transactions on Information Technology in Biomedicine: A Publication of the IEEE Engineering in Medicine and Biology Society} (2010); 14: 734-40.
+    \item Alcoceba-Herrero I., Coco-Martín M.B., Leal-Vega L.
+et al.
+"Randomized Controlled Trial Evaluating the Benefit of a Novel Clinical Decision Support System for the Management of COVID-19 Patients in Home Quarantine: A Study Protocol".
+\textit{International Journal of Environmental Research and Public Health} (2023); 20: 2300.
+    \item van Meenen L.C.C., Groot A.E., Venema E.
+et al.
+"Interhospital Transfer vs.
+Direct Presentation of Patients with a Large Vessel Occlusion Not Eligible for IV Thrombolysis".
+\textit{Journal of Neurology} (2020); 267: 2142-50.
+    \item Forghani R., Gupta R.
+"Case of the Season: Artificial Intelligence in Clinical Practice-Large Vessel Occlusion Triage in Stroke Imaging".
+\textit{Seminars in Roentgenology} (2023); 58: 147-51.
+    \item Jabbour P., Gonzalez L.F., Tjoumakaris S., Randazzo C., Rosenwasser R. "Stroke in the Robotic Era". \textit{World Neurosurgery} (2010); 73: 603-4.
+    \item Larkin H. "On the Spot When Doc Is Not". \textit{Hospitals \& Health Networks} (2008); 82: 41-2.
+    \item Jiang Y., Guarino P., Ma S. et al. "Bayesian Accrual Prediction for Interim Review of Clinical Studies: Open Source R Package and Smartphone Application". \textit{Trials} (2016); 17: 336.
+    \item Lee J.H., Oh B.J., Ahn J.Y.
+et al.
+"Effectiveness of Automatic Acute Stroke Alert System Based on UMLS Mapped Local Terminology Codes at Emergency Department".
+\textit{AMIA ...
+Annual Symposium Proceedings.
+AMIA Symposium} (2008); 1018.
+    \item "Robot Saves ED Stroke Patients, Addresses Subspecialist Shortage". \textit{ED Management: The Monthly Update on Emergency Department Management} (2011); 23: 13-5.
+    \item Nagy Z., Simon P., Sipos E., Kozmann G.
+"The Main Elements of the Information System of the National Stroke Program (Smart Card - Telecommunication - Knowledge Bases)".
+\textit{Medinfo.
+MEDINFO} (1995); 8 Pt 2: 1496-9.
 \end{enumerate}
 }
-
 }

### 2023-10-18 chapters/introduction.tex
Update introduction and discussion

@@
 The application of artificial intelligence in medical decision-making and particularly for diagnosing and recommending treatments is likely to hold significant potential to reduce the effect of these issues and improve patient outcomes.
-However, as such decisions impact critical aspects of human health and well-being, this prospect also raises concerns related to the reliability of artificial intelligence systems \parencite{chen_ethical_2021, shailaja_machine_2018, ahmad_interpretable_2018}.
+However, as such decisions impact critical aspects of human health and well-being, this prospect also raises concerns related to the reliability of systems that use artificial intelligence \parencite{chen_ethical_2021, shailaja_machine_2018, ahmad_interpretable_2018}.
 
-The evolution of technology to support and automate decision-making can be traced back to the development of algorithms and mathematical techniques by ancient civilizations. 
-The Egyptian Rhind Mathematical Papyrus, dated to around 1550 BCE, describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \parencite{georges_universal_2001}.
+The evolution of technology to support decision-making can be traced back to very beginning of recorded history.
+After the agricultural revolution, several ancient civilizations developed mathematical techniques and algorithms to help them manage the newfound complexity of pre-modern society.
+Fittingly, the earliest human whose name is known to recorded history was a Sumerian accountant called Kushim.
+Their signature has been identified on at least 18 clay tablets that deal with transactions and inventory of barley and other ingredients for beer production \parencite{nissen_archaic_1993}.
+Another example is the Egyptian Rhind Mathematical Papyrus dated to around 1550 BCE which describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \parencite{georges_universal_2001}.
 Around the same time, the Babylonians had developed a sophisticated system of mathematics that included algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \parencite{fowler_square_1998}.
 
-These computing tools brought with them novel challenges.
-The Babylonian mathematics, for instance, did not formalize the difference between rational and irrational numbers often leading to approximation errors such as the common practice of approximating $\pi$ with 3 \parencite{georges_universal_2001}.
-Apart from such inherent limitations of the tools, inadequate human proficiency in their use was also a source of error.
-For example, the Plimpton 322 clay tablet (\cref{fig:plimpton_332}) contains several errors, presumably made by a novice in Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
+% \begin{figure}
+%     \centering
+%     \includegraphics[height=0.4\textheight]{kushim_multiplication.png}
+%     \caption[]{ }
+%     \label{fig:kushim_multiplication}
+% \end{figure}
+
+Besides enabling the advancement of early society, these technologies also brought with them novel challenges. 
+In a likely attempt to reduce the effort needed to do Sumerian mathematics, Kushim produced a clay tablet with a multiplication table including fractional representations of common numbers.
+Assumedly unbeknownst to Kushim, this tablet would also come to contain one of the world's first recorded examples of a mathematical error; the statement: $5 \times \dfrac{1/2} = 5$ \parencite{nissen_archaic_1993}.
+The Plimpton 322 clay tablet (\cref{fig:plimpton_332}) contains several errors in its list of Pythagorean triplets, presumably made by a novice in Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
+% Apart from such inherent limitations of the tools, inadequate human proficiency in their use was also a source of error. 
+Besides these human errors, inherent limitations of the tools also posed problems. 
+Babylonian mathematicians did not formalize the difference between rational and irrational numbers which lead to approximation errors.
+For instance, the common practice of approximating $\pi$ with 3 \parencite{georges_universal_2001}.
 % Easy to overlook, the challenges we face in adopting new technologies today are not new but have been faced by humans for thousands of years.
@@
 The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics.
-\parencite[photo credit][]{neugebauer_mathematical_1945}}
+\parencite[photo credit][]{neugebauer_mathematical_1945, cuneiformdigitallibraryinitiativecdli_mct_2005}}
     \label{fig:plimpton_332}

### 2023-10-18 chapters/background.tex
Update introduction and discussion

@@
 \paragraph{Hierarchical models}\label{sec:hierarchical-vae-background}
-The usual mean-field assumption imposed by letting $\zb \sim\mathcal{N}(\mub,\mathbf{I}\sigmab^2)$ can often be more restrictive for model expressiveness than we would like. 
+The mean-field assumption imposed by letting $\zb \sim\mathcal{N}(\mub,\mathbf{I}\sigmab^2)$ can often be more restrictive for model expressiveness than we would like. 
 While learning a full covariance is a simple solution, it is not always sufficiently computationally efficient and, in any case, enables learning only linear covariance between elements of $\zb$. 
-Along with the usual motivations of hierarchical models such as efficient representation, feature disentanglement, and data efficiency, this has led to research into using a hierarchy of several conditionally dependent latent variables.
-This idea is commonly formalized by introducing a set of $L$ latent variables  $\zb=\zb_1, \dots, \zb_L$ and letting the generative model be defined as,
+The drawbacks of mean-field approximation and linear covariance has lead to research into learning hierarchies of several non-linearly dependent latent variables; an idea well in line with the usual motivations behind deep neural networks such as efficient, compositional representation \cite{lecun_deep_2015}.
+Such hierarchical models are usually formalized by introducing a set of $L$ latent variables  $\zb=\zb_1, \dots, \zb_L$ and letting the generative model be defined as,
 %

### 2023-10-18 chapters/discussion.tex
Update bibliography
Update introduction and discussion

@@
 For instance, the wav2vec 2.0 model is well-known to learn representations that have high similarity with word identity and meaning \parencite{pasad_layerwise_2021}.
+
 In comparison, reconstruction in VAEs is done from latent variables that are inferred from the full, unmasked input.
@@
 By comparison to self"=supervised approaches that use masking for representation learning, the missing data setting of VAEs focuses on reconstructing the observed data rather than the missing, which might not lead to the same benefits in representation learning.
+
 The idea of using VAEs to impute missing data was already examined in the seminal paper by \textcite{rezende_stochastic_2014}.
@@
 
-\paragraph{Giving up?}
-\lesstodo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimizing a mutual information term in the ELBO (derive this form).}
-\lesstodo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other).
+\paragraph{Stop representing?}
+
+Two key observations relating to the training of VAEs might lead to the more pessimistic view that VAEs are fundamentally unsuited for representation learning. 
+First, variational autoencoders are trained to maximize the ELBO which is a marginal likelihood objective that is conditionally independent of the latent variables $\zb$.
+Second, as we saw in \cref{eq:variational-autoencoder-elbo-factorized-expected}, the ELBO contains a hidden
+
++ maximum likelihood does not guarantee good representations of unseen data. On the contrary; the two are actually only correlated by inductive biases present in the model type that we care about. 
++ VAEs do maximum likelihood.
+    + but this is true for any model trained with maximum likelihood
+    + in fact masked pre-trained used for many self-supervised methods can be seen as maximum likelihood training.
+
++ mutual information minimization of VAEs leads to difficulty of training and posterior collapse.
+    + only when properly constraining the inference and generative models do VAEs function.
+    + if the combined inference and generative models easily minimize the mutual information, while producing good reconstructions and making the marginal posterior match the prior, then the model will learn to disregard z, whatever is encoded in it.
+    + this is especially prone to happen for strong generative models with those that are autoregressive on the observed variable as the core example.
+Since $p(\xb_t)$ in many cases can be modelled very well by a generative model of the form $p(\xb_t | \xb_{1:t-1})$, the potential improvement from using $p(\xb_t | \xb_{1:t-1}, \zb_{1:t})$, or even $p(\xb_t | \xb_{1:t-1}, \zb_{1:T})$, might be small compared to the magnitude of the mutual information term
+    \cref{eq:variational-autoencoder-elbo-factorized-expected}
+
+\begin{equation} \label{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)}
+    \mathbb{E}_{q_{}(\xb,\zb)} \left[ \mathcal{L}_{\thetab,\phib}(\xb) \right] = 
+    \underset{\text{average reconstruction}}{\underbrace{
+        \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log p_{}(\xb | \zb) \right]
+    }}
+    - 
+    \underset{\text{marginal KL to prior}}{\underbrace{
+    \vphantom{\mathbb{E}_{q_{}(\xb,\zb)}\left[ \log p_{}(\xb | \zb) \right]}{
+        D_{\mathrm{KL}}\left[ q_{}(\zb) \parallel p(\zb) \right]
+    }
+    }} 
+    -
+    \underset{\text{mutual information}}{\underbrace{
+    I_{q_{}(\xb,\zb)}\left[\xb ; \zb \right]
+    }} \enspace . \nonumber
+\end{equation}
+
+\parencite{tomczak_trouble_2022}
+\parencite{huszar_is_2017}
+
+\todo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimizing a mutual information term in the ELBO (derive this form).}
+\todo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other).
 Include reference to \parencite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
 
-\lesstodo[inline]{Out-of-distribution detection on speech?}
+\todo[inline]{Masked pre-training actually corresponds to marginal maximum likelihood optimization such as done by VAES \parencite{moreno-munoz_masked_2023}.
+Since maximum likelihood in no way guarantees good representations, how do self"=supervised methods achieve  this, if they also do maximum likelihood?}
+
+\todo[inline]{Out-of-distribution detection on speech?}
 

### 2023-10-18 chapters/introduction.tex
Update introduction and discussion

@@
 Assumedly unbeknownst to Kushim, this tablet would also come to contain one of the world's first recorded examples of a mathematical error; the statement: $5 \times \dfrac{1/2} = 5$ \parencite{nissen_archaic_1993}.
-The Plimpton 322 clay tablet (\cref{fig:plimpton_332}) contains several errors in its list of Pythagorean triplets, presumably made by a novice in Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
+Later, in its list of Pythagorean triplets, the Plimpton 322 clay tablet (\cref{fig:plimpton_332}) also has several errors, presumably made by a novice in Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
 % Apart from such inherent limitations of the tools, inadequate human proficiency in their use was also a source of error. 

### 2023-10-18 chapters/introduction.tex
Fix fraction

@@
 In a likely attempt to reduce the effort needed to do Sumerian mathematics, Kushim produced a clay tablet with a multiplication table including fractional representations of common numbers.
-Assumedly unbeknownst to Kushim, this tablet would also come to contain one of the world's first recorded examples of a mathematical error; the statement: $5 \times \dfrac{1/2} = 5$ \parencite{nissen_archaic_1993}.
+Assumedly unbeknownst to Kushim, this tablet would also come to contain one of the world's first recorded examples of a mathematical error; the statement: $5 \times \frac{1}{2} = 5$ \parencite{nissen_archaic_1993}.
 Later, in its list of Pythagorean triplets, the Plimpton 322 clay tablet (\cref{fig:plimpton_332}) also has several errors, presumably made by a novice in Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.

### 2023-10-19 appendices/supplementary_retrospective.tex
Update introduction

@@
 This allowed us to test the significance of the observed pairings on the chosen statistics, i.e.
-whether a significant difference was observed in the test statistics depending on whether the call taker or model made the prediction.
+whether a significant difference was observed in the test statistics depending on whether the call-taker or model made the prediction.
 We used this approach to test whether
@@
 \begin{itemize}
-    \item the model performed better than the call takers on the 2021 test set,
+    \item the model performed better than the call-takers on the 2021 test set,
     \item including 112 training data improved the model's performance on the 2021 test set.
@@
 This approach allowed us to test the significance of the observed sample assignments, i.e.
-whether there was a significant difference in test statistics depending on whether the predictions were assigned to the model or call taker.
+whether there was a significant difference in test statistics depending on whether the predictions were assigned to the model or call-taker.
 We used this approach to test whether.
@@
     \item the model performed better on the 65+ group than on the 18-64 group on the 2021 test set,
-    \item the call taker performed better on men than women on the 2021 test set,
-    \item the call taker performed better on the 65+ group than on the 18-64 group on the 2021 test set.
+    \item the call-taker performed better on men than women on the 2021 test set,
+    \item the call-taker performed better on the 65+ group than on the 18-64 group on the 2021 test set.
 \end{itemize}
@@
 
-We computed \emph{bootstrapped confidence intervals} for the statistics by resampling (with replacement) the predictions made by model or call taker on a relevant subgroup and recomputing the relevant statistics for each bootstrap sample.
+We computed \emph{bootstrapped confidence intervals} for the statistics by resampling (with replacement) the predictions made by model or call-taker on a relevant subgroup and recomputing the relevant statistics for each bootstrap sample.
 This process established a bootstrap distribution of the statistic that was then used to estimate the standard error and compute the confidence intervals.
@@
 \begin{enumerate}[label=\Roman*.]
-    \item When is the model prediction presented to the call taker?
-    \item How does prediction influence the diagnostic code the call taker assigns to the call?
+    \item When is the model prediction presented to the call-taker?
+    \item How does prediction influence the diagnostic code the call-taker assigns to the call?
 \end{enumerate}
@@
 \begin{enumerate}
-    \item Notify the call taker of potential false positive or negative stroke cases after the call ends.
-    \item Notify the call taker of potential false positive or negative stroke cases during the call.
+    \item Notify the call-taker of potential false positive or negative stroke cases after the call ends.
+    \item Notify the call-taker of potential false positive or negative stroke cases during the call.
 \end{enumerate}
@@
 
-As we can only assume how call takers are influenced by model predictions (II), precisely evaluating the hypothetical performance of call takers when supported by a machine learning framework is impossible.
+As we can only assume how call-takers are influenced by model predictions (II), precisely evaluating the hypothetical performance of call-takers when supported by a machine learning framework is impossible.
 Furthermore, option 2 may influence the conversation, further complicating matters.
-Therefore, we report the results combining the call taker and the model under the following two assumptions:
+Therefore, we report the results combining the call-taker and the model under the following two assumptions:
 %
 \begin{enumerate}[label=\Alph*.]
-    \item Call-takers change any stroke prediction from negative to positive if the model predicts a positive (call takers mirror model positives).
-    \item Call-takers change any stroke prediction from positive to negative if the model predicts a negative (call takers mirror model negatives).
+    \item Call-takers change any stroke prediction from negative to positive if the model predicts a positive (call-takers mirror model positives).
+    \item Call-takers change any stroke prediction from positive to negative if the model predicts a negative (call-takers mirror model negatives).
 \end{enumerate}
@@
 We also report the results of the model itself (C).
-This method corresponds to call takers mirroring the model predictions exactly.
-This is not feasible in practice, although technically possible, because the conversation and instructions given to patients may conflict with the actions taken by the call taker after hanging up.
+This method corresponds to call-takers mirroring the model predictions exactly.
+This is not feasible in practice, although technically possible, because the conversation and instructions given to patients may conflict with the actions taken by the call-taker after hanging up.
 Method 1.C is identical to the method employed in the main text, and we have copied the same results here for easier comparison.
@@
 As expected (\cref{tab_retrospective:tableA9}), method 2.C (raw model predictions during calls) yielded slightly worse results than 1.C (raw model predictions after calls).
-Compared with method C, method A (call takers mirror model positives) led to increased sensitivity and decreased PPV, whereas method B (call takers mirror model negatives) led to decreased sensitivity and increased PPV, as expected.
-The numerical changes compared with method C are quite large because, in our simulation, the call taker is assumed to strictly follow methods A or B without divergence.
+Compared with method C, method A (call-takers mirror model positives) led to increased sensitivity and decreased PPV, whereas method B (call-takers mirror model negatives) led to decreased sensitivity and increased PPV, as expected.
+The numerical changes compared with method C are quite large because, in our simulation, the call-taker is assumed to strictly follow methods A or B without divergence.
 
@@
     \centering
-    \caption{Overall performance of model, call takers and simulated combinations of model and call takers on MH-1813 test data.}
+    \caption{Overall performance of model, call-takers and simulated combinations of model and call-takers on MH-1813 test data.}
     \label{tab_retrospective:tableA9}
@@
 
-        Predictor & Call taker & \multicolumn{2}{c|}{Model} & \multicolumn{4}{c}{Call taker supported by the model (simulated)} \\
+        Predictor & Call-taker & \multicolumn{2}{c|}{Model} & \multicolumn{4}{c}{Call-taker supported by the model (simulated)} \\
         \midrule
@@
 
-Method 1.A (call takers mirror model positives after a call) yields a better F1-score, sensitivity, PPV, and FOR than call takers alone, although at the cost of a slightly higher FPR.
-This stands in contrast to methods 1.B, 2.A, and 2.B where either the sensitivity or the PPV is worse for the combined system than for call takers alone.
-Regardless, the F1-score (harmonic mean of sensitivity and PPV) is higher for all methods of combining call takers and model (1.A through 2.C).
+Method 1.A (call-takers mirror model positives after a call) yields a better F1-score, sensitivity, PPV, and FOR than call-takers alone, although at the cost of a slightly higher FPR.
+This stands in contrast to methods 1.B, 2.A, and 2.B where either the sensitivity or the PPV is worse for the combined system than for call-takers alone.
+Regardless, the F1-score (harmonic mean of sensitivity and PPV) is higher for all methods of combining call-takers and model (1.A through 2.C).
 

### 2023-10-19 chapters/background.tex
Update introduction

@@
 This is the same distinction made by \textcite{graham_denoising_2023} and \textcite{liu_unsupervised_2023a}.
-    It is important to note, however, that in other works, the distinction is made based on whether the model is trained on OOD data or not \parencite{hendrycks_baseline_2017,liu_energy-based_2020}.
+    It is important to note, however, that in other works, the distinction is made based on whether the model is trained on OOD data or not \parencite{hendrycks_baseline_2017,liu_energybased_2020}.
 This difference in nomenclature is currently unresolved in the literature.
@@
 from the \emph{probabilities} given by the predictive distribution $p(y|\xb)$ \parencite{hendrycks_baseline_2017,hendrycks_scaling_2022}, 
-from the \emph{logits} of $p(y|\xb)$ \parencite{hendrycks_scaling_2022,liu_energy-based_2020}, or 
+from the \emph{logits} of $p(y|\xb)$ \parencite{hendrycks_scaling_2022,liu_energybased_2020}, or 
 use a latent \emph{representation} of the input \parencite{lee_simple_2018,li_anomaly_2019,ndiour_outofdistribution_2020,cook_outlier_2020,zaeemzadeh_outofdistribution_2021}, similar to unsupervised methods.
@@
 % No OOD data, logits
-Although some works use the maximum softmax probability as a score for OOD detection \parencite{hendrycks_deep_2019,ren_likelihood_2019}, several works have noted that it is not generally reliable \parencite{hendrycks_scaling_2022,liu_energy-based_2020}.
-\textcite{liu_energy-based_2020} make an interesting argument as to why based on the energy $E(\xb; f)$ of a softmax classifier $f(\xb)$ \parencite{lecun_tutorial_2006},
+Although some works use the maximum softmax probability as a score for OOD detection \parencite{hendrycks_deep_2019,ren_likelihood_2019}, several works have noted that it is not generally reliable \parencite{hendrycks_scaling_2022,liu_energybased_2020}.
+\textcite{liu_energybased_2020} make an interesting argument as to why based on the energy $E(\xb; f)$ of a softmax classifier $f(\xb)$ \parencite{lecun_tutorial_2006},
 %

### 2023-10-19 chapters/discussion.tex
Update introduction

@@
 
-Two key observations relating to the training of VAEs might lead to the more pessimistic view that VAEs are fundamentally unsuited for representation learning. 
-First, variational autoencoders are trained to maximize the ELBO which is a marginal likelihood objective that is conditionally independent of the latent variables $\zb$.
-Second, as we saw in \cref{eq:variational-autoencoder-elbo-factorized-expected}, the ELBO contains a hidden
+Two key observations might lead to the more pessimistic view that VAEs are fundamentally unsuited for representation learning. 
+First, variational autoencoders are trained to maximize the ELBO which is a marginal likelihood objective conditionally independent of the latent variables $\zb$.
+Second, as we saw in \cref{eq:variational-autoencoder-elbo-factorized-expected}, the ELBO contains can be written to reveal that its maximization leads to minimization of the mutual information between the observed and latent variables.
+In the following we will discuss each of these observations in turn.
 

### 2023-10-19 chapters/introduction.tex
Update introduction

@@
 
-In an era where technology increasingly intertwines with healthcare, artificial intelligence is on the verge of becoming part of standard medical practice. 
+In an era where technology increasingly intertwines with healthcare, artificial intelligence is on the verge of becoming part of standard medical practice \parencite{shailaja_machine_2018}. 
 Growing administrative burdens, aging populations, and rapid scientific progress in medicine and medical devices have created a need to rethink how medical professionals are best enabled to successfully do their job.
 The application of artificial intelligence in medical decision-making and particularly for diagnosing and recommending treatments is likely to hold significant potential to reduce the effect of these issues and improve patient outcomes.
-However, as such decisions impact critical aspects of human health and well-being, this prospect also raises concerns related to the reliability of systems that use artificial intelligence \parencite{chen_ethical_2021, shailaja_machine_2018, ahmad_interpretable_2018}.
+However, as such decisions impact critical aspects of human health and well-being, this prospect also raises concerns related to the reliability of systems that use artificial intelligence \parencite{chen_ethical_2021, ahmad_interpretable_2018}.
 
-The evolution of technology to support decision-making can be traced back to very beginning of recorded history.
-After the agricultural revolution, several ancient civilizations developed mathematical techniques and algorithms to help them manage the newfound complexity of pre-modern society.
-Fittingly, the earliest human whose name is known to recorded history was a Sumerian accountant called Kushim.
-Their signature has been identified on at least 18 clay tablets that deal with transactions and inventory of barley and other ingredients for beer production \parencite{nissen_archaic_1993}.
+The use of technology to support decision-making can be traced back to very beginning of recorded history.
+After the agricultural revolution, several ancient civilizations developed mathematical techniques and algorithms to help them manage the newfound complexity of society.
+Fittingly, the earliest human whose name is still known today was a Sumerian accountant called Kushim.
+Their signature has been identified on at least 18 clay tablets dated to 3400-3000 BCE that deal with transactions and inventory of barley and other ingredients for beer production \parencite{nissen_archaic_1993}.
 Another example is the Egyptian Rhind Mathematical Papyrus dated to around 1550 BCE which describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \parencite{georges_universal_2001}.
@@
 Besides enabling the advancement of early society, these technologies also brought with them novel challenges. 
-In a likely attempt to reduce the effort needed to do Sumerian mathematics, Kushim produced a clay tablet with a multiplication table including fractional representations of common numbers.
+In a likely attempt to ease the strenuous work of doing Sumerian mathematics, Kushim produced a clay tablet with a multiplication table that included fractional representations of common numbers.
 Assumedly unbeknownst to Kushim, this tablet would also come to contain one of the world's first recorded examples of a mathematical error; the statement: $5 \times \frac{1}{2} = 5$ \parencite{nissen_archaic_1993}.
@@
 % Apart from such inherent limitations of the tools, inadequate human proficiency in their use was also a source of error. 
-Besides these human errors, inherent limitations of the tools also posed problems. 
-Babylonian mathematicians did not formalize the difference between rational and irrational numbers which lead to approximation errors.
+Besides these human errors, limitations inherent to the tools themselves also posed problems. 
+Babylonian mathematicians for instance did not formalize the difference between rational and irrational numbers, which lead to approximation errors.
 For instance, the common practice of approximating $\pi$ with 3 \parencite{georges_universal_2001}.
@@
 
-With the digital revolution continuing into the 21st century and the worldwide spread of the internet, the amount of data available to decision-makers has grown exponentially. 
-This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that focuses on developing computer algorithms that can learn from data to recognize patterns and make predictions that guide decision-making.
-Enabled by a massive increase in the performance and availability of parallel computing resources, such methods have set new standards for the abilities of computer systems in various domains, such as speech recognition, computer vision, and natural language processing \parencite{lecun_deep_2015}.
-In recent years, machine learning has made its way into the healthcare sector, where it has shown promising results in tasks such as medical imaging \parencite{lundervold_overview_2019}, drug discovery \parencite{chen_rise_2018}, and clinical decision support \parencite{cite15, cite14}.
+With the digital revolution continuing into the 21st century and the worldwide spread of the internet, the amount of data and its availability have grown exponentially. 
+This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence.
+Machine learning is the study of computer algorithms that can learn functions from data and make predictions to guide decision-making.
+Enabled by a massive increase in the performance and availability of parallel computing resources, such methods have set new standards for the abilities of computer systems at various tasks usually reserved for humans, such as vision, speech recognition, and natural language understanding \parencite{lecun_deep_2015}.
+In recent years, machine learning has also made its way into the healthcare sector, where it has shown promising results on tasks within medical imaging \parencite{lundervold_overview_2019}, drug discovery \parencite{chen_rise_2018}, and clinical decision support \parencite{cite15, cite14}.
 
-\begin{figure}
+\begin{figure}[t]
     \centering
@@
     \caption[A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.]{ A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
-The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \parencite[photo credit][]{nationaltransportationsafetyboardnhsa_collision_2019}.}
+The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \parencite[photo credit]{nationaltransportationsafetyboardnhsa_collision_2019}.}
     % \caption{Aerial view of crash location showing path of pedestrian as she attempted to cross N.
@@
 
-However, these technologies come with the same risks as previous generations did, and in some cases, may amplify them. 
-Limitations of the technology itself, human error, and the complexity of the real world are all factors that can lead to failures.
-A pertinent example is the use of machine learning in autonomous driving systems:
+% While machine learning has immense potential for positive impact, as any of it also comes with risks of error, much the same as its predecessors.
+
+% While applications in healthcare are `nascent' and generally move forward slowly, machine learning has been fast adopted for self-driving vehicles.
+
+% As with other technologies that came before it, limitations of the machine learning itself, human error, and the complexity of the real world are all factors that can lead to failures. 
+% However, compared to earlier, the sheer number of potential applications of machine learning, and the weight of some of them, serve to amplify both potential benefits and risks.
+
+% As for all technologies that came before it, limitations of machine learning algorithms, the complexity of the real world, and human error can all lead to failures. 
+% As for all technologies that came before it, machine learning systems can result in error, whether due to the complexity of the real world, human error, or limitations inherent itself.
+
+% However, the vast array of applications and their large potential impact amplifies these effects, compared to earlier technologies.
+
+% Similar to technologies that came before it, machine learning brings risks alongside its benefits. 
+% % The complexity of the real world, limitations inherent to machine learning itself, and human mistakes can all lead to errors, while the nature of some applications serves to amplify their consequences, compared to some earlier technologies.
+% % Real-world complexity, limitations of machine learning itself, and human error can all lead to system errors whose consequences might be amplified compared to earlier technologies due to the nature of some applications, such as in healthcare or autonomous driving.
+% % Real-world complexity, machine learning limitations, and human errors can magnify risks in fields like healthcare and autonomous driving
+% Risks, as well as benefits, might be amplified in machine learning compared to earlier technologies due to the nature of some of the tasks that machine learning enables augmenting.
+% While applications in healthcare are nascent and generally move forward in small steps, autonomous driving has fast adopted machine learning and presents a pertinent example:
+
+Machine learning, like previous technologies, carries both benefits and risks. 
+However, some of these risks may be heightened in machine learning due to the unique tasks it enables in domains such as healthcare and autonomous driving. 
+While applications in healthcare are nascent and generally move forward in small steps, autonomous driving has fast adopted machine learning and presents a pertinent example: 
+% A pertinent example is autonomous driving systems which were fast equipped with machine learning technology: 
 In 2018, the first incident of a self-driving car killing a pedestrian took place in Tempe, Arizona. 
@@
 % revealed far fewer crashes, with only 2 of 20 incidents over 1 million miles meeting reporting criteria, it is interesting to note that the cars are all operated in the sunny states of California and Arizona, where driving conditions vary relatively little \parencite{hawkins_waymo_2023}.
-While driver-assistance systems undoubtedly help reduce accidents in many cases, and has great future potential, the issues of current systems highlight the need for machine learning systems that are robust in rare, adverse or difficult situations.
+While driver-assistance systems undoubtedly help reduce accidents in many cases, and has great future potential, the issues of current systems highlight the need for machine learning systems that are robust in rare, adverse, and difficult situations.
 
-Recently, European policymakers proposed a set of requirements to ensure that machine learning applications are safe, reliable, and trustworthy \parencite{europeancommission_briefing_2021}. 
-This AI Act groups applications into three categories based on the potential risk they pose to society: limited, high, and unacceptable.
-Limited risk applications, such as in computer games, must meet transparency requirements, and technologies such as public facial recognition are banned due to unacceptable risk.
-High risk applications include autonomous vehicles and systems within healthcare. To avoid critical errors, these must meet a set of strict requirements focusing on anti-discrimination and robustness.
+Recently, European policymakers proposed legislation to enforce that machine learning applications live up to such standards \parencite{europeancommission_briefing_2021}. 
+This so-called AI Act considers three categories of applications based on the potential risk they pose to society: limited, high, and unacceptable. 
+Limited risk applications such as in computer games must meet transparency criteria, while autonomous vehicles and systems within healthcare fall in the category of high risk applications.
+To avoid critical errors, these must meet a set of strict requirements focusing on anti-discrimination and robustness.
+Facial recognition is an example of an application deemed to pose unacceptable risk to society.
+% Limited risk applications, such as in computer games, must meet transparency requirements, and technologies such as public facial recognition are banned due to unacceptable risk.
 
-One of the key ways to improve the safety, reliability and trustworthiness of machine learning systems is to equip them with the ability accurately estimate the uncertainty of their predictions. 
-While any model will inevitably make a wrong prediction, by accurately estimating the uncertainty it can communicate to other systems or humans when it is not to be trusted, and so, reduce the risk that wrong predictions lead to catastrophic failures.
+% Miscalibrated models are unreliable, especially in high risk applications, and although no model is perfect, the inability of a human, or a model, to assess the certainty with which a prediction was made, makes it impossible to know how to weigh it in a decision-making process.
+% A connection can be made back to the incident with the Uber self-driving car.
+
+One of the key advances needed to ensure that high-risk applications of machine learning comply with such requirements is accurate uncertainty estimation of their predictions. 
+The multitude of rapid misclassifications made by the Uber self-driving car in Tempe is a likely indication that the model responsible for object-classification was overconfident in its predictions and unable to represent or communicate uncertainty to other systems in the car, or the safety driver.
+As a consequence, the model's many incoherent predictions were taken at face value and subsystems in the car acted on them without appropriate consideration to their reliability.
+While any model will inevitably make a wrong prediction, accurately estimating the associated uncertainty and communicating it to other systems or humans is necessary to reduce the risk that wrong predictions lead to catastrophic failures.
 Fittingly, a recent study by the European Parliamentary Research Services \parencite{europeanparliament_artificial_2022} concluded: 
@@
 The research project was partially funded by a grant from the national Danish Innovation Fund (grant no.\@ 0153-00167B) and defined and carried out in collaboration with the Danish company Corti, which develops computer software for medical decision support.
-In the rest of this chapter, we will provide a high-level introduction to these topics and discuss the motivation behind the research project.
-
-% \todo[inline]{Can we find a better example than the Rhind Papyrus, or an additional one?
-It could be more on-point with decision-making and more well-known.
-Ideally, weave in an example of a failure of such technology.}
+In the rest of this chapter, we will provide a high-level introduction to these topics and to Corti and discuss the motivation behind the research project.
 
@@
 The classifier might misclassify a conversation due to medically ambiguous symptoms, a general lack of information given in the conversation, or transcription errors made by the speech recognizer. 
-How to best make such models accurately represent and estimate the uncertainty of given predictions, and how to use it to improve the value of such a system, especially in cooperation with the call taker, is an open question.
+How to best make such models accurately represent and estimate the uncertainty of given predictions, and how to use it to improve the value of such a system, especially in cooperation with the call-taker, is an open question.
 What is clear, however, is that the ability of such a system to accurately estimate the uncertainty of given predictions is crucial for its safe and reliable deployment in the real world.
@@
 
-Miscalibrated models are unreliable, especially in high risk applications, and although no model is perfect, the inability of a human, or a model, to assess the certainty with which a prediction was made, makes it impossible to know how to weigh it in a decision-making process.
-% A connection can be made back to the incident with the Uber self-driving car. 
-The multitude of rapid misclassifications made by the Uber self-driving car in Tempe is likely an indication that the model responsible for object-classification was overconfident in its predictions and unable to represent or communicate uncertainty to other systems in the car, or the safety driver.
+% Miscalibrated models are unreliable, especially in high risk applications, and although no model is perfect, the inability of a human, or a model, to assess the certainty with which a prediction was made, makes it impossible to know how to weigh it in a decision-making process.
+% % A connection can be made back to the incident with the Uber self-driving car. 
+% The multitude of rapid misclassifications made by the Uber self-driving car in Tempe is likely an indication that the model responsible for object-classification was overconfident in its predictions and unable to represent or communicate uncertainty to other systems in the car, or the safety driver.
 Instead, the model's predictions were taken at face value and subsystems in the car acted on them without appropriate consideration to their reliability.
@@
 
-% In practice, we are interested in the predictive uncertainty, so we need to include both aleatoric and epistemic parts in the uncertainty estimate. 
-% Since sources of epistemic uncertainty are by definition those not represented in the data, it is not generally possible to model them directly, as it is for aleatoric uncertainty.
-This makes epistemic uncertainty more difficult to model and quantify than aleatoric uncertainty.
-% % Methods for estimating epistemic uncertainty include Bayesian neural networks \parencite{mackay_practical_1992, neal_bayesian_1995}, ensembling \parencite{gal_dropout_2016,lakshminarayanan_simple_2017} and a suite of recent research into out-of-distribution detection (see \cref{sec:out-of-distribution-detection}).
-% In contrast to aleatoric uncertainty, which is usually represented by a distribution over the data, epistemic uncertainty is often represented by an implicit or explicit distribution over the model parameters, such as in Bayesian neural networks \parencite{mackay_practical_1992, neal_bayesian_1995} and ensemble models \parencite{gal_dropout_2016,lakshminarayanan_simple_2017}.
-In most cases, such methods require multiple forward passes through the model to estimate the predictive uncertainty making them computationally expensive and impractical for real-time applications.
-Within the field of out-of-distribution detection, the epistemic uncertainty is generally not expressed in model parameters, but e.g.
-by a distribution over data representations, output probabilities with special classes for out-of-distribution data, or by distances in feature spaces.
-In such cases, the uncertainty can often be estimated in a single forward pass through the model.
-
 \section{Thesis scope and outline}

### 2023-10-19 chapters/main-contributions.tex
Update introduction

@@
 We shall return to this in the discussion of \cref{chp:paper-review} in \cref{sec:discussion-paper-review}.
+\todo[inline]{Edit this if we did not discuss this.}
 An early version of this review paper included a review of variational autoencoders for speech representation learning and a comparison between the two modelling paradigms \parencite{borgholt_brief_2022}.

### 2023-10-19 chapters/paper_hierarchical.tex
Update introduction

@@
 
-We also note that since the hierarchical VAE is a probabilistic graphical latent variable model, it lends itself very naturally to manipulation at the feature level \parencite{kingma_semi-supervised_2014, maaloe_auxiliary_2016, maaloe_semi-supervised_2017}.
+We also note that since the hierarchical VAE is a probabilistic graphical latent variable model, it lends itself very naturally to manipulation at the feature level \parencite{kingma_semi-supervised_2014, maaloe_auxiliary_2016, maaloe_semisupervised_2017}.
 This property sets it apart from other generative models that do not explicitly define such a hierarchy of features.

### 2023-10-19 chapters/paper_retrospective.tex
Fix many spelling and language errors
Update introduction

@@
     \centering
-    \caption[Overall stroke recognition performance of model compared to call takers.]{Overall performance on MH-1813 test data, performance without 1-1-2 training data, and performance on data from 2021 without diagnostic categories as well as performance on MH-1813 test data based on demographic subgroups (age/sex) [mean (95\% CI)].}
+    \caption[Overall stroke recognition performance of model compared to call-takers.]{Overall performance on MH-1813 test data, performance without 1-1-2 training data, and performance on data from 2021 without diagnostic categories as well as performance on MH-1813 test data based on demographic subgroups (age/sex) [mean (95\% CI)].}
     \label{tab_retrospective:table2-main-results}
@@
     \caption[Prediction confusion matrices for stroke recognition.]{Prediction confusion matrices.
-Confusion matrices of predictions for call takers and the model on the test set.
+Confusion matrices of predictions for call-takers and the model on the test set.
 Numbers for the model are given as the rounded mean over eleven runs.}
@@
 
-\section{Data availability}
+\section*{Data availability}
 
@@
 
-\section{Code availability}
+\section*{Code availability}
 
@@
 
-\section{Declaration of interests}
+\section*{Declaration of interests}
 

### 2023-10-19 chapters/paper_review/04b_analysis.tex
Update introduction

@@
 
-A similar insight was obtained in \parencite{zhou_comparison_2020}, which compared VQ-VAE and
+A similar insight was obtained in \parencite{zhou_comparison_2021}, which compared VQ-VAE and
 vq-wav2vec with respect to their ability to discover phonetic units.

### 2023-10-22 chapters/background.tex
Work on discussion, background and introduction

@@
 \begin{equation} \label{eq:kullback-leibler-divergence}
-    D_{KL}(p(x)\parallel q(x)) = \int_{\mathcal{X}} p(x) \log \left(\frac{p(x)}{q(x)}\right) \enspace .
+    D_{\mathrm{KL}}(p(x)\parallel q(x)) = \int_{\mathcal{X}} p(x) \log \left(\frac{p(x)}{q(x)}\right) \enspace .
 \end{equation}
 %
-The Kullback-Leibler divergence $D_{KL}(p(x)\parallel q(x))$, also known as the \emph{information gain}, is often interpreted as the amount of additional information required to represent events from a distribution with density $p(x)$ using a code optimized for a distribution with density $q(x)$.
+The Kullback-Leibler divergence $D_{\mathrm{KL}}(p(x)\parallel q(x))$, also known as the \emph{information gain}, is often interpreted as the amount of additional information required to represent events from a distribution with density $p(x)$ using a code optimized for a distribution with density $q(x)$.
 In other words, it measures how surprised one would be if they used distribution $q$ to represent events from distribution $p$, i.e.
@@
 % \paragraph{Deep generative models} 
-VAEs belong to the class of deep generative models which are models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution. 
+Variational autoencoders (VAEs) belong to the class of deep generative models.
+These are models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution.
 There exist at least six main classes of deep generative models: variational autoencoders \parencite{kingma_autoencoding_2014,rezende_stochastic_2014,ranganath_hierarchical_2016,vahdat_nvae_2020,child_very_2021}, normalizing flow models \parencite{dinh_nice_2015,rezende_variational_2015,dinh_density_2017,kingma_glow_2018,grathwohl_ffjord_2018}, diffusion models \parencite{sohl-dickstein_deep_2015,song_generative_2019, ho_denoising_2020, vahdat_scorebased_2021}, energy-based models \parencite{lecun_tutorial_2006,hinton_fast_2006,salakhutdinov_efficient_2010,du_implicit_2019}, autoregressive models \parencite{oord_conditional_2016,oord_wavenet_2016,radford_improving_2018}, and generative adversarial networks \parencite{goodfellow_generative_2014,arjovsky_wasserstein_2017,brock_large_2019,karras_stylebased_2019}.
@@
 % A variational autoencoder (VAE) \parencite{kingma_autoencoding_2014,rezende_stochastic_2014} is a type of deep generative model and an instance of a latent variable model. 
-In the following, we will introduce VAEs starting from the definition of the group of latent variable models to which they belong.
+In the following, we will introduce VAEs starting from the definition of latent variable models.
 We then derive the marginal likelihood objective used to train VAEs, discuss its different components, and highlight some central challenges facing VAEs.
@@
 
-The marginal distribution $p(\xb)$ involves an integral over all possible values of the latent variable $\zb$ that is often analytically intractable.
+The marginal distribution $p(\xb)$ involves an integral over all possible values of the latent variable $\zb$ and is often analytically intractable.
 This is the case when the latent variable space is high-dimensional or the conditional distribution $p(\xb|\zb)$ is complex, as with non-conjugate models or neural networks.
@@
 
-\subsection{Variational autoencoders} \label{sec:variational-autoencoders}
-To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to parameterize the generative model $p_\thetab(\xb,\zb)$ using a neural network with parameters $\thetab$.
-Specifically, they let $p(\xb,\zb) = p_\thetab(\xb|\zb)p(\zb)$ and use a neural network for the generative model $p_\thetab(\xb|\zb)$ and any distribution that can be efficiently sampled and evaluated for the prior $p(\zb)$.
+\subsection{Variational inference} \label{sec:variational-autoencoders}
+To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to factor the joint distribution and parameterize it using a neural network with parameters $\thetab$.
+Specifically, they let $p_\thetab(\xb,\zb) = p_\thetab(\xb|\zb)p(\zb)$ where $p(\zb)$ is the prior.
 This choice makes both the marginal and posterior distributions intractable and so, training and inference require approximate methods. 
@@
 \paragraph{Evidence lower bound}
-VAEs amortize the cost of variational inference by parameterizing $q_\phib(\zb|\xb)$ as a single neural network which makes it possible to efficiently compute approximate posterior distributions for any input $\xb$.
-The approximate posterior can also be used to compute the so-called evidence lower bound (ELBO) on the marginal distribution $p(\xb)$,
+VAEs amortize the cost of variational inference by parameterizing $q_\phib(\zb|\xb)$ as a single neural network.
+This makes it possible to efficiently compute approximate posterior distributions for any input $\xb$.
+The approximate posterior can also be used to compute a lower bound on the marginal distribution $p(\xb)$, the so-called evidence lower bound (ELBO),
 %
@@
 %
-Although analytically evaluating the expectation is intractable, it can be approximated by sampling from the variational approximation $q_\phib(\zb|\xb)$. 
-Since both the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ are parameterized by neural networks that can be efficiently evaluated, and the prior $p(\zb)$ is simple, this makes it possible to compute the ELBO for any input $\xb$.
+Although analytically evaluating the ELBO is intractable, it can be approximated by sampling and used to train VAEs. We will return to this after examining the ELBO a bit.
 
@@
 This reveals a commonly used interpretation of the ELBO: 
-The expected log-likelihood of the data under the generative model $p_\thetab(\xb|\zb)$ acts as a reconstruction loss encouraging the model to make $q_\phib(\zb|\xb)$ as peaky as possible to make reconstruction easy.
-The negative KL-divergence between $q_\phib(\zb|\xb)$ and the prior $p(\zb)$ acts as a regularizer that forces the approximate posterior to be within the support of the prior (reverse KL-divergence). 
-A well-fitted VAE will have a large expected log-likelihood of the data under the generative model enabled by an informative latent variable.
-However, an optimal KL-divergence of zero entails that $q_\phib(\zb|\xb) = p(\zb)$ which means that $\zb$ is independent of $\xb$.
-This local optimum is referred to as posterior collapse and is particularly prone to happen for strong generative models, e.g.
-with autoregressive dependencies $p(\xb_{t}|\xb_{1:t},\zb)$.
+The expected log-likelihood of the data under the generative model $p_\thetab(\xb|\zb)$ acts as a reconstruction loss that encourages the model to make $q_\phib(\zb|\xb)$ as peaky as possible to make reconstruction easy.
+The negative KL-divergence between the posterior $q_\phib(\zb|\xb)$ and the prior $p(\zb)$ acts as a regularizer that forces the approximate posterior to be within the support of the prior (reverse KL-divergence).
+This helps prevent overfitting but importantly also makes ancestral sampling from the prior produce high quality data samples.
+
+By maximizing $\mathcal{L}_{\thetab,\phib}(\xb)$, we must minimize the KL-divergence.
+However, an optimal KL-divergence of zero requires $q_\phib(\zb|\xb) = p(\zb)$ which would mean that $\zb$ has become independent of $\xb$.
+This state is a poor local optimum and is usually referred to as posterior collapse.
+It is particularly prone to happen for strong generative models for instance with autoregressive dependencies $p(\xb_{t}|\xb_{1:t},\zb)$.
 Some works have proposed to mitigate posterior collapse by tempering the KL-divergence term \parencite{alemi_fixing_2018,higgins_vvae_2017} or by adding additional terms to the ELBO \parencite{zhao_infovae_2018}.
+A well-fitted VAE will therefore have maximized the expected log-likelihood of the data under the generative model $p_\thetab(\xb|\zb)$ by learning an informative latent variable and thus have a nonzero KL-divergence $D_\text{KL}\left( q_\phib(\zb|\xb) \,||\, p(\zb) \right)$.
+
+\subsection{Maximum likelihood estimation}
+To train a VAE by maximum likelihood, we must approximate the expectation in the ELBO \cref{eq:variational-autoencoder-elbo} and compute its gradients w.r.t. $\phib$ and $\thetab$.
 
 \paragraph{Sampling}
-For some choices of prior and approximate posterior, the KL-divergence term can be computed analytically, but we generally must estimate the reconstruction loss by sampling from the approximate posterior.
+Since we have freedom to select the approximate posterior, we can choose it such that it is easy to sample. 
+We can then form a Monte Carlo estimate of the ELBO as
+% 
+\begin{equation} \label{eq:variational-autoencoder-elbo-sampling-expanded}
+    \mathcal{L}_{\thetab,\phib}(\xb)
+    = \frac{1}{K} \sum_{k=1}^K \log p_\thetab(\xb|\zb_k) + p(\zb_k) - \log q_\phib(\zb_k|\xb) \enspace , 
+\end{equation}
+%
+where $\zb_k \sim q_\phib(\zb|\xb)$ are samples from the approximate posterior. 
+For some choices of prior and approximate posterior, such as Gaussians, the KL-divergence term of the ELBO can be computed analytically. 
 This Monte Carlo estimator can be written as,
 %
-\begin{equation} \label{eq:variational-autoencoder-elbo-sampling}
+\begin{equation} \label{eq:variational-autoencoder-elbo-sampling-KL}
     \mathcal{L}_{\thetab,\phib}(\xb)
@@
 %
-where $\zb_k \sim q_\phib(\zb|\xb)$ are samples from the approximate posterior. 
 To reduce the variance of the estimate, \textcite{burda_importance_2016} propose to use importance sampling to reweigh the samples.
@@
 \paragraph{Reparameterization} 
-Via the ELBO, we want to use maximum likelihood estimation to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$.
+We want to use maximum likelihood estimation to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$. 
 Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
@@
 %
-The most important instantiation of this reparameterization is the diagonal Gaussian case, where $\zb = \mu_\phib(\xb) + \sigma_\phib(\xb) \odot \epsilonb$ with $\epsilonb\sim\mathcal{N}(0,\mathbf{I})$.
-By assuming the covariance matrix is diagonal, this is a mean-field approximation which imposes independence between the latent variables in $\zb$.
+The most important instantiation of this reparameterization is the diagonal Gaussian case, where $g_\phib(\xb,\epsilonb) = \mu_\phib(\xb) + \sigma_\phib(\xb) \odot \epsilonb$ with $\epsilonb\sim\mathcal{N}(0,\mathbf{I})$.
+By assuming the covariance matrix is diagonal, this is a mean-field approximation which imposes independence between the dimensions of $\zb$.
 
@@
 
-\paragraph{Hierarchical models}\label{sec:hierarchical-vae-background}
-The mean-field assumption imposed by letting $\zb \sim\mathcal{N}(\mub,\mathbf{I}\sigmab^2)$ can often be more restrictive for model expressiveness than we would like. 
+\subsection{Hierarchical models}\label{sec:hierarchical-vae-background}
+The mean-field assumption imposed by $\zb \sim\mathcal{N}(\mub,\mathbf{I}\sigmab^2)$ can often be more restrictive for model expressiveness than we would like. 
 While learning a full covariance is a simple solution, it is not always sufficiently computationally efficient and, in any case, enables learning only linear covariance between elements of $\zb$. 
@@
 
-\paragraph{Mutual information interpretation}
-Training by stochastic gradient descent training involves sampling mini-batches of data points $\xb$ from the empirical training data distribution $\hat{p}(\xb)$ and evaluating the gradient of the ELBO with respect to the parameters $\thetab$ and $\phib$ for each mini-batch.
-To gain some additional insight into the objective that we are optimizing, we will exploit this expectation over the data distribution $\hat{p}(\xb)$ to form a factorization similar to \cref{eq:variational-autoencoder-elbo-factorized} but with the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$ in the KL-divergence \parencite{tomczak_trouble_2022}.
-The marginal likelihood under this expectation is lower bounded by the same expectation over the ELBO,
+\subsection{Mutual information interpretation} \label{subsec: mutual-information-interpretation-of-elbo}
+Training VAEs by stochastic gradient descent involves first sampling a mini-batch of data points $\xb$ from the empirical distribution of training data $\hat{p}(\xb)$ and then evaluating the gradient of the ELBO with respect to the parameters $\thetab$ and $\phib$.
+To gain some additional insight into the objective that we are optimizing, we will use this expectation over the data distribution $\hat{p}(\xb)$ to form a factorization similar to \cref{eq:variational-autoencoder-elbo-factorized} but with the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$ in the KL-divergence \parencite{tomczak_trouble_2022}.
+Under the expectation over the data distribution, the marginal likelihood is lower bounded by that expectation over the ELBO,
 %
@@
     \mathbb{E}_{\hat{p}(\xb)}\left[ \log p(\xb) \right] 
-    &\geq \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log \frac{p_{}(\xb|\zb)p(\zb)}{q_{}(\zb|\xb)} \right] \enspace ,
+    &\geq \mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log \frac{p_{\thetab}(\xb|\zb)p(\zb)}{q_{\phib}(\zb|\xb)} \right] \enspace ,
 \end{align}
 %
-where we have defined $q(\xb) = \hat{p}(\xb)$ and $q_{}(\xb,\zb) = q_{}(\zb|\xb)\hat{p}(\xb)$ and dropped the parameter subscripts for ease of notation. 
-This allows us to focus on the KL-divergence in \cref{eq:variational-autoencoder-elbo-factorized} which we can rewrite as follows,
+% where we have defined $q(\xb) = \hat{p}(\xb)$ and $q_{\phib}(\xb,\zb) = q_{\phib}(\zb|\xb)\hat{p}(\xb)$ and dropped the parameter subscripts for ease of notation. 
+where we have defined $q_{\phib}(\xb,\zb) = q_{\phib}(\zb|\xb)\hat{p}(\xb)$ for ease of notation. This is simply \cref{eq:variational-autoencoder-elbo} under the extra expectation. 
+We leave the reconstruction loss in \cref{eq:variational-autoencoder-elbo-factorized} is as and focus on the KL-divergence to the prior. We rewrite it as follows,
 %
 \begin{align} \label{eq:variational-autoencoder-kl-factor-expected}
-    \mathbb{E}_{\hat{p}(\xb)} \left[ D_\text{KL}\left( q_{}(\zb|\xb) \,||\, p(\zb) \right) \right]
-    &= \mathbb{E}_{q_{}(\zb,\xb)} \left[ \log \frac{q_{}(\zb|\xb)}{p(\zb)}\right] \nonumber \\
-    % &= - \mathbb{E}_{q_{}(\zb)}\left[ \log p(\zb) \right] + \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log q_{}(\zb | \xb) \right] \nonumber \\
-    &= \mathbb{E}_{q_{}(\zb)}\left[ \log \frac{1}{p(\zb)} \right] + \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log \frac{q_{}(\xb,\zb)}{q(\xb)} \right] \nonumber \\
-    % &= - \mathbb{E}_{q_{}(\zb)}\left[ \log p(\zb) \right] + \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log \frac{q_{}(\xb,\zb)}{q(\xb)} \right] \nonumber \\ 
-    % &\qquad + \mathbb{E}_{q_{}(\zb)}\left[ \log q_{}(\zb) \right] - \mathbb{E}_{q_{}(\zb)}\left[ \log q_{}(\zb) \right] \nonumber \\
-    &= \mathbb{E}_{q_{}(\zb)}\left[ \log \frac{q_{}(\zb)}{p(\zb)} \right] - \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log \frac{q_{}(\xb,\zb)}{q(\xb)q_{}(\zb)} \right] \nonumber \\
-    &= D_\mathrm{KL}\left( q_{}(\zb) \parallel p(\zb) \right) - I_{q_{}(\xb,\zb)}[\xb;\zb] \enspace .
+    \mathbb{E}_{\hat{p}(\xb)} \left[ D_\text{KL}\left( q_{\phib}(\zb|\xb) \,||\, p(\zb) \right) \right]
+    &= \mathbb{E}_{q_{\phib}(\zb,\xb)} \left[ \log \frac{q_{\phib}(\zb|\xb)}{p(\zb)}\right] \nonumber \\
+    % &= - \mathbb{E}_{q_{\phib}(\zb)}\left[ \log p(\zb) \right] + \mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log q_{\phib}(\zb | \xb) \right] \nonumber \\
+    &= \mathbb{E}_{q_{\phib}(\zb)}\left[ \log \frac{1}{p(\zb)} \right] + \mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log \frac{q_{\phib}(\xb,\zb)}{q(\xb)} \right] \nonumber \\
+    % &= - \mathbb{E}_{q_{\phib}(\zb)}\left[ \log p(\zb) \right] + \mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log \frac{q_{\phib}(\xb,\zb)}{q(\xb)} \right] \nonumber \\ 
+    % &\qquad + \mathbb{E}_{q_{\phib}(\zb)}\left[ \log q_{\phib}(\zb) \right] - \mathbb{E}_{q_{\phib}(\zb)}\left[ \log q_{\phib}(\zb) \right] \nonumber \\
+    &= \mathbb{E}_{q_{\phib}(\zb)}\left[ \log \frac{q_{\phib}(\zb)}{p(\zb)} \right] + \mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log \frac{q_{\phib}(\xb,\zb)}{q(\xb)q_{\phib}(\zb)} \right] \nonumber \\
+    &= D_\mathrm{KL}\left( q_{\phib}(\zb) \parallel p(\zb) \right) + I_{q_{\phib}(\xb,\zb)}[\xb;\zb] \enspace .
 \end{align}
@@
 \begin{equation} \label{eq:variational-autoencoder-elbo-factorized-expected}
-    \mathbb{E}_{q_{}(\xb,\zb)} \left[ \mathcal{L}_{\thetab,\phib}(\xb) \right] = 
+    \mathbb{E}_{\hat{p}(\xb)} \left[ \mathcal{L}_{\thetab,\phib}(\xb) \right] = 
     \underset{\text{average reconstruction}}{\underbrace{
-        \mathbb{E}_{q_{}(\xb,\zb)}\left[ \log p_{}(\xb | \zb) \right]
+        \mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log p_{\thetab}(\xb | \zb) \right]
     }}
@@
     \underset{\text{marginal KL to prior}}{\underbrace{
-    \vphantom{\mathbb{E}_{q_{}(\xb,\zb)}\left[ \log p_{}(\xb | \zb) \right]}{
-        D_{\mathrm{KL}}\left[ q_{}(\zb) \parallel p(\zb) \right]
+    \vphantom{\mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log p_{\thetab}(\xb | \zb) \right]}{
+        D_{\mathrm{KL}}\left[ q_{\phib}(\zb) \parallel p(\zb) \right]
     }
@@
     \underset{\text{mutual information}}{\underbrace{
-    I_{q_{}(\xb,\zb)}\left[\xb ; \zb \right]
+    I_{q_{\phib}(\xb,\zb)}\left[\xb ; \zb \right]
     }} \enspace . \nonumber
@@
 In this form, the KL-divergence term of \cref{eq:variational-autoencoder-elbo-factorized} has been factored into the marginal KL-divergence of the aggregated posterior to the prior and the mutual information between the data and the latent variables.
-The marginal KL-divergence is minimized by making the aggregated posterior $q_{}(\zb)$ match the prior $p(\zb)$.
+The marginal KL-divergence is minimized by making the aggregated posterior $q_{\phib}(\zb)$ match the prior $p(\zb)$.
 Contrary to the KL-divergence term of \cref{eq:variational-autoencoder-elbo-factorized}, driving the marginal KL-divergence to zero does not enforce independence between $\xb$ and $\zb$.
-However, this alternative form reveals that the ELBO is maximized by minimizing the mutual information between $\xb$ and $\zb$, $I_{q_{}(\xb,\zb)}\left[\xb ; \zb \right]$. 
+However, this alternative form reveals that the ELBO is maximized by minimizing the mutual information between $\xb$ and $\zb$, $I_{q_{\phib}(\xb,\zb)}\left[\xb ; \zb \right]$. 
 This result indicates that the usual interpretation of $\zb$ as a representation of $\xb$ is not a fundamental property arising from the ELBO and challenges the usefulness of variational autoencoders as representation learners.

### 2023-10-22 chapters/discussion.tex
Work on discussion, background and introduction

--- 2023-10-15 chapters/discussion.tex
@@
 
-Having focused on VAEs for out-of-distribution detection in \cref{part:unsupervised-uncertainty-estimation} and on self"=supervised methods for representation learning in \cref{part:self-supervised-speech-representation-learning}, this discussion will bridge a gap between the two approaches and consider how VAEs might be enabled to learn representations more competitive with self"=supervised methods.
+Having focused on VAEs for out-of-distribution detection in \cref{part:unsupervised-uncertainty-estimation} and on self"=supervised methods for representation learning in \cref{part:self-supervised-speech-representation-learning}, this discussion will bridge a gap between the two approaches and consider how VAEs might be enabled to learn representations more competitive with self"=supervised methods for downstream tasks.
 Furthermore, since the work in \cref{part:medical-applications} only had limited focus on uncertainty, we will also present and discuss the use of calibration techniques for the stroke recognition model drawing connections to how such systems are perceived and used in practice.
@@
 %
-In this thesis we studied two different approaches to speech representation learning: VAEs in \cref{chp:paper-hierarchical,chp:paper-modelagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-review}.
+In this thesis we studied two different approaches to speech representation learning: VAEs in \cref{chp:paper-hierarchical,chp:paper-mo?delagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-review}.
 As we saw, the probabilistic formulation of VAEs provides benefits for their application to uncertainty quantification, although competitive methods that use self"=supervised foundation models have also been proposed \textcite{xiao_we_2021,hendrycks_using_2019,bergman_classificationbased_2020}.
@@
 For instance, between \cref{tab: phoneme recognition (PER),tab:unsupervisedASR} self"=supervised methods can be seen to outperform VAEs for phoneme recognition while self"=supervised methods are also reported for word-level speech recognition.
-In the following, we will discuss potential directions of future research that might help shed light on why VAEs representations underperform on downstream tasks and how they might be improved.
+In the following, we will discuss potential directions of future research that might help shed light on why representations from VAEs underperform on downstream tasks and how they might be improved.
 
-\paragraph{Designing the latent space} 
+\subsection{Giving up?}
+
+Two key observations might lead to the pessimistic view that VAEs are fundamentally unsuited for representation learning. 
+First, VAEs are trained to maximize the ELBO which is a marginal likelihood objective conditionally independent of the latent variables $\zb$.
+Second, as we saw in \cref{eq:variational-autoencoder-elbo-factorized-expected}, the ELBO contains can be written to reveal that its maximization leads to minimization of the mutual information between the observed and latent variables.
+In the following we will discuss each of these observations.
+
+\paragraph{Marginal likelihood maximization} For any latent variable model $p_{\thetab}(\xb,\zb)$ we can use the posterior distribution $p_{\thetab}(\zb|\xb)$ to map a given observation $\xb$ to its latent representation $\zb$.
+We hope this representation is useful, for instance in the sense that it can improve performance on downstream tasks compared to other methods.
+Essential to the usefulness is the posterior distribution $p_{\thetab}(\zb|\xb)$.
+
+On the other hand, when we use maximum likelihood to train a latent variable model, we maximize the log-marginal likelihood $\log p(\xb)$.
+This can be equivalently phrased as maximizing the KL-divergence of the true data distribution $p_{\text{data}}(\xb)$ to the learned marginal, $D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb))$\footnote{%
+Maximizing the likelihood $\sum_{n=1}^N \log p_\theta(\xb)$ w.r.t. parameters $\thetab$ is asymptotically equivalent to minimizing $D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb))$. 
+%
+\begin{align}
+    \argmin_\thetab D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb)) 
+    &= \argmin_\thetab \mathbb{E}_{p_{\text{data}}(\xb)} \left[ \log \frac{p_{\text{data}}(\xb)}{p_{\theta}(\xb)} \right] \nonumber \\
+    % &= \argmin_\thetab \mathbb{E}_{p_{\text{data}}(\xb)} \left[ \log p_{\text{data}}(\xb) - \log p_{\theta}(\xb) \right] \nonumber \\
+    &= \argmax_\thetab \mathbb{E}_{p_{\text{data}}(\xb)} \left[ \log p_{\theta}(\xb) \right] \nonumber \\
+    &= \argmax_\thetab \lim_{n\rightarrow\infty} \underset{\text{log-likelihood}}{\underbrace{\sum_{n=1}^N \log p_{\theta}(\xb_n)}} \enspace . \nonumber
+\end{align}
+%
+The last equation follows from the law of large numbers.
+% \begin{align}
+%     \mathcal{L}(\thetab; p_{\mathrm{data}}(\xb))
+%     &= \lim_{N\rightarrow\infty} \frac{1}{N} \sum\limits_{n=1}^{N} \log p_{\thetab}(\xb) \\
+%     &= \mathbb{E}_{p_{\mathrm{data}}(\xb)} \left[ \log p_{\thetab}(\xb) \right] \\
+%     &= \mathbb{E}_{p_{\mathrm{data}}(\xb)} \left[ \log p_{\thetab}(\xb) - \log p_{\mathrm{data}}(\xb) + \log p_{\mathrm{data}}(\xb) \right] \\
+%     &= \mathbb{E}_{p_{\mathrm{data}}(\xb)} \left[ \log p_{\mathrm{data}}(\xb) \right] + \mathbb{E}_{p_{\mathrm{data}}(\xb)} \left[ \log\frac{ p_{\thetab}(\xb)}{p_{\mathrm{data}}(\xb) }\right] \\
+%     &= - \mathbb{H}[p_{\mathrm{data}}(\xb)] - D_\mathrm{KL}\left( p_{\mathrm{data}}(\xb)\mid\mid p(\xb) \right)
+% \end{align}
+}. 
+The first key observation then is that the marginal $p_{\thetab}(\xb)$, which we maximize, and the posterior $p_{\thetab}(\zb|\xb)$, which we want to produce useful representations, are separate properties of a latent variable model.
+Any combination uniquely defines a valid latent variable model.
+Hence, maximizing the marginal is a completely useless criteria for representation learning, regardless of the measure of usefulness \parencite{huszar_is_2017}.
+
+\paragraph{Mutual information minimization}
+As we saw in \cref{subsec: mutual-information-interpretation-of-elbo} ...
+
+\begin{equation} \label{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)}
+    \mathbb{E}_{q_{\phib}(\xb,\zb)} \left[ \mathcal{L}_{\thetab,\phib}(\xb) \right] = 
+    \underset{\text{average reconstruction}}{\underbrace{
+        \mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log p_{\thetab}(\xb | \zb) \right]
+    }}
+    - 
+    \underset{\text{marginal KL to prior}}{\underbrace{
+    \vphantom{\mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log p_{\thetab}(\xb | \zb) \right]}{
+        D_{\mathrm{KL}}\left[ q_{\phib}(\zb) \parallel p(\zb) \right]
+    }
+    }} 
+    -
+    \underset{\text{mutual information}}{\underbrace{
+    I_{q_{\phib}(\xb,\zb)}\left[\xb ; \zb \right]
+    }} \enspace . \nonumber
+\end{equation}
+
+\parencite{tomczak_trouble_2022}
+
+% The first component of the ELBO is the negative reconstruction error.
+In other words,  𝐳  is sampled from variational posterior for a given  𝐱  and then it is stochastically reconstructed by  𝑝𝜃(𝐱|𝐳) .
+That is, the conditional likelihood is used to calculate the reconstruction error.
+This term tries to make  𝑞𝜙  as peaky as possible.
+If  𝑞𝜙  is non-concentrated, is is hard to learn a stochastic mapping from multiple  𝐳 's to a single  𝐱 .
+
+% The second component of the ELBO,  𝐷KL[𝑞𝜙(𝐳)‖𝑝𝜆(𝐳)] , determines the error between the aggregated posterior and the marginal over  𝐳 (a.k.a.
+the \textit{prior}).
+Since we maximize the ELBO, and the term 𝐷KL[𝑞𝜙(𝐳)‖𝑝𝜆(𝐳)]  is with the negative sign, we want to minimize it.
+Thus, the ELBO tells us that the difference between  𝑞𝜙(𝐳)  and  𝑝𝜆(𝐳) should be as small as possible.
+This makes perfect sense because we want the aggregated posterior to assign the probability mass to the same regions as the prior.
+
+% The last component of the ELBO,  𝕀𝑞𝜙(𝐱,𝐳)[𝐱;𝐳] , is the mutual information between  𝐱  and  𝐳  for the aggregated posterior and the empirical distribution.
+This term appears with the negative sing, hence, the ELBO indicates that this mutual information should be minimized.
+This is a very puzzling result because, in other words, there should be no stochastic dependency between  𝐳  and  𝐱 .
+Commonly,  𝐳  is interpreted as a representation of  𝐱 , however, optimizing the ELBO leads to a completely different outcome!
+𝕀𝑞𝜙(𝐱,𝐳)[𝐱;𝐳]  is a cumbersome component amd it may be an explanation why training latent variable models with variational inference is typically hard.
+
+% In the conclusion, the ELBO navigates the model to fit the aggregated posterior to the prior, and, on the other hand, to learn no stochastic dependency between  𝐱  and  𝐳  while maintaining a one-to-one mapping from  𝐳  to  𝐱 .
+This last contradiction may result in a difficult training of the model.
+As reported in the literature (e.g., (Alemi et al., 2018)), the choice of parameterizations of the distributions play the crucial role and could be the reason for complicating the training.
+For instance, taking a very powerful parameterization of the conditional likelihood,  𝑝𝜃(𝐱|𝐳) , e.g., a deep autoregressive model, gives a model that completely disregards  𝐳 .
+
+\paragraph{How has this ever worked}
+
+Even though these properties indicate that VAEs might not work at all for representation learning, VAEs have consistently been shown to be able to learn meaningful representations \parencite{sonderby_ladder_2016,hsu_unsupervised_2017,maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
+We can understand the reason for this by noting that both of the above scenarios are only achievable if the optimization too unconstrained. 
+Specifically, the maximum likelihood estimation takes place not over all possible models, but a constrained parametric class defined by the inference network $q_\thetab(\zb|\xb) \approx p(\zb|\xb)$ and the generative model $p_\thetab(\xb|\zb)p(\zb) = p_\theta(\xb,\zb)$.
+This introduces a coupling between the marginal $p_{\thetab}(\xb)$ and the posterior $p_{\thetab}(\zb|\xb)$ which makes the maximum likelihood solution unlikely to correspond to completely useless representations \parencite{huszar_is_2017}.
+
+Similarly, maximizing the ELBO in \cref{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)} by exactly minimizing the mutual information term ...
+
+The challenge is then how to impose appropriate constraints to make VAEs learn representations that are competitive with self"=supervised methods.
+
+% + maximum likelihood does not guarantee good representations of unseen data. On the contrary; the two are actually only correlated by inductive biases present in the model type that we care about. 
+% + VAEs do maximum likelihood.
+%     + but this is true for any model trained with maximum likelihood
+%     + in fact masked pre-trained used for many self-supervised methods can be seen as maximum likelihood training.
+
+% + mutual information minimization of VAEs leads to difficulty of training and posterior collapse.
+%     + only when properly constraining the inference and generative models do VAEs function.
+%     + if the combined inference and generative models easily minimize the mutual information, while producing good reconstructions and making the marginal posterior match the prior, then the model will learn to disregard z, whatever is encoded in it.
+%     + this is especially prone to happen for strong generative models with those that are autoregressive on the observed variable as the core example.
+Since $p(\xb_t)$ in many cases can be modelled very well by a generative model of the form $p(\xb_t | \xb_{1:t-1})$, the potential improvement from using $p(\xb_t | \xb_{1:t-1}, \zb_{1:t})$, or even $p(\xb_t | \xb_{1:t-1}, \zb_{1:T})$, might be small compared to the magnitude of the mutual information term
+%     \cref{eq:variational-autoencoder-elbo-factorized-expected}
+
+\todo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimizing a mutual information term in the ELBO (derive this form).}
+\todo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other).
+Include reference to \parencite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
+
+\todo[inline]{Masked pre-training actually corresponds to marginal maximum likelihood optimization such as done by VAES \parencite{moreno-munoz_masked_2023}.
+Since maximum likelihood in no way guarantees good representations, how do self"=supervised methods achieve  this, if they also do maximum likelihood?}
+
+\subsection{Designing the latent space} 
 Architectural improvements and hierarchies of latent variables are two related, and interesting, approaches to learning more semantic features in VAEs. 
@@
 
-\paragraph{Adding a few labels} 
+\subsection{Adding a few labels} 
 Another way to improve the representations learned in VAEs is via semi-supervised learning. 
@@
 
-\paragraph*{Approximating less}
+\subsection{Approximating less}
 VAE are inherently approximate.
@@
 
-\paragraph{Masking} 
+\subsection{Masking} 
 As discussed in \cref{chp:paper-review}, masking is one of the driving techniques behind the success of self"=supervised methods for speech \parencite{devlin_bert_2018,baevski_wav2vec_2020}.
@@
 
-\paragraph{Giving up?}
-\lesstodo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimizing a mutual information term in the ELBO (derive this form).}
-\lesstodo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other).
-Include reference to \parencite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
-
-\lesstodo[inline]{Out-of-distribution detection on speech?}
+\todo[inline]{Out-of-distribution detection on speech?}
 

### 2023-10-22 chapters/introduction.tex
Work on discussion, background and introduction

@@
 
-In an era where technology increasingly intertwines with healthcare, artificial intelligence is on the verge of becoming part of standard medical practice \parencite{shailaja_machine_2018}. 
+In an era where technology increasingly intertwines with healthcare, artificial intelligence is on the verge of becoming part of standard medical practice. 
 Growing administrative burdens, aging populations, and rapid scientific progress in medicine and medical devices have created a need to rethink how medical professionals are best enabled to successfully do their job.
-The application of artificial intelligence in medical decision-making and particularly for diagnosing and recommending treatments is likely to hold significant potential to reduce the effect of these issues and improve patient outcomes.
-However, as such decisions impact critical aspects of human health and well-being, this prospect also raises concerns related to the reliability of systems that use artificial intelligence \parencite{chen_ethical_2021, ahmad_interpretable_2018}.
+The application of artificial intelligence in medical decision-making is likely to hold significant potential to reduce the effect of these issues and improve patient outcomes \parencite{shailaja_machine_2018}.
+However, as such decisions impact critical aspects of human health and well-being, this prospect has also raised concerns relating to the reliability of systems that use artificial intelligence \parencite{chen_ethical_2021, ahmad_interpretable_2018}.
 
@@
 After the agricultural revolution, several ancient civilizations developed mathematical techniques and algorithms to help them manage the newfound complexity of society.
-Fittingly, the earliest human whose name is still known today was a Sumerian accountant called Kushim.
-Their signature has been identified on at least 18 clay tablets dated to 3400-3000 BCE that deal with transactions and inventory of barley and other ingredients for beer production \parencite{nissen_archaic_1993}.
-Another example is the Egyptian Rhind Mathematical Papyrus dated to around 1550 BCE which describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \parencite{georges_universal_2001}.
+Fittingly, the first recorded human name in written history belonged to Kushim, a Sumerian accountant.
+Their signature has been identified on at least 18 clay tablets dated to 3400-3000 BCE, which document transactions and inventory related to barley and other ingredients used in beer production \parencite{nissen_archaic_1993}.
+A later example is the Egyptian Rhind Mathematical Papyrus dated to around 1550 BCE which describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \parencite{georges_universal_2001}.
 Around the same time, the Babylonians had developed a sophisticated system of mathematics that included algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \parencite{fowler_square_1998}.
@@
 Assumedly unbeknownst to Kushim, this tablet would also come to contain one of the world's first recorded examples of a mathematical error; the statement: $5 \times \frac{1}{2} = 5$ \parencite{nissen_archaic_1993}.
-Later, in its list of Pythagorean triplets, the Plimpton 322 clay tablet (\cref{fig:plimpton_332}) also has several errors, presumably made by a novice in Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
+Later, the scribe of the Plimpton 322 clay tablet (\cref{fig:plimpton_332}) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
+% Later, in its list of Pythagorean triplets, the Plimpton 322 clay tablet (\cref{fig:plimpton_332}) also has several errors, presumably made by a novice in Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
 % Apart from such inherent limitations of the tools, inadequate human proficiency in their use was also a source of error. 
 Besides these human errors, limitations inherent to the tools themselves also posed problems. 
-Babylonian mathematicians for instance did not formalize the difference between rational and irrational numbers, which lead to approximation errors.
-For instance, the common practice of approximating $\pi$ with 3 \parencite{georges_universal_2001}.
+The Babylonian number system, for instance, was well-equipped to represent rational numbers but inefficient for approximating irrational numbers. 
+This forced significant approximation errors, for instance the common practice of approximating $\pi$ with 3 \parencite{georges_universal_2001}. 
 % Easy to overlook, the challenges we face in adopting new technologies today are not new but have been faced by humans for thousands of years.
 Today, as in ancient times, it was important to understand the limitations of technology and the need for human expertise in its use.
+
+% Yes, Babylonian mathematics, which developed in ancient Mesopotamia (modern-day Iraq) around 2000 BCE, had a numerical system that was capable of representing some irrational numbers, although they did so indirectly.
+The Babylonians primarily used a base-60 positional numeral system, which was well-suited for representing many rational numbers.
+
+% Irrational numbers, such as the square root of 2 (√2) or π (pi), cannot be exactly represented in any finite numerical system, whether it's base-10 (our modern decimal system) or base-60 (used by the Babylonians).
+However, the Babylonians were able to approximate these irrational numbers to a certain degree using their sexagesimal system.
+
+% For example, to approximate the square root of 2, the Babylonians might have used a fraction like 1;24,51,10 (where the semicolon represents the sexagesimal radix point).
+This would represent an approximation of √2, although not an exact representation.
+Similarly, they could approximate π using a sexagesimal fraction, but it would be an approximation rather than an exact value.
+
+% The Babylonians were skilled mathematicians and astronomers, and they used their numerical system to perform a wide range of calculations, including those involving geometry and timekeeping.
+While they couldn't precisely represent irrational numbers, they could work with these numbers to a certain level of accuracy within the constraints of their base-60 system
 

### 2023-10-23 chapters/paper_benchmarking.tex
Improve table layout

@@
 \begin{table}
-    \caption[Model likelihoods and phoneme error rate for TIMIT.]{ We report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines (left) and phoneme error rate (PER) of different representations for phoneme recognition on TIMIT (right).}
+    \caption[Model likelihoods and phoneme error rate for TIMIT.]{ We report model likelihoods $\mathcal{L}$ for TIMIT represented as a $\SI{16}{bit}$ $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines (left) and phoneme error rate (PER) of different representations for phoneme recognition on TIMIT (right).}
     \hfill
-    \begin{subtable}[t]{.49\textwidth}
+    \begin{subtable}[t]{.485\textwidth}
     \centering
@@
     \hfill
-    \begin{subtable}[t]{.49\textwidth}
+    \begin{subtable}[t]{.485\textwidth}
     \centering

### 2023-10-24 chapters/background.tex
Restructure VAE background [WIP]

@@
 \section{Out-of-distribution detection} \label{sec:out-of-distribution-detection}
+\todo[inline]{Maybe this section should come after the VAE background?}
 % 
@@
 % In the literature, the task is also referred to as open set recognition \parencite{bendale_open_2016, cardoso_weightless_2017, panaredabusto_open_2017}, outlier detection \parencite{xu_deep_2014, mor_confidence_2018}, or selective prediction \parencite{geifman_selective_2017} .
-OOD detection bears many similarities with anomaly detection, novelty detection, open set recognition, and outlier detection \parencite{yang_generalized_2022}. 
+\paragraph{Related tasks} OOD detection bears many similarities with anomaly detection, novelty detection, open set recognition, and outlier detection \parencite{yang_generalized_2022}. 
 Some of these differences are subtle, and the terminology is not always used consistently in the literature.
@@
 
-Recently, several approaches for deep neural networks have been developed that address the rejection of samples $\xb\notin\mathcal{D}_{\text{in}}$. 
+\paragraph{Supervision} Recently, several approaches for deep neural networks have been developed that address the rejection of OOD samples $\xb\notin\mathcal{D}_{\text{in}}$. 
 One way to categorize the different approaches is based on whether the underlying model is a classifier that estimates a conditional probability distribution $p(y|\xb)$ over some target variable $y$, or a model that learns a probability density $p(\xb)$ over the input itself.
 We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \parencite{graham_denoising_2023,liu_unsupervised_2023a}.
-It is important to note that this distinction relates only to the availability of some target value, $y$ - not whether OOD data is available for supervision\footnote{
+It is important to note that this distinction relates only to the availability of some target value, $y$ -- not whether OOD data is available for supervision\footnote{
     In our overview, we distinguish between supervised and unsupervised out-of-distribution by letting methods be classified as supervised that use any kind of target value $y$, whether it relates to an original (in-distribution) task, available OOD data, or both.
@@
 
-Central to any OOD detection method is the ability to assign a score $s(\xb) \in \mathbb{R}$ to a given input $\xb$ that indicates the degree to which the input is likely to be OOD.
+\paragraph{Scoring} Central to any OOD detection method is the ability to assign a score $s(\xb) \in \mathbb{R}$ to a given input $\xb$ that indicates the degree to which the input is likely to be OOD.
 After defining a score, we typically use a validation set to tune a threshold $\tau$ such that $\xb$ is considered OOD if $s(\xb) > \tau$.
@@
 Other methods including MCD \parencite{yu_unsupervised_2019}, NGC \parencite{wu_ngc_2021}, and UDG \parencite{yang_semantically_2021} use external, unlabeled, noisy data to improve OOD detection performance without requiring cleanly labeled OOD examples.
+
 As we discussed earlier, for many modern applications of machine learning, input data is often high dimensional and complex which makes it difficult to obtain enough representative OOD data to ensure robust OOD detection capabilities.
+This fundamentally limits the usefulness of methods that require samples of real OOD data.
 
@@
 Another approach generates OOD inputs using a generative adversarial network \parencite{lee_training_2018}.
+
 Similar to actual OOD data, the usefulness of synthetic OOD data is also fundamentally limited by the intractability of sampling the complete space of OOD data.
+Again, these methods are fundamentally limited in their generality.
 
@@
 \paragraph{Methods not using OOD data}
-The least informed supervised OOD detection methods do not require instantiations of any kind of OOD data, real or synthetic. 
+The least informed supervised OOD detection methods do not require instantiations of any kind of OOD data, real or synthetic, and do away with the associated limitations.
+
 A baseline approach uses the maximum class probability of $p(y|\xb)$ directly by noting that it tends to be larger for correctly classified examples \parencite{hendrycks_baseline_2017}.
@@
 Other methods that derive the score from the classifier probabilities include \textcite{lakshminarayanan_simple_2017} who propose to use an ensemble of independently trained classifiers to discriminate between in-distribution and OOD data by evaluating the agreement between the classifiers, \textcite{devries_learning_2018} who augment the network with a confidence estimation branch that learns to estimate the confidence of the classifier separately from the probability, and \textcite{huang_importance_2021} who compute the gradient of the KL-divergence of the predictive distribution $p(y|\xb)$ to a uniform distribution noting that the magnitude of gradients is higher for in-distribution data than for OOD data.
-The Variational Information Bottleneck \parencite{alemi_deep_2017} jointly learns a probabilistic latent representation, $\zb$ and $p(y|\xb)$, using a generalized variational autoencoder \parencite{kingma_autoencoding_2014} that tries to maximize the mutual information between $\zb$ and $y$.
+The Variational Information Bottleneck \parencite{alemi_deep_2017} jointly learns a probabilistic latent representation, $p(\zb|\xb)$, and a classifier, $p(y|\xb)$, using a generalized variational autoencoder \parencite{kingma_autoencoding_2014} and tries to maximize the mutual information between $\zb$ and $y$ and minimize it between $\zb$ and $\xb$.
 
@@
 %
-This shows that log of the softmax confidence score is equivalent to the special case of the energy score, where all the logits are shifted by their maximum logit value. 
-The authors then note that the energy $E(\xb; f)$ empirically tends to be larger for OOD data than for in-distribution data, while $f^\text{max}(\xb)$ tends to be smaller. 
+This shows that log of the softmax confidence score is equivalent to the special case of the energy score where all the logits are shifted by their maximum logit value. 
+The authors empirically observe that the energy $E(\xb; f)$ tends to be larger for OOD data than for in-distribution data, while $f^\text{max}(\xb)$ tends to be smaller. 
 They conclude that this shift results in the maximum softmax probability being a biased score for OOD detection and propose to instead use the energy directly.
-This energy-score was improved in ReAct by feature clipping \parencite{sun_react_2021}.
+This energy-score was further improved in ReAct by feature clipping \parencite{sun_react_2021}.
 
@@
 Some methods define OOD classes. 
-\textcite{huang_mos_2021} groups the classes of the target variable $y$ and defines an OOD class for each group.
+\textcite{huang_mos_2021} group the classes of the target variable $y$ and defines an OOD class for each group.
 Each training example is then the correct target for one group and an OOD example for all other groups; a kind of hierarchical version of the OOD class of simpler baselines based on noise augmentation \parencite{ren_likelihood_2019}.
-Similarly, to represent a virtual OOD class, \textcite{wang_vim_2022} generates an additional logit by first computing the residual of the input's latent space representation against a principal feature space and then converting it to a valid logit by matching its mean over training samples to the average maximum logits.
+\todo[inline]{Does the previous sentence belong here or in a different paragraph? It is not really using the latent space.}
+Similarly, to represent a virtual OOD class, \textcite{wang_vim_2022} generate an additional logit by first computing the residual of the input's latent space representation against a principal feature space and then converting it to a valid logit by matching its mean over training samples to the average maximum logits.
 
 Other methods note that the difficulty of detecting OOD data might be attributed to the curse of dimensionality in the learned feature spaces and propose to use dimensionality reduction techniques.
-In \textcite{ndiour_outofdistribution_2020}, apply dimensionality reduction on learned, high-dimensional features to capture the true feature subspace and compute the norm of the difference between the original feature and the pre-image of its low-dimensional manifold embedding.
-\textcite{zaeemzadeh_outofdistribution_2021} forces the ID samples to embed into a union of 1-dimensional subspaces during training and computes the minimum angular distance from the feature to the class-wise subspaces.
+\textcite{ndiour_outofdistribution_2020} apply dimensionality reduction on learned, high-dimensional features to capture the true feature subspace and compute the norm of the difference between the original feature and the pre-image of its low-dimensional manifold embedding.
+\textcite{zaeemzadeh_outofdistribution_2021} force the ID samples to embed into a union of 1-dimensional subspaces during training and computes the minimum angular distance from the feature to the class-wise subspaces.
 NuSA \parencite{cook_outlier_2020} uses projects features onto the column space of the classification weight matrix and computes the ratio of the norm the projected and original features.
@@
 % Summary and weaknesses of supervised OOD detection approaches
+
 A general weakness of all supervised out-of-distribution detection is that in learning the task-specific model $p(y|\xb)$ a model may discard information about $p(\xb)$ which could be useful for out-of-distribution detection.
-Methods that use real OOD data achieve high performance but are limited by the availability of OOD data and how well it covers the large variation in OOD data.
 
@@
 
+\todo[inline]{Briefly hold the readers hand and tell them why VAEs are relevant to the overarching project.}
+
 % A variational autoencoder (VAE) \parencite{kingma_autoencoding_2014,rezende_stochastic_2014} is a type of deep generative model and an instance of a latent variable model. 
 In the following, we will introduce VAEs starting from the definition of latent variable models.
-We then derive the marginal likelihood objective used to train VAEs, discuss its different components, and highlight some central challenges facing VAEs.
+We then derive the marginal likelihood objective used to train VAEs, discuss its different components, and highlight some central challenges of applying VAEs.
 
@@
 This makes it possible to efficiently compute approximate posterior distributions for any input $\xb$.
-The approximate posterior can also be used to compute a lower bound on the marginal distribution $p(\xb)$, the so-called evidence lower bound (ELBO),
+The approximate posterior can also be used to compute a lower bound on the marginal distribution $p_\thetab(\xb)$ via Jensen's inequality, the so-called evidence lower bound (ELBO),
 %
  \begin{align} \label{eq:variational-autoencoder-elbo}
-    \log p(\xb)
-    &= \log \int p_\thetab(\xb|\zb)p(\zb),\textrm{d}\zb \nonumber \\
-    &= \log \int q_\phib(\zb|\xb) \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)}\,\textrm{d}\zb \nonumber \\
-    &\geq \int q_\phib(\zb|\xb) \log \left( \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right) \,\textrm{d}\zb \nonumber \\
-    &\geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}_{\thetab,\phib}(\xb) \enspace .
+    \log p_\thetab(\xb)
+    &= \log \int p_\thetab(\xb,\zb),\textrm{d}\zb \nonumber \\
+    &= \log \int q_\phib(\zb|\xb) \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)}\,\textrm{d}\zb \nonumber \\
+    &\geq \int q_\phib(\zb|\xb) \log \left( \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right) \,\textrm{d}\zb \nonumber \\
+    &\geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}_{}(\xb;\thetab,\phib) \enspace .
+    % &= \log \int p_\thetab(\xb|\zb)p(\zb),\textrm{d}\zb \nonumber \\
+    % &= \log \int q_\phib(\zb|\xb) \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)}\,\textrm{d}\zb \nonumber \\
+    % &\geq \int q_\phib(\zb|\xb) \log \left( \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right) \,\textrm{d}\zb \nonumber \\
+    % &\geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}_{}(\xb;\thetab,\phib) \enspace .
 \end{align}
 %
-Although analytically evaluating the ELBO is intractable, it can be approximated by sampling and used to train VAEs. We will return to this after examining the ELBO a bit.
+Although analytically evaluating the ELBO is intractable, it can be approximated by sampling and used to train VAEs. We will return to this after examining the ELBO in detail.
 
+\paragraph{KL-divergence to prior}
 Since the joint factorizes in VAEs, we can also write the ELBO in \cref{eq:variational-autoencoder-elbo} as,
@@
 \begin{align} \label{eq:variational-autoencoder-elbo-factorized}
-    \mathcal{L}_{\thetab,\phib}(\xb)
+    \mathcal{L}_{}(\xb;\thetab,\phib)
     % &\equiv \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
     &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right] + \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
-    &= \underset{\text{reconstruction loss}}{\underbrace{\mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]}} - \underset{\text{KL-divergence to prior}}{\underbrace{D_\text{KL}\left( q_\phib(\zb|\xb) \,||\, p(\zb) \right)}} \enspace .
+    &= 
+    \underset{\text{reconstruction loss}}{\underbrace{
+        \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]
+    }}
+     - 
+    \underset{\text{KL-divergence to prior}}{
+        \underbrace{\vphantom{\mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]}{D_\text{KL}\left( q_\phib(\zb|\xb) \parallel p(\zb) \right)}
+    }} \enspace .
 \end{align}
@@
 This reveals a commonly used interpretation of the ELBO: 
-The expected log-likelihood of the data under the generative model $p_\thetab(\xb|\zb)$ acts as a reconstruction loss that encourages the model to make $q_\phib(\zb|\xb)$ as peaky as possible to make reconstruction easy.
-The negative KL-divergence between the posterior $q_\phib(\zb|\xb)$ and the prior $p(\zb)$ acts as a regularizer that forces the approximate posterior to be within the support of the prior (reverse KL-divergence).
-This helps prevent overfitting but importantly also makes ancestral sampling from the prior produce high quality data samples.
+The expected log-likelihood of the generative model $\mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]$ acts as a reconstruction loss that encourages the model to make $q_\phib(\zb|\xb)$ as peaky as possible since mapping many $\zb$ to a single $\xb$ is hard.
+The negative KL-divergence between the approximate posterior $q_\phib(\zb|\xb)$ and the prior $p(\zb)$ acts as a regularizer that forces the approximate posterior to be within the support of the prior (reverse KL-divergence).
+This helps prevent overfitting but importantly also enables ancestral sampling from the prior to generate high quality data.
 
-By maximizing $\mathcal{L}_{\thetab,\phib}(\xb)$, we must minimize the KL-divergence.
+By maximizing $\mathcal{L}_{}(\xb;\thetab,\phib)$ we must minimize the KL-divergence.
 However, an optimal KL-divergence of zero requires $q_\phib(\zb|\xb) = p(\zb)$ which would mean that $\zb$ has become independent of $\xb$.
@@
 
+\paragraph{Inference gap} 
+The ELBO is a lower bound on the marginal likelihood, so it is interesting to examine when the bound holds exactly. 
+Without using Jensen's inequality, in the below, we will rewrite the log-marginal likelihood with the goal of arriving at the ELBO of \cref{eq:variational-autoencoder-elbo} plus a factor to fill the inference gap.
+% Since the ELBO is computed as an expectation under the variational approximation to the posterior, $q_\phib(\xb|\zb)$, we will seek to rewrite the log-marginal likelihood into such an expectation, .
+%
+\begin{align}
+    \log p_\thetab(\xb) 
+    &= \int q_\phib(\zb|\xb) \log p_\thetab(\xb) \, \textrm{d}\zb \nonumber \\
+    &= \int q_\phib(\zb|\xb) \log \left(\frac{p_\thetab(\xb,\zb)}{p_\thetab(\zb|\xb)}\right) \, \textrm{d}\zb \nonumber \\
+    &= \int q_\phib(\zb|\xb) \log \left(\frac{p_\thetab(\xb,\zb) q_\phib(\zb|\xb)}{p_\thetab(\zb|\xb) q_\phib(\zb|\xb)}\right) \, \textrm{d}\zb \nonumber \\
+    &= \int q_\phib(\zb|\xb) \log \left(\frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)}\right) \, \textrm{d}\zb + \int q_\phib(\zb|\xb) \log \left(\frac{q_\phib(\zb|\xb)}{p_\thetab(\zb|\xb)}\right) \, \textrm{d}\zb \nonumber \\
+    &=
+    \underset{\text{ELBO}}{\underbrace{
+        \mathcal{L}_{}(\xb;\thetab,\phib)
+    }} 
+    + 
+    \underset{\text{KL-divergence to true posterior}}{\underbrace{
+        \vphantom{\mathcal{L}_{}(\xb;\thetab,\phib)}{D_\mathrm{KL}\left( p_\thetab(\zb|\xb) \parallel q_\phib(\zb|\xb) \right)}
+    }} \enspace .
+\end{align}
+%
+We can see that the inference gap of the ELBO is exactly equal to the KL-divergence of the approximate posterior to the true posterior. 
+This also means that the approximate posterior $q_\phib(\zb|\xb)$ that maximizes the ELBO also best approximates the true posterior, as measured by $D_\mathrm{KL}\left( p(\zb|\xb) \parallel q(\zb|\xb) \right)$.
+
+We can subdivide the inference gap \cite{cremer_inference_2018}: 
+By amortizing the variational inference we introduce a gap compared to the alternative of fitting a posterior per data point, $q^{*}_{\xb}(\zb)$.
+This amortization gap can be reduced by using large capacity amortized posteriors.
+In cases where the true posterior does not belong to the model class of the variational family of posteriors that we fit, the inference gap cannot be reduced to zero and the remaining gap is referred to as the approximation gap.
+
+\todo[inline]{Finish this with the parsimony gap}
+
+A final error is introduced  the maximally attainable likelihood over all the group $G$ of possible models, $\mathcal{l}(G)$
+
+It is important to also note that in fitting a VAE, the approximation gap is not the only source of error.
+
 \subsection{Maximum likelihood estimation}
-To train a VAE by maximum likelihood, we must approximate the expectation in the ELBO \cref{eq:variational-autoencoder-elbo} and compute its gradients w.r.t. $\phib$ and $\thetab$.
+%
+We want to use maximum likelihood estimation to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$. 
+Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
+To do this we therefore must approximate the expectation in the ELBO \cref{eq:variational-autoencoder-elbo} and compute its gradients w.r.t. $\phib$ and $\thetab$.
 
 \paragraph{Sampling}
-Since we have freedom to select the approximate posterior, we can choose it such that it is easy to sample. 
-We can then form a Monte Carlo estimate of the ELBO as
+Since we have freedom to select the approximate posterior, we can choose it such that it is easy to sample and then use that property to form a Monte Carlo estimate of the ELBO as
 % 
 \begin{equation} \label{eq:variational-autoencoder-elbo-sampling-expanded}
-    \mathcal{L}_{\thetab,\phib}(\xb)
-    = \frac{1}{K} \sum_{k=1}^K \log p_\thetab(\xb|\zb_k) + p(\zb_k) - \log q_\phib(\zb_k|\xb) \enspace , 
+    \mathcal{L}_{}(\xb;\thetab,\phib)
+    \approx \frac{1}{K} \sum_{k=1}^K \log p_\thetab(\xb|\zb_k) + \log p(\zb_k) - \log q_\phib(\zb_k|\xb) \enspace , 
 \end{equation}
@@
 \begin{equation} \label{eq:variational-autoencoder-elbo-sampling-KL}
-    \mathcal{L}_{\thetab,\phib}(\xb)
-    = \frac{1}{K} \sum_{k=1}^K \log p_\thetab(\xb|\zb_k) - D_\text{KL}\left( q_\phib(\zb|\xb) \,||\, p(\zb) \right) \enspace ,
+    \mathcal{L}_{}(\xb;\thetab,\phib)
+    \approx \frac{1}{K} \sum_{k=1}^K \log p_\thetab(\xb|\zb_k) - D_\text{KL}\left( q_\phib(\zb|\xb) \,||\, p(\zb) \right) \enspace .
 \end{equation}
@@
 
+\todo[inline]{Derive and introduce the IWAE bound.}
+
 \paragraph{Reparameterization} 
-We want to use maximum likelihood estimation to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$. 
-Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
+% We want to use maximum likelihood estimation to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$.
+% Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
 Computing the gradient of the ELBO with respect to $\thetab$ is straightforward,
@@
     % \nabla_\thetab \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
-    \nabla_\thetab \mathcal{L}_{\thetab,\phib}(\xb)
+    \nabla_\thetab \mathcal{L}_{}(\xb;\thetab,\phib)
     = \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \nabla_\thetab \log p_\thetab(\xb|\zb) \right] \enspace .
@@
 %
-The gradient with respect to the $\phib$ is more challenging since $\phib$ occurs in the expectation itself making it impossible to move the gradient inside the expectation. 
+However, the gradient with respect to the $\phib$ is more challenging. Since $\phib$ occurs in the expectation itself, we may not move the gradient inside the expectation. 
 However, by reparameterizing the latent variable $\zb$ as a deterministic function of the input $\xb$ and a random variable $\epsilonb\sim p(\epsilonb)$, $\zb = g_\phib(\xb,\epsilonb)$, we can write the gradient as a path-wise derivative,
@@
     % \nabla_\phib \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
-    \nabla_\phib \mathcal{L}_{\thetab,\phib}(\xb)
+    \nabla_\phib \mathcal{L}_{}(\xb;\thetab,\phib)
     = \mathbb{E}_{p(\epsilonb)} \left[ \nabla_\phib \log \frac{p_\thetab\left(\xb|g_\phib(\xb,\epsilonb)\right)p\left(g_\phib(\xb,\epsilonb)\right)}{q_\phib\left(g_\phib(\xb,\epsilonb)|\xb\right)} \right] \enspace .
@@
 %
-The most important instantiation of this reparameterization is the diagonal Gaussian case, where $g_\phib(\xb,\epsilonb) = \mu_\phib(\xb) + \sigma_\phib(\xb) \odot \epsilonb$ with $\epsilonb\sim\mathcal{N}(0,\mathbf{I})$.
-By assuming the covariance matrix is diagonal, this is a mean-field approximation which imposes independence between the dimensions of $\zb$.
+The most important instantiation of this reparameterization is the diagonal Gaussian case. 
+By choosing $q_\phib(\zb|\xb) = \mathcal{N}(\zb; \mub_\phib(\xb), \mathbf{I}\sigmab_\phib(\xb))$ it is possible to sample differentiably via the reparameterization trick by setting $g_\phib(\xb,\epsilonb) = \mu_\phib(\xb) + \sigma_\phib(\xb) \odot \epsilonb$ with $\epsilonb\sim\mathcal{N}(0,\mathbf{I})$.
+By assuming the covariance matrix is diagonal, this becomes a mean-field approximation which imposes independence between the dimensions of $\zb$.
 
@@
 While learning a full covariance is a simple solution, it is not always sufficiently computationally efficient and, in any case, enables learning only linear covariance between elements of $\zb$. 
-The drawbacks of mean-field approximation and linear covariance has lead to research into learning hierarchies of several non-linearly dependent latent variables; an idea well in line with the usual motivations behind deep neural networks such as efficient, compositional representation \cite{lecun_deep_2015}.
+The drawbacks of mean-field approximation and linear covariance have lead to research into learning hierarchies of several non-linearly dependent latent variables; an idea well in line with the usual motivations behind deep neural networks such as efficient, compositional representation \cite{lecun_deep_2015}.
 Such hierarchical models are usually formalized by introducing a set of $L$ latent variables  $\zb=\zb_1, \dots, \zb_L$ and letting the generative model be defined as,
@@
     %q_\phib(\zb|\xb)=q_\phib(\zb_1|\xb)q_\phib(\zb_2|\zb_1)\cdots q_\phib(\zb_L|\zb_{L-1})
-    q_\phib(\zb|\xb) = q_\phib(\zb_1|\xb)\textstyle\prod_{i=2}^{L} q_\phib(\zb_i|\zb_{i-1})
+    q_\phib(\zb|\xb) = q_\phib(\zb_1|\xb)\prod_{i=2}^{L} q_\phib(\zb_i|\zb_{i-1})
 \end{equation}
@@
     % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb_{L-1}|\zb_L)\cdots q_\phib(\zb_2|\zb_1),
-    q_\phib(\zb|\xb) = q_\phib(\zb_L|\xb)\textstyle\prod_{i=L-1}^{1} q_\phib(\zb_{i}|\zb_{i+1}) \ .
+    q_\phib(\zb|\xb) = q_\phib(\zb_L|\xb)\prod_{i=L-1}^{1} q_\phib(\zb_{i}|\zb_{i+1}) \ .
 \end{equation}
@@
 \begin{equation} \label{eq:variational-autoencoder-elbo-factorized-expected}
-    \mathbb{E}_{\hat{p}(\xb)} \left[ \mathcal{L}_{\thetab,\phib}(\xb) \right] = 
+    \mathbb{E}_{\hat{p}(\xb)} \left[ \mathcal{L}_{}(\xb;\thetab,\phib) \right]
+    = 
     \underset{\text{average reconstruction}}{\underbrace{
@@
     I_{q_{\phib}(\xb,\zb)}\left[\xb ; \zb \right]
-    }} \enspace . \nonumber
+    }} \. . %\enspace . \nonumber
 \end{equation}

### 2023-10-24 chapters/discussion.tex
Updates to discussion

@@
 This can be equivalently phrased as maximizing the KL-divergence of the true data distribution $p_{\text{data}}(\xb)$ to the learned marginal, $D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb))$\footnote{%
-Maximizing the likelihood $\sum_{n=1}^N \log p_\theta(\xb)$ w.r.t. parameters $\thetab$ is asymptotically equivalent to minimizing $D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb))$. 
+Maximizing the likelihood $\sum_{n=1}^N \log p_\theta(\xb)$ w.r.t. parameters $\thetab$ is asymptotically equivalent to minimizing $D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb))$: 
 %
@@
 %
-The last equation follows from the law of large numbers.
+The last equality follows from the law of large numbers.
 % \begin{align}
@@
 Any combination uniquely defines a valid latent variable model.
-Hence, maximizing the marginal is a completely useless criteria for representation learning, regardless of the measure of usefulness \parencite{huszar_is_2017}.
+Hence, maximizing the marginal is a completely useless criteria for representation learning, regardless of the measure of usefulness \parencite{huszar_is_2017, alemi_fixing_2018}.
 
 \paragraph{Mutual information minimization}
-As we saw in \cref{subsec: mutual-information-interpretation-of-elbo} ...
-
-\begin{equation} \label{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)}
+As we saw in \cref{subsec: mutual-information-interpretation-of-elbo}, optimizing the ELBO involves minimizing the mutual information between the latent and observed variables.
+We reprint equation \cref{eq:variational-autoencoder-elbo-factorized-expected} below for reference.
+%
+\begin{equation} %\label{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)}
     \mathbb{E}_{q_{\phib}(\xb,\zb)} \left[ \mathcal{L}_{\thetab,\phib}(\xb) \right] = 
@@
 \end{equation}
+%
+Optimizing the ELBO can be understood as simultaneously maximizing the average reconstruction likelihood, minimizing the marginal KL-divergence to the prior, and minimizing the mutual information between the observed and latent variables.
 
@@
 
-% The first component of the ELBO is the negative reconstruction error.
-In other words,  𝐳  is sampled from variational posterior for a given  𝐱  and then it is stochastically reconstructed by  𝑝𝜃(𝐱|𝐳) .
-That is, the conditional likelihood is used to calculate the reconstruction error.
-This term tries to make  𝑞𝜙  as peaky as possible.
-If  𝑞𝜙  is non-concentrated, is is hard to learn a stochastic mapping from multiple  𝐳 's to a single  𝐱 .
-
-% The second component of the ELBO,  𝐷KL[𝑞𝜙(𝐳)‖𝑝𝜆(𝐳)] , determines the error between the aggregated posterior and the marginal over  𝐳 (a.k.a.
-the \textit{prior}).
-Since we maximize the ELBO, and the term 𝐷KL[𝑞𝜙(𝐳)‖𝑝𝜆(𝐳)]  is with the negative sign, we want to minimize it.
-Thus, the ELBO tells us that the difference between  𝑞𝜙(𝐳)  and  𝑝𝜆(𝐳) should be as small as possible.
-This makes perfect sense because we want the aggregated posterior to assign the probability mass to the same regions as the prior.
-
-% The last component of the ELBO,  𝕀𝑞𝜙(𝐱,𝐳)[𝐱;𝐳] , is the mutual information between  𝐱  and  𝐳  for the aggregated posterior and the empirical distribution.
-This term appears with the negative sing, hence, the ELBO indicates that this mutual information should be minimized.
-This is a very puzzling result because, in other words, there should be no stochastic dependency between  𝐳  and  𝐱 .
-Commonly,  𝐳  is interpreted as a representation of  𝐱 , however, optimizing the ELBO leads to a completely different outcome!
-𝕀𝑞𝜙(𝐱,𝐳)[𝐱;𝐳]  is a cumbersome component amd it may be an explanation why training latent variable models with variational inference is typically hard.
-
-% In the conclusion, the ELBO navigates the model to fit the aggregated posterior to the prior, and, on the other hand, to learn no stochastic dependency between  𝐱  and  𝐳  while maintaining a one-to-one mapping from  𝐳  to  𝐱 .
-This last contradiction may result in a difficult training of the model.
-As reported in the literature (e.g., (Alemi et al., 2018)), the choice of parameterizations of the distributions play the crucial role and could be the reason for complicating the training.
-For instance, taking a very powerful parameterization of the conditional likelihood,  𝑝𝜃(𝐱|𝐳) , e.g., a deep autoregressive model, gives a model that completely disregards  𝐳 .
-
-\paragraph{How has this ever worked}
+\paragraph{How did this ever work?}
 
@@
 Instead, these models rely on advanced inference networks \parencite{maaloe_biva_2019} and architectural improvements \parencite{vahdat_nvae_2020, child_very_2021} to overcome the challenges of training hierarchical VAEs.
-This seems to indicate that there might be untapped potential in consolidating the work on low-variance gradient estimators and advanced architectures and inference networks.
+This seems to indicate that there might be untapped potential in consolidating the work on low-variance gradient estimators, advanced architectures and inference methods.
 
@@
 
-\section{\Cref{chp:paper-retrospective} revisited: Calibration}
+% \section{\Cref{chp:paper-retrospective} revisited: Calibration}
+\section{Uncertainty of the stroke recognition classifier}
 

### 2023-10-24 chapters/introduction.tex
Add todo

@@
 
+\todo[inline]{Update here with:
+
+I would try to embed a perspective for contemplation.
+Instead of us always talking about what the technology needs to improve on.
+Perhaps, it would also be prudent to ask an open-ended philosophical question: “While our requirements to the technology are in its place, one could ask, whether human beings have the ability to quantify uncertainty.
+In 2023, the prevalence of human errors in driving are much higher than self-driving vehicles.” <- Quick example.
+I’m sure you can find something better.
+
+% - a copilot for medical interviews
+% - decision support
+% - quality assurance
+% - documentation
+% -> many applications
+}
+
 \subsection{Stroke recognition in emergency calls} \label{subsec:motivation-stroke-recognition}
@@
 %, especially if they require outputs unavailable to the model.
-In high-dimensional data with a practically unlimited diversity in out-of-distribution examples, it is impossible to collect enough data to eliminate all potential sources of epistemic uncertainty: By including horses in the training data, the model would likely learn to recognize them, but would still be unable to recognize giraffes, or zebras, or unicorns.
+In high-dimensional data with a practically unlimited diversity in out-of-distribution examples, it is impossible to collect enough data to eliminate all potential sources of epistemic uncertainty.
+By including horses in the training data, the model would likely learn to recognize them, but would still be unable to recognize giraffes, or zebras, or unicorns.
 Ultimately, we must accept that any practical machine learning model will have some sources of epistemic uncertainty.
 
-Since aleatoric uncertainty is present in the data, it can usually be modelled directly.
+Since aleatoric uncertainty is represented by the data, it can usually be modelled directly.
 For instance, in the case of speech recognition, we can model the uncertainty in the transcription of a word by a distribution over the words in the vocabulary.
 But how do we model epistemic uncertainty?
-Since sources of epistemic uncertainty are by definition those not represented in the data, it is not generally possible to model them directly.
+Since sources of epistemic uncertainty are by definition those not represented in the data, it is generally not possible to model them directly.
 This makes epistemic uncertainty more difficult to quantify than aleatoric uncertainty.
-Methods for estimating epistemic uncertainty include Bayesian neural networks \parencite{mackay_practical_1992, neal_bayesian_1995} which represent it by an explicit distribution over learned model parameters.
+Methods for estimating epistemic uncertainty include Bayesian probability and Bayesian neural networks \parencite{mackay_practical_1992, neal_bayesian_1995} which represent it by an explicit distribution over learned model parameters.
 Ensemble methods \parencite{gal_dropout_2016,lakshminarayanan_simple_2017} take a similar approach but use an implicit distribution.
 Other approaches include anomaly detection and the recent field of out-of-distribution detection which, for example, represent epistemic uncertainty by special output classes for out-of-distribution data, distributions over data representations, or distances between them (see \cref{sec:out-of-distribution-detection}).
-In this thesis, we will focus on the latter approach.
+In this thesis, we will focus on out-of-distribution detection for quantifying epistemic uncertainty.
 

### 2023-10-24 chapters/discussion.tex
Revised abstract and changes review paper to brief review

@@
 
-Having focused on VAEs for out-of-distribution detection in \cref{part:unsupervised-uncertainty-estimation} and on self"=supervised methods for representation learning in \cref{part:self-supervised-speech-representation-learning}, this discussion will bridge a gap between the two approaches and consider how VAEs might be enabled to learn representations more competitive with self"=supervised methods for downstream tasks.
+Having focused on VAEs for out-of-distribution detection in \cref{part:unsupervised-uncertainty-estimation} and on self"=supervised methods for representation learning in \cref{part:unsupervised-speech-representation-learning}, this discussion will bridge a gap between the two approaches and consider how VAEs might be enabled to learn representations more competitive with self"=supervised methods for downstream tasks.
 Furthermore, since the work in \cref{part:medical-applications} only had limited focus on uncertainty, we will also present and discuss the use of calibration techniques for the stroke recognition model drawing connections to how such systems are perceived and used in practice.

### 2023-10-24 chapters/introduction.tex
Revised abstract and changes review paper to brief review

@@
 After the agricultural revolution, several ancient civilizations developed mathematical techniques and algorithms to help them manage the newfound complexity of society.
-Fittingly, the first recorded human name in written history belonged to Kushim, a Sumerian accountant.
+Fittingly, the first human name recorded in written history belonged to Kushim, a Sumerian accountant.
 Their signature has been identified on at least 18 clay tablets dated to 3400-3000 BCE, which document transactions and inventory related to barley and other ingredients used in beer production \parencite{nissen_archaic_1993}.
-A later example is the Egyptian Rhind Mathematical Papyrus dated to around 1550 BCE which describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \parencite{georges_universal_2001}.
-Around the same time, the Babylonians had developed a sophisticated system of mathematics that included algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \parencite{fowler_square_1998}.
-
-% \begin{figure}
-%     \centering
-%     \includegraphics[height=0.4\textheight]{kushim_multiplication.png}
-%     \caption[]{ }
-%     \label{fig:kushim_multiplication}
-% \end{figure}
+% A later example is the Egyptian Rhind Mathematical Papyrus dated to around 1550 BCE which describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \parencite{georges_universal_2001}.
+Later, the Babylonians developed a sophisticated system of mathematics that included algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \parencite{fowler_square_1998}.
 
 Besides enabling the advancement of early society, these technologies also brought with them novel challenges. 
-In a likely attempt to ease the strenuous work of doing Sumerian mathematics, Kushim produced a clay tablet with a multiplication table that included fractional representations of common numbers.
-Assumedly unbeknownst to Kushim, this tablet would also come to contain one of the world's first recorded examples of a mathematical error; the statement: $5 \times \frac{1}{2} = 5$ \parencite{nissen_archaic_1993}.
-Later, the scribe of the Plimpton 322 clay tablet (\cref{fig:plimpton_332}) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
+In a likely attempt to ease the strenuous work of doing Sumerian mathematics, Kushim produced a clay tablet with a multiplication table that included fractional representations of common numbers (\cref{fig:kushim_multiplication}).
+Assumedly unbeknownst to Kushim, this tablet would also come to contain one of the world's first recorded examples of a mathematical error; the statement: $5 \times \sfrac{1}{2} = 5$ \parencite{nissen_archaic_1993}.
+Later, the scribe of the Plimpton 322 clay tablet made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011,cuneiformdigitallibraryinitiativecdli_mct_2005}.
 % Later, in its list of Pythagorean triplets, the Plimpton 322 clay tablet (\cref{fig:plimpton_332}) also has several errors, presumably made by a novice in Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
@@
 
+% \begin{figure}[t]
+%     \centering
+%     % \includegraphics[width=0.90\textwidth]{plimpton_322.png}
+%     \includegraphics[height=0.4\textheight]{plimpton_322.png}
+%     \caption[Technology is difficult.
+The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets.]{ Technology is difficult.
+The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics.
+\parencite[photo credit][]{neugebauer_mathematical_1945, cuneiformdigitallibraryinitiativecdli_mct_2005}}
+%     \label{fig:plimpton_332}
+% \end{figure}
+
 \begin{figure}[t]
@@
     % \includegraphics[width=0.90\textwidth]{plimpton_322.png}
-    \includegraphics[height=0.4\textheight]{plimpton_322.png}
+    \includegraphics[height=0.55\textheight]{kushim_multiplication.png}
     \caption[Technology is difficult.
-The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets.]{ Technology is difficult.
-The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics.
-\parencite[photo credit][]{neugebauer_mathematical_1945, cuneiformdigitallibraryinitiativecdli_mct_2005}}
-    \label{fig:plimpton_332}
+The accountant Kushim made mistakenly claimed $5 \times \sfrac{1}{2} = 5$.]{ Technology is difficult.
+Between 3400 and 3000 BCE, the Sumerian accountant Kushim wrote this tablet.
+Besides calculations of basic ingredients required for the production of cereal products, in the top left $6 \times 2$ grid it contains a multiplication table.
+While presumably aimed to ease the strenuous work of doing Sumerian mathematics, unfortunately, the top row states: $5 \times \sfrac{1}{2} = 5$, an unforgiving imprint the worlds first recorded mathematical error \parencite{nissen_archaic_1993, renn_learning_2019}, \parencite[photo credit][]{}.}
+    \label{fig:kushim_multiplication}
 \end{figure}
+\todo[inline]{\cref{fig:kushim_multiplication} is awaiting response from CDLI with proper reference to tablet.}
 
@@
 Refined by Gottfried Leibniz during the latter half of the 17th century, the mechanical calculator marked the beginning of a new era where rudimentary calculations could be automated and performed with greater speed and accuracy than humans.
-In the following centuries, the development of computing machines continued, and in the 1820s, Charles Babbage proposed the first programmable mechanical computer, the Difference Engine.
+In the 1820s, Charles Babbage proposed the first programmable mechanical computer, the Difference Engine.
 Although never completed, it laid the foundation for his later design of the Analytical Engine in 1837 which is considered to be the first general-purpose computer design.
@@
 % the first programmable, electronic, general-purpose digital computers were developed, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}) \parencite{georges_universal_2001}. 
-As manufacturing processes were developed and scaled, electronic computers were adopted widely across society for various applications in science and industry, including healthcare \parencite{georges_universal_2001, harari_sapiens_2011}.
+As the technology developed and manufacturing processes were refined and scaled, electronic computers were adopted widely across society for various applications in science and industry, including healthcare \parencite{georges_universal_2001, harari_sapiens_2011}.
 % The societal changes brought about by the widespread adoption of computer technology through the latter half of the 20th century were so profound that some scholars refer to the era as the information age \parencite{georges_universal_2001, harari_sapiens_2011}.
@@
     % \includegraphics[width=0.95\textwidth]{eniac_programmers.jpeg}
-    \includegraphics[height=0.4\textheight]{eniac_programmers.jpeg}
-    \caption[Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.]{ Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.
+    \includegraphics[height=0.42\textheight]{eniac_programmers.jpeg}
+    \caption[Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC, the world's first general-purpose electronic computer.]{ Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC.
 \parencite[photo credit][]{usarmyresearchlaboratoryarltechnicallibrary_female_1940}}
@@
 With the digital revolution continuing into the 21st century and the worldwide spread of the internet, the amount of data and its availability have grown exponentially. 
-This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence.
-Machine learning is the study of computer algorithms that can learn functions from data and make predictions to guide decision-making.
+This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that studies how computer algorithms can learn functions from data and make predictions to guide decision-making.
 Enabled by a massive increase in the performance and availability of parallel computing resources, such methods have set new standards for the abilities of computer systems at various tasks usually reserved for humans, such as vision, speech recognition, and natural language understanding \parencite{lecun_deep_2015}.
-In recent years, machine learning has also made its way into the healthcare sector, where it has shown promising results on tasks within medical imaging \parencite{lundervold_overview_2019}, drug discovery \parencite{chen_rise_2018}, and clinical decision support \parencite{cite15, cite14}.
+In recent years, machine learning has also made its way into healthcare systems, where it has shown promising results on difficult tasks for instance within medical imaging \parencite{lundervold_overview_2019}, drug discovery \parencite{chen_rise_2018}, and clinical decision support \parencite{cite15, cite14}.
 
@@
     \centering
-    \includegraphics[width=0.90\textwidth]{uber_nhsa_accident.png}
+    \includegraphics[width=0.85\textwidth]{uber_nhsa_accident.png}
     \caption[A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.]{ A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
@@
 
-Recently, European policymakers proposed legislation to enforce that machine learning applications live up to such standards \parencite{europeancommission_briefing_2021}. 
+Recently, European policymakers proposed legislation to enforce that machine learning applications live up to certain standards \parencite{europeancommission_briefing_2021}. 
 This so-called AI Act considers three categories of applications based on the potential risk they pose to society: limited, high, and unacceptable. 
-Limited risk applications such as in computer games must meet transparency criteria, while autonomous vehicles and systems within healthcare fall in the category of high risk applications.
+Autonomous vehicles and systems within healthcare fall in the category of high risk applications.
 To avoid critical errors, these must meet a set of strict requirements focusing on anti-discrimination and robustness.
-Facial recognition is an example of an application deemed to pose unacceptable risk to society.
+% Limited risk applications such as in computer games must meet transparency criteria, while
+% Facial recognition is an example of an application deemed to pose unacceptable risk to society.
 % Limited risk applications, such as in computer games, must meet transparency requirements, and technologies such as public facial recognition are banned due to unacceptable risk.
@@
 As a consequence, the model's many incoherent predictions were taken at face value and subsystems in the car acted on them without appropriate consideration to their reliability.
-While any model will inevitably make a wrong prediction, accurately estimating the associated uncertainty and communicating it to other systems or humans is necessary to reduce the risk that wrong predictions lead to catastrophic failures.
+While any model will inevitably make wrong predictions, accurately estimating the associated uncertainty and communicating it to other systems or humans is necessary to reduce the risk that wrong predictions lead to catastrophic failures.
 Fittingly, a recent study by the European Parliamentary Research Services \parencite{europeanparliament_artificial_2022} concluded: 
@@
 \end{quote}
-%
-\vspace{0.5em}
+% \vspace{0.5em}
 \begin{center}
@@
 \vspace{1em}
-
+%
 % OUTLINE: Primer overview of the thesis. What are the topics and what is the context of the research.
 % \noindent The research presented in this thesis will focus on the use of machine learning in relation to uncertainty estimation, speech processing, and medical conversations. 
-\noindent The research presented in this thesis deals with the use of machine learning for medical conversations with a focus on uncertainty estimation and speech processing. 
+\noindent The research presented in this thesis deals with the use of machine learning for medical interviews with a focus on uncertainty estimation and speech processing. 
 The research project was partially funded by a grant from the national Danish Innovation Fund (grant no.\@ 0153-00167B) and defined and carried out in collaboration with the Danish company Corti, which develops computer software for medical decision support.
@@
 \todo[inline]{Update here with:
-
 I would try to embed a perspective for contemplation.
@@
 In 2023, the prevalence of human errors in driving are much higher than self-driving vehicles.” <- Quick example.
-I’m sure you can find something better.
-
+I’m sure you can find something better.}
 % - a copilot for medical interviews
@@
 % -> many applications
-}
 
@@
 
-\Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. 
+\Cref{part:unsupervised-uncertainty-estimation,part:unsupervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. 
 \Cref{part:unsupervised-uncertainty-estimation} is made up of three papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
-\Cref{part:self-supervised-speech-representation-learning} is a single, comprehensive review paper on self"=supervised speech representation learning \parencite{mohamed_selfsupervised_2022}. 
+\Cref{part:unsupervised-speech-representation-learning} is a single, comprehensive review paper on self"=supervised speech representation learning \parencite{mohamed_selfsupervised_2022}. 
 \Cref{part:medical-applications} consists of two papers on applications of machine learning within the medical domain, specifically recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}.

### 2023-10-24 chapters/main-contributions.tex
Revised abstract and changes review paper to brief review

@@
 
-% The chapters are structured into three parts: {\color{black}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{black}\cref{part:self-supervised-speech-representation-learning}} is concerned with self"=supervised speech representation learning, and {\color{black}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.
+% The chapters are structured into three parts: {\color{black}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{black}\cref{part:unsupervised-speech-representation-learning}} is concerned with self"=supervised speech representation learning, and {\color{black}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.
 Finally, {\color{black}\cref{part:discussion-and-conclusion}} contains a discussion and the conclusion of the thesis.
@@
 
+\todo[inline]{Revise the location of the benchmarking paper and change the SSL review to the SSL+LVM brief review. Revise the text here accordingly.}
+
 \contribitem{chp:paper-benchmarking}
@@
 
-\contribitem{part:self-supervised-speech-representation-learning}
+\contribitem{part:unsupervised-speech-representation-learning}
 This part of the thesis is concerned with self"=supervised speech representation learning and consists of a single paper that provides a substantial review of the field \parencite{mohamed_selfsupervised_2022}.

### 2023-10-24 chapters/paper_hierarchical.tex
Revised abstract and changes review paper to brief review

@@
 
-\todo[inline]{Check out if we can make underbrace like curly brackets to the left of tab_hierarchical:rocauc-ood to decrease height and fit page better.}
+\todo[inline]{Check out if we can make underbrace like curly brackets to the left of \cref{tab_hierarchical:rocauc-ood} to decrease height and fit page better.
+See \url{https://tex.stackexchange.com/questions/62573/using-underbrace-with-table-columns}}
+
 \begin{table}

### 2023-10-25 chapters/discussion.tex
Add todo

@@
 
+\todo[inline]{Unsupervised representation learning for medical applications.
+We tried BERT for stroke but no success.
+Medical coding state-of-the-art uses pretrained BERT, but only marginally improves upon simpler baselines.}
+
 The rapid progress of machine learning that started about a decade ago with the 2012 ImageNet competition and the work of \textcite{krizhevsky_imagenet_2012} all but slowed down during the course of this project.
@@
 %
-In this thesis we studied two different approaches to speech representation learning: VAEs in \cref{chp:paper-hierarchical,chp:paper-mo?delagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-review}.
+In this thesis we studied two different approaches to speech representation learning: VAEs in \cref{chp:paper-hierarchical,chp:paper-modelagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-brief}.
 As we saw, the probabilistic formulation of VAEs provides benefits for their application to uncertainty quantification, although competitive methods that use self"=supervised foundation models have also been proposed \textcite{xiao_we_2021,hendrycks_using_2019,bergman_classificationbased_2020}.
@@
 \subsection{Masking} 
-As discussed in \cref{chp:paper-review}, masking is one of the driving techniques behind the success of self"=supervised methods for speech \parencite{devlin_bert_2018,baevski_wav2vec_2020}.
+As discussed in \cref{chp:paper-brief}, masking is one of the driving techniques behind the success of self"=supervised methods for speech \parencite{devlin_bert_2018,baevski_wav2vec_2020}.
 By removing tokens from the input or an early feature extraction layer and tasking a model with inferring their representations, models are forced to learn how neighboring tokens relate to those masked.
@@
 % % \textcite{wickstrom_relax_2023} propose a method for quantifying uncertainty based on
-
-% \section{\Cref{chp:paper-review} revisited: \dots} \label{sec:discussion-paper-review}
 

### 2023-10-25 chapters/background.tex
Change references and add todos

@@
 Since the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic} are relatively short, we also provide a concise review of the literature on out-of-distribution detection.
+
+\todo[inline]{Briefly hold the readers hand at actualize the importance of OODD.}
 
@@
 
-\todo[inline]{Briefly hold the readers hand and tell them why VAEs are relevant to the overarching project.}
+\moretodo[inline]{Briefly hold the readers hand and tell them why VAEs are relevant to the overarching project.}
 
@@
 This amortization gap can be reduced by using large capacity amortized posteriors.
-In cases where the true posterior does not belong to the model class of the variational family of posteriors that we fit, the inference gap cannot be reduced to zero and the remaining gap is referred to as the approximation gap.
+In case the true posterior does not belong to the model class of the variational family of posteriors that we fit, the inference gap cannot be reduced to zero and the remaining gap is referred to as the approximation gap.
 
-\todo[inline]{Finish this with the parsimony gap}
+\moretodo[inline]{Finish this with the parsimony gap}
 
@@
 
-\todo[inline]{Derive and introduce the IWAE bound.}
+\moretodo[inline]{Derive and introduce the IWAE bound.}
 
@@
 \paragraph{Self"=supervised learning}
+\moretodo[inline]{Revise according to change of review papers.}
 An alternative approach to representation learning is provided by self"=supervised learning \parencite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020,schneider_wav2vec_2019} which is a form of unsupervised learning where the training objective is derived from the data itself.
 While it is fair to say that not all self"=supervised training objectives are as well-motivated as the ELBO is for VAEs, they have shown impressive results within the fields of natural language processing \parencite{devlin_bert_2018,chen_simple_2020}, speech processing \parencite{schneider_wav2vec_2019}, and computer vision \parencite{chen_simple_2020}.
-For a deeper comparison of VAEs and self"=supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{app:paper-brief}. 
-\Cref{chp:paper-review} provides a comprehensive introduction to and review of self"=supervised learning in speech recognition.
+For a deeper comparison of VAEs and self"=supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{chp:paper-brief}. 
+\Cref{app:paper-review} provides a comprehensive introduction to and review of self"=supervised learning in speech recognition.

### 2023-10-25 chapters/main-contributions.tex
Change references and add todos

@@
 
-\todo[inline]{Revise the location of the benchmarking paper and change the SSL review to the SSL+LVM brief review. Revise the text here accordingly.}
+\moretodo[inline]{Revise the location of the benchmarking paper and change the SSL review to the SSL+LVM brief review. Revise the text here accordingly. $\downarrow$}
 
@@
 In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self"=supervised learning for learning representations for unsupervised uncertainty estimation.
-We shall return to this in the discussion of \cref{chp:paper-review} in \cref{sec:discussion-paper-review}.
-\todo[inline]{Edit this if we did not discuss this.}
+We shall return to this in the discussion of \cref{app:paper-review} in \cref{sec:discussion-paper-review}.
+\moretodo[inline]{Edit this if we did not discuss this.}
 An early version of this review paper included a review of variational autoencoders for speech representation learning and a comparison between the two modelling paradigms \parencite{borgholt_brief_2022}.
-Due to a desire to focus on self"=supervised methods, the paper in \cref{chp:paper-review} does not put particular focus on variational methods. 
-The early version is included in \cref{app:paper-brief} for reference.
+Due to a desire to focus on self"=supervised methods, the paper in \cref{app:paper-review} does not put particular focus on variational methods. 
+The early version is included in \cref{chp:paper-brief} for reference.
 
-\contribitem{chp:paper-review}
+\contribitem{app:paper-review}
 This paper provides a comprehensive review of self"=supervised speech representation learning.
@@
 The review also provides a discussion of the current state of the field and promising future directions of research.
+
+\moretodo[inline]{Revise the location of the benchmarking paper and change the SSL review to the SSL+LVM brief review. Revise the text here accordingly. $\uparrow$}
 

### 2023-10-26 chapters/introduction.tex
Add photo credit to Kushim

@@
 Besides calculations of basic ingredients required for the production of cereal products, in the top left $6 \times 2$ grid it contains a multiplication table.
-While presumably aimed to ease the strenuous work of doing Sumerian mathematics, unfortunately, the top row states: $5 \times \sfrac{1}{2} = 5$, an unforgiving imprint the worlds first recorded mathematical error \parencite{nissen_archaic_1993, renn_learning_2019}, \parencite[photo credit][]{}.}
+While presumably aimed to ease the strenuous work of doing Sumerian mathematics, unfortunately, the top row states: $5 \times \sfrac{1}{2} = 5$, an unforgiving imprint the worlds first recorded mathematical error \parencite{nissen_archaic_1993, renn_learning_2019}, \parencite[photo credit][]{cuneiformdigitallibraryinitiativecdli_msvo_2018}.}
     \label{fig:kushim_multiplication}

### 2023-10-26 chapters/background.tex
Updates

@@
 
-The marginal distribution $p(\xb)$ involves an integral over all possible values of the latent variable $\zb$ and is often analytically intractable.
-This is the case when the latent variable space is high-dimensional or the conditional distribution $p(\xb|\zb)$ is complex, as with non-conjugate models or neural networks.
-In these cases, approximate methods such as variational inference \parencite{jordan_introduction_1999} and Markov Chain Monte Carlo (MCMC) \parencite{mohamed_monte_2019} are used to approximately compute the marginal and posterior distributions.
+% In these cases, approximate methods such as variational inference \parencite{jordan_introduction_1999} and Markov Chain Monte Carlo (MCMC) \parencite{mohamed_monte_2019} are used to approximately compute the marginal and posterior distributions.
 
 \subsection{Variational inference} \label{sec:variational-autoencoders}
-To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to factor the joint distribution and parameterize it using a neural network with parameters $\thetab$.
+To be able to model high-dimensional data, we want to use neural networks to represent the generative process. 
+This leads to deep latent variable models which were pioneered by \textcite{kingma_autoencoding_2014,rezende_stochastic_2014}.
+The authors propose to factor the joint distribution and parameterize it using a neural network with parameters $\thetab$ and let $p_\thetab(\xb,\zb) = p_\thetab(\xb|\zb)p(\zb)$ where $p(\zb)$ is the prior.
+
+This choice makes both the marginal and posterior distributions intractable since they involve an integral over all possible values of the latent variable $\zb$. 
+This precludes training with commonly used methods such as expectation maximization and exact maximum likelihood estimation and while MCMC methods are usually too computationally expensive for large neural networks, variational inference provides a scalable alternative \parencite{jordan_introduction_1999}.
+
+% To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to factor the joint distribution and parameterize it using a neural network with parameters $\thetab$.
 Specifically, they let $p_\thetab(\xb,\zb) = p_\thetab(\xb|\zb)p(\zb)$ where $p(\zb)$ is the prior.
-This choice makes both the marginal and posterior distributions intractable and so, training and inference require approximate methods. 
-While MCMC methods are computationally expensive for large neural networks, variational inference provides a more scalable alternative by estimating the posterior distribution $p(\zb|\xb)$ with a variational approximation.
+% This choice makes both the marginal and posterior distributions intractable and so, training and inference require approximate methods. 
+% While MCMC methods are computationally expensive for large neural networks, variational inference provides a more scalable alternative by estimating the posterior distribution $p(\zb|\xb)$ with a variational approximation $q_{\xb}(z)$
 
 \paragraph{Evidence lower bound}
-VAEs amortize the cost of variational inference by parameterizing $q_\phib(\zb|\xb)$ as a single neural network.
-This makes it possible to efficiently compute approximate posterior distributions for any input $\xb$.
-The approximate posterior can also be used to compute a lower bound on the marginal distribution $p_\thetab(\xb)$ via Jensen's inequality, the so-called evidence lower bound (ELBO),
+By estimating the posterior distribution $p(\zb|\xb)$ with a variational approximation $q(\zb)$ we can derive a lower bound on the otherwise intractable marginal log-likelihood $\log p_\theta(\xb)$.
 %
@@
     &= \log \int p_\thetab(\xb,\zb),\textrm{d}\zb \nonumber \\
-    &= \log \int q_\phib(\zb|\xb) \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)}\,\textrm{d}\zb \nonumber \\
-    &\geq \int q_\phib(\zb|\xb) \log \left( \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right) \,\textrm{d}\zb \nonumber \\
-    &\geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}_{}(\xb;\thetab,\phib) \enspace .
+    &= \log \int q(\zb) \frac{p_\thetab(\xb,\zb)}{q(\zb)}\,\textrm{d}\zb \nonumber \\
+    &\geq \int q(\zb) \log \left( \frac{p_\thetab(\xb,\zb)}{q(\zb)} \right) \,\textrm{d}\zb \nonumber \\
+    &\geq \mathbb{E}_{q(\zb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q(\zb)} \right] \equiv \mathcal{L}_{}(\xb;q) \enspace .
     % &= \log \int p_\thetab(\xb|\zb)p(\zb),\textrm{d}\zb \nonumber \\
-    % &= \log \int q_\phib(\zb|\xb) \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)}\,\textrm{d}\zb \nonumber \\
-    % &\geq \int q_\phib(\zb|\xb) \log \left( \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right) \,\textrm{d}\zb \nonumber \\
-    % &\geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}_{}(\xb;\thetab,\phib) \enspace .
+    % &= \log \int q(\zb) \frac{p_\thetab(\xb|\zb)p(\zb)}{q(\zb)}\,\textrm{d}\zb \nonumber \\
+    % &\geq \int q(\zb) \log \left( \frac{p_\thetab(\xb|\zb)p(\zb)}{q(\zb)} \right) \,\textrm{d}\zb \nonumber \\
+    % &\geq \mathbb{E}_{q(\zb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q(\zb)} \right] \equiv \mathcal{L}_{}(\xb;\thetab,\phib) \enspace .
 \end{align}
 %
-Although analytically evaluating the ELBO is intractable, it can be approximated by sampling and used to train VAEs. We will return to this after examining the ELBO in detail.
+Since we can evaluate the generative model $p_\thetab(\xb,\zb)$ and are free to define $q(\zb)$ as we please, this bound is tractable. 
+However, finding an approximate posterior distribution $q(\zb)$ that is ...
+
+\moretodo[inline]{Glue this together..}
+
+Instead, VAEs amortize the cost of variational inference by parameterizing $q_\phib(\zb|\xb)$ as a single neural network.
+This makes it possible to efficiently compute approximate posterior distributions for any input $\xb$.
+The approximate posterior can also be used to compute a lower bound on the marginal distribution $p_\thetab(\xb)$ via Jensen's inequality, the so-called evidence lower bound (ELBO),
+%
+\begin{equation}
+    \log p_\thetab(\xb) \geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}_{}(\xb;\thetab,\phib) \enspace .
+\end{equation}
+%
+Although analytically evaluating the ELBO is still intractable, it can be approximated by sampling and used to train VAEs. We will return to this after examining the ELBO in more detail.
 
@@
 The ELBO is a lower bound on the marginal likelihood, so it is interesting to examine when the bound holds exactly. 
-Without using Jensen's inequality, in the below, we will rewrite the log-marginal likelihood with the goal of arriving at the ELBO of \cref{eq:variational-autoencoder-elbo} plus a factor to fill the inference gap.
-% Since the ELBO is computed as an expectation under the variational approximation to the posterior, $q_\phib(\xb|\zb)$, we will seek to rewrite the log-marginal likelihood into such an expectation, .
+As is standard for variational Bayesian methods, we can take the objective of the VAE to be minimization of the KL-divergence of the approximate posterior to the true posterior, $D_\mathrm{KL}\left( p_\thetab(\zb|\xb) \parallel q_\phib(\zb|\xb) \right)$ and rewrite it as follows.
 %
 \begin{align}
-    \log p_\thetab(\xb) 
-    &= \int q_\phib(\zb|\xb) \log p_\thetab(\xb) \, \textrm{d}\zb \nonumber \\
-    &= \int q_\phib(\zb|\xb) \log \left(\frac{p_\thetab(\xb,\zb)}{p_\thetab(\zb|\xb)}\right) \, \textrm{d}\zb \nonumber \\
-    &= \int q_\phib(\zb|\xb) \log \left(\frac{p_\thetab(\xb,\zb) q_\phib(\zb|\xb)}{p_\thetab(\zb|\xb) q_\phib(\zb|\xb)}\right) \, \textrm{d}\zb \nonumber \\
-    &= \int q_\phib(\zb|\xb) \log \left(\frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)}\right) \, \textrm{d}\zb + \int q_\phib(\zb|\xb) \log \left(\frac{q_\phib(\zb|\xb)}{p_\thetab(\zb|\xb)}\right) \, \textrm{d}\zb \nonumber \\
+    D_\text{KL}\left( p_\theta(\zb|\xb) \parallel q_\phib(\zb|\xb) \right)
+    &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[ \log \frac{p_\theta(\zb|\xb)}{q_\phib(\zb|\xb)} \right] \nonumber \\
+    &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[ \log \frac{p_\theta(\zb,\xb)}{q_\phib(\zb|\xb)p_\theta(\xb)} \right] \nonumber \\
     &= 
     \underset{\text{ELBO}}{\underbrace{
-        \mathcal{L}_{}(\xb;\thetab,\phib)
+        \mathbb{E}_{q_\phib(\zb|\xb)}\left[ \log \frac{p_\theta(\zb,\xb)}{q_\phib(\zb|\xb)} \right]
     }}
-    + 
-    \underset{\text{KL-divergence to true posterior}}{\underbrace{
-        \vphantom{\mathcal{L}_{}(\xb;\thetab,\phib)}{D_\mathrm{KL}\left( p_\thetab(\zb|\xb) \parallel q_\phib(\zb|\xb) \right)}
+    - 
+    \underset{\text{Marginal likelihood}}{\underbrace{
+        \vphantom{\mathbb{E}_{q_\phib(\zb|\xb)}\left[ \log \frac{p_\theta(\zb,\xb)}{q_\phib(\zb|\xb)} \right]}{\log p_\theta(\xb)}
     }} \enspace .
@@
 %
-We can see that the inference gap of the ELBO is exactly equal to the KL-divergence of the approximate posterior to the true posterior. 
-This also means that the approximate posterior $q_\phib(\zb|\xb)$ that maximizes the ELBO also best approximates the true posterior, as measured by $D_\mathrm{KL}\left( p(\zb|\xb) \parallel q(\zb|\xb) \right)$.
+We can see that the inference gap of the ELBO is exactly equal to the KL-divergence of the approximate posterior to the true posterior $D_\text{KL}\left( p_\theta(\zb|\xb) \parallel q_\phib(\zb|\xb) \right)$.
+This also means that the approximate posterior $q_\phib(\zb|\xb)$ that maximizes the ELBO also best approximates the true posterior, as measured by KL-divergence.
 
-We can subdivide the inference gap \cite{cremer_inference_2018}: 
+We can analyze the inference gap further \cite{cremer_inference_2018}. 
 By amortizing the variational inference we introduce a gap compared to the alternative of fitting a posterior per data point, $q^{*}_{\xb}(\zb)$.
-This amortization gap can be reduced by using large capacity amortized posteriors.
+This is referred to as the amortization gap and can be reduced by using large capacity amortized posteriors.
 In case the true posterior does not belong to the model class of the variational family of posteriors that we fit, the inference gap cannot be reduced to zero and the remaining gap is referred to as the approximation gap.
+If the inference gap is closed, the maximum of the ELBO corresponds to the maximum of the maringa
+Finally, the generative model
 
@@
 
+\moretodo[inline]{Maybe add a small graphic illustrating the gaps.}
+
 \subsection{Maximum likelihood estimation}
 %
-We want to use maximum likelihood estimation to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$. 
+VAEs are usually mixed-effect models. Although VAEs use variational Bayes to infer a posterior distribution over latent variables, their parameters are estimated using maximum likelihood. 
+That is, we want to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$ to find point estimates. 
 Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
-To do this we therefore must approximate the expectation in the ELBO \cref{eq:variational-autoencoder-elbo} and compute its gradients w.r.t. $\phib$ and $\thetab$.
+Hence, we must approximate the expectation in the ELBO \cref{eq:variational-autoencoder-elbo} and compute its gradients w.r.t. $\phib$ and $\thetab$.
 
 \paragraph{Sampling}
-Since we have freedom to select the approximate posterior, we can choose it such that it is easy to sample and then use that property to form a Monte Carlo estimate of the ELBO as
+Since we have freedom to select the approximate posterior, we can choose it such that it is easy to sample and then use that property to form a Monte Carlo estimate of the ELBO,
 % 
@@
 To reduce the variance of the estimate, \textcite{burda_importance_2016} propose to use importance sampling to reweigh the samples. 
+Compared to the estimator in \cref{eq:variational-autoencoder-elbo-sampling-expanded,eq:variational-autoencoder-elbo-sampling-KL} which
 
@@
 However, the gradient with respect to the $\phib$ is more challenging. Since $\phib$ occurs in the expectation itself, we may not move the gradient inside the expectation. 
-However, by reparameterizing the latent variable $\zb$ as a deterministic function of the input $\xb$ and a random variable $\epsilonb\sim p(\epsilonb)$, $\zb = g_\phib(\xb,\epsilonb)$, we can write the gradient as a path-wise derivative,
+However, by reparameterizing the latent variable $\zb$ as a deterministic function of the input $\xb$ and a random variable $\epsilonb\sim p(\epsilonb)$, $\zb = g_\phib(\xb,\epsilonb)$, we can write the gradient as a path-wise derivative \parencite{kingma_autoencoding_2014,rezende_stochastic_2014,mohamed_monte_2019},
 %
@@
     % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb_{L-1}|\zb_L)\cdots q_\phib(\zb_2|\zb_1),
-    q_\phib(\zb|\xb) = q_\phib(\zb_L|\xb)\prod_{i=L-1}^{1} q_\phib(\zb_{i}|\zb_{i+1}) \ .
+    q_\phib(\zb|\xb) = q_\phib(\zb_L|\xb)\prod_{i=1}^{L-1} q_\phib(\zb_{i}|\zb_{i+1}) \ .
 \end{equation}

### 2023-10-27 chapters/background.tex
Updates

@@
 In the following, we will introduce VAEs starting from the definition of latent variable models.
-We then derive the marginal likelihood objective used to train VAEs, discuss its different components, and highlight some central challenges of applying VAEs.
+We then derive the marginal likelihood objective bound used to train them, discuss its different components and the approximation error, and highlight some central challenges of applying VAEs.
 
@@
 %
-Since we can evaluate the generative model $p_\thetab(\xb,\zb)$ and are free to define $q(\zb)$ as we please, this bound is tractable. 
-However, finding an approximate posterior distribution $q(\zb)$ that is ...
+Since we can evaluate the generative model $p_\thetab(\xb,\zb)$ and are free to define $q(\zb)$ as we please, this bound can be made tractable.
 
-\moretodo[inline]{Glue this together..}
-
-Instead, VAEs amortize the cost of variational inference by parameterizing $q_\phib(\zb|\xb)$ as a single neural network.
-This makes it possible to efficiently compute approximate posterior distributions for any input $\xb$.
-The approximate posterior can also be used to compute a lower bound on the marginal distribution $p_\thetab(\xb)$ via Jensen's inequality, the so-called evidence lower bound (ELBO),
+The usual approach in variational inference is to fit an approximate posterior $q(\zb)$ per data point.
+However, for large datasets, as is common in natural language processing, this can be too computationally expensive.
+% \moretodo[inline]{Glue this together..}
+Instead, VAEs amortize the cost of variational inference by parameterizing the approximate posterior with a neural network that transforms any data point $\xb$ into the parameters of the corresponding conditional distribution, $q_\phib(\zb|\xb)$.
+This makes it possible to efficiently infer the approximate posterior distribution for any input $\xb$.
+With amortized variational inference, the ELBO can simply be written as,
+% The approximate posterior can also be used to compute a lower bound on the marginal distribution $p_\thetab(\xb)$ via Jensen's inequality, the so-called evidence lower bound (ELBO), 
 %
@@
 %
-Although analytically evaluating the ELBO is still intractable, it can be approximated by sampling and used to train VAEs. We will return to this after examining the ELBO in more detail.
+Although analytically evaluating the ELBO is intractable due to the expectation, it can be approximated by sampling and then used to train VAEs.
+We will return to this after examining the ELBO in more detail.
 

### 2023-10-27 chapters/introduction.tex
Updates

@@
 %
+%
 % OUTLINE: Primer overview of the thesis. What are the topics and what is the context of the research.
@@
 
-\todo[inline]{Update here with:
-I would try to embed a perspective for contemplation.
-Instead of us always talking about what the technology needs to improve on.
-Perhaps, it would also be prudent to ask an open-ended philosophical question: “While our requirements to the technology are in its place, one could ask, whether human beings have the ability to quantify uncertainty.
-In 2023, the prevalence of human errors in driving are much higher than self-driving vehicles.” <- Quick example.
-I’m sure you can find something better.}
+\moretodo[inline]{Update description of Corti. Focus on the current year. }
+
 % - a copilot for medical interviews

### 2023-10-31 chapters/background.tex
Update background with IWAE

@@
 
-\todo[inline]{Come up with a suitable example explaining the intuition behind information entropy.}
+\lesstodo[inline]{Come up with a suitable example explaining the intuition behind information entropy.}
 
@@
 \section{Out-of-distribution detection} \label{sec:out-of-distribution-detection}
-\todo[inline]{Maybe this section should come after the VAE background?}
+\lesstodo[inline]{Maybe this section should come after the VAE background?}
 % 
@@
 
-\todo[inline]{Briefly hold the readers hand at actualize the importance of OODD.}
+\moretodo[inline]{Briefly hold the readers hand at actualize the importance of OODD.}
 
@@
 Each training example is then the correct target for one group and an OOD example for all other groups; a kind of hierarchical version of the OOD class of simpler baselines based on noise augmentation \parencite{ren_likelihood_2019}.
-\todo[inline]{Does the previous sentence belong here or in a different paragraph? It is not really using the latent space.}
+\moretodo[inline]{Does the previous sentence belong here or in a different paragraph? It is not really using the latent space.}
 Similarly, to represent a virtual OOD class, \textcite{wang_vim_2022} generate an additional logit by first computing the residual of the input's latent space representation against a principal feature space and then converting it to a valid logit by matching its mean over training samples to the average maximum logits.
@@
 \subsection{Variational inference} \label{sec:variational-autoencoders}
-To be able to model high-dimensional data, we want to use neural networks to represent the generative process. 
+To model high-dimensional data, we want to use neural networks to represent the generative process. 
 This leads to deep latent variable models which were pioneered by \textcite{kingma_autoencoding_2014,rezende_stochastic_2014}.
-The authors propose to factor the joint distribution and parameterize it using a neural network with parameters $\thetab$ and let $p_\thetab(\xb,\zb) = p_\thetab(\xb|\zb)p(\zb)$ where $p(\zb)$ is the prior.
+The authors propose to factor the joint distribution and parameterize it using a neural network with parameters $\thetab$ and let $p_\thetab(\xb,\zb) = p_\thetab(\xb|\zb)p(\zb)$ where $p(\zb)$ is a prior.
 
 This choice makes both the marginal and posterior distributions intractable since they involve an integral over all possible values of the latent variable $\zb$. 
-This precludes training with commonly used methods such as expectation maximization and exact maximum likelihood estimation and while MCMC methods are usually too computationally expensive for large neural networks, variational inference provides a scalable alternative \parencite{jordan_introduction_1999}.
+This precludes training with commonly used methods such as expectation maximization and exact maximum likelihood estimation.
+Since Markov Chain Monte Carlo methods are usually too computationally expensive for large neural networks, this leads to variational inference as a scalable alternative \parencite{jordan_introduction_1999}.
 
@@
 \paragraph{Evidence lower bound}
-By estimating the posterior distribution $p(\zb|\xb)$ with a variational approximation $q(\zb)$ we can derive a lower bound on the otherwise intractable marginal log-likelihood $\log p_\theta(\xb)$.
+By approximating the intractable posterior distribution $p(\zb|\xb)$ with a variational distribution $q(\zb)$, we can derive a lower bound on the also intractable marginal log-likelihood $\log p_\theta(\xb)$.
 %
- \begin{align} \label{eq:variational-autoencoder-elbo}
+ \begin{align}
     \log p_\thetab(\xb)
     &= \log \int p_\thetab(\xb,\zb),\textrm{d}\zb \nonumber \\
-    &= \log \int q(\zb) \frac{p_\thetab(\xb,\zb)}{q(\zb)}\,\textrm{d}\zb \nonumber \\
+    &= \log \int q(\zb) \frac{p_\thetab(\xb,\zb)}{q(\zb)}\,\textrm{d}\zb  \label{eq:marginal-likelihood-with-variational-approximation} \\
     &\geq \int q(\zb) \log \left( \frac{p_\thetab(\xb,\zb)}{q(\zb)} \right) \,\textrm{d}\zb \nonumber \\
-    &\geq \mathbb{E}_{q(\zb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q(\zb)} \right] \equiv \mathcal{L}_{}(\xb;q) \enspace .
+    &= \mathbb{E}_{q(\zb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q(\zb)} \right] \equiv \mathcal{L}_{}(\xb;\thetab,q) \enspace . \label{eq:latent-variable-model-elbo}
     % &= \log \int p_\thetab(\xb|\zb)p(\zb),\textrm{d}\zb \nonumber \\
@@
 However, for large datasets, as is common in natural language processing, this can be too computationally expensive.
-% \moretodo[inline]{Glue this together..}
-Instead, VAEs amortize the cost of variational inference by parameterizing the approximate posterior with a neural network that transforms any data point $\xb$ into the parameters of the corresponding conditional distribution, $q_\phib(\zb|\xb)$.
+Instead, VAEs amortize the cost of variational inference by parameterizing the approximate posterior with a neural network that transforms any data point $\xb$ into the parameters of the corresponding conditional distribution.
+We denote this by $q_\phib(\zb|\xb)$ where $\phib$ are the parameters of this inference model.
 This makes it possible to efficiently infer the approximate posterior distribution for any input $\xb$.
-With amortized variational inference, the ELBO can simply be written as,
+With amortized variational inference, the ELBO is simply be written as,
 % The approximate posterior can also be used to compute a lower bound on the marginal distribution $p_\thetab(\xb)$ via Jensen's inequality, the so-called evidence lower bound (ELBO), 
 %
-\begin{equation}
-    \log p_\thetab(\xb) \geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}_{}(\xb;\thetab,\phib) \enspace .
+\begin{equation} \label{eq:variational-autoencoder-elbo}
+    \mathcal{L}_{}(\xb;\thetab,\phib) \equiv \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \leq \log p_\thetab(\xb) \enspace .
 \end{equation}
@@
 \paragraph{KL-divergence to prior}
-Since the joint factorizes in VAEs, we can also write the ELBO in \cref{eq:variational-autoencoder-elbo} as,
+Since VAEs factorize the joint distribution, we can also write the ELBO in \cref{eq:variational-autoencoder-elbo} as,
 %
@@
 We can analyze the inference gap further \cite{cremer_inference_2018}. 
-By amortizing the variational inference we introduce a gap compared to the alternative of fitting a posterior per data point, $q^{*}_{\xb}(\zb)$.
-This is referred to as the amortization gap and can be reduced by using large capacity amortized posteriors.
-In case the true posterior does not belong to the model class of the variational family of posteriors that we fit, the inference gap cannot be reduced to zero and the remaining gap is referred to as the approximation gap.
-If the inference gap is closed, the maximum of the ELBO corresponds to the maximum of the maringa
-Finally, the generative model
+By amortizing the variational inference we introduce a gap compared to the alternative of fitting a posterior per data point. %, $q^{*}_{\xb}(\zb)$.
+This is referred to as the \emph{amortization gap} and can be reduced by using large capacity amortized posteriors. 
+In case the true posterior does not belong to the chosen class of variational posteriors, the inference gap cannot be reduced to zero even with infinite capacity.
+The remaining gap is referred to as the \emph{approximation gap}.
+If the inference gap is closed, the ELBO becomes equal to the marginal likelihood, the value of which then depends on how well the generative model is fitted \cite{cremer_inference_2018,frellsen_deep_2019}.
 
-\moretodo[inline]{Finish this with the parsimony gap}
-
-A final error is introduced  the maximally attainable likelihood over all the group $G$ of possible models, $\mathcal{l}(G)$
-
-It is important to also note that in fitting a VAE, the approximation gap is not the only source of error.
-
-\moretodo[inline]{Maybe add a small graphic illustrating the gaps.}
+\lesstodo[inline]{Do we want to refer to the parsimony gap?}
+\lesstodo[inline]{Maybe add a small graphic illustrating the gaps?}
+% A final error is introduced  the maximally attainable likelihood over all the group $G$ of possible models, $\mathcal{l}(G)$ 
+% It is important to also note that in fitting a VAE, the approximation gap is not the only source of error.
 
@@
 \paragraph{Sampling}
-Since we have freedom to select the approximate posterior, we can choose it such that it is easy to sample and then use that property to form a Monte Carlo estimate of the ELBO,
+Since we have freedom to select the approximate posterior, we can choose it such that it is easy to sample and then use that property to form a Monte Carlo estimate of the ELBO, $\mathcal{L}_{}(\xb;\thetab,\phib)$,
 % 
 \begin{equation} \label{eq:variational-autoencoder-elbo-sampling-expanded}
-    \mathcal{L}_{}(\xb;\thetab,\phib)
-    \approx \frac{1}{K} \sum_{k=1}^K \log p_\thetab(\xb|\zb_k) + \log p(\zb_k) - \log q_\phib(\zb_k|\xb) \enspace , 
+    % \mathcal{L}_{}(\xb;\thetab,\phib)
+    % \approx 
+    \widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)
+    =
+    \frac{1}{S} \sum_{s=1}^{S} \log \frac{p_\thetab(\xb|\zb_{s}) p(\zb_{s})}{q_\phib(\zb_{s}|\xb)} \enspace , 
+    % \frac{1}{S} \sum_{s=1}^S \log p_\thetab(\xb|\zb_{s}) + \log p(\zb_{s}) - \log q_\phib(\zb_{s}|\xb) \enspace , 
 \end{equation}
 %
-where $\zb_k \sim q_\phib(\zb|\xb)$ are samples from the approximate posterior. 
-For some choices of prior and approximate posterior, such as Gaussians, the KL-divergence term of the ELBO can be computed analytically. 
-This Monte Carlo estimator can be written as,
+where $\zb_{s} \sim q_\phib(\zb|\xb)$ are samples from the approximate posterior. 
+For some choices of prior and approximate posterior, such as Gaussians, the KL-divergence term of the ELBO can be computed analytically which reduces estimator variance. 
+This estimator can be written as,
 %
 \begin{equation} \label{eq:variational-autoencoder-elbo-sampling-KL}
-    \mathcal{L}_{}(\xb;\thetab,\phib)
-    \approx \frac{1}{K} \sum_{k=1}^K \log p_\thetab(\xb|\zb_k) - D_\text{KL}\left( q_\phib(\zb|\xb) \,||\, p(\zb) \right) \enspace .
+    \widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)
+    \equiv
+    \frac{1}{S} \sum_{s=1}^{S} \log p_\thetab(\xb|\zb_{s}) - D_\text{KL}\left( q_\phib(\zb|\xb) \,||\, p(\zb) \right)\enspace .
 \end{equation}
 %
-To reduce the variance of the estimate, \textcite{burda_importance_2016} propose to use importance sampling to reweigh the samples. 
-Compared to the estimator in \cref{eq:variational-autoencoder-elbo-sampling-expanded,eq:variational-autoencoder-elbo-sampling-KL} which
+Since $\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)$ is the sample average, it is an unbiased estimator of the ELBO, $\mathbb{E}\left[\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)\right] = \mathcal{L}_{S}(\xb;\thetab,\phib)$, and hence a biased estimator of the marginal likelihood, $\log p(\xb)$.
 
-\moretodo[inline]{Derive and introduce the IWAE bound.}
+Whereas we arrive at $\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)$ by first deriving the ELBO as a lower bound on the marginal likelihood and then estimating that bound by Monte Carlo sampling, \textcite{burda_importance_2016} instead propose to directly Monte Carlo sample the expectation in the logarithm of the intractable marginal likelihood \cref{eq:marginal-likelihood-with-variational-approximation},
+%
+\begin{equation}
+    \widehat{\mathcal{L}}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \equiv \log \left( \frac{1}{S}\sum_{s=1}^{S} \frac{p_\thetab(\xb|\zb_{s})p(\zb_{s})}{q_\phib(\zb_{s}|\xb)} \right) = \frac{1}{S} \sum_{s=1}^S \log w_\thetab(\xb,\zb_{s}) \enspace ,
+\end{equation}
+%
+where $w_\thetab(\xb,\zb_{s})$ are the importance weights. Note the difference with the estimator in \cref{eq:variational-autoencoder-elbo-sampling-expanded}. 
+By taking the expectation of this estimator, we can see that it too is a lower bound on the marginal likelihood,
+%
+\begin{align}
+    \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib)
+    &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[ \widehat{\mathcal{L}}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \right] \nonumber \\
+    &= \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \frac{1}{S}\sum_{s=1}^{S} \log \left( w_\thetab(\xb,\zb_{s}) \right) \right] \nonumber \\
+    &\leq \log \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \frac{1}{S}\sum_{s=1}^{S} w_\thetab(\xb,\zb_{s}) \right] \enspace . \nonumber \\
+    &= \log p(\xb) \enspace .
+\end{align}
+%
+From this and the law of large numbers, it can be shown that for all $S$,
+%
+\begin{equation}
+    \log p(\xb) \geq \mathcal{L}^{\text{IWAE}}_{S+1}(\xb;\thetab,\phib) \geq \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \enspace ,
+\end{equation}
+%
+and $\lim_{S\rightarrow\infty} \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \rightarrow \log p(\xb)$, if $w_\thetab(\xb,\zb_{s})$ are bounded. 
+Hence, $\widehat{\mathcal{L}}^{\text{IWAE}}_{S}$ is a biased estimator of $\log p(\xb)$ but with a bias that goes to zero as the number of samples goes to infinity. 
+With $S=1$, the estimator is equivalent to the regular ELBO \cref{eq:variational-autoencoder-elbo} and so is always as tight or tighter than the ELBO \parencite{burda_importance_2016}.
 
@@
 By assuming the covariance matrix is diagonal, this becomes a mean-field approximation which imposes independence between the dimensions of $\zb$. 
+The gradient for $\mathcal{L}^{\text{IWAE}}_{S}$ is derived similarly.
 
@@
 The drawbacks of mean-field approximation and linear covariance have lead to research into learning hierarchies of several non-linearly dependent latent variables; an idea well in line with the usual motivations behind deep neural networks such as efficient, compositional representation \cite{lecun_deep_2015}.
-Such hierarchical models are usually formalized by introducing a set of $L$ latent variables  $\zb=\zb_1, \dots, \zb_L$ and letting the generative model be defined as,
+Such hierarchical models are usually formalized by introducing a set of $L$ latent variables  $\zb=\zb^{(1)}, \dots, \zb^{(L)}$ and letting the generative model be defined as,
 %
 \begin{equation}
-    p_\thetab(\xb|\zb)=p(\xb|\zb_1) p_\thetab(\zb_1|\zb_2) \cdots p_\thetab(\zb_{L-1}|\zb_L) p(\zb_L) \enspace .
+    p_\thetab(\xb|\zb)=p(\xb|\zb^{(1)}) p_\thetab(\zb^{(1)}|\zb^{(2)}) \cdots p_\thetab(\zb^{(L-1)}|\zb^{(L)}) p(\zb^{(L)}) \enspace .
 \end{equation}
 %
-This top-down generative model can be efficiently sampled via ancestral sampling; first $\zb_L$ is drawn from the prior $p(\zb_L)$, and then each $\zb_l$ is drawn from the corresponding $p(\zb_l|\zb_{l+1})$ until we can draw $p(\xb|\zb_1)$.
+This top-down generative model can be efficiently sampled via ancestral sampling; first $\zb^{(L)}$ is drawn from the prior $p(\zb^{(L)})$, and then each $\zb^{(l)}$ is drawn from the corresponding $p(\zb^{(l)}|\zb^{(l+1)})$ until we can draw $p(\xb|\zb^{(1)})$.
 
@@
 \begin{equation}
-    %q_\phib(\zb|\xb)=q_\phib(\zb_1|\xb)q_\phib(\zb_2|\zb_1)\cdots q_\phib(\zb_L|\zb_{L-1})
-    q_\phib(\zb|\xb) = q_\phib(\zb_1|\xb)\prod_{i=2}^{L} q_\phib(\zb_i|\zb_{i-1})
+    %q_\phib(\zb|\xb)=q_\phib(\zb^{(1)}|\xb)q_\phib(\zb^{(2)}|\zb^{(1)})\cdots q_\phib(\zb^{(L)}|\zb^{(L-1)})
+    q_\phib(\zb|\xb) = q_\phib(\zb^{(1)}|\xb)\prod_{i=2}^{L} q_\phib(\zb^{(i)}|\zb^{(i-1)})
 \end{equation}
@@
 \begin{equation}
-    % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb_{L-1}|\zb_L)\cdots q_\phib(\zb_2|\zb_1),
-    q_\phib(\zb|\xb) = q_\phib(\zb_L|\xb)\prod_{i=1}^{L-1} q_\phib(\zb_{i}|\zb_{i+1}) \ .
+    % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb^{(L-1)}|\zb_L)\cdots q_\phib(\zb_2|\zb^{(1)}),
+    q_\phib(\zb|\xb) = q_\phib(\zb^{(L)}|\xb)\prod_{i=1}^{L-1} q_\phib(\zb^{(i)}|\zb^{(i+1)}) \ .
 \end{equation}
@@
 Recent architectural advances have alleviated the posterior collapse problem and made it possible to train VAEs with many latent variables \parencite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
+A key modelling choice in these works is conditioning each latent variable directly on the input via residual connections.
 
@@
 This result indicates that the usual interpretation of $\zb$ as a representation of $\xb$ is not a fundamental property arising from the ELBO and challenges the usefulness of variational autoencoders as representation learners.
-% %
 
@@
 While it is fair to say that not all self"=supervised training objectives are as well-motivated as the ELBO is for VAEs, they have shown impressive results within the fields of natural language processing \parencite{devlin_bert_2018,chen_simple_2020}, speech processing \parencite{schneider_wav2vec_2019}, and computer vision \parencite{chen_simple_2020}.
-For a deeper comparison of VAEs and self"=supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{chp:paper-brief}. 
-\Cref{app:paper-review} provides a comprehensive introduction to and review of self"=supervised learning in speech recognition.
+We provide a review and comparison of self"=supervised methods and variational autoencoders for speech representation learning in \cref{chp:paper-brief}. 
+In \cref{app:paper-review} we provide a comprehensive introduction to and review of self"=supervised learning in speech recognition. 
+% For a deeper comparison of VAEs and self"=supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{chp:paper-brief}.

### 2023-10-31 chapters/introduction.tex
Update thesis scope and outline

@@
 \end{figure}
-\todo[inline]{\cref{fig:kushim_multiplication} is awaiting response from CDLI with proper reference to tablet.}
 
@@
 These methods are generally simple and effective, but not without limitations. 
-To perform the calibration, most of the methods require a held-out validation set on which the model was not trained, Platt scaling and isotonic regression are not naively applicable to multi-class classification problems, and how to calibrate models for structured prediction tasks, such as speech recognition and machine translation, remains an open problem \parencite{astudillo_uncertainty_2010, astudillo_integration_2013, jayashankar_detecting_2020}.
+To perform the calibration, most of the methods require a held-out validation set on which the model was not trained, Platt scaling and isotonic regression are not directly applicable to multi-class classification problems, and how to calibrate models for structured prediction tasks, such as speech recognition and machine translation, remains an open problem \parencite{astudillo_uncertainty_2010, astudillo_integration_2013, jayashankar_detecting_2020}.
 
@@
 This introduction provided a high-level overview of the motivation behind the research project by focusing on the use-cases of stroke recognition and automated medical coding as well as speech processing and the importance of uncertainty quantification.
+
 The remainder of \cref{part:background} consists of two chapters. 
@@
 \Cref{part:unsupervised-uncertainty-estimation,part:unsupervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. 
-\Cref{part:unsupervised-uncertainty-estimation} is made up of three papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}.
-\Cref{part:unsupervised-speech-representation-learning} is a single, comprehensive review paper on self"=supervised speech representation learning \parencite{mohamed_selfsupervised_2022}. 
+\Cref{part:unsupervised-uncertainty-estimation} is made up of two papers dealing with out-of-distribution detection using generative models, including variational autoencoders \parencite{havtorn_hierarchical_2021,bergamin_modelagnostic_2022}.
+\Cref{part:unsupervised-speech-representation-learning} consists of two papers that explore speech representation learning with variational autoencoders and self"=supervised methods \parencite{borgholt_brief_2022,havtorn_benchmarking_2022}.
 \Cref{part:medical-applications} consists of two papers on applications of machine learning within the medical domain, specifically recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}.
 
-Finally, \cref{part:discussion-and-conclusion} concludes the thesis by discussing the presented work and presenting future directions for research.
+Finally, \cref{part:discussion-and-conclusion} concludes the thesis by discussing the presented work and future directions for research.

### 2023-10-31 chapters/main-contributions.tex
Minor edits

@@
 
-The following chapters of the thesis are self-contained studies and therefore detail their own research hypotheses and contributions.
+The chapters of \cref{part:unsupervised-uncertainty-estimation,part:unsupervised-speech-representation-learning,part:medical-applications} are self-contained studies and therefore detail their own research hypotheses and contributions.
 Since each study was written without consideration to the other chapters of the thesis, we here detail them in relation to the overall research project. 
@@
 Later, in the discussion, we shall further consider uncertainty estimation in relation to these two papers.
-\todo[inline]{Revise and add more details to this section. Do we discuss uncertainty for medical coding?}
+\moretodo[inline]{Revise and add more details to this section. Do we discuss uncertainty for medical coding?}
 

### 2023-10-31 chapters/paper_modelagnostic.tex
Fix margin violation

@@
 Historically, \textcite{bishop_novelty_1994} proposed to use a one-sided threshold on the log-likelihoods of a learned model as a decision rule to identify outliers in a dataset.
-However, recently, \textcite{nalisnick_deep_2019,hendrycks_deep_2019} showed that state-of-the-art deep generative models (DGMs) failed in this task, assigning higher a likelihood to out-of-distribution (OOD) data than in-distribution data.
+However, recently, \textcite{nalisnick_deep_2019,hendrycks_deep_2019} showed that state-of-the-art deep generative models (DGMs) failed in this task, assigning higher a likelihood to out-of-distribution (OOD) data than in"-distribution data.
 Most of the recent works focused on proposing new test statistics to alleviate the problem of using the plain likelihood, see \cref{sec_modelagnostic:related_works} for details.

### 2023-11-17 chapters/background.tex
Fix notation for layer-number and timestep

@@
     \begin{equation*}
-        H_{N\rightarrow\infty}(y(x)) =  = - \int_{\mathcal{Y}} \widetilde{w}(y)\,\frac{dy}{dx} \log \frac{\widetilde{w}(y)\frac{dy}{dx}}{\widetilde{q}(y)\frac{dy}{dx}} \, dx = - \int_{\mathcal{Y}} w(y) \log \frac{w(y)}{q(y)} \, dy \enspace .
+        H_{N\rightarrow\infty}(y(x)) =  = - \int_{\mathcal{Y}} \widetilde{w}(y)\,\frac{dy}{dx} \log \left( \frac{\widetilde{w}(y)\frac{dy}{dx}}{\widetilde{q}(y)\frac{dy}{dx}} \right) \, dx = - \int_{\mathcal{Y}} w(y) \log \left( \frac{w(y)}{q(y)} \right) \, dy \enspace .
     \end{equation*}
@@
 Since the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic} are relatively short, we also provide a concise review of the literature on out-of-distribution detection.
-
-\moretodo[inline]{Briefly hold the readers hand at actualize the importance of OODD.}
 
@@
 
-% No OOD data, probs
+% No OOD data
 \paragraph{Methods not using OOD data}
@@
 
+% No OOD data, probs
 A baseline approach uses the maximum class probability of $p(y|\xb)$ directly by noting that it tends to be larger for correctly classified examples \parencite{hendrycks_baseline_2017}.
@@
 Other methods that derive the score from the classifier probabilities include \textcite{lakshminarayanan_simple_2017} who propose to use an ensemble of independently trained classifiers to discriminate between in-distribution and OOD data by evaluating the agreement between the classifiers, \textcite{devries_learning_2018} who augment the network with a confidence estimation branch that learns to estimate the confidence of the classifier separately from the probability, and \textcite{huang_importance_2021} who compute the gradient of the KL-divergence of the predictive distribution $p(y|\xb)$ to a uniform distribution noting that the magnitude of gradients is higher for in-distribution data than for OOD data.
+\textcite{huang_mos_2021} group the classes of the target variable $y$ and define an OOD class for each group.
+Each training example is then the correct target for one group and an OOD example for all other groups; a kind of hierarchical version of the OOD class of simpler baselines based on noise augmentation \parencite{ren_likelihood_2019}.
 The Variational Information Bottleneck \parencite{alemi_deep_2017} jointly learns a probabilistic latent representation, $p(\zb|\xb)$, and a classifier, $p(y|\xb)$, using a generalized variational autoencoder \parencite{kingma_autoencoding_2014} and tries to maximize the mutual information between $\zb$ and $y$ and minimize it between $\zb$ and $\xb$.
 
-\textcite{hsu_generalized_2020} proposes a generalized version of ODIN that removes the need for simulating OOD data.
+\textcite{hsu_generalized_2020} propose a generalized version of ODIN that removes the need for simulating OOD data.
 The authors note that most current methods make a closed world assumption and implicitly condition on the in-domain $\mathcal{D}_{\text{in}}$ in the form of the predictive distribution $p(y|\xb, \mathcal{D}_{\text{in}})$.
@@
 
-% No OOD, feature
+% No OOD data, feature
 The final category of supervised methods derive the score from a latent representation of the input. 
-Some methods define OOD classes. 
-\textcite{huang_mos_2021} group the classes of the target variable $y$ and defines an OOD class for each group.
-Each training example is then the correct target for one group and an OOD example for all other groups; a kind of hierarchical version of the OOD class of simpler baselines based on noise augmentation \parencite{ren_likelihood_2019}.
-\moretodo[inline]{Does the previous sentence belong here or in a different paragraph? It is not really using the latent space.}
-Similarly, to represent a virtual OOD class, \textcite{wang_vim_2022} generate an additional logit by first computing the residual of the input's latent space representation against a principal feature space and then converting it to a valid logit by matching its mean over training samples to the average maximum logits.
-
+For instance, to represent a virtual OOD class, \textcite{wang_vim_2022} generate an additional logit by first computing the residual of the input's latent space representation against a principal feature space and then converting it to a valid logit by matching its mean over training samples to the average maximum logits.
 Other methods note that the difficulty of detecting OOD data might be attributed to the curse of dimensionality in the learned feature spaces and propose to use dimensionality reduction techniques.

### 2023-11-17 chapters/paper_hierarchical.tex
Fix notation for layer-number and timestep

@@
         Reconstruction quality of OOD data is comparable to in-distribution data, resulting in high likelihoods and poor OOD discrimination.
-        By sampling the $k$ bottom-most latent variables from the conditional prior distribution $p(\zb_{\geq l}|\zb_{>l})$ (latent reconstructions) instead of the approximate posterior $q(\zb_{>l}|\zb_{<l})$, the model reconstructs from the training distribution resulting in lower $p(\xb|\zb)$ for OOD data.
+        By sampling the $k$ bottom-most latent variables from the conditional prior distribution $p_\thetab(\zb^{(\geq l)}|\zb^{(>l)})$ (latent reconstructions) instead of the approximate posterior $q_\phib(\zb^{(>l)}|\zb^{(<l)})$, the model reconstructs from the training distribution resulting in lower $p_\thetab(\xb|\zb)$ for OOD data.
     }
@@
 \subsection{Hierarchical variational autoencoders}\label{sec_paper_hierarchical:background-hie-VAE}
-Hierarchical VAEs are a family of probabilistic latent variable models which extends the basic VAE by introducing a hierarchy of $L$ latent variables $\zb=\zb_1, \dots, \zb_L$.
-The most common generative model is defined from the top down as $p_\thetab(\xb|\zb)=p(\xb|\zb_1)p_\thetab(\zb_1|\zb_2)\cdots p_\thetab(\zb_{L-1}|\zb_L)$.
+Hierarchical VAEs are a family of probabilistic latent variable models which extends the basic VAE by introducing a hierarchy of $L$ latent variables $\zb=\zb^{(1)}, \dots, \zb^{(L)}$.
+The most common generative model is defined from the top down as $p_\thetab(\xb|\zb)=p_\thetab(\xb|\zb^{(1)})p_\thetab(\zb^{(1)}|\zb^{(2)})\cdots p_\thetab(\zb^{(L-1)}|\zb^{(L)})$.
 The inference model can then be defined in two ways respectively referred to as \textit{bottom-up} \parencite{burda_importance_2016}
 \begin{equation}
-    %q_\phib(\zb|\xb)=q_\phib(\zb_1|\xb)q_\phib(\zb_2|\zb_1)\cdots q_\phib(\zb_L|\zb_{L-1})
-    q_\phib(\zb|\xb) = q_\phib(\zb_1|\xb)\textstyle\prod_{i=2}^{L} q_\phib(\zb_i|\zb_{i-1})
+    %q_\phib(\zb|\xb)=q_\phib(\zb^{(1)}|\xb)q_\phib(\zb^{(2)}|\zb^{(1)})\cdots q_\phib(\zb^{(L)}|\zb^{(L-1)})
+    q_\phib(\zb|\xb) = q_\phib(\zb^{(1)}|\xb)\textstyle\prod_{i=2}^{L} q_\phib(\zb^{(i)}|\zb^{(i-1)})
 \end{equation}
@@
 \begin{equation}
-    % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb_{L-1}|\zb_L)\cdots q_\phib(\zb_2|\zb_1),
-    q_\phib(\zb|\xb) = q_\phib(\zb_L|\xb)\textstyle\prod_{i=L-1}^{1} q_\phib(\zb_{i}|\zb_{i+1}) \ .
+    % q_\phib(\zb|\xb)=q_\phib(\zb^{(L)}|\xb)q_\phib(\zb^{(L-1)}|\zb^{(L)})\cdots q_\phib(\zb^{(2)}|\zb^{(1)}),
+    q_\phib(\zb|\xb) = q_\phib(\zb^{(L)}|\xb)\textstyle\prod_{i=L-1}^{1} q_\phib(\zb^{(i)}|\zb^{(i+1)}) \ .
 \end{equation}
@@
 \subsection{A bound for semantic OOD detection}
-If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \crefrange{fig_hierarchical:reconstructions-fashionmnist}{fig_hierarchical:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb_1)$ will be high for both in- and out-of-distribution data.
-Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb_1)] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ from \cref{eq_hierarchical:elbo} relies on the KL-term for OOD detection.
+If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \crefrange{fig_hierarchical:reconstructions-fashionmnist}{fig_hierarchical:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb^{(1)})$ will be high for both in- and out-of-distribution data.
+Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb^{(1)})] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ from \cref{eq_hierarchical:elbo} relies on the KL-term for OOD detection.
 For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel p(\zb))$ can be expressed by a hierarchical sum,% over the hierarchy
 % \begin{multline}
-%     \mathcal{L}(\xb) =  \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \log p_\thetab(\xb|\zb_1) \\
-%                      + \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} 
-%                      + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
+%     \mathcal{L}(\xb) =  \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \log p_\thetab(\xb|\zb^{(1)}) \\
+%                      + \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb^{(i)}|\zb^{(i+1)})}{q_\phib(\zb^{(i)}|\zb^{(i-1)})} 
+%                      + \log \frac{p_\thetab(\zb^{(L)})}{q_\phib(\zb^{(L)}|\zb^{(L-1)})} \Big] .
 % \end{multline}
 % \begin{multline}
-%     D_{\mathrm{KL}}( q(\zb|\xb) \parallel p(\zb)) = \\
-%     \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
+%     D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel p(\zb)) = \\
+%     \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb^{(i)}|\zb^{(i+1)})}{q_\phib(\zb^{(i)}|\zb^{(i-1)})} + \log \frac{p_\thetab(\zb^{(L)})}{q_\phib(\zb^{(L)}|\zb^{(L-1)})} \Big] .
 % \end{multline}
 \begin{equation}
-    \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] \ .
+    \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb^{(i)}|\zb^{(i+1)})}{q_\phib(\zb^{(i)}|\zb^{(i-1)})} + \log \frac{p_\thetab(\zb^{(L)})}{q_\phib(\zb^{(L)}|\zb^{(L-1)})} \Big] \ .
 \end{equation}
-In general, the absolute log-ratios grow with $\mathrm{dim}(\zb_i)$ as the individual log probability terms are computed by summing over the dimensionality of $\zb_i$.
-This means that the value of the KL-term is dominated by terms where $\zb_i$ is high-dimensional. We refer to \cref{sec_hierarchical:analysis} for a more detailed argument.
+In general, the absolute log-ratios grow with $\mathrm{dim}(\zb^{(i)})$ as the individual log probability terms are computed by summing over the dimensionality of $\zb^{(i)}$.
+This means that the value of the KL-term is dominated by terms where $\zb^{(i)}$ is high-dimensional. We refer to \cref{sec_hierarchical:analysis} for a more detailed argument.
 Since hierarchical VAEs are generally constructed with a bottleneck type structure, the terms corresponding to latent variables towards the top of the hierarchy will have a vanishing influence on the value of the KL-term.
@@
 \begin{equation}\label{eq_hierarchical:biva->k}
-    \mathcal{L}^{>k} = \mathbb{E}_{p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p_\thetab(\zb_{>k})}{q_\phib(\zb_{>k}|\xb)} \right]
+    \mathcal{L}^{>k} = \mathbb{E}_{p_\thetab(\zb^{(\leq k)}|\zb^{(>k)}) q_\phib(\zb^{(>k)}|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p_\thetab(\zb^{(>k)})}{q_\phib(\zb^{(>k)}|\xb)} \right]
 \end{equation}
@@
 We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq_hierarchical:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$, although this need not hold in general.
-The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\zb_1,\dots,\zb_l \sim p_\thetab(\zb_{\leq k}|\zb_{>k})$ and only the $L-k$ highest from the approximate posterior $\zb_{k+1},\dots,\zb_L \sim q_\phib(\zb_{>k}|\xb)$.
-Importantly, this has the effect that the data likelihood $p(\xb|\zb)$ is dependent on the approximate posterior through a latent variable $\zb_{k+1}$ different from $\zb_1$ for all $k \geq 1$.
-Thereby, the likelihood can be evaluated with a reconstruction from each of the latent variables $\zb_k$ of the hierarchical VAE.
+The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\zb^{(1)},\dots,\zb^{(k)} \sim p_\thetab(\zb^{(\leq k)}|\zb^{(>k)})$ and only the $L-k$ highest from the approximate posterior $\zb^{(k+1)},\dots,\zb^{(L)} \sim q_\phib(\zb^{(>k)}|\xb)$.
+Importantly, this has the effect that the data likelihood $p(\xb|\zb)$ is dependent on the approximate posterior through a latent variable $\zb^{(k+1)}$ different from $\zb^{(1)}$ for all $k \geq 1$.
+Thereby, the likelihood can be evaluated with a reconstruction from each of the latent variables $\zb^{(k)}$ of the hierarchical VAE.
 Hence, we can now test how well the input $\xb$ is reconstructed from each latent variable.
-The notation $\mathcal{L}^{>k}$ highlights that for latent variables $\zb_{>k}$, the bound is the regular ELBO while for the latent variables $\zb_{\leq k}$, the bound is evaluated using the (conditional) prior rather than the approximate posterior as the proposal distribution.
+The notation $\mathcal{L}^{>k}$ highlights that for latent variables $\zb^{(>k)}$, the bound is the regular ELBO while for the latent variables $\zb^{(\leq k)}$, the bound is evaluated using the (conditional) prior rather than the approximate posterior as the proposal distribution.
 
@@
     \mathcal{L}      &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) \parallel p_\thetab(\zb|\xb)\right), \label{eq_hierarchical:likelihoods-as-exact} \\ 
-    \mathcal{L}^{>k} &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k }|\zb_{>k}) q_\phib(\zb_{>k}|\xb) \parallel p_\thetab(\zb|\xb)\right) \nonumber \ .
+    \mathcal{L}^{>k} &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( p_\thetab(\zb^{(\leq k)}|\zb^{(>k)}) q_\phib(\zb^{(>k)}|\xb) \parallel p_\thetab(\zb|\xb)\right) \nonumber \ .
 \end{align}
 Subtracting these cancel out the two data likelihood terms $\log p_\thetab(\xb)$ and only the KL-divergences from the approximate to the true posterior remain,
-\begin{align}
+\begin{align}\label{eq_hierarchical:llr-as-kls}
     LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) \parallel p_\thetab(\zb|\xb)\right) \\
-                 &\quad + D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb) \parallel p_\thetab(\zb|\xb)\right) \ . \notag
-\end{align}\label{eq_hierarchical:llr-as-kls}
+                 &\quad + D_{\mathrm{KL}}\left( p_\thetab(\zb^{(\leq k)}|\zb^{(>k)}) q_\phib(\zb^{(>k)}|\xb) \parallel p_\thetab(\zb|\xb)\right) \ . \notag
+\end{align}
 
@@
 \begin{equation}
-    \mathcal{L}_{S} = \mathbb{E}_{q(\zb|\xb)}\left[ \log \frac{1}{N} \sum_{s=1}^{S} \frac{p(\xb, \zb^{(s)})}{q(\zb^{(s)}|\xb)} \right] \ , \label{eq_hierarchical:iw-bound}
+    \mathcal{L}^{\text{IWAE}}_{S} = \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \log \frac{1}{N} \sum_{s=1}^{S} \frac{p_\thetab(\xb, \zb_{s})}{q_\phib(\zb_{s}|\xb)} \right] \ , \label{eq_hierarchical:iw-bound}
 \end{equation}
-then, in the limit $S\rightarrow\infty$, we have $\mathcal{L}_{S} \rightarrow \log p_\thetab(\xb)$ and the likelihood ratio reduces to
-\begin{equation}
-    LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq k}|\zb_{>k}) q(\zb_{>k}|\xb) \parallel p(\zb|\xb))
-\end{equation}\label{eq_hierarchical:llr-as-kls-iwae-reduced}
+then, in the limit $S\rightarrow\infty$, we have $\mathcal{L}^{\text{IWAE}}_{S} \rightarrow \log p_\thetab(\xb)$ and the likelihood ratio reduces to
+\begin{equation}\label{eq_hierarchical:llr-as-kls-iwae-reduced}
+    LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p_\thetab(\zb^{(\leq k)}|\zb^{(>k)}) q_\phib(\zb^{(>k)}|\xb) \parallel p_\thetab(\zb|\xb))
+\end{equation}
 which, in practice, is well-approximated for a finite $S$.
-We expect this importance weighted likelihood ratio to monotonically improve upon the one in \cref{eq_hierarchical:llr-as-kls} as $S$ increases and the KL-divergence in the regular ELBO that contains terms for which $\zb_i$ is high-dimensional goes to zero.
+We expect this importance weighted likelihood ratio to monotonically improve upon the one in \cref{eq_hierarchical:llr-as-kls} as $S$ increases and the KL-divergence in the regular ELBO that contains terms for which $\zb^{(i)}$ is high-dimensional goes to zero.
 
@@
         \node[obs] (x_inf) {$\xb$};
-        \node[latent,above=.75cm of x_inf](z1_inf){$\zb_1$};
-        \node[latent,above=.75cm of z1_inf](z2_inf){$\zb_2$};
-        % \node[latent,above=.75cm of z2_inf](z3_inf){$\zb_3$};
+        \node[latent,above=.75cm of x_inf](z1_inf){$\zb^{(1)}$};
+        \node[latent,above=.75cm of z1_inf](z2_inf){$\zb^{(2)}$};
+        % \node[latent,above=.75cm of z2_inf](z3_inf){$\zb^{(3)}$};
         \node[above=of z2_inf, yshift=-1.cm] (phi) {$q_\phib(\zb|\xb)$};
@@
         \node[obs,right=0.75cm of x_inf] (x_gen) {$\xb$};
-        \node[latent,above=.75cm of x_gen](z1_gen){$\zb_1$};
-        \node[latent,above=.75cm of z1_gen](z2_gen){$\zb_2$};
-        % \node[latent,above=.75cm of z2_gen](z3_gen){$\zb_3$};
-        \node[above=of z2_gen, yshift=-1.cm] (theta) {$p_\thetab(\xb,\zb)$};
+        \node[latent,above=.75cm of x_gen](z1_gen){$\zb^{(1)}$};
+        \node[latent,above=.75cm of z1_gen](z2_gen){$\zb^{(2)}$};
+        % \node[latent,above=.75cm of z2_gen](z3_gen){$\zb^{(3)}$};
+        \node[right=0.75cm of phi] (theta) {$p_\thetab(\xb,\zb)$};
 
@@
 \paragraph{Models} For each OOD task, we train a simple bottom-up hierarchical VAE with $L$ stochastic layers which we will refer to as ``HVAE''.
-To alleviate posterior collapse we include skip-connections that connect $\zb_i$ to $\zb_{i+2}$ for $i\in\{0, L-2\}$ and $\zb_0\equiv\xb$ in both the inference and generative models \parencite{dieng_avoiding_2019} and employ the \textit{free bits} scheme with $\lambda=2$ \parencite{kingma_improved_2016}.
+To alleviate posterior collapse we include skip-connections that connect $\zb^{(i)}$ to $\zb^{(i+2)}$ for $i\in\{0, L-2\}$ and $\zb^{(0)}\equiv\xb$ in both the inference and generative models \parencite{dieng_avoiding_2019} and employ the \textit{free bits} scheme with $\lambda=2$ \parencite{kingma_improved_2016}.
 We use weight-normalization \parencite{salimans_weight_2016} on all weights and residual networks in the deterministic paths. 
@@
 Switching the in-distribution data for the OOD data in both cases result in correctly detecting the OOD data; an asymmetry also reported by \textcite{nalisnick_deep_2019}.
-\cref{fig_hierarchical:FMNIST-elbo-k0} shows the density of $\mathcal{L}^{>0}$ in bits per dimension \parencite{theis_note_2016} by the model trained on FashionMNIST when evaluated on the FashionMNIST and MNIST test sets.
+\Cref{fig_hierarchical:FMNIST-elbo-k0} shows the density of $\mathcal{L}^{>0}$ in bits per dimension \parencite{theis_note_2016} by the model trained on FashionMNIST when evaluated on the FashionMNIST and MNIST test sets.
 We observe a high degree of overlap, with less separation of the OOD data compared to similar results of autoregressive and flow-based models, like \textcite{xiao_likelihood_2020}.
@@
 We then evaluate the looser $\mathcal{L}^{>k}$ \cref{eq_hierarchical:biva->k} for $k\in\{1,L\}$.
-\cref{fig_hierarchical:FMNIST-elbo-k2} shows the result for $\mathcal{L}^{>2}$, which yielded the highest AUCROC$\uparrow$, only slightly better than random.
+\Cref{fig_hierarchical:FMNIST-elbo-k2} shows the result for $\mathcal{L}^{>2}$, which yielded the highest AUCROC$\uparrow$, only slightly better than random.
 Like \textcite{maaloe_biva_2019}, we see that increasing the value of $k$ generally leads to improved OOD detection.
@@
 
-\cref{tab_hierarchical:rocauc-ood} shows that BIVA improves upon the HVAE model for OOD detection on CIFAR while \cref{tab_hierarchical:bits-per-dim-ood} shows that the BIVA model also improves upon the HVAE in terms of likelihood.
+\Cref{tab_hierarchical:rocauc-ood} shows that BIVA improves upon the HVAE model for OOD detection on CIFAR while \cref{tab_hierarchical:bits-per-dim-ood} shows that the BIVA model also improves upon the HVAE in terms of likelihood.
 We hypothesize that models larger than our implementation of BIVA, with better likelihood scores may perform even better \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
@@
 \subsection{Comparison to baselines}
-\paragraph{Performance} \cref{tab_hierarchical:rocauc-ood} summarize our results compared to baselines based on the commonly used AUROC$\uparrow$, AUPRC$\uparrow$ and FPR80$\downarrow$ metrics.
+\paragraph{Performance} \Cref{tab_hierarchical:rocauc-ood} summarize our results compared to baselines based on the commonly used AUROC$\uparrow$, AUPRC$\uparrow$ and FPR80$\downarrow$ metrics.
 Our method outperforms other generative model-based methods such as WAIC \parencite{choi_waic_2019} with Glow model and performs similarly to the likelihood regret method of \parencite{xiao_likelihood_2020}.

### 2023-11-17 chapters/paper_review/04_benchmarking.tex
Fix notation for layer-number and timestep

@@
 baseline (i.e., using the entire training split of downstream datasets for
-fine"=tuning). \cref{table:method_data_setting,table:method_data_setting_2} summarize experiment settings used in the
+fine"=tuning). \Cref{table:method_data_setting,table:method_data_setting_2} summarize experiment settings used in the
 SSL literature, including the pre"=training corpora, downstream tasks and datasets,

### 2023-11-17 chapters/main-contributions.tex
Update contributions and research questions

@@
 \contribitem{part:unsupervised-uncertainty-estimation}
-This part of the thesis is concerned with unsupervised uncertainty estimation and consists of three papers. 
-The first two papers approach unsupervised uncertainty estimation through the task of out-of-distribution detection using generative models. 
-The first paper proposes a new method for out-of-distribution detection using hierarchical variational autoencoders and the second proposes a model-agnostic method for that can be used with any differentiable generative model with explicit likelihood.
-The third paper proposes a new hierarchical variational autoencoder for speech and benchmarks it against other variational and autoregressive models.
+This part of the thesis is concerned with unsupervised uncertainty estimation and consists of two papers. 
+% The first two papers approach unsupervised uncertainty estimation through the task of out-of-distribution detection using generative models. 
+Both papers focus on using generative models for out-of-distribution detection which is the task of detecting data that are likely to be sampled from a different data generating distribution than the training data.
+In both cases, the contributions are methodological focusing on developing improved methods for 
+% The first paper proposes a new method using hierarchical variational autoencoders while the second develops a model-agnostic, combined statistical test.
 
@@
 
-\moretodo[inline]{Revise the location of the benchmarking paper and change the SSL review to the SSL+LVM brief review. Revise the text here accordingly. $\downarrow$}
+% \moretodo[inline]{Revise the location of the benchmarking paper and change the SSL review to the SSL+LVM brief review. Revise the text here accordingly. $\downarrow$}
+
+\contribitem{part:unsupervised-speech-representation-learning}
+This part deals with unsupervised learning of speech representations and consists of two papers. 
+The contributions of this part lie in analyzing and comparing different approaches to speech representation along with a comprehensive overview of the field.
 
 \contribitem{chp:paper-benchmarking}
-In this paper, we focus on benchmarking hierarchical variational autoencoders for speech modelling. 
-We hypothesize that the improvements in likelihood scores proven by hierarchical models in other domains, most prominently images, can also carry over to speech modelling. 
-To this end, we benchmark a number of existing deterministic and stochastic models and adapt the hierarchical Clockwork VAE developed for video \parencite{saxena_clockwork_2021} to speech data. 
-We show that using a hierarchy of latent variables can improve the likelihood of the model and that the learned latent space captures information relevant for speech recognition, such as phonetic content.
+This chapter develops a novel hierarchical latent variable model for speech, drawing inspiration from the Clockwork VAE \parencite{saxena_clockwork_2021}.
+Comparative benchmarks against alternative latent variable models and autoregressive models for speech highlight improvements to likelihood brought by using hierarchical latent variables.
+Finally, the latent space of the models are analyzed in terms of phonetic content.
+
+\contribitem{chp:paper-brief}
+In this chapter, presents a comprehensive overview of unsupervised neural representation learning for speech.
+Previous research is categorized into self"-supervised methods and probabilistic latent variable models and described in a common notation.
+This description assists in developing a model taxonomy that shapes a discussion of the models' representational power, the associated learning strategies, and the methods used to evaluate them.
+An extended version of this overview paper that focuses exclusively on self"=supervised methods was also published as part of the project \ref{enumerate: paper-review} \parencite{mohamed_selfsupervised_2022}.
+This paper is included in \cref{app:paper-review} for reference.
+
+% \contribitem{chp:paper-benchmarking}
+% In this paper, we focus on benchmarking hierarchical variational autoencoders for speech modelling. 
+% We hypothesize that the improvements in likelihood scores proven by hierarchical models in other domains, most prominently images, can also carry over to speech modelling. 
+% To this end, we benchmark a number of existing deterministic and stochastic models and adapt the hierarchical Clockwork VAE developed for video \parencite{saxena_clockwork_2021} to speech data. 
+% We show that using a hierarchy of latent variables can improve the likelihood of the model and that the learned latent space captures information relevant for speech recognition, such as phonetic content.
 % This paper proposes a new hierarchical variational autoencoder for speech inspired by the Clockwork VAE \parencite{saxena_clockwork_2021}. 
@@
 
-\contribitem{part:unsupervised-speech-representation-learning}
-This part of the thesis is concerned with self"=supervised speech representation learning and consists of a single paper that provides a substantial review of the field \parencite{mohamed_selfsupervised_2022}.
-Self"=supervised learning is a promising approach to speech representation learning that has shown a fast, wide and successful adoption across speech modelling over the course of the last few years. 
-As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self"=supervised representations generally outperform representations learned via probabilistic generative models such as those investigated in \cref{part:unsupervised-uncertainty-estimation}.
-In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self"=supervised learning for learning representations for unsupervised uncertainty estimation.
-We shall return to this in the discussion of \cref{app:paper-review} in \cref{sec:discussion-paper-review}.
-\moretodo[inline]{Edit this if we did not discuss this.}
-An early version of this review paper included a review of variational autoencoders for speech representation learning and a comparison between the two modelling paradigms \parencite{borgholt_brief_2022}.
-Due to a desire to focus on self"=supervised methods, the paper in \cref{app:paper-review} does not put particular focus on variational methods. 
-The early version is included in \cref{chp:paper-brief} for reference.
+% \contribitem{part:unsupervised-speech-representation-learning}
+% This part of the thesis is concerned with self"=supervised speech representation learning and consists of a single paper that provides a substantial review of the field \parencite{mohamed_selfsupervised_2022}.
+% Self"=supervised learning is a promising approach to speech representation learning that has shown a fast, wide and successful adoption across speech modelling over the course of the last few years.
+% As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self"=supervised representations generally outperform representations learned via probabilistic generative models such as those investigated in \cref{part:unsupervised-uncertainty-estimation}.
+% In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self"=supervised learning for learning representations for unsupervised uncertainty estimation.
+% We shall return to this in the discussion of \cref{app:paper-review} in \cref{sec:discussion-paper-review}.
+% \moretodo[inline]{Edit this if we did not discuss this.}
+% An early version of this review paper included a review of variational autoencoders for speech representation learning and a comparison between the two modelling paradigms \parencite{borgholt_brief_2022}.
+% Due to a desire to focus on self"=supervised methods, the paper in \cref{app:paper-review} does not put particular focus on variational methods. 
+% The early version is included in \cref{chp:paper-brief} for reference.
 
-\contribitem{app:paper-review}
-This paper provides a comprehensive review of self"=supervised speech representation learning.
+% \contribitem{app:paper-review}
+% This paper provides a comprehensive review of self"=supervised speech representation learning.
 Previous work is grouped into three main categories of methods: contrastive, autoregressive, and generative, and the methods are compared in terms of their training objectives, model architectures, and performance on downstream tasks.
@@
 
-\moretodo[inline]{Revise the location of the benchmarking paper and change the SSL review to the SSL+LVM brief review. Revise the text here accordingly. $\uparrow$}
+% \moretodo[inline]{Revise the location of the benchmarking paper and change the SSL review to the SSL+LVM brief review. Revise the text here accordingly. $\uparrow$}
 

### 2023-11-17 chapters/discussion.tex
Add to "How did this ever work?"

@@
 Medical coding state-of-the-art uses pretrained BERT, but only marginally improves upon simpler baselines.}
+
+\moretodo[inline]{Add to the discussion somewhere:
+I would try to embed a perspective for contemplation.
+Instead of us always talking about what the technology needs to improve on.
+Perhaps, it would also be prudent to ask an open-ended philosophical question: “While our requirements to the technology are in its place, one could ask, whether human beings have the ability to quantify uncertainty.
+In 2023, the prevalence of human errors in driving are much higher than self-driving vehicles.” <- Quick example.
+I'm sure you can find something better.}
+%
 
@@
 \begin{equation} %\label{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)}
-    \mathbb{E}_{q_{\phib}(\xb,\zb)} \left[ \mathcal{L}_{\thetab,\phib}(\xb) \right] = 
+    \mathbb{E}_{q_{\phib}(\xb,\zb)} \left[ \mathcal{L}_{}(\xb;\thetab,\phib) \right] = 
     \underset{\text{average reconstruction}}{\underbrace{
@@
     I_{q_{\phib}(\xb,\zb)}\left[\xb ; \zb \right]
-    }} \enspace . \nonumber
+    }} \. . \nonumber
 \end{equation}
 %
+Note that we let $q_{\phib}(\xb,\zb) = q_{\phib}(\zb|\xb)\hat{p}(\xb)$ for ease of notation. 
 Optimizing the ELBO can be understood as simultaneously maximizing the average reconstruction likelihood, minimizing the marginal KL-divergence to the prior, and minimizing the mutual information between the observed and latent variables.
@@
 
+\moretodo[inline]{Finish the discussion of the MI interpretation}
+
 \paragraph{How did this ever work?}
@@
 Even though these properties indicate that VAEs might not work at all for representation learning, VAEs have consistently been shown to be able to learn meaningful representations \parencite{sonderby_ladder_2016,hsu_unsupervised_2017,maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
-We can understand the reason for this by noting that both of the above scenarios are only achievable if the optimization too unconstrained. 
-Specifically, the maximum likelihood estimation takes place not over all possible models, but a constrained parametric class defined by the inference network $q_\thetab(\zb|\xb) \approx p(\zb|\xb)$ and the generative model $p_\thetab(\xb|\zb)p(\zb) = p_\theta(\xb,\zb)$.
+We can understand the reason for this by noting that both of the above scenarios can only occur if the optimization is too unconstrained.
+
+In practice, the maximum marginal likelihood estimation takes place not over all possible models, but over a constrained parametric class defined by the inference network $q_\thetab(\zb|\xb) \approx p(\zb|\xb)$ and the generative model $p_\thetab(\xb|\zb)p(\zb) = p_\theta(\xb,\zb)$.
 This introduces a coupling between the marginal $p_{\thetab}(\xb)$ and the posterior $p_{\thetab}(\zb|\xb)$ which makes the maximum likelihood solution unlikely to correspond to completely useless representations \parencite{huszar_is_2017}.
-
-Similarly, maximizing the ELBO in \cref{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)} by exactly minimizing the mutual information term ...
-
+Similarly, maximizing the ELBO in \cref{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)} by exactly minimizing the mutual information term corresponds to a complete posterior collapse that requires a trade-off with the other terms.
+This solution is almost surely a local minimum and is often avoided in practice.
 The challenge is then how to impose appropriate constraints to make VAEs learn representations that are competitive with self"=supervised methods.

### 2023-11-20 chapters/paper_retrospective.tex
Update retrospective paper according to final submission version

@@
 \section*{Abstract}
-Effective stroke treatment depends on recognition by call-takers at prehospital telehealth services.
-This study aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.
-We used calls from 1 January 2015 to 31 December 2020 in Copenhagen to develop a machine learning-based classification pipeline.
-Calls from 2021 were used for testing.
-Calls were transcribed using an automatic speech recognition model and then categorized as stroke or non-stroke using a text classification model.
-Call-takers achieved a sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%).
-The machine learning framework performed significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
+Advanced stroke treatment is time dependent and, therefore, relies on recognition by call-takers at prehospital telehealth services to ensure fast hospitalisation.
+This study aims to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.
+We use calls from 1 January 2015 to 31 December 2020 in Copenhagen to develop a machine learning-based classification pipeline.
+Calls from 2021 are used for testing.
+Calls are first transcribed using an automatic speech recognition model and then categorized as stroke or non-stroke using a text classification model.
+Call-takers achieve a sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%).
+The machine learning framework performs significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
 Thus, a machine learning framework for recognising stroke in prehospital medical helpline calls may become a supportive tool for call-takers, aiding in early and accurate stroke recognition.
@@
 The gateway to ambulance transport and hospital admittance is through prehospital telehealth services, including emergency medical call centres, nurse advice call lines, and out-of-hours health services.
-In the pre-hospital setting, the use of mobile stroke units has made it possible to deliver advanced treatment faster.
-As the mobile stroke unit is only dispatched to patients with a suspected stroke, the impact of mobile stroke units is directly influenced by accurate call-taker recognition of stroke \parencite{cite6,cite7}.
-Call-takers who can rapidly and accurately recognize stroke are therefore crucial in facilitating prompt care in both pre-hospital and in-hospital settings.
+In the pre-hospital setting, the use of mobile stroke units has made it possible to deliver advanced treatment faster \parencite{cite6,cite7}.
+As the mobile stroke unit is only dispatched to patients with a suspected stroke, the impact of mobile stroke unit is directly influenced by accurate call-taker recognition of stroke \parencite{cite6,cite7}.
+Call-takers who can rapidly and accurately recognise stroke are therefore crucial in facilitating prompt care in both pre-hospital and in-hospital settings.
 
@@
 
-We used call recordings and registry data from the Copenhagen Emergency Medical Services (CEMS) and the Danish Stroke Registry (DanStroke)\parencite{cite16} from 2015 to 2020.
-The call recordings were obtained from two call lines: the 1-1-2 emergency line and the medical helpline 1813 (MH-1813).
-These data were used to build a machine learning framework to classify medical helpline calls as stroke or non-stroke.
-Calls were first transcribed using an automatic speech recognition model and then categorised by a text classification model trained as an ensemble of five individual models.
-We compared the performance of the model with that of call-takers using MH-1813 data from 2021.
+In this study, we use call recordings and registry data from the Copenhagen Emergency Medical Services (CEMS) and the Danish Stroke Registry (DanStroke) \parencite{cite16} from 2015 to 2020.
+We obtain call recordings from two call lines: the 1-1-2 emergency line and the medical helpline 1813 (MH-1813).
+We then fit a machine learning framework to classify medical helpline calls as stroke or non-stroke.
+Calls are first transcribed using an automatic speech recognition model and then categorised by a text classification model trained as an ensemble of five individual models.
+We compare the performance of the model with that of call-takers using MH-1813 data from 2021.
 
@@
     \centering
-    \caption[Performance metrics for stroke recognition.]{Performance metrics.
-Performance metrics for the model and call-takers on the test set.
-The model was trained on the training set, which included calls from 2015 to 2020.
-The call-takers were trained on the same data.
-The 2021 data were used for testing.
-The 2021 data without diagnostic categories were excluded}
+    \caption[Population characteristics for each data subset.]{Population characteristics for each data subset.}
     \label{tab_retrospective:table1-population-characteristics}
@@
 
-The classification model outperformed the call-takers (\cref{tab_retrospective:table2-main-results}), with significant differences in all metrics (p < 0.0001).
-Excluding the 1-1-2 call line training data significantly degraded the model's performance (p < 0.0001), despite the domain mismatch with the MH-1813 call line test data.
-The performance on the 2021 calls without a diagnostic category was significantly worse than that of the test set regarding F1-score, sensitivity, false positive rate (FPR), and false omission rate (FOR) (p < 0.0001).
-The difference in positive predictive value (PPV) was not significant (p = 0.298).
+The classification model outperformed the call-takers (\cref{tab_retrospective:table2-main-results}), with significant differences in all metrics (p < 0.0001, paired approximate permutation test).
+Excluding the 1-1-2 call line training data significantly degraded the model's performance (p < 0.0001, paired approximate permutation test), despite the domain mismatch with the MH-1813 call line test data.
+The performance on the 2021 calls without a diagnostic category was significantly worse than that of the test set regarding F1-score, sensitivity, false positive rate (FPR), and false omission rate (FOR) (p < 0.0001, independent approximate permutation test).
+The difference in positive predictive value (PPV) was not significant (p = 0.298, independent approximate permutation test).
 
 The receiver operating characteristic (ROC) curve (\cref{fig_retrospective:figure1-roc-curve}, left) illustrates the potential to increase the sensitivity while maintaining a FPR lower than or equal to that of the call-takers.
-Similarly, the PPV sensitivity curve (\cref{fig_retrospective:figure1-roc-curve}, right) demonstrates that sensitivity can be improved while retaining a PPV higher than that of the call-takers.
-The framework can thus be tuned to a sensitivity of around 73\%, while still having a higher positive predictive value than the human call-taker (\cref{fig_retrospective:figure1-roc-curve}, right) The ensemble model outperformed the individual models regardless of the threshold, except for one that exhibited a slightly better sensitivity at a high FPR exceeding 1.5\%.
+Similarly, the PPV-sensitivity curve (\cref{fig_retrospective:figure1-roc-curve}, right) demonstrates that sensitivity can be improved while retaining a PPV higher than that of the call-takers.
+The framework can thus be tuned to a sensitivity of around 73\%, while still having a higher positive predictive value than the human call-taker (\cref{fig_retrospective:figure1-roc-curve}, right).
+The ensemble model outperformed the individual models regardless of the threshold, except for one that exhibited a slightly better sensitivity at a high FPR exceeding 1.5\%.
 The confusion matrices (\cref{fig_retrospective:figure2-prediction-confusion-matrices}) illustrate the performance differences in absolute numbers, with the model exhibiting more true positives and fewer false positives than the call-takers.
@@
     \centering
-    \caption[Overall stroke recognition performance of model compared to call-takers.]{Overall performance on MH-1813 test data, performance without 1-1-2 training data, and performance on data from 2021 without diagnostic categories as well as performance on MH-1813 test data based on demographic subgroups (age/sex) [mean (95\% CI)].}
+    \caption[Overall stroke recognition performance of model compared to call-takers.]{Overall performance on MH-1813 test data, performance without 1-1-2 training data, and performance on data from 2021 without diagnostic categories as well as performance on MH-1813 based on demographic subgroups (age/sex) [mean (95\% CI)].
+NPV: negative predictive value, PPV: positive predictive value, FOR: false omission rate, CI: confidence interval.}
     \label{tab_retrospective:table2-main-results}
@@
 
-The model and call-takers exhibited significantly higher PPV and F1-score in men than in women (p < 0.0001) (\cref{tab_retrospective:table2-main-results}).
-The model significantly outperformed the call-takers on all metrics for each sex (p < 0.0001).
-The model performed significantly better in the 65+ group than in the 18-64 year group regarding sensitivity, PPV, and F1-score (p < 0.0001).
-Similarly, the call-takers performed significantly better in the 65+ group than in the 18-64 group regarding PPV and F1-score (p < 0.0001).
-Finally, the model significantly outperformed the call-takers on all metrics in both age groups (p < 0.0001).
+The model and call-takers exhibited significantly higher PPV and F1-score in men than in women (p < 0.0001, independent approximate permutation test) (\cref{tab_retrospective:table2-main-results}).
+The model significantly outperformed the call-takers on all metrics for each sex (p < 0.0001, paired approximate permutation test).
+The model performed significantly better in the 65+ group than in the 18-64 year group regarding sensitivity, PPV, and F1-score (p < 0.0001, independent approximate permutation test).
+Similarly, the call-takers performed significantly better in the 65+ group than in the 18-64 group regarding PPV and F1-score (p < 0.0001, independent approximate permutation test).
+Finally, the model significantly outperformed the call-takers on all metrics in both age groups (p < 0.0001, paired approximate permutation test).
 
@@
     \caption[Prediction confusion matrices for stroke recognition.]{Prediction confusion matrices.
-Confusion matrices of predictions for call-takers and the model on the test set.
+Confusion matrices of predictions for call takers and the model on the test set.
 Numbers for the model are given as the rounded mean over eleven runs.}
@@
 
-We performed an occlusion analysis to evaluate the importance of individual words for both positive and negative classifier predictions \cref{tab_retrospective:table3-occlusion-analysis}
+We performed an occlusion analysis to evaluate the importance of individual words for both positive and negative classifier predictions \cref{tab_retrospective:table3-occlusion-analysis}. 
 Among the words with a positive rank score, several words are synonymous with stroke, such as ``blood clot", ``haemorrhagic stroke", and ``stroke".
@@
     \centering
-    \caption[Words with the largest positive and negative ranking score in calls predicted as stroke and non-stroke, respectively.]{English translation of words with the largest positive and negative ranking score in calls predicted as stroke and non-stroke, respectively.}
+    \caption[Words with the largest positive and negative ranking score in calls predicted as stroke and non-stroke, respectively.]{English translation of words with the largest positive and negative ranking score in calls predicted as stroke and non-stroke, respectively.
+For this analysis, we used the model with the median F1-score out of 11 randomly seeded runs.}
     \label{tab_retrospective:table3-occlusion-analysis}
@@
 The effectiveness of the model is fully reliant on the conversation between the call-taker and caller and the call-taker's ability to skillfully triage the patient.
-Instead, the model should be used as a supportive tool for call-takers in the decision-making process, contributing to a higher recognition of patients with stroke and potentially boosting the confidence of call- takers in their decisions.
-A similar machine learning model designed to predict cardiac arrest was tested in a randomised controlled trial (RCT) at CEMS.15 The results highlighted the necessity of incorporating input from call-takers.
+Instead, the model should be used as a supportive tool for call-takers in the decision-making process, contributing to a higher recognition of patients with stroke and potentially boosting the confidence of call-takers in their decisions.
+A similar machine learning model designed to predict cardiac arrest was tested in a randomised controlled trial (RCT) at CEMS \parencite{cite15}.
+The results highlighted the necessity of incorporating input from call-takers.
 The machine learning model for cardiac arrest has subsequently been implemented in daily practice at CEMS, in a setup similar to the one presented in our study.
 However, implementation of our framework requires further investigation.
-The relative performance gap between call-takers and the model was larger in our study than in the cardiac arrest study,15 which may affect the results of a potential RCT.
+The relative performance gap between call-takers and the model was larger in our study than in the cardiac arrest study \parencite{cite15}, which may affect the results of a potential RCT.
 
-To support future work and discussions beyond the scope of this study, \cref{app:supplementary-retrospective} includes the results of a simulation of a live implementation where call-takers are assumed to follow a set of fixed rules based on the output of the machine learning framework.
+To support future work and discussions beyond the scope of this study, the supplementary material includes the results of a simulation of a live implementation where call-takers are assumed to follow a set of fixed rules based on the output of the machine learning framework \cref{app:supplementary-retrospective}.
 For instance, in one simulation call-takers are assumed to change any stroke negative to a positive, if the model predicts a positive.
@@
 First, the model may have learned to mimic call-takers with the same bias.
-Second, women may experience different and more challenging-to-identify symptoms than men \parencite{cite29, cite30}.
+Second, women may experience different and more challenging-to-identify symptoms than men \parencite{cite29,cite30}.
 Third, a higher prevalence of male patients with stroke was observed in the training data.
@@
 Since our model was trained on a representative - and therefore unbalanced - sample, we expect it to behave similarly.
-Future research should look to address these shortcomings, for example by utilising self"=supervised learning on massive amounts of diverse, unlabelled data covering multiple languages, accents, and dialects.
+Future research should look to address these shortcomings, for example by utilising self"-supervised learning on massive amounts of diverse, unlabelled data covering multiple languages, accents, and dialects.
 
@@
 CEMS operates two call lines: the 1-1-2 emergency line, similar to 9-1-1 in the United States, intended for acute conditions.
-The other is the medical helpline 1813 (MH-1813, pronounced ``'eighteen-thirteen") intended for non- life-threatening conditions that cannot wait until a general practitioner is available.18
+The other is the medical helpline 1813 (MH-1813, pronounced ``'eighteen-thirteen") intended for non- life-threatening conditions that cannot wait until a general practitioner is available \parencite{cite18}.
 
 Call-takers for both lines, who are nurses, paramedics, or physicians, can dispatch ambulances.
-The condition suspected by the call-taker is categorised based on a predefined diagnostic index and stored in an electronic record using a computer-aided dispatch (CAD) system.
+The condition suspected by the call-taker is categorised based on a predefined diagnostic index and stored in an electronic record using a CAD system.
 The CAD records are associated with the Danish civil registration number (CPR number) \parencite{cite19} of the patient.
 The CPR number is a unique identification assigned to all Danish residents.
-It is used for interactions with health services and registries, enabling cross- referencing of the data sources used in this study.
+It is used for interactions with health services and registries, enabling cross-referencing of the data sources used in this study.
 The call audio is recorded and stored separately from the CAD recordings using a telephone system.
@@
 Stroke prevalence in calls made to the MH-1813 is lower than that in calls made to 1-1-2.
-Patients with stroke may exhibit different symptoms and symptom severity because MH- 1813 is meant for low-acuity incidents, leading to reduced recognition.
+Patients with stroke may exhibit different symptoms and symptom severity because MH-1813 is meant for low-acuity incidents, leading to reduced recognition.
 In addition, MH-1813 call-takers dispatch high-priority transport less frequently, which may affect optimal treatment timing.
@@
 
-From the CAD medical records, we included all calls that could be matched to a corresponding audio file.
-for 1-1-2 and MH-1813 from 2015 to 2021 for patients older than 18.
+From the CAD medical records, we included all calls that could be matched to a corresponding audio file for 1-1-2 and MH-1813 from 2015 to 2021 for patients older than 18.
 The CAD records were matched with the telephone call recordings based on the call start, call duration, and call-taker identity.
@@
 
-As input for the classification model, each transcript was transformed into a fixed-size bag- of-words vector, which encoded the occurrence of word and character ($n$-grams) in a fixed vocabulary.
+As input for the classification model, each transcript was transformed into a fixed-size bag-of-words vector, which encoded the occurrence of word and character ($n$-grams) in a fixed vocabulary.
 The feature selection procedure is detailed in \cref{app:supplementary-retrospective}.
@@
 Subsequently, we defined the ensemble output score as the average of the centred output scores.
-The exact equations are provided in \cref{app:supplementary-retrospective}.
+The exact equations are provided in \cref{app:supplementary-retrospective} in \cref{eq_retrospective:output_score, eq_retrospective:ensemble_prediction}.
 
@@
 
-The speech recognition model was trained on 3,811 manually transcribed random calls (173 h) from the CEMS as part of a previous project.14 These calls exclusively originated from 1- 1-2 between 2015 and 2018, ensuring no overlap with the test data used for the text classification model.
+The speech recognition model was trained on 3,811 manually transcribed random calls (173 h) from the CEMS as part of a previous project \parencite{cite14}.
+These calls exclusively originated from 1- 1-2 between 2015 and 2018, ensuring no overlap with the test data used for the text classification model.
 The model was trained using a connectionist temporal classification objective \parencite{graves_connectionist_2006}.
@@
 
-We report the F1-score, sensitivity, PPV, false omission rate (FOR, equal to 1- negative predictive value), and false positive rate (FPR, equal to 1 - specificity).
+We report the F1-score, sensitivity, PPV, FOR (equal to 1- negative predictive value), and FPR (equal to 1 - specificity).
 Due to the imbalanced nature of the dataset, the negative predictive value and specificity were greater than 99\% for all cases.
 We reported FOR and FPR instead because such large numerical values exhibit low relative variance, thereby obfuscating comparisons.
-Finally, we report the prediction confusion matrices, receiver operator characteristic (ROC) curve, and PPV- sensitivity curve, commonly known as the precision-recall curve.
+Finally, we report the prediction confusion matrices, ROC curve, and PPV-sensitivity curve, commonly known as the precision-recall curve.
 All results are reported with up to three significant digits.
 
+We present the results with and without 1-1-2 training data, subgroup analyses based on age (18–64/65+) and sex (male/female), and call-takers performance.
+We also report the model performance on calls without a diagnostic category from the test year 2021 to assess potential data bias.
 We tested our results for statistical significance using approximate permutation tests.
@@
 female) we used one-sided independent approximate permutation tests.
-We computed 95\% confidence intervals (CIs) using bootstrapping.
-\parencite{dwass_modified_1957,eden_validity_1933}.
+We computed 95\% confidence intervals (CIs) using bootstrapping \parencite{dwass_modified_1957,eden_validity_1933}.
 In our assessment, we accounted for random variation associated with model training by basing the means, tests, and CIs on the predictions of 11 randomly initialised training runs.
-Statistical significance was defined as a p-value of < 0.05.
+Statistical significance was defined as a p-value of less than 0.05.
 
@@
 
-\subsection{Role of the funding source}
+\section*{Data availability}
 
+The datasets used to evaluate call-taker performance and to train and evaluate the machine learning framework are legally restricted by Danish patient privacy and secrecy laws and are therefore, not publicly available.
+The data can be made available from the publication date but requires a Data Access Agreement, which is examined and approved by the ethics committees that approved this research \footnote{Danish Data Protection Agency, \url{https://www.datatilsynet.dk/english}}.
+For the same reason, the machine learning framework trained in this study is not publicly available; however, instructions on how to train it are included in the main manuscript and the supplementary material.
+
+\section*{Code availability}
+
+The source code can be shared using a Creative Commons NC-ND 4.0, an international licence, upon reasonable written request to the corresponding author, and requires a data use agreement.
+
+\section*{Acknowledgements}
+
+We thank the staff of the CEMS for their role in generating the data used in this study.
+We thank Emilie Grunddal Pedersen, Mette Bjerg Lindhøj, and Jens Morten Haugård for their help and cooperation in accessing the data sources.
+We also thank the Centre for IT and Medical Technology (CIMT) and Corti employees Akihiro Inui and Nathaniel Joselson for their assistance in setting up and using the cloud-computing environment for training and evaluating the machine learning framework of this study.
+Funding for the work was received from Innovation Fund Denmark, Trygfonden, Copenhagen University Hospital - Herlev, Gentofte and University of Copenhagen.
 The grant providers had no role in the study design, data collection, analysis, interpretation, manuscript writing, or publication decision.
@@
 Corti owns the rights to the models and source code.
-
-\section*{Data availability}
-
-The datasets used to evaluate call-taker performance and to train and evaluate the machine learning framework are legally restricted by Danish patient privacy and secrecy laws and are therefore, not publicly available.
-The data can be made available from the publication date but requires a Data Access Agreement, which is examined and approved by the ethics committees that approved this research.
-For the same reason, the machine learning framework trained in this study is not publicly available; however, instructions on how to train it are included in the main manuscript and supplementary material.
-The source code can be shared using a Creative Commons NC-ND 4.0, an international licence, upon reasonable written request to the corresponding author, and requires a data use agreement.
-
-\section*{Code availability}
-
-The source code can be shared using a Creative Commons NC-ND 4.0, an international license, upon reasonable written request to the corresponding author, and requires a data use agreement.
 
@@
 Both positions are unpaid.
-
-\section*{Acknowledgements}
-
-We thank the staff of the CEMS for their role in generating the data used in this study.
-We thank Emilie Grunddal Pedersen, Mette Bjerg Lindhøj, and Jens Morten Haugård for their help and cooperation in accessing the data sources.
-We also thank the Centre for IT and Medical Technology (CIMT) and Corti employees Akihiro Inui and Nathaniel Joselson for their assistance in setting up and using the cloud-computing environment for training and evaluating the machine learning framework of this study.
 
@@
 
+\section*{Competing interests}
+JDH and LB received funding from the Innovation Fund Denmark.
+JDH and LB used Corti and hold stock warrants.
+LM is a co-founder, stockholder, and the Chief Technology Officer of Corti.
+JW received funding from Trygfonden.
+SNFB has no conflicts of interest to declare.
+HC has received funding from the Velux Foundation, Tværsfonden, Helsefonden, Hartmann Fonden, Lundbeck Foundation, and Novo Nordisk Foundation; royalties from Gyldendal; honoraria from Bayer and Bristol Meyers Squibb, and is chair of Action Plan for stroke in Europe Implementation, Co-chair of the Scientific Stroke Panel EAN and Senior Guest Editor of AHA Stroke.
+MS has no conflicts of interest to declare.
+HCC has no conflicts of interest to declare.
+CK received funding from the Novo Nordisk Foundation and is the chair of the Danish Resuscitation Council and vice chair of the Danish Stroke Society.
+Both positions are unpaid.
+
 }

### 2023-11-20 chapters/paper_modelagnostic.tex
Fix notation for layer-number and timestep
Add source reference at the top of every paper chapter

@@
 \label{chp:paper-modelagnostic}
+
+\textit{This chapter is a piece of original research published as part of the project:} \newline
+\begin{center}
+    \begin{enumerate}[leftmargin=8mm,rightmargin=8mm,topsep=0mm,label={[\Alph*]}]
+        \setcounter{enumi}{1}
+        \item \fullcite{bergamin_modelagnostic_2022} \co \label{enumerate: paper-modelagnostic}
+        \end{enumerate}
+\end{center}
+
 \ifthenelse{\equal{\skippapers}{true}}{}{
@@
 Ideally, we would like to have a low Type I and a low Type II error rate, meaning that we are not considering a lot of in-distribution examples as OOD and at the same time considering a lot of outliers as in-distribution.
-\cref{fig_modelagnostic:type1} shows that we can achieve this for low values of $\alpha$.
+\Cref{fig_modelagnostic:type1} shows that we can achieve this for low values of $\alpha$.
 When training on CIFAR, instead, we are able to control the FDR only from a certain significance level (see \cref{appendix_modelagnostic:BH_cifar}).

### 2023-11-20 appendices/supplementary_retrospective.tex
Update retrospective paper supplementary to final submission version

@@
 
-The selection of hyperparameters followed a simple two-stage process using validation data (table A1).
+The selection of hyperparameters followed a simple two-stage process using validation data (table 1).
 First, a manual search was conducted by running different model configurations with varying numbers of epochs, updates per epoch, batch sizes, vectoriser types, and hyperparameters.
@@
 These vectors encode the occurrence of words and character n-grams within a fixed vocabulary.
-We selected vocabulary by first computing the χ2-statistics for all word uni-and bi-grams and character three-, four-, and fivegrams that occurred in more than ten training calls.
-We then retained the M highest-scoring word n-grams and $M$ highest-scoring character n-grams, yielding $2M$ input features, where $M$ represents a tuned hyperparameter.
+We selected vocabulary by first computing the $\chi^2$-statistics for all word uni- and bi-grams and character three-, four-, and five-grams that occurred in more than ten training calls.
+We then retained the $M$ highest-scoring word n-grams and M highest-scoring character n-grams, yielding $2M$ input features, where $M$ represents a tuned hyperparameter.
 By complementing word n-grams with character n-grams, the model can use out-of-vocabulary words not included in the word n-grams and robustly represent words misspelt by the speech recogniser.
@@
 The individual models are broken down in panels B and C.
-The 2D convolutional layers have parameters [input channels, output channels]-[kernel width, kernel height]-[stride width, stride height].
+The 2D convolutional layers have parameters [\textit{input channels, output channels}]-[\textit{kernel width, kernel height}]-[\textit{stride width, stride height}].
 The long short-term memory (LSTM) blocks have parameters hidden units-number of layers.
@@
 This is problematic for two reasons.
-%
-\begin{itemize}
-    \item The lack of a continuous output score prevents the evaluation of the model's performance across a continuous range of thresholds required for plotting the receiver operating characteristic and precisionrecall curves.
-    \item The lack of a continuous output score deteriorates the quality of assessing the effect of different words on ensemble performances (see \cref{sec_retrospective:model_explainability}).
-\end{itemize}
-%
+
+The lack of a continuous output score prevents the evaluation of the model's performance across a continuous range of thresholds required for plotting the receiver operating characteristic and precision-recall curves.
+
+The lack of a continuous output score deteriorates the quality of assessing the effect of different words on ensemble performances (see \cref{sec_retrospective:model_explainability}).
+
 Therefore, we used a different ensemble method, which is briefly described in the main text. Herein, we provide a mathematically rigorous definition of the proposed method.
@@
 %
-\begin{equation}
+\begin{equation} \label{eq_retrospective:output_score}
     p^{(d)} = \frac{1}{N} \sum_{n=1}^N \sigma\left(z^{(n,d)} - t^{(n)}\right) \enspace ,
@@
 %
-where $\sigma(\cdot)$ is the sigmoid function (or standard logistic function). The final ensembel predction $s^{(d)}$ is then simply 
+where $\sigma(\cdot)$ is the sigmoid function (or standard logistic function). The final ensemble prediction $s^{(d)}$ is then simply 
 %
-\begin{equation}
+\begin{equation} \label{eq_retrospective:ensemble_prediction}
     s^{(d)} = I_{p^{(d)} > 0.5}\left(p^{(d)}\right) \enspace ,
@@
 We used standard methods for significance testing and computing the confidence intervals.
-We used approximate methods owing to data size and to maintain computational feasibility \parencite{dwass_modified_1957,eden_validity_1933}
+We used approximate methods owing to data size and to maintain computational feasibility \parencite{dwass_modified_1957,eden_validity_1933}.
 
@@
 
-When 1-1-2 data were not used for training (\cref{tab_retrospective:tableA2}), the model performed significantly better in men than in women on the test set in terms of all metrics (p < 0.0001) and significantly better on the 65+ group than on the 1864 group in terms of all metrics (p < 0.0001).
-As noted in the main text, both these statements were also true when 1-1-2 data were used for the training (\cref{tab_retrospective:table2-main-results}) and that including 1-1-2 data significantly improved overall performance (p < 0.0001).
-
-Compared to the test set (\cref{tab_retrospective:table2-main-results}), model performance on the 2021 calls without diagnostic category was significantly worse in all demographic subgroups (\cref{tab_retrospective:tableA3}) in terms of all metrics (p < 0.0001), except for positive predictive value (PPV) for male (where p = 0.0056 [significant]) and for the false positive rate (FPR) on females and individuals aged 65+ years (where p = 0.213 and p = 0.362, respectively [insignificant]).
+When 1-1-2 data were not used for training (\cref{tab_retrospective:tableA2}), the model performed significantly better in men than in women on the test set in terms of all metrics (p < 0.0001, paired approximate permutation test) and significantly better on the 65+ group than on the 18-64 group in terms of all metrics (p < 0.0001, paired approximate permutation test).
+As noted in the main text, both these statements were also true when 1-1-2 data were used for the training (\cref{tab_retrospective:table2-main-results}) and that including 1-1-2 data significantly improved overall performance (p < 0.0001, paired approximate permutation test).
+Compared to the test set (\cref{tab_retrospective:table2-main-results}), model performance on the 2021 calls without diagnostic category was significantly worse in all demographic subgroups (\cref{tab_retrospective:tableA3}) in terms of all metrics (p < 0.0001, independent approximate permutation test), except for positive predictive value (PPV) for male (where p = 0.0056 [significant], independent approximate permutation test) and for the false positive rate (FPR) on females and individuals aged 65+ years (where p = 0.213 and p = 0.362, respectively [insignificant] , independent approximate permutation test).
 
@@
     \centering
-    \caption{Model performance for sex and age subgroups [mean (95\% CI)] on the MH-1813 test data when 1-1-2 training data is not used for training.}
+    \caption[Model performance grouped by sex and age when 1-1-2 training data is not used for training.]{Model performance grouped by sex and age [mean (95\% CI)] when 1-1-2 training data is not used for training.
+NPV: negative predictive value, PPV: positive predictive value, FOR: false omission rate, CI: confidence interval.}
     \label{tab_retrospective:tableA2}
@@
     \centering
-    \caption{Model performance for sex and age subgroups [mean (95\% CI)] on the 2021 data without diagnostic category.}
+    \caption[Model performance grouped by sex and age on the 2021 data without diagnostic category.]{Model performance grouped by sex and age [mean (95\% CI)] on the 2021 data without diagnostic category.
+NPV: negative predictive value, PPV: positive predictive value, FOR: false omission rate, CI: confidence interval.}
     \label{tab_retrospective:tableA3}
@@
     \caption[Overall performance on MH-1813 test data for the model that also takes patient age and sex as direct inputs.]{Overall performance on MH-1813 test data for the model that also takes patient age and sex as direct inputs.
-We also list the original performance of call-takers and the model (w/o sex and age) from the main manuscript for ease of comparison [mean (95\% CI)].}
+We also list the original performance of call-takers and the model (w/o sex and age) from the main manuscript for ease of comparison [mean (95\% CI)].
+NPV: negative predictive value, PPV: positive predictive value, FOR: false omission rate, CI: confidence interval.}
     \label{tab_retrospective:tableA4}
@@
 
-We note that the performance difference between the original model and the models with age and sex inputs is statistically insignificant (p > 0.05).
+We note that the performance difference between the original model and the models with age and sex inputs is statistically insignificant (p > 0.05, paired approximate permutation test).
 We hypothesize this might be because information regarding the age and sex of the patient is, in many cases, already present in the transcript.
@@
 We note that the model still performed better than the call-takers in terms of F1-score, PPV, and FPR (p < 0.0001) and was on-par in terms of sensitivity and FOR.
+All tests for this comparison were paired approximate permutation tests.
 
 In summary, training on only 1-1-2 data was only somewhat worse than training on only MH-1813 data, and still outperformed call-takers to some degree.
-This indicates that the domain shift between different call lines, even with different acuity levels, is small enough that naive domain transfer of models works fairly well.
+This indicates that the domain shift between different call lines, even with different acuity levels, is small enough that naive domain transfer of models works well.
 
@@
     \centering
-    \caption{Overall performance on MH-1813 test data for the original model, the model without 1-1-2 training data, and the model without 1813 training data along with call-taker performance [mean (95\% CI)].}
+    \caption[Overall performance on MH-1813 test data, performance without 1-1-2 training data, and performance without 1813 training data.]{Overall performance on MH-1813 test data, performance without 1-1-2 training data, and performance without 1813 training data [mean (95\% CI)].
+NPV: negative predictive value, PPV: positive predictive value, FOR: false omission rate, CI: confidence interval.}
     \label{tab_retrospective:tableA5}
@@
 
-        % & \multicolumn{5}{c}{\textit{Without 1-1-2 training data (only MH-1813 data)}} \\
-        % \midrule
-        % Model        & \makecell[c]{32.4 \\ (31.8-33.1)} & \makecell[c]{60.4 \\ (59.3-61.4)} & \makecell[c]{22.2 \\ (21.6-22.7)} & \makecell[c]{0.088 \\ (0.085-0.091)} & \makecell[c]{0.467 \\ (0.460-0.474)} \\
-
-        % & \multicolumn{5}{c}{\textit{Without MH-1813 training data (only 1-1-2 data)}} \\
-        % \midrule
-        % Model        & \makecell[c]{31.4 \\ (30.7-32.1)} & \makecell[c]{50.4 \\ (49.3-51.4)} & \makecell[c]{22.8 \\ (22.2-23.4)} & \makecell[c]{0.110 \\ (0.106-0.113)} & \makecell[c]{0.375 \\ (0.369-0.381)} \\
-
         \bottomrule
@@
 
-% % \todo[inline]{Insert table A4 here}
 \begin{table}
@@
 
-We see that the fine"=tuned BERT model performs slightly worse across F1-score, sensitivity, PPV and FOR (p < 0.0001), but better in terms of FPR (p < 0.0001), compared the MLP model presented in the main manuscript.
+We see that the fine"=tuned BERT model performs slightly worse across F1-score, sensitivity, PPV and FOR (p < 0.0001, paired approximate permutation test), but better in terms of FPR (p < 0.0001, paired approximate permutation test), compared the MLP model presented in the main manuscript.
 As described in the discussion section, we hypothesize that the number of stroke positives was too small for these advanced models to learn more complex patterns than the MLP ensemble.
@@
     \centering
-    \caption[Overall performance on MH-1813 test data for a fine"=tuned BERT model.]{Overall performance on MH-1813 test data for a fine"=tuned BERT model.
-We also list the original performance of call-takers and the ensemble model from the main manuscript for ease of comparison [mean (95\% CI)].}
+    \caption[Overall performance on MH-1813 test data for a fine"=tuned BERT model.]{Overall performance on MH-1813 test data for the fine-tuned BERT model described in the revised discussion of the paper.
+Includes performance of the call-takers and the multi-layer perceptron (MLP) from the main manuscript for ease of comparison [mean (95\% CI)].
+NPV: negative predictive value, PPV: positive predictive value, FOR: false omission rate, CI: confidence interval.}
     \label{tab_retrospective:tableA8}
@@
 We implemented option 2 in such a manner that the model predicted every time 50 new words were transcribed and added to the transcript.
-A stroke positive was triggered only when three consecutive positive predictions were made (i.e.
-without intermediate negative stroke predictions).
+A stroke positive was triggered only when three consecutive positive predictions were made (i.e., without intermediate negative stroke predictions).
 In other words, the sigmoid activation of the model had to remain above 0.5 for three consecutive predictions, for example, after 150, 200, and 250 words were transcribed.
 
-As we can only assume how call-takers are influenced by model predictions (II), precisely evaluating the hypothetical performance of call-takers when supported by a machine learning framework is impossible.
+As we can only assume how call takers are influenced by model predictions (II), precisely evaluating the hypothetical performance of call takers when supported by a machine learning framework is impossible.
 Furthermore, option 2 may influence the conversation, further complicating matters.
-Therefore, we report the results combining the call-taker and the model under the following two assumptions:
+Therefore, we report the results combining the call taker and the model under the following two assumptions:
 %

### 2023-11-20 chapters/paper_review/03_approaches.tex
Fix references to paragraphs

@@
 In the standard BERT masking policy, each token is masked independently at random.
-However, for speech, masking a single \edit{sample or spectrogram frame results in a largely} trivial reconstruction task \edit{since, as discussed in \cref{par:apc}, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore it is common to mask chunks of consecutive frames~\parencite{liu_mockingjay_2020,jiang_further_2021}.
+However, for speech, masking a single \edit{sample or spectrogram frame results in a largely} trivial reconstruction task \edit{since, as discussed in %\cref{par:apc}
+the paragraph on autoregressive prediction, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore it is common to mask chunks of consecutive frames~\parencite{liu_mockingjay_2020,jiang_further_2021}.
 
@@
 \subsubsection{Approaches}
-\paragraph{CPC}
+\paragraph{Contrastive predictive coding}
 \label{par:cpc}
@@
 The Discrete BERT approach~\parencite{baevski_effectiveness_2020} uses a pre"=trained vq-wav2vec model to derive a discrete vocabulary \parencite{baevski_vqwav2vec_2020}.
-The vq-wav2vec model is similar to wav2vec mentioned in \cref{par:cpc} but uses quantization to learn discrete representations.
+The vq-wav2vec model is similar to wav2vec mentioned in %\cref{par:cpc}
+the paragraph on contrastive predictive coding but uses quantization to learn discrete representations. 
 \edit{Specifically, discrete units $c_t$ are first extracted with the vq-wav2vec model $f_1(\cdot)$ and then used as inputs \emph{and} targets in a standard BERT model $f_2(\cdot)$ with a softmax normalized output layer $g(\cdot)$,

### 2023-11-20 chapters/paper_retrospective.tex
Add source reference at the top of every paper chapter
Fix references

@@
 \label{chp:paper-retrospective}
+
+\textit{This chapter is a piece of original research published as part of the project:} \newline
+\begin{center}
+    \begin{enumerate}[leftmargin=8mm,rightmargin=8mm,topsep=0mm,label={[\Alph*]}]
+        \setcounter{enumi}{5}
+        \item \fullcite{wenstrup_retrospective_2023} \shared
+    \end{enumerate}
+\end{center}
+
 \ifthenelse{\equal{\skippapers}{true}}{}{
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{paper_retrospective/figure1.pdf}
+    \includegraphics[width=0.95\textwidth]{paper_retrospective/figure1.pdf}
     \caption[ROC- and PPV-sensitivity curves for stroke recognition model.]{Receiver operator characteristic (ROC) curve and PPV-sensitivity curve.
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{paper_retrospective/figure2.pdf}
+    \includegraphics[width=0.98\textwidth]{paper_retrospective/figure2.pdf}
     \caption[Prediction confusion matrices for stroke recognition.]{Prediction confusion matrices.
@@
 Subsequently, we defined the ensemble output score as the average of the centred output scores.
-The exact equations are provided in \cref{app:supplementary-retrospective} in \cref{eq_retrospective:output_score, eq_retrospective:ensemble_prediction}.
+The exact equations are provided in \cref{eq_retrospective:output_score,eq_retrospective:ensemble_prediction}.
 

### 2023-11-20 chapters/background.tex
Some updates to contributions and background

@@
 While the papers in this thesis were written to be self-contained, page limits and the need to maintain a specific focus forced us to limit the scope of the technical background that each paper provides.
-In this chapter, we present a more in-depth overview of the technical background relevant to the papers in this thesis.
-We start by quantifying the concept of uncertainty relating it to information and probability.
+In this chapter, we present a more in-depth introduction to the technical background relevant to the papers in this thesis.
+We start by quantifying the concept of uncertainty and relating it to information and probability.
 We then introduce the task of out-of-distribution detection and review the existing literature on the topic.
@@
 %
-Although such a purely lexical definition of uncertainty might prompt a philosophical inquiry, it highlights an important connection between uncertainty, information and probability.
+Although such a purely lexical definition of uncertainty might prompt philosophical inquiry, it highlights an important connection between uncertainty, information and probability.
 When information is limited, we can describe the state of the world only with uncertainty; we are simply not certain.
@@
 There exist at least six main classes of deep generative models: variational autoencoders \parencite{kingma_autoencoding_2014,rezende_stochastic_2014,ranganath_hierarchical_2016,vahdat_nvae_2020,child_very_2021}, normalizing flow models \parencite{dinh_nice_2015,rezende_variational_2015,dinh_density_2017,kingma_glow_2018,grathwohl_ffjord_2018}, diffusion models \parencite{sohl-dickstein_deep_2015,song_generative_2019, ho_denoising_2020, vahdat_scorebased_2021}, energy-based models \parencite{lecun_tutorial_2006,hinton_fast_2006,salakhutdinov_efficient_2010,du_implicit_2019}, autoregressive models \parencite{oord_conditional_2016,oord_wavenet_2016,radford_improving_2018}, and generative adversarial networks \parencite{goodfellow_generative_2014,arjovsky_wasserstein_2017,brock_large_2019,karras_stylebased_2019}.
+
+VAEs are interesting for representation learning as well as uncertainty quantification, due to the assumptions they impose on the data modelling task and their probabilistic formulation.
+To some degree, these properties set them apart from other deep generative models that focus more exclusively on data generation, and we chose to focus on them in this thesis.
+% \moretodo[inline]{Briefly hold the readers hand and tell them why VAEs are relevant to the overarching project.}
 % Besides data generation, common properties of deep generative models include the ability to compute an exact or approximate likelihood of input $p(\xb)$, reconstruct an input $\xb$ from its latent representation $\zb$, and represent an input $\xb$ in a latent space.
 These properties enable various methods for their use in OOD detection.
-
-\moretodo[inline]{Briefly hold the readers hand and tell them why VAEs are relevant to the overarching project.}
-
 % A variational autoencoder (VAE) \parencite{kingma_autoencoding_2014,rezende_stochastic_2014} is a type of deep generative model and an instance of a latent variable model. 
@@
     &= \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \frac{1}{S}\sum_{s=1}^{S} \log \left( w_\thetab(\xb,\zb_{s}) \right) \right] \nonumber \\
-    &\leq \log \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \frac{1}{S}\sum_{s=1}^{S} w_\thetab(\xb,\zb_{s}) \right] \enspace . \nonumber \\
+    &\leq \log \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \frac{1}{S}\sum_{s=1}^{S} w_\thetab(\xb,\zb_{s}) \right] \nonumber \\
     &= \log p(\xb) \enspace .
@@
 Hence, $\widehat{\mathcal{L}}^{\text{IWAE}}_{S}$ is a biased estimator of $\log p(\xb)$ but with a bias that goes to zero as the number of samples goes to infinity. 
-With $S=1$, the estimator is equivalent to the regular ELBO \cref{eq:variational-autoencoder-elbo} and so is always as tight or tighter than the ELBO \parencite{burda_importance_2016}.
+With $S=1$, the estimator is equivalent to the regular ELBO \cref{eq:variational-autoencoder-elbo} and so, it is always as tight, or tighter, than the ELBO \parencite{burda_importance_2016}.
 
@@
 % Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
-Computing the gradient of the ELBO with respect to $\thetab$ is straightforward,
+Computing the gradient of the ELBO \cref{eq:variational-autoencoder-elbo} with respect to $\thetab$ is straightforward,
 %
@@
     \nabla_\thetab \mathcal{L}_{}(\xb;\thetab,\phib)
-    = \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \nabla_\thetab \log p_\thetab(\xb|\zb) \right] \enspace .
+    = \nabla_\thetab \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] 
+    = \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \nabla_\thetab \log p_\thetab(\xb,\zb) \right] \enspace .
 \end{equation}
@@
 %
-\begin{equation}
-    % \nabla_\phib \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
+\begin{align}
     \nabla_\phib \mathcal{L}_{}(\xb;\thetab,\phib)
-    = \mathbb{E}_{p(\epsilonb)} \left[ \nabla_\phib \log \frac{p_\thetab\left(\xb|g_\phib(\xb,\epsilonb)\right)p\left(g_\phib(\xb,\epsilonb)\right)}{q_\phib\left(g_\phib(\xb,\epsilonb)|\xb\right)} \right] \enspace .
-\\
-\end{equation}
+    &= \nabla_\phib \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \nonumber \\
+    &= \mathbb{E}_{p(\epsilonb)} \left[ \nabla_\phib \log \frac{p_\thetab\left(\xb,g_\phib(\xb,\epsilonb)\right)}{q_\phib\left(g_\phib(\xb,\epsilonb)|\xb\right)} \right] \enspace .
+\end{align}
 %
@@
     % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb^{(L-1)}|\zb_L)\cdots q_\phib(\zb_2|\zb^{(1)}),
-    q_\phib(\zb|\xb) = q_\phib(\zb^{(L)}|\xb)\prod_{i=1}^{L-1} q_\phib(\zb^{(i)}|\zb^{(i+1)}) \ .
+    q_\phib(\zb|\xb) = q_\phib(\zb^{(L)}|\xb)\prod_{i=1}^{L-1} q_\phib(\zb^{(i)}|\zb^{(i+1)}) \enspace .
 \end{equation}

### 2023-11-20 chapters/main-contributions.tex
Some updates to contributions and background

@@
 \Cref{chp:technical-background} provides additional technical background not included in the individual studies.
-It introduces uncertainty as a concept in the context of information and probability theory, introduces the task of out-of-distribution detection and provides a review of existing work on the problem, and lays out the foundations for variational autoencoders.
+It introduces uncertainty as a concept in the context of information and probability theory and introduces the task of out-of-distribution detection and provides a review of existing work on the problem.
+Finally, it lays out the foundations for variational autoencoders.
 
@@
 % The first two papers approach unsupervised uncertainty estimation through the task of out-of-distribution detection using generative models. 
-Both papers focus on using generative models for out-of-distribution detection which is the task of detecting data that are likely to be sampled from a different data generating distribution than the training data.
-In both cases, the contributions are methodological focusing on developing improved methods for 
+Both papers focus on using generative models for out-of-distribution detection, which is the task of detecting data that are likely to be sampled from a different data generating distribution than the training data.
+In both cases, the contributions are methodological and relate to developing improved methods for out-of-distribution detection.
 % The first paper proposes a new method using hierarchical variational autoencoders while the second develops a model-agnostic, combined statistical test.
@@
 In this work we hypothesize that the likelihood estimate of variational autoencoders is a poor score for out-of-distribution due to an overemphasis on low-level features that generalize between distributions.
-We further hypothesize that a well-formed hierarchy of latent variables provides a tool that can be used to select which features to emphasize for out-of-distribution detection and hence a way to improve the performance of variational autoencoders on this task.
-We proceed to provide empirical and theoretical evidence that low-level features indeed do dominate the likelihood and propose a new method for out-of-distribution detection using hierarchical variational autoencoders.
-The proposed method computes a likelihood-ratio score for out-of-distribution detection that requires data to be in-distribution across all feature-levels. 
+We further hypothesize that a well-formed hierarchy of latent variables provides a tool that can be used to select which features to emphasize for out-of-distribution detection and, hence, a way to improve the performance of variational autoencoders on this task.
+We proceed to provide empirical and theoretical evidence that low-level features do indeed dominate the likelihood score and propose a new method for out-of-distribution detection using hierarchical variational autoencoders based on a likelihood-ratio score that requires data to be in-distribution across all feature-levels.
 The proposed method is computationally efficient, fully unsupervised, and performs well on several out-of-distribution detection benchmarks. 
@@
 \contribitem{chp:paper-modelagnostic}
-In this follow-up work, we note that many methods have been proposed for out-of-distribution detection using generative models, with some tailored for specific model types, and seek to develop a model-agnostic approach.
-We hypothesize that by phrasing the problem as a statistical testing problem and combining different tests, the method's efficacy can be improved and weaknesses inherent to any particular test alleviated.
-From this hypothesis, we develop a method that combines a classical parametric test with the recently introduced typicality test and can be used with any differentiable, explicit likelihood generative model.
-We show that this leads to a more accurate out-of-distribution test. 
+In this follow-up work to \cref{chp:paper-hierarchical}, we note that the set of methods proposed for out-of-distribution detection using generative models is quite large and that many are tailored for specific model types, which suggests that it is possible to develop a model-agnostic approach.
+We hypothesize that by phrasing the task as a statistical testing problem and combining different tests, the method's efficacy can be improved and weaknesses inherent to any particular test can be alleviated.
+From this hypothesis, we combine a classical parametric test with the recently introduced typicality test to develop a method applicable to any differentiable  generative model with explicit likelihood, and show that this leads to a more accurate out-of-distribution test.
 Finally, we discuss the benefits of casting out-of-distribution detection as a statistical testing problem, for instance enabling false positive rate control.
-This is valuable for practical applications, especially in high-risk settings such as medical decision-making.
+This property is valuable in many practical applications, especially in high-risk settings such as medical decision-making.
 
@@
 
+\contribitem{chp:paper-brief}
+In this chapter, we present a comprehensive overview of unsupervised neural representation learning for speech.
+Previous research is categorized into self"-supervised methods and probabilistic latent variable models and described in a common notation.
+This description assists in developing a model taxonomy that shapes a discussion of the models' representational power, the associated learning strategies, and the methods used to evaluate them.
+The discussion points to interesting avenues of future research.
+An extended version of this overview paper that focuses exclusively on self"=supervised methods was also published as part of the project \ref{enumerate: paper-review} \parencite{mohamed_selfsupervised_2022}.
+This paper is included in \cref{app:paper-review} for reference.
+
 \contribitem{chp:paper-benchmarking}
 This chapter develops a novel hierarchical latent variable model for speech, drawing inspiration from the Clockwork VAE \parencite{saxena_clockwork_2021}.
-Comparative benchmarks against alternative latent variable models and autoregressive models for speech highlight improvements to likelihood brought by using hierarchical latent variables.
-Finally, the latent space of the models are analyzed in terms of phonetic content.
-
-\contribitem{chp:paper-brief}
-In this chapter, presents a comprehensive overview of unsupervised neural representation learning for speech.
-Previous research is categorized into self"-supervised methods and probabilistic latent variable models and described in a common notation.
-This description assists in developing a model taxonomy that shapes a discussion of the models' representational power, the associated learning strategies, and the methods used to evaluate them.
-An extended version of this overview paper that focuses exclusively on self"=supervised methods was also published as part of the project \ref{enumerate: paper-review} \parencite{mohamed_selfsupervised_2022}.
-This paper is included in \cref{app:paper-review} for reference.
+A comparative benchmark against alternative latent variable models and autoregressive models for speech highlights the improvements to likelihood brought by using hierarchical latent variables.
+The paper also analyzes the latent spaces learned by the models in terms of phonetic content.
 
@@
 \contribitem{part:medical-applications}
-This part contains studies on machine learning methods applied for tasks in a medical setting.
-While uncertainty estimation is not a central theme to the two papers, the first paper performs a substantial evaluation of the explainability of the proposed model via an occlusion analysis on the text input.
-Later, in the discussion, we shall further consider uncertainty estimation in relation to these two papers.
-\moretodo[inline]{Revise and add more details to this section. Do we discuss uncertainty for medical coding?}
+This part contains two studies on machine learning methods applied for tasks in a medical setting.
+The contributions of the first are methodological focusing on improved comparability and reproducibility of studies on automated medical coding.
+The second is a retrospective study on machine learning-assisted stroke recognition focusing on the potential clinical impact of using machine learning to recognize stroke cases in emergency calls.
+While uncertainty estimation is not a central theme to the two papers, the retrospective study performs a substantial evaluation of the explainability of the proposed model via an occlusion analysis on the text input.
+Later, in the discussion, we shall further consider uncertainty estimation in relation to medical applications focusing on the retrospective study.
+% \moretodo[inline]{Revise and add more details to this section. Do we discuss uncertainty for medical coding?}
 
@@
 We hypothesize that several previous works underperform for reasons more related to suboptimal hyperparameter tuning, incorrect evaluation, and crude data handling than to model design, and that performance and comparability can be improved by addressing these issues.
-With our work, we aim to improve reproducibility of future work and to provide a baseline for future comparisons. 
-We first reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is indeed attributable to weak configuration of the training and crudely sampled train-test splits with many extremely rare classes and several without examples in the training data at all.
+We first reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is indeed attributable to weak configuration of the training and crudely sampled train-test splits with many extremely rare classes - several without examples in the training data.
 We also identify and correct a widespread error in the calculation of the macro F1-score. 
 To compare models, we propose new data splits created with stratified sampling, use identical experimental setups and tune hyperparameters and decision boundaries. 
-By analyzing prediction errors, we confirm the observation of previous work that all models struggle with rare codes, although long documents only have a negligible impact on performance, contrary to previous claims.
-We also present the first comprehensive results on the newly released MIMIC-IV dataset using the reproduced models.
+By analyzing prediction errors, we confirm the observation of previous work that all models struggle with rare codes, although, contrary to previous claims, long documents only have a negligible impact on performance.
+Finally, we present the first comprehensive results on the recently released MIMIC-IV dataset using the reproduced models.
 

### 2023-11-21 chapters/introduction.tex
Update description of Corti

@@
 %
-At the start of this project in 2020, Corti's main product was a software system providing decision-support for call-takers that operate the 1-1-2 emergency line in Copenhagen, Denmark, and the 9-1-1 emergency line in Seattle, Washington, in the United States.
-The system is in continued use in 2023, and provides an interface for navigating a graph-based protocol, defined by the local emergency medical services, which call-takers use to triage patients during emergency calls.
-Calls are transcribed in real-time using automatic speech recognition built for the customer-domain and the system makes suggestions for the call-taker to use in the conversation with the caller, including notifications about forgotten protocol questions \parencite{havtorn_multiqt_2020} and potential cases of cardiac arrests \parencite{cite15, cite14}.
+Corti develops decision support software for healthcare professionals such as general practitioners, medical coders, and call-takers at emergency medical services.
+The system supports the professional similar to a copilot for medical interviews, for instance providing a graph-based protocol for triaging and helping to keep track of key information shared in the conversation.
+
+When used by emergency services for example, calls are transcribed in real-time using automatic speech recognition built for the customer-domain and the system makes suggestions for the call-taker to use in the conversation with the caller, including notifications about forgotten protocol questions \parencite{havtorn_multiqt_2020} and potential cases of cardiac arrests \parencite{cite15, cite14}.
 The system logs the details about the call and the actions taken by the call-taker for later use in review, training and quality assurance.
-
-\moretodo[inline]{Update description of Corti. Focus on the current year. }
 

### 2023-11-21 chapters/background.tex
Improve wording

@@
 An alternative approach to representation learning is provided by self"=supervised learning \parencite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020,schneider_wav2vec_2019} which is a form of unsupervised learning where the training objective is derived from the data itself.
-While it is fair to say that not all self"=supervised training objectives are as well-motivated as the ELBO is for VAEs, they have shown impressive results within the fields of natural language processing \parencite{devlin_bert_2018,chen_simple_2020}, speech processing \parencite{schneider_wav2vec_2019}, and computer vision \parencite{chen_simple_2020}.
+While it is fair to say that not all self"=supervised training objectives have as principled a motivation as the VAE ELBO does, they have shown impressive results within the fields of natural language processing \parencite{devlin_bert_2018,chen_simple_2020}, speech processing \parencite{schneider_wav2vec_2019}, and computer vision \parencite{chen_simple_2020}.
 We provide a review and comparison of self"=supervised methods and variational autoencoders for speech representation learning in \cref{chp:paper-brief}. 

### 2023-11-21 chapters/introduction.tex
Improve wording

@@
 % A later example is the Egyptian Rhind Mathematical Papyrus dated to around 1550 BCE which describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \parencite{georges_universal_2001}.
-Later, the Babylonians developed a sophisticated system of mathematics that included algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods for approximating square roots \parencite{fowler_square_1998}.
+Later, the Babylonians developed a sophisticated system of mathematics that included algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods at approximating square roots \parencite{fowler_square_1998}.
 
@@
 The Babylonian number system, for instance, was well-equipped to represent rational numbers but inefficient for approximating irrational numbers. 
-This forced significant approximation errors, for instance the common practice of approximating $\pi$ with 3 \parencite{georges_universal_2001}. 
+This forced significant approximation errors such as the common practice of approximating $\pi$ with 3 \parencite{georges_universal_2001}. 
 % Easy to overlook, the challenges we face in adopting new technologies today are not new but have been faced by humans for thousands of years.
@@
 While efforts to ease human decision-making have been ongoing for thousands of years, it was not until 1642 that the first mechanical calculator was successfully designed by Blaise Pascal.
-Refined by Gottfried Leibniz during the latter half of the 17th century, the mechanical calculator marked the beginning of a new era where rudimentary calculations could be automated and performed with greater speed and accuracy than humans.
+Refined by Gottfried Leibniz during the latter half of the 17th century, the mechanical calculator marked the beginning of a new era where rudimentary calculations could be automated and performed with greater speed and accuracy than by humans alone.
 In the 1820s, Charles Babbage proposed the first programmable mechanical computer, the Difference Engine.
@@
 With the digital revolution continuing into the 21st century and the worldwide spread of the internet, the amount of data and its availability have grown exponentially. 
-This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence that studies how computer algorithms can learn functions from data and make predictions to guide decision-making.
+This has led to a surge of interest in the scientific field of machine learning, a subfield of artificial intelligence, that studies how computer algorithms can learn functions from data and make predictions to guide decision-making.
 Enabled by a massive increase in the performance and availability of parallel computing resources, such methods have set new standards for the abilities of computer systems at various tasks usually reserved for humans, such as vision, speech recognition, and natural language understanding \parencite{lecun_deep_2015}.
@@
 % revealed far fewer crashes, with only 2 of 20 incidents over 1 million miles meeting reporting criteria, it is interesting to note that the cars are all operated in the sunny states of California and Arizona, where driving conditions vary relatively little \parencite{hawkins_waymo_2023}.
-While driver-assistance systems undoubtedly help reduce accidents in many cases, and has great future potential, the issues of current systems highlight the need for machine learning systems that are robust in rare, adverse, and difficult situations.
+While driver-assistance systems undoubtedly help reduce accidents in many cases, and humans have been argued to be worse drivers still, the issues of current systems highlight the need for machine learning systems that are robust in rare, adverse, and difficult situations \parencite{metz_how_2022}.
+% While driver-assistance systems undoubtedly help reduce accidents in many cases, and has great future potential, the issues of current systems highlight the need for machine learning systems that are robust in rare, adverse, and difficult situations.
 
@@
 Autonomous vehicles and systems within healthcare fall in the category of high risk applications.
-To avoid critical errors, these must meet a set of strict requirements focusing on anti-discrimination and robustness.
+To avoid critical errors, these must meet a set of strict requirements focusing on anti"-discrimination and robustness.
 % Limited risk applications such as in computer games must meet transparency criteria, while
@@
 
-\paragraph{Uncertainty in stroke recognition} A reasonable machine learning pipeline to assist in recognizing stroke cases on calls to emergency services might consist of an automatic speech recognition model that transcribes the conversation and a binary classifier that estimates the probability that the transcript describes a case of stroke.
+\paragraph{Uncertainty in stroke recognition} A reasonable machine learning pipeline to assist in recognizing stroke cases on calls to emergency services could consist of an automatic speech recognition model that transcribes the conversation and a binary classifier that estimates the probability that the transcript describes a case of stroke.
 Such a system might be trained on a dataset calls with verified stroke and non-stroke cases and evaluated against a held-out test set of calls where the call-taker has indicated whether they suspect a stroke.
@@
 Additionally, a single clinical note will usually contain several diagnoses and procedures which must all be inferred from the natural language text \parencite{johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}.
-Furthermore, any single code can have several criteria defined by official guidelines that determine under which conditions it is mutually exclusive with other codes, and when other codes must be coded along with it.
-These properties and more make medical coding a very complex, multi-label classification problem \parencite{centersformedicaremedicaidservicesus_icd10cm_2023}.
+Furthermore, any single code can have several criteria defined by official guidelines that determine under which conditions it is mutually exclusive with other codes, and when other codes must be coded along with it \parencite{centersformedicaremedicaidservicesus_icd10cm_2023}.
+These properties make medical coding a very complex, multi-label classification problem.
 % Official guidelines describe how to code specific diagnoses and procedures and combinations thereof . 
@@
 
-\paragraph{Uncertainty in automated medical coding} A reasonable machine learning pipeline to assist in medical coding might consist of a natural language processing model that extracts relevant information from the clinical note and outputs a probability distribution for each of the potential medical codes.
+\paragraph{Uncertainty in automated medical coding} A reasonable machine learning pipeline to assist in medical coding could consist of a natural language processing model that extracts relevant information from the clinical note and outputs a probability distribution for each of the potential medical codes.
 However, training and using such as system comes with several challenges. 
 For instance, the prevalence of different medical codes in the training data is highly imbalanced; it is common to have several orders of magnitude difference between the frequency of the most frequent code and the least frequent code.
-Since each clinical note is associated with multiple codes that have complex co-occurrence patterns, it is often impossible to exactly correct for this class imbalance by stratified sampling, which is an otherwise common approach to deal with class imbalance in machine learning.
+Since each clinical note is associated with multiple codes that have complex co-occurrence patterns, it is often impossible to exactly correct for this class imbalance by stratified sampling, which is the usual go-to approach to deal with class imbalance in machine learning.
 % This also gives rise to 
@@
 For example, if the model is uncertain about two similar codes, it might be appropriate to ask a human expert to review the note and make the final decision.
-Additionally, the lack of training data for the rarest codes makes them difficult to learn to robustly predict, and so, we might expect the model to often be uncertain for rare codes.
+Additionally, the lack of training data for the rarest codes makes them difficult to learn to robustly predict, and so, we might reasonably expect the model to often be uncertain for rare codes.
 
@@
 Several factors define the reliability of a machine learning model. 
-Besides model performance and accuracy, important factors include the interpretability of how the model functions, the explainability of its predictions, fairness in its treatment of different groups, and robustness to noise, outliers and adversarial attacks.
+Besides model performance and accuracy, important factors include the \emph{interpretability} of how the model functions, the \emph{explainability} of its predictions, \emph{fairness} in its treatment of different groups, and \emph{robustness} to noise, outliers, and adversarial attacks.
 Since many modern machine learning models are deep neural networks with millions or billions of parameters, their size and complexity make them inherently difficult to interpret, and their predictions hard to explain.
@@
 The ability of a model to accurately estimate the uncertainty of its predictions plays an important role in its reliability.
-While an accurate uncertainty estimate provides a certain degree of interpretability, it especially improves robustness by indicating that uncertain predictions are not to be trusted. 
-However, overfitting, memorization and the use of training objectives that are proxies for the evaluation metrics of interest often lead to models that are miscalibrated; that is, the predicted probability of a class does not reflect the true probability of the model being correct \parencite{guo_calibration_2017, kull_temperature_2019}.
+While associating any prediction with an accurate uncertainty estimate can be argued to improve both interpretability, explainability, and fairness, this is generally done through improved robustness to noise, outliers, and adversarial attacks.
+Given an input which cannot be mapped to a single output with certainty, the model can indicate that this prediction should not to be trusted. 
+However, overfitting, memorization, and the use of training objectives that are proxies for the evaluation metrics of interest often lead to models that are miscalibrated; that is, the predicted probability of a class does not reflect the true probability of the model being correct \parencite{guo_calibration_2017, kull_temperature_2019}.
 % This means, for instance, that if for a certain input the model assigns 70\% probability to a class, it would not generally be right 70\% of the time - if we could repeat the prediction multiple times under similar circumstances \parencite{guo_calibration_2017, kull_temperature_2019}.
@@
 For instance, take a perfectly calibrated model trained to classify images of cats and dogs.
-If presented with an image of a horse, it has no option but to distribute 100\% total probability across the cat and dog categories, even though that is clearly wrong.
-Worse yet, to indicate the uncertainty, we might expect the model to assign 50\% probability to each of the cat and dog categories, but that behavior is not guaranteed. 
-Specifically, since no horses were in the training data, the model will not have learned specialized features for horses, nor learned to associate relevant known features with horses.
-So, if a particular horse has features that resemble those learned for a dog more than for a cat, the model might assign arbitrarily high probability to the dog class.
-Therefore, even perfectly calibrated models risks being confidently wrong \parencite{zhou_survey_2022}.
+If presented with an image of a horse, the model has no option but to distribute 100\% total probability across the cat and dog categories, even though that is clearly wrong.
+Worse yet, to indicate uncertainty, we might expect the model to assign 50\% probability to each of the cat and dog categories, but sadly that behavior is not guaranteed. 
+On the contrary, since no horses were in the training data, the model will not have learned specialized features for horses, nor learned to associate any relevant known features with horses.
+So, if a particular horse has features that resemble those learned for a dog more than for a cat, the model might assign arbitrarily high probability to the dog class; and vice versa.
+Therefore, even perfectly calibrated models risk being confidently wrong \parencite{zhou_survey_2022}.
 
@@
 %, especially if they require outputs unavailable to the model.
-In high-dimensional data with a practically unlimited diversity in out-of-distribution examples, it is impossible to collect enough data to eliminate all potential sources of epistemic uncertainty.
+In high-dimensional data with a practically unlimited diversity in out-of-distribution examples, it is impossible to collect enough data to completely eliminate all potential sources of epistemic uncertainty.
 By including horses in the training data, the model would likely learn to recognize them, but would still be unable to recognize giraffes, or zebras, or unicorns.
@@
 
-This introduction provided a high-level overview of the motivation behind the research project by focusing on the use-cases of stroke recognition and automated medical coding as well as speech processing and the importance of uncertainty quantification.
-
+In this introduction we provided a high-level overview of the motivation behind the research project by focusing on the use-cases of stroke recognition and automated medical coding as well as speech processing and the importance of uncertainty quantification.
 The remainder of \cref{part:background} consists of two chapters. 
@@
 \Cref{part:unsupervised-speech-representation-learning} consists of two papers that explore speech representation learning with variational autoencoders and self"=supervised methods \parencite{borgholt_brief_2022,havtorn_benchmarking_2022}.
-\Cref{part:medical-applications} consists of two papers on applications of machine learning within the medical domain, specifically recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}.
-
+\Cref{part:medical-applications} consists of two papers on applications of machine learning within the medical domain, specifically medical coding of clinical notes \parencite{edin_automated_2023} and recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023}.
 Finally, \cref{part:discussion-and-conclusion} concludes the thesis by discussing the presented work and future directions for research.

### 2023-11-22 chapters/background.tex
Update final paragraph

@@
 \paragraph{Self"=supervised learning}
-\moretodo[inline]{Revise according to change of review papers.}
 An alternative approach to representation learning is provided by self"=supervised learning \parencite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020,schneider_wav2vec_2019} which is a form of unsupervised learning where the training objective is derived from the data itself.
@@
 In \cref{app:paper-review} we provide a comprehensive introduction to and review of self"=supervised learning in speech recognition.
-% For a deeper comparison of VAEs and self"=supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{chp:paper-brief}.

### 2023-11-22 chapters/paper_hierarchical.tex
Add source reference at the top of every paper chapter
Fix references
Add citation to top reference
Update initial reference

@@
 \label{chp:paper-hierarchical}
+
+\textit{This chapter is a piece of original research published as part of the project:} \newline
+\begin{center}
+    \begin{enumerate}[leftmargin=8mm,rightmargin=8mm,topsep=0mm,label={[\Alph*]}]
+        \setcounter{enumi}{0}
+        \item \fullcite{havtorn_hierarchical_2021} \main \. \parencite{havtorn_hierarchical_2021}
+        \end{enumerate}
+\end{center}
+
 \ifthenelse{\equal{\skippapers}{true}}{}{
@@
 \begin{figure}
-    %\begin{figure}
     \centering
-    \resizebox{0.25\textwidth}{!}{
+    % \resizebox{0.95\textwidth}{!}{
     \tikz{
@@
         \node[obs] (x_inf) {$\xb$};
-        \node[latent,above=.75cm of x_inf](z1_inf){$\zb^{(1)}$};
-        \node[latent,above=.75cm of z1_inf](z2_inf){$\zb^{(2)}$};
-        % \node[latent,above=.75cm of z2_inf](z3_inf){$\zb^{(3)}$};
-        \node[above=of z2_inf, yshift=-1.cm] (phi) {$q_\phib(\zb|\xb)$};
+        \node[latent,right=.75cm of x_inf](z1_inf){$\zb^{(1)}$};
+        \node[latent,right=.75cm of z1_inf](z2_inf){$\zb^{(2)}$};
+        \node[above=of z2_inf, xshift=-1.cm] (phi) {$q_\phib(\zb|\xb)$};
 
@@
         \edge[]{z1_inf}{z2_inf}; %
-        % \edge[]{z2_inf}{z3_inf}; %
         \edge[dashed, bend left]{x_inf}{z2_inf}; %
-        % \edge[dashed, bend left]{x_inf}{z3_inf}; %
 
@@
         % nodes$
-        \node[obs,right=0.75cm of x_inf] (x_gen) {$\xb$};
-        \node[latent,above=.75cm of x_gen](z1_gen){$\zb^{(1)}$};
-        \node[latent,above=.75cm of z1_gen](z2_gen){$\zb^{(2)}$};
-        % \node[latent,above=.75cm of z2_gen](z3_gen){$\zb^{(3)}$};
-        \node[right=0.75cm of phi] (theta) {$p_\thetab(\xb,\zb)$};
+        \node[latent,right=.75cm of z2_inf](z2_gen){$\zb^{(2)}$};
+        \node[latent,right=.75cm of z2_gen](z1_gen){$\zb^{(1)}$};
+        \node[obs,right=0.75cm of z1_gen] (x_gen) {$\xb$};
+        \node[above=of z1_gen] (theta) {$p_\thetab(\xb,\zb)$};
 
         % edges
-        % \edge[]{z3_gen}{z2_gen}; %
         \edge[]{z2_gen}{z1_gen}; %
@@
         \edge[dashed, bend left]{z2_gen}{x_gen}; %
-        % \edge[dashed, bend left]{z3_gen}{x_gen}; %
     }
-    }
+    % \tikz{
+    %     % inference
+    %     % nodes
+    %     \node[obs] (x_inf) {$\xb$};
+    %     \node[latent,above=.75cm of x_inf](z1_inf){$\zb^{(1)}$};
+    %     \node[latent,above=.75cm of z1_inf](z2_inf){$\zb^{(2)}$};
+    %     % \node[latent,above=.75cm of z2_inf](z3_inf){$\zb^{(3)}$};
+    %     \node[above=of z2_inf, yshift=-1.cm] (phi) {$q_\phib(\zb|\xb)$};
+
+    %     % edges
+    %     \edge[]{x_inf}{z1_inf}; %
+    %     \edge[]{z1_inf}{z2_inf}; %
+    %     % \edge[]{z2_inf}{z3_inf}; %
+    %     \edge[dashed, bend left]{x_inf}{z2_inf}; %
+    %     % \edge[dashed, bend left]{x_inf}{z3_inf}; %
+
+    %     % generative
+    %     % nodes$
+    %     \node[obs,right=0.75cm of x_inf] (x_gen) {$\xb$};
+    %     \node[latent,above=.75cm of x_gen](z1_gen){$\zb^{(1)}$};
+    %     \node[latent,above=.75cm of z1_gen](z2_gen){$\zb^{(2)}$};
+    %     % \node[latent,above=.75cm of z2_gen](z3_gen){$\zb^{(3)}$};
+    %     \node[right=0.75cm of phi] (theta) {$p_\thetab(\xb,\zb)$};
+
+    %     % edges
+    %     % \edge[]{z3_gen}{z2_gen}; %
+    %     \edge[]{z2_gen}{z1_gen}; %
+    %     \edge[]{z1_gen}{x_gen}; %
+    %     \edge[dashed, bend left]{z2_gen}{x_gen}; %
+    %     % \edge[dashed, bend left]{z3_gen}{x_gen}; %
+    % }
+    % }
     \caption[Inference and generative models for a bottom-up hierarchical VAEs.]{ The inference and generative models, $q_\phib$ and $p_\thetab$, for an $L=2$ layered bottom-up hierarchical VAE as the one used in our experiments.
@@
     \label{fig_hierarchical:hvae-graphical-model}
-    % \end{figure}
 %\end{wrapfigure}

### 2023-11-22 chapters/background.tex
Add note on notation for density of random variable

@@
 \begin{equation} \label{eq:corrected-differential-entropy-jaynes}
-    H_{N\rightarrow\infty}(X) \equiv \lim_{N\rightarrow\infty} [H_{N}(X) - \log N] = - \int_{\mathcal{X}} p(x) \log \left(\frac{p(x)}{q(x)}\right) \, dx \enspace ,
+    H_{N\rightarrow\infty}(X) \equiv \lim_{N\rightarrow\infty} [H_{N}(X) - \log N] = - \int_{\mathcal{X}} p_X(x) \log \left(\frac{p_X(x)}{q_X(x)}\right) \, dx \enspace ,
 \end{equation}
 %
-where $H_N(X)$ is the limiting density for discrete points, $q(x)$ is a uniform density over the quantization of the continuous space, and we have subtracted a $\log N$ term that would go to infinity in the limit of infinite points.
+where $H_N(X)$ is the limiting density for discrete points, $q_X(x)$ is a uniform density over the quantization of the continuous space, and we have subtracted a $\log N$ term that would go to infinity in the limit of infinite points.
 By doing this, the information entropy becomes positive, dimensionally correct, and invariant under a change of variables%
@@
         \widetilde{w}(y)\,dy = w(x)\,dx \enspace , \quad
-        \widetilde{q}(y)\,dy = q(x)\,dx \enspace .
+        \widetilde{q}(y)\,dy = q_X(x)\,dx \enspace .
     \end{align*}
@@
 
-The form on the right-hand side of \cref{eq:corrected-differential-entropy-jaynes} can be recognized as the negative Kullback-Leibler divergence of $p(x)$ to $q(x)$ \parencite{kullback_information_1959},
+The form on the right-hand side of \cref{eq:corrected-differential-entropy-jaynes} can be recognized as the negative Kullback-Leibler divergence of $p_X(x)$ to $q_X(x)$ \parencite{kullback_information_1959},
 %
 \begin{equation} \label{eq:kullback-leibler-divergence}
-    D_{\mathrm{KL}}(p(x)\parallel q(x)) = \int_{\mathcal{X}} p(x) \log \left(\frac{p(x)}{q(x)}\right) \enspace .
+    D_{\mathrm{KL}}(p_X(x)\parallel q_X(x)) = \int_{\mathcal{X}} p_X(x) \log \left(\frac{p_X(x)}{q_X(x)}\right) \enspace .
 \end{equation}
 %
-The Kullback-Leibler divergence $D_{\mathrm{KL}}(p(x)\parallel q(x))$, also known as the \emph{information gain}, is often interpreted as the amount of additional information required to represent events from a distribution with density $p(x)$ using a code optimized for a distribution with density $q(x)$.
+The Kullback-Leibler divergence $D_{\mathrm{KL}}(p_X(x)\parallel q_X(x))$, also known as the \emph{information gain}, is often interpreted as the amount of additional information required to represent events from a distribution with density $p_X(x)$ using a code optimized for a distribution with density $q_X(x)$.
 In other words, it measures how surprised one would be if they used distribution $q$ to represent events from distribution $p$, i.e.
@@
 In \cref{chp:paper-hierarchical} we shall see how the Kullback-Leibler divergence emerges in a likelihood-ratio test statistic for out-of-distribution detection using hierarchical variational autoencoders.
+In the remainder of the thesis, we will write the density of a random variable $X$ as $p(x)$, dropping the subscript, as is usual in machine learning. 
 % In the following, we will see that this interpretation of information entropy as a measure of diversity is useful for understanding the concept of uncertainty in machine learning.
@@
 \section{Out-of-distribution detection} \label{sec:out-of-distribution-detection}
-\lesstodo[inline]{Maybe this section should come after the VAE background?}
+% \lesstodo[inline]{Maybe this section should come after the VAE background?}
 % 
@@
 %
-In this section we introduce out-of-distribution (OOD) detection and connect it to the concept of uncertainty. 
-Since the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic} are relatively short, we also provide a concise review of the literature on out-of-distribution detection.
+In \cref{subsec: overview-of-out-of-distribution-detection} we introduce out-of-distribution (OOD) detection and connect it to the concept of uncertainty. 
+Since the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic} are relatively short, we also provide a concise review of the literature on out-of-distribution detection in \cref{subsec: supervised-out-of-distribution-detection, subsec: unsupervised-out-of-distribution-detection}.
 
-\subsection{An overview of the task}
+\subsection{An overview of the task}\label{subsec: overview-of-out-of-distribution-detection}
 %
-OOD detection is the task of identifying test data that is unlikely to originate from the distribution of the training data and, in the context of neural networks, dates back several decades \parencite{bishop_novelty_1994, chang_figure_1993}.
+OOD detection is the task of identifying test data that are unlikely to originate from the distribution of the training data and, in the context of neural networks, dates back several decades \parencite{bishop_novelty_1994, chang_figure_1993}.
 In the general case, we assume that we have a domain of in-distribution data, $\mathcal{D}_{\text{in}}$, and we would like to build a model that can be used to assess whether a test data point $\xb$ originates from that domain or not.
@@
 
-\subsection{Supervised out-of-distribution detection}
+\subsection{Supervised out-of-distribution detection} \label{subsec: supervised-out-of-distribution-detection}
 % OUTLINE:
@@
 
-\subsection{Unsupervised out-of-distribution detection}
+\subsection{Unsupervised out-of-distribution detection} \label{subsec: unsupervised-out-of-distribution-detection}
 % OUTLINE:
@@
 
-\lesstodo[inline]{Do we want to refer to the parsimony gap?}
-\lesstodo[inline]{Maybe add a small graphic illustrating the gaps?}
+% \lesstodo[inline]{Do we want to refer to the parsimony gap?}
+% \lesstodo[inline]{Maybe add a small graphic illustrating the gaps?}
 % A final error is introduced  the maximally attainable likelihood over all the group $G$ of possible models, $\mathcal{l}(G)$ 
@@
 Training VAEs by stochastic gradient descent involves first sampling a mini-batch of data points $\xb$ from the empirical distribution of training data $\hat{p}(\xb)$ and then evaluating the gradient of the ELBO with respect to the parameters $\thetab$ and $\phib$.
-To gain some additional insight into the objective that we are optimizing, we will use this expectation over the data distribution $\hat{p}(\xb)$ to form a factorization similar to \cref{eq:variational-autoencoder-elbo-factorized} but with the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$ in the KL-divergence \parencite{tomczak_trouble_2022}.
+We can note that evaluating the ELBO with a mini-batch sampled from the training data corresponds to taking an expectation over the data distribution $\hat{p}(\xb)$.
+
+To gain some additional insight into the objective that we are optimizing, we will use this expectation $\hat{p}(\xb)$ to form a factorization similar to \cref{eq:variational-autoencoder-elbo-factorized} but with the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$ in the KL-divergence \parencite{tomczak_trouble_2022}.
 Under the expectation over the data distribution, the marginal likelihood is lower bounded by that expectation over the ELBO,
@@
 where we have defined $q_{\phib}(\xb,\zb) = q_{\phib}(\zb|\xb)\hat{p}(\xb)$ for ease of notation. This is simply \cref{eq:variational-autoencoder-elbo} under the extra expectation. 
-We leave the reconstruction loss in \cref{eq:variational-autoencoder-elbo-factorized} is as and focus on the KL-divergence to the prior. We rewrite it as follows,
+We leave the reconstruction loss in \cref{eq:variational-autoencoder-elbo-factorized} is as and focus on the expectation over the KL-divergence to the prior. We rewrite it as follows,
 %
@@
 However, this alternative form reveals that the ELBO is maximized by minimizing the mutual information between $\xb$ and $\zb$, $I_{q_{\phib}(\xb,\zb)}\left[\xb ; \zb \right]$. 
-This result indicates that the usual interpretation of $\zb$ as a representation of $\xb$ is not a fundamental property arising from the ELBO and challenges the usefulness of variational autoencoders as representation learners.
+This result indicates that the usual interpretation of $\zb$ as a representation of $\xb$ is not a fundamental property arising from the ELBO, or from the assumptions from which the VAE was derived.
+This challenges the usefulness of variational autoencoders as representation learners.
 

### 2023-11-22 chapters/discussion.tex
Work on discussion

@@
 I'm sure you can find something better.}
-%
 
 The rapid progress of machine learning that started about a decade ago with the 2012 ImageNet competition and the work of \textcite{krizhevsky_imagenet_2012} all but slowed down during the course of this project.
-Following the papers by \textcite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019}, the field of out-of-distribution detection saw a surge of interest that has since grown yearly, while self"=supervised learning for speech has continued to advance with works such as CPC \parencite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.
+Following the papers by \textcite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019}, the field of out-of-distribution detection saw a surge of interest that has since grown yearly, while representation learning for speech has continued to advance with notable works such as CPC \parencite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.
 
@@
 Two key observations might lead to the pessimistic view that VAEs are fundamentally unsuited for representation learning. 
-First, VAEs are trained to maximize the ELBO which is a marginal likelihood objective conditionally independent of the latent variables $\zb$.
-Second, as we saw in \cref{eq:variational-autoencoder-elbo-factorized-expected}, the ELBO contains can be written to reveal that its maximization leads to minimization of the mutual information between the observed and latent variables.
+First, VAEs are trained to maximize the ELBO which includes a marginal likelihood objective that is independent of the latent variables $\zb$.
+Second, as we saw in \cref{eq:variational-autoencoder-elbo-factorized-expected}, the ELBO can be written to reveal that its maximization leads to minimization of the mutual information between the observed and latent variables.
 In the following we will discuss each of these observations.
@@
 Any combination uniquely defines a valid latent variable model.
-Hence, maximizing the marginal is a completely useless criteria for representation learning, regardless of the measure of usefulness \parencite{huszar_is_2017, alemi_fixing_2018}.
+Hence, maximizing the marginal is a useless criterion for representation learning, regardless of the measure of usefulness \parencite{huszar_is_2017, alemi_fixing_2018}.
+
+However, VAEs are not trained to maximize the marginal likelihood directly, but a lower bound of it, the ELBO,
+%
+\begin{equation} \label{eq:variational-autoencoder-elbo-factorized (discussion copy)}
+    \mathcal{L}_{}(\xb;\thetab,\phib)
+    = 
+    \underset{\text{reconstruction loss}}{\underbrace{
+        \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]
+    }}
+     - 
+    \underset{\text{KL-divergence to prior}}{
+        \underbrace{\vphantom{\mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]}{D_\text{KL}\left( q_\phib(\zb|\xb) \parallel p(\zb) \right)}
+    }} \enspace .
+\end{equation}
+%
+Since the ELBO is not independent of the latent variables, its maximization will typically depend on the latent variables and the posterior $p(\zb|\xb)$.
+Hence, unlike the marginal likelihood, the ELBO is not guaranteed to be a useless criterion for representation learning.
+Nonetheless, being a lower bound of the marginal likelihood that we seek to maximize, seems to indicate that this issue might not be irrelevant to consider.
 
@@
 %
-\begin{equation} %\label{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)}
+\begin{equation} \label{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)}
     \mathbb{E}_{q_{\phib}(\xb,\zb)} \left[ \mathcal{L}_{}(\xb;\thetab,\phib) \right] = 
@@
     I_{q_{\phib}(\xb,\zb)}\left[\xb ; \zb \right]
-    }} \. . \nonumber
+    }} \. . %\nonumber
 \end{equation}
@@
 Note that we let $q_{\phib}(\xb,\zb) = q_{\phib}(\zb|\xb)\hat{p}(\xb)$ for ease of notation. 
-Optimizing the ELBO can be understood as simultaneously maximizing the average reconstruction likelihood, minimizing the marginal KL-divergence to the prior, and minimizing the mutual information between the observed and latent variables.
-
-\parencite{tomczak_trouble_2022}
-
-\moretodo[inline]{Finish the discussion of the MI interpretation}
+% Optimizing the ELBO can be understood as simultaneously maximizing the average reconstruction likelihood, minimizing the marginal KL-divergence to the prior, and minimizing the mutual information between the observed and latent variables.
+We can see that optimizing the ELBO means learning a good one-to-one mapping of $z$ to $x$ (average reconstruction), fitting the aggregated posterior to the prior (marginal KL-divergence), while also not learning any stochastic dependency between $x$ and $z$ (mutual information).
+These objectives are conflicting and especially the minimization of mutual information seems counterproductive to our goals \parencite{tomczak_trouble_2022}.
 
@@
 Similarly, maximizing the ELBO in \cref{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)} by exactly minimizing the mutual information term corresponds to a complete posterior collapse that requires a trade-off with the other terms.
-This solution is almost surely a local minimum and is often avoided in practice.
+For a given model, this solution is almost surely a local minimum, but one that is often avoided in practice.
 The challenge is then how to impose appropriate constraints to make VAEs learn representations that are competitive with self"=supervised methods.
@@
 
-Including masked objectives into the principled probabilistic framework of VAEs might require creative thinking, but it seems likely that relaxing the requirement of exact input reconstruction and enforcing the need of encoding (and decoding) contextual information could lead the way for future types of VAEs to learn representations that are competitive with self"=supervised methods.
+Including masked objectives into the principled probabilistic framework of VAEs might require rethinking the principles themselves, but it seems likely that relaxing the requirement of exact input reconstruction and enforcing the need for encoding (and decoding) contextual information could pave the way for future types of VAEs to learn representations that are competitive with self"=supervised methods.
 
@@
 
+Our work on stroke recognition in \cref{chp:paper-retrospective} focuses on the predictive performance of the ensemble model and an analysis of feature importance, but does not explicitly consider uncertainty estimation.
+As we discussed in \cref{subsec:model-calibration}, such a model can be calibrated to predict probabilities that are aligned with the empirical probability of the model being correct on some validation set.
+Here we present and discuss model calibration for the ensemble model.
+
+We compute the calibration curve by sorting the probabilities predicted on the test set into a number of bins spanning the range from zero to one.
+For each bin $b$, we compute the mean predicted probability $\bar{p}$ and the fraction of examples for which the model predicted correctly $r_{b}$.
+The calibration curve is the drawn from the $\{(\bar{p}_{b}, r_{b})\}_b$ pairs.
+For any given bin, a perfectly calibrated model would have the same fraction of correct predictions as that bin's mean value, $\bar{p}_{b} = r_{b}\,\forall\,b$.
+
 \begin{figure}
@@
 \end{figure}
-
-Our work on stroke recognition in \cref{chp:paper-retrospective} focuses on the predictive performance of the ensemble model and an analysis of feature importance, but does not explicitly consider uncertainty estimation.
-As we discussed in \cref{subsec:model-calibration}, such a model can be calibrated to predict probabilities that are aligned with the empirical probability of the model being correct on some validation set.
-Here we present and discuss model calibration for the ensemble model.
-
-We compute the calibration curve by sorting the probabilities predicted on the test set into a number of bins spanning the range from zero to one.
-For each bin $b$, we compute the mean predicted probability $\bar{p}$ and the fraction of examples for which the model predicted correctly $r_{b}$.
-The calibration curve is the drawn from the $\{(\bar{p}_{b}, r_{b})\}_b$ pairs.
-For any given bin, a perfectly calibrated model would have the same fraction of correct predictions as that bin's mean value, $\bar{p}_{b} = r_{b}\,\forall\,b$.
 
@@
 Calibrated probabilities in the alerts presented to users might be a way to alleviate alarm fatigue.
-Calibrated probabilities would allow users to discern between certain and uncertain predictions and also enable the system to present users only with predictions that have at least a certain probability of being correct.
+Calibrated probabilities would allow users to discern between certain and uncertain predictions and also enable the system to present users only with predictions that have a minimum probability of being correct.
 Similar approaches have been suggested by clinicians and interviews indicate that predictive uncertainty is perceived by experts as a sort of explanation that complements the prediction \cite{tonekaboni_what_2019}.

### 2023-11-24 chapters/paper_retrospective.tex
Add citation to top reference
Update initial reference
Update top reference
Convert retrospective from EN-UK to EN-US

@@
         \setcounter{enumi}{5}
-        \item \fullcite{wenstrup_retrospective_2023} \shared
+        \item \fullcite{wenstrup_retrospective_2023} \shared \hspace{0.1em} \parencite{wenstrup_retrospective_2023}
     \end{enumerate}
@@
 \section*{Abstract}
-Advanced stroke treatment is time dependent and, therefore, relies on recognition by call-takers at prehospital telehealth services to ensure fast hospitalisation.
+Advanced stroke treatment is time dependent and, therefore, relies on recognition by call-takers at prehospital telehealth services to ensure fast hospitalization.
 This study aims to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls.
@@
 The machine learning framework performs significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
-Thus, a machine learning framework for recognising stroke in prehospital medical helpline calls may become a supportive tool for call-takers, aiding in early and accurate stroke recognition.
+Thus, a machine learning framework for recognizing stroke in prehospital medical helpline calls may become a supportive tool for call-takers, aiding in early and accurate stroke recognition.
 
@@
 Effective treatment is time-sensitive, and an optimal outcome is more likely when treatment is administered within the first four and a half hours from stroke onset \parencite{cite4,cite5}.
-The gateway to ambulance transport and hospital admittance is through prehospital telehealth services, including emergency medical call centres, nurse advice call lines, and out-of-hours health services.
+The gateway to ambulance transport and hospital admittance is through prehospital telehealth services, including emergency medical call centers, nurse advice call lines, and out-of-hours health services.
 In the pre-hospital setting, the use of mobile stroke units has made it possible to deliver advanced treatment faster \parencite{cite6,cite7}.
 As the mobile stroke unit is only dispatched to patients with a suspected stroke, the impact of mobile stroke unit is directly influenced by accurate call-taker recognition of stroke \parencite{cite6,cite7}.
-Call-takers who can rapidly and accurately recognise stroke are therefore crucial in facilitating prompt care in both pre-hospital and in-hospital settings.
+Call-takers who can rapidly and accurately recognize stroke are therefore crucial in facilitating prompt care in both pre-hospital and in-hospital settings.
 
@@
 We then fit a machine learning framework to classify medical helpline calls as stroke or non-stroke.
-Calls are first transcribed using an automatic speech recognition model and then categorised by a text classification model trained as an ensemble of five individual models.
+Calls are first transcribed using an automatic speech recognition model and then categorized by a text classification model trained as an ensemble of five individual models.
 We compare the performance of the model with that of call-takers using MH-1813 data from 2021.
@@
 We performed an occlusion analysis to evaluate the importance of individual words for both positive and negative classifier predictions \cref{tab_retrospective:table3-occlusion-analysis}. 
-Among the words with a positive rank score, several words are synonymous with stroke, such as ``blood clot", ``haemorrhagic stroke", and ``stroke".
+Among the words with a positive rank score, several words are synonymous with stroke, such as ``blood clot", ``hemorrhagic stroke", and ``stroke".
 Ambulances are rarely dispatched because the MH-1813 is not intended for emergencies.
@@
         \midrule
-                    & Word, $w$ \textit{(translated)} & Occurences, $D^{(w)}$ & Word, $w$ \textit{(translated)} & Occurences, $D^{(w)}$ \\
+                    & Word, $w$ \textit{(translated)} & Occurrences, $D^{(w)}$ & Word, $w$ \textit{(translated)} & Occurrences, $D^{(w)}$ \\
         \midrule
@@
         13. & Call & 39 & Prescription & 5,450 \\        
-        14. & Numb & 94 & Centimetre & 12,026 \\        
+        14. & Numb & 94 & Centimeter & 12,026 \\        
         15. & Minutes & 763 & The knee & 8,875 \\        
         16. & Difficulties speaking & 44 & The pharmacy & 10,085 \\        
-        17. & Haemorrhagic stroke & 133 & The stomach & 42,105 \\        
+        17. & Hemorrhagic stroke & 133 & The stomach & 42,105 \\        
         18. & Hand & 297 & Psychiatric & 3,688 \\        
@@
 Instead, the model should be used as a supportive tool for call-takers in the decision-making process, contributing to a higher recognition of patients with stroke and potentially boosting the confidence of call-takers in their decisions.
-A similar machine learning model designed to predict cardiac arrest was tested in a randomised controlled trial (RCT) at CEMS \parencite{cite15}.
+A similar machine learning model designed to predict cardiac arrest was tested in a randomized controlled trial (RCT) at CEMS \parencite{cite15}.
 The results highlighted the necessity of incorporating input from call-takers.
@@
 Additionally, certain stroke symptoms are so rare that some call-takers may never encounter them, increasing the risk of false negatives.
-The model was trained on more calls than any single call-taker would handle in a lifetime, enabling it to recognise even rare descriptors of stroke.
-The model is specifically trained to recognise strokes and exclusively learns from actual stroke descriptions, unlike call-takers, who are trained with generalised teaching materials to triage many different conditions.
+The model was trained on more calls than any single call-taker would handle in a lifetime, enabling it to recognize even rare descriptors of stroke.
+The model is specifically trained to recognize strokes and exclusively learns from actual stroke descriptions, unlike call-takers, who are trained with generalized teaching materials to triage many different conditions.
 Therefore, call-takers may not have received specific training for patients with stroke and may never have encountered them.
@@
 We did not have data to estimate potential bias related to race, ethnicity, language, accent, or dialects.
-Previous studies on speech recognition for call centres have indeed found that non-native speakers had a higher rate of transcription errors \parencite{han_deep_2017}.
+Previous studies on speech recognition for call centers have indeed found that non-native speakers had a higher rate of transcription errors \parencite{han_deep_2017}.
 Since our model was trained on a representative - and therefore unbalanced - sample, we expect it to behave similarly.
-Future research should look to address these shortcomings, for example by utilising self"-supervised learning on massive amounts of diverse, unlabelled data covering multiple languages, accents, and dialects.
+Future research should look to address these shortcomings, for example by utilizing self"-supervised learning on massive amounts of diverse, unlabeled data covering multiple languages, accents, and dialects.
 
@@
 Instead, we used the downstream performance of the text classification model when trained in combination with different speech recognition models to choose the best option.
-Since the focus of this study is the ability to correctly recognise stroke, and not the performance of the speech recognition model alone, this approach is better suited.
+Since the focus of this study is the ability to correctly recognize stroke, and not the performance of the speech recognition model alone, this approach is better suited.
 Indeed, the WER might be misleading when choosing a speech recognition model for a specific task.
@@
 “uh” and “uhm”) and make small inflection errors (e.g.
-“clot” instead of “clots”), which results in a relatively high WER, while another model only fails to predict rare, specialised words that are highly indicative of stroke (e.g.
-“haemorrhage” and “thrombolysis”), which results in a relatively low WER.
+“clot” instead of “clots”), which results in a relatively high WER, while another model only fails to predict rare, specialized words that are highly indicative of stroke (e.g.
+“hemorrhage” and “thrombolysis”), which results in a relatively low WER.
 
 Although we believe that the proposed machine learning framework can be further improved, several alternatives have already been explored in the preliminary experimental phase.
-The speech recognition model we used was trained on 1-1-2 calls for a previous project \parencite{cite14}, and so, was specialised to a domain very similar to that of MH-1813.
+The speech recognition model we used was trained on 1-1-2 calls for a previous project \parencite{cite14}, and so, was specialized to a domain very similar to that of MH-1813.
 We also tested an open-source, multilingual model from OpenAI called Whisper \parencite{radford_robust_2023}, but found that performance degraded slightly compared to the model trained on 1-1-2.
-We hypothesise that this is due to Whisper's inability to handle the specific noise conditions and recognise words from a specialised medical vocabulary.
+We hypothesize that this is due to Whisper's inability to handle the specific noise conditions and recognize words from a specialized medical vocabulary.
 
@@
 Although many of these models are freely available to the public, they are primarily trained on English data.
-Only relatively few options exist for the Danish language, none of which are specialised in the medical domain.
+Only relatively few options exist for the Danish language, none of which are specialized in the medical domain.
 We used a monolingual Danish BERT model, which has previously been shown to outperform a multilingual alternative from Google for Danish named entity recognition \parencite{hvingelby_dane_2020}.
 However, this also did not result in a significant performance improvement.
-We hypothesise that the number of ground truth stroke positives was too small for these advanced models to learn more complex patterns than the MLP ensemble.
+We hypothesize that the number of ground truth stroke positives was too small for these advanced models to learn more complex patterns than the MLP ensemble.
 In addition, a self"=supervised model would likely benefit from being pre"=trained on speech or text data from the target domain.
@@
 Call-takers for both lines, who are nurses, paramedics, or physicians, can dispatch ambulances.
-The condition suspected by the call-taker is categorised based on a predefined diagnostic index and stored in an electronic record using a CAD system.
+The condition suspected by the call-taker is categorized based on a predefined diagnostic index and stored in an electronic record using a CAD system.
 The CAD records are associated with the Danish civil registration number (CPR number) \parencite{cite19} of the patient.
@@
 
-All patients with a final diagnosis of stroke or transient ischaemic attack admitted to a Danish hospital within 5 days of symptom onset are recorded in the Danish Stroke Registry \parencite{cite16}, also known as DanStroke.
-This record includes the patient-reported time of onset, stroke type (haemorrhagic, ischaemic, or transient ischaemic attack), and CPR number of the patient.
+All patients with a final diagnosis of stroke or transient ischemic attack admitted to a Danish hospital within 5 days of symptom onset are recorded in the Danish Stroke Registry \parencite{cite16}, also known as DanStroke.
+This record includes the patient-reported time of onset, stroke type (hemorrhagic, ischemic, or transient ischemic attack), and CPR number of the patient.
 The diagnosis is obtained according to the national guidelines \parencite{baluenfeldt_national_2021}, which includes cerebral imaging and full diagnostic workup by neurologists.
-The validity of the Danish stroke registry has been shown to be high \parencite{wildenschild_registration_2013}, and the number of stroke mimics is therefore minimised in our dataset.
+The validity of the Danish stroke registry has been shown to be high \parencite{wildenschild_registration_2013}, and the number of stroke mimics is therefore minimized in our dataset.
 
@@
 CEMS approved the transcription of all calls made to 1-1-2 and MH-1813.
-All electronic records were anonymised before analysis, and the researchers did not inspect the calls manually.
+All electronic records were anonymized before analysis, and the researchers did not inspect the calls manually.
 
@@
 The CAD records were matched with the telephone call recordings based on the call start, call duration, and call-taker identity.
-Due to data incompleteness, and the way the audio data is stored, at CEMS, 2,730,199 contacts could not be matched to their corresponding audiofile, however, 2,361,178 contacts were successfully matched.
+Due to data incompleteness, and the way the audio data is stored, at CEMS, 2,730,199 contacts could not be matched to their corresponding audio file, however, 2,361,178 contacts were successfully matched.
 We found no obvious pattern in the matched and unmatched calls and we included all calls with at matching audio file.
@@
 We allowed a window of 72 hours before and 24 hours after the call starts to account for uncertainty in recording stroke onset time.
-We excluded calls involving subarachnoid haemorrhage cases.
+We excluded calls involving subarachnoid hemorrhage cases.
 Finally, we considered a call to be a call-taker stroke positive when the call-taker selected the stroke diagnostic category during the call and dispatched an ambulance with the appropriate level of response \parencite{cite20}.
@@
 Instead, we first subtracted the threshold from the output score in the logit space (before sigmoid nonlinearity) for each model to obtain the same threshold (0.5).
-Subsequently, we defined the ensemble output score as the average of the centred output scores.
+Subsequently, we defined the ensemble output score as the average of the centered output scores.
 The exact equations are provided in \cref{eq_retrospective:output_score,eq_retrospective:ensemble_prediction}.
@@
 This involved removing all instances of a given word from the input transcript to evaluate its impact on the model output.
-The word was removed before vectorisation, such that all word and character $n$-grams associated with the word were discarded.
+The word was removed before vectorization, such that all word and character $n$-grams associated with the word were discarded.
 Specifically, let $z^{(n, d, w)}$ be the logit output of model $n$ in the ensemble for transcript $d$ when the word $w$ is occluded.
@@
 %
-where $\text{sign}$ represents the sign function. Squaring $i^{(d,w)}$ favours rare features with a high impact over common features with a low impact.
+where $\text{sign}$ represents the sign function. Squaring $i^{(d,w)}$ favors rare features with a high impact over common features with a low impact.
 
@@
 We computed 95\% confidence intervals (CIs) using bootstrapping \parencite{dwass_modified_1957,eden_validity_1933}.
-In our assessment, we accounted for random variation associated with model training by basing the means, tests, and CIs on the predictions of 11 randomly initialised training runs.
+In our assessment, we accounted for random variation associated with model training by basing the means, tests, and CIs on the predictions of 11 randomly initialized training runs.
 Statistical significance was defined as a p-value of less than 0.05.
@@
 
-The source code can be shared using a Creative Commons NC-ND 4.0, an international licence, upon reasonable written request to the corresponding author, and requires a data use agreement.
+The source code can be shared using a Creative Commons NC-ND 4.0, an international license, upon reasonable written request to the corresponding author, and requires a data use agreement.
 
@@
 
-\section*{Declaration of interests}
-
-JDH and LB received funding from the Innovation Fund Denmark.
-JDH and LB used Corti and hold stock warrants.
-LM is a co-founder, stockholder, and the Chief Technology Officer of Corti.
-JW received funding from Trygfonden.
-SNFB has no conflicts of interest to declare.
-HC has received funding from the Velux Foundation, Tværsfonden, Helsefonden, Hartmann Fonden, Lundbeck Foundation, and Novo Nordisk Foundation; royalties from Gyldendal; honoraria from Bayer and Bristol Meyers Squibb, and is chair of Action Plan for stroke in Europe Implementation, Co-chair of the Scientific Stroke Panel EAN and Senior Guest Editor of AHA Stroke.
-MS has no conflicts of interest to declare.
-HCC has no conflicts of interest to declare.
-CK received funding from the Novo Nordisk Foundation and is the chair of the Danish Resuscitation Council and vice chair of the Danish Stroke Society.
-Both positions are unpaid.
-
 \section*{Author contributions}
 
-JW, JDH and LB share lead authorship and contributed equally to the original drafting of the paper, review and editing, data curation, formal analysis, methodology, validation, and conceptualisation.
+JW, JDH and LB share lead authorship and contributed equally to the original drafting of the paper, review and editing, data curation, formal analysis, methodology, validation, and conceptualization.
 JDH and LB contributed to software development and funding acquisition.
-SNFB contributed to the conceptualisation, data curation, funding acquisition, supervision, and writing - review and editing.
-LM contributed to conceptualisation, funding acquisition, project administration, supervision, and writing - review and editing.
+SNFB contributed to the conceptualization, data curation, funding acquisition, supervision, and writing - review and editing.
+LM contributed to conceptualization, funding acquisition, project administration, supervision, and writing - review and editing.
 MS contributed to supervision and writing - review and editing.
 HC contributed to the methodology, supervision, and writing - review and editing.
-CK contributed to the conceptualisation, funding acquisition, methodology, project administration, supervision, and writing - review and editing.
-HCC contributed to the conceptualisation, data curation, funding acquisition, methodology, project administration, supervision, and writing - review and editing.
+CK contributed to the conceptualization, funding acquisition, methodology, project administration, supervision, and writing - review and editing.
+HCC contributed to the conceptualization, data curation, funding acquisition, methodology, project administration, supervision, and writing - review and editing.
 HCC and CK share last authorship.

### 2023-11-24 chapters/background.tex
Lowercase
Improve VAE intro

@@
 Variational autoencoders (VAEs) belong to the class of deep generative models. 
-These are models that learn a probabilistic representation of the input data and can be used to generate new samples from the same distribution.
+These are models that use deep neural networks to approximate the underlying distribution of the unlabeled training data, from which they can then be used to generate new data samples. 
+% These are models that unsupervisedly learn a representation of data and can be used to generate new samples by learning an approximation to the distribution that generated the training data. 
 There exist at least six main classes of deep generative models: variational autoencoders \parencite{kingma_autoencoding_2014,rezende_stochastic_2014,ranganath_hierarchical_2016,vahdat_nvae_2020,child_very_2021}, normalizing flow models \parencite{dinh_nice_2015,rezende_variational_2015,dinh_density_2017,kingma_glow_2018,grathwohl_ffjord_2018}, diffusion models \parencite{sohl-dickstein_deep_2015,song_generative_2019, ho_denoising_2020, vahdat_scorebased_2021}, energy-based models \parencite{lecun_tutorial_2006,hinton_fast_2006,salakhutdinov_efficient_2010,du_implicit_2019}, autoregressive models \parencite{oord_conditional_2016,oord_wavenet_2016,radford_improving_2018}, and generative adversarial networks \parencite{goodfellow_generative_2014,arjovsky_wasserstein_2017,brock_large_2019,karras_stylebased_2019}.
 
-VAEs are interesting for representation learning as well as uncertainty quantification, due to the assumptions they impose on the data modelling task and their probabilistic formulation.
-To some degree, these properties set them apart from other deep generative models that focus more exclusively on data generation, and we chose to focus on them in this thesis.
-% \moretodo[inline]{Briefly hold the readers hand and tell them why VAEs are relevant to the overarching project.}
-% Besides data generation, common properties of deep generative models include the ability to compute an exact or approximate likelihood of input $p(\xb)$, reconstruct an input $\xb$ from its latent representation $\zb$, and represent an input $\xb$ in a latent space.
-These properties enable various methods for their use in OOD detection.
-% A variational autoencoder (VAE) \parencite{kingma_autoencoding_2014,rezende_stochastic_2014} is a type of deep generative model and an instance of a latent variable model. 
+Due to the assumptions they impose on the data modelling task and their probabilistic formulation, VAEs are interesting for representation learning as well as uncertainty quantification.
+To some degree, these properties set them apart from other deep generative models that focus more exclusively on data generation.
 In the following, we will introduce VAEs starting from the definition of latent variable models.
@@
     - 
-    \underset{\text{Marginal likelihood}}{\underbrace{
+    \underset{\text{marginal likelihood}}{\underbrace{
         \vphantom{\mathbb{E}_{q_\phib(\zb|\xb)}\left[ \log \frac{p_\theta(\zb,\xb)}{q_\phib(\zb|\xb)} \right]}{\log p_\theta(\xb)}

### 2023-11-24 chapters/introduction.tex
Update wording

@@
 % \noindent The research presented in this thesis will focus on the use of machine learning in relation to uncertainty estimation, speech processing, and medical conversations. 
-\noindent The research presented in this thesis deals with the use of machine learning for medical interviews with a focus on uncertainty estimation and speech processing. 
+\noindent The research presented in this thesis deals with the use of machine learning for medical interviews with a focus on uncertainty estimation and speech representation learning. 
 The research project was partially funded by a grant from the national Danish Innovation Fund (grant no.\@ 0153-00167B) and defined and carried out in collaboration with the Danish company Corti, which develops computer software for medical decision support.

### 2023-11-24 frontmatter/abstract-da.tex
Revised abstract and changes review paper to brief review
Update Danish abstract

--- 2022-02-27 frontmatter/abstract-da.tex
@@
-%!TEX root = ../Thesis.tex
-\chapter{Resumé (Danish)}
-\blindtext[5]
+%!TEX root = ../thesis.tex
+% LTeX: language=da-DK
+
+\chapter[resumé (abstract in danish)]{Resumé (Abstract in Danish)}
+
+Naturligt sprog spiller en nøglerolle i sundhedssystemer verden over.
+Alligevel har den medicinske interviewproces kun oplevet en lille udvikling sammenlignet med de fremskridt, der er gjort inden for medicinsk teknologi.
+Traditionelt udføres medicinske interviews af sundhedspersonale, der primært afhænger af deres individuelle erfaring for at forstå en situation.
+Med aldrende befolkninger, strenge dokumentationskrav, og fremskridt inden for diagnostiske muligheder og behandlingsmuligheder, er denne tilgang dyr og risikerer at komme til kort, hvilket potentielt kompromitterer nøjagtigheden og kvaliteten af medicinsk behandling.
+
+Nylige fremskridt inden for naturlig sprogbehandling gør det muligt for maskinlæringsmodeller at deltage aktivt i medicinske interviews for at lette administrative byrder, forbedre dokumentationen, og  yde realtidshjælp i dialog.
+Sundhedsplejen adskiller sig dog fra andre domæner på grund af den høje risiko forbundet med selv små fejl.
+Da ingen model er fejlfri, tilskynder dette til at associere modelprædiktioner med robuste usikkerhedsestimater, især i scenarier, der involverer ude-af-fordeling (OAF) data, såsom sjældne sygdomme.
+
+Denne afhandling tager udgangspunkt i hovedhypotesen, at usuperviseret repræsentationslæring er nyttig til usikkerhedsestimering for medicinske opgaver.
+Det giver følgende bidrag:
+%
+\begin{enumerate}[topsep=3pt, partopsep=0pt, itemsep=3pt, parsep=0pt, leftmargin=2em, label=(\alph*)] %, label=(\roman*)]
+    \item En likelihood-ratio score til OAF-detektion med variationelle autoenkodere, der afhjælper den bevist negative effekt af lav-niveau features.
+    \item En statistisk test til OAF-detektion, der kombinerer score- og typikalitetstests og er anvendelig med likelihoods fra enhver differentiabel generativ model.
+    \item Et benchmark for probabilistiske talerepræsentationsmodeller og en ny metode til at lære hierarkiske repræsentationer.
+    \item En oversigt over usuperviseret repræsentationslæring til neural talebehandling og en tilsvarende modeltaksonomi.
+    \item En fejlanalyse og revideret evaluering af state-of-the-art modeller til automatiseret medicinsk kodning på MIMIC-III og IV datasættene.
+    \item Et retrospektivt studie af talebaseret genkendelse af slagtilfælde i præhospitale akuttelefonopkald der viser betydelige forbedringer i forhold til resultatet opnået af opkaldstagere.\end{enumerate}
+%
+Sammenfattende adresserer denne afhandling udfordringer indenfor usikkerhedsestimering og repræsentationslæring til tale, mens den udforsker medicinske anvendelser af maskinlæring. 
+Dets bidrag er afgørende skridt i udviklingen af et operationelt beslutningsstøttesystem til medicinske samtaler, der i sidste ende sigter efter at forbedre kvaliteten af patientbehandlingen ved at understøtte effektiv, informeret beslutningstagning.

### 2023-11-24 frontmatter/abstract-da.tex
Update abstracts

@@
 Naturligt sprog spiller en nøglerolle i sundhedssystemer verden over.
-Alligevel har den medicinske interviewproces kun oplevet en lille udvikling sammenlignet med de fremskridt, der er gjort inden for medicinsk teknologi.
+Alligevel har den medicinske interviewproces kun oplevet en lille udvikling sammenlignet fremskridtene inden for medicinsk teknologi.
 Traditionelt udføres medicinske interviews af sundhedspersonale, der primært afhænger af deres individuelle erfaring for at forstå en situation.
@@
 
-Nylige fremskridt inden for naturlig sprogbehandling gør det muligt for maskinlæringsmodeller at deltage aktivt i medicinske interviews for at lette administrative byrder, forbedre dokumentationen, og  yde realtidshjælp i dialog.
+Nylige fremskridt inden for naturlig sprogbehandling gør det muligt for maskinlæringsmodeller at deltage aktivt i medicinske interviews for at lette administrative byrder, forbedre dokumentation, og assistere i realtid.
 Sundhedsplejen adskiller sig dog fra andre domæner på grund af den høje risiko forbundet med selv små fejl.
@@
 
-Denne afhandling tager udgangspunkt i hovedhypotesen, at usuperviseret repræsentationslæring er nyttig til usikkerhedsestimering for medicinske opgaver.
-Det giver følgende bidrag:
+Denne afhandling tager udgangspunkt i hovedhypotesen, at usuperviseret repræsentationslæring er nyttig til usikkerhedsestimering i medicinske opgaver.
+Den giver følgende bidrag:
 %
@@
     \item En likelihood-ratio score til OAF-detektion med variationelle autoenkodere, der afhjælper den bevist negative effekt af lav-niveau features.
-    \item En statistisk test til OAF-detektion, der kombinerer score- og typikalitetstests og er anvendelig med likelihoods fra enhver differentiabel generativ model.
+    \item En statistisk test til OAF-detektion, der kombinerer score- og typikalitetstests og kan bruges med likelihoods fra enhver differentiabel generativ model.
     \item Et benchmark for probabilistiske talerepræsentationsmodeller og en ny metode til at lære hierarkiske repræsentationer.
@@
     \item En fejlanalyse og revideret evaluering af state-of-the-art modeller til automatiseret medicinsk kodning på MIMIC-III og IV datasættene.
-    \item Et retrospektivt studie af talebaseret genkendelse af slagtilfælde i præhospitale akuttelefonopkald der viser betydelige forbedringer i forhold til resultatet opnået af opkaldstagere.\end{enumerate}
+    \item Et retrospektivt studie af talebaseret genkendelse af stroke i præhospitale akuttelefonopkald der viser betydelig forbedring i forhold til opkaldstagere.
+\end{enumerate}
 %
-Sammenfattende adresserer denne afhandling udfordringer indenfor usikkerhedsestimering og repræsentationslæring til tale, mens den udforsker medicinske anvendelser af maskinlæring. 
-Dets bidrag er afgørende skridt i udviklingen af et operationelt beslutningsstøttesystem til medicinske samtaler, der i sidste ende sigter efter at forbedre kvaliteten af patientbehandlingen ved at understøtte effektiv, informeret beslutningstagning.
+Sammenfattende adresserer denne afhandling udfordringer indenfor usikkerhedsestimering og repræsentationslæring til tale og udforsker medicinske anvendelser af maskinlæring. 
+Dens bidrag er afgørende i udviklingen af et operationelt beslutningsstøttesystem til medicinske interviews, der søger at øge kvaliteten af patientbehandlingen ved at understøtte effektiv, informeret beslutningstagen.

### 2023-11-24 chapters/conclusion.tex
Change references and add todos
Updates

@@
 \Cref{chp:technical-background} provided technical background only covered briefly by the individual studies.
-We first introduced uncertainty as a concept in the context of information and probability theory \cref{sec:uncertainty-information-theory}.
-Then, we defined the task of out-of-distribution detection and reviewed existing work on the problem \cref{sec:out-of-distribution-detection}.
-Finally, we provided technical background for variational autoencoders \cref{sec:variational-autoencoders}.
+We first introduced uncertainty as a concept in the context of information and probability theory (\cref{sec:uncertainty-information-theory}).
+Then, we defined the task of out-of-distribution detection and reviewed existing work on the problem (\cref{sec:out-of-distribution-detection}).
+Finally, we provided technical background for variational autoencoders (\cref{sec:variational-autoencoders}).
 
@@
 
-\cref{chp:paper-review}
+In \cref{chp:paper-brief}
+
 % Chapter 6 presented an overview of unsupervised neural speech representation learning.

### 2023-11-24 chapters/discussion.tex
Updates

@@
 
-\todo[inline]{Unsupervised representation learning for medical applications.
+% \todo[inline]{Unsupervised representation learning for medical applications.
 We tried BERT for stroke but no success.
@@
 
-\moretodo[inline]{Add to the discussion somewhere:
-I would try to embed a perspective for contemplation.
+% \moretodo[inline]{Add to the discussion somewhere:
+% I would try to embed a perspective for contemplation.
 Instead of us always talking about what the technology needs to improve on.
@@
 The rapid progress of machine learning that started about a decade ago with the 2012 ImageNet competition and the work of \textcite{krizhevsky_imagenet_2012} all but slowed down during the course of this project.
-Following the papers by \textcite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019}, the field of out-of-distribution detection saw a surge of interest that has since grown yearly, while representation learning for speech has continued to advance with notable works such as CPC \parencite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.
+Following the papers by \textcite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019}, the field of out-of-distribution detection saw a surge of interest that has since grown yearly.
+Similarly, representation learning for speech has continued to advance with notable works such as CPC \parencite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.
 
-Having focused on VAEs for out-of-distribution detection in \cref{part:unsupervised-uncertainty-estimation} and on self"=supervised methods for representation learning in \cref{part:unsupervised-speech-representation-learning}, this discussion will bridge a gap between the two approaches and consider how VAEs might be enabled to learn representations more competitive with self"=supervised methods for downstream tasks.
-Furthermore, since the work in \cref{part:medical-applications} only had limited focus on uncertainty, we will also present and discuss the use of calibration techniques for the stroke recognition model drawing connections to how such systems are perceived and used in practice.
+\moretodo{Adjust the below to include any discussion on out-of-distribution detection.}
+Having examined VAEs and self"=supervised approaches for representation learning in \cref{part:unsupervised-uncertainty-estimation,part:unsupervised-speech-representation-learning}, in this discussion we will consider how VAEs might be enabled to learn representations more competitive with self"=supervised methods for downstream tasks.
+Furthermore, since the work presented in \cref{part:medical-applications} only had limited focus on uncertainty, we will also present and discuss the use of calibration techniques for the stroke recognition model, drawing connections to how such systems might be perceived and used in practice.
 
@@
 In this thesis we studied two different approaches to speech representation learning: VAEs in \cref{chp:paper-hierarchical,chp:paper-modelagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-brief}.
-As we saw, the probabilistic formulation of VAEs provides benefits for their application to uncertainty quantification, although competitive methods that use self"=supervised foundation models have also been proposed \textcite{xiao_we_2021,hendrycks_using_2019,bergman_classificationbased_2020}.
-Self"=supervised methods are generally superior when it comes to performance on downstream tasks.
-While this of course depends on the task and the amount and type of unlabeled and labeled data, self"=supervised methods for speech are commonly evaluated on downstream tasks that are harder and more complex than those used for VAEs.
-For instance, between \cref{tab: phoneme recognition (PER),tab:unsupervisedASR} self"=supervised methods can be seen to outperform VAEs for phoneme recognition while self"=supervised methods are also reported for word-level speech recognition.
-In the following, we will discuss potential directions of future research that might help shed light on why representations from VAEs underperform on downstream tasks and how they might be improved.
+As we saw in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}, the probabilistic formulation of VAEs provides benefits for their application to uncertainty quantification, although competitive methods based on self"=supervised foundation models have also been successful \parencite{xiao_we_2021,hendrycks_using_2019,bergman_classificationbased_2020}.
+Nonetheless, as discussed in \cref{chp:paper-brief}, self"=supervised methods are superior to VAEs when it comes to performance on most other downstream tasks, such as speech recognition and spoken language understanding.
+While this statement of course depends on the task and the amount and type of unlabeled and labeled data, self"=supervised methods for speech are generally evaluated on downstream tasks that are harder and more complex than those used to evaluate VAEs.
+This is evidenced by the leaderboards on benchmarks such as SUPERB, SLUE and ZeroSpeech \parencite{yang_superb_2021,shon_slue_2021,dunbar_zero_2021} \footnote{One example of comparable evaluation is that between \cref{tab: phoneme recognition (PER),tab:unsupervisedASR} where self"=supervised methods can be seen to outperform VAEs for phoneme recognition while self"=supervised methods are also reported for word-level speech recognition.}.
+In the following, we will discuss potential directions of future research and try to shed light on why representations from VAEs underperform on downstream tasks and how they might be improved.
 
@@
 
-However, VAEs are not trained to maximize the marginal likelihood directly, but a lower bound of it, the ELBO,
+VAEs, however, are not trained to directly maximize the marginal likelihood, but a lower bound of it, the ELBO,
 %
@@
 %
-Since the ELBO is not independent of the latent variables, its maximization will typically depend on the latent variables and the posterior $p(\zb|\xb)$.
-Hence, unlike the marginal likelihood, the ELBO is not guaranteed to be a useless criterion for representation learning.
-Nonetheless, being a lower bound of the marginal likelihood that we seek to maximize, seems to indicate that this issue might not be irrelevant to consider.
+Since the ELBO is not independent of the latent variables, its maximization will typically depend on the latent variables and the posterior $p(\zb|\xb)$ and hence, it is not necessarily a useless criterion for representation learning.
+Nonetheless, being a lower bound of the marginal likelihood suggests it could suffer from similar issues under ``right'' conditions.
 
@@
     I_{q_{\phib}(\xb,\zb)}\left[\xb ; \zb \right]
-    }} \. . %\nonumber
+    }} \. . \nonumber
 \end{equation}
@@
 % Optimizing the ELBO can be understood as simultaneously maximizing the average reconstruction likelihood, minimizing the marginal KL-divergence to the prior, and minimizing the mutual information between the observed and latent variables.
-We can see that optimizing the ELBO means learning a good one-to-one mapping of $z$ to $x$ (average reconstruction), fitting the aggregated posterior to the prior (marginal KL-divergence), while also not learning any stochastic dependency between $x$ and $z$ (mutual information).
+We can see that optimizing the ELBO means learning a good one-to-one mapping from $\zb$ to $\xb$ (average reconstruction), fitting the aggregated posterior to the prior (marginal KL-divergence), while also not learning any stochastic dependency between $\xb$ and $\zb$ (mutual information).
 These objectives are conflicting and especially the minimization of mutual information seems counterproductive to our goals \parencite{tomczak_trouble_2022}.
@@
 
-Even though these properties indicate that VAEs might not work at all for representation learning, VAEs have consistently been shown to be able to learn meaningful representations \parencite{sonderby_ladder_2016,hsu_unsupervised_2017,maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
+Even though these properties suggest that VAEs might not work well for representation learning, VAEs have consistently been shown to be able to learn meaningful representations \parencite{sonderby_ladder_2016,hsu_unsupervised_2017,maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
 We can understand the reason for this by noting that both of the above scenarios can only occur if the optimization is too unconstrained.
 
-In practice, the maximum marginal likelihood estimation takes place not over all possible models, but over a constrained parametric class defined by the inference network $q_\thetab(\zb|\xb) \approx p(\zb|\xb)$ and the generative model $p_\thetab(\xb|\zb)p(\zb) = p_\theta(\xb,\zb)$.
-This introduces a coupling between the marginal $p_{\thetab}(\xb)$ and the posterior $p_{\thetab}(\zb|\xb)$ which makes the maximum likelihood solution unlikely to correspond to completely useless representations \parencite{huszar_is_2017}.
-Similarly, maximizing the ELBO in \cref{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)} by exactly minimizing the mutual information term corresponds to a complete posterior collapse that requires a trade-off with the other terms.
-For a given model, this solution is almost surely a local minimum, but one that is often avoided in practice.
-The challenge is then how to impose appropriate constraints to make VAEs learn representations that are competitive with self"=supervised methods.
+Specifically, the maximum marginal likelihood estimation in practice takes place not over all possible models, but over a constrained parametric class defined by the inference network $q_\phib(\zb|\xb) \approx p(\zb|\xb)$ and the generative model $p_\thetab(\xb,\zb) = p_\thetab(\xb|\zb)p(\zb)$.
+This introduces a coupling between the marginal $p_{\thetab}(\xb)$ and the posterior $p(\zb|\xb)$ which makes the maximum likelihood solution unlikely to correspond to completely useless representations \parencite{huszar_is_2017}.
+Similarly, maximizing the ELBO %in \cref{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)} 
+by exactly minimizing the mutual information term corresponds to a complete posterior collapse that requires a trade-off with the other terms.
+For a given model, this solution is almost surely a local minimum, and one that is often avoided in practice.
+
+For VAEs to learn representations more competitive with self"=supervised methods, the fundamental challenge then seems to be how to impose constraints on the model class and optimization problem to ensure this; a strong coupling between the marginal likelihood and the usefulness of the posterior, and an optimization problem where minimization of the mutual information is very hard, or impossible.
 
@@
 Architectural improvements and hierarchies of latent variables are two related, and interesting, approaches to learning more semantic features in VAEs. 
-Some success has been demonstrated in the image domain with models like BIVA \parencite{maaloe_biva_2019}, NVAE \parencite{vahdat_nvae_2020}, and VD-VAE \parencite{child_very_2021}, with up to 78 stochastic layers, achieving tight likelihoods and high-quality samples, although limited effort has put towards using such models for downstream tasks.
+Some success has been demonstrated in the image domain where models like BIVA \parencite{maaloe_biva_2019}, NVAE \parencite{vahdat_nvae_2020}, and VD-VAE \parencite{child_very_2021} use up to 78 stochastic layers and have achieved tight likelihoods and high-quality samples.
+Nonetheless, only limited effort has been put towards evaluating the usefulness for downstream task of representations learned by these very deep models. 
 For speech, a hierarchical model operating on multiple temporal scales, such as the Clockwork VAE \parencite{saxena_clockwork_2021} adapted for speech in \cref{chp:paper-benchmarking}, might help better capture dependencies at longer ranges and encode more semantic features.
@@
 
-\todo[inline]{Out-of-distribution detection on speech?}
+\moretodo[inline]{Out-of-distribution detection on speech?}
 

### 2023-11-27 frontmatter/acknowledgements.tex
Rename TODO
Add hierarchical VAE section and update main contributions with research questions/hypotheses
Write acknowledgments

@@
 %!TEX root = ../thesis.tex
+
 \chapter[acknowledgements]{Acknowledgements}
-First of all, I want to thank all my supervisors...
+First, I want to thank my supervisors. Jes, you have been an amazing support through this project. 
+Your positivity and insightfulness were essential to me daring to take on this project, and to its success. 
+Lars, you have been an indispensable source of purpose, energy, and direction. I look forward to many more fruitful discussions in the future.
+
+I also want to thank the brilliant researchers at the section for Cognitive Systems that I have had the privilege of working with. 
+Søren, as my co-supervisor, you have been a fantastic discussion partner. Your wit and your ability to pose genuinely significant questions have been invaluable. 
+I want to also convey appreciation for the researchers in my group, for the open, inclusive, and relaxed culture, and for many good discussions, often over coffee and pastries.
+
+I want to thank Lars, Andreas, and Michael for letting me become part of the Corti project after graduating back in 2018. 
+The foundational will to invest and believe in people, and to dare to have bold visions, has made Corti a fantastic place for me to develop, professionally and personally. 
+I want to also thank the members of the machine learning team during the first years of my time at Corti.
+Lasse, Tycho, Marco, Jan, Janek, and Joakim, you all played a part in inspiring and enabling me to dive into this project.
+A special thanks should go to Joakim. Your honest curiosity and dedication has been a wonderful addition to the research team. 
+Finally, I want to thank Lasse.
+Our numerous inspirational discussions and combined efforts through the years have been instrumental in shaping the trajectory of this thesis.
+It has been a privilege to carry out this PhD project in the company of you both.
+
+I want to also thank my friends and my family who have provided much needed diversion throughout the years. 
+Last, but not least, I want to thank my wife, Rikke, who has been an unbelievably patient and absolutely indispensable supporter at every step of this undertaking.
 
 \vspace*{\fill}
-\noindent \textit{The project was funded by Corti and Innovation Fund Denmark through industrial PhD grant no.\@ 0153-00167B).}
+\noindent \textit{The project was funded by Corti and Innovation Fund Denmark through industrial PhD grant no.\@ 0153-00167B.}

### 2023-11-27 chapters/discussion.tex
Minor

@@
 
-\todo[inline]{Masked pre-training actually corresponds to marginal maximum likelihood optimization such as done by VAES \parencite{moreno-munoz_masked_2023}.
+\todo[inline]{Masked pre-training actually corresponds to marginal maximum likelihood optimization such as done by VAEs \parencite{moreno-munoz_masked_2023}.
 Since maximum likelihood in no way guarantees good representations, how do self"=supervised methods achieve  this, if they also do maximum likelihood?}

### 2023-11-27 chapters/introduction.tex
Make some minor updates

@@
 \noindent The research presented in this thesis deals with the use of machine learning for medical interviews with a focus on uncertainty estimation and speech representation learning. 
-The research project was partially funded by a grant from the national Danish Innovation Fund (grant no.\@ 0153-00167B) and defined and carried out in collaboration with the Danish company Corti, which develops computer software for medical decision support.
+The research project was partially funded by a grant from the national Danish Innovation Fund (grant no.\@ 0153-00167B) and defined and carried out in collaboration with the Danish health tech company Corti, which develops computer software for medical decision support.
 In the rest of this chapter, we will provide a high-level introduction to these topics and to Corti and discuss the motivation behind the research project.

### 2023-11-27 chapters/conclusion.tex
Minor
Add to conclusion

@@
 We emphasized that machine learning comes with these same types of risk, in some cases amplifying their impacts, and argued that uncertainty estimation is a key component in ensuring that systems are safe and reliable in practice.
-We also provided the background for the research project by introducing the healthtech company Corti, with which with project has been defined and carried out, and the motivational cases of recognition of stroke cases in emergency calls (\cref{subsec:motivation-stroke-recognition}) and automated medical coding (\cref{subsec:motivation-medical-coding}).
-We described how machine learning systems could be designed to aid healthcare professionals with these tasks and gave examples of potential sources of uncertainty and of how correctly quantifying it might improve the usefulness of such systems.
+We also provided the background for the research project by introducing the health tech company Corti, with which with project has been defined and carried out, and the motivational cases of recognition of stroke cases in emergency calls (\cref{subsec:motivation-stroke-recognition}) and automated medical coding (\cref{subsec:motivation-medical-coding}).
+Furthermore, we described how machine learning systems could be designed to aid healthcare professionals with these tasks and gave examples of potential sources of uncertainty and of how correctly quantifying it might improve the usefulness of such systems.
 Finally, we drew connections between machine learning reliability (\cref{sec:machine-learning-reliability}) and model calibration (\cref{subsec:model-calibration}) and defined the aleatoric, epistemic, and predictive types of uncertainty (\cref{subsec:understanding-uncertainty}).
@@
 
-In \cref{chp:paper-hierarchical}
+In \cref{chp:paper-hierarchical} we showed how hierarchical variational autoencoders can fail at likelihood-based OOD detection due to an overemphasis on low-level features that generalize between different data distributions.
+By exploiting that VAEs tend to learn more abstract features at latent variables high in the hierarchy, we were able to define a likelihood-ratio score for OOD detection that focused more on features unlikely to be shared between datasets.
+% - VAEs can learn useful representations, but it can be hard to know how to do that.
+% - 
+This work also highlights that overlap between distributions of high-dimensional data can severely impede OOD detection.
+Our work shows that a promising direction of future research is into how select which dimensions are the most discriminatory.
+
+% In this work we hypothesize that the likelihood estimate of variational autoencoders is a poor score for out-of-distribution due to an overemphasis on low-level features that generalize between distributions.
+% We further hypothesize that a well-formed hierarchy of latent variables provides a tool that can be used to select which features to emphasize for out-of-distribution detection and, hence, a way to improve the performance of variational autoencoders on this task.
+% We proceed to provide empirical and theoretical evidence that low-level features do indeed dominate the likelihood score and propose a new method for out-of-distribution detection using hierarchical variational autoencoders based on a likelihood-ratio score that requires data to be in-distribution across all feature-levels.
+% The proposed method is computationally efficient, fully unsupervised, and performs well on several out-of-distribution detection benchmarks.
 
@@
 
-In \cref{chp:paper-brief}
+In \cref{chp:paper-brief} we provided an overview of unsupervised neural speech representation learning.
+Such approaches have recently matched supervised methods on many tasks and represent a significant advance in low-resource settings, such as speech recognition for minority languages.
+
+% In this chapter, we present a comprehensive overview of unsupervised neural representation learning for speech.
+Previous research is categorized into self"-supervised methods and probabilistic latent variable models and described in a common notation.
+This description assists in developing a model taxonomy that shapes a discussion of the models' representational power, the associated learning strategies, and the methods used to evaluate them.
+The discussion points to interesting avenues of future research.
 
@@
 
-\cref{chp:paper-automated}
+In \cref{chp:paper-automated}
 
-\cref{chp:paper-retrospective}
+In \cref{chp:paper-retrospective} we studied how machine learning might be used to improve decision-making at emergency services. 
+%
+As discussed in \cref{sec: discussion-stroke-recognition-uncertainty}, predictive uncertainty is likely to be essential to user adoption

### 2023-11-28 chapters/conclusion.tex
Work on conclusion

@@
 
-\Cref{chp:technical-background} provided technical background only covered briefly by the individual studies.
+\vspace{3em}
+\textbf{\Cref{chp:technical-background}} provided technical background only covered briefly by the individual studies.
 We first introduced uncertainty as a concept in the context of information and probability theory (\cref{sec:uncertainty-information-theory}).
@@
 
-In \cref{chp:paper-hierarchical} we showed how hierarchical variational autoencoders can fail at likelihood-based OOD detection due to an overemphasis on low-level features that generalize between different data distributions.
+\textbf{In \cref{chp:paper-hierarchical}} we showed how hierarchical variational autoencoders can fail at likelihood-based OOD detection due to an overemphasis on low-level features that generalize between different data distributions.
 By exploiting that VAEs tend to learn more abstract features at latent variables high in the hierarchy, we were able to define a likelihood-ratio score for OOD detection that focused more on features unlikely to be shared between datasets.
-% - VAEs can learn useful representations, but it can be hard to know how to do that.
-% - 
+% - VAEs can learn useful representations, but it can be hard to impose the appropriate constraints that enable this.
+% - VAEs can learn representations that are useful for OOD detection, but good performance might require selecting an appropriate subset of feature dimensions to use for discriminating.
 This work also highlights that overlap between distributions of high-dimensional data can severely impede OOD detection.
@@
 
-In \cref{chp:paper-modelagnostic}
+\vspace{3em}
+\textbf{In \cref{chp:paper-modelagnostic}}
 
-In \cref{chp:paper-benchmarking} 
+\vspace{3em}
+\textbf{In \cref{chp:paper-benchmarking}} 
 % In chapter 8, stochastic and deterministic generative models were benchmarked in terms of model likelihood.
@@
 
-In \cref{chp:paper-brief} we provided an overview of unsupervised neural speech representation learning.
+\vspace{3em}
+\textbf{In \cref{chp:paper-brief}} we provided an overview of unsupervised neural speech representation learning.
 Such approaches have recently matched supervised methods on many tasks and represent a significant advance in low-resource settings, such as speech recognition for minority languages.
+% Self-supervised learning seems to have better inductive biases, or at least a more forgiving learning problem, than VAEs for the purpose of learning good representations in an unsupervised manner.
+% As we discussed in \cref{chp:discussion}, there are many interesting avenues of future research for improving the ability of VAEs to learn useful representations, including better gradient estimates and masked objectives.
 
@@
 
-In \cref{chp:paper-automated}
+\vspace{3em}
+\textbf{In \cref{chp:paper-automated}}
 
-In \cref{chp:paper-retrospective} we studied how machine learning might be used to improve decision-making at emergency services. 
-%
-As discussed in \cref{sec: discussion-stroke-recognition-uncertainty}, predictive uncertainty is likely to be essential to user adoption
+\vspace{3em}
+\textbf{In \cref{chp:paper-retrospective}} we studied how machine learning might be used to improve decision-making at emergency services. 
+We saw that a model was able to improve significantly on the stroke recognition ability of call-takers alone and that the features it used were sensible and related to symptoms and descriptions of stroke.
+
+In \cref{sec: discussion-stroke-recognition-uncertainty}, we discussed how calibrating the predictive uncertainty of the stroke model is likely to be necessary to enable sustained use of such a model in practice.
+In \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability}, we noted that, as expected, model performance measured by F1-score increased with increasing model certainty which underlines the likely usefulness of uncertainty estimates in helping to match practitioner expectations of such a model.
+Nonetheless, basic metrics of model performance are still obstacles for its practical usefulness.
+Specifically, the rarity of stroke cases lead to a false positive rate and precision that are likely to induce alarm fatigue among its users.
+Similar effects are likely to have influenced the practical impact of a similar system for cardiac arrest detection which showed significant improvements retrospectively \parencite{cite14} which, although matched by the model in a prospective study, did not ultimately result in improved call-taker performance \parencite{cite15}.
+
+Nevertheless, the strong retrospective performance of the stroke recognition model indicates that there is significant potential for augmenting the medical interview to allow better recognizing stroke cases.
+Possible improvements to the system could include utilization of the audio signal to detect speech-related symptoms such mumbling or slurring, or integrating with electronic health-records to cross-reference with patient history.
+Even so, directly predicting the diagnosis from the conversation is not the only path towards practical impact. Simply suggesting which questions to ask
+
+% Explore the utilization of audio data 
+% in conjunction with EHR information to gain a comprehensive understanding of the patient's condition.
+
+% Suggesting Questions:
+
+% Integrate the ML system into the interview framework to dynamically suggest relevant questions based on the ongoing conversation.
+This approach leverages the model's retrospective success, guiding clinicians to inquire about specific symptoms or risk factors that may contribute to more accurate stroke identification.
+
+% Fact/Sanity-Checking with Electronic Health Record (EHR):
+
+% Establish a seamless connection between the ML model and the patient's EHR to perform real-time fact-checking during the interview.
+This ensures the accuracy of the information provided by the patient, enhancing the reliability of the diagnostic process.
+
+% Improved Understanding of the Patient Using Audio and EHR:
+
+% Explore the utilization of audio data in conjunction with EHR information to gain a comprehensive understanding of the patient's condition.
+By analyzing speech patterns, tone, and content, the ML model can contribute valuable insights to the diagnostic process.
+Additionally, cross-referencing this audio data with EHR details enhances the model's ability to discern subtle indicators of stroke risk or occurrence.
+
+% - Suggesting questions
+% - Fact/sanity-checking with electronic health-record
+% - For improved understanding of the patient: Using audio directly, using electronic health-record,
+
+% easily see its retrospective performance transferred to practical impact, but its
+
+% Directly predicting the diagnosis from the conversation may not be the best path
+
+% There are many promising paths forward.

### 2023-11-28 chapters/background.tex
Minor stuff

@@
 
-\subsection{Supervised out-of-distribution detection} \label{subsec: supervised-out-of-distribution-detection}
+\subsection{Supervised out-of-distribution detection}\label{subsec: supervised-out-of-distribution-detection}
 % OUTLINE:
@@
 
-\subsection{Unsupervised out-of-distribution detection} \label{subsec: unsupervised-out-of-distribution-detection}
+\subsection{Unsupervised out-of-distribution detection}\label{subsec: unsupervised-out-of-distribution-detection}
 % OUTLINE:

### 2023-11-28 chapters/conclusion.tex
Work on conclusion

@@
 
-In the introductory \cref{chp:introduction} we laid out how progress-driving technological development has always come with new challenges and risks of error - whether via misuse, misunderstanding or inherent limitations.
+\textbf{In \cref{chp:introduction,chp:technical-background}} we introduced automated medical coding and stroke recognition as motivational cases. 
+We used these cases to exemplify the importance of out-of-distribution detection, and, by extension, representation learning.
+In the context of these cases, we discussed possible machine learning system designs for decision-support and considered potential sources of uncertainty and ideal model behavior.
+As such, the cases have formed a reference point for the thesis as a whole with contributions within out-of-distribution detection as well as representation learning connecting to these applications. 
+As \cref{chp:main-contributions} already details the contributions made by this thesis, we will not further discuss these in this conclusion.
+
+% \textbf{\cref{chp:introduction}} laid out how progress-driving technological development has always come with new challenges and risks of error - whether via misuse, misunderstanding or inherent limitations.
 We emphasized that machine learning comes with these same types of risk, in some cases amplifying their impacts, and argued that uncertainty estimation is a key component in ensuring that systems are safe and reliable in practice.
-We also provided the background for the research project by introducing the health tech company Corti, with which with project has been defined and carried out, and the motivational cases of recognition of stroke cases in emergency calls (\cref{subsec:motivation-stroke-recognition}) and automated medical coding (\cref{subsec:motivation-medical-coding}).
-Furthermore, we described how machine learning systems could be designed to aid healthcare professionals with these tasks and gave examples of potential sources of uncertainty and of how correctly quantifying it might improve the usefulness of such systems.
-Finally, we drew connections between machine learning reliability (\cref{sec:machine-learning-reliability}) and model calibration (\cref{subsec:model-calibration}) and defined the aleatoric, epistemic, and predictive types of uncertainty (\cref{subsec:understanding-uncertainty}).
 
-\vspace{3em}
+% The chapter also provided some background for the research project by introducing the health tech company Corti, with which with project has been defined and carried out, and the motivational cases of recognition of stroke cases in emergency calls (\cref{subsec:motivation-stroke-recognition}) and automated medical coding (\cref{subsec:motivation-medical-coding}).
+Using these cases, we exemplified how uncertainty might arise in practice and how quantifying it might improve the usefulness of machine learning systems built for the tasks.
+% Finally, we drew connections between machine learning reliability (\cref{sec:machine-learning-reliability}) and model calibration (\cref{subsec:model-calibration}) and defined the aleatoric, epistemic, and predictive types of uncertainty (\cref{subsec:understanding-uncertainty}).
+
+% Along with the technical background provided in
+% This introduction formed the basis of
+
+\vspace{1em}
 \textbf{\Cref{chp:technical-background}} provided technical background only covered briefly by the individual studies.
@@
 
+\vspace{1em}
 \textbf{In \cref{chp:paper-hierarchical}} we showed how hierarchical variational autoencoders can fail at likelihood-based OOD detection due to an overemphasis on low-level features that generalize between different data distributions.
@@
 
-\vspace{3em}
+\vspace{1em}
 \textbf{In \cref{chp:paper-modelagnostic}}
 
-\vspace{3em}
+\vspace{1em}
 \textbf{In \cref{chp:paper-benchmarking}} 
@@
 
-\vspace{3em}
+\vspace{1em}
 \textbf{In \cref{chp:paper-brief}} we provided an overview of unsupervised neural speech representation learning.
 Such approaches have recently matched supervised methods on many tasks and represent a significant advance in low-resource settings, such as speech recognition for minority languages.
-% Self-supervised learning seems to have better inductive biases, or at least a more forgiving learning problem, than VAEs for the purpose of learning good representations in an unsupervised manner.
+We found that for the purpose of learning good representations in an unsupervised manner, self-supervised learning seems to have better inductive biases, or at least pose a more forgiving learning problem, than do VAEs.
+As discussed in the paper and in \cref{sec_discussion:representation-learning-with-vaes}, this likely relates more to inductive biases imposed by implicit constraints in the optimization problem and architecture than to the underlying formalism.
+For instance, in discussion about the weaknesses of VAE-based approaches (\cref{sec_discussion:representation-learning-with-vaes}) we concluded that their challenges could not be directly attributed to the maximum marginal likelihood objective.
+Indeed, the masked pre-training objective widely used for successful self"=supervised methods also corresponds to a maximum marginal likelihood objective \parencite{moreno-munoz_masked_2023}.
+
 % As we discussed in \cref{chp:discussion}, there are many interesting avenues of future research for improving the ability of VAEs to learn useful representations, including better gradient estimates and masked objectives.
@@
 
-\vspace{3em}
+\vspace{1em}
 \textbf{In \cref{chp:paper-automated}}
 
-\vspace{3em}
+Future directions
+- Avoiding predicting several mutually exclusive classes, which is a general problem for multi-label
+
+\vspace{1em}
 \textbf{In \cref{chp:paper-retrospective}} we studied how machine learning might be used to improve decision-making at emergency services. 
@@
 Nevertheless, the strong retrospective performance of the stroke recognition model indicates that there is significant potential for augmenting the medical interview to allow better recognizing stroke cases.
-Possible improvements to the system could include utilization of the audio signal to detect speech-related symptoms such mumbling or slurring, or integrating with electronic health-records to cross-reference with patient history.
-Even so, directly predicting the diagnosis from the conversation is not the only path towards practical impact. Simply suggesting which questions to ask
+Possible improvements to the system could include using the audio signal to detect speech-related symptoms such mumbling or slurring, or integrating with electronic health-records to cross-reference with patient history.
+Even so, directly predicting the diagnosis from the conversation is not the only path towards practical impact.
+By suggesting informative questions to the medical professional, a system could help guide the conversation to avoid missing important details, and to improve the precision of the model.
 

### 2023-11-28 chapters/discussion.tex
Make some minor updates
Work on conclusion

@@
 
-\section{Representation learning with variational autoencoders}
+\section{Representation learning with variational autoencoders}\label{sec_discussion:representation-learning-with-vaes}
 %
@@
 Similarly, maximizing the ELBO %in \cref{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)} 
-by exactly minimizing the mutual information term corresponds to a complete posterior collapse that requires a trade-off with the other terms.
-For a given model, this solution is almost surely a local minimum, and one that is often avoided in practice.
+by exactly minimizing the mutual information term corresponds to a complete posterior collapse which necessitates a trade-off with the other terms.
+For a given model, this solution is almost surely a local minimum, and one that can often be avoided in practice.
+% Further evidence that the trouble is not solely attributable, the masked pre-training objective commonly used for self"=supervised methods has been shown to correspond to a maximum likelihood objective \parencite{moreno-munoz_masked_2023}.
 
-For VAEs to learn representations more competitive with self"=supervised methods, the fundamental challenge then seems to be how to impose constraints on the model class and optimization problem to ensure this; a strong coupling between the marginal likelihood and the usefulness of the posterior, and an optimization problem where minimization of the mutual information is very hard, or impossible.
+For VAEs to learn representations more competitive with self"=supervised methods, the fundamental challenge then seems to be how to impose constraints on the model class and optimization problem to ensure a strong coupling between the marginal likelihood and the usefulness of the posterior, and an optimization problem where minimization of the mutual information is very hard, or impossible.
 
@@
 
-\todo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimizing a mutual information term in the ELBO (derive this form).}
-\todo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other).
-Include reference to \parencite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
-
-\todo[inline]{Masked pre-training actually corresponds to marginal maximum likelihood optimization such as done by VAEs \parencite{moreno-munoz_masked_2023}.
+% \todo[inline]{Masked pre-training actually corresponds to marginal maximum likelihood optimization such as done by VAEs \parencite{moreno-munoz_masked_2023}.
 Since maximum likelihood in no way guarantees good representations, how do self"=supervised methods achieve  this, if they also do maximum likelihood?}
@@
 % \section{\Cref{chp:paper-retrospective} revisited: Calibration}
-\section{Uncertainty of the stroke recognition classifier}
+\section{Uncertainty of the stroke recognition classifier} \label{sec: discussion-stroke-recognition-uncertainty}
 
@@
         \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curve_ensemble_and_all_models_uncalibrated.pdf}
-        % \caption{}
-        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
     \end{subfigure}
@@
         \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/histogram_ensemble_and_single_model.pdf}
-        % \caption{}
-        % \label{fig_discussion:histogram_ensemble_and_single_model}
     \end{subfigure}
@@
     \label{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability}
-    % \begin{subfigure}[c]{0.49\columnwidth}
-    %     \centering
-    %     \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/precision_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
-    %     % \caption{}
-    %     % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
-    % \end{subfigure}    
-    % % \hfill
-    % \begin{subfigure}[c]{0.49\columnwidth}
-    %     \centering
-    %     \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/recall_vs_predicted_probability_ensemble_calltaker_with_cis.pdf}
-    %     % \caption{}
-    %     % \label{fig_discussion:histogram_ensemble_and_single_model}
-    % \end{subfigure}
 \end{figure}
@@
 
-% If the stroke recognition ensemble were to be deployed in a randomized controlled trial, s
 Calibrated probabilities in the alerts presented to users might be a way to alleviate alarm fatigue.
@@
 This indicates that selecting, based on certainty, which predictions to present to users might indeed help build trust in the system and ensure a practical impact.
-
-% \lesstodo[inline]{Maybe mention MultiQT paper \parencite{havtorn_multiqt_2020}.}

### 2023-11-29 chapters/conclusion.tex
Minor edits

@@
 
-\textbf{In \cref{chp:introduction,chp:technical-background}} we introduced automated medical coding and stroke recognition as motivational cases. 
+\textbf{In \cref{chp:introduction}} we introduced automated medical coding and stroke recognition as motivational cases. 
 We used these cases to exemplify the importance of out-of-distribution detection, and, by extension, representation learning.
@@
 As such, the cases have formed a reference point for the thesis as a whole with contributions within out-of-distribution detection as well as representation learning connecting to these applications. 
-As \cref{chp:main-contributions} already details the contributions made by this thesis, we will not further discuss these in this conclusion.
+As \textbf{\cref{chp:main-contributions}} already details the contributions made by this thesis, we will not reiterate these in detail in this conclusion.
 
@@
 This work also highlights that overlap between distributions of high-dimensional data can severely impede OOD detection.
-Our work shows that a promising direction of future research is into how select which dimensions are the most discriminatory.
+Our work shows that a promising direction of future research is into how to select which dimensions are the most discriminatory.
 
@@
 % The proposed method is computationally efficient, fully unsupervised, and performs well on several out-of-distribution detection benchmarks.
-
-\vspace{1em}
-\textbf{In \cref{chp:paper-modelagnostic}}
 
@@
 \vspace{1em}
+\textbf{In \cref{chp:paper-modelagnostic}}
+
+\vspace{1em}
 \textbf{In \cref{chp:paper-brief}} we provided an overview of unsupervised neural speech representation learning.
@@
 We found that for the purpose of learning good representations in an unsupervised manner, self-supervised learning seems to have better inductive biases, or at least pose a more forgiving learning problem, than do VAEs.
-As discussed in the paper and in \cref{sec_discussion:representation-learning-with-vaes}, this likely relates more to inductive biases imposed by implicit constraints in the optimization problem and architecture than to the underlying formalism.
+As discussed in \cref{chp:paper-brief,sec_discussion:representation-learning-with-vaes}, this likely relates more to inductive biases imposed by implicit constraints in the optimization problem and architecture than to the underlying formalism.
 For instance, in discussion about the weaknesses of VAE-based approaches (\cref{sec_discussion:representation-learning-with-vaes}) we concluded that their challenges could not be directly attributed to the maximum marginal likelihood objective.
 Indeed, the masked pre-training objective widely used for successful self"=supervised methods also corresponds to a maximum marginal likelihood objective \parencite{moreno-munoz_masked_2023}.
+This also leaves hope ... \moretodo{hope to improve VAEs by certain measures.}
 
@@
 In \cref{sec: discussion-stroke-recognition-uncertainty}, we discussed how calibrating the predictive uncertainty of the stroke model is likely to be necessary to enable sustained use of such a model in practice.
-In \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability}, we noted that, as expected, model performance measured by F1-score increased with increasing model certainty which underlines the likely usefulness of uncertainty estimates in helping to match practitioner expectations of such a model.
-Nonetheless, basic metrics of model performance are still obstacles for its practical usefulness.
+In \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability}, we noted that, as expected, model performance measured by F1-score increased with increasing model certainty which underlines the likely usefulness of uncertainty estimates in matching practitioner expectations of the predictive performance.
+Nonetheless, basic metrics of model performance might still be obstacles for its practical usefulness.
 Specifically, the rarity of stroke cases lead to a false positive rate and precision that are likely to induce alarm fatigue among its users.

### 2023-11-30 chapters/conclusion.tex
Work on conclusion

@@
 
-\textbf{In \cref{chp:introduction}} we introduced automated medical coding and stroke recognition as motivational cases. 
-We used these cases to exemplify the importance of out-of-distribution detection, and, by extension, representation learning.
+\textbf{\Cref{chp:introduction}} introduced the motivational cases of automated medical coding and stroke recognition and used them to exemplify the importance of out-of-distribution detection, and, by extension, representation learning.
 In the context of these cases, we discussed possible machine learning system designs for decision-support and considered potential sources of uncertainty and ideal model behavior.
-As such, the cases have formed a reference point for the thesis as a whole with contributions within out-of-distribution detection as well as representation learning connecting to these applications. 
-As \textbf{\cref{chp:main-contributions}} already details the contributions made by this thesis, we will not reiterate these in detail in this conclusion.
+The cases form a reference point for the thesis as a whole, connecting its contributions within out-of-distribution detection and representation learning back to practical applications. 
+In this conclusion, we will review the studies presented in previous chapters in the context of the discussion of \cref{chp:discussion}, the recent progress in the field, and point to interesting directions of future research.
+As \textbf{\cref{chp:main-contributions}} already details the contributions made by this thesis, we will not reiterate them in detail in this conclusion.
 
@@
 \vspace{1em}
-\textbf{\Cref{chp:technical-background}} provided technical background only covered briefly by the individual studies.
+\textbf{\Cref{chp:technical-background}} provided technical background that could only be covered briefly by the individual studies. 
 We first introduced uncertainty as a concept in the context of information and probability theory (\cref{sec:uncertainty-information-theory}).
-Then, we defined the task of out-of-distribution detection and reviewed existing work on the problem (\cref{sec:out-of-distribution-detection}).
+We then defined the task of out-of-distribution detection and reviewed existing work on the problem (\cref{sec:out-of-distribution-detection}).
 Finally, we provided technical background for variational autoencoders (\cref{sec:variational-autoencoders}).
@@
 \vspace{1em}
-\textbf{In \cref{chp:paper-hierarchical}} we showed how hierarchical variational autoencoders can fail at likelihood-based OOD detection due to an overemphasis on low-level features that generalize between different data distributions.
-By exploiting that VAEs tend to learn more abstract features at latent variables high in the hierarchy, we were able to define a likelihood-ratio score for OOD detection that focused more on features unlikely to be shared between datasets.
+\textbf{\Cref{chp:paper-hierarchical}} showed how hierarchical variational autoencoders can fail at likelihood"-based OOD detection due to an overemphasis on low-level features that generalize between different data distributions.
+In other words, the distributions of latent representations of high-dimensional data overlap, especially for latent variables low in the hierarchy.
+By exploiting that VAEs tend to learn more abstract features at latent variables high in the hierarchy, we were able to define a likelihood-ratio score that focused more on features unlikely to be shared between datasets and performed much better for OOD detection.
+In line with our discussion in \cref{sec_discussion:representation-learning-with-vaes}, these findings show that VAEs can indeed learn useful latent representations, although good performance, at least for OOD detection, might require selecting an appropriate subset of latent representations to use.
+Similar variations among features learned at different layers have also been identified for self"=supervised models and speech representations \parencite{pasad_layerwise_2021}. 
+To obtain good downstream task performance, this suggests that, besides seeking to improve representations overall, future research effort should also be directed towards finding good methods for selecting relevant subsets of latent variables in a given hierarchical VAE for a certain task.
 % - VAEs can learn useful representations, but it can be hard to impose the appropriate constraints that enable this.
 % - VAEs can learn representations that are useful for OOD detection, but good performance might require selecting an appropriate subset of feature dimensions to use for discriminating.
-This work also highlights that overlap between distributions of high-dimensional data can severely impede OOD detection.
-Our work shows that a promising direction of future research is into how to select which dimensions are the most discriminatory.
+% Our work shows that a promising direction of future research is into how to select which dimensions are the most discriminatory.
 
@@
 \vspace{1em}
-\textbf{In \cref{chp:paper-benchmarking}} 
-% In chapter 8, stochastic and deterministic generative models were benchmarked in terms of model likelihood.
-In addition, the chapter presented a hierarchical latent variable model (LVM) for speech inspired by the Clockwork VAE [238].
-In contrast to self-supervised models, LVMs have not yet seen the same success for representation learning.
-As such, they offer little guidance in terms of alleviating the need for labeled data, as discussed in section 1.2.1.
+\textbf{\Cref{chp:paper-modelagnostic}} took a different approach to OOD detection than \cref{chp:paper-hierarchical} and focused on developing a model-agnostic method. 
+We showed that by phrasing OOD detection as a statistical testing problem and combining different tests, orthogonal properties of the individual tests could be leveraged to improve the OOD detection performance over any single test.
 
-% However, since LVMs learn a distribution over the training data, they are forced to encode all aspects of it.
-While this may be redundant for some tasks, others benefit from a broad variety of features.
-A hierarchical model that operates on multiple temporal scales, such as the Clockwork VAE, offers a natural way to encode different feature categories.
-Pronunciation might be learned at lower layers, speaker identity at the upper layers, and semantic features in between.
-Although some work has successfully separated speaker identity from content [114], models that can learn a deep hierarchy of features for speech remains an open challenge.
-The model presented in chapter 8 represents an effort to reignite this area of research.
+The formulation of OOD detection as a statistical test also allows for better guarantees for such systems in practice.
+For instance, as also discussed in \cref{chp:paper-modelagnostic}, the statistical framework enables false positive rate control which is a valuable property in many practical applications especially if they involve high-risk actions such as in medical decision-support.
+
+% In this follow-up work to \cref{chp:paper-hierarchical}, we note that the set of methods proposed for out-of-distribution detection using generative models is quite large and that many are tailored for specific model types, which suggests that it is possible to develop a model-agnostic approach.
+We hypothesize that by phrasing the task as a statistical testing problem and combining different tests, the method's efficacy can be improved and weaknesses inherent to any particular test can be alleviated.
+% From this hypothesis, we combine a classical parametric test with the recently introduced typicality test to develop a method applicable to any differentiable  generative model with explicit likelihood, and show that this leads to a more accurate out-of-distribution test.
+% Finally, we discuss the benefits of casting out-of-distribution detection as a statistical testing problem, for instance enabling false positive rate control.
+This property is valuable in many practical applications, especially in high-risk settings such as medical decision-making.
 
 \vspace{1em}
-\textbf{In \cref{chp:paper-modelagnostic}}
-
-\vspace{1em}
-\textbf{In \cref{chp:paper-brief}} we provided an overview of unsupervised neural speech representation learning.
+\textbf{\Cref{chp:paper-brief}} provided an overview of unsupervised neural speech representation learning.
 Such approaches have recently matched supervised methods on many tasks and represent a significant advance in low-resource settings, such as speech recognition for minority languages.
@@
 Indeed, the masked pre-training objective widely used for successful self"=supervised methods also corresponds to a maximum marginal likelihood objective \parencite{moreno-munoz_masked_2023}.
-This also leaves hope ... \moretodo{hope to improve VAEs by certain measures.}
+
+This also leaves potential for future work to improve the ability of VAEs to learn useful representations.
+As discussed in \cref{chp:paper-brief,sec_discussion:representation-learning-with-vaes}, promising approaches include adopting masked objectives for VAE training, improved architectural designs to impose better inductive biases, and incorporating advances in gradient estimators more widely in the literature \parencite{rainforth_tighter_2019,roeder_sticking_2017,tucker_doubly_2019,bauer_generalized_2021}, especially works on large VAE models \parencite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
 
@@
 \vspace{1em}
-\textbf{In \cref{chp:paper-automated}}
+\textbf{\Cref{chp:paper-benchmarking}} 
+% In chapter 8, stochastic and deterministic generative models were benchmarked in terms of model likelihood.
+In addition, the chapter presented a hierarchical latent variable model (LVM) for speech inspired by the Clockwork VAE [238].
+In contrast to self-supervised models, LVMs have not yet seen the same success for representation learning.
+As such, they offer little guidance in terms of alleviating the need for labeled data, as discussed in section 1.2.1.
+
+% However, since LVMs learn a distribution over the training data, they are forced to encode all aspects of it.
+While this may be redundant for some tasks, others benefit from a broad variety of features.
+A hierarchical model that operates on multiple temporal scales, such as the Clockwork VAE, offers a natural way to encode different feature categories.
+Pronunciation might be learned at lower layers, speaker identity at the upper layers, and semantic features in between.
+Although some work has successfully separated speaker identity from content [114], models that can learn a deep hierarchy of features for speech remains an open challenge.
+The model presented in chapter 8 represents an effort to reignite this area of research.
+
+\vspace{1em}
+\textbf{\Cref{chp:paper-automated}}
 
@@
 \vspace{1em}
-\textbf{In \cref{chp:paper-retrospective}} we studied how machine learning might be used to improve decision-making at emergency services. 
+\textbf{\Cref{chp:paper-retrospective}} studied how machine learning might be used to improve decision-making at emergency services. 
 We saw that a model was able to improve significantly on the stroke recognition ability of call-takers alone and that the features it used were sensible and related to symptoms and descriptions of stroke.

### 2023-12-02 appendices/supplementary_modelagnostic.tex
Update footnote placement

@@
 
-\paragraph{PixelCNN++} For PixelCNN++ we used the code available in this repository\footnote{\href{https://github.com/pclucas14/pixel-cnn-pp}{\texttt{https://github.com/pclucas14/pixel-cnn-pp}}}.
-For the greyscale images, we used one residual block per stage with 32 filters and 5 logistic components in the discretized mixture of logistics.
+\paragraph{PixelCNN++} For PixelCNN++ we used the code available in this repository.\footnote{\href{https://github.com/pclucas14/pixel-cnn-pp}{\texttt{https://github.com/pclucas14/pixel-cnn-pp}}} For the greyscale images, we used one residual block per stage with 32 filters and 5 logistic components in the discretized mixture of logistics.
 For natural images, instead, we used 5 residual blocks per stage with 160 filters and 10 components in the mixture.
@@
 
-\paragraph{Glow} For training Glow models we follow \textcite{kirichenko_why_2020} and their repository\footnote{\href{https://github.com/PolinaKirichenko/flows_ood}{\texttt{https://github.com/PolinaKirichenko/flows\_ood}}}.
-They closely follow \textcite{nalisnick_deep_2019} and \parencite{kingma_glow_2018} implementation for multi-scale Glow, where a scale is defined as the sequence of actorm, invertible $1\times1$ convolution and coupling layers.
+\paragraph{Glow} For training Glow models we follow \textcite{kirichenko_why_2020} and their repository.\footnote{\href{https://github.com/PolinaKirichenko/flows_ood}{\texttt{https://github.com/PolinaKirichenko/flows\_ood}}} They closely follow \textcite{nalisnick_deep_2019} and \parencite{kingma_glow_2018} implementation for multi-scale Glow, where a scale is defined as the sequence of actorm, invertible $1\times1$ convolution and coupling layers.
 While \textcite{kirichenko_why_2020} only considers the RMSProp optimizer, we trained two different models, one using RMSProp and one using Adam with batch-size 32.
@@
 \paragraph{Hierarchical VAE} We follow \parencite{havtorn_hierarchical_2021} for both model architecture design and training choices for our hierarchical VAEs.
-We used their open-sourced repository\footnote{\href{https://github.com/JakobHavtorn/hvae-oodd}{\texttt{https://github.com/JakobHavtorn/hvae-oodd}}}.
-As mentioned in the paper, the HVAE model we used has a bottom-up inference path and a top-down generative path.
+We used their open-sourced repository.\footnote{\href{https://github.com/JakobHavtorn/hvae-oodd}{\texttt{https://github.com/JakobHavtorn/hvae-oodd}}} As mentioned in the paper, the HVAE model we used has a bottom-up inference path and a top-down generative path.
 We trained each model for 1000 epochs using Adam optimizer with learning rate $3e-4$ and a batch-size of $128$.

### 2023-12-02 chapters/background.tex
Update footnote placement

@@
 
-However, the differential entropy does not have several of the desirable properties of the discrete version; it can be negative, it is not invariant under a change of variables, and it is not dimensionally correct%
+However, the differential entropy does not have several of the desirable properties of the discrete version; it can be negative, it is not invariant under a change of variables, and it is not dimensionally correct.%
 \footnote{\label{fn:dimensional-analysis-of-differential-entropy}
@@
     % Taking the position that a mole is quite simply a number (not a number of particles) \parencite{debievre_atomic_1992,baranski_atomic_2012}, and a bit is a number of elementary (information) entities, we could define a bit as a derived unit from the mole as $\SI{1}{bit} = \SI{1}{mol} / 6.02214076\times10^{23}$.
-}. 
+} 
 For these reasons, the differential entropy may not be a suitable measure of information, or uncertainty, for continuous random variables.
@@
 where $H_N(X)$ is the limiting density for discrete points, $q_X(x)$ is a uniform density over the quantization of the continuous space, and we have subtracted a $\log N$ term that would go to infinity in the limit of infinite points.
-By doing this, the information entropy becomes positive, dimensionally correct, and invariant under a change of variables%
+By doing this, the information entropy becomes positive, dimensionally correct, and invariant under a change of variables.%
 \footnote{\label{fn:change-of-variables-invariance-of-corrected-differential-entropy}
@@
     \end{equation*}
-}.
+}
 
@@
 We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \parencite{graham_denoising_2023,liu_unsupervised_2023a}.
-It is important to note that this distinction relates only to the availability of some target value, $y$ -- not whether OOD data is available for supervision\footnote{
+It is important to note that this distinction relates only to the availability of some target value, $y$ -- not whether OOD data is available for supervision.\footnote{
     In our overview, we distinguish between supervised and unsupervised out-of-distribution by letting methods be classified as supervised that use any kind of target value $y$, whether it relates to an original (in-distribution) task, available OOD data, or both.
@@
 This difference in nomenclature is currently unresolved in the literature.
-}. 
+} 
 Supervised OOD detection tries to assess whether the model's prediction $\hat{y}$ via $p(y|\xb)$ should be trusted for a given input $\xb$, whereas unsupervised OOD detection judges whether the input $\xb$ should be trusted and used at all.

### 2023-12-02 chapters/discussion.tex
Update footnote placement

@@
 While this statement of course depends on the task and the amount and type of unlabeled and labeled data, self"=supervised methods for speech are generally evaluated on downstream tasks that are harder and more complex than those used to evaluate VAEs.
-This is evidenced by the leaderboards on benchmarks such as SUPERB, SLUE and ZeroSpeech \parencite{yang_superb_2021,shon_slue_2021,dunbar_zero_2021} \footnote{One example of comparable evaluation is that between \cref{tab: phoneme recognition (PER),tab:unsupervisedASR} where self"=supervised methods can be seen to outperform VAEs for phoneme recognition while self"=supervised methods are also reported for word-level speech recognition.}.
+This is evidenced by the leaderboards on benchmarks such as SUPERB, SLUE and ZeroSpeech \parencite{yang_superb_2021,shon_slue_2021,dunbar_zero_2021}.\footnote{One example of comparable evaluation is that between \cref{tab: phoneme recognition (PER),tab:unsupervisedASR} where self"=supervised methods can be seen to outperform VAEs for phoneme recognition while self"=supervised methods are also reported for word-level speech recognition.}
 In the following, we will discuss potential directions of future research and try to shed light on why representations from VAEs underperform on downstream tasks and how they might be improved.
@@
 On the other hand, when we use maximum likelihood to train a latent variable model, we maximize the log-marginal likelihood $\log p(\xb)$.
-This can be equivalently phrased as maximizing the KL-divergence of the true data distribution $p_{\text{data}}(\xb)$ to the learned marginal, $D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb))$\footnote{%
+This can be equivalently phrased as maximizing the KL-divergence of the true data distribution $p_{\text{data}}(\xb)$ to the learned marginal, $D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb))$.\footnote{%
 Maximizing the likelihood $\sum_{n=1}^N \log p_\theta(\xb)$ w.r.t. parameters $\thetab$ is asymptotically equivalent to minimizing $D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb))$: 
@@
 % \end{align}
-}. 
+} 
 The first key observation then is that the marginal $p_{\thetab}(\xb)$, which we maximize, and the posterior $p_{\thetab}(\zb|\xb)$, which we want to produce useful representations, are separate properties of a latent variable model.

### 2023-12-02 chapters/paper_automated.tex
Fix paragraph error
Add source reference at the top of every paper chapter
Fix references
Add citation to top reference
Update initial reference
Update top reference
Update footnote placement

@@
 \label{chp:paper-automated}
+
+\textit{This chapter is a piece of original research published as part of the project:} \newline
+\begin{center}
+    \begin{enumerate}[leftmargin=8mm,rightmargin=8mm,topsep=0mm,label={[\Alph*]}]
+        \setcounter{enumi}{4}
+        \item \fullcite{edin_automated_2023} \co \hspace{0.1em} \parencite{edin_automated_2023}
+    \end{enumerate}
+\end{center}
+
 \ifthenelse{\equal{\skippapers}{true}}{}{
@@
 Medical coding is the task of assigning diagnosis and procedure codes to free-text medical documentation \parencite{dongAutomatedClinicalCoding2022,tengReviewDeepNeural2022}.
-
 These codes ensure that patients receive the correct level of care and that healthcare providers are accurately compensated for their services.
@@
 However, medical coding is error-prone, and, in many cases, deciding between certain codes can be a subjective matter \parencite{nouraeiAuditNatureImpact2013, lloydPhysicianCodingErrors1985}.
-\textcite{burnsSystematicReviewDischarge2012} systematically reviewed studies assessing the accuracy of human medical coders and found an overall median accuracy of 83.2\% (IQR: 67.3–92.1\%).
+\textcite{burnsSystematicReviewDischarge2012} systematically reviewed studies assessing the accuracy of human medical coders and found an overall median accuracy of 83.2\% (IQR: 67.3-92.1\%).
 \textcite{searleExperimentalEvaluationDevelopment2020} investigated the quality of the human annotations in MIMIC-III and concluded that 35\% of the common codes were under-coded.

### 2023-12-02 chapters/paper_benchmarking.tex
Add source reference at the top of every paper chapter
Fix references
Add citation to top reference
Update initial reference
Update top reference
Update footnote placement

@@
 \label{chp:paper-benchmarking}
+
+\textit{This chapter is a piece of original research published as part of the project:} \newline
+\begin{center}
+    \begin{enumerate}[leftmargin=8mm,rightmargin=8mm,topsep=0mm,label={[\Alph*]}]
+        \setcounter{enumi}{3}
+        \item \fullcite{havtorn_benchmarking_2022} \main \hspace{0.1em} \parencite{havtorn_benchmarking_2022}
+        \end{enumerate}
+\end{center}
+
 \ifthenelse{\equal{\skippapers}{true}}{}{
@@
 \end{itemize}
-At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models\footnote{\href{https://github.com/JakobHavtorn/benchmarking-lvms}{\url{github.com/JakobHavtorn/benchmarking-lvms}}}.
+At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models.\footnote{\href{https://github.com/JakobHavtorn/benchmarking-lvms}{\url{github.com/JakobHavtorn/benchmarking-lvms}}}
 % \footnote{\fontsize{8.5}{0}\selectfont\texttt{ \url{github.com/anonymous/available-later}}}.

### 2023-12-02 chapters/paper_benchmarking_workshop.tex
Update footnote placement

@@
 \paragraph{Scope and related work} 
-At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models\footnote{\texttt{ \href{https://github.com/JakobHavtorn/benchmarking-lvms}{github.com/JakobHavtorn/benchmarking-lvms}}}.
+At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models.\footnote{\texttt{ \href{https://github.com/JakobHavtorn/benchmarking-lvms}{github.com/JakobHavtorn/benchmarking-lvms}}}
 An extended version of this paper is available at \href{https://arxiv.org/abs/2202.12707}{arXiv:2202.12707}.

### 2023-12-02 chapters/paper_brief.tex
Change references and add todos
Add source reference at the top of every paper chapter
Fix references
Add citation to top reference
Update initial reference
Update top reference
Add label
Update footnote placement

--- 2023-10-14 appendices/paper_brief.tex
@@
 \chapter[a brief overview of unsupervised neural speech representation learning]{A Brief Overview of Unsupervised Neural Speech Representation Learning}
-\label{app:paper-brief}
+\label{chp:paper-brief}
+
+\textit{This chapter is a piece of original research published as part of the project:} \newline
+\begin{center}
+    \begin{enumerate}[leftmargin=8mm,rightmargin=8mm,topsep=0mm,label={[\Alph*]}]
+        \setcounter{enumi}{2}
+        \item \fullcite{borgholt_brief_2022} \co \hspace{0.1em} \parencite{borgholt_brief_2022}
+        \end{enumerate}
+\end{center}
+
 \ifthenelse{\equal{\skipappendices}{true}}{}{
@@
     \label{tab:model-taxonomy}
-    \begin{center}
+    % \begin{center}
+    \centering
     \setlength{\tabcolsep}{3pt}
@@
 
-            \textbf{APC} \footnotesize{\parencite{chung_unsupervised_2019}}          & 2019 Oct.
-& \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 5/4
+        \textbf{APC} \footnotesize{\parencite{chung_unsupervised_2019}}          & 2019 Oct. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 5/4
         \textbf{wav2vec} \footnotesize{\parencite{schneider_wav2vec_2019}}       & 2019 Apr.
@@
 & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/12
-            \textbf{SCPC} \footnotesize{\parencite{bhati_segmental_2021}}            & 2021 Jun.
-& \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ % 3/6
+        \textbf{SCPC} \footnotesize{\parencite{bhati_segmental_2021}}            & 2021 Jun. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ % 3/6
         \textbf{HuBERT} \footnotesize{\parencite{hsu_hubert_2021}}               & 2021 Jun.
@@
     }
-    \end{center}
+    % \end{center}
 \end{table}
@@
 
-In brief, LVMs of the VAE-type consist of an approximate posterior, $q(\mathbf{z}|\mathbf{x})$, an observation model, $p(\mathbf{x}|\mathbf{z})$, and a prior, $p(\mathbf{z})$. 
+In brief, LVMs of the VAE type consist of an approximate posterior, $q(\mathbf{z}|\mathbf{x})$, an observation model, $p(\mathbf{x}|\mathbf{z})$, and a prior, $p(\mathbf{z})$. 
 With reference to probabilistic coding theory, the approximate posterior is often referred to as the encoder and the observation model as the decoder \parencite{kingma_autoencoding_2014, rezende_stochastic_2014}.
@@
     \label{tab:lvm-model-primitives}
-    \begin{center}
+    % \begin{center}
+    \centering
     % \renewcommand{\arraystretch}{1.1}
@@
     % }
-    \end{center}
+    % \end{center}
 \end{table}
@@
     \label{tab:lvm-taxonomy}
-    \begin{center}
+    % \begin{center}
+    \centering
     \setlength{\tabcolsep}{3pt}
@@
     }
-    \end{center}
+    % \end{center}
 \end{table*}
@@
 \section{Discussion}
-\label{sec:mtax}
+\label{paper_brief_sec: discussion}
 

### 2023-12-02 chapters/paper_retrospective.tex
Minor formatting
Update footnote placement

@@
 Due to data incompleteness, and the way the audio data is stored, at CEMS, 2,730,199 contacts could not be matched to their corresponding audio file, however, 2,361,178 contacts were successfully matched.
-We found no obvious pattern in the matched and unmatched calls and we included all calls with at matching audio file.
+We found no obvious pattern in the matched and unmatched calls, and we included all calls with at matching audio file.
 Next, a call was regarded as a case of ground truth stroke positive when the CPR number in the CAD record matched that of a DanStroke record, and the patient-reported time of onset was close to the call start time.
@@
 The datasets used to evaluate call-taker performance and to train and evaluate the machine learning framework are legally restricted by Danish patient privacy and secrecy laws and are therefore, not publicly available.
-The data can be made available from the publication date but requires a Data Access Agreement, which is examined and approved by the ethics committees that approved this research \footnote{Danish Data Protection Agency, \url{https://www.datatilsynet.dk/english}}.
-For the same reason, the machine learning framework trained in this study is not publicly available; however, instructions on how to train it are included in the main manuscript and the supplementary material.
+The data can be made available from the publication date but requires a Data Access Agreement, which is examined and approved by the ethics committees that approved this research.\footnote{Danish Data Protection Agency, \url{https://www.datatilsynet.dk/english}} For the same reason, the machine learning framework trained in this study is not publicly available; however, instructions on how to train it are included in the main manuscript and the supplementary material.
 

### 2023-12-02 chapters/conclusion.tex
Work on conclusion

@@
 \vspace{1em}
-\textbf{\Cref{chp:paper-benchmarking}} 
-% In chapter 8, stochastic and deterministic generative models were benchmarked in terms of model likelihood.
-In addition, the chapter presented a hierarchical latent variable model (LVM) for speech inspired by the Clockwork VAE [238].
-In contrast to self-supervised models, LVMs have not yet seen the same success for representation learning.
-As such, they offer little guidance in terms of alleviating the need for labeled data, as discussed in section 1.2.1.
-
-% However, since LVMs learn a distribution over the training data, they are forced to encode all aspects of it.
-While this may be redundant for some tasks, others benefit from a broad variety of features.
-A hierarchical model that operates on multiple temporal scales, such as the Clockwork VAE, offers a natural way to encode different feature categories.
-Pronunciation might be learned at lower layers, speaker identity at the upper layers, and semantic features in between.
-Although some work has successfully separated speaker identity from content [114], models that can learn a deep hierarchy of features for speech remains an open challenge.
-The model presented in chapter 8 represents an effort to reignite this area of research.
+\textbf{\Cref{chp:paper-benchmarking}} conducted a comprehensive evaluation of stochastic and deterministic generative models, focusing on their model likelihood. 
+The chapter also introduced a novel hierarchical VAE type model for speech, by drawing inspiration from the Clockwork VAE \parencite{saxena_clockwork_2021}. 
+Despite the supposed limitations of VAEs for representation learning as compared to self"=supervised methods (\cref{chp:discussion}), hierarchical VAE models for sequence data that operate on multiple temporal scales, provide a natural framework for encoding distinct feature categories.
+For instance, pronunciation features may be learned at lower layers, speaker identity at upper layers, and semantic features in intermediate layers.
+Despite successful attempts at isolating speaker identity from content in some existing work \parencite{hsu_unsupervised_2017}, developing VAE models with the capacity to learn a deep hierarchy of features for speech persists.
+The model presented in \cref{chp:paper-benchmarking} is an attempt at this challenge.
+% Despite the current limitations, VAEs possess a unique ability to learn a distribution over the training data, requiring them to encode diverse aspects of the dataset.
+This inclusivity, while potentially redundant for certain tasks, proves advantageous for others that benefit from a wide array of features.
 
 \vspace{1em}
-\textbf{\Cref{chp:paper-automated}}
+\textbf{\Cref{chp:paper-automated}} examined existing works on automated medical coding and found that, in any cases, training was suboptimal, and evaluation standards were biased.
+We performed a revised comparison of the selected models and provided updated conclusions on the relative performance of models, and the impact of rare codes and long discharge note documents.
 
-Future directions
-- Avoiding predicting several mutually exclusive classes, which is a general problem for multi-label
+The practical impact of the work of \cref{chp:paper-automated}, and the work in the field of medical coding in general, depend on a number of factors. 
+Much of the field has focused on the MIMIC datasets which originate from the emergency department and ICU of a single hospital.
+These datasets have enabled much of the progress in the field, but their singular data source also reduces the generality of the derived results and risks biasing the directions of research deemed most impactful \parencite{tengReviewDeepNeural2022, venkateshAutomatingOverburdenedClinical2023,johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}.
+Furthermore, the complexity and multi-label nature of medical coding has lead to a high prevalence of label errors in MIMIC-III \parencite{searleExperimentalEvaluationDevelopment2020}.
+While difficult to remove entirely, having multiple diverse datasets would also alleviate the risk of biases in such errors leading to misguided conclusions.
+Besides gathering more data, as also mentioned in \cref{chp:paper-automated}, evaluation methods, such as human-in-the-loop, could be useful to increase the reliability of results.
+
+A weakness of medical coding models is that performance varies widely between classes.
+The long-tailed distribution of code-frequency is particularly challenging and leads to underperformance on rare codes.
+Practical applications could still benefit from such models though.
+By limiting model predictions to a subset of codes for which they perform well, practitioners could focus on harder to code cases.
+This selection of cases in practical applications of medical coding could also benefit from research into selective prediction \parencite{geifman_selective_2017}.
+Although pre"=training is an obvious approach to improve performance when little data is available, the improvements observed by using pre"=trained models for medical coding have been limited \parencite{jiDoesMagicBERT2021,gaoLimitationsTransformersClinical2021,michalopoulosICDBigBirdContextualEmbedding2022,pascualBERTbasedAutomaticICD2021,zhangBERTXMLLargeScale2020} compared to the effect of pre"=training in other domains \parencite{mohamed_selfsupervised_2022, linPretrainedTransformersText2021,baevski_wav2vec_2020,devlin_bert_2018,dosovitskiy_image_2021}.
+This suggests an untapped potential for future research into more targeted pre"=training for medical coding.
+
+While our work in \cref{chp:paper-automated} focused on unimodal medical coding models, it is highly likely that future state-of-the-art models will augment discharge summaries with multi-modal inputs such as medical code descriptions, synonyms, and hierarchies \parencite{kimReadAttendCode2021, mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, caoHyperCoreHyperbolicCograph2020, baoMedicalCodePrediction2021, yuanCodeSynonymsMatter2022,caoHyperCoreHyperbolicCograph2020, xieEHRCodingMultiscale2019}.
+A main reason for this is that coding standards are updated regularly.
+The ICD standard for instance sees revisions and new codes added yearly, with local adoption following national guidelines \parencite{centerfordiseasecontrolandpreventioncdc_international_2023}.
+Leveraging such modalities will enable adapting models to updated coding standards without having to gather large amounts of data.
+
+% The use of supportive systems for medical coding in practice relies on
+
+% - Errors in the gold-standard data: \textcite{searleExperimentalEvaluationDevelopment2020} investigated the quality of the human annotations in MIMIC-III and concluded that 35\% of the common codes were under-coded.
+Such errors and subjectivity in manual medical coding make model training and evaluation challenging and suggests that additional evaluation methods using, e.g., a human-in-the-loop, could be useful to increase the reliability of results.
+
+% Future directions
+% - Avoiding predicting several mutually exclusive classes, which is a general problem for multi-label classification.
+
+% - research should focus on improving performance on rare codes while, in the shorter term, developing methods to detect codes that are too challenging for automated coding and, therefore, should be coded manually
+
+% - Fully leveraging pre"=training (while PLM-ICD outperforms the other models in this paper, the improvements are limited compared to the effect of pre"=training in other domains \parencite{mohamed_selfsupervised_2022, linPretrainedTransformersText2021,baevski_wav2vec_2020,devlin_bert_2018,dosovitskiy_image_2021})
+% Notably, there have been several unsuccessful attempts at using pre"=trained transformers for medical coding \parencite{jiDoesMagicBERT2021,gaoLimitationsTransformersClinical2021,michalopoulosICDBigBirdContextualEmbedding2022,pascualBERTbasedAutomaticICD2021,zhangBERTXMLLargeScale2020}
+
+% - Generalization to other hospitals.
+discharge summaries from outpatient care are often easier to code than summaries from inpatient care as they are shorter with fewer codes per document \parencite{zhangBERTXMLLargeScale2020, liuEffectiveConvolutionalAttention2021, tsengAdministrativeCostsAssociated2018}
 
 \vspace{1em}
-\textbf{\Cref{chp:paper-retrospective}} studied how machine learning might be used to improve decision-making at emergency services. 
+\textbf{\Cref{chp:paper-retrospective}} studied how machine learning might be used to improve decision-making at emergency services in relation to stroke detection. 
 We saw that a model was able to improve significantly on the stroke recognition ability of call-takers alone and that the features it used were sensible and related to symptoms and descriptions of stroke.

### 2023-12-03 chapters/conclusion.tex
Fix decision-support -> decision support

@@
 \textbf{\Cref{chp:introduction}} introduced the motivational cases of automated medical coding and stroke recognition and used them to exemplify the importance of out-of-distribution detection, and, by extension, representation learning.
-In the context of these cases, we discussed possible machine learning system designs for decision-support and considered potential sources of uncertainty and ideal model behavior.
+In the context of these cases, we discussed possible machine learning system designs for decision support and considered potential sources of uncertainty and ideal model behavior.
 The cases form a reference point for the thesis as a whole, connecting its contributions within out-of-distribution detection and representation learning back to practical applications. 
@@
 The formulation of OOD detection as a statistical test also allows for better guarantees for such systems in practice.
-For instance, as also discussed in \cref{chp:paper-modelagnostic}, the statistical framework enables false positive rate control which is a valuable property in many practical applications especially if they involve high-risk actions such as in medical decision-support.
+For instance, as also discussed in \cref{chp:paper-modelagnostic}, the statistical framework enables false positive rate control which is a valuable property in many practical applications especially if they involve high-risk actions such as in medical decision support.
 

### 2023-12-23 chapters/discussion.tex
Updates

@@
 While this statement of course depends on the task and the amount and type of unlabeled and labeled data, self"=supervised methods for speech are generally evaluated on downstream tasks that are harder and more complex than those used to evaluate VAEs.
-This is evidenced by the leaderboards on benchmarks such as SUPERB, SLUE and ZeroSpeech \parencite{yang_superb_2021,shon_slue_2021,dunbar_zero_2021}.\footnote{One example of comparable evaluation is that between \cref{tab: phoneme recognition (PER),tab:unsupervisedASR} where self"=supervised methods can be seen to outperform VAEs for phoneme recognition while self"=supervised methods are also reported for word-level speech recognition.}
+This is evidenced by the leaderboards on benchmarks such as SUPERB, SLUE and ZeroSpeech \parencite{yang_superb_2021,shon_slue_2021,dunbar_zero_2021}.\footnote{One example of comparable evaluation is that between \cref{tab: phoneme recognition (PER),tab:unsupervisedASR} where self"=supervised methods can be seen to outperform VAEs for phoneme recognition while self"=supervised methods are additionally reported for word-level speech recognition.}
 In the following, we will discuss potential directions of future research and try to shed light on why representations from VAEs underperform on downstream tasks and how they might be improved.
@@
 
-Including masked objectives into the principled probabilistic framework of VAEs might require rethinking the principles themselves, but it seems likely that relaxing the requirement of exact input reconstruction and enforcing the need for encoding (and decoding) contextual information could pave the way for future types of VAEs to learn representations that are competitive with self"=supervised methods.
+Including masked objectives into the principled probabilistic framework of VAEs might require rethinking the principles themselves.
+Nonetheless, it seems likely that relaxing the requirement of exact input reconstruction and enforcing the need for encoding (and decoding) contextual information could pave the way for future types of VAEs to learn representations that are competitive with self"=supervised methods.
 
-\moretodo[inline]{Out-of-distribution detection on speech?}
+\section{Out-of-distribution detection for speech}
+%
+% \moretodo[inline]{Out-of-distribution detection on speech?
+%
+% Outline:
+% Intro paragraph drawing from OOD detection on images.
+% Differences and unique challenges. Generally a harder problem than OOD detection for images
+
+The studies of \cref{part:unsupervised-uncertainty-estimation} dealt with OOD detection in the context of image classification, somewhat casually ignoring the fact that this thesis is mainly interested in audio and speech data.
+Image classification has been the de facto standard task for benchmarking OOD detection methods
+
+For certain image distributions, it might be possible to deem an image OOD due to the color or intensity of a single pixel. 
+However, for most complex distributions we are usually interested in a test relating to more semantic properties of the data - for instance the animal class, as discussed in the introduction (\cref{subsec:understanding-uncertainty}).
+
+For speech, this property is similar; a single amplitude value is unlikely to provide much information about the distributional origin of a speech sample.
+Instead, whether a sample is OOD must be inferred from correlations and patterns found in amplitudes along the temporal dimension.
+
+One study \parencite{bukhsh_outofdistribution_2023} ... (OOD detection on audio)
+
+% - Temporal scale, can be OOD in local segments but not entire sequence. 
+% - Harder to work with: Images provide immediate, human-interpretable feedback but audio requires playback which takes time.
+%
 

### 2023-12-23 chapters/introduction.tex
Add some files
Fix decision-support -> decision support
Updates

@@
 % - Why should they care about this topic? 
-% - What is the historical context of decision-support, what are the practical implications of it, and what happens when it fails.
+% - What is the historical context of decision support, what are the practical implications of it, and what happens when it fails.
 
@@
     \centering
-    \includegraphics[width=0.85\textwidth]{uber_nhsa_accident.png}
+    \includegraphics[width=0.85\textwidth]{uber_nhsa_accident_cropped.png}
     \caption[A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.]{ A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018.
@@
 % \noindent The research presented in this thesis will focus on the use of machine learning in relation to uncertainty estimation, speech processing, and medical conversations. 
-\noindent The research presented in this thesis deals with the use of machine learning for medical interviews with a focus on uncertainty estimation and speech representation learning. 
+\noindent The research presented in this thesis deals with the use of machine learning for medical interviews with a focus on uncertainty estimation and representation learning. 
 The research project was partially funded by a grant from the national Danish Innovation Fund (grant no.\@ 0153-00167B) and defined and carried out in collaboration with the Danish health tech company Corti, which develops computer software for medical decision support.
@@
 % - What is the context of the research and why is it important? 
-% - What are the current applications of machine learning in the domain (decision-support in healthcare)?
+% - What are the current applications of machine learning in the domain (decision support in healthcare)?
 % - What are the practical implications of the research? 

### 2023-12-23 frontmatter/abstract-da.tex
Updates

@@
 Sundhedsplejen adskiller sig dog fra andre domæner på grund af den høje risiko forbundet med selv små fejl.
-Da ingen model er fejlfri, tilskynder dette til at associere modelprædiktioner med robuste usikkerhedsestimater, især i scenarier, der involverer ude-af-fordeling (OAF) data, såsom sjældne sygdomme.
+Da ingen model er fejlfri, tilskynder dette til at associere modelprædiktioner med robuste usikkerhedsestimater, især i scenarier, der involverer ude-af-fordeling (UAF) data, såsom sjældne sygdomme.
 
@@
 \begin{enumerate}[topsep=3pt, partopsep=0pt, itemsep=3pt, parsep=0pt, leftmargin=2em, label=(\alph*)] %, label=(\roman*)]
-    \item En likelihood-ratio score til OAF-detektion med variationelle autoenkodere, der afhjælper den bevist negative effekt af lav-niveau features.
-    \item En statistisk test til OAF-detektion, der kombinerer score- og typikalitetstests og kan bruges med likelihoods fra enhver differentiabel generativ model.
+    \item En likelihood-ratio score til UAF-detektion med variationelle autoenkodere, der afhjælper den bevist negative effekt af lav-niveau features.
+    \item En statistisk test til UAF-detektion, der kombinerer score- og typikalitetstests og kan bruges med likelihoods fra enhver differentiabel generativ model.
     \item Et benchmark for probabilistiske talerepræsentationsmodeller og en ny metode til at lære hierarkiske repræsentationer.
@@
 %
+
 Sammenfattende adresserer denne afhandling udfordringer indenfor usikkerhedsestimering og repræsentationslæring til tale og udforsker medicinske anvendelser af maskinlæring. 

### 2023-12-30 chapters/background.tex
Work on conclusion
Updates to discussion

@@
 \paragraph{Related tasks} OOD detection bears many similarities with anomaly detection, novelty detection, open set recognition, and outlier detection \parencite{yang_generalized_2022}.
+%, and selective prediction \parencite{yang_generalized_2022, geifman_selective_2017}.
 Some of these differences are subtle, and the terminology is not always used consistently in the literature.
@@
 Drawing inspiration from Robust Principal Component Analysis, \textcite{zhou_anomaly_2017} propose to first split the input data into a dense low-rank factor and a sparse factor, assuming that outliers are caught in the sparse factor, and then use a deep autoencoder to reconstruct the dense factor.
-\textcite{zong_deep_2018} jointly learn a deep autoencoder and a Gaussian Mixture Model on the learned hidden representations and draw parallels of their method to neural variational inference \parencite{mnih_neural_2014}.
+\textcite{zong_deep_2018} jointly learn a deep autoencoder and a Gaussian Mixture Model on the learned hidden representations and point out similarities of their method to neural variational inference \parencite{mnih_neural_2014}.
 Similarly to the ensemble-based methods for supervised OOD detection, \textcite{chen_outlier_2017} propose to use an ensemble of autoencoders to reconstruct the input and use the median reconstruction error as the OOD score.
 
-Although generative adversarial networks do not have the ability to encode a given data point, methods have been proposed to invert the generator to find a latent representation that can be used to reconstruct the input and use the reconstruction error as well as a discriminator score for OOD detection \parencite{schlegl_unsupervised_2017,li_madgan_2019}.
+Although generative adversarial networks do not have the ability to encode a given data point, methods have been proposed to invert the generator to enable finding a latent representation.
+This representation can then be used to reconstruct the input and a reconstruction error, or a discriminator score, can be used for OOD detection \parencite{schlegl_unsupervised_2017,li_madgan_2019}.
 

### 2023-12-30 chapters/discussion.tex
Updates to discussion

@@
 % Differences and unique challenges. Generally a harder problem than OOD detection for images
+Somewhat casually ignoring the fact that this thesis is mainly interested in audio and speech data, the studies of \cref{part:unsupervised-uncertainty-estimation} dealt with OOD detection in the context of image classification.
+The reason for this is that image classification has been the de facto standard task for benchmarking OOD detection methods since the first papers on the topic \parencite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019,kirichenko_why_2020}.
+In this section, we shall briefly discuss OOD detection in the context of audio data.
 
-The studies of \cref{part:unsupervised-uncertainty-estimation} dealt with OOD detection in the context of image classification, somewhat casually ignoring the fact that this thesis is mainly interested in audio and speech data.
-Image classification has been the de facto standard task for benchmarking OOD detection methods
-
-For certain image distributions, it might be possible to deem an image OOD due to the color or intensity of a single pixel. 
-However, for most complex distributions we are usually interested in a test relating to more semantic properties of the data - for instance the animal class, as discussed in the introduction (\cref{subsec:understanding-uncertainty}).
-
-For speech, this property is similar; a single amplitude value is unlikely to provide much information about the distributional origin of a speech sample.
+In certain cases, it might be possible to classify an image OOD due to the color or intensity of a single pixel. 
+For most distributions, however, we are interested in a test that is sensitive to more semantic properties of the data - for instance the animal class, as discussed in the introduction (\cref{subsec:understanding-uncertainty}).
+For speech, this property holds too; a single amplitude value is unlikely to provide much information about the distributional origin of a speech sample.
+In fact, audio samples are one-dimensional which likely makes them less discriminatory compared to image pixels that generally have multiple channels e.g.
+for color.
 Instead, whether a sample is OOD must be inferred from correlations and patterns found in amplitudes along the temporal dimension.
 
-One study \parencite{bukhsh_outofdistribution_2023} ... (OOD detection on audio)
+Few studies have explored OOD detection specifically for speech.
+One recent study has explored OOD detection for speech \parencite{bukhsh_outofdistribution_2023}.
+The authors propose a supervised approach that uses a $k$NN classifier fitted to representations of the training data.
+The approach is tuned to a fixed recall on the training data and operates on full length audio samples.
+This method is similar to image methods that fit a Gaussian mixture to an appropriate layer \parencite{lee_simple_2018, xiao_we_2021}.
+In the early work of \textcite{hendrycks_baseline_2017}, the method (based on softmax probabilities) is also benchmarked on audio datasets including TIMIT and Aurora-2 \parencite{garofolo_timit_1993,pearce_aurora_2000}.
+
+% - Harder to work with: Images provide immediate, human-interpretable feedback but audio requires playback which takes time.
+The studies on OOD detection for audio reflect its similarity to OOD detection for images. 
+Nonetheless, OOD detection on audio comes with unique properties.
+Where images are immediately interpretable to a researcher, audio requires playback, often in real-time, to present its contents.
+This fact alone seems likely to play a role in OOD detection being understudied for audio data by simply making it more time-consuming.
+On top of this, audio is generally variable length and requires more memory than images.
 
 % - Temporal scale, can be OOD in local segments but not entire sequence.
-% - Harder to work with: Images provide immediate, human-interpretable feedback but audio requires playback which takes time.
-%
-
-% \subsection{Uncertainty estimation with self-supervised methods}
-% %
-
-% \textcite{pasad_layerwise_2021} found that layers 6-7 hold the most information about phonetic content and word identity and meaning for \texttt{wav2vec2-base}. 
-% For \texttt{wav2vec2-large} the phonetic content is highest in layers 11 and 18-19 with a drop in between, while word identity and meaning are highest between layers 12 and 18.
-
-% Uncertainty of representations versus uncertainty of downstream task outputs. 
-% % \textcite{wickstrom_relax_2023} propose a method for quantifying uncertainty based on
-
-% \lesstodo[inline]{Are self"=supervised speech representations useful for unsupervised uncertainty estimation?
-\parencite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}
-% % Since the main focus within self"=supervised learning has been on improving downstream task performance, very limited work, if any, has investigated self"=supervised representations in terms of uncertainty estimation.
-% % However, in the context of medical applications where data can be abundant but labels are sparse, unsupervised uncertainty estimation is a very interesting direction for future work.
-% \lesstodo[inline]{OOD data: Generalization or detection (https://arxiv.org/pdf/2110.11334.pdf)?}
 

### 2023-12-31 chapters/background.tex
Revisions

@@
 %
-VAEs are usually mixed-effect models. Although VAEs use variational Bayes to infer a posterior distribution over latent variables, their parameters are estimated using maximum likelihood. 
-That is, we want to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$ to find point estimates. 
+VAEs are almost always defined as mixed-effect models: VAEs use variational Bayes to infer a posterior distribution over latent variables, but their parameters are estimated using maximum likelihood. 
+Hence, parameter estimation entails jointly optimizing the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$ to find point estimates.
 Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
-Hence, we must approximate the expectation in the ELBO \cref{eq:variational-autoencoder-elbo} and compute its gradients w.r.t. $\phib$ and $\thetab$.
+Furthermore, we must approximate the expectation over the approximate posterior in the ELBO \cref{eq:variational-autoencoder-elbo} and compute its gradients w.r.t. $\phib$ and $\thetab$. 
+We will look into how batching affects the form of the ELBO in \cref{subsec: mutual-information-interpretation-of-elbo} and move on with posterior sampling and gradient estimation here.
 
@@
 
-Whereas we arrive at $\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)$ by first deriving the ELBO as a lower bound on the marginal likelihood and then estimating that bound by Monte Carlo sampling, \textcite{burda_importance_2016} instead propose to directly Monte Carlo sample the expectation in the logarithm of the intractable marginal likelihood \cref{eq:marginal-likelihood-with-variational-approximation},
+In the above we arrived at $\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)$ by first deriving the ELBO of \cref{eq:latent-variable-model-elbo} as a lower bound on the marginal likelihood and then approximating that bound by Monte Carlo sampling.
+Instead, \textcite{burda_importance_2016} propose to Monte Carlo sample the expectation inside the logarithm of \cref{eq:marginal-likelihood-with-variational-approximation} directly, without first deriving the ELBO.
+They obtain
 %
@@
 %
-where $w_\thetab(\xb,\zb_{s})$ are the importance weights. Note the difference with the estimator in \cref{eq:variational-autoencoder-elbo-sampling-expanded}. 
+where $w_\thetab(\xb,\zb_{s})$ can be seen as importance weights.
+Note that the location of the logarithm is the only difference with the estimator in \cref{eq:variational-autoencoder-elbo-sampling-expanded}.
 By taking the expectation of this estimator, we can see that it too is a lower bound on the marginal likelihood,

### 2023-12-31 chapters/discussion.tex
Finished discussion on OODD for speech

@@
 % ~5 pages
-%
-% OUTLINE:
-% - paper-hierarchical:
-%   - Suitability of VAEs for representation learning (minimization of mutual information and sensitivity to implicit prior such as architecture)
-% - paper-benchmarking
-%   - Inferiority of probabilistic methods compared to self"=supervised learning. 
-% -
-
-% \todo[inline]{Unsupervised representation learning for medical applications.
-We tried BERT for stroke but no success.
-Medical coding state-of-the-art uses pretrained BERT, but only marginally improves upon simpler baselines.}
-
-% \moretodo[inline]{Add to the discussion somewhere:
-% I would try to embed a perspective for contemplation.
-Instead of us always talking about what the technology needs to improve on.
-Perhaps, it would also be prudent to ask an open-ended philosophical question: “While our requirements to the technology are in its place, one could ask, whether human beings have the ability to quantify uncertainty.
-In 2023, the prevalence of human errors in driving are much higher than self-driving vehicles.” <- Quick example.
-I'm sure you can find something better.}
 
@@
 \moretodo{Adjust the below to include any discussion on out-of-distribution detection.}
-Having examined VAEs and self"=supervised approaches for representation learning in \cref{part:unsupervised-uncertainty-estimation,part:unsupervised-speech-representation-learning}, in this discussion we will consider how VAEs might be enabled to learn representations more competitive with self"=supervised methods for downstream tasks.
-Furthermore, since the work presented in \cref{part:medical-applications} only had limited focus on uncertainty, we will also present and discuss the use of calibration techniques for the stroke recognition model, drawing connections to how such systems might be perceived and used in practice.
+Having examined VAEs and self"=supervised approaches for representation learning in \cref{part:unsupervised-uncertainty-estimation,part:unsupervised-speech-representation-learning}, in this discussion we will consider how VAEs might be enabled to learn representations more competitive with self"=supervised methods for downstream tasks (\cref{sec_discussion:representation-learning-with-vaes}).
+We will also complement the works on OOD detection in \cref{part:unsupervised-uncertainty-estimation}, which focused on image data, with a discussion about similarities with OOD detection for audio and speech data (\cref{sec_discussion:out-of-distribution-detection-for-speech}).
+Furthermore, since the work presented in \cref{part:medical-applications} only had limited focus on uncertainty, we will also present and discuss the use of calibration techniques for the stroke recognition model, drawing connections to how such systems might be perceived and used in practice (\cref{sec: discussion-stroke-recognition-uncertainty}).
 
@@
 
-\section{Out-of-distribution detection for speech}
+\section{Out-of-distribution detection for speech}\label{sec_discussion:out-of-distribution-detection-for-speech}
 %
@@
 % Differences and unique challenges. Generally a harder problem than OOD detection for images
-Somewhat casually ignoring the fact that this thesis is mainly interested in audio and speech data, the studies of \cref{part:unsupervised-uncertainty-estimation} dealt with OOD detection in the context of image classification.
+Somewhat casually ignoring the fact that this thesis is mainly interested in audio and speech data, the studies of \cref{part:unsupervised-uncertainty-estimation} dealt with OOD detection only in the context of image classification.
 The reason for this is that image classification has been the de facto standard task for benchmarking OOD detection methods since the first papers on the topic \parencite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019,kirichenko_why_2020}.
-In this section, we shall briefly discuss OOD detection in the context of audio data.
+In this section, we shall provide some perspective on OOD detection in the context of audio data.
 
@@
 Instead, whether a sample is OOD must be inferred from correlations and patterns found in amplitudes along the temporal dimension.
+In this way, the task of OOD detection has a similar formulation across different data types.
 
@@
 This method is similar to image methods that fit a Gaussian mixture to an appropriate layer \parencite{lee_simple_2018, xiao_we_2021}.
-In the early work of \textcite{hendrycks_baseline_2017}, the method (based on softmax probabilities) is also benchmarked on audio datasets including TIMIT and Aurora-2 \parencite{garofolo_timit_1993,pearce_aurora_2000}.
+In the early work of \textcite{hendrycks_baseline_2017}, the OOD detection method (based on softmax probabilities) is also benchmarked on audio datasets including TIMIT and Aurora-2 \parencite{garofolo_timit_1993,pearce_aurora_2000}.
+These works strongly suggest that with good data representations, we can use similar methods for OOD detection on both audio and images.
 
 % - Harder to work with: Images provide immediate, human-interpretable feedback but audio requires playback which takes time.
-The studies on OOD detection for audio reflect its similarity to OOD detection for images. 
-Nonetheless, OOD detection on audio comes with unique properties.
-Where images are immediately interpretable to a researcher, audio requires playback, often in real-time, to present its contents.
-This fact alone seems likely to play a role in OOD detection being understudied for audio data by simply making it more time-consuming.
-On top of this, audio is generally variable length and requires more memory than images.
+% The studies on OOD detection for audio reflect its similarity to OOD detection for images. 
+Despite these similarities of the OOD detection task and its methods across data types, audio comes with unique properties. 
+Where images are immediately interpretable to a researcher by visual inspection, audio requires playback, often in real-time, to present its contents.
+This fact alone is likely to have reduced interest in audio data by simply making it more time-consuming to work with than images.
+On top of this, audio is generally variable length and requires more memory than images which increases complexity and the amount of compute required to obtain equivalent results compared to images.
+As such, audio-based OOD detection is likely to benefit from standardized frameworks like HuggingFace \parencite{wolf_huggingface_2020,lhoest_datasets_2021} which have made eased working with irregular data types.
 
 % - Temporal scale, can be OOD in local segments but not entire sequence. 
+Another fundamental difference between audio and images is that audio is additive; two signals can be added together to form a new, mixed signal that is meaningful in the sense that it could also have been recorded.
+Addition of two images on the other hand produces an image that is not generally representative of what might be photographed. 
+This property of images has led to the practice of evaluating OOD detection performance by comparing examples between different datasets (e.g. MNIST versus FashionMNIST). 
+The additive property of audio allows complementing this dataset-versus-dataset evaluation with an approach where in-distribution examples are augmented to be OOD by mixing with unknown sources.
+This was explored by \textcite{hendrycks_baseline_2017} but has not seen a wide exploration.
+
+Finally, most work on OOD detection has focused on an example-centric approach where an entire example is either in- or out-of-distribution. 
+For audio, however, it is easy to create examples where local segments are OOD while other parts are not. 
+Using the additive property, we could add a loud Morse signal on top of a dialogue to change a segment of in-distribution audio to OOD. 
+Similarly, silence might be used as a kind of padding to insert entirely OOD segments in between in-distribution parts without addition and without significant boundary effects. 
+For images, the equivalent approach is similar to regularization techniques such as CutMix \parencite{yun_cutmix_2019} which paste images on top of each other.
+But this usually creates strong boundary errors, which might be used or learned to easily detect such synthetic OOD examples.
+Examples that contain locally OOD parts might occur naturally too; both in the image and audio domains.
+All kinds of unknown sound can be considered OOD and hence any recording might be locally OOD.
+In image segmentation, unknown object classes might similarly be seen as locally OOD parts of an example.
+Future research might examine further how detection of locally OOD examples differs from the dominant example-centric approach.
 
@@
 
-\begin{figure}
+\begin{figure}[t!]
     \centering

### 2024-01-01 chapters/paper_hierarchical.tex
Update top reference
Change image to png
Fix commas after equations

@@
 
-\textit{This chapter is a piece of original research published as part of the project:} \newline
+\textit{This chapter is a piece of original research by the candidate, previously published elsewhere and reprinted here with permission:} \newline
 \begin{center}
@@
         \setcounter{enumi}{0}
-        \item \fullcite{havtorn_hierarchical_2021} \main \. \parencite{havtorn_hierarchical_2021}
+        \item \fullcite{havtorn_hierarchical_2021} \main \hspace{0.1em} \parencite{havtorn_hierarchical_2021}
         \end{enumerate}
@@
     %\vspace{0.17cm}
-    \includegraphics[width=1\columnwidth]{paper_hierarchical/biva_reconstructions.pdf}
+    \includegraphics[width=1\columnwidth]{paper_hierarchical/biva_reconstructions.png}
     % https://docs.google.com/drawings/d/1UXdA3c18Oek9_env1vEB9mqOAHZ-tu0ijWmvqmapWCk/edit?usp=sharing
@@
 \begin{equation}\label{eq_hierarchical:elbo}
-    \log p_\thetab(\xb) \geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}(\xb; \thetab, \phib)\ .
+    \log p_\thetab(\xb) \geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}(\xb; \thetab, \phib) \enspace .
 \end{equation}
@@
     % q_\phib(\zb|\xb)=q_\phib(\zb^{(L)}|\xb)q_\phib(\zb^{(L-1)}|\zb^{(L)})\cdots q_\phib(\zb^{(2)}|\zb^{(1)}),
-    q_\phib(\zb|\xb) = q_\phib(\zb^{(L)}|\xb)\textstyle\prod_{i=L-1}^{1} q_\phib(\zb^{(i)}|\zb^{(i+1)}) \ .
+    q_\phib(\zb|\xb) = q_\phib(\zb^{(L)}|\xb)\textstyle\prod_{i=L-1}^{1} q_\phib(\zb^{(i)}|\zb^{(i+1)}) \enspace .
 \end{equation}
@@
 \begin{equation}
-    \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb^{(i)}|\zb^{(i+1)})}{q_\phib(\zb^{(i)}|\zb^{(i-1)})} + \log \frac{p_\thetab(\zb^{(L)})}{q_\phib(\zb^{(L)}|\zb^{(L-1)})} \Big] \ .
+    \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb^{(i)}|\zb^{(i+1)})}{q_\phib(\zb^{(i)}|\zb^{(i-1)})} + \log \frac{p_\thetab(\zb^{(L)})}{q_\phib(\zb^{(L)}|\zb^{(L-1)})} \Big] \enspace .
 \end{equation}
@@
     \mathcal{L}      &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) \parallel p_\thetab(\zb|\xb)\right), \label{eq_hierarchical:likelihoods-as-exact} \\ 
-    \mathcal{L}^{>k} &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( p_\thetab(\zb^{(\leq k)}|\zb^{(>k)}) q_\phib(\zb^{(>k)}|\xb) \parallel p_\thetab(\zb|\xb)\right) \nonumber \ .
+    \mathcal{L}^{>k} &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( p_\thetab(\zb^{(\leq k)}|\zb^{(>k)}) q_\phib(\zb^{(>k)}|\xb) \parallel p_\thetab(\zb|\xb)\right) \nonumber \enspace .
 \end{align}
@@
     LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) \parallel p_\thetab(\zb|\xb)\right) \\
-                 &\quad + D_{\mathrm{KL}}\left( p_\thetab(\zb^{(\leq k)}|\zb^{(>k)}) q_\phib(\zb^{(>k)}|\xb) \parallel p_\thetab(\zb|\xb)\right) \ . \notag
+                 &\quad + D_{\mathrm{KL}}\left( p_\thetab(\zb^{(\leq k)}|\zb^{(>k)}) q_\phib(\zb^{(>k)}|\xb) \parallel p_\thetab(\zb|\xb)\right) \enspace . \notag
 \end{align}

### 2024-01-01 chapters/paper_modelagnostic.tex
Fix references
Add citation to top reference
Update initial reference
Update top reference
Fix commas after equations

@@
         \setcounter{enumi}{1}
-        \item \fullcite{bergamin_modelagnostic_2022} \co \label{enumerate: paper-modelagnostic}
+        \item \fullcite{bergamin_modelagnostic_2022} \co \hspace{0.1em} \parencite{bergamin_modelagnostic_2022}
         \end{enumerate}
@@
 \begin{equation}
-    \textup{MMD}_\Phi(P,Q) = \| E_{X\sim P}[\Phi(X)] - E_{Y\sim Q}[\Phi(Y)] \|_\mathcal{H}. \hspace{-.1em}
+    \textup{MMD}_\Phi(P,Q) = \| E_{X\sim P}[\Phi(X)] - E_{Y\sim Q}[\Phi(Y)] \|_\mathcal{H} \enspace . \hspace{-.1em}
 \end{equation}
@@
     \label{eq_modelagnostic:mmd}
-    \textup{MMD}_\Phi\left(\frac{1}{m}\sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \left\|\frac{1}{m}\sum_{i=1}^{m} \Phi(x_i) -  \frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)\right\|_\mathcal{H},
+    \textup{MMD}_\Phi\left(\frac{1}{m}\sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \left\|\frac{1}{m}\sum_{i=1}^{m} \Phi(x_i) -  \frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)\right\|_\mathcal{H} \enspace ,
 \end{equation}
@@
     \label{eq_modelagnostic:Fisher_score}
-    \Phi_\textup{Fisher} (x) = I(\theta)^{-\frac{1}{2}} \nabla \log p_\theta (x),
+    \Phi_\textup{Fisher} (x) = I(\theta)^{-\frac{1}{2}} \nabla \log p_\theta (x) \enspace ,
 \end{equation}
@@
     \label{eq_modelagnostic:mmd_fisher}
-    \textup{MMD}_{\Phi_\textup{Fisher}}\left(\frac{1}{m} \sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \\ \left\|\frac{I(\theta)^{-\frac{1}{2}}}{m} \sum_{i=1}^{m}  \nabla \log p_\theta (x_i) - \frac{I(\theta)^{-\frac{1}{2}}}{n}  \sum_{i=1}^{n}  \nabla \log p_\theta (\tilde{x}_i)\right\|_2.
+    \textup{MMD}_{\Phi_\textup{Fisher}}\left(\frac{1}{m} \sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \\ \left\|\frac{I(\theta)^{-\frac{1}{2}}}{m} \sum_{i=1}^{m}  \nabla \log p_\theta (x_i) - \frac{I(\theta)^{-\frac{1}{2}}}{n}  \sum_{i=1}^{n}  \nabla \log p_\theta (\tilde{x}_i)\right\|_2 \enspace .
 \end{multline}
@@
 %While $\Phi_\textup{Typical}$ is not as well motivated as the Fisher kernel, we found that it generally gives good results. 
-While $\Phi_\textup{Typical}$ is not as well motivated as a kernel as $\Phi_\textup{Fisher}$, the concepts of typicality and typical set can be used to explain unintuitive behaviours of probability distributions in high-dimensional space as highlighted by \textcite{nalisnick_deep_2019}.
+While $\Phi_\textup{Typical}$ is not as well motivated as a kernel as $\Phi_\textup{Fisher}$, the concepts of typicality and typical set can be used to explain unintuitive behaviors of probability distributions in high-dimensional space as highlighted by \textcite{nalisnick_deep_2019}.
 We also found that using this kernel generally gives good results for OOD tasks.
@@
 Given $T_1,\dots, T_k$ and corresponding $p$-values $p_1,\dots, p_k$, Fisher's method combines the $p$-values into a test statistic $X^2$ defined as
+%
 \begin{equation}
     \label{eq_modelagnostic:Fisher_method}
-    X^2 \sim -2 \sum_{j=1}^{k} \ln (p_j).
+    X^2 \sim -2 \sum_{j=1}^{k} \ln (p_j) \enspace .
 \end{equation}
+%
 In case all null-hypotheses are accepted, the resulting test statistic $X^2$ follows a chi-squared distribution with $2k$ degrees of freedom. 
@@
 
-Assume that we have two independent OOD data samples that follow a product of truncated normal distributions, with density proportional to $$\mathcal{N}(x | 0,I_D) \cdot  \mathbf{1}\{x_1>0,\ldots,x_D>0\}.$$
-
+Assume that we have two independent OOD data samples that follow a product of truncated normal distributions, with density proportional to
+%
+\begin{equation}
+    \mathcal{N}(x | 0,I_D) \cdot  \mathbf{1}\{x_1>0,\ldots,x_D>0\} \enspace .
+\end{equation}
+%
 We denote by $T_\text{score}^\text{ood},T_\text{score}^\text{id}$ and $T_\text{typicality}^\text{ood},T_\text{typicality}^\text{id}$ the statistics obtained when confronted with either OOD data from the truncated normal, or the in-distribution data.
@@
 \begin{equation}
-T_\text{score}^\text{ood} - T_\text{score}^\text{id} \sim 2 D \mu_\text{TN}^2,
+T_\text{score}^\text{ood} - T_\text{score}^\text{id} \sim 2 D \mu_\text{TN}^2 \enspace ,
 \end{equation}

### 2024-01-01 chapters/paper_retrospective.tex
Fix commas after equations

@@
 
-\begin{figure}[t]
+\begin{table}[t]
     \centering
-    \includegraphics[width=0.98\textwidth]{paper_retrospective/figure2.pdf}
     \caption[Prediction confusion matrices for stroke recognition.]{Prediction confusion matrices.
@@
 Numbers for the model are given as the rounded mean over eleven runs.}
+    \includegraphics[width=0.98\textwidth]{paper_retrospective/figure2.pdf}
     \label{fig_retrospective:figure2-prediction-confusion-matrices}
-\end{figure}
+\end{table}
 
@@
 We report the F1-score, sensitivity, PPV, FOR (equal to 1- negative predictive value), and FPR (equal to 1 - specificity).
-Due to the imbalanced nature of the dataset, the negative predictive value and specificity were greater than 99\% for all cases.
+Due to the unbalanced nature of the dataset, the negative predictive value and specificity were greater than 99\% for all cases.
 We reported FOR and FPR instead because such large numerical values exhibit low relative variance, thereby obfuscating comparisons.

### 2024-01-01 chapters/introduction.tex
Polish introduction

@@
 Fittingly, the first human name recorded in written history belonged to Kushim, a Sumerian accountant.
-Their signature has been identified on at least 18 clay tablets dated to 3400-3000 BCE, which document transactions and inventory related to barley and other ingredients used in beer production \parencite{nissen_archaic_1993}.
+Their signature has been identified on at least 18 clay tablets dated to 3400-3000 BCE, documenting transactions and inventory related to barley and other ingredients used in beer production \parencite{nissen_archaic_1993}.
 % A later example is the Egyptian Rhind Mathematical Papyrus dated to around 1550 BCE which describes various mathematical techniques for doing calculations involving multiplication, division, and fractions, and how to use them for practical purposes like measuring land, constructing buildings, and sharing resources \parencite{georges_universal_2001}.
-Later, the Babylonians developed a sophisticated system of mathematics that included algorithms for solving quadratic equations, calculating areas of shapes, and numerical methods at approximating square roots \parencite{fowler_square_1998}.
+Later, the Babylonians developed a sophisticated system of mathematics that included algorithms for solving quadratic equations and calculating areas of shapes, and numerical methods for approximating square roots \parencite{fowler_square_1998}.
 
@@
 Assumedly unbeknownst to Kushim, this tablet would also come to contain one of the world's first recorded examples of a mathematical error; the statement: $5 \times \sfrac{1}{2} = 5$ \parencite{nissen_archaic_1993}.
-Later, the scribe of the Plimpton 322 clay tablet made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011,cuneiformdigitallibraryinitiativecdli_mct_2005}.
+Later, the scribe of the Plimpton 322 clay tablet also made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011,cuneiformdigitallibraryinitiativecdli_mct_2005}.
 % Later, in its list of Pythagorean triplets, the Plimpton 322 clay tablet (\cref{fig:plimpton_332}) also has several errors, presumably made by a novice in Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011}.
@@
 Besides these human errors, limitations inherent to the tools themselves also posed problems. 
-The Babylonian number system, for instance, was well-equipped to represent rational numbers but inefficient for approximating irrational numbers. 
+The Babylonian number system, for instance, was well-equipped to represent rational numbers but inefficient at approximating irrational numbers. 
 This forced significant approximation errors such as the common practice of approximating $\pi$ with 3 \parencite{georges_universal_2001}. 
@@
 % the first programmable, electronic, general-purpose digital computers were developed, starting with the ENIAC in 1945 (\cref{fig:eniac_programmers}) \parencite{georges_universal_2001}. 
-As the technology developed and manufacturing processes were refined and scaled, electronic computers were adopted widely across society for various applications in science and industry, including healthcare \parencite{georges_universal_2001, harari_sapiens_2011}.
+As the technology matured and manufacturing processes were refined and scaled, electronic computers were adopted widely across society for various applications in science and industry, including healthcare \parencite{georges_universal_2001, harari_sapiens_2011}.
 % The societal changes brought about by the widespread adoption of computer technology through the latter half of the 20th century were so profound that some scholars refer to the era as the information age \parencite{georges_universal_2001, harari_sapiens_2011}.
@@
 Enabled by a massive increase in the performance and availability of parallel computing resources, such methods have set new standards for the abilities of computer systems at various tasks usually reserved for humans, such as vision, speech recognition, and natural language understanding \parencite{lecun_deep_2015}.
-In recent years, machine learning has also made its way into healthcare systems, where it has shown promising results on difficult tasks for instance within medical imaging \parencite{lundervold_overview_2019}, drug discovery \parencite{chen_rise_2018}, and clinical decision support \parencite{cite15, cite14}.
+In recent years, machine learning has also made its way into healthcare systems, where it has shown promising results on difficult tasks such as medical imaging \parencite{lundervold_overview_2019}, drug discovery \parencite{chen_rise_2018}, and clinical decision support \parencite{cite15, cite14}.
 
@@
 Machine learning, like previous technologies, carries both benefits and risks. 
-However, some of these risks may be heightened in machine learning due to the unique tasks it enables in domains such as healthcare and autonomous driving. 
+However, machine learning might exacerbate some of these risks due to the unique tasks it enables in domains such as healthcare and autonomous driving. 
 While applications in healthcare are nascent and generally move forward in small steps, autonomous driving has fast adopted machine learning and presents a pertinent example: 
@@
 
-Recently, European policymakers proposed legislation to enforce that machine learning applications live up to certain standards \parencite{europeancommission_briefing_2021}. 
+Recently, European policymakers proposed legislation to enforce that applied machine learning meets certain standards \parencite{europeancommission_briefing_2021}. 
 This so-called AI Act considers three categories of applications based on the potential risk they pose to society: limited, high, and unacceptable. 
@@
 
-One of the key advances needed to ensure that high-risk applications of machine learning comply with such requirements is accurate uncertainty estimation of their predictions. 
-The multitude of rapid misclassifications made by the Uber self-driving car in Tempe is a likely indication that the model responsible for object-classification was overconfident in its predictions and unable to represent or communicate uncertainty to other systems in the car, or the safety driver.
-As a consequence, the model's many incoherent predictions were taken at face value and subsystems in the car acted on them without appropriate consideration to their reliability.
-While any model will inevitably make wrong predictions, accurately estimating the associated uncertainty and communicating it to other systems or humans is necessary to reduce the risk that wrong predictions lead to catastrophic failures.
+One of the key advances needed to ensure that high-risk applications of machine learning comply with such requirements is accurate uncertainty estimation of machine predictions. 
+The multitude of rapid misclassifications made by the Uber self-driving car in Tempe is a likely indication that the model responsible for object-classification was overconfident in its predictions and unable to represent or communicate this uncertainty to other subsystems in the car, or the safety driver.
+As a consequence, the model's many incoherent predictions were taken at face value and other subsystems acted on them without appropriate consideration to their reliability.
+While any model will inevitably make wrong predictions, accurately estimating the associated uncertainty and communicating it to reliant systems or humans is necessary to reduce the risk that wrong predictions lead to catastrophic failures.
 Fittingly, a recent study by the European Parliamentary Research Services \parencite{europeanparliament_artificial_2022} concluded: 
@@
 The research project was partially funded by a grant from the national Danish Innovation Fund (grant no.\@ 0153-00167B) and defined and carried out in collaboration with the Danish health tech company Corti, which develops computer software for medical decision support.
-In the rest of this chapter, we will provide a high-level introduction to these topics and to Corti and discuss the motivation behind the research project.
+In the rest of this chapter, we will provide a high-level introduction to these topics, and to Corti, and discuss the motivation behind the research project.
 
@@
 This is likely due to the complexity of stroke cases which exhibit a wide range of symptoms that can be difficult to recognize over the phone.
-Additionally, stroke cases are relatively rare, occurring in only $0.25\%$ of all calls made to the Copenhagen 1-1-2 emergency line in 2021.
+Additionally, stroke cases are relatively rare, occurring in only $0.25\%$ of all calls made to the Copenhagen 1813 medical helpline in 2021.
 This makes it difficult for call-takers to gain practical experience with stroke cases which compounds to the fundamental difficulty of recognizing them.
@@
 Such a system might be trained on a dataset calls with verified stroke and non-stroke cases and evaluated against a held-out test set of calls where the call-taker has indicated whether they suspect a stroke.
-Due to the low prevalence of stroke cases, the dataset would likely be imbalanced, and the number of stroke cases limited.
+Due to the low prevalence of stroke cases, the dataset would be highly unbalanced, and the number of stroke cases limited.
 % The performance of such a system would depend on the quality of the automatic speech recognition model, the quality of the classifier, and the quality of the data used to train them. 
@@
 The process of medical coding is time-consuming and error-prone due to the vast number and high complexity of medical diagnoses and procedures.
-For instance, the widely used International Classification of Diseases (ICD) standard consist of 55,000 medical codes in version 10 and 85,000 in version 11 \parencite{worldhealthorganisationwho_international_2023}.
+For instance, the widely used International Classification of Diseases (ICD) standard consists of 55,000 medical codes in version 10 and 85,000 in version 11 \parencite{worldhealthorganisationwho_international_2023}.
 Additionally, a single clinical note will usually contain several diagnoses and procedures which must all be inferred from the natural language text \parencite{johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}.
@@
 However, training and using such as system comes with several challenges. 
-For instance, the prevalence of different medical codes in the training data is highly imbalanced; it is common to have several orders of magnitude difference between the frequency of the most frequent code and the least frequent code.
+For instance, the prevalence of different medical codes in the training data is highly unbalanced; it is common to have several orders of magnitude difference between the frequency of the most frequent code and the least frequent code.
 Since each clinical note is associated with multiple codes that have complex co-occurrence patterns, it is often impossible to exactly correct for this class imbalance by stratified sampling, which is the usual go-to approach to deal with class imbalance in machine learning.
 % This also gives rise to 
-Furthermore, some codes are highly similar, for instance the ICD-10 codes Z87.891 ``Personal history of nicotine dependence'' and F17.210 ``Nicotine dependence, cigarettes, uncomplicated'', but in practice, only one of them should be assigned to a given note.
+Furthermore, some codes are highly similar, for instance the ICD-10 codes Z87.891 ``Personal history of nicotine dependence'' and F17.210 ``Nicotine dependence, cigarettes, uncomplicated''.
+In practice, however, only one of them should be assigned to a given note.
 As with the stroke recognition system, we would like the model to express high degree of uncertainty when it makes a mistake.
@@
 The ability of a model to accurately estimate the uncertainty of its predictions plays an important role in its reliability.
-While associating any prediction with an accurate uncertainty estimate can be argued to improve both interpretability, explainability, and fairness, this is generally done through improved robustness to noise, outliers, and adversarial attacks.
-Given an input which cannot be mapped to a single output with certainty, the model can indicate that this prediction should not to be trusted. 
+While associating any prediction with an accurate uncertainty estimate can be argued to improve both interpretability, explainability, and fairness, the main goal is to improve the model's robustness to noise, outliers, and adversarial attacks.
+Given an input which cannot be mapped to a single output with certainty, the model can then indicate that its prediction should not to be trusted. 
 However, overfitting, memorization, and the use of training objectives that are proxies for the evaluation metrics of interest often lead to models that are miscalibrated; that is, the predicted probability of a class does not reflect the true probability of the model being correct \parencite{guo_calibration_2017, kull_temperature_2019}.
@@
 If presented with an image of a horse, the model has no option but to distribute 100\% total probability across the cat and dog categories, even though that is clearly wrong.
-Worse yet, to indicate uncertainty, we might expect the model to assign 50\% probability to each of the cat and dog categories, but sadly that behavior is not guaranteed. 
+Worse yet, to indicate maximal uncertainty, we might expect the model to assign 50\% probability to each of the cat and dog categories, but sadly that behavior is not guaranteed. 
 On the contrary, since no horses were in the training data, the model will not have learned specialized features for horses, nor learned to associate any relevant known features with horses.
@@
 An example of aleatoric uncertainty is the uncertainty in the transcription of a word due to a noisy environment or overlapping, slurred or mumbling speech. 
-Unknown words are arguably sources of epistemic uncertainty although a good speech recognition model may generalize well if the word's spelling and pronunciation follow the same patterns as the words in the training data.
-Another example of epistemic uncertainty is the occurrence of truly out-of-distribution examples, such as the horse in the image classifier example from earlier: inputs unlike anything the model has seen during training.
-%, especially if they require outputs unavailable to the model.
+Unknown words are arguably sources of epistemic uncertainty although a good speech recognition model might generalize well if the word's spelling and pronunciation follow the same patterns as in the training data.
+Another example of epistemic uncertainty is the occurrence of truly out-of-distribution examples such as the horse in the image classifier example from earlier: inputs unlike anything the model has seen during training, especially if they require outputs unavailable to the model.
 In high-dimensional data with a practically unlimited diversity in out-of-distribution examples, it is impossible to collect enough data to completely eliminate all potential sources of epistemic uncertainty.
@@
 
-Since aleatoric uncertainty is represented by the data, it can usually be modelled directly.
+As mentioned, aleatoric uncertainty is represented by the data, and so can usually be modelled directly.
 For instance, in the case of speech recognition, we can model the uncertainty in the transcription of a word by a distribution over the words in the vocabulary.
@@
 Other approaches include anomaly detection and the recent field of out-of-distribution detection which, for example, represent epistemic uncertainty by special output classes for out-of-distribution data, distributions over data representations, or distances between them (see \cref{sec:out-of-distribution-detection}).
-In this thesis, we will focus on out-of-distribution detection for quantifying epistemic uncertainty.
+In this thesis, we will focus on quantifying epistemic uncertainty using out-of-distribution detection.
 
@@
 
-In this introduction we provided a high-level overview of the motivation behind the research project by focusing on the use-cases of stroke recognition and automated medical coding as well as speech processing and the importance of uncertainty quantification.
+In this introduction we provided a high-level overview of the motivation behind the research project by focusing on the use-cases of stroke recognition and automated medical coding, as well as speech processing and the importance of uncertainty quantification.
 The remainder of \cref{part:background} consists of two chapters. 
@@
 \Cref{part:unsupervised-uncertainty-estimation,part:unsupervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. 
-\Cref{part:unsupervised-uncertainty-estimation} is made up of two papers dealing with out-of-distribution detection using generative models, including variational autoencoders \parencite{havtorn_hierarchical_2021,bergamin_modelagnostic_2022}.
+\Cref{part:unsupervised-uncertainty-estimation} is made up of two papers that deal with out-of-distribution detection using generative models, including variational autoencoders \parencite{havtorn_hierarchical_2021,bergamin_modelagnostic_2022}.
 \Cref{part:unsupervised-speech-representation-learning} consists of two papers that explore speech representation learning with variational autoencoders and self"=supervised methods \parencite{borgholt_brief_2022,havtorn_benchmarking_2022}.
-\Cref{part:medical-applications} consists of two papers on applications of machine learning within the medical domain, specifically medical coding of clinical notes \parencite{edin_automated_2023} and recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023}.
+\Cref{part:medical-applications} consists of two papers on applications of machine learning within the medical domain: medical coding of clinical notes \parencite{edin_automated_2023} and recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023}.
 Finally, \cref{part:discussion-and-conclusion} concludes the thesis by discussing the presented work and future directions for research.

### 2024-01-01 chapters/main-contributions.tex
Make some minor updates
Fix decision-support -> decision support
Polish contributions and research hypotheses

@@
 The chapters of \cref{part:unsupervised-uncertainty-estimation,part:unsupervised-speech-representation-learning,part:medical-applications} are self-contained studies and therefore detail their own research hypotheses and contributions.
-Since each study was written without consideration to the other chapters of the thesis, we here detail them in relation to the overall research project. 
-This constitutes the research hypotheses and contributions of the thesis.
+Since each study was written without consideration to the other chapters of this thesis, we here detail them in relation to the overall research project. 
+This constitutes the research hypotheses and the contributions of the thesis.
 
@@
 This part deals with unsupervised learning of speech representations and consists of two papers.
-The contributions of this part lie in analyzing and comparing different approaches to speech representation along with a comprehensive overview of the field.
+Speech representations are fundamental to any practical system for decision support as well as for uncertainty quantification on speech data.
+The contributions of this part lie in analyzing and comparing different approaches to speech representation and in providing a comprehensive overview of the field.
 
@@
 In this chapter, we present a comprehensive overview of unsupervised neural representation learning for speech.
-Previous research is categorized into self"-supervised methods and probabilistic latent variable models and described in a common notation.
+Previous research is categorized into self"=supervised methods and probabilistic latent variable models and described in a common notation.
 This description assists in developing a model taxonomy that shapes a discussion of the models' representational power, the associated learning strategies, and the methods used to evaluate them.
 The discussion points to interesting avenues of future research.
-An extended version of this overview paper that focuses exclusively on self"=supervised methods was also published as part of the project \ref{enumerate: paper-review} \parencite{mohamed_selfsupervised_2022}.
-This paper is included in \cref{app:paper-review} for reference.
+An extended version of this overview paper that focuses exclusively on self"=supervised methods was also published as part of the project \ref{enumerate: paper-review} \parencite{mohamed_selfsupervised_2022} and this paper is included in \cref{app:paper-review} for reference.
 
@@
 While uncertainty estimation is not a central theme to the two papers, the retrospective study performs a substantial evaluation of the explainability of the proposed model via an occlusion analysis on the text input.
-Later, in the discussion, we shall further consider uncertainty estimation in relation to medical applications focusing on the retrospective study.
+Later, in the discussion, we further consider uncertainty estimation in relation to medical applications focusing on the retrospective study (\cref{sec: discussion-stroke-recognition-uncertainty}).
 % \moretodo[inline]{Revise and add more details to this section. Do we discuss uncertainty for medical coding?}
@@
 We hypothesize that several previous works underperform for reasons more related to suboptimal hyperparameter tuning, incorrect evaluation, and crude data handling than to model design, and that performance and comparability can be improved by addressing these issues.
-We first reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is indeed attributable to weak configuration of the training and crudely sampled train-test splits with many extremely rare classes - several without examples in the training data.
+We first reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is indeed attributable to weak configuration of the training and poorly sampled train-test splits with many extremely rare classes - several without examples in the training data.
 We also identify and correct a widespread error in the calculation of the macro F1-score. 

### 2024-01-01 chapters/background.tex
Polish background

@@
     \end{align*}
-    Plugging this into the differential entropy \cref{eq:corrected-differential-entropy-jaynes} we arrive at the original expression but now with $y$ as the independent variable,
+    Plugging this into \cref{eq:corrected-differential-entropy-jaynes} we arrive at the original expression but now with $y$ as the independent variable,
     \begin{equation*}
@@
 Identifying the Kullback-Leibler divergence as a measure of this relative information entropy provides a natural point of convergence for this section since it is equivalently defined for both discrete and continuous random variables, contrary to the differential entropy.
-In this view, information entropy is a quantity that distills the distribution of a random variable into a single number that describes the diversity of the potential outcomes of the random variable.
+In this view, information entropy is a quantity that distills the distribution of a random variable into a single number describing the diversity of the potential outcomes of the random variable.
 
@@
 In \cref{subsec: overview-of-out-of-distribution-detection} we introduce out-of-distribution (OOD) detection and connect it to the concept of uncertainty. 
-Since the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic} are relatively short, we also provide a concise review of the literature on out-of-distribution detection in \cref{subsec: supervised-out-of-distribution-detection, subsec: unsupervised-out-of-distribution-detection}.
+Since the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic} are relatively short, we also provide a concise review of the literature on out-of-distribution detection in \cref{subsec: supervised-out-of-distribution-detection} and \ref{subsec: unsupervised-out-of-distribution-detection}.
 

### 2024-01-01 chapters/background.tex
Polish background

@@
 In the general case, we assume that we have a domain of in-distribution data, $\mathcal{D}_{\text{in}}$, and we would like to build a model that can be used to assess whether a test data point $\xb$ originates from that domain or not.
-Since such OOD data is unknown to the model, it is, by definition, a source of epistemic uncertainty. This makes OOD detection closely related to uncertainty quantification. 
+Since OOD data is unknown to the model, it is, by definition, a source of epistemic uncertainty. This makes the task of OOD detection an instance of uncertainty quantification. 
 % Only in cases where strong prior knowledge of the data is incorporated into a model, can we expect it to be able to make reliable predictions on OOD data.
@@
 To provide some clarity, we will briefly outline the taxonomy of \textcite{yang_generalized_2022}:
-Outlier detection most clearly differs from the other tasks by directly processing all observations and aiming to select outliers from a single contaminated dataset. 
+Outlier detection most clearly differs from the other tasks by aiming to select outliers from a single contaminated dataset by processing all observations together. 
 The remaining tasks differ in whether they detect both covariate and semantic shift and require the simultaneous classification of in-distribution classes.
 %, and whether the in-distribution data contains more than one class.
-Anomaly detection deals with multiclass data and detects both covariate and semantic shift without requiring simultaneous classification of in-distribution classes whereas novelty detection, open set recognition and OOD detection are usually concerned with semantic shifts.
-Although a vague difference, novelty detection usually defined in terms of a class of normal data, while OOD detection centers around the distribution of the training data.
+Anomaly detection deals with multiclass data and detects both covariate and semantic shift without requiring simultaneous classification of in-distribution classes whereas novelty detection, open set recognition and OOD detection are usually concerned only with semantic shifts.
+Although a vague difference, novelty detection is usually defined in terms of a class of normal data, while OOD detection centers around the distribution of the training data.
 Different from novelty detection, OOD detection methods sometimes draw on examples of OOD data or require the simultaneous classification of in-distribution classes, although this is not always the case.
@@
 % Another closely related field is that of network-based uncertainty estimation \parencite{neal_bayesian_1995, hernandez-lobato_probabilistic_2015, gawlikowski_survey_2023}. %[14, 10, 36, 20]. 
-Although we share the sentiment of \textcite{yang_generalized_2022}, who propose to unify the tasks as ``generalized out-of-distribution detection'', this thesis will follow the nomenclature used in the works most related to ours and refer to the task as out-of-distribution detection.
+Although we share the sentiment of \textcite{yang_generalized_2022}, who propose to unify all these tasks as ``generalized out-of-distribution detection'', this thesis will follow the nomenclature used in the works most related to ours and refer to the task as out-of-distribution detection.
 
@@
 
-\paragraph{Scoring} Central to any OOD detection method is the ability to assign a score $s(\xb) \in \mathbb{R}$ to a given input $\xb$ that indicates the degree to which the input is likely to be OOD.
+\paragraph{Scoring} Central to any OOD detection method is the ability to assign a score $s(\xb) \in \mathbb{R}$ to a given input $\xb$ to indicate the degree to which the input is likely to be OOD.
 After defining a score, we typically use a validation set to tune a threshold $\tau$ such that $\xb$ is considered OOD if $s(\xb) > \tau$.
@@
 
-There are several ways to define a score $s(\xb)$, and they be used to further categorize OOD detection methods. 
+There are several general ways to define a score $s(\xb)$, and they can be used to further categorize OOD detection methods. 
 Unsupervised methods often derive the score 
@@
 from the \emph{logits} of $p(y|\xb)$ \parencite{hendrycks_scaling_2022,liu_energybased_2020}, or 
-use a latent \emph{representation} of the input \parencite{lee_simple_2018,li_anomaly_2019,ndiour_outofdistribution_2020,cook_outlier_2020,zaeemzadeh_outofdistribution_2021}, similar to unsupervised methods.
+use a hidden \emph{representation} of the input \parencite{lee_simple_2018,li_anomaly_2019,ndiour_outofdistribution_2020,cook_outlier_2020,zaeemzadeh_outofdistribution_2021}, similar to unsupervised methods.
 In the following sections, we provide a more in-depth overview of the different approaches to OOD detection following the above categorization into supervised and unsupervised methods.
@@
 %
-In the supervised setting the supervision can come from different sources depending on how much we know, or assume to know, about the OOD data.
-Taking supervision from the target variable $y$ requires few assumptions while to representative OOD data.
+In the supervised setting, we take the supervision from the target variable $y$ to be unrelated to the nature of the OOD data. 
+However, a number of works on supervised OOD detection use real or synthetic OOD data to augment the training objective of the classifier $p(y|\xb)$ to better detect OOD data.
+For that reason, in the following, we will categorize the supervised OOD methods depending on whether they require the use of OOD data or not, and if they do, whether the used OOD data is real or synthetic.
 
@@
 In a similar vein, \textcite{dhamija_reducing_2018} propose losses designed to maximize the entropy of $p(y|\xb)$ and decrease feature magnitudes for OOD data sampled from other datasets. 
-\textcite{ruff_deep_2020} use semi"=supervised learning and learn representations of in-distribution data that concentrate close to a centroid in latent space while labeled outliers are pushed away from the centroid.
+\textcite{ruff_deep_2020} use semi"=supervised learning and learn representations of in-distribution data that concentrate close to a centroid in latent space, while labeled outliers are pushed away from the centroid.
 Other methods including MCD \parencite{yu_unsupervised_2019}, NGC \parencite{wu_ngc_2021}, and UDG \parencite{yang_semantically_2021} use external, unlabeled, noisy data to improve OOD detection performance without requiring cleanly labeled OOD examples.
 
-As we discussed earlier, for many modern applications of machine learning, input data is often high dimensional and complex which makes it difficult to obtain enough representative OOD data to ensure robust OOD detection capabilities.
+% As we discussed earlier, for many modern applications of machine learning, input data is often high dimensional and complex which makes it difficult to obtain enough representative OOD data to ensure robust OOD detection capabilities.
 This fundamentally limits the usefulness of methods that require samples of real OOD data.
@@
 
-Similar to actual OOD data, the usefulness of synthetic OOD data is also fundamentally limited by the intractability of sampling the complete space of OOD data.
+% Similar to actual OOD data, the usefulness of synthetic OOD data is also fundamentally limited by the intractability of sampling the complete space of OOD data.
 Again, these methods are fundamentally limited in their generality.
@@
 \paragraph{Methods not using OOD data}
-The least informed supervised OOD detection methods do not require instantiations of any kind of OOD data, real or synthetic, and do away with the associated limitations.
-
+% The least informed supervised OOD detection methods do not require instantiations of any kind of OOD data, real or synthetic, and do away with the associated limitations.
 % No OOD data, probs
-A baseline approach uses the maximum class probability of $p(y|\xb)$ directly by noting that it tends to be larger for correctly classified examples \parencite{hendrycks_baseline_2017}.
-Another baseline method proposes that a high entropy of $p(y|\xb)$ indicates an OOD input \parencite{ren_likelihood_2019}
-Other methods that derive the score from the classifier probabilities include \textcite{lakshminarayanan_simple_2017} who propose to use an ensemble of independently trained classifiers to discriminate between in-distribution and OOD data by evaluating the agreement between the classifiers, \textcite{devries_learning_2018} who augment the network with a confidence estimation branch that learns to estimate the confidence of the classifier separately from the probability, and \textcite{huang_importance_2021} who compute the gradient of the KL-divergence of the predictive distribution $p(y|\xb)$ to a uniform distribution noting that the magnitude of gradients is higher for in-distribution data than for OOD data.
+A group of methods do not rely on OOD data for accurate detection. 
+An early baseline approach uses the maximum class probability of $p(y|\xb)$ directly by noting that it tends to be larger for correctly classified examples \parencite{hendrycks_baseline_2017}.
+Another baseline method proposes that a high entropy of $p(y|\xb)$ indicates an OOD input \parencite{ren_likelihood_2019}.
+Other methods that derive the score from the classifier probabilities include that of \textcite{lakshminarayanan_simple_2017} who propose to use an ensemble of independently trained classifiers to discriminate between in-distribution and OOD data by evaluating the agreement between the classifiers, but also \textcite{devries_learning_2018} who augment the network with a confidence estimation branch that learns to estimate the confidence of the classifier separately from the probability, and \textcite{huang_importance_2021} who compute the gradient of the KL-divergence of the predictive distribution $p(y|\xb)$ to a uniform distribution noting that the magnitude of gradients is higher for in-distribution data than for OOD data.
 \textcite{huang_mos_2021} group the classes of the target variable $y$ and define an OOD class for each group.
 Each training example is then the correct target for one group and an OOD example for all other groups; a kind of hierarchical version of the OOD class of simpler baselines based on noise augmentation \parencite{ren_likelihood_2019}.
-The Variational Information Bottleneck \parencite{alemi_deep_2017} jointly learns a probabilistic latent representation, $p(\zb|\xb)$, and a classifier, $p(y|\xb)$, using a generalized variational autoencoder \parencite{kingma_autoencoding_2014} and tries to maximize the mutual information between $\zb$ and $y$ and minimize it between $\zb$ and $\xb$.
+The Variational Information Bottleneck \parencite{alemi_deep_2017} jointly learns a probabilistic latent representation, $p(\zb|\xb)$, and a classifier, $p(y|\xb)$, using a generalized variational autoencoder and tries to maximize the mutual information between $\zb$ and $y$ and minimize it between $\zb$ and $\xb$.
 
@@
 % No OOD data, feature
-The final category of supervised methods derive the score from a latent representation of the input. 
+The final category of supervised methods derive the score from a hidden representation of the input. 
 For instance, to represent a virtual OOD class, \textcite{wang_vim_2022} generate an additional logit by first computing the residual of the input's latent space representation against a principal feature space and then converting it to a valid logit by matching its mean over training samples to the average maximum logits.
@@
 \textcite{zaeemzadeh_outofdistribution_2021} force the ID samples to embed into a union of 1-dimensional subspaces during training and computes the minimum angular distance from the feature to the class-wise subspaces.
-NuSA \parencite{cook_outlier_2020} uses projects features onto the column space of the classification weight matrix and computes the ratio of the norm the projected and original features.
+NuSA \parencite{cook_outlier_2020} projects features onto the column space of the classification weight matrix and computes the ratio of the norm the projected and original features.
 \textcite{lee_simple_2018} fit a multivariate Gaussian distribution to the activations of the penultimate layer of a pre"=trained classifier and use the Mahalanobis distance to this distribution to evaluate whether inputs are OOD.
@@
 
-A general weakness of all supervised out-of-distribution detection is that in learning the task-specific model $p(y|\xb)$ a model may discard information about $p(\xb)$ which could be useful for out-of-distribution detection.
+\paragraph{A note on limitations}
+As we discussed earlier, for modern applications of machine learning, input data is often high dimensional and complex which makes it difficult to obtain enough representative OOD data to ensure robust OOD detection capabilities.
+This inescapable property limits the practical usefulness of methods that use real OOD data, as well as those that synthesize it.
+Furthermore, a general weakness of all supervised out-of-distribution detection is that in learning the task-specific model $p(y|\xb)$ a model may discard information about $p(\xb)$ which could be useful for out-of-distribution detection.
+While augmenting training objectives and (re)training models specifically for OOD detection might alleviate this issue, it does not remove the issue.
 
@@
 \textcite{bishop_novelty_1994} first proposed to use the likelihood assigned to data by a generative model as a measure for detecting anomalous data.
-Simply put, since the likelihood measures ``how probable the data is'', OOD data is expected to give lower likelihoods than in-distribution data.
+Simply put, since the likelihood measures ``how probable the model is to explain the data'', OOD data is expected to give lower likelihoods than in-distribution data.
 Although, this method originally gave encouraging results, the advent of deep generative models and their application to high-dimensional data has lead many to observe likelihoods for OOD data that are higher than for in-distribution data \parencite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019,kirichenko_why_2020}.
@@
 \textcite{zong_deep_2018} jointly learn a deep autoencoder and a Gaussian Mixture Model on the learned hidden representations and point out similarities of their method to neural variational inference \parencite{mnih_neural_2014}.
-Similarly to the ensemble-based methods for supervised OOD detection, \textcite{chen_outlier_2017} propose to use an ensemble of autoencoders to reconstruct the input and use the median reconstruction error as the OOD score.
+Similar to the ensemble-based methods for supervised OOD detection, \textcite{chen_outlier_2017} propose to use an ensemble of autoencoders to reconstruct the input and use the median reconstruction error as the OOD score.
 
-Although generative adversarial networks do not have the ability to encode a given data point, methods have been proposed to invert the generator to enable finding a latent representation.
+Although generative adversarial networks do not have the ability to encode a given data point, methods have been proposed to invert the generator to find an associated latent representation.
 This representation can then be used to reconstruct the input and a reconstruction error, or a discriminator score, can be used for OOD detection \parencite{schlegl_unsupervised_2017,li_madgan_2019}.
@@
 In the following, we will introduce VAEs starting from the definition of latent variable models.
-We then derive the marginal likelihood objective bound used to train them, discuss its different components and the approximation error, and highlight some central challenges of applying VAEs.
+We then derive the marginal likelihood objective bound used to train them, discuss its different components and approximation error, and highlight some central challenges of applying VAEs.
 
@@
 
-This choice makes both the marginal and posterior distributions intractable since they involve an integral over all possible values of the latent variable $\zb$. 
+Since the marginal and posterior distributions involve integrals over all possible values of the latent variable $\zb$, using neural networks makes both intractable. 
 This precludes training with commonly used methods such as expectation maximization and exact maximum likelihood estimation.
-Since Markov Chain Monte Carlo methods are usually too computationally expensive for large neural networks, this leads to variational inference as a scalable alternative \parencite{jordan_introduction_1999}.
+Since Markov Chain Monte Carlo methods are usually too computationally expensive for large neural networks, this leaves variational inference as the scalable alternative \parencite{jordan_introduction_1999}.
 
@@
 \paragraph{Evidence lower bound}
-By approximating the intractable posterior distribution $p(\zb|\xb)$ with a variational distribution $q(\zb)$, we can derive a lower bound on the also intractable marginal log-likelihood $\log p_\theta(\xb)$.
+By approximating the intractable posterior distribution $p(\zb|\xb)$ with a variational distribution $q(\zb)$, we can derive a lower bound on the also intractable marginal log-likelihood, $\log p_\theta(\xb)$.
 %
@@
 This makes it possible to efficiently infer the approximate posterior distribution for any input $\xb$.
-With amortized variational inference, the ELBO is simply be written as,
+Using amortized variational inference, the ELBO is written as,
 % The approximate posterior can also be used to compute a lower bound on the marginal distribution $p_\thetab(\xb)$ via Jensen's inequality, the so-called evidence lower bound (ELBO), 
@@
 This reveals a commonly used interpretation of the ELBO: 
-The expected log-likelihood of the generative model $\mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]$ acts as a reconstruction loss that encourages the model to make $q_\phib(\zb|\xb)$ as peaky as possible since mapping many $\zb$ to a single $\xb$ is hard.
+The expected log-likelihood of the generative model $\mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]$ acts as a reconstruction loss; since mapping many $\zb$ to a single $\xb$ is hard, it encourages the model to make $q_\phib(\zb|\xb)$ as peaky as possible.
 The negative KL-divergence between the approximate posterior $q_\phib(\zb|\xb)$ and the prior $p(\zb)$ acts as a regularizer that forces the approximate posterior to be within the support of the prior (reverse KL-divergence).
@@
 However, an optimal KL-divergence of zero requires $q_\phib(\zb|\xb) = p(\zb)$ which would mean that $\zb$ has become independent of $\xb$.
-This state is a poor local optimum and is usually referred to as posterior collapse.
+This state is a poor local optimum and is usually referred to as \emph{posterior collapse}.
 It is particularly prone to happen for strong generative models for instance with autoregressive dependencies $p(\xb_{t}|\xb_{1:t},\zb)$.
@@
 The ELBO is a lower bound on the marginal likelihood, so it is interesting to examine when the bound holds exactly. 
-As is standard for variational Bayesian methods, we can take the objective of the VAE to be minimization of the KL-divergence of the approximate posterior to the true posterior, $D_\mathrm{KL}\left( p_\thetab(\zb|\xb) \parallel q_\phib(\zb|\xb) \right)$ and rewrite it as follows.
+As is standard for variational Bayesian methods, we can equivalently take the objective of the VAE to be the minimization of the KL-divergence of the approximate posterior to the true posterior, $D_\mathrm{KL}\left( p_\thetab(\zb|\xb) \parallel q_\phib(\zb|\xb) \right)$.
+We can then rewrite this objective as follows.
 %
@@
 %
-We can see that the inference gap of the ELBO is exactly equal to the KL-divergence of the approximate posterior to the true posterior $D_\text{KL}\left( p_\theta(\zb|\xb) \parallel q_\phib(\zb|\xb) \right)$.
+We can see that the difference between the ELBO and the marginal likelihood is exactly equal to the KL-divergence of the approximate posterior to the true posterior $D_\text{KL}\left( p_\theta(\zb|\xb) \parallel q_\phib(\zb|\xb) \right)$.
+This difference is called the \emph{inference gap}.
 This also means that the approximate posterior $q_\phib(\zb|\xb)$ that maximizes the ELBO also best approximates the true posterior, as measured by KL-divergence.
@@
 The remaining gap is referred to as the \emph{approximation gap}.
-If the inference gap is closed, the ELBO becomes equal to the marginal likelihood, the value of which then depends on how well the generative model is fitted \cite{cremer_inference_2018,frellsen_deep_2019}.
+The inference gap is the sum of these two contributions.
+When the inference gap is closed, the ELBO becomes equal to the marginal likelihood, the value of which then depends on how well the generative model is fitted \cite{cremer_inference_2018,frellsen_deep_2019}.
 
@@
 Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
-Furthermore, we must approximate the expectation over the approximate posterior in the ELBO \cref{eq:variational-autoencoder-elbo} and compute its gradients w.r.t. $\phib$ and $\thetab$. 
+Furthermore, we must approximate the expectation over the approximate posterior in the ELBO \cref{eq:variational-autoencoder-elbo} to compute its gradients w.r.t. $\phib$ and $\thetab$. 
 We will look into how batching affects the form of the ELBO in \cref{subsec: mutual-information-interpretation-of-elbo} and move on with posterior sampling and gradient estimation here.
@@
 For some choices of prior and approximate posterior, such as Gaussians, the KL-divergence term of the ELBO can be computed analytically which reduces estimator variance. 
-This estimator can be written as,
+Such estimators can be written as,
 %
@@
 %
-Since $\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)$ is the sample average, it is an unbiased estimator of the ELBO, $\mathbb{E}\left[\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)\right] = \mathcal{L}_{S}(\xb;\thetab,\phib)$, and hence a biased estimator of the marginal likelihood, $\log p(\xb)$.
+Since $\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)$ is the sample average, it is an unbiased estimator of the ELBO, $\mathbb{E}\left[\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)\right] = \mathcal{L}_{S}(\xb;\thetab,\phib)$, and hence a biased estimator of the marginal likelihood, $\log_\thetab p(\xb)$.
 
@@
     &\leq \log \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \frac{1}{S}\sum_{s=1}^{S} w_\thetab(\xb,\zb_{s}) \right] \nonumber \\
-    &= \log p(\xb) \enspace .
+    &= \log p_\thetab(\xb) \enspace .
 \end{align}
@@
 \begin{equation}
-    \log p(\xb) \geq \mathcal{L}^{\text{IWAE}}_{S+1}(\xb;\thetab,\phib) \geq \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \enspace ,
+    \log p_\thetab(\xb) \geq \mathcal{L}^{\text{IWAE}}_{S+1}(\xb;\thetab,\phib) \geq \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \enspace ,
 \end{equation}
 %
-and $\lim_{S\rightarrow\infty} \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \rightarrow \log p(\xb)$, if $w_\thetab(\xb,\zb_{s})$ are bounded. 
-Hence, $\widehat{\mathcal{L}}^{\text{IWAE}}_{S}$ is a biased estimator of $\log p(\xb)$ but with a bias that goes to zero as the number of samples goes to infinity. 
+and $\lim_{S\rightarrow\infty} \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \rightarrow \log p_\thetab(\xb)$, if $w_\thetab(\xb,\zb_{s})$ are bounded. 
+Hence, $\widehat{\mathcal{L}}^{\text{IWAE}}_{S}$ is a biased estimator of $\log p_\thetab(\xb)$ but with a bias that goes to zero as the number of samples goes to infinity. 
 With $S=1$, the estimator is equivalent to the regular ELBO \cref{eq:variational-autoencoder-elbo} and so, it is always as tight, or tighter, than the ELBO \parencite{burda_importance_2016}.
@@
 
-To gain some additional insight into the objective that we are optimizing, we will use this expectation $\hat{p}(\xb)$ to form a factorization similar to \cref{eq:variational-autoencoder-elbo-factorized} but with the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$ in the KL-divergence \parencite{tomczak_trouble_2022}.
+To gain some additional insight into the objective that we are optimizing, we will use this expectation over $\hat{p}(\xb)$ to form a factorization similar to \cref{eq:variational-autoencoder-elbo-factorized} but with the aggregated posterior $q_\phib(\zb) = \int q_\phib(\zb|\xb)\hat{p}(\xb)\,\mathrm{d}\xb$ in the KL-divergence \parencite{tomczak_trouble_2022}.
 Under the expectation over the data distribution, the marginal likelihood is lower bounded by that expectation over the ELBO,
@@
 \begin{align} \label{eq:variational-autoencoder-elbo-expected}
-    \mathbb{E}_{\hat{p}(\xb)}\left[ \log p(\xb) \right] 
+    \mathbb{E}_{\hat{p}(\xb)}\left[ \log p_\thetab(\xb) \right] 
     &\geq \mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log \frac{p_{\thetab}(\xb|\zb)p(\zb)}{q_{\phib}(\zb|\xb)} \right] \enspace ,
@@
 where we have defined $q_{\phib}(\xb,\zb) = q_{\phib}(\zb|\xb)\hat{p}(\xb)$ for ease of notation. This is simply \cref{eq:variational-autoencoder-elbo} under the extra expectation. 
-We leave the reconstruction loss in \cref{eq:variational-autoencoder-elbo-factorized} is as and focus on the expectation over the KL-divergence to the prior. We rewrite it as follows,
+We leave the reconstruction loss in \cref{eq:variational-autoencoder-elbo-factorized} as is and focus on the expectation over the KL-divergence to the prior. We rewrite it as follows,
 %

### 2024-01-01 chapters/main-contributions.tex
Polish background

@@
 The discussion points to interesting avenues of future research.
-An extended version of this overview paper that focuses exclusively on self"=supervised methods was also published as part of the project \ref{enumerate: paper-review} \parencite{mohamed_selfsupervised_2022} and this paper is included in \cref{app:paper-review} for reference.
+An extended version of this overview paper that focuses exclusively on self"=supervised methods was also published as part of the project \ref{enumerate: paper-review} \parencite{mohamed_selfsupervised_2022} and is included in \cref{app:paper-review} for reference.
 

### 2024-01-02 chapters/discussion.tex
Polish discussion

@@
 Having examined VAEs and self"=supervised approaches for representation learning in \cref{part:unsupervised-uncertainty-estimation,part:unsupervised-speech-representation-learning}, in this discussion we will consider how VAEs might be enabled to learn representations more competitive with self"=supervised methods for downstream tasks (\cref{sec_discussion:representation-learning-with-vaes}).
-We will also complement the works on OOD detection in \cref{part:unsupervised-uncertainty-estimation}, which focused on image data, with a discussion about similarities with OOD detection for audio and speech data (\cref{sec_discussion:out-of-distribution-detection-for-speech}).
-Furthermore, since the work presented in \cref{part:medical-applications} only had limited focus on uncertainty, we will also present and discuss the use of calibration techniques for the stroke recognition model, drawing connections to how such systems might be perceived and used in practice (\cref{sec: discussion-stroke-recognition-uncertainty}).
+We will also complement the works on OOD detection in \cref{part:unsupervised-uncertainty-estimation}, which focused on image data, with a discussion about the OOD detection for audio and speech data (\cref{sec_discussion:out-of-distribution-detection-for-speech}).
+Since the work presented in \cref{part:medical-applications} only had limited focus on uncertainty, we will also present and discuss the use of calibration techniques for the stroke recognition model, drawing connections to how such systems might be perceived and used in practice (\cref{sec: discussion-stroke-recognition-uncertainty}).
 
@@
 %
-In this thesis we studied two different approaches to speech representation learning: VAEs in \cref{chp:paper-hierarchical,chp:paper-modelagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-brief}.
+In this thesis we studied two different approaches to speech representation learning: VAEs in \cref{chp:paper-hierarchical,chp:paper-modelagnostic,chp:paper-brief,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-brief}.
 As we saw in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}, the probabilistic formulation of VAEs provides benefits for their application to uncertainty quantification, although competitive methods based on self"=supervised foundation models have also been successful \parencite{xiao_we_2021,hendrycks_using_2019,bergman_classificationbased_2020}.
-Nonetheless, as discussed in \cref{chp:paper-brief}, self"=supervised methods are superior to VAEs when it comes to performance on most other downstream tasks, such as speech recognition and spoken language understanding.
+Nonetheless, as discussed in \cref{chp:paper-brief}, self"=supervised methods are superior to VAEs when it comes to performance on most downstream tasks, such as speech recognition and spoken language understanding.
 While this statement of course depends on the task and the amount and type of unlabeled and labeled data, self"=supervised methods for speech are generally evaluated on downstream tasks that are harder and more complex than those used to evaluate VAEs.
@@
 Two key observations might lead to the pessimistic view that VAEs are fundamentally unsuited for representation learning. 
-First, VAEs are trained to maximize the ELBO which includes a marginal likelihood objective that is independent of the latent variables $\zb$.
+First, VAEs are trained to maximize the ELBO which, when tight, is a marginal likelihood objective that is independent of the latent variables $\zb$.
 Second, as we saw in \cref{eq:variational-autoencoder-elbo-factorized-expected}, the ELBO can be written to reveal that its maximization leads to minimization of the mutual information between the observed and latent variables.
@@
 %
-Since the ELBO is not independent of the latent variables, its maximization will typically depend on the latent variables and the posterior $p(\zb|\xb)$ and hence, it is not necessarily a useless criterion for representation learning.
+Since the ELBO is not independent of the latent variables, its maximization will typically depend on the latent variables and the posterior $p(\zb|\xb)$.
 Nonetheless, being a lower bound of the marginal likelihood suggests it could suffer from similar issues under ``right'' conditions.
@@
 We can see that optimizing the ELBO means learning a good one-to-one mapping from $\zb$ to $\xb$ (average reconstruction), fitting the aggregated posterior to the prior (marginal KL-divergence), while also not learning any stochastic dependency between $\xb$ and $\zb$ (mutual information).
-These objectives are conflicting and especially the minimization of mutual information seems counterproductive to our goals \parencite{tomczak_trouble_2022}.
+These objectives are conflicting and especially the minimization of mutual information seems counterproductive to our goal of learning useful representations \parencite{tomczak_trouble_2022}.
 
@@
 Even though these properties suggest that VAEs might not work well for representation learning, VAEs have consistently been shown to be able to learn meaningful representations \parencite{sonderby_ladder_2016,hsu_unsupervised_2017,maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
-We can understand the reason for this by noting that both of the above scenarios can only occur if the optimization is too unconstrained.
+We can understand the reason for this by noting that both of the above scenarios can only occur if the optimization is sufficiently unconstrained.
 
@@
 This introduces a coupling between the marginal $p_{\thetab}(\xb)$ and the posterior $p(\zb|\xb)$ which makes the maximum likelihood solution unlikely to correspond to completely useless representations \parencite{huszar_is_2017}.
+Particularly successful approaches have made this coupling stronger by sharing parameters between the inference and generative models \parencite{sonderby_ladder_2016,maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
 Similarly, maximizing the ELBO %in \cref{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)} 
-by exactly minimizing the mutual information term corresponds to a complete posterior collapse which necessitates a trade-off with the other terms.
+by exactly minimizing the mutual information term would correspond to a complete posterior collapse which necessitates a trade-off with the other terms.
 For a given model, this solution is almost surely a local minimum, and one that can often be avoided in practice.
@@
 
-For VAEs to learn representations more competitive with self"=supervised methods, the fundamental challenge then seems to be how to impose constraints on the model class and optimization problem to ensure a strong coupling between the marginal likelihood and the usefulness of the posterior, and an optimization problem where minimization of the mutual information is very hard, or impossible.
+Following this discussion, we can highlight the main challenges for VAEs to learn representations more competitive with self"=supervised methods.
+We must better understand how to impose constraints on the model class and optimization problem in a structured way in order to 1) ensure a strong coupling between the marginal likelihood and the usefulness of the posterior, and 2) make minimization of the mutual information term very difficult.
+In the following, we will discuss some possible approaches.
 
@@
 Although some work has successfully separated speaker identity from content by modifying the latent variables and their dependencies \parencite{hsu_unsupervised_2017}, models that can learn a deep hierarchy of features for speech remains an open challenge.
-The work presented in \cref{chp:paper-benchmarking} represents an effort to make progress towards such a model.
+The work presented in \cref{chp:paper-benchmarking} represents an effort to make progress towards such a model, but more work remains.
 
 \subsection{Adding a few labels} 
-Another way to improve the representations learned in VAEs is via semi-supervised learning. 
-Here, a few labels are used to inform which patterns that are learned from a large, mostly unlabeled, data set. 
+Another way to improve the representations learned in VAEs is via semi"=supervised learning. 
+Here, a few labels are used to inform which patterns are learned from a large, mostly unlabeled, data set. 
 This is usually done by defining a new stochastic variable as the target and deriving a semi-supervised version of the ELBO that accommodates using the labels when they are available, or marginalizing the target variable when they are not.
@@
 Despite being theoretically appealing, joint objectives for semi-supervised learning in VAEs have practical drawbacks.
-By training on labeled and unlabeled data simultaneously, semi-supervised VAEs are often less flexible than are self"=supervised methods that divide the training into two separate phases.
-By first fitting a general foundation model in an expensive pre"=training phase it can then later be fine"=tuned for many downstream tasks, at relatively low cost.
+By training on labeled and unlabeled data simultaneously, semi-supervised VAEs are often less flexible than self"=supervised methods which divide the training into two separate phases.
+By first fitting a general foundation model in an expensive pre"=training phase that can later be fine"=tuned, self"=supervised foundation models can be fine"=tuned for several downstream tasks, at relatively low cost.
 VAEs, on the other hand, must often learn the labeled downstream tasks while simultaneously training on the unlabeled data.
 This is computationally expensive, but also requires retraining with the unlabeled data, when new supervised data becomes available or a new task is added.
+Improvements in semi-supervised learning with VAEs might lead to more useful representations, but by requiring labels it remains less attractive than fully unsupervised learning.
 
 \subsection{Approximating less}
-VAE are inherently approximate.
+VAEs are inherently approximate.
 Training is performed on a lower bound of the likelihood and inference is variational, amortized and for non-hierarchical models, mean field.
 As a consequence, the gradient itself is estimated inexactly, usually by only one posterior sample, leading to non-negligible variance.
-For that reason, it seems plausible that using tighter bounds and reducing gradient variance improve representation learning in VAEs. 
+For that reason, it seems plausible that using tighter bounds and reducing gradient variance might indirectly yield more useful representations by improving learning in VAEs.
+
 Importance weighting the ELBO \parencite{burda_importance_2016} provides a tight bound on the likelihood, reduces gradient variance, and induces a complex implicit posterior distribution. 
-Its use has become standard when reporting likelihood benchmarks but as we described in \cref{sec:variational-autoencoders}, \textcite{rainforth_tighter_2019} demonstrated that using it during training introduces high gradient variance for the inference network hurting its ability to learn useful representations.
-Later works largely solved this issue \parencite{roeder_sticking_2017,tucker_doubly_2019,bauer_generalized_2021} and showed that the reduced variance leads to improved likelihoods.
+Its use has become standard when reporting likelihood benchmarks but as we described in \cref{sec:variational-autoencoders}, \textcite{rainforth_tighter_2019} demonstrated that using it during training introduces high gradient variance for the inference network, hurting its ability to learn useful representations.
+However, Later works largely solved this issue and showed that the reduced variance leads to improved likelihoods \parencite{roeder_sticking_2017,tucker_doubly_2019,bauer_generalized_2021}.
 
@@
 Instead, these models rely on advanced inference networks \parencite{maaloe_biva_2019} and architectural improvements \parencite{vahdat_nvae_2020, child_very_2021} to overcome the challenges of training hierarchical VAEs.
-This seems to indicate that there might be untapped potential in consolidating the work on low-variance gradient estimators, advanced architectures and inference methods.
+This indicates that there might be untapped potential in consolidating the work on low-variance gradient estimators, advanced architectures and inference methods, which could help improve representation learning in VAEs.
 
 \subsection{Masking} 
-As discussed in \cref{chp:paper-brief}, masking is one of the driving techniques behind the success of self"=supervised methods for speech \parencite{devlin_bert_2018,baevski_wav2vec_2020}.
-By removing tokens from the input or an early feature extraction layer and tasking a model with inferring their representations, models are forced to learn how neighboring tokens relate to those masked.
+As discussed in \cref{chp:paper-brief}, masking is one of the driving forces behind the success of self"=supervised methods for speech \parencite{devlin_bert_2018,baevski_wav2vec_2020}.
+By removing tokens from the input, or an early feature extraction layer, and tasking a model with inferring their representations, models are forced to learn how neighboring tokens relate to those that are masked.
 Depending on the size of the mask, these dependencies can be more or less semantic in nature.
@@
 Furthermore, VAE reconstruction almost always targets the direct input in order to learn the distribution over the training data and to enable generating new samples.
-However, compared to many self"=supervised methods that target intermediate, learned representations, this forces VAEs to encode all aspects of the input that are important to its distribution.
+However, compared to many self"=supervised methods that target intermediate, learned representations, this forces VAEs to encode all aspects of the input that are important to its representation in the observed space.
 As we saw in \cref{chp:paper-hierarchical}, this will generally include low-level features that are necessary for accurate input reconstruction.
@@
 The model is then trained to infer the latent variable from the observed data and reconstruction also deals only with the observed data.
-By comparison to self"=supervised approaches that use masking for representation learning, the missing data setting of VAEs focuses on reconstructing the observed data rather than the missing, which might not lead to the same benefits in representation learning.
-
+By comparison to self"=supervised approaches that use masking for representation learning, VAE training in the missing data setting focuses on reconstructing the observed data rather than the missing, which might not lead to the same benefits in representation learning.
 The idea of using VAEs to impute missing data was already examined in the seminal paper by \textcite{rezende_stochastic_2014}.
@@
 Including masked objectives into the principled probabilistic framework of VAEs might require rethinking the principles themselves.
-Nonetheless, it seems likely that relaxing the requirement of exact input reconstruction and enforcing the need for encoding (and decoding) contextual information could pave the way for future types of VAEs to learn representations that are competitive with self"=supervised methods.
+Nonetheless, relaxing the requirement of exact input reconstruction and enforcing the need for encoding (and decoding) contextual information is one of the strongest candidates to pave the way for future types of VAEs to learn representations that are competitive with self"=supervised methods.
 
@@
 
-In certain cases, it might be possible to classify an image OOD due to the color or intensity of a single pixel. 
+In certain cases, it might be possible to classify an image as OOD based only on the color or intensity of a single pixel. 
 For most distributions, however, we are interested in a test that is sensitive to more semantic properties of the data - for instance the animal class, as discussed in the introduction (\cref{subsec:understanding-uncertainty}).
 For speech, this property holds too; a single amplitude value is unlikely to provide much information about the distributional origin of a speech sample.
-In fact, audio samples are one-dimensional which likely makes them less discriminatory compared to image pixels that generally have multiple channels e.g.
+In fact, audio samples are one-dimensional which likely makes them even less discriminative compared to image pixels which generally have multiple channels e.g.
 for color.
@@
 
-Few studies have explored OOD detection specifically for speech.
-One recent study has explored OOD detection for speech \parencite{bukhsh_outofdistribution_2023}.
-The authors propose a supervised approach that uses a $k$NN classifier fitted to representations of the training data.
+Nevertheless, few studies have explored OOD detection specifically for speech. 
+In one recent study, the authors propose a supervised approach that uses a $k$NN classifier fitted to representations of the training data \parencite{bukhsh_outofdistribution_2023}.
 The approach is tuned to a fixed recall on the training data and operates on full length audio samples.
-This method is similar to image methods that fit a Gaussian mixture to an appropriate layer \parencite{lee_simple_2018, xiao_we_2021}.
+This method is similar to image methods that fit a Gaussian mixture to an appropriate model layer \parencite{lee_simple_2018, xiao_we_2021}.
 In the early work of \textcite{hendrycks_baseline_2017}, the OOD detection method (based on softmax probabilities) is also benchmarked on audio datasets including TIMIT and Aurora-2 \parencite{garofolo_timit_1993,pearce_aurora_2000}.
@@
 % The studies on OOD detection for audio reflect its similarity to OOD detection for images. 
-Despite these similarities of the OOD detection task and its methods across data types, audio comes with unique properties. 
-Where images are immediately interpretable to a researcher by visual inspection, audio requires playback, often in real-time, to present its contents.
+Despite the similarities of the OOD detection task and its methods between data types, audio comes with unique properties. 
+Where images are immediately interpretable to a researcher by visual inspection, audio requires playback, often in real-time, to reveal its contents.
 This fact alone is likely to have reduced interest in audio data by simply making it more time-consuming to work with than images.
-On top of this, audio is generally variable length and requires more memory than images which increases complexity and the amount of compute required to obtain equivalent results compared to images.
-As such, audio-based OOD detection is likely to benefit from standardized frameworks like HuggingFace \parencite{wolf_huggingface_2020,lhoest_datasets_2021} which have made eased working with irregular data types.
+On top of this, audio is generally variable length and requires more computer memory than images which increases complexity and the amount of compute required to obtain equivalent results compared to images.
+As such, audio-based OOD detection is likely to benefit from progress in standardized frameworks for machine learning like PyTorch \parencite{paszke_automatic_2017} and HuggingFace \parencite{wolf_huggingface_2020,lhoest_datasets_2021} which make it increasingly easier to work with irregular data types.
 
 % - Temporal scale, can be OOD in local segments but not entire sequence. 
-Another fundamental difference between audio and images is that audio is additive; two signals can be added together to form a new, mixed signal that is meaningful in the sense that it could also have been recorded.
+Another fundamental difference between audio and images is that audio is additive; two signals can be added together to form a new, mixed signal that is meaningful in the sense that it might as well be a true recording.
 Addition of two images on the other hand produces an image that is not generally representative of what might be photographed. 
 This property of images has led to the practice of evaluating OOD detection performance by comparing examples between different datasets (e.g. MNIST versus FashionMNIST). 
-The additive property of audio allows complementing this dataset-versus-dataset evaluation with an approach where in-distribution examples are augmented to be OOD by mixing with unknown sources.
-This was explored by \textcite{hendrycks_baseline_2017} but has not seen a wide exploration.
+Interestingly, the additive property of audio allows complementing this dataset-versus-dataset evaluation with an approach where in-distribution examples are augmented to be OOD by mixing with unknown sources.
+This was initially explored by \textcite{hendrycks_baseline_2017} but has not seen a wide investigation.
 
-Finally, most work on OOD detection has focused on an example-centric approach where an entire example is either in- or out-of-distribution. 
+Finally, most work on OOD detection has focused on an example-centric approach where an entire example must be classified as either in- or out-of-distribution. 
 For audio, however, it is easy to create examples where local segments are OOD while other parts are not. 
@@
 Similarly, silence might be used as a kind of padding to insert entirely OOD segments in between in-distribution parts without addition and without significant boundary effects. 
-For images, the equivalent approach is similar to regularization techniques such as CutMix \parencite{yun_cutmix_2019} which paste images on top of each other.
-But this usually creates strong boundary errors, which might be used or learned to easily detect such synthetic OOD examples.
+For images, the equivalent approach is similar to the successful regularization technique CutMix \parencite{yun_cutmix_2019} which pastes unrelated images into an example image.
+In the context of OOD detection, this approach creates strong boundary errors, which might be used or learned to easily detect such synthetic OOD images.
 Examples that contain locally OOD parts might occur naturally too; both in the image and audio domains.
@@
 Since the ensemble's output probability is computed as the harmonic mean of the five constituent model probabilities, it can never exceed the maximum probability predicted between the constituent models.
-This property tends to make ensemble probabilities less extreme and, since the constituent models are overconfident, this results in better calibration (see also the histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).
+This property tends to make ensemble probabilities less extreme and, since the constituent models are overconfident, this results in better calibration.
 
@@
 Relative improvement in Brier score compared to uncalibrated (Brier skill score): Logistic = $0.4830$, isotonic = $0.4924$.} and that the predicted probabilities are shifted towards smaller values.
-Since stroke cases have low prevalence, high probability is predicted only for a few examples (see histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).
+Since stroke cases have low prevalence, high probability is predicted only for a few examples (see the histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).
 This leads to a lack of data for the calibration fits at high predicted probabilities which can be seen to result in poor generalization to the test set, especially for the nonparametric isotonic regression.
@@
 With a precision of $24.9\%$ (95\% confidence interval $24.3-25.5\%$) it will on average be wrong three out of four times it predicts a stroke.
-This unfortunately risks alarm fatigue among its potential users diminishing its effect in practice.
+This unfortunately risks alarm fatigue among its potential users, diminishing its effect in practice.
 
-Calibrated probabilities in the alerts presented to users might be a way to alleviate alarm fatigue.
-Calibrated probabilities would allow users to discern between certain and uncertain predictions and also enable the system to present users only with predictions that have a minimum probability of being correct.
-Similar approaches have been suggested by clinicians and interviews indicate that predictive uncertainty is perceived by experts as a sort of explanation that complements the prediction \cite{tonekaboni_what_2019}.
+Calibrated probabilities in the alerts presented to users might be a way to alleviate the problem of alarm fatigue.
+They would allow users to discern between certain and uncertain predictions and also enable the system to present users only with predictions that have a minimum probability of being correct.
+Similar approaches have been suggested by clinicians and interviews indicate that predictive uncertainty is perceived by experts as a kind of explanation that complements the prediction \cite{tonekaboni_what_2019}.
 In \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability} we show the F1-score of the ensemble model and the call-takers computed on subsets of the data created by binning the calibrated probabilities.
 We note that, as might be expected, both call-taker and ensemble model performance increase with increased model certainty.
-This indicates that selecting, based on certainty, which predictions to present to users might indeed help build trust in the system and ensure a practical impact.
+This indicates that selecting, based on certainty, which predictions to present to users might indeed help build trust in the system, reduce alarm fatigue, and ensure a practical impact.

### 2024-01-02 chapters/background.tex
Minor edits and polishing

@@
 
-\section{Variational autoencoders}
+\section{Variational autoencoders}\label{sec:variational-autoencoders}
 %
@@
 
-\subsection{Variational inference} \label{sec:variational-autoencoders}
+\subsection{Variational inference}
 To model high-dimensional data, we want to use neural networks to represent the generative process. 

### 2024-01-02 frontmatter/acknowledgements.tex
Minor edits and polishing

@@
 Your positivity and insightfulness were essential to me daring to take on this project, and to its success. 
-Lars, you have been an indispensable source of purpose, energy, and direction. I look forward to many more fruitful discussions in the future.
+Lars, you have been a completely indispensable source of purpose, energy, and direction throughout my time with Corti. I look forward to many more fruitful discussions in the future.
 
@@
 A special thanks should go to Joakim. Your honest curiosity and dedication has been a wonderful addition to the research team. 
-Finally, I want to thank Lasse.
-Our numerous inspirational discussions and combined efforts through the years have been instrumental in shaping the trajectory of this thesis.
+I also want to thank Lasse.
+The numerous inspirational discussions with you and our combined efforts through the years have been instrumental in shaping this thesis.
 It has been a privilege to carry out this PhD project in the company of you both.
 
-I want to also thank my friends and my family who have provided much needed diversion throughout the years. 
-Last, but not least, I want to thank my wife, Rikke, who has been an unbelievably patient and absolutely indispensable supporter at every step of this undertaking.
+I want to also thank my friends and my family who have provided much needed diversion and encouragement throughout the years. 
+Last, but not least, I want to thank my wife, Rikke, who has been an unbelievably patient and absolutely essential supporter at every step of this undertaking.
+I am excited about what lies ahead for us and our daughter Vigga, and I cannot wait to build our future together.
 

### 2024-01-02 preamble/etc.tex
Make project run in VS Code
Sync-up
Add benchmarking paper
Add review paper (WIP)
Minor updates
Updates
Work on variational autoencoders
Minor edits and polishing

@@
 \usepackage{multicol}
+\usepackage{multirow}
+\usepackage{makecell}
 \usepackage{booktabs}
-\usepackage{multirow}
 \usepackage{enumerate}
 \usepackage{enumitem}
-\usepackage{layouts}
+% \usepackage{layouts}
 \usepackage[showframe=\showframe,xetex]{geometry}
+\usepackage{adjustbox}
+\usepackage{siunitx}
+\usepackage{nomencl}
+\usepackage[acronym,toc]{glossaries}
+
+% Hyphenation
+\hyphenation{ma-skin-læ-rings-mo-del}
+\hyphenation{ma-skin-læ-rings-mo-del-ler}
+\hyphenation{ty-pi-ka-li-tets-tests}
+\hyphenation{ta-le-re-præ-sen-ta-tions-model-ler}
+\hyphenation{me-to-de}
+\hyphenation{me-di-cin-ske}
+
+\DeclareMathOperator*{\argmax}{arg\,max}
+\DeclareMathOperator*{\argmin}{arg\,min}
 
@@
     shapes,
+    shapes.multipart,
     topaths,
@@
 }
+\input{preamble/bayesnet_extras}
+
+\definecolor{customgreen} {RGB}{217	234	212}
+\definecolor{customblue}  {RGB}{205	226	242}
+\definecolor{custommorered}{RGB}{218 46 42}
+\definecolor{customred}{RGB}{255 182 173}%{255 149 150}
+
+\colorlet{observed-color}{customgreen}
+\colorlet{latent-color}{customblue}
+\colorlet{deterministic-color}{gray!15}
+\colorlet{deterministic-skip-color}{custommorered}
+\colorlet{shared-function-color}{blue}
 

### 2024-01-02 chapters/conclusion.tex
Polish conclusion

@@
 The cases form a reference point for the thesis as a whole, connecting its contributions within out-of-distribution detection and representation learning back to practical applications. 
-In this conclusion, we will review the studies presented in previous chapters in the context of the discussion of \cref{chp:discussion}, the recent progress in the field, and point to interesting directions of future research.
+In this conclusion, we will review the studies presented in previous chapters in the context of the discussion of \cref{chp:discussion} and the recent progress in the field, and point to interesting directions of future research.
 As \textbf{\cref{chp:main-contributions}} already details the contributions made by this thesis, we will not reiterate them in detail in this conclusion.
@@
 \vspace{1em}
-\textbf{\Cref{chp:technical-background}} provided technical background that could only be covered briefly by the individual studies. 
+\textbf{\Cref{chp:technical-background}} provided in-depth technical background that could only be covered briefly by the individual studies. 
 We first introduced uncertainty as a concept in the context of information and probability theory (\cref{sec:uncertainty-information-theory}).
@@
 \textbf{\Cref{chp:paper-hierarchical}} showed how hierarchical variational autoencoders can fail at likelihood"-based OOD detection due to an overemphasis on low-level features that generalize between different data distributions.
-In other words, the distributions of latent representations of high-dimensional data overlap, especially for latent variables low in the hierarchy.
+In other words, the latent representations of high-dimensional data from different distributions overlap, especially for latent variables low in the hierarchy.
 By exploiting that VAEs tend to learn more abstract features at latent variables high in the hierarchy, we were able to define a likelihood-ratio score that focused more on features unlikely to be shared between datasets and performed much better for OOD detection.
@@
 We showed that by phrasing OOD detection as a statistical testing problem and combining different tests, orthogonal properties of the individual tests could be leveraged to improve the OOD detection performance over any single test.
-
+%
 The formulation of OOD detection as a statistical test also allows for better guarantees for such systems in practice.
@@
 For instance, in discussion about the weaknesses of VAE-based approaches (\cref{sec_discussion:representation-learning-with-vaes}) we concluded that their challenges could not be directly attributed to the maximum marginal likelihood objective.
-Indeed, the masked pre-training objective widely used for successful self"=supervised methods also corresponds to a maximum marginal likelihood objective \parencite{moreno-munoz_masked_2023}.
+Indeed, the masked pre-training objective widely used for successful self"=supervised methods has also been identified to correspond to a maximum marginal likelihood objective \parencite{moreno-munoz_masked_2023}.
 
 This also leaves potential for future work to improve the ability of VAEs to learn useful representations.
-As discussed in \cref{chp:paper-brief,sec_discussion:representation-learning-with-vaes}, promising approaches include adopting masked objectives for VAE training, improved architectural designs to impose better inductive biases, and incorporating advances in gradient estimators more widely in the literature \parencite{rainforth_tighter_2019,roeder_sticking_2017,tucker_doubly_2019,bauer_generalized_2021}, especially works on large VAE models \parencite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
+As discussed in \cref{chp:paper-brief,sec_discussion:representation-learning-with-vaes}, promising approaches include adopting masked objectives for VAE training, improving architectural designs to impose better inductive biases, and incorporating advances in gradient estimators more widely \parencite{rainforth_tighter_2019,roeder_sticking_2017,tucker_doubly_2019,bauer_generalized_2021}, especially works on large VAE models \parencite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.
 
@@
 The chapter also introduced a novel hierarchical VAE type model for speech, by drawing inspiration from the Clockwork VAE \parencite{saxena_clockwork_2021}. 
-Despite the supposed limitations of VAEs for representation learning as compared to self"=supervised methods (\cref{chp:discussion}), hierarchical VAE models for sequence data that operate on multiple temporal scales, provide a natural framework for encoding distinct feature categories.
-For instance, pronunciation features may be learned at lower layers, speaker identity at upper layers, and semantic features in intermediate layers.
+Despite the limitations of VAEs for representation learning as compared to self"=supervised methods (\cref{chp:discussion}), for sequence data, hierarchical VAE models that operate on multiple temporal scales provide a natural framework for encoding distinct feature categories.
+For instance, pronunciation features might be learned at lower layers, speaker identity at upper layers, and semantic features in intermediate layers.
 Despite successful attempts at isolating speaker identity from content in some existing work \parencite{hsu_unsupervised_2017}, developing VAE models with the capacity to learn a deep hierarchy of features for speech persists.
@@
 \vspace{1em}
-\textbf{\Cref{chp:paper-automated}} examined existing works on automated medical coding and found that, in any cases, training was suboptimal, and evaluation standards were biased.
+\textbf{\Cref{chp:paper-automated}} examined existing works on automated medical coding and found that in many cases training was suboptimal and evaluation standards were biased.
 We performed a revised comparison of the selected models and provided updated conclusions on the relative performance of models, and the impact of rare codes and long discharge note documents.
@@
 Furthermore, the complexity and multi-label nature of medical coding has lead to a high prevalence of label errors in MIMIC-III \parencite{searleExperimentalEvaluationDevelopment2020}.
-While difficult to remove entirely, having multiple diverse datasets would also alleviate the risk of biases in such errors leading to misguided conclusions.
-Besides gathering more data, as also mentioned in \cref{chp:paper-automated}, evaluation methods, such as human-in-the-loop, could be useful to increase the reliability of results.
+While difficult to remove entirely, having multiple diverse datasets would also alleviate the risk that the biases of such errors lead to misguided conclusions.
+Besides gathering more data, improved evaluation methods, such as human-in-the-loop, could be useful to increase the reliability of results.
 
-A weakness of medical coding models is that performance varies widely between classes.
+A common weakness of medical coding models is that performance varies widely between classes.
 The long-tailed distribution of code-frequency is particularly challenging and leads to underperformance on rare codes.
@@
 This selection of cases in practical applications of medical coding could also benefit from research into selective prediction \parencite{geifman_selective_2017}.
-Although pre"=training is an obvious approach to improve performance when little data is available, the improvements observed by using pre"=trained models for medical coding have been limited \parencite{jiDoesMagicBERT2021,gaoLimitationsTransformersClinical2021,michalopoulosICDBigBirdContextualEmbedding2022,pascualBERTbasedAutomaticICD2021,zhangBERTXMLLargeScale2020} compared to the effect of pre"=training in other domains \parencite{mohamed_selfsupervised_2022, linPretrainedTransformersText2021,baevski_wav2vec_2020,devlin_bert_2018,dosovitskiy_image_2021}.
+Although pre"=training is an obvious approach to improve performance when little data is available, compared to the effect of pre"=training in other domains \parencite{mohamed_selfsupervised_2022, linPretrainedTransformersText2021,baevski_wav2vec_2020,devlin_bert_2018,dosovitskiy_image_2021}, the improvements observed by using pre"=trained models for medical coding have been limited \parencite{jiDoesMagicBERT2021,gaoLimitationsTransformersClinical2021,michalopoulosICDBigBirdContextualEmbedding2022,pascualBERTbasedAutomaticICD2021,zhangBERTXMLLargeScale2020}.
 This suggests an untapped potential for future research into more targeted pre"=training for medical coding.
@@
 While our work in \cref{chp:paper-automated} focused on unimodal medical coding models, it is highly likely that future state-of-the-art models will augment discharge summaries with multi-modal inputs such as medical code descriptions, synonyms, and hierarchies \parencite{kimReadAttendCode2021, mullenbachExplainablePredictionMedical2018, vuLabelAttentionModel2020, caoHyperCoreHyperbolicCograph2020, baoMedicalCodePrediction2021, yuanCodeSynonymsMatter2022,caoHyperCoreHyperbolicCograph2020, xieEHRCodingMultiscale2019}.
-A main reason for this is that coding standards are updated regularly.
+This is particularly useful since coding standards are updated regularly. 
 The ICD standard for instance sees revisions and new codes added yearly, with local adoption following national guidelines \parencite{centerfordiseasecontrolandpreventioncdc_international_2023}. 
-Leveraging such modalities will enable adapting models to updated coding standards without having to gather large amounts of data.
+Leveraging such modalities will enable adapting models to updated coding standards without having to gather large amounts of new data.
 
@@
 
-In \cref{sec: discussion-stroke-recognition-uncertainty}, we discussed how calibrating the predictive uncertainty of the stroke model is likely to be necessary to enable sustained use of such a model in practice.
-In \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability}, we noted that, as expected, model performance measured by F1-score increased with increasing model certainty which underlines the likely usefulness of uncertainty estimates in matching practitioner expectations of the predictive performance.
+In \cref{sec: discussion-stroke-recognition-uncertainty}, we discussed how calibrating the predictive uncertainty of the stroke model is likely to be necessary to ensure sustained use of such a model in practice.
+In \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability}, we noted that, as expected, model performance measured by F1-score increased with increasing model certainty which underlines the likely usefulness of uncertainty estimates in better matching practitioner expectations of the predictive performance.
 Nonetheless, basic metrics of model performance might still be obstacles for its practical usefulness.
-Specifically, the rarity of stroke cases lead to a false positive rate and precision that are likely to induce alarm fatigue among its users.
-Similar effects are likely to have influenced the practical impact of a similar system for cardiac arrest detection which showed significant improvements retrospectively \parencite{cite14} which, although matched by the model in a prospective study, did not ultimately result in improved call-taker performance \parencite{cite15}.
+Specifically, the rarity of stroke cases lead to relatively high false positive rate and low precision, likely to induce alarm fatigue among its users.
+Similar effects are likely to have influenced the practical impact of a similar system for cardiac arrest detection which also showed significant improvements in a retrospective study \parencite{cite14}.
+Although the model later matched the retrospective results in a prospective study, it did not ultimately result in improved call-taker performance \parencite{cite15}.
 

### 2024-01-02 chapters/conclusion.tex
Final updates

@@
 Even so, directly predicting the diagnosis from the conversation is not the only path towards practical impact.
-By suggesting informative questions to the medical professional, a system could help guide the conversation to avoid missing important details, and to improve the precision of the model.
+By suggesting informative questions to the medical professional, a system could also help guide the course of conversation to avoid missing important details and keep an overview, and to in turn improve the performance of the model.
+Ultimately, machine learning has proven capable of contributing meaningfully to medical conversations aiming to improve patient outcomes.
+Future work seems poised to make a significant positive impact on the healthcare industry over the years to come.
 

### 2024-01-02 chapters/introduction.tex
Final updates

@@
 In a likely attempt to ease the strenuous work of doing Sumerian mathematics, Kushim produced a clay tablet with a multiplication table that included fractional representations of common numbers (\cref{fig:kushim_multiplication}).
-Assumedly unbeknownst to Kushim, this tablet would also come to contain one of the world's first recorded examples of a mathematical error; the statement: $5 \times \sfrac{1}{2} = 5$ \parencite{nissen_archaic_1993}.
+Assumedly unbeknownst to Kushim, this tablet would also come to contain the world's first recorded example of a mathematical error; the statement: $5 \times \sfrac{1}{2} = 5$ \parencite{nissen_archaic_1993}.
 Later, the scribe of the Plimpton 322 clay tablet also made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics \parencite{neugebauer_mathematical_1945,britton_plimpton_2011,cuneiformdigitallibraryinitiativecdli_mct_2005}.
@@
     \caption[Technology is difficult.
-The accountant Kushim made mistakenly claimed $5 \times \sfrac{1}{2} = 5$.]{ Technology is difficult.
+The Sumerian accountant Kushim mistakenly added $5 \times \sfrac{1}{2} = 5$ to their multiplication table.]{ Technology is difficult.
 Between 3400 and 3000 BCE, the Sumerian accountant Kushim wrote this tablet.
-Besides calculations of basic ingredients required for the production of cereal products, in the top left $6 \times 2$ grid it contains a multiplication table.
-While presumably aimed to ease the strenuous work of doing Sumerian mathematics, unfortunately, the top row states: $5 \times \sfrac{1}{2} = 5$, an unforgiving imprint the worlds first recorded mathematical error \parencite{nissen_archaic_1993, renn_learning_2019}, \parencite[photo credit][]{cuneiformdigitallibraryinitiativecdli_msvo_2018}.}
+Besides calculations of basic ingredients required for the production of cereal products, in the top left 6-by-2 grid it contains a multiplication table.
+While presumably aimed to ease the strenuous work of doing Sumerian mathematics, the top row unfortunately states: $5 \times \sfrac{1}{2} = 5$, an unforgiving imprint of the world's first recorded mathematical error \parencite{nissen_archaic_1993, renn_learning_2019}, \parencite[photo credit][]{cuneiformdigitallibraryinitiativecdli_msvo_2018}.}
     \label{fig:kushim_multiplication}

### 2024-01-02 chapters/conclusion.tex
Final updates

@@
 
-\textbf{\Cref{chp:introduction}} introduced the motivational cases of automated medical coding and stroke recognition and used them to exemplify the importance of out-of-distribution detection, and, by extension, representation learning.
+\textbf{\Cref{chp:introduction}} introduced the motivational cases of automated medical coding and stroke recognition and used them to exemplify the importance of out"=of"=distribution detection, and, by extension, representation learning.
 In the context of these cases, we discussed possible machine learning system designs for decision support and considered potential sources of uncertainty and ideal model behavior.
-The cases form a reference point for the thesis as a whole, connecting its contributions within out-of-distribution detection and representation learning back to practical applications. 
+The cases form a reference point for the thesis as a whole, connecting its contributions within out"=of"=distribution detection and representation learning back to practical applications. 
 In this conclusion, we will review the studies presented in previous chapters in the context of the discussion of \cref{chp:discussion} and the recent progress in the field, and point to interesting directions of future research.

### 2024-01-02 chapters/paper_review/03_approaches.tex
Fix commas after equations
Final updates

@@
 \begin{gather}
-    H_{[1,t]} = f_1(X_{[1,t]}), \\
-    H^\prime_{[t+k,T]} = f_2(X_{[t+k,T]}), \\
+    H_{[1,t]} = f_1(X_{[1,t]}) \enspace , \\
+    H^\prime_{[t+k,T]} = f_2(X_{[t+k,T]}) \enspace , \\
     \hat{X}_{[t+1,t+k-1]} = g(h_t, h^\prime_{t+k}) \enspace .
@@
 \begin{align}
-    H &= f(m(X)), \\
-    \hat{x}_t &= g(h_{t}), \\
-    \mathcal{L}_t &= \lVert \hat{x}_{t} - x_{t} \rVert_1 .
+    H &= f(m(X)) \enspace , \\
+    \hat{x}_t &= g(h_{t}) \enspace , \\
+    \mathcal{L}_t &= \lVert \hat{x}_{t} - x_{t} \rVert_1 \enspace .
 \end{align}
@@
 However, for speech, masking a single \edit{sample or spectrogram frame results in a largely} trivial reconstruction task \edit{since, as discussed in %\cref{par:apc}
-the paragraph on autoregressive prediction, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore it is common to mask chunks of consecutive frames~\parencite{liu_mockingjay_2020,jiang_further_2021}.
+the paragraph on autoregressive prediction, the smoothness of audio signals may encourage the model to learn to simply interpolate neighboring frames.} Therefore, it is common to mask chunks of consecutive frames~\parencite{liu_mockingjay_2020,jiang_further_2021}.
 
@@
 Some studies explore alternatives to masking the input directly.
-In non"autoregressive predictive coding (NPC)~\parencite{liu_nonautoregressive_2020}, time masking is introduced through masked convolution blocks.
-Taking inspiration from XLNet~\parencite{yang_xlnet_2019}, it has also been suggested that the input be reconstructed from a shuffled version~\parencite{song_speechxlnet_2020} to address the discrepancy between pre"=training and fine"=tuning of masking-based approaches.
+In non"=autoregressive predictive coding (NPC)~\parencite{liu_nonautoregressive_2020}, time masking is introduced through masked convolution blocks.
+Taking inspiration from XLNet~\parencite{yang_xlnet_2019}, it has also been suggested that the input be reconstructed from a shuffled version~\parencite{song_speechxlnet_2020} to address the discrepancy between pre"=training and fine"=tuning of masking"=based approaches.
 

### 2024-02-12 presentation/departments.tex
Start presentation

% departments.tex
% This file is a part of the DTU letter package and contains the file path for
% the grahic file, and text name for the different departments.
%
% Changelog
% 2010-04-07 Added % at the end of each line to make it possible to use the definitions in the documentation
% 2010-04-09 Added the 5th mandatory argument (long text name)
% 2010-04-23 Moved the new argument, #6, to #4 and added all info. However two graphic files are missing and I have therefore made a test: if the graphic file is missing the administration logo is used.  
% 2013-02-11 Added compute and diplom department entries.  Added check to see that the department macros are defined, otherwise an error is printed.
% 
%\makeDepartmentInfo{<danish|english>}dtu_dep_name_logo/{<dedtu_dep_logo/partmentname>}{<dtu_frise/text graphic file name>}{<Big department logo file name>}{<department text name>}{<department long text name>}
%

\RequirePackage{etoolbox}

\newcommand\setDepartmentNameLogo[1]{\def\@departmentNameLogo{#1}}%
\newcommand\thedepartmentNameLogo{\@departmentNameLogo}%
\newcommand\setDepartmentNameText[1]{\def\@departmentNameText{#1}}%
\newcommand\thedepartmentNameText{\@departmentNameText}%
\newcommand\setDepartmentLongNameText[1]{\def\@departmentLongNameText{#1}}%
\newcommand\thedepartmentLongNameText{\@departmentLongNameText}%
\newcommand\setDepartmentLogo[1]{\def\@departmentLogo{#1}}%
\newcommand\thedepartmentLogo{\@departmentLogo}%
\newcommand\setDepartmentFrise[1]{\def\@departmentFrise{#1}}%
\newcommand\thedepartmentFrise{\@departmentFrise}%
%
\newcommand\createDepartment[1]{%
\expandafter\def\csname department@#1\endcsname{#1}}%
%
\newcommand\aliasDepartment[2]{%
\expandafter\def\csname department@#2\endcsname{#1}}%
%
\ifundef{\makeDepartmentInfo}{%
	\newcommand\makeDepartmentInfo[7]{%
		\def\@departmentcmd{\csname department@#2\endcsname}
		\createDepartment{#2}
		\expandafter\def\csname namelogo#1@\@departmentcmd\endcsname{\setDepartmentNameLogo{#3}}%
		\expandafter\def\csname deplogo#1@\@departmentcmd\endcsname{\setDepartmentLogo{#4}}%
		\expandafter\def\csname depfrise#1@\@departmentcmd\endcsname{\setDepartmentFrise{#5}}%
		\expandafter\def\csname nametext#1@\@departmentcmd\endcsname{\setDepartmentNameText{#6}}%
		\expandafter\def\csname namelongtext#1@\@departmentcmd\endcsname{\setDepartmentLongNameText{#7}}%

	}%
}{}%
%
\newcommand\activateDepartmentInfo[2]{%
	\ifcsname department@#2\endcsname%
		\def\@departmentcmd{\csname department@#2\endcsname}%
	\else%
		\def\@departmentcmd{\department@admin}%
	\fi%
	\csname namelogo#1@\@departmentcmd\endcsname% TODO test if command exists before executing it
	\csname nametext#1@\@departmentcmd\endcsname%
	\csname namelongtext#1@\@departmentcmd\endcsname%
	\csname deplogo#1@\@departmentcmd\endcsname%
	\csname depfrise#1@\@departmentcmd\endcsname%
	% \fromdepartment{\thedepartmentLongNameText}
}%
%
\makeDepartmentInfo{danish}{aqua}{dtu_dep_name_logo/tex_aqua_dk}{dtu_dep_logo/tex_dtu_aqua_a}{dtu_frise/tex_dtu_aqua_frise}{DTU Aqua}{Institut for Akvatiske Ressourcer}%
\makeDepartmentInfo{english}{aqua}{dtu_dep_name_logo/tex_aqua_uk}{dtu_dep_logo/tex_dtu_aqua_a_uk}{dtu_frise/tex_dtu_aqua_frise}{DTU Aqua}{National Institute of Aquatic Resources}%

\makeDepartmentInfo{danish}{byg}{dtu_dep_name_logo/tex_byg_dk}{dtu_dep_logo/tex_dtu_byg_a}{dtu_frise/tex_dtu_byg_frise}{DTU Byg}{Institut for Byggeri og Anl\ae g}%
\makeDepartmentInfo{english}{byg}{dtu_dep_name_logo/tex_byg_uk}{dtu_dep_logo/tex_dtu_byg_a_uk}{dtu_frise/tex_dtu_byg_frise}{DTU Civil Engineering}{Department of Civil Engineering}%

\makeDepartmentInfo{danish}{compute}{dtu_dep_name_logo/tex_compute_uk}{dtu_dep_logo/tex_dtu_compute_a}{dtu_frise/tex_dtu_frise}{DTU Compute}{Institut for Matematik og Computer Science}
\makeDepartmentInfo{english}{compute}{dtu_dep_name_logo/tex_compute_uk}{dtu_dep_logo/tex_dtu_compute_a_uk}{dtu_frise/tex_dtu_frise}{DTU Compute}{Department of Applied Mathematics and Computer Science}

\makeDepartmentInfo{danish}{elektro}{dtu_dep_name_logo/tex_elektro_dk}{dtu_dep_logo/tex_dtu_elektro_a}{dtu_frise/tex_dtu_frise}{DTU Elektro}{Institut for Elektroteknologi}
\makeDepartmentInfo{english}{elektro}{dtu_dep_name_logo/tex_elektro_uk}{dtu_dep_logo/tex_dtu_elektro_a_uk}{dtu_frise/tex_dtu_frise}{DTU Electrical Engineering}{Department of Electrical Engineering}

\makeDepartmentInfo{danish}{energi}{dtu_dep_name_logo/tex_energikonvertering_dk}{dtu_dep_logo/tex_dtu_energi_a}{dtu_frise/tex_dtu_energi_frise}{DTU Energi}{Institut for Energikonvertering og -lagring}
\makeDepartmentInfo{english}{energi}{dtu_dep_name_logo/tex_energikonvertering_uk}{dtu_dep_logo/tex_dtu_energi_a_uk}{dtu_frise/tex_dtu_energi_frise}{DTU Energy}{Department of Energy Conversion and Storage}

\makeDepartmentInfo{danish}{fotonik}{dtu_dep_name_logo/tex_fotonik_dk}{dtu_dep_logo/tex_dtu_fotonik_a}{dtu_frise/tex_dtu_frise}{DTU Fotonik}{Institut for Fotonik}
\makeDepartmentInfo{english}{fotonik}{dtu_dep_name_logo/tex_fotonik_uk}{dtu_dep_logo/tex_dtu_fotonik_a_uk}{dtu_frise/tex_dtu_frise}{DTU Fotonik}{Department of Photonics Engineering}

\makeDepartmentInfo{danish}{fysik}{dtu_dep_name_logo/tex_fysik_dk}{dtu_dep_logo/tex_dtu_fysik_a}{dtu_frise/tex_dtu_fysik_frise}{DTU Fysik}{Institut for Fysik}
\makeDepartmentInfo{english}{fysik}{dtu_dep_name_logo/tex_fysik_uk}{dtu_dep_logo/tex_dtu_fysik_a_uk}{dtu_frise/tex_dtu_fysik_frise}{DTU Physics}{Department of Physics}

\makeDepartmentInfo{danish}{food}{dtu_dep_name_logo/tex_fodevareinstituttet_dk}{dtu_dep_logo/tex_dtu_fdevareinstituttet_a}{dtu_frise/tex_dtu_frise}{DTU F\o devareinstituttet}{F\o devareinstituttet}
\makeDepartmentInfo{english}{food}{dtu_dep_name_logo/tex_fodevareinstituttet_uk}{dtu_dep_logo/tex_dtu_fdevareinstituttet_a_uk}{dtu_frise/tex_dtu_frise}{DTU Food}{National Food Institute}

\makeDepartmentInfo{danish}{kemi}{dtu_dep_name_logo/tex_kemi_dk}{dtu_dep_logo/tex_dtu_kemi_a}{dtu_frise/tex_dtu_kemi_frise}{DTU Kemi}{Institut for Kemi}
\makeDepartmentInfo{english}{kemi}{dtu_dep_name_logo/tex_kemi_uk}{dtu_dep_logo/tex_dtu_kemi_a_uk}{dtu_frise/tex_dtu_kemi_frise}{DTU Chemistry}{Department of Chemistry}

\makeDepartmentInfo{danish}{kemiteknik}{dtu_dep_name_logo/tex_kemiteknik_dk}{dtu_dep_logo/tex_dtu_kemiteknik_a}{dtu_frise/tex_dtu_kemiteknik_frise}{DTU Kemiteknik}{Institut for Kemiteknik}
\makeDepartmentInfo{english}{kemiteknik}{dtu_dep_name_logo/tex_kemiteknik_uk}{dtu_dep_logo/tex_dtu_kemiteknik_a_uk}{dtu_frise/tex_dtu_kemiteknik_frise}{DTU Chemical Engineering}{Department of Chemical and Biochemical Engineering}

\makeDepartmentInfo{danish}{management}{dtu_dep_name_logo/tex_management_dk}{dtu_dep_logo/tex_dtu_management_a}{dtu_frise/tex_dtu_frise}{DTU Management}{Institut for Systemer, Produktion og Ledelse}
\makeDepartmentInfo{english}{management}{dtu_dep_name_logo/tex_management_uk}{dtu_dep_logo/tex_dtu_management_a_uk}{dtu_frise/tex_dtu_frise}{DTU Management Engineering}{Department of Management Engineering}

\makeDepartmentInfo{danish}{mekanik}{dtu_dep_name_logo/tex_mekanik_dk}{dtu_dep_logo/tex_dtu_mekanik_a}{dtu_frise/tex_dtu_mek_frise}{DTU Mekanik}{Institut for Mekanisk Teknologi}
\makeDepartmentInfo{english}{mekanik}{dtu_dep_name_logo/tex_mekanik_uk}{dtu_dep_logo/tex_dtu_mekanik_a_uk}{dtu_frise/tex_dtu_mek_frise}{DTU Mechanical Engineering}{Department of Mechanical Engineering}

\makeDepartmentInfo{danish}{miljo}{dtu_dep_name_logo/tex_miljo_dk}{dtu_dep_logo/tex_dtu_milj_a}{dtu_frise/tex_dtu_miljoe_frise}{DTU Milj\o}{Institut for Vand og Milj\o teknologi}
\makeDepartmentInfo{english}{miljo}{dtu_dep_name_logo/tex_miljo_uk}{dtu_dep_logo/tex_dtu_milj_a_uk}{dtu_frise/tex_dtu_miljoe_frise}{DTU Environment}{Department of Environmental Engineering}

\makeDepartmentInfo{danish}{nanotek}{dtu_dep_name_logo/tex_nanotek_dk}{dtu_dep_logo/tex_dtu_nanotek_a}{dtu_frise/tex_dtu_frise}{DTU Nanotek}{Institut for Mikro- og Nanoteknologi}
\makeDepartmentInfo{english}{nanotek}{dtu_dep_name_logo/tex_nanotek_uk}{dtu_dep_logo/tex_dtu_nanotek_a_uk}{dtu_frise/tex_dtu_frise}{DTU Nanotech}{Department of Micro- and Nanotechnology}

\makeDepartmentInfo{danish}{space}{dtu_dep_name_logo/tex_space_dk}{dtu_dep_logo/tex_dtu_space_a}{dtu_frise/tex_dtu_space_frise}{DTU Space}{Institut for Rumforskning og Rumteknologi}
\makeDepartmentInfo{english}{space}{dtu_dep_name_logo/tex_space_uk}{dtu_dep_logo/tex_dtu_space_a_uk}{dtu_frise/tex_dtu_space_frise}{DTU Space}{National Space Institute}

\makeDepartmentInfo{danish}{systembiologi}{dtu_dep_name_logo/}{tedtu_dep_logo/x_dtu_systembiologi_a}{dtu_frise/tex_dtu_frise}{DTU Systembiologi}{Institut for Systembiologi}
\makeDepartmentInfo{english}{systembiologi}{dtu_dep_name_logo/}{tedtu_dep_logo/x_dtu_systembiologi_a_uk}{dtu_frise/tex_dtu_frise}{DTU Systems Biology}{Department of Systems Biology}

\makeDepartmentInfo{danish}{transport}{dtu_dep_name_logo/tex_transport_dk}{dtu_dep_logo/tex_dtu_transport_a}{dtu_frise/tex_dtu_transport_frise}{DTU Transport}{Institut for Transport}
\makeDepartmentInfo{english}{transport}{dtu_dep_name_logo/tex_transport_uk}{dtu_dep_logo/tex_dtu_transport_a_uk}{dtu_frise/tex_dtu_transport_frise}{DTU Transport}{Department of Transport}

\makeDepartmentInfo{danish}{vaterinaerinstituttet}{dtu_dep_name_logo/tex_veterinaertinstituttet_dk}{dtu_dep_logo/tex_dtu_veterinerinstituttet_a}{dtu_frise/tex_dtu_vet_frise}{DTU Veterin\ae rinstituttet}{Veterin\ae rinstituttet}
\makeDepartmentInfo{english}{vaterinaerinstituttet}{dtu_dep_name_logo/tex_veterinaertinstituttet_uk}{dtu_dep_logo/tex_dtu_veterinerinstituttet_a_uk}{dtu_frise/tex_dtu_vet_frise}{DTU Vet}{National Veterinary Institute}

\makeDepartmentInfo{danish}{vindenergi}{dtu_dep_name_logo/tex_vindenergi_dk}{dtu_dep_logo/tex_dtu_vindenergi_a}{dtu_frise/tex_dtu_vindenergi_frise}{DTU Vindenergi}{Institut for Vindenergi}
\makeDepartmentInfo{english}{vindenergi}{dtu_dep_name_logo/tex_vindenergi_uk}{dtu_dep_logo/tex_dtu_vindenergi_a_uk}{dtu_frise/tex_dtu_vindenergi_frise}{DTU Wind Energy}{Department of Wind Energy}

% Extra
\makeDepartmentInfo{danish}{bibliotek}{dtu_dep_name_logo/tex_bibliotek_dk}{dtu_dep_logo/tex_dtu_bibliotek_a}{dtu_frise/tex_dtu_bibliotek_frise}{DTU Bibliotek}{Danmarks Tekniske Informationcenter}%
\makeDepartmentInfo{english}{bibliotek}{dtu_dep_name_logo/tex_bibliotek_uk}{dtu_dep_logo/tex_dtu_bibliotek_uk_a}{dtu_frise/tex_dtu_bibliotek_frise}{DTU Library}{Technical Information Center of Denmark}%

\makeDepartmentInfo{danish}{admin}{dtu_dep_name_logo/tex_dtu_navn_dk}{dtu_dep_logo/}{tex_dtu_frisedtu_frise/}{Danmarks Tekniske Universitet}{}%
\makeDepartmentInfo{english}{admin}{dtu_dep_name_logo/tex_dtu_navn_uk}{dtu_dep_logo/}{tex_dtu_frisedtu_frise/}{Technical University of Denmark}{}%

\makeDepartmentInfo{danish}{riso}{dtu_dep_name_logo/tex_riso_dk}{dtu_dep_logo/tex_ris_dtu_a}{dtu_frise/tex_dtu_frise}{Ris\o\ DTU}{Nationallaboratoriet for B\ae redygtig Energi}
\makeDepartmentInfo{english}{riso}{dtu_dep_name_logo/tex_riso_uk}{dtu_dep_logo/tex_ris_dtu_a_uk}{dtu_frise/tex_dtu_frise}{Ris\o\ DTU}{National Laboratory for Sustainable Energy}

\makeDepartmentInfo{danish}{diplom}{dtu_dep_name_logo/tex_diplom_dk}{dtu_dep_logo/tex_dtu_diplom_a}{dtu_frise/tex_dtu_frise}{Center for Diplomingeni\o ruddannelse}{DTU Diplom}
\makeDepartmentInfo{english}{diplom}{dtu_dep_name_logo/tex_diplom_dk}{dtu_dep_logo/tex_dtu_diplom_a_uk}{dtu_frise/tex_dtu_frise}{Center for Diplomingeni\o ruddannelse}{DTU Diplom}

### 2024-02-12 presentation/parts/introduction.tex
Start presentation

% !TEX root = ../presentation.tex
% !BIB program = biber
% !TEX program = xelatex

### 2024-02-12 presentation/parts/paper_hierarchical.tex
Start presentation

% !TEX root = ../presentation.tex
% !BIB program = biber
% !TEX program = xelatex

\section{Out-of-distribution detection}

\frame{
    \frametitle{Defining OOD detection}
    \begin{columns}

    \begin{column}{0.5\textwidth}

    Out-of-distribution (OOD) detection is about enabling models to distinguish the training data distribution $p(\xb)$ from any other distribution $\tilde{p}(\xb)$.

    \vspace{3mm}

    We are concerned with doing this on a per-observation basis, i.e. answering the question:

    \vspace{3mm}

    \begin{center}
        "Was $\xb$ sampled from $p(\xb)$ or not?"
    \end{center}

    \vspace{3mm}

    % It is related to \textbf{anomaly detection} and \textbf{adversarial example detection}.
    \end{column}

    \begin{column}{0.5\textwidth}

    \begin{figure}[\textwidth]
        \centering
        \begin{tikzpicture}
        \begin{axis}[
          domain=0:13,samples=100,
          height=4.5cm, width=8cm,
          xtick=\empty,
          every axis plot/.append style={very thick,smooth}
        ]

        \addplot [name path=p1, color=red!50!black] {gauss(5,1)};
        \addplot [name path=p2, color=black] {gauss(8,1)};
        \path[name path=axis] (axis cs:0,0) -- (axis cs:10,0);
        \addlegendentry{$p(\xb)$}
        \addlegendentry{$\tilde{p}(\xb)$}

        \end{axis}
        \end{tikzpicture}
    \end{figure}
    \begin{figure}[\textwidth]
        \centering
        \begin{tikzpicture}
        \begin{axis}[
          domain=0:13,samples=100,
          height=4.5cm, width=8cm,
          every axis plot/.append style={very thick,smooth}
        ]

        \addplot [name path=p1, color=red!50!black] {gauss(4,0.5)};
        \addplot [name path=p2, color=black] {gauss(9,0.5)};
        \path[name path=axis] (axis cs:0,0) -- (axis cs:10,0);
        \addlegendentry{$p(\xb)$}
        \addlegendentry{$\tilde{p}(\xb)$}

        \end{axis}
        \end{tikzpicture}
    \end{figure}
    \end{column}

    \end{columns}
}

\frame{
    \frametitle{Problem and Contributions}
    \begin{itemize}
        \item Deep generative models often fail at OOD detection task when using their likelihood estimate as the score function \cite{nalisnick_deep_2019} by, perhaps surprisingly, assigning \textbf{higher likelihoods} to the OOD data.
        \item Contributions:
        \begin{itemize}
            \item We present a fast and fully unsupervised method for OOD detection competitive with the state-of-the-art
            \item We provide evidence that out-of-distribution detection fails due to learned low-level features that generalize across datasets.
        \end{itemize}  
    \end{itemize}
}

\frame{
    \frametitle{In distribution?}
    \begin{figure}[\textwidth]
        \centering
        \includegraphics[scale=0.6]{figures/dataexamples_shuffled.pdf}
    \end{figure}
}

\frame{
    \frametitle{Out of distribution?}
    \begin{figure}[\textwidth]
        \centering
        \includegraphics[scale=0.6]{figures/dataexamples.pdf}
    \end{figure}
}

\section{Models}

\begin{frame}{Hierarchical VAE}
  \begin{columns}
      \begin{column}{0.7\textwidth}
          We choose the hierarchical VAE as our model \cite{kingma_autoencoding_2014, rezende_stochastic_2014}.
          \begin{equation*}
              p_\theta(\xb) = \int p_\theta(\xb,\zb) \text{d}\zb = \int p_\theta(\xb|\zb)p_\theta(\zb) \text{d}\zb
          \end{equation*}
          Specifically we use

          \begin{enumerate}
              \item a three-layered hierarchical VAE with bottom-up inference and deterministic skip-connections for both inference and generation.
              \begin{align*}
                  \text{Generative model: }\quad p_\theta(\xb|\zb) &= p_\theta(\xb|\zb_1) p_\theta(\zb_1|\zb_2) p(\zb_3),\\
                  \text{Inference model: }\quad q_\phi(\zb|\xb) &= q_\phi(\zb_1|\xb)q_\phi(\zb_2|\zb_1)q_\phi(\zb_3|\zb_2).
              \end{align*}
              \item a ten-layered layered Bidirectional-Inference Variational Autoencoder (BIVA) \cite{maaloe_biva_2019}.
          \end{enumerate} 
      \end{column}
      \begin{column}{0.3\textwidth}
          \begin{figure}[.5\textwidth]
          \tikz{
          % inference
          % nodes
          \node[obs] (x_inf) {$\xb$};%
          \node[latent,above=.75cm of x_inf](z1_inf){$\zb_1$}; %
          \node[latent,above=.75cm of z1_inf](z2_inf){$\zb_2$}; %
          \node[latent,above=.75cm of z2_inf](z3_inf){$\zb_3$}; %
          \node[above=of z3_inf, yshift=-1.cm] (phi) {$q_\phi(\zb|\xb)$};

          % edges
          \edge[]{x_inf}{z1_inf};
          \edge[]{z1_inf}{z2_inf};
          \edge[]{z2_inf}{z3_inf};
          \edge[dashed, bend left]{x_inf}{z2_inf};
          \edge[dashed, bend left]{x_inf}{z3_inf};

          % generative
          % nodes$
          \node[obs,right=0.75cm of x_inf] (x_gen) {$\xb$};%
          \node[latent,above=.75cm of x_gen](z1_gen){$\zb_1$}; %
          \node[latent,above=.75cm of z1_gen](z2_gen){$\zb_2$}; %
          \node[latent,above=.75cm of z2_gen](z3_gen){$\zb_3$}; %
          \node[above=of z3_gen, yshift=-1.cm] (theta) {$p_\theta(\xb,\zb)$};

          % edges
          \edge[]{z3_gen}{z2_gen};
          \edge[]{z2_gen}{z1_gen};
          \edge[]{z1_gen}{x_gen};
          \edge[dashed, bend left]{z2_gen}{x_gen};
          \edge[dashed, bend left]{z3_gen}{x_gen};
          }
          \end{figure}
      \end{column}
  \end{columns}
\end{frame}

\section{The Problem}

\frame{
    \frametitle{What is wrong with the ELBO for OOD detection?}
    We can split the ELBO into two terms
    \begin{equation}
        \mathcal{L}(\xb;\theta,\phi) 
        = \mathbb{E}_{q_\phi(\zb|\xb)} \left[ \log \frac{p_\theta(\xb,\zb)}{q_\phi(\zb|\xb)} \right] 
        = \underbrace{\mathbb{E}_{q_\phi(\zb|\xb)}[\log p_\theta(\xb|\zb)]}_{\text{reconstruction likelihood}} - \underbrace{D_{\mathrm{KL}}( q_\phi(\zb|\xb) || p(\zb))}_{\text{regularization penalty}} \ .
    \end{equation}
    The first term is high if the data is well-explained by $\zb$.

    \vspace{3mm}
    The second term we can rewrite as,
    \begin{equation}
        D_{\mathrm{KL}}( q_\phi(\zb|\xb) || p(\zb)) = \mathbb{E}_{q_\phi(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\theta(\zb_i|\zb_{i+1})}{q_\phi(\zb_i|\zb_{i-1})} + \log \frac{p_\theta(\zb_L)}{q_\phi(\zb_L|\zb_{L-1})} \Big] \ .
    \end{equation}
    The absolute log-ratios grow with $\mathrm{dim}(\zb_i)$ since the log probability terms are computed by summing over the dimensionality of $\zb_i$.

    \vspace{3mm}
}

\frame{
    \frametitle{What do the lowest latent variables code for?}

    Absolute Pearson correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST. 
    \vspace{0.3cm}

    Correlation computed between the representations of the two different models given the same data, FashionMNIST (top) and MNIST (bottom).

    \begin{figure}[\textwidth]
        \centering
        \includegraphics[scale=0.75]{../graphics/paper_hierarchical/feature-correlation-heatmap2.pdf}
        % \caption{Caption}
        % \label{fig:my_label}
    \end{figure}
}

\section{The $\mathcal{L}^{>k}$ likelihood bound}

\frame{
    \frametitle{An alternative likelihood bound, $\mathcal{L}^{>k}$}
    An alternative version of the ELBO that only partially uses the approximate posterior can be written as \cite{maaloe_biva_2019}
    \begin{equation}
        \mathcal{L}^{>k}(\xb; \theta, \phi) = \mathbb{E}_{p_\theta(\zb_{\leq k}|\zb{>k})q_\phi(\zb_{>k}|\xb)} \left[ \log \frac{p_\theta(\xb|\zb)p_\theta(\zb_{>k})}{q_\phi(\zb_{>k}|\xb)} \right]
    \end{equation}

    Here, we have replaced the approximate posterior $q_\phi(\zb|\xb)$ with a different proposal distribution that combines part of the approximate posterior with the conditional prior, namely

    $$p_\theta(\zb_{\leq k}|\zb_{>k})q_\phi(\zb_{>k}|\xb)$$

    This bound uses the conditional prior for the lowest latent variables in the hierarchy.
}

\section{Likelihood ratio}

\frame{
    \frametitle{Likelihood ratios}
    We can use our new bound to compute the score used in a standard likelihood ratio test \cite{buse_likelihood_1982}.
    \begin{equation}\label{eq:llr-as-difference-in-likelihoods}
        LLR^{>k}(\xb) \equiv \mathcal{L}(\xb) - \mathcal{L}^{>k}(\xb) \ .
    \end{equation}
    We can inspect what this likelihood-ratio measures by considering the exact form of our bounds.
    \begin{align}
        \mathcal{L}      &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) || p_\theta(\zb|\xb)\right), \label{eq:likelihoods-as-exact} \\ 
        \mathcal{L}^{>k} &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) || p_\theta(\zb|\xb)\right) \notag \ .
    \end{align}
    In the likelihood ratio the reconstruction terms cancel out and only the KL-divergences from the approximate to the true posterior remain.
    \begin{align}\label{eq:llr-as-kls}
        LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) || p_\theta(\zb|\xb)\right) \\
                     &\quad + D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) || p_\theta(\zb|\xb)\right) \ . \notag
    \end{align}
}

\frame{
    \frametitle{Importance sampling the ELBO}
    The well-known importance weighted autoencoder (IWAE) bound is tight with the true likelihood in the limit of infinite samples, $S\rightarrow\infty$ \cite{burda_importance_2016},
    \begin{equation}\label{eq:iw-bound}
        \mathcal{L}_{S} = \mathbb{E}_{q(\zb|\xb)}\left[ \log \frac{1}{N} \sum_{s=1}^{S} \frac{p(\xb, \zb^{(s)})}{q(\zb^{(s)}|\xb)} \right] \leq \log p_\theta(\xb) \ ,
    \end{equation}

    \vspace{3mm}

    Consequently, by importance sampling the ELBO, the associated KL-divergence associated vanishes and our likelihood ratio reduces to the KL-divergence associated with $\mathcal{L}^{>k}$.

    \begin{equation}\label{eq:llr-as-kls-iwae-reduced}
        LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq }|\zb_{>k}) q(\zb_{>k}|\xb) || p(\zb|\xb)) \ .
    \end{equation}

    We can now see that $LLR^{>k}_{S}(\xb)$ performs OOD detection based on the top-most latent variables.
}

\frame{
    \frametitle{Results with $LLR^{>k}$}
    \begin{figure}[\textwidth]
    \centering
    \begin{subfigure}[c]{0.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../graphics/paper_hierarchical/roc-FashionMNIST-test-MNIST-test-ll-and-llr-IW250_sohau.pdf}
        \caption{FashionMNIST HVAE evaluated on MNIST}
    \end{subfigure}
    \hfill
    \begin{subfigure}[c]{0.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../graphics/paper_hierarchical/roc-biva-CIFAR10-SVHN-ll-and-llr_sohau.pdf}
        \caption{CIFAR10 BIVA evaluated on SVHN}
    \end{subfigure}
    \end{figure}
}

\frame{
    \frametitle{Results with $LLR^{>k}$}
    The score has good performance across many different datasets.

    \begin{columns}

    \begin{column}{0.5\textwidth}

        \begin{table}[t]
            \centering
            \resizebox{0.9\textwidth}{!}{%
            \begin{tabular}{llrrr}
                \toprule
                 OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on CIFAR10}} \\
                 \midrule
        SVHN          &  $LLR^{>2}$  &  0.811  &  0.837  &  0.394  \\
        CIFAR10       &  $LLR^{>1}$  &  0.469  &  0.479  &  0.835  \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on SVHN}} \\
                 \midrule
        CIFAR10            &  $LLR^{>1}$       &  $0.939$  &  $0.950$  &  $0.052$  \\
        SVHN               &  $LLR^{>1}$       &  $0.489$  &  $0.484$  &  $0.799$  \\
            \bottomrule
            \end{tabular}
            }
        \end{table}
    \end{column}

    \begin{column}{0.5\textwidth}
        \begin{table}[t]
            \centering
            \resizebox{0.9\textwidth}{!}{%
            \begin{tabular}{llrrr}
                \toprule
                 OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on FashionMNIST}} \\
                 \midrule
        MNIST                    & $LLR^{>1}$             &  0.986  &  0.987  &  0.011 \\
        notMNIST                 &  $LLR^{>1}$            &  0.998  &  0.998  &  0.000 \\
        KMNIST                   &  $LLR^{>1}$            &  0.974  &  0.977  &  0.017 \\
        Omniglot28x28            &  $LLR^{>2}$            &  1.000  &  1.000  &  0.000 \\
        Omniglot28x28Inverted    &  $LLR^{>1}$            &  0.954  &  0.954  &  0.050 \\
        SmallNORB28x28           &  $LLR^{>2}$            &  0.999  &  0.999  &  0.002 \\
        SmallNORB28x28Inverted   &  $LLR^{>2}$            &  0.941  &  0.946  &  0.069 \\
        FashionMNIST             &  $LLR^{>1}$            &  0.488  &  0.496  &  0.811 \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on MNIST}} \\
                 \midrule
        FashionMNIST                   &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
        notMNIST                       &  $LLR^{>1}$  &  $1.000$  &  $0.999$  &  $0.000$ \\
        KMNIST                         &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
        Omniglot28x28                  &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
        Omniglot28x28Inverted          &  $LLR^{>1}$  &  $0.944$  &  $0.953$  &  $0.057$ \\
        SmallNORB28x28                 &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
        SmallNORB28x28Inverted         &  $LLR^{>1}$  &  $0.985$  &  $0.987$  &  $0.000$ \\
        MNIST                          &  $LLR^{>2}$  &  $0.515$  &  $0.507$  &  $0.792$ \\
                 \bottomrule
            \end{tabular}
            }
            % \caption{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
            % \label{tab:additional-results-fashionmnist}
        \end{table}
    \end{column}

    \end{columns}
}

% \frame{
%     \frametitle{Results with $LLR^{>k}$}
%     % The score has good performance across many different datasets.

%     \begin{columns}

%     \begin{column}{0.5\textwidth}

%         \begin{table}[t]
%             \centering
%             \resizebox{0.6\textwidth}{!}{%
%             \begin{tabular}{llr}
%                 \toprule
%                  OOD dataset & Metric & AUROC$\uparrow$ \\
%                  \midrule
%                  \multicolumn{3}{c}{\textbf{Trained on CIFAR10}} \\
%                  \midrule
%         SVHN          &  $LLR^{>2}$  &  0.811 \\
%         CIFAR10       &  $LLR^{>1}$  &  0.469 \\
%                  \midrule
%                  \multicolumn{3}{c}{\textbf{Trained on SVHN}} \\
%                  \midrule
%         CIFAR10            &  $LLR^{>1}$       &  0.939 \\
%         SVHN               &  $LLR^{>1}$       &  0.489 \\
%             \bottomrule
%             \end{tabular}
%             }
%         \end{table}
%     \end{column}

%     \begin{column}{0.5\textwidth}
%         \begin{table}[t]
%             \centering
%             \resizebox{0.75\textwidth}{!}{%
%             \begin{tabular}{llr}
%                 \toprule
%                  OOD dataset & Metric & AUROC$\uparrow$ \\
%                  \midrule
%                  \multicolumn{3}{c}{\textbf{Trained on FashionMNIST}} \\
%                  \midrule
%         MNIST                    & $LLR^{>1}$             &  0.986 \\
%         notMNIST                 &  $LLR^{>1}$            &  0.998 \\
%         KMNIST                   &  $LLR^{>1}$            &  0.974 \\
%         Omniglot28x28            &  $LLR^{>2}$            &  1.000 \\
%         Omniglot28x28Inverted    &  $LLR^{>1}$            &  0.954 \\
%         SmallNORB28x28           &  $LLR^{>2}$            &  0.999 \\
%         SmallNORB28x28Inverted   &  $LLR^{>2}$            &  0.941 \\
%         FashionMNIST             &  $LLR^{>1}$            &  0.488 \\
%                  \midrule
%                  \multicolumn{3}{c}{\textbf{Trained on MNIST}} \\
%                  \midrule
%         FashionMNIST                   &  $LLR^{>1}$  &  0.999 \\
%         notMNIST                       &  $LLR^{>1}$  &  1.000 \\
%         KMNIST                         &  $LLR^{>1}$  &  0.999 \\
%         Omniglot28x28                  &  $LLR^{>1}$  &  1.000 \\
%         Omniglot28x28Inverted          &  $LLR^{>1}$  &  0.944 \\
%         SmallNORB28x28                 &  $LLR^{>1}$  &  1.000 \\
%         SmallNORB28x28Inverted         &  $LLR^{>1}$  &  0.985 \\
%         MNIST                          &  $LLR^{>2}$  &  0.515 \\
%                  \bottomrule
%             \end{tabular}
%             }
%             % \caption{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
%             % \label{tab:additional-results-fashionmnist}
%         \end{table}
%     \end{column}

%    \end{columns}
%}

### 2024-02-13 presentation/departments.tex
More presentation work

@@
 \makeDepartmentInfo{danish}{compute}{dtu_dep_name_logo/tex_compute_uk}{dtu_dep_logo/tex_dtu_compute_a}{dtu_frise/tex_dtu_frise}{DTU Compute}{Institut for Matematik og Computer Science}
-\makeDepartmentInfo{english}{compute}{dtu_dep_name_logo/tex_compute_uk}{dtu_dep_logo/tex_dtu_compute_a_uk}{dtu_frise/tex_dtu_frise}{DTU Compute}{Department of Applied Mathematics and Computer Science}
+\makeDepartmentInfo{english}{compute}{dtu_dep_name_logo/tex_compute_uk}{dtu_dep_logo/tex_dtu_compute_a_uk}{dtu_frise/tex_dtu_frise}{DTU Compute and Corti}{Department of Applied Mathematics and Computer Science}
 

### 2024-02-14 presentation/parts/introduction.tex
More presentation work
Updates to presentation

@@
 % !TEX program = xelatex
+
+\section{Introduction}
+
+\begin{frame}
+    \frametitle{Healthcare}
+    \begin{quotation}
+        \noindent\centering
+        Healthcare is the improvement of health via the {\color{dtured}prevention}, {\color{dtured}diagnosis}, {\color{dtured}treatment}, {\color{dtured}amelioration} or {\color{dtured}cure} of {\color{theme-green}disease}, {\color{theme-green}illness}, {\color{theme-green}injury}, and {\color{theme-green}other physical and mental impairments} in people.
+    \end{quotation}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Medical dialogue}
+    \begin{figure}
+        \centering
+        \includegraphics[width=0.7\paperwidth]{figures/speech_bubble.pdf}
+    \end{figure}
+
+    \note[item]{}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Medical dialogue}
+    \begin{enumerate}
+        \item Failure of communication is a leading cause of medical error contributing to two out of three adverse events \cite{starmer_changes_2014}.
+        \item Between 9\% and 16.6\% of all hospital admissions had preventable adverse outcomes (AU, UK, NZ, DK) \cite{carver_medical_2024}.
+    \end{enumerate}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Types of dialogue}
+    \begin{itemize}
+        \item Context: Controlled / Chaotic
+        \item Domain: Specialized / General
+        \item Person: Nurse, doctor, midwife, caregiver, psychiatrist, insurance professional
+        \item Purpose: Triaging, diagnosis, treatment, follow-up, documentation, coding, billing
+    \end{itemize}
+
+    \note[item]{}
+\end{frame}

### 2024-02-18 presentation/parts/introduction.tex
Add slides on retrospective

@@
 
-    \note[item]{}
+    \note[item]{Healthcare sector is a complex system}
 \end{frame}
@@
 \begin{frame}
-    \frametitle{Medical dialogue}
-    \begin{enumerate}
-        \item Failure of communication is a leading cause of medical error contributing to two out of three adverse events \cite{starmer_changes_2014}.
-        \item Between 9\% and 16.6\% of all hospital admissions had preventable adverse outcomes (AU, UK, NZ, DK) \cite{carver_medical_2024}.
-    \end{enumerate}
+    \frametitle{Errors in medical dialogue}
+    \begin{itemize}
+        \item<1,2>Communication is everywhere in healthcare. 
+        \item <1,2>It is complex, involving multiple participants, different contexts, and different purposes.
+        \item <2> {\color{dtured} } Failure of communication is a leading cause of medical error contributing to two out of three adverse events \cite{starmer_changes_2014}.
+        \item <2> A considerable fraction of all hospital admissions had preventable adverse outcomes (9\% to 16.6\% in AU, NZ, UK, DK) \cite{carver_medical_2024}.
+    \end{itemize}
+\end{frame}
+
+% \begin{frame}
+%     \frametitle{Types of dialogue}
+%     \begin{itemize}
+%         \item Domain: Specialized / General
+%         \item Context: Controlled / Chaotic
+%         \item Person: Nurse, doctor, midwife, caregiver, psychiatrist, insurance professional
+%         \item Purpose: Triaging, diagnosis, treatment, follow-up, documentation, coding, billing
+%     \end{itemize}
+% \end{frame}
+
+\begin{frame}
+    \frametitle{Documenting medical encounters}
+    \begin{itemize}
+        \item <1,2> Documentation is a central part of healthcare.
+        \item <1,2> E.g. patient records, insurance claims, billing, research, training, legal purposes.
+        \vspace{1em}
+        \item <2> {\color{dtured}Time}: Physicians spend 34-37\% of their time on documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
+        \item <2> {\color{dtured}Quality}: Discharge summaries rarely meet all timeline, transmission, and content criteria.
+\cite{horwitz_comprehensive_2013}\footnote{Outpatient visits, Yale-New Haven Hospital.}
+    \end{itemize}
+
+    \note[item]{Documentation is essential for a number of purposes but is very time-consuming and imperfect}
 \end{frame}
@@
 \begin{frame}
-    \frametitle{Types of dialogue}
+    \frametitle{How might machine learning help?}
     \begin{itemize}
-        \item Context: Controlled / Chaotic
-        \item Domain: Specialized / General
-        \item Person: Nurse, doctor, midwife, caregiver, psychiatrist, insurance professional
-        \item Purpose: Triaging, diagnosis, treatment, follow-up, documentation, coding, billing
+        \item <1> {\color{dtured}Assist} with documentation.
+        \item <1> {\color{dtured}Augment} communication.
+        \item <1> {\color{dtured}Improve} decision-making.
+        \vspace{1em}
+        \item <1> {\color{dtured}Reduce} errors.
+        \item <1> {\color{dtured}Save} time.
+    \end{itemize}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Reliability of machine learning}
+    \begin{itemize}
+        \item <1> {\color{dtured}Data}: Privacy, quality, quantity, diversity.
+        \item <1> {\color{dtured}Interpretability}: Trust, ethics, regulation.
+        \item <1> {\color{dtured}Explainability}: Transparency, accountability.
+        \item <1> {\color{dtured}Robustness}: Adversarial attacks, distribution shift.
+        \item <1> {\color{dtured}Bias}: Fairness, transparency.
+        \item <1> {\color{dtured}Complexity}: Context, domain, language, culture.
     \end{itemize}
 
-    \note[item]{}
 \end{frame}

### 2024-02-20 appendices/supplementary_retrospective.tex
Final updates
Fix mixed-up column names in supplementary

@@
 
-        Predictor & Call-taker & \multicolumn{2}{c|}{Model} & \multicolumn{4}{c}{Call-taker supported by the model (simulated)} \\
+            {Predictor} & {Call-taker} & \multicolumn{2}{c|}{{Model}} & \multicolumn{4}{c}{{Call-taker supported by the model (simulated)}} \\
             \midrule
-        When & - & After call & During call & After call & During call & After call & During call \\
+            {When} & {During call} & {After call} & {During call} & {After call} & {During call} & {After call} & {During call} \\
             \midrule
-        Method & - & 1.C & 2.C & 1.A & 1.B & 2.A & 2.B \\
+            {Method} & - & - & - & {neg $\rightarrow$ pos} & {neg $\rightarrow$ pos} & {pos $\rightarrow$ neg} & {pos $\rightarrow$ neg} \\
             \midrule
 
-        \makecell[l]{F1-score [\%] $\uparrow$}                   & \makecell[c]{25.8 \\ (23.7-27.9)} & \makecell[c]{35.7 \\ (35.0-36.4)} & \makecell[c]{33.1 \\ (32.4-33.7)} & \makecell[c]{28.9 \\ (28.3-29.5)} & \makecell[c]{33.3 \\ (32.5-34.1)} & \makecell[c]{27.6 \\ (27.0-28.1)} & \makecell[c]{32.7 \\ (31.8-33.5)} \\
+            \makecell[l]{{F1-score} [\%] $\uparrow$}                   & \makecell[c]{25.8 \\ (23.7-27.9)} & \makecell[c]{35.7 \\ (35.0-36.4)} & \makecell[c]{33.1 \\ (32.4-33.7)} & \makecell[c]{28.9 \\ (28.3-29.5)} & \makecell[c]{27.6 \\ (27.0-28.1)}  & \makecell[c]{33.3 \\ (32.5-34.1)}& \makecell[c]{32.7 \\ (31.8-33.5)} \\
             \midrule
-        \makecell[l]{Sensitivity [\%] $\uparrow$}                & \makecell[c]{52.7 \\ (49.2-56.4)} & \makecell[c]{63.0 \\ (62.0-64.1)} & \makecell[c]{58.7 \\ (57.7-59.8)} & \makecell[c]{72.4 \\ (71.5-73.3)} & \makecell[c]{43.4 \\ (42.3-44.5)} & \makecell[c]{72.3 \\ (71.4-73.3)} & \makecell[c]{39.1 \\ (38.1-40.1)} \\
+            \makecell[l]{{Sensitivity} [\%] $\uparrow$}                & \makecell[c]{52.7 \\ (49.2-56.4)} & \makecell[c]{63.0 \\ (62.0-64.1)} & \makecell[c]{58.7 \\ (57.7-59.8)} & \makecell[c]{72.4 \\ (71.5-73.3)} & \makecell[c]{72.3 \\ (71.4-73.3)} & \makecell[c]{43.4 \\ (42.3-44.5)} & \makecell[c]{39.1 \\ (38.1-40.1)} \\
             \midrule
-        \makecell[l]{PPV [\%] $\uparrow$}                        & \makecell[c]{17.1 \\ (15.5-18.6)} & \makecell[c]{24.9 \\ (24.3-25.5)} & \makecell[c]{23.0 \\ (22.5-23.6)} & \makecell[c]{18.0 \\ (17.6-18.4)} & \makecell[c]{27.0 \\ (26.3-27.8)} & \makecell[c]{17.0 \\ (16.7-17.4)} & \makecell[c]{28.1 \\ (27.3-28.9)} \\
+            \makecell[l]{{PPV} [\%] $\uparrow$}                        & \makecell[c]{17.1 \\ (15.5-18.6)} & \makecell[c]{24.9 \\ (24.3-25.5)} & \makecell[c]{23.0 \\ (22.5-23.6)} & \makecell[c]{18.0 \\ (17.6-18.4)} & \makecell[c]{17.0 \\ (16.7-17.4)} & \makecell[c]{27.0 \\ (26.3-27.8)} & \makecell[c]{28.1 \\ (27.3-28.9)} \\
             \midrule
-        \makecell[l]{FOR [\%] $\downarrow$ \\ (1 - NPV)}         & \makecell[c]{0.105 \\ (0.094-0.116)} & \makecell[c]{0.082 \\ (0.079-0.085)} & \makecell[c]{0.091 \\ (0.088-0.094)} & \makecell[c]{0.061 \\ (0.059-0.064)} & \makecell[c]{0.125 \\ (0.121-0.129)} & \makecell[c]{0.061 \\ (0.059-0.064)} & \makecell[c]{0.134 \\ (0.131-0.138)} \\
+            \makecell[l]{{FOR} [\%] $\downarrow$ \\ (1 - NPV)}         & \makecell[c]{0.105 \\ (0.094-0.116)} & \makecell[c]{0.082 \\ (0.079-0.085)} & \makecell[c]{0.091 \\ (0.088-0.094)} & \makecell[c]{0.061 \\ (0.059-0.064)} & \makecell[c]{0.061 \\ (0.059-0.064)} & \makecell[c]{0.125 \\ (0.121-0.129)} & \makecell[c]{0.134 \\ (0.131-0.138)} \\
             \midrule
-        \makecell[l]{FPR [\%] $\downarrow$ \\ (1 - specificity)} & \makecell[c]{0.565 \\ (0.539-0.590)} & \makecell[c]{0.419 \\ (0.413-0.426)} & \makecell[c]{0.432 \\ (0.426-0.439)} & \makecell[c]{0.726 \\ (0.717-0.735)} & \makecell[c]{0.258 \\ (0.253-0.263)} & \makecell[c]{0.776 \\ (0.767-0.786)} & \makecell[c]{0.221 \\ (0.216-0.226)} \\
+            \makecell[l]{{FPR} [\%] $\downarrow$ \\ (1 - specificity)} & \makecell[c]{0.565 \\ (0.539-0.590)} & \makecell[c]{0.419 \\ (0.413-0.426)} & \makecell[c]{0.432 \\ (0.426-0.439)} & \makecell[c]{0.726 \\ (0.717-0.735)} & \makecell[c]{0.776 \\ (0.767-0.786)} & \makecell[c]{0.258 \\ (0.253-0.263)} & \makecell[c]{0.221 \\ (0.216-0.226)} \\
 

### 2024-02-20 presentation/parts/introduction.tex
Work on presentation

@@
         \vspace{1em}
-        \item <2> {\color{dtured}Time}: Physicians spend 34-37\% of their time on documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
-        \item <2> {\color{dtured}Quality}: Discharge summaries rarely meet all timeline, transmission, and content criteria.
+        \item <2> {\color{dtured}Time-consuming}: Physicians spend 34-37\% of their time on documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
+        \item <2> {\color{dtured}Varying quality}: Discharge summaries rarely meet all timeline, transmission, and content criteria.
 \cite{horwitz_comprehensive_2013}\footnote{Outpatient visits, Yale-New Haven Hospital.}

### 2024-02-20 presentation/parts/presentation_hierarchical.tex
More presentation work
Updates to presentation
WIP
Work on presentation

--- 2024-02-12 presentation/parts/paper_hierarchical.tex
@@
 \section{Out-of-distribution detection}
+% \subsection{Out-of-distribution detection}
 
@@
         \begin{itemize}
+            \item We provide evidence that out-of-distribution detection fails due to learned low-level features that generalize across datasets.
             \item We present a fast and fully unsupervised method for OOD detection competitive with the state-of-the-art
-            \item We provide evidence that out-of-distribution detection fails due to learned low-level features that generalize across datasets.
         \end{itemize}  
@@
 
-\section{Models}
+\section{Latent variable models}
 
@@
 
-\section{The Problem}
+\section{Identifying the issue}
 
@@
     \frametitle{Importance sampling the ELBO}
-    The well-known importance weighted autoencoder (IWAE) bound is tight with the true likelihood in the limit of infinite samples, $S\rightarrow\infty$ \cite{burda_importance_2016},
+    The importance weighted autoencoder (IWAE) bound is tight with the true likelihood in the limit of infinite samples, $S\rightarrow\infty$ \cite{burda_importance_2016},
     \begin{equation}\label{eq:iw-bound}
@@
 
-    Consequently, by importance sampling the ELBO, the associated KL-divergence associated vanishes and our likelihood ratio reduces to the KL-divergence associated with $\mathcal{L}^{>k}$.
+    Consequently, by importance sampling the ELBO, the associated KL-divergence vanishes and our likelihood ratio reduces to the KL-divergence of $\mathcal{L}^{>k}$.
 
@@
 
-    We can now see that $LLR^{>k}_{S}(\xb)$ performs OOD detection based on the top-most latent variables.
+    $LLR^{>k}_{S}(\xb)$ performs KL-divergence-based OOD detection using top-most latent variables.
 }
@@
 
+\begin{frame}
+    \frametitle{Results on FashionMNIST/MNIST}
+    \begin{table}
+        \centering
+        \resizebox*{!}{0.87\textheight}{%
+        \begin{tabular}{lrrr}
+            \toprule
+             Method & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
+             \midrule
+             \multicolumn{4}{c}{\textbf{FashionMNIST (in) / MNIST (out)}} \\
+             \midrule
+             \multicolumn{4}{l}{\textbf{Use prior knowledge of OOD}} \\
+    Backgr. contrast. LR (PixelCNN) {\parencite{ren_likelihood_2019}}               & $0.994$ & $0.993$ & $0.001$ \\
+    Backgr. contrast. LR (VAE) {\parencite{choi_waic_2019}}                    & $0.924$ & - & - \\
+    Binary classifier {\parencite{ren_likelihood_2019}}                              & $0.455$ & $0.505$ & $0.886$ \\ % 6
+    $p(\hat{y} | \xb)$ with OOD as noise class {\parencite{ren_likelihood_2019}}     & $0.877$ & $0.871$ & $0.195$ \\ % 7
+    $p(\hat{y} | \xb)$ with calibration on OOD {\parencite{ren_likelihood_2019}}     & $0.904$ & $0.895$ & $0.139$ \\ % 8
+    Input complexity ($S$, Glow) \parencite{hendrycks_deep_2019}                    & $0.998$ & - & - \\
+    Input complexity ($S$, PixelCNN++) \parencite{hendrycks_deep_2019}              & $0.967$ & - & - \\
+             \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
+    $p(\hat{y} | \xb)$ {\parencite{ren_likelihood_2019, hendrycks_baseline_2017}}                        & $0.734$ & $0.702$ & $0.506$ \\
+    Entropy of $p(y | \xb)$ {\parencite{ren_likelihood_2019}}                        & $0.746$ & $0.726$ & $0.448$ \\
+    ODIN {\parencite{ren_likelihood_2019, liang_enhancing_2018}}                                       & $0.752$ & $0.763$ & $0.432$ \\
+    VIB \parencite{alemi_uncertainty_2018, choi_waic_2019}                                          & $0.941$ & - & - \\
+    Mahalanobis distance, CNN {\parencite{ren_likelihood_2019}}                     & $0.942$ & $0.928$ & $0.088$ \\
+    Mahalanobis distance, DenseNet {\parencite{lee_simple_2018}}                & $0.986$ & - & - \\
+    Ensemble, 20 classifiers {\parencite{ren_likelihood_2019, lakshminarayanan_simple_2017}}                  & $0.857$ & $0.849$ & $0.240$ \\
+             \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
+             \multicolumn{4}{l}{\textit{- Ensembles}} \\
+    WAIC, 5 models, VAE {\parencite{choi_waic_2019}}                          & $0.766$ & - & - \\
+    WAIC, 5 models, PixelCNN {\parencite{ren_likelihood_2019}}                      & $0.221$ & $0.401$ & $0.911$ \\
+            \multicolumn{4}{l}{\textit{- Not ensembles}} \\
+    Likelihood regret \parencite{xiao_likelihood_2020}                               & $\mathbf{0.988}$ & - & - \\
+    $\mathcal{L}^{>0}$ + HVAE (ours)                    & $0.268$ & $0.363$ & $0.882$ \\
+    $\mathcal{L}^{>1}$ + HVAE (ours)                    & $0.593$ & $0.591$ & $0.658$ \\
+    $\mathcal{L}^{>2}$ + HVAE (ours)                    & $0.712$ & $0.750$ & $0.548$ \\
+    $LLR^{>1}$ + HVAE (ours)                            & $0.964$ & $0.961$ & $0.036$ \\
+    $LLR^{>1}_{250}$ + HVAE (ours)                      & $0.984$ & $\mathbf{0.984}$ & $\mathbf{0.013}$ \\
+             \bottomrule
+        \end{tabular}%
+        }
+    \end{table}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Results on CIFAR10/SVHN}
+    \begin{table}
+        \centering
+        \resizebox*{!}{0.87\textheight}{%
+        \begin{tabular}{lrrr}
+            \toprule
+             Method & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
+             \midrule
+             \multicolumn{4}{c}{\textbf{CIFAR10 (in) / SVHN (out)}} \\
+             \midrule
+             \multicolumn{4}{l}{\textbf{Use prior knowledge of OOD}} \\
+    Backgr. contrast. LR (PixelCNN) {\parencite{ren_likelihood_2019}}               & $0.930$ & $0.881$ & $0.066$ \\
+    Backgr. contrast. LR (VAE) {\parencite{xiao_likelihood_2020}}                    & $0.265$ & - & - \\
+    Outlier exposure {\parencite{hendrycks_deep_2019}}                              & $0.984$ & - & - \\
+    Input complexity ($S$, Glow) \parencite{serra_input_2020}                   & $0.950$ & - & - \\
+    Input complexity ($S$, PixelCNN++) \parencite{serra_input_2020}             & $0.929$ & - & - \\
+    Input complexity ($S$, HVAE) (Ours) \parencite{serra_input_2020}\textsuperscript{\ref{footnote:serra_hierarchical_likelihood}} & $0.833$ & $0.855$ & $0.344$ \\
+    \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
+    Mahalanobis distance {\parencite{lee_simple_2018}}                          & $0.991$ & - & -  \\
+             \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
+             \multicolumn{4}{l}{\textit{- Ensembles}} \\
+    WAIC, 5 models, Glow {\parencite{choi_waic_2019}}                          & $1.000$ & - & - \\
+    WAIC, 5 models, PixelCNN {\parencite{ren_likelihood_2019}}                      & $0.628$ & $0.616$ & $0.657$ \\
+             \multicolumn{4}{l}{\textit{- Not ensembles}} \\
+    Likelihood regret \parencite{xiao_likelihood_2020}                               & $0.875$ & - & - \\
+    $LLR^{>2}$ + HVAE (ours)                            & $0.811$ & $0.837$ & $0.394$ \\
+    $LLR^{>2}$ + BIVA (ours)                            & $\mathbf{0.891}$ & $\mathbf{0.875}$ & $\mathbf{0.172}$ \\
+             \bottomrule
+        \end{tabular}%
+        }
+    \end{table}
+\end{frame}
+
 \frame{
-    \frametitle{Results with $LLR^{>k}$}
-    The score has good performance across many different datasets.
-
+    \frametitle{Results on diverse datasets}
     \begin{columns}
-
     \begin{column}{0.5\textwidth}
@@
     \end{column}
-
     \begin{column}{0.5\textwidth}

### 2024-02-21 presentation/departments.tex
WIP

--- 2024-02-12 presentation/departments.tex
@@
 \makeDepartmentInfo{danish}{compute}{dtu_dep_name_logo/tex_compute_uk}{dtu_dep_logo/tex_dtu_compute_a}{dtu_frise/tex_dtu_frise}{DTU Compute}{Institut for Matematik og Computer Science}
-\makeDepartmentInfo{english}{compute}{dtu_dep_name_logo/tex_compute_uk}{dtu_dep_logo/tex_dtu_compute_a_uk}{dtu_frise/tex_dtu_frise}{DTU Compute}{Department of Applied Mathematics and Computer Science}
+\makeDepartmentInfo{english}{compute}{dtu_dep_name_logo/tex_compute_uk}{dtu_dep_logo/tex_dtu_compute_a_uk}{dtu_frise/tex_dtu_frise}{}{Department of Applied Mathematics and Computer Science}
 

### 2024-02-21 presentation/parts/overview.tex
WIP

% !TEX root = ../presentation.tex
% !BIB program = biber
% !TEX program = xelatex

\begin{frame}
    \frametitle{Overview: Thesis}
    \begin{columns}
        \begin{column}{0.6\textwidth}
            \begin{table}
                \resizebox*{\textwidth}{!}{%
                \begin{tabular}{l l}
                    \textsc{chapter 1-3}                   & \makecell[l]{Introduction, Research Questions, and Background} \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]
                    \textsc{chapter 4}                     & \makecell[l]{Hierarchical VAEs Know What They Don't Know} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    \makecell[l]{\textsc{chapter 5}\\}     & \makecell[l]{Model-Agnostic Out-of-Distribution Detection\\ Using Combined Statistical Tests} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    \makecell[l]{\textsc{chapter 6}\\}     & \makecell[l]{A Brief Overview of Unsupervised Speech \\Representation Learning} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    \textsc{chapter 7}     & \makecell[l]{Benchmarking Latent Variable Models for Speech} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    \makecell[l]{\textsc{chapter 8}\\}     & \makecell[l]{Automated Medical Coding on MIMIC-III and \\MIMIC-IV: A Critical Review and Replicability Study} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    \makecell[l]{\textsc{chapter 9}\\}     & \makecell[l]{A Retrospective Study on Machine Learning-\\Assisted Stroke Recognition for Medical Helpline Calls} \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]
                    \textsc{chapter 10}                    & \makecell[l]{Discussion and Conclusion} \\
                \end{tabular}
                }
            \end{table}
        \end{column}
        \begin{column}{0.4\textwidth}
        \end{column}
    \end{columns}
\end{frame}

### 2024-02-21 presentation/parts/selection.tex
WIP

% !TEX root = ../presentation.tex
% !BIB program = biber
% !TEX program = xelatex

\begin{frame}
    \frametitle{Overview: Thesis}
    \begin{columns}
        \begin{column}{0.6\textwidth}

            \hfsetfillcolor{presentation_green}
            \hfsetbordercolor{presentation_green}

            \begin{table}
                \resizebox*{\textwidth}{!}{%
                \begin{tabular}{l l l l}
                    & \textsc{chapter 1-3}   & \makecell[l]{Introduction, Research Questions, and Background} & \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]
                    \tikzmarkin<2>[below right offset={0.1,-0.4},above left offset={-0.1,0.4}]{a}
                    & \textsc{chapter 4}     & \makecell[l]{Hierarchical VAEs Know What They Don't Know} & \tikzmarkend{a}\\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    \tikzmarkin<3>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{b}
                    & \makecell[tl]{\textsc{chapter 5}}     & \makecell[l]{Model-Agnostic Out-of-Distribution Detection\\ Using Combined Statistical Tests} & \tikzmarkend{b} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    \tikzmarkin<4>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{c}
                    & \makecell[tl]{\textsc{chapter 6}}     & \makecell[l]{A Brief Overview of Unsupervised Speech \\Representation Learning} & \tikzmarkend{c} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    \tikzmarkin<5>[below right offset={0.1,-0.4},above left offset={-0.1,0.4}]{d}
                    & \textsc{chapter 7}     & \makecell[l]{Benchmarking Latent Variable Models for Speech} & \tikzmarkend{d} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    \tikzmarkin<6>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{e}
                    & \makecell[tl]{\textsc{chapter 8}}     & \makecell[l]{Automated Medical Coding on MIMIC-III and \\MIMIC-IV: A Critical Review and Replicability Study} & \tikzmarkend{e} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    \tikzmarkin<7>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{f}
                    & \makecell[tl]{\textsc{chapter 9}}     & \makecell[l]{A Retrospective Study on Machine Learning-\\Assisted Stroke Recognition for Medical Helpline Calls} & \tikzmarkend{f} \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]
                    & \textsc{chapter 10}    & \makecell[l]{Discussion and Conclusion} & \\
                \end{tabular}
                }
            \end{table}
        \end{column}
        \begin{column}{0.4\textwidth}
            \begin{figure}
                \centering
                \begin{overprint}
                    \onslide<1>
                    \begin{tikzpicture}
                        \tikzstyle{mybox} = [fill, rounded corners, minimum width=5cm, minimum height=0.8cm]
                        \tikzstyle{myrect} = [fill, minimum height=0.6cm]

                        \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
                            \node [myrect, fill=presentation_orange_shaded] at (\x,5) {};
                        \node [] at (0,5) {\textsc{predictions}};

                        \node [mybox, fill=black!10] at (0,4) (downstream) {\textsc{downstream model}};

                        \node [mybox, fill=presentation_blue] at (0,3) (text) {};
                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
                            \node [myrect, fill=presentation_blue_shaded] at (\x,3) {};
                        \node [] at (0,3) {\textsc{text}};

                        \node [mybox, fill=black!10] at (0,2) (recognition) {\textsc{speech recognition}};

                        \node [mybox, fill=black!10] at (0,1) (representation) {\textsc{representation learning}};

                        \node [mybox, fill=presentation_green] at (0,0) (speech) {};
                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
                            \node [myrect, fill=presentation_green_shaded] at (\x,0) {};
                        \node [] at (0,0) {\textsc{speech}};

                        \node [draw, rectangle, rounded corners, line width=0.1cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
                        \node [above] at (uncertainty.north) {\textsc{uncertainty}};
                    \end{tikzpicture}

                    \onslide<2>  % Hierarchical
                    \begin{tikzpicture}
                        \tikzstyle{mybox} = [fill, rounded corners, minimum width=5cm, minimum height=0.8cm]
                        \tikzstyle{myrect} = [fill, minimum height=0.6cm]

                        \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
                            \node [nearly transparent, myrect, fill=presentation_orange_shaded] at (\x,5) {};
                        \node [nearly transparent] at (0,5) {\textsc{predictions}};

                        \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\textsc{downstream model}};

                        \node [nearly transparent, mybox, fill=presentation_blue] at (0,3) (text) {};
                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
                            \node [nearly transparent, myrect, fill=presentation_blue_shaded] at (\x,3) {};
                        \node [nearly transparent] at (0,3) {\textsc{text}};

                        \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\textsc{speech recognition}};

                        \node [mybox, fill=black!10] at (0,1) (representation) {\textsc{representation learning}};

                        \node [nearly transparent, mybox, fill=presentation_green] at (0,0) (speech) {};
                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
                            \node [nearly transparent, myrect, fill=presentation_green_shaded] at (\x,0) {};
                        \node [nearly transparent] at (0,0) {\textsc{speech}};

                        \node [draw, rectangle, rounded corners, line width=0.1cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
                        \node [above] at (uncertainty.north) {\textsc{uncertainty}};
                    \end{tikzpicture}

                    \onslide<3>  % Model-Agnostic
                    \begin{tikzpicture}
                        \tikzstyle{mybox} = [fill, rounded corners, minimum width=5cm, minimum height=0.8cm]
                        \tikzstyle{myrect} = [fill, minimum height=0.6cm]

                        \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
                            \node [nearly transparent, myrect, fill=presentation_orange_shaded] at (\x,5) {};
                        \node [nearly transparent] at (0,5) {\textsc{predictions}};

                        \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\textsc{downstream model}};

                        \node [nearly transparent, mybox, fill=presentation_blue] at (0,3) (text) {};
                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
                            \node [nearly transparent, myrect, fill=presentation_blue_shaded] at (\x,3) {};
                        \node [nearly transparent] at (0,3) {\textsc{text}};

                        \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\textsc{speech recognition}};

                        \node [mybox, fill=black!10] at (0,1) (representation) {\textsc{representation learning}};

                        \node [nearly transparent, mybox, fill=presentation_green] at (0,0) (speech) {};
                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
                            \node [nearly transparent, myrect, fill=presentation_green_shaded] at (\x,0) {};
                        \node [nearly transparent] at (0,0) {\textsc{speech}};

                        \node [draw, rectangle, rounded corners, line width=0.1cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
                        \node [above] at (uncertainty.north) {\textsc{uncertainty}};
                    \end{tikzpicture}

                    \onslide<4>  % Brief
                    \begin{tikzpicture}
                        \tikzstyle{mybox} = [fill, rounded corners, minimum width=5cm, minimum height=0.8cm]
                        \tikzstyle{myrect} = [fill, minimum height=0.6cm]

                        \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
                            \node [nearly transparent, myrect, fill=presentation_orange_shaded] at (\x,5) {};
                        \node [nearly transparent] at (0,5) {\textsc{predictions}};

                        \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\textsc{downstream model}};

                        \node [nearly transparent, mybox, fill=presentation_blue] at (0,3) (text) {};
                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
                            \node [nearly transparent, myrect, fill=presentation_blue_shaded] at (\x,3) {};
                        \node [nearly transparent] at (0,3) {\textsc{text}};

                        \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\textsc{speech recognition}};

                        \node [mybox, fill=black!10] at (0,1) (representation) {\textsc{representation learning}};

                        \node [mybox, fill=presentation_green] at (0,0) (speech) {};
                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
                            \node [myrect, fill=presentation_green_shaded] at (\x,0) {};
                        \node [] at (0,0) {\textsc{speech}};

                        \node [nearly transparent, draw, rectangle, rounded corners, line width=0.1cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
                        \node [nearly transparent, above] at (uncertainty.north) {\textsc{uncertainty}};
                    \end{tikzpicture}

                    % \onslide<6>
                    % \begin{tikzpicture}
                    %     % Define box style
                    %     \tikzstyle{mybox} = [fill, rounded corners, minimum width=5cm, minimum height=0.8cm]
                    %     % Define rectangle style
                    %     \tikzstyle{myrect} = [fill, minimum width=0.25cm, minimum height=0.6cm]

                    %     \node [mybox, fill=presentation_orange] at (0,5) (predictions) {Predictions};
                    %     \node [mybox, fill=black!10] at (0,4) (downstream) {Downstream Model};

                    %     \node [mybox, fill=presentation_blue] at (0,3) (text) {};
                    %     \node [myrect, fill=presentation_blue_shaded] at (-2.0,3) {};
                    %     \node [myrect, fill=presentation_blue_shaded] at (-1.5,3) {};
                    %     \node [myrect, fill=presentation_blue_shaded] at (-1.0,3) {};
                    %     \node [myrect, fill=presentation_blue_shaded] at (-0.5,3) {};
                    %     \node [myrect, fill=presentation_blue_shaded] at (0,3) {};
                    %     \node [myrect, fill=presentation_blue_shaded] at (0.5,3) {};
                    %     \node [myrect, fill=presentation_blue_shaded] at (1.0,3) {};
                    %     \node [myrect, fill=presentation_blue_shaded] at (1.5,3) {};
                    %     \node [myrect, fill=presentation_blue_shaded] at (2.0,3) {};
                    %     \node at (0,3) {Text};

                    %     \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {Speech Recognition};
                    %     \node [nearly transparent, mybox, fill=black!10] at (0,1) (representation) {Representation Learning};
                    %     %
                    %     \node [nearly transparent, mybox, fill=presentation_green] at (0,0) (speech) {};
                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (-2.0,0) {};
                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (-1.5,0) {};
                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (-1.0,0) {};
                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (-0.5,0) {};
                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (0,0) {};
                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (0.5,0) {};
                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (1.0,0) {};
                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (1.5,0) {};
                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (2.0,0) {};
                    %     \node [nearly transparent] at (0,0) {Speech};
                    % \end{tikzpicture}

                    \end{overprint}
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}
    \frametitle{Overview: Presentation}
    \begin{columns}
        \begin{column}{0.6\textwidth}

            \hfsetfillcolor{presentation_green}
            \hfsetbordercolor{presentation_green}

            \begin{table}
                \resizebox*{\textwidth}{!}{%
                \begin{tabular}{l l l l}
                    & \textsc{chapter 1-3}   & {\color<2->{black!50}}\makecell[l]{Introduction, Research Questions, and Background} & \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]
                    \tikzmarkin<2>[below right offset={0.1,-0.4},above left offset={-0.1,0.4}]{a}
                    & \textsc{chapter 4}     & \makecell[l]{Hierarchical VAEs Know What They Don't Know} & \tikzmarkend{a}\\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    % \tikzmarkin<3>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{b}
                    % & \makecell[tl]{\textsc{chapter 5}}     & \makecell[l]{Model-Agnostic Out-of-Distribution Detection\\ Using Combined Statistical Tests} & \tikzmarkend{b} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    \tikzmarkin<4>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{c}
                    & \makecell[tl]{\textsc{chapter 6}}     & \makecell[l]{A Brief Overview of Unsupervised Speech \\Representation Learning} & \tikzmarkend{c} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    % \tikzmarkin<5>[below right offset={0.1,-0.4},above left offset={-0.1,0.4}]{d}
                    % & \textsc{chapter 7}     & \makecell[l]{Benchmarking Latent Variable Models for Speech} & \tikzmarkend{d} \\
                    % \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    % \tikzmarkin<6>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{e}
                    % & \makecell[tl]{\textsc{chapter 8}}     & \makecell[l]{Automated Medical Coding on MIMIC-III and \\MIMIC-IV: A Critical Review and Replicability Study} & \tikzmarkend{e} \\
                    % \addlinespace[0.5em]
                    \addlinespace[0.5em]
                    \tikzmarkin<7>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{f}
                    & \makecell[tl]{\textsc{chapter 9}}     & \makecell[l]{A Retrospective Study on Machine Learning-\\Assisted Stroke Recognition for Medical Helpline Calls} & \tikzmarkend{f} \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]
                    & \textsc{chapter 10}    & \makecell[l]{Discussion and Conclusion} & \\
                \end{tabular}
                }
            \end{table}
        \end{column}
    \end{columns}
\end{frame}

### 2024-02-22 presentation/parts/overview.tex
WIP

@@
                 \resizebox*{\textwidth}{!}{%
-                \begin{tabular}{l l}
-                    \textsc{chapter 1-3}                   & \makecell[l]{Introduction, Research Questions, and Background} \\
+                \begin{tabular}{l l l l}
+
+                    & {\color<2->{black}   \makecell[l]{\textsc{\itshape chapter 1-3}}}       & {\color<2->{black}    \bfseries \scshape \Large \makecell[l]{introduction, research questions, and background}} & \\
                     \addlinespace[0.5em]
@@
                     \addlinespace[0.5em]
-                    \textsc{chapter 4}                     & \makecell[l]{Hierarchical VAEs Know What They Don't Know} \\
+
+                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 4}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{hierarchical vaes know what they don't know}} & \\
                     \addlinespace[0.5em]
                     \addlinespace[0.5em]
-                    \makecell[l]{\textsc{chapter 5}\\}     & \makecell[l]{Model-Agnostic Out-of-Distribution Detection\\ Using Combined Statistical Tests} \\
+
+                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \\
                     \addlinespace[0.5em]
                     \addlinespace[0.5em]
-                    \makecell[l]{\textsc{chapter 6}\\}     & \makecell[l]{A Brief Overview of Unsupervised Speech \\Representation Learning} \\
+
+                    & {\color<2->{black!20}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \\
                     \addlinespace[0.5em]
                     \addlinespace[0.5em]
-                    \textsc{chapter 7}     & \makecell[l]{Benchmarking Latent Variable Models for Speech} \\
+
+                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \\
                     \addlinespace[0.5em]
                     \addlinespace[0.5em]
-                    \makecell[l]{\textsc{chapter 8}\\}     & \makecell[l]{Automated Medical Coding on MIMIC-III and \\MIMIC-IV: A Critical Review and Replicability Study} \\
+
+                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \\
                     \addlinespace[0.5em]
                     \addlinespace[0.5em]
-                    \makecell[l]{\textsc{chapter 9}\\}     & \makecell[l]{A Retrospective Study on Machine Learning-\\Assisted Stroke Recognition for Medical Helpline Calls} \\
+
+                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \\
                     \addlinespace[0.5em]
@@
                     \addlinespace[0.5em]
-                    \textsc{chapter 10}                    & \makecell[l]{Discussion and Conclusion} \\
+
+                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 10}}}        & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{discussion and conclusion}} & \\
                 \end{tabular}

### 2024-02-22 presentation/parts/selection.tex
WIP

@@
                 \begin{tabular}{l l l l}
-                    & \textsc{chapter 1-3}   & \makecell[l]{Introduction, Research Questions, and Background} & \\
+
+                    & {\color<2->{black!20} \textsc{chapter 1-3}}   & {\color<2->{black!20}\makecell[l]{Introduction, Research Questions, and Background}} & \\
                     \addlinespace[0.5em]
@@
                     \addlinespace[0.5em]
-                    \tikzmarkin<2>[below right offset={0.1,-0.4},above left offset={-0.1,0.4}]{a}
-                    & \textsc{chapter 4}     & \makecell[l]{Hierarchical VAEs Know What They Don't Know} & \tikzmarkend{a}\\
+
+                    \tikzmarkin<2>{hierarchical-1}(0.1,-0.3)(-0.1,0.4)
+                    & \textsc{chapter 4}     & Hierarchical VAEs Know What They Don't Know & \tikzmarkend{hierarchical-1} \\
                     \addlinespace[0.5em]
                     \addlinespace[0.5em]
-                    \tikzmarkin<3>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{b}
-                    & \makecell[tl]{\textsc{chapter 5}}     & \makecell[l]{Model-Agnostic Out-of-Distribution Detection\\ Using Combined Statistical Tests} & \tikzmarkend{b} \\
+
+                    \tikzmarkin<3>{modelagnostic-1}(0.1,-0.4)(-0.1,0.6)
+                    & \makecell[l]{\textsc{chapter 5}\\\\}     & \makecell[l]   {Model-Agnostic Out-of-Distribution Detection\\ Using Combined Statistical Tests} & \tikzmarkend{modelagnostic-1} \\
                     \addlinespace[0.5em]
                     \addlinespace[0.5em]
-                    \tikzmarkin<4>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{c}
-                    & \makecell[tl]{\textsc{chapter 6}}     & \makecell[l]{A Brief Overview of Unsupervised Speech \\Representation Learning} & \tikzmarkend{c} \\
+
+                    \tikzmarkin<4>{brief-1}(0.1,-0.4)(-0.1,0.6)
+                    & \makecell[l]{\textsc{chapter 6}\\\\}     & \makecell[l]{A Brief Overview of Unsupervised Speech \\Representation Learning} & \tikzmarkend{brief-1} \\
                     \addlinespace[0.5em]
                     \addlinespace[0.5em]
-                    \tikzmarkin<5>[below right offset={0.1,-0.4},above left offset={-0.1,0.4}]{d}
-                    & \textsc{chapter 7}     & \makecell[l]{Benchmarking Latent Variable Models for Speech} & \tikzmarkend{d} \\
+
+                    \tikzmarkin<5>{benchmarking-1}(0.1,-0.3)(-0.1,0.4)
+                    & \textsc{chapter 7}     & \makecell[l]{Benchmarking Latent Variable Models for Speech} & \tikzmarkend{benchmarking-1} \\
                     \addlinespace[0.5em]
                     \addlinespace[0.5em]
-                    \tikzmarkin<6>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{e}
-                    & \makecell[tl]{\textsc{chapter 8}}     & \makecell[l]{Automated Medical Coding on MIMIC-III and \\MIMIC-IV: A Critical Review and Replicability Study} & \tikzmarkend{e} \\
+
+                    \tikzmarkin<6>{automated-1}(0.1,-0.4)(-0.1,0.6)
+                    & \makecell[l]{\textsc{chapter 8}\\\\}     & \makecell[l]{Automated Medical Coding on MIMIC-III and \\MIMIC-IV: A Critical Review and Replicability Study} & \tikzmarkend{automated-1} \\
                     \addlinespace[0.5em]
                     \addlinespace[0.5em]
-                    \tikzmarkin<7>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{f}
-                    & \makecell[tl]{\textsc{chapter 9}}     & \makecell[l]{A Retrospective Study on Machine Learning-\\Assisted Stroke Recognition for Medical Helpline Calls} & \tikzmarkend{f} \\
+
+                    \tikzmarkin<7>{retrospective-1}(0.1,-0.4)(-0.1,0.6)
+                    & \makecell[l]{\textsc{chapter 9}\\\\}     & \makecell[l]{A Retrospective Study on Machine Learning-\\Assisted Stroke Recognition for Medical Helpline Calls} & \tikzmarkend{retrospective-1} \\
                     \addlinespace[0.5em]
@@
                     \addlinespace[0.5em]
-                    & \textsc{chapter 10}    & \makecell[l]{Discussion and Conclusion} & \\
+
+                    & {\color<2->{black!20} \textsc{chapter 10}}    & {\color<2->{black!20} \makecell[l]{Discussion and Conclusion}} & \\
                 \end{tabular}
@@
         \begin{column}{0.4\textwidth}
+
+            \tikzstyle{mybox} = [fill, rounded corners, minimum width=140, minimum height=20]
+            \tikzstyle{myrect} = [fill]
+
             \begin{figure}
@@
                 \begin{overprint}
-                    \onslide<1>
+                    \onslide<1>  % Overview
                     \begin{tikzpicture}
-                        \tikzstyle{mybox} = [fill, rounded corners, minimum width=5cm, minimum height=0.8cm]
-                        \tikzstyle{myrect} = [fill, minimum height=0.6cm]
 
                         \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
-                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
-                            \node [myrect, fill=presentation_orange_shaded] at (\x,5) {};
-                        \node [] at (0,5) {\textsc{predictions}};
+                        \foreach \i in {-11,...,11} {
+                            \fill[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [] at (0,5) {\small\textsc{predictions}};
 
-                        \node [mybox, fill=black!10] at (0,4) (downstream) {\textsc{downstream model}};
+                        \node [mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};
 
                         \node [mybox, fill=presentation_blue] at (0,3) (text) {};
-                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
-                            \node [myrect, fill=presentation_blue_shaded] at (\x,3) {};
-                        \node [] at (0,3) {\textsc{text}};
+                        \foreach \i in {-11,...,11} {
+                            \fill[myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [] at (0,3) {\small\textsc{text}};
 
-                        \node [mybox, fill=black!10] at (0,2) (recognition) {\textsc{speech recognition}};
+                        \node [mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};
 
-                        \node [mybox, fill=black!10] at (0,1) (representation) {\textsc{representation learning}};
+                        \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};
 
                         \node [mybox, fill=presentation_green] at (0,0) (speech) {};
-                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
-                            \node [myrect, fill=presentation_green_shaded] at (\x,0) {};
-                        \node [] at (0,0) {\textsc{speech}};
+                        \foreach \i in {-11,...,11} {
+                            \fill[myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
+                        }
+                        \node [] at (0,0) {\small\textsc{speech}};
 
-                        \node [draw, rectangle, rounded corners, line width=0.1cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
-                        \node [above] at (uncertainty.north) {\textsc{uncertainty}};
+                        \node [draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
+                        \node [above] at (uncertainty.north) {\small\textsc{uncertainty}};
+
                     \end{tikzpicture}
@@
                     \onslide<2>  % Hierarchical
-                    \begin{tikzpicture}
-                        \tikzstyle{mybox} = [fill, rounded corners, minimum width=5cm, minimum height=0.8cm]
-                        \tikzstyle{myrect} = [fill, minimum height=0.6cm]
+                    \begin{tikzpicture}[x=1cm,y=1cm]
+                        \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
+                        \foreach \i in {-11,...,11} {
+                            \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [nearly transparent] at (0,5) {\small\textsc{predictions}};
 
-                        \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
-                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
-                            \node [nearly transparent, myrect, fill=presentation_orange_shaded] at (\x,5) {};
-                        \node [nearly transparent] at (0,5) {\textsc{predictions}};
-
-                        \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\textsc{downstream model}};
+                        \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};
 
                         \node [nearly transparent, mybox, fill=presentation_blue] at (0,3) (text) {};
-                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
-                            \node [nearly transparent, myrect, fill=presentation_blue_shaded] at (\x,3) {};
-                        \node [nearly transparent] at (0,3) {\textsc{text}};
+                        \foreach \i in {-11,...,11} {
+                            \fill[nearly transparent, myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [nearly transparent] at (0,3) {\small\textsc{text}};
 
-                        \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\textsc{speech recognition}};
+                        \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};
 
-                        \node [mybox, fill=black!10] at (0,1) (representation) {\textsc{representation learning}};
+                        \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};
 
                         \node [nearly transparent, mybox, fill=presentation_green] at (0,0) (speech) {};
-                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
-                            \node [nearly transparent, myrect, fill=presentation_green_shaded] at (\x,0) {};
-                        \node [nearly transparent] at (0,0) {\textsc{speech}};
+                        \foreach \i in {-11,...,11} {
+                            \fill[nearly transparent, myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
+                        }
+                        \node [nearly transparent] at (0,0) {\small\textsc{speech}};
 
-                        \node [draw, rectangle, rounded corners, line width=0.1cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
-                        \node [above] at (uncertainty.north) {\textsc{uncertainty}};
+                        \node [draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
+                        \node [above] at (uncertainty.north) {\small\textsc{uncertainty}};
                     \end{tikzpicture}
@@
                     \onslide<3>  % Model-Agnostic
-                    \begin{tikzpicture}
-                        \tikzstyle{mybox} = [fill, rounded corners, minimum width=5cm, minimum height=0.8cm]
-                        \tikzstyle{myrect} = [fill, minimum height=0.6cm]
+                    \begin{tikzpicture}[x=1cm,y=1cm]
+                        \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
+                        \foreach \i in {-11,...,11} {
+                            \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [nearly transparent] at (0,5) {\small\textsc{predictions}};
 
-                        \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
-                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
-                            \node [nearly transparent, myrect, fill=presentation_orange_shaded] at (\x,5) {};
-                        \node [nearly transparent] at (0,5) {\textsc{predictions}};
-
-                        \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\textsc{downstream model}};
+                        \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};
 
                         \node [nearly transparent, mybox, fill=presentation_blue] at (0,3) (text) {};
-                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
-                            \node [nearly transparent, myrect, fill=presentation_blue_shaded] at (\x,3) {};
-                        \node [nearly transparent] at (0,3) {\textsc{text}};
+                        \foreach \i in {-11,...,11} {
+                            \fill[nearly transparent, myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [nearly transparent] at (0,3) {\small\textsc{text}};
 
-                        \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\textsc{speech recognition}};
+                        \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};
 
-                        \node [mybox, fill=black!10] at (0,1) (representation) {\textsc{representation learning}};
+                        \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};
 
                         \node [nearly transparent, mybox, fill=presentation_green] at (0,0) (speech) {};
-                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
-                            \node [nearly transparent, myrect, fill=presentation_green_shaded] at (\x,0) {};
-                        \node [nearly transparent] at (0,0) {\textsc{speech}};
+                        \foreach \i in {-11,...,11} {
+                            \fill[nearly transparent, myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
+                        }
+                        \node [nearly transparent] at (0,0) {\small\textsc{speech}};
 
-                        \node [draw, rectangle, rounded corners, line width=0.1cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
-                        \node [above] at (uncertainty.north) {\textsc{uncertainty}};
+                        \node [draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
+                        \node [above] at (uncertainty.north) {\small\textsc{uncertainty}};
                     \end{tikzpicture}
@@
                     \onslide<4>  % Brief
-                    \begin{tikzpicture}
-                        \tikzstyle{mybox} = [fill, rounded corners, minimum width=5cm, minimum height=0.8cm]
-                        \tikzstyle{myrect} = [fill, minimum height=0.6cm]
+                    \begin{tikzpicture}[x=1cm,y=1cm]
+                        \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
+                        \foreach \i in {-11,...,11} {
+                            \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [nearly transparent] at (0,5) {\small\textsc{predictions}};
 
-                        \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
-                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
-                            \node [nearly transparent, myrect, fill=presentation_orange_shaded] at (\x,5) {};
-                        \node [nearly transparent] at (0,5) {\textsc{predictions}};
-
-                        \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\textsc{downstream model}};
+                        \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};
 
                         \node [nearly transparent, mybox, fill=presentation_blue] at (0,3) (text) {};
-                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
-                            \node [nearly transparent, myrect, fill=presentation_blue_shaded] at (\x,3) {};
-                        \node [nearly transparent] at (0,3) {\textsc{text}};
+                        \foreach \i in {-11,...,11} {
+                            \fill[nearly transparent, myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [nearly transparent] at (0,3) {\small\textsc{text}};
 
-                        \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\textsc{speech recognition}};
+                        \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};
 
-                        \node [mybox, fill=black!10] at (0,1) (representation) {\textsc{representation learning}};
+                        \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};
 
                         \node [mybox, fill=presentation_green] at (0,0) (speech) {};
-                        \foreach \x in {-2.0,-1.5,-1.0,-0.5,0,0.5,1.0,1.5,2.0}
-                            \node [myrect, fill=presentation_green_shaded] at (\x,0) {};
-                        \node [] at (0,0) {\textsc{speech}};
+                        \foreach \i in {-11,...,11} {
+                            \fill[myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
+                        }
+                        \node [] at (0,0) {\small\textsc{speech}};
 
-                        \node [nearly transparent, draw, rectangle, rounded corners, line width=0.1cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
-                        \node [nearly transparent, above] at (uncertainty.north) {\textsc{uncertainty}};
+                        \node [nearly transparent, draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
+                        \node [nearly transparent, above] at (uncertainty.north) {\small\textsc{uncertainty}};
                     \end{tikzpicture}
 
-                    % \onslide<6>
-                    % \begin{tikzpicture}
-                    %     % Define box style
-                    %     \tikzstyle{mybox} = [fill, rounded corners, minimum width=5cm, minimum height=0.8cm]
-                    %     % Define rectangle style
-                    %     \tikzstyle{myrect} = [fill, minimum width=0.25cm, minimum height=0.6cm]
+                    \onslide<5>  % Benchmarking
+                    \begin{tikzpicture}[x=1cm,y=1cm]
+                        \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
+                        \foreach \i in {-11,...,11} {
+                            \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [nearly transparent] at (0,5) {\small\textsc{predictions}};
 
-                    %     \node [mybox, fill=presentation_orange] at (0,5) (predictions) {Predictions};
-                    %     \node [mybox, fill=black!10] at (0,4) (downstream) {Downstream Model};
+                        \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};
 
-                    %     \node [mybox, fill=presentation_blue] at (0,3) (text) {};
-                    %     \node [myrect, fill=presentation_blue_shaded] at (-2.0,3) {};
-                    %     \node [myrect, fill=presentation_blue_shaded] at (-1.5,3) {};
-                    %     \node [myrect, fill=presentation_blue_shaded] at (-1.0,3) {};
-                    %     \node [myrect, fill=presentation_blue_shaded] at (-0.5,3) {};
-                    %     \node [myrect, fill=presentation_blue_shaded] at (0,3) {};
-                    %     \node [myrect, fill=presentation_blue_shaded] at (0.5,3) {};
-                    %     \node [myrect, fill=presentation_blue_shaded] at (1.0,3) {};
-                    %     \node [myrect, fill=presentation_blue_shaded] at (1.5,3) {};
-                    %     \node [myrect, fill=presentation_blue_shaded] at (2.0,3) {};
-                    %     \node at (0,3) {Text};
+                        \node [mybox, fill=presentation_blue] at (0,3) (text) {};
+                        \foreach \i in {-11,...,11} {
+                            \fill[myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [] at (0,3) {\small\textsc{text}};
 
-                    %     \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {Speech Recognition};
-                    %     \node [nearly transparent, mybox, fill=black!10] at (0,1) (representation) {Representation Learning};
-                    %     %
-                    %     \node [nearly transparent, mybox, fill=presentation_green] at (0,0) (speech) {};
-                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (-2.0,0) {};
-                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (-1.5,0) {};
-                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (-1.0,0) {};
-                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (-0.5,0) {};
-                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (0,0) {};
-                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (0.5,0) {};
-                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (1.0,0) {};
-                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (1.5,0) {};
-                    %     \node [nearly transparent, myrect, fill=presentation_green_shaded] at (2.0,0) {};
-                    %     \node [nearly transparent] at (0,0) {Speech};
-                    % \end{tikzpicture}
+                        \node [mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};
+
+                        \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};
+
+                        \node [mybox, fill=presentation_green] at (0,0) (speech) {};
+                        \foreach \i in {-11,...,11} {
+                            \fill[myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
+                        }
+                        \node [] at (0,0) {\small\textsc{speech}};
+
+                        \node [nearly transparent, draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
+                        \node [nearly transparent, above] at (uncertainty.north) {\small\textsc{uncertainty}};
+                    \end{tikzpicture}
+
+                    \onslide<6>  % Automated
+                    \begin{tikzpicture}[x=1cm,y=1cm]
+                        \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
+                        \foreach \i in {-11,...,11} {
+                            \fill[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [] at (0,5) {\small\textsc{predictions}};
+
+                        \node [mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};
+
+                        \node [mybox, fill=presentation_blue] at (0,3) (text) {};
+                        \foreach \i in {-11,...,11} {
+                            \fill[myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [] at (0,3) {\small\textsc{text}};
+
+                        \node [mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};
+
+                        \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};
+
+                        \node [mybox, fill=presentation_green] at (0,0) (speech) {};
+                        \foreach \i in {-11,...,11} {
+                            \fill[myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
+                        }
+                        \node [] at (0,0) {\small\textsc{speech}};
+
+                        \node [nearly transparent, draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
+                        \node [nearly transparent, above] at (uncertainty.north) {\small\textsc{uncertainty}};
+                    \end{tikzpicture}
+
+                    \onslide<7>  % Retrospective
+                    \begin{tikzpicture}[x=1cm,y=1cm]
+                        \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
+                        \foreach \i in {-11,...,11} {
+                            \fill[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [] at (0,5) {\small\textsc{predictions}};
+
+                        \node [mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};
+
+                        \node [mybox, fill=presentation_blue] at (0,3) (text) {};
+                        \foreach \i in {-11,...,11} {
+                            \fill[myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
+                        }
+                        \node [] at (0,3) {\small\textsc{text}};
+
+                        \node [mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};
+
+                        \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};
+
+                        \node [mybox, fill=presentation_green] at (0,0) (speech) {};
+                        \foreach \i in {-11,...,11} {
+                            \fill[myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
+                        }
+                        \node [] at (0,0) {\small\textsc{speech}};
+
+                        \node [draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
+                        \node [above] at (uncertainty.north) {\small\textsc{uncertainty}};
+                    \end{tikzpicture}
 
@@
                     % \tikzmarkin<3>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{b}
-                    % & \makecell[tl]{\textsc{chapter 5}}     & \makecell[l]{Model-Agnostic Out-of-Distribution Detection\\ Using Combined Statistical Tests} & \tikzmarkend{b} \\
+                    % & \makecell[l]{\textsc{chapter 5}}     & \makecell[l]{Model-Agnostic Out-of-Distribution Detection\\ Using Combined Statistical Tests} & \tikzmarkend{b} \\
                     \addlinespace[0.5em]
@@
                     \tikzmarkin<4>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{c}
-                    & \makecell[tl]{\textsc{chapter 6}}     & \makecell[l]{A Brief Overview of Unsupervised Speech \\Representation Learning} & \tikzmarkend{c} \\
+                    & \makecell[l]{\textsc{chapter 6}}     & \makecell[l]{A Brief Overview of Unsupervised Speech \\Representation Learning} & \tikzmarkend{c} \\
                     \addlinespace[0.5em]
@@
                     % \tikzmarkin<6>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{e}
-                    % & \makecell[tl]{\textsc{chapter 8}}     & \makecell[l]{Automated Medical Coding on MIMIC-III and \\MIMIC-IV: A Critical Review and Replicability Study} & \tikzmarkend{e} \\
+                    % & \makecell[l]{\textsc{chapter 8}}     & \makecell[l]{Automated Medical Coding on MIMIC-III and \\MIMIC-IV: A Critical Review and Replicability Study} & \tikzmarkend{e} \\
                     % \addlinespace[0.5em]
@@
                     \tikzmarkin<7>[below right offset={0.1,-0.6},above left offset={-0.1,0.6}]{f}
-                    & \makecell[tl]{\textsc{chapter 9}}     & \makecell[l]{A Retrospective Study on Machine Learning-\\Assisted Stroke Recognition for Medical Helpline Calls} & \tikzmarkend{f} \\
+                    & \makecell[l]{\textsc{chapter 9}}     & \makecell[l]{A Retrospective Study on Machine Learning-\\Assisted Stroke Recognition for Medical Helpline Calls} & \tikzmarkend{f} \\
                     \addlinespace[0.5em]

### 2024-02-23 chapters/discussion.tex
Move figures in hierarchical

@@
 Instead, these models rely on advanced inference networks \parencite{maaloe_biva_2019} and architectural improvements \parencite{vahdat_nvae_2020, child_very_2021} to overcome the challenges of training hierarchical VAEs.
-This indicates that there might be untapped potential in consolidating the work on low-variance gradient estimators, advanced architectures and inference methods, which could help improve representation learning in VAEs.
+This indicates that there might be untapped potential in consolidating the work on low-variance gradient estimators, advanced architectures and inference methods to improve representation learning in VAEs.
 

### 2024-02-23 chapters/paper_hierarchical.tex
Move figures in hierarchical

@@
 %
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\columnwidth]{paper_hierarchical/reconstructions-front-page-4-samples-2-recons2.pdf}
-    \caption[Reconstructions of a hierarchical VAE trained on FashionMNIST.]{ Reconstructions using a hierarchical VAE trained on FashionMNIST.
-        Reconstruction quality of OOD data is comparable to in-distribution data, resulting in high likelihoods and poor OOD discrimination.
-        By sampling the $k$ bottom-most latent variables from the conditional prior distribution $p_\thetab(\zb^{(\geq l)}|\zb^{(>l)})$ (latent reconstructions) instead of the approximate posterior $q_\phib(\zb^{(>l)}|\zb^{(<l)})$, the model reconstructs from the training distribution resulting in lower $p_\thetab(\xb|\zb)$ for OOD data.
-    }
-    \label{fig_hierarchical:reconstructions-fashionmnist}
-\end{figure}
 %
@@
     \centering
-    \includegraphics[width=1\columnwidth]{paper_hierarchical/feature-correlation-heatmap2.pdf}
-    \vspace{0mm}
-    \caption[Absolute correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST.]{
-    Absolute correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST.
-    We compute the correlation between the representations of the two different models given the same data, FashionMNIST (top) and MNIST (bottom).
-    %We reduce the representations to a single number by summing over all dimensions and compute the correlation between these numbers for 256 examples.
-    %To reduce de-correlation due to the stochastic latent variables, we encode using the mode of the approximate posterior distributions.
+    \includegraphics[width=0.7\columnwidth]{paper_hierarchical/reconstructions-front-page-4-samples-2-recons2.pdf}
+    \caption[Reconstructions of a hierarchical VAE trained on FashionMNIST.]{ Reconstructions using a hierarchical VAE trained on FashionMNIST.
+        Reconstruction quality of OOD data is comparable to in-distribution data, resulting in high likelihoods and poor OOD discrimination.
+        By sampling the $k$ bottom-most latent variables from the conditional prior distribution $p_\thetab(\zb^{(\geq l)}|\zb^{(>l)})$ (latent reconstructions) instead of the approximate posterior $q_\phib(\zb^{(>l)}|\zb^{(<l)})$, the model reconstructs from the training distribution resulting in lower $p_\thetab(\xb|\zb)$ for OOD data.
     }
-    \vspace{0mm}
-    \label{fig_hierarchical:correlations-heatmap}
+    \label{fig_hierarchical:reconstructions-fashionmnist}
 \end{figure}
-
+%
 The inability to detect out-of-distribution data with deep generative models is surprising.
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=1\columnwidth]{paper_hierarchical/feature-correlation-heatmap2.pdf}
+    \vspace{0mm}
+    \caption[Absolute correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST.]{
+    Absolute correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST.
+    We compute the correlation between the representations of the two different models given the same data, FashionMNIST (top) and MNIST (bottom).
+    %We reduce the representations to a single number by summing over all dimensions and compute the correlation between these numbers for 256 examples.
+    %To reduce de-correlation due to the stochastic latent variables, we encode using the mode of the approximate posterior distributions.
+    }
+    \vspace{0mm}
+    \label{fig_hierarchical:correlations-heatmap}
+\end{figure}
+
 To shed further light on the impact of semantic versus low-level features, we look at model reconstructions of images with a hierarchical VAE (\cref{fig_hierarchical:biva-reconstructions-celeba}).
@@
 This suggests that as we fall back to the conditional priors of each layer, we are pushed closer to local modes of the modeled distribution.
+
+Finally, we look at reconstructions of out-of-distribution data.
+\Cref{fig_hierarchical:reconstructions-fashionmnist} illustrates that MNIST data is surprisingly well reconstructed by a hierarchical VAE trained on FashionMNIST.
+Similar results have been found elsewhere \parencite{xiao_likelihood_2020}.
+We repeat the previous experiment and replace inference distributions by their corresponding conditional prior, and now observe that reconstructions from higher latent layers become increasingly similar to the data on which the model was trained.
+The reliance on conditional priors seems to prevent accurate reconstruction of out-of-distribution data.
+Some details are lost on in-distribution data too, but the distinction between that and out-of-distribution data becomes more clear.
+
+\textbf{These observations lead to our main hypothesis.}
+The lowest latent variables in a hierarchical VAE learn generic features that can describe a wide range of data.
+This enables the model to achieve high rates of compression and high likelihoods, even on out-of-distribution data as long as the learned low-level features are appropriate.
+We further suggest that OOD data are in-distribution with respect to these low-level features, but not with respect to semantic ones.
 
@@
 \end{figure}
-Finally, we look at reconstructions of out-of-distribution data.
-\cref{fig_hierarchical:reconstructions-fashionmnist} illustrates that MNIST data is surprisingly well reconstructed by a hierarchical VAE trained on FashionMNIST.
-Similar results have been found elsewhere \parencite{xiao_likelihood_2020}.
-We repeat the previous experiment and replace inference distributions by their corresponding conditional prior, and now observe that reconstructions from higher latent layers become increasingly similar to the data on which the model was trained.
-The reliance on conditional priors seems to prevent accurate reconstruction of out-of-distribution data.
-Some details are lost on in-distribution data too, but the distinction between that and out-of-distribution data becomes more clear.
-
-\textbf{These observations lead to our main hypothesis.}
-The lowest latent variables in a hierarchical VAE learn generic features that can describe a wide range of data.
-This enables the model to achieve high rates of compression and high likelihoods, even on out-of-distribution data as long as the learned low-level features are appropriate.
-We further suggest that OOD data are in-distribution with respect to these low-level features, but not with respect to semantic ones.
 
@@
         \node[latent,right=.75cm of z1_inf](z2_inf){$\zb^{(2)}$};
-        \node[above=of z2_inf, xshift=-1.cm] (phi) {$q_\phib(\zb|\xb)$};
+        \node[above=0.46cm of z1_inf] (phi) {$q_\phib(\zb|\xb)$};
 
@@
         \node[obs,right=0.75cm of z1_gen] (x_gen) {$\xb$};
-        \node[above=of z1_gen] (theta) {$p_\thetab(\xb,\zb)$};
+        \node[above=0.5cm of z1_gen] (theta) {$p_\thetab(\xb,\zb)$};
 

### 2024-02-24 presentation/parts/introduction.tex
WIP

@@
         \item <1,2>It is complex, involving multiple participants, different contexts, and different purposes.
-        \item <2> {\color{dtured} } Failure of communication is a leading cause of medical error contributing to two out of three adverse events \cite{starmer_changes_2014}.
-        \item <2> A considerable fraction of all hospital admissions had preventable adverse outcomes (9\% to 16.6\% in AU, NZ, UK, DK) \cite{carver_medical_2024}.
+        \vspace{1em}
+        \item <2> {\color{dtured} Adverse events}: Failure of communication is a leading cause of medical error contributing to two out of three adverse events \cite{starmer_changes_2014}.
+        \item <2> {\color{dtured} Preventability}: A considerable fraction of all hospital admissions had preventable adverse outcomes (9\% to 16.6\% in AU, NZ, UK, DK) \cite{carver_medical_2024}.
     \end{itemize}
@@
     \begin{itemize}
-        \item <1> {\color{dtured}Assist} with documentation.
-        \item <1> {\color{dtured}Augment} communication.
-        \item <1> {\color{dtured}Improve} decision-making.
+        \item <1-> {\color{dtured}Assist} with documentation.
+        \item <1-> {\color{dtured}Augment} communication.
+        \item <1-> {\color{dtured}Improve} decision-making.
         \vspace{1em}
-        \item <1> {\color{dtured}Reduce} errors.
-        \item <1> {\color{dtured}Save} time.
+        \item <2-> {\color{dtured}Reduce} errors.
+        \item <2-> {\color{dtured}Free up} time for patient care.
     \end{itemize}
@@
 \begin{frame}
-    \frametitle{Reliability of machine learning}
+    \frametitle{Reliability of machine learning systems}
     \begin{itemize}
-        \item <1> {\color{dtured}Data}: Privacy, quality, quantity, diversity.
-        \item <1> {\color{dtured}Interpretability}: Trust, ethics, regulation.
-        \item <1> {\color{dtured}Explainability}: Transparency, accountability.
-        \item <1> {\color{dtured}Robustness}: Adversarial attacks, distribution shift.
-        \item <1> {\color{dtured}Bias}: Fairness, transparency.
-        \item <1> {\color{dtured}Complexity}: Context, domain, language, culture.
+        \item <1-> {\color{dtured}Data}: Privacy, quality, quantity, diversity, bias.
+        \item <2-> {\color{dtured}Task}: Context, domain, language, culture.
+        \item <3-> {\color{dtured}Interpretability}: Trust, ethics, regulation, transparency, accountability.
+        \item <3-> {\color{dtured}Explainability}: 
+        \item <4-> {\color{dtured}Fairness} in treatment of different groups of people.
+        \item <5-> {\color{dtured}Robustness} to noise, outliers, adversarial attacks, and distribution shift.
     \end{itemize}

### 2024-02-24 presentation/parts/overview_brief.tex
WIP

% !TEX root = ../presentation.tex
% !BIB program = biber
% !TEX program = xelatex

\section{Overview}

\begin{frame}
    \frametitle{Presentation}
    \begin{columns}
        \begin{column}{0.6\textwidth}
            \begin{table}
                \resizebox{\textwidth}{!}{%
                \begin{tabular}{l l l l}
                    & {\color<1->{black!20}   \makecell[l]{\textsc{\itshape chapter 1-3}}} & {\color<1->{black!20}    \bfseries \scshape \Large \makecell[l]{introduction, research questions, and background}} & \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]

                    \tikzmarkin<1->{hierarchical-8}(0.1,-0.3)(-0.1,0.4)
                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 4}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{hierarchical vaes know what they don't know}} & \tikzmarkend{hierarchical-8} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    % \tikzmarkin<1->{modelagnostic-8}(0.1,-0.4)(-0.1,0.6)
                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-8} \\
                    % \addlinespace[0.5em]
                    % \addlinespace[0.5em]

                    \tikzmarkin<1->{brief-8}(0.1,-0.4)(-0.1,0.6)
                    & {\color<-1>{black!20}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<-1>{black!20} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-8} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    % \tikzmarkin<1->{benchmarking-8}(0.1,-0.3)(-0.1,0.4)
                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-8} \\
                    % \addlinespace[0.5em]
                    % \addlinespace[0.5em]

                    % \tikzmarkin<1->{automated-8}(0.1,-0.4)(-0.1,0.6)
                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-8} \\
                    % \addlinespace[0.5em]
                    % \addlinespace[0.5em]

                    \tikzmarkin<1->{retrospective-8}(0.1,-0.4)(-0.1,0.6)
                    & {\color<1->{black!20}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-8} \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]

                    & {\color<1->{black!20}\makecell[l]{\textsc{\itshape chapter 10}}}        & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{discussion and conclusion}} & \\
                \end{tabular}
                }
            \end{table}
        \end{column}
        \begin{column}{0.4\textwidth}
            \tikzstyle{mybox} = [fill, rounded corners, minimum width=140, minimum height=20]
            \tikzstyle{myrect} = [fill]
            \begin{figure}
                \centering
                \begin{overprint}
                    \onslide<1>\tikzhierarchical
                    \onslide<2>\tikzbrief
                \end{overprint}
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

### 2024-02-24 presentation/parts/overview_discussion.tex
WIP

--- 2024-02-24 presentation/parts/overview_brief.tex
@@
                     \tikzmarkin<1->{hierarchical-8}(0.1,-0.3)(-0.1,0.4)
-                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 4}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{hierarchical vaes know what they don't know}} & \tikzmarkend{hierarchical-8} \\
+                    & {\color<1->{black!20}\makecell[l]{\textsc{\itshape chapter 4}}}         & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{hierarchical vaes know what they don't know}} & \tikzmarkend{hierarchical-8} \\
                     \addlinespace[0.5em]
@@
                     \tikzmarkin<1->{brief-8}(0.1,-0.4)(-0.1,0.6)
-                    & {\color<-1>{black!20}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<-1>{black!20} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-8} \\
+                    & {\color<1->{black!20}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-8} \\
                     \addlinespace[0.5em]
@@
                     \tikzmarkin<1->{retrospective-8}(0.1,-0.4)(-0.1,0.6)
-                    & {\color<1->{black!20}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-8} \\
+                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-8} \\
                     \addlinespace[0.5em]
@@
 
-                    & {\color<1->{black!20}\makecell[l]{\textsc{\itshape chapter 10}}}        & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{discussion and conclusion}} & \\
+                    & {\color<-1>{black!20}\makecell[l]{\textsc{\itshape chapter 10}}}        & {\color<-1>{black!20} \bfseries \scshape \Large \makecell[l]{discussion and conclusion}} & \\
                 \end{tabular}
@@
                 \begin{overprint}
-                    \onslide<1>\tikzhierarchical
-                    \onslide<2>\tikzbrief
+                    \onslide<1>\tikzretrospective
+                    \onslide<2>\tikzoverview
                 \end{overprint}

### 2024-02-24 presentation/parts/overview_hierarchical.tex
WIP

% !TEX root = ../presentation.tex
% !BIB program = biber
% !TEX program = xelatex

\section{Overview}

\begin{frame}
    \frametitle{Presentation}
    \begin{columns}
        \begin{column}{0.6\textwidth}
            \begin{table}
                \resizebox{\textwidth}{!}{%
                \begin{tabular}{l l l l}
                    & {\color<2->{black!20}   \makecell[l]{\textsc{\itshape chapter 1-3}}} & {\color<2->{black!20}    \bfseries \scshape \Large \makecell[l]{introduction, research questions, and background}} & \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]

                    \tikzmarkin<1->{hierarchical-3}(0.1,-0.3)(-0.1,0.4)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 4}}}         & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{hierarchical vaes know what they don't know}} & \tikzmarkend{hierarchical-3} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    % \tikzmarkin<1->{modelagnostic-3}(0.1,-0.4)(-0.1,0.6)
                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-3} \\
                    % \addlinespace[0.5em]
                    % \addlinespace[0.5em]

                    \tikzmarkin<1->{brief-3}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black!20}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-3} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    % \tikzmarkin<1->{benchmarking-3}(0.1,-0.3)(-0.1,0.4)
                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-3} \\
                    % \addlinespace[0.5em]
                    % \addlinespace[0.5em]

                    % \tikzmarkin<1->{automated-3}(0.1,-0.4)(-0.1,0.6)
                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-3} \\
                    % \addlinespace[0.5em]
                    % \addlinespace[0.5em]

                    \tikzmarkin<1->{retrospective-3}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-3} \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]

                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 10}}}        & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{discussion and conclusion}} & \\
                \end{tabular}
                }
            \end{table}
        \end{column}
        \begin{column}{0.4\textwidth}
            \tikzstyle{mybox} = [fill, rounded corners, minimum width=140, minimum height=20]
            \tikzstyle{myrect} = [fill]
            \begin{figure}
                \centering
                \begin{overprint}
                    \onslide<1>\tikzoverview
                    \onslide<2>\tikzhierarchical
                \end{overprint}
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

### 2024-02-24 presentation/parts/overview_presentation.tex
WIP

% !TEX root = ../presentation.tex
% !BIB program = biber
% !TEX program = xelatex

\hfsetfillcolor{presentation_green}
\hfsetbordercolor{presentation_green}

\newcommand{\tikzoverview}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [] at (0,5) {\small\textsc{predictions}};

            \node [mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};

            \node [mybox, fill=presentation_blue] at (0,3) (text) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
            }
            \node [] at (0,3) {\small\textsc{text}};

            \node [mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};

            \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};

            \node [mybox, fill=presentation_green] at (0,0) (speech) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
            }
            \node [] at (0,0) {\small\textsc{speech}};

            \node [draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
            \node [above] at (uncertainty.north) {\small\textsc{uncertainty}};
        \end{tikzpicture}
    }
}

\newcommand{\tikzhierarchical}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

            \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};

            \node [nearly transparent, mybox, fill=presentation_blue] at (0,3) (text) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,3) {\small\textsc{text}};

            \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};

            \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};

            \node [nearly transparent, mybox, fill=presentation_green] at (0,0) (speech) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,0) {\small\textsc{speech}};

            \node [draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
            \node [above] at (uncertainty.north) {\small\textsc{uncertainty}};
        \end{tikzpicture}
    }
}

\newcommand{\tikzmodelagnostic}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

            \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};

            \node [nearly transparent, mybox, fill=presentation_blue] at (0,3) (text) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,3) {\small\textsc{text}};

            \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};

            \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};

            \node [nearly transparent, mybox, fill=presentation_green] at (0,0) (speech) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,0) {\small\textsc{speech}};

            \node [draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
            \node [above] at (uncertainty.north) {\small\textsc{uncertainty}};
        \end{tikzpicture}
    }
}

\newcommand{\tikzbrief}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

            \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};

            \node [nearly transparent, mybox, fill=presentation_blue] at (0,3) (text) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,3) {\small\textsc{text}};

            \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};

            \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};

            \node [mybox, fill=presentation_green] at (0,0) (speech) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
            }
            \node [] at (0,0) {\small\textsc{speech}};

            \node [nearly transparent, draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
            \node [nearly transparent, above] at (uncertainty.north) {\small\textsc{uncertainty}};
        \end{tikzpicture}
    }
}

\newcommand{\tikzbenchmarking}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

            \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};

            \node [mybox, fill=presentation_blue] at (0,3) (text) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
            }
            \node [] at (0,3) {\small\textsc{text}};

            \node [mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};

            \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};

            \node [mybox, fill=presentation_green] at (0,0) (speech) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
            }
            \node [] at (0,0) {\small\textsc{speech}};

            \node [nearly transparent, draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
            \node [nearly transparent, above] at (uncertainty.north) {\small\textsc{uncertainty}};
        \end{tikzpicture}
    }
}

\newcommand{\tikzautomated}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

            \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};

            \node [nearly transparent, mybox, fill=presentation_blue] at (0,3) (text) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,3) {\small\textsc{text}};

            \node [nearly transparent, mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};

            \node [nearly transparent, mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};

            \node [nearly transparent, mybox, fill=presentation_green] at (0,0) (speech) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,0) {\small\textsc{speech}};

            \node [draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
            \node [above] at (uncertainty.north) {\small\textsc{uncertainty}};
        \end{tikzpicture}
    }
}

\newcommand{\tikzretrospective}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [] at (0,5) {\small\textsc{predictions}};

            \node [mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};

            \node [mybox, fill=presentation_blue] at (0,3) (text) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
            }
            \node [] at (0,3) {\small\textsc{text}};

            \node [mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};

            \node [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};

            \node [mybox, fill=presentation_green] at (0,0) (speech) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
            }
            \node [] at (0,0) {\small\textsc{speech}};

            \node [draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
            \node [above] at (uncertainty.north) {\small\textsc{uncertainty}};
        \end{tikzpicture}
    }
}

\section{Overview}

% Go through the papers of the thesis while showing the graphic of their themes.
\begin{frame}
    \frametitle{Thesis}
    \begin{columns}
        \begin{column}{0.6\textwidth}
            \begin{table}
                \resizebox{\textwidth}{!}{%
                \begin{tabular}{l l l l}
                    & {\color<2->{black!20}   \makecell[l]{\textsc{\itshape chapter 1-3}}} & {\color<2->{black!20}    \bfseries \scshape \Large \makecell[l]{introduction, research questions, and background}} & \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]

                    \tikzmarkin<2>{hierarchical-1}(0.1,-0.3)(-0.1,0.4)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 4}}}         & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{hierarchical vaes know what they don't know}} & \tikzmarkend{hierarchical-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<3>{modelagnostic-1}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<4>{brief-1}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<5>{benchmarking-1}(0.1,-0.3)(-0.1,0.4)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<6>{automated-1}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<7>{retrospective-1}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-1} \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]

                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 10}}}        & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{discussion and conclusion}} & \\
                \end{tabular}
                }
            \end{table}
        \end{column}
        \begin{column}{0.4\textwidth}
            \tikzstyle{mybox} = [fill, rounded corners, minimum width=140, minimum height=20]
            \tikzstyle{myrect} = [fill]
            \begin{figure}
                \centering
                \begin{overprint}
                    \onslide<1>\tikzoverview
                    \onslide<2>\tikzhierarchical
                    \onslide<3>\tikzmodelagnostic
                    \onslide<4>\tikzbrief
                    \onslide<5>\tikzbenchmarking
                    \onslide<6>\tikzautomated
                    \onslide<7>\tikzretrospective
                \end{overprint}
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

\hfsetfillcolor{white}
\hfsetbordercolor{white}

% Grey out scoped-out papers in the overview
\begin{frame}
    \frametitle{Presentation}
    \begin{columns}
        \begin{column}{0.6\textwidth}
            \begin{table}
                \resizebox{\textwidth}{!}{%
                \begin{tabular}{l l l l}
                    & {\color<2->{black}   \makecell[l]{\textsc{\itshape chapter 1-3}}} & {\color<2->{black}    \bfseries \scshape \Large \makecell[l]{introduction, research questions, and background}} & \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]

                    \tikzmarkin<1->{hierarchical-2}(0.1,-0.3)(-0.1,0.4)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 4}}}         & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{hierarchical vaes know what they don't know}} & \tikzmarkend{hierarchical-2} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<1->{modelagnostic-2}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-2} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<1->{brief-2}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-2} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<1->{benchmarking-2}(0.1,-0.3)(-0.1,0.4)
                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-2} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<1->{automated-2}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-2} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<1->{retrospective-2}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-2} \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]

                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 10}}}        & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{discussion and conclusion}} & \\
                \end{tabular}
                }
            \end{table}
        \end{column}
        \begin{column}{0.4\textwidth}
            \tikzstyle{mybox} = [fill, rounded corners, minimum width=140, minimum height=20]
            \tikzstyle{myrect} = [fill]
            \begin{figure}
                \centering
                \begin{overprint}
                    \onslide<1>\tikzoverview
                \end{overprint}
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

### 2024-02-24 presentation/parts/overview_retrospective.tex
WIP

--- 2024-02-24 presentation/parts/overview_discussion.tex
@@
                     \tikzmarkin<1->{brief-8}(0.1,-0.4)(-0.1,0.6)
-                    & {\color<1->{black!20}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-8} \\
+                    & {\color<2->{black!20}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-8} \\
                     \addlinespace[0.5em]
@@
                     \tikzmarkin<1->{retrospective-8}(0.1,-0.4)(-0.1,0.6)
-                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-8} \\
+                    & {\color<-1>{black!20}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<-1>{black!20} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-8} \\
                     \addlinespace[0.5em]
@@
 
-                    & {\color<-1>{black!20}\makecell[l]{\textsc{\itshape chapter 10}}}        & {\color<-1>{black!20} \bfseries \scshape \Large \makecell[l]{discussion and conclusion}} & \\
+                    & {\color<1->{black!20}\makecell[l]{\textsc{\itshape chapter 10}}}        & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{discussion and conclusion}} & \\
                 \end{tabular}
@@
                 \begin{overprint}
-                    \onslide<1>\tikzretrospective
-                    \onslide<2>\tikzoverview
+                    \onslide<1>\tikzbrief
+                    \onslide<2>\tikzretrospective
                 \end{overprint}

### 2024-02-25 presentation/parts/introduction.tex
WIP

--- 2024-02-18 presentation/parts/introduction.tex
@@
 \begin{frame}
+    % Background on Corti and the problem of medical communication, diagnosis, and documentation.
+    \frametitle{Background}
+    \begin{columns}
+        \begin{column}{0.6\textwidth}
+            FILL ME OUT. \\
+            Background on Corti, motivation for project.
+        \end{column}
+        \begin{column}{0.4\textwidth}
+        \end{column}
+    \end{columns}
+\end{frame}
+
+\begin{frame}
     \frametitle{Healthcare}
@@
         \noindent\centering
-        Healthcare is the improvement of health via the {\color{dtured}prevention}, {\color{dtured}diagnosis}, {\color{dtured}treatment}, {\color{dtured}amelioration} or {\color{dtured}cure} of {\color{theme-green}disease}, {\color{theme-green}illness}, {\color{theme-green}injury}, and {\color{theme-green}other physical and mental impairments} in people.
+        Healthcare is the improvement of health via the \highlight{prevention}, \highlight{diagnosis}, \highlight{treatment}, \highlight{amelioration} or \highlight{cure} of {\color{theme-green}disease}, {\color{theme-green}illness}, {\color{theme-green}injury}, and {\color{theme-green}other physical and mental impairments} in people.
     \end{quotation}
@@
         \item <1,2>It is complex, involving multiple participants, different contexts, and different purposes.
-        \item <2> {\color{dtured} } Failure of communication is a leading cause of medical error contributing to two out of three adverse events \cite{starmer_changes_2014}.
-        \item <2> A considerable fraction of all hospital admissions had preventable adverse outcomes (9\% to 16.6\% in AU, NZ, UK, DK) \cite{carver_medical_2024}.
+        \vspace{1em}
+        \item <2> \highlight{Adverse events}: Failure of communication is a leading cause of medical error contributing to two out of three adverse events \cite{starmer_changes_2014}.
+        \item <2> \highlight{Preventability}: A considerable fraction of all hospital admissions had preventable adverse outcomes (9\% to 16.6\% in AU, NZ, UK, DK) \cite{carver_medical_2024}.
     \end{itemize}
@@
         \vspace{1em}
-        \item <2> {\color{dtured}Time}: Physicians spend 34-37\% of their time on documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
-        \item <2> {\color{dtured}Quality}: Discharge summaries rarely meet all timeline, transmission, and content criteria.
+        \item <2> \highlight{Time-consuming}: Physicians spend 34-37\% of their time on documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
+        \item <2> \highlight{Varying quality}: Discharge summaries rarely meet all timeline, transmission, and content criteria.
 \cite{horwitz_comprehensive_2013}\footnote{Outpatient visits, Yale-New Haven Hospital.}
@@
     \begin{itemize}
-        \item <1> {\color{dtured}Assist} with documentation.
-        \item <1> {\color{dtured}Augment} communication.
-        \item <1> {\color{dtured}Improve} decision-making.
+        \item <1-> \highlight{Assist} with documentation.
+        \item <1-> \highlight{Augment} communication.
+        \item <1-> \highlight{Improve} decision-making.
         \vspace{1em}
-        \item <1> {\color{dtured}Reduce} errors.
-        \item <1> {\color{dtured}Save} time.
+        \item <2-> \highlight{Reduce} number of impact of medical errors and adverse events.
+        \item <2-> \highlight{Free up} time spent on documentation for patient care.
     \end{itemize}
@@
 \begin{frame}
-    \frametitle{Reliability of machine learning}
+    \frametitle{Reliability of machine learning systems}
     \begin{itemize}
-        \item <1> {\color{dtured}Data}: Privacy, quality, quantity, diversity.
-        \item <1> {\color{dtured}Interpretability}: Trust, ethics, regulation.
-        \item <1> {\color{dtured}Explainability}: Transparency, accountability.
-        \item <1> {\color{dtured}Robustness}: Adversarial attacks, distribution shift.
-        \item <1> {\color{dtured}Bias}: Fairness, transparency.
-        \item <1> {\color{dtured}Complexity}: Context, domain, language, culture.
+        \item <1-> \highlight{Data}: Quality, quantity, diversity, bias, privacy, ethics.
+        \item <2-> \highlight{Task}: Context, domain, language, culture, purpose.
+        \vspace{1em}
+        \item <3-> \highlight{Interpretability} of how a model works (transparency, accountability, regulation).
+        \item <4-> \highlight{Explainability} of model predictions (trust, understanding, feedback).
+        \item <5-> \highlight{Fairness} in treatment of different groups of people.
+        \item <6-> \highlight{Robustness} to noise, outliers, adversarial attacks, and distribution shift.
     \end{itemize}
-
 \end{frame}

### 2024-02-25 presentation/parts/overview_brief.tex
WIP

@@
 
-                    \tikzmarkin<1->{hierarchical-8}(0.1,-0.3)(-0.1,0.4)
-                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 4}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{hierarchical vaes know what they don't know}} & \tikzmarkend{hierarchical-8} \\
+                    \tikzmarkin<1->{hierarchical-7}(0.1,-0.3)(-0.1,0.4)
+                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 4}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{hierarchical vaes know what they don't know}} & \tikzmarkend{hierarchical-7} \\
                     \addlinespace[0.5em]
@@
 
-                    % \tikzmarkin<1->{modelagnostic-8}(0.1,-0.4)(-0.1,0.6)
-                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-8} \\
+                    % \tikzmarkin<1->{modelagnostic-7}(0.1,-0.4)(-0.1,0.6)
+                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-7} \\
                     % \addlinespace[0.5em]
@@
 
-                    \tikzmarkin<1->{brief-8}(0.1,-0.4)(-0.1,0.6)
-                    & {\color<-1>{black!20}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<-1>{black!20} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-8} \\
+                    \tikzmarkin<1->{brief-7}(0.1,-0.4)(-0.1,0.6)
+                    & {\color<-1>{black!20}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<-1>{black!20} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-7} \\
                     \addlinespace[0.5em]
@@
 
-                    % \tikzmarkin<1->{benchmarking-8}(0.1,-0.3)(-0.1,0.4)
-                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-8} \\
+                    % \tikzmarkin<1->{benchmarking-7}(0.1,-0.3)(-0.1,0.4)
+                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-7} \\
                     % \addlinespace[0.5em]
@@
 
-                    % \tikzmarkin<1->{automated-8}(0.1,-0.4)(-0.1,0.6)
-                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-8} \\
+                    % \tikzmarkin<1->{automated-7}(0.1,-0.4)(-0.1,0.6)
+                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-7} \\
                     % \addlinespace[0.5em]
@@
 
-                    \tikzmarkin<1->{retrospective-8}(0.1,-0.4)(-0.1,0.6)
-                    & {\color<1->{black!20}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-8} \\
+                    \tikzmarkin<1->{retrospective-7}(0.1,-0.4)(-0.1,0.6)
+                    & {\color<1->{black!20}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-7} \\
                     \addlinespace[0.5em]

### 2024-02-25 presentation/parts/overview_discussion.tex
WIP

@@
 
-                    \tikzmarkin<1->{hierarchical-8}(0.1,-0.3)(-0.1,0.4)
-                    & {\color<1->{black!20}\makecell[l]{\textsc{\itshape chapter 4}}}         & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{hierarchical vaes know what they don't know}} & \tikzmarkend{hierarchical-8} \\
+                    \tikzmarkin<1->{hierarchical-6}(0.1,-0.3)(-0.1,0.4)
+                    & {\color<1->{black!20}\makecell[l]{\textsc{\itshape chapter 4}}}         & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{hierarchical vaes know what they don't know}} & \tikzmarkend{hierarchical-6} \\
                     \addlinespace[0.5em]
@@
 
-                    % \tikzmarkin<1->{modelagnostic-8}(0.1,-0.4)(-0.1,0.6)
-                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-8} \\
+                    % \tikzmarkin<1->{modelagnostic-6}(0.1,-0.4)(-0.1,0.6)
+                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-6} \\
                     % \addlinespace[0.5em]
@@
 
-                    \tikzmarkin<1->{brief-8}(0.1,-0.4)(-0.1,0.6)
-                    & {\color<1->{black!20}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-8} \\
+                    \tikzmarkin<1->{brief-6}(0.1,-0.4)(-0.1,0.6)
+                    & {\color<1->{black!20}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<1->{black!20} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-6} \\
                     \addlinespace[0.5em]
@@
 
-                    % \tikzmarkin<1->{benchmarking-8}(0.1,-0.3)(-0.1,0.4)
-                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-8} \\
+                    % \tikzmarkin<1->{benchmarking-6}(0.1,-0.3)(-0.1,0.4)
+                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-6} \\
                     % \addlinespace[0.5em]
@@
 
-                    % \tikzmarkin<1->{automated-8}(0.1,-0.4)(-0.1,0.6)
-                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-8} \\
+                    % \tikzmarkin<1->{automated-6}(0.1,-0.4)(-0.1,0.6)
+                    % & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-6} \\
                     % \addlinespace[0.5em]
@@
 
-                    \tikzmarkin<1->{retrospective-8}(0.1,-0.4)(-0.1,0.6)
-                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-8} \\
+                    \tikzmarkin<1->{retrospective-6}(0.1,-0.4)(-0.1,0.6)
+                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-6} \\
                     \addlinespace[0.5em]

### 2024-02-25 presentation/parts/overview_presentation.tex
WIP

@@
         \begin{tikzpicture}[x=1cm,y=1cm]
-            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
+            \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
             \foreach \i in {-11,...,11} {
-                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
+                \fill[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
             }
-            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};
+            \node [] at (0,5) {\small\textsc{predictions}};
 
-            \node [nearly transparent, mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};
+            \node [mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};
 
-            \node [nearly transparent, mybox, fill=presentation_blue] at (0,3) (text) {};
+            \node [mybox, fill=presentation_blue] at (0,3) (text) {};
             \foreach \i in {-11,...,11} {
-                \fill[nearly transparent, myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
+                \fill[myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
             }
-            \node [nearly transparent] at (0,3) {\small\textsc{text}};
+            \node [] at (0,3) {\small\textsc{text}};
 
@@
 
-            \node [draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
-            \node [above] at (uncertainty.north) {\small\textsc{uncertainty}};
+            \node [nearly transparent, draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
+            \node [nearly transparent, above] at (uncertainty.north) {\small\textsc{uncertainty}};
         \end{tikzpicture}
@@
                     \tikzmarkin<1->{modelagnostic-2}(0.1,-0.4)(-0.1,0.6)
-                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-2} \\
+                    & {\color<2->{dtured!40}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{dtured!40} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-2} \\
                     \addlinespace[0.5em]
@@
                     \tikzmarkin<1->{benchmarking-2}(0.1,-0.3)(-0.1,0.4)
-                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-2} \\
+                    & {\color<2->{dtured!40}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{dtured!40} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-2} \\
                     \addlinespace[0.5em]
@@
                     \tikzmarkin<1->{automated-2}(0.1,-0.4)(-0.1,0.6)
-                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-2} \\
+                    & {\color<2->{dtured!40}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{dtured!40} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-2} \\
                     \addlinespace[0.5em]
@@
                 \begin{overprint}
-                    \onslide<1>\tikzoverview
+                    \onslide<1->\tikzoverview
                 \end{overprint}

### 2024-02-25 presentation/parts/presentation_hierarchical.tex
WIP

@@
 
-\section{Out-of-distribution detection}
-% \subsection{Out-of-distribution detection}
+\section{hierarchical vaes know what they don't know}
 
@@
     \begin{columns}
-
         \begin{column}{0.5\textwidth}
-
-    Out-of-distribution (OOD) detection is about enabling models to distinguish the training data distribution $p(\xb)$ from any other distribution $\tilde{p}(\xb)$.
-
+            Enable models to distinguish the training data distribution $p(\xb)$ from any other distribution $\tilde{p}(\xb)$.
             \vspace{3mm}
-
-    We are concerned with doing this on a per-observation basis, i.e. answering the question:
-
+            Do this for any given single observation, i.e. answer the question:
             \vspace{3mm}
-
             \begin{center}
@@
             \end{center}
-
-    \vspace{3mm}
-
-    % It is related to \textbf{anomaly detection} and \textbf{adversarial example detection}.
         \end{column}
-
         \begin{column}{0.5\textwidth}
-
             \begin{figure}[\textwidth]
@@
                         \begin{axis}[
+                            name=axistop,
                             domain=0:13,samples=100,
@@
                         ]
-
                             \addplot [name path=p1, color=red!50!black] {gauss(5,1)};
@@
                             \addlegendentry{$\tilde{p}(\xb)$}
-
                         \end{axis}
-        \end{tikzpicture}
-    \end{figure}
-    \begin{figure}[\textwidth]
-        \centering
-        \begin{tikzpicture}
+                        \path (axistop.south) ++ (0,-0.4cm) coordinate (test);
                         \begin{axis}[
+                            name=axisbottom,
+                            anchor=north,
+                            shift=(test),
                             domain=0:13,samples=100,
@@
                         ]
-
                             \addplot [name path=p1, color=red!50!black] {gauss(4,0.5)};
@@
                             \addlegendentry{$\tilde{p}(\xb)$}
-
                         \end{axis}
@@
         \end{column}
-
     \end{columns}
@@
 
-\section{Latent variable models}
+% \section{Latent variable models}
 
@@
 
-\section{Identifying the issue}
+\begin{frame}
+    \frametitle{Out-of-distribution detection with hierarchical VAEs}
+    \begin{columns}
+        \begin{column}{0.4\textwidth}
+            \begin{itemize}
+                \item Generative models learn to approximate the \textbf{data distribution} $p(\xb)$.
+                \item The likelihood of the model given a sample $\xb$ is a measure of how well the model \textbf{explains the data}.
+                \item \textbf{Model likelihood} have long been thought of as useful for OOD detection \cite{bishop_novelty_1994}.
+            \end{itemize}
+        \end{column}
+        \begin{column}{0.6\textwidth}
+            \begin{overprint}
+                \onslide<2>
+                \begin{figure}
+                    \centering
+                    \includegraphics[width=\textwidth]{../graphics/paper_hierarchical/densities-FashionMNIST-test-MNIST-test-bpp-k-0_sohau.pdf}
+                \end{figure}
+            \end{overprint}
+        \end{column}
+    \end{columns}
+\end{frame}
+
+% \section{Identifying the issue}
 
@@
             Input complexity ($S$, PixelCNN++) \parencite{serra_input_2020}             & $0.929$ & - & - \\
-    Input complexity ($S$, HVAE) (Ours) \parencite{serra_input_2020}\textsuperscript{\ref{footnote:serra_hierarchical_likelihood}} & $0.833$ & $0.855$ & $0.344$ \\
+            Input complexity ($S$, HVAE) (Ours) \parencite{serra_input_2020} & $0.833$ & $0.855$ & $0.344$ \\
             \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
@@
         \begin{column}{0.5\textwidth}
-
             \begin{table}[t]
@@
                 }
-            % \caption{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
-            % \label{tab:additional-results-fashionmnist}
             \end{table}
         \end{column}
-
     \end{columns}

### 2024-02-25 presentation/parts/presentation_ladder_vae.tex
WIP

% !TEX root = ../presentation.tex
% !BIB program = biber
% !TEX program = xelatex

\section{vae background and the ladder variational autoencoder (lvae)}

\begin{frame}{The Ladder Variational Autoencoder (LVAE)}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            This is a Ladder VAE with three latent variables.
        \end{column}
        \begin{column}{0.5\textwidth}
            \begin{figure}[\textwidth]
            \def\col{blue}
            \tikz{
            % INFERENCE
            % nodes
            \node[obs] (x_inf) {$\xb$};%
            \node[det,above=.75cm of x_inf] (d1){$\db_1$}; %
            \node[det,above=.75cm of d1] (d2){$\db_2$}; %
            \node[det,above=.75cm of d2] (d3){$\db_3$}; %

            \node[latent,right=.5cm of d1] (z1_inf) {$\zb_1$}; %
            \node[latent,right=.5cm of d2] (z2_inf) {$\zb_2$}; %
            \node[latent,right=.5cm of d3] (z3_inf) {$\zb_3$}; %
            \node[above=of z3_inf, xshift=-.7cm, yshift=-1.cm] (phi) {$q_\phi(\zb|\xb)$};

            % edges
            \edge[]{x_inf}{d1}
            \edge[\col]{z2_inf}{z1_inf}
            \edge[]{d1}{d2}
            \edge[\col]{z3_inf}{z2_inf}
            \edge[]{d2}{d3}
            \edge[]{d1}{z1_inf}
            \edge[]{d2}{z2_inf}
            \edge[]{d3}{z3_inf}

            % GENERATIVE
            % nodes
            \node[latent,right=.75cm of z3_inf] (z3_gen) {$\zb_3$}; %
            \node[latent,right=.75cm of z2_inf] (z2_gen) {$\zb_2$}; %
            \node[latent,right=.75cm of z1_inf] (z1_gen) {$\zb_1$}; %
            \node[obs,below=.75cm of z1_gen] (x_gen) {$\xb$};
            \node[above=of z3_gen, yshift=-1.cm] (theta) {$p_\thetab(\xb,\zb)$};

            % edges
            \edge[\col]{z3_gen}{z2_gen}
            \edge[\col]{z2_gen}{z1_gen}
            \edge[]{z1_gen}{x_gen}

            }
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Overview}
    Understanding the Ladder VAE is not so much about understanding the specific architecture as it is about:
    \begin{itemize}
        \item[a)] understanding the reason for choosing it,
        \item[b)] which other options were available,
        \item[c)] and why they don't work as well.
    \end{itemize} 
\end{frame}

\begin{frame}{A generative model from latent variables}
    \begin{columns}[c]
        \begin{column}{.7\textwidth}
            Suppose data $\xb \sim p (\xb)$ is generated via some underlying \textit{latent} variable $\zb$. Then
            \begin{equation}
                p(\xb) = \int p(\xb,\zb) \text{d}\zb = \int p(\xb|\zb)p(\zb) \text{d}\zb.
            \end{equation}
            We would like to 
            \begin{itemize}
                \item[a)] \textit{infer} the latent variables $\zb$ given $\xb$
                \item[b)] \textit{generate} the observed variable $\xb$ given $\zb$
            \end{itemize}
        \end{column}
        \begin{column}{.3\textwidth}
            \begin{figure}[\textwidth]
                \tikz{
                    % GENERATIVE
                    % nodes$
                    \node[obs] (x_gen) {$\xb$};%
                    \node[latent,above=.75cm of x_gen](z1_gen){$\zb$}; %
                    \node[above=of z1_gen, yshift=-1.cm] (phi) {$p(\xb,\zb)$};

                    % edges
                    \edge[]{z1_gen}{x_gen};
                }
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Exact inference}
    \begin{columns}
        \begin{column}{0.7\textwidth}
            We can choose some simple model for $p(\xb,\zb)$ (denoted $p_\thetab$ and parameterized by $\thetab$).
            \vspace{3mm}

            Bayes theorem then gives us the true model posterior,
            \begin{equation}
                p_\thetab(\zb|\xb) = \frac{p_\thetab(\xb,\zb)}{p_\thetab(\xb)} = \frac{p_\thetab(\xb|\zb)p_\thetab(\zb)}{\int p_\thetab(\xb|\zb)p_\thetab(\zb) \text{d}\zb}.
            \end{equation}
            But this only works if we can integrate over the latent variable $\zb$ and the model $p_\thetab(\xb|\zb)$.
        \end{column}
        \begin{column}{.3\textwidth}
            \begin{figure}[\textwidth]
                \tikz{
                    % GENERATIVE
                    % nodes$
                    \node[obs] (x_gen) {$\xb$};%
                    \node[latent,above=.75cm of x_gen](z1_gen){$\zb$}; %
                    \node[above=of z1_gen, yshift=-1.cm] (phi) {$p(\xb,\zb)$};

                    % edges
                    \edge[]{z1_gen}{x_gen};
                }
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Exact inference becomes intractable}
    Suppose we want to model complex data where $\vert\xb\vert = D \gg 1$, such as images, audio or graphs.
    \vspace{3mm}

    We might choose to parameterize $p(\xb,\zb)$ with a neural network model with parameters $\thetab$ and try to \textit{learn} $p_\thetab(\xb) \approx p(\xb)$.

    \begin{equation}
        p_\thetab(\xb, \zb) = p_\thetab(\xb|\zb)p(\zb)
    \end{equation}

    where we could choose $p(\zb)=\mathcal{N}(\0,\1)$.
    \vspace{3mm}

    However, this comes at the cost of making integrals over the latent variables intractable and hence, \textbf{we can no longer directly compute} $p_\thetab(\xb)$.
\end{frame}

\begin{frame}{Variational inference and the ELBO}
    Introduce some \textit{variational} distribution $q_\phi(\zb|\xb)$ (parameterized by $\phi$)
    \begin{align}
        \log p(\xb) &= \log \int p_\thetab(\xb,\zb) \,\text{d}\zb \notag \\
                    &= \log \int q_\phi(\zb|\xb) \frac{p_\thetab(\xb,\zb)}{q_\phi(\zb|\xb)} \,\text{d}\zb \notag \\
                    &\geq \int q_\phi(\zb|\xb) \log \frac{p_\thetab(\xb,\zb)}{q_\phi(\zb|\xb)} \,\text{d}\zb \notag \\
                    &= \mathbb{E}_{q_\phi(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phi(\zb|\xb)} \right] \notag \\
                    &= \mathbb{E}_{q_\phi(\zb|\xb)} \left[ \log p_\thetab(\xb|\zb) + \log p_\thetab(\zb) - \log q_\phi(\zb|\xb) \right] \notag\\
                    &\equiv \underbrace{\mathcal{L}(\xb;\thetab,\phi)}_{\text{evidence lower bound}}
    \end{align}
\end{frame}

\begin{frame}{The variational autoencoder (VAE)}
    \begin{columns}
        \begin{column}{0.8\textwidth}
        Suppose we parameterize $q_\phi(\zb|\xb)$ by a model with parameters $\phi$.
        Then we can optimize both $\{\thetab,\phi\}$ jointly by maximizing the ELBO.

        \begin{equation}
            \{\thetab^{*},\phi^{*}\} = \underset{\{\thetab, \phi\}}{\arg\max} \, \mathcal{L}(\xb;\thetab,\phi)
        \end{equation}

        The result is the VAE \cite{kingma_autoencoding_2014} consisting of
        \begin{itemize}
            \item[a)] an \textit{inference model} (or \textit{encoder}) $q\phi(\zb|\xb)$ which approximates the intractable true posterior $p(\zb|\xb)$.
            \item[b)] a \textit{generative model} (or \textit{decoder}) $p_\thetab(\xb|\zb)$ which can generate new samples from the prior $p(\zb)$, or "reconstruct" with proposals from $q_\phi(\zb|\xb)$.
        \end{itemize}

        \end{column}

        \begin{column}{0.2\textwidth}
            \begin{figure}[.5\textwidth]
            \tikz{
                % inference
                % nodes
                \node[obs] (x_inf) {$\xb$};%
                \node[latent,above=.75cm of x_inf](z1_inf){$\zb$}; %
                \node[above=of z1_inf, yshift=-1.cm] (phi) {$q_\phi(\zb|\xb)$};

                % edges
                \edge[]{x_inf}{z1_inf};

                % generative
                % nodes$
                \node[obs,right=0.75cm of x_inf] (x_gen) {$\xb$};%
                \node[latent,above=.75cm of x_gen](z1_gen){$\zb$}; %
                \node[above=of z1_gen, yshift=-1.cm] (theta) {$p_\thetab(\xb,\zb)$};

                % edges
                \edge[]{z1_gen}{x_gen};
            }
            \end{figure}

        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Limitations of VAEs}
    \begin{columns}
        \begin{column}{0.8\textwidth}
            The VAE uses a mean field approximation for most common variational posteriors.
            \begin{equation}
                q(\zb) = \prod_{i} q(z_i)
            \end{equation}
            \begin{itemize}
                \item Assumes independence between latent variables.
                \item Model cannot learn covariance between latents.
                \item This limits expressivity as we cannot expect to always match the true posterior well.
            \end{itemize}
            {\small \textit{E.g. a vehicle's color is often dependent on its type (fire truck, police car, bus, taxi etc.) but these are coded independently in $\zb$}.}
        \end{column}
        \begin{column}{0.2\textwidth}
            \begin{figure}[.5\textwidth]
            \tikz{
                % inference
                % nodes
                \node[obs] (x_inf) {$\xb$};%
                \node[latent,above=.75cm of x_inf](z1_inf){$\zb$}; %
                \node[above=of z1_inf, yshift=-1.cm] (phi) {$q_\phi(\zb|\xb)$};

                % edges
                \edge[]{x_inf}{z1_inf};

                % generative
                % nodes$
                \node[obs,right=0.75cm of x_inf] (x_gen) {$\xb$};%
                \node[latent,above=.75cm of x_gen](z1_gen){$\zb$}; %
                \node[above=of z1_gen, yshift=-1.cm] (theta) {$p_\thetab(\xb,\zb)$};

                % edges
                \edge[]{z1_gen}{x_gen};
            }
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Hierarchical VAE}
    \begin{columns}
        \begin{column}{0.8\textwidth}
            To avoid these limitations, we can introduce a hierarchy of additional latent variables $\zb=\zb_1,\dots,\zb_L$. For $L=3$,
            \begin{equation}
                p_\thetab(\xb|\zb) = p_\thetab(\xb|\zb_1) p_\thetab(\zb_1|\zb_2) p(\zb_3).
            \end{equation}
            We can straightforwardly generalize the inference model,
            \begin{equation}
                q_\phi(\zb|\xb) = q_\phi(\zb_1|\xb)q_\phi(\zb_2|\zb_1)q_\phi(\zb_3|\zb_2).
            \end{equation}
            This is called \textit{bottom-up} inference.
        \end{column}
        \begin{column}{0.2\textwidth}
            \begin{figure}[.5\textwidth]
                \tikz{
                    % INFERENCE
                    % nodes
                    \node[obs] (x_inf) {$\xb$};%
                    \node[latent,above=.75cm of x_inf](z1_inf){$\zb_1$}; %
                    \node[latent,above=.75cm of z1_inf](z2_inf){$\zb_2$}; %
                    \node[latent,above=.75cm of z2_inf](z3_inf){$\zb_3$}; %
                    \node[above=of z3_inf, yshift=-1.cm] (phi) {$q_\phi(\zb|\xb)$};

                    % edges
                    \edge[]{x_inf}{z1_inf};
                    \edge[]{z1_inf}{z2_inf};
                    \edge[]{z2_inf}{z3_inf};

                    % GENERATIVE
                    % nodes
                    \node[obs,right=0.75cm of x_inf] (x_gen) {$\xb$};%
                    \node[latent,above=.75cm of x_gen](z1_gen){$\zb_1$}; %
                    \node[latent,above=.75cm of z1_gen](z2_gen){$\zb_2$}; %
                    \node[latent,above=.75cm of z2_gen](z3_gen){$\zb_3$}; %
                    \node[above=of z3_gen, yshift=-1.cm] (phi) {$p_\thetab(\xb,\zb)$};

                    % edges
                    \edge[]{z3_gen}{z2_gen};
                    \edge[]{z2_gen}{z1_gen};
                    \edge[]{z1_gen}{x_gen};
                }
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Challenges of hierarchical VAEs}
    Consider a simple model $p_{\text{simple}}(\xb)$ without any latent variables. We can rewrite the likelihood as,
    \begin{align*}
        \mathbb{E}_{p(\xb)} \left[ \log p_{\text{simple}}(\xb) \right] &= \mathbb{E}_{p(\xb)} \left[ \log \left( p(\xb) \frac{p_{\text{simple}}(\xb)}{p(\xb)} \right) \right] \\
                                                                    %  &= -\mathcal{H}(p) - D_\text{KL} \left( p(\xb) || p_\text{simple}(\xb) \right)
                                                                    &= -\underbrace{\mathcal{H}(p)}_{\text{data entropy}} - \underbrace{D_\text{KL} \left( p(\xb) || p_\text{simple}(\xb) \right)}_{\text{divergence from data distribution}}
    \end{align*}
    where $\mathcal{H}(p)=\mathbb{E}_{p(\xb)}[\log p(\xb)]$.
\end{frame}

\begin{frame}{Challenges of hierarchical VAEs}
    \begin{columns}
        \begin{column}{1\textwidth}
            Let's do the same for a VAE $p_\thetab(\xb)$ with ELBO given by
            \begin{equation}
                \log p_\thetab(\xb) \geq \mathbb{E}_{q_\phi(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p_\thetab(\zb)}{q_\phi(\zb|\xb)} \right].
            \end{equation}
            The expectation over the data becomes,
            \begin{align}
                \mathbb{E}_{p(\xb)} \left[ \log p_\thetab(\xb) \right] \geq& \mathbb{E}_{p(\xb)} \left[ \log \left( p(\xb) \frac{p_\thetab(\xb)}{p(\xb)} \right) \right] \notag \\
                                                                    =& -\mathcal{H}(p) - D_\text{KL} \left( p(\xb) || p_\thetab(\xb) \right) \notag \\
                                                                        & \quad - \underbrace{\mathbb{E}_{p(\xb)} \left[ D_\text{KL}\left( q_\phi(\zb|\xb) || p_\thetab(\zb|\xb) \right) \right]}_{\text{divergence from true posterior}}
            \end{align}
            Compared to the simple model, we incur an \textbf{additional cost} for the latent variable given by the divergence from the true model posterior.
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Challenges of hierarchical VAEs}
    For a hierarchical VAE with two latent variables
    \begin{align}
        \mathbb{E}_{p(\xb)} \left[ \log p_\thetab(\xb) \right] \geq& \mathbb{E}_{p(\xb)} \left[ \log \left( p(\xb) \frac{p_\thetab(\xb)}{p(\xb)} \right) \right] \notag \\
                                                            =& -\mathcal{H}(p) - D_\text{KL} \left( p(\xb) || p_\thetab(\xb) \right) \notag \\
                                                                & \quad - \mathbb{E}_{p(\xb)} \mathbb{E}_{q_\phi(\zb_1|\xb)} \left[ D_\text{KL}\left( q_\phi(\zb_2|\zb_1) || p_\thetab(\zb_2|\zb_1) \right) \right] \notag \\
                                                                & \quad - \mathbb{E}_{p(\xb)} \left[ D_\text{KL}\left( q_\phi(\zb_1|\xb) || p_\thetab(\zb_1|\xb) \right) \right]
    \end{align}

    The cost is \textbf{incurred for each additional latent} we add to the hierarchy.
    \vspace{3mm}

    Hence, each latent will only be used by the model if we get an \textbf{opposite equal or larger improvement} in the ELBO.
\end{frame}

\begin{frame}{{\tiny (Failing to)} Dodge the challenges}
    \begin{columns}
        \begin{column}{0.77\textwidth}
            There's a few common ways to try and dodge these issues:
            \begin{itemize}
                \item Skip connections \cite{dieng_avoiding_2019, maaloe_biva_2019}
                \item Free bits in the KL-terms \cite{kingma_improved_2016}
                \item Deterministic warmup \cite{sonderby_ladder_2016}
                \item Batch normalization/weight normalization \cite{ioffe_batch_2015, salimans_weight_2016}
            \end{itemize}

            \vspace{5mm}
            None of them work for more than around 5 latent variables.
        \end{column}
        \begin{column}{0.3\textwidth}
            \begin{figure}[.5\textwidth]
                \tikz{
                    % INFERENCE
                    % nodes
                    \node[obs] (x_inf) {$\xb$};%
                    \node[latent,above=.75cm of x_inf](z1_inf){$\zb_1$}; %
                    \node[latent,above=.75cm of z1_inf](z2_inf){$\zb_2$}; %
                    \node[latent,above=.75cm of z2_inf](z3_inf){$\zb_3$}; %
                    \node[above=of z3_inf, yshift=-1.cm] (phi) {$q_\phi(\zb|\xb)$};

                    % edges
                    \edge[]{x_inf}{z1_inf};
                    \edge[]{z1_inf}{z2_inf};
                    \edge[]{z2_inf}{z3_inf};
                    \edge[dashed, bend left]{x_inf}{z2_inf};
                    \edge[dashed, bend left]{x_inf}{z3_inf};

                    % GENERATIVE
                    % nodes
                    \node[obs,right=0.75cm of x_inf] (x_gen) {$\xb$};%
                    \node[latent,above=.75cm of x_gen](z1_gen){$\zb_1$}; %
                    \node[latent,above=.75cm of z1_gen](z2_gen){$\zb_2$}; %
                    \node[latent,above=.75cm of z2_gen](z3_gen){$\zb_3$}; %
                    \node[above=of z3_gen, yshift=-1.cm] (phi) {$p_\thetab(\xb,\zb)$};

                    % edges
                    \edge[]{z3_gen}{z2_gen};
                    \edge[]{z2_gen}{z1_gen};
                    \edge[]{z1_gen}{x_gen};
                    \edge[dashed, bend left]{z2_gen}{x_gen};
                    \edge[dashed, bend left]{z3_gen}{x_gen};
                }
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{The Ladder VAE}
    \begin{columns}
        \begin{column}{0.6\textwidth}
            The LVAE introduces a \textit{top-down} inference path,
            \begin{equation}
                q_\phi(\zb|\xb) = q_\phi(\zb_L|\xb) \prod_{i=1}^{L-1} q_\phi(\zb_{i}|\zb_{i+1}).
            \end{equation}
            It is aided by a \textbf{deterministic bottom-up path} and \textbf{{\color{blue} parameter sharing}} between the inference and generative models.
        \end{column}
        \begin{column}{0.4\textwidth}
            \begin{figure}[\textwidth]
            \def\col{blue}
            \tikz{
                % INFERENCE
                % nodes
                \node[obs] (x_inf) {$\xb$};%
                \node[det,above=.75cm of x_inf] (d1){$\db_1$}; %
                \node[det,above=.75cm of d1] (d2){$\db_2$}; %
                \node[det,above=.75cm of d2] (d3){$\db_3$}; %

                \node[latent,right=.5cm of d1] (z1_inf) {$\zb_1$}; %
                \node[latent,right=.5cm of d2] (z2_inf) {$\zb_2$}; %
                \node[latent,right=.5cm of d3] (z3_inf) {$\zb_3$}; %
                \node[above=of z3_inf, xshift=-.7cm, yshift=-1.cm] (phi) {$q_\phi(\zb|\xb)$};

                % edges
                \edge[]{x_inf}{d1}
                \edge[\col]{z2_inf}{z1_inf}
                \edge[]{d1}{d2}
                \edge[\col]{z3_inf}{z2_inf}
                \edge[]{d2}{d3}
                \edge[]{d1}{z1_inf}
                \edge[]{d2}{z2_inf}
                \edge[]{d3}{z3_inf}

                % GENERATIVE
                % nodes
                \node[latent,right=.75cm of z3_inf] (z3_gen) {$\zb_3$}; %
                \node[latent,right=.75cm of z2_inf] (z2_gen) {$\zb_2$}; %
                \node[latent,right=.75cm of z1_inf] (z1_gen) {$\zb_1$}; %
                \node[obs,below=.75cm of z1_gen] (x_gen) {$\xb$};
                \node[above=of z3_gen, yshift=-1.cm] (theta) {$p_\thetab(\xb,\zb)$};

                % edges
                \edge[\col]{z3_gen}{z2_gen}
                \edge[\col]{z2_gen}{z1_gen}
                \edge[]{z1_gen}{x_gen}
            }
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Benchmarks}
    \begin{table}[]
        \centering
        \begin{tabular}{lr}
        \toprule
        Model & $\geq \log p(\xb)$ \\
        \midrule
        VAE + NF ($L=1$) & -85.10 \\
        IWAE ($L=2$, $K=1$) & -85.33 \\
        IWAE ($L=2$, $K=50$) & -82.90 \\
        VAE + VGP ($L=2$) & -81.90 \\
        \midrule
        LVAE ($L=5$) & -82.12 \\
        LVAE + finetuning ($L=5$) & -81.84 \\
        LVAE + finetuning ($L=5$, $K=10$) & -81.74 \\
        \bottomrule
        \end{tabular}
        \caption{Results on dynamically binarized MNIST.}
    \end{table}
\end{frame}

\begin{frame}{Latent space activation}
    \begin{figure}[\textwidth]
        \centering
        \includegraphics[scale=0.85]{figures/ladder_vae_latent_activations.pdf}
        \caption{Element-wise KL-divergences in each latent variable $q(\zb_i|\zb_{i-1})$ with $\zb_0\equiv \xb$.}
    \end{figure}
\end{frame}

\begin{frame}{Latent space representation}
    \begin{figure}[\textwidth]
        \centering
        \includegraphics[scale=1]{figures/ladder_vae_pca.pdf}
        \caption{PCA of latent samples from $q(\zb_i|\zb_{i-1})$ with $\zb_0\equiv \xb$.}
    \end{figure}
\end{frame}

\begin{frame}{Recent work}
    \begin{figure}[\textwidth]
        \centering
        \includegraphics[scale=0.7]{figures/child_faces.pdf}
        \caption{Samples from the generative model of \cite{child_very_2021} with more than 70 latents.}
    \end{figure}
\end{frame}

### 2024-02-27 presentation/parts/introduction.tex
WIP

@@
 
-\begin{frame}
+% \begin{frame}
+%     \frametitle{Medical dialogue}
+%     \begin{figure}
+%         \centering
+%         \includegraphics[width=0.7\paperwidth]{figures/speech_bubble.pdf}
+%     \end{figure}
+
+%     \note[item]{Healthcare sector is a complex system}
+% \end{frame}
+% \begin{frame}
+%     \frametitle{Medical dialogue}
+%     \begin{figure}
+%         \centering
+%         \includegraphics[width=0.95\textheight]{figures/corti_sketch_conversations.png}
+%     \end{figure}
+% \end{frame}
+
+\begin{frame}[t]
     \frametitle{Medical dialogue}
+    \begin{columns}[t]
+        \begin{column}{0.4\textwidth}
+            \begin{itemize}
+                \item <2> General practitioner
+                \item <2> Nurse
+                \item <2> Midwife
+                \item <2> Emergency medical dispatcher
+                \item <2> Paramedic
+                \item <2> Emergency room
+                \item <2> Health insurance
+            \end{itemize}
+        \end{column}
+        \begin{column}{0.6\textwidth}
             \begin{figure}
                 \centering
-        \includegraphics[width=0.7\paperwidth]{figures/speech_bubble.pdf}
+                \includegraphics[width=0.9\textwidth]{figures/corti_sketch_conversations.png}
             \end{figure}
-
-    \note[item]{Healthcare sector is a complex system}
+        \end{column}
+    \end{columns}
 \end{frame}
@@
     \frametitle{Errors in medical dialogue}
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
             \begin{itemize}
-        \item<1,2>Communication is everywhere in healthcare. 
-        \item <1,2>It is complex, involving multiple participants, different contexts, and different purposes.
+                \item <1-> Communication is everywhere in healthcare. 
+                \item <1-> It is complex, involving multiple participants, different contexts, and different purposes.
                 \vspace{1em}
-        \item <2> \highlight{Adverse events}: Failure of communication is a leading cause of medical error contributing to two out of three adverse events \cite{starmer_changes_2014}.
-        \item <2> \highlight{Preventability}: A considerable fraction of all hospital admissions had preventable adverse outcomes (9\% to 16.6\% in AU, NZ, UK, DK) \cite{carver_medical_2024}.
+                \item <2-> \highlight{Adverse events}: Failure of communication is a leading cause of medical error contributing to two out of three adverse events \cite{starmer_changes_2014}.
+                \item <2-> \highlight{Preventability}: A considerable fraction of all hospital admissions have preventable adverse outcomes\footnote<2>{9\% to 16.6\% in AU, NZ, UK, DK.} \cite{carver_medical_2024}.
             \end{itemize}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+                \centering
+                \includegraphics[width=0.95\textwidth]{figures/corti_sketch_emergency.png}
+            \end{figure}
+        \end{column}
+    \end{columns}
 \end{frame}
-
-% \begin{frame}
-%     \frametitle{Types of dialogue}
-%     \begin{itemize}
-%         \item Domain: Specialized / General
-%         \item Context: Controlled / Chaotic
-%         \item Person: Nurse, doctor, midwife, caregiver, psychiatrist, insurance professional
-%         \item Purpose: Triaging, diagnosis, treatment, follow-up, documentation, coding, billing
-%     \end{itemize}
-% \end{frame}
 
@@
     \frametitle{Documenting medical encounters}
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
             \begin{itemize}
@@
                 \vspace{1em}
-        \item <2> \highlight{Time-consuming}: Physicians spend 34-37\% of their time on documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
-        \item <2> \highlight{Varying quality}: Discharge summaries rarely meet all timeline, transmission, and content criteria.
-\cite{horwitz_comprehensive_2013}\footnote{Outpatient visits, Yale-New Haven Hospital.}
+                \item <2> \highlight{Time-consuming}: Physicians spend 34-37\% of their time on documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote<2>{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
+                \item <2> \highlight{Varying quality}: Discharge summaries almost never meet \emph{all} timeline, transmission, and content criteria.
+\cite{horwitz_comprehensive_2013}\footnote<2>{Outpatient visits, Yale-New Haven Hospital.}
             \end{itemize}
-
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+                \centering
+                \includegraphics[width=0.95\textwidth]{figures/corti_sketch_responders.png}
+            \end{figure}
+        \end{column}
+    \end{columns}
     \note[item]{Documentation is essential for a number of purposes but is very time-consuming and imperfect}
@@
     \frametitle{How might machine learning help?}
+    \begin{columns}
+        \begin{column}{0.35\textwidth}
             \begin{itemize}
@@
             \end{itemize}
+        \end{column}
+        \begin{column}{0.65\textwidth}
+            \centering
+            \includegraphics[width=0.95\textwidth]{figures/corti_sketch_office.png}
+        \end{column}
+    \end{columns}
 \end{frame}

### 2024-02-27 presentation/parts/presentation_hierarchical.tex
WIP

@@
 \frame{
-    \frametitle{Problem and Contributions}
-    \begin{itemize}
-        \item Deep generative models often fail at OOD detection task when using their likelihood estimate as the score function \cite{nalisnick_deep_2019} by, perhaps surprisingly, assigning \textbf{higher likelihoods} to the OOD data.
-        \item Contributions:
-        \begin{itemize}
-            \item We provide evidence that out-of-distribution detection fails due to learned low-level features that generalize across datasets.
-            \item We present a fast and fully unsupervised method for OOD detection competitive with the state-of-the-art
-        \end{itemize}  
-    \end{itemize}
-}
-
-\frame{
     \frametitle{In distribution?}
@@
     \end{figure}
+}
+
+\frame{
+    \frametitle{Problem and Contributions}
+    \begin{itemize}
+        \item Deep generative models often fail at OOD detection task when using their likelihood estimate as the score function \cite{nalisnick_deep_2019} by, perhaps surprisingly, assigning \highlight{higher likelihoods to OOD data}.
+        \item Contributions:
+        \begin{itemize}
+            \item We provide evidence that out-of-distribution detection fails due to learned low-level features that generalize across datasets.
+            \item We present a new score for OOD detection with hierarchical VAEs that alleviates this issue.
+        \end{itemize}  
+    \end{itemize}
 }
@@
     \frametitle{Out-of-distribution detection with hierarchical VAEs}
-    \begin{columns}
+    \begin{columns}[t]
         \begin{column}{0.4\textwidth}
@@
                 \item The likelihood of the model given a sample $\xb$ is a measure of how well the model \textbf{explains the data}.
-                \item \textbf{Model likelihood} have long been thought of as useful for OOD detection \cite{bishop_novelty_1994}.
+                \item \textbf{Model likelihood} has long been thought of as useful for OOD detection \cite{bishop_novelty_1994}.
             \end{itemize}
@@
 
+\begin{frame}
+    \frametitle{Selecting the value of $k$}
+    \begin{itemize}
+        \item Use validation OOD dataset(s).
+        \item Compute $LLR^{>k}$ for different values of $k$ and select the one that maximizes the AUROC.
+        \item Compute feature correlations for different values of $k$ and select $k$ at the drop.
+    \end{itemize}
+    \begin{figure}
+        \centering
+        \includegraphics[width=0.9\textwidth]{../graphics/paper_hierarchical/feature-correlation-heatmap2.pdf}
+    \end{figure}
+\end{frame}
+
 \frame{
@@
     \begin{columns}
-        \begin{column}{0.5\textwidth}
+        \begin{column}{0.45\textwidth}
+            \hfill
             \begin{table}[t]
@@
             \end{table}
+            \hfill
         \end{column}
-        \begin{column}{0.5\textwidth}
+        \begin{column}{0.55\textwidth}
+            \hfill
             \begin{table}[t]
@@
                     SmallNORB28x28Inverted   &  $LLR^{>2}$            &  0.941  &  0.946  &  0.069 \\
-                    FashionMNIST             &  $LLR^{>1}$            &  0.488  &  0.496  &  0.811 \\
+   {\color{black!60}FashionMNIST}             &  $LLR^{>1}$            &  0.488  &  0.496  &  0.811 \\
                     \midrule
@@
                     SmallNORB28x28Inverted         &  $LLR^{>1}$  &  $0.985$  &  $0.987$  &  $0.000$ \\
-                    MNIST                          &  $LLR^{>2}$  &  $0.515$  &  $0.507$  &  $0.792$ \\
+   {\color{black!60}MNIST}                          &  $LLR^{>2}$  &  $0.515$  &  $0.507$  &  $0.792$ \\
                     \bottomrule
@@
             \end{table}
+            \hfill
         \end{column}
@@
 }
+
+\begin{frame}
+    \frametitle{Conclusions}
+    \begin{itemize}
+        \item Key observations: 
+        \begin{itemize}
+            \item The likelihood of a generative model is not a good score for OOD detection \cite{nalisnick_deep_2019}.
+            \item Strong correlations between some latent variables for different datasets.
+            \item Reconstructions of OOD data are good when using full approximate posterior.
+        \end{itemize}
+        \item[\highlight{$\rightarrow$}] Proposed a new score, $LLR^{>k}$, that uses the conditional prior for the top-most latent variables in the hierarchy.
+    \end{itemize}
+\end{frame}
 

### 2024-02-28 presentation/parts/introduction.tex
WIP

@@
 \section{Introduction}
-
-\begin{frame}
-    % Background on Corti and the problem of medical communication, diagnosis, and documentation.
-    \frametitle{Background}
-    \begin{columns}
-        \begin{column}{0.6\textwidth}
-            FILL ME OUT. \\
-            Background on Corti, motivation for project.
-        \end{column}
-        \begin{column}{0.4\textwidth}
-        \end{column}
-    \end{columns}
-\end{frame}
 
@@
         \begin{column}{0.6\textwidth}
-            \begin{figure}
+            \vspace{-2em}
+            \begin{figure}[t]
                 \centering
@@
                 \vspace{1em}
-                \item <2-> \highlight{Reduce} number of impact of medical errors and adverse events.
+                \item <2-> \highlight{Reduce} the impact of medical errors and adverse events.
                 \item <2-> \highlight{Free up} time spent on documentation for patient care.
@@
         \item <5-> \highlight{Fairness} in treatment of different groups of people.
-        \item <6-> \highlight{Robustness} to noise, outliers, adversarial attacks, and distribution shift.
+        \item <6-> \highlight{Robustness} to noise, outliers, distribution shift, and adversarial attacks.
     \end{itemize}

### 2024-02-29 presentation/parts/introduction.tex
WIP

@@
 \end{frame}
+
+\tikzstyle{mybox} = [fill, rounded corners, minimum width=140, minimum height=20]
+\tikzstyle{myrect} = [fill]
+
+\begin{frame}
+    \frametitle{Building a decision-support system}
+    \begin{columns}
+        \begin{column}{0.6\textwidth}
+            Modular approach:
+            \begin{itemize}
+                \item \highlight{Input}: Speech or text (potentially video, tabular data, images, etc.).
+                \item \highlight{Parsing the input}: Representation learning and speech recognition.
+                \item \highlight{Output}: Predictions (suggesting, summarizing, classifying, etc.).
+            \end{itemize}
+        \end{column}
+        \begin{column}{0.4\textwidth}
+            \resizebox{!}{0.8\textheight}{%
+            \begin{tikzpicture}[x=1cm,y=1cm]
+                \node<6> [mybox, fill=presentation_orange] at (0,5) (predictions) {};
+                \foreach \i in {-11,...,11} {
+                    \fill<6>[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
+                }
+                \node<6> [] at (0,5) {\small\textsc{predictions}};
+
+                \node<5> [mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};
+
+                \node<4> [mybox, fill=presentation_blue] at (0,3) (text) {};
+                \foreach \i in {-11,...,11} {
+                    \fill<4>[myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
+                }
+                \node<4> [] at (0,3) {\small\textsc{text}};
+
+                \node<3> [mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};
+
+                \node<2> [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};
+
+                \node<1> [mybox, fill=presentation_green] at (0,0) (speech) {};
+                \foreach \i in {-11,...,11} {
+                    \fill<1>[myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
+                }
+                \node<1> [] at (0,0) {\small\textsc{speech}};
+
+                \node<7> [draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
+                \node<7> [above] at (uncertainty.north) {\small\textsc{uncertainty}};
+            \end{tikzpicture}
+        }
+        \end{column}
+    \end{columns}
+\end{frame}

### 2024-02-29 presentation/parts/introduction.tex
WIP

--- 2024-02-28 presentation/parts/introduction.tex
@@
 \end{frame}
+
+\tikzstyle{mybox} = [fill, rounded corners, minimum width=140, minimum height=20]
+\tikzstyle{myrect} = [fill]
+
+\begin{frame}
+    \frametitle{Building a decision-support system}
+    \begin{columns}
+        \begin{column}{0.6\textwidth}
+            Modular approach:
+            \begin{itemize}
+                \item <2-> \highlight{Source data}: Speech or text (potentially images, video, electronic health records, etc.).
+                \item <3-> \highlight{Foundation modelling}: Converting the input into representations useful for downstream tasks.
+                \item <4-> \highlight{Downstream tasks}: Suggesting, summarizing, classifying, transcribing, translating, etc.
+                \item <5-> \highlight{Uncertainty}: Estimating the reliability of data, representations, predictions.
+            \end{itemize}
+        \end{column}
+        \begin{column}{0.4\textwidth}
+            \resizebox{0.8\textwidth}{!}{%
+            \begin{tikzpicture}[x=1cm,y=1cm]
+                \node [mybox, fill=white] at (0,5) (predictions_white) {};
+                \node<4-> [mybox, fill=presentation_orange] at (0,5) (predictions) {};
+                \foreach \i in {-11,...,11} {
+                    \fill<4->[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
+                }
+                \node<4-> [] at (0,5) {\small\textsc{predictions}};
+
+                \node<4-> [mybox, fill=black!10] at (0,4) (downstream) {\small\textsc{downstream model}};
+
+                \node [mybox, fill=white] at (0,3) (text_white) {};
+                \node<2-> [mybox, fill=presentation_blue] at (0,3) (text) {};
+                \foreach \i in {-11,...,11} {
+                    \fill<2->[myrect, fill=presentation_blue_shaded] (0.2*\i,2.8) rectangle ++(0.1,0.4);
+                }
+                \node<2-> [] at (0,3) {\small\textsc{text}};
+
+                \node [mybox, fill=white] at (0,2) (recognition_white) {};
+                \node<3-> [mybox, fill=black!10] at (0,2) (recognition) {\small\textsc{speech recognition}};
+
+                \node [mybox, fill=white] at (0,1) (representation_white) {};
+                \node<3-> [mybox, fill=black!10] at (0,1) (representation) {\small\textsc{representation learning}};
+
+                \node [mybox, fill=white] at (0,0) (speech_white) {};
+                \node<2-> [mybox, fill=presentation_green] at (0,0) (speech) {};
+                \foreach \i in {-11,...,11} {
+                    \fill<2->[myrect, fill=presentation_green_shaded] (0.2*\i,-0.2) rectangle ++(0.1,0.4);
+                }
+                \node<2-> [] at (0,0) {\small\textsc{speech}};
+
+                \node [draw, rectangle, rounded corners, line width=0.15cm, white, fit=(speech_white)(representation_white)(recognition_white)(text_white)(predictions_white), inner sep=10pt] (uncertainty_white) {};
+                \node [above, white] at (uncertainty_white.north) {\color{white}\small\textsc{uncertainty}};
+
+                \node<5-> [draw, rectangle, rounded corners, line width=0.15cm, presentation_red, fit=(speech)(representation)(recognition)(text)(predictions), inner sep=10pt] (uncertainty) {};
+                \node<5-> [above] at (uncertainty.north) {\small\textsc{uncertainty}};
+            \end{tikzpicture}
+        }
+        \end{column}
+    \end{columns}
+\end{frame}

### 2024-03-02 presentation/parts/introduction.tex
WIP

@@
 \begin{frame}
-    \frametitle{Healthcare}
+    \frametitle{What is healthcare?}
     \begin{quotation}
@@
     \end{quotation}
+
+    \note[item]{Machine learning for healthcare can touch on most of these aspects. \cite{abati_conditional_2020}}
 \end{frame}
@@
 % \end{frame}
-% \begin{frame}
-%     \frametitle{Medical dialogue}
-%     \begin{figure}
-%         \centering
-%         \includegraphics[width=0.95\textheight]{figures/corti_sketch_conversations.png}
-%     \end{figure}
-% \end{frame}
 
 \begin{frame}[t]
-    \frametitle{Medical dialogue}
+    \frametitle{Medical dialogue - a challenging domain}
     \begin{columns}[t]
@@
             \begin{itemize}
-                \item <2> General practitioner
-                \item <2> Nurse
-                \item <2> Midwife
-                \item <2> Emergency medical dispatcher
-                \item <2> Paramedic
-                \item <2> Emergency room
-                \item <2> Health insurance
+                \item <1> General practitioner
+                \item <1> Nurse
+                \item <1> Midwife
+                \item <1> Emergency medical dispatcher
+                \item <1> Paramedic
+                \item <1> Emergency room
+                \item <1> Health insurance
             \end{itemize}
@@
     \end{columns}
+
+    \note[item]{We focus on medical communication.}
+    \note[item]{Involves many different parties.}
+    \note[item]{Different contexts and purposes.}
+    \note[item]{Busy emergency room, calls to emergency medical dispatchers, visits at general practitioners, etc.}
 \end{frame}
@@
     \end{columns}
+
+    \note[item]{So yes communication can be complex and noisy, but what are the consequences?}
+    \note[item]{Adverse events are a leading cause of medical error.}
+    \note[item]{But many of these are preventable.}
 \end{frame}
@@
             \begin{itemize}
-                \item <1,2> Documentation is a central part of healthcare.
-                \item <1,2> E.g. patient records, insurance claims, billing, research, training, legal purposes.
+                \item <1-> Documentation is a central part of healthcare.
+                \item <1-> E.g. patient records, insurance claims, billing, research, training, legal purposes.
                 \vspace{1em}
-                \item <2> \highlight{Time-consuming}: Physicians spend 34-37\% of their time on documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote<2>{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
-                \item <2> \highlight{Varying quality}: Discharge summaries almost never meet \emph{all} timeline, transmission, and content criteria.
-\cite{horwitz_comprehensive_2013}\footnote<2>{Outpatient visits, Yale-New Haven Hospital.}
+                \item <2-> \highlight{Time-consuming}: Physicians spend 34-37\% of their time on documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote<2->{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
+                \item <2-> \highlight{Varying quality}: Discharge summaries almost never meet \emph{all} timeline, transmission, and content criteria.
+\cite{horwitz_comprehensive_2013}\footnote<2->{Outpatient visits, Yale-New Haven Hospital.}
             \end{itemize}
@@
     \end{columns}
+
+    % \note[item] Another aspect of medical communication is documentation.
+    % \note[item] This is a central part of healthcare.
     \note[item]{Documentation is essential for a number of purposes but is very time-consuming and imperfect}
+
 \end{frame}

### 2024-03-02 presentation/parts/titlepage.tex
WIP

% !TEX root = presentation.tex
% !BIB program = biber
% !TEX program = xelatex

\frame{
    \maketitle

    \note{
        \begin{itemize}
            \item Welcome to my PhD defense.
            \item Thank you to the moderator and the assessment committee for taking part today.
            \item I will present my work on uncertainty estimation in AI systems for medical domains.
            \item I will start with an overview of the thesis followed by a brief introduction.
            \item Then I will present a selection of the research chapters.
            \item Finally, I will discuss the broader implications of the work.
        \end{itemize}
    }
}

### 2024-03-02 presentation/parts/presentation_introduction.tex
WIP

--- 2024-03-02 presentation/parts/introduction.tex
@@
         \noindent\centering
-        Healthcare is the improvement of health via the \highlight{prevention}, \highlight{diagnosis}, \highlight{treatment}, \highlight{amelioration} or \highlight{cure} of {\color{theme-green}disease}, {\color{theme-green}illness}, {\color{theme-green}injury}, and {\color{theme-green}other physical and mental impairments} in people.
+        \textbf{Healthcare} is the improvement of health via the \highlight{prevention}, \highlight{diagnosis}, \highlight{treatment}, \highlight{amelioration} or \highlight{cure} of {\color{theme-green}disease}, {\color{theme-green}illness}, {\color{theme-green}injury}, and {\color{theme-green}other physical and mental impairments} in people.
     \end{quotation}
+    \begin{flushright}
+        \footnotesize
+        --- Wikipedia
+    \end{flushright}
 
-    \note[item]{Machine learning for healthcare can touch on most of these aspects. \cite{abati_conditional_2020}}
+    \note[item]{Machine learning for healthcare touches on most of these aspects. Fx:}
+    \note[item]{
+        \begin{itemize}
+            \item \textbf{Prevention} by predicting risk of disease allowing for early intervention.
+            \item \textbf{Diagnosis} by detecting stroke in emergency calls.
+            \item \textbf{Treatment} by suggesting medication.
+            \item Even \textbf{Cure} by helping to find new cures or treatments for diseases.
+        \end{itemize}
+    }
 \end{frame}
 
-% \begin{frame}
-%     \frametitle{Medical dialogue}
-%     \begin{figure}
-%         \centering
-%         \includegraphics[width=0.7\paperwidth]{figures/speech_bubble.pdf}
-%     \end{figure}
-
-%     \note[item]{Healthcare sector is a complex system}
-% \end{frame}
-
 \begin{frame}[t]
-    \frametitle{Medical dialogue - a challenging domain}
+    \frametitle{Medical dialogue}
     \begin{columns}[t]
         \begin{column}{0.4\textwidth}
+            Every interaction within a healthcare system is a form of \highlight{medical dialogue}:
             \begin{itemize}
@@
                 \vspace{1em}
-                \item <2-> \highlight{Adverse events}: Failure of communication is a leading cause of medical error contributing to two out of three adverse events \cite{starmer_changes_2014}.
-                \item <2-> \highlight{Preventability}: A considerable fraction of all hospital admissions have preventable adverse outcomes\footnote<2>{9\% to 16.6\% in AU, NZ, UK, DK.} \cite{carver_medical_2024}.
+                \item <2-> \highlight{Adverse events}: Failure of communication contributes to two out of three adverse events \cite{starmer_changes_2014}.
+                \item <2-> \highlight{Preventability}: Many adverse outcomes are preventable \cite{carver_medical_2024}.
+                % \footnote<2>{9\% to 16.6\% in AU, NZ, UK, DK.} \cite{carver_medical_2024}.
             \end{itemize}
@@
     \note[item]{So yes communication can be complex and noisy, but what are the consequences?}
-    \note[item]{Adverse events are a leading cause of medical error.}
-    \note[item]{But many of these are preventable.}
+    \note[item]{Failure of communication is a leading cause of adverse events.}
+    \note[item]{Adverse events are episodes of medical error that result in harm to the patient.}
+    \note[item]{Luckily, many of these are preventable, although exact numbers vary.}
+    \note[item]{Better communication could help reduce these numbers.}
 \end{frame}
@@
 
-    % \note[item] Another aspect of medical communication is documentation.
-    % \note[item] This is a central part of healthcare.
-    \note[item]{Documentation is essential for a number of purposes but is very time-consuming and imperfect}
-
+    \note[item]{Another central aspect of medical communication is documentation.}
+    \note[item]{Essential for a number of purposes}
+    \note[item]{But, it is time-consuming and of varying quality.}
+    \note[item]{(Ambulatory $\equiv$ outpatient care, Tertiary $\equiv$ specialized care)}
 \end{frame}
@@
     \end{columns}
+
+    \note[item]{Machine learning can help in many ways.}
+    \note[item]{The main goal of augmenting communication is:
+        \begin{itemize}
+            \item Reduce the impact of medical errors and adverse events.
+            \item Free up time spent on documentation for patient care.
+        \end{itemize}
+    }
 \end{frame}

### 2024-03-03 presentation/parts/overview_presentation.tex
WIP

@@
 \newcommand{\tikzoverview}{%
-    \resizebox{!}{0.8\textheight}{%
+    \resizebox{0.8\textwidth}{!}{%
         \begin{tikzpicture}[x=1cm,y=1cm]
@@
 \newcommand{\tikzhierarchical}{%
-    \resizebox{!}{0.8\textheight}{%
+    \resizebox{0.8\textwidth}{!}{%
         \begin{tikzpicture}[x=1cm,y=1cm]
@@
 \newcommand{\tikzmodelagnostic}{%
-    \resizebox{!}{0.8\textheight}{%
+    \resizebox{0.8\textwidth}{!}{%
         \begin{tikzpicture}[x=1cm,y=1cm]
@@
 \newcommand{\tikzbrief}{%
-    \resizebox{!}{0.8\textheight}{%
+    \resizebox{0.8\textwidth}{!}{%
         \begin{tikzpicture}[x=1cm,y=1cm]
@@
 \newcommand{\tikzbenchmarking}{%
-    \resizebox{!}{0.8\textheight}{%
+    \resizebox{0.8\textwidth}{!}{%
         \begin{tikzpicture}[x=1cm,y=1cm]
@@
 \newcommand{\tikzautomated}{%
-    \resizebox{!}{0.8\textheight}{%
+    \resizebox{0.8\textwidth}{!}{%
         \begin{tikzpicture}[x=1cm,y=1cm]
@@
 \newcommand{\tikzretrospective}{%
-    \resizebox{!}{0.8\textheight}{%
+    \resizebox{0.8\textwidth}{!}{%
         \begin{tikzpicture}[x=1cm,y=1cm]
@@
 
-                    \tikzmarkin<3>{modelagnostic-1}(0.1,-0.4)(-0.1,0.6)
+                    \tikzmarkin<2>{modelagnostic-1}(0.1,-0.4)(-0.1,0.6)
                     & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-1} \\
@@
 
-                    \tikzmarkin<4>{brief-1}(0.1,-0.4)(-0.1,0.6)
+                    \tikzmarkin<3>{brief-1}(0.1,-0.4)(-0.1,0.6)
                     & {\color<2->{black}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-1} \\
@@
 
-                    \tikzmarkin<5>{benchmarking-1}(0.1,-0.3)(-0.1,0.4)
+                    \tikzmarkin<4>{benchmarking-1}(0.1,-0.3)(-0.1,0.4)
                     & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-1} \\
@@
 
-                    \tikzmarkin<6>{automated-1}(0.1,-0.4)(-0.1,0.6)
+                    \tikzmarkin<5>{automated-1}(0.1,-0.4)(-0.1,0.6)
                     & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-1} \\
@@
 
-                    \tikzmarkin<7>{retrospective-1}(0.1,-0.4)(-0.1,0.6)
+                    \tikzmarkin<6>{retrospective-1}(0.1,-0.4)(-0.1,0.6)
                     & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-1} \\
@@
         \begin{column}{0.4\textwidth}
-            \tikzstyle{mybox} = [fill, rounded corners, minimum width=140, minimum height=20]
-            \tikzstyle{myrect} = [fill]
             \begin{figure}
@@
                     \onslide<2>\tikzhierarchical
-                    \onslide<3>\tikzmodelagnostic
-                    \onslide<4>\tikzbrief
-                    \onslide<5>\tikzbenchmarking
-                    \onslide<6>\tikzautomated
-                    \onslide<7>\tikzretrospective
+                    % \onslide<2>\tikzmodelagnostic
+                    \onslide<3>\tikzbrief
+                    \onslide<4>\tikzbenchmarking
+                    \onslide<5>\tikzautomated
+                    \onslide<6>\tikzretrospective
                 \end{overprint}
@@
     \end{columns}
+
+    \note[item]<1->{So how does the thesis tackle this problem of building such a system?}
+    \note[item]<2->{First we examine how a certain class of generative models can be used to detect data that is different from training data.}
+    \note[item]<2->{In our second paper we generalize this to be model-agnostic and phrased as a statistical test.}
+    \note[item]<3->{We then take a step back and provide an overview of modern methods for unsupervised speech representation learning.}
+    \note[item]<4->{We then benchmark some of these methods, generative latent variable models, on likelihood and automatic speech recognition.}
+    \note[item]<5->{We then move to a different domain, medical coding, and provide a critical review of recent works.}
+    \note[item]<6->{Finally, we provide a retrospective study on a machine learning-assisted stroke recognition system for medical helpline calls.}
     \end{frame}

### 2024-03-03 presentation/parts/presentation_hierarchical.tex
WIP

@@
             \vspace{3mm}
+
             Do this for any given single observation, i.e. answer the question:
+
             \vspace{3mm}
@@
     \end{columns}
+
+    \note[item]{Define OOD}
+    \note[item]{In 1D it is easy to see when this is possible and when it is not.}
+    \note[item]{In higher dimensions it is not as easy to see.}
 }
@@
     \end{figure}
+
+    \note[item]{Datasets can overlap quite a bit in their raw data space.}
 }
 
-\frame{
-    \frametitle{Problem and Contributions}
+\begin{frame}
+    \frametitle{Out-of-distribution detection with generative models}
+    \begin{columns}[t]
+        \begin{column}{0.4\textwidth}
             \begin{itemize}
-        \item Deep generative models often fail at OOD detection task when using their likelihood estimate as the score function \cite{nalisnick_deep_2019} by, perhaps surprisingly, assigning \highlight{higher likelihoods to OOD data}.
-        \item Contributions:
-        \begin{itemize}
-            \item We provide evidence that out-of-distribution detection fails due to learned low-level features that generalize across datasets.
-            \item We present a new score for OOD detection with hierarchical VAEs that alleviates this issue.
+                \item Generative models learn to approximate the \textbf{data distribution} $p(\xb)$.
+                \item The likelihood of the model given a sample $\xb$ is a measure of how well the model \textbf{explains the data}.
+                \item \textbf{Model likelihood} has long been thought of as useful for OOD detection \cite{bishop_novelty_1994}.
             \end{itemize}
-    \end{itemize}
-}
+        \end{column}
+        \begin{column}{0.6\textwidth}
+            \begin{overprint}
+                \onslide<2>
+                {\footnotesize HVAE trained on FashionMNIST evaluated on FashionMNIST and MNIST.}
+                \begin{figure}
+                    \centering
+                    \includegraphics[width=0.8\textwidth]{../graphics/paper_hierarchical/densities-FashionMNIST-test-MNIST-test-bpp-k-0_sohau.pdf}
+                \end{figure}
+            \end{overprint}
+        \end{column}
+    \end{columns}
+\end{frame}
+
+% \frame{
+%     \frametitle{Problem and Contributions}
+%     \begin{itemize}
+%         \item Deep generative models often fail at OOD detection task when using their likelihood estimate as the score function \cite{nalisnick_deep_2019} by, perhaps surprisingly, assigning \highlight{higher likelihoods to OOD data}.
+%         \item Contributions:
+%         \begin{itemize}
+%             \item We provide evidence that out-of-distribution detection fails due to learned low-level features that generalize across datasets.
+%             \item We present a new score for OOD detection with hierarchical VAEs that alleviates this issue.
+%         \end{itemize}  
+%     \end{itemize}
+% }
 
@@
 
-\begin{frame}
-    \frametitle{Out-of-distribution detection with hierarchical VAEs}
-    \begin{columns}[t]
-        \begin{column}{0.4\textwidth}
-            \begin{itemize}
-                \item Generative models learn to approximate the \textbf{data distribution} $p(\xb)$.
-                \item The likelihood of the model given a sample $\xb$ is a measure of how well the model \textbf{explains the data}.
-                \item \textbf{Model likelihood} has long been thought of as useful for OOD detection \cite{bishop_novelty_1994}.
-            \end{itemize}
-        \end{column}
-        \begin{column}{0.6\textwidth}
-            \begin{overprint}
-                \onslide<2>
-                \begin{figure}
-                    \centering
-                    \includegraphics[width=\textwidth]{../graphics/paper_hierarchical/densities-FashionMNIST-test-MNIST-test-bpp-k-0_sohau.pdf}
-                \end{figure}
-            \end{overprint}
-        \end{column}
-    \end{columns}
-\end{frame}
-
 % \section{Identifying the issue}
@@
         = \mathbb{E}_{q_\phi(\zb|\xb)} \left[ \log \frac{p_\theta(\xb,\zb)}{q_\phi(\zb|\xb)} \right] 
-        = \underbrace{\mathbb{E}_{q_\phi(\zb|\xb)}[\log p_\theta(\xb|\zb)]}_{\text{reconstruction likelihood}} - \underbrace{D_{\mathrm{KL}}( q_\phi(\zb|\xb) || p(\zb))}_{\text{regularization penalty}} \ .
+        = \underbrace{\mathbb{E}_{q_\phi(\zb|\xb)}[\log p_\theta(\xb|\zb)]}_{\text{reconstruction likelihood}} - \underbrace{D_{\mathrm{KL}}( q_\phi(\zb|\xb) \parallel p(\zb))}_{\text{regularization penalty}} \ .
     \end{equation}
@@
     \begin{equation}
-        D_{\mathrm{KL}}( q_\phi(\zb|\xb) || p(\zb)) = \mathbb{E}_{q_\phi(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\theta(\zb_i|\zb_{i+1})}{q_\phi(\zb_i|\zb_{i-1})} + \log \frac{p_\theta(\zb_L)}{q_\phi(\zb_L|\zb_{L-1})} \Big] \ .
+        D_{\mathrm{KL}}( q_\phi(\zb|\xb) \parallel p(\zb)) = \mathbb{E}_{q_\phi(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\theta(\zb_i|\zb_{i+1})}{q_\phi(\zb_i|\zb_{i-1})} + \log \frac{p_\theta(\zb_L)}{q_\phi(\zb_L|\zb_{L-1})} \Big] \ .
     \end{equation}
-    The absolute log-ratios grow with $\mathrm{dim}(\zb_i)$ since the log probability terms are computed by summing over the dimensionality of $\zb_i$.
+    Since the individual terms are computed by summing over the dimensionality of $\zb_i$, the absolute log-ratios grow with $\mathrm{dim}(\zb_i)$.
 
     \vspace{3mm}
+
+    \note[item]{There is a tendency for the lower-most latent variables to be weighted higher than the higher ones.}
 }
@@
     \end{figure}
+
+    \note[item]{Strong evidence that the lowest latent variables are generalizing across datasets.}
 }
 
-\section{The $\mathcal{L}^{>k}$ likelihood bound}
+% \section{The $\mathcal{L}^{>k}$ likelihood bound}
 
@@
     This bound uses the conditional prior for the lowest latent variables in the hierarchy.
+
+    \note[item]{So can we come up with a new bound that does not use the lowest latent variables in the same way?}
+    \note[item]{So we could use this for OOD detection (as done in BIVA).}
 }
 
-\section{Likelihood ratio}
+% \section{Likelihood ratio}
 
@@
     \begin{align}
-        \mathcal{L}      &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) || p_\theta(\zb|\xb)\right), \label{eq:likelihoods-as-exact} \\ 
-        \mathcal{L}^{>k} &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) || p_\theta(\zb|\xb)\right) \notag \ .
+        \mathcal{L}      &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) \parallel p_\theta(\zb|\xb)\right), \label{eq:likelihoods-as-exact} \\ 
+        \mathcal{L}^{>k} &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) \parallel p_\theta(\zb|\xb)\right) \notag \ .
     \end{align}
@@
     \begin{align}\label{eq:llr-as-kls}
-        LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) || p_\theta(\zb|\xb)\right) \\
-                     &\quad + D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) || p_\theta(\zb|\xb)\right) \ . \notag
+        LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) \parallel p_\theta(\zb|\xb)\right) \\
+                     &\quad + D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) \parallel p_\theta(\zb|\xb)\right) \ . \notag
     \end{align}
+
+    \note[item]{Write likelihood-ratio using the exact form of the bounds including intractable KL-divergence.}
 }
@@
 
-    Consequently, by importance sampling the ELBO, the associated KL-divergence vanishes and our likelihood ratio reduces to the KL-divergence of $\mathcal{L}^{>k}$.
+    Consequently, by importance sampling the ELBO, $D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) \parallel p_\theta(\zb|\xb)\right) \rightarrow 0$ and our likelihood ratio reduces to the KL-divergence of $\mathcal{L}^{>k}$.
 
     \begin{equation}\label{eq:llr-as-kls-iwae-reduced}
-        LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq }|\zb_{>k}) q(\zb_{>k}|\xb) || p(\zb|\xb)) \ .
+        LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq }|\zb_{>k}) q(\zb_{>k}|\xb) \parallel p(\zb|\xb)) \ .
     \end{equation}
@@
     $LLR^{>k}_{S}(\xb)$ performs KL-divergence-based OOD detection using top-most latent variables.
+
+    \note[item]{We can use the IWAE bound force the KL-divergence of the original ELBO towards zero.}
+    \note[item]{Then, the likelihood ratio reduces to the KL-divergence of $\mathcal{L}^{>k}$.}
+    \note[item]{We can hope that using the conditional prior for the lowest latent variables will make the KL-divergence more sensitive to OOD data in the top-most latent variables.}
 }
-
-\begin{frame}
-    \frametitle{Selecting the value of $k$}
-    \begin{itemize}
-        \item Use validation OOD dataset(s).
-        \item Compute $LLR^{>k}$ for different values of $k$ and select the one that maximizes the AUROC.
-        \item Compute feature correlations for different values of $k$ and select $k$ at the drop.
-    \end{itemize}
-    \begin{figure}
-        \centering
-        \includegraphics[width=0.9\textwidth]{../graphics/paper_hierarchical/feature-correlation-heatmap2.pdf}
-    \end{figure}
-\end{frame}
 
@@
 }
+
+\begin{frame}
+    \frametitle{Selecting the value of $k$}
+    \begin{itemize}
+        \item Use validation OOD dataset(s).
+        \item Compute $LLR^{>k}$ for different values of $k$ and select the one that maximizes the AUROC.
+        \item Compute feature correlations for different values of $k$ and select $k$ at the drop.
+    \end{itemize}
+    \begin{figure}
+        \centering
+        \includegraphics[width=0.9\textwidth]{../graphics/paper_hierarchical/feature-correlation-heatmap2.pdf}
+    \end{figure}
+\end{frame}
 
@@
 
-\frame{
-    \frametitle{Results on diverse datasets}
-    \begin{columns}
-        \begin{column}{0.45\textwidth}
-            \hfill
-            \begin{table}[t]
-                \centering
-                \resizebox{0.9\textwidth}{!}{%
-                \begin{tabular}{llrrr}
-                    \toprule
-                    OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
-                    \midrule
-                    \multicolumn{5}{c}{\textbf{Trained on CIFAR10}} \\
-                    \midrule
-                    SVHN          &  $LLR^{>2}$  &  0.811  &  0.837  &  0.394  \\
-                    CIFAR10       &  $LLR^{>1}$  &  0.469  &  0.479  &  0.835  \\
-                    \midrule
-                    \multicolumn{5}{c}{\textbf{Trained on SVHN}} \\
-                    \midrule
-                    CIFAR10            &  $LLR^{>1}$       &  $0.939$  &  $0.950$  &  $0.052$  \\
-                    SVHN               &  $LLR^{>1}$       &  $0.489$  &  $0.484$  &  $0.799$  \\
-                    \bottomrule
-                \end{tabular}
-                }
-            \end{table}
-            \hfill
-        \end{column}
-        \begin{column}{0.55\textwidth}
-            \hfill
-            \begin{table}[t]
-                \centering
-                \resizebox{0.9\textwidth}{!}{%
-                \begin{tabular}{llrrr}
-                    \toprule
-                    OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
-                    \midrule
-                    \multicolumn{5}{c}{\textbf{Trained on FashionMNIST}} \\
-                    \midrule
-                    MNIST                    & $LLR^{>1}$             &  0.986  &  0.987  &  0.011 \\
-                    notMNIST                 &  $LLR^{>1}$            &  0.998  &  0.998  &  0.000 \\
-                    KMNIST                   &  $LLR^{>1}$            &  0.974  &  0.977  &  0.017 \\
-                    Omniglot28x28            &  $LLR^{>2}$            &  1.000  &  1.000  &  0.000 \\
-                    Omniglot28x28Inverted    &  $LLR^{>1}$            &  0.954  &  0.954  &  0.050 \\
-                    SmallNORB28x28           &  $LLR^{>2}$            &  0.999  &  0.999  &  0.002 \\
-                    SmallNORB28x28Inverted   &  $LLR^{>2}$            &  0.941  &  0.946  &  0.069 \\
-   {\color{black!60}FashionMNIST}             &  $LLR^{>1}$            &  0.488  &  0.496  &  0.811 \\
-                    \midrule
-                    \multicolumn{5}{c}{\textbf{Trained on MNIST}} \\
-                    \midrule
-                    FashionMNIST                   &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
-                    notMNIST                       &  $LLR^{>1}$  &  $1.000$  &  $0.999$  &  $0.000$ \\
-                    KMNIST                         &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
-                    Omniglot28x28                  &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-                    Omniglot28x28Inverted          &  $LLR^{>1}$  &  $0.944$  &  $0.953$  &  $0.057$ \\
-                    SmallNORB28x28                 &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
-                    SmallNORB28x28Inverted         &  $LLR^{>1}$  &  $0.985$  &  $0.987$  &  $0.000$ \\
-   {\color{black!60}MNIST}                          &  $LLR^{>2}$  &  $0.515$  &  $0.507$  &  $0.792$ \\
-                    \bottomrule
-                \end{tabular}
-                }
-            \end{table}
-            \hfill
-        \end{column}
-    \end{columns}
-}
-
 \begin{frame}
@@
         \end{itemize}
-        \item[\highlight{$\rightarrow$}] Proposed a new score, $LLR^{>k}$, that uses the conditional prior for the top-most latent variables in the hierarchy.
+        \item[\highlight{$\rightarrow$}] Proposed a likelihood-ratio score, $LLR^{>k}$, that uses the conditional prior for the top-most latent variables in the hierarchy.
+        \item[\highlight{$\rightarrow$}] Showed that $LLR^{>k}$ outperforms the standard likelihood score across many different datasets and is competetive with state-of-the-art methods.
     \end{itemize}

### 2024-03-03 presentation/parts/presentation_introduction.tex
WIP

@@
             \item \textbf{Treatment} by suggesting medication.
-            \item Even \textbf{Cure} by helping to find new cures or treatments for diseases.
         \end{itemize}
@@
         \begin{column}{0.4\textwidth}
-            Every interaction within a healthcare system is a form of \highlight{medical dialogue}:
+            Central to an \highlight{interaction} within a healthcare system is the \highlight{medical dialogue}:
+            % Every interaction within a healthcare system is a form of \highlight{medical dialogue}:
             \begin{itemize}
@@
         \begin{itemize}
-            \item Reduce the impact of medical errors and adverse events.
-            \item Free up time spent on documentation for patient care.
+            \item Reduce the impact.
+            \item Free up time.
         \end{itemize}
@@
         \item <1-> \highlight{Data}: Quality, quantity, diversity, bias, privacy, ethics.
-        \item <2-> \highlight{Task}: Context, domain, language, culture, purpose.
+        \item <1-> \highlight{Task}: Context, domain, language, culture, purpose.
         \vspace{1em}
-        \item <3-> \highlight{Interpretability} of how a model works (transparency, accountability, regulation).
-        \item <4-> \highlight{Explainability} of model predictions (trust, understanding, feedback).
-        \item <5-> \highlight{Fairness} in treatment of different groups of people.
-        \item <6-> \highlight{Robustness} to noise, outliers, distribution shift, and adversarial attacks.
+        \item <2-> \highlight{Interpretability} of how a model works (transparency, accountability, regulation).
+        \item <2-> \highlight{Explainability} of model predictions (trust, understanding, feedback).
+        \item <2-> \highlight{Fairness} in treatment of different groups of people.
+        \item <2-> \highlight{Robustness} to noise, outliers, distribution shift, and adversarial attacks.
     \end{itemize}
+
+    \note[item]{So what are the challenges holding us back in implementing such systems?}
+    \note[item]{
+        \begin{itemize}
+            \item Strong requirements of data.
+            \item Tasks that span different contexts, domains, languages, and cultures.
+            \item Regulatory requirements for transparency and accountability.
+            \item Trust and understanding of predictions.
+        \end{itemize}
+    }
 \end{frame}
@@
         \begin{column}{0.6\textwidth}
-            Modular approach:
+            % Modular approach:
             \begin{itemize}
@@
     \end{columns}
+
+    \note[item]{We will take a modular approach to building a decision-support system.}
+    \note[item]{First we need source data}
+    \note[item]{Then we need to convert it into representations useful for downstream tasks.}
+    \note[item]{Then we can perform the downstream tasks.}
+    \note[item]{Finally, we need to estimate the reliability of our data, representations, and predictions.}
 \end{frame}

### 2024-03-03 presentation/parts/presentation_hierarchical.tex
WIP

@@
 
-\section{hierarchical vaes know what they don't know}
+\section[Hierarchical VAEs Know What They Don't Know]{hierarchical vaes know what they don't know}
 
@@
     \end{figure}
+
+    \note[item]{Datasets can overlap quite a bit in their raw data space.}
+    \note[item]{What we usually care about is a more semantic notion of similarity.}
 }
@@
     \note[item]{Datasets can overlap quite a bit in their raw data space.}
+    \note[item]{What we usually care about is a more semantic notion of similarity.}
 }
@@
     The first term is high if the data is well-explained by $\zb$.
-
-    \vspace{3mm}
     The second term we can rewrite as,
@@
     Since the individual terms are computed by summing over the dimensionality of $\zb_i$, the absolute log-ratios grow with $\mathrm{dim}(\zb_i)$.
+
+    Since the lower-most latent variables are usually higher dimensional than top ones, these are weighted higher in the ELBO.
 
@@
 \frame{
-    \frametitle{Results with $LLR^{>k}$}
+    \frametitle{ROC curves with $\mathcal{L}^{>k}$ and $LLR^{>k}$}
     \begin{figure}[\textwidth]
@@
 
-\begin{frame}
-    \frametitle{Selecting the value of $k$}
-    \begin{itemize}
-        \item Use validation OOD dataset(s).
-        \item Compute $LLR^{>k}$ for different values of $k$ and select the one that maximizes the AUROC.
-        \item Compute feature correlations for different values of $k$ and select $k$ at the drop.
-    \end{itemize}
-    \begin{figure}
-        \centering
-        \includegraphics[width=0.9\textwidth]{../graphics/paper_hierarchical/feature-correlation-heatmap2.pdf}
-    \end{figure}
-\end{frame}
+% \begin{frame}
+%     \frametitle{Selecting the value of $k$}
+%     \begin{itemize}
+%         \item Use validation OOD dataset(s).
+%         \item Compute $LLR^{>k}$ for different values of $k$ and select the one that maximizes the AUROC.
+%         \item Compute feature correlations for different values of $k$ and select $k$ at the drop.
+%     \end{itemize}
+%     \begin{figure}
+%         \centering
+%         \includegraphics[width=0.9\textwidth]{../graphics/paper_hierarchical/feature-correlation-heatmap2.pdf}
+%     \end{figure}
+% \end{frame}
 
-\begin{frame}
-    \frametitle{Results on FashionMNIST/MNIST}
-    \begin{table}
-        \centering
-        \resizebox*{!}{0.87\textheight}{%
-        \begin{tabular}{lrrr}
-            \toprule
-            Method & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
-            \midrule
-            \multicolumn{4}{c}{\textbf{FashionMNIST (in) / MNIST (out)}} \\
-            \midrule
-            \multicolumn{4}{l}{\textbf{Use prior knowledge of OOD}} \\
-            Backgr. contrast. LR (PixelCNN) {\parencite{ren_likelihood_2019}}               & $0.994$ & $0.993$ & $0.001$ \\
-            Backgr. contrast. LR (VAE) {\parencite{choi_waic_2019}}                    & $0.924$ & - & - \\
-            Binary classifier {\parencite{ren_likelihood_2019}}                              & $0.455$ & $0.505$ & $0.886$ \\ % 6
-            $p(\hat{y} | \xb)$ with OOD as noise class {\parencite{ren_likelihood_2019}}     & $0.877$ & $0.871$ & $0.195$ \\ % 7
-            $p(\hat{y} | \xb)$ with calibration on OOD {\parencite{ren_likelihood_2019}}     & $0.904$ & $0.895$ & $0.139$ \\ % 8
-            Input complexity ($S$, Glow) \parencite{hendrycks_deep_2019}                    & $0.998$ & - & - \\
-            Input complexity ($S$, PixelCNN++) \parencite{hendrycks_deep_2019}              & $0.967$ & - & - \\
-            \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
-            $p(\hat{y} | \xb)$ {\parencite{ren_likelihood_2019, hendrycks_baseline_2017}}                        & $0.734$ & $0.702$ & $0.506$ \\
-            Entropy of $p(y | \xb)$ {\parencite{ren_likelihood_2019}}                        & $0.746$ & $0.726$ & $0.448$ \\
-            ODIN {\parencite{ren_likelihood_2019, liang_enhancing_2018}}                                       & $0.752$ & $0.763$ & $0.432$ \\
-            VIB \parencite{alemi_uncertainty_2018, choi_waic_2019}                                          & $0.941$ & - & - \\
-            Mahalanobis distance, CNN {\parencite{ren_likelihood_2019}}                     & $0.942$ & $0.928$ & $0.088$ \\
-            Mahalanobis distance, DenseNet {\parencite{lee_simple_2018}}                & $0.986$ & - & - \\
-            Ensemble, 20 classifiers {\parencite{ren_likelihood_2019, lakshminarayanan_simple_2017}}                  & $0.857$ & $0.849$ & $0.240$ \\
-            \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
-            \multicolumn{4}{l}{\textit{- Ensembles}} \\
-            WAIC, 5 models, VAE {\parencite{choi_waic_2019}}                          & $0.766$ & - & - \\
-            WAIC, 5 models, PixelCNN {\parencite{ren_likelihood_2019}}                      & $0.221$ & $0.401$ & $0.911$ \\
-            \multicolumn{4}{l}{\textit{- Not ensembles}} \\
-            Likelihood regret \parencite{xiao_likelihood_2020}                               & $\mathbf{0.988}$ & - & - \\
-            $\mathcal{L}^{>0}$ + HVAE (ours)                    & $0.268$ & $0.363$ & $0.882$ \\
-            $\mathcal{L}^{>1}$ + HVAE (ours)                    & $0.593$ & $0.591$ & $0.658$ \\
-            $\mathcal{L}^{>2}$ + HVAE (ours)                    & $0.712$ & $0.750$ & $0.548$ \\
-            $LLR^{>1}$ + HVAE (ours)                            & $0.964$ & $0.961$ & $0.036$ \\
-            $LLR^{>1}_{250}$ + HVAE (ours)                      & $0.984$ & $\mathbf{0.984}$ & $\mathbf{0.013}$ \\
-             \bottomrule
-        \end{tabular}%
-        }
-    \end{table}
-\end{frame}
+% \begin{frame}
+%     \frametitle{Results on FashionMNIST/MNIST}
+%     \begin{table}
+%         \centering
+%         \resizebox*{!}{0.87\textheight}{%
+%         \begin{tabular}{lrrr}
+%             \toprule
+%             Method & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
+%             \midrule
+%             \multicolumn{4}{c}{\textbf{FashionMNIST (in) / MNIST (out)}} \\
+%             \midrule
+%             \multicolumn{4}{l}{\textbf{Use prior knowledge of OOD}} \\
+%             Backgr. contrast. LR (PixelCNN) {\parencite{ren_likelihood_2019}}               & $0.994$ & $0.993$ & $0.001$ \\
+%             Backgr. contrast. LR (VAE) {\parencite{choi_waic_2019}}                    & $0.924$ & - & - \\
+%             Binary classifier {\parencite{ren_likelihood_2019}}                              & $0.455$ & $0.505$ & $0.886$ \\ % 6
+%             $p(\hat{y} | \xb)$ with OOD as noise class {\parencite{ren_likelihood_2019}}     & $0.877$ & $0.871$ & $0.195$ \\ % 7
+%             $p(\hat{y} | \xb)$ with calibration on OOD {\parencite{ren_likelihood_2019}}     & $0.904$ & $0.895$ & $0.139$ \\ % 8
+%             Input complexity ($S$, Glow) \parencite{hendrycks_deep_2019}                    & $0.998$ & - & - \\
+%             Input complexity ($S$, PixelCNN++) \parencite{hendrycks_deep_2019}              & $0.967$ & - & - \\
+%             \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
+%             $p(\hat{y} | \xb)$ {\parencite{ren_likelihood_2019, hendrycks_baseline_2017}}                        & $0.734$ & $0.702$ & $0.506$ \\
+%             Entropy of $p(y | \xb)$ {\parencite{ren_likelihood_2019}}                        & $0.746$ & $0.726$ & $0.448$ \\
+%             ODIN {\parencite{ren_likelihood_2019, liang_enhancing_2018}}                                       & $0.752$ & $0.763$ & $0.432$ \\
+%             VIB \parencite{alemi_uncertainty_2018, choi_waic_2019}                                          & $0.941$ & - & - \\
+%             Mahalanobis distance, CNN {\parencite{ren_likelihood_2019}}                     & $0.942$ & $0.928$ & $0.088$ \\
+%             Mahalanobis distance, DenseNet {\parencite{lee_simple_2018}}                & $0.986$ & - & - \\
+%             Ensemble, 20 classifiers {\parencite{ren_likelihood_2019, lakshminarayanan_simple_2017}}                  & $0.857$ & $0.849$ & $0.240$ \\
+%             \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
+%             \multicolumn{4}{l}{\textit{- Ensembles}} \\
+%             WAIC, 5 models, VAE {\parencite{choi_waic_2019}}                          & $0.766$ & - & - \\
+%             WAIC, 5 models, PixelCNN {\parencite{ren_likelihood_2019}}                      & $0.221$ & $0.401$ & $0.911$ \\
+%             \multicolumn{4}{l}{\textit{- Not ensembles}} \\
+%             Likelihood regret \parencite{xiao_likelihood_2020}                               & $\mathbf{0.988}$ & - & - \\
+%             $\mathcal{L}^{>0}$ + HVAE (ours)                    & $0.268$ & $0.363$ & $0.882$ \\
+%             $\mathcal{L}^{>1}$ + HVAE (ours)                    & $0.593$ & $0.591$ & $0.658$ \\
+%             $\mathcal{L}^{>2}$ + HVAE (ours)                    & $0.712$ & $0.750$ & $0.548$ \\
+%             $LLR^{>1}$ + HVAE (ours)                            & $0.964$ & $0.961$ & $0.036$ \\
+%             $LLR^{>1}_{250}$ + HVAE (ours)                      & $0.984$ & $\mathbf{0.984}$ & $\mathbf{0.013}$ \\
+%              \bottomrule
+%         \end{tabular}%
+%         }
+%     \end{table}
+% \end{frame}
 
@@
             \item Strong correlations between some latent variables for different datasets.
-            \item Reconstructions of OOD data are good when using full approximate posterior.
+            % \item Reconstructions of OOD data are good when using full approximate posterior.
         \end{itemize}
-        \item[\highlight{$\rightarrow$}] Proposed a likelihood-ratio score, $LLR^{>k}$, that uses the conditional prior for the top-most latent variables in the hierarchy.
-        \item[\highlight{$\rightarrow$}] Showed that $LLR^{>k}$ outperforms the standard likelihood score across many different datasets and is competetive with state-of-the-art methods.
+        \item[\highlight{$\rightarrow$}] Proposed a likelihood-ratio score, $LLR^{>k}$, that uses the conditional prior for the top-most latent variables in the hierarchy and showed its effectiveness.
+        % \item[\highlight{$\rightarrow$}] Showed that $LLR^{>k}$ outperforms the standard likelihood score across many different datasets and is competetive with state-of-the-art methods.
     \end{itemize}

### 2024-03-03 presentation/parts/presentation_introduction.tex
WIP

@@
 
-\section{Introduction}
+\section{introduction}
 
@@
 
-\begin{frame}
-    \frametitle{Reliability of machine learning systems}
-    \begin{itemize}
-        \item <1-> \highlight{Data}: Quality, quantity, diversity, bias, privacy, ethics.
-        \item <1-> \highlight{Task}: Context, domain, language, culture, purpose.
-        \vspace{1em}
-        \item <2-> \highlight{Interpretability} of how a model works (transparency, accountability, regulation).
-        \item <2-> \highlight{Explainability} of model predictions (trust, understanding, feedback).
-        \item <2-> \highlight{Fairness} in treatment of different groups of people.
-        \item <2-> \highlight{Robustness} to noise, outliers, distribution shift, and adversarial attacks.
-    \end{itemize}
+% \begin{frame}
+%     \frametitle{Reliability of machine learning systems}
+%     \begin{itemize}
+%         \item <1-> \highlight{Data}: Quality, quantity, diversity, bias, privacy, ethics.
+%         \item <1-> \highlight{Task}: Context, domain, language, culture, purpose.
+%         \vspace{1em}
+%         \item <2-> \highlight{Interpretability} of how a model works (transparency, accountability, regulation).
+%         \item <2-> \highlight{Explainability} of model predictions (trust, understanding, feedback).
+%         \item <2-> \highlight{Fairness} in treatment of different groups of people.
+%         \item <2-> \highlight{Robustness} to noise, outliers, distribution shift, and adversarial attacks.
+%     \end{itemize}
 
-    \note[item]{So what are the challenges holding us back in implementing such systems?}
-    \note[item]{
-        \begin{itemize}
-            \item Strong requirements of data.
-            \item Tasks that span different contexts, domains, languages, and cultures.
-            \item Regulatory requirements for transparency and accountability.
-            \item Trust and understanding of predictions.
-        \end{itemize}
-    }
-\end{frame}
+%     \note[item]{So what are the challenges holding us back in implementing such systems?}
+%     \note[item]{
+%         \begin{itemize}
+%             \item Strong requirements of data.
+%             \item Tasks that span different contexts, domains, languages, and cultures.
+%             \item Regulatory requirements for transparency and accountability.
+%             \item Trust and understanding of predictions.
+%         \end{itemize}
+%     }
+% \end{frame}
 

### 2024-03-04 presentation/parts/overview_presentation.tex
WIP

@@
 
-\section{Overview}
+\section*{overview}
 
@@
 
-    \note[item]<1->{So how does the thesis tackle this problem of building such a system?}
-    \note[item]<2->{First we examine how a certain class of generative models can be used to detect data that is different from training data.}
-    \note[item]<2->{In our second paper we generalize this to be model-agnostic and phrased as a statistical test.}
-    \note[item]<3->{We then take a step back and provide an overview of modern methods for unsupervised speech representation learning.}
-    \note[item]<4->{We then benchmark some of these methods, generative latent variable models, on likelihood and automatic speech recognition.}
-    \note[item]<5->{We then move to a different domain, medical coding, and provide a critical review of recent works.}
-    \note[item]<6->{Finally, we provide a retrospective study on a machine learning-assisted stroke recognition system for medical helpline calls.}
+    \note[item]{So how does the thesis tackle this problem of building such a system?}
+    \note{
+        \begin{itemize}
+            \item [Chp. 4] Examine how a certain class of generative models can be used to detect data that is different from training data. 
+            \item [Chp. 5] Generalize this to be model-agnostic and phrased as a statistical test.
+            \item [Chp. 6] Step back and provide an overview of modern methods for unsupervised speech representation learning.
+            \item [Chp. 7] Benchmark some of these methods, generative latent variable models, on likelihood and automatic speech recognition.
+            \item [Chp. 8] Move to a different domain, medical coding, and provide a critical review of recent works.
+            \item [Chp. 9] Provide a retrospective study on a machine learning-assisted stroke recognition system for medical helpline calls.
+        \end{itemize}
+    }
     \end{frame}

### 2024-03-04 presentation/parts/presentation_hierarchical.tex
WIP

@@
   \end{columns}
+
+  \note[item]{We want to explain why HVAEs fail at OOD detection when using their likelihood.}
+    \note[item]{And how to fix them.}
 \end{frame}
@@
 \frame{
-    \frametitle{What do the lowest latent variables code for?}
+    \frametitle{What do the lowest latent variables represent?}
 
@@
 
-% \section{The $\mathcal{L}^{>k}$ likelihood bound}
-
-\frame{
-    \frametitle{An alternative likelihood bound, $\mathcal{L}^{>k}$}
-    An alternative version of the ELBO that only partially uses the approximate posterior can be written as \cite{maaloe_biva_2019}
-    \begin{equation}
-        \mathcal{L}^{>k}(\xb; \theta, \phi) = \mathbb{E}_{p_\theta(\zb_{\leq k}|\zb{>k})q_\phi(\zb_{>k}|\xb)} \left[ \log \frac{p_\theta(\xb|\zb)p_\theta(\zb_{>k})}{q_\phi(\zb_{>k}|\xb)} \right]
-    \end{equation}
-
-    Here, we have replaced the approximate posterior $q_\phi(\zb|\xb)$ with a different proposal distribution that combines part of the approximate posterior with the conditional prior, namely
-
-    $$p_\theta(\zb_{\leq k}|\zb_{>k})q_\phi(\zb_{>k}|\xb)$$
-
-    This bound uses the conditional prior for the lowest latent variables in the hierarchy.
-
-    \note[item]{So can we come up with a new bound that does not use the lowest latent variables in the same way?}
-    \note[item]{So we could use this for OOD detection (as done in BIVA).}
-}
-
-% \section{Likelihood ratio}
-
 \frame{
     \frametitle{Likelihood ratios}
-    We can use our new bound to compute the score used in a standard likelihood ratio test \cite{buse_likelihood_1982}.
-    \begin{equation}\label{eq:llr-as-difference-in-likelihoods}
-        LLR^{>k}(\xb) \equiv \mathcal{L}(\xb) - \mathcal{L}^{>k}(\xb) \ .
+    We suggest to define a likelihood ratio score \cite{buse_likelihood_1982} using the ELBO $\mathcal{L}(\xb)$ and a relaxed bound $\mathcal{L}^{>k}(\xb)$.
+    %
+    \begin{equation}
+        LLR^{>k}(\xb) \equiv \mathcal{L}(\xb) - \mathcal{L}^{>k}(\xb) \ ,
     \end{equation}
-    We can inspect what this likelihood-ratio measures by considering the exact form of our bounds.
+    %
+    where the exact form of the bounds is,
+    %
     \begin{align}
@@
     \end{align}
-    In the likelihood ratio the reconstruction terms cancel out and only the KL-divergences from the approximate to the true posterior remain.
-    \begin{align}\label{eq:llr-as-kls}
+    %
+    In the likelihood ratio $log p_\thetab(\xb)$ cancels out and only the KL-divergences from the approximate to the true posterior remain.
+    %
+    \begin{align}
         LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) \parallel p_\theta(\zb|\xb)\right) \\
@@
         \end{itemize}
-        \item[\highlight{$\rightarrow$}] Proposed a likelihood-ratio score, $LLR^{>k}$, that uses the conditional prior for the top-most latent variables in the hierarchy and showed its effectiveness.
+        \item[\highlight{$\rightarrow$}] Provide explanation for why the likelihood fails fro OODD for HVAEs.
+        \item[\highlight{$\rightarrow$}] Proposed a likelihood-ratio score, $LLR^{>k}$, that uses the conditional prior for the bottom-most latent variables in the hierarchy and showed its effectiveness.
         % \item[\highlight{$\rightarrow$}] Showed that $LLR^{>k}$ outperforms the standard likelihood score across many different datasets and is competetive with state-of-the-art methods.

### 2024-03-04 presentation/parts/presentation_introduction.tex
WIP

@@
 
-\begin{frame}
-    \frametitle{What is healthcare?}
-    \begin{quotation}
-        \noindent\centering
-        \textbf{Healthcare} is the improvement of health via the \highlight{prevention}, \highlight{diagnosis}, \highlight{treatment}, \highlight{amelioration} or \highlight{cure} of {\color{theme-green}disease}, {\color{theme-green}illness}, {\color{theme-green}injury}, and {\color{theme-green}other physical and mental impairments} in people.
-    \end{quotation}
-    \begin{flushright}
-        \footnotesize
-        --- Wikipedia
-    \end{flushright}
+% \begin{frame}
+%     \frametitle{What is healthcare?}
+%     \begin{quotation}
+%         \noindent\centering
+%         \textbf{Healthcare} is the improvement of health via the \highlight{prevention}, \highlight{diagnosis}, \highlight{treatment}, \highlight{amelioration} or \highlight{cure} of {\color{theme-green}disease}, {\color{theme-green}illness}, {\color{theme-green}injury}, and {\color{theme-green}other physical and mental impairments} in people.
+%     \end{quotation}
+%     \begin{flushright}
+%         \footnotesize
+%         --- Wikipedia
+%     \end{flushright}
 
-    \note[item]{Machine learning for healthcare touches on most of these aspects. Fx:}
-    \note[item]{
-        \begin{itemize}
-            \item \textbf{Prevention} by predicting risk of disease allowing for early intervention.
-            \item \textbf{Diagnosis} by detecting stroke in emergency calls.
-            \item \textbf{Treatment} by suggesting medication.
-        \end{itemize}
-    }
-\end{frame}
+%     \note[item]{Machine learning for healthcare touches on most of these aspects. Fx:}
+%     \note[item]{
+%         \begin{itemize}
+%             \item \textbf{Prevention} by predicting risk of disease allowing for early intervention.
+%             \item \textbf{Diagnosis} by detecting stroke in emergency calls.
+%             \item \textbf{Treatment} by suggesting medication.
+%         \end{itemize}
+%     }
+% \end{frame}
 

### 2024-03-04 presentation/parts/presentation_introduction.tex
WIP

@@
     \begin{columns}[t]
-        \begin{column}{0.4\textwidth}
+        \begin{column}{0.43\textwidth}
             Central to an \highlight{interaction} within a healthcare system is the \highlight{medical dialogue}:
@@
         \end{column}
-        \begin{column}{0.6\textwidth}
+        \begin{column}{0.57\textwidth}
             \vspace{-2em}
@@
             \begin{itemize}
-                \item <1-> Documentation is a central part of healthcare.
-                \item <1-> E.g. patient records, insurance claims, billing, research, training, legal purposes.
+                \item <1-> Almost every patient interaction has to be \highlight{documented}.
+                \item <1-> Patient records, insurance claims, billing, research, training, legal purposes.
                 \vspace{1em}
-                \item <2-> \highlight{Time-consuming}: Physicians spend 34-37\% of their time on documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote<2->{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
+                \item <2-> \highlight{Time-consuming}: Physicians spend 34-37\% of their time on writing documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote<2->{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
                 \item <2-> \highlight{Varying quality}: Discharge summaries almost never meet \emph{all} timeline, transmission, and content criteria.
@@
                 \item <3-> \highlight{Foundation modelling}: Converting the input into representations useful for downstream tasks.
-                \item <4-> \highlight{Downstream tasks}: Suggesting, summarizing, classifying, transcribing, translating, etc.
+                \item <4-> \highlight{Downstream tasks}: Suggesting questions, summarizing conversations, classifying illnesses, translating, etc.
                 \item <5-> \highlight{Uncertainty}: Estimating the reliability of data, representations, predictions.

### 2024-03-04 presentation/parts/presentation_hierarchical.tex
WIP

@@
                 \onslide<2>
-                {\footnotesize HVAE trained on FashionMNIST evaluated on FashionMNIST and MNIST.}
                 \begin{figure}
@@
                 \end{figure}
+                \centering\footnotesize Hierarchical VAE trained on FashionMNIST.
             \end{overprint}
@@
     Absolute Pearson correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST.
-    \vspace{0.3cm}
 
@@
     %
-    In the likelihood ratio $log p_\thetab(\xb)$ cancels out and only the KL-divergences from the approximate to the true posterior remain.
+    In the likelihood ratio $\log p_\thetab(\xb)$ cancels out and only the KL-divergences from the approximate to the true posterior remain.
     %
@@
         \end{itemize}
-        \item[\highlight{$\rightarrow$}] Provide explanation for why the likelihood fails fro OODD for HVAEs.
+        \item[\highlight{$\rightarrow$}] Provide explanation for why the likelihood fails for OODD for HVAEs.
         \item[\highlight{$\rightarrow$}] Proposed a likelihood-ratio score, $LLR^{>k}$, that uses the conditional prior for the bottom-most latent variables in the hierarchy and showed its effectiveness.

### 2024-03-05 presentation/parts/overview_presentation.tex
WIP
Done

@@
     \note{
-        \begin{itemize}
-            \item [Chp. 4] Examine how a certain class of generative models can be used to detect data that is different from training data. 
-            \item [Chp. 5] Generalize this to be model-agnostic and phrased as a statistical test.
-            \item [Chp. 6] Step back and provide an overview of modern methods for unsupervised speech representation learning.
-            \item [Chp. 7] Benchmark some of these methods, generative latent variable models, on likelihood and automatic speech recognition.
-            \item [Chp. 8] Move to a different domain, medical coding, and provide a critical review of recent works.
-            \item [Chp. 9] Provide a retrospective study on a machine learning-assisted stroke recognition system for medical helpline calls.
-        \end{itemize}
+        \begin{enumerate}
+            \item Examine how a certain class of generative models can be used to detect data that is different from training data. 
+            \item Generalize this to be model-agnostic and phrased as a statistical test.
+            \item Step back and provide an overview of modern methods for unsupervised speech representation learning.
+            \item Benchmark some of these methods, generative latent variable models, on likelihood and automatic speech recognition.
+            \item Move to a different domain, medical coding, and provide a critical review of recent works.
+            \item Provide a retrospective study on a machine learning-assisted stroke recognition system for medical helpline calls.
+        \end{enumerate}
     }
@@
     \end{columns}
+
+    \note[item]{We want to keep papers that represent each of the three main themes of the thesis: (Uncertainty estimation, representation learning, and clinical applications)}
 \end{frame}
